### 2008-02-04 htdocs/sections.tex
Documentation sources
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@166 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.1 2008-02-04 12:46:14 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{Note to Readers}

This document is intended as a starting point for a full-blown
specification of the next version of the Ludwig code, and everything
to do with it. It is (manifestly) not complete.

\subsection{Some History}

Early Lattice Gas and Lattice Boltzmann. Early Lattice Boltzmann at Edinburgh.
Ludwig.

\subsection{Aims}

We aim to provide a robust and portable code, written in C, which
can be used to perform serial and scalable parallel simulations of
complex fluid systems based around the Lattice Boltzmann method.

\subsection{Availability}

We now have a NeSCForge project account available at

\texttt{http://forge.nesc.ac.uk/projects/ludwig/}

This provides controlled access to known
persons and provide CVS, bug tracking, and so on. It is
\textbf{essential} that people sign up to this, as we must
move to a single supported version.

The NeSCForge CVS is not available to the public.

\subsection{Ongoing Development}

\subsubsection{Some notes}

\begin{enumerate}

\item
Serial timer is exhausted at about 35 minutes.

\item
Size of the colloid message buffers needs to be computed robustly
at run time, or, allocated on a just-in-time basis with knowledge
of exact size required. The former would mean knowing the number
of particles before the call to CCOM\_init\_halo.

\end{enumerate}

\subsubsection{Other Agenda Items}
\begin{enumerate}

\item
The particle code needs a critic from someone other than myself.
Again we need to abstract it away a bit more.

\item
Two-dimensional model (D2Q9).

\item
Finite Difference for order parameter.

\item
I/O via XML/HDF5.

\end{enumerate}

\section{User Input}

By default the executable will look for an input file at run time.
By default, this file is named \texttt{input} and should be in the
current working directory. The alternative is to specify the name
of the file explicitly on the command line:
\begin{verbatim}
$ ./Ludwig.exe input_file
\end{verbatim}

The contents of the input file are made up of \textit{key value}
pairs with control the run-time behaviour. Blank lines and lines
begining with \# are ignored as comments. The following describes
the effect of various keys. Most keys have a default value which
will be used by the code if the corresponding key is not present
(or commented out) in the input file.

\subsection{Basic run parameters}

\inputkey{N\_cycles}

The number of lattice Boltzmann time steps to execute. Default
value:

\texttt{N\_cycles 0}

\inputkey{t\_start}

By default the code will start from time step $t = 0$. If you 
wish to restart from a previously saved configuration, set
the appropriate value for \texttt{t\_start}. The code will
execute \texttt{N\_cycles} steps starting from this point.

\inputkey{input\_config}

The file name stub for input configuration required for restarts.
For example, to restart from the file config.out100 previously
produced at the end of a 100 step run use
\begin{verbatim}
t_start 100
input_config config.out100
\end{verbatim}

If you are using parallel I/O, the filename stub is required; the
file extensions are computed automatically.

\inputkey{size}

Controls the number of lattice points in the $(x, y, z)$ directions,
respectively. Default value:

\texttt{size 64\_64\_64}

\inputkey{grid}

In parallel, this sets the processor Cartesian decomposition, i.e.,
the number of processors in each dimension of the Cartesian communicator.
The default value may be implementation dependent as it is that returned
by \texttt{MPI\_Dims\_create()}. For
example:
\begin{verbatim}
size 64_64_64
grid 4_2_1
\end{verbatim}
gives a total (physical) system size of 64 lattice sites in each
direction, decomposed across 4 processors in the x-direction, 2 in
the y-direction and 1 (no decomposition) in the z-direction. The
total number of processors must therefore be 8. The local domain
size per processor is then 16x32x64.

\subsection{Fluid properties}

\inputkey{viscosity}

Sets the LB fluid shear viscosity $\eta$ (and related relaxation times).
Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
(i.e., realaxation time equal to unity).

\inputkey{viscosity\_bulk}

Sets the LB bulk viscosity $\zeta$ (and related relaxation times).
This can be use if you are concerned with incompressiblity
violations, in which case $\zeta >> \eta$ can be useful. The
default value is $\zeta = \eta$.

\inputkey{temperature}

The LB 'temperature' using fluctuating hydrodynamics. Safe values
are $0.0001 > kT > 0$.

\inputkey{rho0}

Sets the mean density of the fluid. The default value is unity, and
it is recommended that the value should not be changed.

\subsection{Binary fluid parameters}

The following apply for the symmetric binary fluid problem with
order parameter $\phi$ and free energy (see section X) \cite{viv}
\begin{equation}
 F[\phi] = 
\int dr \left(
{\scriptstyle \frac{1}{2}}A\phi^2
+ {\scriptstyle \frac{1}{4}}B\phi^4
+ {\scriptstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right)
\end{equation}

\inputkey{A}

The parameter $A$. Note $A < 0$.

\inputkey{B}

The parameter $B$. Note $B = -A$ for common usage, although this is
not enforced.

\inputkey{K}

The parameter $\kappa$

\inputkey{C}

The surface free energy parameter {C}. Currently unused.

\inputkey{H}

The surface free energy parameter {H}. Currently unused.

\inputkey{mobility}

Sets the order parameter mobility in the LB framework.

\inputkey{phi0}

The mean compositional order parameter roughly $-0.5 < \phi < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

\inputkey{noise}

The magnitude of the initial fluctuations in $\phi$ used to
initiate spinodal decompsotion.

\subsection{Colloidal particles}

There are quite a few options here.

\inputkey{colloid\_init}

This determines how the code initialises particles (or not). The
default is to have no particles (value \texttt{no\_colloids}). There
are two other options at the moment.

\texttt{colloid\_init    fixed\_number\_monodisperse}

initialises a fixed number of particles with randomly selected
positions. Note that there is no gaurentee that the particles
do not overlap. The number of particles is determined by
the value of the key \texttt{colloid\_no}.

\texttt{colloid\_init   fixed\_volume\_fraction\_monodisperse}

initialises a given volume fraction of particles on a (roughly)
bcc lattice. The particles will not overlap. The volume fraction
is set using the key \texttt{colloid\_vf}.

\inputkey{colloid\_a0}

Sets the nominal radius of the particle on the lattice $a_0$. This is
used to construct the links for BBL. The smallest acceptable LB
particle is $a_0 = 1.25$ \cite{nl02}. There are 'magic' values
including 1.25, 2.3, 3.7, 4.7 which minimise discretisation effects.

\inputkey{colloid\_ah}

Sets the hydrodynamic radius of the particles $a_h$. For all
reasonable applications choose $a_h = a_0$.

\inputkey{colloid\_mc\_steps}

It is possible to have the code follow the initialisation of
particle positions by a number of Monte Carlo steps. This can
help to remove overlaps (fixed\_number\_monodisperse) or
melt the lattice (fixed\_volume\_fraction\_monodisperse).

\inputkey{colloid\_gravity}

Sets a body force on each paarticle. Useful, for example, for
sedimentation. For example,

\inputkey{cell\_list\_min}

This sets the minimum allowable width of the cell list used to
store the particles. This should be as small as possible for
efficiency purposes but must be large enough to capture all
possible interactions.

\texttt{colloid\_gravity  0.0\_0.0\_-0.0001}

gives a constant force in the negative $z$-direction on each particle.
The code automatically computes the compensating body force on fluid nodes
required to give no net change in the total momentum of the system at
each time step.

\subsection{Active particles}

Not in general use yet.

\subsection{Particle-particle interactions}

\subsection{Boundary conditions}

\subsection{Input/Output}

\inputkey{freq\_statistics}

sets the frequency for reporting general information from
the code. (In general, the function
 \texttt{is\_statistics\_step()} returns true at this frequency.

\inputkey{freq\_measure}

sets the frequency of recording of colloid/order parameter output.
(In general, the function \texttt{is\_measurement\_step()} returns
true at this frequency.)

\inputkey{freq\_config}

set frequency of full configuration (checkpoint) dumps suitable
for restart.

\inputkey{config\_at\_end}

enforces full configuration dump at the end of execution. The
default is 'yes'.

\subsection{Lees-Edwards planes}

\inputkey{N\_LE\_plane}

sets the total number of Lees-Edwards planes. Default is zero.

\inputkey{LE\_plane\_vel}

sets the velocity of each plane relative to the lattice.

\inputkey{LE\_init\_profile}

if set to 1, the fluid velocity is initialise to reflect a steady
state shear flow appropriate for the number of planes at the
given velocity. If set to zero, the fluid is initialised with
zero velocity.

\subsection{Miscellaneous}

\inputkey{random\_seed}

sets the random number generator seed.

\section{The Lattice Boltzmann Method}

We review here the lattice Boltzmann method applied to single
phase problems.

\subsection{The Navier Stokes Equation}

We seek to solve the isothermal Navier-Stokes equations which, often
written in vector form, express mass conservation
\begin{equation}
\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
\label{eq_mass1}
\end{equation}
and the conservation of momentum
\begin{equation}
\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
-\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
\label{eq_momentum1}
\end{equation}
Equation~\ref{eq_mass1} expresses the local rate of change of the
density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
Equation~\ref{eq_momentum1} expresses Newton's second law for
momentum, where the terms on the right hand side represent the
force on the fluid.

For this work, it is more convenient to rewrite these equations
in tensor notation, where Cartesian coordinates ${x,y,z}$ are
represented by indices $\alpha$ and $\beta$, viz
\begin{equation}
\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
\end{equation}
and
\begin{equation}
\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
= -\nabla_\alpha p
+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
\end{equation}
Here, repeated indices are understood to be summed over.
The conservation law is seem better if the forcing terms of the
right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
so that
\begin{equation}
\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
\end{equation}
In this case the expanded expression for the stress tensor is
\begin{equation}
\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
\end{equation}
where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
equations have 10 degrees of freedom (or hydrodynamic modes) being
$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.

\subsection{The Distribution Function and its Moments}

In lattice Boltzmann, the density and velocity of the continuum fluid
are replaced by their discrete counterparrts, the distribution function
$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
It is possible to relate the hydrodynamic modes to distribution
function via its moments, so
\begin{equation}
\rho = \sum_i f_i,  \,
\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
\end{equation}
here, the index of the summation is over the number of discrete
velocities used $n$, which in three dimensions is generally
denoted by D3Q$n$. We restrict our discussion to D3Q15 and
D3Q19 (see below).

In general, the number of modes supported in $n$ and these can
be written as
\begin{equation}
M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
\end{equation}
where $m_i^a$ is the corresponding eigenvector of the collision
maxtrix (see section blah). For example, if $m_i^a = 1$ then the
mode $M^a$ is the density. Correspondingly, the distribution
function can be related to the modes $M^a$ via
\begin{equation}
f_i() = \sum_a w_i m_i^a M^a() N^a
\end{equation}

\subsubsection{Hydrodynamic modes}

\subsubsection{D3Q15 ghost modes}

\subsubsection{D3Q19 ghost modes}

\subsection{The Lattice Boltzmann Equation}

\subsection{The Collision}

\subsection{External Body Forces}

\subsection{Fluctuating LBE}

It is possible \cite{adhikari2005} to simulate fluctuating
hydrodynamics for an isothermal fluid via the inclusion of
a fluctuating stress $\sigma_{\alpha\beta}$:
\begin{equation}
\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
\end{equation}
The fluctuation-dissipation theorem relates the magnitude of this
random stress to the isothermal temperature and the viscosity.

In the LBE, this translates to the addition of a random contribution
$\xi_i$ to the distribution at the collision stage, so that
\begin{equation}
\ldots + \xi_i.
\end{equation}

For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
i.e., those with dissipation, the fluctuating part may be written
\begin{equation}
\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
\end{equation}
where $\hat{\xi}^a$ is a noise termwhich has a variance determined
by the relaxation time for given mode
\begin{equation}
\left< \hat{\xi}^a \hat{\xi}^b \right> =
\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
\left< \delta M^a \delta M^b \right>.
\label{eq_fvar}
\end{equation}

\subsubsection{Fluctuating stress}

For the stress, the random contribution to the distributions is
\begin{equation}
\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
\end{equation}
where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
variates drawn from a Gaussian distribution with variance given
by equation~\ref{eq_fvar}. In the case that the shear and bulk
viscosities are the same, i.e., there is a single relaxation
time, then the variances of the six independent components of
the matrix are given by
\begin{equation}
\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
\frac{2\tau + 1}{\tau^2}
(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
\end{equation}

\subsubsection{Fluctuating ghost modes: D3Q15}

\subsubsection{Fluctuating ghost modes: D3Q19}

\subsection{The Propagation}

\subsection{Solid bodies and Bounce-Back on Links}

A very general method for the representation of solid objects
within the LB approach was put forward by Ladd \cite{l94a, l94b}.
Solid objects (of any shape) are defined by a boundary surface
which intersects some of the velocity vectors $\mathbf{c}_i$
joining lattice nodes. Sites inside are designated solid, while
sites outside remain fluid. The correct boundary condition is
defined by identifying \textit{links} between fluid and solid
sites, which allows those elements of the distribution which would
cross the boundary at the propagation step to be ``bounced-back''
into the fluid. This bounce-back on links is an efficient method
to obtain the  correct hydrodynamic interaction between solid
and fluid.

\subsubsection{Boundary walls}

\subsection{Colloids}

\subsubsection{General}

Colloidal particles are assumed to be spherical with a geometrical
centre $\mathbf{r}_c$, which is also the centre of mass. The
centre is allowed to move continuously across the lattice
with velocity $\mathbf{U}$; the particle has an angular velocity
$\mathbf{\Omega}$. The surface of the colloid is defined by an
input radius, $a_0$, which determines which lattice nodes are
inside or outside the colloid. (The hydrodynamic properties of
the colloid are specified by a different radius $a_h$ --- more
of this later.) The boundary links are then the set of vectors
joining lattice nodes which intersect the spherical surface
$\{\mathbf{c}_b\}$. A schematic picture is presented in
Figure~\ref{fig_coll2}. Note that a lattice node exactly at
the solid-fluid interface is defined to be outside the colloid.

In the original approach of Ladd, fluid occupied nodes both inside
and outside the particle. The effect of the ``internal fluid'' is
known to be restricted to short time scales (compared to the
characteristic time $a_0^2/\nu$), on which the fluid inside the
particle relaxes to a solid body rotation \cite{heemels}. There
are a number of problems related to the internal fluid, so we
use fully solid particles.

Nguyen and Ladd.

For each boundary link, it is useful to define a boundary vector,
$\mathbf{r}_b$, which connects the centre of the colloid to the
appropriate boundary node (note that the length of this vector is
not necessarily equal to the input radius of the colloid).

\begin{figure}[tb]
\begin{center}
\includegraphics{xfig/colloidhalflinksnew.eps}
\end{center}
\caption{Boundary links for a colloidal particle in two dimensions
with D2Q9 (half the particle is shown). Links join fluid sites (red)
to solid sites (green) and
intersect the circular shell radius $a_0$. Boundary nodes (circles)
lie exactly half way between pairs of lattice nodes. The generalisation
to three dimensions is straightforward.}
\label{fig_coll2}
\end{figure}

A \textit{boundary node} halfway between fluid nodes which are joined
by a boundary link. The position of the boundary node is always
exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
\Delta t$) regardless of the actual position of the intersection
of the colloid surface and the boundary link.
The colloid therefore has a discrete representation which becomes a
better approximation to the sphere as the input radius becomes larger;
this approximation is known to be reasonably good for $a_0 > 5\Delta x$.

\subsubsection{Bounce-back on links}

The standard boundary condition required for the solid-fluid
interface of a moving particle is described by Ladd \cite{l94b},
and is generally refered to as bounce-back on links (BBL).
A boundary link is defined as joining a node $\mathbf{r}$
inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
If the post-collision distributions are denoted by $f^\ast$, then
the distributions must be reflected at the solid surface so that
\begin{equation}
\label{eq:bbl1}
f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
\end{equation}
where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
at the boundary
\begin{equation}
\label{eq:ub}
\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
\end{equation}
is determined by the particle linear velocity $\mathbf{U}$ and angular
velocity $\mathbf{\Omega}$. The change in momentum described by

The force exerted on a
single link is
\begin{equation}
\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
\end{equation}
with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.

The total hydrodynamic force on the particle is then found by taking
the sum of
$\mathbf{F}_b$ over all the boundary links defining the particle.
The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
which again is summed over all links to give the total torque on the colloid.

Note that $\rho$ in the above equations is the density of the fluid
at the appropriate fluid node for the link. As the density fluctuations
in the fluid are small compared with the mean density $\rho_0$, the
density in the correction to the bounce-back can be replaced by the
mean $\rho_0$.

\subsubsection{Dynamics}
Having computed the total force and torque on an individual colloid,
is is possible to update the the linear velocities
\begin{equation}
m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
\end{equation}
where the mass of the colloid is related to the input radius by
$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
velocity
\begin{equation}
I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
\mathbf{T}(t),
\end{equation}
where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
However, this explicit update is generally found to have poor stability
properties \cite{l94b, nl02}. The alternative is to use a velocity update
which is implicit \cite{heemels,nl02}.

The total force and torque on a colloid can be split into velocity-dependent
and -independent parts by combining Equations~(\ref{eq:ub})
and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
In this way, the decomposition is
\begin{eqnarray}
\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
-\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
\label{eq:d3}
\\
\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
-\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
\label{eq:d4}
\end{eqnarray}
The velocity independent parts of the force and the torque
(appropriate for a colloid at rest) are
\begin{eqnarray}
\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
\\
\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
(\mathbf{r}_b\times \mathbf{c}_b)
\end{eqnarray}
The matrices $\boldsymbol{\zeta}$ are interpreted as drag
coefficients and can be written as
\begin{eqnarray}
\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
\\
\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
\\
\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
\\
\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
(\mathbf{r}_b \times \mathbf{c}_b).
\end{eqnarray}
As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is

\begin{equation}
\label{eq:k14}
\left( \begin{array}{rrr}
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\

\end{array} \right).
\end{equation}
If the particle has a symmetric distribution of boundary links (the
special case where the centre of mass is on a symmetry point of the
lattice) the $\boldsymbol{\zeta}^{FU}$ and
$\boldsymbol{\zeta}^{T\Omega}$ matrices are
diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
and $\boldsymbol{\zeta}^{TU}$ matrices are zero.

The velocity updates can now be rewritten using an
implicit update, which results in
\begin{eqnarray}
\label{eq:k15}
m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
\\
\label{eq:k16}
I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
\end{eqnarray}
In general all the matrix elements are non-zero, in which case
equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
6$\times$6 matrix inversion.

Finally, the position of the particle can be updated: an Euler
forward step is taken using the updated velocity
\begin{equation}
\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
\end{equation}

\subsubsection{Changes in particle shape}

One consequence is that the discrete shape of the particle fluctuates
as the colloid centre moves relative to the
lattice. This means that the size of the particle as seen by the
fluid changes despite the fact that the input radius of the colloid
is fixed.  However, these fluctuations are small for input radii
greater than about 5 lattice units \cite{l96a}. Furthermore, the
fluctuations may be greater for some input radii than others
\cite{nl02}.

In the standard approach, changes in the map of boundary links
are accomodated by the internal fluid. If the particle shape
changes so that an internal node is exposed, the internal fluid
at that site rejoins the fluid proper with the solid body momentum
(assuming the internal fluid is relaxed to solid-body rotation).
If a fluid node is covered by particle movement, then it simply
joins the internal fluid, and will relax to solid-body rotation.
Without internal fluid, changes in particle shape
in which fluid nodes are either covered or uncovered must be
accompanied by the removal or addition
of fluid with appropriate properties at the nodes in question.
It is essential that this
is done in a way which minimises the perturbation to the fluid
flow.

If a fluid node at $\mathbf{r}$ is covered by the movement of a
solid particle,
the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
explicitly so that
the overall mass density is unchanged. At the same time, the
colloid must assume the momentum lost by the fluid
$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.

Similarly, when a fluid node is exposed by particle movement,
fluid must be replaced with the appropriate distribution, density
and velocity. If the
new density is $\rho$ (to be determined by some method) then
the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
\Delta x^3$ which must be balanced elsewhere. If the new fluid
is assumed to have velocity $\mathbf{u}$ then the particle must
give up momentum $\Delta x^3 \rho\mathbf{u}$.

Following Nguyen and Ladd \cite{nl02}, these changes owing
to change in particle shape are implemented by added a small
correction to the bounce-back at the following time step.
The excess mass from a covered fluid site is redistributed over all the
boundary nodes by redefined the mometum transfer at bounce-back
\begin{equation}
f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
\label{eq:q1}
\end{equation}
where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
force on the particle owing to the change in shape is then
that lost by the fluid plus the contributions from the final term
in (\ref{eq:q1}) summed over all the boundary links
\begin{equation}
\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
\mathbf{c}_b.
\label{eq:q2}
\end{equation}

Note that for a closed surface,
$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
does not contribute to the net force and torque on the particle.
The corresponding change in torque is
\begin{equation}
\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
\times \sum_i f_i \mathbf{c}_i
+ \frac{\Delta x^3}{\Delta t}
\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
\label{eq:q2a}
\end{equation}
where $\mathbf{r}_c$ is the boundary vector at the position where the
fluid has been removed. Again, for a closed surface, the
redistribution of mass does not contribute to the net torque
on the particle.

Likewise, the contribution to the bounce-back from newly uncovered nodes
is
\begin{equation}
- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
\label{eq:q3}
\end{equation}
leading to a change in force of
\begin{equation}
\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
\label{eq:q4}
\end{equation}
with a corresponding change in the torque.
If more than one lattice node is either covered or uncovered by
the movement of the particle, then these contributions add in a simple
way. The overall contributions can be added to the
velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).

\subsubsection{Particles near contact}

Particles near contact may not pocess a full set of boundary
links (see Figure~\ref{fig:f5}). This leads to a potential
non-conservation of mass associated with the particle motion
which must be corrected.

\begin{figure}[tb]
\begin{center}
\includegraphics{xfig/colloidclose.eps}
\end{center}
\caption{Two colloids close to contact are missing links in the
region of closest approach owing to a lack of fluid nodes in the
interstice.}
\label{fig:f5}
\end{figure}
Again following Nguyen and Ladd \cite{nl02}, mass conservation is
enforced explicitly using the following procedure. There is a mass
transfer associated with the bounce-back at each link of
$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
The net mass transport into the particle is the sum of this
quantity over all the links, and can be written
\begin{equation}
\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
\end{equation}
For any partilce with a full set of boundary links, both
$\sum_b w_{c_b} \mathbf{c}_b$
and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
However, if the set of boundary links is imcomplete
(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
a correction is required. This is again made by redistributing the
excess or deficit of mass over the other existing boundary nodes.
The additional contribution to the bounce-back here is
$-w_{c_b} \rho_0 \Delta M_s/A$, where
$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).

This contribution
now adds to the velocity-dependent part of the force and torque and
leads to a slightly different form of the drag matrices
\begin{eqnarray}
\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
\\
\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
\overline{\mathbf{r}_b\times\mathbf{c}_b}),
\\
\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
(\mathbf{c}_b - \overline{\mathbf{c}_b}),
\\
\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
\end{eqnarray}
The mean quantities
\begin{equation}
\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
\end{equation}
and
\begin{equation}
\overline{\mathbf{r}_b\times\mathbf{c}_b} =
\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
\end{equation}
A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
can be replaced by their deviation from the mean. This provides a
convenient way to compute the drag matrices (maintaining symmetry)
and computing the correct force and torque on the particle when close
to contact.

\subsubsection{Rotational motions}

For a number of applications it is useful to update not only
the psoition of a spherical particle, but its orientation. This
is important, e.g., when considering magnetic particles. The
best way to implementment the required rotations is by
considering quaternions.

A quaternion can be thought of as an extended complex number
\begin{equation}
q = w + xi + yj + zk
\end{equation}
where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
of a quaternion is
\begin{equation}
N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
\end{equation}
Addition and subtraction of quaternions proceeds as normal, but
care is required for multiplication, which is associative but not
commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
$\bf{v}$ is a normal 3-vector, then multiplication can be written
\begin{equation}
q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
+ {\bf v}_1 \times {\bf v}_2),
\end{equation}
where the usual scalar and vector products are used. Note that the
conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
$qq^\star = N^2(q)$.
It can be shown that a rotation of an arbitrary vector though an
angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
be represented by the quarternion
\begin{equation}
q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
\end{equation}
which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
then the rotated vector ${\bf v}^{'}$ can be written as
\begin{equation}
(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
\end{equation}
which can be expanded as
\begin{equation}
{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
\end{equation}

\subsection{Colloid-Colloid interactions}

\subsubsection{Lubrication corrections}

\subsubsection{Hard sphere interaction}

Particles do not overlap, i.e., there is always
a hard-sphere interaction which is a function of the separation $h$:
\begin{equation}
v^{hs}(h) = \left\{
\begin{array}{ll}
\infty & h \leq 0,\\
0   & h > 0.
\end{array} \right.
\end{equation}
The hard-sphere interaction does not give rise to a force.

\subsubsection{Soft-sphere interaction}

A simple soft-sphere interaction is available, the basic form
of which is:
\begin{equation}
v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
\end{equation}
where $\epsilon$ sets the energy scale, and $\sigma$ sets the
characteristic width. The steepness of the potential is set by
the exponent $\nu (> 0)$.

To prevent the need for calculation of long-range interactions,
the soft-sphere potential is truncated in a ``cut-and-shift''
approach. This is done in such a way as to smoothly match
both the potential and the force at the cut-off distance
$r_c$. For $r < r_c$ the potential is then
\begin{equation}
v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
\label{eq_ss_shift}
\end{equation}
Clearly, this potential is not exactly what we first thought of as
soft-sphere. Matching the potential smoothly ensures conservation
of energy, while matching the force smoothly prevents potential
instabilities in the molecular dynamics update.

The soft-sphere potential can be useful as a mechanism to
keep the particles from touching (which risks hard-sphere interactions),
in which case is is possible to compute the interaction as a function
of the separation $h$, instead of $r$. The force between
two particles is computed via the derivative of equation~\ref{eq_ss_shift}
with respect to $r$:
\begin{equation}
\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
\end{equation}
where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
particle $i$ to the centre of $j$.

\subsubsection{Leonard-Jones interaction}

\section{Lattice Boltzmann for a Binary Fluid}

\subsection{The Cahn-Hilliard Equation}

\subsection{The Choice of Free Energy}

\subsection{A Second Distribution Function}

\subsection{The Collision}

\subsection{Bounce-Back on Links}

\section{Code Design}

\subsection{Some Comments}

In designing a code in C the main object of which is to do high
performance computing, there are some inevitable tensions with
software engineering best-practice. In particular, complete
encapsulation of data in the spirit of using abstract data types
tends to cause problems of performance.

For example, encapsulation of the lattice Boltzmann distribution
$f_i$, together with appropriate ``get'' and ``set'' functions for
access when approapriate, is possible. There could then be a
complete separation of interface and implementation of the
distribution (other than that each individual $f_i$ is represented
by a \texttt{double}). However, experience suggests that
this imposes severe impediment to optimisation and hence
unacceptable performance degradation.

Schemes do exist by which one can approximate the object
oriented programming style in C, and so achieve a separation
of interface and implementation. However, this tends to put
a heavy burden on the use of \texttt{void} pointers, itself
a recipe for possible disaster. A truly object-oriented
language such as Java would be preferable.

So, for the time being, some relaxation of encapsulation is required.
Exposing a pointer to the distributions which allows them to be accessed
directly where required. This can be done merely by declaring the
necessary variables \texttt{extern}. In practice, such deviations
from encapsulation can be minimised in the lattice sector of the
code. In particlar, \texttt{extern} declarations should not be
required in \texttt{main()}, i.e., it should be able to write a
new application without worrying about details of implementation.

\subsubsection{Objects}

Somewhat more difficult is the part of the code involving
colloids, which would really favour a truly object-oriented
approach.

\subsubsection{Serial and parallel}

The design decission has been taken to support execution of the
code in both serial and parallel. The default compilation
assumes that a serial executable is required, while  MPI code
is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).

There is no shared memory inplementation at this time.

\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}

\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}

\begin{figure}
\begin{center}
%\includegraphics{xfig/sketch.eps}
\end{center}
\caption{A scheme by which different parts of the code might be
encapsulated and in such a way to allow reliable development and
testing. As this is C, will do not adopt a truly ``object'' langauge.
Broadly, different modular parts should only depend on stuff above
in the diagram. So, at the top, there is the parallel environment
where communication occurs within MPI\_COMM\_WORLD. Built on top
of this is the Cartesian Communicator responsible for halo swaps
and so forth. The LE code may have a separate Communicator again. 
(There's an argument to say the diagram should be the other way up!)}
\end{figure}

\section{Code Implementation}

\subsection{Stuff}

\subsection{Coordinate System}

\subsubsection{The global coordinate system}

The user specifies the system size in lattice units at run time,
which sets the extent of memory which must be allocated for the
Cartesian lattice. The extent of the physical coordinate system
then reflects the lattice size directly (see Figure \ref{fig_c1}). 
Each lattice site, or node,  has coordinates corresponding to its
integer position in the system and can be thought of as being
at the centre of a control volume one lattice unit on a side.
This means that the nominal edges of the
system are offset from the lattice sites themselves by half a
lattice unit in each direction.

\begin{figure}[h]
\begin{center}
\includegraphics{xfig/fig_c1.eps}
\end{center}
\caption{The global Cartesian coordinate system reflects the choice
of lattice size; lattices sites are located at
$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
(the Figure only shows the labels in the $y-$direction for simplicity).
In the Figure, the lattice sites in the domain proper are the full
circles; halo sites, which are used to effect periodic boundary
conditions, are included as open cirles}
\label{fig_c1}
\end{figure}

For particles, the centres of which can move continuously across the
lattice, the extents of the global coordinate system are then
$L_{min} < x < L + L_{min}$ and so on.

\subsubsection{Addressing the lattice: the local coordinate system}

\subsection{Solid walls}

General solid objects which are fixed for the duration of
the simulation, such as porous media, can be defined
by assigning solid/fluid status to each lattice node. 
The links required between solid and fluid nodes can
then be identified for the given model.

\textit{Comment.} In general, a solid object will not
affect the periodic nature of the simulation. A general
solid object cannot move, i.e., the boundary velocity
must be zero.

\textit{Special case.} In the case where the solid object
represents a flat boundary wall at the edge of the system
in one or more directions, the periodicity is broken in
the corresponding directions.

\vfill
\pagebreak

\section{Code Interface}

\subsection{The Parallel Environment}

\texttt{\#include ``pe.h''}

As the code is to be available in both serial and parallel versions,
it is desirable to abstract the basic environment to some extent.
The parallel environment is intended to support MPI communication
within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
consistent picture in serial (i.e., rank zero is the single process).

\texttt{void pe\_init(int argc, char ** argv)}\\
Responsible for initialisation of MPI, and hence must be first
execuatable statement. The arguments are those to \texttt{main}
which will be passed through to \texttt{MPI\_Init()}.

\texttt{int pe\_rank(void)}\\
Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
serial.

\texttt{int pe\_size(void)}\\
Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
1 in serial.

\texttt{void pe\_finalise(void)}\\
Responsible for \texttt{MPI\_Finalize()}, and must be the last
executable statement.

\texttt{void info(const char * fmt, ...)}\\
Behaves like \texttt{printf}, but only produces output on the
root process.

\texttt{void verbose(const char * fmt, ...)}\\
Behaves as \texttt{printf} and reports the rank of the
calling process in \texttt{MPI\_COMM\_WORLD}.

\texttt{void fatal(const char * fmt, ...)}\\
Prints an error message and identifies
the rank of the calling process before terminating the program.

\subsection{Run Time Input}

\texttt{\#include ``runtime.h''}

The ability to set a wide range of parameters at run time is
very convenient, and is supported via an input file in which the
user specifies parameters as key value pairs. The code maintains
a list of key value pairs from which the user can retrieve the
appropriate values for given keys.

\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
This reads the named input file and sets up the list of keys.

\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
Set the \texttt{value} associated with \texttt{key}, if present. The number
of key matches found is returned (i.e., 0 or 1).

\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
Set the \texttt{value} associated with \texttt{key}, if present. The
number of key matches is returned.

\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
Set \texttt{value} to the string assocaited with \texttt{key}, if
present. Returns the number of key matches found.

\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
If \texttt{key} is present, set the associated 3-vector of integers.
Returns the number of key matches found.

\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
double v[])}\\
If \texttt{key} is present, set the associated 3-vector of doubles.
Returns the number of key matches found.

\textit{Comment:} We probably need a function to report unused keys, so
the user can be alerted if a key has unexpectedly been ignored.

\subsection{Random Number Generators}

\texttt{\#include ``ran.h''}

Note that the random number generation in parallel is
decomposition dependent.

\texttt{void rand\_init(void)}\\
Responsible for initialising RNG state.

\texttt{double rand\_serial\_uniform(void)}\\
Return a single variate uniformly distributed on $[0,1]$ from
the serial generator.

\texttt{double rand\_serial\_gaussian(void)}\\
Return a single variate with Gaussian distribution about zero
and having unit variance from the serial generator.

\textit{double rand\_parallel\_uniform(void)}\\
Return a single variate uniformly distributed on $[0,1]$
from the serial generator.

\texttt{double rand\_parallel\_gaussian(void)}\\
Return a single variate with Gaussian distribution about zero
and having unit variance from the parallel generator.

\textit{Comment:} The serial generator will produce the
same sequence on any MPI process, while the parallel generator
will produce different sequences. The latter is intended for
generating, specifically, fluctuations, where the need for
decomposition-independance is questionable.

\subsection{Timers}

\texttt{\#include ``timer.h''}

Performance figures require timing for different sections
of the code. The code provides a set of routines to time
standard and user-defined sections of code. Timers are
started and stopped by call to the appropriate routines.

\texttt{TIMER\_init()}\\
This initialises all the timers to zero (e.g., at the start of
the program).

\texttt{TIMER\_start(const int t\_id)}\\
Start the given timer.

\texttt{TIMER\_stop(const int t\_id)}\\
Stop the given timer.

\texttt{TIMER\_statistics(void)}\\
Prints the current state of all the active timers to standard
output on the root process in a digestible form. In serial,
the minimum, maximum, and mean time for each timer is reported
in seconds (based on the number of start/stop cycles). In
parallel, the same figures are computed on each process and the
result of the corresponding \texttt{MPI\_Reduce()} operation is
printed on the root process.

The code provides a number of timers which would time standard
parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
total execution time, and a number of spare timers for users to
use at will.

\textit{Comment:}
No intra-run variance (i.e., between time steps) is provided, as the
calculation required is considerably more complex to generalise, and
requires memory proportional to the number of time steps used.

\textit{Comment:}
Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
provided to insert calls to the above functions for optional timing
of performance-sensitive regions
of the code in conjunction with the preprocessor flag
\texttt{-D\_TIMING\_MACRO\_ON\_}.

\subsection{Model Time Control}

This provides time step control, event frequency (?)

\texttt{void init\_control(void)}\\
Initialise time controls.

\texttt{int get\_step(void)}\\
Return the current model time step.

\texttt{int next\_step(void)}\\
Increments the current time step. Returns number of steps left
to run (i.e., 0 when finished).

\subsection{Coordinate System}

\texttt{\#include ``coords.h''}

A three-dimensional Cartesian coordinate system is used throughout,
with the exact system size usually chosen by the user at run time.
The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
used to identify the coordinate directions.

This also provides access to the MPI Cartesian Communicator which
is responsible for most communication in the model.

\texttt{int N\_total(const int dim)}\\
Returns the total number of lattice sites in the given Cartesian direction.

\texttt{double L(const int dim)}\\
Returns the length of the system in the given Cartesian direction
(which is the same as the number of lattice sites).

\texttt{double Lmin(const int dim)}\\
Returns coordinate position of left-hand edge of system in given
Cartesian direction.

\texttt{int is\_periodic(const int dim)}\\
Returns 1 if given Cartesian direction is periodic, else 0.

\texttt{void cart\_init(void)}\\
Responsible for setting up the Cartesian decomposition.

\texttt{int cart\_rank(void)}\\
Returns rank in the Cartesian communicator.

\texttt{MPI\_Comm cart\_comm(void)}\\
Returns the MPI Communicator handle for the Cartesian communicator.

\texttt{int cart\_size(const int dim)}\\
Returns the size of the Cartesian communicator grid in the given
coordinate direcxtion.

\texttt{int cart\_coord(const int dim)}\\
Returns the coordinate of the current process in the Cartesian
Communicator.

\texttt{int cart\_neigh(const int dir, const int dim)}\\
Returns the rank of the neighbouring process in the
Cartesian communicator.

\subsection{Colloids}

\texttt{\#include ``colloids.h''}

These functions provide the basic interface for use
of colloidal particles in the code.

\subsubsection{Colloid structure}

The basic colloid struct is defined ...

\subsubsection{Colloid interface}

\texttt{void colloids\_init(void)}\\
Call once for initialisation. This initialises the cell list,
so the coordinate system must be initialised via a call to
\texttt{coords\_init()} first.

\texttt{void colloids\_finish(void)}\\
Call once for finalisation. Frees all colloid-related memory.

\texttt{Colloid * allocate\_colloid(void)}\\
Returns a pointer to a newly allocated \texttt{Colloid} object
or fails gracefully.

\texttt{void free\_colloid(Colloid * p\_colloid)}\\
Deallocates all memory associated with \texttt{p\_colloid} or
fails gracefully.

\texttt{struct c\_link allocate\_boundary\_link(void)}\\
Returns a pointer to a newly allocated colloid boundary
link or fails gracefully.

\texttt{void free\_boundary\_link(struct c\_link *)}\\
Deallocates memory associated with a link or fails gracefully.

\texttt{void colloids\_memory\_report(void)}\\
The module keeps track of how much memory has been allocated in
total for colloids and boundary links. This reports the current
usage.

\texttt{get\_Ncell\_total()}\\

\texttt{get\_Ncell\_local()}\\

\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
This returns a pointer to the Colloid at the head of the list at
cell list location (i, j, k) or \texttt{NULL} if there is none.

\texttt{void cell\_insert(Colloid * p\_colloid)}\\
This inserts a colloid at the head of the appropriate list
depending upon its position. This is the only way a colloid
can be added to the list.

\texttt{update\_cells(void)}\\
Performs the update of the cell list in response to updated
particle positions.

\texttt{IVector cell\_coords(FVector r)}\\

\subsection{Speculation}

\subsection{The Free Energy}

It is imagined that a single file will encapsulate all the details
of a given free energy. The interface for different choices of
free energy should be the same.

\texttt{FE\_init()}

Responsible for querying the runtime input key value pair list to
find appropriate user parameters for the free energy, or
falling back to default values.

\texttt{double FE\_get\_sigma()}

Return the surface tension. There should be a similar routine
for the interfacial thickness.

\texttt{double FE\_get\_mu(const type location)}

Return the chemical potential at the single lattice
node \texttt{location}.

\textit{Comment:} The computation of quantities such as the chemical
potential will a) affect performance and b) require halo values of
the order parameter (depending on the highest derivative
in the order parameter present). Performance here should be the
first consideration, so this needs to be handled with care.

\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}

Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
given location.

\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
the number of dimensions. If we are to provide explicit support for
D2Q9, \texttt{ND} will be 2, otherwise 3.

\section{Implementation Notes}

\subsection{Serial Version}

\subsection{Efficiently Scaling Parallel Computing}

We should be thinking about the 100,000 processor regime, and
correspondingly large system sizes. To avoid embarrassment in this regime,
some issues which don't usually arise are relevant, for example:

\begin{enumerate}
\item
Large integer quantities such as
\begin{verbatim}
   N_total.x*N_total.y*N_total.z
\end{verbatim}
are likely to cause integer overflow (if default integers are 4 byte).
\item
Local data structures whose size is proportional
to the number of processors are contraindicated, e.g.,
anything like
\begin{verbatim}
p = (foo *) malloc(nprocs*sizeof(foo));
\end{verbatim}
is to be avoided. At the moment, we are in good shape on this
(apart from some functions in \texttt{misc.c}).
\item
I/O is likely become a very important issue, so may need to think
more about this.
\item
We should probably introduce a standard set of benchmarks which can
be scaled up in some systematic way, so that we can keep track of
performance over time.

\end{enumerate}

\section{Model Details}

\subsection{D3Q15}

These are the eigenvalues and eigenvectors of teh collision
matrix used for D3Q15

\begin{table}
\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
\hline\hline
$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
\hline
$\rho$ & - &
 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
1 &$\mathbf{1}$ \\
\hline
$\rho c_{ix}$ & - &
 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
3  & $c_{ix}$ \\
\hline
$\rho c_{iy}$ & - &
 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
3  &$c_{iy}$ \\
\hline
$\rho c_{iz}$ & - &
 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
3  & $c_{iz}$ \\
\hline
$Q_{xx}$ & 1/3 &
-1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
9/2  & $c_{ix} c_{ix} - c_s^2$ \\
\hline
$Q_{yy}$ & 1/3 &
-1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
\hline
$Q_{zz}$ & 1/3 &
-1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
\hline
$Q_{xy}$ & - &
 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
9  & $c_{ix} c_{iy}$ \\
\hline
$Q_{yz}$ & - &
 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
9  & $c_{iy} c_{iz}$ \\
\hline
$Q_{zx}$ & - &
 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
9  & $c_{iz} c_{ix}$ \\
\hline\hline
$\chi^1$ & - &
-2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
1/2 & $\chi^1$ \\
\hline
$J_{ix}$ & - &
 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
3/2 & $\chi^1 \rho c_{ix}$\\
\hline
$J_{iy}$ & - &
 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
3/2 & $\chi^1 \rho c_{iy}$\\
\hline
$J_{iz}$ & - &
 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
3/2 & $\chi^1 \rho c_{iz}$\\
\hline
$\chi^3$ & - &
 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
9 & $c_{ix} c_{iy} c_{iz}$ \\
\hline\hline
$w_i$ & 1/72 &
$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
 & $w_i$\\
\hline\hline
\end{tabular}
\end{table}

\subsection{D3Q19}

\begin{table}
\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
\hline\hline
$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
\hline
$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
& 1\\
\hline
$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
& 3 \\
\hline
$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
& 3\\
\hline
$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
& 3\\
\hline
$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
& 9/2\\
\hline
$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
& 9/2\\
\hline
$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
& 9/2\\
\hline
$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
& 9\\
\hline
$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
& 9\\
\hline
$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
& 9\\
\hline\hline
$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
& 3/4\\
\hline
$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
& 3/2\\
\hline
$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
& 3/2\\
\hline
$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
& 3/2\\
\hline
$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
& 9/4\\
\hline
$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
& 9/2\\
\hline
$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
& 9/2\\
\hline
$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
& 9/2\\
\hline
$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
& 1/2\\
\hline\hline
$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
& \\
\hline\hline
\end{tabular}
\end{table}

\pagebreak

\begin{thebibliography}{99}

\bibitem{adhikari2005}
Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
Fluctuating Lattice Boltzmann,
\textit{Europhys. Lett.}, \textbf{71}, 473 2005.

\bibitem{ald98}
Aidun, C.K., Y. Lu, and E.-J. Ding,
Direct analysis of particulate suspensions with inertia using the
discrete Boltzmann equation,
\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.

\bibitem{cj98}
Cichocki, B., and R.B. Jones,
Image representation of a spherical particle near a hard wall,
\textit{Physica A}, \textbf{258}, 273, 1998.

\bibitem{cpc}
Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.

\bibitem{h59}
Hasimoto, H., On the periodic fundamental solutions of the Stokes
equation and their application to viscous flow past a cubic array
of spheres.
\textit{J. Fluid Mech.}, \textbf{5}, 317.

\bibitem{heemels}
Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
particles using the lattice-Boltzmann method,
\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.

\bibitem{jo84}
Jeffrey, D.J., and Y. Onishi,
Calculation of the resistance and mobility functions for the two
unequal rigid spheres in low-Reynolds-number flow,
\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.

\bibitem{viv}
Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
P. Bladon,
Inertial effects in three dimensional spinodal decomposition of
a symmetric binary fluid mixture: A lattice Boltzmann study,
\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).

\bibitem{l94a}
Ladd, A.J.C., Numerical simulations of particulate suspensions
via a discretised Boltzmann equation. Part 1. Theoretical foundation,
\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.

\bibitem{l94b}
Ladd, A.J.C., Numerical simulations of particulate suspensions
via a discretised Boltzmann equation. Part 2. Numerical results,
\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.

\bibitem{l96a}
Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
spheres,
\textit{Phys. Fluids}, \textbf{9}, 491. 1996.

\bibitem{l96b}
Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
of non-Brownian spheres,
\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.

\bibitem{lv01}
Ladd, A.J.C., and R. Verberg,
Lattice-Boltzmann simulations of particle-fluid suspensions,
\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001

\bibitem{nl02}
Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
lattice-Boltzmann simulations of particle suspensions,
\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.

\bibitem{r95}
Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
Cambridge University Press, 1995.

\bibitem{succi}
S. Succi, \textit{The lattice Boltzmann equation and beyond},
2001.

\end{thebibliography}

\end{document}

### 2008-02-25 htdocs/sections.tex
Added information on solid wall options
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@187 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.1 2008-02-04 12:46:14 kevin Exp $
+%  $Id: sections.tex,v 1.2 2008-02-25 09:53:37 kevin Exp $
 %
@@
 gives a total (physical) system size of 64 lattice sites in each
-direction, decomposed across 4 processors in the x-direction, 2 in
+vdirection, decomposed across 4 processors in the x-direction, 2 in
 the y-direction and 1 (no decomposition) in the z-direction. The
@@
 
+Solid planar boundary walls can be added at the extremities of the
+system (i.e., the volume of fluid is extactly that specified by
+the \texttt{size} keyword). The wall can be in one, two or all three
+coordinate direcctions, i.e., a 'slab' geometry, and 'duct' geometry
+or a 'box' geometry, respectively.
+
+The boundary walls are switched on via the keyword value pair
+
+\inputkey{boundary\_walls\_on [no | yes]}
+
+If \texttt{yes} then the geometry is specified with the keyword
+as one of
+
+\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}
+
+Note that these are the only valid selections. \texttt{1\_1\_0}
+gives periodic boundaries in the $x-$ and $y-$directions, with
+walls at either edge in the $z-$direction. \texttt{1\_0\_0}
+gives periodic conditions in the $x-$ direction only, while
+\texttt{0\_0\_0} is a fully enclosed box.
+
+In the case of the slab geometry, two further values are used so
+that shear may be imparted to the system:
+
+\inputkey{boundary\_speed\_bottom}
+
+\inputkey{boundary\_speed\_top}
+
+These values set the $x-$component of the wall velocity at
+$z = z_{min}$ and $z = z_{max}$, respectively. Like all
+velocities, these values must respect the low Mach number
+constraint $u_x < c_s$.
 \subsection{Input/Output}

### 2008-08-29 htdocs/sections.tex
Added porous media information
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@406 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.2 2008-02-25 09:53:37 kevin Exp $
+%  $Id: sections.tex,v 1.3 2008-08-29 09:28:08 kevin Exp $
 %
@@
 
-\subsection{Ongoing Development}
-
-\subsubsection{Some notes}
-
-\begin{enumerate}
-
-\item
-Serial timer is exhausted at about 35 minutes.
-
-\item
-Size of the colloid message buffers needs to be computed robustly
-at run time, or, allocated on a just-in-time basis with knowledge
-of exact size required. The former would mean knowing the number
-of particles before the call to CCOM\_init\_halo.
-
-\end{enumerate}
-
-\subsubsection{Other Agenda Items}
-\begin{enumerate}
-
-\item
-The particle code needs a critic from someone other than myself.
-Again we need to abstract it away a bit more.
-
-\item
-Two-dimensional model (D2Q9).
-
-\item
-Finite Difference for order parameter.
-
-\item
-I/O via XML/HDF5.
-
-\end{enumerate}
+\subsection{Download}
 
@@
 zero velocity.
+
+\subsection{Porous Media}
+
+Porous media calculations can be undertaken when appropriate
+solid/fluid status information is supplied. The are a number
+of switches available in the input file:
+
+\inputkey{porous\_media\_file}
+
+specifies the file stub name to be read at the start of execution.
+(If the stub name is \texttt{file} then the code will expect to
+find \texttt{file.001-001} in the current directory.
+
+\inputkey{porous\_media\_format}
+
+is either \texttt{ASCII} or \texttt{BINARY} as appropriate. Note that
+in parallel, a single data file can be supplied, but it must be binary.
+The default is \texttt{BINARY}.
+
+\inputkey{porous\_media\_type}
+
+is either \texttt{status\_only} (the default) or \texttt{status\_with\_h}.
+
+\subsubsection{File format}
+
+In all cases the status (fluid/solid) information is represented by a
+single \texttt{char} (or integer), which must be supplied via the
+porous media file. A single file should contain data matching the
+current system size, and have the $z-$direction running fastest,
+followed by the $y-$direction. Note that in non-periodic directions,
+the structure must be 'closed', i.e., all the points at the edge
+should be solid.
+
+Fluid sites are designated by \texttt{0} and boundary or solid sites
+by \texttt{1}. These data should be of type \texttt{char} in binary,
+and may be integer in ASCII.
+
+Where wetting information is required, the free energy parameter $H$
+can be supplied by using the \texttt{status\_with\_h} switch. In this
+case, the \texttt{char} status is augmented by a single \texttt{double}
+value which is the local value of $H$. The order is then
+$s_1,h_1, s_2, h_2, \ldots$.
+
+\subsubsection{Permeability calculations}
+
+Single fluid permeability calculations for a given proous structure
+can be undertaken by driving a flow via the fluid body force. Note
+that the structure must be periodic in the direction of the force
+(this may mean duplicating a 'reflected' version of a given sample
+to create the correct input). The body force can be specified so
+that, e.g., to drive a flow in the $x-$direction
+
+\texttt{force 0.001\_0.000\_0.000}
+
+The force should not be so large that the maximum velocity generated
+threatens the Mach number constraint. To get a measurement of the
+flow at equilibrium, the calculation should be run at least the
+momentum diffusion time for the system ($L^2/\eta$ in LB time steps).
+
+The net flow can be measured by combining the statistics for the
+total momemtum and the total density (which is equal to the volume
+with $\rho_0 = 1$).
+
+Note that in the case of porous media with narrow channels at the grid
+scale, the wall velocity can be dependent on the viscosity of the fluid.
+This is an artefact of the bounce-back on links and will result in a
+viscosity-dependent permeability (see, e.g., \cite{lipanmiller}). To minimise
+this effect, it is recommended to to on the correction to the ghost
+mode relaxation time proposed by Ginzburg and d'Humi\`eres \cite{ginzburg}
+which corrects the viscosity-dependence of the apparent boundary
+position. This is done via the input switches
+
+\texttt{ghost\_modes on\\
+ginzburg\_dhumieres on}
+
+and should recover a viscosity-independent premeability measurement
+\cite{lipanmiller,chunladd}.
+
+\subsubsection{Capillaries}
+
+The utility \texttt{capillary.c} can be used to generate various
+porous media input files corresponding to various simple square
+or circular capillaries.
 
@@
 
+\bibitem{chunladd}
+B. Chun and A.J.C. Ladd,
+Interpolated boundary condition for lattice Boltzmann simulations in
+narrow gaps,
+\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
+
 \bibitem{cj98}
@@
 \textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
+
+\bibitem{ginzburg}
+I. Ginzburg and D. d'Humi\`eres,
+Multireflection boundary conditions for lattice Boltzmann models,
+\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
 
@@
 Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001
+\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
+
+\bibitem{lipanmiller}
+H. Li, C. Pan, and C.T. Miller,
+Pore-scale investigation of viscous coupling effects for two-phase
+flow in porous media,
+\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
 

### 2008-08-29 htdocs/sections.tex
Added porous media section correction.
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@407 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.3 2008-08-29 09:28:08 kevin Exp $
+%  $Id: sections.tex,v 1.4 2008-08-29 09:32:59 kevin Exp $
 %
@@
 to create the correct input). The body force can be specified so
-that, e.g., to drive a flow in the $x-$direction
+that, e.g., to drive a flow in the positive $x-$direction
 
@@
 viscosity-dependent permeability (see, e.g., \cite{lipanmiller}). To minimise
-this effect, it is recommended to to on the correction to the ghost
+this effect, it is recommended to switch on the correction to the ghost
 mode relaxation time proposed by Ginzburg and d'Humi\`eres \cite{ginzburg}
@@
 
-and should recover a viscosity-independent premeability measurement
+and should recover a viscosity-independent permeability measurement
 \cite{lipanmiller,chunladd}.

### 2008-09-20 trunk/htdocs/sections.tex
Added notes on capillary conductance
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@414 40876441-3496-0410-a613-a2fca35b2ae2

--- 2008-08-29 htdocs/sections.tex
@@
 %
-%  $Id: sections.tex,v 1.4 2008-08-29 09:32:59 kevin Exp $
+%  $Id: sections.tex,v 1.5 2008-09-20 10:39:58 kevin Exp $
 %
@@
 
-\subsubsection{Capillaries}
+\subsubsection{Example: circular capillaries}
 
 The utility \texttt{capillary.c} can be used to generate various
-porous media input files corresponding to various simple square
-or circular capillaries.
+porous media input files corresponding to various simple circular
+or square capillaries.
+
+For an infinite  capillary of circular cross section radius $a$,
+the conductance is known analytically\cite{papanastasiou}. The
+flow per unit area $J$ is given by
+\begin{equation}
+J = - \frac{1}{8\eta} \frac{\partial p}{ \partial x} a^2. 
+\end{equation}
+If the pressure gradient is replaced by a uniform body force
+$-\partial p/ \partial x = \rho g$, then one can define a
+viscosity-independent conductance $C$ via $ J = C \rho g / \eta$,
+i.e., $C = a^2/8$. So, by measuring the rate of change of flow
+with applied body force, one can compute an estimate of the conductance
+to compare with this result.
+
+If the total momentum is measured from the code for a given force
+(in say the $x$-direction), then the flux per unit area is
+$J = \sum_i \rho v_x / \sum_i \rho = \sum_i \rho v_x / V$
+where the sum is over all fluid lattice sites $i$, with $V$
+the fluid volume. The rate of change of flux with applied force
+$\Delta J / \Delta\rho g$ can then be combined with the viscosity
+to give the conductance.
+
+\subsubsection{Example: rectangular capillaries}
+
+An analtyical expression is also available for the conductance
+of a square or rectugular capillary. For an infinite capillary
+of square cross section width $w \times h$ [WHICH IS LONGER?],
+the volume flux per unit area $J$ is expected to be
+\cite{papanastasiou,edo1}
+\begin{equation}
+J = - \frac{1}{3\eta} \frac{\partial p}{\partial x} (h/2)^2
+\left[
+1 - 6(h/w) \sum_{k=1}^{\infty} \frac{\tanh(\alpha_k w/h)}{\alpha_k^5} 
+\right]
+\end{equation}
+where $\alpha_k = (2k - 1)\pi/2$. The pressure gradient
+$-\partial p / \partial x$ is replaced by the body force $\rho g$ and
+one can difine the viscosity-independnet conductance $C$ via
+$J = C\rho g / \eta$. The calculation procedes as above.
 
@@
 
+\bibitem{papanastasiou}
+T. paapnastasiou, G. Georgiou, and A. Alexandrou,
+\textit{Viscous Fluid Flow},
+CRC Press, Boca Raton, Florida, 2000.
+
 \bibitem{r95}
@@
 S. Succi, \textit{The lattice Boltzmann equation and beyond},
-2001.
+Oxford University Press, Oxford, 2001.
+
+\bibitem{edo1}
+M. Venuroli and E.S. Boek,
+Two-dimensional lattice-Boltzmann simulations of single phase
+flow in a pseudo two-dimensional micromodel,
+\textit{Physica A}, \textbf{362}, 23, 2006.
 

### 2008-11-10 htdocs/sections.tex
Some details on I/O
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@431 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.4 2008-08-29 09:32:59 kevin Exp $
+%  $Id: sections.tex,v 1.6 2008-11-10 11:23:58 kevin Exp $
 %
@@
 constraint $u_x < c_s$.
-\subsection{Input/Output}
+
+\subsection{Input/Output and Parallel I/O}
 
@@
 
+\inputkey{freq\_phi}
+
+set frequency of order parameter output.
+
+\inputkey{phi\_format}
+
+either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
+
+\inputkey{freq\_vel}
+
+set frequancy of velocity output.
+
+\inputkey{vel\_format}
+
+either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
+
 \inputkey{config\_at\_end}
@@
 sets the total number of Lees-Edwards planes. Default is zero.
+The placing is as followings. The number of planes $n$ must
+divide the lattice size in the $x$-direction to give an integer
+$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
 
@@
 
-sets the velocity of each plane relative to the lattice.
+sets the velocity of each plane relative to the lattice. All planes
+have the same, constant, velocity.
 
@@
 sets the random number generator seed.
+
+\subsection{Dealing with the Parallel I/O files}
+
+When running in parallel, output of lattice-based (and particle)
+quantities takes place in parallel. As a consequence, the order
+of the output in the files is not in the same order as would be
+the case in serial. This means the parallel output must be
+manipulated into the correct order before any analysis.
+
+Parallel I/O subdivides the processors into one or more groups,
+each of which write data to a separate file. Infomation on the
+content of each of these files is contained in `meta-data' file
+for each quantity, which are produced automatically at the start
+of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
+If there are 8 I/O groups, the files will be named \texttt{008-001}
+to \texttt{008-008} for the different groups.
+
+The actual data is stored for each relevant time step as, e.g.,
+\texttt{phi-000100.001-001} where the extention refers to the
+I/O. To recombined a single, or multiple, data files into a
+single file in the correct order, a utility program
+\texttt{extract.c} is provided.
+
+The exectutable takes two arguments, e.g.,
+
+\inputkey{./extract phi.001-001.meta phi-000100.001-001}
+
+where  the first is the first meta-data file in the relevant series,
+and the second is the first daa file for the relevant time step. The
+program will combine the relevant quantities in the correct order
+and produce a singel new file \texttt{phi-000100} (in the above case).
+
+\subsection{Restarting}
+
+When restarting in parallel, the processor decomposition must be
+preserved. An attempt to change the processor decomposition will
+cause the configuration files to be read in the incorrect order,
+and results will be wrong.
 

### 2008-12-03 htdocs/sections.tex
Added initial notes on surfactant free energy
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@444 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.6 2008-11-10 11:23:58 kevin Exp $
+%  $Id: sections.tex,v 1.7 2008-12-03 20:19:27 kevin Exp $
 %
@@
 
-The following apply for the symmetric binary fluid problem with
-order parameter $\phi$ and free energy (see section X) \cite{viv}
+The are currently three choices of free energy relating to
+various systems.
+
+\subsubsection{Symmetric binary fluid}
+
+\inputkey{free\_energy symmetric}
+
+The following applies for the binary fluid problem with
+compositional order parameter $\phi$ and free energy
+(excluding the term in the density $\rho$):
 \begin{equation}
@@
 \int dr \left(
-{\scriptstyle \frac{1}{2}}A\phi^2
-+ {\scriptstyle \frac{1}{4}}B\phi^4
-+ {\scriptstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right)
-\end{equation}
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
+\end{equation}
+
+This is described in some detail
+by Kendon et al \cite{viv}. The first two terms represent the bulk
+contribution, whereas the term in $\kappa$ penalisies curvature in
+the interface.
 
@@
 
-The parameter $\kappa$
-
-\inputkey{C}
-
-The surface free energy parameter {C}. Currently unused.
-
-\inputkey{H}
-
-The surface free energy parameter {H}. Currently unused.
+The parameter $\kappa$, which is positive.
+
+\subsubsection{Brazovskii free energy}
+
+This is similar to the symetric, but with one extra term which
+requires description.
+\begin{equation}
+ F[\phi] = 
+\int dr \left(
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
++ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
+\end{equation}
+
+\subsubsection{Surfactant model free energy}
+
+\inputkey{free\_energy surfactant}
+
+This is for use with a two order parameter model the first of
+which is the composition, as for the symmetric free energy,
+while the second $\psi$ represents surfactant concentration.
+The free energy density is made up of a number of terms:
+
+\begin{equation}
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+\end{equation}
+is the standard symmetric term related to the composition as above;
+
+\begin{equation}
+D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
+\end{equation}
+represents the energy of surfactant in the bluk phase with surfactant
+concentration $0 < \psi < 1$ with a single parameter $D$;
+\begin{equation}
+-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
+-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
+\end{equation}
+where the term in $\epsilon$ represents the energy reduction by
+adsorbing surfactant at the interface and the term in $\beta$
+provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
+\begin{equation}
++{\textstyle \frac{1}{2}} W \psi \phi^2
+\end{equation}
+which is included to stabilise the interface \cite{theissen}.
+
+The corresponding keys for the input file are:
+
+\inputkey{D}
+
+is the bulk surfactant contribution parameter
+
+\inputkey{epsilon}
+
+is the parameter $\epsilon$
+
+\inputkey{W}
+
+is the parameter $W$
+
+\inputkey{beta}
+
+is the value of the parameter $\beta$ (not implemented yet).
+
+\inputkey{phi\_b}
+
+the initial, uniform, background surfactant concentration $\psi_b$.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
+\hline
+Name & $\eta$ & $-A,B$ & $\kappa$ & $D$ & $\epsilon$ & $W$
+     & $M$ & $M_\psi$ \\
+\hline
+Run001 & 0.025 & 0.000625 & 0.004 & 0.0001 & 0.004 & 0.0003  & 2.0 & 0.1\\
+\hline
+\end{tabular}
+\caption{Acceptable parameters for the surfactant free energy model.
+Note that $\epsilon = \kappa$.}
+\end{center}
+\end{table}
+
+\subsection{Other Binary Fluid Parameters}
 
@@
 
-Sets the order parameter mobility in the LB framework.
+Sets the compositional order parameter mobility $M$ (uniform in space)
+which appears in the Cahn-Hilliard equation.
+
+\inputkey{mobilty\_psi}
+
+Sets the (uniform) mobility for the surfactant as appropriate for
+the Cahn-Hilliard equation.
 
@@
 
-The mean compositional order parameter roughly $-0.5 < \phi < 0.5$.
+The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
 The default value is zero, i.e., a symmetric 50:50 mixture by volume.
+
+\inputkey{phi\_initialisation}
+
+Determines how the compositional order parameter is initialisated
+at the start of the run. If set to \texttt{spinodal} the value
+is set to $\phi_0$ as set above plus or minus a random noise, the
+magnitude of which is set by the value of the \texttt{noise} key.
+If set to \texttt{block} a one-dimensional profile is set up in
+the $z$-direction representing two blocks of fluid with
+$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
+$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
+appropriate width. The $\phi = -1$ section is in the middle.
 
@@
 of a square or rectugular capillary. For an infinite capillary
-of square cross section width $w \times h$ [WHICH IS LONGER?],
+of square cross section width $w \times h$ (where $h$ is the
+longer),
 the volume flux per unit area $J$ is expected to be

### 2009-01-29 htdocs/sections.tex
Added information on surfactant free energy etc.
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@462 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.7 2008-12-03 20:19:27 kevin Exp $
+%  $Id: sections.tex,v 1.8 2009-01-29 12:25:43 kevin Exp $
 %
@@
 
+\section{Surfactants}
+
+This section describes the approach taken to study surfactants. Here,
+we always adopt the hydrid method, i.e., LB for the hydrodynmic
+equations and finite difference for the associated Cahn-Hilliard
+equations. Coupling between the two sectors is via the velocity
+field and the external force on the fluid computed via the
+divergence of the `chemical stress'.
+
+In this model the surfactant appears as a concentration field
+$\psi(\mathbf{r};t)$, alongside the compositional order parameter
+$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
+two corresponding Cahn-Hilliard equations: first,
+
+\begin{equation}
+\partial_t \phi + \partial_\alpha \phi u_\alpha = 
+\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
+\end{equation}
+and second
+\begin{equation}
+\partial_t \psi + \partial_\alpha \psi u_\alpha = 
+\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
+\end{equation}
+
+\subsection{Free Energy}
+
+A number of different free energies have been used to study
+surfactants [cite Yeomans]. Here, the free energy is based on
+that used by van der Graff and van der Sman and others [cite].
+There are a number of different contributions to the free
+energy density:
+\begin{equation}
+f = f_\phi + f_\psi +f_1 + f_{ex}.
+\end{equation}
+The first is exactly that used for the binary fluid, i.e.,
+\begin{equation}
+f_\phi (\phi, \partial_\alpha\phi) =
+{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
++ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
+\end{equation}
+The surface tension and the interfacial width are determined from the
+choice of parameters $A,B,\kappa$ as before.
+The second term in the free energu density represents the entropic
+energy of mixing of the surfactant with the bulk phases
+\begin{equation}
+f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
+\end{equation}
+where the parameter $kT$ sets the energy scale for the whole system.
+The surfactant order parameter, $\psi$, is unity for a saturated
+interface and zero for a plain interface.
+
+The surfactant lowers the energy of the interface where it is adsorbed
+there, represented by the term
+\begin{equation}
+F_1(\psi, \partial_\alpha \phi) 
+= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
+  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
+\end{equation}
+where $\epsilon$ and $\beta$ are parameters. Note that only the
+first term appears in \cite{vandergraaf}, while the second term is
+based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
+the surfactant profile at the interface is symmetric (so-called
+Langmuir adsoption) whereas the term in $\beta$ induces an
+asymmetry (Frumpkin adsorption).
+Finally,  $f_{ex}$ is an enthalpic contribution introduced 
+to stabilise the interface \cite{theissengompper}
+\begin{equation}
+f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
+\end{equation}
+with $W$ a parameter.
+
+\subsection{Chemical potentials}
+
+The respective chemical potentials are obtained from the functional
+derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
+$\mu_\psi = \delta F / \delta \psi$. We obtain:
+\begin{equation}
+\mu_\phi =
+A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
++ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
++ \epsilon \partial_\alpha \phi \partial_\alpha \psi
++ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
++ \beta \psi^2 \partial_\alpha^2 \phi,
+\end{equation}
+and
+\begin{equation}
+\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
++ {\textstyle \frac{1}{2}}W\phi^2
+- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
+- \beta\psi (\partial_\alpha \phi)^2.
+\end{equation}
+As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
++ \tilde{P}_{\alpha\beta}$ with the isotrpic part being determined as (e.g.,
+\cite{theissengompper})
+\begin{equation}
+p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
+\end{equation}
+There is no closed form for the tensor part, which must be choosen to
+comply with the constraint that in mechanical equilibrium
+\begin{equation}
+\partial_\alpha P_{\alpha\beta} = \phi\partial_\alpha\mu_\phi +
+\psi\partial_\alpha\mu_\psi.
+\end{equation}
+In this way we obtain for the isotropic contribution
+\begin{eqnarray}
+p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
+- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
+- \kappa \phi \partial_\alpha^2 \phi
+- kT \ln(1-\psi) + W\psi\phi^2
++ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
+\nonumber\\
++ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
++ \beta \phi \psi^2 \partial_\alpha^2 \phi
+- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
+\end{eqnarray}
+and for the tensor part
+\begin{equation}
+\tilde{P}_{\alpha\beta} =
+(\kappa - \epsilon \psi - \beta\psi^2) 
+\partial_\alpha \phi \partial_\beta \phi.
+\end{equation}
+
 \section{Code Design}
@@
 
+\bibitem{vandergraaf}
+R.G.M. van der Sman and S. van der Graaf,
+Diffuse interface model of surfactant adsorption onto flat and
+droplet interfaces,
+\textit{Rheol. Acta} \textbf{46} 3 (2006).
+
+\bibitem{diamant}
+H. Diamant and D. Andelman,
+Kinetics of surfactant adsorption at fluid/fluid
+interfaces: non-ionic surfactants,
+\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
+
+\bibitem{theissengompper}
+O. theissen and G. Gompper,
+Lattice Boltzmann study of spontaneous emulsification,
+\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
+
 \end{thebibliography}

### 2009-02-23 htdocs/sections.tex
Correction to surfactant p0
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@471 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.8 2009-01-29 12:25:43 kevin Exp $
+%  $Id: sections.tex,v 1.9 2009-02-23 18:54:09 kevin Exp $
 %
@@
 choice of parameters $A,B,\kappa$ as before.
-The second term in the free energu density represents the entropic
+The second term in the free energy density
+represents the entropic
 energy of mixing of the surfactant with the bulk phases
@@
 \begin{equation}
-F_1(\psi, \partial_\alpha \phi) 
+f_1(\psi, \partial_\alpha \phi) 
 = -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
@@
 asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is an enthalpic contribution introduced 
+Finally,  $f_{ex}$ is [van der Sman claims]
+an enthalpic contribution introduced 
 to stabilise the interface \cite{theissengompper}
@@
 As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotrpic part being determined as (e.g.,
++ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
 \cite{theissengompper})
@@
 \end{equation}
-There is no closed form for the tensor part, which must be choosen to
+There is no closed form for the tensor part, which must be chosen to
 comply with the constraint that in mechanical equilibrium
 \begin{equation}
-\partial_\alpha P_{\alpha\beta} = \phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi.
+\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
+\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
 \end{equation}
@@
 - kT \ln(1-\psi) + W\psi\phi^2
+\nonumber\\
 + \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-\nonumber\\
++ \epsilon\phi\psi \partial_\alpha^2 \phi
 + 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
@@
 \end{eqnarray}
-and for the tensor part
+and for the tensor part we can have
 \begin{equation}

### 2009-05-06 trunk/htdocs/surfactant.tex
Surfactant information.
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@525 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  surfactant.tex
%
%  Some notes on surfactant work
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Surfactants}

\subsection{Note on Experimental Measurements}

Experimental measurements have generally been based on mechanical
methods which allow the determination of an interfacial tension
(for recent reviews, see \cite{chang,eastoe}). The degree of surface
activity is then accessed by measuring the the interfacial tension
$\sigma$ as a function of the bulk concentration $\psi_b$, and then
using the Gibbs equation
\begin{equation}
\psi_{0,\mathrm{eq}} = -\frac{1}{nRT}
\left( \frac{\partial \sigma}{\partial \ln \psi_b} \right)_T
\end{equation}
where $\psi_{0,\mathrm{eq}}$ is the equilibrium surface excess
concentration. The curve of $\psi_{0,\mathrm{eq}}$ against $\psi_b$
so obtained is thus relevant for a fixed temperature and is generally
refered to as the \textit{adsorption isotherm}. More recently, the
method of neutron reflection has been used to to access the interfacial
concentration directly (with good agreement with the mechanical
methods).

\subsubsection{Dynamic surface tension}

A newly formed interface (e.g., after shaking) in a surfactant
solution will have interfacial tension close to the bare value
$\sigma_0$. As the surfactant concentration at the interface
approaches its equilibrium value, the interfacial tension is
reduced. The quantity of interest is then the \textit{dynamical
surface tension} $\sigma(t)$. This process takes place on
timescales typically 10$^{-3}$--1 second, and is important in
many industrial processes such as coatings and formation of
soap lather.

At equilibrium, the final change in interfacial tension can
be obtained from Gibbs' equation together with a suitable
isotherm relating $\psi_0$ and $\psi_b$.

\subsubsection{Adsorption isotherms}

A simple choice of isotherm is Henry's law:
\begin{equation}
\psi_{0,\mathrm{eq}} = K_H \psi_b
\label{eq:iso:henry}
\end{equation}
where $K_H$ is the equilibrium adsorption constant. Note this has the
dimensions of length, and represents the thickness of bulk solution
holding the same quantity of surfactant as an interface with
concentration $\psi_{0,\mathrm{eq}}$. In this way it is also a
measure of the strength of surface activity for a given surfactant.
Henry's law has a number of drawbacks: there is no upper limit
on $\psi_0$, and is only valid at low concentrations.

A more practical choice is the Langmuir equilibrium isotherm
\begin{equation}
\psi_{0,\mathrm{eq}} = \psi^\star \frac{K_L \psi_b}{1 + K_L \psi_b}
\label{eq:iso:langmuir}
\end{equation}
where $\psi^\star$ is the maximum concentration supported at the
interface and the Langmuir equilibrium adsorption constant is $K_L$
(with units mol$^{-1}$~m$^3$). At low concentrations $ K_L\psi_b << 1$,
the Langmuir isotherm reduces to the Henry isotherm with
$K_H = \psi^\star K_L$.

One further case of interest here is the Frumkin isotherm
\begin{equation}
\psi_{0, \mathrm{eq}} = \psi^\star
\frac{K_F \psi_b}{K_F\psi_b + e^{-B\psi_{0,\mathrm{eq}} / \psi^\star}}
\label{eq:iso:frumkin}
\end{equation}
where the Frumkin equilibrium adsorption constant $K_F$ again
has units of mol$^{-1}$~m$^3$. The dimensionless parameter $B$
 is a measure of the degree to which the surfactant behaviour
at the interface is non-ideal. For $B=0$, the Frumkin isotherm
reduces to that of Langmuir.

\subsection{Note on the Theory of Diamant and Andelman 1996}

As we ultimately adopt a free energy approach, we first review one
free energy based work --- that of Diamant and Andelman
\cite{diamant96}. This introduces a uniform system with interface
of zero thickness at $z = 0$. A uniform dilute solution of
surfactant occupies $z > 0$. The excess free energy per unit area
of interface is then written
\begin{equation}
\Delta\sigma [\psi] = \int_0^\infty  \big\{ \Delta f[\psi (z)] +
f_0 [\psi (z) ] \delta(z)  \big\} \mathrm{d}z. 
\end{equation}
The first term here is a contribution from the bulk while the
second is that from the interface.

The uniform bulk solution has an ideal entropy of mixing
\begin{equation}
\Delta f(\psi) = (1/a^3)
\big\{ kT[\psi \ln\psi - \psi - (\psi_b \ln\psi_b - \psi_b)]
- \mu_b (\psi - \psi_b) \big\}.
\label{eq:da2}
\end{equation}
Here, $a$ is the characteristic size of the surfactant molecules.
The chemical potential $\mu_b$ and volume fraction of surfactant
in the bulk are fixed (with $\psi_b < 1$). The final term in
Eq.~\ref{eq:da2} is a contact term with a reservoir at $z = \infty$.

At the interface, we have a surfactant concentration $\psi_0$, which
is no longer small:
\begin{equation}
f_0 (\psi_0) = (1/a^2) \big\{
kT [ \psi_0 \ln\psi_0 + (1 - \psi_0) \ln(1 - \psi_0)]
- \alpha\psi_0 - {\textstyle\frac{1}{2}}\beta\psi_0^2 - \mu_1\psi_0 \big\}.
\end{equation} 
The first term is the full entropy of mixing. The second term represents
the reduction ($\alpha$ is positive) in surface energy owing to the
presence of surfactant. The third term represents an attractive
interaction between surfactants at the interface ($beta$ again positive).
The final term is a contact term between the interface and the adjacent
solution (at $z\rightarrow 0$).

\subsubsection{Equilibrium}

At equilibrium, where the chemical potential is the same throughout
the system $\mu (z) = \mu_b$, the DA recover the adsorption isotherm
\begin{equation}
\psi_{0,\mathrm{eq}} = \frac{\psi_b}{\psi_b
+ e^{-(\alpha + \beta\psi_{0,\mathrm{eq}})/kT}}.
\label{eq:iso:diamant}
\end{equation}
Comparison with the experiment forms Eq.~\ref{eq:iso:langmuir} and
Eq.~\ref{eq:iso:frumkin} reveals that $K_F$ is equivalent to
$e^{\alpha /kT}$ in the Frumkin isotherm. In addition, the equilibrium
equation of state is
\begin{equation}
\Delta \gamma_{\mathrm{eq}} = (1/a^2) [ kT \ln (1 - \psi_0) 
+ {\textstyle\frac{1}{2}}\beta \psi_0^2 ].
\end{equation}

\subsubsection{Out of Equilibrium}

DA96 go on to derive a variant of the Ward-Tordai equation \cite{wardtordai}
which describes the time evolution of the concentration of surfactant
at the interface:
\begin{equation}
\psi_0 (t) = \sqrt{D/\pi a^2}
\Bigg[ 2\psi_b \sqrt{t} 
- \int_0^\infty \frac{\psi_1 (\tau)}{\sqrt{t - \tau}} \mathrm{d}\tau \Bigg]
+ 2\psi_b - \psi_1.
\label{eq:wardtordai}
\end{equation}
When the time-scale for equilibration at the interface is fast compared
with that in the bulk, $\psi_0$ responds immediately to changes in
$\psi_1$. This is the case of \textit{diffusion limited adsorption} (DLA).
Here, the diffusive problem Eq.~\ref{eq:wardtordai} can be closed
using the isotherm Eq.~\ref{eq:iso:diamant} with $\psi_b$ replaced by
$\psi_1$. This can be solved numerically for $\psi_0 (t)$. If one further
makes the assumption that the equilibrium equation of state holds
approximately out of equililbrium, then one can also recover
$\Delta\gamma (t)$, that is, the dynamic surface tension.

\subsection{Current Model}

The current model is based the standard symmetric binary fluid
free energy to represent two fluids separated by a finite
interface of thickness $\xi_0$. This free energy is the usual
functional of the compositional order parameter $\phi$. This is
referred to as
\begin{equation}
f_\phi = {\textstyle\frac{1}{2}}A\phi^2
+ {\textstyle\frac{1}{4}}B\phi^4
+ {\textstyle\frac{1}{2}}\kappa (\partial_\alpha \phi)^2.
\end{equation}
Note that here, the excess free energy related to $f_\phi$ may
be integrated analytically to provide an expression for the
intefacial tension $\sigma_0 = 4\kappa\phi^\star/3\xi_0$ at
equilibrium.
(In one dimension the equiibrium profile
is $\phi(z) = \phi^\star \tanh(z/\xi_0)$ with $\phi^\star = \sqrt{-A/B}$
and $\xi_0^2 = -2\kappa/A$.)

To this we add a separate order parameter to represent the
surfactant $\psi$ \cite{smangraaf}. The free energy density
is made up of two parts following DA96:
\begin{equation}
f_{\psi} = kT[\psi\ln\psi  + (1 - \psi) \ln (1 - \psi)]
\end{equation}
represents the entropy of mixing, valid in both the bulk and at
the interface. The interfacial part of the free energy density is
\begin{equation}
f_{\psi,i} = -{\textstyle\frac{1}{2}}\epsilon\psi (\partial_\alpha \phi)^2
- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2.
\end{equation}

We note that the excess free energy can no longer be integrated
exactly. However, it can be integrated numerically for a
given profile $\psi(z)$ (see below).

\subsubsection{Equilibrium}

We can compute the chemical potential associated with each order
parameter by taking the functional derivative with respect to
$\phi$ and $\psi$ respectively:
\begin{equation}
\mu_\phi = A\phi + B\phi^3 - \kappa \partial_\alpha^2 \phi
+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
+ \epsilon \psi \partial_\alpha^2 \phi
+ 2\beta\psi \partial_\alpha \phi \partial_\alpha \psi
+ \beta\psi^2 \partial_\alpha^2 \phi,
\label{eq:mu:phi}
\end{equation}
and
\begin{equation}
\mu_\psi = kT[\ln\psi - \ln(1 - \psi)]
- {\textstyle \frac{1}{2}} \epsilon (\partial_\alpha \phi)^2
- \beta\psi (\partial_\alpha \phi)^2.
\label{eq:mu:psi}
\end{equation}

At equilibrium, we can set the chemical potential in the bulk and
at the interface to be equal. We assume that the presence of
surfactant does not affect the form of the equilibrium
composition profile so that in one dimension
$\phi(z) = \phi^\star \tanh(z/\xi_0)$ as before.
In the bulk we assume the concentration $\psi_b << 1$ so that
Eq.~\ref{eq:mu:psi} gives $\mu_{\psi,b} = kT\ln \psi_b$. At
the interface we ahve $\mu_{\phi,i} = kT[\ln\psi_0 - \ln(1 - \psi_0)]
- (1/2)\epsilon(\partial_\alpha \phi)^2
- \beta\psi_0 (\partial_\alpha \phi)^2$.
If we equate $\mu_{\psi, b}$ and $\mu_{\psi,i}$ we can recover the
Frumkin isotherm
\begin{equation}
\psi_{0,\mathrm{eq}} = \frac{\psi_b}{\psi_b +
e^{-({\scriptscriptstyle \frac{1}{2}}\epsilon + \beta\psi_{0,\mathrm{eq}})
(\phi^\star)^2 / \xi_0^2 kT}}.
\end{equation}
This is equivalent to Eq.~\ref{eq:iso:diamant} of DA96, where the
interfacial width has has appeared in place of $a$ (absorped into
the definition of the free energy density by DA96).

In the following we have $D = M_\psi \psi ( 1 - \psi)$. The
initial interfacial width is $\xi = 1.13$ (narrower than the
equilibrium value). This is a one-dimensional system of
length 100 lattice units. We choose to fix the binary fluid
parameters, and fix $\epsilon$ in relation to $\kappa$. The
isotherm is then varied by adjusting the value of $kT$.

\subsubsection{Parameters with $\mathbf{\xi_0 = 1.13}$, $\epsilon = \kappa/4$}

Here we keep the standard binary fluid parameters, having interfacial
width $\xi_0 = 1.13$.

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline
Name & $\eta$ & $-A,B$ & $\kappa$ & $kT$ & $\epsilon$ & $\beta$ & $W$
     & $M$ & $M_\psi$\\
\hline
R008a & 1/6 & 0.0625 & 0.04 & 1.6965$\times 10^{-3}$ & 0.01 & 0.0000 & 0.0 
      & 0.15 & 1.0 \\
R008d &  &  &  & & & 0.0025 &  &  &  \\
R008f &  &  &  & & & 0.0050 &  &  &  \\
\hline
R006a & 1/6 & 0.0625 & 0.04 & 8.48231$\times 10^{-4}$ & 0.01 & 0.0000 & 0.0 
      & 0.15 & 1.0 \\
R006d & &  &  &  & & 0.0025 &  &  &  \\
R006f & &  &  &  & & 0.0050 &  &  &  \\
\hline
\end{tabular}
\caption{Parameters}
\end{center}
\end{table}

\subsubsection{Parameters with $\mathbf{\xi_0 = 2.26}$, $\epsilon = \kappa/4$}

Here we relax the interfaical width by a factor of 2, but
keep $A,-B$ as before so that the surface tension is increased
by a factor of $\sqrt{2}$.

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline
Name & $\eta$ & $-A,B$ & $\kappa$ & $kT$ & $\epsilon$ & $\beta$ & $W$
     & $M$ & $M_\psi$\\
\hline
R002a & 1/6 & 0.0625 & 0.16 & 1.6965$\times 10^{-3}$ & 0.04 & 0.00 & 0.0 
      & 0.15 & 1.0 \\
R002d &  &  &  & & & 0.01 &  &  &  \\
R002f &  &  &  & & & 0.02 &  &  &  \\
\hline
R003a & 1/6 & 0.0625 & 0.16 & 5.6549$\times 10^{-4}$ & 0.04 & 0.00 & 0.0 
      & 0.15 & 1.0 \\

R003d & &  & &  &  & 0.01 &  &  & 1.0\\
R003e & &  & &  &  & 0.01 &  &  & 0.5\\
R003f & &  & &  &  & 0.02 &  &  & 1.0\\
\hline
R004a & 1/6 & 0.0625 & 0.16 & 8.48231$\times 10^{-4}$ & 0.04 & 0.00 & 0.0 
      & 0.15 & 1.0 \\
R004d & &  &  &  & & 0.01 &  &  &  \\
R004f & &  &  &  & & 0.02 &  &  &  \\

\hline
\end{tabular}
\caption{Parameters for the surfactant free energy model. There are
three values of the (Langmuir) isotherm $(1/K_L) = 0.1$ (R002a),
0.01 (R004a) and 0.001 (R003a).
See section X for notes on $W$.}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Name & $\bar{\psi}$ & $\psi_b$ & $\psi_{0\mathrm{eq}}$
     & $\sigma / \sigma_0$  & Comment \\
\hline
R008a &  $2.0\times 10^{-1}$ & $1.859\times 10^{-1}$ & $5.143\times10^{-1}$
      & 0.76 & ok\\
R008a &  $1.0\times 10^{-1}$ & $9.093\times 10^{-2}$ & $3.033\times10^{-1}$
      & 0.80 & ok\\
R008a &  $5.0\times 10^{-2}$ & $4.490\times 10^{-2}$ & $1.643\times10^{-1}$
      & 0.82 & ok\\
R008a &  $2.5\times 10^{-2}$ & $2.231\times 10^{-2}$ & $8.540\times10^{-2}$
      & 0.84 & ok\\
R008a &  $1.0\times 10^{-2}$ & $8.888\times 10^{-3}$ & $3.494\times10^{-2}$
      & 0.86 & ok\\
R008a &  $1.0\times 10^{-3}$ & $8.867\times 10^{-4}$ & $3.540\times10^{-3}$
      & 0.87 & ok\\
\hline
R008d &  $2.0\times 10^{-1}$ &  &  &  & unstable \\
R008d &  $1.0\times 10^{-1}$ &  &  &  & unstable \\
R008d &  $5.0\times 10^{-2}$ &  &  &  & unstable \\
R008d &  $2.5\times 10^{-2}$ &  &  &  & no equilibrium \\
R008d &  $1.0\times 10^{-2}$ & $8.455\times 10^{-3}$ & $4.535\times10^{-2}$
      & 0.85 & ok \\
R008d &  $1.0\times 10^{-3}$ & $8.835\times 10^{-4}$ & $3.617\times10^{-3}$
      & 0.87 & ok \\
\hline
R006a &  $2.0\times 10^{-1}$ & $1.713\times 10^{-1}$ & $8.510\times10^{-1}$
      & 0.71 & ok\\
R006a &  $1.0\times 10^{-1}$ & $7.556\times 10^{-2}$ & $6.657\times10^{-1}$
      & 0.74 & ok\\
R006a &  $5.0\times 10^{-2}$ & $3.406\times 10^{-2}$ & $4.211\times10^{-1}$
      & 0.78 & ok\\
R006a &  $2.5\times 10^{-2}$ & $1.616\times 10^{-2}$ & $2.310\times10^{-1}$
      & 0.81 & ok\\
R006a &  $1.0\times 10^{-2}$ & $6.286\times 10^{-3}$ & $9.653\times10^{-2}$
      & 0.84 & ok\\
R006a &  $1.0\times 10^{-3}$ & $6.192\times 10^{-4}$ & $9.868\times10^{-3}$
      & 0.87 & ok\\
\hline
R006d &  $2.0\times 10^{-1}$ & $1.664\times 10^{-1}$ & $9.791\times10^{-1}$
      & 0.64 & ok\\
R006d &  $1.0\times 10^{-1}$ &  & &  & unstable\\
R006d &  $5.0\times 10^{-2}$ &  & &  & unstable\\
R006d &  $2.5\times 10^{-2}$ &  & &  & unstable\\
R006d &  $1.0\times 10^{-2}$ & $2.9\times 10^{-4}$ & $4.75\times10^{-1}$
      & 0.80 & unsymmetric\\
R006d &  $1.0\times 10^{-3}$ & $5.822\times 10^{-4}$ & $1.077\times10^{-2}$
      & 0.87 & ok \\
\hline
\end{tabular}
\label{table:xi113}
\caption{Equilibrium results for $\xi_0 = 1.13$, $\epsilon = \kappa/4$.}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Name & $\bar{\psi}$ & $\psi_b$ & $\psi_{0\mathrm{eq}}$
     & $\sigma / \sigma_0$  & Comment \\
\hline
R002a &  $2.0\times 10^{-1}$ & $1.701\times 10^{-1}$ & $7.423\times10^{-1}$
      & 0.88 & ok\\
R002a &  $1.0\times 10^{-1}$ & $8.008\times 10^{-2}$ & $5.023\times10^{-1}$
      & 0.90 & ok\\
R002a &  $5.0\times 10^{-2}$ & $3.871\times 10^{-2}$ & $2.886\times10^{-1}$
      & 0.93 & ok\\
R002a &  $2.5\times 10^{-2}$ & $1.904\times 10^{-2}$ & $1.530\times10^{-1}$
      & 0.95 & ok\\
R002a &  $1.0\times 10^{-2}$ & $7.456\times 10^{-2}$ & $6.305\times10^{-2}$
      & 0.96 & ok \\
R002a &  $1.0\times 10^{-3}$ & $7.506\times 10^{-3}$ & $6.407\times10^{-3}$
      & 0.97 & ok \\
\hline
R002d &  $2.0\times 10^{-1}$ & $1.598\times 10^{-1}$ & $9.443\times10^{-1}$
      & 0.85 & ok \\
R002d &  $1.0\times 10^{-1}$ & $7.267\times 10^{-2}$ & $7.540\times10^{-1}$
      & 0.88 & ok \\
R002d &  $5.0\times 10^{-2}$ & $3.615\times 10^{-2}$ & $3.894\times10^{-1}$
      & 0.92 & ok \\
R002d &  $2.5\times 10^{-2}$ & $1.841\times 10^{-2}$ & $1.776\times10^{-1}$
      & 0.94 & ok \\
R002d &  $1.0\times 10^{-2}$ & $7.449\times 10^{-3}$ & $6.680\times10^{-2}$
      & 0.96 & ok \\
R002d &  $1.0\times 10^{-3}$ & $7.497\times 10^{-4}$ & $6.444\times10^{-3}$
      & 0.97 & ok \\
\hline
R002f &  $2.0\times 10^{-1}$ & $1.512\times 10^{-1}$ & $9.972\times10^{-1}$
      & 0.80 &  ok\\
R002f &  $1.0\times 10^{-1}$ & $5.994\times 10^{-2}$ & $9.751\times10^{-1}$
      & 0.80 & ok \\
R002f &  $5.0\times 10^{-2}$ & $3.002\times 10^{-2}$ & $6.947\times10^{-1}$
      & 0.90 & ok \\
R002f &  $2.5\times 10^{-2}$ & $1.746\times 10^{-2}$ & $2.202\times10^{-1}$
      & 0.95 & ok \\
R002f &  $1.0\times 10^{-2}$ & $7.338\times 10^{-3}$ & $7.132\times10^{-2}$
      & 0.96 & ok \\
R002f &  $1.0\times 10^{-3}$ & $7.487\times 10^{-4}$ & $6.480\times10^{-3}$
      & 0.97 &  ok\\
\hline
R003a &  $1.0\times 10^{-1}$ & $4.168\times 10^{-2}$ & $9.979\times10^{-1}$
      & 0.85 & ok\\
R003a &  $1.0\times 10^{-2}$ & $5.7\times 10^{-4}$ & $3.563\times10^{-1}$
      & 0.95 & ok\\
R003a &  $1.0\times 10^{-3}$ & $5.7\times 10^{-5}$ & $3.590\times10^{-2}$
      & 0.97 & ok\\
\hline
R003d & $1.0\times 10^{-1}$& & & & unstable\\
R003e & $1.0\times 10^{-1}$& & & & unstable\\
R003f & $1.0\times 10^{-1}$& & & & unstable\\
\hline
R004a &  $1.0\times 10^{-1}$ & $5.475\times 10^{-2}$ & $9.476\times10^{-1}$
      & 0.87 & ok\\
R004a &  $5.0\times 10^{-2}$ & $1.960\times 10^{-2}$ & $7.935\times10^{-1}$
      & 0.89 & ok\\
R004a &  $2.5\times 10^{-2}$ & $8.113\times 10^{-3}$ & $5.029\times10^{-1}$
      & 0.93 & ok\\
R004a &  $1.0\times 10^{-2}$ & $3.043\times 10^{-3}$ & $2.160\times10^{-1}$
      & 0.95 & ok\\
R004a &  $1.0\times 10^{-3}$ & $2.999\times 10^{-4}$ & $2.191\times10^{-2}$
      & 0.97 & ok\\
\hline
R004d &  $1.0\times 10^{-1}$ & $4.365\times 10^{-2}$ & $9.995\times10^{-1}$
      & 0.82 & ok\\
R004d &  $5.0\times 10^{-2}$ & $9.260\times 10^{-3}$ & $9.829\times10^{-1}$
      & 0.84 & ok\\
R004d &  $2.5\times 10^{-2}$ & $3.896\times 10^{-3}$ & $8.405\times10^{-1}$
      & 0.91 & ok\\
R004d &  $1.0\times 10^{-2}$ & $2.198\times 10^{-3}$ & $2.854\times10^{-1}$
      & 0.95 & ok\\
R004d &  $1.0\times 10^{-3}$ & $2.927\times 10^{-4}$ & $2.246\times10^{-2}$
      & 0.97 & ok\\
\hline
R004f &  $1.0\times 10^{-1}$ &  &    &  & unstable\\
R004f &  $5.0\times 10^{-2}$ & $9.183\times 10^{-3}$ & $1.000\times10^{0}$
      & 0.80 & ok\\
R004f &  $2.5\times 10^{-2}$ & &   &  & no equilibrium\\
R004f &  $1.0\times 10^{-2}$ & $1.045\times 10^{-3}$ & $3.965\times10^{-1}$
      & 0.94 & ok\\
\hline
\end{tabular}
\label{table:xi226}
\caption{Equilibrium results for $\xi_0 = 2.26$, $\epsilon = \kappa/4$.}
\end{center}
\end{table}

\subsubsection{Dynamics}

Solve the Ward-Tordai equation.

\vfill
\pagebreak

### 2009-05-07 trunk/htdocs/sections.tex
Updates to references, tables
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@528 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.5 2008-09-20 10:39:58 kevin Exp $
+%  $Id: sections.tex,v 1.10 2009-05-07 15:29:05 kevin Exp $
 %
@@
 
-The parameter $\kappa$
-
-\inputkey{C}
-
-The surface free energy parameter {C}. Currently unused.
-
-\inputkey{H}
-
-The surface free energy parameter {H}. Currently unused.
+The parameter $\kappa$, which is positive.
+
+\subsubsection{Brazovskii free energy}
+
+This is similar to the symetric, but with one extra term which
+requires description.
+\begin{equation}
+ F[\phi] = 
+\int dr \left(
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
++ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
+\end{equation}
+
+\subsubsection{Surfactant model free energy}
+
+\inputkey{free\_energy surfactant}
+
+This is for use with a two order parameter model the first of
+which is the composition, as for the symmetric free energy,
+while the second $\psi$ represents surfactant concentration.
+The free energy density is made up of a number of terms:
+
+\begin{equation}
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+\end{equation}
+is the standard symmetric term related to the composition as above;
+
+\begin{equation}
+D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
+\end{equation}
+represents the energy of surfactant in the bluk phase with surfactant
+concentration $0 < \psi < 1$ with a single parameter $D$;
+\begin{equation}
+-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
+-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
+\end{equation}
+where the term in $\epsilon$ represents the energy reduction by
+adsorbing surfactant at the interface and the term in $\beta$
+provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
+\begin{equation}
++{\textstyle \frac{1}{2}} W \psi \phi^2
+\end{equation}
+which is included to stabilise the interface \cite{theissengompper}.
+
+The corresponding keys for the input file are:
+
+\inputkey{D}
+
+is the bulk surfactant contribution parameter
+
+\inputkey{epsilon}
+
+is the parameter $\epsilon$
+
+\inputkey{W}
+
+is the parameter $W$
+
+\inputkey{beta}
+
+is the value of the parameter $\beta$ (not implemented yet).
+
+\inputkey{phi\_b}
+
+the initial, uniform, background surfactant concentration $\psi_b$.
+
+\subsection{Other Binary Fluid Parameters}
 
@@
 \mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
+\label{eq:fb}
 \end{equation}
@@
 
+\section{Surfactants}
+
+This section describes the approach taken to study surfactants. Here,
+we always adopt the hydrid method, i.e., LB for the hydrodynmic
+equations and finite difference for the associated Cahn-Hilliard
+equations. Coupling between the two sectors is via the velocity
+field and the external force on the fluid computed via the
+divergence of the `chemical stress'.
+
+In this model the surfactant appears as a concentration field
+$\psi(\mathbf{r};t)$, alongside the compositional order parameter
+$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
+two corresponding Cahn-Hilliard equations: first,
+
+\begin{equation}
+\partial_t \phi + \partial_\alpha \phi u_\alpha = 
+\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
+\end{equation}
+and second
+\begin{equation}
+\partial_t \psi + \partial_\alpha \psi u_\alpha = 
+\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
+\end{equation}
+
+\subsection{Free Energy}
+
+A number of different free energies have been used to study
+surfactants [cite Yeomans]. Here, the free energy is based on
+that used by van der Graff and van der Sman and others [cite].
+There are a number of different contributions to the free
+energy density:
+\begin{equation}
+f = f_\phi + f_\psi +f_1 + f_{ex}.
+\end{equation}
+The first is exactly that used for the binary fluid, i.e.,
+\begin{equation}
+f_\phi (\phi, \partial_\alpha\phi) =
+{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
++ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
+\end{equation}
+The surface tension and the interfacial width are determined from the
+choice of parameters $A,B,\kappa$ as before.
+The second term in the free energy density
+represents the entropic
+energy of mixing of the surfactant with the bulk phases
+\begin{equation}
+f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
+\end{equation}
+where the parameter $kT$ sets the energy scale for the whole system.
+The surfactant order parameter, $\psi$, is unity for a saturated
+interface and zero for a plain interface.
+
+The surfactant lowers the energy of the interface where it is adsorbed
+there, represented by the term
+\begin{equation}
+f_1(\psi, \partial_\alpha \phi) 
+= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
+  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
+\end{equation}
+where $\epsilon$ and $\beta$ are parameters. Note that only the
+first term appears in \cite{vandergraaf}, while the second term is
+based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
+the surfactant profile at the interface is symmetric (so-called
+Langmuir adsoption) whereas the term in $\beta$ induces an
+asymmetry (Frumpkin adsorption).
+Finally,  $f_{ex}$ is [van der Sman claims]
+an enthalpic contribution introduced 
+to stabilise the interface \cite{theissengompper}
+\begin{equation}
+f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
+\end{equation}
+with $W$ a parameter.
+
+\subsection{Chemical potentials}
+
+The respective chemical potentials are obtained from the functional
+derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
+$\mu_\psi = \delta F / \delta \psi$. We obtain:
+\begin{equation}
+\mu_\phi =
+A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
++ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
++ \epsilon \partial_\alpha \phi \partial_\alpha \psi
++ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
++ \beta \psi^2 \partial_\alpha^2 \phi,
+\end{equation}
+and
+\begin{equation}
+\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
++ {\textstyle \frac{1}{2}}W\phi^2
+- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
+- \beta\psi (\partial_\alpha \phi)^2.
+\end{equation}
+As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
++ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
+\cite{theissengompper})
+\begin{equation}
+p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
+\end{equation}
+There is no closed form for the tensor part, which must be chosen to
+comply with the constraint that in mechanical equilibrium
+\begin{equation}
+\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
+\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
+\end{equation}
+In this way we obtain for the isotropic contribution
+\begin{eqnarray}
+p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
+- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
+- \kappa \phi \partial_\alpha^2 \phi
+- kT \ln(1-\psi) + W\psi\phi^2
+\nonumber\\
++ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
++ \epsilon\phi\psi \partial_\alpha^2 \phi
++ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
++ \beta \phi \psi^2 \partial_\alpha^2 \phi
+- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
+\end{eqnarray}
+and for the tensor part we can have
+\begin{equation}
+\tilde{P}_{\alpha\beta} =
+(\kappa - \epsilon \psi - \beta\psi^2) 
+\partial_\alpha \phi \partial_\beta \phi.
+\end{equation}
+
 \section{Code Design}
@@
 
+\bibitem{chang}
+C.-H. Chang and E.I. Franses,
+Adsorption dynamics of surfactants at the air/water interface:
+a critical review of mathematical models, data, and mechanisms,
+\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
+
 \bibitem{cpc}
@@
 \textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
+
+\bibitem{diamant}
+H. Diamant and D. Andelman,
+Kinetics of surfactant adsorption at fluid/fluid
+interfaces: non-ionic surfactants,
+\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
+
+\bibitem{diamant96}
+H. Diamant and D. Andelman,
+Kinetics of surfactant adsorption at fluid-fluid interfaces,
+\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
+
+\bibitem{eastoe}
+J. Eastoe and J.S. Dalton,
+Dynamic surface tension and adsorption mechanims of surfactants
+at the air-water interface,
+\textit{Advances in Colloid and Interface Science}, \textbf{85}
+13, 2000.
 
@@
 
+\bibitem{vandergraaf}
+R.G.M. van der Sman and S. van der Graaf,
+Diffuse interface model of surfactant adsorption onto flat and
+droplet interfaces,
+\textit{Rheol. Acta} \textbf{46} 3 (2006).
+
+\bibitem{theissengompper}
+O. Theissen and G. Gompper,
+Lattice Boltzmann study of spontaneous emulsification,
+\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
+
+\bibitem{wardtordai}
+A.F.H. Ward and L. Tordai,
+\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
+
 \end{thebibliography}

### 2009-05-07 trunk/htdocs/surfactant.tex
Updates to references, tables
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@528 40876441-3496-0410-a613-a2fca35b2ae2

@@
 To this we add a separate order parameter to represent the
-surfactant $\psi$ \cite{smangraaf}. The free energy density
+surfactant $\psi$ \cite{vandergraaf}. The free energy density
 is made up of two parts following DA96:
@@
 R008d &  &  &  & & & 0.0025 &  &  &  \\
-R008f &  &  &  & & & 0.0050 &  &  &  \\
 \hline
@@
 R006d & &  &  &  & & 0.0025 &  &  &  \\
-R006f & &  &  &  & & 0.0050 &  &  &  \\
+\hline
+R010a & 0.025 & 0.00625 & 0.004 & 1.6965$\times 10^{-4}$ & 0.001
+      & 0.0000 & 0.0 & 2.0 & 2.0 \\
+R010d & &  &  &  & & 0.0005 &  &  &  \\
+\hline
+R012a & 0.025 & 0.00625 & 0.004 & 8.48231$\times 10^{-5}$ & 0.001
+      & 0.0000 & 0.0 & 2.0 & 2.0 \\
+R012d & &  &  &  & & 0.0005 &  &  &  \\
 \hline
 \end{tabular}
-\caption{Parameters}
+\caption{Parameters for $\xi_0 = 1.13$ and $\epsilon = \kappa/4$.}
 \end{center}
@@
 \hline
+R010a & $2.0\times 10^{-1}$ & $1.859\times 10^{-1}$ & $5.143\times10^{-1}$
+      & 0.76 & ok \\
+R010a & $1.0\times 10^{-1}$ & $9.093\times 10^{-2}$ & $3.033\times10^{-1}$
+      & 0.80 &  ok\\
+R010a & $5.0\times 10^{-2}$ & $4.490\times 10^{-2}$ & $1.643\times10^{-1}$
+      & 0.82 & ok \\
+R010a & $2.5\times 10^{-2}$ & $2.231\times 10^{-2}$ & $8.540\times10^{-2}$
+      & 0.84 & ok \\
+R010a & $1.0\times 10^{-2}$ & $8.888\times 10^{-3}$ & $3.494\times10^{-2}$
+      & 0.86 & ok \\
+R010a & $1.0\times 10^{-3}$ & $8.867\times 10^{-4}$ & $3.540\times10^{-3}$
+      & 0.87 & ok \\
+\hline
+R010d & $2.0\times 10^{-1}$ & $1.715\times 10^{-1}$ & $8.685\times10^{-1}$
+      & 0.62 & ok \\
+R010d & $1.0\times 10^{-1}$ & $8.327\times 10^{-2}$ & $4.893\times10^{-1}$
+      & 0.73 & ok \\
+R010d & $5.0\times 10^{-2}$ & $4.317\times 10^{-2}$ & $2.061\times10^{-1}$
+      & 0.81 & ok \\
+R010d & $2.5\times 10^{-2}$ & $2.190\times 10^{-2}$ & $9.524\times10^{-2}$
+      & 0.84 & ok \\
+R010d & $1.0\times 10^{-2}$ & $8.824\times 10^{-3}$ & $3.646\times10^{-2}$
+      & 0.86 & ok \\
+R010d & $1.0\times 10^{-3}$ & $8.861\times 10^{-4}$ & $3.555\times10^{-3}$
+      & 0.87 & ok \\
+\hline
+R012a & $2.0\times 10^{-1}$ & $1.72\times 10^{-1}$ & $8.51\times10^{-1}$
+      & 0.71 &  not quite at eqbm.\\
+R012a & $1.0\times 10^{-1}$ & $7.5\times 10^{-2}$ & $6.63\times10^{-1}$
+      & 0.74 &  not quite at eqbm.\\
+R012a & $5.0\times 10^{-2}$ & $3.4\times 10^{-2}$ & $4.15\times10^{-1}$
+      & 0.78 &  not quite\\
+R012a & $2.5\times 10^{-2}$ & $1.6\times 10^{-2}$ & $2.26\times10^{-1}$
+      & 0.82 &  not quite\\
+R012a & $1.0\times 10^{-2}$ & $6.4\times 10^{-3}$ & $9.4\times10^{-2}$
+      & 0.85 & not quite \\
+R012a & $1.0\times 10^{-3}$ & $6.3\times 10^{-4}$ & $9.6\times10^{-3}$
+      & 0.87 &  not quite\\
+\hline
 \end{tabular}
 \label{table:xi113}
-\caption{Equilibrium results for $\xi_0 = 1.13$, $\epsilon = \kappa/4$.}
+\caption{Equilibrium results for $\xi_0 = 1.13$, $\epsilon = \kappa/4$
+at 2 million time steps.
+Sets R006 and R008 have viscosity $\eta = 1/6$, while R010 and R012
+have $\eta = 0.025$, showing very little sensitivity to viscosity.}
 \end{center}
@@
 
+\subsubsection{Parameters with $\xi_0 = 3.39$, $\epsilon = kappa/4$}
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline
+Name & $\eta$ & $-A,B$ & $\kappa$ & $kT$ & $\epsilon$ & $\beta$ & $W$
+     & $M$ & $M_\psi$\\
+\hline
+R014a & 1/6 & 0.020833 & 0.12 & 5.6587$\times 10^{-4}$ & 0.03 & 0.0000 & 0.0 
+      & 0.15 & 2.0 \\
+R014d &  &  &  & & & 0.0075 &  &  &  \\
+\hline
+R016a & 1/6 & 0.020833 & 0.12 & $2.82743\times 10^{-4}$ & 0.03 & 0.0000 & 0.0 
+      & 0.15 & 2.0 \\
+R016d & &  &  &  & & 0.0075 &  &  &  \\
+\hline
+R018a & 0.025 & 0.0020833 & 0.012 & 5.6587$\times 10^{-5}$ & 0.003
+      & 0.0000 & 0.0 & 2.0 & 2.0 \\
+R018d & &  &  &  & & 0.00075 &  &  &  \\
+\hline
+R020a & 0.025 & 0.0020833 & 0.012 & 2.82743$\times 10^{-5}$ & 0.003
+      & 0.0000 & 0.0 & 2.0 & 2.0 \\
+R020d & &  &  &  & & 0.0075 &  &  &  \\
+\hline
+\end{tabular}
+\caption{Parameters for $\xi_0 = 3.39$ and $\epsilon = \kappa/4$.}
+\end{center}
+\end{table}
+
 \subsubsection{Dynamics}

### 2009-05-12 trunk/htdocs/surfactant.tex
Added surfactant figure
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@543 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+In the following we retain $\xi_0 = 2.26$ but keep the same surface
+tension cf. $\xi_0 = 1.13$, that is 0.047.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline
+Name & $\eta$ & $-A,B$ & $\kappa$ & $kT$ & $\epsilon$ & $\beta$ & $W$
+     & $M$ & $M_\psi$\\
+\hline
+R022a & 1/6 & 0.03125 & 0.08 & 8.48231$\times 10^{-4}$ & 0.02 & 0.000 & 0.0 
+      & 0.15 & 2.0 \\
+R022d &  &  &  & & & 0.005 &  &  &  \\
+\hline
+R024a & 1/6 & 0.03125 & 0.08 & 4.24115$\times 10^{-4}$ & 0.02 & 0.00 & 0.0 
+      & 0.15 & 2.0 \\
+R024d & &  &  &  & & 0.005 &  &  &  \\
+\hline
+R026a & 0.025 & 0.003125 & 0.008 & 8.48231$\times 10^{-4}$ & 0.002 & 0.00
+      & 0.0 & 2.0 & 2.0 \\
+R026d & &  &  &  & & 0.0005 &  &  &  \\
+\hline
+R028a & 0.025 & 0.003125 & 0.008 & 4.24115$\times 10^{-4}$ & 0.002 & 0.00
+      & 0.0 & 2.0 & 2.0 \\
+R028d & &  &  &  & & 0.0005 &  &  &  \\
+\hline
+\end{tabular}
+\caption{Equilibrium results for $\xi_0 = 2.26$, $\epsilon = \kappa/4$, and
+$\sigma_0 = 0.047$.}
+\end{center}
+\end{table}
+
 \begin{table}
@@
 have $\eta = 0.025$, showing very little sensitivity to viscosity.}
+\end{center}
+\end{table}
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+Name & $\bar{\psi}$ & $\psi_b$ & $\psi_{0\mathrm{eq}}$
+     & $\sigma / \sigma_0$  & Comment \\
+\hline
+R022a &  $2.0\times 10^{-1}$ & $1.701\times 10^{-1}$ & $7.423\times10^{-1}$
+      & 0.88 & ok\\
+R022a &  $1.0\times 10^{-1}$ & $8.008\times 10^{-2}$ & $5.023\times10^{-1}$
+      & 0.90  & ok\\
+R022a &  $5.0\times 10^{-2}$ & $3.871\times 10^{-2}$ & $2.886\times10^{-1}$
+      & 0.93 & ok\\
+R022a &  $2.5\times 10^{-1}$ & $1.904\times 10^{-2}$ & $1.530\times10^{-1}$
+      & 0.95 & ok\\
+R022a &  $1.0\times 10^{-2}$ & $7.546\times 10^{-3}$ & $6.305\times10^{-2}$
+      & 0.96 & ok \\
+R022a &  $1.0\times 10^{-3}$ & $7.506\times 10^{-4}$ & $6.407\times10^{-3}$
+      & 0.97 & ok\\
+\hline
+R024a &  $2.0\times 10^{-1}$ & $1.471\times 10^{-1}$ & $9.854\times10^{-1}$
+      & 0.84 & ok\\
+R024a &  $1.0\times 10^{-1}$ & $5.475\times 10^{-2}$ & $9.476\times10^{-1}$
+      & 0.87 & ok\\
+R024a &  $5.0\times 10^{-2}$ & $1.960\times 10^{-2}$ & $7.935\times10^{-1}$
+      & 0.89 & ok\\
+R024a &  $2.5\times 10^{-2}$ & $8.113\times 10^{-3}$ & $5.029\times10^{-1}$
+      &  0.93 & ok\\
+R024a &  $1.0\times 10^{-2}$ & $3.042\times 10^{-3}$ & $2.160\times10^{-1}$
+      &  0.95 & ok\\
+R024a &  $1.0\times 10^{-3}$ & $2.998\times 10^{-4}$ & $2.191\times10^{-2}$
+      &  0.97 & ok\\
+\hline
+\end{tabular}
+\label{table:xi226a}
+\caption{Equilibrium results for $\xi_0 = 2.26$, $\epsilon = \kappa/4$
+and $\sigma_0 = 0.047$ at 2 million time steps.}
 \end{center}
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|}
+\hline
+Name & $\bar{\psi}$ & $\psi_b$ & $\psi_{0\mathrm{eq}}$
+     & $\sigma / \sigma_0$  & Comment \\
+\hline
+R014a &  $2.0\times 10^{-1}$ & $1.551\times 10^{-1}$ & $7.718\times10^{-1}$
+      & 0.90 & ok\\
+R014a &  $1.0\times 10^{-1}$ & $7.072\times 10^{-2}$ & $5.193\times10^{-1}$
+      & 0.92 & ok\\
+R014a &  $5.0\times 10^{-2}$ & $3.384\times 10^{-2}$ & $2.889\times10^{-1}$
+      & 0.94 & ok\\
+R014a &  $2.5\times 10^{-2}$ & $1.661\times 10^{-2}$ & $1.494\times10^{-1}$
+      & 0.96 & ok\\
+R014a &  $1.0\times 10^{-2}$ & $6.578\times 10^{-3}$ & $6.060\times10^{-2}$
+      & 0.98 & ok\\
+R014a &  $1.0\times 10^{-3}$ & $6.542\times 10^{-4}$ & $6.098\times10^{-3}$
+      & 0.99 & ok\\
+\hline
+R016a &  $2.0\times 10^{-1}$ & $1.231\times 10^{-1}$ & $9.897\times10^{-1}$
+      & 0.88 & ok\\
+R016a &  $1.0\times 10^{-1}$ & $3.779\times 10^{-2}$ & $9.652\times10^{-1}$
+      & 0.89 & ok\\
+R016a &  $5.0\times 10^{-2}$ & $1.101\times 10^{-2}$ & $7.988\times10^{-1}$
+      & 0.92 & ok\\
+R016a &  $2.5\times 10^{-2}$ & $4.87\times 10^{-3}$ & $4.57\times10^{-1}$
+      & 0.95 & ok\\
+R016a &  $1.0\times 10^{-2}$ & $1.973\times 10^{-3}$ & $1.813\times10^{-1}$
+      & 0.97 & ok\\
+R016a &  $1.0\times 10^{-3}$ & $2.016\times 10^{-4}$ & $1.77\times10^{-2}$
+      & 0.99 & ok\\
+\hline
+\end{tabular}
+\label{table:xi226}
+\caption{Equilibrium results for $\xi_0 = 3.39$, $\epsilon = \kappa/4$.}
+\end{center}
+\end{table}
+
+\begin{figure}
+\begin{center}
+\includegraphics{surfactant_fig1.eps}
+\end{center}
+\caption{Equilibrium results for one-dimensional simulations with
+varying interfacial width $\xi_0$, all other physical parameters
+being the same. In particular, $\epsilon = \kappa/4$ and $\beta = 0$.
+The
+left-hand column shows equilibrium isotherms with $1/K_L$ = 0.1 (red)
+and $1/K_L$ = 0.01 (green). The circles are simulations and the lines
+the theoretical prediction. The right-hand column shows the
+corresponding reduction in surface tension at equilibrium
+as a function of $\psi_{0,\mathrm{eq}}$. This reduction is scaled by
+the bare interface surface tension $\sigma_0$. The exact result here
+is based on a direct numerical integration of the excess free energy
+across an equilibrium interfacial profile. The points are computed
+by carrying out an analogous numerical integration of the simulated
+profile at the given resolution. Note that the representation of
+the surfactant porfile is particularly poor at $\xi_0 = 1.13$ as
+there turn out to be an even number of lattice points where
+$\psi >> \psi_b$ so that the peak value is missed.}
+\end{figure}
+
 \subsubsection{Dynamics}

### 2009-05-20 htdocs/sections.tex
Added section of matching units in porous media
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@560 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.9 2009-02-23 18:54:09 kevin Exp $
+%  $Id: sections.tex,v 1.11 2009-05-20 17:02:38 kevin Exp $
 %
@@
 \end{equation}
-which is included to stabilise the interface \cite{theissen}.
+which is included to stabilise the interface \cite{theissengompper}.
 
@@
 the initial, uniform, background surfactant concentration $\psi_b$.
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
-\hline
-Name & $\eta$ & $-A,B$ & $\kappa$ & $D$ & $\epsilon$ & $W$
-     & $M$ & $M_\psi$ \\
-\hline
-Run001 & 0.025 & 0.000625 & 0.004 & 0.0001 & 0.004 & 0.0003  & 2.0 & 0.1\\
-\hline
-\end{tabular}
-\caption{Acceptable parameters for the surfactant free energy model.
-Note that $\epsilon = \kappa$.}
-\end{center}
-\end{table}
 
@@
 $J = C\rho g / \eta$. The calculation procedes as above.
+
+\subsubsection{Matching lattice and real units in porous media}
+
+Following Succi \cite{succi} (Chapter~8) the following argument
+can be made to match lattice quantities and real physical
+quantities for a given system of interest. Suppose we have
+a structure with characteristic pore size $h = 1$ micron
+and characteristic sample size of 1 cm. If the fluid
+(e.g., water) has viscosity $\eta = 10^{-3}$ Pa s and density
+$\rho = 10^3$ kg~s$^{-1}$, and a typical flow is
+$U \sim 10^{-3}$ m~s$^{-1}$, then the pore scale Reynolds
+number is $Re_h \sim 10^{-3}$.
+
+If the speed of sound in water is 1500~m~s$^{-1}$,
+then the corresponding time scale related to the
+speed of sound and the pore size is $T = h/ c_s$,
+which is $\sim 10^{-10}-10^{-9}$~s. (Exact matching
+is acheived via the dimensionless group $\Delta x / c_s\Delta t$,
+with the lattice speed of sound $1/\sqrt{3}$).
+
+This, effectively very small, lattice
+time step is clearly a problem in reaching any respectable 
+macroscopic time in a simulation.
+One step that can be taken is to increase both the lattice
+viscosity and the lattice velocity so that the Reynolds number
+remains unchanged. (Note that the lattice viscosity matching
+the real viscoity might be $(\eta/\rho)T/h^2$ which, using the
+above figures, is $\sim 10^{-4}$.) This can gain 2--3 orders
+of magnitude while keeping the viscosity below $O(1)$. Ultimately,
+it may also be necessary to raise the Reynolds number, which may
+be considered small compared to unity as long as it remains
+in practice less than $O(10^{-1})$ \cite{cates_scaling}.
 
@@
 
+\bibitem{cates_scaling}
+M. E. Cates, J.-C. Desplat, P. Stansell, A.J. Wagner, K. Stratford,
+R. Adhikari, and I. Pagonabarraga,
+Physical and Computational Scaling Issues in Lattice Boltzmann
+Simulations of Binary Fluid Mixtures,
+\textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
+
 \bibitem{chunladd}
@@
 
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
 \bibitem{theissengompper}
-O. theissen and G. Gompper,
+O. Theissen and G. Gompper,
 Lattice Boltzmann study of spontaneous emulsification,
@@
 
+\bibitem{wardtordai}
+A.F.H. Ward and L. Tordai,
+\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
+
 \end{thebibliography}

### 2010-01-21 trunk/htdocs/sections.tex
Some stuff on Lees Edwards
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@784 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.10 2009-05-07 15:29:05 kevin Exp $
+%  $Id: sections.tex,v 1.12 2010-01-21 18:20:51 kevin Exp $
 %
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% LE Implementation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Lees Edwards Sliding Period Boundary Conditions}
+
+\subsubsection{The LE module}
+
+\subsubsection{Distributions crossing the LE planes}
+
+With the planes conceptually half-way between lattice sites, the
+propagation moves some of the distributions (namely, those with
+$c_x = \pm1$) between different sliding blocks. While the collision
+stage is unaffected, action must be taken to adjust these
+distributions when the LE boundary conditions are active. This
+is done in a two-stage process described by Adhikari \textit{et al.}
+\cite{adhikari_desplat}, implemented between the collision and
+propagation.
+
+\textbf{a. Reprojection} The post-collision distributions with
+$c_x = \pm 1$ at sites adjacent to a boundary are modified to
+take account of the velocity jump $\pm u^{LE}_y$ between the sliding
+blocks. In terms of the hydrodynamic moments we have:
+\begin{eqnarray}
+\rho &\rightarrow& \rho', \\
+\rho u_\alpha &\rightarrow& (\rho u_\alpha)' \pm \rho' u^{LE}_\alpha, \\
+S_{\alpha\beta} &\rightarrow&
+S'_{\alpha\beta} \pm (\rho u_\alpha)' u^{LE}_\beta
+\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
+\end{eqnarray}
+If we work with the changes to the moemnts, then the density is
+unaffected $\delta\rho =  0$, the velocity is changed by
+$\delta u_\alpha = \pm u^{LE}_\alpha$, with an analogous expression for
+the change in the stress
+\begin{equation}
+\delta S_{\alpha\beta} =
+\pm (\rho u_\alpha)' u^{LE}_\beta
+\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
+\end{equation}
+We can then work out the change in the distributions by reprojecting
+via Eq.\ref{eq:}, i.e.,
+\begin{equation}
+f_i \rightarrow f_i' + w_i \Bigg(
+\frac{\rho \delta u_\alpha c_{i\alpha}}{c_s^2} +
+\frac{\delta S_{\alpha\beta}Q_{i\alpha\beta}}{2c_s^4} \Bigg).
+\end{equation}
+A similar set of expressions are given for the order parameter
+distributions in Adhikari \textit{et al.} \cite{adhikari_desplat}.
+
+\textbf{b. Interpolation of reprojected distributions}
+As the sliding blocks are displaced continuously with
+$\delta y = u_y^{LE}t$, which is not in general a whole number of
+lattice spacings, an interpolation is also required before
+propagation can take place. Consider the situation (Fig.~\ref{fig_le2}) from
+a stationary frame where the frame above has translated a small
+distance to the right ($\delta y < \Delta y$). In the stationary frame
+we must interpolate the ditributions with $c_x = 1$ to a
+`departure point' from which the propagation will move the
+distribution exactly to the appropriate lattice site in the moving
+frame. Schemeatically,
+\begin{equation}
+f_i'(x, y + \delta y, z) = 
+ (1 - \delta y) f_i^\star(x, y, z) +
+\delta y f_i^\star(x, y + \Delta y, z),
+\end{equation}
+where $f_i'$ here represents the interpolated value and $f_i^\star$
+is the reprojected post-collision quantity.
+
+In the case where the displacement $\delta y > 1$, the relevant
+lattice sites involved in the interpolation are displaced to the right
+by the integral part of $\delta y$, while the relative weight given to
+the two sites involves
+the fractional part of $\delta y$. To take account of the periodic
+boundary conditions
+in the $y$-direction, all displacements are modulo $L_y$.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics{xfig/fig_le2.eps}
+\end{center}
+\caption{Schematic diagram showing interpolation of distributions in the
+stationary plane (below) in preparation for propagation into the moving
+frame (above). For a small displacement $\delta y = u^{LE} t$, the
+interpolation takes place between lattice sites in the stationary frame
+(green, left). Propogation from the interpolated positions (red) can
+then take place into the moving frame (right).}
+\label{fig_le2}
+\end{figure}
+
+In parallel, exactly the same calculation is undertaken. However,
+communication is required to take account of the processor
+decomposition in the $y$-direction. This requires the identification
+of the two processes which hold the relevant information as a function
+of displacement.
+
+Note that the interpolated values are stored in the appropriate lattice
+site in the rest frame that ensures the normal propagation stage moves all
+distributions to the correct destination in memory.
+
+\subsubsection{Order parameter gradients}
+
+When the order parameter gradient is computed at a given point
+near the LE boundaries, the stencil may extend out of the stationary
+frame. The process of interpolation is slightly different to that
+for the distributions.
+
+Here, we need to interpolate values in the moving frame to the position
+which is seen by the rest frame. Figure~\ref{fig_le3} shows a five-point
+stencil to calculate the gradient at the central point $(x,y,z)$ in the
+rest frame. An interpolation of the $\phi$-field of the form
+\begin{equation}
+\phi'(x + \Delta x, y - \delta y, z) =
+\delta y \phi(x + \Delta x,  y - \Delta y, z) + 
+(1 - \delta y) \phi(x + \Delta x, y, z)
+\end{equation}
+must then take place. In practice, the interpolated values are stored
+in the offset buffer. As the gradient calculation requires $\phi$
+values in the halo regions, interpolated values for the buffer halo
+region are also required.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics{xfig/fig_le3.eps}
+\end{center}
+\caption{Schematic diagram showing a five-point stencil to compute a
+gradient at the central point in the stationary frame. The upper point
+of the stencil is in the moving frame and so must be interpolated to
+the correct position (red sites).}
+\label{fig_le3}
+\end{figure}
+
+In parallel, the interpolation requires data from at most two adjacent
+processes in the along-plane $y-$direction as long as the $\phi$ halo
+regions are up-to-date. The rank of the processes involved is
+determined by the displacement $\delta y$ as a function of time.
+
+\subsubsection{Velocities and finite-difference fluxes}
+
+For the finite-difference implementation, the velocity field at
+the cell boundaries is required. At the planes, this means an
+interpolation which takes the same form as that for
+$\phi$ is needed.
+
+In computing the advective fluxes between cells, we must then
+allow for the presence of the planes. This is done by computing
+separately the 'east' and 'west' face fluxes in the $x$-direction
+(the flux that crosses the plane). Away from the planes, face-flux
+uniqueness $f_{ijk}^e = f_{i+1jk}^w$ ensures conservation of order
+parameter. However, at the planes, the non-linear conbination of
+velocity field and order parameter field does not in general give
+rise to consistent fluxes. In order to restore conservation, we
+reconcile east and west face fluxes at the plane by taking an
+average
+\begin{eqnarray}
+f_{ijk}^{e\star}&=&{\textstyle\frac{1}{2}}(f_{ijk}^e + f_{i+1j'k}^w),\quad 
+f_{i+1j'k}^w  = \delta y f_{i+1 j-1 k}^w + (1 - \delta y) f_{i+1jk}^w,\\
+f_{i+1 j k}^{w\star} &=& {\textstyle\frac{1}{2}}(f_{ij'k}^e + f_{i+1jk}^w),\quad
+f_{ij'k}^e = (1 - \delta y) f_{ijk}^e + \delta y f_{ij+1k}^e.
+\end{eqnarray}
+In this way, we average the east face flux in the rest frame $f_{ijk}^e$
+with the interpolated value of the west face flux in the moving frame
+$f_{i+1j'k}^w$, and vice-versa. One can then easily show that when
+integrated over the length of the plane, the average fluxes are
+consistent, i.e.,
+\begin{equation}
+\sum_j (f_{ijk}^{e\star} - f_{i+1jk}^{w\star}) = 0 \quad\forall k,
+\end{equation}
+thus ensuring order parameter conservation.
+
+A similar correction is required when computing the force on the fluid
+directly via the divergence of the thermodynamic stress
+$F_\alpha = \nabla_\beta P_{\alpha\beta}^{th}$. This is implementated
+by computing fluxes of momentum at cell faces in each direction, and
+then taking the divergence to compute the force. At the planes, an
+averaging procedure is again used to ensure conservation of momentum.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% End LE implementation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \vfill
@@
 
+\bibitem{adhikari_desplat}
+Adhikari, R. J.-C. Desplat, and K. Stratford,
+Sliding periodic boundary conditions for lattice Boltzmann and lattice
+kinetic equations,
+\texttt{arXiv:cond-mat/0503175v1} (2005).
+
 \bibitem{adhikari2005}
@@
 \textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
+
+\bibitem{cates_scaling}
+M. E. Cates, J.-C. Desplat, P. Stansell, A.J. Wagner, K. Stratford,
+R. Adhikari, and I. Pagonabarraga,
+Physical and Computational Scaling Issues in Lattice Boltzmann
+Simulations of Binary Fluid Mixtures,
+\textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
 

### 2010-02-13 htdocs/sections.tex
Added D2Q9 table.
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@803 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.11 2009-05-20 17:02:38 kevin Exp $
+%  $Id: sections.tex,v 1.13 2010-02-13 15:32:51 kevin Exp $
 %
@@
 
+\subsection{D2Q9}
+
+These are the eigenvalues and eigenvectors of the collision
+matrix used for D2Q9
+
+\begin{table}
+\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
+\hline\hline
+$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
+\hline
+$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
+\hline
+$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
+\hline
+$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
+\hline
+$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
+& $c_{ix} c_{ix} - c_s^2$ \\
+\hline
+$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
+\hline
+$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
+& $c_{iy} c_{iy} - c_s^2$ \\
+\hline\hline
+$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
+\hline
+$J_{ix}$ & - & 0 &  4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
+& $\chi^1 \rho c_{ix}$\\
+\hline
+$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
+& $\chi^1 \rho c_{iy}$\\
+\hline\hline
+$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
+\hline\hline
+\end{tabular}
+\end{table}
+
 \subsection{D3Q15}

### 2010-04-30 trunk/htdocs/sections.tex
Added notes on polar active gel from Davide
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@993 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.12 2010-01-21 18:20:51 kevin Exp $
+%  $Id: sections.tex,v 1.14 2010-04-30 11:00:50 kevin Exp $
 %
@@
 
+\section{Polar Active Gel}
+
+Here we consider a polar vector field model to study active gels. 
+The order parameter is this time a vector $P_\alpha$
+(with variable magnitude) as there is no longer head-tail symmetry
+in the system as was the case for the ${\bf Q}$ tensor formulation. 
+The equation of motion as:
+\begin{equation}
+\partial_t P_\alpha  \partial_\beta (u_{\beta} + wP_{\beta}) P_{\alpha}
+= \lambda D_{\alpha\beta} P_{\beta} - \omega_{\alpha\beta} P_{\beta}
++\Gamma' h_{\alpha}.
+\end{equation}
+
+In this equation, $w$ is an active term, due to swimming, which causes
+self-advection of the order parameter, while $\lambda$ is a material
+dependent constant -- positive for rod-like molecules. If $|\lambda|>1$
+the liquid crystalline passive phase is flow-aligning, otherwise it is
+flow-tumbling.
+
+Note that the convection for the Leslie-Ericksen is that the
+velocity gradient tensor should be writen $W_{\alpha\beta}
+= \partial_\alpha u_\beta$ (and not the other way round, which
+is usual).
+
+The molecular field is given by
+$h_{\alpha}=-\delta {\cal F}_{\rm pol}/\delta p_{\alpha}$
+where ${\cal F}_{\rm pol}$ is the free energy for a polar active nematic,
+whose density we can take to be:
+\begin{equation}
+f =  \frac{\alpha_2}{2} |{\mathbf P}|^2 + \frac{\alpha_4}{4}|{\mathbf P}|^4
++ \frac{K}{2}\left(\partial_\alpha P_{\beta}\right)^2 + 
+\frac{K_{LC}}{2}\left[\partial_\alpha (P_{\beta}P_{\gamma})\right]^2 
+\end{equation}
+where $K$ and $K_{LC}$ are elastic constants, $\alpha_4>0$, and $\alpha_2$ 
+can be positive (isotropic phase) or negative (polar nematic phase).
+
+The molecular field is explicitly given by:
+\begin{equation}
+h_{\alpha}=-\alpha_2 P_{\alpha} -\alpha_4 P_{\alpha} P_{\beta}P_{\beta}+
+K \partial_{\beta}\partial_{\beta} P_{\alpha}+2 K_{LC}
+P_{\gamma}\partial_{\beta}\partial_{\beta} (P_{\gamma}P_{\alpha}).
+\end{equation}
+
+The Navier-Stokes equation is:
+\begin{equation}
+\rho \left(\partial_t+v_{\beta}\partial_{\beta}\right) v_{\alpha} =
+\partial_{\beta} \Pi_{\alpha\beta} 
++\eta \partial_{\beta}\partial_{\beta} v_{\alpha}.
+\end{equation}
+
+The stress tensor this time is 
+\begin{eqnarray}
+\Pi_{\alpha\beta} =
+\frac{1}{2}\left(P_\alpha h_\beta - P_\beta h_\alpha \right)
+- \lambda\left( (P_\alpha h_\beta + P_\beta h_\alpha)/2 
+- (1/3)P_\gamma P_\gamma \delta_{\alpha\beta} \right)
+\\
+-  \zeta ( P_\alpha P_\beta - (1/3)P_\gamma P_\gamma \delta_{\alpha\beta})
+- K \partial_\alpha P_\gamma \partial_\beta P_\gamma.
+\end{eqnarray}
+The active term in the stress tensor, proportional
+to $\zeta$, has the same form for apolar and polar active gels.
+
 \section{Code Design}

### 2010-10-21 trunk/htdocs/user.tex
I've separated out a 'user guide' from the main sections, all of which needs work
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1156 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  user.tex
%
%  The 'user' section which holds an overview, and discusses
%  compilation and running.
%
%  $Id: user.tex,v 1.1 2010-10-21 18:05:50 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2010 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{Overview}

We aim to provide a robust and portable code, written in C, which
can be used to perform serial and scalable parallel simulations of
complex fluid systems based around hydrodynamics via the Lattice
Boltzmann method. Complex fluids are introduced via a free energy,
of which a number of different examples are available. The preferred
method of dealing with the corresponding order parameter equations
is now by using finite difference. However, for the case of a binary
fluid, the `full LB' approach, using two distributions, is retained
as an option for the time being.

\subsection{Availability}

We have a NeSCForge project account available at

\texttt{http://forge.nesc.ac.uk/projects/ludwig/}

This provides controlled access to known persons and provide CVS, bug
tracking, and so on. The NeSCForge CVS is not available to the public.

\textbf{IMPORTANT:} NeSCForge is scheduled to go out of service on 20th
December 2010. KS will announce an alternative as soon as possible.
This may mean some disruption.

To download, e.g., the \texttt{src} module, you need
\begin{verbatim}
cvs -z3 -d:ext:developername@forge.nesc.ac.uk:/cvsroot/ludwig co src
\end{verbatim}
where \texttt{developername} is replaced by your Nescforge user name.
(Note that this requires secure shell; you may need to set the
environment variable \texttt{CVS\_RSH=ssh} if you do not appear
to be getting any response.)

Currently, you need to download at least the cvs modules \texttt{src},
and \texttt{mpi\_s}. The \texttt{mpi\_s} module contains a `stub'
library to replace MPI in serial, and is required to run in serial.
This documentation is in the module
\texttt{htdocs}, unit tests are in \texttt{tests}, and the utilities
are in the module \texttt{util}.

\subsection{Compilation}

Locate your local C compiler. The following example uses \texttt{gcc}
in serial, and \texttt{mpicc} in parallel. For MPI in particular,
local details may vary.

\subsubsection{Serial}

First, compile the MPI stub library in the \texttt{mpi\_s}
directory. Do this by editing the Makefile, and checking the compiler
is appropriate. You should be able to build the library and run the
tests using, e.g.,:

\begin{verbatim}
$ make libc
gcc -Wall -I.  -c mpi_serial.c
ar -cru libmpi.a mpi_serial.o
$ make testc
gcc -Wall   mpi_tests.c -L. -lmpi
./a.out
Running mpi_s tests...
Finished mpi_s tests ok.
\end{verbatim}

Now compile the main code in the \texttt{src} directory. Again
edit the Makefile to check the compiler. At this stage, you also
need to choose the LB model using exactly one of the preprocessor
options for D2Q9 (fluid only problems), D3Q15, or D3Q19. Now type,
e.g., using D3Q19:

\begin{verbatim}
$ make serial
gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d2q9.c
gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d3q15.c
gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d3q19.c
gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c model.c
...
\end{verbatim}

This should provide an executable \texttt{Ludwig.exe} which is linked
against the MPI stub library.

\subsubsection{Parallel}

Here, you do not need to compile the stub library. Simply compile
the main code with, e.g.,:
\begin{verbatim}
$ make mpi
mpicc -D_D3Q19_ -O3 -Wall -I. -c d2q9.c
mpicc -D_D3Q19_ -O3 -Wall -I. -c d3q15.c
mpicc -D_D3Q19_ -O3 -Wall -I. -c d3q19.c
mpicc -D_D3Q19_ -O3 -Wall -I. -c model.c
...
\end{verbatim}

Invoke the executable with the MPI launcher in the usual way for your
system, e.g., \texttt{mpirun -np 8 ./Ludwig.exe} for 8 processors,
and so on.

\subsubsection{C assertions}

The code makes quite a lot of use of standard C assertions, which
are useful to prevent errors. They do result in a considerably
slower execution in some instances, so production runs should
switch off the assertions with the standard \texttt{NDEBUG}
preprocessor flag.

\section{User Input}

The executable will look for an input file at run time.
By default, this file is named \texttt{input} and should be in the
current working directory. The alternative is to specify the name
of the file explicitly on the command line:
\begin{verbatim}
$ ./Ludwig.exe input_file
\end{verbatim}
A reference input file \texttt{input.ref} is provided as a template.

The contents of the input file are made up of \textit{key value}
pairs with control the run-time behaviour. Blank lines and lines
begining with \# are ignored as comments. The following describes
the effect of various keys. Most keys have a default value which
will be used by the code if the corresponding key is not present
(or commented out) in the input file.

\subsection{Setting up the input}

Decide what free energy you want and check the relevant sections
of this dcoumentation. If you want a simple fluid only, the free
energy should be set to \texttt{none}.

You should set at least the basic fluid parameters, the system
size, and number of steps to be run.

\textbf{IMPORTANT:} Although many incompatible choices of input
parameters are trapped at run time, you must be careful in how
the input file is set out. If you mistakenly set or unset some
options in the input, erroneous results will surely result...

\subsection{Basic run parameters}

\inputkey{N\_cycles}

The number of lattice Boltzmann time steps to execute. Default
value: 0.

\inputkey{N\_start}

By default the code will start from time step $t = 0$. If you 
wish to restart from a previously saved configuration, set
the appropriate value for \texttt{N\_start}. The code will
execute \texttt{N\_cycles} steps starting from this point.

\inputkey{size}

Controls the number of lattice points in the $(x, y, z)$ directions,
respectively. Default value: 64\_64\_64. If a two-dimensional system
is required, the $z$-direction can be set to 1.

\inputkey{grid}

In parallel, this sets the processor Cartesian decomposition, i.e.,
the number of processors in each dimension of the Cartesian communicator.
The default value may be implementation dependent as it is that returned
by \texttt{MPI\_Dims\_create()}. For
example:
\begin{verbatim}
size 64_64_64
grid 4_2_1
\end{verbatim}
gives a total (physical) system size of 64 lattice sites in each
direction, decomposed across 4 processors in the x-direction, 2 in
the y-direction and 1 (no decomposition) in the z-direction. The
total number of processors must therefore be 8. The local domain
size per processor is then 16x32x64.

\subsection{Fluid properties}

\inputkey{free\_energy}

This sets the free energy in use, and hence has many consequences.
The default value is \texttt{none}, i.e., use a simple fluid only.
The available choices are for symmetric binary fluids, Brazovskii
smectics, polar active gels, and liquid crystals. The surfactant
model is currently being rebuilt.

\inputkey{viscosity}

Sets the LB fluid shear viscosity $\eta$ (and related relaxation time).
Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
(i.e., relaxation time equal to unity).

\inputkey{viscosity\_bulk}

Sets the LB bulk viscosity $\zeta$ (and related relaxation time).
This can be useful if you are concerned with incompressiblity
violations, in which case $\zeta >> \eta$ can be useful. The
default value is $\zeta = \eta$.

\inputkey{isothermal\_fluctuations}

This switches on the fluctuationing hydrodynamics. Default is off.

\inputkey{temperature}

The LB 'temperature' using fluctuating hydrodynamics. Safe values
are $0.0001 > kT > 0$.

\inputkey{ghost\_modes}

Allows you to switch the ghost modes off in the collision stage.
Default is on.

\subsection{Symmetric binary fluid}

\inputkey{free\_energy symmetric}

The following applies for the binary fluid problem with
compositional order parameter $\phi$ and free energy
(excluding the term in the density $\rho$):
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
\end{equation}

This is described in some detail
by Kendon et al \cite{viv}. The first two terms represent the bulk
contribution, whereas the term in $\kappa$ penalises curvature in
the interface.

\inputkey{A} The parameter $A$. Note $A < 0$.

\inputkey{B} The parameter $B$. Note $B = -A$ for common usage,
although this is not enforced.

\inputkey{K} The parameter $\kappa$, which is positive.

The order parameter evolution is determined by a Cahn-Hilliard equation
with mobility set by

\inputkey{mobility} Sets mobility $M$ (uniform in space).

The following parameters control the initialisation of the order
parameter.

\inputkey{phi\_initialisation}

Determines how the compositional order parameter is initialised
at the start of the run. If set to \texttt{spinodal} the value
is set to $\phi_0$ as set above plus or minus a random noise, the
magnitude of which is set by the value of the \texttt{noise} key.
If set to \texttt{block} a one-dimensional profile is set up in
the $z$-direction representing two blocks of fluid with
$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
appropriate width. The $\phi = -1$ section is in the middle.

\inputkey{phi0}

The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

\inputkey{noise}

The magnitude of the initial fluctuations in $\phi$ used to
initiate spinodal decomposition.

\subsubsection{Binary fluid using two distributions}

\inputkey{symmetric\_lb}

This is the special case where the composition is represented
by a second LB distribution, and an appropriate lattice kinetic
equation approximating the Cahn-Hilliard equation is solved.
In this case the above parameters have the same meaning.

\subsection{Brazovskii}

\inputkey{free\_energy brazovskii}

This is similar to the symmetric free energy, but with one extra term
in a higher derivative of $\phi$.
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
\end{equation}

The parameters now include $C$. For $A<0$, phase separation occurs
with a result depending on $\kappa$: ones get two symmetric phases
for $\kappa >0$ (cf. symmetric) or a lamellar phase for
$\kappa < 0$.

\inputkey{A} Bulk parameter $A < 0$.

\inputkey{B} Bulk parameter $B = -A$.

\inputkey{K} Negative for lamellar phase.

\inputkey{C} Positive.

Other parameters, including the mobility, are set as for the
symmetric free energy (see above).

\subsection{Surfactant in binary fluid}

\inputkey{free\_energy surfactant}

Currently under reconstruction.

This is for use with a two order parameter model the first of
which is the composition, as for the symmetric free energy,
while the second $\psi$ represents surfactant concentration.
The free energy density is made up of a number of terms:

\begin{equation}
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
\end{equation}
is the standard symmetric term related to the composition.
These parameters are represented in the input as

\inputkey{surf\_A}
\inputkey{surf\_B}
\inputkey{surf\_kappa}

\begin{equation}
D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
\end{equation}
represents the energy of surfactant in the bulk phase with surfactant
concentration $0 < \psi < 1$ with a single parameter $D$;
\begin{equation}
-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
\end{equation}
where the term in $\epsilon$ represents the energy reduction by
adsorbing surfactant at the interface and the term in $\beta$
provides an additional favourable term modelling the fact that
the molecules like to line up; finally, there is an extra term
\begin{equation}
+{\textstyle \frac{1}{2}} W \psi \phi^2
\end{equation}
which is included to stabilise the interface \cite{theissengompper}.

The corresponding keys for the input file are:

\inputkey{D} is the bulk surfactant contribution parameter

\inputkey{epsilon} is the parameter $\epsilon$

\inputkey{W} is the parameter $W$

\inputkey{beta} is the value of the parameter $\beta$.

\inputkey{phi\_b} the initial, uniform, background surfactant
concentration $\psi_b$.

\inputkey{mobility\_psi} Sets the (uniform) mobility for the surfactant
as appropriate for the Cahn-Hilliard equation.

\subsection{Polar Active Gel}

\inputkey{free\_energy polar\_active}

This models a polar active gel with vector order parameter $P_\alpha$.
The free energy density is:

The corresponding input parameters are:

\inputkey{polar\_active\_a}

\inputkey{polar\_active\_b}

\inputkey{polar\_active\_k}

\inputkey{polar\_active\_klc}

\inputkey{polar\_active\_zeta}

\inputkey{polar\_active\_lambda}

\inputkey{leslie\_ericksen\_gamma}

\inputkey{leslie\_ericksen\_swim}

\subsection{Liquid Crystal}

\inputkey{free\_energy lc\_blue\_phase}

Here, we have a tensor order parameter $Q_{\alpha\beta}$.
The free energy density is
\begin{eqnarray}
f = {\textstyle\frac{1}{2}}A_0(1 - \gamma/3)Q^2_{\alpha\beta}
  - {\textstyle\frac{1}{3}}A_0 \gamma
Q_{\alpha\beta}Q_{\beta\delta}Q_{\delta\alpha}
 + {\textstyle\frac{1}{4}}A_0 \gamma (Q^2_{\alpha\beta})^2
\nonumber
\\
+ {\textstyle\frac{1}{2}} \Big(
\kappa_0 (\epsilon_{\alpha\delta\sigma} \partial_\delta Q_{\sigma\beta} +
2q_0 Q_{\alpha\beta})^2 + \kappa_1(\partial_\alpha Q_{\alpha\beta})^2 \Big)
\end{eqnarray}

The corresponding input parameters are:

\inputkey{lc\_a0} The bulk free energy parameter $A_0$

\inputkey{lc\_gamma} The bulk free energy parameter $\gamma$

\inputkey{lc\_q0} The pitch wavenumber $q_0 = 2\pi / p$, where p
is the pitch.

\inputkey{lc\_kappa0} Elastic constant $\kappa_0$ in distortion free energy

\inputkey{lc\_kappa1} Elastic constant $\kappa_1$ in distortion free energy

Note that the code currently enforces the `one elastic constant'
approximation ($\kappa_0 = \kappa_1$), so both these values must
be equal in the input. This constraint may be relaxed in the future.

\inputkey{lc\_Gamma} The rotational diffusion constant appearing in
the Beris Edwards equation (note \texttt{lc\_gamma} and \texttt{lc\_Gamma}
are different).

\inputkey{lc\_xi} Aspect ratio for rod lie molecules appearing in the
stress tensor.

\subsubsection{Initial LC order parameter $Q_{\alpha\beta}$}

\inputkey{lc\_q\_initialisation}

There are a number of useful choices here. The key value
\texttt{twist} initialises a simple cholesteric with the
helical axis in the $z-$direction;  \texttt{o8m}
gives equilibrium BPI, \texttt{o2} gives BPII. The appropriate
choices for the other parameters must be made correspondingly.

An additional \texttt{lc\_q\_init\_amplitude} value sets the
scalar order parameter  `amplitude' in each case.

\subsubsection{The redshift}

\inputkey{lc\_init\_redshift}

Sets the initial value of the redshift parameter. For example,
for the \texttt{o8m} initialisation, the redshift can be set to 0.83,
and for the \texttt{o2} initialisation, the redshift can be set to 0.91.
The default redshift is unity.

\subsection{Colloidal particles}

The are a number of important options required for colloids. If
none of these keys are present, no colloids will be used.

\inputkey{colloid\_init}

This determines how the code initialises particles (or not). As stated, the
default is to have no particles (value \texttt{no\_colloids}). There
are two other options at the moment.

There are two options:

\inputkey{colloid\_init    from\_file}

requires a preprepared file of colloid information. To understand
how to create such a file in the correct format,
see the example program \texttt{colloid\_file.c} in the util
directory.

\inputkey{colloid\_init   random}

This may be useful to initialise a limited number of colloids at
low volume fraction. Note that there is no guarantee that the
particles do not overlap, in which case the initialisation will
terminate. The number of particles and their properties
are determined by the value of the keys described below.

\inputkey{colloid\_type}

This controls the the type of particle. Appropriate values are
\texttt{inactive}, giving a standard fully resolved LB colloid;
\texttt{active} switches on the correction to BBL appropriate
for active particles (expecting parameters $b_0$ and $b_1$ to be
set); \texttt{subgrid} gives unresolved particles (radius $< 1$
lattice unit) and no BBL.

If more than one particle is used, all positions are set at random.

\inputkey{colloid\_gravity  0.0\_0.0\_-0.0001}

Sets a body force on each particle. Useful, for example, for
sedimentation. For example, the above gives a constant force in
the negative $z$-direction on each particle.
The code automatically computes the compensating body force on fluid
nodes required to give no net change in the total momentum of the
system at each time step.

\inputkey{colloid\_cell\_min}

This sets the minimum cell list width. The cell list has a number
of purposes: it is used to compute colloid-colloid interactions
at $O(N)$, but it is also used in the message passing episodes
for the colloids. This key must be set and there must be at least
two cells per processor domain.

\subsubsection{Initialisation from file}

A file containing initial state properties for one or more colloids
can be read at run file. This file must be of the correct from, but
can be ASCII or binary. To help to create such a file, an example
is given in \texttt{./util/colloid\_file.c}. This file will be
read correctly in both serial and in parallel. Colloid positions
should always be in terms of the global coordinates. By default,
the input expected is \texttt{config.cds.init.001-001}.

All colloid output is in serial (ie., a single file is produced)
at the moment. Again, it may be ASCII or binary.

\subsubsection{Initialisation at random}

\inputkey{colloid\_random\_no} Sets the total number of particles

\inputkey{colloid\_random\_a0}

\inputkey{colloid\_random\_ah}

\inputkey{colloid\_random\_dh}

The parameter $a_0$ the nominal radius of the particle on the lattice.
This is used to construct the links for BBL. The smallest acceptable LB
particle is $a_0 = 1.25$ (approximately). There are 'magic' values
including 1.25, 2.3, 3.7, 4.77 which minimise discretisation effects.
The hydrodynamic radius of the particles is $a_h$. For all
reasonable applications choose $a_h = a_0$. This is the radius for
all physical purposes. The \texttt{colloid\_random\_dh} parameter
specifies the smallest allowed surface-surface separation if two or
more particles (or walls) are required. This is useful to prevent
very close particles experiencing large and destabilising interactions
at start time.

The other state for the initial colloids can be set uniformly by
means of the keys:

\inputkey{colloid\_random\_v0} initial velocity vector

\inputkey{colloid\_random\_w0} initial angular velocity vector

\inputkey{colloid\_random\_s0} initial spin vector

\inputkey{colloid\_random\_m0} initial direction of motion vector for active
particles

\inputkey{colloid\_random\_b1} active parameter $b_1$

\inputkey{colloid\_random\_b2} active parameter $b_2$ (should be combined
with the \texttt{active} particle type key as discussed above).

There is one special case. If you want exactly one particle, you may
set its initial position with

\inputkey{colloid\_random\_r0} the initial position vector.

\subsection{Boundary conditions}

Solid planar boundary walls can be added at the extremities of the
system (i.e., the volume of fluid is exactly that specified by
the \texttt{size} keyword). The wall can be in one, two or all three
coordinate directions, i.e., a 'slab' geometry, and 'duct' geometry
or a 'box' geometry, respectively.

The boundary walls are switched on via the keyword value pair

\inputkey{boundary\_walls\_on [no | yes]}

If \texttt{yes} then the geometry is specified with the keyword
as one of

\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}

Note that these are the only valid selections. \texttt{1\_1\_0}
gives periodic boundaries in the $x-$ and $y-$directions, with
walls at either edge in the $z-$direction. \texttt{1\_0\_0}
gives periodic conditions in the $x-$ direction only, while
\texttt{0\_0\_0} is a fully enclosed box.

In the case of the slab geometry, two further values are used so
that shear may be imparted to the system:

\inputkey{boundary\_speed\_bottom}

\inputkey{boundary\_speed\_top}

These values set the $x-$component of the wall velocity at
$z = z_{min}$ and $z = z_{max}$, respectively. Like all
velocities, these values must respect the low Mach number
constraint $u_x < c_s$.

\subsection{Lees-Edwards planes}

\inputkey{N\_LE\_plane}

sets the total number of Lees-Edwards planes. Default is zero.
The placing is as followings. The number of planes $n$ must
divide the lattice size in the $x$-direction to give an integer
$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.

\inputkey{LE\_plane\_vel}

sets the velocity of each plane relative to the lattice. All planes
have the same, constant, velocity.

\inputkey{LE\_init\_profile}

if set to 1, the fluid velocity is initialise to reflect a steady
state shear flow appropriate for the number of planes at the
given velocity. If set to zero, the fluid is initialised with
zero velocity.

Note that when the Lees Edwards boundaries are in place, output
files containing lattice quantities must be `unrolled' to remove
the time-dependent displacement of the planes. This is done using
the extraction utility (see section on parallel I/O).

\subsection{Input/Output}

\subsubsection{Serial I/O and standard output}

A variety of information is printed to standard output during the
course of the simulation. This should allow the user to keep track
of basic progress in terms of integrated quantities, total mass,
total momentum and so on. Frequency of output is controlled by

\inputkey{freq\_statistics}

which is an integer. Note that these global statistics require
global communication, and so can affect performance in parallel.
It is recommended that the frequency is kept to a minimum for
large systems (e.g., every 500 or 1000 time steps). This
consideration is true for most types of output (see section
on parallel I/O).

\subsubsection{Configurations}

It may be appropriate to save entire model configurations from
time to time, and certainly when a restart is wanted. For each lattice
quantity, in addition to the data themselves, the program produces a
single meta-data file (ending in
\texttt{.meta}) which is used to describe these data. Colloid
data does not require a meta-data file at the moment.

\inputkey{freq\_config}

This is an integer and sets the frequency of full configuration dumps.
These may be large depending on the size of the system.
Full configurations should consist of the LB distributions,
order parameter (if present) and colloids (if present).

\inputkey{config\_at\_end [yes | no]}

forces (or switches off) a configuration dump at the end of the run.
The default is to produce a configuration at the end of the run.

\subsubsection{Other lattice quantities}

A number of other lattice-based quantities are available as output.

\inputkey{freq\_measure}

sets the frequency of recording of colloid/order parameter output.
This includes the combined scalar order parameter and director
field output in the case of the liquid crystal free energy.

\inputkey{freq\_phi}

set frequency of order parameter output.

\inputkey{freq\_vel}

set frequency of velocity output. Note that all the hydrodynamic
quantities can be reconstructed from the distribution output, if
available. If only the velocity field is required, this method
allows a considerable saving in storage.

\subsubsection{I/O format}

\inputkey{vel\_format}

\inputkey{phi\_format}

May be either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
Configuration I/O is always in binary. It is recommended to use
binary output for speed and disk space considerations.

\subsubsection{Parallel I/O}

\subsection{Miscellaneous}

\inputkey{random\_seed}

sets the random number generator seed.

\subsection{Dealing with the Parallel I/O files}

When running in parallel, output of lattice-based (and particle)
quantities takes place in parallel. As a consequence, the order
of the output in the files is not in the same order as would be
the case in serial. This means the parallel output must be
manipulated into the correct order before any analysis.

Parallel I/O subdivides the processors into one or more groups,
each of which write data to a separate file. Information on the
content of each of these files is contained in `meta-data' file
for each quantity, which are produced automatically at the start
of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
If there are 8 I/O groups, the files will be named \texttt{008-001}
to \texttt{008-008} for the different groups.

The actual data is stored for each relevant time step as, e.g.,
\texttt{phi-000100.001-001} where the extension refers to the
I/O. To recombined a single, or multiple, data files into a
single file in the correct order, a utility program
\texttt{extract.c} is provided.

The executable takes two arguments, e.g.,

\inputkey{./extract phi.001-001.meta phi-000100.001-001}

where  the first is the first meta-data file in the relevant series,
and the second is the first data file for the relevant time step. The
program will combine the relevant quantities in the correct order
and produce a single new file \texttt{phi-000100} (in the above case).

\subsection{Restarting}

When restarting in parallel, the processor decomposition must be
preserved. An attempt to change the processor decomposition will
cause the configuration files to be read in the incorrect order,
and results will be wrong.

### 2010-11-29 trunk/htdocs/porous.tex
Seaparte file for porous media stuff
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1171 40876441-3496-0410-a613-a2fca35b2ae2

--- 2009-05-07 trunk/htdocs/sections.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  sections.tex
-%
-%  Content for Ludwig documentation.
-%
-%  $Id: sections.tex,v 1.10 2009-05-07 15:29:05 kevin Exp $
-%
-%  Edinburgh Soft Matter and Statistical Physics Group and
-%  Edinburgh Parallel Computing Centre
-%
-%  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-\subsection{Note to Readers}
-
-This document is intended as a starting point for a full-blown
-specification of the next version of the Ludwig code, and everything
-to do with it. It is (manifestly) not complete.
-
-\subsection{Some History}
-
-Early Lattice Gas and Lattice Boltzmann. Early Lattice Boltzmann at Edinburgh.
-Ludwig.
-
-\subsection{Aims}
-
-We aim to provide a robust and portable code, written in C, which
-can be used to perform serial and scalable parallel simulations of
-complex fluid systems based around the Lattice Boltzmann method.
-
-\subsection{Availability}
-
-We now have a NeSCForge project account available at
-
-\texttt{http://forge.nesc.ac.uk/projects/ludwig/}
-
-This provides controlled access to known
-persons and provide CVS, bug tracking, and so on. It is
-\textbf{essential} that people sign up to this, as we must
-move to a single supported version.
-
-The NeSCForge CVS is not available to the public.
-
-\subsection{Download}
-
-\section{User Input}
-
-By default the executable will look for an input file at run time.
-By default, this file is named \texttt{input} and should be in the
-current working directory. The alternative is to specify the name
-of the file explicitly on the command line:
-\begin{verbatim}
-$ ./Ludwig.exe input_file
-\end{verbatim}
-
-The contents of the input file are made up of \textit{key value}
-pairs with control the run-time behaviour. Blank lines and lines
-begining with \# are ignored as comments. The following describes
-the effect of various keys. Most keys have a default value which
-will be used by the code if the corresponding key is not present
-(or commented out) in the input file.
-
-\subsection{Basic run parameters}
-
-\inputkey{N\_cycles}
-
-The number of lattice Boltzmann time steps to execute. Default
-value:
-
-\texttt{N\_cycles 0}
-
-\inputkey{t\_start}
-
-By default the code will start from time step $t = 0$. If you 
-wish to restart from a previously saved configuration, set
-the appropriate value for \texttt{t\_start}. The code will
-execute \texttt{N\_cycles} steps starting from this point.
-
-\inputkey{input\_config}
-
-The file name stub for input configuration required for restarts.
-For example, to restart from the file config.out100 previously
-produced at the end of a 100 step run use
-\begin{verbatim}
-t_start 100
-input_config config.out100
-\end{verbatim}
-
-If you are using parallel I/O, the filename stub is required; the
-file extensions are computed automatically.
-
-\inputkey{size}
-
-Controls the number of lattice points in the $(x, y, z)$ directions,
-respectively. Default value:
-
-\texttt{size 64\_64\_64}
-
-\inputkey{grid}
-
-In parallel, this sets the processor Cartesian decomposition, i.e.,
-the number of processors in each dimension of the Cartesian communicator.
-The default value may be implementation dependent as it is that returned
-by \texttt{MPI\_Dims\_create()}. For
-example:
-\begin{verbatim}
-size 64_64_64
-grid 4_2_1
-\end{verbatim}
-gives a total (physical) system size of 64 lattice sites in each
-vdirection, decomposed across 4 processors in the x-direction, 2 in
-the y-direction and 1 (no decomposition) in the z-direction. The
-total number of processors must therefore be 8. The local domain
-size per processor is then 16x32x64.
-
-\subsection{Fluid properties}
-
-\inputkey{viscosity}
-
-Sets the LB fluid shear viscosity $\eta$ (and related relaxation times).
-Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
-(i.e., realaxation time equal to unity).
-
-\inputkey{viscosity\_bulk}
-
-Sets the LB bulk viscosity $\zeta$ (and related relaxation times).
-This can be use if you are concerned with incompressiblity
-violations, in which case $\zeta >> \eta$ can be useful. The
-default value is $\zeta = \eta$.
-
-\inputkey{temperature}
-
-The LB 'temperature' using fluctuating hydrodynamics. Safe values
-are $0.0001 > kT > 0$.
-
-\inputkey{rho0}
-
-Sets the mean density of the fluid. The default value is unity, and
-it is recommended that the value should not be changed.
-
-\subsection{Binary fluid parameters}
-
-The are currently three choices of free energy relating to
-various systems.
-
-\subsubsection{Symmetric binary fluid}
-
-\inputkey{free\_energy symmetric}
-
-The following applies for the binary fluid problem with
-compositional order parameter $\phi$ and free energy
-(excluding the term in the density $\rho$):
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
-\end{equation}
-
-This is described in some detail
-by Kendon et al \cite{viv}. The first two terms represent the bulk
-contribution, whereas the term in $\kappa$ penalisies curvature in
-the interface.
-
-\inputkey{A}
-
-The parameter $A$. Note $A < 0$.
-
-\inputkey{B}
-
-The parameter $B$. Note $B = -A$ for common usage, although this is
-not enforced.
-
-\inputkey{K}
-
-The parameter $\kappa$, which is positive.
-
-\subsubsection{Brazovskii free energy}
-
-This is similar to the symetric, but with one extra term which
-requires description.
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
-\end{equation}
-
-\subsubsection{Surfactant model free energy}
-
-\inputkey{free\_energy surfactant}
-
-This is for use with a two order parameter model the first of
-which is the composition, as for the symmetric free energy,
-while the second $\psi$ represents surfactant concentration.
-The free energy density is made up of a number of terms:
-
-\begin{equation}
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-\end{equation}
-is the standard symmetric term related to the composition as above;
-
-\begin{equation}
-D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
-\end{equation}
-represents the energy of surfactant in the bluk phase with surfactant
-concentration $0 < \psi < 1$ with a single parameter $D$;
-\begin{equation}
--{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
--{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
-\end{equation}
-where the term in $\epsilon$ represents the energy reduction by
-adsorbing surfactant at the interface and the term in $\beta$
-provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
-\begin{equation}
-+{\textstyle \frac{1}{2}} W \psi \phi^2
-\end{equation}
-which is included to stabilise the interface \cite{theissengompper}.
-
-The corresponding keys for the input file are:
-
-\inputkey{D}
-
-is the bulk surfactant contribution parameter
-
-\inputkey{epsilon}
-
-is the parameter $\epsilon$
-
-\inputkey{W}
-
-is the parameter $W$
-
-\inputkey{beta}
-
-is the value of the parameter $\beta$ (not implemented yet).
-
-\inputkey{phi\_b}
-
-the initial, uniform, background surfactant concentration $\psi_b$.
-
-\subsection{Other Binary Fluid Parameters}
-
-\inputkey{mobility}
-
-Sets the compositional order parameter mobility $M$ (uniform in space)
-which appears in the Cahn-Hilliard equation.
-
-\inputkey{mobilty\_psi}
-
-Sets the (uniform) mobility for the surfactant as appropriate for
-the Cahn-Hilliard equation.
-
-\inputkey{phi0}
-
-The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
-The default value is zero, i.e., a symmetric 50:50 mixture by volume.
-
-\inputkey{phi\_initialisation}
-
-Determines how the compositional order parameter is initialisated
-at the start of the run. If set to \texttt{spinodal} the value
-is set to $\phi_0$ as set above plus or minus a random noise, the
-magnitude of which is set by the value of the \texttt{noise} key.
-If set to \texttt{block} a one-dimensional profile is set up in
-the $z$-direction representing two blocks of fluid with
-$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
-$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
-appropriate width. The $\phi = -1$ section is in the middle.
-
-\inputkey{noise}
-
-The magnitude of the initial fluctuations in $\phi$ used to
-initiate spinodal decompsotion.
-
-\subsection{Colloidal particles}
-
-There are quite a few options here.
-
-\inputkey{colloid\_init}
-
-This determines how the code initialises particles (or not). The
-default is to have no particles (value \texttt{no\_colloids}). There
-are two other options at the moment.
-
-\texttt{colloid\_init    fixed\_number\_monodisperse}
-
-initialises a fixed number of particles with randomly selected
-positions. Note that there is no gaurentee that the particles
-do not overlap. The number of particles is determined by
-the value of the key \texttt{colloid\_no}.
-
-\texttt{colloid\_init   fixed\_volume\_fraction\_monodisperse}
-
-initialises a given volume fraction of particles on a (roughly)
-bcc lattice. The particles will not overlap. The volume fraction
-is set using the key \texttt{colloid\_vf}.
-
-\inputkey{colloid\_a0}
-
-Sets the nominal radius of the particle on the lattice $a_0$. This is
-used to construct the links for BBL. The smallest acceptable LB
-particle is $a_0 = 1.25$ \cite{nl02}. There are 'magic' values
-including 1.25, 2.3, 3.7, 4.7 which minimise discretisation effects.
-
-\inputkey{colloid\_ah}
-
-Sets the hydrodynamic radius of the particles $a_h$. For all
-reasonable applications choose $a_h = a_0$.
-
-\inputkey{colloid\_mc\_steps}
-
-It is possible to have the code follow the initialisation of
-particle positions by a number of Monte Carlo steps. This can
-help to remove overlaps (fixed\_number\_monodisperse) or
-melt the lattice (fixed\_volume\_fraction\_monodisperse).
-
-\inputkey{colloid\_gravity}
-
-Sets a body force on each paarticle. Useful, for example, for
-sedimentation. For example,
-
-\inputkey{cell\_list\_min}
-
-This sets the minimum allowable width of the cell list used to
-store the particles. This should be as small as possible for
-efficiency purposes but must be large enough to capture all
-possible interactions.
-
-\texttt{colloid\_gravity  0.0\_0.0\_-0.0001}
-
-gives a constant force in the negative $z$-direction on each particle.
-The code automatically computes the compensating body force on fluid nodes
-required to give no net change in the total momentum of the system at
-each time step.
-
-\subsection{Active particles}
-
-Not in general use yet.
-
-\subsection{Particle-particle interactions}
-
-\subsection{Boundary conditions}
-
-Solid planar boundary walls can be added at the extremities of the
-system (i.e., the volume of fluid is extactly that specified by
-the \texttt{size} keyword). The wall can be in one, two or all three
-coordinate direcctions, i.e., a 'slab' geometry, and 'duct' geometry
-or a 'box' geometry, respectively.
-
-The boundary walls are switched on via the keyword value pair
-
-\inputkey{boundary\_walls\_on [no | yes]}
-
-If \texttt{yes} then the geometry is specified with the keyword
-as one of
-
-\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}
-
-Note that these are the only valid selections. \texttt{1\_1\_0}
-gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction. \texttt{1\_0\_0}
-gives periodic conditions in the $x-$ direction only, while
-\texttt{0\_0\_0} is a fully enclosed box.
-
-In the case of the slab geometry, two further values are used so
-that shear may be imparted to the system:
-
-\inputkey{boundary\_speed\_bottom}
-
-\inputkey{boundary\_speed\_top}
-
-These values set the $x-$component of the wall velocity at
-$z = z_{min}$ and $z = z_{max}$, respectively. Like all
-velocities, these values must respect the low Mach number
-constraint $u_x < c_s$.
-
-\subsection{Input/Output and Parallel I/O}
-
-\inputkey{freq\_statistics}
-
-sets the frequency for reporting general information from
-the code. (In general, the function
- \texttt{is\_statistics\_step()} returns true at this frequency.
-
-\inputkey{freq\_measure}
-
-sets the frequency of recording of colloid/order parameter output.
-(In general, the function \texttt{is\_measurement\_step()} returns
-true at this frequency.)
-
-\inputkey{freq\_config}
-
-set frequency of full configuration (checkpoint) dumps suitable
-for restart.
-
-\inputkey{freq\_phi}
-
-set frequency of order parameter output.
-
-\inputkey{phi\_format}
-
-either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-
-\inputkey{freq\_vel}
-
-set frequancy of velocity output.
-
-\inputkey{vel\_format}
-
-either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-
-\inputkey{config\_at\_end}
-
-enforces full configuration dump at the end of execution. The
-default is 'yes'.
-
-\subsection{Lees-Edwards planes}
-
-\inputkey{N\_LE\_plane}
-
-sets the total number of Lees-Edwards planes. Default is zero.
-The placing is as followings. The number of planes $n$ must
-divide the lattice size in the $x$-direction to give an integer
-$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
-
-\inputkey{LE\_plane\_vel}
-
-sets the velocity of each plane relative to the lattice. All planes
-have the same, constant, velocity.
-
-\inputkey{LE\_init\_profile}
-
-if set to 1, the fluid velocity is initialise to reflect a steady
-state shear flow appropriate for the number of planes at the
-given velocity. If set to zero, the fluid is initialised with
-zero velocity.
+#############################################################################
+#
+#  porous.tex
+#
+#  Some considerations for porous media.
+#
+#  $Id: porous.tex,v 1.1 2010-11-29 16:45:57 kevin Exp $
+#
+#  Edinburgh Soft Matter and Statistical Physics Group and
+#  Edinburgh Parallel Computing Centre
+#
+#  Kevin Stratford (kevin@epcc.ed.ac.uk)
+#  (c) 2010 The University of Edinburgh
+#
+#############################################################################
 
@@
 
-\subsection{Miscellaneous}
+\subsubsection{Matching lattice and real units in porous media}
 
-\inputkey{random\_seed}
+Following Succi \cite{succi} (Chapter~8) the following argument
+can be made to match lattice quantities and real physical
+quantities for a given system of interest. Suppose we have
+a structure with characteristic pore size $h = 1$ micron
+and characteristic sample size of 1 cm. If the fluid
+(e.g., water) has viscosity $\eta = 10^{-3}$ Pa s and density
+$\rho = 10^3$ kg~s$^{-1}$, and a typical flow is
+$U \sim 10^{-3}$ m~s$^{-1}$, then the pore scale Reynolds
+number is $Re_h \sim 10^{-3}$.
 
-sets the random number generator seed.
+If the speed of sound in water is 1500~m~s$^{-1}$,
+then the corresponding time scale related to the
+speed of sound and the pore size is $T = h/ c_s$,
+which is $\sim 10^{-10}-10^{-9}$~s. (Exact matching
+is acheived via the dimensionless group $\Delta x / c_s\Delta t$,
+with the lattice speed of sound $1/\sqrt{3}$).
 
-\subsection{Dealing with the Parallel I/O files}
-
-When running in parallel, output of lattice-based (and particle)
-quantities takes place in parallel. As a consequence, the order
-of the output in the files is not in the same order as would be
-the case in serial. This means the parallel output must be
-manipulated into the correct order before any analysis.
-
-Parallel I/O subdivides the processors into one or more groups,
-each of which write data to a separate file. Infomation on the
-content of each of these files is contained in `meta-data' file
-for each quantity, which are produced automatically at the start
-of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
-If there are 8 I/O groups, the files will be named \texttt{008-001}
-to \texttt{008-008} for the different groups.
-
-The actual data is stored for each relevant time step as, e.g.,
-\texttt{phi-000100.001-001} where the extention refers to the
-I/O. To recombined a single, or multiple, data files into a
-single file in the correct order, a utility program
-\texttt{extract.c} is provided.
-
-The exectutable takes two arguments, e.g.,
-
-\inputkey{./extract phi.001-001.meta phi-000100.001-001}
-
-where  the first is the first meta-data file in the relevant series,
-and the second is the first daa file for the relevant time step. The
-program will combine the relevant quantities in the correct order
-and produce a singel new file \texttt{phi-000100} (in the above case).
-
-\subsection{Restarting}
-
-When restarting in parallel, the processor decomposition must be
-preserved. An attempt to change the processor decomposition will
-cause the configuration files to be read in the incorrect order,
-and results will be wrong.
-
-\section{The Lattice Boltzmann Method}
-
-We review here the lattice Boltzmann method applied to single
-phase problems.
-
-\subsection{The Navier Stokes Equation}
-
-We seek to solve the isothermal Navier-Stokes equations which, often
-written in vector form, express mass conservation
-\begin{equation}
-\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
-\label{eq_mass1}
-\end{equation}
-and the conservation of momentum
-\begin{equation}
-\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
--\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
-+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
-\label{eq_momentum1}
-\end{equation}
-Equation~\ref{eq_mass1} expresses the local rate of change of the
-density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
-mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
-Equation~\ref{eq_momentum1} expresses Newton's second law for
-momentum, where the terms on the right hand side represent the
-force on the fluid.
-
-For this work, it is more convenient to rewrite these equations
-in tensor notation, where Cartesian coordinates ${x,y,z}$ are
-represented by indices $\alpha$ and $\beta$, viz
-\begin{equation}
-\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
-\end{equation}
-and
-\begin{equation}
-\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
-= -\nabla_\alpha p
-+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
-+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
-\end{equation}
-Here, repeated indices are understood to be summed over.
-The conservation law is seem better if the forcing terms of the
-right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
-so that
-\begin{equation}
-\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
-\end{equation}
-In this case the expanded expression for the stress tensor is
-\begin{equation}
-\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
-+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
-\end{equation}
-where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
-$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
-modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.
-
-\subsection{The Distribution Function and its Moments}
-
-In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
-\begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
-\end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
-
-In general, the number of modes supported in $n$ and these can
-be written as
-\begin{equation}
-M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
-\end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
-\begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
-\end{equation}
-
-\subsubsection{Hydrodynamic modes}
-
-\subsubsection{D3Q15 ghost modes}
-
-\subsubsection{D3Q19 ghost modes}
-
-\subsection{The Lattice Boltzmann Equation}
-
-\subsection{The Collision}
-
-\subsection{External Body Forces}
-
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsubsection{Fluctuating ghost modes: D3Q15}
-
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Boundary walls}
-
-\subsection{Colloids}
-
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
-\section{Code Design}
-
-\subsection{Some Comments}
-
-In designing a code in C the main object of which is to do high
-performance computing, there are some inevitable tensions with
-software engineering best-practice. In particular, complete
-encapsulation of data in the spirit of using abstract data types
-tends to cause problems of performance.
-
-For example, encapsulation of the lattice Boltzmann distribution
-$f_i$, together with appropriate ``get'' and ``set'' functions for
-access when approapriate, is possible. There could then be a
-complete separation of interface and implementation of the
-distribution (other than that each individual $f_i$ is represented
-by a \texttt{double}). However, experience suggests that
-this imposes severe impediment to optimisation and hence
-unacceptable performance degradation.
-
-Schemes do exist by which one can approximate the object
-oriented programming style in C, and so achieve a separation
-of interface and implementation. However, this tends to put
-a heavy burden on the use of \texttt{void} pointers, itself
-a recipe for possible disaster. A truly object-oriented
-language such as Java would be preferable.
-
-So, for the time being, some relaxation of encapsulation is required.
-Exposing a pointer to the distributions which allows them to be accessed
-directly where required. This can be done merely by declaring the
-necessary variables \texttt{extern}. In practice, such deviations
-from encapsulation can be minimised in the lattice sector of the
-code. In particlar, \texttt{extern} declarations should not be
-required in \texttt{main()}, i.e., it should be able to write a
-new application without worrying about details of implementation.
-
-\subsubsection{Objects}
-
-Somewhat more difficult is the part of the code involving
-colloids, which would really favour a truly object-oriented
-approach.
-
-\subsubsection{Serial and parallel}
-
-The design decission has been taken to support execution of the
-code in both serial and parallel. The default compilation
-assumes that a serial executable is required, while  MPI code
-is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).
-
-There is no shared memory inplementation at this time.
-
-\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}
-
-\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}
-
-\begin{figure}
-\begin{center}
-%\includegraphics{xfig/sketch.eps}
-\end{center}
-\caption{A scheme by which different parts of the code might be
-encapsulated and in such a way to allow reliable development and
-testing. As this is C, will do not adopt a truly ``object'' langauge.
-Broadly, different modular parts should only depend on stuff above
-in the diagram. So, at the top, there is the parallel environment
-where communication occurs within MPI\_COMM\_WORLD. Built on top
-of this is the Cartesian Communicator responsible for halo swaps
-and so forth. The LE code may have a separate Communicator again. 
-(There's an argument to say the diagram should be the other way up!)}
-\end{figure}
-
-\section{Code Implementation}
-
-\subsection{Stuff}
-
-\subsection{Coordinate System}
-
-\subsubsection{The global coordinate system}
-
-The user specifies the system size in lattice units at run time,
-which sets the extent of memory which must be allocated for the
-Cartesian lattice. The extent of the physical coordinate system
-then reflects the lattice size directly (see Figure \ref{fig_c1}). 
-Each lattice site, or node,  has coordinates corresponding to its
-integer position in the system and can be thought of as being
-at the centre of a control volume one lattice unit on a side.
-This means that the nominal edges of the
-system are offset from the lattice sites themselves by half a
-lattice unit in each direction.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_c1.eps}
-\end{center}
-\caption{The global Cartesian coordinate system reflects the choice
-of lattice size; lattices sites are located at
-$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
-(the Figure only shows the labels in the $y-$direction for simplicity).
-In the Figure, the lattice sites in the domain proper are the full
-circles; halo sites, which are used to effect periodic boundary
-conditions, are included as open cirles}
-\label{fig_c1}
-\end{figure}
-
-For particles, the centres of which can move continuously across the
-lattice, the extents of the global coordinate system are then
-$L_{min} < x < L + L_{min}$ and so on.
-
-\subsubsection{Addressing the lattice: the local coordinate system}
-
-\subsection{Solid walls}
-
-General solid objects which are fixed for the duration of
-the simulation, such as porous media, can be defined
-by assigning solid/fluid status to each lattice node. 
-The links required between solid and fluid nodes can
-then be identified for the given model.
-
-\textit{Comment.} In general, a solid object will not
-affect the periodic nature of the simulation. A general
-solid object cannot move, i.e., the boundary velocity
-must be zero.
-
-\textit{Special case.} In the case where the solid object
-represents a flat boundary wall at the edge of the system
-in one or more directions, the periodicity is broken in
-the corresponding directions.
-
-\vfill
-\pagebreak
-
-\section{Code Interface}
-
-\subsection{The Parallel Environment}
-
-\texttt{\#include ``pe.h''}
-
-As the code is to be available in both serial and parallel versions,
-it is desirable to abstract the basic environment to some extent.
-The parallel environment is intended to support MPI communication
-within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
-consistent picture in serial (i.e., rank zero is the single process).
-
-\texttt{void pe\_init(int argc, char ** argv)}\\
-Responsible for initialisation of MPI, and hence must be first
-execuatable statement. The arguments are those to \texttt{main}
-which will be passed through to \texttt{MPI\_Init()}.
-
-\texttt{int pe\_rank(void)}\\
-Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
-serial.
-
-\texttt{int pe\_size(void)}\\
-Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
-1 in serial.
-
-\texttt{void pe\_finalise(void)}\\
-Responsible for \texttt{MPI\_Finalize()}, and must be the last
-executable statement.
-
-\texttt{void info(const char * fmt, ...)}\\
-Behaves like \texttt{printf}, but only produces output on the
-root process.
-
-\texttt{void verbose(const char * fmt, ...)}\\
-Behaves as \texttt{printf} and reports the rank of the
-calling process in \texttt{MPI\_COMM\_WORLD}.
-
-\texttt{void fatal(const char * fmt, ...)}\\
-Prints an error message and identifies
-the rank of the calling process before terminating the program.
-
-\subsection{Run Time Input}
-
-\texttt{\#include ``runtime.h''}
-
-The ability to set a wide range of parameters at run time is
-very convenient, and is supported via an input file in which the
-user specifies parameters as key value pairs. The code maintains
-a list of key value pairs from which the user can retrieve the
-appropriate values for given keys.
-
-\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
-This reads the named input file and sets up the list of keys.
-
-\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The number
-of key matches found is returned (i.e., 0 or 1).
-
-\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The
-number of key matches is returned.
-
-\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
-Set \texttt{value} to the string assocaited with \texttt{key}, if
-present. Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
-If \texttt{key} is present, set the associated 3-vector of integers.
-Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
-double v[])}\\
-If \texttt{key} is present, set the associated 3-vector of doubles.
-Returns the number of key matches found.
-
-\textit{Comment:} We probably need a function to report unused keys, so
-the user can be alerted if a key has unexpectedly been ignored.
-
-\subsection{Random Number Generators}
-
-\texttt{\#include ``ran.h''}
-
-Note that the random number generation in parallel is
-decomposition dependent.
-
-\texttt{void rand\_init(void)}\\
-Responsible for initialising RNG state.
-
-\texttt{double rand\_serial\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$ from
-the serial generator.
-
-\texttt{double rand\_serial\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the serial generator.
-
-\textit{double rand\_parallel\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$
-from the serial generator.
-
-\texttt{double rand\_parallel\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the parallel generator.
-
-\textit{Comment:} The serial generator will produce the
-same sequence on any MPI process, while the parallel generator
-will produce different sequences. The latter is intended for
-generating, specifically, fluctuations, where the need for
-decomposition-independance is questionable.
-
-\subsection{Timers}
-
-\texttt{\#include ``timer.h''}
-
-Performance figures require timing for different sections
-of the code. The code provides a set of routines to time
-standard and user-defined sections of code. Timers are
-started and stopped by call to the appropriate routines.
-
-\texttt{TIMER\_init()}\\
-This initialises all the timers to zero (e.g., at the start of
-the program).
-
-\texttt{TIMER\_start(const int t\_id)}\\
-Start the given timer.
-
-\texttt{TIMER\_stop(const int t\_id)}\\
-Stop the given timer.
-
-\texttt{TIMER\_statistics(void)}\\
-Prints the current state of all the active timers to standard
-output on the root process in a digestible form. In serial,
-the minimum, maximum, and mean time for each timer is reported
-in seconds (based on the number of start/stop cycles). In
-parallel, the same figures are computed on each process and the
-result of the corresponding \texttt{MPI\_Reduce()} operation is
-printed on the root process.
-
-The code provides a number of timers which would time standard
-parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
-total execution time, and a number of spare timers for users to
-use at will.
-
-\textit{Comment:}
-No intra-run variance (i.e., between time steps) is provided, as the
-calculation required is considerably more complex to generalise, and
-requires memory proportional to the number of time steps used.
-
-\textit{Comment:}
-Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
-provided to insert calls to the above functions for optional timing
-of performance-sensitive regions
-of the code in conjunction with the preprocessor flag
-\texttt{-D\_TIMING\_MACRO\_ON\_}.
-
-\subsection{Model Time Control}
-
-This provides time step control, event frequency (?)
-
-\texttt{void init\_control(void)}\\
-Initialise time controls.
-
-\texttt{int get\_step(void)}\\
-Return the current model time step.
-
-\texttt{int next\_step(void)}\\
-Increments the current time step. Returns number of steps left
-to run (i.e., 0 when finished).
-
-\subsection{Coordinate System}
-
-\texttt{\#include ``coords.h''}
-
-A three-dimensional Cartesian coordinate system is used throughout,
-with the exact system size usually chosen by the user at run time.
-The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
-used to identify the coordinate directions.
-
-This also provides access to the MPI Cartesian Communicator which
-is responsible for most communication in the model.
-
-\texttt{int N\_total(const int dim)}\\
-Returns the total number of lattice sites in the given Cartesian direction.
-
-\texttt{double L(const int dim)}\\
-Returns the length of the system in the given Cartesian direction
-(which is the same as the number of lattice sites).
-
-\texttt{double Lmin(const int dim)}\\
-Returns coordinate position of left-hand edge of system in given
-Cartesian direction.
-
-\texttt{int is\_periodic(const int dim)}\\
-Returns 1 if given Cartesian direction is periodic, else 0.
-
-\texttt{void cart\_init(void)}\\
-Responsible for setting up the Cartesian decomposition.
-
-\texttt{int cart\_rank(void)}\\
-Returns rank in the Cartesian communicator.
-
-\texttt{MPI\_Comm cart\_comm(void)}\\
-Returns the MPI Communicator handle for the Cartesian communicator.
-
-\texttt{int cart\_size(const int dim)}\\
-Returns the size of the Cartesian communicator grid in the given
-coordinate direcxtion.
-
-\texttt{int cart\_coord(const int dim)}\\
-Returns the coordinate of the current process in the Cartesian
-Communicator.
-
-\texttt{int cart\_neigh(const int dir, const int dim)}\\
-Returns the rank of the neighbouring process in the
-Cartesian communicator.
-
-\subsection{Colloids}
-
-\texttt{\#include ``colloids.h''}
-
-These functions provide the basic interface for use
-of colloidal particles in the code.
-
-\subsubsection{Colloid structure}
-
-The basic colloid struct is defined ...
-
-\subsubsection{Colloid interface}
-
-\texttt{void colloids\_init(void)}\\
-Call once for initialisation. This initialises the cell list,
-so the coordinate system must be initialised via a call to
-\texttt{coords\_init()} first.
-
-\texttt{void colloids\_finish(void)}\\
-Call once for finalisation. Frees all colloid-related memory.
-
-\texttt{Colloid * allocate\_colloid(void)}\\
-Returns a pointer to a newly allocated \texttt{Colloid} object
-or fails gracefully.
-
-\texttt{void free\_colloid(Colloid * p\_colloid)}\\
-Deallocates all memory associated with \texttt{p\_colloid} or
-fails gracefully.
-
-\texttt{struct c\_link allocate\_boundary\_link(void)}\\
-Returns a pointer to a newly allocated colloid boundary
-link or fails gracefully.
-
-\texttt{void free\_boundary\_link(struct c\_link *)}\\
-Deallocates memory associated with a link or fails gracefully.
-
-\texttt{void colloids\_memory\_report(void)}\\
-The module keeps track of how much memory has been allocated in
-total for colloids and boundary links. This reports the current
-usage.
-
-\texttt{get\_Ncell\_total()}\\
-
-\texttt{get\_Ncell\_local()}\\
-
-\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
-This returns a pointer to the Colloid at the head of the list at
-cell list location (i, j, k) or \texttt{NULL} if there is none.
-
-\texttt{void cell\_insert(Colloid * p\_colloid)}\\
-This inserts a colloid at the head of the appropriate list
-depending upon its position. This is the only way a colloid
-can be added to the list.
-
-\texttt{update\_cells(void)}\\
-Performs the update of the cell list in response to updated
-particle positions.
-
-\texttt{IVector cell\_coords(FVector r)}\\
-
-\subsection{Speculation}
-
-\subsection{The Free Energy}
-
-It is imagined that a single file will encapsulate all the details
-of a given free energy. The interface for different choices of
-free energy should be the same.
-
-\texttt{FE\_init()}
-
-Responsible for querying the runtime input key value pair list to
-find appropriate user parameters for the free energy, or
-falling back to default values.
-
-\texttt{double FE\_get\_sigma()}
-
-Return the surface tension. There should be a similar routine
-for the interfacial thickness.
-
-\texttt{double FE\_get\_mu(const type location)}
-
-Return the chemical potential at the single lattice
-node \texttt{location}.
-
-\textit{Comment:} The computation of quantities such as the chemical
-potential will a) affect performance and b) require halo values of
-the order parameter (depending on the highest derivative
-in the order parameter present). Performance here should be the
-first consideration, so this needs to be handled with care.
-
-\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}
-
-Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
-given location.
-
-\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
-the number of dimensions. If we are to provide explicit support for
-D2Q9, \texttt{ND} will be 2, otherwise 3.
-
-\section{Implementation Notes}
-
-\subsection{Serial Version}
-
-\subsection{Efficiently Scaling Parallel Computing}
-
-We should be thinking about the 100,000 processor regime, and
-correspondingly large system sizes. To avoid embarrassment in this regime,
-some issues which don't usually arise are relevant, for example:
-
-\begin{enumerate}
-\item
-Large integer quantities such as
-\begin{verbatim}
-   N_total.x*N_total.y*N_total.z
-\end{verbatim}
-are likely to cause integer overflow (if default integers are 4 byte).
-\item
-Local data structures whose size is proportional
-to the number of processors are contraindicated, e.g.,
-anything like
-\begin{verbatim}
-p = (foo *) malloc(nprocs*sizeof(foo));
-\end{verbatim}
-is to be avoided. At the moment, we are in good shape on this
-(apart from some functions in \texttt{misc.c}).
-\item
-I/O is likely become a very important issue, so may need to think
-more about this.
-\item
-We should probably introduce a standard set of benchmarks which can
-be scaled up in some systematic way, so that we can keep track of
-performance over time.
-
-\end{enumerate}
-
-\section{Model Details}
-
-\subsection{D3Q15}
-
-These are the eigenvalues and eigenvectors of teh collision
-matrix used for D3Q15
-
-\begin{table}
-\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - &
- 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
-1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
-3  & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - &
- 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
-3  &$c_{iy}$ \\
-\hline
-$\rho c_{iz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
-3  & $c_{iz}$ \\
-\hline
-$Q_{xx}$ & 1/3 &
--1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
-9/2  & $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{yy}$ & 1/3 &
--1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
-\hline
-$Q_{zz}$ & 1/3 &
--1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
-9  & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
-9  & $c_{iy} c_{iz}$ \\
-\hline
-$Q_{zx}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
-9  & $c_{iz} c_{ix}$ \\
-\hline\hline
-$\chi^1$ & - &
--2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
-1/2 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
-3/2 & $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - &
- 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iy}$\\
-\hline
-$J_{iz}$ & - &
- 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iz}$\\
-\hline
-$\chi^3$ & - &
- 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
-9 & $c_{ix} c_{iy} c_{iz}$ \\
-\hline\hline
-$w_i$ & 1/72 &
-$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
- & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q19}
-
-\begin{table}
-\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
-\hline\hline
-$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
-\hline
-$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
-         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
-& 1\\
-\hline
-$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
-         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
-& 3 \\
-\hline
-$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
-         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
-& 3\\
-\hline
-$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
-         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
-& 3\\
-\hline
-$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
-& 9/2\\
-\hline
-$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
-         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
-& 9\\
-\hline\hline
-$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
-         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 3/4\\
-\hline
-$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
-         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
-         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
-& 3/2\\
-\hline
-$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
-& 9/4\\
-\hline
-$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
-& 9/2\\
-\hline
-$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
-         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 1/2\\
-\hline\hline
-$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
-& \\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{adhikari2005}
-Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
-Fluctuating Lattice Boltzmann,
-\textit{Europhys. Lett.}, \textbf{71}, 473 2005.
-
-\bibitem{ald98}
-Aidun, C.K., Y. Lu, and E.-J. Ding,
-Direct analysis of particulate suspensions with inertia using the
-discrete Boltzmann equation,
-\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
-
-\bibitem{chunladd}
-B. Chun and A.J.C. Ladd,
-Interpolated boundary condition for lattice Boltzmann simulations in
-narrow gaps,
-\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
-
-\bibitem{cj98}
-Cichocki, B., and R.B. Jones,
-Image representation of a spherical particle near a hard wall,
-\textit{Physica A}, \textbf{258}, 273, 1998.
-
-\bibitem{chang}
-C.-H. Chang and E.I. Franses,
-Adsorption dynamics of surfactants at the air/water interface:
-a critical review of mathematical models, data, and mechanisms,
-\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
-
-\bibitem{cpc}
-Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
-LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
-\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
-
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
-\bibitem{diamant96}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid-fluid interfaces,
-\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
-
-\bibitem{eastoe}
-J. Eastoe and J.S. Dalton,
-Dynamic surface tension and adsorption mechanims of surfactants
-at the air-water interface,
-\textit{Advances in Colloid and Interface Science}, \textbf{85}
-13, 2000.
-
-\bibitem{ginzburg}
-I. Ginzburg and D. d'Humi\`eres,
-Multireflection boundary conditions for lattice Boltzmann models,
-\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
-
-\bibitem{h59}
-Hasimoto, H., On the periodic fundamental solutions of the Stokes
-equation and their application to viscous flow past a cubic array
-of spheres.
-\textit{J. Fluid Mech.}, \textbf{5}, 317.
-
-\bibitem{heemels}
-Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
-particles using the lattice-Boltzmann method,
-\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.
-
-\bibitem{jo84}
-Jeffrey, D.J., and Y. Onishi,
-Calculation of the resistance and mobility functions for the two
-unequal rigid spheres in low-Reynolds-number flow,
-\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.
-
-\bibitem{viv}
-Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
-P. Bladon,
-Inertial effects in three dimensional spinodal decomposition of
-a symmetric binary fluid mixture: A lattice Boltzmann study,
-\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).
-
-\bibitem{l94a}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 1. Theoretical foundation,
-\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.
-
-\bibitem{l94b}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 2. Numerical results,
-\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.
-
-\bibitem{l96a}
-Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
-spheres,
-\textit{Phys. Fluids}, \textbf{9}, 491. 1996.
-
-\bibitem{l96b}
-Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
-of non-Brownian spheres,
-\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.
-
-\bibitem{lv01}
-Ladd, A.J.C., and R. Verberg,
-Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
-
-\bibitem{lipanmiller}
-H. Li, C. Pan, and C.T. Miller,
-Pore-scale investigation of viscous coupling effects for two-phase
-flow in porous media,
-\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
-
-\bibitem{nl02}
-Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
-lattice-Boltzmann simulations of particle suspensions,
-\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.
-
-\bibitem{papanastasiou}
-T. paapnastasiou, G. Georgiou, and A. Alexandrou,
-\textit{Viscous Fluid Flow},
-CRC Press, Boca Raton, Florida, 2000.
-
-\bibitem{r95}
-Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
-Cambridge University Press, 1995.
-
-\bibitem{succi}
-S. Succi, \textit{The lattice Boltzmann equation and beyond},
-Oxford University Press, Oxford, 2001.
-
-\bibitem{edo1}
-M. Venuroli and E.S. Boek,
-Two-dimensional lattice-Boltzmann simulations of single phase
-flow in a pseudo two-dimensional micromodel,
-\textit{Physica A}, \textbf{362}, 23, 2006.
-
-\bibitem{vandergraaf}
-R.G.M. van der Sman and S. van der Graaf,
-Diffuse interface model of surfactant adsorption onto flat and
-droplet interfaces,
-\textit{Rheol. Acta} \textbf{46} 3 (2006).
-
-\bibitem{theissengompper}
-O. Theissen and G. Gompper,
-Lattice Boltzmann study of spontaneous emulsification,
-\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
-
-\bibitem{wardtordai}
-A.F.H. Ward and L. Tordai,
-\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
-
-\end{thebibliography}
-
-\end{document}
+This, effectively very small, lattice
+time step is clearly a problem in reaching any respectable 
+macroscopic time in a simulation.
+One step that can be taken is to increase both the lattice
+viscosity and the lattice velocity so that the Reynolds number
+remains unchanged. (Note that the lattice viscosity matching
+the real viscoity might be $(\eta/\rho)T/h^2$ which, using the
+above figures, is $\sim 10^{-4}$.) This can gain 2--3 orders
+of magnitude while keeping the viscosity below $O(1)$. Ultimately,
+it may also be necessary to raise the Reynolds number, which may
+be considered small compared to unity as long as it remains
+in practice less than $O(10^{-1})$ \cite{cates_scaling}.

### 2010-11-29 htdocs/sigma.tex
Check in
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1173 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sigma.tex
%
%  A rather long section on the calibration of surface tension
%  for the binary fluid models.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Calibration of the Surface Tension}

The binary fluid model introduces an interfacial tension $\sigma$ which
can be predicted from the choice of free energy \cite{swift,kendon}.
In particular, Kendon et al. 2000 described the process of calibrating
the actual interfacial tension of the model fluid. This was based on
a measurement of the Laplace pressure difference between the fluid
inside and outside a spherical droplet via
\begin{equation}
\Delta p = 2 \sigma / R
\end{equation}
where $R$ is the radius of the spherical drop at equilibrium. This
measurement appears quite difficult in practice owing to tendency of
any initial drop to evaporate \cite{yue2007}.

\subsection{New measurement}

Here, we prefer a measure based on the expected profile of the
order parameter $\phi$ and the free energy density at a flat interface.
In order to sample all orientations of the interface with respect to
the lattice we again take a large droplet of fluid A initialised at
rest in fluid B. We assume that the interface is locally flat and so
the procedure holds.

If a droplet with radius $r_0$ is constructed with an initial interfacial
profile of $\phi(r) = \tanh{(r - r_0) / \xi_0}$
with $r$ in the radial direction,  then the free energy density will be
(with $-A = B$)
\begin{equation}
 e(r) = -{\scriptstyle \frac{1}{4}} |A| [1 - \mathrm{sech}^4 (r-r_0)/\xi_0]
+ {\scriptstyle \frac{1}{2}}
 (\kappa/\xi_0^2)\, \mathrm{sech}^4{(r-r_0)/\xi_0}.
\end{equation}
Obtaining a fit to $\xi_0$ from the measured interfacial profile of
$\phi (r)$ and to $ e_0 = \kappa / 2 \xi_0^2$ from the peak in the measured
profile of free energy
density allows an estimate of the interfacial tension to be made:
\begin{equation}
\sigma = 2(8/9)^{1/2} \xi_0 |e_0|.
\end{equation}
This measurement is taken after the initial profile has been allowed
to relax for at least 10 times the larger of the diffusion time
$\xi_0^2 / M$ or the momentum diffusion time $\xi_0^2 / \eta$.

We have measured the surface tension using the following approaches.

\textit{Method 1}.

LB for two distributions with relaxation of order parameter
following Kendon etal. The force on the fluid is applied
via the equilibrium stress following Swift et al.

\textit{Method 2}.

As method 1 but using the realxation of order parameter
following Stratford etal.

\textit{Method 3}.

As Method 1 but with force via divergence of the chemical stress.

\textit{Method 4}.

As Method 3 but with force via divergence of the chemical stress.

\textit{Method 5}.

Using finite difference (1st order upwind) for the order parameter.

\textit{Method 6}.

Using finite difference (3rd order upwind).

\subsection{Results}

We follow the parameter sets in Kendon et al Table~2. These
are reproduced here in Table~\ref{tab:r1}. Measured values for
Run028 parameters are show in Table~\ref{tab:newr028}.

\begin{table}
\begin{center}
\begin{tabular}{llllllll}
\hline
Run & $-A, B$ & $\kappa$ & $\eta$ & $M$ & $\sigma$  & $L_0$ & $T_0$ \\
\hline
Run028 & 0.083 & 0.053 & 1.41 & 0.05  & 0.0625 & 31.8 & 718 \\
Run022 & 0.0625 & 0.04 & 0.5 & 0.25   & 0.0471 & 5.31 & 56.3 \\
Run029 & 0.0625 & 0.04 & 0.2 & 0.15   & 0.0471 & 0.849 & 3.61 \\
Run020 & 0.00625 & 0.004 & 0.025 & 2.00 & 0.00471 & 0.133 & 0.704 \\
Run030 & 0.00625 & 0.004 & 0.0065 & 1.25 & 0.00471 & 0.00829 & 0.0124 \\
Run019 & 0.003125 & 0.002 & 0.0014 & 4.00 & 0.00236 & 0.000831 & 0.000493 \\
Run032 & 0.00125  & 0.0008 & 0.0005 & 5.0 & 0.000943 & 0.000265 & 0.000141\\
\hline  
\end{tabular}
\end{center}
\caption{Table showing common parameter values following Kendon et al.
with the theoretical interfacial tension, $L_0$ and $T_0$. Note that
the order parameter mobility is used here (cf $\tilde{M} = 2M$) in
Kendon et al.}
\label{tab:r1}
\end{table}

\begin{table}
\begin{tabular}{lllllll}
\hline
Method & $\xi_0$ & $e_{min}$ & $e_{max}$ & $\sigma$  & $\phi_{min}$
& $\phi_{max}$\\
\hline
\multicolumn{7}{c}{Run028}\\
\hline
Theory   & 1.13 & -0.0207 &  0.0207 & 0.0625 & -1.0 & 1.0 \\
Method 1 & 1.00 & -0.0207 & 0.0128 & 0.063(3) & -1.0091 & 1.0028 \\
Method 2 & 1.00 & -0.0207 & 0.0126 & 0.063(1) & -1.0113 & 0.9993 \\
Method 3 & && & & & \\
Method 4 & && & & & \\
Method 5 & && & & & \\
Method 6 & --& --& --& --& --& --\\
\hline
\multicolumn{7}{c}{Run029}\\
\hline
Theory & 1.13 & -0.0156 & 0.0156 & 0.0471  & -1.0 & 1.0\\
Method 2 & 0.99 & -0.0156 & 0.010(?) & 0.048(4) & -1.0110 & 1.0003\\ 
Method 6 & 1.02 & -0.0156 & 0.0099 & 0.049(0) & -1.0048 & 1.0002\\
\hline
\multicolumn{7}{c}{Run030}\\
\hline
Theory & 1.13 & -0.00156 & 0.00156 & 0.00471 & -1.0 & 1.0\\
Method 2 & 1.04 & -0.00156 & 0.0008(?) & 0.0047(1) & -1.0167 & 1.0063\\
Method 6 & 1.08 & -0.00156 & 0.0007(?) & 0.0048(0) & -1.0028 & 1.0031\\
\hline
\multicolumn{7}{c}{Run019}\\
\hline
Theory & 1.13 & -0.000781 & 0.000781 & 0.00236 & -1.0 & 1.0\\
Method 2 & 1.29(?) & -0.000781 & 0.00015(?) & 0.0022(8) & -1.0312 & 1.0242\\
Method 6 & 1.23(?) & 0.0010(?) &&  0.0024(?) & -1.0085 & 1.0019\\
\hline 
\multicolumn{7}{c}{Run032}\\
\hline
Theory   & 1.13 &  -0.000312 & 0.000312 & 0.000943  & -1.0 &1.0\\
Method 2 & 1.07  & -0.000312& 0.000157 & 0.00094(7) & -1.0511 & 1.0311\\
Method 6 & 1.6(5)& -0.000312 & not st. & 0.0010(7) &-1.0056 & 1.0010\\
\hline
\end{tabular}
\caption{Results for the different methods for Run208 parameters.}
\end{table}

\section{Upwind Advection Schemes}

The solution to the Cahn-Hilliard equation for the order parameter
\begin{equation}
\partial_t \phi + \partial_\alpha (u_\alpha \phi + M\partial_\alpha \mu) = 0
\label{eq:ch}
\end{equation}
assumes that the valocity field $u_\alpha$ is known, along with the
order parameter mobility $M$. Adopting a divergence form ensures that
any finite difference scheme conserves the total order parameter in
the system.

Considering just the advective part of \ref{eq:ch}, a finite difference
approach boils down to finding an interpolation of $\phi(\mathbf{r})$ to
the faces of a control volume surrounding the lattice site at $\mathbf{r}$.
So, in one dimension we have
\begin{equation}
\phi_i^{n+1} = \phi_i^n + \frac{\Delta t}{\Delta x} (u_w \phi_w - u_e \phi_e)
\label{eq:phifd}
\end{equation}
where subscripts $w$ and $e$ refer to compass directions with index $i$
incresing eastward. A first order upwind scheme approximates the
interfacial value $\phi_w$ depending on the direction of the velocity
at the face, viz:
\begin{equation}
\phi_w = \left\{
\begin{array}{ll} \phi_i & u_w < 0, \\ \phi_{i-1} & u_w >= 0.
\end{array} \right.
\end{equation}
This choice is conditionally stable for the Euler forward time step
of Eq.~(\ref{eq:phifd}), but highly dissipative.

Better accuracy requires
a higher-order approximation to the interfacial value of $\phi$.

\subsection{Uniformly third order scheme}

We follow Leonard et al. \cite{utopia} in adopting a uniformly third
order approximation dependent upon both the normal and tangential
interfacial Courant numbers. It is extended here to three dimensions.

Consider the two dimensional problem of Figure \ref{fig:utopia}.

The flux at face $W$ is then
\[\begin{array}{ll}
c_x \Big\{ &
1/2(\phi_c + \phi_w) - 1/2c_x (\phi_c - \phi_w)
 - 1/6 (1 -c_x^2)(\phi_c - 2\phi_w + \phi_{ww})\\
& - 1/2c_y (\phi_w - \phi_{sw})\\
& - c_y (1/4 - 1/3c_x)(\phi_c - \phi_w - \phi_s + \phi_{sw})
 - c_y (1/4 - 1/6c_y)(\phi_{nw} - 2\phi_w + \phi_{sw})\\
& - 1/2c_z (\phi_w - \phi_{wd}) \\
& - c_z (1/4 - 1/3c_x) (\phi_c - \phi_w -\phi_{cd} + \phi_{wd})
 - c_z (1/4 - 1/6 c_y) (\phi_{wu} - 2 \phi_w + \phi_{wd}) \\
& + c_y c_z \big[
1/3(\phi_w - \phi_{wd}) - 1/3 (\phi_{sw} - \phi_{swd}) \\
& + (1/6 - 1/4c_x) (\phi_c - \phi_w -\phi_s + \phi_{sw}
    -(\phi_{cd} - \phi_{wd} - \phi_{sd} + \phi_{swd}))\\
& + (1/6 - 1/8c_y)(\phi_{nw} - 2 \phi_w + \phi_{sw} 
                - (\phi_{nwd} - 2\phi_{wd} + \phi_{swd}))\\
& + (1/6 - 1/8c_z)(\phi_{wu} - 2\phi_w + \phi_{wd}
                 - (\phi_{swu} - 2\phi_{sw} + \phi_{swd}))\big] \Big\}
\end{array}\]

\subsection{Gory details}

The flux-integral method of Leonard et al. assumes that within a
given cell, the order parameter is piecewise quadratic. So, for
cell $W$ we have
\begin{eqnarray}
\phi (\zeta, \eta, \theta) = \phi_W - {\scriptstyle \frac{1}{24}} 
(\phi_C + \phi_{SW} + \phi_{WW} + \phi_{NW} + \phi_{WD} + \phi_{WU}
- 6\phi_W)\\
+ {\scriptstyle \frac{1}{2}} (\phi_C - \phi_{WW}) \zeta
+ {\scriptstyle \frac{1}{2}} (\phi_C - 2\phi_{W} + \phi_{WW}) \zeta^2\\
+ {\scriptstyle \frac{1}{2}} (\phi_{NW} - \phi_{SW}) \eta
+ {\scriptstyle \frac{1}{2}} (\phi_{NW} - 2\phi_{W} + \phi_{SW}) \eta^2\\
+ {\scriptstyle \frac{1}{2}} (\phi_{WU} - \phi_{WD}) \theta
+ {\scriptstyle \frac{1}{2}} (\phi_{WU} - 2\phi_{W} + \phi_{WD}) \theta^2
\end{eqnarray}

\section{Lattice kinetic equation viewed as finite difference}

For binary fluid problems the second distribution $g_i (\mathbf{x};t)$,
representing the composition, obeys the evolution equation
\begin{equation}
g_i (\mathbf{x}; t + \Delta t )
= g_i^\star (\mathbf{x} - \mathbf{c}_i \Delta t; t)
\label{lke}
\end{equation}
where the star indicates the post-collision distribution. Here we consider
the distribution to be set by requiring the first moment
\begin{equation}
j_\alpha = \sum_i g_i^\star (\mathbf{x};t) c_{i\alpha} = \phi u_\alpha
\label{eq:moment1}
\end{equation}
and the second moment as
\begin{equation}
\Phi_{\alpha\beta} = \sum_i g_i^\star (\mathbf{x}; t) c_{i\alpha} c_{i\beta}
= \phi u_\alpha u_\beta + 2M\mu (\mathbf{x};t) \delta_{\alpha\beta}.
\end{equation}
Here the mobility $M$ enters through the terms related to the chemical
potential $\mu (\mathbf{x};t)$. The distribution is then set using the
reprojection
\begin{equation}
g_i^\star (\mathbf{x}; t) = \delta_{i0}\phi
+ w_i (j_\alpha u_{i\alpha} / c_s^2
+      \Phi_{\alpha\beta} Q_{i\alpha\beta} / 2c_s^4),
\end{equation}
where the $\delta_{i0}$ moves $\phi$ mostly into the non-propagating
distribution.

For a uniform velocity field $u_\alpha = (u_x, u_y, u_z)$ it is possible
to expand \ref{eq:lke} in a finite difference form which includes three
parts: the advective part in $\phi$ related to the first moment, a
dissipative contribution in $\phi$ related to the $\phi u_\alpha u_\beta$
term in the second moment, and a part related to the
 diffusion of the chemical potential.

\subsection{Advective terms}

Combining \ref{eq:lke} and \ref{eq:moment1} and taking the sum over
the distributions we have
\begin{equation}
\sum_i g_i (\mathbf{x};t) = \sum_i \Big( g_0^\star (\mathbf{x};t)  +
(1/c_s^2)  w_i \phi (\mathbf{x} - \mathbf{c}_i \Delta t;t)
u_\alpha c_{i\alpha} \Big).
\end{equation}
Introducing a finite finiterence notation $\phi_{ijk}^n
= \sum_{i'} g_{i'}^\star (\mathbf{x}; t)$ where the indices $i,j,k$
represent the spatial discretistation, and the superscript $n$ represents
the discrete time level we have
\begin{eqnarray}
\phi_{ijk}^{n+1}  = \phi_{ijk}^n - u_x w_1/c_s^2 \Big\{
\phi_{i+1 j k}^n - \phi_{i-1 j k}^n \\
+ 1/2(
\phi_{i+1 j+1 k}^n - \phi_{i-1 j-1 k}^n +
\phi_{i+1 j-1 k}^n - \phi_{i-1 j+1 k}^n +
\phi_{i+1 j k+1}^n - \phi_{i-1 j k-1}^n +
\phi_{i+1 j k-1}^n - \phi_{i-1 j k-+}^n)
\Big\} \\ +
u_y w_1 / c_s^2 \Big\{
\phi_{i j+1 k}^n - \phi_{i j-1 k}^n \\
+ 1/2 (
\phi_{i+1 j+1 k}^n - \phi_{i-1 j-1 k}^n +
\phi_{i-1 j+1 k}^n - \phi_{i+1 j-1 k}^n +
\phi_{i j+1 k+1}^n - \phi_{i j-1 k-1}^n +
\phi_{i j+1 k-1}^n - \phi_{i j-1 k+1}^n)
\Big\} \\ +
u_z w_1 / c_s^2 \Big\{
\phi_{i j k+1}^n - \phi_{i j k-1}^n \\
+ 1/2 (
\phi_{i+1 j k+1}^n - \phi_{i-1 j k-1}^n +
\phi_{i-1 j k+1}^n - \phi_{i+1 j k-1}^n +
\phi_{i j+1 k+1}^n - \phi_{i j-1 k-1}^n +
\phi_{i j-1 k+1}^n - \phi_{i j+1 k-1}^n) \Big\}.
\end{eqnarray}
We can regcognise here second order centred difference approximations to
the first derivative in $\phi$ in the coordinate direction in the
diagonal directions (cf \cite{sescu2008} with 'isotropy correction factor'
$\beta = 1/2$ in three dimensions).

There are a number of points to note about this. First, there are
terms in $\sum_i g_0^\star$ which are not included above, but are dealt
with below. Second, while the spatial discretisation looks ok, this is
forward in time, which would classically be undesirable. Dissipation
is required to prevent dispersive errors dominating. (Although with
general flow field, the coefficients of the finite difference form are
altered in ways which could provide some upwind bias.)

\subsection{Dissipative terms}

The dissipative terms from the $\phi u_\alpha u_\beta$ term in the
second moment give
\begin{equation}
\sum_{i'} g_{i'} (\mathbf{x}; t + \Delta t) =
1/2c_s^4 \sum_{i'} w_i \phi (\mathbf{x} - \mathbf{c}_i \Delta t; t)
u_\alpha u_\beta Q_{i'\alpha\beta}.
\end{equation}
With a little effort this may be expanded to give and evolution
equation with the following difusive tendency terms:
\begin{eqnarray}
w_1/2c_s^4  \Big\{
(u_x^2 - c_s^2u^2)(\phi_{i+1 j k} - 2\phi_{ijk} + \phi_{i-1 j k}) \\+
(u_y^2 - c_s^2u^2)(\phi_{i j+1 k} - 2\phi_{ijk} + \phi_{i j-1 k}) \\+
(u_z^2 - c_s^2u^2)(\phi_{i j k+1} - 2\phi_{ijk} + \phi_{i j k-1}) \Big\}\\
+ w_2/2c_s^4 \Big\{
\big[ (u_x^2 + u_y^2) - c_s^2u^2\big]
(\phi_{i+1 j+1 k} - 2\phi_{ijk} + \phi_{i-1 j-1 k}) \\+
\big[ (u_x^2 - u_y^2) - c_s^2u^2 \big]
(\phi_{i+1 j-1 k} - 2\phi_{ijk} + \phi_{i-1 j+1 k}) \\+
\big[ (u_x^2 + u_z^2) - c_s^2u^2 \big]
(\phi_{i-1 j k-1} - 2\phi_{ijk} + \phi_{i+1 j k+1}) \\+
\big[ (u_x^2 - u_z^2) - c_s^2u^2 \big]
(\phi_{i+1 j k-1} - 2\phi_{ijk} + \phi_{i-1 j k+1}) \\+
\big[ (u_y^2 + u_z^2) - c_s^2u^2 \big]
(\phi_{i j-1 k-1} -2\phi_{ijk} + \phi_{i j+1 k+1}) \\+
\big[ (u_y^2 - u_z^2) - c_s^2u^2 \big]
(\phi_{i j+1 k-1} - 2\phi_{ijk} + \phi_{i j-1 k+1}) \Big\}.
\end{eqnarray}
It can be see that the coefficients of these diffusive terms is
related to the aspect ratio of the flow, and that the coefficents
of given terms could become negative.

\subsection{Chemical potential term}

Finally, the terms in the chemical potential give
\begin{equation}
\sum_{i'} g(\mathbf{x};t + \Delta t) =
\sum_{i'} w_i 2M\mu(\mathbf{x} - \mathbf{c}_ \Delta t)
\delta_{\alpha\beta} Q_{i\alpha\beta}.
\end{equation}
Noting that the contraction $\delta_{\alpha\beta} Q_{i\alpha\beta} =
-1, 0, 1$ for velocities with weights $w_0, w_1, w_2$ respectively,
we get
\begin{eqnarray}
\phi_{ijk}^{n+1} = \phi_{ijk}^n -  w_0 2M u_{ijk}/2c_s^4
+ 2M w_2/2c_s^4 \Big\{
\mu_{i+1 j+1 k} + \mu_{i-1 j-1 k} + \mu_{i+1 j-1 k} + \mu_{i-1 j+1 k} + \\
\mu_{i+1 j k+1} + \mu_{i-1 j k-1} + \mu_{i+1 j k-1} + \mu_{i-1 j k+1} +
\mu_{i j+1 k+1} + \mu_{i j-1 k-1} + \mu_{i j+1 k-1} + \mu_{i j-1 k+1} \Big\}.
\end{eqnarray}
This can be combined to give an easily recognisable finite difference
solution for the diffusion equation
\begin{eqnarray}
\phi_{ijk}^{n+1} = \phi_{ijk}^n + 2M w_2 / 2 c_s^4
\Big\{
\mu_{i+1 j+1 k} - 2\mu_{ijk} + \mu_{i-1 j-1 k} + 
\mu_{i+1 j-1 k} - 2\mu_{ijk} + \mu_{i-1 j+1 k} + \\
\mu_{i+1 j k+1} - 2\mu_{ijk} + \mu_{i-1 j k-1} +
\mu_{i+1 j k-1} - 2\mu_{ijk} + \mu_{i-1 j k+1} + \\
\mu_{i j+1 k+1} - 2\mu_{ijk} + \mu_{i j-1 k-1} + 
\mu_{i j+1 k-1} - 2\mu_{ijk} + \mu_{i j-1 k+1} \Big\}.
\end{eqnarray}
Note that this uses the twelve stencil points in $\mu$ related to
the velocities with weight $w_2$ and so should have very good isotropy
properties. The final prefactor is $M$ (check).

### 2010-12-07 trunk/htdocs/user.tex
Added the liquid crystal activity constant zeta
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1159 40876441-3496-0410-a613-a2fca35b2ae2
Updated intructions to replace CVS by SVN
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1181 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: user.tex,v 1.1 2010-10-21 18:05:50 kevin Exp $
+%  $Id: user.tex,v 1.2 2010-11-02 17:38:48 jlintuvu Exp $
 %
@@
 
-We have a NeSCForge project account available at
+Under the auspices of the CCPs (Collaborative Computational Projects),
+and in particular CCP5, we have a CCPForge project account at
 
-\texttt{http://forge.nesc.ac.uk/projects/ludwig/}
+\texttt{http://ccpforge.cse.rl.ac.uk/}
 
-This provides controlled access to known persons and provide CVS, bug
-tracking, and so on. The NeSCForge CVS is not available to the public.
+This provides controlled access to known persons and uses version
+control via SVN, and includes bug
+tracking, and so on. The CCPForge SVN is not available to the public
+at the present time. You need to create an account at CCPForge, and
+let me or Juho know the username so you can be added to the access list.
 
-\textbf{IMPORTANT:} NeSCForge is scheduled to go out of service on 20th
-December 2010. KS will announce an alternative as soon as possible.
-This may mean some disruption.
+To download the entire code, you need
+\begin{verbatim}
+svn checkout --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig
+\end{verbatim}
+where \texttt{<user>} is replaced by your CCPForge user name.
+Note that this will download all the existing tags and branches
+into a new directory named \texttt{ludwig}, as well as the main code!
+On success, you should see three directories
+\begin{verbatim}
+$ ls ludwig
+branches  tags      trunk
+\end{verbatim}
+The main code is in the \texttt{trunk} subdirectory. This documentation is
+found in the file \texttt{trunk/htdocs/ludwig.tex}, unit tests are in
+the directory \texttt{trunk/tests}, and the utilities are in
+\texttt{trunk/util}, and so on.
 
-To download, e.g., the \texttt{src} module, you need
+If you do not want all the tags and branches (which take up a certain amount
+of space), you can specify that you
+want only the trunk, e.g.,
 \begin{verbatim}
-cvs -z3 -d:ext:developername@forge.nesc.ac.uk:/cvsroot/ludwig co src
+svn co --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk
 \end{verbatim}
-where \texttt{developername} is replaced by your Nescforge user name.
-(Note that this requires secure shell; you may need to set the
-environment variable \texttt{CVS\_RSH=ssh} if you do not appear
-to be getting any response.)
+in which case you will get a new directory called \texttt{trunk} containing
+the source files. All the instructions below refer to paths relative to
+this trunk directory.
 
-Currently, you need to download at least the cvs modules \texttt{src},
-and \texttt{mpi\_s}. The \texttt{mpi\_s} module contains a `stub'
-library to replace MPI in serial, and is required to run in serial.
-This documentation is in the module
-\texttt{htdocs}, unit tests are in \texttt{tests}, and the utilities
-are in the module \texttt{util}.
+If you have problems, it may be worth while looking at the issue
+tracker on the project web page. It would be useful if all issues
+could be reported via this mechanism. If you e-mail me, I will just
+insert your e-mail into the issue tracker!
 
@@
 
+\inputkey{lc\_active\_zeta} Apolar liquid crystal activity constant.
+
 \subsubsection{Initial LC order parameter $Q_{\alpha\beta}$}

### 2010-12-09 trunk/htdocs/user.tex
Keywords again
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1183 40876441-3496-0410-a613-a2fca35b2ae2
Try svn propset svn:keywords "Id HeadURL" user.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1184 40876441-3496-0410-a613-a2fca35b2ae2
Back to just Id
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1186 40876441-3496-0410-a613-a2fca35b2ae2
Info on svnversion
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1194 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: user.tex,v 1.2 2010-11-02 17:38:48 jlintuvu Exp $
+%  $Id$
 %
@@
 
+The other thing required by the Makefile is the utility
+\texttt{svnversion}, which allows the compile-time SVN version number to
+be included in the executable (and hence the output of the program). This
+is useful to remember which version you were using when looking at
+results at a later date. If
+\texttt{svnversion} is not available (and it
+should be if you have downloaded the code via SVN), you may get an
+error message, but the code should still compile.
+
 \subsubsection{Serial}
@@
 $ make serial
+gcc  -D SVN_REVISION='"'`svnversion`'"' -c svn.c
 gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d2q9.c

### 2011-01-14 branches/CUDA/htdocs/sections.tex
adding branch for CUDA development code
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1221 40876441-3496-0410-a613-a2fca35b2ae2

--- 2010-04-30 trunk/htdocs/sections.tex
@@
 %
-%  $Id: sections.tex,v 1.14 2010-04-30 11:00:50 kevin Exp $
+%  $Id: sections.tex,v 1.15 2010-10-21 18:05:50 kevin Exp $
 %
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-\subsection{Note to Readers}
-
-This document is intended as a starting point for a full-blown
-specification of the next version of the Ludwig code, and everything
-to do with it. It is (manifestly) not complete.
-
-\subsection{Some History}
-
-Early Lattice Gas and Lattice Boltzmann. Early Lattice Boltzmann at Edinburgh.
-Ludwig.
-
-\subsection{Aims}
-
-We aim to provide a robust and portable code, written in C, which
-can be used to perform serial and scalable parallel simulations of
-complex fluid systems based around the Lattice Boltzmann method.
-
-\subsection{Availability}
-
-We now have a NeSCForge project account available at
-
-\texttt{http://forge.nesc.ac.uk/projects/ludwig/}
-
-This provides controlled access to known
-persons and provide CVS, bug tracking, and so on. It is
-\textbf{essential} that people sign up to this, as we must
-move to a single supported version.
-
-The NeSCForge CVS is not available to the public.
-
-\subsection{Download}
-
-\section{User Input}
-
-By default the executable will look for an input file at run time.
-By default, this file is named \texttt{input} and should be in the
-current working directory. The alternative is to specify the name
-of the file explicitly on the command line:
-\begin{verbatim}
-$ ./Ludwig.exe input_file
-\end{verbatim}
-
-The contents of the input file are made up of \textit{key value}
-pairs with control the run-time behaviour. Blank lines and lines
-begining with \# are ignored as comments. The following describes
-the effect of various keys. Most keys have a default value which
-will be used by the code if the corresponding key is not present
-(or commented out) in the input file.
-
-\subsection{Basic run parameters}
-
-\inputkey{N\_cycles}
-
-The number of lattice Boltzmann time steps to execute. Default
-value:
-
-\texttt{N\_cycles 0}
-
-\inputkey{t\_start}
-
-By default the code will start from time step $t = 0$. If you 
-wish to restart from a previously saved configuration, set
-the appropriate value for \texttt{t\_start}. The code will
-execute \texttt{N\_cycles} steps starting from this point.
-
-\inputkey{input\_config}
-
-The file name stub for input configuration required for restarts.
-For example, to restart from the file config.out100 previously
-produced at the end of a 100 step run use
-\begin{verbatim}
-t_start 100
-input_config config.out100
-\end{verbatim}
-
-If you are using parallel I/O, the filename stub is required; the
-file extensions are computed automatically.
-
-\inputkey{size}
-
-Controls the number of lattice points in the $(x, y, z)$ directions,
-respectively. Default value:
-
-\texttt{size 64\_64\_64}
-
-\inputkey{grid}
-
-In parallel, this sets the processor Cartesian decomposition, i.e.,
-the number of processors in each dimension of the Cartesian communicator.
-The default value may be implementation dependent as it is that returned
-by \texttt{MPI\_Dims\_create()}. For
-example:
-\begin{verbatim}
-size 64_64_64
-grid 4_2_1
-\end{verbatim}
-gives a total (physical) system size of 64 lattice sites in each
-vdirection, decomposed across 4 processors in the x-direction, 2 in
-the y-direction and 1 (no decomposition) in the z-direction. The
-total number of processors must therefore be 8. The local domain
-size per processor is then 16x32x64.
-
-\subsection{Fluid properties}
-
-\inputkey{viscosity}
-
-Sets the LB fluid shear viscosity $\eta$ (and related relaxation times).
-Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
-(i.e., realaxation time equal to unity).
-
-\inputkey{viscosity\_bulk}
-
-Sets the LB bulk viscosity $\zeta$ (and related relaxation times).
-This can be use if you are concerned with incompressiblity
-violations, in which case $\zeta >> \eta$ can be useful. The
-default value is $\zeta = \eta$.
-
-\inputkey{temperature}
-
-The LB 'temperature' using fluctuating hydrodynamics. Safe values
-are $0.0001 > kT > 0$.
-
-\inputkey{rho0}
-
-Sets the mean density of the fluid. The default value is unity, and
-it is recommended that the value should not be changed.
-
-\subsection{Binary fluid parameters}
-
-The are currently three choices of free energy relating to
-various systems.
-
-\subsubsection{Symmetric binary fluid}
-
-\inputkey{free\_energy symmetric}
-
-The following applies for the binary fluid problem with
-compositional order parameter $\phi$ and free energy
-(excluding the term in the density $\rho$):
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
-\end{equation}
-
-This is described in some detail
-by Kendon et al \cite{viv}. The first two terms represent the bulk
-contribution, whereas the term in $\kappa$ penalisies curvature in
-the interface.
-
-\inputkey{A}
-
-The parameter $A$. Note $A < 0$.
-
-\inputkey{B}
-
-The parameter $B$. Note $B = -A$ for common usage, although this is
-not enforced.
-
-\inputkey{K}
-
-The parameter $\kappa$, which is positive.
-
-\subsubsection{Brazovskii free energy}
-
-This is similar to the symetric, but with one extra term which
-requires description.
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
-\end{equation}
-
-\subsubsection{Surfactant model free energy}
-
-\inputkey{free\_energy surfactant}
-
-This is for use with a two order parameter model the first of
-which is the composition, as for the symmetric free energy,
-while the second $\psi$ represents surfactant concentration.
-The free energy density is made up of a number of terms:
-
-\begin{equation}
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-\end{equation}
-is the standard symmetric term related to the composition as above;
-
-\begin{equation}
-D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
-\end{equation}
-represents the energy of surfactant in the bluk phase with surfactant
-concentration $0 < \psi < 1$ with a single parameter $D$;
-\begin{equation}
--{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
--{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
-\end{equation}
-where the term in $\epsilon$ represents the energy reduction by
-adsorbing surfactant at the interface and the term in $\beta$
-provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
-\begin{equation}
-+{\textstyle \frac{1}{2}} W \psi \phi^2
-\end{equation}
-which is included to stabilise the interface \cite{theissengompper}.
-
-The corresponding keys for the input file are:
-
-\inputkey{D}
-
-is the bulk surfactant contribution parameter
-
-\inputkey{epsilon}
-
-is the parameter $\epsilon$
-
-\inputkey{W}
-
-is the parameter $W$
-
-\inputkey{beta}
-
-is the value of the parameter $\beta$ (not implemented yet).
-
-\inputkey{phi\_b}
-
-the initial, uniform, background surfactant concentration $\psi_b$.
-
-\subsection{Other Binary Fluid Parameters}
-
-\inputkey{mobility}
-
-Sets the compositional order parameter mobility $M$ (uniform in space)
-which appears in the Cahn-Hilliard equation.
-
-\inputkey{mobilty\_psi}
-
-Sets the (uniform) mobility for the surfactant as appropriate for
-the Cahn-Hilliard equation.
-
-\inputkey{phi0}
-
-The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
-The default value is zero, i.e., a symmetric 50:50 mixture by volume.
-
-\inputkey{phi\_initialisation}
-
-Determines how the compositional order parameter is initialisated
-at the start of the run. If set to \texttt{spinodal} the value
-is set to $\phi_0$ as set above plus or minus a random noise, the
-magnitude of which is set by the value of the \texttt{noise} key.
-If set to \texttt{block} a one-dimensional profile is set up in
-the $z$-direction representing two blocks of fluid with
-$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
-$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
-appropriate width. The $\phi = -1$ section is in the middle.
-
-\inputkey{noise}
-
-The magnitude of the initial fluctuations in $\phi$ used to
-initiate spinodal decompsotion.
-
-\subsection{Colloidal particles}
-
-There are quite a few options here.
-
-\inputkey{colloid\_init}
-
-This determines how the code initialises particles (or not). The
-default is to have no particles (value \texttt{no\_colloids}). There
-are two other options at the moment.
-
-\texttt{colloid\_init    fixed\_number\_monodisperse}
-
-initialises a fixed number of particles with randomly selected
-positions. Note that there is no gaurentee that the particles
-do not overlap. The number of particles is determined by
-the value of the key \texttt{colloid\_no}.
-
-\texttt{colloid\_init   fixed\_volume\_fraction\_monodisperse}
-
-initialises a given volume fraction of particles on a (roughly)
-bcc lattice. The particles will not overlap. The volume fraction
-is set using the key \texttt{colloid\_vf}.
-
-\inputkey{colloid\_a0}
-
-Sets the nominal radius of the particle on the lattice $a_0$. This is
-used to construct the links for BBL. The smallest acceptable LB
-particle is $a_0 = 1.25$ \cite{nl02}. There are 'magic' values
-including 1.25, 2.3, 3.7, 4.7 which minimise discretisation effects.
-
-\inputkey{colloid\_ah}
-
-Sets the hydrodynamic radius of the particles $a_h$. For all
-reasonable applications choose $a_h = a_0$.
-
-\inputkey{colloid\_mc\_steps}
-
-It is possible to have the code follow the initialisation of
-particle positions by a number of Monte Carlo steps. This can
-help to remove overlaps (fixed\_number\_monodisperse) or
-melt the lattice (fixed\_volume\_fraction\_monodisperse).
-
-\inputkey{colloid\_gravity}
-
-Sets a body force on each paarticle. Useful, for example, for
-sedimentation. For example,
-
-\inputkey{cell\_list\_min}
-
-This sets the minimum allowable width of the cell list used to
-store the particles. This should be as small as possible for
-efficiency purposes but must be large enough to capture all
-possible interactions.
-
-\texttt{colloid\_gravity  0.0\_0.0\_-0.0001}
-
-gives a constant force in the negative $z$-direction on each particle.
-The code automatically computes the compensating body force on fluid nodes
-required to give no net change in the total momentum of the system at
-each time step.
-
-\subsection{Active particles}
-
-Not in general use yet.
-
-\subsection{Particle-particle interactions}
-
-\subsection{Boundary conditions}
-
-Solid planar boundary walls can be added at the extremities of the
-system (i.e., the volume of fluid is extactly that specified by
-the \texttt{size} keyword). The wall can be in one, two or all three
-coordinate direcctions, i.e., a 'slab' geometry, and 'duct' geometry
-or a 'box' geometry, respectively.
-
-The boundary walls are switched on via the keyword value pair
-
-\inputkey{boundary\_walls\_on [no | yes]}
-
-If \texttt{yes} then the geometry is specified with the keyword
-as one of
-
-\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}
-
-Note that these are the only valid selections. \texttt{1\_1\_0}
-gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction. \texttt{1\_0\_0}
-gives periodic conditions in the $x-$ direction only, while
-\texttt{0\_0\_0} is a fully enclosed box.
-
-In the case of the slab geometry, two further values are used so
-that shear may be imparted to the system:
-
-\inputkey{boundary\_speed\_bottom}
-
-\inputkey{boundary\_speed\_top}
-
-These values set the $x-$component of the wall velocity at
-$z = z_{min}$ and $z = z_{max}$, respectively. Like all
-velocities, these values must respect the low Mach number
-constraint $u_x < c_s$.
-
-\subsection{Input/Output and Parallel I/O}
-
-\inputkey{freq\_statistics}
-
-sets the frequency for reporting general information from
-the code. (In general, the function
- \texttt{is\_statistics\_step()} returns true at this frequency.
-
-\inputkey{freq\_measure}
-
-sets the frequency of recording of colloid/order parameter output.
-(In general, the function \texttt{is\_measurement\_step()} returns
-true at this frequency.)
-
-\inputkey{freq\_config}
-
-set frequency of full configuration (checkpoint) dumps suitable
-for restart.
-
-\inputkey{freq\_phi}
-
-set frequency of order parameter output.
-
-\inputkey{phi\_format}
-
-either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-
-\inputkey{freq\_vel}
-
-set frequancy of velocity output.
-
-\inputkey{vel\_format}
-
-either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-
-\inputkey{config\_at\_end}
-
-enforces full configuration dump at the end of execution. The
-default is 'yes'.
-
-\subsection{Lees-Edwards planes}
-
-\inputkey{N\_LE\_plane}
-
-sets the total number of Lees-Edwards planes. Default is zero.
-The placing is as followings. The number of planes $n$ must
-divide the lattice size in the $x$-direction to give an integer
-$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
-
-\inputkey{LE\_plane\_vel}
-
-sets the velocity of each plane relative to the lattice. All planes
-have the same, constant, velocity.
-
-\inputkey{LE\_init\_profile}
-
-if set to 1, the fluid velocity is initialise to reflect a steady
-state shear flow appropriate for the number of planes at the
-given velocity. If set to zero, the fluid is initialised with
-zero velocity.
-
-\subsection{Porous Media}
-
-Porous media calculations can be undertaken when appropriate
-solid/fluid status information is supplied. The are a number
-of switches available in the input file:
-
-\inputkey{porous\_media\_file}
-
-specifies the file stub name to be read at the start of execution.
-(If the stub name is \texttt{file} then the code will expect to
-find \texttt{file.001-001} in the current directory.
-
-\inputkey{porous\_media\_format}
-
-is either \texttt{ASCII} or \texttt{BINARY} as appropriate. Note that
-in parallel, a single data file can be supplied, but it must be binary.
-The default is \texttt{BINARY}.
-
-\inputkey{porous\_media\_type}
-
-is either \texttt{status\_only} (the default) or \texttt{status\_with\_h}.
-
-\subsubsection{File format}
-
-In all cases the status (fluid/solid) information is represented by a
-single \texttt{char} (or integer), which must be supplied via the
-porous media file. A single file should contain data matching the
-current system size, and have the $z-$direction running fastest,
-followed by the $y-$direction. Note that in non-periodic directions,
-the structure must be 'closed', i.e., all the points at the edge
-should be solid.
-
-Fluid sites are designated by \texttt{0} and boundary or solid sites
-by \texttt{1}. These data should be of type \texttt{char} in binary,
-and may be integer in ASCII.
-
-Where wetting information is required, the free energy parameter $H$
-can be supplied by using the \texttt{status\_with\_h} switch. In this
-case, the \texttt{char} status is augmented by a single \texttt{double}
-value which is the local value of $H$. The order is then
-$s_1,h_1, s_2, h_2, \ldots$.
-
-\subsubsection{Permeability calculations}
-
-Single fluid permeability calculations for a given proous structure
-can be undertaken by driving a flow via the fluid body force. Note
-that the structure must be periodic in the direction of the force
-(this may mean duplicating a 'reflected' version of a given sample
-to create the correct input). The body force can be specified so
-that, e.g., to drive a flow in the positive $x-$direction
-
-\texttt{force 0.001\_0.000\_0.000}
-
-The force should not be so large that the maximum velocity generated
-threatens the Mach number constraint. To get a measurement of the
-flow at equilibrium, the calculation should be run at least the
-momentum diffusion time for the system ($L^2/\eta$ in LB time steps).
-
-The net flow can be measured by combining the statistics for the
-total momemtum and the total density (which is equal to the volume
-with $\rho_0 = 1$).
-
-Note that in the case of porous media with narrow channels at the grid
-scale, the wall velocity can be dependent on the viscosity of the fluid.
-This is an artefact of the bounce-back on links and will result in a
-viscosity-dependent permeability (see, e.g., \cite{lipanmiller}). To minimise
-this effect, it is recommended to switch on the correction to the ghost
-mode relaxation time proposed by Ginzburg and d'Humi\`eres \cite{ginzburg}
-which corrects the viscosity-dependence of the apparent boundary
-position. This is done via the input switches
-
-\texttt{ghost\_modes on\\
-ginzburg\_dhumieres on}
-
-and should recover a viscosity-independent permeability measurement
-\cite{lipanmiller,chunladd}.
-
-\subsubsection{Example: circular capillaries}
-
-The utility \texttt{capillary.c} can be used to generate various
-porous media input files corresponding to various simple circular
-or square capillaries.
-
-For an infinite  capillary of circular cross section radius $a$,
-the conductance is known analytically\cite{papanastasiou}. The
-flow per unit area $J$ is given by
-\begin{equation}
-J = - \frac{1}{8\eta} \frac{\partial p}{ \partial x} a^2. 
-\end{equation}
-If the pressure gradient is replaced by a uniform body force
-$-\partial p/ \partial x = \rho g$, then one can define a
-viscosity-independent conductance $C$ via $ J = C \rho g / \eta$,
-i.e., $C = a^2/8$. So, by measuring the rate of change of flow
-with applied body force, one can compute an estimate of the conductance
-to compare with this result.
-
-If the total momentum is measured from the code for a given force
-(in say the $x$-direction), then the flux per unit area is
-$J = \sum_i \rho v_x / \sum_i \rho = \sum_i \rho v_x / V$
-where the sum is over all fluid lattice sites $i$, with $V$
-the fluid volume. The rate of change of flux with applied force
-$\Delta J / \Delta\rho g$ can then be combined with the viscosity
-to give the conductance.
-
-\subsubsection{Example: rectangular capillaries}
-
-An analtyical expression is also available for the conductance
-of a square or rectugular capillary. For an infinite capillary
-of square cross section width $w \times h$ (where $h$ is the
-longer),
-the volume flux per unit area $J$ is expected to be
-\cite{papanastasiou,edo1}
-\begin{equation}
-J = - \frac{1}{3\eta} \frac{\partial p}{\partial x} (h/2)^2
-\left[
-1 - 6(h/w) \sum_{k=1}^{\infty} \frac{\tanh(\alpha_k w/h)}{\alpha_k^5} 
-\right]
-\end{equation}
-where $\alpha_k = (2k - 1)\pi/2$. The pressure gradient
-$-\partial p / \partial x$ is replaced by the body force $\rho g$ and
-one can difine the viscosity-independnet conductance $C$ via
-$J = C\rho g / \eta$. The calculation procedes as above.
-
-\subsubsection{Matching lattice and real units in porous media}
-
-Following Succi \cite{succi} (Chapter~8) the following argument
-can be made to match lattice quantities and real physical
-quantities for a given system of interest. Suppose we have
-a structure with characteristic pore size $h = 1$ micron
-and characteristic sample size of 1 cm. If the fluid
-(e.g., water) has viscosity $\eta = 10^{-3}$ Pa s and density
-$\rho = 10^3$ kg~s$^{-1}$, and a typical flow is
-$U \sim 10^{-3}$ m~s$^{-1}$, then the pore scale Reynolds
-number is $Re_h \sim 10^{-3}$.
-
-If the speed of sound in water is 1500~m~s$^{-1}$,
-then the corresponding time scale related to the
-speed of sound and the pore size is $T = h/ c_s$,
-which is $\sim 10^{-10}-10^{-9}$~s. (Exact matching
-is acheived via the dimensionless group $\Delta x / c_s\Delta t$,
-with the lattice speed of sound $1/\sqrt{3}$).
-
-This, effectively very small, lattice
-time step is clearly a problem in reaching any respectable 
-macroscopic time in a simulation.
-One step that can be taken is to increase both the lattice
-viscosity and the lattice velocity so that the Reynolds number
-remains unchanged. (Note that the lattice viscosity matching
-the real viscoity might be $(\eta/\rho)T/h^2$ which, using the
-above figures, is $\sim 10^{-4}$.) This can gain 2--3 orders
-of magnitude while keeping the viscosity below $O(1)$. Ultimately,
-it may also be necessary to raise the Reynolds number, which may
-be considered small compared to unity as long as it remains
-in practice less than $O(10^{-1})$ \cite{cates_scaling}.
-
-\subsection{Miscellaneous}
-
-\inputkey{random\_seed}
-
-sets the random number generator seed.
-
-\subsection{Dealing with the Parallel I/O files}
-
-When running in parallel, output of lattice-based (and particle)
-quantities takes place in parallel. As a consequence, the order
-of the output in the files is not in the same order as would be
-the case in serial. This means the parallel output must be
-manipulated into the correct order before any analysis.
-
-Parallel I/O subdivides the processors into one or more groups,
-each of which write data to a separate file. Infomation on the
-content of each of these files is contained in `meta-data' file
-for each quantity, which are produced automatically at the start
-of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
-If there are 8 I/O groups, the files will be named \texttt{008-001}
-to \texttt{008-008} for the different groups.
-
-The actual data is stored for each relevant time step as, e.g.,
-\texttt{phi-000100.001-001} where the extention refers to the
-I/O. To recombined a single, or multiple, data files into a
-single file in the correct order, a utility program
-\texttt{extract.c} is provided.
-
-The exectutable takes two arguments, e.g.,
-
-\inputkey{./extract phi.001-001.meta phi-000100.001-001}
-
-where  the first is the first meta-data file in the relevant series,
-and the second is the first daa file for the relevant time step. The
-program will combine the relevant quantities in the correct order
-and produce a singel new file \texttt{phi-000100} (in the above case).
-
-\subsection{Restarting}
-
-When restarting in parallel, the processor decomposition must be
-preserved. An attempt to change the processor decomposition will
-cause the configuration files to be read in the incorrect order,
-and results will be wrong.
 

### 2011-01-14 branches/CUDA/htdocs/user.tex
adding branch for CUDA development code
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1221 40876441-3496-0410-a613-a2fca35b2ae2

--- 2010-12-09 trunk/htdocs/user.tex
@@
 and results will be wrong.
+
+\section{Parallel Performance}
+
+\subsection{General Comments}
+
+The basic LB calculation should scale well in parallel, that is,
+the time taken for fixed problem size will decrease linearly
+with the number of MPI processes (strong scaling), or a larger
+problem can be run on a proportionally larger number processes
+in the same time (weak scaling). To ensure performance is retained
+in parallel, it is useful to understand some basic considerations.
+Some of these are discussed below.
+
+\textit{Ludwig} has been run successfully on up to 131,072 MPI
+tasks with close to ideal scaling, and there is no problem in
+principle to run larger decompositions.
+
+\subsection{Things to Remember and Things to Avoid}
+
+There are a number of simple issues to get right:
+\begin{itemize}
+\item
+Make sure the assertions are switched off and appropriate optimisations
+are switched on for production runs (see Makefile).
+\item
+Do not use too small a local domain size. Typically, 16$^3$ to 32$^3$
+cubic local domains will give resaonble scaling. Anything smaller may
+be inefficient. The best size may depend on the exact nature of the
+calculation and the hardware. Note that this limitation will always
+ultimately limit strong scaling.
+\item
+Don't have diagnsotic output (freq\_statistics) more often than is
+necessary. This output requires a global communication which is slow.
+\item
+Use BINARY output options; ASCII output is slower, and produces larger
+files.
+\item
+Do not use serial I/O. This will have an increasingly high overhead as
+the number of processors is increases (it's probably reasonable to
+128--256 MPI tasks). This is because each MPI tasks writes its data,
+in turn, to a single file; this in inherently serial. To avoid this,
+you need to specify the I/O decomposition for the various output
+quantities of interest. This splits the system into regular blocks
+(in a similar approach to the MPI decomposition), the data for which
+are written to separate files in parallel. For example, if the MPI
+decomposition is 8\_8\_8 (512 MPI tasks), then a reasonable I/O
+decomposition may be 2\_2\_2 or 4\_4\_4. The only way to find out
+what works best is to try some tests for the problem size you want to
+address.
+\item
+Don't use the Ewald sum! This simply does not scale.
+An alternative algorithm is required.
+\end{itemize}

### 2011-01-14 branches/CUDA/mpi_s/mpi.h

@@ -1,0 +1,192 @@
+/*****************************************************************************
+ *
+ *  mpi.h
+ *
+ *  Serial interface to MPI routines serving as a replacement in serial.
+ *
+ *  Broadly, point-to-point operations are disallowed, while
+ *  global operations do nothing.
+ *
+ *  From an idea appearing in LAMMPS.
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *
+ *****************************************************************************/
+
+#ifndef _MPI_SERIAL
+#define _MPI_SERIAL
+
+#ifdef __cplusplus
+extern "C" {
+#endif
+
+/* Datatypes */
+
+typedef int MPI_Handle;
+typedef MPI_Handle MPI_Comm;
+typedef MPI_Handle MPI_Datatype;
+typedef MPI_Handle MPI_Request;
+typedef MPI_Handle MPI_Op;
+typedef MPI_Handle MPI_Errhandler;
+
+typedef struct {
+  int MPI_SOURCE;
+  int MPI_TAG;
+} MPI_Status;
+
+typedef MPI_Handle MPI_Aint;
+
+/* Defined constants (see Annex A.2) */
+
+/* Return codes */
+
+enum return_codes {MPI_SUCCESS};
+
+/* Assorted constants */
+
+#define MPI_PROC_NULL     -9
+#define MPI_ANY_SOURCE    -10
+#define MPI_ANY_TAG       -11
+#define MPI_BOTTOM         0x0000
+#define MPI_UNDEFINED     -999
+
+/* Error-handling specifiers */
+
+enum error_specifiers {MPI_ERRORS_ARE_FATAL, MPI_ERRORS_RETURN};
+
+enum elementary_datatypes {MPI_CHAR,
+			   MPI_SHORT,
+			   MPI_INT,
+			   MPI_LONG,
+			   MPI_UNSIGNED_CHAR,
+			   MPI_UNSIGNED_SHORT,
+			   MPI_UNSIGNED,
+			   MPI_UNSIGNED_LONG,
+			   MPI_FLOAT,
+			   MPI_DOUBLE,
+			   MPI_LONG_DOUBLE,
+			   MPI_BYTE,
+			   MPI_PACKED};
+
+enum collective_operations {MPI_MAX,
+			    MPI_MIN,
+			    MPI_SUM,
+			    MPI_PROD,
+			    MPI_MAXLOC,
[…]

### 2011-01-14 branches/CUDA/src/colloid_io_rt.c

@@ -1,0 +1,93 @@
+/*****************************************************************************
+ *
+ *  colloid_io_rt.c
+ *
+ *  Run time colloid I/O settings.
+ *
+ *  $Id: colloid_io_rt.c,v 1.3 2010-10-19 07:45:28 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <string.h>
+
+#include "pe.h"
+#include "runtime.h"
+#include "cio.h"
+
+/*****************************************************************************
+ *
+ *  colloid_io_run_time
+ *
+ *  The decomposition is pending.
+ *
+ *****************************************************************************/
+
+void colloid_io_run_time(void) {
+
+  int  nuser;
+  char tmp[256];
+
+  /* Defaults */
+
+  colloid_io_format_input_serial_set();
+  colloid_io_format_input_ascii_set();
+  colloid_io_format_output_ascii_set();
+
+  info("\n");
+  info("Colloid I/O settings\n");
+  info("--------------------\n");
+
+  nuser = RUN_get_string_parameter("colloid_io_format_input", tmp, 256);
+
+  if (nuser == 0) {
+    info("Input format:       ascii serial\n");
+  }
+
+  if (strncmp("ASCII",  tmp, 5) == 0 ) {
+    colloid_io_format_input_ascii_set();
+    info("Input format:  ascii\n");
+  }
+
+  if (strncmp("ASCII_SERIAL",  tmp, 12) == 0 ) {
+    colloid_io_format_input_ascii_set();
+    colloid_io_format_input_serial_set();
+    info("Input file:    serial single file\n");
+  }
+
+  if (strncmp("BINARY", tmp, 6) == 0 ) {
+    colloid_io_format_input_binary_set();
+    info("Input format:  binary\n");
+  }
+
+  if (strncmp("BINARY_SERIAL", tmp, 6) == 0 ) {
+    colloid_io_format_input_binary_set();
+    colloid_io_format_input_serial_set();
+    info("Input file:    serial single file\n");
+  }
+
+  nuser = RUN_get_string_parameter("colloid_io_format_output", tmp, 256);
+
+  if (nuser == 0) {
[…]

### 2011-01-14 branches/CUDA/src/distribution_rt.h

@@ -1,0 +1,20 @@
+/*****************************************************************************
+ *
+ *  distribution_rt.h
+ *
+ *  $Id: distribution_rt.h,v 1.2 2010-10-15 12:40:02 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef DISTRIBUTION_RT_H
+#define DISTRIBUTION_RT_H
+
+void distribution_run_time(void);
+
+#endif

### 2011-01-14 branches/CUDA/src/interaction.c

@@ -1,0 +1,801 @@
+/*****************************************************************************
+ *
+ *  interaction.c
+ *
+ *  Colloid potentials and colloid-colloid interactions.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <float.h>
+#include <math.h>
+
+#include "pe.h"
+#include "timer.h"
+#include "coords.h"
+#include "ran.h"
+#include "runtime.h"
+
+#include "bbl.h"
+#include "build.h"
+#include "physics.h"
+#include "magnetic_field.h"
+#include "magnetic_field_rt.h"
+#include "potential.h"
+
+#include "colloids.h"
+#include "interaction.h"
+#include "model.h"
+#include "site_map.h"
+#include "cio.h"
+#include "control.h"
+#include "subgrid.h"
+#include "stats_colloid.h"
+
+#include "util.h"
+#include "colloid_sums.h"
+#include "colloids_halo.h"
+#include "colloids_init.h"
+#include "ewald.h"
+
+static void    COLL_overlap(colloid_t *, colloid_t *);
+static void    COLL_set_fluid_gravity(void);
+
+void lubrication_sphere_sphere(double a1, double a2,
+			       const double u1[3], const double u2[3],
+			       const double r12[3], double f[3]);
+static void    coll_position_update(void);
+static void    lubrication_init(void);
+static void    colloid_forces_check(void);
+
+static int    gravity_ = 0;            /* Switch */
+static double g_[3] = {0.0, 0.0, 0.0}; /* External gravitational force */
+
+struct lubrication_struct {
+  int corrections_on;
+  double cutoff_norm;  /* Normal */
+  double cutoff_tang;  /* Tangential */
+} lubrication;
+
+static double epotential_;
+
+/*****************************************************************************
+ *
+ *  COLL_update
+ *
+ *  The colloid positions have been updated. The following are required:
+ *    (1) deal with processor/periodic boundaries
[…]

### 2011-01-14 branches/CUDA/tests/test_assumptions.c

@@ -1,0 +1,166 @@
+/*****************************************************************************
+ *
+ *  test_assumptions.c
+ *
+ *  Test basic model assumptions, portability issues.
+ *
+ *  Also testing util.h stuff at the moment.
+ *
+ *  $Id: test_assumptions.c,v 1.8 2010-11-02 17:51:22 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  (c) 2008 The University of Edinburgh 
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *
+ *****************************************************************************/
+
+#include <math.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <assert.h>
+
+#include "tests.h"
+#include "util.h"
+
+void (* p_function)(void);
+void test_util(void);
+
+int main(int argc, char ** argv) {
+
+  int n;
+  int * p_int;
+
+  printf("Testing assumptions...\n");
+
+  /* All integers in the code should be declared 'int', which
+   * we expect to be (at the least) 4 bytes. */
+
+  printf("Checking sizeof(int) is 4 bytes... ");
+  assert(sizeof(int) == 4);
+  printf("yes\n");
+
+  printf("Checking sizeof(long int) is >= 4 bytes... ");
+  assert(sizeof(long int) >= 4);
+  printf("yes (%ld bytes)\n", sizeof(long int));
+
+  /* All floating point types in the code should be double,
+   * which must be 8 bytes. */
+
+  printf("Checking sizeof(float) is 4 bytes... ");
+  assert(sizeof(float) == 4);
+  printf("yes\n");
+
+  printf("Checking sizeof(double) is 8 bytes... ");
+  assert(sizeof(double) == 8);
+  printf("yes\n");
+
+  printf("Checking sizeof(void *) ...\n");
+  printf("%lu\n", sizeof(void *));
+
+  printf("Checking FILENAME_MAX >= 128 characters ... ");
+  assert(FILENAME_MAX >= 128);
+  printf("yes (%d characters)\n", FILENAME_MAX);
+  printf("Checking BUFSIZ... (%d bytes)\n", BUFSIZ);
+
[…]

### 2011-01-14 branches/CUDA/tests/test_model.c

@@ -1,0 +1,619 @@
+/*****************************************************************************
+ *
+ *  test_model.c
+ *
+ *  Unit test for the currently compiled model (D3Q15 or D3Q19).
+ *
+ *  $Id: test_model.c,v 1.10 2010-11-02 17:51:22 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) The University of Edinburgh (2010)
+ *
+ *****************************************************************************/
+
+#include <math.h>
+#include <stdlib.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "util.h"
+#include "model.h"
+#include "tests.h"
+
+static void test_model_constants(void);
+static void test_model_velocity_set(void);
+static void test_model_distributions(void);
+static void test_model_halo_swap(void);
+static void test_model_reduced_halo_swap(void);
+static  int test_model_is_domain(const int ic, const int jc, const int kc);
+
+int main(int argc, char ** argv) {
+
+  pe_init(argc, argv);
+
+  info("Testing D%1dQ%d\n", NDIM, NVEL);
+
+  /* Test model structure (coordinate-independent stuff) */
+
+  test_model_constants();
+  test_model_velocity_set();
+
+  /* Now test actual distributions */
+
+  coords_init();
+
+  test_model_distributions();
+  test_model_halo_swap();
+  test_model_reduced_halo_swap();
+
+  info("\nModel tests passed ok.\n\n");
+
+  coords_finish();
+  pe_finalise();
+
+  return 0;
+}
+
+/*****************************************************************************
+ *
+ *  test_model_constants
+ *
+ *  Check the various constants associated with the reduced halo swap.
+ *
+ *****************************************************************************/
+
[…]

### 2011-01-14 branches/CUDA/tests/test_sigma.c

@@ -1,0 +1,510 @@
+/*****************************************************************************
+ *
+ *  test_sigma.c
+ *
+ *  System test to compute the surface tension via measurement of the
+ *  Laplace pressure difference between the inside an outside of a
+ *  stationary droplet.
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdio.h>
+#include <float.h>
+#include <math.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "lattice.h"
+#include "model.h"
+#include "collision.h"
+#include "utilities.h"
+#include "propagation.h"
+#include "free_energy.h"
+#include "phi.h"
+#include "phi_stats.h"
+#include "phi_gradients.h"
+#include "phi_cahn_hilliard.h"
+#include "physics.h"
+
+#include "timer.h"
+#include "runtime.h"
+
+struct drop_t {
+  double radius;
+  double xi0;
+  double centre[3];
+  double phimax;
+  double laplace_in, laplace_out;
+};
+
+void drop_init(struct drop_t);
+void drop_locate_centre(struct drop_t *);
+void drop_locate_radius(struct drop_t *);
+void drop_locate_profile(struct drop_t, const int);
+void drop_laplace_pressure_difference(struct drop_t *);
+void drop_relax(int);
+double sigma(struct drop_t);
+
+int main (int argc, char ** argv) {
+
+  struct drop_t drop0;
+
+  pe_init(argc, argv);
+
+  RUN_read_input_file("test_sigma_input");
+
+  coords_init();
+  init_free_energy();
+
+  drop0.radius = L(X)/4.0;
+  drop0.xi0 = 2.0*interfacial_width();
+  drop0.centre[X] = L(X)/2.0;
+  drop0.centre[Y] = L(Y)/2.0;
+  drop0.centre[Z] = L(Z)/2.0;
+  drop0.phimax    = 1.0;
+
+  MODEL_init();
+
+  /* Compute the surface tension for a given parameter set */
+  sigma(drop0);
+
+  TIMER_statistics();
+  phi_finish();
+  pe_finalise();
+
+  return 0;
+}
+
+/*****************************************************************************
+ *
+ *  sigma
[…]

### 2011-01-14 branches/CUDA/tests/trytests.sh

@@ -1,0 +1,119 @@
+##############################################################################
+#
+#  try.sh
+#
+#  This file runs the unit tests and the regression tests.
+#  It is intended for my nightly test, so may not be very
+#  portable.
+#
+#  $Id$
+#
+#  Edinburgh Soft Matter and Statistical Physics Group and
+#  Edinburgh Parallel Computing Centre
+#
+#  Kevin Stratford (kevin@epcc.ed.ac.uk)
+#  (c) 2010 The University of Edinburgh
+#
+##############################################################################
+#!/bin/bash
+
+# Serial unit tests
+
+cd ../mpi_s
+make clean
+make libc
+make testc
+
+cd ../src
+make clean
+make lib
+cd ../tests
+make clean
+make do_tests
+
+# Serial regressions
+
+cd ../src
+make serial
+
+echo Running autocorrelation test
+./Ludwig.exe ../tests/regression/test_auto1_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_auto1_srl_d3q19.ref
+
+echo Running LE1 test
+./Ludwig.exe ../tests/regression/test_le1_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_le1_srl_d3q19.ref
+
+echo Running LE2 test
+./Ludwig.exe ../tests/regression/test_le2_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_le2_srl_d3q19.ref1
+
+echo Running LE3 test
+./Ludwig.exe ../tests/regression/test_le3_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_le3_srl_d3q19.ref
+
+echo Running LE4 test
+./Ludwig.exe ../tests/regression/test_le4_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_le4_srl_d3q19.ref
+
+echo Running LE5 test
+./Ludwig.exe ../tests/regression/test_le5_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_le5_srl_d3q19.ref
+
+echo Running LE6 test
+./Ludwig.exe ../tests/regression/test_le6_input > /tmp/junk
+diff /tmp/junk ../tests/regression/test_le6_srl_d3q19.ref
+
+echo Running spin03 serial test
[…]

### 2011-01-19 htdocs/user.tex
Keyword tags require updating
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1182 40876441-3496-0410-a613-a2fca35b2ae2
Try svn propset svn:keywords "Author Date" user.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1185 40876441-3496-0410-a613-a2fca35b2ae2
Comments on calibration in user.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1228 40876441-3496-0410-a613-a2fca35b2ae2

--- 2010-12-07 trunk/htdocs/user.tex
@@
 
+\subsubsection{Hydrodynamic radius}
+
+There is an important issue to understand about the (nominally)
+spherical particles in LB. The input radius $a_0$ is used to
+construct the link boundary conditions. However, discretisation
+errors in the boundary conditions means that the `real' particle
+size may be different. This real radius is the hydrodynamic radius $a_h$.
+As a further complication, the hydrodynamic radius can also
+be a function of viscosity for a given input radius. Some values
+are given in Table~X. This values are computed using a calibration
+procedure described by Nguyen and Ladd \cite{nguyen2001}.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+\hline
+$\eta$ & \multicolumn{3}{c|}{$a_0$}\\ \cline{2-4} 
+       & 1.25 & 2.30 & 4.77 \\
+\hline
+1/6  & 1.09 (1.05) & 2.23 (2.20) & 4.76 \\
+1/100  & 1.40 (1.34) & 2.50 (2.46) & 5.01 \\
+1/1000  & 1.63 & 2.71 & 5.23 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{A table of calibrated hydrodynamic radii as a function of input
+radius and fluid viscosity $\eta$. These figures are for D3Q19 (and D3Q15
+in brackets).}
+\end{table}
+
+If a calibration is required for a given radius and viscosity,
+use the \texttt{calibration on} key pair in the input file. Some
+examples are given in the \texttt{tests/calibration} directory.
+An apppropriate
+number of time steps must be used to allow the system to get over
+the initial transient (we use the momentum diffusion time for the
+system $L^2/\eta$) and to allow at least one
+particle Stokes time ($a/U$). The input radius and hydrodynamic
+radius should be set the same in the input file.
+
+% PENDING Ginzburg d'Humieres correct to ghost mode relaxation times
+
 \subsection{Boundary conditions}

### 2011-01-26 trunk/htdocs/user.tex
Added comments on LE constraints to user.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1240 40876441-3496-0410-a613-a2fca35b2ae2

@@
 the extraction utility (see section on parallel I/O).
+
+Note that if something more complicated is required, for example,
+an equilibration at zero shear followed by a run with shear, one
+needs to be careful how this is arranged. The code works out the
+current displacement of the planes by computing $U_{LE} t$, where
+$t$ is the current time step. A shear run should then start from
+$t = 0$, otherwise an offset is required. At the moment this cannot
+be done via the input.
+
+\subsubsection{Lees-Edwards planes in parallel}
+
+There are a couple of additional constraints to use the Lees-Edwards
+planes in parallel. In particular, the planes cannot fall at a
+processor boundary in the $x$-direction. This means you should
+arrange an integer number of planes per process in the $x$-direction.
+(For example, use one plane per process; this will also ensure the number
+of planes
+still evenly divides the total system size.)
+This will interleave the planes with the processor decomposition.
+The $y$-direction and $z$-direction may be decomposed without
+further constraint.
+
+Note that this means a simulation with one plane will only work
+if there is one process in the $x$ decomposition.
 

### 2011-02-11 htdocs/porous.tex
Updates to tex documents
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1271 40876441-3496-0410-a613-a2fca35b2ae2

--- 2009-05-07 trunk/htdocs/sections.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
-%  sections.tex
+%  porous.tex
 %
-%  Content for Ludwig documentation.
+%  Some considerations for porous media.
 %
-%  $Id: sections.tex,v 1.10 2009-05-07 15:29:05 kevin Exp $
+%  $Id$
 %
@@
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
+%  (c) 2011 The University of Edinburgh
 %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Introduction}
-
-\subsection{Note to Readers}
-
-This document is intended as a starting point for a full-blown
-specification of the next version of the Ludwig code, and everything
-to do with it. It is (manifestly) not complete.
-
-\subsection{Some History}
-
-Early Lattice Gas and Lattice Boltzmann. Early Lattice Boltzmann at Edinburgh.
-Ludwig.
-
-\subsection{Aims}
-
-We aim to provide a robust and portable code, written in C, which
-can be used to perform serial and scalable parallel simulations of
-complex fluid systems based around the Lattice Boltzmann method.
-
-\subsection{Availability}
-
-We now have a NeSCForge project account available at
-
-\texttt{http://forge.nesc.ac.uk/projects/ludwig/}
-
-This provides controlled access to known
-persons and provide CVS, bug tracking, and so on. It is
-\textbf{essential} that people sign up to this, as we must
-move to a single supported version.
-
-The NeSCForge CVS is not available to the public.
-
-\subsection{Download}
-
-\section{User Input}
-
-By default the executable will look for an input file at run time.
-By default, this file is named \texttt{input} and should be in the
-current working directory. The alternative is to specify the name
-of the file explicitly on the command line:
-\begin{verbatim}
-$ ./Ludwig.exe input_file
-\end{verbatim}
-
-The contents of the input file are made up of \textit{key value}
-pairs with control the run-time behaviour. Blank lines and lines
-begining with \# are ignored as comments. The following describes
-the effect of various keys. Most keys have a default value which
-will be used by the code if the corresponding key is not present
-(or commented out) in the input file.
-
-\subsection{Basic run parameters}
-
-\inputkey{N\_cycles}
-
-The number of lattice Boltzmann time steps to execute. Default
-value:
-
-\texttt{N\_cycles 0}
-
-\inputkey{t\_start}
-
-By default the code will start from time step $t = 0$. If you 
-wish to restart from a previously saved configuration, set
-the appropriate value for \texttt{t\_start}. The code will
-execute \texttt{N\_cycles} steps starting from this point.
-
-\inputkey{input\_config}
-
-The file name stub for input configuration required for restarts.
-For example, to restart from the file config.out100 previously
-produced at the end of a 100 step run use
-\begin{verbatim}
-t_start 100
-input_config config.out100
-\end{verbatim}
-
-If you are using parallel I/O, the filename stub is required; the
-file extensions are computed automatically.
-
-\inputkey{size}
-
-Controls the number of lattice points in the $(x, y, z)$ directions,
-respectively. Default value:
-
-\texttt{size 64\_64\_64}
-
-\inputkey{grid}
-
-In parallel, this sets the processor Cartesian decomposition, i.e.,
-the number of processors in each dimension of the Cartesian communicator.
-The default value may be implementation dependent as it is that returned
-by \texttt{MPI\_Dims\_create()}. For
-example:
-\begin{verbatim}
-size 64_64_64
-grid 4_2_1
-\end{verbatim}
-gives a total (physical) system size of 64 lattice sites in each
-vdirection, decomposed across 4 processors in the x-direction, 2 in
-the y-direction and 1 (no decomposition) in the z-direction. The
-total number of processors must therefore be 8. The local domain
-size per processor is then 16x32x64.
-
-\subsection{Fluid properties}
-
-\inputkey{viscosity}
-
-Sets the LB fluid shear viscosity $\eta$ (and related relaxation times).
-Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
-(i.e., realaxation time equal to unity).
-
-\inputkey{viscosity\_bulk}
-
-Sets the LB bulk viscosity $\zeta$ (and related relaxation times).
-This can be use if you are concerned with incompressiblity
-violations, in which case $\zeta >> \eta$ can be useful. The
-default value is $\zeta = \eta$.
-
-\inputkey{temperature}
-
-The LB 'temperature' using fluctuating hydrodynamics. Safe values
-are $0.0001 > kT > 0$.
-
-\inputkey{rho0}
-
-Sets the mean density of the fluid. The default value is unity, and
-it is recommended that the value should not be changed.
-
-\subsection{Binary fluid parameters}
-
-The are currently three choices of free energy relating to
-various systems.
-
-\subsubsection{Symmetric binary fluid}
-
-\inputkey{free\_energy symmetric}
-
-The following applies for the binary fluid problem with
-compositional order parameter $\phi$ and free energy
-(excluding the term in the density $\rho$):
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
-\end{equation}
-
-This is described in some detail
-by Kendon et al \cite{viv}. The first two terms represent the bulk
-contribution, whereas the term in $\kappa$ penalisies curvature in
-the interface.
-
-\inputkey{A}
-
-The parameter $A$. Note $A < 0$.
-
-\inputkey{B}
-
-The parameter $B$. Note $B = -A$ for common usage, although this is
-not enforced.
-
-\inputkey{K}
-
-The parameter $\kappa$, which is positive.
-
-\subsubsection{Brazovskii free energy}
-
-This is similar to the symetric, but with one extra term which
-requires description.
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
-\end{equation}
-
-\subsubsection{Surfactant model free energy}
-
-\inputkey{free\_energy surfactant}
-
-This is for use with a two order parameter model the first of
-which is the composition, as for the symmetric free energy,
-while the second $\psi$ represents surfactant concentration.
-The free energy density is made up of a number of terms:
-
-\begin{equation}
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-\end{equation}
-is the standard symmetric term related to the composition as above;
-
-\begin{equation}
-D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
-\end{equation}
-represents the energy of surfactant in the bluk phase with surfactant
-concentration $0 < \psi < 1$ with a single parameter $D$;
-\begin{equation}
--{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
--{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
-\end{equation}
-where the term in $\epsilon$ represents the energy reduction by
-adsorbing surfactant at the interface and the term in $\beta$
-provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
-\begin{equation}
-+{\textstyle \frac{1}{2}} W \psi \phi^2
-\end{equation}
-which is included to stabilise the interface \cite{theissengompper}.
-
-The corresponding keys for the input file are:
-
-\inputkey{D}
-
-is the bulk surfactant contribution parameter
-
-\inputkey{epsilon}
-
-is the parameter $\epsilon$
-
-\inputkey{W}
-
-is the parameter $W$
-
-\inputkey{beta}
-
-is the value of the parameter $\beta$ (not implemented yet).
-
-\inputkey{phi\_b}
-
-the initial, uniform, background surfactant concentration $\psi_b$.
-
-\subsection{Other Binary Fluid Parameters}
-
-\inputkey{mobility}
-
-Sets the compositional order parameter mobility $M$ (uniform in space)
-which appears in the Cahn-Hilliard equation.
-
-\inputkey{mobilty\_psi}
-
-Sets the (uniform) mobility for the surfactant as appropriate for
-the Cahn-Hilliard equation.
-
-\inputkey{phi0}
-
-The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
-The default value is zero, i.e., a symmetric 50:50 mixture by volume.
-
-\inputkey{phi\_initialisation}
-
-Determines how the compositional order parameter is initialisated
-at the start of the run. If set to \texttt{spinodal} the value
-is set to $\phi_0$ as set above plus or minus a random noise, the
-magnitude of which is set by the value of the \texttt{noise} key.
-If set to \texttt{block} a one-dimensional profile is set up in
-the $z$-direction representing two blocks of fluid with
-$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
-$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
-appropriate width. The $\phi = -1$ section is in the middle.
-
-\inputkey{noise}
-
-The magnitude of the initial fluctuations in $\phi$ used to
-initiate spinodal decompsotion.
-
-\subsection{Colloidal particles}
-
-There are quite a few options here.
-
-\inputkey{colloid\_init}
-
-This determines how the code initialises particles (or not). The
-default is to have no particles (value \texttt{no\_colloids}). There
-are two other options at the moment.
-
-\texttt{colloid\_init    fixed\_number\_monodisperse}
-
-initialises a fixed number of particles with randomly selected
-positions. Note that there is no gaurentee that the particles
-do not overlap. The number of particles is determined by
-the value of the key \texttt{colloid\_no}.
-
-\texttt{colloid\_init   fixed\_volume\_fraction\_monodisperse}
-
-initialises a given volume fraction of particles on a (roughly)
-bcc lattice. The particles will not overlap. The volume fraction
-is set using the key \texttt{colloid\_vf}.
-
-\inputkey{colloid\_a0}
-
-Sets the nominal radius of the particle on the lattice $a_0$. This is
-used to construct the links for BBL. The smallest acceptable LB
-particle is $a_0 = 1.25$ \cite{nl02}. There are 'magic' values
-including 1.25, 2.3, 3.7, 4.7 which minimise discretisation effects.
-
-\inputkey{colloid\_ah}
-
-Sets the hydrodynamic radius of the particles $a_h$. For all
-reasonable applications choose $a_h = a_0$.
-
-\inputkey{colloid\_mc\_steps}
-
-It is possible to have the code follow the initialisation of
-particle positions by a number of Monte Carlo steps. This can
-help to remove overlaps (fixed\_number\_monodisperse) or
-melt the lattice (fixed\_volume\_fraction\_monodisperse).
-
-\inputkey{colloid\_gravity}
-
-Sets a body force on each paarticle. Useful, for example, for
-sedimentation. For example,
-
-\inputkey{cell\_list\_min}
-
-This sets the minimum allowable width of the cell list used to
-store the particles. This should be as small as possible for
-efficiency purposes but must be large enough to capture all
-possible interactions.
-
-\texttt{colloid\_gravity  0.0\_0.0\_-0.0001}
-
-gives a constant force in the negative $z$-direction on each particle.
-The code automatically computes the compensating body force on fluid nodes
-required to give no net change in the total momentum of the system at
-each time step.
-
-\subsection{Active particles}
-
-Not in general use yet.
-
-\subsection{Particle-particle interactions}
-
-\subsection{Boundary conditions}
-
-Solid planar boundary walls can be added at the extremities of the
-system (i.e., the volume of fluid is extactly that specified by
-the \texttt{size} keyword). The wall can be in one, two or all three
-coordinate direcctions, i.e., a 'slab' geometry, and 'duct' geometry
-or a 'box' geometry, respectively.
-
-The boundary walls are switched on via the keyword value pair
-
-\inputkey{boundary\_walls\_on [no | yes]}
-
-If \texttt{yes} then the geometry is specified with the keyword
-as one of
-
-\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}
-
-Note that these are the only valid selections. \texttt{1\_1\_0}
-gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction. \texttt{1\_0\_0}
-gives periodic conditions in the $x-$ direction only, while
-\texttt{0\_0\_0} is a fully enclosed box.
-
-In the case of the slab geometry, two further values are used so
-that shear may be imparted to the system:
-
-\inputkey{boundary\_speed\_bottom}
-
-\inputkey{boundary\_speed\_top}
-
-These values set the $x-$component of the wall velocity at
-$z = z_{min}$ and $z = z_{max}$, respectively. Like all
-velocities, these values must respect the low Mach number
-constraint $u_x < c_s$.
-
-\subsection{Input/Output and Parallel I/O}
-
-\inputkey{freq\_statistics}
-
-sets the frequency for reporting general information from
-the code. (In general, the function
- \texttt{is\_statistics\_step()} returns true at this frequency.
-
-\inputkey{freq\_measure}
-
-sets the frequency of recording of colloid/order parameter output.
-(In general, the function \texttt{is\_measurement\_step()} returns
-true at this frequency.)
-
-\inputkey{freq\_config}
-
-set frequency of full configuration (checkpoint) dumps suitable
-for restart.
-
-\inputkey{freq\_phi}
-
-set frequency of order parameter output.
-
-\inputkey{phi\_format}
-
-either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-
-\inputkey{freq\_vel}
-
-set frequancy of velocity output.
-
-\inputkey{vel\_format}
-
-either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-
-\inputkey{config\_at\_end}
-
-enforces full configuration dump at the end of execution. The
-default is 'yes'.
-
-\subsection{Lees-Edwards planes}
-
-\inputkey{N\_LE\_plane}
-
-sets the total number of Lees-Edwards planes. Default is zero.
-The placing is as followings. The number of planes $n$ must
-divide the lattice size in the $x$-direction to give an integer
-$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
-
-\inputkey{LE\_plane\_vel}
-
-sets the velocity of each plane relative to the lattice. All planes
-have the same, constant, velocity.
-
-\inputkey{LE\_init\_profile}
-
-if set to 1, the fluid velocity is initialise to reflect a steady
-state shear flow appropriate for the number of planes at the
-given velocity. If set to zero, the fluid is initialised with
-zero velocity.
-
-\subsection{Porous Media}
+\section{Porous Media}
 
@@
 
-\subsubsection{File format}
+\subsection{File format}
 
@@
 
-\subsubsection{Permeability calculations}
+\subsection{Permeability calculations}
 
@@
 This is an artefact of the bounce-back on links and will result in a
-viscosity-dependent permeability (see, e.g., \cite{lipanmiller}). To minimise
-this effect, it is recommended to switch on the correction to the ghost
-mode relaxation time proposed by Ginzburg and d'Humi\`eres \cite{ginzburg}
-which corrects the viscosity-dependence of the apparent boundary
-position. This is done via the input switches
+viscosity-dependent permeability (see, e.g., \cite{lipanmiller}).
+To minimise this effect, e.g.,  Ginzburg and d'Humi\`eres \cite{ginzburg}
+corrects the viscosity-dependence of the apparent boundary position
+using a three relaxation time scheme. This is currently under
+investigation.
 
-\texttt{ghost\_modes on\\
-ginzburg\_dhumieres on}
-
-and should recover a viscosity-independent permeability measurement
-\cite{lipanmiller,chunladd}.
-
-\subsubsection{Example: circular capillaries}
+\subsection{Example: circular capillaries}
 
@@
 
-\subsubsection{Example: rectangular capillaries}
+\subsection{Example: rectangular capillaries}
 
@@
 
-\subsection{Miscellaneous}
+\subsection{Matching lattice and real units in porous media}
 
-\inputkey{random\_seed}
+Following Succi \cite{succi} (Chapter~8) the following argument
+can be made to match lattice quantities and real physical
+quantities for a given system of interest. Suppose we have
+a structure with characteristic pore size $h = 1$ micron
+and characteristic sample size of 1 cm. If the fluid
+(e.g., water) has viscosity $\eta = 10^{-3}$ Pa s and density
+$\rho = 10^3$ kg~s$^{-1}$, and a typical flow is
+$U \sim 10^{-3}$ m~s$^{-1}$, then the pore scale Reynolds
+number is $Re_h \sim 10^{-3}$.
 
-sets the random number generator seed.
+If the speed of sound in water is 1500~m~s$^{-1}$,
+then the corresponding time scale related to the
+speed of sound and the pore size is $T = h/ c_s$,
+which is $\sim 10^{-10}-10^{-9}$~s. (Exact matching
+is acheived via the dimensionless group $\Delta x / c_s\Delta t$,
+with the lattice speed of sound $1/\sqrt{3}$).
 
-\subsection{Dealing with the Parallel I/O files}
-
-When running in parallel, output of lattice-based (and particle)
-quantities takes place in parallel. As a consequence, the order
-of the output in the files is not in the same order as would be
-the case in serial. This means the parallel output must be
-manipulated into the correct order before any analysis.
-
-Parallel I/O subdivides the processors into one or more groups,
-each of which write data to a separate file. Infomation on the
-content of each of these files is contained in `meta-data' file
-for each quantity, which are produced automatically at the start
-of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
-If there are 8 I/O groups, the files will be named \texttt{008-001}
-to \texttt{008-008} for the different groups.
-
-The actual data is stored for each relevant time step as, e.g.,
-\texttt{phi-000100.001-001} where the extention refers to the
-I/O. To recombined a single, or multiple, data files into a
-single file in the correct order, a utility program
-\texttt{extract.c} is provided.
-
-The exectutable takes two arguments, e.g.,
-
-\inputkey{./extract phi.001-001.meta phi-000100.001-001}
-
-where  the first is the first meta-data file in the relevant series,
-and the second is the first daa file for the relevant time step. The
-program will combine the relevant quantities in the correct order
-and produce a singel new file \texttt{phi-000100} (in the above case).
-
-\subsection{Restarting}
-
-When restarting in parallel, the processor decomposition must be
-preserved. An attempt to change the processor decomposition will
-cause the configuration files to be read in the incorrect order,
-and results will be wrong.
-
-\section{The Lattice Boltzmann Method}
-
-We review here the lattice Boltzmann method applied to single
-phase problems.
-
-\subsection{The Navier Stokes Equation}
-
-We seek to solve the isothermal Navier-Stokes equations which, often
-written in vector form, express mass conservation
-\begin{equation}
-\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
-\label{eq_mass1}
-\end{equation}
-and the conservation of momentum
-\begin{equation}
-\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
--\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
-+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
-\label{eq_momentum1}
-\end{equation}
-Equation~\ref{eq_mass1} expresses the local rate of change of the
-density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
-mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
-Equation~\ref{eq_momentum1} expresses Newton's second law for
-momentum, where the terms on the right hand side represent the
-force on the fluid.
-
-For this work, it is more convenient to rewrite these equations
-in tensor notation, where Cartesian coordinates ${x,y,z}$ are
-represented by indices $\alpha$ and $\beta$, viz
-\begin{equation}
-\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
-\end{equation}
-and
-\begin{equation}
-\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
-= -\nabla_\alpha p
-+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
-+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
-\end{equation}
-Here, repeated indices are understood to be summed over.
-The conservation law is seem better if the forcing terms of the
-right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
-so that
-\begin{equation}
-\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
-\end{equation}
-In this case the expanded expression for the stress tensor is
-\begin{equation}
-\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
-+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
-\end{equation}
-where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
-$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
-modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.
-
-\subsection{The Distribution Function and its Moments}
-
-In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
-\begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
-\end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
-
-In general, the number of modes supported in $n$ and these can
-be written as
-\begin{equation}
-M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
-\end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
-\begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
-\end{equation}
-
-\subsubsection{Hydrodynamic modes}
-
-\subsubsection{D3Q15 ghost modes}
-
-\subsubsection{D3Q19 ghost modes}
-
-\subsection{The Lattice Boltzmann Equation}
-
-\subsection{The Collision}
-
-\subsection{External Body Forces}
-
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsubsection{Fluctuating ghost modes: D3Q15}
-
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Boundary walls}
-
-\subsection{Colloids}
-
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
-\section{Code Design}
-
-\subsection{Some Comments}
-
-In designing a code in C the main object of which is to do high
-performance computing, there are some inevitable tensions with
-software engineering best-practice. In particular, complete
-encapsulation of data in the spirit of using abstract data types
-tends to cause problems of performance.
-
-For example, encapsulation of the lattice Boltzmann distribution
-$f_i$, together with appropriate ``get'' and ``set'' functions for
-access when approapriate, is possible. There could then be a
-complete separation of interface and implementation of the
-distribution (other than that each individual $f_i$ is represented
-by a \texttt{double}). However, experience suggests that
-this imposes severe impediment to optimisation and hence
-unacceptable performance degradation.
-
-Schemes do exist by which one can approximate the object
-oriented programming style in C, and so achieve a separation
-of interface and implementation. However, this tends to put
-a heavy burden on the use of \texttt{void} pointers, itself
-a recipe for possible disaster. A truly object-oriented
-language such as Java would be preferable.
-
-So, for the time being, some relaxation of encapsulation is required.
-Exposing a pointer to the distributions which allows them to be accessed
-directly where required. This can be done merely by declaring the
-necessary variables \texttt{extern}. In practice, such deviations
-from encapsulation can be minimised in the lattice sector of the
-code. In particlar, \texttt{extern} declarations should not be
-required in \texttt{main()}, i.e., it should be able to write a
-new application without worrying about details of implementation.
-
-\subsubsection{Objects}
-
-Somewhat more difficult is the part of the code involving
-colloids, which would really favour a truly object-oriented
-approach.
-
-\subsubsection{Serial and parallel}
-
-The design decission has been taken to support execution of the
-code in both serial and parallel. The default compilation
-assumes that a serial executable is required, while  MPI code
-is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).
-
-There is no shared memory inplementation at this time.
-
-\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}
-
-\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}
-
-\begin{figure}
-\begin{center}
-%\includegraphics{xfig/sketch.eps}
-\end{center}
-\caption{A scheme by which different parts of the code might be
-encapsulated and in such a way to allow reliable development and
-testing. As this is C, will do not adopt a truly ``object'' langauge.
-Broadly, different modular parts should only depend on stuff above
-in the diagram. So, at the top, there is the parallel environment
-where communication occurs within MPI\_COMM\_WORLD. Built on top
-of this is the Cartesian Communicator responsible for halo swaps
-and so forth. The LE code may have a separate Communicator again. 
-(There's an argument to say the diagram should be the other way up!)}
-\end{figure}
-
-\section{Code Implementation}
-
-\subsection{Stuff}
-
-\subsection{Coordinate System}
-
-\subsubsection{The global coordinate system}
-
-The user specifies the system size in lattice units at run time,
-which sets the extent of memory which must be allocated for the
-Cartesian lattice. The extent of the physical coordinate system
-then reflects the lattice size directly (see Figure \ref{fig_c1}). 
-Each lattice site, or node,  has coordinates corresponding to its
-integer position in the system and can be thought of as being
-at the centre of a control volume one lattice unit on a side.
-This means that the nominal edges of the
-system are offset from the lattice sites themselves by half a
-lattice unit in each direction.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_c1.eps}
-\end{center}
-\caption{The global Cartesian coordinate system reflects the choice
-of lattice size; lattices sites are located at
-$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
-(the Figure only shows the labels in the $y-$direction for simplicity).
-In the Figure, the lattice sites in the domain proper are the full
-circles; halo sites, which are used to effect periodic boundary
-conditions, are included as open cirles}
-\label{fig_c1}
-\end{figure}
-
-For particles, the centres of which can move continuously across the
-lattice, the extents of the global coordinate system are then
-$L_{min} < x < L + L_{min}$ and so on.
-
-\subsubsection{Addressing the lattice: the local coordinate system}
-
-\subsection{Solid walls}
-
-General solid objects which are fixed for the duration of
-the simulation, such as porous media, can be defined
-by assigning solid/fluid status to each lattice node. 
-The links required between solid and fluid nodes can
-then be identified for the given model.
-
-\textit{Comment.} In general, a solid object will not
-affect the periodic nature of the simulation. A general
-solid object cannot move, i.e., the boundary velocity
-must be zero.
-
-\textit{Special case.} In the case where the solid object
-represents a flat boundary wall at the edge of the system
-in one or more directions, the periodicity is broken in
-the corresponding directions.
-
-\vfill
-\pagebreak
-
-\section{Code Interface}
-
-\subsection{The Parallel Environment}
-
-\texttt{\#include ``pe.h''}
-
-As the code is to be available in both serial and parallel versions,
-it is desirable to abstract the basic environment to some extent.
-The parallel environment is intended to support MPI communication
-within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
-consistent picture in serial (i.e., rank zero is the single process).
-
-\texttt{void pe\_init(int argc, char ** argv)}\\
-Responsible for initialisation of MPI, and hence must be first
-execuatable statement. The arguments are those to \texttt{main}
-which will be passed through to \texttt{MPI\_Init()}.
-
-\texttt{int pe\_rank(void)}\\
-Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
-serial.
-
-\texttt{int pe\_size(void)}\\
-Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
-1 in serial.
-
-\texttt{void pe\_finalise(void)}\\
-Responsible for \texttt{MPI\_Finalize()}, and must be the last
-executable statement.
-
-\texttt{void info(const char * fmt, ...)}\\
-Behaves like \texttt{printf}, but only produces output on the
-root process.
-
-\texttt{void verbose(const char * fmt, ...)}\\
-Behaves as \texttt{printf} and reports the rank of the
-calling process in \texttt{MPI\_COMM\_WORLD}.
-
-\texttt{void fatal(const char * fmt, ...)}\\
-Prints an error message and identifies
-the rank of the calling process before terminating the program.
-
-\subsection{Run Time Input}
-
-\texttt{\#include ``runtime.h''}
-
-The ability to set a wide range of parameters at run time is
-very convenient, and is supported via an input file in which the
-user specifies parameters as key value pairs. The code maintains
-a list of key value pairs from which the user can retrieve the
-appropriate values for given keys.
-
-\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
-This reads the named input file and sets up the list of keys.
-
-\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The number
-of key matches found is returned (i.e., 0 or 1).
-
-\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The
-number of key matches is returned.
-
-\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
-Set \texttt{value} to the string assocaited with \texttt{key}, if
-present. Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
-If \texttt{key} is present, set the associated 3-vector of integers.
-Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
-double v[])}\\
-If \texttt{key} is present, set the associated 3-vector of doubles.
-Returns the number of key matches found.
-
-\textit{Comment:} We probably need a function to report unused keys, so
-the user can be alerted if a key has unexpectedly been ignored.
-
-\subsection{Random Number Generators}
-
-\texttt{\#include ``ran.h''}
-
-Note that the random number generation in parallel is
-decomposition dependent.
-
-\texttt{void rand\_init(void)}\\
-Responsible for initialising RNG state.
-
-\texttt{double rand\_serial\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$ from
-the serial generator.
-
-\texttt{double rand\_serial\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the serial generator.
-
-\textit{double rand\_parallel\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$
-from the serial generator.
-
-\texttt{double rand\_parallel\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the parallel generator.
-
-\textit{Comment:} The serial generator will produce the
-same sequence on any MPI process, while the parallel generator
-will produce different sequences. The latter is intended for
-generating, specifically, fluctuations, where the need for
-decomposition-independance is questionable.
-
-\subsection{Timers}
-
-\texttt{\#include ``timer.h''}
-
-Performance figures require timing for different sections
-of the code. The code provides a set of routines to time
-standard and user-defined sections of code. Timers are
-started and stopped by call to the appropriate routines.
-
-\texttt{TIMER\_init()}\\
-This initialises all the timers to zero (e.g., at the start of
-the program).
-
-\texttt{TIMER\_start(const int t\_id)}\\
-Start the given timer.
-
-\texttt{TIMER\_stop(const int t\_id)}\\
-Stop the given timer.
-
-\texttt{TIMER\_statistics(void)}\\
-Prints the current state of all the active timers to standard
-output on the root process in a digestible form. In serial,
-the minimum, maximum, and mean time for each timer is reported
-in seconds (based on the number of start/stop cycles). In
-parallel, the same figures are computed on each process and the
-result of the corresponding \texttt{MPI\_Reduce()} operation is
-printed on the root process.
-
-The code provides a number of timers which would time standard
-parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
-total execution time, and a number of spare timers for users to
-use at will.
-
-\textit{Comment:}
-No intra-run variance (i.e., between time steps) is provided, as the
-calculation required is considerably more complex to generalise, and
-requires memory proportional to the number of time steps used.
-
-\textit{Comment:}
-Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
-provided to insert calls to the above functions for optional timing
-of performance-sensitive regions
-of the code in conjunction with the preprocessor flag
-\texttt{-D\_TIMING\_MACRO\_ON\_}.
-
-\subsection{Model Time Control}
-
-This provides time step control, event frequency (?)
-
-\texttt{void init\_control(void)}\\
-Initialise time controls.
-
-\texttt{int get\_step(void)}\\
-Return the current model time step.
-
-\texttt{int next\_step(void)}\\
-Increments the current time step. Returns number of steps left
-to run (i.e., 0 when finished).
-
-\subsection{Coordinate System}
-
-\texttt{\#include ``coords.h''}
-
-A three-dimensional Cartesian coordinate system is used throughout,
-with the exact system size usually chosen by the user at run time.
-The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
-used to identify the coordinate directions.
-
-This also provides access to the MPI Cartesian Communicator which
-is responsible for most communication in the model.
-
-\texttt{int N\_total(const int dim)}\\
-Returns the total number of lattice sites in the given Cartesian direction.
-
-\texttt{double L(const int dim)}\\
-Returns the length of the system in the given Cartesian direction
-(which is the same as the number of lattice sites).
-
-\texttt{double Lmin(const int dim)}\\
-Returns coordinate position of left-hand edge of system in given
-Cartesian direction.
-
-\texttt{int is\_periodic(const int dim)}\\
-Returns 1 if given Cartesian direction is periodic, else 0.
-
-\texttt{void cart\_init(void)}\\
-Responsible for setting up the Cartesian decomposition.
-
-\texttt{int cart\_rank(void)}\\
-Returns rank in the Cartesian communicator.
-
-\texttt{MPI\_Comm cart\_comm(void)}\\
-Returns the MPI Communicator handle for the Cartesian communicator.
-
-\texttt{int cart\_size(const int dim)}\\
-Returns the size of the Cartesian communicator grid in the given
-coordinate direcxtion.
-
-\texttt{int cart\_coord(const int dim)}\\
-Returns the coordinate of the current process in the Cartesian
-Communicator.
-
-\texttt{int cart\_neigh(const int dir, const int dim)}\\
-Returns the rank of the neighbouring process in the
-Cartesian communicator.
-
-\subsection{Colloids}
-
-\texttt{\#include ``colloids.h''}
-
-These functions provide the basic interface for use
-of colloidal particles in the code.
-
-\subsubsection{Colloid structure}
-
-The basic colloid struct is defined ...
-
-\subsubsection{Colloid interface}
-
-\texttt{void colloids\_init(void)}\\
-Call once for initialisation. This initialises the cell list,
-so the coordinate system must be initialised via a call to
-\texttt{coords\_init()} first.
-
-\texttt{void colloids\_finish(void)}\\
-Call once for finalisation. Frees all colloid-related memory.
-
-\texttt{Colloid * allocate\_colloid(void)}\\
-Returns a pointer to a newly allocated \texttt{Colloid} object
-or fails gracefully.
-
-\texttt{void free\_colloid(Colloid * p\_colloid)}\\
-Deallocates all memory associated with \texttt{p\_colloid} or
-fails gracefully.
-
-\texttt{struct c\_link allocate\_boundary\_link(void)}\\
-Returns a pointer to a newly allocated colloid boundary
-link or fails gracefully.
-
-\texttt{void free\_boundary\_link(struct c\_link *)}\\
-Deallocates memory associated with a link or fails gracefully.
-
-\texttt{void colloids\_memory\_report(void)}\\
-The module keeps track of how much memory has been allocated in
-total for colloids and boundary links. This reports the current
-usage.
-
-\texttt{get\_Ncell\_total()}\\
-
-\texttt{get\_Ncell\_local()}\\
-
-\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
-This returns a pointer to the Colloid at the head of the list at
-cell list location (i, j, k) or \texttt{NULL} if there is none.
-
-\texttt{void cell\_insert(Colloid * p\_colloid)}\\
-This inserts a colloid at the head of the appropriate list
-depending upon its position. This is the only way a colloid
-can be added to the list.
-
-\texttt{update\_cells(void)}\\
-Performs the update of the cell list in response to updated
-particle positions.
-
-\texttt{IVector cell\_coords(FVector r)}\\
-
-\subsection{Speculation}
-
-\subsection{The Free Energy}
-
-It is imagined that a single file will encapsulate all the details
-of a given free energy. The interface for different choices of
-free energy should be the same.
-
-\texttt{FE\_init()}
-
-Responsible for querying the runtime input key value pair list to
-find appropriate user parameters for the free energy, or
-falling back to default values.
-
-\texttt{double FE\_get\_sigma()}
-
-Return the surface tension. There should be a similar routine
-for the interfacial thickness.
-
-\texttt{double FE\_get\_mu(const type location)}
-
-Return the chemical potential at the single lattice
-node \texttt{location}.
-
-\textit{Comment:} The computation of quantities such as the chemical
-potential will a) affect performance and b) require halo values of
-the order parameter (depending on the highest derivative
-in the order parameter present). Performance here should be the
-first consideration, so this needs to be handled with care.
-
-\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}
-
-Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
-given location.
-
-\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
-the number of dimensions. If we are to provide explicit support for
-D2Q9, \texttt{ND} will be 2, otherwise 3.
-
-\section{Implementation Notes}
-
-\subsection{Serial Version}
-
-\subsection{Efficiently Scaling Parallel Computing}
-
-We should be thinking about the 100,000 processor regime, and
-correspondingly large system sizes. To avoid embarrassment in this regime,
-some issues which don't usually arise are relevant, for example:
-
-\begin{enumerate}
-\item
-Large integer quantities such as
-\begin{verbatim}
-   N_total.x*N_total.y*N_total.z
-\end{verbatim}
-are likely to cause integer overflow (if default integers are 4 byte).
-\item
-Local data structures whose size is proportional
-to the number of processors are contraindicated, e.g.,
-anything like
-\begin{verbatim}
-p = (foo *) malloc(nprocs*sizeof(foo));
-\end{verbatim}
-is to be avoided. At the moment, we are in good shape on this
-(apart from some functions in \texttt{misc.c}).
-\item
-I/O is likely become a very important issue, so may need to think
-more about this.
-\item
-We should probably introduce a standard set of benchmarks which can
-be scaled up in some systematic way, so that we can keep track of
-performance over time.
-
-\end{enumerate}
-
-\section{Model Details}
-
-\subsection{D3Q15}
-
-These are the eigenvalues and eigenvectors of teh collision
-matrix used for D3Q15
-
-\begin{table}
-\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - &
- 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
-1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
-3  & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - &
- 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
-3  &$c_{iy}$ \\
-\hline
-$\rho c_{iz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
-3  & $c_{iz}$ \\
-\hline
-$Q_{xx}$ & 1/3 &
--1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
-9/2  & $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{yy}$ & 1/3 &
--1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
-\hline
-$Q_{zz}$ & 1/3 &
--1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
-9  & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
-9  & $c_{iy} c_{iz}$ \\
-\hline
-$Q_{zx}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
-9  & $c_{iz} c_{ix}$ \\
-\hline\hline
-$\chi^1$ & - &
--2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
-1/2 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
-3/2 & $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - &
- 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iy}$\\
-\hline
-$J_{iz}$ & - &
- 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iz}$\\
-\hline
-$\chi^3$ & - &
- 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
-9 & $c_{ix} c_{iy} c_{iz}$ \\
-\hline\hline
-$w_i$ & 1/72 &
-$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
- & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q19}
-
-\begin{table}
-\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
-\hline\hline
-$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
-\hline
-$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
-         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
-& 1\\
-\hline
-$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
-         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
-& 3 \\
-\hline
-$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
-         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
-& 3\\
-\hline
-$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
-         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
-& 3\\
-\hline
-$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
-& 9/2\\
-\hline
-$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
-         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
-& 9\\
-\hline\hline
-$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
-         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 3/4\\
-\hline
-$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
-         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
-         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
-& 3/2\\
-\hline
-$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
-& 9/4\\
-\hline
-$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
-& 9/2\\
-\hline
-$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
-         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 1/2\\
-\hline\hline
-$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
-& \\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{adhikari2005}
-Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
-Fluctuating Lattice Boltzmann,
-\textit{Europhys. Lett.}, \textbf{71}, 473 2005.
-
-\bibitem{ald98}
-Aidun, C.K., Y. Lu, and E.-J. Ding,
-Direct analysis of particulate suspensions with inertia using the
-discrete Boltzmann equation,
-\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
-
-\bibitem{chunladd}
-B. Chun and A.J.C. Ladd,
-Interpolated boundary condition for lattice Boltzmann simulations in
-narrow gaps,
-\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
-
-\bibitem{cj98}
-Cichocki, B., and R.B. Jones,
-Image representation of a spherical particle near a hard wall,
-\textit{Physica A}, \textbf{258}, 273, 1998.
-
-\bibitem{chang}
-C.-H. Chang and E.I. Franses,
-Adsorption dynamics of surfactants at the air/water interface:
-a critical review of mathematical models, data, and mechanisms,
-\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
-
-\bibitem{cpc}
-Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
-LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
-\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
-
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
-\bibitem{diamant96}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid-fluid interfaces,
-\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
-
-\bibitem{eastoe}
-J. Eastoe and J.S. Dalton,
-Dynamic surface tension and adsorption mechanims of surfactants
-at the air-water interface,
-\textit{Advances in Colloid and Interface Science}, \textbf{85}
-13, 2000.
-
-\bibitem{ginzburg}
-I. Ginzburg and D. d'Humi\`eres,
-Multireflection boundary conditions for lattice Boltzmann models,
-\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
-
-\bibitem{h59}
-Hasimoto, H., On the periodic fundamental solutions of the Stokes
-equation and their application to viscous flow past a cubic array
-of spheres.
-\textit{J. Fluid Mech.}, \textbf{5}, 317.
-
-\bibitem{heemels}
-Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
-particles using the lattice-Boltzmann method,
-\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.
-
-\bibitem{jo84}
-Jeffrey, D.J., and Y. Onishi,
-Calculation of the resistance and mobility functions for the two
-unequal rigid spheres in low-Reynolds-number flow,
-\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.
-
-\bibitem{viv}
-Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
-P. Bladon,
-Inertial effects in three dimensional spinodal decomposition of
-a symmetric binary fluid mixture: A lattice Boltzmann study,
-\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).
-
-\bibitem{l94a}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 1. Theoretical foundation,
-\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.
-
-\bibitem{l94b}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 2. Numerical results,
-\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.
-
-\bibitem{l96a}
-Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
-spheres,
-\textit{Phys. Fluids}, \textbf{9}, 491. 1996.
-
-\bibitem{l96b}
-Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
-of non-Brownian spheres,
-\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.
-
-\bibitem{lv01}
-Ladd, A.J.C., and R. Verberg,
-Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
-
-\bibitem{lipanmiller}
-H. Li, C. Pan, and C.T. Miller,
-Pore-scale investigation of viscous coupling effects for two-phase
-flow in porous media,
-\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
-
-\bibitem{nl02}
-Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
-lattice-Boltzmann simulations of particle suspensions,
-\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.
-
-\bibitem{papanastasiou}
-T. paapnastasiou, G. Georgiou, and A. Alexandrou,
-\textit{Viscous Fluid Flow},
-CRC Press, Boca Raton, Florida, 2000.
-
-\bibitem{r95}
-Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
-Cambridge University Press, 1995.
-
-\bibitem{succi}
-S. Succi, \textit{The lattice Boltzmann equation and beyond},
-Oxford University Press, Oxford, 2001.
-
-\bibitem{edo1}
-M. Venuroli and E.S. Boek,
-Two-dimensional lattice-Boltzmann simulations of single phase
-flow in a pseudo two-dimensional micromodel,
-\textit{Physica A}, \textbf{362}, 23, 2006.
-
-\bibitem{vandergraaf}
-R.G.M. van der Sman and S. van der Graaf,
-Diffuse interface model of surfactant adsorption onto flat and
-droplet interfaces,
-\textit{Rheol. Acta} \textbf{46} 3 (2006).
-
-\bibitem{theissengompper}
-O. Theissen and G. Gompper,
-Lattice Boltzmann study of spontaneous emulsification,
-\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
-
-\bibitem{wardtordai}
-A.F.H. Ward and L. Tordai,
-\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
-
-\end{thebibliography}
-
-\end{document}
+This, effectively very small, lattice
+time step is clearly a problem in reaching any respectable 
+macroscopic time in a simulation.
+One step that can be taken is to increase both the lattice
+viscosity and the lattice velocity so that the Reynolds number
+remains unchanged. (Note that the lattice viscosity matching
+the real viscoity might be $(\eta/\rho)T/h^2$ which, using the
+above figures, is $\sim 10^{-4}$.) This can gain 2--3 orders
+of magnitude while keeping the viscosity below $O(1)$. Ultimately,
+it may also be necessary to raise the Reynolds number, which may
+be considered small compared to unity as long as it remains
+in practice less than $O(10^{-1})$ \cite{cates_scaling}.

### 2011-02-11 htdocs/sections.tex
Updates to tex documents
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1271 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  $Id: sections.tex,v 1.13 2010-02-13 15:32:51 kevin Exp $
+%  $Id$
 %
@@
 \begin{center}
-\includegraphics{xfig/fig_le2.eps}
+%\includegraphics{xfig/fig_le2.eps}
 \end{center}
@@
 \begin{center}
-\includegraphics{xfig/fig_le3.eps}
+%\includegraphics{xfig/fig_le3.eps}
 \end{center}

### 2011-02-11 htdocs/user.tex
Updates to tex documents
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1271 40876441-3496-0410-a613-a2fca35b2ae2

@@
 and so on.
+
+\subsubsection{ANSI C}
+
+The code is ANSI C (1989), with the single exception of calls to the
+library function \texttt{erfc()} in the Ewald sum code. This
+can sometimes cause a problem, which can usually be solved with
+a compiler-specific work-around.
 
@@
 
+\inputkey{lc\_redshift\_update [0|1]}
+
+Switches on (default is off) the dynamic computation of the redshift
+from the order parameter gradient terms in the free energy. The
+caluclation is performed at every time step, and override any
+initial value specified in the above. It is also robust to restarts.
+
 \subsection{Colloidal particles}
@@
 
+\subsubsection{Interactions and the cell list}
+
+In each local domain, colloidal particles are stored in a cell list.
+This is a common structure in molecular dynamics-like problems. Here,
+it also serves as the basis for parallel communication of colloid
+information, so there are a number of constraints which it is important
+to understand. In particular, parallel communication imposes the
+constraint that there be at least 2 cells in the local domain in
+each coordinate direction.
+
+The size of the cells is computed automatically by the code based on
+the local domain size, particle size, and the cut-off distance of any
+pairwise interactions that are relevant. This will usually mean trying
+to maximise the number of cells (to reduce pairwise interaction
+calculation). The user must set the minimum allowable cell width.
+
 \inputkey{colloid\_cell\_min}
 
-This sets the minimum cell list width. The cell list has a number
-of purposes: it is used to compute colloid-colloid interactions
-at $O(N)$, but it is also used in the message passing episodes
-for the colloids. This key must be set and there must be at least
-two cells per processor domain.
+This sets the minimum cell list width. This key must be set and there
+must be at least two cells per processor domain. If the minimum does
+not capture the specified interactions, a fatal error will result.
+For spherical particles, the minimum width will typically be
+$2a_h + h_c$ where $h_c$ is an interaction cut-off distance.
+
+In the case where very large particles are required, it can be useful
+to switch of the cell list. This is done via
+
+\inputkey{colloid\_cell\_list\_interactions no}
+
+The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
+there is effectively one cell in each local domain. This option
+should not be required in normal circumstances.
 
@@
 
+The code works out the
+current displacement of the planes by computing $U_{LE} t$, where
+$t$ is the current time step. A shear run should then start from
+$t = 0$, otherwise an offset is required.
+
+\inputkey{LE\_time\_offset}
+
+It is often convenient to run an equilibration with no shear, and
+then to start an experiment after some number of steps. This
+key allows you to offset the start of the Lees-Edwards motion.
+It should then take the value of the start time (in time steps)
+corresponding to the restart at the end of the equilibration
+period. You also need to take this value into account when unrolling
+the output.
+
+\subsubsection{Lees-Edwards planes in parallel}
+
+There are a couple of additional constraints to use the Lees-Edwards
+planes in parallel. In particular, the planes cannot fall at a
+processor boundary in the $x$-direction. This means you should
+arrange an integer number of planes per process in the $x$-direction.
+(For example, use one plane per process; this will also ensure the number
+of planes
+still evenly divides the total system size.)
+This will interleave the planes with the processor decomposition.
+The $y$-direction and $z$-direction may be decomposed without
+further constraint.
+
+Note that this means a simulation with one plane will only work
+if there is one process in the $x$ decomposition.
+
 \subsection{Input/Output}
@@
 
+In serial, each output quantity appears in one file (\texttt{phi-}
+and so on). In parallel, this is also the case by default. However,
+this does not scale, owing to the fact that each process has to
+write to the same file in turn (ie., in serial). The solution is
+to make the I/O parallel. This is done by splitting the domain into
+different ``I/O grids'' each of which writes to a separate file.
+
+The default I/O grid can be set using, e.g.,
+
+\inputkey{default\_io\_grid 2\_2\_2}
+
+meaning that the domain is decomposed into a 2x2x2 Cartesian
+decomposition for I/O purposes. This will give rise to a total
+of 8 files per I/O event with extensions \texttt{.008-001} to
+\texttt{.008-008}. These files must be recombined if analysis
+is required. However, restarting can be performed with the
+parallel files.
+
+Individual lattice quantities can use separate I/O grids. This
+may be particularly useful for the distribution data, which are
+much larger than the other lattice quantities. This means a
+finer decomposition may be appropriate. Clearly, the size of the
+I/O decomposition cannot exceed the processor decomposition.
+See the advice in section \cite{section-advice} for guidelines.
+
 \subsection{Miscellaneous}
@@
 \subsection{Things to Remember and Things to Avoid}
+\ref{section-advice}
 

### 2011-03-04 htdocs/user.tex
Added note on random restart in user.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1291 40876441-3496-0410-a613-a2fca35b2ae2
Shear initialisation comment in user.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1294 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+\subsubsection{Restarting at random!}
+
+Be careful if you are restarting a simulation that used the random
+initialisation at $t=0$. You must change to
+
+\texttt{colloid\_init  from\_file}
+
+to avoid generating a completely new set of random positions
+(which will not be consistent with the density and order parameter
+fields at the restart time). This is almost certainly not what you
+intended!
+
 \subsubsection{Hydrodynamic radius}
@@
 constraint $u_x < c_s$.
+
+There is a further option for use with the special case in which
+top and bottom walls in the $z-direction$ are used to drive a shear
+flow (typically with the top speed positive and the bottom speed
+negative). The key
+
+\inputkey{boundary\_shear\_init}
+
+may be switched on (set to 1) to provide initialisation of the
+LB distributions appropriate for the specified shear rate based
+on a linear velocity profile in the $z-$direction. The default
+is to initialise the fluid at rest.
 

### 2011-05-10 htdocs/user.tex
Added new instructions for walls in user guide
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1349 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-The boundary walls are switched on via the keyword value pair
-
-\inputkey{boundary\_walls\_on [no | yes]}
-
-If \texttt{yes} then the geometry is specified with the keyword
-as one of
-
-\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}
-
-Note that these are the only valid selections. \texttt{1\_1\_0}
-gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction. \texttt{1\_0\_0}
+The boundary walls are switched on via the keyword
+
+\inputkey{boundary\_walls}
+
+and with a complementatary specification of the key \texttt{periodicity}.
+
+For example, walls at $x = x_{\min}$  and $x = x_{\max}$  are specified
+by
+
+\inputkey{boundary\_walls 1\_0\_0}
+
+\inputkey{periodicity 0\_1\_1}
+
+Both must be present. Note that there are a number of constraints on
+valid choices, and the code will stop if there is an invalid choice, or
+the two keys above do not match. Common choices are:
+\texttt{boundary\_walls 0\_0\_1},
+which gives periodic boundaries in the $x-$ and $y-$directions, with
+walls at either edge in the $z-$direction; \texttt{0\_1\_1}
 gives periodic conditions in the $x-$ direction only, while
-\texttt{0\_0\_0} is a fully enclosed box.
-
-In the case of the slab geometry, two further values are used so
-that shear may be imparted to the system:
+\texttt{1\_1\_1} is a fully enclosed box.
+
+In the case of the slab geometry with walls at $z = z_{\min}$ and
+$z= z_{\max}$ (only),
+two further values are used so that shear may be imparted to the system:
 
@@
 These values set the $x-$component of the wall velocity at
-$z = z_{min}$ and $z = z_{max}$, respectively. Like all
+$z = z_{\min}$ and $z = z_{\max}$, respectively. Like all
 velocities, these values must respect the low Mach number

### 2011-05-11 trunk/htdocs/user.tex
docs: added comments on wall lubrication
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1362 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+\subsubsection{Interactions and the cell list}
+
+In each local domain, colloidal particles are stored in a cell list.
+This is a common structure in molecular dynamics-like problems. Here,
+it also serves as the basis for parallel communication of colloid
+information, so there are a number of constraints which it is important
+to understand. In particular, parallel communication imposes the
+constraint that there be at least 2 cells in the local domain in
+each coordinate direction.
+
+The size of the cells is computed automatically by the code based on
+the local domain size, particle size, and the cut-off distance of any
+pairwise interactions that are relevant. This will usually mean trying
+to maximise the number of cells (to reduce pairwise interaction
+calculation). The user must set the minimum allowable cell width.
+
 \inputkey{colloid\_cell\_min}
 
-This sets the minimum cell list width. The cell list has a number
-of purposes: it is used to compute colloid-colloid interactions
-at $O(N)$, but it is also used in the message passing episodes
-for the colloids. This key must be set and there must be at least
-two cells per processor domain.
+This sets the minimum cell list width. This key must be set and there
+must be at least two cells per processor domain. If the minimum does
+not capture the specified interactions, a fatal error will result.
+For spherical particles, the minimum width will typically be
+$2a_h + h_c$ where $h_c$ is an interaction cut-off distance.
+
+In the case where very large particles are required, it can be useful
+to switch of the cell list. This is done via
+
+\inputkey{colloid\_cell\_list\_interactions no}
+
+The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
+there is effectively one cell in each local domain. (In practice,
+the constraint is slightly tighter than this if larger halo regions
+are in effect.) This option
+should not be required in normal circumstances.
 
@@
 
-The boundary walls are switched on via the keyword value pair
+The boundary walls are switched on via the keyword
 
-\inputkey{boundary\_walls\_on [no | yes]}
+\inputkey{boundary\_walls}
 
-If \texttt{yes} then the geometry is specified with the keyword
-as one of
+and with a complementatary specification of the key \texttt{periodicity}.
 
-\inputkey{periodicity [1\_1\_1 | 1\_1\_0 | 1\_0\_0 | 0\_0\_0]}
+For example, walls at $x = x_{\min}$  and $x = x_{\max}$  are specified
+by
 
-Note that these are the only valid selections. \texttt{1\_1\_0}
-gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction. \texttt{1\_0\_0}
+\inputkey{boundary\_walls 1\_0\_0}
+
+\inputkey{periodicity 0\_1\_1}
+
+Both must be present. Note that there are a number of constraints on
+valid choices, and the code will stop if there is an invalid choice, or
+the two keys above do not match. Common choices are:
+\texttt{boundary\_walls 0\_0\_1},
+which gives periodic boundaries in the $x-$ and $y-$directions, with
+walls at either edge in the $z-$direction; \texttt{0\_1\_1}
 gives periodic conditions in the $x-$ direction only, while
-\texttt{0\_0\_0} is a fully enclosed box.
+\texttt{1\_1\_1} is a fully enclosed box.
 
-In the case of the slab geometry, two further values are used so
-that shear may be imparted to the system:
+In the case of the slab geometry with walls at $z = z_{\min}$ and
+$z= z_{\max}$ (only),
+two further values are used so that shear may be imparted to the system:
 
@@
 These values set the $x-$component of the wall velocity at
-$z = z_{min}$ and $z = z_{max}$, respectively. Like all
+$z = z_{\min}$ and $z = z_{\max}$, respectively. Like all
 velocities, these values must respect the low Mach number
 constraint $u_x < c_s$.
+
+There is a further option for use with the special case in which
+top and bottom walls in the $z-direction$ are used to drive a shear
+flow (typically with the top speed positive and the bottom speed
+negative). The key
+
+\inputkey{boundary\_shear\_init}
+
+may be switched on (set to 1) to provide initialisation of the
+LB distributions appropriate for the specified shear rate based
+on a linear velocity profile in the $z-$direction. The default
+is to initialise the fluid at rest.
+
+\subsubsection{Flat wall lubrication correction}
+
+It is possible to prevent colloids colliding with the flat boundary walls
+by adding the correction to the normal component of the lubrication
+force. The general form of this correction is
+\begin{equation}
+\mathbf{F}_{\mathrm{lub}} = -6\pi \eta a_h^2
+%({\textstyle \frac{1}{h} - \frac{1}{h_c}})
+(1/h - 1/h_c)
+\mathbf{u} . \mathbf{r}.
+\end{equation}
+Here, $h$ is the surface to surface separation, and $h_c$ is the cutoff
+distance beyond which the correction is not applied. The relative
+velocity (assuming the wall is not moving) and normal separation are
+$\mathbf{u}$ and $\mathbf{r}$ respectively.
+
+The cutoff distance can be set from the input via
+
+\inputkey{boundary\_lubrication\_rcnormal 0.1}
+
+where the value is (strictly) set via calibration, cf. the hydrodynamic
+radius. In general, the cutoff is between zero and 0.5 lattice spacing,
+and can be smaller for larger particles. Because the wall velocity is
+fixed, this force correction should cause no stability issues in the
+colloid velocity update (provided a colloid's initial position is not
+very close to the wall). By default the cutoff $h_c = 0$.
 
@@
 
-Note that if something more complicated is required, for example,
-an equilibration at zero shear followed by a run with shear, one
-needs to be careful how this is arranged. The code works out the
+The code works out the
 current displacement of the planes by computing $U_{LE} t$, where
 $t$ is the current time step. A shear run should then start from
-$t = 0$, otherwise an offset is required. At the moment this cannot
-be done via the input.
+$t = 0$, otherwise an offset is required.
+
+\inputkey{LE\_time\_offset}
+
+It is often convenient to run an equilibration with no shear, and
+then to start an experiment after some number of steps. This
+key allows you to offset the start of the Lees-Edwards motion.
+It should then take the value of the start time (in time steps)
+corresponding to the restart at the end of the equilibration
+period. You also need to take this value into account when unrolling
+the output.
 

### 2011-06-21 htdocs/plan.tex
New documentation source files
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1417 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  plan.tex
%
%  Software Configuration Management Plan
%
%  $Id$
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2011 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% This is a section to be included in ludwig.tex

\section{Software Configuration Management Plan}

\subsection{Introduction}

\subsubsection{Purpose}

This section contains information on the \textit{Software Configuration
Management Plan} for the \textit{Ludwig} code.
It is to provide users of the code a background on the way in which
the code is developed, how bugs are dealt with, the way in which
new features may be added, and so on. It is therefore a part
of the process by which the code is maintained. The Software Configuration
Management Plan will be referred to as `The Plan' throughout the remainder
of the section.

\subsubsection{Scope}

The \textit{Ludwig} code is designed to study the hydrodynamic properties
of simple and complex fluids based around the numerical solution of the
Navier-Stokes equations. Complex fluids are dealt with via a free-energy
based finite difference approach, which couples explicitly to the
hydrodynamics. Examples of complex fluids include binary mixtures,
surfactants, liquid crystals, and active gels. Solid objects such as
porous media may be included, as well as moving particles (colloids).
Components of \textit{Ludwig} include the main program, unit and
regression tests, and a small number of utility programs used to prepare
input and post-process output. The Plan is limited to these
software components.

\textit{Ludwig} is specifically designed for parallel computers and
relies on the Message Passing Interface \cite{mpi-standard}. The
code will also run in serial, and is designed to be independent of
the platform it is run on. The Plan is therefore not concerned with
hardware or system management activities.

This document is meant to be fairly informal, but will be updated as
the need arises. 
\textit{Ludwig} is an on-going research project, and relies mainly
on funding for the United Kingdom Engineering and Physical Sciences Research
Council to provide support for staff time to work on maintenance and
development. As such, we cannot commit to any and all user requests!

\subsubsection{Definition of key terms}

\subsubsection{References}

This section is based on the IEEE standard for Software Configuration
Management Plans IEEE 828-2005 \cite{ieee-208}, and follows the format
set out therein.
Relevant references are included and can be found in the main References
section at the end of the document.

\subsection{Management}

\subsubsection{Organisations}

\textit{Ludwig} is developed by a team based in the School of Physics at
The University of Edinburgh. This involves two groups: the Soft Matter
Physics Group under the leadership of Prof. Mike Cates FRS, and EPCC in
the person of Kevin Stratford. We collaborate with a number of workers at
different Universities around the world on different aspects of the code.

\subsubsection{Responsibilities}

While the Soft Matter Physics Group is responsible for the overall
scientific direction and development of the code and concomitant
priorities, all
the Software Configuration Management activities will be the
responsibility of EPCC.

\subsubsection{Procedures}

There are currently no external constraints on The Plan.

\subsubsection{Process Management}

EPCC is responsible for the Software Configuration Management process.
It is anticipated that the cost of this process will be negligible, and
will not require monitoring. There is currently no independent
surveillance of activities to ensure compliance with The Plan.

\subsection{Activities}

\subsubsection{Identification}

The material under control of the project includes three sets of files:
this documentation, the source code, and the test suite which is used
to monitor the status of the code.
Control of project files is via a single on-line repository which
uses the subversion (SVN) revision control system. The SVN repository
is currently located at

\texttt{http://ccpforge.cse.rl.ac.uk/}

which is maintained at the Rutherford Appleton Laboratory on behalf
of Collaborative Computational Physics 5 (The Computer Simulation of
Condensed Phases). Subversion identifies different revisions by a
unique revision number. A given version of a file is then uniquely
identified by its path in the repository as it exists at a given
revision number. For example, SVN revision 1416 contains the
following directories:

\texttt{/trunk/current} a precursor code not under control

\texttt{/trunk/htdocs} source files for this documentation

\texttt{/trunk/mpi\_s} MPI stub library (scheduled to move)

\texttt{/trunk/src} source code

\texttt{/trunk/tests} unit and regression test suite

\texttt{/trunk/util} utility programs

At present access is restricted. However, we have committed to making
the code available to the public by October 2014. We expect this to
be via the same repository, referred to as CCPForge.
All members of the \textit{Ludwig} project are currently registered with
a CCPForge user account, and fall into one of three categories:
Administrator, Developer, and User. Only the first two categories have
permission to change files in the repository. Users have read-only
access. The exact mechanism for public access is yet to be decided,
but may still require users to register with CCPForge.

As CCPForge is not physically under our control, we undertake a
complete dump of the repository itself to a local resource on a
weekly basis for security.

\subsubsection{Control}

Requests for changes or additions to the code should be made via
the issue tracker at CCPForge. The project team will then decide
whether the change is possible. If so, the implementation and
testing of the change will take place, and the new code committed
back to the repository.

\subsubsection{Status}

Can we account for the status of the code and tests?

\subsubsection{Evaluation and review}

How do we evaluate and audit the code?

\subsubsection{Interface control}

At the moment, we do not explicitly interface to any code outside The
Plan (except MPI). We could extend coverage to include data export
to e.g., Paraview \cite{paraview}.

\subsubsection{Subcontractor / vendor control}

None required.

\subsubsection{Release management and delivery}

There is currently no formal release mechanism. If such a mechanism were
introduced, management and  delivery would be via the SVN repository.

\subsection{Schedules}

Sequence of events for SCM process.

\subsection{Resources}

For each active, state what resources (infrastructure, software tools,
personnel etc) are required.

\subsection{Plan Maintenance}

The Plan is currently in a draft form. A complete version of The Plan
will be in place to coincide with the public release.

### 2011-06-21 trunk/htdocs/anchoring.tex
New documentation source files
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1417 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  anchoring.tex
%
%  $Id$
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2011 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Cholesteric Anchoring}

We have a fluid free energy density
\begin{equation}
f = 
{\textstyle\frac{1}{2}} \kappa_0 (\partial_\beta Q_{\alpha\beta})^2
+ {\textstyle\frac{1}{2}}
 \kappa_1 (\epsilon_{\alpha\gamma\sigma} \partial_\gamma
Q_{\sigma\beta} + 2q_0 Q_{\alpha\beta})^2
\end{equation}
where we have ignored the bulk terms for the time being, but we retain
two elastic constants $\kappa_0$ and $\kappa_1$ in the distortion term.
The cholesteric pitch $p = 2\pi/q_0$.

There is also a surface term (an area density)
\begin{equation}
f_s = {\textstyle\frac{1}{2}} w (Q_{\alpha\beta} - Q_{\alpha\beta}^0)^2
\end{equation}
where $Q^0_{\alpha\beta}$ is the preferred order parameter configuration
at the surface (to be either normal or planar).

The boundary condition we wish to apply is, for the bulk term
\begin{equation}
n_\gamma \frac{\partial f}{\partial Q_{\alpha\beta,\gamma}} = 0
\end{equation}
where $Q_{\alpha\beta,\gamma} = \partial_\gamma Q_{\alpha\beta}$. Here,
$n_\gamma$ is the outward unit normal at the surface (pointing into the
fluid). With the
addition of the surface term we obtain
\begin{equation}
n_\gamma \frac{\partial f}{\partial Q_{\alpha\beta,\gamma}}
+ \frac{\partial f_s}{\partial Q_{\alpha\beta}} = 0.
\end{equation}
The derivative of the distortion free energy with respect to
$Q_{\alpha\beta,\gamma}$ gives
\begin{equation}
\kappa_0 n_\beta \partial_\gamma Q_{\alpha\gamma}
+ \kappa_1 n_\gamma
(\partial_\gamma Q_{\alpha\beta} - \partial_\alpha Q_{\gamma\beta})
- 2\kappa_1 q_0 n_\gamma \epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}.
\end{equation}
However, if we want a symmetric form (the derivative with respect to
$Q_{\beta\alpha,\gamma}$ is just as good) we can write this as
\begin{eqnarray}
{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
+ n_\beta \partial_\gamma Q_{\alpha\gamma})
+ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
+ \partial_\beta Q_{\gamma\alpha})
\nonumber
\\
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha}).
\end{eqnarray}
If we add the derivative of the surface free energy,
where we assume that the preferred orientation $Q_{\alpha\beta}^0$ is
independent of $Q_{\alpha\beta}$, we get a full boundary condition
for the gradient of the tensor order parameter at the solid-fluid boundary:
\begin{eqnarray}
{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
+ n_\beta \partial_\gamma Q_{\alpha\gamma})
+ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
+ \partial_\beta Q_{\gamma\alpha})
\nonumber
\\
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
- w(Q_{\alpha\beta} - Q_{\alpha\beta}^0) = 0.
\label{cholesteric_bc}
\end{eqnarray}

\subsection{Discrete implementation}

In three dimensions, the boundary condition Eq.~\ref{cholesteric_bc}
contains (potentially) 15 unknown derivatives
$\partial_\gamma Q_{\alpha\beta}$
corresponding to the 5 independent elements of the order parameter
tensor $Q_{xx}$, $Q_{xy}$, $Q_{xz}$, $Q_{yy}$, and $Q_{yz}$. At a
flat surface with, e.g., $n = (1, 0, 0)$, this is reduced to five unknown
gradients $\partial_x Q_{\alpha\beta}$ at the
boundary if we assume the tangential gradients $\partial_y Q_{\alpha\beta}$
and $\partial_z Q_{\alpha\beta}$ can be approximated
using the standard differencing method involving only fluid values of
$Q_{\alpha\beta}$. We proceed by
computing the constant terms
$$
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
- w(Q_{\alpha\beta} - Q_{\alpha\beta}^0)
$$
using $Q_{\alpha\beta}$ from the adjacent fluid site, and an estimate
of $Q^0_{\alpha\beta}$ for the appropriate anchoring type. To these
constant terms are added the tangential gradients.
The gradients at the surface are then computed by solving a 5x5
linear algebra problem for $\partial_x Q_{\alpha\beta}$. This
allows the full gradient at the adjacent fluid site to be constructed.

At concave edges or corners, where it is not possible to compute the
tangential gradients from the usual stencil as for a flat interface,
a different approach is required. We note than an attempt to solve a
10x10 or 15x15 linear algebra problem for the full set of unknown
gradients has proven unreliable: in the case where $n_\gamma$ is
normal to the coordinate directions there is simply no solution
available. Instead we adopt an iterative approach.

Consider the case of an edge where the solid neighbours
are in the $x$ and $y$ directions.  We first make an approximation
to the tangential gradients in the $x$ and $y$ directions by using
a one-sided derivative from the fluid points available. This
tangential approximation in $y$ is then used to find a first
estimate of the gradients $\partial_x Q_{\alpha\beta}$ using the
procedure described above. The tangential estimate in $x$ is then
used to obtain $\partial_y Q_{\alpha\beta}$. These surface gradients
are then used to improve the estimate of the tangential gradients,
and the process is iterated.

{\small
\begin{table}
\begin{center}

\begin{tabular}{|c|ccccc|}

\hline
&
$Q_{xx,x}$ & $Q_{xy,x}$ & $Q_{xz,x}$ & $Q_{yy,x}$ & $Q_{yz,x}$ \\
\hline
$Q_{xx}$ &
$\kappa_0 n_x$ & $-\kappa_1 n_y$ & $-\kappa_1 n_z$ & & \\
$Q_{xy}$ &
$\kappa_0 n_y$ & $\kappa' n_x$ & & $-\kappa_1 n_y$  & $-\kappa_1 n_z$  \\
$Q_{xz}$ &
$\kappa' n_z$ & & $\kappa' n_x$ & $\kappa_1 n_z$ & $-\kappa_1 n_y$ \\
$Q_{yy}$ &
 & $\kappa_0 n_y$ & & $\kappa_1 n_x$ & \\
$Q_{yz}$ &
 & $\kappa_0 n_z$ & $\kappa_0 n_y$ & & $2\kappa_1 n_x$ \\
\hline
\hline
&
$Q_{xx,y}$ & $Q_{xy,y}$ & $Q_{xz,y}$ & $Q_{yy,y}$ & $Q_{yz,y}$ \\
\hline
$Q_{xx}$ &
$\kappa_1 n_y$ & $\kappa_0 n_x$ & & & \\
$Q_{xy}$ &
$-\kappa_1 n_x$ & $\kappa' n_y$ & $-\kappa_1 n_z$ & $\kappa_0 n_x$ & \\
$Q_{xz}$ &
 & $\kappa_0 n_z$ & $2\kappa_1 n_y$ & & $\kappa_0 n_x$ \\
$Q_{yy}$ &
 & $-\kappa_1 n_x$ & & $\kappa_0 n_y$ & $-\kappa_1 n_z$\\
$Q_{yz}$ &
$\kappa_1 n_z$ & & $-\kappa_1 n_x$ & $\kappa' n_z$ & $\kappa' n_y$\\
\hline
\hline
&
$Q_{xx,z}$ & $Q_{xy,z}$ & $Q_{xz,z}$ & $Q_{yy,z}$ & $Q_{yz,z}$ \\
\hline
$Q_{xx}$ &
$\kappa_1 n_z$ & & $\kappa_0 n_x$ & & \\
$Q_{xy}$ &
 & $2\kappa_1 n_z$ & $\kappa_0 n_y$ & & $\kappa_0 n_x$ \\
$Q_{xz}$ &
$-\kappa' n_x$ & $-\kappa_1 n_y$ & $\kappa' n_z$ & $-\kappa_0 n_x$ & \\
$Q_{yy}$ &
 & & & $\kappa_1 n_z$ & $\kappa_0 n_y$  \\
$Q_{yz}$ &
$-\kappa_0 n_y$ & $-\kappa_1 n_x$ & & $-\kappa' n_y$ & $\kappa' n_z$ \\
\hline
\end{tabular}
\end{center}
\caption{Coefficients of the various derivatives of the order parameter
tensor appearing in five equations for the independent elements of the
order parameter. Note $\kappa_0 + \kappa_1 = \kappa'$ and all the
coefficients have been multiplied by a factor of 2.}
\label{tab:cholesteric_bcs}
\end{table} 
}

### 2011-06-21 trunk/htdocs/references.tex
New documentation source files
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1417 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-02-11 htdocs/sections.tex
@@
 %
-%  sections.tex
+%  references.tex
 %
-%  Content for Ludwig documentation.
+%  Bibliography
 %
@@
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
+%  (c) 2011 The University of Edinburgh
 %
@@
 
-\section{The Lattice Boltzmann Method}
-
-We review here the lattice Boltzmann method applied to single
-phase problems.
-
-\subsection{The Navier Stokes Equation}
-
-We seek to solve the isothermal Navier-Stokes equations which, often
-written in vector form, express mass conservation
-\begin{equation}
-\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
-\label{eq_mass1}
-\end{equation}
-and the conservation of momentum
-\begin{equation}
-\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
--\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
-+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
-\label{eq_momentum1}
-\end{equation}
-Equation~\ref{eq_mass1} expresses the local rate of change of the
-density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
-mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
-Equation~\ref{eq_momentum1} expresses Newton's second law for
-momentum, where the terms on the right hand side represent the
-force on the fluid.
-
-For this work, it is more convenient to rewrite these equations
-in tensor notation, where Cartesian coordinates ${x,y,z}$ are
-represented by indices $\alpha$ and $\beta$, viz
-\begin{equation}
-\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
-\end{equation}
-and
-\begin{equation}
-\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
-= -\nabla_\alpha p
-+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
-+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
-\end{equation}
-Here, repeated indices are understood to be summed over.
-The conservation law is seem better if the forcing terms of the
-right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
-so that
-\begin{equation}
-\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
-\end{equation}
-In this case the expanded expression for the stress tensor is
-\begin{equation}
-\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
-+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
-\end{equation}
-where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
-$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
-modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.
-
-\subsection{The Distribution Function and its Moments}
-
-In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
-\begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
-\end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
-
-In general, the number of modes supported in $n$ and these can
-be written as
-\begin{equation}
-M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
-\end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
-\begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
-\end{equation}
-
-\subsubsection{Hydrodynamic modes}
-
-\subsubsection{D3Q15 ghost modes}
-
-\subsubsection{D3Q19 ghost modes}
-
-\subsection{The Lattice Boltzmann Equation}
-
-\subsection{The Collision}
-
-\subsection{External Body Forces}
-
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsubsection{Fluctuating ghost modes: D3Q15}
-
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Boundary walls}
-
-\subsection{Colloids}
-
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
-\section{Polar Active Gel}
-
-Here we consider a polar vector field model to study active gels. 
-The order parameter is this time a vector $P_\alpha$
-(with variable magnitude) as there is no longer head-tail symmetry
-in the system as was the case for the ${\bf Q}$ tensor formulation. 
-The equation of motion as:
-\begin{equation}
-\partial_t P_\alpha  \partial_\beta (u_{\beta} + wP_{\beta}) P_{\alpha}
-= \lambda D_{\alpha\beta} P_{\beta} - \omega_{\alpha\beta} P_{\beta}
-+\Gamma' h_{\alpha}.
-\end{equation}
-
-In this equation, $w$ is an active term, due to swimming, which causes
-self-advection of the order parameter, while $\lambda$ is a material
-dependent constant -- positive for rod-like molecules. If $|\lambda|>1$
-the liquid crystalline passive phase is flow-aligning, otherwise it is
-flow-tumbling.
-
-Note that the convection for the Leslie-Ericksen is that the
-velocity gradient tensor should be writen $W_{\alpha\beta}
-= \partial_\alpha u_\beta$ (and not the other way round, which
-is usual).
-
-The molecular field is given by
-$h_{\alpha}=-\delta {\cal F}_{\rm pol}/\delta p_{\alpha}$
-where ${\cal F}_{\rm pol}$ is the free energy for a polar active nematic,
-whose density we can take to be:
-\begin{equation}
-f =  \frac{\alpha_2}{2} |{\mathbf P}|^2 + \frac{\alpha_4}{4}|{\mathbf P}|^4
-+ \frac{K}{2}\left(\partial_\alpha P_{\beta}\right)^2 + 
-\frac{K_{LC}}{2}\left[\partial_\alpha (P_{\beta}P_{\gamma})\right]^2 
-\end{equation}
-where $K$ and $K_{LC}$ are elastic constants, $\alpha_4>0$, and $\alpha_2$ 
-can be positive (isotropic phase) or negative (polar nematic phase).
-
-The molecular field is explicitly given by:
-\begin{equation}
-h_{\alpha}=-\alpha_2 P_{\alpha} -\alpha_4 P_{\alpha} P_{\beta}P_{\beta}+
-K \partial_{\beta}\partial_{\beta} P_{\alpha}+2 K_{LC}
-P_{\gamma}\partial_{\beta}\partial_{\beta} (P_{\gamma}P_{\alpha}).
-\end{equation}
-
-The Navier-Stokes equation is:
-\begin{equation}
-\rho \left(\partial_t+v_{\beta}\partial_{\beta}\right) v_{\alpha} =
-\partial_{\beta} \Pi_{\alpha\beta} 
-+\eta \partial_{\beta}\partial_{\beta} v_{\alpha}.
-\end{equation}
-
-The stress tensor this time is 
-\begin{eqnarray}
-\Pi_{\alpha\beta} =
-\frac{1}{2}\left(P_\alpha h_\beta - P_\beta h_\alpha \right)
-- \lambda\left( (P_\alpha h_\beta + P_\beta h_\alpha)/2 
-- (1/3)P_\gamma P_\gamma \delta_{\alpha\beta} \right)
-\\
--  \zeta ( P_\alpha P_\beta - (1/3)P_\gamma P_\gamma \delta_{\alpha\beta})
-- K \partial_\alpha P_\gamma \partial_\beta P_\gamma.
-\end{eqnarray}
-The active term in the stress tensor, proportional
-to $\zeta$, has the same form for apolar and polar active gels.
-
-\section{Code Design}
-
-\subsection{Some Comments}
-
-In designing a code in C the main object of which is to do high
-performance computing, there are some inevitable tensions with
-software engineering best-practice. In particular, complete
-encapsulation of data in the spirit of using abstract data types
-tends to cause problems of performance.
-
-For example, encapsulation of the lattice Boltzmann distribution
-$f_i$, together with appropriate ``get'' and ``set'' functions for
-access when approapriate, is possible. There could then be a
-complete separation of interface and implementation of the
-distribution (other than that each individual $f_i$ is represented
-by a \texttt{double}). However, experience suggests that
-this imposes severe impediment to optimisation and hence
-unacceptable performance degradation.
-
-Schemes do exist by which one can approximate the object
-oriented programming style in C, and so achieve a separation
-of interface and implementation. However, this tends to put
-a heavy burden on the use of \texttt{void} pointers, itself
-a recipe for possible disaster. A truly object-oriented
-language such as Java would be preferable.
-
-So, for the time being, some relaxation of encapsulation is required.
-Exposing a pointer to the distributions which allows them to be accessed
-directly where required. This can be done merely by declaring the
-necessary variables \texttt{extern}. In practice, such deviations
-from encapsulation can be minimised in the lattice sector of the
-code. In particlar, \texttt{extern} declarations should not be
-required in \texttt{main()}, i.e., it should be able to write a
-new application without worrying about details of implementation.
-
-\subsubsection{Objects}
-
-Somewhat more difficult is the part of the code involving
-colloids, which would really favour a truly object-oriented
-approach.
-
-\subsubsection{Serial and parallel}
-
-The design decission has been taken to support execution of the
-code in both serial and parallel. The default compilation
-assumes that a serial executable is required, while  MPI code
-is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).
-
-There is no shared memory inplementation at this time.
-
-\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}
-
-\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}
-
-\begin{figure}
-\begin{center}
-%\includegraphics{xfig/sketch.eps}
-\end{center}
-\caption{A scheme by which different parts of the code might be
-encapsulated and in such a way to allow reliable development and
-testing. As this is C, will do not adopt a truly ``object'' langauge.
-Broadly, different modular parts should only depend on stuff above
-in the diagram. So, at the top, there is the parallel environment
-where communication occurs within MPI\_COMM\_WORLD. Built on top
-of this is the Cartesian Communicator responsible for halo swaps
-and so forth. The LE code may have a separate Communicator again. 
-(There's an argument to say the diagram should be the other way up!)}
-\end{figure}
-
-\section{Code Implementation}
-
-\subsection{Stuff}
-
-\subsection{Coordinate System}
-
-\subsubsection{The global coordinate system}
-
-The user specifies the system size in lattice units at run time,
-which sets the extent of memory which must be allocated for the
-Cartesian lattice. The extent of the physical coordinate system
-then reflects the lattice size directly (see Figure \ref{fig_c1}). 
-Each lattice site, or node,  has coordinates corresponding to its
-integer position in the system and can be thought of as being
-at the centre of a control volume one lattice unit on a side.
-This means that the nominal edges of the
-system are offset from the lattice sites themselves by half a
-lattice unit in each direction.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_c1.eps}
-\end{center}
-\caption{The global Cartesian coordinate system reflects the choice
-of lattice size; lattices sites are located at
-$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
-(the Figure only shows the labels in the $y-$direction for simplicity).
-In the Figure, the lattice sites in the domain proper are the full
-circles; halo sites, which are used to effect periodic boundary
-conditions, are included as open cirles}
-\label{fig_c1}
-\end{figure}
-
-For particles, the centres of which can move continuously across the
-lattice, the extents of the global coordinate system are then
-$L_{min} < x < L + L_{min}$ and so on.
-
-\subsubsection{Addressing the lattice: the local coordinate system}
-
-\subsection{Solid walls}
-
-General solid objects which are fixed for the duration of
-the simulation, such as porous media, can be defined
-by assigning solid/fluid status to each lattice node. 
-The links required between solid and fluid nodes can
-then be identified for the given model.
-
-\textit{Comment.} In general, a solid object will not
-affect the periodic nature of the simulation. A general
-solid object cannot move, i.e., the boundary velocity
-must be zero.
-
-\textit{Special case.} In the case where the solid object
-represents a flat boundary wall at the edge of the system
-in one or more directions, the periodicity is broken in
-the corresponding directions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LE Implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Lees Edwards Sliding Period Boundary Conditions}
-
-\subsubsection{The LE module}
-
-\subsubsection{Distributions crossing the LE planes}
-
-With the planes conceptually half-way between lattice sites, the
-propagation moves some of the distributions (namely, those with
-$c_x = \pm1$) between different sliding blocks. While the collision
-stage is unaffected, action must be taken to adjust these
-distributions when the LE boundary conditions are active. This
-is done in a two-stage process described by Adhikari \textit{et al.}
-\cite{adhikari_desplat}, implemented between the collision and
-propagation.
-
-\textbf{a. Reprojection} The post-collision distributions with
-$c_x = \pm 1$ at sites adjacent to a boundary are modified to
-take account of the velocity jump $\pm u^{LE}_y$ between the sliding
-blocks. In terms of the hydrodynamic moments we have:
-\begin{eqnarray}
-\rho &\rightarrow& \rho', \\
-\rho u_\alpha &\rightarrow& (\rho u_\alpha)' \pm \rho' u^{LE}_\alpha, \\
-S_{\alpha\beta} &\rightarrow&
-S'_{\alpha\beta} \pm (\rho u_\alpha)' u^{LE}_\beta
-\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
-\end{eqnarray}
-If we work with the changes to the moemnts, then the density is
-unaffected $\delta\rho =  0$, the velocity is changed by
-$\delta u_\alpha = \pm u^{LE}_\alpha$, with an analogous expression for
-the change in the stress
-\begin{equation}
-\delta S_{\alpha\beta} =
-\pm (\rho u_\alpha)' u^{LE}_\beta
-\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
-\end{equation}
-We can then work out the change in the distributions by reprojecting
-via Eq.\ref{eq:}, i.e.,
-\begin{equation}
-f_i \rightarrow f_i' + w_i \Bigg(
-\frac{\rho \delta u_\alpha c_{i\alpha}}{c_s^2} +
-\frac{\delta S_{\alpha\beta}Q_{i\alpha\beta}}{2c_s^4} \Bigg).
-\end{equation}
-A similar set of expressions are given for the order parameter
-distributions in Adhikari \textit{et al.} \cite{adhikari_desplat}.
-
-\textbf{b. Interpolation of reprojected distributions}
-As the sliding blocks are displaced continuously with
-$\delta y = u_y^{LE}t$, which is not in general a whole number of
-lattice spacings, an interpolation is also required before
-propagation can take place. Consider the situation (Fig.~\ref{fig_le2}) from
-a stationary frame where the frame above has translated a small
-distance to the right ($\delta y < \Delta y$). In the stationary frame
-we must interpolate the ditributions with $c_x = 1$ to a
-`departure point' from which the propagation will move the
-distribution exactly to the appropriate lattice site in the moving
-frame. Schemeatically,
-\begin{equation}
-f_i'(x, y + \delta y, z) = 
- (1 - \delta y) f_i^\star(x, y, z) +
-\delta y f_i^\star(x, y + \Delta y, z),
-\end{equation}
-where $f_i'$ here represents the interpolated value and $f_i^\star$
-is the reprojected post-collision quantity.
-
-In the case where the displacement $\delta y > 1$, the relevant
-lattice sites involved in the interpolation are displaced to the right
-by the integral part of $\delta y$, while the relative weight given to
-the two sites involves
-the fractional part of $\delta y$. To take account of the periodic
-boundary conditions
-in the $y$-direction, all displacements are modulo $L_y$.
-
-\begin{figure}[h]
-\begin{center}
-%\includegraphics{xfig/fig_le2.eps}
-\end{center}
-\caption{Schematic diagram showing interpolation of distributions in the
-stationary plane (below) in preparation for propagation into the moving
-frame (above). For a small displacement $\delta y = u^{LE} t$, the
-interpolation takes place between lattice sites in the stationary frame
-(green, left). Propogation from the interpolated positions (red) can
-then take place into the moving frame (right).}
-\label{fig_le2}
-\end{figure}
-
-In parallel, exactly the same calculation is undertaken. However,
-communication is required to take account of the processor
-decomposition in the $y$-direction. This requires the identification
-of the two processes which hold the relevant information as a function
-of displacement.
-
-Note that the interpolated values are stored in the appropriate lattice
-site in the rest frame that ensures the normal propagation stage moves all
-distributions to the correct destination in memory.
-
-\subsubsection{Order parameter gradients}
-
-When the order parameter gradient is computed at a given point
-near the LE boundaries, the stencil may extend out of the stationary
-frame. The process of interpolation is slightly different to that
-for the distributions.
-
-Here, we need to interpolate values in the moving frame to the position
-which is seen by the rest frame. Figure~\ref{fig_le3} shows a five-point
-stencil to calculate the gradient at the central point $(x,y,z)$ in the
-rest frame. An interpolation of the $\phi$-field of the form
-\begin{equation}
-\phi'(x + \Delta x, y - \delta y, z) =
-\delta y \phi(x + \Delta x,  y - \Delta y, z) + 
-(1 - \delta y) \phi(x + \Delta x, y, z)
-\end{equation}
-must then take place. In practice, the interpolated values are stored
-in the offset buffer. As the gradient calculation requires $\phi$
-values in the halo regions, interpolated values for the buffer halo
-region are also required.
-
-\begin{figure}[h]
-\begin{center}
-%\includegraphics{xfig/fig_le3.eps}
-\end{center}
-\caption{Schematic diagram showing a five-point stencil to compute a
-gradient at the central point in the stationary frame. The upper point
-of the stencil is in the moving frame and so must be interpolated to
-the correct position (red sites).}
-\label{fig_le3}
-\end{figure}
-
-In parallel, the interpolation requires data from at most two adjacent
-processes in the along-plane $y-$direction as long as the $\phi$ halo
-regions are up-to-date. The rank of the processes involved is
-determined by the displacement $\delta y$ as a function of time.
-
-\subsubsection{Velocities and finite-difference fluxes}
-
-For the finite-difference implementation, the velocity field at
-the cell boundaries is required. At the planes, this means an
-interpolation which takes the same form as that for
-$\phi$ is needed.
-
-In computing the advective fluxes between cells, we must then
-allow for the presence of the planes. This is done by computing
-separately the 'east' and 'west' face fluxes in the $x$-direction
-(the flux that crosses the plane). Away from the planes, face-flux
-uniqueness $f_{ijk}^e = f_{i+1jk}^w$ ensures conservation of order
-parameter. However, at the planes, the non-linear conbination of
-velocity field and order parameter field does not in general give
-rise to consistent fluxes. In order to restore conservation, we
-reconcile east and west face fluxes at the plane by taking an
-average
-\begin{eqnarray}
-f_{ijk}^{e\star}&=&{\textstyle\frac{1}{2}}(f_{ijk}^e + f_{i+1j'k}^w),\quad 
-f_{i+1j'k}^w  = \delta y f_{i+1 j-1 k}^w + (1 - \delta y) f_{i+1jk}^w,\\
-f_{i+1 j k}^{w\star} &=& {\textstyle\frac{1}{2}}(f_{ij'k}^e + f_{i+1jk}^w),\quad
-f_{ij'k}^e = (1 - \delta y) f_{ijk}^e + \delta y f_{ij+1k}^e.
-\end{eqnarray}
-In this way, we average the east face flux in the rest frame $f_{ijk}^e$
-with the interpolated value of the west face flux in the moving frame
-$f_{i+1j'k}^w$, and vice-versa. One can then easily show that when
-integrated over the length of the plane, the average fluxes are
-consistent, i.e.,
-\begin{equation}
-\sum_j (f_{ijk}^{e\star} - f_{i+1jk}^{w\star}) = 0 \quad\forall k,
-\end{equation}
-thus ensuring order parameter conservation.
-
-A similar correction is required when computing the force on the fluid
-directly via the divergence of the thermodynamic stress
-$F_\alpha = \nabla_\beta P_{\alpha\beta}^{th}$. This is implementated
-by computing fluxes of momentum at cell faces in each direction, and
-then taking the divergence to compute the force. At the planes, an
-averaging procedure is again used to ensure conservation of momentum.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% End LE implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \vfill
-\pagebreak
-
-\section{Code Interface}
-
-\subsection{The Parallel Environment}
-
-\texttt{\#include ``pe.h''}
-
-As the code is to be available in both serial and parallel versions,
-it is desirable to abstract the basic environment to some extent.
-The parallel environment is intended to support MPI communication
-within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
-consistent picture in serial (i.e., rank zero is the single process).
-
-\texttt{void pe\_init(int argc, char ** argv)}\\
-Responsible for initialisation of MPI, and hence must be first
-execuatable statement. The arguments are those to \texttt{main}
-which will be passed through to \texttt{MPI\_Init()}.
-
-\texttt{int pe\_rank(void)}\\
-Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
-serial.
-
-\texttt{int pe\_size(void)}\\
-Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
-1 in serial.
-
-\texttt{void pe\_finalise(void)}\\
-Responsible for \texttt{MPI\_Finalize()}, and must be the last
-executable statement.
-
-\texttt{void info(const char * fmt, ...)}\\
-Behaves like \texttt{printf}, but only produces output on the
-root process.
-
-\texttt{void verbose(const char * fmt, ...)}\\
-Behaves as \texttt{printf} and reports the rank of the
-calling process in \texttt{MPI\_COMM\_WORLD}.
-
-\texttt{void fatal(const char * fmt, ...)}\\
-Prints an error message and identifies
-the rank of the calling process before terminating the program.
-
-\subsection{Run Time Input}
-
-\texttt{\#include ``runtime.h''}
-
-The ability to set a wide range of parameters at run time is
-very convenient, and is supported via an input file in which the
-user specifies parameters as key value pairs. The code maintains
-a list of key value pairs from which the user can retrieve the
-appropriate values for given keys.
-
-\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
-This reads the named input file and sets up the list of keys.
-
-\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The number
-of key matches found is returned (i.e., 0 or 1).
-
-\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The
-number of key matches is returned.
-
-\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
-Set \texttt{value} to the string assocaited with \texttt{key}, if
-present. Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
-If \texttt{key} is present, set the associated 3-vector of integers.
-Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
-double v[])}\\
-If \texttt{key} is present, set the associated 3-vector of doubles.
-Returns the number of key matches found.
-
-\textit{Comment:} We probably need a function to report unused keys, so
-the user can be alerted if a key has unexpectedly been ignored.
-
-\subsection{Random Number Generators}
-
-\texttt{\#include ``ran.h''}
-
-Note that the random number generation in parallel is
-decomposition dependent.
-
-\texttt{void rand\_init(void)}\\
-Responsible for initialising RNG state.
-
-\texttt{double rand\_serial\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$ from
-the serial generator.
-
-\texttt{double rand\_serial\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the serial generator.
-
-\textit{double rand\_parallel\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$
-from the serial generator.
-
-\texttt{double rand\_parallel\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the parallel generator.
-
-\textit{Comment:} The serial generator will produce the
-same sequence on any MPI process, while the parallel generator
-will produce different sequences. The latter is intended for
-generating, specifically, fluctuations, where the need for
-decomposition-independance is questionable.
-
-\subsection{Timers}
-
-\texttt{\#include ``timer.h''}
-
-Performance figures require timing for different sections
-of the code. The code provides a set of routines to time
-standard and user-defined sections of code. Timers are
-started and stopped by call to the appropriate routines.
-
-\texttt{TIMER\_init()}\\
-This initialises all the timers to zero (e.g., at the start of
-the program).
-
-\texttt{TIMER\_start(const int t\_id)}\\
-Start the given timer.
-
-\texttt{TIMER\_stop(const int t\_id)}\\
-Stop the given timer.
-
-\texttt{TIMER\_statistics(void)}\\
-Prints the current state of all the active timers to standard
-output on the root process in a digestible form. In serial,
-the minimum, maximum, and mean time for each timer is reported
-in seconds (based on the number of start/stop cycles). In
-parallel, the same figures are computed on each process and the
-result of the corresponding \texttt{MPI\_Reduce()} operation is
-printed on the root process.
-
-The code provides a number of timers which would time standard
-parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
-total execution time, and a number of spare timers for users to
-use at will.
-
-\textit{Comment:}
-No intra-run variance (i.e., between time steps) is provided, as the
-calculation required is considerably more complex to generalise, and
-requires memory proportional to the number of time steps used.
-
-\textit{Comment:}
-Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
-provided to insert calls to the above functions for optional timing
-of performance-sensitive regions
-of the code in conjunction with the preprocessor flag
-\texttt{-D\_TIMING\_MACRO\_ON\_}.
-
-\subsection{Model Time Control}
-
-This provides time step control, event frequency (?)
-
-\texttt{void init\_control(void)}\\
-Initialise time controls.
-
-\texttt{int get\_step(void)}\\
-Return the current model time step.
-
-\texttt{int next\_step(void)}\\
-Increments the current time step. Returns number of steps left
-to run (i.e., 0 when finished).
-
-\subsection{Coordinate System}
-
-\texttt{\#include ``coords.h''}
-
-A three-dimensional Cartesian coordinate system is used throughout,
-with the exact system size usually chosen by the user at run time.
-The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
-used to identify the coordinate directions.
-
-This also provides access to the MPI Cartesian Communicator which
-is responsible for most communication in the model.
-
-\texttt{int N\_total(const int dim)}\\
-Returns the total number of lattice sites in the given Cartesian direction.
-
-\texttt{double L(const int dim)}\\
-Returns the length of the system in the given Cartesian direction
-(which is the same as the number of lattice sites).
-
-\texttt{double Lmin(const int dim)}\\
-Returns coordinate position of left-hand edge of system in given
-Cartesian direction.
-
-\texttt{int is\_periodic(const int dim)}\\
-Returns 1 if given Cartesian direction is periodic, else 0.
-
-\texttt{void cart\_init(void)}\\
-Responsible for setting up the Cartesian decomposition.
-
-\texttt{int cart\_rank(void)}\\
-Returns rank in the Cartesian communicator.
-
-\texttt{MPI\_Comm cart\_comm(void)}\\
-Returns the MPI Communicator handle for the Cartesian communicator.
-
-\texttt{int cart\_size(const int dim)}\\
-Returns the size of the Cartesian communicator grid in the given
-coordinate direcxtion.
-
-\texttt{int cart\_coord(const int dim)}\\
-Returns the coordinate of the current process in the Cartesian
-Communicator.
-
-\texttt{int cart\_neigh(const int dir, const int dim)}\\
-Returns the rank of the neighbouring process in the
-Cartesian communicator.
-
-\subsection{Colloids}
-
-\texttt{\#include ``colloids.h''}
-
-These functions provide the basic interface for use
-of colloidal particles in the code.
-
-\subsubsection{Colloid structure}
-
-The basic colloid struct is defined ...
-
-\subsubsection{Colloid interface}
-
-\texttt{void colloids\_init(void)}\\
-Call once for initialisation. This initialises the cell list,
-so the coordinate system must be initialised via a call to
-\texttt{coords\_init()} first.
-
-\texttt{void colloids\_finish(void)}\\
-Call once for finalisation. Frees all colloid-related memory.
-
-\texttt{Colloid * allocate\_colloid(void)}\\
-Returns a pointer to a newly allocated \texttt{Colloid} object
-or fails gracefully.
-
-\texttt{void free\_colloid(Colloid * p\_colloid)}\\
-Deallocates all memory associated with \texttt{p\_colloid} or
-fails gracefully.
-
-\texttt{struct c\_link allocate\_boundary\_link(void)}\\
-Returns a pointer to a newly allocated colloid boundary
-link or fails gracefully.
-
-\texttt{void free\_boundary\_link(struct c\_link *)}\\
-Deallocates memory associated with a link or fails gracefully.
-
-\texttt{void colloids\_memory\_report(void)}\\
-The module keeps track of how much memory has been allocated in
-total for colloids and boundary links. This reports the current
-usage.
-
-\texttt{get\_Ncell\_total()}\\
-
-\texttt{get\_Ncell\_local()}\\
-
-\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
-This returns a pointer to the Colloid at the head of the list at
-cell list location (i, j, k) or \texttt{NULL} if there is none.
-
-\texttt{void cell\_insert(Colloid * p\_colloid)}\\
-This inserts a colloid at the head of the appropriate list
-depending upon its position. This is the only way a colloid
-can be added to the list.
-
-\texttt{update\_cells(void)}\\
-Performs the update of the cell list in response to updated
-particle positions.
-
-\texttt{IVector cell\_coords(FVector r)}\\
-
-\subsection{Speculation}
-
-\subsection{The Free Energy}
-
-It is imagined that a single file will encapsulate all the details
-of a given free energy. The interface for different choices of
-free energy should be the same.
-
-\texttt{FE\_init()}
-
-Responsible for querying the runtime input key value pair list to
-find appropriate user parameters for the free energy, or
-falling back to default values.
-
-\texttt{double FE\_get\_sigma()}
-
-Return the surface tension. There should be a similar routine
-for the interfacial thickness.
-
-\texttt{double FE\_get\_mu(const type location)}
-
-Return the chemical potential at the single lattice
-node \texttt{location}.
-
-\textit{Comment:} The computation of quantities such as the chemical
-potential will a) affect performance and b) require halo values of
-the order parameter (depending on the highest derivative
-in the order parameter present). Performance here should be the
-first consideration, so this needs to be handled with care.
-
-\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}
-
-Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
-given location.
-
-\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
-the number of dimensions. If we are to provide explicit support for
-D2Q9, \texttt{ND} will be 2, otherwise 3.
-
-\section{Implementation Notes}
-
-\subsection{Serial Version}
-
-\subsection{Efficiently Scaling Parallel Computing}
-
-We should be thinking about the 100,000 processor regime, and
-correspondingly large system sizes. To avoid embarrassment in this regime,
-some issues which don't usually arise are relevant, for example:
-
-\begin{enumerate}
-\item
-Large integer quantities such as
-\begin{verbatim}
-   N_total.x*N_total.y*N_total.z
-\end{verbatim}
-are likely to cause integer overflow (if default integers are 4 byte).
-\item
-Local data structures whose size is proportional
-to the number of processors are contraindicated, e.g.,
-anything like
-\begin{verbatim}
-p = (foo *) malloc(nprocs*sizeof(foo));
-\end{verbatim}
-is to be avoided. At the moment, we are in good shape on this
-(apart from some functions in \texttt{misc.c}).
-\item
-I/O is likely become a very important issue, so may need to think
-more about this.
-\item
-We should probably introduce a standard set of benchmarks which can
-be scaled up in some systematic way, so that we can keep track of
-performance over time.
-
-\end{enumerate}
-
-\section{Model Details}
-
-\subsection{D2Q9}
-
-These are the eigenvalues and eigenvectors of the collision
-matrix used for D2Q9
-
-\begin{table}
-\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
-\hline
-$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
-& $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
-& $c_{iy} c_{iy} - c_s^2$ \\
-\hline\hline
-$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - & 0 &  4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{iy}$\\
-\hline\hline
-$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q15}
-
-These are the eigenvalues and eigenvectors of teh collision
-matrix used for D3Q15
-
-\begin{table}
-\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - &
- 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
-1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
-3  & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - &
- 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
-3  &$c_{iy}$ \\
-\hline
-$\rho c_{iz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
-3  & $c_{iz}$ \\
-\hline
-$Q_{xx}$ & 1/3 &
--1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
-9/2  & $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{yy}$ & 1/3 &
--1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
-\hline
-$Q_{zz}$ & 1/3 &
--1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
-9  & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
-9  & $c_{iy} c_{iz}$ \\
-\hline
-$Q_{zx}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
-9  & $c_{iz} c_{ix}$ \\
-\hline\hline
-$\chi^1$ & - &
--2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
-1/2 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
-3/2 & $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - &
- 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iy}$\\
-\hline
-$J_{iz}$ & - &
- 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iz}$\\
-\hline
-$\chi^3$ & - &
- 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
-9 & $c_{ix} c_{iy} c_{iz}$ \\
-\hline\hline
-$w_i$ & 1/72 &
-$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
- & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q19}
-
-\begin{table}
-\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
-\hline\hline
-$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
-\hline
-$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
-         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
-& 1\\
-\hline
-$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
-         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
-& 3 \\
-\hline
-$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
-         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
-& 3\\
-\hline
-$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
-         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
-& 3\\
-\hline
-$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
-& 9/2\\
-\hline
-$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
-         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
-& 9\\
-\hline\hline
-$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
-         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 3/4\\
-\hline
-$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
-         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
-         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
-& 3/2\\
-\hline
-$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
-& 9/4\\
-\hline
-$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
-& 9/2\\
-\hline
-$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
-         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 1/2\\
-\hline\hline
-$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
-& \\
-\hline\hline
-\end{tabular}
-\end{table}
-
 \pagebreak
@@
 
+\bibitem{ieee-208}
+IEEE Standard 208-2005, IEEE Standard for Software Configuration Management
+Plans. See \texttt{http://ieeexplore.ieee.org/xpl/standards.jsp}
+(accessed 2011).
+
 \bibitem{jo84}
@@
 
-\bibitem{nl02}
+\bibitem{mpi-standard}
+Message Passing Interface Forum. MPI: A Message Passing Interface Standard
+Version 1.3 (2008).
+
+\bibitem{nguyen-ladd2002}
 Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
@@
 CRC Press, Boca Raton, Florida, 2000.
+
+\bibitem{paraview}
+Paraview. See \texttt{http://www.paraview.org/}. Accessed 2011.
 
@@
 \end{thebibliography}
-
-\end{document}

### 2011-08-19 papers/smectics/smectics.tex
Smectics paper added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1444 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%This is the article LaTeX template for RSC journals
%Copyright The Royal Society of Chemistry 2010
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance} 
\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}

\pagestyle{fancy}

\newcommand{\update}[1]{{\color{red} #1}}
\newcommand{\e}[1]{\cdot10^{#1}}
\newcommand{\gd}{\dot{\gamma}}

\begin{document}

\thispagestyle{plain}
\fancypagestyle{plain}{
\fancyhead[L]{\includegraphics[height=8pt]{LH.pdf}}
\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{CH.pdf}}
\fancyhead[R]{\includegraphics[height=10pt]{RH.pdf}\vspace{-0.2cm}}
\renewcommand{\headrulewidth}{1pt}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand\footnoterule{\vspace*{1pt}% 
\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
\setcounter{secnumdepth}{5}

\makeatletter 
\def\subsubsection{\@startsection{subsubsection}{3}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\bf}} 
\def\paragraph{\@startsection{paragraph}{4}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\textit}} 
\renewcommand\@biblabel[1]{#1}            
\renewcommand\@makefntext[1]% 
{\noindent\makebox[0pt][r]{\@thefnmark\,}#1}
\makeatother 
\renewcommand{\figurename}{\small{Fig.}~}
\sectionfont{\large}
\subsectionfont{\normalsize}

\fancyfoot{}
\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{LF.pdf}}
\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{RF.pdf}}
\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}
\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}

\twocolumn[
  \begin{@twocolumnfalse}
\noindent\LARGE{\textbf{Nonlinear Rheology of Layered Smectic Liquid Crystals}}
\vspace{0.6cm}

\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{a}$}, K. Stratford\textit{$^{b}$} and M.E. Cates \textit{$^{a}$}}
}\vspace{0.5cm}
%Please note that \ast indicates the corresponding author(s) but no footnote text is required.

\noindent\textit{\small{\textbf{Received Xth XXXXXXXXXX 20XX, Accepted Xth XXXXXXXXX 20XX\newline
First published on the web Xth XXXXXXXXXX 200X}}}

\noindent \textbf{\small{DOI: 10.1039/b000000x}}
\vspace{0.6cm}
%Please do not change this text.

\noindent \normalsize{The abstract should be a single paragraph which summarises the content of the article. Any references in the abstract should be written out in full \textit{e.g.} [Surname \textit{et al., Journal Title}, 2000, \textbf{35}, 3523].}
\vspace{0.5cm}
 \end{@twocolumnfalse}
  ]

%Footnotes
%Please use \dag to cite the ESI in the main text of the article.
%If you article does not have ESI please remove the the \dag symbol from the title and the above footnotetext.

\footnotetext{\textit{$^{a}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
\footnotetext{\textit{$^{b}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}

%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required

\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P. Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}

\section{Introduction}
In our everyday lives we are surrounded by complex fluids such as foams, colloidal dispersions, microemulsions, polymeric fluids, liquid crystals and many biofluids.
Their intricate microstructure gives rise to unusual rheological properties that are not found in simple Newtonian fluids and make them indispensable materials in numerous industrial applications.
Despite their abundance in the processing, energy or chemical industry complex fluids remain difficult to model as they normally consists of two or more separate phases which interact with each other and the solvent.
This in turn causes stresses and emerging flow properties like the viscosity to depend highly on the fow state and on the structure of the microphases. 
In order to capture their rich phenomenology and properties ab initio without using constitutive relations that only mimick their behaviour, computationally expensive approaches are necessary.
It is only now that we have the resources to simulate nonlinear flow of complex fluids on large scales\cite{Saksena09}.\\
In the present work we focus on a specific class of complex fluids, whose members share the common feature to undergo an order-disorder transition from an isotropic-disordered state at high temperature to a microphase separated state at low temperature.
They can be described with a Brazovskii free energy functional \cite{Brazovskii75} and comprise systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.
Especially the latter have attracted considerable interest because of their intricate morphology.
Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}. 
Moreover, in some regimes shear-induced transitions from the microphase separated state to ordered phases like the lamellar one occur \cite{Cates89,Koppi93,Fredrickson94}.
Lyotropic liquid crystals are known to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93}.
Shear-induced ordering takes also place in smectic liquid crystals \cite{Larson93,Hamley99}, making these systems an interesting object for rheological studies.\\
The purpose of this paper is twofold.
First we investigate the morphology and rheology of a two-dimensional lamellar system in steady shear flow.
This system has been studied extensively \cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}, but because of computational constraints almost all results were limited to two dimensions and smaller strains.
Here we quantify rheological properties and try to relate them to the defect pattern and the microstructure of the fluid.
In the second part of the analysis we compare the results in two and three dimensions in order to see how the dimensionality of the system influences the defect dynamics and phase kinetics.
\\ 
The governing equations in our model are a convection-diffusion equation for the scalar order parameter, which represents the density difference of the two phases and the Navier-Stokes equation with a stress tensor that contains extra contributions due to the presence of the lamellar interfaces.
This approach allows for stress-strain rate relations to emerge rather than just putting them in by means of consitutive equations.
In our approach the dynamical equations are solved using a combination of finite difference scheme for the convection-diffusion equation and the lattice Boltzmann method (LBM) \cite{Succi} for the Navier-Stokes equation.
The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries[], where it is superior to conventional methods of computational fluid dynamics.\\ 
This paper is organised as follows. In Sec.II we introduce the model, including its equilibrium properties, the full stress tensor and the dynamical equations.
We briefly sketch how we derive structural information from the order parameter and describe our defect quantification procedure, which measures the local alignment of the lamellar structure.
We present results of the two dimensional case in In Sec.III, discuss the morphology, average and local shear stress densities and first normal stress differences and their relation to the defect structure.
We also show structure factors for various flow states and flow curves.
In Sec. IV we compare our 2D-results with results in three dimensions. 
A summary with conclusions closes the paper in Sec.V.

\section{Model and Definitions}

The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimizes the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.
\begin{eqnarray}
{\cal F}[\phi]&=&\int d^dx f(\phi({\bm x}))\nonumber\\
&=&\int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\bm \nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\label{FE-functional}
\end{eqnarray}
The parameters $b$ and $c$ are positive to ensure stability. 
For $a>0$ there is only one solution $\phi\equiv0$ which minimizes the functional and corresponds to the disordered state.
For $a<0$ a microphase separation occurs and depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) because the formation of interfaces become energetically favourable.\\
The dynamics of the binary fluid is governed by a convection-diffusion equation with a diffusive current proportional to the chemical potential $\mu$, which is the functional derivative of ${\cal F}$ with respect to $\phi$.
\begin{equation}\label{OPEOM}
\partial_t \phi +{\bm \nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}
The coefficient $M$ herein plays the role of a mobility coefficient.\\
If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys a Navier-Stokes equation
\begin{equation}\label{NSE}
\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}
The stress tensor reads
\begin{equation}
\sigma_{\alpha \beta}=-P^{th}_{\alpha \beta} +\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha).
\end{equation}
It consists of a hydrodynamic part, which is proportonial to the viscosity $\eta$ of the solvent and a thermodynamical part $P^{th}$ that depends on the order parameter.
The latter can be calculated from the free energy functional:
\begin{equation}
P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + D_{\alpha \beta}(\phi).
\end{equation}  
In this expression the symmetric tensor $D_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee the mechanical equilibrium \cite{Evans79}.
The thermodynamic pressure tensor is given by
\begin{equation}
P^{th}_{\alpha \beta}=p\, \delta_{\alpha \beta} + P^{chem}_{\alpha \beta}
\end{equation}
with an isotropic pressure contribution
\begin{eqnarray}
p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\bm \nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}
and an anisotropic part
\begin{equation}
P^{chem}_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

The Navier-Stokes Eq.\ref{NSE} was solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
The velocity field were then put into the convection-diffusion equation Eq.\ref{OPEOM} and solved at the same timestep. 
We assumed periodic boundary conditions along the flow and vorticity direction and Lees-Edwards boundary conditions (BC) \cite{Wagner02} along the direction of the velocity gradient.
The total number of Lees-Edwards planes was varied according to the size of the simulation so that the distance of two planes was always 16 lattice sites.
To impose simple shear flow, the planes in the upper half of the velocity gradient direction were moved at a constant speed in positive flow direction, whereas the planes in the lower hemisphere were moved in negative flow direction.\\
Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
\begin{equation}
C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
\end{equation}
with $\phi({\mathbf k})$ as normalised Fourier-transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
In order to analyse the relation between rheological quantities and the morphology of the system it was necessary to quantify the number of defects.
For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
More explicitely the alignment factor $X_i$ is given as
\begin{equation}\label{alignment}
X_i=\frac{1}{N_{nn}} \sum_{j=1,N_{nn}}\left(\frac{\nabla \phi_i \cdot \nabla \phi_j}{|\nabla\phi_i||\nabla\phi_j|}\right)^2. 
\end{equation}
$N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes up smaller positive values.
By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates from what would be recognised by the eye as an aligned configuration.
We found a threshold of $\xi=0.8$ to work quite well for our requirements.
Unfortunately this defect detecting protocol is not able to assign only a single defect site to a disturbed region in the order parameter structure as normally more than one site close to a defect would have an alignment factor below the threshold.
However, as the number of sites per defect is roughly constant throughout the sample, it is still useful for quantifying the degree of alignment and comparing different systems or flow conditions.\\
By relating the number of defects sites with $X_i$ below the threshold to the total number of sites it is possible to define a defect density:
\begin{equation}\label{defectdensity}
\rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
\end{equation}
The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
The exact value of the defect density depends clearly on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.

\subsection*{Parameter Mapping to Physical Units}

In the following we describe how the simulation units can be mapped onto real physical units.
It is necessary to calibrate the scales for mass density, length and time or alternatively as we do for mass density, length and energy density.\\
As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the reference fluid $\rho_0$ all to unity in LB units (LBU).
It is straightforward to determine the unit of mass. 
A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.\\
In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space.\\
The energy scale is slightly more difficult to calibrate.
We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
This formula reads 
\begin{equation}
b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
\end{equation}
whereas $\rho_c$ is the number density of chains per unit volume,
often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
Hence the LB unit of energy density is about $1\e{6} \rm{Pa}$.\\
We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=1\e{-3+6+14} \rm{LBU}\simeq1\e{17} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq3\e{8}$LBU and one LB timestep $\Delta t$ corresponds to roughly 3ns.\\
Finally we want to get an estimated scale for shear rates. 
A common way to quantify them is the dimensionless Peclet number, which measures the transport of a particle as the ratio of the rate of advection to the rate of diffusion.
The Peclet number can be defined as $Pe_0=\gd\, D^2/D_0$, where $\gd$ is the rate of strain, $D$ is a characteristic length, e.g. the diameter of a particle and $D_0$ is the diffusivity.
The highest shear rate we apply is $2\e{-4}$ in LBU, basically reciprocal LB time units.
With the above value for $\Delta t$ we can convert this into about $5\e{4}{\rm s}^{-1}$.
If we suppose the Stokes-Einstein diffusion constant is a good approximation for $D_0$, we get for an object of the size of $D\simeq5\e{-8}$ at room temperature in a low viscous solevent like water values of about $D_0=1\e{-11} \rm{m}^2\rm{s}^{-1}$.
This results in Peclet number of the order ${\cal O}(10)$ and means shear rates $\gd={\cal O}(10^{-4})$ in LBU correspond to a Peclet numbers $Pe_0={\cal O}(10)$.

\section{Results}

Our discussion will be in two parts, beginning with simulations in two dimensions for a system A and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
The second part of our discussion consits of simulations in three dimensions using identical parameters to system A and a box size $N_x\times N_y\times N_z=128\times256\times128$.
We will refer to this system as system C.
Table \ref{tab1} at the end of this paragraph provides an overview of all simulation parameters.

\subsection*{Results in Two Dimensions}
There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}. 
We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run703_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_500.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
\caption{Order parameter $\phi$ of system A: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom). Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
\label{fig1}
\end{figure}
This effect has been reported before \cite{Gonnella97} and was more pronounced for softer systems with smaller Landau parameters, although not quite to the same extent as in the above reference.
Under shear higher viscosities caused the lamellae to be much less homogenous in shape and thickness.\\
In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the stiffer system A with $-a=b=5\e{-4}$ and the softer system B with $-a=b=5\e{-5}$, respectively.\\
The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
The equilibrium structure is completely isotropic and exhibits a lamellar spacing of roughly 10 lattice sites.
If shear flow is imposed, the lamellae start to align with the flow.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_run707_320.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run710_800.jpg}\\
\includegraphics[angle=0,width=0.5\textwidth]{phi_run765_1280.jpg}
\caption{Order parameter $\phi$ of system B: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom)}
\label{fig2}
\end{figure}
In this paragraph the flow direction and the direction of the velocity gradient are x and y, respectively.
The apparent flow due to the effect of the Lees-Edwards planes is always in positive x-direction in the upper half and in negative x-direction in the lower half of the box.
After a short startup phase in which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emarge and the spacing deviates from its equilibrium value.
These regions are advected with the flow while they undergo an affine stretching.
The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $1\e{-4}$.
In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
However, the reported results differ from our observations in the following aspects.
First of all, the shear-induced structures emerge at the centre and remain located, while ours occur in bandlike patterns that are wrapped around the periodic boundaries.
Moreover those bear the characteristics of a shear banding region exhibiting a reduced local shear rate.
These differences may be a direct consequence of the different boundary conditions.
While we applied Lees-Edwards BC along y and periodic boundary conditions along x for all quantities, \cite{Xu06b} seem to have used sort of hybrid scheme.
\begin{figure}[!]
\centering
\hspace*{-0.1cm}\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{shear_str_run704_400.jpg}
\includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A at $t=4\e{5}$}
\label{fig3}
\end{figure}
\begin{figure*}[!]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
\caption{Structure factor $C({\bm k})$ of system A: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right). The sections shown correspond to wavevectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
\label{fig4}
\end{figure*}
To our understanding they applied Lees-Edwards BC for the order parameter only, but placed solid moving walls at the upper and lower boundaries in the lattice Boltzmann part.
Identical to our setup periodic BC in flow direction were applied for both order parameter and momentum flux.
The it might be that \cite{Xu06b} have not simulated bulk flow behaviour but rather flow in a confined geometry and the shear-induced structure they observe might be a consequence of this difference.\\ 
After switching off of the shear flow the expanded regions contract and the compressed regions expand and the system returns to a state with homogenous lamellar width, seemingly close to the equilibrium value.
The final state for which the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
The lamellar interfaces remain more or less oriented, but due to the expansion and contraction of the lamellae some individual domains show large angles with the previous flow direction. 
The length of the interfaces varies considerably, which is appears to be another residue of the inhomogeneities in the flowing state.\\
Fig. \ref{fig2} shows equivalent states for the softer system B.
The characteristic results of Fig. \ref{fig1} are retained, but the interfaces in all states tend to be longer.
The lamellar alignment in flow direction after switch off is more pronounced.\\ 
Fig. \ref{fig3} relates the inhomogeneities in the morphology to rheological quantities.
The pictures show the order parameter $\phi$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
Where the system is extendend and the lamellar width is larger compared to the equilibrium state, both shear stress and first normal stress differences have low positive or negative values.
Large positive values occur in the regions where the lamellae are squeezed.
The peak values in lattice units are in the region of $-2\e{-4}$ and $3\e{-4}$ for the shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
Despite the local negativity of quantities, the overall average are still positive for all times.
Very similar results with smaller peak magnitudes are obtained for system B, which appears somewhat more homogenous.\\
Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the three states depicted in Fig. \ref{fig1}.
At first isotropic after the microphase separation and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
In steady shear flow the structure factor develops two distinct though rather broad peaks, which are situated at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
It is interesting to see that the Fourier-vector of the peak position pics up a small component in x-direction, indicating on average a slight tilt of the structure.
The tilting orients the lamellae away from the direction of flow and towards the velocity gradient direction, which is the direction of extension in simple shear flow.
The position of the peak is almost constant throughout the entire simulation.
The average lamellar distance decreases from about 10 lattice sites in the quiescent state to about 8.5 lattice sites.
However, after switching off the system returns its equilibrium lattice spacing.
The peak intensity rises due to the fact that the interfaces remain oriented along the flow direction but more interestingly the small x-component of the peak position disappears as the lamellar spacing becomes homogenous again.
This gives some evidence that the small x-component is indeed connected to extensional flow in the regions with negative first normal stress differences.\\
With the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify defects in the lamellar structure.
We would like to locate regions where there is no homogenous order parameter changes because it changes sign and the interfaces have a large local curvature.
Fig. \ref{fig5} shows an example of the time evolution of system A at low shear.
All lattice sites with $X_i\le0.8$ are marked red and obviously coincide with interfacial endpoints and disturbances.
The top picture shows the situation at the beginning of the simulation with the lowest shear rate $\dot{\gamma}=1.25\e{-5}$.
The system has been pre-sheared at consecutively lowered shear rates ranging from $\gd=1\e{-4}$ to $2.5\e{-5}$, whereas the final configuration of the previous run was always used as initial configuration of the following run.
There are regions which exhibit some degree of flow alignment and others, where the angle between the lamellae and the flow direction remains rather large.
The bottom picture of Fig. \ref{fig5} shows the structure at the end of the run.
The total strain difference between both images is about $10^{5}\%$. 
Obviously many defects have annealed and the interfaces are well oriented in flow direction.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
\caption{Defect structure of system A: The pictures show results of a smaller simulation at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$. The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}). Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
\label{fig5}
\end{figure}
This degree of defect annihilation is only found below a critical shear rate where defects are advected smoothly on well aligned lamellae, approach each other and cancel.
Above the critical shear rate the lamellae buckle quite similar to what has been reported in \cite{Gonnella98}, the structure shows a tendency to trip over itself which creates new defects.
The critical value is different in both systems, roughly between $\dot{\gamma}=2.5-5\e{-5}$ in the stiffer system A and higher at around $\dot{\gamma}=1-2\e{-4}$ in the softer system B.
Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A for a sequence of decreasing shear rates.
The runs with the two highest shear rates were started from the quiescent equilibrated state, whereas all others used the final configuration of the preceding run with higher shear rate as initial configuration.
Here and in the following figures horizontal lines are time averages for the particular quantities where we think the system has reached a steady state. 
The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
Interestingly once the shear rate has been decreased to the next lower rate, both systems A and B show an trend to anneal defects instantaneously or after a short latency time if the shear rate is rather low.
If the shear rate is turned up again, the lamellae buckle and the structure begins to trip over itself, once again creating new defects.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted). Horizontal lines indicate averages over time for the particular quantities, where we assume the system has reached a steady state. The inset gives data for the 3D-system C and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.} 
\label{fig6}
\end{figure}
This suggests that there is a dynamical equilibrium between defect creation and annihilation above a critical shear rate, while below the critical value defects get not anymore widely generated and the system anneals its defects and forms properly aligned lamellae.
In order to test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
The defect density goes indeed up (not shown) and reaches similar values to those shown in Fig. \ref{fig6}.
However it takes some time before a steady state is reached as the applied shear rates are rather low.\\
Fig \ref{fig7} shows the average total shear stress density $\langle \sigma_{xy}\rangle$, i.e. the total shear stress per unit lattice volume.
In the following we will drop the backets $\langle\dots\rangle$ for simplicity and write $\langle \sigma_{xy}\rangle\equiv  \sigma_{xy}$.
While for the highest applied shear rate the hydrodynamic part forms the largest contribution to the total stress density, a regime where the chemical stress dominates is entered on decreasing the shear rate.
At low shear rates stress fluctuations become more stretched in time and get smaller, what is slightly hidden by the logarithmic scale of Fig. \ref{fig7}
Their magnitude is about three times smaller in the larger simulation shown in Fig. \ref{fig1}, \ref{fig2}.
Fortunately we found that the average shear stress density over time is to a good degree independent of the system size and therefore it is justified to simulate longer times by using a smaller system size without flawing the results.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
\caption{Average total shear stress $\sigma_{xy}$ versus time for system A: The horizontal lines are steady state time averages}
\label{fig7}
\end{figure}
Furthermore the stress fluctuations represent a physical aspect as they can be related to the inhomogeneities of the order parameter.
System A shows clear shear thinning behaviour and seems to obey a power law (see Fig. \ref{fig11}).
For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the time averaged apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
Stress fluctuations cause temporary variations of these mean values of up to a factor 3 at the lowest shear rate.
In contrast to system A system B does not exhibit a clear monotonous shear thinning tendency.
The apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow more a horizontal trend when the shear rate is decreased.
The stress fluctuations and time averaged shear stress densities are slightly smaller than in the stiffer system A with one exception.
For the highest shear rate $\gd=2\e{-4}$ the shear stress turns out to be larger than in system A.\\
The deviatoric part of the average chemical pressure density $\langle P_{xy}\rangle\equiv P_{xy}$ for system A is shown in the main picture of Fig. \ref{fig8}.
The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
System B shows a very steep decrease of $P_{xy}$ during the simulation with $\gd=1\e{-4}$, which is because of the defect annihilation.\\
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A. The inset shows data for the 3D-system (system C) and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
\label{fig8}
\end{figure}
Fig. \ref{fig9} illustrates the relation between $P_{xy}$ and the defect density $\rho_D$.
The black line represents a fit to the system A, which shows a linear dependence almost down to the lowest shear rate.
The result for the softer system B is slightly more obscured as a consequence of the ordering and the tendency to anneal defects.
Our quantifying protocol is based on averaging over a larger number of defects sites and is prone to inaccuracy in a regime where defects are rather scarce.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs defect density for system A ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)} 
\label{fig9}
\end{figure}
The main picture in Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A.
At first until $\gd=5\e{-5}$ the first normal stress differences decrease monotonously with decreasing shear rate.
At lower shear rates, however, the behaviour is far from being monotonous and $N_1$ rises again.
This makes it difficult or even impossible to define reasonable time averages as $N_1$ may still evolve at the end of the run.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{xx}-\sigma_{yy}\rangle$ vs time for system A1. The inset shows data for the 3D-system (system C) and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
\label{fig10}
\end{figure}
This feature is actually less erratic than it seems at first sight and can be rationalised. 
A look at Fig. \ref{fig6} reveals that $N_1$ rises in the same way as the defect density drops.
A pair of defects can cancel if one of them is an endpoint of a black lamella surrounded by a branching white lamella and the other one is a white endpoint surrounded by a branching black lamella.
When these two defects anneal the lamella spacing increases locally by half a lamellar width.
The interfaces have to expand in y-direction as the system is not frustrated enough to create an entire new layer in order to get the lamellar width back closer to its equilibrium value.
Hence the normal stresses $\sigma_{yy}$ become negative and consecutively smaller, leading to gradually larger values of $N_1$.
It should be stressed that only $\sigma_{yy}$ seems to evolve anomalously slowly and all other components of the stress tensor reach state state values more rapidly.
System B again behaves different in the sense that $N_1$ continues to fall for lower shear rates despite there is also ongoing annealing of defects.
This difference is a direct effect of the softness of system B, whose smaller Landau coefficients prevent the normal stresses $\sigma_{yy}$ from taking up very negative values and $N_1$ from rising.
Nevertheless close to the end of the simulations with the two lowest shear rates $N_1$ is still slightly changing which suggests that this slowly evolving behaviour is also due to ongoing defect annihilation.
Because both trends in system A and B are consequences of the impossibility to optimize the lamellar width, larger simulations should lead to more defined steady state values for $N_1$.
However, we were not able to check this as this proves computationally cosly.\\
In Fig. \ref{fig11} we show flowcurves derived from time averages of $\sigma_xy$ and $N_1$ as a function of shear rate.
The time average of shear stress density in systems A follow a power law $\sigma_{xy}=\dot{\gamma}^m$ with an exponent $m\simeq 0.87$ and gives evidence for shear thinning behaviour.
In case of system B the best fit for the lowest three shear rate is obtained for an exponent $m=1$, indicationg more or less constant viscosity and no shear thinning.
It is not very surprising as the chemical contribution to the total stress almost vanishes for shear rates $\dot{\gamma}\le1\e{-4}$ due to the significant decrease of the number of defects.
The observed power law behaviour is simply that of the hydrodynamic part of the stress, which is linear in our case.\\
A comparison with experimental work remains difficult because the topology of defects differ in two and three dimensions.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
\caption{Flowcurves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)} 
\label{fig11}
\end{figure}
Our two-dimensional system is translational invariant in the third dimension.
So every branch point or endpoint of a lamella has the topology of an edge defect in the 3D-counterpart of the structure.
Moreover screw defects do not exist in our picture as they can only occur in more than two dimensions.
A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
Interestingly the authors identify exactly the motion of screw defects as the key mechanism responsible for the layer tilting and defect formation. 
For the power law dependence of the shear stress on the shear rate they find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.\\
Contrary to shear stresses first normal stress differences do not level off because of ongoing defect annihilation.
Neither system A nor B is in a steady state at the end of the simulation and therefore the data for $N_1$ is only shown for the sake of completeness.

\subsection*{Results in Three Dimensions}

The computational costs of a detailed analysis of the nonlinear rheology in three dimensions are still substantial as for steady state quantities it is necessary to simulate large strains.
Therefore we performed only a limited number of simulations for selected parameters. 
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.45\textwidth]{phi_run786_320.jpg}
\includegraphics[angle=0,width=0.45\textwidth]{phi_run788_960.jpg}
\includegraphics[angle=0,width=0.45\textwidth]{phi_run792_960.jpg}
\caption{Order parameter $\phi$ in two and three dimensions: The top and centre picture show system C after equilibration at timestep $t=3.2\e{5}$ and in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at $t=9.6\e{5}$, where shear-induced layering takes place. The image at the bottom depicts the corresponding two dimensional system A of the same size in flow and velocity gradient direction.} 
\label{fig12}
\end{figure}
For our three dimensional simulations, in the following referred to as system C, we used parameters identical to those of the stiffer 2D-system A, but reduced the system size to $N_x\times N_y \times N_z=128\times256\times128$ lattice sites.
The directions of vorticity, flow and velocity gradient are now x, y and z, respectively.\\ 
The top picture in Fig. \ref{fig12} shows the order parameter $\phi$ after equilibration at time step $t=3.2\e{5}$ prior to shearing, which served as initial configuration for the runs with shear rate $\gd=1\e{-4}$ and $2\e{-4}$.
A corresponding configuration, which was equilibrated for another $3.2\e{5}$ timesteps was used for the run with $\gd=\e{-5}$.
As before black and white correspond to $\phi=-1$ and $\phi=1$, repectively, whereas gray scales give intermediate values of $\phi$.
The opacity of $phi$ has been linearly reduced from its full value at $\phi=-1$ to 0 at $\phi=1$ in order to reveal the lamellar structure close to the  boundaries.
In comparison to the 2D-case the isosurfaces of the cuboid, taken as quasi-two-dimensional equivalent in 3D, exhibit a slightly different morphology with shorter protrusions and an overall pattern, which appears somewhat more irregular.
Nonetheless the quiescent and equilibrated system C adopts the same lamellar spacing of about 10 lattice sites as system A.Fig. \ref{fig13} shows a slice through $C({\bm k})$ along $k_y=0$ (cuts along $k_x,k_z=0$ look identical), which resembles the left picture in Fig. \ref{fig4} and differs only in the magniude.\\ 
However, in shear flow system A and C behave very differently.
Even for the highest shear rate we applied and contrary to the 2D-case system C orders readily into a regular lamellar structure.
The centre picture of Fig.\ref{fig12} shows an example at timestep $t=9.6\e{5}$ and shear rate $\gd=1\e{-4}$, which we picked as this was the least ordered one of all three runs.
With a total number of 13 layers the lamellar spacing is still close to the equilibrium value of roughly 10 lattice sites.
Besides some irregularities at the front boundary there are obviously edge defects, caused by a branching of the lamellae and visible at the very top and bottom in the foreground and more towards the centre in the background.
\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run786_320.jpg}
\caption{Structure factor $C({\mathbf k})$ of quiescent system C, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$. Shown are wavevectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
\label{fig13}
\end{figure}
\begin{figure*}[!]
\centering
\includegraphics[angle=0,width=0.33\textwidth]{ck_x-slice_run788_960.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_y-slice_run788_960.jpg}
\includegraphics[angle=0,width=0.33\textwidth]{ck_z-slice_run788_960.jpg}
\caption{Structure factor $C({\mathbf k})$ of system C in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wavevectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$. Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
\label{fig14}
\end{figure*}
The edge defects emerge from the spontaneous shear-induced ordering of the phase separated structure but remain very localised at first.
As a result of shear advection the interfaces spread and close gaps at later times to finally form either continuous lamellae or uniform edges, which are oriented along the flow, virtually stable and do not anneal.
The fact that separated regions with similar order parameter values can now also connect along a third direction and expell regions with opposite order parameter, seems to play a key role pertaining to the formation of a homogenously layered structure.
The bottom picture in Fig. \ref{fig12} gives an idea about how the dimesionality affects the morphology of layered smectics and binary fluids.
It shows the two-dimensional system A as slice of system C along the direction of vorticity.
The difference between both systems is striking and even more surprising as both results were obtained with the same parameter setting and identical equilibration and shear histories.
The inset of Fig. \ref{fig6} quantifies the time it takes system C to order into a layererd defect-scarce structure.
Within a few $10^5$ timesteps the defect density of system C decreases to magnitudes, which we obtained in the 2D-case only for shear rates about one decade lower and times about a factor 10 longer.\\
Scarce defects lead to smaller chemical pressure, which is shown in the insets of Fig. \ref{fig8} for the component $P_{xy}$.
The hydrodynamic part of the stress is by far dominating and shear thinning occurs to an ultimate degree where the apparent viscosity of the fluid is that of the solvent and the chemical contribution has practically vanished.
A relation between the deviatoric chemical pressure tensor and the defect density like in Fig. \ref{fig9} can therefore not be established.
The inset of Fig.\ref{fig10} shows first normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$.
For all shear rates $\gd5\e{-5}-2\e{-4}$ system C takes similar values, which are monotonously approached and about one decade smaller than those for $N_1$ of system A.\\ 
Finally we want to glimpse at the structure factor of system C in shear flow.
Fig. \ref{fig14} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
It shows data in flow-velocity-gradient plane and compares directly with the left picture in Fig. \ref{fig4}.
The pictures in the centre and on the right give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
Note the scale of the logarithmic color map in the centre picture.
Despite a certain similarity with the 2D-structure after switch off in the rightmost picture of Fig. \ref{fig4}, the structure factor in three dimensions has a very pronounced peak at $k_z\simeq\pm0.2\pi/l$.
It appears to be split into two parts, but this is an effect of the relatively sparse discretization and caused by the visualization software.
Although strongly preaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
Fig. \ref{fig15} depicts the structure factor in 3D, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
It reveals a banana-shaped spatial dependence with some small appendices in flow direction at $k_x=0$.

\begin{figure}[!]
\centering
\includegraphics[angle=0,width=0.45\textwidth]{ck_run788_960_zoom.jpg}
\caption{Structure factor $C({\mathbf k})$ of system C in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
\label{fig15}
\end{figure}

\begin{table}[htp]
\small
\caption{Simulation parameters: The systems were first equilibrated from spinodal decomposition at zero shear rate (EQ), followed by various shear rates, ranging from $\dot{\gamma}=2\cdot10^{-4}$ till $1.25\cdot10^{-5}$. For some runs a final switch off (SO) at zero shear rate was performed. The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
  \label{tab1}
%  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lllll}
  \begin{tabular*}{0.495\textwidth}{lllll}
\hline
run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
\hline
system A & & & &\\
\hline 
703 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
704 & $1024\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
705 & $1024\times512\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
\hline
767 & $512\times256\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
775 & $512\times256\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
769 & $512\times256\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
771 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
772 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
774 & $512\times256\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
\hline
792 & $256\times128\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
system B & & & &\\
\hline
707 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
710 & $1024\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
765 & $1024\times512\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
\hline
776 & $512\times256\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
777 & $512\times256\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
778 & $512\times256\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
780 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
781 & $512\times256\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
782 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
\hline
system C & & & &\\
\hline
786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
\hline
\end{tabular*}
\end{table}

\section{Conclusions}
We have performed large scale simulations of a lamellar system in steady shear flow  in two and three dimensions.
We were able to relate the morphology of the order parameter in the quiescent state to results from previous studies of the same system using similar parameters.\\
In two dimensions and steady state simple shear flow at Peclet numbers ${\cal O}(10)$ we did not observe a homogenous lamellar structure for shear rates above a critical threshold, which depends on the exact parametrization of the free energy functional.
The morphology was rather inhomogenous instead, exhibiting regions that are under pressure with lamellar spacings smaller than the equilibrium value and others that were stretched.
We could relate the structure to typical values of rheological quantities.
Areas under extensional flow coincide with locally negative or small positive shear stresses and first normal stress differences, whereas both go to positive peak values in the compressed domains.
Below a critical shear rate the systems showed no tendency anymore to trip over itself and defects annealed as they could approach each other in adjacent lamellae.
The characteritic time of this process was set by the shear rate and the total strain.
The deviatoric part of the chemical pressure was correlated to the number of defect sites, wheras the latter was measured by means of an alignment factor, which quantified the localby curvature of the lamellar interfaces.
For the stiffer system A we could verify a linear relation between the defect density and the average chemical contribution to the shear stress.\\ 
In system A first normal stress differences decreased as long as a dynamic equilibrium between defect creation an annihilation existed.
Below the critical threshold this was no longer the case and pronounced defect annealing caused a rise in the first normal stress differences as the average lamellar spacing increased.
We could establish a shear thinning regime for the more defect ladden system A.
System B in contrary was so perfectly annealed in the steady state that virtually shear stresses were down at the hydrodynamic level and the relation between shear stress and shear rate was trivially linear.\\
In three dimensions we simulated shear strains which have not been attained so far.
Interestingly for the same parameters we observe striking differences between system A in 2D and system C in 3D.
It seems that the dynamics of the defects in two and three dimensions is rather incommensurable.
Even at the lowest applied shear rates the chemical stress has decayed to a level where the apparent viscosity was virtually identical to the solvent.
Structure factor data confirmed the very regular layered ordering of the structure and a lamellar spacing close to the equilibrium value.

\section{Acknowledgements}

\footnotesize{
\bibliography{smectics} %your .bib file
\bibliographystyle{rsc} %the RSC's .bst file
}

\end{document}

### 2011-08-19 papers/bprheo/bprheo.tex
Bprheo paper added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1445 40876441-3496-0410-a613-a2fca35b2ae2

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,pre,twocolumn,groupedaddress]{revtex4-1}
%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
\bibliographystyle{apsrev4-1}

\usepackage{graphicx}
\usepackage{amsmath}

\begin{document}

% Use the \preprint command to place your local institutional report
% number in the upper righthand corner of the title page in preprint mode.
% Multiple \preprint commands are allowed.
% Use the 'preprintnumbers' class option to override journal defaults
% to display numbers if necessary
%\preprint{}

%Title of paper
\title{Bulk Flow Behaviour of Cubic Blue Phases}

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current
% author. Explanatory text should go in the []'s, actual e-mail
% address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro foreach each type of information

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.
\author{O. Henrich$^{1,3}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}
\affiliation{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}

%\email[]{}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}

%Collaboration name if desired (requires use of superscriptaddress
%option in \documentclass). \noaffiliation is required (may also be
%used with the \author command).
%\collaboration can be followed by \email, \homepage, \thanks as well.
%\collaboration{}

%\noaffiliation

\date{\today}

\begin{abstract}
\end{abstract}

% insert suggested PACS numbers in braces on next line
\pacs{}
% insert suggested keywords - APS authors don't need to do this
%\keywords{}

%\maketitle must follow title, authors, abstract, \pacs, and \keywords
\maketitle

% body of paper here - Use proper section commands
% References should be done using the \cite, \ref, and \label commands
\section{Introduction}
% Put \label in argument of \section for cross-referencing
%\section{\label{}}
\section{Model and Methods}

Our approach is based on a well-established model for hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes the order structure of the liquid crystal by a traceless, symmetric tensor order parameter ${\bf Q}({\bf r})$. 
In uniaxial approximation the tensor order paramter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$, whereas ${\bf n}$ represents the local orientation of the liquid crystal molecules and brackets indicate mesoscopic averaging over a large number individual molecules.
Its largest eigenvalue $\frac{2}{3}q$ with $0<q<1$ characterizes the degree of order around the corresponding axis. 
The equilibrium properties of the system are determined by a Landau-de Gennes free energy density $f$ and the functional ${\cal F}[Q]=\int d^3{\bf r} f({\bf Q})$.
The free energy density consists of two different parts, namely a bulk contribution $f_b$ and a gradient term $f_g$.

\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{eqn1}
\end{eqnarray}

The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ and features a the first-order transition from the isotropic to the ordered phase at $\gamma>2.7$.
The second part comprises the elastic contributions to the free energy due to splay-, bend- and twist-deformations and assume for simplicity that they have all the same elastic constant $K$.
The pitch length $p$ of the cholesteric liquid crystal is related to the wavervector $q_0=2\pi/p$.
In order to specify a thermodynamic state it is useful to introduce two dimensionless quantities, the effective temperature $\tau$ and chirality $\kappa$.
They are given by

\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber,
\end{eqnarray}

whereas the chirality qqnatifies the ratio between gradient and bulk free energy.\\
The equation of motion of the tensor order parameter reads

\begin{equation}\label{}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\bf Q} - {\bf S}({\bf W},{\bf Q}) = \Gamma {\bf H}.
\label{eqn2}
\end{equation}

The first term on the lefthand side of Eq.(\ref{eqn2}) is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$:

\begin{eqnarray}
{\bf S}({\bf W}, {\bf Q}) &=& (\xi {\bf A} + {\boldsymbol \Omega})({\bf Q}+\frac{\bf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\bf Q}+\frac{\bf I}{3})(\xi {\bf A}  - {\boldsymbol \Omega})-2 \xi ({\bf Q}+\frac{\bf I}{3})Tr({\bf Q W}),
\label{eqn3}
\end{eqnarray}

where $Tr$ denotes the trace operation and ${\bf A}=({\bf W}+{\bf W}^T)/2$ and ${\boldsymbol \Omega}=({\bf W}-{\bf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
The molecular details of the liquid crystal such as the aspect ratio of its mesogens are embodied in the constant $\xi$, also know as the tumbling parameter.
The righthand side relates the rate of change to the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\bf Q$.

\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{eqn4}
\end{equation}

The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q^2/\Gamma$.\\
The time evolution of density and velocity are goverend by the continuity equation Eq. (\ref{eqn5})

\begin{eqnarray}
\partial_t \rho + \partial_\alpha (\rho u_\alpha)=0
\label{eqn5}
\end{eqnarray}

and the Navier-Stokes equation Eq. (\ref{eqn6})

\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}\nonumber\\
&+&\hspace*{-2cm}\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}. 
\label{eqn6}
\end{eqnarray}

At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq. (\ref{eqn6}) can be safely neglected.
The pressure tensor reads explicitely

\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}

It comprises the terms that give rise to an additional stress in the ordered state.
In the isotropic state when ${\bf Q}\equiv 0$ Eq. (\ref{eqn7}) is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Denniston:2004, Marenduzzo:2008} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the hydrodynamic equations (\ref{eqn5}) and (\ref{eqn6}) and a finite difference scheme for the tensor order parameter equation (\ref{eqn2}).

\section{Results and Discussion}

In the following we report results on the bulk flow behaviour of BPI and BPII for thermodynamic states characterized by $\tau=-0.5$ and $\kappa=1.0$ (BPI) and $\kappa=2.0$ (BPII), respectively.
For these parameters both blue phases are equilibrium phases in the quiescent state.
We have also performed simulations at higher temperatures and chiralities and for metastable configurations, which led to very similar results.
Quantitative differences occured only in that sense that the disclination network dissolved more easily for a particular flow rate if the system was closer to the cholesteric-isotropic phase boundary and where the order parameter takes lower values and free energy density is higher.\\
The tensor order parameter was initialized with the high-chirality solution of Eq. \ref{eqn1}, which has the correct topology of the actual blue phase.
The initial configurations were equilibrated for $5000$ LB time steps, where we allowed for optimization of the unit cell size \cite{Alexander:2006}, a feature that is often referred to as 'redshift'.
After the equilibration sequence the unit cell size was fixed and the shear flow was switched on.
For BPII we used a pitch length of 32 lattice sites, but in case of BPI it was necessary to increase the resolution to 64 sites per pitch length to get accurate results of the flow induced transformations in the disclination network.
In order to impose a flow profile we used Lees-Edwards boundary conditions \cite{Wagner:2002}, whereas the top (bottom) half of planes moved in positive (negative) y-direction.
We applied a correction for spurious momentum currents that emerge due to interpolations accross the Lees-Edwards planes \cite{Henrich:2011b}.
The imposed flow rates ranged over two orders of magnitude from about $1.25\cdot 10^{-6}$ to $6\cdot10^{-4}$ in LB units.\\
In the visualizations that follow the orientation of the principal coordinate axes in x-, y- and z-direction always coincides with the direction of the velocity gradient direction, the flow and the vorticity, respectively.
Disclination lines are plotted as isosurface of the scalar order parameter $q$

\subsection{Blue Phase II}

We start the discussion with BPII, because its flow behaviour is simpler and probably more easily anticipated.
Fig \ref{fig1} shows the disclination network undergoing a flow-induced transformation.
The periodically recurring structure remains very regular and apart from the quasi-affine transformation relatively close to its quiescent appearance with individual disclination lines breaking up and reconnecting further downstream.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\cdot 10^{-4}$. The horizontal y-direction is the flow direction whereas the velocity gradient is oriented along the vertical x-direction.}
\label{fig1}
\end{figure*}

Interestingly, while being steadily sheared in y-direction, the entire disclination network is slowly displaced in z-direction, the direction of vorticity.
This feature is most clearly observed when the structure is viewed along the flow direction.
Fig. \ref{fig2} depicts a sequence of close-ups of the disclination network as it travels in negative z-direction, i.e. from the foreground towards the background of the picture. 
The velocity $v_z$, represented by the colour code, exhibits a recurring pattern of positive and negative values.
Peak velocities occur close to disclination lines where the molecular order is suppressed or in regions where two disclination lines approach each other and the network is about to break up or reform.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-149k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-160k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-167k_run902.png}\\
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-173k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-182k_run902.png}
\includegraphics[width=0.32\textwidth]{slice-disc-v_z-192k_run902.png}\\
\caption{A section of $2\times2\times1$ unit cells of BPII seen along the direction of vorticity. The colour code gives the magnitude and direction of the velocity component $v_z$ perpendicular to the plane.}
\label{fig2}
\end{figure*}

Plotting only x- and z-component of the local velocity allows to visualize the spatial dependence of the secondary flow in the blue phase with respect to the disclination network.
Fig. \ref{fig3} shows characteristic regions of upward and downward 'flow', which are normally superimposed by the primary shear flow in y-direction.
Remarkably, the secondary flow is directly connected to the sense of the underlying cholesteric helix as changing from left- to right-handed helix also inverts the sense of motion of the network.
%%%%%%%%%%%%%%%%%%%%

The lefthand picture depicts the situation the movement of the disclination network is downwards, while helix inversion leads to an upward motion in the righthand picture.
More precisely the orientation of the velocities with respect to the bands is exactly inverted in both cases.

A closer look at the shape of the disclination lines reveals that also the network seems to be inverted. 
The total magnitude of the secondary flow is in the range of a few percent of the primary flow and the images in Fig. \ref{fig3} give actually the situation close to the peak point.\\

A secondary flow pattern with distinctive regions of positive and negative velocity components $v_z$ emerges and reoccurs periodically.

\begin{figure*}[h]
\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run902.png}
\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run903.png}
\caption{Secondary flow in BPII visualized by means of the velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity of the cholesteric helix. The images depict an individual snapshot of the periodically recurring sequence in the steady state.}
\label{fig3}
\end{figure*}

It is interesting to study the flow behaviour from a more quantitative point of view by looking at the shear stress dependence on flow rate and time.

The effective viscosity $\eta_{eff}$ over time is shown in Fig. \ref{fig4}. 
For all apart from the highest flow rate it oscillates sinusoidally, exhibiting symmetric minima and maxima and decreasing mean values, indicating slightly shear-thinning behaviour.
The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig1}.
However, due to the motion of the network in z-direction it turns out that it takes four periods altogether for the network to move one unit cell along the direction of vorticity and to complete one cycle.
The inset shows the deviatoric part of the chemical stress tensor as a function of shear rate.
The dependence is that of a power law $\sigma_{eff, xy}=a \dot{\gamma}^b$ with $a=0.34(6116), b=0.95(2654)$ for low and intermediate shear rates. 
The error bars indicate the maximum and minimum stress during one cycle.
The idependence is almost linear and gives further evidence of the remarkably small degree of shear-thinning and also reflected in the very regular and undisturbed affine transformation of the disclination network.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{stress_bp2.pdf}
\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPII in the steady state. }
\label{fig4}
\end{figure}

Above a critical shear rate, which is $\dot{\gamma}=3.1-6.2\cdot 10^{-4}$ in our case, the rotational diffusion coefficient $\Gamma$ in Eq. \ref{eqn2} is too small and the disclination network adopts a completely different conformation in the steady state.
Fig. \ref{fig5} shows the director field and the position of disclination plane.
The degree of twist is the same as that of the underlying cholesteric, but the orientation of the director on either side of the planes differs by almost 90 degrees, giving rise to four distinct layers, which are separated by disclination planes.
Even in this flow-induced state the disclinations move and retain the above described sense of motion in the direction of vorticity.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{dir-op-y_225k_run949.png}
\caption{BPII at high shear rate: director field and the disclination planes seen along the flow direction. The BPII breaks up and flows in helical layers separated from each other by a disclination plane}
\label{fig5}
\end{figure}

\subsection{Blue Phase I}

The flow behaviour of BPI deviates significantly from that of BPII in a number of points.
First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig1}, but this time seen along the direction of flow.
A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
The initial and final state in this sequence correspond to half a cycle. 
This can be easily verified by comparing the position of the disclination network at the left or right boundary.
BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI seen along the flow direction in steady shear flow at $\dot{\gamma}=1.56\cdot 10^{-4}$, showing the downward movement of the disclination network in the direction of vorticity.}
\caption{\label{fig6}}
\end{figure*}

The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.

\begin{figure*}[h]
\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run914.png}
\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run922.png}
\caption{Secondary flow in BPI: velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity.}
\label{fig7}
\end{figure*}

Despite qualitative similarities between the BPI and BPII, the quantitative analysis reveals striking differences.
In Fig. \ref{fig8} we show the effective viscosity $\eta_{eff}$ over time and flow curves $\sigma_{chem}(\dot{\gamma})$. 
BPI exhibits pronounced shear-thinning contrary to BPII, whereas the shape of the curves in no longer symmetric.
In fact on approaching the critical shear rate that led in the case of BPII to the disruption of the network, the curves become increasingly complex but remain fully periodic.
The effective viscosity drops by about an order of magnitude between the lowest and the highest shear rates.
Note that the scale has been adapted to allow for the different magnitudes.
The inset can be directly compared with that of Fig. \ref{fig4}.
Interestingly the difference between minima and maxima of the chemical shear stress is much larger than for BPII and for the lowest shear rate we applied the absolute minimum takes even negative values, which are not shown in Fig. \ref{fig4}.
Our understanding of this phenomenon is an interplay of the elastic forces and the individual order paramter configuration of the liquid crystal.
If the system is close to a local minimum of the free energy but moving away from it the increasing strain pumps elastic energy into the system. 
This results in positive chemical stresses as the system tries to withstand the deformation.
However, if the system has sufficiently transformed and is about to overcome a local maximum, it will be pulled towards the next local minumum, which can lead temporarily to negative chemical stresses. 
The inset shows that the chemical stress decreases monotonically with decreasing flow rate, but does not follow a simple power law dependence.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{stress_bp1.pdf}
\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPI in the steady state.}
\label{fig8}
\end{figure}

A Fourier analysis of the time series of the deviatoric chemical stress $\sigma_{eff,xy}$ or alternatively the free energy density $f$ can provide a further insight into the flow behaviour of BPI.
Fig. \ref{fig9} shows the Fourier spectrum of the time series of the chemical stress.
For simplicity we have always doubled the shear rate, so that the first overtone ocoalesces with the fundamental mode of the next highest shear rate. 
At low shear rates we observe a strong contribution of the first harmonic, which is directly related to the recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
The magnitude of higher harmonics decreases monotonously with increasing frequency.
Interestingly the inset of Fig. \ref{fig9} shows that at intermediate shear rates the contribution of higher harmonics becomes more pronounced until after a kind of gap and similar break up the fundamental mode dominates again and the former characteristics are restored.
Remarkably the break up in case of BPI occurs in exactly the same flow rate window as BPII.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{spectrum_bp1.pdf}
\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{fig9}
\end{figure}

Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.

\begin{figure}[h]
\includegraphics[width=0.5\textwidth]{dir-op-z_395k_run916.png}
\includegraphics[width=0.5\textwidth]{dir-op-x_395k_run916.png}
\caption{BPI at high shear rate: director field and the disclination planes seen along the direction of voritcity and the gradient direction. BPI forms rolls with $\pi/2$ disclinations that are advected with the general flow.}
\label{fig10}
\end{figure}

\section{Conclusions}

% If in two-column mode, this environment will change to single-column
% format so that long equations can be displayed. Use
% sparingly.
%\begin{widetext}
% put long equation here
%\end{widetext}

% figures should be put into the text as floats.
% Use the graphics or graphicx packages (distributed with LaTeX2e)
% and the \includegraphics macro defined in those packages.
% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz,
% and Frank Mittelbach for instance.
%
% Here is an example of the general form of a figure:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Use the figure* environment if the figure should span across the
% entire page. There is no need to do explicit centering.

% \begin{figure}
% \includegraphics{}%
% \caption{\label{}}
% \end{figure}

% Surround figure environment with turnpage environment for landscape
% figure
% \begin{turnpage}
% \begin{figure}
% \includegraphics{}%
% \caption{\label{}}
% \end{figure}
% \end{turnpage}

% tables should appear as floats within the text
%
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
% The ruledtabular enviroment adds doubled rules to table and sets a
% reasonable default table settings.
% Use the table* environment to get a full-width table in two-column
% Add \usepackage{longtable} and the longtable (or longtable*}
% environment for nicely formatted long tables. Or use the the [H]
% placement option to break a long table (with less control than 
% in longtable).
% \begin{table}%[H] add [H] placement to break table across pages
% \caption{\label{}}
% \begin{ruledtabular}
% \begin{tabular}{}
% Lines of table here ending with \\
% \end{tabular}
% \end{ruledtabular}
% \end{table}

% Surround table environment with turnpage environment for landscape
% table
% \begin{turnpage}
% \begin{table}
% \caption{\label{}}
% \begin{ruledtabular}
% \begin{tabular}{}
% \end{tabular}
% \end{ruledtabular}
% \end{table}
% \end{turnpage}

% Specify following sections are appendices. Use \appendix* if there
% only one appendix.
\appendix*

\begin{table*}
\begin{tabular}{|c|| c | c| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $\tau$ & $\kappa$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
1.95 & -0.5 & 1.0 &0.1388 & -2.49 &2.50 &4.51 &-124.65 &124.82 &3.62 &-1.62 &1.89 &3.51 \\
3.91 & -0.5 & 1.0 &0.1388 & -3.31 &3.35 &4.21 &-248.69 &248.74 &3.90 &-2.88 &2.56 &4.39 \\
7.81 & -0.5 & 1.0 &0.1388 &-6.49 &6.56 &7.99 &-497.22 &496.16 &4.46 &-7.46 &5.31 &6.81 \\ 
15.63 & -0.5 & 1.0 &0.1388 &-3.08 &3.17 &10.49 &-992.46 &993.18 &10.36 &-3.57 &2.87 &10.54 \\
15.63 & -0.5 & 1.0 &-0.1388 &-3.08 &3.17 &10.49 &-992.46 &993.18 &10.36 &-2.87 &3.57 &10.54 \\
31.25 & -0.5 & 1.0 &0.1388 &-4.08 &4.46 &15.94 & -1988.36 &1986.71 &16.85 &-12.16 &11.37 &19.38\\
62.50 & -0.5 & 1.0 &0.1388 & -62.63 & 62.13 & 24.68 & -4039.41 &3995.3  & 20.57 &-110.76  &73.52 & 33.26 \\
\hline
BPII \\
\hline
1.95 & -0.5 & 2.0 & 0.1963 & -1.09 &1.07 & 0.76 &  -62.04  &61.60 & 0.97 & -1.55 &1.64 & 0.81\\
3.91 & -0.5 & 2.0 & 0.1963&-1.66 &1.70 & 1.45 &-123.93 &123.18 & 1.76 &-2.73 &3.09 &1.47\\
7.81 &  -0.5 & 2.0 &0.1963& -2.62 &2.71 & 2.67 &-247.67 &246.45 & 3.20 &-4.77 &5.78 &2.74\\
15.63 & -0.5 & 2.0 & 0.1963&-3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-7.66 &10.00 &4.33\\
15.63 & -0.5 & 2.0 &-0.1963& -3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-10.00 &7.66 &4.33\\
31.25 & -0.5 & 2.0 &0.1963 &-1.79 &1.82 &5.86 &-988.73 &985.93 &8.96  &-11.04 &14.39 &6.35\\
62.50 & -0.5 & 2.0 &0.1963 & -1.20 & 1.20 & 0.38 & -1969.41  & 1969.41 & 3.07 &-1.48 &0.96 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge$200000 for BPI and t$\ge$30000 in case of BPII.
\label{table1}}
\end{table*}
%\section{}

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
Thanks folks!
\end{acknowledgments}

% Create the reference section using BibTeX:
\bibliography{bprheo}

\end{document}
%
% ****** End of file apstemplate.tex ******

### 2011-08-23 papers/lmcproc/iopguidelines/IOPLaTeXGuidelines.tex
LMC proceeding added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1448 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title[Author guidelines for IOP journals in  \LaTeXe]{How to prepare and submit an article for 
publication in an IOP journal using \LaTeXe}

\author{}

\address{IOP 
Publishing, Dirac
House, Temple Back, Bristol BS1 6BE, UK}
\ead{custserv@iop.org}
\begin{abstract}
This document describes the  preparation of an article using \LaTeXe\ and 
\verb"iopart.cls" (the IOP \LaTeXe\ preprint class file).
This class file is designed to help 
authors produce preprints in a form suitable for submission to any of the
journals published by IOP Publishing.
Authors submitting to any IOP journal, i.e.\ 
both single- and double-column ones, should follow the guidelines set out here. 
On acceptance, their TeX code will be converted to 
the appropriate format for the journal concerned.

\end{abstract}

%Uncomment for PACS numbers title message
%\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
%\vspace{2pc}
%\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
% Comment out if separate title page not required
\maketitle

\section{Introduction: file preparation and submission}
The \verb"iopart" \LaTeXe\ article class file is provided to help authors prepare articles for submission to the IOP journals. Submission is not restricted to those using of this class file and articles prepared using any other class or style files can also be submitted. Use of the \verb"iopart" class file does, however, help to speed the publication process and it is hoped it will make it easy for authors to prepare their articles. This document describes and demonstrates the use of \verb"iopart.cls" to produce a preprint for submission and refereeing. The detailed formatting for print and web use will be undertaken by IOP as part of the production process after the article has been accepted for publication, but use of \verb"iopart" does make this simpler.

\subsection{What you will need to supply}
Full details on how to submit files to a particular IOP journal are contained in the document {\it Guidelines for authors} which is discussed in section \ref{agide}. Here we mention the key points you need to consider when preparing your files for submission.

\subsubsection{{\bf Text}} The \LaTeX\ source code of your paper with all figures included in 
the source code (see section \ref{figinc}). Although we recommend that authors use the \verb"iopart" class file, articles prepared using almost any  version of \TeX\ or \LaTeX\ 
can be handled.  
\begin{itemize}
\item {\bf Using \LaTeX\ packages} Most \LaTeXe\ packages can be used if they are 
available in the normal distribution of \LaTeXe; however, if it is essential to use 
a non-standard package then any extra files needed to process the article must 
also be sent in. Authors should be aware that the final version will be printed on 
a different page size and using different fonts to the preprint version so that any 
special effects used should not contain material that is not easily scalable.
Alterations to the source code will be made during the production process in order to 
conform to IOP house style and journal format.
\end{itemize}
\subsubsection{\bf Figures} Figures must be included in an article as encapsulated PostScript files
(see section \ref{figinc}) or created using standard \LaTeX\ drawing commands. 
 Please name all figure files using the guidelines in section \ref{fname}. For more detail
on preparation of EPS figure files, please refer to IOP's graphics guidelines (\gguide) 
which can be downloaded from authors.iop.org.

\begin{itemize}
\item {\bf Including figures\label{fig1}} All figures can be included within the body of the text 
at an appropriate point or grouped together with their captions at the end of the article. A standard graphics inclusion package such as \verb"graphicx" should be used for figure inclusion.
Authors should avoid using special effects generated by including verbatim
PostScript code. Wherever possible, please try to use standard \LaTeX\ tools 
and packages.
\end{itemize}

\subsubsection{\bf BiBTeX \label{bibby}}  If you use BiBTeX to prepare your
references, our preferred  \verb".bst" styles are:

\begin{itemize}
\item for the numerical (Vancouver) reference style we recommend that authors use 
 \verb"unsrt.bst";
\item for alphabetic (Harvard) style reference we recommend that authors use
the \verb"jphysicsB.bst" BiBTeX style file together with the harvard \LaTeX\ package. 
\verb"jphysicsB.bst" is supplied as part of the \verb"harvard.sty" package which can be 
downloaded from www.ctan.org.
\end{itemize}

Please make sure that you include your .bib bibliographic database file(s) and any 
.bst style file(s) you have used (other than unsrt.bst or jphysicsB.bst).

\subsubsection{\bf Copyrighted material} If you wish to illustrate your article using material for which you do not own the copyright then you must seek permission from the copyright holder, usually both the author and the publisher. It is the author's responsibility to obtain copyright permissions and this should be done prior to submitting your article. If you have obtained permission, please provide full details of the permission granted---for example, copies of the text of any e-mails or a copy of any letters you may have received. Figure captions must include an acknowledgment of the original source of the material even when permission to use has been obtained. Full details on copyright can be found in IOP's {\it Guidelines for authors} which can be accessed online at authors.iop.org.

\subsection{Naming your files}
Please name all your files, both figures and text as follows:
\begin{itemize}
\item use only characters from the set a to z, A to Z, 0 to 9 and underscore (\_); 
\item do not use spaces in file names;
\item include an extension to indicate the file type (e.g., .tex, .eps, .txt, etc);
\item do not use any accented characters in file names; for example, \'a, \^e, \~n, \"o, etc because 
they can cause difficulties when processing your files.
\end{itemize}
\subsubsection{\label{fname}Naming your figure files} In addition to the above points, please give each figure file a name which indicates the number of the figure it contains; for example, figure1.eps, figure2a.eps etc. If the figure file contains a figure with multiple parts, for example figure 2(a) to 2(e), give it a name such as figure2a\_2e.eps, and so forth.
\subsection{Submitting your article files to an IOP journal}
For most IOP journals you send your files directly to us by web, e-mail or FTP; however, there are a number of exceptions to this. For certain IOP journals you need to send your files to a different institution or to an external editor. How to find the submission address for a particular journal is discussed in section 1.4, below.
\subsection{Where to send your files \label{agide}}
Full details of how to submit files to a particular IOP journal are contained in IOP's {\it Guidelines for authors} which can be accessed online by going to authors.iop.org. From this document's table of contents, select the `How to submit' link to access full details of the submission address for a particular journal.

\subsection{Obtaining the class file and documentation}
The  \LaTeXe\ class file and guidelines can be downloaded from \verb"authors.iop.org".
Files are available in zipped format for PCs, tar compressed format files for
Unix or as a stuffit archive for Macintosh.

\section{Preparing your article}

Please follow these guidelines as closely as possible, particularly with regard to the preparation of the reference list.

At the start of the \LaTeX\ source please include 
commented material to identify the journal, author and reference number if 
known. The first non-commented line should be 
\verb"\documentclass[12pt]{iopart}"  to load the preprint class 
file. Omitting \verb"[12pt]" produces an article with the normal journal
page and type sizes. The start of the article text is signalled by 
\verb"\begin{document}".
Authors of very long articles may find it convenient to separate 
their article into a series of files each containing a section, each of which is called 
in turn by the primary file.

Authors may add their own macros 
at the start of an article provided they do not overwrite existing definitions and 
that they send copies of their new macros with their text file. Macros for the individual paper not included 
in a style file should be inserted in the preamble to the paper with comments to
describe any complex or non-obvious ones.
\verb"iopart" can be used with other package files such
as those loading the AMS extension fonts 
\verb"msam" and \verb"msbm" (these fonts provide the 
blackboard bold alphabet and various extra maths symbols as well as symbols useful in figure 
captions); an extra style file \verb"iopams.sty" is provided to load these
packages and provide extra definitions for bold Greek letters.

\subsection{Double-column journals}
Authors writing for double-column journals should use the \verb"iopart" class file. Conversion from the single-column format to the double-column output required for printing will be done during the production process. 
However, authors should bear in mind that all mathematical formulae will need to be fitted
into the width of a single column, so individual lines of equations should not occupy more than two thirds of the line width in this preprint form.

\section{The title and abstract page} 
The code for setting the title page information is slightly different from
the normal default in \LaTeX.

\subsection{Titles and article types}
The title is set using the command
\verb"\title{#1}", where \verb"#1" is the title of the article. The
first letter 
of the title should be capitalized with the rest in lower case. 
Mathematical expressions within the title may be left in light-face type.

If the title is too long to use as a running head at the top of each page (apart from the
first) a short
form can be provided as an optional argument (in square brackets)
before the full title, i.e.\ \verb"\title[Short title]{Full title}".

For article types other than papers, \verb"iopart.cls"
has a generic heading \verb"\article[Short title]{TYPE}{Full title}" 
and the specific definitions given in table~\ref{arttype}. In each case (apart from Letters
to the Editor) an 
optional argument can be used immediately after the control sequence name
to specify the short title; where no short title is given the full title
will be used as the running head. 
For Letters use \verb"\letter{Full title}", no short title is required as 
the running head is automatically defined to be {\it Letter to the Editor}.
The generic heading could be used for 
articles such as those presented at a conference or workshop, e.g.
\small\begin{verbatim}
\article[Short title]{WORKSHOP ON HIGH-ENERGY PHYSICS}{Title}
\end{verbatim}\normalsize
Footnotes to titles may be given, though acknowledgment of 
funding should be included in the acknowledgments section rather than here. A footnote can be included by using \verb"\footnote{Text of footnote.}" immediately after the title.

\begin{table}
\caption{\label{arttype}Types of article defined in the {\tt iopart.cls} 
class file.}
\footnotesize\rm
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus12pt}}l}}
\br
Command&Type&Heading on first page\\
\mr
\verb"\title{#1}"&Paper&---\\
\verb"\review{#1}"&Review&REVIEW\\
\verb"\topical{#1}"&Topical review&TOPICAL REVIEW\\
\verb"\comment{#1}"&Comment&COMMENT\\
\verb"\note{#1}"&Note&NOTE\\
\verb"\paper{#1}"&Paper&---\\
\verb"\prelim{#1}"&Preliminary communication&PRELIMINARY COMMUNICATION\\
\verb"\rapid{#1}"&Rapid communication&RAPID COMMUNICATION\\
\verb"\letter{#1}"&Letter&LETTER TO THE EDITOR\\
\verb"\article{#1}{#2}"&Other articles&Whatever is entered as {\tt 
\#1}\\
\br
\end{tabular*}
\end{table}

\subsection{Authors' names and addresses}
The next information required is the list of authors' names and 
their affiliations. For the authors' names type \verb"\author{#1}", 
where \verb"#1" is the 
list of all authors' names. The style for the names is initials then
family name, with a comma after all but the last 
two names, which are separated by `and'. Initials should {\it not} be followed by full stops. First (given) names may be used if 
desired and Chinese-style names included in the form they should be printed in. If the authors are at different addresses a superscripted number, e.g. $^1$, \verb"$^1$", should be used after each 
name to reference the author to his/her address. 
If an author has additional information to appear as a footnote, such as 
a permanent address, a normal \LaTeX\ footnote command
should be given after the family name and address marker 
with this extra information.

The addresses of the authors' affiliations follow the list of authors. 
Each address is set by using
\verb"\address{#1}" with the address as the single parameter in braces. 
If there is more 
than one address then the appropriate superscripted number, followed by a space, should come at the start of
the address.

Please also add the e-mail addresses for at least one of the authors. This is done by inserting the 
command \verb"\ead{#1}" after the postal address(es) where \verb"#1" is the e-mail address.  
See section~\ref{startsample} for sample coding. For more than one e-mail address, please use the command 
\verb"\eads{\mailto{#1}, \mailto{#2}}" with \verb"\mailto" surrounding each e-mail address.

\subsection{The abstract}
The abstract follows the addresses and
should give readers concise information about the content 
of the article and indicate the main results obtained and conclusions 
drawn. It should be self-contained with no reference to 
figures, tables, equations, bibliographic references etc
included and should not normally exceed 200 
words.
To indicate the start 
of the abstract type \verb"\begin{abstract}" followed by the text of the 
abstract (not in braces).  The abstract should normally be restricted 
to a single paragraph and is terminated by the command
\verb"\end{abstract}"

\subsection{Subject classification numbers}
Any Physics and Astronomy Classification System (PACS) 
codes or Mathematics Subject Classification (MSC) scheme 
numbers should come immediately after the abstract. 
Classification codes can greatly help in the choice of suitable 
referees and allocation of articles to subject areas. 
For {\it Inverse Problems} and {\it Nonlinearity} authors may 
use either PACS or MSC codes.

PACS or MSC numbers are included after the abstract 
using \verb"\pacs{#1}" and \verb"\ams{#1}" respectively.

After any classification numbers the command
\verb"\submitto{#1}" can be inserted, where \verb"#1" is the journal name written in full or the appropriate control sequence as
given in table~\ref{jlab1}. This command is not essential to the running of the file.

\subsubsection{Information on PACS and MSC}
For more information on PACS and MSC see
\begin{itemize}
\item MSC: http://www.ams.org/msc
\item PACS: http://www.aip.org/pacs
\end{itemize}

\subsection{Keywords}
Keywords should be provided for submissions to {\it Measurement Science and Technology, Physical Biology, Physiological Measurement}, and both parts of {\it Journal of Optics\/}. Add these as a new paragraph starting \verb"\noindent{\it Keywords\/}:" after the end of the abstract.

\subsection{Making a separate title page}
The command \verb"\maketitle" forces a page break after the point where it
is inserted and so to keep the header material on a separate page from the
body of the text insert \verb"\maketitle" or \verb"\newpage" before the start of the text. 
If \verb"\maketitle" is not included the text of the
article will start immediately after the abstract.

\subsection{Sample coding for the start of an article}
\label{startsample}
The code for the start of a title page of a typical paper might read:
\small\begin{verbatim}
\documentclass[12pt]{iopart}
\begin{document}
\title[The anomalous magnetic moment of the 
neutrino]{The anomalous magnetic moment of the 
neutrino and its relation to the solar neutrino problem}

\author{P J Smith$^1$, T M Collins$^2$, 
R J Jones$^3$\footnote{Present address:
Department of Physics, University of Bristol, Tyndalls Park Road, 
Bristol BS8 1TS, UK.} and Janet Williams$^3$}

\address{$^1$ Mathematics Faculty, Open University, 
Milton Keynes MK7~6AA, UK}
\address{$^2$ Department of Mathematics, 
Imperial College, Prince Consort Road, London SW7~2BZ, UK}
\address{$^3$ Department of Computer Science, 
University College London, Gower Street, London WC1E~6BT, UK}
\ead{williams@ucl.ac.uk}

\begin{abstract}
...
\end{abstract}
\pacs{1315, 9440T}
\submitto{\JPG}
\maketitle
\end{verbatim}
\normalsize

\section{The text}
\subsection{Sections, subsections and subsubsections}
The text of articles may be divided into sections, subsections and, where necessary, 
subsubsections. To start a new section, end the previous paragraph and 
then include \verb"\section" followed by the section heading within braces. 
Numbering of sections is done {\it automatically} in the headings: 
sections will be numbered 1, 2, 3, etc, subsections will be numbered 
2.1, 2.2,  3.1, etc, and subsubsections will be numbered 2.3.1, 2.3.2, 
etc.  Cross references to other sections in the text should, where
possible, be made using 
labels (see section~\ref{xrefs}) but can also
be made manually. See section~\ref{eqnum} for information on the numbering of displayed equations. Subsections and subsubsections are 
similar to sections but 
the commands are \verb"\subsection" and \verb"\subsubsection" respectively. 
Sections have a bold heading, subsections an italic heading and 
subsubsections an italic heading with the text following on directly.
\small\begin{verbatim}
\section{This is the section title}
\subsection{This is the subsection title}
\end{verbatim}\normalsize

The first section is normally an introduction,  which should state clearly 
the object of the work, its scope and the main advances reported, with 
brief references to relevant results by other workers. In long papers it is 
helpful to indicate the way in which the paper is arranged and the results 
presented.

Footnotes should be avoided whenever possible and can often be included in the text as phrases or sentences in parentheses. If required, they should be used only for brief notes that do not fit conveniently into the text. The use of 
displayed mathematics in footnotes should be avoided wherever possible and no equations within a footnote should be numbered. 
The standard \LaTeX\ macro \verb"\footnote" should be used.

\subsection{Acknowledgments}
Authors wishing to acknowledge assistance or encouragement from 
colleagues, special work by technical staff or financial support from 
organizations should do so in an unnumbered `Acknowledgments' section 
immediately following the last numbered section of the paper. The 
command \verb"\ack" sets the acknowledgments heading as an unnumbered
section. For Letters 
\verb"\ack" does not set a heading but leaves a line space and does not 
indent the next paragraph.

\subsection{Appendices}
Technical detail that it is necessary to include, but that interrupts 
the flow of the article, may be consigned to an appendix. 
Any appendices should be included at the end of the main text of the paper, after the acknowledgments section (if any) but before the reference list.
If there are 
two or more appendices they should be called Appendix A, Appendix B, etc. 
Numbered equations will be in the form (A.1), (A.2), etc,
figures will appear as figure A1, figure B1, etc and tables as table A1,
table B1, etc.

The command \verb"\appendix" is used to signify the start of the
appendices. Thereafter \verb"\section", \verb"\subsection", etc, will 
give headings appropriate for an appendix. To obtain a simple heading of 
`Appendix' use the code \verb"\section*{Appendix}". If it contains
numbered equations, figures or tables the command \verb"\appendix" should
precede it and \verb"\setcounter{section}{1}" must follow it.

\subsection{Some matters of style}
It will help the readers if your article is written in a clear,
consistent and concise manner. The Production Department at IOP
Publishing will try to make sure that your work is presented to its
readers in the best possible way without sacrificing the individuality of
your writing. Some recommended 
points to note, however, are the following.
\begin{enumerate}
\item Authors are often inconsistent in the use of `ize' and `ise' endings.
We recommend using `-ize' spellings (diagonalize, 
renormalization, minimization, etc) but there are some common 
exceptions to this, for example: devise, 
promise and advise.

\item English spellings are mandatory (colour, flavour, behaviour, tunnelling, artefact, focused, focusing, fibre, etc) except in the following journals where either American or English spellings are acceptable: {\it Physical Biology}, {\it Smart Materials and Structures} and {\it Journal of Micromechanics and Microengineering}.

\item The words table and figure should be written 
in full and {\bf not} abbreviaged to tab. and fig. Do not include `eq.'\, `equation' etc before an equation number or `ref.'\, `reference' etc before a reference number.
\end{enumerate}

Please check your article carefully for accuracy, consistency and clarity before
submission. Remember that your article will probably be read by many
people whose native language is not English and who may not therefore 
be aware of many of the subtle meanings of words or idiomatic phases
present in the English language. It therefore helps if you try to keep
sentences as short and simple as possible.

\section{Mathematics}
\subsection{Two-line constructions}
The great advantage of \LaTeX\ 
over other text processing systems is its 
ability to handle mathematics to almost any degree of complexity. However, 
in order to produce an article suitable for publication both within a print journal and online, 
authors should exercise some restraint on the constructions used. Some equations using very small characters which are clear in a preprint style article may be difficult read in a smaller format or online at a relatively low resolution.
For simple fractions in the text the solidus \verb"/", as in 
$\lambda/2\pi$, should be used instead of \verb"\frac" or \verb"\over", 
care being taken to use parentheses where necessary to avoid ambiguity, 
for example to distinguish between $1/(n-1)$ and $1/n-1$. Exceptions to 
this are the proper fractions $\frac12$, $\frac13$, $\frac34$, 
etc, which are better left in this form. In displayed equations 
horizontal lines are preferable to solidi provided the equation is 
kept within a height of two lines. A two-line solidus should be 
avoided where possible; the construction $(\ldots)^{-1}$ should be 
used instead; for example use:
\begin{equation*}
\frac{1}{M_{\rm a}}\left(\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}\right)^{-1}\qquad\mbox{instead of}\qquad
\frac{1}{M_{\rm a}}\biggl/\int^\infty_0{\rm d}
\omega\;\frac{|S_o|^2}{N}.
\end{equation*}

\subsection{Roman and italic in mathematics}
In mathematics mode \LaTeX\ automatically sets variables in an italic 
font. In most cases authors should accept this italicization. However, 
there are some cases where it is better to use a Roman font; for 
instance, IOP journals use a Roman d for a differential d, a Roman e 
for an exponential e and a Roman i for the square root of $-1$. To 
accommodate this and to simplify the  typing of equations we have 
provided some extra definitions. \verb"\rmd", \verb"\rme" and \verb"\rmi" 
now give Roman d, e and i respectively for use in equations, 
e.g.\ $\rmi x\rme^{2x}\rmd x/\rmd y$ 
is obtained by typing \verb"$\rmi x\rme^{2x}\rmd x/\rmd y$".

Certain other common mathematical functions, such as cos, sin, det and 
ker, should appear in Roman type. \LaTeX\ provides macros for 
most of these functions 
(in the cases above, \verb"\cos", \verb"\sin", \verb"\det" and \verb"\ker" 
respectively); we have also provided 
additional definitions for $\Tr$, $\tr$ and 
$\Or$ (\verb"\Tr", \verb"\tr" and \verb"\Or", respectively).

Subscripts and superscripts should be in Roman type if they are labels 
rather than variables or characters that take values. For example in the 
equation
\[
\epsilon_m=-g\mu_{\rm n}Bm
\]
$m$, the $z$ component of the nuclear spin, is italic because it can have 
different values whereas n is Roman because it 
is a label meaning nuclear ($\mu_{\rm n}$ 
is the nuclear magneton).

\subsection{Displayed equations in double-column journals}
Authors should bear in mind that all mathematical formulae in double-column journals will need to be fitted
into the width of a single column, so individual lines of equations should not occupy more than two thirds of the line width in this preprint form.

\subsection{Special characters for mathematics}
Bold italic characters can be used in our journals to signify vectors (rather
than using an upright bold or an over arrow). To obtain this effect
use \verb"\bi{#1}" within maths mode, e.g. $\bi{ABCdef}$. If upright 
bold characters are required in maths use \verb"\mathbf{#1}" within maths
mode, e.g. $\mathbf{XYZabc}$. The calligraphic (script) uppercase alphabet
is obtained with \verb"\mathcal{AB}" or \verb"\cal{CD}" 
($\mathcal{AB}\cal{CD}$).

The American Mathematical Society provides a series of extra symbol fonts
to use with \LaTeX\ and packages containing the character definitions to
use these fonts. Authors wishing to use Fraktur 
\ifiopams$\mathfrak{ABC}$ \fi
or Blackboard Bold \ifiopams$\mathbb{XYZ}$ \fi can include the appropriate
AMS package (e.g. amsgen, amsfonts, amsbsy, amssymb) with a 
\verb"\usepackage" command or add the command \verb"\usepackage{iopams}"
which loads the four AMS packages mentioned above and also provides
definitions for extra bold characters (all Greek letters and some
additional other symbols).

The package iopams uses the definition \verb"\boldsymbol" in amsbsy
which allows individual non-alphabetical symbols and Greek letters to be 
made bold within equations.
The bold Greek lowercase letters \ifiopams$\balpha \ldots\bomega$,\fi 
are obtained with the commands 
\verb"\balpha" \dots\ \verb"\bomega" (but note that
bold eta\ifiopams, $\bfeta$,\fi\ is \verb"\bfeta" rather than \verb"\beta")
and the capitals\ifiopams, $\bGamma\ldots\bOmega$,\fi\ with commands 
\verb"\bGamma" \dots\
\verb"\bOmega". Bold versions of the following symbols are
predefined in iopams: 
bold partial\ifiopams, $\bpartial$,\fi\ \verb"\bpartial",
bold `ell'\ifiopams, $\bell$,\fi\  \verb"\bell", 
bold imath\ifiopams, $\bimath$,\fi\  \verb"\bimath", 
bold jmath\ifiopams, $\bjmath$,\fi\  \verb"\bjmath", 
bold infinity\ifiopams, $\binfty$,\fi\ \verb"\binfty", 
bold nabla\ifiopams, $\bnabla$,\fi\ \verb"\bnabla", 
bold centred dot\ifiopams, $\bdot$,\fi\  \verb"\bdot", other 
characters are made bold using 
\verb"\boldsymbol{\symbolname}".

Please do not use the style file \verb"amsmath.sty" (part of the AMSTeX package) in conjunction with \verb"iopart.cls". This will result in several errors. To make use of the macros defined in \verb"amsmath.sty", we have provided the file \verb"setstack.sty" which reproduces the following useful macros from \verb"amsmath.sty":
\small\begin{verbatim}
\overset \underset \sideset \substack \boxed   \leftroot
\uproot  \dddot    \ddddot  \varrow   \harrow
\end{verbatim}\normalsize

Table~\ref{math-tab2} lists some other macros for use in 
mathematics with a brief description of their purpose.

\begin{table}
\caption{\label{math-tab2}Other macros defined in IOP macros for use in maths.}
\begin{tabular*}{\textwidth}{@{}l*{15}{@{\extracolsep{0pt plus
12pt}}l}}
\br
Macro&Result&Description\\
\mr
\verb"\fl"&&Start line of equation full left\\
\verb"\case{#1}{#2}"&$\case{\#1}{\#2}$&Text style fraction in display\\
\verb"\Tr"&$\Tr$&Roman Tr (Trace)\\
\verb"\tr"&$\tr$&Roman tr (trace)\\
\verb"\Or"&$\Or$&Roman O (of order of)\\
\verb"\tdot{#1}"&$\tdot{x}$&Triple dot over character\\
\verb"\lshad"&$\lshad$&Text size left shadow bracket\\
\verb"\rshad"&$\rshad$&Text size right shadow bracket\\
\br
\end{tabular*}
\end{table}

\subsection{Alignment of displayed equations}
The normal style for aligning displayed equations is to align them left rather than centre them. The class file automatically does this and indents each line of a display. To make any line start at the left margin of the page, add \verb"\fl" at its start (to indicate full left).

Using the \verb"eqnarray" environment equations will naturally be aligned left without the use of any ampersands for alignment, see equations (\ref{eq1}) and (\ref{eq2})
\begin{eqnarray}
\alpha + \beta =\gamma^2 \label{eq1}\\
\alpha^2 + 2\gamma + \cos\theta = \delta \label{eq2} 
\end{eqnarray}
This is the normal equation style for our journals.

Where some secondary alignment is needed, for instance a second part of an equation on a second line, a single ampersand is added at the point of alignment in each line  (see  (\ref{eq3}) and (\ref{eq4})).
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} \label{eq3}\\
 & = \delta\theta PQ \cos\gamma. \label{eq4} 
\end{eqnarray}

Two points of alignment are possible using two ampersands for alignment (see  (\ref{eq5}) and (\ref{eq6})).  Note in this case extra space \verb"\qquad" is added before the second ampersand in the longest line (the top one) to separate the condition from the equation. 
\begin{eqnarray}
\alpha &=2\gamma^2 + \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta}\qquad& \theta > 1 \label{eq5}\\
 & = \delta\theta PQ \cos\gamma &\theta \leq 1\label{eq6} 
\end{eqnarray}

For a long equation which has to be split over more than one line the first line should start at the left margin, this is achieved by inserting \verb"\fl" (full left) at the start of the line, the use of the alignment parameter \verb"&" is not necessary unless some secondary alignment is needed.
\begin{eqnarray}
\fl \alpha + 2\gamma^2 = \cos\theta + \frac{XY \sin\theta}{X+ Y\cos\theta} +  \frac{XY \sin\theta}{X- Y\cos\theta} +
+ \left(\frac{XY \sin\theta}{X+ Y\cos\theta}\right)^2 \nonumber\\
+  \left(\frac{XY \sin\theta}{X- Y\cos\theta}\right)^2\label{eq7} 
\end{eqnarray}

The Plain \TeX\ command \verb"\eqalign" can be used within an \verb"equation" environment to obtain a multiline equation with a single centred number, for example
\begin{equation}
\eqalign{\alpha + \beta =\gamma^2 \cr
\alpha^2 + 2\gamma + \cos\theta = \delta.}\label{eq2} 
\end{equation}

\subsection{Miscellaneous points}
Exponential expressions, especially those containing subscripts or 
superscripts, are clearer if the notation $\exp(\ldots)$ is used except for 
simple examples. For instance $\exp[\rmi(kx-\omega t)]$ and $\exp(z^2)$ are 
preferred to $\e^{\rmi(kx-\omega t)}$ and $\e^{z^2}$, but 
$\e^x$ 
is acceptable. Similarly the square root sign $\sqrt{\phantom{b}}$ should 
only be used with relatively
simple expressions, e.g.\ $\sqrt2$ and $\sqrt{a^2+b^2}$;
in other cases the 
power $1/2$ should be used; for example, $[(x^2+y^2)/xy(x-y)]^{1/2}$.

It is important to distinguish between $\ln = \log_\e$ and $\lg 
=\log_{10}$. Braces, brackets and parentheses should be used in the 
following order: $\{[(\;)]\}$. The same ordering of brackets should be 
used within each size. However, this ordering can be ignored if the
brackets have a 
special meaning (e.g.\ if they denote an average or a function).  Decimal 
fractions should 
always be preceded by a zero: for example 0.123 {\bf not} .123. For long 
numbers commas are not inserted but instead a thin space is added after 
every third character away from the position of the decimal point unless 
this leaves a single separated character: e.g.\ $60\,000$, $0.123\,456\,78$ 
but 4321 and 0.7325.

Equations should be followed by a full stop (periods) when at the end
of a sentence.

\subsection{Equation numbering}
\label{eqnum}
\LaTeX\ provides facilities for automatically numbering equations 
and these should be used where possible. Sequential numbering (1), (2), 
etc, is the default numbering system although, if the command
\verb"\eqnobysec" is included in the preamble, equation numbering
by section is obtained, e.g.\ 
(2.1), (2.2), etc. In articles with several appendices equation numbering
by section is useful in the appendices even when sequential numbering has
been used throughout the main body of the text and is switched on by the 
\verb"\appendix" command. Equation numbering by section {\it must} 
be used for {\it Reports on Progress in 
Physics}. When referring to an equation in the text it is not normally necessary to include the word equation before the number, 
which should be parentheses. Do not 
use abbreviations such as eqn or eq.
When cross-referencing is used, \verb"(\ref{<label>})"
 will produce `(\verb"<eqnum>")',
\verb"\eref{<label>}" produces `(\verb"<eqnum>")' and
\verb"\Eref{<label>}" produces `Equation (\verb"<eqnum>")', 
where \verb"<label>" is the label to produce equation number \verb"<eqnum>".

Sometimes it is useful to number equations as parts of the same
basic equation. This can be accomplished by inserting the 
commands \verb"\numparts" before the equations concerned and 
\verb"\endnumparts" when reverting to the normal sequential numbering.
The equations below show the previous equations numbered as separate parts
using \verb"\numparts ... \endnumparts" and the \verb"eqnarray"
environment
\numparts
\begin{eqnarray}
T_{11}&=(1+P_\e)I_{\uparrow\uparrow}-(1-P_\e)
I_{\uparrow\downarrow}\\
T_{-1-1}&=(1+P_\e)I_{\downarrow\downarrow}-(1-P_\e)I_{\uparrow\downarrow}\\
S_{11}&=(3+P_\e)I_{\downarrow\uparrow}-(3-P_e)I_{\uparrow\uparrow}\\
S_{-1-1}&=(3+P_\e)I_{\uparrow\downarrow}-(3-P_\e)
I_{\downarrow\downarrow}
\end{eqnarray}
\endnumparts

\subsection{Miscellaneous extra commands for displayed equations}
The \verb"\cases" command of Plain \TeX\ is available
for use with \LaTeX\ but has been amended slightly to 
increase the space between the equation and the condition. 
\Eref{cases} 
demonstrates simply the output from the \verb"\cases" command
\begin{equation}
\label{cases}
X=\cases{1&for $x \ge 0$\\
-1&for $x<0$\\}
\end{equation}
%The code used was:
%\small\begin{verbatim}
%\begin{equation}
%\label{cases}
%X=\cases{1&for $x \ge 0$\\
%-1&for $x<0$\\}
%\end{equation}
%\end{verbatim}
%\normalsize

To obtain text style fractions within displayed maths the command 
\verb"\case{#1}{#2}" can be used (see equations (2) and (5)) instead
of the usual \verb"\frac{#1}{#2}" command or \verb"{#1 \over #2}".

When two or more short equations are on the same line they should be 
separated by a `qquad space' (\verb"\qquad"), rather than
\verb"\quad" or any combination of \verb"\,", \verb"\>", \verb"\;" 
and \verb"\ ".

\section{Referencing\label{except}}
Two different styles of referencing are in common use: 
the Harvard alphabetical system and the Vancouver numerical system. 
All IOP journals allow the use of the Harvard or Vancouver system 
(authors should use the style they are comfortable with), 
apart from the following journals

\begin{itemize}
\item {\it Physics in Medicine and Biology} 
\item {\it Physiological Measurement}
\end{itemize}

\noindent for which authors must use the Harvard referencing style.

\subsection{Alphabetical (Harvard) system}
In the Harvard system the name of the author appears in the text together 
with the year of publication. As appropriate, either the date or the name 
and date are included within parentheses. Where there are only two authors 
both names should be given in the text; if there are more than two 
authors only the first name should appear followed by `{\it et al}' 
(which can be obtained by typing \verb"\etal"). When two or 
more references to work by one author or group of authors occur for the 
same year they should be identified by including a, b, etc after the date 
(e.g.\ 1986a). If several references to different pages of the same article 
occur the appropriate page number may be given in the text, e.g.\ Kitchen 
(1982, p 39).

The reference list at the end of an article consists of an 
unnumbered `References' section containing an
alphabetical listing by authors' names and in date order for each 
author or group of identical authors. The reference list in the 
preprint style is started by including the command \verb"\section*{References}" and then
\verb"\begin{harvard}".
Individual references start with \verb"\item[]" and the reference list is completed with \verb"\end{harvard}".
There is also a shortened form of the coding; \verb"\section*{References}"
and \verb"\begin{harvard}" can be replaced by the single command
\verb"\References" and \verb"\end{harvard}" can be shortened to
\verb"\endrefs".

\subsection{Numerical (Vancouver) system}
In the numerical system references are numbered sequentially 
throughout the text. The numbers occur within square brackets and one 
number can be used to designate several references. A numerical 
reference list in the \verb"iopart" style is started by including the 
command \verb"\section*{References}" and then
\verb"\begin{thebibliography}{<num>}", where \verb"<num>" is the largest
number in the reference list (or any other number with the same number
of digits).  The 
reference list gives the references in 
numerical, not alphabetical, order and is completed by
\verb"\end{thebibliography}". Short forms of the commands are again
available: \verb"\Bibliography{<num>}" can be used at the start of the
references section and \verb"\endbib" at the end.

A variant of this system is to use labels instead of numbers within 
square brackets, this method is allowed although not recommended.

\subsection{BiBTeX\label{bibtex}}
Note that for the journals lised at the start of section {\ref{except}} authors should use the
appropriate \verb".bst" file to produce references in the correct style---see section \ref{bibby} for
a list of reccommended \verb".bst" files.

\subsection{References, general}
A complete reference should provide the reader with enough information to 
locate the article concerned. Titles of journal articles can also be included and are required for {\it Inverse problems, Journal of Neural Engineering, Measurement Science and Technology, Physical Biology, Physics in Medicine and Biology\/} and {\it Physiological Measurement}. Final page numbers of references are required for {\it Reports on Progress in Physics\/} and {\it Physiological Measurement}.

Up to ten authors may be given in a particular reference; where 
there are more than ten only the first should be given followed by 
`{\it et al}'. If an author is unsure of an abbreviation and the 
journal is not given in appendix B, it is best to leave the title in full.

The terms {\it loc.\ cit.}\ and {\it ibid}.\ should not be used. 
Unpublished conferences and reports should generally not be included 
in the reference list if a published version of the work exists. Articles in the course of publication should 
include the article title and the journal of publication, if known. 
A reference to a thesis submitted for a higher degree may be included 
if it has not been superseded by a published 
paper but the institution where the work was done should be included.

The basic structure of a reference in the reference list is the same in both the alphabetical and numerical systems, the only difference the code at the start of the reference. Alphabetic references are preceded by \verb"\item[]", numeric by \verb"\bibitem{label}" or just \verb"\item" to generate a number or \verb"\nonum" where a reference is not the first in a group of references under the same number.

Cross referencing between the text and the 
reference list is most useful in the numeric system but is not necessary for alphabetic referencing 
in the Harvard system as adding or deleting a reference 
does not normally change any of the other references. 
Care should, however, be taken to make sure that the text citations match the reference list and vice versa.

Note that footnotes to the text should not be 
included in reference list, but should appear at the bottom of the relevant page by using the \verb"\footnote" command.

\subsection{References to journal articles}
A normal reference to a journal article contains three changes of 
font:
the authors and date appear in Roman type, the journal title in 
italic, the volume number in bold and the page numbers in Roman again. 
A typical journal entry would be:

\smallskip
\begin{harvard}
\item[] Cisneros A 1971 {\it Astrophys.\ Space Sci.} {\bf 10} 87
\end{harvard}
\smallskip

\noindent which would be obtained by typing, within the references
environment 
\small\begin{verbatim}
\item[] Cisneros A 1971 {\it Astrophys. Space Sci.} {\bf 10} 87
\end{verbatim}\normalsize

Features to note are the following.

\begin{enumerate}
\item The authors should be in the form surname (with only the first 
letter capitalized) {\bf followed} by the initials with {\bf no} 
periods after the initials. Authors should be separated by a comma 
except for the last two which should be separated by `and' with no 
comma preceding it. For journals that accept titles of articles in the 
reference list,  the title should be in Roman (upright)
lower case letters, except for an initial 
capital, and should follow the date.

\item The year of publication follows the authors and is not in parentheses. An article title (in Roman) would follow the year.

\item The journal is in italic and is abbreviated. \ref{jlabs} 
gives a list of macros that will give the correct abbreviation for 
many of the common journals. If a journal has several parts denoted by 
different letters the part letter should be inserted after the journal in Roman type, e.g.\ 
{\it Phys.\ Rev.\ \rm A}.

\item The volume number is bold; the page number is Roman.
 Both the initial and final page numbers should be given where possible. The final page number should be in 
the shortest possible form and separated from the initial page number by an en rule (\verb"--"), e.g.\ 1203--14.

\item Where there are two or more references with identical authors, 
the authors' names should be repeated for the second and subsequent references. Each individual publication should be presented as a separate reference, although in the numerical system one number can be used for several references. This facilitates linking in the electronic version of the journal. 
\end{enumerate}
\subsection{Electronic journal references}
These often do not always follow the conventional {\it year-journal-volume-page numbers} pattern. Some examples are:

\numrefs{1}
\item Carlip S and Vera R 1998 {\it Phys. Rev.} D {\bf 58} 011345 
\item Davies K and Brown G 1997 {\it J. High Energy Phys.} JHEP12(1997)002
\item Hannestad S 2005 {\it J. Cosmol. Astropart. Phys.} JCAP02(2005)011
\item Hilhorst H J 2005 {\it J. Stat. Mech.} L02003
\item Gundlach C 1999 {\it Liv. Rev. Rel.} 1994-4
\endnumrefs

\subsection{Preprint references}
Preprint may be referenced but if the article concerned has been published in a peer-reviewed journal, that reference should take precedence. If only a preprint reference can be given, it is helpful to include the article title. Examples are:
\vskip6pt
\numrefs{1}
\item Neilson D and Choptuik M 2000 {\it Class. Quantum Grav.} {\bf 17} 761 ({\it Preprint} gr-qc/9812053)
\item Harrison M 1999 Dipheomorphism-invariant manifolds {\it Preprint} hep-th/9909196
\endnumrefs

\subsection{References to books, conference proceedings and reports}

References to books, proceedings and reports are similar, but have only two
changes of font. The authors and date of publication are in Roman, the 
title of the book is in italic, and the editors, publisher, 
town of publication 
and page number are in Roman. A typical reference to a book and a
conference paper might be

\smallskip
\begin{harvard}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc.\
Int.\ Conf.\ on Nuclear Physics (Munich)} vol~1 (Amsterdam:  
North-Holland/American Elsevier) p~517
\end{harvard}
\smallskip

\noindent which would be obtained by typing
\small\begin{verbatim}
\item[] Dorman L I 1975 {\it Variations of Galactic Cosmic Rays} 
(Moscow: Moscow State University Press) p~103
\item[] Caplar R and Kulisic P 1973 {\it Proc. Int. Conf. on Nuclear 
Physics (Munich)} vol~1 (Amsterdam: North-Holland/American 
Elsevier) p~517
\end{verbatim}\normalsize
\noindent respectively.

Features to note are the following.
\begin{enumerate}
\item Book titles are in italic and should be spelt out in full with 
initial capital letters for all except minor words. Words such as 
Proceedings, Symposium, International, Conference, Second, etc should 
be abbreviated to Proc., Symp., Int., Conf., 2nd, 
respectively, but the rest of the title should be given in full, 
followed by the date of the conference and the 
town or city where the conference was held. For 
Laboratory Reports the Laboratory should be spelt out wherever 
possible, e.g.\ {\it Argonne National Laboratory Report}.

\item The volume number as, for example, vol~2, should be followed by 
the editors, if any, in a form such as ed~A~J~Smith and P~R~Jones. Use 
\etal if there are more than two editors. Next comes the town of 
publication and publisher, within brackets and separated by a colon, 
and finally the page numbers preceded by p if only one number is given 
or pp if both the initial and final numbers are given.

\item If a book is part of a series (for examples, {\it Springer Tracts in Modern Physics\/}), the series title and volume number is given in parentheses after the book title. Whereas for an individual volume in a multivolume set, the set title is given first, then the volume title. 
\end{enumerate}
\smallskip
\begin{harvard}
\item[]Morse M 1996 Supersonic beam sources {\it Atomic Molecular and Optical Physics\/} ({\it Experimental Methods in the Physical Sciences\/} vol 29) ed F B Dunning and R Hulet (San Diego: Academic)  
\item[]Fulco C E, Liverman C T and Sox H C (eds) 2000 {\it Gulf War and Health\/} vol 1 {\it Depleted Uranium, Pyridostigmine Bromide, Sarin, and Vaccines\/} (Washington, DC: The National Academies Press)
\end{harvard}

\section{Cross referencing\label{xrefs}}
The facility to cross reference items in the text is very useful when 
composing articles as the precise form of the article may be uncertain at the start 
and  revisions and amendments may subsequently be made. 
\LaTeX\ provides excellent facilities for doing cross-referencing
and these can be very useful in preparing articles.

\subsection{References}
\label{refs}
Cross referencing is useful for numeric reference lists because, if it 
is used, adding 
another reference to the list does not then involve renumbering all 
subsequent references. It is not necessary for referencing 
in the Harvard system where the final reference list is alphabetical 
and normally no other changes are necessary when a reference is added or
deleted.
Two passes are necessary initially to get the cross references right 
but once they are correct a single run is usually sufficient provided an 
\verb".aux" file is available and the file 
is run to the end each time.
\verb"\cite" and \verb"\bibitem" are used to link citations in the text
with their entry in the reference list;
if the 
reference list contains an entry \verb"\bibitem{label}", 
then \verb"\bibitem{label}" 
will produce the correct number in the reference list and 
\verb"\cite{label}" will produce the number within square brackets in the 
text. \verb"label" may contain alphabetic letters, 
or punctuation characters but must not contain spaces or commas. It is also
recommended that the underscore character \_{} is not used in cross
referencing. 
Thus labels for the form 
\verb"eq:partial", \verb"fig:run1", \verb"eq:dy'", 
etc, may be used. When several 
references occur together in the text \verb"\cite" may be used with 
multiple labels with commas but no spaces separating them; 
the output will be the 
numbers within a single pair of square brackets with a comma and a 
thin space separating the numbers. Thus \verb"\cite{label1,label2,label4}"
would give [1,\,2,\,4]. Note that no attempt is made by the style file to sort the 
labels and no shortening of groups of consecutive numbers is done.
Authors should therefore try to use multiple labels in the correct 
order.

The numbers for the cross referencing are generated in the order the 
references appear in the reference list, so that if the entries in the 
list are not in the order in which the references appear in the text 
then the 
numbering within the text will not be sequential. To correct this 
change the ordering of the entries in the reference list and then 
rerun {\it twice}.

\subsection{Equation numbers, sections, subsections, figures and 
tables}
Labels for equation numbers, sections, subsections, figures and tables 
are all defined with the \verb"\label{label}" command and cross references 
to them are made with the \verb"\ref{label}" command.

Any section, subsection, subsubsection, appendix or subappendix 
command defines a section type label, e.g. 1, 2.2, A2, A1.2 depending 
on context. A typical article might have in the code of its introduction 
`The results are discussed in section\verb"~\ref{disc}".' and
the heading for the discussion section would be:
\small\begin{verbatim}
\section{Results}\label{disc}
\end{verbatim}\normalsize
Labels to sections, etc, may occur anywhere within that section except
within another numbered environment. 
Within a maths environment labels can be used to tag equations which are 
referred to within the text.

In addition to the standard \verb"\ref{<label>}" the abbreviated
forms given in the \tref{abrefs}
are available for reference to standard parts of the text

\Table{\label{abrefs}Alternatives to the normal references $\backslash${\tt ref} 
and the text generated by
them. Note it is not normally necessary to include the word equation
before an equation number except where the number starts a sentence. The
versions producing an initial capital should only be used at the start of
sentences.} 
\br
Reference&Text produced\\
\mr
\verb"\eref{<label>}"&(\verb"<num>")\\
\verb"\Eref{<label>}"&Equation (\verb"<num>")\\
\verb"\fref{<label>}"&figure \verb"<num>"\\
\verb"\Fref{<label>}"&Figure \verb"<num>"\\
\verb"\sref{<label>}"&section \verb"<num>"\\
\verb"\Sref{<label>}"&Section \verb"<num>"\\
\verb"\tref{<label>}"&table \verb"<num>"\\
\verb"\Tref{<label>}"&Table \verb"<num>"\\
\br
\endTable

\section{Tables and table captions}
Tables are numbered serially and referred to in the text 
by number (table 1, etc, {\bf not} tab. 1). Each table should have an 
explanatory caption which should be as concise as possible. If a table 
is divided into parts these should be labelled \pt(a), \pt(b), 
\pt(c), etc but there should be only one caption for the whole 
table, not separate ones for each part.

In the preprint style the tables may be included in the text 
or listed separately after the reference list 
starting on a new page.

\subsection{The basic table format}
The standard form for a table is:
\small\begin{verbatim}
\begin{table}
\caption{\label{label}Table caption.}
\begin{indented}
\item[]\begin{tabular}{@{}llll}
\br
Head 1&Head 2&Head 3&Head 4\\
\mr
1.1&1.2&1.3&1.4\\
2.1&2.2&2.3&2.4\\
\br
\end{tabular}
\end{indented}
\end{table}
\end{verbatim}\normalsize

Points to note are:
\begin{enumerate}
\item The caption comes before the table. It should have a period at
the end.

\item Tables are normally set in a smaller type than the text.
The normal style is for tables to be indented. This is accomplished
by using \verb"\begin{indented}" \dots\ \verb"\end{indented}"
and putting \verb"\item[]" before the start of the tabular environment.
Omit these
commands for any tables which will not fit on the page when indented.

\item The default is for columns to be aligned left and 
adding \verb"@{}" omits the extra space before the first column.

\item Tables have only horizontal rules and no vertical ones. The rules at
the top and bottom are thicker than internal rules and are set with
\verb"\br" (bold rule). 
The rule separating the headings from the entries is set with
\verb"\mr" (medium rule).

\item Numbers in columns should be aligned on the decimal point;
to help do this a control sequence \verb"\lineup" has been defined 
which sets \verb"\0" equal to a space the size of a digit, \verb"\m"
to be a space the width of a minus sign, and \verb"\-" to be a left
overlapping minus sign. \verb"\-" is for use in text mode while the other
two commands may be used in maths or text.
(\verb"\lineup" should only be used within a table
environment after the caption so that \verb"\-" has its normal meaning
elsewhere.) See table~\ref{tabone} for an example of a table where
\verb"\lineup" has been used.
\end{enumerate}

\begin{table}
\caption{\label{tabone}A simple example produced using the standard table commands 
and $\backslash${\tt lineup} to assist in aligning columns on the 
decimal point. The width of the 
table and rules is set automatically by the 
preamble.}

\begin{indented}
\lineup
\item[]\begin{tabular}{@{}*{7}{l}}
\br                              
$\0\0A$&$B$&$C$&\m$D$&\m$E$&$F$&$\0G$\cr 
\mr
\0\023.5&60  &0.53&$-20.2$&$-0.22$ &\01.7&\014.5\cr
\0\039.7&\-60&0.74&$-51.9$&$-0.208$&47.2 &146\cr 
\0123.7 &\00 &0.75&$-57.2$&\m---   &---  &---\cr 
3241.56 &60  &0.60&$-48.1$&$-0.29$ &41   &\015\cr 
\br
\end{tabular}
\end{indented}
\end{table}

\subsection{Simplified coding and extra features for tables}
The basic coding format can be simplified using extra commands provided in
the \verb"iopart" class file. The commands up to and including 
the start of the tabular environment
can be replaced by
\small\begin{verbatim}
\Table{\label{label}Table caption}
\end{verbatim}\normalsize
this also activates the definitions within \verb"\lineup".
The final three lines can also be reduced to \verb"\endTable" or
\verb"\endtab". Similarly for a table which does not fit in when indented
\verb"\fulltable{\label{label}caption}" \dots\ \verb"\endfulltable" or \verb"\endtab"
can be used. \LaTeX\ optional positional parameters can, if desired, be added after 
\verb"\Table{\label{label}caption}" and \verb"\fulltable{\label{label}caption}".

\verb"\centre{#1}{#2}" can be used to centre a heading 
\verb"#2" over \verb"#1" 
columns and \verb"\crule{#1}" puts a rule across 
\verb"#1" columns. A negative 
space \verb"\ns" is usually useful to reduce the space between a centred 
heading and a centred rule. \verb"\ns" should occur immediately after the 
\verb"\\" of the row containing the centred heading (see code for
\tref{tabl3}). A small space can be 
inserted between rows of the table 
with \verb"\ms" and a half line space with \verb"\bs" 
(both must follow a \verb"\\" but should not have a 
\verb"\\" following them).

\Table{\label{tabl3}A table with headings spanning two columns and containing notes. 
To improve the 
visual effect a negative skip ($\backslash${\tt ns})
has been put in between the lines of the 
headings. Commands set-up by $\backslash${\tt lineup} are used to aid 
alignment in columns. $\backslash${\tt lineup} is defined within
the $\backslash${\tt Table} definition.}
\br
&&&\centre{2}{Separation energies}\\
\ns
&Thickness&&\crule{2}\\
Nucleus&(mg\,cm$^{-2}$)&Composition&$\gamma$, n (MeV)&$\gamma$, 2n (MeV)\\
\mr
$^{181}$Ta&$19.3\0\pm 0.1^{\rm a}$&Natural&7.6&14.2\\
$^{208}$Pb&$\03.8\0\pm 0.8^{\rm b}$&99\%\ enriched&7.4&14.1\\
$^{209}$Bi&$\02.86\pm 0.01^{\rm b}$&Natural&7.5&14.4\\
\br
\end{tabular}
\item[] $^{\rm a}$ Self-supporting.
\item[] $^{\rm b}$ Deposited over Al backing.
\end{indented}
\end{table}

Units should not normally be given within the body of a table but 
given in brackets following the column heading; however, they can be 
included in the caption for long column headings or complicated units. 
Where possible tables should not be broken over pages. 
If a table has related notes these should appear directly below the table
rather than at the bottom of the page. Notes can be designated with
footnote symbols (preferable when there are only a few notes) or
superscripted small roman letters. The notes are set to the same width as
the table and in normal tables follow after \verb"\end{tabular}", each
note preceded by \verb"\item[]". For a full width table \verb"\noindent"
should precede the note rather than \verb"\item[]". To simplify the coding 
\verb"\tabnotes" can, if desired, replace \verb"\end{tabular}" and 
\verb"\endtabnotes" replaces
\verb"\end{indented}\end{table}".

If all the tables are grouped at the end of a document
the command \verb"\Tables" is used to start a new page and 
set a heading `Tables and table captions'. If the tables follow an appendix then add the command \verb"\noappendix" to revert to normal style numbering.

\section{Figures and figure captions}

Figures (with their captions) can be incorporated into the text at the appropriate position or grouped together
at the end of the article. If the figures are at the end of the article and follow an appendix then add the command \verb"\noappendix" to revert to normal style numbering.

\subsection{Using copyright material}
If you wish to illustrate your paper using material for which you 
do not own the copyright then you must seek permission from 
the copyright holder, usually both the author and the publisher of the previous work. It is the author's responsibility to obtain 
copyright permissions and this should be done
prior to submitting your paper. 
If you have obtained permission, please provide full details of 
the permission granted---for example, copies of the text of any e-mails 
or a copy of any letters you may have received. Figure captions must include an acknowledgment of the original source of the material even when permission to use has been obtained.

\subsection{Inclusion of graphics files\label{figinc}}
Using the \verb"graphicx" package graphics files can 
be included within figure and center environments at an 
appropriate point within the text using code such as:
\small\begin{verbatim}
\includegraphics{file.eps}
\end{verbatim}\normalsize
The \verb"graphicx" package supports various optional arguments
to control the appearance of the figure. Other similar 
packages can also be used (e.g. \verb"graphics", \verb"epsf"). 
For more detail about graphics inclusion see the documentation 
of the \verb"graphicx" package, refer to one of the books on \LaTeX\ {\cite{book1}}
or download some of the excellent free documentation available via the Comprehensive
TeX Archive Network (CTAN) http://www.ctan.org---{in particular see \cite{eps}}.
IOP's graphics guidelines, \gguide, provide further information on preparing EPS files---a copy
should have accompanied this document but it may be download from authors.iop.org.

\subsection{Captions}
Each figure should have a brief caption describing it and, if 
necessary, interpreting the various lines and symbols on the figure. 
As much lettering as possible should be removed from the figure itself 
and included in the caption. If a figure has parts, these should be 
labelled ($a$), ($b$), ($c$), etc and all parts should be described 
within a single caption. \Tref{blobs} gives the definitions for describing 
symbols and lines often used within figure captions (more symbols are 
available when using the optional packages loading the AMS extension fonts).

\subsection{Supplementary Data}
All of our journals encourage authors to submit supplementary data attachments to 
enhance the online versions of published research articles. Supplementary data 
enhancements typically consist of video clips, animations or
data files, tables of extra information or extra figures. They can 
add to the reader's understanding and present results in attractive ways that go 
beyond what can be presented in the print version of the journal. 
The printed journal remains the archival version, and supplementary data items are 
supplements which enhance a reader's understanding of the paper but are not 
essential to that understanding. For electronic-only journals, supplementary data attachments 
may be used to convey essential information. Guidelines on supplementary data file formats 
are contained in the document {\it Guidelines for authors} which can be accessed 
online by going to authors.iop.org.

\begin{table}[t]
\caption{\label{blobs}Control sequences to describe lines and symbols in figure 
captions.}
\begin{indented}
\item[]\begin{tabular}{@{}lllll}
\br
Control sequence&Output&&Control sequence&Output\\
\mr
\verb"\dotted"&\dotted        &&\verb"\opencircle"&\opencircle\\
\verb"\dashed"&\dashed        &&\verb"\opentriangle"&\opentriangle\\
\verb"\broken"&\broken&&\verb"\opentriangledown"&\opentriangledown\\
\verb"\longbroken"&\longbroken&&\verb"\fullsquare"&\fullsquare\\
\verb"\chain"&\chain          &&\verb"\opensquare"&\opensquare\\
\verb"\dashddot"&\dashddot    &&\verb"\fullcircle"&\fullcircle\\
\verb"\full"&\full            &&\verb"\opendiamond"&\opendiamond\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage

\appendix
\section{List of macros for formatting text, figures and tables}

\begin{table}[hb]
\caption{Macros available for use in text. Parameters in square brackets are optional.}
\footnotesize\rm
\begin{tabular}{@{}*{7}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\title[#1]{#2}"&Title of article and short title (optional)\\
\verb"\paper[#1]{#2}"&Title of paper and short title (optional)\\
\verb"\letter{#1}"&Title of Letter to the Editor\\
\verb"\comment[#1]{#2}"&Title of Comment and short title (optional)\\
\verb"\topical[#1]{#2}"&Title of Topical Review and short title 
(optional)\\
\verb"\review[#1]{#2}"&Title of review article and short title (optional)\\
\verb"\note[#1]{#2}"&Title of Note and short title (optional)\\
\verb"\prelim[#1]{#2}"&Title of Preliminary Communication \& short title\\
\verb"\author{#1}"&List of all authors\\
\verb"\article[#1]{#2}{#3}"&Type and title of other articles and 
short title (optional)\\
\verb"\address{#1}"&Address of author\\
\verb"\pacs{#1}"&PACS classification codes\\
\verb"\pacno{#1}"&Single PACS classification code\\
\verb"\ams{#1}"&Mathematics Classification Scheme\\
\verb"\submitto{#1}"&`Submitted to' message\\
\verb"\maketitle"&Creates title page\\
\verb"\begin{abstract}"&Start of abstract\\
\verb"\end{abstract}"&End of abstract\\
\verb"\nosections"&Inserts space before text when no sections\\
\verb"\section{#1}"&Section heading\\
\verb"\subsection{#1}"&Subsection heading\\
\verb"\subsubsection{#1}"&Subsubsection heading\\
\verb"\appendix"&Start of appendixes\\
\verb"\ack"&Acknowledgments heading\\
\verb"\References"&Heading for reference list\\
\verb"\begin{harvard}"&Start of alphabetic reference list\\
\verb"\end{harvard}"&End of alphabetic reference list\\
\verb"\begin{thebibliography}{#1}"&Start of numeric reference list\\
\verb"\end{thebibliography}"&End of numeric reference list\\
\verb"\etal"&\etal for text and reference lists\\
\verb"\nonum"&Unnumbered entry in numerical reference list\\
\br
\end{tabular}
\end{table}

\begin{table}
\caption{Macros defined within {\tt iopart.cls}
for use with figures and tables.}
\begin{indented}
\item[]\begin{tabular}{@{}l*{15}{l}}
\br
Macro name&Purpose\\
\mr
\verb"\Figures"&Heading for list of figure captions\\
\verb"\Figure{#1}"&Figure caption\\
\verb"\Tables"&Heading for tables and table captions\\
\verb"\Table{#1}"&Table caption\\
\verb"\fulltable{#1}"&Table caption for full width table\\
\verb"\endTable"&End of table created with \verb"\Table"\\
\verb"\endfulltab"&End of table created with \verb"\fulltable"\\
\verb"\endtab"&End of table\\
\verb"\br"&Bold rule for tables\\
\verb"\mr"&Medium rule for tables\\
\verb"\ns"&Small negative space for use in table\\
\verb"\centre{#1}{#2}"&Centre heading over columns\\
\verb"\crule{#1}"&Centre rule over columns\\
\verb"\lineup"&Set macros for alignment in columns\\
\verb"\m"&Space equal to width of minus sign\\
\verb"\-"&Left overhanging minus sign\\
\verb"\0"&Space equal to width of a digit\\
\br
\end{tabular}
\end{indented}
\end{table}

\clearpage
\section{Control sequences for journal
abbreviations}
\label{jlabs}
\begin{table}[hb]
\caption{\label{jlab1}Abbreviations for the IOP journals.}
\begin{indented}
\item[]
\begin{tabular}{@{}lll}
\br
Macro name&{\rm Short form of journal title}&Years relevant\\
\mr
{\it Current journals}\\
\mr
\verb"\CQG"&Class. Quantum Grav.\\
\verb"\EJP"&Eur. J. Phys.\\
\verb"\IP"&Inverse Problems\\
\verb"\JHEP"&J. High Energy Phys.&1999 and onwards\\
\verb"\JMM"&J. of Michromech. and Microeng.\\
\verb"\JNE"&J. Neural Eng.\\
\verb"\JOA"&J. Opt. A: Pure and Applied Opt.&1998 and onwards\\
\verb"\JOB"&J. Opt. B: Quantum and Semiclass. Opt.&1999 and onwards\\
\verb"\JPA"&J. Phys. A: Math. Gen.\\
\verb"\jpb"&J. Phys. B: At. Mol. Opt. Phys.&1988 and onwards\\
\verb"\JPCM"&J. Phys: Condens. Matter&1989 and onwards\\
\verb"\JPD"&J. Phys. D: Appl. Phys.\\
\verb"\jpg"&J. Phys. G: Nucl. Part. Phys.&1989 and onwards\\
\verb"\MSMSE"&Modelling Simul. Mater. Sci. Eng.\\
\verb"\MST"&Meas. Sci. Technol.&1990 and onwards\\
\verb"\NJP"&New J. Phys.&1999 and onwards\\
\verb"\NL"&Nonlinearity\\
\verb"\NT"&Nanotechnology\\
\verb"\PB"&Phys. Biol.\\
\verb"\PM"&Physiol. Meas.\\
\verb"\PMB"&Phys. Med. Biol.\\
\verb"\PPCF"&Plasma Physics and Controlled Fusion\\
\verb"\PSST"&Plasma Sources Sci. Technol.\\
\verb"\RPP"&Rep. Prog. Phys.\\
\verb"\SST"&Semicond. Sci. Technol.\\
\verb"\SMS"&Smart Mater. Struct.\\
\verb"\SUST"&Supercond. Sci. Technol.\\
\mr
{\it No Longer Published}\\
\mr
\verb"\JPB"&J. Phys. B: At. Mol. Phys.&1968--1987\\
\verb"\JPC"&J. Phys. C: Solid State Phys.&1968--1988\\
\verb"\JPE"&J. Phys. E: Sci. Instrum.&1968--1989\\
\verb"\JPF"&J. Phys. F: Met. Phys.&1971--1988\\
\verb"\JPG"&J. Phys. G: Nucl. Phys.&1975--1988\\
\verb"\PAO"&Pure and Applied Opt.&1992--1998\\
\verb"\QO"&Quantum Opt.&1989--1994\\
\verb"\QSO"&Quantum and Semiclass. Opt.&1995--1998\\
\br
\end{tabular}
\end{indented}
\end{table}

\begin{table}

\caption{\label{jlab2}Abbreviations for some more 
common non-IOP journals.}
\begin{indented}
\item[]\begin{tabular}{@{}ll}
\br
Macro name&{\rm Short form of journal}\\
\mr
\verb"\AC"&Acta Crystallogr.\\
\verb"\AM"&Acta Metall.\\
\verb"\AP"&Ann. Phys., Lpz\\
\verb"\APNY"&Ann. Phys., NY\\
\verb"\APP"&Ann. Phys., Paris\\
\verb"\CJP"&Can. J. Phys.\\
\verb"\GRG"&Gen. Rel. Grav.\\
\verb"\JAP"&J. Appl. Phys.\\
\verb"\JCP"&J. Chem. Phys.\\
\verb"\JJAP"&Japan. J. Appl. Phys.\\
\verb"\JMMM"&J. Magn. Magn. Mater.\\
\verb"\JMP"&J. Math. Phys.\\
\verb"\JOSA"&J. Opt. Soc. Am.\\
\verb"\JP"&J. Physique\\
\verb"\JPhCh"&J. Phys. Chem.\\
\verb"\JPSJ"&J. Phys. Soc. Japan\\
\verb"\JQSRT"&J. Quant. Spectrosc. Radiat. Transfer\\
\verb"\NC"&Nuovo Cimento\\
\verb"\NIM"&Nucl. Instrum. Methods\\
\verb"\NP"&Nucl. Phys.\\
\verb"\PF"&Phys. Fluids\\
\verb"\PL"&Phys. Lett.\\
\verb"\PR"&Phys. Rev.\\
\verb"\PRL"&Phys. Rev. Lett.\\
\verb"\PRS"&Proc. R. Soc.\\
\verb"\PS"&Phys. Scr.\\
\verb"\PSS"&Phys. Status Solidi\\
\verb"\PTRS"&Phil. Trans. R. Soc.\\
\verb"\RMP"&Rev. Mod. Phys.\\
\verb"\RSI"&Rev. Sci. Instrum.\\
\verb"\SSC"&Solid State Commun.\\
\verb"\SPJ"&Sov. Phys.--JETP\\
\verb"\ZP"&Z. Phys.\\
\br
\end{tabular}
\end{indented}
\end{table}

\section*{References}
\begin{thebibliography}{10}
\bibitem{book1} Goosens M, Rahtz S and Mittelbach F 1997 {\it The \LaTeX\ Graphics Companion\/} 
(Reading, MA: Addison-Wesley)
\bibitem{eps} Reckdahl K 1997 {\it Using Imported Graphics in \LaTeX\ } (search CTAN for the file `epslatex.pdf')
\end{thebibliography}

\end{document}

### 2011-08-30 trunk/papers/lmcproc/lmcproc.tex
Move papers directory one level down
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1454 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
\usepackage{iopams} 
\usepackage{graphicx, subfigure}

\begin{document}

\title[]{Shear Flow of Cubic Blue Phases in Confined Geometries}

\author{O. Henrich$^{1,3}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}

\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
\ead{o.henrich@ucl.ac.uk}

\begin{abstract}
\end{abstract}

%Uncomment for PACS numbers title message
%\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
%\vspace{2pc}
%\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
% Comment out if separate title page not required
\maketitle

\section{Introduction}
% Put \label in argument of \section for cross-referencing
%\section{\label{}}
\section{Model and Methods}

%Our approach is based on a well-established model for hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes the order structure of the liquid crystal by a traceless, symmetric tensor order parameter ${\bf Q}({\bf r})$. 
%In uniaxial approximation the tensor order paramter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$, whereas ${\bf n}$ represents the local orientation of the liquid crystal molecules and brackets indicate mesoscopic averaging over a large number individual molecules.
%Its largest eigenvalue $\frac{2}{3}q$ with $0<q<1$ characterizes the degree of order around the corresponding axis. 
%The equilibrium properties of the system are determined by a Landau-de Gennes free energy density $f$ and the functional ${\cal F}[Q]=\int d^3{\bf r} f({\bf Q})$.
%The free energy density consists of two different parts, namely a bulk contribution $f_b$ and a gradient term $f_g$.

%\begin{eqnarray}
%f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
%f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{eqn1}
%\end{eqnarray}

%The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ and features a the first-order transition from the isotropic to the ordered phase at $\gamma>2.7$.
%The second part comprises the elastic contributions to the free energy due to splay-, bend- and twist-deformations and assume for simplicity that they have all the same elastic constant $K$.
%The pitch length $p$ of the cholesteric liquid crystal is related to the wavervector $q_0=2\pi/p$.
%In order to specify a thermodynamic state it is useful to introduce two dimensionless quantities, the effective temperature $\tau$ and chirality $\kappa$.
%They are given by

%\begin{eqnarray}
%\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
%\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber,
%\end{eqnarray}

%whereas the chirality qqnatifies the ratio between gradient and bulk free energy.\\
%The equation of motion of the tensor order parameter reads

%\begin{equation}\label{}
%\left(\partial_t+ u_\alpha \partial_\alpha \right){\bf Q} - {\bf S}({\bf W},{\bf Q}) = \Gamma {\bf H}.
%\label{eqn2}
%\end{equation}

%The first term on the lefthand side of Eq.(\ref{eqn2}) is a material derivative, which describes the rate of change of a quantity moving along with the flow.
%The second term accounts for the rate of change due to velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$:

%\begin{eqnarray}
%{\bf S}({\bf W}, {\bf Q}) &=& (\xi {\bf A} + {\boldsymbol \Omega})({\bf Q}+\frac{\bf I}{3})\nonumber\\
%& &\hspace*{-1.5cm}+ ({\bf Q}+\frac{\bf I}{3})(\xi {\bf A}  - {\boldsymbol \Omega})-2 \xi ({\bf Q}+\frac{\bf I}{3})Tr({\bf Q W}),
%\label{eqn3}
%\end{eqnarray}

%where $Tr$ denotes the trace operation and ${\bf A}=({\bf W}+{\bf W}^T)/2$ and ${\boldsymbol \Omega}=({\bf W}-{\bf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
%The molecular details of the liquid crystal such as the aspect ratio of its mesogens are embodied in the constant $\xi$, also know as the tumbling parameter.
%The righthand side relates the rate of change to the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\bf Q$.

%\begin{equation}
%{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
%\label{eqn4}
%\end{equation}

%The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q^2/\Gamma$.\\
%The time evolution of density and velocity are goverend by the continuity equation Eq. (\ref{eqn5})

%\begin{eqnarray}
%\partial_t \rho + \partial_\alpha (\rho u_\alpha)=0
%\label{eqn5}
%\end{eqnarray}

%and the Navier-Stokes equation Eq. (\ref{eqn6})

%\begin{eqnarray}
%\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}\nonumber\\
%&+&\hspace*{-2cm}\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}. 
%\label{eqn6}
%\end{eqnarray}

%At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq. (\ref{eqn6}) can be safely neglected.
%The pressure tensor reads explicitely

%\begin{eqnarray}
%\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
%&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
%&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
%\label{eqn7}
%\end{eqnarray}

%It comprises the terms that give rise to an additional stress in the ordered state.
%In the isotropic state when ${\bf Q}\equiv 0$ Eq. (\ref{eqn7}) is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
%The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Denniston:2004, Marenduzzo:2008} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the hydrodynamic equations (\ref{eqn5}) and (\ref{eqn6}) and a finite difference scheme for the tensor order parameter equation (\ref{eqn2}).

\section{Results and Discussion}

\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_100k_x.png}
\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_100k_y.png}\\
\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_300k_x.png}
\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_300k_y.png}
\caption{}
\label{fig1}
\end{figure}

\begin{figure*}[h]
\centering
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_25k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_75k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_125k.png}
\caption{}
\label{fig2}
\end{figure*}

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{stress_bp2_fbc.pdf}
\caption{}
\label{fig3}
\end{figure*}

\begin{figure*}[h]
\centering
\begin{minipage}[b]{0.3\textwidth}
\subfigure[]{\includegraphics[width=\textwidth]{disc_bp1_t-0o5_k1o0_400k_1uc.png}}
\subfigure[]{\includegraphics[width=\textwidth]{disc_bp1_t-0o5_k1o0_400k_2uc.png}}
\end{minipage}
\subfigure[]{\includegraphics[width=0.3\textwidth]{disc_bp1_t-0o5_k1o0_400k_3uc.png}} 
\subfigure[]{\includegraphics[width=0.3\textwidth]{disc_bp1_t-0o5_k1o0_400k_4uc.png}}
\caption{}
\label{fig4}
\end{figure*}

\begin{figure*}[h]
\centering
\includegraphics[width=\textwidth]{stress_bp1_fbc.pdf}
\caption{}
\label{fig5}
\end{figure*}

\section{Conclusions}

% If you have acknowledgments, this puts in the proper section head.
\ack
Thanks folks!

% Create the reference section using BibTeX:

\section*{References}
\bibliography{lmcproc}

\end{document}

### 2011-08-30 trunk/papers/smectics/smectics.tex
Move papers directory one level down
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1454 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-08-19 papers/smectics/smectics.tex
@@
 
-\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{a}$}, K. Stratford\textit{$^{b}$} and M.E. Cates \textit{$^{a}$}}
+\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{b}$}, K. Stratford\textit{$^{c}$} and M.E. Cates \textit{$^{b}$}}
 }\vspace{0.5cm}
@@
 
-\noindent \normalsize{The abstract should be a single paragraph which summarises the content of the article.
-Any references in the abstract should be written out in full \textit{e.g.} [Surname \textit{et al., Journal Title}, 2000, \textbf{35}, 3523].}
+\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, threfore in the nonlinear regime.
+In two dimensions we find that, according to the thermodynamic parameters, shear may either align the system and stabilise an almost regular lamellar phase, or induce the nucleation and proliferation of defects in the smectic layers.
+In this last case, defects may be progressively annealed by slowly decreasing the applied shear rate, so that there are marked memory effects and the rheology is history-dependent.
+Finally, selected simulations in three dimensions show shear-induced ordering rather than defect nucleation, thus pointing out to a dramatic effect of dimensionality on the physics of the system.}
 \vspace{0.5cm}
@@
 
-\footnotetext{\textit{$^{a}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
-\footnotetext{\textit{$^{b}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
+\footnotetext{\textit{$^{a}$~University College of London, London, UK}}
+\footnotetext{\textit{$^{b}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
+\footnotetext{\textit{$^{c}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
 
@@
 
-\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P.
+%\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P.
 Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}
@@
 \section{Introduction}
-In our everyday lives we are surrounded by complex fluids such as foams, colloidal dispersions, microemulsions, polymeric fluids, liquid crystals and many biofluids.
-Their intricate microstructure gives rise to unusual rheological properties that are not found in simple Newtonian fluids and make them indispensable materials in numerous industrial applications.
-Despite their abundance in the processing, energy or chemical industry complex fluids remain difficult to model as they normally consists of two or more separate phases which interact with each other and the solvent.
-This in turn causes stresses and emerging flow properties like the viscosity to depend highly on the fow state and on the structure of the microphases. 
-In order to capture their rich phenomenology and properties ab initio without using constitutive relations that only mimick their behaviour, computationally expensive approaches are necessary.
-It is only now that we have the resources to simulate nonlinear flow of complex fluids on large scales\cite{Saksena09}.\\
-In the present work we focus on a specific class of complex fluids, whose members share the common feature to undergo an order-disorder transition from an isotropic-disordered state at high temperature to a microphase separated state at low temperature.
-They can be described with a Brazovskii free energy functional \cite{Brazovskii75} and comprise systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.
-Especially the latter have attracted considerable interest because of their intricate morphology.
-Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
-Moreover, in some regimes shear-induced transitions from the microphase separated state to ordered phases like the lamellar one occur \cite{Cates89,Koppi93,Fredrickson94}.
-Lyotropic liquid crystals are known to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93}.
-Shear-induced ordering takes also place in smectic liquid crystals \cite{Larson93,Hamley99}, making these systems an interesting object for rheological studies.\\
-The purpose of this paper is twofold.
-First we investigate the morphology and rheology of a two-dimensional lamellar system in steady shear flow.
-This system has been studied extensively \cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}, but because of computational constraints almost all results were limited to two dimensions and smaller strains.
-Here we quantify rheological properties and try to relate them to the defect pattern and the microstructure of the fluid.
-In the second part of the analysis we compare the results in two and three dimensions in order to see how the dimensionality of the system influences the defect dynamics and phase kinetics.
-\\ 
+Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurth etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid.
+Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibers constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character.
+Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in biosensing.
+
+Rheology gives a powerful way to characterise macroscopic physical properties of a soft material such as its viscosity, elastic and loss moduli, which quantify, for instance, its rigidity, response to applied forces or stresses etc.
+While in a Newtonian, viscous fluid, shearing gives rise to a stress which is linearly proportional to the applied shear rate -- the slope being the viscosity -- in complex fluids flow is often non-Newtonian and such a simple linear relation no longer holds.
+Thus, viscoelastic complex fluids may be shear thickening or shear thinning, if their effective viscosity increases or decreases with applied shear.
+They may even spontaneously phase separate when flowing, giving rise to shear bands which are familiar in the rheology of liquid crystals in the vicinity of the isotropic-nematic transition, and in wormlike micelles.
+
+The rheological properties of a non-Newtonian fluid are often linked to the microstrucure of the material.
+For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate.
+Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals.
+While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1.
+Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics.
+Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via preshearing, or not.
+In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample.
+An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state.
+This leads to increased dissipation and hence to a larger viscosity.
+A well-known example of permeation flow is the one which occurs e.g.
+in cholesterics which are sheared or pushed along the direction of the cholesteric helix.
+However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
+Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve.
+For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches.
+Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.
+
+Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically.
+This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations.
+At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally.
+In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear.
+For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
+We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters.
+In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.
+
+%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
+
 The governing equations in our model are a convection-diffusion equation for the scalar order parameter, which represents the density difference of the two phases and the Navier-Stokes equation with a stress tensor that contains extra contributions due to the presence of the lamellar interfaces.
-This approach allows for stress-strain rate relations to emerge rather than just putting them in by means of consitutive equations.
 In our approach the dynamical equations are solved using a combination of finite difference scheme for the convection-diffusion equation and the lattice Boltzmann method (LBM) \cite{Succi} for the Navier-Stokes equation.
 The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
-The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries[], where it is superior to conventional methods of computational fluid dynamics.\\
+The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.
+
+Our main finding is that as anticipated above depending on system parameters we find that a smectic phase may lead to vastly different rheological responses. 
+When thermodynamic forces are weaker with respect to viscous, flow-induced, ones, shear quickly orders the system into a lamellar phase, with only a few defects.
+The rheology in this regime is Newtonian, and the shear stress is linear in the applied shear rate.
+If, on the other hand, thermodynamic forces are larger, we observe that shear actually leads to the nucleation of new defects, and the rheology is highly nonlinear, with the shear rate proportional to $\dot{\gamma}^{\alpha}$, with $\alpha\simeq 0.87$, leading to a shear thinning behaviour.
+Such a behaviour is qualitatively reminiscent of the one found experimentally in XXX, where the numerical value of $\alpha$ was however different.
+In this regime we also find that the rheology is history dependent, as the defect density decreases for instance if the shear rate is ramped down gradually.
+Finally, intriguingly, dimensionality appears to play a key role: within our 3D simulations, we have indeed not been able to enter the nonlinear rheology regime with defect nucleation, and we always observe a shear-induced ordering into a lamellar phase.
+More research and parameter steering will be needed to make this observation more quantitative.
+We note that sheared smectics have been studied extensively via simulations in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}, but because of computational constraints almost all results were limited to two dimensions and smaller strains than the ones considered here.
+Furthermore in this work we use Lees-Edwards boundary conditions which allow us to probe the true bulk rheology of the system, at variance with previous work which employed solid walls and different boundary conditions, which introduce finite size and boundary effects into the physics.
+
 This paper is organised as follows. In Sec.II we introduce the model, including its equilibrium properties, the full stress tensor and the dynamical equations.

### 2011-09-28 papers/smectics/smectics.tex
Added appendix on SPBC
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1456 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+\section{Appendix: sliding periodic boundary conditions}
+
+The simulations of shear flow are undertaken using a mixed finite
+difference / lattice Boltzmann approach, where the former is used
+to advance the order parameter in time via the Cahn-Hilliard equation,
+and the later is used to solve the Navier-Stokes equations.
+Coupling between the two is
+via the velocity field, which advects the order parameter, and
+the thermodynamic pressure tensor (6), the divergence of which
+contributes a body force on the fluid. The sliding periodic
+boundary conditions of Lees and Edwards\cite{leesedwards} may be
+applied to the lattice Boltzmann part of the calculation
+\cite{wagnerpagonabarraga,adhikari2005} to generate shear in the system. We
+describe here how the sliding periodic boundaries are extended to
+include the finite difference part of the calculation.
+
+We consider a finite difference mesh where the order parameter
+values are co-located with the density and velocity fields provided
+by lattice Boltzmann. Away from the sliding periodic boundaries, the
+finite difference discretisation is as one would expect.
+In order to ensure conservation of order parameter, as expressed by
+\begin{equation}
+\partial_t \phi + \partial_\alpha (u_\alpha \phi - M\partial_\alpha \mu) = 0,
+\end{equation}
+we consider each lattice point to be surrounded by a cubic control volume or
+cell
+of dimensions $\Delta x = 1$ in two or three dimensions.
+We compute convective and diffusive fluxes at the cell boundaries using
+linear interpolation of the normal velocity component to the cell face.
+For the advective fluxes, the interfacial values of the order parameter
+are computed via cubic interpolation weighted in the upwind direction.
+This is combined with an Euler forward step for the time integration:
+a small Courant number ($u\Delta t/\Delta x$), and hence numerically
+stability, is gauranteed by the small Mach number constraint
+standard in lattice Boltzmann. The gradient of the chemical potential
+is computed at each cell face using values of $\mu$ computed at the
+lattice point either side.
+
+A body force on the fluid in the Navier-Stokes equations may be
+included in the lattice Boltzmann approach. Again, to ensure conservation
+of momentum, the divergence of the thermodynamic presure tensor is
+computed based on cell-face values. In practice, this requires a
+finite difference stencil
+that entends at least $\pm 3$ lattice points in each coordinate direction
+(the thermodynamic pressure involves a fourth derivative of the order
+parameter).
+
+The sliding periodic boundaries lie (conceptually) along plane
+co-incident with the cell faces. Where the finite difference stencil 
+overlaps a such a plane, the relative movement between the two sides
+of the plane must be accounted for when computing derivatives of the
+order parameter, and hence fluxes of order parameter and momentum,
+near the planes. This means interpolating values of the order parameter
+on each side of the plane to positions corresponding to the lattice
+locations on the other side so that the finite difference stencil
+can be used normally. For this study, where there are high derivatives
+of the order parameter in the free energy, it is important that this
+interpolation is at least cubic. Linear interpolation of the order
+parameter leads to unphysical oscillations of physical quantities
+such as the shear stress related to the magnitude of the velocity jump
+across the sliding boundary.
+With cubic interpolation, such artefacts are absent. The non-linear
+nature of this interpoation means that fluxes of both order parameter
+and momentum computed at the plane do not exectly match on each side.
+However, this may be corrected at each sliding boundary at each time
+step, restoring global order parameter and momentum conservation.
+
 \footnotesize{

### 2011-10-10 papers/smectics/smectics.tex
Model part and 3D-result part reworked
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1457 40876441-3496-0410-a613-a2fca35b2ae2

@@
   \begin{@twocolumnfalse}
-\noindent\LARGE{\textbf{Nonlinear Rheology of Layered Smectic Liquid Crystals}}
+\noindent\LARGE{\textbf{Large-scale Simulation of Layered Binary Fluids and Smectic Liquid Crystals in Shear Flow}}
 \vspace{0.6cm}
 
-\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{b}$}, K. Stratford\textit{$^{c}$} and M.E. Cates \textit{$^{b}$}}
+\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.V. Coveney{$^{a}$} and M.E. Cates \textit{$^{c}$}}
 }\vspace{0.5cm}
@@
 
-\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, threfore in the nonlinear regime.
+\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, therefore in the nonlinear regime.
 In two dimensions we find that, according to the thermodynamic parameters, shear may either align the system and stabilise an almost regular lamellar phase, or induce the nucleation and proliferation of defects in the smectic layers.
@@
 
-\footnotetext{\textit{$^{a}$~University College of London, London, UK}}
-\footnotetext{\textit{$^{b}$~SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
-\footnotetext{\textit{$^{c}$~EPCC, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh, EH9 3JZ, Scotland.}}
+\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, UK}}
+\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK}}
+\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, UK}}
 
@@
 \section{Introduction}
-Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurth etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid.
-Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibers constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character.
-Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in biosensing.
+Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurt etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid.
+Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibres constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character.
+Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in bio-sensing.
 
@@
 
-The rheological properties of a non-Newtonian fluid are often linked to the microstrucure of the material.
+The rheological properties of a non-Newtonian fluid are often linked to the microstructure of the material.
 For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate.
@@
 Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics.
-Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via preshearing, or not.
+Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via pre-shearing, or not.
 In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample.
@@
 We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters.
-In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like miscelles, surfactants, smectic liquid crystals and block-copolymer.
+In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like micelles, surfactants, smectic liquid crystals and block-copolymer.
 
-%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organize in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
+%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organise in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
 
@@
 
-The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimizes the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.
+The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.
+
 \begin{eqnarray}
@@
 \end{eqnarray}
-The parameters $b$ and $c$ are positive to ensure stability. 
-For $a>0$ there is only one solution $\phi\equiv0$ which minimizes the functional and corresponds to the disordered state.
-For $a<0$ a microphase separation occurs and depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) because the formation of interfaces become energetically favourable.\\
-The dynamics of the binary fluid is governed by a convection-diffusion equation with a diffusive current proportional to the chemical potential $\mu$, which is the functional derivative of ${\cal F}$ with respect to $\phi$.
+
+In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
+For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
+For $a<0$ a microphase separation occurs.
+Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
+The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$, whereas the interfacial width is roughly $\sqrt{2|\kappa|/|a|}$. 
+This entails that by varying $a$ (and $b\sim |a|$) by about on decade, the segregation of the two phases varies by about a factor $3$.\\
+The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.
+
 \begin{equation}\label{OPEOM}
@@
 \end{equation}
-The coefficient $M$ herein plays the role of a mobility coefficient.\\
-If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys a Navier-Stokes equation
+
+The coefficient $M$ herein plays the role of a mobility coefficient and the diffusivity of the system can be approximated by $D_0\simeq M\;b$.\\
+If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
+
 \begin{equation}\label{NSE}
@@
 \end{equation}
-The stress tensor reads
+
+The stress tensor in this equation reads
+
 \begin{equation}
-\sigma_{\alpha \beta}=-P^{th}_{\alpha \beta} +\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha).
+\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-P^{th}_{\alpha \beta}.
 \end{equation}
-It consists of a hydrodynamic part, which is proportonial to the viscosity $\eta$ of the solvent and a thermodynamical part $P^{th}$ that depends on the order parameter.
-The latter can be calculated from the free energy functional:
+
+It consists of a hydrodynamic part, which is proportional to the background viscosity $\eta$ and velocity gradients and a thermodynamical part $P^{th}$ that depends on the order parameter and gradients thereof:
+
 \begin{equation}
-P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + D_{\alpha \beta}(\phi).
+P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + {\cal D}_{\alpha \beta}(\phi).
 \end{equation}
-In this expression the symmetric tensor $D_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee the mechanical equilibrium \cite{Evans79}.
+
+In this last expression the symmetric tensor ${\cal D}_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee mechanical equilibrium \cite{Evans79}.
 The thermodynamic pressure tensor is given by
+
 \begin{equation}
@@
 \end{equation}
+
 with an isotropic pressure contribution
+
 \begin{eqnarray}
@@
 \end{eqnarray}
+
 and an anisotropic part
+
 \begin{equation}
@@
 The Navier-Stokes Eq.\ref{NSE} was solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
-The velocity field were then put into the convection-diffusion equation Eq.\ref{OPEOM} and solved at the same timestep. 
+The velocity field was then used in the advection-diffusion equation Eq.\ref{OPEOM}, which was solved by a finite-difference method at the same timestep. 
 We assumed periodic boundary conditions along the flow and vorticity direction and Lees-Edwards boundary conditions (BC) \cite{Wagner02} along the direction of the velocity gradient.
@@
 Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
+
 \begin{equation}
@@
 \end{equation}
-with $\phi({\mathbf k})$ as normalised Fourier-transform of the order parameter.
+
+with $\phi({\mathbf k})$ being the normalised Fourier-transform of the order parameter.
 The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
-In order to analyse the relation between rheological quantities and the morphology of the system it was necessary to quantify the number of defects.
+For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
 For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
-More explicitely the alignment factor $X_i$ is given as
+More explicitly the alignment factor $X_i$ is given by
+
 \begin{equation}\label{alignment}
@@
 \end{equation}
+
 $N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
 For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
-At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes up smaller positive values.
-By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates from what would be recognised by the eye as an aligned configuration.
-We found a threshold of $\xi=0.8$ to work quite well for our requirements.
-Unfortunately this defect detecting protocol is not able to assign only a single defect site to a disturbed region in the order parameter structure as normally more than one site close to a defect would have an alignment factor below the threshold.
-However, as the number of sites per defect is roughly constant throughout the sample, it is still useful for quantifying the degree of alignment and comparing different systems or flow conditions.\\
-By relating the number of defects sites with $X_i$ below the threshold to the total number of sites it is possible to define a defect density:
-\begin{equation}\label{defectdensity}
+At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
+By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates preceivably from what would be recognised as an aligned configuration exhibiting lamellar order.
+We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
+Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment factor below the threshold.
+However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
+By normalising the number of defects sites with $X_i$ below the threshold a defect density can be defined:
+
+\begin{equation}\label{defect density}
 \rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
 \end{equation}
+
 The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
-The exact value of the defect density depends clearly on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.
+The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.
 
@@
 Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
-Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space.\\
+Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space or $1 {\rm m}=2\cdot10^8$ LBU.\\
 The energy scale is slightly more difficult to calibrate.
@@
 With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
-Hence the LB unit of energy density is about $1\e{6} \rm{Pa}$.\\
+Hence the LB unit of energy density is about $10^{6} \rm{Pa}$ or $1\rm{Pa}=10^{-6}$ LBU.\\
 We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
-As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=1\e{-3+6+14} \rm{LBU}\simeq1\e{17} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq3\e{8}$LBU and one LB timestep $\Delta t$ corresponds to roughly 3ns.\\
+As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=(1\e{-3}) \cdot (1\e{6}) \cdot (4\e{16})\rm{LBU}\simeq4\e{19} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq6.3\e{9}$ LBU and one LB timestep $\Delta t$ corresponds to roughly $0.15 \rm{ns}$.\\
 Finally we want to get an estimated scale for shear rates. 
-A common way to quantify them is the dimensionless Peclet number, which measures the transport of a particle as the ratio of the rate of advection to the rate of diffusion.
-The Peclet number can be defined as $Pe_0=\gd\, D^2/D_0$, where $\gd$ is the rate of strain, $D$ is a characteristic length, e.g. the diameter of a particle and $D_0$ is the diffusivity.
-The highest shear rate we apply is $2\e{-4}$ in LBU, basically reciprocal LB time units.
-With the above value for $\Delta t$ we can convert this into about $5\e{4}{\rm s}^{-1}$.
-If we suppose the Stokes-Einstein diffusion constant is a good approximation for $D_0$, we get for an object of the size of $D\simeq5\e{-8}$ at room temperature in a low viscous solevent like water values of about $D_0=1\e{-11} \rm{m}^2\rm{s}^{-1}$.
-This results in Peclet number of the order ${\cal O}(10)$ and means shear rates $\gd={\cal O}(10^{-4})$ in LBU correspond to a Peclet numbers $Pe_0={\cal O}(10)$.
+A common way is to express the ratio of transport via advection to that via diffusion by means of the Peclet number.
+The Peclet number can be defined as $Pe=\gd\, L^2/D_0$, whereas $\gd$ is the strain rate, $L$ is a characteristic length scale, e.g. the lamellar width and $D_0$ is the diffusivity.
+Because the Peclet number is a dimensionless quantity, we can perform this estimation conveniently in lattice units.
+The highest shear rate we apply are of the order $1\e{-4}$ in reciprocal LB time units.
+With the above given values for the lamellar width of $L\simeq10$ and diffusivity of system A $D_0=M\;b\simeq 0.25 \cdot 5\e{-4}=1.25\e{-4}$ this results in Peclet numbers $Pe=(1\e{-4})\cdot(1\e{2})\cdot(1.25\e{-4})^{-1}=80={\cal O}(10^2)$.
+Hence, the conversion between shear rates and Peclet numbers is roughly $\gd=10^{-4}\cong Pe=10^2$.
 
@@
 
-Our discussion will be in two parts, beginning with simulations in two dimensions for a system A and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
+Our discussion will be in two parts, beginning with simulations in two dimensions for a system A-2D and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
 The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
 By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
-The second part of our discussion consits of simulations in three dimensions using identical parameters to system A and a box size $N_x\times N_y\times N_z=128\times256\times128$.
-We will refer to this system as system C.
-Table \ref{tab1} at the end of this paragraph provides an overview of all simulation parameters.
+The second part of our discussion consists of simulations in three dimensions using identical parameters to system A-2D and a box size $N_x\times N_y\times N_z=128\times256\times128$.
+We will refer to this system as system A-3D.
+Table \ref{tab1} provides an overview of all simulation parameters.
 
-\subsection*{Results in Two Dimensions}
-There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
-Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}.
-We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
-Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
-The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
-Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
-Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{phi_run703_320.jpg}\\
-\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_500.jpg}\\
-\includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
-\caption{Order parameter $\phi$ of system A: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
-Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
-\label{fig1}
-\end{figure}
-This effect has been reported before \cite{Gonnella97} and was more pronounced for softer systems with smaller Landau parameters, although not quite to the same extent as in the above reference.
-Under shear higher viscosities caused the lamellae to be much less homogenous in shape and thickness.\\
-In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the stiffer system A with $-a=b=5\e{-4}$ and the softer system B with $-a=b=5\e{-5}$, respectively.\\
-The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
-The equilibrium structure is completely isotropic and exhibits a lamellar spacing of roughly 10 lattice sites.
-If shear flow is imposed, the lamellae start to align with the flow.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{phi_run707_320.jpg}\\
-\includegraphics[angle=0,width=0.5\textwidth]{phi_run710_800.jpg}\\
-\includegraphics[angle=0,width=0.5\textwidth]{phi_run765_1280.jpg}
-\caption{Order parameter $\phi$ of system B: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom)}
-\label{fig2}
-\end{figure}
-In this paragraph the flow direction and the direction of the velocity gradient are x and y, respectively.
-The apparent flow due to the effect of the Lees-Edwards planes is always in positive x-direction in the upper half and in negative x-direction in the lower half of the box.
-After a short startup phase in which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emarge and the spacing deviates from its equilibrium value.
-These regions are advected with the flow while they undergo an affine stretching.
-The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $1\e{-4}$.
-In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
-However, the reported results differ from our observations in the following aspects.
-First of all, the shear-induced structures emerge at the centre and remain located, while ours occur in bandlike patterns that are wrapped around the periodic boundaries.
-Moreover those bear the characteristics of a shear banding region exhibiting a reduced local shear rate.
-These differences may be a direct consequence of the different boundary conditions.
-While we applied Lees-Edwards BC along y and periodic boundary conditions along x for all quantities, \cite{Xu06b} seem to have used sort of hybrid scheme.
-\begin{figure}[!]
-\centering
-\hspace*{-0.1cm}\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_400.jpg}
-\includegraphics[angle=0,width=0.49\textwidth]{shear_str_run704_400.jpg}
-\includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
-\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A at $t=4\e{5}$}
-\label{fig3}
-\end{figure}
-\begin{figure*}[!]
-\centering
-\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
-\caption{Structure factor $C({\bm k})$ of system A: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right).
-The sections shown correspond to wavevectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
-\label{fig4}
-\end{figure*}
-To our understanding they applied Lees-Edwards BC for the order parameter only, but placed solid moving walls at the upper and lower boundaries in the lattice Boltzmann part.
-Identical to our setup periodic BC in flow direction were applied for both order parameter and momentum flux.
-The it might be that \cite{Xu06b} have not simulated bulk flow behaviour but rather flow in a confined geometry and the shear-induced structure they observe might be a consequence of this difference.\\
-After switching off of the shear flow the expanded regions contract and the compressed regions expand and the system returns to a state with homogenous lamellar width, seemingly close to the equilibrium value.
-The final state for which the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
-The lamellar interfaces remain more or less oriented, but due to the expansion and contraction of the lamellae some individual domains show large angles with the previous flow direction. 
-The length of the interfaces varies considerably, which is appears to be another residue of the inhomogeneities in the flowing state.\\
-Fig. \ref{fig2} shows equivalent states for the softer system B.
-The characteristic results of Fig. \ref{fig1} are retained, but the interfaces in all states tend to be longer.
-The lamellar alignment in flow direction after switch off is more pronounced.\\ 
-Fig. \ref{fig3} relates the inhomogeneities in the morphology to rheological quantities.
-The pictures show the order parameter $\phi$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
-Where the system is extendend and the lamellar width is larger compared to the equilibrium state, both shear stress and first normal stress differences have low positive or negative values.
-Large positive values occur in the regions where the lamellae are squeezed.
-The peak values in lattice units are in the region of $-2\e{-4}$ and $3\e{-4}$ for the shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
-Despite the local negativity of quantities, the overall average are still positive for all times.
-Very similar results with smaller peak magnitudes are obtained for system B, which appears somewhat more homogenous.\\
-Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the three states depicted in Fig. \ref{fig1}.
-At first isotropic after the microphase separation and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
-In steady shear flow the structure factor develops two distinct though rather broad peaks, which are situated at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
-It is interesting to see that the Fourier-vector of the peak position pics up a small component in x-direction, indicating on average a slight tilt of the structure.
-The tilting orients the lamellae away from the direction of flow and towards the velocity gradient direction, which is the direction of extension in simple shear flow.
-The position of the peak is almost constant throughout the entire simulation.
-The average lamellar distance decreases from about 10 lattice sites in the quiescent state to about 8.5 lattice sites.
-However, after switching off the system returns its equilibrium lattice spacing.
-The peak intensity rises due to the fact that the interfaces remain oriented along the flow direction but more interestingly the small x-component of the peak position disappears as the lamellar spacing becomes homogenous again.
-This gives some evidence that the small x-component is indeed connected to extensional flow in the regions with negative first normal stress differences.\\
-With the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify defects in the lamellar structure.
-We would like to locate regions where there is no homogenous order parameter changes because it changes sign and the interfaces have a large local curvature.
-Fig. \ref{fig5} shows an example of the time evolution of system A at low shear.
-All lattice sites with $X_i\le0.8$ are marked red and obviously coincide with interfacial endpoints and disturbances.
-The top picture shows the situation at the beginning of the simulation with the lowest shear rate $\dot{\gamma}=1.25\e{-5}$.
-The system has been pre-sheared at consecutively lowered shear rates ranging from $\gd=1\e{-4}$ to $2.5\e{-5}$, whereas the final configuration of the previous run was always used as initial configuration of the following run.
-There are regions which exhibit some degree of flow alignment and others, where the angle between the lamellae and the flow direction remains rather large.
-The bottom picture of Fig. \ref{fig5} shows the structure at the end of the run.
-The total strain difference between both images is about $10^{5}\%$. 
-Obviously many defects have annealed and the interfaces are well oriented in flow direction.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
-\caption{Defect structure of system A: The pictures show results of a smaller simulation at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
-The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
-Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
-\label{fig5}
-\end{figure}
-This degree of defect annihilation is only found below a critical shear rate where defects are advected smoothly on well aligned lamellae, approach each other and cancel.
-Above the critical shear rate the lamellae buckle quite similar to what has been reported in \cite{Gonnella98}, the structure shows a tendency to trip over itself which creates new defects.
-The critical value is different in both systems, roughly between $\dot{\gamma}=2.5-5\e{-5}$ in the stiffer system A and higher at around $\dot{\gamma}=1-2\e{-4}$ in the softer system B.
-Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A for a sequence of decreasing shear rates.
-The runs with the two highest shear rates were started from the quiescent equilibrated state, whereas all others used the final configuration of the preceding run with higher shear rate as initial configuration.
-Here and in the following figures horizontal lines are time averages for the particular quantities where we think the system has reached a steady state. 
-The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
-Interestingly once the shear rate has been decreased to the next lower rate, both systems A and B show an trend to anneal defects instantaneously or after a short latency time if the shear rate is rather low.
-If the shear rate is turned up again, the lamellae buckle and the structure begins to trip over itself, once again creating new defects.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
-\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
-Horizontal lines indicate averages over time for the particular quantities, where we assume the system has reached a steady state.
-The inset gives data for the 3D-system C and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
-\label{fig6}
-\end{figure}
-This suggests that there is a dynamical equilibrium between defect creation and annihilation above a critical shear rate, while below the critical value defects get not anymore widely generated and the system anneals its defects and forms properly aligned lamellae.
-In order to test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
-The defect density goes indeed up (not shown) and reaches similar values to those shown in Fig. \ref{fig6}.
-However it takes some time before a steady state is reached as the applied shear rates are rather low.\\
-Fig \ref{fig7} shows the average total shear stress density $\langle \sigma_{xy}\rangle$, i.e. the total shear stress per unit lattice volume.
-In the following we will drop the backets $\langle\dots\rangle$ for simplicity and write $\langle \sigma_{xy}\rangle\equiv  \sigma_{xy}$.
-While for the highest applied shear rate the hydrodynamic part forms the largest contribution to the total stress density, a regime where the chemical stress dominates is entered on decreasing the shear rate.
-At low shear rates stress fluctuations become more stretched in time and get smaller, what is slightly hidden by the logarithmic scale of Fig. \ref{fig7}
-Their magnitude is about three times smaller in the larger simulation shown in Fig. \ref{fig1}, \ref{fig2}.
-Fortunately we found that the average shear stress density over time is to a good degree independent of the system size and therefore it is justified to simulate longer times by using a smaller system size without flawing the results.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
-\caption{Average total shear stress $\sigma_{xy}$ versus time for system A: The horizontal lines are steady state time averages}
-\label{fig7}
-\end{figure}
-Furthermore the stress fluctuations represent a physical aspect as they can be related to the inhomogeneities of the order parameter.
-System A shows clear shear thinning behaviour and seems to obey a power law (see Fig. \ref{fig11}).
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the time averaged apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
-Stress fluctuations cause temporary variations of these mean values of up to a factor 3 at the lowest shear rate.
-In contrast to system A system B does not exhibit a clear monotonous shear thinning tendency.
-The apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow more a horizontal trend when the shear rate is decreased.
-The stress fluctuations and time averaged shear stress densities are slightly smaller than in the stiffer system A with one exception.
-For the highest shear rate $\gd=2\e{-4}$ the shear stress turns out to be larger than in system A.\\
-The deviatoric part of the average chemical pressure density $\langle P_{xy}\rangle\equiv P_{xy}$ for system A is shown in the main picture of Fig. \ref{fig8}.
-The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
-System B shows a very steep decrease of $P_{xy}$ during the simulation with $\gd=1\e{-4}$, which is because of the defect annihilation.\\
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A.
-The inset shows data for the 3D-system (system C) and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
-\label{fig8}
-\end{figure}
-Fig. \ref{fig9} illustrates the relation between $P_{xy}$ and the defect density $\rho_D$.
-The black line represents a fit to the system A, which shows a linear dependence almost down to the lowest shear rate.
-The result for the softer system B is slightly more obscured as a consequence of the ordering and the tendency to anneal defects.
-Our quantifying protocol is based on averaging over a larger number of defects sites and is prone to inaccuracy in a regime where defects are rather scarce.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs defect density for system A ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)} 
-\label{fig9}
-\end{figure}
-The main picture in Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A.
-At first until $\gd=5\e{-5}$ the first normal stress differences decrease monotonously with decreasing shear rate.
-At lower shear rates, however, the behaviour is far from being monotonous and $N_1$ rises again.
-This makes it difficult or even impossible to define reasonable time averages as $N_1$ may still evolve at the end of the run.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
-\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{xx}-\sigma_{yy}\rangle$ vs time for system A1.
-The inset shows data for the 3D-system (system C) and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
-\label{fig10}
-\end{figure}
-This feature is actually less erratic than it seems at first sight and can be rationalised. 
-A look at Fig. \ref{fig6} reveals that $N_1$ rises in the same way as the defect density drops.
-A pair of defects can cancel if one of them is an endpoint of a black lamella surrounded by a branching white lamella and the other one is a white endpoint surrounded by a branching black lamella.
-When these two defects anneal the lamella spacing increases locally by half a lamellar width.
-The interfaces have to expand in y-direction as the system is not frustrated enough to create an entire new layer in order to get the lamellar width back closer to its equilibrium value.
-Hence the normal stresses $\sigma_{yy}$ become negative and consecutively smaller, leading to gradually larger values of $N_1$.
-It should be stressed that only $\sigma_{yy}$ seems to evolve anomalously slowly and all other components of the stress tensor reach state state values more rapidly.
-System B again behaves different in the sense that $N_1$ continues to fall for lower shear rates despite there is also ongoing annealing of defects.
-This difference is a direct effect of the softness of system B, whose smaller Landau coefficients prevent the normal stresses $\sigma_{yy}$ from taking up very negative values and $N_1$ from rising.
-Nevertheless close to the end of the simulations with the two lowest shear rates $N_1$ is still slightly changing which suggests that this slowly evolving behaviour is also due to ongoing defect annihilation.
-Because both trends in system A and B are consequences of the impossibility to optimize the lamellar width, larger simulations should lead to more defined steady state values for $N_1$.
-However, we were not able to check this as this proves computationally cosly.\\
-In Fig. \ref{fig11} we show flowcurves derived from time averages of $\sigma_xy$ and $N_1$ as a function of shear rate.
-The time average of shear stress density in systems A follow a power law $\sigma_{xy}=\dot{\gamma}^m$ with an exponent $m\simeq 0.87$ and gives evidence for shear thinning behaviour.
-In case of system B the best fit for the lowest three shear rate is obtained for an exponent $m=1$, indicationg more or less constant viscosity and no shear thinning.
-It is not very surprising as the chemical contribution to the total stress almost vanishes for shear rates $\dot{\gamma}\le1\e{-4}$ due to the significant decrease of the number of defects.
-The observed power law behaviour is simply that of the hydrodynamic part of the stress, which is linear in our case.\\
-A comparison with experimental work remains difficult because the topology of defects differ in two and three dimensions.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
-\caption{Flowcurves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)}
-\label{fig11}
-\end{figure}
-Our two-dimensional system is translational invariant in the third dimension.
-So every branch point or endpoint of a lamella has the topology of an edge defect in the 3D-counterpart of the structure.
-Moreover screw defects do not exist in our picture as they can only occur in more than two dimensions.
-A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
-Interestingly the authors identify exactly the motion of screw defects as the key mechanism responsible for the layer tilting and defect formation. 
-For the power law dependence of the shear stress on the shear rate they find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.\\
-Contrary to shear stresses first normal stress differences do not level off because of ongoing defect annihilation.
-Neither system A nor B is in a steady state at the end of the simulation and therefore the data for $N_1$ is only shown for the sake of completeness.
-
-\subsection*{Results in Three Dimensions}
-
-The computational costs of a detailed analysis of the nonlinear rheology in three dimensions are still substantial as for steady state quantities it is necessary to simulate large strains.
-Therefore we performed only a limited number of simulations for selected parameters. 
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.45\textwidth]{phi_run786_320.jpg}
-\includegraphics[angle=0,width=0.45\textwidth]{phi_run788_960.jpg}
-\includegraphics[angle=0,width=0.45\textwidth]{phi_run792_960.jpg}
-\caption{Order parameter $\phi$ in two and three dimensions: The top and centre picture show system C after equilibration at timestep $t=3.2\e{5}$ and in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at $t=9.6\e{5}$, where shear-induced layering takes place.
-The image at the bottom depicts the corresponding two dimensional system A of the same size in flow and velocity gradient direction.}
-\label{fig12}
-\end{figure}
-For our three dimensional simulations, in the following referred to as system C, we used parameters identical to those of the stiffer 2D-system A, but reduced the system size to $N_x\times N_y \times N_z=128\times256\times128$ lattice sites.
-The directions of vorticity, flow and velocity gradient are now x, y and z, respectively.\\ 
-The top picture in Fig.
-\ref{fig12} shows the order parameter $\phi$ after equilibration at time step $t=3.2\e{5}$ prior to shearing, which served as initial configuration for the runs with shear rate $\gd=1\e{-4}$ and $2\e{-4}$.
-A corresponding configuration, which was equilibrated for another $3.2\e{5}$ timesteps was used for the run with $\gd=\e{-5}$.
-As before black and white correspond to $\phi=-1$ and $\phi=1$, repectively, whereas gray scales give intermediate values of $\phi$.
-The opacity of $phi$ has been linearly reduced from its full value at $\phi=-1$ to 0 at $\phi=1$ in order to reveal the lamellar structure close to the  boundaries.
-In comparison to the 2D-case the isosurfaces of the cuboid, taken as quasi-two-dimensional equivalent in 3D, exhibit a slightly different morphology with shorter protrusions and an overall pattern, which appears somewhat more irregular.
-Nonetheless the quiescent and equilibrated system C adopts the same lamellar spacing of about 10 lattice sites as system A.Fig.
-\ref{fig13} shows a slice through $C({\bm k})$ along $k_y=0$ (cuts along $k_x,k_z=0$ look identical), which resembles the left picture in Fig.
-\ref{fig4} and differs only in the magniude.\\
-However, in shear flow system A and C behave very differently.
-Even for the highest shear rate we applied and contrary to the 2D-case system C orders readily into a regular lamellar structure.
-The centre picture of Fig.\ref{fig12} shows an example at timestep $t=9.6\e{5}$ and shear rate $\gd=1\e{-4}$, which we picked as this was the least ordered one of all three runs.
-With a total number of 13 layers the lamellar spacing is still close to the equilibrium value of roughly 10 lattice sites.
-Besides some irregularities at the front boundary there are obviously edge defects, caused by a branching of the lamellae and visible at the very top and bottom in the foreground and more towards the centre in the background.
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run786_320.jpg}
-\caption{Structure factor $C({\mathbf k})$ of quiescent system C, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
-Shown are wavevectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
-\label{fig13}
-\end{figure}
-\begin{figure*}[!]
-\centering
-\includegraphics[angle=0,width=0.33\textwidth]{ck_x-slice_run788_960.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_y-slice_run788_960.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_z-slice_run788_960.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system C in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wavevectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
-Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
-\label{fig14}
-\end{figure*}
-The edge defects emerge from the spontaneous shear-induced ordering of the phase separated structure but remain very localised at first.
-As a result of shear advection the interfaces spread and close gaps at later times to finally form either continuous lamellae or uniform edges, which are oriented along the flow, virtually stable and do not anneal.
-The fact that separated regions with similar order parameter values can now also connect along a third direction and expell regions with opposite order parameter, seems to play a key role pertaining to the formation of a homogenously layered structure.
-The bottom picture in Fig. \ref{fig12} gives an idea about how the dimesionality affects the morphology of layered smectics and binary fluids.
-It shows the two-dimensional system A as slice of system C along the direction of vorticity.
-The difference between both systems is striking and even more surprising as both results were obtained with the same parameter setting and identical equilibration and shear histories.
-The inset of Fig. \ref{fig6} quantifies the time it takes system C to order into a layererd defect-scarce structure.
-Within a few $10^5$ timesteps the defect density of system C decreases to magnitudes, which we obtained in the 2D-case only for shear rates about one decade lower and times about a factor 10 longer.\\
-Scarce defects lead to smaller chemical pressure, which is shown in the insets of Fig. \ref{fig8} for the component $P_{xy}$.
-The hydrodynamic part of the stress is by far dominating and shear thinning occurs to an ultimate degree where the apparent viscosity of the fluid is that of the solvent and the chemical contribution has practically vanished.
-A relation between the deviatoric chemical pressure tensor and the defect density like in Fig. \ref{fig9} can therefore not be established.
-The inset of Fig.\ref{fig10} shows first normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$.
-For all shear rates $\gd5\e{-5}-2\e{-4}$ system C takes similar values, which are monotonously approached and about one decade smaller than those for $N_1$ of system A.\\ 
-Finally we want to glimpse at the structure factor of system C in shear flow.
-Fig. \ref{fig14} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
-The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
-It shows data in flow-velocity-gradient plane and compares directly with the left picture in Fig. \ref{fig4}.
-The pictures in the centre and on the right give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
-Note the scale of the logarithmic color map in the centre picture.
-Despite a certain similarity with the 2D-structure after switch off in the rightmost picture of Fig.
-\ref{fig4}, the structure factor in three dimensions has a very pronounced peak at $k_z\simeq\pm0.2\pi/l$.
-It appears to be split into two parts, but this is an effect of the relatively sparse discretization and caused by the visualization software.
-Although strongly preaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
-Fig. \ref{fig15} depicts the structure factor in 3D, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
-It reveals a banana-shaped spatial dependence with some small appendices in flow direction at $k_x=0$.
-
-\begin{figure}[!]
-\centering
-\includegraphics[angle=0,width=0.45\textwidth]{ck_run788_960_zoom.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system C in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
-\label{fig15}
-\end{figure}
-
-\begin{table}[htp]
+\begin{table}[htp!]
 \small
@@
 792 & $256\times128\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+\hline
+786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
+787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
+788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
+789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
+\hline
 system B & & & &\\
@@
 \hline
-786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
-787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
-788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
-789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
+998 & $256\times128\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
+1001 & $256\times128\times1$ & 320-640& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
+1002 & $128\times256\times128$ & 320-640 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
 \hline
@@
 \end{table}
+
+\subsection*{Results in Two Dimensions}
+There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
+Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}.
+We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
+Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
+The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
+Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
+Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{phi_run703_320.jpg}\\
+\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_500.jpg}\\
+\includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
+\caption{Order parameter $\phi$ of system A-2D: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
+Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
+\label{fig1}
+\end{figure}
+
+This effect has been reported before \cite{Gonnella97} and was more pronounced for softer systems with smaller Landau parameters, although not quite to the same extent as in the above reference.
+Under shear higher viscosities caused the lamellae to be much less homogenous in shape and thickness.\\
+In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the stiffer system A-2D with $-a=b=5\e{-4}$ and the softer system B with $-a=b=5\e{-5}$, respectively.\\
+The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
+The equilibrium structure is completely isotropic and exhibits a lamellar spacing of roughly 10 lattice sites.
+If shear flow is imposed, the lamellae start to align with the flow.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{phi_run707_320.jpg}\\
+\includegraphics[angle=0,width=0.5\textwidth]{phi_run710_800.jpg}\\
+\includegraphics[angle=0,width=0.5\textwidth]{phi_run765_1280.jpg}
+\caption{Order parameter $\phi$ of system B: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom)}
+\label{fig2}
+\end{figure}
+
+In this paragraph the flow direction and the direction of the velocity gradient are x and y, respectively.
+The apparent flow due to the effect of the Lees-Edwards planes is always in positive x-direction in the upper half and in negative x-direction in the lower half of the box.
+After a short startup phase in which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing deviates from its equilibrium value.
+These regions are advected with the flow while they undergo an affine stretching.
+The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $1\e{-4}$.
+In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
+However, the reported results differ from our observations in the following aspects.
+First of all, the shear-induced structures emerge at the centre and remain located, while ours occur in band-like patterns that are wrapped around the periodic boundaries.
+Moreover those bear the characteristics of a shear banding region exhibiting a reduced local shear rate.
+These differences may be a direct consequence of the different boundary conditions.
+While we applied Lees-Edwards BC along y and periodic boundary conditions along x for all quantities, \cite{Xu06b} seem to have used sort of hybrid scheme.
+
+\begin{figure}[htp!]
+\centering
+\hspace*{-0.1cm}\includegraphics[angle=0,width=0.5\textwidth]{phi_run704_400.jpg}
+\includegraphics[angle=0,width=0.49\textwidth]{shear_str_run704_400.jpg}
+\includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
+\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
+\label{fig3}
+\end{figure}
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.33\textwidth]{ck_run703_320.jpg}
+\includegraphics[angle=0,width=0.33\textwidth]{ck_run704_500.jpg}
+\includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
+\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right).
+The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
+\label{fig4}
+\end{figure}
+
+To our understanding they applied Lees-Edwards BC for the order parameter only, but placed solid moving walls at the upper and lower boundaries in the lattice Boltzmann part.
+Identical to our setup periodic BC in flow direction were applied for both order parameter and momentum flux.
+The it might be that \cite{Xu06b} have not simulated bulk flow behaviour but rather flow in a confined geometry and the shear-induced structure they observe might be a consequence of this difference.\\
+After switching off of the shear flow the expanded regions contract and the compressed regions expand and the system returns to a state with homogenous lamellar width, seemingly close to the equilibrium value.
+The final state for which the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
+The lamellar interfaces remain more or less oriented, but due to the expansion and contraction of the lamellae some individual domains show large angles with the previous flow direction. 
+The length of the interfaces varies considerably, which is appears to be another residue of the inhomogeneities in the flowing state.\\
+Fig. \ref{fig2} shows equivalent states for the softer system B.
+The characteristic results of Fig. \ref{fig1} are retained, but the interfaces in all states tend to be longer.
+The lamellar alignment in flow direction after switch off is more pronounced.\\ 
+Fig. \ref{fig3} relates the inhomogeneities in the morphology to rheological quantities.
+The pictures show the order parameter $\phi$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
+Where the system is extended and the lamellar width is larger compared to the equilibrium state, both shear stress and first normal stress differences have low positive or negative values.
+Large positive values occur in the regions where the lamellae are squeezed.
+The peak values in lattice units are in the region of $-2\e{-4}$ and $3\e{-4}$ for the shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
+Despite the local negativity of quantities, the overall average are still positive for all times.
+Very similar results with smaller peak magnitudes are obtained for system B, which appears somewhat more homogenous.\\
+Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the three states depicted in Fig. \ref{fig1}.
+At first isotropic after the microphase separation and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
+In steady shear flow the structure factor develops two distinct though rather broad peaks, which are situated at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
+It is interesting to see that the Fourier-vector of the peak position pics up a small component in x-direction, indicating on average a slight tilt of the structure.
+The tilting orients the lamellae away from the direction of flow and towards the velocity gradient direction, which is the direction of extension in simple shear flow.
+The position of the peak is almost constant throughout the entire simulation.
+The average lamellar distance decreases from about 10 lattice sites in the quiescent state to about 8.5 lattice sites.
+However, after switching off the system returns its equilibrium lattice spacing.
+The peak intensity rises due to the fact that the interfaces remain oriented along the flow direction but more interestingly the small x-component of the peak position disappears as the lamellar spacing becomes homogenous again.
+This gives some evidence that the small x-component is indeed connected to extensional flow in the regions with negative first normal stress differences.\\
+With the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify defects in the lamellar structure.
+We would like to locate regions where there is no homogenous order parameter changes because it changes sign and the interfaces have a large local curvature.
+Fig. \ref{fig5} shows an example of the time evolution of system A-2D at low shear.
+All lattice sites with $X_i\le0.8$ are marked red and obviously coincide with interfacial endpoints and disturbances.
+The top picture shows the situation at the beginning of the simulation with the lowest shear rate $\dot{\gamma}=1.25\e{-5}$.
+The system has been pre-sheared at consecutively lowered shear rates ranging from $\gd=1\e{-4}$ to $2.5\e{-5}$, whereas the final configuration of the previous run was always used as initial configuration of the following run.
+There are regions which exhibit some degree of flow alignment and others, where the angle between the lamellae and the flow direction remains rather large.
+The bottom picture of Fig. \ref{fig5} shows the structure at the end of the run.
+The total strain difference between both images is about $10^{5}\%$. 
+Obviously many defects have annealed and the interfaces are well oriented in flow direction.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
+\caption{Defect structure of system A-2D: The pictures show results of a smaller simulation at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
+The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
+Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
+\label{fig5}
+\end{figure}
+
+This degree of defect annihilation is only found below a critical shear rate where defects are advected smoothly on well aligned lamellae, approach each other and cancel.
+Above the critical shear rate the lamellae buckle quite similar to what has been reported in \cite{Gonnella98}, the structure shows a tendency to trip over itself which creates new defects.
+The critical value is different in both systems, roughly between $\dot{\gamma}=2.5-5\e{-5}$ in the stiffer system A-2D and higher at around $\dot{\gamma}=1-2\e{-4}$ in the softer system B.
+Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates.
+The runs with the two highest shear rates were started from the quiescent equilibrated state, whereas all others used the final configuration of the preceding run with higher shear rate as initial configuration.
+Here and in the following figures horizontal lines are time averages for the particular quantities where we think the system has reached a steady state. 
+The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
+Interestingly once the shear rate has been decreased to the next lower rate, both systems A and B show an trend to anneal defects instantaneously or after a short latency time if the shear rate is rather low.
+If the shear rate is turned up again, the lamellae buckle and the structure begins to trip over itself, once again creating new defects.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
+\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
+Horizontal lines indicate averages over time for the particular quantities, where we assume the system has reached a steady state.
+The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\label{fig6}
+\end{figure}
+
+This suggests that there is a dynamical equilibrium between defect creation and annihilation above a critical shear rate, while below the critical value defects get not anymore widely generated and the system anneals its defects and forms properly aligned lamellae.
+In order to test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
+The defect density goes indeed up (not shown) and reaches similar values to those shown in Fig. \ref{fig6}.
+However it takes some time before a steady state is reached as the applied shear rates are rather low.\\
+Fig \ref{fig7} shows the average total shear stress density $\langle \sigma_{xy}\rangle$, i.e. the total shear stress per unit lattice volume.
+In the following we will drop the brackets $\langle\dots\rangle$ for simplicity and write $\langle \sigma_{xy}\rangle\equiv  \sigma_{xy}$.
+While for the highest applied shear rate the hydrodynamic part forms the largest contribution to the total stress density, a regime where the chemical stress dominates is entered on decreasing the shear rate.
+At low shear rates stress fluctuations become more stretched in time and get smaller, what is slightly hidden by the logarithmic scale of Fig. \ref{fig7}
+Their magnitude is about three times smaller in the larger simulation shown in Fig. \ref{fig1}, \ref{fig2}.
+Fortunately we found that the average shear stress density over time is to a good degree independent of the system size and therefore it is justified to simulate longer times by using a smaller system size without flawing the results.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
+\caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages}
+\label{fig7}
+\end{figure}
+
+Furthermore the stress fluctuations represent a physical aspect as they can be related to the inhomogeneities of the order parameter.
+System A shows clear shear thinning behaviour and seems to obey a power law (see Fig. \ref{fig11}).
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the time averaged apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
+Stress fluctuations cause temporary variations of these mean values of up to a factor 3 at the lowest shear rate.
+In contrast to system A-2D system B does not exhibit a clear monotonous shear thinning tendency.
+The apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow more a horizontal trend when the shear rate is decreased.
+The stress fluctuations and time averaged shear stress densities are slightly smaller than in the stiffer system A-2D with one exception.
+For the highest shear rate $\gd=2\e{-4}$ the shear stress turns out to be larger than in system A-2D.\\
+The deviatoric part of the average chemical pressure density $\langle P_{xy}\rangle\equiv P_{xy}$ for system A-2D is shown in the main picture of Fig. \ref{fig8}.
+The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
+System B shows a very steep decrease of $P_{xy}$ during the simulation with $\gd=1\e{-4}$, which is because of the defect annihilation.\\
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
+\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\label{fig8}
+\end{figure}
+
+Fig. \ref{fig9} illustrates the relation between $P_{xy}$ and the defect density $\rho_D$.
+The black line represents a fit to the system A-2D, which shows a linear dependence almost down to the lowest shear rate.
+The result for the softer system B is slightly more obscured as a consequence of the ordering and the tendency to anneal defects.
+Our quantifying protocol is based on averaging over a larger number of defects sites and is prone to inaccuracy in a regime where defects are rather scarce.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
+\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs. defect density for system A-2D and B-2D } 
+\label{fig9}
+\end{figure}
+
+The main picture in Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A-2D.
+At first until $\gd=5\e{-5}$ the first normal stress differences decrease monotonously with decreasing shear rate.
+At lower shear rates, however, the behaviour is far from being monotonous and $N_1$ rises again.
+This makes it difficult or even impossible to define reasonable time averages as $N_1$ may still evolve at the end of the run.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
+\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{xx}-\sigma_{yy}\rangle$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\label{fig10}
+\end{figure}
+
+This feature is actually less erratic than it seems at first sight and can be rationalised. 
+A look at Fig. \ref{fig6} reveals that $N_1$ rises in the same way as the defect density drops.
+A pair of defects can cancel if one of them is an endpoint of a black lamella surrounded by a branching white lamella and the other one is a white endpoint surrounded by a branching black lamella.
+When these two defects anneal the lamella spacing increases locally by half a lamellar width.
+The interfaces have to expand in y-direction as the system is not frustrated enough to create an entire new layer in order to get the lamellar width back closer to its equilibrium value.
+Hence the normal stresses $\sigma_{yy}$ become negative and consecutively smaller, leading to gradually larger values of $N_1$.
+It should be stressed that only $\sigma_{yy}$ seems to evolve anomalously slowly and all other components of the stress tensor reach state state values more rapidly.
+System B again behaves different in the sense that $N_1$ continues to fall for lower shear rates despite there is also ongoing annealing of defects.
+This difference is a direct effect of the softness of system B, whose smaller Landau coefficients prevent the normal stresses $\sigma_{yy}$ from taking up very negative values and $N_1$ from rising.
+Nevertheless close to the end of the simulations with the two lowest shear rates $N_1$ is still slightly changing which suggests that this slowly evolving behaviour is also due to ongoing defect annihilation.
+Because both trends in system A-2D and B are consequences of the impossibility to optimise the lamellar width, larger simulations should lead to more defined steady state values for $N_1$.
+However, we were not able to check this as this proves computationally costly.\\
+In Fig. \ref{fig11} we show flow curves derived from time averages of $\sigma_xy$ and $N_1$ as a function of shear rate.
+The time average of shear stress density in systems A follow a power law $\sigma_{xy}=\dot{\gamma}^m$ with an exponent $m\simeq 0.87$ and gives evidence for shear thinning behaviour.
+In case of system B the best fit for the lowest three shear rate is obtained for an exponent $m=1$, indicating more or less constant viscosity and no shear thinning.
+It is not very surprising as the chemical contribution to the total stress almost vanishes for shear rates $\dot{\gamma}\le1\e{-4}$ due to the significant decrease of the number of defects.
+The observed power law behaviour is simply that of the hydrodynamic part of the stress, which is linear in our case.\\
+A comparison with experimental work remains difficult because the topology of defects differ in two and three dimensions.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
+\caption{Flow curves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)}
+\label{fig11}
+\end{figure}
+
+Our two-dimensional system is translational invariant in the third dimension.
+So every branch point or endpoint of a lamella has the topology of an edge defect in the 3D-counterpart of the structure.
+Moreover screw defects do not exist in our picture as they can only occur in more than two dimensions.
+A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
+Interestingly the authors identify exactly the motion of screw defects as the key mechanism responsible for the layer tilting and defect formation. 
+For the power law dependence of the shear stress on the shear rate they find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.\\
+Contrary to shear stresses first normal stress differences do not level off because of ongoing defect annihilation.
+Neither system A-2D nor B is in a steady state at the end of the simulation and therefore the data for $N_1$ is only shown for the sake of completeness.
+
+\subsection*{Results in Three Dimensions}
+
+Because steady states require to account for large strains, the computational costs of a similar analysis of nonlinear rheology in three-dimensions are still prohibitive.
+In the following we compare 3D-systems with 2D-systems smaller than those of the previous paragraph but large enough to show qualitatively similar structure under flow.
+
+\begin{figure*}[htp!]
+\centering
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1002_320k.png}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1002_640k.png}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1001_640k.png}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1003_960k.png}
+\caption{Order parameter $\phi$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\cdot10^{-4}$) at $t=6.4\e{5}$.
+The opacity of one phase has been reduced to enable a better view into the lamellar structure.
+The bottom left picture shows System C-2D at the same time step and shear rate.
+The bottom right picture is system A-2D at $\dot{\gamma}=1\cdot10^{-4}$ and time step $t=9.6\e{5}$.
+Results for A-3D are morphologically identical to those above.}
+\label{fig12}
+\end{figure*}
+
+The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which has a slightly larger segregation than system B.
+The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $256\times128\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
+Two snapshots of system C-2D and A-2D in shear flow are shown in the bottom pictures in Fig. \ref{fig12}.
+They feature the typical fluctuating pattern of breaking and merging lamellae that was described in the previous paragraph and which is particularly pronounced in the case of system C-2D.
+The top picture visualise results for system C-3D with the same free energy parameters as C-2D.
+The quiescent state of system C-3D after equilibration is shown at the left with a morphology virtually identical to that of A-3D.
+The lamellar width of roughly 10 lattice sites is more or less the same as in two dimensions.
+A slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig. \ref{fig4}.
+However, in shear flow considerable differences arise between the two- and three-dimensional case.
+Contrary to the result in 2D, the three dimensional systems order readily into a regular lamellar structure after a transient period as shown in Fig. \ref{fig12} for system C-3D. 
+We opted for this system as the differences between the 3D- and 2D-system are the most striking we observed.
+Interestingly, these findings are consistent for even the highest shear rates, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run786_320.jpg}
+\caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
+Shown are wave vectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
+\label{fig13}
+\end{figure}
+
+\begin{figure*}[htp!]
+\centering
+\includegraphics[angle=0,width=0.33\textwidth]{ck_x-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.33\textwidth]{ck_y-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.33\textwidth]{ck_z-slice_run788_960.jpg}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
+Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
+\label{fig14}
+\end{figure*}
+
+Nevertheless, the order is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers.
+The Burgers vector is oriented along the gradient direction with edges along the flow direction.
+This forms a structure that is translational invariant along the flow direction and in the steady state the location of the edge defects does not change and the defects to not anneal.
+A closer look at the kinetic evolution of the order parameter reveals that regions separated by different values of order parameter connect along a third direction by expelling regions with opposite order parameter.
+This is a direct consequence of the additional dimension and seems to play a key role during the formation of homogeneously layered structure.\\
+It is instructive to quantify the time it takes system A-3D to emerge as lamellar ordered structure starting from an isotropic initial configuration.  
+The inset of Fig. \ref{fig6} gives the defect density as defined by Eq. \ref{defect density} versus time for shear rates $\dot{gamma}=5\cdot10^{-5}-2\cdot10^{-4}$. 
+Within a few $10^5$ timesteps the defect density decreases to magnitudes, which we obtained in case of system A-2D for shear rates about one decade lower.
+The total strain where the curves level off and ordering is achieved scales with the shear rate and is about $\gamma=1.3$ ($\dot{gamma}=2\cdot10^{-4}$), $\gamma=0.8$ ($\dot{gamma}=1\cdot10^{-4}$) and $\gamma=0.6$ ($\dot{gamma}=5\cdot10^{-5}$).
+ \\
+A lower number of defects yields smaller chemical pressure, which is shown for the deviatortic part $P_{xy}$ in the inset of Fig. \ref{fig8}.
+This contribution to the total stress is dwarfed by the hydrodynamic part and system A-3D exhibits shear thinning almost to an ultimate degree.
+A relation between the magnitude of the deviatoric part of the chemical pressure and the defect density like that in Fig. \ref{fig9} in the two-dimensional case cannot be deduced.\\
+First normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig10}.
+For all shear rates $\gd5\e{-5}-2\e{-4}$ system A-3D takes similar values, which are monotonously approached and about one decade smaller than those of system A-2D.\\ 
+Finally we want to glimpse at the structure factor of system A-3D in shear flow.
+Fig. \ref{fig14} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
+The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
+It shows data in flow-velocity-gradient plane and can be directly compared with the left picture in Fig. \ref{fig4}.
+The other images give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
+Note that for the centre picture a logarithmic colour map had to be adopted due to the strong ordering in 3D.
+There is a certain similarity between the 2D-structure factor {\it after} switch-off (Fig.
+\ref{fig4}) and the 3D-structure, but the latter shows a pronounced peak at $k_z\simeq\pm0.2\pi/l$ also during shear flow, which is slightly concealed by the logarithmic scale.The split nature is an effect of the relatively sparse discretization and was caused by the visualisation software.
+Although strongly peaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
+Fig. \ref{fig15} depicts a close up of the 3D-structure factor, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
+It reveals a banana-shaped spatial dependence with some small appendices in flow direction at $k_x=0$.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[angle=0,width=0.45\textwidth]{ck_run788_960_zoom.jpg}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
+\label{fig15}
+\end{figure}
 
@@
 Below a critical shear rate the systems showed no tendency anymore to trip over itself and defects annealed as they could approach each other in adjacent lamellae.
-The characteritic time of this process was set by the shear rate and the total strain.
-The deviatoric part of the chemical pressure was correlated to the number of defect sites, wheras the latter was measured by means of an alignment factor, which quantified the localby curvature of the lamellar interfaces.
-For the stiffer system A we could verify a linear relation between the defect density and the average chemical contribution to the shear stress.\\ 
-In system A first normal stress differences decreased as long as a dynamic equilibrium between defect creation an annihilation existed.
+The characteristic time of this process was set by the shear rate and the total strain.
+The deviatoric part of the chemical pressure was correlated to the number of defect sites, whereas the latter was measured by means of an alignment factor, which quantified the local curvature of the lamellar interfaces.
+For the stiffer system A-2D we could verify a linear relation between the defect density and the average chemical contribution to the shear stress.\\ 
+In system A-2D first normal stress differences decreased as long as a dynamic equilibrium between defect creation an annihilation existed.
 Below the critical threshold this was no longer the case and pronounced defect annealing caused a rise in the first normal stress differences as the average lamellar spacing increased.
-We could establish a shear thinning regime for the more defect ladden system A.
+We could establish a shear thinning regime for the more defect laden system A-2D.
 System B in contrary was so perfectly annealed in the steady state that virtually shear stresses were down at the hydrodynamic level and the relation between shear stress and shear rate was trivially linear.\\
 In three dimensions we simulated shear strains which have not been attained so far.
-Interestingly for the same parameters we observe striking differences between system A in 2D and system C in 3D.
+Interestingly for the same parameters we observe striking differences between system A-2D in 2D and system A-3D in 3D.
 It seems that the dynamics of the defects in two and three dimensions is rather incommensurable.
@@
 applied to the lattice Boltzmann part of the calculation
-\cite{wagnerpagonabarraga,adhikari2005} to generate shear in the system. We
+\cite{Wagner02,adhikari2005} to generate shear in the system. We
 describe here how the sliding periodic boundaries are extended to
@@
 a small Courant number ($u\Delta t/\Delta x$), and hence numerically
-stability, is gauranteed by the small Mach number constraint
+stability, is guaranteed by the small Mach number constraint
 standard in lattice Boltzmann. The gradient of the chemical potential
@@
 included in the lattice Boltzmann approach. Again, to ensure conservation
-of momentum, the divergence of the thermodynamic presure tensor is
+of momentum, the divergence of the thermodynamic pressure tensor is
 computed based on cell-face values. In practice, this requires a
@@
 With cubic interpolation, such artefacts are absent. The non-linear
-nature of this interpoation means that fluxes of both order parameter
-and momentum computed at the plane do not exectly match on each side.
+nature of this interpolation means that fluxes of both order parameter
+and momentum computed at the plane do not exactly match on each side.
 However, this may be corrected at each sliding boundary at each time

### 2011-10-12 trunk/papers/smectics/smectics.tex
2D-result paragraph reworked - conclusions pending
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1460 40876441-3496-0410-a613-a2fca35b2ae2

@@
   \begin{@twocolumnfalse}
-\noindent\LARGE{\textbf{Nonlinear Rheology of Layered Smectic Liquid Crystals}}
+\noindent\LARGE{\textbf{Shear Flow of Lamellar Mesophases and Smectic Liquid Crystals}}
 \vspace{0.6cm}
 
-\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, D. Marenduzzo\textit{$^{b}$}, K. Stratford\textit{$^{c}$} and M.E. Cates \textit{$^{b}$}}
+\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.V. Coveney{$^{a}$} and M.E. Cates \textit{$^{c}$}}
 }\vspace{0.5cm}
@@
 
-Our discussion will be in two parts, beginning with simulations in two dimensions for a system A and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
-The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
-By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
-The second part of our discussion consits of simulations in three dimensions using identical parameters to system A and a box size $N_x\times N_y\times N_z=128\times256\times128$.
-We will refer to this system as system C.
-Table \ref{tab1} at the end of this paragraph provides an overview of all simulation parameters.
+Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
+The system size was $N_x\times N_y\times N_z=512\time1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
+Phase segregation was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq.
+\ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
+By this we obtained a slightly different stress response with the system with lower segregation level appearing somwhat sofer.
+The lamellar distance was identical in both cases.\\
+The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
+Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
+Table \ref{tab1} gives an overview of all simulation parameters.
+
+\begin{table}[htp]
+\small
+\caption{Simulation parameters: Starting from spinodal decomposition the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\cdot10^{-5}$ to $2\cdot10^{-4}$.
+In some cases we performed a final switch off run (SO) at zero shear rate.
+The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
+\label{tab1}
+\begin{tabular*}{0.495\textwidth}{lllll}
+\hline
+run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
+\hline
+system A & & & &\\
+\hline 
+703 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
+704 & $512\times1024\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
+705 & $512\times1024\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
+\hline
+767 & $256\times512\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
+775 & $256\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
+769 & $256\times512\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
+771 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
+772 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
+774 & $256\times512\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
+\hline
+792 & $128\times256\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+\hline
+786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
+787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
+788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
+789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
+\hline
+system B & & & &\\
+\hline
+707 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
+710 & $512\times1024\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
+765 & $512\times1024\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
+\hline
+776 & $256\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
+777 & $256\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
+778 & $256\times512\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
+780 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
+781 & $256\times512\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
+782 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
+\hline
+system C & & & &\\
+\hline
+998 & $128\times256\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
+1001 & $128\times256\times1$ & 320-960& $1\cdot10^{-4}$ & $2\cdot10^{-4}$ \\
+1000 & $128\times256\times1$ & 320-1280& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
+1003 & $128\times256\times128$ & 320-960 & $1\cdot10^{-4}$ & $2\cdot10^{-4}$\\
+1002 & $128\times256\times128$ & 320-1280 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
+\hline
+\end{tabular*}
+\end{table}
 
 \subsection*{Results in Two Dimensions}
-There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
-Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}.
-We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
+
+Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq.
+\ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
+Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
+In previous studies \cite{Kendon01,Xu03, Xu06b} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
+We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
 Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
 The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
-Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
-Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.
-\begin{figure}[!]
+Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems with lower segregation have lower average order parameters and are generally closer to the critical point.
+Higher viscosities lead to shorter interfaces, whereas lower values typically invoke longer lamellae.
+This effect has been reported before \cite{Gonnella97}.
+It turned out to be more pronounced for low segregation, although not as quite to the same extent as in the above reference.
+In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
+In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
+The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
+The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
+If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
+
+\begin{figure}[htp]
 \centering
@@
 \includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
-\caption{Order parameter $\phi$ of system A: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
+\caption{Order parameter $\phi$ of system A-2D: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
 Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
@@
 \end{figure}
-This effect has been reported before \cite{Gonnella97} and was more pronounced for softer systems with smaller Landau parameters, although not quite to the same extent as in the above reference.
-Under shear higher viscosities caused the lamellae to be much less homogenous in shape and thickness.\\
-In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the stiffer system A with $-a=b=5\e{-4}$ and the softer system B with $-a=b=5\e{-5}$, respectively.\\
-The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
-The equilibrium structure is completely isotropic and exhibits a lamellar spacing of roughly 10 lattice sites.
-If shear flow is imposed, the lamellae start to align with the flow.
-\begin{figure}[!]
+
+\begin{figure}[htp]
 \centering
@@
 \end{figure}
-In this paragraph the flow direction and the direction of the velocity gradient are x and y, respectively.
-The apparent flow due to the effect of the Lees-Edwards planes is always in positive x-direction in the upper half and in negative x-direction in the lower half of the box.
-After a short startup phase in which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emarge and the spacing deviates from its equilibrium value.
-These regions are advected with the flow while they undergo an affine stretching.
-The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $1\e{-4}$.
+
+The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
+The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
+After a short transient phase during which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing begins to deviate from its well-defined equilibrium value.
+These regions are advected with the flow while they continuously undergo affine transformations.
+The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $\gd= 1\e{-4}$.
 In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
 However, the reported results differ from our observations in the following aspects.
-First of all, the shear-induced structures emerge at the centre and remain located, while ours occur in bandlike patterns that are wrapped around the periodic boundaries.
-Moreover those bear the characteristics of a shear banding region exhibiting a reduced local shear rate.
-These differences may be a direct consequence of the different boundary conditions.
-While we applied Lees-Edwards BC along y and periodic boundary conditions along x for all quantities, \cite{Xu06b} seem to have used sort of hybrid scheme.
-\begin{figure}[!]
+Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
+Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
+These differences may be simply a direct consequence of the different boundary conditions that were applied.
+While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.
+
+\begin{figure}[htp!]
 \centering
@@
 \includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
-\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A at $t=4\e{5}$}
+\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
 \label{fig3}
 \end{figure}
-\begin{figure*}[!]
+
+\begin{figure}[htp!]
 \centering
@@
 \includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
-\caption{Structure factor $C({\bm k})$ of system A: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right).
-The sections shown correspond to wavevectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
+\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right).
+The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
 \label{fig4}
-\end{figure*}
-To our understanding they applied Lees-Edwards BC for the order parameter only, but placed solid moving walls at the upper and lower boundaries in the lattice Boltzmann part.
-Identical to our setup periodic BC in flow direction were applied for both order parameter and momentum flux.
-The it might be that \cite{Xu06b} have not simulated bulk flow behaviour but rather flow in a confined geometry and the shear-induced structure they observe might be a consequence of this difference.\\
-After switching off of the shear flow the expanded regions contract and the compressed regions expand and the system returns to a state with homogenous lamellar width, seemingly close to the equilibrium value.
-The final state for which the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
-The lamellar interfaces remain more or less oriented, but due to the expansion and contraction of the lamellae some individual domains show large angles with the previous flow direction. 
-The length of the interfaces varies considerably, which is appears to be another residue of the inhomogeneities in the flowing state.\\
-Fig. \ref{fig2} shows equivalent states for the softer system B.
-The characteristic results of Fig. \ref{fig1} are retained, but the interfaces in all states tend to be longer.
-The lamellar alignment in flow direction after switch off is more pronounced.\\ 
-Fig. \ref{fig3} relates the inhomogeneities in the morphology to rheological quantities.
-The pictures show the order parameter $\phi$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
-Where the system is extendend and the lamellar width is larger compared to the equilibrium state, both shear stress and first normal stress differences have low positive or negative values.
-Large positive values occur in the regions where the lamellae are squeezed.
-The peak values in lattice units are in the region of $-2\e{-4}$ and $3\e{-4}$ for the shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
-Despite the local negativity of quantities, the overall average are still positive for all times.
-Very similar results with smaller peak magnitudes are obtained for system B, which appears somewhat more homogenous.\\
-Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the three states depicted in Fig. \ref{fig1}.
-At first isotropic after the microphase separation and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
-In steady shear flow the structure factor develops two distinct though rather broad peaks, which are situated at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
-It is interesting to see that the Fourier-vector of the peak position pics up a small component in x-direction, indicating on average a slight tilt of the structure.
-The tilting orients the lamellae away from the direction of flow and towards the velocity gradient direction, which is the direction of extension in simple shear flow.
-The position of the peak is almost constant throughout the entire simulation.
-The average lamellar distance decreases from about 10 lattice sites in the quiescent state to about 8.5 lattice sites.
-However, after switching off the system returns its equilibrium lattice spacing.
-The peak intensity rises due to the fact that the interfaces remain oriented along the flow direction but more interestingly the small x-component of the peak position disappears as the lamellar spacing becomes homogenous again.
-This gives some evidence that the small x-component is indeed connected to extensional flow in the regions with negative first normal stress differences.\\
-With the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify defects in the lamellar structure.
-We would like to locate regions where there is no homogenous order parameter changes because it changes sign and the interfaces have a large local curvature.
-Fig. \ref{fig5} shows an example of the time evolution of system A at low shear.
-All lattice sites with $X_i\le0.8$ are marked red and obviously coincide with interfacial endpoints and disturbances.
-The top picture shows the situation at the beginning of the simulation with the lowest shear rate $\dot{\gamma}=1.25\e{-5}$.
-The system has been pre-sheared at consecutively lowered shear rates ranging from $\gd=1\e{-4}$ to $2.5\e{-5}$, whereas the final configuration of the previous run was always used as initial configuration of the following run.
-There are regions which exhibit some degree of flow alignment and others, where the angle between the lamellae and the flow direction remains rather large.
-The bottom picture of Fig. \ref{fig5} shows the structure at the end of the run.
-The total strain difference between both images is about $10^{5}\%$. 
-Obviously many defects have annealed and the interfaces are well oriented in flow direction.
-\begin{figure}[!]
+\end{figure}
+
+To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
+It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
+This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\ 
+After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
+The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
+Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
+The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
+Fig. \ref{fig2} shows equivalent states for system B-2D with lower segregation.
+All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
+Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
+In Fig. \ref{fig3} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
+The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
+In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
+Large positive values of both quantities coincide with regions where the lamellae are under compression.
+Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
+Despite being locall negative, the ensemble averages of these quantities are always positive.
+Very similar results with smaller peak magnitudes are obtained for system B-2D.
+Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
+Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
+In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
+It is interesting to see that the inhomogenous morphology of the lamellar structure is also clearly encoded in the structure factor. 
+The Fourier-vector at the peak has a small component in x-direction, indicating the average tilt of the lamellae away from the flow direction in the direction of extensional flow.
+The position of the peak is almost constant throughout the entire simulation and the average lamellar spacing decreases from about 10 lattice sites in equilibrium to about 8.5 lattice sites.
+If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
+The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.
+By introducing the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
+Fig. \ref{fig5} shows an example of the time evolution of system A-2D at low shear rate.
+All lattice sites with $X_i\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
+The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
+There are regions which show some sort of flow alignment and others which don't.
+The bottom picture of Fig. \ref{fig5} shows a later state with a total strain that has increased by another $10^{5}\%$. 
+Obviously many defects have annealed and the interfaces are well aligned with the flow direction.
+
+\begin{figure}[htp!]
 \centering
@@
 \includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
-\caption{Defect structure of system A: The pictures show results of a smaller simulation at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
+\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
 The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
@@
 \end{figure}
-This degree of defect annihilation is only found below a critical shear rate where defects are advected smoothly on well aligned lamellae, approach each other and cancel.
-Above the critical shear rate the lamellae buckle quite similar to what has been reported in \cite{Gonnella98}, the structure shows a tendency to trip over itself which creates new defects.
-The critical value is different in both systems, roughly between $\dot{\gamma}=2.5-5\e{-5}$ in the stiffer system A and higher at around $\dot{\gamma}=1-2\e{-4}$ in the softer system B.
-Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A for a sequence of decreasing shear rates.
-The runs with the two highest shear rates were started from the quiescent equilibrated state, whereas all others used the final configuration of the preceding run with higher shear rate as initial configuration.
-Here and in the following figures horizontal lines are time averages for the particular quantities where we think the system has reached a steady state. 
+
+Defect annihilation to this large extent is only found below a specific shear rate.
+A prerequisite seems to be that the defects can smootly approach each other and cancel, which is not possible above the instability as the lamellae buckle similar to \cite{Gonnella98}, show a tendency to trip over themselves which results in the creation of new defects.
+The critical value depends on the segregation and differs in systems A-2D and B-2D.
+It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
+Fig.
+\ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$.
+The runs with the two highest shear rates were started from the equilibrium configuration, whereas all others were pre-sheared using the final configuration of the preceding run as initial configuration in order to save computational rexources.
 The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
-Interestingly once the shear rate has been decreased to the next lower rate, both systems A and B show an trend to anneal defects instantaneously or after a short latency time if the shear rate is rather low.
-If the shear rate is turned up again, the lamellae buckle and the structure begins to trip over itself, once again creating new defects.
-\begin{figure}[!]
+Interestingly, while stepping down in shear rate both systems A-2D and B-2D anneal defects almost instantaneously.
+If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur immediately.
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
-\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
-Horizontal lines indicate averages over time for the particular quantities, where we assume the system has reached a steady state.
-The inset gives data for the 3D-system C and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
+Horizontal lines indicate time averages of the particular quantities, where we assume the system has reached a steady state.
+The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
 \label{fig6}
 \end{figure}
-This suggests that there is a dynamical equilibrium between defect creation and annihilation above a critical shear rate, while below the critical value defects get not anymore widely generated and the system anneals its defects and forms properly aligned lamellae.
-In order to test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
-The defect density goes indeed up (not shown) and reaches similar values to those shown in Fig. \ref{fig6}.
-However it takes some time before a steady state is reached as the applied shear rates are rather low.\\
-Fig \ref{fig7} shows the average total shear stress density $\langle \sigma_{xy}\rangle$, i.e. the total shear stress per unit lattice volume.
-In the following we will drop the backets $\langle\dots\rangle$ for simplicity and write $\langle \sigma_{xy}\rangle\equiv  \sigma_{xy}$.
-While for the highest applied shear rate the hydrodynamic part forms the largest contribution to the total stress density, a regime where the chemical stress dominates is entered on decreasing the shear rate.
-At low shear rates stress fluctuations become more stretched in time and get smaller, what is slightly hidden by the logarithmic scale of Fig. \ref{fig7}
-Their magnitude is about three times smaller in the larger simulation shown in Fig. \ref{fig1}, \ref{fig2}.
-Fortunately we found that the average shear stress density over time is to a good degree independent of the system size and therefore it is justified to simulate longer times by using a smaller system size without flawing the results.
-\begin{figure}[!]
+
+This finding is in further support of a dynamical equilibrium between defect creation and annihilation above a critical threshold, wheras below the critical value defects cancel and lamellae align.
+To test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
+The defect density goes indeed up (not shown) and reaches similar values as before.\\
+Fig \ref{fig7} shows the ensemble averaged total shear stress $\langle \sigma_{xy}\rangle \equiv\sigma_{xy} $, i.e. the total shear stress per unit lattice volume.
+At low shear rates stress fluctuations become actually more stretched in time and smaller, a fact that is obscured by the logarithmic scales.
+In larger simulations such as those shown in Fig. \ref{fig1} and Fig. \ref{fig2} they are also about three times smaller.
+Nevertheless we found that the averaged shear stress is to a very good degree independent of the system size, which makes it possible to simulate longer times using smaller system sizes without flawing the results.
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
-\caption{Average total shear stress $\sigma_{xy}$ versus time for system A: The horizontal lines are steady state time averages}
+\caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
+ }
 \label{fig7}
 \end{figure}
-Furthermore the stress fluctuations represent a physical aspect as they can be related to the inhomogeneities of the order parameter.
-System A shows clear shear thinning behaviour and seems to obey a power law (see Fig. \ref{fig11}).
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the time averaged apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
-Stress fluctuations cause temporary variations of these mean values of up to a factor 3 at the lowest shear rate.
-In contrast to system A system B does not exhibit a clear monotonous shear thinning tendency.
-The apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow more a horizontal trend when the shear rate is decreased.
-The stress fluctuations and time averaged shear stress densities are slightly smaller than in the stiffer system A with one exception.
-For the highest shear rate $\gd=2\e{-4}$ the shear stress turns out to be larger than in system A.\\
-The deviatoric part of the average chemical pressure density $\langle P_{xy}\rangle\equiv P_{xy}$ for system A is shown in the main picture of Fig. \ref{fig8}.
+
+System A-2D shows clear shear thinning behaviour and seems to obey a power law (see below Fig. \ref{fig11}).
+Values of the apparent viscosity are given in the caption of Fig. \ref{fig7}.
+Contrary to system A-2D system B-2D does not exhibit a clear tendency of shear thinning.
+Here, the apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow a more horizontal trend.\\
+We want to focus now on the deviatoric part of the chemical pressure $\langle P^{chem}_{xy}\rangle\equiv P^{chem}_{xy}$.
+This is shown for system A-2D in Fig. \ref{fig8}.
 The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
-System B shows a very steep decrease of $P_{xy}$ during the simulation with $\gd=1\e{-4}$, which is because of the defect annihilation.\\
-\begin{figure}[!]
+In System B-2D $P^{chem}_{xy}$ decreases sharply during the run with $\gd=1\e{-4}$ (not shown) because at this shear rate pronounced defect annihilation sets in.\\
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A.
-The inset shows data for the 3D-system (system C) and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
 \label{fig8}
 \end{figure}
-Fig. \ref{fig9} illustrates the relation between $P_{xy}$ and the defect density $\rho_D$.
-The black line represents a fit to the system A, which shows a linear dependence almost down to the lowest shear rate.
-The result for the softer system B is slightly more obscured as a consequence of the ordering and the tendency to anneal defects.
-Our quantifying protocol is based on averaging over a larger number of defects sites and is prone to inaccuracy in a regime where defects are rather scarce.
-\begin{figure}[!]
+
+Fig. \ref{fig9} illustrates dependence of $P_{xy}$ on the defect density $\rho_D$.
+System A-2D features a linear relation between both quantities that breaks only down at the lowest defect density or rather shear rate.
+The result for the softer system B-2D are less clear due to the ordering and annihilation of defects.
+It is very likely that our quantifying procedure, which is based on averaging over a larger number of defects sites,is prone to inaccuracy in this regime.
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs defect density for system A ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)} 
+\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs. defect density for system A-2D and B-2D } 
 \label{fig9}
 \end{figure}
-The main picture in Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A.
-At first until $\gd=5\e{-5}$ the first normal stress differences decrease monotonously with decreasing shear rate.
-At lower shear rates, however, the behaviour is far from being monotonous and $N_1$ rises again.
-This makes it difficult or even impossible to define reasonable time averages as $N_1$ may still evolve at the end of the run.
-\begin{figure}[!]
+
+Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A-2D, where we followed the same protocol as before.
+Initially, down till $\gd=5\e{-5}$ $N_1$ decreases monotonously with decreasing shear rate.
+However, at lower shear rates the behaviour is not monotonous anymore and $N_1$ rises again.
+Hence, it is difficult to define a reasonable time average for $N_1$ as for these low shear rates the system may still evolve at the end of the run.
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
-\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{xx}-\sigma_{yy}\rangle$ vs time for system A1.
-The inset shows data for the 3D-system (system C) and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{yy}-\sigma_{xx}\rangle$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
 \label{fig10}
 \end{figure}
-This feature is actually less erratic than it seems at first sight and can be rationalised. 
-A look at Fig. \ref{fig6} reveals that $N_1$ rises in the same way as the defect density drops.
-A pair of defects can cancel if one of them is an endpoint of a black lamella surrounded by a branching white lamella and the other one is a white endpoint surrounded by a branching black lamella.
-When these two defects anneal the lamella spacing increases locally by half a lamellar width.
-The interfaces have to expand in y-direction as the system is not frustrated enough to create an entire new layer in order to get the lamellar width back closer to its equilibrium value.
-Hence the normal stresses $\sigma_{yy}$ become negative and consecutively smaller, leading to gradually larger values of $N_1$.
-It should be stressed that only $\sigma_{yy}$ seems to evolve anomalously slowly and all other components of the stress tensor reach state state values more rapidly.
-System B again behaves different in the sense that $N_1$ continues to fall for lower shear rates despite there is also ongoing annealing of defects.
-This difference is a direct effect of the softness of system B, whose smaller Landau coefficients prevent the normal stresses $\sigma_{yy}$ from taking up very negative values and $N_1$ from rising.
-Nevertheless close to the end of the simulations with the two lowest shear rates $N_1$ is still slightly changing which suggests that this slowly evolving behaviour is also due to ongoing defect annihilation.
-Because both trends in system A and B are consequences of the impossibility to optimize the lamellar width, larger simulations should lead to more defined steady state values for $N_1$.
-However, we were not able to check this as this proves computationally cosly.\\
-In Fig. \ref{fig11} we show flowcurves derived from time averages of $\sigma_xy$ and $N_1$ as a function of shear rate.
-The time average of shear stress density in systems A follow a power law $\sigma_{xy}=\dot{\gamma}^m$ with an exponent $m\simeq 0.87$ and gives evidence for shear thinning behaviour.
-In case of system B the best fit for the lowest three shear rate is obtained for an exponent $m=1$, indicationg more or less constant viscosity and no shear thinning.
-It is not very surprising as the chemical contribution to the total stress almost vanishes for shear rates $\dot{\gamma}\le1\e{-4}$ due to the significant decrease of the number of defects.
-The observed power law behaviour is simply that of the hydrodynamic part of the stress, which is linear in our case.\\
-A comparison with experimental work remains difficult because the topology of defects differ in two and three dimensions.
-\begin{figure}[!]
+
+This feature which appears somewhat erratic at first has actually a simple explanation. 
+The time sequence of the defect density in Fig. \ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$ and linked to the defect annihilation.
+When two defects meet and cancel the lamellar width increases locally by a certain amount.
+This causes frustration in the system as the incommensurability between the layers is not large enough to create a new lamellar layer and allow the system to snap back towards its equilibrium width.
+Hence the normal stresses $\sigma_{xx}$ become increasingly smaller and may even attain negative values, leading to gradually larger values of $N_1$.
+This explanation is further sustained by the fact that only $\sigma_{xx}$ seems to evolve anomalously slowly and all other components of the stress tensor reach their steady state values more quickly.
+We expect that these anomalous trends become considerably smaller in larger systems.
+System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
+This is likely to be a direct consequence of the lower segregation in system B-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
+In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
+System A-2D gives rise to shear thinning with the average shear stress of systems A-2D following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
+For system B-2D the best fit is actually obtained for an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning.
+This is because due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$ the chemical stress $P^{chem}_{xy}$ almost vanishes and the total stress is dominated by the hydrodynamic part.\\
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
-\caption{Flowcurves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)}
+\caption{Flow curves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)}
 \label{fig11}
 \end{figure}
-Our two-dimensional system is translational invariant in the third dimension.
-So every branch point or endpoint of a lamella has the topology of an edge defect in the 3D-counterpart of the structure.
-Moreover screw defects do not exist in our picture as they can only occur in more than two dimensions.
-A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
-Interestingly the authors identify exactly the motion of screw defects as the key mechanism responsible for the layer tilting and defect formation. 
-For the power law dependence of the shear stress on the shear rate they find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.\\
-Contrary to shear stresses first normal stress differences do not level off because of ongoing defect annihilation.
-Neither system A nor B is in a steady state at the end of the simulation and therefore the data for $N_1$ is only shown for the sake of completeness.
 
@@
 
-The computational costs of a detailed analysis of the nonlinear rheology in three dimensions are still substantial as for steady state quantities it is necessary to simulate large strains.
-Therefore we performed only a limited number of simulations for selected parameters. 
-\begin{figure}[!]
+Because steady states require to account for large strains, the computational costs of a similar analysis of nonlinear rheology in three-dimensions are still prohibitive.
+In the following we compare 3D-systems with 2D-systems smaller than those of the previous paragraph but large enough to show qualitatively similar structure under flow.
+
+\begin{figure*}[htp!]
 \centering
-\includegraphics[angle=0,width=0.45\textwidth]{phi_run786_320.jpg}
-\includegraphics[angle=0,width=0.45\textwidth]{phi_run788_960.jpg}
-\includegraphics[angle=0,width=0.45\textwidth]{phi_run792_960.jpg}
-\caption{Order parameter $\phi$ in two and three dimensions: The top and centre picture show system C after equilibration at timestep $t=3.2\e{5}$ and in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at $t=9.6\e{5}$, where shear-induced layering takes place.
-The image at the bottom depicts the corresponding two dimensional system A of the same size in flow and velocity gradient direction.}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1002_320k.png}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1002_640k.png}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1001_640k.png}
+\includegraphics[angle=0,width=0.45\textwidth]{phi_run1003_960k.png}
+\caption{Order parameter $\phi$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\cdot10^{-4}$) at $t=6.4\e{5}$.
+The opacity of one phase has been reduced to enable a better view into the lamellar structure.
+The bottom left picture shows System C-2D at the same time step and shear rate.
+The bottom right picture is system A-2D at $\dot{\gamma}=1\cdot10^{-4}$ and time step $t=9.6\e{5}$.
+Results for A-3D are morphologically identical to those above.}
 \label{fig12}
-\end{figure}
-For our three dimensional simulations, in the following referred to as system C, we used parameters identical to those of the stiffer 2D-system A, but reduced the system size to $N_x\times N_y \times N_z=128\times256\times128$ lattice sites.
-The directions of vorticity, flow and velocity gradient are now x, y and z, respectively.\\ 
-The top picture in Fig.
-\ref{fig12} shows the order parameter $\phi$ after equilibration at time step $t=3.2\e{5}$ prior to shearing, which served as initial configuration for the runs with shear rate $\gd=1\e{-4}$ and $2\e{-4}$.
-A corresponding configuration, which was equilibrated for another $3.2\e{5}$ timesteps was used for the run with $\gd=\e{-5}$.
-As before black and white correspond to $\phi=-1$ and $\phi=1$, repectively, whereas gray scales give intermediate values of $\phi$.
-The opacity of $phi$ has been linearly reduced from its full value at $\phi=-1$ to 0 at $\phi=1$ in order to reveal the lamellar structure close to the  boundaries.
-In comparison to the 2D-case the isosurfaces of the cuboid, taken as quasi-two-dimensional equivalent in 3D, exhibit a slightly different morphology with shorter protrusions and an overall pattern, which appears somewhat more irregular.
-Nonetheless the quiescent and equilibrated system C adopts the same lamellar spacing of about 10 lattice sites as system A.Fig.
-\ref{fig13} shows a slice through $C({\bm k})$ along $k_y=0$ (cuts along $k_x,k_z=0$ look identical), which resembles the left picture in Fig.
-\ref{fig4} and differs only in the magniude.\\
-However, in shear flow system A and C behave very differently.
-Even for the highest shear rate we applied and contrary to the 2D-case system C orders readily into a regular lamellar structure.
-The centre picture of Fig.\ref{fig12} shows an example at timestep $t=9.6\e{5}$ and shear rate $\gd=1\e{-4}$, which we picked as this was the least ordered one of all three runs.
-With a total number of 13 layers the lamellar spacing is still close to the equilibrium value of roughly 10 lattice sites.
-Besides some irregularities at the front boundary there are obviously edge defects, caused by a branching of the lamellae and visible at the very top and bottom in the foreground and more towards the centre in the background.
-\begin{figure}[!]
+\end{figure*}
+
+The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which has a slightly larger segregation than system B.
+The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $128\times256\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
+Two snapshots of system C-2D and A-2D in shear flow are shown in the bottom pictures in Fig. \ref{fig12}.
+They feature the typical fluctuating pattern of breaking and merging lamellae that was described in the previous paragraph and which is particularly pronounced in the case of system C-2D.
+The top picture visualise results for system C-3D with the same free energy parameters as C-2D.
+The quiescent state of system C-3D after equilibration is shown at the left with a morphology virtually identical to that of A-3D.
+The lamellar width of roughly 10 lattice sites is more or less the same as in two dimensions.
+A slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig. \ref{fig4}.
+However, in shear flow considerable differences arise between the two- and three-dimensional case.
+Contrary to the result in 2D, the three dimensional systems order readily into a regular lamellar structure after a transient period as shown in Fig. \ref{fig12} for system C-3D. 
+We opted for this system as the differences between the 3D- and 2D-system are the most striking we observed.
+Interestingly, these findings are consistent for even the highest shear rates, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
+
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run786_320.jpg}
-\caption{Structure factor $C({\mathbf k})$ of quiescent system C, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
-Shown are wavevectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
+\caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
+Shown are wave vectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
 \label{fig13}
 \end{figure}
-\begin{figure*}[!]
+
+\begin{figure*}[htp!]
 \centering
@@
 \includegraphics[angle=0,width=0.33\textwidth]{ck_z-slice_run788_960.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system C in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wavevectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
 Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
@@
 \end{figure*}
-The edge defects emerge from the spontaneous shear-induced ordering of the phase separated structure but remain very localised at first.
-As a result of shear advection the interfaces spread and close gaps at later times to finally form either continuous lamellae or uniform edges, which are oriented along the flow, virtually stable and do not anneal.
-The fact that separated regions with similar order parameter values can now also connect along a third direction and expell regions with opposite order parameter, seems to play a key role pertaining to the formation of a homogenously layered structure.
-The bottom picture in Fig. \ref{fig12} gives an idea about how the dimesionality affects the morphology of layered smectics and binary fluids.
-It shows the two-dimensional system A as slice of system C along the direction of vorticity.
-The difference between both systems is striking and even more surprising as both results were obtained with the same parameter setting and identical equilibration and shear histories.
-The inset of Fig. \ref{fig6} quantifies the time it takes system C to order into a layererd defect-scarce structure.
-Within a few $10^5$ timesteps the defect density of system C decreases to magnitudes, which we obtained in the 2D-case only for shear rates about one decade lower and times about a factor 10 longer.\\
-Scarce defects lead to smaller chemical pressure, which is shown in the insets of Fig. \ref{fig8} for the component $P_{xy}$.
-The hydrodynamic part of the stress is by far dominating and shear thinning occurs to an ultimate degree where the apparent viscosity of the fluid is that of the solvent and the chemical contribution has practically vanished.
-A relation between the deviatoric chemical pressure tensor and the defect density like in Fig. \ref{fig9} can therefore not be established.
-The inset of Fig.\ref{fig10} shows first normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$.
-For all shear rates $\gd5\e{-5}-2\e{-4}$ system C takes similar values, which are monotonously approached and about one decade smaller than those for $N_1$ of system A.\\ 
-Finally we want to glimpse at the structure factor of system C in shear flow.
+
+Nevertheless, the order is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers.
+The Burgers vector is oriented along the gradient direction with edges along the flow direction.
+This forms a structure that is translational invariant along the flow direction and in the steady state the location of the edge defects does not change and the defects to not anneal.
+A closer look at the kinetic evolution of the order parameter reveals that regions separated by different values of order parameter connect along a third direction by expelling regions with opposite order parameter.
+This is a direct consequence of the additional dimension and seems to play a key role during the formation of homogeneously layered structure.\\
+It is instructive to quantify the time it takes system A-3D to emerge as lamellar ordered structure starting from an isotropic initial configuration.  
+The inset of Fig. \ref{fig6} gives the defect density as defined by Eq. \ref{defect density} versus time for shear rates $\dot{gamma}=5\cdot10^{-5}-2\cdot10^{-4}$. 
+Within a few $10^5$ timesteps the defect density decreases to magnitudes, which we obtained in case of system A-2D for shear rates about one decade lower.
+The total strain where the curves level off and ordering is achieved scales with the shear rate and is about $\gamma=1.3$ ($\dot{gamma}=2\cdot10^{-4}$), $\gamma=0.8$ ($\dot{gamma}=1\cdot10^{-4}$) and $\gamma=0.6$ ($\dot{gamma}=5\cdot10^{-5}$).
+ \\
+A lower number of defects yields smaller chemical pressure, which is shown for the deviatortic part $P_{xy}$ in the inset of Fig. \ref{fig8}.
+This contribution to the total stress is dwarfed by the hydrodynamic part and system A-3D exhibits shear thinning almost to an ultimate degree.
+A relation between the magnitude of the deviatoric part of the chemical pressure and the defect density like that in Fig. \ref{fig9} in the two-dimensional case cannot be deduced.\\
+First normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig10}.
+For all shear rates $\gd5\e{-5}-2\e{-4}$ system A-3D takes similar values, which are monotonously approached and about one decade smaller than those of system A-2D.\\ 
+Finally we want to glimpse at the structure factor of system A-3D in shear flow.
 Fig. \ref{fig14} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
 The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
-It shows data in flow-velocity-gradient plane and compares directly with the left picture in Fig. \ref{fig4}.
-The pictures in the centre and on the right give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
-Note the scale of the logarithmic color map in the centre picture.
-Despite a certain similarity with the 2D-structure after switch off in the rightmost picture of Fig.
-\ref{fig4}, the structure factor in three dimensions has a very pronounced peak at $k_z\simeq\pm0.2\pi/l$.
-It appears to be split into two parts, but this is an effect of the relatively sparse discretization and caused by the visualization software.
-Although strongly preaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
-Fig. \ref{fig15} depicts the structure factor in 3D, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
+It shows data in flow-velocity-gradient plane and can be directly compared with the left picture in Fig. \ref{fig4}.
+The other images give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
+Note that for the centre picture a logarithmic colour map had to be adopted due to the strong ordering in 3D.
+There is a certain similarity between the 2D-structure factor {\it after} switch-off (Fig.
+\ref{fig4}) and the 3D-structure, but the latter shows a pronounced peak at $k_z\simeq\pm0.2\pi/l$ also during shear flow, which is slightly concealed by the logarithmic scale.The split nature is an effect of the relatively sparse discretization and was caused by the visualisation software.
+Although strongly peaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
+Fig. \ref{fig15} depicts a close up of the 3D-structure factor, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
 It reveals a banana-shaped spatial dependence with some small appendices in flow direction at $k_x=0$.
 
-\begin{figure}[!]
+\begin{figure}[htp!]
 \centering
 \includegraphics[angle=0,width=0.45\textwidth]{ck_run788_960_zoom.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system C in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
 \label{fig15}
@@
 
-\begin{table}[htp]
-\small
-\caption{Simulation parameters: The systems were first equilibrated from spinodal decomposition at zero shear rate (EQ), followed by various shear rates, ranging from $\dot{\gamma}=2\cdot10^{-4}$ till $1.25\cdot10^{-5}$.
-For some runs a final switch off (SO) at zero shear rate was performed.
-The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
-  \label{tab1}
-%  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lllll}
-  \begin{tabular*}{0.495\textwidth}{lllll}
-\hline
-run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
-\hline
-system A & & & &\\
-\hline 
-703 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
-704 & $1024\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
-705 & $1024\times512\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
-\hline
-767 & $512\times256\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
-775 & $512\times256\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
-769 & $512\times256\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
-771 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
-772 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
-774 & $512\times256\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
-\hline
-792 & $256\times128\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
-system B & & & &\\
-\hline
-707 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
-710 & $1024\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
-765 & $1024\times512\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
-\hline
-776 & $512\times256\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
-777 & $512\times256\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
-778 & $512\times256\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
-780 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
-781 & $512\times256\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
-782 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
-\hline
-system C & & & &\\
-\hline
-786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
-787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
-788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
-789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
-\hline
-\end{tabular*}
-\end{table}
+\section{Discussion and Conclusions}
 
-\section{Conclusions}
 We have performed large scale simulations of a lamellar system in steady shear flow  in two and three dimensions.
@@
 
+A comparison of these results with experimental work can be only limited because the topology is different in two and three dimensions.
+Our 2D-systems correspond to are translational invariant along the third dimension.
+Every branching point or endpoint of a lamella has the topology of an edge defect in the 3D-continuation of the structure.
+A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
+Interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, floding  and formation of defects. 
+For the power law dependence of the shear stress on the shear rate Lu et al.
+find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
+Our value of $m=0.87$ appears to be slightly too large.
+A possible explanation for this is that screw defects, which do not exist in our 2D-picture, play a key role for this theory.\\
+
 \section{Acknowledgements}
+We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time at the Argonne Leadership Computing Facility, Argonne, U.S.A. as well as on HECToR, U.K.
+We like to thank A. Morozov for useful discussions. 
+M.E.C. holds a Royal Society Research Professorship.
+
+\section{Appendix: Sliding Periodic Boundary Conditions}
+
+The simulations of shear flow are undertaken using a mixed finite
+difference / lattice Boltzmann approach, where the former is used
+to advance the order parameter in time via the Cahn-Hilliard equation,
+and the later is used to solve the Navier-Stokes equations.
+Coupling between the two is
+via the velocity field, which advects the order parameter, and
+the thermodynamic pressure tensor (6), the divergence of which
+contributes a body force on the fluid. The sliding periodic
+boundary conditions of Lees and Edwards\cite{leesedwards} may be
+applied to the lattice Boltzmann part of the calculation
+\cite{Wagner02,Adhikari05} to generate shear in the system. We
+describe here how the sliding periodic boundaries are extended to
+include the finite difference part of the calculation.
+
+We consider a finite difference mesh where the order parameter
+values are co-located with the density and velocity fields provided
+by lattice Boltzmann. Away from the sliding periodic boundaries, the
+finite difference discretisation is as one would expect.
+In order to ensure conservation of order parameter, as expressed by
+\begin{equation}
+\partial_t \phi + \partial_\alpha (u_\alpha \phi - M\partial_\alpha \mu) = 0,
+\end{equation}
+we consider each lattice point to be surrounded by a cubic control volume or
+cell
+of dimensions $\Delta x = 1$ in two or three dimensions.
+We compute convective and diffusive fluxes at the cell boundaries using
+linear interpolation of the normal velocity component to the cell face.
+For the advective fluxes, the interfacial values of the order parameter
+are computed via cubic interpolation weighted in the upwind direction.
+This is combined with an Euler forward step for the time integration:
+a small Courant number ($u\Delta t/\Delta x$), and hence numerically
+stability, is guaranteed by the small Mach number constraint
+standard in lattice Boltzmann. The gradient of the chemical potential
+is computed at each cell face using values of $\mu$ computed at the
+lattice point either side.
+
+A body force on the fluid in the Navier-Stokes equations may be
+included in the lattice Boltzmann approach. Again, to ensure conservation
+of momentum, the divergence of the thermodynamic pressure tensor is
+computed based on cell-face values. In practice, this requires a
+finite difference stencil
+that entends at least $\pm 3$ lattice points in each coordinate direction
+(the thermodynamic pressure involves a fourth derivative of the order
+parameter).
+
+The sliding periodic boundaries lie (conceptually) along plane
+co-incident with the cell faces. Where the finite difference stencil 
+overlaps a such a plane, the relative movement between the two sides
+of the plane must be accounted for when computing derivatives of the
+order parameter, and hence fluxes of order parameter and momentum,
+near the planes. This means interpolating values of the order parameter
+on each side of the plane to positions corresponding to the lattice
+locations on the other side so that the finite difference stencil
+can be used normally. For this study, where there are high derivatives
+of the order parameter in the free energy, it is important that this
+interpolation is at least cubic. Linear interpolation of the order
+parameter leads to unphysical oscillations of physical quantities
+such as the shear stress related to the magnitude of the velocity jump
+across the sliding boundary.
+With cubic interpolation, such artefacts are absent. The non-linear
+nature of this interpolation means that fluxes of both order parameter
+and momentum computed at the plane do not exactly match on each side.
+However, this may be corrected at each sliding boundary at each time
+step, restoring global order parameter and momentum conservation.
 

### 2011-10-14 trunk/papers/smectics/smectics.tex
smectics introduction (with line breaks)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1461 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \section{Introduction}
-Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurt etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid.
-Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibres constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character.
-Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in bio-sensing.
 
-Rheology gives a powerful way to characterise macroscopic physical properties of a soft material such as its viscosity, elastic and loss moduli, which quantify, for instance, its rigidity, response to applied forces or stresses etc.
-While in a Newtonian, viscous fluid, shearing gives rise to a stress which is linearly proportional to the applied shear rate -- the slope being the viscosity -- in complex fluids flow is often non-Newtonian and such a simple linear relation no longer holds.
-Thus, viscoelastic complex fluids may be shear thickening or shear thinning, if their effective viscosity increases or decreases with applied shear.
-They may even spontaneously phase separate when flowing, giving rise to shear bands which are familiar in the rheology of liquid crystals in the vicinity of the isotropic-nematic transition, and in wormlike micelles.
+Instances of soft complex fluids abound. Examples include colloidal
+suspensions, where particles or droplets are suspended in a viscous
+fluid, living tissues, and liquid crystalline fluids. Well known
+examples of colloidal suspensions include paint, cosmetics, and
+food products such as ketchup, mayonnaise, and yogurt.
+The intracellular network of actin fibres constituting the
+cytoskeleton of mammalian cells is a polymeric fluid.
+The same is true of the DNA making up the bacterial nucleoid and
+eukaryotic chromosomes, and the viral genome of bacteriophages,
+where enhanced confinement imparts a strong liquid crystalline
+character to the system.
+Liquid crystalline fluids themselves, with or without colloidal
+inclusions, are of of wide technological importance in devices
+such as electronic displays and sensors.
 
-The rheological properties of a non-Newtonian fluid are often linked to the microstructure of the material.
-For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate.
-Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals.
-While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1.
-Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics.
-Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via pre-shearing, or not.
-In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample.
-An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state.
-This leads to increased dissipation and hence to a larger viscosity.
-A well-known example of permeation flow is the one which occurs e.g.
-in cholesterics which are sheared or pushed along the direction of the cholesteric helix.
-However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
-Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve.
-For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches.
-Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.
+The behaviour under flow, or rheology, of such materials is important
+to understand, as it affects many industrial and manufacturing processes.
+At the same time, rheological measurements provide a powerful way to
+characterise macroscopic properties of a soft material.
+The response to applied forces or stresses can be used to quantify,
+e.g., the viscosity, or elastic and loss moduli of soft materials.
+On shearing, a simple Newtonian fluid exhibits a stress which is
+linearly proportional to the applied shear rate; the
+constant of proportionality is just the viscosity. In marked
+contrast, complex fluids flow is often non-Newtonian and such a simple
+linear relation no longer holds. On shearing, the effective viscosity
+of a complex viscoelastic fluid may increase or decrease, corresponding
+to shear thinkening or shear thinning behaviour. Complex fluids may even
+spontaneously phase separate when flowing, giving rise to shear bands.
+These are familiar in the rheology of liquid crystals in the vicinity of
+the isotropic-nematic transition, and in wormlike micelles.
 
-Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically.
-This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations.
-At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally.
-In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear.
-For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
-We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters.
-In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like micelles, surfactants, smectic liquid crystals and block-copolymer.
+The rheological properties of a non-Newtonian fluid are typically linked
+to the microstructure of the material. Particularly relevant to our
+current work is the rheology of layered systems, such as smectics and
+cholesterics. Here, experiments may show different rheological results
+depending on whether the system has been ordered prior to measurement
+(for example, by pre-shearing).
+For instance, in the flow of polydomain liquid crystals, which are not
+uniformly ordered, the existence of regions with conflicting director
+patterns leads to a non-linear relation between shear stress and
+applied shear rate.
+Another example of strongly non-Newtonian rheology is region I flow
+in polymeric nematic liquid crystals. While our understanding of the
+physics of region I flow is far from complete, it is thought that a
+defect network may form and lead to much enhanced viscosity, with
+marked shear thinning. In both these cases, stiffening is usually
+associated to the presence of defects in the sample.
+
+Another intriguing type of flow which is possible in layered systems
+is permeation,
+where individual molecules in the fluid translate while keeping the
+defect pattern intact in steady state. This leads to increased dissipation
+and hence to a larger viscosity. A well-known example of permeation flow
+is found in cholesterics which are sheared or pushed along the direction
+of the cholesteric helix. However, one would also expect permeation flows
+to be possible when a non-translationally invariant defect pattern is
+present in the system; this might be the case in quenched smectics, for
+instance. In addition to their non-linear shear-stress curve, lamellar
+systems are also of interest for a number of other phenomena.
+In block-copolymers one may observe, under some conditions, shear-induced
+phase transitions from the microphase separated state to ordered phases
+like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. It would be
+desirable to extend our theoretical understanding of such switches.
+Further, lyotropic layered liquid crystals have sometimes been observed
+to form an onion phase \cite{Panizza96,Iwashita07} which can happen either
+spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under
+shear \cite{Diat93}. The mechanism underlying such pattern formation has
+remained elusive to date.
+
+Despite a wealth of experimental results, and the importance of rheology
+to many commercial applications, theoretical progress in understanding
+flow behaviour of complex fluids has remained a challenge. This is largely
+because of the non-linear nature of the equations of motion, and the
+coupling of the flow to different microstructures. These either require
+drastic approximations or, for practical purposes, resort to computer
+simulation. Even here, realistic 3-dimensional calculations are
+computationally expensive.
+In this work, our focus is the physics of smectics under shear.
+While sheared smectics have been studied extensively via simulations
+in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
+computational constraints have meant almost all these results are limited
+to two dimensions. Further, a number of these studies have used solid walls
+to drive shear in the hydrodymamics and, inconsistently, sliding periodic
+boundaries for the composition variable. Such simulations have a number of
+drawbacks: solid walls to drive shear limits the attanable shear rate in
+larger systems and so exacerbates finite size effects; mixed boundary
+conditions make it difficult to disentangle bulk and boundary effects
+in the rheology. In this work, we try to overcome all of these issues.
+(One notable example which does tackle a large 3-dimensional simulation
+of a complex fluid under shear involved gyroid phases~\cite{Saksena09}.)
+
+We describe smectics via a Brazovskii free energy functional
+\cite{Brazovskii75}, which provides a robust framework to study systems
+like micelles, surfactants, smectic liquid crystals and block-copolymers.
+The governing equations in the model are a convection-diffusion equation
+for the scalar order parameter, which represents the density difference
+of two fluid components, and the Navier-Stokes equation describing
+hydrodynamics. A stress tensor that contains extra contributions owing
+to the presence of interfaces between the components is included.
+These dynamical equations are solved using a combination of a finite
+difference scheme for the convection-diffusion equation, and the
+lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation.
+We use consistent sliding periodic (Lees-Edwards) boundary conditions for
+both convection-diffusion and Navier-Stokes equations. This allows us to
+probe the true bulk rheology of the smectic phase for large systems and
+high shear rates, as quantified by an appropriate Peclet number.
+
+% KS: this can be moved to existing appendix
+%The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
+%The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.
 
@@
 
-The governing equations in our model are a convection-diffusion equation for the scalar order parameter, which represents the density difference of the two phases and the Navier-Stokes equation with a stress tensor that contains extra contributions due to the presence of the lamellar interfaces.
-In our approach the dynamical equations are solved using a combination of finite difference scheme for the convection-diffusion equation and the lattice Boltzmann method (LBM) \cite{Succi} for the Navier-Stokes equation.
-The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
-The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.
+% KS: this following paragraph doesn't quite make sense to me yet.
 
-Our main finding is that as anticipated above depending on system parameters we find that a smectic phase may lead to vastly different rheological responses. 
-When thermodynamic forces are weaker with respect to viscous, flow-induced, ones, shear quickly orders the system into a lamellar phase, with only a few defects.
-The rheology in this regime is Newtonian, and the shear stress is linear in the applied shear rate.
+Our main finding is that, depending on system parameters, a smectic phase
+shows markedly different rheological responses. At lower Peclet
+numbers, where thermodynamic effects are relatively strong compared with
+flow effects,  shear quickly orders the system into a lamellar phase,
+with only a few defects. The rheology in this regime is Newtonian, and
+the shear stress is linear in the applied shear rate.
 If, on the other hand, thermodynamic forces are larger, we observe that shear actually leads to the nucleation of new defects, and the rheology is highly nonlinear, with the shear rate proportional to $\dot{\gamma}^{\alpha}$, with $\alpha\simeq 0.87$, leading to a shear thinning behaviour.
@@
 More research and parameter steering will be needed to make this observation more quantitative.
-We note that sheared smectics have been studied extensively via simulations in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}, but because of computational constraints almost all results were limited to two dimensions and smaller strains than the ones considered here.
-Furthermore in this work we use Lees-Edwards boundary conditions which allow us to probe the true bulk rheology of the system, at variance with previous work which employed solid walls and different boundary conditions, which introduce finite size and boundary effects into the physics.
 
-This paper is organised as follows. In Sec.II we introduce the model, including its equilibrium properties, the full stress tensor and the dynamical equations.
-We briefly sketch how we derive structural information from the order parameter and describe our defect quantification procedure, which measures the local alignment of the lamellar structure.
-We present results of the two dimensional case in In Sec.III, discuss the morphology, average and local shear stress densities and first normal stress differences and their relation to the defect structure.
+This paper is organised as follows. In Section II we introduce the model,
+including its equilibrium properties, the full stress tensor and the
+dynamical equations. We briefly sketch how we derive structural information
+from the order parameter and describe our defect quantification procedure,
+which measures the local alignment of the lamellar structure.
+We present results of the two-dimensional case in in Section III and
+discuss the morphology, average and local shear stress densities and
+first normal stress differences and their relation to the defect structure.
 We also show structure factors for various flow states and flow curves.
-In Sec. IV we compare our 2D-results with results in three dimensions. 
-A summary with conclusions closes the paper in Sec.V.
+We also compare our two-dimensional results with results in three dimensions. 
+A summary with conclusions closes the paper in Section V.
 

### 2011-10-17 trunk/papers/smectics/smectics.tex
Conclusions and references edited
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1464 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-Instances of soft complex fluids abound. Examples include colloidal
-suspensions, where particles or droplets are suspended in a viscous
+Instances of soft complex fluids are abound in our everyday world. 
+Examples include colloidal suspensions, 
+where particles or droplets are suspended in a viscous
 fluid, living tissues, and liquid crystalline fluids. Well known
@@
 Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
-The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$, whereas the interfacial width is roughly $\sqrt{2|\kappa|/|a|}$. 
-This entails that by varying $a$ (and $b\sim |a|$) by about on decade, the segregation of the two phases varies by about a factor $3$.\\
+The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$. 
+By varying the Landau-parameters $a$ (and $b\sim |a|$) we change the distance from the critical point and the magnitude of the order-parameter.
+The segregation between both phases was found to dependent neither solely on $a,b$ nor only on $\kappa,c$. 
+It is rather the magnitude of all parameters, which has to be similar in order to produce a lamellar phase, that controls the level of segregation.\\
 The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.
@@
 For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
-For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
-More explicitly the alignment factor $X_i$ is given by
+For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average projection of the normalised order parameter gradients onto the gradient at a specific site.
+More explicitly the alignment function $X$ is given by
 
 \begin{equation}\label{alignment}
-X_i=\frac{1}{N_{nn}} \sum_{j=1,N_{nn}}\left(\frac{\nabla \phi_i \cdot \nabla \phi_j}{|\nabla\phi_i||\nabla\phi_j|}\right)^2. 
+X({\mathbf r})=\frac{1}{V'({\mathbf r})} \int_{V'({\mathbf r})} d{\mathbf r'} \left(\frac{\nabla \phi ({\mathbf r}) \cdot \nabla \phi ({\mathbf r'})}{|\nabla\phi ({\mathbf r}) ||\nabla\phi ({\mathbf r'}) |}\right)^2.
 \end{equation}
 
-$N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
-For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
+$V'$ is a small volume that contains $\mathbf r$, which in practice can be chosen to contain  and its nearest neighbours.
+For smooth alignment and small curvature of the lamellae $X$ is very close to unity.
 At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
@@
 We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
-Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment factor below the threshold.
+Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment function below the threshold.
 However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
-By normalising the number of defects sites with $X_i$ below the threshold a defect density can be defined:
+By normalising the number of defects sites with $X$ below the threshold a defect density can be defined:
 
 \begin{equation}\label{defect density}
-\rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
+\rho_D({\mathbf r})=\frac{1}{V} \int d{\mathbf r'} \theta(\xi-X(\mathbf r')).  
 \end{equation}
 
-The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
+The function $\theta$ is the Heaviside function and $V$ is the volume of the entire system.
 The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.
@@
 Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
-The system size was $N_x\times N_y\times N_z=512\time1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
-Phase segregation was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq.
+The system size was $N_x\times N_y\times N_z=512\times 1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
+The distance from the critical point was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq.
 \ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
-By this we obtained a slightly different stress response with the system with lower segregation level appearing somwhat sofer.
+By this we obtained a slightly different stress response with the system with lower order appearing somwhat sofer.
 The lamellar distance was identical in both cases.\\
@@
 Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
-In previous studies \cite{Kendon01,Xu03, Xu06b} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
+In previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
 We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
@@
 The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
-Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems with lower segregation have lower average order parameters and are generally closer to the critical point.
-Higher viscosities lead to shorter interfaces, whereas lower values typically invoke longer lamellae.
-This effect has been reported before \cite{Gonnella97}.
-It turned out to be more pronounced for low segregation, although not as quite to the same extent as in the above reference.
+Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems B has a lower average order parameter and is closer to the critical point.
+Higher viscosities lead to shorter interfaces, whereas lower values typically yield longer lamellae.
+This effect has been reported before \cite{Gonnella97} and turned out to be more pronounced for the softer system B-2D, although not as quite to the same extent as in the former case.
 In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
@@
 
+The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
+The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
+After a short transient phase during which the order parameter remains rather homogenous, expanded and compressed regions emerge with a lamellar spacing that deviates from the equilibrium value.
+This is depicted in the centre picture of Fig. \ref{fig1} for shear rate $\gd= 1\e{-4}$.
+The inhomogeneities remain present in the steady state, are advected by the flow and undergo continuously affine transformations.
+A buckling instability above a critical shear rate has been described earlier by Gonnella et al. \cite{Gonnella98} and Kumaran et al. \cite{Kumaran2001}.
+It was suggested that the density of defects defects get introduced into a lamellar liquid crystal due to dilation and annealing occurs due to application of shear.
+This would lead to irregularities in the sample a dynamical process where the steady state defect density is determined by balance between the rate of introduction and rate of annealing of defects.
+In a more recent paper Kumaran et al. state that they observe similar inhomogeneities only above a certain system size \cite{Kumaran2011}, which agrees well with our observation.
+We found that shear alignment is more likely to take place in smaller systems of $N_x\times N_y\times N_z=128\times 256\times 1$.
+However, this trend is not necessarily followed at higher shear rates, a fact that we will discuss later in connection with our results in three dimensions.\\
+Kumaran et al. also identified a second mechanism of defect creation that is due to layer compression and splitting.
+Although we have not checked this in detail, this is also very likely the case in our results as the lamellar pattern is strongly compressed and heterogenous as shown in Fig. \ref{fig1}.\\
+After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
+The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
+Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
+The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
+Fig. \ref{fig2} shows equivalent states for system B-2D, which is closer to the critical point.
+All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
+Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
+Another shear-induced structure in binary lamellar systems was reported in \cite{Xu06b}.
+The reported results differ from our observations in some aspects.
+Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
+Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
+These differences may be simply a direct consequence of the different boundary conditions that were applied.
+While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.\\
+To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
+It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
+This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\
+
 \begin{figure}[htp]
@@
 
-The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
-The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
-After a short transient phase during which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing begins to deviate from its well-defined equilibrium value.
-These regions are advected with the flow while they continuously undergo affine transformations.
-The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $\gd= 1\e{-4}$.
-In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
-However, the reported results differ from our observations in the following aspects.
-Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
-Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
-These differences may be simply a direct consequence of the different boundary conditions that were applied.
-While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.
+By introducing the alignment function $X$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
+Fig. \ref{fig3} shows an example of the time evolution of system A-2D at low shear rate.
+All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
+The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
+There are regions which show some sort of flow alignment and others which don't.
+The bottom picture of Fig. \ref{fig3} shows a later state with a total strain that has increased by another $10^{5}\%$. 
+Obviously many defects have annealed and the interfaces are well aligned with the flow direction.
 
-\begin{figure}[htp!]
+\begin{figure}[htp]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
+\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
+The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
+Sites where the alignment function $X$ was smaller than $0.8$ are highlighted red.}
+\label{fig3}
+\end{figure}
+
+Defect annihilation to this large extent is only found below a specific shear rate.
+A prerequisite seems to be that the defects can smootly approach each other and cancel, which is prevented by the buckling instability above a criticl shear rate.
+Then the system shows a tendency to trip over itself, which results in the creation of new defects.
+The critical value differs in systems A-2D and B-2D and is related to the magnitude of the order parameter and the distance from the critical point.
+It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
+
+\begin{figure}[htp]
 \centering
@@
 \caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
-\label{fig3}
+\label{fig4}
 \end{figure}
 
-\begin{figure}[htp!]
+In Fig. \ref{fig4} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
+The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
+In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
+Large positive values of both quantities coincide with regions where the lamellae are under compression.
+Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
+Despite being locall negative, the ensemble averages of these quantities are always positive.
+Very similar results with smaller peak magnitudes are obtained for system B-2D.
+
+\begin{figure}[htp]
 \centering
@@
 The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
-\label{fig4}
+\label{fig5}
 \end{figure}
 
-To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
-It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
-This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\ 
-After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
-The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
-Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
-The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
-Fig. \ref{fig2} shows equivalent states for system B-2D with lower segregation.
-All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
-Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
-In Fig. \ref{fig3} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
-The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
-In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
-Large positive values of both quantities coincide with regions where the lamellae are under compression.
-Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
-Despite being locall negative, the ensemble averages of these quantities are always positive.
-Very similar results with smaller peak magnitudes are obtained for system B-2D.
-Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
+Fig. \ref{fig5} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
 Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
@@
 If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
-The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.
-By introducing the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
-Fig. \ref{fig5} shows an example of the time evolution of system A-2D at low shear rate.
-All lattice sites with $X_i\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
-The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
-There are regions which show some sort of flow alignment and others which don't.
-The bottom picture of Fig. \ref{fig5} shows a later state with a total strain that has increased by another $10^{5}\%$. 
-Obviously many defects have annealed and the interfaces are well aligned with the flow direction.
+The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.\\
 
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
-\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
-The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
-Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
-\label{fig5}
-\end{figure}
-
-Defect annihilation to this large extent is only found below a specific shear rate.
-A prerequisite seems to be that the defects can smootly approach each other and cancel, which is not possible above the instability as the lamellae buckle similar to \cite{Gonnella98}, show a tendency to trip over themselves which results in the creation of new defects.
-The critical value depends on the segregation and differs in systems A-2D and B-2D.
-It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
 Fig.
@@
 
-\begin{figure}[htp!]
+\begin{figure}[htp]
 \centering
@@
 System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
-This is likely to be a direct consequence of the lower segregation in system B-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
+This is likely to be a direct consequence of the lower order parameter in system B-2D and its softness with respect to system A-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
 In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
@@
 
-The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which has a slightly larger segregation than system B.
+The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which is intermediate between A and B.
 The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $128\times256\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
@@
 The lamellar width of roughly 10 lattice sites is more or less the same as in two dimensions.
-A slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig. \ref{fig4}.
+A slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig. \ref{fig5}.
 However, in shear flow considerable differences arise between the two- and three-dimensional case.
@@
 Interestingly, these findings are consistent for even the highest shear rates, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
+This is consistent with the picture that layer undulation due to an instability in three-dimensional smectic liquid crystal layers emerges above a critical shear rates, which is well above those we have applied in the present study \cite{Guo2002,Soddemann2004, Stewart2009}.
 
@@
 \centering
-\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run786_320.jpg}
+\includegraphics[angle=0,width=0.40\textwidth]{ck_y-slice_run786_320.jpg}
 \caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
@@
 
-\begin{figure*}[htp!]
+\begin{figure}[htp!]
 \centering
-\includegraphics[angle=0,width=0.33\textwidth]{ck_x-slice_run788_960.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_y-slice_run788_960.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_z-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.35\textwidth]{ck_x-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.35\textwidth]{ck_z-slice_run788_960.jpg}
 \caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
@@
 \label{fig14}
-\end{figure*}
+\end{figure}
 
-Nevertheless, the order is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers.
+However, the order is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers.
 The Burgers vector is oriented along the gradient direction with edges along the flow direction.
@@
 The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
-It shows data in flow-velocity-gradient plane and can be directly compared with the left picture in Fig. \ref{fig4}.
+It shows data in flow-velocity-gradient plane and can be directly compared with the left picture in Fig. \ref{fig5}.
 The other images give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
@@
 There is a certain similarity between the 2D-structure factor {\it after} switch-off (Fig.
-\ref{fig4}) and the 3D-structure, but the latter shows a pronounced peak at $k_z\simeq\pm0.2\pi/l$ also during shear flow, which is slightly concealed by the logarithmic scale.The split nature is an effect of the relatively sparse discretization and was caused by the visualisation software.
+\ref{fig5}) and the 3D-structure, but the latter shows a pronounced peak at $k_z\simeq\pm0.2\pi/l$ also during shear flow, which is slightly concealed by the logarithmic scale.The split nature is an effect of the relatively sparse discretization and was caused by the visualisation software.
 Although strongly peaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
@@
 
-\section{Discussion and Conclusions}
-
-We have performed large scale simulations of a lamellar system in steady shear flow  in two and three dimensions.
-We were able to relate the morphology of the order parameter in the quiescent state to results from previous studies of the same system using similar parameters.\\
-In two dimensions and steady state simple shear flow at Peclet numbers ${\cal O}(10)$ we did not observe a homogenous lamellar structure for shear rates above a critical threshold, which depends on the exact parametrization of the free energy functional.
-The morphology was rather inhomogenous instead, exhibiting regions that are under pressure with lamellar spacings smaller than the equilibrium value and others that were stretched.
-We could relate the structure to typical values of rheological quantities.
-Areas under extensional flow coincide with locally negative or small positive shear stresses and first normal stress differences, whereas both go to positive peak values in the compressed domains.
-Below a critical shear rate the systems showed no tendency anymore to trip over itself and defects annealed as they could approach each other in adjacent lamellae.
-The characteristic time of this process was set by the shear rate and the total strain.
-The deviatoric part of the chemical pressure was correlated to the number of defect sites, whereas the latter was measured by means of an alignment factor, which quantified the local curvature of the lamellar interfaces.
-For the stiffer system A-2D we could verify a linear relation between the defect density and the average chemical contribution to the shear stress.\\ 
-In system A-2D first normal stress differences decreased as long as a dynamic equilibrium between defect creation an annihilation existed.
-Below the critical threshold this was no longer the case and pronounced defect annealing caused a rise in the first normal stress differences as the average lamellar spacing increased.
-We could establish a shear thinning regime for the more defect laden system A-2D.
-System B in contrary was so perfectly annealed in the steady state that virtually shear stresses were down at the hydrodynamic level and the relation between shear stress and shear rate was trivially linear.\\
-In three dimensions we simulated shear strains which have not been attained so far.
-Interestingly for the same parameters we observe striking differences between system A-2D in 2D and system A-3D in 3D.
-It seems that the dynamics of the defects in two and three dimensions is rather incommensurable.
-Even at the lowest applied shear rates the chemical stress has decayed to a level where the apparent viscosity was virtually identical to the solvent.
-Structure factor data confirmed the very regular layered ordering of the structure and a lamellar spacing close to the equilibrium value.
-
-A comparison of these results with experimental work can be only limited because the topology is different in two and three dimensions.
+A comparison of these results with experimental work is possbile although the topology is different in two and three dimensions.
 Our 2D-systems correspond to are translational invariant along the third dimension.
@@
 A possible explanation for this is that screw defects, which do not exist in our 2D-picture, play a key role for this theory.\\
+
+\section{Conclusions}
+
+We studied nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model for binary liquid-crystalline fluids, which is based on a Brazovskii-free-energy-functional.
+The lamellar structure of the order parameter in the quiescent state agrees well with previous results using the same model in two dimensions and similar parameters.
+When flowing the morphology of the order parameter depends crucially on the system size.
+At system sizes $N_x\times N_y\times N_z=128\times256\times 1$ the system showed a strong tendency to align with the shear flow at low shear rates $\gd\le5\e{-5}$.
+At higher shear rates and for larger system sizes of $N_x\times N_y\times N_z=256\times512\times 1$ and above heterogenous lamellar structures with strongly extended and compressed regions were present in the steady states.
+A similar dependence of these charactistic patterns on the system size has been recently reported in \cite{Kumaran2011}.
+The patterns occured in distinctive bands and were oriented along the direction of extension and perpendicular to the direction of compression.
+The closer the system was to the critical point the more this feature gave way to flow alignment.
+The extended regions with lamellar spacing larger than at equilibrium were local regions with either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
+Depending on the applied shear rate and in agreement with other studies we found two different dynamical regimes. 
+Below a critical value, which was basically controlled by the magnitude of the order parameter and the distance from the critical point, alignment in flow direction and annealing of defects to a very large degree was obtained.
+Beyond the critical shear rate a buckling instability occured.
+The systems continuously tripped over itself, which resulted in a dynamical equilibrium between defect annihilation and creation of new defects.
+The shear stress in system A-2D follwed a clear power law dependence on the strain rate like $\sigma_{xy}=\gamma^\alpha$ with $\alpha=0.87$, giving rise to shear-thinning behaviour.
+Interestingly, the magnitude of chemical contribution $P_{xy}$ to the shear stress could be directly related to the density of defects.
+In system A-2D first normal stress differences became smaller with decreasing shear rate as long as the system was above the buckling instability.
+Below the critical shear rate continued defect annealing caused a rise in the first normal stress differences due to incommensurabilities in the lamellar width.\\
+We simulated flow behaviour of the same systems in three dimensions for the first time and found striking differences between the two- and three-dimensional versions of systems A and C.
+The 3D-system showed none of the distictive bands along the extensional direction and no buckling instability was found in the range of applied shear rates.
+The structure, initially amorphous after spinodal decomposition, ordered into a lamellar array with edge defects oriented along the flow directions.
+This is in total contrast to the 2D-case, where numerous defects appeared that would correspond to edge defects that are oriented perpendicular to the flow in 3D.
+Our results agree with theoretical values and simulations that predict a layer undulation instability at shear rates about two orders of magnitude larger than the ones we have applied.
+This constitutes a surprising change in results due to the different dimensonality and topoloogy of the strucutre despite identical parametrisation of the model.
 

### 2011-10-18 papers/smectics/smectics.tex
2D-results update
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1465 40876441-3496-0410-a613-a2fca35b2ae2

@@
   \begin{@twocolumnfalse}
-\noindent\LARGE{\textbf{Large-scale Simulation of Layered Binary Fluids and Smectic Liquid Crystals in Shear Flow}}
+\noindent\LARGE{\textbf{Shear Flow of Lamellar Mesophases and Smectic Liquid Crystals}}
 \vspace{0.6cm}
@@
 \section{Introduction}
-Instances of soft complex fluids surrounding us abound: for instance paint, cosmetics and food products such as ketchup, mayonnaise, yogurt etc are all well-known examples of colloidal suspensions, where particles or droplets are suspended in a viscous fluid.
-Biological systems also include a large range of soft materials: for instance the intracellular network of actin fibres constituting the cytoskeleton of mammalian cells is a polymeric fluid, as is the DNA making up the bacterial nucleoid or eukaryotic chromosomes, or the viral genome of bacteriophages, where enhanced confinement imparts strong liquid crystalline character.
-Soft materials such as liquid crystalline fluids, with or without colloidal inclusions, are also of importance technologically, for instance when they are used to build up display devices or in bio-sensing.
 
-Rheology gives a powerful way to characterise macroscopic physical properties of a soft material such as its viscosity, elastic and loss moduli, which quantify, for instance, its rigidity, response to applied forces or stresses etc.
-While in a Newtonian, viscous fluid, shearing gives rise to a stress which is linearly proportional to the applied shear rate -- the slope being the viscosity -- in complex fluids flow is often non-Newtonian and such a simple linear relation no longer holds.
-Thus, viscoelastic complex fluids may be shear thickening or shear thinning, if their effective viscosity increases or decreases with applied shear.
-They may even spontaneously phase separate when flowing, giving rise to shear bands which are familiar in the rheology of liquid crystals in the vicinity of the isotropic-nematic transition, and in wormlike micelles.
+Instances of soft complex fluids are abound in our everyday world. 
+Examples include colloidal suspensions, 
+where particles or droplets are suspended in a viscous
+fluid, living tissues, and liquid crystalline fluids. Well known
+examples of colloidal suspensions include paint, cosmetics, and
+food products such as ketchup, mayonnaise, and yogurt.
+The intracellular network of actin fibres constituting the
+cytoskeleton of mammalian cells is a polymeric fluid.
+The same is true of the DNA making up the bacterial nucleoid and
+eukaryotic chromosomes, and the viral genome of bacteriophages,
+where enhanced confinement imparts a strong liquid crystalline
+character to the system.
+Liquid crystalline fluids themselves, with or without colloidal
+inclusions, are of of wide technological importance in devices
+such as electronic displays and sensors.
 
-The rheological properties of a non-Newtonian fluid are often linked to the microstructure of the material.
-For instance, in the flow of polydomain liquid crystals, which are not uniformly ordered, the existence of regions with conflicting director patterns leads to a non-linear relation between shear stress and applied shear rate.
-Another intriguing example of strongly non-Newtonian rheology is region I flow in polymeric nematic liquid crystals.
-While our understanding of the physics of region I flow is far from complete, it is thought that a defect network may form and lead to much enhanced viscosity, with marked shear thinning, $\eta \sim \dot{\gamma}^{\alpha}$, and an exponent $\alpha$ typically between 0 and 1.
-Particularly relevant to our current work is the rheology of layered systems, such as smectics and cholesterics.
-Experiments involving any of such soft materials typically lead to vastly different results, according to whether the system has been ordered prior to the rheological study, for instance via pre-shearing, or not.
-In the latter case the measured viscosities are much larger, and just as in polydomain or region I nematic flow, this stiffening is usually associated to the presence of defects in the sample.
-An intriguing flow mode which is possible in layered system is permeation, through which the individual molecules in the fluid displace keeping the defect pattern intact in steady state.
-This leads to increased dissipation and hence to a larger viscosity.
-A well-known example of permeation flow is the one which occurs e.g.
-in cholesterics which are sheared or pushed along the direction of the cholesteric helix.
-However, one would expect permeation flows to be possible also when a non-translationally invariant defect pattern is present in the system, so that it may occur to some extent also in quenched smectics for instance.
-Lamellar rheology is also of interest for a number of additional intriguing phenomenological observations beyond their nonlinear shear-stress curve.
-For instance, in block-copolymers one may observe, under some conditions, shear-induced phase transitions from the microphase separated state to ordered phases like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}, and it would be desirable to extend our theoretical understanding of such switches.
-Furthermore, lyotropic layered liquid crystals have sometimes been observed to form an onion phase \cite{Panizza96,Iwashita07} which can happen either spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under shear \cite{Diat93} -- the mechanism underlying such pattern formation has remained elusive to date.
+The behaviour under flow, or rheology, of such materials is important
+to understand, as it affects many industrial and manufacturing processes.
+At the same time, rheological measurements provide a powerful way to
+characterise macroscopic properties of a soft material.
+The response to applied forces or stresses can be used to quantify,
+e.g., the viscosity, or elastic and loss moduli of soft materials.
+On shearing, a simple Newtonian fluid exhibits a stress which is
+linearly proportional to the applied shear rate; the
+constant of proportionality is just the viscosity. In marked
+contrast, complex fluids flow is often non-Newtonian and such a simple
+linear relation no longer holds. On shearing, the effective viscosity
+of a complex viscoelastic fluid may increase or decrease, corresponding
+to shear thinkening or shear thinning behaviour. Complex fluids may even
+spontaneously phase separate when flowing, giving rise to shear bands.
+These are familiar in the rheology of liquid crystals in the vicinity of
+the isotropic-nematic transition, and in wormlike micelles.
 
-Despite the wealth of experimentally available results, and the importance of rheology to for instance industrial applications, flow behaviour remains challenging to study theoretically.
-This is mainly in view of the highly non-linear equations of motion which require either drastic approximations or challenging computer simulations.
-At the same time, realistic modelling eventually requires the simulation of a fully 3-dimensional flow, which is particularly expensive computationally.
-In this work, we aim to contribute to fill in the gap towards realistic large scale computational studies of complex fluids, and our focus is the physics of smectics under shear.
-For another example of large scale simulations for complex fluids, see Ref.~\cite{Saksena09} where sheared gyroid phases were simulated.
-We shall concentrate here on intermediate values of the shear rate for which, as we shall see, we may observe either an approximately linear or markedly non-linear behaviour of the system, according to the thermodynamic parameters.
-In our treatment, we describe smectics via a Brazovskii free energy functional \cite{Brazovskii75}, which provides a robust framework to study systems like micelles, surfactants, smectic liquid crystals and block-copolymer.
+The rheological properties of a non-Newtonian fluid are typically linked
+to the microstructure of the material. Particularly relevant to our
+current work is the rheology of layered systems, such as smectics and
+cholesterics. Here, experiments may show different rheological results
+depending on whether the system has been ordered prior to measurement
+(for example, by pre-shearing).
+For instance, in the flow of polydomain liquid crystals, which are not
+uniformly ordered, the existence of regions with conflicting director
+patterns leads to a non-linear relation between shear stress and
+applied shear rate.
+Another example of strongly non-Newtonian rheology is region I flow
+in polymeric nematic liquid crystals. While our understanding of the
+physics of region I flow is far from complete, it is thought that a
+defect network may form and lead to much enhanced viscosity, with
+marked shear thinning. In both these cases, stiffening is usually
+associated to the presence of defects in the sample.
+
+Another intriguing type of flow which is possible in layered systems
+is permeation,
+where individual molecules in the fluid translate while keeping the
+defect pattern intact in steady state. This leads to increased dissipation
+and hence to a larger viscosity. A well-known example of permeation flow
+is found in cholesterics which are sheared or pushed along the direction
+of the cholesteric helix. However, one would also expect permeation flows
+to be possible when a non-translationally invariant defect pattern is
+present in the system; this might be the case in quenched smectics, for
+instance. In addition to their non-linear shear-stress curve, lamellar
+systems are also of interest for a number of other phenomena.
+In block-copolymers one may observe, under some conditions, shear-induced
+phase transitions from the microphase separated state to ordered phases
+like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. It would be
+desirable to extend our theoretical understanding of such switches.
+Further, lyotropic layered liquid crystals have sometimes been observed
+to form an onion phase \cite{Panizza96,Iwashita07} which can happen either
+spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under
+shear \cite{Diat93}. The mechanism underlying such pattern formation has
+remained elusive to date.
+
+Despite a wealth of experimental results, and the importance of rheology
+to many commercial applications, theoretical progress in understanding
+flow behaviour of complex fluids has remained a challenge. This is largely
+because of the non-linear nature of the equations of motion, and the
+coupling of the flow to different microstructures. These either require
+drastic approximations or, for practical purposes, resort to computer
+simulation. Even here, realistic 3-dimensional calculations are
+computationally expensive.
+In this work, our focus is the physics of smectics under shear.
+While sheared smectics have been studied extensively via simulations
+in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
+computational constraints have meant almost all these results are limited
+to two dimensions. Further, a number of these studies have used solid walls
+to drive shear in the hydrodymamics and, inconsistently, sliding periodic
+boundaries for the composition variable. Such simulations have a number of
+drawbacks: solid walls to drive shear limits the attanable shear rate in
+larger systems and so exacerbates finite size effects; mixed boundary
+conditions make it difficult to disentangle bulk and boundary effects
+in the rheology. In this work, we try to overcome all of these issues.
+(One notable example which does tackle a large 3-dimensional simulation
+of a complex fluid under shear involved gyroid phases~\cite{Saksena09}.)
+
+We describe smectics via a Brazovskii free energy functional
+\cite{Brazovskii75}, which provides a robust framework to study systems
+like micelles, surfactants, smectic liquid crystals and block-copolymers.
+The governing equations in the model are a convection-diffusion equation
+for the scalar order parameter, which represents the density difference
+of two fluid components, and the Navier-Stokes equation describing
+hydrodynamics. A stress tensor that contains extra contributions owing
+to the presence of interfaces between the components is included.
+These dynamical equations are solved using a combination of a finite
+difference scheme for the convection-diffusion equation, and the
+lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation.
+We use consistent sliding periodic (Lees-Edwards) boundary conditions for
+both convection-diffusion and Navier-Stokes equations. This allows us to
+probe the true bulk rheology of the smectic phase for large systems and
+high shear rates, as quantified by an appropriate Peclet number.
+
+% KS: this can be moved to existing appendix
+%The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
+%The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.
 
@@
 
-Our discussion will be in two parts, beginning with simulations in two dimensions for a system A-2D and a slightly softer system B of size $N_x\times N_y=1024\times 512$ and $N_x \times N_y=512\times 256$ respectively for longtime runs.
-The stiffness of the system was controlled by modifying the Landau parameters $a$ and $b$ of the bulk terms in the free energy functional \ref{FE-functional} with the parameters $\kappa$ and $c$ multiplying the gradient terms held constant.
-By this we obtained slightly different segregation levels in both systems A and B, whereas the lamellar distance was virtually the same.\\
-The second part of our discussion consists of simulations in three dimensions using identical parameters to system A-2D and a box size $N_x\times N_y\times N_z=128\times256\times128$.
-We will refer to this system as system A-3D.
-Table \ref{tab1} provides an overview of all simulation parameters.
+Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
+The system size was $N_x\times N_y\times N_z=512\times 1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
+The distance from the critical point was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq.
+\ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
+By this we obtained a slightly different stress response with the system with lower order appearing somwhat sofer.
+The lamellar distance was identical in both cases.\\
+The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
+Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
+Table \ref{tab1} gives an overview of all simulation parameters.
 
-\begin{table}[htp!]
+\begin{table}[htp]
 \small
-\caption{Simulation parameters: The systems were first equilibrated from spinodal decomposition at zero shear rate (EQ), followed by various shear rates, ranging from $\dot{\gamma}=2\cdot10^{-4}$ till $1.25\cdot10^{-5}$.
-For some runs a final switch off (SO) at zero shear rate was performed.
+\caption{Simulation parameters: Starting from spinodal decomposition the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\cdot10^{-5}$ to $2\cdot10^{-4}$.
+In some cases we performed a final switch off run (SO) at zero shear rate.
 The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
 \label{tab1}
-%  \begin{tabular*}{0.48\textwidth}{@{\extracolsep{\fill}}lllll}
 \begin{tabular*}{0.495\textwidth}{lllll}
@@
 \hline 
-703 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
-704 & $1024\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
-705 & $1024\times512\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
+703 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
+704 & $512\times1024\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
+705 & $512\times1024\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
 \hline
-767 & $512\times256\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
-775 & $512\times256\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
-769 & $512\times256\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
-771 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
-772 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
-774 & $512\times256\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
+767 & $256\times512\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
+775 & $256\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
+769 & $256\times512\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
+771 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
+772 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
+774 & $256\times512\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
 \hline
-792 & $256\times128\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+792 & $128\times256\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
 \hline
@@
 \hline
-707 & $1024\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
-710 & $1024\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
-765 & $1024\times512\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
+707 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
+710 & $512\times1024\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
+765 & $512\times1024\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
 \hline
-776 & $512\times256\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
-777 & $512\times256\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
-778 & $512\times256\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
-780 & $512\times256\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
-781 & $512\times256\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
-782 & $512\times256\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
+776 & $256\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
+777 & $256\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
+778 & $256\times512\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
+780 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
+781 & $256\times512\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
+782 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
 \hline
@@
 \hline
-998 & $256\times128\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
-1001 & $256\times128\times1$ & 320-640& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+998 & $128\times256\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
+1001 & $128\times256\times1$ & 320-960& $1\cdot10^{-4}$ & $2\cdot10^{-4}$ \\
+1000 & $128\times256\times1$ & 320-1280& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
 999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
-1002 & $128\times256\times128$ & 320-640 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
+1003 & $128\times256\times128$ & 320-960 & $1\cdot10^{-4}$ & $2\cdot10^{-4}$\\
+1002 & $128\times256\times128$ & 320-1280 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
 \hline
@@
 \subsection*{Results in Two Dimensions}
-There is some freedom regarding the parametrizaton of binary fluids using the Landau-Brazovskii free energy approach \ref{FE-functional}.
-Hence we stuck in the present work to parameters, which have been found to cover the relevant morphology of lamellar systems and produce stable outcomes in previous studies \cite{Kendon01,Xu03, Xu06b}.
-We performed simulations for a variety of parameter sets with Landau parameters $1\cdot10^{-3}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
+
+Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq.
+\ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
+Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
+In previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
+We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
 Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
 The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
-Smaller Landau parameters slowed down the ripening speed of the lamellar structure, which is consistent with the fact that the system is closer to the critical point where the phase microphase separation occurs.
-Higher viscosities led to shorter interfaces, whereas values around the lower threshold invoked longer interfaces.
+Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems B has a lower average order parameter and is closer to the critical point.
+Higher viscosities lead to shorter interfaces, whereas lower values typically yield longer lamellae.
+This effect has been reported before \cite{Gonnella97} and turned out to be more pronounced for the softer system B-2D, although not as quite to the same extent as in the former case.
+In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
+In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
+The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
+The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
+If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
 
-\begin{figure}[htp!]
+\begin{figure}[htp]
 \centering
@@
 
-This effect has been reported before \cite{Gonnella97} and was more pronounced for softer systems with smaller Landau parameters, although not quite to the same extent as in the above reference.
-Under shear higher viscosities caused the lamellae to be much less homogenous in shape and thickness.\\
-In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the stiffer system A-2D with $-a=b=5\e{-4}$ and the softer system B with $-a=b=5\e{-5}$, respectively.\\
-The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
-The equilibrium structure is completely isotropic and exhibits a lamellar spacing of roughly 10 lattice sites.
-If shear flow is imposed, the lamellae start to align with the flow.
+The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
+The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
+After a short transient phase during which the order parameter remains rather homogenous, expanded and compressed regions emerge with a lamellar spacing that deviates from the equilibrium value.
+This is depicted in the centre picture of Fig. \ref{fig1} for shear rate $\gd= 1\e{-4}$.
+The inhomogeneities remain present in the steady state, are advected by the flow and undergo continuously affine transformations.
+A buckling instability above a critical shear rate has been described earlier by Gonnella et al. \cite{Gonnella98} and Kumaran et al. \cite{Kumaran2001}.
+It was suggested that the density of defects defects get introduced into a lamellar liquid crystal due to dilation and annealing occurs due to application of shear.
+This would lead to irregularities in the sample a dynamical process where the steady state defect density is determined by balance between the rate of introduction and rate of annealing of defects.
+In a more recent paper Kumaran et al. state that they observe similar inhomogeneities only above a certain system size \cite{Kumaran2011}, which agrees well with our observation.
+We found that shear alignment is more likely to take place in smaller systems of $N_x\times N_y\times N_z=128\times 256\times 1$.
+However, this trend is not necessarily present at higher shear rates, a fact that we will discuss later pertaining to our simulations in three dimensions.\\
+Kumaran et al. applied only one shear rate and focussed on the dependence of shear alignment on the Schmidt number $Sc$, defined as the ratio between kinematic viscosity and diffusivity.
+Their values for $Sc$ are of the order ${\cal O}(1)$ (ours are roughly ${\cal O}(10^2-10^3)$), which is surprisingly low given the fact that the appearance of their lamellae does not seem to be dominated by heavy fluctuations.
+However, identical to our results they observe enhanced shear alignement at lower Schmidt numbers, i.e. higher diffusivities.\\
+Kumaran et al. also identified a second mechanism of defect creation that is due to layer compression and splitting.
+Although we have not checked this in detail, this is also very likely the case in our results as the lamellar pattern is strongly compressed and heterogenous as shown in Fig. \ref{fig1}.\\
+After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
+The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
+Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
+The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
+Fig. \ref{fig2} shows equivalent states for system B-2D, which is closer to the critical point.
+All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
+Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
+Another shear-induced structure in binary lamellar systems was reported in \cite{Xu06b}.
+The reported results differ from our observations in some aspects.
+Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
+Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
+These differences may be simply a direct consequence of the different boundary conditions that were applied.
+While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.\\
+To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
+It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
+This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\
 
-\begin{figure}[htp!]
+\begin{figure}[htp]
 \centering
@@
 
-In this paragraph the flow direction and the direction of the velocity gradient are x and y, respectively.
-The apparent flow due to the effect of the Lees-Edwards planes is always in positive x-direction in the upper half and in negative x-direction in the lower half of the box.
-After a short startup phase in which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing deviates from its equilibrium value.
-These regions are advected with the flow while they undergo an affine stretching.
-The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $1\e{-4}$.
-In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
-However, the reported results differ from our observations in the following aspects.
-First of all, the shear-induced structures emerge at the centre and remain located, while ours occur in band-like patterns that are wrapped around the periodic boundaries.
-Moreover those bear the characteristics of a shear banding region exhibiting a reduced local shear rate.
-These differences may be a direct consequence of the different boundary conditions.
-While we applied Lees-Edwards BC along y and periodic boundary conditions along x for all quantities, \cite{Xu06b} seem to have used sort of hybrid scheme.
+By introducing the alignment function $X$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
+Fig. \ref{fig3} shows an example of the time evolution of system A-2D at low shear rate.
+All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
+The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
+There are regions which show some sort of flow alignment and others which don't.
+The bottom picture of Fig. \ref{fig3} shows a later state with a total strain that has increased by another $10^{5}\%$. 
+Obviously many defects have annealed and the interfaces are well aligned with the flow direction.
 
-\begin{figure}[htp!]
+\begin{figure}[htp]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
+\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
+The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
+Sites where the alignment function $X$ was smaller than $0.8$ are highlighted red.}
+\label{fig3}
+\end{figure}
+
+Defect annihilation to this large extent is only found below a specific shear rate.
+A prerequisite seems to be that the defects can smootly approach each other and cancel, which is prevented by the buckling instability above a criticl shear rate.
+Then the system shows a tendency to trip over itself, which results in the creation of new defects.
+The critical value differs in systems A-2D and B-2D and is related to the magnitude of the order parameter and the distance from the critical point.
+It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
+
+\begin{figure}[htp]
 \centering
@@
 The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
-\label{fig4}
-\end{figure}
-
-To our understanding they applied Lees-Edwards BC for the order parameter only, but placed solid moving walls at the upper and lower boundaries in the lattice Boltzmann part.
-Identical to our setup periodic BC in flow direction were applied for both order parameter and momentum flux.
-The it might be that \cite{Xu06b} have not simulated bulk flow behaviour but rather flow in a confined geometry and the shear-induced structure they observe might be a consequence of this difference.\\
-After switching off of the shear flow the expanded regions contract and the compressed regions expand and the system returns to a state with homogenous lamellar width, seemingly close to the equilibrium value.
-The final state for which the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
-The lamellar interfaces remain more or less oriented, but due to the expansion and contraction of the lamellae some individual domains show large angles with the previous flow direction. 
-The length of the interfaces varies considerably, which is appears to be another residue of the inhomogeneities in the flowing state.\\
-Fig. \ref{fig2} shows equivalent states for the softer system B.
-The characteristic results of Fig. \ref{fig1} are retained, but the interfaces in all states tend to be longer.
-The lamellar alignment in flow direction after switch off is more pronounced.\\ 
-Fig. \ref{fig3} relates the inhomogeneities in the morphology to rheological quantities.
-The pictures show the order parameter $\phi$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
-Where the system is extended and the lamellar width is larger compared to the equilibrium state, both shear stress and first normal stress differences have low positive or negative values.
-Large positive values occur in the regions where the lamellae are squeezed.
-The peak values in lattice units are in the region of $-2\e{-4}$ and $3\e{-4}$ for the shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
-Despite the local negativity of quantities, the overall average are still positive for all times.
-Very similar results with smaller peak magnitudes are obtained for system B, which appears somewhat more homogenous.\\
-Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the three states depicted in Fig. \ref{fig1}.
-At first isotropic after the microphase separation and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
-In steady shear flow the structure factor develops two distinct though rather broad peaks, which are situated at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
-It is interesting to see that the Fourier-vector of the peak position pics up a small component in x-direction, indicating on average a slight tilt of the structure.
-The tilting orients the lamellae away from the direction of flow and towards the velocity gradient direction, which is the direction of extension in simple shear flow.
-The position of the peak is almost constant throughout the entire simulation.
-The average lamellar distance decreases from about 10 lattice sites in the quiescent state to about 8.5 lattice sites.
-However, after switching off the system returns its equilibrium lattice spacing.
-The peak intensity rises due to the fact that the interfaces remain oriented along the flow direction but more interestingly the small x-component of the peak position disappears as the lamellar spacing becomes homogenous again.
-This gives some evidence that the small x-component is indeed connected to extensional flow in the regions with negative first normal stress differences.\\
-With the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify defects in the lamellar structure.
-We would like to locate regions where there is no homogenous order parameter changes because it changes sign and the interfaces have a large local curvature.
-Fig. \ref{fig5} shows an example of the time evolution of system A-2D at low shear.
-All lattice sites with $X_i\le0.8$ are marked red and obviously coincide with interfacial endpoints and disturbances.
-The top picture shows the situation at the beginning of the simulation with the lowest shear rate $\dot{\gamma}=1.25\e{-5}$.
-The system has been pre-sheared at consecutively lowered shear rates ranging from $\gd=1\e{-4}$ to $2.5\e{-5}$, whereas the final configuration of the previous run was always used as initial configuration of the following run.
-There are regions which exhibit some degree of flow alignment and others, where the angle between the lamellae and the flow direction remains rather large.
-The bottom picture of Fig. \ref{fig5} shows the structure at the end of the run.
-The total strain difference between both images is about $10^{5}\%$. 
-Obviously many defects have annealed and the interfaces are well oriented in flow direction.
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
-\caption{Defect structure of system A-2D: The pictures show results of a smaller simulation at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
-The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
-Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
 \label{fig5}
@@
 
-This degree of defect annihilation is only found below a critical shear rate where defects are advected smoothly on well aligned lamellae, approach each other and cancel.
-Above the critical shear rate the lamellae buckle quite similar to what has been reported in \cite{Gonnella98}, the structure shows a tendency to trip over itself which creates new defects.
-The critical value is different in both systems, roughly between $\dot{\gamma}=2.5-5\e{-5}$ in the stiffer system A-2D and higher at around $\dot{\gamma}=1-2\e{-4}$ in the softer system B.
-Fig. \ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates.
-The runs with the two highest shear rates were started from the quiescent equilibrated state, whereas all others used the final configuration of the preceding run with higher shear rate as initial configuration.
-Here and in the following figures horizontal lines are time averages for the particular quantities where we think the system has reached a steady state. 
+Fig. \ref{fig5} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
+Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
+In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
+It is interesting to see that the inhomogenous morphology of the lamellar structure is also clearly encoded in the structure factor. 
+The Fourier-vector at the peak has a small component in x-direction, indicating the average tilt of the lamellae away from the flow direction in the direction of extensional flow.
+The position of the peak is almost constant throughout the entire simulation and the average lamellar spacing decreases from about 10 lattice sites in equilibrium to about 8.5 lattice sites.
+If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
+The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.\\
+
+Fig.
+\ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$.
+The runs with the two highest shear rates were started from the equilibrium configuration, whereas all others were pre-sheared using the final configuration of the preceding run as initial configuration in order to save computational rexources.
 The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
-Interestingly once the shear rate has been decreased to the next lower rate, both systems A and B show an trend to anneal defects instantaneously or after a short latency time if the shear rate is rather low.
-If the shear rate is turned up again, the lamellae buckle and the structure begins to trip over itself, once again creating new defects.
+Interestingly, while stepping down in shear rate both systems A-2D and B-2D anneal defects almost instantaneously.
+If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur immediately.
 
-\begin{figure}[htp!]
+\begin{figure}[htp]
 \centering
@@
 \caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
-Horizontal lines indicate averages over time for the particular quantities, where we assume the system has reached a steady state.
+Horizontal lines indicate time averages of the particular quantities, where we assume the system has reached a steady state.
 The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
@@
 
-This suggests that there is a dynamical equilibrium between defect creation and annihilation above a critical shear rate, while below the critical value defects get not anymore widely generated and the system anneals its defects and forms properly aligned lamellae.
-In order to test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
-The defect density goes indeed up (not shown) and reaches similar values to those shown in Fig. \ref{fig6}.
-However it takes some time before a steady state is reached as the applied shear rates are rather low.\\
-Fig \ref{fig7} shows the average total shear stress density $\langle \sigma_{xy}\rangle$, i.e. the total shear stress per unit lattice volume.
-In the following we will drop the brackets $\langle\dots\rangle$ for simplicity and write $\langle \sigma_{xy}\rangle\equiv  \sigma_{xy}$.
-While for the highest applied shear rate the hydrodynamic part forms the largest contribution to the total stress density, a regime where the chemical stress dominates is entered on decreasing the shear rate.
-At low shear rates stress fluctuations become more stretched in time and get smaller, what is slightly hidden by the logarithmic scale of Fig. \ref{fig7}
-Their magnitude is about three times smaller in the larger simulation shown in Fig. \ref{fig1}, \ref{fig2}.
-Fortunately we found that the average shear stress density over time is to a good degree independent of the system size and therefore it is justified to simulate longer times by using a smaller system size without flawing the results.
+This finding is in further support of a dynamical equilibrium between defect creation and annihilation above a critical threshold, wheras below the critical value defects cancel and lamellae align.
+To test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
+The defect density goes indeed up (not shown) and reaches similar values as before.\\
+Fig \ref{fig7} shows the ensemble averaged total shear stress $\langle \sigma_{xy}\rangle \equiv\sigma_{xy} $, i.e. the total shear stress per unit lattice volume.
+At low shear rates stress fluctuations become actually more stretched in time and smaller, a fact that is obscured by the logarithmic scales.
+In larger simulations such as those shown in Fig. \ref{fig1} and Fig. \ref{fig2} they are also about three times smaller.
+Nevertheless we found that the averaged shear stress is to a very good degree independent of the system size, which makes it possible to simulate longer times using smaller system sizes without flawing the results.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
-\caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages}
+\caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
+ }
 \label{fig7}
@@
 
-Furthermore the stress fluctuations represent a physical aspect as they can be related to the inhomogeneities of the order parameter.
-System A shows clear shear thinning behaviour and seems to obey a power law (see Fig. \ref{fig11}).
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the time averaged apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
-Stress fluctuations cause temporary variations of these mean values of up to a factor 3 at the lowest shear rate.
-In contrast to system A-2D system B does not exhibit a clear monotonous shear thinning tendency.
-The apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow more a horizontal trend when the shear rate is decreased.
-The stress fluctuations and time averaged shear stress densities are slightly smaller than in the stiffer system A-2D with one exception.
-For the highest shear rate $\gd=2\e{-4}$ the shear stress turns out to be larger than in system A-2D.\\
-The deviatoric part of the average chemical pressure density $\langle P_{xy}\rangle\equiv P_{xy}$ for system A-2D is shown in the main picture of Fig. \ref{fig8}.
+System A-2D shows clear shear thinning behaviour and seems to obey a power law (see below Fig. \ref{fig11}).
+Values of the apparent viscosity are given in the caption of Fig. \ref{fig7}.
+Contrary to system A-2D system B-2D does not exhibit a clear tendency of shear thinning.
+Here, the apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow a more horizontal trend.\\
+We want to focus now on the deviatoric part of the chemical pressure $\langle P^{chem}_{xy}\rangle\equiv P^{chem}_{xy}$.
+This is shown for system A-2D in Fig. \ref{fig8}.
 The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
-System B shows a very steep decrease of $P_{xy}$ during the simulation with $\gd=1\e{-4}$, which is because of the defect annihilation.\\
+In System B-2D $P^{chem}_{xy}$ decreases sharply during the run with $\gd=1\e{-4}$ (not shown) because at this shear rate pronounced defect annihilation sets in.\\
 
@@
 
-Fig. \ref{fig9} illustrates the relation between $P_{xy}$ and the defect density $\rho_D$.
-The black line represents a fit to the system A-2D, which shows a linear dependence almost down to the lowest shear rate.
-The result for the softer system B is slightly more obscured as a consequence of the ordering and the tendency to anneal defects.
-Our quantifying protocol is based on averaging over a larger number of defects sites and is prone to inaccuracy in a regime where defects are rather scarce.
+Fig. \ref{fig9} illustrates dependence of $P_{xy}$ on the defect density $\rho_D$.
+System A-2D features a linear relation between both quantities that breaks only down at the lowest defect density or rather shear rate.
+The result for the softer system B-2D are less clear due to the ordering and annihilation of defects.
+It is very likely that our quantifying procedure, which is based on averaging over a larger number of defects sites,is prone to inaccuracy in this regime.
 
@@
 
-The main picture in Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A-2D.
-At first until $\gd=5\e{-5}$ the first normal stress differences decrease monotonously with decreasing shear rate.
-At lower shear rates, however, the behaviour is far from being monotonous and $N_1$ rises again.
-This makes it difficult or even impossible to define reasonable time averages as $N_1$ may still evolve at the end of the run.
+Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A-2D, where we followed the same protocol as before.
+Initially, down till $\gd=5\e{-5}$ $N_1$ decreases monotonously with decreasing shear rate.
+However, at lower shear rates the behaviour is not monotonous anymore and $N_1$ rises again.
+Hence, it is difficult to define a reasonable time average for $N_1$ as for these low shear rates the system may still evolve at the end of the run.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
-\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{xx}-\sigma_{yy}\rangle$ vs time for system A-2D.
+\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{yy}-\sigma_{xx}\rangle$ vs time for system A-2D.
 The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
@@
 
-This feature is actually less erratic than it seems at first sight and can be rationalised. 
-A look at Fig. \ref{fig6} reveals that $N_1$ rises in the same way as the defect density drops.
-A pair of defects can cancel if one of them is an endpoint of a black lamella surrounded by a branching white lamella and the other one is a white endpoint surrounded by a branching black lamella.
-When these two defects anneal the lamella spacing increases locally by half a lamellar width.
-The interfaces have to expand in y-direction as the system is not frustrated enough to create an entire new layer in order to get the lamellar width back closer to its equilibrium value.
-Hence the normal stresses $\sigma_{yy}$ become negative and consecutively smaller, leading to gradually larger values of $N_1$.
-It should be stressed that only $\sigma_{yy}$ seems to evolve anomalously slowly and all other components of the stress tensor reach state state values more rapidly.
-System B again behaves different in the sense that $N_1$ continues to fall for lower shear rates despite there is also ongoing annealing of defects.
-This difference is a direct effect of the softness of system B, whose smaller Landau coefficients prevent the normal stresses $\sigma_{yy}$ from taking up very negative values and $N_1$ from rising.
-Nevertheless close to the end of the simulations with the two lowest shear rates $N_1$ is still slightly changing which suggests that this slowly evolving behaviour is also due to ongoing defect annihilation.
-Because both trends in system A-2D and B are consequences of the impossibility to optimise the lamellar width, larger simulations should lead to more defined steady state values for $N_1$.
-However, we were not able to check this as this proves computationally costly.\\
-In Fig. \ref{fig11} we show flow curves derived from time averages of $\sigma_xy$ and $N_1$ as a function of shear rate.
-The time average of shear stress density in systems A follow a power law $\sigma_{xy}=\dot{\gamma}^m$ with an exponent $m\simeq 0.87$ and gives evidence for shear thinning behaviour.
-In case of system B the best fit for the lowest three shear rate is obtained for an exponent $m=1$, indicating more or less constant viscosity and no shear thinning.
-It is not very surprising as the chemical contribution to the total stress almost vanishes for shear rates $\dot{\gamma}\le1\e{-4}$ due to the significant decrease of the number of defects.
-The observed power law behaviour is simply that of the hydrodynamic part of the stress, which is linear in our case.\\
-A comparison with experimental work remains difficult because the topology of defects differ in two and three dimensions.
+This feature which appears somewhat erratic at first has actually a simple explanation. 
+The time sequence of the defect density in Fig. \ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$ and linked to the defect annihilation.
+When two defects meet and cancel the lamellar width increases locally by a certain amount.
+This causes frustration in the system as the incommensurability between the layers is not large enough to create a new lamellar layer and allow the system to snap back towards its equilibrium width.
+Hence the normal stresses $\sigma_{xx}$ become increasingly smaller and may even attain negative values, leading to gradually larger values of $N_1$.
+This explanation is further sustained by the fact that only $\sigma_{xx}$ seems to evolve anomalously slowly and all other components of the stress tensor reach their steady state values more quickly.
+We expect that these anomalous trends become considerably smaller in larger systems.
+System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
+This is likely to be a direct consequence of the lower order parameter in system B-2D and its softness with respect to system A-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
+In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
+System A-2D gives rise to shear thinning with the average shear stress of systems A-2D following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
+For system B-2D the best fit is actually obtained for an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning.
+This is because due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$ the chemical stress $P^{chem}_{xy}$ almost vanishes and the total stress is dominated by the hydrodynamic part.\\
 
@@
 \end{figure}
-
-Our two-dimensional system is translational invariant in the third dimension.
-So every branch point or endpoint of a lamella has the topology of an edge defect in the 3D-counterpart of the structure.
-Moreover screw defects do not exist in our picture as they can only occur in more than two dimensions.
-A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
-Interestingly the authors identify exactly the motion of screw defects as the key mechanism responsible for the layer tilting and defect formation. 
-For the power law dependence of the shear stress on the shear rate they find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.\\
-Contrary to shear stresses first normal stress differences do not level off because of ongoing defect annihilation.
-Neither system A-2D nor B is in a steady state at the end of the simulation and therefore the data for $N_1$ is only shown for the sake of completeness.
 
@@
 
-The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which has a slightly larger segregation than system B.
-The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $256\times128\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
+The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which is intermediate between A and B.
+The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $128\times256\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
 Two snapshots of system C-2D and A-2D in shear flow are shown in the bottom pictures in Fig. \ref{fig12}.
@@
 
+A comparison of these results with experimental work is possbile although the topology is different in two and three dimensions.
+Our 2D-systems correspond to are translational invariant along the third dimension.
+Every branching point or endpoint of a lamella has the topology of an edge defect in the 3D-continuation of the structure.
+A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
+Interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, floding  and formation of defects. 
+For the power law dependence of the shear stress on the shear rate Lu et al.
+find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
+Our value of $m=0.87$ appears to be slightly too large.
+A possible explanation for this is that screw defects, which do not exist in our 2D-picture, play a key role for this theory.\\
+
 \section{Conclusions}
-We have performed large scale simulations of a lamellar system in steady shear flow  in two and three dimensions.
-We were able to relate the morphology of the order parameter in the quiescent state to results from previous studies of the same system using similar parameters.\\
-In two dimensions and steady state simple shear flow at Peclet numbers ${\cal O}(10)$ we did not observe a homogenous lamellar structure for shear rates above a critical threshold, which depends on the exact parametrization of the free energy functional.
-The morphology was rather inhomogenous instead, exhibiting regions that are under pressure with lamellar spacings smaller than the equilibrium value and others that were stretched.
-We could relate the structure to typical values of rheological quantities.
-Areas under extensional flow coincide with locally negative or small positive shear stresses and first normal stress differences, whereas both go to positive peak values in the compressed domains.
-Below a critical shear rate the systems showed no tendency anymore to trip over itself and defects annealed as they could approach each other in adjacent lamellae.
-The characteristic time of this process was set by the shear rate and the total strain.
-The deviatoric part of the chemical pressure was correlated to the number of defect sites, whereas the latter was measured by means of an alignment factor, which quantified the local curvature of the lamellar interfaces.
-For the stiffer system A-2D we could verify a linear relation between the defect density and the average chemical contribution to the shear stress.\\ 
-In system A-2D first normal stress differences decreased as long as a dynamic equilibrium between defect creation an annihilation existed.
-Below the critical threshold this was no longer the case and pronounced defect annealing caused a rise in the first normal stress differences as the average lamellar spacing increased.
-We could establish a shear thinning regime for the more defect laden system A-2D.
-System B in contrary was so perfectly annealed in the steady state that virtually shear stresses were down at the hydrodynamic level and the relation between shear stress and shear rate was trivially linear.\\
-In three dimensions we simulated shear strains which have not been attained so far.
-Interestingly for the same parameters we observe striking differences between system A-2D in 2D and system A-3D in 3D.
-It seems that the dynamics of the defects in two and three dimensions is rather incommensurable.
-Even at the lowest applied shear rates the chemical stress has decayed to a level where the apparent viscosity was virtually identical to the solvent.
-Structure factor data confirmed the very regular layered ordering of the structure and a lamellar spacing close to the equilibrium value.
+
+We studied nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model for binary liquid-crystalline fluids, which is based on a Brazovskii-free-energy-functional.
+The lamellar structure of the order parameter in the quiescent state agrees well with previous results using the same model in two dimensions and similar parameters.
+When flowing the morphology of the order parameter depends crucially on the system size.
+At system sizes $N_x\times N_y\times N_z=128\times256\times 1$ the system showed a strong tendency to align with the shear flow at low shear rates $\gd\le5\e{-5}$.
+At higher shear rates and for larger system sizes of $N_x\times N_y\times N_z=256\times512\times 1$ and above heterogenous lamellar structures with strongly extended and compressed regions were present in the steady states.
+A similar dependence of these charactistic patterns on the system size has been recently reported in \cite{Kumaran2011}.
+The patterns occured in distinctive bands and were oriented along the direction of extension and perpendicular to the direction of compression.
+The closer the system was to the critical point the more this feature gave way to flow alignment.
+The extended regions with lamellar spacing larger than at equilibrium were local regions with either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
+Depending on the applied shear rate and in agreement with other studies we found two different dynamical regimes. 
+Below a critical value, which was basically controlled by the magnitude of the order parameter and the distance from the critical point, alignment in flow direction and annealing of defects to a very large degree was obtained.
+Beyond the critical shear rate a buckling instability occured.
+The systems continuously tripped over itself, which resulted in a dynamical equilibrium between defect annihilation and creation of new defects.
+The shear stress in system A-2D follwed a clear power law dependence on the strain rate like $\sigma_{xy}=\gamma^\alpha$ with $\alpha=0.87$, giving rise to shear-thinning behaviour.
+Interestingly, the magnitude of chemical contribution $P_{xy}$ to the shear stress could be directly related to the density of defects.
+In system A-2D first normal stress differences became smaller with decreasing shear rate as long as the system was above the buckling instability.
+Below the critical shear rate continued defect annealing caused a rise in the first normal stress differences due to incommensurabilities in the lamellar width.\\
+We simulated flow behaviour of the same systems in three dimensions for the first time and found striking differences between the two- and three-dimensional versions of systems A and C.
+The 3D-system showed none of the distictive bands along the extensional direction and no buckling instability was found in the range of applied shear rates.
+The structure, initially amorphous after spinodal decomposition, ordered into a lamellar array with edge defects oriented along the flow directions.
+This is in total contrast to the 2D-case, where numerous defects appeared that would correspond to edge defects that are oriented perpendicular to the flow in 3D.
+Our results agree with theoretical values and simulations that predict a layer undulation instability at shear rates about two orders of magnitude larger than the ones we have applied.
+This constitutes a surprising change in results due to the different dimensonality and topoloogy of the strucutre despite identical parametrisation of the model.
 
 \section{Acknowledgements}
+We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time at the Argonne Leadership Computing Facility, Argonne, U.S.A. as well as on HECToR, U.K.
+We like to thank A. Morozov for useful discussions. 
+M.E.C. holds a Royal Society Research Professorship.
 
-\section{Appendix: sliding periodic boundary conditions}
+\section{Appendix: Sliding Periodic Boundary Conditions}
 
@@
 applied to the lattice Boltzmann part of the calculation
-\cite{Wagner02,adhikari2005} to generate shear in the system. We
+\cite{Wagner02,Adhikari05} to generate shear in the system. We
 describe here how the sliding periodic boundaries are extended to

### 2011-11-03 papers/smectics/smectics.tex
Smectics draft with Mike's comments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1478 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{float}
+%\usepackage{rsc}
 
@@
 \newcommand{\update}[1]{{\color{red} #1}}
-\newcommand{\e}[1]{\cdot10^{#1}}
+\newcommand{\e}[1]{\times10^{#1}}
 \newcommand{\gd}{\dot{\gamma}}
@@
 
+%\newcommand{\Pe}{\hbox\rm Pe}
 \twocolumn[
   \begin{@twocolumnfalse}
-\noindent\LARGE{\textbf{Shear Flow of Lamellar Mesophases and Smectic Liquid Crystals}}
+\noindent\LARGE{\textbf{Rheology of Lamellar Liquid Crystals in Two and Three Dimensions: A Simulation Study}}
+%: A Simulation Study}}
 \vspace{0.6cm}
 
-\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.V. Coveney{$^{a}$} and M.E. Cates \textit{$^{c}$}}
+\noindent\large{\textbf{O. Henrich\textit{$^{a,c}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.~V. Coveney{$^{a}$} and M.~E. Cates \textit{$^{c}$}}
 }\vspace{0.5cm}
@@
 
-\noindent \normalsize{In this work we present large scale computer simulations of the bulk rheology of a smectic liquid crystal at intermediate to large values of the shear flow, therefore in the nonlinear regime.
-In two dimensions we find that, according to the thermodynamic parameters, shear may either align the system and stabilise an almost regular lamellar phase, or induce the nucleation and proliferation of defects in the smectic layers.
-In this last case, defects may be progressively annealed by slowly decreasing the applied shear rate, so that there are marked memory effects and the rheology is history-dependent.
-Finally, selected simulations in three dimensions show shear-induced ordering rather than defect nucleation, thus pointing out to a dramatic effect of dimensionality on the physics of the system.}
+\noindent \normalsize{We present large scale computer simulations of the nonlinear bulk rheology of lamellar phases (smectic liquid crystals) at moderate to large values of the shear rate (Peclet numbers $10-100$), in both two and three dimensions.
+In two dimensions we find that modest shear rates align the system and stabilise an almost regular lamellar phase, but high shear rates induce the nucleation and proliferation of defects, which in steady state is balanced by the annihilation of defects of opposite sign.
+The shear rate $\dot\gamma_c$ at onset of this second regime is controlled by thermodynamic and kinetic parameters; we offer a scaling analysis that relates $\dot\gamma_c$ to a critical ``capillary number'' involving those variables.
+Within the defect proliferation regime, the defects may be partially annealed by slowly decreasing the applied shear rate; this causes marked memory effects, and history-dependent rheology.
+Simulations in three dimensions show instead shear-induced ordering even at the highest shear rates studied here.
+This suggests that $\dot\gamma_c$ shifts markedly upward on increasing dimensionality.
+This may in part reflect the reduced constraints on defect motion, allowing them to find and annihilate each other more easily.
+Residual edge defects in the 3D aligned state mostly point along the flow velocity, an orientation impossible in two dimensions.}
 \vspace{0.5cm}
@@
 
-\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, UK}}
-\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK}}
-\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, UK}}
+\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, INSERT ADDRESS HERE, UK}}
+\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
+\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
 
@@
 
+SADLY, I THINK IT WOULD BE BEST TO REWRITE THIS PAPER WITH THE FLOW DIRECTION ALONG X, THE GRADIENT ALONG Y AND THE VORTICITY ALONG Z AS IS COMMONLY USED IN THE RHEOLOGICAL LITERATURE!
+I HAVE NOT STARTED TO MAKE THIS CHANGE AS IT WOULD CONFUSE THINGS TOO MUCH AND THE JOB REQUIRES CARE.
+ALSO THE CHEMICAL PRESSURE AND THE STRESS HAVE OPPOSITE SIGNS YET THE FIGURES SHOW THEM HAVING THE SAME SIGN.
+WHAT HAS HAPPENED?
+
 \section{Introduction}
 
-Instances of soft complex fluids are abound in our everyday world. 
-Examples include colloidal suspensions, 
-where particles or droplets are suspended in a viscous
-fluid, living tissues, and liquid crystalline fluids. Well known
-examples of colloidal suspensions include paint, cosmetics, and
-food products such as ketchup, mayonnaise, and yogurt.
-The intracellular network of actin fibres constituting the
-cytoskeleton of mammalian cells is a polymeric fluid.
-The same is true of the DNA making up the bacterial nucleoid and
-eukaryotic chromosomes, and the viral genome of bacteriophages,
-where enhanced confinement imparts a strong liquid crystalline
-character to the system.
-Liquid crystalline fluids themselves, with or without colloidal
-inclusions, are of of wide technological importance in devices
+Instances of soft complex fluids abound in our everyday lives: examples include paint (a colloidal suspension), mayonnaise (an emulsion), and engine oil (a polymer solution).
+Many soft materials, including the amphiphiles used in skin creams, fabric softeners, and some pharmaceutical products, form lamellar mesophases (also known as smectic liquid crystals) whose perfectly ordered state consists of a regular stack of flat layers.
+As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent [deGennes book].
+Such phases are of wide technological importance in devices
 such as electronic displays and sensors.
 
-The behaviour under flow, or rheology, of such materials is important
-to understand, as it affects many industrial and manufacturing processes.
-At the same time, rheological measurements provide a powerful way to
-characterise macroscopic properties of a soft material.
-The response to applied forces or stresses can be used to quantify,
-e.g., the viscosity, or elastic and loss moduli of soft materials.
-On shearing, a simple Newtonian fluid exhibits a stress which is
-linearly proportional to the applied shear rate; the
-constant of proportionality is just the viscosity. In marked
-contrast, complex fluids flow is often non-Newtonian and such a simple
-linear relation no longer holds. On shearing, the effective viscosity
-of a complex viscoelastic fluid may increase or decrease, corresponding
-to shear thinkening or shear thinning behaviour. Complex fluids may even
-spontaneously phase separate when flowing, giving rise to shear bands.
-These are familiar in the rheology of liquid crystals in the vicinity of
-the isotropic-nematic transition, and in wormlike micelles.
+The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
+For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity.
+Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression.
+However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress.
+Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear.
+On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
+to shear thickening or shear thinning behaviour [Larson book].
+Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
+Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) [P D Olmsted, Rheologica Acta 47, 283-300 (2008)].
 
-The rheological properties of a non-Newtonian fluid are typically linked
-to the microstructure of the material. Particularly relevant to our
-current work is the rheology of layered systems, such as smectics and
-cholesterics. Here, experiments may show different rheological results
-depending on whether the system has been ordered prior to measurement
-(for example, by pre-shearing).
-For instance, in the flow of polydomain liquid crystals, which are not
-uniformly ordered, the existence of regions with conflicting director
-patterns leads to a non-linear relation between shear stress and
-applied shear rate.
-Another example of strongly non-Newtonian rheology is region I flow
-in polymeric nematic liquid crystals. While our understanding of the
-physics of region I flow is far from complete, it is thought that a
-defect network may form and lead to much enhanced viscosity, with
-marked shear thinning. In both these cases, stiffening is usually
-associated to the presence of defects in the sample.
+Lamellar liquid crystals provide an important paradigm for non-Newtonian fluids in which the microstructure of the material is strongly coupled to the flow.
+Crucially, the defect texture may depend on past history -- for instance, the strain rate in any prior flow that was suddenly switched off.
+This rheological hysteresis means that the precise flow protocol, including any pre-shearing applied prior to starting the experiment, needs to be specified.
+Such hysteresis can arise in polydomain liquid crystals of various types (not just smectics) such as nematics in which the existence and evolution of regions with misaligned directors creates rheological nonlinearity.
+A closely related example is the so-called `region I flow' in polymeric nematic liquid crystals, where it is thought that a defect network arises, with enhanced viscosity, but with
+marked shear thinning as this structure is sheared out at high flow rates [Larson book -- or Davide to supply references].
 
-Another intriguing type of flow which is possible in layered systems
+An intriguing complication in addressing the dynamics of layered systems
 is permeation,
 where individual molecules in the fluid translate while keeping the
-defect pattern intact in steady state. This leads to increased dissipation
-and hence to a larger viscosity. A well-known example of permeation flow
+layerwise order intact. In steady state, permeation leads to increased dissipation
+and hence to a large viscosity in any geometry where flow necessitates such motion [deGennes book]. (A well-known example of permeation flow
 is found in cholesterics which are sheared or pushed along the direction
-of the cholesteric helix. However, one would also expect permeation flows
-to be possible when a non-translationally invariant defect pattern is
-present in the system; this might be the case in quenched smectics, for
-instance. In addition to their non-linear shear-stress curve, lamellar
-systems are also of interest for a number of other phenomena.
-In block-copolymers one may observe, under some conditions, shear-induced
-phase transitions from the microphase separated state to ordered phases
-like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. It would be
-desirable to extend our theoretical understanding of such switches.
-Further, lyotropic layered liquid crystals have sometimes been observed
-to form an onion phase \cite{Panizza96,Iwashita07} which can happen either
-spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under
-shear \cite{Diat93}. The mechanism underlying such pattern formation has
-remained elusive to date.
+of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction.
+Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07}[ADD; M Gradzielski, JPCM 15, R655-R697 (2003)].
+(The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear.
+Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation [A G Zilman and R Granek, EPJB 11, 593-608 (1999), L Courbin, J P Delville, J Rouch and P Panizza, PRL 89, 148305 (2002)]\cite{Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion [J Zipfel, J Berghause, P Lindner, W Richtering, J Phys Chem B 103, 2841-2849 (1999), A Leon, D Bonn, J Meunier, A Al-Kahlwaji, O Greffer, H Kellay, PRL 84, 1355-1338 (2000)].
+Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work.
+Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
+phase transitions from a homogeneous state into to ordered phases
+like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}.
+However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}.
+In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
+once defects are present).
 
-Despite a wealth of experimental results, and the importance of rheology
-to many commercial applications, theoretical progress in understanding
-flow behaviour of complex fluids has remained a challenge. This is largely
-because of the non-linear nature of the equations of motion, and the
-coupling of the flow to different microstructures. These either require
-drastic approximations or, for practical purposes, resort to computer
-simulation. Even here, realistic 3-dimensional calculations are
-computationally expensive.
-In this work, our focus is the physics of smectics under shear.
+Despite a wealth of experimental results, and the importance of flow behaviour
+to many commercial applications, mechanistic understanding
+the rheology of lamellar phases remains limited. This is largely
+because of the nonlinear nature of the equations of motion, and the
+coupling of the flow to different microstructures. These complications either require
+drastic approximations or, as pursued here, use of computer
+simulation. However, once defects are present, such simulations are computationally expensive. 
 While sheared smectics have been studied extensively via simulations
@@
 computational constraints have meant almost all these results are limited
-to two dimensions. Further, a number of these studies have used solid walls
-to drive shear in the hydrodymamics and, inconsistently, sliding periodic
-boundaries for the composition variable. Such simulations have a number of
-drawbacks: solid walls to drive shear limits the attanable shear rate in
-larger systems and so exacerbates finite size effects; mixed boundary
-conditions make it difficult to disentangle bulk and boundary effects
-in the rheology. In this work, we try to overcome all of these issues.
-(One notable example which does tackle a large 3-dimensional simulation
-of a complex fluid under shear involved gyroid phases~\cite{Saksena09}.)
+to two dimensions.
+Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology.
+Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs.
+This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems.
+Moreover, solid walls have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.
 
-We describe smectics via a Brazovskii free energy functional
-\cite{Brazovskii75}, which provides a robust framework to study systems
-like micelles, surfactants, smectic liquid crystals and block-copolymers.
-The governing equations in the model are a convection-diffusion equation
-for the scalar order parameter, which represents the density difference
-of two fluid components, and the Navier-Stokes equation describing
-hydrodynamics. A stress tensor that contains extra contributions owing
-to the presence of interfaces between the components is included.
-These dynamical equations are solved using a combination of a finite
+In this work, we try to overcome these technical obstacles, presenting systematic results for large two dimensional simulations with fully periodic boundary conditions.
+Guided by those results, we additionally report some very large three dimensional simulations at selected points in the parameter space.
+These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals~\cite{Saksena09}.
+[ADD: R S Saksena and P V Coveney, Proc.
+Roy.
+Soc.
+A.
+465, 1977-2002 (2009)).]
+
+We describe smectics below via a widely-used free energy functional [G Gompper and M Schick, Phase Transitions and Critical Phenomena vol 16, C Domb and J L Lebowitz, Eds., Academic Press NY 1994; then cite all the lamellar simulation papers using this or similar functionals].
+This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations).
+The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$.
+This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid and interfaces between these layers.
+A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks.
+However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents the smectic state.
+
+The governing equations in our model are a convection-diffusion equation
+for the scalar order parameter, and the Navier-Stokes equation describing
+the hydrodynamics of fluid flow.
+The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy.
+Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
+Our dynamical equations are solved using a combination of a finite
 difference scheme for the convection-diffusion equation, and the
-lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation.
-We use consistent sliding periodic (Lees-Edwards) boundary conditions for
-both convection-diffusion and Navier-Stokes equations. This allows us to
-probe the true bulk rheology of the smectic phase for large systems and
-high shear rates, as quantified by an appropriate Peclet number.
+lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation
+[CITE OUR PREVIOUS PAPERS USING HYBRID LB, ALSO THAT OF GONELLA ET AL WHO FIRST USED IT FOR SMECTICS].
+We use sliding periodic (Lees-Edwards) boundary conditions for
+both the convection-diffusion and Navier-Stokes equations.
+Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses.
+(As it happens, we do not observe shear bands in the range of shear rates we address here.)
+
+%This simulation strategy allows us to probe the bulk rheology of large systems at moderate to high shear rates, as quantified by an appropriate Peclet number, defined below.
 
@@
 
-Our main finding is that, depending on system parameters, a smectic phase
-shows markedly different rheological responses. At lower Peclet
-numbers, where thermodynamic effects are relatively strong compared with
-flow effects,  shear quickly orders the system into a lamellar phase,
-with only a few defects. The rheology in this regime is Newtonian, and
-the shear stress is linear in the applied shear rate.
-If, on the other hand, thermodynamic forces are larger, we observe that shear actually leads to the nucleation of new defects, and the rheology is highly nonlinear, with the shear rate proportional to $\dot{\gamma}^{\alpha}$, with $\alpha\simeq 0.87$, leading to a shear thinning behaviour.
-Such a behaviour is qualitatively reminiscent of the one found experimentally in XXX, where the numerical value of $\alpha$ was however different.
-In this regime we also find that the rheology is history dependent, as the defect density decreases for instance if the shear rate is ramped down gradually.
-Finally, intriguingly, dimensionality appears to play a key role: within our 3D simulations, we have indeed not been able to enter the nonlinear rheology regime with defect nucleation, and we always observe a shear-induced ordering into a lamellar phase.
-More research and parameter steering will be needed to make this observation more quantitative.
+Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
+of shear, where thermodynamic forces are relatively strong compared with
+flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
+with only a few defects.
+The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate.
+(There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow.
+This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other.
+The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters; a comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
+Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning.
+The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
+Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE?], where the numerical value of $m$ was however different.
+In this strongly non-Newtonian regime we also find that the rheology is history dependent.
 
-This paper is organised as follows. In Section II we introduce the model,
+Finally, and intriguingly, we find that dimensionality appears to play a key role.
+Within our 3D simulations, even when performed with parameter values identical to those used in 2D, we have so far not been able to enter the nonlinear flow regime associated with defect proliferation.
+Instead, we always observe shear-induced ordering into a lamellar phase with quasi-Newtonian rheology.
+Further research, requiring careful and systematic parameter steering, will be needed to map out quantitatively the onset of the defect proliferation regime, if one exists, in three dimensions.
+
+This paper is organized as follows. In Section \ref{themodel} we introduce the model,
 including its equilibrium properties, the full stress tensor and the
 dynamical equations. We briefly sketch how we derive structural information
-from the order parameter and describe our defect quantification procedure,
-which measures the local alignment of the lamellar structure.
-We present results of the two-dimensional case in in Section III and
-discuss the morphology, average and local shear stress densities and
-first normal stress differences and their relation to the defect structure.
-We also show structure factors for various flow states and flow curves.
-We also compare our two-dimensional results with results in three dimensions. 
-A summary with conclusions closes the paper in Section V.
+from the order parameter and describe a method for quantitative analysis of the defect density.
+Section \ref{parametermap} describes the mapping between our simulation parameters and those for physical systems of interest.
+We present results of the two-dimensional case in Section \ref{results2D}, discussing
+the morphology as a function of flow rate. We also consider average and local shear stress densities and
+first normal stress differences, and explore their relation to the defect structure. In addition we report Fourier-space information (structure factors) under flow.
+In Section \ref{results3D} we present and compare our results for 3D simulations which, by computational necessity, are less comprehensive than in 2D. 
+A summary with conclusions closes the paper in Section \ref{conclusions}.
 
-\section{Model and Definitions}
+\section{Model and Definitions} \label{themodel}
 
-The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.
+The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75} [ADD: Gommper and Schick].
+%
+\begin{eqnarray}
+{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
+\end{eqnarray}
+%
+In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature [Kendon et al JFM, relevant Gonnella papers].
+This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform.
+The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.
 
+The parameter range relevant to lamellar phases has $\kappa < 0$ and $c>0$.
+Then, if we suppress the quartic term, the free energy density of a sinusoidal lamellar state $\phi({\bf r}) = \phi_k\cos({\bf k}.{\bf r})$ obeys $\bar f = (a+\kappa k^2+ c k^4)\phi_k^2/4$.
+This is minimized by choosing $k=q\equiv\sqrt{|\kappa|/2c}$; the free energy density is then $\bar f = (a-\kappa^2/4c)\phi_q^2/4$ so that ordering will occur for $a<a_c=\kappa^2/4c$.
+The amplitude $\phi_q$ is then unbounded without the (positive) quartic term, since $\bar f$ can in that manner become infinitely negative.
+Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{(a_c-a)/b}$, creating also nonlinear harmonics in the density profile.
+The final free energy density obeys $\bar f \simeq (a_c-a)^2/b$.
+By similar arguments, the characteristic elastic modulus associated with microstructural deformations is
+%
+\begin{equation}
+\bar g\sim \frac{(a_c-a)a_c}{4b}
+\label{barg}
+\end{equation}
+%
+which is the energy density of the gradient terms only.
+The terms involving $\phi^2$ and $\phi^4$ in Eq.~\ref{FE-functional} are unaltered by affine deformations which conserve density locally; hence those terms do not contribute to $\bar g$.
+
+The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
+%
+\begin{equation}\label{OPEOM}
+\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
+\end{equation}
+%
+Here $M$ is a mobility coefficient.
+A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ [CONFIRM] (this is negative, signifying the instability).
+In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$.
+This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.
+
+Treating the fluid as incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
+%
+\begin{equation}\label{NSE}
+\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta}. 
+\end{equation}
+%
+where $\rho$ is the mass density. The stress tensor in this equation reads
+%
+\begin{equation}
+\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-p\delta_{\alpha\beta} - P_{\alpha \beta}. \label{fullstress}
+\end{equation}
+%
+This consists of a Newtonian hydrodynamic part, which is proportional to the background viscosity $\eta$ and the symmetrized velocity gradient, a hydrostatic pressure term, which includes an isotropic stress contribution from the order parameter
+%
 \begin{eqnarray}
-{\cal F}[\phi]&=&\int d^dx f(\phi({\bm x}))\nonumber\\
-&=&\int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\bm \nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\label{FE-functional}
-\end{eqnarray}
-
-In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
-For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
-For $a<0$ a microphase separation occurs.
-Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
-The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$. 
-By varying the Landau-parameters $a$ (and $b\sim |a|$) we change the distance from the critical point and the magnitude of the order-parameter.
-The segregation between both phases was found to dependent neither solely on $a,b$ nor only on $\kappa,c$. 
-It is rather the magnitude of all parameters, which has to be similar in order to produce a lamellar phase, that controls the level of segregation.\\
-The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.
-
-\begin{equation}\label{OPEOM}
-\partial_t \phi +{\bm \nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
-\end{equation}
-
-The coefficient $M$ herein plays the role of a mobility coefficient and the diffusivity of the system can be approximated by $D_0\simeq M\;b$.\\
-If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
-
-\begin{equation}\label{NSE}
-\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha = \partial_\beta \sigma_{\alpha \beta}. 
-\end{equation}
-
-The stress tensor in this equation reads
-
-\begin{equation}
-\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-P^{th}_{\alpha \beta}.
-\end{equation}
-
-It consists of a hydrodynamic part, which is proportional to the background viscosity $\eta$ and velocity gradients and a thermodynamical part $P^{th}$ that depends on the order parameter and gradients thereof:
-
-\begin{equation}
-P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + {\cal D}_{\alpha \beta}(\phi).
-\end{equation}
-
-In this last expression the symmetric tensor ${\cal D}_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee mechanical equilibrium \cite{Evans79}.
-The thermodynamic pressure tensor is given by
-
-\begin{equation}
-P^{th}_{\alpha \beta}=p\, \delta_{\alpha \beta} + P^{chem}_{\alpha \beta}
-\end{equation}
-
-with an isotropic pressure contribution
-
-\begin{eqnarray}
-p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\bm \nabla}\phi)^2\right)\nonumber\\
+\Delta p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
 & &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
 \end{eqnarray}
+%
+and finally another term consisting of the order-parameter contribution to the deviatoric stress
+\cite{Evans79} [REFERENCE]
+%
+\begin{equation}
+P_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
+\end{equation}
+CONFIRM SIGN.
 
-and an anisotropic part
+In our hybrid LB scheme, the Navier-Stokes equation (\ref{NSE}) is solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
+The velocity field is then used in the advection-diffusion equation Eq.\ref{OPEOM}, which is solved by a finite-difference method at the same timestep.
+We assume periodic boundary conditions along the flow and vorticity direction and Lees-Edwards sliding periodic boundary conditions\cite{leesedwards,Wagner02} along the direction of the velocity gradient.
+Following \cite{Wagner02}, we introduce
+additional internal Lees-Edwards planes to overcome an upper limit on shear rate that would otherwise be inversely proportional to the system size in the gradient direction.
+(This restriction arises because fluid velocities must be small compared to the sound speed in the frame of the lattice; the additional planes create a stack of lattice slabs whose velocity is nearly matched to the local fluid velocity.)
+The total number of Lees-Edwards planes was varied according to the size of the simulation so that the separation of two planes was always 16 lattice sites.
 
-\begin{equation}
-P^{chem}_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
-\end{equation}
-
-The Navier-Stokes Eq.\ref{NSE} was solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
-The velocity field was then used in the advection-diffusion equation Eq.\ref{OPEOM}, which was solved by a finite-difference method at the same timestep. 
-We assumed periodic boundary conditions along the flow and vorticity direction and Lees-Edwards boundary conditions (BC) \cite{Wagner02} along the direction of the velocity gradient.
-The total number of Lees-Edwards planes was varied according to the size of the simulation so that the distance of two planes was always 16 lattice sites.
-To impose simple shear flow, the planes in the upper half of the velocity gradient direction were moved at a constant speed in positive flow direction, whereas the planes in the lower hemisphere were moved in negative flow direction.\\
 Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
-
+%
 \begin{equation}
 C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
-\end{equation}
+\label{S}\end{equation}
+%
+with $\phi({\mathbf k})$ the normalized Fourier transform of the order parameter.
+The transform itself was calculated using the FFTW-library \cite{FFTW}.
+In principle an ensemble- or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.
 
-with $\phi({\mathbf k})$ being the normalised Fourier-transform of the order parameter.
-The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
-For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
+For the analysis of the relation between rheological quantities and the morphology of the system we need to to quantify the number of defects.
 For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average projection of the normalised order parameter gradients onto the gradient at a specific site.
-More explicitly the alignment function $X$ is given by
-
+More explicitly, 
+%
 \begin{equation}\label{alignment}
@@
 \end{equation}
+%
+Here $V'$ is a small volume that contains $\mathbf r$, which in practice can be chosen to contain $\mathbf r$ and its nearest neighbours only.
 
-$V'$ is a small volume that contains $\mathbf r$, which in practice can be chosen to contain  and its nearest neighbours.
-For smooth alignment and small curvature of the lamellae $X$ is very close to unity.
-At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
-By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates preceivably from what would be recognised as an aligned configuration exhibiting lamellar order.
-We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
-Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment function below the threshold.
-However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
-By normalising the number of defects sites with $X$ below the threshold a defect density can be defined:
-
+For ordered regions with small curvature of the lamellar interfaces, $X$ is very close to unity.
+Wherever lamellae terminate at edge dislocations, and in regions where their orientation changes abruptly, $X$ departs from unity, taking smaller positive values.
+By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as as an aligned lamellar configuration.
+We found a threshold value of $\xi=0.8$ to be well suited for our requirements.
+This protocol does not directly count the number of defects, since at each defect more than one site has an alignment function below the threshold.
+However, as the number of sub-threshold sites per defect is reasonably constant, the degree of order under flow can usefully be quantified by defining a ``defect density" as the system-wide fraction of sub-threshold sites:
+%
 \begin{equation}\label{defect density}
@@
 \end{equation}
+%
+Here $\theta$ is the Heaviside function.
+The conversion factor between $\rho_D$ and the defect number density (in 2D) or line density (in 3D) depends on the lamellar wavenumber $q$ and the choice of threshold parameter $\xi$; we do not make such conversions here and consider $\rho_D$ itself as an appropriate measure of disorder.
 
-The function $\theta$ is the Heaviside function and $V$ is the volume of the entire system.
-The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.
+\section{Parameter Mapping to Physical Units}
+\label{parametermap}
 
-\subsection*{Parameter Mapping to Physical Units}
+The parameter space of the model is quite large.
+Although in exploratory work we performed studies across a wider range of parameter values, for simplicity in the results presented here we restrict attention to fixed values of $\kappa = -6\times 10^{-4}$LBU and $c = 7.6\times 10^{-4}$LBU (here LBU denotes LB units as defined below).
+These are chosen to give a lamellar wavelength of $L = 10$ LBU (that is, $q = \sqrt{|\kappa|/2c} = 2\pi/10$LBU) which offers suitable spatial resolution.
+Moreover the microphase separation point $a_c = \kappa^2/4c = 1.2\times 10^{-4}$LBU is then a fixed constant.
 
-In the following we describe how the simulation units can be mapped onto real physical units.
-It is necessary to calibrate the scales for mass density, length and time or alternatively as we do for mass density, length and energy density.\\
-As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the reference fluid $\rho_0$ all to unity in LB units (LBU).
-It is straightforward to determine the unit of mass. 
-A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
-Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.\\
-In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
-Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
-Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space or $1 {\rm m}=2\cdot10^8$ LBU.\\
-The energy scale is slightly more difficult to calibrate.
-We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
-This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
-The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
-The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
-Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
-This formula reads 
+As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the fluid $\rho$ all to unity to define LB units (LBU).
+We need to calibrate these units (or equivalently, length, energy density and time) against real physical units.
+We do this first for thermotropic smectics, then consider the lyotropic case.
+
+To calibrate the LB length unit we equate $L$ the lamellar wavelength with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm.
+Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{10}$LBU.
+Free energy density parameters such as $a_c \simeq 10^{-4}$LBU should then be of order $\alpha^2k_BT/L^3$ where $\alpha =10$ is a molecular aspect ratio.
+Thus the LB unit of energy density is about $6\times 10^{10}$Pa, i.e.
+1 Pa$= 1.6 \times 10^{-11}$ LBU.
+
+Turning now to the fluid viscosity $\eta$, after pilot runs spanning a range of several decades, we selected $8.33\times 10^{-2}$LBU for the production runs reported here.
+This represents the baseline viscosity of the isotropic phase close to the transition (not the higher, effective viscosity of the smectic once formed).
+Requiring this to be about ten times the viscosity of water (which is $\eta_w = 8.9\times 10^{-4}$Pa s) we find that the LBU unit of viscosity is about $0.1$Pa s.
+Given the previous result for the energy density, we infer that 1s$= 6\times 10^{11}$ LBU.
+The LBU timestep is thus around $1.6\times 10^{-12}$s.
+
+As detailed previously, an estimate of the molecular diffusivity is $\tilde D = 2Mb$; we set $M = 0.25$LBU throughout this study, so that $\tilde D$ lies in the range $2.5\times 10^{-5} - 2.5\times 10^{-4}$LBU.
+Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{11}$LBU) gives $D\simeq 10^{-11}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.
+
+Recall that the mass density of our fluid is unity in LBU.
+However, combining the above conversion factors for length,  energy density, and time shows the physical and simulation units of mass density to be related via 1 kg m$^{-3} \equiv 1$ Pa s$^2$ m$^{-2} \simeq 10^{-8}$LBU.
+Thus our simulations address a fluid of mass density $10^8$kg m$^{-3}$ which is $10^5$ times larger than the true value for a typical material.
+The reason for this is that the LB algorithm uses fluid inertia to update the dynamics and the discretization should therefore involve replacing the real fluid with the densest one possible, while ensuring that the Reynolds number $Re \sim \rho\dot\gamma L^2/\eta$ (which for real smectics is exceedingly small) remains small enough for the simulation to faithfully describe the low $Re$ limit.
+As discussed elsewhere [CODEF,SOFT MATTER] $Re$ values up to about 0.1 are then admissable; our simulations comply with this requirement.
+
+Finally, our shear rates lie in the range 
+$10^{-5}-10^{-4}$LBU, so that the (dimensionless) Peclet number, $Pe =\dot\gamma L^2/\tilde D$, which expresses the relative importance of advective to diffusive transport, lies in the range $20-1600$ CONFIRM: WRONG VALUES GIVEN PREVIOUSLY?.
+These are moderate to high values where significant structural rearrangement can be expected.
+However, for thermotropic smectics, given our time unit of 1LBU$=1.6\times 10^{-12}$s, they correspond to strain rates of order $10^6$s$^{-1}$.
+This lies well above the range typically addressable with rheological experiments, but far below that addressed recently in molecular dynamics simulations [Guo,Sodderman,Stewart].
+
+The shear rates needed to achieve this range of $Pe$ falls dramatically when lyotropic smectics are considered.
+For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm.
+The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$) [Lyotropic Review].
+The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e.
+1 Pa$= 1.6 \times 10^{-6}$ LBU.
+Both the viscosity $\eta$ and the molecular diffusivity $\tilde D$ remain comparable (within one order of magnitude) to the thermotropic values; from the viscosity we then determine that 1s$=6\times 10^6$LBU or 1LB timestep $=1.6\times 10^{-7}$s.
+Accordingly the strain rates relevant to the $Pe$ range addressed here are of order 100 s$^{-1}$ and well within experimental reach.
+Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still irrelevantly small.
+
+To further simplify our parameter exploration, throughout what follows we restrict attention to  states with $b = -a$.
+This convention has been widely used to study bulk spinodal decomposition in binary fluids where it can be absorbed in a rescaling of the order parameter [KENDON JFM]: put differently, $b=-a$ ensures that coexisting bulk ordered phases have $\phi = \pm\phi_0$ with $\phi_0=1$.
+In lamellar states of $b= -a$ the amplitude of the lamellar ordering need not be unity but varies like $\phi_q \sim \sqrt{1+a_c/b}$ as stated previously.
+Thus the chosen value of $b$ can be thought of as controlling whether the peak order parameter variation is close to that arising for bulk coexistence (when $b\gg a_c$), or whether the amplitude $\phi_q$ is instead much larger ($b\ll a_c$) and determined by a competition between the gradient terms and a relatively weak quartic nonlinearity.
+In this study we choose three values, $b = 5\times 10^{-4}, 1\times 10^{-4}$ and $5\times 10^{-5}$LBU so that the dimensionless quantity $\beta \equiv b/a_c$ is respectively $2.1,0.42,0.21$, lying near the crossover between these regimes.
+The observed $\phi_q$ range is about $1.2-1.5$, somewhat narrower than the range $1.1-1.8$ predicted from $\phi_q \sim \sqrt{1+a_c/b}$.
+However the latter estimate, which does not treat the nonlinear term exactly, is only a guideline.
+
+Alongside the Reynolds number (which we deem irrelevant), $\beta$, and the Peclet number, we can identify two further dimensionless numbers of interest.
+One is the Schmidt number $Sc = \eta/\rho\tilde D$, which is large when momentum can diffuse much more rapidly than molecules -- as applies in almost all dense fluids.
+In our simulations $Sc$ is of order $100--1000$ which is safely large.
+(The work of [KumaranRamanPRE2011] instead addresses Schmidt numbers in the range $0.2-5$ where different physics may apply.) The second parameter controls the strength of feedback between distortions of the $\phi$ field and the fluid motion.
+This
+can be expressed in terms of a characteristic value of the Peclet number, $Pe^*$, at which the free energy density in the order parameter field that couples to deformation (Eq.\ref{barg}), $\bar g=(a_c+b)a_c/4b$, is comparable to the viscous stress $\sigma_v=\eta\dot\gamma = (\eta \tilde D/L^2)Pe$.
+Equating these gives
 \begin{equation}
-b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
+Pe^* =\frac{(a_c+b)a_c}{4b^2}\,\frac{L^2}{\eta M}\label{pestar}
 \end{equation}
-whereas $\rho_c$ is the number density of chains per unit volume,
-often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
-Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
-With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
-Hence the LB unit of energy density is about $10^{6} \rm{Pa}$ or $1\rm{Pa}=10^{-6}$ LBU.\\
-We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
-As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=(1\e{-3}) \cdot (1\e{6}) \cdot (4\e{16})\rm{LBU}\simeq4\e{19} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq6.3\e{9}$ LBU and one LB timestep $\Delta t$ corresponds to roughly $0.15 \rm{ns}$.\\
-Finally we want to get an estimated scale for shear rates. 
-A common way is to express the ratio of transport via advection to that via diffusion by means of the Peclet number.
-The Peclet number can be defined as $Pe=\gd\, L^2/D_0$, whereas $\gd$ is the strain rate, $L$ is a characteristic length scale, e.g. the lamellar width and $D_0$ is the diffusivity.
-Because the Peclet number is a dimensionless quantity, we can perform this estimation conveniently in lattice units.
-The highest shear rate we apply are of the order $1\e{-4}$ in reciprocal LB time units.
-With the above given values for the lamellar width of $L\simeq10$ and diffusivity of system A $D_0=M\;b\simeq 0.25 \cdot 5\e{-4}=1.25\e{-4}$ this results in Peclet numbers $Pe=(1\e{-4})\cdot(1\e{2})\cdot(1.25\e{-4})^{-1}=80={\cal O}(10^2)$.
-Hence, the conversion between shear rates and Peclet numbers is roughly $\gd=10^{-4}\cong Pe=10^2$.
+With our chosen parameters (in LBU, $a_c=1.2\times 10^{-4}, L = 10, \eta = 8.3\times 10^{-2}, M = 0.25$) we find $Pe^* \sim 1400$ at $\beta = 2.1$ and $Pe^* \sim 40000$ at $\beta = 0.21$.
+The maximum values of $Pe/Pe^*$ that we address below are of order 0.1 and 0.04 respectively in the two cases.
 
-\section{Results}
+For later convenience we may define a new dimensionless number,
+which we can call a capillary number. (The capillary number is conventionally defined as the ratio of viscous to interfacial forces in binary fluid flow.) We define this as:
+%
+\begin{equation}
+Ca = \frac{Pe}{Pe^*} = \frac{4\eta\dot\gamma b}{(a_c+b)a_c}
+\label{capillary}
+\end{equation}
+%
+We will argue below that this could be a more relevant choice than $Pe$ for understanding the dependence of morphology on flow rate in smectic systems, at least when $Pe^*\gg 1$.
 
-Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
-The system size was $N_x\times N_y\times N_z=512\times 1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
-The distance from the critical point was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq.
-\ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
-By this we obtained a slightly different stress response with the system with lower order appearing somwhat sofer.
-The lamellar distance was identical in both cases.\\
-The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
-Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
-Table \ref{tab1} gives an overview of all simulation parameters.
+\section{Results and Discussion}
+
+Below we first present detailed results for two parameter sets ($\beta = 2.1, 0.21$) in two dimensions which we call systems A-2D and B-2D.
+System lattice sizes and shear rates are reported in Table \ref{tab1}; all remaining parameters are the same in both systems as detailed above.
+%As expected from the preceding discussion of $Pe^*(\beta)$, the stress response at a given flow rate is softer in the second system. 
+A third parameter set ($\beta = 0.42$, system C-2D) gave results broadly similar to B-2D.
+Note that we will later report simulations in three dimensions for systems A-3D and C-3D which share parameters with A-2D and C-2D (see Table \ref{tab1}).
 
@@
 \small
-\caption{Simulation parameters: Starting from spinodal decomposition the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\cdot10^{-5}$ to $2\cdot10^{-4}$.
+\caption{Simulation parameters: After initialization from a noisy uniform state, the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\times10^{-5}$ to $2\times10^{-4}$LBU.
 In some cases we performed a final switch off run (SO) at zero shear rate.
-The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
+The parameters $\kappa=-6\times10^{-4}, c=7.6\times10^{-4}, M=0.25$ and $\eta=8.33\times10^{-2}$ were the same throughout all simulations.
+ADD PE and CA COLUMNS!}
 \label{tab1}
@@
 \hline
-run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
+run \# & $N_x\times N_y\times N_z$ & time $/10^3$ & $10^5b$ &  $10^5\dot\gamma$\\
 \hline
-system A & & & &\\
+A & & & &\\
 \hline 
-703 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
-704 & $512\times1024\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
-705 & $512\times1024\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
+703 & $512\times1024\times1$ & 0-320 & $50$ & 0 - EQ  \\
+704 & $512\times1024\times1$ & 320-640& $50$ & $10$\\
+705 & $512\times1024\times1$ & 640-960& $50$ & 0 - SO \\
 \hline
-767 & $256\times512\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
-775 & $256\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
-769 & $256\times512\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
-771 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
-772 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
-774 & $256\times512\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
+767 & $256\times512\times1$ & 0-320 & $50$  & 0 - EQ\\
+775 & $256\times512\times1$ & 320-640& $50$ & $20$ \\
+769 & $256\times512\times1$ & 320-2560& $50$ & $10$ \\
+771 & $256\times512\times1$ & 1280-3840& $50$ & $ 5$ \\
+772 & $256\times512\times1$ & 5120-7680& $50$ & $ 2.5$ \\
+774 & $256\times512\times1$ & 5120-10240& $50$ & $1.25$ \\
 \hline
-792 & $128\times256\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+792 & $128\times256\times1$ & 320-1600& $50$ & $10$ \\
 \hline
-786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
-787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
-788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
-789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
+786 & $128\times256\times128$ & 0-640 & $50$ & 0 - EQ\\
+787 & $128\times256\times128$ & 320-640 & $50$ & $20$\\
+788 & $128\times256\times128$ & 320-960 & $50$ & $10$\\
+789 & $128\times256\times128$ & 640-1280 & $50$ & $5$\\
 \hline
-system B & & & &\\
+B & & & &\\
 \hline
-707 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
-710 & $512\times1024\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
-765 & $512\times1024\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
+707 & $512\times1024\times1$ & 0-320 & $5$ & 0 - EQ \\
+710 & $512\times1024\times1$ & 320-960& $5$ & $10$ \\
+765 & $512\times1024\times1$ & 960-1280& $5$ & 0 - SO \\
 \hline
-776 & $256\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
-777 & $256\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
-778 & $256\times512\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
-780 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
-781 & $256\times512\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
-782 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
+776 & $256\times512\times1$ & 0-320 & $5$ & 0 - EQ\\
+777 & $256\times512\times1$ & 320-960& $5$ & $20$ \\
+778 & $256\times512\times1$ & 320-2560& $5$ & $10$ \\
+780 & $256\times512\times1$ & 1280-3840& $5$ & $5$ \\
+781 & $256\times512\times1$ & 2560-6400& $5$ & $2.5$ \\
+782 & $256\times512\times1$ & 5120-7680& $5$ & $1.25$ \\
 \hline
-system C & & & &\\
+C & & & &\\
 \hline
-998 & $128\times256\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
-1001 & $128\times256\times1$ & 320-960& $1\cdot10^{-4}$ & $2\cdot10^{-4}$ \\
-1000 & $128\times256\times1$ & 320-1280& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
-999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
-1003 & $128\times256\times128$ & 320-960 & $1\cdot10^{-4}$ & $2\cdot10^{-4}$\\
-1002 & $128\times256\times128$ & 320-1280 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
+998 & $128\times256\times1$ & 0-320& $10$ & 0 - EQ \\
+1001 & $128\times256\times1$ & 320-960& $10$ & $20$ \\
+1000 & $128\times256\times1$ & 320-1280& $10$ & $10$ \\
+999 & $128\times256\times128$ & 0-320 & $10$ & 0 - EQ\\
+1003 & $128\times256\times128$ & 320-960 & $10$ & $20$\\
+1002 & $128\times256\times128$ & 320-1280 & $10$ & $10$\\
 \hline
@@
 
-\subsection*{Results in Two Dimensions}
+\subsection{Two Dimensions}
 
-Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq.
-\ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
-Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
-In previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
-We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
-Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
-The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
-Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems B has a lower average order parameter and is closer to the critical point.
-Higher viscosities lead to shorter interfaces, whereas lower values typically yield longer lamellae.
-This effect has been reported before \cite{Gonnella97} and turned out to be more pronounced for the softer system B-2D, although not as quite to the same extent as in the former case.
-In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
-In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
-The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
-The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
-If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
+\label{results2D}
+Before presenting specific results for the parameter sets detailed above, we briefly describe the effect of wider parameter variations.
+Within the Landau-Brazovskii free energy functional (Eq.~\ref{FE-functional}) a range of parameter variations can produce stable lamellar phases, as explored in previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011}.
+The nonequilibrium morphology of the lamellar structure, and the phase ordering kinetics (starting from a slightly noisy, uniform initial state) do however depend on the choice made for these, and also for the viscosity.
+Formation and ripening of the lamellar structure is slower when $\beta$ is small, consistent with the reduced initial driving force for ordering in this case.
+Higher viscosities lead to more locally disordered structures in which there is a shorter separation between kinks or edge defects as one moves along the lamellae; conversely lower viscosities typically yield ``longer" lamellae.
+This effect has been reported before \cite{Gonnella97}; we found it to be more pronounced for softer (lower $\beta$) systems.
+At fixed $\dot\gamma$, higher values of $\eta$ were found to cause greater inhomogeneity of lamellar shape and thickness, consistent with the fact that $Ca\propto\eta$ (the larger the shear stress, the larger the lamellar deformation).
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
-\caption{Order parameter $\phi$ of system A-2D: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
-Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
+\caption{Order parameter pattern $\phi({\bf r})$ of system A-2D (thresholded to enhance visibility): after equilibration (timestep $t=3.2\e{5}$, top), in steady shear at $\dot{\gamma}=1\times10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
+Gray scaling from black to white corresponds to values of $-1\le\phi\le1$.}
 \label{fig1}
@@
 
-The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
-The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
-After a short transient phase during which the order parameter remains rather homogenous, expanded and compressed regions emerge with a lamellar spacing that deviates from the equilibrium value.
-This is depicted in the centre picture of Fig. \ref{fig1} for shear rate $\gd= 1\e{-4}$.
-The inhomogeneities remain present in the steady state, are advected by the flow and undergo continuously affine transformations.
-A buckling instability above a critical shear rate has been described earlier by Gonnella et al. \cite{Gonnella98} and Kumaran et al. \cite{Kumaran2001}.
-It was suggested that the density of defects defects get introduced into a lamellar liquid crystal due to dilation and annealing occurs due to application of shear.
-This would lead to irregularities in the sample a dynamical process where the steady state defect density is determined by balance between the rate of introduction and rate of annealing of defects.
-In a more recent paper Kumaran et al. state that they observe similar inhomogeneities only above a certain system size \cite{Kumaran2011}, which agrees well with our observation.
-We found that shear alignment is more likely to take place in smaller systems of $N_x\times N_y\times N_z=128\times 256\times 1$.
-However, this trend is not necessarily present at higher shear rates, a fact that we will discuss later pertaining to our simulations in three dimensions.\\
-Kumaran et al. applied only one shear rate and focussed on the dependence of shear alignment on the Schmidt number $Sc$, defined as the ratio between kinematic viscosity and diffusivity.
-Their values for $Sc$ are of the order ${\cal O}(1)$ (ours are roughly ${\cal O}(10^2-10^3)$), which is surprisingly low given the fact that the appearance of their lamellae does not seem to be dominated by heavy fluctuations.
-However, identical to our results they observe enhanced shear alignement at lower Schmidt numbers, i.e. higher diffusivities.\\
-Kumaran et al. also identified a second mechanism of defect creation that is due to layer compression and splitting.
-Although we have not checked this in detail, this is also very likely the case in our results as the lamellar pattern is strongly compressed and heterogenous as shown in Fig. \ref{fig1}.\\
-After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
-The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
-Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
-The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
-Fig. \ref{fig2} shows equivalent states for system B-2D, which is closer to the critical point.
-All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
-Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
-Another shear-induced structure in binary lamellar systems was reported in \cite{Xu06b}.
-The reported results differ from our observations in some aspects.
-Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
-Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
-These differences may be simply a direct consequence of the different boundary conditions that were applied.
-While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.\\
-To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
-It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
-This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\
+We turn now to system A-2D ($\beta = 2.1$) The top panel in Fig. \ref{fig1} shows the order parameter pattern $\phi$ at the end of the equilibration phase prior to shearing.
+Although the true equilibrium structure would show homogeneous alignment, this cannot be reached in a reasonable time scale unless shear is applied.
+The resulting metastable structure is amorphous and isotropic, but locally exhibits the same lamellar spacing $L = 10$ LBU as the fully ordered equilibrium state.
+If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
+A typical snapshot in the resulting steady state at $Pe = 80$ $(Ca = 0.06)$ CONFIRM is shown in the middle panel of Fig.~\ref{fig1}.
+In this and all subsequent images, the applied flow velocity is horizontal and the velocity gradient vertical, with the flow directed to the right in the upper half and to the left in the lower half of the picture.
+After a short transient during which the order parameter remains relatively homogenous (not shown) there emerge extended regions in which the smectic layers are either expanded and compressed, so that $L$ locally deviates from the equilibrium value.
+These mesoscopic inhomogeneities are inclined at an angle to the flow and remain continuously present in the steady state, but are not stationary: their local features are advected by the flow and evolve continuously by deformation.
+
+We found that after the shear flow was switched off the lamellar spacing returned to its equilibrium value everywhere, with expanded regions contracting and vice versa.
+The final state, after all relaxation has virtually come to a standstill, is shown in the bottom panel of Fig.~\ref{fig1}.
+Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression undergone during the previous flow, some regions retain strong misalignment which does not anneal away.
+The separation between defects as one moves along the lamellae varies considerably in space; both short and long separations can be seen.
+This applies both in the aligned layers and in layers that are misaligned with the flow.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{phi_run765_1280.jpg}
-\caption{Order parameter $\phi$ of system B: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom)}
+\caption{Order parameter $\phi({\bf r})$ of system B-2D: after equilibration (timestep $t=3.2\e{5}$, top), in steady shear at $\dot{\gamma}=1\times10^{-4}$ ($t=8\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).}
 \label{fig2}
@@
 
-By introducing the alignment function $X$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
-Fig. \ref{fig3} shows an example of the time evolution of system A-2D at low shear rate.
-All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
-The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
-There are regions which show some sort of flow alignment and others which don't.
-The bottom picture of Fig. \ref{fig3} shows a later state with a total strain that has increased by another $10^{5}\%$. 
-Obviously many defects have annealed and the interfaces are well aligned with the flow direction.
+Fig. \ref{fig2} shows equivalent states for system B-2D ($\beta = 0.21$).
+The Peclet number during the shearing phase is now $Pe = 800$ $(Ca = 0.02)$ CONFIRM.
+All the generic results of Fig.~\ref{fig1} are the same, but the defect density, both during flow and after switchoff, is significantly reduced.
+(However the bands of misalignment in the flowing state are, if anything, even more apparent.) Consequently, the lamellar alignment along the flow direction after switch off is also more pronounced.
+
+It is notable that at the shear rate shown in these figures, the more ordered system is B-2D which has higher $Pe$ than A-2D but smaller $Ca$.
+Varying $\dot\gamma$ within system A-2D or B-2D establishes unambiguously that the steady state order decreases with shear rate at fixed thermodynamic parameters (Fig.\ref{fig6} below).
+This points strongly to $Ca$ as the more relevant measure of flow rate when discussing onset of the disordered phase.
+
+In principle the dependence of the state of order on flow rate could be smooth, or could entail a nonequilibrium phase transition -- for instance if the perfectly ordered state ceased to be stable above a certain shear rate \cite{Gonnella98,Kumaran2001} [Ramaswamy,Granek].
+In practice, however, starting as we do from an initially amorphous lamellar texture, a perfectly ordered state is generally not reached for large systems CONFIRM even if it might be stable once formed.
+In any case it seems likely that at high $Pe$, any defects introduced into the smectic state by deformation of an initially imperfect structure (whose layers are then dilated and compressed by the flow) cannot anneal away or annihilate each other rapidly.
+Such irregularities then build up until the steady state defect density is determined by a balance between their rate of introduction and rate of annealing.
+
+Recently Kumaran and Raman studied a single shear rate and varied the Schmidt number $Sc$ and system size, reporting enhanced alignment on decreasing $Sc$ and in small systems \cite{Kumaran2011}.
+Our comparative studies on systems A and C have confirmed both trends (although, as mentioned previously, our $Sc$ values are much higher).
+The same authors also discuss a distinct mechanism of defect creation caused by layer compression and splitting.
+Although we have not checked this in detail, this may be consistent with our observation of a strongly compressed and heterogeneous morphology, with bands oriented normal to the compression axis, as shown in Figs.~\ref{fig1} and \ref{fig2}.
+
+A somewhat different morphology in sheared lamellar systems was reported by Xu et al.\cite{Xu06b}.
+This morphology exhibits macroscopic inhomogeneity between the centre of the simulated domain and the regions near the walls, accompanied by variations in the local shear rate.
+This contrasts with our own work which shows translationally invariant statistics in all directions.
+These differences may be a result of Xu et al.\cite{Xu06b} using mixed boundary conditions, as discussed already in the introduction, in which hard walls for the fluid flow are combined with periodic BCs on the order parameter field.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
-\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
-The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
-Sites where the alignment function $X$ was smaller than $0.8$ are highlighted red.}
+\caption{Defect structure of system A-2D:  at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\times10^{-5}$.
+The system has been pre-sheared at stepwise decreasing rates (see Table \ref{tab1}).
+Sites where the alignment function $X$ is smaller than $0.8$ are highlighted red.}
 \label{fig3}
@@
 
-Defect annihilation to this large extent is only found below a specific shear rate.
-A prerequisite seems to be that the defects can smootly approach each other and cancel, which is prevented by the buckling instability above a criticl shear rate.
-Then the system shows a tendency to trip over itself, which results in the creation of new defects.
-The critical value differs in systems A-2D and B-2D and is related to the magnitude of the order parameter and the distance from the critical point.
-It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
+To quantify our observed morphology more closely, we use the alignment function $X$ introduced in Eq.~\ref{alignment}.
+Fig.~\ref{fig3} shows snapshots of the time evolution of system A-2D at modest shear rate.
+All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints (edge dislocations) or other local defects.
+The upper panel gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, immediately following a pre-shearing protocol in which the strain rate was ramped down stepwise from higher values (see Table \ref{tab1}) IS THIS CROPPED OR THE WHOLE SYSTEM?
+WHAT SIZE?.
+While large aligned regions are present there are other regions of high defect density.
+The lower panel of Fig.~\ref{fig3} shows a later state with a total strain incremented by a further $10^{5}\%$. 
+Obviously many defects have annealed away and the smectic layers are now rather well aligned with the flow direction.
+
+Defect annihilation to this large extent is only found at low enough shear rates.
+A prerequisite seems to be that the defects can smoothly approach each other and then annihilate.
+This is prevented by the buckling of layers above a certain shear rate, beyond which the system shows a tendency to trip over itself, which results in the creation of new defects.
+The onset of defect proliferation arises at characteristic flow rates $\dot{\gamma}_c=2.5-5\e{-5}$LBU in system A-2D and around $\dot{\gamma}_c=1-2\e{-4}$LBU in B-2D.
+(Here the subscript $c$ could stand for ``critical" or for ``crossover" depending on whether there is a singular transition or not; see above.)
+These values are for
+systems of size $256\times512\times1$; we found evidence of an increasing trend with system size, but the extremely long run-times involved ($\sim 10^7$ timesteps) precluded quantitative exploration.
+These results correspond respectively to values of the Peclet number, $Pe_c$, obeying $20<Pe_c<40$ (system A) and $800<Pe_c<1600$ (system B).
+However, in terms of $Ca = Pe/Pe^*$ one has $0.014<Ca_c<0.028$ (system A) and $0.02<Ca_c<0.04$ (system B).
+Given the 40-fold difference in $Pe_c$, the similarity of $Ca_c$ in the two different systems is remarkable.
+
+Boldly extrapolating across parameter regimes that we have not explored in any detail, our results motivate the hypothesis that the onset of defect proliferation occurs generically in 2D at $Ca_c\simeq 0.03$.
+Note that $Ca$, which comprises the ratio of a viscous stress to an elastic modulus, can also be viewed as a strain.
+In that context one can imagine that defects proliferate when the deformation of the smectic exceeds a yield strain value (for which a threshold of a few percent is reasonable).
+This makes sense if, at $Ca \simeq Ca_c$, the stress contributed from the order parameter is (a) comparable to the viscous stress from the background fluid -- essentially the force balance argument made above -- and (b) primarily elastic in character.
+The latter in turn requires that $Pe_c$ is large and hence that $Pe^*$ is larger still.
+A different mechanism might apply at small $Pe^*$, since in that case layers relax rapidly by diffusion on the time scale set by $Ca_c$.
 
@@
 \includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
-\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
+\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$ (see Table \ref{tab1}).}
 \label{fig4}
@@
 
-In Fig. \ref{fig4} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
-The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
-In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
+In Fig.~\ref{fig4} we explore the correlation between the inhomogeneous morphology in the flowing state and locally defined rheological quantities.
+The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N_1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$ ($Pe = XX$, $Ca = YY$).
+(Here and below we adopt a COMPLETELY UNCONVENTIONAL coordinate system where $y$ is along the flow direction and $x$ the velocity gradient.)
+In regions where the lamellar width is larger than in the equilibrium state, shear stresses and first normal stress differences show low positive or negative values.
 Large positive values of both quantities coincide with regions where the lamellae are under compression.
-Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
-Despite being locall negative, the ensemble averages of these quantities are always positive.
-Very similar results with smaller peak magnitudes are obtained for system B-2D.
+Peak values (in LBU) are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
+Despite having locally negative regions, spatial and ensemble averages of these quantities are always positive.
+Very similar results, but with smaller peak magnitudes, are obtained for system B-2D at the same shear rate.
 
@@
 \includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
-\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right).
-The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
+\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (top); in steady shear at $\dot{\gamma}=1\times10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (bottom).
+The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU.
+LABEL THE AXES OTHERWISE X AND Y DIRECTIONS ARE NOT SPECIFIED.}
 \label{fig5}
@@
 
-Fig. \ref{fig5} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
-Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
-In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
-It is interesting to see that the inhomogenous morphology of the lamellar structure is also clearly encoded in the structure factor. 
-The Fourier-vector at the peak has a small component in x-direction, indicating the average tilt of the lamellae away from the flow direction in the direction of extensional flow.
-The position of the peak is almost constant throughout the entire simulation and the average lamellar spacing decreases from about 10 lattice sites in equilibrium to about 8.5 lattice sites.
-If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
-The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.\\
+Fig.~\ref{fig5} shows structure factors $C({\bm k})$ for the same states as Fig.~\ref{fig1}.
+$C({\bm k})$ is initially isotropic after layer formation and equilibration in the system at rest with, as expected, a distinct peak at $|{\bm k}|\simeq \pm2\pi/10$ LBU (recall $L = 10$ LBU).
+In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi, k_y\simeq\pm 0.22 \pi$.
+Thus the inhomogenous morphology of the lamellar structure is clearly encoded in the structure factor. 
+The wavevector at the peak has a small component in the $x$-direction, indicating the average tilt of the lamellae away from the flow direction towards the elongational axis of the flow (which in shear is at $\pi/4$ to the flow direction).
 
-Fig.
-\ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$.
-The runs with the two highest shear rates were started from the equilibrium configuration, whereas all others were pre-sheared using the final configuration of the preceding run as initial configuration in order to save computational rexources.
+When the shear flow is switched off, the system relaxes again to its equilibrium layer spacing and the small tilt of the $C({\bm k})$ signal decreases to an undetectable level: this is consistent with the fact that, in Figs.~\ref{fig1} and \ref{fig2} (bottom panels) it is easy to see when the system has been sheared horizontally, but not in which sense.
+The intensity of the peak is slightly enhanced in the final state due to the improved orientation of the layers along the flow direction.
+
+Fig.~\ref{fig6} shows the time evolution of the defects density $\rho_D$ of system A-2D in the sequence of stepwise decreasing shear rates alluded to previously (starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$, Table \ref{tab1}).
+The runs with the two highest shear rates were started from the amorphous, post-equilibration configuration, whereas all others were pre-sheared using the final configuration of the preceding run as the initial state.
 The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
-Interestingly, while stepping down in shear rate both systems A-2D and B-2D anneal defects almost instantaneously.
-If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur immediately.
+Interestingly, following each step down in shear rate, both systems A-2D and B-2D continuously anneal to lower defect densities on a timescale $\simeq 10^6$LBU CONFIRM FOR B.
+If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur on a timescale $XXX$LBU.
+This finding offers further support for a dynamical equilibrium between defect creation and annihilation above a critical (or crossover) shear rate, below which defects progressively annihiliate and near-perfect alignment eventually results.
+To test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
+The defect density indeed goes up (not shown) and reaches similar values to before.
+Note however that the defect annealing time (of order $10^6$LBU) is quite long compared to any inherent relaxation time; also, the final state after switchoff strongly depends on the previous flow.
+Thus the rheology and microstructure of these materials is inherently history dependent.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
-\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
-Horizontal lines indicate time averages of the particular quantities, where we assume the system has reached a steady state.
-The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D in a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
+Horizontal lines indicate time averages after entering the apparent steady state (where present).
+The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right) CONFIRM.}
 \label{fig6}
@@
 
-This finding is in further support of a dynamical equilibrium between defect creation and annihilation above a critical threshold, wheras below the critical value defects cancel and lamellae align.
-To test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
-The defect density goes indeed up (not shown) and reaches similar values as before.\\
-Fig \ref{fig7} shows the ensemble averaged total shear stress $\langle \sigma_{xy}\rangle \equiv\sigma_{xy} $, i.e. the total shear stress per unit lattice volume.
-At low shear rates stress fluctuations become actually more stretched in time and smaller, a fact that is obscured by the logarithmic scales.
-In larger simulations such as those shown in Fig. \ref{fig1} and Fig. \ref{fig2} they are also about three times smaller.
-Nevertheless we found that the averaged shear stress is to a very good degree independent of the system size, which makes it possible to simulate longer times using smaller system sizes without flawing the results.
+Fig.~\ref{fig7} shows the spatially [CONFIRM] averaged total shear stress $\langle \sigma_{xy}({\bf r})\rangle \equiv\sigma_{xy} $.
+At low shear rates, stress fluctuations have longer temporal correlations but smaller amplitudes.
+(The latter dependence is obscured by the logarithmic vertical scale in this plot.) The root-mean-square temporal fluctuations can be reduced by increasing system size (by a factor 3 or so for the runs shown in Fig.
+\ref{fig1} and Fig.
+\ref{fig2}).
+In contrast, the averaged shear stress is almost independent of the system size, which makes it possible to simulate longer times using smaller system sizes while still giving accurate predictions for the bulk rheological behaviour.
+We pursue this strategy here.
 
@@
 \caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$, we find $\eta_{app}/\eta\equiv\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
  }
@@
 
-System A-2D shows clear shear thinning behaviour and seems to obey a power law (see below Fig. \ref{fig11}).
-Values of the apparent viscosity are given in the caption of Fig. \ref{fig7}.
-Contrary to system A-2D system B-2D does not exhibit a clear tendency of shear thinning.
-Here, the apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow a more horizontal trend.\\
-We want to focus now on the deviatoric part of the chemical pressure $\langle P^{chem}_{xy}\rangle\equiv P^{chem}_{xy}$.
-This is shown for system A-2D in Fig. \ref{fig8}.
-The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
-In System B-2D $P^{chem}_{xy}$ decreases sharply during the run with $\gd=1\e{-4}$ (not shown) because at this shear rate pronounced defect annihilation sets in.\\
-
 \begin{figure}[htp!]
@@
 \includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A-2D.
-The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\caption{Average order parameter contribution to the pressure tensor $P_{xy}$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).}
 \label{fig8}
 \end{figure}
-
-Fig. \ref{fig9} illustrates dependence of $P_{xy}$ on the defect density $\rho_D$.
-System A-2D features a linear relation between both quantities that breaks only down at the lowest defect density or rather shear rate.
-The result for the softer system B-2D are less clear due to the ordering and annihilation of defects.
-It is very likely that our quantifying procedure, which is based on averaging over a larger number of defects sites,is prone to inaccuracy in this regime.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs. defect density for system A-2D and B-2D } 
+\caption{Average chemical contribution to the pressure tensor $ P_{xy}$ vs. defect density for system A-2D. REMOVE B-2D. PROBABLY MERGE 7, 8, 9 INTO SINGLE FIGURE.} 
 \label{fig9}
@@
 
-Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A-2D, where we followed the same protocol as before.
-Initially, down till $\gd=5\e{-5}$ $N_1$ decreases monotonously with decreasing shear rate.
-However, at lower shear rates the behaviour is not monotonous anymore and $N_1$ rises again.
-Hence, it is difficult to define a reasonable time average for $N_1$ as for these low shear rates the system may still evolve at the end of the run.
+System A-2D shows clear shear thinning behaviour; values of its apparent viscosity at various shear rates are given in the caption of Fig.~\ref{fig7}.
+The corresponding range of $Ca$ is XXXX-YYYY.
+Using Eq.\ref{fullstress}, we can identify the stress contribution arising from the deviatoric part of the order parameter pressure tensor
+$\langle P_{xy}({\bf r})\rangle\equiv P_{xy}$; this is shown in Fig.~\ref{fig8} RESOLVE SIGN ISSUE.
+The temporal fluctuations are in phase with those of the total stress in Fig.~\ref{fig7} and decrease with decreasing shear rate.
+Fig. \ref{fig9} plots $P_{xy}$ against the defect density $\rho_D$; this shows a linear relation across almost the whole range of shear rates.
+
+In contrast, system B-2D does not exhibit clear shear thinning; the apparent viscosities obey $\eta_{app}/\eta=1.57, 1.62, 1.55, 1.42, 1.87$ for the same set of shear rates as used in Fig.~\ref{fig7}.
+To within the measured accuracy, system B-2D is rheologically quasi-Newtonian.
+However, the range of $Ca$ in this case is lower, ZZZZ-TTTT, and the order parameter stress creates only a relatively modest uplift to the baseline viscosity.
+More significantly, $Ca\le Ca_c$ throughout this range CONFIRM so that the steady states found in system B-2D are far more ordered than in A-2D.
+This results from a defect annihilation process, occurring in the early stages of ramp-down, which we found to be accompanied by a sharp drop in $P_{xy}$ (not shown).
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
-\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{yy}-\sigma_{xx}\rangle$ vs time for system A-2D.
-The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
+\caption{Spatially averaged first normal stress differences $N_1= \sigma_{yy}-\sigma_{xx}$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).
+Horizontal solid lines are time averages.}
 \label{fig10}
@@
 
-This feature which appears somewhat erratic at first has actually a simple explanation. 
-The time sequence of the defect density in Fig. \ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$ and linked to the defect annihilation.
-When two defects meet and cancel the lamellar width increases locally by a certain amount.
-This causes frustration in the system as the incommensurability between the layers is not large enough to create a new lamellar layer and allow the system to snap back towards its equilibrium width.
-Hence the normal stresses $\sigma_{xx}$ become increasingly smaller and may even attain negative values, leading to gradually larger values of $N_1$.
-This explanation is further sustained by the fact that only $\sigma_{xx}$ seems to evolve anomalously slowly and all other components of the stress tensor reach their steady state values more quickly.
-We expect that these anomalous trends become considerably smaller in larger systems.
-System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
-This is likely to be a direct consequence of the lower order parameter in system B-2D and its softness with respect to system A-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
-In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
-System A-2D gives rise to shear thinning with the average shear stress of systems A-2D following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
-For system B-2D the best fit is actually obtained for an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning.
-This is because due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$ the chemical stress $P^{chem}_{xy}$ almost vanishes and the total stress is dominated by the hydrodynamic part.\\
+Fig.~\ref{fig10} shows averaged first normal stress differences $\langle N_1({\bf r})\rangle \equiv N_1$ versus time for system A-2D, where we followed the same stepwise decreasing flow protocol as before.
+For the first several steps $N_1$ decreases monotonously with decreasing shear rate.
+However, at the lowest shear rates the behaviour is not monotic: $N_1$ eventually rises again. (The resulting time averages, shown as horizontal lines, do not therefore reflect true steady states.)
+This somewhat surprising feature has a simple explanation. 
+The time sequence of the defect density in Fig.
+\ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$, and linked to the onset of defect annihilation (which happens only at low shear rates).
+When two defects meet and cancel, the local lamellar width alters by a certain amount.
+This causes frustration as the system is unable to snap back locally towards its equilibrium width.
+Hence the normal stresses $\sigma_{xx}$ falls, and may even attain negative values, leading to increasing values of $N_1$.
+This explanation is further supported by the fact that $\sigma_{xx}$ evolves anomalously slowly, whereas all other components of the stress tensor reach their steady state values more quickly.
+We have not explored the dependence on system size of this normal stress anomaly, but it may be partly a finite size effect caused by the relatively modest number of smectic layers within the simulated volume.
+IS THIS WHAT YOU MEANT?
+The normal stress in system B-2D (not shown) behaves differently: in this case $N_1$ falls monotonically during the defect annihilation phase. 
+It is not clear whether this difference should be attributed to the change in $\beta, Ca$ or $Pe$ between the two systems, or indeed a combination of those changes.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
-\caption{Flow curves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)}
+\caption{Flow curves: Time-averaged shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D ($\eta=8.33\times10^{-2}, -a=b=5\times 10^{-4}$).
+LABEL THIS PROPERLY}
 \label{fig11}
@@
 
-\subsection*{Results in Three Dimensions}
+Fig.~\ref{fig11} shows `flow curves' representing the shear rate dependence of time averaged stresses $\sigma_{xy}$ and $N_1$.
+This is calculated from the stepwise ramp data as detailed previously, so the $N_1$ results at low shear rates should not be taken as definitive CONFIRM.
+System A-2D shows shear thinning as already discussed, apparently following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
+A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}; interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, folding  and formation of defects.
+They report an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
+Our value of $m=0.87$ for the disordered flow regime is larger.
+One possible explanation for this discrepancy is that screw defects, which do not exist in our 2D simulations, play a key role for this theory.
+On the other hand the fitted power law must be treated with great caution since the fit includes data from both $\dot\gamma$ both above and below $\dot\gamma_c$ CONFIRM.
+It is somewhat surprising that the strong morphological changes around $\dot\gamma_c$ give no visible anomaly or feature in the flow curves.
+CONFIRM AGAIN -- THIS IS VERY ODD.
+%
+For system B-2D the data is consistent with an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning (as previously described). 
+This appears to be due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$; the order parameter stress $P_{xy}$ is then dominated by the background viscosity term $\eta\dot\gamma$.
+THAT MAKES THE RESULT FOR A-2D STILL MORE SURPRISING, NO?
 
-Because steady states require to account for large strains, the computational costs of a similar analysis of nonlinear rheology in three-dimensions are still prohibitive.
-In the following we compare 3D-systems with 2D-systems smaller than those of the previous paragraph but large enough to show qualitatively similar structure under flow.
+In neither system does $\sigma_{xy}(\dot\gamma)$ offer any clear signature of a yield stress, as would be expected to arise for a frozen-in defect texture in the static state.
+This can be understood if, as detailed previously, that texture is replaced by a fully aligned state at the lowest shear rates studied here: with the layering wavevector perpendicular to the flow, such a state should have a purely viscous response.
+On the other hand, one would expect $N_1$ also to vanish in that case; if this happens, it is at lower strain rates than encountered here.
+WE HAVE TO UNDERSTAND WHY THE NORMAL STRESS IS SO LARGE, ESPECIALLY IN B-2D WHERE YOU ARE CLAIMING THAT THE CHEMICAL STRESS IS NEGLIGIBLE!!
+
+\subsection{Three Dimensions}
+\label{results3D}
+
+The computational cost of repeating all the above studies in three dimensional systems of equivalent size remains prohibitive, particularly when addressing steady state properties which emerge only after a million or more timesteps.
+We therefore focus in 3D on smaller systems (Table \ref{tab1}), but still large enough in principle to show qualitatively similar structure under flow.
+Our studies use two sets of parameters.
+System A-3D is identicaly to A-2D except for the introduction of the third dimension.
+System C-3D has parameters intermediate between A-2D and B-2D (Table \ref{tab1}) and was chosen in preference to the latter because
+preliminary studies on B-3D showed very rapid alignment into a fully ordered state.
+CONFIRM THAT, OR GIVE ANOTHER GOOD REASON.
+(We see below however that in fact this is a rather general outcome in 3D.)
 
@@
 \includegraphics[angle=0,width=0.45\textwidth]{phi_run1003_960k.png}
-\caption{Order parameter $\phi$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\cdot10^{-4}$) at $t=6.4\e{5}$.
+\caption{Order parameter $\phi({\bf r})$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\times10^{-4}$) at $t=6.4\e{5}$.
 The opacity of one phase has been reduced to enable a better view into the lamellar structure.
-The bottom left picture shows System C-2D at the same time step and shear rate.
-The bottom right picture is system A-2D at $\dot{\gamma}=1\cdot10^{-4}$ and time step $t=9.6\e{5}$.
-Results for A-3D are morphologically identical to those above.}
+The bottom left picture shows system C-2D at the same time step and shear rate.
+The bottom right picture is system A-2D at $\dot{\gamma}=1\times10^{-4}$ and time step $t=9.6\e{5}$.
+Results for A-3D are morphologically indistinguishable from those for C-3D.
+In all these images the flow direction is horizontal, to the right in the upper half of the sample and to the left in the lower half.
+CONFIRM}
 \label{fig12}
 \end{figure*}
-
-The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which is intermediate between A and B.
-The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $128\times256\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
-Two snapshots of system C-2D and A-2D in shear flow are shown in the bottom pictures in Fig. \ref{fig12}.
-They feature the typical fluctuating pattern of breaking and merging lamellae that was described in the previous paragraph and which is particularly pronounced in the case of system C-2D.
-The top picture visualise results for system C-3D with the same free energy parameters as C-2D.
-The quiescent state of system C-3D after equilibration is shown at the left with a morphology virtually identical to that of A-3D.
-The lamellar width of roughly 10 lattice sites is more or less the same as in two dimensions.
-A slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig. \ref{fig5}.
-However, in shear flow considerable differences arise between the two- and three-dimensional case.
-Contrary to the result in 2D, the three dimensional systems order readily into a regular lamellar structure after a transient period as shown in Fig. \ref{fig12} for system C-3D. 
-We opted for this system as the differences between the 3D- and 2D-system are the most striking we observed.
-Interestingly, these findings are consistent for even the highest shear rates, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
-This is consistent with the picture that layer undulation due to an instability in three-dimensional smectic liquid crystal layers emerges above a critical shear rates, which is well above those we have applied in the present study \cite{Guo2002,Soddemann2004, Stewart2009}.
 
@@
 \caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
-Shown are wave vectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
+Shown are wave vectors on the interval $k_x, k_z\in[-\pi/2 ,\pi/2 ]$.
+Data for C-3D (not shown) is almost the same.
+CONFIRM}
 \label{fig13}
@@
 \includegraphics[angle=0,width=0.35\textwidth]{ck_z-slice_run788_960.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\times10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 $, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 ,\pi/2]$.
 Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
@@
 
-However, the order is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers.
+Fig.~\ref{fig12} shows the post-equilibration amorphous structure and the steady-state structure under flow in system C-3D (A-3D is similar) at system size $N_x\times N_y \times N_z=128\times256\times128$.
+The latter is compared with C-2D and A-2D in a system of the same $x$ and $y$ dimensions ($128\times256\times1$).
+The quiescent metastable state of system C-3D after equilibration is more or less the same as it would be in two dimensions.
+Indeed, a slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig.
+~\ref{fig5}.
+However, under shear at $\dot\gamma = 2\times 10^{-4}$ ($Pe = XXX$, $Ca = YYY$) this structure anneals readily into a regular lamellar structure after a transient period of about XXX timesteps.
+This occurs for even the highest shear rates applied, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
+The same behaviour was seen CONFIRM in system A-3D for which the highest flow rate used corresponds to $Pe = XXX$ and $Ca = YYY$.
+(In contrast, the 2D results at the same flow rates and system sizes feature the typical fluctuating pattern of breaking and merging lamellae that was described previously and is visible in the lower panels of Fig.~\ref{fig12}.) The structure factor seen in 3D is accordingly close to that of a perfectly ordered lamellar state (see Fig.~\ref{fig14}) in contrast to the corresponding results in 2D (Fig.~\ref{fig5}).
+These findings suggest that any characteristic strain rate $\dot\gamma_c$ at which defects proliferate must be markedly higher in 3D than in 2D.
+Specifically, if $Ca$ is the relevant control parameter, our results for A-3D imply that $Ca_c > YYY$ in three dimensions.
+
+Note however that the order we observe is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers. WHAT DOES THAT MEAN?
 The Burgers vector is oriented along the gradient direction with edges along the flow direction.
-This forms a structure that is translational invariant along the flow direction and in the steady state the location of the edge defects does not change and the defects to not anneal.
+This forms a structure that is translationally invariant along the flow direction and in the final steady state the location of the edge defects does not change.
+Notably, the corresponding orientation of edge dislocations does not exist in two dimensions.
 A closer look at the kinetic evolution of the order parameter reveals that regions separated by different values of order parameter connect along a third direction by expelling regions with opposite order parameter.
-This is a direct consequence of the additional dimension and seems to play a key role during the formation of homogeneously layered structure.\\
-It is instructive to quantify the time it takes system A-3D to emerge as lamellar ordered structure starting from an isotropic initial configuration.  
-The inset of Fig. \ref{fig6} gives the defect density as defined by Eq. \ref{defect density} versus time for shear rates $\dot{gamma}=5\cdot10^{-5}-2\cdot10^{-4}$. 
-Within a few $10^5$ timesteps the defect density decreases to magnitudes, which we obtained in case of system A-2D for shear rates about one decade lower.
-The total strain where the curves level off and ordering is achieved scales with the shear rate and is about $\gamma=1.3$ ($\dot{gamma}=2\cdot10^{-4}$), $\gamma=0.8$ ($\dot{gamma}=1\cdot10^{-4}$) and $\gamma=0.6$ ($\dot{gamma}=5\cdot10^{-5}$).
- \\
-A lower number of defects yields smaller chemical pressure, which is shown for the deviatortic part $P_{xy}$ in the inset of Fig. \ref{fig8}.
-This contribution to the total stress is dwarfed by the hydrodynamic part and system A-3D exhibits shear thinning almost to an ultimate degree.
-A relation between the magnitude of the deviatoric part of the chemical pressure and the defect density like that in Fig. \ref{fig9} in the two-dimensional case cannot be deduced.\\
+This is a direct consequence of the additional dimension and seems to play a key role in promoting the formation of the homogeneously layered structure.
+I DO NOT UNDERSTAND ALL THIS: DOES IT CORRESPOND TO MOTION OF THE ORIENTED EDGE DISLOCATIONS? IF NOT, WHAT ROLE DO THOSE PLAY DURING THE ANNEALING?
+EXPLAIN CAREFULLY WHY THIS MECHANISM SPEEDS THINGS UP IN 3D BUT NOT IN 2D.
+
+We next quantify the time system A-3D takes to anneal into its final ordered structure starting from an isotropic initial configuration.
+(These and subsequent measurements are reported for A-3D only; C-3D is qualitatively similar CONFIRM.)
+The inset of Fig.~\ref{fig6} gives the defect density as defined by Eq.~\ref{defect density} versus time for shear rates $\dot{\gamma}=5\times10^{-5}-2\times10^{-4}$. 
+Within a few hundred thousand LB timesteps the defect density decreases by two orders of magnitude, resembling the behaviour seen in system A-2D at shear rates one decade smaller.
+The total strain where the curves level off, and ordering is achieved, scales with the shear rate and is about $\gamma=1.3$ (at $\dot{\gamma}=2\times10^{-4}$), $\gamma=0.8$ (at $\dot{\gamma}=1\times10^{-4}$) and $\gamma=0.6$ (at $\dot{\gamma}=5\times10^{-5}$).
+CONFIRM THESE STRAIN NUMBERS, THEY SEEM UNBELIEVABLY SMALL!!!
+
+The reduced number of defects is reflected in the magnitude of the order parameter stress contribution $P_{xy}$; the 3D behaviour is shown in the inset of Fig.~\ref{fig8}.
+The final contribution to the total stress is created by a small number of remaining defects (hence the inconsistent trend of the tiny plateau value with strain rate, which is not a concern here CONFIRM) and is quite negligible compared to the background viscosity contribution $\eta\dot\gamma$.
+The drop in $P_{xy}$ by three orders of magnitude during the defect annihilation process can formally be viewed as shear thinning of a quite extreme degree -- albeit one masked, in any macroscopic measurement of $\sigma_{xy}$, by the Newtonian background.
+
 First normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig10}.
-For all shear rates $\gd5\e{-5}-2\e{-4}$ system A-3D takes similar values, which are monotonously approached and about one decade smaller than those of system A-2D.\\ 
+THIS IS THE WRONG DEFINITION EVEN WITH THE CRAZY COORDINATE SYSTEM ADOPTED IN THIS PAPER.
+N1 IS FLOW MINUS GRADIENT NOT FLOW MINUS VORTICITY.
+For all shear rates addressed here system A-3D attains similar values of the normal stress; these are approached monotonically and about one decade smaller than those seen in system A-2D.
+The mechanism reported previously for the buildup of normal stress on approaching the ordered state in 2D systems seemingly does not apply in 3D, perhaps because of the much more efficient pathways to relaxation and mutual annihilation of defects.
+
+CAN WE PLEASE MEASURE THE NORMAL STRESS IN A COMPLETELY DEFECT FREE STATE AND COMPARE WITH THE OBSERVED VALUES.
+DOES THE PLATEAU CORRESPOND TO A FEW REMAINING DEFECTS OR DOES IT STEM FROM THE PERFECT BEHAVIOUR??
+
+\if{ 
 Finally we want to glimpse at the structure factor of system A-3D in shear flow.
@@
 \includegraphics[angle=0,width=0.45\textwidth]{ck_run788_960_zoom.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\times10^{-4}$.}
 \label{fig15}
@@
 
-A comparison of these results with experimental work is possbile although the topology is different in two and three dimensions.
-Our 2D-systems correspond to are translational invariant along the third dimension.
-Every branching point or endpoint of a lamella has the topology of an edge defect in the 3D-continuation of the structure.
-A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
-Interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, floding  and formation of defects. 
-For the power law dependence of the shear stress on the shear rate Lu et al.
-find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
-Our value of $m=0.87$ appears to be slightly too large.
-A possible explanation for this is that screw defects, which do not exist in our 2D-picture, play a key role for this theory.\\
+}\fi
 
-\section{Conclusions}
+\section{Conclusions}\label{conclusions}
 
-We studied nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model for binary liquid-crystalline fluids, which is based on a Brazovskii-free-energy-functional.
-The lamellar structure of the order parameter in the quiescent state agrees well with previous results using the same model in two dimensions and similar parameters.
-When flowing the morphology of the order parameter depends crucially on the system size.
-At system sizes $N_x\times N_y\times N_z=128\times256\times 1$ the system showed a strong tendency to align with the shear flow at low shear rates $\gd\le5\e{-5}$.
-At higher shear rates and for larger system sizes of $N_x\times N_y\times N_z=256\times512\times 1$ and above heterogenous lamellar structures with strongly extended and compressed regions were present in the steady states.
-A similar dependence of these charactistic patterns on the system size has been recently reported in \cite{Kumaran2011}.
-The patterns occured in distinctive bands and were oriented along the direction of extension and perpendicular to the direction of compression.
-The closer the system was to the critical point the more this feature gave way to flow alignment.
-The extended regions with lamellar spacing larger than at equilibrium were local regions with either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
-Depending on the applied shear rate and in agreement with other studies we found two different dynamical regimes. 
-Below a critical value, which was basically controlled by the magnitude of the order parameter and the distance from the critical point, alignment in flow direction and annealing of defects to a very large degree was obtained.
-Beyond the critical shear rate a buckling instability occured.
-The systems continuously tripped over itself, which resulted in a dynamical equilibrium between defect annihilation and creation of new defects.
-The shear stress in system A-2D follwed a clear power law dependence on the strain rate like $\sigma_{xy}=\gamma^\alpha$ with $\alpha=0.87$, giving rise to shear-thinning behaviour.
-Interestingly, the magnitude of chemical contribution $P_{xy}$ to the shear stress could be directly related to the density of defects.
-In system A-2D first normal stress differences became smaller with decreasing shear rate as long as the system was above the buckling instability.
-Below the critical shear rate continued defect annealing caused a rise in the first normal stress differences due to incommensurabilities in the lamellar width.\\
-We simulated flow behaviour of the same systems in three dimensions for the first time and found striking differences between the two- and three-dimensional versions of systems A and C.
-The 3D-system showed none of the distictive bands along the extensional direction and no buckling instability was found in the range of applied shear rates.
-The structure, initially amorphous after spinodal decomposition, ordered into a lamellar array with edge defects oriented along the flow directions.
-This is in total contrast to the 2D-case, where numerous defects appeared that would correspond to edge defects that are oriented perpendicular to the flow in 3D.
-Our results agree with theoretical values and simulations that predict a layer undulation instability at shear rates about two orders of magnitude larger than the ones we have applied.
-This constitutes a surprising change in results due to the different dimensonality and topoloogy of the strucutre despite identical parametrisation of the model.
+Many earlier studies have proposed a picture in which a layer undulation instability in three-dimensional smectic liquid crystals emerges above a critical shear rate [A G Zilman and R Granek, EPJB 11, 593-608 (1999), L Courbin, J P Delville, J Rouch and P Panizza, PRL 89, 148305 (2002)]\cite{Gonnella98,Kumaran2001} [Ramaswamy].
+Molecular dynamics simulations on thermotropic smectics place this in a regime of flow rates well above those we have applied in the present study \cite{Guo2002,Soddemann2004, Stewart2009}  CONFIRM THIS WITH NEW PARAMETER MAPPING -- IT COULD BE WRONG.
+The status of this instability and its effects on rheology in experiments remains somewhat unclear.
+Also unclear is its status in two dimensions.
+
+In the present work we have studied the nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model.
+After quenching from a homogeneous phase a metastable amorphous lamellar structure appears, similar to previous simulations in two dimensions using comparable models and parameters [REFS].
+Also in accord with some previous studies [REF], under flow in two dimensions we found the morphology of the order parameter to show two distinct regimes: one with very few defects which arises at low shear rates, and a defect rich phase at higher ones.
+We found that the critical (or crossover) strain rate $\dot\gamma_c$ separating these phases depends somewhat on system size, but much more strongly on thermodynamic and kinetic parameters.
+We have argued that this dependence might be summarized in terms of a capillary number $Ca$ (Eq.~\ref{capillary}) that is the ratio of viscous stress to the gradient free energy density of the lamellar phase.
+This controls a steady-state elastic strain in the smectic which, when it exceeds some characteristic value $Ca_c$, might cause yielding of layers and defect proliferation; our best estimate is $Ca_c \simeq 0.03$.
+However we emphasize that the evidence for $Ca$ as a control parameter is currently very limited and it would take a painstaking exploration of parameter space to confirm this picture.
+
+A notable feature of the disordered flow phase is our observation of striated heterogeneity in the state of smectic order, with strongly dilated and compressed regions oriented along the extensional axis of the flow.
+These are present in steady state but are nonstationary.
+The extended regions with lamellar spacing larger than at equilibrium exhibit either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
+[PLEASE EXPLAIN THE FOLLOWING SENTENCE
+A similar dependence of these charactistic patterns on the system size has been recently reported in \cite{Kumaran2011}.]
+These distortions of the order parameter contribute a term $P_{xy}$ to the shear stress whose magnitude could be directly related to the density of defects.
+In one system we found shear thinning described by a power law dependence of stress on strain rate, $\sigma_{xy}=\gamma^m$ with $m=0.87$; however the simplicity of this result may be fortuitous since the range of strain rates studied spans $\dot\gamma_c$ CONFIRM.
+In another system, at strain rates primarily below $\dot\gamma_c$, the shear rheology was quasi-Newtonian ($m=1$) although a significant first normal stress difference $N_1$ was also present.
+
+We then simulated the flow behaviour of directly comparable systems in three dimensions, and found striking differences between the two- and three-dimensional response under conditions of matched $\dot\gamma$.
+The 3D-system showed none of the distictive dilation/compression bands seen in 2D; moreover the microstructure, initially amorphous after quenching from the homogeneous phase, rapidly ordered under shear into a lamellar array with a low density of edge defects, all of which were oriented along the flow directions.
+This contrasts with the 2D case where numerous edge dislocations are present in the disordered phase ($\dot\gamma > \dot\gamma_c$) but, by virtue of the 2D geometry, all are oriented along the vorticity direction.
+(In saying this, we imagine a 3D continuation of the 2D structure which is translationally invariant along that direction.) Our 3D results do not rule out a transition or crossover from ordered to disordered structures, but this is not seen within the range of flow rates addressed here.
+This means that, if such a transition is controlled by capillary number, its value at onset in 3D obeys $Ca_c>YYY$, a value notably higher than that observed in 2D.
+Since identical parameter sets were chosen, the same applies to any other dimensionless number that might control the onset of the defect-rich phase.
+
+Finally, we note that one recent theory of smectic rheology [LU] suggests this is controlled by screw dislocations which have no 2D analogue.
+Moreover, the residual dislocations seen in the 3D ordered phase have an orientation that also does not exist in 2D.
+It should be remembered that, in 3D, edge dislocations and screw dislocations are interconvertible: they have the same topological status and the nomenclature merely describes whether the defect line is oriented normal or parallel to the smectic layers [CHAIKIN+LUBENSKY].
+The ability of 3D defect lines to reorient as well as translate underscores their greater dynamical freedom compared to similar defects in a 2D geometry.
+It is thus tempting to attribute the marked difference in dynamics between 2D and 3D systems to the reduced constraints on defect motion that arise in 3D, allowing them to find and annihilate each other more easily.
 
@@
 \footnotesize{
+RSC STYLE FILE APPARENTLY STRIPS OUT MIDDLE INITIALS. THIS IS OFFENSIVE, PARTICULARLY TO AUTHORS LIKE AJ WAGNER WITH COMMON NAMES.
 \bibliography{smectics} %your .bib file
@@
 \end{document}
+
+\if{
+In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
+Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
+Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space or $1 {\rm m}=2\times10^8$ LBU.
+
+The energy scale is slightly more difficult to calibrate.
+We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
+This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
+The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
+The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
+Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
+This formula reads 
+\begin{equation}
+b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
+\end{equation}
+whereas $\rho_c$ is the number density of chains per unit volume,
+often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
+Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
+With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
+Hence the LB unit of energy density is about $10^{6} \rm{Pa}$ or $1\rm{Pa}=10^{-6}$ LBU.
+
+It is straightforward to determine the unit of mass. 
+A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
+Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.
+
+We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
+As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=(1\e{-3}) \times (1\e{6}) \times (4\e{16})\rm{LBU}\simeq4\e{19} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq6.3\e{9}$ LBU and one LB timestep $\Delta t$ corresponds to roughly $0.15 \rm{ns}$.
+}\fi

### 2011-11-11 papers/lmcproc/lmcproc.tex
Davide's comments added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1481 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt,twoside]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
\usepackage{iopams} 
\usepackage{graphicx, subfigure}
\usepackage{balance}
\usepackage{bm}
\usepackage{float}
\newcommand{\ex}[1]{\times10^{#1}}

\begin{document}

\title[]{Confined Cubic Blue Phases under Shear}
\author{O. Henrich$^{1}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}

\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
\ead{o.henrich@ucl.ac.uk}

\begin{abstract}
We study the behaviour of confined cubic blue phases under a shear flow, via lattice Boltzmann simulations. We focus on the two experimentally observed
phases, blue phase I and blue phase II. The disclination network of blue phase II continuously breaks and reforms under shear, 
leading to an oscillatory stress response in time. The oscillations are only regular for very thin samples. For thicker samples, the shear leads to a ``stick-slip'' motion
of part of the network along the vorticity direction. Blue phase I responds vastly differently: its defect network undergoes seemingly
chaotic rearrangements under shear, irrespective of system size. 
\end{abstract}

%Uncomment for PACS numbers title message
%\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
%\vspace{2pc}
%\noindent{\it Keywords}: Article preparation, IOP journals
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
% Comment out if separate title page not required
%\maketitle

\section{Introduction}
% Put \label in argument of \section for cross-referencing
%\section{\label{}}
Cholesterics are liquid crystals in which the local director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
As it is topologically impossible to continuously patch 3D space with double-twist cylinders, defects arise. These may organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}. There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, is likely amorphous~\cite{Henrich:2011a}).\\
BPs have been long considered as purely of academic interest due to their very narrow range of stability, but this has drastically changed since the creation of polymer- and temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the possibility of novel applications.
During the last few years considerable progress has been achieved regarding the behaviour of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under external fields \cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011}, and in the presence of colloidal particles \cite{Ravnik:2011a}. 
The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. However, our understanding of their dynamical behaviour under flow remains
very limited. The aim of this work is to contribute to fill this gap by studying, for the first time, the response of confined BP samples to a shear flow. \\
Flow in cholesterics is both strongly non-Newtonian and highly anisotropic. If a cholesteric helix is subjected to, for instance, a Poiseuille flow along the helical axis, for small pressure differences flow occurs mainly through ``permeation'', first investigated by Helfrich \cite{Helfrich:1969}. In the permeation mode the liquid crystal flows by leaving the director field virtually unchanged, and this leads to high dissipation and large viscosities. Marenduzzo et al. \cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. They also found that a strong secondary flow appears.
Rey \cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along the vorticity axis and found that at low Ericksen number travelling twist waves appear which lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and leaves a flow-induced nematic phase. 
Rey also studied cholesterics subjected to steady and low frequency small amplitude oscillatory shear for different helix orientations \cite{Rey:2000, Rey:2002}, where he found that splay–bend–twist deformations were excited when the helix is aligned along the flow direction. Splay–bend deformation occured when the helix is aligned along the velocity gradient, and only twist deformations when the helix is aligned along the vorticity axis. \\
Dupuis et al. \cite{Dupuis:2005} performed the first numerical investigation of BP rheology in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic array of doubly twisted cylinders.
Under small forcing, the network opposed the flow giving rise to a significant increase in apparent viscosity.
Upon increasing the forcing they found clear evidence of shear thinning.
In the crossover region they predicted a novel oscillatory regime where the network continuously breaks and reforms as portions of the disclinations in the center of the channel move to neighboring cells and relink with the parts of the network left behind by the flow. With respect to the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than in cholesterics in the permeation mode, which is in agreement with experiments experiments \cite{Zapotocky:1999, Ramos:2002}.\\
This paper is organised as follows: 
In Sec. \ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics of liquid crystals, which is a generalisation of the Eriksen and Leslie's theory of nematodynamics~\cite{deGennes}.
In Sec. \ref{sec_results} we describe the simulation methodology and boundary conditions we applied.
Results of a confined BPII in rectilinear shear flow are presented in Sec. \ref{sec_bpii} with particular focus on the dependence of the thickness of the sample and thermodynamic state.
In Sec. \ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.

\section{Model}\label{sec_model}

Our approach is based on a well-established model for the hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes their state in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation the order parameter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$ with ${\mathbf n}$ the director field.
Brackets indicate mesoscopic averaging over a large number of individual molecules.
The largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$ characterizes the local degree of orientational order.
The thermodynamical properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$ --
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to \cite{Alexander:2006}.\\
A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
which are given by
\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
\end{eqnarray}

The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{eqn2}
\end{equation}
The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})Tr({\mathbf Q W}),
\label{eqn3}
\end{eqnarray}
where $Tr$ denotes the tensorial trace and ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and ${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively. $\xi$ is a constant depending on the molecular details of the liquid crystals.
Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our runs.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$.
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{eqn4}
\end{equation}
The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$.\\
The time evolution of density and velocity are respectively governed by the continuity and the following 
Navier-Stokes equation Eq.\ref{eqn6}, 
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}. 
\label{eqn6}
\end{eqnarray}

At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq.\ref{eqn6} can be safely neglected.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}
In the isotropic state when ${\bf Q}\equiv 0$ Eq.\ref{eqn7} is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Marenduzzo:2007} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the continuity equation and Eq.\ref{eqn6} and a finite difference scheme for the equation of motion of the tensor order parameter Eq.\ref{eqn2}.
More details on the algorithm can be found in \cite{Denniston:2001, Denniston:2004}.

\section{Methodology and Results}\label{sec_results}

In the following we present results from simulations of confined cubic blue phases in steady shear flow.
Our system consists of $n_y\times n_z = 4\times4$ unit cells in y- and z-direction (flow and vorticity direction respectively) and a varying number $n_x$ of unit cells along a gap in x-direction (which is the flow gradient direction in our geometry).
As is common in BP simulation work~\cite{Henrich:2010b,Henrich:2011a}, we initialised our simulations with analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, and equilibrated these (for 3000 LB timesteps) with periodic boundary conditions prior to starting the shear. During the equilibration run the optimal redshift $r$ was recalculated
at every timestep.

Shear flow was then started by applying shear boundary conditions to the LB distribution functions as in~\cite{Denniston:2004}. As regards ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities. Periodic boundary conditions were retained along the flow direction and the direction of vorticity.
This set-up is similar to the one Dupuis et al. used to simulate Poiseuille flow \cite{Dupuis:2005}. The pinning of the defect network at the boundaries may be realised in practice by impurities or surface defects.
During shear flow the unit cell size was kept fixed and a constant redshift equal to the value at the end of the equilibration protocol was assumed.\\ 
The total number of unit cells in our simulations was $n_x\times n_y \times n_z= 1, \dots, 4 \times 4 \times 4$ for both blue phases, using a resolution of 32 lattice Boltzmann units (LBU) per pitch length (16 per unit cell) for BPII and 64 LBU per pitch length (32 per unit cell) for BPI.  
This provided enough spatial resolution to track even complex rearrangements of the disclination networks.
A timestep and lattice spacing in LBU corresponds to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI-units.
The LBU of stress equals to about $10^8 Pa$. 
More details about the conversion from LBU to SI-units can be found in \cite{Henrich:2010b,Henrich:2011a}.

\subsection{Blue Phase II}\label{sec_bpii}

\begin{figure}[t]
\centering
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_25k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_50k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_75k_y.png}
\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_100k_y.png}
\caption{Shear flow of BPII at different gap sizes: The pictures show a section of the entire disclination network of $n_x=1 \mbox{ (yellow)}, 2 \mbox{ (green)}, 3 \mbox{ (red) and } 4 \mbox{ (blue)}$ unit cells along the gap and $n_y \times n_z =2 \times 2$ unit cells in flow and vorticity direction. The thermodynamic parameters temperature and chirality are $\tau=-0.5$ and chirality $\kappa=2$, respectively, where BPII is the equlibrium phase. The velocity gradient is oriented along the vertical direction with the top (bottom) wall moving into (out of) the viewing plane. The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ for all gap sizes.}
\label{fig1}
\end{figure}

We first report the results for BPII.
Fig.\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
(We prefer the perspective along the flow direction because the network topology is clearer then.)
Temperature and chirality are $\tau=-0.5$ and $\kappa=2$ respectively, and for these parameters BPII is the equilibrium phase.
The velocity of the walls has been adapted so that the same shear rate $\dot{\gamma}=7.8125\ex{-5}$ is achieved for all gap sizes.
Shear rates four times larger and smaller were also applied and produced very similar results, apart from the flow velocities which scale up accordingly.
For all gap sizes we observe an affine transformation in gradient-flow plane that leads to a break-up of the network at strains of about half a unit cell size.
Shorlty after the break-up the network reforms further downstream, resulting in a periodic, regular structure that is very close to the equilibrium
BPII disclination network. Similar oscillations have been previously reported for
BPs in Poiseuille flow, with a gap of one unit cell size \cite{Dupuis:2005}.\\
It is interesting to compare the results of different gap sizes.
For $n_x=1$ we observe that the BPII-network is slightly deformed with respect to the quiescent state, but reforms always at the same position.
At gap sizes of more than one unit cell the BPII network distorts more under the same shear rate, 
suggesting that the structure within the sample is highly affected by the anchoring at the boundaries.
Interestingly, when the sample is thicker than one unit cell a movement of the network in negative vorticity direction arises, 
which is not observed for $n_x=1$.
This network motion does not proceeds at a constant speed, rather it proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up, and are then 
quickly released as the network ``slips'' along the vorticity direction.
Changing from a left- to a right-handed helix inverts the direction of motion of the network.
Hence the ``stick-slip'' motion is related to the chiral nature of the helix. This phenomenon may also bear some similarities to
the travelling wave motion in cholesterics sheared along the vorticity direction 
below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
It is interesting to note that the secondary flow in the BP
does {\em not} match the direction of the ``stick-slip'' motion of the network, which is then not connected to mass flow
(in this respect this phenomenon is close to a permeation flow).

\begin{figure*}[t]
\centering
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_25k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_75k.png}
\includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_125k.png}
\caption{BPII in shear flow for different thermodynamic parameters: The images show the network as it appears along the flow direction at the reference state (blue, $\tau=-0.5, \kappa=2$), a high temperature state (red, $\tau=0.5, \kappa=2$) and a low chirality state (green, $\tau=-0.5, \kappa=1$), where BPII is metastable. The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all three cases. The pictures show a central section consisting of 8 unit cells from a larger run with $n_x=4$ gap size.}
\label{fig2}
\end{figure*}

The speed of the network in vorticity direction depends on the thermodynamic state of the blue phase.
A comparison between the state from Fig. \ref{fig1} with a ``high temperature'' state at $\tau=0.5, \kappa=2$ and a ``low chirality state'' at $\tau=-0.5, \kappa=1$
is shown in Fig.\ref{fig2}. Note that in both these states BPII is now a metastable phase. 
The images in Fig.~\ref{fig2} give a sequence of initially overlapping disclination networks that separate during the course of the simulation.
Of all three states the high temperature state is the fastest one, followed by the reference state and the low chirality state.

\begin{figure*}[t]
\centering
\includegraphics[width=0.75\textwidth]{stress_bp2_fbc.pdf}
\caption{BPII: Ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$ vs. time in the reference state ($\tau=-0.5, \kappa=2$) for gaps ranging from one to four unit cells across.}
\label{fig3}
\end{figure*}

The total network displacement along the vorticity direction is more than a unit cell within $10^5$ timesteps, for the ``high-temperature'' state.
Typical average and maximum values of the scalar order parameter are $\bar{q}_s=0.338, q_{s,max}=0.383$ (low chirality state), 
$\bar{q}_s=0.206, q_{s,max}=0.320$ (high temperature state) and $\bar{q}_s=0.304, q_{s,max}=0.408$ (reference state, in Fig.~\ref{fig1}).
The average order parameter of the ``low-chirality'' state is by about $10\%$ higher compared to the reference state, 
whereas the relative differerence between the mean and the maximum value is about a factor $2.5$ smaller.
We propose that the speed of the networks is different because permeation, which should be required for the ``stick-slip'' motion, may occur more easily in states with less order (``high-temperature'' state) and gentler spatial variation (``low-chirality'' state). Furthermore, the uneven, ``stick-slip'' nature of the network motion is possibly due to the anchoring at the boundaries which prevents network displacements close to there.

Fig. \ref{fig3} shows the spatially averaged shear stress $\langle \Pi_{xy}\rangle$ versus time, in the reference state and for different gap sizes.
For gaps accommodating only one unit cell a regular periodic pattern emerges shortly after a transient start-up phase.
Every cycle is related to a break-up and reformation of the network.
The time-average in the steady state is about $\Pi_{xy}\simeq5.62\ex{-5}$ in LBU and yields an apparent viscosity $\eta_{app}=(1 +\Pi_{xy}/\eta\dot{\gamma})\simeq 1.86$. 
When the gap contains $n_x\ge2$ unit cells, the average stress shows less regular oscillations, and fluctuates around a mean value which is about $10-15\%$ smaller, roughly $\eta_{app}=1.65$ for $n_x=4$. The fact that the average stress stabilises for $n_x\ge2$ indicates that bulk flow behaviour should occur for those thicknesses, but not in the single unit cell case.
Different shear rates lead to similar generic features.

\subsection{Blue Phase I}\label{sec_bpi}

We now compare and contrast the results just obtained with those pertaining to the rheology of blue phase I. The topology of the disclination network in BPI in equilibrium is very different from that of BPII: the main difference is the lackin BPI of junction points where several (four in BPII) disclination lines of topological charge -1/2 meet. \\
In the following we give representative results for BPI at $\tau=-0.5$ and $\kappa=1$, where it is the equilibrium phase.
Fig.\ref{fig4} shows the disclination network at timestep $t=4\ex{5}$ for different gap sizes, with the viewing direction again chosen along the flow direction.
For clarity we only show $n_x \times n_y \times n_z=1,\dots4\times1\times2$ unit cells.

\begin{figure*}[h]
\centering
\begin{minipage}[b]{0.3\textwidth}
\subfigure[]{\includegraphics[width=\textwidth]{disc_bp1_t-0o5_k1o0_400k_1uc.png}}
\subfigure[]{\includegraphics[width=\textwidth]{disc_bp1_t-0o5_k1o0_400k_2uc.png}}
\end{minipage}
\subfigure[]{\includegraphics[width=0.3\textwidth]{disc_bp1_t-0o5_k1o0_400k_3uc.png}} 
\subfigure[]{\includegraphics[width=0.3\textwidth]{disc_bp1_t-0o5_k1o0_400k_4uc.png}}
\caption{Confined BPI in shear flow at $\tau=-0.5, \kappa=1$ and and gap sizes of one (a, yellow), two (b, green), three (c, red) and four (d, blue) unit cells. The pictures show the defect network along the flow direction at timestep $t=4\ex{5}$ at the same flow rate as above.}
\label{fig4}
\end{figure*}

\begin{figure*}[h]
\centering
\includegraphics[width=0.75\textwidth]{stress_bp1_fbc.pdf}
\caption{BPI: Ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$ vs. time in the reference state ($\tau=-0.5, \kappa=1$) for gaps ranging from one to four unit cells across.}
\label{fig5}
\end{figure*}

Contrary to BPII, the disclination network of BPI changes greatly under a shear flow, and the typical snapshots in Fig.~\ref{fig4} bear no relation to the equilibrium BPI configuration. Furthermore, there is no sign of the ``stick-slip'' motion of the defects observed in BPII. While no regular cubic structure emerges under flow, it is interesting to note that the network develops a periodic modulation along the vorticity direction (horizontal in Fig.~\ref{fig4}). This periodicity arises for all gap sizes. Whether this feature is a consequence of frustration with the anchoring at the boundaries, or is a prelude to a slow ordering of the network under flow, we cannot yet tell.

The ensemble averages of the shear stress $\Pi_{xy}$ for different gap sizes is shown in Fig.\ref{fig5} and gives more evidence of the chaotic nature of the flow of BPI.
The magnitude of the average shear stress as well as the drop when increasing the gap size seem are both comparable to those observed with BPI.

\section{Discussions and Conclusions}

In conclusion, we presented the first supra-unit cell simulations of confined cubic blue phases in simple shear flow.
The rheological behaviour of BPI and BPII was found to be strikingly different.

Our simulations show that the disclination network of BPII continuously breaks and reforms while the network undergoes an affine deformation under shear. This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient directions, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
For larger gap sizes of two and more unit cells a ``stick-slip'' motion of the network arises, along the vorticity direction. The associated flow is not directly linked to mass flow, and is therefore somewhat akin to permeation flow in cholesterics. The average speed of this network translation depends on the thermodynamic parameters, and appears to be higher for higher temperature and lower chirality.

Contrary to BPII, BPI shows no regular break-up and relinking of the unit cells and its disclination network undergoes chaotic rearrangements under shear.
This behaviour was found for all gap sizes, and may be linked to the largely different topology in the equilibrium disclination patterns in BPI and BPII.
We will adress the rheology of BPI more in detail in an upcoming publication \cite{Henrich:2012}.

% If you have acknowledgments, this puts in the proper section head.
\ack
We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time on HECToR, U.K.
M.E.C. holds a Royal Society Research Professorship.

% Create the reference section using BibTeX:

\section*{References}

\bibliography{lmcproc}
\bibliographystyle{unsrt}
\end{document}

### 2011-11-11 papers/lmcproc/lmcproc.tex
Various minor changes to text.
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1486 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
+\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, The University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, The University of Edinburgh, UK}
 \ead{o.henrich@ucl.ac.uk}
@@
 \begin{abstract}
-We study the behaviour of confined cubic blue phases under a shear flow, via lattice Boltzmann simulations. We focus on the two experimentally observed
+We study the behaviour of confined cubic blue phases under shear flow
+via lattice Boltzmann simulations. We focus on the two experimentally observed
 phases, blue phase I and blue phase II. The disclination network of blue phase II continuously breaks and reforms under shear, 
 leading to an oscillatory stress response in time. The oscillations are only regular for very thin samples. For thicker samples, the shear leads to a ``stick-slip'' motion
-of part of the network along the vorticity direction. Blue phase I responds vastly differently: its defect network undergoes seemingly
+of part of the network along the vorticity direction.
+Blue phase I responds very differently: its defect network undergoes seemingly
 chaotic rearrangements under shear, irrespective of system size. 
@@
 The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
-As it is topologically impossible to continuously patch 3D space with double-twist cylinders, defects arise.
-These may organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}.
-There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, is likely amorphous~\cite{Henrich:2011a}).\\
-BPs have been long considered as purely of academic interest due to their very narrow range of stability, but this has drastically changed since the creation of polymer- and temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the possibility of novel applications.
+As it is topologically impossible to patch continuously 3D space with double-twist cylinders, defects arise.
+These organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}.
+There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, is thought to be amorphous~\cite{Henrich:2011a}).
+
+BPs have been long considered as purely of academic interest due to their very narrow range of stability. This view has changed since the creation of
+polymer-stabilised and temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the possibility of novel applications.
 During the last few years considerable progress has been achieved regarding the behaviour of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under external fields \cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011}, and in the presence of colloidal particles \cite{Ravnik:2011a}.
 The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. However, our understanding of their dynamical behaviour under flow remains
-very limited. The aim of this work is to contribute to fill this gap by studying, for the first time, the response of confined BP samples to a shear flow. \\
+very limited. The aim of this work is to address this issue by studying,
+for the first time, the response of confined BP samples to a shear flow.
+
 Flow in cholesterics is both strongly non-Newtonian and highly anisotropic.
-If a cholesteric helix is subjected to, for instance, a Poiseuille flow along the helical axis, for small pressure differences flow occurs mainly through ``permeation'', first investigated by Helfrich \cite{Helfrich:1969}.
-In the permeation mode the liquid crystal flows by leaving the director field virtually unchanged, and this leads to high dissipation and large viscosities.
+For example, if a cholesteric helix is subjected to a Poiseuille flow along
+the helical axis, small pressure differences drive flow mainly through
+``permeation'', first investigated by Helfrich \cite{Helfrich:1969}.
+In the permeation mode the liquid crystal flows while leaving the director
+field virtually unchanged, which leads to high dissipation and large
+viscosities.
 Marenduzzo et al.
@@
 They also found that a strong secondary flow appears.
-Rey \cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along the vorticity axis and found that at low Ericksen number travelling twist waves appear which lead to the rotation of the cholesteric helix.
+Rey \cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which lead to the rotation of the cholesteric helix.
 At higher forcing, the helix uncoils and leaves a flow-induced nematic phase.
-Rey also studied cholesterics subjected to steady and low frequency small amplitude oscillatory shear for different helix orientations \cite{Rey:2000, Rey:2002}, where he found that splay–bend–twist deformations were excited when the helix is aligned along the flow direction.
-Splay–bend deformation occured when the helix is aligned along the velocity gradient, and only twist deformations when the helix is aligned along the vorticity axis.
-\\
+Rey also studied cholesterics subjected to both steady flow and low frequency
+small amplitude oscillatory shear for different helix orientations
+\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
+excited when the helix was aligned along the flow direction, splay/bend
+deformation occured when the helix was aligned along the velocity gradient,
+but only twist deformations
+appeared when the helix wass aligned along the vorticity axis.
+
 Dupuis et al.
@@
 In the crossover region they predicted a novel oscillatory regime where the network continuously breaks and reforms as portions of the disclinations in the center of the channel move to neighboring cells and relink with the parts of the network left behind by the flow.
-With respect to the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than in cholesterics in the permeation mode, which is in agreement with experiments experiments \cite{Zapotocky:1999, Ramos:2002}.\\
-This paper is organised as follows: 
-In Sec. \ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics of liquid crystals, which is a generalisation of the Eriksen and Leslie's theory of nematodynamics~\cite{deGennes}.
-In Sec. \ref{sec_results} we describe the simulation methodology and boundary conditions we applied.
-Results of a confined BPII in rectilinear shear flow are presented in Sec. \ref{sec_bpii} with particular focus on the dependence of the thickness of the sample and thermodynamic state.
-In Sec. \ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.
+Compared with the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than in cholesterics in the permeation mode, which is in agreement with experiments
+\cite{Zapotocky:1999, Ramos:2002}.
+
+This paper is organised as follows.
+In Sec.~\ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics of liquid crystals, which is a generalisation of the Eriksen and Leslie's theory of nematodynamics~\cite{deGennes}.
+In Sec.~\ref{sec_results} we describe the simulation methodology and the
+boundary conditions we applied.
+Results for a confined BPII in rectilinear shear flow are presented in
+Sec.~\ref{sec_bpii} with particular focus on the dependence of the flow
+on the thickness of the sample and thermodynamic state.
+In Sec.~\ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.
 
@@
 
-Our approach is based on a well-established model for the hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes their state in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$.
-In the uniaxial approximation the order parameter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$ with ${\mathbf n}$ the director field.
+Our approach is based on a well-established model for the hydrodynamics of
+cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999}. The model
+describes their state in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
+In the uniaxial approximation, the order parameter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$ with ${\mathbf n}$ the director field.
 Brackets indicate mesoscopic averaging over a large number of individual molecules.
 The largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$ characterizes the local degree of orientational order.
-The thermodynamical properties of the liquid crystal are determined by a free energy
+The thermodynamic properties of the liquid crystal are determined by a free energy
 ${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
 \begin{eqnarray}
-f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
-f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{FE}
+f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
+f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
 \end{eqnarray}
@@
 The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
-The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to \cite{Alexander:2006}.\\
+The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to \cite{Alexander:2006}.
+
 A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
@@
 \end{eqnarray}
-
 The dynamical evolution of the order parameter is given by the equation 
@@
 where $Tr$ denotes the tensorial trace and ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and ${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
-$\xi$ is a constant depending on the molecular details of the liquid crystals.
-Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our runs.
+$\xi$ is a constant depending on the molecular details of the liquid crystal.
+Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our simulations.
 ${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$.
@@
 \begin{eqnarray}
-\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}.
+\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
+= \partial_\beta \Pi_{\alpha \beta}
++\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
++(1+3\frac{P_0}{\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
 \label{eqn6}
@@
 \end{eqnarray}
-In the isotropic state when ${\bf Q}\equiv 0$ Eq.\ref{eqn7} is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
+In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
 The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Marenduzzo:2007} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the continuity equation and Eq.\ref{eqn6} and a finite difference scheme for the equation of motion of the tensor order parameter Eq.\ref{eqn2}.
@@
 Shear flow was then started by applying shear boundary conditions to the LB distribution functions as in~\cite{Denniston:2004}.
-As regards ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities.
+For ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities.
 Periodic boundary conditions were retained along the flow direction and the direction of vorticity.
@@
 The pinning of the defect network at the boundaries may be realised in practice by impurities or surface defects.
-During shear flow the unit cell size was kept fixed and a constant redshift equal to the value at the end of the equilibration protocol was assumed.\\ 
-The total number of unit cells in our simulations was $n_x\times n_y \times n_z= 1, \dots, 4 \times 4 \times 4$ for both blue phases, using a resolution of 32 lattice Boltzmann units (LBU) per pitch length (16 per unit cell) for BPII and 64 LBU per pitch length (32 per unit cell) for BPI.
-This provided enough spatial resolution to track even complex rearrangements of the disclination networks.
+During shear flow the unit cell size was kept fixed and a constant redshift equal to the value at the end of the equilibration protocol was assumed.
+
+The total number of unit cells in our simulations of both BPI and BPII
+was $n_x\times n_y \times n_z= \{1,2,3,4\} \times 4\times 4$.
+A resolution of 32 lattice Boltzmann units (LBU) per pitch length
+(16 per unit cell) was used for BPII, and 64 LBU per pitch length
+(32 per unit cell) used for BPI.  
+This provided enough spatial resolution to track even complex rearrangements
+of the disclination networks in each case.
 A timestep and lattice spacing in LBU corresponds to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI-units.
-The LBU of stress equals to about $10^8 Pa$. 
+The LBU of stress equals to about $10^8$~Pa. 
 More details about the conversion from LBU to SI-units can be found in \cite{Henrich:2010b,Henrich:2011a}.
@@
 \begin{figure}[t]
-\centering
 \includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_25k_y.png}
@@
 \includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_100k_y.png}
-\caption{Shear flow of BPII at different gap sizes: The pictures show a section of the entire disclination network of $n_x=1 \mbox{ (yellow)}, 2 \mbox{ (green)}, 3 \mbox{ (red) and } 4 \mbox{ (blue)}$ unit cells along the gap and $n_y \times n_z =2 \times 2$ unit cells in flow and vorticity direction.
-The thermodynamic parameters temperature and chirality are $\tau=-0.5$ and chirality $\kappa=2$, respectively, where BPII is the equlibrium phase.
-The velocity gradient is oriented along the vertical direction with the top (bottom) wall moving into (out of) the viewing plane.
-The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ for all gap sizes.}
+\caption{Shear flow of BPII at different gap sizes. The pictures show
+snapshots of a subset of the entire disclination network in a system
+with walls at top and bottom, and the flow direction into/out of the
+page. The different number of unit cells in the velocity gradient
+direction have been superposed and are shown in different colours:
+$n_x=1 \mbox{ (yellow)}, n_x=2 \mbox{ (green)}, n_x=3 \mbox{ (red) and }
+n_x=4 \mbox{ (blue)}$. All cases show $n_y \times n_z =2 \times 2$ unit
+cells in flow and vorticity directions. The temperature and chirality are
+$\tau=-0.5$ and $\kappa=2$, respectively, where BPII is the equlibrium phase.
+The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all cases.}
 \label{fig1}
@@
 We first report the results for BPII.
-Fig.\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
-(We prefer the perspective along the flow direction because the network topology is clearer then.)
+Fig.~\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
+(We prefer the perspective along the flow direction because the network topology is clearer.)
 Temperature and chirality are $\tau=-0.5$ and $\kappa=2$ respectively, and for these parameters BPII is the equilibrium phase.
@@
 BPII disclination network. Similar oscillations have been previously reported for
-BPs in Poiseuille flow, with a gap of one unit cell size \cite{Dupuis:2005}.\\
-It is interesting to compare the results of different gap sizes.
-For $n_x=1$ we observe that the BPII-network is slightly deformed with respect to the quiescent state, but reforms always at the same position.
-At gap sizes of more than one unit cell the BPII network distorts more under the same shear rate, 
-suggesting that the structure within the sample is highly affected by the anchoring at the boundaries.
-Interestingly, when the sample is thicker than one unit cell a movement of the network in negative vorticity direction arises, 
+BPs in Poiseuille flow, with a gap of one unit cell size \cite{Dupuis:2005}.
+
+It is interesting to compare the results for different gap sizes.
+For $n_x=1$ we observe that the BPII-network is slightly deformed with
+respect to the quiescent state, but always reforms at the same position.
+For $n_x > 1$, the BPII network distorts more under the same shear rate, 
+suggesting that the structure within the sample is strongly affected by
+the anchoring at the boundaries.
+Interestingly, when the sample is thicker than one unit cell, a movement
+of the network in the negative vorticity direction arises
 which is not observed for $n_x=1$.
-This network motion does not proceeds at a constant speed, rather it proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up, and are then 
-quickly released as the network ``slips'' along the vorticity direction.
-Changing from a left- to a right-handed helix inverts the direction of motion of the network.
-Hence the ``stick-slip'' motion is related to the chiral nature of the helix. This phenomenon may also bear some similarities to
-the travelling wave motion in cholesterics sheared along the vorticity direction 
+This network motion does not proceed at a constant speed, but rather
+proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up,
+and are then quickly released as the network ``slips'' along the vorticity
+direction. Changing from a left-handed to a right-handed helix inverts the
+direction of motion of the network.
+Hence the ``stick-slip'' motion is related to the chiral nature of the helix.
+This phenomenon may bear some similarity to the travelling wave motion in
+cholesterics sheared along the vorticity direction 
 below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
@@
 \includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_125k.png}
-\caption{BPII in shear flow for different thermodynamic parameters: The images show the network as it appears along the flow direction at the reference state (blue, $\tau=-0.5, \kappa=2$), a high temperature state (red, $\tau=0.5, \kappa=2$) and a low chirality state (green, $\tau=-0.5, \kappa=1$), where BPII is metastable.
+\caption{BPII in shear flow for different thermodynamic parameters. The
+snapshots show the evolution of the network with the flow direction into/out of
+the page.  The reference state (blue, $\tau=-0.5, \kappa=2$), a high
+temperature state (red, $\tau=0.5, \kappa=2$) and a low chirality state
+(green, $\tau=-0.5, \kappa=1$) are shown; BPII is metastable in the later
+two parameter sets.
 The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all three cases.
-The pictures show a central section consisting of 8 unit cells from a larger run with $n_x=4$ gap size.}
+The pictures show a central cubic section consisting of 8 unit cells cut
+from a larger simulation with $n_x=4$.}
 \label{fig2}
@@
 
-The speed of the network in vorticity direction depends on the thermodynamic state of the blue phase.
-A comparison between the state from Fig. \ref{fig1} with a ``high temperature'' state at $\tau=0.5, \kappa=2$ and a ``low chirality state'' at $\tau=-0.5, \kappa=1$
-is shown in Fig.\ref{fig2}. Note that in both these states BPII is now a metastable phase. 
-The images in Fig.~\ref{fig2} give a sequence of initially overlapping disclination networks that separate during the course of the simulation.
-Of all three states the high temperature state is the fastest one, followed by the reference state and the low chirality state.
+The speed of the network in the vorticity direction depends on the
+thermodynamic state of the blue phase.
+A comparison between the state from Fig.~\ref{fig1} with a
+``high temperature'' state at $\tau=0.5, \kappa=2$ and a
+``low chirality state'' at $\tau=-0.5, \kappa=1$ is shown in
+Fig.~\ref{fig2}. Note that in both these states BPII is a metastable phase. 
+The images in Fig.~\ref{fig2} show a sequence of initially overlapping
+disclination networks which separate during the course of the simulation.
+The network moves fastest in the high temperature state, is slower in the
+reference state and slower still in the low chirality state.
 
@@
 \includegraphics[width=0.75\textwidth]{stress_bp2_fbc.pdf}
-\caption{BPII: Ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$ vs. time in the reference state ($\tau=-0.5, \kappa=2$) for gaps ranging from one to four unit cells across.}
+\caption{BPII: ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$
+versus time in the reference state ($\tau=-0.5, \kappa=2$) for gaps ranging
+from one to four unit cells.}
 \label{fig3}
@@
 
-The total network displacement along the vorticity direction is more than a unit cell within $10^5$ timesteps, for the ``high-temperature'' state.
-Typical average and maximum values of the scalar order parameter are $\bar{q}_s=0.338, q_{s,max}=0.383$ (low chirality state), 
-$\bar{q}_s=0.206, q_{s,max}=0.320$ (high temperature state) and $\bar{q}_s=0.304, q_{s,max}=0.408$ (reference state, in Fig.~\ref{fig1}).
-The average order parameter of the ``low-chirality'' state is by about $10\%$ higher compared to the reference state, 
-whereas the relative differerence between the mean and the maximum value is about a factor $2.5$ smaller.
+The total network displacement along the vorticity direction is more than
+a unit cell in $10^5$ timesteps for the ``high-temperature'' state.
+Typical average and maximum values of the scalar order parameter are
+$\bar{q}_s=0.338$, $q_{s,max}=0.383$ (low chirality state), 
+$\bar{q}_s=0.206$, $q_{s,max}=0.320$ (high temperature state) and
+$\bar{q}_s=0.304$, $q_{s,max}=0.408$ (reference state, in Fig.~\ref{fig1}).
+The average scalar order parameter of the ``low-chirality'' state is about
+$10\%$ higher compared with the reference state. The  relative differerence
+between the mean and the maximum value is about a factor $2.5$ smaller.
 We propose that the speed of the networks is different because permeation, which should be required for the ``stick-slip'' motion, may occur more easily in states with less order (``high-temperature'' state) and gentler spatial variation (``low-chirality'' state).
-Furthermore, the uneven, ``stick-slip'' nature of the network motion is possibly due to the anchoring at the boundaries which prevents network displacements close to there.
+Furthermore, the uneven, ``stick-slip'' nature of the network motion is possibly due to the anchoring at the boundaries which prevents network displacements at the wall.
 
@@
 
-We now compare and contrast the results just obtained with those pertaining to the rheology of blue phase I.
-The topology of the disclination network in BPI in equilibrium is very different from that of BPII: the main difference is the lackin BPI of junction points where several (four in BPII) disclination lines of topological charge -1/2 meet.
-\\
-In the following we give representative results for BPI at $\tau=-0.5$ and $\kappa=1$, where it is the equilibrium phase.
-Fig.\ref{fig4} shows the disclination network at timestep $t=4\ex{5}$ for different gap sizes, with the viewing direction again chosen along the flow direction.
-For clarity we only show $n_x \times n_y \times n_z=1,\dots4\times1\times2$ unit cells.
+We now compare and contrast the results just obtained for BPII with those
+pertaining to the rheology of blue phase I. In equilibrium, the topology
+of the disclination network in BPI is very different from that of BPII.
+The main difference is presence of junction points where four disclination
+lines of topological charge -1/2 meet in BPII; no such junctions occur in BPI.
+
+In the following, we give representative results for BPI at $\tau=-0.5$
+and $\kappa=1$, where it is the equilibrium phase.
+Fig.~\ref{fig4} shows the disclination network at timestep $t=4\ex{5}$
+for different gap sizes, with the viewing direction again chosen along
+the flow direction.
+For clarity we only show $n_x \times n_y \times n_z=\{1,2,3,4\}\times1\times2$
+unit cells.
 
@@
 \includegraphics[width=0.75\textwidth]{stress_bp1_fbc.pdf}
-\caption{BPI: Ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$ vs. time in the reference state ($\tau=-0.5, \kappa=1$) for gaps ranging from one to four unit cells across.}
+\caption{BPI: ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$
+versus time in the reference state ($\tau=-0.5, \kappa=1$) for gaps
+ranging from one to four unit cells.}
 \label{fig5}
@@
 
-Contrary to BPII, the disclination network of BPI changes greatly under a shear flow, and the typical snapshots in Fig.~\ref{fig4} bear no relation to the equilibrium BPI configuration.
+In contrast to BPII, the disclination network of BPI changes significantly
+under a shear flow, and typical snapshots in Fig.~\ref{fig4} bear no
+relation to the equilibrium BPI configuration.
 Furthermore, there is no sign of the ``stick-slip'' motion of the defects observed in BPII.
@@
 This periodicity arises for all gap sizes.
-Whether this feature is a consequence of frustration with the anchoring at the boundaries, or is a prelude to a slow ordering of the network under flow, we cannot yet tell.
+Whether this feature is a consequence of frustration caused by the anchoring at the boundaries, or is a prelude to a slow ordering of the network under flow, we cannot yet tell.
 
-The ensemble averages of the shear stress $\Pi_{xy}$ for different gap sizes is shown in Fig.\ref{fig5} and gives more evidence of the chaotic nature of the flow of BPI.
-The magnitude of the average shear stress as well as the drop when increasing the gap size seem are both comparable to those observed with BPI.
+The ensemble averages of the shear stress $\Pi_{xy}$ for different gap sizes
+is shown in Fig.~\ref{fig5} and gives more evidence of the chaotic nature of
+the flow of BPI. The magnitude of the average shear stress is comparable to
+that seen in BPII; the fall in amplitude of the oscillations with increasing
+gap size is also similar.
 
@@
 
-In conclusion, we presented the first supra-unit cell simulations of confined cubic blue phases in simple shear flow.
+In conclusion, we have presented the first supra-unit cell simulations of
+confined cubic blue phases in simple shear flow.
 The rheological behaviour of BPI and BPII was found to be strikingly different.
@@
 Our simulations show that the disclination network of BPII continuously breaks and reforms while the network undergoes an affine deformation under shear.
-This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient directions, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
+This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient direction, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
 For larger gap sizes of two and more unit cells a ``stick-slip'' motion of the network arises, along the vorticity direction.
@@
 
-Contrary to BPII, BPI shows no regular break-up and relinking of the unit cells and its disclination network undergoes chaotic rearrangements under shear.
+In contrast to BPII, BPI shows no regular break-up and relinking of the unit cells and its disclination network undergoes chaotic rearrangements under shear.
 This behaviour was found for all gap sizes, and may be linked to the largely different topology in the equilibrium disclination patterns in BPI and BPII.
@@
 \ack
-We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time on HECToR, U.K.
+We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time on HECToR.
 M.E.C. holds a Royal Society Research Professorship.

### 2011-11-14 papers/lmcproc/lmcproc.tex
Comments from Mike added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1487 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+% TODO
+% OLIVER: Please check Olmsted reference: Mike says ``*not* in this paper''
+% at point of reference
+% DAVIDE (?): Page 6, sentence ``It is interesting to note that the
+% secondary flow in the BP does not match the direction...'' Mike says
+% ``incomprehensible''
+% ANYONE: Please add reference ''to explain strange p_0/rho bit in
+% equation 5''. 
+% OLIVER (lucky you): Please adjust to use standard rheological x,y,z
+% Cartesian coordinates. (y,z,x) at moment?
+
 \begin{document}
@@
 
-\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, The University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, The University of Edinburgh, UK}
+\address{$^1$ Centre for Computational Science, Department of Chemistry,
+University College London, WC1H 0AJ, United Kingdom\\
+$^2$ Edinburgh Parallel Computing Centre and
+$^3$ School of Physics and Astronomy, The King's Buildings,
+The University of Edinburgh, EH9 3JZ, United Kingdom}
 \ead{o.henrich@ucl.ac.uk}
@@
 %\section{\label{}}
-Cholesterics are liquid crystals in which the local director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
+Cholesterics are liquid crystals in which the local nematic director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
 The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
-As it is topologically impossible to patch continuously 3D space with double-twist cylinders, defects arise.
+As it is topologically impossible to cover continuously 3D space with double-twist cylinders, defects arise.
 These organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}.
@@
 
-Flow in cholesterics is both strongly non-Newtonian and highly anisotropic.
+Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
 For example, if a cholesteric helix is subjected to a Poiseuille flow along
@@
 \cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
-excited when the helix was aligned along the flow direction, splay/bend
-deformation occured when the helix was aligned along the velocity gradient,
+excited when the helix was aligned along the flow direction; splay/bend
+deformation occured when the helix was aligned along the velocity gradient;
 but only twist deformations
@@
 In the crossover region they predicted a novel oscillatory regime where the network continuously breaks and reforms as portions of the disclinations in the center of the channel move to neighboring cells and relink with the parts of the network left behind by the flow.
-Compared with the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than in cholesterics in the permeation mode, which is in agreement with experiments
+Compared with the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than
+for cholesterics in the permeation mode, which is in agreement with experiments
 \cite{Zapotocky:1999, Ramos:2002}.
@@
 This paper is organised as follows.
-In Sec.~\ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics of liquid crystals, which is a generalisation of the Eriksen and Leslie's theory of nematodynamics~\cite{deGennes}.
+In Sec.~\ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics
+of liquid crystals, which is a generalisation of the Ericksen and Leslie's
+theory of nematodynamics~\cite{deGennes}.
 In Sec.~\ref{sec_results} we describe the simulation methodology and the
@@
 cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999}. The model
-describes their state in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
-In the uniaxial approximation, the order parameter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$ with ${\mathbf n}$ the director field.
-Brackets indicate mesoscopic averaging over a large number of individual molecules.
-The largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$ characterizes the local degree of orientational order.
+describes their state in terms of a traceless, symmetric tensor order
+parameter ${\mathbf Q}({\mathbf r})$. 
+In the uniaxial approximation, the order parameter is given by
+$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta})$
+with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
+order. More generally,
+the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
+characterizes the local degree of orientational order.
 The thermodynamic properties of the liquid crystal are determined by a free energy
@@
 The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
-The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$ --
+The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
 we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
@@
 {\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
-& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})Tr({\mathbf Q W}),
+& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
+\mathrm{Tr}({\mathbf Q W}),
 \label{eqn3}
 \end{eqnarray}
-where $Tr$ denotes the tensorial trace and ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and ${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
+where $\mathbf{Tr}$ denotes the tensorial trace, and
+${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
+${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
 $\xi$ is a constant depending on the molecular details of the liquid crystal.
 Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our simulations.
-${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$.
+${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
 \begin{equation}
-{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
+{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
+\mathbf{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
 \label{eqn4}
 \end{equation}
-The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$.\\
-The time evolution of density and velocity are respectively governed by the continuity and the following 
-Navier-Stokes equation Eq.\ref{eqn6}, 
+The rotational diffusion constant $\Gamma$ in Eq.~\ref{eqn2} is proportional
+to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
+\cite{deGennes}.
+
+The time evolution of the fluid density and velocity are respectively governed
+by the continuity equation $\dot{\rho} = -\nabla_\alpha(\rho u_\alpha)$, and
+the following Navier-Stokes equation, 
 \begin{eqnarray}
@@
 \end{eqnarray}
-
-At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq.\ref{eqn6} can be safely neglected.
+At low flow rates the fluid can be considered as incompressible, so that the
+last term on the right hand side of Eq.~\ref{eqn6} remains small.
 $\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
@@
 \end{eqnarray}
-In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
-The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Marenduzzo:2007} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the continuity equation and Eq.\ref{eqn6} and a finite difference scheme for the equation of motion of the tensor order parameter Eq.\ref{eqn2}.
-More details on the algorithm can be found in \cite{Denniston:2001, Denniston:2004}.
+In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
+scalar pressure which, in a system at rest, is constant to a very good
+approximation.
+
+The system of coupled partial differential equations Eqs.~\ref{eqn2}
+and~\ref{eqn6} is solved by means of a
+hybrid scheme \cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
+with predictor-corrector scheme for the continuity equation and
+Eq.~\ref{eqn6}, and a finite difference scheme for the equation of motion of
+the tensor order parameter Eq.\ref{eqn2}. More details on the algorithm can
+be found in \cite{Denniston:2001, Denniston:2004}.
 
@@
 
-In the following we present results from simulations of confined cubic blue phases in steady shear flow.
-Our system consists of $n_y\times n_z = 4\times4$ unit cells in y- and z-direction (flow and vorticity direction respectively) and a varying number $n_x$ of unit cells along a gap in x-direction (which is the flow gradient direction in our geometry).
+We present results from simulations of confined cubic blue
+phases in steady shear flow.
+Our system consists of $n_y\times n_z = 4\times4$ unit cells in $y$-direction
+and $z$-direction (flow and vorticity direction respectively) and a varying
+number $n_x$ of unit cells in $x$-direction (which is the flow gradient
+direction in our geometry).
 As is common in BP simulation work~\cite{Henrich:2010b,Henrich:2011a}, we initialised our simulations with analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, and equilibrated these (for 3000 LB timesteps) with periodic boundary conditions prior to starting the shear.
@@
 of the disclination networks in each case.
-A timestep and lattice spacing in LBU corresponds to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI-units.
-The LBU of stress equals to about $10^8$~Pa. 
-More details about the conversion from LBU to SI-units can be found in \cite{Henrich:2010b,Henrich:2011a}.
+The timestep and lattice spacing in LBU corresponds roughly to
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+is equal to about $10^8$~Pa. 
+More details about the conversion from LBU to SI units can be found in
+\cite{Henrich:2010b,Henrich:2011a}.
 
@@
 $\tau=-0.5$ and $\kappa=2$, respectively, where BPII is the equlibrium phase.
-The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all cases.}
+The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ LBU in all cases.}
 \label{fig1}
@@
 two parameter sets.
-The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in all three cases.
+The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ LBU in all three cases.
 The pictures show a central cubic section consisting of 8 unit cells cut
@@
 
-The speed of the network in the vorticity direction depends on the
+The speed of the network migration in the vorticity direction depends on the
 thermodynamic state of the blue phase.
@@
 Fig.~\ref{fig2}. Note that in both these states BPII is a metastable phase. 
-The images in Fig.~\ref{fig2} show a sequence of initially overlapping
+Fig.~\ref{fig2} shows a sequence of initially overlapping
 disclination networks which separate during the course of the simulation.
@@
 The total network displacement along the vorticity direction is more than
-a unit cell in $10^5$ timesteps for the ``high-temperature'' state.
+a unit cell in $10^5$ timesteps for the high temperature state.
 Typical average and maximum values of the scalar order parameter are
@@
 $\bar{q}_s=0.304$, $q_{s,max}=0.408$ (reference state, in Fig.~\ref{fig1}).
-The average scalar order parameter of the ``low-chirality'' state is about
-$10\%$ higher compared with the reference state. The  relative differerence
+The average scalar order parameter of the low chirality state is about
+$10\%$ higher compared than reference state. The  relative differerence
 between the mean and the maximum value is about a factor $2.5$ smaller.
-We propose that the speed of the networks is different because permeation, which should be required for the ``stick-slip'' motion, may occur more easily in states with less order (``high-temperature'' state) and gentler spatial variation (``low-chirality'' state).
-Furthermore, the uneven, ``stick-slip'' nature of the network motion is possibly due to the anchoring at the boundaries which prevents network displacements at the wall.
+We propose that the speed of network migration is different because
+permeation, which should be required for the stick-slip motion, may occur
+more easily in states with less order (high temperature state) and gentler
+spatial variation (low chirality state). Furthermore, the uneven stick-slip
+nature of the network motion is possibly due to the anchoring at the
+boundaries which prevents network displacements at the wall.
 
@@
 Every cycle is related to a break-up and reformation of the network.
-The time-average in the steady state is about $\Pi_{xy}\simeq5.62\ex{-5}$ in LBU and yields an apparent viscosity $\eta_{app}=(1 +\Pi_{xy}/\eta\dot{\gamma})\simeq 1.86$. 
-When the gap contains $n_x\ge2$ unit cells, the average stress shows less regular oscillations, and fluctuates around a mean value which is about $10-15\%$ smaller, roughly $\eta_{app}=1.65$ for $n_x=4$.
-The fact that the average stress stabilises for $n_x\ge2$ indicates that bulk flow behaviour should occur for those thicknesses, but not in the single unit cell case.
-Different shear rates lead to similar generic features.
+The time-average in the steady state is about $\Pi_{xy}\simeq5.62\ex{-5}$
+in LBU and yields an apparent viscosity
+$\eta_{app}=(1 +\Pi_{xy}/\eta\dot{\gamma})\simeq 1.86$~LBU. This is
+significantly than our baseline Newtonian viscosity $\eta = 0.83$.
+When the gap contains $n_x\ge2$ unit cells, the average stress shows
+less regular oscillations, and fluctuates around a mean value which is
+about $10-15\%$ smaller, roughly $\eta_{app}=1.65$ for $n_x=4$. The fact
+that the average stress stabilises for $n_x\ge2$ indicates that bulk flow
+behaviour is already reached for those thicknesses, but not in the single
+unit cell case. Different shear rates lead to similar generic features.
 
@@
 
-We now compare and contrast the results just obtained for BPII with those
-pertaining to the rheology of blue phase I. In equilibrium, the topology
+We now compare and contrast the results obtained for BPII with those
+pertaining to the rheology of BPI. In equilibrium, the topology
 of the disclination network in BPI is very different from that of BPII.
-The main difference is presence of junction points where four disclination
-lines of topological charge -1/2 meet in BPII; no such junctions occur in BPI.
-
-In the following, we give representative results for BPI at $\tau=-0.5$
+The main difference is presence in BPII of junction points where four
+disclination lines of topological charge -1/2 meet; no such junctions
+occur in BPI. We give representative results for BPI at $\tau=-0.5$
 and $\kappa=1$, where it is the equilibrium phase.
@@
 This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient direction, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
-For larger gap sizes of two and more unit cells a ``stick-slip'' motion of the network arises, along the vorticity direction.
-The associated flow is not directly linked to mass flow, and is therefore somewhat akin to permeation flow in cholesterics.
-The average speed of this network translation depends on the thermodynamic parameters, and appears to be higher for higher temperature and lower chirality.
+For larger gap sizes of two and more unit cells a ``stick-slip'' motion of
+the network arises, along the vorticity direction. The associated network
+migration is not directly linked to mass flow, and is therefore somewhat
+akin to permeation flow in cholesterics. The average speed of this network
+migration depends on the thermodynamic parameters, and appears to be higher
+for higher temperature and lower chirality.
 
 In contrast to BPII, BPI shows no regular break-up and relinking of the unit cells and its disclination network undergoes chaotic rearrangements under shear.
-This behaviour was found for all gap sizes, and may be linked to the largely different topology in the equilibrium disclination patterns in BPI and BPII.
-We will adress the rheology of BPI more in detail in an upcoming publication \cite{Henrich:2012}.
+This behaviour was found for all gap sizes, and may be linked to the
+strongly different topology in the equilibrium disclination patterns in
+BPI and BPII. We will adress the rheology of BPI more in detail in an
+upcoming publication \cite{Henrich:2012}.
 

### 2011-11-14 papers/lmcproc/lmcproc.tex
Correct minor typos
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1488 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \end{eqnarray}
-where $\mathbf{Tr}$ denotes the tensorial trace, and
+where $\mathrm{Tr}$ denotes the tensorial trace, and
 ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
@@
 {\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
-\mathbf{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
+\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
 \label{eqn4}
@@
 The time evolution of the fluid density and velocity are respectively governed
-by the continuity equation $\dot{\rho} = -\nabla_\alpha(\rho u_\alpha)$, and
-the following Navier-Stokes equation, 
+by the continuity equation
+$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
+the following Navier-Stokes equation:
 \begin{eqnarray}

### 2011-11-15 papers/lmcproc/lmcproc.tex
Removed 2 todos
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1489 40876441-3496-0410-a613-a2fca35b2ae2

@@
 % at point of reference
-% DAVIDE (?): Page 6, sentence ``It is interesting to note that the
-% secondary flow in the BP does not match the direction...'' Mike says
-% ``incomprehensible''
-% ANYONE: Please add reference ''to explain strange p_0/rho bit in
-% equation 5''. 
 % OLIVER (lucky you): Please adjust to use standard rheological x,y,z
@@
 +\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
-+(1+3\frac{P_0}{\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
++(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
 \label{eqn6}
 \end{eqnarray}
+The final term in Eq.~\ref{eqn6} arises from the Chapman-Enskog expansion
+of the LB equations \cite{Denniston:2001}.
 At low flow rates the fluid can be considered as incompressible, so that the
@@
 \begin{eqnarray}
-\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
+\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
+-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
 &+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
@@
 below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
-It is interesting to note that the secondary flow in the BP
-does {\em not} match the direction of the ``stick-slip'' motion of the network, which is then not connected to mass flow
-(in this respect this phenomenon is close to a permeation flow).
 
@@
 This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient direction, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
-For larger gap sizes of two and more unit cells a ``stick-slip'' motion of
-the network arises, along the vorticity direction. The associated network
+For larger gap sizes (two and more unit cells) a ``stick-slip'' motion of
+the network arises along the vorticity direction. The associated network
 migration is not directly linked to mass flow, and is therefore somewhat

### 2011-11-15 trunk/papers/lmcproc/lmcproc.tex
Removed all todos - ready for submission
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1490 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-\documentclass[12pt]{iopart}
+\documentclass[12pt,twoside]{iopart}
 \newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
@@
 \usepackage{graphicx, subfigure}
+\usepackage{balance}
+\usepackage{bm}
+\usepackage{float}
+\newcommand{\ex}[1]{\times10^{#1}}
 
@@
 
-\title[]{Shear Flow of Cubic Blue Phases in Confined Geometries}
+\title[]{Confined Cubic Blue Phases under Shear}
+\author{O. Henrich$^{1}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}
 
-\author{O. Henrich$^{1,3}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}
-
-\address{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
+\address{$^1$ Centre for Computational Science, Department of Chemistry,
+University College London, WC1H 0AJ, United Kingdom\\
+$^2$ Edinburgh Parallel Computing Centre and
+$^3$ School of Physics and Astronomy, The King's Buildings,
+The University of Edinburgh, EH9 3JZ, United Kingdom}
 \ead{o.henrich@ucl.ac.uk}
@@
 \begin{abstract}
+We study the behaviour of confined cubic blue phases under shear flow
+via lattice Boltzmann simulations. We focus on the two experimentally observed
+phases, blue phase I and blue phase II. The disclination network of blue phase II continuously breaks and reforms under shear, 
+leading to an oscillatory stress response in time. The oscillations are only regular for very thin samples. For thicker samples, the shear leads to a ``stick-slip'' motion
+of part of the network along the vorticity direction.
+Blue phase I responds very differently: its defect network undergoes seemingly
+chaotic rearrangements under shear, irrespective of system size. 
 \end{abstract}
@@
 % Comment out if separate title page not required
-\maketitle
+%\maketitle
 
@@
 %\section{\label{}}
-\section{Model and Methods}
+Cholesterics are liquid crystals in which the local nematic director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
+The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
+As it is topologically impossible to cover continuously 3D space with double-twist cylinders, defects arise.
+These organise into a variety of regular periodic lattices, giving rise to the so-called cubic blue phases (BPs) \cite{Grebel:1984,Wright:1989}.
+There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, is thought to be amorphous~\cite{Henrich:2011a}).
 
-%Our approach is based on a well-established model for hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes the order structure of the liquid crystal by a traceless, symmetric tensor order parameter ${\bf Q}({\bf r})$.
-%In uniaxial approximation the tensor order paramter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$, whereas ${\bf n}$ represents the local orientation of the liquid crystal molecules and brackets indicate mesoscopic averaging over a large number individual molecules.
-%Its largest eigenvalue $\frac{2}{3}q$ with $0<q<1$ characterizes the degree of order around the corresponding axis. 
-%The equilibrium properties of the system are determined by a Landau-de Gennes free energy density $f$ and the functional ${\cal F}[Q]=\int d^3{\bf r} f({\bf Q})$.
-%The free energy density consists of two different parts, namely a bulk contribution $f_b$ and a gradient term $f_g$.
+BPs have been long considered as purely of academic interest due to their very narrow range of stability. This view has changed since the creation of
+polymer-stabilised and temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the possibility of novel applications.
+During the last few years considerable progress has been achieved regarding the behaviour of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under external fields \cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011}, and in the presence of colloidal particles \cite{Ravnik:2011a}.
+The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. However, our understanding of their dynamical behaviour under flow remains
+very limited. The aim of this work is to address this issue by studying,
+for the first time, the response of confined BP samples to a shear flow.
 
-%\begin{eqnarray}
-%f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
-%f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{eqn1}
-%\end{eqnarray}
+Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
+For example, if a cholesteric helix is subjected to a Poiseuille flow along
+the helical axis, small pressure differences drive flow mainly through
+``permeation'', first investigated by Helfrich \cite{Helfrich:1969}.
+In the permeation mode the liquid crystal flows while leaving the director
+field virtually unchanged, which leads to high dissipation and large
+viscosities.
+Marenduzzo et al.
+\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and showed the importance of the boundary conditions in determining the apparent viscosity of the fluid.
+They also found that a strong secondary flow appears.
+Rey \cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which lead to the rotation of the cholesteric helix.
+At higher forcing, the helix uncoils and leaves a flow-induced nematic phase.
+Rey also studied cholesterics subjected to both steady flow and low frequency
+small amplitude oscillatory shear for different helix orientations
+\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
+excited when the helix was aligned along the flow direction; splay/bend
+deformation occured when the helix was aligned along the velocity gradient;
+but only twist deformations
+appeared when the helix was aligned along the vorticity axis.
 
-%The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ and features a the first-order transition from the isotropic to the ordered phase at $\gamma>2.7$.
-%The second part comprises the elastic contributions to the free energy due to splay-, bend- and twist-deformations and assume for simplicity that they have all the same elastic constant $K$.
-%The pitch length $p$ of the cholesteric liquid crystal is related to the wavervector $q_0=2\pi/p$.
-%In order to specify a thermodynamic state it is useful to introduce two dimensionless quantities, the effective temperature $\tau$ and chirality $\kappa$.
-%They are given by
+Dupuis et al.
+\cite{Dupuis:2005} performed the first numerical investigation of BP rheology in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic array of doubly twisted cylinders.
+Under small forcing, the network opposed the flow giving rise to a significant increase in apparent viscosity.
+Upon increasing the forcing they found clear evidence of shear thinning.
+In the crossover region they predicted a novel oscillatory regime where the network continuously breaks and reforms as portions of the disclinations in the center of the channel move to neighboring cells and relink with the parts of the network left behind by the flow.
+Compared with the cholesteric case, the viscosity still decreases with forcing (the system shear thins) but much less than
+for cholesterics in the permeation mode, which is in agreement with experiments
+\cite{Zapotocky:1999, Ramos:2002}.
 
-%\begin{eqnarray}
-%\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
-%\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber,
-%\end{eqnarray}
+This paper is organised as follows.
+In Sec.~\ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics
+of liquid crystals, which is a generalisation of the Ericksen and Leslie's
+theory of nematodynamics~\cite{deGennes}.
+In Sec.~\ref{sec_results} we describe the simulation methodology and the
+boundary conditions we applied.
+Results for a confined BPII in rectilinear shear flow are presented in
+Sec.~\ref{sec_bpii} with particular focus on the dependence of the flow
+on the thickness of the sample and thermodynamic state.
+In Sec.~\ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.
 
-%whereas the chirality qqnatifies the ratio between gradient and bulk free energy.\\
-%The equation of motion of the tensor order parameter reads
+\section{Model}\label{sec_model}
 
-%\begin{equation}\label{}
-%\left(\partial_t+ u_\alpha \partial_\alpha \right){\bf Q} - {\bf S}({\bf W},{\bf Q}) = \Gamma {\bf H}.
-%\label{eqn2}
-%\end{equation}
+Our approach is based on a well-established model for the hydrodynamics of
+cholesteric liquid crystals \cite{Beris:1994}. The model
+describes their state in terms of a traceless, symmetric tensor order
+parameter ${\mathbf Q}({\mathbf r})$. 
+In the uniaxial approximation, the order parameter is given by
+$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta})$
+with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
+order. More generally,
+the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
+characterizes the local degree of orientational order.
+The thermodynamic properties of the liquid crystal are determined by a free energy
+${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
+\begin{eqnarray}
+f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
+f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
+\end{eqnarray}
+The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
+The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
+we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
+The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
+The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to \cite{Alexander:2006}.
 
-%The first term on the lefthand side of Eq.(\ref{eqn2}) is a material derivative, which describes the rate of change of a quantity moving along with the flow.
-%The second term accounts for the rate of change due to velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$:
+A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
+which are given by
+\begin{eqnarray}
+\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
+\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
+\end{eqnarray}
+The dynamical evolution of the order parameter is given by the equation 
+\begin{equation}
+\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
+\label{eqn2}
+\end{equation}
+The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
+The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
+\begin{eqnarray}
+{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
+& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
+\mathrm{Tr}({\mathbf Q W}),
+\label{eqn3}
+\end{eqnarray}
+where $\mathrm{Tr}$ denotes the tensorial trace, and
+${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
+${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
+$\xi$ is a constant depending on the molecular details of the liquid crystal.
+Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our simulations.
+${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
+\begin{equation}
+{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
+\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
+\label{eqn4}
+\end{equation}
+The rotational diffusion constant $\Gamma$ in Eq.~\ref{eqn2} is proportional
+to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
+\cite{deGennes}.
 
-%\begin{eqnarray}
-%{\bf S}({\bf W}, {\bf Q}) &=& (\xi {\bf A} + {\boldsymbol \Omega})({\bf Q}+\frac{\bf I}{3})\nonumber\\
-%& &\hspace*{-1.5cm}+ ({\bf Q}+\frac{\bf I}{3})(\xi {\bf A}  - {\boldsymbol \Omega})-2 \xi ({\bf Q}+\frac{\bf I}{3})Tr({\bf Q W}),
-%\label{eqn3}
-%\end{eqnarray}
+The time evolution of the fluid density and velocity are respectively governed
+by the continuity equation
+$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
+the following Navier-Stokes equation:
+\begin{eqnarray}
+\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
+= \partial_\beta \Pi_{\alpha \beta}
++\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
++(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
+\label{eqn6}
+\end{eqnarray}
+The final term in Eq.~\ref{eqn6} arises from the Chapman-Enskog expansion
+of the LB equations \cite{Denniston:2001}.
+At low flow rates the fluid can be considered as incompressible, so that the
+last term on the right hand side of Eq.~\ref{eqn6} remains small.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
+The pressure tensor reads explicitly
+\begin{eqnarray}
+\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
+-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
+&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
+&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
+\label{eqn7}
+\end{eqnarray}
+In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
+scalar pressure which, in a system at rest, is constant to a very good
+approximation.
 
-%where $Tr$ denotes the trace operation and ${\bf A}=({\bf W}+{\bf W}^T)/2$ and ${\boldsymbol \Omega}=({\bf W}-{\bf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
-%The molecular details of the liquid crystal such as the aspect ratio of its mesogens are embodied in the constant $\xi$, also know as the tumbling parameter.
-%The righthand side relates the rate of change to the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\bf Q$.
+The system of coupled partial differential equations Eqs.~\ref{eqn2}
+and~\ref{eqn6} is solved by means of a
+hybrid scheme \cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
+with predictor-corrector scheme for the continuity equation and
+Eq.~\ref{eqn6}, and a finite difference scheme for the equation of motion of
+the tensor order parameter Eq.\ref{eqn2}. More details on the algorithm can
+be found in \cite{Denniston:2001, Denniston:2004}.
 
-%\begin{equation}
-%{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
-%\label{eqn4}
-%\end{equation}
+\section{Methodology and Results}\label{sec_results}
 
-%The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q^2/\Gamma$.\\
-%The time evolution of density and velocity are goverend by the continuity equation Eq. (\ref{eqn5})
+We present results from simulations of confined cubic blue
+phases in steady shear flow.
+Our system consists of $n_x\times n_z = 4\times4$ unit cells in $x$-direction
+and $z$-direction (flow and vorticity direction respectively) and a varying
+number $n_y$ of unit cells in $y$-direction (which is the flow gradient
+direction in our geometry).
+As is common in BP simulation work~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, and equilibrated these (for 3000 LB timesteps) with periodic boundary conditions prior to starting the shear.
+During the equilibration run the optimal redshift $r$ was recalculated
+at every timestep.
 
-%\begin{eqnarray}
-%\partial_t \rho + \partial_\alpha (\rho u_\alpha)=0
-%\label{eqn5}
-%\end{eqnarray}
+Shear flow was then started by applying shear boundary conditions to the LB distribution functions as in~\cite{Denniston:2004}.
+For ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities.
+Periodic boundary conditions were retained along the flow direction and the direction of vorticity.
+This set-up is similar to the one Dupuis et al.
+used to simulate Poiseuille flow \cite{Dupuis:2005}.
+The pinning of the defect network at the boundaries may be realised in practice by impurities or surface defects.
+During shear flow the unit cell size was kept fixed and a constant redshift equal to the value at the end of the equilibration protocol was assumed.
 
-%and the Navier-Stokes equation Eq. (\ref{eqn6})
+The total number of unit cells in our simulations of both BPI and BPII
+was $n_x\times n_y \times n_z= 4 \times \{1,2,3,4\} \times 4$.
+A resolution of 32 lattice Boltzmann units (LBU) per pitch length
+(16 per unit cell) was used for BPII, and 64 LBU per pitch length
+(32 per unit cell) used for BPI.  
+This provided enough spatial resolution to track even complex rearrangements
+of the disclination networks in each case.
+The timestep and lattice spacing in LBU corresponds roughly to
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+is equal to about $10^8$~Pa. 
+More details about the conversion from LBU to SI units can be found in
+\cite{Henrich:2011a,Henrich:2010b}.
 
-%\begin{eqnarray}
-%\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}\nonumber\\
-%&+&\hspace*{-2cm}\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}. 
-%\label{eqn6}
-%\end{eqnarray}
+\subsection{Blue Phase II}\label{sec_bpii}
 
-%At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq. (\ref{eqn6}) can be safely neglected.
-%The pressure tensor reads explicitely
-
-%\begin{eqnarray}
-%\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
-%&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
-%&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
-%\label{eqn7}
-%\end{eqnarray}
-
-%It comprises the terms that give rise to an additional stress in the ordered state.
-%In the isotropic state when ${\bf Q}\equiv 0$ Eq. (\ref{eqn7}) is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
-%The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Denniston:2004, Marenduzzo:2008} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the hydrodynamic equations (\ref{eqn5}) and (\ref{eqn6}) and a finite difference scheme for the tensor order parameter equation (\ref{eqn2}).
-
-\section{Results and Discussion}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_100k_x.png}
-\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_100k_y.png}\\
-\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_300k_x.png}
-\includegraphics[width=0.45\textwidth]{disc_bp2_t-0o5_k2o0_300k_y.png}
-\caption{}
+\begin{figure}[t]
+\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_25k_y.png}
+\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_50k_y.png}
+\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_75k_y.png}
+\includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_100k_y.png}
+\caption{Shear flow of BPII at different gap sizes. The pictures show
+snapshots of a subset of the entire disclination network in a system
+with walls at top and bottom, and the flow direction into/out of the
+page. The different number of unit cells in the velocity gradient
+direction have been superposed and are shown in different colours:
+$n_y=1 \mbox{ (yellow)}, n_y=2 \mbox{ (green)}, n_y=3 \mbox{ (red) and }
+n_y=4 \mbox{ (blue)}$. All cases show $n_x \times n_z =2 \times 2$ unit
+cells in flow and vorticity directions. The temperature and chirality are
+$\tau=-0.5$ and $\kappa=2$, respectively, where BPII is the equlibrium phase.
+The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ LBU in all cases.}
 \label{fig1}
@@
 
-\begin{figure*}[h]
+We first report the results for BPII.
+Fig.~\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
+(We prefer the perspective along the flow direction because the network topology is clearer.)
+Temperature and chirality are $\tau=-0.5$ and $\kappa=2$ respectively, and for these parameters BPII is the equilibrium phase.
+The velocity of the walls has been adapted so that the same shear rate $\dot{\gamma}=7.8125\ex{-5}$ is achieved for all gap sizes.
+Shear rates four times larger and smaller were also applied and produced very similar results, apart from the flow velocities which scale up accordingly.
+For all gap sizes we observe an affine transformation in gradient-flow plane that leads to a break-up of the network at strains of about half a unit cell size.
+Shorlty after the break-up the network reforms further downstream, resulting in a periodic, regular structure that is very close to the equilibrium
+BPII disclination network. Similar oscillations have been previously reported for
+BPs in Poiseuille flow, with a gap of one unit cell size \cite{Dupuis:2005}.
+
+It is interesting to compare the results for different gap sizes.
+For $n_y=1$ we observe that the BPII-network is slightly deformed with
+respect to the quiescent state, but always reforms at the same position.
+For $n_y > 1$, the BPII network distorts more under the same shear rate, 
+suggesting that the structure within the sample is strongly affected by
+the anchoring at the boundaries.
+Interestingly, when the sample is thicker than one unit cell, a movement
+of the network in the positive vorticity direction arises
+which is not observed for $n_y=1$.
+This network motion does not proceed at a constant speed, but rather
+proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up,
+and are then quickly released as the network ``slips'' along the vorticity
+direction. Changing from a left-handed to a right-handed helix inverts the
+direction of motion of the network.
+Hence the ``stick-slip'' motion is related to the chiral nature of the helix.
+This phenomenon may bear some similarity to the travelling wave motion in
+cholesterics sheared along the vorticity direction 
+below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
+
+\begin{figure*}[t]
 \centering
@@
 \includegraphics[width=0.3\textwidth]{disc_bp2_tk_scan_125k.png}
-\caption{}
+\caption{BPII in shear flow for different thermodynamic parameters. The
+snapshots show the evolution of the network with the flow direction into/out of
+the page.  The reference state (blue, $\tau=-0.5, \kappa=2$), a high
+temperature state (red, $\tau=0.5, \kappa=2$) and a low chirality state
+(green, $\tau=-0.5, \kappa=1$) are shown; BPII is metastable in the later
+two parameter sets.
+The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ LBU in all three cases.
+The pictures show a central cubic section consisting of 8 unit cells cut
+from a larger simulation with $n_y=4$.}
 \label{fig2}
@@
 
-\begin{figure*}[h]
+The speed of the network migration in the vorticity direction depends on the
+thermodynamic state of the blue phase.
+A comparison between the state from Fig.~\ref{fig1} with a
+``high temperature'' state at $\tau=0.5, \kappa=2$ and a
+``low chirality state'' at $\tau=-0.5, \kappa=1$ is shown in
+Fig.~\ref{fig2}. Note that in both these states BPII is a metastable phase. 
+Fig.~\ref{fig2} shows a sequence of initially overlapping
+disclination networks which separate during the course of the simulation.
+The network moves fastest in the high temperature state, is slower in the
+reference state and slower still in the low chirality state.
+
+\begin{figure*}[t]
 \centering
-\includegraphics[width=\textwidth]{stress_bp2_fbc.pdf}
-\caption{}
+\includegraphics[width=0.725\textwidth]{stress_bp2_fbc.pdf}
+\caption{BPII: ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$
+versus time in the reference state ($\tau=-0.5, \kappa=2$) for gaps ranging
+from one to four unit cells.}
 \label{fig3}
 \end{figure*}
+
+The total network displacement along the vorticity direction is more than
+a unit cell in $10^5$ timesteps for the high temperature state.
+Typical average and maximum values of the scalar order parameter are
+$\bar{q}_s=0.338$, $q_{s,max}=0.383$ (low chirality state), 
+$\bar{q}_s=0.206$, $q_{s,max}=0.320$ (high temperature state) and
+$\bar{q}_s=0.304$, $q_{s,max}=0.408$ (reference state, in Fig.~\ref{fig1}).
+The average scalar order parameter of the low chirality state is about
+$10\%$ higher compared than reference state. The  relative differerence
+between the mean and the maximum value is about a factor $2.5$ smaller.
+We propose that the speed of network migration is different because
+permeation, which should be required for the stick-slip motion, may occur
+more easily in states with less order (high temperature state) and gentler
+spatial variation (low chirality state). Furthermore, the uneven stick-slip
+nature of the network motion is possibly due to the anchoring at the
+boundaries which prevents network displacements at the wall.
+
+Fig. \ref{fig3} shows the spatially averaged shear stress $\langle \Pi_{xy}\rangle$ versus time, in the reference state and for different gap sizes.
+For gaps accommodating only one unit cell a regular periodic pattern emerges shortly after a transient start-up phase.
+Every cycle is related to a break-up and reformation of the network.
+The time-average in the steady state is about $\Pi_{xy}\simeq5.62\ex{-5}$
+in LBU and yields an apparent viscosity
+$\eta_{app}=(1 +\Pi_{xy}/\eta\dot{\gamma})\simeq 1.86$~LBU. This is
+significantly than our baseline Newtonian viscosity $\eta = 0.83$.
+When the gap contains $n_y\ge2$ unit cells, the average stress shows
+less regular oscillations, and fluctuates around a mean value which is
+about $10-15\%$ smaller, roughly $\eta_{app}=1.65$ for $n_y=4$. The fact
+that the average stress stabilises for $n_y\ge2$ indicates that bulk flow
+behaviour is already reached for those thicknesses, but not in the single
+unit cell case. Different shear rates lead to similar generic features.
+
+\subsection{Blue Phase I}\label{sec_bpi}
+
+We now compare and contrast the results obtained for BPII with those
+pertaining to the rheology of BPI. In equilibrium, the topology
+of the disclination network in BPI is very different from that of BPII.
+The main difference is presence in BPII of junction points where four
+disclination lines of topological charge -1/2 meet; no such junctions
+occur in BPI. We give representative results for BPI at $\tau=-0.5$
+and $\kappa=1$, where it is the equilibrium phase.
+Fig.~\ref{fig4} shows the disclination network at timestep $t=4\ex{5}$
+for different gap sizes, with the viewing direction again chosen along
+the flow direction.
+For clarity we only show $n_x \times n_y \times n_z= 1 \times \{1,2,3,4\} \times 2$
+unit cells.
 
@@
 \subfigure[]{\includegraphics[width=0.3\textwidth]{disc_bp1_t-0o5_k1o0_400k_4uc.png}}
-\caption{}
+\caption{Confined BPI in shear flow at $\tau=-0.5, \kappa=1$ and and gap sizes of one (a, yellow), two (b, green), three (c, red) and four (d, blue) unit cells.
+The pictures show the defect network along the flow direction at timestep $t=4\ex{5}$ at the same flow rate as above.}
 \label{fig4}
@@
 \centering
-\includegraphics[width=\textwidth]{stress_bp1_fbc.pdf}
-\caption{}
+\includegraphics[width=0.725\textwidth]{stress_bp1_fbc.pdf}
+\caption{BPI: ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$
+versus time in the reference state ($\tau=-0.5, \kappa=1$) for gaps
+ranging from one to four unit cells.}
 \label{fig5}
@@
 
-\section{Conclusions}
+In contrast to BPII, the disclination network of BPI changes significantly
+under a shear flow, and typical snapshots in Fig.~\ref{fig4} bear no
+relation to the equilibrium BPI configuration.
+Furthermore, there is no sign of the ``stick-slip'' motion of the defects observed in BPII.
+While no regular cubic structure emerges under flow, it is interesting to note that the network develops a periodic modulation along the vorticity direction (horizontal in Fig.~\ref{fig4}).
+This periodicity arises for all gap sizes.
+Whether this feature is a consequence of frustration caused by the anchoring at the boundaries, or is a prelude to a slow ordering of the network under flow, we cannot yet tell.
+
+The ensemble averages of the shear stress $\Pi_{xy}$ for different gap sizes
+is shown in Fig.~\ref{fig5} and gives more evidence of the chaotic nature of
+the flow of BPI. The magnitude of the average shear stress is comparable to
+that seen in BPII; the fall in amplitude of the oscillations with increasing
+gap size is also similar.
+
+\section{Discussions and Conclusions}
+
+In conclusion, we have presented the first supra-unit cell simulations of
+confined cubic blue phases in simple shear flow.
+The rheological behaviour of BPI and BPII was found to be strikingly different.
+
+Our simulations show that the disclination network of BPII continuously breaks and reforms while the network undergoes an affine deformation under shear.
+This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient direction, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
+For larger gap sizes (two and more unit cells) a ``stick-slip'' motion of
+the network arises along the vorticity direction. The associated network
+migration is not directly linked to mass flow, and is therefore somewhat
+akin to permeation flow in cholesterics. The average speed of this network
+migration depends on the thermodynamic parameters, and appears to be higher
+for higher temperature and lower chirality.
+
+In contrast to BPII, BPI shows no regular break-up and relinking of the unit cells and its disclination network undergoes chaotic rearrangements under shear.
+This behaviour was found for all gap sizes, and may be linked to the
+strongly different topology in the equilibrium disclination patterns in
+BPI and BPII. We will adress the rheology of BPI more in detail in an
+upcoming publication \cite{Henrich:2012}.
 
@@
 \ack
-Thanks folks!
+We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time on HECToR.
+M.E.C. holds a Royal Society Research Professorship.
 
@@
 \section*{References}
+
 \bibliography{lmcproc}
-
+\bibliographystyle{unsrt}
 \end{document}

### 2011-11-16 papers/lmcproc/lmcproc.tex
Peter's corrections added - 1st submission
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1491 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-% TODO
-% OLIVER: Please check Olmsted reference: Mike says ``*not* in this paper''
-% at point of reference
-% OLIVER (lucky you): Please adjust to use standard rheological x,y,z
-% Cartesian coordinates. (y,z,x) at moment?
-
 \begin{document}
@@
 %Uncomment for PACS numbers title message
-%\pacs{00.00, 20.00, 42.10}
+\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
 % Keywords required only for MST, PB, PMB, PM, JOA, JOB? 
@@
 %\section{\label{}}
-Cholesterics are liquid crystals in which the local nematic director field spontaneously twist in thermodynamic equilibrium~\cite{deGennes}.
+Cholesterics are liquid crystals in which the local nematic director field spontaneously twists in thermodynamic equilibrium~\cite{deGennes}.
 The preferred configuration close to the isotropic boundary features twist around two perpendicular axes, as opposed to just one axis in the regular cholesteric state, and the corresponding deformation is denoted a ``double-twist cylinder''.
@@
 excited when the helix was aligned along the flow direction; splay/bend
-deformation occured when the helix was aligned along the velocity gradient;
+deformation occurred when the helix was aligned along the velocity gradient;
 but only twist deformations
-appeared when the helix wass aligned along the vorticity axis.
+appeared when the helix was aligned along the vorticity axis.
 
@@
 
-This paper is organised as follows.
+The purpose of the present paper is to contribute to the understanding of the 
+flow behaviour of cholesteric blue phases by reporting results of the first
+supra-unit cell simulations. This paper is organised as follows.
 In Sec.~\ref{sec_model} we introduce the Beris-Edwards model for hydrodynamics
-of liquid crystals, which is a generalisation of the Ericksen and Leslie's
+of liquid crystals, which is a generalisation of Ericksen's and Leslie's
 theory of nematodynamics~\cite{deGennes}.
@@
 on the thickness of the sample and thermodynamic state.
-In Sec.~\ref{sec_bpi} we show corresponding results for BPI, which indicates fundamental differences in the flow behaviour of BPI and BPII.
+In Sec.~\ref{sec_bpi} we show corresponding results for BPI, which indicate 
+fundamental differences in the flow behaviour of BPI and BPII.
 
@@
 
-Our approach is based on a well-established model for the hydrodynamics of
-cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999}. The model
-describes their state in terms of a traceless, symmetric tensor order
-parameter ${\mathbf Q}({\mathbf r})$. 
+Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
+cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
+in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
 In the uniaxial approximation, the order parameter is given by
-$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta})$
+$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
 with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
@@
 \end{eqnarray}
-where $\mathrm{Tr}$ denotes the tensorial trace, and
+where $\mathrm{Tr}$ denotes the tensorial trace, while 
 ${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
-${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
+${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively.
 $\xi$ is a constant depending on the molecular details of the liquid crystal.
-Flow alignment occurs if $\xi \cos{2\theta}=\frac{3q_s}{2+q_s}$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our simulations.
+Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by setting $\xi=0.7$ in our simulations.
 ${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
@@
 last term on the right hand side of Eq.~\ref{eqn6} remains small.
-$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
 The pressure tensor reads explicitly
@@
 phases in steady shear flow.
-Our system consists of $n_y\times n_z = 4\times4$ unit cells in $y$-direction
+Our system consists of $n_x\times n_z = 4\times4$ unit cells in $x$-
 and $z$-direction (flow and vorticity direction respectively) and a varying
-number $n_x$ of unit cells in $x$-direction (which is the flow gradient
+number $n_y$ of unit cells in $y$-direction (which is the flow gradient
 direction in our geometry).
-As is common in BP simulation work~\cite{Henrich:2010b,Henrich:2011a}, we initialised our simulations with analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, and equilibrated these (for 3000 LB timesteps) with periodic boundary conditions prior to starting the shear.
-During the equilibration run the optimal redshift $r$ was recalculated
-at every timestep.
+As is common in BP simulation studies ~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, and equilibrated these (for 3000 LB timesteps) with periodic boundary conditions prior to starting the shear.
+During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
 
@@
 For ${\mathbf Q}$, the equilibrated order parameter was advected on the top and bottom boundary, according to the imposed wall velocities.
-Periodic boundary conditions were retained along the flow direction and the direction of vorticity.
+Periodic boundary conditions were retained along the flow and vorticity direction.
 This set-up is similar to the one Dupuis et al.
@@
 The total number of unit cells in our simulations of both BPI and BPII
-was $n_x\times n_y \times n_z= \{1,2,3,4\} \times 4\times 4$.
+was $n_x\times n_y \times n_z= 4 \times \{1,2,3,4\} \times 4$.
 A resolution of 32 lattice Boltzmann units (LBU) per pitch length
@@
 More details about the conversion from LBU to SI units can be found in
-\cite{Henrich:2010b,Henrich:2011a}.
+\cite{Henrich:2011a,Henrich:2010b}.
 
@@
 \includegraphics[width=0.24\textwidth]{bp2_t-0o5_k2_100k_y.png}
-\caption{Shear flow of BPII at different gap sizes. The pictures show
+\caption{Shear flow of BPII at different gap sizes. The images show
 snapshots of a subset of the entire disclination network in a system
@@
 direction have been superposed and are shown in different colours:
-$n_x=1 \mbox{ (yellow)}, n_x=2 \mbox{ (green)}, n_x=3 \mbox{ (red) and }
-n_x=4 \mbox{ (blue)}$. All cases show $n_y \times n_z =2 \times 2$ unit
+$n_y=1 \mbox{ (yellow)}, n_y=2 \mbox{ (green)}, n_y=3 \mbox{ (red) and }
+n_y=4 \mbox{ (blue)}$. All cases show $n_x \times n_z =2 \times 2$ unit
 cells in flow and vorticity directions. The temperature and chirality are
 $\tau=-0.5$ and $\kappa=2$, respectively, where BPII is the equlibrium phase.
-The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ LBU in all cases.}
+The strain rate is $\dot{\gamma}=7.8125\ex{-5}$ in reciprocal LB time units in all cases.}
 \label{fig1}
@@
 We first report the results for BPII.
-Fig.~\ref{fig1} shows the disclination network in steady shear flow for different number of unit cells between the walls.
+Fig.~\ref{fig1} shows the disclination network in steady shear flow for different numbers of unit cells between the walls.
 (We prefer the perspective along the flow direction because the network topology is clearer.)
@@
 Shear rates four times larger and smaller were also applied and produced very similar results, apart from the flow velocities which scale up accordingly.
-For all gap sizes we observe an affine transformation in gradient-flow plane that leads to a break-up of the network at strains of about half a unit cell size.
-Shorlty after the break-up the network reforms further downstream, resulting in a periodic, regular structure that is very close to the equilibrium
+For all gap sizes we observe an affine transformation in the gradient-flow plane that leads to a break-up of the network at strains of about half a unit cell size.
+Shortly after the break-up the network reforms further downstream, resulting in a periodic, regular structure that is very close to the equilibrium
 BPII disclination network. Similar oscillations have been previously reported for
@@
 It is interesting to compare the results for different gap sizes.
-For $n_x=1$ we observe that the BPII-network is slightly deformed with
+For $n_y=1$ we observe that the BPII-network is slightly deformed with
 respect to the quiescent state, but always reforms at the same position.
-For $n_x > 1$, the BPII network distorts more under the same shear rate, 
+For $n_y > 1$, the BPII network distorts more under the same shear rate, 
 suggesting that the structure within the sample is strongly affected by
@@
 Interestingly, when the sample is thicker than one unit cell, a movement
-of the network in the negative vorticity direction arises
-which is not observed for $n_x=1$.
+of the network in the positive vorticity direction arises
+which is not observed for $n_y=1$.
 This network motion does not proceed at a constant speed, but rather
@@
 The pictures show a central cubic section consisting of 8 unit cells cut
-from a larger simulation with $n_x=4$.}
+from a larger simulation with $n_y=4$.}
 \label{fig2}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{stress_bp2_fbc.pdf}
+\includegraphics[width=0.725\textwidth]{stress_bp2_fbc.pdf}
 \caption{BPII: ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$
@@
 The average scalar order parameter of the low chirality state is about
-$10\%$ higher compared than reference state. The  relative differerence
+$10\%$ higher than that of the reference state. The  relative differerence
 between the mean and the maximum value is about a factor $2.5$ smaller.
@@
 $\eta_{app}=(1 +\Pi_{xy}/\eta\dot{\gamma})\simeq 1.86$~LBU. This is
-significantly than our baseline Newtonian viscosity $\eta = 0.83$.
-When the gap contains $n_x\ge2$ unit cells, the average stress shows
+significantly higher than our baseline Newtonian viscosity $\eta = 0.8333$.
+When the gap contains $n_y\ge2$ unit cells, the average stress shows
 less regular oscillations, and fluctuates around a mean value which is
-about $10-15\%$ smaller, roughly $\eta_{app}=1.65$ for $n_x=4$. The fact
-that the average stress stabilises for $n_x\ge2$ indicates that bulk flow
+about $10-15\%$ smaller, roughly $\eta_{app}=1.65$ for $n_y=4$. The fact
+that the average stress stabilises for $n_y\ge2$ indicates that bulk flow
 behaviour is already reached for those thicknesses, but not in the single
@@
 of the disclination network in BPI is very different from that of BPII.
-The main difference is presence in BPII of junction points where four
+The main difference is the presence in BPII of junction points where four
 disclination lines of topological charge -1/2 meet; no such junctions
@@
 the flow direction.
-For clarity we only show $n_x \times n_y \times n_z=\{1,2,3,4\}\times1\times2$
+For clarity we only show $n_x \times n_y \times n_z= 1 \times \{1,2,3,4\} \times 2$
 unit cells.
@@
 \subfigure[]{\includegraphics[width=0.3\textwidth]{disc_bp1_t-0o5_k1o0_400k_4uc.png}}
-\caption{Confined BPI in shear flow at $\tau=-0.5, \kappa=1$ and and gap sizes of one (a, yellow), two (b, green), three (c, red) and four (d, blue) unit cells.
+\caption{Confined BPI in shear flow at $\tau=-0.5, \kappa=1$ and gap sizes of one (a, yellow), two (b, green), three (c, red) and four (d, blue) unit cells.
 The pictures show the defect network along the flow direction at timestep $t=4\ex{5}$ at the same flow rate as above.}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{stress_bp1_fbc.pdf}
+\includegraphics[width=0.725\textwidth]{stress_bp1_fbc.pdf}
 \caption{BPI: ensemble-averaged shear stress $\langle \Pi_{xy}\rangle$
@@
 This behaviour was found for all gap sizes, and may be linked to the
-strongly different topology in the equilibrium disclination patterns in
-BPI and BPII. We will adress the rheology of BPI more in detail in an
+strongly different topology in the equilibrium disclination patterns of
+BPI and BPII. We will address the rheology of BPI in more detail in an
 upcoming publication \cite{Henrich:2012}.
@@
 \ack
-We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time on HECToR.
+We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173,
+the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
 M.E.C. holds a Royal Society Research Professorship.

### 2011-12-01 papers/smectics/smectics.tex
Update OH
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1492 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{float}
-%\usepackage{rsc}
 
@@
 
-\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, INSERT ADDRESS HERE, UK}}
+\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}}
 \footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
@@
 Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}
-
-SADLY, I THINK IT WOULD BE BEST TO REWRITE THIS PAPER WITH THE FLOW DIRECTION ALONG X, THE GRADIENT ALONG Y AND THE VORTICITY ALONG Z AS IS COMMONLY USED IN THE RHEOLOGICAL LITERATURE!
-I HAVE NOT STARTED TO MAKE THIS CHANGE AS IT WOULD CONFUSE THINGS TOO MUCH AND THE JOB REQUIRES CARE.
-ALSO THE CHEMICAL PRESSURE AND THE STRESS HAVE OPPOSITE SIGNS YET THE FIGURES SHOW THEM HAVING THE SAME SIGN.
-WHAT HAS HAPPENED?
 
@@
 Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
-Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) [P D Olmsted, Rheologica Acta 47, 283-300 (2008)].
+Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.
 
@@
 A closely related example is the so-called `region I flow' in polymeric nematic liquid crystals, where it is thought that a defect network arises, with enhanced viscosity, but with
-marked shear thinning as this structure is sheared out at high flow rates [Larson book -- or Davide to supply references].
+marked shear thinning as this structure is sheared out at high flow rates \cite{Larson}.
 
@@
 layerwise order intact. In steady state, permeation leads to increased dissipation
-and hence to a large viscosity in any geometry where flow necessitates such motion [deGennes book]. (A well-known example of permeation flow
+and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
 is found in cholesterics which are sheared or pushed along the direction
 of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction.
-Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07}[ADD; M Gradzielski, JPCM 15, R655-R697 (2003)].
+Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}.
 (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear.
-Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation [A G Zilman and R Granek, EPJB 11, 593-608 (1999), L Courbin, J P Delville, J Rouch and P Panizza, PRL 89, 148305 (2002)]\cite{Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion [J Zipfel, J Berghause, P Lindner, W Richtering, J Phys Chem B 103, 2841-2849 (1999), A Leon, D Bonn, J Meunier, A Al-Kahlwaji, O Greffer, H Kellay, PRL 84, 1355-1338 (2000)].
+Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}.
 Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work.
@@
 Guided by those results, we additionally report some very large three dimensional simulations at selected points in the parameter space.
-These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals~\cite{Saksena09}.
-[ADD: R S Saksena and P V Coveney, Proc.
-Roy.
-Soc.
-A.
-465, 1977-2002 (2009)).]
-
-We describe smectics below via a widely-used free energy functional [G Gompper and M Schick, Phase Transitions and Critical Phenomena vol 16, C Domb and J L Lebowitz, Eds., Academic Press NY 1994; then cite all the lamellar simulation papers using this or similar functionals].
+These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals \cite{Saksena09a, Saksena09b}.
+
+We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}.
 This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations).
@@
 difference scheme for the convection-diffusion equation, and the
-lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation
-[CITE OUR PREVIOUS PAPERS USING HYBRID LB, ALSO THAT OF GONELLA ET AL WHO FIRST USED IT FOR SMECTICS].
+lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation \cite{Marenduzzo07, Henrich10a,  Henrich10b, Henrich11a,Gonnella97}.
 We use sliding periodic (Lees-Edwards) boundary conditions for
@@
 
-% KS: this following paragraph doesn't quite make sense to me yet.
+KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.
 
@@
 The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
-Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE?], where the numerical value of $m$ was however different.
+Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE], where the numerical value of $m$ was however different.
 In this strongly non-Newtonian regime we also find that the rheology is history dependent.
@@
 
-The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75} [ADD: Gommper and Schick].
+The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75, Gompper}.
 %
@@
 %
-In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature [Kendon et al JFM, relevant Gonnella papers].
+In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature \cite{Kendon01,Gonnella97, Gonnella98, Xu03, Xu05, Xu06a, Xu06b}.
 This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform.
@@
 The amplitude $\phi_q$ is then unbounded without the (positive) quartic term, since $\bar f$ can in that manner become infinitely negative.
-Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{(a_c-a)/b}$, creating also nonlinear harmonics in the density profile.
+Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{4(a_c-a)/3b}$, creating also nonlinear harmonics in the density profile.
 The final free energy density obeys $\bar f \simeq (a_c-a)^2/b$.
@@
 Here $M$ is a mobility coefficient.
-A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ [CONFIRM] (this is negative, signifying the instability).
+A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ (this is negative, signifying the instability).
 In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$.
@@
 \begin{eqnarray}
-\Delta p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
+p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
 & &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
@@
 and finally another term consisting of the order-parameter contribution to the deviatoric stress
-\cite{Evans79} [REFERENCE]
+\cite{Yang76,Evans79} [REFERENCE OK?]
 %
@@
 \end{equation}
-CONFIRM SIGN.
 
@@
 To calibrate the LB length unit we equate $L$ the lamellar wavelength with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm.
-Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{10}$LBU.
+Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{9}$LBU.
 Free energy density parameters such as $a_c \simeq 10^{-4}$LBU should then be of order $\alpha^2k_BT/L^3$ where $\alpha =10$ is a molecular aspect ratio.
@@
 As detailed previously, an estimate of the molecular diffusivity is $\tilde D = 2Mb$; we set $M = 0.25$LBU throughout this study, so that $\tilde D$ lies in the range $2.5\times 10^{-5} - 2.5\times 10^{-4}$LBU.
-Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{11}$LBU) gives $D\simeq 10^{-11}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.
+Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{12}$LBU) gives $D\simeq 10^{-10}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.
 
@@
 The reason for this is that the LB algorithm uses fluid inertia to update the dynamics and the discretization should therefore involve replacing the real fluid with the densest one possible, while ensuring that the Reynolds number $Re \sim \rho\dot\gamma L^2/\eta$ (which for real smectics is exceedingly small) remains small enough for the simulation to faithfully describe the low $Re$ limit.
-As discussed elsewhere [CODEF,SOFT MATTER] $Re$ values up to about 0.1 are then admissable; our simulations comply with this requirement.
+As discussed elsewhere \cite{Cates04, Cates09} $Re$ values up to about 0.1 are then admissable; our simulations comply with this requirement.
 
 Finally, our shear rates lie in the range 
-$10^{-5}-10^{-4}$LBU, so that the (dimensionless) Peclet number, $Pe =\dot\gamma L^2/\tilde D$, which expresses the relative importance of advective to diffusive transport, lies in the range $20-1600$ CONFIRM: WRONG VALUES GIVEN PREVIOUSLY?.
+$10^{-5}-10^{-4}$LBU, so that the (dimensionless) Peclet number, $Pe =\dot\gamma L^2/\tilde D$, which expresses the relative importance of advective to diffusive transport, lies in the range $5-800$.
 These are moderate to high values where significant structural rearrangement can be expected.
 However, for thermotropic smectics, given our time unit of 1LBU$=1.6\times 10^{-12}$s, they correspond to strain rates of order $10^6$s$^{-1}$.
-This lies well above the range typically addressable with rheological experiments, but far below that addressed recently in molecular dynamics simulations [Guo,Sodderman,Stewart].
+This lies well above the range typically addressable with rheological experiments, but below that recently given in a theoretical study \cite{Stewart2009}.
 
@@
 For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm.
-The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$) [Lyotropic Review].
+The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$) [REFERENCE LYOTROPIC REVIEW ?].
 The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e.
@@
 Both the viscosity $\eta$ and the molecular diffusivity $\tilde D$ remain comparable (within one order of magnitude) to the thermotropic values; from the viscosity we then determine that 1s$=6\times 10^6$LBU or 1LB timestep $=1.6\times 10^{-7}$s.
-Accordingly the strain rates relevant to the $Pe$ range addressed here are of order 100 s$^{-1}$ and well within experimental reach.
+Accordingly the strain rates relevant to the $Pe$ range addressed here are of order $10^2$ s$^{-1}$ and well within experimental reach, but below that addressed in molecular dynamics simulations \cite{Guo2002,Soddemann2004}.
 Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still irrelevantly small.
@@
 To further simplify our parameter exploration, throughout what follows we restrict attention to  states with $b = -a$.
-This convention has been widely used to study bulk spinodal decomposition in binary fluids where it can be absorbed in a rescaling of the order parameter [KENDON JFM]: put differently, $b=-a$ ensures that coexisting bulk ordered phases have $\phi = \pm\phi_0$ with $\phi_0=1$.
+This convention has been widely used to study bulk spinodal decomposition in binary fluids where it can be absorbed in a rescaling of the order parameter \cite{Kendon01}: put differently, $b=-a$ ensures that coexisting bulk ordered phases have $\phi = \pm\phi_0$ with $\phi_0=1$.
 In lamellar states of $b= -a$ the amplitude of the lamellar ordering need not be unity but varies like $\phi_q \sim \sqrt{1+a_c/b}$ as stated previously.
 Thus the chosen value of $b$ can be thought of as controlling whether the peak order parameter variation is close to that arising for bulk coexistence (when $b\gg a_c$), or whether the amplitude $\phi_q$ is instead much larger ($b\ll a_c$) and determined by a competition between the gradient terms and a relatively weak quartic nonlinearity.
-In this study we choose three values, $b = 5\times 10^{-4}, 1\times 10^{-4}$ and $5\times 10^{-5}$LBU so that the dimensionless quantity $\beta \equiv b/a_c$ is respectively $2.1,0.42,0.21$, lying near the crossover between these regimes.
+In this study we choose three values, $b = 5\times 10^{-4}, 1\times 10^{-4}$ and $5\times 10^{-5}$LBU so that the dimensionless quantity $\beta \equiv b/a_c$ is respectively $4.2,0.83,0.42$, lying near the crossover between these regimes.
 The observed $\phi_q$ range is about $1.2-1.5$, somewhat narrower than the range $1.1-1.8$ predicted from $\phi_q \sim \sqrt{1+a_c/b}$.
@@
 Alongside the Reynolds number (which we deem irrelevant), $\beta$, and the Peclet number, we can identify two further dimensionless numbers of interest.
-One is the Schmidt number $Sc = \eta/\rho\tilde D$, which is large when momentum can diffuse much more rapidly than molecules -- as applies in almost all dense fluids.
-In our simulations $Sc$ is of order $100--1000$ which is safely large.
-(The work of [KumaranRamanPRE2011] instead addresses Schmidt numbers in the range $0.2-5$ where different physics may apply.) The second parameter controls the strength of feedback between distortions of the $\phi$ field and the fluid motion.
+One is the Schmidt number $Sc = \eta/\rho\tilde D$, which is large when momentum can diffuse much more rapidly than molecules - as applies in almost all dense fluids.
+In our simulations $Sc$ is of order $100-1000$ which is safely large.
+(The work of Kumaran \cite{Kumaran2011} instead addresses Schmidt numbers in the range $0.2-5$ where different physics may apply.) The second parameter controls the strength of feedback between distortions of the $\phi$ field and the fluid motion.
 This
@@
 \begin{equation}
-Pe^* =\frac{(a_c+b)a_c}{4b^2}\,\frac{L^2}{\eta M}\label{pestar}
+Pe^* =\frac{(a_c+b)a_c}{8b^2}\,\frac{L^2}{\eta M}\label{pestar}
 \end{equation}
-With our chosen parameters (in LBU, $a_c=1.2\times 10^{-4}, L = 10, \eta = 8.3\times 10^{-2}, M = 0.25$) we find $Pe^* \sim 1400$ at $\beta = 2.1$ and $Pe^* \sim 40000$ at $\beta = 0.21$.
-The maximum values of $Pe/Pe^*$ that we address below are of order 0.1 and 0.04 respectively in the two cases.
+With our chosen parameters (in LBU, $a_c=1.2\times 10^{-4}, L = 10, \eta = 8.3\times 10^{-2}, M = 0.25$) we find $Pe^* \sim 180$ at $\beta = 4.2$ and $Pe^* \sim 5000$ at $\beta = 0.42$.
+The maximum values of $Pe/Pe^*$ that we address below are of order 0.2 and 0.08 respectively in the two cases.
 
@@
 
-Below we first present detailed results for two parameter sets ($\beta = 2.1, 0.21$) in two dimensions which we call systems A-2D and B-2D.
+Below we first present detailed results for two parameter sets ($\beta = 4.2, 0.42$) in two dimensions which we call systems A-2D and B-2D.
 System lattice sizes and shear rates are reported in Table \ref{tab1}; all remaining parameters are the same in both systems as detailed above.
 %As expected from the preceding discussion of $Pe^*(\beta)$, the stress response at a given flow rate is softer in the second system. 
-A third parameter set ($\beta = 0.42$, system C-2D) gave results broadly similar to B-2D.
+A third parameter set ($\beta = 0.83$, system C-2D) gave results broadly similar to B-2D.
 Note that we will later report simulations in three dimensions for systems A-3D and C-3D which share parameters with A-2D and C-2D (see Table \ref{tab1}).
@@
 \small
-\caption{Simulation parameters: After initialization from a noisy uniform state, the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\times10^{-5}$ to $2\times10^{-4}$LBU.
-In some cases we performed a final switch off run (SO) at zero shear rate.
-The parameters $\kappa=-6\times10^{-4}, c=7.6\times10^{-4}, M=0.25$ and $\eta=8.33\times10^{-2}$ were the same throughout all simulations.
-ADD PE and CA COLUMNS!}
+\begin{tabular*}{0.52\textwidth}{l|l|l|l|l|l|l}
+\hline
+run \# & $N_x\times N_y\times N_z$ & time $/10^5$ & $10^5b$ & Ca & Pe & $10^5\dot\gamma$\\
+\hline
+A & & & & & &\\
+\hline 
+703 & $1024\times512\times1$ & 0-3.2 & $50$ &0 &0 & 0 \\
+704 & $1024\times512\times1$ & 3.2-6.4& $50$ & 0.224 & 40 &  $10$\\
+705 & $1024\times512\times1$ & 6.4-9.6& $50$ &0 & 0 &  0 \\
+\hline
+767 & $512\times256\times1$ & 0-3.2 & $50$ &0 & 0 &  0\\
+775 & $512\times256\times1$ & 3.2-6.4& $50$ &0.448 & 80 &  $20$ \\
+769 & $512\times256\times1$ & 3.2-25.6& $50$ &0.224 & 40 &  $10$ \\
+771 & $512\times256\times1$ & 12.8-38.4& $50$ &0.112 & 20 &  $ 5$ \\
+772 & $512\times256\times1$ & 51.2-76.8& $50$ &0.056 & 10 &  $ 2.5$ \\
+774 & $512\times256\times1$ & 51.2-102.4& $50$ &0.028 & 5 &  $1.25$ \\
+\hline
+792 & $256\times128\times1$ & 3.2-16.0& $50$ &0.22 & 40 &  $10$ \\
+\hline
+786 & $256\times128\times128$ & 0-6.4 & $50$ &0 & 0 &  0\\
+787 & $256\times128\times128$ & 3.2-6.4 & $50$ &0.448 & 80 &  $20$\\
+788 & $256\times128\times128$ & 3.2-9.6 & $50$ &0.224 & 40 &  $10$\\
+789 & $256\times128\times128$ & 6.4-12.8 & $50$ &0.112 & 20 &  $5$\\
+\hline
+B & & & & & &\\
+\hline
+707 & $1024\times512\times1$ & 0-3.2 & $5$ &0 & 0 &  0 \\
+710 & $1024\times512\times1$ & 3.2-9.6& $5$ &0.082 & 400 &  $10$ \\
+765 & $1024\times512\times1$ & 9.6-12.8& $5$ &0 & 0 &  0 \\
+\hline
+776 & $512\times256\times1$ & 0-3.2 & $5$ & & 0 &  0 \\
+777 & $512\times256\times1$ & 3.2-9.6& $5$ &0.164 & 800 &  $20$ \\
+778 & $512\times256\times1$ & 3.2-25.6& $5$ &0.082 & 400 &  $10$ \\
+780 & $512\times256\times1$ & 12.8-38.4& $5$ &0.041 & 200 &  $5$ \\
+781 & $512\times256\times1$ & 25.6-64.0& $5$ &0.020 & 100 &  $2.5$ \\
+782 & $512\times256\times1$ & 51.2-76.8& $5$ &0.010 & 50 &  $1.25$ \\
+\hline
+C & & & & & &\\
+\hline
+998 & $256\times128\times1$ & 0-3.2& $10$ &0 & 0 & 0 \\
+1001 & $256\times128\times1$ & 3.2-9.6& $10$ &0.252 & 400 &  $20$ \\
+1000 & $256\times128\times1$ & 3.2-12.8& $10$ &0.126 & 200 &  $10$ \\
+\hline
+999 & $256\times128\times128$ & 0-3.2 & $10$ &0 & 0 &  0 \\
+1003 & $256\times128\times128$ & 3.2-9.6 & $10$ &0.252 & 400 &  $20$\\
+1002 & $256\times128\times128$ & 3.2-12.8 & $10$ &0.126 & 200 &  $10$\\
+\hline
+\end{tabular*}
+
+\caption{Simulation parameters: After initialization from a noisy uniform state, the systems were first equilibrated at zero shear rate, followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\times10^{-5}$ to $2\times10^{-4}$LBU.
+In some cases we performed a final switch off run at zero shear rate.
+The other model parameters $\kappa=-6\times10^{-4}, c=7.6\times10^{-4}, M=0.25$ and $\eta=8.33\times10^{-2}$ were the same throughout all simulations.}
 \label{tab1}
-\begin{tabular*}{0.495\textwidth}{lllll}
-\hline
-run \# & $N_x\times N_y\times N_z$ & time $/10^3$ & $10^5b$ &  $10^5\dot\gamma$\\
-\hline
-A & & & &\\
-\hline 
-703 & $512\times1024\times1$ & 0-320 & $50$ & 0 - EQ  \\
-704 & $512\times1024\times1$ & 320-640& $50$ & $10$\\
-705 & $512\times1024\times1$ & 640-960& $50$ & 0 - SO \\
-\hline
-767 & $256\times512\times1$ & 0-320 & $50$  & 0 - EQ\\
-775 & $256\times512\times1$ & 320-640& $50$ & $20$ \\
-769 & $256\times512\times1$ & 320-2560& $50$ & $10$ \\
-771 & $256\times512\times1$ & 1280-3840& $50$ & $ 5$ \\
-772 & $256\times512\times1$ & 5120-7680& $50$ & $ 2.5$ \\
-774 & $256\times512\times1$ & 5120-10240& $50$ & $1.25$ \\
-\hline
-792 & $128\times256\times1$ & 320-1600& $50$ & $10$ \\
-\hline
-786 & $128\times256\times128$ & 0-640 & $50$ & 0 - EQ\\
-787 & $128\times256\times128$ & 320-640 & $50$ & $20$\\
-788 & $128\times256\times128$ & 320-960 & $50$ & $10$\\
-789 & $128\times256\times128$ & 640-1280 & $50$ & $5$\\
-\hline
-B & & & &\\
-\hline
-707 & $512\times1024\times1$ & 0-320 & $5$ & 0 - EQ \\
-710 & $512\times1024\times1$ & 320-960& $5$ & $10$ \\
-765 & $512\times1024\times1$ & 960-1280& $5$ & 0 - SO \\
-\hline
-776 & $256\times512\times1$ & 0-320 & $5$ & 0 - EQ\\
-777 & $256\times512\times1$ & 320-960& $5$ & $20$ \\
-778 & $256\times512\times1$ & 320-2560& $5$ & $10$ \\
-780 & $256\times512\times1$ & 1280-3840& $5$ & $5$ \\
-781 & $256\times512\times1$ & 2560-6400& $5$ & $2.5$ \\
-782 & $256\times512\times1$ & 5120-7680& $5$ & $1.25$ \\
-\hline
-C & & & &\\
-\hline
-998 & $128\times256\times1$ & 0-320& $10$ & 0 - EQ \\
-1001 & $128\times256\times1$ & 320-960& $10$ & $20$ \\
-1000 & $128\times256\times1$ & 320-1280& $10$ & $10$ \\
-999 & $128\times256\times128$ & 0-320 & $10$ & 0 - EQ\\
-1003 & $128\times256\times128$ & 320-960 & $10$ & $20$\\
-1002 & $128\times256\times128$ & 320-1280 & $10$ & $10$\\
-\hline
-\end{tabular*}
 \end{table}
@@
 
+We turn now to system A-2D ($\beta = 4.2$) The top panel in Fig. \ref{fig1} shows the order parameter pattern $\phi$ at the end of the equilibration phase prior to shearing.
+Although the true equilibrium structure would show homogeneous alignment, this cannot be reached in a reasonable time scale unless shear is applied.
+The resulting metastable structure is amorphous and isotropic, but locally exhibits the same lamellar spacing $L = 10$ LBU as the fully ordered equilibrium state.
+If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
+A typical snapshot in the resulting steady state at $Pe = 40$ $(Ca = 0.22)$ is shown in the middle panel of Fig.~\ref{fig1}.
+In this and all subsequent images, the applied flow velocity is horizontal and the velocity gradient vertical, with the flow directed to the right in the upper half and to the left in the lower half of the picture.
+After a short transient during which the order parameter remains relatively homogenous (not shown) there emerge extended regions in which the smectic layers are either expanded or compressed, so that $L$ locally deviates from the equilibrium value.
+These mesoscopic inhomogeneities are inclined at an angle to the flow and remain continuously present in the steady state, but are not stationary: their local features are advected by the flow and evolve continuously by deformation.
 \begin{figure}[htp]
@@
 
-We turn now to system A-2D ($\beta = 2.1$) The top panel in Fig. \ref{fig1} shows the order parameter pattern $\phi$ at the end of the equilibration phase prior to shearing.
-Although the true equilibrium structure would show homogeneous alignment, this cannot be reached in a reasonable time scale unless shear is applied.
-The resulting metastable structure is amorphous and isotropic, but locally exhibits the same lamellar spacing $L = 10$ LBU as the fully ordered equilibrium state.
-If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
-A typical snapshot in the resulting steady state at $Pe = 80$ $(Ca = 0.06)$ CONFIRM is shown in the middle panel of Fig.~\ref{fig1}.
-In this and all subsequent images, the applied flow velocity is horizontal and the velocity gradient vertical, with the flow directed to the right in the upper half and to the left in the lower half of the picture.
-After a short transient during which the order parameter remains relatively homogenous (not shown) there emerge extended regions in which the smectic layers are either expanded and compressed, so that $L$ locally deviates from the equilibrium value.
-These mesoscopic inhomogeneities are inclined at an angle to the flow and remain continuously present in the steady state, but are not stationary: their local features are advected by the flow and evolve continuously by deformation.
-
 We found that after the shear flow was switched off the lamellar spacing returned to its equilibrium value everywhere, with expanded regions contracting and vice versa.
@@
 
-Fig. \ref{fig2} shows equivalent states for system B-2D ($\beta = 0.21$).
-The Peclet number during the shearing phase is now $Pe = 800$ $(Ca = 0.02)$ CONFIRM.
+Fig. \ref{fig2} shows equivalent states for system B-2D ($\beta = 0.42$).
+The Peclet number during the shearing phase is now $Pe = 400$ $(Ca = 0.082)$.
 All the generic results of Fig.~\ref{fig1} are the same, but the defect density, both during flow and after switchoff, is significantly reduced.
@@
 
-In principle the dependence of the state of order on flow rate could be smooth, or could entail a nonequilibrium phase transition -- for instance if the perfectly ordered state ceased to be stable above a certain shear rate \cite{Gonnella98,Kumaran2001} [Ramaswamy,Granek].
-In practice, however, starting as we do from an initially amorphous lamellar texture, a perfectly ordered state is generally not reached for large systems CONFIRM even if it might be stable once formed.
+In principle the dependence of the state of order on flow rate could be smooth, or could entail a nonequilibrium phase transition -- for instance if the perfectly ordered state ceased to be stable above a certain shear rate \cite{Gonnella98,Zilman99,Kumaran2001} [REFERENCE RAMASWAMY?].
+In practice, however, starting as we do from an initially amorphous lamellar texture, a perfectly ordered state is generally not reached for large systems even if it might be stable once formed.
 In any case it seems likely that at high $Pe$, any defects introduced into the smectic state by deformation of an initially imperfect structure (whose layers are then dilated and compressed by the flow) cannot anneal away or annihilate each other rapidly.
@@
 The system has been pre-sheared at stepwise decreasing rates (see Table \ref{tab1}).
-Sites where the alignment function $X$ is smaller than $0.8$ are highlighted red.}
+Sites where the alignment function $X$ is smaller than $0.8$ are highlighted red.
+The system size was $N_x\times N_y \times N_z=512\times 256 \times 1$.}
 \label{fig3}
@@
 All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints (edge dislocations) or other local defects.
-The upper panel gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, immediately following a pre-shearing protocol in which the strain rate was ramped down stepwise from higher values (see Table \ref{tab1}) IS THIS CROPPED OR THE WHOLE SYSTEM?
-WHAT SIZE?.
+The upper panel gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, immediately following a pre-shearing protocol in which the strain rate was ramped down stepwise from higher values (see Table \ref{tab1}).
 While large aligned regions are present there are other regions of high defect density.
@@
 This is prevented by the buckling of layers above a certain shear rate, beyond which the system shows a tendency to trip over itself, which results in the creation of new defects.
-The onset of defect proliferation arises at characteristic flow rates $\dot{\gamma}_c=2.5-5\e{-5}$LBU in system A-2D and around $\dot{\gamma}_c=1-2\e{-4}$LBU in B-2D.
+The onset of defect proliferation arises at characteristic flow rates $\dot{\gamma}_c=1.25-2.5\e{-5}$LBU in system A-2D and around $\dot{\gamma}_c=1-2\e{-4}$LBU in B-2D.
 (Here the subscript $c$ could stand for ``critical" or for ``crossover" depending on whether there is a singular transition or not; see above.)
 These values are for
-systems of size $256\times512\times1$; we found evidence of an increasing trend with system size, but the extremely long run-times involved ($\sim 10^7$ timesteps) precluded quantitative exploration.
-These results correspond respectively to values of the Peclet number, $Pe_c$, obeying $20<Pe_c<40$ (system A) and $800<Pe_c<1600$ (system B).
-However, in terms of $Ca = Pe/Pe^*$ one has $0.014<Ca_c<0.028$ (system A) and $0.02<Ca_c<0.04$ (system B).
-Given the 40-fold difference in $Pe_c$, the similarity of $Ca_c$ in the two different systems is remarkable.
-
-Boldly extrapolating across parameter regimes that we have not explored in any detail, our results motivate the hypothesis that the onset of defect proliferation occurs generically in 2D at $Ca_c\simeq 0.03$.
+systems of size $512\times256\times1$; we found evidence of an increasing trend with system size, but the extremely long run-times involved ($\sim 10^7$ LB timesteps) precluded quantitative exploration.
+These results correspond respectively to values of the Peclet number, $Pe_c$, obeying $5<Pe_c<10$ (system A) and $400<Pe_c<800$ (system B).
+However, in terms of $Ca = Pe/Pe^*$ one has $0.056<Ca_c<0.112$ (system A) and $0.082<Ca_c<0.164$ (system B).
+Given the 80-fold difference in $Pe_c$, the similarity of $Ca_c$ in the two different systems is remarkable.
+
+Boldly extrapolating across parameter regimes that we have not explored in any detail, our results motivate the hypothesis that the onset of defect proliferation occurs generically in 2D at $Ca_c\simeq 0.08$.
 Note that $Ca$, which comprises the ratio of a viscous stress to an elastic modulus, can also be viewed as a strain.
@@
 In Fig.~\ref{fig4} we explore the correlation between the inhomogeneous morphology in the flowing state and locally defined rheological quantities.
-The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N_1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$ ($Pe = XX$, $Ca = YY$).
-(Here and below we adopt a COMPLETELY UNCONVENTIONAL coordinate system where $y$ is along the flow direction and $x$ the velocity gradient.)
+The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N_1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$ ($Pe = 40$, $Ca = 0.22$).
+Here and below we adopt a coordinate system where $x$ is along the flow direction and $y$ the velocity gradient.
 In regions where the lamellar width is larger than in the equilibrium state, shear stresses and first normal stress differences show low positive or negative values.
@@
 \caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (top); in steady shear at $\dot{\gamma}=1\times10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (bottom).
-The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU.
-LABEL THE AXES OTHERWISE X AND Y DIRECTIONS ARE NOT SPECIFIED.}
+The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU with $k_x$ and $k_y$ along the horizontal and vertical coodinate direction, respectively.}
 \label{fig5}
@@
 The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
-Interestingly, following each step down in shear rate, both systems A-2D and B-2D continuously anneal to lower defect densities on a timescale $\simeq 10^6$LBU CONFIRM FOR B.
-If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur on a timescale $XXX$LBU.
+Interestingly, following each step down in shear rate, both systems A-2D and B-2D continuously anneal to lower defect densities on a timescale $\simeq 10^6$LBU.
+If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur on a timescale corresponding to a stain of about 100\%.
 This finding offers further support for a dynamical equilibrium between defect creation and annihilation above a critical (or crossover) shear rate, below which defects progressively annihiliate and near-perfect alignment eventually results.
@@
 Horizontal lines indicate time averages after entering the apparent steady state (where present).
-The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right) CONFIRM.}
+The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).}
 \label{fig6}
@@
 
-Fig.~\ref{fig7} shows the spatially [CONFIRM] averaged total shear stress $\langle \sigma_{xy}({\bf r})\rangle \equiv\sigma_{xy} $.
+The top picture in Fig.~\ref{fig7} shows the spatially averaged total shear stress $\langle \sigma_{xy}({\bf r})\rangle \equiv\sigma_{xy} $.
 At low shear rates, stress fluctuations have longer temporal correlations but smaller amplitudes.
@@
 \centering
-\includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
-\caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$, we find $\eta_{app}/\eta\equiv\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
- }
+\includegraphics[angle=0,width=0.45\textwidth]{S_xy_tot_t_5e-4.pdf}
+\includegraphics[angle=0,width=0.45\textwidth]{P_xy_chem_t_5e-4.pdf}
+\includegraphics[angle=0,width=0.465\textwidth]{P_xy_defect_density.pdf}
+\caption{Top: average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$, we find $\eta_{app}/\eta\equiv\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$; centre: average order parameter contribution to the pressure tensor $-P_{xy}$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right); bottom: average chemical contribution to the pressure tensor $-P_{xy}$ vs.
+defect density for system A-2D.}
 \label{fig7}
@@
 
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
-\caption{Average order parameter contribution to the pressure tensor $P_{xy}$ vs time for system A-2D.
-The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).}
-\label{fig8}
-\end{figure}
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
-\caption{Average chemical contribution to the pressure tensor $ P_{xy}$ vs. defect density for system A-2D. REMOVE B-2D. PROBABLY MERGE 7, 8, 9 INTO SINGLE FIGURE.} 
-\label{fig9}
-\end{figure}
-
-System A-2D shows clear shear thinning behaviour; values of its apparent viscosity at various shear rates are given in the caption of Fig.~\ref{fig7}.
-The corresponding range of $Ca$ is XXXX-YYYY.
+System A-2D shows clear shear thinning behaviour; values of its apparent viscosity at various shear rates are given in the caption.
+The corresponding range of $Ca$ is 0.028-0.448.
 Using Eq.\ref{fullstress}, we can identify the stress contribution arising from the deviatoric part of the order parameter pressure tensor
-$\langle P_{xy}({\bf r})\rangle\equiv P_{xy}$; this is shown in Fig.~\ref{fig8} RESOLVE SIGN ISSUE.
-The temporal fluctuations are in phase with those of the total stress in Fig.~\ref{fig7} and decrease with decreasing shear rate.
-Fig. \ref{fig9} plots $P_{xy}$ against the defect density $\rho_D$; this shows a linear relation across almost the whole range of shear rates.
+$\langle - P_{xy}({\bf r})\rangle\equiv - P_{xy}$; this is shown in the centre graph in Fig.~\ref{fig7}.
+The temporal fluctuations are in phase with those of the total stress and decrease with decreasing shear rate.
+The bottom image in Fig. \ref{fig7} plots $- P_{xy}$ against the defect density $\rho_D$; this shows a linear relation across almost the whole range of shear rates.
 
@@
 To within the measured accuracy, system B-2D is rheologically quasi-Newtonian.
-However, the range of $Ca$ in this case is lower, ZZZZ-TTTT, and the order parameter stress creates only a relatively modest uplift to the baseline viscosity.
-More significantly, $Ca\le Ca_c$ throughout this range CONFIRM so that the steady states found in system B-2D are far more ordered than in A-2D.
+However, the range of $Ca$ in this case is lower, 0.01-0.164, and the order parameter stress creates only a relatively modest uplift to the baseline viscosity.
+More significantly, $Ca\le Ca_c$ for all but the highest applied shear rate so that the steady states found in system B-2D are far more ordered than in A-2D.
 This results from a defect annihilation process, occurring in the early stages of ramp-down, which we found to be accompanied by a sharp drop in $P_{xy}$ (not shown).
@@
 \includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
-\caption{Spatially averaged first normal stress differences $N_1= \sigma_{yy}-\sigma_{xx}$ vs time for system A-2D.
+\caption{Spatially averaged first normal stress differences $N_1= \sigma_{xx}-\sigma_{yy}$ vs time for system A-2D.
 The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).
 Horizontal solid lines are time averages.}
-\label{fig10}
-\end{figure}
-
-Fig.~\ref{fig10} shows averaged first normal stress differences $\langle N_1({\bf r})\rangle \equiv N_1$ versus time for system A-2D, where we followed the same stepwise decreasing flow protocol as before.
+\label{fig8}
+\end{figure}
+
+Fig.~\ref{fig8} shows averaged first normal stress differences $\langle N_1({\bf r})\rangle \equiv N_1$ versus time for system A-2D, where we followed the same stepwise decreasing flow protocol as before.
 For the first several steps $N_1$ decreases monotonously with decreasing shear rate.
@@
 We have not explored the dependence on system size of this normal stress anomaly, but it may be partly a finite size effect caused by the relatively modest number of smectic layers within the simulated volume.
-IS THIS WHAT YOU MEANT?
 The normal stress in system B-2D (not shown) behaves differently: in this case $N_1$ falls monotonically during the defect annihilation phase. 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
-\caption{Flow curves: Time-averaged shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D ($\eta=8.33\times10^{-2}, -a=b=5\times 10^{-4}$).
-LABEL THIS PROPERLY}
-\label{fig11}
-\end{figure}
-
-Fig.~\ref{fig11} shows `flow curves' representing the shear rate dependence of time averaged stresses $\sigma_{xy}$ and $N_1$.
-This is calculated from the stepwise ramp data as detailed previously, so the $N_1$ results at low shear rates should not be taken as definitive CONFIRM.
+\caption{Flow curves: Time-averaged shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D and B-2D.
+The dash-dotted lines are power-law fits to the data.}
+\label{fig9}
+\end{figure}
+
+Fig.~\ref{fig9} shows `flow curves' representing the shear rate dependence of time averaged stresses $\sigma_{xy}$ and $N_1$.
+This is calculated from the stepwise ramp data as detailed previously, so the $N_1$ results at low shear rates should not be taken as definitive.
 System A-2D shows shear thinning as already discussed, apparently following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
@@
 One possible explanation for this discrepancy is that screw defects, which do not exist in our 2D simulations, play a key role for this theory.
-On the other hand the fitted power law must be treated with great caution since the fit includes data from both $\dot\gamma$ both above and below $\dot\gamma_c$ CONFIRM.
-It is somewhat surprising that the strong morphological changes around $\dot\gamma_c$ give no visible anomaly or feature in the flow curves.
-CONFIRM AGAIN -- THIS IS VERY ODD.
+On the other hand the fitted power law must be treated with great caution since the fit includes data from both $\dot\gamma$ both above and below $\dot\gamma_c$.
+It is somewhat surprising that the morphological changes around $1.25\e{-5}<\dot\gamma_c <2.5\e{-5}$ in system A give no visible anomaly or feature in the flow curves.
+This may be partly explained with a more crossover-like behaviour of system A at the lowest applied shear rate.
 %
@@
 This appears to be due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$; the order parameter stress $P_{xy}$ is then dominated by the background viscosity term $\eta\dot\gamma$.
-THAT MAKES THE RESULT FOR A-2D STILL MORE SURPRISING, NO?
 
@@
 On the other hand, one would expect $N_1$ also to vanish in that case; if this happens, it is at lower strain rates than encountered here.
+
 WE HAVE TO UNDERSTAND WHY THE NORMAL STRESS IS SO LARGE, ESPECIALLY IN B-2D WHERE YOU ARE CLAIMING THAT THE CHEMICAL STRESS IS NEGLIGIBLE!!
+
+OH: I have performed simulations with a perfectly aligned state with different layer spacing. The viscous stress vanishes completely in all cases.
+However, the first normal stress difference seems to depend sensitively on the layer spacing.
+
+For a layer spacing equal to the Brazovskii equilibrium wavelength  (corresponding to 26 lamellae across the gradient direction at $N_x \times N_y \times N_z = 512 \times 256 \times 1$) $N_1$ is very small and positive of order $10^-6$.
+Interestingly, neither $P^{chem}_{xy}$ nor $N_1$ alter when the flow is switched on, but remain virtually constant.
+
+If the wavelength of the initial state is changed by 5-10\%, the number of layers decreases or increases by 2.
+Again, the viscous stress is zero, but in the compressed / expanded state we get positive /negative $N_1=P^{chem}_{yy}-P^{chem}_{xx}$ (using the conventional definition of flow-minus-gradient and the fact that the chemical pressure has a negative sign).
+The magnitude of $N_1$ is higher for system B than for system A. This can be understood with difference in Brazvskii amplitude (A: $\phi_q\simeq1.28$, B: $\phi_q\simeq 2$).
+For system B $N_1$ is almost of order $10^{-4}$!
+
+Therfore, our findings for $N_1$ correspond to a slightly compressed system with a layer spacing smaller than the equilibrium one.
+The fact that $N_1$ is already very small if equilibrium spacing is used suggests that the plateau is caused by finite size effects.
+
+The reason for the sensitivity of $N_1$ towards compression and expansion is not really clear yet to me. 
+It appears that $P_xx$ (x being the flow direction) is much more sensitive towards changes in the layer spacing than $P_yy$.
+Intuitively I would have expected it to be the other way round as changes in the layer spacing affect the y-gradients more than the x-gradients (after all the structure is translationally invariant in x-direction).
+At the moment I can only think of an explanation that draws on something beyond the single-mode expansion where gradients in flow direction pick up wavelengths in the velocity-gradient-direction.
 
@@
 
-The computational cost of repeating all the above studies in three dimensional systems of equivalent size remains prohibitive, particularly when addressing steady state properties which emerge only after a million or more timesteps.
+The computational cost of repeating all the above studies in three dimensional systems of equivalent size remains prohibitive, particularly when addressing steady state properties which emerge only after a million or more LB timesteps.
 We therefore focus in 3D on smaller systems (Table \ref{tab1}), but still large enough in principle to show qualitatively similar structure under flow.
@@
 System A-3D is identicaly to A-2D except for the introduction of the third dimension.
-System C-3D has parameters intermediate between A-2D and B-2D (Table \ref{tab1}) and was chosen in preference to the latter because
-preliminary studies on B-3D showed very rapid alignment into a fully ordered state.
-CONFIRM THAT, OR GIVE ANOTHER GOOD REASON.
+System C-3D has parameters intermediate between A-2D and B-2D (Table \ref{tab1}) and was chosen in preference to the latter because B-2D ordered completely if the system size was matched to a slice of B-3D.
 (We see below however that in fact this is a rather general outcome in 3D.)
@@
 Results for A-3D are morphologically indistinguishable from those for C-3D.
-In all these images the flow direction is horizontal, to the right in the upper half of the sample and to the left in the lower half.
-CONFIRM}
-\label{fig12}
+In all these images the flow direction is horizontal, to the right in the upper half of the sample and to the left in the lower half.}
+\label{fig10}
 \end{figure*}
@@
 \includegraphics[angle=0,width=0.40\textwidth]{ck_y-slice_run786_320.jpg}
-\caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
-Shown are wave vectors on the interval $k_x, k_z\in[-\pi/2 ,\pi/2 ]$.
-Data for C-3D (not shown) is almost the same.
-CONFIRM}
-\label{fig13}
+\caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_x=0$ after equilibration at $t=3.2\e{5}$.
+Shown are wave vectors on the interval $k_y, k_z\in[-\pi/2 ,\pi/2 ]$ with $k_y$ and $k_z$ being the horizontal and vertical direction, respectively.
+Data for C-3D (not shown) is virtually the same.}
+\label{fig11}
 \end{figure}
@@
 \includegraphics[angle=0,width=0.35\textwidth]{ck_z-slice_run788_960.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\times10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 $, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 ,\pi/2]$.
-Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
-\label{fig14}
-\end{figure}
-
-Fig.~\ref{fig12} shows the post-equilibration amorphous structure and the steady-state structure under flow in system C-3D (A-3D is similar) at system size $N_x\times N_y \times N_z=128\times256\times128$.
-The latter is compared with C-2D and A-2D in a system of the same $x$ and $y$ dimensions ($128\times256\times1$).
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\times10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=3\pi/16$ and $k_z=0$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 ,\pi/2]$.}
+\label{fig12}
+\end{figure}
+
+Fig.~\ref{fig10} shows the post-equilibration amorphous structure and the steady-state structure under flow in system C-3D (A-3D is similar) at system size $N_x\times N_y \times N_z=256\times128\times128$.
+The latter is compared with C-2D and A-2D in a system of the same $x$ and $y$ dimensions ($256\times128\times1$).
 The quiescent metastable state of system C-3D after equilibration is more or less the same as it would be in two dimensions.
-Indeed, a slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig.
+Indeed, a slice through the structure factor along $k_x=0$ (cuts along $k_y,k_z=0$ look identical) is shown in Fig. \ref{fig11} and resembles that in Fig.
 ~\ref{fig5}.
-However, under shear at $\dot\gamma = 2\times 10^{-4}$ ($Pe = XXX$, $Ca = YYY$) this structure anneals readily into a regular lamellar structure after a transient period of about XXX timesteps.
+However, under shear at $\dot\gamma = 2\times 10^{-4}$ ($Pe = 400$, $Ca = 0.252$) this structure anneals readily into a regular lamellar structure after a transient period of about $6\e{5}$ LB timesteps.
 This occurs for even the highest shear rates applied, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
-The same behaviour was seen CONFIRM in system A-3D for which the highest flow rate used corresponds to $Pe = XXX$ and $Ca = YYY$.
-(In contrast, the 2D results at the same flow rates and system sizes feature the typical fluctuating pattern of breaking and merging lamellae that was described previously and is visible in the lower panels of Fig.~\ref{fig12}.) The structure factor seen in 3D is accordingly close to that of a perfectly ordered lamellar state (see Fig.~\ref{fig14}) in contrast to the corresponding results in 2D (Fig.~\ref{fig5}).
+The same behaviour was seen in system A-3D for which the highest flow rate used corresponds to $Pe = 80$ and $Ca = 0.448$.
+(In contrast, the 2D results at the same flow rates and system sizes feature the typical fluctuating pattern of breaking and merging lamellae that was described previously and is visible in the lower panels of Fig.~\ref{fig10}.) The structure factor seen in 3D is accordingly close to that of a perfectly ordered lamellar state (see Fig.~\ref{fig12}) in contrast to the corresponding results in 2D (Fig.~\ref{fig5}).
 These findings suggest that any characteristic strain rate $\dot\gamma_c$ at which defects proliferate must be markedly higher in 3D than in 2D.
-Specifically, if $Ca$ is the relevant control parameter, our results for A-3D imply that $Ca_c > YYY$ in three dimensions.
-
-Note however that the order we observe is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers. WHAT DOES THAT MEAN?
+Specifically, if $Ca$ is the relevant control parameter, our results for A-3D imply that $Ca_c > 0.448 $ in three dimensions.
+
+[DESCRIPTION CLEAR?] Note however that the order we observe is not completely perfect as a mismatch of the layers occurs leading to branching points and edges (e.g.
+at the centre of the ordered 3D-system in Fig.~\ref{fig10}).
 The Burgers vector is oriented along the gradient direction with edges along the flow direction.
@@
 Notably, the corresponding orientation of edge dislocations does not exist in two dimensions.
-A closer look at the kinetic evolution of the order parameter reveals that regions separated by different values of order parameter connect along a third direction by expelling regions with opposite order parameter.
-This is a direct consequence of the additional dimension and seems to play a key role in promoting the formation of the homogeneously layered structure.
-I DO NOT UNDERSTAND ALL THIS: DOES IT CORRESPOND TO MOTION OF THE ORIENTED EDGE DISLOCATIONS? IF NOT, WHAT ROLE DO THOSE PLAY DURING THE ANNEALING?
-EXPLAIN CAREFULLY WHY THIS MECHANISM SPEEDS THINGS UP IN 3D BUT NOT IN 2D.
+[TRYING TO EXPLAIN WHAT'S HAPPENING IN 3D. CLEAR?]
+
+A closer look at the kinetic evolution of the order parameter reveals that the additional dimension seems to play a key role in promoting the formation of the homogeneously layered structure.
+The ordering process takes place in two consecutive steps: 
+Firstly, the equilibrated state undergoes a transition to a semi-ordered state, which consists of small patches of layered regions with the layer normals in gradient direction.
+Initially, the length of the lamellae is quite short but gets longer in flow direction as the simulation proceeds and the structure is advected.
+It remains, however, relatively short in vorticity direction with the individual layers being disconnected.
+In a second step the layers merge by linking up in vorticity direction and expelling regions with opposite order parameter, which is only possible due to the additional third dimension.
 
 We next quantify the time system A-3D takes to anneal into its final ordered structure starting from an isotropic initial configuration.
-(These and subsequent measurements are reported for A-3D only; C-3D is qualitatively similar CONFIRM.)
+(These and subsequent measurements are reported for A-3D only; C-3D is qualitatively similar.)
 The inset of Fig.~\ref{fig6} gives the defect density as defined by Eq.~\ref{defect density} versus time for shear rates $\dot{\gamma}=5\times10^{-5}-2\times10^{-4}$. 
 Within a few hundred thousand LB timesteps the defect density decreases by two orders of magnitude, resembling the behaviour seen in system A-2D at shear rates one decade smaller.
-The total strain where the curves level off, and ordering is achieved, scales with the shear rate and is about $\gamma=1.3$ (at $\dot{\gamma}=2\times10^{-4}$), $\gamma=0.8$ (at $\dot{\gamma}=1\times10^{-4}$) and $\gamma=0.6$ (at $\dot{\gamma}=5\times10^{-5}$).
-CONFIRM THESE STRAIN NUMBERS, THEY SEEM UNBELIEVABLY SMALL!!!
+The total strain where the curves level off, and ordering is achieved, scales with the shear rate and is about $\gamma=160$ (at $\dot{\gamma}=2\times10^{-4}$), $\gamma=120$ (at $\dot{\gamma}=1\times10^{-4}$) and $\gamma=65$ (at $\dot{\gamma}=5\times10^{-5}$).
 
 The reduced number of defects is reflected in the magnitude of the order parameter stress contribution $P_{xy}$; the 3D behaviour is shown in the inset of Fig.~\ref{fig8}.
-The final contribution to the total stress is created by a small number of remaining defects (hence the inconsistent trend of the tiny plateau value with strain rate, which is not a concern here CONFIRM) and is quite negligible compared to the background viscosity contribution $\eta\dot\gamma$.
+The final contribution to the total stress is created by a small number of remaining defects (hence the inconsistent trend of the tiny plateau value with strain rate, which is not a concern here) and is quite negligible compared to the background viscosity contribution $\eta\dot\gamma$.
 The drop in $P_{xy}$ by three orders of magnitude during the defect annihilation process can formally be viewed as shear thinning of a quite extreme degree -- albeit one masked, in any macroscopic measurement of $\sigma_{xy}$, by the Newtonian background.
 
-First normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig10}.
-THIS IS THE WRONG DEFINITION EVEN WITH THE CRAZY COORDINATE SYSTEM ADOPTED IN THIS PAPER.
-N1 IS FLOW MINUS GRADIENT NOT FLOW MINUS VORTICITY.
+First normal stress differences $N_1=\langle\sigma_{xx}-\sigma{yy}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig8}.
 For all shear rates addressed here system A-3D attains similar values of the normal stress; these are approached monotonically and about one decade smaller than those seen in system A-2D.
@@
 Finally we want to glimpse at the structure factor of system A-3D in shear flow.
-Fig. \ref{fig14} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
+Fig. \ref{fig12} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
 The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
@@
 Although strongly peaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
-Fig. \ref{fig15} depicts a close up of the 3D-structure factor, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
+Fig. \ref{fig13} depicts a close up of the 3D-structure factor, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
 It reveals a banana-shaped spatial dependence with some small appendices in flow direction at $k_x=0$.
@@
 \caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\times10^{-4}$.}
-\label{fig15}
+\label{fig13}
 \end{figure}
@@
 
-Many earlier studies have proposed a picture in which a layer undulation instability in three-dimensional smectic liquid crystals emerges above a critical shear rate [A G Zilman and R Granek, EPJB 11, 593-608 (1999), L Courbin, J P Delville, J Rouch and P Panizza, PRL 89, 148305 (2002)]\cite{Gonnella98,Kumaran2001} [Ramaswamy].
-Molecular dynamics simulations on thermotropic smectics place this in a regime of flow rates well above those we have applied in the present study \cite{Guo2002,Soddemann2004, Stewart2009}  CONFIRM THIS WITH NEW PARAMETER MAPPING -- IT COULD BE WRONG.
+Many earlier studies have proposed a picture in which a layer undulation instability in three-dimensional smectic liquid crystals emerges above a critical shear rate \cite{Zilman99,Courbin02,Gonnella98,Kumaran2001} [Ramaswamy].
+Molecular dynamics simulations on lyotropic smectics place this in a regime of flow rates well above those we have applied in the present study \cite{Guo2002,Soddemann2004}.
+The critical shear rate for thermotropic smectics that has been given in a theoretical approach \cite{Stewart2009} is slightly above ours [CHECKED].
 The status of this instability and its effects on rheology in experiments remains somewhat unclear.
@@
 In the present work we have studied the nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model.
-After quenching from a homogeneous phase a metastable amorphous lamellar structure appears, similar to previous simulations in two dimensions using comparable models and parameters [REFS].
-Also in accord with some previous studies [REF], under flow in two dimensions we found the morphology of the order parameter to show two distinct regimes: one with very few defects which arises at low shear rates, and a defect rich phase at higher ones.
+After quenching from a homogeneous phase a metastable amorphous lamellar structure appears, similar to previous simulations in two dimensions using comparable models and parameters [REFERENCES].
+Also in accord with some previous studies [REFERENCES], under flow in two dimensions we found the morphology of the order parameter to show two distinct regimes: one with very few defects which arises at low shear rates, and a defect rich phase at higher ones.
 We found that the critical (or crossover) strain rate $\dot\gamma_c$ separating these phases depends somewhat on system size, but much more strongly on thermodynamic and kinetic parameters.
 We have argued that this dependence might be summarized in terms of a capillary number $Ca$ (Eq.~\ref{capillary}) that is the ratio of viscous stress to the gradient free energy density of the lamellar phase.
-This controls a steady-state elastic strain in the smectic which, when it exceeds some characteristic value $Ca_c$, might cause yielding of layers and defect proliferation; our best estimate is $Ca_c \simeq 0.03$.
+This controls a steady-state elastic strain in the smectic which, when it exceeds some characteristic value $Ca_c$, might cause yielding of layers and defect proliferation; our best estimate is $Ca_c \simeq 0.03$ (system A) and $0.01$ (system B).
 However we emphasize that the evidence for $Ca$ as a control parameter is currently very limited and it would take a painstaking exploration of parameter space to confirm this picture.
@@
 The extended regions with lamellar spacing larger than at equilibrium exhibit either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
-[PLEASE EXPLAIN THE FOLLOWING SENTENCE
-A similar dependence of these charactistic patterns on the system size has been recently reported in \cite{Kumaran2011}.]
+[EXPLANATION CLEAR?]
+A similar dependence of the layer alignment in flow direction depending on the system size has been recently reported in \cite{Kumaran2011}.
+Smaller systems with 16 layers across the gradient direction, which is comparable with our systems consisting of  $N_x\times N_y \times N_z=256\times128\times1$ sites, showed a larger tendency to align in flow direction.
+For larger systems there was a clear trend to align at an angle with the flow direction.
 These distortions of the order parameter contribute a term $P_{xy}$ to the shear stress whose magnitude could be directly related to the density of defects.
-In one system we found shear thinning described by a power law dependence of stress on strain rate, $\sigma_{xy}=\gamma^m$ with $m=0.87$; however the simplicity of this result may be fortuitous since the range of strain rates studied spans $\dot\gamma_c$ CONFIRM.
+In one system we found shear thinning described by a power law dependence of stress on strain rate, $\sigma_{xy}=\gamma^m$ with $m=0.87$; however the simplicity of this result may be fortuitous since the range of strain rates studied spans $\dot\gamma_c$.
 In another system, at strain rates primarily below $\dot\gamma_c$, the shear rheology was quasi-Newtonian ($m=1$) although a significant first normal stress difference $N_1$ was also present.
@@
 (In saying this, we imagine a 3D continuation of the 2D structure which is translationally invariant along that direction.) Our 3D results do not rule out a transition or crossover from ordered to disordered structures, but this is not seen within the range of flow rates addressed here.
-This means that, if such a transition is controlled by capillary number, its value at onset in 3D obeys $Ca_c>YYY$, a value notably higher than that observed in 2D.
+This means that, if such a transition is controlled by capillary number, its value at onset in 3D obeys $Ca_c>0.448$, a value notably higher than that observed in 2D.
 Since identical parameter sets were chosen, the same applies to any other dimensionless number that might control the onset of the defect-rich phase.
 
-Finally, we note that one recent theory of smectic rheology [LU] suggests this is controlled by screw dislocations which have no 2D analogue.
+Finally, we note that one recent theory of smectic rheology \cite{Lu08} suggests this is controlled by screw dislocations which have no 2D analogue.
 Moreover, the residual dislocations seen in the 3D ordered phase have an orientation that also does not exist in 2D.
-It should be remembered that, in 3D, edge dislocations and screw dislocations are interconvertible: they have the same topological status and the nomenclature merely describes whether the defect line is oriented normal or parallel to the smectic layers [CHAIKIN+LUBENSKY].
+It should be remembered that, in 3D, edge dislocations and screw dislocations are interconvertible: they have the same topological status and the nomenclature merely describes whether the defect line is oriented normal or parallel to the smectic layers \cite{ChaikinLubensky}.
 The ability of 3D defect lines to reorient as well as translate underscores their greater dynamical freedom compared to similar defects in a 2D geometry.
@@
 \section{Acknowledgements}
-We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time at the Argonne Leadership Computing Facility, Argonne, U.S.A. as well as on HECToR, U.K.
+We acknowledge support by EPSRC Grants No.
+EP/E045316 and No.
+EP/E030173, the MAPPER EU-FP7 project (grant no.
+RI-261507) and computing time at the Argonne Leadership Computing Facility, Argonne, U.S.A.
+as well as on HECToR, U.K.
 We like to thank A. Morozov for useful discussions. 
@@
 \footnotesize{
-RSC STYLE FILE APPARENTLY STRIPS OUT MIDDLE INITIALS. THIS IS OFFENSIVE, PARTICULARLY TO AUTHORS LIKE AJ WAGNER WITH COMMON NAMES.
 \bibliography{smectics} %your .bib file
@@
 \if{
-In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
+In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig10} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
 Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.

### 2011-12-09 trunk/papers/smectics/smectics.tex
Next iteration
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1494 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-10-14 trunk/papers/smectics/smectics.tex
@@
 
+%\documentclass[8.5pt,twoside,twocolumn]{article}
+%\oddsidemargin -1.2cm
+%\evensidemargin -1.2cm
+%\textwidth 18cm
+%\headheight 1.0in
+%\topmargin -3.5cm
+%\textheight 22cm
+%\usepackage[super,sort&compress,comma]{natbib} 
+%\usepackage{mhchem}
+%\usepackage{times,mathptm}
+%\usepackage{sectsty}
+%\usepackage{balance} 
+%\usepackage{graphicx} %eps figures can be used instead
+%\usepackage{lastpage}
+%\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
+%\usepackage{fancyhdr}
+%\usepackage{bm}
+%\usepackage{float}
+
+%\pagestyle{fancy}
 \documentclass[8.5pt,twoside,twocolumn]{article}
@@
 \usepackage{balance}
+
 \usepackage{graphicx} %eps figures can be used instead
@@
 \usepackage{bm}
-\usepackage{float}
-
 \pagestyle{fancy}
+\usepackage{color}
 
 \newcommand{\update}[1]{{\color{red} #1}}
-\newcommand{\e}[1]{\cdot10^{#1}}
+\newcommand{\e}[1]{\times10^{#1}}
 \newcommand{\gd}{\dot{\gamma}}
@@
 \fancypagestyle{plain}{
-\fancyhead[L]{\includegraphics[height=8pt]{LH.pdf}}
-\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{CH.pdf}}
-\fancyhead[R]{\includegraphics[height=10pt]{RH.pdf}\vspace{-0.2cm}}
+%\fancyhead[L]{\includegraphics[height=8pt]{headers/LH.pdf}}
+%\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{headers/CH.pdf}}
+%\fancyhead[R]{\includegraphics[height=10pt]{headers/RH.pdf}\vspace{-0.2cm}}
+\renewcommand{\headrulewidth}{1pt}}
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\renewcommand\footnoterule{\vspace*{1pt}% 
+\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
+\setcounter{secnumdepth}{5}
+
+\makeatletter 
+\def\subsubsection{\@startsection{subsubsection}{3}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\bf}} 
+\def\paragraph{\@startsection{paragraph}{4}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\textit}} 
+\renewcommand\@biblabel[1]{#1}            
+\renewcommand\@makefntext[1]% 
+{\noindent\makebox[0pt][r]{\@thefnmark\,}#1}
+\makeatother 
+\renewcommand{\figurename}{\small{Fig.}~}
+\sectionfont{\large}
+\subsectionfont{\normalsize}
+
+\fancyfoot{}
+%\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{headers/LF.pdf}}
+%\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{headers/RF.pdf}}
+%\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{headers/RF.pdf}}
+\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
+\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
+\fancyhead{}
+\renewcommand{\headrulewidth}{1pt} 
+\renewcommand{\footrulewidth}{1pt}
+
+\setlength{\arrayrulewidth}{1pt}
+\setlength{\columnsep}{6.5mm}
+\setlength\bibsep{1pt}
+
+\if{\thispagestyle{plain}
+\fancypagestyle{plain}{
+%\fancyhead[L]{\includegraphics[height=8pt]{LH.pdf}}
+%\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{CH.pdf}}
+%\fancyhead[R]{\includegraphics[height=10pt]{RH.pdf}\vspace{-0.2cm}}
 \renewcommand{\headrulewidth}{1pt}}
@@
 \setlength\bibsep{1pt}
+}\fi
 
+%\newcommand{\Pe}{\hbox\rm Pe}
 \twocolumn[
   \begin{@twocolumnfalse}
-\noindent\LARGE{\textbf{Shear Flow of Lamellar Mesophases and Smectic Liquid Crystals}}
+\noindent\LARGE{\textbf{Rheology of Lamellar Liquid Crystals in Two and Three Dimensions: A Simulation Study}}
+%: A Simulation Study}}
 \vspace{0.6cm}
 
-\noindent\large{\textbf{O. Henrich\textit{$^{a}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.V. Coveney{$^{a}$} and M.E. Cates \textit{$^{c}$}}
+\noindent\large{\textbf{O. Henrich\textit{$^{a,c}$}, K. Stratford\textit{$^{b}$}, D. Marenduzzo\textit{$^{c}$}, P.~V. Coveney{$^{a}$} and M.~E. Cates \textit{$^{c}$}}
 }\vspace{0.5cm}
@@
 
-\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, UK}}
-\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK}}
-\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, UK}}
+\footnotetext{\textit{$^{a}$ Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}}
+\footnotetext{\textit{$^{b}$ Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
+\footnotetext{\textit{$^{c}$ SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
 
@@
 
-Instances of soft complex fluids abound. Examples include colloidal
-suspensions, where particles or droplets are suspended in a viscous
-fluid, living tissues, and liquid crystalline fluids. Well known
-examples of colloidal suspensions include paint, cosmetics, and
-food products such as ketchup, mayonnaise, and yogurt.
-The intracellular network of actin fibres constituting the
-cytoskeleton of mammalian cells is a polymeric fluid.
-The same is true of the DNA making up the bacterial nucleoid and
-eukaryotic chromosomes, and the viral genome of bacteriophages,
-where enhanced confinement imparts a strong liquid crystalline
-character to the system.
-Liquid crystalline fluids themselves, with or without colloidal
-inclusions, are of of wide technological importance in devices
+Instances of soft complex fluids abound in our everyday lives: examples include paint (a colloidal suspension), mayonnaise (an emulsion), and engine oil (a polymer solution).
+Many soft materials, including the amphiphiles used in skin creams, fabric softeners, and some pharmaceutical products, form lamellar mesophases (also known as smectic liquid crystals) whose perfectly ordered state consists of a regular stack of flat layers.
+As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent~\cite{deGennes}.
+Such phases are of wide technological importance in devices
 such as electronic displays and sensors.
 
-The behaviour under flow, or rheology, of such materials is important
-to understand, as it affects many industrial and manufacturing processes.
-At the same time, rheological measurements provide a powerful way to
-characterise macroscopic properties of a soft material.
-The response to applied forces or stresses can be used to quantify,
-e.g., the viscosity, or elastic and loss moduli of soft materials.
-On shearing, a simple Newtonian fluid exhibits a stress which is
-linearly proportional to the applied shear rate; the
-constant of proportionality is just the viscosity. In marked
-contrast, complex fluids flow is often non-Newtonian and such a simple
-linear relation no longer holds. On shearing, the effective viscosity
-of a complex viscoelastic fluid may increase or decrease, corresponding
-to shear thinkening or shear thinning behaviour. Complex fluids may even
-spontaneously phase separate when flowing, giving rise to shear bands.
-These are familiar in the rheology of liquid crystals in the vicinity of
-the isotropic-nematic transition, and in wormlike micelles.
+The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
+For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity.
+Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression.
+However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress.
+Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear.
+On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
+to shear thickening or shear thinning behaviour~\cite{Larson}.
+Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
+Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.
 
-The rheological properties of a non-Newtonian fluid are typically linked
-to the microstructure of the material. Particularly relevant to our
-current work is the rheology of layered systems, such as smectics and
-cholesterics. Here, experiments may show different rheological results
-depending on whether the system has been ordered prior to measurement
-(for example, by pre-shearing).
-For instance, in the flow of polydomain liquid crystals, which are not
-uniformly ordered, the existence of regions with conflicting director
-patterns leads to a non-linear relation between shear stress and
-applied shear rate.
-Another example of strongly non-Newtonian rheology is region I flow
-in polymeric nematic liquid crystals. While our understanding of the
-physics of region I flow is far from complete, it is thought that a
-defect network may form and lead to much enhanced viscosity, with
-marked shear thinning. In both these cases, stiffening is usually
-associated to the presence of defects in the sample.
+Lamellar liquid crystals provide an important paradigm for non-Newtonian fluids in which the microstructure of the material is strongly coupled to the flow.
+Crucially, the defect texture may depend on past history -- for instance, the strain rate in any prior flow that was suddenly switched off.
+This rheological hysteresis means that the precise flow protocol, including any pre-shearing applied prior to starting the experiment, needs to be specified.
+Such hysteresis can arise in polydomain liquid crystals of various types (not just smectics) such as nematics in which the existence and evolution of regions with misaligned directors creates rheological nonlinearity.
+A closely related example is the so-called `region I flow' in polymeric nematic liquid crystals, where it is thought that a defect network arises, with enhanced viscosity, but with
+marked shear thinning as this structure is sheared out at high flow rates \cite{Larson}.
 
-Another intriguing type of flow which is possible in layered systems
+An intriguing complication in addressing the dynamics of layered systems
 is permeation,
 where individual molecules in the fluid translate while keeping the
-defect pattern intact in steady state. This leads to increased dissipation
-and hence to a larger viscosity. A well-known example of permeation flow
+layerwise order intact. In steady state, permeation leads to increased dissipation
+and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
 is found in cholesterics which are sheared or pushed along the direction
-of the cholesteric helix. However, one would also expect permeation flows
-to be possible when a non-translationally invariant defect pattern is
-present in the system; this might be the case in quenched smectics, for
-instance. In addition to their non-linear shear-stress curve, lamellar
-systems are also of interest for a number of other phenomena.
-In block-copolymers one may observe, under some conditions, shear-induced
-phase transitions from the microphase separated state to ordered phases
-like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. It would be
-desirable to extend our theoretical understanding of such switches.
-Further, lyotropic layered liquid crystals have sometimes been observed
-to form an onion phase \cite{Panizza96,Iwashita07} which can happen either
-spontaneously \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04} or under
-shear \cite{Diat93}. The mechanism underlying such pattern formation has
-remained elusive to date.
+of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction.
+Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}.
+(The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear.
+Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}.
+Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work.
+Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
+phase transitions from a homogeneous state into ordered phases
+like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}.
+However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}.
+In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
+once defects are present).
 
-Despite a wealth of experimental results, and the importance of rheology
-to many commercial applications, theoretical progress in understanding
-flow behaviour of complex fluids has remained a challenge. This is largely
-because of the non-linear nature of the equations of motion, and the
-coupling of the flow to different microstructures. These either require
-drastic approximations or, for practical purposes, resort to computer
-simulation. Even here, realistic 3-dimensional calculations are
-computationally expensive.
-In this work, our focus is the physics of smectics under shear.
+Despite a wealth of experimental results, and the importance of flow behaviour
+to many commercial applications, mechanistic understanding
+the rheology of lamellar phases remains limited. This is largely
+because of the nonlinear nature of the equations of motion, and the
+coupling of the flow to different microstructures. These complications either require
+drastic approximations or, as pursued here, use of computer
+simulation. However, once defects are present, such simulations are computationally expensive. 
 While sheared smectics have been studied extensively via simulations
@@
 computational constraints have meant almost all these results are limited
-to two dimensions. Further, a number of these studies have used solid walls
-to drive shear in the hydrodymamics and, inconsistently, sliding periodic
-boundaries for the composition variable. Such simulations have a number of
-drawbacks: solid walls to drive shear limits the attanable shear rate in
-larger systems and so exacerbates finite size effects; mixed boundary
-conditions make it difficult to disentangle bulk and boundary effects
-in the rheology. In this work, we try to overcome all of these issues.
-(One notable example which does tackle a large 3-dimensional simulation
-of a complex fluid under shear involved gyroid phases~\cite{Saksena09}.)
+to two dimensions.
+Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology.
+Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs.
+This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems.
+Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.
 
-We describe smectics via a Brazovskii free energy functional
-\cite{Brazovskii75}, which provides a robust framework to study systems
-like micelles, surfactants, smectic liquid crystals and block-copolymers.
-The governing equations in the model are a convection-diffusion equation
-for the scalar order parameter, which represents the density difference
-of two fluid components, and the Navier-Stokes equation describing
-hydrodynamics. A stress tensor that contains extra contributions owing
-to the presence of interfaces between the components is included.
-These dynamical equations are solved using a combination of a finite
+In this work, we try to overcome these technical obstacles, presenting systematic results for large two dimensional simulations with fully periodic boundary conditions.
+Guided by those results, we additionally report some very large three dimensional simulations at selected points in the parameter space.
+These 3D simulations are, we believe, of unprecedented scale for lamellar phases, although even larger ones have recently been used to address the rheology of cubic liquid crystals \cite{Saksena09a, Saksena09b}.
+
+We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}.
+This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations).
+The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$.
+This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid, with interfaces between these layers.
+A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks.
+However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents a thermotropic smectic state.
+
+The governing equations in our model are a convection-diffusion equation
+for the scalar order parameter, and the Navier-Stokes equation describing
+the hydrodynamics of fluid flow.
+The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy.
+Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
+Our dynamical equations are solved using a combination of a finite
 difference scheme for the convection-diffusion equation, and the
-lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation.
-We use consistent sliding periodic (Lees-Edwards) boundary conditions for
-both convection-diffusion and Navier-Stokes equations. This allows us to
-probe the true bulk rheology of the smectic phase for large systems and
-high shear rates, as quantified by an appropriate Peclet number.
+lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation \cite{Marenduzzo07, Henrich10a,  Henrich10b, Henrich11a,Gonnella97}.
+We use sliding periodic (Lees-Edwards) boundary conditions for
+both the convection-diffusion and the Navier-Stokes equations.
+Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses.
+(As it happens, we do not observe shear bands in the range of shear rates we address here.)
+
+%This simulation strategy allows us to probe the bulk rheology of large systems at moderate to high shear rates, as quantified by an appropriate Peclet number, defined below.
 
@@
 
-% KS: this following paragraph doesn't quite make sense to me yet.
+%KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.
 
-Our main finding is that, depending on system parameters, a smectic phase
-shows markedly different rheological responses. At lower Peclet
-numbers, where thermodynamic effects are relatively strong compared with
-flow effects,  shear quickly orders the system into a lamellar phase,
-with only a few defects. The rheology in this regime is Newtonian, and
-the shear stress is linear in the applied shear rate.
-If, on the other hand, thermodynamic forces are larger, we observe that shear actually leads to the nucleation of new defects, and the rheology is highly nonlinear, with the shear rate proportional to $\dot{\gamma}^{\alpha}$, with $\alpha\simeq 0.87$, leading to a shear thinning behaviour.
-Such a behaviour is qualitatively reminiscent of the one found experimentally in XXX, where the numerical value of $\alpha$ was however different.
-In this regime we also find that the rheology is history dependent, as the defect density decreases for instance if the shear rate is ramped down gradually.
-Finally, intriguingly, dimensionality appears to play a key role: within our 3D simulations, we have indeed not been able to enter the nonlinear rheology regime with defect nucleation, and we always observe a shear-induced ordering into a lamellar phase.
-More research and parameter steering will be needed to make this observation more quantitative.
+Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
+of shear, where thermodynamic forces are relatively strong compared with
+flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
+with only a few defects.
+The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate.
+(There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow.
+This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other.
+The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters.
+A comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
+Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning.
+The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
+Such a behaviour is qualitatively reminiscent of the one found experimentally in~\cite{Lu08}, where the numerical value of $m$ was however different.
+In this strongly non-Newtonian regime we also find that the rheology is history dependent.
 
-This paper is organised as follows. In Section II we introduce the model,
+Finally, and intriguingly, we find that dimensionality appears to play a key role.
+Within our 3D simulations, even when performed with parameter values identical to those used in 2D, we have so far not been able to enter the nonlinear flow regime associated with defect proliferation.
+Instead, we always observe shear-induced ordering into a lamellar phase with quasi-Newtonian rheology.
+Further research, requiring careful and systematic parameter steering, will be needed to map out quantitatively the onset of the defect proliferation regime, if one exists, in three dimensions.
+
+This paper is organized as follows. In Section \ref{themodel} we introduce the model,
 including its equilibrium properties, the full stress tensor and the
 dynamical equations. We briefly sketch how we derive structural information
-from the order parameter and describe our defect quantification procedure,
-which measures the local alignment of the lamellar structure.
-We present results of the two-dimensional case in in Section III and
-discuss the morphology, average and local shear stress densities and
-first normal stress differences and their relation to the defect structure.
-We also show structure factors for various flow states and flow curves.
-We also compare our two-dimensional results with results in three dimensions. 
-A summary with conclusions closes the paper in Section V.
+from the order parameter and describe a method for quantitative analysis of the defect density.
+Section \ref{parametermap} describes the mapping between our simulation parameters and those for physical systems of interest.
+We present results for the two-dimensional case in Section \ref{results2D}, discussing
+the morphology as a function of flow rate. We also consider average and local shear stress densities and
+first normal stress differences, and explore their relation to the defect structure. In addition we report Fourier-space information (structure factors) under flow.
+In Section \ref{results3D} we present and compare our results for 3D simulations which, by computational necessity, are less comprehensive than in 2D. 
+A summary with conclusions closes the paper in Section \ref{conclusions}.
 
-\section{Model and Definitions}
+\section{Model and Definitions} \label{themodel}
 
-The equilibrium phase is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75,deGennes}.
+The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75, Gompper}.
+%
+\begin{eqnarray}
+{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
+\end{eqnarray}
+%
+In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature \cite{Kendon01,Gonnella97, Gonnella98, Xu03, Xu05, Xu06a, Xu06b}.
+This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform.
+The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.
 
+The parameter range relevant to lamellar phases has $\kappa < 0$ and $c>0$.
+Then, if we suppress the quartic term, the free energy density of a sinusoidal lamellar state $\phi({\bf r}) = \phi_k\cos({\bf k}.{\bf r})$ obeys $\bar f = (a+\kappa k^2+ c k^4)\phi_k^2/4$.
+This is minimized by choosing $k=q\equiv\sqrt{|\kappa|/2c}$; the free energy density is then $\bar f = (a-\kappa^2/4c)\phi_q^2/4$ so that ordering will occur for $a<a_c=\kappa^2/4c$.
+The amplitude $\phi_q$ is then unbounded without the (positive) quartic term, since $\bar f$ can in that manner become infinitely negative.
+Restoring the quartic term now causes saturation at $\phi_q\sim \sqrt{4(a_c-a)/3b}$, creating also nonlinear harmonics in the density profile.
+The final free energy density obeys $\bar f \simeq (a_c-a)^2/b$.
+By similar arguments, the characteristic elastic modulus associated with microstructural deformations is
+%
+\begin{equation}
+\bar g\sim \frac{(a_c-a)a_c}{4b}
+\label{barg}
+\end{equation}
+%
+which is the energy density of the gradient terms only.
+The terms involving $\phi^2$ and $\phi^4$ in Eq.~\ref{FE-functional} are unaltered by affine deformations which conserve density locally; hence those terms do not contribute to $\bar g$.
+
+The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
+%
+\begin{equation}\label{OPEOM}
+\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
+\end{equation}
+%
+Here $M$ is a mobility coefficient.
+A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ (this is negative, signifying the instability).
+In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$.
+This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.
+
+Treating the fluid as incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
+%
+\begin{equation}\label{NSE}
+\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta}. 
+\end{equation}
+%
+where $\rho$ is the mass density. The stress tensor in this equation reads
+%
+\begin{equation}
+\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-p\delta_{\alpha\beta} - P_{\alpha \beta}. \label{fullstress}
+\end{equation}
+%
+This consists of a Newtonian hydrodynamic part, which is proportional to the background viscosity $\eta$ and the symmetrized velocity gradient, a hydrostatic pressure term, which includes an isotropic stress contribution from the order parameter
+%
 \begin{eqnarray}
-{\cal F}[\phi]&=&\int d^dx f(\phi({\bm x}))\nonumber\\
-&=&\int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\bm \nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\label{FE-functional}
+p_{\phi}&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
+& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
 \end{eqnarray}
-
-In order to ensure thermodynamic stability the parameters $b$ and $c$ have to be positive.
-For $a>0$ there is only one unique solution $\phi\equiv0$ which minimises the functional Eq. \ref{FE-functional} and wich reprsents the the disordered state.
-For $a<0$ a microphase separation occurs.
-Depending on the sign of $\kappa$ either two homogenous phases coexist ($\kappa>0$) or a transition to an ordered lamellar phase state takes place ($\kappa<0$) as the formation of interfaces become energetically favourable.
-The lamellar width is approximately given by $2\pi /\sqrt{|\kappa|/2 c}$, whereas the interfacial width is roughly $\sqrt{2|\kappa|/|a|}$. 
-This entails that by varying $a$ (and $b\sim |a|$) by about on decade, the segregation of the two phases varies by about a factor $3$.\\
-The dynamics of the binary fluid is governed by an advection-diffusion equation with a diffusive current that is proportional to the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$.
-
-\begin{equation}\label{OPEOM}
-\partial_t \phi +{\bm \nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
+%
+and finally another term consisting of the order-parameter contribution to the deviatoric pressure~\cite{Swift96,Yang76,Evans79} 
+%
+\begin{equation}
+P_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
 \end{equation}
 
-The coefficient $M$ herein plays the role of a mobility coefficient and the diffusivity of the system can be approximated by $D_0\simeq M\;b$.\\
-If the fluid is incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
+In our hybrid LB scheme, the Navier-Stokes equation (\ref{NSE}) is solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
+The velocity field is then used in the advection-diffusion equation Eq.\ref{OPEOM}, which is solved by a finite-difference method at the same timestep.
+We assume periodic boundary conditions along the flow and vorticity directions and Lees-Edwards sliding periodic boundary conditions\cite{leesedwards,Wagner02} along the direction of the velocity gradient.
+Following \cite{Wagner02}, we introduce
+additional internal Lees-Edwards planes to overcome an upper limit on shear rate that would otherwise be inversely proportional to the system size in the gradient direction.
+(This restriction arises because fluid velocities must be small compared to the sound speed in the frame of the lattice; the additional planes create a stack of lattice slabs whose velocity is nearly matched to the local fluid velocity.)
+The total number of Lees-Edwards planes was varied according to the size of the simulation so that the separation of two planes was always 16 lattice sites.
 
-\begin{equation}\label{NSE}
-\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha = \partial_\beta \sigma_{\alpha \beta}. 
-\end{equation}
-
-The stress tensor in this equation reads
-
-\begin{equation}
-\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-P^{th}_{\alpha \beta}.
-\end{equation}
-
-It consists of a hydrodynamic part, which is proportional to the background viscosity $\eta$ and velocity gradients and a thermodynamical part $P^{th}$ that depends on the order parameter and gradients thereof:
-
-\begin{equation}
-P_{\alpha\beta}^{th}=\left(\phi\frac{\delta {\cal F}}{\delta \phi} -f(\phi)\right) \delta_{\alpha\beta} + {\cal D}_{\alpha \beta}(\phi).
-\end{equation}
-
-In this last expression the symmetric tensor ${\cal D}_{\alpha\beta})$ has been added to make the pressure tensor divergence free and to guarantee mechanical equilibrium \cite{Evans79}.
-The thermodynamic pressure tensor is given by
-
-\begin{equation}
-P^{th}_{\alpha \beta}=p\, \delta_{\alpha \beta} + P^{chem}_{\alpha \beta}
-\end{equation}
-
-with an isotropic pressure contribution
-
-\begin{eqnarray}
-p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\bm \nabla}\phi)^2\right)\nonumber\\
-& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
-\end{eqnarray}
-
-and an anisotropic part
-
-\begin{equation}
-P^{chem}_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
-\end{equation}
-
-The Navier-Stokes Eq.\ref{NSE} was solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
-The velocity field was then used in the advection-diffusion equation Eq.\ref{OPEOM}, which was solved by a finite-difference method at the same timestep. 
-We assumed periodic boundary conditions along the flow and vorticity direction and Lees-Edwards boundary conditions (BC) \cite{Wagner02} along the direction of the velocity gradient.
-The total number of Lees-Edwards planes was varied according to the size of the simulation so that the distance of two planes was always 16 lattice sites.
-To impose simple shear flow, the planes in the upper half of the velocity gradient direction were moved at a constant speed in positive flow direction, whereas the planes in the lower hemisphere were moved in negative flow direction.\\
 Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
-
+%
 \begin{equation}
 C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
+\label{S}\end{equation}
+%
+with $\phi({\mathbf k})$ the normalized Fourier transform of the order parameter.
+The transform itself was calculated using the FFTW-library \cite{FFTW}.
+In principle an ensemble- or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.
+
+For the analysis of the relation between rheological quantities and the morphology of the system we need to to quantify the number of defects.
+For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average orientation of the order parameter gradients, as follows
+%
+\begin{equation}\label{alignment}
+X({\mathbf r})=\frac{1}{V'({\mathbf r})} \int_{V'({\mathbf r})} d{\mathbf r'} \left(\frac{\nabla \phi ({\mathbf r}) \cdot \nabla \phi ({\mathbf r'})}{|\nabla\phi ({\mathbf r}) ||\nabla\phi ({\mathbf r'}) |}\right)^2.
 \end{equation}
+%
+Here $V'$ is a small volume that contains $\mathbf r$; in practice $V'$ can be chosen to contain $\mathbf r$ and its nearest neighbours only.
 
-with $\phi({\mathbf k})$ being the normalised Fourier-transform of the order parameter.
-The transform itself was calculated using the FFTW-library \cite{FFTW}.\\
-For the analysis of the relation between rheological quantities and the morphology of the system it was necessary to find a means to quantify the number of defects.
-For this purpose we defined an alignment factor $X_i$, which measures the average projection of the normalised order parameter gradients onto the gradient at site $i$.
-More explicitly the alignment factor $X_i$ is given by
+For ordered regions with small curvature of the lamellar interfaces, $X$ is very close to unity.
+Wherever lamellae terminate at edge dislocations, and in regions where their orientation changes abruptly, $X$ departs from unity, taking smaller positive values.
+By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as an aligned lamellar configuration.
+We found a threshold value of $\xi=0.8$ to be well suited for our requirements.
+This protocol does not directly count the number of defects, since at each defect more than one site has an alignment function below the threshold.
+However, as the number of sub-threshold sites per defect is reasonably constant, the degree of order under flow can usefully be quantified by defining a ``defect density" as the system-wide fraction of sub-threshold sites:
+%
+\begin{equation}\label{defect density}
+\rho_D({\mathbf r})=\frac{1}{V} \int d{\mathbf r'} \theta(\xi-X(\mathbf r')).  
+\end{equation}
+%
+Here $\theta$ is the Heaviside function.
+The conversion factor between $\rho_D$ and the defect number density (in 2D) or line density (in 3D) depends on the lamellar wavenumber $q$ and the choice of threshold parameter $\xi$; we do not make such conversions here and consider $\rho_D$ itself as an appropriate measure of disorder.
 
-\begin{equation}\label{alignment}
-X_i=\frac{1}{N_{nn}} \sum_{j=1,N_{nn}}\left(\frac{\nabla \phi_i \cdot \nabla \phi_j}{|\nabla\phi_i||\nabla\phi_j|}\right)^2. 
+\section{Parameter Mapping to Physical Units}
+\label{parametermap}
+
+The parameter space of the model is quite large.
+Although in exploratory work we performed studies across a wider range of parameter values, for simplicity in the results presented here we restrict attention to fixed values of $\kappa = -6\times 10^{-4}$LBU and $c = 7.6\times 10^{-4}$LBU (here LBU denotes LB units as defined below).
+These are chosen to give a lamellar wavelength of $L = 10$ LBU (that is, $q = \sqrt{|\kappa|/2c} = 2\pi/10$LBU) which offers suitable spatial resolution.
+Moreover the microphase separation point, set by $a_c = \kappa^2/4c = 1.2\times 10^{-4}$LBU, is then a fixed constant.
+
+As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the fluid $\rho$ all to unity to define LB units (LBU).
+We need to calibrate these units (or equivalently, length, energy density and time) against real physical units.
+We do this first for thermotropic smectics, then consider the lyotropic case.
+
+To calibrate the LB length unit, we equate $L$ the lamellar wavelength, with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm.
+Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{9}$LBU.
+Free energy density parameters such as $a_c \simeq 10^{-4}$LBU should then be of order $\alpha^2k_BT/L^3$ where $\alpha =10$ is a molecular aspect ratio.
+Thus the LB unit of energy density is about $6\times 10^{10}$Pa, i.e.
+1 Pa$= 1.6 \times 10^{-11}$ LBU.
+
+Turning now to the fluid viscosity $\eta$, after pilot runs spanning a range of several decades, we selected $8.33\times 10^{-2}$LBU for the production runs reported here.
+This represents the baseline viscosity of the isotropic phase close to the transition (not the higher, effective viscosity of the smectic once formed).
+Requiring this to be about ten times the viscosity of water (which is $\eta_w = 8.9\times 10^{-4}$Pa s) we find that the LBU unit of viscosity is about $0.1$Pa s.
+Given the previous result for the energy density, we infer that 1s$= 6\times 10^{11}$ LBU.
+The LBU timestep is thus around $1.6\times 10^{-12}$s.
+
+As detailed previously, an estimate of the molecular diffusivity is $\tilde D = 2Mb$; we set $M = 0.25$LBU throughout this study, so that $\tilde D$ lies in the range $2.5\times 10^{-5} - 2.5\times 10^{-4}$LBU.
+Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{11}$LBU) gives $D\simeq 10^{-11}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.
+
+Recall that the mass density of our fluid is unity in LBU.
+However, combining the above conversion factors for length,  energy density, and time shows the physical and simulation units of mass density to be related via 1 kg m$^{-3} \equiv 1$ Pa s$^2$ m$^{-2} \simeq 10^{-8}$LBU.
+Thus our simulations address a fluid of mass density $10^6$kg m$^{-3}$ which is $10^3$ times larger than the true value for a typical material.
+The reason for this is that the LB algorithm uses fluid inertia to update the dynamics and the discretization should therefore involve replacing the real fluid with the densest one possible, while ensuring that the Reynolds number $Re \sim \rho\dot\gamma L^2/\eta$ (which for real smectics is exceedingly small) remains small enough for the simulation to faithfully describe the low $Re$ limit.
+As discussed elsewhere \cite{Cates04, Cates09} $Re$ values up to about 0.1 are then admissable; our simulations comply with this requirement.
+
+Finally, our shear rates lie in the range 
+$10^{-5}-10^{-4}$LBU, so that the (dimensionless) Peclet number, $Pe =\dot\gamma L^2/\tilde D$, which expresses the relative importance of advective to diffusive transport, lies in the range $5-800$.
+These are moderate to high values where significant structural rearrangement can be expected.
+However, for thermotropic smectics, given our time unit of 1LBU$=1.6\times 10^{-12}$s, they correspond to strain rates of order $10^6$s$^{-1}$.
+This lies well above the range typically addressable with rheological experiments, but below that recently given in a theoretical study \cite{Stewart2009}.
+
+The shear rates needed to achieve this range of $Pe$ falls dramatically when lyotropic smectics are considered.
+For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm.
+The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$)~\cite{Safran}.
+The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e.
+1 Pa$= 1.6 \times 10^{-6}$ LBU.
+Both the viscosity $\eta$ and the molecular diffusivity $\tilde D$ remain comparable (within one order of magnitude) to the thermotropic values; from the viscosity we then determine that 1s$=6\times 10^6$LBU or 1LB timestep $=1.6\times 10^{-7}$s.
+Accordingly the strain rates relevant to the $Pe$ range addressed here are of order $10^2$ s$^{-1}$ and well within experimental reach, but below that addressed in molecular dynamics simulations \cite{Guo2002,Soddemann2004}.
+Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still sufficiently small~\cite{Cates04}.
+
+To further simplify our parameter exploration, throughout what follows we restrict attention to  states with $b = -a$.
+This convention has been widely used to study bulk spinodal decomposition in binary fluids where it can be absorbed in a rescaling of the order parameter \cite{Kendon01}: put differently, $b=-a$ ensures that coexisting bulk ordered phases have $\phi = \pm\phi_0$ with $\phi_0=1$.
+In lamellar states of $b= -a$ the amplitude of the lamellar ordering need not be unity but varies like $\phi_q \sim \sqrt{1+a_c/b}$ as stated previously.
+Thus the chosen value of $b$ can be thought of as controlling whether the peak order parameter variation is close to that arising for bulk coexistence (when $b\gg a_c$), or whether the amplitude $\phi_q$ is instead much larger ($b\ll a_c$) and determined by a competition between the gradient terms and a relatively weak quartic nonlinearity.
+In this study we choose three values, $b = 5\times 10^{-4}, 1\times 10^{-4}$ and $5\times 10^{-5}$LBU so that the dimensionless quantity $\beta \equiv b/a_c$ is respectively $4.2,0.83,0.42$, lying near the crossover between these regimes.
+The observed $\phi_q$ range is about $1.2-1.5$, somewhat narrower than the range $1.1-1.8$ predicted from $\phi_q \sim \sqrt{1+a_c/b}$.
+However the latter estimate, which does not treat the nonlinear term exactly, is only a guideline.
+
+Alongside the Reynolds number (which we deem irrelevant), $\beta$, and the Peclet number, we can identify two further dimensionless numbers of interest.
+One is the Schmidt number $Sc = \eta/\rho\tilde D$, which is large when momentum can diffuse much more rapidly than molecules - as applies in almost all dense fluids.
+In our simulations $Sc$ is of order $100-1000$ which is safely large.
+(The work of Kumaran \cite{Kumaran2011} instead addresses Schmidt numbers in the range $0.2-5$ where different physics may apply.) The second parameter controls the strength of feedback between distortions of the $\phi$ field and the fluid motion.
+This
+can be expressed in terms of a characteristic value of the Peclet number, $Pe^*$, at which the free energy density in the order parameter field that couples to deformation (Eq.\ref{barg}), $\bar g=(a_c+b)a_c/4b$, is comparable to the viscous stress $\sigma_v=\eta\dot\gamma = (\eta \tilde D/L^2)Pe$.
+Equating these gives
+\begin{equation}
+Pe^* =\frac{(a_c+b)a_c}{8b^2}\,\frac{L^2}{\eta M}\label{pestar}
 \end{equation}
+With our chosen parameters (in LBU, $a_c=1.2\times 10^{-4}, L = 10, \eta = 8.3\times 10^{-2}, M = 0.25$) we find $Pe^* \sim 180$ at $\beta = 4.2$ and $Pe^* \sim 5000$ at $\beta = 0.42$.
+The maximum values of $Pe/Pe^*$ that we address below are of order 0.2 and 0.08 respectively in the two cases.
 
-$N_{nn}$ is the number of nearest neighbours ($N_{nn}=8$ in 2D and $N_{nn}=26$ in 3D).
-For smooth alignment and small curvature of the lamellae $X_i$ is very close to unity.
-At the endpoints of the lamellae or in regions where their orientation changes abruptly it departs form unity and takes smaller positive values.
-By setting a threshold $\xi$ it is possible to detect domains where the order parameter structure deviates preceivably from what would be recognised as an aligned configuration exhibiting lamellar order.
-We found a threshold value of $\xi=0.8$ to be well suitable for our requirements.
-Unfortunately, this defect detecting protocol is not able to assign only single sites to defect regions as naturally more than one site would have an alignment factor below the threshold.
-However, as the number of sites per defect is on average more or less constant throughout the sample, this protocol provides still a useful way to quantify the degree of order and alignment in various flow conditions.\\
-By normalising the number of defects sites with $X_i$ below the threshold a defect density can be defined:
+For later convenience we may define a new dimensionless number,
+which we can call a capillary number. (The capillary number is conventionally defined as the ratio of viscous to interfacial forces in binary fluid flow.) We define this as:
+%
+\begin{equation}
+Ca = \frac{Pe}{Pe^*} = \frac{4\eta\dot\gamma b}{(a_c+b)a_c}
+\label{capillary}
+\end{equation}
+%
+We will argue below that this could be a more relevant choice than $Pe$ for understanding the dependence of morphology on flow rate in smectic systems, at least when $Pe^*\gg 1$.
 
-\begin{equation}\label{defect density}
-\rho_D=\frac{1}{N} \sum_{i=1,N} \theta(\xi-X_i).  
-\end{equation}
+\section{Results and Discussion}
 
-The function $\theta$ is the Heaviside function and $N=N_x\times N_y \times N_z$ the total number of sites in the simulation.
-The exact value of the defect density clearly depends on the lamellar width and of course on the threshold $\xi$ for the alignment parameter.
-
-\subsection*{Parameter Mapping to Physical Units}
-
-In the following we describe how the simulation units can be mapped onto real physical units.
-It is necessary to calibrate the scales for mass density, length and time or alternatively as we do for mass density, length and energy density.\\
-As usual we set the lattice spacing $l$, time step $\Delta t$ and mass density of the reference fluid $\rho_0$ all to unity in LB units (LBU).
-It is straightforward to determine the unit of mass. 
-A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
-Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.\\
-In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig12} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
-Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
-Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space or $1 {\rm m}=2\cdot10^8$ LBU.\\
-The energy scale is slightly more difficult to calibrate.
-We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
-This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
-The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
-The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
-Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
-This formula reads 
-\begin{equation}
-b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
-\end{equation}
-whereas $\rho_c$ is the number density of chains per unit volume,
-often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
-Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
-With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
-Hence the LB unit of energy density is about $10^{6} \rm{Pa}$ or $1\rm{Pa}=10^{-6}$ LBU.\\
-We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
-As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=(1\e{-3}) \cdot (1\e{6}) \cdot (4\e{16})\rm{LBU}\simeq4\e{19} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq6.3\e{9}$ LBU and one LB timestep $\Delta t$ corresponds to roughly $0.15 \rm{ns}$.\\
-Finally we want to get an estimated scale for shear rates. 
-A common way is to express the ratio of transport via advection to that via diffusion by means of the Peclet number.
-The Peclet number can be defined as $Pe=\gd\, L^2/D_0$, whereas $\gd$ is the strain rate, $L$ is a characteristic length scale, e.g. the lamellar width and $D_0$ is the diffusivity.
-Because the Peclet number is a dimensionless quantity, we can perform this estimation conveniently in lattice units.
-The highest shear rate we apply are of the order $1\e{-4}$ in reciprocal LB time units.
-With the above given values for the lamellar width of $L\simeq10$ and diffusivity of system A $D_0=M\;b\simeq 0.25 \cdot 5\e{-4}=1.25\e{-4}$ this results in Peclet numbers $Pe=(1\e{-4})\cdot(1\e{2})\cdot(1.25\e{-4})^{-1}=80={\cal O}(10^2)$.
-Hence, the conversion between shear rates and Peclet numbers is roughly $\gd=10^{-4}\cong Pe=10^2$.
-
-\section{Results}
-
-Our discussion will be in two parts, beginning with simulations in two dimensions for systems A-2D and B-2D.
-The system size was $N_x\times N_y\times N_z=512\time1024\times 1$ and $N_x \times N_y \times N_z=256\times512\times 1$ respectively for the longtime runs in the nonlinear rheological regime.
-Phase segregation was controlled by modifying the Landau parameters $a$ and $b$ in the bulk free energy term of Eq.
-\ref{FE-functional} while keeping the parameters $\kappa$ and $c$ controlling the gradient terms constant.
-By this we obtained a slightly different stress response with the system with lower segregation level appearing somwhat sofer.
-The lamellar distance was identical in both cases.\\
-The second part of our discussion consists of simulations in three dimensions, referred to as A-3D and C-3D using identical parameters to system A-2D and C-2D and $N_x\times N_y\times N_z=128\times256\times128$.
-Note that velocity gradient and flow direction are oriented along the x- and y-axis, respectively, which is the {\it vertical} and the {\it horizontal} dimension in the 2D-plots. 
-Table \ref{tab1} gives an overview of all simulation parameters.
+Below we first present detailed results for two parameter sets ($\beta = 4.2, 0.42$) in two dimensions which we call systems A-2D and B-2D.
+System lattice sizes and shear rates are reported in Table \ref{tab1}; all remaining parameters are the same in both systems as detailed above.
+%As expected from the preceding discussion of $Pe^*(\beta)$, the stress response at a given flow rate is softer in the second system. 
+A third parameter set ($\beta = 0.83$, system C-2D) gave results broadly similar to B-2D.
+Note that we will later report simulations in three dimensions for systems A-3D and C-3D which share parameters with A-2D and C-2D (see Table \ref{tab1}).
 
@@
 \small
-\caption{Simulation parameters: Starting from spinodal decomposition the systems were first equilibrated at zero shear rate (EQ), followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\cdot10^{-5}$ to $2\cdot10^{-4}$.
-In some cases we performed a final switch off run (SO) at zero shear rate.
-The parameters $\kappa=-6\cdot10^{-4}, c=7.6\cdot10^{-4}, M=0.25$ and $\eta=8.33\cdot10^{-2}$ were the same throughout all simulations.}
-\label{tab1}
-\begin{tabular*}{0.495\textwidth}{lllll}
+\begin{tabular*}{0.52\textwidth}{l|l|l|l|l|l|l}
 \hline
-run \# & $N_x\times N_y\times N_z$ & time $\cdot 10^3$ & -a=b &  $\dot\gamma$\\
+run \# & $N_x\times N_y\times N_z$ & time $/10^5$ & $10^5b$ & Ca & Pe & $10^5\dot\gamma$\\
 \hline
-system A & & & &\\
+A & & & & & &\\
 \hline 
-703 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-4}$ & 0 - EQ  \\
-704 & $512\times1024\times1$ & 320-640& $5\cdot10^{-4}$ & $1\cdot10^{-4}$\\
-705 & $512\times1024\times1$ & 640-960& $5\cdot10^{-4}$ & 0 - SO-4 \\
+1 & $1024\times512\times1$ & 0-3.2 & $50$ &0 &0 & 0 \\
+%703
+2 & $1024\times512\times1$ & 3.2-6.4& $50$ & 0.224 & 40 &  $10$\\
+%704
+3 & $1024\times512\times1$ & 6.4-9.6& $50$ &0 & 0 &  0 \\
+%705
 \hline
-767 & $256\times512\times1$ & 0-320 & $5\cdot10^{-4}$  & 0 - EQ\\
-775 & $256\times512\times1$ & 320-640& $5\cdot10^{-4}$ & $2\cdot 10^{-4}$ \\
-769 & $256\times512\times1$ & 320-2560& $5\cdot10^{-4}$ & $1\cdot 10^{-4}$ \\
-771 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-4}$ & $ 5\cdot10^{-5}$ \\
-772 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-4}$ & $ 2.5\cdot10^{-5}$ \\
-774 & $256\times512\times1$ & 5120-10240& $5\cdot10^{-4}$ & $1.25\cdot10^{-5}$ \\
+4 & $512\times256\times1$ & 0-3.2 & $50$ &0 & 0 &  0\\
+%767
+5 & $512\times256\times1$ & 3.2-6.4& $50$ &0.448 & 80 &  $20$ \\
+%775
+6 & $512\times256\times1$ & 3.2-25.6& $50$ &0.224 & 40 &  $10$ \\
+%769
+7 & $512\times256\times1$ & 12.8-38.4& $50$ &0.112 & 20 &  $ 5$ \\
+%771
+8 & $512\times256\times1$ & 51.2-76.8& $50$ &0.056 & 10 &  $ 2.5$ \\
+%772
+9 & $512\times256\times1$ & 51.2-102.4& $50$ &0.028 & 5 &  $1.25$ \\
+%774
 \hline
-792 & $128\times256\times1$ & 320-1600& $5\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
+10 & $256\times128\times1$ & 3.2-16.0& $50$ &0.22 & 40 &  $10$ \\
+%792
 \hline
-786 & $128\times256\times128$ & 0-640 & $5\cdot10^{-4}$ & 0 - EQ\\
-787 & $128\times256\times128$ & 320-640 & $5\cdot10^{-4}$ & $2\cdot 10^{-4}$\\
-788 & $128\times256\times128$ & 320-960 & $5\cdot10^{-4}$ & $1\cdot 10^{-4}$\\
-789 & $128\times256\times128$ & 640-1280 & $5\cdot10^{-4}$ & $5\cdot 10^{-5}$\\
+11 & $256\times128\times128$ & 0-6.4 & $50$ &0 & 0 &  0\\
+%786
+12 & $256\times128\times128$ & 3.2-6.4 & $50$ &0.448 & 80 &  $20$\\
+%787
+13 & $256\times128\times128$ & 3.2-9.6 & $50$ &0.224 & 40 &  $10$\\
+%788
+14 & $256\times128\times128$ & 6.4-12.8 & $50$ &0.112 & 20 &  $5$\\
+%789
 \hline
-system B & & & &\\
+B & & & & & &\\
 \hline
-707 & $512\times1024\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ \\
-710 & $512\times1024\times1$ & 320-960& $5\cdot10^{-5}$ & $1\cdot10^{-4}$ \\
-765 & $512\times1024\times1$ & 960-1280& $5\cdot10^{-5}$ & 0 - SO-4 \\
+15 & $1024\times512\times1$ & 0-3.2 & $5$ &0 & 0 &  0 \\
+%707
+16 & $1024\times512\times1$ & 3.2-9.6& $5$ &0.082 & 400 &  $10$ \\
+%710
+17 & $1024\times512\times1$ & 9.6-12.8& $5$ &0 & 0 &  0 \\
+%765
 \hline
-776 & $256\times512\times1$ & 0-320 & $5\cdot10^{-5}$ & 0 - EQ\\
-777 & $256\times512\times1$ & 320-960& $5\cdot10^{-5}$ & $2\cdot 10^{-4}$ \\
-778 & $256\times512\times1$ & 320-2560& $5\cdot10^{-5}$ & $1\cdot 10^{-4}$ \\
-780 & $256\times512\times1$ & 1280-3840& $5\cdot10^{-5}$ & $5\cdot 10^{-5}$ \\
-781 & $256\times512\times1$ & 2560-6400& $5\cdot10^{-5}$ & $2.5\cdot 10^{-5}$ \\
-782 & $256\times512\times1$ & 5120-7680& $5\cdot10^{-5}$ & $1.25\cdot 10^{-5}$ \\
+18 & $512\times256\times1$ & 0-3.2 & $5$ & & 0 &  0 \\
+%776
+19 & $512\times256\times1$ & 3.2-9.6& $5$ &0.164 & 800 &  $20$ \\
+%777
+20 & $512\times256\times1$ & 3.2-25.6& $5$ &0.082 & 400 &  $10$ \\
+%778
+21 & $512\times256\times1$ & 12.8-38.4& $5$ &0.041 & 200 &  $5$ \\
+%780
+22 & $512\times256\times1$ & 25.6-64.0& $5$ &0.020 & 100 &  $2.5$ \\
+%781
+23 & $512\times256\times1$ & 51.2-76.8& $5$ &0.010 & 50 &  $1.25$ \\
+%782
 \hline
-system C & & & &\\
+C & & & & & &\\
 \hline
-998 & $128\times256\times1$ & 0-320& $1\cdot10^{-4}$ & 0 - EQ \\
-1001 & $128\times256\times1$ & 320-960& $1\cdot10^{-4}$ & $2\cdot10^{-4}$ \\
-1000 & $128\times256\times1$ & 320-1280& $1\cdot10^{-4}$ & $1\cdot10^{-4}$ \\
-999 & $128\times256\times128$ & 0-320 & $1\cdot10^{-4}$ & 0 - EQ\\
-1003 & $128\times256\times128$ & 320-960 & $1\cdot10^{-4}$ & $2\cdot10^{-4}$\\
-1002 & $128\times256\times128$ & 320-1280 & $1\cdot10^{-4}$ & $1\cdot10^{-4}$\\
+24 & $256\times128\times1$ & 0-3.2& $10$ &0 & 0 & 0 \\
+%998
+25 & $256\times128\times1$ & 3.2-9.6& $10$ &0.252 & 400 &  $20$ \\
+%1001
+26 & $256\times128\times1$ & 3.2-12.8& $10$ &0.126 & 200 &  $10$ \\
+%1000
+\hline
+27 & $256\times128\times128$ & 0-3.2 & $10$ &0 & 0 &  0 \\
+%999
+28 & $256\times128\times128$ & 3.2-9.6 & $10$ &0.252 & 400 &  $20$\\
+%1003
+29 & $256\times128\times128$ & 3.2-12.8 & $10$ &0.126 & 200 &  $10$\\
+%1002
 \hline
 \end{tabular*}
+
+\caption{Simulation parameters: After initialization from a noisy uniform state, the systems were first equilibrated at zero shear rate, followed by a sequence of various shear rates, ranging from $\dot{\gamma}=1.25\times10^{-5}$ to $2\times10^{-4}$LBU.
+In some cases we performed a final switch off run at zero shear rate.
+The other model parameters $\kappa=-6\times10^{-4}, c=7.6\times10^{-4}, M=0.25$ and $\eta=8.33\times10^{-2}$ were the same throughout all simulations.}
+\label{tab1}
 \end{table}
 
-\subsection*{Results in Two Dimensions}
+\subsection{Two Dimensions}
 
-Lamellar phases occur within the Landau-Brazovskii free energy functional approach Eq.
-\ref{FE-functional} due to the interplay of gain in free energy by virtue of phase separation and loss due to local curvature.
-Hence, a range of parametrizations can produce lamellar phases with almost identical outcome.
-In previous studies \cite{Kendon01,Xu03, Xu06b} parameters have been identified which cover the relevant morphology of lamellar phases and to produce stable outcomes. 
-We performed simulations for Landau parameters $5\cdot10^{-4}\le-a=b\le5\cdot10^{-5}$ and viscosities $8.33\cdot10^{-4}\le\eta\le 0.833$.
-Reasonable numerical stability was attained for all Landau parameters in this range and for viscosities $\eta\ge 8.33\e{-3}$.
-The morphology of the lamellar structure and the kinetics of the spinodal decomposition depended on the specific choice though.
-Smaller Landau parameters slowed down the ripening of the lamellar structure, which is consistent with the fact that systems with lower segregation have lower average order parameters and are generally closer to the critical point.
-Higher viscosities lead to shorter interfaces, whereas lower values typically invoke longer lamellae.
-This effect has been reported before \cite{Gonnella97}.
-It turned out to be more pronounced for low segregation, although not as quite to the same extent as in the above reference.
-In shear flow high values of $\eta$ entailed more inhomogenous lamellae in terms of shape and thickness.\\
-In the following we will show characteristic results for viscosity $\eta=8.33\e{-2}$ and the systems A-2D and B-2D, respectively.
-The top picture in Fig. \ref{fig1} shows the order parameter $\phi$ at the end of the equilibration phase prior to shearing.
-The equilibrium structure is completely isotropic and exhibits a lamellar spacing $L$ of roughly 10 lattice sites.
-If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
+\label{results2D}
+Before presenting specific results for the parameter sets detailed above, we briefly describe the effect of wider parameter variations.
+Within the Landau-Brazovskii free energy functional (Eq.~\ref{FE-functional}) a range of parameter variations can produce stable lamellar phases, as explored in previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011}.
+The nonequilibrium morphology of the lamellar structure, and the phase ordering kinetics (starting from a slightly noisy, uniform initial state) do however depend on the choice made for these, and also for the viscosity.
+Formation and ripening of the lamellar structure is slower when $\beta$ is small, consistent with the reduced initial driving force for ordering in this case.
+Higher viscosities lead to more locally disordered structures in which there is a shorter separation between kinks or edge defects as one moves along the lamellae; conversely lower viscosities typically yield ``longer" lamellae.
+This effect has been reported before \cite{Gonnella97}; we found it to be more pronounced for softer (lower $\beta$) systems.
+At fixed $\dot\gamma$, higher values of $\eta$ were found to cause greater inhomogeneity of lamellar shape and thickness, consistent with the fact that $Ca\propto\eta$ (the larger the shear stress, the larger the lamellar deformation).
 
+We turn now to system A-2D ($\beta = 4.2$). The top panel in Fig. \ref{fig1} shows the order parameter pattern $\phi$ at the end of the equilibration phase prior to shearing.
+Although the true equilibrium structure would show homogeneous alignment, this cannot be reached in a reasonable time scale unless shear is applied.
+The resulting metastable structure is amorphous and isotropic, but locally exhibits the same lamellar spacing $L = 10$ LBU as the fully ordered equilibrium state.
+If shear flow is imposed, the lamellae start to align in the flow direction, which induces a morphological changes to the structure.
+A typical snapshot in the resulting steady state at $Pe = 40$ $(Ca = 0.22)$ is shown in the middle panel of Fig.~\ref{fig1}.
+In this and all subsequent images, the applied flow velocity is horizontal and the velocity gradient vertical, with the flow directed to the right in the upper half and to the left in the lower half of the picture.
+After a short transient during which the order parameter pattern remains relatively homogenous (not shown) there emerge extended regions in which the smectic layers are either expanded or compressed, so that $L$ locally deviates from the equilibrium value.
+These mesoscopic inhomogeneities are inclined at an angle to the flow and remain continuously present in the steady state, but are not stationary: their local features are advected by the flow and evolve continuously by deformation.
 \begin{figure}[htp]
@@
 \includegraphics[angle=0,width=0.5\textwidth]{phi_run705_1280.jpg}
-\caption{Order parameter $\phi$ of system A-2D: after equilibration (timestep $t=3.2\e{5}$, top), at $\dot{\gamma}=1\cdot10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
-Gray scaling from black to white corresponds to values of $-1\le\phi\le1$}
+\caption{Order parameter pattern $\phi({\bf r})$ of system A-2D (thresholded to enhance visibility): after equilibration (timestep $t=3.2\e{5}$, top), in steady shear at $\dot{\gamma}=1\times10^{-4}$ ($t=5\e{5}$, centre) and after switch-off ($t=1.28\e{6}$, bottom).
+Gray scaling from black to white corresponds to values of $-1\le\phi\le1$.}
 \label{fig1}
 \end{figure}
+
+We found that after the shear flow was switched off the lamellar spacing returned to its equilibrium value everywhere, with expanded regions contracting and vice versa.
+The final state, after all relaxation has (virtually) come to a standstill, is shown in the bottom panel of Fig.~\ref{fig1}.
+Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression undergone during the previous flow, some regions retain strong misalignment which does not anneal away.
+The separation between defects as one moves along the lamellae varies considerably in space; both short and long separations can be seen.
+This applies both to the aligned layers and to layers that are misaligned with the flow.
 
@@
 
-The flow direction and the direction of the velocity gradient are {\it vertical} x- and {\it horizontal} y-direction.
-The relative flow imposed by the Lees-Edwards planes is always in positive y-direction in the upper half and in negative y-direction in the lower half of the images.
-After a short transient phase during which the order parameter remains rather homogenous, regions of reduced and increased lamellar width emerge and the spacing begins to deviate from its well-defined equilibrium value.
-These regions are advected with the flow while they continuously undergo affine transformations.
-The centre picture in Fig. \ref{fig1} shows an example at an intermediate time step at shear rate $\gd= 1\e{-4}$.
-In a previous work on binary fluids and lamellar systems in shear flow \cite{Xu06b} another shear-induced structure was found.
-However, the reported results differ from our observations in the following aspects.
-Xu et al. shear-induced structures emerge at the centre and remain localised, while ours do occur in band-like patterns that are wrapped around the periodic boundaries of our system.
-Xu et al. also reported a shear banding region with reduced local shear rate, which is absent in our case.
-These differences may be simply a direct consequence of the different boundary conditions that were applied.
-While we applied Lees-Edwards BCs to all quantities, Xu and co-workers seem to have used a kind of hybrid scheme.
+Fig. \ref{fig2} shows equivalent states for system B-2D ($\beta = 0.42$).
+The Peclet number during the shearing phase is now $Pe = 400$ $(Ca = 0.082)$.
+All the generic results of Fig.~\ref{fig1} are the same, but the defect density, both during flow and after switchoff, is significantly reduced.
+(However the bands of misalignment in the flowing state are, if anything, even more apparent.) Consequently, the lamellar alignment along the flow direction after switch off is also more complete.
 
-\begin{figure}[htp!]
+It is notable that at the shear rate shown in these figures, the more ordered system is B-2D which has higher $Pe$ than A-2D but smaller $Ca$.
+Varying $\dot\gamma$ within system A-2D or B-2D establishes unambiguously that the steady state order decreases with shear rate at fixed thermodynamic parameters (Fig.\ref{fig6} below).
+This points strongly to $Ca$ as the more relevant measure of flow rate when discussing onset of the disordered phase.
+
+In principle the dependence of the state of order on flow rate could be smooth, or could entail a nonequilibrium phase transition -- for instance if the perfectly ordered state ceased to be stable above a certain shear rate \cite{Gonnella98,Zilman99,Kumaran2001,Ramaswamy92}.
+In practice, however, starting as we do from an initially amorphous lamellar texture, a perfectly ordered state is generally not reached for large systems even if it might be stable once formed.
+In any case it seems likely that at high $Pe$, any defects introduced into the smectic state by deformation of an initially imperfect structure (whose layers are then dilated and compressed by the flow) cannot anneal away or annihilate each other rapidly.
+Such irregularities then build up until the steady state defect density is determined by a balance between their rate of introduction and rate of annealing.
+
+Recently Kumaran and Raman studied a single shear rate and varied the Schmidt number $Sc$ and system size, reporting enhanced alignment on decreasing $Sc$ and in small systems \cite{Kumaran2011}.
+Our comparative studies on systems A and C have confirmed both trends (although, as mentioned previously, our $Sc$ values are much higher).
+The same authors also discuss a distinct mechanism of defect creation caused by layer compression and splitting.
+Although we have not checked this in detail, this may be consistent with our observation of a strongly compressed and heterogeneous morphology, with bands oriented normal to the compression axis, as shown in Figs.~\ref{fig1} and \ref{fig2}.
+
+A somewhat different morphology in sheared lamellar systems was reported by Xu et al.\cite{Xu06b}.
+This morphology exhibits macroscopic inhomogeneity between the centre of the simulated domain and the regions near the walls, accompanied by variations in the local shear rate.
+This contrasts with our own work which shows translationally invariant statistics in all directions.
+These differences may be a result of Xu et al.\cite{Xu06b} using mixed boundary conditions, as discussed already in the introduction, in which hard walls for the fluid flow are combined with periodic BCs on the order parameter field.
+
+\begin{figure}[htp]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
+\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
+\caption{Defect structure of system A-2D: at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\times10^{-5}$.
+The system has been pre-sheared at stepwise decreasing rates (see Table \ref{tab1}).
+Sites where the alignment function $X$ is smaller than $0.8$ are highlighted red.
+The system size was $N_x\times N_y \times N_z=512\times 256 \times 1$.}
+\label{fig3}
+\end{figure}
+
+To quantify our observed morphology more closely, we use the alignment function $X$ introduced in Eq.~\ref{alignment}.
+Fig.~\ref{fig3} shows snapshots of the time evolution of system A-2D at modest shear rate.
+All lattice sites with $X({\mathbf r})\le0.8$ are marked in red and they obviously coincide with lamellar endpoints (edge dislocations) or other local defects.
+The upper panel gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, immediately following a pre-shearing protocol in which the strain rate was ramped down stepwise from higher values (see Table \ref{tab1}).
+While large aligned regions are present there are other regions of high defect density.
+The lower panel of Fig.~\ref{fig3} shows a later state with a total strain incremented by a further $6400\%$ lattice units (i.e. by $64$ times the system size). 
+Obviously many defects have annealed away and the smectic layers are now rather well aligned with the flow direction.
+
+Defect annihilation to this large extent is only found at low enough shear rates.
+A prerequisite seems to be that the defects can smoothly approach each other and then annihilate.
+This is prevented by the buckling of layers above a certain shear rate, beyond which the system shows a tendency to trip over itself, which results in the creation of new defects.
+The onset of defect proliferation arises at characteristic flow rates $\dot{\gamma}_c=1.25-2.5\e{-5}$LBU in system A-2D and around $\dot{\gamma}_c=1-2\e{-4}$LBU in B-2D.
+(Here the subscript $c$ could stand for ``critical" or for ``crossover" depending on whether there is a singular transition or not; see above.)
+These values are for
+systems of size $512\times256\times1$; we found evidence of an increasing trend with system size, but the extremely long run-times involved ($\sim 10^7$ LB timesteps) precluded quantitative exploration.
+These results correspond respectively to values of the Peclet number, $Pe_c$, obeying $5<Pe_c<10$ (system A) and $400<Pe_c<800$ (system B).
+However, in terms of $Ca = Pe/Pe^*$ one has $0.056<Ca_c<0.112$ (system A) and $0.082<Ca_c<0.164$ (system B).
+Given the 80-fold difference in $Pe_c$, the similarity of $Ca_c$ in the two different systems is remarkable.
+
+Boldly extrapolating across parameter regimes that we have not explored in any detail, our results motivate the hypothesis that the onset of defect proliferation occurs generically in 2D at $Ca_c\simeq 0.08$.
+Note that $Ca$, which comprises the ratio of a viscous stress to an elastic modulus, can also be viewed as a strain.
+In that context one can imagine that defects proliferate when the steady-state elastic deformation of the smectic exceeds a yield strain value (for which a threshold of a few percent is reasonable).
+This makes sense if, at $Ca \simeq Ca_c$, the stress contributed from the order parameter is (a) comparable to the viscous stress from the background fluid -- essentially the force balance argument made above -- and (b) primarily elastic in character.
+The latter in turn requires that $Pe_c$ is large and hence that $Pe^*$ is larger still.
+A different mechanism might apply at small $Pe^*$, since in that case layers relax rapidly by diffusion on the time scale set by $Ca_c$.
+
+\begin{figure}[htp]
 \centering
@@
 \includegraphics[angle=0,width=0.49\textwidth]{N1_run704_400.jpg}
-\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$}
-\label{fig3}
+\caption{Order parameter $\phi({\mathbf r})$ (top), shear stress $\sigma_{xy}({\mathbf r})$ (centre) and first normal stress differences $N_1({\mathbf r})$ (bottom) of system A-2D at $t=4\e{5}$ (see Table \ref{tab1}).}
+\label{fig4}
 \end{figure}
 
-\begin{figure}[htp!]
+In Fig.~\ref{fig4} we explore the correlation between the inhomogeneous morphology in the flowing state and locally defined rheological quantities.
+The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N_1({\bm r})=\sigma_{xx}({\bm r})-\sigma_{yy}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$ ($Pe = 40$, $Ca = 0.22$).
+Here and below we adopt a coordinate system where $x$ is along the flow direction and $y$ the velocity gradient.
+In regions where the lamellar width is larger than in the equilibrium state, shear stresses and first normal stress differences show low positive or negative values.
+Large positive values of both quantities coincide with regions where the lamellae are under compression.
+Peak values (in LBU) are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
+Despite having locally negative regions, spatial and ensemble averages of these quantities are always positive.
+Very similar results, but with smaller peak magnitudes, are obtained for system B-2D at the same shear rate.
+
+\begin{figure}[htp]
 \centering
@@
 \includegraphics[angle=0,width=0.33\textwidth]{ck_run705_1280.jpg}
-\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (left), at $\dot{\gamma}=1\cdot10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (right).
-The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2 l,\pi/2 l]$.}
-\label{fig4}
+\caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (top); in steady shear at $\dot{\gamma}=1\times10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (bottom).
+The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU with $k_x$ and $k_y$ along the horizontal and vertical coodinate direction, respectively.}
+\label{fig5}
 \end{figure}
 
-To our understanding Lees-Edwards BC were applied to the order parameter only, whereas solid moving walls were placed in the LB-part at the upper and lower boundaries.
-It might be that in \cite{Xu06b} actually not bulk flow behaviour was simulated, but rather flow in a confined geometry. 
-This could also explain why Xu et al. always observe a shear-induced structure at the centre of the system.\\ 
-After the shear flow has been switched off the lamellar spacing became again more homogenous and the equilibrium value was restored with expanded regions contracting and vice versa.
-The final state after the relaxation has virtually come to a standstill is shown in the bottom picture of Fig. \ref{fig1}. 
-Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression level during flow some regions may have large angles after relaxation. 
-The length of the interfaces varies considerably, which is another residue of the inhomogeneities present in shear flow.\\
-Fig. \ref{fig2} shows equivalent states for system B-2D with lower segregation.
-All generic results of Fig. \ref{fig1} are the same, but the interfaces appear to be slightly longer.
-Consequently, the lamellar alignment in flow direction after switch off is also more pronounced.\\ 
-In Fig. \ref{fig3} we suggest a relation between the inhomogeneous morphology in the flowing state and typical rheological quantities.
-The pictures show the order parameter $\phi({\bm r})$, shear stress density $\sigma_{xy}({\bm r})$ and first normal stress difference density $N1({\bm r})=\sigma_{yy}({\bm r})-\sigma_{xx}({\bm r})$ of system A-2D in steady shear flow at shear rate $\dot{\gamma}=1\e{-4}$.
-In regions where the lamellar width is larger than in the equilibrium state shear stresses and first normal stress differences show low positive or negative values.
-Large positive values of both quantities coincide with regions where the lamellae are under compression.
-Peak values in LBU are in the region of $-2\e{-4}$ and $3\e{-4}$ for shear stress and  $-3.5\e{-4}$ and $7\e{-4}$ for the normal stress differences.
-Despite being locall negative, the ensemble averages of these quantities are always positive.
-Very similar results with smaller peak magnitudes are obtained for system B-2D.
-Fig. \ref{fig4} shows structure factors $C({\bm k})$ for the same states as Fig. \ref{fig1}.
-Initially isotropic after spinodal decomposition and equilibration, $C({\bm k})$ has a distinct peak at $|{\bm k}|\simeq \pm0.2\pi/l$.
-In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi/l, k_y\simeq\pm 0.22 \pi/l$.
-It is interesting to see that the inhomogenous morphology of the lamellar structure is also clearly encoded in the structure factor. 
-The Fourier-vector at the peak has a small component in x-direction, indicating the average tilt of the lamellae away from the flow direction in the direction of extensional flow.
-The position of the peak is almost constant throughout the entire simulation and the average lamellar spacing decreases from about 10 lattice sites in equilibrium to about 8.5 lattice sites.
-If the shear flow is switched off the system adopts again its equilibrium value and the small tilt of the signal vanishes completely, indicating that this is a purely dynamical phenomenon. 
-The intensity of the peak is slightly larger due to the fact that the interfaces remain oriented along the flow direction.
-By introducing the alignment factor $X_i$ in Eq. \ref{alignment} we suggest a method to identify and quantify defects in the lamellar structure.
-Fig. \ref{fig5} shows an example of the time evolution of system A-2D at low shear rate.
-All lattice sites with $X_i\le0.8$ are marked in red and they obviously coincide with lamellar endpoints or local disturbances.
-The top picture gives the situation at the lowest applied shear rate $\dot{\gamma}=1.25\e{-5}$, whereas the system has been pre-sheared at consecutively decreasing shear rates between $\gd=1\e{-4}$ and $2.5\e{-5}$.
-There are regions which show some sort of flow alignment and others which don't.
-The bottom picture of Fig. \ref{fig5} shows a later state with a total strain that has increased by another $10^{5}\%$. 
-Obviously many defects have annealed and the interfaces are well aligned with the flow direction.
+Fig.~\ref{fig5} shows structure factors $C({\bm k})$ for the same states as Fig.~\ref{fig1}.
+$C({\bm k})$ is initially isotropic after layer formation and equilibration in the system at rest with, as expected, a distinct peak at $|{\bm k}|\simeq \pm2\pi/10$ LBU (recall $L = 10$ LBU).
+In steady shear flow the structure factor develops two distinct but rather broad peaks at $k_x\simeq\mp 0.012\pi, k_y\simeq\pm 0.22 \pi$.
+Thus the inhomogenous morphology of the lamellar structure is clearly encoded in the structure factor. 
+The wavevector at the peak has a small component in the $x$-direction, indicating the average tilt of the lamellae away from the flow direction towards the elongational axis of the flow (which in shear is at $\pi/4$ to the flow direction).
+
+When the shear flow is switched off, the system relaxes again to its equilibrium layer spacing and the small tilt of the $C({\bm k})$ signal decreases to an undetectable level: this is consistent with the fact that, in Figs.~\ref{fig1} and \ref{fig2} (bottom panels) it is easy to see when the system has been sheared horizontally, but not in which sense.
+The intensity of the peak is slightly enhanced in the final state due to the improved orientation of the layers along the flow direction.
+
+Fig.~\ref{fig6} shows the time evolution of the defects density $\rho_D$ of system A-2D in the sequence of stepwise decreasing shear rates alluded to previously (starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$, Table \ref{tab1}).
+The runs with the two highest shear rates were started from the amorphous, post-equilibration configuration, whereas all others were pre-sheared using the final configuration of the preceding run as the initial state.
+The total strain between the beginning and end of every individual run was at least $3\times 10^{3}\%$.
+Interestingly, following each step down in shear rate, both systems A-2D and B-2D continuously anneal to lower defect densities on a timescale $\simeq 10^6$LBU.
+If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur on a timescale corresponding to a stain of about 100\%.
+This finding offers further support for a dynamical equilibrium between defect creation and annihilation above a critical (or crossover) shear rate, below which defects progressively annihiliate and near-perfect alignment eventually results.
+To test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
+The defect density indeed goes up (not shown) and reaches similar values to before.
+Note however that the defect annealing time (of order $10^6$LBU) is quite long compared to any inherent relaxation time; also, the final state after switchoff strongly depends on the previous flow.
+Thus the rheology and microstructure of these materials is inherently history dependent.
+
+\begin{figure}[htp]
+\centering
+\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
+\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D in a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
+Horizontal lines indicate time averages after entering the apparent steady state (where present).
+The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).}
+\label{fig6}
+\end{figure}
+
+The top picture in Fig.~\ref{fig7} shows the spatially averaged total shear stress $\langle \sigma_{xy}({\bf r})\rangle \equiv\sigma_{xy} $.
+At low shear rates, stress fluctuations have longer temporal correlations but smaller amplitudes.
+(The latter dependence is obscured by the logarithmic vertical scale in this plot.) The root-mean-square temporal fluctuations can be reduced by increasing system size (by a factor 3 or so for the runs shown in Fig.
+\ref{fig1} and Fig.
+\ref{fig2}).
+In contrast, the averaged shear stress is almost independent of the system size, which makes it possible to simulate longer times using smaller system sizes while still giving accurate predictions for the bulk rheological behaviour.
+We pursue this strategy here.
 
@@
 \centering
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_5120.jpg}
-\includegraphics[angle=0,width=0.5\textwidth]{phi_defects_run774_10240.jpg}
-\caption{Defect structure of system A-2D: The pictures show results at timestep $t=5.12\e{6}$ (top) and $1.024\e{7}$ (bottom) with an imposed shear rate $\dot{\gamma}=1.25\cdot10^{-5}$.
-The system has been pre-sheared at various higher, subsequently lowered shear rates ranging from $\dot{\gamma}=2\e{-4}$ to the above value (see Tab.\ref{tab1}).
-Sites where the alignment factor $X_i$ was smaller than $0.8$ are highlighted red.}
-\label{fig5}
-\end{figure}
-
-Defect annihilation to this large extent is only found below a specific shear rate.
-A prerequisite seems to be that the defects can smootly approach each other and cancel, which is not possible above the instability as the lamellae buckle similar to \cite{Gonnella98}, show a tendency to trip over themselves which results in the creation of new defects.
-The critical value depends on the segregation and differs in systems A-2D and B-2D.
-It is roughly between $\dot{\gamma}=2.5-5\e{-5}$ in A-2D and around $\dot{\gamma}=1-2\e{-4}$ in B-2D.\\
-Fig.
-\ref{fig6} shows the  complete time evolution of the defects density $\rho_D$ of system A-2D for a sequence of decreasing shear rates starting at $\gd=2\e{-4}$ and ranging down to $\gd=2.5\e{-5}$.
-The runs with the two highest shear rates were started from the equilibrium configuration, whereas all others were pre-sheared using the final configuration of the preceding run as initial configuration in order to save computational rexources.
-The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
-Interestingly, while stepping down in shear rate both systems A-2D and B-2D anneal defects almost instantaneously.
-If the shear rate is turned up again (not shown), the lamellae buckle and new defects recur immediately.
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{defect_density_5e-4.pdf}
-\caption{Defect density $\rho_D$ vs time: The main picture shows data for system A-2D of a sequence of simulations with consecutively lowered shear rate $\dot{\gamma}=2\e{-4}$ (red solid), $1\e{-4}$ (green long-dashed),  $5\e{-5}$ (blue short-dashed),  $2.5\e{-5}$ (magenta dotted) and  $1.25\e{-5}$ (cyan dashed-dotted).
-Horizontal lines indicate time averages of the particular quantities, where we assume the system has reached a steady state.
-The inset gives data for the system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
-\label{fig6}
-\end{figure}
-
-This finding is in further support of a dynamical equilibrium between defect creation and annihilation above a critical threshold, wheras below the critical value defects cancel and lamellae align.
-To test this assumption we performed simulations starting from a very well aligned state and increased the shear rate.
-The defect density goes indeed up (not shown) and reaches similar values as before.\\
-Fig \ref{fig7} shows the ensemble averaged total shear stress $\langle \sigma_{xy}\rangle \equiv\sigma_{xy} $, i.e. the total shear stress per unit lattice volume.
-At low shear rates stress fluctuations become actually more stretched in time and smaller, a fact that is obscured by the logarithmic scales.
-In larger simulations such as those shown in Fig. \ref{fig1} and Fig. \ref{fig2} they are also about three times smaller.
-Nevertheless we found that the averaged shear stress is to a very good degree independent of the system size, which makes it possible to simulate longer times using smaller system sizes without flawing the results.
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{S_xy_tot_t_5e-4.pdf}
-\caption{Average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$ the apparent viscosities are $\eta_{app}=\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$.
- }
+\includegraphics[angle=0,width=0.45\textwidth]{S_xy_tot_t_5e-4.pdf}
+\includegraphics[angle=0,width=0.45\textwidth]{P_xy_chem_t_5e-4.pdf}
+\includegraphics[angle=0,width=0.465\textwidth]{P_xy_defect_density.pdf}
+\caption{Top: average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$, we find $\eta_{app}/\eta\equiv\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$; centre: average order parameter contribution to the pressure tensor $-P_{xy}$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right); bottom: average chemical contribution to the pressure tensor $-P_{xy}$ vs.
+defect density for system A-2D.}
 \label{fig7}
@@
 
-System A-2D shows clear shear thinning behaviour and seems to obey a power law (see below Fig. \ref{fig11}).
-Values of the apparent viscosity are given in the caption of Fig. \ref{fig7}.
-Contrary to system A-2D system B-2D does not exhibit a clear tendency of shear thinning.
-Here, the apparent viscosities are $\eta_{app}=1.57, 1.62, 1.55, 1.42, 1.87$ for the above shear rates and follow a more horizontal trend.\\
-We want to focus now on the deviatoric part of the chemical pressure $\langle P^{chem}_{xy}\rangle\equiv P^{chem}_{xy}$.
-This is shown for system A-2D in Fig. \ref{fig8}.
-The fluctuations are in phase with those of the total stress in Fig. \ref{fig7} and decrease with decreasing shear rate.
-In System B-2D $P^{chem}_{xy}$ decreases sharply during the run with $\gd=1\e{-4}$ (not shown) because at this shear rate pronounced defect annihilation sets in.\\
+System A-2D shows clear shear thinning behaviour; values of its apparent viscosity at various shear rates are given in the caption of Fig.
+7.
+The corresponding range of $Ca$ is 0.028-0.448.
+Using Eq.\ref{fullstress}, we can identify the stress contribution arising from the deviatoric part of the order parameter pressure tensor
+$\langle - P_{xy}({\bf r})\rangle\equiv - P_{xy}$; this is shown in the centre graph in Fig.~\ref{fig7}.
+The temporal fluctuations are in phase with those of the total stress and decrease with decreasing shear rate.
+The bottom image in Fig. \ref{fig7} plots $- P_{xy}$ against the defect density $\rho_D$; this shows a linear relation across almost the whole range of shear rates.
 
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{P_xy_chem_t_5e-4.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs time for system A-2D.
-The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
-\label{fig8}
-\end{figure}
-
-Fig. \ref{fig9} illustrates dependence of $P_{xy}$ on the defect density $\rho_D$.
-System A-2D features a linear relation between both quantities that breaks only down at the lowest defect density or rather shear rate.
-The result for the softer system B-2D are less clear due to the ordering and annihilation of defects.
-It is very likely that our quantifying procedure, which is based on averaging over a larger number of defects sites,is prone to inaccuracy in this regime.
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[angle=0,width=0.5\textwidth]{P_xy_defect_density.pdf}
-\caption{Average chemical contribution to the pressure tensor $\langle P_{xy}^{chem}\rangle$ vs. defect density for system A-2D and B-2D } 
-\label{fig9}
-\end{figure}
-
-Fig. \ref{fig10} shows averaged first normal stress differences $\langle N_1\rangle \equiv N_1$ versus time for system A-2D, where we followed the same protocol as before.
-Initially, down till $\gd=5\e{-5}$ $N_1$ decreases monotonously with decreasing shear rate.
-However, at lower shear rates the behaviour is not monotonous anymore and $N_1$ rises again.
-Hence, it is difficult to define a reasonable time average for $N_1$ as for these low shear rates the system may still evolve at the end of the run.
+In contrast, system B-2D does not exhibit clear shear thinning; the apparent viscosities obey $\eta_{app}/\eta=1.57, 1.62, 1.55, 1.42, 1.87$ for the same set of shear rates as used in Fig.~\ref{fig7}.
+To within the measured accuracy, system B-2D is thus rheologically quasi-Newtonian.
+However, the range of $Ca$ in this case is lower, 0.01-0.164, and the order parameter stress creates only a relatively modest uplift to the baseline viscosity.
+More significantly, $Ca\le Ca_c$ for all but the highest applied shear rate so that the steady states found in system B-2D are far more ordered than in A-2D.
+This results from a defect annihilation process, occurring in the early stages of ramp-down, which we found to be accompanied by a sharp drop in $P_{xy}$ (not shown).
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{N1_t_5e-4.pdf}
-\caption{Average first normal stress differences $\langle N_1 \rangle= \langle \sigma_{yy}-\sigma_{xx}\rangle$ vs time for system A-2D.
-The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\cdot10^{-4}, 1\cdot10^{-4}$ and $5\cdot10^{-5}$.}
-\label{fig10}
+\caption{Spatially averaged first normal stress differences $N_1= \sigma_{xx}-\sigma_{yy}$ vs time for system A-2D.
+The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right).
+Horizontal solid lines are time averages.}
+\label{fig8}
 \end{figure}
 
-This feature which appears somewhat erratic at first has actually a simple explanation. 
-The time sequence of the defect density in Fig. \ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$ and linked to the defect annihilation.
-When two defects meet and cancel the lamellar width increases locally by a certain amount.
-This causes frustration in the system as the incommensurability between the layers is not large enough to create a new lamellar layer and allow the system to snap back towards its equilibrium width.
-Hence the normal stresses $\sigma_{xx}$ become increasingly smaller and may even attain negative values, leading to gradually larger values of $N_1$.
-This explanation is further sustained by the fact that only $\sigma_{xx}$ seems to evolve anomalously slowly and all other components of the stress tensor reach their steady state values more quickly.
-We expect that these anomalous trends become considerably smaller in larger systems.
-System B-2D behaves differently in the sense that $N_1$ does not follow this trend but rather continues to fall for lower shear rates. 
-This is likely to be a direct consequence of the lower segregation in system B-2D, which gives rise to positive normal stresses $\sigma_{xx}$ preventing $N_1$ from increasing.\\
-In Fig. \ref{fig11} the shear rate dependence of $\sigma_{xy}$ and $N_1$ is shown. 
-System A-2D gives rise to shear thinning with the average shear stress of systems A-2D following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
-For system B-2D the best fit is actually obtained for an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning.
-This is because due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$ the chemical stress $P^{chem}_{xy}$ almost vanishes and the total stress is dominated by the hydrodynamic part.\\
+Fig.~\ref{fig8} shows averaged first normal stress differences $\langle N_1({\bf r})\rangle \equiv N_1$ versus time for system A-2D, where we report the same stepwise decreasing flow protocol as before.
+For the first several steps $N_1$ decreases monotonously with decreasing shear rate.
+However, at the lowest shear rates the behaviour is not monotic: $N_1$ eventually rises again. (The resulting time averages, shown as horizontal lines, do not therefore reflect true steady states.)
+This somewhat surprising feature has a simple explanation. 
+The time sequence of the defect density in Fig.
+\ref{fig6} reveals that the rise in $N_1$ is correlated with a drop in $\rho_D$, and linked to the onset of defect annihilation (which happens only at low shear rates).
+When two defects meet and cancel, the local lamellar width alters by a certain amount.
+This causes frustration as the system is unable to snap back locally towards its equilibrium layer spacing.
+Hence the normal stress difference $N_1$ $\sigma_{xx}-\sigma_{yy}$ fails to decay to zero, and remains positive at large times, as appropriate for a system of metastable and (on average) compressed smectic layers.
+The time dependence of the normal stress difference is also anomalously slow, whereas the shear stress reaches their steady state values more quickly.
+The relatively large steady state value of $N_1$ and this slow dynamics may partly be a finite size effect caused by the relatively modest number of smectic layers within the simulated volume.
+The normal stress in system B-2D (not shown) behaves differently: in this case $N_1$ falls monotonically during the defect annihilation phase. 
+It is not clear whether this difference should be attributed to the change in $\beta, Ca$ or $Pe$ between the two systems, or indeed a combination of those changes.
 
@@
 \includegraphics[angle=0,width=0.5\textwidth]{S_xy_N1_gammadot.pdf}
-\caption{Flow curves: Average shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D ($\eta=8.33\cdot10^{-2}, -a=b=5\cdot 10^{-4}$)}
-\label{fig11}
+\caption{Flow curves: Time-averaged shear stress $\langle \sigma_{xy}\rangle$ and first normal stress differences $\langle N_1 \rangle$ vs shear rate $\dot{\gamma}$ for system A-2D and B-2D.
+The dash-dotted lines are power-law fits to the data.}
+\label{fig9}
 \end{figure}
 
-\subsection*{Results in Three Dimensions}
+Fig.~\ref{fig9} shows `flow curves' representing the shear rate dependence of time averaged stresses $\sigma_{xy}$ and $N_1$.
+This is calculated from the stepwise ramp data as detailed previously, so the $N_1$ results at low shear rates should not be taken as definitive.
+System A-2D shows shear thinning as already discussed, apparently following a power law $\sigma_{xy}=\dot{\gamma}^m$ with exponent $m\simeq 0.87$.
+A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}; interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, folding  and formation of defects.
+They report an experimental value of $m\simeq 1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
+Our value of $m=0.87$ for the disordered flow regime is larger.
+One possible explanation for this discrepancy is that screw defects, which do not exist in our 2D simulations, play a key role for this theory.
+On the other hand our fitted power law must be treated with great caution since the fit includes data from $\dot\gamma$ both above and below $\dot\gamma_c$.
+It is somewhat surprising that the morphological changes around $1.25\e{-5}<\dot\gamma_c <2.5\e{-5}$ in system A give no visible anomaly or feature in the flow curves.
+%This may be partly explained with a more crossover-like behaviour of system A at the lowest applied shear rate.
+%
+For system B-2D the data is consistent with an exponent $m=1$, indicating more or less constant viscosity and no pronounced shear thinning (as previously described). 
+This appears to be due to the decreasing number of defects at shear rates below $\dot{\gamma}\le2\e{-4}$; the order parameter stress $P_{xy}$ is then dominated by the background viscosity term $\eta\dot\gamma$.
 
-Because steady states require to account for large strains, the computational costs of a similar analysis of nonlinear rheology in three-dimensions are still prohibitive.
-In the following we compare 3D-systems with 2D-systems smaller than those of the previous paragraph but large enough to show qualitatively similar structure under flow.
+{In neither system does $\sigma_{xy}(\dot\gamma)$ offer any clear signature of a yield stress. There may be two ways to understand this. Either, as detailed
+previously, this is because at the lowest shear rates studied here the selected
+state is a fully aligned one with the layering wavevector perpendicular to
+the flow: such a state should have a purely viscous response.
+However, a defect texture may also lead to a viscous response in the linear regime if flow occurs via permeation~\cite{deGennes} (i.e.
+by fluid flow not altering appreciably the order parameter pattern).
 
+To further characterise the small shear rate regimes in our simulations, we have performed additional simulations on fully aligned states, with different layer spacings.
+While the shear stress contribution in this case is negligible as $\dot{\gamma}\to 0$, $N_1$ does not vanish for vanishing shear rate, but depends sensitively on the nonequilibrium layer spacing in the range of shear rates addressed here.
+This can be described as a ``quasi-Newtonian'' behaviour.
+As expected, $N_1$ is positive for overcompressed layers, and negative for overdilated layers.
+The values of $N_1$ obtained in system B-2D (see Fig.~9) are comparable to those observed in a quiescent lamellar smectic system prepared by compressing the layers by about 5-10\%.}
+
+\subsection{Three Dimensions}
+\label{results3D}
+
+The computational cost of repeating all the above studies in three dimensional systems of equivalent size remains prohibitive, particularly when addressing steady state properties which emerge only after a million or more LB timesteps.
+We therefore focus in 3D on smaller systems (Table \ref{tab1}), but still large enough in principle to show qualitatively similar structure under flow.
+Our studies use two sets of parameters.
+System A-3D is identical to A-2D except for the introduction of the third dimension.
+System C-3D has parameters intermediate between A-2D and B-2D (Table \ref{tab1}) and was chosen in preference to the latter because B-2D ordered completely if the system size was matched to a slice of B-3D.
+(We see below however that in fact this is a rather general outcome in 3D.)
+
+{
 \begin{figure*}[htp!]
@@
 \includegraphics[angle=0,width=0.45\textwidth]{phi_run1003_960k.png}
-\caption{Order parameter $\phi$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\cdot10^{-4}$) at $t=6.4\e{5}$.
+\caption{Order parameter $\phi({\bf r})$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\times10^{-4}$) at $t=6.4\e{5}$ LBU.
 The opacity of one phase has been reduced to enable a better view into the lamellar structure.
-The bottom left picture shows System C-2D at the same time step and shear rate.
-The bottom right picture is system A-2D at $\dot{\gamma}=1\cdot10^{-4}$ and time step $t=9.6\e{5}$.
-Results for A-3D are morphologically identical to those above.}
-\label{fig12}
+The bottom left picture shows system C-2D at the same time step and shear rate.
+The bottom right picture is system A-2D at $\dot{\gamma}=1\times10^{-4}$ and time step $t=9.6\e{5}$ LBU.
+Results for A-3D are morphologically indistinguishable from those for C-3D.
+In all these images the flow direction is horizontal, to the right in the upper half of the sample and to the left in the lower half.}
+\label{fig10}
 \end{figure*}
-
-The parameters sets are those of system A - we refer to A-3D for the three-dimensional run - and system C, which has a slightly larger segregation than system B.
-The system size is $N_x\times N_y \times N_z=128\times256\times128$ lattice sites and $128\times256\times1$, respectively with the directions of flow along the long dimension and the velocity gradient along x in the 3D-case.
-Two snapshots of system C-2D and A-2D in shear flow are shown in the bottom pictures in Fig. \ref{fig12}.
-They feature the typical fluctuating pattern of breaking and merging lamellae that was described in the previous paragraph and which is particularly pronounced in the case of system C-2D.
-The top picture visualise results for system C-3D with the same free energy parameters as C-2D.
-The quiescent state of system C-3D after equilibration is shown at the left with a morphology virtually identical to that of A-3D.
-The lamellar width of roughly 10 lattice sites is more or less the same as in two dimensions.
-A slice through the structure factor along $k_y=0$ (cuts along $k_x,k_z=0$ look identical) is shown in Fig. \ref{fig13} and resembles that in Fig. \ref{fig4}.
-However, in shear flow considerable differences arise between the two- and three-dimensional case.
-Contrary to the result in 2D, the three dimensional systems order readily into a regular lamellar structure after a transient period as shown in Fig. \ref{fig12} for system C-3D. 
-We opted for this system as the differences between the 3D- and 2D-system are the most striking we observed.
-Interestingly, these findings are consistent for even the highest shear rates, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
+}
 
@@
 \centering
-\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run786_320.jpg}
-\caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_y=0$ after equilibration at $t=3.2\e{5}$.
-Shown are wave vectors on the interval $k_x, k_z\in[-\pi/2 l,\pi/2 l]$.}
-\label{fig13}
+\includegraphics[angle=0,width=0.40\textwidth]{ck_y-slice_run786_320.jpg}
+\caption{Structure factor $C({\mathbf k})$ of quiescent system A-3D, showing a cut along $k_x=0$ after equilibration at $t=3.2\e{5}$.
+Shown are wave vectors on the interval $k_y, k_z\in[-\pi/2 ,\pi/2 ]$ with $k_y$ and $k_z$ being the horizontal and vertical direction, respectively.
+Data for C-3D (not shown) is virtually the same.}
+\label{fig11}
 \end{figure}
 
-\begin{figure*}[htp!]
+\begin{figure}[htp!]
 \centering
-\includegraphics[angle=0,width=0.33\textwidth]{ck_x-slice_run788_960.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_y-slice_run788_960.jpg}
-\includegraphics[angle=0,width=0.33\textwidth]{ck_z-slice_run788_960.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\cdot10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=0$ and $k_z=3\pi/16 l$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 l,\pi/2 l]$.
-Note that in these images y is the flow direction, whereas the direction of the velocity gradient and vorticity are z and x, respectively.}
-\label{fig14}
-\end{figure*}
+\includegraphics[angle=0,width=0.35\textwidth]{ck_x-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.35\textwidth]{ck_y-slice_run788_960.jpg}
+\includegraphics[angle=0,width=0.35\textwidth]{ck_z-slice_run788_960.jpg}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\times10^{-4}$) at t=960k.
+Results are shown in the planes at $k_x=0$ (top panel), $k_z=0$ (middle panel), and $k_y=3\pi/16$ (middle panel).
+The horizontal and vertical directions are respectively along $k_z$ and $k_y$ (top panel), $k_x$ and $k_y$ (middle panel), $k_x$ and $k_z$ (bottom panel).
+The wave-vectors in the horizontal and vertical directions are in $[-\pi/2 ,\pi/2]$ for all panels.
+The structure factor is akin to that expected of a regular lamellar phase.}
+\label{fig12}
+\end{figure}
 
-Nevertheless, the order is not completely perfect and edge dislocations emerge spontaneously in pairs as a consequence of a mismatch between some of the layers.
+Fig.~\ref{fig10} shows the post-equilibration amorphous structure and the steady-state structure under flow in system C-3D (A-3D is similar) at system size $N_x\times N_y \times N_z=256\times128\times128$.
+The latter is compared with C-2D and A-2D in a system of the same $x$ and $y$ dimensions ($256\times128\times1$).
+The quiescent metastable state of system C-3D after equilibration is more or less the same as it would be in two dimensions.
+Indeed, a slice through the structure factor along $k_x=0$ (cuts along $k_y,k_z=0$ look identical) is shown in Fig. \ref{fig11} and resembles that in Fig.
+~\ref{fig5}.
+However, under shear at $\dot\gamma = 2\times 10^{-4}$ ($Pe = 400$, $Ca = 0.252$) this structure anneals readily into a regular lamellar structure after a transient period of about $6\e{5}$ LB timesteps.
+This occurs for even the highest shear rates applied, indicating that there is a very strong tendency in three dimensions to form ordered lamellar arrays.
+The same behaviour was seen in system A-3D for which the highest flow rate used corresponds to $Pe = 80$ and $Ca = 0.448$.
+(In contrast, the 2D results at the same flow rates and system sizes feature the typical fluctuating pattern of breaking and merging lamellae that was described previously and is visible in the lower panels of Fig.~\ref{fig10}.) The structure factor seen in 3D (see Fig.~\ref{fig12}) is accordingly close to that of a perfectly ordered lamellar state, in contrast to the corresponding results in 2D (Fig.~\ref{fig5}).
+These findings suggest that any characteristic strain rate $\dot\gamma_c$ at which defects proliferate must be markedly higher in 3D than in 2D.
+Specifically, if $Ca$ is the relevant control parameter, our results for A-3D imply that $Ca_c > 0.448 $ in three dimensions.
+
+Note however that the order we observe is not completely perfect as a mismatch of the layers occurs leading to branching points and edges (e.g.
+at the centre of the ordered 3D-system in Fig.~\ref{fig10}).
 The Burgers vector is oriented along the gradient direction with edges along the flow direction.
-This forms a structure that is translational invariant along the flow direction and in the steady state the location of the edge defects does not change and the defects to not anneal.
-A closer look at the kinetic evolution of the order parameter reveals that regions separated by different values of order parameter connect along a third direction by expelling regions with opposite order parameter.
-This is a direct consequence of the additional dimension and seems to play a key role during the formation of homogeneously layered structure.\\
-It is instructive to quantify the time it takes system A-3D to emerge as lamellar ordered structure starting from an isotropic initial configuration.  
-The inset of Fig. \ref{fig6} gives the defect density as defined by Eq. \ref{defect density} versus time for shear rates $\dot{gamma}=5\cdot10^{-5}-2\cdot10^{-4}$. 
-Within a few $10^5$ timesteps the defect density decreases to magnitudes, which we obtained in case of system A-2D for shear rates about one decade lower.
-The total strain where the curves level off and ordering is achieved scales with the shear rate and is about $\gamma=1.3$ ($\dot{gamma}=2\cdot10^{-4}$), $\gamma=0.8$ ($\dot{gamma}=1\cdot10^{-4}$) and $\gamma=0.6$ ($\dot{gamma}=5\cdot10^{-5}$).
- \\
-A lower number of defects yields smaller chemical pressure, which is shown for the deviatortic part $P_{xy}$ in the inset of Fig. \ref{fig8}.
-This contribution to the total stress is dwarfed by the hydrodynamic part and system A-3D exhibits shear thinning almost to an ultimate degree.
-A relation between the magnitude of the deviatoric part of the chemical pressure and the defect density like that in Fig. \ref{fig9} in the two-dimensional case cannot be deduced.\\
-First normal stress differences $N_1=\langle\sigma_{yy}-\sigma{zz}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig10}.
-For all shear rates $\gd5\e{-5}-2\e{-4}$ system A-3D takes similar values, which are monotonously approached and about one decade smaller than those of system A-2D.\\ 
+This forms a structure that is translationally invariant along the flow direction and in the final steady state the location of the edge defects does not change.
+Notably, the corresponding orientation of edge dislocations does not exist in two dimensions.
+
+{A closer look at the kinetic evolution of the order parameter reveals that the additional dimension appears to play a key role in promoting the formation of the homogeneously layered structure.
+The ordering process takes place in two consecutive steps.
+Firstly, the equilibrated state undergoes a transition to a semi-ordered state, which consists of small patches of layered regions, with the layer normals along gradient direction.
+The length of the lamellae is initially quite short, but gets longer along the flow direction as the simulation proceeds and the structure is advected.
+It remains, however, relatively short along the vorticity direction, with the individual layers being disconnected.
+In a second step the layers merge by linking up along the vorticity direction.
+This is possible as the layers can be locally translated along the flow gradient direction, so that the average order parameter within a plane can change and is not conserved as in the two-dimensional case.
+}
+
+We next quantify the time system A-3D takes to anneal into its final ordered structure starting from an isotropic initial configuration.
+(These and subsequent measurements are reported for A-3D only; C-3D is qualitatively similar.)
+The inset of Fig.~\ref{fig6} gives the defect density as defined by Eq.~\ref{defect density} versus time for shear rates $\dot{\gamma}=5\times10^{-5}-2\times10^{-4}$. 
+Within a few hundred thousand LB timesteps the defect density decreases by two orders of magnitude, resembling the behaviour seen in system A-2D at shear rates one decade smaller.
+The total strain where the curves level off, and ordering is achieved is about $\gamma=160$ (at $\dot{\gamma}=2\times10^{-4}$), $\gamma=120$ (at $\dot{\gamma}=1\times10^{-4}$) and $\gamma=65$ (at $\dot{\gamma}=5\times10^{-5}$).
+
+The reduced number of defects is reflected in the magnitude of the order parameter stress contribution $P_{xy}$; the 3D behaviour is shown in the inset of Fig.~\ref{fig8}.
+The final contribution to the total stress is created by a small number of remaining defects (hence the inconsistent trend of the tiny plateau value with strain rate, which is not a concern here) and is quite negligible compared to the background viscosity contribution $\eta\dot\gamma$.
+The drop in $P_{xy}$ by three orders of magnitude during the defect annihilation process can formally be viewed as shear thinning of a quite extreme degree -- albeit one masked, in any macroscopic measurement of $\sigma_{xy}$, by the Newtonian background.
+
+First normal stress differences $N_1=\langle\sigma_{xx}-\sigma_{yy}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig8}.
+For all shear rates addressed here system A-3D attains similar values of the normal stress; these are approached monotonically and about one decade smaller than those seen in system A-2D.
+The mechanism reported previously for the buildup of normal stress on approaching the ordered state in 2D systems seemingly does not apply in 3D, perhaps because of the much more efficient pathways to relaxation and mutual annihilation of defects.
+
+\if{ 
 Finally we want to glimpse at the structure factor of system A-3D in shear flow.
-Fig. \ref{fig14} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
+Fig. \ref{fig12} shows $C({\bm k})$ at shear rate $\gd=1\e{-4}$ for Fourier vectors on the interval $[-\pi/2 l, \pi/ 2 l]$.
 The leftmost picture is a cut along $k_x=0$ in with $k_y$ in horizontal and $k_z$ in vertical direction.
-It shows data in flow-velocity-gradient plane and can be directly compared with the left picture in Fig. \ref{fig4}.
+It shows data in flow-velocity-gradient plane and can be directly compared with the left picture in Fig. \ref{fig5}.
 The other images give cuts along $k_y=0$ in $k_x-k_z$-plane and along $k_z$ in $k_x-k_y$-plane. 
@@
 There is a certain similarity between the 2D-structure factor {\it after} switch-off (Fig.
-\ref{fig4}) and the 3D-structure, but the latter shows a pronounced peak at $k_z\simeq\pm0.2\pi/l$ also during shear flow, which is slightly concealed by the logarithmic scale.The split nature is an effect of the relatively sparse discretization and was caused by the visualisation software.
+\ref{fig5}) and the 3D-structure, but the latter shows a pronounced peak at $k_z\simeq\pm0.2\pi/l$ also during shear flow, which is slightly concealed by the logarithmic scale.The split nature is an effect of the relatively sparse discretization and was caused by the visualisation software.
 Although strongly peaked, the spatial dependence of the signal is best demonstrated in a 3D-representation. 
-Fig. \ref{fig15} depicts a close up of the 3D-structure factor, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
+Fig. \ref{fig13} depicts a close up of the 3D-structure factor, whereas the red area corresponds to an isosurface of about $C({\bm k})\simeq 10^3$.
 It reveals a banana-shaped spatial dependence with some small appendices in flow direction at $k_x=0$.
@@
 \includegraphics[angle=0,width=0.45\textwidth]{ck_run788_960_zoom.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\cdot10^{-4}$.}
-\label{fig15}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow: The picture gives a 3D-representation of the data in shear flow at $t=9.6\e{5}$ and $\dot{\gamma}=1\times10^{-4}$.}
+\label{fig13}
 \end{figure}
 
-\section{Discussion and Conclusions}
+}\fi
 
-We have performed large scale simulations of a lamellar system in steady shear flow  in two and three dimensions.
-We were able to relate the morphology of the order parameter in the quiescent state to results from previous studies of the same system using similar parameters.\\
-In two dimensions and steady state simple shear flow at Peclet numbers ${\cal O}(10)$ we did not observe a homogenous lamellar structure for shear rates above a critical threshold, which depends on the exact parametrization of the free energy functional.
-The morphology was rather inhomogenous instead, exhibiting regions that are under pressure with lamellar spacings smaller than the equilibrium value and others that were stretched.
-We could relate the structure to typical values of rheological quantities.
-Areas under extensional flow coincide with locally negative or small positive shear stresses and first normal stress differences, whereas both go to positive peak values in the compressed domains.
-Below a critical shear rate the systems showed no tendency anymore to trip over itself and defects annealed as they could approach each other in adjacent lamellae.
-The characteristic time of this process was set by the shear rate and the total strain.
-The deviatoric part of the chemical pressure was correlated to the number of defect sites, whereas the latter was measured by means of an alignment factor, which quantified the local curvature of the lamellar interfaces.
-For the stiffer system A-2D we could verify a linear relation between the defect density and the average chemical contribution to the shear stress.\\ 
-In system A-2D first normal stress differences decreased as long as a dynamic equilibrium between defect creation an annihilation existed.
-Below the critical threshold this was no longer the case and pronounced defect annealing caused a rise in the first normal stress differences as the average lamellar spacing increased.
-We could establish a shear thinning regime for the more defect laden system A-2D.
-System B in contrary was so perfectly annealed in the steady state that virtually shear stresses were down at the hydrodynamic level and the relation between shear stress and shear rate was trivially linear.\\
-In three dimensions we simulated shear strains which have not been attained so far.
-Interestingly for the same parameters we observe striking differences between system A-2D in 2D and system A-3D in 3D.
-It seems that the dynamics of the defects in two and three dimensions is rather incommensurable.
-Even at the lowest applied shear rates the chemical stress has decayed to a level where the apparent viscosity was virtually identical to the solvent.
-Structure factor data confirmed the very regular layered ordering of the structure and a lamellar spacing close to the equilibrium value.
+\section{Conclusions}\label{conclusions}
 
-A comparison of these results with experimental work can be only limited because the topology is different in two and three dimensions.
-Our 2D-systems correspond to are translational invariant along the third dimension.
-Every branching point or endpoint of a lamella has the topology of an edge defect in the 3D-continuation of the structure.
-A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}
-Interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, floding  and formation of defects. 
-For the power law dependence of the shear stress on the shear rate Lu et al.
-find an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
-Our value of $m=0.87$ appears to be slightly too large.
-A possible explanation for this is that screw defects, which do not exist in our 2D-picture, play a key role for this theory.\\
+Many earlier studies have proposed a picture in which a layer undulation instability in three-dimensional smectic liquid crystals emerges above a critical shear rate \cite{Zilman99,Courbin02,Gonnella98,Kumaran2001,Ramaswamy92}.
+Molecular dynamics simulations on lyotropic smectics place this in a regime of flow rates well above those we have applied in the present study \cite{Guo2002,Soddemann2004}.
+The critical shear rate for thermotropic smectics that has been given in a theoretical approach \cite{Stewart2009} is slightly above ours.
+The status of this instability and its effects on rheology in experiments remains somewhat unclear.
+Also unclear is its status in two dimensions.
+
+In the present work we have studied the nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model.
+After quenching from a homogeneous phase a metastable amorphous lamellar structure appears, similar to previous simulations in two dimensions using comparable models and parameters~\cite{Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}.
+Also in accord with some previous studies, under flow in two dimensions we found the morphology of the order parameter to show two distinct regimes: one with very few defects which arises at low shear rates, and a defect rich phase at higher ones.
+We found that the critical (or crossover) strain rate $\dot\gamma_c$ separating these phases depends somewhat on system size, but much more strongly on thermodynamic and kinetic parameters.
+We have argued that this dependence might be summarized in terms of a capillary number $Ca$ (Eq.~\ref{capillary}) that is the ratio of viscous stress to the gradient free energy density of the lamellar phase.
+This controls a steady-state elastic strain in the smectic which, when it exceeds some characteristic value $Ca_c$, might cause yielding of layers and defect proliferation; our estimate is $0.056<Ca_c<0.112$ (system A) and $0.082<Ca_c<0.164$ (system B).
+However we emphasize that the evidence for $Ca$ as a control parameter is currently very limited and it would take a painstaking exploration of parameter space to confirm this picture.
+
+A notable feature of the disordered flow phase is our observation of striated heterogeneity in the state of smectic order, with strongly dilated and compressed regions oriented along the extensional axis of the flow.
+These are present in steady state but are nonstationary.
+The extended regions with lamellar spacing larger than at equilibrium exhibit either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
+(A similar dependence of the layer alignment in flow direction depending on the system size has been recently reported in \cite{Kumaran2011}.)
+Smaller systems with 16 layers across the gradient direction, which is comparable with our systems consisting of  $N_x\times N_y \times N_z=256\times128\times1$ sites, showed a larger tendency to align in flow direction.
+For larger systems there was a clear trend to align at an angle with the flow direction.
+These distortions of the order parameter contribute a term in $P_{xy}$ to the shear stress whose magnitude could be directly related to the density of defects.
+In one system we found shear thinning described by a power law dependence of stress on strain rate, $\sigma_{xy}=\gamma^m$ with $m=0.87$; however the simplicity of this result may be fortuitous since the range of strain rates studied spans $\dot\gamma_c$.
+In another system, at strain rates primarily below $\dot\gamma_c$, the shear rheology was quasi-Newtonian ($m=1$) although a significant first normal stress difference $N_1$ was also present.
+
+We then simulated the flow behaviour of directly comparable systems in three dimensions, and found striking differences between the two- and three-dimensional response under conditions of matched $\dot\gamma$.
+The 3D-system showed none of the distictive dilation/compression bands seen in 2D; moreover the microstructure, initially amorphous after quenching from the homogeneous phase, rapidly ordered under shear into a lamellar array with a low density of edge defects, all of which were oriented along the flow directions.
+This contrasts with the 2D case where numerous edge dislocations are present in the disordered phase ($\dot\gamma > \dot\gamma_c$) but, by virtue of the 2D geometry, all are oriented along the vorticity direction.
+(In saying this, we imagine a 3D continuation of the 2D structure which is translationally invariant along that direction.) Our 3D results do not rule out a transition or crossover from ordered to disordered structures, but this is not seen within the range of flow rates addressed here.
+This means that, if such a transition is controlled by capillary number, its value at onset in 3D obeys $Ca_c>0.448$, a value notably higher than that observed in 2D.
+Since identical parameter sets were chosen, the same applies to any other dimensionless number that might control the onset of the defect-rich phase.
+
+Finally, we note that one recent theory of smectic rheology \cite{Lu08} suggests this is controlled by screw dislocations which have no 2D analogue.
+Moreover, the residual dislocations seen in the 3D ordered phase have an orientation that also does not exist in 2D.
+It should be remembered that, in 3D, edge dislocations and screw dislocations are interconvertible: they have the same topological status and the nomenclature merely describes whether the defect line is oriented normal or parallel to the smectic layers \cite{ChaikinLubensky}.
+The ability of 3D defect lines to reorient as well as translate underscores their greater dynamical freedom compared to similar defects in a 2D geometry.
+It is thus tempting to attribute the marked difference in dynamics between 2D and 3D systems to the reduced constraints on defect motion that arise in 3D, allowing them to find and annihilate each other more easily.
 
 \section{Acknowledgements}
-We acknowledge support by EPSRC Grants No. EP/E045316 and No. EP/E030173 and computing time at the Argonne Leadership Computing Facility, Argonne, U.S.A. as well as on HECToR, U.K.
-We like to thank A. Morozov for useful discussions. 
+We acknowledge support by EPSRC Grants No.
+EP/E045316 and No.
+EP/E030173, the MAPPER EU-FP7 project (grant no.
+RI-261507) and computing time at the Argonne Leadership Computing Facility, Argonne, U.S.A.
+as well as on HECToR, U.K.
+We thank A. Morozov for useful discussions. 
 M.E.C. holds a Royal Society Research Professorship.
@@
 to advance the order parameter in time via the Cahn-Hilliard equation,
-and the later is used to solve the Navier-Stokes equations.
+and the latter is used to solve the Navier-Stokes equations.
 Coupling between the two is
@@
 \end{document}
+
+\if{
+In order to get a length scale we compare the average distance of two lamellae in Fig.\ref{fig10} with characteristic values of diblock copolymers in the lamellar phase or smectic liquid crystals.
+Typical values for the lamellar width are roughly 50nm, which make about 10 LBU in our simulation.
+Therefore the unit length $l$ in LBU corresponds to about 5nm in physical space or $1 {\rm m}=2\times10^8$ LBU.
+
+The energy scale is slightly more difficult to calibrate.
+We use the fact that both diblock copolymer melts and smectic liquid crystals belong to the Brazovskii class.
+This has been already pointed out by Leibler \cite{Leibler80} in his seminal work on microphase separation in block copolymers.
+The Landau free energy functional we are using can be transformed into a Brazovskii-type of form and coefficients can be compared \cite{Xu05, Fredrickson89}.
+The energy density scale in our functional \ref{FE-functional} is set by the coefficients $-a=b$ of the quadratic and quartic term, respectively.
+Referring to \cite{Fredrickson89} we can derive an expression that relates $b$ to Leibler's original parameters.
+This formula reads 
+\begin{equation}
+b=\frac{u_0 k_B T}{6}=\frac{c^4\lambda k_B T}{6} \rho_c,
+\end{equation}
+whereas $\rho_c$ is the number density of chains per unit volume,
+often expressed as $\rho_c=\sqrt{N/6^3} R_g^{-3}$ with $N$ as roughly the number of Kuhn segments and $R_g$ as radius of gyration.
+Typical values are $N\simeq10^4$ and $R_g\simeq 100{\rm nm}$.
+With $c=1.1019$ and $\lambda=106.18$ taken from the same reference and assuming $k_B T\simeq 4\e{-21} {\rm J}$ at room temperature, we get an estimate for $b$ of roughly $5\e{2} {\rm J m}^{-3}$, which is $b=5\e{-4}$ in LBU.
+Hence the LB unit of energy density is about $10^{6} \rm{Pa}$ or $1\rm{Pa}=10^{-6}$ LBU.
+
+It is straightforward to determine the unit of mass. 
+A typical block copolymers or liquid crystal has a mass density the region of $10^3 {\rm kg m}^{-3}$, which is of the order ${\cal O}(1)$ in LBU.
+Hence $1 {\rm kg m}^{-3}=10^{-3}$ LBU.
+
+We can now combine the scales for energy density, length and mass density to get an estimate for the time scale.
+As ${\rm s}^2=\rm{kg}\, \rm{m}^{-3}\, \rm{Pa}^{-1}\rm{m}^2=(1\e{-3}) \times (1\e{6}) \times (4\e{16})\rm{LBU}\simeq4\e{19} \,{\rm LBU}$, we derive that $1 \rm{s}\simeq6.3\e{9}$ LBU and one LB timestep $\Delta t$ corresponds to roughly $0.15 \rm{ns}$.
+}\fi

### 2011-12-09 trunk/papers/smectics/smectics.tex
Ready to submit
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1495 40876441-3496-0410-a613-a2fca35b2ae2

@@
 While large aligned regions are present there are other regions of high defect density.
-The lower panel of Fig.~\ref{fig3} shows a later state with a total strain incremented by a further $6400\%$ lattice units (i.e. by $64$ times the system size). 
+The lower panel of Fig.~\ref{fig3} shows a later state with a total strain incremented by a further $6400\%$ lattice units (i.e. by a displacement $64$ times the system size). 
 Obviously many defects have annealed away and the smectic layers are now rather well aligned with the flow direction.
@@
 \caption{Top: average total shear stress $\sigma_{xy}$ versus time for system A-2D: The horizontal lines are steady state time averages.
-For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$, we find $\eta_{app}/\eta\equiv\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$; centre: average order parameter contribution to the pressure tensor $-P_{xy}$ vs time for system A-2D.
+For shear rate $\dot{\gamma}=1.25\e{-5}, 2.5\e{-5}, 5\e{-5}, 1\e{-4}, 2\e{-4}$, we find $\eta_{app}/\eta\equiv\sigma_{xy}/ \eta \dot{\gamma}=2.49, 2.25, 2.18, 1.93, 1.74$; centre: average order parameter contribution to the stress tensor $-P_{xy}$ vs time for system A-2D.
 The inset shows data for system A-3D and shear rates $\dot{\gamma}=2\times10^{-4}, 1\times10^{-4}$ and $5\times10^{-5}$ (left to right); bottom: average chemical contribution to the pressure tensor $-P_{xy}$ vs.
@@
 This causes frustration as the system is unable to snap back locally towards its equilibrium layer spacing.
-Hence the normal stress difference $N_1$ $\sigma_{xx}-\sigma_{yy}$ fails to decay to zero, and remains positive at large times, as appropriate for a system of metastable and (on average) compressed smectic layers.
+Hence the normal stress difference $N_1=\sigma_{xx}-\sigma_{yy}$ fails to decay to zero, and remains positive at large times, as appropriate for a system of metastable and (on average) compressed smectic layers.
 The time dependence of the normal stress difference is also anomalously slow, whereas the shear stress reaches their steady state values more quickly.

### 2011-12-12 trunk/papers/smectics/smectics.tex
Final corrections from Peter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1496 40876441-3496-0410-a613-a2fca35b2ae2

@@
 like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}.
-However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}.
+However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) plays a central role in this transition~\cite{Cates89}.
 In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
@@
 Despite a wealth of experimental results, and the importance of flow behaviour
-to many commercial applications, mechanistic understanding
+to many commercial applications, mechanistic understanding of
 the rheology of lamellar phases remains limited. This is largely
@@
 This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems.
-Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.
+Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that makes it difficult to disentangle bulk and boundary effects.
 
@@
 
-For the analysis of the relation between rheological quantities and the morphology of the system we need to to quantify the number of defects.
+For the analysis of the relation between rheological quantities and the morphology of the system we need to quantify the number of defects.
 For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average orientation of the order parameter gradients, as follows
@@
 By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as an aligned lamellar configuration.
-We found a threshold value of $\xi=0.8$ to be well suited for our requirements.
+We found a threshold value of $\xi=0.8$ to be well suited to our requirements.
 This protocol does not directly count the number of defects, since at each defect more than one site has an alignment function below the threshold.
@@
 
-The shear rates needed to achieve this range of $Pe$ falls dramatically when lyotropic smectics are considered.
+The shear rates needed to achieve this range of $Pe$ fall dramatically when lyotropic smectics are considered.
 For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm.
@@
 \caption{Structure factor $C({\bm k})$ of system A-2D: Shown are states after equilibration at $t=3.2\e{5}$ (top); in steady shear at $\dot{\gamma}=1\times10^{-4}$ and $t=5\e{5}$ (centre) and after switch-off at $t=1.28\e{6}$ (bottom).
-The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU with $k_x$ and $k_y$ along the horizontal and vertical coodinate direction, respectively.}
+The sections shown correspond to wave vectors on the interval $k_x, k_y \in [-\pi/2,\pi/2 ]$ LBU with $k_x$ and $k_y$ along the horizontal and vertical coordinate direction, respectively.}
 \label{fig5}

### 2011-12-19 trunk/papers/smectics/smectics.tex
Another clean-up
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1498 40876441-3496-0410-a613-a2fca35b2ae2
Minor updates to text
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1499 40876441-3496-0410-a613-a2fca35b2ae2

@@
 to shear thickening or shear thinning behaviour~\cite{Larson}.
-Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
+In addition, if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
 Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.
@@
 
-We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}.
+Below, we describe smectics via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}.
 This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations).
@@
 %Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organise in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
-
-%KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.
 
@@
 \begin{eqnarray}
-{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
+{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}.
+\nonumber\label{FE-functional}
 \end{eqnarray}
@@
 \begin{equation}\label{OPEOM}
-\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
+\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu.
 \end{equation}
@@
 \begin{equation}\label{NSE}
-\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta}. 
+\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta},
 \end{equation}
@@
 The total number of Lees-Edwards planes was varied according to the size of the simulation so that the separation of two planes was always 16 lattice sites.
+We relegate further comments on the implementation of the Lees-Edwards boundary
+conditions to an Appendix.
 
@@
 The transform itself was calculated using the FFTW-library \cite{FFTW}.
-In principle an ensemble- or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.
+In principle an ensemble-average or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.
 
@@
 \begin{equation}
-Pe^* =\frac{(a_c+b)a_c}{8b^2}\,\frac{L^2}{\eta M}\label{pestar}
+Pe^* =\frac{(a_c+b)a_c}{8b^2}\,\frac{L^2}{\eta M}.
+\label{pestar}
 \end{equation}
@@
 
-Below we first present detailed results for two parameter sets ($\beta = 4.2, 0.42$) in two dimensions which we call systems A-2D and B-2D.
+We first present detailed results for two parameter sets ($\beta = 4.2, 0.42$) in two dimensions which we call systems A-2D and B-2D.
 System lattice sizes and shear rates are reported in Table \ref{tab1}; all remaining parameters are the same in both systems as detailed above.
@@
 A third parameter set ($\beta = 0.83$, system C-2D) gave results broadly similar to B-2D.
-Note that we will later report simulations in three dimensions for systems A-3D and C-3D which share parameters with A-2D and C-2D (see Table \ref{tab1}).
+We will later report simulations in three dimensions for systems A-3D and C-3D which share parameters with A-2D and C-2D (see Table \ref{tab1}).
 
@@
 Within the Landau-Brazovskii free energy functional (Eq.~\ref{FE-functional}) a range of parameter variations can produce stable lamellar phases, as explored in previous studies \cite{Kendon01,Kumaran2001,Xu03, Xu06b,Kumaran2011}.
-The nonequilibrium morphology of the lamellar structure, and the phase ordering kinetics (starting from a slightly noisy, uniform initial state) do however depend on the choice made for these, and also for the viscosity.
+The nonequilibrium morphology of the lamellar structure, and the phase ordering kinetics (starting from a slightly noisy, uniform initial state) do however depend on the choice made for these parameters, and also for the viscosity.
 Formation and ripening of the lamellar structure is slower when $\beta$ is small, consistent with the reduced initial driving force for ordering in this case.
@@
 A prerequisite seems to be that the defects can smoothly approach each other and then annihilate.
-This is prevented by the buckling of layers above a certain shear rate, beyond which the system shows a tendency to trip over itself, which results in the creation of new defects.
+This is prevented by the buckling of layers above a certain shear rate; the undulations so formed ultimately pinch-off and break, resulting in the creation of new defects.
 The onset of defect proliferation arises at characteristic flow rates $\dot{\gamma}_c=1.25-2.5\e{-5}$LBU in system A-2D and around $\dot{\gamma}_c=1-2\e{-4}$LBU in B-2D.
@@
 The quiescent metastable state of system C-3D after equilibration is more or less the same as it would be in two dimensions.
-Indeed, a slice through the structure factor along $k_x=0$ (cuts along $k_y,k_z=0$ look identical) is shown in Fig. \ref{fig11} and resembles that in Fig.
-~\ref{fig5}.
+Indeed, a slice through the structure factor along $k_x=0$ (cuts along $k_y,k_z=0$ look identical) is shown in Fig. \ref{fig11} and resembles that in
+Fig.~\ref{fig5}.
 However, under shear at $\dot\gamma = 2\times 10^{-4}$ ($Pe = 400$, $Ca = 0.252$) this structure anneals readily into a regular lamellar structure after a transient period of about $6\e{5}$ LB timesteps.
@@
 
-First normal stress differences $N_1=\langle\sigma_{xx}-\sigma_{yy}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig8}.
+First normal stress differences $N_1=\langle\sigma_{xx}-\sigma_{yy}\rangle$ for system A-3D are shown in the inset of Fig.~\ref{fig8}.
 For all shear rates addressed here system A-3D attains similar values of the normal stress; these are approached monotonically and about one decade smaller than those seen in system A-2D.
@@
 Since identical parameter sets were chosen, the same applies to any other dimensionless number that might control the onset of the defect-rich phase.
-
 Finally, we note that one recent theory of smectic rheology \cite{Lu08} suggests this is controlled by screw dislocations which have no 2D analogue.

### 2011-12-19 papers/smectics/smectics.tex
Corrected label for eq. 1
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1500 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{bm}
-\usepackage{float}
-
 \pagestyle{fancy}
+\usepackage{color}
 
@@
 \setlength\bibsep{1pt}
+}\fi
 
@@
 Many soft materials, including the amphiphiles used in skin creams, fabric softeners, and some pharmaceutical products, form lamellar mesophases (also known as smectic liquid crystals) whose perfectly ordered state consists of a regular stack of flat layers.
-As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent [deGennes book].
+As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent~\cite{deGennes}.
 Such phases are of wide technological importance in devices
@@
 On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
-to shear thickening or shear thinning behaviour [Larson book].
-Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
+to shear thickening or shear thinning behaviour~\cite{Larson}.
+In addition, if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
 Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.
@@
 Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
-phase transitions from a homogeneous state into to ordered phases
+phase transitions from a homogeneous state into ordered phases
 like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}.
-However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}.
+However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) plays a central role in this transition~\cite{Cates89}.
 In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
@@
 This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems.
-Moreover, solid walls have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.
+Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that makes it difficult to disentangle bulk and boundary effects.
 
@@
 
-We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}.
+Below, we describe smectics via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}.
 This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations).
 The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$.
-This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid and interfaces between these layers.
+This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid, with interfaces between these layers.
 A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks.
-However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents the smectic state.
+However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents a thermotropic smectic state.
 
@@
 We use sliding periodic (Lees-Edwards) boundary conditions for
-both the convection-diffusion and Navier-Stokes equations.
+both the convection-diffusion and the Navier-Stokes equations.
 Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses.
@@
 %Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organise in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
-
-KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.
 
@@
 This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other.
-The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters; a comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
+The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters.
+A comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
 Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning.
 The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
-Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE], where the numerical value of $m$ was however different.
+Such a behaviour is qualitatively reminiscent of the one found experimentally in~\cite{Lu08}, where the numerical value of $m$ was however different.
 In this strongly non-Newtonian regime we also find that the rheology is history dependent.
@@
 Section \ref{parametermap} describes the mapping between our simulation parameters and those for physical systems of interest.
-We present results of the two-dimensional case in Section \ref{results2D}, discussing
+We present results for the two-dimensional case in Section \ref{results2D}, discussing
 the morphology as a function of flow rate. We also consider average and local shear stress densities and
@@
 \begin{eqnarray}
-{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
+{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f
+\label{FE-functional}
+\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}.
+\nonumber
 \end{eqnarray}
@@
 \begin{eqnarray}
-p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
+p_{\phi}&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
 & &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
@@
 %
-and finally another term consisting of the order-parameter contribution to the deviatoric stress
-\cite{Yang76,Evans79} [REFERENCE OK?]
+and finally another term consisting of the order-parameter contribution to the deviatoric pressure~\cite{Swift96,Yang76,Evans79} 
 %
@@
 The velocity field is then used in the advection-diffusion equation Eq.\ref{OPEOM}, which is solved by a finite-difference method at the same timestep.
-We assume periodic boundary conditions along the flow and vorticity direction and Lees-Edwards sliding periodic boundary conditions\cite{leesedwards,Wagner02} along the direction of the velocity gradient.
+We assume periodic boundary conditions along the flow and vorticity directions and Lees-Edwards sliding periodic boundary conditions\cite{leesedwards,Wagner02} along the direction of the velocity gradient.
 Following \cite{Wagner02}, we introduce
@@
 The transform itself was calculated using the FFTW-library \cite{FFTW}.
-In principle an ensemble- or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.
-
-For the analysis of the relation between rheological quantities and the morphology of the system we need to to quantify the number of defects.
-For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average projection of the normalised order parameter gradients onto the gradient at a specific site.
-More explicitly, 
+In principle an ensemble-average or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.
+
+For the analysis of the relation between rheological quantities and the morphology of the system we need to quantify the number of defects.
+For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average orientation of the order parameter gradients, as follows
 %
@@
 %
-Here $V'$ is a small volume that contains $\mathbf r$, which in practice can be chosen to contain $\mathbf r$ and its nearest neighbours only.
+Here $V'$ is a small volume that contains $\mathbf r$; in practice $V'$ can be chosen to contain $\mathbf r$ and its nearest neighbours only.
 
@@
 Wherever lamellae terminate at edge dislocations, and in regions where their orientation changes abruptly, $X$ departs from unity, taking smaller positive values.
-By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as as an aligned lamellar configuration.
-We found a threshold value of $\xi=0.8$ to be well suited for our requirements.
+By setting a threshold $X=\xi$ it is possible to detect regions where the order parameter structure deviates significantly from what one would recognize visually as an aligned lamellar configuration.
+We found a threshold value of $\xi=0.8$ to be well suited to our requirements.
 This protocol does not directly count the number of defects, since at each defect more than one site has an alignment function below the threshold.
@@
 These are chosen to give a lamellar wavelength of $L = 10$ LBU (that is, $q = \sqrt{|\kappa|/2c} = 2\pi/10$LBU) which offers suitable spatial resolution.
-Moreover the microphase separation point $a_c = \kappa^2/4c = 1.2\times 10^{-4}$LBU is then a fixed constant.
+Moreover the microphase separation point, set by $a_c = \kappa^2/4c = 1.2\times 10^{-4}$LBU, is then a fixed constant.
 
@@
 
-To calibrate the LB length unit we equate $L$ the lamellar wavelength with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm.
+To calibrate the LB length unit, we equate $L$ the lamellar wavelength, with that of a typical thermotropic smectic: $10$ LBU $= L = 4$nm.
 Hence the LB length unit is about $0.4$nm; and 1m$= 2.5\times 10^{9}$LBU.
@@
 As detailed previously, an estimate of the molecular diffusivity is $\tilde D = 2Mb$; we set $M = 0.25$LBU throughout this study, so that $\tilde D$ lies in the range $2.5\times 10^{-5} - 2.5\times 10^{-4}$LBU.
-Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{12}$LBU) gives $D\simeq 10^{-10}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.
+Taking a mid-range value of  $\tilde D = 10^{-4}$LBU and using the above conversion factors (1 m$= 2.5\times 10^9$LBU, 1s $= 6\times 10^{11}$LBU) gives $D\simeq 10^{-11}$m$^2$s$^{-1}$ which is a reasonable mesogenic value.
 
@@
 However, combining the above conversion factors for length,  energy density, and time shows the physical and simulation units of mass density to be related via 1 kg m$^{-3} \equiv 1$ Pa s$^2$ m$^{-2} \simeq 10^{-8}$LBU.
-Thus our simulations address a fluid of mass density $10^8$kg m$^{-3}$ which is $10^5$ times larger than the true value for a typical material.
+Thus our simulations address a fluid of mass density $10^6$kg m$^{-3}$ which is $10^3$ times larger than the true value for a typical material.
 The reason for this is that the LB algorithm uses fluid inertia to update the dynamics and the discretization should therefore involve replacing the real fluid with the densest one possible, while ensuring that the Reynolds number $Re \sim \rho\dot\gamma L^2/\eta$ (which for real smectics is exceedingly small) remains small enough for the simulation to faithfully describe the low $Re$ limit.
@@
 
-The shear rates needed to achieve this range of $Pe$ falls dramatically when lyotropic smectics are considered.
+The shear rates needed to achieve this range of $Pe$ fall dramatically when lyotropic smectics are considered.
 For lyotropics one has $L\sim 40$nm; the LB length unit is thus about 4nm.
-The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$) [REFERENCE LYOTROPIC REVIEW ?].
+The factor $\alpha$ in the free energy density is now of order unity (since the main contributions are entropic and for flexible membranes there is about one relevant degree of freedom per cube of side $L$)~\cite{Safran}.
 The LB unit of energy density is then about $6\times 10^{5}$Pa, i.e.
@@
 Accordingly the strain rates relevant to the $Pe$ range addressed here are of order $10^2$ s$^{-1}$ and well within experimental reach, but below that addressed in molecular dynamics simulations \cite{Guo2002,Soddemann2004}.
-Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still irrelevantly small.
+Note that for lyotropics we are effectively simulating a fluid of even higher mass density than for thermotropics; but since the simulated Reynolds numbers are unaffected by the change in parameter mapping, they are still sufficiently small~\cite{Cates04}.
 
@@
 \hline 
-703 & $1024\times512\times1$ & 0-3.2 & $50$ &0 &0 & 0 \\
-704 & $1024\times512\times1$ & 3.2-6.4& $50$ & 0.224 & 40 &  $10$\\
-705 & $1024\times512\times1$ & 6.4-9.6& $50$ &0 & 0 &  0 \\
-\hline
-767 & $512\times256\times1$ & 0-3.2 & $50$ &0 & 0 &  0\\
-775 & $512\times256\times1$ & 3.2-6.4& $50$ &0.448 & 80 &  $20$ \\
-769 & $512\times256\times1$ & 3.2-25.6& $50$ &0.224 & 40 &  $10$ \\
-771 & $512\times256\times1$ & 12.8-38.4& $50$ &0.112 & 20 &  $ 5$ \\
-772 & $512\times256\times1$ & 51.2-76.8& $50$ &0.056 & 10 &  $ 2.5$ \\
-774 & $512\times256\times1$ & 51.2-102.4& $50$ &0.028 & 5 &  $1.25$ \\
-\hline
-792 & $256\times128\times1$ & 3.2-16.0& $50$ &0.22 & 40 &  $10$ \\
-\hline
-786 & $256\times128\times128$ & 0-6.4 & $50$ &0 & 0 &  0\\
-787 & $256\times128\times128$ & 3.2-6.4 & $50$ &0.448 & 80 &  $20$\\
-788 & $256\times128\times128$ & 3.2-9.6 & $50$ &0.224 & 40 &  $10$\\
-789 & $256\times128\times128$ & 6.4-12.8 & $50$ &0.112 & 20 &  $5$\\
+1 & $1024\times512\times1$ & 0-3.2 & $50$ &0 &0 & 0 \\
+%703
+2 & $1024\times512\times1$ & 3.2-6.4& $50$ & 0.224 & 40 &  $10$\\
+%704
+3 & $1024\times512\times1$ & 6.4-9.6& $50$ &0 & 0 &  0 \\
+%705
+\hline
+4 & $512\times256\times1$ & 0-3.2 & $50$ &0 & 0 &  0\\
+%767
+5 & $512\times256\times1$ & 3.2-6.4& $50$ &0.448 & 80 &  $20$ \\
+%775
+6 & $512\times256\times1$ & 3.2-25.6& $50$ &0.224 & 40 &  $10$ \\
+%769
+7 & $512\times256\times1$ & 12.8-38.4& $50$ &0.112 & 20 &  $ 5$ \\
+%771
+8 & $512\times256\times1$ & 51.2-76.8& $50$ &0.056 & 10 &  $ 2.5$ \\
+%772
+9 & $512\times256\times1$ & 51.2-102.4& $50$ &0.028 & 5 &  $1.25$ \\
+%774
+\hline
+10 & $256\times128\times1$ & 3.2-16.0& $50$ &0.22 & 40 &  $10$ \\
+%792
+\hline
+11 & $256\times128\times128$ & 0-6.4 & $50$ &0 & 0 &  0\\
+%786
+12 & $256\times128\times128$ & 3.2-6.4 & $50$ &0.448 & 80 &  $20$\\
+%787
+13 & $256\times128\times128$ & 3.2-9.6 & $50$ &0.224 & 40 &  $10$\\
+%788
+14 & $256\times128\times128$ & 6.4-12.8 & $50$ &0.112 & 20 &  $5$\\
+%789
 \hline
@@
 \hline
-707 & $1024\times512\times1$ & 0-3.2 & $5$ &0 & 0 &  0 \\
-710 & $1024\times512\times1$ & 3.2-9.6& $5$ &0.082 & 400 &  $10$ \\
-765 & $1024\times512\times1$ & 9.6-12.8& $5$ &0 & 0 &  0 \\
-\hline
-776 & $512\times256\times1$ & 0-3.2 & $5$ & & 0 &  0 \\
-777 & $512\times256\times1$ & 3.2-9.6& $5$ &0.164 & 800 &  $20$ \\
-778 & $512\times256\times1$ & 3.2-25.6& $5$ &0.082 & 400 &  $10$ \\
-780 & $512\times256\times1$ & 12.8-38.4& $5$ &0.041 & 200 &  $5$ \\
-781 & $512\times256\times1$ & 25.6-64.0& $5$ &0.020 & 100 &  $2.5$ \\
-782 & $512\times256\times1$ & 51.2-76.8& $5$ &0.010 & 50 &  $1.25$ \\
+15 & $1024\times512\times1$ & 0-3.2 & $5$ &0 & 0 &  0 \\
+%707
+16 & $1024\times512\times1$ & 3.2-9.6& $5$ &0.082 & 400 &  $10$ \\
+%710
+17 & $1024\times512\times1$ & 9.6-12.8& $5$ &0 & 0 &  0 \\
+%765
+\hline
+18 & $512\times256\times1$ & 0-3.2 & $5$ & & 0 &  0 \\
+%776
+19 & $512\times256\times1$ & 3.2-9.6& $5$ &0.164 & 800 &  $20$ \\
+%777
+20 & $512\times256\times1$ & 3.2-25.6& $5$ &0.082 & 400 &  $10$ \\
+%778
+21 & $512\times256\times1$ & 12.8-38.4& $5$ &0.041 & 200 &  $5$ \\
+%780
+22 & $512\times256\times1$ & 25.6-64.0& $5$ &0.020 & 100 &  $2.5$ \\
+%781
+23 & $512\times256\times1$ & 51.2-76.8& $5$ &0.010 & 50 &  $1.25$ \\
+%782
 \hline
@@
 \hline
-998 & $256\times128\times1$ & 0-3.2& $10$ &0 & 0 & 0 \\
-1001 & $256\times128\times1$ & 3.2-9.6& $10$ &0.252 & 400 &  $20$ \\
-1000 & $256\times128\times1$ & 3.2-12.8& $10$ &0.126 & 200 &  $10$ \\
-\hline
-999 & $256\times128\times128$ & 0-3.2 & $10$ &0 & 0 &  0 \\
-1003 & $256\times128\times128$ & 3.2-9.6 & $10$ &0.252 & 400 &  $20$\\
-1002 & $256\times128\times128$ & 3.2-12.8 & $10$ &0.126 & 200 &  $10$\\
+24 & $256\times128\times1$ & 0-3.2& $10$ &0 & 0 & 0 \\
+%998
+25 & $256\times128\times1$ & 3.2-9.6& $10$ &0.252 & 400 &  $20$ \\
+%1001
+26 & $256\times128\times1$ & 3.2-12.8& $10$ &0.126 & 200 &  $10$ \\
+%1000
+\hline
+27 & $256\times128\times128$ & 0-3.2 & $10$ &0 & 0 &  0 \\
+%999
+28 & $256\times128\times128$ & 3.2-9.6 & $10$ &0.252 & 400 &  $20$\\
+%1003
+29 & $256\times128\times128$ & 3.2-12.8 & $10$ &0.126 & 200 &  $10$\\
+%1002
 \hline
@@
 
-We turn now to system A-2D ($\beta = 4.2$) The top panel in Fig. \ref{fig1} shows the order parameter pattern $\phi$ at the end of the equilibration phase prior to shearing.
+We turn now to system A-2D ($\beta = 4.2$). The top panel in Fig. \ref{fig1} shows the order parameter pattern $\phi$ at the end of the equilibration phase prior to shearing.
 Although the true equilibrium structure would show homogeneous alignment, this cannot be reached in a reasonable time scale unless shear is applied.
 The resulting metastable structure is amorphous and isotropic, but locally exhibits the same lamellar spacing $L = 10$ LBU as the fully ordered equilibrium state.
-If shear flow is imposed, the lamellae start to align in flow direction, which induces a morphological changes to the structure.
+If shear flow is imposed, the lamellae start to align in the flow direction, which induces a morphological changes to the structure.
 A typical snapshot in the resulting steady state at $Pe = 40$ $(Ca = 0.22)$ is shown in the middle panel of Fig.~\ref{fig1}.
 In this and all subsequent images, the applied flow velocity is horizontal and the velocity gradient vertical, with the flow directed to the right in the upper half and to the left in the lower half of the picture.
-After a short transient during which the order parameter remains relatively homogenous (not shown) there emerge extended regions in which the smectic layers are either expanded or compressed, so that $L$ locally deviates from the equilibrium value.
+After a short transient during which the order parameter pattern remains relatively homogenous (not shown) there emerge extended regions in which the smectic layers are either expanded or compressed, so that $L$ locally deviates from the equilibrium value.
 These mesoscopic inhomogeneities are inclined at an angle to the flow and remain continuously present in the steady state, but are not stationary: their local features are advected by the flow and evolve continuously by deformation.
@@
 We found that after the shear flow was switched off the lamellar spacing returned to its equilibrium value everywhere, with expanded regions contracting and vice versa.
-The final state, after all relaxation has virtually come to a standstill, is shown in the bottom panel of Fig.~\ref{fig1}.
+The final state, after all relaxation has (virtually) come to a standstill, is shown in the bottom panel of Fig.~\ref{fig1}.
 Most of the lamellae remain oriented along the flow direction, but depending on the local expansion and compression undergone during the previous flow, some regions retain strong misalignment which does not anneal away.
 The separation between defects as one moves along the lamellae varies considerably in space; both short and long separations can be seen.
-This applies both in the aligned layers and in layers that are misaligned with the flow.
+This applies both to the aligned layers and to layers that are misaligned with the flow.
 
@@
 All the generic results of Fig.~\ref{fig1} are the same, but the defect density, both during flow and after switchoff, is significantly reduced.
-(However the bands of misalignment in the flowing state are, if anything, even more apparent.) Consequently, the lamellar alignment along the flow direction after switch off is also more pronounced.
+(However the bands of misalignment in the flowing state are, if anything, even more apparent.) Consequently, the lamellar alignment along the flow direction after switch off is also more complete.
 
@@
 
-In principle the dependence of the state of order on flow rate could be smooth, or could entail a nonequilibrium phase transition -- for instance if the perfectly ordered state ceased to be stable above a certain shear rate \cite{Gonnella98,Zilman99,Kumaran2001} [REFERENCE RAMASWAMY?].
+In principle the dependence of the state of order on flow rate could be smooth, or could entail a nonequilibrium phase transition -- for instance if the perfectly ordered state ceased to be stable above a certain shear rate \cite{Gonnella98,Zilman99,Kumaran2001,Ramaswamy92}.
 In practice, however, starting as we do from an initially amorphous lamellar texture, a perfectly ordered state is generally not reached for large systems even if it might be stable once formed.
@@
 While large aligned regions are present there are other regions of high defect density.
-The lower panel of Fig.~\ref{fig3} shows a later state with a total strain incremented by a further $10^{5}\%$. 
+The lower panel of Fig.~\ref{fig3} shows a later state with a total strain incremented by a further $6400\%$ lattice units (i.e. by a displacement $64$ times the system size). 
 Obviously many defects have annealed away and the smectic layers are now rather well aligned with the flow direction.
@@
 Note that $Ca$, which comprises the ratio of a viscous stress to an elastic modulus, can also be viewed as a strain.
-In that context one can imagine that defects proliferate when the deformation of the smectic exceeds a yield strain value (for which a threshold of a few percent is reasonable).
+In that context one can imagine that defects proliferate when the steady-state elastic deformation of the smectic exceeds a yield strain value (for which a threshold of a few percent is reasonable).
 This makes sense if, at $Ca \simeq Ca_c$, the stress contributed from the order parameter is (a) comparable to the viscous stress from the background fluid -- essentially the force balance argument made above -- and (b) primarily elastic in character.
@@
 The runs with the two highest shear rates were started from the amorphous, post-equilibration configuration, whereas all others were pre-sheared using the final configuration of the preceding run as the initial state.
-The total strain between the beginning and end of every individual run was at least $10^{5}\%$.
+The total strain between the beginning and end of every individual run was at least $3\times 10^{3}\%$.
 Interestingly, following each step down in shear rate, both systems A-2D and B-2D continuously anneal to lower defect densities on a timescale $\simeq 10^6$LBU.
@@
 
-System A-2D shows clear shear thinning behaviour; values of its apparent viscosity at various shear rates are given in the caption.
+System A-2D shows clear shear thinning behaviour; values of its apparent viscosity at various shear rates are given in the caption of Fig.
+7.
 The corresponding range of $Ca$ is 0.028-0.448.
@@
 In contrast, system B-2D does not exhibit clear shear thinning; the apparent viscosities obey $\eta_{app}/\eta=1.57, 1.62, 1.55, 1.42, 1.87$ for the same set of shear rates as used in Fig.~\ref{fig7}.
-To within the measured accuracy, system B-2D is rheologically quasi-Newtonian.
+To within the measured accuracy, system B-2D is thus rheologically quasi-Newtonian.
 However, the range of $Ca$ in this case is lower, 0.01-0.164, and the order parameter stress creates only a relatively modest uplift to the baseline viscosity.
@@
 
-Fig.~\ref{fig8} shows averaged first normal stress differences $\langle N_1({\bf r})\rangle \equiv N_1$ versus time for system A-2D, where we followed the same stepwise decreasing flow protocol as before.
+Fig.~\ref{fig8} shows averaged first normal stress differences $\langle N_1({\bf r})\rangle \equiv N_1$ versus time for system A-2D, where we report the same stepwise decreasing flow protocol as before.
 For the first several steps $N_1$ decreases monotonously with decreasing shear rate.
@@
 When two defects meet and cancel, the local lamellar width alters by a certain amount.
-This causes frustration as the system is unable to snap back locally towards its equilibrium width.
-Hence the normal stresses $\sigma_{xx}$ falls, and may even attain negative values, leading to increasing values of $N_1$.
-This explanation is further supported by the fact that $\sigma_{xx}$ evolves anomalously slowly, whereas all other components of the stress tensor reach their steady state values more quickly.
-We have not explored the dependence on system size of this normal stress anomaly, but it may be partly a finite size effect caused by the relatively modest number of smectic layers within the simulated volume.
+This causes frustration as the system is unable to snap back locally towards its equilibrium layer spacing.
+Hence the normal stress difference $N_1=\sigma_{xx}-\sigma_{yy}$ fails to decay to zero, and remains positive at large times, as appropriate for a system of metastable and (on average) compressed smectic layers.
+The time dependence of the normal stress difference is also anomalously slow, whereas the shear stress reaches their steady state values more quickly.
+The relatively large steady state value of $N_1$ and this slow dynamics may partly be a finite size effect caused by the relatively modest number of smectic layers within the simulated volume.
 The normal stress in system B-2D (not shown) behaves differently: in this case $N_1$ falls monotonically during the defect annihilation phase. 
@@
 A recent experimental and theoretical work studies the nonlinear relation between shear stress and shear rate in lyotropic lamellar phases \cite{Lu08}; interestingly the authors identify the motion of screw defects as being responsible for the layer tilting, folding  and formation of defects.
-They report an experimental value of $m\simeq=1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
+They report an experimental value of $m\simeq 1/1.44\simeq 0.694$, which is in good agreement with their theoretical prediction of $m=2/3$.
 Our value of $m=0.87$ for the disordered flow regime is larger.
 One possible explanation for this discrepancy is that screw defects, which do not exist in our 2D simulations, play a key role for this theory.
-On the other hand the fitted power law must be treated with great caution since the fit includes data from both $\dot\gamma$ both above and below $\dot\gamma_c$.
+On the other hand our fitted power law must be treated with great caution since the fit includes data from $\dot\gamma$ both above and below $\dot\gamma_c$.
 It is somewhat surprising that the morphological changes around $1.25\e{-5}<\dot\gamma_c <2.5\e{-5}$ in system A give no visible anomaly or feature in the flow curves.
-This may be partly explained with a more crossover-like behaviour of system A at the lowest applied shear rate.
+%This may be partly explained with a more crossover-like behaviour of system A at the lowest applied shear rate.
 %
@@
 
-In neither system does $\sigma_{xy}(\dot\gamma)$ offer any clear signature of a yield stress, as would be expected to arise for a frozen-in defect texture in the static state.
-This can be understood if, as detailed previously, that texture is replaced by a fully aligned state at the lowest shear rates studied here: with the layering wavevector perpendicular to the flow, such a state should have a purely viscous response.
-On the other hand, one would expect $N_1$ also to vanish in that case; if this happens, it is at lower strain rates than encountered here.
-
-WE HAVE TO UNDERSTAND WHY THE NORMAL STRESS IS SO LARGE, ESPECIALLY IN B-2D WHERE YOU ARE CLAIMING THAT THE CHEMICAL STRESS IS NEGLIGIBLE!!
-
-OH: I have performed simulations with a perfectly aligned state with different layer spacing. The viscous stress vanishes completely in all cases.
-However, the first normal stress difference seems to depend sensitively on the layer spacing.
-
-For a layer spacing equal to the Brazovskii equilibrium wavelength  (corresponding to 26 lamellae across the gradient direction at $N_x \times N_y \times N_z = 512 \times 256 \times 1$) $N_1$ is very small and positive of order $10^-6$.
-Interestingly, neither $P^{chem}_{xy}$ nor $N_1$ alter when the flow is switched on, but remain virtually constant.
-
-If the wavelength of the initial state is changed by 5-10\%, the number of layers decreases or increases by 2.
-Again, the viscous stress is zero, but in the compressed / expanded state we get positive /negative $N_1=P^{chem}_{yy}-P^{chem}_{xx}$ (using the conventional definition of flow-minus-gradient and the fact that the chemical pressure has a negative sign).
-The magnitude of $N_1$ is higher for system B than for system A. This can be understood with difference in Brazvskii amplitude (A: $\phi_q\simeq1.28$, B: $\phi_q\simeq 2$).
-For system B $N_1$ is almost of order $10^{-4}$!
-
-Therfore, our findings for $N_1$ correspond to a slightly compressed system with a layer spacing smaller than the equilibrium one.
-The fact that $N_1$ is already very small if equilibrium spacing is used suggests that the plateau is caused by finite size effects.
-
-The reason for the sensitivity of $N_1$ towards compression and expansion is not really clear yet to me. 
-It appears that $P_xx$ (x being the flow direction) is much more sensitive towards changes in the layer spacing than $P_yy$.
-Intuitively I would have expected it to be the other way round as changes in the layer spacing affect the y-gradients more than the x-gradients (after all the structure is translationally invariant in x-direction).
-At the moment I can only think of an explanation that draws on something beyond the single-mode expansion where gradients in flow direction pick up wavelengths in the velocity-gradient-direction.
+{In neither system does $\sigma_{xy}(\dot\gamma)$ offer any clear signature of a yield stress. There may be two ways to understand this. Either, as detailed
+previously, this is because at the lowest shear rates studied here the selected
+state is a fully aligned one with the layering wavevector perpendicular to
+the flow: such a state should have a purely viscous response.
+However, a defect texture may also lead to a viscous response in the linear regime if flow occurs via permeation~\cite{deGennes} (i.e.
+by fluid flow not altering appreciably the order parameter pattern).
+
+To further characterise the small shear rate regimes in our simulations, we have performed additional simulations on fully aligned states, with different layer spacings.
+While the shear stress contribution in this case is negligible as $\dot{\gamma}\to 0$, $N_1$ does not vanish for vanishing shear rate, but depends sensitively on the nonequilibrium layer spacing in the range of shear rates addressed here.
+This can be described as a ``quasi-Newtonian'' behaviour.
+As expected, $N_1$ is positive for overcompressed layers, and negative for overdilated layers.
+The values of $N_1$ obtained in system B-2D (see Fig.~9) are comparable to those observed in a quiescent lamellar smectic system prepared by compressing the layers by about 5-10\%.}
 
@@
 Our studies use two sets of parameters.
-System A-3D is identicaly to A-2D except for the introduction of the third dimension.
+System A-3D is identical to A-2D except for the introduction of the third dimension.
 System C-3D has parameters intermediate between A-2D and B-2D (Table \ref{tab1}) and was chosen in preference to the latter because B-2D ordered completely if the system size was matched to a slice of B-3D.
@@
 
+{
 \begin{figure*}[htp!]
@@
 \includegraphics[angle=0,width=0.45\textwidth]{phi_run1003_960k.png}
-\caption{Order parameter $\phi({\bf r})$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\times10^{-4}$) at $t=6.4\e{5}$.
+\caption{Order parameter $\phi({\bf r})$ in two and three dimensions: Top pictures show system C-3D after equilibration (left) and in shear flow (right, $\dot{\gamma}=2\times10^{-4}$) at $t=6.4\e{5}$ LBU.
 The opacity of one phase has been reduced to enable a better view into the lamellar structure.
 The bottom left picture shows system C-2D at the same time step and shear rate.
-The bottom right picture is system A-2D at $\dot{\gamma}=1\times10^{-4}$ and time step $t=9.6\e{5}$.
+The bottom right picture is system A-2D at $\dot{\gamma}=1\times10^{-4}$ and time step $t=9.6\e{5}$ LBU.
 Results for A-3D are morphologically indistinguishable from those for C-3D.
@@
 \end{figure*}
+}
 
@@
 \includegraphics[angle=0,width=0.35\textwidth]{ck_z-slice_run788_960.jpg}
-\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\times10^{-4}$) at t=960k: Cuts through along $k_x=0,k_y=3\pi/16$ and $k_z=0$, respectively are shown for wave-vectors $k_x, k_y, k_z\in[-\pi/2 ,\pi/2]$.}
+\caption{Structure factor $C({\mathbf k})$ of system A-3D in shear flow ($\dot{\gamma}=1\times10^{-4}$) at t=960k.
+Results are shown in the planes at $k_x=0$ (top panel), $k_z=0$ (middle panel), and $k_y=3\pi/16$ (middle panel).
+The horizontal and vertical directions are respectively along $k_z$ and $k_y$ (top panel), $k_x$ and $k_y$ (middle panel), $k_x$ and $k_z$ (bottom panel).
+The wave-vectors in the horizontal and vertical directions are in $[-\pi/2 ,\pi/2]$ for all panels.
+The structure factor is akin to that expected of a regular lamellar phase.}
 \label{fig12}
@@
 The same behaviour was seen in system A-3D for which the highest flow rate used corresponds to $Pe = 80$ and $Ca = 0.448$.
-(In contrast, the 2D results at the same flow rates and system sizes feature the typical fluctuating pattern of breaking and merging lamellae that was described previously and is visible in the lower panels of Fig.~\ref{fig10}.) The structure factor seen in 3D is accordingly close to that of a perfectly ordered lamellar state (see Fig.~\ref{fig12}) in contrast to the corresponding results in 2D (Fig.~\ref{fig5}).
+(In contrast, the 2D results at the same flow rates and system sizes feature the typical fluctuating pattern of breaking and merging lamellae that was described previously and is visible in the lower panels of Fig.~\ref{fig10}.) The structure factor seen in 3D (see Fig.~\ref{fig12}) is accordingly close to that of a perfectly ordered lamellar state, in contrast to the corresponding results in 2D (Fig.~\ref{fig5}).
 These findings suggest that any characteristic strain rate $\dot\gamma_c$ at which defects proliferate must be markedly higher in 3D than in 2D.
@@
 
-[DESCRIPTION CLEAR?] Note however that the order we observe is not completely perfect as a mismatch of the layers occurs leading to branching points and edges (e.g.
+Note however that the order we observe is not completely perfect as a mismatch of the layers occurs leading to branching points and edges (e.g.
 at the centre of the ordered 3D-system in Fig.~\ref{fig10}).
@@
 Notably, the corresponding orientation of edge dislocations does not exist in two dimensions.
-[TRYING TO EXPLAIN WHAT'S HAPPENING IN 3D. CLEAR?]
-
-A closer look at the kinetic evolution of the order parameter reveals that the additional dimension seems to play a key role in promoting the formation of the homogeneously layered structure.
-The ordering process takes place in two consecutive steps: 
-Firstly, the equilibrated state undergoes a transition to a semi-ordered state, which consists of small patches of layered regions with the layer normals in gradient direction.
-Initially, the length of the lamellae is quite short but gets longer in flow direction as the simulation proceeds and the structure is advected.
-It remains, however, relatively short in vorticity direction with the individual layers being disconnected.
-In a second step the layers merge by linking up in vorticity direction and expelling regions with opposite order parameter, which is only possible due to the additional third dimension.
+
+{A closer look at the kinetic evolution of the order parameter reveals that the additional dimension appears to play a key role in promoting the formation of the homogeneously layered structure.
+The ordering process takes place in two consecutive steps.
+Firstly, the equilibrated state undergoes a transition to a semi-ordered state, which consists of small patches of layered regions, with the layer normals along gradient direction.
+The length of the lamellae is initially quite short, but gets longer along the flow direction as the simulation proceeds and the structure is advected.
+It remains, however, relatively short along the vorticity direction, with the individual layers being disconnected.
+In a second step the layers merge by linking up along the vorticity direction.
+This is possible as the layers can be locally translated along the flow gradient direction, so that the average order parameter within a plane can change and is not conserved as in the two-dimensional case.
+}
 
@@
 Within a few hundred thousand LB timesteps the defect density decreases by two orders of magnitude, resembling the behaviour seen in system A-2D at shear rates one decade smaller.
-The total strain where the curves level off, and ordering is achieved, scales with the shear rate and is about $\gamma=160$ (at $\dot{\gamma}=2\times10^{-4}$), $\gamma=120$ (at $\dot{\gamma}=1\times10^{-4}$) and $\gamma=65$ (at $\dot{\gamma}=5\times10^{-5}$).
+The total strain where the curves level off, and ordering is achieved is about $\gamma=160$ (at $\dot{\gamma}=2\times10^{-4}$), $\gamma=120$ (at $\dot{\gamma}=1\times10^{-4}$) and $\gamma=65$ (at $\dot{\gamma}=5\times10^{-5}$).
 
@@
 
-First normal stress differences $N_1=\langle\sigma_{xx}-\sigma{yy}\rangle$ for system A-3D are shown in the inset of Fig.\ref{fig8}.
+First normal stress differences $N_1=\langle\sigma_{xx}-\sigma_{yy}\rangle$ for system A-3D are shown in the inset of Fig.~\ref{fig8}.
 For all shear rates addressed here system A-3D attains similar values of the normal stress; these are approached monotonically and about one decade smaller than those seen in system A-2D.
 The mechanism reported previously for the buildup of normal stress on approaching the ordered state in 2D systems seemingly does not apply in 3D, perhaps because of the much more efficient pathways to relaxation and mutual annihilation of defects.
-
-CAN WE PLEASE MEASURE THE NORMAL STRESS IN A COMPLETELY DEFECT FREE STATE AND COMPARE WITH THE OBSERVED VALUES.
-DOES THE PLATEAU CORRESPOND TO A FEW REMAINING DEFECTS OR DOES IT STEM FROM THE PERFECT BEHAVIOUR??
 
@@
 
-Many earlier studies have proposed a picture in which a layer undulation instability in three-dimensional smectic liquid crystals emerges above a critical shear rate \cite{Zilman99,Courbin02,Gonnella98,Kumaran2001} [Ramaswamy].
+Many earlier studies have proposed a picture in which a layer undulation instability in three-dimensional smectic liquid crystals emerges above a critical shear rate \cite{Zilman99,Courbin02,Gonnella98,Kumaran2001,Ramaswamy92}.
 Molecular dynamics simulations on lyotropic smectics place this in a regime of flow rates well above those we have applied in the present study \cite{Guo2002,Soddemann2004}.
-The critical shear rate for thermotropic smectics that has been given in a theoretical approach \cite{Stewart2009} is slightly above ours [CHECKED].
+The critical shear rate for thermotropic smectics that has been given in a theoretical approach \cite{Stewart2009} is slightly above ours.
 The status of this instability and its effects on rheology in experiments remains somewhat unclear.
@@
 In the present work we have studied the nonlinear rheology of lamellar systems in two and three dimensions by means of large-scale simulations of a well-established model.
-After quenching from a homogeneous phase a metastable amorphous lamellar structure appears, similar to previous simulations in two dimensions using comparable models and parameters [REFERENCES].
-Also in accord with some previous studies [REFERENCES], under flow in two dimensions we found the morphology of the order parameter to show two distinct regimes: one with very few defects which arises at low shear rates, and a defect rich phase at higher ones.
+After quenching from a homogeneous phase a metastable amorphous lamellar structure appears, similar to previous simulations in two dimensions using comparable models and parameters~\cite{Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b}.
+Also in accord with some previous studies, under flow in two dimensions we found the morphology of the order parameter to show two distinct regimes: one with very few defects which arises at low shear rates, and a defect rich phase at higher ones.
 We found that the critical (or crossover) strain rate $\dot\gamma_c$ separating these phases depends somewhat on system size, but much more strongly on thermodynamic and kinetic parameters.
 We have argued that this dependence might be summarized in terms of a capillary number $Ca$ (Eq.~\ref{capillary}) that is the ratio of viscous stress to the gradient free energy density of the lamellar phase.
-This controls a steady-state elastic strain in the smectic which, when it exceeds some characteristic value $Ca_c$, might cause yielding of layers and defect proliferation; our best estimate is $Ca_c \simeq 0.03$ (system A) and $0.01$ (system B).
+This controls a steady-state elastic strain in the smectic which, when it exceeds some characteristic value $Ca_c$, might cause yielding of layers and defect proliferation; our estimate is $0.056<Ca_c<0.112$ (system A) and $0.082<Ca_c<0.164$ (system B).
 However we emphasize that the evidence for $Ca$ as a control parameter is currently very limited and it would take a painstaking exploration of parameter space to confirm this picture.
@@
 The extended regions with lamellar spacing larger than at equilibrium exhibit either negative or small positive shear stress and first normal stress difference, whereas positive peak values were found in the compressed domains.
-[EXPLANATION CLEAR?]
-A similar dependence of the layer alignment in flow direction depending on the system size has been recently reported in \cite{Kumaran2011}.
+(A similar dependence of the layer alignment in flow direction depending on the system size has been recently reported in \cite{Kumaran2011}.)
 Smaller systems with 16 layers across the gradient direction, which is comparable with our systems consisting of  $N_x\times N_y \times N_z=256\times128\times1$ sites, showed a larger tendency to align in flow direction.
 For larger systems there was a clear trend to align at an angle with the flow direction.
-These distortions of the order parameter contribute a term $P_{xy}$ to the shear stress whose magnitude could be directly related to the density of defects.
+These distortions of the order parameter contribute a term in $P_{xy}$ to the shear stress whose magnitude could be directly related to the density of defects.
 In one system we found shear thinning described by a power law dependence of stress on strain rate, $\sigma_{xy}=\gamma^m$ with $m=0.87$; however the simplicity of this result may be fortuitous since the range of strain rates studied spans $\dot\gamma_c$.
@@
 as well as on HECToR, U.K.
-We like to thank A. Morozov for useful discussions. 
+We thank A. Morozov for useful discussions. 
 M.E.C. holds a Royal Society Research Professorship.

### 2012-02-10 trunk/papers/smectics/smectics.tex
Revised version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1507 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-%This simulation strategy allows us to probe the bulk rheology of large systems at moderate to high shear rates, as quantified by an appropriate Peclet number, defined below.
-
-% KS: this can be moved to existing appendix
-%The LBM solves the Boltzmann equation on a regular lattice for a representative fluid that obey simplified dynamics in terms of a free propagation step of quasi-particles followed by a collision step, in which the particles interact.
-%The LBM solves the Navier-Stokes equation in the limit of small Mach numbers $Ma\ll1$ and has been largely applied to fluid flow in complex geometries, where it is superior to conventional methods of computational fluid dynamics.
-
-%Copolymer consist of type-A polymers covalently bonded to type-B polymers and can self-organise in a variety of phases, ranging from spherical phases over hexagonal cylindrical phases to lamellar phases or bicontinuous gyroid phases \cite{Bates90}.
-
 Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
@@
 This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other.
+
+The defects do not appear as a direct result of spontaneous thermal fluctuations.
+Instead they are a result of an undulation instability in our (deterministic) nonlinear dynamics.
+This occurs when viscous forces begin to dominate interfacial forces causing layers to pinch off and break, creating new defects in the process.
+The onset of a nontrivial topological pattern under deterministic evolution equations in a complex fluid is similar to the case of sheared binary fluids undergoing phase separation \cite{Stratford07}.
+
 The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters.
@@
 \begin{eqnarray}
-{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}.
-\nonumber\label{FE-functional}
+{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f
+\label{FE-functional}
+\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}.
+\nonumber
 \end{eqnarray}
@@
 \label{parametermap}
+
+The Landau-Brazovskii model belongs to the class of phenomenological models of liquid-crystalline materials.
+This means it features an isotropic-to-lamellar phase transition without explaining the underlying microscopic mechanism.
+Hence there is a many-to-one mapping between physical parameters and those of the phenomenological model.
+However, in the context of any particular system, it is possible to indicate physical parameters for which the chosen model is a good representation, and we do this below.
 
@@
 We will argue below that this could be a more relevant choice than $Pe$ for understanding the dependence of morphology on flow rate in smectic systems, at least when $Pe^*\gg 1$.
+\vfill
 
@@
 System lattice sizes and shear rates are reported in Table \ref{tab1}; all remaining parameters are the same in both systems as detailed above.
-%As expected from the preceding discussion of $Pe^*(\beta)$, the stress response at a given flow rate is softer in the second system. 
 A third parameter set ($\beta = 0.83$, system C-2D) gave results broadly similar to B-2D.
 We will later report simulations in three dimensions for systems A-3D and C-3D which share parameters with A-2D and C-2D (see Table \ref{tab1}).
+Note that our simulations in two dimensions are equivalent to assuming, in three dimensions, that the system maintains translational invariance in the vorticity direction.
+We find this is an increasingly good approximation at very low shear rates.
 
@@
 A prerequisite seems to be that the defects can smoothly approach each other and then annihilate.
-This is prevented by the buckling of layers above a certain shear rate; the undulations so formed ultimately pinch-off and break, resulting in the creation of new defects.
+This is prevented by the buckling of layers above a certain shear rate; the undulations so formed ultimately pinch off and break, resulting in the creation of new defects.
 The onset of defect proliferation arises at characteristic flow rates $\dot{\gamma}_c=1.25-2.5\e{-5}$LBU in system A-2D and around $\dot{\gamma}_c=1-2\e{-4}$LBU in B-2D.

### 2012-02-14 papers/lmcproc/lmcproc.tex
Revised submission
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1508 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{float}
+\usepackage{color}
 \newcommand{\ex}[1]{\times10^{#1}}
+\newcommand{\update}[1]{{\color{red} #1}}
 
@@
 
-\section{Discussions and Conclusions}
+\section{Conclusions}
 
@@
 
+We hope that our results will stimulate further experimental work on 
+the rheology of blue phases, similar to those reported e.g. in \cite{Negita:98}
+but targeted to micron-sized samples, which would provide a direct test
+of our simulations.
+
+We finally note that the anchoring we have chosen does not lead to a conflict
+with the bulk ordering of blue phases. In liquid crystal cells, it is 
+customary to treat the surface so as to favour normal or planar anchoring
+of the director field at the boundary. When used in very thin samples,
+these boundary conditions lead to the formation of new disclination 
+networks whose rheology it would be interesting to investigate~\cite{Fukuda:2010a,Fukuda:2010b,Tiribocchi:2011a}.
+
 % If you have acknowledgments, this puts in the proper section head.

### 2012-02-29 trunk/htdocs/user.tex
Some updates to tex docs
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1514 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \texttt{active} switches on the correction to BBL appropriate
-for active particles (expecting parameters $b_0$ and $b_1$ to be
+for active particles (expecting parameters $b_1$ and $b_2$ to be
 set); \texttt{subgrid} gives unresolved particles (radius $< 1$
@@
 at the moment. Again, it may be ASCII or binary.
+
+\subsubsection{The colloid state}
+
+The full data structure for the colloid state is defined in
+\texttt{src/colloid.h} and the corresponding routines which read
+and write the values are in \texttt{src/colloid.c}
+(note colloid singular!). Most of the state corresponds clearly
+with physical properties, although there are a number of 'structural'
+elements which have no physical meaning. These include the
+\texttt{rebuild} flag, which instructs the code to reconstruct the
+boundary links if necessary, and \texttt{deltaphi} which is involved
+in accounting for conserved order parameter in the \texttt{symmetric\_lb}
+free energy. Further, not all the elements are relevant is all
+circumstances: e.g., the magnetic dipole is only relevant is magnetic
+interactions are required, and squirmer parameters are only relevant
+to active particles.
+
+Note that it is possible in principle to have magnetic active particles,
+in which case the dipole direction (\texttt{s}) and the direction of
+motion vector (\texttt{m}) are allowed to be distinct.
 
@@
 
-\inputkey{colloid\_random\_s0} initial spin vector
+\inputkey{colloid\_random\_s0} initial spin vector (only relevant in magnetic
+field)
+
+The
 
@@
 
+The squirmer parameters specify the surface boundary conditions in
+an approach originating with Lighthill \cite{lighthill} and Blake
+\cite{blake}. Briefly, $b_1$ sets the propulsion speed
+($U = {\frac{2}{3}} b_1$), while $b_2$ sets the
+particle stresslet. The ratio $\beta = b_2/b_1$ determines the
+mode of the squirmer motion; see \cite{isaac} for further details.
+
 There is one special case. If you want exactly one particle, you may
@@
 \inputkey{colloid\_random\_r0} the initial position vector.
+
+This position must be within the current system, or no particle will
+be created.
 
@@
 are given in Table~X. This values are computed using a calibration
-procedure described by Nguyen and Ladd \cite{nguyen2001}.
+procedure described by Nguyen and Ladd \cite{nguyen-ladd2002}.
 
@@
 
+\subsubsection{Colloid I/O}
+
+Colloid I/O is controlled via a series of key value pairs. These are:
+
+\inputkey{colloid\_io\_freq}
+
+gives output at this frequency in time steps. Output is also generated
+at configuration steps (and measurement steps at the moment).
+
+\inputkey{colloid\_io\_grid} integer vector
+
+sets the colloid I/O grid. It must be accomodated by the current
+choice of \texttt{grid}.
+
+\inputkey{colloid\_io\_format\_input}
+
+One of \texttt{ASCII}, \texttt{ASCII\_SERIAL}, \texttt{BINARY}, or
+\texttt{BINARY\_SERIAL}
+
+\inputkey{colloid\_io\_format\_output}
+
+For output, only \texttt{ASCII} or \texttt{BINARY} are available.
+
 \subsubsection{I/O format}
@@
 Configuration I/O is always in binary. It is recommended to use
-binary output for speed and disk space considerations.
+binary output for speed and disk space considerations. For a local
+sub-domain, the order of the output in the file follows the standard
+loop order in the code, i.e., with the $z$-direction running fastest,
+then $y$, and with $x$ running slowest. For vector quantities
+(anything with more components than a scalar), the vector at each
+lattice site appears contiguously.
 
@@
 I/O decomposition cannot exceed the processor decomposition.
-See the advice in section \cite{section-advice} for guidelines.
+See the advice in section \ref{section-advice} for guidelines.
 
@@
 \subsection{Things to Remember and Things to Avoid}
-\ref{section-advice}
+\label{section-advice}
 
@@
 \item
-Make sure the assertions are switched off and appropriate optimisations
-are switched on for production runs (see Makefile).
+Make sure the assertions are switched off (via preprocessor option
+\texttt{-DNDEBUG}) and appropriate optimisations
+for the current compiler are switched on for production runs (see Makefile).
 \item

### 2012-03-21 trunk/src/blue_phase_rt.c
BPIII initialisation added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1544 40876441-3496-0410-a613-a2fca35b2ae2

@@ -242,6 +242,7 @@ void blue_phase_rt_initial_conditions(void) {
   char key1[FILENAME_MAX];
 
   double nhat[3] = {1.0, 0.0, 0.0};
+  double nhat2[3] = {64.0, 3.0, 1.0};
 
   info("\n");
 
@@ -315,6 +316,15 @@ void blue_phase_rt_initial_conditions(void) {
   if (strcmp(key1, "dtc") == 0) {
     info("Initialising Q_ab using DTC\n");
     blue_phase_DTC_init();
+  }
+
+  if (strcmp(key1, "bp3") == 0) {
+    info("Initialising Q_ab using BPIII\n");
+    RUN_get_double_parameter_vector("lc_init_bp3", nhat2);
+    info("BPIII specifications: N_DTC=%g,  R_DTC=%g,  ", nhat2[0], nhat2[1]);
+    if (nhat2[2] == 0) info("isotropic environment\n");
+    if (nhat2[2] == 1) info("cholesteric environment\n");
+    blue_phase_BPIII_init(nhat2);
   }
 
   if (strcmp(key1, "random") == 0) {

### 2012-03-21 htdocs/user.tex
BPIII initialisation added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1544 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \texttt{twist} initialises a simple cholesteric with the
-helical axis in the $z-$direction;  \texttt{o8m}
-gives equilibrium BPI, \texttt{o2} gives BPII. The appropriate
-choices for the other parameters must be made correspondingly.
+helical axis in the $z-$direction, whereas the values 
+\texttt{cholesteric\_x\_y\_z} provide according initialisations 
+with twist along the other axes.
+The value \texttt{o8m} gives equilibrium BPI, \texttt{o2} gives BPII.
+Another, cubic body-centered BP is obtained by the value \texttt{o5}.
+Hexagonal BPs are initialised with \texttt{h2d, h3da} and \texttt{h3db}.\\
+BPIII emerges from randomly positioned and oriented DTCs 
+embedded into an isotropic or cholesteric environment.
+It is initialised with the key value \texttt{bp3}, whereas the number
+$N$ and radius $R$ of the DTCs and the type of the environment 
+$ENV=\{0,1\}$, (0 = isotropic, 1 = cholesteric), is specified with the key value
+\texttt{lc\_init\_bp3} $N\_R\_ENV$.\\
+The appropriate choices for the other parameters must be made correspondingly.
 
@@
 The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
-there is effectively one cell in each local domain. This option
+there is effectively one cell in each local domain. (In practice,
+the constraint is slightly tighter than this if larger halo regions
+are in effect.) This option
 should not be required in normal circumstances.
@@
 
+\subsubsection{Flat wall lubrication correction}
+
+It is possible to prevent colloids colliding with the flat boundary walls
+by adding the correction to the normal component of the lubrication
+force. The general form of this correction is
+\begin{equation}
+\mathbf{F}_{\mathrm{lub}} = -6\pi \eta a_h^2
+%({\textstyle \frac{1}{h} - \frac{1}{h_c}})
+(1/h - 1/h_c)
+\mathbf{u} . \mathbf{r}.
+\end{equation}
+Here, $h$ is the surface to surface separation, and $h_c$ is the cutoff
+distance beyond which the correction is not applied. The relative
+velocity (assuming the wall is not moving) and normal separation are
+$\mathbf{u}$ and $\mathbf{r}$ respectively.
+
+The cutoff distance can be set from the input via
+
+\inputkey{boundary\_lubrication\_rcnormal 0.1}
+
+where the value is (strictly) set via calibration, cf. the hydrodynamic
+radius. In general, the cutoff is between zero and 0.5 lattice spacing,
+and can be smaller for larger particles. Because the wall velocity is
+fixed, this force correction should cause no stability issues in the
+colloid velocity update (provided a colloid's initial position is not
+very close to the wall). By default the cutoff $h_c = 0$.
+
 \subsection{Lees-Edwards planes}

### 2012-03-22 trunk/htdocs/user.tex
Update BPIII configuration
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1547 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \texttt{twist} initialises a simple cholesteric with the
-helical axis in the $z-$direction;  \texttt{o8m}
-gives equilibrium BPI, \texttt{o2} gives BPII. The appropriate
-choices for the other parameters must be made correspondingly.
+helical axis in the $z-$direction, whereas the values 
+\texttt{cholesteric\_x\_y\_z} provide similar initialisations 
+with twist along the other axes.
+The values \texttt{o8m} and \texttt{o2} give equilibrim BPI and BPII.
+Another, cubic body-centered BP is obtained by the value \texttt{o5}.
+Hexagonal BPs are initialised with \texttt{h2d, h3da} and \texttt{h3db}.\\
+BPIII emerges from a start configuration of randomly positioned 
+and oriented DTCs embedded into an isotropic or cholesteric environment.
+This is initialised with the key value \texttt{bp3}, whereas the number
+$N$ and radius $R$ of the DTCs and the type of the environment 
+$ENV=\{0,1\}$, (0 = isotropic, 1 = cholesteric) is specified 
+with the key value \texttt{lc\_init\_bp3} $N\_R\_ENV$. 
+Note that calculations can become surprisingly unstable if the 
+start configuration consits of too many and too large DTC-regions 
+with respect to system size and decomposition.\\
+The appropriate choices for the other parameters must be made correspondingly.
 
 An additional \texttt{lc\_q\_init\_amplitude} value sets the
-scalar order parameter  `amplitude' in each case.
+scalar order parameter  `amplitude' in each case. Typical
+values are given in the input file. Values which are too large can
+equally lead to instability.
 

### 2012-06-28 htdocs/user.tex
Comment on LB mobility
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1694 40876441-3496-0410-a613-a2fca35b2ae2

@@
 In this case the above parameters have the same meaning.
+
+Note that the relaxation time for the second distribution is
+related to the mobility by $tau_\phi = M\rho_0/\Delta t + 1/2$.
 
@@
 helical axis in the $z-$direction, whereas the values 
-\texttt{cholesteric\_x\_y\_z} provide according initialisations 
+\texttt{cholesteric\_x\_y\_z} provide similar initialisations 
 with twist along the other axes.
-The value \texttt{o8m} gives equilibrium BPI, \texttt{o2} gives BPII.
+The values \texttt{o8m} and \texttt{o2} give equilibrim BPI and BPII.
 Another, cubic body-centered BP is obtained by the value \texttt{o5}.
 Hexagonal BPs are initialised with \texttt{h2d, h3da} and \texttt{h3db}.\\
-BPIII emerges from randomly positioned and oriented DTCs 
-embedded into an isotropic or cholesteric environment.
-It is initialised with the key value \texttt{bp3}, whereas the number
+BPIII emerges from a start configuration of randomly positioned 
+and oriented DTCs embedded into an isotropic or cholesteric environment.
+This is initialised with the key value \texttt{bp3}, whereas the number
 $N$ and radius $R$ of the DTCs and the type of the environment 
-$ENV=\{0,1\}$, (0 = isotropic, 1 = cholesteric), is specified with the key value
-\texttt{lc\_init\_bp3} $N\_R\_ENV$.\\
+$ENV=\{0,1\}$, (0 = isotropic, 1 = cholesteric) is specified 
+with the key value \texttt{lc\_init\_bp3} $N\_R\_ENV$. 
+Note that calculations can become surprisingly unstable if the 
+start configuration consits of too many and too large DTC-regions 
+with respect to system size and decomposition.\\
 The appropriate choices for the other parameters must be made correspondingly.
@@
 An additional \texttt{lc\_q\_init\_amplitude} value sets the
-scalar order parameter  `amplitude' in each case.
+scalar order parameter  `amplitude' in each case. Typical
+values are given in the input file. Values which are too large can
+equally lead to instability.
 

### 2012-07-18 branches/CUDA_old/htdocs/hybrid.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1727 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  hybrid.tex
%
%  Description of  LB-FD hybrid considerations.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{BBL and hybrid dynamics}

[I assume this will be preceded by some description of the
governing equations for the tensor order parameter and their
treatment.]

The well-established procedure of bounce-back on links (BBL)
\cite{Ladd04, nguyen} has been employed for the representation
of the colloids as moving solid objects within the LBM.
In the hybrid LB/FD approach used here, BBL is retained for
the distributions, which are simply reflected at the solid-fluid
surface with a correction which depends on the local surface
velocity (see Figure Xa). The resulting change in momentum is
summed over the links to give the net hydrodynamic
force on the colloid, which is then used to update the
particle velocity, and
hence position, in a molecular dynamics-like step. Boundary
conditions for the finite difference equations for the order
parameter tensor are dealt with in a different way.

First, we note that the assignment of solid and fluid lattice
nodes for the order parameter follows that for the density:
inside and outside are distinguished using the nominal radius
of the colloid $a_0$ and its position. It is useful, in addition,
to think about a series of control volumes surrounding each lattice
node whose faces are aligned with the lattice (Figure Xb). A set of
these
faces constitute the solid-fluid boundary in the hybrid picture.

Boundary conditions for $Q_{\alpha\beta}$ are of two types:
homeotropic, where the director  $n_\alpha^0$ is aligned with
the local unit normal to the surface $\hat{n}_\alpha$, and planar,
where the director lies in the plane of the tangent to the
surface. For either choice of director at the surface, we
may set the corresponding value of $Q_{\alpha\beta}$ at
lattice nodes immediately inside the surface via
\begin{equation}
Q_{\alpha\beta}^0 = S^0 (n_\alpha^0 n_\beta^0
- {\scriptstyle\frac{1}{3}}\delta_{\alpha\beta})
\end{equation}
where the constant $S^0$ controls the degree of surface order.
This allows us
to compute, at all fluid nodes, the derivatives
$\nabla_\gamma Q_{\alpha\beta}$ and $\nabla^2 Q_{\alpha\beta}$
using the same finite difference stencil. This allows the
molecular field and hence the diffusive terms in the Beris
Edwards equation to be computed.

Also appearing in the Beris-Edwards equations is the velocity
gradient tensor, which can be handled in a similar fashion
close to the colloid. The velocity field at solid nodes
immediately inside the colloid surface are set to the solid
body velocity $\bf{u} + \bf{\Omega} \times \bf{r}$. Again,
the velocity gradient tensor $\partial_\alpha u_\beta$ may
be computed using the same stencil at all fluid nodes.

Advective fluxes of order parameter are computed at the faces
of the control volumes, and the boundary condition is zero
normal flux at solid-fluid interfaces. Note that the colloid
is assumed to be stationary in assigning these fluxes.

The force on the fluid originating from the order parameter
is computed via the discrete divergence of the stress
$P_{\alpha\beta}$. In the fluid, this is implemented by
interpolating $P_{\alpha\beta}$ to the control volume faces
and taking differences between faces in each direction. This
method has the advantage that, with the introduction of colloids,
an interpolation/extrapolation of $P_{\alpha\beta}$ to the
solid-fluid boundary is possible. This allows one to compute the
divergence of the stress at fluid nodes adjacent to the colloid
in the normal way. At the same time, the discrete equivalent of
\begin{equation}
F_\alpha^\mathrm{coll} = \int P_{\alpha\beta} \hat{n}_\beta dS
\end{equation}
by summing $P_{\alpha\beta}$ over the relevant solid-fluid control
volume faces. By construction, this ensures that momentum lost by
the fluid is gained by the colloid, i.e., global momentum is
conserved.

Finally, movement of the colloid across the lattice is accompanied
by changes in its discrete shape. The events necessitate the
removal or replacement of fluid information. For the replacement
of fluid at newly exposed lattice nodes, this means an
interpolation of nearby order parameter values in the fluid to
provide the new information. This is analogous to what is done
for the LB distributions.

\begin{figure}[h]
\begin{center}

%\input{/home/kevin/doc/rgrid/xfig/hybrid1.epic}
\input{hybrid2.epic}
\end{center}
\caption{The colloid (represented by the solid circle) moves continuously
across the lattice. Lattice sites inside are designated solid, and those
outside fluid (open and closed points, respectively). In the lattice
Boltzmann picture (left) the surface is defined by a set of links
$f_b$, which involve discrete vectors $\mathbf{c}_b \Delta t$ which
connect fluid and solid sites. For the order parameter (right), the
colloid is represented by the set of faces, e.g., that between sites
$i,j$ and $i+1,j$ with unit normal $\hat{n_x}$. Discretisation effects
are found to be negigible for radii greater than about 5 lattice units.}
\end{figure}

### 2012-07-18 branches/CUDA/htdocs/porous.tex
creating new CUDA branch from trunk
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1728 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-02-11 htdocs/porous.tex
@@
 $s_1,h_1, s_2, h_2, \ldots$.
+
+An example of how to construct a porous media file is provided in
+\texttt{util/capillary.c}, which builds an appropriate file for
+a square or circular capillary tube. Please see the comments in
+the file for further details. Note that the allowed
+solid/fluid status values are defined in \texttt{src/site\_map.h}.
+A solid boundary is \texttt{BOUNDARY}, while fluid is \texttt{FLUID}.
 

### 2012-07-18 branches/CUDA/papers/droplet-breakup/ms_revised.tex
creating new CUDA branch from trunk
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1728 40876441-3496-0410-a613-a2fca35b2ae2

%%
%%  This manuscript has appeared as
%%  Physical Review E, \textbf{85} 020403(R) 2012.
%%
%%  This file includes the supplimentary material, although the
%%  movies have not been retained here.
%%
\documentclass[aps,prl,superscriptaddress,twocolumn,showpacs,floatfix]{revtex4-1}
\bibliographystyle{apsrev4-1}

\usepackage{graphicx}% Include figure files
\usepackage{subfigure}
\usepackage{color}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\newcommand{\Boc}{Bo$_{\hbox{\rm c}}$~}
\newcommand{\Bocm}{Bo$_{\hbox{\rm c}}$}

\begin{document}
\title{Field-Induced Breakup of Emulsion Droplets Stabilized by Colloidal Particles}

\author{E. Grace Kim}
 \affiliation{Max-Planck-Institute for Intelligent systems, Heisenbergstr. 3, 70569 Stuttgart, Germany}
\author{Kevin Stratford}
\affiliation{Edinburgh Parallel Computing Centre, The University of Edinburgh, JCMB, The King's Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
\author{Paul S. Clegg} \author{Michael E. Cates}
\affiliation{SUPA, School of Physics and Astronomy, The University of Edinburgh, JCMB, The King's Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

 \date{\today}
 \begin{abstract}
 We simulate the response of a particle-stabilized emulsion droplet in an external force field, such as gravity, acting equally on all $N$ particles.  We show that the field strength required for breakup (at fixed initial area fraction) decreases markedly with droplet size, because the forces act cumulatively, not individually, to detach the interfacial particles. The breakup mode involves the collective destabilization of a solidified particle raft occupying the lower part of the droplet, leading to a critical force per particle that scales approximately as $N^{-1/2}$.
 \end{abstract}
 \pacs{82.70.Kj (emulsions); 82.70.Dd (colloids); 81.40.Np (failure of materials)}
 \maketitle

%\section{Introduction} \label{introduction}

The stabilization of emulsions is an important issue in industry, where it controls the functionality and shelf-life of many products \cite{miller:2008/a}. Alongside traditional surfactants, colloidal particles or nanoparticles have long been used to stabilize emulsion droplets (Pickering emulsions) \cite{pickering:1907/a}, with several emerging applications that exploit the nonequilibrium nature of their adsorption to the fluid-fluid interface \cite{Leal,binksnat,lin}. However this also poses obstacles to formulation: to maintain emulsion stability, it is essential that the colloidal particles remain sequestered on the interface, even when the system is subject to external challenges \cite{Aveyard,Leal}.

Brownian motion is generally insufficient to cause detachment \cite{Aveyard,Kevin}, so we ignore it here. In contrast, external body forces acting equally on each particle can, if strong enough, overcome the capillary forces holding them in place. Destabilizing forces can be caused by magnetic or electric fields or their gradients, and also by gravity. For simplicity we refer mainly to gravity in what follows.

A groundbreaking experimental study was performed by Melle {\it et al.}~\cite{fuller:2005/a}, using decane droplets in water stabilized by paramagnetic particles. They demonstrated, using a simple bar magnet, controlled field-gradient induced breakup of an otherwise highly stable emulsion.  A second study uses instead the dielectrophoretic force caused by an electric field gradient \cite{aubry:2010/a}. A third recent study has found complex dynamics in droplets stabilized by supra-colloidal (400$\mu$m) particles under gravity \cite{Joe}. (Scaling by the Bond number, defined below, indicates that the same physics should arise for, e.g., 1$\mu$m particles under bench-top centrifugation.) Although field-induced breakup is simultaneously a threat to an emulsion's stability and a promising route to its controlled breakage, the dependence of the critical field for breakup on the droplet size is far from understood.

Previously we performed lattice Boltzmann (LB) simulations on emulsion structures (primarily bicontinuous \cite{Kevin,Paul}) stabilized by magnetic colloids in the presence of a magnetic field gradient \cite{kim:one,kim:two}. There we argued that the
effect of the force does not act separately on each particle but is cumulative across the structure. (This explains the very low field-gradients required in \cite{fuller:2005/a}.) By that argument, a major fraction of the weight of all particles 
could be loaded onto a single `keystone' particle at the droplet bottom which should then detach, initiating breakup. 
In this paper, we study in detail the breakup dynamics of isolated emulsion droplets coated with $N$ (essentially hard-sphere) particles under gravity. 
We find a scaling of the critical force with $N$ that is inconsistent with the `keystone' argument, and associated instead with a failure mode in which a raft of jammed particles detaches from the droplet base taking liquid with it.

\textit{Control parameters:}
For a single particle adsorbed onto a flat horizontal interface between two fluids of equal density, detachment is controlled by the Bond number \cite{clift:1979/a}
${\rm Bo}= {3}{F}/ {(4\pi\sigma a)} $
where $a$ is the particle radius, $\sigma$ the fluid-fluid interfacial tension, and $F$ the normal detachment force (in gravity, $F$ is the buoyancy force on the particle).
Detachment of an isolated particle ($N=1$) occurs at a critical Bond number \Bocm $(1)$ which depends on the fluid-solid interfacial tensions through the contact angle $\theta$. In this work we consider only neutral wetting ($\theta  = 90^\circ$) for which \Bocm $(1)= 3/4$ \cite{stratford:2005/a,derjaguin:1946/a}. The critical Bond number for detachment of one particle from a fluid sphere of radius $R$ should also depend on $r\equiv R/a$; to allow for this we define \Bocm $(1,r)$ such that \Bocm $(1,\infty) =$ \Bocm $(1)$.  The numerically determined \Bocm $(1,r)$ additionally reflects various sources of discretization error (e.g., the fact that in LB the particle radius is not much larger than the interfacial width \cite{supmat}).
%(In general, \Bocm $(1,r) <$ \Bocm $(1)$.)

\textit{Simulations:}
The LB methodology used in this work is by now standard \cite{kim:one,stratford:2005/a,Bounce} and summarized in \cite{supmat}.
The initial condition in our simulations comprises a fluid droplet of radius $R$, immersed in another fluid of equal mass density, with $N$ neutrally wetting particles present on its surface. We choose a near-constant surface area fraction $\alpha=N\pi a^2/4\pi R^2 = N/ 4r^2 = 0.53-0.54$; this is high enough to represent a stable Pickering emulsion droplet but low enough to allow particles to rearrange on application of even a weak body force. The chosen $N$ values are $N=42,50,102,162,194$ and $700$.
Each droplet is placed in a box, with periodic boundary conditions (a repeat with closed-box boundaries for $N=700$ gave near-identical results), whose LB lattice size (depending on $R$) lies between $64^2 \times 128$ and $200^2\times 400$.

We have examined three different initial arrangements of particles: one disordered, one a regular hexagonal packing and one a regular square packing (subject to the defects required to tile these onto a sphere). For the ordered configurations, initial particle coordinates lie on the vertices of a suitable polyhedron. 
The particles and droplet are then relaxed in the absence of body forces ($F=0$) to allow the interfacial structure to equilibrate.
Once this is done, a vertical body force $F$ is switched on; this acts directly on each particle. To prevent overall translation of the droplet's center of mass, a balancing buoyancy force $NF$ is applied as a uniform upward force density
on the fluid nodes occupied by the droplet interior. The structure of the droplet is then evolved by the LB algorithm which faithfully includes the effects of fluid flow and hydrodynamic interactions.

\begin{figure}[tbph]
\centering
\subfigure[]{\label{fig:order_initial}\includegraphics[height=0.2\textwidth]
{fig1a.eps}}
\subfigure[]{\label{fig:order_stable01}\includegraphics[height=0.2\textwidth]
{fig1b.eps}}
\subfigure[]{\label{fig:order_stable02}\includegraphics[height=0.2\textwidth]
{fig1c.eps}}
\subfigure[]{\label{fig:order_unstable}\includegraphics[height=0.2\textwidth]
{fig1d.eps}}\\
\subfigure[]{\label{fig:disorder_initial}\includegraphics[height=0.2\textwidth]
{fig1e.eps}}
\subfigure[]{\label{fig:disorder_stable01}
\includegraphics[height=0.2\textwidth]{fig1f.eps}}
\subfigure[]{\label{fig:disorder_stable02}
\includegraphics[height=0.2\textwidth]{fig1g.eps}}
\subfigure[]{\label{fig:disorder_unstable}
\includegraphics[height=0.2\textwidth]{fig1h.eps}} \\
\subfigure[]{\label{fig:rN700initial}
\includegraphics[height=0.2\textwidth]{fig1i.eps}}
\subfigure[]{\label{fig:rN700_stable01}
\includegraphics[height=0.2\textwidth]{fig1j.eps}}
\subfigure[]{\label{fig:rN700_stable02}
\includegraphics[height=0.2\textwidth]{fig1k.eps}}
\subfigure[]{\label{fig:rN700_unstable}
\includegraphics[height=0.2\textwidth]{fig1l.eps}} 
\caption{(Color online.) \small{(a-h): Snapshots at $N=194$. (a) The ordered initial configuration; (b)-(d) are snapshots at $t=37t_S$ evolved from (a) with $\rm Bo=0.050, 0.051$ and $0.052$ respectively. Stable states are shown in (b) $\rm Bo=0.050$ with $N_{c}=80$, and (c) $\rm Bo=0.051$ with $N_{c}=78$. (d) Unstable state found at $\rm Bo=0.052$ with $N_{c}=121$ at $t=t_d=37t_S$.
(e) The disordered initial configuration; (f)-(h) are snapshots at $t=30t_S$ evolved from  (e) with $\rm Bo=0.050, 0.053$ and $0.055$ respectively. Stable states are shown in (f) $\rm Bo=0.050$ with $N_{c}=74$, and (g) $\rm Bo=0.053$ with $N_{c}=80$. (h) Unstable state found at $\rm Bo=0.055$ with $N_{c}=122$ at $t=t_d=30t_S$.
(i)-(l): Snapshots for N=700. (i) The disordered initial configuration; (j)-(l) are snapshots at $t=14.5t_S$ from the configuration (i) with $\rm Bo=0.030, 0.033$ and $0.034$ respectively. Stable states are shown in (j) $\rm Bo=0.030$ with $N_{c}=193$, and (k)$\rm Bo=0.033$ with $N_{c}=315$. (l) Unstable state found at $\rm Bo=0.034$ with $N_{c}=342$ at $t=t_d=14.5t_S$. }
Particles separated by $h<0.04a$ from a neighbor are shown in red.}
\label{fig:stable/unstable}
\end{figure}

\textit{Results:}
We define the critical Bond number \Bocm $(N,r)$ as the threshold beyond which one or more particles break away from a droplet covered by $N$ particles. For Bo $<$ \Bocm $(N,r)$ the droplet instead achieves a final steady state that is distorted but not ruptured. Since we work at constant $\alpha = N/4r^2 \simeq 0.53$, we now drop the argument $r$ and refer simply to \Bocm $(N)$.
In practice we find particle ejection always precedes disconnection of a droplet into two pieces and therefore used the time of first detachment, $t_d$, as a measure of when breakup occurs. (As time unit we adopt the Stokes time $t_S =6\pi\eta a^2/F$.)

Starting from an equilibrated spherical droplet and switching on the external forces, particles start to move downwards (while the buoyancy force holds the droplet up). Upper sections of the droplet become bare \cite{fuller:2005/a,Joe} and a dense interfacial `sediment' is created towards the bottom of the droplet. As the sediment builds, forces are transmitted to the particles beneath. The lower part of the droplet can then become highly distorted. 
%
In contrast to droplets initiated from a disordered configuration, which maintain a prolate spheroidal shape throughout the stable regime, the ordered arrangements give almost symmetric force patterns in which the most unstable particle lies at the tip of a cusp that develops at the droplet bottom. For Bo $<$ \Bocm $(N)$, such a cusped configuration can remain stable indefinitely. Fig.~\ref{fig:stable/unstable} shows snapshots for both stable and unstable 
cases 
%(Bo just below and just above \Bocm $(N)$) 
at the time $t_d$. Shown in red are those particles whose interparticle separation $h<0.04 a$. This criterion identifies the main load-bearing particles in the system \cite{supmat}.

Although the body force changes only by two percent between the highest stable and lowest unstable value examined in Fig.~\ref{fig:stable/unstable}, there is a clear difference in their configurations at the detachment time $t_d$. In both cases, 
close-packed clusters of $N_c$ load-bearing particles are observed at the bottom of droplets; while the geometry of these regions depends on initial conditions as detailed above, in the last stable droplet configuration  $N_c\simeq 0.4N$ for the ordered and disordered cases. In the unstable state, immediately prior to particle ejection, $N_c\simeq 0.6N$ is rather larger but again nearly independent of the initial condition.
In Table \ref{table:N194unstable}
we report $N_{c}(t_d)$ for several values of Bo just above the threshold, in the case of ordered (square) and disordered packings with $N=194$.  We find $N_{c}>N/2$ in all unstable cases, with rather little dependence on Bo (once the critical value is exceeded). As expected, however, a stronger body force (larger Bo) causes the detachment time $t_d$ to decrease markedly.

\begin{table}
\begin{center}
\begin{tabular}{c c c c }
\hline\hline
Initial state & $\rm Bo$    & $t_{d}/t_S$ & $N_{c}(t_d)$ \\ \hline
Ordered         & $0.052^\dag$ & $37$  & 121 \\
              & $0.053$ & $27$  & 125 \\
              & $0.054$ & $21$   & 114 \\ 
              & $0.055$ & $18$   & 130 \\ \hline
Disordered      & $0.055^\dag$ & $30$   & 122 \\ 
              & $0.060$ & $15$   & 126 \\
              & $0.065$ & $10$   & 127 \\ \hline

\end{tabular}   
\caption[smallcaption]{The time $t_{d}$ of first particle detachment and the size $N_{c}$ of the load-bearing cluster immediately prior to $t_d$ for unstable droplets of $N=194$ with ordered and disordered initial states. Superscript $^\dag$ denotes \Bocm $(N)$ (numerically, the smallest Bo found for which breakup occurs).}
\label{table:N194unstable}
\end{center}
\end{table}

Fig.~\ref{fig:Ncluster} depicts the time evolution of $N_{c}$ in the cases with $N=194$. This shows the buildup of the sedimented cluster after switching on gravity, followed by a period of saturation. Beyond $t=13t_S$, stable droplets reach steady state with roughly constant $N_{c}$, although disordered cases show fluctuations due to reorganization of the randomly packed 2D sediment. In the unstable cases, such reorganization is always seen during the plateau region; following a series of these plastic events $N_{c}$ starts increasing steadily until breakup occurs.

\begin{figure}[tbph]
\centering
\vspace{.15in}
{\includegraphics[width=0.45\textwidth]{figure2_ks.eps}}
\caption{(Color online.) The evolving cluster size, $N_{c}(t)$ for the simulations shown in Fig.~\ref{fig:stable/unstable} (a-h).  Left: $N_{c}(t)$ for the ordered initial configuration: black circles, $\rm Bo=0.050$, blue triangles $\rm Bo=0.051$, red squares $\rm Bo=0.052$. Right:  disordered initial configuration: black circles  $\rm Bo=0.050$, blue triangles $\rm Bo=0.053$, red squares $\rm Bo=0.055$. For unstable cases, the plots end upon particle ejection.}
\label{fig:Ncluster}
\end{figure}

\begin{figure}[tbph]
\centering
\vspace{.1in}
\includegraphics[width=0.3\textwidth]{figure3_ks.eps}
\caption{(Color online.) Log-log plot of critical Bond number for a droplet covered with $N$ particles, \Bocm $(N)$, vs $N$. Blue diamonds: square ordered; green squares: disordered; black circles: hexagonal ordered. Blue dot-dashed line: best fit of all data to  ${\rm Bo}_{\hbox{\rm c}}(N) = A N^{-1/2}$ ($A = 0.71$). Red dotted line: best fit to ${\rm Bo}_{\hbox{\rm c}}(N) = A N^{B}$ ($A = 0.38; B=-0.37$). Black dashed line: bound found from $\lambda=3/2$ (see text).}
\label{fig:scalingN0.5}
\end{figure}

\textit{Scaling of critical Bond number:}
The above mechanism entails asssembly of a close-packed loadbearing cluster, involving roughly half the particles, which either maintains itself indefinitely (Bo $<$ \Bocm $(N)$) or, after a series of plastic rearrangements, 
finally gives way, causing particle ejection and droplet rupture (Bo $>$ \Bocm $(N)$). The droplet prior to this event is distorted significantly; its shape, while dependent in detail on the initial condition, is roughly independent of $N$, and involves order-one departures from a sphere. There is no sign of extreme cylindrical elongation as can arise in shear flow \cite{Bentley,Lyn}.

Although breakup is preceded by particle ejection, with all parameters studied here we do not see a stream of particles being ejected from the lower tip of the droplet as might be expected were the sediment of loadbearing particles to behave as a fluid. Such an outcome would allow the weight of all $N$ particles to be borne by a single `keystone' particle at the bottom of the droplet. Ejection of that particle, possibly nucleating an instability of the entire structure, could then be expected at a critical Bond number scaling as \Bocm $\sim$\Bocm $(1,r)/N$ \cite{kim:one}. For our nearly hard-sphere particles\cite{supmat}, this is not observed here \cite{Foot2}; we instead find a scaling closer to $N^{-1/2}$ (see Fig.~\ref{fig:scalingN0.5}). This is because the cluster has formed a jammed structure in which forces are transmitted laterally as well as vertically. This semi-rigid assembly then becomes collectively unstable (see Movies at \cite{supmat}).

Given the destabilization mode observed, we now estimate \Boc by addressing the force balance on the close-packed sediment, viewed as a solid cluster. (See \cite{Oettel,Stone} for related arguments.) This has weight $FN_c\simeq FN/2$ and to a first approximation occupies the lower half of a droplet that is only moderately distorted. Accordingly at Bo $=$ \Boc this weight is just balanced by the surface tension force acting vertically across the equator of the deformed droplet, $2\pi\sigma \tilde R$ where $\tilde R \simeq R/2$ is the radius at the equator. Bearing in mind that the initial surface coverage (here held constant) obeys $\alpha = Na^2/4R^2 = N/4r^2$ we then have the estimate \Bocm $(N,r) \simeq 3\tilde R/Na$ or, introducing a negotiable geometrical constant $\lambda \simeq 0.75$,
$
%\begin{equation}\label{eq:effective_capillary}
{\rm Bo}_{\hbox{\rm c}}(N,r)  = {\lambda}/{\sqrt{\alpha N}}.
%\nonumber
%\end{equation}
$
The predicted scaling \Bocm $\sim N^{-1/2}$ is tested in Fig.~\ref{fig:scalingN0.5} which plots all data for \Boc against $N$ on a log-log plot. The best fit (treating data for all initial conditions equally)  has $\lambda = 0.52$.

Table \ref{table:criticalBo} shows all the \Boc values found in this study alongside run-by-run values of $\lambda$. The residual increasing trend of $\lambda$ with $N$ is consistent with the fact that the pre-breakup droplet shape becomes more spherical for larger $N$. Indeed the quantity $\lambda R/\tilde R$, with $\pi\tilde R^2$ defined as the cross sectional area of the deformed droplet at the height of the uppermost load-bearing particle, has no residual trend (see Table \ref{table:criticalBo})  \cite{Footnote}. A free fit to the exponent gives \Bocm $\sim N^{-0.37}$, or $\tilde R/R\sim N^{0.13}$. We have no explanation for this weak residual trend, but note that it cannot persist to very large $N$ unless droplets become oblate. Excluding that outcome, a bound \Bocm $\ge 3/(2\sqrt{\alpha N})$ at large $N$ is found by balancing the tensile force $2\pi\sigma R$ in an undeformed spherical configuration against the weight $FN/2$ of its lower hemisphere. (Any prolate deformation reduces the tensile force, decreasing \Boc.) This bound is plotted in Fig.~\ref{fig:scalingN0.5}; extrapolating our $N^{-0.37}$ fit, saturation is expected only at $N\simeq 10^4$.

\begin{table}
\begin{center}
\begin{tabular}{c  r c r c c r r r r}
\hline
%\hline
 & $N$ & ${\rm Bo}_{\hbox{\rm c}}(1,r)$ & ${\rm Bo}_{\hbox{\rm c}}(N)$ & $N_{c}(t_d)$ & $t_{d}/t_S$ & $\lambda$ & $\tilde R/R$ & $\lambda R/\tilde R$\\ \hline
S & $50$ & $0.55$ & $0.085$ & $29$ & $103$ & $0.44$  & $0.52$ & 0.84\\
        & $102$& $0.57$ & $0.068$ & $54$ & $38.5$ & $0.50$  & $0.65$ & 0.77\\
        & $194$& $0.60$ & $0.052$ & $121$ & $37$& $0.53$  & $0.55$ & 0.97\\ \hline
H & $42$ & $0.54$ & $0.10$ & $23$ & $38$ & $0.47$  & $0.49$ & 0.95\\
& $162$& $0.59$ & $0.057$ & $88$ & $38.5$ & $0.53$  & $0.69$ & 0.77\\ \hline
D& $50$ & $0.55$ & $0.090$ & $31$ & $59$ & $0.46$ &  $0.63$ & 0.73\\
        & $102$& $0.57$ & $0.070$ & $52$ & $21$ &  $0.52$  & $0.59$ & 0.88\\
        & $194$& $0.60$ & $0.055$ & $122$ & $30$& $0.56$  & $0.60$ & 0.94\\ 
        & $700$& $0.63$ & $0.034$ & $342$ & $14.5$& $0.65$  & $0.77$ & 0.85\\ \hline
\end{tabular}
\caption[smallcaption]{Critical Bond numbers (\Bocm $(N)$ for a droplet coated with $N$ particles at initial coverage $\alpha \simeq 0.53$, and \Bocm $(1,r)$ for a single particle on a droplet of the same radius); size of the load-bearing cluster prior to breakup; detachment time; breakup parameter $\lambda$; and shape parameter $\tilde R/R$. Initial configurations S, square, H, hexagonal, D, disordered. }
\label{table:criticalBo}
\end{center}
\end{table}

\textit{Conclusion:}
We have investigated by LB simulations the destabilization of an emulsion droplet coated with hard spheres under a body force that acts equally on all the particles
\cite{fuller:2005/a,aubry:2010/a,Joe}. We confirmed that the force required for destabilization is far less than would be expected if the force were to cause independent detachment of the particles from the fluid-fluid interface \cite{kim:one}. However, it is also far greater than would be expected if the weight of all $N$ particles were to accumulate as a net detachment force on a single `keystone' particle at the bottom of the droplet. That picture, which implicitly considers the stabilizing particle layer to comprise a 2D fluid, must now be replaced by one in which the lower part of that layer has solidified. A simple geometrical analysis then suggests a critical Bond number scaling as $N^{-1/2}$ which gives a satisfactory first account of the simulation data.

\textit{Acknowledgments:}
EGK thanks Siegfried Dietrich for hospitality in Stuttgart. We thank Jan Guzowski and Joe Tavacoli for useful discussions. MEC is funded by the Royal Society.

\begin{thebibliography}{100}

\bibitem{miller:2008/a}
%C. A. Miller and P. Neogi, \textit{Interfacial Phenomena, Equilibrium and Dynamic Effects}, Taylor \& Francis Group LLC: CRC Press, \textbf{2008}.
F. Leal-Calderon, V. Schmitt and J. Bibette, \textit{Emulsion Science: Basic Principles}, 2nd Ed, Springer, NY (2007).

\bibitem{pickering:1907/a}
S. U. Pickering, J. Chem. Soc. \textbf{91}, 2001 (1907).

\bibitem{Leal}
F. Leal-Calderon, and V. Schmitt, Curr. Opin. Colloid Interface Sci.\textbf{13}, 217 (2008).

\bibitem{binksnat} B. P. Binks and R. Murakami, Nat. Mater. \textbf{5}, 865 (2006).

\bibitem{lin} S. H. Kim et al, Chem. Mater. \textbf{19}, 4751 (2007).

\bibitem{Aveyard}  
R. Aveyard, B. P.Binks, and J. H. Clint, Adv. Colloid Interface Sci. \textbf{100}, 503 (2003).

\bibitem{Kevin}
K. Stratford, R. Adhikari, I. Pagonabarraga, J.-C. Desplat, and M. E. Cates,  Science \textbf{309}, 2198 (2005).

\bibitem{fuller:2005/a}
S. Melle, L. Mauricio, and G. G. Fuller, Langmuir \textbf{21}, 2158 (2005).

\bibitem{aubry:2010/a}
K. Hwang, P. Singh, and N. Aubry, Electrophoresis \textbf{31}, 850 (2010).

\bibitem{Joe}
J. W. Tavacoli, J. H. J. Thijssen, and P. S. Clegg, Soft Matter \textbf{7}, 7969 (2011).

\bibitem{Paul}
J. W. Tavacoli, J. H. J. Thijssen, A. B. Schofield, and P. S. Clegg, Adv. Funct. Mater. \textbf{21}, 2020 (2011).

\bibitem{kim:one}
E. Kim, K. Stratford, and M. E. Cates, Langmuir \textbf{26}, 7928 (2010).
\bibitem{kim:two} 
E. Kim, Ph.D. Thesis, University of Edinburgh (2009).

\bibitem{clift:1979/a}
R. Clift, J. R. Grace, and M. E. Weber, \textit{Bubbles Drops and Particles}, New York: Academic Press (1979).

\bibitem{stratford:2005/a}
K. Stratford, R. Adhikari, I. Pagonabarraga, and J. -C. Desplat, J. Stat. Phys. \textbf{121}, 163 (2005).

\bibitem{derjaguin:1946/a}
B. V. Derjaguin, Dokl. Akad. Nauk. USSR  \textbf{51}, 517 (1946).

\bibitem{supmat} The simulation methodology, additional results, and movies are available online at XXXXXXXX.

\bibitem{Bounce} N. Q. Nguyen and A. J. C. Ladd, Phys. Rev. E \textbf{66}, 046708 (2002).

\bibitem{kendon:2001/a}
V. M. Kendon, M. E. Cates, I. Pagonabarraga, J. -C. Desplat, and P. Bladon, J. Fluid. Mech. \textbf{440}, 147 (2001).

\bibitem{Bentley}
B. J. Bentley, and L. G. Leal, J. Fluid. Mech. \textbf{167}, 241 (1986).

\bibitem{Lyn}
A. J. Wagner, L. M. Wilson, and M. E. Cates, Phys. Rev. E \textbf{68}, 045301 (2003).

\bibitem{Oettel}
A. Dominguez, M. Oettel, and S. Dietrich, J. Chem. Phys. \textbf{128}, 114904 (2008).

\bibitem{Stone} A. B. Subramaniam, et al, 
%M. Abkarian, L. Mahadevan and H. A. Stone, 
Langmuir \textbf{22}, 10204 (2006); M. Abkarian et al, Phys. Rev. Lett. \textbf{99}, 188301 (2007).

\bibitem{Foot2} For hard spheres the collective breakup mode is more unstable than the keystone detachment mode. Remarkably, this can be reversed by halving the capillary radius of the particles, while maintaining the range of the hard-core interparticle forces. This barely affects force balance for the collective failure mode, but reduces four-fold the maximum force the keystone can sustain; ejection of individual particles now happens first \cite{supmat}.

\bibitem{Footnote} The deviation might also be linked to \Bocm $(1,r)$, which increases weakly with $N$; however, our collective mechanism presents no obvious role for this quantity.

\end{thebibliography}

\newpage

\section{Field-Induced Breakup of Emulsion Droplets Stabilized by Colloidal
Particles: Additional Online Material}

\section{Simulation method} \label{method}

All simulations are performed using the LB method on a $D3Q19$ lattice for
a binary fluid, with the bounce-back-on links method used to handle momentum exchange with solid particles \cite{Bounce}, as extended to the binary fluid case in \cite{stratford:2005/a}. The fluid-fluid interfacial tension and width are determined by the parameters of Ginzburg-Landau free energy functional $F[\psi]= \int dV \left( \frac{A}{2}\psi^2 + \frac{B}{4}\psi^4 +  \frac{\kappa}{2}(\nabla \psi)^2 \right)$, where $\psi$ is a compositional order parameter.
Following \cite{kendon:2001/a,kim:one} we express these parameters in lattice units (LU) as $-A=B=0.0011832$ and $\kappa=0.0023664$; the interfacial thickness and the nominal interfacial tension are deduced as $\xi=2.0$ LU and $\sigma_0=1.5776\times10^{-3}$ LU respectively. The fluid viscosity is set at $\eta=0.1$ LU. Lattice units are as defined in \cite{kim:one} but for our purposes are arbitrary: all that matters is the Bond number. Note that the Reynolds number, which controls the relative importance of inertia compared to viscosity, remains small throughout our simulations which represent the Stokes flow limit \cite{kim:one}. The sedimentation velocity of one particle $v_s$ can be used to define a capillary number Ca $=\eta v_s/\sigma = 2{\rm Bo}/9$ which remains small in all our simulations.

From LB benchmark tests with this parameter set, the numerical value of surface tension is determined as $\sigma = 1.51 \times 10^{-3}$ LU which is 4\% less than the nominal value $\sigma_0$. (Discrepancies of this scale are common in LB \cite{kendon:2001/a}.) Monodisperse neutral-wetting particles are introduced with hydrodynamic radius $a_h=3.9$ LU. This choice is imposed by the numerical resources available; note however that $a_h$ is only twice the interfacial width $\xi$. For that reason, one expects some discrepancy between the values of \Bocm $(1,r)$ seen in simulations and those arising for a structureless interface ($\xi\to 0$). The values tabulated in the main text are those found by numerical simulations on a single particle residing on a droplet of radius $R = ra$. We have however confirmed that for a flat interface ($r\to\infty$), our methods recover the expected result: indeed we find \Bocm $(1) = 0.74\simeq 3/4$. Particle ejection is defined by the surface-average value of $\psi$ becoming sufficiently different from zero.

Interparticle forces are mediated by a soft-core potential $U^{sc}(h)$ chosen to prevent the surface-to-surface interparticle separation $h$ from becoming too small. (To model capillary forces correctly, it is important that the lattice Boltzmann nodes surrounding each particle mainly represent fluid not solid sites.) We choose $U^{sc}(h)=u(h)-u(h_c)-(h-h_c)(\frac{du}{dh})_{h=h_c}$, where $u(h)=\gamma (h_0/h)$, and $h_c$ is the maximum range of the interaction; we select $\gamma=10^{-4}$, $h_0=0.1$, and $h_c=0.25$ \cite{kim:one}. In practice, the particles behave as nearly impenetrable spheres of radius $a \simeq a_h\simeq 4$ LU. Finally, to allow for the discretization error in the hydrodynamic forces at small separations, a lubrication force (proportional to the normal component of the interparticle velocity) acts whenever $h<0.3$ LU. The criterion $h < 0.15$ LU $= 0.04a$ is used to identify load-bearing particle contacts; this lies within the range of the soft-core potential defined above.

\section{Additional Results}

%Table \ref{table:N194unstable} is discussed in the main text.

Figure \ref{fig:order/random} shows snapshots of the particle configurations in much smaller droplets ($N=50$), prepared initially in ordered (square lattice) and disordered states, at two different later times. These are for the lowest unstable Bo ($\simeq$ \Bocm $(N)$) so that both droplets will subsequently break. The left pictures in Figures \ref{fig:order/random}(a) and \ref{fig:order/random}(b) are configurations where $N_{c}$ has reached the near-saturation plateau value (compare Fig.~\ref{fig:Ncluster}) showing semi-crystalline clusters at the droplet bottom. The right pictures in Figures \ref{fig:order/random}(a) and \ref{fig:order/random}(b) are the configurations immediately before the first particle detachment. In each case there is a neck close to the droplet equator, with load-bearing particles below the neck. Although $N$ here is quite small, these configurations are remarkably similar to those seen in Fig.~\ref{fig:stable/unstable} for $N=194$ and even $N = 700$.

Figure \ref{fig:N194aH7.8/unstable} shows the breakup mode for an emulsion droplet ($N=194$) in which the hard-core interaction radius of the particles is unchanged but the radius they occupy on the interface is halved. For a given gravitational force per particle, this does not alter the force balance argument for the collective detachment of the lower hemisphere. However it does reduce four-fold the maximum total force that can be transmitted to the bottom-most particle without this detaching individually from the interface. In this case, a `keystone' type failure mode is seen, in which particles stream off the bottom of the droplet without entrainment of the enclosed fluid. Note that similar physics could arise for partially wetting particles with a contact angle far from 90 degrees: in this case also, the radius of the interfacial disc covered by each particle is significantly less than its hard sphere interaction radius. This again reduces the collective force sustainable without individual detachment, and therefore lowers stability with respect to the keystone detachment mode without having the same effect on the collective mode (in which a naive force balance again equates the total mass in the lower hemisphere to $2\pi R\sigma$, independent of contact angle).

\section{Movies}

The following online movies are available showing the time evolution of stable and unstable droplets for various sizes and initial conditions.

Movie 1. $N = 194$, ordered initial condition, stable (parameters as in Fig.1b).

Movie 2. $N = 194$, ordered initial condition, unstable (parameters as in Fig.1d).

Movie 3. $N = 194$, disordered initial condition, stable (parameters as in Fig.1f).

Movie 4. $N = 194$, disordered initial condition, unstable (parameters as in Fig.1h).

Movie 5. $N = 700$, disordered initial condition, stable (parameters as in Fig.1j).

Movie 6. $N = 700$, disordered initial condition, unstable (parameters as in Fig.1l).

Movie 7. $N=194$, ordered initial condition, unstable to `keystone' type failure mode (parameters in Fig. 5c).  
\vfill\eject

\begin{figure}[tbph]
\centering
\vspace{.5in}
\subfigure[]{\includegraphics[height=0.3\textwidth]{fig4a.eps}}
\subfigure[]{\includegraphics[height=0.3\textwidth]{fig4b.eps}}
\vspace{.05in}
\caption{Comparison of particle-packing between ordered and disordered initial configurations at different time steps. (a) Snapshots of a droplet at $t=89 t_S$ on the left and at $t_{d}=100 t_S$ on the right from initial ordered configuration. (b) Snapshots of a droplet at $t=51 t_S$ (left) and at $t_{d}=59 t_S$ (right) from disordered configuration.}
\label{fig:order/random}
\end{figure}

\begin{figure}[tbph]
\centering
\subfigure[]{\label{fig:aH7.8_initial}\includegraphics[height=0.2\textwidth]{fig5a.eps}}
\subfigure[]{\label{fig:order_stable01A}\includegraphics[height=0.2\textwidth]{fig5b.eps}}
\subfigure[]{\label{fig:order_stable02A}\includegraphics[height=0.2\textwidth]{fig5c.eps}}
\subfigure[]{\label{fig:order_unstableA}\includegraphics[height=0.2\textwidth]{fig5d.eps}}\\
\caption{Snapshots for droplets at $N=194$ with hard core interaction radius twice the capillary radius, showing keystone detachment mode in place of the collective breakup. (The capillary radius is used in the definition of the Bond number; the hard core radius is used for the visualization.) (a) The ordered initial configuration; (b)-(d) are snapshots at $t=11.4t_S$ evolved from (a) with $\rm Bo=0.048, 0.064$ and $0.11$ respectively. (b) $\rm Bo=0.048$ with $N_{c}=157$ and $N_{f}=2$, (c) $\rm Bo=0.064$ with $N_{c}=160$ and  $N_{f}=8$. (d) $\rm Bo=0.11$ with $N_{c}=157$ and $N_{f}=21$.}
\label{fig:N194aH7.8/unstable}
\end{figure}

\end{document}

### 2012-07-18 branches/CUDA/current/utilities/io_sort.c

@@ -1,0 +1,475 @@
+/****************************************************************************
+ *
+ *  io_sort.c
+ *
+ *  Invoke with three arguments, e.g.,
+ *
+ *  ./a.out type_id nio_groups file_stub
+ *
+ *  To sort parallel I/O into a single ordered file for analysis.
+ *
+ *  Type 1 Disclination locations
+ *         3*int (being ic jc kc)
+ *
+ *  Type 2 Order parameter etc on lattice
+ *         3*int + 10*double
+ *
+ *  Type 3 Stress on lattice
+ *         3*int + 9*double (being ic jc kc Sxx Sxy ... Szz)
+
+ *  Type 4 Director on lattice
+ *         3*int + 3*double (being ic jc kc dx dy dz)
+ *
+ *  Expected file names e.g.,: file_stub-n-nio
+ *         n = current io group
+ *         nio = total number of io groups (or, number of files per step)
+ *
+ *  The output will appear in a new file 'file_stub'
+ *
+ ****************************************************************************/
+
+#include <stdio.h>
+#include <stdlib.h>
+
+/* System size (could be read at run time) */
+
+int lx_ = 128;
+int ly_ = 128;
+int lz_ = 32;
+
+float * data_out_;
+int nio_;
+
+
+int get_global_index(int, int, int);
+void read_disclination(const char *);
+void write_disclination(const char *);
+void read_order_velo(const char *);
+void write_order_velo(const char *);
+void read_stress(const char *);
+void write_stress(const char *);
+void read_director(const char *);
+void write_director(const char *);
+
+int main (int argc, char ** argv) {
+
+  int n;
+  int nbuffer;
+  int id_type;
+  char file_name[FILENAME_MAX];
+  char file_stub[FILENAME_MAX];
+
+  void (* read_function)(const char *);
+  void (* write_function)(const char *);
+
+  /* Read command line */
+
+  if (argc != 4) {
+    printf("check command line arguments\n");
+    exit(-1);
+  }
+
+  id_type = atoi(argv[1]);
+  nio_     = atoi(argv[2]);
+  sprintf(file_stub, argv[3]);
+
+  printf("Input type is %d\n", id_type);
+  printf("Input number I/O groups is %d\n", nio_);
+  printf("Input file stub is %s\n", file_stub);
+
[…]

### 2012-07-18 branches/CUDA/src/colloids_halo.c

@@ -1,0 +1,433 @@
+/*****************************************************************************
+ *
+ *  colloids_halo.c
+ *
+ *  Halo exchange of colloid state information.
+ *
+ *  $Id: colloids_halo.c,v 1.4 2010-11-29 17:03:16 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdlib.h>
+#include <string.h>
+#include <float.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "colloids.h"
+#include "colloids_halo.h"
+
+static colloid_state_t * send_;   /* Send buffer */
+static colloid_state_t * recv_;   /* Recv buffer */
+static const int tagf_ = 1061;
+static const int tagb_ = 1062;
+
+static void colloids_halo_load(int dim, const int nsend[2]);
+static void colloids_halo_unload(int nrecv);
+static void colloids_halo_number(int dim, int nsend[2], int nrecv[2]);
+static void colloids_halo_irecv(int dim, int nrecv[2], MPI_Request req[2]);
+static void colloids_halo_isend(int dim, int nsend[2], MPI_Request req[2]);
+static  int colloids_halo_load_list(int ic, int jc, int kc,
+				    const double rperiod[3], int noff);
+
+/*****************************************************************************
+ *
+ *  colloids_halo_state
+ *
+ *  A limited choice of periodic conditions is available:
+ *    {1, 1, 1}, {1, 1, 0}, {1, 0, 0}, or {0, 0, 0}
+ *
+ *  These are trapped explicitly in serial here. In parallel,
+ *  communications at the boundaries involve MPI_PROC_NULL and are
+ *  trapped later by setting the number of incoming colloids to zero.
+ *
+ *****************************************************************************/
+
+void colloids_halo_state(void) {
+
+  colloids_halo_dim(X);
+  colloids_halo_dim(Y);
+  colloids_halo_dim(Z);
+
[…]

### 2012-07-18 branches/CUDA/src/gradient_2d_5pt_fluid.c

@@ -1,0 +1,467 @@
+/*****************************************************************************
+ *
+ *  gradient_2d_5pt_fluid.c
+ *
+ *  Gradient operations for 2D five point stencil.
+ *
+ *                    (ic, jc+1)
+ *         (ic-1, jc) (ic, jc  ) (ic+1, jc)
+ *                    (ic, jc-1)
+ *
+ *  d_x phi = [phi(ic+1,jc) - phi(ic-1,jc)] / 2
+ *  d_y phi = [phi(ic,jc+1) - phi(ic,jc-1)] / 2
+ *
+ *  nabla^2 phi = phi(ic+1,jc) + phi(ic-1,jc) + phi(ic,jc+1) + phi(ic,jc-1)
+ *              - 4 phi(ic,jc)
+ *
+ *  Corrections for Lees-Edwards planes and plane wall in X are included.
+ *
+ *  $Id: gradient_2d_5pt_fluid.c,v 1.2 2010-10-15 12:40:03 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  (c) The University of Edinburgh (2010)
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdlib.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "leesedwards.h"
+#include "wall.h"
+#include "gradient.h"
+#include "gradient_2d_5pt_fluid.h"
+
+static void gradient_2d_5pt_fluid_operator(const int nop,
+					   const double * field,
+					   double * grad,
+					   double * delsq,
+					   const int nextra);
+static void gradient_2d_5pt_fluid_le_correction(const int nop,
+						const double * field,
+						double * grad,
+						double * delsq,
+						int nextra);
+static void gradient_2d_5pt_fluid_wall_correction(const int nop,
+						  const double * field,
+						  double * grad,
+						  double * delsq,
+						  const int nextra);
+
+/*****************************************************************************
+ *
+ *  gradient_2d_5pt_fluid_init
+ *
+ *****************************************************************************/
+
+void gradient_2d_5pt_fluid_init(void) {
+
[…]

### 2012-07-18 branches/CUDA/src/ran.c

@@ -1,0 +1,240 @@
+/*****************************************************************************
+ *
+ *  random.c
+ *
+ *  Random number generators for serial and parallel.
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *
+ *****************************************************************************/
+
+#include <math.h>
+
+#include "pe.h"
+#include "ran.h"
+#include "runtime.h"
+
+struct lecuyer {
+  double rspare;                  /* Spare Gaussian random number */
+  int    ispare;                  /* Flag for ditto */
+  int    rstate[5];               /* State space (32-bit) */
+};
+
+static struct lecuyer p_rng;      /* Parallel generator */
+static struct lecuyer s_rng;      /* Serial generator */
+
+static double ran_gaussian(struct lecuyer *);
+static double ran_lecuyer(struct lecuyer *);
+
+#define _rmodulus 4.656612873077393e-10
+#define _m        2147483647
+#define _a1       107374182
+#define _q1       20
+#define _r1       7
+#define _a5       104480
+#define _q5       20554
+#define _r5       1727
+
+/****************************************************************************
+ *
+ *  ran_init
+ *
+ *  Initialise both serial and parallel random states.
+ *
+ ****************************************************************************/
+
+void ran_init( void ) {
+
+  int n;
+  int scalar_seed = 7361237;
+
+  /* Look for "random_seed" in the user input, or use a default. */ 
+
+  n = RUN_get_int_parameter("random_seed", &scalar_seed);
+
+  if (n == 0) {
+    info("[Default] Random number seed: %d\n", scalar_seed);
+  }
+  else {
+    info("[User   ] Random number seed: %d\n", scalar_seed);
+  }
+
+  /* Serial generator */
+
+  s_rng.ispare = 0;
+
+  s_rng.rstate[0] = scalar_seed;
+  s_rng.rstate[1] = 1;
+  s_rng.rstate[2] = 0;
+  s_rng.rstate[3] = 3;
+  s_rng.rstate[4] = 4;
+
+  /* Parallel generator */
+
+  p_rng.ispare = 0;
+
+  p_rng.rstate[0] = scalar_seed;
+  p_rng.rstate[1] = pe_size();
[…]

### 2012-07-18 branches/CUDA/tests/test_phi_ch.c

@@ -1,0 +1,230 @@
+/*****************************************************************************
+ *
+ *  test_phi_ch.c
+ *
+ *  Order parameter advection and the Cahn Hilliard equation.
+ *  NOT A COMPLETE TEST.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdio.h>
+#include <math.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "leesedwards.h"
+#include "lattice.h"
+#include "site_map.h"
+
+#include "phi.h"
+#include "phi_gradients.h"
+#include "gradient_3d_7pt_fluid.h"
+#include "phi_cahn_hilliard.h"
+#include "symmetric.h"
+
+#include "util.h"
+
+static void test_advection(void);
+static void test_set_velocity(const double *);
+static void test_set_drop(void);
+static void test_drop_difference(void);
+
+int main (int argc, char ** argv) {
+
+  MPI_Init(&argc, &argv);
+  pe_init();
+  coords_nhalo_set(2);
+  coords_init();
+  le_init();
+  site_map_init();
+  phi_nop_set(1);
+  phi_init();
+  phi_gradients_init();
+  hydrodynamics_init();
+  gradient_3d_7pt_fluid_init();
+
+  test_advection();
+
+  phi_finish();
+  pe_finalise();
+  MPI_Finalize();
+
+  return 0;
+}
+
+/*****************************************************************************
+ *
+ *  test_advection
+ *
+ *  Advect a droplet at constant velocity once across the lattice.
+ *
+ *****************************************************************************/
+
+void test_advection() {
+
+  int n, ntmax;
+  double u[3];
+
+  u[X] = -0.25;
+  u[Y] = 0.25;
+  u[Z] = -0.25;
+  ntmax = -L(X)/u[X];
+  info("Time steps: %d\n", ntmax);
+
+  test_set_velocity(u);
+  test_set_drop();
+
+  /* Steps */
+
[…]

### 2012-08-02 papers/dsfdproc/dsfdproc.tex
Added source of electrokinetics proceeding
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1738 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[aps,pre,preprint,groupedaddress]{revtex4-1}
\bibliographystyle{apsrev4-1}

\usepackage[pdftex]{graphicx}
\usepackage{amsmath}
\newcommand{\e}[1]{\cdot10^{#1}}
\let\endtitlepage\relax

\begin{document}

\title{Report HPC-Europa2 Transnational Access Programme}

\author{Oliver Henrich$^1$ and Kevin Stratford$^2$}

\affiliation{
$^1$Centre for Computational Science, University College London, United Kingdom\\
$^2$Edinburgh Parallel Computing Centre, University of Edinburgh, United Kingdom
}

\date{\today}
\begin{abstract}
This is the intermim report of our joint visit from 12 April - 11 May 2012 to Barcelona Supercomputing Center and Prof Ignacio Pagonabarraga, Departament de Fisica Fonamental, Universitat de Barcelona, Spain.
\end{abstract}

\maketitle

\section{Week 1 (including period 12-20 April)}

We have introduced an object to hold the scalar electric potential,
and the charge densities for a general number of species. This
includes specification of diffusivities, valencies and unit charge.
The corresponding parallel halo swaps and I/O routines have been
added and subjected to unit tests.

In discussions with Pagonabarraga, we decided to prioritise the
implemenation of the the Poisson solver (successive overrelaxation,
SOR) so that more simple validation problems can be attempted as
soon as possible. The SOR solver has been implemented and tested
against an exact matrix solver for a one-dimensional problem in
both serial and parallel. The SOR solver assumes a spatially
uniform permittivity. We will extend this to a spatially varying
permittivity at a later stage.

A start has been made on the finite difference solution of the
Nernst Planck equation (advection/diffusion equation for charged
species). This will be completed and tested next week. This puts
us slightly ahead of schedule cf the original work plan.

Basic tests of the SOR solver have been already performed.
We have identified a number of other suitable test scenarios to 
verify the implemention of the Nernst-Planck equation in 
combination with the SOR solver.
We decided to test the Gouy-Chapman theory
for electric double layers in front of a charged wall \cite{Lyklema},
the liquid junction potential emerging between two electrolytes
of slighty different concentration and diffusivity of the charged species \cite{Mafe},
electro-osmotic flow in a slit pore  
\cite{Capuani, Rotenberg} and Debye-H\"uckel theory for charged 
colloidal particles and small enough potentials \cite{Lyklema}.

\section{Week 2 (23-27 April)}

We completed the numerical implementation of the Nernst-Planck
Equation which is solved via a simple finite difference method.
The solution includes advection (although there is no fluid
fluid fluid present in the test problems). We have included
no normal flux boundary conditions at solid fluid interfaces
to ensure the conservation of charged species. The solver
has been tested in both serial and in parallel.

A first validation test is a flat surface carrying a surface charge $\sigma$
with counterions and symmetic electrolyte. This is a one-dimensional, 
electroneutral problem of a diffusive electric double layer which has 
an analytical solution \cite{Lyklema}. 
The approximation for low electrostatic 
potentials reads

\begin{equation}\label{gouychapman}
\Psi(x) = \Psi_D \exp(-\kappa\, x)
\end{equation}
with $\kappa$ as inverse Debye length 
\begin{equation}
\kappa = l_D^{-1} = \sqrt{8\pi\, l_B \, I}.
\end{equation}
The Bjerrum length $l_B$ is given by 
\begin{equation}
l_B = \frac{\beta\,e^2}{4\pi\,\varepsilon}
\end{equation}

with $\beta^{-1}=k_B T$, $e$ as unit charge and $\varepsilon=\varepsilon_0\varepsilon_r$ 
as dielectric permittivity.
The parameter

\begin{equation}
I = \frac{1}{2}\sum_k z_k^2\; \rho_{B,k}
\end{equation}

is the ionic strength 
of the electrolyte with $z_k$ as valencies of species $k$
($z_\pm=\pm 1$ for simple symmetric electrolyte) and $\rho_{B,k}$ as 
bulk charge density of species $k$ far away from the wall.
The Stern potential $\Psi_D$ at the surface of the wall is related 
to the surface charge $\sigma$ via

\begin{eqnarray}
\Psi_D&=&\frac{2}{\beta \,e} \sinh^{-1}\left(-\sigma\,p\right)\\
\Psi_D&\simeq&\frac{2}{\beta \,e} \ln\left(-\sigma \, p + \sqrt{(\sigma\,p)^2+1}\right)\\
p &=& \frac{1}{\sqrt{8\, \varepsilon\, \beta^{-1} \,\rho_B}}.
\end{eqnarray}

The quantity $\rho_B=\rho_{B,+}=\rho_{B,-}$ is 
the average bulk charge density of the electrolyte.

We solved the Gouy-Chapman problem for a 
system consisting of $L_x \times L_y \times L_z=64\times4\times4$
lattice sites. Periodic boundary conditions were used at all sides 
and a no-flux boundary conditions was set at $L_x=1$ and $L_x=64$.

The parameters were chosen as follows (all in simulation units):
unit charge $e=1$, temperature $k_B T=\beta^{-1}=3.333\e{-5}$, valency $z_{\pm}=\pm1$, 
dielectric permittivity $\varepsilon=3.3\e{3}$, diffusivities 
of the species $D_{\pm}=1\e{-2}$ and initial densities $\rho_{0,\pm}=1\e{-2}$.
The entire system was electro-neutral and had a Bjerrum length $l_B=0.723$.

At first we tested the linear case applying a small positive surface charge
$\sigma=3.125\e{-2}$, which led to bulk charge density $\rho_{B,+}=\rho_{B,-}=1.044\e{-2}$, 
Debye length $l_D=2.295$ and surface potential $\Psi_D=2.136\e{-5}$.
The potential was initialised with zero and had a value $\Psi_c=-2.364\e{-6}$ 
in the centre of the system after equilibration which we subtracted in the 
following analysis.

We reduced the initial density of the electrolyte to $\rho_{0,\pm}=1\e{-3}$,
which resulted in bulk charge densities  
$\rho_{B,+}=1.298\e{-3}, \rho_{B,-}=1.370\e{-3}$, 
Debye length $l_D=6.420$, surface potential $\Psi_D=5.451\e{-5}$
and centre potential $\Psi_c=-1.256\e{-5}$.
The comparison with the approximate solution Eq. \ref{gouychapman} is shown 
in Fig. \ref{fig1}.

\begin{figure}[h!t]
\includegraphics[width=0.495\textwidth]{test1.pdf}
\includegraphics[width=0.495\textwidth]{test2.pdf}
\caption{Gouy-Chapman theory for electric double layers: The plots compare the simulation results for the electric potential with the analytical solution for $\rho_{0,\pm}=1\e{-2}$ (left) and $\rho_{0,\pm}=1\e{-3}$ (right).} 
\label{fig1} 
\end{figure}

For larger surface charges the low-potential assumption which led to Eq. \ref{gouychapman}
is no longer valid and the nonlinear nature of the Poisson-Boltzmann 
equation becomes evident.
Fig. \ref{fig2} shows the results for a surface charge $\sigma=9.375\e{-1}$
and electrolyte density $\rho_{0,\pm}=3\e{-3}$. We obtained for   
bulk charge densities $\rho_{B,+}=4.443\e{-3}$ and $\rho_{B,-}=4.461\e{-3}$, 
Debye length $l_D=3.514$, the surface potential $\Psi_D=2.267\e{-4}$
and the centre potential $\Psi_c=-3.395\e{-5}$.

\begin{figure}[h!t]
\includegraphics[width=0.495\textwidth]{test3.pdf}
\includegraphics[width=0.495\textwidth]{test3-rho.pdf}
\caption{Nonlinear regime: For larger surface charges the solution deviates from the low-potential approximation Eq. \ref{gouychapman} to the Gouy-Chapman theory. The righthand picture shows the charge distribution across the gap.} 
\label{fig2} 
\end{figure}

A number of minor bugs in the parallel implementation of the
SOR solver were resolved, and the unit tests updated appropriately.

Some effort has been used to create a comprehensive test suite
for the basic functionality involved in these routines. These
tests are valid both in serial and in parallel.

Work is now under way to couple the new Nernst-Planck and
Poisson routines with the main code. This includes work to
include charged colloids.

\section{Week 3 (30 April - 4 May)}

We have now coupled the Poisson solver and the Nernst Planck solver
to the main code and checked for correct operation. This has included
the run time intialisation of the electrokinetic-related quatities
from the input file. We have also integrated the parallel I/O for
the electrokinetic lattice quantities to the main code, again with
appropriate initialisation from input.

We proceeded with another test. The liquid junction potential 
is a charge separation process that 
occurs when electrolytes with slightly different concentrations
whose species have different diffusivities are brought into contact.
Charges from the regions of higher concentration diffuse   
into the parts with lower concentration. Due to the difference 
in diffusivity they migrate at different speeds, leaving parts of
the system charged. This leads to a build-up of a potential
which balances the diffusive flux.

After the initial build-up phase the potential decreases slowly 
again until the charge concentration has become homogeneous throughout 
the system. Both timescales of emergence and decay of the potential
can be separated by chosing a sufficiently large system size.

This problem allowed us to verify the correct temporal 
behaviour of the Nernst-Planck equation solver by resolving the transient 
dynamics without having to account for advective terms.

For simplicity we considered systems of size 
$L_x\times L_y\times Lz=128\times4\times4$ and 
$L_x\times L_y\times Lz=256\times4\times4$ with
periodic boundary conditions at either end.
The two halfs were electroneutral and had ionic concentrations 
$\rho_{L,\pm}=\rho_{0,\pm} + \delta\rho$ and 
$\rho_{R,\pm}=\rho_{0,\pm} - \delta\rho$ 
with $\rho_{0,\pm}=1\e{-2}$ and $\delta\rho = 0.01$.

The potential difference between both sides during the build-up 
obeys approximately

\begin{equation}
\Delta\Psi(t)\simeq\Delta\Psi_P \left\{1-\exp\left(-\frac{t}{\tau_e}\right)\right\}\\
\end{equation}

with

\begin{equation}
\Delta\Psi_P=\frac{(D_+ D_-)}{\beta e (D_+ + D_-)} \frac{\delta\rho}{\rho_0}\\ 
\end{equation}

as saturation value of the potential difference.
The saturation time scale is given by

\begin{equation}
\tau_e=\frac{\varepsilon}{\beta \, e^2 (D_+ + D_-) \rho_0}.
\end{equation}

A more exact solution can be derived in the limit $l_D/L_x<<1, N\to\infty$:

\begin{eqnarray}
\Delta\Psi(t)&=&\Delta\Psi_P \left\{1-\exp\left(-\frac{t}{\tau_e}\right)\right\}\frac{4}{\pi}\left\{\sum_{m=1}^N \frac{\sin^3(m\pi/2)}{m} \exp\left(-\frac{ m^2\, t}{\tau_d}\right)\right\}\\
\tau_d&=&\frac{L^2}{2\pi^2 (D_+ + D_-)}.\label{taud}
\end{eqnarray}

It contains as well the dependence on the decay timescale $\tau_d$.
Only odd indices $m$ contribute to the sum:

\begin{eqnarray}
\sum_{m=1}^N \frac{\sin^3(m\pi/2)}{m} \exp\left(-\frac{m^2\, t}{\tau_d}\right)&=&\nonumber\\
&&\hspace*{-4cm} \exp\left(-\frac{t}{\tau_d}\right)-\frac{1}{3} \exp\left(-\frac{9 t}{\tau_d}\right)+\frac{1}{5}\exp\left(-\frac{25t}{\tau_d}\right)-\frac{1}{9}\exp\left(-\frac{81 t}{\tau_d}\right)+\ldots
\end{eqnarray}

A complete discussion of the solution can be found in \cite{Mafe}. 
There, the upper limit of significant modes has been also estimated as $N_{max} = L/\pi l_D$.
Note the factor 2 difference between Eq. \ref{taud} and the corresponding expression in \cite{Mafe}.
\begin{figure}[h!t]
\includegraphics[width=0.495\textwidth]{test_lj_zoom1.pdf}
\includegraphics[width=0.495\textwidth]{test_lj_zoom3.pdf}
\caption{Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue). The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$. Deviations can be seen which are due
to the finite size of our system and the approximate nature of the analytical solution.} 
\label{fig3} 
\end{figure}

The following parameters were used:
dielectric permittivity $\epsilon=3.3\e{3}$, temperature $\beta^{-1}=3.333\e{-5}$, unit charge $e=1$, valency $z_\pm=\pm1$, diffusivities $D_+=0.0125$ and $D_-=0.0075$.
We obtained
$\Delta\Psi_P=1.6667\e{-7}$, $\tau_e=550$, $\tau_d=41501.2\, (L_x=128)$, $\tau_d=166004.6\, (L_x=256)$ and $\tau_d=664018.5 (L_x=512)$.
The results for the potential difference over time are shown in Fig. \ref{fig3}.

\begin{figure}[h!t]
\includegraphics[width=0.9\textwidth]{test_lj_decay1.pdf}
\caption{Rescaled plot of the decay: Times have been Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue). The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$. Deviations can be seen which are due
to the finite size of our system and the approximate nature of the analytical solution.} 
\label{fig4} 
\end{figure}

Fig. \ref{fig4} shows results with times rescaled to the decay 
time scale $\tau_d$ (cf. Eq. \ref{taud}). Obviously the 
deviations we observe are not due to the limited system size 
and have a more systematic origin.

The curves coincide if 
the theoretic limit for $\tau_d$ is rescaled by a factor $1.067$,
suggesting the effective system length for this sort of setup is
actually about 3\% larger than the numerical value.

A reason for this might be that the analytical solution was gained
for an infinitely large system with constant charge concentrations,
vanishing currents at both ends and finite diffusive zone of size $L_x$.
In our situation the entire system is within the diffusive zone.
This may lead to smaller effective diffusivities or larger effective
system sizes.
Interestingly, all runs with solid walls at both ends resulted in 
oscillatory behaviour and an effective system size of $2L_x$.

We have run some simple scaling tests for the electrokinetic problem
with zero fluid velocity. A system of $128^3$ lattice sites was used
as representative of a small production system. The code was run on
the Mare Nostrum machine in Barcelona for a small number of time
steps to assess performance on up to 512 MPI tasks. The results are
shown in Fig. \ref{fig5}.

While the scaling for the Nernst Planck part of
the calulcation are reasonable, it can be seen that the SOR routine
used to solve the Poisson equation is not really acceptable. This
was not unexpected. Some improvements might be possible, but the
algorithm is essentailly unsuitable for large systems. We will
investigate the use of an alternative (e.g., multigrid), but this
is beyond the scope of this HPC Europa visit.

In discussion with Pagonabarraga, we have also made some progress
in deciding how to modify a number of other parts of the code to
allow for the presence of the electric potential. These include
the free energy calculation, which can be generalised to compute
both electric contributions and coupling between potential and
compositional order parameter. A flexible solution may require
some refactoring.

We have also implemented a fixed external electric field in the
Nernst Planck equation. This will be tested during the last week
of the visit.

\begin{figure}[h!t]
\includegraphics[width=0.9\textwidth]{scaling.pdf}
\caption{Scaling of a system of $128^3$ lattice sites on up to 512 MPI tasks. 
While the scaling for the Nernst Planck part of the calulcation are reasonable, 
it can be seen that the SOR routine used to solve the Poisson equation deviates
strongly from ideal behaviour.} 
\label{fig5} 
\end{figure}

\section{Week 4 (7 May - 11 May)}

To test the implementation with all couplings to external and 
internal forces we consider a forced charged fluid in a slit
of size $L_z$. An electrostatic field $E_{||}$ is allied
parallel to the walls. The entire system is electroneutral with 
each wall having the surface charge density $\sigma$ 
and compensationg counterions with total charge $2 \sigma A_{wall}$
in the fluid.

In equilibrium the charge density at a distance $x$ from the wall obeys

\begin{equation}
\rho(x)=\frac{\rho_0}{\cos^2(K\,x)}
\end{equation}

with

\begin{equation}
\rho_0=K^2/2\pi l_B
\end{equation}

and

\begin{eqnarray}
K \,L_x \tan\left(\frac{K\, L_x}{2}\right)&=&\pi\, l_B\, L_x\, 2\sigma\label{kex} \\
K \,L_x&\simeq&\sqrt{4\pi \,l_B\,L_x\,2\sigma}\label{klin}.
\end{eqnarray}

The liniarised version Eq. \ref{klin} has only a limited range of applicabilty.
We solved Eq. \ref{kex} numerically and found solutions 
$K=0.01959\; (\sigma=0.003125)$ and $K=0.03311\; (\sigma=0.00125)$, 
which is reasonably far away from the 
theoretical limit $K_{max}=\pi/L_x$ set by the tangent.

Note the factor 2 difference on the lhs of Eq. \ref{kex} with respect 
to \cite{Capuani, Rotenberg}. There is also a factor $L_x$ missing on 
the rhs of Eq. \ref{klin}.

The steady state velocity of the fluid can be derived from the 
force balance of the gradient of the stresses and the electrostatic
forces:

\begin{eqnarray}
v_y(x)&=&\hat{v} \ln\left(\frac{\cos(K\,x)}{\cos(K\,L_x/2)}\right)\label{vy}\\
\hat{v}&=&\frac{e \,E_{||}\rho_o}{\eta\, K^2}=\frac{e \,E_{||}}{2\pi\eta l_B}\label{vhat}
\end{eqnarray}

The result for two different charge densities is shown in Fig. \ref{fig6}.
The accuracy is acceptable with deviations for high surface 
charged potentially being caused by the chosen discretisation or 
by the numerical solution of Eq. \ref{kex} approaching the limit 
of $\pi/L_z\simeq0.049$.

\begin{figure}[h!t]
\includegraphics[width=0.9\textwidth]{test_eo.pdf}
\caption{Steady state flow profile across a slit of width $L_x=63$ for an applied field of magnitude $e \beta E L_x=1.89$ for two different charge densities $\sigma=0.003125$ (red) and $\sigma=0.0125$ (blue), Bjerrum length $l_B=0.7234$, viscosity $\eta=0.1$ and unit charge $e=1$. The dashed black lines correspond to theoretical prediction according to Eqs. \ref{vy} and \ref{vhat}.} 
\label{fig6} 
\end{figure}

Finally, we have tested the implementation for a single fixed
colloid and compared the result with Debye-H\"uckel theory. A result
is shown in Fig.~\ref{fig7}.
We used the following parametrisation:

$L_x \times L_y \times L_z=64\times64\times64$,
$D_+=D_-=0.01$,
$e=1$, 
$z_{\pm}=1$,
$\beta^{-1}=3.333\e{-5}$,
$\varepsilon=3.3\e{3}$,
$l_B=0.723$.

For a central and fixed colloid of radius $R_c=7.5$ carrying a positive unit charge
$q_{c,+}=1.0, q_{c,-}=0$ we obtained 
$\rho_{c,+}=5.58\e{-4}$,
$\rho_{el}=\rho_{B,\pm}=1\e{-2}$, 
$\Psi_D=8.836\e{-7}$ for $2\,R_c=16$.

For a higher larger positive charge $q_{c,+}=4.0$ we got
$\rho_{c,+}=2.23\e{-3}$
$\rho_{el}=\rho_{B,\pm}=5\e{-3}$ 
$\Psi_D=4.993\e{-6}$ for $2\,R_c=16$.

\begin{figure}[h!t]
\includegraphics[angle=-90,width=0.9\textwidth]{test_dh1.pdf}\\
\includegraphics[angle=-90,width=0.9\textwidth]{test_dh2.pdf}
\caption{Debye-H\"uckel theory for positively charged colloid: the picture
shows a cut through the center of a single colloid in a peridoic system.
The potential is shown in red, the Stern potential in blue, and the negative 
charge density in green. The colloid is in the center.}
\label{fig7}
\end{figure}

\clearpage
\begin{thebibliography}{99}

\bibitem{Lyklema} J. Lyklema {\em Fundamentals of Interface and Colloid Science} Academic Press (1995).
\bibitem{Mafe} S. Maf\'e, J.A. Manzanares, J. Pellicer, {\em J. Electroanal. Chem.} {\bf 241}, 57-77 (1988).
\bibitem{Capuani} F. Capuani, I. Pagonabarraga, D. Frenkel, {\em J. Chem. Phys.} {\bf 121}, 973-986 (2004).
\bibitem{Rotenberg} B. Rotenberg, I. Pagonabarraga, D. Frenkel, {\em Farad. Discuss.} {\bf 144}, 223-243 (2010).

\end{thebibliography}

\end{document}

### 2012-08-08 trunk/papers/bprheo/bprheo.tex
Added new pictures and removed others
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1744 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-08-19 papers/bprheo/bprheo.tex
@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,pre,twocolumn,groupedaddress]{revtex4-1}
+\documentclass[aps,pre,onecolumn,groupedaddress]{revtex4-1}
 %\documentclass[aps,pre,preprint,superscriptaddress]{revtex4-1}
@@
 
+\newcommand{\e}[1]{\times10^{#1}}
+
 \begin{document}
@@
 %Title of paper
-\title{Bulk Flow Behaviour of Cubic Blue Phases}
+\title{Rheology of Cubic Blue Phases}
 
@@
 % \affiliation can be followed by \email, \homepage, \thanks as well.
-\author{O. Henrich$^{1,3}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}
-\affiliation{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
+\author{O. Henrich$^{1,2}$, K. Stratford$^1$, , P.V. Coveney$^2$, M.E. Cates $^3$ and D. Marenduzzo$^3$}
+\affiliation{$^1$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^2$ Centre for Computational Science, University College London, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
 
@@
 
-In the following we report results on the bulk flow behaviour of BPI and BPII for thermodynamic states characterized by $\tau=-0.5$ and $\kappa=1.0$ (BPI) and $\kappa=2.0$ (BPII), respectively.
+In the following we report our results of simulations of the bulk flow behaviour of cubic BPI and BPI.
+The thermodynamic states characterized by $\tau=-0.5$ and $\kappa=1.0$ (BPI) and $\kappa=2.0$ (BPII), respectively.
 For these parameters both blue phases are equilibrium phases in the quiescent state.
-We have also performed simulations at higher temperatures and chiralities and for metastable configurations, which led to very similar results.
-Quantitative differences occured only in that sense that the disclination network dissolved more easily for a particular flow rate if the system was closer to the cholesteric-isotropic phase boundary and where the order parameter takes lower values and free energy density is higher.\\
-The tensor order parameter was initialized with the high-chirality solution of Eq. \ref{eqn1}, which has the correct topology of the actual blue phase.
-The initial configurations were equilibrated for $5000$ LB time steps, where we allowed for optimization of the unit cell size \cite{Alexander:2006}, a feature that is often referred to as 'redshift'.
-After the equilibration sequence the unit cell size was fixed and the shear flow was switched on.
-For BPII we used a pitch length of 32 lattice sites, but in case of BPI it was necessary to increase the resolution to 64 sites per pitch length to get accurate results of the flow induced transformations in the disclination network.
+Simulations of metastable states at higher temperatures and different chiralities showed the same generic behaviour.
+The only observable quantitative difference was that networks closer to the cholesteric-isotropic phase boundary with typically higher free energies and lower average order parameter proved more unstable with respect to the external driving forces.\\
+The order parameter was initialized with an analytical solution of Eq. \ref{eqn1} in the limit of infinite chirality which has already the correct topology of the blue phase.
+These initial configurations were then equilibrated for $5000$ LB time steps, where we also allowed for the a priori unknown 'redshift' of the pitch, the optimization of the unit cell size (cf.
+\cite{Alexander:2006}).
+After equilibration the unit cell size was kept constant throughout the rest of the runs with the shear flow switched on.
+In order to guarantee that complex configurational changes in the blue phase networks are properly resolved we opted for a unit cell size of 16 lattice sites in case of BPII and increased the unit cell size to 32 sites in case of BPI.
+Runs with higher resolution confirmed that this is an appropriate minimal resolution to cover all kinematic details of the network.
+
 In order to impose a flow profile we used Lees-Edwards boundary conditions \cite{Wagner:2002}, whereas the top (bottom) half of planes moved in positive (negative) y-direction.
@@
 We start the discussion with BPII, because its flow behaviour is simpler and probably more easily anticipated.
-Fig \ref{fig1} shows the disclination network undergoing a flow-induced transformation.
+Fig \ref{fig11} shows the disclination network undergoing a flow-induced transformation.
 The periodically recurring structure remains very regular and apart from the quasi-affine transformation relatively close to its quiescent appearance with individual disclination lines breaking up and reconnecting further downstream.
@@
 \includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\cdot 10^{-4}$.
-The horizontal y-direction is the flow direction whereas the velocity gradient is oriented along the vertical x-direction.}
-\label{fig1}
+\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$.
+The velocity gradient is oriented along the vertical direction, whereas the horizontal direction is the flow direction.
+The flow takes place in such as way that network moves to the right in the upper half and to the left in the lower half.}
+\label{bp2-med}
 \end{figure*}
@@
 This feature is most clearly observed when the structure is viewed along the flow direction.
-Fig. \ref{fig2} depicts a sequence of close-ups of the disclination network as it travels in negative z-direction, i.e. from the foreground towards the background of the picture. 
-The velocity $v_z$, represented by the colour code, exhibits a recurring pattern of positive and negative values.
-Peak velocities occur close to disclination lines where the molecular order is suppressed or in regions where two disclination lines approach each other and the network is about to break up or reform.
 
-\begin{figure*}[h]
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-149k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-160k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-167k_run902.png}\\
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-173k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-182k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-192k_run902.png}\\
-\caption{A section of $2\times2\times1$ unit cells of BPII seen along the direction of vorticity.
-The colour code gives the magnitude and direction of the velocity component $v_z$ perpendicular to the plane.}
-\label{fig2}
-\end{figure*}
-
-Plotting only x- and z-component of the local velocity allows to visualize the spatial dependence of the secondary flow in the blue phase with respect to the disclination network.
 Fig. \ref{fig3} shows characteristic regions of upward and downward 'flow', which are normally superimposed by the primary shear flow in y-direction.
 Remarkably, the secondary flow is directly connected to the sense of the underlying cholesteric helix as changing from left- to right-handed helix also inverts the sense of motion of the network.
-%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{figure*}[h]
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run902.png}
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run903.png}
-\caption{Secondary flow in BPII visualized by means of the velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity of the cholesteric helix.
-The images depict an individual snapshot of the periodically recurring sequence in the steady state.}
-\label{fig3}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
+\caption{Velocity patterns in BPII for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
+The colour code gives the magnitude and sign of the z-component.
+The vertical and horizontal direction are the gradient and vorticity direction, respectivelty.
+Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
+\label{bp2-velo}
 \end{figure*}
@@
 For all apart from the highest flow rate it oscillates sinusoidally, exhibiting symmetric minima and maxima and decreasing mean values, indicating slightly shear-thinning behaviour.
-The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig1}.
+The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig11}.
 However, due to the motion of the network in z-direction it turns out that it takes four periods altogether for the network to move one unit cell along the direction of vorticity and to complete one cycle.
@@
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{stress_bp2.pdf}
-\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPII in the steady state. }
-\label{fig4}
+\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
+\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ of BPII over time (left) and strain (right). The inset shows the flow curve $\sigma_{chem}(\dot{\gamma})$.}
+\label{bp2-rheo}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{dir-op-y_225k_run949.png}
-\caption{BPII at high shear rate: director field and the disclination planes seen along the flow direction.
-The BPII breaks up and flows in helical layers separated from each other by a disclination plane}
-\label{fig5}
+\caption{BPII at low shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom).
+The configuration is a travelling helical wave along the vorticity direction.
+The configuration is translationally invariant in flow and in gradient direction.
+The colour code indicates the orientation in x-y direction.}
+\label{bp2-low}
 \end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
+\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
+\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
+\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
+\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom).
+The configuration is a travelling helical wave along the vorticity direction.
+The configuration is translationally invariant in flow and in gradient direction.
+The colour code indicates the orientation in x-y direction.}
+\label{bp2-high}
+\end{figure}
+
+\clearpage
 
@@
 This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
-Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig1}, but this time seen along the direction of flow.
+Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig11}, but this time seen along the direction of flow.
 A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
@@
 \includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
-\caption{Disclination network of BPI seen along the flow direction in steady shear flow at $\dot{\gamma}=1.56\cdot 10^{-4}$, showing the downward movement of the disclination network in the direction of vorticity.}
-\caption{\label{fig6}}
+\caption{Disclination network of BPI in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$.
+The horizontal direction is the vorticity direction, whereas the velocity gradient is oriented along the vertical direction.}
+\label{bp1-med}
 \end{figure*}
@@
 \begin{figure*}[h]
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run914.png}
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run922.png}
-\caption{Secondary flow in BPI: velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity.}
-\label{fig7}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
+The colour code gives the magnitude and sign of the z-component.
+The vertical and horizontal direction are the gradient and vorticity direction, respectivelty.
+Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
+\label{bp1-velo}
 \end{figure*}
@@
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{stress_bp1.pdf}
-\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPI in the steady state.}
-\label{fig8}
+\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
+\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\sigma_{chem}(\dot{\gamma})$.}
+\label{bp1-rheo}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{spectrum_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
 \caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
-\label{fig9}
+\label{bp1-spec}
 \end{figure}
@@
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{dir-op-z_395k_run916.png}
-\includegraphics[width=0.5\textwidth]{dir-op-x_395k_run916.png}
-\caption{BPI at high shear rate: director field and the disclination planes seen along the direction of voritcity and the gradient direction.
-BPI forms rolls with $\pi/2$ disclinations that are advected with the general flow.}
-\label{fig10}
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-600k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-775k_run1115.png}
+\caption{Disclination network of BPI at low shear flow:}
+\label{bp1-low}
+\end{figure}
+
+\begin{figure}[h]
+\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
+\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
+\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
+\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
+The colour code gives the magnitude of the in-plane component of the director field.
+The data is translationally invariant along the vorticity direction.}
+\label{bp1-high}
 \end{figure}
@@
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge$200000 for BPI and t$\ge$30000 in case of BPII.
+\caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
 \label{table1}}

### 2012-08-10 trunk/papers/bprheo/bprheo.tex
Back up bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1745 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,pre,onecolumn,groupedaddress]{revtex4-1}
-%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4-1}
-%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
+\documentclass[aps,pre,reprint,superscriptaddress]{revtex4}
+%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
+%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}
 
@@
 % \affiliation can be followed by \email, \homepage, \thanks as well.
-\author{O. Henrich$^{1,2}$, K. Stratford$^1$, , P.V. Coveney$^2$, M.E. Cates $^3$ and D. Marenduzzo$^3$}
-\affiliation{$^1$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^2$ Centre for Computational Science, University College London, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
+\author{O. Henrich}
+\email[Corresponding author: ]{o.henrich@ucl.ac.uk}
+\affiliation{Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+\affiliation{Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}
 
-%\email[]{}
+\author{K. Stratford}
+\affiliation{Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+
+\author{P.V. Coveney}
+\affiliation{Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}
+
+\author{M.E. Cates}
+\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+
+\author{D. Marenduzzo}
+\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+
 %\homepage[]{Your web page}
@@
 \begin{abstract}
+bla
 \end{abstract}
@@
 % insert suggested PACS numbers in braces on next line
-\pacs{}
+\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
 % insert suggested keywords - APS authors don't need to do this
@@
 %\section{\label{}}
+
+Cholesterics are liquid crystals in which the local nematic director field spontaneously 
+twists in thermodynamic equilibrium~\cite{deGennes}. The preferred configuration close 
+to the isotropic boundary features twist around two perpendicular axes, as opposed to 
+just one axis in the regular cholesteric state, and the corresponding deformation is 
+denoted a ``double-twist cylinder''.
+As it is topologically impossible to cover continuously 3D space with double-twist 
+cylinders, defects arise. These organise into a variety of regular periodic lattices, 
+giving rise to the so-called cubic blue phases (BPs)~\cite{Grebel:1984,Wright:1989}. 
+There are two experimentally observed cubic blue phases, BPI and BPII (a third, BPIII, 
+is thought to be amorphous~\cite{Henrich:2011a}).
+
+BPs have been long considered as purely of academic interest due to their very narrow 
+range of stability. This view has changed since the creation of polymer-stabilised and 
+temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the 
+possibility of novel applications.
+During the last few years considerable progress has been achieved regarding the behaviour 
+of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under 
+external fields ~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
+and in the presence of colloidal particles~\cite{Ravnik:2011a}.
+The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. 
+However, our understanding of their dynamical behaviour under flow remains
+very limited. The aim of this work is to address this issue by studying,
+for the first time, the response of confined BP samples to a shear flow.
+
+Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
+For example, if a cholesteric helix is subjected to a Poiseuille flow along
+the helical axis, small pressure differences drive flow mainly through
+``permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
+In the permeation mode the liquid crystal flows while leaving the director
+field virtually unchanged, which leads to high dissipation and large
+viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
+shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
+showed the importance of the boundary conditions in determining the apparent viscosity of the fluid. 
+They also found that a strong secondary flow appears.
+Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
+the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
+lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and 
+leaves a flow-induced nematic phase.
+Rey also studied cholesterics subjected to both steady flow and low frequency
+small amplitude oscillatory shear for different helix orientations
+~\cite{Rey:2000, Rey:2002}. He found that splay/bend/twist deformations were
+excited when the helix was aligned along the flow direction; splay/bend
+deformation occurred when the helix was aligned along the velocity gradient;
+but only twist deformations
+appeared when the helix was aligned along the vorticity axis.
+
+Dupuis et al.~\cite{Dupuis:2005} performed the first numerical investigation of BP rheology 
+in Poiseuille flow, starting from equilibrium structures of BPI and BPII and a periodic 
+array of doubly twisted cylinders.
+Under small forcing, the network opposed the flow giving rise to a significant 
+increase in apparent viscosity.
+Upon increasing the forcing they found clear evidence of shear thinning.
+In the crossover region they predicted a novel oscillatory regime where the network 
+continuously breaks and reforms as portions of the disclinations in the center of the channel 
+move to neighboring cells and relink with the parts of the network left behind by the flow. 
+Compared with the cholesteric case, the viscosity still decreases with forcing 
+(the system shear thins) but much less than
+for cholesterics in the permeation mode, which is in agreement with experiments
+~\cite{Zapotocky:1999, Ramos:2002}.
+
 \section{Model and Methods}
 
-Our approach is based on a well-established model for hydrodynamics of cholesteric liquid crystals \cite{Beris:1994,Olmsted:1999} that describes the order structure of the liquid crystal by a traceless, symmetric tensor order parameter ${\bf Q}({\bf r})$.
-In uniaxial approximation the tensor order paramter is given by $Q_{\alpha \beta}=\langle n_\alpha n_\beta - 1/3\; \delta_{\alpha\beta}\rangle$, whereas ${\bf n}$ represents the local orientation of the liquid crystal molecules and brackets indicate mesoscopic averaging over a large number individual molecules.
-Its largest eigenvalue $\frac{2}{3}q$ with $0<q<1$ characterizes the degree of order around the corresponding axis. 
-The equilibrium properties of the system are determined by a Landau-de Gennes free energy density $f$ and the functional ${\cal F}[Q]=\int d^3{\bf r} f({\bf Q})$.
-The free energy density consists of two different parts, namely a bulk contribution $f_b$ and a gradient term $f_g$.
+Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
+cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
+in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
+In the uniaxial approximation, the order parameter is given by
+$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
+with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
+order. More generally,
+the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
+characterizes the local degree of orientational order.
+The thermodynamic properties of the liquid crystal are determined by a free energy
+${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
+\begin{eqnarray}
+f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
+f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
+\end{eqnarray}
+The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
+The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
+we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
+The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
+The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.
 
-\begin{eqnarray}
-f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2\nonumber\\
-f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2\label{eqn1}
-\end{eqnarray}
-
-The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ and features a the first-order transition from the isotropic to the ordered phase at $\gamma>2.7$.
-The second part comprises the elastic contributions to the free energy due to splay-, bend- and twist-deformations and assume for simplicity that they have all the same elastic constant $K$.
-The pitch length $p$ of the cholesteric liquid crystal is related to the wavervector $q_0=2\pi/p$.
-In order to specify a thermodynamic state it is useful to introduce two dimensionless quantities, the effective temperature $\tau$ and chirality $\kappa$.
-They are given by
-
+A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
+which are given by
 \begin{eqnarray}
 \tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
-\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber,
+\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
 \end{eqnarray}
-
-whereas the chirality qqnatifies the ratio between gradient and bulk free energy.\\
-The equation of motion of the tensor order parameter reads
-
-\begin{equation}\label{}
-\left(\partial_t+ u_\alpha \partial_\alpha \right){\bf Q} - {\bf S}({\bf W},{\bf Q}) = \Gamma {\bf H}.
+The dynamical evolution of the order parameter is given by the equation 
+\begin{equation}
+\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
 \label{eqn2}
 \end{equation}
-
-The first term on the lefthand side of Eq.(\ref{eqn2}) is a material derivative, which describes the rate of change of a quantity moving along with the flow.
-The second term accounts for the rate of change due to velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$:
-
+The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
+The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
 \begin{eqnarray}
-{\bf S}({\bf W}, {\bf Q}) &=& (\xi {\bf A} + {\boldsymbol \Omega})({\bf Q}+\frac{\bf I}{3})\nonumber\\
-& &\hspace*{-1.5cm}+ ({\bf Q}+\frac{\bf I}{3})(\xi {\bf A}  - {\boldsymbol \Omega})-2 \xi ({\bf Q}+\frac{\bf I}{3})Tr({\bf Q W}),
+{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
+& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
+\mathrm{Tr}({\mathbf Q W}),
 \label{eqn3}
 \end{eqnarray}
-
-where $Tr$ denotes the trace operation and ${\bf A}=({\bf W}+{\bf W}^T)/2$ and ${\boldsymbol \Omega}=({\bf W}-{\bf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradients, respectively.
-The molecular details of the liquid crystal such as the aspect ratio of its mesogens are embodied in the constant $\xi$, also know as the tumbling parameter.
-The righthand side relates the rate of change to the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\bf Q$.
-
+where $\mathrm{Tr}$ denotes the tensorial trace, while 
+${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
+${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
+is a constant depending on the molecular details of the liquid crystal.
+Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by 
+setting $\xi=0.7$ in our simulations.
+${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
 \begin{equation}
-{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\, Tr\left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
+{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
+\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
 \label{eqn4}
 \end{equation}
+The rotational diffusion constant $\Gamma$ in Eq.~\ref{eqn2} is proportional
+to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
+\cite{deGennes}.
 
-The rotational diffusion constant $\Gamma$ is proportional to the inverse of the rotational viscosity $\gamma_1=2 q^2/\Gamma$.\\
-The time evolution of density and velocity are goverend by the continuity equation Eq. (\ref{eqn5})
-
+The time evolution of the fluid density and velocity are respectively governed
+by the continuity equation
+$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
+the following Navier-Stokes equation:
 \begin{eqnarray}
-\partial_t \rho + \partial_\alpha (\rho u_\alpha)=0
-\label{eqn5}
-\end{eqnarray}
-
-and the Navier-Stokes equation Eq. (\ref{eqn6})
-
-\begin{eqnarray}
-\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha&=&\partial_\beta \Pi_{\alpha \beta}\nonumber\\
-&+&\hspace*{-2cm}\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha +(1+3\frac{P_0}{\partial \rho} )\partial_\mu u_\mu \delta_{\alpha \beta})\}. 
+\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
+= \partial_\beta \Pi_{\alpha \beta}
++\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
++(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
 \label{eqn6}
 \end{eqnarray}
-
-At low flow rates the fluid can be considered as incompressible, so that the last term on the righthand side of Eq. (\ref{eqn6}) can be safely neglected.
-The pressure tensor reads explicitely
-
+The final term in Eq.~\ref{eqn6} arises from the Chapman-Enskog expansion
+of the LB equations \cite{Denniston:2001}.
+At low flow rates the fluid can be considered as incompressible, so that the
+last term on the right hand side of Eq.~\ref{eqn6} remains small.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
+The pressure tensor reads explicitly
 \begin{eqnarray}
-\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
+\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
+-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
 &+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
@@
 \end{eqnarray}
+In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
+scalar pressure which, in a system at rest, is constant to a very good
+approximation.
 
-It comprises the terms that give rise to an additional stress in the ordered state.
-In the isotropic state when ${\bf Q}\equiv 0$ Eq. (\ref{eqn7}) is reduced to the scalar pressure, which is constant to a very good approximation.\\ 
-The system of coupled partial differential equations is solved by means of a hybrid scheme \cite{Denniston:2004, Marenduzzo:2008} that uses a lattice Boltzmann algorithm with predictor-corrector scheme for the hydrodynamic equations (\ref{eqn5}) and (\ref{eqn6}) and a finite difference scheme for the tensor order parameter equation (\ref{eqn2}).
+The system of coupled partial differential equations Eqs.~\ref{eqn2}
+and~\ref{eqn6} is solved by means of a
+hybrid scheme~\cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
+with predictor-corrector scheme for the continuity equation and
+Eq.~\ref{eqn6}, and a finite difference scheme for the equation of motion of
+the tensor order parameter Eq.\ref{eqn2}. More details on the algorithm can
+be found in~\cite{Denniston:2001, Denniston:2004}.
 
@@
 
-In the following we report our results of simulations of the bulk flow behaviour of cubic BPI and BPI.
-The thermodynamic states characterized by $\tau=-0.5$ and $\kappa=1.0$ (BPI) and $\kappa=2.0$ (BPII), respectively.
-For these parameters both blue phases are equilibrium phases in the quiescent state.
-Simulations of metastable states at higher temperatures and different chiralities showed the same generic behaviour.
-The only observable quantitative difference was that networks closer to the cholesteric-isotropic phase boundary with typically higher free energies and lower average order parameter proved more unstable with respect to the external driving forces.\\
-The order parameter was initialized with an analytical solution of Eq. \ref{eqn1} in the limit of infinite chirality which has already the correct topology of the blue phase.
-These initial configurations were then equilibrated for $5000$ LB time steps, where we also allowed for the a priori unknown 'redshift' of the pitch, the optimization of the unit cell size (cf.
-\cite{Alexander:2006}).
-After equilibration the unit cell size was kept constant throughout the rest of the runs with the shear flow switched on.
-In order to guarantee that complex configurational changes in the blue phase networks are properly resolved we opted for a unit cell size of 16 lattice sites in case of BPII and increased the unit cell size to 32 sites in case of BPI.
-Runs with higher resolution confirmed that this is an appropriate minimal resolution to cover all kinematic details of the network.
+In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
+phases BPI and BPII. The thermodynamic states are characterised by an effective temperature 
+$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy terms.
+We opted for states sufficiently far away from the chiral-nematic-isotropic 
+where both quiescent blue phases are at equilibrium, namely
+$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
+Simulations at metastable states at higher and lower temperatures and different chiralities showed 
+no differences in their generic flow behaviour.
+However, states closer to the phase boundary featuring higher free energy densities and 
+smaller average order parameters show a tendency to break up more easily under external driving forces.
 
-In order to impose a flow profile we used Lees-Edwards boundary conditions \cite{Wagner:2002}, whereas the top (bottom) half of planes moved in positive (negative) y-direction.
-We applied a correction for spurious momentum currents that emerge due to interpolations accross the Lees-Edwards planes \cite{Henrich:2011b}.
-The imposed flow rates ranged over two orders of magnitude from about $1.25\cdot 10^{-6}$ to $6\cdot10^{-4}$ in LB units.\\
-In the visualizations that follow the orientation of the principal coordinate axes in x-, y- and z-direction always coincides with the direction of the velocity gradient direction, the flow and the vorticity, respectively.
-Disclination lines are plotted as isosurface of the scalar order parameter $q$
+As is common in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with 
+analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, 
+and equilibrated these (for 5000 LB timesteps) prior to starting the shear. 
+During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
+After equilibration the redshift parameter was kept constant throughout the rest of the simulation.
+We chose a unit cell size of 16 lattice sites for BPII and 32 sites in case of BPI. 
+Runs with higher resolution confirmed that this was an appropriate minimal resolution to cover 
+the complex conformational changes and kinematic details of the blue phase networks.
+
+Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
+the top (bottom) half of the planes moving in positive (negative) x-direction.
+We applied a correction for spurious momentum currents that emerge due to interpolations accross 
+the Lees-Edwards planes \cite{Henrich:2012a}.
+The shear rates were varied over more than two orders of magnitude from about 
+$2.44\cdot 10^{-6}$ to $6.3\cdot10^{-4}$ in LB units.
+The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
+along the y- and z-axis.
+Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
+are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
+The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+is equal to about $10^8$~Pa.
+More details about the conversion from LBU to SI units can be found in
+\cite{Henrich:2011a,Henrich:2010b}.
 
@@
 
-We start the discussion with BPII, because its flow behaviour is simpler and probably more easily anticipated.
-Fig \ref{fig11} shows the disclination network undergoing a flow-induced transformation.
-The periodically recurring structure remains very regular and apart from the quasi-affine transformation relatively close to its quiescent appearance with individual disclination lines breaking up and reconnecting further downstream.
+We start the discussion with BPII. Fig \ref{bp2-med} shows the disclination network 
+undergoing an affine transformation under shear. The disclination lines break up and 
+reconnect further downstream forming a periodically recurring pattern. 
+Apart from the affine transformation the general appearance is very close to that 
+of the quiescent equilibrium phase.
 
@@
 \includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$.
-The velocity gradient is oriented along the vertical direction, whereas the horizontal direction is the flow direction.
-The flow takes place in such as way that network moves to the right in the upper half and to the left in the lower half.}
+\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
+of shear-induced transformations in the steady state at 
+$\dot{\gamma}=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
+is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
+The boundary conditions are applied flow takes place in such as way that network moves 
+to the right in the upper half and to the left in the lower half.}
 \label{bp2-med}
@@
 
-Interestingly, while being steadily sheared in y-direction, the entire disclination network is slowly displaced in z-direction, the direction of vorticity.
-This feature is most clearly observed when the structure is viewed along the flow direction.
+Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
+the gradient direction in the direction of vorticity.
+Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
+geometries~\cite{Henrich:2012b} the movement is steady. It happens to be in such a way that the 
+positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
+making the network travel along the vorticity direction.
 
-Fig. \ref{fig3} shows characteristic regions of upward and downward 'flow', which are normally superimposed by the primary shear flow in y-direction.
-Remarkably, the secondary flow is directly connected to the sense of the underlying cholesteric helix as changing from left- to right-handed helix also inverts the sense of motion of the network.
-
-The lefthand picture depicts the situation the movement of the disclination network is downwards, while helix inversion leads to an upward motion in the righthand picture.
-More precisely the orientation of the velocities with respect to the bands is exactly inverted in both cases.
-
-A closer look at the shape of the disclination lines reveals that also the network seems to be inverted. 
-The total magnitude of the secondary flow is in the range of a few percent of the primary flow and the images in Fig. \ref{fig3} give actually the situation close to the peak point.\\
-
-A secondary flow pattern with distinctive regions of positive and negative velocity components $v_z$ emerges and reoccurs periodically.
+Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
+Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
+This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
+two much smaller components $v_y$ and $v_z$.
+The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
+Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
+When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
+and the sense of motion also changes.
+Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
+individual cycles.
+Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
+All values for the two runs with inverted helicity are identical apart from a change of sign of the z-components.
 
@@
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
-\caption{Velocity patterns in BPII for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
-The colour code gives the magnitude and sign of the z-component.
-The vertical and horizontal direction are the gradient and vorticity direction, respectivelty.
-Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
+\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
+underlying cholesteric helix: 
+The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, i.e. the velocity has been projected onto a 
+plane perpendicular to the flow direction. The vertical and horizontal direction are the gradient and vorticity 
+direction, respectively.
+The colour code gives the magnitude and sign of the component in vorticity direction.
+The snapshot shows a typical frame during a periodically recurring sequence.
+The network on the left with positive helicity travels rightwards, whereas the one one the right
+moves leftwards.
+The recurrence period in vorticity direction is six times longer than the time it takes the network 
+to reconnect in flow direction.}
 \label{bp2-velo}
@@
 
+\clearpage
+
 \section{Conclusions}
@@
 \caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
-\label{table1}}
+\label{tab1}}
 \end{table*}

### 2012-08-14 trunk/papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1749 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \newcommand{\e}[1]{\times10^{#1}}
+\newcommand{\gd}{\dot{\gamma}}
 
@@
 \begin{abstract}
-bla
+We study the behaviour of cubic blue phases under shear flow
+via lattice Boltzmann simulations. We focus on the two experimentally observed
+phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
+II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
 \end{abstract}
@@
 The shear rates were varied over more than two orders of magnitude from about 
-$2.44\cdot 10^{-6}$ to $6.3\cdot10^{-4}$ in LB units.
+$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
 The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
@@
 of shear-induced transformations in the steady state at 
-$\dot{\gamma}=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
+$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
 is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
@@
 Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
-geometries~\cite{Henrich:2012b} the movement is steady. It happens to be in such a way that the 
+geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
 positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
-making the network travel along the vorticity direction.
+making the network travel along the z-direction.
+The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.
 
@@
 Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
-All values for the two runs with inverted helicity are identical apart from a change of sign of the z-components.
+All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
+There is only a slight imbalance between the maximum and minimum velocity in z-direction.
+Hence the movement of the network in vorticity direction is predominantly permeative.
+Note that the imbalance is exaclty retained if the helicity is inverted.
 
@@
 \caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
-underlying cholesteric helix: 
-The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, i.e. the velocity has been projected onto a 
-plane perpendicular to the flow direction. The vertical and horizontal direction are the gradient and vorticity 
-direction, respectively.
+underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
+i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
+The vertical and horizontal direction are the gradient and vorticity direction, respectively.
 The colour code gives the magnitude and sign of the component in vorticity direction.
@@
 The network on the left with positive helicity travels rightwards, whereas the one one the right
-moves leftwards.
-The recurrence period in vorticity direction is six times longer than the time it takes the network 
+moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
 to reconnect in flow direction.}
@@
 
-It is interesting to study the flow behaviour from a more quantitative point of view by looking at the shear stress dependence on flow rate and time.
+If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
+descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
+The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
+a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
+invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
+before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
+shear rates \cite{Rey:1996a, Rey:1996b}.
 
-The effective viscosity $\eta_{eff}$ over time is shown in Fig. \ref{fig4}. 
-For all apart from the highest flow rate it oscillates sinusoidally, exhibiting symmetric minima and maxima and decreasing mean values, indicating slightly shear-thinning behaviour.
-The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig11}.
-However, due to the motion of the network in z-direction it turns out that it takes four periods altogether for the network to move one unit cell along the direction of vorticity and to complete one cycle.
-The inset shows the deviatoric part of the chemical stress tensor as a function of shear rate.
-The dependence is that of a power law $\sigma_{eff, xy}=a \dot{\gamma}^b$ with $a=0.34(6116), b=0.95(2654)$ for low and intermediate shear rates. 
-The error bars indicate the maximum and minimum stress during one cycle.
-The idependence is almost linear and gives further evidence of the remarkably small degree of shear-thinning and also reflected in the very regular and undisturbed affine transformation of the disclination network.
-
-\begin{figure}[h]
-\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ of BPII over time (left) and strain (right). The inset shows the flow curve $\sigma_{chem}(\dot{\gamma})$.}
-\label{bp2-rheo}
-\end{figure}
-
-Above a critical shear rate, which is $\dot{\gamma}=3.1-6.2\cdot 10^{-4}$ in our case, the rotational diffusion coefficient $\Gamma$ in Eq.
-\ref{eqn2} is too small and the disclination network adopts a completely different conformation in the steady state.
-Fig. \ref{fig5} shows the director field and the position of disclination plane.
-The degree of twist is the same as that of the underlying cholesteric, but the orientation of the director on either side of the planes differs by almost 90 degrees, giving rise to four distinct layers, which are separated by disclination planes.
-Even in this flow-induced state the disclinations move and retain the above described sense of motion in the direction of vorticity.
-
-\begin{figure}[h]
-\caption{BPII at low shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom).
-The configuration is a travelling helical wave along the vorticity direction.
-The configuration is translationally invariant in flow and in gradient direction.
-The colour code indicates the orientation in x-y direction.}
-\label{bp2-low}
-\end{figure}
+In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
+However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
+such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
+Due to a lack of computational resources we have not investigated these aspects any further.
 
@@
 \caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom).
-The configuration is a travelling helical wave along the vorticity direction.
-The configuration is translationally invariant in flow and in gradient direction.
+The images show 1D-cuts as the configuration is translationally invariant in flow and gradient direction.
+The configuration is a travelling helical wave along the vorticity or z-direction.
 The colour code indicates the orientation in x-y direction.}
@@
 \end{figure}
+
+A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
+The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
+the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
+the picture on the right shows the same data versus accummulated total strain.
+For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
+is due to the periodic break-up and reconnection of the network under shear flow.
+
+The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
+The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
+a very good approximation, exibiting a remarkably small degree of shear-thinning.
+This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.
+
+\begin{figure*}[h]
+\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
+\label{bp2-rheo}
+\end{figure*}
 
@@
 
-The flow behaviour of BPI deviates significantly from that of BPII in a number of points.
+The flow behaviour of BPI at low shear rates is is similar deviates significantly from that of BPII in a number of points.
+
+\begin{figure}[h]
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-600k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-775k_run1115.png}
+\caption{Disclination network of BPI at low shear flow:}
+\label{bp1-low}
+\end{figure}
+
 First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
@@
 
+Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
+Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
+Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
+They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.
+
+\begin{figure}[h]
+\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
+\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
+\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
+\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
+The colour code gives the magnitude of the in-plane component of the director field.
+The data is translationally invariant along the vorticity direction.}
+\label{bp1-high}
+\end{figure}
+
 \begin{figure}[h]
@@
 \begin{figure}[h]
+\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
 \includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
@@
 \label{bp1-spec}
-\end{figure}
-
-Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
-Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
-Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
-They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.
-
-\begin{figure}[h]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-600k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-775k_run1115.png}
-\caption{Disclination network of BPI at low shear flow:}
-\label{bp1-low}
-\end{figure}
-
-\begin{figure}[h]
-\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
-\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
-\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
-The colour code gives the magnitude of the in-plane component of the director field.
-The data is translationally invariant along the vorticity direction.}
-\label{bp1-high}
 \end{figure}
@@
 \begin{table*}
-\begin{tabular}{|c|| c | c| c || c |c |c||c| c| c||c| c| c|}
+\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
 \hline
-$\dot{\gamma}$ & $\tau$ & $\kappa$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
 \hline
@@
 \hline
-1.95 & -0.5 & 1.0 &0.1388 & -2.49 &2.50 &4.51 &-124.65 &124.82 &3.62 &-1.62 &1.89 &3.51 \\
-3.91 & -0.5 & 1.0 &0.1388 & -3.31 &3.35 &4.21 &-248.69 &248.74 &3.90 &-2.88 &2.56 &4.39 \\
-7.81 & -0.5 & 1.0 &0.1388 &-6.49 &6.56 &7.99 &-497.22 &496.16 &4.46 &-7.46 &5.31 &6.81 \\ 
-15.63 & -0.5 & 1.0 &0.1388 &-3.08 &3.17 &10.49 &-992.46 &993.18 &10.36 &-3.57 &2.87 &10.54 \\
-15.63 & -0.5 & 1.0 &-0.1388 &-3.08 &3.17 &10.49 &-992.46 &993.18 &10.36 &-2.87 &3.57 &10.54 \\
-31.25 & -0.5 & 1.0 &0.1388 &-4.08 &4.46 &15.94 & -1988.36 &1986.71 &16.85 &-12.16 &11.37 &19.38\\
-62.50 & -0.5 & 1.0 &0.1388 & -62.63 & 62.13 & 24.68 & -4039.41 &3995.3  & 20.57 &-110.76  &73.52 & 33.26 \\
+0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
+7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
 \hline
@@
 \hline
-1.95 & -0.5 & 2.0 & 0.1963 & -1.09 &1.07 & 0.76 &  -62.04  &61.60 & 0.97 & -1.55 &1.64 & 0.81\\
-3.91 & -0.5 & 2.0 & 0.1963&-1.66 &1.70 & 1.45 &-123.93 &123.18 & 1.76 &-2.73 &3.09 &1.47\\
-7.81 &  -0.5 & 2.0 &0.1963& -2.62 &2.71 & 2.67 &-247.67 &246.45 & 3.20 &-4.77 &5.78 &2.74\\
-15.63 & -0.5 & 2.0 & 0.1963&-3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-7.66 &10.00 &4.33\\
-15.63 & -0.5 & 2.0 &-0.1963& -3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-10.00 &7.66 &4.33\\
-31.25 & -0.5 & 2.0 &0.1963 &-1.79 &1.82 &5.86 &-988.73 &985.93 &8.96  &-11.04 &14.39 &6.35\\
-62.50 & -0.5 & 2.0 &0.1963 & -1.20 & 1.20 & 0.38 & -1969.41  & 1969.41 & 3.07 &-1.48 &0.96 &0.38 \\
+0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
 \hline
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
+\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
 \label{tab1}}

### 2012-08-28 papers/bprheo/bprheo.tex
Test commit
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1446 40876441-3496-0410-a613-a2fca35b2ae2
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1763 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,pre,twocolumn,groupedaddress]{revtex4-1}
-%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4-1}
-%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
+\documentclass[aps,pre,reprint,superscriptaddress]{revtex4}
+%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
+%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}
 
@@
 \usepackage{amsmath}
+
+\newcommand{\e}[1]{\times10^{#1}}
+\newcommand{\gd}{\dot{\gamma}}
 
@@
 % \affiliation can be followed by \email, \homepage, \thanks as well.
-\author{O. Henrich$^{1,3}$, K. Stratford$^2$, D. Marenduzzo$^3$, P.V. Coveney$^1$ and M.E. Cates $^3$}
-\affiliation{$^1$ Centre for Computational Science, University College London, UK\\$^2$ Edinburgh Parallel Computing Centre, University of Edinburgh, UK\\$^3$ School of Physics and Astronomy, University of Edinburgh, UK}
 
-%\email[]{}
+\author{O. Henrich}
+\email[Corresponding author: ]{o.henrich@ucl.ac.uk}
+\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
+
+\author{K. Stratford}
+\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+
+\author{P.V. Coveney}
+\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
+
+\author{M.E. Cates}
+\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+
+\author{D. Marenduzzo}
+\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+
 %\homepage[]{Your web page}
@@
 \begin{abstract}
+We study the behaviour of cubic blue phases under shear flow
+via lattice Boltzmann simulations. We focus on the two experimentally observed
+phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
+II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
 \end{abstract}
@@
 
-In the following we report results on the bulk flow behaviour of BPI and BPII for thermodynamic states characterized by $\tau=-0.5$ and $\kappa=1.0$ (BPI) and $\kappa=2.0$ (BPII), respectively.
-For these parameters both blue phases are equilibrium phases in the quiescent state.
-We have also performed simulations at higher temperatures and chiralities and for metastable configurations, which led to very similar results.
-Quantitative differences occured only in that sense that the disclination network dissolved more easily for a particular flow rate if the system was closer to the cholesteric-isotropic phase boundary and where the order parameter takes lower values and free energy density is higher.\\
-The tensor order parameter was initialized with the high-chirality solution of Eq. \ref{eqn1}, which has the correct topology of the actual blue phase.
-The initial configurations were equilibrated for $5000$ LB time steps, where we allowed for optimization of the unit cell size \cite{Alexander:2006}, a feature that is often referred to as 'redshift'.
-After the equilibration sequence the unit cell size was fixed and the shear flow was switched on.
-For BPII we used a pitch length of 32 lattice sites, but in case of BPI it was necessary to increase the resolution to 64 sites per pitch length to get accurate results of the flow induced transformations in the disclination network.
-In order to impose a flow profile we used Lees-Edwards boundary conditions \cite{Wagner:2002}, whereas the top (bottom) half of planes moved in positive (negative) y-direction.
-We applied a correction for spurious momentum currents that emerge due to interpolations accross the Lees-Edwards planes \cite{Henrich:2011b}.
-The imposed flow rates ranged over two orders of magnitude from about $1.25\cdot 10^{-6}$ to $6\cdot10^{-4}$ in LB units.\\
-In the visualizations that follow the orientation of the principal coordinate axes in x-, y- and z-direction always coincides with the direction of the velocity gradient direction, the flow and the vorticity, respectively.
-Disclination lines are plotted as isosurface of the scalar order parameter $q$
+In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
+phases BPI and BPII. The thermodynamic states are characterised by an effective temperature 
+$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy terms.
+We opted for states sufficiently far away from the chiral-nematic-isotropic 
+where both quiescent blue phases are at equilibrium, namely
+$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
+Simulations at metastable states at higher and lower temperatures and different chiralities showed 
+no differences in their generic flow behaviour.
+However, states closer to the phase boundary featuring higher free energy densities and 
+smaller average order parameters show a tendency to break up more easily under external driving forces.
+
+As is common in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with 
+analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, 
+and equilibrated these (for 5000 LB timesteps) prior to starting the shear. 
+During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
+After equilibration the redshift parameter was kept constant throughout the rest of the simulation.
+We chose a unit cell size of 16 lattice sites for BPII and 32 sites in case of BPI. 
+Runs with higher resolution confirmed that this was an appropriate minimal resolution to cover 
+the complex conformational changes and kinematic details of the blue phase networks.
+
+Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
+the top (bottom) half of the planes moving in positive (negative) x-direction.
+We applied a correction for spurious momentum currents that emerge due to interpolations accross 
+the Lees-Edwards planes \cite{Henrich:2012a}.
+The shear rates were varied over more than two orders of magnitude from about 
+$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
+The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
+along the y- and z-axis.
+Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
+are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
+The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+is equal to about $10^8$~Pa.
+More details about the conversion from LBU to SI units can be found in
+\cite{Henrich:2011a,Henrich:2010b}.
 
@@
 
-We start the discussion with BPII, because its flow behaviour is simpler and probably more easily anticipated.
-Fig \ref{fig1} shows the disclination network undergoing a flow-induced transformation.
-The periodically recurring structure remains very regular and apart from the quasi-affine transformation relatively close to its quiescent appearance with individual disclination lines breaking up and reconnecting further downstream.
+We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
+BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
+intersect and form a characteristic network of nodes.
+
+\subsubsection{Low and intermediate shear rates}
+
+Fig \ref{bp2-med} shows the disclination network 
+undergoing an affine transformation under shear. The disclination lines break up and 
+reconnect further downstream forming a periodically recurring pattern. 
+The recurrence time in flow direction is $\tau_F=p/(2\, v_{LE})=p/(2\,\gd\, \Delta y_{LE})= 1/\gd$.
+Apart from the affine transformation the general appearance is very close to that 
+of the quiescent equilibrium phase.
 
@@
 \includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\cdot 10^{-4}$.
-The horizontal y-direction is the flow direction whereas the velocity gradient is oriented along the vertical x-direction.}
-\label{fig1}
+\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
+of shear-induced transformations in the steady state at 
+$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
+is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
+The boundary conditions impose flow in such a way that network moves 
+to the right in the upper half and to the left in the lower half of the picture.}
+\label{bp2-med}
 \end{figure*}
 
-Interestingly, while being steadily sheared in y-direction, the entire disclination network is slowly displaced in z-direction, the direction of vorticity.
-This feature is most clearly observed when the structure is viewed along the flow direction.
-Fig. \ref{fig2} depicts a sequence of close-ups of the disclination network as it travels in negative z-direction, i.e. from the foreground towards the background of the picture. 
-The velocity $v_z$, represented by the colour code, exhibits a recurring pattern of positive and negative values.
-Peak velocities occur close to disclination lines where the molecular order is suppressed or in regions where two disclination lines approach each other and the network is about to break up or reform.
+Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
+the gradient direction in the direction of vorticity.
+Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
+geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
+positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
+making the network travel along the z-direction.
+The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.
+
+A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
+The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
+the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
+the picture on the right shows the same data versus accummulated total strain.
+For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
+is due to the periodic break-up and reconnection of the network under shear flow.
+
+The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
+The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
+a very good approximation, exibiting a remarkably small degree of shear-thinning.
+This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.
 
 \begin{figure*}[h]
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-149k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-160k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-167k_run902.png}\\
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-173k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-182k_run902.png}
-\includegraphics[width=0.32\textwidth]{slice-disc-v_z-192k_run902.png}\\
-\caption{A section of $2\times2\times1$ unit cells of BPII seen along the direction of vorticity.
-The colour code gives the magnitude and direction of the velocity component $v_z$ perpendicular to the plane.}
-\label{fig2}
+\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
+\label{bp2-rheo}
 \end{figure*}
 
-Plotting only x- and z-component of the local velocity allows to visualize the spatial dependence of the secondary flow in the blue phase with respect to the disclination network.
-Fig. \ref{fig3} shows characteristic regions of upward and downward 'flow', which are normally superimposed by the primary shear flow in y-direction.
-Remarkably, the secondary flow is directly connected to the sense of the underlying cholesteric helix as changing from left- to right-handed helix also inverts the sense of motion of the network.
-%%%%%%%%%%%%%%%%%%%%
-
-The lefthand picture depicts the situation the movement of the disclination network is downwards, while helix inversion leads to an upward motion in the righthand picture.
-More precisely the orientation of the velocities with respect to the bands is exactly inverted in both cases.
-
-A closer look at the shape of the disclination lines reveals that also the network seems to be inverted. 
-The total magnitude of the secondary flow is in the range of a few percent of the primary flow and the images in Fig. \ref{fig3} give actually the situation close to the peak point.\\
-
-A secondary flow pattern with distinctive regions of positive and negative velocity components $v_z$ emerges and reoccurs periodically.
+Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
+Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
+This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
+two much smaller components $v_y$ and $v_z$.
+The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
+Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
+When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
+and the sense of motion also changes.
+Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
+individual cycles.
+Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
+All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
+There is only a slight imbalance between the maximum and minimum velocity in z-direction.
+Hence the movement of the network in vorticity direction is predominantly permeative.
+Note that the imbalance is exaclty retained if the helicity is inverted.
 
 \begin{figure*}[h]
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run902.png}
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-160k_run903.png}
-\caption{Secondary flow in BPII visualized by means of the velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity of the cholesteric helix.
-The images depict an individual snapshot of the periodically recurring sequence in the steady state.}
-\label{fig3}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
+\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
+underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
+i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
+The vertical and horizontal direction are the gradient and vorticity direction, respectively.
+The colour code gives the magnitude and sign of the component in vorticity direction.
+The snapshot shows a typical frame during a periodically recurring sequence.
+The network on the left with positive helicity travels rightwards, whereas the one one the right
+moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
+to reconnect in flow direction.}
+\label{bp2-velo}
 \end{figure*}
 
-It is interesting to study the flow behaviour from a more quantitative point of view by looking at the shear stress dependence on flow rate and time.
+\subsubsection{High shear rates}
 
-The effective viscosity $\eta_{eff}$ over time is shown in Fig. \ref{fig4}. 
-For all apart from the highest flow rate it oscillates sinusoidally, exhibiting symmetric minima and maxima and decreasing mean values, indicating slightly shear-thinning behaviour.
-The period reflects the time it takes for the disclination network to break up and reconnect as shown in Fig. \ref{fig1}.
-However, due to the motion of the network in z-direction it turns out that it takes four periods altogether for the network to move one unit cell along the direction of vorticity and to complete one cycle.
-The inset shows the deviatoric part of the chemical stress tensor as a function of shear rate.
-The dependence is that of a power law $\sigma_{eff, xy}=a \dot{\gamma}^b$ with $a=0.34(6116), b=0.95(2654)$ for low and intermediate shear rates. 
-The error bars indicate the maximum and minimum stress during one cycle.
-The idependence is almost linear and gives further evidence of the remarkably small degree of shear-thinning and also reflected in the very regular and undisturbed affine transformation of the disclination network.
+If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
+descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
+The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
+a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
+invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
+before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
+shear rates \cite{Rey:1996a, Rey:1996b}.
+
+In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
+However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
+such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
+Due to a lack of computational resources we have not investigated these aspects any further.
 
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{stress_bp2.pdf}
-\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPII in the steady state. }
-\label{fig4}
+\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
+\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
+\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
+\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
+\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+The configuration is a helical wave that travels along the direction of vorticity.
+It is translationally invariant in gradient and flow direction.}
+\label{bp2-high}
 \end{figure}
-
-Above a critical shear rate, which is $\dot{\gamma}=3.1-6.2\cdot 10^{-4}$ in our case, the rotational diffusion coefficient $\Gamma$ in Eq.
-\ref{eqn2} is too small and the disclination network adopts a completely different conformation in the steady state.
-Fig. \ref{fig5} shows the director field and the position of disclination plane.
-The degree of twist is the same as that of the underlying cholesteric, but the orientation of the director on either side of the planes differs by almost 90 degrees, giving rise to four distinct layers, which are separated by disclination planes.
-Even in this flow-induced state the disclinations move and retain the above described sense of motion in the direction of vorticity.
-
-\begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{dir-op-y_225k_run949.png}
-\caption{BPII at high shear rate: director field and the disclination planes seen along the flow direction.
-The BPII breaks up and flows in helical layers separated from each other by a disclination plane}
-\label{fig5}
-\end{figure}
+\clearpage
 
@@
 
-The flow behaviour of BPI deviates significantly from that of BPII in a number of points.
-First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
-This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
-Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig1}, but this time seen along the direction of flow.
-A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
-The initial and final state in this sequence correspond to half a cycle. 
-This can be easily verified by comparing the position of the disclination network at the left or right boundary.
-BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.
+In BPI the disclination lines are arranged in a body-centered-cubic array.
+Contrary to BPII, they are well separated and do not intersect. 
+This difference in topology leads to some stiking differences in the flow behaviour of both blue phases 
+as we will see in the following.
+
+We can identify three different regimes in the range of applied shear rates:
+At high shear rates a breakup of the network structure occurs, similar to that seen in BPII. 
+At intermediate shear rates BPI features seemingly chaotic but periodically recurring conformations.
+At low shear rates the network undergoes first an affine transformation, but becomes unstable
+at larger strains and adopts an amorphous configuration with well separated disclination lines.
+
+\subsubsection{Intermediate shear rates}
+
+Fig. \ref{bp1-med} shows a typical sequence of shear-induced conformations.
+The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
+
+There is again an offset in the position of the network at the begin and the end of each cycle. 
+But neither the magnitude nor the direction of the offset seem to obey a clear rule.
+They are seemingly chaotic and different for every shear rate.
 
@@
 \includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
-\caption{Disclination network of BPI seen along the flow direction in steady shear flow at $\dot{\gamma}=1.56\cdot 10^{-4}$, showing the downward movement of the disclination network in the direction of vorticity.}
-\caption{\label{fig6}}
+\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page.
+The network is displaced during every cycle, but the magnitude and direction of the offset are seemingly chaotic.}
+\label{bp1-med}
 \end{figure*}
 
-The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
-Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
-Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
-
-\begin{figure*}[h]
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run914.png}
-\includegraphics[width=0.45\textwidth]{v_xz-v_z-360k_run922.png}
-\caption{Secondary flow in BPI: velocity components $v_x$ and $v_z$ for positive (left) and negative (right) helicity.}
-\label{fig7}
-\end{figure*}
-
-Despite qualitative similarities between the BPI and BPII, the quantitative analysis reveals striking differences.
-In Fig. \ref{fig8} we show the effective viscosity $\eta_{eff}$ over time and flow curves $\sigma_{chem}(\dot{\gamma})$. 
-BPI exhibits pronounced shear-thinning contrary to BPII, whereas the shape of the curves in no longer symmetric.
-In fact on approaching the critical shear rate that led in the case of BPII to the disruption of the network, the curves become increasingly complex but remain fully periodic.
-The effective viscosity drops by about an order of magnitude between the lowest and the highest shear rates.
-Note that the scale has been adapted to allow for the different magnitudes.
-The inset can be directly compared with that of Fig. \ref{fig4}.
-Interestingly the difference between minima and maxima of the chemical shear stress is much larger than for BPII and for the lowest shear rate we applied the absolute minimum takes even negative values, which are not shown in Fig.
-\ref{fig4}.
-Our understanding of this phenomenon is an interplay of the elastic forces and the individual order paramter configuration of the liquid crystal.
-If the system is close to a local minimum of the free energy but moving away from it the increasing strain pumps elastic energy into the system. 
-This results in positive chemical stresses as the system tries to withstand the deformation.
-However, if the system has sufficiently transformed and is about to overcome a local maximum, it will be pulled towards the next local minumum, which can lead temporarily to negative chemical stresses.
-The inset shows that the chemical stress decreases monotonically with decreasing flow rate, but does not follow a simple power law dependence.
+The apparent viscosity of BPI in simple shear flow, depicted in Fig. \ref{bp1-rheo}, is larger for The degree of shear thinning, also visible in the inset in terms of average shear stress over
+flow rate
 
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{stress_bp1.pdf}
-\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ over time and flow curve $\sigma_{chem}(\dot{\gamma})$ (inset) of BPI in the steady state.}
-\label{fig8}
+\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
+\label{bp1-rheo}
 \end{figure}
+
+Runs at smaller system sizes suggest that the individual conformation 
+can depend sensitively on the system size. Fig. \ref{bp1-2uc4uc}
+
+\begin{figure}[h]
+\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{}
+\label{bp1-2uc4uc}
+\end{figure}
+
+\begin{equation}
+X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N)
+\end{equation}
 
@@
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{spectrum_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
 \caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
-\label{fig9}
+\label{bp1-spec}
 \end{figure}
 
-Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
-Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
-Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
-They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.
+%%%%%%%%%%%%%%%%%%%
+The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
+Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
+Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
+
+\begin{figure*}[h]
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
+The colour code gives the magnitude and sign of the z-component.
+The vertical and horizontal direction are the gradient and vorticity direction, respectivelty.
+Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
+\label{bp1-velo}
+\end{figure*}
+
+\subsubsection{Low shear rates}
+
+At low shear rates the conformational changes of BPI in shear flow are very similar to 
+that of BPII descibed above. Fig. \ref{bp1-low} shows how the network undergoes an affine 
+transformation during one cycle. However, contrary to BPII the disclination lines of BPI 
+do not resemble those of the quiescent state, although they have the same topological
+structure. At one particular point (top-left and top-centre) 
+the disclination network are intertwined helices with the helical axis aligned
+along the gradient direction. This shear-induced conformation forms shortly after the
+flow has been switched on at strains $\gamma\simeq1$.
+During one cycle there is no perceivable movement of the network in vorticity direction.
 
 \begin{figure}[h]
-\includegraphics[width=0.5\textwidth]{dir-op-z_395k_run916.png}
-\includegraphics[width=0.5\textwidth]{dir-op-x_395k_run916.png}
-\caption{BPI at high shear rate: director field and the disclination planes seen along the direction of voritcity and the gradient direction.
-BPI forms rolls with $\pi/2$ disclinations that are advected with the general flow.}
-\label{fig10}
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\caption{Disclination network of BPI at low shear rate: Depicted is a sequence of affine, shear-induced transformations at shear rate $\dot{\gamma}=4.88\e{-6}$ and time step $t=4.0, 4.0, 6.0, 7.25, 7.50, 7.75\e{5}$ (top-left to bottom-right).
+The horizontal and the vertical direction are the flow and the gradient direction in all but the top-centre picure.
+There, the network is depicted in gradient direction with the direction of vorticity oriented along the vertical axis.}
+\label{bp1-low}
 \end{figure}
+
+\subsubsection{High shear rates}
+
+\begin{figure}[h]
+\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
+\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
+\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
+\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
+The colour code gives the magnitude of the in-plane component of the director field.
+The data is translationally invariant along the vorticity direction.}
+\label{bp1-high}
+\end{figure}
+
+\clearpage
 
@@
 \hline
-1.95 & -0.5 & 2.0 & 0.1963 & -1.09 &1.07 & 0.76 &  -62.04  &61.60 & 0.97 & -1.55 &1.64 & 0.81\\
-3.91 & -0.5 & 2.0 & 0.1963&-1.66 &1.70 & 1.45 &-123.93 &123.18 & 1.76 &-2.73 &3.09 &1.47\\
-7.81 &  -0.5 & 2.0 &0.1963& -2.62 &2.71 & 2.67 &-247.67 &246.45 & 3.20 &-4.77 &5.78 &2.74\\
-15.63 & -0.5 & 2.0 & 0.1963&-3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-7.66 &10.00 &4.33\\
-15.63 & -0.5 & 2.0 &-0.1963& -3.13 &3.31 &4.24 &-494.85 &493.09 & 5.64 &-10.00 &7.66 &4.33\\
-31.25 & -0.5 & 2.0 &0.1963 &-1.79 &1.82 &5.86 &-988.73 &985.93 &8.96  &-11.04 &14.39 &6.35\\
-62.50 & -0.5 & 2.0 &0.1963 & -1.20 & 1.20 & 0.38 & -1969.41  & 1969.41 & 3.07 &-1.48 &0.96 &0.38 \\
+0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
 \hline
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow: Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge$200000 for BPI and t$\ge$30000 in case of BPII.
-\label{table1}}
+\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
+\label{tab1}}
 \end{table*}

### 2012-08-29 trunk/papers/2012-sandwich/support.tex
Sandwich material added
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1768 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%
%%%%% Supporting material for Sandwich paper.
%%%%% Compile with pdflatex as figures are not eps.
%%%%%
\documentclass[12pt,twoside]{article}
\usepackage{graphicx}

\setlength{\hoffset}{-1in}
\setlength{\voffset}{-1in}

\setlength{\topmargin}{2cm}
\setlength{\evensidemargin}{3.1cm}
\setlength{\oddsidemargin}{3.1cm}

\setlength{\textwidth}{14.8cm}
\setlength{\textheight}{23cm}

\setlength{\parindent}{0pt}
\setlength{\parskip}{\medskipamount}
\pagestyle{empty}

\begin{document}

{\bf Supporting Online Material: Sandwich Title}

Here, the details of the simulation method, protocol, and parameters are
set out. The matching of simulation units to physical units is also
discussed.

{\bf 1. Thermodynamics}

The thermodynamics of the cholesteric liquid crystal can be described
by means of a  Landau-de Gennes free energy functional $\cal F$,
whose density is written as $f$:
\begin{eqnarray}
{\cal F}[{\bf Q}]&=&\int d^3{\bf r} f({\bf Q}({\bf r})).
\end{eqnarray}
This free energy density $f = f({\bf Q})$ may be expanded in powers of the
order parameter ${\bf Q}$ and its gradients; ${\bf Q}$ is a traceless 
and symmetric tensor which we will denote from now on in subscript
notation as $Q_{\alpha\beta}$.
The largest eigenvalue $q$ and corresponding eigenvector $n_\alpha$
of $Q_{\alpha\beta}$ describe the local strength and major orientation axis
of molecular order.
The theory based on the tensor $Q_{\alpha\beta}$, rather than a theory based
solely
on the director field $n_\alpha ({\bf r})$, allows treatment of disclinations
(defect lines) in whose cores $n_\alpha$ is undefined.

Explicitly, the free energy density is:
\begin{eqnarray}
f(Q_{\alpha\beta}) &=& {\textstyle \frac{1}{2}} A_0
 \left(1- {\textstyle \frac{1}{3}}\gamma \right)Q^2_{\alpha \beta}
-{\textstyle \frac{1}{3}}A_0\gamma Q_{\alpha \beta} Q_{\beta \gamma}Q_{\gamma \alpha} 
+ {\textstyle \frac{1}{4}} A_0\gamma (Q^2_{\alpha \beta})^2  \nonumber\\
&+& {\textstyle \frac{1}{2}} K(\varepsilon_{\alpha \gamma \delta}
\partial_\gamma Q_{\delta \beta} + 2 q_0 Q_{\alpha \beta})^2
+ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2
\label{free}
\end{eqnarray}
Here, repeated indices are summed over, while terms of the form
$Q^2_{\alpha\beta}$ should be expanded to read $Q_{\alpha\beta}Q_{\alpha\beta}$;
$\varepsilon_{\alpha\gamma\delta}$ is the permutation tensor.
The first three terms are a bulk free energy density whose overall scale
is set by $A_0$ (discussed further below); $\gamma$ is a control parameter
related to a reduced temperature. Varying the latter in the absence of
chiral terms ($q_0=0$) gives an isotropic-nematic transition at
$\gamma = 2.7$ with a mean-field spinodal instability at $\gamma = 3$.

The remaining terms of the free energy density in Eq.~\ref{free} describe
distortions of
the order parameter field. It is conventional \cite{blue1,deGennes} to
assume that splay, bend and twist deformations of the director are equally
costly, that is, there is a single elastic constant $K$. The parameter $q_0$
is related to the helical pitch length $p$ via $q_0=2\pi/p$,
describing one full turn of the director in the cholesteric phase.

There are two dimensionless numbers, which are commonly referred to as
$\kappa$, the chirality, and  $\tau$, the reduced temperature
\cite{blue1}.  In terms of the parameters in the free energy density,
these are:
\begin{eqnarray}\label{cntrl-param} 
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\label{tau}\\
\kappa&=&\sqrt{\frac{108\ K\, q_0^2}{A_0\, \gamma}}\label{kappa}.
\end{eqnarray}
If the free energy density Eq.~\ref{free} is made dimensionless, $\tau$
appears as prefactor of the term  quadratic in $Q_{\alpha\beta}$,
whereas $\kappa$
quantifies the ratio between bulk and  gradient free energy terms. The
chirality and reduced temperature may be used to characterise an equilibrium
phase diagram of the blue phases in bulk cholesteric liquid crystals
\cite{blue1,oliver1}.
In this work, the chirality and reduced temperature are always set to lie
within the blue phase I region of the equilibrium phase diagram.

In addition to the fluid free density density $f(Q_{\alpha\beta})$, a surface
free energy density is also present to represent the energetic cost of
anchoring at a solid surface. If the preferred order parameter tensor
at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
density (per unit area) is
\begin{equation}
f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
\end{equation}
where $W$ is a constant determining the strength of the anchoring.
The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
discussed below. For a colloid of radius $a$, the strength of the surface
anchoring compared which the bulk fluid elastic constant may be quantified
by the dimensionless parameter $Wa/K$. For small values of this parameter,
the presence of a colloid surface should have little impact on the local
fluid LC ordering.

% NO REDSHIFT

{\bf 2. Dynamics}

{\bf 2.1 Order parameter}

A framework for the dynamics of liquid crystals is provided by the 
Beris-Edwards model \cite{beris}, in which the time evolution of the
tensor order parameter obeys
\begin{equation}
\label{eom}
\left(\partial_t+ u_\nu \partial_\nu \right) Q_{\alpha\beta} - S_{\alpha\beta}
= \Gamma H_{\alpha\beta}.
\end{equation}
In the absence of flow, Eq.~\ref{eom} describes a relaxation towards
equilibrium on a timescale determined by a collective rotational diffusion 
constant $\Gamma$. This relaxation is driven by the molecular field
$H_{\alpha\beta}$, which is the functional derivative of the free energy
with respect to the order parameter \cite{beris}:
\begin{equation}
H_{\alpha\beta} = -\frac{\delta {\cal F}} {\delta Q_{\alpha\beta}} 
+ {\textstyle \frac{1}{3}} \delta_{\alpha\beta} 
\mathrm{Tr}\left(\frac{\delta {\cal F}}{\delta Q_{\alpha\beta}}\right).
\end{equation}

The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
derivative for rod-like molecules \cite{beris}. It couples the order
parameter to the symmetric and antisymmetric parts of the velocity 
gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
$\Omega_{\alpha\beta}$ are defined as 
\begin{equation}
A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
\end{equation}
and
\begin{equation}
\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
\end{equation}
This full coupling term is then
\begin{eqnarray}
\label{coupling-term}
S_{\alpha\beta}  &=&
(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
+
(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
(\xi A_{\nu\beta} - \Omega_{\nu\beta})
\nonumber\\ 
&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
Q_{\nu\nu} W_{\nu\nu}.
\end{eqnarray}
CORRECT TRACE?!
Here, $\xi$ is a material-dependent `tumbling parameter' that controls the
relative importance of rotational and elongational flow for molecular
alignment. In all cases, the value used here is $\xi = 0.7$, which is within
the `flow aligning' regime, where molecules align at a fixed angle
(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
The value of the collective rotational diffusion used in all simulations is
$\Gamma = 0.5$ in simulation units.

{\bf 2.2 Hydrodynamics}

The momentum evolution obeys a Navier-Stokes equation driven by the divergence
of a generalised stress $P_{\alpha\beta}$:
\begin{equation}
\label{nse}
\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
=\partial_\beta P_{\alpha\beta}
\end{equation}
The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
both viscous and thermodynamic components:
\begin{eqnarray}
P_{\alpha \beta}&=&
- p_0 \delta_{\alpha \beta} 
+ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
\nonumber\\
&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
+ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
-\xi \left(Q_{\alpha \gamma}
+ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
\nonumber\\ 
&+& 2 \xi  \left(Q_{\alpha \beta}
+ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
-\partial_\alpha Q_{\gamma \nu}
\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\end{eqnarray}
A lattice Boltzmann flow solver is used, where the isotropic pressure $p_0$
and viscous terms are managed directly by the solver (as in a simple Newtonian
fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
is treated as a local force density on the fluid. In all simulations
reported the mean fluid density $\rho = 1$ and the viscosity
$\eta = 0.01$ (simulations in confined geometry also used enhanced
bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).

{\bf 3. Surface boundary conditions}

Hydrodynamic boundary conditions for solid objects are handled via
the lattice Boltzmann solver. In particular, the standard method of
bounce-back on links is used for both colloids and confining walls
\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
are set out in the following.

{\bf 3.1 Colloids}

In all the simulations reported here the preferred orientation of the
director field at the colloid surface is normal to the surface. The
required nematic director at the particle surface in a given location
$\hat{n}^0_\alpha$ may then be determined from geometry alone,
and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
\begin{equation}
Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
\label{eq:surf_q}
\end{equation}
The magnitude of surface order $q^0$ is set by
\begin{equation}
q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
+ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
\label{eq:surf_q0}
\end{equation}
CHECK which corresponds to SOMETHING [REFERENCE].

In all simulations the colloidal (hard sphere) radius is $a = 2.3$. To
prevent colloids overlapping at the hard sphere radius, an addition
short range soft potential is included. This takes the form of
$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
units. Both potential and resulting force are smoothly matched to zero at
a cut-off distance of $h_c = 0.25$ simulation units.

{\bf 3.2 Confining walls}

The confining solid walls, which are flat and stationary, have the
same normal boundary condition as at colloid surface. In addition,
degenerate planar anchoring is applied at the walls, where the
preferred nematic director may lay in any orientation parallel
to the surface. The preferred direction is determined explicitly
by projecting the local fluid order parameter to the plane of the
wall \cite{fournier2005}. Eq.~\ref{eq:surf_q} and Eq.~\ref{eq:surf_q0}
are then used as before.

To prevent the colloids being forced into the wall, a correction
to the lubrication force on the colloid is added for very low colloid
surface to wall surface separations $h < 0.5$ lattice spacing. The
correction is based on the analytic expression for the lubrication
between a sphere and a flat surface.

{\bf 4. Protocol and parameter details}

{\bf 4.1 Bulk blue phase I}

The parameters for the free energy are chosen to be representative
of blue phase I: chirality $\kappa = 0.7$ and reduced temperature
$\tau = 0$.
The full free energy parameters are shown in Table~\ref{table:params}.
The order parameter tensor $Q_{\alpha\beta}$ for bulk blue phase I is
initialised from an approximation in high chirality limit
\cite{blue1,oliver1}. Systems of 128$^3$ lattice sites are
used with a pitch length of $p = 32\sqrt{2}$, which accommodates 4$^3$
unit cells with fully periodic boundary conditions.
Simulations at
different solid volume fractions (1\% and 5\%), and different surface
anchoring strengths ($Wa/K = 0.23$ and $Wa/k = 2.3$) representing
``weak'' and ``strong'' anchoring are run for 500,000 simulation
time steps (and 700,000 time steps for 5\% solid volume fraction).
The initial distribution of the colloid positions is set at random
for the required solid volume fraction; the colloids are initially
at rest.

To generate a disordered network of disclination lines, a ``quench''
is performed in which the order parameter is initialised via a
locally chosen random director (and using an equation of the form
Eq.~\ref{eq:surf_q} with an initial amplitude of order $q = -0.2$).
In order to avoid the mean-field spinodal point at $\gamma = 3.0$,
the parameters are adjusted slightly while maintaining the same
pitch $p = 32\sqrt{2}$. The chirality and reduced temperature (see
Table~\ref{table:params}) remain appropriate for blue phase I.
Colloids are added
as before and the simulations run for 500,000 or 700,000 simulation
steps.

{\bf 4.2 Confined blue phase I}

Here, a narrow sandwich of fluid is placed between flat walls in
perpendicular to the narrow coordinate direction, and with periodic
boundaries in the other two directions. The system size used in all
cases is 256$^2 \times$56 lattice sites. Each simulation is a ``quench''
in a similar manner to that used in the bulk: the order parameter is
initialised randomly. The free energy parameters (see Table~\ref{table:params})
are again appropriate for blue phase I (chirality $\kappa = 0.8$ and
reduced temperature $\tau = -0.25$). The cholesteric pitch is set
to be $p = 64$, providing a slight better resolution that the bulk
simulations.

The surface anchoring for colloids is as before: always normal, but with
$Wa/K = 0.23$ and $Wa/K = 2.3$. The colloid solid volume fractions are
1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
in simulation units. All simulations are run for 1.6 million simulations
time steps unless otherwise stated CHECK.

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $Wa/K$\\
\hline 
Bulk BPI & 0.01 & 3.000 & 0.007061 & 32$\sqrt{2}$ & 0.7000 & 0.0000 &
0.23, 2.3\\
\hline
Bulk Quench & 0.01 & 3.100 & 0.007061 & 32$\sqrt{2}$ & 0.6886 & -0.2903 &
0.23, 2.3 \\
\hline
Confined & 0.01 & 3.0857 & 0.01897 & 64 & 0.8 & -0.25 & 0.23, 2.3\\
\hline
\end{tabular}
\end{center}
\caption{Free energy parameters used in the simulations.}
\label{table:params}
\end{table}

In all cases visualisation of the defect lines is carried out via
an isosurface of the scalar order parameter $q$ determined from the
largest eigenvalue of the local order parameter tensor. A low value
of $q$ (typically 0.12) unambiguously identifies the disclinations.

{\bf 5. Parameter mapping to physical units}

ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK

ISSUE: no redshift?

ISSUE: do we want some free energy numbers?

PENDING

{\bf 6. Supporting Figures and Animations}

\begin{figure}
\begin{center}
\includegraphics[scale=0.35]{s1.jpg}
\end{center}
\caption{As for main manuscript Fig.~1 (initialised to equilibrium blue
phase I), but for the entire simulation
system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
volume fraction and strong anchoring. The view direction in the
main figure is from the left here. There is a reference
particle at the bottom left in each panel which does not take part
in the simulation.}
\end{figure}

\begin{figure}
\includegraphics[scale=0.35]{s2.jpg}
\caption{As for main manuscript Fig.~2 (initialised via a ``quench''
to generate a disordered network), but for the entire simulation
system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
volume fraction and strong anchoring. Again, there is a reference
particle at the bottom left in each case which does not take part
in the simulation.}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.42]{s3.jpg}
\end{center}
\caption{As for main manuscript Fig.~3 (confined geometry with normal
anchoring walls at both top and bottom), but for the entire simulation
system of 256$^2\times$56 lattice sites. Each shows a top view and
side view of the same simulation state. (a) solid volume fraction 1\%
and weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 4\% solid volume fraction and weak anchoring; (d) 4\% solid
volume fraction and strong anchoring.}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.42]{s4.jpg}
\end{center}
\caption{As for main manuscript Fig.~4 (confined geometry with planar
anchoring walls at both top and bottom), but for the entire simulation
system of 256$^2\times$56 lattice sites. Each shows a top view and
side view of the same simulation state. (a) colloid solid volume fraction 1\%
and weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 4\% solid volume fraction and weak anchoring; (d) 4\% solid
volume fraction and strong anchoring.}
\end{figure}

\newpage

{\bf Supporting References}

\begin{thebibliography}{99}

\bibitem{blue1}
D.C. Wright and N.D. Mermin,
%  Crystalline liquids -- the blue phases.
{\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).

%\bibitem{marendu1}
%D. Marenduzzo, E.  Orlandini, M. E. Cates, J. M. Yeomans,
%{\rm Steady-state hydrodynamic instabilities of active liquid crystals:
%Hybrid lattice Boltzmann simulations.}
%{\it Phys. Rev. E} {\bf 76}, 031921 (2007).

%\bibitem{marendu2}
%O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates, 
%{\rm Domain growth in cholesteric blue phases:
%hybrid lattice Boltzmann simulations}.
%{\it Comput. Math. Appl.}, doi:10.1016/j.camwa.2009.08.047[dx.doi.org] (2009).

%\bibitem{marendu3}
%M. E. Cates, O. Henrich, D. Marenduzzo, K. Stratford,
%{\rm  Lattice Boltzmann simulations of liquid crystalline fluids:
%active gels and blue phases.}
%{\it Soft Matter} {\bf 5}, 3791-3800 (2009).

\bibitem{oliver1}
O. Henrich, D. Marenduzzo, K. Stratford, and M.E. Cates,
%Thermodynamics of blue phases in electric fields
\textit{Phys. Rev. E}, \textbf{81} 031706 (2010).

\bibitem{beris}
A. N. Beris, B. J. Edwards, 
{\it Thermodynamics of Flowing Systems with Internal Microstructure}
(Oxford Univ. Press, New York, 1994).

\bibitem{deGennes}
P.-G. de Gennes, J. Prost,
{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).

\bibitem{ladd94}
A.J.C. Ladd,
\textit{J. Fluid Mech.} \textbf{271}, 285 (1994); \textit{ibid.} \textbf{271}
311 (1994).

\bibitem{nguyen2002}
N.-Q. Nguyen and A.J.C. Ladd,
\textit{Phys. Rev. E} \textbf{66} 046708 (2002).

\bibitem{fournier2005}
J.-B. Fournier and P. Galatola,
\textit{Europhys. Lett.} \textbf{72}, 403 (2005).

\end{thebibliography}

\end{document}

% MY NOTES

KEVIN

All particle radius $a = 2.3$.

Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = 10\eta$;
temperature $T =2.133333 10^{-9}$.

Free energy:
$A_0 = 0.01$; $\gamma = 3.085143$; $q_0 = 0.09817477$ (and $p = 64$);
$\kappa_1 = \kappa_2 = 0.018971857$;
Chirality $\bar{\kappa} = 0.8$; $\tau = -0.25$

Amplitude order $ 0.0001$; Computed surface order $0.3509236$;
Rotational diffusion $\Gamma = 0.5$
Initialisation is 'random'.

Wall surface $W = 0.018971957$, ie., $W/\kappa$ = 1 length

Weak colloid anchoring $W = 0.0018971957$, ie., $W/\kappa = 0.1$ length
and $Wa/\kappa = 0.23$.

Strong colloid anchoring $W = 0.037943914$, ie., $W/\kappa = 2.0$ length
or $Wa/\kappa = 4.6$.

JUHO (BPI)

Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = \eta$;
Temperature: $T = 2.133333 10^{-9}$.

Free energy: $A_0 = 0.01$; $\gamma = 3.0$; $q_0 = 0.13884$; (and $p = 45.25$
is $32\sqrt{2}$);
$\kappa_1 = \kappa_2 = 0.00706095$;
Chirality $\bar{\kappa} = 0.7$ Reduced temperature $\tau = 0.0$.

Amplitude of order $-0.2$; Computed surface order:
Rotational diffusion $\Gamma = 0.5$

Wall irrelevant

Weak colloid anchoring $W = 0.000706095$, ie., $W/\kappa = 0.1$ length
and $Wa / \kappa = 2.3$.

Strong colloid anchoring $W = 0.00706095$, i.e., $W/\kappa = 1.0$
and $Wa/\kappa = 1.0$.

JUHO (QUENCH)

Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = \eta$;
Temperature: $T = 2.133333 10^{-9}$.

Free energy: $A_0 = 0.01$; $\gamma = 3.1$; $q_0 = 0.13884$; (and $p = 45.25$
is $32\sqrt{2}$);
$\kappa_1 = \kappa_2 = 0.00706095$;
Chirality $\bar{\kappa} = 0.68862$ Reduced temperature $\tau = -2.90322$.

### 2012-09-04 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1776 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %  Add 'showkeys' option to make keywords appear
-\documentclass[aps,pre,reprint,superscriptaddress]{revtex4}
+\documentclass[aps,pre,reprint,superscriptaddress, twocolumn]{revtex4}
 %\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
@@
 \begin{eqnarray}
-f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
+f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2\nonumber\\
+&-&\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
 f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
@@
 \left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
-\label{eqn2}
+\label{op-eom}
 \end{equation}
-The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
+The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
 The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
@@
 \mathrm{Tr}({\mathbf Q W}),
-\label{eqn3}
+\label{sw}
 \end{eqnarray}
@@
 \mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
-\label{eqn4}
+\label{molfield}
 \end{equation}
-The rotational diffusion constant $\Gamma$ in Eq.~\ref{eqn2} is proportional
+The rotational diffusion constant $\Gamma$ in Eq.~\ref{op-eom} is proportional
 to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
@@
 \partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
-= \partial_\beta \Pi_{\alpha \beta}
-+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
+&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha u_\beta \nonumber\\
+& &\hspace*{-1.5cm} +\; \partial_\beta u_\alpha
 +(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
-\label{eqn6}
+\label{NSE}
 \end{eqnarray}
-The final term in Eq.~\ref{eqn6} arises from the Chapman-Enskog expansion
+The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
 of the LB equations \cite{Denniston:2001}.
 At low flow rates the fluid can be considered as incompressible, so that the
-last term on the right hand side of Eq.~\ref{eqn6} remains small.
+last term on the right hand side of Eq.~\ref{NSE} remains small.
 $\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
@@
 \Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
--\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
-&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
-&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
-\label{eqn7}
+-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
+&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
+&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
+- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
+\label{Pi}
 \end{eqnarray}
-In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
+In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} is reduced to the
 scalar pressure which, in a system at rest, is constant to a very good
@@
 
-The system of coupled partial differential equations Eqs.~\ref{eqn2}
-and~\ref{eqn6} is solved by means of a
+The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation accoring to
+\begin{equation}
+X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
+\label{spectrum}
+\end{equation}
+wheras we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
+$N$ was the number of points in the time series, typically one for every 100 LB time steps.
+
+The system of coupled partial differential equations Eqs.~\ref{op-eom}
+and~\ref{NSE} is solved by means of a
 hybrid scheme~\cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
 with predictor-corrector scheme for the continuity equation and
-Eq.~\ref{eqn6}, and a finite difference scheme for the equation of motion of
-the tensor order parameter Eq.\ref{eqn2}. More details on the algorithm can
+Eq.~\ref{NSE}, and a finite difference scheme for the equation of motion of
+the tensor order parameter Eq.\ref{op-eom}. More details on the algorithm can
 be found in~\cite{Denniston:2001, Denniston:2004}.
@@
 
-\subsubsection{Low and intermediate shear rates}
-
-Fig \ref{bp2-med} shows the disclination network 
-undergoing an affine transformation under shear. The disclination lines break up and 
-reconnect further downstream forming a periodically recurring pattern. 
-The recurrence time in flow direction is $\tau_F=p/(2\, v_{LE})=p/(2\,\gd\, \Delta y_{LE})= 1/\gd$.
-Apart from the affine transformation the general appearance is very close to that 
-of the quiescent equilibrium phase.
-
-\begin{figure*}[h]
-\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
-of shear-induced transformations in the steady state at 
-$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
-is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
-The boundary conditions impose flow in such a way that network moves 
-to the right in the upper half and to the left in the lower half of the picture.}
-\label{bp2-med}
-\end{figure*}
-
-Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
-the gradient direction in the direction of vorticity.
-Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
-geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
-positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
-making the network travel along the z-direction.
-The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.
-
 A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
@@
 
-\begin{figure*}[h]
+\begin{figure}[ht!]
 \includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
@@
 \label{bp2-rheo}
-\end{figure*}
+\end{figure}
+
+\subsubsection{Low and intermediate shear rates}
+
+Fig \ref{bp2-med} shows the disclination network 
+undergoing an affine transformation under shear. The disclination lines break up and 
+reconnect further downstream forming a periodically recurring pattern. 
+The recurrence time in flow direction is $\tau_F=p/(2\, v_{LE})=p/(2\,\gd\, \Delta y_{LE})= 1/\gd$.
+Apart from the affine transformation the general appearance is very close to that 
+of the quiescent equilibrium phase.
+
+\begin{figure}[ht]
+\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
+\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
+\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
+\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
+of shear-induced transformations in the steady state at 
+$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
+is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
+The boundary conditions impose flow in such a way that network moves 
+to the right in the upper half and to the left in the lower half of the picture.}
+\label{bp2-med}
+\end{figure}
+
+Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
+the gradient direction in the direction of vorticity.
+Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
+geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
+positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
+making the network travel along the z-direction.
+The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.
 
@@
 
-\begin{figure*}[h]
+\begin{figure*}[ht]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
@@
 
-\begin{figure}[h]
+\begin{figure*}[h]
 \includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
@@
 \label{bp2-high}
-\end{figure}
+\end{figure*}
+
 \clearpage
@@
 Contrary to BPII, they are well separated and do not intersect. 
-This difference in topology leads to some stiking differences in the flow behaviour of both blue phases 
-as we will see in the following.
+There is strong evidence that these topological differences result in 
+a very different rheological response of BPI. In fact BPI exhibits some 
+peculiarities that are absent in BPII.
 
-We can identify three different regimes in the range of applied shear rates:
-At high shear rates a breakup of the network structure occurs, similar to that seen in BPII. 
-At intermediate shear rates BPI features seemingly chaotic but periodically recurring conformations.
-At low shear rates the network undergoes first an affine transformation, but becomes unstable
-at larger strains and adopts an amorphous configuration with well separated disclination lines.
+Just as before in a kind of overview we present rheological key quantities first and 
+will address the detail aspects further below. 
+Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
+strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
+are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
+relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
+the apparent viscosity varies by almost a factor 8 during one cycle.
+The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
+follow a simple power law.
+
+The inset of shows the flow curve, i.e. shear stress over shear rate.
+The data points are averages over one cycle, whereas the error bars indicate the maximum and
+minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
+At high shear rates the network breaks up similar to BPII. The director field in this state
+is effectively two-dimensional and translationally invariant along the neutral direction.
+At intermediate shear rates the BPI network flows with periodically recurring conformations.
+These complex flow-induced conformations are accompanied by large fluctuations that increase 
+in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
+of the liquid crystal during one cycle is actually negative. 
+Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
+Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
+end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.
+
+\begin{figure}[ht]
+\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right).
+The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.
+Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle).
+The error bars give the minimum and maximum shear stress during one cycle.}
+\label{bp1-rheo}
+\end{figure}
+
+The Fourier analysis of the time-dependent deviatoric stress $\Pi_{xy}$ 
+offers another perspective on this intricate flow behaviour at high and intermediate shear rates. 
+The spectrum $X_\omega$ according to Eq. \ref{spectrum} is shown in Fig. \ref{bp1-spectrum}.
+We observe a strong contribution of the first harmonic, which is directly related to the 
+recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ 
+via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
+As the shear rates we applied always differ by a factor 2 the Fourier coefficients of the 
+individual runs with different shear rate coincide in such a way that the
+the fundamental mode is the same frequency as the first overtone of the next lower shear rate
+or the third overtone of the next-to-next lower shear rate 
+(viz. peak positions at $\gd=7.81, 3.91$ and $1.95\e{-5}$).
+Moreover, the magnitude of higher Fourier modes decreases monotonously.   
+However, the inset shows that this characteristic frequency relation is violated 
+for shear rates above $\gd=7.81\e{-5}$.
+Although the velocity of the primary flow component in x-direcition, listed in Tab. \ref{tab1},
+ is very well in accordance with the imposed shear rate, the fundamental mode appears 
+now at the same frequency.
+The contribution of higher Fourier modes increases and even exceeds that of
+the fundamental mode, which gives the shear stress generally a more jittering appearance.
+At the highest imposed shear rate where the breakup occurs the Fourier spectrum 
+becomes simple again and features a dominating fundamental mode at the 
+expected frequency just as described above for $\gd\le7.81\e{-5}$.
+The breakup occurs at exactly the same flow rate as those of BPII, 
+suggesting that this can be traced back to the rotational dynamics 
+of the order parameter.
+
+\begin{figure}[ht]
+\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
+\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate.
+The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed.
+Instead the spectral contribution of higher modes grows until finally the breakup takes place.}
+\label{bp1-spectrum}
+\end{figure}
+
+\subsubsection{Low shear rates}
+
+The rheological response of BPI at low shear rate is strikingly 
+different from that of BPII.
+We want to explain these differences on the basis of a typical 
+appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
+Three different stages can be distinguished. 
+With the onset of shearing the disclination lines 
+in BPI get more and more squeezed together which is incommensurable with 
+the defect topology of this blue phase. 
+Consequently, the network adopts a flow-induced conformation which consists 
+of intertwined helices that stretch during the affine transformation.
+This helical conformation that emerges already at strains $\gamma\simeq1$ 
+is shown in Fig. \ref{bp1-low} (top row).
+At a particular point in the cycle a non-helical configuration is
+observed (bottom left) before the network quickly returns
+to its helical state.
+There is no perceivable movement of the network in vorticity 
+direction at any stage of the cycle.
+
+However, this mode of flow proves unstable as after a few cycles 
+distortions appear which lead to further destabilisation.
+Finally, the network
+
+\begin{figure*}[ht]
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).}
+\label{bp1-low}
+\end{figure*}
+
+\begin{figure}[ht]
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{}
+\label{bp1-fe-yield}
+\end{figure}
 
@@
 
-\begin{figure*}[h]
+\begin{figure*}[ht]
 \includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
@@
 
-The apparent viscosity of BPI in simple shear flow, depicted in Fig. \ref{bp1-rheo}, is larger for The degree of shear thinning, also visible in the inset in terms of average shear stress over
-flow rate
-
-\begin{figure}[h]
-\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
-\label{bp1-rheo}
-\end{figure}
-
-Runs at smaller system sizes suggest that the individual conformation 
-can depend sensitively on the system size. Fig. \ref{bp1-2uc4uc}
-
-\begin{figure}[h]
-\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{}
-\label{bp1-2uc4uc}
-\end{figure}
-
-\begin{equation}
-X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N)
-\end{equation}
-
-A Fourier analysis of the time series of the deviatoric chemical stress $\sigma_{eff,xy}$ or alternatively the free energy density $f$ can provide a further insight into the flow behaviour of BPI.
-Fig. \ref{fig9} shows the Fourier spectrum of the time series of the chemical stress.
-For simplicity we have always doubled the shear rate, so that the first overtone ocoalesces with the fundamental mode of the next highest shear rate. 
-At low shear rates we observe a strong contribution of the first harmonic, which is directly related to the recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
-The magnitude of higher harmonics decreases monotonously with increasing frequency.
-Interestingly the inset of Fig.
-\ref{fig9} shows that at intermediate shear rates the contribution of higher harmonics becomes more pronounced until after a kind of gap and similar break up the fundamental mode dominates again and the former characteristics are restored.
-Remarkably the break up in case of BPI occurs in exactly the same flow rate window as BPII.
-
-\begin{figure}[h]
-\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
-\label{bp1-spec}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%
-The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
-Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
-Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
-
-\begin{figure*}[h]
+\begin{figure*}[ht]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
@@
 
-\subsubsection{Low shear rates}
+Runs at smaller system sizes suggest that the individual conformation 
+can depend sensitively on the system size. Fig. \ref{bp1-low-fe}
 
-At low shear rates the conformational changes of BPI in shear flow are very similar to 
-that of BPII descibed above. Fig. \ref{bp1-low} shows how the network undergoes an affine 
-transformation during one cycle. However, contrary to BPII the disclination lines of BPI 
-do not resemble those of the quiescent state, although they have the same topological
-structure. At one particular point (top-left and top-centre) 
-the disclination network are intertwined helices with the helical axis aligned
-along the gradient direction. This shear-induced conformation forms shortly after the
-flow has been switched on at strains $\gamma\simeq1$.
-During one cycle there is no perceivable movement of the network in vorticity direction.
+\begin{figure}[ht]
+\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
+\caption{}
+\label{bp0-2uc4uc-fe}
+\end{figure}
 
-\begin{figure}[h]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
-\caption{Disclination network of BPI at low shear rate: Depicted is a sequence of affine, shear-induced transformations at shear rate $\dot{\gamma}=4.88\e{-6}$ and time step $t=4.0, 4.0, 6.0, 7.25, 7.50, 7.75\e{5}$ (top-left to bottom-right).
-The horizontal and the vertical direction are the flow and the gradient direction in all but the top-centre picure.
-There, the network is depicted in gradient direction with the direction of vorticity oriented along the vertical axis.}
-\label{bp1-low}
+\begin{figure}[ht]
+
+\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
+\label{bp1-spec}
 \end{figure}
+
+The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
+Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
+Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
 
@@
 
-\begin{figure}[h]
+\begin{figure*}[ht]
 \includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
@@
 The colour code gives the magnitude of the in-plane component of the director field.
-The data is translationally invariant along the vorticity direction.}
+The data is translationally invariant in vorticity direction.}
 \label{bp1-high}
-\end{figure}
+\end{figure*}
 

### 2012-09-06 trunk/papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1779 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{amsmath}
+\usepackage{float}
 
@@
 \newcommand{\gd}{\dot{\gamma}}
+\newcommand{\com}[1]{{\bf {#1}}}
 
@@
 % \affiliation can be followed by \email, \homepage, \thanks as well.
+
 \author{O. Henrich}
 \email[Corresponding author: ]{o.henrich@ucl.ac.uk}
-\affiliation{Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-\affiliation{Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}
+\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
 
 \author{K. Stratford}
-\affiliation{Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
 
 \author{P.V. Coveney}
-\affiliation{Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}
+\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
 
 \author{M.E. Cates}
-\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
 
 \author{D. Marenduzzo}
-\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
+\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
 
@@
 % insert suggested keywords - APS authors don't need to do this
-%\keywords{}
+\keywords{blue phase, cholesterics, chiral nematics, rheology}
 
@@
 \section{Results and Discussion}
+
+\begin{table*}[htpb]
+\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
+\hline
+$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+\hline
+BPI \\
+\hline
+0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
+7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
+\hline
+BPII \\
+\hline
+0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+\hline
+\end{tabular}
+\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
+\label{tab1}}
+\end{table*}
 
@@
 
-We start the discussion with BPII. Fig \ref{bp2-med} shows the disclination network 
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
+\label{bp2-rheo}
+\end{figure}
+
+We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
+BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
+intersect and form a characteristic network of nodes.
+
+A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
+The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
+the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
+the picture on the right shows the same data versus accummulated total strain.
+For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
+is due to the periodic break-up and reconnection of the network under shear flow.
+
+The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
+The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
+a very good approximation, exibiting a remarkably small degree of shear-thinning.
+This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.
+
+\subsubsection{Low and intermediate shear rates}
+
+Fig \ref{bp2-med} shows the disclination network 
 undergoing an affine transformation under shear. The disclination lines break up and 
 reconnect further downstream forming a periodically recurring pattern. 
-Apart from the affine transformation the general appearance is very close to that 
-of the quiescent equilibrium phase.
+The recurrence time in flow direction $\tau_F=p/(2\, v_{LE})= 1/\gd$.
 
-\begin{figure*}[h]
-\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
+\begin{figure}[htpb]
+\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
+\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
+\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
 \caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
@@
 $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
-is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
-The boundary conditions are applied flow takes place in such as way that network moves 
-to the right in the upper half and to the left in the lower half.}
+is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. The boundary conditions impose flow in such a way that network moves 
+to the right in the upper half and to the left in the lower half of the picture.}
 \label{bp2-med}
+\end{figure}
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
+\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
+underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
+i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
+The vertical and horizontal direction are the gradient and vorticity direction, respectively.
+The colour code gives the magnitude and sign of the component in vorticity direction.
+The snapshot shows a typical frame during a periodically recurring sequence.
+The network on the left with positive helicity travels rightwards, whereas the one one the right
+moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
+to reconnect in flow direction.}
+\label{bp2-velo}
 \end{figure*}
 
+Apart from the affine transformation the general appearance is very close to that 
+of the quiescent equilibrium phase.
 Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
@@
 making the network travel along the z-direction.
-The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.
+
+The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
+i.e. it is takes exactly six unit cell displacements in flow directions 
+until the network has moved one unit cell in voriticity direction.
 
@@
 
-\begin{figure*}[h]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
-\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
-underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
-i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
-The vertical and horizontal direction are the gradient and vorticity direction, respectively.
-The colour code gives the magnitude and sign of the component in vorticity direction.
-The snapshot shows a typical frame during a periodically recurring sequence.
-The network on the left with positive helicity travels rightwards, whereas the one one the right
-moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
-to reconnect in flow direction.}
-\label{bp2-velo}
-\end{figure*}
+\subsubsection{High shear rates}
 
-If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
+If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation in shear flow that we
 descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
@@
 
-In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
+In our scenario the break-up occurs for shear rates between $3.12\e{-4}\le\gd_c\le 6.25\e{-4}$.
 However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
@@
 
-\begin{figure}[h]
-\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom).
-The images show 1D-cuts as the configuration is translationally invariant in flow and gradient direction.
-The configuration is a travelling helical wave along the vorticity or z-direction.
-The colour code indicates the orientation in x-y direction.}
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
+\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+The configuration is a helical wave that travels along the direction of vorticity.
+It is translationally invariant in gradient and flow direction.}
 \label{bp2-high}
@@
 
-A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
-The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
-the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
-the picture on the right shows the same data versus accummulated total strain.
-For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is due to the periodic break-up and reconnection of the network under shear flow.
+\subsection{Blue Phase I}
 
-The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
-The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
-a very good approximation, exibiting a remarkably small degree of shear-thinning.
-This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.
+In BPI the disclination lines are arranged in a body-centered-cubic array.
+Contrary to BPII, they are well separated and do not intersect. 
+There is strong evidence that these topological differences result in 
+a very different rheological response of BPI. In fact BPI exhibits some 
+peculiarities that are absent in BPII.
 
-\begin{figure*}[h]
-\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
-\label{bp2-rheo}
+Just as before in a kind of overview we present rheological key quantities first and 
+will address the detail aspects further below. 
+Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
+strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
+are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
+relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
+the apparent viscosity varies by almost a factor 8 during one cycle.
+The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
+follow a simple power law.
+
+The inset of shows the flow curve, i.e. shear stress over shear rate.
+The data points are averages over one cycle, whereas the error bars indicate the maximum and
+minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
+At high shear rates the network breaks up similar to BPII. The director field in this state
+is effectively two-dimensional and translationally invariant along the neutral direction.
+At intermediate shear rates the BPI network flows with periodically recurring conformations.
+These complex flow-induced conformations are accompanied by large fluctuations that increase 
+in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
+of the liquid crystal during one cycle is actually negative. 
+Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
+Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
+end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right).
+The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.
+Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle).
+The error bars give the minimum and maximum shear stress during one cycle.}
+\label{bp1-rheo}
+\end{figure}
+
+The Fourier analysis of the time-dependent deviatoric stress $\Pi_{xy}$ 
+offers another perspective on this intricate flow behaviour at high and intermediate shear rates. 
+The spectrum $X_\omega$ according to Eq. \ref{spectrum} is shown in Fig. \ref{bp1-spectrum}.
+We observe a strong contribution of the first harmonic, which is directly related to the 
+recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ 
+via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
+As the shear rates we applied always differ by a factor 2 the Fourier coefficients of the 
+individual runs with different shear rate coincide in such a way that the
+the fundamental mode is the same frequency as the first overtone of the next lower shear rate
+or the third overtone of the next-to-next lower shear rate 
+(viz. peak positions at $\gd=7.81, 3.91$ and $1.95\e{-5}$).
+Moreover, the magnitude of higher Fourier modes decreases monotonously.   
+However, the inset shows that this characteristic frequency relation is violated 
+for shear rates above $\gd=7.81\e{-5}$.
+Although the velocity of the primary flow component in x-direcition, listed in Tab. \ref{tab1},
+ is very well in accordance with the imposed shear rate, the fundamental mode appears 
+now at the same frequency.
+The contribution of higher Fourier modes increases and even exceeds that of
+the fundamental mode, which gives the shear stress generally a more jittering appearance.
+At the highest imposed shear rate where the breakup occurs the Fourier spectrum 
+becomes simple again and features a dominating fundamental mode at the 
+expected frequency just as described above for $\gd\le7.81\e{-5}$.
+The breakup occurs at exactly the same flow rate as those of BPII, 
+suggesting that this can be traced back to the rotational dynamics 
+of the order parameter.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
+\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate.
+The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed.
+Instead the spectral contribution of higher modes grows until finally the breakup takes place.
+The frequency is given in units of $10^2\cdot 2\pi/T$ with $T$ as period of one cycle in LB time steps.}
+\label{bp1-spectrum}
+\end{figure}
+
+\subsubsection{Low shear rates}
+
+The rheological response of BPI at low shear rate which exhibits a
+dynamic yield stress is strikingly different from that of BPII.
+An explanation for this strange behaviour can be found by 
+looking more closely at the average deviatoric stress and 
+free energy density, as shown in Fig. \ref{bp1-fe-yield}.
+When the quiescent and equilibrated BPI network begins to flow
+the shear stress increases steeply and goes through a global maximum.
+Shortly after it goes negative (the total shear stress is still
+positive as the backgound viscosity has been subracted) and positive
+again, forming a complete cycle. 
+This cycle is repeated for a couple of times before 
+the amorphous configuration emerges and any regular pattern
+disappears.
+The negative branch in the deviatoric stress is connected
+to a local maximum followed by a local minimum in the average free energy.
+Hence, the region of negative stress can be 
+regarded as a phase during which elastic forces excert a 
+'pull' onto the network while it tries to reach a more 
+favourable configuration. However, the low flow rate
+causes this unstable intermediate state to be exposed to these
+pulling elastic forces for a relatively long time, long enough to 
+create small distortions which finally break the symmetry of the
+flow-induced configuration.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density.
+This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
+\label{bp1-fe-yield}
+\end{figure}
+
+We want to explain these differences on the basis of a typical 
+appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
+Three different stages can be distinguished. 
+With the onset of shearing the disclination lines 
+in BPI get more and more squeezed together which is incommensurable with 
+the defect topology of this blue phase. 
+Consequently, the network adopts a flow-induced conformation which consists 
+of intertwined helices that stretch during the affine transformation.
+This helical conformation that emerges already at strains $\gamma\simeq1$ 
+is shown in Fig. \ref{bp1-low} (top row).
+At a particular point in the cycle a non-helical configuration is
+observed (bottom left) before the network quickly returns
+to its helical state.
+There is no perceivable movement of the network in vorticity 
+direction at any stage of the cycle.
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).
+All pictures have the vie along the vorticity direction apart from the one at the top centre which shows the network in gradient direction.}
+\label{bp1-low}
 \end{figure*}
 
-\clearpage
+However, this mode of flow proves unstable as after a few cycles 
+distortions appear which lead to further destabilisation.
+Finally, the network loses any residual order and 
+transforms into an amorphous state with virtually constant
+yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LU.
 
-\subsection{Blue Phase I}
+\subsubsection{Intermediate shear rates}
 
-The flow behaviour of BPI at low shear rates is is similar deviates significantly from that of BPII in a number of points.
+Intermediate shear rates in the sense of this study of BPI are characterised 
+by the missing negative branch in the deviatoric stress. It is unclear 
+why this feature is missing at higher flow rates. A possible explanation is that 
+above a certain flow rate the network simply skips the comparably small 
+window of negative stress in Fig. \ref{bp1-fe-yield} and finds a quicker 
+route to get back on the positive branch.
 
-\begin{figure}[h]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-600k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-775k_run1115.png}
-\caption{Disclination network of BPI at low shear flow:}
-\label{bp1-low}
-\end{figure}
+Fig. \ref{bp1-med} shows snapshots of BPI in steady shear flow. 
+Contrary to BPII the flow-induced conformations of BPI differ significanly 
+from the appearance of BPI at equilibrium. The depicted 
+state is rather surprisingly still kinematically connected to the equilibrium state.
+When the shear flow is switched off the flow-induced configuration reverts
+to its original configuration of a BPI network at rest.
+%%%%%%%%%%%%
+%The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
+%%%%%%%%%%%
+After each cycle the position of the network is displaced with respect to the beginning of the cycle.
+However, neither the magnitude nor the direction of the offset obey a clear rule,
+are different and non-integer for every shear rate in this regime of flow. 
+This suggests the described phenomenon constitutes an instance of rheochaos in this 
+regime of liquid crystalline flow.
 
-First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
-This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
-Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig11}, but this time seen along the direction of flow.
-A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
-The initial and final state in this sequence correspond to half a cycle. 
-This can be easily verified by comparing the position of the disclination network at the left or right boundary.
-BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.
-
-\begin{figure*}[h]
+\begin{figure*}[htpb]
 \includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
@@
 \includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
-\caption{Disclination network of BPI in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$.
-The horizontal direction is the vorticity direction, whereas the velocity gradient is oriented along the vertical direction.}
+\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page.
+The network is displaced during every cycle, but the magnitude and direction of the offset appear to be chaotic.}
 \label{bp1-med}
@@
 
-The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
-Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
-Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
+Fig.
+\ref{bp1-velo} depicts the secondary velocity components $v_y$ and $v_z$ in the same fashion as Fig.
+\ref{bp2-velo} for BPII.
+Again, the orientation of the components are as mirror images of each other if the helicity of the cholesteric changes from left- to right-handed.
+The emerging pattern is different from the one observed before in BPII, but agrees with
+the previous result in the sense that secondary velocities are largest in the regions
+where large conformational changes take place.
 
-\begin{figure*}[h]
+\begin{figure*}[htpb]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a snapshot of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
 The colour code gives the magnitude and sign of the z-component.
@@
 
-Despite qualitative similarities between the BPI and BPII, the quantitative analysis reveals striking differences.
-In Fig. \ref{fig8} we show the effective viscosity $\eta_{eff}$ over time and flow curves $\sigma_{chem}(\dot{\gamma})$. 
-BPI exhibits pronounced shear-thinning contrary to BPII, whereas the shape of the curves in no longer symmetric.
-In fact on approaching the critical shear rate that led in the case of BPII to the disruption of the network, the curves become increasingly complex but remain fully periodic.
-The effective viscosity drops by about an order of magnitude between the lowest and the highest shear rates.
-Note that the scale has been adapted to allow for the different magnitudes.
-The inset can be directly compared with that of Fig. \ref{fig4}.
-Interestingly the difference between minima and maxima of the chemical shear stress is much larger than for BPII and for the lowest shear rate we applied the absolute minimum takes even negative values, which are not shown in Fig.
-\ref{fig4}.
-Our understanding of this phenomenon is an interplay of the elastic forces and the individual order paramter configuration of the liquid crystal.
-If the system is close to a local minimum of the free energy but moving away from it the increasing strain pumps elastic energy into the system. 
-This results in positive chemical stresses as the system tries to withstand the deformation.
-However, if the system has sufficiently transformed and is about to overcome a local maximum, it will be pulled towards the next local minumum, which can lead temporarily to negative chemical stresses.
-The inset shows that the chemical stress decreases monotonically with decreasing flow rate, but does not follow a simple power law dependence.
+The complexity of flow-induced states and the seemingly chaotic but 
+periodic movement of the BPI network raises the question how robust 
+these results are with respect to the system size.
+Although the pitch length of 64 LU and the equivalent of 32 LU per
+unit cell provide enough resolution to track the motion of the 
+network in flow it is not clear whether 64 unit cells in total are enough to 
+mimick true bulk behaviour.
 
-Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
-Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
-Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
-They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.
+Unfortunately the computational costs are still prohibitive for a run 
+with 512 unit cells, an equivalent of 8 unit cells along each coordinate 
+direction. Therefore we opted for a comparison with a smaller run consiting
+of 8 unit cells, 2 along each dimension.
+Fig. \ref{bp1-2uc4uc} shows a comparison of the apparent viscosity $\eta_{app}$ 
+of runs with 8 and 64 unit cells at flow rates 
+$\gd=9.76\e{-6}$ and $1.95\e{-5}$, respectively.
+We do not visualise the conformations directly but rather use the fact that 
+they are coded in the shape of each individual curve. Obviously there 
+are stiking differences between 
+both system sizes, but the recurrence period is hardly affected. In the slower
+run the smaller system apparently follows the same branch for a while but then
+deviates from the larger run. 
+This result indicates that we cannot know with certainty if the particular flow patterns 
+we find at runs with 64 unit cells do not suffer from the imposed periodicity.
+Nevertheless, the chaotic behaviour and the recurrency seem to be a generic
+feature of BPI in this flow regime.
 
-\begin{figure}[h]
-\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
-\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
-\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
+\caption{Apparent viscosity $\eta_{app}$ for 8 (solid lines) and 64 unit cells (dashed lines)
+at $\gd=9.76\e{-6}$ and $1.95\e{-5}$. The shape of each curve differs largely, but the recurrence frequencies are hardly affected by the system size.}
+\label{bp1-2uc4uc}
+\end{figure}
+
+\subsubsection{High shear rates}
+
+\begin{figure}[htpb]
+%\includegraphics[width=0.4\textwidth]{dir+z-300k_run916.png}
+\includegraphics[width=0.4\textwidth]{dir+z-301k_run916.png}
+\includegraphics[width=0.4\textwidth]{dir+z-302k_run916.png}
 \caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
 The colour code gives the magnitude of the in-plane component of the director field.
-The data is translationally invariant along the vorticity direction.}
+The data is translationally invariant in vorticity direction.}
 \label{bp1-high}
@@
 
-\begin{figure}[h]
-\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\sigma_{chem}(\dot{\gamma})$.}
-\label{bp1-rheo}
-\end{figure}
-
-A Fourier analysis of the time series of the deviatoric chemical stress $\sigma_{eff,xy}$ or alternatively the free energy density $f$ can provide a further insight into the flow behaviour of BPI.
-Fig. \ref{fig9} shows the Fourier spectrum of the time series of the chemical stress.
-For simplicity we have always doubled the shear rate, so that the first overtone ocoalesces with the fundamental mode of the next highest shear rate. 
-At low shear rates we observe a strong contribution of the first harmonic, which is directly related to the recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
-The magnitude of higher harmonics decreases monotonously with increasing frequency.
-Interestingly the inset of Fig.
-\ref{fig9} shows that at intermediate shear rates the contribution of higher harmonics becomes more pronounced until after a kind of gap and similar break up the fundamental mode dominates again and the former characteristics are restored.
-Remarkably the break up in case of BPI occurs in exactly the same flow rate window as BPII.
-
-\begin{figure}[h]
-\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
-\label{bp1-spec}
-\end{figure}
+A breakup of the network at large flow rates is also observed
+in BPI. Interestingly the window of critical shear rates 
+$3.125\e{-4}\le\gd_c\le 6.25\e{-4}$  is the same we find for BPII.  
+Contrary to BPII, which flows in form of a travelling helical wave
+(viz. Fig. \ref{bp2-high}) the preferred configuration is now
+a quasi two-dimensional state, shown in Fig. \ref{bp1-high} as 
+slices through the director field along the voritcity direction.      
+The sequence shows that BPI undergoes the affine transformation 
+in gradient-flow plane often seen in simple shear flow.
+There are distinctive regions where the predominant orientation 
+of the director is along the vorticity direction and others
+where it lies in flow-gradient plane. Occasionally while 
+flowing past each other this produces regions that 
+resemble local double twist.
 
@@
 % only one appendix.
-\appendix*
+%\appendix
 
-\begin{table*}
-\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
-\hline
-$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
-\hline
-BPI \\
-\hline
-0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
-3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
-\hline
-BPII \\
-\hline
-0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
-62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
-\hline
-\end{tabular}
-\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
-\label{tab1}}
-\end{table*}
 %\section{}

### 2012-09-11 papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1782 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+\begin{table*}[htpb]
+\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
+\hline
+$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+\hline
+BPI \\
+\hline
+0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
+7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
+\hline
+BPII \\
+\hline
+0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+\hline
+\end{tabular}
+\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
+\label{tab1}
+\end{table*}
+
 In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
-phases BPI and BPII. The thermodynamic states are characterised by an effective temperature 
-$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy terms.
-We opted for states sufficiently far away from the chiral-nematic-isotropic 
-where both quiescent blue phases are at equilibrium, namely
+phases BPI and BPII. 
+The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+is equal to about $10^8$~Pa. Further details about the conversion 
+from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.
+
+The thermodynamic states are characterised by an effective temperature 
+$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy.
+For the vast majority of simulations we opted for equilibrium states that are sufficiently 
+far away from the cholesteric-isotropic transition, namely
 $\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
-Simulations at metastable states at higher and lower temperatures and different chiralities showed 
-no differences in their generic flow behaviour.
-However, states closer to the phase boundary featuring higher free energy densities and 
-smaller average order parameters show a tendency to break up more easily under external driving forces.
+We performed also a number of simulations at metastable states at higher and lower temperatures 
+and different at different chiralities, but didn't find any significant differences in the 
+generic flow behaviour that is described below.
+However, for thermodynamic states that are closer to the phase boundary the disclination 
+network broke up into non-BP states at lower shear rates. This is not surprising as
+these states have on average higher free energy densities and smaller order parameters
+than those specified above and can therefore not withstand the same external forces.
 
-As is common in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with 
-analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, 
-and equilibrated these (for 5000 LB timesteps) prior to starting the shear. 
-During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
-After equilibration the redshift parameter was kept constant throughout the rest of the simulation.
-We chose a unit cell size of 16 lattice sites for BPII and 32 sites in case of BPI. 
-Runs with higher resolution confirmed that this was an appropriate minimal resolution to cover 
-the complex conformational changes and kinematic details of the blue phase networks.
+As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
+analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit 
+and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
+During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
+After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow.
+We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, which is the equivalent 
+of 4 unit cells along each coordinate direction and a total of 64 unit cells in our simulation box.
+Runs with higher resolution confirmed that the resolution was adequate to track  
+all kinematic details of the blue phase networks in shear flow.
 
-Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
-the top (bottom) half of the planes moving in positive (negative) x-direction.
-We applied a correction for spurious momentum currents that emerge due to interpolations accross 
+Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
+the top (bottom) half of the planes moving in positive (negative) x-direction and the 
+velocity gradient along the y-direction.
+We applied a correction for spurious momentum currents that emerge due to required interpolations across 
 the Lees-Edwards planes \cite{Henrich:2012a}.
 The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
-The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
-along the y- and z-axis.
+$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
+For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
+They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
+BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
 Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
 are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
-$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
-is equal to about $10^8$~Pa.
-More details about the conversion from LBU to SI units can be found in
-\cite{Henrich:2011a,Henrich:2010b}.
 
@@
 
-We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
-BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
-intersect and form a characteristic network of nodes.
-
-A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
-The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
-the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
-the picture on the right shows the same data versus accummulated total strain.
-For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is due to the periodic break-up and reconnection of the network under shear flow.
-
-The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
-The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
-a very good approximation, exibiting a remarkably small degree of shear-thinning.
-This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.
-
-\begin{figure}[ht!]
+\begin{figure}[htpb]
 \includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom).
+For clarity the individual curves have been shifted along the time and strain axis.
+The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) and break up of the network into a travelling helical wave (BPII-2, solid circle).}
 \label{bp2-rheo}
@@
 
-\subsubsection{Low and intermediate shear rates}
+We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
+than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
 
-Fig \ref{bp2-med} shows the disclination network 
-undergoing an affine transformation under shear. The disclination lines break up and 
-reconnect further downstream forming a periodically recurring pattern. 
-The recurrence time in flow direction is $\tau_F=p/(2\, v_{LE})=p/(2\,\gd\, \Delta y_{LE})= 1/\gd$.
-Apart from the affine transformation the general appearance is very close to that 
-of the quiescent equilibrium phase.
+As it might be easier to enter a discussion with the full picture in mind, we provide first a 
+quantitative analysis of the flow behaviour at all applied shear rates.
+Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
+deviatoric stress normalised to the isotropic background viscosity: 
+$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
+A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
+any additional contribution from the liquid crystal.
+The top picutre shows the date over time, whereas the graph at the bottom gives 
+the same data versus accummulated strain.
+For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
+is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
 
-\begin{figure}[ht]
+The inset shows a kind of flow curve, defined as time averages of 
+the deviatoric stress tensor $\bar{Pi}_{xy}$ as a function of shear rate $\gd$.
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
+For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
+$a=0.35, b=0.95$ describes the data to a very good approximation. 
+Hence, the degree of shear-thinning is remarkably small, but may depend
+on the exact direction in which the strain is applied, an aspect that we     
+leave for future work.
+
+For shear rates $\gd\ge6.25\e{-4}$ the network of disclination lines
+breaks up and oscillations in the stress signal are absent.
+In this regime of shear rates, to which we refer as BPII-2, the 
+blue phase dissolves into a simple cholesteric liquid 
+crystal. The mode of flow is a travelling helical wave with the helical axis 
+oriented along the vorticity direction.
+
+In the next paragraph we will investigate this in more detail by looking at the 
+kinetics of the disclination network.
+
+\subsubsection{Regime BPII-1: low and intermediate shear rates}
+
+Fig. \ref{bp2-med} shows the disclination network in shear flow as
+it undergoes an affine transformation. The disclination lines break up and 
+reconnect further downstream forming a periodically recurring pattern with 
+recurrence time $\tau_F=1/\gd$ in flow direction.
+
+\begin{figure}[htpb]
 \includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
@@
 \includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
-of shear-induced transformations in the steady state at 
-$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
-is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
-The boundary conditions impose flow in such a way that network moves 
-to the right in the upper half and to the left in the lower half of the picture.}
+\caption{Disclination network of BPII in shear flow: 
+The pictures show a typical sequence of snapshots in the steady state 
+at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
+gradient is oriented along the vertical direction, whereas the 
+horizontal direction is the flow direction. Lees-Edwards boundary 
+conditions have been imposed in such a way that network moves to the 
+right in the upper half and to the left in the lower half of the picture.}
 \label{bp2-med}
@@
 
-Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
-the gradient direction in the direction of vorticity.
-Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
-geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
-positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
-making the network travel along the z-direction.
-The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.
-
-Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
-Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
-This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
-Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
-When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
-and the sense of motion also changes.
-Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
-individual cycles.
-Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
-All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
-There is only a slight imbalance between the maximum and minimum velocity in z-direction.
-Hence the movement of the network in vorticity direction is predominantly permeative.
-Note that the imbalance is exaclty retained if the helicity is inverted.
-
-\begin{figure*}[ht]
+\begin{figure*}[htpb]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
@@
 The network on the left with positive helicity travels rightwards, whereas the one one the right
-moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
-to reconnect in flow direction.}
+moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network to reconnect in flow direction.}
 \label{bp2-velo}
@@
 
-\subsubsection{High shear rates}
+The general appearance of the flowing network is apart from the affine transformation 
+very close to that of the quiescent blue phase at equilibrium. It is the same for 
+all shear rates in regime BPII-1 and a first explanation why shear-thinning is so weak.
 
-If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
-descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
-The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
-a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
-invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
-before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
-shear rates \cite{Rey:1996a, Rey:1996b}.
+Interestingly, while being displaced with the flow the entire network moves perpendicular 
+to the flow and velocity gradient in the vorticity direction.
+A similar behaviour has been recently observed for blue phases in shear flow and 
+confined geometries ~\cite{Henrich:2012b}.
+But contrary to the stick-slip motion that has been reported there the movement is 
+steady. It occurs in such a manner that the positions of breakup and reconnection 
+of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
+travel along the z-direction.
+The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
+i.e. it takes a displacement of six unit cells in flow directions 
+until the network has moved one unit cell in voriticity direction.
 
-In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
-However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
-such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
-Due to a lack of computational resources we have not investigated these aspects any further.
+Fig. \ref{bp2-velo} shows the disclination network sometime between reconnecting 
+and the next breakup with superimposed velocity vectors.
+Shown are what we may call secondary velocity components, obtained by projecting the
+velocity onto a plain perpendicular to the flow direction.
+This allows to get rid of the dominating velocity component $v_x$ and to
+visualise the two much smaller components $v_y$ and $v_z$.
+The magnitude of the secondary components is typically in the range of a 
+few percent of the primary flow component.
 
-\begin{figure*}[h]
-\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+Characteristic bands are visible, which are oriented along the vorticity direction and
+normally hidden behind the primary flow component.
+When the helicity of the underlying cholesteric phase is changed from left- to a 
+righthanded the pattern becomes a mirror image of the former one. The sign of 
+$v_z$ change and also the sense of motion is inverted.
+Further quantitative evidence for a direct link between the sense of motion 
+and the helicity can be gained by time-averaging over individual cycles.
+Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
+of the velocity components.
+All values for the two runs with inverted helicity are identical apart 
+from a change of sign in the z-components.
+There is only a slight imbalance between the maximum and minimum velocities 
+in z-direction, suggesting that the movement of the network in vorticity 
+direction is permeative. Note that this imbalance is retained under inversion of
+helicity.
+
+\subsubsection{Regime BPII-2: high shear rates}
+
+If the flow rate exceeds a certain critical value BPII does not undergo the 
+affine transformation descibed above and adopts another mode of flow. 
+Fig. \ref{bp2-high} shows a sequence of cuts through the director 
+field at high shear rate. The entire blue phase network has broken up 
+into a simple cholesteric helix and flows as a travelling 
+helical wave with the helical axis oriented along the vorticity direction.
+The state is translationally invariant in flow and gradient direction.
+and is the preferred mode of flow of a cholesteric 
+liquid crystal at low Ericksen numbers below the uncoiling transition 
+to a nematic, flow-aligned state at even higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
+
+In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$.
+At the moment we do not have a hard criterium for the critical
+shear rate at hand. However, it seems reasonable that the critical shear rate 
+depends on thermodynamic and kinetic parameters such as the effective 
+temperature $\tau$, the chirality $\kappa$ and the rotational diffusion 
+constant $\Gamma$.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
+\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
+$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. 
 The configuration is a helical wave that travels along the direction of vorticity.
@@
 \label{bp2-high}
-\end{figure*}
-
-\clearpage
+\end{figure}
 
@@
 
-In BPI the disclination lines are arranged in a body-centered-cubic array.
-Contrary to BPII, they are well separated and do not intersect. 
-There is strong evidence that these topological differences result in 
-a very different rheological response of BPI. In fact BPI exhibits some 
-peculiarities that are absent in BPII.
+BPI has body-centered symmetry and contrary to BPII the disclination lines
+ are well separated and do not intersect.
+We believe that this topological difference is responsible for most of
+the differences between BPI and BPII regarding their rheological response.
 
-Just as before in a kind of overview we present rheological key quantities first and 
-will address the detail aspects further below. 
-Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
-strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
-are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
-relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
-the apparent viscosity varies by almost a factor 8 during one cycle.
-The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
-follow a simple power law.
+Again, we present key aspects in a kind of overview before we 
+address some features in more detail. Fig. \ref{bp1-rheo} shows the 
+apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
+as in Fig. \ref{bp2-rheo}.
+For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
+the disclination network occurs similar to and at the same critical shear rate as 
+the one observed in BPII-2 but with an order structure that is a quasi 
+two-dimensional.
 
-The inset of shows the flow curve, i.e. shear stress over shear rate.
-The data points are averages over one cycle, whereas the error bars indicate the maximum and
-minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
-At high shear rates the network breaks up similar to BPII. The director field in this state
-is effectively two-dimensional and translationally invariant along the neutral direction.
-At intermediate shear rates the BPI network flows with periodically recurring conformations.
-These complex flow-induced conformations are accompanied by large fluctuations that increase 
-in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
-of the liquid crystal during one cycle is actually negative. 
-Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
-Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
-end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.
+In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
+deviatoric stress that are not sinusoidal, but show an increasingly complex
+pattern up to the breakup point.
+For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
+they become  unregular and finally disappear after a few
+cycles. Shear-thinning is slightly stronger than in BPII. 
+The average shear stress during one cycle can be fitted reasonably 
+well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
 
-\begin{figure}[ht]
+\begin{figure}[htpb]
 \includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right).
-The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.
-Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle).
-The error bars give the minimum and maximum shear stress during one cycle.}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (top) 
+and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
+Depending on the configuration in steady shear flow three different regimes 
+can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
+steady flow with recurring patterns (BPI-2, open circles) and 
+breakup of the disclination network (BPI-3, solid circle). 
+The error bars represent the minimum and maximum shear stress 
+during one cycle. For clarity the individual curves have been shifted along
+the time and strain axis.}
 \label{bp1-rheo}
@@
 
-\begin{figure}[ht]
+\begin{figure}[htpb]
 \includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
@@
 The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed.
-Instead the spectral contribution of higher modes grows until finally the breakup takes place.}
+Instead the spectral contribution of higher modes grows until finally the breakup takes place.
+The frequency is given in units of $10^2\cdot 2\pi/T$ with $T$ as period of one cycle in LB time steps.}
 \label{bp1-spectrum}
@@
 
-\subsubsection{Low shear rates}
+\subsubsection{Regime BPI-1: low shear rates}
 
-The rheological response of BPI at low shear rate is strikingly 
-different from that of BPII.
+The rheological response of BPI at low shear rate which exhibits a
+dynamic yield stress is strikingly different from that of BPII.
+An explanation for this strange behaviour can be found by 
+looking more closely at the average deviatoric stress and 
+free energy density, as shown in Fig. \ref{bp1-fe-yield}.
+When the quiescent and equilibrated BPI network begins to flow
+the shear stress increases steeply and goes through a global maximum.
+Shortly after it goes negative (the total shear stress is still
+positive as the backgound viscosity has been subracted) and positive
+again, forming a complete cycle. 
+This cycle is repeated for a couple of times before 
+the amorphous configuration emerges and any regular pattern
+disappears.
+The negative branch in the deviatoric stress is connected
+to a local maximum followed by a local minimum in the average free energy.
+Hence, the region of negative stress can be 
+regarded as a phase during which elastic forces excert a 
+'pull' onto the network while it tries to reach a more 
+favourable configuration. However, the low flow rate
+causes this unstable intermediate state to be exposed to these
+pulling elastic forces for a relatively long time, long enough to 
+create small distortions which finally break the symmetry of the
+flow-induced configuration.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density.
+This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
+\label{bp1-fe-yield}
+\end{figure}
+
 We want to explain these differences on the basis of a typical 
@@
 
-However, this mode of flow proves unstable as after a few cycles 
-distortions appear which lead to further destabilisation.
-Finally, the network
-
-\begin{figure*}[ht]
+\begin{figure*}[htpb]
 \includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
@@
 \includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
-\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).}
+\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).
+All pictures have the vie along the vorticity direction apart from the one at the top centre which shows the network in gradient direction.}
 \label{bp1-low}
@@
 
-\begin{figure}[ht]
-\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+However, this mode of flow proves unstable as after a few cycles 
+distortions appear which lead to further destabilisation.
+Finally, the network loses any residual order and 
+transforms into an amorphous state with virtually constant
+yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LU.
+
+\subsubsection{Regime BPI-2: intermediate shear rates}
+
+Intermediate shear rates in the sense of this study of BPI are characterised 
+by the missing negative branch in the deviatoric stress. It is unclear 
+why this feature is missing at higher flow rates. A possible explanation is that 
+above a certain flow rate the network simply skips the comparably small 
+window of negative stress in Fig. \ref{bp1-fe-yield} and finds a quicker 
+route to get back on the positive branch.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
 \caption{}
-\label{bp1-fe-yield}
+\label{bp1-1_bp1-2}
 \end{figure}
 
-\subsubsection{Intermediate shear rates}
+Fig. \ref{bp1-med} shows snapshots of BPI in steady shear flow. 
+Contrary to BPII the flow-induced conformations of BPI differ significanly 
+from the appearance of BPI at equilibrium. The depicted 
+state is rather surprisingly still kinematically connected to the equilibrium state.
+When the shear flow is switched off the flow-induced configuration reverts
+to its original configuration of a BPI network at rest.
+%%%%%%%%%%%%
+%The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
+%%%%%%%%%%%
+After each cycle the position of the network is displaced with respect to the beginning of the cycle.
+However, neither the magnitude nor the direction of the offset obey a clear rule,
+are different and non-integer for every shear rate in this regime of flow. 
+This suggests the described phenomenon constitutes an instance of rheochaos in this 
+regime of liquid crystalline flow.
 
-Fig. \ref{bp1-med} shows a typical sequence of shear-induced conformations.
-The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
-
-There is again an offset in the position of the network at the begin and the end of each cycle. 
-But neither the magnitude nor the direction of the offset seem to obey a clear rule.
-They are seemingly chaotic and different for every shear rate.
-
-\begin{figure*}[ht]
+\begin{figure*}[htpb]
 \includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
@@
 \caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page.
-The network is displaced during every cycle, but the magnitude and direction of the offset are seemingly chaotic.}
+The network is displaced during every cycle, but the magnitude and direction of the offset appear to be chaotic.}
 \label{bp1-med}
@@
 
-\begin{figure*}[ht]
+Fig.
+\ref{bp1-velo} depicts the secondary velocity components $v_y$ and $v_z$ in the same fashion as Fig.
+\ref{bp2-velo} for BPII.
+Again, the orientation of the components are as mirror images of each other if the helicity of the cholesteric changes from left- to right-handed.
+The emerging pattern is different from the one observed before in BPII, but agrees with
+the previous result in the sense that secondary velocities are largest in the regions
+where large conformational changes take place.
+
+\begin{figure*}[htpb]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a snapshot of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
 The colour code gives the magnitude and sign of the z-component.
@@
 
-Runs at smaller system sizes suggest that the individual conformation 
-can depend sensitively on the system size. Fig. \ref{bp1-low-fe}
+The complexity of flow-induced states and the seemingly chaotic but 
+periodic movement of the BPI network raises the question how robust 
+these results are with respect to the system size.
+Although the pitch length of 64 LU and the equivalent of 32 LU per
+unit cell provide enough resolution to track the motion of the 
+network in flow it is not clear whether 64 unit cells in total are enough to 
+mimick true bulk behaviour.
 
-\begin{figure}[ht]
+Unfortunately the computational costs are still prohibitive for a run 
+with 512 unit cells, an equivalent of 8 unit cells along each coordinate 
+direction. Therefore we opted for a comparison with a smaller run consiting
+of 8 unit cells, 2 along each dimension.
+Fig. \ref{bp1-2uc4uc} shows a comparison of the apparent viscosity $\eta_{app}$ 
+of runs with 8 and 64 unit cells at flow rates 
+$\gd=9.76\e{-6}$ and $1.95\e{-5}$, respectively.
+We do not visualise the conformations directly but rather use the fact that 
+they are coded in the shape of each individual curve. Obviously there 
+are stiking differences between 
+both system sizes, but the recurrence period is hardly affected. In the slower
+run the smaller system apparently follows the same branch for a while but then
+deviates from the larger run. 
+This result indicates that we cannot know with certainty if the particular flow patterns 
+we find at runs with 64 unit cells do not suffer from the imposed periodicity.
+Nevertheless, the chaotic behaviour and the recurrency seem to be a generic
+feature of BPI in this flow regime.
+
+\begin{figure}[htpb]
 \includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{}
-\label{bp0-2uc4uc-fe}
+\caption{Apparent viscosity $\eta_{app}$ for 8 (solid lines) and 64 unit cells (dashed lines)
+at $\gd=9.76\e{-6}$ and $1.95\e{-5}$. The shape of each curve differs largely, but the recurrence frequencies are hardly affected by the system size.}
+\label{bp1-2uc4uc}
 \end{figure}
 
-\begin{figure}[ht]
+\subsubsection{Regime BPI-3: high shear rates}
 
-\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
-\label{bp1-spec}
-\end{figure}
-
-The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
-Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
-Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
-
-\subsubsection{High shear rates}
-
-\begin{figure*}[ht]
-\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
-\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
-\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
+\begin{figure}[htpb]
+%\includegraphics[width=0.4\textwidth]{dir+z-300k_run916.png}
+\includegraphics[width=0.4\textwidth]{dir+z-301k_run916.png}
+\includegraphics[width=0.4\textwidth]{dir+z-302k_run916.png}
 \caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
@@
 \label{bp1-high}
-\end{figure*}
+\end{figure}
+
+A breakup of the network at large flow rates is also observed
+in BPI. Interestingly the window of critical shear rates 
+$3.125\e{-4}\le\gd_c\le 6.25\e{-4}$  is the same we find for BPII.  
+Contrary to BPII, which flows in form of a travelling helical wave
+(viz. Fig. \ref{bp2-high}) the preferred configuration is now
+a quasi two-dimensional state, shown in Fig. \ref{bp1-high} as 
+slices through the director field along the voritcity direction.      
+The sequence shows that BPI undergoes the affine transformation 
+in gradient-flow plane often seen in simple shear flow.
+There are distinctive regions where the predominant orientation 
+of the director is along the vorticity direction and others
+where it lies in flow-gradient plane. Occasionally while 
+flowing past each other this produces regions that 
+resemble local double twist.
 

### 2012-09-14 trunk/papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1783 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
-\label{tab1}}
+\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
+\label{tab1}
 \end{table*}
@@
 In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
-phases BPI and BPII. The thermodynamic states are characterised by an effective temperature 
-$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy terms.
-We opted for states sufficiently far away from the chiral-nematic-isotropic 
-where both quiescent blue phases are at equilibrium, namely
+phases BPI and BPII. 
+The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+is equal to about $10^8$~Pa. Further details about the conversion 
+from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.
+
+The thermodynamic states are characterised by an effective temperature 
+$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy.
+For the vast majority of simulations we opted for equilibrium states that are sufficiently 
+far away from the cholesteric-isotropic transition, namely
 $\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
-Simulations at metastable states at higher and lower temperatures and different chiralities showed 
-no differences in their generic flow behaviour.
-However, states closer to the phase boundary featuring higher free energy densities and 
-smaller average order parameters show a tendency to break up more easily under external driving forces.
+We performed also a number of simulations at metastable states at higher and lower temperatures 
+and different at different chiralities, but didn't find any significant differences in the 
+generic flow behaviour that is described below.
+However, for thermodynamic states that are closer to the phase boundary the disclination 
+network broke up into non-BP states at lower shear rates. This is not surprising as
+these states have on average higher free energy densities and smaller order parameters
+than those specified above and can therefore not withstand the same external forces.
 
-As is common in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our simulations with 
-analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit, 
-and equilibrated these (for 5000 LB timesteps) prior to starting the shear. 
-During the equilibration run the optimal redshift $r$ was recalculated at every timestep.
-After equilibration the redshift parameter was kept constant throughout the rest of the simulation.
-We chose a unit cell size of 16 lattice sites for BPII and 32 sites in case of BPI. 
-Runs with higher resolution confirmed that this was an appropriate minimal resolution to cover 
-the complex conformational changes and kinematic details of the blue phase networks.
+As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
+analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit 
+and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
+During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
+After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow.
+We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, which is the equivalent 
+of 4 unit cells along each coordinate direction and a total of 64 unit cells in our simulation box.
+Runs with higher resolution confirmed that the resolution was adequate to track  
+all kinematic details of the blue phase networks in shear flow.
 
-Simple shear flow is imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
-the top (bottom) half of the planes moving in positive (negative) x-direction.
-We applied a correction for spurious momentum currents that emerge due to interpolations accross 
+Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
+the top (bottom) half of the planes moving in positive (negative) x-direction and the 
+velocity gradient along the y-direction.
+We applied a correction for spurious momentum currents that emerge due to required interpolations across 
 the Lees-Edwards planes \cite{Henrich:2012a}.
 The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $6.3\times\e{-4}$ in LB units.
-The shear flow was imposed along the x-axis, whereas velocity gradient and vorticity are oriented 
-along the y- and z-axis.
+$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
+For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
+They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
+BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
 Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
 are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
-$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
-is equal to about $10^8$~Pa.
-More details about the conversion from LBU to SI units can be found in
-\cite{Henrich:2011a,Henrich:2010b}.
 
@@
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). 
+For clarity the individual curves have been shifted along the time and strain axis. 
+The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
+regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) 
+and break up of the network into a travelling helical wave (BPII-2, solid circle).}
 \label{bp2-rheo}
@@
 
-We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
-BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
-intersect and form a characteristic network of nodes.
+We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
+than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
+Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
+Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
+deviatoric stress normalised to the isotropic background viscosity: 
+$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
+A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
+any additional contribution from the liquid crystal.
+The top picutre shows the date over time, whereas the graph at the bottom gives 
+the same data versus accummulated strain.
+For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
+is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
 
-A quantitative analysis of the flow behaviour of BPII is shown in Fig. \ref{bp2-rheo}.
-The apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ over time normalised to
-the isotropic background viscosity $\eta_{app}$ is given in the left image, whereas
-the picture on the right shows the same data versus accummulated total strain.
-For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is due to the periodic break-up and reconnection of the network under shear flow.
+The inset shows a kind of flow curve, defined as time averages of 
+the deviatoric stress tensor $\bar{Pi}_{xy}$ as a function of shear rate $\gd$.
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
+For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
+$a=0.35, b=0.95$ describes the data to a very good approximation. 
+Hence, the degree of shear-thinning is remarkably small, but may depend
+on the exact direction in which the strain is applied, an aspect that we     
+leave for future work.
 
-The inset shows the time average of the deviatoric part of the stress tensor $\bar{Pi}_{xy}$ as a function of shear rate.
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
-The data for low and intermediate $\gd$ obeys a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.34(99), b=0.95(38)$ to
-a very good approximation, exibiting a remarkably small degree of shear-thinning.
-This probably reflects the very regular and rather undisturbed affine transformation of the disclination network.
+For shear rates $\gd\ge6.25\e{-4}$ the network of disclination lines
+breaks up and oscillations in the stress signal are absent.
+In this regime of shear rates, to which we refer as BPII-2, the 
+blue phase dissolves into a simple cholesteric liquid 
+crystal. The mode of flow is a travelling helical wave with the helical axis 
+oriented along the vorticity direction.
 
-\subsubsection{Low and intermediate shear rates}
+In the next paragraph we will investigate this in more detail by looking at the 
+kinetics of the disclination network.
 
-Fig \ref{bp2-med} shows the disclination network 
-undergoing an affine transformation under shear. The disclination lines break up and 
-reconnect further downstream forming a periodically recurring pattern. 
-The recurrence time in flow direction $\tau_F=p/(2\, v_{LE})= 1/\gd$.
+\subsubsection{Regime BPII-1: low and intermediate shear rates}
+
+Fig. \ref{bp2-med} shows the disclination network in shear flow as
+it undergoes an affine transformation. The disclination lines break up and 
+reconnect further downstream forming a periodically recurring pattern with 
+recurrence time $\tau_F=1/\gd$ in flow direction.
 
@@
 \includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
-of shear-induced transformations in the steady state at 
-$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
-is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. The boundary conditions impose flow in such a way that network moves 
-to the right in the upper half and to the left in the lower half of the picture.}
+\caption{Disclination network of BPII in shear flow: 
+The pictures show a typical sequence of snapshots in the steady state 
+at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
+gradient is oriented along the vertical direction, whereas the 
+horizontal direction is the flow direction. Lees-Edwards boundary 
+conditions have been imposed in such a way that network moves to the 
+right in the upper half and to the left in the lower half of the picture.}
 \label{bp2-med}
@@
 
-Apart from the affine transformation the general appearance is very close to that 
-of the quiescent equilibrium phase.
-Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
-the gradient direction in the direction of vorticity.
-Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
-geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
-positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
-making the network travel along the z-direction.
+The general appearance of the flowing network is apart from the affine transformation 
+very close to that of the quiescent blue phase at equilibrium. It is the same for 
+all shear rates in regime BPII-1 and a first explanation why shear-thinning is so weak.
 
+Interestingly, while being displaced with the flow the entire network moves perpendicular 
+to the flow and velocity gradient in the vorticity direction.
+A similar behaviour has been recently observed for blue phases in shear flow and 
+confined geometries ~\cite{Henrich:2012b}.
+But contrary to the stick-slip motion that has been reported there the movement is 
+steady. It occurs in such a manner that the positions of breakup and reconnection 
+of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
+travel along the z-direction.
 The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
-i.e. it is takes exactly six unit cell displacements in flow directions 
+i.e. it takes a displacement of six unit cells in flow directions 
 until the network has moved one unit cell in voriticity direction.
 
-Fig. \ref{bp2-velo} shows the disclination network at a characteristic time step between reconnecting and another breakup.
-Superimposed are vectors which represent the velocity projected onto a plain perpendicular to the flow direction.
-This projection makes it possible to get rid of the dominating velocity component $v_x$ and permits visualisation the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the minor components is typically in the range of a few percent of the primary flow component.
-Characteristic bands oriented along the vorticity direction form, which are normally hidden behind the primary flow in x-direction.
-When the helicity of the underlying cholesteric phase is changed from a left- to a righthanded helix the sign of $v_z$ 
-and the sense of motion also changes.
-Further quantitative evidence for a direct link between flow direction and helicity can be found by time-averaging over
-individual cycles.
-Tab. \ref{tab1} gives minima, maxima, averages and standard deviations of the velocity components.
-All values for the two runs with inverted helicity are identical apart from a change of sign in the z-components.
-There is only a slight imbalance between the maximum and minimum velocity in z-direction.
-Hence the movement of the network in vorticity direction is predominantly permeative.
-Note that the imbalance is exaclty retained if the helicity is inverted.
+Fig. \ref{bp2-velo} shows the disclination network sometime between reconnecting 
+and the next breakup with superimposed velocity vectors.
+Shown are what we may call secondary velocity components, obtained by projecting the
+velocity onto a plain perpendicular to the flow direction.
+This allows to get rid of the dominating velocity component $v_x$ and to
+visualise the two much smaller components $v_y$ and $v_z$.
+The magnitude of the secondary components is typically in the range of a 
+few percent of the primary flow component.
 
-\subsubsection{High shear rates}
+Characteristic bands are visible, which are oriented along the vorticity direction and
+normally hidden behind the primary flow component.
+When the helicity of the underlying cholesteric phase is changed from left- to a 
+righthanded the pattern becomes a mirror image of the former one. The sign of 
+$v_z$ change and also the sense of motion is inverted.
+Further quantitative evidence for a direct link between the sense of motion 
+and the helicity can be gained by time-averaging over individual cycles.
+Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
+of the velocity components.
+All values for the two runs with inverted helicity are identical apart 
+from a change of sign in the z-components.
+There is only a slight imbalance between the maximum and minimum velocities 
+in z-direction, suggesting that the movement of the network in vorticity 
+direction is permeative. Note that this imbalance is retained under inversion of
+helicity.
 
-If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation in shear flow that we
-descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
-The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
-a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
-invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
-before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
-shear rates \cite{Rey:1996a, Rey:1996b}.
+\subsubsection{Regime BPII-2: high shear rates}
 
-In our scenario the break-up occurs for shear rates between $3.12\e{-4}\le\gd_c\le 6.25\e{-4}$.
-However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
-such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
-Due to a lack of computational resources we have not investigated these aspects any further.
+If the flow rate exceeds a certain critical value BPII does not undergo the 
+affine transformation descibed above and adopts another mode of flow. 
+Fig. \ref{bp2-high} shows a sequence of cuts through the director 
+field at high shear rate. The entire blue phase network has broken up 
+into a simple cholesteric helix and flows as a travelling 
+helical wave with the helical axis oriented along the vorticity direction.
+The state is translationally invariant in flow and gradient direction.
+and is the preferred mode of flow of a cholesteric 
+liquid crystal at low Ericksen numbers below the uncoiling transition 
+to a nematic, flow-aligned state at even higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
+
+In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$.
+At the moment we do not have a hard criterium for the critical
+shear rate at hand. However, it seems reasonable that the critical shear rate 
+depends on thermodynamic and kinetic parameters such as the effective 
+temperature $\tau$, the chirality $\kappa$ and the rotational diffusion 
+constant $\Gamma$.
 
@@
 \includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
+$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. 
 The configuration is a helical wave that travels along the direction of vorticity.
@@
 
-In BPI the disclination lines are arranged in a body-centered-cubic array.
-Contrary to BPII, they are well separated and do not intersect. 
-There is strong evidence that these topological differences result in 
-a very different rheological response of BPI. In fact BPI exhibits some 
-peculiarities that are absent in BPII.
+BPI has body-centered symmetry and contrary to BPII the disclination lines
+ are well separated and do not intersect.
+We believe that this topological difference is responsible for most of
+the differences between BPI and BPII regarding their rheological response.
 
-Just as before in a kind of overview we present rheological key quantities first and 
-will address the detail aspects further below. 
-Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
-strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
-are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
-relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
-the apparent viscosity varies by almost a factor 8 during one cycle.
-The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
-follow a simple power law.
+We present again key aspects in a kind of overview before we 
+address some features in more detail. Fig. \ref{bp1-rheo} shows the 
+apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
+as in Fig. \ref{bp2-rheo}.
+For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
+the disclination network occurs similar to and at the same critical shear rate as 
+the one observed in BPII-2 but with an order structure that is a quasi 
+two-dimensional.
 
-The inset of shows the flow curve, i.e. shear stress over shear rate.
-The data points are averages over one cycle, whereas the error bars indicate the maximum and
-minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
-At high shear rates the network breaks up similar to BPII. The director field in this state
-is effectively two-dimensional and translationally invariant along the neutral direction.
-At intermediate shear rates the BPI network flows with periodically recurring conformations.
-These complex flow-induced conformations are accompanied by large fluctuations that increase 
-in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
-of the liquid crystal during one cycle is actually negative. 
-Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
-Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
-end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.
+In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
+deviatoric stress that are not sinusoidal, but show an increasingly complex
+pattern up to the breakup point.
+For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
+they become  unregular and finally disappear after a few
+cycles. Shear-thinning is slightly stronger than in BPII. 
+The average shear stress during one cycle can be fitted reasonably 
+well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
 
@@
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right).
-The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.
-Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle).
-The error bars give the minimum and maximum shear stress during one cycle.}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (top) 
+and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
+Depending on the configuration in steady shear flow three different regimes 
+can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
+steady flow with recurring patterns (BPI-2, open circles) and 
+breakup of the disclination network (BPI-3, solid circle). 
+The error bars represent the minimum and maximum shear stress 
+during one cycle. For clarity the individual curves have been shifted along
+the time and strain axis.}
 \label{bp1-rheo}
@@
 
-The Fourier analysis of the time-dependent deviatoric stress $\Pi_{xy}$ 
-offers another perspective on this intricate flow behaviour at high and intermediate shear rates. 
-The spectrum $X_\omega$ according to Eq. \ref{spectrum} is shown in Fig. \ref{bp1-spectrum}.
-We observe a strong contribution of the first harmonic, which is directly related to the 
-recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ 
-via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
-As the shear rates we applied always differ by a factor 2 the Fourier coefficients of the 
-individual runs with different shear rate coincide in such a way that the
-the fundamental mode is the same frequency as the first overtone of the next lower shear rate
-or the third overtone of the next-to-next lower shear rate 
-(viz. peak positions at $\gd=7.81, 3.91$ and $1.95\e{-5}$).
-Moreover, the magnitude of higher Fourier modes decreases monotonously.   
-However, the inset shows that this characteristic frequency relation is violated 
-for shear rates above $\gd=7.81\e{-5}$.
-Although the velocity of the primary flow component in x-direcition, listed in Tab. \ref{tab1},
- is very well in accordance with the imposed shear rate, the fundamental mode appears 
-now at the same frequency.
-The contribution of higher Fourier modes increases and even exceeds that of
-the fundamental mode, which gives the shear stress generally a more jittering appearance.
-At the highest imposed shear rate where the breakup occurs the Fourier spectrum 
-becomes simple again and features a dominating fundamental mode at the 
-expected frequency just as described above for $\gd\le7.81\e{-5}$.
-The breakup occurs at exactly the same flow rate as those of BPII, 
-suggesting that this can be traced back to the rotational dynamics 
-of the order parameter.
+In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
+we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
+The spectrum is shown in Fig. \ref{bp1-spectrum}.
+
+There are two aspects which are worth mentioning. First of all, we observe a strong contribution of 
+the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
+However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
+significantly larger. 
+More surprisingly the frequency of the fundamental mode is the same 
+for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is completely 
+contrasted by what is observed at lower flow rates. By doubling the flow rate from 
+$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the position of the peaks doubles in the spectrum 
+moves to the double frequency. This means that the fundamental mode at flowrate $\gd=7.81\e{-5}$ 
+coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on,
+which is the expected behaviour. At higher shear rates in regime BPI-3 this lock-in of 
+flow modes disappears again.
+
+Interestingly the lock-in has only a minor effect on the flow velocities themselves.
+The average velocities in Tab. \ref{tab1} reveal that only the magnitude of some of the secondary 
+flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
+The latter is directly linked to the imposed shear rate.
 
@@
 \includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate.
-The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed.
-Instead the spectral contribution of higher modes grows until finally the breakup takes place.
-The frequency is given in units of $10^2\cdot 2\pi/T$ with $T$ as period of one cycle in LB time steps.}
+\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture and inset show spectral
+components and the lock-in of flow modes, visible in the fixed peak position for certain flow rates. 
+The inset represents a blow-up of the same data. The frequency is given in radians.}
 \label{bp1-spectrum}
@@
 
-\subsubsection{Low shear rates}
+\subsubsection{Regime BPI-1: low shear rates}
 
@@
 
-\subsubsection{Intermediate shear rates}
+\subsubsection{Regime BPI-2: intermediate shear rates}
 
@@
 route to get back on the positive branch.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
+\caption{}
+\label{bp1-1_bp1-2}
+\end{figure}
 
@@
 
-\subsubsection{High shear rates}
+\subsubsection{Regime BPI-3: high shear rates}
 

### 2012-09-28 papers/bprheo/bprheo.tex
Updated bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1786 40876441-3496-0410-a613-a2fca35b2ae2

@@
 They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
-BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
+BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
 Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
@@
 For clarity the individual curves have been shifted along the time and strain axis. 
-The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) and break up of the network into a travelling helical wave (BPII-2, solid circle).}
+The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
+regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) 
+and break up of the network into a travelling helical wave (BPII-2, solid circle).}
 \label{bp2-rheo}
@@
 than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
-
-As it might be easier to enter a discussion with the full picture in mind, we provide first a 
-quantitative analysis of the flow behaviour at all applied shear rates.
+Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
 Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
@@
 
-Again, we present key aspects in a kind of overview before we 
+We present again key aspects in a kind of overview before we 
 address some features in more detail. Fig. \ref{bp1-rheo} shows the 
@@
 
-The rheological response of BPI at low shear rate which exhibits a
-dynamic yield stress is strikingly different from that of BPII.
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density.
+This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
+\label{bp1-fe-yield}
+\end{figure}
+
+The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
+is strikingly different from that of BPII and exhibits a
+dynamic yield stress.
 An explanation for this strange behaviour can be found by 
@@
 disappears.
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$): 
+Depicted is the transition from flow-induced, intertwined helices that 
+undergo a recurring affine transformation (top left and centre at 
+time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate 
+conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) 
+to a distorted state (bottom centre at time step $t=1.7\e{6}$) and 
+finally to an amorphous network (bottom right at $t=2.2\e{6}$). 
+All pictures have the vie along the vorticity direction apart from the 
+one at the top centre which shows the network in gradient direction.}
+\label{bp1-low}
+\end{figure*}
+
 The negative branch in the deviatoric stress is connected
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density.
-This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
-\label{bp1-fe-yield}
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
+\caption{Deviatoric stress $\Pi_{xy}$ over strain at the transition between regime
+BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
+and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
+where the stress goes negative (viz. Fig. \ref{bp1-fe-yield}) the stress is 
+throughout positive, but exhibits still very large fluctuations during one cycle, 
+which are absent in regime BPI-2 at higher shear rates.}
+\label{bp1-1_bp1-2}
 \end{figure}
@@
 
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
-\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).
-All pictures have the vie along the vorticity direction apart from the one at the top centre which shows the network in gradient direction.}
-\label{bp1-low}
-\end{figure*}
-
 However, this mode of flow proves unstable as after a few cycles 
@@
 transforms into an amorphous state with virtually constant
-yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LU.
+yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
 
@@
 
-Intermediate shear rates in the sense of this study of BPI are characterised 
-by the missing negative branch in the deviatoric stress. It is unclear 
-why this feature is missing at higher flow rates. A possible explanation is that 
-above a certain flow rate the network simply skips the comparably small 
-window of negative stress in Fig. \ref{bp1-fe-yield} and finds a quicker 
-route to get back on the positive branch.
+Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) which features yield stress 
+behaviour at very low flow rates and the onset of power-law behaviour 
+at slightly larger shear rates (viz. Fig. \ref{bp1-rheo}) lies another 
+region ($3.91\e{-5}\le\gd< 6.25\e{-4}$) where the network flows with periodically 
+recurring patterns. The transition between these two regimes is shown
+in Fig. \ref{bp1-1_bp1-2} in terms of deviatoric stress over total strain.
 
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{}
-\label{bp1-1_bp1-2}
-\end{figure}
+The reason for this qualitative difference is rather obscur. A possible 
+explanation is that due to the larger flow rate the network kinetically 
+bypasses the relatively large stress fluctuations during one cycle which
+are typical in regime BPI-1. For even lower $\gd$ the deviatoric stress
+goes actually negative (viz. Fig. \ref{bp1-fe-yield}). This destabilises
+the periodic network very quickly and leads to an amorphous configuration
+in the steady state.
 
@@
 these results are with respect to the system size.
-Although the pitch length of 64 LU and the equivalent of 32 LU per
+Although the pitch length of 64 LBU and the equivalent of 32 LBU per
 unit cell provide enough resolution to track the motion of the 
@@
 \begin{figure}[htpb]
+\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
 \includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Apparent viscosity $\eta_{app}$ for 8 (solid lines) and 64 unit cells (dashed lines)
-at $\gd=9.76\e{-6}$ and $1.95\e{-5}$. The shape of each curve differs largely, but the recurrence frequencies are hardly affected by the system size.}
+\caption{}
 \label{bp1-2uc4uc}

### 2012-10-02 papers/2012-sandwich/CholColloids.3.tex
added tex file for sandwich paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1788 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[12pt]{article}

\usepackage{graphicx,color}

\begin{document}

\title{Self-assembled soft solids in colloid-cholesteric mixtures}
\author{Ourselves \\ {\small Here and there}}
\date{}

\maketitle

{\bf Dispersion of particle inside liquid crystals have been shown to form new materials with tunable elastic and electrooptic properties~\cite{stark}. When the host is a blue phase~\cite{mermin}, a chiral liquid crystal with full three-dimensional periodicity, it has been proposed that the particles should self assemble into colloidal crystals which can selectively absorb light~\cite{miha,lavrentovich}. 
Here we present computer simulations which show that such colloid-cholesteric mixtures give actually rise to a much wider range of new soft solids than previously anticipated, encompassing regular crystals, glasses, cellular solids, isolated clusters, twisted rings and undulating colloidal ropes. The final self-assembled structure can be controlled by tuning the colloid concentration and the anchoring strength, that is the extent to which particles perturb the liquid crystalline alignment nearby. An important role is also played by the dynamical schedule with which colloids are introduced in the mixture, and by the ordering imparted through confining walls. The soft solids we find are metastable, and should be relevant to the design of new generation switchable multistable devices, which may be used to create, for instance, novel display and electronic paper technologies~\cite{epaper}.}

%Coles abstract/first paragraph:
%A promising approach to the fabrication of materials with
%nanoscale features is the transfer of liquid-crystalline structure
%to polymers1–11. However, this has not been achieved in systems
%with full three-dimensional periodicity. Here we demonstrate
%the fabrication of self-assembled three-dimensional nanostructures
%by polymer templating blue phase I, a chiral liquid
%crystal with cubic symmetry. Blue phase I was photopolymerized
%and the remaining liquid crystal removed to create
%a porous free-standing cast, which retains the chiral threedimensional
%structure of the blue phase, yet contains no chiral
%additive molecules. The cast may in turn be used as a hard template
%for the fabrication of new materials. By refilling the cast
%with an achiral nematic liquid crystal, we created templated
%blue phases that have unprecedented thermal stability in the
%range 􀀀125 to 125 C, and that act as both mirrorless lasers
%and switchable electro-optic devices. Blue-phase templated
%materials will facilitate advances in device architectures for
%photonics applications in particular.

%introduction on colloids in liquid crystals
%why it is interesting to consider a cholesteric liquid crystal or blue phase
%new materials: gels, photonic crystals and glasses

When spherical particles are mixed into a nematic liquid crystal, 
they disrupt the orientational order in the fluid and create defects,
or disclination lines, close to their surface. As these local drops in
the order increase the overall free energy of the system, it is advantageous
to gather many such deformations in one place, rather than distribute them 
uniformly in 3D throughout the fluid. Therefore these particles 
have a generic tendency to aggregate. Such aggregates may further 
self-assemble into lines~\cite{wiresmiha}, 2D crystals~\cite{zumer}, 
planar structures~\cite{tanaka}, or 3D amorphous glasses~\cite{tiffany}.
Besides being of fundamental interest to materials science, these
structures have tunable elastic and optical properties, hence 
they offer exciting prospects for applications as biosensors~\cite{abbott}, or
as new-generation display devices~\cite{colloiddevice}.

But one may also choose to disperse colloidal particles in a cholesteric, 
rather than nematic, liquid crystal. The molecules making up this material are
chiral, and this causes their average orientation to rotate in space in
a helical fashion, rather than stay constant as in nematics. 
This typically creates a 1-dimensional periodic structure whose wavelength 
is the pitch, $p$. The periodicity can however become 3D in the so-called 
blue phases (BPs). BPs arise because the molecule orientation -- the so-called 
director field -- can twist along more than one direction (see 
the cartoon in Fig. 1). Such double twist 
regions are energetically favoured due to the molecular chirality of the fluid,
however they lead to frustration because it is impossible to tile the whole 3D
space by putting together double twist regions without introducting
topological defects -- singularities where the director field is undefined.
These defects line up to form disclinations, which in blue phases may
either form 3D periodic regular crystals (the case of BPI and BPII) or be 
disordered, as in BPIII, aka the blue fog. BPs therefore are stabilised by
the tendency of chiral molecules to twist, and destabilised by the
elastic cost of forming topological defects and distorting the orientation.
This competition limited the temperature range in which they could be
observed in the past to a tiny neighbourhood or the order-disorder
transition. Current advances have increased the stability range by orders
of magnitude, up to an amazing 250 K, which has paved the road to
applications of BPs in display devices~\cite{kikuchi,coleswidetrange,bpdevice}.

%what is the use?? biosensors and photonic crystals then?? bistable
%devices?

%why use cholesteric/blue phases? because they provide defect
%network even in the absence of particles!

Besides being remarkable materials {\it per se}, BPs make potentially
interesting hosts for dispersing colloidal particles as well. One reason
is that BPs form a disclination network even in the absence of
particles, so that they can potentially template their self-assembly. 
Indeed, as proved recently by simulations~\cite{miha}, one can imagine that 
when dispersing nanoparticles (which interact only weakly with the molecules
of a BP), such particles will be attracted to the defects in the BP
network. That is because, by covering up such disclinations, 
the system does not have to pay anymore the elastic cost associated with
defect formation, hence the free energy of the mixture decreases.
As the defect network is highly ordered in BPs, the particles will tend
to form a regular colloidal crystal, with the periodicity of BPs. 
Because, in turn, its associated wavelength lies in the visible range, 
the resulting material should have a photonic bandgap (albeit probably
an incomplete one, as the BPs themselves do~\cite{bplasers}), and can thus be used to 
control light propagation.
Another reason to study colloid-BP mixtures is that even in the absence
of particles there are several coexisting metastable minima in the system,
corresponding to different topologies of the defect network~\cite{adriano}. 
Adding particles
is likely to add more features in the free energy profiles, and if this
can be done in a controlled way the resulting system will be uniquely
placed to provide a bistable or multistable device, provided a strategy
can be found to kinetically access this or that minimum. Multistability
is highly sought after in new generation display devices, because it
endows samples with memory. For instance, a metastable state will not
change much even after the field which was needed to reach it is removed,
and this leads to huge saving in energy consumption.

%what new here? phase transitions between different soft solids! tunability
%in the bulk: transition between photonic crystal, isolated clusters
%and self-quenched (blue phase) glass
%with confinement: transition between arrested loops, sheets 
%and cellular solids

We show in Fig. 2 results from large scale simulations which suggest the
existence of multiple phases for mixtures of colloidal particles 
(size $2R=100$ nm) of finite volume fraction in a BPI disclination network
(elastic constant $K= XXX$ pN, unit cell size XXX nm). 
We show in Fig. 1 the final states of four simulations (see Supporting 
Information for 
details) in which we first equilibrated 
a BPI network of disclination, for a choice of parameters (see Supporting
Information) 
for which such a state is the thermodynamically stable one, and then 
dispersed colloidal
particles, making up either 1\% (Fig. 2A and 2B) or 5\% (Fig. 2C and 2D)
of the total volume of the mixture. Briefly, our algorithm couples molecular
dynamics for the colloidal particles to continuum hydrodynamic theory
for the evolution of the tensorial order parameter of 
the blue phases (see Supporting Information). 
The liquid crystal director field tends
to align perpendicular to the surface of the colloidal particle, and
the strength of the anchoring is $W$.

The value of the dimensionless ratio between the anchoring energy, $WR^2$, 
and the stored elastic energy due to the
nearby distortion in the ordering, $KR$, is well known to 
play an important role in the physics of a single colloidal particle
in a nematic fluid~\cite{stark}, where it controls the transition
between different topologies in the disclination accompanying the particle.
We find that this ratio, $WR/K$, is also a key contributor, together
with the colloidal volume fraction, to determine which soft solid assembles
starting from the random dispersion of particles in an ordered BP.

%results from computer simulations in bulk cholesterics
%small anchoring particles go to defect lines
%large anchoring particles form either clusters or percolated networks 
%Fig. 1 on blue phases

When $WR/K$ is small (0.23 in Fig. 1A and 1C), the particles are
attracted to the disclination network, and cover up the defect lines
orderly to form a regular structure, similar to the photonic crystal
predicted in~\cite{miha}. Interestingly, the weak but finite anchoring
appears to lead to an elastic force between particles on a disclination lines
which sets a typical distance between them. In this
small $WR/K$ regime, it is the preexisting BP network which templates
the formation of a colloidal structure. It is likely that polymer-stabilised 
blue phases~\cite{kikuchi} 
work by a similar principle, based on a weak interaction between 
the monomers in the chain and the surrounding liquid crystal orientation.

For $WR/K>1$ (see Fig. 2B and 2D), this picture changes dramatically.
In this case the colloidal particles disrupt significantly the
orientational order in the fluid, and this leads to the 
creation of new disclination close to their surface, similar 
to the Saturn rings which appear in nematics~\cite{stark}. The sharp drop in
liquid crystalline order within the BP leads now to a 
strong interparticle attraction, which restructures the disclination
network completely, such that the final result is quite distinct from
the original BPI topology. Most notably, the aggregates favour
the formation of defect junctions, where four disclinations meet
(similar to the structure of BPII rather than BPI). The 
particle clusters are disjoint for low volume fraction, but 
they interact elastically via the connecting disclinations.
As the volume fraction becomes larger, the aggregates join up
to form a percolating colloidal cluster which the BP disclinations
follow. All the structures in Fig. 1 should be soft solids, in
the sense that their zero-frequency elastic modulus, $G'$ should
be non-zero if probed via oscillatory rheology (unlike pure BPs where
$G'=0$ at zero frequency as the system can flow with a finite
viscosity via permeation~\cite{permeation1,permeation2}). We expect that the
structures in Fig. 1A-C should be weak gels, as here the 
defect network percolates but the particles do not, similarly
to what found in the cholesteric-nanoparticle gel stabilised
in~\cite{lubensky} (with $G'<1$ Pa). The ``cellular solid'' 
structure in Fig. 1D should be much more resistent mechanically, 
due to the formation of a thick percolating particle network. The 
structure is now akin to that of the self-quenched glass found in 
colloid-nematic composites~\cite{tiffany}, but for much higher volume 
fraction (over 20\% as opposed to below 5\% here).

One may wonder how easy it is in practice to grow very regular blue
phase structures in the absence of colloidal nanoparticles. 
Before the technological advances of~\cite{kikuchi,coleswidetrange}
this was challenging and required fine tuning of temperature
and chirality: old experimental phase diagrams 
show that in a wider parameter range BPIII, which is amorphous, forms
in place of the ordered cubic BPI and BPII~\cite{mermin} (see 
typical experimental phase diagram in Fig. 1). 
Recent simulations also suggest that
domain growth in BPs favours disordered structures, and that the
nucleation of an ordered disclination crystal is probably hierarchical
and slow~\cite{domaingrowth,bp3}. 
What happens then when we disperse then colloids into a glassy
defect network, such as that of BPIII? Fig. 3 addresses this question,
for the same parameters used in Fig. 2. For small anchoring ($WR/K<1$),
the particles once again dress the disclination lines, but this time they
give rise to a glassy structure rather than to a crystal, as such
is the nature of the underlying network which acts as a template.
Increasing the volume fraction again leads to a more even coverage with
a rather well defined particle spacing (compare Fig. 3A and 3B). 
For large anchoring ($WR/K>1$), on the other hand, the pattern is very
much alike that seen in Fig. 1: now it is the particles which dictate
the self-assembly, and we observe isolated clusters at low volume 
fraction, and cellular solids for higher colloid concentration
(see Fig. 3C and 3D). As expected, the free energy of the soft solids
obtained by dispersing nanoparticles inside these disordered networks
is considerably higher than the one found with the stable regular BPI
structure (see Supporting Information, Fig. S1). 
Nevertheless, all these structures
are metastable, as the elastic forces keeping them together dwarf thermal
energy from Brownian motion (see the scales of energies and energy
differences in Fig. S1).

%sandwiches: effect of anchoring on the structure of the soft solid
%soft solid: G'\ne 0

The existence of several locally stable competing solids with 
very different symmetry and, arguably, physical properties, 
is of interest to technology, as it makes these materials ideal
candidate to build multistable devices, which can retain memory of
more than one metastable state in the absence of an applied field.
Multistability is indeed a must-have feature for new generation switchable 
electrooptic devices for use in energy-saving electronics-free
appliances, smart glass, and e-paper to cite just a few examples.
However, state-of-the-art liquid crystal display devices are
rarely so large that the bulk description in Fig. 2 and 3 is relevant to
them. Typically, the fluid is strongly confined in these devices, and the
boundary effects cannot be disregarded. Therefore in what follows we describe
how colloidal suspensions self assemble in thin sandwiches, where
the confining walls are treated so as to favour planar or 
perpendicular (homeotropic) anchoring of the liquid crystal.

Fig. 4 shows the case of normal anchoring at the wall, again for
small (Fig. 4A and 4C) and large (Fig. 4B and 4D) values of
$WR/K$. Initially the liquid crystal is quenched from the isotropic
to the ordered phase as done for Fig. 3, for similar parameter values 
as previously used. Therefore, a metastable glassy disclination network
arises once again (see Supplementary Material): the conflict between the
normal anchoring at the wall and the tendency of the liquid crystal to
twist in the bulk further hinders the formation of a regular structure.
The structures formed by dispersing nanoparticles within this network
are somewhat similar to the ones already observed in the bulk: at low
values of $WR/K$ the particles simply hop on the network (Fig. 4A and 4C), 
whereas for large anchoring at the colloid surface we observe isolated
clusters (Fig. 4B) and percolated cellular solids (Fig. 4D), for
``small'' (1\%) and ``large'' (5\%) particle volume fraction respectively.
There are however important differences, due to the key role played
by the boundaries. 
First, the normal anchoring at the wall recruits particles there,
so that the colloids aggregate either on the planes or towards the middle of
the sample (see the side view of the snapshots in Fig. 4). Second,
the clusters formed for strong anchoring (see Fig. 4D in particular) 
are planar, similarly to those observed in
colloid-cholesteric mixtures for large particle
size~\cite{niek}.

The case of planar wall anchoring is treated in Fig. 5 and leads to
striking differences, and yet more self-assembled structures. 
Without colloids, the planar anchoring now leads to
the formation of a regular cholesteric helix,
with axis perpendicular to the boundary walls. 
This twisted texture forms by annealing away an initial loop defect network,
which shrink and disappear to leave a defect-free sample. However,
dispersing particles with small $WR/K$ arrests the loop shrinking process,
and particles stabilise twisted figure-of-eight rings for dilute
suspensions, or plectonemic structures for more concentrated ones (see Fig. 5A
and 5C respectively). 
For larger $WR/K$, structures form thick fibers, with diameter about
5-fold larger than that of single colloids (see Fig. 5B and 5D). 
These colloidal ropes are undulating due to the chiral structure of the 
underlying fluid, and develop branching points upon increasing the particle
density, to eventually converge again into cellular solids once the suspension
becomes concentrated enough. The wall-induced ordering now is incompatible
with the homeotropic anchoring at the colloid surface, and particle
are this time repelled from the wall (see the side views of the
configurations in Fig. 5).

%Conclusions

The surprising wealth of structures obtained in Figs. 2-5 is the complex 
consequence of a simple competition between two self-assembly principles. 
First, blue phases and cholesteric defect networks provide a three-dimensional 
template of disclinations to which colloidal particles are attracted, in order 
to lower the free energy of the mixture. Second, colloidal intrusions in a 
liquid crystal lower the orientational order locally, hence are subjected to a 
generic tendency to aggregate into clusters to minimise such elastic
distortions. Which of these principles dominates depends on the value of a key 
dimensionless parameter in the system, $WR/K$, which measures the strength of 
the anchoring of the liquid crystal director field at the colloidal surface. 
If this ratio is low, the templating principle dominates: according to
the wall coating and the initial disclination network chosen (BPI or BPIII) 
we get colloidal crystals, glasses or twisted rings. If $WR/K$ is high,
flocculation wins over templating, and we end up with isolated aggregates,
percolating cellular solids, or helical colloidal ropes according to
boundary properties and the volume fraction of the dispersion. All these
soft solids will have different, and tunable, elastic and optical properties,
hence are highly promising for the design of future multistable liquid
crystal devices, which are at the forefront of current technological 
developments in this field.

\begin{thebibliography}{99}
\bibitem{stark} H. Stark, {\it Phys. Rep.} {\bf 351}, 387-474 (2001).
\bibitem{mermin} Wright D.~C., Mermin, N.~D. Crystalline liquids: The blue phases. {\it Rev. Mod. Phys.} {\bf 61}, 385–432 (1989).
\bibitem{miha} M. Ravnik, G.~P. Alexander, J.~M. Yeomans,  
S. Zumer, {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188-5192 (2011).
\bibitem{lavrentovich} O.~D. Lavrentovich, {\it Proc. Natl. Acad. Sci.
USA} {\bf 108}, 5143-5144 (2011).
\bibitem{epaper} J. Heikenfeld, P. Drzaic,  J.~S. Yeo, T. Koch.
A critical review of the present and future prospects for electronic paper.
{\it J. Soc. Inf. Display} {\bf 19}, 129-156 (2011).
\bibitem{zumer} I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik,
S. Zumer, {\it Science} {\bf 313}, 954-958 (2006).
\bibitem{tiffany} T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo,
W.~C.~K. Poon, {\it Science} {\bf 333}, 79-83 (2011).
\bibitem{abbott} I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo, N.~L. Abbott, {\it Science} {\bf 332}, 1297-1300 (2011).
\bibitem{colloiddevice} K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton,
R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}
{\bf 35}, 4419 (1987).
\bibitem{kikuchi} H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama.
Polymer-stabilized liquid crystal blue phases.
{\it Nat. Mat.} {\bf 1}, 64-68 (2002).
\bibitem{coleswidetrange} H.~J. Coles, M.~N. Pivnenko.
Liquid crystal 'blue phases' with a wide temperature range.
{\it Nature} {\bf 436}, 997-1000 (2005).
\bibitem{bpdevice} Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
{\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
\bibitem{bplasers} W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri.
Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
phase II. {\it Nat. Mat.} {\bf 1}, 111-113 (2002).
\bibitem{permeation1} 	W. Helfrich. Capillary flow of cholesteric
and smectic liquid crystals. {\it Phys. Rev. Lett.} {\bf 23}, 372-374 (1969).
\bibitem{permeation2} A. Dupuis, D. Marenduzz, E. Orlandini, J.~M. Yeomans.
Rheology of cholesteric blue phases. {\it Phys. Rev. Lett.} {\bf 95},
097801 (2005).
\bibitem{lubensky} M. Zapotocky, L. Ramos, P. Poulin, T.~C. Lubensky, 
D.~A. Weitz, {\it Science} {\bf 283}, 209-212 (1999).
\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles, {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates,
{\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic, {\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
\bibitem{tanaka} T. Araki, H. Tanaka, {\it Phys. Rev. Lett.} {\bf 97}, 
127801 (2006).
\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka,
{\it Nat. Mat.} {\bf 10}, 303-309 (2011).
\bibitem{adriano} A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini,
F. Salvadore, {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
\bibitem{fukuda} J.-I. Fukuda, S. Zumer, {\it Nat. Comm.} {\bf 2}, 246 (2011).
\bibitem{niek} Hijnen N., Wood T.~A., Wilson D., Clegg P.~S. 
Self-Organization of Particles with Planar Surface Anchoring in a Cholesteric
 Liquid Crystal. {\it Langmuir} {\bf 26}, 13502-13510 (2010).
\end{thebibliography}

\newpage

\begin{figure}
\includegraphics[width=\textwidth]{s0.png}
\caption{Schematics of the molecular arrangement in a 
cholesteric helix (A) and within a double twist cylinder (B, the
axis of the cylinder is perpendicular to the plane of the page).
The disclinations line arrangement in the unit cell of BPI is shown in
(C), while a reference phase diagram (from Ref.~\cite{crooker}) is
shown in (D).}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s1.jpg}
\caption{Snapshots of the steady states obtained after dispersing
a suspension of colloidal nanoparticles inside blue phase I, initially
in its thermodynamic equilibrium structure. Structures correspond
to (A) $WR/K=0.23$ and $\phi=1\%$, (B) $WR/K=2.3$ and $\phi=1\%$,
(C) $WR/K=0.23$ and $\phi=5\%$, (D) $WR/K=2.3$ and $\phi=5\%$.
The anchoring of the director field to the colloidal surface is normal.
[For the full parameter list used to generate Figs. 2-5, 
in simulation units, see Supporting Information.] }
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s2.jpg}
\caption{Same as Fig. 2, but now the disclination network
inside which the nanoparticles are dispersed is 
amorphous, and we prepared it through
a quench from the isotropic into the blue phase  I region.}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s3.jpg}
\caption{Snapshots of the steady states obtained after dispersing
a suspension of colloidal nanoparticles inside a sandwich filled with
cholesteric liquid crystal, with normal anchoring of the director field
at the boundaries. Both top and side views are provided. 
Structures correspond
to (A) $WR/K=XXX$ and $\phi=1\%$, (B) $WR/K=XXX$ and $\phi=1\%$,
(C) $WR/K=XXX$ and $\phi=2\%$, (D) $WR/K=XXX$ and $\phi=2\%$.
}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s4.jpg}
\caption{Same as Fig. 4, but with planar anchoring of the director 
field to the boundary planes.}
\end{figure}

\end{document}

### 2012-10-11 papers/bprheo/bprheo.tex
New draft BP rheology paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1790 40876441-3496-0410-a613-a2fca35b2ae2

@@
 the helical axis, small pressure differences drive flow mainly through
-``permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
+''permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
 In the permeation mode the liquid crystal flows while leaving the director
@@
 In the crossover region they predicted a novel oscillatory regime where the network 
-continuously breaks and reforms as portions of the disclinations in the center of the channel 
-move to neighboring cells and relink with the parts of the network left behind by the flow. 
+continuously breaks and reforms as portions of the disclinations in the centre of the channel 
+move to neighbouring cells and relink with the parts of the network left behind by the flow. 
 Compared with the cholesteric case, the viscosity still decreases with forcing 
@@
 the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
-characterizes the local degree of orientational order.
+characterises the local degree of orientational order.
 The thermodynamic properties of the liquid crystal are determined by a free energy
@@
 
-The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation accoring to
+The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
 \begin{equation}
@@
 \end{equation}
-wheras we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
+whereas we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
 $N$ was the number of points in the time series, typically one for every 100 LB time steps.
@@
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
+\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minimum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
 \label{tab1}
@@
 As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
-analytical solutions that minimize the free energy functional Eq.\ref{FE} in the high-chirality limit 
+analytical solutions that minimise the free energy functional Eq.\ref{FE} in the high-chirality limit 
 and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
@@
 $\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
-For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
+For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
 They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
@@
 The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
-regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) 
+regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
 and break up of the network into a travelling helical wave (BPII-2, solid circle).}
@@
 any additional contribution from the liquid crystal.
-The top picutre shows the date over time, whereas the graph at the bottom gives 
-the same data versus accummulated strain.
+The top picture shows the date over time, whereas the graph at the bottom gives 
+the same data versus accumulated strain.
 For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
@@
 i.e. it takes a displacement of six unit cells in flow directions 
-until the network has moved one unit cell in voriticity direction.
+until the network has moved one unit cell in vorticity direction.
 
@@
 When the helicity of the underlying cholesteric phase is changed from left- to a 
-righthanded the pattern becomes a mirror image of the former one. The sign of 
+right-handed the pattern becomes a mirror image of the former one. The sign of 
 $v_z$ change and also the sense of motion is inverted.
@@
 If the flow rate exceeds a certain critical value BPII does not undergo the 
-affine transformation descibed above and adopts another mode of flow. 
+affine transformation  above and adopts another mode of flow. 
 Fig. \ref{bp2-high} shows a sequence of cuts through the director 
@@
 
-In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$.
-At the moment we do not have a hard criterium for the critical
-shear rate at hand. However, it seems reasonable that the critical shear rate 
-depends on thermodynamic and kinetic parameters such as the effective 
-temperature $\tau$, the chirality $\kappa$ and the rotational diffusion 
-constant $\Gamma$.
+In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
+corresponds to Ericksen numbers of ${\cal E}\simeq10$, based on our
+parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
+and length scales ($p/2$).
 
@@
 For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
-they become  unregular and finally disappear after a few
+they become  irregularunregular and finally disappear after a few
 cycles. Shear-thinning is slightly stronger than in BPII. 
@@
 Shortly after it goes negative (the total shear stress is still
-positive as the backgound viscosity has been subracted) and positive
+positive as the background viscosity has been subtracted) and positive
 again, forming a complete cycle. 
@@
 Hence, the region of negative stress can be 
-regarded as a phase during which elastic forces excert a 
+regarded as a phase during which elastic forces exert a 
 'pull' onto the network while it tries to reach a more 
@@
 observed (bottom left) before the network quickly returns
-to its helical state.
-There is no perceivable movement of the network in vorticity 
-direction at any stage of the cycle.
+to its helical state. There is no perceivable movement of the 
+network in vorticity direction at any stage of the cycle.
 
@@
 
-Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) which features yield stress 
-behaviour at very low flow rates and the onset of power-law behaviour 
-at slightly larger shear rates (viz. Fig. \ref{bp1-rheo}) lies another 
-region ($3.91\e{-5}\le\gd< 6.25\e{-4}$) where the network flows with periodically 
-recurring patterns. The transition between these two regimes is shown
-in Fig. \ref{bp1-1_bp1-2} in terms of deviatoric stress over total strain.
+Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) at 
+at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
+and Ericksen numbers $0.5\le{\cal E}<10$ 
+lies another region where the network flows with periodically 
+recurring flow-induced conformations (viz. Fig. \ref{bp1-rheo}). 
+We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
+is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress over total strain. 
+Obviously a qualitative difference between these two is the absence of 
+large stress fluctuations that occur during one typical cycle in regime BPI-1.
+Remember that for even lower $\gd$ the deviatoric stress becomes 
+temporarily negative (viz. Fig. \ref{bp1-fe-yield}), which 
+caused destabilisation of the periodic network and led to the 
+amorphous configuration in the steady state. Hence, an explanation 
+for this phenomenon could be that in BPI-2 the network finds a 
+kinetic path to bypass these large amplitude fluctuations and 
+prevents the network from destabilising.
 
-The reason for this qualitative difference is rather obscur. A possible 
-explanation is that due to the larger flow rate the network kinetically 
-bypasses the relatively large stress fluctuations during one cycle which
-are typical in regime BPI-1. For even lower $\gd$ the deviatoric stress
-goes actually negative (viz. Fig. \ref{bp1-fe-yield}). This destabilises
-the periodic network very quickly and leads to an amorphous configuration
-in the steady state.
+Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
+BPI in steady shear flow. Contrary to BPII-1 at these shear rates
+, BPI-2 does not resemble the equilibrium configuration
+undergoing an affine transformation. It features rather intricate, flow-induced 
+conformations. Surprisingly this state is kinetically connected to the equilibrium 
+state as when the shear flow is switched off the flow-induced configuration 
+reverts to a perfect quiescent BPI unless it got stuck in a local free energy 
+minimum.
 
-Fig. \ref{bp1-med} shows snapshots of BPI in steady shear flow. 
-Contrary to BPII the flow-induced conformations of BPI differ significanly 
-from the appearance of BPI at equilibrium. The depicted 
-state is rather surprisingly still kinematically connected to the equilibrium state.
-When the shear flow is switched off the flow-induced configuration reverts
-to its original configuration of a BPI network at rest.
-%%%%%%%%%%%%
-%The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
-%%%%%%%%%%%
-After each cycle the position of the network is displaced with respect to the beginning of the cycle.
-However, neither the magnitude nor the direction of the offset obey a clear rule,
-are different and non-integer for every shear rate in this regime of flow. 
-This suggests the described phenomenon constitutes an instance of rheochaos in this 
-regime of liquid crystalline flow.
+The recurrence period of conformations with visually similar appearance is 
+less straightforward to determine. In flow direction a multiple of $1/\gd$,
+i.e. $\tau_F=n/\gd, n\in\cal{N}$ generally holds, e.g. $n=1,2,4$ for $\gd=3.91\e{-5},
+7.81\e{-5},1.563\e{-4}$. The analysis is complicated by the fact
+that for $n>1$ thermodynamically equivalent, mirror-image 
+conformations arise after shorter times. For instance at $\gd=7.81\e{-5}$
+the conformations at times $t$ and $t+1/\gd$ are mirror-images, whereas
+those at $t$ and $t+2/\gd$ coalesce in flow direction.
+
+However, there is no such thing as a definite recurrence period $\tau_V$ 
+in vorticity direction. The direction and magnitude of the displacement
+in other directions (i.e. both in gradient and vorticity direction) 
+is random, non-integer and depends sensitively on the imposed shear rate,
+suggesting BPI-2 constitutes an instance of rheochaos.
 
@@
 \includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
-\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page.
-The network is displaced during every cycle, but the magnitude and direction of the offset appear to be chaotic.}
+\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
+The sequence shows a typical cycle of shear-induced transformations in the 
+steady state at $\gd=1.56\e{-4}$ and time steps 
+$t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
+During every cycle the network is displaced in gradient and vorticity direction, 
+but the magnitude and direction of the offset appear to be chaotic and do not 
+obey a simple rule.}
 \label{bp1-med}
@@
 
-Fig.
-\ref{bp1-velo} depicts the secondary velocity components $v_y$ and $v_z$ in the same fashion as Fig.
-\ref{bp2-velo} for BPII.
-Again, the orientation of the components are as mirror images of each other if the helicity of the cholesteric changes from left- to right-handed.
-The emerging pattern is different from the one observed before in BPII, but agrees with
-the previous result in the sense that secondary velocities are largest in the regions
-where large conformational changes take place.
+Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ for left- and right-handed cholesteric
+helix. The emerging pattern is similar to that of BPII from 
+Fig. \ref{bp2-velo}, but differs in the aspect that there is no
+arrangement in bands. However, just as in case of BPII both patterns
+are mirror images and feature the highest velocities in the regions where 
+the largest conformational changes take place.
 
@@
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a snapshot of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero.
-The colour code gives the magnitude and sign of the z-component.
-The vertical and horizontal direction are the gradient and vorticity direction, respectivelty.
-Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
+The pictures show a snapshot of the periodically recurring patterns in the 
+secondary velocity components $(0,v_y,v_z)$, i.e. the component $v_x$ in flow direction 
+has been projected out. The colour code gives the magnitude and sign 
+of the z-component. The vertical and horizontal direction are the gradient and 
+vorticity direction, respectively, with the flow into/out of the plane in the 
+upper/lower half of the images.}
 \label{bp1-velo}
@@
 
-The complexity of flow-induced states and the seemingly chaotic but 
-periodic movement of the BPI network raises the question how robust 
-these results are with respect to the system size.
-Although the pitch length of 64 LBU and the equivalent of 32 LBU per
-unit cell provide enough resolution to track the motion of the 
-network in flow it is not clear whether 64 unit cells in total are enough to 
-mimick true bulk behaviour.
+The intricacy and periodically recurrence of the flow-induced conformations 
+in BPI raises the question how robust these results are with respect to 
+system size. Although a pitch length of 64 LBU provides enough resolution 
+for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
+in total are actually sufficient to avoid finite size effects.
+Unfortunately the computational costs of simulations, given that large strains have to
+be simulated in order to reach a steady state. Therefore we opted for a 
+comparison with smaller runs.
 
-Unfortunately the computational costs are still prohibitive for a run 
-with 512 unit cells, an equivalent of 8 unit cells along each coordinate 
-direction. Therefore we opted for a comparison with a smaller run consiting
-of 8 unit cells, 2 along each dimension.
-Fig. \ref{bp1-2uc4uc} shows a comparison of the apparent viscosity $\eta_{app}$ 
-of runs with 8 and 64 unit cells at flow rates 
-$\gd=9.76\e{-6}$ and $1.95\e{-5}$, respectively.
-We do not visualise the conformations directly but rather use the fact that 
-they are coded in the shape of each individual curve. Obviously there 
-are stiking differences between 
-both system sizes, but the recurrence period is hardly affected. In the slower
-run the smaller system apparently follows the same branch for a while but then
-deviates from the larger run. 
-This result indicates that we cannot know with certainty if the particular flow patterns 
-we find at runs with 64 unit cells do not suffer from the imposed periodicity.
-Nevertheless, the chaotic behaviour and the recurrency seem to be a generic
-feature of BPI in this flow regime.
+Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
+deviatoric stress $\Pi_{xy}$ of to different runs, one consisting of 8 
+unit cells and the other one with 64 unit cells.
+Obviously there are large deviations in both the visual appearance of the 
+network as well as in the time-dependence of the deviatoric stress.
+It takes BPI also considerably longer in the smaller run to reach the 
+steady state with its flow-induced recurring disclination network.
+
+These results indicate that we cannot know with certainty if the individual 
+flow-induced conformations that we find for 64 unit cells are not affected 
+by imposed periodicity. Nevertheless, the comparison suggests that 
+the recurrence period and the complexity of the network in flow are  
+generic features of BPI in this regime of flow.
 
@@
 \includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{}
+\caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
+The dark/red network in top picture shows a snapshot of run that consists 
+of $2^3=8$ unit cells, whereas the light/cyan network displays the 
+central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
+The bottom picture shows the deviatoric stress over time. Interestingly, 
+an intermediate sized run with $3^3=27$ unit cells led to results that 
+were identical to the smaller run with 8 unit cells. The shear rate was $\gd=7.81\e{-5}$.
+}
 \label{bp1-2uc4uc}
@@
 \includegraphics[width=0.4\textwidth]{dir+z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis.
+\caption{Director field of BPI at high shear rate: The pictures show a slice 
+of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction 
+with the flow/gradient direction oriented along the horizontal/vertical axis. 
 The colour code gives the magnitude of the in-plane component of the director field. 
@@
 
-A breakup of the network at large flow rates is also observed
-in BPI. Interestingly the window of critical shear rates 
-$3.125\e{-4}\le\gd_c\le 6.25\e{-4}$  is the same we find for BPII.  
-Contrary to BPII, which flows in form of a travelling helical wave
-(viz. Fig. \ref{bp2-high}) the preferred configuration is now
-a quasi two-dimensional state, shown in Fig. \ref{bp1-high} as 
-slices through the director field along the voritcity direction.      
-The sequence shows that BPI undergoes the affine transformation 
-in gradient-flow plane often seen in simple shear flow.
-There are distinctive regions where the predominant orientation 
-of the director is along the vorticity direction and others
-where it lies in flow-gradient plane. Occasionally while 
-flowing past each other this produces regions that 
-resemble local double twist.
+At high flow rates we observe a break-up of the flow-induced 
+disclination network of regime BPI-2. However, 
+contrary to BPII, where the configuration is a travelling helical 
+wave (viz. Fig. \ref{bp2-high}), BPI prefers a quasi 
+two-dimensional, translationally invariant state, which is shown in 
+Fig. \ref{bp1-high}. The snapshots visualise nicely the affine 
+transformation of simple shear flow in gradient-flow plane.
+There are distinctive regions where the director field is 
+predominantly oriented along the vorticity direction and 
+other regions where it is oriented rather in flow-gradient plane. 
+Occasionally while moving past each other these rolls resemble 
+regions of local double twist.
+Interestingly the critical shear rate for the break-up lies in the same 
+window as that of transition between BPII-1 and BPII-2, 
+ $3.125\e{-4}\le\gd_c\le 6.25\e{-4}$ and occurs at similar 
+Ericksen numbers ${\cal E}\simeq 10$.
 
@@
 \section{Conclusions}
+
+In summary, our work constitutes the first simulation of bulk flow behaviour 
+of cubic BPs in simple shear flow. Although experimental evidence of our results 
+is currently not available, we believe this work can shed some light into the 
+flow properties of complex liquid-crystalline phases.
+
+We were able to characterise the rheology of cubic BPs and identified 
+two different flow regimes for BPII and three different ones in case of BPI. 
+Below Ericksen numbers ${\cal O}(10)$ BPII exhibits weak shear-thinning and 
+obeys a power-law.  The BPII disclination network breaks up and reconnects 
+in the flow, which leads to a periodically recurring dependence 
+of the deviatoric shear stress. The flow-induced conformation look generally
+very similar to the quiescent network at equilibrium. 
+While being affinely transformed due to the shear flow
+the disclination network moves in the vorticity direction. The sense of
+motion is directly linked to the helicity of the underlying cholesteric phase.
+
+At larger Ericksen numbers the flowing BPII network breaks up completely into 
+a simple cholesteric phase and flows as a travelling helical wave.
+
+Interestingly, BPI shows a flow behaviour that is very different to that of BPII.
+This is a direct consequence of the topological differences in both disclination 
+network. Below Ericksen numbers ${\cal O}(0.1)$ the BPI network cannot flow and 
+retain a regular appearance. Shortly after the onset of the shearing 
+very large stress fluctuations occur. At the lowest shear rates $\gd<9.76\e{-6}$
+the contribution of the network to the total deviatoric stress is actually temporarily 
+negative during one cycle. These fluctuations destabilise the network and 
+eventually trigger its dissolution into an amorphous state with 
+dynamic yield stress.
+
+At slightly larger Ericksen numbers $0.1<{\cal E}<10$ shear thinning is 
+slightly stronger than for BPII, but BPI does not show simple power-law 
+behaviour. 
+Instead, it appears BPI kinematically bypasses the large stress fluctuations and 
+flows with very complex, periodically recurring flow-induced conformations.
+The offset of the network after each cycle is random in magnitude and direction.
+However, despite their complex appearance the flow-induced conformations are 
+topologically connected to the quiescent BPI and switching off of the 
+shear flow often restores a perfect equilibrium.
+It is possible to interpret these findings as an instance of 
+deterministic rheochaos.
+
+Just as in case of BPII, the BPI network breaks up at Ericksen numbers ${\cal O}(10)$,
+but adopts a quasi-two-dimensional configuration that is translationally 
+invariant in vorticity direction.
 
@@
 \begin{acknowledgments}
-Thanks folks!
+We acknowledge support by EPSRC grant nos. EP/E045316 and EP/E030173, 
+the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
+We thank Peter J. Collings for very stimulating discussions. 
+MEC holds a Royal Society Research Professorship.
 \end{acknowledgments}

### 2012-10-19 trunk/htdocs/references.tex
Expanded porous media section
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1792 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
+
+\bibitem{batchelor}
+G.K. Batchelor
+\textit{An Introduction to Fluid Mechanics},
+Cambridge University Press (1967).
+
+\bibitem{blake}
+J.R. Blake, A spherical envelope approach to ciliary propusion,
+\textit{J. Fluid Mech.}, \textbf{46}, 199, 1971.
 
@@
 
+\bibitem{lighthill}
+M.J. Lighthill,
+On the squirming motion of nearly spherical deformable bodies through
+liquid at very small Reynolds numbers,
+\textit{Comm. Pure Appl. Math.}, \textbf{5}, 109, 1952.
+
+\bibitem{isaac}
+I. Llopis Fust\'e,
+\textit{Hydrodynamic cooperativity in micro-swimmer suspensions},
+Ph.D. Thesis, University of Barcelona, 2008.
+
 \bibitem{mpi-standard}

### 2012-10-29 trunk/papers/bprheo/bprheo.tex
Applied Kevin's corrections in the text part
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1800 40876441-3496-0410-a613-a2fca35b2ae2

@@
 In the uniaxial approximation, the order parameter is given by
-$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
-with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
+$Q_{\alpha \beta}= q_s ( \hat{n}_\alpha \hat{n}_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
+with $\hat{{\mathbf n}}$ the director field and $q_s$ the amplitude of nematic
 order. More generally,
 the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
-characterizes the local degree of orientational order.
+characterises the local degree of orientational order.
 The thermodynamic properties of the liquid crystal are determined by a free energy
@@
 The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
-The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.
+The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure~\cite{Alexander:2006}.
 
-A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
-which are given by
-\begin{eqnarray}
-\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
-\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
-\end{eqnarray}
+A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature 
+\begin{equation}
+\tau=\frac{27(1-\gamma/3)}{\gamma}
+\end{equation}
+and chirality 
+\begin{equation}
+\kappa=\sqrt{\frac{108 K q_0^2}{A_0 \gamma}},
+\end{equation}
+which measures the ratio of gradient to bulk free energy.
+
 The dynamical evolution of the order parameter is given by the equation 
 \begin{equation}
-\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
+\left(\partial_t+ v_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
 \label{op-eom}
@@
 The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
-The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
+The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta v_\alpha$,
 \begin{eqnarray}
@@
 setting $\xi=0.7$ in our simulations.
-${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
+${\mathbf H}$ is the molecular field, which is the functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
 \begin{equation}
@@
 by the continuity equation
-$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
+$\partial_t \rho = -\partial_\alpha(\rho v_\alpha)$, and
 the following Navier-Stokes equation:
 \begin{eqnarray}
-\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
-&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha u_\beta \nonumber\\
-& &\hspace*{-1.5cm} +\; \partial_\beta u_\alpha
-+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
+\partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
+&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha v_\beta \nonumber\\
+& &\hspace*{-1.5cm} +\; \partial_\beta v_\alpha
++(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}\}. 
 \label{NSE}
@@
 The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
-of the LB equations \cite{Denniston:2001}.
+of the Lattice Boltzmann (LB) equations \cite{Denniston:2001}.
 At low flow rates the fluid can be considered as incompressible, so that the
 last term on the right hand side of Eq.~\ref{NSE} remains small.
-$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units.
 The pressure tensor reads explicitly
@@
 
-The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation accoring to
+It is possible to define a dimensionless number that describes the deformation
+of the director field under flow. The so-called Ericksen number
+is defined as 
+\begin{equation}
+Er=\frac{\eta V L}{K}
+\end{equation}
+with $\eta$ and $K$ being the viscosity and elastic constant and
+$V$ and $L$ as typical velocity and length scale.
+
+The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
 \begin{equation}
@@
 \end{equation}
-wheras we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
+where we subtract the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
 $N$ was the number of points in the time series, typically one for every 100 LB time steps.
@@
 and~\ref{NSE} is solved by means of a
-hybrid scheme~\cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
-with predictor-corrector scheme for the continuity equation and
-Eq.~\ref{NSE}, and a finite difference scheme for the equation of motion of
-the tensor order parameter Eq.\ref{op-eom}. More details on the algorithm can
+hybrid scheme~\cite{Marenduzzo:2007}. 
+More details on the algorithm can
 be found in~\cite{Denniston:2001, Denniston:2004}.
@@
 The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
-$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LBU of stress
+$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LB unit of stress
 is equal to about $10^8$~Pa. Further details about the conversion 
@@
 
-The thermodynamic states are characterised by an effective temperature 
-$\tau$ and the chirality $\kappa$, which measures the ratio of gradient to bulk free energy.
-For the vast majority of simulations we opted for equilibrium states that are sufficiently 
-far away from the cholesteric-isotropic transition, namely
-$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ (BPII), respectively.
+For the vast majority of simulations we opted for thermodynamic states that are 
+well inside the equilibrium region of the individual blue phase and far away 
+from the cholesteric-isotropic transition. Temperature and chirality were 
+$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
+For these parameters the total free energy density $f$ remained always negative, 
+even at the highest flow rate and despite larger gradient contributions $f_g$.
 We performed also a number of simulations at metastable states at higher and lower temperatures 
 and different at different chiralities, but didn't find any significant differences in the 
-generic flow behaviour that is described below.
+general flow behaviour that is described below.
 However, for thermodynamic states that are closer to the phase boundary the disclination 
@@
 Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
-the top (bottom) half of the planes moving in positive (negative) x-direction and the 
+the top (bottom) part of the system flowing in positive (negative) x-direction and the 
 velocity gradient along the y-direction.
-We applied a correction for spurious momentum currents that emerge due to required interpolations across 
-the Lees-Edwards planes \cite{Henrich:2012a}.
 The shear rates were varied over more than two orders of magnitude from about 
 $\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
-For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
+For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
 They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
-BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
+BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
 Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
@@
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). 
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII versus time (top) and strain (bottom). 
 For clarity the individual curves have been shifted along the time and strain axis. 
 The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
-regular periodic breakup and reconnecting of the discination lines (BPII-1, open circles) 
+regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
 and break up of the network into a travelling helical wave (BPII-2, solid circle).}
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
 \label{bp2-rheo}
@@
 We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
-than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
+than that of BPI. BPII has simple cubic symmetry and the disclination lines intersect and form a characteristic network of nodes.
 Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
@@
 any additional contribution from the liquid crystal.
-The top picutre shows the date over time, whereas the graph at the bottom gives 
-the same data versus accummulated strain.
+The top picture shows the data over time, whereas the graph at the bottom gives 
+the same data versus accumulated strain.
 For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
+is because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
 
-The inset shows a kind of flow curve, defined as time averages of 
-the deviatoric stress tensor $\bar{Pi}_{xy}$ as a function of shear rate $\gd$.
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
+The inset shows a flow curve, defined as time averages of 
+the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
 For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
@@
 
-In the next paragraph we will investigate this in more detail by looking at the 
-kinetics of the disclination network.
+In the next paragraph we will investigate some of these aspects in more detail 
+by looking at the kinetics of the disclination network.
 
@@
 \caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
-underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
-i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
+underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
+The view is in x-direction. This means the velocity ${\mathbf v}$ 
+has been projected onto a plane perpendicular to the flow direction. 
 The vertical and horizontal direction are the gradient and vorticity direction, respectively.
@@
 
-Interestingly, while being displaced with the flow the entire network moves perpendicular 
-to the flow and velocity gradient in the vorticity direction.
-A similar behaviour has been recently observed for blue phases in shear flow and 
+Interestingly, while being displaced with the flow the entire network moves 
+in the vorticity direction.  A similar behaviour has been recently 
+observed for blue phases in shear flow in 
 confined geometries ~\cite{Henrich:2012b}.
-But contrary to the stick-slip motion that has been reported there the movement is 
+In contrast to the stick-slip motion that has been reported there the movement is 
 steady. It occurs in such a manner that the positions of breakup and reconnection 
@@
 i.e. it takes a displacement of six unit cells in flow directions 
-until the network has moved one unit cell in voriticity direction.
+until the network has moved one unit cell in vorticity direction.
 
-Fig. \ref{bp2-velo} shows the disclination network sometime between reconnecting 
+Fig. \ref{bp2-velo} shows the disclination network between reconnecting 
 and the next breakup with superimposed velocity vectors.
 Shown are what we may call secondary velocity components, obtained by projecting the
-velocity onto a plain perpendicular to the flow direction.
-This allows to get rid of the dominating velocity component $v_x$ and to
-visualise the two much smaller components $v_y$ and $v_z$.
+velocity onto a plane perpendicular to the flow direction.
+This gets rid of the dominating velocity component $v_x$ and 
+allows to visualise the two much smaller components $v_y$ and $v_z$.
 The magnitude of the secondary components is typically in the range of a 
@@
 
-Characteristic bands are visible, which are oriented along the vorticity direction and
-normally hidden behind the primary flow component.
-When the helicity of the underlying cholesteric phase is changed from left- to a 
-righthanded the pattern becomes a mirror image of the former one. The sign of 
-$v_z$ change and also the sense of motion is inverted.
+Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
+oriented along the vorticity direction. They are obscured by the
+primary flow component if the above described projection procedure
+is not applied.
+When the helicity of the underlying cholesteric phase is changed from 
+left-handed to right-handed, 
+the pattern becomes a mirror image of the former one. The sign of 
+$v_z$ changes and the sense of motion is inverted.
 Further quantitative evidence for a direct link between the sense of motion 
@@
 
-In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$.
-At the moment we do not have a hard criterium for the critical
-shear rate at hand. However, it seems reasonable that the critical shear rate 
-depends on thermodynamic and kinetic parameters such as the effective 
-temperature $\tau$, the chirality $\kappa$ and the rotational diffusion 
-constant $\Gamma$.
+In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
+corresponds to Ericksen numbers of ${\it Er}\simeq10$, based on our
+parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
+and length scales ($p/2$).
 
@@
 
-BPI has body-centered symmetry and contrary to BPII the disclination lines
+BPI has body-centered symmetry and unlike BPII. The disclination lines
  are well separated and do not intersect.
@@
 
-We present again key aspects in a kind of overview before we 
+We present again key aspects in an overview before we 
 address some features in more detail. Fig. \ref{bp1-rheo} shows the 
-apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
+apparent viscosity $\eta_{app}$ versus time and strain for the same shear rates
 as in Fig. \ref{bp2-rheo}.
 For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
-the disclination network occurs similar to and at the same critical shear rate as 
-the one observed in BPII-2 but with an order structure that is a quasi 
+the disclination network occurs. This is the same critical shear rate as 
+the one observed in BPII-2. But this time the configuration is quasi 
 two-dimensional.
@@
 For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
-they become  unregular and finally disappear after a few
-cycles. Shear-thinning is slightly stronger than in BPII. 
+these oscillations become very erratic. There is no regular recurrence and 
+after a few cycles the stress pattern is completely irregular. 
+Shear-thinning is slightly stronger than in BPII. 
 The average shear stress during one cycle can be fitted reasonably 
@@
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (top) 
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI versus time (top) 
 and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
@@
 
-There are two aspects which are worth mentioning. First of all, we observe a strong contribution of 
+There are two aspects which are worth mentioning. First, we observe a strong contribution of 
 the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
@@
 More surprisingly the frequency of the fundamental mode is the same 
-for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is completely 
-contrasted by what is observed at lower flow rates. By doubling the flow rate from 
-$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the position of the peaks doubles in the spectrum 
+for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
+in contrast to what is observed at lower flow rates. By doubling the flow rate from 
+$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
 moves to the double frequency. This means that the fundamental mode at flowrate $\gd=7.81\e{-5}$ 
 coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on,
-which is the expected behaviour. At higher shear rates in regime BPI-3 this lock-in of 
+which is the expected behaviour. At higher shear rates in regime BPI-3 this 'lock-in' of 
 flow modes disappears again.
@@
 \includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture and inset show spectral
-components and the lock-in of flow modes, visible in the fixed peak position for certain flow rates. 
-The inset represents a blow-up of the same data. The frequency is given in radians.}
+\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
+spectral components and the 'lock-in' of flow modes, 
+visible in the fixed peak position for certain flow rates.  
+The inset shows the same data on an expanded scale. 
+The frequency is given in radians.}
 \label{bp1-spectrum}
@@
 
-The rheological response of BPI at low shear rate which exhibits a
-dynamic yield stress is strikingly different from that of BPII.
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{Average deviatoric stress and free energy density of BPI at low shear rate: 
+The negative branch in the stress is related to a local maximum and a following local 
+minimum in average free energy density. This creates unstable conditions that lead 
+to a dissolving regular network and an amorphous state which features a yield stress.}
+\label{bp1-fe-yield}
+\end{figure}
+
+The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
+is strikingly different from that of BPII and exhibits a
+dynamic yield stress.
 An explanation for this strange behaviour can be found by 
@@
 Shortly after it goes negative (the total shear stress is still
-positive as the backgound viscosity has been subracted) and positive
+positive as the background viscosity has been subtracted) and then positive
 again, forming a complete cycle. 
 This cycle is repeated for a couple of times before 
-the amorphous configuration emerges and any regular pattern
-disappears.
+any regular pattern disappears and the amorphous configuration 
+emerges.
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
+\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
+\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
+Depicted is the transition from flow-induced, intertwined helices that 
+undergo a recurring affine transformation (top left and centre, both at 
+time step $t=4.0\e{5}$; top right at time step $t=7.0\e{5}$) with an intermediate 
+conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) 
+to a distorted state (bottom centre at time step $t=1.7\e{6}$) and 
+finally to an amorphous network (bottom right at $t=2.2\e{6}$). 
+The picture at the top centre shows the network in gradient direction.
+All other pictures have the view along the vorticity direction.}
+\label{bp1-low}
+\end{figure*}
+
 The negative branch in the deviatoric stress is connected
@@
 Hence, the region of negative stress can be 
-regarded as a phase during which elastic forces excert a 
-'pull' onto the network while it tries to reach a more 
+regarded as a phase during which elastic forces exert a 
+'pull' on the network while it tries to reach a more 
 favourable configuration. However, the low flow rate
 causes this unstable intermediate state to be exposed to these
-pulling elastic forces for a relatively long time, long enough to 
+elastic forces for a relatively long time, long enough to 
 create small distortions which finally break the symmetry of the
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density.
-This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
-\label{bp1-fe-yield}
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
+\caption{Deviatoric stress $\Pi_{xy}$ versus strain at the transition between regime
+BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
+and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
+where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
+throughout positive, but exhibits very large fluctuations during one cycle, 
+which are absent in regime BPI-2 at higher shear rates.}
+\label{bp1-1_bp1-2}
 \end{figure}
@@
 observed (bottom left) before the network quickly returns
-to its helical state.
-There is no perceivable movement of the network in vorticity 
-direction at any stage of the cycle.
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
-\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).
-All pictures have the vie along the vorticity direction apart from the one at the top centre which shows the network in gradient direction.}
-\label{bp1-low}
-\end{figure*}
+to its helical state. There is no perceivable movement of the 
+network in vorticity direction at any stage of the cycle.
 
@@
 Finally, the network loses any residual order and 
-transforms into an amorphous state with virtually constant
-yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LU.
+transforms into an amorphous state with almost constant
+yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
 
@@
 
-Intermediate shear rates in the sense of this study of BPI are characterised 
-by the missing negative branch in the deviatoric stress. It is unclear 
-why this feature is missing at higher flow rates. A possible explanation is that 
-above a certain flow rate the network simply skips the comparably small 
-window of negative stress in Fig. \ref{bp1-fe-yield} and finds a quicker 
-route to get back on the positive branch.
+Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) and 
+at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
+and Ericksen numbers $0.5\le{\it Er}<10$ 
+lies another region where the network flows with periodically 
+recurring conformations (Fig. \ref{bp1-rheo}). 
+We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
+is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress versus total strain. 
+A qualitative difference between these two is the absence of 
+large stress fluctuations that occur during one typical cycle in regime BPI-1.
+Remember that for even lower $\gd$ the deviatoric stress becomes 
+temporarily negative (Fig. \ref{bp1-fe-yield}), which 
+caused destabilisation of the periodic network and led to the 
+amorphous configuration in the steady state. Hence, an explanation 
+for this phenomenon could be that in BPI-2 the network finds a 
+kinetic path to bypass these large amplitude fluctuations that 
+prevents the network from destabilising.
 
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{}
-\label{bp1-1_bp1-2}
-\end{figure}
+Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
+BPI in steady shear flow. Contrary to BPII-1 at these shear rates
+, BPI-2 does not resemble the equilibrium configuration
+undergoing an affine transformation. It features rather intricate, flow-induced 
+conformations. Surprisingly this state is kinetically connected to the equilibrium 
+state as when the shear flow is switched off the flow-induced configuration 
+reverts to a perfect quiescent BPI unless it got stuck in a local free energy 
+minimum.
 
-Fig. \ref{bp1-med} shows snapshots of BPI in steady shear flow. 
-Contrary to BPII the flow-induced conformations of BPI differ significanly 
-from the appearance of BPI at equilibrium. The depicted 
-state is rather surprisingly still kinematically connected to the equilibrium state.
-When the shear flow is switched off the flow-induced configuration reverts
-to its original configuration of a BPI network at rest.
-%%%%%%%%%%%%
-%The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
-%%%%%%%%%%%
-After each cycle the position of the network is displaced with respect to the beginning of the cycle.
-However, neither the magnitude nor the direction of the offset obey a clear rule,
-are different and non-integer for every shear rate in this regime of flow. 
-This suggests the described phenomenon constitutes an instance of rheochaos in this 
-regime of liquid crystalline flow.
+The recurrence period of conformations with visually similar appearance is 
+less straightforward to determine. In the flow direction relation like
+$\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
+7.81\e{-5},1.563\e{-4}$ for example. The analysis is complicated by the fact
+that for $n>1$ thermodynamically equivalent, mirror-image 
+conformations arise after shorter times. For instance at $\gd=7.81\e{-5}$
+the conformations at times $t$ and $t+1/\gd$ are mirror-images, whereas
+those at $t$ and $t+2/\gd$ coalesce in flow direction.
+
+However, there is no definite recurrence period $\tau_V$ 
+in the vorticity direction. The direction of the displacement
+in other directions (i.e. both in gradient and vorticity direction) 
+is not simply along coordinate directions. 
+The magnitude of the displacement show no clear pattern, is non-integer 
+and depends sensitively on the imposed shear rate,
+suggesting BPI-2 constitutes an instance of 'rheochaos' [ref].
 
@@
 
-Fig.
-\ref{bp1-velo} depicts the secondary velocity components $v_y$ and $v_z$ in the same fashion as Fig.
-\ref{bp2-velo} for BPII.
-Again, the orientation of the components are as mirror images of each other if the helicity of the cholesteric changes from left- to right-handed.
-The emerging pattern is different from the one observed before in BPII, but agrees with
-the previous result in the sense that secondary velocities are largest in the regions
-where large conformational changes take place.
+Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ for left- and right-handed cholesteric
+helix. The emerging pattern is similar to that of BPII from 
+Fig. \ref{bp2-velo}, but differs in that there is no
+arrangement in bands. However, just as in case of BPII, both patterns
+are mirror images and feature the highest velocities in the regions where 
+the largest conformational changes take place.
 
@@
 
-The complexity of flow-induced states and the seemingly chaotic but 
-periodic movement of the BPI network raises the question how robust 
-these results are with respect to the system size.
-Although the pitch length of 64 LU and the equivalent of 32 LU per
-unit cell provide enough resolution to track the motion of the 
-network in flow it is not clear whether 64 unit cells in total are enough to 
-mimick true bulk behaviour.
+The intricacy and periodic recurrence of the flow-induced conformations 
+in BPI raises the question: how robust these results are with respect to 
+system size? Although a pitch length of 64 LBU provides enough resolution 
+for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
+in total are actually sufficient to avoid finite size effects.
+Unfortunately the computational costs of larger simulations are still
+prohibitive, given that large strains have to
+be simulated in order to reach a steady state. Therefore we opted for a 
+comparison with smaller runs.
 
-Unfortunately the computational costs are still prohibitive for a run 
-with 512 unit cells, an equivalent of 8 unit cells along each coordinate 
-direction. Therefore we opted for a comparison with a smaller run consiting
-of 8 unit cells, 2 along each dimension.
-Fig. \ref{bp1-2uc4uc} shows a comparison of the apparent viscosity $\eta_{app}$ 
-of runs with 8 and 64 unit cells at flow rates 
-$\gd=9.76\e{-6}$ and $1.95\e{-5}$, respectively.
-We do not visualise the conformations directly but rather use the fact that 
-they are coded in the shape of each individual curve. Obviously there 
-are stiking differences between 
-both system sizes, but the recurrence period is hardly affected. In the slower
-run the smaller system apparently follows the same branch for a while but then
-deviates from the larger run. 
-This result indicates that we cannot know with certainty if the particular flow patterns 
-we find at runs with 64 unit cells do not suffer from the imposed periodicity.
-Nevertheless, the chaotic behaviour and the recurrency seem to be a generic
-feature of BPI in this flow regime.
+Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
+deviatoric stress $\Pi_{xy}$ of to different runs, one consisting of 8 
+unit cells and the other one with 64 unit cells.
+Obviously there are large deviations in both the visual appearance of the 
+network as well as in the time-dependence of the deviatoric stress.
+In the smaller run with $2^3=8$ unit cells takes BPI also considerably 
+to reach the 
+steady state with its flow-induced recurring disclination network.
+
+These results indicate that we cannot know with certainty if the individual 
+flow-induced conformations that we find for 64 unit cells are not affected 
+by imposed periodicity. Nevertheless, the comparison suggests that 
+the recurrence period and the complexity of the network in flow are  
+generic features of BPI in this regime of flow.
 
 \begin{figure}[htpb]
+\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
 \includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Apparent viscosity $\eta_{app}$ for 8 (solid lines) and 64 unit cells (dashed lines)
-at $\gd=9.76\e{-6}$ and $1.95\e{-5}$. The shape of each curve differs largely, but the recurrence frequencies are hardly affected by the system size.}
+\caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
+The dark (red) network in top picture shows a snapshot of run that consists 
+of $2^3=8$ unit cells, whereas the light (cyan) network displays the 
+central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
+The bottom picture shows the deviatoric stress over time. Interestingly, 
+an intermediate sized run with $3^3=27$ unit cells led to results that 
+were identical to the smaller run with 8 unit cells. The shear rate was $\gd=7.81\e{-5}$.
+}
 \label{bp1-2uc4uc}
@@
 
-A breakup of the network at large flow rates is also observed
-in BPI. Interestingly the window of critical shear rates 
-$3.125\e{-4}\le\gd_c\le 6.25\e{-4}$  is the same we find for BPII.  
-Contrary to BPII, which flows in form of a travelling helical wave
-(viz. Fig. \ref{bp2-high}) the preferred configuration is now
-a quasi two-dimensional state, shown in Fig. \ref{bp1-high} as 
-slices through the director field along the voritcity direction.      
-The sequence shows that BPI undergoes the affine transformation 
-in gradient-flow plane often seen in simple shear flow.
-There are distinctive regions where the predominant orientation 
-of the director is along the vorticity direction and others
-where it lies in flow-gradient plane. Occasionally while 
-flowing past each other this produces regions that 
-resemble local double twist.
+At high flow rates we observe a break-up of the flow-induced 
+disclination network of regime BPI-2. However, 
+contrary to BPII, where the configuration is a travelling helical 
+wave (viz. Fig. \ref{bp2-high}), BPI prefers a quasi 
+two-dimensional, translationally invariant state, which is shown in 
+Fig. \ref{bp1-high}. The snapshots visualise the affine 
+transformation of simple shear flow in the gradient-flow plane.
+There are distinctive regions where the director field is 
+predominantly oriented along the vorticity direction and 
+other regions where it is oriented rather in flow-gradient plane. 
+Occasionally while moving past each other these rolls resemble 
+regions of local double twist.
+Interestingly the critical shear rate for the break-up lies in the same 
+window as that of transition between BPII-1 and BPII-2, 
+ $3.125\e{-4}\le\gd_c\le 6.25\e{-4}$ and occurs at similar 
+Ericksen numbers ${\it Er}\simeq 10$.
 
@@
 \section{Conclusions}
+
+In summary, our work constitutes the first simulation of bulk flow behaviour 
+of cubic BPs in simple shear flow. Although experimental evidence to support
+our results is currently not available, we believe this work can shed some light into the 
+flow properties of complex liquid-crystalline phases.
+
+We were able to characterise the rheology of cubic BPs and identified 
+two different flow regimes for BPII and three different ones in case of BPI. 
+Below Ericksen numbers ${\cal O}(10)$ BPII exhibits weak shear-thinning and 
+obeys a power-law.  The BPII disclination network breaks up and reconnects 
+in the flow, which leads to a periodically recurring dependence 
+of the deviatoric shear stress. The flow-induced conformation look generally
+very similar to the quiescent network at equilibrium. 
+While being affinely transformed due to the shear flow
+the disclination network moves in the vorticity direction. The sense of
+motion is directly linked to the helicity of the underlying cholesteric phase.
+
+At larger Ericksen numbers the flowing BPII network breaks up completely into 
+a simple cholesteric phase and flows as a travelling helical wave.
+
+Interestingly, BPI shows a flow behaviour that is very different to that of BPII.
+This is a direct consequence of the topological differences in both disclination 
+network. Below Ericksen numbers ${\cal O}(0.1)$ the BPI network cannot flow and 
+retain a regular appearance. Shortly after the onset of the shearing 
+very large stress fluctuations occur. At the lowest shear rates $\gd<9.76\e{-6}$
+the contribution of the network to the total deviatoric stress is actually temporarily 
+negative during one cycle. These fluctuations destabilise the network and 
+eventually trigger its dissolution into an amorphous state with 
+dynamic yield stress.
+
+At slightly larger Ericksen numbers $0.1<{\it Er}<10$ shear thinning is 
+slightly stronger than for BPII, but BPI does not show simple power-law 
+behaviour. 
+Instead, it appears BPI kinematically bypasses the large stress fluctuations and 
+flows with very complex, periodically recurring flow-induced conformations.
+The offset of the network after each cycle is random in magnitude and direction.
+However, despite their complex appearance the flow-induced conformations are 
+topologically connected to the quiescent BPI and switching off of the 
+shear flow often restores a perfect equilibrium.
+It is possible to interpret these findings as an instance of 
+deterministic rheochaos.
+
+Just as in case of BPII, the BPI network breaks up at Ericksen numbers ${\cal O}(10)$,
+but adopts a quasi-two-dimensional configuration that is translationally 
+invariant in vorticity direction.
 

### 2012-11-02 papers/bprheo/bprheo.tex
(Partly) updated pictures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1805 40876441-3496-0410-a613-a2fca35b2ae2

@@
 In the uniaxial approximation, the order parameter is given by
-$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
-with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
+$Q_{\alpha \beta}= q_s ( \hat{n}_\alpha \hat{n}_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
+with $\hat{{\mathbf n}}$ the director field and $q_s$ the amplitude of nematic
 order. More generally,
@@
 
+It is possible to define a dimensionless number that describes the deformation
+of the director field under flow. The so-called Ericksen number
+is defined as 
+\begin{equation}
+Er=\frac{\eta V L}{K}
+\end{equation}
+with $\eta$ and $K$ being the viscosity and elastic constant and
+$V$ and $L$ as typical velocity and length scale.
+
 The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
@@
 \end{equation}
-whereas we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
+where we subtract the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
 $N$ was the number of points in the time series, typically one for every 100 LB time steps.
@@
 Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
-the top (bottom) half of the planes moving in positive (negative) x-direction and the 
+the top (bottom) part of the system flowing in positive (negative) x-direction and the 
 velocity gradient along the y-direction.
-We applied a correction for spurious momentum currents that emerge due to required interpolations across 
-the Lees-Edwards planes \cite{Henrich:2012a}.
 The shear rates were varied over more than two orders of magnitude from about 
@@
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (top) and strain (bottom). 
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII versus time (top) and strain (bottom). 
 For clarity the individual curves have been shifted along the time and strain axis. 
@@
 and break up of the network into a travelling helical wave (BPII-2, solid circle).}
+The error bars indicate the maximum and minimum stresses that occur during one cycle.
 \label{bp2-rheo}
@@
 any additional contribution from the liquid crystal.
-The top picture shows the date over time, whereas the graph at the bottom gives 
+The top picture shows the data over time, whereas the graph at the bottom gives 
 the same data versus accumulated strain.
 For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
+is because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
 
-The inset shows a kind of flow curve, defined as time averages of 
-the deviatoric stress tensor $\bar{Pi}_{xy}$ as a function of shear rate $\gd$.
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
+The inset shows a flow curve, defined as time averages of 
+the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
 For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
+\includegraphics[width=0.41\textwidth]{disc-160k_run902.png}
+\includegraphics[width=0.41\textwidth]{disc-164k_run902.png}
+\includegraphics[width=0.41\textwidth]{disc-165k_run902.png}
 \caption{Disclination network of BPII in shear flow: 
@@
 
-Fig. \ref{bp2-velo} shows the disclination network sometime between reconnecting 
+Fig. \ref{bp2-velo} shows the disclination network between reconnecting 
 and the next breakup with superimposed velocity vectors.
 Shown are what we may call secondary velocity components, obtained by projecting the
-velocity onto a plain perpendicular to the flow direction.
-This allows to get rid of the dominating velocity component $v_x$ and to
-visualise the two much smaller components $v_y$ and $v_z$.
+velocity onto a plane perpendicular to the flow direction.
+This gets rid of the dominating velocity component $v_x$ and 
+allows to visualise the two much smaller components $v_y$ and $v_z$.
 The magnitude of the secondary components is typically in the range of a 
@@
 
-Characteristic bands are visible, which are oriented along the vorticity direction and
-normally hidden behind the primary flow component.
-When the helicity of the underlying cholesteric phase is changed from left- to a 
-right-handed the pattern becomes a mirror image of the former one. The sign of 
-$v_z$ change and also the sense of motion is inverted.
+Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
+oriented along the vorticity direction. They are obscured by the
+primary flow component if the above described projection procedure
+is not applied.
+When the helicity of the underlying cholesteric phase is changed from 
+left-handed to right-handed, 
+the pattern becomes a mirror image of the former one. The sign of 
+$v_z$ changes and the sense of motion is inverted.
 Further quantitative evidence for a direct link between the sense of motion 
@@
 In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
-corresponds to Ericksen numbers of ${\cal E}\simeq10$, based on our
+corresponds to Ericksen numbers of ${\it Er}\simeq10$, based on our
 parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
@@
 For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
-they become  irregularunregular and finally disappear after a few
-cycles. Shear-thinning is slightly stronger than in BPII. 
+these oscillations become very erratic. There is no regular recurrence and 
+after a few cycles the stress pattern is completely irregular. 
+Shear-thinning is slightly stronger than in BPII. 
 The average shear stress during one cycle can be fitted reasonably 
@@
 \includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density.
-This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
+\caption{Average deviatoric stress and free energy density of BPI at low shear rate: 
+The negative branch in the stress is related to a local maximum and a following local 
+minimum in average free energy density. This creates unstable conditions that lead 
+to a dissolving regular network and an amorphous state which features a yield stress.}
 \label{bp1-fe-yield}
@@
 Shortly after it goes negative (the total shear stress is still
-positive as the background viscosity has been subtracted) and positive
+positive as the background viscosity has been subtracted) and then positive
 again, forming a complete cycle. 
 This cycle is repeated for a couple of times before 
-the amorphous configuration emerges and any regular pattern
-disappears.
+any regular pattern disappears and the amorphous configuration 
+emerges.
 
@@
 \includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
-\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$): 
+\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
 Depicted is the transition from flow-induced, intertwined helices that 
-undergo a recurring affine transformation (top left and centre at 
-time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate 
+undergo a recurring affine transformation (top left and centre, both at 
+time step $t=4.0\e{5}$; top right at time step $t=7.0\e{5}$) with an intermediate 
 conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) 
@@
 finally to an amorphous network (bottom right at $t=2.2\e{6}$). 
-All pictures have the vie along the vorticity direction apart from the 
-one at the top centre which shows the network in gradient direction.}
+The picture at the top centre shows the network in gradient direction.
+All other pictures have the view along the vorticity direction.}
 \label{bp1-low}
@@
 regarded as a phase during which elastic forces exert a 
-'pull' onto the network while it tries to reach a more 
+'pull' on the network while it tries to reach a more 
 favourable configuration. However, the low flow rate
 causes this unstable intermediate state to be exposed to these
-pulling elastic forces for a relatively long time, long enough to 
+elastic forces for a relatively long time, long enough to 
 create small distortions which finally break the symmetry of the
@@
 \includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{Deviatoric stress $\Pi_{xy}$ over strain at the transition between regime
+\caption{Deviatoric stress $\Pi_{xy}$ versus strain at the transition between regime
 BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
 and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
-where the stress goes negative (viz. Fig. \ref{bp1-fe-yield}) the stress is 
-throughout positive, but exhibits still very large fluctuations during one cycle, 
+where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
+throughout positive, but exhibits very large fluctuations during one cycle, 
 which are absent in regime BPI-2 at higher shear rates.}
@@
 Finally, the network loses any residual order and 
-transforms into an amorphous state with virtually constant
+transforms into an amorphous state with almost constant
 yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
@@
 
-Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) at 
+Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) and 
 at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
-and Ericksen numbers $0.5\le{\cal E}<10$ 
+and Ericksen numbers $0.5\le{\it Er}<10$ 
 lies another region where the network flows with periodically 
-recurring flow-induced conformations (viz. Fig. \ref{bp1-rheo}). 
+recurring conformations (Fig. \ref{bp1-rheo}). 
 We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
-is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress over total strain. 
-Obviously a qualitative difference between these two is the absence of 
+is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress versus total strain. 
+A qualitative difference between these two is the absence of 
 large stress fluctuations that occur during one typical cycle in regime BPI-1.
 Remember that for even lower $\gd$ the deviatoric stress becomes 
-temporarily negative (viz. Fig. \ref{bp1-fe-yield}), which 
+temporarily negative (Fig. \ref{bp1-fe-yield}), which 
 caused destabilisation of the periodic network and led to the 
@@
 for this phenomenon could be that in BPI-2 the network finds a 
-kinetic path to bypass these large amplitude fluctuations and 
+kinetic path to bypass these large amplitude fluctuations that 
 prevents the network from destabilising.
@@
 The recurrence period of conformations with visually similar appearance is 
-less straightforward to determine. In flow direction a multiple of $1/\gd$,
-i.e. $\tau_F=n/\gd, n\in\cal{N}$ generally holds, e.g. $n=1,2,4$ for $\gd=3.91\e{-5},
-7.81\e{-5},1.563\e{-4}$. The analysis is complicated by the fact
+less straightforward to determine. In the flow direction relation like
+$\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
+7.81\e{-5},1.563\e{-4}$ for example. The analysis is complicated by the fact
 that for $n>1$ thermodynamically equivalent, mirror-image 
@@
 
-However, there is no such thing as a definite recurrence period $\tau_V$ 
-in vorticity direction. The direction and magnitude of the displacement
+However, there is no definite recurrence period $\tau_V$ 
+in the vorticity direction. The direction of the displacement
 in other directions (i.e. both in gradient and vorticity direction) 
-is random, non-integer and depends sensitively on the imposed shear rate,
-suggesting BPI-2 constitutes an instance of rheochaos.
+is not simply along coordinate directions. 
+The magnitude of the displacement show no clear pattern, is non-integer 
+and depends sensitively on the imposed shear rate,
+suggesting BPI-2 constitutes an instance of 'rheochaos' [ref].
 
@@
 helix. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in the aspect that there is no
-arrangement in bands. However, just as in case of BPII both patterns
+Fig. \ref{bp2-velo}, but differs in that there is no
+arrangement in bands. However, just as in case of BPII, both patterns
 are mirror images and feature the highest velocities in the regions where 
@@
 
-The intricacy and periodically recurrence of the flow-induced conformations 
-in BPI raises the question how robust these results are with respect to 
-system size. Although a pitch length of 64 LBU provides enough resolution 
+The intricacy and periodic recurrence of the flow-induced conformations 
+in BPI raises the question: how robust these results are with respect to 
+system size? Although a pitch length of 64 LBU provides enough resolution 
 for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
 in total are actually sufficient to avoid finite size effects.
-Unfortunately the computational costs of simulations, given that large strains have to
+Unfortunately the computational costs of larger simulations are still
+prohibitive, given that large strains have to
 be simulated in order to reach a steady state. Therefore we opted for a 
@@
 network as well as in the time-dependence of the deviatoric stress.
-It takes BPI also considerably longer in the smaller run to reach the 
+In the smaller run with $2^3=8$ unit cells takes BPI also considerably 
+to reach the 
 steady state with its flow-induced recurring disclination network.
@@
 \caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
-The dark/red network in top picture shows a snapshot of run that consists 
-of $2^3=8$ unit cells, whereas the light/cyan network displays the 
+The dark (red) network in top picture shows a snapshot of run that consists 
+of $2^3=8$ unit cells, whereas the light (cyan) network displays the 
 central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
@@
 %\includegraphics[width=0.4\textwidth]{dir+z-300k_run916.png}
-\includegraphics[width=0.4\textwidth]{dir+z-301k_run916.png}
-\includegraphics[width=0.4\textwidth]{dir+z-302k_run916.png}
+%\includegraphics[width=0.4\textwidth]{dir+z-301k_run916.png}
+\includegraphics[width=0.4\textwidth]{dir3d-z-302k_run916.png}
 \caption{Director field of BPI at high shear rate: The pictures show a slice 
-of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction 
+of $L_x\times L_y \times L_Z= 32\times32\times 32$ lattice sites along the vorticity direction 
 with the flow/gradient direction oriented along the horizontal/vertical axis. 
@@
 two-dimensional, translationally invariant state, which is shown in 
-Fig. \ref{bp1-high}. The snapshots visualise nicely the affine 
-transformation of simple shear flow in gradient-flow plane.
+Fig. \ref{bp1-high}. The snapshots visualise the affine 
+transformation of simple shear flow in the gradient-flow plane.
 There are distinctive regions where the director field is 
@@
  $3.125\e{-4}\le\gd_c\le 6.25\e{-4}$ and occurs at similar 
-Ericksen numbers ${\cal E}\simeq 10$.
+Ericksen numbers ${\it Er}\simeq 10$.
+
+\subsection{Transition to 1D Uniaxial Twist}
+
+\subsection{Uncoiling Transition to Nematic State}
 
@@
 In summary, our work constitutes the first simulation of bulk flow behaviour 
-of cubic BPs in simple shear flow. Although experimental evidence of our results 
-is currently not available, we believe this work can shed some light into the 
+of cubic BPs in simple shear flow. Although experimental evidence to support
+our results is currently not available, we believe this work can shed some light into the 
 flow properties of complex liquid-crystalline phases.
@@
 
-At slightly larger Ericksen numbers $0.1<{\cal E}<10$ shear thinning is 
+At slightly larger Ericksen numbers $0.1<{\it Er}<10$ shear thinning is 
 slightly stronger than for BPII, but BPI does not show simple power-law 

### 2012-11-06 branches/CUDA/BookGPU/Chapters/chapter1/ch1.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1807 40876441-3496-0410-a613-a2fca35b2ae2

\chapterauthor{Raphaël Couturier}{Femto-ST Institute, University of Franche-Comte}

\chapter{Presentation of the GPU architecture and of the Cuda environment}
\label{chapter1}

\section{Introduction}\label{ch1:intro}

This chapter introduces the Graphics  Processing Unit (GPU) architecture and all
the concepts needed to understand how GPUs  work and can be used to speed up the
execution of some algorithms. First of all this chapter gives a brief history of
the development  of Graphics  card until they  have been  used in order  to make
general   purpose   computation.    Then   the   architecture  of   a   GPU   is
illustrated.  There  are  many  fundamental  differences between  a  GPU  and  a
tradition  processor. In  order  to benefit  from the  power  of a  GPU, a  Cuda
programmer needs to use threads. They have some particularities which enable the
Cuda model to be efficient and scalable when some constraints are addressed.

\section{Brief history of Video Card}

Video  cards or Graphics  cards have  been introduced  in personal  computers to
produce  high quality graphics  faster than  classical Central  Processing Units
(CPU) and  to alleviate CPU from this  task. In general, display  tasks are very
repetitive and very specific.  Hence,  some manufacturers have produced more and
more sophisticated video cards, providing 2D accelerations then 3D accelerations,
then some  light transforms. Video cards  own their own memory  to perform their
computation.  For at least two decades, every personal computer has had a video
card which is simple for  desktop computers or which provides many accelerations
for game and/or  graphic oriented computers.  In the  latter case, graphic cards
may be more expensive than a CPU.

Since  2000, video  cards have  allowed  users to  apply arithmetic  operations
simultaneously on a sequence of  pixels, also later called stream processing. In
this case, the information of the pixels (color, location and other information) are
combined in order  to produce a pixel  color that can be displayed  on a screen.
Simultaneous  computations are  provided  by shaders  which calculate  rendering
effects on  graphics hardware with a  high degree of  flexibility. These shaders
handles the stream data with pipelines.

Some researchers  tried to  apply those operations  on other  data, representing
something different  from pixels,  and consequently this  resulted in  the first
uses of video cards for  performing general purpose computation. The programming
model  was not  easy  to use  at  all and  was very  dependent  of the  hardware
constraints.   More precisely  it consisted  in using  either DirectX  of OpenGL
functions  providing  an  interface  to  some classical  operations  for  videos
operations  (memory  transfers,  texture  manipulation,  ...).   Floating  point
operations were  most of the  time unimaginable.  Obviously when  something went
wrong, programmers had no way (and neither the tools) to detect it.

\section{GPGPU}

In order  to benefit from the computing  power of more recent  video cards, Cuda
was first proposed in 2007 by  NVidia. It unifies the programming model for some
of  their most performant  video cards.   Cuda~\cite{ch1:cuda} has  quickly been
considered by  the scientific community as  a great advance  for general purpose
graphics processing unit (GPGPU)  computing.  Of course other programming models
have been  proposed. The  other well-known alternative  is OpenCL which  aims at
proposing an alternative to Cuda  and which is multi-platform and portable. This
is a  great advantage since  it is even  possible to execute OpenCL  programs on
traditional CPUs.  The main drawback is that it is less tight with the hardware
and  consequently sometimes  provides  less efficient  programs. Moreover,  Cuda
benefits from  more mature compilation and optimization  procedures.  Other less
known environments  have been proposed, but  most of them have  been stopped, for
example  we can  cite: FireStream  by ATI  which is  not maintained  anymore and
replaced by  OpenCL, BrookGPU by  Standford University~\cite{ch1:Buck:2004:BGS}.
Another environment based  on pragma (insertion of pragma  directives inside the
code to  help the compiler to generate  efficient code) is call  OpenACC.  For a
comparison with OpenCL, interested readers may refer to~\cite{ch1:CMR:12}.

\section{Architecture of current GPUs}

The architecture  \index{architecture of  a GPU} of  current GPUs  is constantly
evolving.  Nevertheless  some trends remain constant  throughout this evolution.
Processing units composing a GPU are  far more simple than a traditional CPU but
it is much easier to integrate many computing units inside a GPU card than to do
so with many cores inside a CPU. This is due to the fact that the cores of a GPU are
simpler than the cores of a CPU.  In  2012, the most powerful GPUs own more than 500
cores       and       the       most       powerful      CPUs       have       8
cores. Figure~\ref{ch1:fig:comparison_cpu_gpu} shows  the number of cores inside
a  CPU  and  inside a  GPU.   In  fact,  in  a  current NVidia  GPU,  there  are
multiprocessors which have 32 cores (for example on Fermi cards). The core clock
of CPU is  generally around 3GHz and  the one of GPU is  about 1.5GHz.  Although
the core clock of GPU cores is slower, the amount of cores inside a GPU provides
more computational power.  This measure is commonly represented by the number of
floating point operation  per seconds. Nowadays the most powerful  GPUs provide more
than   1TFlops,  i.e.    $10^{12}$   floating  point   operations  per   second.
Nevertheless  GPUs are very  efficient to  perform some  operations but  not all
kinds of operations. They are very efficient to execute repetitive work in which
only  the data  change. It  is important  to keep  in mind  that multiprocessors
inside a GPU have 32 cores. Later we will see that these 32 cores need to do the
same work to get maximum performance.

\begin{figure}[b!]
\centerline{\includegraphics[]{Chapters/chapter1/figures/nb_cores_CPU_GPU.pdf}}
\caption{Comparison of number of cores in a CPU and in a GPU.}
%[Comparison of number of cores in a CPU and in a GPU]
\label{ch1:fig:comparison_cpu_gpu}
\end{figure}

On the most powerful  GPU cards, called Fermi, multiprocessors  are called streaming
multiprocessors  (SM). Each  SM contains  32  cores and  is able  to perform  32
floating points or integer operations on  32 bits numbers per clock or 16 floating
points  on  64 bits number  per  clock. SMs  have  their  own registers,  execution
pipelines and caches.  On Fermi architecture,  there are 64Kb shared memory + L1
cache  and 32,536 32bits  registers per  SM. More  precisely the  programmer can
decide what amount  of shared memory and  L1 cache SM can use.  The constraint is
that the sum of both amounts should be less or equal to 64Kb.

Threads are used to  benefit from the important number of cores  of a GPU. Those
threads    are   different    from    traditional   threads    for   CPU.     In
chapter~\ref{chapter2},  some  examples of  GPU  programming  will explicit  the
details of  the GPU  threads. However,  threads are gathered  into blocks  of 32
threads, called ``warps''. Those warps  are important when designing an algorithm
for GPU.

Another big  difference between CPU and GPU  is the latency of  memory.  In CPU,
everything is optimized  to obtain a low latency  architecture. This is possible
through  the  use  of  cache  memories. Moreover,  nowadays  CPUs  perform  many
performance optimizations  such as speculative execution  which roughly speaking
consists in executing  a small part of  code in advance even if  later this work
reveals itself  to be  useless. On the  contrary, GPUs  do not have  low latency
memory.   In comparison GPUs  have small  cache memories.  Nevertheless the
architecture of GPUs is optimized  for throughput computation and it takes into
account the memory latency.

\begin{figure}[b!]
\centerline{\includegraphics[scale=0.7]{Chapters/chapter1/figures/low_latency_vs_high_throughput.pdf}}
\caption{Comparison of low latency of CPU and high throughput of GPU.}
\label{ch1:fig:latency_throughput}
\end{figure}

Figure~\ref{ch1:fig:latency_throughput}  illustrates   the  main  difference  of
memory latency between a CPU and a  GPU. In a CPU, tasks ``ti'' are executed one
by one with a short memory latency to get the data to process. After some tasks,
there is  a context switch  that allows the  CPU to run  concurrent applications
and/or multi-threaded  applications. Memory latencies  are longer in a  GPU, the
the  principle  to   obtain  a  high  throughput  is  to   have  many  tasks  to
compute. Later we  will see that those tasks are called  threads with Cuda. With
this  principle, as soon  as a  task is  finished the  next one  is ready  to be
executed  while the  wait for  data for  the previous  task is  overlapped by
computation of other tasks.

\section{Kinds of parallelism}

Many  kinds  of parallelism  are  amiable according  to  the  type of  hardware.
Roughly  speaking, there  are three  classes of  parallelism: instruction-level
parallelism, data parallelism and task parallelism.

Instruction-level parallelism consists in re-ordering some instructions in order
to execute  some of them in parallel  without changing the result  of the code.
In  modern CPUs, instruction  pipelines allow  processor to  execute instructions
faster.   With   a  pipeline  a  processor  can   execute  multiple  instructions
simultaneously due  to the fact that  the output of a  task is the  input of the
next one.

Data parallelism consists  in executing the same program  with different data on
different computing  units.  Of course,  no dependency should exist  between the
data. For example, it is easy  to parallelize loops without dependency using the
data parallelism paradigm. This paradigm  is linked with the Single Instructions
Multiple Data (SIMD)  architecture. This is the kind  of parallelism provided by
GPUs.

Task parallelism is the common parallelism  achieved out on clusters and grids and
high performance  architectures where different tasks are  executed by different
computing units.

\section{Cuda Multithreading}

The data parallelism  of Cuda is more precisely based  on the Single Instruction
Multiple Thread (SIMT) model. This is due to the fact that a programmer accesses
to  the cores  by the  intermediate of  threads. In  the Cuda  model,  all cores
execute the  same set of  instructions but with  different data. This  model has
similarities with the vector programming  model proposed for vector machines through
the  1970s into  the  90s, notably  the  various Cray  platforms.   On the  Cuda
architecture, the  performance is  led by the  use of  a huge number  of threads
(from thousands up to  to millions). The particularity of the  model is that there
is no  context switching as in  CPUs and each  thread has its own  registers. In
practice,  threads  are executed  by  SM  and are  gathered  into  groups of  32
threads.  Those  groups  are  called  ``warps''. Each  SM  alternatively  executes
``active warps''  and warps becoming temporarily  inactive due to  waiting of data
(as shown in Figure~\ref{ch1:fig:latency_throughput}).

The key to scalability in the Cuda model is the use of a huge number of threads.
In practice, threads are not only gathered  in warps but also in thread blocks. A
thread block is executed  by only one SM and it cannot  migrate. The typical size of
a thread block is a number power of two (for example: 64, 128, 256 or 512).

In this  case, without changing anything inside  a Cuda code, it  is possible to
run your  code with  a small Cuda  device or  the most performing Tesla  Cuda cards.
Blocks are  executed in any order depending  on the number of  SMs available. So
the  programmer  must  conceive  its  code  having this  issue  in  mind.   This
independence between thread blocks provides the scalability of Cuda codes.

\begin{figure}[b!]
\centerline{\includegraphics[scale=0.65]{Chapters/chapter1/figures/scalability.pdf}}
\caption{Scalability of GPU.}
\label{ch1:fig:scalability}
\end{figure}

A kernel is a function which  contains a block of instructions that are executed
by the  threads of a GPU.   When the problem considered  is a two  dimensional or three
dimensional  problem,  it is  possible  to group  thread  blocks  into a grid.   In
practice, the number of  thread blocks and the size of thread  blocks is given as
parameters  to  each  kernel.   Figure~\ref{ch1:fig:scalability}  illustrates  an
example of a kernel composed of 8 thread blocks. Then this kernel is executed on
a small device containing only 2 SMs.  So in  this case, blocks are executed 2
by 2 in any order.  If the kernel is executed on a larger Cuda device containing
4 SMs, blocks are executed 4 by 4 simultaneously.  The execution times should be
approximately twice faster in the latter  case. Of course, that depends on other
parameters that will be described later.

Thread blocks provide a way to cooperation  in the sense that threads of the same
block   cooperatively    load   and   store   blocks   of    memory   they   all
use. Synchronizations of threads in the same block are possible (but not between
threads of different  blocks). Threads of the same block  can also share results
in order  to compute a  single result. In chapter~\ref{chapter2},  some examples
will explicit that.

\section{Memory hierarchy}

The memory hierarchy of  GPUs\index{memory~hierarchy} is different from the CPUs
one.  In practice,  there are registers\index{memory~hierarchy!registers}, local
memory\index{memory~hierarchy!local~memory},                               shared
memory\index{memory~hierarchy!shared~memory},                               cache
memory\index{memory~hierarchy!cache~memory}              and              global
memory\index{memory~hierarchy!global~memory}.

As  previously  mentioned each  thread  can access  its  own  registers.  It  is
important to keep in mind that the  number of registers per block is limited. On
recent cards,  this number is  limited to 64Kb  per SM.  Access to  registers is
very fast, so it is a good idea to use them whenever possible.

Likewise each thread can access local  memory which, in practice, is much slower
than registers.  Local memory is automatically used by the compiler when all the
registers are  occupied. So the  best idea is  to optimize the use  of registers
even if this implies to reduce the number of threads per block.

\begin{figure}[hbtp!]
\centerline{\includegraphics[scale=0.60]{Chapters/chapter1/figures/memory_hierarchy.pdf}}
\caption{Memory hierarchy of a GPU.}
\label{ch1:fig:memory_hierarchy}
\end{figure}

Shared memory allows  cooperation between threads of the  same block.  This kind
of memory is fast because it requires to be manipulated manually and its size is
limited.  It is accessible during the execution of a kernel. So the principle is
to fill the shared  memory at the start of the kernel  with global data that are
used very  frequently, then threads can  access it for  their computation.  They
can obviously change  the content of this shared  memory either with computation
or load of  other data and they can  store its content in the  global memory. So
shared memory can  be seen as a cache memory  manageable manually. This requires
obviously an effort from the programmer.

On  recent cards,  the programmer  may decide  what amount  of cache  memory and
shared memory is attributed to a kernel. The cache memory is a L1 cache which is
directly  managed by  the GPU.  Sometimes,  this cache  provides very  efficient
result and sometimes the use of shared memory is a better solution.

Figure~\ref{ch1:fig:memory_hierarchy}  illustrates  the  memory hierarchy  of  a
GPU. Threads are represented on the top  of the figure. They can access to their
own registers  and their local memory. Threads  of the same block  can access to
the shared memory of this block. The cache memory is not represented here but it
is local  to a thread. Then  each block can access  to the global  memory of the
GPU.

 \section{Conclusion}

In this chapter,  a brief presentation of the video card,  which has later been
used to perform computation, has been  given. The architecture of a GPU has been
illustrated focusing on the particularity of GPUs in term of parallelism, memory
latency and  threads. In order to design  an efficient algorithm for  GPU, it is
essential to have all these parameters in mind.

\putbib[Chapters/chapter1/biblio]

### 2012-11-06 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1807 40876441-3496-0410-a613-a2fca35b2ae2

\chapterauthor{Alan Gray and Kevin Stratford}{EPCC, The University of Edinburgh}

\chapter{Ludwig}

%\putbib[biblio]

\section{Introduction}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface MPI \cite{mpi-standard}. However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address verly large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers as many supercomputing services are moving
toward clusters of GPU accelerated nodes.

The \textit{Ludwig} code \cite{desplat} is a LB package developed
specifically for complex fluid problems, for example mixtures,
surfactants, liquid crystals, and particle
suspensions~\cite{aidun2010}.  In this chapter we present the steps
required to allow Ludwig to efficiently exploit many NVIDIA GPUs in
parallel.  We show that Ludwig scales excellently to at least the
thousand GPU level (the largest resource available at the time of
writing) with indications that the code will scale to much larger
systems as they come on-line. We first consider the problem of a
binary fluid mixture (e.g., oil and water) modelled using the LB
approach, and we go on to describe how we deal with the complexities
arising when colloidal particles are introduced to interact with the
fluid.

In Section \ref{ch14:sec:background} we give background details to
introduce the the methods being employed. In Section
\ref{ch14:sec:singlegpu}, we describe the adaptations made to Ludwig
to allow it to efficiently exploit the GPU architecture. In Section
\ref{ch14:sec:singlegpu}, we extend this description to the work
required to allow exploitation of many GPUs in parallel. We present
performance and scaling results, for the binary fluid case, in Section
\ref{ch14:sec:performance}, before going, in Section
\ref{ch14:sec:particles}, to describe how we minimise the overheads
involved with introducing colloidal particles.

\section{Background}\label{ch14:sec:background}

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/basiclattice}
\caption{The lattice Boltzmann approach discritises space into a 3D lattice (left). The fluid is represented at each lattice site (right), by the {\it distribution} vector: each component corresponding to a specific velocity direction. Here we are interested in the {\it D3Q19} model, in which there are 3 spatial directions and 19 velocity components (including a zero component).}
\label{ch14:fig:basiclattice}
\end{figure}

A symmetric binary fluid, in which both components have the same
density and viscosity, may be described by introducing a composition
variable or \textit{order parameter} $\phi(\mathbf{r})$. This is
a function of position $\mathbf{r}$ and describes the relative
proportions of the two fluids present locally. To describe the
thermodynamics of the system it is possible to write down a
\textit{free energy} which is a functional of the order parameter.
The equation of motion for $\phi$ is then
\begin{equation}
\partial_t \phi + \mathbf{\nabla}.(\mathbf{u}\phi) = 
- \mathbf{\nabla} . (M \mathbf{\nabla}\mu),
\label{cahn-hilliard}
\end{equation}
where $\mu$ is the chemical potential related to the functional derivative
of the free energy, and $M$ is a mobility. This couples
to the fluid velocity field $\mathbf{u}(\mathbf{r})$, whose evolution
obeys the Navier-Stokes equations describing the conservation of mass
(or density $\rho$) and momentum
\begin{equation}
\rho  [ \partial_t \mathbf{u} + (\mathbf{u}.\mathbf{\nabla})\mathbf{u} ]
= -\nabla p + \eta \nabla^2 \mathbf{u} + \mathbf{f}(\mathbf{r}),
\end{equation}
where $p$ is the isotropic pressure and $\eta$ is the viscosity.
A force density
$\mathbf{f}(\mathbf{r})$  describes the force exerted by a
curved fluid interface on the fluid. It is sufficient to note
here that this force depends locally on the order parameter and its
derivatives $\nabla\phi$ and $\nabla^2 \phi$. For a full
description the interested reader is referred to, e.g.,
\cite{bray1994,kendon2001}.

The LB approach makes use of a regular 3-dimensional
lattice (see Figure \ref{ch14:fig:basiclattice}) with discrete spacing $\Delta r$. It also makes use of a
discrete velocity space $\mathbf{c}_i$, where the $\mathbf{c}_i$
are chosen to capture the correct symmetries of the Navier-Stokes
equations. A typical choice, used here, is the so-called D3Q19
basis in three dimensions where there is one velocity such that
$\mathbf{c} \Delta t$ is zero, along with six extending to the nearest
neighbour
lattice sites, and twelve extending to the next-nearest neighbour sites
($\Delta t$ being the discrete time step). The fundamental object
in LB is then the distribution function $f_i (\mathbf{r};t)$ whose
moments are related to the local hydrodynamic quantities: the fluid
density, momentum, and stress. The time evolution of the distribution
function is described by a discrete Boltzmann equation
\begin{equation}
f_i(\mathbf{r} + \mathbf{c}_i \Delta t; t) - f_i(\mathbf{r}; t) 
= - {\cal L}_{ij} f_j(\mathbf{r};t).
\end{equation}
It is convenient to think of this in two stages. First, the right hand
side represents the action of a collision operator ${\cal L}_{ij}$,
which is local to each lattice site and relaxes the distribution toward
a local equilibrium at a rate ultimately related to the fluid viscosity.
Second, the left hand side represents a propagation step (sometimes referred
to as streaming step), in which each element $i$ of the distribution is
displaced $\mathbf{c}_i \Delta t$, i.e., one lattice spacing in the
appropriate direction per discrete time step.

Computationally, we store a vector of 19 double precision floating
point values at each lattice site for the distribution function $f_i$.
The collision operation, which is local at each lattice site, may be
thought of as follows. The matrix-vector multiplication $L_{ji}f_i$ is
used to transform the distributions into the hydrodynamic quantities,
where $L_{ij}$ is a constant 19x19 matrix related to the choice of
$\mathbf{c}_i$. The non-conserved hydrodynamic quantities are then
relaxed toward their (known) equilibrium values, and are transformed
back to new post-collision distributions via the inverse transformation
$L^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
floating point multiplications per lattice site. In contrast, the
propagation stage consists solely of moving the distribution values
between lattice sites, and involves no floating point operations.

To represent the composition variable, we introduce a second distribution
function $g_i (\mathbf{r};t)$ whose first moment is $\phi$, second moment
is the order parameter flux, and so on \cite{swift1996}.
The time evolution of this second distribution follows an analogous
collision and propagation process to that described for $f_i$, and
so provides a solution to Eq.~\ref{cahn-hilliard} \cite{stratford-jsp2005}.

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption{Left: the lattice is decomposed between MPI tasks. For clarity we show a 2D decomposition of a 3D lattice, but in practice we decompose in all 3 dimensions. Halo cells are added to each sudomain, as shown on the right for a single slice, which store data retrieved from remote neighbours in the halo exchange.}
\label{ch14:fig:decomphalo}
\end{figure}

As illustrated in Figure \ref{ch14:fig:decomphalo}, Ludwig is
parallelized (across CPU cores) using a regular 3D domain decomposition
where communications are performed via message passing using MPI.  We
note that while the collision is local for $f_i$, computation of the
force terms requires the evaluation of the derivatives of $\phi$ via a
standard finite different stencil.  This gives rise to the need for a
halo exchange involving the scalar $\phi$ before the collision.
Before the propagation stage, a halo exchange in both $f_i$ and $g_i$
is required.

For a 3D halo exchange, 2D planes of data are exchanged in each of the
positive and negative directions, for each of the 3 dimensions, i.e. 6
planes are sent and received by each process.  For the CPU version of
Ludwig, MPI datatype functionality is used to designate these planes.
The $\phi$ halo exchange involves scalar values at each site, whereas
the distribution halo exchange involves a higher volume of data since
we must transfer multiple components of each velocity vector. However
the number of velocity components included in a given transfer can be
filtered from 19 to 5 (for D3Q19), since only those elements
propagating outward from a local domain play a role in the propagation
stage. The specific components required depend on the direction of
transfer, and again MPI datatype functionality is used to designate
this.

\section{GPU Implementation}\label{ch14:sec:singlegpu}

In this section we describe the methods used to enable Ludwig
for the GPU architecture and describe some key optimizations that
significantly enhance performance.

To ensure that \textit{Ludwig} remains suitable for CPU-based systems,
we augment rather than modify the original code. All new GPU-related
functionality are implemented in an additive manner, mirroring the
modular structure of the original code. The new GPU modules have
interfaces similar to the original CPU functions, and the ability to
switch between versions is achieved at compile-time via C
pre-processing directives.  All GPU compute kernels are implemented using
CUDA.  Wrapper routines are developed to specify the
decomposition, invoke the CUDA kernels and perform the necessary data
management operations: the latter are facilitated through newly
developed modules for creation and destruction of device memory
structures, the copying of data between host and device, and where
necessary the packing and buffering of data.

The most important code section, in terms of performance, is the
collision stage. But it is important to offload all other
computational components of the timestep. Kernels with relatively low
computational demand, such as the propagation stage and operations
involving the order parameter, are also GPU enabled: this allows data
to remain resident on the GPU, and avoids expensive host-device data
transfer at each timestep (which would reduce the benefit of GPU
acceleration).  In order to port to the GPU in an incremental fashion,
while retaining the ability to test for correctness throughout the
process, we require a suite of data management functions. For each
data structure (such as the distribution, which we will use as an
example), we maintain a copy in both the CPU and GPU memory spaces.
Functions such as \verb+put_dist_on_gpu()+, and
\verb+get_dist_from_gpu()+, which abstract underlying CUDA
functionality, allow us to update the GPU copy from the host copy and
vice versa\footnote {Within the implementation of these functions we
  have to work around data encapsulation issues. In the code, the use
  of the file scope identifier \texttt{static} prevents direct access
  to data; instead access functions are provided.  For the GPU code,
  it was necessary to use these existing access functions to copy
  these data to equivalent arrays on the device (via temporary buffers
  on the host).  Within the CUDA kernel, the data access function
  calls are substituted with array accesses.  The process is reversed
  when encapsulated data needs to be transferred back to the host.}.
This makes it easy, in the function implementing the Ludwig timestep,
to switch between the CPU and GPU for different components in the
timestep: we can develop GPU functionality incrementally while
retaining a code that runs correctly (albeit slowly due to data
transfer overheads).  Once every component in the timestep is
GPU-enabled, then it becomes possible to move such data transfers
outside the timestep loop, and just use the GPU-resident versions
throughout the simulation.

To achieve optimal performance, it is vital to fully exploit
parallelism inherent to the NVIDIA architecture, particularly for
those matrix-vector operations within the dominant collision kernel. A
simplified schematic of a serial implementation of such a matrix-vector
operations is as follows:
{\footnotesize
\begin{verbatim}
// loop over lattice sites
for ( is = 0; is < N; is++){
  ...
  //load distribution from site-major structure
  for (i = 0; i < 19; i++)    
    f[i]=f_all[is][i]
  ...
  //perform matrix-vector multiplication  
  for (i = 0; i < 19; i++){    
    a[i] = 0.0;    
    for (j = 0; j < 19; j++){      
      a[i] += f[j]*L[i][j];   
    }
  }
  ...
}
\end{verbatim}
}
The GPU architecture features a hierarchy of parallelism. At the
lowest level, groups of 32 threads called {\it warps} operate in
lock-step on different data elements - this is SIMD style vector-level
parallelism. Multiple warps are combined into a thread block (in which
communication and synchronisation are possible), and multiple blocks can
run concurrently across the SMs in the GPU (with no communication or
synchronisation possible across blocks).  To decompose on the GPU, we
must choose which of the above loops to assign to which level of
parallelism.

There exists parallelism within the matrix-vector operation, but the
length of each {\it distribution} vector, at 19, is smaller than the
warp size and typical thread block sizes. So therefore we simply
decompose the loop over lattice sites to all levels of parallelism,
i.e.  we use a separate CUDA thread for every lattice site, and each
thread performs the full matrix-vector operation. We find that a block
size of 256 performs best: we decompose our lattice domain into groups
of 256 sites, and assign each group to a block of CUDA threads.

Since the matrix \verb+L+ is constant across CUDA threads (sites), it
can reside in the fast {\it constant} on-chip device memory. For each
\verb+is+, \verb+L+ is multiplied by the distribution function at that
site; the distributions for the complete lattice are stored in
\verb+f_all+ residing in off-chip device global memory.

Two distinct optimizations were found to improve the performance of this code section, and hence the application as a whole, on
the GPU.  First, the original code uses the output vector \verb+a+ to
hold each temporary summation in the inner loop.  The code was adapted
so that a temporary scalar variable was used instead. This allows
caching of intermediate values on-chip and thus avoids expensive
global memory accesses within the inner loop.  Secondly, it can be
seen the original data layout of \verb+f_all+ is site-major, i.e. for
each site, consecutive velocity components are contiguous in memory.
This layout is optimal when executing on the CPU, but is sub-optimal
on the GPU when combined with our CUDA decomposition. An architectural
constraint of NVIDIA GPUs means that optimal global memory bandwidth
is only achieved when data is structured such that threads within a
{\it half-warp} (a group of 16 threads) load data from the same memory
segment in a single transaction: this is referred to as {\it memory
  coalescing}.  Otherwise, memory accesses are serialized,
significantly degrading performance. The coalescing criteria is met
when consecutive threads read consecutive memory addresses. Therefore
the alternative transposed velocity-major layout of \verb+f_all+ is
more optimal, i.e.  for each velocity component, consecutive site
values are contiguous in memory.  \textit{Ludwig} was modified to
allow a choice of distribution data layout at compilation time: the
original (most optimal for the CPU) or that permitting coalesced
accesses on the GPU.

A schematic of our optimised CUDA code, to be compared with the
listing above, is
{\footnotesize
\begin{verbatim}
// calculate is from CUDA thread and block indexes
...
//load distribution from velocity-major structure
for (i = 0; i < 19; i++)    
   f[i]=f_all[i][is]
...
//perform matrix-vector multiplication  
for (i = 0; i < 19; i++){    
 double a_tmp = 0.0;    
 for (j = 0; j < 19; j++){      
   a_tmp += f[j]*L[i][j];   
 }
 a[i] = a_tmp;    
}
...
\end{verbatim}
}

\section{Multi-GPU Implementation}\label{ch14:sec:parallelgpu}

To execute on multiple GPUs in parallel we decompose the problem using
the existing high-level parallelisaton framework: each MPI task acts
as a host to a separate GPU, with each GPU kernel operating on the
subset of the lattice volume corresponding to the partition assigned
to that task.  As described in Section \ref{ch14:sec:background}, halo
data exchanges are required at each timestep.  Significant
developments are made for the GPU adaptation of this communication
phase to achieve good inter-GPU communication performance and allow
scaling to many GPUs.

The
traditional code uses MPI datatype functionality to designate the six
planes of lattice data to be sent/received, and performs the
communications {\it in-place}, i.e.  there is no explicit buffering of
data in the application.  For the GPU adaptation, the halo planes must
be exchanged between GPUs, on which MPI functionality is not
available, so transfers must be staged through the host CPUs. Buffer
packing/unpacking of the halo planes is explicitly performed using
CUDA kernels, and CUDA memory copy calls are used to transfer data
from a device to its host (using pinned host memory) through the PCI-e
bus (and vice versa). MPI calls are used to transfer the buffers
between hosts. The CPU version of Ludwig has a mechanism, using MPI built-in
data types, to reduce the amount of data communicated by around a
factor of four by only sending those elements of the distribution
propagating outward from a local domain. For the GPU adaptation, this
filtering of data is explicitly performed (again since MPI
functionality is not available on the GPU) in the buffer
packing/unpacking kernels.  Special care had to be taken for those
lattice sites on the corners of the halo planes, which had to be sent
in more than one direction (since the propagation includes accesses to
diagonal neighbours).

The fact that we are transferring multiple buffers allows us to
overlap the communication stages, which each stress different parts of
the hardware, and hence reduce overall communication overhead.  For
example, after the planes corresponding to the first dimension have
been retrieved by the host, these can be exchanged using MPI between
hosts at the same time as the packing and retrieving of the second
dimension planes. The operations are not completely independent across
dimensions because of updates to corner data (see above). We use a
separate CUDA stream for each dimension and overlap to the extent
allowed. When overlapping is invoked, we observe that some of the
host-device communication time can be effectively ``hidden'' behind
the host-host MPI communication time, resulting in an overall speedup.
The effect is more pronounced for the smaller local lattice size,
perhaps due to less CPU memory bandwidth contention. Hence the
optimization is particularly valuable to aid strong scaling, where the
local system size reduces as the number of GPUs increases.

\section{Performance}\label{ch14:sec:performance}

We compare the performance of the CPU and GPU versions of Ludwig using
state of the art Cray supercomputing facilities. The Cray XE6
architecture \cite{xe6} features 2 AMD Opteron CPUs per node, with
nodes interconnected using Cray Gemini technology. For our baseline
CPU results we use the HECToR UK academic supercomputing facility
\cite{hector}. The Cray XK6 \cite{xk6} is very similar to the Cray
XE6, but with one CPU per node replaced with an NVIDIA X2090 GPU. Each
node in the system therefore features a single Opteron CPU acting as a
host to a single GPU. The inter-node interconnect architecture is the
same as for the Cray XE6: all communications are performed via the
host CPUs.  For our GPU performance tests we use a prototype Cray XK6
system, internal to Cray, with 936 nodes (GPUs).

\begin{figure}[!t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/length_scale}
\caption{The dependence of the characteristic length scale of a binary
  fluid mixture on the simulation timestep number. Superimposed are
  three snapshots showing increased mixture separation with time:
  these are obtained through the order parameter and each show one
  eighth of the total simulation size, which is $1548\times 1764\times 1568$. }\label{ch14:fig:length_scale}
\end{figure}

\begin{figure}[!t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
\caption{The weak (top) and strong (bottom) scaling of Ludwig.  Closed
  shapes denote results using the traditional version run on the Cray
  XE6 (using two 16-core AMD Interlagos CPUs per node), while open
  shapes denote results using the GPU version on the Cray XK6 (using a
  single NVIDIA X2090 GPU per node). Top: the benchmark time is shown
  where the problem size per node is constant. Bottom: performance is
  shown where, for each of the four problem sizes presented, the
  results are scaled by the lattice volume, and all results are
  normalized by the same arbitrary constant.  }
\label{ch14:fig:scaling}
\end{figure}

The benchmark used involves the separation of an immiscible binary
fluid, where the characteristic length scale of fluid domains grows in
time in a well understood way \cite{kendon2001}. Figure
\ref{ch14:fig:length_scale} shows the characteristic length scale as a
function of time (both in LB units) for a system of size $1548\times
1764\times 1568$ run for around 12 hours on the Cray XK6 using 936
GPUs. The insets show a visualization of the fluid domains at
different times.  For the following comparisons we restrict our
benchmark runs to 100 timesteps: this is representative of the full
run.

The top of Figure \ref{ch14:fig:scaling} shows the benchmark time,
where the problem size is increased with the number of nodes. This
shows weak scaling: the time taken by a perfectly scaling code would
remain constant. The figure plots the time against the number of {\it
  nodes}: for the GPU version we utilize the one GPU per Cray XK6 node
whilst for the traditional version we fully utilize two 16-core CPUs
per Cray XE6 node, i.e. one GPU is compared with 32 CPU cores.  We use
a single MPI task per Cray XK6 node to host the GPU, and 32 MPI tasks
to fully utilize the CPUs in the Cray XE6 node.  Note that our CPU
version is highly optimised: in particular it has been tuned for
optimal utilisation of the SIMD vector units in each CPU core.
The new GPU version scales almost perfectly: the scaling advantage
over the CPU version is likely caused by the fact that the GPU version
has a factor of 32 less MPI tasks per node, hence communications
require a larger number of smaller data buffers. The performance
advantage of the GPU version ranges from a factor of around 1.5 to
around 1.8.

We present a strong scaling comparison in the bottom of Figure
\ref{ch14:fig:scaling}, where we measure the performance (inverse
runtime) varying the number of nodes utilized (where we used up to 768
nodes), for several fixed problem sizes (where for each size we scale
results by the lattice volume to allow plotting on the same graph).
It can be seen that, for each of the four problem sizes, the GPU
adaptation on the Cray XK6 scales better (i.e.  the series are closer
to linearity) than the traditional version on the Cray XE6.  It can be
seen that ideal scaling is being approached by the larger problem
sizes.  The higher performance of the GPU version is manifested in the
larger gradients of the Cray XK6 series.

\section{Inclusion of Particles}\label{ch14:sec:particles}

\begin{figure}[t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
\caption{A 2D illustration of colloidal particles (grey circles)
  interacting with the lattice, where each lattice point is
  represented by a cross. Left: each particle spans multiple sites.
  Right: the particle-facing distribution components are moved inside
  the particle with the direction reversed, such that the ``bounce
  back'' will be completed when the fluid propagates.}
\label{ch14:fig:bbl}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=6cm]{Chapters/chapter14/figures/particlemove}
\caption{A 2D illustration of a colloidal particle moving across the
    lattice. The old and new positions are represented by the open and
    filled circles respectively.}
\label{ch14:fig:particlemove}
\end{figure}

It is of interest, in many situations, to include colloidal particles
in the simulation **more from Kevin***. These are by represented by
moving spheres that interact with the fluid through transfer of
momenta: fluid coming into contact with a particle will bounce off it,
giving the particle a kick in the process. As illustrated on the left
of Figure \ref{ch14:fig:bbl} (which shows 2 dimensions only for
clarity - in practice, the particles will be represented by spheres on
a 3D lattice, and will be much larger relative to the lattice
spacing), typically each particle spans multiple lattice sites, and
therefore ``intercepts'' multiple inter-site links. Ludwig stores
particle information using nested linked lists: a list of particles,
each with a list of all the links intercepted by that particle (plus
other quantities). The fluid-particle interaction proceeds using the
so-called {\it bounce back on links} procedure during each timestep. As
illustrated in the right of Figure \ref{ch14:fig:bbl}, for each of the
intercepting links, those velocity components of the distribution that
face inwards to each particle are moved from the site
outside the particle to the site inside the particle, and
reversed in direction. This is done before the propagation stage,
which then propagates these same velocity components back out of the
particles: the velocity bounces back from the particles. The
momenta transfer from fluid to particles is also determined
from the distribution values.

A simplified schematic of the implementation on the CPU is as follows,
noting that multiplicative coefficients are required:
{\footnotesize
\begin{verbatim}
For each particle:
  For each link:
    Calculate coefficient
    Bounce back fluid using coefficient
    Update particle momentum
\end{verbatim}
}
This process is relatively inexpensive on the CPU, since the total
number of particles (and intercepted links) is always relatively
small, but it does require access to distribution data which we want
to keep resident on the GPU throughout the simulation. The overhead of
transferring the full distribution to the CPU and back every timestep
would be too high. But on the other hand, the full codebase required
to manage particles is quite complex and comprehensive: completely
porting it to the GPU, such that the particles effectively also remain
resident on the GPU, would be a large effort and would cause a major
diversion of CPU and GPU codebases. Further complications would arise
when running on multiple GPUs due to the fact that particles can move
between parallel subdomains. Our solution, which minimises overhead,
is to keep the particles resident on CPU, but offload only their interaction with the fluid to the GPU, as we describe below.

On the CPU, the code in the above listing can be restructured into
three distinct stages as (introducing arrays to temporrily store
indices and data):
{\footnotesize
\begin{verbatim}
1) For each particle, for each link:
     Store site and velocity indices associated with link
     Calculate and store coefficient
2) For each particle, for each link:
     Update velocity data using stored values
     Calculate and store particle momenta updates
3) For each particle, for each link:
     Update particle momentum using stored values
\end{verbatim}
}
Stage 2 is only one that that accesses fluid data: this can therefore
be moved to the GPU, while stages 1 and 3 remain on the CPU. Stage 2
therefore becomes, for the GPU implementation
{\footnotesize
\begin{verbatim}
Transfer coefficient and link index arrays to GPU
Run CUDA kernel:
  Update fluid data using coefficient and link index arrays
  Calculate and store particle momenta updates from fluid data
Transfer particle momenta updates array back to CPU
\end{verbatim}
}
Since the total number of links involves is relatively small, these
partial data transfers have minimal overhead.  A complication arises
due to the fact that the particles move through the lattice, albeit
relatively slowly, and therefore the list of links that they intercept
occasionally changes (see Figure \ref{ch14:fig:particlemove}).  When a
particle moves off a lattice site, the fluid on that site must be
reconstructed. Similarly, the fluid on the newly obstructed site must
be removed. For the same reasons given above, we want to avoid moving
all the code dealing with this complication to the GPU, whilst
minimising overheads. Since, on any one timestep, the number of
lattice sites affected is relatively low, we implement functionality
to perform partial data transfers of the distribution. Only those
sites affected are packed into a buffer on the GPU, transferred to the
CPU and unpacked into the CPU version of the distribution. Then, the
existing code can be used to update those affected sites, before the
reverse partial data transfer is performed to update the GPU version
of data.

Our methods for managing particles allow us to exploit the GPU for the
intensive fluid simulation whilst maintaining the complex code
required to manage particles on the CPU, with overheads minimised
through only transferring those small subsets of data corresponding to
the points of interaction.

\section{Summary}

The \textit{Ludwig} LB application, which specializes
in complex fluid problems, has undergone the developments required to
efficiently use large-scale parallel GPU accelerated supercomputers.
The new functionality augments the original code, such that the
package now supports both traditional CPU-based and GPU-accelerated
systems.

Through use of the NVIDIA CUDA programming language the computational
kernels in the Ludwig timestep were offloaded to the GPU, including
not just those responsible for the majority of the compute time but
any that accessed the main data structures, to allow data to be kept
resident on the GPU and avoid expensive data transfers. We described
steps taken to optimise the performance on the GPU architecture through
reducing off-chip memory accesses and including the restructuring of
data layout in memory to ensure the available memory bandwidth was
fully exploited.

A new halo-exchange communication phase for the code was developed to
allow efficient parallel scaling to many GPUs. The traditional version
relies on MPI datatype functionality for CPU to CPU communication; we
have explicitly written buffer packing/unpacking and transfer
operations to for GPU to GPU data exchange, staged through the host
CPUs. This includes the filtering of velocity components to only
transfer those required in a specific direction, hence reducing the
overall volume of data transferred. We used CUDA stream functionality
to overlap separate stages within the communication phase, and reduce
the overall communication time. The effect was measured to be more
profound for smaller local system sizes, i.e. it is especially useful
in aiding strong scaling.

We compared performance, for a binary fluid benchmark, of the new GPU
adaptation using up to 936 GPUs in parallel on prototype Cray XK6 GPU
accelerated hardware, comparing to runs on the counterpart CPU based
Cray XE6 using the (optimized) traditional version. We compare on a
node-by-node basis, i.e. each GPU against 2 CPUs (or 32 cores),
ensuring full utilization of each CPU.  The performance of the GPU
version on the Cray XK6 version is around a factor of 1.5-1.8 better
than the traditional version on the Cray XE6, depending on how many
nodes is used in the comparison. The weak scaling of the GPU version
is excellent up to the 936 GPUs tested. The scaling advantage over the
traditional version is likely due to the fact that we have a factor of
32 less MPI processes per node. The ability to strong scale depends on
the physical system size; as the size increases ideal scaling is
approached. The strong scaling of the GPU version is better than that
for the CPU version, for each of the sizes tested.

We described work to enable colloidal particles in the simulation with
minimal overhead. We implemented these in such a way that we avoid a
major diversion of the CPU and GPU codebases, whilst minimising data
transfers between the CPU and GPU at each timestep. We keep the
majority of the (relatively inexpensive) particle related code on the
CPU, while offloading only those parts responsible for the interaction
with the fluid to the GPU (where the fluid data resides).

Our resulting package is able to exploit the largest of GPU
accelerated architectures to address a wide range of complex problems.

%\section{Introduction}
%% no \IEEEPARstart
%This demo file is intended to serve as a ``starter file''
%for IEEE conference papers produced under \LaTeX\ using
%IEEEtran.cls version 1.7 and later.
%% You must have at least 2 lines in the paragraph with the drop letter
%% (should never be an issue)
%I wish you the best of success.

%\hfill mds

%\hfill January 11, 2007

%\subsection{Subsection Heading Here}
%Subsection text here.

%\subsubsection{Subsubsection Heading Here}
%Subsubsection text here.

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation Results}
%\label{fig_sim}
%\end{figure}

% Note that IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command, the
% \label for the overall figure must come after \caption.
% \hfil must be used as a separator to get equal spacing.
% The subfigure.sty package works much the same way, except \subfigure is
% used instead of \subfloat.
%
%\begin{figure*}[!t]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that IEEE does not put floats in the very first column - or typically
% anywhere on the first page for that matter. Also, in-text middle ("here")
% positioning is not used. Most IEEE journals/conferences use top floats
% exclusively. Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the \fnbelowfloat
% command of the stfloats package.

% conference papers do not normally have an appendix

% use section* for acknowledgement
\section*{Acknowledgments}
This work has been supported by the CRESTA project that has received
funding from the European Community's Seventh Framework Programme
(ICT-2011.9.13) under Grant Agreement no. 287703.

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\begin{thebibliography}{1}

%\bibitem{IEEEhowto:kopka}
%H.~Kopka and P.~W. Daly, \emph{A Guide to \LaTeX}, 3rd~ed.\hskip 1em plus
%  0.5em minus 0.4em\relax Harlow, England: Addison-Wesley, 1999.

\bibitem{succi-book}
S. Succi, \textit{The lattice Boltzmann equation and beyond},
Oxford University Press, Oxford, 2001.

\bibitem{mpi-standard}
Message Passing Interface Forum, http://www.mpi-forum.org

\bibitem{desplat}
Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
\textit{LUDWIG: A parallel lattice-Boltzmann code for complex fluids}.
Comput. Phys. Comms., \textbf{134}, 273, 2001.

\bibitem{aidun2010}
C.K. Aidun and J.R. Clausen,
\textit{Lattice Botlzmann method for complex flows},
Ann. Rev. Fluid Mech., \textbf{42} 439--472 (2010).

\bibitem{bray1994}
A.J. Bray,
\textit{Theory of phase-ordering kinetics},
Adv. Phys., \textbf{43} 357--459 (1994).

\bibitem{kendon2001}
V.M. Kendon, M.E. Cates, I. Pangonabarraga, J.-C. Desplat, and P. Bladon,
\textit{Inertial effects in the three-dimensional spinodal decomposition of a
symmetric binary fluid mixture; a lattice Boltzmann study},
J. Fluid Mech., \textbf{440}, 147--203 (2001).

\bibitem{swift1996}
M.R. Swift, E. Orlandini, W.R. Osborn, and J.M. Yeomans,
\textit{Lattice Boltzmann simulations of liquid-gas and binary
fluid systems},
Phys. Rev. E, \textbf{54}, 5041--5052 (1996).

\bibitem{stratford-jsp2005}
K. Stratford, R. Adhikari, I. Pagonabarraga, and J.-C. Desplat,
\textit{Lattice Boltzmann for Binary Fluids with Suspended Colloids},
J. Stat. Phys. \textbf{121}, 163 (2005).

\bibitem{xe6}
Cray XE6 Product Brochure, available from http://www.cray.com/Products/XE/Resources.aspx (2010)

\bibitem{hector} HECToR home page, http://www.hector.ac.uk

\bibitem{xk6}
Cray XK6 Product Brochure, available from http://www.cray.com/Products/XK6/XK6.aspx (2011)

%\bibitem{wei2004}
%X. Wei, W. Li, K. M\"uller, and A.E. Kaufman,
%\textit{The lattice Blotzmann method for simulating gaseous phenomena},
%IEEE Transactions on Visualization and Computer Graphics,
%\textbf{10}, 164--176 (2004).
%% Apparently first LB via GPU; a serious contribution using single fluid
%%d3q19 single relaxation time.

%\bibitem{zhu2006}
%H. Zhu, X. Liu, Y. Liu, and E. Wu,
%\textit{Simulation of miscible binary mixtures based on lattice Boltzmann method},
%Comp. Anim. Virtual Worlds, \textbf{17}, 403--410 (2006).
%% Single relaxation (MRT mentioned) time d3q19 apparently sound although
%%not a lot of detail. Pre-cuda so graphics code.

%\bibitem{zhao2007}
%Y. Zhao,
%\textit{Lattice Boltzmann based PDE solver on the GPU},
%Visual Comput., doi 10.1007/s00371-0070191-y (2007).

%\bibitem{toelke}
%J. T\"olke,
%\textit{Implementation of a lattice Boltzmann kernel using the Compute Unified
%Device Architecture developed by NVIDIA},
%Comput. Visual. Sci. doi:10.1007/s00791-008-0120-2 (2008).

%\bibitem{fan2004}
%Z. Fan, F. Qiu, A. Kaufman, and S. Yoakum-Stover,
%\textit{GPU cluster for high performance computing},
%Proceedings of ACM/IEEE Supercomputing Conference, pp. 47--59,
%IEEE Computer Society Press, Pittsburgh, PA (2004).

%\bibitem{myre2011}
%J. Myre, S.D.C. Walsh, D. Lilja, and M.O. Saar,
%\textit{Performance analysis of single-phase, multiphase, and multicomponent
%lattice Boltzmann fluid flow simulations on GPU clusters},
%Concurrency Computat.: Pract. Exper., \textbf{23}, 332--350 (2011).

%\bibitem{obrecht2011}
%C. Obrecht, F. Kuznik, B. Tourancheau, and J.-J. Roux,
%\textit{Multi-GPU implementation of the lattice Boltzmann method},
%Comput. Math. with Applications,
%doi:10.1016/j.camwa.2011.02.020 (2011).

%\bibitem{bernaschi2010}
%M. Bernaschi, M. Fatica, S. Melchionna, S. Succi, and E. Kaxiras,
%\textit{A flexible high-performance lattice Boltzmann GPU code for the
%simulations of fluid flow in complex geometries},
%Concurrency Computat.: Pract. Exper., \textbf{22}, 1--14 (2010).

%\bibitem{xian2011}
%W. Xian and A. Takayuki,
%\textit{Multi-GPU performance of incompressible flow computation by
%lattice Boltzmann method on GPU cluster},
%Parallel Comput., doi:10.1016/j.parco.2011.02.007 (2011).

%\bibitem{simd} Intel AVX: New Frontiers in Performance Improvements and Energy Efficiency, Intel article, http://software.intel.com/en-us/articles (2009)

%\bibitem{amd} AMD 6200 series home page  
%http://www.amd.com/us/products/server/
%processors/6000-series-platform/6200/Pages/6200-series-processors.aspx (accessed April 2012)

%\bibitem{nvidia} NVIDIA Tesla home page, http://www.nvidia.com/object/
%tesla\_computing\_solutions.html (accessed April 2012)

%\bibitem{titan} \textit{ORNL awards contract to Cray for Titan supercomputer}, Oak Ridge National Laboratory Press Release mr20111011-00, http://www.ornl.gov/info/press\_releases/ (2011)

\end{thebibliography}

% that's all folks
\end{document}

### 2012-11-06 branches/CUDA/BookGPU/Chapters/chapter2/ch2.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1807 40876441-3496-0410-a613-a2fca35b2ae2

\chapterauthor{Raphaël Couturier}{Femto-ST Institute, University of Franche-Comte}

\chapter{Introduction to Cuda}
\label{chapter2}

\section{Introduction}
\label{ch2:intro}

In this chapter  we give some simple examples of Cuda  programming.  The goal is
not to provide an exhaustive presentation of all the functionalities of Cuda but
rather to give some basic elements. Of  course, readers that do not know Cuda are
invited  to read  other  books that  are  specialized on  Cuda programming  (for
example: \cite{ch2:Sanders:2010:CEI}).

\section{First example}
\label{ch2:1ex}

This first example is  intented to show how to build a  very simple program with
Cuda.  Its goal  is to perform the sum  of two arrays and put the  result into a
third array.  A Cuda program consists in  a C code which calls Cuda kernels that
are executed on a GPU. The listing of this code is in Listing~\ref{ch2:lst:ex1}.

As GPUs have  their own memory, the first step consists  in allocating memory on
the GPU.  A call to  \texttt{cudaMalloc}\index{Cuda~functions!cudaMalloc} allows
to allocate memory on the GPU. The first parameter of this function is a pointer
on a memory on the device (i.e. the GPU). In this example, \texttt{d\_} is added
on each variable allocated  on the GPU, meaning this variable is  on the GPU. The
second parameter represents the size of the allocated variables, this size is expressed in
bits.

In this example, we  want to compare the execution time of  the additions of two
arrays in  CPU and  GPU. So  for both these  operations, a  timer is  created to
measure the  time. Cuda proposes to  manipulate timers quite  easily.  The first
step is to create the timer\index{Cuda~functions!timer}, then to start it and at
the end to stop it. For each of these operations a dedicated function is used.

In  order to  compute  the same  sum  with a  GPU, the  first  step consists  in
transferring the data from the CPU (considered as the host with Cuda) to the GPU
(considered as the  device with Cuda).  A call  to \texttt{cudaMemcpy} allows to
copy the content of an array allocated in the host to the device when the fourth
parameter                                 is                                 set
to  \texttt{cudaMemcpyHostToDevice}\index{Cuda~functions!cudaMemcpy}.  The first
parameter of the function is the  destination array, the second is the
source  array and  the third  is the  number of  elements to  copy  (expressed in
bytes).

Now the  GPU contains the  data needed to  perform the addition.   In sequential
programming, such  addition is  achieved out  with a loop  on all  the elements.
With a GPU,  it is possible to perform  the addition of all the  elements of the
two  arrays in  parallel (if  the number  of blocks  and threads  per  blocks is
sufficient).   In Listing\ref{ch2:lst:ex1}  at the  beginning, a  simple kernel,
called \texttt{addition} is defined to  compute in parallel the summation of the
two     arrays.      With     Cuda,     a     kernel     starts     with     the
keyword   \texttt{\_\_global\_\_}   \index{Cuda~keywords!\_\_shared\_\_}   which
indicates that this kernel can be called from the C code.  The first instruction
in this kernel is used to compute the variable \texttt{tid} which represents the
thread index.   This thread index\index{thread  index} is computed  according to
the           values            of           the           block           index
(called  \texttt{blockIdx} \index{Cuda~keywords!blockIdx}  in Cuda)  and  of the
thread   index   (called   \texttt{blockIdx}\index{Cuda~keywords!threadIdx}   in
Cuda). Blocks of threads and thread  indexes can be decomposed into 1 dimension,
2 dimensions or  3 dimensions.  According to the  dimension of manipulated data,
the appropriate dimension  can be useful. In our example,  only one dimension is
used.   Then using notation  \texttt{.x} we  can access  to the  first dimension
(\texttt{.y}  and \texttt{.z}  respectively allow to access  to the  second and
third dimension).   The variable \texttt{blockDim}\index{Cuda~keywords!blockDim}
gives the size of each block.

\lstinputlisting[label=ch2:lst:ex1,caption=A simple example]{Chapters/chapter2/ex1.cu}

\section{Second example: using CUBLAS}
\label{ch2:2ex}

The Basic Linear Algebra Subprograms  (BLAS) allows programmers to use efficient
routines  that are  often  required. Those  routines  are heavily  used in  many
scientific applications  and are optimized for  vector operations, matrix-vector
operations                           and                           matrix-matrix
operations~\cite{ch2:journals/ijhpca/Dongarra02}. Some  of those operations seem
to be  easy to  implement with Cuda.   Nevertheless, as  soon as a  reduction is
needed, implementing an efficient reduction routine with Cuda is far from being
simple. Roughly speaking, a reduction operation\index{reduction~operation} is an
operation  which combines  all the  elements of  an array  and extracts  a number
computed with all the  elements. For example, a sum, a maximum  or a dot product
are reduction operations.

In this second example, we consider that  we have two vectors $A$ and $B$. First
of all, we want to compute the sum  of both vectors in a vector $C$. Then we want
to compute the  scalar product between $1/C$ and $1/A$. This  is just an example
which has no direct interest except to show how to program it with Cuda.

Listing~\ref{ch2:lst:ex2} shows this example with Cuda. The first kernel for the
addition  of two  arrays  is exactly  the same  as  the one  described in  the
previous example.

The  kernel  to  compute the  opposite  of  the  elements  of  an array  is  very
simple. For  each thread index,  the inverse of  the array replaces  the initial
array.

In the main function,  the beginning is very similar to the  one in the previous
example.  First,  the user is  askef to define  the number of elements.   Then a
call  to \texttt{cublasCreate}  allows  to initialize  the  cublas library.   It
creates a handle. Then all the arrays  are allocated in the host and the device,
as in the  previous example.  Both arrays $A$ and $B$  are initialized.  The CPU
computation is performed  and the time for this CPU  computation is measured. In
order to  compute the same result  on the GPU, first  of all, data  from the CPU
need to be  copied into the memory of  the GPU. For that, it is  possible to use
cublas   function   \texttt{cublasSetVector}.    This   function   has   several
arguments. More precisely, the first  argument represents the number of elements
to transfer, the second arguments is the size of each element, the third element
represents the source  of the array to  transfer (in the GPU), the  fourth is an
offset between each element of the source  (usually this value is set to 1), the
fifth is  the destination (in the  GPU) and the  last is an offset  between each
element  of the  destination. Then  we call  the kernel  \texttt{addition} which
computes the  sum of all elements  of arrays $A$ and  $B$.  The \texttt{inverse}
kernel  is called twice,  once to  inverse elements  of array  $C$ and  once for
$A$. Finally,  we call the  function \texttt{cublasDdot} which computes  the dot
product  of two  vectors.   To use  this  routine, we  must  specify the  handle
initialized by  Cuda, the number  of elements to  consider, then each  vector is
followed by the offset between every  element.  After the GPU computation, it is
possible to check that both computation produce the same result.

\lstinputlisting[label=ch2:lst:ex2,caption=A simple example with cublas]{Chapters/chapter2/ex2.cu}

\section{Third example: matrix-matrix multiplication}
\label{ch2:3ex}

Matrix-matrix multiplication is an operation  which is quite easy to parallelize
with a GPU. If we consider that  a matrix is represented using a two dimensional
array, $A[i][j]$ represents the element of  the $i^{th}$ row and of the $j^{th}$
column. In  many cases, it is  easier to manipulate a  1D array instead  of a 2D
array.   With Cuda,  even if  it is  possible to  manipulate 2D  arrays,  in the
following we present an example based on a 1D array. For the sake of simplicity,
we  consider we  have  a square  matrix of  size  \texttt{size}.  So  with a  1D
array,  \texttt{A[i*size+j]} allows  us to  have access  to the  element  of the
$i^{th}$ row and of the $j^{th}$ column.

With  a sequential  programming, the  matrix multiplication  is  performed using
three loops. We assume that $A$, $B$  represent two square matrices and the
result   of    the   multiplication    of   $A   \times    B$   is    $C$.   The
element \texttt{C[i*size+j]} is computed as follows:
\begin{equation}
C[i*size+j]=\sum_{k=0}^{size-1} A[i*size+k]*B[k*size+j];
\end{equation}

In Listing~\ref{ch2:lst:ex3},  the CPU computation  is performed using  3 loops,
one  for $i$,  one for  $j$  and one  for $k$.   In  order to  perform the  same
computation on a  GPU, a naive solution consists in  considering that the matrix
$C$ is split into  2 dimensional blocks.  The size of each  block must be chosen
such as the number of threads per block is inferior to $1,024$.

In Listing~\ref{ch2:lst:ex3},  we consider that  a block contains 16  threads in
each   dimension,  the   variable  \texttt{width}   is  used   for   that.   The
variable \texttt{nbTh} represents the number of threads per block. So, to be able
to compute the matrix-matrix product on a GPU, each block of threads is assigned
to compute the result  of the product for the elements of  this block.  The main
part of the code is quite similar to the previous code.  Arrays are allocated in
the  CPU and  the GPU.   Matrices $A$  and $B$  are randomly  initialized.  Then
arrays are  transferred inside the  GPU memory with call  to \texttt{cudaMemcpy}.
So the first step for each thread of a block is to compute the corresponding row
and   column.    With   a    2   dimensional   decomposition,   \texttt{int   i=
blockIdx.y*blockDim.y+ threadIdx.y;} allows us to compute the corresponding line
and  \texttt{int  j=   blockIdx.x*blockDim.x+  threadIdx.x;}  the  corresponding
column. Then each  thread has to compute the  sum of the product of  the line of
$A$   by   the  column   of   $B$.    In  order   to   use   a  register,   the
kernel  \texttt{matmul}  uses a  variable  called  \texttt{sum}  to compute  the
sum. Then the result is set into  the matrix at the right place. The computation
of  CPU matrix-matrix multiplication  is performed  as described  previously.  A
timer measures  the time.   In order to  use 2 dimensional  blocks, \texttt{dim3
dimGrid(size/width,size/width);} allows us  to create \texttt{size/width} blocks
in each  dimension.  Likewise,  \texttt{dim3 dimBlock(width,width);} is  used to
create \texttt{width} thread  in each dimension. After that,  the kernel for the
matrix  multiplication is  called. At  the end  of the  listing, the  matrix $C$
computed by the GPU is transferred back  into the CPU and we check if both matrices
C computed by the CPU and the GPU are identical with a precision of $10^{-4}$.

With $1,024  \times 1,024$ matrices,  on a C2070M  Tesla card, this  code takes
$37.68$ms to perform the multiplication. With an Intel Xeon E31245 at $3.30$GHz, it
takes $2465$ms  without any parallelization (using only  one core). Consequently
the speed up  between the CPU and GPU  version is about $65$ which  is very good
regarding the difficulty of parallelizing this code.

\lstinputlisting[label=ch2:lst:ex3,caption=simple Matrix-matrix multiplication with cuda]{Chapters/chapter2/ex3.cu}

\section{Conclusion}
In this chapter, three simple Cuda examples have been  presented. They are
quite  simple. As we  cannot  present  all the  possibilities  of  the  Cuda
programming, interested  readers  are  invited  to  consult  Cuda  programming
introduction books if some issues regarding the Cuda programming are not clear.

\putbib[Chapters/chapter2/biblio]

### 2012-11-06 branches/CUDA/BookGPU/frontmatter/Foreword.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1807 40876441-3496-0410-a613-a2fca35b2ae2

\chapter*{Foreward}
I am delighted to introduce the first book on Multimedia Data Mining.  When I came to know about this book project undertaken by two of the most active young researchers in the field, I was pleased that this book is coming in early stage of a field that will need it more than most fields do.  In most emerging research fields, a book can play a significant role in bringing some maturity to the field.  Research fields advance through research papers.  In research papers, however, only a limited perspective could be provided about the field, its application potential, and the techniques required and already developed in the field.  A book gives such a chance.  I liked the idea that there will be a book that will try to unify the field by bringing in disparate topics already available in several papers that are not easy to find and understand.  I was supportive of this book project even before I had seen any material on it.  The project was a brilliant and a bold idea by two active researchers.  Now that I have it on my screen, it appears to be even a better idea.

Multimedia started gaining recognition in 1990s as a field.  Processing, storage, communication, and capture and display technologies had advanced enough that researchers and technologists started building approaches to combine information in multiple types of signals such as audio, images, video, and  text.  Multimedia computing and communication techniques recognize correlated information in multiple sources as well as insufficiency of information in any individual source.    By properly selecting sources to provide complementary information, such systems aspire, much like human perception system, to create a holistic picture of a situation using only partial information from separate sources.

Data mining is a direct outgrowth of progress in data storage and processing speeds.  When it became possible to store large volume of data and run different statistical computations to explore all possible and even unlikely correlations among data, the field of data mining was born.  Data mining allowed people to hypothesize relationships among data entities and explore support for those.  This field has been put to applications in many diverse domains and keeps getting more applications.  In fact many new fields are direct outgrowth of data mining and it is likely to become a powerful computational tool.\vadjust{\vfill\pagebreak}

\thispagestyle{empty}

### 2012-11-06 branches/CUDA/BookGPU/frontmatter/Preface.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1807 40876441-3496-0410-a613-a2fca35b2ae2

\chapter*{Preface}

Approximately 17 million people in the USA (6{\%} of the
population) and 140 million people worldwide (this number is
expected to rise to almost 300 million by the year 2025) suffer
from \textit{diabetes mellitus}. Currently, there a few dozens of
commercialised devices for detecting blood glucose levels [1].
However, most of them are invasive. The development of a
noninvasive method would considerably improve the quality of life
for diabetic patients, facilitate their compliance for glucose
monitoring, and reduce complications and mortality associated with
this disease. Noninvasive and continuous monitoring of glucose
concentration in blood and tissues is one of the most challenging
and exciting applications of optics in medicine. The major
difficulty in development and clinical application of optical
noninvasive blood glucose sensors is associated with very low
signal produced by glucose molecules. This results in low
sensitivity and specificity of glucose monitoring by optical
methods and needs a lot of efforts to overcome this difficulty.

A wide range of optical technologies have been designed in
attempts to develop robust noninvasive methods for glucose
sensing. The methods include infrared absorption, near-infrared
scattering, Raman, fluorescent, and thermal gradient
spectroscopies, as well as polarimetric, polarization
heterodyning, photonic crystal, optoacoustic, optothermal, and
optical coherence tomography (OCT) techniques [1-31].

For example, the polarimetric quantification of glucose is based
on the phenomenon of optical rotatory dispersion, whereby a chiral
molecule in an aqueous solution rotates the plane of linearly
polarized light passing through the solution. The angle of
rotation depends linearly on the concentration of the chiral
species, the pathlength through the sample, and the molecule
specific rotation. However, polarization sensitive optical
technique makes it difficult to measure \textit{in vivo} glucose
concentration in blood through the skin because of the strong
light scattering which causes light depolarization. For this
reason, the anterior chamber of the eye has been suggested as a
sight well suited for polarimetric measurements, since scattering
in the eye is generally very low compared to that in other
tissues, and a high correlation exists between the glucose in the
blood and in the aqueous humor. The high accuracy of anterior eye
chamber measurements is also due to the low concentration of
optically active aqueous proteins within the aqueous humor.

On the other hand, the concept of noninvasive blood glucose
sensing using the scattering properties of blood and tissues as an
alternative to spectral absorption and polarization methods for
monitoring of physiological glucose concentrations in diabetic
patients has been under intensive discussion for the last decade.
Many of the considered  effects, such as changing of the size,
refractive index, packing, and aggregation of RBC under glucose
variation, are important for glucose monitoring in diabetic
patients. Indeed, at physiological concentrations of glucose,
ranging from 40 to 400 mg/dl, the role of some of the effects may
be modified, and some other effects, such as glucose penetration
inside the RBC and the followed hemoglobin glycation, may be
important [30-32].

Noninvasive determination of glucose was attempted using light
scattering of skin tissue components measured by a
spatially-resolved diffuse reflectance or NIR fre\-quen\-cy-domain
reflectance techniques. Both approaches are based on change in
glucose concentration, which affects the refractive index mismatch
between the interstitial fluid and tissue fibers, and hence
reduces scattering coefficient. A glucose clamp experiment showed
that reduced scattering coefficient measured in the visible range
qualitatively tracked changes in blood glucose concentration for
the volunteer with diabetes studied.

### 2012-11-07 trunk/papers/bprheo/bprheo.tex
Updated pictures and Tab.1
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1808 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \begin{table*}[htpb]
-\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
+\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
 \hline
-$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
 \hline
@@
 \hline
-0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
-3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
+AN & 0.244 & 0.08& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+AN & 0.488 & 0.17 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+AN & 0.976 & 0.33 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+AN & 1.95 & 0.67 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
 \hline
+3D-PRC & 3.91 & 1.33 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
+3D-PRC & 7.81 & 2.67 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+3D-PRC & 15.63 & 5.33 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+3D-PRC & 15.63 & 5.33 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+3D-PRC & 31.25 &10.67 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+3D-PRC & 46.87 & 15.99 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+\hline
+2D-DTR & 62.50 & 21.33& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
+2D-DTR &78.125 &26.66 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+\hline
+1D-CH & 93.75 &32.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
+1D-CH & 125.0 & 42.66& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+\hline
+FAN & 187.5 & 63.99& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+\hline
+
 BPII \\
 \hline
-0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
-62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+3D-PRC & 0.244 &0.04 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+3D-PRC & 0.488 &0.08 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+3D-PRC & 0.976 &0.17 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+3D-PRC & 1.95 &0.33  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+3D-PRC &3.91 & 0.67& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+3D-PRC &7.81 &1.33 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+3D-PRC &15.63 & 2.67& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+3D-PRC & 15.63 & 2.67& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+3D-PRC & 31.25 &5.34 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+\hline
+1D-THW & 46.87 &8.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
+1D-THW & 62.50 & 10.67& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+\hline
+1D-CH & 78.125 & 13.33 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
+1D-CH & 93.75 &16.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
+\hline
+FAN & 125.0 &21.33 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
 \hline
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minimum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
+\caption{Minima, maxima and standard deviation of time-averaged velocity components 
+for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
+shear flow: All values are given in $10^{-5}$ lattice units. The transient dynamics 
+has not been included in the averaging procedure. The regimes comprise periodically 
+recurring conformation (3D-PRC), two-dimensional double twist rolls (2D-DTR), 
+one-dimensional helix (1D-CH), one-dimensional travelling helical wave (1D-THW) 
+and a flow-aligned nematic state (FAN).}
 \label{tab1}
@@
 \begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115.png}\\
+\includegraphics[width=0.245\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-700k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2200k_run1115.png}
 \caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
@@
 \begin{figure}[htpb]
-%\includegraphics[width=0.4\textwidth]{dir+z-300k_run916.png}
-\includegraphics[width=0.4\textwidth]{dir+z-301k_run916.png}
-\includegraphics[width=0.4\textwidth]{dir+z-302k_run916.png}
+\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
 \caption{Director field of BPI at high shear rate: The pictures show a slice 
-of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction 
+of $L_x\times L_y \times L_Z= 32\times32\times 32$ lattice sites along the vorticity direction 
 with the flow/gradient direction oriented along the horizontal/vertical axis. 
@@
 Ericksen numbers ${\it Er}\simeq 10$.
+
+\subsection{Transition to cholesteric helix and flow-aligned nematic state}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
+\caption{}
+\label{cholflow}
+\end{figure}
 

### 2012-11-08 trunk/papers/bprheo/bprheo.tex
Updated text - version for Davide to check out
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1809 40876441-3496-0410-a613-a2fca35b2ae2

@@
 $V$ and $L$ as typical velocity and length scale. 
+In the present work $L=p_0/2$ was assumed to be the unit cell size,
+whereas $V$ was taken to be the velocity difference
+across one unit cell, i.e. $V=\dot{\gamma} p_0/2$.
 
@@
 \hline
-AN & 0.244 & 0.08& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-AN & 0.488 & 0.17 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-AN & 0.976 & 0.33 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-AN & 1.95 & 0.67 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
-\hline
-3D-PRC & 3.91 & 1.33 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-3D-PRC & 7.81 & 2.67 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-3D-PRC & 15.63 & 5.33 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-3D-PRC & 15.63 & 5.33 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-3D-PRC & 31.25 &10.67 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-3D-PRC & 46.87 & 15.99 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
-\hline
-2D-DTR & 62.50 & 21.33& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
-2D-DTR &78.125 &26.66 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
-\hline
-1D-CH & 93.75 &32.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
-1D-CH & 125.0 & 42.66& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
-\hline
-FAN & 187.5 & 63.99& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+\hline
+BPI-2 & 3.91 & 0.67 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
+BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+\hline
+BPI-3 & 62.50 & 10.67& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
+BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+\hline
+CHG & 93.75 &16.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
+CHG & 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+\hline
+FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
 \hline
@@
 \hline
-3D-PRC & 0.244 &0.04 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-3D-PRC & 0.488 &0.08 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-3D-PRC & 0.976 &0.17 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-3D-PRC & 1.95 &0.33  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-3D-PRC &3.91 & 0.67& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-3D-PRC &7.81 &1.33 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-3D-PRC &15.63 & 2.67& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-3D-PRC & 15.63 & 2.67& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-3D-PRC & 31.25 &5.34 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
-\hline
-1D-THW & 46.87 &8.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
-1D-THW & 62.50 & 10.67& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
-\hline
-1D-CH & 78.125 & 13.33 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
-1D-CH & 93.75 &16.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
-\hline
-FAN & 125.0 &21.33 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
+BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+\hline
+BPII-2 & 46.87 &4.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
+BPII-2 & 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+\hline
+CHG & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
+CHG & 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
+\hline
+FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
 \hline
@@
 shear flow: All values are given in $10^{-5}$ lattice units. The transient dynamics 
-has not been included in the averaging procedure. The regimes comprise periodically 
-recurring conformation (3D-PRC), two-dimensional double twist rolls (2D-DTR), 
-one-dimensional helix (1D-CH), one-dimensional travelling helical wave (1D-THW) 
+has not been included in the averaging procedure. The regimes comprise the formation
+of an amorphous network (BPI-1), periodically recurring conformations (BPI-2 and BPII-1), 
+two-dimensional double twist rolls (BPI-3), a one-dimensional travelling helical wave (BPII-2), 
+a cholesteric helix oriented along the gradient direction (CHG), 
 and a flow-aligned nematic state (FAN).}
@@
 
-For shear rates $\gd\ge6.25\e{-4}$ the network of disclination lines
+For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ the network of disclination lines
 breaks up and oscillations in the stress signal are absent.
@@
 blue phase dissolves into a simple cholesteric liquid 
-crystal. The mode of flow is a travelling helical wave with the helical axis 
-oriented along the vorticity direction.
+crystal. The helical axis is oriented in vorticity direction and 
+the mode of flow is a travelling helical wave.
+
+At shear rates $7.81\e{-4}\lesssim\gd9.37\e{-4}$ the helix changes orientation.
+The helical axis is now oriented along the gradient direction with
+the liquid crystal flowing in 'nematic planes'. For even higher shear rates
+the system undergoes a transition to a flow-aligned nematic state.
 
@@
 and is the preferred mode of flow of a cholesteric 
-liquid crystal at low Ericksen numbers below the uncoiling transition 
-to a nematic, flow-aligned state at even higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
-
-In our scenario the break-up occurs for shear rates $\gd>3.12\e{-4}$, which
-corresponds to Ericksen numbers of ${\it Er}\simeq10$, based on our
-parameters ($K=0.02, \eta=0.66$), typical velocities (viz. Tab. \ref{tab1}) 
-and length scales ($p/2$).
+liquid crystal in vorticity mode at low Ericksen numbers, i.e. below 
+the uncoiling transition to a nematic, flow-aligned state at even 
+higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
+
+In our scenario the break-up occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
+corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
+parameters for BPII ($K=0.02, \eta=0.666$, Tab. \ref{tab1} and $p_0/2=16$).
 
@@
 
+We observe another configuration between the travelling helical wave and the 
+flow-aligned nematic state at shear rates $7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$, 
+which has so far not been reported to our knowledge \com{CHECK}.
+We will address this in section \ref{cholflow}.
+
 \subsection{Blue Phase I}
@@
 the differences between BPI and BPII regarding their rheological response. 
-
 We present again key aspects in an overview before we 
-address some features in more detail. Fig. \ref{bp1-rheo} shows the 
-apparent viscosity $\eta_{app}$ versus time and strain for the same shear rates
-as in Fig. \ref{bp2-rheo}.
-For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
-the disclination network occurs. This is the same critical shear rate as 
-the one observed in BPII-2. But this time the configuration is quasi 
-two-dimensional.
-
-In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
+address some features in more detail.
+
+Fig. \ref{bp1-rheo} shows the apparent viscosity $\eta_{app}$ versus time and 
+strain for the same shear rates as in Fig. \ref{bp2-rheo}.
+For shear rates $9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ we observe the same cholesteric 
+configuration with the helical axis along the gradient direction as for BPII (see
+section \ref{cholflow}).
+
+In regime BPI-3 $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ a breakup of 
+the blue phase disclination network occurs. This is about the same critical shear rate 
+where we observe the transition to BPII-2. However, for BPI the configuration is not
+a travelling helical wave, but rather a quasi two-dimensional state consisting of
+double-twist rolls.
+
+At shear rates $3.91\e{-5}\lesssim\gd\lesssim4.68\e{-4}$ we find flow with 
+periodically recurring conformations. We refer to this regime as BPI-2. 
+The oscillations in the 
 deviatoric stress that are not sinusoidal, but show an increasingly complex
 pattern up to the breakup point.
-For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
-these oscillations become very erratic. There is no regular recurrence and 
+For even lower shear rates $\gd\lesssim3.91\e{-5}$ in regime BPI-1 
+these oscillations become very erratic and unstable. 
+There is no regular recurrence and 
 after a few cycles the stress pattern is completely irregular.
+Finally, the system ends up in an amorphous configuration.
 Shear-thinning is slightly stronger than in BPII.
@@
 \caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
-Depicted is the transition from flow-induced, intertwined helices that 
-undergo a recurring affine transformation (top left and centre, both at 
-time step $t=4.0\e{5}$; top right at time step $t=7.0\e{5}$) with an intermediate 
-conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) 
-to a distorted state (bottom centre at time step $t=1.7\e{6}$) and 
-finally to an amorphous network (bottom right at $t=2.2\e{6}$). 
-The picture at the top centre shows the network in gradient direction.
-All other pictures have the view along the vorticity direction.}
+Depicted is the transition from the quiescent state to flow-induced, intertwined 
+helices that undergo a recurring affine transformation. The top row
+shows a section of one unit cell for early times 
+$t=1\e{4}, 1.8\e{5}, 2.0\e{5}$ and $2.1\e{5}$. The bottom 
+pictures on the far left, centre left and centre right 
+show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
+which the system passes for several cycles before it ends up in
+an amorphous state (bottom row far right at time step $t=2.2\e{6}$).}
 \label{bp1-low}
@@
 at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
-and Ericksen numbers $0.5\le{\it Er}<10$ 
+and Ericksen numbers $0.67\le{\it Er}<8$ 
 lies another region where the network flows with periodically 
@@
 \includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a slice 
-of $L_x\times L_y \times L_Z= 32\times32\times 32$ lattice sites along the vorticity direction 
-with the flow/gradient direction oriented along the horizontal/vertical axis. 
-The colour code gives the magnitude of the in-plane component of the director field. 
+\caption{Director field of BPI at high shear rate: The pictures show a region 
+of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
+direction oriented along the horizontal/vertical axis. 
+The colour code gives the magnitude of the z-component of the director field. 
 The data is translationally invariant in vorticity direction.}
@@
 
-At high flow rates we observe a break-up of the flow-induced 
+At high flow rates $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ 
+we observe a break-up of the flow-induced 
 disclination network of regime BPI-2. However, 
-contrary to BPII, where the configuration is a travelling helical 
+in contrast to BPII, where the configuration is a travelling helical 
 wave (viz. Fig. \ref{bp2-high}), BPI prefers a quasi 
@@
 regions of local double twist.
-Interestingly the critical shear rate for the break-up lies in the same 
-window as that of transition between BPII-1 and BPII-2, 
- $3.125\e{-4}\le\gd_c\le 6.25\e{-4}$ and occurs at similar 
-Ericksen numbers ${\it Er}\simeq 10$.
-
-\subsection{Transition to cholesteric helix and flow-aligned nematic state}
+Interestingly the critical Ericksen number for the break-up lies in 
+about the same window as that of transition between BPII-1 and BPII-2.
+
+\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{cholflow}
+
+At shear rates beyond the regimes BPI-3 and BPII-2 and below the transition to a 
+flow-aligned nematic state we found another regime where 
+both blue phases adopt the same configuration in steady shear flow, 
+independent of their initial state.
+The director field of this configuration is shown in Fig. \ref{cholflow}.
+It consists of a simple cholesteric helix with the helical axis oriented 
+along the gradient direction. The state is translational invariant in 
+flow and vorticity direction and therefore also one-dimensional. 
+However, in contrast to the travelling helical wave of BPII-2 the 
+director field is now static and there is no tumbling motion. 
+While flowing the director retains its relative orientation.
+Consequently, the shear stress is lower than for
+the 2D double-twist rolls of BPI-3 and the travelling helical wave
+(Fig. \ref{bp2-rheo} and \ref{bp1-rheo}) as there is less dissipation.
 
@@
 \includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
-\caption{}
+\caption{Flow induced cholesteric state: The orientation of the helix is 
+along the gradient direction. The director field is quasi-static during 
+the flow, which occurs in 'nematic layers'. For clarity the picture shows 
+the director field along the y-direction for selected sites in the x-z-plane.}
 \label{cholflow}
@@
 
-\clearpage
+%\clearpage
 
@@
 two different flow regimes for BPII and three different ones in case of BPI. 
-Below Ericksen numbers ${\cal O}(10)$ BPII exhibits weak shear-thinning and 
+Below Ericksen numbers $Er\simeq3$ BPII exhibits weak shear-thinning and 
 obeys a power-law.  The BPII disclination network breaks up and reconnects 
@@
 
-At larger Ericksen numbers the flowing BPII network breaks up completely into 
-a simple cholesteric phase and flows as a travelling helical wave.
+At larger Ericksen numbers $4\lesssim Er\lesssim 10$ the flowing BPII network 
+breaks up in a two-step scenario. First a simple cholesteric helix forms with the helical axis 
+in vorticity direction. The flow is in vorticity mode, i.e. as a travelling helical wave. 
+At slightly larger shear rates another cholesteric conformation emerges.
+However, this time the orientation of the helix is in gradient direction.
+The flow is in quasi-nematic layers with static orientation of the directior field.
 
@@
 This is a direct consequence of the topological differences in both disclination 
-network. Below Ericksen numbers ${\cal O}(0.1)$ the BPI network cannot flow and 
-retain a regular appearance. Shortly after the onset of the shearing 
-very large stress fluctuations occur. At the lowest shear rates $\gd<9.76\e{-6}$
-the contribution of the network to the total deviatoric stress is actually temporarily 
-negative during one cycle. These fluctuations destabilise the network and 
+network. Below Ericksen numbers $Er<0.4$ the BPI network cannot flow and 
+retain a regular appearance. It dissolves eventually into an amorphous network and 
+features yield-stress behaviour. The reason is that shortly after the onset of the shearing 
+very large stress fluctuations occur. At the lowest shear rates
+the contribution of the network to the total deviatoric stress is temporarily 
+actually negative during one cycle. These fluctuations destabilise the network and 
 eventually trigger its dissolution into an amorphous state with 
@@
 
-At slightly larger Ericksen numbers $0.1<{\it Er}<10$ shear thinning is 
+At slightly larger Ericksen numbers $0.4<{\it Er}<8$ shear thinning is 
 slightly stronger than for BPII, but BPI does not show simple power-law 
@@
 
-Just as in case of BPII, the BPI network breaks up at Ericksen numbers ${\cal O}(10)$,
-but adopts a quasi-two-dimensional configuration that is translationally 
-invariant in vorticity direction.
+Just as in case of BPII, the BPI network breaks up at larger Ericksen numbers in
+a two-step process. First at $10\lesssim\gd\lesssim 15$, it adopts a 
+quasi-two-dimensional configuration that is translationally invariant in vorticity direction. 
+Then at $16\lesssim Er \lesssim 22$ it forms cholestric helix along the gradient 
+direction that emerges from BPII.  
+Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
+is a flow-aligned nematic state.
 

### 2012-11-12 trunk/papers/bprheo/bprheo.tex
update Davide
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1811 40876441-3496-0410-a613-a2fca35b2ae2

@@
 II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
+For larger shear rates, the network breaks up into a cholesteric helix, lying either along the
+vorticity or the flow gradient direction according to the imposed flow.
+BPI leads to a very different response. Here, oscillations are only possible for intermediate 
+shear rates -- very slow flow leads to a transition into an amorphous network 
+with a residual stress. The flow-induced oscillation regime features a novel arrangement
+of double helical disclinations which rotate due to the shear. Large flow can, 
+just as in the BPII case, lead to the rupture of the network and the stabilisation
+of less ordered cholesteric structures. Our results provide the first 
+theoretical investigation of sheared blue phases in large systems, which
+are relevant to understanding the bulk rheology of these materials. 
 \end{abstract}
@@
 
+Our work differs from these earlier efforts as it addresses shear flows, 
+and focuses on flow-induced reconstruction and nonequilibrium transition
+between different blue phase networks, which appear at intermediate or
+high shear. Our simulations employ Lees-Edwards boundary conditions, which
+are naturally suited to address these regimes, and bulk as opposed
+to boundary-dominated flow. 
+Our simulations are large scale and parallel, so that we can study
+significantly larger systems than previously possible, so that our
+results can be tested with bulk rheology experiments.
+
+Our paper is organised as follows. 
+In Section II, we describe the method we use and review
+the hydrodynamic equation of motion which we aim to solve. In Section III, 
+we report our numerical results, separating them into subsections relative to
+blue phase I and blue phase II, and corresponding to low, intermediate, and
+high shear. Finally, we draw our conclusions in Section IV.
+
 \section{Model and Methods}
@@
 The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
-we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
+we work for simplicity in the one-elastic constant approximation. The wavevector $q_0$ is equal to $2\pi/p_0$, where $p_0$ is the cholesteric pitch.
 The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
-The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure~\cite{Alexander:2006}.
-
-A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature 
+The ``redshift'' $r=p/p_0$ is adjusted during the simulation, to optimise the free energy density
+at all times -- this is done by following the procedure
+previously described in~\cite{Alexander:2006}.
+
+A thermodynamic state is specified by two dimensionless quantities: the effective temperature 
 \begin{equation}
@@
 \end{equation}
-The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
+The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity advected by the flow.
 The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta v_\alpha$,
+and is explicitly given by
 \begin{eqnarray}
@@
 is a constant depending on the molecular details of the liquid crystal.
-Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by 
+Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the 
+co-called Leslie angle: we select this case by 
 setting $\xi=0.7$ in our simulations.
@@
 \end{eqnarray}
-In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} is reduced to the
+In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
 scalar pressure which, in a system at rest, is constant to a very good
@@
 
-It is possible to define a dimensionless number that describes the deformation
-of the director field under flow. The so-called Ericksen number
-is defined as 
+It is possible to define a further 
+dimensionless number that describes the deformation
+of the director field under flow. This so-called Ericksen number
+is given by 
 \begin{equation}
-Er=\frac{\eta V L}{K}
+Er=\frac{\eta v l}{K}
 \end{equation}
 with $\eta$ and $K$ being the viscosity and elastic constant and
-$V$ and $L$ as typical velocity and length scale. 
-In the present work $L=p_0/2$ was assumed to be the unit cell size,
-whereas $V$ was taken to be the velocity difference
-across one unit cell, i.e. $V=\dot{\gamma} p_0/2$.
+$v$ and $l$ as typical velocity and length scale. 
+In the present work $l=p_0/2$ was assumed to be the size of the BP unit cell,
+whereas $v$ was taken to be the velocity difference
+across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.
 
@@
 phases BPI and BPII. 
-The timestep and lattice spacing in lattice Boltzmann units (LBU) corresponds roughly to
-$\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units. The LB unit of stress
+The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
+approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
 is equal to about $10^8$~Pa. Further details about the conversion 
@@
 
-For the vast majority of simulations we opted for thermodynamic states that are 
-well inside the equilibrium region of the individual blue phase and far away 
-from the cholesteric-isotropic transition. Temperature and chirality were 
+For typical simulations reported in this work, 
+we chose as initial conditions thermodynamic states that are 
+well inside the equilibrium region of the individual blue phase, and far away 
+from the cholesteric-isotropic transition. Thus, temperature and chirality were 
 $\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
-For these parameters the total free energy density $f$ remained always negative, 
-even at the highest flow rate and despite larger gradient contributions $f_g$.
-We performed also a number of simulations at metastable states at higher and lower temperatures 
-and different at different chiralities, but didn't find any significant differences in the 
+For these parameters the total free energy density $f$ remained always negative at all flow rates
+simulated (so that the system is always in a liquid crystalline state). 
+We performed also selected simulations at metastable states at higher and lower temperatures, 
+and at different chiralities, but did not find any significant differences in the 
 general flow behaviour that is described below.
@@
 these states have on average higher free energy densities and smaller order parameters
-than those specified above and can therefore not withstand the same external forces.
+than those specified above and cannot therefore withstand the same external forces
+before breaking down.
 
@@
 During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
-After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow.
-We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, which is the equivalent 
-of 4 unit cells along each coordinate direction and a total of 64 unit cells in our simulation box.
-Runs with higher resolution confirmed that the resolution was adequate to track  
+After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow (as 
+it is no longer justified to optimise the free energy in a nonequilibrium scenario).
+We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, and we considered in both cases 
+4 unit cells along each coordinate direction, for a total of 64 unit cells in our simulation box.
+Runs with higher resolution confirmed that this choice was adequate to track  
 all kinematic details of the blue phase networks in shear flow.
@@
 $\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
-For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
-They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
+For clarity we classify various flow regimes, two in case of BPII and three in case of BPI. 
+They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$),
 BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
-Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
+Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical choices 
 are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
@@
 For clarity the individual curves have been shifted along the time and strain axis. 
-The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
+The inset in the top panel 
+shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
 regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
@@
 
-We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
-than that of BPI. BPII has simple cubic symmetry and the disclination lines intersect and form a characteristic network of nodes.
-Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
+We start our discussion with BPII as its flow behaviour is somewhat simpler 
+than that of BPI. BPII has simple cubic symmetry and the disclination lines 
+intersect and form a characteristic network of nodes.
+Ahead of the more detailed discussion we provide first a general overview of the 
+flow behaviour at all applied shear rates.
 Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
@@
 $\eta_{app}=\Pi_{xy}/\eta\,\gd$.
-A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
-any additional contribution from the liquid crystal.
+A numerical value of $\eta_{app}=1$ corresponds to a fully Newtonian flow,
+without any additional contribution from the liquid crystal.
 The top picture shows the data over time, whereas the graph at the bottom gives 
@@
 For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.
-
-The inset shows a flow curve, defined as time averages of 
+is because of the periodic breakup and reconnecting of the network in shear flow. 
+We refer to this regime as BPII-1. We note that permeation flows, which would
+{\em not} lead to oscillatory response, are effectively disallowed by the combination
+of the geometry and boundary condition we consider, because at the Lees-Edwards 
+plane the network needs to reconnect with its Galilean ``image'' travelling
+at the speed specified by the shear. This means that  our results are effectively
+restricted to intermediate or high flow, where permeation flows are 
+absent~\cite{Marenduzzo:2006a,Marenduzzo:2006b} and the response of the network
+is indeed oscillatory irrespective of the boundary conditions used at the wall. 
+We also note that, if the boundaries are such that
+the director field pattern sticks to the shear planes, then our analysis
+would apply and be realistic for arbitrarily low shear as well.
+
+The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
 the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
@@
 $a=0.35, b=0.95$ describes the data to a very good approximation. 
-Hence, the degree of shear-thinning is remarkably small, but may depend
-on the exact direction in which the strain is applied, an aspect that we     
-leave for future work.
+Hence, the degree of shear-thinning is remarkably small in BPII, in
+the range of shear rates which we have explored here.
 
 For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ the network of disclination lines
-breaks up and oscillations in the stress signal are absent.
+breaks up completely, and oscillations in the stress signal are absent.
 In this regime of shear rates, to which we refer as BPII-2, the 
 blue phase dissolves into a simple cholesteric liquid 
-crystal. The helical axis is oriented in vorticity direction and 
-the mode of flow is a travelling helical wave.
+crystal. The helical axis is along the vorticity direction and
+the flow excites travelling helical waves, such as those 
+predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis.
 
@@
 
-In the next paragraph we will investigate some of these aspects in more detail 
+In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
 by looking at the kinetics of the disclination network.
@@
 Fig. \ref{bp2-med} shows the disclination network in shear flow as
-it undergoes an affine transformation. The disclination lines break up and 
-reconnect further downstream forming a periodically recurring pattern with 
-recurrence time $\tau_F=1/\gd$ in flow direction.
+it undergoes an affine transformation in the BPII-1 flow regime. 
+The disclination lines break up and 
+reconnect further downstream, forming a periodically recurring pattern. The
+period needed for a pattern to break up and 
+reform along the flow direction is $\tau_F=1/\gd$.
 
@@
 underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
-The view is in x-direction. This means the velocity ${\mathbf v}$ 
+The view is along the x-direction. This means the velocity ${\mathbf v}$ 
 has been projected onto a plane perpendicular to the flow direction. 
@@
 The snapshot shows a typical frame during a periodically recurring sequence.
-The network on the left with positive helicity travels rightwards, whereas the one one the right
-moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network to reconnect in flow direction.}
+The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
+with reversed helicity, moves leftward. 
+The periodicity of the motion along the vorticity direction is six times longer than the time 
+it takes the network to reconnect along the flow direction.}
 \label{bp2-velo}
@@
 
-The general appearance of the flowing network is apart from the affine transformation 
-very close to that of the quiescent blue phase at equilibrium. It is the same for 
-all shear rates in regime BPII-1 and a first explanation why shear-thinning is so weak.
+The general appearance of the flowing network is, apart from the affine transformation, 
+very close to that of the quiescent blue phase at equilibrium. This holds for all
+shear rates in the BPII-1 regime.
 
 Interestingly, while being displaced with the flow the entire network moves 
-in the vorticity direction.  A similar behaviour has been recently 
-observed for blue phases in shear flow in 
-confined geometries ~\cite{Henrich:2012b}.
-In contrast to the stick-slip motion that has been reported there the movement is 
-steady. It occurs in such a manner that the positions of breakup and reconnection 
-of the network, visible in Fig. \ref{bp2-med}, are slightly offset and make the network 
-travel along the z-direction.
-The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
-i.e. it takes a displacement of six unit cells in flow directions 
-until the network has moved one unit cell in vorticity direction.
-
-Fig. \ref{bp2-velo} shows the disclination network between reconnecting 
-and the next breakup with superimposed velocity vectors.
-Shown are what we may call secondary velocity components, obtained by projecting the
+along the vorticity direction as well.  A similar behaviour has been recently 
+observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
+In contrast to the stick-slip motion that has been reported there,
+in this case the movement is steady. 
+The vorticity motion occurs in such a way that the positions of breakup and reconnection 
+point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
+to travel along the z-direction. The periodicity of the motion along the vorticity direction
+is $\tau_V=6\tau_F$, 
+i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
+for the network to move one unit cell along $z$ (the vorticity direction).
+
+Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
+cycle, with superimposed velocity vectors.
+The plot shows the secondary velocity components, obtained by projecting the
 velocity onto a plane perpendicular to the flow direction.
-This gets rid of the dominating velocity component $v_x$ and 
-allows to visualise the two much smaller components $v_y$ and $v_z$.
+This gets rid of the dominating velocity component along the flow, $v_x$, and 
+allows to visualise the patterns in the 
+two much smaller components $v_y$ and $v_z$.
 The magnitude of the secondary components is typically in the range of a 
-few percent of the primary flow component.
+few percent of the primary flow component, for the shear rates simulated here.
 
 Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
-oriented along the vorticity direction. They are obscured by the
-primary flow component if the above described projection procedure
-is not applied.
-When the helicity of the underlying cholesteric phase is changed from 
-left-handed to right-handed, 
-the pattern becomes a mirror image of the former one. The sign of 
-$v_z$ changes and the sense of motion is inverted.
+oriented along the vorticity direction. The direction of the flow
+in each of the bands depends on the helicity of the underlying cholesteric
+liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
+patterns for left-handed and right-handed BPs are one the mirror image of
+the other. 
 Further quantitative evidence for a direct link between the sense of motion 
@@
 There is only a slight imbalance between the maximum and minimum velocities 
-in z-direction, suggesting that the movement of the network in vorticity 
-direction is permeative. Note that this imbalance is retained under inversion of
-helicity.
+in z-direction, suggesting that the flow of the network along the vorticity 
+direction is permeative.
 
@@
 The configuration is a helical wave that travels along the direction of vorticity.
-It is translationally invariant in gradient and flow direction.}
+It is translationally invariant along the gradient and flow direction.}
 \label{bp2-high}
@@
 flow-aligned nematic state at shear rates $7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$, 
-which has so far not been reported to our knowledge \com{CHECK}.
-We will address this in section \ref{cholflow}.
+where the helical axis is now along the flow gradient direction (the so-called
+Grandjean texture).
+We will address this in section \ref{cholflow}, as it is common to BPI and BPII.
 
@@
 
-BPI has body-centered symmetry and unlike BPII. The disclination lines
- are well separated and do not intersect.
+BPI has body-centered symmetry and, unlike BPII, the disclination lines
+characterising its equilibrium structure are well separated and do not intersect.
 We believe that this topological difference is responsible for most of
 the differences between BPI and BPII regarding their rheological response. 
-We present again key aspects in an overview before we 
-address some features in more detail.
+We present again key aspects in an overview before we address specific features 
+in more detail.
 
 Fig. \ref{bp1-rheo} shows the apparent viscosity $\eta_{app}$ versus time and 
-strain for the same shear rates as in Fig. \ref{bp2-rheo}.
-For shear rates $9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ we observe the same cholesteric 
+strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
+shows that there is more shear thinning in BPI than in BPII, for
+the shear rate range we considered. 
+Throughout the flow range we studies,
+the average shear stress versus shear rate can be fitted reasonably 
+well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
+
+For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
+response versus time, and in the network morphology -- we label this 
+regime BPI-2. 
+The BPI-2 regime now does not persist to arbitrarily low shear: there the
+oscillations become irregular and seemingly chaotic, and after a few
+cycles the disclination pattern becomes amorphous. We refer to this 
+flow regime as BPI-1. 
+The BPI-2 regime is also unstable if the shear is increased {\em above} 
+a critical value -- close to this breakup point the oscillations in the 
+deviatoric stress show an increasingly complex
+pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
+In regime BPI-3 $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ 
+we observe that the BPI network transforms under shear into
+a  quasi two-dimensional state consisting of
+double-twist rolls.
+Finally, for shear rates $9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ we observe the same cholesteric 
 configuration with the helical axis along the gradient direction as for BPII (see
 section \ref{cholflow}).
-
-In regime BPI-3 $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ a breakup of 
-the blue phase disclination network occurs. This is about the same critical shear rate 
-where we observe the transition to BPII-2. However, for BPI the configuration is not
-a travelling helical wave, but rather a quasi two-dimensional state consisting of
-double-twist rolls.
-
-At shear rates $3.91\e{-5}\lesssim\gd\lesssim4.68\e{-4}$ we find flow with 
-periodically recurring conformations. We refer to this regime as BPI-2. 
-The oscillations in the 
-deviatoric stress that are not sinusoidal, but show an increasingly complex
-pattern up to the breakup point.
-For even lower shear rates $\gd\lesssim3.91\e{-5}$ in regime BPI-1 
-these oscillations become very erratic and unstable. 
-There is no regular recurrence and 
-after a few cycles the stress pattern is completely irregular.
-Finally, the system ends up in an amorphous configuration.
-Shear-thinning is slightly stronger than in BPII.
-The average shear stress during one cycle can be fitted reasonably 
-well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
 
@@
 significantly larger. 
-More surprisingly the frequency of the fundamental mode is the same 
+Second, more surprisingly, the frequency of the fundamental mode is the same 
 for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
@@
 $\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
-moves to the double frequency. This means that the fundamental mode at flowrate $\gd=7.81\e{-5}$ 
-coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on,
-which is the expected behaviour. At higher shear rates in regime BPI-3 this 'lock-in' of 
-flow modes disappears again.
+moves to double the frequency. This means that the fundamental mode at flow rate $\gd=7.81\e{-5}$ 
+coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on. 
+At higher shear rates in regime BPI-3 this 'lock-in' of flow modes disappears again.
 
@@
 visible in the fixed peak position for certain flow rates.  
-The inset shows the same data on an expanded scale. 
-The frequency is given in radians.}
+The inset shows a zoom of the same data over 
+a portion of the frequency range.
+The frequency is given in inverse time, in simulation units. TRUE?}
 \label{bp1-spectrum}
@@
 minimum in average free energy density. This creates unstable conditions that lead 
-to a dissolving regular network and an amorphous state which features a yield stress.}
+to a dissolving regular network and an amorphous state which 
+seemingly features a yield stress.}
 \label{bp1-fe-yield}
@@
 The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
-is strikingly different from that of BPII and exhibits a
-dynamic yield stress.
-An explanation for this strange behaviour can be found by 
-looking more closely at the average deviatoric stress and 
-free energy density, as shown in Fig. \ref{bp1-fe-yield}.
+is strikingly different from that of BPII and appears to show a
+yield stress. An explanation for this behaviour can be found by 
+looking more closely at the average deviatoric stress (where the
+background viscosity contribution has been subtracted) and 
+free energy density as a function of time, as shown in Fig. \ref{bp1-fe-yield}.
 When the quiescent and equilibrated BPI network begins to flow
-the shear stress increases steeply and goes through a global maximum.
-Shortly after it goes negative (the total shear stress is still
-positive as the background viscosity has been subtracted) and then positive
-again, forming a complete cycle. 
-This cycle is repeated for a couple of times before 
-any regular pattern disappears and the amorphous configuration 
-emerges.
+the shear stress increases steeply and goes through a maximum.
+Shortly after it goes negative (the total shear stress 
+with the background viscosity contribution remains positive) 
+to become positive again later on, forming a complete cycle.
+The amplitude of the excursions in these stress oscillations and
+the presence of a part of the period where the thermodynamic
+contribution to the stress is negative suggest that the
+BPI network is subject to large forces which eventually 
+cause the flow-induced collapse to the amorphous state with
+an apparent yield stress. We should underscore that this,
+low shear, regime, might be brought about by the boundary
+conditions, which in practice lead to the network sticking to
+the Lees-Edwards plane. With walls and free boundary conditions,
+one may imagine that, for very low wall velocity, 
+permeative flows may be possible which leave the BPI 
+structure almost unchanged, as in cholesterics sheared
+along their axis~\cite{Marenduzzo:2006b}.
 
@@
 
-The negative branch in the deviatoric stress is connected
-to a local maximum followed by a local minimum in the average free energy.
-Hence, the region of negative stress can be 
+\if{The negative branch in the deviatoric stress is connected
+to a local maximum followed by a local minimum in the 
+average free energy. Hence, the region of negative stress can be 
 regarded as a phase during which elastic forces exert a 
@@
 create small distortions which finally break the symmetry of the
-flow-induced configuration.
+flow-induced configuration.}\fi
 
@@
 
-We want to explain these differences on the basis of a typical 
-appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
+To gain further insight into the flow-induced deformation, and eventual
+breakdown, of BPI, it is instructive to follow the dynamics of 
+the network under a slow flow -- this is depicted in Fig. \ref{bp1-low}.
 Three different stages can be distinguished. 
-With the onset of shearing the disclination lines 
-in BPI get more and more squeezed together which is incommensurable with 
-the defect topology of this blue phase. 
+Just after the onset of the shear flow, the disclination lines 
+in BPI get more and more squeezed together, and this is incompatible with
+the defect topology in this phase.
 Consequently, the network adopts a flow-induced conformation which consists 
@@
 is shown in Fig. \ref{bp1-low} (top row).
-At a particular point in the cycle a non-helical configuration is
-observed (bottom left) before the network quickly returns
-to its helical state. There is no perceivable movement of the 
-network in vorticity direction at any stage of the cycle.
-
-However, this mode of flow proves unstable as after a few cycles 
-distortions appear which lead to further destabilisation.
-Finally, the network loses any residual order and 
+At this point the original BPI has already been significantly deformed.
+Shortly afterwards regular oscillations set up temporarily,
+where the disclinations form double helices which tilt and realign under the shear
+TRUE? There is no perceivable movement of the 
+network along the vorticity direction at any stage of the cycle.
+
+This mode of flow proves unstable as after a few cycles 
+distortions appear which lead to further destabilisation -- presumably in
+view of the large stress fluctuations discussed above. 
+Finally, the network loses order and 
 transforms into an amorphous state with almost constant
 yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
+If the shear is stopped, this flow-induced amorphous state
+gets stuck into a metastable phase, and cannot
+find its way back to the original BPI structure.
 
@@
 We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
-is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress versus total strain. 
-A qualitative difference between these two is the absence of 
-large stress fluctuations that occur during one typical cycle in regime BPI-1.
-Remember that for even lower $\gd$ the deviatoric stress becomes 
+is clear when looking at Fig. \ref{bp1-1_bp1-2} 
+which shows the deviatoric stress versus total strain. 
+A qualitative difference between these two is the absence in BPI-2 of the 
+large stress fluctuations that occur during the early cycles in regime BPI-1.
+[Recall that for even lower $\gd$ the deviatoric thermodynamic stress becomes 
 temporarily negative (Fig. \ref{bp1-fe-yield}), which 
 caused destabilisation of the periodic network and led to the 
-amorphous configuration in the steady state. Hence, an explanation 
-for this phenomenon could be that in BPI-2 the network finds a 
-kinetic path to bypass these large amplitude fluctuations that 
-prevents the network from destabilising.
+amorphous configuration in the steady state.] Hence, an explanation
+for the existence of the regular BPI-2 oscillations 
+could be that these are fast enough to bypass or suppress 
+large stress fluctuations -- this leads to
+a stabilisation of the oscillations found only transiently in
+the BPI-1 regime.
 
 Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
-BPI in steady shear flow. Contrary to BPII-1 at these shear rates
-, BPI-2 does not resemble the equilibrium configuration
+BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
+BPI-2 does not resemble the equilibrium configuration
 undergoing an affine transformation. It features rather intricate, flow-induced 
-conformations. Surprisingly this state is kinetically connected to the equilibrium 
-state as when the shear flow is switched off the flow-induced configuration 
-reverts to a perfect quiescent BPI unless it got stuck in a local free energy 
-minimum.
-
-The recurrence period of conformations with visually similar appearance is 
-less straightforward to determine. In the flow direction relation like
+conformations. Surprisingly, if the shear is switched off in
+the BPI-2 regime at some point during an oscillations, 
+most of the times the flow-induced configuration 
+reverts to a perfect quiescent BPI -- in stark contrast with
+what happens in BPI-1, for lower shear rates.
+
+The period after which network conformations recur is not
+easy to work out for BPI. Along the flow direction a relation of the form
 $\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
@@
 in the vorticity direction. The direction of the displacement
-in other directions (i.e. both in gradient and vorticity direction) 
-is not simply along coordinate directions. 
-The magnitude of the displacement show no clear pattern, is non-integer 
-and depends sensitively on the imposed shear rate,
-suggesting BPI-2 constitutes an instance of 'rheochaos' [ref].
+perpendicular to the flow direction is not along a coordinate
+axis. This direction, and the magnitude of the secondary
+network movement depends sensitively on the imposed shear rate. 
+This suggests that BPI-2 constitutes an instance of 
+'rheochaos'~\cite{rheochaos}.
 
@@
 $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
-During every cycle the network is displaced in gradient and vorticity direction, 
-but the magnitude and direction of the offset appear to be chaotic and do not 
-obey a simple rule.}
+During every cycle the network is displaced also along the gradient and vorticity direction, 
+but the magnitude and direction of the offset does not appear to obey a 
+simple rule.}
 \label{bp1-med}
@@
 Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ for left- and right-handed cholesteric
-helix. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in that there is no
-arrangement in bands. However, just as in case of BPII, both patterns
+components $v_y$ and $v_z$ for left- and right-handed chirality of
+the blue phase. The emerging pattern is similar to that of BPII from 
+Fig. \ref{bp2-velo}, but differs in that there is no band
+arrangement. However, just as in the case of BPII, both patterns
 are mirror images and feature the highest velocities in the regions where 
@@
 The intricacy and periodic recurrence of the flow-induced conformations 
-in BPI raises the question: how robust these results are with respect to 
+in BPI raises the question: how robust are these results to variations in 
 system size? Although a pitch length of 64 LBU provides enough resolution 
@@
 unit cells and the other one with 64 unit cells.
-Obviously there are large deviations in both the visual appearance of the 
+We observe that there are deviations in both the visual appearance of the 
 network as well as in the time-dependence of the deviatoric stress.
-In the smaller run with $2^3=8$ unit cells takes BPI also considerably 
-to reach the 
-steady state with its flow-induced recurring disclination network.
-
+Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
+to reach the steady state with its flow-induced recurring disclination network. 
 These results indicate that we cannot know with certainty if the individual 
 flow-induced conformations that we find for 64 unit cells are not affected 
-by imposed periodicity. Nevertheless, the comparison suggests that 
+by the imposed periodicity -- quite possibly the details of the oscillations
+would actually differ in an even large simulation. 
+Nevertheless, the comparison suggests that 
 the recurrence period and the complexity of the network in flow are  
@@
 central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
-The bottom picture shows the deviatoric stress over time. Interestingly, 
+The bottom picture shows the deviatoric stress over time. [Interestingly, 
 an intermediate sized run with $3^3=27$ unit cells led to results that 
-were identical to the smaller run with 8 unit cells. The shear rate was $\gd=7.81\e{-5}$.
+were virtually identical to the smaller run with 8 unit cells.] 
+The shear rate was $\gd=7.81\e{-5}$.
 }
@@
 The colour code gives the magnitude of the z-component of the director field. 
-The data is translationally invariant in vorticity direction.}
+The data is translationally invariant along the vorticity direction.}
 \label{bp1-high}
@@
 
-At high flow rates $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ 
+At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$, 
 we observe a break-up of the flow-induced 
@@
 in contrast to BPII, where the configuration is a travelling helical 
-wave (viz. Fig. \ref{bp2-high}), BPI prefers a quasi 
+wave (viz. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
 two-dimensional, translationally invariant state, which is shown in 
@@
 independent of their initial state.
-The director field of this configuration is shown in Fig. \ref{cholflow}.
+The director field of this configuration is shown in Fig. \ref{fig:cholflow}.
 It consists of a simple cholesteric helix with the helical axis oriented 
-along the gradient direction. The state is translational invariant in 
-flow and vorticity direction and therefore also one-dimensional. 
+along the velocity gradient direction. The state is translational invariant 
+along the flow and vorticity direction and therefore also one-dimensional. 
 However, in contrast to the travelling helical wave of BPII-2 the 
@@
 the flow, which occurs in 'nematic layers'. For clarity the picture shows 
-the director field along the y-direction for selected sites in the x-z-plane.}
-\label{cholflow}
+the director field along the y-direction for selected sites on the x-z-plane.}
+\label{fig:cholflow}
 \end{figure}
@@
 of cubic BPs in simple shear flow. Although experimental evidence to support
-our results is currently not available, we believe this work can shed some light into the 
-flow properties of complex liquid-crystalline phases.
+our results is currently not available, we believe this work can shed some light 
+into the flow properties of complex liquid-crystalline phases.
 
@@
 in the flow, which leads to a periodically recurring dependence 
-of the deviatoric shear stress. The flow-induced conformation look generally
+of the deviatoric shear stress. The flow-induced conformation looks generally
 very similar to the quiescent network at equilibrium. 
@@
 At larger Ericksen numbers $4\lesssim Er\lesssim 10$ the flowing BPII network 
-breaks up in a two-step scenario. First a simple cholesteric helix forms with the helical axis 
-in vorticity direction. The flow is in vorticity mode, i.e. as a travelling helical wave. 
-At slightly larger shear rates another cholesteric conformation emerges.
-However, this time the orientation of the helix is in gradient direction.
-The flow is in quasi-nematic layers with static orientation of the directior field.
+breaks up in a two-step manner. First a simple cholesteric helix forms with the helical axis 
+along the vorticity direction. The flow excites a travelling helical wave,
+as predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}.
+Upon increasing the shear, the helix rotates and this time orients along the
+gradient direction, where it can support larger flows due to the translational
+symmetry along the flow direction. Still larger flows break this residual
+cholesteric order and leave a standard flow-aligned nematic staete.
 
@@
 This is a direct consequence of the topological differences in both disclination 
-network. Below Ericksen numbers $Er<0.4$ the BPI network cannot flow and 
-retain a regular appearance. It dissolves eventually into an amorphous network and 
+network. Below Ericksen numbers $Er<0.4$ the BPI network cannot both flow and 
+retain a regular appearance. It eventually collapses into an amorphous network and 
 features yield-stress behaviour. The reason is that shortly after the onset of the shearing 
 very large stress fluctuations occur. At the lowest shear rates
-the contribution of the network to the total deviatoric stress is temporarily 
-actually negative during one cycle. These fluctuations destabilise the network and 
-eventually trigger its dissolution into an amorphous state with 
-dynamic yield stress.
-
-At slightly larger Ericksen numbers $0.4<{\it Er}<8$ shear thinning is 
-slightly stronger than for BPII, but BPI does not show simple power-law 
-behaviour. 
-Instead, it appears BPI kinematically bypasses the large stress fluctuations and 
+the thermodynamic contribution to the deviatoric (shear) stress is temporarily 
+negative during one cycle. These fluctuations destabilise the network and 
+eventually trigger the transition into an amorphous state with a residual yield
+stress.
+
+At slightly larger Ericksen numbers $0.4<{\it Er}<8$,
+BPI kinematically bypasses the large stress fluctuations and 
 flows with very complex, periodically recurring flow-induced conformations.
-The offset of the network after each cycle is random in magnitude and direction.
-However, despite their complex appearance the flow-induced conformations are 
-topologically connected to the quiescent BPI and switching off of the 
-shear flow often restores a perfect equilibrium.
-It is possible to interpret these findings as an instance of 
-deterministic rheochaos.
-
-Just as in case of BPII, the BPI network breaks up at larger Ericksen numbers in
-a two-step process. First at $10\lesssim\gd\lesssim 15$, it adopts a 
-quasi-two-dimensional configuration that is translationally invariant in vorticity direction. 
-Then at $16\lesssim Er \lesssim 22$ it forms cholestric helix along the gradient 
-direction that emerges from BPII.  
-Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
+These conformations entail regularly arrange helical disclinations which 
+rotate due to the shear -- a new flow-induced state which was not previously
+predicted. 
+The offset of the network after each oscillation is along a direction which,
+while being a multiple of the unit cell size along all three coordinate
+axis, depends on system size and shear rate in a seemingly chaotic fashion.
+Despite their complex appearance, the flow-induced conformations are 
+topologically connected to the quiescent BPI and after switching off of the 
+shear flow a defect-free blue phase reforms. It is tempting to
+interpret all these finding as an instance of 
+deterministic rheological chaos.
+
+Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
+a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
+quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
+Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
+direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
 is a flow-aligned nematic state.

### 2012-11-13 papers/bprheo/bprheo.tex
Version for Mike
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1813 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \author{O. Henrich}
-\email[Corresponding author: ]{o.henrich@ucl.ac.uk}
+\email[Corresponding author: ]{ohenrich@epcc.ed.ac.uk}
 \affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
@@
 
-It is possible to define a dimensionless number that describes the deformation
-of the director field under flow. The so-called Ericksen number
-is defined as 
+It is possible to define a further 
+dimensionless number that describes the deformation
+of the director field under flow. This so-called Ericksen number
+is given by 
 \begin{equation}
-Er=\frac{\eta V L}{K}
+Er=\frac{\eta v l}{K}
 \end{equation}
 with $\eta$ and $K$ being the viscosity and elastic constant and
-$V$ and $L$ as typical velocity and length scale.
+$v$ and $l$ as typical velocity and length scale. 
+In the present work $l=p_0/2$ was assumed to be the size of the BP unit cell,
+whereas $v$ was taken to be the velocity difference
+across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.
 
@@
 \begin{table*}[htpb]
-\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
-\hline
-$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
+\hline
+& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
 \hline
@@
 \hline
-0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
-3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
-\hline
+BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+\hline
+BPI-2 & 3.91 & 0.67 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
+BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+\hline
+BPI-3 & 62.50 & 10.67& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
+BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+\hline
+CHG & 93.75 &16.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
+CHG & 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+\hline
+FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+\hline
+
 BPII \\
 \hline
-0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
-62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+\hline
+BPII-2 & 46.87 &4.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
+BPII-2 & 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+\hline
+CHG & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
+CHG & 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
+\hline
+FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
 \hline
 \end{tabular}
-\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minimum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding any transient dynamics.}
+\caption{Minima, maxima and standard deviation of time-averaged velocity components 
+for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
+shear flow: All values are given in $10^{-5}$ lattice units. The transient dynamics 
+has not been included in the averaging procedure. The regimes comprise the formation
+of an amorphous network (BPI-1), periodically recurring conformations (BPI-2 and BPII-1), 
+two-dimensional double twist rolls (BPI-3), a one-dimensional travelling helical wave (BPII-2), 
+a cholesteric helix oriented along the gradient direction (CHG), 
+and a flow-aligned nematic state (FAN).}
 \label{tab1}
@@
 The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
-For clarity we classify various sub-regimes, two in case of BPII and three in case of BPI. 
-They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
-BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
-Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
-are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
+$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ in LB units.
+For clarity we classify various flow regimes, two in case of BPII and three in case of BPI. 
+They are BPII-1 ($\gd \lesssim 3.125\e{-4}$), BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$),
+BPI-1 ($\gd \lesssim 1.95\e{-5}$), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$) and BPI-3 ($6.25\e{-4}\lesssim gd\lesssim 7.813\e{-4}$).
+Beyond these regimes both BPI and BPII form common configurations regardless 
+of the initial state, namely a cholesteric helix with helical axis 
+oriented along the direction of the velocity gradient (CHG), 
+and a flow-aligned nematic state (FAN).  
+Disclination lines are plotted as isosurface of the scalar order parameter $q$. 
+Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
 
@@
 $a=0.35, b=0.95$ describes the data to a very good approximation. 
-Hence, the degree of shear-thinning is remarkably small, but may depend
-on the exact direction in which the strain is applied, an aspect that we     
-leave for future work.
-
-For shear rates $\gd\ge6.25\e{-4}$ the network of disclination lines
-breaks up and oscillations in the stress signal are absent.
+Hence, the degree of shear-thinning is remarkably small in BPII, in
+the range of shear rates which we have explored here.
+
+For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ the network of disclination lines
+breaks up completely, and oscillations in the stress signal are absent.
 In this regime of shear rates, to which we refer as BPII-2, the 
 blue phase dissolves into a simple cholesteric liquid 
-crystal. The mode of flow is a travelling helical wave with the helical axis 
-oriented along the vorticity direction.
-
-In the next paragraph we will investigate some of these aspects in more detail 
+crystal. The helical axis is along the vorticity direction and
+the flow excites travelling helical waves, such as those 
+predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis.
+
+At shear rates $7.81\e{-4}\lesssim\gd \lesssim 9.37\e{-4}$ the helix changes orientation.
+The helical axis is now oriented along the gradient direction with
+the liquid crystal flowing in 'nematic planes'. For even higher shear rates
+the system undergoes a transition to a flow-aligned nematic state.
+
+In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
 by looking at the kinetics of the disclination network.
@@
 
+We observe another configuration between the travelling helical wave and the 
+flow-aligned nematic state at shear rates $7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$, 
+where the helical axis is now along the flow gradient direction (the so-called
+Grandjean texture).
+We will address this in section \ref{cholflow}, as it is common to BPI and BPII.
+
 \subsection{Blue Phase I}
 
-BPI has body-centered symmetry and unlike BPII. The disclination lines
- are well separated and do not intersect.
+BPI has body-centered symmetry and, unlike BPII, the disclination lines
+characterising its equilibrium structure are well separated and do not intersect.
 We believe that this topological difference is responsible for most of
 the differences between BPI and BPII regarding their rheological response. 
-
-We present again key aspects in an overview before we 
-address some features in more detail. Fig. \ref{bp1-rheo} shows the 
-apparent viscosity $\eta_{app}$ versus time and strain for the same shear rates
-as in Fig. \ref{bp2-rheo}.
-For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
-the disclination network occurs. This is the same critical shear rate as 
-the one observed in BPII-2. But this time the configuration is quasi 
-two-dimensional.
-
-In regime BPI-2 at lower shear rates periodic oscillations emerge in the 
-deviatoric stress that are not sinusoidal, but show an increasingly complex
-pattern up to the breakup point.
-For even lower shear rates $\gd\le3.91\e{-5}$ in regime BPI-1 
-these oscillations become very erratic. There is no regular recurrence and 
-after a few cycles the stress pattern is completely irregular. 
-Shear-thinning is slightly stronger than in BPII. 
-The average shear stress during one cycle can be fitted reasonably 
+We present again key aspects in an overview before we address specific features 
+in more detail.
+
+Fig. \ref{bp1-rheo} shows the apparent viscosity $\eta_{app}$ versus time and 
+strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
+shows that there is more shear thinning in BPI than in BPII, for
+the shear rate range we considered. 
+Throughout the flow range we studies,
+the average shear stress versus shear rate can be fitted reasonably 
 well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
+
+For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
+response versus time, and in the network morphology -- we label this 
+regime BPI-2. 
+The BPI-2 regime now does not persist to arbitrarily low shear: there the
+oscillations become irregular and seemingly chaotic, and after a few
+cycles the disclination pattern becomes amorphous. We refer to this 
+flow regime as BPI-1. 
+The BPI-2 regime is also unstable if the shear is increased {\em above} 
+a critical value -- close to this breakup point the oscillations in the 
+deviatoric stress show an increasingly complex
+pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
+In regime BPI-3 $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ 
+we observe that the BPI network transforms under shear into
+a  quasi two-dimensional state consisting of
+double-twist rolls.
+Finally, for shear rates $9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ we observe the same cholesteric 
+configuration with the helical axis along the gradient direction as for BPII (see
+section \ref{cholflow}).
 
@@
 
-In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
-we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
-The spectrum is shown in Fig. \ref{bp1-spectrum}.
-
-There are two aspects which are worth mentioning. First, we observe a strong contribution of 
-the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
-However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
-significantly larger. 
-More surprisingly the frequency of the fundamental mode is the same 
-for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
-in contrast to what is observed at lower flow rates. By doubling the flow rate from 
-$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
-moves to the double frequency. This means that the fundamental mode at flowrate $\gd=7.81\e{-5}$ 
-coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on,
-which is the expected behaviour. At higher shear rates in regime BPI-3 this 'lock-in' of 
-flow modes disappears again.
-
-Interestingly the lock-in has only a minor effect on the flow velocities themselves.
-The average velocities in Tab. \ref{tab1} reveal that only the magnitude of some of the secondary 
-flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
-The latter is directly linked to the imposed shear rate.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
-spectral components and the 'lock-in' of flow modes, 
-visible in the fixed peak position for certain flow rates.  
-The inset shows the same data on an expanded scale. 
-The frequency is given in radians.}
-\label{bp1-spectrum}
-\end{figure}
-
 \subsubsection{Regime BPI-1: low shear rates}
@@
 
-The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
-is strikingly different from that of BPII and exhibits a
-dynamic yield stress.
-An explanation for this strange behaviour can be found by 
-looking more closely at the average deviatoric stress and 
-free energy density, as shown in Fig. \ref{bp1-fe-yield}.
+The rheological response of BPI at low shear rate $\gd\lesssim1.95\e{-5}$ 
+is strikingly different from that of BPII and appears to show a
+yield stress. An explanation for this behaviour can be found by 
+looking more closely at the average deviatoric stress (where the
+background viscosity contribution has been subtracted) and 
+free energy density as a function of time, as shown in Fig. \ref{bp1-fe-yield}.
 When the quiescent and equilibrated BPI network begins to flow
-the shear stress increases steeply and goes through a global maximum.
-Shortly after it goes negative (the total shear stress is still
-positive as the background viscosity has been subtracted) and then positive
-again, forming a complete cycle. 
-This cycle is repeated for a couple of times before 
-any regular pattern disappears and the amorphous configuration 
-emerges.
+the shear stress increases steeply and goes through a maximum.
+Shortly after it goes negative (the total shear stress 
+with the background viscosity contribution remains positive) 
+to become positive again later on, forming a complete cycle.
+The amplitude of the excursions in these stress oscillations and
+the presence of a part of the period where the thermodynamic
+contribution to the stress is negative suggest that the
+BPI network is subject to large forces which eventually 
+cause the flow-induced collapse to the amorphous state with
+an apparent yield stress. We should underscore that this,
+low shear, regime, might be brought about by the boundary
+conditions, which in practice lead to the network sticking to
+the Lees-Edwards plane. With walls and free boundary conditions,
+one may imagine that, for very low wall velocity, 
+permeative flows may be possible which leave the BPI 
+structure almost unchanged, as in cholesterics sheared
+along their axis~\cite{Marenduzzo:2006b}.
 
 \begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
-\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
-\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115.png}\\
+\includegraphics[width=0.245\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-700k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2200k_run1115.png}
 \caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
-Depicted is the transition from flow-induced, intertwined helices that 
-undergo a recurring affine transformation (top left and centre, both at 
-time step $t=4.0\e{5}$; top right at time step $t=7.0\e{5}$) with an intermediate 
-conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) 
-to a distorted state (bottom centre at time step $t=1.7\e{6}$) and 
-finally to an amorphous network (bottom right at $t=2.2\e{6}$). 
-The picture at the top centre shows the network in gradient direction.
-All other pictures have the view along the vorticity direction.}
+Depicted is the transition from the quiescent state to flow-induced, intertwined 
+helices that undergo a recurring affine transformation. The top row
+shows a section of one unit cell for early times 
+$t=1\e{4}, 1.8\e{5}, 2.0\e{5}$ and $2.1\e{5}$. The bottom 
+pictures on the far left, centre left and centre right 
+show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
+which the system passes for several cycles before it ends up in
+an amorphous state (bottom row far right at time step $t=2.2\e{6}$).}
 \label{bp1-low}
@@
 is shown in Fig. \ref{bp1-low} (top row).
-At a particular point in the cycle a non-helical configuration is
-observed (bottom left) before the network quickly returns
-to its helical state. There is no perceivable movement of the 
-network in vorticity direction at any stage of the cycle.
-
-However, this mode of flow proves unstable as after a few cycles 
-distortions appear which lead to further destabilisation.
-Finally, the network loses any residual order and 
+At this point the original BPI has already been significantly deformed.
+Shortly afterwards regular oscillations set up temporarily,
+where the disclinations form double helices which tilt and realign under the shear.
+There is no perceivable movement of the 
+network along the vorticity direction at any stage of the cycle.
+
+This mode of flow proves unstable as after a few cycles 
+distortions appear which lead to further destabilisation -- presumably in
+view of the large stress fluctuations discussed above. 
+Finally, the network loses order and 
 transforms into an amorphous state with almost constant
 yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
+If the shear is stopped, this flow-induced amorphous state
+gets stuck into a metastable phase, and cannot
+find its way back to the original BPI structure.
 
@@
 
-Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) and 
-at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
-and Ericksen numbers $0.5\le{\it Er}<10$ 
+Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$) and 
+at slightly larger shear rates $3.91\e{-5}\lesssim\gd\lesssim 4.687\e{-4}$
+and Ericksen numbers $0.67\lesssim{\it Er}\lesssim8$ 
 lies another region where the network flows with periodically 
@@
 We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
-is shown in Fig. \ref{bp1-1_bp1-2} by means of the deviatoric stress versus total strain. 
-A qualitative difference between these two is the absence of 
-large stress fluctuations that occur during one typical cycle in regime BPI-1.
-Remember that for even lower $\gd$ the deviatoric stress becomes 
+is clear when looking at Fig. \ref{bp1-1_bp1-2} 
+which shows the deviatoric stress versus total strain. 
+A qualitative difference between these two is the absence in BPI-2 of the 
+large stress fluctuations that occur during the early cycles in regime BPI-1.
+[Recall that for even lower $\gd$ the deviatoric thermodynamic stress becomes 
 temporarily negative (Fig. \ref{bp1-fe-yield}), which 
 caused destabilisation of the periodic network and led to the 
-amorphous configuration in the steady state. Hence, an explanation 
-for this phenomenon could be that in BPI-2 the network finds a 
-kinetic path to bypass these large amplitude fluctuations that 
-prevents the network from destabilising.
-
-Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
-BPI in steady shear flow. Contrary to BPII-1 at these shear rates
-, BPI-2 does not resemble the equilibrium configuration
-undergoing an affine transformation. It features rather intricate, flow-induced 
-conformations. Surprisingly this state is kinetically connected to the equilibrium 
-state as when the shear flow is switched off the flow-induced configuration 
-reverts to a perfect quiescent BPI unless it got stuck in a local free energy 
-minimum.
-
-The recurrence period of conformations with visually similar appearance is 
-less straightforward to determine. In the flow direction relation like
-$\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
-7.81\e{-5},1.563\e{-4}$ for example. The analysis is complicated by the fact
-that for $n>1$ thermodynamically equivalent, mirror-image 
-conformations arise after shorter times. For instance at $\gd=7.81\e{-5}$
-the conformations at times $t$ and $t+1/\gd$ are mirror-images, whereas
-those at $t$ and $t+2/\gd$ coalesce in flow direction.
-
-However, there is no definite recurrence period $\tau_V$ 
-in the vorticity direction. The direction of the displacement
-in other directions (i.e. both in gradient and vorticity direction) 
-is not simply along coordinate directions. 
-The magnitude of the displacement show no clear pattern, is non-integer 
-and depends sensitively on the imposed shear rate,
-suggesting BPI-2 constitutes an instance of 'rheochaos' [ref].
+amorphous configuration in the steady state.] Hence, an explanation
+for the existence of the regular BPI-2 oscillations 
+could be that these are fast enough to bypass or suppress 
+large stress fluctuations -- this leads to
+a different order structure and to a stabilisation 
+compared to the oscillations found transiently in
+the BPI-1 regime.
 
@@
 
+Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
+BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
+BPI-2 does not resemble the equilibrium configuration
+undergoing an affine transformation. It features rather intricate, flow-induced 
+conformations. Surprisingly, if the shear is switched off in
+the BPI-2 regime at some point during an oscillations, 
+most of the times the flow-induced configuration 
+reverts to a perfect quiescent BPI -- in stark contrast with
+what happens in BPI-1, for lower shear rates.
+
+The period after which network conformations recur is not
+easy to work out for BPI. Along the flow direction a relation of the form
+$\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
+7.81\e{-5},1.563\e{-4}$ for example. The analysis is complicated by the fact
+that for $n>1$ thermodynamically equivalent, mirror-image 
+conformations arise after shorter times. For instance at $\gd=7.81\e{-5}$
+the conformations at times $t$ and $t+1/\gd$ are mirror-images, whereas
+those at $t$ and $t+2/\gd$ coalesce in flow direction.
+
+However, there is no definite recurrence period $\tau_V$ 
+in the vorticity direction. The direction of the displacement
+perpendicular to the flow direction is not along a coordinate
+axis. This direction, and the magnitude of the secondary
+network movement depends sensitively on the imposed shear rate. 
+This suggests that BPI-2 constitutes an instance of 
+'rheochaos'~\cite{rheochaos}.
+
 Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ for left- and right-handed cholesteric
-helix. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in that there is no
-arrangement in bands. However, just as in case of BPII, both patterns
+components $v_y$ and $v_z$ for left- and right-handed chirality of
+the blue phase. The emerging pattern is similar to that of BPII from 
+Fig. \ref{bp2-velo}, but differs in that there is no band
+arrangement. However, just as in the case of BPII, both patterns
 are mirror images and feature the highest velocities in the regions where 
@@
 
+In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
+we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
+The spectrum is shown in Fig. \ref{bp1-spectrum}.
+
+There are two aspects which are worth mentioning. First, we observe a strong contribution of 
+the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
+However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
+significantly larger. 
+Second, more surprisingly, the frequency of the fundamental mode is the same 
+for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
+in contrast to what is observed at lower flow rates. By doubling the flow rate from 
+$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
+moves to double the frequency. This means that the fundamental mode at flow rate $\gd=7.81\e{-5}$ 
+coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on. 
+At higher shear rates in regime BPI-3 this 'lock-in' of flow modes disappears again.
+
+Interestingly the lock-in has only a minor effect on the flow velocities themselves.
+The average velocities in Tab. \ref{tab1} reveal that only the magnitude of some of the secondary 
+flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
+The latter is directly linked to the imposed shear rate.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
+\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
+spectral components and the 'lock-in' of flow modes, 
+visible in the fixed peak position for certain flow rates.  
+The inset shows a zoom of the same data over 
+a portion of the frequency range.
+The frequency is given in inverse time, in simulation units. TRUE?
+OH: I WOULD SAY NO AS IT IS IN UNITS OF 2$\pi$ (RADIANS?).}
+\label{bp1-spectrum}
+\end{figure}
+
 The intricacy and periodic recurrence of the flow-induced conformations 
-in BPI raises the question: how robust these results are with respect to 
+in BPI raises the question: how robust are these results to variations in 
 system size? Although a pitch length of 64 LBU provides enough resolution 
@@
 
-\begin{figure}[htpb]
-%\includegraphics[width=0.4\textwidth]{dir+z-300k_run916.png}
-%\includegraphics[width=0.4\textwidth]{dir+z-301k_run916.png}
-\includegraphics[width=0.4\textwidth]{dir3d-z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a slice 
-of $L_x\times L_y \times L_Z= 32\times32\times 32$ lattice sites along the vorticity direction 
-with the flow/gradient direction oriented along the horizontal/vertical axis. 
-The colour code gives the magnitude of the in-plane component of the director field. 
-The data is translationally invariant in vorticity direction.}
-\label{bp1-high}
-\end{figure}
-
-At high flow rates we observe a break-up of the flow-induced 
+At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$, 
+we observe a break-up of the flow-induced 
 disclination network of regime BPI-2. However, 
-contrary to BPII, where the configuration is a travelling helical 
-wave (viz. Fig. \ref{bp2-high}), BPI prefers a quasi 
+in contrast to BPII, where the configuration is a travelling helical 
+wave (viz. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
 two-dimensional, translationally invariant state, which is shown in 
@@
 regions of local double twist.
-Interestingly the critical shear rate for the break-up lies in the same 
-window as that of transition between BPII-1 and BPII-2, 
- $3.125\e{-4}\le\gd_c\le 6.25\e{-4}$ and occurs at similar 
-Ericksen numbers ${\it Er}\simeq 10$.
-
-\subsection{Transition to 1D Uniaxial Twist}
-
-\subsection{Uncoiling Transition to Nematic State}
-
-\clearpage
+Interestingly the critical Ericksen number for the break-up lies in 
+about the same window as that of transition between BPII-1 and BPII-2.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
+\caption{Director field of BPI at high shear rate: The pictures show a region 
+of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
+direction oriented along the horizontal/vertical axis. 
+The colour code gives the magnitude of the z-component of the director field. 
+The data is translationally invariant along the vorticity direction.}
+\label{bp1-high}
+\end{figure}
+
+\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{cholflow}
+
+At shear rates beyond the regimes BPI-3 and BPII-2 and below the transition to a 
+flow-aligned nematic state we found another regime where 
+both blue phases adopt the same configuration in steady shear flow, 
+independent of their initial state.
+The director field of this configuration is shown in Fig. \ref{fig:cholflow}.
+It consists of a simple cholesteric helix with the helical axis oriented 
+along the velocity gradient direction. The state is translational invariant 
+along the flow and vorticity direction and therefore also one-dimensional. 
+However, in contrast to the travelling helical wave of BPII-2 the 
+director field is now static and there is no tumbling motion. 
+While flowing the director retains its relative orientation.
+Consequently, the shear stress is lower than for
+the 2D double-twist rolls of BPI-3 and the travelling helical wave
+(Fig. \ref{bp2-rheo} and \ref{bp1-rheo}) as there is less dissipation.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
+\caption{Flow induced cholesteric state: The orientation of the helix is 
+along the gradient direction. The director field is quasi-static during 
+the flow, which occurs in 'nematic layers'. For clarity the picture shows 
+the director field along the y-direction for selected sites on the x-z-plane.}
+\label{fig:cholflow}
+\end{figure}
+
+%\clearpage
 
@@
 two different flow regimes for BPII and three different ones in case of BPI. 
-Below Ericksen numbers ${\cal O}(10)$ BPII exhibits weak shear-thinning and 
+Below Ericksen numbers $Er\simeq3$ BPII exhibits weak shear-thinning and 
 obeys a power-law.  The BPII disclination network breaks up and reconnects 
 in the flow, which leads to a periodically recurring dependence 
-of the deviatoric shear stress. The flow-induced conformation look generally
+of the deviatoric shear stress. The flow-induced conformation looks generally
 very similar to the quiescent network at equilibrium. 
@@
 
-At larger Ericksen numbers the flowing BPII network breaks up completely into 
-a simple cholesteric phase and flows as a travelling helical wave.
+At larger Ericksen numbers $4\lesssim Er\lesssim 10$ the flowing BPII network 
+breaks up in a two-step manner. First a simple cholesteric helix forms with the helical axis 
+along the vorticity direction. The flow excites a travelling helical wave,
+as predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}.
+Upon increasing the shear, the helix rotates and this time orients along the
+gradient direction, where it can support larger flows due to the translational
+symmetry along the flow direction. Still larger flows break this residual
+cholesteric order and leave a standard flow-aligned nematic staete.
 
@@
 This is a direct consequence of the topological differences in both disclination 
-network. Below Ericksen numbers ${\cal O}(0.1)$ the BPI network cannot flow and 
-retain a regular appearance. Shortly after the onset of the shearing 
-very large stress fluctuations occur. At the lowest shear rates $\gd<9.76\e{-6}$
-the contribution of the network to the total deviatoric stress is actually temporarily 
+network. Below Ericksen numbers $Er\lesssim0.4$ the BPI network cannot both flow and 
+retain a regular appearance. It eventually collapses into an amorphous network and 
+features yield-stress behaviour. The reason is that shortly after the onset of the shearing 
+very large stress fluctuations occur. At the lowest shear rates
+the thermodynamic contribution to the deviatoric (shear) stress is temporarily 
 negative during one cycle. These fluctuations destabilise the network and 
-eventually trigger its dissolution into an amorphous state with 
-dynamic yield stress.
-
-At slightly larger Ericksen numbers $0.1<{\it Er}<10$ shear thinning is 
-slightly stronger than for BPII, but BPI does not show simple power-law 
-behaviour. 
-Instead, it appears BPI kinematically bypasses the large stress fluctuations and 
+eventually trigger the transition into an amorphous state with a residual yield
+stress.
+
+At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim8$,
+BPI kinematically bypasses the large stress fluctuations and 
 flows with very complex, periodically recurring flow-induced conformations.
-The offset of the network after each cycle is random in magnitude and direction.
-However, despite their complex appearance the flow-induced conformations are 
-topologically connected to the quiescent BPI and switching off of the 
-shear flow often restores a perfect equilibrium.
-It is possible to interpret these findings as an instance of 
-deterministic rheochaos.
-
-Just as in case of BPII, the BPI network breaks up at Ericksen numbers ${\cal O}(10)$,
-but adopts a quasi-two-dimensional configuration that is translationally 
-invariant in vorticity direction.
+These conformations entail regularly arrange helical disclinations which 
+rotate due to the shear -- a new flow-induced state which was not previously
+predicted. 
+The offset of the network after each oscillation is along a direction which,
+while being a multiple of the unit cell size along all three coordinate
+axis, depends on system size and shear rate in a seemingly chaotic fashion.
+Despite their complex appearance, the flow-induced conformations are 
+topologically connected to the quiescent BPI and after switching off of the 
+shear flow a defect-free blue phase reforms. It is tempting to
+interpret all these finding as an instance of 
+deterministic rheological chaos.
+
+Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
+a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
+quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
+Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
+direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
+is a flow-aligned nematic state.
 

### 2012-11-26 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
updates to gpu chapter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1820 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\chapter{Ludwig}
+\chapter{Ludwig: GPU implementation of a complex fluid lattice Boltzmann
+application}
 
@@
 a combination of distributed domain decomposition and the Message
-Passing Interface MPI \cite{mpi-standard}. However, the potential
+Passing Interface (MPI). However, the potential
 performance benefits offered by GPUs has motivated a new `mixed-mode'
-approach to address verly large problems. Here, fine-grained
+approach to address very large problems. Here, fine-grained
 parallelism is implemented on the GPU, while MPI is reserved for
 larger-scale parallelism.  This mixed mode is of increasing interest
-to application programmers as many supercomputing services are moving
-toward clusters of GPU accelerated nodes.
+to application programmers at a time when many supercomputing services
+are moving
+toward clusters of GPU accelerated nodes. The design questions which
+arise when developing a lattice Boltzmann code for this type of
+heterogeneous system are therefore worth studying. However, similar
+questions also recur in many other types of stencil-based algorithms.
 
-The \textit{Ludwig} code \cite{desplat} is a LB package developed
-specifically for complex fluid problems, for example mixtures,
-surfactants, liquid crystals, and particle
-suspensions~\cite{aidun2010}.  In this chapter we present the steps
-required to allow Ludwig to efficiently exploit many NVIDIA GPUs in
-parallel.  We show that Ludwig scales excellently to at least the
-thousand GPU level (the largest resource available at the time of
-writing) with indications that the code will scale to much larger
-systems as they come on-line. We first consider the problem of a
-binary fluid mixture (e.g., oil and water) modelled using the LB
-approach, and we go on to describe how we deal with the complexities
-arising when colloidal particles are introduced to interact with the
-fluid.
+The first applications of LB on GPUs were to achieve fluid-like
+effects in computer animation, rather than scientific applications per
+se.  These early works include simple fluids~\cite{wei2004}, miscible
+two-component flow~\cite{zhu2006}, and various image processing tasks
+based on the use of partial differential equations~\cite{zhao2007}.
+While these early works used relatively low level graphics APIs, the
+first CUDA runtime interface implementation was a two-dimensional
+simple fluid problem~\cite{toelke2010}.  Following pioneering work on
+clusters of GPUs coupled via MPI to study air
+pollution~\cite{fan2004}, more recent work has included mixed OpenMP
+and CUDA~\cite{myre2011}, Posix threads and CUDA~\cite{obrecht2011},
+and MPI and CUDA for increasingly large GPU clusters
+\cite{bernaschi2010,xian2011,feichtinger2011}. The heterogeneous
+nature of these systems has also spurred interest in approaches
+including automatic code generation \cite{walshsaar2012} and autotuning
+\cite{williams2011} to aid application performance.
 
-In Section \ref{ch14:sec:background} we give background details to
-introduce the the methods being employed. In Section
-\ref{ch14:sec:singlegpu}, we describe the adaptations made to Ludwig
-to allow it to efficiently exploit the GPU architecture. In Section
-\ref{ch14:sec:singlegpu}, we extend this description to the work
-required to allow exploitation of many GPUs in parallel. We present
-performance and scaling results, for the binary fluid case, in Section
-\ref{ch14:sec:performance}, before going, in Section
-\ref{ch14:sec:particles}, to describe how we minimise the overheads
-involved with introducing colloidal particles.
+Many of these authors make use of
+another attractive feature of LB: the ability to include fixed
+solid-fluid boundary conditions as a straightforward addition to
+the algorithm to study, for example, flow in
+porous media. This points to an important
+application area for LB methods: that of complex fluids.
+Complex fluids include mixtures, surfactants, liquid crystals,
+and particle suspensions, and typically require additional physics
+beyond the bare Navier-Stokes equations to provide a full
+description~\cite{aidun2010}. The representation of this extra
+physics raises additional design questions for the application
+programmer. Here, we consider the \textit{Ludwig} code \cite{desplat},
+an LB application developed specifically for complex fluids
+(\textit{Ludwig} was named for Boltzmann, 1844--1906).
+We will present the steps
+required to allow Ludwig to exploit efficiently both a single
+GPU, and also many GPUs in parallel.  We show that
+Ludwig scales excellently to at least the one thousand GPU level
+(the largest resource available at the time of writing) with
+indications that the code will scale to much larger systems as
+they become available. In addition, we consider the steps required
+to represent fully-resolved moving solid particles (usually referred
+to as colloids in the context of complex fluids). Such particles
+need to have their surface resolved on the lattice scale in order
+to include relevant surface physics, and must be able to move: e.g.,
+to execute Brownian motion in response to random forces from the
+fluid. Standard methods are available to represent such particles
+(e.g., \cite{ladd1994,nguyen2002}) which are amenable to effective
+domain decomposition and message passing \cite{stratford2008}.
+We describe below the additional considerations which arise
+for such moving particles when developing an
+implementation on a GPU.
+
+In the following section we provide a brief overview of the lattice
+Boltzmann method, and mention some of the general issues which can
+influence performance in the CPU context.
+In Section \ref{ch14:sec:singlegpu}, we then describe the alterations
+which are required to exploit the GPU architecture effectively, and
+highlight how the resulting code differs from the CPU version. In
+Section \ref{ch14:sec:parallelgpu}, we extend this description to
+include the steps required to allow exploitation of many GPUs in
+parallel while retaining effective scaling. We also present results for
+a typical benchmark for a fluid-only problem in Section
+\ref{ch14:sec:parallelgpu} to demonstrate the success of the approach.
+In Section \ref{ch14:sec:particles}, we describe the design choices
+which are relevant to a GPU implementation of moving particles. Finally,
+we include in the summary a number of general observations on software
+engineering and maintenance which arise from our experience.
 
@@
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=12cm]{Chapters/chapter14/figures/basiclattice}
-\caption{The lattice Boltzmann approach discritises space into a 3D lattice (left).
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=12cm]{Chapters/chapter14/figures/basiclattice}
+%\caption{The lattice Boltzmann approach discritises space into a 3D lattice (left).
 The fluid is represented at each lattice site (right), by the {\it distribution} vector: each component corresponding to a specific velocity direction.
 Here we are interested in the {\it D3Q19} model, in which there are 3 spatial directions and 19 velocity components (including a zero component).}
-\label{ch14:fig:basiclattice}
-\end{figure}
+%\label{ch14:fig:basiclattice}
+%\end{figure}
 
-A symmetric binary fluid, in which both components have the same
-density and viscosity, may be described by introducing a composition
-variable or \textit{order parameter} $\phi(\mathbf{r})$. This is
-a function of position $\mathbf{r}$ and describes the relative
-proportions of the two fluids present locally. To describe the
-thermodynamics of the system it is possible to write down a
-\textit{free energy} which is a functional of the order parameter.
-The equation of motion for $\phi$ is then
-\begin{equation}
-\partial_t \phi + \mathbf{\nabla}.(\mathbf{u}\phi) = 
-- \mathbf{\nabla} . (M \mathbf{\nabla}\mu),
-\label{cahn-hilliard}
-\end{equation}
-where $\mu$ is the chemical potential related to the functional derivative
-of the free energy, and $M$ is a mobility. This couples
-to the fluid velocity field $\mathbf{u}(\mathbf{r})$, whose evolution
+For a general complex fluid problem the starting point is the
+fluid velocity field $\mathbf{u}(\mathbf{r})$, whose evolution
 obeys the Navier-Stokes equations describing the conservation of mass
-(or density $\rho$) and momentum
+(or density $\rho$), and momentum:
 \begin{equation}
@@
 where $p$ is the isotropic pressure and $\eta$ is the viscosity.
-A force density
-$\mathbf{f}(\mathbf{r})$  describes the force exerted by a
-curved fluid interface on the fluid. It is sufficient to note
-here that this force depends locally on the order parameter and its
-derivatives $\nabla\phi$ and $\nabla^2 \phi$. For a full
-description the interested reader is referred to, e.g.,
-\cite{bray1994,kendon2001}.
+A local force $\mathbf{f}(\mathbf{r})$ provides a means for coupling
+to other complex fluid constituents, e.g., it might represent the force
+exerted on the fluid by a curved interface between different phases or
+components.
 
-The LB approach makes use of a regular 3-dimensional
-lattice (see Figure \ref{ch14:fig:basiclattice}) with discrete spacing $\Delta r$. It also makes use of a
+The LB approach makes use of a regular three-dimensional
+lattice (see Figure \ref{ch14:fig:decomphalo}) with discrete spacing
+$\Delta r$. It also makes use of a
 discrete velocity space $\mathbf{c}_i$, where the $\mathbf{c}_i$
@@
 
-Computationally, we store a vector of 19 double precision floating
-point values at each lattice site for the distribution function $f_i$.
-The collision operation, which is local at each lattice site, may be
-thought of as follows. The matrix-vector multiplication $L_{ji}f_i$ is
-used to transform the distributions into the hydrodynamic quantities,
-where $L_{ij}$ is a constant 19x19 matrix related to the choice of
+More specifically, we store a vector of 19 double precision floating
+point values at each lattice site for the distribution function
+$f_i(\mathbf{r};t)$.
+The collision operation ${\cal L}_{ij}$, which is local at each lattice
+site, may be thought of as follows. A matrix-vector multiplication
+${\cal M}_{ij}f_j$ is used to transform the distributions into the
+hydrodynamic quantities, where ${\cal M}_{ij}$ is a constant 19x19
+matrix related to the choice of
 $\mathbf{c}_i$. The non-conserved hydrodynamic quantities are then
@@
 back to new post-collision distributions via the inverse transformation
-$L^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
+${\cal M}^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
 floating point multiplications per lattice site. In contrast, the
@@
 
-To represent the composition variable, we introduce a second distribution
-function $g_i (\mathbf{r};t)$ whose first moment is $\phi$, second moment
-is the order parameter flux, and so on \cite{swift1996}.
-The time evolution of this second distribution follows an analogous
-collision and propagation process to that described for $f_i$, and
-so provides a solution to Eq.~\ref{cahn-hilliard} \cite{stratford-jsp2005}.
+In the CPU version, the \textit{Ludwig} implementation stores one
+time level of distribution values $f_i(\mathbf{r}; t)$. This distribution
+is stored as a flat array of C data type \texttt{double}, and laid out so
+that all elements of the velocity are contiguous at a given site (often
+referred to as ``array-of-structures'' order). This is appropriate
+for sums over
+the distribution required locally at the collision stage as illustrated
+schematically in Listing~\ref{ch14:listing1}. (The temporary scalar
+\texttt{a\_tmp} allows caching of the intermediate accumulated value
+in the innermost loop.)
+A variety of standard sequential optimisations are relevant for the
+collision stage (loop unrolling, inclusion of SIMD operations, and
+so on \cite{wellein2006}). 
+For example, the collision stage in \textit{Ludwig} includes explicit
+SIMD code, which is useful if the compiler on a given platform cannot
+identify it.
+The propagation stage is separate, and is organised as a ``pull''
+(again, various optimisation have been considered, e.g., 
+\cite{pohl2003,mattila2007,wittmann2012}).
+No further optimisation is done
+here beyond ensuring that the ordering of the discrete velocities allows
+memory access to be as efficient as possible. While these optimisations
+are important, it should be remembered that for some complex fluid
+problems, the hydrodynamics embodied in the LB calculation is a relatively
+small part of the total computational cost (at the 10\%-level in some cases).
+This means optimisation effort may be better concentrated elsewhere.
+
+\begin{lstlisting}[float, label=ch14:listing1,
+caption = Collision schematic for CPU.]
+/* loop over lattice sites */
+for (is = 0; is < nsites; is++) {
+  ...
+  /* load distribution from ``array-of-structures'' */
+  for (i = 0; i < 19; i++)    
+    f[i] = f_aos[19*is + i]
+  ...
+  /* perform matrix-vector multiplication */  
+  for (i = 0; i < 19; i++) {    
+    a_tmp = 0.0;    
+    for (j = 0; j < 19; j++) {      
+      a_tmp += f[j]*M[i][j];   
+    }
+    a[i] = a_tmp;
+  }
+  ...
+}
+\end{lstlisting}
 
@@
 \includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
-\caption{Left: the lattice is decomposed between MPI tasks.
-For clarity we show a 2D decomposition of a 3D lattice, but in practice we decompose in all 3 dimensions.
-Halo cells are added to each sudomain, as shown on the right for a single slice, which store data retrieved from remote neighbours in the halo exchange.}
+\caption{Left: the lattice is decomposed between MPI tasks. For clarity
+we show a 2D decomposition of a 3D lattice, but in practice we decompose
+in all 3 dimensions. Halo cells are added to each sub-domain (as shown on
+the right for a single slice) which store data retrieved from remote
+neighbours in the halo exchange. CAN WE ADD: Lower right: the D3Q19
+velocity set required 5 ``outgoing'' elements to be transferred at
+each edge.}
 \label{ch14:fig:decomphalo}
@@
 
-As illustrated in Figure \ref{ch14:fig:decomphalo}, Ludwig is
-parallelized (across CPU cores) using a regular 3D domain decomposition
-where communications are performed via message passing using MPI.  We
-note that while the collision is local for $f_i$, computation of the
-force terms requires the evaluation of the derivatives of $\phi$ via a
-standard finite different stencil.  This gives rise to the need for a
-halo exchange involving the scalar $\phi$ before the collision.
-Before the propagation stage, a halo exchange in both $f_i$ and $g_i$
-is required.
+The regular 3D decomposition is illustrated in Fig.~\ref{ch14:fig:decomphalo}.
+Each local sub-domain is surrounded by a halo, or ghost, region of width one
+lattice site. While the collision is local, elements of the distribution
+must be exchanged at the edges of the domains to facilitate the propagation.
+To achieve the full 3D halo exchange, the standard approach of shifting the
+relevant data in each co-ordinate direction in turn is adopted. This
+requires appropriate synchronisation, i.e., a receive in the the first
+co-ordinate direction must be complete before a send in the second direction
+involving relevant data can take place, and so on. We note that only
+``outgoing'' elements of the distribution need to be sent at each edge.
+For the D3Q19 model, this reduces the volume of data traffic from 19 to
+5 of the $f_i(\mathbf{r};t)$ per lattice site at each edge. In the CPU
+version, the necessary transfers are implemented in place using
+a vector of appropriately strided MPI datatypes for each direction.
 
-For a 3D halo exchange, 2D planes of data are exchanged in each of the
-positive and negative directions, for each of the 3 dimensions, i.e. 6
-planes are sent and received by each process.  For the CPU version of
-Ludwig, MPI datatype functionality is used to designate these planes.
-The $\phi$ halo exchange involves scalar values at each site, whereas
-the distribution halo exchange involves a higher volume of data since
-we must transfer multiple components of each velocity vector. However
-the number of velocity components included in a given transfer can be
-filtered from 19 to 5 (for D3Q19), since only those elements
-propagating outward from a local domain play a role in the propagation
-stage. The specific components required depend on the direction of
-transfer, and again MPI datatype functionality is used to designate
-this.
+\section{Single GPU Implementation}\label{ch14:sec:singlegpu}
 
-\section{GPU Implementation}\label{ch14:sec:singlegpu}
+In this section we describe the steps taken to enable \textit{Ludwig}
+for the GPU. There are a number of crucial issues: first, the
+minimisation of data traffic between host and device; second, the
+maximisation of available parallelism and third, the issue of
+memory coalescing. We discuss each of these in turn.
 
-In this section we describe the methods used to enable Ludwig
-for the GPU architecture and describe some key optimizations that
-significantly enhance performance.
+While the most important section of the code in terms of floating-point
+performance is the collision stage, this cannot be the only consideration
+for a GPU implementation. It is essential to offload all computational
+activity which involves the main data structures (such as the distribution)
+to the GPU. This includes kernels with relatively low computational demand,
+such as the propagation stage. All relevant data then remain resident on
+the GPU, which avoids expensive host-device data transfer at each iteration
+of the algorithm. Such transfers would negate any benefit of GPU acceleration.
+We note that for a complex fluid code, this requirement can extend to a
+considerable number of kernels, although we limit the discussion to
+collision and propagation for brevity.
 
-To ensure that \textit{Ludwig} remains suitable for CPU-based systems,
-we augment rather than modify the original code. All new GPU-related
-functionality are implemented in an additive manner, mirroring the
-modular structure of the original code. The new GPU modules have
-interfaces similar to the original CPU functions, and the ability to
-switch between versions is achieved at compile-time via C
-pre-processing directives.  All GPU compute kernels are implemented using
-CUDA.  Wrapper routines are developed to specify the
-decomposition, invoke the CUDA kernels and perform the necessary data
-management operations: the latter are facilitated through newly
-developed modules for creation and destruction of device memory
-structures, the copying of data between host and device, and where
-necessary the packing and buffering of data.
-
-The most important code section, in terms of performance, is the
-collision stage. But it is important to offload all other
-computational components of the timestep. Kernels with relatively low
-computational demand, such as the propagation stage and operations
-involving the order parameter, are also GPU enabled: this allows data
-to remain resident on the GPU, and avoids expensive host-device data
-transfer at each timestep (which would reduce the benefit of GPU
-acceleration).  In order to port to the GPU in an incremental fashion,
-while retaining the ability to test for correctness throughout the
-process, we require a suite of data management functions. For each
-data structure (such as the distribution, which we will use as an
-example), we maintain a copy in both the CPU and GPU memory spaces.
-Functions such as \verb+put_dist_on_gpu()+, and
-\verb+get_dist_from_gpu()+, which abstract underlying CUDA
-functionality, allow us to update the GPU copy from the host copy and
-vice versa\footnote {Within the implementation of these functions we
-  have to work around data encapsulation issues. In the code, the use
-  of the file scope identifier \texttt{static} prevents direct access
-  to data; instead access functions are provided.  For the GPU code,
-  it was necessary to use these existing access functions to copy
-  these data to equivalent arrays on the device (via temporary buffers
-  on the host).  Within the CUDA kernel, the data access function
-  calls are substituted with array accesses.  The process is reversed
-  when encapsulated data needs to be transferred back to the host.}.
-This makes it easy, in the function implementing the Ludwig timestep,
-to switch between the CPU and GPU for different components in the
-timestep: we can develop GPU functionality incrementally while
-retaining a code that runs correctly (albeit slowly due to data
-transfer overheads).  Once every component in the timestep is
-GPU-enabled, then it becomes possible to move such data transfers
-outside the timestep loop, and just use the GPU-resident versions
-throughout the simulation.
-
-To achieve optimal performance, it is vital to fully exploit
-parallelism inherent to the NVIDIA architecture, particularly for
-those matrix-vector operations within the dominant collision kernel. A
-simplified schematic of a serial implementation of such a matrix-vector
-operations is as follows:
-{\footnotesize
-\begin{verbatim}
-// loop over lattice sites
-for ( is = 0; is < N; is++){
-  ...
-  //load distribution from site-major structure
-  for (i = 0; i < 19; i++)    
-    f[i]=f_all[is][i]
-  ...
-  //perform matrix-vector multiplication  
-  for (i = 0; i < 19; i++){    
-    a[i] = 0.0;    
-    for (j = 0; j < 19; j++){      
-      a[i] += f[j]*L[i][j];   
-    }
-  }
-  ...
-}
-\end{verbatim}
-}
+To achieve optimal performance, it is vital to exploit fully the
+parallelism inherent in the GPU architecture, particularly for
+those matrix-vector operations within the collision kernel.
 The GPU architecture features a hierarchy of parallelism. At the
-lowest level, groups of 32 threads called {\it warps} operate in
-lock-step on different data elements - this is SIMD style vector-level
+lowest level, groups of 32 threads (warps) operate in
+lock-step on different data elements: this is SIMD style vector-level
 parallelism. Multiple warps are combined into a thread block (in which
 communication and synchronisation are possible), and multiple blocks can
-run concurrently across the SMs in the GPU (with no communication or
+run concurrently across the streaming multiprocessors in the GPU
+(with no communication or
 synchronisation possible across blocks).  To decompose on the GPU, we
-must choose which of the above loops to assign to which level of
+must choose which part of the collision to assign to which level of
 parallelism.
 
-There exists parallelism within the matrix-vector operation, but the
-length of each {\it distribution} vector, at 19, is smaller than the
-warp size and typical thread block sizes. So therefore we simply
-decompose the loop over lattice sites to all levels of parallelism,
-i.e.  we use a separate CUDA thread for every lattice site, and each
-thread performs the full matrix-vector operation. We find that a block
-size of 256 performs best: we decompose our lattice domain into groups
-of 256 sites, and assign each group to a block of CUDA threads.
+While there exists parallelism within the matrix-vector operation, the
+length of each vector (here, 19) is smaller than the warp size and typical
+thread block sizes. So we simply decompose the loop over lattice sites
+to all levels of parallelism, i.e., we use a separate CUDA thread for
+every lattice site, and each thread performs the full matrix-vector
+operation. We find that a block size of 256 performs well on current
+devices: we therefore decompose
+the lattice into groups of 256 sites, and assign each group to a block
+of CUDA threads. As the matrix ${\cal M}_{ij}$ is constant, it is
+assigned to the fast {\it constant} on-chip device memory.
 
-Since the matrix \verb+L+ is constant across CUDA threads (sites), it
-can reside in the fast {\it constant} on-chip device memory. For each
-\verb+is+, \verb+L+ is multiplied by the distribution function at that
-site; the distributions for the complete lattice are stored in
-\verb+f_all+ residing in off-chip device global memory.
+For the propagation stage, the GPU implementation adds a second time
+level of distribution values. The data-dependencies inherent in the
+propagation mean that the in-place propagation of the CPU version
+cannot be parallelised effectively without the additional time level.
+As both time levels may remain resident on the GPU, this is not a
+significant overhead.
 
-Two distinct optimizations were found to improve the performance of this code section, and hence the application as a whole, on
-the GPU.  First, the original code uses the output vector \verb+a+ to
-hold each temporary summation in the inner loop.  The code was adapted
-so that a temporary scalar variable was used instead. This allows
-caching of intermediate values on-chip and thus avoids expensive
-global memory accesses within the inner loop.  Secondly, it can be
-seen the original data layout of \verb+f_all+ is site-major, i.e. for
-each site, consecutive velocity components are contiguous in memory.
-This layout is optimal when executing on the CPU, but is sub-optimal
-on the GPU when combined with our CUDA decomposition. An architectural
-constraint of NVIDIA GPUs means that optimal global memory bandwidth
-is only achieved when data is structured such that threads within a
+An architectural
+constraint of GPUs means that optimal global memory bandwidth
+is only achieved when data are structured such that threads within a
 {\it half-warp} (a group of 16 threads) load data from the same memory
-segment in a single transaction: this is referred to as {\it memory
-  coalescing}.  Otherwise, memory accesses are serialized,
-significantly degrading performance. The coalescing criteria is met
-when consecutive threads read consecutive memory addresses. Therefore
-the alternative transposed velocity-major layout of \verb+f_all+ is
-more optimal, i.e.  for each velocity component, consecutive site
-values are contiguous in memory.  \textit{Ludwig} was modified to
-allow a choice of distribution data layout at compilation time: the
-original (most optimal for the CPU) or that permitting coalesced
-accesses on the GPU.
+segment in a single transaction: this is memory coalescing. It can be
+seen that the array-of-structures ordering used for the distribution
+in the CPU code would not be suitable for coalescing; in fact, it would
+result in serialised memory accesses and relative poor performance.
+To meet the coalescing criteria and allow consecutive threads to read
+consecutive memory addresses on the GPU, we transpose the layout of the
+distribution so that, for each velocity component, consecutive sites
+are contiguous in memory (``structure-of-arrays'' order). A schematic of
+the GPU collision code is given in Listing~\ref{ch14:listing2}.
 
-A schematic of our optimised CUDA code, to be compared with the
-listing above, is
-{\footnotesize
-\begin{verbatim}
-// calculate is from CUDA thread and block indexes
+\begin{lstlisting}[float, label=ch14:listing2,
+caption = Collision schematic for GPU.]
+/* compute current site index 'is' from CUDA thread and block */
+/* indices and load distribution from "structure-of-arrays" */
+
+for (i = 0; i < 19; i++)    
+  f[i] = f_soa[nsites*i + is]
+
+/* perform matrix-vector multiplication as before */
 ...
-//load distribution from velocity-major structure
-for (i = 0; i < 19; i++)    
-   f[i]=f_all[i][is]
-...
-//perform matrix-vector multiplication  
-for (i = 0; i < 19; i++){    
- double a_tmp = 0.0;    
- for (j = 0; j < 19; j++){      
-   a_tmp += f[j]*L[i][j];   
- }
- a[i] = a_tmp;    
-}
-...
-\end{verbatim}
-}
+\end{lstlisting}
 
-\section{Multi-GPU Implementation}\label{ch14:sec:parallelgpu}
+\textit{Ludwig} was modified to allow a choice of distribution data layout
+at compilation time depending on the target architecture: CPU or GPU. We
+defer some further comments on software engineering aspects of the code to
+the summary.
 
-To execute on multiple GPUs in parallel we decompose the problem using
-the existing high-level parallelisaton framework: each MPI task acts
-as a host to a separate GPU, with each GPU kernel operating on the
-subset of the lattice volume corresponding to the partition assigned
-to that task.  As described in Section \ref{ch14:sec:background}, halo
-data exchanges are required at each timestep.  Significant
-developments are made for the GPU adaptation of this communication
-phase to achieve good inter-GPU communication performance and allow
-scaling to many GPUs.
+\section{Multiple GPU Implementation}\label{ch14:sec:parallelgpu}
 
-The
-traditional code uses MPI datatype functionality to designate the six
-planes of lattice data to be sent/received, and performs the
-communications {\it in-place}, i.e.  there is no explicit buffering of
-data in the application.  For the GPU adaptation, the halo planes must
-be exchanged between GPUs, on which MPI functionality is not
-available, so transfers must be staged through the host CPUs. Buffer
-packing/unpacking of the halo planes is explicitly performed using
-CUDA kernels, and CUDA memory copy calls are used to transfer data
-from a device to its host (using pinned host memory) through the PCI-e
-bus (and vice versa). MPI calls are used to transfer the buffers
-between hosts. The CPU version of Ludwig has a mechanism, using MPI built-in
-data types, to reduce the amount of data communicated by around a
-factor of four by only sending those elements of the distribution
-propagating outward from a local domain. For the GPU adaptation, this
-filtering of data is explicitly performed (again since MPI
-functionality is not available on the GPU) in the buffer
-packing/unpacking kernels.  Special care had to be taken for those
-lattice sites on the corners of the halo planes, which had to be sent
-in more than one direction (since the propagation includes accesses to
-diagonal neighbours).
+To execute on multiple GPUs, we use the same domain decomposition
+and message passing framework as the CPU version. Within each
+sub-domain (allocated to one MPI task) the GPU implementation
+proceeds as described in the previous section. The only additional
+complication is that halo transfers between GPUs must be staged
+through the host (in future, direct GPU to GPU data transfers via
+MPI may be possible, obviating the need for these steps). This
+means host MPI sends must be preceded by appropriate device to host
+transfers and host MPI receives must be followed by corresponding host
+to device transfers.
 
-The fact that we are transferring multiple buffers allows us to
-overlap the communication stages, which each stress different parts of
-the hardware, and hence reduce overall communication overhead.  For
-example, after the planes corresponding to the first dimension have
-been retrieved by the host, these can be exchanged using MPI between
-hosts at the same time as the packing and retrieving of the second
-dimension planes. The operations are not completely independent across
-dimensions because of updates to corner data (see above). We use a
-separate CUDA stream for each dimension and overlap to the extent
-allowed. When overlapping is invoked, we observe that some of the
-host-device communication time can be effectively ``hidden'' behind
-the host-host MPI communication time, resulting in an overall speedup.
-The effect is more pronounced for the smaller local lattice size,
-perhaps due to less CPU memory bandwidth contention. Hence the
-optimization is particularly valuable to aid strong scaling, where the
-local system size reduces as the number of GPUs increases.
+%To execute on multiple GPUs in parallel we decompose the problem using
+%the existing high-level parallelisaton framework: each MPI task acts
+%as a host to a separate GPU, with each GPU kernel operating on the
+%subset of the lattice volume corresponding to the partition assigned
+%to that task.  As described in Section \ref{ch14:sec:background}, halo
+%data exchanges are required at each timestep.  Significant
+%developments are made for the GPU adaptation of this communication
+%phase to achieve good inter-GPU communication performance and allow
+%scaling to many GPUs.
 
-\section{Performance}\label{ch14:sec:performance}
+%The CPU code uses strided MPI datatypes to define the six
+%planes of lattice data to be sent or received, and performs the
+%communications {\it in-place}, i.e.  there is no explicit buffering of
+%data in the application.  For the GPU adaptation, the halo planes must
+%be exchanged between GPUs, on which MPI functionality is not
+%available, so transfers must be staged through the host CPUs. Buffer
+%packing/unpacking of the halo planes is explicitly performed using
+%CUDA kernels, and CUDA memory copy calls are used to transfer data
+%from a device to its host (using pinned host memory) through the PCI-e
+%bus (and vice versa). MPI calls are used to transfer the buffers
+%between hosts. The CPU version of Ludwig has a mechanism, using MPI built-in
+%data types, to reduce the amount of data communicated by around a
+%factor of four by only sending those elements of the distribution
+%propagating outward from a local domain. For the GPU adaptation, this
+%filtering of data is explicitly performed (again since MPI
+%functionality is not available on the GPU) in the buffer
+%packing/unpacking kernels.  Special care had to be taken for those
+%lattice sites on the corners of the halo planes, which had to be sent
+%in more than one direction (since the propagation includes accesses to
+%diagonal neighbours).
 
-We compare the performance of the CPU and GPU versions of Ludwig using
-state of the art Cray supercomputing facilities. The Cray XE6
-architecture \cite{xe6} features 2 AMD Opteron CPUs per node, with
-nodes interconnected using Cray Gemini technology. For our baseline
-CPU results we use the HECToR UK academic supercomputing facility
-\cite{hector}. The Cray XK6 \cite{xk6} is very similar to the Cray
-XE6, but with one CPU per node replaced with an NVIDIA X2090 GPU. Each
-node in the system therefore features a single Opteron CPU acting as a
-host to a single GPU. The inter-node interconnect architecture is the
-same as for the Cray XE6: all communications are performed via the
-host CPUs.  For our GPU performance tests we use a prototype Cray XK6
-system, internal to Cray, with 936 nodes (GPUs).
+In practice, this data movement requires additional GPU kernels to
+pack and unpack the relevant data before and after corresponding MPI
+calls. However, the standard shift algorithm, in which each co-ordinate
+direction is treated in turn, does provide some scope for the
+overlapping of different operations.
+For example, after the data for the first co-ordinate direction have been
+retrieved by the host, these can be exchanged using MPI between
+hosts at the same time as kernels for packing and retrieving of
+data for the second co-ordinate direction are
+executed. This
+overlapping must respect the synchronisation required to ensure
+that data values at the corners of the sub-domain are transferred
+correctly.
+We use a separate CUDA stream for each co-ordinate direction:
+this allows some of the host-device communication time to be
+effectively ``hidden'' behind the host-host MPI communication,
+resulting in an overall speedup.
+The improvement is more pronounced for the smaller local lattice size,
+perhaps because of less CPU memory bandwidth contention. The
+overlapping is then a particularly valuable aid to strong scaling,
+where the local system size decreases as the number of GPUs increases.
 
-\begin{figure}[!t]
-\centering
-\includegraphics[width=10cm]{Chapters/chapter14/figures/length_scale}
-\caption{The dependence of the characteristic length scale of a binary
-  fluid mixture on the simulation timestep number. Superimposed are
-  three snapshots showing increased mixture separation with time:
-  these are obtained through the order parameter and each show one
-  eighth of the total simulation size, which is $1548\times 1764\times 1568$. }\label{ch14:fig:length_scale}
-\end{figure}
+%\section{Performance}\label{ch14:sec:performance}
+
+To demonstrate the effectiveness of this approach, we compare the
+performance of both CPU and GPU versions of \textit{Ludwig}. To
+test the complex fluid nature of the code, the problem is
+actually an immiscible fluid mixture which uses a second distribution
+function to introduce a composition variable. The interested reader
+is referred to \cite{ch14:stratford-jsp2005} for further details. The
+largest total problem size used is $1548\times 1764\times 1568$ ?PLEASE CHECK EXACTLY.
+The CPU system features a Cray XE6 architecture with 2 16-core AMD Opteron
+CPUs per node, and with nodes interconnected using Cray Gemini technology.
+For GPU results, a Cray XK6 system is used: this is very similar to the
+XE6, but has one CPU per node replaced with an NVIDIA X2090 GPU. Each
+node in the GPU system therefore features a single Opteron CPU acting
+as a host to a single GPU. The inter-node interconnect architecture is
+the same as for the Cray XE6. The GPU performance tests use a prototype
+Cray XK6 system with 936 nodes (the LARGEST AVAILABLE AT TIME OF WRITING?).
+To provide a fair comparison, we compare scaling on a \textit{per node}
+basis. That is, we compare 1 fully occupied 32-core CPU node
+(running 32 MPI tasks) with 1 GPU node (host running 1 MPI task,
+and 1 device). We believe this is representative of the true ``cost'' of
+a simulation in terms of accounting budgets, or electricity.
+
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=10cm]{Chapters/chapter14/figures/length_scale}
+%\caption{The dependence of the characteristic length scale of a binary
+%  fluid mixture on the simulation timestep number. Superimposed are
+%  three snapshots showing increased mixture separation with time:
+%  these are obtained through the order parameter and each show one
+%  eighth of the total simulation size, which is $1548\times 1764\times 1568$. }\label{ch14:fig:length_scale}
+%\end{figure}
 
@@
 \caption{The weak (top) and strong (bottom) scaling of Ludwig.  Closed
-  shapes denote results using the traditional version run on the Cray
+  shapes denote results using the CPU version run on the Cray
   XE6 (using two 16-core AMD Interlagos CPUs per node), while open
@@
 
-The benchmark used involves the separation of an immiscible binary
-fluid, where the characteristic length scale of fluid domains grows in
-time in a well understood way \cite{kendon2001}. Figure
-\ref{ch14:fig:length_scale} shows the characteristic length scale as a
-function of time (both in LB units) for a system of size $1548\times
-1764\times 1568$ run for around 12 hours on the Cray XK6 using 936
-GPUs. The insets show a visualization of the fluid domains at
-different times.  For the following comparisons we restrict our
-benchmark runs to 100 timesteps: this is representative of the full
-run.
+%The top of Figure \ref{ch14:fig:scaling} shows the benchmark time,
+%where the problem size is increased with the number of nodes. This
+%shows weak scaling: the time taken by a perfectly scaling code would
+%remain constant. The figure plots the time against the number of {\it
+%  nodes}: for the GPU version we utilize the one GPU per Cray XK6 node
+%whilst for the CPU version we fully utilize two 16-core CPUs
+%per Cray XE6 node, i.e. one GPU is compared with 32 CPU cores.  We use
+%a single MPI task per Cray XK6 node to host the GPU, and 32 MPI tasks
+%to fully utilize the CPUs in the Cray XE6 node.  Note that our CPU
+%version is highly optimised: in particular it has been tuned for
+%optimal utilisation of the SIMD vector units in each CPU core.
 
-The top of Figure \ref{ch14:fig:scaling} shows the benchmark time,
-where the problem size is increased with the number of nodes. This
-shows weak scaling: the time taken by a perfectly scaling code would
-remain constant. The figure plots the time against the number of {\it
-  nodes}: for the GPU version we utilize the one GPU per Cray XK6 node
-whilst for the traditional version we fully utilize two 16-core CPUs
-per Cray XE6 node, i.e. one GPU is compared with 32 CPU cores.  We use
-a single MPI task per Cray XK6 node to host the GPU, and 32 MPI tasks
-to fully utilize the CPUs in the Cray XE6 node.  Note that our CPU
-version is highly optimised: in particular it has been tuned for
-optimal utilisation of the SIMD vector units in each CPU core.
-The new GPU version scales almost perfectly: the scaling advantage
-over the CPU version is likely caused by the fact that the GPU version
-has a factor of 32 less MPI tasks per node, hence communications
-require a larger number of smaller data buffers. The performance
+Figure \ref{ch14:fig:scaling} shows the results of both weak and
+strong scaling tests (top and bottom panels, respectively). For
+weak scaling, where the local sub-domain size is kept fixed
+(here a relatively large 196$^3$ lattice), the time taken by
+an ideal code would remain constant. It can be seen that while
+scaling of the CPU version shows a pronounced upward slope, the
+GPU version scales almost perfectly. The advantage of the GPU
+version over the CPU version is the fact that the GPU version
+has a factor of 32 fewer MPI tasks per node, so communications
+require a signifcantly smaller number of larger data transfers.
+The performance
 advantage of the GPU version ranges from a factor of around 1.5 to
-around 1.8.
+around 1.8. Careful study by other authors \cite{williams2011} have
+found the absolute performance (in terms of floating point
+operations per second, or floating point operations per Watt) to
+be remarkably similar between architectures.
 
-We present a strong scaling comparison in the bottom of Figure
-\ref{ch14:fig:scaling}, where we measure the performance (inverse
-runtime) varying the number of nodes utilized (where we used up to 768
-nodes), for several fixed problem sizes (where for each size we scale
-results by the lattice volume to allow plotting on the same graph).
-It can be seen that, for each of the four problem sizes, the GPU
-adaptation on the Cray XK6 scales better (i.e.  the series are closer
-to linearity) than the traditional version on the Cray XE6.  It can be
-seen that ideal scaling is being approached by the larger problem
-sizes.  The higher performance of the GPU version is manifested in the
-larger gradients of the Cray XK6 series.
+Perhaps of more interest is the strong scaling picture (lower
+panel in Figure \ref{ch14:fig:scaling}) where the performance
+as a function of the number of nodes is measured for fixed problem
+size. We consider four different fixed problem sizes on both CPU
+(up to 1024 nodes) and GPU (up to 768 nodes). To allow comparison,
+the results are scaled by total system size in each case. For strong
+scaling, the disparity in the number of MPI tasks is clearly revealed
+in the failure of the CPU version to provide any significant benefit
+beyond a modest number of nodes as communication overheads dominate.
+In contrast, the GPU version shows
+reasonably robust scaling even for the smaller system sizes and good
+scaling for the larger systems.
 
-\section{Inclusion of Particles}\label{ch14:sec:particles}
+\section{Moving Solid Particles}\label{ch14:sec:particles}
 
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=6cm]{Chapters/chapter14/figures/particlemove}
-\caption{A 2D illustration of a colloidal particle moving across the
-    lattice. The old and new positions are represented by the open and
-    filled circles respectively.}
-\label{ch14:fig:particlemove}
-\end{figure}
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=6cm]{Chapters/chapter14/figures/particlemove}
+%\caption{A 2D illustration of a colloidal particle moving across the
+%    lattice. The old and new positions are represented by the open and
+%    filled circles respectively.}
+%\label{ch14:fig:particlemove}
+%\end{figure}
 
-It is of interest, in many situations, to include colloidal particles
-in the simulation **more from Kevin***. These are by represented by
-moving spheres that interact with the fluid through transfer of
-momenta: fluid coming into contact with a particle will bounce off it,
-giving the particle a kick in the process. As illustrated on the left
-of Figure \ref{ch14:fig:bbl} (which shows 2 dimensions only for
-clarity - in practice, the particles will be represented by spheres on
-a 3D lattice, and will be much larger relative to the lattice
-spacing), typically each particle spans multiple lattice sites, and
-therefore ``intercepts'' multiple inter-site links. Ludwig stores
-particle information using nested linked lists: a list of particles,
-each with a list of all the links intercepted by that particle (plus
-other quantities). The fluid-particle interaction proceeds using the
-so-called {\it bounce back on links} procedure during each timestep. As
-illustrated in the right of Figure \ref{ch14:fig:bbl}, for each of the
-intercepting links, those velocity components of the distribution that
-face inwards to each particle are moved from the site
-outside the particle to the site inside the particle, and
-reversed in direction. This is done before the propagation stage,
-which then propagates these same velocity components back out of the
-particles: the velocity bounces back from the particles. The
-momenta transfer from fluid to particles is also determined
-from the distribution values.
+% BEING NEW SOLID TEXT
 
-A simplified schematic of the implementation on the CPU is as follows,
-noting that multiplicative coefficients are required:
-{\footnotesize
-\begin{verbatim}
-For each particle:
-  For each link:
-    Calculate coefficient
-    Bounce back fluid using coefficient
-    Update particle momentum
-\end{verbatim}
-}
-This process is relatively inexpensive on the CPU, since the total
-number of particles (and intercepted links) is always relatively
-small, but it does require access to distribution data which we want
-to keep resident on the GPU throughout the simulation. The overhead of
-transferring the full distribution to the CPU and back every timestep
-would be too high. But on the other hand, the full codebase required
-to manage particles is quite complex and comprehensive: completely
-porting it to the GPU, such that the particles effectively also remain
-resident on the GPU, would be a large effort and would cause a major
-diversion of CPU and GPU codebases. Further complications would arise
-when running on multiple GPUs due to the fact that particles can move
-between parallel subdomains. Our solution, which minimises overhead,
-is to keep the particles resident on CPU, but offload only their interaction with the fluid to the GPU, as we describe below.
+The introduction of moving solid particles poses an additional hurdle
+to efficient GPU implementation of an LB code such as \textit{Ludwig}.
+In this section, we give a brief overview of the essential features
+of the CPU implementation, and how the considerations raised in the
+previous sections --- maximising parallelism and minimising both
+host to device and GPU to GPU data movement --- shape the design decisions
+for a GPU implementation. We restrict our discussion to a simple fluid
+in this section; additional solid-fluid boundary conditions (e.g.,
+wetting at a fluid-fluid-solid contact line) usually arise elsewhere
+in the calculation and broadly independent of the hydrodynamic boundary
+conditions which are described in what follows.
 
-On the CPU, the code in the above listing can be restructured into
-three distinct stages as (introducing arrays to temporrily store
-indices and data):
-{\footnotesize
-\begin{verbatim}
-1) For each particle, for each link:
-     Store site and velocity indices associated with link
-     Calculate and store coefficient
-2) For each particle, for each link:
-     Update velocity data using stored values
-     Calculate and store particle momenta updates
-3) For each particle, for each link:
-     Update particle momentum using stored values
-\end{verbatim}
-}
-Stage 2 is only one that that accesses fluid data: this can therefore
-be moved to the GPU, while stages 1 and 3 remain on the CPU. Stage 2
-therefore becomes, for the GPU implementation
-{\footnotesize
-\begin{verbatim}
-Transfer coefficient and link index arrays to GPU
-Run CUDA kernel:
-  Update fluid data using coefficient and link index arrays
-  Calculate and store particle momenta updates from fluid data
-Transfer particle momenta updates array back to CPU
-\end{verbatim}
-}
-Since the total number of links involves is relatively small, these
-partial data transfers have minimal overhead.  A complication arises
-due to the fact that the particles move through the lattice, albeit
-relatively slowly, and therefore the list of links that they intercept
-occasionally changes (see Figure \ref{ch14:fig:particlemove}).  When a
-particle moves off a lattice site, the fluid on that site must be
-reconstructed. Similarly, the fluid on the newly obstructed site must
-be removed. For the same reasons given above, we want to avoid moving
-all the code dealing with this complication to the GPU, whilst
-minimising overheads. Since, on any one timestep, the number of
-lattice sites affected is relatively low, we implement functionality
-to perform partial data transfers of the distribution. Only those
-sites affected are packed into a buffer on the GPU, transferred to the
-CPU and unpacked into the CPU version of the distribution. Then, the
-existing code can be used to update those affected sites, before the
-reverse partial data transfer is performed to update the GPU version
-of data.
+Moving solid particles (here, spheres) are defined by a centre position
+which is allowed to move continuously across the space of the lattice,
+and a fixed radius which is typically a few lattice spacings $\Delta r$.
+The surface of the particle is defined by a series of \textit{links}
+where a discrete velocity propagation $\mathbf{c}_i \Delta t$ would
+intercept or cut the spherical shell (see Fig.~\ref{ch14:fig:bbl}).
+Hydrodynamic boundary conditions are then implemented via the standard
+approach of bounce-back on links \cite{ladd1994, nguyen2002}, where the
+relevant post-collision distribution values are reversed at the
+propagation stage with an appropriate correction to allow for the
+solid body motion. The exchange of momentum at each link must then be
+accumulated around the entire particle surface to provide the net
+hydrodynamic force
+and torque on the sphere. The particle motion can then be updated in a
+molecular dynamics-like step.
 
-Our methods for managing particles allow us to exploit the GPU for the
-intensive fluid simulation whilst maintaining the complex code
-required to manage particles on the CPU, with overheads minimised
-through only transferring those small subsets of data corresponding to
-the points of interaction.
+For the CPU implementation, a number of additional MPI communications
+are required: (1) to exchange the centre position, radius, etc of each
+particle as it moves and (2), to allow the accumulation of the force
+and torque for each particle (the links of which may be distributed
+between up to 8 MPI tasks in three dimensions). Appropriate marshaling
+of these data can provide an effective parallelisation for relatively
+dense particle suspensions of up to  40\% solid volume fraction
+\cite{stratford2008}. As a final consideration, fluid distributions
+must be removed and replaced consistently as a given particle moves
+across the lattice and changes its discrete shape.
+
+With these features in mind, we can identify a number of competing
+concerns which are relevant to a GPU implementation:
+\begin{enumerate}
+\item
+Minimisation of host-device data transfer would argue for moving the
+entire particle code to the GPU. However, the code in question involves
+largely conditional logic (e.g., identifying cut surface links) and
+irregular memory accesses (e.g., access to distribution elements around
+a spherical particle). These operations seems poorly suited to effective
+parallelisation on the GPU. As an additional complication, the sums
+required over the particle surface would involve potentially tricky
+and inefficient reductions in GPU memory.
+\item
+The alternative is to retain the relevant code on the CPU. While the
+transfer of the entire distribution $f_i(\mathbf{r};t)$ between host
+and device at each time step is unconscionable owing to memory bandwidth
+considerations, the transfer of only relevant distribution information
+to allow bounce-back on links is possible. At modest solid volumes
+fractions only a very small fraction of the distribution data is
+involved in bounce-back (e.g., a 2\% solid volume fraction of particles
+radius 2.3$\Delta r$ would involve approximately 1\% of the distribution
+data). This option also has the advantage that no further host-device
+data transfers are necessary to allow the MPI exchanges required for
+particle information.
+\end{enumerate}
+
+ALAN PLEASE CHECK:
+We have implemented the second option as follows. For each sub-domain,
+a list of boundary-cutting links is assembled on the CPU which includes
+the identity of the relevant element of the distribution. This list,
+together with the particle information required to compute the correct
+bounce-back term, are transferred to the GPU. The updates to the relevant
+elements of the distribution can then take place on the GPU. The
+corresponding information to compute the update of the particle dynamics
+is returned
+to the CPU, where the reduction over the surface links is computed.
+The change of particle shape may be dealt with in a similar manner:
+the relatively small number of updates required at any one time step
+(or however frequently the particle position is updated) can be
+marshaled to the GPU as necessary.
+CAN WE SAY SOMETHING LIKE: This preliminary implementation
+is found to be effective on problems involving up to X per cent solid
+volume fraction.
+
+Finally, we note that the CPU version actually avoids the collision calculation
+at solid lattice points by consulting a look-up table of solid/fluid
+status. On the GPU, it is perhaps preferable to perform the collision
+stage everywhere instead of moving the look-up table to the GPU and
+introducing the associated logic.
+Ultimately, the GPU might favour other boundary methods which treat solid and
+fluid on a somewhat more equal basis, for example, the immersed boundary
+method \cite{ch14:immersed,ch14:immersed-lb} or smoothed profile method
+\cite{ch14:spm}.
+However, the approach adopted here  allows us to exploit
+the GPU for the intensive fluid simulation whilst maintaining the complex
+code required for particles on the CPU. Overheads of CPU-GPU transfer are
+minimised by transferring only those data relevant to the hydrodynamic
+interaction implemented via bounce-back on links.
+
+% END NEW SOLID TEXT
+
+%It is of interest, in many situations, to include colloidal particles
+%in the simulation **more from Kevin***. These are by represented by
+%moving spheres that interact with the fluid through transfer of
+%momenta: fluid coming into contact with a particle will bounce off it,
+%giving the particle a kick in the process. As illustrated on the left
+%of Figure \ref{ch14:fig:bbl} (which shows 2 dimensions only for
+%clarity - in practice, the particles will be represented by spheres on
+%a 3D lattice, and will be much larger relative to the lattice
+%spacing), typically each particle spans multiple lattice sites, and
+%therefore ``intercepts'' multiple inter-site links. Ludwig stores
+%particle information using nested linked lists: a list of particles,
+%each with a list of all the links intercepted by that particle (plus
+%other quantities). The fluid-particle interaction proceeds using the
+%so-called {\it bounce back on links} procedure during each timestep. As
+%illustrated in the right of Figure \ref{ch14:fig:bbl}, for each of the
+%intercepting links, those velocity components of the distribution that
+%face inwards to each particle are moved from the site
+%outside the particle to the site inside the particle, and
+%reversed in direction. This is done before the propagation stage,
+%which then propagates these same velocity components back out of the
+%particles: the velocity bounces back from the particles. The
+%momenta transfer from fluid to particles is also determined
+%from the distribution values.
+
+%A simplified schematic of the implementation on the CPU is as follows,
+%noting that multiplicative coefficients are required:
+%{\footnotesize
+%\begin{verbatim}
+%For each particle:
+%  For each link:
+%    Calculate coefficient
+%    Bounce back fluid using coefficient
+%    Update particle momentum
+%\end{verbatim}
+%}
+%This process is relatively inexpensive on the CPU, since the total
+%number of particles (and intercepted links) is always relatively
+%small, but it does require access to distribution data which we want
+%to keep resident on the GPU throughout the simulation. The overhead of
+%transferring the full distribution to the CPU and back every timestep
+%would be too high. But on the other hand, the full codebase required
+%to manage particles is quite complex and comprehensive: completely
+%porting it to the GPU, such that the particles effectively also remain
+%resident on the GPU, would be a large effort and would cause a major
+%diversion of CPU and GPU codebases. Further complications would arise
+%when running on multiple GPUs due to the fact that particles can move
+%between parallel subdomains. Our solution, which minimises overhead,
+%is to keep the particles resident on CPU, but offload only their
+%interaction with the fluid to the GPU, as we describe below.
+
+%On the CPU, the code in the above listing can be restructured into
+%three distinct stages as (introducing arrays to temporrily store
+%indices and data):
+%{\footnotesize
+%\begin{verbatim}
+%1) For each particle, for each link:
+%     Store site and velocity indices associated with link
+%     Calculate and store coefficient
+%2) For each particle, for each link:
+%     Update velocity data using stored values
+%     Calculate and store particle momenta updates
+%3) For each particle, for each link:
+%     Update particle momentum using stored values
+%\end{verbatim}
+%}
+%Stage 2 is only one that that accesses fluid data: this can therefore
+%be moved to the GPU, while stages 1 and 3 remain on the CPU. Stage 2
+%therefore becomes, for the GPU implementation
+%{\footnotesize
+%\begin{verbatim}
+%Transfer coefficient and link index arrays to GPU
+%Run CUDA kernel:
+%  Update fluid data using coefficient and link index arrays
+%  Calculate and store particle momenta updates from fluid data
+%Transfer particle momenta updates array back to CPU
+%\end{verbatim}
+%}
+%Since the total number of links involves is relatively small, these
+%partial data transfers have minimal overhead.  A complication arises
+%due to the fact that the particles move through the lattice, albeit
+%relatively slowly, and therefore the list of links that they intercept
+%occasionally changes.  When a
+%particle moves off a lattice site, the fluid on that site must be
+%reconstructed. Similarly, the fluid on the newly obstructed site must
+%be removed. For the same reasons given above, we want to avoid moving
+%all the code dealing with this complication to the GPU, whilst
+%minimising overheads. Since, on any one timestep, the number of
+%lattice sites affected is relatively low, we implement functionality
+%to perform partial data transfers of the distribution. Only those
+%sites affected are packed into a buffer on the GPU, transferred to the
+%CPU and unpacked into the CPU version of the distribution. Then, the
+%existing code can be used to update those affected sites, before the
+%reverse partial data transfer is performed to update the GPU version
+%of data.
+
+It is perhaps interesting at this point to make some more general
+observations on the software engineering challenge presented when
+extending an existing CPU code to the GPU. The already complex
+task of maintaining the code in a portable fashion while also
+maintaining performance is currently formidable. To help this process,
+we have followed a number of basic principles. First,
+in order to port to the GPU in an incremental fashion,
+we have tried to maintain the modular structure of the CPU where
+possible. For each data structure, such as the distribution, a separate
+analogue is maintained in both the CPU and GPU memory spaces. However,
+the GPU copy does not include the complete CPU structure: in
+particular, non-intrinsic datatypes such as MPI datatypes are not
+required on the GPU. Functions to marshal data between CPU and GPU
+are provided for each data structure, abstracting the underlying
+CUDA implementation. (This reasonably lightweight abstraction layer
+could also allow an OpenCL implementation to be developed.) 
+This makes it easy to switch between the CPU and GPU for different
+components in the code, which is useful in development
+and testing. GPU functionality can be added incrementally while
+retaining a code that runs correctly (albeit slowly due to data
+transfer overheads).  Once all relevant components are moved to
+the GPU, it becomes possible to remove such data transfers and
+keep the entire problem resident on the device.
+
+%To ensure that \textit{Ludwig} remains suitable for CPU-based systems,
+%we augment rather than modify the original code. All new GPU-related
+%functionality are implemented in an additive manner, mirroring the
+%modular structure of the original code. The new GPU modules have
+%interfaces similar to the original CPU functions, and the ability to
+%switch between versions is achieved at compile-time via C
+%pre-processing directives.  All GPU compute kernels are implemented using
+%CUDA.  Wrapper routines are developed to specify the
+%decomposition, invoke the CUDA kernels and perform the necessary data
+%management operations: the latter are facilitated through newly
+%developed modules for creation and destruction of device memory
+%structures, the copying of data between host and device, and where
+%necessary the packing and buffering of data.
 
@@
 in complex fluid problems, has undergone the developments required to
-efficiently use large-scale parallel GPU accelerated supercomputers.
-The new functionality augments the original code, such that the
-package now supports both traditional CPU-based and GPU-accelerated
-systems.
-
-Through use of the NVIDIA CUDA programming language the computational
-kernels in the Ludwig timestep were offloaded to the GPU, including
-not just those responsible for the majority of the compute time but
-any that accessed the main data structures, to allow data to be kept
-resident on the GPU and avoid expensive data transfers. We described
-steps taken to optimise the performance on the GPU architecture through
-reducing off-chip memory accesses and including the restructuring of
-data layout in memory to ensure the available memory bandwidth was
-fully exploited.
+use large-scale parallel GPU accelerated supercomputers effectively.
+The new functionality augments the original code, so that
+\textit{Ludwig} can run on either CPU-based or GPU-accelerated systems. 
+Using the NVIDIA CUDA programming language, all the computational
+kernels were offloaded to the GPU. This included not only those
+responsible for the majority of the compute time, but any that
+accessed the relevant data structures. This allows data to be kept
+resident on the GPU and avoids expensive data transfers. We described
+steps taken to optimise the performance on the GPU architecture by
+reducing off-chip memory accesses and restructuring the data layout
+in memory to ensure the available memory bandwidth was exploited fully.
 
 A new halo-exchange communication phase for the code was developed to
-allow efficient parallel scaling to many GPUs. The traditional version
+allow efficient parallel scaling to many GPUs. The CPU version
 relies on MPI datatype functionality for CPU to CPU communication; we
@@
 in aiding strong scaling.
-
-We compared performance, for a binary fluid benchmark, of the new GPU
-adaptation using up to 936 GPUs in parallel on prototype Cray XK6 GPU
-accelerated hardware, comparing to runs on the counterpart CPU based
-Cray XE6 using the (optimized) traditional version. We compare on a
-node-by-node basis, i.e. each GPU against 2 CPUs (or 32 cores),
-ensuring full utilization of each CPU.  The performance of the GPU
-version on the Cray XK6 version is around a factor of 1.5-1.8 better
-than the traditional version on the Cray XE6, depending on how many
-nodes is used in the comparison. The weak scaling of the GPU version
-is excellent up to the 936 GPUs tested. The scaling advantage over the
-traditional version is likely due to the fact that we have a factor of
-32 less MPI processes per node. The ability to strong scale depends on
-the physical system size; as the size increases ideal scaling is
-approached. The strong scaling of the GPU version is better than that
-for the CPU version, for each of the sizes tested.
 
@@
 \section*{Acknowledgments}
-This work has been supported by the CRESTA project that has received
+AG was supported by the CRESTA project, which has received
 funding from the European Community's Seventh Framework Programme
-(ICT-2011.9.13) under Grant Agreement no. 287703.
+(ICT-2011.9.13) under Grant Agreement no. 287703. KS was supported
+by UK EPSRC under grant EV/J007404/1.
 
@@
 
-\bibitem{mpi-standard}
-Message Passing Interface Forum, http://www.mpi-forum.org
+%\bibitem{mpi-standard}
+%Message Passing Interface Forum, http://www.mpi-forum.org
 
@@
 C.K. Aidun and J.R. Clausen,
-\textit{Lattice Botlzmann method for complex flows},
+\textit{Lattice Boltzmann method for complex flows},
 Ann. Rev. Fluid Mech., \textbf{42} 439--472 (2010).
 
-\bibitem{bray1994}
-A.J. Bray,
-\textit{Theory of phase-ordering kinetics},
-Adv. Phys., \textbf{43} 357--459 (1994).
+%\bibitem{bray1994}
+%A.J. Bray,
+%\textit{Theory of phase-ordering kinetics},
+%Adv. Phys., \textbf{43} 357--459 (1994).
 
-\bibitem{kendon2001}
-V.M. Kendon, M.E. Cates, I. Pangonabarraga, J.-C. Desplat, and P. Bladon,
-\textit{Inertial effects in the three-dimensional spinodal decomposition of a
-symmetric binary fluid mixture; a lattice Boltzmann study},
-J. Fluid Mech., \textbf{440}, 147--203 (2001).
+%\bibitem{kendon2001}
+%V.M. Kendon, M.E. Cates, I. Pangonabarraga, J.-C. Desplat, and P. Bladon,
+%\textit{Inertial effects in the three-dimensional spinodal decomposition of a
+%symmetric binary fluid mixture; a lattice Boltzmann study},
+%J. Fluid Mech., \textbf{440}, 147--203 (2001).
 
-\bibitem{swift1996}
-M.R. Swift, E. Orlandini, W.R. Osborn, and J.M. Yeomans,
-\textit{Lattice Boltzmann simulations of liquid-gas and binary
-fluid systems},
-Phys. Rev. E, \textbf{54}, 5041--5052 (1996).
+%\bibitem{swift1996}
+%M.R. Swift, E. Orlandini, W.R. Osborn, and J.M. Yeomans,
+%\textit{Lattice Boltzmann simulations of liquid-gas and binary
+%fluid systems},
+%Phys. Rev. E, \textbf{54}, 5041--5052 (1996).
 
-\bibitem{stratford-jsp2005}
+\bibitem{stratford2008}
+K. Stratford and I. Pagonabarraga,
+Parallel domain decomposition for lattice Boltzmann with moving particles,
+CAMWA REF \textbf{55}, 1585 (2008).
+
+%\bibitem{xe6}
+%Cray XE6 Product Brochure, available from
+%http://www.cray.com/Products/XE/Resources.aspx (2010)
+
+%\bibitem{hector} HECToR home page, http://www.hector.ac.uk
+
+%\bibitem{xk6}
+%Cray XK6 Product Brochure, available from
+%http://www.cray.com/Products/XK6/XK6.aspx (2011)
+
+\bibitem{wei2004}
+X. Wei, W. Li, K. M\"uller, and A.E. Kaufman,
+\textit{The lattice Boltzmann method for simulating gaseous phenomena},
+IEEE Transactions on Visualization and Computer Graphics,
+\textbf{10}, 164--176 (2004).
+%% Apparently first LB via GPU; a serious contribution using single fluid
+%%d3q19 single relaxation time.
+
+\bibitem{zhu2006}
+H. Zhu, X. Liu, Y. Liu, and E. Wu,
+\textit{Simulation of miscible binary mixtures based on lattice Boltzmann method},
+Comp. Anim. Virtual Worlds, \textbf{17}, 403--410 (2006).
+%% Single relaxation (MRT mentioned) time d3q19 apparently sound although
+%%not a lot of detail. Pre-cuda so graphics code.
+
+\bibitem{zhao2007}
+Y. Zhao,
+\textit{Lattice Boltzmann based PDE solver on the GPU},
+Visual Comput., doi 10.1007/s00371-0070191-y (2007).
+
+\bibitem{toelke2010}
+J. T\"olke,
+Implementation of a lattice Boltzmann kernel using the compute unified
+device architecture developed by nVIDIA,
+Comput. Visual Sci. 13 29--39 (2010).
+
+\bibitem{fan2004}
+Z. Fan, F. Qiu, A. Kaufman, and S. Yoakum-Stover,
+\textit{GPU cluster for high performance computing},
+Proceedings of ACM/IEEE Supercomputing Conference, pp. 47--59,
+IEEE Computer Society Press, Pittsburgh, PA (2004).
+
+\bibitem{myre2011}
+J. Myre, S.D.C. Walsh, D. Lilja, and M.O. Saar,
+\textit{Performance analysis of single-phase, multiphase, and multicomponent
+lattice Boltzmann fluid flow simulations on GPU clusters},
+Concurrency Computat.: Pract. Exper., \textbf{23}, 332--350 (2011).
+
+\bibitem{obrecht2011}
+C. Obrecht, F. Kuznik, B. Tourancheau, and J.-J. Roux,
+\textit{Multi-GPU implementation of the lattice Boltzmann method},
+Comput. Math. with Applications,
+doi:10.1016/j.camwa.2011.02.020 (2011).
+
+\bibitem{bernaschi2010}
+M. Bernaschi, M. Fatica, S. Melchionna, S. Succi, and E. Kaxiras,
+\textit{A flexible high-performance lattice Boltzmann GPU code for the
+simulations of fluid flow in complex geometries},
+Concurrency Computat.: Pract. Exper., \textbf{22}, 1--14 (2010).
+
+\bibitem{xian2011}
+W. Xian and A. Takayuki,
+\textit{Multi-GPU performance of incompressible flow computation by
+lattice Boltzmann method on GPU cluster},
+Parallel Comput., doi:10.1016/j.parco.2011.02.007 (2011).
+
+\bibitem{feichtinger2011}
+C. Feichtinger, J. Habich, H. K\"ostler, G. Hager, U. R\"ude, and
+G. Wellein,
+A flexible patch-based lattice Boltzmann parallelization approach
+for heterogeneous GPU-CPU clusters,
+\textit{Parallel Computing} \textbf{37} 536--549 (2011).
+
+\bibitem{wellein2006}
+G. Wellein, T. Zeiser, G Hager, and S. Donath,
+On the single processor performance of simple lattice Boltzmann kernels,
+\textit{Computers and Fluids}, \textbf{35}, 910--919 (2006).
+
+\bibitem{pohl2003}
+T. Pohl, M. Kowarschik, J. Wilke, K. Igelberger, and U. R\"ude,
+Optimization and profiling of the cache performance of parallel
+lattice Boltzmann code,
+\textit{Parallel Process Lett.} \textit{13} 549--560 (2003).
+
+\bibitem{mattila2007}
+K. Mattila, J. Hyv\"aluoma, T. Rossi, M. Aspn\"as and J. Westerholm,
+An efficient swap algorithm for the lattice Boltzmann method,
+\textit{Comput. Phys. Comms.} \textit{176} 200-210 (2007).
+
+\bibitem{wittmann2012}
+M. Wittmann, T. Zeiser, G. Hager, and G. Wellein,
+Comparison of different propagation steps for lattice Boltzmann methods,
+\textit{Comput. Math with Appl.} doi:10.1016/j.camwa.2012.05.002 (2012).
+
+\bibitem{walshsaar2012}
+S.D.C. Walsh and M.O. Saar,
+Developing extensible lattice Boltzmann simulators for general-purpose
+graphics-processing units,
+\textit{Comm. Comput. Phys.}, \textbf{13} 867--879 (2013).
+
+\bibitem{williams2011}
+S. Williams, L. Oliker, J. Carter, and J Shalf,
+Extracting ultra-scale lattice Boltzmann performance via
+hierarchical and distributed auto-tuning,
+\textit{Proc. SC2011}.
+
+\bibitem{ch14:stratford-jsp2005}
 K. Stratford, R. Adhikari, I. Pagonabarraga, and J.-C. Desplat,
@@
 
-\bibitem{xe6}
-Cray XE6 Product Brochure, available from http://www.cray.com/Products/XE/Resources.aspx (2010)
+\bibitem{ladd1994}
+A.J.C. Ladd,
+Numerical simulations of particle suspensions via a discretized
+Boltzmann equation. Part 1. Theoretical foundation,
+\textit{J. Fluid Mech.} \textbf{271} 285--309 (1994);
+Part II. Numerical results,
+\textit{ibid.} \textbf{271} 311--339 (1994).
 
-\bibitem{hector} HECToR home page, http://www.hector.ac.uk
+\bibitem{nguyen2002}
+N.-Q. Nguyen and A.J.C. Ladd,
+Lubrication corrections for lattice Boltzmann simulations of particle
+suspensions,
+\textit{Phys. Rev. E} \textbf{66} 046708 (2002).
 
-\bibitem{xk6}
-Cray XK6 Product Brochure, available from http://www.cray.com/Products/XK6/XK6.aspx (2011)
+\bibitem{ch14:immersed}
+C.S. Peskin,
+Flow patterns around heart valves; a numerical method,
+\textit{J. Comp. Phys.}, \textbf{10}, 252--271 (1972);
+C.S. Peskin,
+The immersed boundary method,
+\textit{Acta Nummerica} \textbf{11} 479--517 (2002).
 
-%\bibitem{wei2004}
-%X. Wei, W. Li, K. M\"uller, and A.E. Kaufman,
-%\textit{The lattice Blotzmann method for simulating gaseous phenomena},
-%IEEE Transactions on Visualization and Computer Graphics,
-%\textbf{10}, 164--176 (2004).
-%% Apparently first LB via GPU; a serious contribution using single fluid
-%%d3q19 single relaxation time.
+\bibitem{ch14:immersed-lb}
+Z.-G. Feng and E.E. Michaelides,
+The immersed boundary-lattice Boltzmann method for solving
+fluid-particles interaction problem,
+\textit{J. Comp. Phys.}, \textbf{195} 602--628 (2004).
 
-%\bibitem{zhu2006}
-%H. Zhu, X. Liu, Y. Liu, and E. Wu,
-%\textit{Simulation of miscible binary mixtures based on lattice Boltzmann method},
-%Comp. Anim. Virtual Worlds, \textbf{17}, 403--410 (2006).
-%% Single relaxation (MRT mentioned) time d3q19 apparently sound although
-%%not a lot of detail. Pre-cuda so graphics code.
-
-%\bibitem{zhao2007}
-%Y. Zhao,
-%\textit{Lattice Boltzmann based PDE solver on the GPU},
-%Visual Comput., doi 10.1007/s00371-0070191-y (2007).
-
-%\bibitem{toelke}
-%J. T\"olke,
-%\textit{Implementation of a lattice Boltzmann kernel using the Compute Unified
-%Device Architecture developed by NVIDIA},
-%Comput. Visual. Sci. doi:10.1007/s00791-008-0120-2 (2008).
-
-%\bibitem{fan2004}
-%Z. Fan, F. Qiu, A. Kaufman, and S. Yoakum-Stover,
-%\textit{GPU cluster for high performance computing},
-%Proceedings of ACM/IEEE Supercomputing Conference, pp. 47--59,
-%IEEE Computer Society Press, Pittsburgh, PA (2004).
-
-%\bibitem{myre2011}
-%J. Myre, S.D.C. Walsh, D. Lilja, and M.O. Saar,
-%\textit{Performance analysis of single-phase, multiphase, and multicomponent
-%lattice Boltzmann fluid flow simulations on GPU clusters},
-%Concurrency Computat.: Pract. Exper., \textbf{23}, 332--350 (2011).
-
-%\bibitem{obrecht2011}
-%C. Obrecht, F. Kuznik, B. Tourancheau, and J.-J. Roux,
-%\textit{Multi-GPU implementation of the lattice Boltzmann method},
-%Comput. Math. with Applications,
-%doi:10.1016/j.camwa.2011.02.020 (2011).
-
-%\bibitem{bernaschi2010}
-%M. Bernaschi, M. Fatica, S. Melchionna, S. Succi, and E. Kaxiras,
-%\textit{A flexible high-performance lattice Boltzmann GPU code for the
-%simulations of fluid flow in complex geometries},
-%Concurrency Computat.: Pract. Exper., \textbf{22}, 1--14 (2010).
-
-%\bibitem{xian2011}
-%W. Xian and A. Takayuki,
-%\textit{Multi-GPU performance of incompressible flow computation by
-%lattice Boltzmann method on GPU cluster},
-%Parallel Comput., doi:10.1016/j.parco.2011.02.007 (2011).
-
-%\bibitem{simd} Intel AVX: New Frontiers in Performance Improvements and Energy Efficiency, Intel article, http://software.intel.com/en-us/articles (2009)
-
-%\bibitem{amd} AMD 6200 series home page  
-%http://www.amd.com/us/products/server/
-%processors/6000-series-platform/6200/Pages/6200-series-processors.aspx (accessed April 2012)
-
-%\bibitem{nvidia} NVIDIA Tesla home page, http://www.nvidia.com/object/
-%tesla\_computing\_solutions.html (accessed April 2012)
-
-%\bibitem{titan} \textit{ORNL awards contract to Cray for Titan supercomputer}, Oak Ridge National Laboratory Press Release mr20111011-00, http://www.ornl.gov/info/press\_releases/ (2011)
+\bibitem{ch14:spm}
+Y. Nakayama and R. Yammamoto,
+Simulation method to resolve hydrodynamic interactions in colloidal
+dispersions,
+\textit{Phys. Rev. E}, \textbf{71} 036707 (2005).
 
@@
 \end{document}
+
+% LocalWords:  CRESTA EPSRC Succi Desplat Pagonabarraga

### 2012-11-28 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
AG updates to gpu chapter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1822 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-In the CPU version, the \textit{Ludwig} implementation stores one
-time level of distribution values $f_i(\mathbf{r}; t)$. This distribution
-is stored as a flat array of C data type \texttt{double}, and laid out so
-that all elements of the velocity are contiguous at a given site (often
-referred to as ``array-of-structures'' order). This is appropriate
-for sums over
-the distribution required locally at the collision stage as illustrated
-schematically in Listing~\ref{ch14:listing1}. (The temporary scalar
-\texttt{a\_tmp} allows caching of the intermediate accumulated value
-in the innermost loop.)
-A variety of standard sequential optimisations are relevant for the
-collision stage (loop unrolling, inclusion of SIMD operations, and
-so on \cite{wellein2006}). 
-For example, the collision stage in \textit{Ludwig} includes explicit
-SIMD code, which is useful if the compiler on a given platform cannot
-identify it.
-The propagation stage is separate, and is organised as a ``pull''
-(again, various optimisation have been considered, e.g., 
-\cite{pohl2003,mattila2007,wittmann2012}).
-No further optimisation is done
-here beyond ensuring that the ordering of the discrete velocities allows
-memory access to be as efficient as possible. While these optimisations
-are important, it should be remembered that for some complex fluid
-problems, the hydrodynamics embodied in the LB calculation is a relatively
-small part of the total computational cost (at the 10\%-level in some cases).
-This means optimisation effort may be better concentrated elsewhere.
+In the CPU version, the \textit{Ludwig} implementation stores one time
+level of distribution values $f_i(\mathbf{r}; t)$. This distribution
+is stored as a flat array of C data type \texttt{double}, and laid out
+so that all elements of the velocity are contiguous at a given site
+(often referred to as ``array-of-structures'' order). This is
+appropriate for sums over the distribution required locally at the
+collision stage as illustrated schematically in
+Listing~\ref{ch14:listing1}: the fact that consecutive loads are from
+consecutive memory addresses allows the prefetcher to fully engage.
+(The temporary scalar \texttt{a\_tmp} allows caching of the
+intermediate accumulated value in the innermost loop.)  A variety of
+standard sequential optimisations are relevant for the collision stage
+(loop unrolling, inclusion of SIMD operations, and so on
+\cite{wellein2006}).  For example, the collision stage in
+\textit{Ludwig} includes explicit SIMD code, which is useful if the
+compiler on a given platform cannot identify it.  The propagation
+stage is separate, and is organised as a ``pull'' (again, various
+optimisation have been considered, e.g.,
+\cite{pohl2003,mattila2007,wittmann2012}).  No further optimisation is
+done here beyond ensuring that the ordering of the discrete velocities
+allows memory access to be as efficient as possible. While these
+optimisations are important, it should be remembered that for some
+complex fluid problems, the hydrodynamics embodied in the LB
+calculation is a relatively small part of the total computational cost
+(at the 10\%-level in some cases).  This means optimisation effort may
+be better concentrated elsewhere.
 
@@
 \includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
-\caption{Left: the lattice is decomposed between MPI tasks. For clarity
-we show a 2D decomposition of a 3D lattice, but in practice we decompose
-in all 3 dimensions. Halo cells are added to each sub-domain (as shown on
-the right for a single slice) which store data retrieved from remote
-neighbours in the halo exchange. CAN WE ADD: Lower right: the D3Q19
-velocity set required 5 ``outgoing'' elements to be transferred at
-each edge.}
+\caption{Left: the lattice is decomposed between MPI tasks. For
+  clarity we show a 2D decomposition of a 3D lattice, but in practice
+  we decompose in all 3 dimensions. Halo cells are added to each
+  sub-domain (as shown on the upper right for a single slice) which store
+  data retrieved from remote neighbours in the halo exchange. Lower
+  right: the D3Q19 velocity set resident on a lattice site;
+  highlighted are the 5 ``outgoing'' elements to be transferred in a
+  specific direction.}
 \label{ch14:fig:decomphalo}
@@
 minimisation of data traffic between host and device; second, the
-maximisation of available parallelism and third, the issue of
-memory coalescing. We discuss each of these in turn.
-
-While the most important section of the code in terms of floating-point
-performance is the collision stage, this cannot be the only consideration
-for a GPU implementation. It is essential to offload all computational
-activity which involves the main data structures (such as the distribution)
-to the GPU. This includes kernels with relatively low computational demand,
-such as the propagation stage. All relevant data then remain resident on
-the GPU, which avoids expensive host-device data transfer at each iteration
-of the algorithm. Such transfers would negate any benefit of GPU acceleration.
-We note that for a complex fluid code, this requirement can extend to a
-considerable number of kernels, although we limit the discussion to
-collision and propagation for brevity.
+optimal mapping of available parallelism onto the architecure and
+third, the issue of memory coalescing. We discuss each of these in
+turn.
+
+While the most important section of the LB in terms of
+floating-point performance is the collision stage, this cannot be the
+only consideration for a GPU implementation. It is essential to
+offload all computational activity which involves the main data
+structures (such as the distribution) to the GPU. This includes
+kernels with relatively low computational demand, such as the
+propagation stage. All relevant data then remain resident on the GPU,
+to avoid expensive host-device data transfer at each iteration of
+the algorithm. Such transfers would negate any benefit of GPU
+acceleration.  We note that for a complex fluid code, this requirement
+can extend to a considerable number of kernels, although we limit the
+discussion to collision and propagation for brevity.
 
@@
 
-To demonstrate the effectiveness of this approach, we compare the
+To demonstrate the effectiveness of our approach, we compare the
 performance of both CPU and GPU versions of \textit{Ludwig}. To
@@
 is referred to \cite{ch14:stratford-jsp2005} for further details. The
-largest total problem size used is $1548\times 1764\times 1568$ ?PLEASE CHECK EXACTLY.
+largest total problem size used is $2548\times 1764\times 1568$.
 The CPU system features a Cray XE6 architecture with 2 16-core AMD Opteron
@@
 the same as for the Cray XE6. The GPU performance tests use a prototype
-Cray XK6 system with 936 nodes (the LARGEST AVAILABLE AT TIME OF WRITING?).
+Cray XK6 system with 936 nodes (the largest available at the time of writing).
 To provide a fair comparison, we compare scaling on a \textit{per node}
@@
 irregular memory accesses (e.g., access to distribution elements around
-a spherical particle). These operations seems poorly suited to effective
+a spherical particle). These operations seem poorly suited to effective
 parallelisation on the GPU. As an additional complication, the sums
@@
 transfer of the entire distribution $f_i(\mathbf{r};t)$ between host
-and device at each time step is unconscionable owing to memory bandwidth
+and device at each time step is unconscionable owing to PCIe bus bandwidth
 considerations, the transfer of only relevant distribution information
 to allow bounce-back on links is possible. At modest solid volumes
-fractions only a very small fraction of the distribution data is
+only a very small fraction of the distribution data is
 involved in bounce-back (e.g., a 2\% solid volume fraction of particles
@@
 
-ALAN PLEASE CHECK:
 We have implemented the second option as follows. For each sub-domain,

### 2012-11-29 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
New fig 3 and summary
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1823 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\chapter{Ludwig: GPU implementation of a complex fluid lattice Boltzmann
+\chapter{Ludwig: multiple GPUs for a complex fluid lattice Boltzmann
 application}
@@
 arise when developing a lattice Boltzmann code for this type of
-heterogeneous system are therefore worth studying. However, similar
+heterogeneous system are therefore worth studying. Furthermore, similar
 questions also recur in many other types of stencil-based algorithms.
@@
 nature of these systems has also spurred interest in approaches
-including automatic code generation \cite{walshsaar2012} and autotuning
+including automatic code generation \cite{walshsaar2012} and auto-tuning
 \cite{williams2011} to aid application performance.
@@
 We will present the steps
-required to allow Ludwig to exploit efficiently both a single
+required to allow \textit{Ludwig} to exploit efficiently both a single
 GPU, and also many GPUs in parallel.  We show that
-Ludwig scales excellently to at least the one thousand GPU level
+\textit{Ludwig} scales excellently to at least the one thousand GPU level
 (the largest resource available at the time of writing) with
@@
 which are relevant to a GPU implementation of moving particles. Finally,
-we include in the summary a number of general observations on software
+we include a number of general observations on software
 engineering and maintenance which arise from our experience.
+A summary is provided in Section~\ref{ch14:sec:summary}.
 
@@
 ${\cal M}^{-1}_{ij}$. This gives rise to the need for a minimum of 2x19$^2$
-floating point multiplications per lattice site. In contrast, the
+floating point multiplications per lattice site. (Additional operations are
+required to implement, for example, the force $\mathbf{f}(\mathbf{r})$.)
+In contrast, the
 propagation stage consists solely of moving the distribution values
@@
 Listing~\ref{ch14:listing1}: the fact that consecutive loads are from
-consecutive memory addresses allows the prefetcher to fully engage.
+consecutive memory addresses allows the prefetcher to engage fully.
 (The temporary scalar \texttt{a\_tmp} allows caching of the
@@
 minimisation of data traffic between host and device; second, the
-optimal mapping of available parallelism onto the architecure and
+optimal mapping of available parallelism onto the architecture and
 third, the issue of memory coalescing. We discuss each of these in
@@
 \includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
-\caption{The weak (top) and strong (bottom) scaling of Ludwig.  Closed
+\caption{The weak (top) and strong (bottom) scaling of \textit{Ludwig}.  Closed
   shapes denote results using the CPU version run on the Cray
@@
 has a factor of 32 fewer MPI tasks per node, so communications
-require a signifcantly smaller number of larger data transfers.
+require a significantly smaller number of larger data transfers.
 The performance
@@
 size. We consider four different fixed problem sizes on both CPU
-(up to 1024 nodes) and GPU (up to 768 nodes). To allow comparison,
+(up to 512 nodes shown) and GPU (up to 768 nodes). To allow comparison,
 the results are scaled by total system size in each case. For strong
@@
 \centering
-\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
-\caption{A 2D illustration of colloidal particles (grey circles)
-  interacting with the lattice, where each lattice point is
-  represented by a cross. Left: each particle spans multiple sites.
-  Right: the particle-facing distribution components are moved inside
-  the particle with the direction reversed, such that the ``bounce
-  back'' will be completed when the fluid propagates.}
+%\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
+\includegraphics[width=10cm]{Chapters/chapter14/figures/colloid_new}
+\caption{
+A two-dimensional schematic picture of spherical particles on the lattice.
+Left: a particle is allowed
+to move continuously across the lattice, and the position of the
+surface defines fluid lattice sites (light blue) and solid lattice
+sites (dark red). The discrete surface is defined by links
+where propagation would intersect the surface (arrows). Note the
+discrete shape of the two particles is different. Right: post-collision
+distributions are reversed at the surface by the process of bounce-back
+on links, which replaces the propagation.
+%A 2D illustration of colloidal particles (grey circles)
+%  interacting with the lattice, where each lattice point is
+%  represented by a cross. Left: each particle spans multiple sites.
+%  Right: the particle-facing distribution components are moved inside
+%  the particle with the direction reversed, such that the ``bounce
+%  back'' will be completed when the fluid propagates.
+}
 \label{ch14:fig:bbl}
@@
 (or however frequently the particle position is updated) can be
-marshaled to the GPU as necessary.
-CAN WE SAY SOMETHING LIKE: This preliminary implementation
-is found to be effective on problems involving up to X per cent solid
-volume fraction.
-
-Finally, we note that the CPU version actually avoids the collision calculation
+marshaled to the GPU as necessary. This preliminary implementation
+is found to be effective on problems involving up to 4\%
+solid volume fraction.
+
+We note that the CPU version actually avoids the collision calculation
 at solid lattice points by consulting a look-up table of solid/fluid
@@
 \section{Summary}
-
-The \textit{Ludwig} LB application, which specializes
-in complex fluid problems, has undergone the developments required to
-use large-scale parallel GPU accelerated supercomputers effectively.
-The new functionality augments the original code, so that
-\textit{Ludwig} can run on either CPU-based or GPU-accelerated systems. 
-Using the NVIDIA CUDA programming language, all the computational
-kernels were offloaded to the GPU. This included not only those
-responsible for the majority of the compute time, but any that
-accessed the relevant data structures. This allows data to be kept
-resident on the GPU and avoids expensive data transfers. We described
-steps taken to optimise the performance on the GPU architecture by
-reducing off-chip memory accesses and restructuring the data layout
-in memory to ensure the available memory bandwidth was exploited fully.
-
-A new halo-exchange communication phase for the code was developed to
-allow efficient parallel scaling to many GPUs. The CPU version
-relies on MPI datatype functionality for CPU to CPU communication; we
-have explicitly written buffer packing/unpacking and transfer
-operations to for GPU to GPU data exchange, staged through the host
-CPUs. This includes the filtering of velocity components to only
-transfer those required in a specific direction, hence reducing the
-overall volume of data transferred. We used CUDA stream functionality
-to overlap separate stages within the communication phase, and reduce
-the overall communication time. The effect was measured to be more
-profound for smaller local system sizes, i.e. it is especially useful
-in aiding strong scaling.
-
-We described work to enable colloidal particles in the simulation with
-minimal overhead. We implemented these in such a way that we avoid a
-major diversion of the CPU and GPU codebases, whilst minimising data
-transfers between the CPU and GPU at each timestep. We keep the
-majority of the (relatively inexpensive) particle related code on the
-CPU, while offloading only those parts responsible for the interaction
-with the fluid to the GPU (where the fluid data resides).
-
-Our resulting package is able to exploit the largest of GPU
-accelerated architectures to address a wide range of complex problems.
-
-%\section{Introduction}
-%% no \IEEEPARstart
-%This demo file is intended to serve as a ``starter file''
-%for IEEE conference papers produced under \LaTeX\ using
-%IEEEtran.cls version 1.7 and later.
-%% You must have at least 2 lines in the paragraph with the drop letter
-%% (should never be an issue)
-%I wish you the best of success.
-
-%\hfill mds
-
-%\hfill January 11, 2007
-
-%\subsection{Subsection Heading Here}
-%Subsection text here.
-
-%\subsubsection{Subsubsection Heading Here}
-%Subsubsection text here.
+\label{ch14:sec:summary}
+
+We have described the steps take to implement, on both single and
+multiple GPUs, the \textit{Ludwig} code which was originally
+designed for complex fluid problems on a conventional CPU
+architecture. We have added the necessary functionality using
+NVIDIA CUDA and discussed the important changes to the main
+data structures. By retaining domain decomposition and message
+passing via MPI, we have demonstrated it is possible to scale
+complex fluid problems to large numbers of GPUs in parallel.
+By following the key design criteria of maximising parallelism
+and minimising host-device data transfers, we have confirmed
+the mixed MPI-GPU approach is viable for scientific applications.
+For the more intricate problem of moving solid particles, we find
+it is possible to retain the more serial elements related to
+particle link operations on the CPU, while offloading only the
+parallel lattice-based operations involving the LB distribution
+to the GPU. Again, this minimises host-device movement of data.
+
+From the software engineering viewpoint, some duplication of code
+to allow efficient implementation on both host and device is
+currently required. This issue might be addressed by approaches
+such as automatic kernel generation, but may also be addressed
+naturally in time as GPU and CPU hardware converge. Portable
+abstractions and APIs, perhaps based on approaches such as OpenCL,
+will also facilitate the development and maintenance of portable
+codes which also exhibit portable performance (perhaps in conjunction
+with automatic tuning approaches).
+
+So, while the challenges in designing portable and efficient scientific
+applications remain very real, this work provides some hope the large
+clusters of GPU machines can be used effectively for a wide range of
+complex fluid problems.
+
+%The \textit{Ludwig} LB application, which specializes
+%in complex fluid problems, has undergone the developments required to
+%use large-scale parallel GPU accelerated supercomputers effectively.
+%The new functionality augments the original code, so that
+%\textit{Ludwig} can run on either CPU-based or GPU-accelerated systems. 
+%Using the NVIDIA CUDA programming language, all the computational
+%kernels were offloaded to the GPU. This included not only those
+%responsible for the majority of the compute time, but any that
+%accessed the relevant data structures. This allows data to be kept
+%resident on the GPU and avoids expensive data transfers. We described
+%steps taken to optimise the performance on the GPU architecture by
+%reducing off-chip memory accesses and restructuring the data layout
+%in memory to ensure the available memory bandwidth was exploited fully.
+
+%A new halo-exchange communication phase for the code was developed to
+%allow efficient parallel scaling to many GPUs. The CPU version
+%relies on MPI datatype functionality for CPU to CPU communication; we
+%have explicitly written buffer packing/unpacking and transfer
+%operations to for GPU to GPU data exchange, staged through the host
+%CPUs. This includes the filtering of velocity components to only
+%transfer those required in a specific direction, hence reducing the
+%overall volume of data transferred. We used CUDA stream functionality
+%to overlap separate stages within the communication phase, and reduce
+%the overall communication time. The effect was measured to be more
+%profound for smaller local system sizes, i.e. it is especially useful
+%in aiding strong scaling.
+
+%We described work to enable colloidal particles in the simulation with
+%minimal overhead. We implemented these in such a way that we avoid a
+%major diversion of the CPU and GPU codebases, whilst minimising data
+%transfers between the CPU and GPU at each timestep. We keep the
+%majority of the (relatively inexpensive) particle related code on the
+%CPU, while offloading only those parts responsible for the interaction
+%with the fluid to the GPU (where the fluid data resides).
+
+%Our resulting package is able to exploit the largest of GPU
+%accelerated architectures to address a wide range of complex problems.
 
@@
 Parallel domain decomposition for lattice Boltzmann with moving particles,
-CAMWA REF \textbf{55}, 1585 (2008).
+\textit{Comput. Math. with Applications} \textbf{55}, 1585 (2008).
 

### 2012-12-10 trunk/htdocs/colloid.tex
Some colloid documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1833 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  colloid.tex
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Colloid Examples}

\subsection{Initialisation from file}

We present a number of examples taken from the regression tests, the inputs
for which are found in
\begin{verbatim}
trunk/tests/regression
\end{verbatim}
For each test there is an input file, and an ASCII file containing the
details of the initial colloid state which is read at run time by means
of the key value pair
\begin{verbatim}
colloid_init    from_file
\end{verbatim}

A file of colloid information may be created with the utility
\begin{verbatim}
util/colloid_file.c
\end{verbatim}
As a minimum, this file must specify:
\begin{enumerate}
\item
A unique integer id for each particle;
\item
a radius and hydrodynamic radius $a_0$ and $a_h$ (use the same value
if unsure);
\item
an initial position ($x_{min} < x < x_{max}$) with $x_{min} = 0.5$
and $x_{max} = L_x + 0.5$ etc, where $L_x$ is the appropriate system
size for the problem at hand;
\item
the initial velocity etc may be safely initialised to zero.
\end{enumerate}
Note that is an inter-particle potential is to be specified at run time,
the initial position of the particles should not be so close that a
large force is experienced at time $t=0$. This can destabilise the
dynamics.

A single file of colloid information is produced in either ASCII or binary
format, which can be read in by specifying the appropriate
\begin{verbatim}
colloid_io_format_input   ASCII_SERIAL
\end{verbatim}
or \texttt{BINARY\_SERIAL} key value in both serial and parallel.

\subsubsection{Very short range potential}

See
\begin{verbatim}
tests/regression/test_spin_solid2_input
tests/regression/config.cds.init.001-001
tests/regression/test_spin_solid2_d3q19.ref*
\end{verbatim}

An example of a moderate volume fraction of particles is given for
a binary fluid undergoing spinodal decomposition. Here,
the capillary interaction between the neutrally wetting colloids
causes a strong effective attraction between particles. It is therefore
necessary to ensure the particles do not collide to the point of
overlapping at their hard-sphere radius. A counterbalancing short-range
soft-sphere potential is then defined in the input file:
\begin{verbatim}
soft_sphere_on 1
soft_sphere_epsilon 0.0004
soft_sphere_sigma 0.1
soft_sphere_nu 1.0
soft_sphere_cutoff 0.25
\end{verbatim}
where, following Eq.~\ref{eq_ss_shift}, we have
$v(r) \sim \epsilon (\sigma /r)^\nu$ ``cut-and-shifted'' so that
both potential and force smoothly match to zero at the cut-off
distance $r_c$ (here $0.25\Delta x$). The energy scale $\epsilon$
will need to be adjusted depending on the exact problem at hand
(here it will be related to the fluid-fluid interfacial tension).

Other relevant parameters here are
\begin{verbatim}
colloid_cell_min 8.0
lubrication_on 0
\end{verbatim}
which control, respectively, the minimum cell list width used to
help identify interactions, and the lubrication correction (here
switched off).

\subsubsection{Short-range potential}

See
\begin{verbatim}
tests/regression/test_yukawa_input
tests/regression/test_yukawa_cds.001-001
tests/regression/test_yukawa_d3q10.ref1
\end{verbatim}

This example involves a number of interacting particles in a simple
fluid including Brownian motion. The potential is Yukawa-like,
representative of a screened Coulomb interaction.

We have the parameters:
\begin{verbatim}
yukawa_on 1
yukawa_epsilon  1.330764285
yukawa_kappa 0.72463768115
yukawa_cutoff 16.0
\end{verbatim}
where the potential is $v(r) \sim \epsilon (-\kappa r) / r$. Again,
both potential and force are smoothly matched to zero at the
cut-off distance $r_c$ (here $r_c = 16\Delta x$). The relatively
long cut-off distance means that the minimum cell list cell width
must be correspondingly large (16 lattice units) which limits the
minimum parallel domain size.

This example also includes Brownian motion by means of fluctuating
hydrodynamics. The relevant parameters in the input are:
\begin{verbatim}
isothermal_fluctuations on 
temperature 0.0002133333
\end{verbatim}
Note that the \texttt{temperature} parameter here specifies
$k_BT = \langle c_x^2\rangle = \langle c_y^2 \rangle = \langle c_z^2 \rangle$
in three dimensions (at equilibrium);
this is reported in the output of the code at run time.

\subsubsection{Long range forces}

An Ewald sum is available for magnetic dipoles.

### 2012-12-10 trunk/htdocs/sigma.tex
Some colloid documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1833 40876441-3496-0410-a613-a2fca35b2ae2

--- 2010-11-29 htdocs/sigma.tex
@@
 \hline
+Rks001 & 0.00625 & 0.004 & 1.41 & 0.05 & 0.0047 & 422 & 1.26e10$^5$\\
+Rks002 & 0.00625 & 0.004 & 0.65 & 0.25 & 0.0047 & 89.7 & 1.24e10$^4$\\
 Run028 & 0.083 & 0.053 & 1.41 & 0.05  & 0.0625 & 31.8 & 718 \\
@@
 the order parameter mobility is used here (cf $\tilde{M} = 2M$) in
-Kendon et al.}
+Kendon et al. (KS values are extra sets used to low 1/$\dot{\gamma}T_0$.)}
 \label{tab:r1}

### 2012-12-13 htdocs/anchoring.tex
Added comment and reference on Euler-Lagrange eq. for anchoring
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1834 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-06-21 trunk/htdocs/anchoring.tex
@@
 
-The boundary condition we wish to apply is, for the bulk term
+The boundary condition we wish to apply are the Euler-Lagrange equations. 
+They are at the surface of the particle
 \begin{equation}
@@
 $n_\gamma$ is the outward unit normal at the surface (pointing into the
-fluid). With the
+fluid) \cite{skarabot}. Note that compared to the regular minimization as  
+explained in \cite{wright} this is a surface integral term that would 
+usually vanish in the bulk.
+With the
 addition of the surface term we obtain

### 2012-12-13 htdocs/references.tex
Some updates to tex docs
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1514 40876441-3496-0410-a613-a2fca35b2ae2
Added comment and reference on Euler-Lagrange eq. for anchoring
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1834 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-06-21 trunk/htdocs/references.tex
@@
 
+\bibitem{skarabot}
+M. Skarabot, M. Ravnik, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman and I. Musevic,
+\textit{Phys. Rev. E} \textbf{76}, 051406 (2007).
+
+\bibitem{wright}
+D.C. Wright and N.D. Mermin,
+\textit{Rev. Mod. Phys.} \textbf{61}, 385 (1989).
+
 \end{thebibliography}

### 2012-12-19 htdocs/anchoring.tex
Added to htdocs on surface anchoring
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1844 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \begin{equation}
-f_s = {\textstyle\frac{1}{2}} w (Q_{\alpha\beta} - Q_{\alpha\beta}^0)^2
+f_s = {\textstyle\frac{1}{2}} w_1 (Q_{\alpha\beta} - Q_{\alpha\beta}^0)^2
 \end{equation}
 where $Q^0_{\alpha\beta}$ is the preferred order parameter configuration
-at the surface (to be either normal or planar).
+at the surface (in the case of normal anchoring or fixed planar anchoring).
+For degenerate planar anchoring,
+following Fournier and Galatola \cite{fournier2005}, we have
+\begin{equation}
+f_s = {\textstyle\frac{1}{2}} w_1 (\tilde{Q}_{\alpha\beta}
+    - \tilde{Q}^\perp_{\alpha\beta})^2
+    + {\textstyle\frac{1}{2}} w_2 (\tilde{Q}_{\alpha\beta}
+                      \tilde{Q}_{\alpha\beta} - S^2_0)^2
+\end{equation}
+where $\tilde{Q}_{\alpha\beta} + (1/3)S_0 \delta_{\alpha\beta}$, $S_0$ is
+is a fixed surface amplitude; to obtain a prefered degenerate
+configuration we take the fluid order parameter $Q_{\alpha\beta}$ and
+project $\tilde{Q}^\perp_{\alpha\beta} = P_{\alpha\gamma}
+\tilde{Q}_{\gamma\sigma}
+P_{\sigma\beta}$, where $P_{\alpha\beta} = \delta_{\alpha\beta}
+- n_\alpha n_\beta$ and $n_\alpha$ is the local unit surface normal.
 
@@
 + \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
-- w(Q_{\alpha\beta} - Q_{\alpha\beta}^0) = 0.
+- w_1(Q_{\alpha\beta} - Q_{\alpha\beta}^0) = 0.
 \label{cholesteric_bc}
 \end{eqnarray}
+Note that for planar generate anchoring, the final term (the `surface
+molecular field') is replaced by
+\begin{equation}
+- w_1 (\tilde{Q}_{\alpha\beta} - \tilde{Q}^\perp_{\alpha\beta})
+- 2w_2 (\tilde{Q}_{\gamma\sigma} \tilde{Q}_{\gamma\sigma} - S^2_0)
+  \tilde{Q}_{\alpha\beta}. 
+\end{equation}
 

### 2013-01-07 trunk/papers/bprheo/bprheo.tex
Davide's revision
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1849 40876441-3496-0410-a613-a2fca35b2ae2

@@
 phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
-II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
-For larger shear rates, the network breaks up into a cholesteric helix, lying either along the
+II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
+For larger shear rates, the structure breaks up into a cholesteric helix, lying either along the
 vorticity or the flow gradient direction according to the imposed flow.
@@
 of less ordered cholesteric structures. Our results provide the first 
-theoretical investigation of sheared blue phases in large systems, which
+theoretical investigation of sheared blue phases in large systems, and
 are relevant to understanding the bulk rheology of these materials. 
@@
 
-Cholesterics are liquid crystals in which the local nematic director field spontaneously 
-twists in thermodynamic equilibrium~\cite{deGennes}. The preferred configuration close 
+Cholesterics are liquid crystals in which the local nematic director field shows spontaneous 
+twist in thermodynamic equilibrium~\cite{deGennes}. 
+The simplest manifestation is the standard cholesteric phase where the
+director precesses around a single helical axis of fixed orientation.
+For highly chiral systems, however,
+the preferred configuration close 
 to the isotropic boundary features twist around two perpendicular axes, as opposed to 
@@
 As it is topologically impossible to cover continuously 3D space with double-twist 
-cylinders, defects arise. These organise into a variety of regular periodic lattices, 
+cylinders, defects arise. The resulting disclination lines (at which the
+nematic director is undefined)
+ organise into a variety of regular periodic lattices, 
 giving rise to the so-called cubic blue phases (BPs)~\cite{Grebel:1984,Wright:1989}. 
@@
 
-BPs have been long considered as purely of academic interest due to their very narrow 
-range of stability. This view has changed since the creation of polymer-stabilised and 
-temperature-stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the 
+BPs were long considered as purely of academic interest due to their very narrow 
+range of stability. This view has changed since the creation of polymer-stabilised and other thermally 
+stabilised BPs~\cite{Kikuchi:2002,Coles:2005}, which has opened up the 
 possibility of novel applications.
@@
 very limited. The aim of this work is to address this issue by studying,
-for the first time, the response of confined BP samples to a shear flow.
+for the first time, the response of large BP samples to a shear flow.
 
 Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
-For example, if a cholesteric helix is subjected to a Poiseuille flow along
-the helical axis, small pressure differences drive flow mainly through
-''permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
+For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
+its helical axis, small pressure differences drive flow mainly through
+``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
 In the permeation mode the liquid crystal flows while leaving the director
@@
 the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
-lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and 
-leaves a flow-induced nematic phase.
+lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase.
 Rey also studied cholesterics subjected to both steady flow and low frequency
@@
 array of doubly twisted cylinders.
-Under small forcing, the network opposed the flow giving rise to a significant 
+Under small forcing, the network opposed the flow, giving rise to a significant 
 increase in apparent viscosity.
@@
 move to neighbouring cells and relink with the parts of the network left behind by the flow. 
-Compared with the cholesteric case, the viscosity still decreases with forcing 
+The viscosity still decreases with forcing 
 (the system shear thins) but much less than
@@
 significantly larger systems than previously possible, so that our
-results can be tested with bulk rheology experiments.
+results can in principle be compared with bulk rheology experiments.
 
@@
 the hydrodynamic equation of motion which we aim to solve. In Section III, 
-we report our numerical results, separating them into subsections relative to
-blue phase I and blue phase II, and corresponding to low, intermediate, and
+we report our numerical results, separating them into subsections referring to
+Blue Phase I and Blue Phase II, and corresponding to low, intermediate, and
 high shear. Finally, we draw our conclusions in Section IV.
@@
 \end{eqnarray}
-The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
-The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
-we work for simplicity in the one-elastic constant approximation. The wavevector $q_0$ is equal to $2\pi/p_0$, where $p_0$ is the cholesteric pitch.
+The first term contains a bulk-free energy constant $A_0$ and the temperature-related parameter $\gamma$ which controls the magnitude of order.
+The second part quantifies the cost of elastic distortions, which is proportional to the elastic constant $K$;
+we work for simplicity in the one-elastic constant approximation~\cite{deGennes}. The wavevector $q_0$ is equal to $2\pi/p_0$, where $p_0$ is the cholesteric pitch.
 The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
-The ``redshift'' $r=p/p_0$ is adjusted during the simulation, to optimise the free energy density
+Indeed, the ``redshift'' $r=p/p_0$ is adjusted during the simulation, to optimise the free energy density
 at all times -- this is done by following the procedure
@@
 
-A thermodynamic state is specified by two dimensionless quantities: the effective temperature 
+A thermodynamic state is specified by two dimensionless quantities: the reduced temperature 
 \begin{equation}
-\tau=\frac{27(1-\gamma/3)}{\gamma}
+\tau=\frac{27(1-\gamma/3)}{\gamma},
 \end{equation}
-and chirality 
+which vanishes at the spinodal point of a nematic ($q_0=0$), 
+and the reduced chirality 
 \begin{equation}
@@
 The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta v_\alpha$,
-and is explicitly given by
+and is explicitly given by~\cite{Beris:1994}
 \begin{eqnarray}
@@
 Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the 
-co-called Leslie angle: we select this case by 
+so-called Leslie angle: we select this case by 
 setting $\xi=0.7$ in our simulations.
-${\mathbf H}$ is the molecular field, which is the functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
+${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that respects the tracelessness of $\mathbf Q$:
 \begin{equation}
@@
 \partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
-&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha v_\beta \nonumber\\
-& &\hspace*{-1.5cm} +\; \partial_\beta v_\alpha
-+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}\}. 
+&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
 \label{NSE}
 \end{eqnarray}
-The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
-of the Lattice Boltzmann (LB) equations \cite{Denniston:2001}.
-At low flow rates the fluid can be considered as incompressible, so that the
-last term on the right hand side of Eq.~\ref{NSE} remains small.
-$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units.
-The pressure tensor reads explicitly
+This emerges from the Chapman-Enskog expansion
+of the Lattice Boltzmann (LB) equations 
+that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta})$ that formally appears
+in this expansion is negligible under the slow flows considered here
+for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}.
+The thermodynamic stress tensor reads explicitly
 \begin{eqnarray}
@@
 \end{eqnarray}
+and is responsible for strong non-Newtonian flow effects.
 In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} reduces to the
-scalar pressure which, in a system at rest, is constant to a very good
-approximation.
-
-It is possible to define a further 
+scalar pressure as would appear in Eq.~\ref{NSE} for a Newtonian fluid.
+
+We next define a 
 dimensionless number that describes the deformation
@@
 \end{equation}
-with $\eta$ and $K$ being the viscosity and elastic constant and
-$v$ and $l$ as typical velocity and length scale. 
-In the present work $l=p_0/2$ was assumed to be the size of the BP unit cell,
-whereas $v$ was taken to be the velocity difference
+with $\eta$ and $K$ defined previously, and 
+$v$ and $l$ a typical velocity and length scale. 
+In the present work $l=p_0/2$ was used as this is the approximate size of
+the BP unit cell. Likewise, $v$ was taken to be the velocity difference
 across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.
 
-The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
+\if{The spectrum $X_\omega$ of the shear stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
 \begin{equation}
@@
 \end{equation}
-where we subtract the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
-$N$ was the number of points in the time series, typically one for every 100 LB time steps.
+where we subtract the average stress $\bar{\Pi}_{xy}$ from the signal.
+$N$ was the number of points in the time series, typically one for every 100 LB time steps. }\fi
+
+In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
+gradient and vorticity direction -- therefore $\Pi_{xy}$ is the shear stress.
 
@@
 
+In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
+phases BPI and BPII. 
+The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
+approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
+is equal to about $10^8$~Pa. Further details about the conversion 
+from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.
+
 \section{Results and Discussion}
+
+For typical simulations reported in this work, 
+we chose as initial conditions thermodynamic states that are 
+well inside the equilibrium region of the individual blue phase, and far away 
+from the cholesteric-isotropic transition. Thus, temperature and chirality were 
+$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
+For these parameters the total free energy density $f$ remained always negative at all flow rates
+simulated. 
+Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}={\mathbf 0}$, this means that our sytem, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
+We performed also selected simulations on metastable states at higher and lower temperatures, 
+and at different chiralities, but did not find any significant differences in the 
+general flow behaviour from that described below.
+The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the disclination 
+network broke up into cholesteric/Newtonian states beyond a critical shear rate
+which was lower.
+This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.
+
+As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
+analytical solutions that minimise the free energy functional Eq.\ref{FE} in the high-chirality limit 
+and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
+During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
+After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow (as 
+it is no longer justified to optimise the free energy in a nonequilibrium scenario).
+We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, and we considered in both cases 
+4 unit cells along each coordinate direction, for a total of 64 unit cells in our simulation box.
+Runs with higher resolution confirmed that this choice was adequate to track  
+all kinematic details of the blue phase networks in shear flow. This includes
+reconstruction of the unit cell not acessible in simulation with only
+few cells~\cite{Dupuis:2005}.
+
+Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
+the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
+velocity gradient along the $y$-direction.
+The shear rates were varied over more than two orders of magnitude from about 
+$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
+For clarity we classify various flow regimes, two in case of BPII and three in case of BPI. 
+They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
+BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
+BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
+Beyond these regimes both BPI and BPII form common configurations regardless 
+of the initial state, namely a cholesteric helix with helical axis 
+oriented along the direction of the velocity gradient (CHG), 
+and a flow-aligned nematic state (FAN).  
+Disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
+Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
+(Note that $q_s$ is small but non-zero at the disclination core.)
+
+\subsection{Blue Phase II}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
+\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd$ (normalised to $\eta$) 
+of BPII versus time (top) and strain (bottom). 
+For clarity the individual curves have been shifted along the time and strain axis. 
+The inset in the top panel 
+shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
+regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
+and break up of the network into a travelling helical wave (BPII-2, solid circle).
+The error bars indicate the maximum and minimum stresses that occur during one cycle.}
+\label{bp2-rheo}
+\end{figure}
+
+We start our discussion with BPII as its flow behaviour is somewhat simpler 
+than that of BPI. BPII has simple cubic symmetry and the disclination lines 
+intersect and form a characteristic network of nodes.
+Ahead of the more detailed discussion we provide first a general overview of the 
+flow behaviour at all applied shear rates.
+Fig. \ref{bp2-rheo} shows the ratio between the apparent viscosity, 
+defined as 
+$\eta_{app}=\langle \Pi_{xy} \rangle/\gd$, and $\eta$.
+A numerical value of $\eta_{app}=1$ corresponds to a fully Newtonian flow,
+without any additional contribution from the liquid crystal.
+The top picture shows the data over time, whereas the graph at the bottom gives 
+the same data versus accumulated strain.
+For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
+is because of the periodic breakup and reconnecting of the network in shear flow. 
+We refer to this regime as BPII-1. As noted in the Appendix, the
+absence at very low shear rate of a permeation mode is due to our
+choice of boundary conditions.
+%{We note that permeation flows, which would
+%{\em not} lead to oscillatory response, are effectively disallowed by the combination
+%of the geometry and boundary condition we consider. This is
+%because the periodic boundary conditions impose macroscopic distortion
+%of the network, and are effective analogous to infinitely distant
+%but anchored wall along the $xy$ plane.
+%This means that  our results are effectively valid for sheared bulk 
+%BPs which are anchored at the walls. Furthermore, permeation flows are 
+%only restricted to very slow flows, and are unstable for intermediate 
+%and high flows, where the response of the network depends less on
+%the details of the boundary conditions used at 
+%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}
+
+The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
+the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
+For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
+$a=0.35, b=0.95$ describes the data to a very good approximation. 
+Hence, the degree of shear-thinning is remarkably small in BPII, in
+the range of shear rates which we have explored here.
+
+For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ ($4\lesssim {\it Er} \lesssim 5.34$) the network of disclination lines
+breaks up completely, and oscillations in the stress signal are absent.
+In this regime of shear rates, to which we refer as BPII-2, the 
+blue phase dissolves into a simple cholesteric liquid 
+crystal. The helical axis is along the vorticity direction and
+the flow excites travelling helical waves, such as those 
+predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis.
+
+At shear rates $7.81\e{-4}\lesssim\gd \lesssim 9.37\e{-4}$ ($6.66\lesssim {\it Er} \lesssim 8$) the helix changes orientation.
+The helical axis is now oriented along the gradient direction with
+the liquid crystal flowing in ``nematic planes''. For even higher shear rates
+the system undergoes a transition to a flow-aligned nematic state.
+
+In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
+by looking at the kinetics of the disclination network.
+
+\subsubsection{Regime BPII-1: low and intermediate shear rates}
+
+Fig. \ref{bp2-med} shows the disclination network in shear flow as
+it undergoes homogeneous shearing in the BPII-1 flow regime. 
+The disclination lines break up and 
+reconnect further downstream, forming a periodically recurring pattern. The
+period needed for a pattern to break up and 
+reform along the flow direction is $\tau_F = 1/\gd$.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.41\textwidth]{disc-160k_run902.png}
+\includegraphics[width=0.41\textwidth]{disc-164k_run902.png}
+\includegraphics[width=0.41\textwidth]{disc-165k_run902.png}
+\caption{Disclination network of BPII in shear flow: 
+The pictures show a typical sequence of snapshots in the steady state 
+at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
+gradient is oriented along the vertical direction, whereas the 
+horizontal direction is the flow direction. Lees-Edwards boundary 
+conditions have been imposed in such a way that network moves to the 
+right in the upper half and to the left in the lower half of the picture.}
+\label{bp2-med}
+\end{figure}
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
+\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
+underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
+The view is along the x-direction. This means the velocity ${\mathbf v}$ 
+has been projected onto a plane perpendicular to the flow direction. 
+The vertical and horizontal direction are the gradient and vorticity direction, respectively.
+The colour code gives the magnitude and sign of the component in vorticity direction.
+The snapshot shows a typical frame during a periodically recurring sequence.
+The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
+with reversed helicity, moves leftward. 
+The periodicity of the motion along the vorticity direction is six times longer than the time 
+it takes the network to reconnect along the flow direction.}
+\label{bp2-velo}
+\end{figure*}
+
+The general appearance of the flowing network is, apart from the homogeneous distortion,
+very close to that of the quiescent blue phase at equilibrium. This holds for all
+shear rates in the BPII-1 regime.
+
+Interestingly, while being displaced with the flow the entire network moves 
+along the vorticity direction as well.  A similar behaviour has been recently 
+observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
+In contrast to the stick-slip motion that has been reported there,
+in this case the movement is steady. 
+The vorticity motion occurs in such a way that the positions of breakup and reconnection 
+point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
+to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
+is $\tau_V=6\tau_F$, 
+i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
+for the network to move one unit cell along $z$ (the vorticity direction).
+
+Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
+cycle, with superimposed velocity vectors.
+The plot shows the secondary velocity components, obtained by projecting the
+velocity onto a plane perpendicular to the flow direction.
+This gets rid of the dominating velocity component along the flow, $v_x$, and 
+allows to visualise the patterns in the 
+two much smaller components $v_y$ and $v_z$.
+The magnitude of the secondary components is typically in the range of a 
+few percent of the primary flow component, for the shear rates 
+and system sizes simulated here.
+
+Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
+oriented along the vorticity direction. The direction of the flow
+in each of the bands depends on the helicity of the underlying cholesteric
+liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
+patterns for left-handed and right-handed BPs are one the mirror image of
+the other. 
+Further quantitative evidence for a direct link between the sense of motion 
+and the helicity can be gained by time-averaging over individual cycles.
+
+Tab. \ref{tab1} in the Appendix 
+gives minima, maxima, averages and standard deviations 
+of the velocity components.
+All values for the two runs with inverted helicity are identical apart 
+from a change of sign in the $z$-components.
+There is only a slight imbalance in magnitude 
+between the maximum and minimum velocities 
+along the $z$-direction. Therefore the averaged flow along the vorticity
+direction, leading to network motion, is very small, and much smaller than
+the local velocity flow. This suggests that the
+flow of the network along the vorticity 
+direction is permeative (i.e. the fluid moves but the network as a whole 
+moves much less).
+
+\subsubsection{Regime BPII-2: high shear rates}
+
+If the flow rate exceeds a certain critical value BPII does not undergo the 
+homogeneous deformation described above and adopts another mode of flow. 
+Fig. \ref{bp2-high} shows a sequence of cuts through the director 
+field at high shear rate. The entire blue phase network has broken up 
+into a simple cholesteric helix and flows as a travelling 
+helical wave with the helical axis oriented along the vorticity direction.
+The state is translationally invariant in the flow and gradient directions.
+It is the preferred mode of flow of a standard cholesteric phase
+in vorticity mode at low Ericksen numbers. This is followed by
+the uncoiling transition to a nematic, flow-aligned state at even 
+higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
+
+In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
+corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
+parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
+see Tab. \ref{tab1} in the Appendix).
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
+\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
+\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
+$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. 
+The configuration is a helical wave that travels along the direction of vorticity.
+It is translationally invariant along the gradient and flow direction.}
+\label{bp2-high}
+\end{figure}
+
+We observe another configuration, lying between the travelling helical wave 
+and the  flow-aligned nematic state at shear rates 
+$7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$ ($6.66\lesssim {\it Er} \lesssim 8$). In this interval
+the helical axis now lies along the flow gradient direction (the so-called
+Grandjean texture~\cite{deGennes}).
+We will address this in section \ref{cholflow}, as it is common to BPI and BPII.
+
+\subsection{Blue Phase I}
+
+BPI has body-centered symmetry and, unlike BPII, the disclination lines
+characterising its equilibrium structure are well separated and do not 
+intersect. (The quiescent state resembles the first frame in Fig.~\ref{bp1-low}
+below.)
+We believe that this topological difference is responsible for most of
+the differences between BPI and BPII regarding their rheological response. 
+We present again key aspects in an overview before we address specific features 
+in more detail.
+
+Fig. \ref{bp1-rheo} shows the ratio between the 
+apparent viscosity $\eta_{app}$ and $\eta$ versus time and 
+strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
+shows that there is more shear thinning in BPI than in BPII, for
+the shear rate range we considered. 
+Through much of the flow range we studied,
+the average shear stress versus shear rate can be fitted reasonably 
+well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
+
+For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
+response versus time, and in the network morphology -- we label this 
+regime BPI-2. 
+The BPI-2 regime now does not persist to arbitrarily low shear: there the
+oscillations become irregular and seemingly chaotic, and after a few
+cycles the disclination pattern becomes amorphous. We refer to this 
+flow regime as BPI-1. 
+The BPI-2 regime is also unstable if the shear is increased {\em above} 
+a critical value -- close to this breakup point the oscillations in the 
+deviatoric stress show an increasingly complex
+pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
+In regime BPI-3 ($6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$) 
+we observe that the BPI network transforms under shear into
+a  quasi two-dimensional state consisting of
+double-twist rolls.
+Finally, for shear rates such that
+$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
+configuration with the helical axis along the gradient direction as 
+reported above for BPII-2 (see
+section \ref{cholflow}).
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
+\caption{Apparent viscosity (normalised by $\eta$) $\eta_{app}/\eta=\langle \Pi_{xy}\rangle/\dot{\gamma}$ of BPI versus time (top) 
+and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
+Depending on the configuration in steady shear flow three different regimes 
+can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
+steady flow with recurring patterns (BPI-2, open circles) and 
+breakup of the disclination network (BPI-3, solid circle). 
+The error bars represent the minimum and maximum shear stress 
+during one cycle. For clarity the individual curves have been shifted along
+the time and strain axis.}
+\label{bp1-rheo}
+\end{figure}
+
+\subsubsection{Regime BPI-1: low shear rates}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
+\caption{Average deviatoric stress and free energy density of BPI at low shear rate: 
+The negative branch in the stress is related to a local maximum and a following local 
+minimum in average free energy density. This creates unstable conditions that lead 
+to a dissolving regular network and an amorphous state which 
+seemingly features a yield stress.}
+\label{bp1-fe-yield}
+\end{figure}
+
+The rheological response of BPI at low shear rate, $\gd\lesssim1.95\e{-5}$
+(${\it Er} \lesssim 0.33$), 
+is strikingly different from that of BPII and appears to show a
+yield stress. An explanation for this behaviour can be found by 
+looking more closely at the average deviatoric stress (where the
+background viscosity contribution has been subtracted) and 
+free energy density as a function of time, as shown in Fig. \ref{bp1-fe-yield}.
+When the quiescent and equilibrated BPI network begins to flow
+the shear stress increases steeply and goes through a maximum.
+Shortly after it goes negative (the total shear stress 
+with the background viscosity contribution remains positive) 
+to become positive again later on, forming a complete cycle.
+The amplitude of the excursions in these stress oscillations, and
+the presence of a part of the period where the thermodynamic
+contribution to the stress is negative, suggest that the
+BPI network is subject to large forces. Seemingly, these eventually
+cause the flow-induced collapse to the amorphous state with
+an apparent yield stress. We should underscore that this
+low shear regime, might be brought about by the boundary
+conditions, which in practice is equivalent to having infinitely
+distant walls which the network is anchored to (see discussion
+in the Appendix).
+With free boundary conditions,
+one may imagine that, for very low wall velocity, 
+permeative flows may be possible which allow for
+some slip of the BPI network, as in cholesterics sheared
+along their axis~\cite{Marenduzzo:2006b}.
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115.png}\\
+\includegraphics[width=0.245\textwidth]{disc-xy-400k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-700k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-750k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2200k_run1115.png}
+\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
+Depicted is the transition from the quiescent state to flow-induced, intertwined 
+helices that undergo a recurring structural transformation. The top row
+shows a section of one unit cell for early times 
+$t=1\e{4}, 1.8\e{5}, 2.0\e{5}$ and $2.1\e{5}$. The bottom 
+pictures on the far left, centre left and centre right 
+show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
+which the system passes for several cycles before it ends up in
+an amorphous state (bottom row far right, at time step $t=2.2\e{6}$).}
+\label{bp1-low}
+\end{figure*}
+
+\if{The negative branch in the deviatoric stress is connected
+to a local maximum followed by a local minimum in the 
+average free energy. Hence, the region of negative stress can be 
+regarded as a phase during which elastic forces exert a 
+'pull' on the network while it tries to reach a more 
+favourable configuration. However, the low flow rate
+causes this unstable intermediate state to be exposed to these
+elastic forces for a relatively long time, long enough to 
+create small distortions which finally break the symmetry of the
+flow-induced configuration.}\fi
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
+\caption{Shear stress $\Pi_{xy}$ versus strain near the transition between regime
+BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
+and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
+where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
+throughout positive, but exhibits very large fluctuations during one cycle, 
+which are absent in regime BPI-2 at higher shear rates than shown here.}
+\label{bp1-1_bp1-2}
+\end{figure}
+
+To gain further insight into the flow-induced deformation, and eventual
+breakdown, of BPI, it is instructive to follow the dynamics of 
+the network under a slow flow -- this is depicted in Fig. \ref{bp1-low}.
+Three different stages can be distinguished. 
+Just after the onset of the shear flow, the disclination lines 
+in BPI get more and more squeezed together, and this is incompatible with
+the defect topology in this phase.
+Consequently, the network adopts a flow-induced conformation which consists 
+of intertwined helices that stretch during the affine transformation.
+This helical conformation that emerges already at strains $\gamma\simeq1$ 
+is shown in Fig. \ref{bp1-low} (top row).
+At this point the original BPI has already been significantly deformed.
+Shortly afterwards regular oscillations set up temporarily,
+where the disclinations form double helices which tilt and realign under the shear.
+There is no perceivable movement of the 
+network along the vorticity direction at any stage of the cycle.
+(This contrasts with the BPII-2 results reported earlier.)
+
+This mode of flow proves unstable, as after a few cycles 
+distortions appear which lead to further destabilisation -- presumably in
+view of the large stress fluctuations discussed above. 
+Finally, the network loses order and 
+transforms into an amorphous state with almost constant
+yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ LBU.
+If the shear is stopped, this flow-induced amorphous state
+remains arrested and metastable, and cannot
+find its way back to the original BPI structure.
+
+\subsubsection{Regime BPI-2: intermediate shear rates}
+
+Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$; $ {\it Er}\lesssim 0.33$) and 
+at slightly larger shear rates ($3.91\e{-5}\lesssim\gd\lesssim 4.687\e{-4}$; $0.67\lesssim{\it Er}\lesssim8$)
+lies another region where the network flows with periodically 
+recurring conformations (Fig. \ref{bp1-rheo}). 
+We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
+is clear when looking at Fig. \ref{bp1-1_bp1-2} 
+which shows the deviatoric stress versus total strain. 
+A qualitative difference between these two is the absence in BPI-2 of the 
+large stress fluctuations that occur during the early cycles in regime BPI-1.
+(Recall that for even lower $\gd$ the deviatoric thermodynamic stress becomes 
+temporarily negative (Fig. \ref{bp1-fe-yield}), which 
+caused destabilisation of the periodic network and led to the 
+amorphous configuration in the steady state.)
+Hence, an explanation
+for the existence of the regular BPI-2 oscillations 
+could be that these are fast enough to bypass or suppress 
+large stress fluctuations -- this leads to
+a different order structure and to a stabilisation 
+compared to the oscillations found transiently in
+the BPI-1 regime.
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
+\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
+\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
+\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
+\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
+\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
+\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
+The sequence shows a typical cycle of shear-induced transformations in the 
+steady state at $\gd=1.56\e{-4}$ and time steps 
+$t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
+During every cycle the network is displaced also along the gradient and vorticity direction, 
+but the magnitude and direction of the offset does not appear to obey a 
+simple rule.}
+\label{bp1-med}
+\end{figure*}
+
+Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
+BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
+BPI-2 does not resemble the equilibrium configuration
+undergoing a homogeneous topology-conserving 
+transformation. It features rather intricate, flow-induced 
+conformations. Surprisingly, if the shear is switched off in
+the BPI-2 regime at some point during an oscillation, 
+the flow-induced configuration usually
+reverts to a perfect quiescent BPI -- in stark contrast with
+what happens in BPI-1, for lower shear rates.
+
+Finding the recurrence period for BPI is not simple, as
+this depends subtly on the applied shear rate. The analysis is complex for
+a number of reasons. 
+First, the recurrence period for the stress is not always the same as
+that of the network configuration. 
+For instance, at $\gd=3.9062500\e{-5}$, it takes $2/\gd$ for one
+ stress cycle to complete, but $4/\gd$ for the network to recur --
+this is because the configurations at times $t$ and $t+2/\gd$ are 
+mirror-images. This is by no means generic though: for example
+at $\gd=4.8828125\e{-5}$  it takes $8/\gd$ for one
+stress cycle to complete, but $16/\gd$ for the network to recur. 
+For some higher shear rates, the recurrence period of stress
+and network conformations are instead the same.
+Second, there exist isolated values (or perhaps small ranges) of shear
+rates for which the recurrence period is much longer and far more irregular,
+with seemingly intermittent dynamics (e.g. at $\gd=1.3671875\e{-4}$ the
+network configuration repeats itself after $6/\gd$, $12/\gd$, $22/\gd$).
+Taken together, all this evidence very much suggests that 
+BPI-2 constitutes an instance of ``rheochaos''~\cite{rheochaos}.
+
+Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ for left- and right-handed chirality of
+the blue phase. The emerging pattern is similar to that of BPII from 
+Fig. \ref{bp2-velo}, but differs in that there is no band
+arrangement. However, just as in the case of BPII, both patterns
+are mirror images and feature the highest velocities in the regions where 
+the largest conformational changes take place.
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
+The pictures show a snapshot of the periodically recurring patterns in the 
+secondary velocity components $(0,v_y,v_z)$, i.e. the component $v_x$ in flow direction 
+has been projected out. The colour code gives the magnitude and sign 
+of the $z$-component. The vertical and horizontal direction are the gradient and 
+vorticity direction, respectively, with the flow into/out of the plane in the 
+upper/lower half of the images.}
+\label{bp1-velo}
+\end{figure*}
+
+\if{In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
+we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
+The spectrum is shown in Fig. \ref{bp1-spectrum}.
+
+There are two aspects which are worth mentioning. First, we observe a strong contribution of 
+the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
+However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
+significantly larger. 
+Second, more surprisingly, the frequency of the fundamental mode is the same 
+for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
+in contrast to what is observed at lower flow rates. By doubling the flow rate from 
+$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
+moves to double the frequency. This means that the fundamental mode at flow rate $\gd=7.81\e{-5}$ 
+coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on. 
+At higher shear rates in regime BPI-3 this 'lock-in' of flow modes disappears again.
+
+Interestingly the lock-in has only a minor effect on the flow velocities themselves.
+The average velocities in Tab. \ref{tab1} (see Appendix) reveal that only the magnitude of some of the secondary 
+flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
+The latter is directly linked to the imposed shear rate. }\fi
+
+\if{
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
+\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
+spectral components and the 'lock-in' of flow modes, 
+visible in the fixed peak position for certain flow rates.  
+The inset shows a zoom of the same data over 
+a portion of the frequency range.
+The frequency is given in inverse time, in simulation units. TRUE?
+OH: I WOULD SAY NO AS IT IS IN UNITS OF 2$\pi$ (RADIANS?).}
+\label{bp1-spectrum}
+\end{figure}
+}\fi
+
+The intricacy and periodic recurrence of the flow-induced conformations 
+in BPI raises the question: how robust are these results to variations in 
+system size? Although a pitch length of 64 LBU provides enough resolution 
+for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
+in total are actually sufficient to avoid finite size effects.
+Unfortunately the computational costs of larger simulations are still
+prohibitive, given that large strains have to
+be simulated in order to reach a steady state. Therefore we opted for a 
+comparison with smaller runs.
+
+Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
+deviatoric stress $\Pi_{xy}$ of two different runs, one consisting of 8 
+unit cells and the other one with 64 unit cells.
+We observe that there are deviations in both the visual appearance of the 
+network as well as in the time-dependence of the deviatoric stress.
+Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
+to reach the steady state with its flow-induced recurring disclination network. 
+These results indicate that we cannot know with certainty if the individual 
+flow-induced conformations that we find for 64 unit cells are not affected 
+by the imposed periodicity -- quite possibly the details of the oscillations
+would actually differ in an even large simulation. 
+Nevertheless, the comparison suggests that 
+the recurrence period and the complexity of the network in flow are  
+generic features of BPI in this regime of flow.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
+\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
+\caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
+The dark (red) network in top picture shows a snapshot of run that consists 
+of $2^3=8$ unit cells, whereas the light (cyan) network displays the 
+central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
+The bottom picture shows the deviatoric stress over time. (Interestingly, 
+an intermediate sized run with $3^3=27$ unit cells led to results that 
+were virtually identical to the smaller run with 8 unit cells.) 
+The shear rate was $\gd=7.81\e{-5}$.
+}
+\label{bp1-2uc4uc}
+\end{figure}
+
+\subsubsection{Regime BPI-3: high shear rates}
+
+At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
+we observe a break-up of the flow-induced 
+disclination network of regime BPI-2. However, 
+in contrast to BPII, where the configuration is a travelling helical 
+wave (cf. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
+two-dimensional, translationally invariant state, which is shown in 
+Fig. \ref{bp1-high}. 
+There are distinctive regions where the director field is 
+predominantly oriented along the vorticity direction and 
+other regions where it is oriented rather in flow-gradient plane. 
+Occasionally while moving past each other these rolls resemble 
+regions of local double twist.
+Interestingly the critical Ericksen number for the break-up lies in 
+about the same window as that of transition between BPII-1 and BPII-2.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
+\caption{Director field of BPI at high shear rate: The pictures show a region 
+of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
+direction oriented along the horizontal/vertical axis. 
+The colour code gives the magnitude of the z-component of the director field. 
+The data show that the structure is quasi-2D, with rolling double twist
+patterns in the $xy$ plane, and an approximate translational symmetry 
+along the vorticity direction.}
+\label{bp1-high}
+\end{figure}
+
+\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{cholflow}
+
+At shear rates beyond the regimes BPI-3 and BPII-2, but below the transition to a 
+flow-aligned nematic state at still higher shear rates, we found another regime where 
+both blue phases adopt the same configuration in steady shear flow, 
+independent of their initial state.
+The director field of this configuration is shown in Fig. \ref{fig:cholflow}.
+It consists of a simple cholesteric helix with the helical axis oriented 
+along the velocity gradient direction. The state is translational invariant 
+along the flow and vorticity direction and therefore also one-dimensional. 
+However, in contrast to the travelling helical wave of BPII-2 the 
+director field is now static and there is no tumbling motion. 
+While flowing the director retains its relative orientation.
+Consequently, the shear stress is lower than for
+the 2D double-twist rolls of BPI-3 and the travelling helical wave
+(Fig. \ref{bp2-rheo} and \ref{bp1-rheo}) as there is less dissipation.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
+\caption{Flow induced cholesteric state. The orientation of the helix is 
+along the gradient direction. The director field is quasi-static during 
+the flow, which occurs in ``nematic layers''. Color coding indicates the 
+$x$ component of the director field (see colorbar on the right).
+For clarity only selected sites are shown along the $y$-direction (along which
+the structure is translationally invariant).}
+\label{fig:cholflow}
+\end{figure}
+
+%\clearpage
+
+\section{Conclusions}
+
+In summary, our work constitutes the first simulation of bulk flow behaviour 
+of cubic BPs in simple shear flow.
+
+We were able to characterise the rheology of cubic BPs, and identified 
+two different flow regimes for BPII and three different ones for BPI. 
+Below an Ericksen number of $Er\simeq3$, BPII exhibits weak shear-thinning and 
+obeys a power-law flow curve.  
+The BPII disclination network breaks up and reconnects 
+in the flow, which leads to a periodically recurring dependence 
+of the deviatoric shear stress. The flow-induced conformation looks generally
+very similar to the quiescent network at equilibrium. 
+While being homogeneously transformed due to the shear flow
+the disclination network moves steadily in the vorticity direction
+apparently by a permeation mechanism. The sense of
+motion is directly linked to the helicity of the underlying cholesteric phase.
+
+At larger Ericksen numbers, $4\lesssim Er\lesssim 10$, the flowing BPII network 
+breaks up in a two-step manner. First a simple cholesteric helix forms with the helical axis 
+along the vorticity direction. The flow excites a travelling helical wave,
+as predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}.
+Upon increasing the shear, the helix rotates and orients along the
+gradient direction, where it can support larger flows due to the translational
+symmetry along the flow direction. Still larger flows break this residual
+cholesteric order and leave a standard flow-aligned nematic state,
+at the highest shear rate studied.
+
+Interestingly, BPI shows a flow behaviour that is very different to that of BPII.
+This is a direct consequence of the topological differences between their disclination 
+networks. Below an Ericksen number of $Er\sim 0.4$, the BPI network cannot both flow and 
+retain a regular appearance. It eventually reorganises into an amorphous network that
+features yield-stress behaviour. The apparent reason is that shortly after the onset of the shearing 
+very large stress fluctuations occur. At the lowest shear rates
+the thermodynamic contribution to the deviatoric (shear) stress becomes temporarily 
+negative during each cycle. These fluctuations seem to destabilise the network and 
+eventually trigger the transition into an amorphous state with a residual yield
+stress.
+
+At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim8$,
+BPI kinematically bypasses the large stress fluctuations and 
+flows with very complex, periodically recurring flow-induced conformations.
+These conformations entail regularly arrange helical disclinations which 
+rotate due to the shear -- a new flow-induced state which was not previously
+predicted. 
+The oscillatory dynamics is in some cases intermittent, and the
+recurrence period for the network conformations and stress patterns
+is not given by a simple formula as for BPII.
+Despite their complex appearance, the flow-induced conformations are 
+topologically connected to the quiescent BPI and after switching off of the 
+shear flow a defect-free blue phase reforms. It is tempting to
+interpret all these finding as an instance of 
+deterministic rheological chaos~\cite{fielding}.
+
+Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
+a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
+quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
+Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
+direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
+is a flow-aligned nematic state.
+Although experimental evidence to support
+our results is currently not available, we hope this work will inspire such experiments, and believe it can shed some light 
+on the flow properties of complex liquid-crystalline phases.
+
+% If in two-column mode, this environment will change to single-column
+% format so that long equations can be displayed. Use
+% sparingly.
+%\begin{widetext}
+% put long equation here
+%\end{widetext}
+
+% figures should be put into the text as floats.
+% Use the graphics or graphicx packages (distributed with LaTeX2e)
+% and the \includegraphics macro defined in those packages.
+% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz,
+% and Frank Mittelbach for instance.
+%
+% Here is an example of the general form of a figure:
+% Fill in the caption in the braces of the \caption{} command. Put the label
+% that you will use with \ref{} command in the braces of the \label{} command.
+% Use the figure* environment if the figure should span across the
+% entire page. There is no need to do explicit centering.
+
+% \begin{figure}
+% \includegraphics{}%
+% \caption{\label{}}
+% \end{figure}
+
+% Surround figure environment with turnpage environment for landscape
+% figure
+% \begin{turnpage}
+% \begin{figure}
+% \includegraphics{}%
+% \caption{\label{}}
+% \end{figure}
+% \end{turnpage}
+
+% tables should appear as floats within the text
+%
+% Here is an example of the general form of a table:
+% Fill in the caption in the braces of the \caption{} command. Put the label
+% that you will use with \ref{} command in the braces of the \label{} command.
+% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
+% \begin{tabular}{} command.
+% The ruledtabular enviroment adds doubled rules to table and sets a
+% reasonable default table settings.
+% Use the table* environment to get a full-width table in two-column
+% Add \usepackage{longtable} and the longtable (or longtable*}
+% environment for nicely formatted long tables. Or use the the [H]
+% placement option to break a long table (with less control than 
+% in longtable).
+% \begin{table}%[H] add [H] placement to break table across pages
+% \caption{\label{}}
+% \begin{ruledtabular}
+% \begin{tabular}{}
+% Lines of table here ending with \\
+% \end{tabular}
+% \end{ruledtabular}
+% \end{table}
+
+% Surround table environment with turnpage environment for landscape
+% table
+% \begin{turnpage}
+% \begin{table}
+% \caption{\label{}}
+% \begin{ruledtabular}
+% \begin{tabular}{}
+% \end{tabular}
+% \end{ruledtabular}
+% \end{table}
+% \end{turnpage}
+
+% Specify following sections are appendices. Use \appendix* if there
+% only one appendix.
+%\appendix
+
+%\section{}
+
+% If you have acknowledgments, this puts in the proper section head.
+\begin{acknowledgments}
+We acknowledge support by EPSRC grant nos. EP/E045316 and EP/E030173, 
+the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
+We thank Peter J. Collings for very stimulating discussions. 
+MEC holds a Royal Society Research Professorship.
+\end{acknowledgments}
+
+\appendix
+\section{}
+
+In this Appendix we give some more details on our simulations,
+including the importance and effect of boundary conditions.
+
+Tab. \ref{tab1} lists the parameters chosen for our runs (values of
+the reduced temperature and reduced chirality are given in the caption),
+together with the minima, maxima, and standard deviations of the velocity
+field in the three directions (discussed further in the main text). 
+The first column also describes what flow regime each simulation
+leads to in steady state.
 
@@
 \hline
-BPI \\
+BPI & & & & $\times 10^5$\\
 \hline
@@
 
-BPII \\
+BPII  & & & & $\times 10^5$ \\
 \hline
@@
 for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
-shear flow: All values are given in $10^{-5}$ lattice units. The transient dynamics 
-has not been included in the averaging procedure. The regimes comprise the formation
+shear flow. All velocity values are given in $10^{-5}$ lattice units. The transient dynamics 
+has not been included in the averages. The regimes comprise the formation
 of an amorphous network (BPI-1), periodically recurring conformations (BPI-2 and BPII-1), 
@@
 a cholesteric helix oriented along the gradient direction (CHG), 
-and a flow-aligned nematic state (FAN).}
+and a flow-aligned nematic state (FAN).
+(Last six columns refer to the secondary flow.)}
 \label{tab1}
@@
 
-In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
-phases BPI and BPII. 
-The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
-approximately to $\sim 1 {\rm ns}$ and $\sim 10{\rm nm}$ in SI units, respectively. The LB unit of stress
-is equal to about $10^8$~Pa. Further details about the conversion 
-from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.
-
-For typical simulations reported in this work, 
-we chose as initial conditions thermodynamic states that are 
-well inside the equilibrium region of the individual blue phase, and far away 
-from the cholesteric-isotropic transition. Thus, temperature and chirality were 
-$\tau=-0.5, \kappa=1.0$ in case of BPI and $\tau=-0.5, \kappa=2.0$ for BPII, respectively.
-For these parameters the total free energy density $f$ remained always negative at all flow rates
-simulated (so that the system is always in a liquid crystalline state). 
-We performed also selected simulations at metastable states at higher and lower temperatures, 
-and at different chiralities, but did not find any significant differences in the 
-general flow behaviour that is described below.
-However, for thermodynamic states that are closer to the phase boundary the disclination 
-network broke up into non-BP states at lower shear rates. This is not surprising as
-these states have on average higher free energy densities and smaller order parameters
-than those specified above and cannot therefore withstand the same external forces
-before breaking down.
-
-As usual in BP simulation studies~\cite{Henrich:2011a,Henrich:2010b}, we initialised our runs with 
-analytical solutions that minimise the free energy functional Eq.\ref{FE} in the high-chirality limit 
-and equilibrated these configuration for 5000 LB timesteps before we started the shear flow. 
-During the equilibration sequence the optimal redshift $r$ was calculated and applied at every timestep.
-After equilibration the redshift was kept constant throughout the rest of the simulation with shear flow (as 
-it is no longer justified to optimise the free energy in a nonequilibrium scenario).
-We chose a pitch length of 32 LBU for BPII and 64 LBU in case of BPI, and we considered in both cases 
-4 unit cells along each coordinate direction, for a total of 64 unit cells in our simulation box.
-Runs with higher resolution confirmed that this choice was adequate to track  
-all kinematic details of the blue phase networks in shear flow.
-
-Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
-the top (bottom) part of the system flowing in positive (negative) x-direction and the 
-velocity gradient along the y-direction.
-The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
-For clarity we classify various flow regimes, two in case of BPII and three in case of BPI. 
-They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$),
-BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
-Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical choices 
-are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-
-\subsection{Blue Phase II}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII versus time (top) and strain (bottom). 
-For clarity the individual curves have been shifted along the time and strain axis. 
-The inset in the top panel 
-shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
-regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
-and break up of the network into a travelling helical wave (BPII-2, solid circle).}
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
-\label{bp2-rheo}
-\end{figure}
-
-We start our discussion with BPII as its flow behaviour is somewhat simpler 
-than that of BPI. BPII has simple cubic symmetry and the disclination lines 
-intersect and form a characteristic network of nodes.
-Ahead of the more detailed discussion we provide first a general overview of the 
-flow behaviour at all applied shear rates.
-Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
-deviatoric stress normalised to the isotropic background viscosity: 
-$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
-A numerical value of $\eta_{app}=1$ corresponds to a fully Newtonian flow,
-without any additional contribution from the liquid crystal.
-The top picture shows the data over time, whereas the graph at the bottom gives 
-the same data versus accumulated strain.
-For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is because of the periodic breakup and reconnecting of the network in shear flow. 
-We refer to this regime as BPII-1. We note that permeation flows, which would
-{\em not} lead to oscillatory response, are effectively disallowed by the combination
-of the geometry and boundary condition we consider, because at the Lees-Edwards 
-plane the network needs to reconnect with its Galilean ``image'' travelling
-at the speed specified by the shear. This means that  our results are effectively
-restricted to intermediate or high flow, where permeation flows are 
-absent~\cite{Marenduzzo:2006a,Marenduzzo:2006b} and the response of the network
-is indeed oscillatory irrespective of the boundary conditions used at the wall. 
-We also note that, if the boundaries are such that
-the director field pattern sticks to the shear planes, then our analysis
-would apply and be realistic for arbitrarily low shear as well.
-
-The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
-the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
-For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
-$a=0.35, b=0.95$ describes the data to a very good approximation. 
-Hence, the degree of shear-thinning is remarkably small in BPII, in
-the range of shear rates which we have explored here.
-
-For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ the network of disclination lines
-breaks up completely, and oscillations in the stress signal are absent.
-In this regime of shear rates, to which we refer as BPII-2, the 
-blue phase dissolves into a simple cholesteric liquid 
-crystal. The helical axis is along the vorticity direction and
-the flow excites travelling helical waves, such as those 
-predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
-the vorticity axis.
-
-At shear rates $7.81\e{-4}\lesssim\gd9.37\e{-4}$ the helix changes orientation.
-The helical axis is now oriented along the gradient direction with
-the liquid crystal flowing in 'nematic planes'. For even higher shear rates
-the system undergoes a transition to a flow-aligned nematic state.
-
-In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
-by looking at the kinetics of the disclination network.
-
-\subsubsection{Regime BPII-1: low and intermediate shear rates}
-
-Fig. \ref{bp2-med} shows the disclination network in shear flow as
-it undergoes an affine transformation in the BPII-1 flow regime. 
-The disclination lines break up and 
-reconnect further downstream, forming a periodically recurring pattern. The
-period needed for a pattern to break up and 
-reform along the flow direction is $\tau_F=1/\gd$.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.41\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.41\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.41\textwidth]{disc-165k_run902.png}
-\caption{Disclination network of BPII in shear flow: 
-The pictures show a typical sequence of snapshots in the steady state 
-at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
-gradient is oriented along the vertical direction, whereas the 
-horizontal direction is the flow direction. Lees-Edwards boundary 
-conditions have been imposed in such a way that network moves to the 
-right in the upper half and to the left in the lower half of the picture.}
-\label{bp2-med}
-\end{figure}
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
-\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
-underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$.
-The view is along the x-direction. This means the velocity ${\mathbf v}$ 
-has been projected onto a plane perpendicular to the flow direction. 
-The vertical and horizontal direction are the gradient and vorticity direction, respectively.
-The colour code gives the magnitude and sign of the component in vorticity direction.
-The snapshot shows a typical frame during a periodically recurring sequence.
-The network on the left with right-handed helicity ($q_0>0$ in Eq.~\ref{FE}) travels rightwards, whereas the one one the right,
-with reversed helicity, moves leftward. 
-The periodicity of the motion along the vorticity direction is six times longer than the time 
-it takes the network to reconnect along the flow direction.}
-\label{bp2-velo}
-\end{figure*}
-
-The general appearance of the flowing network is, apart from the affine transformation, 
-very close to that of the quiescent blue phase at equilibrium. This holds for all
-shear rates in the BPII-1 regime.
-
-Interestingly, while being displaced with the flow the entire network moves 
-along the vorticity direction as well.  A similar behaviour has been recently 
-observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
-In contrast to the stick-slip motion that has been reported there,
-in this case the movement is steady. 
-The vorticity motion occurs in such a way that the positions of breakup and reconnection 
-point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
-to travel along the z-direction. The periodicity of the motion along the vorticity direction
-is $\tau_V=6\tau_F$, 
-i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
-for the network to move one unit cell along $z$ (the vorticity direction).
-
-Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
-cycle, with superimposed velocity vectors.
-The plot shows the secondary velocity components, obtained by projecting the
-velocity onto a plane perpendicular to the flow direction.
-This gets rid of the dominating velocity component along the flow, $v_x$, and 
-allows to visualise the patterns in the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the secondary components is typically in the range of a 
-few percent of the primary flow component, for the shear rates simulated here.
-
-Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
-oriented along the vorticity direction. The direction of the flow
-in each of the bands depends on the helicity of the underlying cholesteric
-liquid crystal (i.e. on the sign of $q_0$ in Eq.~\ref{FE}), and the velocity
-patterns for left-handed and right-handed BPs are one the mirror image of
-the other. 
-Further quantitative evidence for a direct link between the sense of motion 
-and the helicity can be gained by time-averaging over individual cycles.
-Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
-of the velocity components.
-All values for the two runs with inverted helicity are identical apart 
-from a change of sign in the z-components.
-There is only a slight imbalance between the maximum and minimum velocities 
-in z-direction, suggesting that the flow of the network along the vorticity 
-direction is permeative.
-
-\subsubsection{Regime BPII-2: high shear rates}
-
-If the flow rate exceeds a certain critical value BPII does not undergo the 
-affine transformation  above and adopts another mode of flow. 
-Fig. \ref{bp2-high} shows a sequence of cuts through the director 
-field at high shear rate. The entire blue phase network has broken up 
-into a simple cholesteric helix and flows as a travelling 
-helical wave with the helical axis oriented along the vorticity direction.
-The state is translationally invariant in flow and gradient direction.
-and is the preferred mode of flow of a cholesteric 
-liquid crystal in vorticity mode at low Ericksen numbers, i.e. below 
-the uncoiling transition to a nematic, flow-aligned state at even 
-higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
-
-In our scenario the break-up occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
-corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
-parameters for BPII ($K=0.02, \eta=0.666$, Tab. \ref{tab1} and $p_0/2=16$).
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
-$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. 
-The configuration is a helical wave that travels along the direction of vorticity.
-It is translationally invariant along the gradient and flow direction.}
-\label{bp2-high}
-\end{figure}
-
-We observe another configuration between the travelling helical wave and the 
-flow-aligned nematic state at shear rates $7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$, 
-where the helical axis is now along the flow gradient direction (the so-called
-Grandjean texture).
-We will address this in section \ref{cholflow}, as it is common to BPI and BPII.
-
-\subsection{Blue Phase I}
-
-BPI has body-centered symmetry and, unlike BPII, the disclination lines
-characterising its equilibrium structure are well separated and do not intersect.
-We believe that this topological difference is responsible for most of
-the differences between BPI and BPII regarding their rheological response. 
-We present again key aspects in an overview before we address specific features 
-in more detail.
-
-Fig. \ref{bp1-rheo} shows the apparent viscosity $\eta_{app}$ versus time and 
-strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
-shows that there is more shear thinning in BPI than in BPII, for
-the shear rate range we considered. 
-Throughout the flow range we studies,
-the average shear stress versus shear rate can be fitted reasonably 
-well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
-
-For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
-response versus time, and in the network morphology -- we label this 
-regime BPI-2. 
-The BPI-2 regime now does not persist to arbitrarily low shear: there the
-oscillations become irregular and seemingly chaotic, and after a few
-cycles the disclination pattern becomes amorphous. We refer to this 
-flow regime as BPI-1. 
-The BPI-2 regime is also unstable if the shear is increased {\em above} 
-a critical value -- close to this breakup point the oscillations in the 
-deviatoric stress show an increasingly complex
-pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
-In regime BPI-3 $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ 
-we observe that the BPI network transforms under shear into
-a  quasi two-dimensional state consisting of
-double-twist rolls.
-Finally, for shear rates $9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ we observe the same cholesteric 
-configuration with the helical axis along the gradient direction as for BPII (see
-section \ref{cholflow}).
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI versus time (top) 
-and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
-Depending on the configuration in steady shear flow three different regimes 
-can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
-steady flow with recurring patterns (BPI-2, open circles) and 
-breakup of the disclination network (BPI-3, solid circle). 
-The error bars represent the minimum and maximum shear stress 
-during one cycle. For clarity the individual curves have been shifted along
-the time and strain axis.}
-\label{bp1-rheo}
-\end{figure}
-
-In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
-we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
-The spectrum is shown in Fig. \ref{bp1-spectrum}.
-
-There are two aspects which are worth mentioning. First, we observe a strong contribution of 
-the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
-However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
-significantly larger. 
-Second, more surprisingly, the frequency of the fundamental mode is the same 
-for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
-in contrast to what is observed at lower flow rates. By doubling the flow rate from 
-$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
-moves to double the frequency. This means that the fundamental mode at flow rate $\gd=7.81\e{-5}$ 
-coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on. 
-At higher shear rates in regime BPI-3 this 'lock-in' of flow modes disappears again.
-
-Interestingly the lock-in has only a minor effect on the flow velocities themselves.
-The average velocities in Tab. \ref{tab1} reveal that only the magnitude of some of the secondary 
-flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
-The latter is directly linked to the imposed shear rate.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
-spectral components and the 'lock-in' of flow modes, 
-visible in the fixed peak position for certain flow rates.  
-The inset shows a zoom of the same data over 
-a portion of the frequency range.
-The frequency is given in inverse time, in simulation units. TRUE?}
-\label{bp1-spectrum}
-\end{figure}
-
-\subsubsection{Regime BPI-1: low shear rates}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{Average deviatoric stress and free energy density of BPI at low shear rate: 
-The negative branch in the stress is related to a local maximum and a following local 
-minimum in average free energy density. This creates unstable conditions that lead 
-to a dissolving regular network and an amorphous state which 
-seemingly features a yield stress.}
-\label{bp1-fe-yield}
-\end{figure}
-
-The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
-is strikingly different from that of BPII and appears to show a
-yield stress. An explanation for this behaviour can be found by 
-looking more closely at the average deviatoric stress (where the
-background viscosity contribution has been subtracted) and 
-free energy density as a function of time, as shown in Fig. \ref{bp1-fe-yield}.
-When the quiescent and equilibrated BPI network begins to flow
-the shear stress increases steeply and goes through a maximum.
-Shortly after it goes negative (the total shear stress 
-with the background viscosity contribution remains positive) 
-to become positive again later on, forming a complete cycle.
-The amplitude of the excursions in these stress oscillations and
-the presence of a part of the period where the thermodynamic
-contribution to the stress is negative suggest that the
-BPI network is subject to large forces which eventually 
-cause the flow-induced collapse to the amorphous state with
-an apparent yield stress. We should underscore that this,
-low shear, regime, might be brought about by the boundary
-conditions, which in practice lead to the network sticking to
-the Lees-Edwards plane. With walls and free boundary conditions,
-one may imagine that, for very low wall velocity, 
-permeative flows may be possible which leave the BPI 
-structure almost unchanged, as in cholesterics sheared
-along their axis~\cite{Marenduzzo:2006b}.
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115.png}\\
-\includegraphics[width=0.245\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-700k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-2200k_run1115.png}
-\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
-Depicted is the transition from the quiescent state to flow-induced, intertwined 
-helices that undergo a recurring affine transformation. The top row
-shows a section of one unit cell for early times 
-$t=1\e{4}, 1.8\e{5}, 2.0\e{5}$ and $2.1\e{5}$. The bottom 
-pictures on the far left, centre left and centre right 
-show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
-which the system passes for several cycles before it ends up in
-an amorphous state (bottom row far right at time step $t=2.2\e{6}$).}
-\label{bp1-low}
-\end{figure*}
-
-\if{The negative branch in the deviatoric stress is connected
-to a local maximum followed by a local minimum in the 
-average free energy. Hence, the region of negative stress can be 
-regarded as a phase during which elastic forces exert a 
-'pull' on the network while it tries to reach a more 
-favourable configuration. However, the low flow rate
-causes this unstable intermediate state to be exposed to these
-elastic forces for a relatively long time, long enough to 
-create small distortions which finally break the symmetry of the
-flow-induced configuration.}\fi
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{Deviatoric stress $\Pi_{xy}$ versus strain at the transition between regime
-BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
-and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
-where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
-throughout positive, but exhibits very large fluctuations during one cycle, 
-which are absent in regime BPI-2 at higher shear rates.}
-\label{bp1-1_bp1-2}
-\end{figure}
-
-To gain further insight into the flow-induced deformation, and eventual
-breakdown, of BPI, it is instructive to follow the dynamics of 
-the network under a slow flow -- this is depicted in Fig. \ref{bp1-low}.
-Three different stages can be distinguished. 
-Just after the onset of the shear flow, the disclination lines 
-in BPI get more and more squeezed together, and this is incompatible with
-the defect topology in this phase.
-Consequently, the network adopts a flow-induced conformation which consists 
-of intertwined helices that stretch during the affine transformation.
-This helical conformation that emerges already at strains $\gamma\simeq1$ 
-is shown in Fig. \ref{bp1-low} (top row).
-At this point the original BPI has already been significantly deformed.
-Shortly afterwards regular oscillations set up temporarily,
-where the disclinations form double helices which tilt and realign under the shear
-TRUE? There is no perceivable movement of the 
-network along the vorticity direction at any stage of the cycle.
-
-This mode of flow proves unstable as after a few cycles 
-distortions appear which lead to further destabilisation -- presumably in
-view of the large stress fluctuations discussed above. 
-Finally, the network loses order and 
-transforms into an amorphous state with almost constant
-yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
-If the shear is stopped, this flow-induced amorphous state
-gets stuck into a metastable phase, and cannot
-find its way back to the original BPI structure.
-
-\subsubsection{Regime BPI-2: intermediate shear rates}
-
-Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) and 
-at slightly larger shear rates $3.91\e{-5}\le\gd< 6.25\e{-4}$
-and Ericksen numbers $0.67\le{\it Er}<8$ 
-lies another region where the network flows with periodically 
-recurring conformations (Fig. \ref{bp1-rheo}). 
-We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
-is clear when looking at Fig. \ref{bp1-1_bp1-2} 
-which shows the deviatoric stress versus total strain. 
-A qualitative difference between these two is the absence in BPI-2 of the 
-large stress fluctuations that occur during the early cycles in regime BPI-1.
-[Recall that for even lower $\gd$ the deviatoric thermodynamic stress becomes 
-temporarily negative (Fig. \ref{bp1-fe-yield}), which 
-caused destabilisation of the periodic network and led to the 
-amorphous configuration in the steady state.] Hence, an explanation
-for the existence of the regular BPI-2 oscillations 
-could be that these are fast enough to bypass or suppress 
-large stress fluctuations -- this leads to
-a stabilisation of the oscillations found only transiently in
-the BPI-1 regime.
-
-Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
-BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
-BPI-2 does not resemble the equilibrium configuration
-undergoing an affine transformation. It features rather intricate, flow-induced 
-conformations. Surprisingly, if the shear is switched off in
-the BPI-2 regime at some point during an oscillations, 
-most of the times the flow-induced configuration 
-reverts to a perfect quiescent BPI -- in stark contrast with
-what happens in BPI-1, for lower shear rates.
-
-The period after which network conformations recur is not
-easy to work out for BPI. Along the flow direction a relation of the form
-$\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
-7.81\e{-5},1.563\e{-4}$ for example. The analysis is complicated by the fact
-that for $n>1$ thermodynamically equivalent, mirror-image 
-conformations arise after shorter times. For instance at $\gd=7.81\e{-5}$
-the conformations at times $t$ and $t+1/\gd$ are mirror-images, whereas
-those at $t$ and $t+2/\gd$ coalesce in flow direction.
-
-However, there is no definite recurrence period $\tau_V$ 
-in the vorticity direction. The direction of the displacement
-perpendicular to the flow direction is not along a coordinate
-axis. This direction, and the magnitude of the secondary
-network movement depends sensitively on the imposed shear rate. 
-This suggests that BPI-2 constitutes an instance of 
-'rheochaos'~\cite{rheochaos}.
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
-\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
-\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
-The sequence shows a typical cycle of shear-induced transformations in the 
-steady state at $\gd=1.56\e{-4}$ and time steps 
-$t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
-During every cycle the network is displaced also along the gradient and vorticity direction, 
-but the magnitude and direction of the offset does not appear to obey a 
-simple rule.}
-\label{bp1-med}
-\end{figure*}
-
-Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ for left- and right-handed chirality of
-the blue phase. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in that there is no band
-arrangement. However, just as in the case of BPII, both patterns
-are mirror images and feature the highest velocities in the regions where 
-the largest conformational changes take place.
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
-The pictures show a snapshot of the periodically recurring patterns in the 
-secondary velocity components $(0,v_y,v_z)$, i.e. the component $v_x$ in flow direction 
-has been projected out. The colour code gives the magnitude and sign 
-of the z-component. The vertical and horizontal direction are the gradient and 
-vorticity direction, respectively, with the flow into/out of the plane in the 
-upper/lower half of the images.}
-\label{bp1-velo}
-\end{figure*}
-
-The intricacy and periodic recurrence of the flow-induced conformations 
-in BPI raises the question: how robust are these results to variations in 
-system size? Although a pitch length of 64 LBU provides enough resolution 
-for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
-in total are actually sufficient to avoid finite size effects.
-Unfortunately the computational costs of larger simulations are still
-prohibitive, given that large strains have to
-be simulated in order to reach a steady state. Therefore we opted for a 
-comparison with smaller runs.
-
-Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
-deviatoric stress $\Pi_{xy}$ of to different runs, one consisting of 8 
-unit cells and the other one with 64 unit cells.
-We observe that there are deviations in both the visual appearance of the 
-network as well as in the time-dependence of the deviatoric stress.
-Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
-to reach the steady state with its flow-induced recurring disclination network. 
-These results indicate that we cannot know with certainty if the individual 
-flow-induced conformations that we find for 64 unit cells are not affected 
-by the imposed periodicity -- quite possibly the details of the oscillations
-would actually differ in an even large simulation. 
-Nevertheless, the comparison suggests that 
-the recurrence period and the complexity of the network in flow are  
-generic features of BPI in this regime of flow.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
-\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
-The dark (red) network in top picture shows a snapshot of run that consists 
-of $2^3=8$ unit cells, whereas the light (cyan) network displays the 
-central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
-The bottom picture shows the deviatoric stress over time. [Interestingly, 
-an intermediate sized run with $3^3=27$ unit cells led to results that 
-were virtually identical to the smaller run with 8 unit cells.] 
-The shear rate was $\gd=7.81\e{-5}$.
-}
-\label{bp1-2uc4uc}
-\end{figure}
-
-\subsubsection{Regime BPI-3: high shear rates}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a region 
-of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
-direction oriented along the horizontal/vertical axis. 
-The colour code gives the magnitude of the z-component of the director field. 
-The data is translationally invariant along the vorticity direction.}
-\label{bp1-high}
-\end{figure}
-
-At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$, 
-we observe a break-up of the flow-induced 
-disclination network of regime BPI-2. However, 
-in contrast to BPII, where the configuration is a travelling helical 
-wave (viz. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
-two-dimensional, translationally invariant state, which is shown in 
-Fig. \ref{bp1-high}. The snapshots visualise the affine 
-transformation of simple shear flow in the gradient-flow plane.
-There are distinctive regions where the director field is 
-predominantly oriented along the vorticity direction and 
-other regions where it is oriented rather in flow-gradient plane. 
-Occasionally while moving past each other these rolls resemble 
-regions of local double twist.
-Interestingly the critical Ericksen number for the break-up lies in 
-about the same window as that of transition between BPII-1 and BPII-2.
-
-\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{cholflow}
-
-At shear rates beyond the regimes BPI-3 and BPII-2 and below the transition to a 
-flow-aligned nematic state we found another regime where 
-both blue phases adopt the same configuration in steady shear flow, 
-independent of their initial state.
-The director field of this configuration is shown in Fig. \ref{fig:cholflow}.
-It consists of a simple cholesteric helix with the helical axis oriented 
-along the velocity gradient direction. The state is translational invariant 
-along the flow and vorticity direction and therefore also one-dimensional. 
-However, in contrast to the travelling helical wave of BPII-2 the 
-director field is now static and there is no tumbling motion. 
-While flowing the director retains its relative orientation.
-Consequently, the shear stress is lower than for
-the 2D double-twist rolls of BPI-3 and the travelling helical wave
-(Fig. \ref{bp2-rheo} and \ref{bp1-rheo}) as there is less dissipation.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
-\caption{Flow induced cholesteric state: The orientation of the helix is 
-along the gradient direction. The director field is quasi-static during 
-the flow, which occurs in 'nematic layers'. For clarity the picture shows 
-the director field along the y-direction for selected sites on the x-z-plane.}
-\label{fig:cholflow}
-\end{figure}
-
-%\clearpage
-
-\section{Conclusions}
-
-In summary, our work constitutes the first simulation of bulk flow behaviour 
-of cubic BPs in simple shear flow. Although experimental evidence to support
-our results is currently not available, we believe this work can shed some light 
-into the flow properties of complex liquid-crystalline phases.
-
-We were able to characterise the rheology of cubic BPs and identified 
-two different flow regimes for BPII and three different ones in case of BPI. 
-Below Ericksen numbers $Er\simeq3$ BPII exhibits weak shear-thinning and 
-obeys a power-law.  The BPII disclination network breaks up and reconnects 
-in the flow, which leads to a periodically recurring dependence 
-of the deviatoric shear stress. The flow-induced conformation looks generally
-very similar to the quiescent network at equilibrium. 
-While being affinely transformed due to the shear flow
-the disclination network moves in the vorticity direction. The sense of
-motion is directly linked to the helicity of the underlying cholesteric phase.
-
-At larger Ericksen numbers $4\lesssim Er\lesssim 10$ the flowing BPII network 
-breaks up in a two-step manner. First a simple cholesteric helix forms with the helical axis 
-along the vorticity direction. The flow excites a travelling helical wave,
-as predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}.
-Upon increasing the shear, the helix rotates and this time orients along the
-gradient direction, where it can support larger flows due to the translational
-symmetry along the flow direction. Still larger flows break this residual
-cholesteric order and leave a standard flow-aligned nematic staete.
-
-Interestingly, BPI shows a flow behaviour that is very different to that of BPII.
-This is a direct consequence of the topological differences in both disclination 
-network. Below Ericksen numbers $Er<0.4$ the BPI network cannot both flow and 
-retain a regular appearance. It eventually collapses into an amorphous network and 
-features yield-stress behaviour. The reason is that shortly after the onset of the shearing 
-very large stress fluctuations occur. At the lowest shear rates
-the thermodynamic contribution to the deviatoric (shear) stress is temporarily 
-negative during one cycle. These fluctuations destabilise the network and 
-eventually trigger the transition into an amorphous state with a residual yield
-stress.
-
-At slightly larger Ericksen numbers $0.4<{\it Er}<8$,
-BPI kinematically bypasses the large stress fluctuations and 
-flows with very complex, periodically recurring flow-induced conformations.
-These conformations entail regularly arrange helical disclinations which 
-rotate due to the shear -- a new flow-induced state which was not previously
-predicted. 
-The offset of the network after each oscillation is along a direction which,
-while being a multiple of the unit cell size along all three coordinate
-axis, depends on system size and shear rate in a seemingly chaotic fashion.
-Despite their complex appearance, the flow-induced conformations are 
-topologically connected to the quiescent BPI and after switching off of the 
-shear flow a defect-free blue phase reforms. It is tempting to
-interpret all these finding as an instance of 
-deterministic rheological chaos.
-
-Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
-a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
-quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
-Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
-direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
-is a flow-aligned nematic state.
-
-% If in two-column mode, this environment will change to single-column
-% format so that long equations can be displayed. Use
-% sparingly.
-%\begin{widetext}
-% put long equation here
-%\end{widetext}
-
-% figures should be put into the text as floats.
-% Use the graphics or graphicx packages (distributed with LaTeX2e)
-% and the \includegraphics macro defined in those packages.
-% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz,
-% and Frank Mittelbach for instance.
-%
-% Here is an example of the general form of a figure:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Use the figure* environment if the figure should span across the
-% entire page. There is no need to do explicit centering.
-
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}
-% \end{figure}
-
-% Surround figure environment with turnpage environment for landscape
-% figure
-% \begin{turnpage}
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}
-% \end{figure}
-% \end{turnpage}
-
-% tables should appear as floats within the text
-%
-% Here is an example of the general form of a table:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
-% \begin{tabular}{} command.
-% The ruledtabular enviroment adds doubled rules to table and sets a
-% reasonable default table settings.
-% Use the table* environment to get a full-width table in two-column
-% Add \usepackage{longtable} and the longtable (or longtable*}
-% environment for nicely formatted long tables. Or use the the [H]
-% placement option to break a long table (with less control than 
-% in longtable).
-% \begin{table}%[H] add [H] placement to break table across pages
-% \caption{\label{}}
-% \begin{ruledtabular}
-% \begin{tabular}{}
-% Lines of table here ending with \\
-% \end{tabular}
-% \end{ruledtabular}
-% \end{table}
-
-% Surround table environment with turnpage environment for landscape
-% table
-% \begin{turnpage}
-% \begin{table}
-% \caption{\label{}}
-% \begin{ruledtabular}
-% \begin{tabular}{}
-% \end{tabular}
-% \end{ruledtabular}
-% \end{table}
-% \end{turnpage}
-
-% Specify following sections are appendices. Use \appendix* if there
-% only one appendix.
-%\appendix
-
-%\section{}
-
-% If you have acknowledgments, this puts in the proper section head.
-\begin{acknowledgments}
-We acknowledge support by EPSRC grant nos. EP/E045316 and EP/E030173, 
-the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
-We thank Peter J. Collings for very stimulating discussions. 
-MEC holds a Royal Society Research Professorship.
-\end{acknowledgments}
+We close with a technical note on the choice of boundary conditions. 
+In all our runs we used Lees-Edwards boundary conditions: this
+is the sheared equivalence of periodic boundary conditions, for both the 
+velocity and the order parameter field. 
+While extremely useful to simulate bulk flow, we
+note that our boundary conditions impose macroscopic distortion
+of the network, and are effective analogous to infinitely distant
+but anchored wall along the $xy$ plane. In other words, due to
+the periodic boundary conditions along the flow gradient direction, 
+there can be no slip in the network deformation (or strain) at any point. 
+Effectively, this disallows permeation flows, which on the other
+hand require a non-zero velocity field and, symultaneously,
+negligible (ideally zero) perturbations of the blue phase network.
+Consequently our results are effectively valid for sheared bulk 
+BPs which are anchored at the walls. However, permeation flows are 
+only restricted to very slow flows, and are unstable for intermediate 
+and high flows~\cite{Dupuis:2005}, 
+where the response of the network depends less on
+the details of the boundary conditions used at the wall.
+\newpage
 

### 2013-01-16 trunk/papers/bprheo/bprheo.tex
Added all comments, ready for first submission
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1855 40876441-3496-0410-a613-a2fca35b2ae2

@@
-%% ****** Start of file apstemplate.tex ****** %
-%%
-%%
-%%   This file is part of the APS files in the REVTeX 4 distribution.
-%%   Version 4.1r of REVTeX, August 2010
-%%
-%%
-%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
-%%
-%%   See the REVTeX 4 README file for restrictions and more information.
-%%
-%
-% This is a template for producing manuscripts for use with REVTEX 4.0
-% Copy this file to another name and then work on that file.
-% That way, you always have this original template file to use.
-%
-% Group addresses by affiliation; use superscriptaddress for long
-% author lists, or if there are many overlapping affiliations.
-% For Phys. Rev. appearance, change preprint to twocolumn.
-% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
-%  Add 'draft' option to mark overfull boxes with black boxes
-%  Add 'showpacs' option to make PACS codes appear
-%  Add 'showkeys' option to make keywords appear
-\documentclass[aps,pre,reprint,superscriptaddress, twocolumn]{revtex4}
-%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
-%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}
-
-% You should use BibTeX and apsrev.bst for references
-% Choosing a journal automatically selects the correct APS
-% BibTeX style file (bst file), so only uncomment the line
-% below if necessary.
-\bibliographystyle{apsrev4-1}
-
-\usepackage{graphicx}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%This is the article LaTeX template for RSC journals
+%Copyright The Royal Society of Chemistry 2010
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\documentclass[8.5pt,twoside,twocolumn]{article}
+\oddsidemargin -1.2cm
+\evensidemargin -1.2cm
+\textwidth 18cm
+\headheight 1.0in
+\topmargin -3.5cm
+\textheight 22cm
+\usepackage[super,sort&compress,comma]{natbib} 
+\usepackage{mhchem}
+\usepackage{times,mathptmx}
+%\usepackage{times}
+% feel free not to use mathptmx if it causes difficulties
+\usepackage{sectsty}
+\usepackage{balance}
+
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{float}
-
 \newcommand{\e}[1]{\times10^{#1}}
@@
 
+\usepackage{graphicx} %eps figures can be used instead
+\usepackage{lastpage}
+\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+
 \begin{document}
 
-% Use the \preprint command to place your local institutional report
-% number in the upper righthand corner of the title page in preprint mode.
-% Multiple \preprint commands are allowed.
-% Use the 'preprintnumbers' class option to override journal defaults
-% to display numbers if necessary
-%\preprint{}
-
-%Title of paper
-\title{Rheology of Cubic Blue Phases}
-
-% repeat the \author .. \affiliation  etc. as needed
-% \email, \thanks, \homepage, \altaffiliation all apply to the current
-% author. Explanatory text should go in the []'s, actual e-mail
-% address or url should go in the {}'s for \email and \homepage.
-% Please use the appropriate macro foreach each type of information
-
-% \affiliation command applies to all authors since the last
-% \affiliation command. The \affiliation command should follow the
-% other information
-% \affiliation can be followed by \email, \homepage, \thanks as well.
-
-\author{O. Henrich}
-\email[Corresponding author: ]{ohenrich@epcc.ed.ac.uk}
-\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
-
-\author{K. Stratford}
-\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-
-\author{P.V. Coveney}
-\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
-
-\author{M.E. Cates}
-\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-
-\author{D. Marenduzzo}
-\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-
-%\homepage[]{Your web page}
-%\thanks{}
-%\altaffiliation{}
-
-%Collaboration name if desired (requires use of superscriptaddress
-%option in \documentclass). \noaffiliation is required (may also be
-%used with the \author command).
-%\collaboration can be followed by \email, \homepage, \thanks as well.
-%\collaboration{}
-
-%\noaffiliation
-
-\date{\today}
-
-\begin{abstract}
-We study the behaviour of cubic blue phases under shear flow
+\thispagestyle{plain}
+\fancypagestyle{plain}{
+\fancyhead[L]{\includegraphics[height=8pt]{headers/LH}}
+\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{headers/CH}}
+\fancyhead[R]{\includegraphics[height=10pt]{headers/RH}\vspace{-0.2cm}}
+\renewcommand{\headrulewidth}{1pt}}
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\renewcommand\footnoterule{\vspace*{1pt}% 
+\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
+\setcounter{secnumdepth}{5}
+
+\makeatletter 
+\def\subsubsection{\@startsection{subsubsection}{3}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\bf}} 
+\def\paragraph{\@startsection{paragraph}{4}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\textit}} 
+\renewcommand\@biblabel[1]{#1}            
+\renewcommand\@makefntext[1]% 
+{\noindent\makebox[0pt][r]{\@thefnmark\,}#1}
+\makeatother 
+\renewcommand{\figurename}{\small{Fig.}~}
+\sectionfont{\large}
+\subsectionfont{\normalsize}
+
+\fancyfoot{}
+\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{headers/LF}}
+\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{headers/RF}}
+\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{headers/RF}}
+\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
+\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
+\fancyhead{}
+\renewcommand{\headrulewidth}{1pt} 
+\renewcommand{\footrulewidth}{1pt}
+\setlength{\arrayrulewidth}{1pt}
+\setlength{\columnsep}{6.5mm}
+\setlength\bibsep{1pt}
+
+\twocolumn[
+  \begin{@twocolumnfalse}
+\noindent\LARGE{\textbf{Rheology of Cubic Blue Phases}}
+\vspace{0.6cm}
+
+\noindent\large{\textbf{Oliver Henrich $^{\ast}$\textit{$^{a,b}$}, Kevin Stratford \textit{$^{a}$}, Peter V.
+Coveney \textit{$^{b}$}, Michael E.
+Cates \textit{$^{c}$}, Davide Marenduzzo \textit{$^{c}$}}}\vspace{0.5cm}
+%Please note that \ast indicates the corresponding author(s) but no footnote text is required.
+
+\noindent\textit{\small{\textbf{Received Xth XXXXXXXXXX 20XX, Accepted Xth XXXXXXXXX 20XX\newline
+First published on the web Xth XXXXXXXXXX 200X}}}
+
+\noindent \textbf{\small{DOI: }}
+\vspace{0.6cm}
+%Please do not change this text.
+
+\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
 via lattice Boltzmann simulations. We focus on the two experimentally observed
@@
 shear rates -- very slow flow leads to a transition into an amorphous network 
-with a residual stress. The flow-induced oscillation regime features a novel arrangement
-of double helical disclinations which rotate due to the shear. Large flow can, 
+with an apparent yield stress. The flow-induced oscillation regime features a novel arrangement
+of double helical disclinations which rotate due to the shear. Large flow rates can, 
 just as in the BPII case, lead to the rupture of the network and the stabilisation
@@
 theoretical investigation of sheared blue phases in large systems, and
-are relevant to understanding the bulk rheology of these materials. 
-\end{abstract}
-
-% insert suggested PACS numbers in braces on next line
-\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
-% insert suggested keywords - APS authors don't need to do this
-\keywords{blue phase, cholesterics, chiral nematics, rheology}
-
-%\maketitle must follow title, authors, abstract, \pacs, and \keywords
-\maketitle
-
-% body of paper here - Use proper section commands
-% References should be done using the \cite, \ref, and \label commands
+are relevant to understanding the bulk rheology of these materials.}
+\vspace{0.5cm}
+ \end{@twocolumnfalse}
+  ]
+
+%\footnotetext{\dag~Electronic Supplementary Information (ESI) available: [details of any supplementary information available should be included here]. See DOI: 10.1039/b000000x/}
+
+%Please use \dag to cite the ESI in the main text of the article.
+%If you article does not have ESI please remove the the \dag symbol from the title and the above footnotetext.
+
+\footnotetext{\textit{$^{a}$~EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK; E-mail: ohenrich@epcc.ed.ac.uk}}
+\footnotetext{\textit{$^{b}$~Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}}
+\footnotetext{\textit{$^{c}$~SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
+
+%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required
+
+%\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P.
+Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}
+
 \section{Introduction}
@@
 and in the presence of colloidal particles~\cite{Ravnik:2011a}.
-The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. 
+The kinetics of BP domain growth have been recently addressed ~\cite{Henrich:2010b}. 
 However, our understanding of their dynamical behaviour under flow remains
@@
 
-Our work differs from these earlier efforts as it addresses shear flows, 
+Our work differs from these earlier efforts as it addresses homogeneous shear flows, 
 and focuses on flow-induced reconstruction and nonequilibrium transition
@@
 Our simulations are large scale and parallel, so that we can study
-significantly larger systems than previously possible, so that our
+significantly larger systems than previously possible, and our
 results can in principle be compared with bulk rheology experiments.
@@
 In Section II, we describe the method we use and review
-the hydrodynamic equation of motion which we aim to solve. In Section III, 
+the hydrodynamic equations of motion which we aim to solve. In Section III, 
 we report our numerical results, separating them into subsections referring to
@@
 The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
-Indeed, the ``redshift'' $r=p/p_0$ is adjusted during the simulation, to optimise the free energy density
-at all times -- this is done by following the procedure
+Indeed, the ``redshift'' $r=p/p_0$ is adjusted during the equilibration phase of the 
+simulation, to optimise the free energy density
+before shearing begins -- this is done by following the procedure
 previously described in~\cite{Alexander:2006}.
@@
 is a constant depending on the molecular details of the liquid crystal.
-Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the 
+Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution for $\theta$, the 
 so-called Leslie angle: we select this case by 
@@
 of the Lattice Boltzmann (LB) equations 
-that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta})$ that formally appears
+that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
 in this expansion is negligible under the slow flows considered here
 for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
-$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
 The thermodynamic stress tensor reads explicitly
@@
 simulated. 
-Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}={\mathbf 0}$, this means that our sytem, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
+Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our sytem, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
 We performed also selected simulations on metastable states at higher and lower temperatures, 
@@
 general flow behaviour from that described below.
-The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the disclination 
-network broke up into cholesteric/Newtonian states beyond a critical shear rate
-which was lower.
+The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
+rate at wich the disclination 
+network broke up into chiral nematic states was lower.
 This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.
@@
 all kinematic details of the blue phase networks in shear flow. This includes
-reconstruction of the unit cell not acessible in simulation with only
+reconstruction of the unit cell not accessible in simulation with only
 few cells~\cite{Dupuis:2005}.
@@
 $\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
-For clarity we classify various flow regimes, two in case of BPII and three in case of BPI. 
+For clarity we classify various flow regimes, two in the case of BPII and three in the 
+case of BPI. 
 They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
@@
 BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
-Beyond these regimes both BPI and BPII form common configurations regardless 
-of the initial state, namely a cholesteric helix with helical axis 
+Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
+both BPI and BPII form common configurations regardless 
+of the initial state. These are a cholesteric helix with helical axis 
 oriented along the direction of the velocity gradient (CHG), 
 and a flow-aligned nematic state (FAN).  
-Disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
+As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
 Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-(Note that $q_s$ is small but non-zero at the disclination core.)
+(Note that $q_s$ is small but non-zero at the disclination core. The director is undefinded there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
 
@@
 shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
-regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
-and break up of the network into a travelling helical wave (BPII-2, solid circle).
+regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles), 
+break up of the network into a travelling helical wave (BPII-2, solid circle), 
+cholesteric helix with axis in gradient direction (CHG, open squares)
+and flow-aligned nematic state (FAN, solid square).
 The error bars indicate the maximum and minimum stresses that occur during one cycle.}
@@
 $\eta_{app}=\langle \Pi_{xy} \rangle/\gd$, and $\eta$.
-A numerical value of $\eta_{app}=1$ corresponds to a fully Newtonian flow,
+A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
 without any additional contribution from the liquid crystal.
@@
 The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
-the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
+the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
 For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
@@
 
-\subsubsection{Regime BPII-1: low and intermediate shear rates}
+\subsubsection{Regime BPII-1: low and intermediate shear rates }
 
@@
 
+The general appearance of the flowing network is, apart from the homogeneous distortion,
+very close to that of the quiescent blue phase at equilibrium. This holds for all
+shear rates in the BPII-1 regime.
+
+Interestingly, while being displaced with the flow the entire network moves 
+along the vorticity direction (z) as well.  A similar behaviour has been recently 
+observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
+In contrast to the stick-slip motion that has been reported there,
+in this case the movement is steady. 
+The vorticity motion occurs in such a way that the positions of breakup and reconnection 
+point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
+to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
+is $\tau_V=6\tau_F$, 
+i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
+for the network to move one unit cell along $z$ (the vorticity direction).
+
+Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
+cycle, with superimposed velocity vectors.
+The plot shows the secondary velocity components, obtained by projecting the
+velocity onto a plane perpendicular to the flow direction.
+This gets rid of the dominating velocity component along the flow, $v_x$, and 
+allows to visualise the patterns in the 
+two much smaller components $v_y$ and $v_z$.
+The magnitude of the secondary components is typically in the range of a 
+few percent of the primary flow component, for the shear rates 
+and system sizes simulated here.
+
 \begin{figure}[htpb]
-\includegraphics[width=0.41\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.41\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.41\textwidth]{disc-165k_run902.png}
+\center
+\includegraphics[width=0.35\textwidth]{disc-160k_run902.png}
+\includegraphics[width=0.35\textwidth]{disc-164k_run902.png}
+\includegraphics[width=0.35\textwidth]{disc-165k_run902.png}
 \caption{Disclination network of BPII in shear flow: 
@@
 
-The general appearance of the flowing network is, apart from the homogeneous distortion,
-very close to that of the quiescent blue phase at equilibrium. This holds for all
-shear rates in the BPII-1 regime.
-
-Interestingly, while being displaced with the flow the entire network moves 
-along the vorticity direction as well.  A similar behaviour has been recently 
-observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
-In contrast to the stick-slip motion that has been reported there,
-in this case the movement is steady. 
-The vorticity motion occurs in such a way that the positions of breakup and reconnection 
-point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
-to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
-is $\tau_V=6\tau_F$, 
-i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
-for the network to move one unit cell along $z$ (the vorticity direction).
-
-Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
-cycle, with superimposed velocity vectors.
-The plot shows the secondary velocity components, obtained by projecting the
-velocity onto a plane perpendicular to the flow direction.
-This gets rid of the dominating velocity component along the flow, $v_x$, and 
-allows to visualise the patterns in the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the secondary components is typically in the range of a 
-few percent of the primary flow component, for the shear rates 
-and system sizes simulated here.
-
 Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
@@
 
-Tab. \ref{tab1} in the Appendix 
+Table \ref{tab1} in the Appendix 
 gives minima, maxima, averages and standard deviations 
@@
 
-\subsubsection{Regime BPII-2: high shear rates}
+\subsubsection{Regime BPII-2: high shear rates }
 
@@
 parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
-see Tab. \ref{tab1} in the Appendix).
+see Table \ref{tab1} in the Appendix).
 
@@
 $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+The colour code gives the alignment of the director field in flow direction (x) with red fully aligned and blue out of plane.
 The configuration is a helical wave that travels along the direction of vorticity.
@@
 Grandjean texture~\cite{deGennes}).
-We will address this in section \ref{cholflow}, as it is common to BPI and BPII.
+We will address this regime in section \ref{cholflow}, as it is common to BPI and BPII.
 
@@
 
-BPI has body-centered symmetry and, unlike BPII, the disclination lines
+BPI has body-centered cubic symmetry and, unlike BPII, the disclination lines
 characterising its equilibrium structure are well separated and do not 
@@
 a critical value -- close to this breakup point the oscillations in the 
-deviatoric stress show an increasingly complex
+shear stress show an increasingly complex
 pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
@@
 \includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity (normalised by $\eta$) $\eta_{app}/\eta=\langle \Pi_{xy}\rangle/\dot{\gamma}$ of BPI versus time (top) 
+\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
 and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
@@
 can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
-steady flow with recurring patterns (BPI-2, open circles) and 
-breakup of the disclination network (BPI-3, solid circle). 
+steady flow with recurring patterns (BPI-2, open circles), 
+breakup of the disclination network (BPI-3, solid circle), 
+cholesteric helix with axis in gradient direction (CHG, open squares) 
+and flow-aligned nematic state (FAN, solid square). 
 The error bars represent the minimum and maximum shear stress 
@@
 
-\subsubsection{Regime BPI-1: low shear rates}
+\subsubsection{Regime BPI-1: low shear rates }
 
@@
 \includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{Average deviatoric stress and free energy density of BPI at low shear rate: 
-The negative branch in the stress is related to a local maximum and a following local 
+\caption{Average thermodynamic shear stress and free energy density of BPI at shear rate 
+$\gd=4.88\e{-5}, Er=0.08$: The negative branch in the stress is related to a local maximum and a following local 
 minimum in average free energy density. This creates unstable conditions that lead 
-to a dissolving regular network and an amorphous state which 
+to reconstruction of the defect lattice into an amorphous network state which 
 seemingly features a yield stress.}
@@
 yield stress. An explanation for this behaviour can be found by 
-looking more closely at the average deviatoric stress (where the
+looking more closely at the average shear stress (where the
 background viscosity contribution has been subtracted) and 
@@
 BPI network is subject to large forces. Seemingly, these eventually
-cause the flow-induced collapse to the amorphous state with
+cause the flow-induced collapse to an amorphous network state with
 an apparent yield stress. We should underscore that this
-low shear regime, might be brought about by the boundary
-conditions, which in practice is equivalent to having infinitely
-distant walls which the network is anchored to (see discussion
+low shear regime, might depend on our choice of boundary
+conditions, which in practice are equivalent to having infinitely
+distant walls to which the network is anchored (see discussion
 in the Appendix).
-With free boundary conditions,
-one may imagine that, for very low wall velocity, 
-permeative flows may be possible which allow for
+With free boundary conditions and for very low wall velocity, 
+permeative flows might instead allow for
 some slip of the BPI network, as in cholesterics sheared
 along their axis~\cite{Marenduzzo:2006b}.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
+\caption{Thermodynamic shear stress $\Pi_{xy}$ versus strain near the transition between regime
+BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
+and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
+where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
+throughout positive, but exhibits very large fluctuations during one cycle, 
+which are absent in regime BPI-2 at higher shear rates than shown here.}
+\label{bp1-1_bp1-2}
+\end{figure}
 
@@
 
-\if{The negative branch in the deviatoric stress is connected
+\if{The negative branch in the shear stress is connected
 to a local maximum followed by a local minimum in the 
@@
 
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{Shear stress $\Pi_{xy}$ versus strain near the transition between regime
-BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
-and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
-where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
-throughout positive, but exhibits very large fluctuations during one cycle, 
-which are absent in regime BPI-2 at higher shear rates than shown here.}
-\label{bp1-1_bp1-2}
-\end{figure}
-
 To gain further insight into the flow-induced deformation, and eventual
@@
 Consequently, the network adopts a flow-induced conformation which consists 
-of intertwined helices that stretch during the affine transformation.
+of intertwined helices that stretch during the shear transformation.
 This helical conformation that emerges already at strains $\gamma\simeq1$ 
@@
 view of the large stress fluctuations discussed above. 
-Finally, the network loses order and 
-transforms into an amorphous state with almost constant
-yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ LBU.
+Finally, the system loses order and 
+transforms into an amorphous network state with almost constant
+stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ LBU.
 If the shear is stopped, this flow-induced amorphous state
@@
 
-\subsubsection{Regime BPI-2: intermediate shear rates}
-
-Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$; $ {\it Er}\lesssim 0.33$) and 
+\subsubsection{Regime BPI-2: intermediate shear rates }
+
+Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$; $ {\it Er}\lesssim 0.33$) but 
 at slightly larger shear rates ($3.91\e{-5}\lesssim\gd\lesssim 4.687\e{-4}$; $0.67\lesssim{\it Er}\lesssim8$)
@@
 is clear when looking at Fig. \ref{bp1-1_bp1-2} 
-which shows the deviatoric stress versus total strain. 
+which shows the shear stress versus total strain. 
 A qualitative difference between these two is the absence in BPI-2 of the 
 large stress fluctuations that occur during the early cycles in regime BPI-1.
-(Recall that for even lower $\gd$ the deviatoric thermodynamic stress becomes 
+(Recall that for even lower $\gd$ the thermodynamic shear stress becomes 
 temporarily negative (Fig. \ref{bp1-fe-yield}), which 
@@
 \label{bp1-med}
+\end{figure*}
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
+\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
+The pictures show a snapshot of the periodically recurring patterns in the 
+secondary velocity components $(v_y,v_z)$. 
+The colour code gives the magnitude and sign 
+of the $z$-component. The vertical and horizontal direction are the gradient and 
+vorticity direction, respectively, with the flow into/out of the plane in the 
+upper/lower half of the images.}
+\label{bp1-velo}
 \end{figure*}
@@
 Taken together, all this evidence very much suggests that 
-BPI-2 constitutes an instance of ``rheochaos''~\cite{rheochaos}.
+the BPI-2 range includes at least some instances of ``rheochaos''~\cite{rheochaos,Cates:2002}.
 
@@
 the blue phase. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in that there is no band
-arrangement. However, just as in the case of BPII, both patterns
+Fig. \ref{bp2-velo}, but differs in that there is no simple
+separation into bands with opposite secondary velocity. 
+However, just as in the case of BPII, both patterns
 are mirror images and feature the highest velocities in the regions where 
@@
 
-\begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
-The pictures show a snapshot of the periodically recurring patterns in the 
-secondary velocity components $(0,v_y,v_z)$, i.e. the component $v_x$ in flow direction 
-has been projected out. The colour code gives the magnitude and sign 
-of the $z$-component. The vertical and horizontal direction are the gradient and 
-vorticity direction, respectively, with the flow into/out of the plane in the 
-upper/lower half of the images.}
-\label{bp1-velo}
-\end{figure*}
-
 \if{In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
-we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
+we performed a Fourier analysis of the shear stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
 The spectrum is shown in Fig. \ref{bp1-spectrum}.
@@
 \includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
+\caption{Fourier analysis of the shear stress $\Pi_{xy}$ of BPI: The main picture shows 
 spectral components and the 'lock-in' of flow modes, 
@@
 
-Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
-deviatoric stress $\Pi_{xy}$ of two different runs, one consisting of 8 
-unit cells and the other one with 64 unit cells.
-We observe that there are deviations in both the visual appearance of the 
-network as well as in the time-dependence of the deviatoric stress.
-Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
-to reach the steady state with its flow-induced recurring disclination network. 
-These results indicate that we cannot know with certainty if the individual 
-flow-induced conformations that we find for 64 unit cells are not affected 
-by the imposed periodicity -- quite possibly the details of the oscillations
-would actually differ in an even large simulation. 
-Nevertheless, the comparison suggests that 
-the recurrence period and the complexity of the network in flow are  
-generic features of BPI in this regime of flow.
-
 \begin{figure}[htpb]
@@
 \includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
+\caption{Disclination network and shear stress $\Pi_{xy}$ in BPI: 
 The dark (red) network in top picture shows a snapshot of run that consists 
@@
 central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
-The bottom picture shows the deviatoric stress over time. (Interestingly, 
+The bottom picture shows the shear stress over time. (Interestingly, 
 an intermediate sized run with $3^3=27$ unit cells led to results that 
@@
 
-\subsubsection{Regime BPI-3: high shear rates}
+Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
+shear stress $\Pi_{xy}$ of two different runs, one consisting of 8 
+unit cells and the other one with 64 unit cells.
+We observe that there are deviations in the visual appearance of the 
+network as well as in the time-dependence of the shear stress.
+Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
+to reach the steady state with its flow-induced recurring disclination network. 
+These results indicate that we cannot know with certainty whether the individual 
+flow-induced conformations that we find for 64 unit cells are affected 
+by the imposed periodicity -- quite possibly the details of the oscillations
+would differ in an even larger simulation. 
+Nevertheless, the comparison suggests that 
+the recurrence period and the complexity of the network in flow are  
+generic features of BPI in this regime of flow.
+
+\subsubsection{Regime BPI-3: high shear rates }
 
@@
 
-In summary, our work constitutes the first simulation of bulk flow behaviour 
+In summary, our work constitutes the first large scale simulation of bulk flow behaviour 
 of cubic BPs in simple shear flow.
@@
 Below an Ericksen number of $Er\simeq3$, BPII exhibits weak shear-thinning and 
-obeys a power-law flow curve.  
+obeys a power-law flow curve with exponent close to unity.  
 The BPII disclination network breaks up and reconnects 
 in the flow, which leads to a periodically recurring dependence 
-of the deviatoric shear stress. The flow-induced conformation looks generally
+of the shear stress. The flow-induced conformation looks generally
 very similar to the quiescent network at equilibrium. 
@@
 Upon increasing the shear, the helix rotates and orients along the
-gradient direction, where it can support larger flows due to the translational
-symmetry along the flow direction. Still larger flows break this residual
+gradient direction, where it can support larger flow rates due to the translational
+symmetry along the flow direction. Still larger flow rates break this residual
 cholesteric order and leave a standard flow-aligned nematic state,
@@
 very large stress fluctuations occur. At the lowest shear rates
-the thermodynamic contribution to the deviatoric (shear) stress becomes temporarily 
+the thermodynamic contribution to the shear stress becomes temporarily 
 negative during each cycle. These fluctuations seem to destabilise the network and 
@@
 shear flow a defect-free blue phase reforms. It is tempting to
-interpret all these finding as an instance of 
-deterministic rheological chaos~\cite{fielding}.
+interpret the unsteady oscillations seen in this regime as an instance of 
+deterministic rheological chaos~\cite{fielding, Cates:2002}.
 
@@
 % If you have acknowledgments, this puts in the proper section head.
-\begin{acknowledgments}
+\section*{Acknowledgments}
 We acknowledge support by EPSRC grant nos. EP/E045316 and EP/E030173, 
 the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
-We thank Peter J. Collings for very stimulating discussions. 
+This work was granted access to the HPC resource of CSC, Finland, made available 
+within DECI by the PRACE-2IP, funded by the European Community's 7th Framework 
+Program under grand agreement No. RI-283493. We thank Peter J. Collings 
+for very stimulating discussions. 
 MEC holds a Royal Society Research Professorship.
-\end{acknowledgments}
 
 \appendix
-\section{}
+\section*{Appendix}
 
@@
 
-Tab. \ref{tab1} lists the parameters chosen for our runs (values of
+Table \ref{tab1} lists the parameters chosen for our runs (values of
 the reduced temperature and reduced chirality are given in the caption),
@@
 Effectively, this disallows permeation flows, which on the other
-hand require a non-zero velocity field and, symultaneously,
-negligible (ideally zero) perturbations of the blue phase network.
+hand require a non-zero velocity field and, simultaneously,
+much lower (ideally zero) perturbations of the blue phase network.
 Consequently our results are effectively valid for sheared bulk 
-BPs which are anchored at the walls. However, permeation flows are 
-only restricted to very slow flows, and are unstable for intermediate 
-and high flows~\cite{Dupuis:2005}, 
+BPs which are anchored at the walls. In practice, permeation flows are 
+only restricted to very low flow rates, and are unstable for intermediate 
+and fast flows~\cite{Dupuis:2005}, 
 where the response of the network depends less on
@@
 
-% Create the reference section using BibTeX:
-\bibliography{bprheo}
+\footnotesize{
+\bibliography{bprheo} %your .bib file
+\bibliographystyle{rsc} %the RSC's .bst file
+}
 
 \end{document}
-%
-% ****** End of file apstemplate.tex ******

### 2013-01-20 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
submitten version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1872 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\section{Introduction}
+\section{Introduction} \index{lattice Boltzmann} \index{complex fluids}
 The lattice Boltzmann (LB) method (for an overview see, e.g.,
@@
 
-\section{Background}\label{ch14:sec:background}
+\section{Background}\label{ch14:sec:background} \index{lattice Boltzmann} \index{complex fluids} \index{MPI}
 
@@
 \includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
-\caption{Left: the lattice is decomposed between MPI tasks. For
-  clarity we show a 2D decomposition of a 3D lattice, but in practice
-  we decompose in all 3 dimensions. Halo cells are added to each
-  sub-domain (as shown on the upper right for a single slice) which store
-  data retrieved from remote neighbours in the halo exchange. Lower
-  right: the D3Q19 velocity set resident on a lattice site;
-  highlighted are the 5 ``outgoing'' elements to be transferred in a
-  specific direction.}
+\caption[Parallel decomposition of D3Q19 lattice.]{Left: the lattice is
+  decomposed between MPI tasks. For clarity we show a 2D decomposition
+  of a 3D lattice, but in practice we decompose in all 3 dimensions.
+  Halo cells are added to each sub-domain (as shown on the upper right
+  for a single slice) which store data retrieved from remote
+  neighbours in the halo exchange. Lower right: the D3Q19 velocity set
+  resident on a lattice site; highlighted are the 5 ``outgoing''
+  elements to be transferred in a specific direction.}
 \label{ch14:fig:decomphalo}
@@
 
-\section{Single GPU Implementation}\label{ch14:sec:singlegpu}
+\section{Single GPU Implementation}\label{ch14:sec:singlegpu} \index{CUDA}
 
@@
 
-\section{Multiple GPU Implementation}\label{ch14:sec:parallelgpu}
+\section{Multiple GPU Implementation}\label{ch14:sec:parallelgpu} \index{multi-GPU programming} \index{MPI} \index{parallel scaling}
 
@@
 \includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
-\caption{The weak (top) and strong (bottom) scaling of \textit{Ludwig}.  Closed
-  shapes denote results using the CPU version run on the Cray
-  XE6 (using two 16-core AMD Interlagos CPUs per node), while open
-  shapes denote results using the GPU version on the Cray XK6 (using a
-  single NVIDIA X2090 GPU per node). Top: the benchmark time is shown
-  where the problem size per node is constant. Bottom: performance is
-  shown where, for each of the four problem sizes presented, the
-  results are scaled by the lattice volume, and all results are
-  normalized by the same arbitrary constant.  }
+\caption[Scaling of Ludwig to many GPUs in parallel.]{The weak (top)
+  and strong (bottom) scaling of \textit{Ludwig}.  Closed shapes
+  denote results using the CPU version run on the Cray XE6 (using two
+  16-core AMD Interlagos CPUs per node), while open shapes denote
+  results using the GPU version on the Cray XK6 (using a single NVIDIA
+  X2090 GPU per node). Top: the benchmark time is shown where the
+  problem size per node is constant. Bottom: performance is shown
+  where, for each of the four problem sizes presented, the results are
+  scaled by the lattice volume, and all results are normalized by the
+  same arbitrary constant.  }
 \label{ch14:fig:scaling}
@@
 
-\section{Moving Solid Particles}\label{ch14:sec:particles}
+\section{Moving Solid Particles}\label{ch14:sec:particles} \index{particle-fluid interaction} \index{molecular dynamics}
 
@@
 \includegraphics[width=10cm]{Chapters/chapter14/figures/colloid_new}
-\caption{
+\caption[Interaction of spherical particles with lattice Boltzmann fluid.]{
 A two-dimensional schematic picture of spherical particles on the lattice.

### 2013-01-21 trunk/papers/bprheo/bprheo.tex
Submitted version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1874 40876441-3496-0410-a613-a2fca35b2ae2

@@
 of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under 
-external fields ~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
+external fields~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
 and in the presence of colloidal particles~\cite{Ravnik:2011a}.
-The kinetics of BP domain growth have been recently addressed ~\cite{Henrich:2010b}. 
+The kinetics of BP domain growth have been recently addressed~\cite{Henrich:2010b}. 
 However, our understanding of their dynamical behaviour under flow remains
@@
 This emerges from the Chapman-Enskog expansion
-of the Lattice Boltzmann (LB) equations 
+of the lattice Boltzmann (LB) equations 
 that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
@@
 and~\ref{NSE} is solved by means of a
-hybrid scheme~\cite{Marenduzzo:2007}. 
-More details on the algorithm can
-be found in~\cite{Denniston:2001, Denniston:2004}.
+hybrid scheme~\cite{Marenduzzo:2007}, which combines different numerical schemes
+and provides a way to address the multiscale nature of the problem.
+More details on the algorithm and the boundary conditions can
+be found in~\cite{Denniston:2001, Denniston:2004, Henrich:2012a}.
 
@@
 phases BPI and BPII.
+Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
 The timestep and lattice spacing in lattice Boltzmann units (LBU) can be mapped
@@
 simulated. 
-Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our sytem, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
+Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
 We performed also selected simulations on metastable states at higher and lower temperatures, 
@@
 The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
-rate at wich the disclination 
+rate at which the disclination 
 network broke up into chiral nematic states was lower.
@@
 Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-(Note that $q_s$ is small but non-zero at the disclination core. The director is undefinded there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
+(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
 
@@
 at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
-gradient is oriented along the vertical direction, whereas the 
-horizontal direction is the flow direction. Lees-Edwards boundary 
-conditions have been imposed in such a way that network moves to the 
+gradient is oriented along the vertical direction (y), whereas the 
+horizontal direction (x) is the flow direction. Lees-Edwards boundary 
+conditions have been imposed in such a way that the network moves to the 
 right in the upper half and to the left in the lower half of the picture.}
@@
 $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
-The colour code gives the alignment of the director field in flow direction (x) with red fully aligned and blue out of plane.
+The colour code gives the alignment of the director field in the flow direction (x) with red fully aligned and blue out of plane.
 The configuration is a helical wave that travels along the direction of vorticity.
@@
 
-BPI has body-centered cubic symmetry and, unlike BPII, the disclination lines
+BPI has body-centred cubic symmetry and, unlike BPII, the disclination lines
 characterising its equilibrium structure are well separated and do not 
@@
 an apparent yield stress. We should underscore that this
-low shear regime, might depend on our choice of boundary
-conditions, which in practice are equivalent to having infinitely
+low shear regime might depend on our choice of boundary
+conditions which in practice are equivalent to having infinitely
 distant walls to which the network is anchored (see discussion
@@
 Unfortunately the computational costs of larger simulations are still
-prohibitive, given that large strains have to
+quite prohibitive (in the region of ${\it O}(10^5)$ core hours per run at
+doubled system size), given that large strains have to
 be simulated in order to reach a steady state. Therefore we opted for a 
@@
 other regions where it is oriented rather in flow-gradient plane. 
-Occasionally while moving past each other these rolls resemble 
+Occasionally, while moving past each other, these rolls resemble 
 regions of local double twist.
@@
 \caption{Director field of BPI at high shear rate: The pictures show a region 
-of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
+of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
 direction oriented along the horizontal/vertical axis. 
@@
 along the gradient direction. The director field is quasi-static during 
-the flow, which occurs in ``nematic layers''. Color coding indicates the 
+the flow, which occurs in ``nematic layers''. Colour coding indicates the 
 $x$ component of the director field (see colorbar on the right).
@@
 Despite their complex appearance, the flow-induced conformations are 
-topologically connected to the quiescent BPI and after switching off of the 
+topologically connected to the quiescent BPI and after switching off the 
 shear flow a defect-free blue phase reforms. It is tempting to
@@
 quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
-Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
+Then, at $16\lesssim Er \lesssim 22$, it forms a cholesteric helix along the flow gradient 
 direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
@@
 % \begin{tabular}{} command.
-% The ruledtabular enviroment adds doubled rules to table and sets a
+% The ruledtabular environment adds doubled rules to table and sets a
 % reasonable default table settings.

### 2013-01-31 papers/2012-sandwich/CholColloids.tex
Deleted depreciated version and added Mike's rewrite
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1901 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[12pt]{article}

\usepackage{graphicx,color}

\begin{document}

\title{Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
\author{Ourselves \\ {\small Here and there}}
\date{}

\maketitle

{\bf Colloidal particles dispersed in liquid crystals can form new materials with tunable elastic and electro-optic properties~\cite{stark}. In a host comprising  a periodic `blue phase' chiral liquid crystal~\cite{mermin}, the particles should template into colloidal crystals \cite{miha} with potential uses in photonics, metamaterials, and transformational optics~\cite{lavrentovich}. 
Here we show by computer simulation that colloid/cholesteric mixtures can give rise to a much wider range of soft-solid composites than previously anticipated. These include regular crystals, glasses, cellular solids, isolated clusters, twisted rings and undulating colloidal ropes. The final structure can be tuned through the particle concentration, and by varying the surface interactions of the cholesteric host with both the colloids and any confining walls. 
Remarkably, we find that many of these new materials are metastable: two or more structures can arise under identical thermodynamic conditions. The observed structure depends on not only on the formulation protocol, but also on the field history [CONFIRM!].  
This new class of soft materials should thus be relevant to the design of switchable, multistable devices for optical technologies including, for example, e-paper~\cite{epaper}.}

When spherical colloidal particles are mixed into a nematic liquid crystal, 
they disrupt the orientational order in the fluid and create defects (disclination lines) in the nematic close to their surfaces. To minimize their free energy cost, it is generally advantageous to for defects to be shared:
therefore the colloidal particles have a generic tendency to aggregate. Such aggregates may further 
self-assemble into lines~\cite{wiresmiha}, 2D crystals~\cite{zumer}, 
planar structures~\cite{tanaka}, or 3D amorphous glasses~\cite{tiffany}.
Besides being of fundamental interest to materials science, these
structures have tunable elastic and optical properties~\cite{stark}. Hence 
they offer exciting prospects for applications as biosensors~\cite{abbott}, or
as new-generation display devices~\cite{colloiddevice}.

But one may also choose to disperse colloidal particles in a cholesteric, 
rather than nematic, liquid crystal. The molecules making up this material are
chiral, and this causes their average orientation (described by a director field) to rotate in space in
a helical fashion, rather than remain uniform as in nematics. 
This typically creates a 1-dimensional periodic structure, called the cholesteric phase, whose wavelength 
is the pitch, $p$. The periodicity can however become fully three dimensional in the so-called blue phases (BPs)~\cite{mermin}. BPs arise because the director field can twist around more than one direction at once (see 
the cartoon in Fig.~1). Such double twist 
regions (generally cylinders: DTCs) are energetically favoured at high enough molecular chirality, but
lead to geometric frustration because it is impossible to tile the whole 3D
space with DTCs without also introducing disclination lines: singular topological defects on which the director field is undefined. In blue phases these disclination lines themselves either form a 3D periodic regular lattice (in so called BPI and BPII), or remain disordered (BPIII)~\cite{bp3}. 
%
BPs are stabilised by
the tendency of chiral molecules to twist, but destabilised by the
cost of forming topological defects (which locally diminish the molecular alignment). In conventional cholesteric materials~\cite{mermin} this balance is only achievable in a narrow temperature range of a few degrees, around the onset of liquid crystalline ordering. However, recent advances in formulation have widened this stability range enormously (up to 250K - CONFIRM) paving the way to
applications of BPs in display devices~\cite{kikuchi,coleswidetrange,bpdevice}.

Besides being remarkable materials in themselves, BPs offer significant promise as hosts for dispersing colloidal particles. Because BPs contain a disclination network even in the absence of
particles, they can potentially template colloidal self-assembly. 
This was proved recently by simulations of BP-dispersed nanoparticles which interact only weakly with the molecules of a BP (to give a so-called `weak anchoring' regime)~\cite{miha}. Such particles are attracted to the disclination network in BPs: by covering up the defect cores, the high elastic energy cost of those regions is avoided. The pre-existing order of the defect network then templates particles into a regular colloidal crystal of the same periodicity. 
Because this structure has a wavelength in the visible range, 
the resulting material should inherit (and likely enhance) the incomplete photonic bandgap of the parent blue phase, as exploited for instance in laser devices~\cite{bplasers}; for related possibilities see \cite{lavrentovich}.

A second compelling reason to study colloid-BP composites is that, even without particles, BPs can show several competing metastable free energy minima,
corresponding to different topologies of the defect network~\cite{adriano}, with a strong dependence on applied electric or magnetic fields~\cite{henrichfield}.
% 
Adding particles is likely to further enrich the free energy landscape, creating composites that might be promising candidates for bistable or multistable devices, in which energy is needed only to switch optical properties and not to maintain them. (This is the e-paper paradigm~\cite{epaper}, and can lead to huge energy savings.)
To explore such metastability, and switching strategies between states, we require multi-unit-cell, time-dependent simulations. These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{domaingrowth,bp3,henrichfield}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}. Our methodology and simulation parameters are summarized in Supporting Information.

We show in Fig.~2 the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and and then dispersed colloidal particles within it. Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r=XXX$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface). For the latter we choose $w = 0.23, 2.3$ which are at the lower and upper end of the typical experimental range [CONFIRM, GIVE REF.]. This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.

For a single colloidal particle
in a nematic fluid, $w$ controls the formation of a local topological defect at the particle surface~\cite{stark}, with weak distortion of the nearby fluid at small $w$. In the colloid-BP system, $w$ also plays a determining role in the final composite structure. When $w$ is small (Fig.~ 2A,C), the attraction of particles to the disclination lattice causes this to become covered by particles, without disruption of its long-range order. Thus a templated colloidal crystal is created, as first reported in~\cite{miha}. Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which sets a typical spacing between them along the disclination lines. It seems likely that polymer-stabilised  blue phases~\cite{kikuchi} 
work by a similar principle to this small $w$ case, with a weak segregation of monomers towards the defect regions [IS THIS WHAT YOU MEANT?].

For large $w$ (Fig.~2B,D), the picture changes dramatically.
Each colloidal particle now disrupts significantly the order of the nearby fluid, with defect formation close to its surface~\cite{stark}. The resulting strong disruption of
local liquid crystalline order around each particle leads now to a 
strong interparticle attraction. This restructures the disclination
network completely, destroying the long range order of
the original BPI topology. Most notably, particle aggregation favours
the formation of defect junctions, where four disclinations meet. This structural motif is present in the unit cell of BPII rather than BPI (and also seen in the amorphous BPIII~\cite{bp3}). The 
particle clusters are disjoint at low volume fractions ($\phi\simeq 1\%$) but 
they interact elastically via the connecting disclinations. At larger $\phi$, the aggregates join up to form a percolating colloidal cluster which essentially templates the disclinations rather than vice versa.

All the structures in Fig.~2 should be soft solids, of nonzero elastic modulus $G$ at low frequency. This contrasts with pure BPs, where
$G=0$ as the fluid can flow with a finite
viscosity via permeation of molecules through a fixed defect lattice~\cite{permeation1,permeation2}. We expect the
structures in Fig.~2A-C to be only weak gels, as here the 
defect network percolates but particle contacts do not. This is broadly similar to the cholesteric-nanoparticle gel reported
in~\cite{lubensky} (with $G<1$ Pa). In contrast, the ``cellular solid'' 
structure in Fig.~2D should be much more resistant mechanically, 
due to the formation of a thick percolating particle network. The 
structure is now akin to that of the self-quenched glass found in 
colloid-nematic composites~\cite{tiffany} at much higher volume 
fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa. [WHAT IS CELLULAR ABOUT THIS STRUCTURE?]

A more practical protocol for including particles in a BPI phase is to disperse the particles in an isotropic phase, and then quenching this into the temperature range where BPI is stable. Here we know that, without particles, the kinetics can favour disordered (BPIII-like) intermediates which may be long lived or even metastable~\cite{domaingrowth}. Fig.~3 shows outcomes of such a quench protocol with the same parameter sets as in Fig.~2. For small $w$, the particles again decorate the disclination lines, but are templated onto an amorphous network rather than an ordered one.
Increasing the volume fraction (Fig.~3B) leads to a more even coverage with
a better defined particle spacing. 
For large $w$, on the other hand, strong interparticle forces again dictate
the self-assembly. At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~3C) whereas at higher $\phi$ a percolating network of this strands (the ``cellular solid, Fig.~3D) is again seen. Perhaps unsurprisingly, the free energy of these amorphous colloid-BP composites is considerably higher than for the decorated BPI structure (see Supporting Information, Fig.~S1). Nevertheless, all these structures
are metastable, as the elastic forces keeping them together dwarf thermal
energy from Brownian motion (see the scales of energies and energy
differences in Fig.~S1). [TECHNICALLY IN THE METASTABLE STATE WITHOUT BROWNIAN TERMS THERE IS NO NET FORCE ON ANY PARTICLE. IS THIS ACTUALLY TRUE? IF NOT, I'M CONCERNED A REFEREE COULD JUST SAY WE NEED TO RUN FOR LONGER]

In display devices and other optical applications, it is typical for liquid crystals to be strongly confined (and subject to strong boundary effects), in contrast to the bulk geometries considered above. We therefore next describe
how colloidal suspensions self assemble in thin sandwiches (thickness $h = XXX \lambda$), whose confining walls can favour either tangential (planar) or normal (homeotropic) orientation of the director field. For simplicity we here impose one or the other as a strict boundary condition [CONFIRM].

Fig.~4 shows the case of normal wall anchoring on varying $\phi$ and $w$. We quench from the isotropic
to the ordered phase as was done in Fig.~3, and again find a metastable, amorphous disclination network (see Supporting Information). The resulting colloid-cholesteric composites 
are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~4A,C), 
whereas for large $w$ we observe isolated
clusters (Fig.~4B) and percolated cellular solids (Fig.~4D), for
$\phi = 1\%$ and $5\%$ respectively. There are, however, significant 
effects of the confining walls.
First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls and also in the midplane (see side-views in Fig.~4). Second,
the clusters formed at large $w$ are themselves anisotropic with an oblate habit. These are similarly to those reported experimentally for large colloids in a cholesteric (but non-BP) host~\cite{niek}.

Fig.~5 shows comparable simulations for planar wall anchoring. 
Without colloids, planar anchoring results in a regular cholesteric helix,
with axis perpendicular to the boundary walls. 
This twisted texture develops via a set of defect loops,
which shrink and disappear to leave a defect-free sample. However,
dispersing particles with small $w$ arrests the loop shrinking process,
stabilising twisted figure-of-eight rings (Fig.~5A) at small $\phi$, and plectonemic structures at larger $\phi$ (Fig.~5C).
For larger $w$, we observe thick colloidal ``ropes" about 5 colloids across
(Fig.~5B,D). These ropes undulate to follow the chiral structure of the 
underlying fluid. On increasing the density they develop branch points and finally again approach a ``cellular solid" morphology. Note that planar wall anchoring is incompatible with the normal anchoring on the colloids (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~5).

FIELD SWITCHING PARA HERE
\if{The existence of several locally stable competing solids with 
very different symmetry and, arguably, physical properties, 
is of interest to technology, as it makes these materials ideal
candidate to build multistable devices, which can retain memory of
more than one metastable state in the absence of an applied field.
Multistability is indeed a must-have feature for new generation switchable 
electrooptic devices for use in energy-saving electronics-free
appliances, smart glass, and e-paper to cite just a few examples.
}\fi

The surprising wealth of colloid-BP composites obtained in Figs.~2-5 is the complex 
consequence of a simple competition between two self-assembly principles. 
First, blue phases provide a three-dimensional 
template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}. Second, colloidal intrusions in a 
liquid crystal raise the free energy locally; sharing this cost creates a 
generic tendency to aggregate into clusters~\cite{tiffany}. Which of these factors dominates depends on $w$, a dimensionless measure of the anchoring strength of the director field at the colloidal surface. 
For $w\ll 1$, the templating principle dominates. Colloidal crystals, gels or twisted rings can then arise. For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating cellular solids, or helical colloidal ropes. In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls.
MORE ON FIELD SWITCH HERE
This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid
crystal devices, which are at the forefront of current technological 
developments in this field. We have not explored in this work their switchability by flow, but this could again be inherited from the underlying BP dynamics \cite{flowswitch}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.

\begin{thebibliography}{99}
\bibitem{stark} H. Stark, {\it Phys. Rep.} {\bf 351}, 387-474 (2001).
\bibitem{mermin} Wright D.~C., Mermin, N.~D. Crystalline liquids: The blue phases. {\it Rev. Mod. Phys.} {\bf 61}, 385–432 (1989).
\bibitem{miha} M. Ravnik, G.~P. Alexander, J.~M. Yeomans,  
S. Zumer, {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188-5192 (2011).
\bibitem{lavrentovich} O.~D. Lavrentovich, {\it Proc. Natl. Acad. Sci.
USA} {\bf 108}, 5143-5144 (2011).
\bibitem{epaper} J. Heikenfeld, P. Drzaic,  J.~S. Yeo, T. Koch.
A critical review of the present and future prospects for electronic paper.
{\it J. Soc. Inf. Display} {\bf 19}, 129-156 (2011).
\bibitem{zumer} I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik,
S. Zumer, {\it Science} {\bf 313}, 954-958 (2006).
\bibitem{tiffany} T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo,
W.~C.~K. Poon, {\it Science} {\bf 333}, 79-83 (2011).
\bibitem{abbott} I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo, N.~L. Abbott, {\it Science} {\bf 332}, 1297-1300 (2011).
\bibitem{colloiddevice} K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton,
R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006). REPLACE WITH A SUITABLE REVIEW IF POSSIBLE?
\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}
{\bf 35}, 4419 (1987).
\bibitem{kikuchi} H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama.
Polymer-stabilized liquid crystal blue phases.
{\it Nat. Mat.} {\bf 1}, 64-68 (2002).
\bibitem{coleswidetrange} H.~J. Coles, M.~N. Pivnenko.
Liquid crystal 'blue phases' with a wide temperature range.
{\it Nature} {\bf 436}, 997-1000 (2005).
\bibitem{bpdevice} Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
{\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
\bibitem{bplasers} W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri.
Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
phase II. {\it Nat. Mat.} {\bf 1}, 111-113 (2002).
\bibitem{permeation1} 	W. Helfrich. Capillary flow of cholesteric
and smectic liquid crystals. {\it Phys. Rev. Lett.} {\bf 23}, 372-374 (1969).
\bibitem{permeation2} A. Dupuis, D. Marenduzz, E. Orlandini, J.~M. Yeomans.
Rheology of cholesteric blue phases. {\it Phys. Rev. Lett.} {\bf 95},
097801 (2005).
\bibitem{lubensky} M. Zapotocky, L. Ramos, P. Poulin, T.~C. Lubensky, 
D.~A. Weitz, {\it Science} {\bf 283}, 209-212 (1999).
\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles, {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates,
{\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic, {\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
\bibitem{tanaka} T. Araki, H. Tanaka, {\it Phys. Rev. Lett.} {\bf 97}, 
127801 (2006).
\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka,
{\it Nat. Mat.} {\bf 10}, 303-309 (2011).
\bibitem{adriano} A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini,
F. Salvadore, {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
\bibitem{fukuda} J.-I. Fukuda, S. Zumer, {\it Nat. Comm.} {\bf 2}, 246 (2011).
\bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
Self-Organization of Particles with Planar Surface Anchoring in a Cholesteric
 Liquid Crystal. {\it Langmuir} {\bf 26}, 13502-13510 (2010).
\bibitem{henrichfield} Henrich Field PRE 
\bibitem{juho1} juho: microrheology of cholesterics
\bibitem{juho2} juho: dimers in cholesterics
\bibitem{flowswitch} Davide et al on flow reorganization of BPs
\bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang, Developing optofluidic technology through the fusion of microfluidics and optics {\it Nature} {\bf 442}, 381-386 (2006).

\end{thebibliography}

\newpage

\begin{figure}
\includegraphics[width=\textwidth]{s0.png}
\caption{Schematics of the molecular arrangement in a 
cholesteric helix (A) and within a double twist cylinder (B, the
axis of the cylinder is perpendicular to the plane of the page).
The disclinations line arrangement in the unit cell of BPI is shown in
(C), while a reference phase diagram (from Ref.~\cite{crooker}) is
shown in (D). [I SEE NO REASON FOR INCLUDING PHASE DIAGRAM HERE. PUT IN SI IF NEEDED -- BUT SURELY IT IS VITAL TO APPLICATIONS THAT THIS IS *NOT* THE PHASE DIAGRAM FOR COLES OR KIKUCHI!]}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s1.jpg}
\caption{Snapshots of the steady states obtained after dispersing
a suspension of colloidal nanoparticles within a pre-equilibrated  BPI phase.  Structures correspond
to (A) $w=0.23$ and $\phi=1\%$, (B) $w=2.3$ and $\phi=1\%$,
(C) $w=0.23$ and $\phi=5\%$, (D) $w=2.3$ and $\phi=5\%$.
The anchoring of the director field to the colloidal surface is normal. [ISN'T THIS AUTOMATIC FOR $w>0$?]
[For the full parameter list used to generate Figs.~2-5 see Supporting Information.] [FIX COLOURS AND VISIBILITY FOR ALL FIGURES.]}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s2.jpg}
\caption{Steady states obtained by dispersing colloids in an isotropic phase and then quenching into the range where BPI is stable, leading to formation of an amorphous disclination network. Parameters otherwise as in Fig.~2.}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s3.jpg}
\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, with normal anchoring of the director field
at the boundaries, and then quenched into a regime where BPI [CONFIRM] is stable in the bulk. Both top and side views are provided. 
Structures correspond
to (A) $w=XXX$ and $\phi=1\%$, (B) $w=XXX$ and $\phi=1\%$,
(C) $w=XXX$ and $\phi=2\%$, (D) $w=XXX$ and $\phi=2\%$.
}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{s4.jpg}
\caption{As Fig.~4, but with planar anchoring of the director 
field at the boundaries.}
\end{figure}

\end{document}

### 2013-01-31 trunk/htdocs/colloid.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1902 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\section{Colloid Examples}
+\section{Colloids}
 
-\subsection{Initialisation from file}
+\subsection{General}
+
+Colloidal particles are assumed to be spherical with a geometrical
+centre $\mathbf{r}_c$, which is also the centre of mass. The
+centre is allowed to move continuously across the lattice
+with velocity $\mathbf{U}$; the particle has an angular velocity
+$\mathbf{\Omega}$. The surface of the colloid is defined by an
+input radius, $a_0$, which determines which lattice nodes are
+inside or outside the colloid. (The hydrodynamic properties of
+the colloid are specified by a different radius $a_h$ --- more
+of this later.) The boundary links are then the set of vectors
+joining lattice nodes which intersect the spherical surface
+$\{\mathbf{c}_b\}$. A schematic picture is presented in
+Figure~\ref{fig_coll2}. Note that a lattice node exactly at
+the solid-fluid interface is defined to be outside the colloid.
+
+In the original approach of Ladd, fluid occupied nodes both inside
+and outside the particle. The effect of the ``internal fluid'' is
+known to be restricted to short time scales (compared to the
+characteristic time $a_0^2/\nu$), on which the fluid inside the
+particle relaxes to a solid body rotation \cite{heemels}. There
+are a number of problems related to the internal fluid, so we
+use fully solid particles.
+
+Nguyen and Ladd.
+
+For each boundary link, it is useful to define a boundary vector,
+$\mathbf{r}_b$, which connects the centre of the colloid to the
+appropriate boundary node (note that the length of this vector is
+not necessarily equal to the input radius of the colloid).
+
+\begin{figure}[tb]
+\begin{center}
+\includegraphics{xfig/colloidhalflinksnew.eps}
+\end{center}
+\caption{Boundary links for a colloidal particle in two dimensions
+with D2Q9 (half the particle is shown). Links join fluid sites (red)
+to solid sites (green) and
+intersect the circular shell radius $a_0$. Boundary nodes (circles)
+lie exactly half way between pairs of lattice nodes. The generalisation
+to three dimensions is straightforward.}
+\label{fig_coll2}
+\end{figure}
+
+A \textit{boundary node} halfway between fluid nodes which are joined
+by a boundary link. The position of the boundary node is always
+exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
+\Delta t$) regardless of the actual position of the intersection
+of the colloid surface and the boundary link.
+The colloid therefore has a discrete representation which becomes a
+better approximation to the sphere as the input radius becomes larger;
+this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
+
+\subsection{Bounce-back on links}
+
+The standard boundary condition required for the solid-fluid
+interface of a moving particle is described by Ladd \cite{l94b},
+and is generally refered to as bounce-back on links (BBL).
+A boundary link is defined as joining a node $\mathbf{r}$
+inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
+If the post-collision distributions are denoted by $f^\ast$, then
+the distributions must be reflected at the solid surface so that
+\begin{equation}
+\label{eq:bbl1}
+f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
+- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
+\end{equation}
+where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
+at the boundary
+\begin{equation}
+\label{eq:ub}
+\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
+\end{equation}
+is determined by the particle linear velocity $\mathbf{U}$ and angular
+velocity $\mathbf{\Omega}$. The change in momentum described by
+
+The force exerted on a
+single link is
+\begin{equation}
+\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
+t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
+\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
+\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
+\label{eq:fb}
+\end{equation}
+with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
+
+The total hydrodynamic force on the particle is then found by taking
+the sum of
+$\mathbf{F}_b$ over all the boundary links defining the particle.
+The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
+which again is summed over all links to give the total torque on the colloid.
+
+Note that $\rho$ in the above equations is the density of the fluid
+at the appropriate fluid node for the link. As the density fluctuations
+in the fluid are small compared with the mean density $\rho_0$, the
+density in the correction to the bounce-back can be replaced by the
+mean $\rho_0$.
+
+\subsection{Dynamics}
+Having computed the total force and torque on an individual colloid,
+is is possible to update the the linear velocities
+\begin{equation}
+m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
+\end{equation}
+where the mass of the colloid is related to the input radius by
+$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
+velocity
+\begin{equation}
+I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
+\mathbf{T}(t),
+\end{equation}
+where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
+However, this explicit update is generally found to have poor stability
+properties \cite{l94b, nl02}. The alternative is to use a velocity update
+which is implicit \cite{heemels,nl02}.
+
+The total force and torque on a colloid can be split into velocity-dependent
+and -independent parts by combining Equations~(\ref{eq:ub})
+and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
+In this way, the decomposition is
+\begin{eqnarray}
+\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
+-\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
+\label{eq:d3}
+\\
+\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
+-\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
+\label{eq:d4}
+\end{eqnarray}
+The velocity independent parts of the force and the torque
+(appropriate for a colloid at rest) are
+\begin{eqnarray}
+\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
+\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
+- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
+\\
+\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
+\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
+- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
+(\mathbf{r}_b\times \mathbf{c}_b)
+\end{eqnarray}
+The matrices $\boldsymbol{\zeta}$ are interpreted as drag
+coefficients and can be written as
+\begin{eqnarray}
+\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
+\\
+\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
+\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
+\\
+\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
+\\
+\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
+\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
+(\mathbf{r}_b \times \mathbf{c}_b).
+\end{eqnarray}
+As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
+
+\begin{equation}
+\label{eq:k14}
+\left( \begin{array}{rrr}
+\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
+\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
+\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
+
+\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
+\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
+\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
+
+\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
+\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
+\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
+
+\end{array} \right).
+\end{equation}
+If the particle has a symmetric distribution of boundary links (the
+special case where the centre of mass is on a symmetry point of the
+lattice) the $\boldsymbol{\zeta}^{FU}$ and
+$\boldsymbol{\zeta}^{T\Omega}$ matrices are
+diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
+and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
+
+The velocity updates can now be rewritten using an
+implicit update, which results in
+\begin{eqnarray}
+\label{eq:k15}
+m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
+\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
+- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
+- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
+\\
+\label{eq:k16}
+I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
+\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
+- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
+- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
+\end{eqnarray}
+In general all the matrix elements are non-zero, in which case
+equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
+6$\times$6 matrix inversion.
+
+Finally, the position of the particle can be updated: an Euler
+forward step is taken using the updated velocity
+\begin{equation}
+\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
+\end{equation}
+
+\subsection{Changes in particle shape}
+
+One consequence is that the discrete shape of the particle fluctuates
+as the colloid centre moves relative to the
+lattice. This means that the size of the particle as seen by the
+fluid changes despite the fact that the input radius of the colloid
+is fixed.  However, these fluctuations are small for input radii
+greater than about 5 lattice units \cite{l96a}. Furthermore, the
+fluctuations may be greater for some input radii than others
+\cite{nl02}.
+
+In the standard approach, changes in the map of boundary links
+are accomodated by the internal fluid. If the particle shape
+changes so that an internal node is exposed, the internal fluid
+at that site rejoins the fluid proper with the solid body momentum
+(assuming the internal fluid is relaxed to solid-body rotation).
+If a fluid node is covered by particle movement, then it simply
+joins the internal fluid, and will relax to solid-body rotation.
+Without internal fluid, changes in particle shape
+in which fluid nodes are either covered or uncovered must be
+accompanied by the removal or addition
+of fluid with appropriate properties at the nodes in question.
+It is essential that this
+is done in a way which minimises the perturbation to the fluid
+flow.
+
+If a fluid node at $\mathbf{r}$ is covered by the movement of a
+solid particle,
+the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
+$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
+explicitly so that
+the overall mass density is unchanged. At the same time, the
+colloid must assume the momentum lost by the fluid
+$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
+
+Similarly, when a fluid node is exposed by particle movement,
+fluid must be replaced with the appropriate distribution, density
+and velocity. If the
+new density is $\rho$ (to be determined by some method) then
+the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
+\Delta x^3$ which must be balanced elsewhere. If the new fluid
+is assumed to have velocity $\mathbf{u}$ then the particle must
+give up momentum $\Delta x^3 \rho\mathbf{u}$.
+
+Following Nguyen and Ladd \cite{nl02}, these changes owing
+to change in particle shape are implemented by added a small
+correction to the bounce-back at the following time step.
+The excess mass from a covered fluid site is redistributed over all the
+boundary nodes by redefined the mometum transfer at bounce-back
+\begin{equation}
+f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
+- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
++ \frac{w_{c_b} \rho_0\Delta M_c }{A},
+\label{eq:q1}
+\end{equation}
+where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
+force on the particle owing to the change in shape is then
+that lost by the fluid plus the contributions from the final term
+in (\ref{eq:q1}) summed over all the boundary links
+\begin{equation}
+\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
++ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
+\mathbf{c}_b.
+\label{eq:q2}
+\end{equation}
+
+Note that for a closed surface,
+$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
+side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
+does not contribute to the net force and torque on the particle.
+The corresponding change in torque is
+\begin{equation}
+\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
+\times \sum_i f_i \mathbf{c}_i
++ \frac{\Delta x^3}{\Delta t}
+\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
+\label{eq:q2a}
+\end{equation}
+where $\mathbf{r}_c$ is the boundary vector at the position where the
+fluid has been removed. Again, for a closed surface, the
+redistribution of mass does not contribute to the net torque
+on the particle.
+
+Likewise, the contribution to the bounce-back from newly uncovered nodes
+is
+\begin{equation}
+- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
+\label{eq:q3}
+\end{equation}
+leading to a change in force of
+\begin{equation}
+\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
+\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
+\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
+\label{eq:q4}
+\end{equation}
+with a corresponding change in the torque.
+If more than one lattice node is either covered or uncovered by
+the movement of the particle, then these contributions add in a simple
+way. The overall contributions can be added to the
+velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
+Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
+
+\subsection{Particles near contact}
+
+Particles near contact may not pocess a full set of boundary
+links (see Figure~\ref{fig:f5}). This leads to a potential
+non-conservation of mass associated with the particle motion
+which must be corrected.
+
+\begin{figure}[tb]
+\begin{center}
+\includegraphics{xfig/colloidclose.eps}
+\end{center}
+\caption{Two colloids close to contact are missing links in the
+region of closest approach owing to a lack of fluid nodes in the
+interstice.}
+\label{fig:f5}
+\end{figure}
+Again following Nguyen and Ladd \cite{nl02}, mass conservation is
+enforced explicitly using the following procedure. There is a mass
+transfer associated with the bounce-back at each link of
+$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
+The net mass transport into the particle is the sum of this
+quantity over all the links, and can be written
+\begin{equation}
+\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
+\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
+  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
+\end{equation}
+For any partilce with a full set of boundary links, both
+$\sum_b w_{c_b} \mathbf{c}_b$
+and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
+However, if the set of boundary links is imcomplete
+(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
+a correction is required. This is again made by redistributing the
+excess or deficit of mass over the other existing boundary nodes.
+The additional contribution to the bounce-back here is
+$-w_{c_b} \rho_0 \Delta M_s/A$, where
+$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
+
+This contribution
+now adds to the velocity-dependent part of the force and torque and
+leads to a slightly different form of the drag matrices
+\begin{eqnarray}
+\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
+\\
+\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
+\overline{\mathbf{r}_b\times\mathbf{c}_b}),
+\\
+\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
+(\mathbf{c}_b - \overline{\mathbf{c}_b}),
+\\
+\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
+(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
+\end{eqnarray}
+The mean quantities
+\begin{equation}
+\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
+\end{equation}
+and
+\begin{equation}
+\overline{\mathbf{r}_b\times\mathbf{c}_b} =
+\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
+\end{equation}
+A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
+$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
+can be replaced by their deviation from the mean. This provides a
+convenient way to compute the drag matrices (maintaining symmetry)
+and computing the correct force and torque on the particle when close
+to contact.
+
+\subsection{Rotational motions}
+
+For a number of applications it is useful to update not only
+the psoition of a spherical particle, but its orientation. This
+is important, e.g., when considering magnetic particles. The
+best way to implementment the required rotations is by
+considering quaternions.
+
+A quaternion can be thought of as an extended complex number
+\begin{equation}
+q = w + xi + yj + zk
+\end{equation}
+where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
+of a quaternion is
+\begin{equation}
+N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
+\end{equation}
+Addition and subtraction of quaternions proceeds as normal, but
+care is required for multiplication, which is associative but not
+commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
+$\bf{v}$ is a normal 3-vector, then multiplication can be written
+\begin{equation}
+q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
++ {\bf v}_1 \times {\bf v}_2),
+\end{equation}
+where the usual scalar and vector products are used. Note that the
+conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
+$qq^\star = N^2(q)$.
+It can be shown that a rotation of an arbitrary vector though an
+angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
+be represented by the quarternion
+\begin{equation}
+q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
+\end{equation}
+which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
+then the rotated vector ${\bf v}^{'}$ can be written as
+\begin{equation}
+(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
+\end{equation}
+which can be expanded as
+\begin{equation}
+{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
+{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
+\end{equation}
+
+\subsection{Colloid-Colloid interactions}
+
+\subsubsection{Lubrication corrections}
+
+\subsubsection{Hard sphere interaction}
+
+Particles do not overlap, i.e., there is always
+a hard-sphere interaction which is a function of the separation $h$:
+\begin{equation}
+v^{hs}(h) = \left\{
+\begin{array}{ll}
+\infty & h \leq 0,\\
+0   & h > 0.
+\end{array} \right.
+\end{equation}
+The hard-sphere interaction does not give rise to a force.
+
+\subsubsection{Soft-sphere interaction}
+
+A simple soft-sphere interaction is available, the basic form
+of which is:
+\begin{equation}
+v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
+\end{equation}
+where $\epsilon$ sets the energy scale, and $\sigma$ sets the
+characteristic width. The steepness of the potential is set by
+the exponent $\nu (> 0)$.
+
+To prevent the need for calculation of long-range interactions,
+the soft-sphere potential is truncated in a ``cut-and-shift''
+approach. This is done in such a way as to smoothly match
+both the potential and the force at the cut-off distance
+$r_c$. For $r < r_c$ the potential is then
+\begin{equation}
+v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
+\label{eq_ss_shift}
+\end{equation}
+Clearly, this potential is not exactly what we first thought of as
+soft-sphere. Matching the potential smoothly ensures conservation
+of energy, while matching the force smoothly prevents potential
+instabilities in the molecular dynamics update.
+
+The soft-sphere potential can be useful as a mechanism to
+keep the particles from touching (which risks hard-sphere interactions),
+in which case is is possible to compute the interaction as a function
+of the separation $h$, instead of $r$. The force between
+two particles is computed via the derivative of equation~\ref{eq_ss_shift}
+with respect to $r$:
+\begin{equation}
+\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
+\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
+\end{equation}
+where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
+particle $i$ to the centre of $j$.
+
+\subsubsection{Leonard-Jones interaction}
+
+\subsection{Cholesteric Anchoring}
+
+We have a fluid free energy density
+\begin{equation}
+f = 
+{\textstyle\frac{1}{2}} \kappa_0 (\partial_\beta Q_{\alpha\beta})^2
++ {\textstyle\frac{1}{2}}
+ \kappa_1 (\epsilon_{\alpha\gamma\sigma} \partial_\gamma
+Q_{\sigma\beta} + 2q_0 Q_{\alpha\beta})^2
+\end{equation}
+where we have ignored the bulk terms for the time being, but we retain
+two elastic constants $\kappa_0$ and $\kappa_1$ in the distortion term.
+The cholesteric pitch $p = 2\pi/q_0$.
+
+There is also a surface term (an area density)
+\begin{equation}
+f_s = {\textstyle\frac{1}{2}} w_1 (Q_{\alpha\beta} - Q_{\alpha\beta}^0)^2
+\end{equation}
+where $Q^0_{\alpha\beta}$ is the preferred order parameter configuration
+at the surface (in the case of normal anchoring or fixed planar anchoring).
+For degenerate planar anchoring,
+following Fournier and Galatola \cite{fournier2005}, we have
+\begin{equation}
+f_s = {\textstyle\frac{1}{2}} w_1 (\tilde{Q}_{\alpha\beta}
+    - \tilde{Q}^\perp_{\alpha\beta})^2
+    + {\textstyle\frac{1}{2}} w_2 (\tilde{Q}_{\alpha\beta}
+                      \tilde{Q}_{\alpha\beta} - S^2_0)^2
+\end{equation}
+where $\tilde{Q}_{\alpha\beta} + (1/3)S_0 \delta_{\alpha\beta}$, $S_0$ is
+is a fixed surface amplitude; to obtain a prefered degenerate
+configuration we take the fluid order parameter $Q_{\alpha\beta}$ and
+project $\tilde{Q}^\perp_{\alpha\beta} = P_{\alpha\gamma}
+\tilde{Q}_{\gamma\sigma}
+P_{\sigma\beta}$, where $P_{\alpha\beta} = \delta_{\alpha\beta}
+- n_\alpha n_\beta$ and $n_\alpha$ is the local unit surface normal.
+
+The boundary condition we wish to apply are the Euler-Lagrange equations. 
+They are at the surface of the particle
+\begin{equation}
+n_\gamma \frac{\partial f}{\partial Q_{\alpha\beta,\gamma}} = 0
+\end{equation}
+where $Q_{\alpha\beta,\gamma} = \partial_\gamma Q_{\alpha\beta}$. Here,
+$n_\gamma$ is the outward unit normal at the surface (pointing into the
+fluid) \cite{skarabot}. Note that compared to the regular minimization as  
+explained in \cite{wright} this is a surface integral term that would 
+usually vanish in the bulk.
+With the
+addition of the surface term we obtain
+\begin{equation}
+n_\gamma \frac{\partial f}{\partial Q_{\alpha\beta,\gamma}}
++ \frac{\partial f_s}{\partial Q_{\alpha\beta}} = 0.
+\end{equation}
+The derivative of the distortion free energy with respect to
+$Q_{\alpha\beta,\gamma}$ gives
+\begin{equation}
+\kappa_0 n_\beta \partial_\gamma Q_{\alpha\gamma}
++ \kappa_1 n_\gamma
+(\partial_\gamma Q_{\alpha\beta} - \partial_\alpha Q_{\gamma\beta})
+- 2\kappa_1 q_0 n_\gamma \epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}.
+\end{equation}
+However, if we want a symmetric form (the derivative with respect to
+$Q_{\beta\alpha,\gamma}$ is just as good) we can write this as
+\begin{eqnarray}
+{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
++ n_\beta \partial_\gamma Q_{\alpha\gamma})
++ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
+- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
++ \partial_\beta Q_{\gamma\alpha})
+\nonumber
+\\
+- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
++ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha}).
+\end{eqnarray}
+If we add the derivative of the surface free energy,
+where we assume that the preferred orientation $Q_{\alpha\beta}^0$ is
+independent of $Q_{\alpha\beta}$, we get a full boundary condition
+for the gradient of the tensor order parameter at the solid-fluid boundary:
+\begin{eqnarray}
+{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
++ n_\beta \partial_\gamma Q_{\alpha\gamma})
++ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
+- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
++ \partial_\beta Q_{\gamma\alpha})
+\nonumber
+\\
+- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
++ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
+- w_1(Q_{\alpha\beta} - Q_{\alpha\beta}^0) = 0.
+\label{cholesteric_bc}
+\end{eqnarray}
+Note that for planar generate anchoring, the final term (the `surface
+molecular field') is replaced by
+\begin{equation}
+- w_1 (\tilde{Q}_{\alpha\beta} - \tilde{Q}^\perp_{\alpha\beta})
+- 2w_2 (\tilde{Q}_{\gamma\sigma} \tilde{Q}_{\gamma\sigma} - S^2_0)
+  \tilde{Q}_{\alpha\beta}. 
+\end{equation}
+
+\subsubsection{Discrete implementation}
+
+In three dimensions, the boundary condition Eq.~\ref{cholesteric_bc}
+contains (potentially) 15 unknown derivatives
+$\partial_\gamma Q_{\alpha\beta}$
+corresponding to the 5 independent elements of the order parameter
+tensor $Q_{xx}$, $Q_{xy}$, $Q_{xz}$, $Q_{yy}$, and $Q_{yz}$. At a
+flat surface with, e.g., $n = (1, 0, 0)$, this is reduced to five unknown
+gradients $\partial_x Q_{\alpha\beta}$ at the
+boundary if we assume the tangential gradients $\partial_y Q_{\alpha\beta}$
+and $\partial_z Q_{\alpha\beta}$ can be approximated
+using the standard differencing method involving only fluid values of
+$Q_{\alpha\beta}$. We proceed by
+computing the constant terms
+$$
+- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
++ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
+- w(Q_{\alpha\beta} - Q_{\alpha\beta}^0)
+$$
+using $Q_{\alpha\beta}$ from the adjacent fluid site, and an estimate
+of $Q^0_{\alpha\beta}$ for the appropriate anchoring type. To these
+constant terms are added the tangential gradients.
+The gradients at the surface are then computed by solving a 5x5
+linear algebra problem for $\partial_x Q_{\alpha\beta}$. This
+allows the full gradient at the adjacent fluid site to be constructed.
+
+At concave edges or corners, where it is not possible to compute the
+tangential gradients from the usual stencil as for a flat interface,
+a different approach is required. We note than an attempt to solve a
+10x10 or 15x15 linear algebra problem for the full set of unknown
+gradients has proven unreliable: in the case where $n_\gamma$ is
+normal to the coordinate directions there is simply no solution
+available. Instead we adopt an iterative approach.
+
+Consider the case of an edge where the solid neighbours
+are in the $x$ and $y$ directions.  We first make an approximation
+to the tangential gradients in the $x$ and $y$ directions by using
+a one-sided derivative from the fluid points available. This
+tangential approximation in $y$ is then used to find a first
+estimate of the gradients $\partial_x Q_{\alpha\beta}$ using the
+procedure described above. The tangential estimate in $x$ is then
+used to obtain $\partial_y Q_{\alpha\beta}$. These surface gradients
+are then used to improve the estimate of the tangential gradients,
+and the process is iterated.
+
+{\small
+\begin{table}
+\begin{center}
+
+\begin{tabular}{|c|ccccc|}
+
+\hline
+&
+$Q_{xx,x}$ & $Q_{xy,x}$ & $Q_{xz,x}$ & $Q_{yy,x}$ & $Q_{yz,x}$ \\
+\hline
+$Q_{xx}$ &
+$\kappa_0 n_x$ & $-\kappa_1 n_y$ & $-\kappa_1 n_z$ & & \\
+$Q_{xy}$ &
+$\kappa_0 n_y$ & $\kappa' n_x$ & & $-\kappa_1 n_y$  & $-\kappa_1 n_z$  \\
+$Q_{xz}$ &
+$\kappa' n_z$ & & $\kappa' n_x$ & $\kappa_1 n_z$ & $-\kappa_1 n_y$ \\
+$Q_{yy}$ &
+ & $\kappa_0 n_y$ & & $\kappa_1 n_x$ & \\
+$Q_{yz}$ &
+ & $\kappa_0 n_z$ & $\kappa_0 n_y$ & & $2\kappa_1 n_x$ \\
+\hline
+\hline
+&
+$Q_{xx,y}$ & $Q_{xy,y}$ & $Q_{xz,y}$ & $Q_{yy,y}$ & $Q_{yz,y}$ \\
+\hline
+$Q_{xx}$ &
+$\kappa_1 n_y$ & $\kappa_0 n_x$ & & & \\
+$Q_{xy}$ &
+$-\kappa_1 n_x$ & $\kappa' n_y$ & $-\kappa_1 n_z$ & $\kappa_0 n_x$ & \\
+$Q_{xz}$ &
+ & $\kappa_0 n_z$ & $2\kappa_1 n_y$ & & $\kappa_0 n_x$ \\
+$Q_{yy}$ &
+ & $-\kappa_1 n_x$ & & $\kappa_0 n_y$ & $-\kappa_1 n_z$\\
+$Q_{yz}$ &
+$\kappa_1 n_z$ & & $-\kappa_1 n_x$ & $\kappa' n_z$ & $\kappa' n_y$\\
+\hline
+\hline
+&
+$Q_{xx,z}$ & $Q_{xy,z}$ & $Q_{xz,z}$ & $Q_{yy,z}$ & $Q_{yz,z}$ \\
+\hline
+$Q_{xx}$ &
+$\kappa_1 n_z$ & & $\kappa_0 n_x$ & & \\
+$Q_{xy}$ &
+ & $2\kappa_1 n_z$ & $\kappa_0 n_y$ & & $\kappa_0 n_x$ \\
+$Q_{xz}$ &
+$-\kappa' n_x$ & $-\kappa_1 n_y$ & $\kappa' n_z$ & $-\kappa_0 n_x$ & \\
+$Q_{yy}$ &
+ & & & $\kappa_1 n_z$ & $\kappa_0 n_y$  \\
+$Q_{yz}$ &
+$-\kappa_0 n_y$ & $-\kappa_1 n_x$ & & $-\kappa' n_y$ & $\kappa' n_z$ \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Coefficients of the various derivatives of the order parameter
+tensor appearing in five equations for the independent elements of the
+order parameter. Note $\kappa_0 + \kappa_1 = \kappa'$ and all the
+coefficients have been multiplied by a factor of 2.}
+\label{tab:cholesteric_bcs}
+\end{table} 
+}
+
+\subsection{User input}
+
+The are a number of important options required for colloids. If
+none of these keys are present, no colloids will be used.
+
+\inputkey{colloid\_init}
+
+This determines how the code initialises particles (or not). As stated, the
+default is to have no particles (value \texttt{no\_colloids}). There
+are two other options at the moment.
+
+There are two options:
+
+\inputkey{colloid\_init    from\_file}
+
+requires a preprepared file of colloid information. To understand
+how to create such a file in the correct format,
+see the example program \texttt{colloid\_file.c} in the util
+directory.
+
+\inputkey{colloid\_init   random}
+
+This may be useful to initialise a limited number of colloids at
+low volume fraction. Note that there is no guarantee that the
+particles do not overlap, in which case the initialisation will
+terminate. The number of particles and their properties
+are determined by the value of the keys described below.
+
+\inputkey{colloid\_type}
+
+This controls the the type of particle. Appropriate values are
+\texttt{inactive}, giving a standard fully resolved LB colloid;
+\texttt{active} switches on the correction to BBL appropriate
+for active particles (expecting parameters $b_1$ and $b_2$ to be
+set); \texttt{subgrid} gives unresolved particles (radius $< 1$
+lattice unit) and no BBL.
+
+If more than one particle is used, all positions are set at random.
+
+\inputkey{colloid\_gravity  0.0\_0.0\_-0.0001}
+
+Sets a body force on each particle. Useful, for example, for
+sedimentation. For example, the above gives a constant force in
+the negative $z$-direction on each particle.
+The code automatically computes the compensating body force on fluid
+nodes required to give no net change in the total momentum of the
+system at each time step.
+
+\subsubsection{Interactions and the cell list}
+
+In each local domain, colloidal particles are stored in a cell list.
+This is a common structure in molecular dynamics-like problems. Here,
+it also serves as the basis for parallel communication of colloid
+information, so there are a number of constraints which it is important
+to understand. In particular, parallel communication imposes the
+constraint that there be at least 2 cells in the local domain in
+each coordinate direction.
+
+The size of the cells is computed automatically by the code based on
+the local domain size, particle size, and the cut-off distance of any
+pairwise interactions that are relevant. This will usually mean trying
+to maximise the number of cells (to reduce pairwise interaction
+calculation). The user must set the minimum allowable cell width.
+
+\inputkey{colloid\_cell\_min}
+
+This sets the minimum cell list width. This key must be set and there
+must be at least two cells per processor domain. If the minimum does
+not capture the specified interactions, a fatal error will result.
+For spherical particles, the minimum width will typically be
+$2a_h + h_c$ where $h_c$ is an interaction cut-off distance.
+
+In the case where very large particles are required, it can be useful
+to switch of the cell list. This is done via
+
+\inputkey{colloid\_cell\_list\_interactions no}
+
+The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
+there is effectively one cell in each local domain. (In practice,
+the constraint is slightly tighter than this if larger halo regions
+are in effect.) This option
+should not be required in normal circumstances.
+
+\subsubsection{Initialisation from file}
+
+A file containing initial state properties for one or more colloids
+can be read at run file. This file must be of the correct from, but
+can be ASCII or binary. To help to create such a file, an example
+is given in \texttt{./util/colloid\_file.c}. This file will be
+read correctly in both serial and in parallel. Colloid positions
+should always be in terms of the global coordinates. By default,
+the input expected is \texttt{config.cds.init.001-001}.
+All colloid output is in serial (ie., a single file is produced)
+at the moment. Again, it may be ASCII or binary.
 
@@
 of the key value pair
-\begin{verbatim}
-colloid_init    from_file
-\end{verbatim}
+
+\inputkey{colloid\_init from\_file}
 
@@
 format, which can be read in by specifying the appropriate
-\begin{verbatim}
-colloid_io_format_input   ASCII_SERIAL
-\end{verbatim}
-or \texttt{BINARY\_SERIAL} key value in both serial and parallel.
+
+\inputkey{colloid\_io\_format\_input ASCII\_SERIAL} or \inputkey{BINARY\_SERIAL}
+
+key value in both serial and parallel.
 
@@
 soft-sphere potential is then defined in the input file:
-\begin{verbatim}
-soft_sphere_on 1
-soft_sphere_epsilon 0.0004
-soft_sphere_sigma 0.1
-soft_sphere_nu 1.0
-soft_sphere_cutoff 0.25
-\end{verbatim}
+
+\inputkey{soft\_sphere\_on 1}
+
+\inputkey{soft\_sphere\_epsilon 0.0004}
+
+\inputkey{soft\_sphere\_sigma 0.1}
+
+\inputkey{soft\_sphere\_nu 1.0}
+
+\inputkey{soft\_sphere\_cutoff 0.25}
+
 where, following Eq.~\ref{eq_ss_shift}, we have
@@
 Other relevant parameters here are
-\begin{verbatim}
-colloid_cell_min 8.0
-lubrication_on 0
-\end{verbatim}
+
+\inputkey{colloid\_cell\_min 8.0}
+
+\inputkey{lubrication\_on 0}
+
 which control, respectively, the minimum cell list width used to
@@
 We have the parameters:
-\begin{verbatim}
-yukawa_on 1
-yukawa_epsilon  1.330764285
-yukawa_kappa 0.72463768115
-yukawa_cutoff 16.0
-\end{verbatim}
+
+\inputkey{yukawa\_on 1}
+
+\inputkey{yukawa\_epsilon  1.330764285}
+
+\inputkey{yukawa\_kappa 0.72463768115}
+
+\inputkey{yukawa\_cutoff 16.0}
+
 where the potential is $v(r) \sim \epsilon (-\kappa r) / r$. Again,
@@
 hydrodynamics. The relevant parameters in the input are:
-\begin{verbatim}
-isothermal_fluctuations on 
-temperature 0.0002133333
-\end{verbatim}
+
+\inputkey{isothermal\_fluctuations on}
+
+\inputkey{temperature 0.0002133333}
+
 Note that the \texttt{temperature} parameter here specifies
@@
 An Ewald sum is available for magnetic dipoles.
+
+\subsubsection{The colloid state}
+
+The full data structure for the colloid state is defined in
+\texttt{src/colloid.h} and the corresponding routines which read
+and write the values are in \texttt{src/colloid.c}
+(note colloid singular!). Most of the state corresponds clearly
+with physical properties, although there are a number of 'structural'
+elements which have no physical meaning. These include the
+\texttt{rebuild} flag, which instructs the code to reconstruct the
+boundary links if necessary, and \texttt{deltaphi} which is involved
+in accounting for conserved order parameter in the \texttt{symmetric\_lb}
+free energy. Further, not all the elements are relevant is all
+circumstances: e.g., the magnetic dipole is only relevant is magnetic
+interactions are required, and squirmer parameters are only relevant
+to active particles.
+
+Note that it is possible in principle to have magnetic active particles,
+in which case the dipole direction (\texttt{s}) and the direction of
+motion vector (\texttt{m}) are allowed to be distinct.
+
+\subsubsection{Initialisation at random}
+
+\inputkey{colloid\_random\_no} Sets the total number of particles
+
+\inputkey{colloid\_random\_a0}
+
+\inputkey{colloid\_random\_ah}
+
+\inputkey{colloid\_random\_dh}
+
+The parameter $a_0$ the nominal radius of the particle on the lattice.
+This is used to construct the links for BBL. The smallest acceptable LB
+particle is $a_0 = 1.25$ (approximately). There are 'magic' values
+including 1.25, 2.3, 3.7, 4.77 which minimise discretisation effects.
+The hydrodynamic radius of the particles is $a_h$. For all
+reasonable applications choose $a_h = a_0$. This is the radius for
+all physical purposes. The \texttt{colloid\_random\_dh} parameter
+specifies the smallest allowed surface-surface separation if two or
+more particles (or walls) are required. This is useful to prevent
+very close particles experiencing large and destabilising interactions
+at start time.
+
+The other state for the initial colloids can be set uniformly by
+means of the keys:
+
+\inputkey{colloid\_random\_v0} initial velocity vector
+
+\inputkey{colloid\_random\_w0} initial angular velocity vector
+
+\inputkey{colloid\_random\_s0} initial spin vector (only relevant in magnetic
+field)
+
+The
+
+\inputkey{colloid\_random\_m0} initial direction of motion vector for active
+particles
+
+\inputkey{colloid\_random\_b1} active parameter $b_1$
+
+\inputkey{colloid\_random\_b2} active parameter $b_2$ (should be combined
+with the \texttt{active} particle type key as discussed above).
+
+The squirmer parameters specify the surface boundary conditions in
+an approach originating with Lighthill \cite{lighthill} and Blake
+\cite{blake}. Briefly, $b_1$ sets the propulsion speed
+($U = {\frac{2}{3}} b_1$), while $b_2$ sets the
+particle stresslet. The ratio $\beta = b_2/b_1$ determines the
+mode of the squirmer motion; see \cite{isaac} for further details.
+
+There is one special case. If you want exactly one particle, you may
+set its initial position with
+
+\inputkey{colloid\_random\_r0} the initial position vector.
+
+This position must be within the current system, or no particle will
+be created.
+
+\subsubsection{Restarting at random!}
+
+Be careful if you are restarting a simulation that used the random
+initialisation at $t=0$. You must change to
+
+\inputkey{colloid\_init  from\_file}
+
+to avoid generating a completely new set of random positions
+(which will not be consistent with the density and order parameter
+fields at the restart time). This is almost certainly not what you
+intended!
+
+\subsubsection{Hydrodynamic radius}
+
+There is an important issue to understand about the (nominally)
+spherical particles in LB. The input radius $a_0$ is used to
+construct the link boundary conditions. However, discretisation
+errors in the boundary conditions means that the `real' particle
+size may be different. This real radius is the hydrodynamic radius $a_h$.
+As a further complication, the hydrodynamic radius can also
+be a function of viscosity for a given input radius. Some values
+are given in Table~X. This values are computed using a calibration
+procedure described by Nguyen and Ladd \cite{nguyen-ladd2002}.
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|}
+\hline
+$\eta$ & \multicolumn{3}{c|}{$a_0$}\\ \cline{2-4} 
+       & 1.25 & 2.30 & 4.77 \\
+\hline
+1/6  & 1.09 (1.05) & 2.23 (2.20) & 4.76 \\
+1/100  & 1.40 (1.34) & 2.50 (2.46) & 5.01 \\
+1/1000  & 1.63 & 2.71 & 5.23 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{A table of calibrated hydrodynamic radii as a function of input
+radius and fluid viscosity $\eta$. These figures are for D3Q19 (and D3Q15
+in brackets).}
+\end{table}
+
+If a calibration is required for a given radius and viscosity,
+use the \texttt{calibration on} key pair in the input file. Some
+examples are given in the \texttt{tests/calibration} directory.
+An apppropriate
+number of time steps must be used to allow the system to get over
+the initial transient (we use the momentum diffusion time for the
+system $L^2/\eta$) and to allow at least one
+particle Stokes time ($a/U$). The input radius and hydrodynamic
+radius should be set the same in the input file.
+
+% PENDING Ginzburg d'Humieres correct to ghost mode relaxation times
+
+\subsubsection{Flat wall lubrication correction}
+
+It is possible to prevent colloids colliding with the flat boundary walls
+by adding the correction to the normal component of the lubrication
+force. The general form of this correction is
+\begin{equation}
+\mathbf{F}_{\mathrm{lub}} = -6\pi \eta a_h^2
+%({\textstyle \frac{1}{h} - \frac{1}{h_c}})
+(1/h - 1/h_c)
+\mathbf{u} . \mathbf{r}.
+\end{equation}
+Here, $h$ is the surface to surface separation, and $h_c$ is the cutoff
+distance beyond which the correction is not applied. The relative
+velocity (assuming the wall is not moving) and normal separation are
+$\mathbf{u}$ and $\mathbf{r}$ respectively.
+
+The cutoff distance can be set from the input via
+
+\inputkey{boundary\_lubrication\_rcnormal 0.1}
+
+where the value is (strictly) set via calibration, cf. the hydrodynamic
+radius. In general, the cutoff is between zero and 0.5 lattice spacing,
+and can be smaller for larger particles. Because the wall velocity is
+fixed, this force correction should cause no stability issues in the
+colloid velocity update (provided a colloid's initial position is not
+very close to the wall). By default the cutoff $h_c = 0$.

### 2013-01-31 trunk/htdocs/liquidcrystal.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1902 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-05-10 htdocs/user.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  user.tex
-%
-%  The 'user' section which holds an overview, and discusses
-%  compilation and running.
-%
-%  $Id$
-%
-%  Edinburgh Soft Matter and Statistical Physics Group and
-%  Edinburgh Parallel Computing Centre
-%
-%  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2010 The University of Edinburgh
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Liquid Crystal}
 
-\section{Introduction}
-
-\subsection{Overview}
-
-We aim to provide a robust and portable code, written in C, which
-can be used to perform serial and scalable parallel simulations of
-complex fluid systems based around hydrodynamics via the Lattice
-Boltzmann method. Complex fluids are introduced via a free energy,
-of which a number of different examples are available. The preferred
-method of dealing with the corresponding order parameter equations
-is now by using finite difference. However, for the case of a binary
-fluid, the `full LB' approach, using two distributions, is retained
-as an option for the time being.
-
-\subsection{Availability}
-
-Under the auspices of the CCPs (Collaborative Computational Projects),
-and in particular CCP5, we have a CCPForge project account at
-
-\texttt{http://ccpforge.cse.rl.ac.uk/}
-
-This provides controlled access to known persons and uses version
-control via SVN, and includes bug
-tracking, and so on. The CCPForge SVN is not available to the public
-at the present time. You need to create an account at CCPForge, and
-let me or Juho know the username so you can be added to the access list.
-
-To download the entire code, you need
-\begin{verbatim}
-svn checkout --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig
-\end{verbatim}
-where \texttt{<user>} is replaced by your CCPForge user name.
-Note that this will download all the existing tags and branches
-into a new directory named \texttt{ludwig}, as well as the main code!
-On success, you should see three directories
-\begin{verbatim}
-$ ls ludwig
-branches  tags      trunk
-\end{verbatim}
-The main code is in the \texttt{trunk} subdirectory. This documentation is
-found in the file \texttt{trunk/htdocs/ludwig.tex}, unit tests are in
-the directory \texttt{trunk/tests}, and the utilities are in
-\texttt{trunk/util}, and so on.
-
-If you do not want all the tags and branches (which take up a certain amount
-of space), you can specify that you
-want only the trunk, e.g.,
-\begin{verbatim}
-svn co --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk
-\end{verbatim}
-in which case you will get a new directory called \texttt{trunk} containing
-the source files. All the instructions below refer to paths relative to
-this trunk directory.
-
-If you have problems, it may be worth while looking at the issue
-tracker on the project web page. It would be useful if all issues
-could be reported via this mechanism. If you e-mail me, I will just
-insert your e-mail into the issue tracker!
-
-\subsection{Compilation}
-
-Locate your local C compiler. The following example uses \texttt{gcc}
-in serial, and \texttt{mpicc} in parallel. For MPI in particular,
-local details may vary.
-
-The other thing required by the Makefile is the utility
-\texttt{svnversion}, which allows the compile-time SVN version number to
-be included in the executable (and hence the output of the program). This
-is useful to remember which version you were using when looking at
-results at a later date. If
-\texttt{svnversion} is not available (and it
-should be if you have downloaded the code via SVN), you may get an
-error message, but the code should still compile.
-
-\subsubsection{Serial}
-
-First, compile the MPI stub library in the \texttt{mpi\_s}
-directory. Do this by editing the Makefile, and checking the compiler
-is appropriate. You should be able to build the library and run the
-tests using, e.g.,:
-
-\begin{verbatim}
-$ make libc
-gcc -Wall -I.  -c mpi_serial.c
-ar -cru libmpi.a mpi_serial.o
-$ make testc
-gcc -Wall   mpi_tests.c -L. -lmpi
-./a.out
-Running mpi_s tests...
-Finished mpi_s tests ok.
-\end{verbatim}
-
-Now compile the main code in the \texttt{src} directory. Again
-edit the Makefile to check the compiler. At this stage, you also
-need to choose the LB model using exactly one of the preprocessor
-options for D2Q9 (fluid only problems), D3Q15, or D3Q19. Now type,
-e.g., using D3Q19:
-
-\begin{verbatim}
-$ make serial
-gcc  -D SVN_REVISION='"'`svnversion`'"' -c svn.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d2q9.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d3q15.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d3q19.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c model.c
-...
-\end{verbatim}
-
-This should provide an executable \texttt{Ludwig.exe} which is linked
-against the MPI stub library.
-
-\subsubsection{Parallel}
-
-Here, you do not need to compile the stub library. Simply compile
-the main code with, e.g.,:
-\begin{verbatim}
-$ make mpi
-mpicc -D_D3Q19_ -O3 -Wall -I. -c d2q9.c
-mpicc -D_D3Q19_ -O3 -Wall -I. -c d3q15.c
-mpicc -D_D3Q19_ -O3 -Wall -I. -c d3q19.c
-mpicc -D_D3Q19_ -O3 -Wall -I. -c model.c
-...
-\end{verbatim}
-
-Invoke the executable with the MPI launcher in the usual way for your
-system, e.g., \texttt{mpirun -np 8 ./Ludwig.exe} for 8 processors,
-and so on.
-
-\subsubsection{ANSI C}
-
-The code is ANSI C (1989), with the single exception of calls to the
-library function \texttt{erfc()} in the Ewald sum code. This
-can sometimes cause a problem, which can usually be solved with
-a compiler-specific work-around.
-
-\subsubsection{C assertions}
-
-The code makes quite a lot of use of standard C assertions, which
-are useful to prevent errors. They do result in a considerably
-slower execution in some instances, so production runs should
-switch off the assertions with the standard \texttt{NDEBUG}
-preprocessor flag.
-
-\section{User Input}
-
-The executable will look for an input file at run time.
-By default, this file is named \texttt{input} and should be in the
-current working directory. The alternative is to specify the name
-of the file explicitly on the command line:
-\begin{verbatim}
-$ ./Ludwig.exe input_file
-\end{verbatim}
-A reference input file \texttt{input.ref} is provided as a template.
-
-The contents of the input file are made up of \textit{key value}
-pairs with control the run-time behaviour. Blank lines and lines
-begining with \# are ignored as comments. The following describes
-the effect of various keys. Most keys have a default value which
-will be used by the code if the corresponding key is not present
-(or commented out) in the input file.
-
-\subsection{Setting up the input}
-
-Decide what free energy you want and check the relevant sections
-of this dcoumentation. If you want a simple fluid only, the free
-energy should be set to \texttt{none}.
-
-You should set at least the basic fluid parameters, the system
-size, and number of steps to be run.
-
-\textbf{IMPORTANT:} Although many incompatible choices of input
-parameters are trapped at run time, you must be careful in how
-the input file is set out. If you mistakenly set or unset some
-options in the input, erroneous results will surely result...
-
-\subsection{Basic run parameters}
-
-\inputkey{N\_cycles}
-
-The number of lattice Boltzmann time steps to execute. Default
-value: 0.
-
-\inputkey{N\_start}
-
-By default the code will start from time step $t = 0$. If you 
-wish to restart from a previously saved configuration, set
-the appropriate value for \texttt{N\_start}. The code will
-execute \texttt{N\_cycles} steps starting from this point.
-
-\inputkey{size}
-
-Controls the number of lattice points in the $(x, y, z)$ directions,
-respectively. Default value: 64\_64\_64. If a two-dimensional system
-is required, the $z$-direction can be set to 1.
-
-\inputkey{grid}
-
-In parallel, this sets the processor Cartesian decomposition, i.e.,
-the number of processors in each dimension of the Cartesian communicator.
-The default value may be implementation dependent as it is that returned
-by \texttt{MPI\_Dims\_create()}. For
-example:
-\begin{verbatim}
-size 64_64_64
-grid 4_2_1
-\end{verbatim}
-gives a total (physical) system size of 64 lattice sites in each
-direction, decomposed across 4 processors in the x-direction, 2 in
-the y-direction and 1 (no decomposition) in the z-direction. The
-total number of processors must therefore be 8. The local domain
-size per processor is then 16x32x64.
-
-\subsection{Fluid properties}
-
-\inputkey{free\_energy}
-
-This sets the free energy in use, and hence has many consequences.
-The default value is \texttt{none}, i.e., use a simple fluid only.
-The available choices are for symmetric binary fluids, Brazovskii
-smectics, polar active gels, and liquid crystals. The surfactant
-model is currently being rebuilt.
-
-\inputkey{viscosity}
-
-Sets the LB fluid shear viscosity $\eta$ (and related relaxation time).
-Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
-(i.e., relaxation time equal to unity).
-
-\inputkey{viscosity\_bulk}
-
-Sets the LB bulk viscosity $\zeta$ (and related relaxation time).
-This can be useful if you are concerned with incompressiblity
-violations, in which case $\zeta >> \eta$ can be useful. The
-default value is $\zeta = \eta$.
-
-\inputkey{isothermal\_fluctuations}
-
-This switches on the fluctuationing hydrodynamics. Default is off.
-
-\inputkey{temperature}
-
-The LB 'temperature' using fluctuating hydrodynamics. Safe values
-are $0.0001 > kT > 0$.
-
-\inputkey{ghost\_modes}
-
-Allows you to switch the ghost modes off in the collision stage.
-Default is on.
-
-\subsection{Symmetric binary fluid}
-
-\inputkey{free\_energy symmetric}
-
-The following applies for the binary fluid problem with
-compositional order parameter $\phi$ and free energy
-(excluding the term in the density $\rho$):
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
-\end{equation}
-
-This is described in some detail
-by Kendon et al \cite{viv}. The first two terms represent the bulk
-contribution, whereas the term in $\kappa$ penalises curvature in
-the interface.
-
-\inputkey{A} The parameter $A$. Note $A < 0$.
-
-\inputkey{B} The parameter $B$. Note $B = -A$ for common usage,
-although this is not enforced.
-
-\inputkey{K} The parameter $\kappa$, which is positive.
-
-The order parameter evolution is determined by a Cahn-Hilliard equation
-with mobility set by
-
-\inputkey{mobility} Sets mobility $M$ (uniform in space).
-
-The following parameters control the initialisation of the order
-parameter.
-
-\inputkey{phi\_initialisation}
-
-Determines how the compositional order parameter is initialised
-at the start of the run. If set to \texttt{spinodal} the value
-is set to $\phi_0$ as set above plus or minus a random noise, the
-magnitude of which is set by the value of the \texttt{noise} key.
-If set to \texttt{block} a one-dimensional profile is set up in
-the $z$-direction representing two blocks of fluid with
-$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
-$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
-appropriate width. The $\phi = -1$ section is in the middle.
-
-\inputkey{phi0}
-
-The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
-The default value is zero, i.e., a symmetric 50:50 mixture by volume.
-
-\inputkey{noise}
-
-The magnitude of the initial fluctuations in $\phi$ used to
-initiate spinodal decomposition.
-
-\subsubsection{Binary fluid using two distributions}
-
-\inputkey{symmetric\_lb}
-
-This is the special case where the composition is represented
-by a second LB distribution, and an appropriate lattice kinetic
-equation approximating the Cahn-Hilliard equation is solved.
-In this case the above parameters have the same meaning.
-
-\subsection{Brazovskii}
-
-\inputkey{free\_energy brazovskii}
-
-This is similar to the symmetric free energy, but with one extra term
-in a higher derivative of $\phi$.
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
-\end{equation}
-
-The parameters now include $C$. For $A<0$, phase separation occurs
-with a result depending on $\kappa$: ones get two symmetric phases
-for $\kappa >0$ (cf. symmetric) or a lamellar phase for
-$\kappa < 0$.
-
-\inputkey{A} Bulk parameter $A < 0$.
-
-\inputkey{B} Bulk parameter $B = -A$.
-
-\inputkey{K} Negative for lamellar phase.
-
-\inputkey{C} Positive.
-
-Other parameters, including the mobility, are set as for the
-symmetric free energy (see above).
-
-\subsection{Surfactant in binary fluid}
-
-\inputkey{free\_energy surfactant}
-
-Currently under reconstruction.
-
-This is for use with a two order parameter model the first of
-which is the composition, as for the symmetric free energy,
-while the second $\psi$ represents surfactant concentration.
-The free energy density is made up of a number of terms:
-
-\begin{equation}
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-\end{equation}
-is the standard symmetric term related to the composition.
-These parameters are represented in the input as
-
-\inputkey{surf\_A}
-\inputkey{surf\_B}
-\inputkey{surf\_kappa}
-
-\begin{equation}
-D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
-\end{equation}
-represents the energy of surfactant in the bulk phase with surfactant
-concentration $0 < \psi < 1$ with a single parameter $D$;
-\begin{equation}
--{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
--{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
-\end{equation}
-where the term in $\epsilon$ represents the energy reduction by
-adsorbing surfactant at the interface and the term in $\beta$
-provides an additional favourable term modelling the fact that
-the molecules like to line up; finally, there is an extra term
-\begin{equation}
-+{\textstyle \frac{1}{2}} W \psi \phi^2
-\end{equation}
-which is included to stabilise the interface \cite{theissengompper}.
-
-The corresponding keys for the input file are:
-
-\inputkey{D} is the bulk surfactant contribution parameter
-
-\inputkey{epsilon} is the parameter $\epsilon$
-
-\inputkey{W} is the parameter $W$
-
-\inputkey{beta} is the value of the parameter $\beta$.
-
-\inputkey{phi\_b} the initial, uniform, background surfactant
-concentration $\psi_b$.
-
-\inputkey{mobility\_psi} Sets the (uniform) mobility for the surfactant
-as appropriate for the Cahn-Hilliard equation.
-
-\subsection{Polar Active Gel}
-
-\inputkey{free\_energy polar\_active}
-
-This models a polar active gel with vector order parameter $P_\alpha$.
-The free energy density is:
-
-The corresponding input parameters are:
-
-\inputkey{polar\_active\_a}
-
-\inputkey{polar\_active\_b}
-
-\inputkey{polar\_active\_k}
-
-\inputkey{polar\_active\_klc}
-
-\inputkey{polar\_active\_zeta}
-
-\inputkey{polar\_active\_lambda}
-
-\inputkey{leslie\_ericksen\_gamma}
-
-\inputkey{leslie\_ericksen\_swim}
-
-\subsection{Liquid Crystal}
+\subsection{User input}
 
@@
 initial value specified in the above. It is also robust to restarts.
-
-\subsection{Colloidal particles}
-
-The are a number of important options required for colloids. If
-none of these keys are present, no colloids will be used.
-
-\inputkey{colloid\_init}
-
-This determines how the code initialises particles (or not). As stated, the
-default is to have no particles (value \texttt{no\_colloids}). There
-are two other options at the moment.
-
-There are two options:
-
-\inputkey{colloid\_init    from\_file}
-
-requires a preprepared file of colloid information. To understand
-how to create such a file in the correct format,
-see the example program \texttt{colloid\_file.c} in the util
-directory.
-
-\inputkey{colloid\_init   random}
-
-This may be useful to initialise a limited number of colloids at
-low volume fraction. Note that there is no guarantee that the
-particles do not overlap, in which case the initialisation will
-terminate. The number of particles and their properties
-are determined by the value of the keys described below.
-
-\inputkey{colloid\_type}
-
-This controls the the type of particle. Appropriate values are
-\texttt{inactive}, giving a standard fully resolved LB colloid;
-\texttt{active} switches on the correction to BBL appropriate
-for active particles (expecting parameters $b_0$ and $b_1$ to be
-set); \texttt{subgrid} gives unresolved particles (radius $< 1$
-lattice unit) and no BBL.
-
-If more than one particle is used, all positions are set at random.
-
-\inputkey{colloid\_gravity  0.0\_0.0\_-0.0001}
-
-Sets a body force on each particle. Useful, for example, for
-sedimentation. For example, the above gives a constant force in
-the negative $z$-direction on each particle.
-The code automatically computes the compensating body force on fluid
-nodes required to give no net change in the total momentum of the
-system at each time step.
-
-\subsubsection{Interactions and the cell list}
-
-In each local domain, colloidal particles are stored in a cell list.
-This is a common structure in molecular dynamics-like problems. Here,
-it also serves as the basis for parallel communication of colloid
-information, so there are a number of constraints which it is important
-to understand. In particular, parallel communication imposes the
-constraint that there be at least 2 cells in the local domain in
-each coordinate direction.
-
-The size of the cells is computed automatically by the code based on
-the local domain size, particle size, and the cut-off distance of any
-pairwise interactions that are relevant. This will usually mean trying
-to maximise the number of cells (to reduce pairwise interaction
-calculation). The user must set the minimum allowable cell width.
-
-\inputkey{colloid\_cell\_min}
-
-This sets the minimum cell list width. This key must be set and there
-must be at least two cells per processor domain. If the minimum does
-not capture the specified interactions, a fatal error will result.
-For spherical particles, the minimum width will typically be
-$2a_h + h_c$ where $h_c$ is an interaction cut-off distance.
-
-In the case where very large particles are required, it can be useful
-to switch of the cell list. This is done via
-
-\inputkey{colloid\_cell\_list\_interactions no}
-
-The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
-there is effectively one cell in each local domain. This option
-should not be required in normal circumstances.
-
-\subsubsection{Initialisation from file}
-
-A file containing initial state properties for one or more colloids
-can be read at run file. This file must be of the correct from, but
-can be ASCII or binary. To help to create such a file, an example
-is given in \texttt{./util/colloid\_file.c}. This file will be
-read correctly in both serial and in parallel. Colloid positions
-should always be in terms of the global coordinates. By default,
-the input expected is \texttt{config.cds.init.001-001}.
-
-All colloid output is in serial (ie., a single file is produced)
-at the moment. Again, it may be ASCII or binary.
-
-\subsubsection{Initialisation at random}
-
-\inputkey{colloid\_random\_no} Sets the total number of particles
-
-\inputkey{colloid\_random\_a0}
-
-\inputkey{colloid\_random\_ah}
-
-\inputkey{colloid\_random\_dh}
-
-The parameter $a_0$ the nominal radius of the particle on the lattice.
-This is used to construct the links for BBL. The smallest acceptable LB
-particle is $a_0 = 1.25$ (approximately). There are 'magic' values
-including 1.25, 2.3, 3.7, 4.77 which minimise discretisation effects.
-The hydrodynamic radius of the particles is $a_h$. For all
-reasonable applications choose $a_h = a_0$. This is the radius for
-all physical purposes. The \texttt{colloid\_random\_dh} parameter
-specifies the smallest allowed surface-surface separation if two or
-more particles (or walls) are required. This is useful to prevent
-very close particles experiencing large and destabilising interactions
-at start time.
-
-The other state for the initial colloids can be set uniformly by
-means of the keys:
-
-\inputkey{colloid\_random\_v0} initial velocity vector
-
-\inputkey{colloid\_random\_w0} initial angular velocity vector
-
-\inputkey{colloid\_random\_s0} initial spin vector
-
-\inputkey{colloid\_random\_m0} initial direction of motion vector for active
-particles
-
-\inputkey{colloid\_random\_b1} active parameter $b_1$
-
-\inputkey{colloid\_random\_b2} active parameter $b_2$ (should be combined
-with the \texttt{active} particle type key as discussed above).
-
-There is one special case. If you want exactly one particle, you may
-set its initial position with
-
-\inputkey{colloid\_random\_r0} the initial position vector.
-
-\subsubsection{Restarting at random!}
-
-Be careful if you are restarting a simulation that used the random
-initialisation at $t=0$. You must change to
-
-\texttt{colloid\_init  from\_file}
-
-to avoid generating a completely new set of random positions
-(which will not be consistent with the density and order parameter
-fields at the restart time). This is almost certainly not what you
-intended!
-
-\subsubsection{Hydrodynamic radius}
-
-There is an important issue to understand about the (nominally)
-spherical particles in LB. The input radius $a_0$ is used to
-construct the link boundary conditions. However, discretisation
-errors in the boundary conditions means that the `real' particle
-size may be different. This real radius is the hydrodynamic radius $a_h$.
-As a further complication, the hydrodynamic radius can also
-be a function of viscosity for a given input radius. Some values
-are given in Table~X. This values are computed using a calibration
-procedure described by Nguyen and Ladd \cite{nguyen2001}.
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-\hline
-$\eta$ & \multicolumn{3}{c|}{$a_0$}\\ \cline{2-4} 
-       & 1.25 & 2.30 & 4.77 \\
-\hline
-1/6  & 1.09 (1.05) & 2.23 (2.20) & 4.76 \\
-1/100  & 1.40 (1.34) & 2.50 (2.46) & 5.01 \\
-1/1000  & 1.63 & 2.71 & 5.23 \\
-\hline
-\end{tabular}
-\end{center}
-\caption{A table of calibrated hydrodynamic radii as a function of input
-radius and fluid viscosity $\eta$. These figures are for D3Q19 (and D3Q15
-in brackets).}
-\end{table}
-
-If a calibration is required for a given radius and viscosity,
-use the \texttt{calibration on} key pair in the input file. Some
-examples are given in the \texttt{tests/calibration} directory.
-An apppropriate
-number of time steps must be used to allow the system to get over
-the initial transient (we use the momentum diffusion time for the
-system $L^2/\eta$) and to allow at least one
-particle Stokes time ($a/U$). The input radius and hydrodynamic
-radius should be set the same in the input file.
-
-% PENDING Ginzburg d'Humieres correct to ghost mode relaxation times
-
-\subsection{Boundary conditions}
-
-Solid planar boundary walls can be added at the extremities of the
-system (i.e., the volume of fluid is exactly that specified by
-the \texttt{size} keyword). The wall can be in one, two or all three
-coordinate directions, i.e., a 'slab' geometry, and 'duct' geometry
-or a 'box' geometry, respectively.
-
-The boundary walls are switched on via the keyword
-
-\inputkey{boundary\_walls}
-
-and with a complementatary specification of the key \texttt{periodicity}.
-
-For example, walls at $x = x_{\min}$  and $x = x_{\max}$  are specified
-by
-
-\inputkey{boundary\_walls 1\_0\_0}
-
-\inputkey{periodicity 0\_1\_1}
-
-Both must be present. Note that there are a number of constraints on
-valid choices, and the code will stop if there is an invalid choice, or
-the two keys above do not match. Common choices are:
-\texttt{boundary\_walls 0\_0\_1},
-which gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction; \texttt{0\_1\_1}
-gives periodic conditions in the $x-$ direction only, while
-\texttt{1\_1\_1} is a fully enclosed box.
-
-In the case of the slab geometry with walls at $z = z_{\min}$ and
-$z= z_{\max}$ (only),
-two further values are used so that shear may be imparted to the system:
-
-\inputkey{boundary\_speed\_bottom}
-
-\inputkey{boundary\_speed\_top}
-
-These values set the $x-$component of the wall velocity at
-$z = z_{\min}$ and $z = z_{\max}$, respectively. Like all
-velocities, these values must respect the low Mach number
-constraint $u_x < c_s$.
-
-There is a further option for use with the special case in which
-top and bottom walls in the $z-direction$ are used to drive a shear
-flow (typically with the top speed positive and the bottom speed
-negative). The key
-
-\inputkey{boundary\_shear\_init}
-
-may be switched on (set to 1) to provide initialisation of the
-LB distributions appropriate for the specified shear rate based
-on a linear velocity profile in the $z-$direction. The default
-is to initialise the fluid at rest.
-
-\subsection{Lees-Edwards planes}
-
-\inputkey{N\_LE\_plane}
-
-sets the total number of Lees-Edwards planes. Default is zero.
-The placing is as followings. The number of planes $n$ must
-divide the lattice size in the $x$-direction to give an integer
-$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
-
-\inputkey{LE\_plane\_vel}
-
-sets the velocity of each plane relative to the lattice. All planes
-have the same, constant, velocity.
-
-\inputkey{LE\_init\_profile}
-
-if set to 1, the fluid velocity is initialise to reflect a steady
-state shear flow appropriate for the number of planes at the
-given velocity. If set to zero, the fluid is initialised with
-zero velocity.
-
-Note that when the Lees Edwards boundaries are in place, output
-files containing lattice quantities must be `unrolled' to remove
-the time-dependent displacement of the planes. This is done using
-the extraction utility (see section on parallel I/O).
-
-The code works out the
-current displacement of the planes by computing $U_{LE} t$, where
-$t$ is the current time step. A shear run should then start from
-$t = 0$, otherwise an offset is required.
-
-\inputkey{LE\_time\_offset}
-
-It is often convenient to run an equilibration with no shear, and
-then to start an experiment after some number of steps. This
-key allows you to offset the start of the Lees-Edwards motion.
-It should then take the value of the start time (in time steps)
-corresponding to the restart at the end of the equilibration
-period. You also need to take this value into account when unrolling
-the output.
-
-\subsubsection{Lees-Edwards planes in parallel}
-
-There are a couple of additional constraints to use the Lees-Edwards
-planes in parallel. In particular, the planes cannot fall at a
-processor boundary in the $x$-direction. This means you should
-arrange an integer number of planes per process in the $x$-direction.
-(For example, use one plane per process; this will also ensure the number
-of planes
-still evenly divides the total system size.)
-This will interleave the planes with the processor decomposition.
-The $y$-direction and $z$-direction may be decomposed without
-further constraint.
-
-Note that this means a simulation with one plane will only work
-if there is one process in the $x$ decomposition.
-
-\subsection{Input/Output}
-
-\subsubsection{Serial I/O and standard output}
-
-A variety of information is printed to standard output during the
-course of the simulation. This should allow the user to keep track
-of basic progress in terms of integrated quantities, total mass,
-total momentum and so on. Frequency of output is controlled by
-
-\inputkey{freq\_statistics}
-
-which is an integer. Note that these global statistics require
-global communication, and so can affect performance in parallel.
-It is recommended that the frequency is kept to a minimum for
-large systems (e.g., every 500 or 1000 time steps). This
-consideration is true for most types of output (see section
-on parallel I/O).
-
-\subsubsection{Configurations}
-
-It may be appropriate to save entire model configurations from
-time to time, and certainly when a restart is wanted. For each lattice
-quantity, in addition to the data themselves, the program produces a
-single meta-data file (ending in
-\texttt{.meta}) which is used to describe these data. Colloid
-data does not require a meta-data file at the moment.
-
-\inputkey{freq\_config}
-
-This is an integer and sets the frequency of full configuration dumps.
-These may be large depending on the size of the system.
-Full configurations should consist of the LB distributions,
-order parameter (if present) and colloids (if present).
-
-\inputkey{config\_at\_end [yes | no]}
-
-forces (or switches off) a configuration dump at the end of the run.
-The default is to produce a configuration at the end of the run.
-
-\subsubsection{Other lattice quantities}
-
-A number of other lattice-based quantities are available as output.
-
-\inputkey{freq\_measure}
-
-sets the frequency of recording of colloid/order parameter output.
-This includes the combined scalar order parameter and director
-field output in the case of the liquid crystal free energy.
-
-\inputkey{freq\_phi}
-
-set frequency of order parameter output.
-
-\inputkey{freq\_vel}
-
-set frequency of velocity output. Note that all the hydrodynamic
-quantities can be reconstructed from the distribution output, if
-available. If only the velocity field is required, this method
-allows a considerable saving in storage.
-
-\subsubsection{I/O format}
-
-\inputkey{vel\_format}
-
-\inputkey{phi\_format}
-
-May be either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-Configuration I/O is always in binary. It is recommended to use
-binary output for speed and disk space considerations.
-
-\subsubsection{Parallel I/O}
-
-In serial, each output quantity appears in one file (\texttt{phi-}
-and so on). In parallel, this is also the case by default. However,
-this does not scale, owing to the fact that each process has to
-write to the same file in turn (ie., in serial). The solution is
-to make the I/O parallel. This is done by splitting the domain into
-different ``I/O grids'' each of which writes to a separate file.
-
-The default I/O grid can be set using, e.g.,
-
-\inputkey{default\_io\_grid 2\_2\_2}
-
-meaning that the domain is decomposed into a 2x2x2 Cartesian
-decomposition for I/O purposes. This will give rise to a total
-of 8 files per I/O event with extensions \texttt{.008-001} to
-\texttt{.008-008}. These files must be recombined if analysis
-is required. However, restarting can be performed with the
-parallel files.
-
-Individual lattice quantities can use separate I/O grids. This
-may be particularly useful for the distribution data, which are
-much larger than the other lattice quantities. This means a
-finer decomposition may be appropriate. Clearly, the size of the
-I/O decomposition cannot exceed the processor decomposition.
-See the advice in section \cite{section-advice} for guidelines.
-
-\subsection{Miscellaneous}
-
-\inputkey{random\_seed}
-
-sets the random number generator seed.
-
-\subsection{Dealing with the Parallel I/O files}
-
-When running in parallel, output of lattice-based (and particle)
-quantities takes place in parallel. As a consequence, the order
-of the output in the files is not in the same order as would be
-the case in serial. This means the parallel output must be
-manipulated into the correct order before any analysis.
-
-Parallel I/O subdivides the processors into one or more groups,
-each of which write data to a separate file. Information on the
-content of each of these files is contained in `meta-data' file
-for each quantity, which are produced automatically at the start
-of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
-If there are 8 I/O groups, the files will be named \texttt{008-001}
-to \texttt{008-008} for the different groups.
-
-The actual data is stored for each relevant time step as, e.g.,
-\texttt{phi-000100.001-001} where the extension refers to the
-I/O. To recombined a single, or multiple, data files into a
-single file in the correct order, a utility program
-\texttt{extract.c} is provided.
-
-The executable takes two arguments, e.g.,
-
-\inputkey{./extract phi.001-001.meta phi-000100.001-001}
-
-where  the first is the first meta-data file in the relevant series,
-and the second is the first data file for the relevant time step. The
-program will combine the relevant quantities in the correct order
-and produce a single new file \texttt{phi-000100} (in the above case).
-
-\subsection{Restarting}
-
-When restarting in parallel, the processor decomposition must be
-preserved. An attempt to change the processor decomposition will
-cause the configuration files to be read in the incorrect order,
-and results will be wrong.
-
-\section{Parallel Performance}
-
-\subsection{General Comments}
-
-The basic LB calculation should scale well in parallel, that is,
-the time taken for fixed problem size will decrease linearly
-with the number of MPI processes (strong scaling), or a larger
-problem can be run on a proportionally larger number processes
-in the same time (weak scaling). To ensure performance is retained
-in parallel, it is useful to understand some basic considerations.
-Some of these are discussed below.
-
-\textit{Ludwig} has been run successfully on up to 131,072 MPI
-tasks with close to ideal scaling, and there is no problem in
-principle to run larger decompositions.
-
-\subsection{Things to Remember and Things to Avoid}
-\ref{section-advice}
-
-There are a number of simple issues to get right:
-\begin{itemize}
-\item
-Make sure the assertions are switched off and appropriate optimisations
-are switched on for production runs (see Makefile).
-\item
-Do not use too small a local domain size. Typically, 16$^3$ to 32$^3$
-cubic local domains will give resaonble scaling. Anything smaller may
-be inefficient. The best size may depend on the exact nature of the
-calculation and the hardware. Note that this limitation will always
-ultimately limit strong scaling.
-\item
-Don't have diagnsotic output (freq\_statistics) more often than is
-necessary. This output requires a global communication which is slow.
-\item
-Use BINARY output options; ASCII output is slower, and produces larger
-files.
-\item
-Do not use serial I/O. This will have an increasingly high overhead as
-the number of processors is increases (it's probably reasonable to
-128--256 MPI tasks). This is because each MPI tasks writes its data,
-in turn, to a single file; this in inherently serial. To avoid this,
-you need to specify the I/O decomposition for the various output
-quantities of interest. This splits the system into regular blocks
-(in a similar approach to the MPI decomposition), the data for which
-are written to separate files in parallel. For example, if the MPI
-decomposition is 8\_8\_8 (512 MPI tasks), then a reasonable I/O
-decomposition may be 2\_2\_2 or 4\_4\_4. The only way to find out
-what works best is to try some tests for the problem size you want to
-address.
-\item
-Don't use the Ewald sum! This simply does not scale.
-An alternative algorithm is required.
-\end{itemize}

### 2013-01-31 trunk/htdocs/polaractive.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1902 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-01-14 branches/CUDA/htdocs/sections.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  sections.tex
-%
-%  Content for Ludwig documentation.
-%
-%  $Id: sections.tex,v 1.15 2010-10-21 18:05:50 kevin Exp $
-%
-%  Edinburgh Soft Matter and Statistical Physics Group and
-%  Edinburgh Parallel Computing Centre
-%
-%  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The Lattice Boltzmann Method}
-
-We review here the lattice Boltzmann method applied to single
-phase problems.
-
-\subsection{The Navier Stokes Equation}
-
-We seek to solve the isothermal Navier-Stokes equations which, often
-written in vector form, express mass conservation
-\begin{equation}
-\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
-\label{eq_mass1}
-\end{equation}
-and the conservation of momentum
-\begin{equation}
-\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
--\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
-+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
-\label{eq_momentum1}
-\end{equation}
-Equation~\ref{eq_mass1} expresses the local rate of change of the
-density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
-mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
-Equation~\ref{eq_momentum1} expresses Newton's second law for
-momentum, where the terms on the right hand side represent the
-force on the fluid.
-
-For this work, it is more convenient to rewrite these equations
-in tensor notation, where Cartesian coordinates ${x,y,z}$ are
-represented by indices $\alpha$ and $\beta$, viz
-\begin{equation}
-\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
-\end{equation}
-and
-\begin{equation}
-\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
-= -\nabla_\alpha p
-+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
-+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
-\end{equation}
-Here, repeated indices are understood to be summed over.
-The conservation law is seem better if the forcing terms of the
-right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
-so that
-\begin{equation}
-\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
-\end{equation}
-In this case the expanded expression for the stress tensor is
-\begin{equation}
-\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
-+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
-\end{equation}
-where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
-$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
-modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.
-
-\subsection{The Distribution Function and its Moments}
-
-In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
-\begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
-\end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
-
-In general, the number of modes supported in $n$ and these can
-be written as
-\begin{equation}
-M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
-\end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
-\begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
-\end{equation}
-
-\subsubsection{Hydrodynamic modes}
-
-\subsubsection{D3Q15 ghost modes}
-
-\subsubsection{D3Q19 ghost modes}
-
-\subsection{The Lattice Boltzmann Equation}
-
-\subsection{The Collision}
-
-\subsection{External Body Forces}
-
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsubsection{Fluctuating ghost modes: D3Q15}
-
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Boundary walls}
-
-\subsection{Colloids}
-
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
 \section{Polar Active Gel}
@@
 
-\section{Code Design}
+\subsection{User input}
 
-\subsection{Some Comments}
+\inputkey{free\_energy polar\_active}
 
-In designing a code in C the main object of which is to do high
-performance computing, there are some inevitable tensions with
-software engineering best-practice. In particular, complete
-encapsulation of data in the spirit of using abstract data types
-tends to cause problems of performance.
+This models a polar active gel with vector order parameter $P_\alpha$.
+The free energy density is:
 
-For example, encapsulation of the lattice Boltzmann distribution
-$f_i$, together with appropriate ``get'' and ``set'' functions for
-access when approapriate, is possible. There could then be a
-complete separation of interface and implementation of the
-distribution (other than that each individual $f_i$ is represented
-by a \texttt{double}). However, experience suggests that
-this imposes severe impediment to optimisation and hence
-unacceptable performance degradation.
+The corresponding input parameters are:
 
-Schemes do exist by which one can approximate the object
-oriented programming style in C, and so achieve a separation
-of interface and implementation. However, this tends to put
-a heavy burden on the use of \texttt{void} pointers, itself
-a recipe for possible disaster. A truly object-oriented
-language such as Java would be preferable.
+\inputkey{polar\_active\_a}
 
-So, for the time being, some relaxation of encapsulation is required.
-Exposing a pointer to the distributions which allows them to be accessed
-directly where required. This can be done merely by declaring the
-necessary variables \texttt{extern}. In practice, such deviations
-from encapsulation can be minimised in the lattice sector of the
-code. In particlar, \texttt{extern} declarations should not be
-required in \texttt{main()}, i.e., it should be able to write a
-new application without worrying about details of implementation.
+\inputkey{polar\_active\_b}
 
-\subsubsection{Objects}
+\inputkey{polar\_active\_k}
 
-Somewhat more difficult is the part of the code involving
-colloids, which would really favour a truly object-oriented
-approach.
+\inputkey{polar\_active\_klc}
 
-\subsubsection{Serial and parallel}
+\inputkey{polar\_active\_zeta}
 
-The design decission has been taken to support execution of the
-code in both serial and parallel. The default compilation
-assumes that a serial executable is required, while  MPI code
-is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).
+\inputkey{polar\_active\_lambda}
 
-There is no shared memory inplementation at this time.
+\inputkey{leslie\_ericksen\_gamma}
 
-\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}
-
-\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}
-
-\begin{figure}
-\begin{center}
-%\includegraphics{xfig/sketch.eps}
-\end{center}
-\caption{A scheme by which different parts of the code might be
-encapsulated and in such a way to allow reliable development and
-testing. As this is C, will do not adopt a truly ``object'' langauge.
-Broadly, different modular parts should only depend on stuff above
-in the diagram. So, at the top, there is the parallel environment
-where communication occurs within MPI\_COMM\_WORLD. Built on top
-of this is the Cartesian Communicator responsible for halo swaps
-and so forth. The LE code may have a separate Communicator again. 
-(There's an argument to say the diagram should be the other way up!)}
-\end{figure}
-
-\section{Code Implementation}
-
-\subsection{Stuff}
-
-\subsection{Coordinate System}
-
-\subsubsection{The global coordinate system}
-
-The user specifies the system size in lattice units at run time,
-which sets the extent of memory which must be allocated for the
-Cartesian lattice. The extent of the physical coordinate system
-then reflects the lattice size directly (see Figure \ref{fig_c1}). 
-Each lattice site, or node,  has coordinates corresponding to its
-integer position in the system and can be thought of as being
-at the centre of a control volume one lattice unit on a side.
-This means that the nominal edges of the
-system are offset from the lattice sites themselves by half a
-lattice unit in each direction.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_c1.eps}
-\end{center}
-\caption{The global Cartesian coordinate system reflects the choice
-of lattice size; lattices sites are located at
-$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
-(the Figure only shows the labels in the $y-$direction for simplicity).
-In the Figure, the lattice sites in the domain proper are the full
-circles; halo sites, which are used to effect periodic boundary
-conditions, are included as open cirles}
-\label{fig_c1}
-\end{figure}
-
-For particles, the centres of which can move continuously across the
-lattice, the extents of the global coordinate system are then
-$L_{min} < x < L + L_{min}$ and so on.
-
-\subsubsection{Addressing the lattice: the local coordinate system}
-
-\subsection{Solid walls}
-
-General solid objects which are fixed for the duration of
-the simulation, such as porous media, can be defined
-by assigning solid/fluid status to each lattice node. 
-The links required between solid and fluid nodes can
-then be identified for the given model.
-
-\textit{Comment.} In general, a solid object will not
-affect the periodic nature of the simulation. A general
-solid object cannot move, i.e., the boundary velocity
-must be zero.
-
-\textit{Special case.} In the case where the solid object
-represents a flat boundary wall at the edge of the system
-in one or more directions, the periodicity is broken in
-the corresponding directions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LE Implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Lees Edwards Sliding Period Boundary Conditions}
-
-\subsubsection{The LE module}
-
-\subsubsection{Distributions crossing the LE planes}
-
-With the planes conceptually half-way between lattice sites, the
-propagation moves some of the distributions (namely, those with
-$c_x = \pm1$) between different sliding blocks. While the collision
-stage is unaffected, action must be taken to adjust these
-distributions when the LE boundary conditions are active. This
-is done in a two-stage process described by Adhikari \textit{et al.}
-\cite{adhikari_desplat}, implemented between the collision and
-propagation.
-
-\textbf{a. Reprojection} The post-collision distributions with
-$c_x = \pm 1$ at sites adjacent to a boundary are modified to
-take account of the velocity jump $\pm u^{LE}_y$ between the sliding
-blocks. In terms of the hydrodynamic moments we have:
-\begin{eqnarray}
-\rho &\rightarrow& \rho', \\
-\rho u_\alpha &\rightarrow& (\rho u_\alpha)' \pm \rho' u^{LE}_\alpha, \\
-S_{\alpha\beta} &\rightarrow&
-S'_{\alpha\beta} \pm (\rho u_\alpha)' u^{LE}_\beta
-\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
-\end{eqnarray}
-If we work with the changes to the moemnts, then the density is
-unaffected $\delta\rho =  0$, the velocity is changed by
-$\delta u_\alpha = \pm u^{LE}_\alpha$, with an analogous expression for
-the change in the stress
-\begin{equation}
-\delta S_{\alpha\beta} =
-\pm (\rho u_\alpha)' u^{LE}_\beta
-\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
-\end{equation}
-We can then work out the change in the distributions by reprojecting
-via Eq.\ref{eq:}, i.e.,
-\begin{equation}
-f_i \rightarrow f_i' + w_i \Bigg(
-\frac{\rho \delta u_\alpha c_{i\alpha}}{c_s^2} +
-\frac{\delta S_{\alpha\beta}Q_{i\alpha\beta}}{2c_s^4} \Bigg).
-\end{equation}
-A similar set of expressions are given for the order parameter
-distributions in Adhikari \textit{et al.} \cite{adhikari_desplat}.
-
-\textbf{b. Interpolation of reprojected distributions}
-As the sliding blocks are displaced continuously with
-$\delta y = u_y^{LE}t$, which is not in general a whole number of
-lattice spacings, an interpolation is also required before
-propagation can take place. Consider the situation (Fig.~\ref{fig_le2}) from
-a stationary frame where the frame above has translated a small
-distance to the right ($\delta y < \Delta y$). In the stationary frame
-we must interpolate the ditributions with $c_x = 1$ to a
-`departure point' from which the propagation will move the
-distribution exactly to the appropriate lattice site in the moving
-frame. Schemeatically,
-\begin{equation}
-f_i'(x, y + \delta y, z) = 
- (1 - \delta y) f_i^\star(x, y, z) +
-\delta y f_i^\star(x, y + \Delta y, z),
-\end{equation}
-where $f_i'$ here represents the interpolated value and $f_i^\star$
-is the reprojected post-collision quantity.
-
-In the case where the displacement $\delta y > 1$, the relevant
-lattice sites involved in the interpolation are displaced to the right
-by the integral part of $\delta y$, while the relative weight given to
-the two sites involves
-the fractional part of $\delta y$. To take account of the periodic
-boundary conditions
-in the $y$-direction, all displacements are modulo $L_y$.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_le2.eps}
-\end{center}
-\caption{Schematic diagram showing interpolation of distributions in the
-stationary plane (below) in preparation for propagation into the moving
-frame (above). For a small displacement $\delta y = u^{LE} t$, the
-interpolation takes place between lattice sites in the stationary frame
-(green, left). Propogation from the interpolated positions (red) can
-then take place into the moving frame (right).}
-\label{fig_le2}
-\end{figure}
-
-In parallel, exactly the same calculation is undertaken. However,
-communication is required to take account of the processor
-decomposition in the $y$-direction. This requires the identification
-of the two processes which hold the relevant information as a function
-of displacement.
-
-Note that the interpolated values are stored in the appropriate lattice
-site in the rest frame that ensures the normal propagation stage moves all
-distributions to the correct destination in memory.
-
-\subsubsection{Order parameter gradients}
-
-When the order parameter gradient is computed at a given point
-near the LE boundaries, the stencil may extend out of the stationary
-frame. The process of interpolation is slightly different to that
-for the distributions.
-
-Here, we need to interpolate values in the moving frame to the position
-which is seen by the rest frame. Figure~\ref{fig_le3} shows a five-point
-stencil to calculate the gradient at the central point $(x,y,z)$ in the
-rest frame. An interpolation of the $\phi$-field of the form
-\begin{equation}
-\phi'(x + \Delta x, y - \delta y, z) =
-\delta y \phi(x + \Delta x,  y - \Delta y, z) + 
-(1 - \delta y) \phi(x + \Delta x, y, z)
-\end{equation}
-must then take place. In practice, the interpolated values are stored
-in the offset buffer. As the gradient calculation requires $\phi$
-values in the halo regions, interpolated values for the buffer halo
-region are also required.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_le3.eps}
-\end{center}
-\caption{Schematic diagram showing a five-point stencil to compute a
-gradient at the central point in the stationary frame. The upper point
-of the stencil is in the moving frame and so must be interpolated to
-the correct position (red sites).}
-\label{fig_le3}
-\end{figure}
-
-In parallel, the interpolation requires data from at most two adjacent
-processes in the along-plane $y-$direction as long as the $\phi$ halo
-regions are up-to-date. The rank of the processes involved is
-determined by the displacement $\delta y$ as a function of time.
-
-\subsubsection{Velocities and finite-difference fluxes}
-
-For the finite-difference implementation, the velocity field at
-the cell boundaries is required. At the planes, this means an
-interpolation which takes the same form as that for
-$\phi$ is needed.
-
-In computing the advective fluxes between cells, we must then
-allow for the presence of the planes. This is done by computing
-separately the 'east' and 'west' face fluxes in the $x$-direction
-(the flux that crosses the plane). Away from the planes, face-flux
-uniqueness $f_{ijk}^e = f_{i+1jk}^w$ ensures conservation of order
-parameter. However, at the planes, the non-linear conbination of
-velocity field and order parameter field does not in general give
-rise to consistent fluxes. In order to restore conservation, we
-reconcile east and west face fluxes at the plane by taking an
-average
-\begin{eqnarray}
-f_{ijk}^{e\star}&=&{\textstyle\frac{1}{2}}(f_{ijk}^e + f_{i+1j'k}^w),\quad 
-f_{i+1j'k}^w  = \delta y f_{i+1 j-1 k}^w + (1 - \delta y) f_{i+1jk}^w,\\
-f_{i+1 j k}^{w\star} &=& {\textstyle\frac{1}{2}}(f_{ij'k}^e + f_{i+1jk}^w),\quad
-f_{ij'k}^e = (1 - \delta y) f_{ijk}^e + \delta y f_{ij+1k}^e.
-\end{eqnarray}
-In this way, we average the east face flux in the rest frame $f_{ijk}^e$
-with the interpolated value of the west face flux in the moving frame
-$f_{i+1j'k}^w$, and vice-versa. One can then easily show that when
-integrated over the length of the plane, the average fluxes are
-consistent, i.e.,
-\begin{equation}
-\sum_j (f_{ijk}^{e\star} - f_{i+1jk}^{w\star}) = 0 \quad\forall k,
-\end{equation}
-thus ensuring order parameter conservation.
-
-A similar correction is required when computing the force on the fluid
-directly via the divergence of the thermodynamic stress
-$F_\alpha = \nabla_\beta P_{\alpha\beta}^{th}$. This is implementated
-by computing fluxes of momentum at cell faces in each direction, and
-then taking the divergence to compute the force. At the planes, an
-averaging procedure is again used to ensure conservation of momentum.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% End LE implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\vfill
-\pagebreak
-
-\section{Code Interface}
-
-\subsection{The Parallel Environment}
-
-\texttt{\#include ``pe.h''}
-
-As the code is to be available in both serial and parallel versions,
-it is desirable to abstract the basic environment to some extent.
-The parallel environment is intended to support MPI communication
-within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
-consistent picture in serial (i.e., rank zero is the single process).
-
-\texttt{void pe\_init(int argc, char ** argv)}\\
-Responsible for initialisation of MPI, and hence must be first
-execuatable statement. The arguments are those to \texttt{main}
-which will be passed through to \texttt{MPI\_Init()}.
-
-\texttt{int pe\_rank(void)}\\
-Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
-serial.
-
-\texttt{int pe\_size(void)}\\
-Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
-1 in serial.
-
-\texttt{void pe\_finalise(void)}\\
-Responsible for \texttt{MPI\_Finalize()}, and must be the last
-executable statement.
-
-\texttt{void info(const char * fmt, ...)}\\
-Behaves like \texttt{printf}, but only produces output on the
-root process.
-
-\texttt{void verbose(const char * fmt, ...)}\\
-Behaves as \texttt{printf} and reports the rank of the
-calling process in \texttt{MPI\_COMM\_WORLD}.
-
-\texttt{void fatal(const char * fmt, ...)}\\
-Prints an error message and identifies
-the rank of the calling process before terminating the program.
-
-\subsection{Run Time Input}
-
-\texttt{\#include ``runtime.h''}
-
-The ability to set a wide range of parameters at run time is
-very convenient, and is supported via an input file in which the
-user specifies parameters as key value pairs. The code maintains
-a list of key value pairs from which the user can retrieve the
-appropriate values for given keys.
-
-\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
-This reads the named input file and sets up the list of keys.
-
-\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The number
-of key matches found is returned (i.e., 0 or 1).
-
-\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The
-number of key matches is returned.
-
-\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
-Set \texttt{value} to the string assocaited with \texttt{key}, if
-present. Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
-If \texttt{key} is present, set the associated 3-vector of integers.
-Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
-double v[])}\\
-If \texttt{key} is present, set the associated 3-vector of doubles.
-Returns the number of key matches found.
-
-\textit{Comment:} We probably need a function to report unused keys, so
-the user can be alerted if a key has unexpectedly been ignored.
-
-\subsection{Random Number Generators}
-
-\texttt{\#include ``ran.h''}
-
-Note that the random number generation in parallel is
-decomposition dependent.
-
-\texttt{void rand\_init(void)}\\
-Responsible for initialising RNG state.
-
-\texttt{double rand\_serial\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$ from
-the serial generator.
-
-\texttt{double rand\_serial\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the serial generator.
-
-\textit{double rand\_parallel\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$
-from the serial generator.
-
-\texttt{double rand\_parallel\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the parallel generator.
-
-\textit{Comment:} The serial generator will produce the
-same sequence on any MPI process, while the parallel generator
-will produce different sequences. The latter is intended for
-generating, specifically, fluctuations, where the need for
-decomposition-independance is questionable.
-
-\subsection{Timers}
-
-\texttt{\#include ``timer.h''}
-
-Performance figures require timing for different sections
-of the code. The code provides a set of routines to time
-standard and user-defined sections of code. Timers are
-started and stopped by call to the appropriate routines.
-
-\texttt{TIMER\_init()}\\
-This initialises all the timers to zero (e.g., at the start of
-the program).
-
-\texttt{TIMER\_start(const int t\_id)}\\
-Start the given timer.
-
-\texttt{TIMER\_stop(const int t\_id)}\\
-Stop the given timer.
-
-\texttt{TIMER\_statistics(void)}\\
-Prints the current state of all the active timers to standard
-output on the root process in a digestible form. In serial,
-the minimum, maximum, and mean time for each timer is reported
-in seconds (based on the number of start/stop cycles). In
-parallel, the same figures are computed on each process and the
-result of the corresponding \texttt{MPI\_Reduce()} operation is
-printed on the root process.
-
-The code provides a number of timers which would time standard
-parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
-total execution time, and a number of spare timers for users to
-use at will.
-
-\textit{Comment:}
-No intra-run variance (i.e., between time steps) is provided, as the
-calculation required is considerably more complex to generalise, and
-requires memory proportional to the number of time steps used.
-
-\textit{Comment:}
-Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
-provided to insert calls to the above functions for optional timing
-of performance-sensitive regions
-of the code in conjunction with the preprocessor flag
-\texttt{-D\_TIMING\_MACRO\_ON\_}.
-
-\subsection{Model Time Control}
-
-This provides time step control, event frequency (?)
-
-\texttt{void init\_control(void)}\\
-Initialise time controls.
-
-\texttt{int get\_step(void)}\\
-Return the current model time step.
-
-\texttt{int next\_step(void)}\\
-Increments the current time step. Returns number of steps left
-to run (i.e., 0 when finished).
-
-\subsection{Coordinate System}
-
-\texttt{\#include ``coords.h''}
-
-A three-dimensional Cartesian coordinate system is used throughout,
-with the exact system size usually chosen by the user at run time.
-The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
-used to identify the coordinate directions.
-
-This also provides access to the MPI Cartesian Communicator which
-is responsible for most communication in the model.
-
-\texttt{int N\_total(const int dim)}\\
-Returns the total number of lattice sites in the given Cartesian direction.
-
-\texttt{double L(const int dim)}\\
-Returns the length of the system in the given Cartesian direction
-(which is the same as the number of lattice sites).
-
-\texttt{double Lmin(const int dim)}\\
-Returns coordinate position of left-hand edge of system in given
-Cartesian direction.
-
-\texttt{int is\_periodic(const int dim)}\\
-Returns 1 if given Cartesian direction is periodic, else 0.
-
-\texttt{void cart\_init(void)}\\
-Responsible for setting up the Cartesian decomposition.
-
-\texttt{int cart\_rank(void)}\\
-Returns rank in the Cartesian communicator.
-
-\texttt{MPI\_Comm cart\_comm(void)}\\
-Returns the MPI Communicator handle for the Cartesian communicator.
-
-\texttt{int cart\_size(const int dim)}\\
-Returns the size of the Cartesian communicator grid in the given
-coordinate direcxtion.
-
-\texttt{int cart\_coord(const int dim)}\\
-Returns the coordinate of the current process in the Cartesian
-Communicator.
-
-\texttt{int cart\_neigh(const int dir, const int dim)}\\
-Returns the rank of the neighbouring process in the
-Cartesian communicator.
-
-\subsection{Colloids}
-
-\texttt{\#include ``colloids.h''}
-
-These functions provide the basic interface for use
-of colloidal particles in the code.
-
-\subsubsection{Colloid structure}
-
-The basic colloid struct is defined ...
-
-\subsubsection{Colloid interface}
-
-\texttt{void colloids\_init(void)}\\
-Call once for initialisation. This initialises the cell list,
-so the coordinate system must be initialised via a call to
-\texttt{coords\_init()} first.
-
-\texttt{void colloids\_finish(void)}\\
-Call once for finalisation. Frees all colloid-related memory.
-
-\texttt{Colloid * allocate\_colloid(void)}\\
-Returns a pointer to a newly allocated \texttt{Colloid} object
-or fails gracefully.
-
-\texttt{void free\_colloid(Colloid * p\_colloid)}\\
-Deallocates all memory associated with \texttt{p\_colloid} or
-fails gracefully.
-
-\texttt{struct c\_link allocate\_boundary\_link(void)}\\
-Returns a pointer to a newly allocated colloid boundary
-link or fails gracefully.
-
-\texttt{void free\_boundary\_link(struct c\_link *)}\\
-Deallocates memory associated with a link or fails gracefully.
-
-\texttt{void colloids\_memory\_report(void)}\\
-The module keeps track of how much memory has been allocated in
-total for colloids and boundary links. This reports the current
-usage.
-
-\texttt{get\_Ncell\_total()}\\
-
-\texttt{get\_Ncell\_local()}\\
-
-\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
-This returns a pointer to the Colloid at the head of the list at
-cell list location (i, j, k) or \texttt{NULL} if there is none.
-
-\texttt{void cell\_insert(Colloid * p\_colloid)}\\
-This inserts a colloid at the head of the appropriate list
-depending upon its position. This is the only way a colloid
-can be added to the list.
-
-\texttt{update\_cells(void)}\\
-Performs the update of the cell list in response to updated
-particle positions.
-
-\texttt{IVector cell\_coords(FVector r)}\\
-
-\subsection{Speculation}
-
-\subsection{The Free Energy}
-
-It is imagined that a single file will encapsulate all the details
-of a given free energy. The interface for different choices of
-free energy should be the same.
-
-\texttt{FE\_init()}
-
-Responsible for querying the runtime input key value pair list to
-find appropriate user parameters for the free energy, or
-falling back to default values.
-
-\texttt{double FE\_get\_sigma()}
-
-Return the surface tension. There should be a similar routine
-for the interfacial thickness.
-
-\texttt{double FE\_get\_mu(const type location)}
-
-Return the chemical potential at the single lattice
-node \texttt{location}.
-
-\textit{Comment:} The computation of quantities such as the chemical
-potential will a) affect performance and b) require halo values of
-the order parameter (depending on the highest derivative
-in the order parameter present). Performance here should be the
-first consideration, so this needs to be handled with care.
-
-\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}
-
-Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
-given location.
-
-\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
-the number of dimensions. If we are to provide explicit support for
-D2Q9, \texttt{ND} will be 2, otherwise 3.
-
-\section{Implementation Notes}
-
-\subsection{Serial Version}
-
-\subsection{Efficiently Scaling Parallel Computing}
-
-We should be thinking about the 100,000 processor regime, and
-correspondingly large system sizes. To avoid embarrassment in this regime,
-some issues which don't usually arise are relevant, for example:
-
-\begin{enumerate}
-\item
-Large integer quantities such as
-\begin{verbatim}
-   N_total.x*N_total.y*N_total.z
-\end{verbatim}
-are likely to cause integer overflow (if default integers are 4 byte).
-\item
-Local data structures whose size is proportional
-to the number of processors are contraindicated, e.g.,
-anything like
-\begin{verbatim}
-p = (foo *) malloc(nprocs*sizeof(foo));
-\end{verbatim}
-is to be avoided. At the moment, we are in good shape on this
-(apart from some functions in \texttt{misc.c}).
-\item
-I/O is likely become a very important issue, so may need to think
-more about this.
-\item
-We should probably introduce a standard set of benchmarks which can
-be scaled up in some systematic way, so that we can keep track of
-performance over time.
-
-\end{enumerate}
-
-\section{Model Details}
-
-\subsection{D2Q9}
-
-These are the eigenvalues and eigenvectors of the collision
-matrix used for D2Q9
-
-\begin{table}
-\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
-\hline
-$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
-& $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
-& $c_{iy} c_{iy} - c_s^2$ \\
-\hline\hline
-$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - & 0 &  4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{iy}$\\
-\hline\hline
-$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q15}
-
-These are the eigenvalues and eigenvectors of teh collision
-matrix used for D3Q15
-
-\begin{table}
-\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - &
- 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
-1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
-3  & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - &
- 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
-3  &$c_{iy}$ \\
-\hline
-$\rho c_{iz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
-3  & $c_{iz}$ \\
-\hline
-$Q_{xx}$ & 1/3 &
--1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
-9/2  & $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{yy}$ & 1/3 &
--1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
-\hline
-$Q_{zz}$ & 1/3 &
--1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
-9  & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
-9  & $c_{iy} c_{iz}$ \\
-\hline
-$Q_{zx}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
-9  & $c_{iz} c_{ix}$ \\
-\hline\hline
-$\chi^1$ & - &
--2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
-1/2 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
-3/2 & $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - &
- 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iy}$\\
-\hline
-$J_{iz}$ & - &
- 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iz}$\\
-\hline
-$\chi^3$ & - &
- 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
-9 & $c_{ix} c_{iy} c_{iz}$ \\
-\hline\hline
-$w_i$ & 1/72 &
-$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
- & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q19}
-
-\begin{table}
-\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
-\hline\hline
-$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
-\hline
-$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
-         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
-& 1\\
-\hline
-$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
-         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
-& 3 \\
-\hline
-$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
-         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
-& 3\\
-\hline
-$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
-         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
-& 3\\
-\hline
-$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
-& 9/2\\
-\hline
-$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
-         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
-& 9\\
-\hline\hline
-$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
-         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 3/4\\
-\hline
-$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
-         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
-         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
-& 3/2\\
-\hline
-$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
-& 9/4\\
-\hline
-$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
-& 9/2\\
-\hline
-$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
-         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 1/2\\
-\hline\hline
-$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
-& \\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{adhikari_desplat}
-Adhikari, R. J.-C. Desplat, and K. Stratford,
-Sliding periodic boundary conditions for lattice Boltzmann and lattice
-kinetic equations,
-\texttt{arXiv:cond-mat/0503175v1} (2005).
-
-\bibitem{adhikari2005}
-Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
-Fluctuating Lattice Boltzmann,
-\textit{Europhys. Lett.}, \textbf{71}, 473 2005.
-
-\bibitem{ald98}
-Aidun, C.K., Y. Lu, and E.-J. Ding,
-Direct analysis of particulate suspensions with inertia using the
-discrete Boltzmann equation,
-\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
-
-\bibitem{cates_scaling}
-M. E. Cates, J.-C. Desplat, P. Stansell, A.J. Wagner, K. Stratford,
-R. Adhikari, and I. Pagonabarraga,
-Physical and Computational Scaling Issues in Lattice Boltzmann
-Simulations of Binary Fluid Mixtures,
-\textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
-
-\bibitem{chunladd}
-B. Chun and A.J.C. Ladd,
-Interpolated boundary condition for lattice Boltzmann simulations in
-narrow gaps,
-\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
-
-\bibitem{cj98}
-Cichocki, B., and R.B. Jones,
-Image representation of a spherical particle near a hard wall,
-\textit{Physica A}, \textbf{258}, 273, 1998.
-
-\bibitem{chang}
-C.-H. Chang and E.I. Franses,
-Adsorption dynamics of surfactants at the air/water interface:
-a critical review of mathematical models, data, and mechanisms,
-\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
-
-\bibitem{cpc}
-Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
-LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
-\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
-
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
-\bibitem{diamant96}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid-fluid interfaces,
-\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
-
-\bibitem{eastoe}
-J. Eastoe and J.S. Dalton,
-Dynamic surface tension and adsorption mechanims of surfactants
-at the air-water interface,
-\textit{Advances in Colloid and Interface Science}, \textbf{85}
-13, 2000.
-
-\bibitem{ginzburg}
-I. Ginzburg and D. d'Humi\`eres,
-Multireflection boundary conditions for lattice Boltzmann models,
-\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
-
-\bibitem{h59}
-Hasimoto, H., On the periodic fundamental solutions of the Stokes
-equation and their application to viscous flow past a cubic array
-of spheres.
-\textit{J. Fluid Mech.}, \textbf{5}, 317.
-
-\bibitem{heemels}
-Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
-particles using the lattice-Boltzmann method,
-\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.
-
-\bibitem{jo84}
-Jeffrey, D.J., and Y. Onishi,
-Calculation of the resistance and mobility functions for the two
-unequal rigid spheres in low-Reynolds-number flow,
-\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.
-
-\bibitem{viv}
-Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
-P. Bladon,
-Inertial effects in three dimensional spinodal decomposition of
-a symmetric binary fluid mixture: A lattice Boltzmann study,
-\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).
-
-\bibitem{l94a}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 1. Theoretical foundation,
-\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.
-
-\bibitem{l94b}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 2. Numerical results,
-\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.
-
-\bibitem{l96a}
-Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
-spheres,
-\textit{Phys. Fluids}, \textbf{9}, 491. 1996.
-
-\bibitem{l96b}
-Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
-of non-Brownian spheres,
-\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.
-
-\bibitem{lv01}
-Ladd, A.J.C., and R. Verberg,
-Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
-
-\bibitem{lipanmiller}
-H. Li, C. Pan, and C.T. Miller,
-Pore-scale investigation of viscous coupling effects for two-phase
-flow in porous media,
-\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
-
-\bibitem{nl02}
-Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
-lattice-Boltzmann simulations of particle suspensions,
-\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.
-
-\bibitem{papanastasiou}
-T. paapnastasiou, G. Georgiou, and A. Alexandrou,
-\textit{Viscous Fluid Flow},
-CRC Press, Boca Raton, Florida, 2000.
-
-\bibitem{r95}
-Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
-Cambridge University Press, 1995.
-
-\bibitem{succi}
-S. Succi, \textit{The lattice Boltzmann equation and beyond},
-Oxford University Press, Oxford, 2001.
-
-\bibitem{edo1}
-M. Venuroli and E.S. Boek,
-Two-dimensional lattice-Boltzmann simulations of single phase
-flow in a pseudo two-dimensional micromodel,
-\textit{Physica A}, \textbf{362}, 23, 2006.
-
-\bibitem{vandergraaf}
-R.G.M. van der Sman and S. van der Graaf,
-Diffuse interface model of surfactant adsorption onto flat and
-droplet interfaces,
-\textit{Rheol. Acta} \textbf{46} 3 (2006).
-
-\bibitem{theissengompper}
-O. Theissen and G. Gompper,
-Lattice Boltzmann study of spontaneous emulsification,
-\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
-
-\bibitem{wardtordai}
-A.F.H. Ward and L. Tordai,
-\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
-
-\end{thebibliography}
-
-\end{document}
+\inputkey{leslie\_ericksen\_swim}

### 2013-01-31 trunk/htdocs/surfactant.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1902 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \section{Surfactants}
+
+This section describes the approach taken to study surfactants. Here,
+we always adopt the hydrid method, i.e., LB for the hydrodynmic
+equations and finite difference for the associated Cahn-Hilliard
+equations. Coupling between the two sectors is via the velocity
+field and the external force on the fluid computed via the
+divergence of the `chemical stress'.
+
+In this model the surfactant appears as a concentration field
+$\psi(\mathbf{r};t)$, alongside the compositional order parameter
+$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
+two corresponding Cahn-Hilliard equations: first,
+
+\begin{equation}
+\partial_t \phi + \partial_\alpha \phi u_\alpha = 
+\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
+\end{equation}
+and second
+\begin{equation}
+\partial_t \psi + \partial_\alpha \psi u_\alpha = 
+\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
+\end{equation}
+
+\subsection{Free Energy}
+
+A number of different free energies have been used to study
+surfactants [cite Yeomans]. Here, the free energy is based on
+that used by van der Graff and van der Sman and others [cite].
+There are a number of different contributions to the free
+energy density:
+\begin{equation}
+f = f_\phi + f_\psi +f_1 + f_{ex}.
+\end{equation}
+The first is exactly that used for the binary fluid, i.e.,
+\begin{equation}
+f_\phi (\phi, \partial_\alpha\phi) =
+{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
++ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
+\end{equation}
+The surface tension and the interfacial width are determined from the
+choice of parameters $A,B,\kappa$ as before.
+The second term in the free energy density
+represents the entropic
+energy of mixing of the surfactant with the bulk phases
+\begin{equation}
+f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
+\end{equation}
+where the parameter $kT$ sets the energy scale for the whole system.
+The surfactant order parameter, $\psi$, is unity for a saturated
+interface and zero for a plain interface.
+
+The surfactant lowers the energy of the interface where it is adsorbed
+there, represented by the term
+\begin{equation}
+f_1(\psi, \partial_\alpha \phi) 
+= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
+  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
+\end{equation}
+where $\epsilon$ and $\beta$ are parameters. Note that only the
+first term appears in \cite{vandergraaf}, while the second term is
+based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
+the surfactant profile at the interface is symmetric (so-called
+Langmuir adsoption) whereas the term in $\beta$ induces an
+asymmetry (Frumpkin adsorption).
+Finally,  $f_{ex}$ is [van der Sman claims]
+an enthalpic contribution introduced 
+to stabilise the interface \cite{theissengompper}
+\begin{equation}
+f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
+\end{equation}
+with $W$ a parameter.
+
+\subsection{Chemical potentials}
+
+The respective chemical potentials are obtained from the functional
+derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
+$\mu_\psi = \delta F / \delta \psi$. We obtain:
+\begin{equation}
+\mu_\phi =
+A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
++ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
++ \epsilon \partial_\alpha \phi \partial_\alpha \psi
++ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
++ \beta \psi^2 \partial_\alpha^2 \phi,
+\end{equation}
+and
+\begin{equation}
+\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
++ {\textstyle \frac{1}{2}}W\phi^2
+- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
+- \beta\psi (\partial_\alpha \phi)^2.
+\end{equation}
+As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
++ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
+\cite{theissengompper})
+\begin{equation}
+p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
+\end{equation}
+There is no closed form for the tensor part, which must be chosen to
+comply with the constraint that in mechanical equilibrium
+\begin{equation}
+\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
+\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
+\end{equation}
+In this way we obtain for the isotropic contribution
+\begin{eqnarray}
+p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
+- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
+- \kappa \phi \partial_\alpha^2 \phi
+- kT \ln(1-\psi) + W\psi\phi^2
+\nonumber\\
++ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
++ \epsilon\phi\psi \partial_\alpha^2 \phi
++ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
++ \beta \phi \psi^2 \partial_\alpha^2 \phi
+- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
+\end{eqnarray}
+and for the tensor part we can have
+\begin{equation}
+\tilde{P}_{\alpha\beta} =
+(\kappa - \epsilon \psi - \beta\psi^2) 
+\partial_\alpha \phi \partial_\beta \phi.
+\end{equation}
 
@@
 
+\subsubsection{Parameters with $\xi_0 = 3.39$, $\epsilon = kappa/4$}
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+\hline
+Name & $\eta$ & $-A,B$ & $\kappa$ & $kT$ & $\epsilon$ & $\beta$ & $W$
+     & $M$ & $M_\psi$\\
+\hline
+R027a & 1/6 & 0.020833 & 0.12 & 5.6587$\times 10^{-4}$ & 0.06 & 0.0000 & 0.0 
+      & 0.15 & 2.0 \\
+\hline
+R028a & 1/6 & 0.020833 & 0.12 & $2.82743\times 10^{-4}$ & 0.06 & 0.0000 & 0.0 
+      & 0.15 & 2.0 \\
+\hline
+R029a & 1/6 & 0.020833 & 0.12 & 5.6587$\times 10^{-4}$ & 0.12 & 0.0000 & 0.0 
+      & 0.15 & 2.0 \\
+\hline
+R030a & 1/6 & 0.020833 & 0.12 & $2.82743\times 10^{-4}$ & 0.12 & 0.0000 & 0.0 
+      & 0.15 & 2.0 \\
+\hline
+\end{tabular}
+\caption{Parameters for $\xi_0 = 3.39$ and $\epsilon = \kappa/2$ and
+$\epsilon = \kappa$.}
+\end{center}
+\end{table}
+
 \begin{figure}
@@
 
+\subsection{Calibration of the Surface Tension}
+
+The binary fluid model introduces an interfacial tension $\sigma$ which
+can be predicted from the choice of free energy \cite{swift,kendon}.
+In particular, Kendon et al. 2000 described the process of calibrating
+the actual interfacial tension of the model fluid. This was based on
+a measurement of the Laplace pressure difference between the fluid
+inside and outside a spherical droplet via
+\begin{equation}
+\Delta p = 2 \sigma / R
+\end{equation}
+where $R$ is the radius of the spherical drop at equilibrium. This
+measurement appears quite difficult in practice owing to tendency of
+any initial drop to evaporate \cite{yue2007}.
+
+\subsubsection{New measurement}
+
+Here, we prefer a measure based on the expected profile of the
+order parameter $\phi$ and the free energy density at a flat interface.
+In order to sample all orientations of the interface with respect to
+the lattice we again take a large droplet of fluid A initialised at
+rest in fluid B. We assume that the interface is locally flat and so
+the procedure holds.
+
+If a droplet with radius $r_0$ is constructed with an initial interfacial
+profile of $\phi(r) = \tanh{(r - r_0) / \xi_0}$
+with $r$ in the radial direction,  then the free energy density will be
+(with $-A = B$)
+\begin{equation}
+ e(r) = -{\scriptstyle \frac{1}{4}} |A| [1 - \mathrm{sech}^4 (r-r_0)/\xi_0]
++ {\scriptstyle \frac{1}{2}}
+ (\kappa/\xi_0^2)\, \mathrm{sech}^4{(r-r_0)/\xi_0}.
+\end{equation}
+Obtaining a fit to $\xi_0$ from the measured interfacial profile of
+$\phi (r)$ and to $ e_0 = \kappa / 2 \xi_0^2$ from the peak in the measured
+profile of free energy
+density allows an estimate of the interfacial tension to be made:
+\begin{equation}
+\sigma = 2(8/9)^{1/2} \xi_0 |e_0|.
+\end{equation}
+This measurement is taken after the initial profile has been allowed
+to relax for at least 10 times the larger of the diffusion time
+$\xi_0^2 / M$ or the momentum diffusion time $\xi_0^2 / \eta$.
+
+We have measured the surface tension using the following approaches.
+
+\textit{Method 1}.
+
+LB for two distributions with relaxation of order parameter
+following Kendon etal. The force on the fluid is applied
+via the equilibrium stress following Swift et al.
+
+\textit{Method 2}.
+
+As method 1 but using the realxation of order parameter
+following Stratford etal.
+
+\textit{Method 3}.
+
+As Method 1 but with force via divergence of the chemical stress.
+
+\textit{Method 4}.
+
+As Method 3 but with force via divergence of the chemical stress.
+
+\textit{Method 5}.
+
+Using finite difference (1st order upwind) for the order parameter.
+
+\textit{Method 6}.
+
+Using finite difference (3rd order upwind).
+
+\subsubsection{Results}
+
+We follow the parameter sets in Kendon et al Table~2. These
+are reproduced here in Table~\ref{tab:r1}. Measured values for
+Run028 parameters are show in Table~\ref{tab:newr028}.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{llllllll}
+\hline
+Run & $-A, B$ & $\kappa$ & $\eta$ & $M$ & $\sigma$  & $L_0$ & $T_0$ \\
+\hline
+Rks001 & 0.00625 & 0.004 & 1.41 & 0.05 & 0.0047 & 422 & 1.26e10$^5$\\
+Rks002 & 0.00625 & 0.004 & 0.65 & 0.25 & 0.0047 & 89.7 & 1.24e10$^4$\\
+Run028 & 0.083 & 0.053 & 1.41 & 0.05  & 0.0625 & 31.8 & 718 \\
+Run022 & 0.0625 & 0.04 & 0.5 & 0.25   & 0.0471 & 5.31 & 56.3 \\
+Run029 & 0.0625 & 0.04 & 0.2 & 0.15   & 0.0471 & 0.849 & 3.61 \\
+Run020 & 0.00625 & 0.004 & 0.025 & 2.00 & 0.00471 & 0.133 & 0.704 \\
+Run030 & 0.00625 & 0.004 & 0.0065 & 1.25 & 0.00471 & 0.00829 & 0.0124 \\
+Run019 & 0.003125 & 0.002 & 0.0014 & 4.00 & 0.00236 & 0.000831 & 0.000493 \\
+Run032 & 0.00125  & 0.0008 & 0.0005 & 5.0 & 0.000943 & 0.000265 & 0.000141\\
+\hline  
+\end{tabular}
+\end{center}
+\caption{Table showing common parameter values following Kendon et al.
+with the theoretical interfacial tension, $L_0$ and $T_0$. Note that
+the order parameter mobility is used here (cf $\tilde{M} = 2M$) in
+Kendon et al. (KS values are extra sets used to low 1/$\dot{\gamma}T_0$.)}
+\label{tab:r1}
+\end{table}
+
+\begin{table}
+\begin{tabular}{lllllll}
+\hline
+Method & $\xi_0$ & $e_{min}$ & $e_{max}$ & $\sigma$  & $\phi_{min}$
+& $\phi_{max}$\\
+\hline
+\multicolumn{7}{c}{Run028}\\
+\hline
+Theory   & 1.13 & -0.0207 &  0.0207 & 0.0625 & -1.0 & 1.0 \\
+Method 1 & 1.00 & -0.0207 & 0.0128 & 0.063(3) & -1.0091 & 1.0028 \\
+Method 2 & 1.00 & -0.0207 & 0.0126 & 0.063(1) & -1.0113 & 0.9993 \\
+Method 3 & && & & & \\
+Method 4 & && & & & \\
+Method 5 & && & & & \\
+Method 6 & --& --& --& --& --& --\\
+\hline
+\multicolumn{7}{c}{Run029}\\
+\hline
+Theory & 1.13 & -0.0156 & 0.0156 & 0.0471  & -1.0 & 1.0\\
+Method 2 & 0.99 & -0.0156 & 0.010(?) & 0.048(4) & -1.0110 & 1.0003\\ 
+Method 6 & 1.02 & -0.0156 & 0.0099 & 0.049(0) & -1.0048 & 1.0002\\
+\hline
+\multicolumn{7}{c}{Run030}\\
+\hline
+Theory & 1.13 & -0.00156 & 0.00156 & 0.00471 & -1.0 & 1.0\\
+Method 2 & 1.04 & -0.00156 & 0.0008(?) & 0.0047(1) & -1.0167 & 1.0063\\
+Method 6 & 1.08 & -0.00156 & 0.0007(?) & 0.0048(0) & -1.0028 & 1.0031\\
+\hline
+\multicolumn{7}{c}{Run019}\\
+\hline
+Theory & 1.13 & -0.000781 & 0.000781 & 0.00236 & -1.0 & 1.0\\
+Method 2 & 1.29(?) & -0.000781 & 0.00015(?) & 0.0022(8) & -1.0312 & 1.0242\\
+Method 6 & 1.23(?) & 0.0010(?) &&  0.0024(?) & -1.0085 & 1.0019\\
+\hline 
+\multicolumn{7}{c}{Run032}\\
+\hline
+Theory   & 1.13 &  -0.000312 & 0.000312 & 0.000943  & -1.0 &1.0\\
+Method 2 & 1.07  & -0.000312& 0.000157 & 0.00094(7) & -1.0511 & 1.0311\\
+Method 6 & 1.6(5)& -0.000312 & not st. & 0.0010(7) &-1.0056 & 1.0010\\
+\hline
+\end{tabular}
+\caption{Results for the different methods for Run208 parameters.}
+\end{table}
+
+\subsection{User input}
+
+\inputkey{free\_energy surfactant}
+
+Currently under reconstruction.
+
+This is for use with a two order parameter model the first of
+which is the composition, as for the symmetric free energy,
+while the second $\psi$ represents surfactant concentration.
+The free energy density is made up of a number of terms:
+
+\begin{equation}
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+\end{equation}
+is the standard symmetric term related to the composition.
+These parameters are represented in the input as
+
+\inputkey{surf\_A}
+\inputkey{surf\_B}
+\inputkey{surf\_kappa}
+
+\begin{equation}
+D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
+\end{equation}
+represents the energy of surfactant in the bulk phase with surfactant
+concentration $0 < \psi < 1$ with a single parameter $D$;
+\begin{equation}
+-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
+-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
+\end{equation}
+where the term in $\epsilon$ represents the energy reduction by
+adsorbing surfactant at the interface and the term in $\beta$
+provides an additional favourable term modelling the fact that
+the molecules like to line up; finally, there is an extra term
+\begin{equation}
++{\textstyle \frac{1}{2}} W \psi \phi^2
+\end{equation}
+which is included to stabilise the interface \cite{theissengompper}.
+
+The corresponding keys for the input file are:
+
+\inputkey{D} is the bulk surfactant contribution parameter
+
+\inputkey{epsilon} is the parameter $\epsilon$
+
+\inputkey{W} is the parameter $W$
+
+\inputkey{beta} is the value of the parameter $\beta$.
+
+\inputkey{phi\_b} the initial, uniform, background surfactant
+concentration $\psi_b$.
+
+\inputkey{mobility\_psi} Sets the (uniform) mobility for the surfactant
+as appropriate for the Cahn-Hilliard equation.
+
 \vfill

### 2013-01-31 htdocs/code.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1902 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-02-11 htdocs/sections.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  sections.tex
-%
-%  Content for Ludwig documentation.
-%
-%  $Id$
-%
-%  Edinburgh Soft Matter and Statistical Physics Group and
-%  Edinburgh Parallel Computing Centre
-%
-%  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{The Lattice Boltzmann Method}
-
-We review here the lattice Boltzmann method applied to single
-phase problems.
-
-\subsection{The Navier Stokes Equation}
-
-We seek to solve the isothermal Navier-Stokes equations which, often
-written in vector form, express mass conservation
-\begin{equation}
-\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
-\label{eq_mass1}
-\end{equation}
-and the conservation of momentum
-\begin{equation}
-\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
--\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
-+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
-\label{eq_momentum1}
-\end{equation}
-Equation~\ref{eq_mass1} expresses the local rate of change of the
-density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
-mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
-Equation~\ref{eq_momentum1} expresses Newton's second law for
-momentum, where the terms on the right hand side represent the
-force on the fluid.
-
-For this work, it is more convenient to rewrite these equations
-in tensor notation, where Cartesian coordinates ${x,y,z}$ are
-represented by indices $\alpha$ and $\beta$, viz
-\begin{equation}
-\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
-\end{equation}
-and
-\begin{equation}
-\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
-= -\nabla_\alpha p
-+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
-+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
-\end{equation}
-Here, repeated indices are understood to be summed over.
-The conservation law is seem better if the forcing terms of the
-right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
-so that
-\begin{equation}
-\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
-\end{equation}
-In this case the expanded expression for the stress tensor is
-\begin{equation}
-\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
-+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
-\end{equation}
-where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
-$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
-modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.
-
-\subsection{The Distribution Function and its Moments}
-
-In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
-\begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
-\end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
-
-In general, the number of modes supported in $n$ and these can
-be written as
-\begin{equation}
-M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
-\end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
-\begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
-\end{equation}
-
-\subsubsection{Hydrodynamic modes}
-
-\subsubsection{D3Q15 ghost modes}
-
-\subsubsection{D3Q19 ghost modes}
-
-\subsection{The Lattice Boltzmann Equation}
-
-\subsection{The Collision}
-
-\subsection{External Body Forces}
-
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsubsection{Fluctuating ghost modes: D3Q15}
-
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Boundary walls}
-
-\subsection{Colloids}
-
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
-\section{Polar Active Gel}
-
-Here we consider a polar vector field model to study active gels. 
-The order parameter is this time a vector $P_\alpha$
-(with variable magnitude) as there is no longer head-tail symmetry
-in the system as was the case for the ${\bf Q}$ tensor formulation. 
-The equation of motion as:
-\begin{equation}
-\partial_t P_\alpha  \partial_\beta (u_{\beta} + wP_{\beta}) P_{\alpha}
-= \lambda D_{\alpha\beta} P_{\beta} - \omega_{\alpha\beta} P_{\beta}
-+\Gamma' h_{\alpha}.
-\end{equation}
-
-In this equation, $w$ is an active term, due to swimming, which causes
-self-advection of the order parameter, while $\lambda$ is a material
-dependent constant -- positive for rod-like molecules. If $|\lambda|>1$
-the liquid crystalline passive phase is flow-aligning, otherwise it is
-flow-tumbling.
-
-Note that the convection for the Leslie-Ericksen is that the
-velocity gradient tensor should be writen $W_{\alpha\beta}
-= \partial_\alpha u_\beta$ (and not the other way round, which
-is usual).
-
-The molecular field is given by
-$h_{\alpha}=-\delta {\cal F}_{\rm pol}/\delta p_{\alpha}$
-where ${\cal F}_{\rm pol}$ is the free energy for a polar active nematic,
-whose density we can take to be:
-\begin{equation}
-f =  \frac{\alpha_2}{2} |{\mathbf P}|^2 + \frac{\alpha_4}{4}|{\mathbf P}|^4
-+ \frac{K}{2}\left(\partial_\alpha P_{\beta}\right)^2 + 
-\frac{K_{LC}}{2}\left[\partial_\alpha (P_{\beta}P_{\gamma})\right]^2 
-\end{equation}
-where $K$ and $K_{LC}$ are elastic constants, $\alpha_4>0$, and $\alpha_2$ 
-can be positive (isotropic phase) or negative (polar nematic phase).
-
-The molecular field is explicitly given by:
-\begin{equation}
-h_{\alpha}=-\alpha_2 P_{\alpha} -\alpha_4 P_{\alpha} P_{\beta}P_{\beta}+
-K \partial_{\beta}\partial_{\beta} P_{\alpha}+2 K_{LC}
-P_{\gamma}\partial_{\beta}\partial_{\beta} (P_{\gamma}P_{\alpha}).
-\end{equation}
-
-The Navier-Stokes equation is:
-\begin{equation}
-\rho \left(\partial_t+v_{\beta}\partial_{\beta}\right) v_{\alpha} =
-\partial_{\beta} \Pi_{\alpha\beta} 
-+\eta \partial_{\beta}\partial_{\beta} v_{\alpha}.
-\end{equation}
-
-The stress tensor this time is 
-\begin{eqnarray}
-\Pi_{\alpha\beta} =
-\frac{1}{2}\left(P_\alpha h_\beta - P_\beta h_\alpha \right)
-- \lambda\left( (P_\alpha h_\beta + P_\beta h_\alpha)/2 
-- (1/3)P_\gamma P_\gamma \delta_{\alpha\beta} \right)
-\\
--  \zeta ( P_\alpha P_\beta - (1/3)P_\gamma P_\gamma \delta_{\alpha\beta})
-- K \partial_\alpha P_\gamma \partial_\beta P_\gamma.
-\end{eqnarray}
-The active term in the stress tensor, proportional
-to $\zeta$, has the same form for apolar and polar active gels.
-
 \section{Code Design}
@@
 \end{figure}
+
+\clearpage
+\vfill\pagebreak
 
@@
 
-\vfill
-\pagebreak
+\clearpage
+\vfill\pagebreak
 
@@
 D2Q9, \texttt{ND} will be 2, otherwise 3.
+
+\clearpage
+\vfill\pagebreak
 
@@
 \end{enumerate}
-
-\section{Model Details}
-
-\subsection{D2Q9}
-
-These are the eigenvalues and eigenvectors of the collision
-matrix used for D2Q9
-
-\begin{table}
-\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
-\hline
-$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
-& $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
-& $c_{iy} c_{iy} - c_s^2$ \\
-\hline\hline
-$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - & 0 &  4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{iy}$\\
-\hline\hline
-$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q15}
-
-These are the eigenvalues and eigenvectors of teh collision
-matrix used for D3Q15
-
-\begin{table}
-\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - &
- 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
-1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
-3  & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - &
- 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
-3  &$c_{iy}$ \\
-\hline
-$\rho c_{iz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
-3  & $c_{iz}$ \\
-\hline
-$Q_{xx}$ & 1/3 &
--1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
-9/2  & $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{yy}$ & 1/3 &
--1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
-\hline
-$Q_{zz}$ & 1/3 &
--1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
-9  & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
-9  & $c_{iy} c_{iz}$ \\
-\hline
-$Q_{zx}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
-9  & $c_{iz} c_{ix}$ \\
-\hline\hline
-$\chi^1$ & - &
--2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
-1/2 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
-3/2 & $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - &
- 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iy}$\\
-\hline
-$J_{iz}$ & - &
- 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iz}$\\
-\hline
-$\chi^3$ & - &
- 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
-9 & $c_{ix} c_{iy} c_{iz}$ \\
-\hline\hline
-$w_i$ & 1/72 &
-$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
- & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q19}
-
-\begin{table}
-\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
-\hline\hline
-$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
-\hline
-$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
-         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
-& 1\\
-\hline
-$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
-         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
-& 3 \\
-\hline
-$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
-         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
-& 3\\
-\hline
-$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
-         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
-& 3\\
-\hline
-$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
-& 9/2\\
-\hline
-$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
-         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
-& 9\\
-\hline\hline
-$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
-         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 3/4\\
-\hline
-$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
-         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
-         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
-& 3/2\\
-\hline
-$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
-& 9/4\\
-\hline
-$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
-& 9/2\\
-\hline
-$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
-         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 1/2\\
-\hline\hline
-$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
-& \\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{adhikari_desplat}
-Adhikari, R. J.-C. Desplat, and K. Stratford,
-Sliding periodic boundary conditions for lattice Boltzmann and lattice
-kinetic equations,
-\texttt{arXiv:cond-mat/0503175v1} (2005).
-
-\bibitem{adhikari2005}
-Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
-Fluctuating Lattice Boltzmann,
-\textit{Europhys. Lett.}, \textbf{71}, 473 2005.
-
-\bibitem{ald98}
-Aidun, C.K., Y. Lu, and E.-J. Ding,
-Direct analysis of particulate suspensions with inertia using the
-discrete Boltzmann equation,
-\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
-
-\bibitem{cates_scaling}
-M. E. Cates, J.-C. Desplat, P. Stansell, A.J. Wagner, K. Stratford,
-R. Adhikari, and I. Pagonabarraga,
-Physical and Computational Scaling Issues in Lattice Boltzmann
-Simulations of Binary Fluid Mixtures,
-\textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
-
-\bibitem{chunladd}
-B. Chun and A.J.C. Ladd,
-Interpolated boundary condition for lattice Boltzmann simulations in
-narrow gaps,
-\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
-
-\bibitem{cj98}
-Cichocki, B., and R.B. Jones,
-Image representation of a spherical particle near a hard wall,
-\textit{Physica A}, \textbf{258}, 273, 1998.
-
-\bibitem{chang}
-C.-H. Chang and E.I. Franses,
-Adsorption dynamics of surfactants at the air/water interface:
-a critical review of mathematical models, data, and mechanisms,
-\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
-
-\bibitem{cpc}
-Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
-LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
-\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
-
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
-\bibitem{diamant96}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid-fluid interfaces,
-\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
-
-\bibitem{eastoe}
-J. Eastoe and J.S. Dalton,
-Dynamic surface tension and adsorption mechanims of surfactants
-at the air-water interface,
-\textit{Advances in Colloid and Interface Science}, \textbf{85}
-13, 2000.
-
-\bibitem{ginzburg}
-I. Ginzburg and D. d'Humi\`eres,
-Multireflection boundary conditions for lattice Boltzmann models,
-\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
-
-\bibitem{h59}
-Hasimoto, H., On the periodic fundamental solutions of the Stokes
-equation and their application to viscous flow past a cubic array
-of spheres.
-\textit{J. Fluid Mech.}, \textbf{5}, 317.
-
-\bibitem{heemels}
-Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
-particles using the lattice-Boltzmann method,
-\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.
-
-\bibitem{jo84}
-Jeffrey, D.J., and Y. Onishi,
-Calculation of the resistance and mobility functions for the two
-unequal rigid spheres in low-Reynolds-number flow,
-\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.
-
-\bibitem{viv}
-Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
-P. Bladon,
-Inertial effects in three dimensional spinodal decomposition of
-a symmetric binary fluid mixture: A lattice Boltzmann study,
-\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).
-
-\bibitem{l94a}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 1. Theoretical foundation,
-\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.
-
-\bibitem{l94b}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 2. Numerical results,
-\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.
-
-\bibitem{l96a}
-Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
-spheres,
-\textit{Phys. Fluids}, \textbf{9}, 491. 1996.
-
-\bibitem{l96b}
-Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
-of non-Brownian spheres,
-\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.
-
-\bibitem{lv01}
-Ladd, A.J.C., and R. Verberg,
-Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
-
-\bibitem{lipanmiller}
-H. Li, C. Pan, and C.T. Miller,
-Pore-scale investigation of viscous coupling effects for two-phase
-flow in porous media,
-\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
-
-\bibitem{nl02}
-Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
-lattice-Boltzmann simulations of particle suspensions,
-\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.
-
-\bibitem{papanastasiou}
-T. paapnastasiou, G. Georgiou, and A. Alexandrou,
-\textit{Viscous Fluid Flow},
-CRC Press, Boca Raton, Florida, 2000.
-
-\bibitem{r95}
-Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
-Cambridge University Press, 1995.
-
-\bibitem{succi}
-S. Succi, \textit{The lattice Boltzmann equation and beyond},
-Oxford University Press, Oxford, 2001.
-
-\bibitem{edo1}
-M. Venuroli and E.S. Boek,
-Two-dimensional lattice-Boltzmann simulations of single phase
-flow in a pseudo two-dimensional micromodel,
-\textit{Physica A}, \textbf{362}, 23, 2006.
-
-\bibitem{vandergraaf}
-R.G.M. van der Sman and S. van der Graaf,
-Diffuse interface model of surfactant adsorption onto flat and
-droplet interfaces,
-\textit{Rheol. Acta} \textbf{46} 3 (2006).
-
-\bibitem{theissengompper}
-O. Theissen and G. Gompper,
-Lattice Boltzmann study of spontaneous emulsification,
-\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
-
-\bibitem{wardtordai}
-A.F.H. Ward and L. Tordai,
-\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
-
-\end{thebibliography}
-
-\end{document}

### 2013-01-31 htdocs/electrokinetic.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1902 40876441-3496-0410-a613-a2fca35b2ae2

--- 2012-08-02 papers/dsfdproc/dsfdproc.tex
@@
-\documentclass[aps,pre,preprint,groupedaddress]{revtex4-1}
-\bibliographystyle{apsrev4-1}
+\section{Electrokinetics}
 
-\usepackage[pdftex]{graphicx}
-\usepackage{amsmath}
-\newcommand{\e}[1]{\cdot10^{#1}}
-\let\endtitlepage\relax
+\subsection{Sress tensor in the electrosymmetic model}
 
-\begin{document}
+\subsubsection{Definitions}
 
-\title{Report HPC-Europa2 Transnational Access Programme}
+The total free energy functional is a function of the composition $\phi$, the density of the ionic species $\rho_\alpha$,
 
-\author{Oliver Henrich$^1$ and Kevin Stratford$^2$}
+\beq
+{\cal F}[\phi,\{\rho_\alpha\}] = \int d{\bf r} \;F\left(\phi({\bf r}),\{\rho_\alpha({\bf r})\}\right). 
+\eeq
 
-\affiliation{
-$^1$Centre for Computational Science, University College London, United Kingdom\\
-$^2$Edinburgh Parallel Computing Centre, University of Edinburgh, United Kingdom
-}
+The free energy density can be factorised into a contribution from the composition and an ionic contribution according to
 
-\date{\today}
-\begin{abstract}
-This is the intermim report of our joint visit from 12 April - 11 May 2012 to Barcelona Supercomputing Center and Prof Ignacio Pagonabarraga, Departament de Fisica Fonamental, Universitat de Barcelona, Spain.
-\end{abstract}
+\beqa
+F&=&F^{mix} +F^{ion}\\
+F^{mix}&=& \frac{1}{2} \,A \,\phi^2 + \frac{1}{4} \,B \,\phi^4+ \frac{1}{2} \kappa(\bm{\nabla}\phi)^2\\
+F^{ion,ex}&=& \sum_{\alpha=\pm} \rho_\alpha \left(V^{solv}_\alpha-\mu_\alpha +\frac{z_\alpha e}{2}\Psi \right).
+\eeqa
 
-\maketitle
+In the ionic contribution we have omitted the ideal gas contribution of the ions.
+The solvation energy is given by
 
-\section{Week 1 (including period 12-20 April)}
+\beqa
+V^{solv}_\alpha({\bf r})=\Delta\mu_\alpha\frac{1+\phi({\bf r})}{2}
+\eeqa
 
-We have introduced an object to hold the scalar electric potential,
-and the charge densities for a general number of species. This
-includes specification of diffusivities, valencies and unit charge.
-The corresponding parallel halo swaps and I/O routines have been
-added and subjected to unit tests.
+It is convenient to define the total free energy without electric field $F_0$, i.e. with vanishing electric potential:
 
-In discussions with Pagonabarraga, we decided to prioritise the
-implemenation of the the Poisson solver (successive overrelaxation,
-SOR) so that more simple validation problems can be attempted as
-soon as possible. The SOR solver has been implemented and tested
-against an exact matrix solver for a one-dimensional problem in
-both serial and parallel. The SOR solver assumes a spatially
-uniform permittivity. We will extend this to a spatially varying
-permittivity at a later stage.
+\beqa
+F_0&=&F^{mix} +F^{ion,ex}|_{\Psi=0}
+\eeqa
 
-A start has been made on the finite difference solution of the
-Nernst Planck equation (advection/diffusion equation for charged
-species). This will be completed and tested next week. This puts
-us slightly ahead of schedule cf the original work plan.
+The Poisson equation reads
+\beqa
+\bm{\nabla}\left(\varepsilon({\bf r})\bm{\nabla}\Psi({\bf r})\right)=-\sum_{\alpha=\pm} e \,z_\alpha \,\rho_\alpha ({\bf r})
+\eeqa
 
-Basic tests of the SOR solver have been already performed.
-We have identified a number of other suitable test scenarios to 
+whereas the electric field is given by
+
+\beq
+{\bm E}=-{\bm \nabla}\Psi.
+\eeq
+
+The permittivity depends on the composition $\phi$ via
+
+\beqa
+\varepsilon({\bf r})=\bar{\varepsilon} \,(1-\gamma\,\phi({\bf r}))
+\eeqa
+
+The electric displacement field
+
+\beq
+{\bm D}({\bf r})=\varepsilon({\bf r}){\bm E}({\bf r})
+\eeq
+
+The 'co-energy' density
+
+\beq
+\tilde{F}=F_0-\frac{\varepsilon}{2}\bm{E}^2
+\eeq
+
+has the meaning of a Legendre transform of the free energy density $F_0$.
+
+\subsubsection{Electromechanical stress tensor}
+
+The electromechanical stress tensor is generally defined as \cite{Landau-ED, Melcher}.
+
+\beqa
+\sigma_{i j}=\varepsilon E_i E_j + \delta_{i j}\left( \tilde{F} - \phi \frac{\delta \tilde{\cal F}}{\delta\phi} - \sum_{\alpha=\pm} \rho_\alpha \frac{\delta \tilde{\cal F}}{\delta\rho_\alpha}\right).
+\eeqa
+
+The first functional derivative yields
+
+\beqa
+\phi \frac{\delta\tilde{\cal F}}{\delta\phi}&=&\phi\left(\frac{\delta {\cal F}_0}{\delta\phi}-\frac{{\bm E}^2}{2} \frac{\partial \varepsilon}{\partial\phi}\right)\\
+&=&\phi\left(\frac{\partial F_0}{\partial\phi}-{\bm \nabla}\left(\frac{\partial F_0}{\partial {\bm \nabla}\phi}\right)-\frac{{\bm E}^2}{2} \frac{\partial \varepsilon}{\partial\phi}\right)\\
+&=&A\,\phi^2+B\,\phi^4-\kappa\phi({\bm \nabla}^2\phi)+\frac{\phi}{2}\sum_{\alpha=\pm} \rho_\alpha \Delta\mu_\alpha+\phi\,\frac{\bar{\varepsilon}\gamma}{2}{\bm E}^2\\
+&=&A\,\phi^2+B\,\phi^4-\kappa\phi({\bm \nabla}^2\phi)+\frac{\phi}{2}\sum_{\alpha=\pm} \rho_\alpha \Delta\mu_\alpha-\frac{\varepsilon-\bar{\varepsilon}}{2}{\bm E}^2,
+\eeqa
+
+where we used the above dependence of the permittivity on the composition.
+The second functional derivative is given by
+
+\beqa
+\rho_\alpha \frac{\delta \tilde{\cal F}}{\delta\rho_\alpha}&=&\rho_\alpha\left(\frac{\delta {\cal F}_0}{\delta\rho_\alpha}\right)\\
+&=&\rho_\alpha(V^{solv}_\alpha-\mu_\alpha)
+\eeqa
+
+The co-energy density and the functional derivatives with respect to composition $\phi$ and ionic densities $\rho_\alpha$ add up to
+
+\beqa
+\tilde{F} - \phi \frac{\delta \tilde{F}}{\delta\phi} - \sum_{\alpha=\pm} \rho_\alpha \frac{\delta \tilde{F}}{\delta\rho_\alpha}=&&\nonumber\\
+&&\hspace*{-4cm}=-\frac{\varepsilon}{2}\bm{E}^2+\frac{A}{2} \phi^2 + \frac{B}{4} \phi^4+ \frac{\kappa}{2} (\bm{\nabla}\phi)^2 \nonumber\\
+&&\hspace*{-3cm}-A\phi^2-B\phi^4+ \kappa\phi({\bm \nabla}^2\phi) -\frac{\phi}{2}\sum_{\alpha=\pm} \rho_\alpha \Delta\mu_\alpha\nonumber\\
+&&\hspace*{-3cm}-\phi\,\frac{\bar{\varepsilon}\gamma}{2}{\bm E}^2-\sum_{\alpha=\pm}\rho_\alpha(V^{solv}_\alpha-\mu_\alpha)\\
+&&\hspace*{-4cm}=-\frac{\varepsilon}{2}\bm{E}^2-\frac{A}{2}\phi^2 -\frac{3 B}{4} \phi^4+ \frac{\kappa}{2}(\bm{\nabla}\phi)^2 + \kappa\phi({\bm \nabla}^2\phi) \nonumber\\
+&&\hspace*{-3cm}-\frac{\phi}{2}\sum_{\alpha=\pm} \rho_\alpha \Delta\mu_\alpha -\phi\,\frac{\bar{\varepsilon}\gamma}{2}{\bm E}^2
+\eeqa
+
+This yields
+
+\beqa
+\sigma_{ij}&=&\varepsilon E_i E_j - \delta_{i j}\left(\frac{\varepsilon}{2}\bm{E}^2+\frac{A}{2}\phi^2 +\frac{3\,B}{4}\phi^4 - \frac{\kappa}{2} (\bm{\nabla}\phi)^2\right.\nonumber\\
+&&\left.- \kappa\phi({\bm \nabla}^2\phi) +\frac{\phi}{2}\sum_{\alpha=\pm} \rho_\alpha \Delta\mu_\alpha +\phi\,\frac{\bar{\varepsilon}\gamma}{2}{\bm E}^2\right).\nonumber\\
+\eeqa
+
+In order to satisfy the condition of mechanical equilibrium $\nabla_j \sigma_{ij}=0$ a symmetric contribution has to be added to the above stress tensor. The following lines try to elucidate this.\\
+
+Taking the gradient of the stress tensor gives
+
+\beqa
+\nabla_j\sigma_{i j}&=& (\nabla_j \varepsilon E_j) E_i + \varepsilon E_j \nabla_j E_i + \nabla_i\left( \tilde{F} - \phi \frac{\delta \tilde{\cal F}}{\delta\phi} - \sum_{\alpha=\pm} \rho_\alpha \frac{\delta \tilde{\cal F}}{\delta\rho_\alpha}\right)\\
+&=& \left({\bm \nabla}\cdot{\bm D}\right) E_i+ D_j \nabla_j E_i +\nabla_i F_0 - D_j \nabla_i E_j \nonumber\\
+&&\hspace*{0.5cm} -\nabla_i\left(\phi \frac{\delta \tilde{\cal F}}{\delta\phi} +\sum_{\alpha=\pm} \rho_\alpha \frac{\delta \tilde{\cal F}}{\delta\rho_\alpha}\right)\\
+&=& \left({\bm \nabla}\cdot{\bm D}\right) E_i + D_j \left(\nabla_j E_i-\nabla_i E_j\right) + \nabla_i F_0\nonumber\\
+&&\hspace*{0.5cm} -\nabla_i\left(\phi \mu_\phi +\sum_{\alpha=\pm} \rho_\alpha \mu_{\rho_\alpha}\right)
+\eeqa
+
+According to Gauss' law the first term is
+
+\beq
+\left({\bm \nabla}\cdot{\bm D}\right)E_i = \rho_f E_i=\sum_{\alpha=\pm} z_\alpha e \rho_\alpha E_i,
+\eeq
+
+whereas $\rho_f$ is the free charge density.
+Moreover, due to Faraday's law the curl in the second bracket vanishes.
+This leads us to the following intermediate result:
+
+\beqa
+\nabla_j\sigma_{i j}&=& \sum_{\alpha=\pm}z_\alpha e \rho_\alpha E_i + \nabla_i F_0 - \nabla_i\left(\phi \mu_\phi +\sum_{\alpha=\pm} \rho_\alpha \mu_{\rho_\alpha}\right)\\
+&=&\sum_{\alpha=\pm} z_\alpha e \rho_\alpha E_i + \frac{\partial F_0}{\partial \phi} \nabla_i \phi + \frac{\partial F_0}{\partial \nabla_j \phi} \nabla_j \nabla_i \phi +\sum_{\alpha=\pm} \left\{\frac{\partial F_0}{\partial \rho_\alpha}\right\} \nabla_i \rho_\alpha\nonumber\\
+&&\hspace*{0.5cm} -\nabla_i\left(\phi \mu_\phi +\sum_{\alpha=\pm} \rho_\alpha \mu_{\rho_\alpha}\right)\\
+&=& \sum_{\alpha=\pm} z_\alpha e \rho_\alpha E_i + \left\{\frac{\partial F_0}{\partial \phi} - \nabla_j \left(\frac{\partial F_0}{\partial \nabla_j \phi}\right)\right\} \nabla_i \phi +\nabla_j\left(\frac{\partial F_0}{\partial \nabla_j \phi} \nabla_i \phi\right)\nonumber\\
+&&\hspace*{0.5cm} +\sum_{\alpha=\pm} \left\{\frac{\partial F_0}{\partial \rho_\alpha} \right\}\nabla_i \rho_\alpha-\nabla_i\left(\phi \mu_\phi +\sum_{\alpha=\pm} \rho_\alpha \mu_{\rho_\alpha}\right).
+\eeqa
+
+If we replace the terms in curly brackets the chemical potentials $\mu_\phi$ and $\mu_{\rho_\alpha}$ and use the fact that gradients of the chemical potential vanish in equilibrium, we end up with
+
+\beqa
+\nabla_j\sigma_{i j}&=& \sum_{\alpha=\pm}z_\alpha e \rho_\alpha E_i + \nabla_j\left(\frac{\partial F_0}{\partial \nabla_j \phi} \nabla_i \phi\right)
+\eeqa
+
+The first term is the external force on the charges, which is zero if we have no counterions and an equal amount of positve and negative ions. Therefore, we have to add a term
+
+\beq
+-\left(\frac{\partial F_0}{\partial \nabla_j \phi} \nabla_i \phi\right)
+\eeq
+
+to the stress tensor to fulfill the equilibrium condition. This is equivalent to similar derivations in \cite{Landau-EL}.\\
+
+The total stress tensor reads
+
+\beqa
+\sigma_{ij}&=&\varepsilon E_i E_j - \kappa (\nabla_i\phi)(\nabla_j\phi)  - \delta_{i j}\left(\frac{\varepsilon}{2}\bm{E}^2+\frac{A}{2}\phi^2 +\frac{3\,B}{4}\phi^4\right.\nonumber\\
+&&\left. - \frac{\kappa}{2} (\bm{\nabla}\phi)^2- \kappa\phi({\bm \nabla}^2\phi) +\frac{\phi}{2}\sum_{\alpha=\pm} \rho_\alpha \Delta\mu_\alpha +\phi\,\frac{\bar{\varepsilon}\gamma}{2}{\bm E}^2\right).
+\eeqa
+
+\subsection{Examples}
+
+A number of regression tests 
 verify the implemention of the Nernst-Planck equation in 
-combination with the SOR solver.
-We decided to test the Gouy-Chapman theory
+combination with the SOR solver. They can be found in
+
+\begin{verbatim}
+trunk/tests/regression
+\end{verbatim}
+
+The original tests we carried out during the first
+implementation comprise the Gouy-Chapman theory
 for electric double layers in front of a charged wall \cite{Lyklema},
@@
 of slighty different concentration and diffusivity of the charged species \cite{Mafe},
-electro-osmotic flow in a slit pore  
-\cite{Capuani, Rotenberg} and Debye-H\"uckel theory for charged 
+electro-osmotic flow in a slit pore  \cite{Capuani, Rotenberg} 
+and Debye-H\"uckel theory for charged 
 colloidal particles and small enough potentials \cite{Lyklema}.
+They are described in the following paragraphs.
 
-\section{Week 2 (23-27 April)}
+\subsubsection{Gouy-Chapman}
 
-We completed the numerical implementation of the Nernst-Planck
-Equation which is solved via a simple finite difference method.
-The solution includes advection (although there is no fluid
-fluid fluid present in the test problems). We have included
-no normal flux boundary conditions at solid fluid interfaces
-to ensure the conservation of charged species. The solver
-has been tested in both serial and in parallel.
-
-A first validation test is a flat surface carrying a surface charge $\sigma$
+This validation test is a flat surface carrying a surface charge $\sigma$
 with counterions and symmetic electrolyte. This is a one-dimensional, 
@@
 potentials reads
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{./pics/test1.pdf}
+\includegraphics[width=0.495\textwidth]{./pics/test2.pdf}
+\caption{Gouy-Chapman theory for electric double layers: The plots compare the simulation results for the electric potential with the analytical solution for $\rho_{0,\pm}=1\e{-2}$ (left) and $\rho_{0,\pm}=1\e{-3}$ (right).}
+\label{fig1} 
+\end{figure}
 
@@
 
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{./pics/test3.pdf}
+\includegraphics[width=0.495\textwidth]{./pics/test3-rho.pdf}
+\caption{Nonlinear regime: For larger surface charges the solution deviates from the low-potential approximation Eq.
+\ref{gouychapman} to the Gouy-Chapman theory.
+The righthand picture shows the charge distribution across the gap.}
+\label{fig2} 
+\end{figure}
+
 The parameters were chosen as follows (all in simulation units):
@@
 
-\begin{figure}[h!t]
-\includegraphics[width=0.495\textwidth]{test1.pdf}
-\includegraphics[width=0.495\textwidth]{test2.pdf}
-\caption{Gouy-Chapman theory for electric double layers: The plots compare the simulation results for the electric potential with the analytical solution for $\rho_{0,\pm}=1\e{-2}$ (left) and $\rho_{0,\pm}=1\e{-3}$ (right).}
-\label{fig1} 
-\end{figure}
-
 For larger surface charges the low-potential assumption which led to Eq. \ref{gouychapman}
@@
 
-\begin{figure}[h!t]
-\includegraphics[width=0.495\textwidth]{test3.pdf}
-\includegraphics[width=0.495\textwidth]{test3-rho.pdf}
-\caption{Nonlinear regime: For larger surface charges the solution deviates from the low-potential approximation Eq.
-\ref{gouychapman} to the Gouy-Chapman theory.
-The righthand picture shows the charge distribution across the gap.}
-\label{fig2} 
-\end{figure}
+\subsubsection{Liquid-junction potential}
 
-A number of minor bugs in the parallel implementation of the
-SOR solver were resolved, and the unit tests updated appropriately.
-
-Some effort has been used to create a comprehensive test suite
-for the basic functionality involved in these routines. These
-tests are valid both in serial and in parallel.
-
-Work is now under way to couple the new Nernst-Planck and
-Poisson routines with the main code. This includes work to
-include charged colloids.
-
-\section{Week 3 (30 April - 4 May)}
-
-We have now coupled the Poisson solver and the Nernst Planck solver
-to the main code and checked for correct operation. This has included
-the run time intialisation of the electrokinetic-related quatities
-from the input file. We have also integrated the parallel I/O for
-the electrokinetic lattice quantities to the main code, again with
-appropriate initialisation from input.
-
-We proceeded with another test. The liquid junction potential 
+The liquid junction potential 
 is a charge separation process that 
@@
 dynamics without having to account for advective terms.
+
+\begin{figure}[h!t]
+\includegraphics[width=0.495\textwidth]{./pics/test_lj_zoom1.pdf}
+\includegraphics[width=0.495\textwidth]{./pics/test_lj_zoom3.pdf}
+\caption{Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue).
+The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$.
+Deviations can be seen which are presumably due
+to the approximate nature of the analytical solution and the fact that we
+have a different asymptotic flux close to the boundary - the theoretical 
+analysis assumes no flux at a boundary which is infinitely far away from
+the diffusive region.} 
+\label{fig3} 
+\end{figure}
 
@@
 
+\begin{figure}[htpb]
+\includegraphics[width=0.9\textwidth]{./pics/test_lj_decay1.pdf}
+\caption{Rescaled plot of the decay: Times have been Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue).
+The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$.
+Deviations can be seen which are due
+to the approximate nature of the analytical solution and the diffusive fluxes close to the boundary.} 
+\label{fig4} 
+\end{figure}
+
 A more exact solution can be derived in the limit $l_D/L_x<<1, N\to\infty$:
@@
 Note the factor 2 difference between Eq. \ref{taud} and the corresponding expression in \cite{Mafe}.
-\begin{figure}[h!t]
-\includegraphics[width=0.495\textwidth]{test_lj_zoom1.pdf}
-\includegraphics[width=0.495\textwidth]{test_lj_zoom3.pdf}
-\caption{Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue).
-The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$.
-Deviations can be seen which are due
-to the finite size of our system and the approximate nature of the analytical solution.} 
-\label{fig3} 
-\end{figure}
 
@@
 The results for the potential difference over time are shown in Fig. \ref{fig3}.
-
-\begin{figure}[h!t]
-\includegraphics[width=0.9\textwidth]{test_lj_decay1.pdf}
-\caption{Rescaled plot of the decay: Times have been Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue).
-The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$.
-Deviations can be seen which are due
-to the finite size of our system and the approximate nature of the analytical solution.} 
-\label{fig4} 
-\end{figure}
 
@@
 
-A reason for this might be that the analytical solution was gained
+A reason for this might be the approximate nature of the analytical solution 
+and the fact that it was gained
 for an infinitely large system with constant charge concentrations,
@@
 
-We have run some simple scaling tests for the electrokinetic problem
-with zero fluid velocity. A system of $128^3$ lattice sites was used
-as representative of a small production system. The code was run on
-the Mare Nostrum machine in Barcelona for a small number of time
-steps to assess performance on up to 512 MPI tasks. The results are
-shown in Fig. \ref{fig5}.
-
-While the scaling for the Nernst Planck part of
-the calulcation are reasonable, it can be seen that the SOR routine
-used to solve the Poisson equation is not really acceptable. This
-was not unexpected. Some improvements might be possible, but the
-algorithm is essentailly unsuitable for large systems. We will
-investigate the use of an alternative (e.g., multigrid), but this
-is beyond the scope of this HPC Europa visit.
-
-In discussion with Pagonabarraga, we have also made some progress
-in deciding how to modify a number of other parts of the code to
-allow for the presence of the electric potential. These include
-the free energy calculation, which can be generalised to compute
-both electric contributions and coupling between potential and
-compositional order parameter. A flexible solution may require
-some refactoring.
-
-We have also implemented a fixed external electric field in the
-Nernst Planck equation. This will be tested during the last week
-of the visit.
-
-\begin{figure}[h!t]
-\includegraphics[width=0.9\textwidth]{scaling.pdf}
-\caption{Scaling of a system of $128^3$ lattice sites on up to 512 MPI tasks. 
-While the scaling for the Nernst Planck part of the calulcation are reasonable, 
-it can be seen that the SOR routine used to solve the Poisson equation deviates
-strongly from ideal behaviour.} 
-\label{fig5} 
-\end{figure}
-
-\section{Week 4 (7 May - 11 May)}
+\subsubsection{Electroosmotic Flow}
 
@@
 
-\begin{figure}[h!t]
-\includegraphics[width=0.9\textwidth]{test_eo.pdf}
+\begin{figure}[htpb]
+\includegraphics[width=0.9\textwidth]{./pics/test_eo.pdf}
 \caption{Steady state flow profile across a slit of width $L_x=63$ for an applied field of magnitude $e \beta E L_x=1.89$ for two different charge densities $\sigma=0.003125$ (red) and $\sigma=0.0125$ (blue), Bjerrum length $l_B=0.7234$, viscosity $\eta=0.1$ and unit charge $e=1$.
@@
 
-Finally, we have tested the implementation for a single fixed
+\subsubsection{Debye-H\"uckel Theory}
+
+We have tested the implementation for a single fixed
 colloid and compared the result with Debye-H\"uckel theory. A result
@@
 
-\begin{figure}[h!t]
-\includegraphics[angle=-90,width=0.9\textwidth]{test_dh1.pdf}\\
-\includegraphics[angle=-90,width=0.9\textwidth]{test_dh2.pdf}
+\begin{figure}[htpb]
+\includegraphics[angle=-90,width=0.9\textwidth]{./pics/test_dh1.pdf}\\
+\includegraphics[angle=-90,width=0.9\textwidth]{./pics/test_dh2.pdf}
 \caption{Debye-H\"uckel theory for positively charged colloid: the picture
@@
 \end{figure}
+\clearpage
 
+\subsubsection{Scaling of the SOR}
+
+We have run some simple scaling tests for the electrokinetic problem
+with zero fluid velocity. A system of $128^3$ lattice sites was used
+as representative of a small production system. The code was run on
+the Mare Nostrum machine in Barcelona for a small number of time
+steps to assess performance on up to 512 MPI tasks. The results are
+shown in Fig. \ref{fig5}.
+
+While the scaling for the Nernst Planck part of
+the calulcation are reasonable, it can be seen that the SOR routine
+used to solve the Poisson equation is not really acceptable. This
+was not unexpected. Some improvements might be possible, but the
+algorithm is essentailly unsuitable for large systems.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.9\textwidth]{./pics/scaling.pdf}
+\caption{Scaling of a system of $128^3$ lattice sites on up to 512 MPI tasks. 
+While the scaling for the Nernst Planck part of the calulcation are reasonable, 
+it can be seen that the SOR routine used to solve the Poisson equation deviates
+strongly from ideal behaviour.} 
+\label{fig5} 
+\end{figure}
 \clearpage
-\begin{thebibliography}{99}
 
-\bibitem{Lyklema} J. Lyklema {\em Fundamentals of Interface and Colloid Science} Academic Press (1995).
-\bibitem{Mafe} S. Maf\'e, J.A. Manzanares, J. Pellicer, {\em J. Electroanal. Chem.} {\bf 241}, 57-77 (1988).
-\bibitem{Capuani} F. Capuani, I. Pagonabarraga, D. Frenkel, {\em J. Chem. Phys.} {\bf 121}, 973-986 (2004).
-\bibitem{Rotenberg} B. Rotenberg, I. Pagonabarraga, D. Frenkel, {\em Farad. Discuss.} {\bf 144}, 223-243 (2010).
-
-\end{thebibliography}
-
-\end{document}
+\subsection{User input}

### 2013-01-31 htdocs/references.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1902 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+\addcontentsline{toc}{section}{References}
+
+\bibliographystyle{plain}
 \begin{thebibliography}{99}
@@
 \textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
+
+\bibitem{fournier2005}
+J.-B. Fournier and P. Galatola,
+Modeling planar degenerate wetting and anchoring in nematic liquid
+crystals,
+\textit{Europhys. Lett.}, \textbf{72} 403--409 (2005).
 
@@
 
+\bibitem{Lyklema} J. Lyklema {\em Fundamentals of Interface and Colloid Science} Academic Press     (1995).
+\bibitem{Mafe} S. Maf\'e, J.A. Manzanares, J. Pellicer, {\textit J. Electroanal. Chem.} {\textbf 241}, 5    7-77 (1988).
+\bibitem{Capuani} F. Capuani, I. Pagonabarraga, D. Frenkel, {\textit J. Chem. Phys.} {\textbf 121}, 973-    986 (2004).
+\bibitem{Rotenberg} B. Rotenberg, I. Pagonabarraga, D. Frenkel, {\textit Farad. Discuss.} {\textbf 144},     223-243 (2010).
+\bibitem{Landau-ED} L.D. Landau, E.M. Lifshitz, {\textit Electrodynamics of Continuous Media}, \S 15    , 2nd ed., Pergamon Press, Oxford, UK (1984).
+\bibitem{Melcher} J.R. Melcher, {\textit Continuum Electromechanics}, \S 3.10, MIT Press, Cambridge,     MA, USA (1981).\\
+downloadable from:\\
+\url{http://ocw.mit.edu/ans7870/resources/melcher/resized/cem_811.pdf}
+\bibitem{Landau-EL} L.D. Landau, E.M. Lifshitz, {\textit Theory of Elasticity}, \S 3 \& \S 16, 3rd e    d., Butterworth-Heinemann, Oxford, UK (1986).
+
 \end{thebibliography}

### 2013-02-27 htdocs/colloid.tex
Additional comments on cholesteric BC
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1921 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-01-31 trunk/htdocs/colloid.tex
@@
 In three dimensions, the boundary condition Eq.~\ref{cholesteric_bc}
-contains (potentially) 15 unknown derivatives
+provides six equations containing (potentially) 18 unknown derivatives
 $\partial_\gamma Q_{\alpha\beta}$
-corresponding to the 5 independent elements of the order parameter
-tensor $Q_{xx}$, $Q_{xy}$, $Q_{xz}$, $Q_{yy}$, and $Q_{yz}$. At a
-flat surface with, e.g., $n = (1, 0, 0)$, this is reduced to five unknown
-gradients $\partial_x Q_{\alpha\beta}$ at the
+corresponding to the 6 independent elements of the order parameter
+tensor $Q_{xx}$, $Q_{xy}$, $Q_{xz}$, $Q_{yy}$, $Q_{yz}$, $Q_{zz}$.
+Note that the equation corresponding to $Q_{zz}$ must appear to
+retain isotropy. However, $Q_{zz}$ itself, and its derivatives,
+ may be replaced via the
+constraint on the trace of $Q_{\alpha\beta}$. We can therefore either
+solve a fully determined system including $Q_{zz}$, and then impose
+tracelessness on the result, or replace $Q_{zz}$ and solve six
+equations for five unknowns, with the sixth equation acting as the
+constraint. These methods provide the same answer for cases where
+the surface normal is along the coordinate directions.
+
+At a flat surface with, e.g., $n = (1, 0, 0)$, the number of unknowns
+are the gradients $\partial_x Q_{\alpha\beta}$ at the
 boundary if we assume the tangential gradients $\partial_y Q_{\alpha\beta}$
@@
 
-\begin{tabular}{|c|ccccc|}
+\begin{tabular}{|c|cccccc|}
 
@@
 &
-$Q_{xx,x}$ & $Q_{xy,x}$ & $Q_{xz,x}$ & $Q_{yy,x}$ & $Q_{yz,x}$ \\
+$Q_{xx,x}$ & $Q_{xy,x}$ & $Q_{xz,x}$ & $Q_{yy,x}$ & $Q_{yz,x}$ & $Q_{zz,x}$\\
 \hline
 $Q_{xx}$ &
-$\kappa_0 n_x$ & $-\kappa_1 n_y$ & $-\kappa_1 n_z$ & & \\
+$\kappa_0 n_x$ & $-\kappa_1 n_y$ & $-\kappa_1 n_z$ & & &\\
 $Q_{xy}$ &
-$\kappa_0 n_y$ & $\kappa' n_x$ & & $-\kappa_1 n_y$  & $-\kappa_1 n_z$  \\
+$\kappa_0 n_y$ & $\kappa' n_x$ & & $-\kappa_1 n_y$  & $-\kappa_1 n_z$ & \\
 $Q_{xz}$ &
-$\kappa' n_z$ & & $\kappa' n_x$ & $\kappa_1 n_z$ & $-\kappa_1 n_y$ \\
+$\kappa_0 n_z$ & & $\kappa' n_x$ & & $-\kappa_1 n_y$ &$ -\kappa_1 n_z$\\
 $Q_{yy}$ &
- & $\kappa_0 n_y$ & & $\kappa_1 n_x$ & \\
+ & $\kappa_0 n_y$ & & $\kappa_1 n_x$ & &\\
 $Q_{yz}$ &
- & $\kappa_0 n_z$ & $\kappa_0 n_y$ & & $2\kappa_1 n_x$ \\
+ & $\kappa_0 n_z$ & $\kappa_0 n_y$ & & $2\kappa_1 n_x$ & \\
+$Q_{zz}$ &
+ & & $\kappa_0 n_z$ & & & $\kappa_1 n_x$\\
 \hline
@@
 &
-$Q_{xx,y}$ & $Q_{xy,y}$ & $Q_{xz,y}$ & $Q_{yy,y}$ & $Q_{yz,y}$ \\
+$Q_{xx,y}$ & $Q_{xy,y}$ & $Q_{xz,y}$ & $Q_{yy,y}$ & $Q_{yz,y}$ & $Q_{zz,y}$\\
 \hline
 $Q_{xx}$ &
-$\kappa_1 n_y$ & $\kappa_0 n_x$ & & & \\
+$\kappa_1 n_y$ & $\kappa_0 n_x$ & & & &\\
 $Q_{xy}$ &
-$-\kappa_1 n_x$ & $\kappa' n_y$ & $-\kappa_1 n_z$ & $\kappa_0 n_x$ & \\
+$-\kappa_1 n_x$ & $\kappa' n_y$ & $-\kappa_1 n_z$ & $\kappa_0 n_x$ & &\\
 $Q_{xz}$ &
- & $\kappa_0 n_z$ & $2\kappa_1 n_y$ & & $\kappa_0 n_x$ \\
+ & $\kappa_0 n_z$ & $2\kappa_1 n_y$ & & $\kappa_0 n_x$ & \\
 $Q_{yy}$ &
- & $-\kappa_1 n_x$ & & $\kappa_0 n_y$ & $-\kappa_1 n_z$\\
+ & $-\kappa_1 n_x$ & & $\kappa_0 n_y$ & $-\kappa_1 n_z$ & \\
 $Q_{yz}$ &
-$\kappa_1 n_z$ & & $-\kappa_1 n_x$ & $\kappa' n_z$ & $\kappa' n_y$\\
+ & & $-\kappa_1 n_x$ & $\kappa_0 n_z$ & $\kappa' n_y$ & $-\kappa_1 n_z$\\
+$Q_{zz}$ &
+ & & & & $\kappa_0 n_z$ & $\kappa_1 n_y$\\
 \hline
@@
 &
-$Q_{xx,z}$ & $Q_{xy,z}$ & $Q_{xz,z}$ & $Q_{yy,z}$ & $Q_{yz,z}$ \\
+$Q_{xx,z}$ & $Q_{xy,z}$ & $Q_{xz,z}$ & $Q_{yy,z}$ & $Q_{yz,z}$ & $Q_{zz,z}$\\
 \hline
 $Q_{xx}$ &
-$\kappa_1 n_z$ & & $\kappa_0 n_x$ & & \\
+$\kappa_1 n_z$ & & $\kappa_0 n_x$ & & & \\
 $Q_{xy}$ &
- & $2\kappa_1 n_z$ & $\kappa_0 n_y$ & & $\kappa_0 n_x$ \\
+ & $2\kappa_1 n_z$ & $\kappa_0 n_y$ & & $\kappa_0 n_x$ & \\
 $Q_{xz}$ &
-$-\kappa' n_x$ & $-\kappa_1 n_y$ & $\kappa' n_z$ & $-\kappa_0 n_x$ & \\
+$-\kappa_1 n_x$ & $-\kappa_1 n_y$ & $\kappa' n_z$ & & & $\kappa_0 n_x$  \\
 $Q_{yy}$ &
- & & & $\kappa_1 n_z$ & $\kappa_0 n_y$  \\
+ & & & $\kappa_1 n_z$ & $\kappa_0 n_y$ & \\
 $Q_{yz}$ &
-$-\kappa_0 n_y$ & $-\kappa_1 n_x$ & & $-\kappa' n_y$ & $\kappa' n_z$ \\
+ & $-\kappa_1 n_x$ & & $-\kappa_1 n_y$ & $\kappa' n_z$ & $\kappa_0 n_y$ \\
+$Q_{zz}$ &
+ & & $-\kappa_1 n_x$ & & $-\kappa_1 n_y $ & $\kappa_0 n_z$\\
 \hline
@@
 \caption{Coefficients of the various derivatives of the order parameter
-tensor appearing in five equations for the independent elements of the
-order parameter. Note $\kappa_0 + \kappa_1 = \kappa'$ and all the
-coefficients have been multiplied by a factor of 2.}
+tensor appearing in six equations for the elements of the
+order parameter (including $Q_{zz}$).
+Note $\kappa_0 + \kappa_1 = \kappa'$ and all the
+coefficients have been multiplied by a factor of 2 in the off-diagonal
+equations.}
 \label{tab:cholesteric_bcs}

### 2013-03-11 trunk/papers/bprheo/bprheo.tex
More detail on method in section 2
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1938 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
-gradient and vorticity direction -- therefore $\Pi_{xy}$ is the shear stress.
-
-The system of coupled partial differential equations Eqs.~\ref{op-eom}
-and~\ref{NSE} is solved by means of a
-hybrid scheme~\cite{Marenduzzo:2007}, which combines different numerical schemes
-and provides a way to address the multiscale nature of the problem.
-More details on the algorithm and the boundary conditions can
-be found in~\cite{Denniston:2001, Denniston:2004, Henrich:2012a}.
+The system of coupled partial differential equations~\ref{op-eom}
+and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
+which uses a combination of lattice Boltzmann and finite difference schemes.
+(This is in contrast with some earlier methods using solely
+LB~\cite{Denniston:2001, Denniston:2004}.)
+The Navier-Stokes
+equation is solved via the lattice Boltzmann approach, using a standard
+three-dimensional model with 19 discrete velocities (D3Q19).
+A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
+used and the time step is $\Delta t = 1$ in lattice units.
+Coupling to the thermodynamic sector is via a
+local body force computed as the divergence of the thermodynamic
+stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
+of the time evolution of $\mathbf{Q}$ via a standard finite difference
+method using the same grid and the same time step as the LB. The system
+is periodic in all three coordinate directions. Sliding
+boundary conditions, or Lees-Edwards planes, are explicitly implemented
+in both hydrodynamic and thermodynamic sectors to impose shear on the
+system. On
+crossing a Lees-Edwards boundary a Galilean transformation is applied
+with a velocity increment which is fixed in time (and limited in size by the
+low Mach number constraint of LB). An appropriate transformation
+of LB distributions which propagate across a boundary~\cite{Wagner:2002}
+must be applied at each time step, and appropriate
+adjustment to $W_{\alpha\beta}$ is required to compute cross-plane
+gradients of the velocity field used in the update to ${\mathbf Q}$.
+In both cases interpolation of the
+relevant quantities is required to cope with the relative displacement
+of neighbouring lattice sites separated by a sliding plane (the
+displacement may be a fraction of a lattice unit at any given time
+step). The use of multiple sliding planes equally spaced in a single
+system allows
+the overall shear rate to be maintained indefinitely as the system
+becomes larger
+in the velocity gradient direction (in contrast with the use of solid
+walls to impose shear). This hybrid method with Lees-Edwards planes
+has been used successfully to study e.g., smectics~\cite{Henrich:2012a}.
 
@@
 is equal to about $10^8$~Pa. Further details about the conversion 
-from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.
+from LBU to SI units can be found elsewhere~\cite{Henrich:2011a,Henrich:2010b}.
+In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
+gradient and vorticity direction; $\Pi_{xy}$ is therefore the shear stress.
 
@@
 Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
-We performed also selected simulations on metastable states at higher and lower temperatures, 
+We also performed selected simulations on metastable states at higher and lower temperatures, 
 and at different chiralities, but did not find any significant differences in the 
@@
 
-Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
+Simple shear flow was imposed by means of the Lees-Edwards boundary
+conditions with
 the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
@@
 In this Appendix we give some more details on our simulations,
-including the importance and effect of boundary conditions.
+including the relationship between boundary conditions and
+permeation flow.
 

### 2013-04-20 trunk/htdocs/plan.tex
Added minutes of recent coordination meeting
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1955 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-06-21 htdocs/plan.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[11pt,twoside]{article}
 
-% This is a section to be included in ludwig.tex
+\usepackage{amsmath}
+\usepackage{moreverb}
+\usepackage{lscape}
+\usepackage{epic}
+\usepackage[pdftex]{graphicx}
+\usepackage{bm}
+\usepackage{hyperref}
+\usepackage{color}
+
+\setlength{\hoffset}{-1in}
+\setlength{\voffset}{-1in}
+
+\setlength{\topmargin}{2cm}
+\setlength{\evensidemargin}{3.1cm}
+\setlength{\oddsidemargin}{3.1cm}
+
+\setlength{\textwidth}{14.8cm}
+\setlength{\textheight}{23cm}
+
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{\smallskipamount}
+
+\newcommand{\inputkey}[1]{\framebox{\textbf{\texttt{#1}}}}
+\newcommand{\e}[1]{\cdot10^{#1}}
+\newcommand{\beq}{\begin{equation}}
+\newcommand{\eeq}{\end{equation}}
+\newcommand{\beqa}{\begin{eqnarray}}
+\newcommand{\eeqa}{\end{eqnarray}}
+\newcommand{\com}[1]{\textcolor{red}{#1}}
+\newcommand{\cur}[1]{{\textit{#1}}}
+
+\begin{document}
+
+\setcounter{page}{1}
+
+\tableofcontents
+
+\newpage
+
+\setcounter{page}{1}
 
@@
 will be in place to coincide with the public release.
+
+\newpage
+\section{Coordination Meetings}
+
+We may want to keep a log here of various issues and points we discussed.
+
+\subsection{Meeting 16 April 2013}
+
+Alan, Kevin, Davide and Oliver were participating.
+We had a general discussion about the scientific agenda for the 
+next years, the status quo of the code, necessary development 
+steps and open issues.
+
+\subsubsection{Scientific Agenda}
+
+The scientific agenda with Ludwig for the next few years could look like this:
+
+\begin{itemize}
+\item Colloids:\\
+On the CPU colloids in LC are fully supported and the boundary condition for 
+the anchoring has been recently corrected. This functionality should 
+not require any further work on the CPU.
+
+The GPU version could be interesting for very large colloids with respect to 
+the pitch length. An upper limit of currently $2048^3$ lattice sites was mentioned.
+
+The GPU version could be also interesting for hydrodynamic aspects of 
+glass formers. The typical model system would be a 20:80 Kob-Anderson 
+mixture of radii 1:1.4.
+
+\item Liquid crystal droplets and emulsions:\\
+LC-emulsions are currently being developed in the hybrid code as Ludwig 
+proved to be unstable. This may be due to spurious currents that do not
+occur in the hybrid code as some terms are not modelled as divergence of a stress tensor.  
+It would be maybe necessary to merge the models with Ludwig if possible 
+once clarity has been gained how 3D-droplets can be modelled and
+what causes the instability in Ludwig.
+
+\item Electrokinetic phenomena:\\
+The electrokinetics branch is currently emerging. A first project will be
+on spinodal decomposition in the presence of charges. We will continure
+with asymmetrically charged ionic liquids. Modelling the electrophoretic
+behaviour of bacteria via colloids that have a conducting cell wall will
+be another line of research.
+
+\item Polymers:\\
+Polymers are currently not supported, but are on the agenda. 
+At the moment we don't have connectivity between the colloids,
+which could form a polymer-bead-spring model. As Ludwig supports subgrid particles 
+having polymers consisting of pointlike particles would be an interesting 
+alternative in more dilute regimes.
+
+A possible route to polymers in Ludwig could be Burkhard Duenweg's 
+approach (reference missing here).
+
+Another possible realisation could be to develp an add-on for 
+LAMMPS which is based on Ludwig and LAMMPS would be used for the entire MD-part. 
+At the moment it is not clear how the coupling could be achieved. Similar 
+ideas have already surfaced (see F.E. Mackay, C. Denniston, 
+'Coupling MD particles to a lattice-Boltzman fluid through the use of 
+conservative forces', J. Comput. Phys. 237, 289-298 (2013).)
+
+\end{itemize}
+
+\subsubsection{Status Quo and Open Issues}
+
+There are currently three major branches in the repository:
+
+\begin{itemize}
+\item Main branch:\\
+We are still working on fixes for a few issues with main branch. 
+More precisely the anchoring of the LC at the 
+surface of the colloids resulted in anisotropic states and cyclic permutation 
+of the Cartesian coordinates did not lead to the same results. 
+This issue has been fixed and appropriate unit test were suggested.
+The question has been raised how we want to test new functionalities in general.
+A possible solution might be to think about a test case against which
+the new functionality can be tested, e.g. isotropy of the free energy density 
+against permutation of the coordinates in the above case.
+
+Another issue was the bug in the trace calculation which also has been removed.  
+Also, incompressibility is now enforced after each time step.
+
+We are still in the process of comparing Ludwig's output 
+with Davide's hybrid code. This flagged 
+up another problem related to the predictor-corrector algorithm which is 
+currently not implemented in Ludwig, resulting in different convergence 
+behaviour of specific test case presumable due to a lack of accuracy.  
+Possible solutions are RK2, RK4 or semi-implicit methods.
+
+The fluid-solid interaction via bounce-back is perhaps not state-of-the-art
+anymore. In a discussion with Ignacio Pagonabarraga earlier this year
+it turned out that alternative realisations like Yamamoto's approach 
+(Y. Nakayama, R. Yamamoto, PRE 71, 036707 (2005).) with
+a smooth profile may have advantages, but could make it more difficult 
+to impose a variety of boundary conditions. Immersed boundary conditions
+are generally unsuitable for complex fluids.
+
+\item GPU branch:\\
+The GPU branch has diverged from the main branch during summer 2012.
+We are currently thinking how it can be merged with the other branches.
+
+\item Testcharge branch:\\
+The testcharge branch with its electrokinetic functionality is still 
+in statu nascendi. We have derived a Maxwell stress tensor whose divergence
+is giving the force input into Ludwig. This still has to be tested.
+
+The testcharge branch has seen a major technology refresh and will be merged with
+the main branch.
+
+\end{itemize}
+
+\subsubsection{Upcoming Development Steps}
+
+\begin{itemize}
+\item We need to improve the accuracy in the finite difference part of the code. 
+A first approach will use an RK2 or RK4 integrator because of simplicity.
+
+\item The solution of the Poisson equation in the electrokinetic branch needs to be generalised
+so that non-homogeneous dielectric permittivities can be used. We are currently
+trying to identify possible solutions. An iterative procedure will have to be
+found. 
+Moreover, we need to test the implementation of the Maxwell stress tensor against the 
+force input via gradients of the chemical potential.
+
+Later this year the testcharge branch will be merged with the main branch in a major
+technical update.
+
+\item The merging of the GPU and the main branch requires suitable and coherent concepts
+of abstraction to avoid code duplication. The dummy approach used for the serial
+version of the code, which replaces MPI directives with a MPI neutral dummy, will
+prove difficult to copy as CUDA C has some limitations.
+
+\item It would be desireable to increase the internal user base so that issues 
+can be spotted more easily.
+
+\item For the planned public release in 2014/2015 we need to think about an appropriate
+license.
+
+\end{itemize}
+
+\end{document}

### 2013-05-29 papers/bprheo_r/bprheo.tex
Added new directory with revised bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1991 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-03-11 trunk/papers/bprheo/bprheo.tex
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
+\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
 \caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd$ (normalised to $\eta$) 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
+\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_b.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_c.pdf}
 \caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 

### 2013-05-30 trunk/papers/bprheo_r/bprheo.tex
Update paper bprheo_r 30 May
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1992 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-05-29 papers/bprheo_r/bprheo.tex
@@
 \usepackage{balance} 
+\usepackage{color}
 
@@
 \newcommand{\com}[1]{{\bf {#1}}}
+\newcommand{\rev}[1]{{\textcolor{red}{#1}}}
 
@@
 for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
-$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.6666$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
 The thermodynamic stress tensor reads explicitly
@@
 $v$ and $l$ a typical velocity and length scale. 
-In the present work $l=p_0/2$ was used as this is the approximate size of
+In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
 the BP unit cell. Likewise, $v$ was taken to be the velocity difference
-across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.
+across one unit cell, i.e. $v=\dot{\gamma} l$. The elastic constant was taken as $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
 
@@
 defined as 
-$\eta_{app}=\langle \Pi_{xy} \rangle/\gd$, and $\eta$.
+$\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and $\eta$.
 A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
@@
 \center
-\includegraphics[width=0.35\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.35\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.35\textwidth]{disc-165k_run902.png}
+\includegraphics[width=0.35\textwidth]{disc-160k_run902r.png}
+\includegraphics[width=0.35\textwidth]{disc-164k_run902r.png}
+\includegraphics[width=0.35\textwidth]{disc-165k_run902r.png}
 \caption{Disclination network of BPII in shear flow: 
@@
 \begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902r.png}
+\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903r.png}
 \caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
@@
 
-If the flow rate exceeds a certain critical value BPII does not undergo the 
-homogeneous deformation described above and adopts another mode of flow. 
-Fig. \ref{bp2-high} shows a sequence of cuts through the director 
-field at high shear rate. The entire blue phase network has broken up 
-into a simple cholesteric helix and flows as a travelling 
-helical wave with the helical axis oriented along the vorticity direction.
-The state is translationally invariant in the flow and gradient directions.
-It is the preferred mode of flow of a standard cholesteric phase
-in vorticity mode at low Ericksen numbers. This is followed by
-the uncoiling transition to a nematic, flow-aligned state at even 
-higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
-
-In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
-corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
-parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
-see Table \ref{tab1} in the Appendix).
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
-$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+%If the flow rate exceeds a certain critical value BPII does not undergo the 
+%homogeneous deformation described above and adopts another mode of flow. 
+%Fig. \ref{bp2-high} shows a sequence of cuts through the director 
+%field at high shear rate. The entire blue phase network has broken up 
+%into a simple cholesteric helix and flows as a travelling 
+%helical wave with the helical axis oriented along the vorticity direction.
+%The state is translationally invariant in the flow and gradient directions.
+%It is the preferred mode of flow of a standard cholesteric phase
+%in vorticity mode at low Ericksen numbers. This is followed by
+%the uncoiling transition to a nematic, flow-aligned state at even 
+%higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
+
+%In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
+%corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
+%parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
+%see Table \ref{tab1} in the Appendix).
+
+%\begin{figure}[htpb]
+%\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
+%\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
+%$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
 The colour code gives the alignment of the director field in the flow direction (x) with red fully aligned and blue out of plane.
-The configuration is a helical wave that travels along the direction of vorticity.
-It is translationally invariant along the gradient and flow direction.}
-\label{bp2-high}
-\end{figure}
+%The configuration is a helical wave that travels along the direction of vorticity.
+%It is translationally invariant along the gradient and flow direction.}
+%\label{bp2-high}
+%\end{figure}
 
@@
 \hline
-BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+BPI-1 &0.24 &0.02 & 0.1388 & & & & & & & & & \\
+BPI-1 &0.49 &0.04 & 0.1388 & & & & & & & & & \\
+BPI-1 &0.98 &0.08 & 0.1388 & & & & & & & & & \\
+BPI-1 &1.95 &0.17 & 0.1388 & & & & & & & & & \\
+%BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+%BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+%BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+%BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
 \hline
-BPI-2 & 3.91 & 0.67 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+BPI-2 &3.91 &0.33 & 0.1388 & & & & & & & & & \\
+BPI-2 &7.81 &0.67 & 0.1388 & & & & & & & & & \\
+BPI-2 &15.63 &1.33 & 0.1388 & & & & & & & & & \\
+BPI-2 &15.63 &1.33 & -0.1388 & & & & & & & & & \\
+BPI-2 &23.44 &2.00 & 0.1388 & & & & & & & & & \\
+%BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+%BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+%BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+%BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+%BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
 \hline
-BPI-3 & 62.50 & 10.67& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
-BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+BPI-3 &31.25 &2.66 & 0.1388 & & & & & & & & & \\
+BPI-3 &46.87 &4.00 & 0.1388 & & & & & & & & & \\
+BPI-3 &54.69 &4.67 & 0.1388 & & & & & & & & & \\
+BPI-3 &62.50 &5.33 & 0.1388 & & & & & & & & & \\
+BPI-3 &78.13 &6.01 & 0.1388 & & & & & & & & & \\
+%BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
 \hline
-CHG & 93.75 &16.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
-CHG & 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+GJT (BPI-4) &93.75 &8.00 & 0.1388 & & & & & & & & & \\
+GJT (BPI-4) &125.0 &10.66 & 0.1388 & & & & & & & & & \\
+GJT (BPI-4) &187.5 &16.00 & 0.1388 & & & & & & & & & \\
+%GJT (BPI-4)& 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
 \hline
-FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+FAN & 250.0 &21.33 & 0.1388 & & & & & & & & & \\
+FAN & 375.0 &32.00 & 0.1388 & & & & & & & & & \\
+%FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
 \hline
@@
 \hline
-BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+BPII-1 &0.24 &0.02 & 0.1963 & & & & & & & & & \\
+BPII-1 &0.49 &0.04 & 0.1963 & & & & & & & & & \\
+BPII-1 &0.98 &0.08 & 0.1963 & & & & & & & & & \\
+BPII-1 &1.95 &0.17 & 0.1963 & & & & & & & & & \\
+BPII-1 &3.91 &0.33 & 0.1963 & & & & & & & & & \\
+BPII-1 &7.81 &0.67 & 0.1963 & & & & & & & & & \\
+BPII-1 &15.63 &1.33 & 0.1963 & & & & & & & & & \\
+BPII-1 &15.63 &1.33 & -0.1963 & & & & & & & & & \\
+BPII-1 &31.25 &2.66 & 0.1963 & & & & & & & & & \\
+BPII-1 &39.06 &3.33 & 0.1963 & & & & & & & & & \\
+%BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
+%BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
+%BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
+%BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
+%BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
+%BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
+%BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
+%BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
+%BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
 \hline
-BPII-2 & 46.87 &4.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
-BPII-2 & 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+GJT (BPII-2)&46.87 &4.00 & 0.1963 & & & & & & & & & \\
+GJT (BPII-2)&54.69 &4.67 & 0.1963 & & & & & & & & & \\
+GJT (BPII-2)&62.5 &5.33 & 0.1963 & & & & & & & & & \\
+GJT (BPII-2)&78.13 &6.01 & 0.1963 & & & & & & & & & \\
+GJT (BPII-2)&93.75 &8.00 & 0.1963 & & & & & & & & & \\
+GJT (BPII-2)&125.0 &10.66 & 0.1963 & & & & & & & & & \\
+%GJT (BPII-2)& 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+%GJT (BPII-2) & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
+%GJT (BPII-2)& 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
 \hline
-CHG & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
-CHG & 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
-\hline
-FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
+FAN &187.5 &16.00 & 0.1963 &  & & &  &  &  & & & \\
+FAN &250.0 &21.33 & 0.1963 &  & & &  &  &  & & & \\
+%FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
 \hline

### 2013-05-31 trunk/papers/bprheo_r/bprheo.tex
Update bprheo paper 31 May
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1993 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \begin{figure*}[htpb]
-\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115.png}\\
-\includegraphics[width=0.245\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-700k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-2200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115r.png}\\
+\includegraphics[width=0.245\textwidth]{disc-xy-1650k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-1800k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2000k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2400k_run1115r.png}
 \caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
@@
 pictures on the far left, centre left and centre right 
-show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
+show the situation at later time steps $t=1.65, 1.8$ and $2.0\e{6}$,
 which the system passes for several cycles before it ends up in
-an amorphous state (bottom row far right, at time step $t=2.2\e{6}$).}
+an amorphous state (bottom row far right, at time step $t=2.4\e{6}$).}
 \label{bp1-low}
@@
 \begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
-\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
-\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
+\includegraphics[width=0.32\textwidth]{disc-364k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-366k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-368k_run914r.png}\\
+\includegraphics[width=0.32\textwidth]{disc-370k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-372k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-374k_run914r.png}
 \caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
@@
 steady state at $\gd=1.56\e{-4}$ and time steps 
-$t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
-During every cycle the network is displaced also along the gradient and vorticity direction, 
-but the magnitude and direction of the offset does not appear to obey a 
-simple rule.}
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
+During every cycle the network is displaced also along the gradient and vorticity direction.}
 \label{bp1-med}
@@
 \begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-364k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-366k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-368k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-370k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-372k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-374k_run914r.png}
+\caption{Velocity patterns in BPI for positive helicity: 
 The pictures show a snapshot of the periodically recurring patterns in the 
-secondary velocity components $(v_y,v_z)$. 
-The colour code gives the magnitude and sign 
-of the $z$-component. The vertical and horizontal direction are the gradient and 
-vorticity direction, respectively, with the flow into/out of the plane in the 
-upper/lower half of the images.}
+secondary velocity components $(v_y,v_z)$ at $\gd=1.56\e{-4}$ and time steps 
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$, i.e. for the same shear rate
+and time steps as in Fig.\ref{bp1-med}. The colour code gives the magnitude and sign 
+of the $z$-component. For negative helicity the sense of motion of the 
+network and the secondary velocity components are inversed just as in Fig. \ref{bp2-velo}. 
+Note that the viewing direction has been adapted in
+each snapshot to give the clearest view on the band-like pattern between
+the disclination lines.}
 \label{bp1-velo}
@@
 
-\begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a region 
-of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
-direction oriented along the horizontal/vertical axis. 
-The colour code gives the magnitude of the z-component of the director field. 
-The data show that the structure is quasi-2D, with rolling double twist
-patterns in the $xy$ plane, and an approximate translational symmetry 
-along the vorticity direction.}
-\label{bp1-high}
-\end{figure}
+%\begin{figure}[htpb]
+%\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
+%\caption{Director field of BPI at high shear rate: The pictures show a region 
+%of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
+%direction oriented along the horizontal/vertical axis. 
+%The colour code gives the magnitude of the z-component of the director field. 
+%The data show that the structure is quasi-2D, with rolling double twist
+%patterns in the $xy$ plane, and an approximate translational symmetry 
+%along the vorticity direction.}
+%\label{bp1-high}
+%\end{figure}
 

### 2013-06-03 papers/bprheo_r/bprheo.tex
Updated all pictures for bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1994 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \pagestyle{fancy}
+
+\newif\ifold
+\oldtrue
 
@@
 
-If the flow rate exceeds a certain critical value BPII does not undergo the 
+% New conformation is presumably metastable %
+\ifold{If the flow rate exceeds a certain critical value BPII does not undergo the 
 homogeneous deformation described above and adopts another mode of flow. 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
+%\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
 \caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
@@
 \end{figure}
+\fi
 
@@
 
-\if{The negative branch in the shear stress is connected
+\ifold{The negative branch in the shear stress is connected
 to a local maximum followed by a local minimum in the 
@@
 
-\if{In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
+\ifold{In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
 we performed a Fourier analysis of the shear stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
@@
 
-The intricacy and periodic recurrence of the flow-induced conformations 
+% No differences in retake %
+\ifold{The intricacy and periodic recurrence of the flow-induced conformations 
 in BPI raises the question: how robust are these results to variations in 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
-\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
+%\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
+%\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
 \caption{Disclination network and shear stress $\Pi_{xy}$ in BPI: 
@@
 generic features of BPI in this regime of flow.
+\fi
 
@@
 
-At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
+\ifold At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
 we observe a break-up of the flow-induced 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
+%\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
 \caption{Director field of BPI at high shear rate: The pictures show a region 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
+\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
 \caption{Flow induced cholesteric state. The orientation of the helix is 
@@
 \section*{Acknowledgments}
-We acknowledge support by EPSRC grant nos. EP/E045316 and EP/E030173, 
-the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
+
 This work was granted access to the HPC resource of CSC, Finland, made available 
 within DECI by the PRACE-2IP, funded by the European Community's 7th Framework 
-Program under grand agreement No. RI-283493. We thank Peter J. Collings 
-for very stimulating discussions. 
+Program under grand agreement No. RI-283493. We also acknowledge support by 
+EPSRC grant nos. EP/E045316 and EP/E030173 and the MAPPER EU-FP7 project 
+(grant no. RI-261507).We thank Peter J. Collings for stimulating discussions. 
 MEC holds a Royal Society Research Professorship.

### 2013-06-04 trunk/papers/bprheo_r/bprheo.tex
Moving revision into place
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1995 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \pagestyle{fancy}
+
+\newif\ifold
+%\oldtrue
 
@@
 
-%If the flow rate exceeds a certain critical value BPII does not undergo the 
-%homogeneous deformation described above and adopts another mode of flow. 
-%Fig. \ref{bp2-high} shows a sequence of cuts through the director 
-%field at high shear rate. The entire blue phase network has broken up 
-%into a simple cholesteric helix and flows as a travelling 
-%helical wave with the helical axis oriented along the vorticity direction.
-%The state is translationally invariant in the flow and gradient directions.
-%It is the preferred mode of flow of a standard cholesteric phase
-%in vorticity mode at low Ericksen numbers. This is followed by
-%the uncoiling transition to a nematic, flow-aligned state at even 
-%higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
-
-%In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
-%corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
-%parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
-%see Table \ref{tab1} in the Appendix).
-
-%\begin{figure}[htpb]
+% New conformation is presumably metastable %
+\ifold{If the flow rate exceeds a certain critical value BPII does not undergo the 
+homogeneous deformation described above and adopts another mode of flow. 
+Fig. \ref{bp2-high} shows a sequence of cuts through the director 
+field at high shear rate. The entire blue phase network has broken up 
+into a simple cholesteric helix and flows as a travelling 
+helical wave with the helical axis oriented along the vorticity direction.
+The state is translationally invariant in the flow and gradient directions.
+It is the preferred mode of flow of a standard cholesteric phase
+in vorticity mode at low Ericksen numbers. This is followed by
+the uncoiling transition to a nematic, flow-aligned state at even 
+higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
+
+In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
+corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
+parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
+see Table \ref{tab1} in the Appendix).
+
+\begin{figure}[htpb]
 %\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
@@
 %\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
-%\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
-%$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
+\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
+$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
 The colour code gives the alignment of the director field in the flow direction (x) with red fully aligned and blue out of plane.
-%The configuration is a helical wave that travels along the direction of vorticity.
-%It is translationally invariant along the gradient and flow direction.}
-%\label{bp2-high}
-%\end{figure}
+The configuration is a helical wave that travels along the direction of vorticity.
+It is translationally invariant along the gradient and flow direction.}
+\label{bp2-high}
+\end{figure}
+\fi
 
@@
 
-%\begin{figure}[htpb]
+\begin{figure}[htpb]
 %\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
-%\caption{Director field of BPI at high shear rate: The pictures show a region 
-%of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
-%direction oriented along the horizontal/vertical axis. 
-%The colour code gives the magnitude of the z-component of the director field. 
-%The data show that the structure is quasi-2D, with rolling double twist
-%patterns in the $xy$ plane, and an approximate translational symmetry 
-%along the vorticity direction.}
-%\label{bp1-high}
-%\end{figure}
+\caption{Director field of BPI at high shear rate: The pictures show a region 
+of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
+direction oriented along the horizontal/vertical axis. 
+The colour code gives the magnitude of the z-component of the director field. 
+The data show that the structure is quasi-2D, with rolling double twist
+patterns in the $xy$ plane, and an approximate translational symmetry 
+along the vorticity direction.}
+\label{bp1-high}
+\end{figure}
+\fi
 
@@
 \hline
-BPII-1 &0.24 &0.02 & 0.1963 & & & & & & & & & \\
-BPII-1 &0.49 &0.04 & 0.1963 & & & & & & & & & \\
-BPII-1 &0.98 &0.08 & 0.1963 & & & & & & & & & \\
-BPII-1 &1.95 &0.17 & 0.1963 & & & & & & & & & \\
-BPII-1 &3.91 &0.33 & 0.1963 & & & & & & & & & \\
-BPII-1 &7.81 &0.67 & 0.1963 & & & & & & & & & \\
-BPII-1 &15.63 &1.33 & 0.1963 & & & & & & & & & \\
-BPII-1 &15.63 &1.33 & -0.1963 & & & & & & & & & \\
-BPII-1 &31.25 &2.66 & 0.1963 & & & & & & & & & \\
-BPII-1 &39.06 &3.33 & 0.1963 & & & & & & & & & \\
-%BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-%BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-%BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-%BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-%BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-%BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-%BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-%BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-%BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+BPII-1 &0.24 &0.02 & 0.1963 &-7.8 &7.7 &0.1 &-0.2 &0.2 &0.1 &-0.3 &0.3 &0.2 \\
+BPII-1 &0.49 &0.04 & 0.1963 &-15.5 &15.4 &0.2 &-0.3 &0.4 &0.2 &-0.5 &0.4 &0.2 \\
+BPII-1 &0.98 &0.08 & 0.1963 &-31.0 &30.8 &0.3 &-0.6 &0.6 &0.4 &-0.9 &0.8 &0.4 \\
+BPII-1 &1.95 &0.17 & 0.1963 &-62.0 &61.6 &0.6 &-1.2 &1.2 &0.8 &-1.8 &1.6 &0.8 \\
+BPII-1 &3.91 &0.33 & 0.1963 &-123.9 &123.3 &1.2 &-2.1 &2.1 &1.6 &-3.4 &3.0 &1.5 \\
+BPII-1 &7.81 &0.67 & 0.1963 &-247.5 &246.8 &2.0 &-3.8 &3.8 &3.0 &-6.5 &5.5 &2.6 \\
+BPII-1 &15.63 &1.33 & 0.1963 &-494.3 &493.7 &2.5 &-5.5 &-5.5 &4.0 &\bf{-10.9} &\bf{8.7} &\bf{3.9} \\
+BPII-1 &15.63 &1.33 & -0.1963 &-494.3 &493.7 &2.5 &-5.5 &5.5 &4.0 &\bf{-8.7} &\bf{10.9} &\bf{3.9} \\
+BPII-1 &31.25 &2.66 & 0.1963 &-987.5 &987.1 &3.5 &-6.9 &6.9 &5.1 &-17.7 &13.1 &5.0 \\
+BPII-1 &39.06 &3.33 & 0.1963 &1233.9 &1233.6 &3.7 &-6.9 &6.9 &5.1 &-20.0 &14.5 &5.2 \\
 \hline
-GJT (BPII-2)&46.87 &4.00 & 0.1963 & & & & & & & & & \\
-GJT (BPII-2)&54.69 &4.67 & 0.1963 & & & & & & & & & \\
-GJT (BPII-2)&62.5 &5.33 & 0.1963 & & & & & & & & & \\
-GJT (BPII-2)&78.13 &6.01 & 0.1963 & & & & & & & & & \\
-GJT (BPII-2)&93.75 &8.00 & 0.1963 & & & & & & & & & \\
-GJT (BPII-2)&125.0 &10.66 & 0.1963 & & & & & & & & & \\
-%GJT (BPII-2)& 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
-%GJT (BPII-2) & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
-%GJT (BPII-2)& 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
+GJT (BPII-2)&46.87 &4.00 & 0.1963 &-1479.8 &1477.0 &- &- &- &- &-10.0 &9.6  &- \\
+GJT (BPII-2)&54.69 &4.67 & 0.1963 &-1720.5 &1720.8 &- &- &- &- &-11.5 &11.2 &- \\
+GJT (BPII-2)&62.5 &5.33 & 0.1963  &-1966.7 &1966.2 &- &- &- &- &-13.3 &12.7 &- \\
+GJT (BPII-2)&78.13 &6.01 & 0.1963 &-2458.3 &2457.8 &- &- &- &- &-16.9 &15.7 &- \\
+GJT (BPII-2)&93.75 &8.00 & 0.1963 &-2949.7 &2949.7 &- &- &- &- &-20.3 &18.4 &- \\
+GJT (BPII-2)&125.0 &10.66 &0.1963 &-3932.8 &3932.8 &- &- &- &- &-26.6 &22.4 &- \\
 \hline
-FAN &187.5 &16.00 & 0.1963 &  & & &  &  &  & & & \\
-FAN &250.0 &21.33 & 0.1963 &  & & &  &  &  & & & \\
-%FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
+FAN &187.5 &16.00 & 0.1963 &-5906.2  &5906.2 &- &-  &-  &-  &-0.6 &0.4 &- \\
+FAN &250.0 &21.33 & 0.1963 &-7874.9  &7874.9 &- &-  &-  &-  &-0.5 &0.3 &- \\
 \hline

### 2013-06-05 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1999 40876441-3496-0410-a613-a2fca35b2ae2

@@
-%% ****** Start of file apstemplate.tex ****** %
-%%
-%%
-%%   This file is part of the APS files in the REVTeX 4 distribution.
-%%   Version 4.1r of REVTeX, August 2010
-%%
-%%
-%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
-%%
-%%   See the REVTeX 4 README file for restrictions and more information.
-%%
-%
-% This is a template for producing manuscripts for use with REVTEX 4.0
-% Copy this file to another name and then work on that file.
-% That way, you always have this original template file to use.
-%
-% Group addresses by affiliation; use superscriptaddress for long
-% author lists, or if there are many overlapping affiliations.
-% For Phys. Rev. appearance, change preprint to twocolumn.
-% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
-%  Add 'draft' option to mark overfull boxes with black boxes
-%  Add 'showpacs' option to make PACS codes appear
-%  Add 'showkeys' option to make keywords appear
-\documentclass[aps,pre,reprint,superscriptaddress, twocolumn]{revtex4}
-%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
-%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}
-
-% You should use BibTeX and apsrev.bst for references
-% Choosing a journal automatically selects the correct APS
-% BibTeX style file (bst file), so only uncomment the line
-% below if necessary.
-\bibliographystyle{apsrev4-1}
-
-\usepackage{graphicx}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%This is the article LaTeX template for RSC journals
+%Copyright The Royal Society of Chemistry 2010
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\documentclass[8.5pt,twoside,twocolumn]{article}
+\oddsidemargin -1.2cm
+\evensidemargin -1.2cm
+\textwidth 18cm
+\headheight 1.0in
+\topmargin -3.5cm
+\textheight 22cm
+\usepackage[super,sort&compress,comma]{natbib} 
+\usepackage{mhchem}
+\usepackage{times,mathptmx}
+%\usepackage{times}
+% feel free not to use mathptmx if it causes difficulties
+\usepackage{sectsty}
+\usepackage{balance} 
+\usepackage{color}
+
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{float}
-
 \newcommand{\e}[1]{\times10^{#1}}
@@
 \newcommand{\com}[1]{{\bf {#1}}}
+\newcommand{\rev}[1]{{\textcolor{red}{#1}}}
+
+\usepackage{graphicx} %eps figures can be used instead
+\usepackage{lastpage}
+\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+
+\newif\ifold
+%\oldtrue
 
@@
 
-% Use the \preprint command to place your local institutional report
-% number in the upper righthand corner of the title page in preprint mode.
-% Multiple \preprint commands are allowed.
-% Use the 'preprintnumbers' class option to override journal defaults
-% to display numbers if necessary
-%\preprint{}
-
-%Title of paper
-\title{Rheology of Cubic Blue Phases}
-
-% repeat the \author .. \affiliation  etc. as needed
-% \email, \thanks, \homepage, \altaffiliation all apply to the current
-% author. Explanatory text should go in the []'s, actual e-mail
-% address or url should go in the {}'s for \email and \homepage.
-% Please use the appropriate macro foreach each type of information
-
-% \affiliation command applies to all authors since the last
-% \affiliation command. The \affiliation command should follow the
-% other information
-% \affiliation can be followed by \email, \homepage, \thanks as well.
-
-\author{O. Henrich}
-\email[Corresponding author: ]{ohenrich@epcc.ed.ac.uk}
-\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
-
-\author{K. Stratford}
-\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-
-\author{P.V. Coveney}
-\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}
-
-\author{M.E. Cates}
-\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-
-\author{D. Marenduzzo}
-\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
-
-%\homepage[]{Your web page}
-%\thanks{}
-%\altaffiliation{}
-
-%Collaboration name if desired (requires use of superscriptaddress
-%option in \documentclass). \noaffiliation is required (may also be
-%used with the \author command).
-%\collaboration can be followed by \email, \homepage, \thanks as well.
-%\collaboration{}
-
-%\noaffiliation
-
-\date{\today}
-
-\begin{abstract}
-We study the behaviour of cubic blue phases under shear flow
+\thispagestyle{plain}
+\fancypagestyle{plain}{
+\fancyhead[L]{\includegraphics[height=8pt]{headers/LH}}
+\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{headers/CH}}
+\fancyhead[R]{\includegraphics[height=10pt]{headers/RH}\vspace{-0.2cm}}
+\renewcommand{\headrulewidth}{1pt}}
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\renewcommand\footnoterule{\vspace*{1pt}% 
+\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
+\setcounter{secnumdepth}{5}
+
+\makeatletter 
+\def\subsubsection{\@startsection{subsubsection}{3}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\bf}} 
+\def\paragraph{\@startsection{paragraph}{4}{10pt}{-1.25ex plus -1ex minus -.1ex}{0ex plus 0ex}{\normalsize\textit}} 
+\renewcommand\@biblabel[1]{#1}            
+\renewcommand\@makefntext[1]% 
+{\noindent\makebox[0pt][r]{\@thefnmark\,}#1}
+\makeatother 
+\renewcommand{\figurename}{\small{Fig.}~}
+\sectionfont{\large}
+\subsectionfont{\normalsize}
+
+\fancyfoot{}
+\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{headers/LF}}
+\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{headers/RF}}
+\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{headers/RF}}
+\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
+\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
+\fancyhead{}
+\renewcommand{\headrulewidth}{1pt} 
+\renewcommand{\footrulewidth}{1pt}
+\setlength{\arrayrulewidth}{1pt}
+\setlength{\columnsep}{6.5mm}
+\setlength\bibsep{1pt}
+
+\twocolumn[
+  \begin{@twocolumnfalse}
+\noindent\LARGE{\textbf{Rheology of Cubic Blue Phases}}
+\vspace{0.6cm}
+
+\noindent\large{\textbf{Oliver Henrich $^{\ast}$\textit{$^{a,b}$}, Kevin Stratford \textit{$^{a}$}, Peter V.
+Coveney \textit{$^{b}$}, Michael E.
+Cates \textit{$^{c}$}, Davide Marenduzzo \textit{$^{c}$}}}\vspace{0.5cm}
+%Please note that \ast indicates the corresponding author(s) but no footnote text is required.
+
+\noindent\textit{\small{\textbf{Received Xth XXXXXXXXXX 20XX, Accepted Xth XXXXXXXXX 20XX\newline
+First published on the web Xth XXXXXXXXXX 200X}}}
+
+\noindent \textbf{\small{DOI: }}
+\vspace{0.6cm}
+%Please do not change this text.
+
+\noindent \normalsize{We study the behaviour of cubic blue phases under shear flow 
 via lattice Boltzmann simulations. We focus on the two experimentally observed
 phases, Blue Phase I (BPI) and Blue Phase II (BPII). The disclination network of Blue Phase 
-II continuously breaks and reforms under shear, leading to an oscillatory stress response in time.
-For larger shear rates, the network breaks up into a cholesteric helix, lying either along the
-vorticity or the flow gradient direction according to the imposed flow.
+II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
+For larger shear rates, the structure breaks up \rev{into a Grandjean texture with a 
+cholesteric helix lying along the flow gradient direction according to the imposed flow}.
 BPI leads to a very different response. Here, oscillations are only possible for intermediate
 shear rates -- very slow flow leads to a transition into an amorphous network 
-with a residual stress. The flow-induced oscillation regime features a novel arrangement
-of double helical disclinations which rotate due to the shear. Large flow can, 
-just as in the BPII case, lead to the rupture of the network and the stabilisation
-of less ordered cholesteric structures. Our results provide the first 
-theoretical investigation of sheared blue phases in large systems, which
-are relevant to understanding the bulk rheology of these materials. 
-\end{abstract}
-
-% insert suggested PACS numbers in braces on next line
-\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
-% insert suggested keywords - APS authors don't need to do this
-\keywords{blue phase, cholesterics, chiral nematics, rheology}
-
-%\maketitle must follow title, authors, abstract, \pacs, and \keywords
-\maketitle
-
-% body of paper here - Use proper section commands
-% References should be done using the \cite, \ref, and \label commands
+with an apparent yield stress. \rev{Larger shear rates can make the network sensitive
+to flow-induced transitions to metastable and amorphous states. For even
+larger flow rates a Grandjean texture is observed, just as in the case of BPII.
+At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.}
+Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
+\vspace{0.5cm}
+ \end{@twocolumnfalse}
+  ]
+
+%\footnotetext{\dag~Electronic Supplementary Information (ESI) available: [details of any supplementary information available should be included here]. See DOI: 10.1039/b000000x/}
+
+%Please use \dag to cite the ESI in the main text of the article.
+%If you article does not have ESI please remove the the \dag symbol from the title and the above footnotetext.
+
+\footnotetext{\textit{$^{a}$~EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK; E-mail: ohenrich@epcc.ed.ac.uk}}
+\footnotetext{\textit{$^{b}$~Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}}
+\footnotetext{\textit{$^{c}$~SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}}
+
+%additional addresses can be cited as above using the lower-case letters, c, d, e... If all authors are from the same address, no letter is required
+
+%\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P.
+Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}
+
 \section{Introduction}
@@
 of BPs in confined geometries~\cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}, under 
-external fields ~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
+external fields~\cite{Alexander:2008,Fukuda:2009,Henrich:2010a,Castles:2010,Tiribocchi:2011a}, 
 and in the presence of colloidal particles~\cite{Ravnik:2011a}.
-The kinetics of BP domain growth have been recently addressed in~\cite{Henrich:2010b}. 
+The kinetics of BP domain growth have been recently addressed~\cite{Henrich:2010b}. 
 However, our understanding of their dynamical behaviour under flow remains
 very limited. The aim of this work is to address this issue by studying,
-for the first time, the response of confined BP samples to a shear flow.
+for the first time, the response of large BP samples to a shear flow.
 
 Flow response in cholesterics is both strongly non-Newtonian and highly anisotropic.
-For example, if a cholesteric helix is subjected to a Poiseuille flow along
-the helical axis, small pressure differences drive flow mainly through
-''permeation'', first investigated by Helfrich~\cite{Helfrich:1969}.
+For example, if a standard cholesteric phase is subjected to a Poiseuille flow along
+its helical axis, small pressure differences drive flow mainly through
+``permeation'', as first investigated by Helfrich~\cite{Helfrich:1969}.
 In the permeation mode the liquid crystal flows while leaving the director
@@
 They also found that a strong secondary flow appears.
+\rev{Early experiments with cholesteric liquid crystals
+showed that flow can also give rise to conformational transitions. \cite{Press:1978}.}
 Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
 the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
-lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils and 
-leaves a flow-induced nematic phase.
+lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. \rev{These result, however, were derived under the 
+assumption that the molecules rotate only in flow-gradient plane whilst
+the orientation of the cholesteric helix remained unchanged.}  
 Rey also studied cholesterics subjected to both steady flow and low frequency
@@
 Our simulations are large scale and parallel, so that we can study
-significantly larger systems than previously possible, so that our
-results can be tested with bulk rheology experiments.
+significantly larger systems than previously possible, and our
+results can in principle be compared with bulk rheology experiments.
 
@@
 In Section II, we describe the method we use and review
-the hydrodynamic equation of motion which we aim to solve. In Section III, 
-we report our numerical results, separating them into subsections relative to
-blue phase I and blue phase II, and corresponding to low, intermediate, and
+the hydrodynamic equations of motion which we aim to solve. In Section III, 
+we report our numerical results, separating them into subsections referring to
+Blue Phase I and Blue Phase II, and corresponding to low, intermediate, and
 high shear. Finally, we draw our conclusions in Section IV.
@@
 \end{eqnarray}
-The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
-The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
-we work for simplicity in the one-elastic constant approximation. The wavevector $q_0$ is equal to $2\pi/p_0$, where $p_0$ is the cholesteric pitch.
+The first term contains a bulk-free energy constant $A_0$ and the temperature-related parameter $\gamma$ which controls the magnitude of order.
+The second part quantifies the cost of elastic distortions, which is proportional to the elastic constant $K$;
+we work for simplicity in the one-elastic constant approximation~\cite{deGennes}. The wavevector $q_0$ is equal to $2\pi/p_0$, where $p_0$ is the cholesteric pitch.
 The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
-The ``redshift'' $r=p/p_0$ is adjusted during the simulation, to optimise the free energy density
-at all times -- this is done by following the procedure
+Indeed, the ``redshift'' $r=p/p_0$ is adjusted during the equilibration phase of the 
+simulation, to optimise the free energy density
+before shearing begins -- this is done by following the procedure
 previously described in~\cite{Alexander:2006}.
 
-A thermodynamic state is specified by two dimensionless quantities: the effective temperature 
+A thermodynamic state is specified by two dimensionless quantities: the reduced temperature 
 \begin{equation}
-\tau=\frac{27(1-\gamma/3)}{\gamma}
+\tau=\frac{27(1-\gamma/3)}{\gamma},
 \end{equation}
-and chirality 
+which vanishes at the spinodal point of a nematic ($q_0=0$), 
+and the reduced chirality 
 \begin{equation}
@@
 is a constant depending on the molecular details of the liquid crystal.
-Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the 
-co-called Leslie angle: we select this case by 
+Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution for $\theta$, the 
+so-called Leslie angle: we select this case by 
 setting $\xi=0.7$ in our simulations.
-${\mathbf H}$ is the molecular field, which is the functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
+${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that respects the tracelessness of $\mathbf Q$:
 \begin{equation}
@@
 \partial_t v_\alpha +\rho \,v_\beta \partial_\beta v_\alpha
-&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha v_\beta \nonumber\\
-& &\hspace*{-1.5cm} +\; \partial_\beta v_\alpha
-+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}\}. 
+&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta [ \partial_\alpha v_\beta +\; \partial_\beta v_\alpha].
 \label{NSE}
 \end{eqnarray}
-The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
-of the Lattice Boltzmann (LB) equations \cite{Denniston:2001}.
-At low flow rates the fluid can be considered as incompressible, so that the
-last term on the right hand side of Eq.~\ref{NSE} remains small.
-$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LB units.
-The pressure tensor reads explicitly
+This emerges from the Chapman-Enskog expansion
+of the lattice Boltzmann (LB) equations 
+that we solve numerically. A further term $\eta(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu v_\mu \delta_{\alpha \beta}$ that formally appears
+in this expansion is negligible under the slow flows considered here
+for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
+$\eta$ is an isotropic background viscosity which is set to $\eta=0.6666$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
+The thermodynamic stress tensor reads explicitly
 \begin{eqnarray}
@@
 \end{equation}
-with $\eta$ and $K$ being the viscosity and elastic constant and
-$v$ and $l$ as typical velocity and length scale. 
-In the present work $l=p_0/2$ was assumed to be the size of the BP unit cell,
-whereas $v$ was taken to be the velocity difference
-across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.
-
-The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
-\begin{equation}
-X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
-\label{spectrum}
-\end{equation}
-where we subtract the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
-$N$ was the number of points in the time series, typically one for every 100 LB time steps.
-
-The system of coupled partial differential equations Eqs.~\ref{op-eom}
-and~\ref{NSE} is solved by means of a
-hybrid scheme~\cite{Marenduzzo:2007}. 
-More details on the algorithm can
-be found in~\cite{Denniston:2001, Denniston:2004}.
-
-\section{Results and Discussion}
-
-\begin{table*}[htpb]
-\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
-\hline
-& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
-\hline
-BPI \\
-\hline
-BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
-\hline
-BPI-2 & 3.91 & 0.67 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
-\hline
-BPI-3 & 62.50 & 10.67& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
-BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
-\hline
-CHG & 93.75 &16.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
-CHG & 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
-\hline
-FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
-\hline
-
-BPII \\
-\hline
-BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
-\hline
-BPII-2 & 46.87 &4.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
-BPII-2 & 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
-\hline
-CHG & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
-CHG & 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
-\hline
-FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
-\hline
-\end{tabular}
-\caption{Minima, maxima and standard deviation of time-averaged velocity components 
-for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
-shear flow: All values are given in $10^{-5}$ lattice units. The transient dynamics 
-has not been included in the averaging procedure. The regimes comprise the formation
-of an amorphous network (BPI-1), periodically recurring conformations (BPI-2 and BPII-1), 
-two-dimensional double twist rolls (BPI-3), a one-dimensional travelling helical wave (BPII-2), 
-a cholesteric helix oriented along the gradient direction (CHG), 
-and a flow-aligned nematic state (FAN).}
-\label{tab1}
-\end{table*}
+with $\eta$ and $K$ defined previously, and 
+$v$ and $l$ a typical velocity and length scale. 
+In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
+the BP unit cell. Likewise, $v$ was taken to be the velocity difference
+across one unit cell, i.e. $v=\dot{\gamma} l$. \rev{The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.}
+
+The system of coupled partial differential equations~\ref{op-eom}
+and~\ref{NSE} is solved by means of a hybrid method~\cite{Marenduzzo:2007}
+which uses a combination of lattice Boltzmann and finite difference schemes.
+(This is in contrast with some earlier methods using solely
+LB~\cite{Denniston:2001, Denniston:2004}.)
+The Navier-Stokes
+equation is solved via the lattice Boltzmann approach, using a standard
+three-dimensional model with 19 discrete velocities (D3Q19).
+A regular lattice with spacing $\Delta x = \Delta y = \Delta z = 1$ is
+used and the time step is $\Delta t = 1$ in lattice units.
+Coupling to the thermodynamic sector is via a
+local body force computed as the divergence of the thermodynamic
+stress Eq.~\ref{Pi}; the resulting velocity field is used in the computation
+of the time evolution of $\mathbf{Q}$ via a standard finite difference
+method using the same grid and the same time step as the LB. The system
+is periodic in all three coordinate directions. Sliding
+boundary conditions, or Lees-Edwards planes, are explicitly implemented
+in both hydrodynamic and thermodynamic sectors to impose shear on the
+system. On
+crossing a Lees-Edwards boundary a Galilean transformation is applied
+with a velocity increment which is fixed in time (and limited in size by the
+low Mach number constraint of LB). An appropriate transformation
+of LB distributions which propagate across a boundary~\cite{Wagner:2002}
+must be applied at each time step, and appropriate
+adjustment to $W_{\alpha\beta}$ is required to compute cross-plane
+gradients of the velocity field used in the update to ${\mathbf Q}$.
+In both cases interpolation of the
+relevant quantities is required to cope with the relative displacement
+of neighbouring lattice sites separated by a sliding plane (the
+displacement may be a fraction of a lattice unit at any given time
+step). The use of multiple sliding planes equally spaced in a single
+system allows
+the overall shear rate to be maintained indefinitely as the system
+becomes larger
+in the velocity gradient direction (in contrast with the use of solid
+walls to impose shear). This hybrid method with Lees-Edwards planes
+has been used successfully to study e.g., smectics~\cite{Henrich:2012a}.
 
@@
 is equal to about $10^8$~Pa. Further details about the conversion 
-from LBU to SI units can be found in ~\cite{Henrich:2011a,Henrich:2010b}.
+from LBU to SI units can be found elsewhere~\cite{Henrich:2011a,Henrich:2010b}.
+In what follows, $x$, $y$ and $z$ denote respectively the velocity, velocity
+gradient and vorticity direction; $\Pi_{xy}$ is therefore the shear stress.
+
+\section{Results and Discussion}
 
@@
 For these parameters the total free energy density $f$ remained always negative at all flow rates
-simulated (so that the system is always in a liquid crystalline state). 
-We performed also selected simulations at metastable states at higher and lower temperatures, 
+simulated. 
+Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
+We also performed selected simulations on metastable states at higher and lower temperatures, 
 and at different chiralities, but did not find any significant differences in the 
-general flow behaviour that is described below.
-However, for thermodynamic states that are closer to the phase boundary the disclination 
-network broke up into non-BP states at lower shear rates. This is not surprising as
-these states have on average higher free energy densities and smaller order parameters
-than those specified above and cannot therefore withstand the same external forces
-before breaking down.
+general flow behaviour from that described below.
+The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
+rate at which the disclination 
+network broke up into chiral nematic states was lower.
+This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.
 
@@
 Runs with higher resolution confirmed that this choice was adequate to track  
-all kinematic details of the blue phase networks in shear flow.
-
-Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
-the top (bottom) part of the system flowing in positive (negative) x-direction and the 
-velocity gradient along the y-direction.
+all kinematic details of the blue phase networks in shear flow. This includes
+reconstruction of the unit cell not accessible in simulation with only
+few cells~\cite{Dupuis:2005}.
+
+Simple shear flow was imposed by means of the Lees-Edwards boundary
+conditions with
+the top (bottom) part of the system flowing in the positive (negative) $x$-direction and the 
+velocity gradient along the $y$-direction.
 The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ in LB units.
-For clarity we classify various flow regimes, two in case of BPII and three in case of BPI. 
-They are BPII-1 ($\gd \lesssim 3.125\e{-4}$), BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$),
-BPI-1 ($\gd \lesssim 1.95\e{-5}$), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$) and BPI-3 ($6.25\e{-4}\lesssim gd\lesssim 7.813\e{-4}$).
-Beyond these regimes both BPI and BPII form common configurations regardless 
-of the initial state, namely a cholesteric helix with helical axis 
-oriented along the direction of the velocity gradient (CHG), 
-and a flow-aligned nematic state (FAN).  
-Disclination lines are plotted as isosurface of the scalar order parameter $q$. 
+$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
+\rev{For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
+case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory response (PRC), 
+amorphous networks (AN), amorphous and metastable states (AMS) as well as those featuring a Grandjean texture (GJ) and 
+a flow-aligned nematic state (FAN). The last two occur regardless of the initial state. 
+They are for Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
+BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
+BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) and
+for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
+BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
+BPI-3 (AMS) ($3.125\e{-4}\lesssim \gd\lesssim 7.813\e{-4}; 2.67 \lesssim {\it Er} \lesssim 6.01$),
+BPI-4 (GJ) ($9.375\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 8\lesssim {\it Er} \lesssim 16)$ and
+BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.34\lesssim {\it Er}$).}
+
+As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
 Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
+(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII versus time (top) and strain (bottom). 
-For clarity the individual curves have been shifted along the time and strain axis. 
-The inset in the top panel 
-shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
-regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles) 
-and break up of the network into a travelling helical wave (BPII-2, solid circle).}
-The error bars indicate the maximum and minimum stresses that occur during one cycle.
-\label{bp2-rheo}
+\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
+\caption{\rev{Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
+whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
+breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
+The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.}}
+\label{bp2-flowcurve}
+\end{figure}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
+\caption{\rev{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ (normalised to the background viscosity $\eta$) 
+of BPII versus strain $\gamma = \gd\cdot t$ in the regimes BPII-1 (top, PRC) and BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
+The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.}} 
+\label{bp2-appvisc}
 \end{figure}
@@
 flow behaviour at all applied shear rates.
-Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
-deviatoric stress normalised to the isotropic background viscosity: 
-$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
-A numerical value of $\eta_{app}=1$ corresponds to a fully Newtonian flow,
+
+Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
+the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
+\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
+power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
+the data to a very good approximation.
+This holds true even in regime BPII-2, where the network breaks up and 
+the liquid crystal adopts a Grandjean structure. The coefficents of the power law fit are
+here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
+remarkably small in BPII, in the range of shear rates which we have explored here.}
+
+\rev{Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
+defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
+over total strain.
+A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
 without any additional contribution from the liquid crystal.
-The top picture shows the data over time, whereas the graph at the bottom gives 
-the same data versus accumulated strain.
-For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is because of the periodic breakup and reconnecting of the network in shear flow. 
-We refer to this regime as BPII-1. We note that permeation flows, which would
-{\em not} lead to oscillatory response, are effectively disallowed by the combination
-of the geometry and boundary condition we consider, because at the Lees-Edwards 
-plane the network needs to reconnect with its Galilean ``image'' travelling
-at the speed specified by the shear. This means that  our results are effectively
-restricted to intermediate or high flow, where permeation flows are 
-absent~\cite{Marenduzzo:2006a,Marenduzzo:2006b} and the response of the network
-is indeed oscillatory irrespective of the boundary conditions used at the wall. 
-We also note that, if the boundaries are such that
-the director field pattern sticks to the shear planes, then our analysis
-would apply and be realistic for arbitrarily low shear as well.
-
-The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
-the deviatoric stress tensor $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
-For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
-$a=0.35, b=0.95$ describes the data to a very good approximation. 
-Hence, the degree of shear-thinning is remarkably small in BPII, in
-the range of shear rates which we have explored here.
-
-For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ the network of disclination lines
-breaks up completely, and oscillations in the stress signal are absent.
-In this regime of shear rates, to which we refer as BPII-2, the 
-blue phase dissolves into a simple cholesteric liquid 
-crystal. The helical axis is along the vorticity direction and
-the flow excites travelling helical waves, such as those 
-predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
-the vorticity axis.
-
-At shear rates $7.81\e{-4}\lesssim\gd \lesssim 9.37\e{-4}$ the helix changes orientation.
-The helical axis is now oriented along the gradient direction with
-the liquid crystal flowing in 'nematic planes'. For even higher shear rates
-the system undergoes a transition to a flow-aligned nematic state.
-
-In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
+The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
+in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
+absence at very low shear rate of a permeation mode is probably due to our
+choice of boundary conditions.}
+
+%{We note that permeation flows, which would
+%{\em not} lead to oscillatory response, are effectively disallowed by the combination
+%of the geometry and boundary condition we consider. This is
+%because the periodic boundary conditions impose macroscopic distortion
+%of the network, and are effective analogous to infinitely distant
+%but anchored wall along the $xy$ plane.
+%This means that  our results are effectively valid for sheared bulk 
+%BPs which are anchored at the walls. Furthermore, permeation flows are 
+%only restricted to very slow flows, and are unstable for intermediate 
+%and high flows, where the response of the network depends less on
+%the details of the boundary conditions used at 
+%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}
+
+\rev{We observe another configuration, lying between those with periodically recurring conformations 
+and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
+In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
+The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
+and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.}
+
+\rev{The picture at the bottom of Fig. \ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
+the flow-aligned nematic state. We will address these regimes in section \ref{cholflow}, as they are common to BPI and BPII.}
+
+\rev{The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
+to the tumbling motion of the local director field, which is absent in the Grandjean texture.
+For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.}
+
+In the next sections we investigate the BPII-1 flow regime in more detail 
 by looking at the kinetics of the disclination network.
 
-\subsubsection{Regime BPII-1: low and intermediate shear rates}
-
-Fig. \ref{bp2-med} shows the disclination network in shear flow as
-it undergoes an affine transformation in the BPII-1 flow regime. 
+\subsubsection{Regime BPII-1: low and intermediate shear rates }
+
+Fig. \ref{bp2-1-disc} shows the disclination network in shear flow as
+it undergoes homogeneous shearing in the BPII-1 flow regime. 
 The disclination lines break up and 
@@
 period needed for a pattern to break up and 
-reform along the flow direction is $\tau_F=1/\gd$.
+reform along the flow direction is $\tau_F = 1/\gd$.
+
+The general appearance of the flowing network is, apart from the homogeneous distortion,
+very close to that of the quiescent blue phase at equilibrium. This holds for all
+shear rates in the BPII-1 regime.
+
+Interestingly, while being displaced with the flow the entire network moves 
+along the vorticity direction (z) as well.  A similar behaviour has been recently 
+observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
+In contrast to the stick-slip motion that has been reported there,
+in this case the movement is steady. 
+The vorticity motion occurs in such a way that the positions of breakup and reconnection 
+point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
+to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
+is $\tau_V=6\tau_F$, 
+i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
+for the network to move one unit cell along $z$ (the vorticity direction).
+
+Fig. \ref{bp2-1-velo} shows the disclination network in the middle of a breakup-reconnection
+cycle, with superimposed velocity vectors.
+The plot shows the secondary velocity components, obtained by projecting the
+velocity onto a plane perpendicular to the flow direction.
+This gets rid of the dominating velocity component along the flow, $v_x$, and 
+allows to visualise the patterns in the 
+two much smaller components $v_y$ and $v_z$.
+The magnitude of the secondary components is typically in the range of a 
+few percent of the primary flow component, for the shear rates 
+and system sizes simulated here.
 
 \begin{figure}[htpb]
-\includegraphics[width=0.41\textwidth]{disc-160k_run902.png}
-\includegraphics[width=0.41\textwidth]{disc-164k_run902.png}
-\includegraphics[width=0.41\textwidth]{disc-165k_run902.png}
+\center
+\includegraphics[width=0.35\textwidth]{disc-160k_run902r.png}
+\includegraphics[width=0.35\textwidth]{disc-164k_run902r.png}
+\includegraphics[width=0.35\textwidth]{disc-165k_run902r.png}
 \caption{Disclination network of BPII in shear flow: 
@@
 at $\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity 
-gradient is oriented along the vertical direction, whereas the 
-horizontal direction is the flow direction. Lees-Edwards boundary 
-conditions have been imposed in such a way that network moves to the 
+gradient is oriented along the vertical direction (y), whereas the 
+horizontal direction (x) is the flow direction. Lees-Edwards boundary 
+conditions have been imposed in such a way that the network moves to the 
 right in the upper half and to the left in the lower half of the picture.}
-\label{bp2-med}
+\label{bp2-1-disc}
 \end{figure}
@@
 it takes the network to reconnect along the flow direction.}
-\label{bp2-velo}
+\label{bp2-1-velo}
 \end{figure*}
-
-The general appearance of the flowing network is, apart from the affine transformation, 
-very close to that of the quiescent blue phase at equilibrium. This holds for all
-shear rates in the BPII-1 regime.
-
-Interestingly, while being displaced with the flow the entire network moves 
-along the vorticity direction as well.  A similar behaviour has been recently 
-observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
-In contrast to the stick-slip motion that has been reported there,
-in this case the movement is steady. 
-The vorticity motion occurs in such a way that the positions of breakup and reconnection 
-point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
-to travel along the z-direction. The periodicity of the motion along the vorticity direction
-is $\tau_V=6\tau_F$, 
-i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
-for the network to move one unit cell along $z$ (the vorticity direction).
-
-Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
-cycle, with superimposed velocity vectors.
-The plot shows the secondary velocity components, obtained by projecting the
-velocity onto a plane perpendicular to the flow direction.
-This gets rid of the dominating velocity component along the flow, $v_x$, and 
-allows to visualise the patterns in the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the secondary components is typically in the range of a 
-few percent of the primary flow component, for the shear rates simulated here.
 
@@
 and the helicity can be gained by time-averaging over individual cycles.
-Tab. \ref{tab1} gives minima, maxima, averages and standard deviations 
+
+Table \ref{tab1} in the Appendix 
+gives minima, maxima, averages and standard deviations 
 of the velocity components.
 All values for the two runs with inverted helicity are identical apart 
-from a change of sign in the z-components.
-There is only a slight imbalance between the maximum and minimum velocities 
-in z-direction, suggesting that the flow of the network along the vorticity 
-direction is permeative.
-
-\subsubsection{Regime BPII-2: high shear rates}
-
-If the flow rate exceeds a certain critical value BPII does not undergo the 
-affine transformation  above and adopts another mode of flow. 
-Fig. \ref{bp2-high} shows a sequence of cuts through the director 
-field at high shear rate. The entire blue phase network has broken up 
-into a simple cholesteric helix and flows as a travelling 
-helical wave with the helical axis oriented along the vorticity direction.
-The state is translationally invariant in flow and gradient direction.
-and is the preferred mode of flow of a cholesteric 
-liquid crystal in vorticity mode at low Ericksen numbers, i.e. below 
-the uncoiling transition to a nematic, flow-aligned state at even 
-higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
-
-In our scenario the break-up occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
-corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
-parameters for BPII ($K=0.02, \eta=0.666$, Tab. \ref{tab1} and $p_0/2=16$).
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
-$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. 
-The configuration is a helical wave that travels along the direction of vorticity.
-It is translationally invariant along the gradient and flow direction.}
-\label{bp2-high}
-\end{figure}
-
-We observe another configuration between the travelling helical wave and the 
-flow-aligned nematic state at shear rates $7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$, 
-where the helical axis is now along the flow gradient direction (the so-called
-Grandjean texture).
-We will address this in section \ref{cholflow}, as it is common to BPI and BPII.
+from a change of sign in the $z$-components.
+There is only a slight imbalance in magnitude 
+between the maximum and minimum velocities 
+along the $z$-direction. Therefore the averaged flow along the vorticity
+direction, leading to network motion, is very small, and much smaller than
+the local velocity flow. This suggests that the
+flow of the network along the vorticity 
+direction is permeative (i.e. the fluid moves but the network as a whole 
+moves much less).
 
@@
 
-BPI has body-centered symmetry and, unlike BPII, the disclination lines
-characterising its equilibrium structure are well separated and do not intersect.
+BPI has body-centred cubic symmetry and, unlike BPII, the disclination lines
+characterising its equilibrium structure are well separated and do not 
+intersect. (The quiescent state resembles the first frame in Fig.~\ref{bp1-1-disc}
+below.)
 We believe that this topological difference is responsible for most of
@@
 
-Fig. \ref{bp1-rheo} shows the apparent viscosity $\eta_{app}$ versus time and 
+Fig. \ref{bp1-rheo} shows the ratio between the 
+apparent viscosity $\eta_{app}$ and $\eta$ versus time and 
 strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
@@
 the shear rate range we considered. 
-Throughout the flow range we studies,
+Through much of the flow range we studied,
 the average shear stress versus shear rate can be fitted reasonably 
@@
 a critical value -- close to this breakup point the oscillations in the 
-deviatoric stress show an increasingly complex
+shear stress show an increasingly complex
 pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
-In regime BPI-3 $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ 
+In regime BPI-3 ($6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$) 
 we observe that the BPI network transforms under shear into
@@
 double-twist rolls.
-Finally, for shear rates $9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ we observe the same cholesteric 
-configuration with the helical axis along the gradient direction as for BPII (see
+Finally, for shear rates such that
+$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
+configuration with the helical axis along the gradient direction as 
+reported above for BPII-2 (see
 section \ref{cholflow}).
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI versus time (top) 
-and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
-Depending on the configuration in steady shear flow three different regimes 
-can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
-steady flow with recurring patterns (BPI-2, open circles) and 
-breakup of the disclination network (BPI-3, solid circle). 
+\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
+\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
+\rev{Depending on the configuration in steady shear flow five different regimes 
+can be identified: amorphous BP network with yield stress (BPI-1, solid squares), 
+steady flow with periodically recurring patterns (BPI-2, open squares), 
+breakup of the disclination network into amorphous state (BPI-3, open circles)
+and metastable configurations (BPI-3, solid circles), 
+cholesteric helix with axis in gradient direction (BPI-4, open triangles) 
+and flow-aligned nematic state (BPI-5, solid triangles). 
 The error bars represent the minimum and maximum shear stress 
-during one cycle. For clarity the individual curves have been shifted along
-the time and strain axis.}
+during one cycle in regime BPI-2.}}
+\label{bp1-flowcurve}
+\end{figure}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_b.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_c.pdf}
+\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
+and strain (bottom).}
 \label{bp1-rheo}
@@
 
-The rheological response of BPI at low shear rate $\gd\lesssim1.95\e{-5}$ 
+The rheological response of BPI at low shear rate, $\gd\lesssim1.95\e{-5}$
+(${\it Er} \lesssim 0.33$), 
 is strikingly different from that of BPII and appears to show a
 yield stress. An explanation for this behaviour can be found by 
-looking more closely at the average deviatoric stress (where the
+looking more closely at the average shear stress (where the
 background viscosity contribution has been subtracted) and 
@@
 to become positive again later on, forming a complete cycle.
-The amplitude of the excursions in these stress oscillations and
+The amplitude of the excursions in these stress oscillations, and
 the presence of a part of the period where the thermodynamic
-contribution to the stress is negative suggest that the
-BPI network is subject to large forces which eventually 
-cause the flow-induced collapse to the amorphous state with
-an apparent yield stress. We should underscore that this,
-low shear, regime, might be brought about by the boundary
-conditions, which in practice lead to the network sticking to
-the Lees-Edwards plane. With walls and free boundary conditions,
-one may imagine that, for very low wall velocity, 
-permeative flows may be possible which leave the BPI 
-structure almost unchanged, as in cholesterics sheared
+contribution to the stress is negative, suggest that the
+BPI network is subject to large forces. Seemingly, these eventually
+cause the flow-induced collapse to an amorphous network state with
+an apparent yield stress. We should underscore that this
+low shear regime might depend on our choice of boundary
+conditions which in practice are equivalent to having infinitely
+distant walls to which the network is anchored (see discussion
+in the Appendix).
+With free boundary conditions and for very low wall velocity, 
+permeative flows might instead allow for
+some slip of the BPI network, as in cholesterics sheared
 along their axis~\cite{Marenduzzo:2006b}.
 
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
+\caption{Thermodynamic shear stress $\Pi_{xy}$ versus strain near the transition between regime
+BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
+and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
+where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
+throughout positive, but exhibits very large fluctuations during one cycle, 
+which are absent in regime BPI-2 at higher shear rates than shown here.}
+\label{bp1-1_bp1-2}
+\end{figure}
+
 \begin{figure*}[htpb]
-\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115.png}\\
-\includegraphics[width=0.245\textwidth]{disc-xy-400k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-700k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-750k_run1115.png}
-\includegraphics[width=0.245\textwidth]{disc-xy-2200k_run1115.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-10k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-180k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-200k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-210k_run1115r.png}\\
+\includegraphics[width=0.245\textwidth]{disc-xy-1650k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-1800k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2000k_run1115r.png}
+\includegraphics[width=0.245\textwidth]{disc-xy-2400k_run1115r.png}
 \caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$: 
 Depicted is the transition from the quiescent state to flow-induced, intertwined 
-helices that undergo a recurring affine transformation. The top row
+helices that undergo a recurring structural transformation. The top row
 shows a section of one unit cell for early times 
@@
 pictures on the far left, centre left and centre right 
-show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
+show the situation at later time steps $t=1.65, 1.8$ and $2.0\e{6}$,
 which the system passes for several cycles before it ends up in
-an amorphous state (bottom row far right at time step $t=2.2\e{6}$).}
-\label{bp1-low}
+an amorphous state (bottom row far right, at time step $t=2.4\e{6}$).}
+\label{bp1-1-disc}
 \end{figure*}
 
-\if{The negative branch in the deviatoric stress is connected
+%%%  %%%%%%%%%%%%%
+\ifold{The negative branch in the shear stress is connected
 to a local maximum followed by a local minimum in the 
@@
 
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{Deviatoric stress $\Pi_{xy}$ versus strain at the transition between regime
-BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
-and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
-where the stress goes negative (Fig. \ref{bp1-fe-yield}) the stress is 
-throughout positive, but exhibits very large fluctuations during one cycle, 
-which are absent in regime BPI-2 at higher shear rates.}
-\label{bp1-1_bp1-2}
-\end{figure}
-
 To gain further insight into the flow-induced deformation, and eventual
 breakdown, of BPI, it is instructive to follow the dynamics of 
-the network under a slow flow -- this is depicted in Fig. \ref{bp1-low}.
+the network under a slow flow -- this is depicted in Fig. \ref{bp1-1-disc}.
 Three different stages can be distinguished. 
@@
 Consequently, the network adopts a flow-induced conformation which consists 
-of intertwined helices that stretch during the affine transformation.
+of intertwined helices that stretch during the shear transformation.
 This helical conformation that emerges already at strains $\gamma\simeq1$ 
-is shown in Fig. \ref{bp1-low} (top row).
+is shown in Fig. \ref{bp1-1-disc} (top row).
 At this point the original BPI has already been significantly deformed.
@@
 network along the vorticity direction at any stage of the cycle.
-
-This mode of flow proves unstable as after a few cycles 
+(This contrasts with the BPII-2 results reported earlier.)
+
+This mode of flow proves unstable, as after a few cycles 
 distortions appear which lead to further destabilisation -- presumably in
 view of the large stress fluctuations discussed above. 
-Finally, the network loses order and 
-transforms into an amorphous state with almost constant
-yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.
+Finally, the system loses order and 
+transforms into an amorphous network state with almost constant
+stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ LBU.
 If the shear is stopped, this flow-induced amorphous state
-gets stuck into a metastable phase, and cannot
+remains arrested and metastable, and cannot
 find its way back to the original BPI structure.
 
-\subsubsection{Regime BPI-2: intermediate shear rates}
-
-Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$) and 
-at slightly larger shear rates $3.91\e{-5}\lesssim\gd\lesssim 4.687\e{-4}$
-and Ericksen numbers $0.67\lesssim{\it Er}\lesssim8$ 
+\subsubsection{Regime BPI-2: intermediate shear rates }
+
+Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$; $ {\it Er}\lesssim 0.33$) but 
+at slightly larger shear rates ($3.91\e{-5}\lesssim\gd\lesssim 4.687\e{-4}$; $0.67\lesssim{\it Er}\lesssim8$)
 lies another region where the network flows with periodically 
@@
 is clear when looking at Fig. \ref{bp1-1_bp1-2} 
-which shows the deviatoric stress versus total strain. 
+which shows the shear stress versus total strain. 
 A qualitative difference between these two is the absence in BPI-2 of the 
 large stress fluctuations that occur during the early cycles in regime BPI-1.
-[Recall that for even lower $\gd$ the deviatoric thermodynamic stress becomes 
+(Recall that for even lower $\gd$ the thermodynamic shear stress becomes 
 temporarily negative (Fig. \ref{bp1-fe-yield}), which 
 caused destabilisation of the periodic network and led to the 
-amorphous configuration in the steady state.] Hence, an explanation
+amorphous configuration in the steady state.)
+Hence, an explanation
 for the existence of the regular BPI-2 oscillations 
@@
 steady state at $\gd=1.56\e{-4}$ and time steps 
-$t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
-During every cycle the network is displaced also along the gradient and vorticity direction, 
-but the magnitude and direction of the offset does not appear to obey a 
-simple rule.}
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
+During every cycle the network is displaced also along the gradient and vorticity direction.}
 \label{bp1-med}
+\end{figure*}
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-364k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-366k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-368k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-370k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-372k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-374k_run914r.png}
+\caption{Velocity patterns in BPI for positive helicity: 
+The pictures show a snapshot of the periodically recurring patterns in the 
+secondary velocity components $(v_y,v_z)$ at $\gd=1.56\e{-4}$ and time steps 
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$, i.e. for the same shear rate
+and time steps as in Fig.\ref{bp1-med}. The colour code gives the magnitude and sign 
+of the $z$-component. For negative helicity the sense of motion of the 
+network and the secondary velocity components are inversed just as in Fig. \ref{bp2-velo}. 
+Note that the viewing direction has been adapted in
+each snapshot to give the clearest view on the band-like pattern between
+the disclination lines.}
+\label{bp1-velo}
 \end{figure*}
@@
 BPI-2 does not resemble the equilibrium configuration
-undergoing an affine transformation. It features rather intricate, flow-induced 
+undergoing a homogeneous topology-conserving 
+transformation. It features rather intricate, flow-induced 
 conformations. Surprisingly, if the shear is switched off in
-the BPI-2 regime at some point during an oscillations, 
-most of the times the flow-induced configuration 
+the BPI-2 regime at some point during an oscillation, 
+the flow-induced configuration usually
 reverts to a perfect quiescent BPI -- in stark contrast with
@@
 
-The period after which network conformations recur is not
-easy to work out for BPI. Along the flow direction a relation of the form
-$\tau_F=n/\gd$ holds with $n=1,2,4$ for $\gd=3.91\e{-5},
-7.81\e{-5},1.563\e{-4}$ for example. The analysis is complicated by the fact
-that for $n>1$ thermodynamically equivalent, mirror-image 
-conformations arise after shorter times. For instance at $\gd=7.81\e{-5}$
-the conformations at times $t$ and $t+1/\gd$ are mirror-images, whereas
-those at $t$ and $t+2/\gd$ coalesce in flow direction.
-
-However, there is no definite recurrence period $\tau_V$ 
-in the vorticity direction. The direction of the displacement
-perpendicular to the flow direction is not along a coordinate
-axis. This direction, and the magnitude of the secondary
-network movement depends sensitively on the imposed shear rate. 
-This suggests that BPI-2 constitutes an instance of 
-'rheochaos'~\cite{rheochaos}.
+Finding the recurrence period for BPI is not simple, as
+this depends subtly on the applied shear rate. The analysis is complex for
+a number of reasons. 
+First, the recurrence period for the stress is not always the same as
+that of the network configuration. 
+For instance, at $\gd=3.9062500\e{-5}$, it takes $2/\gd$ for one
+ stress cycle to complete, but $4/\gd$ for the network to recur --
+this is because the configurations at times $t$ and $t+2/\gd$ are 
+mirror-images. This is by no means generic though: for example
+at $\gd=4.8828125\e{-5}$  it takes $8/\gd$ for one
+stress cycle to complete, but $16/\gd$ for the network to recur. 
+For some higher shear rates, the recurrence period of stress
+and network conformations are instead the same.
+Second, there exist isolated values (or perhaps small ranges) of shear
+rates for which the recurrence period is much longer and far more irregular,
+with seemingly intermittent dynamics (e.g. at $\gd=1.3671875\e{-4}$ the
+network configuration repeats itself after $6/\gd$, $12/\gd$, $22/\gd$).
+Taken together, all this evidence very much suggests that 
+the BPI-2 range includes at least some instances of ``rheochaos''~\cite{rheochaos,Cates:2002}.
 
@@
 
-\begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
-The pictures show a snapshot of the periodically recurring patterns in the 
-secondary velocity components $(0,v_y,v_z)$, i.e. the component $v_x$ in flow direction 
-has been projected out. The colour code gives the magnitude and sign 
-of the z-component. The vertical and horizontal direction are the gradient and 
-vorticity direction, respectively, with the flow into/out of the plane in the 
-upper/lower half of the images.}
-\label{bp1-velo}
-\end{figure*}
-
-In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
-we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
-The spectrum is shown in Fig. \ref{bp1-spectrum}.
-
-There are two aspects which are worth mentioning. First, we observe a strong contribution of 
-the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
-However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
-significantly larger. 
-Second, more surprisingly, the frequency of the fundamental mode is the same 
-for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
-in contrast to what is observed at lower flow rates. By doubling the flow rate from 
-$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
-moves to double the frequency. This means that the fundamental mode at flow rate $\gd=7.81\e{-5}$ 
-coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on. 
-At higher shear rates in regime BPI-3 this 'lock-in' of flow modes disappears again.
-
-Interestingly the lock-in has only a minor effect on the flow velocities themselves.
-The average velocities in Tab. \ref{tab1} reveal that only the magnitude of some of the secondary 
-flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
-The latter is directly linked to the imposed shear rate.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture shows 
-spectral components and the 'lock-in' of flow modes, 
-visible in the fixed peak position for certain flow rates.  
-The inset shows a zoom of the same data over 
-a portion of the frequency range.
-The frequency is given in inverse time, in simulation units. TRUE?
-OH: I WOULD SAY NO AS IT IS IN UNITS OF 2$\pi$ (RADIANS?).}
-\label{bp1-spectrum}
-\end{figure}
-
-The intricacy and periodic recurrence of the flow-induced conformations 
+%%% No differences in revison between different system sizes %%%
+\ifold{The intricacy and periodic recurrence of the flow-induced conformations 
 in BPI raises the question: how robust are these results to variations in 
@@
 
-Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
-deviatoric stress $\Pi_{xy}$ of to different runs, one consisting of 8 
-unit cells and the other one with 64 unit cells.
-We observe that there are deviations in both the visual appearance of the 
-network as well as in the time-dependence of the deviatoric stress.
-Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
-to reach the steady state with its flow-induced recurring disclination network. 
-These results indicate that we cannot know with certainty if the individual 
-flow-induced conformations that we find for 64 unit cells are not affected 
-by the imposed periodicity -- quite possibly the details of the oscillations
-would actually differ in an even large simulation. 
-Nevertheless, the comparison suggests that 
-the recurrence period and the complexity of the network in flow are  
-generic features of BPI in this regime of flow.
-
 \begin{figure}[htpb]
-\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
-\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Disclination network and deviatoric stress $\Pi_{xy}$ in BPI: 
+%\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
+%\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
+\caption{Disclination network and shear stress $\Pi_{xy}$ in BPI: 
 The dark (red) network in top picture shows a snapshot of run that consists 
@@
 central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
-The bottom picture shows the deviatoric stress over time. [Interestingly, 
+The bottom picture shows the shear stress over time. (Interestingly, 
 an intermediate sized run with $3^3=27$ unit cells led to results that 
-were virtually identical to the smaller run with 8 unit cells.] 
+were virtually identical to the smaller run with 8 unit cells.) 
 The shear rate was $\gd=7.81\e{-5}$.
@@
 
-\subsubsection{Regime BPI-3: high shear rates}
-
-At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$, 
+Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
+shear stress $\Pi_{xy}$ of two different runs, one consisting of 8 
+unit cells and the other one with 64 unit cells.
+We observe that there are deviations in the visual appearance of the 
+network as well as in the time-dependence of the shear stress.
+Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
+to reach the steady state with its flow-induced recurring disclination network. 
+These results indicate that we cannot know with certainty whether the individual 
+flow-induced conformations that we find for 64 unit cells are affected 
+by the imposed periodicity -- quite possibly the details of the oscillations
+would differ in an even larger simulation. 
+Nevertheless, the comparison suggests that 
+the recurrence period and the complexity of the network in flow are  
+generic features of BPI in this regime of flow.
+\fi
+
+\subsubsection{Regime BPI-3: high shear rates }
+
+%%% Absent in revison - new 2D conformation is persumably metastable %%%
+\ifold At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
 we observe a break-up of the flow-induced 
@@
 in contrast to BPII, where the configuration is a travelling helical 
-wave (viz. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
+wave (cf. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
 two-dimensional, translationally invariant state, which is shown in 
-Fig. \ref{bp1-high}. The snapshots visualise the affine 
-transformation of simple shear flow in the gradient-flow plane.
+Fig. \ref{bp1-high}. 
 There are distinctive regions where the director field is 
@@
 about the same window as that of transition between BPII-1 and BPII-2. 
-
-\begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
+\fi
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{dir3d+y-450k_run916r.png}
+\includegraphics[width=0.32\textwidth]{dir3d+y-510k_run916r.png}
+\includegraphics[width=0.32\textwidth]{dir3d+y-800k_run916r.png}
 \caption{Director field of BPI at high shear rate: The pictures show a region 
-of $L_x\times L_y \times L_Z= 32^3$ lattice sites with the flow/gradient 
+of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
 direction oriented along the horizontal/vertical axis. 
 The colour code gives the magnitude of the z-component of the director field. 
-The data is translationally invariant along the vorticity direction.}
+The data show that the structure is quasi-2D, with rolling double twist
+patterns in the $xy$ plane, and an approximate translational symmetry 
+along the vorticity direction.}
 \label{bp1-high}
-\end{figure}
+\end{figure*}
 
@@
 
-At shear rates beyond the regimes BPI-3 and BPII-2 and below the transition to a 
-flow-aligned nematic state we found another regime where 
+At shear rates beyond the regimes BPI-3 and BPII-2, but below the transition to a 
+flow-aligned nematic state at still higher shear rates, we found another regime where 
 both blue phases adopt the same configuration in steady shear flow, 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
-\caption{Flow induced cholesteric state: The orientation of the helix is 
+\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
+\caption{Flow induced cholesteric state. The orientation of the helix is 
 along the gradient direction. The director field is quasi-static during 
-the flow, which occurs in 'nematic layers'. For clarity the picture shows 
-the director field along the y-direction for selected sites on the x-z-plane.}
+the flow, which occurs in ``nematic layers''. Colour coding indicates the 
+$x$ component of the director field (see colorbar on the right).
+For clarity only selected sites are shown along the $y$-direction (along which
+the structure is translationally invariant).}
 \label{fig:cholflow}
@@
 
-In summary, our work constitutes the first simulation of bulk flow behaviour 
-of cubic BPs in simple shear flow. Although experimental evidence to support
-our results is currently not available, we believe this work can shed some light 
-into the flow properties of complex liquid-crystalline phases.
-
-We were able to characterise the rheology of cubic BPs and identified 
-two different flow regimes for BPII and three different ones in case of BPI. 
-Below Ericksen numbers $Er\simeq3$ BPII exhibits weak shear-thinning and 
-obeys a power-law.  The BPII disclination network breaks up and reconnects 
+In summary, our work constitutes the first large scale simulation of bulk flow behaviour 
+of cubic BPs in simple shear flow.
+
+We were able to characterise the rheology of cubic BPs, and identified 
+two different flow regimes for BPII and three different ones for BPI. 
+Below an Ericksen number of $Er\simeq3$, BPII exhibits weak shear-thinning and 
+obeys a power-law flow curve with exponent close to unity.  
+The BPII disclination network breaks up and reconnects 
 in the flow, which leads to a periodically recurring dependence 
-of the deviatoric shear stress. The flow-induced conformation looks generally
+of the shear stress. The flow-induced conformation looks generally
 very similar to the quiescent network at equilibrium. 
-While being affinely transformed due to the shear flow
-the disclination network moves in the vorticity direction. The sense of
+While being homogeneously transformed due to the shear flow
+the disclination network moves steadily in the vorticity direction
+apparently by a permeation mechanism. The sense of
 motion is directly linked to the helicity of the underlying cholesteric phase.
 
-At larger Ericksen numbers $4\lesssim Er\lesssim 10$ the flowing BPII network 
+At larger Ericksen numbers, $4\lesssim Er\lesssim 10$, the flowing BPII network 
 breaks up in a two-step manner. First a simple cholesteric helix forms with the helical axis 
@@
 as predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}.
-Upon increasing the shear, the helix rotates and this time orients along the
-gradient direction, where it can support larger flows due to the translational
-symmetry along the flow direction. Still larger flows break this residual
-cholesteric order and leave a standard flow-aligned nematic staete.
+Upon increasing the shear, the helix rotates and orients along the
+gradient direction, where it can support larger flow rates due to the translational
+symmetry along the flow direction. Still larger flow rates break this residual
+cholesteric order and leave a standard flow-aligned nematic state,
+at the highest shear rate studied.
 
 Interestingly, BPI shows a flow behaviour that is very different to that of BPII.
-This is a direct consequence of the topological differences in both disclination 
-network. Below Ericksen numbers $Er\lesssim0.4$ the BPI network cannot both flow and 
-retain a regular appearance. It eventually collapses into an amorphous network and 
-features yield-stress behaviour. The reason is that shortly after the onset of the shearing 
+This is a direct consequence of the topological differences between their disclination 
+networks. Below an Ericksen number of $Er\sim 0.4$, the BPI network cannot both flow and 
+retain a regular appearance. It eventually reorganises into an amorphous network that
+features yield-stress behaviour. The apparent reason is that shortly after the onset of the shearing 
 very large stress fluctuations occur. At the lowest shear rates
-the thermodynamic contribution to the deviatoric (shear) stress is temporarily 
-negative during one cycle. These fluctuations destabilise the network and 
+the thermodynamic contribution to the shear stress becomes temporarily 
+negative during each cycle. These fluctuations seem to destabilise the network and 
 eventually trigger the transition into an amorphous state with a residual yield
@@
 predicted. 
-The offset of the network after each oscillation is along a direction which,
-while being a multiple of the unit cell size along all three coordinate
-axis, depends on system size and shear rate in a seemingly chaotic fashion.
+The oscillatory dynamics is in some cases intermittent, and the
+recurrence period for the network conformations and stress patterns
+is not given by a simple formula as for BPII.
 Despite their complex appearance, the flow-induced conformations are 
-topologically connected to the quiescent BPI and after switching off of the 
+topologically connected to the quiescent BPI and after switching off the 
 shear flow a defect-free blue phase reforms. It is tempting to
-interpret all these finding as an instance of 
-deterministic rheological chaos.
+interpret the unsteady oscillations seen in this regime as an instance of 
+deterministic rheological chaos~\cite{fielding, Cates:2002}.
 
@@
 quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
-Then, at $16\lesssim Er \lesssim 22$, it forms a cholestric helix along the flow gradient 
+Then, at $16\lesssim Er \lesssim 22$, it forms a cholesteric helix along the flow gradient 
 direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
 is a flow-aligned nematic state.
+Although experimental evidence to support
+our results is currently not available, we hope this work will inspire such experiments, and believe it can shed some light 
+on the flow properties of complex liquid-crystalline phases.
 
@@
 % If you have acknowledgments, this puts in the proper section head.
-\begin{acknowledgments}
-We acknowledge support by EPSRC grant nos. EP/E045316 and EP/E030173, 
-the MAPPER EU-FP7 project (grant no. RI-261507) and computing time on HECToR.
-We thank Peter J. Collings for very stimulating discussions. 
+\section*{Acknowledgments}
+
+This work was granted access to the HPC resource of CSC, Finland, made available 
+within DECI by the PRACE-2IP, funded by the European Community's 7th Framework 
+Program under grand agreement No. RI-283493. We also acknowledge support by 
+EPSRC grant nos. EP/E045316 and EP/E030173 and the MAPPER EU-FP7 project 
+(grant no. RI-261507).We thank Peter J. Collings for stimulating discussions. 
 MEC holds a Royal Society Research Professorship.
-\end{acknowledgments}
-
-% Create the reference section using BibTeX:
-\bibliography{bprheo}
+
+\appendix
+\section*{Appendix}
+
+In this Appendix we give some more details on our simulations,
+including the relationship between boundary conditions and
+permeation flow.
+
+Table \ref{tab1} lists the parameters chosen for our runs (values of
+the reduced temperature and reduced chirality are given in the caption),
+together with the minima, maxima, and standard deviations of the velocity
+field in the three directions (discussed further in the main text). 
+The first column also describes what flow regime each simulation
+leads to in steady state.
+
+\begin{table*}[htpb]
+\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
+\hline
+& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+\hline
+BPI & & & & $\times 10^5$\\
+\hline
+BPI-1 (AN) &0.24 &0.02 & 0.1388 & & & & & & & & & \\
+BPI-1 (AN) &0.49 &0.04 & 0.1388 & & & & & & & & & \\
+BPI-1 (AN) &0.98 &0.08 & 0.1388 & & & & & & & & & \\
+BPI-1 (AN) &1.95 &0.17 & 0.1388 & & & & & & & & & \\
+%BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
+%BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
+%BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
+%BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+\hline
+BPI-2 (PRC) &3.91 &0.33 & 0.1388 & & & & & & & & & \\
+BPI-2 (PRC) &7.81 &0.67 & 0.1388 & & & & & & & & & \\
+BPI-2 (PRC) &15.63 &1.33 & 0.1388 & & & & & & & & & \\
+BPI-2 (PRC) &15.63 &1.33 & -0.1388 & & & & & & & & & \\
+BPI-2 (PRC) &23.44 &2.00 & 0.1388 & & & & & & & & & \\
+%BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
+%BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
+%BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
+%BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
+%BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+\hline
+BPI-3 (AMS) &31.25 &2.66 & 0.1388 & & & & & & & & & \\
+BPI-3 (AMS) &46.87 &4.00 & 0.1388 & & & & & & & & & \\
+BPI-3 (AMS) &54.69 &4.67 & 0.1388 & & & & & & & & & \\
+BPI-3 (AMS) &62.50 &5.33 & 0.1388 & & & & & & & & & \\
+BPI-3 (AMS) &78.13 &6.01 & 0.1388 & & & & & & & & & \\
+%BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+\hline
+BPI-4 (GJ) &93.75 &8.00 & 0.1388 & & & & & & & & & \\
+BPI-4 (GJ) &125.0 &10.66 & 0.1388 & & & & & & & & & \\
+BPI-4 (GJ) &187.5 &16.00 & 0.1388 & & & & & & & & & \\
+%GJT (BPI-4)& 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+\hline
+BPI-5 (FAN) & 250.0 &21.33 & 0.1388 & & & & & & & & & \\
+BPI-5 (FAN) & 375.0 &32.00 & 0.1388 & & & & & & & & & \\
+%FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+\hline
+
+BPII  & & & & $\times 10^5$ \\
+\hline
+BPII-1 (PRC) &0.24 &0.02 & 0.1963 &-7.8 &7.7 &0.1 &-0.2 &0.2 &0.1 &-0.3 &0.3 &0.2 \\
+BPII-1 (PRC) &0.49 &0.04 & 0.1963 &-15.5 &15.4 &0.2 &-0.3 &0.4 &0.2 &-0.5 &0.4 &0.2 \\
+BPII-1 (PRC) &0.98 &0.08 & 0.1963 &-31.0 &30.8 &0.3 &-0.6 &0.6 &0.4 &-0.9 &0.8 &0.4 \\
+BPII-1 (PRC) &1.95 &0.17 & 0.1963 &-62.0 &61.6 &0.6 &-1.2 &1.2 &0.8 &-1.8 &1.6 &0.8 \\
+BPII-1 (PRC) &3.91 &0.33 & 0.1963 &-123.9 &123.3 &1.2 &-2.1 &2.1 &1.6 &-3.4 &3.0 &1.5 \\
+BPII-1 (PRC) &7.81 &0.67 & 0.1963 &-247.5 &246.8 &2.0 &-3.8 &3.8 &3.0 &-6.5 &5.5 &2.6 \\
+BPII-1 (PRC) &15.63 &1.33 & 0.1963 &-494.3 &493.7 &2.5 &-5.5 &-5.5 &4.0 &\bf{-10.9} &\bf{8.7} &\bf{3.9} \\
+BPII-1 (PRC) &15.63 &1.33 & -0.1963 &-494.3 &493.7 &2.5 &-5.5 &5.5 &4.0 &\bf{-8.7} &\bf{10.9} &\bf{3.9} \\
+BPII-1 (PRC) &31.25 &2.67 & 0.1963 &-987.5 &987.1 &3.5 &-6.9 &6.9 &5.1 &-17.7 &13.1 &5.0 \\
+BPII-1 (PRC) &39.06 &3.33 & 0.1963 &1233.9 &1233.6 &3.7 &-6.9 &6.9 &5.1 &-20.0 &14.5 &5.2 \\
+\hline
+BPII-2 (GJ) &46.87 &4.00 & 0.1963 &-1479.8 &1477.0 &- &- &- &- &-10.0 &9.6  &- \\
+BPII-2 (GJ) &54.69 &4.67 & 0.1963 &-1720.5 &1720.8 &- &- &- &- &-11.5 &11.2 &- \\
+BPII-2 (GJ) &62.5 &5.33 & 0.1963  &-1966.7 &1966.2 &- &- &- &- &-13.3 &12.7 &- \\
+BPII-2 (GJ) &78.13 &6.01 & 0.1963 &-2458.3 &2457.8 &- &- &- &- &-16.9 &15.7 &- \\
+BPII-2 (GJ) &93.75 &8.00 & 0.1963 &-2949.7 &2949.7 &- &- &- &- &-20.3 &18.4 &- \\
+BPII-2 (GJ) &125.0 &10.67 &0.1963 &-3932.8 &3932.8 &- &- &- &- &-26.6 &22.4 &- \\
+\hline
+BPII-3 (FAN) &187.5 &16.00 & 0.1963 &-5906.2  &5906.2 &- &-  &-  &-  &-0.6 &0.4 &- \\
+BPII-3 (FAN) &250.0 &21.33 & 0.1963 &-7874.9  &7874.9 &- &-  &-  &-  &-0.5 &0.3 &- \\
+\hline
+\end{tabular}
+\caption{\rev{Minima, maxima and standard deviation of time-averaged velocity components 
+for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
+shear flow. All velocity values are given in $10^{-5}$ lattice units. The transient dynamics 
+has not been included in the averages. The regimes comprise the formation
+of an amorphous network (BPI-1), periodically recurring conformations with oscillatory response (BPI-2 and BPII-1), 
+amorphous and metastable steady states (BPI-3), a Grandjean texture (BPII-2 and BPI-4),
+and a flow-aligned nematic state (BPI-5 and BPII-3).
+(Last six columns refer to the secondary flow.)}}
+\label{tab1}
+\end{table*}
+
+We close with a technical note on the choice of boundary conditions. 
+In all our runs we used Lees-Edwards boundary conditions: this
+is the sheared equivalence of periodic boundary conditions, for both the 
+velocity and the order parameter field. 
+While extremely useful to simulate bulk flow, we
+note that our boundary conditions impose macroscopic distortion
+of the network, and are effective analogous to infinitely distant
+but anchored wall along the $xy$ plane. In other words, due to
+the periodic boundary conditions along the flow gradient direction, 
+there can be no slip in the network deformation (or strain) at any point. 
+Effectively, this disallows permeation flows, which on the other
+hand require a non-zero velocity field and, simultaneously,
+much lower (ideally zero) perturbations of the blue phase network.
+Consequently our results are effectively valid for sheared bulk 
+BPs which are anchored at the walls. In practice, permeation flows are 
+only restricted to very low flow rates, and are unstable for intermediate 
+and fast flows~\cite{Dupuis:2005}, 
+where the response of the network depends less on
+the details of the boundary conditions used at the wall.
+\newpage
+
+\footnotesize{
+\bibliography{bprheo} %your .bib file
+\bibliographystyle{rsc} %the RSC's .bst file
+}
 
 \end{document}
-%
-% ****** End of file apstemplate.tex ******

### 2013-06-06 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2000 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
+Characteristic bands are visible in Fig. \ref{bp2-1-velo}, which are 
 oriented along the vorticity direction. The direction of the flow
@@
 
-Table \ref{tab1} in the Appendix 
+Table~\ref{tab1} in the Appendix 
 gives minima, maxima, averages and standard deviations 
@@
 between the maximum and minimum velocities 
-along the $z$-direction. Therefore the averaged flow along the vorticity
-direction, leading to network motion, is very small, and much smaller than
+along the $z$-direction. \rev{Therefore the averaged flow along the vorticity
+direction is very small, and much smaller than
 the local velocity flow. This suggests that the
 flow of the network along the vorticity 
-direction is permeative (i.e. the fluid moves but the network as a whole 
-moves much less).
+direction is permeative. This claim is further supported by the characteristic 
+velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
+The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
+whereas the network as a whole moves only in one direction.}
 
@@
 
-Fig. \ref{bp1-rheo} shows the ratio between the 
-apparent viscosity $\eta_{app}$ and $\eta$ versus time and 
-strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
-shows that there is more shear thinning in BPI than in BPII, for
-the shear rate range we considered. 
-Through much of the flow range we studied,
-the average shear stress versus shear rate can be fitted reasonably 
-well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
+\rev{Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
+shear rates as in Fig.~\ref{bp2-flowcurve}. 
+The plot shows that there is more shear thinning in BPI than in BPII, for
+the shear rate range we considered.}
 
@@
 regime BPI-2. 
+\rev{This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
+ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.}
+
 The BPI-2 regime now does not persist to arbitrarily low shear: there the
 oscillations become irregular and seemingly chaotic, and after a few
-cycles the disclination pattern becomes amorphous. We refer to this 
-flow regime as BPI-1. 
+cycles the disclination pattern becomes amorphous.
+\rev{For these very low shear rates the flow curve in Fig. \ref{bp1-flowcurve}
+develops a horizontal branch, which suggest the existence of a yield stress.} 
+We refer to this flow regime as BPI-1.
+
 The BPI-2 regime is also unstable if the shear is increased {\em above} 
-a critical value -- close to this breakup point the oscillations in the 
-shear stress show an increasingly complex
-pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
-In regime BPI-3 ($6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$) 
-we observe that the BPI network transforms under shear into
-a  quasi two-dimensional state consisting of
-double-twist rolls.
+a critical value 
+\rev{in the region of Ericksen numbers $Er \simeq O(1)$
+where viscous and elastic forces are of the same magnitude. 
+Close to this breakup point the network seems to become
+increasingly sensitive towards small perturbations in the flow field
+that cause irregularities in the order parameter structure.
+These distortions couple back to the flow and form a possible
+route towards "rheochaos"~\cite{rheochaos,Cates:2002}.}
+
+We also observe that
+
+regime BPI-3 ($3.125\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $2.67 \lesssim {\it Er} \lesssim 6.01$)
+
 Finally, for shear rates such that
@@
 and strain (bottom).}
-\label{bp1-rheo}
+\label{bp1-appvisc}
 \end{figure}
@@
 
-%%%  %%%%%%%%%%%%%
-\ifold{The negative branch in the shear stress is connected
-to a local maximum followed by a local minimum in the 
-average free energy. Hence, the region of negative stress can be 
-regarded as a phase during which elastic forces exert a 
-'pull' on the network while it tries to reach a more 
-favourable configuration. However, the low flow rate
-causes this unstable intermediate state to be exposed to these
-elastic forces for a relatively long time, long enough to 
-create small distortions which finally break the symmetry of the
-flow-induced configuration.}\fi
-
 To gain further insight into the flow-induced deformation, and eventual
@@
 Taken together, all this evidence very much suggests that 
-the BPI-2 range includes at least some instances of ``rheochaos''~\cite{rheochaos,Cates:2002}.
+the BPI-2 range includes at least some instances of "rheochaos"~\cite{rheochaos,Cates:2002}.
 
@@
 \hline
-BPI-3 (AMS) &31.25 &2.66 & 0.1388 & & & & & & & & & \\
+BPI-3 (AMS) &31.25 &2.67 & 0.1388 & & & & & & & & & \\
 BPI-3 (AMS) &46.87 &4.00 & 0.1388 & & & & & & & & & \\
@@
 BPI-4 (GJ) &93.75 &8.00 & 0.1388 & & & & & & & & & \\
-BPI-4 (GJ) &125.0 &10.66 & 0.1388 & & & & & & & & & \\
+BPI-4 (GJ) &125.0 &10.67 & 0.1388 & & & & & & & & & \\
 BPI-4 (GJ) &187.5 &16.00 & 0.1388 & & & & & & & & & \\

### 2013-06-09 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2002 40876441-3496-0410-a613-a2fca35b2ae2

@@
 The error bars represent the minimum and maximum shear stress 
-during one cycle in regime BPI-2.}}
+during one cycle in regime BPI-2.
+The two dashed lines represent fits to the data in regime BPI-2 with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) and in regime BPI-4 with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}}
 \label{bp1-flowcurve}

### 2013-06-27 trunk/papers/bprheo/bprheo.tex
Update bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2028 40876441-3496-0410-a613-a2fca35b2ae2

@@
 II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
-For larger shear rates, the structure breaks up into a cholesteric helix, lying either along the
-vorticity or the flow gradient direction according to the imposed flow.
+For larger shear rates, the structure breaks up 
+\rev{
+into a Grandjean texture with a 
+cholesteric helix lying along the flow gradient direction according to the imposed flow.
+}
 BPI leads to a very different response. Here, oscillations are only possible for intermediate
 shear rates -- very slow flow leads to a transition into an amorphous network 
-with an apparent yield stress. The flow-induced oscillation regime features a novel arrangement
-of double helical disclinations which rotate due to the shear. Large flow rates can, 
-just as in the BPII case, lead to the rupture of the network and the stabilisation
-of less ordered cholesteric structures. Our results provide the first 
-theoretical investigation of sheared blue phases in large systems, and
-are relevant to understanding the bulk rheology of these materials.}
+with an apparent yield stress. Larger shear rates can make the network sensitive
+to flow-induced transitions to metastable and amorphous states. For even
+larger flow rates a Grandjean texture is observed, just as in the case of BPII.
+At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
+Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
 \vspace{0.5cm}
@@
 They also found that a strong secondary flow appears.
+\rev{
+Early experiments with cholesteric liquid crystals
+showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
+}
 Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
@@
 lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
+\rev{
+These result, however, were derived under the 
+assumption that the molecules rotate only in flow-gradient plane whilst
+the orientation of the cholesteric helix remained unchanged.
+}  
 Rey also studied cholesterics subjected to both steady flow and low frequency
@@
 $v$ and $l$ a typical velocity and length scale. 
-In the present work $l=p_0/2$ was used as this is the approximate size of
+In the present work $l=p_0/2=\pi/q_0$ was used as this is the approximate size of
 the BP unit cell. Likewise, $v$ was taken to be the velocity difference
-across one unit cell, i.e. $v=\dot{\gamma} p_0/2$.
-
-\if{The spectrum $X_\omega$ of the shear stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation according to
-\begin{equation}
-X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
-\label{spectrum}
-\end{equation}
-where we subtract the average stress $\bar{\Pi}_{xy}$ from the signal.
-$N$ was the number of points in the time series, typically one for every 100 LB time steps. }\fi
+across one unit cell, i.e. $v=\dot{\gamma} l$. 
+\rev{
+The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
+}
 
@@
 $\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
-For clarity we classify various flow regimes, two in the case of BPII and three in the 
-case of BPI. 
-They are BPII-1 ($\gd \lesssim 3.125\e{-4}$, corresponding to ${\it Er} \lesssim 2.67$), 
-BPII-2 ($4.687\lesssim \gd\lesssim 6.25\e{-4}$; $4\lesssim {\it Er} \lesssim 5.34$), ),
-BPI-1 ($\gd \lesssim 1.95\e{-5}$; ${\it Er} \lesssim 0.33$), ), BPI-2 ($3.91\e{-5}\lesssim \gd \lesssim 4.687\e{-4}$; $0.67\lesssim {\it Er} \lesssim 8$) and BPI-3 ($6.25\e{-4}\lesssim \gd\lesssim 7.813\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$).
-Beyond these regimes, and prior to the ultimate breakdown into an isotropic phase, 
-both BPI and BPII form common configurations regardless 
-of the initial state. These are a cholesteric helix with helical axis 
-oriented along the direction of the velocity gradient (CHG), 
-and a flow-aligned nematic state (FAN).  
+\rev{
+For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
+case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
+amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
+may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
+GJ and FAN occur regardless of the initial state.
+}
+
+\rev{
+These regimes are characterised by the following flow rates: 
+For Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
+BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
+BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) 
+and for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
+BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
+BPI-3 (AN) ($3.125\e{-4}\lesssim \gd\lesssim 4.69\e{-4}; 2.67 \lesssim {\it Er} \lesssim 4$),
+BPI-4 (GJ and FGJ) ($5.47\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 4.67\lesssim {\it Er} \lesssim 16)$ and
+BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.33\lesssim {\it Er}$).
+}
+
 As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
-\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd$ (normalised to $\eta$) 
-of BPII versus time (top) and strain (bottom). 
-For clarity the individual curves have been shifted along the time and strain axis. 
-The inset in the top panel 
-shows the flow curve $\bar{\Pi}_{xy}(\gd)$ and two different regimes of flow: 
-regular periodic breakup and reconnecting of the disclination lines (BPII-1, open circles), 
-break up of the network into a travelling helical wave (BPII-2, solid circle), 
-cholesteric helix with axis in gradient direction (CHG, open squares)
-and flow-aligned nematic state (FAN, solid square).
-The error bars indicate the maximum and minimum stresses that occur during one cycle.}
-\label{bp2-rheo}
+\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
+\caption{
+\rev{
+Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
+whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
+breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
+The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
+}
+}
+\label{bp2-flowcurve}
+\end{figure}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
+\caption{
+\rev{
+Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
+of BPII versus strain $\gamma = \gd\cdot t$  
+normalised to the background viscosity $\eta$.
+The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
+The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
+}
+} 
+\label{bp2-appvisc}
 \end{figure}
@@
 flow behaviour at all applied shear rates.
-Fig. \ref{bp2-rheo} shows the ratio between the apparent viscosity, 
-defined as 
-$\eta_{app}=\langle \Pi_{xy} \rangle/\gd$, and $\eta$.
+
+Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
+the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
+\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
+power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
+the data to a very good approximation.
+This holds true even in regime BPII-2, where the network breaks up and 
+the liquid crystal adopts a Grandjean structure. The coefficents of the power law fit are
+here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
+remarkably small in BPII, in the range of shear rates which we have explored here.}
+
+\rev{
+Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
+defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
+over total strain.
 A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
 without any additional contribution from the liquid crystal.
-The top picture shows the data over time, whereas the graph at the bottom gives 
-the same data versus accumulated strain.
-For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
-is because of the periodic breakup and reconnecting of the network in shear flow. 
-We refer to this regime as BPII-1. As noted in the Appendix, the
-absence at very low shear rate of a permeation mode is due to our
+The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
+in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
+absence at very low shear rate of a permeation mode is probably due to our
 choice of boundary conditions.
+}
+
 %{We note that permeation flows, which would
@@
 
-The inset to Fig. 1 (top) shows a flow curve, defined as time averages of 
-the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
-For all but the largest shear rates a power law fit $\bar{\Pi}_{xy}=a \gd^b$ with 
-$a=0.35, b=0.95$ describes the data to a very good approximation. 
-Hence, the degree of shear-thinning is remarkably small in BPII, in
-the range of shear rates which we have explored here.
-
-For shear rates $4.68\e{-4}\lesssim\gd\lesssim6.25\e{-4}$ ($4\lesssim {\it Er} \lesssim 5.34$) the network of disclination lines
-breaks up completely, and oscillations in the stress signal are absent.
-In this regime of shear rates, to which we refer as BPII-2, the 
-blue phase dissolves into a simple cholesteric liquid 
-crystal. The helical axis is along the vorticity direction and
-the flow excites travelling helical waves, such as those 
-predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
-the vorticity axis.
-
-At shear rates $7.81\e{-4}\lesssim\gd \lesssim 9.37\e{-4}$ ($6.66\lesssim {\it Er} \lesssim 8$) the helix changes orientation.
-The helical axis is now oriented along the gradient direction with
-the liquid crystal flowing in ``nematic planes''. For even higher shear rates
-the system undergoes a transition to a flow-aligned nematic state.
-
-In the next sections we investigate the BPII-1 and BPII-2 flow regimes in more detail 
+\rev{
+We observe another configuration, lying between those with periodically recurring conformations 
+and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
+In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
+The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
+and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
+}
+
+\rev{
+The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
+the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
+}
+
+\rev{
+The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
+to the tumbling motion of the local director field, which is absent in the Grandjean texture.
+For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
+}
+
+In the next sections we investigate the BPII-1 flow regime in more detail 
 by looking at the kinetics of the disclination network.
@@
 
-Fig. \ref{bp2-med} shows the disclination network in shear flow as
+Fig. \ref{bp2-1-disc} shows the disclination network in shear flow as
 it undergoes homogeneous shearing in the BPII-1 flow regime. 
@@
 The vorticity motion occurs in such a way that the positions of breakup and reconnection 
-point in the network, visible in Fig. \ref{bp2-med}, are slightly offset and allow the network 
+point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
 to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
@@
 
-Fig. \ref{bp2-velo} shows the disclination network in the middle of a breakup-reconnection
+Fig. \ref{bp2-1-velo} shows the disclination network in the middle of a breakup-reconnection
 cycle, with superimposed velocity vectors.
@@
 right in the upper half and to the left in the lower half of the picture.}
-\label{bp2-med}
+\label{bp2-1-disc}
 \end{figure}
@@
 it takes the network to reconnect along the flow direction.}
-\label{bp2-velo}
+\label{bp2-1-velo}
 \end{figure*}
 
-Characteristic bands are visible in Fig. \ref{bp2-velo}, which are 
+Characteristic bands are visible in Fig. \ref{bp2-1-velo}, which are 
 oriented along the vorticity direction. The direction of the flow
@@
 between the maximum and minimum velocities 
-along the $z$-direction. Therefore the averaged flow along the vorticity
-direction, leading to network motion, is very small, and much smaller than
+along the $z$-direction. 
+\rev{
+Therefore the averaged flow along the vorticity
+direction is very small, and much smaller than
 the local velocity flow. This suggests that the
 flow of the network along the vorticity 
-direction is permeative (i.e. the fluid moves but the network as a whole 
-moves much less).
-
-\subsubsection{Regime BPII-2: high shear rates }
-
-If the flow rate exceeds a certain critical value BPII does not undergo the 
-homogeneous deformation described above and adopts another mode of flow. 
-Fig. \ref{bp2-high} shows a sequence of cuts through the director 
-field at high shear rate. The entire blue phase network has broken up 
-into a simple cholesteric helix and flows as a travelling 
-helical wave with the helical axis oriented along the vorticity direction.
-The state is translationally invariant in the flow and gradient directions.
-It is the preferred mode of flow of a standard cholesteric phase
-in vorticity mode at low Ericksen numbers. This is followed by
-the uncoiling transition to a nematic, flow-aligned state at even 
-higher shear rates ~\cite{Rey:1996a, Rey:1996b}.
-
-In our scenario the BP to cholesteric transition occurs for shear rates $\gd\simeq 4.68\e{-4}$, which
-corresponds to Ericksen numbers of ${\it Er}\simeq4$, based on our
-parameters for BPII ($K=0.02, \eta=0.666$, and $p_0/2=16$;
-see Table \ref{tab1} in the Appendix).
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
-\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
-\caption{Director field of BPII at high shear rate: The images show snapshots at time steps 
-$t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page.
-The colour code gives the alignment of the director field in the flow direction (x) with red fully aligned and blue out of plane.
-The configuration is a helical wave that travels along the direction of vorticity.
-It is translationally invariant along the gradient and flow direction.}
-\label{bp2-high}
-\end{figure}
-
-We observe another configuration, lying between the travelling helical wave 
-and the  flow-aligned nematic state at shear rates 
-$7.81\e{-4}\lesssim\gd\lesssim9.37\e{-4}$ ($6.66\lesssim {\it Er} \lesssim 8$). In this interval
-the helical axis now lies along the flow gradient direction (the so-called
-Grandjean texture~\cite{deGennes}).
-We will address this regime in section \ref{cholflow}, as it is common to BPI and BPII.
+direction is permeative. This claim is further supported by the characteristic 
+velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
+The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
+whereas the network as a whole moves only in one direction.
+}
 
@@
 characterising its equilibrium structure are well separated and do not 
-intersect. (The quiescent state resembles the first frame in Fig.~\ref{bp1-low}
+intersect. (The quiescent state resembles the first frame in Fig.~\ref{bp1-1-disc}
 below.)
@@
 
-Fig. \ref{bp1-rheo} shows the ratio between the 
-apparent viscosity $\eta_{app}$ and $\eta$ versus time and 
-strain for the same shear rates as in Fig. \ref{bp2-rheo}. The plot
-shows that there is more shear thinning in BPI than in BPII, for
-the shear rate range we considered. 
-Through much of the flow range we studied,
-the average shear stress versus shear rate can be fitted reasonably 
-well by a power law $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.019, b=0.63$.
-
-For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
+\rev{
+Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
+shear rates as in Fig.~\ref{bp2-flowcurve}, whereas Fig. \ref{bp1-appvisc} shows 
+the apparent viscosity ration $\eta_{app}/\eta$ over total strain.
+}
+
+For intermediate (but not low) shear rates
+($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more find regular oscillations in the stress
 response versus time, and in the network morphology -- we label this 
 regime BPI-2. 
+\rev{
+This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
+ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.
+}
+
 The BPI-2 regime now does not persist to arbitrarily low shear: there the
 oscillations become irregular and seemingly chaotic, and after a few
-cycles the disclination pattern becomes amorphous. We refer to this 
-flow regime as BPI-1. 
+cycles the disclination pattern becomes amorphous.
+\rev{
+For these very low shear rates 
+($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
+the flow curve in Fig. \ref{bp1-flowcurve}
+apparently develops a horizontal yield stress branch. 
+We refer to this flow regime as BPI-1.
+}
+
 The BPI-2 regime is also unstable if the shear is increased {\em above} 
-a critical value -- close to this breakup point the oscillations in the 
-shear stress show an increasingly complex
-pattern, until the network breaks up completely and a new regime, BPI-3, is entered.
-In regime BPI-3 ($6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $10.67 \lesssim {\it Er} \lesssim 13.33$) 
-we observe that the BPI network transforms under shear into
-a  quasi two-dimensional state consisting of
-double-twist rolls.
+a critical value 
+\rev{
+in the region of Ericksen numbers $Er \simeq O(1)$.
+This is a regime 
+($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
+where viscous and elastic forces have more or less 
+the same magnitude we labelled BPI-3. 
+In this window of strain rates the network seems to become
+increasingly sensitive towards small perturbations in the flow field
+that can cause irregularities in the order parameter structure.
+Due to the back flow, i.e. the coupling of flow and order in our model, 
+these distortions form a possible route towards an instance of 
+"rheochaos"~\cite{rheochaos,Cates:2002}.
+}
+
+\rev{
 Finally, for shear rates such that
-$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
-configuration with the helical axis along the gradient direction as 
+$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
+and before breakup into a flow-aligned nematic occurs 
+(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
+we observe the same cholesteric Grandjean configuration with the 
+helical axis along the gradient direction as 
 reported above for BPII-2 (see
-section \ref{cholflow}).
+section \ref{gj-fan}). This regime is referred to as BPI-4.
+}
 
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
-\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
-\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
-and strain (bottom). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. 
-Depending on the configuration in steady shear flow three different regimes 
-can be identified: amorphous BP network with yield stress (BPI-1, triangles), 
-steady flow with recurring patterns (BPI-2, open circles), 
-breakup of the disclination network (BPI-3, solid circle), 
-cholesteric helix with axis in gradient direction (CHG, open squares) 
-and flow-aligned nematic state (FAN, solid square). 
+\includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
+\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
+\rev{
+Depending on the configuration in steady shear flow five different regimes 
+can be identified: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
+steady flow with periodically recurring patterns (BPI-2, open squares), 
+breakup of the disclination network into amorphous state (BPI-3, open circles)
+\com{Calaculate another shear rate to get a regime},
+cholesteric helix with axis in gradient direction, also referred to
+as Grandjean texture and frustrated metastable states that transform into 
+a Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
+and flow-aligned nematic state (BPI-5, solid triangles). 
 The error bars represent the minimum and maximum shear stress 
-during one cycle. For clarity the individual curves have been shifted along
-the time and strain axis.}
-\label{bp1-rheo}
+during one cycle in regime BPI-2.
+The two dashed lines show fits to the data in regime BPI-2 with 
+$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
+and in regime BPI-4 with 
+$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
+}
+\label{bp1-flowcurve}
+\end{figure}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
+\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
+\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
+\caption{
+\rev{
+Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised by the background viscosity $\eta$: 
+The top picture shows regime BPI-1 (AN) where
+the BP network transforms into an amorphous network featuring a yield stress.
+The centre picture shows regimes BPI-2 (PRC) and BPI-3 (AN), 
+whereas the picture at the bottom
+depicts data of the Grandjean tecture (GJ and FGJ if frustrated) in regime BPI-4 
+and the flow-aligned nematic state (BPI-5, FAN). 
+Note that the $\eta_{app}/\eta$ is negative for the lowest 
+flow rates as the (negative) contribution of the elastic forces
+of the network dominates the total stress.
+}
+}
+\label{bp1-appvisc}
 \end{figure}
@@
 \includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
-\caption{Average thermodynamic shear stress and free energy density of BPI at shear rate 
-$\gd=4.88\e{-5}, Er=0.08$: The negative branch in the stress is related to a local maximum and a following local 
-minimum in average free energy density. This creates unstable conditions that lead 
-to reconstruction of the defect lattice into an amorphous network state which 
+\caption{Average thermodynamic shear stress and free energy density of 
+BPI at shear rate $\gd=4.88\e{-5}, Er=0.08$: 
+The negative branch in the stress is related to a local maximum 
+and a following local minimum in average free energy density. 
+This creates unstable conditions that lead to reconstruction of the 
+defect lattice into an amorphous network state which 
 seemingly features a yield stress.}
@@
 the shear stress increases steeply and goes through a maximum.
-Shortly after it goes negative (the total shear stress 
-with the background viscosity contribution remains positive) 
+Shortly after it goes negative 
 to become positive again later on, forming a complete cycle.
@@
 low shear regime might depend on our choice of boundary
-conditions which in practice are equivalent to having infinitely
-distant walls to which the network is anchored (see discussion
-in the Appendix).
-With free boundary conditions and for very low wall velocity, 
+conditions which in practice are equivalent to 
+\rev{
+a liquid crystal that is in flow
+contrary to a free-standing network that might withstand
+very small flow rates (see discussion in the Appendix).
+With free boundary conditions and for very low flow velocity, 
 permeative flows might instead allow for
@@
 along their axis~\cite{Marenduzzo:2006b}.
+}
 
@@
 \includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
-\caption{Thermodynamic shear stress $\Pi_{xy}$ versus strain near the transition between regime
-BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
+\caption{Thermodynamic shear stress $\Pi_{xy}$ versus strain near the 
+transition between regime BPI-1 and BPI-2. 
+The shear rates are $\gd=1.95\e{-5}$ (red solid) 
 and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
@@
 pictures on the far left, centre left and centre right 
-show the situation at later time steps $t=4.0, 7.0$ and $7.5\e{5}$,
+show the situation at later time steps $t=1.65, 1.8$ and $2.0\e{6}$,
 which the system passes for several cycles before it ends up in
-an amorphous state (bottom row far right, at time step $t=2.2\e{6}$).}
-\label{bp1-low}
+an amorphous state (bottom row far right, at time step $t=2.4\e{6}$).}
+\label{bp1-1-disc}
 \end{figure*}
-
-\if{The negative branch in the shear stress is connected
-to a local maximum followed by a local minimum in the 
-average free energy. Hence, the region of negative stress can be 
-regarded as a phase during which elastic forces exert a 
-'pull' on the network while it tries to reach a more 
-favourable configuration. However, the low flow rate
-causes this unstable intermediate state to be exposed to these
-elastic forces for a relatively long time, long enough to 
-create small distortions which finally break the symmetry of the
-flow-induced configuration.}\fi
 
@@
 breakdown, of BPI, it is instructive to follow the dynamics of 
-the network under a slow flow -- this is depicted in Fig. \ref{bp1-low}.
+the network under a slow flow -- this is depicted in Fig. \ref{bp1-1-disc}.
 Three different stages can be distinguished. 
@@
 This helical conformation that emerges already at strains $\gamma\simeq1$ 
-is shown in Fig. \ref{bp1-low} (top row).
+is shown in Fig. \ref{bp1-1-disc} (top row).
 At this point the original BPI has already been significantly deformed.
@@
 where the disclinations form double helices which tilt and realign under the shear.
-There is no perceivable movement of the 
-network along the vorticity direction at any stage of the cycle.
-(This contrasts with the BPII-2 results reported earlier.)
 
@@
 transforms into an amorphous network state with almost constant
-stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ LBU.
+stress in the region of $\Pi_{xy}\simeq 1-2\e{-5}$ LBU.
 If the shear is stopped, this flow-induced amorphous state
@@
 Adjacent to regime BPI-1 ($\gd\lesssim1.95\e{-5}$; $ {\it Er}\lesssim 0.33$) but 
-at slightly larger shear rates ($3.91\e{-5}\lesssim\gd\lesssim 4.687\e{-4}$; $0.67\lesssim{\it Er}\lesssim8$)
+at slightly larger shear rates ($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}$; $0.67\lesssim{\it Er}\lesssim2$)
 lies another region where the network flows with periodically 
-recurring conformations (Fig. \ref{bp1-rheo}). 
+recurring conformations (Fig. \ref{bp1-2-disc}). 
 We refer to this region as BPI-2. The transition between BPI-1 and BPI-2 
@@
 caused destabilisation of the periodic network and led to the 
-amorphous configuration in the steady state.)
+amorphous configuration in the steady state of BPI-1.)
 Hence, an explanation
@@
 large stress fluctuations -- this leads to
-a different order structure and to a stabilisation 
+\rev{
+different topological reconncetions with different order structure
+}
+and to a stabilisation 
 compared to the oscillations found transiently in
@@
 steady state at $\gd=1.56\e{-4}$ and time steps 
-$t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. 
-During every cycle the network is displaced also along the gradient and vorticity direction, 
-but the magnitude and direction of the offset does not appear to obey a 
-simple rule.}
-\label{bp1-med}
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
+During every cycle the network also is displaced 
+\rev{along the vorticity direction just as
+in regime BPII-1.
+}
+}
+\label{bp1-2-disc}
 \end{figure*}
@@
 \begin{figure*}[htpb]
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
-\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
-\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: 
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-364k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-366k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-368k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-370k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-372k_run914r.png}
+\includegraphics[width=0.32\textwidth]{v_yz-v_z-374k_run914r.png}
+\caption{Velocity patterns in BPI for positive helicity: 
 The pictures show a snapshot of the periodically recurring patterns in the 
-secondary velocity components $(v_y,v_z)$. 
-The colour code gives the magnitude and sign 
-of the $z$-component. The vertical and horizontal direction are the gradient and 
-vorticity direction, respectively, with the flow into/out of the plane in the 
-upper/lower half of the images.}
-\label{bp1-velo}
+secondary velocity components $(v_y,v_z)$ at $\gd=1.56\e{-4}$ and time steps 
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$, i.e. for the same shear rate
+and time steps as in Fig.\ref{bp1-2-disc}. The colour code gives the magnitude and sign 
+of the $z$-component. For negative helicity the sense of motion of the 
+network and the secondary velocity components are inversed just as in Fig. \ref{bp2-1-velo}. 
+\rev{
+Note that the viewing direction is changed along an orbit in x-y plane
+ to give the clearest view on the velocity pattern between
+the disclination lines.
+}
+}
+\label{bp1-2-velo}
 \end{figure*}
 
-Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
+Fig. \ref{bp1-2-disc} shows snapshots of the periodically recurring 
 BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
@@
 undergoing a homogeneous topology-conserving 
-transformation. It features rather intricate, flow-induced 
-conformations. Surprisingly, if the shear is switched off in
+transformation. 
+\rev{
+It features intricate, flow-induced 
+conformations consisting of well-separated ondulating 
+disclination lines similar to those intertwined helices 
+in Fig.~\ref{bp1-1-disc} at early times.
+}
+
+\com{Need to check this claim:} 
+\rev{
+Surprisingly, if the shear is switched off in
 the BPI-2 regime at some point during an oscillation, 
@@
 reverts to a perfect quiescent BPI -- in stark contrast with
-what happens in BPI-1, for lower shear rates.
-
-Finding the recurrence period for BPI is not simple, as
-this depends subtly on the applied shear rate. The analysis is complex for
-a number of reasons. 
-First, the recurrence period for the stress is not always the same as
-that of the network configuration. 
-For instance, at $\gd=3.9062500\e{-5}$, it takes $2/\gd$ for one
- stress cycle to complete, but $4/\gd$ for the network to recur --
-this is because the configurations at times $t$ and $t+2/\gd$ are 
-mirror-images. This is by no means generic though: for example
-at $\gd=4.8828125\e{-5}$  it takes $8/\gd$ for one
-stress cycle to complete, but $16/\gd$ for the network to recur. 
-For some higher shear rates, the recurrence period of stress
-and network conformations are instead the same.
-Second, there exist isolated values (or perhaps small ranges) of shear
-rates for which the recurrence period is much longer and far more irregular,
-with seemingly intermittent dynamics (e.g. at $\gd=1.3671875\e{-4}$ the
-network configuration repeats itself after $6/\gd$, $12/\gd$, $22/\gd$).
-Taken together, all this evidence very much suggests that 
-the BPI-2 range includes at least some instances of ``rheochaos''~\cite{rheochaos,Cates:2002}.
-
-Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ for left- and right-handed chirality of
-the blue phase. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in that there is no simple
-separation into bands with opposite secondary velocity. 
-However, just as in the case of BPII, both patterns
-are mirror images and feature the highest velocities in the regions where 
-the largest conformational changes take place.
-
-\if{In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
-we performed a Fourier analysis of the shear stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
-The spectrum is shown in Fig. \ref{bp1-spectrum}.
-
-There are two aspects which are worth mentioning. First, we observe a strong contribution of 
-the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
-However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
-significantly larger. 
-Second, more surprisingly, the frequency of the fundamental mode is the same 
-for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is 
-in contrast to what is observed at lower flow rates. By doubling the flow rate from 
-$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the peak position in the spectrum 
-moves to double the frequency. This means that the fundamental mode at flow rate $\gd=7.81\e{-5}$ 
-coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on. 
-At higher shear rates in regime BPI-3 this 'lock-in' of flow modes disappears again.
-
-Interestingly the lock-in has only a minor effect on the flow velocities themselves.
-The average velocities in Tab. \ref{tab1} (see Appendix) reveal that only the magnitude of some of the secondary 
-flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
-The latter is directly linked to the imposed shear rate. }\fi
-
-\if{
+what happens in BPI-1, for lower shear rates or BPI-3, for higher shear rates. 
+}
+
+Fig. \ref{bp1-2-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ positive chirality of
+the blue phase at the same time steps shown in Fig.~\ref{bp1-2-disc}. 
+The emerging pattern is similar to that of BPII, shown in
+Fig. \ref{bp2-1-velo}. 
+\rev{
+There is also separation into fluctuating bands 
+along the vorticity direction with opposite secondary velocity.
+The secondary velocity component $v_z$ also changes sign upon
+inverting the chirality of the underlying cholesteric helix.}
+
+\rev{The recurrence period needs to be determined again 
+}
+
+\subsubsection{Regime BPI-3: high shear rates }
+
+\rev{
+At high flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
+($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
+periodically recurring disclination network of regime BPI-2. 
+Fig. \ref{bp1-3-disc} shows a time sequence of this breakup, beginning
+with a regular disclination network with helices aligned along the 
+flow direction.
+}
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-600k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-675k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-700k_run915r.png}\\
+\includegraphics[width=0.32\textwidth]{disc-725k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-1000k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-1200k_run915r.png}
+\caption{}
+\label{bp1-3-disc}
+\end{figure*}
+
+Interestingly, the critical Ericksen number for this breakup are in 
+the range of 2-4 and therefore in a regime where viscous forces begin to 
+dominate of elastic forces. 
+Therefore the BPI-3 regime includes at least some instances of "rheochaos"~\cite{rheochaos,Cates:2002}.
+
+\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{gj-fan}
+
+At shear rates beyond the regimes BPI-3 and BPII-2, but below the transition to a 
+flow-aligned nematic state at still higher shear rates, we found another regime where 
+both blue phases adopt the same configuration in steady shear flow, 
+independent of their initial state.
+\rev{
+The director field of this configuration, also known as
+Grandjean texture, is shown in Fig. \ref{grandjean}.
+}
+It consists of a simple cholesteric helix with the helical axis oriented 
+along the velocity gradient direction. The state is translational invariant 
+along the flow and vorticity direction and therefore quasi-one-dimensional. 
+\rev{
+However, in contrast to other states that we identified as being metastable
+there is no tumbling motion of the director. 
+While flowing the director retains its relative orientation.
+Consequently, the shear stress is lower than for
+e.g. the travelling helical wave (see \cite{Rey:1996a, Rey:1996b} for instance)
+as there is less dissipation.
+}
+
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
-\caption{Fourier analysis of the shear stress $\Pi_{xy}$ of BPI: The main picture shows 
-spectral components and the 'lock-in' of flow modes, 
-visible in the fixed peak position for certain flow rates.  
-The inset shows a zoom of the same data over 
-a portion of the frequency range.
-The frequency is given in inverse time, in simulation units. TRUE?
-OH: I WOULD SAY NO AS IT IS IN UNITS OF 2$\pi$ (RADIANS?).}
-\label{bp1-spectrum}
+\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
+\caption{Flow induced cholesteric state or Grandjean texture: 
+The orientation of the helix is along the gradient direction. 
+The director field is quasi-static during the flow, which occurs 
+in ``nematic layers''. Colour coding indicates the 
+$x$ component of the director field (see colorbar on the right).
+For clarity only selected sites are shown along the $y$-direction.
+The structure is translationally invariant along the other directions.}
+\label{grandjean}
 \end{figure}
-}\fi
-
-The intricacy and periodic recurrence of the flow-induced conformations 
-in BPI raises the question: how robust are these results to variations in 
-system size? Although a pitch length of 64 LBU provides enough resolution 
-for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
-in total are actually sufficient to avoid finite size effects.
-Unfortunately the computational costs of larger simulations are still
-quite prohibitive (in the region of ${\it O}(10^5)$ core hours per run at
-doubled system size), given that large strains have to
-be simulated in order to reach a steady state. Therefore we opted for a 
-comparison with smaller runs.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
-\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Disclination network and shear stress $\Pi_{xy}$ in BPI: 
-The dark (red) network in top picture shows a snapshot of run that consists 
-of $2^3=8$ unit cells, whereas the light (cyan) network displays the 
-central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
-The bottom picture shows the shear stress over time. (Interestingly, 
-an intermediate sized run with $3^3=27$ unit cells led to results that 
-were virtually identical to the smaller run with 8 unit cells.) 
-The shear rate was $\gd=7.81\e{-5}$.
-}
-\label{bp1-2uc4uc}
-\end{figure}
-
-Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
-shear stress $\Pi_{xy}$ of two different runs, one consisting of 8 
-unit cells and the other one with 64 unit cells.
-We observe that there are deviations in the visual appearance of the 
-network as well as in the time-dependence of the shear stress.
-Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
-to reach the steady state with its flow-induced recurring disclination network. 
-These results indicate that we cannot know with certainty whether the individual 
-flow-induced conformations that we find for 64 unit cells are affected 
-by the imposed periodicity -- quite possibly the details of the oscillations
-would differ in an even larger simulation. 
-Nevertheless, the comparison suggests that 
-the recurrence period and the complexity of the network in flow are  
-generic features of BPI in this regime of flow.
-
-\subsubsection{Regime BPI-3: high shear rates }
-
-At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
-we observe a break-up of the flow-induced 
-disclination network of regime BPI-2. However, 
-in contrast to BPII, where the configuration is a travelling helical 
-wave (cf. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
-two-dimensional, translationally invariant state, which is shown in 
-Fig. \ref{bp1-high}. 
-There are distinctive regions where the director field is 
-predominantly oriented along the vorticity direction and 
-other regions where it is oriented rather in flow-gradient plane. 
-Occasionally, while moving past each other, these rolls resemble 
-regions of local double twist.
-Interestingly the critical Ericksen number for the break-up lies in 
-about the same window as that of transition between BPII-1 and BPII-2.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d-z-302k_run916.png}
-\caption{Director field of BPI at high shear rate: The pictures show a region 
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{dir3d+y-450k_run916r.png}
+\includegraphics[width=0.32\textwidth]{dir3d+y-510k_run916r.png}
+\includegraphics[width=0.32\textwidth]{dir3d+y-800k_run916r.png}
+\caption{Director field of BPI-4 at high shear rate: The pictures show a region 
 of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
@@
 \label{bp1-high}
-\end{figure}
-
-\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{cholflow}
-
-At shear rates beyond the regimes BPI-3 and BPII-2, but below the transition to a 
-flow-aligned nematic state at still higher shear rates, we found another regime where 
-both blue phases adopt the same configuration in steady shear flow, 
-independent of their initial state.
-The director field of this configuration is shown in Fig. \ref{fig:cholflow}.
-It consists of a simple cholesteric helix with the helical axis oriented 
-along the velocity gradient direction. The state is translational invariant 
-along the flow and vorticity direction and therefore also one-dimensional. 
-However, in contrast to the travelling helical wave of BPII-2 the 
-director field is now static and there is no tumbling motion. 
-While flowing the director retains its relative orientation.
-Consequently, the shear stress is lower than for
-the 2D double-twist rolls of BPI-3 and the travelling helical wave
-(Fig. \ref{bp2-rheo} and \ref{bp1-rheo}) as there is less dissipation.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.45\textwidth]{dir3d+y-200k_run1179.png}
-\caption{Flow induced cholesteric state. The orientation of the helix is 
-along the gradient direction. The director field is quasi-static during 
-the flow, which occurs in ``nematic layers''. Colour coding indicates the 
-$x$ component of the director field (see colorbar on the right).
-For clarity only selected sites are shown along the $y$-direction (along which
-the structure is translationally invariant).}
-\label{fig:cholflow}
-\end{figure}
+\end{figure*}
 
@@
 \hline
-BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+BPI-1 (AN) &0.24 &0.02 & 0.1388 &-17.7 &16.7 &3.7 &-3.6 &3.6 &3.5 &-3.9 &3.1 &2.8 \\
+BPI-1 (AN) &0.49 &0.04 & 0.1388 &-33.4 &32.0 &4.3 &-3.0 &3.1 &3.2 &-3.7 &2.5 &3.1 \\
+BPI-1 (AN) &0.98 &0.08 & 0.1388 &-64.4 &63.6 &4.7 &-3.6 &3.4 &4.8 &-5.3 &-4.2 &4.4 \\
+BPI-1 (AN) &1.95 &0.17 & 0.1388 &-126.2 &127.0 &6.8 &-3.8 &4.8 &6.7 &-5.2 &5.1 &6.2 \\
 \hline
-BPI-2 & 3.91 & 0.67 & 0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
-BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+BPI-2 (PRC) &3.91 &0.33 & 0.1388 &-248.6 &248.3 &2.8 &-5.4 &5.6 &5.1 &-3.8 &4.6 &3.9 \\
+BPI-2 (PRC) &7.81 &0.67 & 0.1388 &-497.5 &496.4 &5.2 &-8.5 &8.0 &8.4 &-5.9 &7.8 &6.8 \\
+BPI-2 (PRC) &15.63 &1.33 & 0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-13.0} &\bf{15.6} &\bf{11.2} \\
+BPI-2 (PRC) &15.63 &1.33 & -0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-15.6} &\bf{13.0} &\bf{11.2} \\
+BPI-2 (PRC) &23.44 &2.00 & 0.1388 &-1492.2 &1485.5 &9.6 &-4.8 &3.7 &12.3 &-5.7 &4.6 &14.8 \\
 \hline
-BPI-3 & 62.50 & 10.67& 0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
-BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+BPI-3 (AN) &31.25 &2.67 & 0.1388 &-1990.6 &1983.2 &17.6 &-4.5 &6.3 &13.3 &-8.3 &10.0 &15.9 \\
+BPI-3 (AN) &46.87 &4.00 & 0.1388 &-2988.2 &2985.5 &42.4 &-10.5 &12.3 &28.4 &-15.5 &15.2 &26.8 \\
 \hline
-CHG & 93.75 &16.00 & 0.1388 & -5955.12 &5948.84 & 0.71 & -0.002 & 0.002 & -  & -16.45 & 16.03 &0.62 \\
-CHG & 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+BPI-4 (GJ) &54.69 &4.67 & 0.1388 &-3475.1 &3473.4 &- &- &- &- &-8.9 &8.8 &- \\
+BPI-4 (FGJ) &62.50 &5.33 & 0.1388 &-4101.7 &4122.7 &- &- &- &- &-29.6 &20.0 &- \\
+BPI-4 (FGJ) &78.13 &6.01 & 0.1388 &-5140.0 &5092.3 &- &- &- &- &-35.8 &24.6 &- \\
+BPI-4 (GJ) &93.75 &8.00 & 0.1388 &-5955.4 &5946.1 &-  &- &- &- &-15.5 &15.1 &- \\
+BPI-4 (GJ) &125.0 &10.67 & 0.1388 &-7935.0 &7933.7 &- &- &- &- &-21.3 &20.1 &- \\
+BPI-4 (GJ) &187.5 &16.00 & 0.1388 &-11901.7 &11901.7 &- &- &- &- &-32.8 &29.1 &- \\
 \hline
-FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+BPI-5 (FAN) & 250.0 &21.33 & 0.1388 &-15875.0 &15875.0 &- &- &- &- &-0.2 &-0.1 &- \\
+BPI-5 (FAN) & 375.0 &32.00 & 0.1388 &-23812.5 &23812.5 &- &- &- &- &-0.3 &0.1 &- \\
 \hline
@@
 \hline
-BPII-1 & 0.244 &0.02 & 0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
-BPII-1 & 0.488 &0.04 & 0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
-BPII-1 & 0.976 &0.08 & 0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
-BPII-1 & 1.95 &0.17  & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
-BPII-1 &3.91 & 0.33& 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
-BPII-1 &7.81 &0.67 & 0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
-BPII-1 &15.63 & 1.33& 0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
-BPII-1 & 15.63 & 1.33& -0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
-BPII-1 & 31.25 &2.67 & 0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
+BPII-1 (PRC) &0.24 &0.02 & 0.1963 &-7.8 &7.7 &0.1 &-0.2 &0.2 &0.1 &-0.3 &0.3 &0.2 \\
+BPII-1 (PRC) &0.49 &0.04 & 0.1963 &-15.5 &15.4 &0.2 &-0.3 &0.4 &0.2 &-0.5 &0.4 &0.2 \\
+BPII-1 (PRC) &0.98 &0.08 & 0.1963 &-31.0 &30.8 &0.3 &-0.6 &0.6 &0.4 &-0.9 &0.8 &0.4 \\
+BPII-1 (PRC) &1.95 &0.17 & 0.1963 &-62.0 &61.6 &0.6 &-1.2 &1.2 &0.8 &-1.8 &1.6 &0.8 \\
+BPII-1 (PRC) &3.91 &0.33 & 0.1963 &-123.9 &123.3 &1.2 &-2.1 &2.1 &1.6 &-3.4 &3.0 &1.5 \\
+BPII-1 (PRC) &7.81 &0.67 & 0.1963 &-247.5 &246.8 &2.0 &-3.8 &3.8 &3.0 &-6.5 &5.5 &2.6 \\
+BPII-1 (PRC) &15.63 &1.33 & 0.1963 &-494.3 &493.7 &2.5 &-5.5 &-5.5 &4.0 &\bf{-10.9} &\bf{8.7} &\bf{3.9} \\
+BPII-1 (PRC) &15.63 &1.33 & -0.1963 &-494.3 &493.7 &2.5 &-5.5 &5.5 &4.0 &\bf{-8.7} &\bf{10.9} &\bf{3.9} \\
+BPII-1 (PRC) &31.25 &2.67 & 0.1963 &-987.5 &987.1 &3.5 &-6.9 &6.9 &5.1 &-17.7 &13.1 &5.0 \\
+BPII-1 (PRC) &39.06 &3.33 & 0.1963 &1233.9 &1233.6 &3.7 &-6.9 &6.9 &5.1 &-20.0 &14.5 &5.2 \\
 \hline
-BPII-2 & 46.87 &4.00 & 0.1388 & -1476.75 & 1476.84 &1.69 &-0.35  &0.35  & 0.17 &-0.23 &0.23  & 0.22\\
-BPII-2 & 62.50 & 5.34& 0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
+BPII-2 (GJ) &46.87 &4.00 & 0.1963 &-1479.8 &1477.0 &- &- &- &- &-10.0 &9.6  &- \\
+BPII-2 (GJ) &54.69 &4.67 & 0.1963 &-1720.5 &1720.8 &- &- &- &- &-11.5 &11.2 &- \\
+BPII-2 (GJ) &62.5 &5.33 & 0.1963  &-1966.7 &1966.2 &- &- &- &- &-13.3 &12.7 &- \\
+BPII-2 (GJ) &78.13 &6.01 & 0.1963 &-2458.3 &2457.8 &- &- &- &- &-16.9 &15.7 &- \\
+BPII-2 (GJ) &93.75 &8.00 & 0.1963 &-2949.7 &2949.7 &- &- &- &- &-20.3 &18.4 &- \\
+BPII-2 (GJ) &125.0 &10.67 &0.1963 &-3932.8 &3932.8 &- &- &- &- &-26.6 &22.4 &- \\
 \hline
-CHG & 78.13 & 6.66 &  0.1388 &-2458.60 & 2458.61  &0.37 &- &- &- &-18.54 &16.94 &0.32 \\
-CHG & 93.75 &8.00 & 0.1388 &-2950.2  &2950.2 &0.50 &- &- &- &-22.69 & 19.72 & 0.04\\
-\hline
-FAN & 125.0 &10.67 & 0.1388 &-3937.31  & 3937.32 &0.28 &-  &-  &-  &-1.79 &1.15  &0.14 \\
+BPII-3 (FAN) &187.5 &16.00 & 0.1963 &-5906.2  &5906.2 &- &-  &-  &-  &-0.6 &0.4 &- \\
+BPII-3 (FAN) &250.0 &21.33 & 0.1963 &-7874.9  &7874.9 &- &-  &-  &-  &-0.5 &0.3 &- \\
 \hline
 \end{tabular}
-\caption{Minima, maxima and standard deviation of time-averaged velocity components 
+\caption{
+\rev{Minima, maxima and standard deviation of time-averaged velocity components 
 for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
 shear flow. All velocity values are given in $10^{-5}$ lattice units. The transient dynamics 
-has not been included in the averages. The regimes comprise the formation
-of an amorphous network (BPI-1), periodically recurring conformations (BPI-2 and BPII-1), 
-two-dimensional double twist rolls (BPI-3), a one-dimensional travelling helical wave (BPII-2), 
-a cholesteric helix oriented along the gradient direction (CHG), 
-and a flow-aligned nematic state (FAN).
+prior to regular oscillations or network breakup has not been included into the averages. 
+The regimes comprise the formation of an amorphous network (AN, BPI-1), 
+periodically recurring conformations with oscillatory stress response (PRC, BPI-2 and BPII-1), 
+amorphous networks at Ericksen numbers ${Er\simeq O(1)}$ (AN, BPI-3), 
+a Grandjean texture that may be frustrated (GJ \& FGJ, BPII-2 and BPI-4),
+and a flow-aligned nematic state (FAN, BPI-5 and BPII-3).
 (Last six columns refer to the secondary flow.)}
+}
 \label{tab1}

### 2013-06-28 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2029 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-06-06 papers/bprheo/bprheo.tex
@@
 II continuously breaks and reforms under steady shear, leading to an oscillatory stress response in time.
-For larger shear rates, the structure breaks up \rev{into a Grandjean texture with a 
-cholesteric helix lying along the flow gradient direction according to the imposed flow}.
+For larger shear rates, the structure breaks up 
+\rev{
+into a Grandjean texture with a 
+cholesteric helix lying along the flow gradient direction according to the imposed flow.
+}
 BPI leads to a very different response. Here, oscillations are only possible for intermediate
 shear rates -- very slow flow leads to a transition into an amorphous network 
-with an apparent yield stress. \rev{Larger shear rates can make the network sensitive
+with an apparent yield stress. Larger shear rates can make the network sensitive
 to flow-induced transitions to metastable and amorphous states. For even
 larger flow rates a Grandjean texture is observed, just as in the case of BPII.
-At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.}
+At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
 Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
@@
 %Please use \dag to cite the ESI in the main text of the article.
-%If you article does not have ESI please remove the the \dag symbol from the title and the above footnotetext.
+%If you article does not have ESI please remove the \dag symbol from the title and the above footnotetext.
 
@@
 %\footnotetext{\ddag~Additional footnotes to the title and authors can be included \emph{e.g.}\ `Present address:' or `These authors contributed equally to this work' as above using the symbols: \ddag, \textsection, and \P.
-Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the the correct place in the list.}
+Please place the appropriate symbol next to the author's name and include a \texttt{\textbackslash footnotetext} entry in the correct place in the list.}
 
@@
 They also found that a strong secondary flow appears.
-\rev{Early experiments with cholesteric liquid crystals
-showed that flow can also give rise to conformational transitions. \cite{Press:1978}.}
+\rev{
+Early experiments with cholesteric liquid crystals
+showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
+}
 Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
 the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
-lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. \rev{These result, however, were derived under the 
+lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
+\rev{
+These result, however, were derived under the 
 assumption that the molecules rotate only in flow-gradient plane whilst
-the orientation of the cholesteric helix remained unchanged.}  
+the orientation of the cholesteric helix remained unchanged.
+}  
 Rey also studied cholesterics subjected to both steady flow and low frequency
@@
 in this expansion is negligible under the slow flows considered here
-for which the the fluid motion is almost incompressible~\cite{Denniston:2001}.
+for which the fluid motion is almost incompressible~\cite{Denniston:2001}.
 $\eta$ is an isotropic background viscosity which is set to $\eta=0.6666$ in LB units (these are discussed below, see~\cite{Henrich:2011a,Henrich:2010b}).
@@
 the BP unit cell. Likewise, $v$ was taken to be the velocity difference
-across one unit cell, i.e. $v=\dot{\gamma} l$. \rev{The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.}
+across one unit cell, i.e. $v=\dot{\gamma} l$. 
+\rev{
+The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
+}
 
@@
 $\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
-\rev{For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
-case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory response (PRC), 
-amorphous networks (AN), amorphous and metastable states (AMS) as well as those featuring a Grandjean texture (GJ) and 
-a flow-aligned nematic state (FAN). The last two occur regardless of the initial state. 
-They are for Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
+\rev{
+For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
+case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
+amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
+may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
+GJ and FAN occur regardless of the initial state.
+}
+
+\rev{
+These regimes are characterised by the following flow rates: 
+For Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
 BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
-BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) and
-for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
+BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) 
+and for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
 BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
-BPI-3 (AMS) ($3.125\e{-4}\lesssim \gd\lesssim 7.813\e{-4}; 2.67 \lesssim {\it Er} \lesssim 6.01$),
-BPI-4 (GJ) ($9.375\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 8\lesssim {\it Er} \lesssim 16)$ and
-BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.34\lesssim {\it Er}$).}
+BPI-3 (AN) ($3.125\e{-4}\lesssim \gd\lesssim 4.69\e{-4}; 2.67 \lesssim {\it Er} \lesssim 4$),
+BPI-4 (GJ and FGJ) ($5.47\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 4.67\lesssim {\it Er} \lesssim 16)$ and
+BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.33\lesssim {\it Er}$).
+}
 
@@
 \includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
-\caption{\rev{Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
+\caption{
+\rev{
+Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
 whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
 breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
-The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.}}
+The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
+}
+}
 \label{bp2-flowcurve}
@@
 \includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
-\caption{\rev{Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ (normalised to the background viscosity $\eta$) 
-of BPII versus strain $\gamma = \gd\cdot t$ in the regimes BPII-1 (top, PRC) and BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
-The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.}} 
+\caption{
+\rev{
+Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
+of BPII versus strain $\gamma = \gd\cdot t$  
+normalised to the background viscosity $\eta$.
+The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
+The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
+}
+} 
 \label{bp2-appvisc}
@@
 This holds true even in regime BPII-2, where the network breaks up and 
-the liquid crystal adopts a Grandjean structure. The coefficents of the power law fit are
+the liquid crystal adopts a Grandjean structure. The coefficients of the power law fit are
 here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
@@
 
-\rev{Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
+\rev{
+Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
 defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
@@
 absence at very low shear rate of a permeation mode is probably due to our
-choice of boundary conditions.}
+choice of boundary conditions.
+}
 
@@
 
-\rev{We observe another configuration, lying between those with periodically recurring conformations 
+\rev{
+We observe another configuration, lying between those with periodically recurring conformations 
 and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
@@
 The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
-and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.}
-
-\rev{The picture at the bottom of Fig. \ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
-the flow-aligned nematic state. We will address these regimes in section \ref{cholflow}, as they are common to BPI and BPII.}
-
-\rev{The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
+}
+
+\rev{
+The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
+the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
+}
+
+\rev{
+The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
 the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
 to the tumbling motion of the local director field, which is absent in the Grandjean texture.
-For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.}
+For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
+}
 
@@
 between the maximum and minimum velocities 
-along the $z$-direction. \rev{Therefore the averaged flow along the vorticity
+along the $z$-direction. 
+\rev{
+Therefore the averaged flow along the vorticity
 direction is very small, and much smaller than
@@
 The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
-whereas the network as a whole moves only in one direction.}
+whereas the network as a whole moves only in one direction.
+}
 
@@
 
-\rev{Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
-shear rates as in Fig.~\ref{bp2-flowcurve}. 
-The plot shows that there is more shear thinning in BPI than in BPII, for
-the shear rate range we considered.}
-
-For intermediate (but not low) shear rates, we once more find regular oscillations in the stress
+\rev{
+Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
+shear rates as in Fig.~\ref{bp2-flowcurve}, whereas Fig. \ref{bp1-appvisc} shows 
+the apparent viscosity ration $\eta_{app}/\eta$ over total strain.
+}
+
+For intermediate (but not low) shear rates
+($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more find regular oscillations in the stress
 response versus time, and in the network morphology -- we label this 
 regime BPI-2. 
-\rev{This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
-ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.}
+\rev{
+This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
+ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.
+}
 
@@
 cycles the disclination pattern becomes amorphous.
-\rev{For these very low shear rates the flow curve in Fig. \ref{bp1-flowcurve}
-develops a horizontal branch, which suggest the existence of a yield stress.} 
+\rev{
+For these very low shear rates 
+($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
+the flow curve in Fig. \ref{bp1-flowcurve}
+apparently develops a horizontal yield stress branch. 
 We refer to this flow regime as BPI-1.
+}
 
@@
 a critical value 
-\rev{in the region of Ericksen numbers $Er \simeq O(1)$
-where viscous and elastic forces are of the same magnitude. 
-Close to this breakup point the network seems to become
+\rev{
+in the region of Ericksen numbers $Er \simeq O(1)$.
+This is a regime 
+($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
+where viscous and elastic forces have more or less 
+the same magnitude we labelled BPI-3. 
+In this window of strain rates the network seems to become
 increasingly sensitive towards small perturbations in the flow field
-that cause irregularities in the order parameter structure.
-These distortions couple back to the flow and form a possible
-route towards "rheochaos"~\cite{rheochaos,Cates:2002}.}
-
-We also observe that
-
-regime BPI-3 ($3.125\e{-4}\lesssim\gd\lesssim7.81\e{-4}$; $2.67 \lesssim {\it Er} \lesssim 6.01$)
-
+that can cause irregularities in the order parameter structure.
+Due to the back flow, i.e. the coupling of flow and order in our model, 
+these distortions form a possible route towards an instance of 
+"rheochaos"~\cite{rheochaos,Cates:2002}.
+}
+
+\rev{
 Finally, for shear rates such that
-$9.37\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($16 \lesssim {\it Er} \lesssim 21.33$) we observe the same cholesteric 
-configuration with the helical axis along the gradient direction as 
+$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
+and before breakup into a flow-aligned nematic occurs 
+(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
+we observe the same cholesteric Grandjean configuration with the 
+helical axis along the gradient direction as 
 reported above for BPII-2 (see
-section \ref{cholflow}).
+section \ref{gj-fan}). This regime is referred to as BPI-4.
+}
 
@@
 \caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
-\rev{Depending on the configuration in steady shear flow five different regimes 
-can be identified: amorphous BP network with yield stress (BPI-1, solid squares), 
+\rev{
+Depending on the configuration in steady shear flow five different regimes 
+can be identified: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
 steady flow with periodically recurring patterns (BPI-2, open squares), 
 breakup of the disclination network into amorphous state (BPI-3, open circles)
-and metastable configurations (BPI-3, solid circles), 
-cholesteric helix with axis in gradient direction (BPI-4, open triangles) 
+\com{Calculate another shear rate to get a regime},
+cholesteric helix with axis in gradient direction, also referred to
+as Grandjean texture and frustrated metastable states that transform into 
+a Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
 and flow-aligned nematic state (BPI-5, solid triangles). 
 The error bars represent the minimum and maximum shear stress 
-during one cycle in regime BPI-2.}}
+during one cycle in regime BPI-2.
+The two dashed lines show fits to the data in regime BPI-2 with 
+$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.84}$ (left) 
+and in regime BPI-4 with 
+$\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
+}
 \label{bp1-flowcurve}
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_a.pdf}\\
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_b.pdf}\\
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp1_c.pdf}
-\caption{Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma}$ (normalised by $\eta$) of BPI versus time (top) 
-and strain (bottom).}
+\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_a.pdf}\\
+\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_b.pdf}\\
+\includegraphics[width=0.45\textwidth]{app_visc_strain_bp1_c.pdf}
+\caption{
+\rev{
+Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised by the background viscosity $\eta$: 
+The top picture shows regime BPI-1 (AN) where
+the BP network transforms into an amorphous network featuring a yield stress.
+The centre picture shows regimes BPI-2 (PRC) and BPI-3 (AN), 
+whereas the picture at the bottom
+depicts data of the Grandjean texture (GJ and FGJ if frustrated) in regime BPI-4 
+and the flow-aligned nematic state (BPI-5, FAN). 
+Note that the $\eta_{app}/\eta$ is negative for the lowest 
+flow rates as the (negative) contribution of the elastic forces
+of the network dominates the total stress.
+}
+}
 \label{bp1-appvisc}
@@
 large stress fluctuations -- this leads to
-a different order structure and to a stabilisation 
+\rev{
+different topological reconnections with different order structure
+}
+and to a stabilisation 
 compared to the oscillations found transiently in
@@
 $t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
-During every cycle the network is displaced also along the gradient and vorticity direction.}
-\label{bp1-med}
+During every cycle the network also is displaced 
+\rev{along the vorticity direction just as
+in regime BPII-1.
+}
+}
+\label{bp1-2-disc}
 \end{figure*}
@@
 $t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$, i.e. for the same shear rate
-and time steps as in Fig.\ref{bp1-med}. The colour code gives the magnitude and sign 
+and time steps as in Fig.\ref{bp1-2-disc}. The colour code gives the magnitude and sign 
 of the $z$-component. For negative helicity the sense of motion of the 
-network and the secondary velocity components are inversed just as in Fig. \ref{bp2-velo}. 
-Note that the viewing direction has been adapted in
-each snapshot to give the clearest view on the band-like pattern between
-the disclination lines.}
-\label{bp1-velo}
+network and the secondary velocity components are inversed just as in Fig. \ref{bp2-1-velo}. 
+\rev{
+Note that the viewing direction is changed along an orbit in x-y plane
+ to give the clearest view on the velocity pattern between
+the disclination lines.
+}
+}
+\label{bp1-2-velo}
 \end{figure*}
 
-Fig. \ref{bp1-med} shows snapshots of the periodically recurring 
+Fig. \ref{bp1-2-disc} shows snapshots of the periodically recurring 
 BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
-BPI-2 does not resemble the equilibrium configuration
+BPI-2 does not resemble so much the equilibrium configuration
 undergoing a homogeneous topology-conserving 
-transformation. It features rather intricate, flow-induced 
-conformations. Surprisingly, if the shear is switched off in
+transformation. 
+\rev{
+It features intricate, flow-induced 
+conformations consisting of well-separated undulating 
+disclination lines similar to those intertwined helices 
+in Fig.~\ref{bp1-1-disc} at early times.
+}
+
+\com{Need to check:} 
+\rev{
+Surprisingly, if the shear is switched off in
 the BPI-2 regime at some point during an oscillation, 
@@
 reverts to a perfect quiescent BPI -- in stark contrast with
-what happens in BPI-1, for lower shear rates.
-
-Finding the recurrence period for BPI is not simple, as
-this depends subtly on the applied shear rate. The analysis is complex for
-a number of reasons. 
-First, the recurrence period for the stress is not always the same as
-that of the network configuration. 
-For instance, at $\gd=3.9062500\e{-5}$, it takes $2/\gd$ for one
- stress cycle to complete, but $4/\gd$ for the network to recur --
-this is because the configurations at times $t$ and $t+2/\gd$ are 
-mirror-images. This is by no means generic though: for example
-at $\gd=4.8828125\e{-5}$  it takes $8/\gd$ for one
-stress cycle to complete, but $16/\gd$ for the network to recur. 
-For some higher shear rates, the recurrence period of stress
-and network conformations are instead the same.
-Second, there exist isolated values (or perhaps small ranges) of shear
-rates for which the recurrence period is much longer and far more irregular,
-with seemingly intermittent dynamics (e.g. at $\gd=1.3671875\e{-4}$ the
-network configuration repeats itself after $6/\gd$, $12/\gd$, $22/\gd$).
-Taken together, all this evidence very much suggests that 
-the BPI-2 range includes at least some instances of "rheochaos"~\cite{rheochaos,Cates:2002}.
-
-Fig. \ref{bp1-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ for left- and right-handed chirality of
-the blue phase. The emerging pattern is similar to that of BPII from 
-Fig. \ref{bp2-velo}, but differs in that there is no simple
-separation into bands with opposite secondary velocity. 
-However, just as in the case of BPII, both patterns
-are mirror images and feature the highest velocities in the regions where 
-the largest conformational changes take place.
-
-%%% No differences in revison between different system sizes %%%
-\ifold{The intricacy and periodic recurrence of the flow-induced conformations 
-in BPI raises the question: how robust are these results to variations in 
-system size? Although a pitch length of 64 LBU provides enough resolution 
-for tracking the network in flow it is not clear whether $4^3=64$ unit cells 
-in total are actually sufficient to avoid finite size effects.
-Unfortunately the computational costs of larger simulations are still
-quite prohibitive (in the region of ${\it O}(10^5)$ core hours per run at
-doubled system size), given that large strains have to
-be simulated in order to reach a steady state. Therefore we opted for a 
-comparison with smaller runs.
+what happens in BPI-1, for lower shear rates or BPI-3, for higher shear rates. 
+}
+
+\com{Check recurrence period}
+
+Fig. \ref{bp1-2-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ positive chirality of
+the blue phase at the same time steps shown in Fig.~\ref{bp1-2-disc}. 
+The emerging pattern is similar to that of BPII, shown in
+Fig. \ref{bp2-1-velo}. 
+\rev{
+There is also separation into fluctuating bands 
+along the vorticity direction with opposite secondary velocity.
+The secondary velocity component $v_z$ also changes sign upon
+inverting the chirality of the underlying cholesteric helix.
+}
+
+\subsubsection{Regime BPI-3: high shear rates }
+
+\rev{
+At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
+($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
+periodically recurring disclination network in regime BPI-2 and the formation
+of another amorphous network. Fig. \ref{bp1-3-disc} shows a sequence of 
+snapshots. 
+Initially, the disclination lines take the form of a regular, staggered array of helices 
+with the helical axis along the flow direction. 
+However, this mode of flow proves unstable as the regular formation dissolves into 
+an irregular state that features helicoidal disclination lines of varying 
+appearance. This is different from the amorphous network of
+regime BPI-1 (cf. Fig. \ref{disc-bp1-1}) where the disclination lines are also mostly
+well separated but without a clear orientation with respect to the flow. 
+}
+
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-600k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-675k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-700k_run915r.png}\\
+\includegraphics[width=0.32\textwidth]{disc-725k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-1000k_run915r.png}
+\includegraphics[width=0.32\textwidth]{disc-1200k_run915r.png}
+\caption{
+\rev{Disclination network of BPI in regime BPI-2 (high flow rates):
+The sequence shows the break-up of the initially ordered state into 
+an amorphous network at $\gd=3.125\e{-4}$ and time steps
+$t=6.0, 6.75, 7.0, 7.25 \e{5}, 1.0,1.2\e{6}$ in vorticity-gradient plane.
+During every cycle the network also is displaced
+along the vorticity direction.
+}
+}
+\label{bp1-3-disc}
+\end{figure*}
+
+\rev{
+Interestingly, the critical Ericksen number for this observed break-up is in 
+the range of $Er\simeq 2-4$. This is a regime where viscous forces begin to 
+dominate over elastic forces and where the nonlinear backflow mechanism which couples
+order and flow can cause small deviations to grow over time. 
+This suggests that the BPI-3 regime constitutes an instance 
+of "rheochaos"~\cite{rheochaos,Cates:2002}.
+}
+
+\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{gj-fan}
+
+At shear rates beyond the regimes BPI-3 and BPII-2, but below the transition to a 
+flow-aligned nematic state at still higher shear rates, we found another regime where 
+both blue phases adopt the same configuration in steady shear flow, 
+independent of their initial state.
+\rev{
+The director field of this configuration, also known as
+Grandjean texture, is shown in Fig. \ref{grandjean}.
+}
+It consists of a simple cholesteric helix with the helical axis oriented 
+along the velocity gradient direction. The state is translational invariant 
+along the flow and vorticity direction and therefore quasi-one-dimensional. 
+\rev{
+However, in contrast to other states that we identified as being metastable
+there is no tumbling motion of the director. 
+While flowing the director retains its relative orientation.
+Consequently, the shear stress is lower than for
+e.g. the travelling helical wave (see \cite{Rey:1996a, Rey:1996b} for instance)
+as there is less dissipation.
+}
 
 \begin{figure}[htpb]
-%\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
-%\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
-\caption{Disclination network and shear stress $\Pi_{xy}$ in BPI: 
-The dark (red) network in top picture shows a snapshot of run that consists 
-of $2^3=8$ unit cells, whereas the light (cyan) network displays the 
-central section of 8 unit cells in a bigger run with $4^3=64$ unit cells altogether. 
-The bottom picture shows the shear stress over time. (Interestingly, 
-an intermediate sized run with $3^3=27$ unit cells led to results that 
-were virtually identical to the smaller run with 8 unit cells.) 
-The shear rate was $\gd=7.81\e{-5}$.
-}
-\label{bp1-2uc4uc}
+\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
+\caption{Flow induced cholesteric state or Grandjean texture: 
+The orientation of the helix is along the gradient direction. 
+The director field is quasi-static during the flow, which occurs 
+in ''nematic layers''. Colour coding indicates the 
+$x$ component of the director field (see colorbar on the right).
+For clarity only selected sites are shown along the $y$-direction.
+The structure is translationally invariant along the other directions.}
+\label{grandjean}
 \end{figure}
-
-Fig. \ref{bp1-2uc4uc} shows a comparison of the disclination network and 
-shear stress $\Pi_{xy}$ of two different runs, one consisting of 8 
-unit cells and the other one with 64 unit cells.
-We observe that there are deviations in the visual appearance of the 
-network as well as in the time-dependence of the shear stress.
-Perhaps surprisingly, the smaller run with 8 unit cells takes longer 
-to reach the steady state with its flow-induced recurring disclination network. 
-These results indicate that we cannot know with certainty whether the individual 
-flow-induced conformations that we find for 64 unit cells are affected 
-by the imposed periodicity -- quite possibly the details of the oscillations
-would differ in an even larger simulation. 
-Nevertheless, the comparison suggests that 
-the recurrence period and the complexity of the network in flow are  
-generic features of BPI in this regime of flow.
-\fi
-
-\subsubsection{Regime BPI-3: high shear rates }
-
-%%% Absent in revison - new 2D conformation is persumably metastable %%%
-\ifold At high flow rates, $6.25\e{-4}\lesssim\gd\lesssim7.81\e{-4}$ ($10.67\lesssim {\it Er}\lesssim 13.33$), 
-we observe a break-up of the flow-induced 
-disclination network of regime BPI-2. However, 
-in contrast to BPII, where the configuration is a travelling helical 
-wave (cf. Fig. \ref{bp2-high}), BPI breaks down into a quasi 
-two-dimensional, translationally invariant state, which is shown in 
-Fig. \ref{bp1-high}. 
-There are distinctive regions where the director field is 
-predominantly oriented along the vorticity direction and 
-other regions where it is oriented rather in flow-gradient plane. 
-Occasionally, while moving past each other, these rolls resemble 
-regions of local double twist.
-Interestingly the critical Ericksen number for the break-up lies in 
-about the same window as that of transition between BPII-1 and BPII-2. 
-\fi
 
@@
 \includegraphics[width=0.32\textwidth]{dir3d+y-800k_run916r.png}
-\caption{Director field of BPI at high shear rate: The pictures show a region 
+\caption{Director field of BPI-4 at high shear rate: The pictures show a region 
 of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
@@
 
-\subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{cholflow}
-
-At shear rates beyond the regimes BPI-3 and BPII-2, but below the transition to a 
-flow-aligned nematic state at still higher shear rates, we found another regime where 
-both blue phases adopt the same configuration in steady shear flow, 
-independent of their initial state.
-The director field of this configuration is shown in Fig. \ref{fig:cholflow}.
-It consists of a simple cholesteric helix with the helical axis oriented 
-along the velocity gradient direction. The state is translational invariant 
-along the flow and vorticity direction and therefore also one-dimensional. 
-However, in contrast to the travelling helical wave of BPII-2 the 
-director field is now static and there is no tumbling motion. 
-While flowing the director retains its relative orientation.
-Consequently, the shear stress is lower than for
-the 2D double-twist rolls of BPI-3 and the travelling helical wave
-(Fig. \ref{bp2-rheo} and \ref{bp1-rheo}) as there is less dissipation.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
-\caption{Flow induced cholesteric state. The orientation of the helix is 
-along the gradient direction. The director field is quasi-static during 
-the flow, which occurs in ``nematic layers''. Colour coding indicates the 
-$x$ component of the director field (see colorbar on the right).
-For clarity only selected sites are shown along the $y$-direction (along which
-the structure is translationally invariant).}
-\label{fig:cholflow}
-\end{figure}
-
 %\clearpage
@@
 
+\rev{
 We were able to characterise the rheology of cubic BPs, and identified 
-two different flow regimes for BPII and three different ones for BPI. 
+thre different flow regimes for BPII and five different ones for BPI.
+This comprises steady states that BPI and BPII have in common like the 
+Grandjean texture and the flow-aligned nematic state at high shear rates.
+}  
 Below an Ericksen number of $Er\simeq3$, BPII exhibits weak shear-thinning and 
@@
 
+\rev{
 At larger Ericksen numbers, $4\lesssim Er\lesssim 10$, the flowing BPII network 
-breaks up in a two-step manner. First a simple cholesteric helix forms with the helical axis 
-along the vorticity direction. The flow excites a travelling helical wave,
-as predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}.
-Upon increasing the shear, the helix rotates and orients along the
-gradient direction, where it can support larger flow rates due to the translational
-symmetry along the flow direction. Still larger flow rates break this residual
+breaks up into a simple cholesteric helix with the helical axis 
+along the gradient direction, also referred to as Grandjean texture. 
+The travelling helical wave, predicted theoretically in Ref.~\cite{Rey:1996a,Rey:1996b}
+under the assumption that the director is restricted to the flow-gradient plane,
+turns out to be a metastable state of flow with a higher free energy.
+It may also transforms into the Grandjean texture, the steady state with the lowest
+shear stress and free energy in this range of shear rates.
+}
+Still larger flow rates break this residual
 cholesteric order and leave a standard flow-aligned nematic state,
@@
 This is a direct consequence of the topological differences between their disclination 
-networks. Below an Ericksen number of $Er\sim 0.4$, the BPI network cannot both flow and 
+networks. Below an Ericksen number of $Er\sim 0.3$, the BPI network cannot both flow and 
 retain a regular appearance. It eventually reorganises into an amorphous network that
@@
 
-At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim8$,
+At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim2$,
 BPI kinematically bypasses the large stress fluctuations and 
-flows with very complex, periodically recurring flow-induced conformations.
-These conformations entail regularly arrange helical disclinations which 
-rotate due to the shear -- a new flow-induced state which was not previously
-predicted. 
-The oscillatory dynamics is in some cases intermittent, and the
-recurrence period for the network conformations and stress patterns
-is not given by a simple formula as for BPII.
+flows with periodically recurring flow-induced conformations.
+These conformations entail regularly arranged helical disclinations which 
+rotate due to the shear. 
 Despite their complex appearance, the flow-induced conformations are 
 topologically connected to the quiescent BPI and after switching off the 
-shear flow a defect-free blue phase reforms. It is tempting to
+shear flow a defect-free blue phase reforms.
+
+\rev{
+At even larger Ericksen numbers $2\lesssim Er \lesssim 4$ the behaviour of BPI is again different.
+At first, it exhibits regular helicoidal disclination lines oriented along the flow direction.
+After a short time these become irregular and an amorphous network emerges, which has 
+a different appearance than that observed at the lowest flow rates.
+We believe that at these flow rates, where neither viscous nor elastic forces clearly dominate, 
+the coupling between order and flow makes any ordered conformation very vulnerable to small fluctuations.
+}
+It is tempting to
 interpret the unsteady oscillations seen in this regime as an instance of 
@@
 
-Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers in
-a two-step process. First, at $10\lesssim\ Er\lesssim 15$, it adopts a 
-quasi-two-dimensional configuration that is translationally invariant along vorticity direction. 
-Then, at $16\lesssim Er \lesssim 22$, it forms a cholesteric helix along the flow gradient 
-direction.  Finally, at $Er\gtrsim30$ (BPI) and $Er\gtrsim10$ (BPII) the configuration
+Just as in the case of BPII, the BPI network breaks up at larger Ericksen numbers. 
+First, at $5\lesssim\ Er\lesssim 16$, it forms a cholesteric helix along the flow gradient 
+direction, the Grandjean texture. Finally, at $Er\gtrsim21$ (BPI) and $Er\gtrsim16$ (BPII) the configuration
 is a flow-aligned nematic state.
 Although experimental evidence to support
-our results is currently not available, we hope this work will inspire such experiments, and believe it can shed some light 
-on the flow properties of complex liquid-crystalline phases.
-
-% If in two-column mode, this environment will change to single-column
-% format so that long equations can be displayed. Use
-% sparingly.
-%\begin{widetext}
-% put long equation here
-%\end{widetext}
-
-% figures should be put into the text as floats.
-% Use the graphics or graphicx packages (distributed with LaTeX2e)
-% and the \includegraphics macro defined in those packages.
-% See the LaTeX Graphics Companion by Michel Goosens, Sebastian Rahtz,
-% and Frank Mittelbach for instance.
-%
-% Here is an example of the general form of a figure:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Use the figure* environment if the figure should span across the
-% entire page. There is no need to do explicit centering.
-
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}
-% \end{figure}
-
-% Surround figure environment with turnpage environment for landscape
-% figure
-% \begin{turnpage}
-% \begin{figure}
-% \includegraphics{}%
-% \caption{\label{}}
-% \end{figure}
-% \end{turnpage}
-
-% tables should appear as floats within the text
-%
-% Here is an example of the general form of a table:
-% Fill in the caption in the braces of the \caption{} command. Put the label
-% that you will use with \ref{} command in the braces of the \label{} command.
-% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
-% \begin{tabular}{} command.
-% The ruledtabular environment adds doubled rules to table and sets a
-% reasonable default table settings.
-% Use the table* environment to get a full-width table in two-column
-% Add \usepackage{longtable} and the longtable (or longtable*}
-% environment for nicely formatted long tables. Or use the the [H]
-% placement option to break a long table (with less control than 
-% in longtable).
-% \begin{table}%[H] add [H] placement to break table across pages
-% \caption{\label{}}
-% \begin{ruledtabular}
-% \begin{tabular}{}
-% Lines of table here ending with \\
-% \end{tabular}
-% \end{ruledtabular}
-% \end{table}
-
-% Surround table environment with turnpage environment for landscape
-% table
-% \begin{turnpage}
-% \begin{table}
-% \caption{\label{}}
-% \begin{ruledtabular}
-% \begin{tabular}{}
-% \end{tabular}
-% \end{ruledtabular}
-% \end{table}
-% \end{turnpage}
-
-% Specify following sections are appendices. Use \appendix* if there
-% only one appendix.
-%\appendix
-
-%\section{}
-
-% If you have acknowledgments, this puts in the proper section head.
+our results is currently not available, we hope this work will inspire such experiments, 
+and believe it can shed some light on the flow properties of complex liquid-crystalline phases.
+
 \section*{Acknowledgments}
@@
 EPSRC grant nos. EP/E045316 and EP/E030173 and the MAPPER EU-FP7 project 
-(grant no. RI-261507).We thank Peter J. Collings for stimulating discussions. 
+(grant no. RI-261507). We thank Peter J. Collings and Tom C. Lubensky for stimulating discussions. 
 MEC holds a Royal Society Research Professorship.
@@
 \hline
-BPI-1 (AN) &0.24 &0.02 & 0.1388 & & & & & & & & & \\
-BPI-1 (AN) &0.49 &0.04 & 0.1388 & & & & & & & & & \\
-BPI-1 (AN) &0.98 &0.08 & 0.1388 & & & & & & & & & \\
-BPI-1 (AN) &1.95 &0.17 & 0.1388 & & & & & & & & & \\
-%BPI-1 & 0.244 & 0.04& 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
-%BPI-1 & 0.488 & 0.08 & 0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
-%BPI-1 & 0.976 & 0.17 & 0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
-%BPI-1 & 1.95 & 0.33 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
+BPI-1 (AN) &0.24 &0.02 & 0.1388 &-17.7 &16.7 &3.7 &-3.6 &3.6 &3.5 &-3.9 &3.1 &2.8 \\
+BPI-1 (AN) &0.49 &0.04 & 0.1388 &-33.4 &32.0 &4.3 &-3.0 &3.1 &3.2 &-3.7 &2.5 &3.1 \\
+BPI-1 (AN) &0.98 &0.08 & 0.1388 &-64.4 &63.6 &4.7 &-3.6 &3.4 &4.8 &-5.3 &-4.2 &4.4 \\
+BPI-1 (AN) &1.95 &0.17 & 0.1388 &-126.2 &127.0 &6.8 &-3.8 &4.8 &6.7 &-5.2 &5.1 &6.2 \\
 \hline
-BPI-2 (PRC) &3.91 &0.33 & 0.1388 & & & & & & & & & \\
-BPI-2 (PRC) &7.81 &0.67 & 0.1388 & & & & & & & & & \\
-BPI-2 (PRC) &15.63 &1.33 & 0.1388 & & & & & & & & & \\
-BPI-2 (PRC) &15.63 &1.33 & -0.1388 & & & & & & & & & \\
-BPI-2 (PRC) &23.44 &2.00 & 0.1388 & & & & & & & & & \\
-%BPI-2 & 7.81 & 1.33 & 0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
-%BPI-2 & 15.63 & 2.67 & 0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
-%BPI-2 & 15.63 & 2.67 & -0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
-%BPI-2 & 31.25 &5.33 & 0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
-%BPI-2 & 46.87 & 8.00 & 0.1388 & -3025.46 & 3003.05 &21.43 & -56.96 & 56.44  & 18.75 & -12.65 & 82.97  & 27.96\\
+BPI-2 (PRC) &3.91 &0.33 & 0.1388 &-248.6 &248.3 &2.8 &-5.4 &5.6 &5.1 &-3.8 &4.6 &3.9 \\
+BPI-2 (PRC) &7.81 &0.67 & 0.1388 &-497.5 &496.4 &5.2 &-8.5 &8.0 &8.4 &-5.9 &7.8 &6.8 \\
+BPI-2 (PRC) &15.63 &1.33 & 0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-13.0} &\bf{15.6} &\bf{11.2} \\
+BPI-2 (PRC) &15.63 &1.33 & -0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-15.6} &\bf{13.0} &\bf{11.2} \\
+BPI-2 (PRC) &23.44 &2.00 & 0.1388 &-1492.2 &1485.5 &9.6 &-4.8 &3.7 &12.3 &-5.7 &4.6 &14.8 \\
 \hline
-BPI-3 (AMS) &31.25 &2.67 & 0.1388 & & & & & & & & & \\
-BPI-3 (AMS) &46.87 &4.00 & 0.1388 & & & & & & & & & \\
-BPI-3 (AMS) &54.69 &4.67 & 0.1388 & & & & & & & & & \\
-BPI-3 (AMS) &62.50 &5.33 & 0.1388 & & & & & & & & & \\
-BPI-3 (AMS) &78.13 &6.01 & 0.1388 & & & & & & & & & \\
-%BPI-3 &78.13 &13.33 & 0.1388 & -4968.11 &4973.75 &20.89 &-60.61 &58.75 &23.66 &-84.93 &116.61 &34.75 \\
+BPI-3 (AN) &31.25 &2.67 & 0.1388 &-1990.6 &1983.2 &17.6 &-4.5 &6.3 &13.3 &-8.3 &10.0 &15.9 \\
+BPI-3 (AN) &46.87 &4.00 & 0.1388 &-2988.2 &2985.5 &42.4 &-10.5 &12.3 &28.4 &-15.5 &15.2 &26.8 \\
 \hline
-BPI-4 (GJ) &93.75 &8.00 & 0.1388 & & & & & & & & & \\
-BPI-4 (GJ) &125.0 &10.67 & 0.1388 & & & & & & & & & \\
-BPI-4 (GJ) &187.5 &16.00 & 0.1388 & & & & & & & & & \\
-%GJT (BPI-4)& 125.0 & 21.33& 0.1388 & -7934.97 & 7935.41 & 0.36 & -0.005 & 0.005 & - &-22.73 & 21.12 & 0.07\\
+BPI-4 (GJ) &54.69 &4.67 & 0.1388 &-3475.1 &3473.4 &- &- &- &- &-8.9 &8.8 &- \\
+BPI-4 (FGJ) &62.50 &5.33 & 0.1388 &-4101.7 &4122.7 &- &- &- &- &-29.6 &20.0 &- \\
+BPI-4 (FGJ) &78.13 &6.01 & 0.1388 &-5140.0 &5092.3 &- &- &- &- &-35.8 &24.6 &- \\
+BPI-4 (GJ) &93.75 &8.00 & 0.1388 &-5955.4 &5946.1 &-  &- &- &- &-15.5 &15.1 &- \\
+BPI-4 (GJ) &125.0 &10.67 & 0.1388 &-7935.0 &7933.7 &- &- &- &- &-21.3 &20.1 &- \\
+BPI-4 (GJ) &187.5 &16.00 & 0.1388 &-11901.7 &11901.7 &- &- &- &- &-32.8 &29.1 &- \\
 \hline
-BPI-5 (FAN) & 250.0 &21.33 & 0.1388 & & & & & & & & & \\
-BPI-5 (FAN) & 375.0 &32.00 & 0.1388 & & & & & & & & & \\
-%FAN & 187.5 & 32.00& 0.1388 &-11906.4  &119064 &0.36 &- &- &- &-0.68 & 0.38 & 0.05\\
+BPI-5 (FAN) & 250.0 &21.33 & 0.1388 &-15875.0 &15875.0 &- &- &- &- &-0.2 &-0.1 &- \\
+BPI-5 (FAN) & 375.0 &32.00 & 0.1388 &-23812.5 &23812.5 &- &- &- &- &-0.3 &0.1 &- \\
 \hline
@@
 \end{tabular}
-\caption{\rev{Minima, maxima and standard deviation of time-averaged velocity components 
+\caption{
+\rev{Minima, maxima and standard deviation of time-averaged velocity components 
 for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
 shear flow. All velocity values are given in $10^{-5}$ lattice units. The transient dynamics 
-has not been included in the averages. The regimes comprise the formation
-of an amorphous network (BPI-1), periodically recurring conformations with oscillatory response (BPI-2 and BPII-1), 
-amorphous and metastable steady states (BPI-3), a Grandjean texture (BPII-2 and BPI-4),
-and a flow-aligned nematic state (BPI-5 and BPII-3).
-(Last six columns refer to the secondary flow.)}}
+prior to regular oscillations or network breakup has not been included into the averages. 
+The regimes comprise the formation of an amorphous network (AN, BPI-1), 
+periodically recurring conformations with oscillatory stress response (PRC, BPI-2 and BPII-1), 
+amorphous networks at Ericksen numbers ${Er\simeq O(1)}$ (AN, BPI-3), 
+a Grandjean texture that may be frustrated (GJ \& FGJ, BPII-2 and BPI-4),
+and a flow-aligned nematic state (FAN, BPI-5 and BPII-3).
+(Last six columns refer to the secondary flow.)}
+}
 \label{tab1}

### 2013-07-02 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2034 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{fancyhdr}
+
 \pagestyle{fancy}
@@
 into a Grandjean texture with a 
-cholesteric helix lying along the flow gradient direction according to the imposed flow.
-}
-BPI leads to a very different response. Here, oscillations are only possible for intermediate
+cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
+}
+Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
 shear rates -- very slow flow leads to a transition into an amorphous network 
-with an apparent yield stress. Larger shear rates can make the network sensitive
-to flow-induced transitions to metastable and amorphous states. For even
-larger flow rates a Grandjean texture is observed, just as in the case of BPII.
-At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
-Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
+with an apparent yield stress. 
+\rev{
+Larger shear rates can make the network sensitive
+to flow-induced transitions to metastable and amorphous states.
+}
+For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
+At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
+Our results provide the first theoretical investigation of sheared blue phases in large systems, 
+and are relevant to understanding the bulk rheology of these materials.}
 \vspace{0.5cm}
@@
 \rev{
-These result, however, were derived under the 
+However, this result was derived under the 
 assumption that the molecules rotate only in flow-gradient plane whilst
@@
 is periodic in all three coordinate directions. Sliding
-boundary conditions, or Lees-Edwards planes, are explicitly implemented
+boundary conditions, or Lees-Edwards planes ~\cite{Wagner:2002}, are explicitly implemented
 in both hydrodynamic and thermodynamic sectors to impose shear on the
-system. On
-crossing a Lees-Edwards boundary a Galilean transformation is applied
-with a velocity increment which is fixed in time (and limited in size by the
-low Mach number constraint of LB). An appropriate transformation
-of LB distributions which propagate across a boundary~\cite{Wagner:2002}
-must be applied at each time step, and appropriate
-adjustment to $W_{\alpha\beta}$ is required to compute cross-plane
-gradients of the velocity field used in the update to ${\mathbf Q}$.
-In both cases interpolation of the
-relevant quantities is required to cope with the relative displacement
-of neighbouring lattice sites separated by a sliding plane (the
-displacement may be a fraction of a lattice unit at any given time
-step). The use of multiple sliding planes equally spaced in a single
-system allows
-the overall shear rate to be maintained indefinitely as the system
-becomes larger
-in the velocity gradient direction (in contrast with the use of solid
-walls to impose shear). This hybrid method with Lees-Edwards planes
-has been used successfully to study e.g., smectics~\cite{Henrich:2012a}.
-
-In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
-phases BPI and BPII.
+system. 
+\rev{
+The same hybrid method with Lees-Edwards planes
+has been used successfully to study binary mixtures, particle suspensions~\cite{MacMeccan:2009, Aidun:2010}, 
+emulsions~\cite{Frijters:2012} and smectic liquid crystals~\cite{Henrich:2012a}.
+A more detailed discussion of our use of Lees-Edwards boundary conditions can be found in the Appendix 
+of this paper.
+}
+
+In the following we report results of simulations of the bulk flow behaviour of the cubic Blue 
+Phases BPI and BPII.
 Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
@@
 velocity gradient along the $y$-direction.
-The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
-\rev{
-For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
-case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
-amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
+The shear rates were varied over more than three orders of magnitude from about 
+$\gd=2.44\times \e{-6}$ to $3.75\times\e{-3}$ LBU.
+\rev{
+For clarity we classify various flow regimes, namely three in the case of Blue Phase II and five in the 
+case of Blue Phase I. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
+amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) which 
 may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
-GJ and FAN occur regardless of the initial state.
+GJ and FAN occur in both blue phases as steady states regardless of the initial state.
 }
@@
 As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
-Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
+Typical choices are $q_s=0.17$ for BPI and $q_s=0.15$ for BPII.
+(Note that $q_s$ is small but non-zero at the disclination core. 
+The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
 
 \subsection{Blue Phase II}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
-\caption{
-\rev{
-Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
-whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
-breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
-The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
-}
-}
-\label{bp2-flowcurve}
-\end{figure}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
-\caption{
-\rev{
-Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
-of BPII versus strain $\gamma = \gd\cdot t$  
-normalised to the background viscosity $\eta$.
-The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
-The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
-}
-} 
-\label{bp2-appvisc}
-\end{figure}
 
@@
 
-Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
-the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
-\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
-power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
-the data to a very good approximation.
-This holds true even in regime BPII-2, where the network breaks up and 
-the liquid crystal adopts a Grandjean structure. The coefficients of the power law fit are
-here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
-remarkably small in BPII, in the range of shear rates which we have explored here.}
-
-\rev{
-Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
-defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
-over total strain.
-A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
-without any additional contribution from the liquid crystal.
-The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
-in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
-absence at very low shear rate of a permeation mode is probably due to our
-choice of boundary conditions.
-}
-
-%{We note that permeation flows, which would
-%{\em not} lead to oscillatory response, are effectively disallowed by the combination
-%of the geometry and boundary condition we consider. This is
-%because the periodic boundary conditions impose macroscopic distortion
-%of the network, and are effective analogous to infinitely distant
-%but anchored wall along the $xy$ plane.
-%This means that  our results are effectively valid for sheared bulk 
-%BPs which are anchored at the walls. Furthermore, permeation flows are 
-%only restricted to very slow flows, and are unstable for intermediate 
-%and high flows, where the response of the network depends less on
-%the details of the boundary conditions used at 
-%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}
-
-\rev{
-We observe another configuration, lying between those with periodically recurring conformations 
-and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
-In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
-The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
-and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
-}
-
-\rev{
-The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
-the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
-}
-
-\rev{
-The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
-the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
-to the tumbling motion of the local director field, which is absent in the Grandjean texture.
-For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
-}
-
-In the next sections we investigate the BPII-1 flow regime in more detail 
-by looking at the kinetics of the disclination network.
-
-\subsubsection{Regime BPII-1: low and intermediate shear rates }
-
-Fig. \ref{bp2-1-disc} shows the disclination network in shear flow as
-it undergoes homogeneous shearing in the BPII-1 flow regime. 
-The disclination lines break up and 
-reconnect further downstream, forming a periodically recurring pattern. The
-period needed for a pattern to break up and 
-reform along the flow direction is $\tau_F = 1/\gd$.
-
-The general appearance of the flowing network is, apart from the homogeneous distortion,
-very close to that of the quiescent blue phase at equilibrium. This holds for all
-shear rates in the BPII-1 regime.
-
-Interestingly, while being displaced with the flow the entire network moves 
-along the vorticity direction (z) as well.  A similar behaviour has been recently 
-observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
-In contrast to the stick-slip motion that has been reported there,
-in this case the movement is steady. 
-The vorticity motion occurs in such a way that the positions of breakup and reconnection 
-point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
-to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
-is $\tau_V=6\tau_F$, 
-i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
-for the network to move one unit cell along $z$ (the vorticity direction).
-
-Fig. \ref{bp2-1-velo} shows the disclination network in the middle of a breakup-reconnection
-cycle, with superimposed velocity vectors.
-The plot shows the secondary velocity components, obtained by projecting the
-velocity onto a plane perpendicular to the flow direction.
-This gets rid of the dominating velocity component along the flow, $v_x$, and 
-allows to visualise the patterns in the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the secondary components is typically in the range of a 
-few percent of the primary flow component, for the shear rates 
-and system sizes simulated here.
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
+\caption{
+\rev{
+Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
+whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
+breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
+The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
+}
+}
+\label{bp2-flowcurve}
+\end{figure}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
+\caption{
+\rev{
+Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
+of BPII versus strain $\gamma = \gd\cdot t$  
+normalised to the background viscosity $\eta$.
+The pictures show the regimes BPII-1 (periodically recurring conformations, top), BPII-2 (Grandjean texture, bottom) and BPII-3 (flow-aligned nematic, bottom). 
+The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. total shear-thinning.
+}
+} 
+\label{bp2-appvisc}
+\end{figure}
 
@@
 
-\begin{figure*}[htpb]
+Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
+the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
+\rev{
+Time averages were taken as average over one stress cycle in the steady state.
+Where no oscillations occurred we used the final value at the end of the run.
+For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
+power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
+the data to a very good approximation.
+This holds true even in regime BPII-2, where the network breaks up and 
+the liquid crystal adopts a Grandjean structure with the helical axis oriented along the
+flow gradient direction ($y$). The coefficients of the power law fit are
+here $a=0.68$ and $b=0.94$. Hence, for the range of shear rates which we have explored here, 
+the degree of shear-thinning is remarkably small in BPII.
+}
+
+\rev{
+Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
+defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
+over total strain.
+A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
+without any additional stress contribution from the liquid crystal.
+The top picture shows data for regime BPII-1, where the periodic break-up and reconnecting of the network 
+in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
+absence of a permeation mode at very low shear rates may be due to our
+choice of boundary conditions. We refer the reader to the Appendix for a detailed discussion.
+}
+
+\rev{
+We observe another configuration, lying between those with periodically recurring conformations 
+and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
+In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
+The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
+and the liquid crystal is flowing in ``nematic planes'' with the director predominantly oriented in flow-vorticity plane and having only
+a small angle with the flow direction.
+The picture at the bottom of Fig.~\ref{bp2-appvisc} shows data for this regime, to which we refer as BPII-2. The 
+curves for the two largest flow rates are for BPII-3, the flow-aligned nematic state. 
+We will address these regimes in section \ref{gj-fan}, as they are common to both cubic blue phases BPI and BPII.
+}
+
+\rev{
+The travelling helical waves which was identified in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis turns out to be a metastable flow state as it features larger free energies, higher stresses and dissipation due 
+to the tumbling motion of the director field, which is completely absent in the Grandjean texture.
+For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
+}
+
+In the next sections we investigate the BPII-1 flow regime in more detail 
+by looking at the kinetics of the disclination network.
+
+\subsubsection{Regime BPII-1: low and intermediate shear rates }
+
+\begin{figure}[htbp]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902r.png}
@@
 \label{bp2-1-velo}
-\end{figure*}
+\end{figure}
+
+Fig. \ref{bp2-1-disc} shows the disclination network in shear flow as
+it undergoes homogeneous shearing in the BPII-1 flow regime. 
+The disclination lines break up and 
+reconnect further downstream, forming a periodically recurring pattern. The
+period needed for a pattern to break up and 
+reform along the flow direction is $\tau_F = 1/\gd$.
+
+The general appearance of the flowing network is, apart from the homogeneous distortion,
+very close to that of the quiescent blue phase at equilibrium. This holds for all
+shear rates in the BPII-1 regime.
+
+Interestingly, while being displaced with the flow the entire network moves 
+along the vorticity direction ($z$) as well.  A similar behaviour has been recently 
+observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
+In contrast to the stick-slip motion that has been reported there,
+in this case the movement is steady. 
+The vorticity motion occurs in such a way that the positions of breakup and reconnection 
+point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
+to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
+is $\tau_V=6\tau_F$, 
+i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
+for the network to move one unit cell along $z$ (the vorticity direction).
+
+Fig. \ref{bp2-1-velo} shows the disclination network in the middle of a breakup-reconnection
+cycle, with superimposed velocity vectors.
+The plot shows the secondary velocity components, obtained by projecting the
+velocity onto a plane perpendicular to the flow direction.
+This gets rid of the dominating velocity component along the flow, $v_x$, and 
+allows to visualise the patterns in the 
+two much smaller components $v_y$ and $v_z$.
+The magnitude of the secondary components is typically in the range of a 
+few percent of the primary flow component, for the shear rates 
+and system sizes simulated here.
 
@@
 velocity bands in the secondary flow, visible in Fig.~\ref{bp2-1-velo}. 
-The flow in vorticity direction takes clearly place in positive {\em and} in negative z-direction, 
-whereas the network as a whole moves only in one direction.
+The secondary flow in vorticity direction takes clearly place in positive {\em and} 
+negative z-direction, whereas the movement of the network as a whole is only in one direction.
 }
@@
 Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
-shear rates as in Fig.~\ref{bp2-flowcurve}, whereas Fig. \ref{bp1-appvisc} shows 
-the apparent viscosity ration $\eta_{app}/\eta$ over total strain.
-}
-
-For intermediate (but not low) shear rates
-($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more find regular oscillations in the stress
-response versus time, and in the network morphology -- we label this 
-regime BPI-2. 
-\rev{
-This is also depicted in the centre picture of Fig.~\ref{bp1-appvisc} for the
-ratio of the apparent viscosity $\eta_{app}$ and $\eta$ versus strain.
-}
-
-The BPI-2 regime now does not persist to arbitrarily low shear: there the
-oscillations become irregular and seemingly chaotic, and after a few
-cycles the disclination pattern becomes amorphous.
-\rev{
-For these very low shear rates 
-($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
-the flow curve in Fig. \ref{bp1-flowcurve}
-apparently develops a horizontal yield stress branch. 
-We refer to this flow regime as BPI-1.
-}
-
-The BPI-2 regime is also unstable if the shear is increased {\em above} 
-a critical value 
-\rev{
-in the region of Ericksen numbers $Er \simeq O(1)$.
-This is a regime 
-($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
-where viscous and elastic forces have more or less 
-the same magnitude we labelled BPI-3. 
-In this window of strain rates the network seems to become
-increasingly sensitive towards small perturbations in the flow field
-that can cause irregularities in the order parameter structure.
-Due to the back flow, i.e. the coupling of flow and order in our model, 
-these distortions form a possible route towards an instance of 
-"rheochaos"~\cite{rheochaos,Cates:2002}.
-}
-
-\rev{
-Finally, for shear rates such that
-$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
-and before breakup into a flow-aligned nematic occurs 
-(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
-we observe the same cholesteric Grandjean configuration with the 
-helical axis along the gradient direction as 
-reported above for BPII-2 (see
-section \ref{gj-fan}). This regime is referred to as BPI-4.
+shear rates as in Fig.~\ref{bp2-flowcurve} and Fig. \ref{bp1-appvisc} depicts 
+the apparent viscosity ratio $\eta_{app}/\eta$ over total strain.
 }
@@
 Depending on the configuration in steady shear flow five different regimes 
-can be identified: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
+can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
 steady flow with periodically recurring patterns (BPI-2, open squares), 
-breakup of the disclination network into amorphous state (BPI-3, open circles)
-\com{Calculate another shear rate to get a regime},
+break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
 cholesteric helix with axis in gradient direction, also referred to
-as Grandjean texture and frustrated metastable states that transform into 
-a Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
+as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
+a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
 and flow-aligned nematic state (BPI-5, solid triangles). 
@@
 
+For intermediate (but not too low) shear rates
+($3.91\e{-5}\lesssim\gd\lesssim 2.34\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$), we once more 
+find regular oscillations in the network morphology and in the stress
+response versus time. We label this regime BPI-2. 
+\rev{
+The centre picture of Fig.~\ref{bp1-appvisc} shows the corresponding data for the
+apparent viscosity ratio versus strain.
+}
+
+The BPI-2 regime now does not persist to arbitrarily low shear: there the
+oscillations become irregular and seemingly chaotic, and after a few
+cycles the disclination pattern becomes amorphous.
+\rev{
+For these low shear rates 
+($2.4\e{-6}\lesssim\gd\lesssim 1.95\e{-4}; 0.02\lesssim {\it Er} \lesssim 0.17$) 
+the flow curve in Fig. \ref{bp1-flowcurve}
+develops a horizontal yield stress branch. 
+We refer to this flow regime as BPI-1.
+}
+
+The BPI-2 regime is also unstable if the shear is increased {\em above} 
+a critical value, 
+\rev{
+which is in the region of Ericksen numbers $Er \simeq O(1)$.
+This is a regime 
+($3.13\e{-4}\lesssim\gd\lesssim 4.69\e{-4}; 2.67\lesssim {\it Er} \lesssim 4$) 
+we labelled BPI-3. It is characterised by the balance between viscous and elastic forces.
+This has an influence on the stability of the disclination network as it 
+seems to become increasingly sensitive to small perturbations in the flow field
+which couple back to the order parameter dynamics and vice versa. This phenomenon forms 
+a possible route towards an instance of ``rheochaos'' ~\cite{rheochaos,Cates:2002}.
+}
+
+\rev{
+Finally, for shear rates such that
+$5.47\e{-4}\lesssim\gd\lesssim1.88\e{-3}$ ($4.67\lesssim {\it Er} \lesssim 16$) 
+and before the break-up into a flow-aligned nematic occurs 
+(BPI-5, $2.5\e{-3}\lesssim\gd; 21.33\lesssim {\it Er}$) 
+we observe the same cholesteric Grandjean configuration with the 
+helical axis along the gradient direction as 
+reported above for BPII-2 (see
+section \ref{gj-fan}). This regime is referred to as BPI-4.
+}
+
 \begin{figure}[htpb]
@@
 \rev{
-Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised by the background viscosity $\eta$: 
-The top picture shows regime BPI-1 (AN) where
+Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised to the background viscosity $\eta$: 
+The top picture shows regime BPI-1 (amorphous network) where
 the BP network transforms into an amorphous network featuring a yield stress.
-The centre picture shows regimes BPI-2 (PRC) and BPI-3 (AN), 
+The centre picture shows regimes BPI-2 (periodically recurring conformations) and BPI-3 (amorphous network), 
 whereas the picture at the bottom
-depicts data of the Grandjean texture (GJ and FGJ if frustrated) in regime BPI-4 
-and the flow-aligned nematic state (BPI-5, FAN). 
+depicts data of the (partly frustrated) Grandjean texture in regime BPI-4 
+and the flow-aligned nematic state, labelled BPI-5 in this context. 
 Note that the $\eta_{app}/\eta$ is negative for the lowest 
@@
 low shear regime might depend on our choice of boundary
-conditions which in practice are equivalent to 
-\rev{
-a liquid crystal that is in flow
-contrary to a free-standing network that might withstand
-very small flow rates (see discussion in the Appendix).
-With free boundary conditions and for very low flow velocity, 
-permeative flows might instead allow for
-some slip of the BPI network, as in cholesterics sheared
-along their axis~\cite{Marenduzzo:2006b}.
-}
+conditions. In practice they are equivalent to 
+having infinitely distant walls to which the network is anchored. 
+See Appendix for a full discussion of our boundary conditions.
 
@@
 \end{figure}
+
+To gain further insight into the flow-induced deformation, and eventual
+breakdown, of BPI, it is instructive to follow the dynamics of 
+the network under a slow flow -- this is depicted in Fig. \ref{bp1-1-disc}.
+Three different stages can be distinguished. 
+Just after the onset of the shear flow, the disclination lines 
+in BPI get more and more squeezed together, and this is incompatible with
+the defect topology in this phase.
+Consequently, the network adopts a flow-induced conformation which consists 
+of intertwined helices that stretch during the shear transformation.
+This helical conformation that emerges already at strains $\gamma\simeq1$ 
+is shown in Fig. \ref{bp1-1-disc} (top row).
+At this point the original BPI has already been significantly deformed.
+Shortly afterwards regular oscillations set up temporarily,
+where the disclinations form double helices which tilt and realign under the shear.
+
+This mode of flow proves unstable, as after a few cycles 
+distortions appear which lead to further destabilisation -- presumably in
+view of the large stress fluctuations discussed above. 
+Finally, the system loses order and 
+transforms into an amorphous network state with almost constant
+stress in the region of $\Pi_{xy}\simeq 1-2\e{-5}$ LBU.
+If the shear is stopped, this flow-induced amorphous state
+remains arrested and metastable, and cannot
+find its way back to the original BPI structure.
 
@@
 
-To gain further insight into the flow-induced deformation, and eventual
-breakdown, of BPI, it is instructive to follow the dynamics of 
-the network under a slow flow -- this is depicted in Fig. \ref{bp1-1-disc}.
-Three different stages can be distinguished. 
-Just after the onset of the shear flow, the disclination lines 
-in BPI get more and more squeezed together, and this is incompatible with
-the defect topology in this phase.
-Consequently, the network adopts a flow-induced conformation which consists 
-of intertwined helices that stretch during the shear transformation.
-This helical conformation that emerges already at strains $\gamma\simeq1$ 
-is shown in Fig. \ref{bp1-1-disc} (top row).
-At this point the original BPI has already been significantly deformed.
-Shortly afterwards regular oscillations set up temporarily,
-where the disclinations form double helices which tilt and realign under the shear.
-
-This mode of flow proves unstable, as after a few cycles 
-distortions appear which lead to further destabilisation -- presumably in
-view of the large stress fluctuations discussed above. 
-Finally, the system loses order and 
-transforms into an amorphous network state with almost constant
-stress in the region of $\Pi_{xy}\simeq 1-2\e{-5}$ LBU.
-If the shear is stopped, this flow-induced amorphous state
-remains arrested and metastable, and cannot
-find its way back to the original BPI structure.
+\begin{figure*}[htpb]
+\includegraphics[width=0.32\textwidth]{disc-364k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-366k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-368k_run914r.png}\\
+\includegraphics[width=0.32\textwidth]{disc-370k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-372k_run914r.png}
+\includegraphics[width=0.32\textwidth]{disc-374k_run914r.png}
+\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
+The sequence shows a typical cycle of shear-induced transformations in the 
+steady state at $\gd=1.56\e{-4}$ and time steps 
+$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
+During every cycle the network also is displaced 
+\rev{along the vorticity direction just as BPII
+in regime BPII-1, but at a different rate.
+}
+}
+\label{bp1-2-disc}
+\end{figure*}
 
@@
 for the existence of the regular BPI-2 oscillations 
-could be that these are fast enough to bypass or suppress 
-large stress fluctuations -- this leads to
-\rev{
-different topological reconnections with different order structure
-}
-and to a stabilisation 
-compared to the oscillations found transiently in
-the BPI-1 regime.
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-364k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-366k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-368k_run914r.png}\\
-\includegraphics[width=0.32\textwidth]{disc-370k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-372k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-374k_run914r.png}
-\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
-The sequence shows a typical cycle of shear-induced transformations in the 
-steady state at $\gd=1.56\e{-4}$ and time steps 
-$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
-During every cycle the network also is displaced 
-\rev{along the vorticity direction just as
-in regime BPII-1.
-}
-}
-\label{bp1-2-disc}
-\end{figure*}
+could be that compared to the large oscillations found transiently in
+the BPI-1 regime these oscillations are now fast enough to bypass or suppress 
+large stress fluctuations, which leads to
+\rev{
+different topological reconnections and order structure
+and eventually to a stabilisation of the flow. 
+}
 
@@
 of the $z$-component. For negative helicity the sense of motion of the 
-network and the secondary velocity components are inversed just as in Fig. \ref{bp2-1-velo}. 
-\rev{
-Note that the viewing direction is changed along an orbit in x-y plane
- to give the clearest view on the velocity pattern between
-the disclination lines.
+network and the secondary velocity components is inverted just as in Fig. \ref{bp2-1-velo}. 
+\rev{
+Note that the viewing direction has been changed along an orbit in $xy$-plane
+ to give a clearer view on the velocity pattern.
 }
@@
 in Fig.~\ref{bp1-1-disc} at early times.
-}
-
-\com{Need to check:} 
-\rev{
-Surprisingly, if the shear is switched off in
-the BPI-2 regime at some point during an oscillation, 
-the flow-induced configuration usually
-reverts to a perfect quiescent BPI -- in stark contrast with
-what happens in BPI-1, for lower shear rates or BPI-3, for higher shear rates. 
-}
-
-\com{Check recurrence period}
-
-Fig. \ref{bp1-2-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ positive chirality of
-the blue phase at the same time steps shown in Fig.~\ref{bp1-2-disc}. 
-The emerging pattern is similar to that of BPII, shown in
-Fig. \ref{bp2-1-velo}. 
-\rev{
-There is also separation into fluctuating bands 
-along the vorticity direction with opposite secondary velocity.
-The secondary velocity component $v_z$ also changes sign upon
-inverting the chirality of the underlying cholesteric helix.
-}
-
-\subsubsection{Regime BPI-3: high shear rates }
-
-\rev{
-At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
-($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
-periodically recurring disclination network in regime BPI-2 and the formation
-of another amorphous network. Fig. \ref{bp1-3-disc} shows a sequence of 
-snapshots. 
-Initially, the disclination lines take the form of a regular, staggered array of helices 
-with the helical axis along the flow direction. 
-However, this mode of flow proves unstable as the regular formation dissolves into 
-an irregular state that features helicoidal disclination lines of varying 
-appearance. This is different from the amorphous network of
-regime BPI-1 (cf. Fig. \ref{disc-bp1-1}) where the disclination lines are also mostly
-well separated but without a clear orientation with respect to the flow. 
 }
@@
 
+\com{Still need to check:} 
+\rev{
+Surprisingly, if the shear is switched off in
+the BPI-2 regime at some point during an oscillation, 
+the flow-induced configuration usually
+reverts to a perfect quiescent BPI -- in stark contrast with
+what happens in BPI-1 for lower shear rates or BPI-3 for higher shear rates. 
+}
+
+Fig. \ref{bp1-2-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ of BPI at the same time steps as in Fig.~\ref{bp1-2-disc}. 
+The emerging pattern is similar to that of BPII shown in
+Fig. \ref{bp2-1-velo}. 
+\rev{
+There is also separation into fluctuating bands 
+which are oriented along the vorticity direction.
+The secondary velocity component $v_z$ also changes sign upon
+changing from positive to negative helicity and vice versa.
+The recurrence period is different from the one we observe in BPII.
+If we refer again to $\tau_F$ as the time interval between two consecutive 
+reconnecting events of the network in flow direction, then $\tau_F=2/\gd$. 
+A full reconnection in vorticity direction takes place after the time $\tau_V=4 \tau_F$.
+We believe the reason for these differences is also directly linked to the different
+topology of the defect network. The details of this mechanism will be
+addressed in future work.
+}
+
+\subsubsection{Regime BPI-3: high shear rates }
+
+\rev{
+At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
+($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
+periodically recurring disclination network in regime BPI-2 and the formation
+of another amorphous network. A sequence of snapshots of this is shown in Fig. \ref{bp1-3-disc}. 
+At first the disclination lines take the form of a regular, staggered array of helices 
+with the helical axis lying along the flow direction.
+However, this mode of flow proves equally unstable as the regular formations dissolve into 
+an irregular state with helicoidal disclination lines of varying 
+appearance. The morphology is different from the amorphous network of
+regime BPI-1 (cf. Fig. \ref{disc-bp1-1}) where the disclination lines are 
+well separated but do not have a clear orientation with respect to the flow. 
+}
+
 \rev{
 Interestingly, the critical Ericksen number for this observed break-up is in 
-the range of $Er\simeq 2-4$. This is a regime where viscous forces begin to 
-dominate over elastic forces and where the nonlinear backflow mechanism which couples
-order and flow can cause small deviations to grow over time. 
+the range of $Er\simeq 2-4$. This is a regime where viscous forces have a 
+similar magnitude as elastic forces and where the nonlinear coupling between
+order and flow can cause small deviations to grow quickly over time. 
 This suggests that the BPI-3 regime constitutes an instance 
@@
 \subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{gj-fan}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
+\caption{Flow induced cholesteric state or Grandjean texture in regime BPI-4 and BPII-2: 
+The orientation of the helix is along the gradient direction. 
+The director field is quasi-static during the flow, which occurs 
+in ``nematic layers''. Colour coding indicates the 
+$x$-component of the director field (see colorbar on the right).
+For clarity only selected sites are shown along the $y$-direction.
+The structure is translationally invariant along the other two directions.}
+\label{grandjean}
+\end{figure}
 
@@
 Grandjean texture, is shown in Fig. \ref{grandjean}.
-}
 It consists of a simple cholesteric helix with the helical axis oriented 
-along the velocity gradient direction. The state is translational invariant 
-along the flow and vorticity direction and therefore quasi-one-dimensional. 
-\rev{
-However, in contrast to other states that we identified as being metastable
-there is no tumbling motion of the director. 
-While flowing the director retains its relative orientation.
-Consequently, the shear stress is lower than for
-e.g. the travelling helical wave (see \cite{Rey:1996a, Rey:1996b} for instance)
-as there is less dissipation.
+along the gradient direction ($y$). While flowing the director field is stationary and 
+retains its relative orientation so that the depicted state is translational invariant 
+in flow-vorticity plane and quasi-one-dimensional. 
+There is a small Leslie-type angle with respect to 
+the flow-vorticity plane. 
+Contrary to the other metastable flow states there is no tumbling motion 
+of the director. This leads to lower shear stress and lower dissipation 
+compared to other states like the travelling helical wave ~\cite{Rey:1996a, Rey:1996b}.
 }
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
-\caption{Flow induced cholesteric state or Grandjean texture: 
-The orientation of the helix is along the gradient direction. 
-The director field is quasi-static during the flow, which occurs 
-in ''nematic layers''. Colour coding indicates the 
-$x$ component of the director field (see colorbar on the right).
-For clarity only selected sites are shown along the $y$-direction.
-The structure is translationally invariant along the other directions.}
-\label{grandjean}
+\includegraphics[width=0.245\textwidth]{dir3d+y-450k_run916r.png}\nolinebreak
+\includegraphics[width=0.245\textwidth]{dir3d+y-800k_run916r.png}
+\caption{
+\rev{Director field of BPI-4 at high shear rate: The pictures show a frustrated 
+Grandjean configuration that formed layers along a diagonal direction. 
+The colour code gives the magnitude of the $x$-component in flow direction. 
+}
+}
+\label{bp1-4-frust}
 \end{figure}
 
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{dir3d+y-450k_run916r.png}
-\includegraphics[width=0.32\textwidth]{dir3d+y-510k_run916r.png}
-\includegraphics[width=0.32\textwidth]{dir3d+y-800k_run916r.png}
-\caption{Director field of BPI-4 at high shear rate: The pictures show a region 
-of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
-direction oriented along the horizontal/vertical axis. 
-The colour code gives the magnitude of the z-component of the director field. 
-The data show that the structure is quasi-2D, with rolling double twist
-patterns in the $xy$ plane, and an approximate translational symmetry 
-along the vorticity direction.}
-\label{bp1-high}
-\end{figure*}
-
-%\clearpage
+\rev{The Grandjean configuration does not always emerge perfectly.
+Fig. \ref{bp1-4-frust} shows a time sequence of a frustrated state 
+that undergoes a buckling transition.
+Initially the layers happen to be along a diagonal direction. 
+Because the diagonal state does not have the equilibrium layer spacing the conformation 
+buckles in order to minimise its free energy.
+Due to the periodic boundary conditions and their interlocking effect on the configuration 
+this state cannot transform into a perfect Grandjean texture. However, the tilt angle of the
+layers with respect to the flow-vorticity ($xz$) plane decreases significantly if the 
+strain rate is increased.
+Interestingly, the frustration can occur for any flow rate in regime BPI-4 as shown if Fig.~\ref{bp1-flowcurve}.
+ a consequence of the shearing geometry and the relative orientation of the initial state
+}
 
@@
 We were able to characterise the rheology of cubic BPs, and identified 
-thre different flow regimes for BPII and five different ones for BPI.
+three different flow regimes for BPII and five different ones for BPI.
 This comprises steady states that BPI and BPII have in common like the 
@@
 We believe that at these flow rates, where neither viscous nor elastic forces clearly dominate, 
-the coupling between order and flow makes any ordered conformation very vulnerable to small fluctuations.
+the coupling between order and flow makes any ordered conformation very sensitive towards 
+irregular disturbances caused by small fluctuations.
 }
@@
 \section*{Appendix}
-
-In this Appendix we give some more details on our simulations,
-including the relationship between boundary conditions and
-permeation flow.
 
@@
 
-We close with a technical note on the choice of boundary conditions. 
-In all our runs we used Lees-Edwards boundary conditions: this
-is the sheared equivalence of periodic boundary conditions, for both the 
+\rev{
+In all our runs we used Lees-Edwards boundary conditions \cite{Wagner:2002}.
+This is the sheared equivalence of periodic boundary conditions, for both the 
 velocity and the order parameter field. 
+On crossing a Lees-Edwards boundary a Galilean transformation is applied
+with a velocity increment which is fixed in time (and limited in size by the
+low Mach number constraint of LB). An appropriate transformation
+of LB distributions which propagate across a boundary
+must be applied at each time step, and appropriate
+adjustment to the tensor of velocity gradients $W_{\alpha\beta}$ 
+is required to compute cross-plane gradients of the velocity field 
+used in the update to ${\mathbf Q}$. In both cases interpolation of the
+relevant quantities is required to cope with the relative displacement
+of neighbouring lattice sites separated by a sliding plane (the
+displacement may be a fraction of a lattice unit at any given time
+step) \cite{Henrich:2012a}. The use of multiple sliding planes equally spaced in a single
+system allows the overall shear rate to be maintained indefinitely as the system
+becomes larger in the velocity gradient direction (in contrast with the use of solid
+walls to impose shear).
 While extremely useful to simulate bulk flow, we
-note that our boundary conditions impose macroscopic distortion
-of the network, and are effective analogous to infinitely distant
-but anchored wall along the $xy$ plane. In other words, due to
-the periodic boundary conditions along the flow gradient direction, 
-there can be no slip in the network deformation (or strain) at any point. 
+note that these boundary conditions impose macroscopic distortion
+of the network. They do not allow for a free-standing network as
+in flow-gradient direction there can be no slip in the network deformation 
+or strain at any point.
+This is effectively equivalent to infinitely distant
+walls along the $xy-$plane where the liquid crystal is anchored. 
 Effectively, this disallows permeation flows, which on the other
@@
 the details of the boundary conditions used at the wall.
-\newpage
+With free boundary conditions and for very low flow velocity, 
+permeative flows might instead allow for
+some slip of the BPI network, as in cholesterics sheared
+along their axis~\cite{Marenduzzo:2006a, Marenduzzo:2006b}.
+}
 

### 2013-07-03 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2039 40876441-3496-0410-a613-a2fca35b2ae2

@@
 Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
-shear rates -- very slow flow leads to a transition into an amorphous network 
+shear rates -- 
+\rev{
+very slow flow causes a transition of the initially 
+ordered structure with oscillatory response into an amorphous network 
 with an apparent yield stress. 
-\rev{
-Larger shear rates can make the network sensitive
-to flow-induced transitions to metastable and amorphous states.
+Larger shear rates lead to another amorphous state with different structure of 
+the defect network.
 }
@@
 & &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
-\mathrm{Tr}({\mathbf Q W}),
+\mathrm{Tr}({\mathbf Q \mathbf W}),
 \label{sw}
@@
 the BP unit cell. Likewise, $v$ was taken to be the velocity difference
-across one unit cell, i.e. $v=\dot{\gamma} l$. 
+across one unit cell, i.e. $v=\dot{\gamma}\, l$. 
 \rev{
@@
 \includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
-\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
-\rev{
+\caption{
+\rev{
+Flow curve $\bar{\Pi}_{xy}(\gd)$: 
 Depending on the configuration in steady shear flow five different regimes 
@@
 
-\com{Still need to check:} 
+\com{This is currently under investigation:} 
 \rev{
@@
 the BPI-2 regime at some point during an oscillation, 
-the flow-induced configuration usually
-reverts to a perfect quiescent BPI -- in stark contrast with
+the flow-induced configuration can revert to a perfect quiescent BPI -- in stark contrast with
 what happens in BPI-1 for lower shear rates or BPI-3 for higher shear rates.
+We also see the network being stuck in metastable states with 
+helicoidal disclinations similar to the bottom left picture in Fig.~\ref{bp1-1-disc} 
+and the top left image in Fig.~\ref{bp1-3-disc}.
 }
@@
 appearance. The morphology is different from the amorphous network of
-regime BPI-1 (cf. Fig. \ref{disc-bp1-1}) where the disclination lines are 
+regime BPI-1 (cf. Fig.~\ref{bp1-1-disc}) where the disclination lines are 
 well separated but do not have a clear orientation with respect to the flow. 
@@
 \rev{
-In all our runs we used Lees-Edwards boundary conditions \cite{Wagner:2002}.
+In all our runs we used Lees-Edwards boundary conditions (LEBCs)~\cite{Wagner:2002}.
 This is the sheared equivalence of periodic boundary conditions, for both the 
@@
 displacement may be a fraction of a lattice unit at any given time
-step) \cite{Henrich:2012a}. The use of multiple sliding planes equally spaced in a single
+step) \cite{Henrich:2012a}. It is worth mentioning that apart from 
+the Galilean transformation LEBCs do not impose any further constraint.
+This means that order parameter field is free to follow its own kinetic pathway
+at the Lees-Edwards plane. 
+The use of multiple sliding planes equally spaced in a single
 system allows the overall shear rate to be maintained indefinitely as the system
@@
 or strain at any point.
-This is effectively equivalent to infinitely distant
-walls along the $xy-$plane where the liquid crystal is anchored. 
+The boundary condition is equivalent to infinitely distant
+walls along the $xy-$plane where the liquid crystal is anchored,
+not unlike the situation found in a rheometer with no-slip
+boundary conditions at the walls.
 Effectively, this disallows permeation flows, which on the other
-hand require a non-zero velocity field and, simultaneously,
+hand require a non-zero velocity field and simultaneously
 much lower (ideally zero) perturbations of the blue phase network.
-Consequently our results are effectively valid for sheared bulk 
-BPs which are anchored at the walls. In practice, permeation flows are 
+In practice, permeation flows are 
 only restricted to very low flow rates, and are unstable for intermediate 
@@
 the details of the boundary conditions used at the wall.
-With free boundary conditions and for very low flow velocity, 
+With free boundary conditions that do not impose a macroscopic distortion 
+and for very low flow velocity, 
 permeative flows might instead allow for

### 2013-07-04 trunk/papers/bprheo/bprheo.tex
Update bprheo paper - included Kevin's comments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2040 40876441-3496-0410-a613-a2fca35b2ae2

@@
 into a Grandjean texture with a 
-cholesteric helix lying along the flow gradient direction according to the imposed flow.
-}
-BPI leads to a very different response. Here, oscillations are only possible for intermediate
-shear rates -- very slow flow leads to a transition into an amorphous network 
-with an apparent yield stress. Larger shear rates can make the network sensitive
-to flow-induced transitions to metastable and amorphous states. For even
-larger flow rates a Grandjean texture is observed, just as in the case of BPII.
-At the highest flow rates both cubic blue phases adopt a flow-aligned nematic state.
-Our results provide the first theoretical investigation of sheared blue phases in large systems, and are relevant to understanding the bulk rheology of these materials.}
+cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
+}
+Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
+shear rates -- 
+\rev{
+very slow flow causes a transition of the initially 
+ordered structure with oscillatory response into an amorphous network 
+with an apparent yield stress. 
+Larger shear rates lead to another amorphous state with different structure of 
+the defect network.
+}
+For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
+At the highest imposed flow rates both cubic blue phases adopt a flow-aligned nematic state.
+Our results provide the first theoretical investigation of sheared blue phases in large systems, 
+and are relevant to understanding the bulk rheology of these materials.}
 \vspace{0.5cm}
@@
 field virtually unchanged, which leads to high dissipation and large
-viscosities. Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
+viscosities. \rev{
+Early experiments with cholesteric liquid crystals
+showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
+}
+Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
 shear and Poiseuille flow in cholesteric liquid crystals in the permeation mode, and 
@@
 They also found that a strong secondary flow appears.
-\rev{
-Early experiments with cholesteric liquid crystals
-showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
-}
 Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
@@
 is periodic in all three coordinate directions. Sliding
-boundary conditions, or Lees-Edwards planes, are explicitly implemented
+boundary conditions, or Lees-Edwards planes ~\cite{Wagner:2002}, are explicitly implemented
 in both hydrodynamic and thermodynamic sectors to impose shear on the
-system. On
-crossing a Lees-Edwards boundary a Galilean transformation is applied
-with a velocity increment which is fixed in time (and limited in size by the
-low Mach number constraint of LB). An appropriate transformation
-of LB distributions which propagate across a boundary~\cite{Wagner:2002}
-must be applied at each time step, and appropriate
-adjustment to $W_{\alpha\beta}$ is required to compute cross-plane
-gradients of the velocity field used in the update to ${\mathbf Q}$.
-In both cases interpolation of the
-relevant quantities is required to cope with the relative displacement
-of neighbouring lattice sites separated by a sliding plane (the
-displacement may be a fraction of a lattice unit at any given time
-step). The use of multiple sliding planes equally spaced in a single
-system allows
-the overall shear rate to be maintained indefinitely as the system
-becomes larger
-in the velocity gradient direction (in contrast with the use of solid
-walls to impose shear). This hybrid method with Lees-Edwards planes
-has been used successfully to study e.g., smectics~\cite{Henrich:2012a}.
-
-In the following we report results of simulations of the bulk flow behaviour of the cubic blue 
-phases BPI and BPII.
+system. 
+\rev{
+The same hybrid method with Lees-Edwards planes
+has been used successfully to study binary mixtures, particle suspensions~\cite{MacMeccan:2009, Aidun:2010}, 
+emulsions~\cite{Frijters:2012} and smectic liquid crystals~\cite{Henrich:2012a}.
+A more detailed discussion of our use of Lees-Edwards boundary conditions can be found in the Appendix.
+}
+
+In the following we report results of simulations of the bulk flow behaviour of the cubic Blue 
+Phases BPI and BPII.
 Typical runtimes for system size $L_x\times L_y\times L_z=128^3$ on 512 processes were in the region of 18 to 24 hours.  
@@
 velocity gradient along the $y$-direction.
-The shear rates were varied over more than two orders of magnitude from about 
-$\gd=2.44\times \e{-6}$ to $1.875\times\e{-3}$ LBU.
-\rev{
-For clarity we classify various flow regimes, namely three in the case of BPII and five in the 
-case of BPI. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
-amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) that
+The shear rates were varied over more than three orders of magnitude from about 
+$\gd=2.44\times \e{-6}$ to $3.75\times\e{-3}$ LBU.
+\rev{
+For clarity we classify various flow regimes, namely three in the case of Blue Phase II and five in the 
+case of Blue Phase I. These regimes comprise those with periodically recurring conformations and oscillatory stress response (PRC), 
+amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) which 
 may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
-GJ and FAN occur regardless of the initial state.
-}
-
-\rev{
-These regimes are characterised by the following flow rates: 
-For Blue Phase II BPII-1 (PRC) ($\gd \lesssim 3.91\e{-4}$, corresponding to ${\it Er} \lesssim 3.33$), 
-BPII-2 (GJ) ($4.69\lesssim \gd\lesssim 1.25\e{-3}; 4\lesssim {\it Er} \lesssim 10.67$), 
-BPII-3 (FAN) ($1.875\e{-3}\lesssim \gd; 16 \lesssim {\it Er}$) 
-and for Blue Phase I BPI-1 (AN) ($\gd \lesssim 1.95\e{-5}; {\it Er} \lesssim 0.17$),  
-BPI-2 (PRC) ($3.91\e{-5}\lesssim \gd \lesssim 2.344\e{-4}; 0.33\lesssim {\it Er} \lesssim 2$),
-BPI-3 (AN) ($3.125\e{-4}\lesssim \gd\lesssim 4.69\e{-4}; 2.67 \lesssim {\it Er} \lesssim 4$),
-BPI-4 (GJ and FGJ) ($5.47\e{-4} \lesssim \gd \lesssim 1.875\e{-3}; 4.67\lesssim {\it Er} \lesssim 16)$ and
-BPI-5 (FAN) ($2.5\e{-3}\lesssim \gd, 21.33\lesssim {\it Er}$).
-}
+GJ and FAN occur in both blue phases as steady states regardless of the initial state.
+}
+
+These regimes are characterised by the following approximate flow rates:
+
+\begin{table}[htpb]
+\begin{tabular}{|c|c|c|}
+\hline
+Name & $\gd$ & $Er$ \\
+\hline
+BPI-1 (AN) & $ \lesssim 1.95\e{-5}$ & $\lesssim 0.17$ \\
+BPI-2 (PRC) & $3.91\e{-5} \dots 2.34\e{-4}$ & $0.33 \dots 2$ \\
+BPI-3 (AN) & $3.13\e{-4}\dots 4.69\e{-4}$ & $2.67 \dots 4$ \\
+BPI-4 (GJ, FGJ) & $5.47\e{-4} \dots 1.88\e{-3}$ & $4.67 \dots 16$ \\
+BPI-5 (FAN) & $\gtrsim 2.5\e{-3}$ & $ \gtrsim 21.33$ \\
+\hline
+BPII-1 (PRC) & $\lesssim 3.91\e{-4}$ & $ \lesssim 3.33$\\
+BPII-2 (GJ) & $4.69\dots 1.25\e{-3}$ & $4 \dots 10.67$ \\
+BPII-3 (FAN) &$\gtrsim 1.88\e{-3}$ & $ \gtrsim 16$ \\ 
+\hline
+\end{tabular}
+\label{table2}
+\end{table}
 
 As is standard \cite{Henrich:2010b,Henrich:2012b} disclination lines are represented by plotting an isosurface of the scalar order parameter $q_s$. 
-Typical choices are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.
-(Note that $q_s$ is small but non-zero at the disclination core. The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
+Typical choices are $q_s=0.17$ for BPI and $q_s=0.15$ for BPII.
+(Note that $q_s$ is small but non-zero at the disclination core. 
+The director is undefined there because the largest and second largest eigenvalues of ${\mathbf Q}$ coincide.)
 
 \subsection{Blue Phase II}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
-\caption{
-\rev{
-Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
-whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
-breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
-The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
-}
-}
-\label{bp2-flowcurve}
-\end{figure}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
-\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
-\caption{
-\rev{
-Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
-of BPII versus strain $\gamma = \gd\cdot t$  
-normalised to the background viscosity $\eta$.
-The pictures show the regimes BPII-1 (top, PRC), BPII-2 (bottom, GJ) and BPII-3 (bottom, FAN). 
-The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without additional contribution of the liquid crystal.
-}
-} 
-\label{bp2-appvisc}
-\end{figure}
 
@@
 
-Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
-the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
-\rev{For intermediate down to the lowest shear rates, the regime we refer to as BPII-1, a 
-power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
-the data to a very good approximation.
-This holds true even in regime BPII-2, where the network breaks up and 
-the liquid crystal adopts a Grandjean structure. The coefficents of the power law fit are
-here $a=0.68$ and $b=0.94$. Hence, the degree of shear-thinning is 
-remarkably small in BPII, in the range of shear rates which we have explored here.}
-
-\rev{
-Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
-defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
-over total strain.
-A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
-without any additional contribution from the liquid crystal.
-The top picture shows data for regime BPII-1, where the periodic breakup and reconnecting of the network 
-in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
-absence at very low shear rate of a permeation mode is probably due to our
-choice of boundary conditions.
-}
-
-%{We note that permeation flows, which would
-%{\em not} lead to oscillatory response, are effectively disallowed by the combination
-%of the geometry and boundary condition we consider. This is
-%because the periodic boundary conditions impose macroscopic distortion
-%of the network, and are effective analogous to infinitely distant
-%but anchored wall along the $xy$ plane.
-%This means that  our results are effectively valid for sheared bulk 
-%BPs which are anchored at the walls. Furthermore, permeation flows are 
-%only restricted to very slow flows, and are unstable for intermediate 
-%and high flows, where the response of the network depends less on
-%the details of the boundary conditions used at 
-%the wall~\cite{Marenduzzo:2006a,Marenduzzo:2006b}.}
-
-\rev{
-We observe another configuration, lying between those with periodically recurring conformations 
-and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
-In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
-The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
-and the liquid crystal is flowing in ``nematic planes'' with the director field predominantly oriented in flow-vorticity plane.
-}
-
-\rev{
-The picture at the bottom of Fig.~\ref{bp2-appvisc} gives data for this regime, to which we refer as BPII-2 and for BPII-3,
-the flow-aligned nematic state. We will address these regimes in section \ref{gj-fan}, as they are common to BPI and BPII.
-}
-
-\rev{
-The travelling helical waves which was predicted in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
-the vorticity axis turns out to be actually a metastable state and features larger stress and higher dissipation due 
-to the tumbling motion of the local director field, which is absent in the Grandjean texture.
-For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
-}
-
-In the next sections we investigate the BPII-1 flow regime in more detail 
-by looking at the kinetics of the disclination network.
-
-\subsubsection{Regime BPII-1: low and intermediate shear rates }
-
-Fig. \ref{bp2-1-disc} shows the disclination network in shear flow as
-it undergoes homogeneous shearing in the BPII-1 flow regime. 
-The disclination lines break up and 
-reconnect further downstream, forming a periodically recurring pattern. The
-period needed for a pattern to break up and 
-reform along the flow direction is $\tau_F = 1/\gd$.
-
-The general appearance of the flowing network is, apart from the homogeneous distortion,
-very close to that of the quiescent blue phase at equilibrium. This holds for all
-shear rates in the BPII-1 regime.
-
-Interestingly, while being displaced with the flow the entire network moves 
-along the vorticity direction (z) as well.  A similar behaviour has been recently 
-observed for blue phases in shear flow in confined geometries ~\cite{Henrich:2012b}.
-In contrast to the stick-slip motion that has been reported there,
-in this case the movement is steady. 
-The vorticity motion occurs in such a way that the positions of breakup and reconnection 
-point in the network, visible in Fig. \ref{bp2-1-disc}, are slightly offset and allow the network 
-to travel along the $z$-direction. The periodicity of the motion along the vorticity direction
-is $\tau_V=6\tau_F$, 
-i.e. it takes a displacement of six unit cells along $x$ (the flow direction) 
-for the network to move one unit cell along $z$ (the vorticity direction).
-
-Fig. \ref{bp2-1-velo} shows the disclination network in the middle of a breakup-reconnection
-cycle, with superimposed velocity vectors.
-The plot shows the secondary velocity components, obtained by projecting the
-velocity onto a plane perpendicular to the flow direction.
-This gets rid of the dominating velocity component along the flow, $v_x$, and 
-allows to visualise the patterns in the 
-two much smaller components $v_y$ and $v_z$.
-The magnitude of the secondary components is typically in the range of a 
-few percent of the primary flow component, for the shear rates 
-and system sizes simulated here.
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{flowcurve_bp2.pdf}
+\caption{
+\rev{
+Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
+whereas the error bars indicate the maximum and minimum stresses that occur during one cycle. Above a critical flow rate the network  
+breaks up into a Grandjean texture (open triangles) or a flow-aligned nematic state at even higher flow rates (solid triangles).
+The two dashed lines represent fits to the data with $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.95}$ (left) and $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;0.94}$ (right), respectively.
+}
+}
+\label{bp2-flowcurve}
+\end{figure}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_a.pdf}\\
+\includegraphics[width=0.495\textwidth]{app_visc_strain_bp2_b.pdf}\\
+\caption{
+\rev{
+Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
+of BPII versus strain $\gamma = \gd\cdot t$  
+normalised to the background viscosity $\eta$.
+The pictures show the regimes BPII-1 (periodically recurring conformations, top), BPII-2 (Grandjean texture, bottom) and BPII-3 (flow-aligned nematic, bottom). 
+The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. total shear-thinning.
+}
+} 
+\label{bp2-appvisc}
+\end{figure}
 
@@
 
-\begin{figure*}[htpb]
+Fig. \ref{bp2-flowcurve} shows a flow curve, defined as time averages of 
+the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
+\rev{
+Time averages were taken as average over one stress cycle in the steady state.
+Where no oscillations occurred we used the final value at the end of the run.
+For the lowest up to intermediate shear rates, the regime we refer to as BPII-1, a 
+power law fit $\bar{\Pi}_{xy}=a \gd^b$ with $a=0.30, b=0.95$ describes 
+the data to a very good approximation.
+This holds true even in regime BPII-2, where the network breaks up and 
+the liquid crystal adopts a Grandjean structure with the helical axis oriented along the
+flow gradient direction ($y$). The coefficients of the power law fit are
+here $a=0.68$ and $b=0.94$. Hence, for the range of shear rates which we have explored here, 
+the degree of shear-thinning is remarkably small in BPII.
+}
+
+\rev{
+Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
+defined as $\eta_{app}=\langle \Pi_{xy} \rangle/\gd +\eta$, and the background viscosity $\eta$ 
+over total strain.
+A numerical value of $\eta_{app}/\eta=1$ corresponds to a fully Newtonian flow,
+without any additional stress contribution from the liquid crystal.
+The top picture shows data for regime BPII-1, where the periodic break-up and reconnecting of the network 
+in shear flow causes sinusoidal oscillations in $\eta_{app}$. As noted in the Appendix, the
+absence of a permeation mode at very low shear rates may be due to our
+choice of boundary conditions. We refer the reader to the Appendix for a detailed discussion.
+}
+
+\rev{
+We observe another configuration, lying between those with periodically recurring conformations 
+and the flow-aligned nematic state at shear rates $4.67\e{-4}\lesssim\gd\lesssim1.25\e{-3}$ ($4 \lesssim {\it Er} \lesssim 10.67$). 
+In this interval the network breaks up completely into a simple cholesteric liquid crystal, and oscillations in the stress signal are absent.
+The helical axis lies along the flow gradient direction (the so-called Grandjean texture~\cite{deGennes})
+and the liquid crystal is flowing in ``nematic planes'' with the director predominantly oriented in flow-vorticity plane and having only
+a small angle with the flow direction.
+The picture at the bottom of Fig.~\ref{bp2-appvisc} shows data for this regime, to which we refer as BPII-2. The 
+curves for the two largest flow rates are for BPII-3, the flow-aligned nematic state. 
+We will address these regimes in section \ref{gj-fan}, as they are common to both cubic blue phases BPI and BPII.
+}
+
+\rev{
+The travelling helical wave which was identified in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
+the vorticity axis turns out to be a metastable flow state as it features larger free energies, higher stresses and dissipation due 
+to the tumbling motion of the director field, which are completely absent in the Grandjean texture.
+For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
+}
+
+In the next sections we investigate the BPII-1 flow regime in more detail 
+by looking at the kinetics of the disclination network.
+
+\subsubsection{Regime BPII-1: low and intermediate shear rates }
+
+\begin{figure}[htbp]
 \includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902r.png}
@@
 \includegraphics[width=0.495\textwidth]{flowcurve_bp1.pdf}\\
-\caption{Flow curve $\bar{\Pi}_{xy}(\gd)$: 
-\rev{
+\caption{
+\rev{
+Flow curve $\bar{\Pi}_{xy}(\gd)$: 
 Depending on the configuration in steady shear flow five different regimes 
-can be identified: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
+can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
 steady flow with periodically recurring patterns (BPI-2, open squares), 
-breakup of the disclination network into amorphous state (BPI-3, open circles)
-\com{Calaculate another shear rate to get a regime},
+break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
 cholesteric helix with axis in gradient direction, also referred to
-as Grandjean texture and frustrated metastable states that transform into 
-a Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
+as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
+a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
 and flow-aligned nematic state (BPI-5, solid triangles). 
@@
 \rev{
-Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised by the background viscosity $\eta$: 
-The top picture shows regime BPI-1 (AN) where
+Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised to the background viscosity $\eta$: 
+The top picture shows regime BPI-1 (amorphous network) where
 the BP network transforms into an amorphous network featuring a yield stress.
-The centre picture shows regimes BPI-2 (PRC) and BPI-3 (AN), 
+The centre picture shows regimes BPI-2 (periodically recurring conformations) and BPI-3 (amorphous network), 
 whereas the picture at the bottom
-depicts data of the Grandjean tecture (GJ and FGJ if frustrated) in regime BPI-4 
-and the flow-aligned nematic state (BPI-5, FAN). 
+depicts data of the (partly frustrated) Grandjean texture in regime BPI-4 
+and the flow-aligned nematic state, labelled BPI-5 in this context. 
 Note that the $\eta_{app}/\eta$ is negative for the lowest 
@@
 for the existence of the regular BPI-2 oscillations 
-could be that these are fast enough to bypass or suppress 
-large stress fluctuations -- this leads to
-\rev{
-different topological reconncetions with different order structure
-}
-and to a stabilisation 
-compared to the oscillations found transiently in
-the BPI-1 regime.
-
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{disc-364k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-366k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-368k_run914r.png}\\
-\includegraphics[width=0.32\textwidth]{disc-370k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-372k_run914r.png}
-\includegraphics[width=0.32\textwidth]{disc-374k_run914r.png}
-\caption{Disclination network of BPI in regime BPI-2 (intermediate flow rates): 
-The sequence shows a typical cycle of shear-induced transformations in the 
-steady state at $\gd=1.56\e{-4}$ and time steps 
-$t=3.64, 3.66,3.68,3.70,3.72,3.74\e{5}$ in flow-gradient plane. 
-During every cycle the network also is displaced 
-\rev{along the vorticity direction just as
-in regime BPII-1.
-}
-}
-\label{bp1-2-disc}
-\end{figure*}
+could be that compared to the large oscillations found transiently in
+the BPI-1 regime these oscillations are now fast enough to bypass or suppress 
+large stress fluctuations, which leads to
+\rev{
+different topological reconnections and order structure
+and eventually to a stabilisation of the flow. 
+}
 
@@
 BPI in steady shear flow. Contrary to BPII-1 at these shear rates, 
-BPI-2 does not resemble the equilibrium configuration
+BPI-2 does not resemble so much the equilibrium configuration
 undergoing a homogeneous topology-conserving 
@@
 It features intricate, flow-induced 
-conformations consisting of well-separated ondulating 
+conformations consisting of well-separated undulating 
 disclination lines similar to those intertwined helices 
 in Fig.~\ref{bp1-1-disc} at early times.
-}
-
-\com{Need to check this claim:} 
-\rev{
-Surprisingly, if the shear is switched off in
-the BPI-2 regime at some point during an oscillation, 
-the flow-induced configuration usually
-reverts to a perfect quiescent BPI -- in stark contrast with
-what happens in BPI-1, for lower shear rates or BPI-3, for higher shear rates. 
-}
-
-Fig. \ref{bp1-2-velo} depicts a snapshot of the secondary velocity 
-components $v_y$ and $v_z$ positive chirality of
-the blue phase at the same time steps shown in Fig.~\ref{bp1-2-disc}. 
-The emerging pattern is similar to that of BPII, shown in
-Fig. \ref{bp2-1-velo}. 
-\rev{
-There is also separation into fluctuating bands 
-along the vorticity direction with opposite secondary velocity.
-The secondary velocity component $v_z$ also changes sign upon
-inverting the chirality of the underlying cholesteric helix.}
-
-\rev{The recurrence period needs to be determined again 
-}
-
-\subsubsection{Regime BPI-3: high shear rates }
-
-\rev{
-At high flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
-($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
-periodically recurring disclination network of regime BPI-2. 
-Fig. \ref{bp1-3-disc} shows a time sequence of this breakup, beginning
-with a regular disclination network with helices aligned along the 
-flow direction.
 }
@@
 \includegraphics[width=0.32\textwidth]{disc-1200k_run915r.png}
-\caption{}
+\caption{
+\rev{Disclination network of BPI in regime BPI-2 (high flow rates):
+The sequence shows the break-up of the initially ordered state into 
+an amorphous network at $\gd=3.125\e{-4}$ and time steps
+$t=6.0, 6.75, 7.0, 7.25 \e{5}, 1.0,1.2\e{6}$ in vorticity-gradient plane.
+During every cycle the network also is displaced
+along the vorticity direction.
+}
+}
 \label{bp1-3-disc}
@@
 
-Interestingly, the critical Ericksen number for this breakup are in 
-the range of 2-4 and therefore in a regime where viscous forces begin to 
-dominate of elastic forces. 
-Therefore the BPI-3 regime includes at least some instances of "rheochaos"~\cite{rheochaos,Cates:2002}.
+\rev{
+If the shear is switched off in
+the BPI-2 regime at some point during an oscillation, 
+the flow-induced configuration cannot revert to a perfect quiescent BPI.
+Instead the network is stuck in a metastable states with 
+helicoidal intertwined disclinations similar to those in the bottom left 
+picture of Fig.~\ref{bp1-1-disc} or those in the top left picture of Fig.~\ref{bp1-3-disc}.
+However, the orientation of the helical axis after switch-off is always 
+along the vorticity ($z-$) direction.
+}
+
+Fig. \ref{bp1-2-velo} depicts a snapshot of the secondary velocity 
+components $v_y$ and $v_z$ of BPI at the same time steps as in Fig.~\ref{bp1-2-disc}. 
+The emerging pattern is similar to that of BPII shown in
+Fig. \ref{bp2-1-velo}. 
+\rev{
+There is also separation into fluctuating bands 
+which are oriented along the vorticity direction.
+The secondary velocity component $v_z$ also changes sign upon
+changing from positive to negative helicity and vice versa.
+The recurrence period is different from the one we observe in BPII.
+If we refer again to $\tau_F$ as the time interval between two consecutive 
+reconnecting events of the network in flow direction, then $\tau_F=2/\gd$. 
+A full reconnection in vorticity direction takes place after the time $\tau_V=4 \tau_F$.
+We believe the reason for these differences is also directly linked to the different
+topology of the defect network. The details of this mechanism will be
+addressed in future work.
+}
+
+\subsubsection{Regime BPI-3: high shear rates }
+
+\rev{
+At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
+($2.67\lesssim {\it Er}\lesssim 4$) we observe a break-up of the 
+periodically recurring disclination network in regime BPI-2 and the formation
+of another amorphous network. A sequence of snapshots of this is shown in Fig. \ref{bp1-3-disc}. 
+At first the disclination lines take the form of a regular, staggered array of helices 
+with the helical axis lying along the flow direction.
+However, this mode of flow proves equally unstable as the regular formations dissolve into 
+an irregular state with helicoidal disclination lines of varying 
+appearance. The morphology is different from the amorphous network of
+regime BPI-1 (cf. Fig.~\ref{bp1-1-disc}) where the disclination lines are 
+well separated but do not have a clear orientation with respect to the flow. 
+}
+
+\rev{
+Interestingly, the critical Ericksen number for this observed break-up is in 
+the range of $Er\simeq 2-4$. This is a regime where viscous forces have a 
+similar magnitude as elastic forces and where the nonlinear coupling between
+order and flow can cause small deviations to grow quickly over time. 
+This suggests that the BPI-3 regime constitutes an instance 
+of "rheochaos"~\cite{rheochaos,Cates:2002}.
+}
 
 \subsection{Transition to cholesteric helix and flow-aligned nematic state}\label{gj-fan}
+
+\begin{figure}[htpb]
+\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
+\caption{Flow induced cholesteric state or Grandjean texture in regime BPI-4 and BPII-2: 
+The orientation of the helix is along the gradient direction. 
+The director field is quasi-static during the flow, which occurs 
+in ``nematic layers''. Colour coding indicates the 
+$x$-component of the director field (see colorbar on the right).
+For clarity only selected sites are shown along the $y$-direction.
+The structure is translationally invariant along the other two directions.}
+\label{grandjean}
+\end{figure}
 
@@
 \begin{figure}[htpb]
-\includegraphics[width=0.475\textwidth]{dir3d+y-200k_run1179r.png}
-\caption{Flow induced cholesteric state or Grandjean texture: 
-The orientation of the helix is along the gradient direction. 
-The director field is quasi-static during the flow, which occurs 
-in ``nematic layers''. Colour coding indicates the 
-$x$ component of the director field (see colorbar on the right).
-For clarity only selected sites are shown along the $y$-direction.
-The structure is translationally invariant along the other directions.}
-\label{grandjean}
+\includegraphics[width=0.245\textwidth]{dir3d+y-450k_run916r.png}\nolinebreak
+\includegraphics[width=0.245\textwidth]{dir3d+y-800k_run916r.png}
+\caption{
+\rev{Director field of BPI-4 at high shear rate: The pictures show a frustrated 
+Grandjean configuration that formed layers along a diagonal direction. 
+The colour code gives the magnitude of the $x$-component in flow direction. 
+}
+}
+\label{bp1-4-frust}
 \end{figure}
 
-\begin{figure*}[htpb]
-\includegraphics[width=0.32\textwidth]{dir3d+y-450k_run916r.png}
-\includegraphics[width=0.32\textwidth]{dir3d+y-510k_run916r.png}
-\includegraphics[width=0.32\textwidth]{dir3d+y-800k_run916r.png}
-\caption{Director field of BPI-4 at high shear rate: The pictures show a region 
-of $L_x\times L_y \times L_z= 32^3$ lattice sites with the flow/gradient 
-direction oriented along the horizontal/vertical axis. 
-The colour code gives the magnitude of the z-component of the director field. 
-The data show that the structure is quasi-2D, with rolling double twist
-patterns in the $xy$ plane, and an approximate translational symmetry 
-along the vorticity direction.}
-\label{bp1-high}
-\end{figure*}
-
-%\clearpage
+\rev{The Grandjean configuration does not always emerge perfectly.
+Fig. \ref{bp1-4-frust} shows a time sequence of a frustrated state 
+that undergoes a buckling transition.
+Initially the layers happen to be along a diagonal direction. 
+Because the diagonal state does not have the equilibrium layer spacing the conformation 
+buckles in order to minimise its free energy.
+Due to the periodic boundary conditions and their interlocking effect on the configuration 
+this state cannot transform into a perfect Grandjean texture. However, the tilt angle of the
+layers with respect to the flow-vorticity ($xz$) plane decreases significantly if the 
+strain rate is increased.
+Interestingly, the frustration can occur for any flow rate in regime BPI-4 as shown if Fig.~\ref{bp1-flowcurve}.
+ a consequence of the shearing geometry and the relative orientation of the initial state
+}
 
@@
 
+\rev{
 We were able to characterise the rheology of cubic BPs, and identified 
-two different flow regimes for BPII and three different ones for BPI. 
+three different flow regimes for BPII and five different ones for BPI.
+This comprises steady states that BPI and BPII have in common like the 
+Grandjean texture and the flow-aligned nematic state at high shear rates.
+}  
 Below an Ericksen number of $Er\simeq3$, BPII exhibits weak shear-thinning and 
@@
 
-At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim8$,
+At slightly larger Ericksen numbers $0.4\lesssim Er \lesssim2$,
 BPI kinematically bypasses the large stress fluctuations and 
-flows with very complex, periodically recurring flow-induced conformations.
-These conformations entail regularly arrange helical disclinations which 
-rotate due to the shear -- a new flow-induced state which was not previously
-predicted. 
-The oscillatory dynamics is in some cases intermittent, and the
-recurrence period for the network conformations and stress patterns
-is not given by a simple formula as for BPII.
+flows with periodically recurring flow-induced conformations.
+These conformations entail regularly arranged helical disclinations which 
+rotate due to the shear. 
 Despite their complex appearance, the flow-induced conformations are 
 topologically connected to the quiescent BPI and after switching off the 
-shear flow a defect-free blue phase reforms. It is tempting to
+shear flow a defect-free blue phase reforms.
+
+\rev{
+At even larger Ericksen numbers $2\lesssim Er \lesssim 4$ the behaviour of BPI is again different.
+At first, it exhibits regular helicoidal disclination lines oriented along the flow direction.
+After a short time these become irregular and an amorphous network emerges, which has 
+a different appearance than that observed at the lowest flow rates.
+We believe that at these flow rates, where neither viscous nor elastic forces clearly dominate, 
+the coupling between order and flow makes any ordered conformation very sensitive towards 
+irregular disturbances caused by small fluctuations.
+}
+It is tempting to
 interpret the unsteady oscillations seen in this regime as an instance of 
@@
 \section*{Appendix}
-
-In this Appendix we give some more details on our simulations,
-including the relationship between boundary conditions and
-permeation flow.
-
-Table \ref{tab1} lists the parameters chosen for our runs (values of
-the reduced temperature and reduced chirality are given in the caption),
-together with the minima, maxima, and standard deviations of the velocity
-field in the three directions (discussed further in the main text). 
-The first column also describes what flow regime each simulation
-leads to in steady state.
 
@@
 \end{table*}
-
-We close with a technical note on the choice of boundary conditions. 
-In all our runs we used Lees-Edwards boundary conditions: this
-is the sheared equivalence of periodic boundary conditions, for both the 
+\clearpage
+Table \ref{tab1} lists the parameters chosen for our runs (values of
+the reduced temperature and reduced chirality are given in the caption),
+together with the minima, maxima, and standard deviations of the velocity
+field in the three directions (discussed further in the main text). 
+The first column also describes what flow regime each simulation
+leads to in steady state.
+
+\rev{
+In all our runs we used Lees-Edwards boundary conditions (LEBCs)~\cite{Wagner:2002}.
+This is the sheared equivalent of periodic boundary conditions, for both the 
 velocity and the order parameter field. 
+On crossing a Lees-Edwards boundary a Galilean transformation is applied
+with a velocity increment which is fixed in time (and limited in size by the
+low Mach number constraint of LB). An appropriate transformation
+of LB distributions which propagate across a boundary
+must be applied at each time step, and appropriate
+adjustment to the tensor of velocity gradients $W_{\alpha\beta}$ 
+is required to compute cross-plane gradients of the velocity field 
+used in the update to ${\mathbf Q}$. In both cases interpolation of the
+relevant quantities is required to cope with the relative displacement
+of neighbouring lattice sites separated by a sliding plane (the
+displacement may be a fraction of a lattice unit at any given time
+step) \cite{Henrich:2012a}. It is worth mentioning that apart from 
+the Galilean transformation LEBCs do not impose any further constraint.
+This means that order parameter field is free to follow its own kinetic pathway
+at the Lees-Edwards plane. 
+The use of multiple sliding planes equally spaced in a single
+system allows the overall shear rate to be maintained indefinitely as the system
+becomes larger in the velocity gradient direction (in contrast with the use of solid
+walls to impose shear).
 While extremely useful to simulate bulk flow, we
-note that our boundary conditions impose macroscopic distortion
-of the network, and are effective analogous to infinitely distant
-but anchored wall along the $xy$ plane. In other words, due to
-the periodic boundary conditions along the flow gradient direction, 
-there can be no slip in the network deformation (or strain) at any point. 
+note that these boundary conditions impose macroscopic distortion
+of the network. They do not allow for a free-standing network as
+in flow-gradient direction there can be no slip in the network deformation 
+or strain at any point.
+The boundary condition is equivalent to infinitely distant
+walls along the $xy-$plane where the liquid crystal is anchored,
+not unlike the situation found in a rheometer with no-slip
+boundary conditions at the walls.
 Effectively, this disallows permeation flows, which on the other
-hand require a non-zero velocity field and, simultaneously,
+hand require a non-zero velocity field and simultaneously
 much lower (ideally zero) perturbations of the blue phase network.
-Consequently our results are effectively valid for sheared bulk 
-BPs which are anchored at the walls. In practice, permeation flows are 
+In practice, permeation flows are 
 only restricted to very low flow rates, and are unstable for intermediate 
@@
 the details of the boundary conditions used at the wall.
-\newpage
+With free boundary conditions that do not impose a macroscopic distortion 
+and for very low flow velocity, 
+permeative flows might instead allow for
+some slip of the BPI network, as in cholesterics sheared
+along their axis~\cite{Marenduzzo:2006a, Marenduzzo:2006b}.
+}
 

### 2013-07-04 trunk/papers/bprheo/bprheo.tex
Update bprheo paper -- included Mike's comments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2041 40876441-3496-0410-a613-a2fca35b2ae2

@@
 For larger shear rates, the structure breaks up 
-\rev{
 into a Grandjean texture with a 
-cholesteric helix lying along the flow gradient direction with respect to the imposed flow.
-}
+cholesteric helix lying along the flow gradient direction.
 Blue Phase I leads to a very different response. Here, oscillations are only possible for intermediate
 shear rates -- 
-\rev{
 very slow flow causes a transition of the initially 
-ordered structure with oscillatory response into an amorphous network 
+ordered structure into an amorphous network 
 with an apparent yield stress. 
@@
 the defect network.
-}
 For even larger flow rates the same break-up into a Grandjean texture as for Blue Phase II is observed.
@@
 field virtually unchanged, which leads to high dissipation and large
-viscosities. \rev{
+viscosities. 
 Early experiments with cholesteric liquid crystals
 showed that flow can also give rise to conformational transitions. \cite{Press:1978}.
-}
 Marenduzzo et al.~\cite{Marenduzzo:2006a,Marenduzzo:2006b} simulated 
@@
 They also found that a strong secondary flow appears.
-Rey~\cite{Rey:1996a, Rey:1996b} studied shear in cholesterics oriented with the helix along 
+Rey~\cite{Rey:1996a, Rey:1996b} theoretically studied shear in cholesterics oriented with the helix along 
 the vorticity axis and found that, at low Ericksen number, travelling twist waves appear which 
 lead to the rotation of the cholesteric helix. At higher forcing, the helix uncoils, creating a flow-induced nematic phase. 
-\rev{
 However, this result was derived under the 
@@
 the orientation of the cholesteric helix remained unchanged.
-}  
 Rey also studied cholesterics subjected to both steady flow and low frequency
@@
 across one unit cell, i.e. $v=\dot{\gamma}\, l$. 
-\rev{
 The elastic constants were $K=0.02$ (BPII) and $K=0.04$ (BPI), respectively.
-}
 
@@
 system. 
-\rev{
-The same hybrid method with Lees-Edwards planes
+The same method with Lees-Edwards planes
 has been used successfully to study binary mixtures, particle suspensions~\cite{MacMeccan:2009, Aidun:2010}, 
@@
 A more detailed discussion of our use of Lees-Edwards boundary conditions can be found in the Appendix.
-}
 
@@
 simulated. 
-Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state.
+Since by Eq.~\ref{FE} $f=0$ for an isotropic phase with ${\mathbf Q}\equiv 0$, 
+this means that our system, which is never far from equilibrium locally even under flow, always remains in a liquid crystalline state. 
 We also performed selected simulations on metastable states at higher and lower temperatures, 
@@
 general flow behaviour from that described below.
-The only quantitative difference we found was that, for thermodynamic states that are closer to the phase boundary than the one we focused on (and describe below), the critical shear
+The only quantitative difference we found was that, for thermodynamic states that are closer to the 
+phase boundary than the one we focused on (and describe below), the critical shear
 rate at which the disclination 
-network broke up into chiral nematic states was lower.
-This is expected as these states have on average higher free energy densities and smaller order parameters than those we focus on in what follows, and cannot therefore withstand the same external forces before breaking down.
+network broke up into chiral nematic states was lower. This is expected as these states have on average 
+higher free energy densities and smaller order parameters than those we focus on in what follows, 
+and cannot therefore withstand the same external forces before breaking down.
 
@@
 $\gd=2.44\times \e{-6}$ to $3.75\times\e{-3}$ LBU.
-\rev{
 For clarity we classify various flow regimes, namely three in the case of Blue Phase II and five in the 
@@
 amorphous networks (AN), as well as those featuring a Grandjean texture (GJ) which 
-may be partly frustrated (FGJ) and a flow-aligned nematic state (FAN). 
-GJ and FAN occur in both blue phases as steady states regardless of the initial state.
-}
+may be partly frustrated (FGJ) and a flow-aligned nematic state (FN). 
+GJ and FN occur in both blue phases as steady states regardless of the initial state.
 
@@
 BPI-4 (GJ, FGJ) & $5.47\e{-4} \dots 1.88\e{-3}$ & $4.67 \dots 16$ \\
-BPI-5 (FAN) & $\gtrsim 2.5\e{-3}$ & $ \gtrsim 21.33$ \\
+BPI-5 (FN) & $\gtrsim 2.5\e{-3}$ & $ \gtrsim 21.33$ \\
 \hline
@@
 BPII-2 (GJ) & $4.69\dots 1.25\e{-3}$ & $4 \dots 10.67$ \\
-BPII-3 (FAN) &$\gtrsim 1.88\e{-3}$ & $ \gtrsim 16$ \\ 
+BPII-3 (FN) &$\gtrsim 1.88\e{-3}$ & $ \gtrsim 16$ \\ 
 \hline
@@
 \caption{
-\rev{
 Flowcurve $\bar{\Pi}_{xy}(\gd)$ of BPII. The open squares show the average shear stress in the regime with periodically recurring conformations,
@@
 }
-}
 \label{bp2-flowcurve}
@@
 \caption{
-\rev{
 Apparent viscosity $\eta_{app}=\Pi_{xy}/\gd + \eta$ 
@@
 The pictures show the regimes BPII-1 (periodically recurring conformations, top), BPII-2 (Grandjean texture, bottom) and BPII-3 (flow-aligned nematic, bottom). 
-The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. total shear-thinning.
-}
+The value $\eta_{app}/\eta=1$ corresponds to Newtonian flow without any additional contribution of the liquid crystal, i.e. totally shear-thinned state.
 } 
@@
 the shear stress $\bar{\Pi}_{xy}$ as a function of shear rate $\gd$.
-\rev{
 Time averages were taken as average over one stress cycle in the steady state.
@@
 the degree of shear-thinning is remarkably small in BPII.
-}
-
-\rev{
+
 Fig. \ref{bp2-appvisc} shows the ratio between the apparent viscosity, 
@@
 choice of boundary conditions. We refer the reader to the Appendix for a detailed discussion.
-}
-
-\rev{
+
 We observe another configuration, lying between those with periodically recurring conformations 
@@
 We will address these regimes in section \ref{gj-fan}, as they are common to both cubic blue phases BPI and BPII.
-}
-
-\rev{
+
 The travelling helical wave which was identified in~\cite{Rey:1996a,Rey:1996b} when shearing a cholesteric helix along
@@
 For even higher shear rates the system undergoes a transition to a flow-aligned nematic state.
-}
 
@@
 along the $z$-direction. 
-\rev{
 Therefore the averaged flow along the vorticity
 direction is very small, and much smaller than
-the local velocity flow. This suggests that the
+the local flow velocity. This suggests that the
 flow of the network along the vorticity 
@@
 negative z-direction, whereas the movement of the network as a whole is only in one direction.
-}
 
@@
 
-\rev{
 Fig. \ref{bp1-flowcurve} shows the flow curve of BPI for the same range of
 shear rates as in Fig.~\ref{bp2-flowcurve} and Fig. \ref{bp1-appvisc} depicts 
-the apparent viscosity ratio $\eta_{app}/\eta$ over total strain.
-}
+the apparent viscosity ratio $\eta_{app}/\eta$ as a function of total strain.
 
@@
 \caption{
-\rev{
 Flow curve $\bar{\Pi}_{xy}(\gd)$: 
 Depending on the configuration in steady shear flow five different regimes 
-can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares), 
-steady flow with periodically recurring patterns (BPI-2, open squares), 
-break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles),
+can be distinguished: amorphous BP network with apparent yield stress (BPI-1, solid squares); 
+steady flow with periodically recurring patterns (BPI-2, open squares); 
+break-up of the disclination network into an amorphous state at $Er\simeq O(1)$ (BPI-3, open circles);
 cholesteric helix with axis in gradient direction, also referred to
-as Grandjean texture as well as frustrated metastable Grandjean-like states that can transform into 
-a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles),
+as Grandjean texture, as well as frustrated metastable Grandjean-like states that can transform into 
+a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles);
 and flow-aligned nematic state (BPI-5, solid triangles). 
@@
 $\bar{\Pi}_{xy}(\gd)\propto \gd^{\;1.02}$ (right), respectively.}
-}
 \label{bp1-flowcurve}
@@
 response versus time. We label this regime BPI-2. 
-\rev{
 The centre picture of Fig.~\ref{bp1-appvisc} shows the corresponding data for the
 apparent viscosity ratio versus strain.
-}
 
@@
 cycles the disclination pattern becomes amorphous.
-\rev{
 For these low shear rates 
@@
 We refer to this flow regime as BPI-1.
-}
 
@@
 a critical value, 
-\rev{
 which is in the region of Ericksen numbers $Er \simeq O(1)$.
@@
 a possible route towards an instance of ``rheochaos'' ~\cite{rheochaos,Cates:2002}.
-}
-
-\rev{
+
 Finally, for shear rates such that
@@
 we observe the same cholesteric Grandjean configuration with the 
-helical axis along the gradient direction as 
-reported above for BPII-2 (see
-section \ref{gj-fan}). This regime is referred to as BPI-4.
-}
+helical axis along the gradient direction as for BPII-2. 
+For clarity we refer to this regime as BPI-4. It will be discussed 
+in a separate section \ref{gj-fan}.
 
@@
 \caption{
-\rev{
 Apparent viscosity $\eta_{app}=\langle \Pi_{xy}\rangle/\dot{\gamma} + \eta$ of BPI versus strain normalised to the background viscosity $\eta$: 
@@
 and the flow-aligned nematic state, labelled BPI-5 in this context. 
-Note that the $\eta_{app}/\eta$ is negative for the lowest 
+Note that the $\eta_{app}/\eta$ is sometimes negative for the lowest 
 flow rates as the (negative) contribution of the elastic forces
 of the network dominates the total stress.
-}
 }
@@
 During every cycle the network also is displaced 
-\rev{along the vorticity direction just as BPII
+along the vorticity direction just as BPII
 in regime BPII-1, but at a different rate.
-}
 }
@@
 large stress fluctuations, which leads to
-\rev{
 different topological reconnections and order structure
 and eventually to a stabilisation of the flow.
-}
 
@@
 network and the secondary velocity components is inverted just as in Fig. \ref{bp2-1-velo}. 
-\rev{
 Note that the viewing direction has been changed along an orbit in $xy$-plane
  to give a clearer view on the velocity pattern.
-}
 }
@@
 transformation. 
-\rev{
 It features intricate, flow-induced 
@@
 in Fig.~\ref{bp1-1-disc} at early times.
-}
 
@@
 \caption{
-\rev{Disclination network of BPI in regime BPI-2 (high flow rates):
+Disclination network of BPI in regime BPI-2 (high flow rates):
 The sequence shows the break-up of the initially ordered state into 
@@
 }
-}
 \label{bp1-3-disc}
@@
 
-\rev{
-If the shear is switched off in
-the BPI-2 regime at some point during an oscillation, 
-the flow-induced configuration cannot revert to a perfect quiescent BPI.
-Instead the network is stuck in a metastable states with 
+If the shearing is switched off in
+regime BPI-2 at some point during an oscillation, 
+the flow-induced configuration cannot revert to a quiescent BPI.
+Instead the network is trapped in a metastable state with 
 helicoidal intertwined disclinations similar to those in the bottom left 
 picture of Fig.~\ref{bp1-1-disc} or those in the top left picture of Fig.~\ref{bp1-3-disc}.
-However, the orientation of the helical axis after switch-off is always 
+However, the orientation of the helical axis in now always 
 along the vorticity ($z-$) direction.
-}
 
@@
 Fig. \ref{bp2-1-velo}. 
-\rev{
 There is also separation into fluctuating bands 
 which are oriented along the vorticity direction.
-The secondary velocity component $v_z$ also changes sign upon
+The secondary velocity component $v_z$ changes sign upon
 changing from positive to negative helicity and vice versa.
@@
 addressed in future work.
-}
 
@@
 
-\rev{
 At higher flow rates, $3.125\e{-4}\lesssim\gd\lesssim4.688\e{-4}$ 
@@
 well separated but do not have a clear orientation with respect to the flow.
-}
-
-\rev{
+
 Interestingly, the critical Ericksen number for this observed break-up is in 
 the range of $Er\simeq 2-4$. This is a regime where viscous forces have a 
-similar magnitude as elastic forces and where the nonlinear coupling between
+similar magnitude to elastic forces and where the nonlinear coupling between
 order and flow can cause small deviations to grow quickly over time. 
 This suggests that the BPI-3 regime constitutes an instance 
-of "rheochaos"~\cite{rheochaos,Cates:2002}.
-}
+of "rheochaos"~\cite{rheochaos,Cates:2002} as mentioned previously.
 
@@
 in ``nematic layers''. Colour coding indicates the 
-$x$-component of the director field (see colorbar on the right).
+$x$-component of the director field (see colour bar on the right).
 For clarity only selected sites are shown along the $y$-direction.
@@
 independent of their initial state.
-\rev{
 The director field of this configuration, also known as
@@
 It consists of a simple cholesteric helix with the helical axis oriented 
-along the gradient direction ($y$). While flowing the director field is stationary and 
+along the gradient direction ($y$). While the liquid crystal is flowing 
+the director field is stationary and 
 retains its relative orientation so that the depicted state is translational invariant 
@@
 compared to other states like the travelling helical wave ~\cite{Rey:1996a, Rey:1996b}.
-}
 
@@
 \caption{
-\rev{Director field of BPI-4 at high shear rate: The pictures show a frustrated 
+Director field of BPI-4 at high shear rate: The pictures show a frustrated 
 Grandjean configuration that formed layers along a diagonal direction. 
@@
 }
-}
 \label{bp1-4-frust}
@@
 
-\rev{The Grandjean configuration does not always emerge perfectly.
+The Grandjean configuration does not always emerge perfectly.
 Fig. \ref{bp1-4-frust} shows a time sequence of a frustrated state 
@@
 Because the diagonal state does not have the equilibrium layer spacing the conformation 
-buckles in order to minimise its free energy.
+buckles which reduces its free energy.
 Due to the periodic boundary conditions and their interlocking effect on the configuration 
@@
  a consequence of the shearing geometry and the relative orientation of the initial state
-}
 
@@
 
-\rev{
 We were able to characterise the rheology of cubic BPs, and identified 
 three different flow regimes for BPII and five different ones for BPI.
-This comprises steady states that BPI and BPII have in common like the 
+These include some steady states that BPI and BPII have in common like the 
 Grandjean texture and the flow-aligned nematic state at high shear rates.
-}  
 Below an Ericksen number of $Er\simeq3$, BPII exhibits weak shear-thinning and 
@@
 
-\rev{
 At larger Ericksen numbers, $4\lesssim Er\lesssim 10$, the flowing BPII network 
@@
 It may also transforms into the Grandjean texture, the steady state with the lowest
-shear stress and free energy in this range of shear rates.
-}
+shear stress in this range of shear rates.
 Still larger flow rates break this residual
@@
 
-\rev{
 At even larger Ericksen numbers $2\lesssim Er \lesssim 4$ the behaviour of BPI is again different.
@@
 irregular disturbances caused by small fluctuations.
-}
 It is tempting to
@@
 
-\begin{table*}[htpb]
-\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
-\hline
-& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
-\hline
-BPI & & & & $\times 10^5$\\
-\hline
-BPI-1 (AN) &0.24 &0.02 & 0.1388 &-17.7 &16.7 &3.7 &-3.6 &3.6 &3.5 &-3.9 &3.1 &2.8 \\
-BPI-1 (AN) &0.49 &0.04 & 0.1388 &-33.4 &32.0 &4.3 &-3.0 &3.1 &3.2 &-3.7 &2.5 &3.1 \\
-BPI-1 (AN) &0.98 &0.08 & 0.1388 &-64.4 &63.6 &4.7 &-3.6 &3.4 &4.8 &-5.3 &-4.2 &4.4 \\
-BPI-1 (AN) &1.95 &0.17 & 0.1388 &-126.2 &127.0 &6.8 &-3.8 &4.8 &6.7 &-5.2 &5.1 &6.2 \\
-\hline
-BPI-2 (PRC) &3.91 &0.33 & 0.1388 &-248.6 &248.3 &2.8 &-5.4 &5.6 &5.1 &-3.8 &4.6 &3.9 \\
-BPI-2 (PRC) &7.81 &0.67 & 0.1388 &-497.5 &496.4 &5.2 &-8.5 &8.0 &8.4 &-5.9 &7.8 &6.8 \\
-BPI-2 (PRC) &15.63 &1.33 & 0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-13.0} &\bf{15.6} &\bf{11.2} \\
-BPI-2 (PRC) &15.63 &1.33 & -0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-15.6} &\bf{13.0} &\bf{11.2} \\
-BPI-2 (PRC) &23.44 &2.00 & 0.1388 &-1492.2 &1485.5 &9.6 &-4.8 &3.7 &12.3 &-5.7 &4.6 &14.8 \\
-\hline
-BPI-3 (AN) &31.25 &2.67 & 0.1388 &-1990.6 &1983.2 &17.6 &-4.5 &6.3 &13.3 &-8.3 &10.0 &15.9 \\
-BPI-3 (AN) &46.87 &4.00 & 0.1388 &-2988.2 &2985.5 &42.4 &-10.5 &12.3 &28.4 &-15.5 &15.2 &26.8 \\
-\hline
-BPI-4 (GJ) &54.69 &4.67 & 0.1388 &-3475.1 &3473.4 &- &- &- &- &-8.9 &8.8 &- \\
-BPI-4 (FGJ) &62.50 &5.33 & 0.1388 &-4101.7 &4122.7 &- &- &- &- &-29.6 &20.0 &- \\
-BPI-4 (FGJ) &78.13 &6.01 & 0.1388 &-5140.0 &5092.3 &- &- &- &- &-35.8 &24.6 &- \\
-BPI-4 (GJ) &93.75 &8.00 & 0.1388 &-5955.4 &5946.1 &-  &- &- &- &-15.5 &15.1 &- \\
-BPI-4 (GJ) &125.0 &10.67 & 0.1388 &-7935.0 &7933.7 &- &- &- &- &-21.3 &20.1 &- \\
-BPI-4 (GJ) &187.5 &16.00 & 0.1388 &-11901.7 &11901.7 &- &- &- &- &-32.8 &29.1 &- \\
-\hline
-BPI-5 (FAN) & 250.0 &21.33 & 0.1388 &-15875.0 &15875.0 &- &- &- &- &-0.2 &-0.1 &- \\
-BPI-5 (FAN) & 375.0 &32.00 & 0.1388 &-23812.5 &23812.5 &- &- &- &- &-0.3 &0.1 &- \\
-\hline
-
-BPII  & & & & $\times 10^5$ \\
-\hline
-BPII-1 (PRC) &0.24 &0.02 & 0.1963 &-7.8 &7.7 &0.1 &-0.2 &0.2 &0.1 &-0.3 &0.3 &0.2 \\
-BPII-1 (PRC) &0.49 &0.04 & 0.1963 &-15.5 &15.4 &0.2 &-0.3 &0.4 &0.2 &-0.5 &0.4 &0.2 \\
-BPII-1 (PRC) &0.98 &0.08 & 0.1963 &-31.0 &30.8 &0.3 &-0.6 &0.6 &0.4 &-0.9 &0.8 &0.4 \\
-BPII-1 (PRC) &1.95 &0.17 & 0.1963 &-62.0 &61.6 &0.6 &-1.2 &1.2 &0.8 &-1.8 &1.6 &0.8 \\
-BPII-1 (PRC) &3.91 &0.33 & 0.1963 &-123.9 &123.3 &1.2 &-2.1 &2.1 &1.6 &-3.4 &3.0 &1.5 \\
-BPII-1 (PRC) &7.81 &0.67 & 0.1963 &-247.5 &246.8 &2.0 &-3.8 &3.8 &3.0 &-6.5 &5.5 &2.6 \\
-BPII-1 (PRC) &15.63 &1.33 & 0.1963 &-494.3 &493.7 &2.5 &-5.5 &-5.5 &4.0 &\bf{-10.9} &\bf{8.7} &\bf{3.9} \\
-BPII-1 (PRC) &15.63 &1.33 & -0.1963 &-494.3 &493.7 &2.5 &-5.5 &5.5 &4.0 &\bf{-8.7} &\bf{10.9} &\bf{3.9} \\
-BPII-1 (PRC) &31.25 &2.67 & 0.1963 &-987.5 &987.1 &3.5 &-6.9 &6.9 &5.1 &-17.7 &13.1 &5.0 \\
-BPII-1 (PRC) &39.06 &3.33 & 0.1963 &1233.9 &1233.6 &3.7 &-6.9 &6.9 &5.1 &-20.0 &14.5 &5.2 \\
-\hline
-BPII-2 (GJ) &46.87 &4.00 & 0.1963 &-1479.8 &1477.0 &- &- &- &- &-10.0 &9.6  &- \\
-BPII-2 (GJ) &54.69 &4.67 & 0.1963 &-1720.5 &1720.8 &- &- &- &- &-11.5 &11.2 &- \\
-BPII-2 (GJ) &62.5 &5.33 & 0.1963  &-1966.7 &1966.2 &- &- &- &- &-13.3 &12.7 &- \\
-BPII-2 (GJ) &78.13 &6.01 & 0.1963 &-2458.3 &2457.8 &- &- &- &- &-16.9 &15.7 &- \\
-BPII-2 (GJ) &93.75 &8.00 & 0.1963 &-2949.7 &2949.7 &- &- &- &- &-20.3 &18.4 &- \\
-BPII-2 (GJ) &125.0 &10.67 &0.1963 &-3932.8 &3932.8 &- &- &- &- &-26.6 &22.4 &- \\
-\hline
-BPII-3 (FAN) &187.5 &16.00 & 0.1963 &-5906.2  &5906.2 &- &-  &-  &-  &-0.6 &0.4 &- \\
-BPII-3 (FAN) &250.0 &21.33 & 0.1963 &-7874.9  &7874.9 &- &-  &-  &-  &-0.5 &0.3 &- \\
-\hline
-\end{tabular}
-\caption{
-\rev{Minima, maxima and standard deviation of time-averaged velocity components 
-for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
-shear flow. All velocity values are given in $10^{-5}$ lattice units. The transient dynamics 
-prior to regular oscillations or network breakup has not been included into the averages. 
-The regimes comprise the formation of an amorphous network (AN, BPI-1), 
-periodically recurring conformations with oscillatory stress response (PRC, BPI-2 and BPII-1), 
-amorphous networks at Ericksen numbers ${Er\simeq O(1)}$ (AN, BPI-3), 
-a Grandjean texture that may be frustrated (GJ \& FGJ, BPII-2 and BPI-4),
-and a flow-aligned nematic state (FAN, BPI-5 and BPII-3).
-(Last six columns refer to the secondary flow.)}
-}
-\label{tab1}
-\end{table*}
-\clearpage
 Table \ref{tab1} lists the parameters chosen for our runs (values of
@@
 
-\rev{
 In all our runs we used Lees-Edwards boundary conditions (LEBCs)~\cite{Wagner:2002}.
@@
 note that these boundary conditions impose macroscopic distortion
-of the network. They do not allow for a free-standing network as
-in flow-gradient direction there can be no slip in the network deformation 
-or strain at any point.
+of the network. They do not allow for a free-standing network 
+sustained stationary by permeation along the flow direction.
+Such a state is however forbidden for large system sizes since the 
+local permeation rate would have to increase indefinitely with 
+sample thickness.
 The boundary condition is equivalent to infinitely distant
@@
 boundary conditions at the walls.
-Effectively, this disallows permeation flows, which on the other
-hand require a non-zero velocity field and simultaneously
-much lower (ideally zero) perturbations of the blue phase network.
 In practice, permeation flows are 
@@
 along their axis~\cite{Marenduzzo:2006a, Marenduzzo:2006b}.
+However, as mentioned above, a triply periodic structure cannot
+be maintained stationary in a large system without infinite 
+permeation rate.
+
+\begin{table*}[htpb]
+\begin{tabular}{|c||c|| c || c || c |c |c||c| c| c||c| c| c|}
+\hline
+& $\dot{\gamma}$ & ${\it Er}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
+\hline
+BPI & & & & $\times 10^5$\\
+\hline
+BPI-1 (AN) &0.24 &0.02 & 0.1388 &-17.7 &16.7 &3.7 &-3.6 &3.6 &3.5 &-3.9 &3.1 &2.8 \\
+BPI-1 (AN) &0.49 &0.04 & 0.1388 &-33.4 &32.0 &4.3 &-3.0 &3.1 &3.2 &-3.7 &2.5 &3.1 \\
+BPI-1 (AN) &0.98 &0.08 & 0.1388 &-64.4 &63.6 &4.7 &-3.6 &3.4 &4.8 &-5.3 &-4.2 &4.4 \\
+BPI-1 (AN) &1.95 &0.17 & 0.1388 &-126.2 &127.0 &6.8 &-3.8 &4.8 &6.7 &-5.2 &5.1 &6.2 \\
+\hline
+BPI-2 (PRC) &3.91 &0.33 & 0.1388 &-248.6 &248.3 &2.8 &-5.4 &5.6 &5.1 &-3.8 &4.6 &3.9 \\
+BPI-2 (PRC) &7.81 &0.67 & 0.1388 &-497.5 &496.4 &5.2 &-8.5 &8.0 &8.4 &-5.9 &7.8 &6.8 \\
+BPI-2 (PRC) &15.63 &1.33 & 0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-13.0} &\bf{15.6} &\bf{11.2} \\
+BPI-2 (PRC) &15.63 &1.33 & -0.1388 &-995.0 &993.0 &8.7 &-13.1 &13.8 &13.2 &\bf{-15.6} &\bf{13.0} &\bf{11.2} \\
+BPI-2 (PRC) &23.44 &2.00 & 0.1388 &-1492.2 &1485.5 &9.6 &-4.8 &3.7 &12.3 &-5.7 &4.6 &14.8 \\
+\hline
+BPI-3 (AN) &31.25 &2.67 & 0.1388 &-1990.6 &1983.2 &17.6 &-4.5 &6.3 &13.3 &-8.3 &10.0 &15.9 \\
+BPI-3 (AN) &46.87 &4.00 & 0.1388 &-2988.2 &2985.5 &42.4 &-10.5 &12.3 &28.4 &-15.5 &15.2 &26.8 \\
+\hline
+BPI-4 (GJ) &54.69 &4.67 & 0.1388 &-3475.1 &3473.4 &- &- &- &- &-8.9 &8.8 &- \\
+BPI-4 (FGJ) &62.50 &5.33 & 0.1388 &-4101.7 &4122.7 &- &- &- &- &-29.6 &20.0 &- \\
+BPI-4 (FGJ) &78.13 &6.01 & 0.1388 &-5140.0 &5092.3 &- &- &- &- &-35.8 &24.6 &- \\
+BPI-4 (GJ) &93.75 &8.00 & 0.1388 &-5955.4 &5946.1 &-  &- &- &- &-15.5 &15.1 &- \\
+BPI-4 (GJ) &125.0 &10.67 & 0.1388 &-7935.0 &7933.7 &- &- &- &- &-21.3 &20.1 &- \\
+BPI-4 (GJ) &187.5 &16.00 & 0.1388 &-11901.7 &11901.7 &- &- &- &- &-32.8 &29.1 &- \\
+\hline
+BPI-5 (FN) & 250.0 &21.33 & 0.1388 &-15875.0 &15875.0 &- &- &- &- &-0.2 &-0.1 &- \\
+BPI-5 (FN) & 375.0 &32.00 & 0.1388 &-23812.5 &23812.5 &- &- &- &- &-0.3 &0.1 &- \\
+\hline
+
+BPII  & & & & $\times 10^5$ \\
+\hline
+BPII-1 (PRC) &0.24 &0.02 & 0.1963 &-7.8 &7.7 &0.1 &-0.2 &0.2 &0.1 &-0.3 &0.3 &0.2 \\
+BPII-1 (PRC) &0.49 &0.04 & 0.1963 &-15.5 &15.4 &0.2 &-0.3 &0.4 &0.2 &-0.5 &0.4 &0.2 \\
+BPII-1 (PRC) &0.98 &0.08 & 0.1963 &-31.0 &30.8 &0.3 &-0.6 &0.6 &0.4 &-0.9 &0.8 &0.4 \\
+BPII-1 (PRC) &1.95 &0.17 & 0.1963 &-62.0 &61.6 &0.6 &-1.2 &1.2 &0.8 &-1.8 &1.6 &0.8 \\
+BPII-1 (PRC) &3.91 &0.33 & 0.1963 &-123.9 &123.3 &1.2 &-2.1 &2.1 &1.6 &-3.4 &3.0 &1.5 \\
+BPII-1 (PRC) &7.81 &0.67 & 0.1963 &-247.5 &246.8 &2.0 &-3.8 &3.8 &3.0 &-6.5 &5.5 &2.6 \\
+BPII-1 (PRC) &15.63 &1.33 & 0.1963 &-494.3 &493.7 &2.5 &-5.5 &-5.5 &4.0 &\bf{-10.9} &\bf{8.7} &\bf{3.9} \\
+BPII-1 (PRC) &15.63 &1.33 & -0.1963 &-494.3 &493.7 &2.5 &-5.5 &5.5 &4.0 &\bf{-8.7} &\bf{10.9} &\bf{3.9} \\
+BPII-1 (PRC) &31.25 &2.67 & 0.1963 &-987.5 &987.1 &3.5 &-6.9 &6.9 &5.1 &-17.7 &13.1 &5.0 \\
+BPII-1 (PRC) &39.06 &3.33 & 0.1963 &1233.9 &1233.6 &3.7 &-6.9 &6.9 &5.1 &-20.0 &14.5 &5.2 \\
+\hline
+BPII-2 (GJ) &46.87 &4.00 & 0.1963 &-1479.8 &1477.0 &- &- &- &- &-10.0 &9.6  &- \\
+BPII-2 (GJ) &54.69 &4.67 & 0.1963 &-1720.5 &1720.8 &- &- &- &- &-11.5 &11.2 &- \\
+BPII-2 (GJ) &62.5 &5.33 & 0.1963  &-1966.7 &1966.2 &- &- &- &- &-13.3 &12.7 &- \\
+BPII-2 (GJ) &78.13 &6.01 & 0.1963 &-2458.3 &2457.8 &- &- &- &- &-16.9 &15.7 &- \\
+BPII-2 (GJ) &93.75 &8.00 & 0.1963 &-2949.7 &2949.7 &- &- &- &- &-20.3 &18.4 &- \\
+BPII-2 (GJ) &125.0 &10.67 &0.1963 &-3932.8 &3932.8 &- &- &- &- &-26.6 &22.4 &- \\
+\hline
+BPII-3 (FN) &187.5 &16.00 & 0.1963 &-5906.2  &5906.2 &- &-  &-  &-  &-0.6 &0.4 &- \\
+BPII-3 (FN) &250.0 &21.33 & 0.1963 &-7874.9  &7874.9 &- &-  &-  &-  &-0.5 &0.3 &- \\
+\hline
+\end{tabular}
+\caption{
+Minima, maxima and standard deviation of time-averaged velocity components 
+for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$) in steady 
+shear flow. All velocity values are given in $10^{-5}$ lattice units. 
+The last six columns refer to the secondary flow.
+The transient dynamics
+prior to regular oscillations or network break-up has not been included into the averages. 
+The regimes comprise the formation of an amorphous network (AN, BPI-1), 
+periodically recurring conformations with oscillatory stress response (PRC, BPI-2 and BPII-1), 
+amorphous networks at Ericksen numbers ${Er\simeq O(1)}$ (AN, BPI-3), 
+a Grandjean texture that may be frustrated (GJ \& FGJ, BPII-2 and BPI-4),
+and a flow-aligned nematic state (FN, BPI-5 and BPII-3).
+Bold figures indicate mean minima and maxima in $v_z$ and the standard deviation of the runs with 
+inverted helicity. 
 }
+\label{tab1}
+\end{table*}
 

### 2013-07-09 trunk/papers/bprheo/bprheo.tex
Ready to submit bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2048 40876441-3496-0410-a613-a2fca35b2ae2

@@
 cholesteric helix with axis in gradient direction, also referred to
-as Grandjean texture, as well as frustrated metastable Grandjean-like states that can transform into 
-a regular Grandjean texture upon increasing the flow rate (BPI-4, open triangles and solid circles);
+as Grandjean texture (BPI-4, open triangles), as well as frustrated metastable 
+Grandjean textures (BPI-4, solid circles) that can transform into 
+a regular Grandjean texture upon increasing the flow rate;
 and flow-aligned nematic state (BPI-5, solid triangles). 
@@
 that undergoes a buckling transition.
-Initially the layers happen to be along a diagonal direction. 
+Initially the layers happen to be along a diagonal direction rather than
+along the gradient direction. 
 Because the diagonal state does not have the equilibrium layer spacing the conformation 
@@
 strain rate is increased.
-Interestingly, the frustration can occur for any flow rate in regime BPI-4 as shown if Fig.~\ref{bp1-flowcurve}.
- a consequence of the shearing geometry and the relative orientation of the initial state
+This explains as well why the frustration is more frequent for low shear rates 
+in regime BPI-4, as shown in Fig.~\ref{bp1-flowcurve}.
 

### 2013-07-31 trunk/htdocs/multigrid.tex
Added notes on multigrid algorithm
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2095 40876441-3496-0410-a613-a2fca35b2ae2

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
%\documentclass[aps,prl,preprint,groupedaddress]{revtex4-1}
%\documentclass[aps,prl,preprint,superscriptaddress]{revtex4-1}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4-1}
\documentclass[11pt, oneside, a4paper]{article}

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-1}

\begin{document}

% Use the \preprint command to place your local institutional report
% number in the upper righthand corner of the title page in preprint mode.
% Multiple \preprint commands are allowed.
% Use the 'preprintnumbers' class option to override journal defaults
% to display numbers if necessary
%\preprint{}

%Title of paper
\title{A Parallel Geometric Multigrid Algorithm}

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current
% author. Explanatory text should go in the []'s, actual e-mail
% address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro foreach each type of information

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.
\author{Oliver Henrich$^1$, Kevin Stratford$^1$, S\'ebastien Loisel$^2$\\
$^1$ EPCC, School of Physics and Astronomy,\\
University of Edinburgh, Edinburgh EH9 3JZ, UK\\
$^2$ Department of Mathematics,\\ 
Herriot-Watt University, Edinburgh EH14 4AS \\
}
%\email[]{Your e-mail address}
%\homepage[]{Your web page}
%\thanks{}
%\altaffiliation{}
%\affiliation{}
\date{\today}
%Collaboration name if desired (requires use of superscriptaddress
%option in \documentclass). \noaffiliation is required (may also be
%used with the \author command).
%\collaboration can be followed by \email, \homepage, \thanks as well.
%\collaboration{}
%\noaffiliation

% insert suggested PACS numbers in braces on next line
%\pacs{}
% insert suggested keywords - APS authors don't need to do this
%\keywords{}
%\begin{abstract}
%Blablabla
%\end{abstract}

%\maketitle must follow title, authors, abstract, \pacs, and \keywords
\maketitle

% body of paper here - Use proper section commands
% References should be done using the \cite, \ref, and \label commands
\section{Definitions}

\begin{eqnarray*}
k&\dots&\mbox{grid level with $k=1$ being the coarsest grid}\\
\Omega_k &\dots& \mbox{regular grid with spacing $h(k)$}\\
\partial \Omega_k &\dots& \mbox{boundary of computational domain}\\
m&\dots& \mbox{iteration index}\\
u_k&\dots& \mbox{exact solution on $\Omega_k$}\\ 
v_k^m&\dots& \mbox{approximate solution on $\Omega_k$ after $m$ iterations}\\
e_k^m&\dots& \mbox{error on $\Omega_k$ after $m$ iterations}\\ 
R_h^H, R_k^{k-1}&\dots& \mbox{restriction operator: mapping functions on grid level $k$ with} \\
&& \mbox{spacing $h$ to functions on the next coarser level $k-1$ with spacing $H=2h$}\\ 
I_H^h, I_{k-1}^k&\dots& \mbox{interpolation operator: mapping functions on grid level $k-1$ with }\\
&&\mbox{spacing $H=2h$ to functions on the next finer level $k$ with spacing $h$}\\ 
\end{eqnarray*}

\section{Transfer Operators}

\subsection{Restriction}

We use half weighting (HW) as this restriction operator has reduced communication requirements. In stencil notation the effect of this operator is

\begin{eqnarray*}
f_H(x,y,z) &=& R_h^H f_h(x,y,z) = \frac{1}{12}[ f_h(x-h,y,z)+f_h(x+h,y,z)+f_h(x,y-h,z)\\
&& +f_h(x,y+h,z)+f_h(x,y,z-h)+f_h(x,y,z+h)+6\;f_h(x,y,z)].
\end{eqnarray*}

The operations are independent of each other and can be performed in parallel. The computation can also be overlapped with communication of the halo data.

\subsection{Interpolation}

We use trilinear interpolation. Note that the coordinate indices $x,y$ and $z$ refer to the indices on the fine grid $\Omega_h$.

\begin{equation*}
f_h(x,y,z) = I_H^h f_H(x,y,z)
\end{equation*}

\begin{equation*}
f_h(x,y,z) =\\
\begin{cases}
\;\;\;f_H(x,y,z)& , \mbox{ case } 1\\
\frac{1}{2} [f_H(x,y-h,z)+f_H(x,y+h,z)] &,  \mbox{ case } 2\\
\frac{1}{2} [f_H(x-h,y,z)+f_H(x+h,y,z)] &,  \mbox{ case } 3\\
\frac{1}{4} [f_H(x-h,y-h,z)+f_H(x+h,y-h,z) \\
\;\;\;+f_H(x-h,y+h,z)+f_H(x+h,y+h,z)] &,  \mbox{ case } 4\\
\frac{1}{8} [f_H(x-h,y-h,z-h)+f_H(x+h,y-h,z-h)\\
\;\;\;+f_H(x-h,y+h,z-h)+f_H(x+h,y+h,z-h)\\
\;\;\;+f_H(x-h,y-h,z+h)+f_H(x+h,y-h,z+h)\\
\;\;\;+f_H(x-h,y+h,z+h)+f_H(x+h,y+h,z+h) &, \mbox{ case } 5
\end{cases}
\vspace*{0.5cm}
\end{equation*}

We distinguish the following cases:

\begin{enumerate}
\item points with $x,y,z\in\Omega_H$ 
\item points with $x,z\in \Omega_H$, $y\notin \Omega_H$  
\item points with $y,z\in \Omega_H$, $x\notin \Omega_H$
\item points with $z\in \Omega_H$, $x,y\notin \Omega_H$ 
\item points with $x,y,z\notin \Omega_H$. 
\end{enumerate}

Also these operations are independent of each other and can be performed in parallel. However, it might not be possible to overlap computation and communication.

\section{Relaxation}

The equation we like to solve is the Poisson Equation $A u = f$ with $A=L=-\Delta$. 
We use the standard 7-point discretisation for the Laplacian:

\begin{eqnarray*}
A_h u_h(x,z,z) &=& \frac{1}{h^2} [6 u_h(x,y,z) - u_h(x+h,y,z) - u_h(x-h,y,z)-u_h(x, y+h,z) \\
 &-& u_h(x,y-h,z) - u_h(x,y,z+h) - u_h(x,y,z-h)] = f_h(x,y,z).
\end{eqnarray*}

We use a red-black Gauss-Seidel with overrelaxation parameter. An overrelaxtion parameter of $\omega=1.15$ was reported to lead to an optimal convergence rate. The iteration scheme for the approximation after $m+1$ iterations on grid $\Omega_h$ is

\begin{eqnarray*}
v^{m+1}_h(x,y,z) &=& \frac{1}{6} [h^2\;f_h(x,y,z)\\
&&+\;u_h^m(x+h,y,z)+u_h^m(x-h,y,z)\\
&&+\; u_h^m(x,y+h,z)+u_h^m(x,y-h,z)\\
&&+\; u_h^m(x,y,z+h)+u_h^m(x,y,z-h)]\\
u^{m+1}_h(x,y,z) &\leftarrow& u^m_h(x,y,z)+\omega[v^{m+1}_h(x,y,z)-u^m_h(x,y,z)]
\end{eqnarray*}

where the arrow indicates straight overwriting of the last approximation $u^m_h$ with the new one $u^{m+1}_h$ as it becomes available. All 'red' components on the lefthand side are independent of the 'black' components on the righthand side and can be updated in parallel.

In order overlap computation and communication we relax first the grid points at the boundaries $\partial\Omega_h$ followed by halo swaps via non-blocking MPI calls and compute then the inner grid points $\Omega_h\setminus\partial\Omega_h$.

\section{Coarse-grid communication and solution strategies}

Several coarse-grid communication strategies are suggested in the literature which are supposed to alleviate the communication overhead that emerges when the decreasing ratio of grid points and halo sites starts to have a deteriorating effect on the algorithm.

Probably the easiest way to deal with this issue is to truncate the multigrid hierarchy so that there are always plenty of grid points on each process. This can have a detrimental effect on the overall-convergence of the algorithm as the lowest grid level is probably still too fine to achieve rapid convergence.

Another strategy is to use a fast, parallel iterative solver on the coarstest grid level like conjugate-gradient solver for instance.

It is also possible to gather grid points that are initially allocated on different processes, solve only on a subset of processes with the others idling and scatter afterwards. Alternatively the data from all processes can be distributed over the entire topology by means of an MPI All-to-All call. The identical calculation can then be performed simulataneously on all processes, making further communication after each iteration step unnecessary. The grid can be also further coarsed down to the level where a direct solutin is possible.

\section{Multigrid Cycle}

In the following paragraph we describe a multigrid V-cycle in pseudocode for one iteration of the approximate solution $v_k^m\rightarrow v_k^{m+1}$. For convenience we drop the iteration index $m$.

The multigrid algorithm can be very efficently formulated in terms of a nested procedure. This is because solving $A_k u_k = f_k$ with arbitrary intial guess $v_k$ is equivalent to solving the defect equation $A_{k-1} e_{k-1} = r_{k-1}$ on the next coarser level with $e_{k-1}=0$ as initial guess. This allows mapping onto the original problem with $e_{k-1}$ and $r_{k-1}$ taking now the roles of a solution vector and a right-side vector, respectively. It is thus convenient to re-label the error and residual according to $e_{k-1}\rightarrow v_{k-1}$ and $r_{k-1}\rightarrow f_{k-1}$. This, however, does not affect the meaning of these quantities, in particular that from the first restriction on the solution vector is actually an error and the right-side vector is a residual.

\begin{algorithm}
\caption{Multigrid V-cycle for one iteration step $v_k^m\rightarrow v_k^{m+1}$}
\begin{algorithmic} 
\STATE{}
\STATE{1. Presmoothing:}
\FOR{$\nu_1$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k v_k = f_k$ on $\partial\Omega_k$ with initial guess $v_k$}
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k v_k = f_k$ on $\Omega_k\setminus \partial\Omega_k$  with initial guess $v_k$}
\ENDFOR
\STATE{}
\STATE{2. Compute residual:}
\STATE{$\triangleright$ Compute $r_k=f_k-A_k v_k \rightarrow f_k$}
\STATE{}
\STATE{3. Restrict residual:}
\STATE{$\triangleright$ Compute $f_{k-1}=R^{k-1}_k f_k$}
\STATE{}
\STATE{4. Coarse solve:}
\IF{$k-1>1$}
\FOR{$\nu_1$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_{k-1}v_{k-1}=f_{k-1}$ on $\partial\Omega_{k-1}$ with initial guess $v_{k-1}=0$ }
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_{k-1}v_{k-1}=f_{k-1}$ on $\Omega_{k-1}\setminus \partial\Omega_{k-1}$ with initial guess $v_{k-1}=0$ }
\ENDFOR
\STATE{$\triangleright$ Set $k-1 \rightarrow k$; goto step 2.}
\ENDIF
\STATE{}
\IF{$k-1=1$}
\STATE{$\triangleright$ Scatter all data via MPI All-to-All call}
\STATE{$\triangleright$ On each process solve $A_1 v_1=f_1$ on entire domain $\Omega_1$, either exactly or using a fast iterative solver with inital guess $v_1=0$} 
\ENDIF
\STATE{}
\STATE{5. Apply coarse grid correction to inital guess:}
\STATE{$\triangleright$ Compute $\hat{v}_k=v_k + I_{k-1}^k v_{k-1}$ on $\Omega_k$}
\STATE{}
\STATE{6. Postsmoothing:}
\FOR{$\nu_2$ steps}
\STATE{$\triangleright$ Request halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k \hat{v}_k= f_k$ on $\partial\Omega_k \rightarrow v_k$ new approximate solution}
\STATE{$\triangleright$ Send halo data via non-blocking MPI-call}
\STATE{$\triangleright$ Relax $A_k \hat{v}_k= f_k$ on $\Omega_k\setminus\partial\Omega_k \rightarrow v_k$ new approximate solution}
\ENDFOR
\STATE{}
\end{algorithmic}
\end{algorithm}

%\STATE <text>
%\IF{<condition>} \STATE{<text>} \ENDIF
%\FOR{<condition>} \STATE{<text>} \ENDFOR
%\FOR{<condition> \TO <condition> } \STATE{<text>} \ENDFOR
%\FORALL{<condition>} \STATE{<text>} \ENDFOR
%\WHILE{<condition>} \STATE{<text>} \ENDWHILE
%\REPEAT \STATE{<text>} \UNTIL{<condition>}
%\LOOP \STATE{<text>} \ENDLOOP
%\REQUIRE <text>
%\ENSURE <text>
%\RETURN <text>
%\PRINT <text>
%\COMMENT{<text>}
%\AND, \OR, \XOR, \NOT, \TO, \TRUE, \FALSE

\end{document}

### 2013-08-01 htdocs/multigrid.tex
mgrid comments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2096 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-07-31 trunk/htdocs/multigrid.tex
@@
 
+\section{Speculation}
+
+The following remarks are relevant for the branch of the code living
+in the \texttt{testcharge} directory, where we have SOR (both uniform
+permitivity and heterogenous permitivity) and should shortly have a
+Fourier solver (uniform permeativity) available.
+
+\subsection{Coordinate system}
+
+The existing code has a single coordinaite system which corresponds
+to the fine grid $\Omega_h$ in the multipgrid picture. Details of the
+coordinate system (global and local extent; MPI communicators; 3d-to-1d
+indexing functions, etc) are encapsulated in \texttt{coords.c}.
+\begin{enumerate}
+\item
+As a first step towards a multigrid implementation, it would be
+desirable to ``objectify'' formally a \texttt{coods\_t} data type.
+This would be explicitly referenced in all existing fine grid
+contexts (and is desirable on other grounds as well).
+\item
+For multigrid, one could then instantiate one \texttt{coords\_t}
+object for each level of the grid $\Omega_k$ to allow corresponding
+grid operations at a given level.
+\item
+A second option would be to retain a single \texttt{coords\_t} object,
+but add functionality to recognise a ggrid level $k$ and perform the
+necessary arithmetic/logic assuning uniform refinement. (This is
+probably a less elegant and flexible solution, if more compact.)
+\end{enumerate}
+
+\subsection{Grid data structure(s)}
+
+Each level of the grid $\Omega_k$ requires coordinate information,
+but also local storage for three-dimensional field quantities, eg.,
+$f_k(x,y,z)$. Local field quantities are expected to require a halo
+region of extent \texttt{nhalo} cf. existing field quantities on the
+fine grid. This allows interpolation and restriction operators to be
+evalautated locally.
+
+So, a grid-level structure might contain
+\begin{verbatim}
+  struct grid_s {
+    int k;               /* the grid level */
+    coords_t * coords;   /* coordinate information this level */
+    field_t * f;         /* rhs */
+    ...
+  }
+\end{verbatim}
+A multigrid structure would then contain an array of grids for levels
+$k = 1, \ldots, k_{max}$, information replicated on all MPI tasks:
+\begin{verbatim}
+  struct mgrid_s {
+    int kmax;         /* the number of levels */
+    grid_t ** level;  /* array of grid objects */
+    ...
+  }
+\end{verbatim}
+
+\subsection{The limit of local cosrsening}
+
+For a given uniform domain decomposition of the fine grid, local
+restriction or coarsening can be carried out to the point where
+there is a grid of extent one in the local domain. This limits
+the extent of local coarsening. At this point, further coarsening
+might be based on coarsening the processor grid in the same spirit
+as the computational grid.
+
+\subsection{Issues}
+
+\begin{enumerate}
+\item
+Can we use e.g., PetSc to test some options/performance before starting to
+code?
+\item
+Do we want to limit ourselves to regular system sizes (i.e., 2$^n$)
+and processor decompositions?
+\item
+Can the approach be extended to a threaded model (e.g., GPU)?
+\end{enumerate}
+
 \end{document}

### 2013-08-22 branches/fft-testcharge/docs/README.tex
added Reaadme files
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2140 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[12pt,a4paper]{article}

\usepackage{verbatim}

\begin{document}

%AC%\pagestyle{myheadings}
%AC%\markright{D.~S.~Henty}

%\title{A Latex thesis example}
%\author{D.~S.~Henty}
%\date{\today}

\setcounter{secnumdepth}{3}

\pagenumbering{roman}

%\begin{center}

\title{Ludwig electrokinetics module - FFT solver with P3DFFT:\\ Instructions}
\author{Ruairi Short}
\date{\today}

\maketitle

\tableofcontents
\clearpage

%\end{center}

\section{Installing libraries}
Two libraries need to be installed in order to use the FFT solver in Ludwig.
One of ESSL or FFTW must be installed and P3DFFT built on the appropriate one of these.

\subsection{Install FFTW (optional)}
If you choose to use FFTW and it is not already installed on your system the following is how to install it.
  \begin{itemize}
  \item extract the fftw-3.3.3.tar file of the FFTw library to a directory of your choice. This location will be needed to install P3DFFT.
  \item move to the new directory and run the following commands:
    \begin{verbatim}
      ./configure
      make
      make install
    \end{verbatim}
  \end{itemize}

For an overview on possible arguments to \texttt{configure} see the FFTW documentation: \textit{http://www.fftw.org/fftw3\_doc/}

\subsection{Install P3DFFT}
  This step has been tested on HECToR and BlueGene/Q. Specific instructions are given for these particular machines. For other machines the steps should be similar but it is not guaranteed to be simple.

\subsubsection{HECToR}
  \begin{itemize}
    \item extract the p3dfft.2.5.1.tar file of the library to a folder of your choice.
    \item go to the P3DFFT directory and run the following commands:
      \begin{verbatim}
        module switch PrgEnv-cray PrgEnv-pgi
        $ ./configure --prefix=$INSTALL_DESTINATION
          --enable-pgi --enable-stride1 --enable-fftw 
          --with-fftw=$FFTW_HOME FC=ftn CC=cc
         FCFLAGS=-O3 CFLAGS=-O3
        $ make
        $ make install
      \end{verbatim}
    \item The \texttt{module switch} command ensures the PGI compiler will be used.
    \item \texttt{INSTALL\_DESTINATION} is the location P3DFFT will be installed to. The default is \texttt{/usr/local}
    \item \texttt{FFTW\_HOME} is the location of the FFTW library. On HECToR the FFTW library is already installed and located at \texttt{/opt/fftw/3.3.0.1/interlagos}
    \item copy \texttt{config.h} from the P3DFFT directory to \texttt{INSTALL\_DESTINATION/include}.
  \end{itemize}

To see further flags that can be passed to \texttt{configure} see the P3DFFT users guide and wiki: \textit{http://code.google.com/p/p3dfft/wiki/Install}.

\subsubsection{BlueGene/Q}
  \begin{itemize}
    \item extract the p3dfft.2.5.1.tar file of the library to a folder of your choice.
    \item extract p3dfftbgqedit.tar to the same directory. This will overwrite some of the files in this directory in order to successfully install P3DFFT.
    \item go to the P3DFFT directory and run the following commands:
      \begin{verbatim}
        $ ./configure --prefix=$INSTALL_DESTINATION
          --enable-ibm --enable-stride1 --enable-essl 
          FC=mpixlf90_r CC=mpixlc_r
          FCFLAGS=-O3 CFLAGS=-O3
        $ make
        $ make install
      \end{verbatim}
    \item \texttt{INSTALL\_DESTINATION} is the location P3DFFT will be installed to. The default is \texttt{/usr/local}
    \item The ESSL library is located at \texttt{/opt/ibmmath/lib64/libesslbg.a}. This will be needed later.
    \item copy \texttt{config.h} from the P3DFFT directory to \texttt{INSTALL\_DESTINATION/include}.
  \end{itemize}

To see further flags that can be passed to \texttt{configure} see the P3DFFT users guide and wiki: \textit{http://code.google.com/p/p3dfft/wiki/Install}.

\section{Compiling Ludwig}
The steps in order to succesfully compile Ludwig and the unit tests are now outlined.

\subsection{Edit Makefile}
  The Makefiles in the \texttt{src/} and \texttt{tests/} directories contain the compiler flags necessary to compile Ludwig on HECToR and BlueGene/Q. The appropriate section should be uncommented.\\
  The user must also ensure that the link to the P3DFFT library is correct, along with the links to the FFTW or ESSL libraries.

\subsection{Compile Ludwig}
  Ludwig is compiled in the \texttt{src/} directory using the command:
  \begin{verbatim}
    $ make mpi
  \end{verbatim}
  This wil create an executable \texttt{Ludwig.exe} in this directory.

\subsection{Compile tests}
  To compile the unit tests, in the \texttt{src/} directory run the command
  \begin{verbatim}
    $ make libmpi
  \end{verbatim}
  Then in the \texttt{tests/} directory run
  \begin{verbatim}
    $ make tests_mpi
  \end{verbatim}
  If desired, and the option is available, the tests can be run on the front end by using
  \begin{verbatim}
    $ make do_tests_mpi
  \end{verbatim}
  in the \texttt{tests/} directory. This may take a long time to complete.

\section{Execute program}
  Ludwig can now be executed.
  Ludwig uses an input file to decide what parameters to use upon execution.
  By default it reads input from the file \texttt{input}, though the user can provide a different one at runtime.
  As an example, on HECToR the command
  \begin{verbatim}
    $ aprun -n 8 ./Ludwig.exe input
  \end{verbatim}
  will run Ludwig on 8 cores using the input file \texttt{input}.
  A file, \texttt{input.ref}, has been included as an example input file.
  Also included are the files used to run the Gouy Chapman and liquid junction potential problems, \texttt{GCinput} and \texttt{LJinput} respectively.\\
  Finally sample submission scripts are included for both HECToR and BlueGene/Q, \texttt{hectorludwig.pbs} and \texttt{bgqludwig.pbs} respectively.

\end{document}

### 2013-08-30 papers/2012-sandwich/support.tex
Revised support - Davide, Aug 30
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2142 40876441-3496-0410-a613-a2fca35b2ae2

--- 2012-08-29 trunk/papers/2012-sandwich/support.tex
@@
 
-{\bf Supporting Online Material: Sandwich Title}
+{\bf Supporting Information for: Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
 
-Here, the details of the simulation method, protocol, and parameters are
-set out. The matching of simulation units to physical units is also
-discussed.
+Here, we give more details of the simulation method, protocol, and parameters used.
+We also provide additional figures complementing the main text, and discuss the matching of simulation units to physical units.
 
@@
 The remaining terms of the free energy density in Eq.~\ref{free} describe
-distortions of
-the order parameter field. It is conventional \cite{blue1,deGennes} to
+distortions of the order parameter field. In theoretical work and when
+describing a generic rather than a specific system, it is 
+conventional~\cite{blue1,deGennes} to
 assume that splay, bend and twist deformations of the director are equally
@@
 fluid LC ordering.
+
+The effect of an electric field can be modelled by including the following
+term in the free energy density,
+\begin{equation}
+-(\epsilon_a/12\pi)E_{\alpha}Q_{\alpha\beta}E_{\beta}
+\end{equation} 
+where $E_{\beta}$ are the components of the electric field, and 
+$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
+crystal. The strength of the electric field is quantified via
+a further dimensionless number, 
+\begin{equation}
+{\cal E}^2=(27 \varepsilon_a E_\alpha E_\alpha)/(32 \pi A_0 \gamma).
+\end{equation}
 
@@
 \end{eqnarray}
-CORRECT TRACE?!
-Here, $\xi$ is a material-dependent `tumbling parameter' that controls the
+%CORRECT TRACE?!
+Here, $\xi$ is a material-dependent that controls the
 relative importance of rotational and elongational flow for molecular
-alignment. In all cases, the value used here is $\xi = 0.7$, which is within
+alignment, and determines in practice how the orientational order
+responds to a local shear flow. 
+In all cases, the value used here is $\xi = 0.7$, which is within
 the `flow aligning' regime, where molecules align at a fixed angle
@@
 \end{equation}
-CHECK which corresponds to SOMETHING [REFERENCE].
+which corresponds to the bulk order in a defect-free nematically ordered 
+sample~\cite{denniston} (this is also very close to the value of the order 
+parameter in a cholesteric or in a blue phase away from disclinations).
 
 In all simulations the colloidal (hard sphere) radius is $a = 2.3$. To
-prevent colloids overlapping at the hard sphere radius, an addition
+prevent colloids overlapping at the hard sphere radius, an additional
 short range soft potential is included. This takes the form of
@@
 unit cells with fully periodic boundary conditions.
-Simulations at
-different solid volume fractions (1\% and 5\%), and different surface
-anchoring strengths ($Wa/K = 0.23$ and $Wa/k = 2.3$) representing
+Simulations at different solid volume fractions (1\% and 5\%), and different 
+surface anchoring strengths ($Wa/K = 0.23$ and $Wa/k = 2.3$) representing
 ``weak'' and ``strong'' anchoring are run for 500,000 simulation
@@
 Table~\ref{table:params}) remain appropriate for blue phase I.
-Colloids are added
-as before and the simulations run for 500,000 or 700,000 simulation
-steps.
+Colloids are added as before and the simulations run for 500,000 or 
+700,000 simulation steps.
 
@@
 
-ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
+To get from simulation to physical units, we need to calibrate scales of length,
+energy and time. For similar mappings, see e.g.~\cite{denniston2}.
 
-ISSUE: no redshift?
+The length scale can be set by mapping our half pitch to realistic blue phase
+unit cell size, which are in the 100-500 nm range \cite{blue1}. 
+For bulk simulations (Fig. 1 and Fig. 3 in the main text), a suitable choice
+is that that one simulation unit (lattice site) corresponds to 10 nm. Therefore
+the colloidal size in Figs. 1 and 3 corresponds to $\sim$ 50 nm.
 
-ISSUE: do we want some free energy numbers?
+To get an energy scale, we choose $A_0 \simeq 10^6$ Pa, which is 
+reasonable from Ref.~\cite{blue1} (this is also the choice in
+Ref.~\cite{oliver2}). This leads to a simulation unit of
+free energy (or stress) equal to $10^{8}$Pa in Figs. 1 and 3.
+From the energy and length scales one can map elastic constants to
+physical units: for instance the simulation of Fig. 1 corresponds to
+a liquid crystal with splay, bend and twist (Frank) elastic constants
+equal to 35 pN (or equivalently $K=$70 pN).
+\begin{equation}
+\frac{a}{b^2} = \frac{27}{2 A_0\gamma} \sim 2-5 \times 10^{-6}
+{\rm J}^{-1}\, {\rm m}^{3}.
+\end{equation}
 
-PENDING
+For the timescale calibration we use the formula (see e.g.~\cite{denniston1})
+\begin{equation}
+\gamma_1=\frac{2q^2}{\Gamma}\;.
+\end{equation}
+ which relates the `rotational viscosity' $\gamma_1$, to the ordering strength $q$ and the order parameter mobility $\Gamma$.
+In our simulations, we chose $\Gamma = 0.5$, while the value of $\gamma$ used in Fig. 1 leads to $q=1/2$.   
+For real liquid crystalline materials, $\gamma_1$ lies usually in range $10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness $\gamma_1 = 1$ Pa s (equivalently 10 poise).
+Given the previous mapping for free energy (or stress) units, this leads to a simulation unit of time equal to $10^{-8}$ s.
+The total simulated time in Figs.
+1-3 is therefore in the ms range.
+
+To calibrate the electric field strength 
+we equate the value of the dimensionless parameter ${\cal E}$ in our 
+simulations with that of a hypothetical experiment.
+Given the previous value for $A_0$, and a dielectric anisotropy of 10
+(or equivalently $\epsilon_a=10\epsilon_0$, with $\epsilon_0$ the
+dielectric permittivity of vacuum), we then obtain that an electric field
+of 5 V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.
+
+Given these mappings, all quantities can be easily transformed from
+simulation to physical units and vice versa. For example Fig. 1 and 2
+use an isotropic viscosity $\eta \simeq$ 0.01 and 0.1 in simulation units,
+which translate into $0.01-0.1$ Pa s respectively. This is sensible (if 
+slightly low) for a molecular nematogen in the isotropic phase. 
+The effective viscosity in ordered phases is higher, but this is ensured
+in our model by the coupling to the order parameter.
+We note that the density is set to 1 in our simulations: this
+corresponds to a fluid density which is about a thousand times larger
+than in experiments. This however makes no difference in practice as
+the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
+velocity, in our simulations around $10^{-5}$ in simulation units) 
+remains small enough~\cite{codef}.
+
+To summarize the above, our simulations represent BP-forming materials, with the interpretation of the LB units for length, time, and energy density are close to 10 nm, 10 ns, and 100 MPa respectively.
+
+%ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
+
+%ISSUE: no redshift?
+
+%ISSUE: do we want some free energy numbers?
 
@@
 \end{center}
-\caption{As for main manuscript Fig.~1 (initialised to equilibrium blue
-phase I), but for the entire simulation
-system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
-weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
+\caption{As for main manuscript Fig.~1A-E (initialised to equilibrium blue
+phase I), but for the entire simulation system of 128$^3$ lattice sites. (a) 
+solid volume fraction 1\% and weak anchoring; 
+(b) 1\% solid volume fraction and strong anchoring;
 (c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
@@
 \includegraphics[scale=0.35]{s2.jpg}
-\caption{As for main manuscript Fig.~2 (initialised via a ``quench''
+\caption{As for main manuscript Fig.~1E-H (initialised via a ``quench''
 to generate a disordered network), but for the entire simulation
@@
 volume fraction and strong anchoring.}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{s5.jpg}
+\end{center}
+\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
+(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
+(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
 \end{figure}
@@
 
+\bibitem{denniston}
+C. Denniston, E. Orlandini, J. M. Yeomans, 
+\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
+
+\bibitem{denniston2}
+C. Denniston, D. Marenduzzo, E. Orlandini, J. M.  Yeomans, 
+\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
+
 \bibitem{fournier2005}
@@
 \textit{Europhys. Lett.} \textbf{72}, 403 (2005).
+
+\bibitem{oliver2}
+O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
+\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
 

### 2013-08-30 trunk/papers/2012-sandwich/support.tex
added pair correlation functions
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1770 40876441-3496-0410-a613-a2fca35b2ae2
Revised support, added Fig. S1 - Davide, Aug 30
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2143 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
+To get from simulation to physical units, we need to calibrate scales of length,
+energy and time. For similar mappings, see e.g.~\cite{denniston2}.
 
-ISSUE: no redshift?
+The length scale can be set by mapping our half pitch to realistic blue phase
+unit cell size, which are in the 100-500 nm range \cite{blue1}. 
+For bulk simulations (Fig. 1 and Fig. 3 in the main text), a suitable choice
+is that that one simulation unit (lattice site) corresponds to 10 nm. Therefore
+the colloidal size in Figs. 1 and 3 corresponds to $\sim$ 50 nm.
 
-ISSUE: do we want some free energy numbers?
+To get an energy scale, we choose $A_0 \simeq 10^6$ Pa, which is 
+reasonable from Ref.~\cite{blue1} (this is also the choice in
+Ref.~\cite{oliver2}). This leads to a simulation unit of
+free energy (or stress) equal to $10^{8}$Pa in Figs. 1 and 3.
+From the energy and length scales one can map elastic constants to
+physical units: for instance the simulation of Fig. 1 corresponds to
+a liquid crystal with splay, bend and twist (Frank) elastic constants
+equal to 35 pN (or equivalently $K=$70 pN).
+\begin{equation}
+\frac{a}{b^2} = \frac{27}{2 A_0\gamma} \sim 2-5 \times 10^{-6}
+{\rm J}^{-1}\, {\rm m}^{3}.
+\end{equation}
 
-PENDING
+For the timescale calibration we use the formula (see e.g.~\cite{denniston})
+\begin{equation}
+\gamma_1=\frac{2q^2}{\Gamma}\;.
+\end{equation}
+ which relates the `rotational viscosity' $\gamma_1$, to the ordering strength $q$ and the order parameter mobility $\Gamma$.
+In our simulations, we chose $\Gamma = 0.5$, while the value of $\gamma$ used in Fig. 1 leads to $q=1/2$.   
+For real liquid crystalline materials, $\gamma_1$ lies usually in range $10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness $\gamma_1 = 1$ Pa s (equivalently 10 poise).
+Given the previous mapping for free energy (or stress) units, this leads to a simulation unit of time equal to $10^{-8}$ s.
+The total simulated time in Figs.
+1-3 is therefore in the ms range.
+
+To calibrate the electric field strength 
+we equate the value of the dimensionless parameter ${\cal E}$ in our 
+simulations with that of a hypothetical experiment.
+Given the previous value for $A_0$, and a dielectric anisotropy of 10
+(or equivalently $\epsilon_a=10\epsilon_0$, with $\epsilon_0$ the
+dielectric permittivity of vacuum), we then obtain that an electric field
+of 5 V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.
+
+Given these mappings, all quantities can be easily transformed from
+simulation to physical units and vice versa. For example Fig. 1 and 2
+use an isotropic viscosity $\eta \simeq$ 0.01 and 0.1 in simulation units,
+which translate into $0.01-0.1$ Pa s respectively. This is sensible (if 
+slightly low) for a molecular nematogen in the isotropic phase. 
+The effective viscosity in ordered phases is higher, but this is ensured
+in our model by the coupling to the order parameter.
+We note that the density is set to 1 in our simulations: this
+corresponds to a fluid density which is about a thousand times larger
+than in experiments. This however makes no difference in practice as
+the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
+velocity, in our simulations around $10^{-5}$ in simulation units) 
+remains small enough~\cite{codef}.
+
+To summarize the above, our simulations represent BP-forming materials, with the interpretation of the LB units for length, time, and energy density are close to 10 nm, 10 ns, and 100 MPa respectively.
+
+%ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
+
+%ISSUE: no redshift?
+
+%ISSUE: do we want some free energy numbers?
 
@@
 \begin{center}
+\includegraphics[scale=0.8]{SupplFig1.png}
+\end{center}
+\caption{(A) Schematic representation of the director field
+in a cholesteric (left) and within a double twist cylinder (right).
+(B) Snapshot of the disclination network in an equilibrated 
+blue phase I structure. (C) Same as (B) for an amorphous blue phase III
+network.}
+\end{figure}
+
+\begin{figure}
+\begin{center}
 \includegraphics[scale=0.35]{s1.jpg}
 \end{center}
-\caption{As for main manuscript Fig.~1 (initialised to equilibrium blue
-phase I), but for the entire simulation
-system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
-weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
+\caption{As for main manuscript Fig.~1A-E (initialised to equilibrium blue
+phase I), but for the entire simulation system of 128$^3$ lattice sites. (a) 
+solid volume fraction 1\% and weak anchoring; 
+(b) 1\% solid volume fraction and strong anchoring;
 (c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
@@
 \end{center}
-\caption{As for main manuscript Fig.~3 (confined geometry with normal
+\caption{As for main manuscript Fig.~2A-D (confined geometry with normal
 anchoring walls at both top and bottom), but for the entire simulation
@@
 \end{center}
-\caption{As for main manuscript Fig.~4 (confined geometry with planar
+\caption{As for main manuscript Fig.~2E-H (confined geometry with planar
 anchoring walls at both top and bottom), but for the entire simulation
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=0.95\columnwidth]{s5.jpg}
+\end{center}
+\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
+(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
+(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
+\end{figure}
+
 \newpage
 
-{\bf Supporting References}
+{\bf Captions for Supplementary Movies} \\
+
+%{\bf Supporting References}
 
@@
 \textit{Europhys. Lett.} \textbf{72}, 403 (2005).
+
+\bibitem{oliver2}
+O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
+\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
+
+\bibitem{codef} M. E. Cates,  et al., 
+\textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).
 

### 2013-10-02 papers/2012-sandwich/CholColloids.tex
current version uploaded by Davide (Oct 2)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2152 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \title{Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
-\author{Ourselves \\ {\small Here and there}}
+\author{K. Stratford$^{*}$, O. Henrich$^{*}$, J.~S. Lintuvuori$^{*}$,\\
+M. E. Cates, D. Marenduzzo \\ 
+{\small SUPA, School of Physics and Astronomy, Edinburgh University,} \\
+\small{Edinburgh, EJ9 3JZ, UK}\\
+\small{$^{*}$ equally contributed to this work}}
 \date{}
@@
 Here we show by computer simulation that colloid/cholesteric mixtures can give rise to a much wider range of soft-solid composites than previously anticipated.
-These include regular crystals, glasses, cellular solids, isolated clusters, twisted rings and undulating colloidal ropes.
+These include regular crystals, glasses, percolating gels, isolated clusters, twisted rings and undulating colloidal ropes.
 The final structure can be tuned through the particle concentration, and by varying the surface interactions of the cholesteric host with both the colloids and any confining walls.
 Remarkably, we find that many of these new materials are metastable: two or more structures can arise under identical thermodynamic conditions.
-The observed structure depends on not only on the formulation protocol, but also on the field history [CONFIRM!].
+The observed structure depends not only on the formulation protocol, but also on the field history.
 This new class of soft materials should thus be relevant to the design of switchable, multistable devices for optical technologies including, for example, e-paper~\cite{epaper}.}
@@
 they disrupt the orientational order in the fluid and create defects (disclination lines) in the nematic close to their surfaces.
-To minimize their free energy cost, it is generally advantageous to for defects to be shared:
+To minimize their free energy cost, it is generally advantageous for defects to be shared:
 therefore the colloidal particles have a generic tendency to aggregate. Such aggregates may further 
@@
 they offer exciting prospects for applications as biosensors~\cite{abbott}, or
-as new-generation display devices~\cite{colloiddevice}.
+as new-generation display devices as they can exploit the bistability
+inherent in cholesteric devices~\cite{colloiddevice,choldevice}.
 
@@
 BPs arise because the director field can twist around more than one direction at once (see
-the cartoon in Fig.~1). Such double twist 
+the cartoon in Fig. S1). Such double twist 
 regions (generally cylinders: DTCs) are energetically favoured at high enough molecular chirality, but
@@
 %
-BPs are stabilised by
-the tendency of chiral molecules to twist, but destabilised by the
-cost of forming topological defects (which locally diminish the molecular alignment).
+BPs are stabilised by the tendency of chiral molecules to twist, but destabilised by the cost of forming topological defects (which locally diminish the molecular alignment).
 In conventional cholesteric materials~\cite{mermin} this balance is only achievable in a narrow temperature range of a few degrees, around the onset of liquid crystalline ordering.
-However, recent advances in formulation have widened this stability range enormously (up to 250K - CONFIRM) paving the way to
+However, recent advances in formulation have widened this stability range enormously (up to 250K) paving the way to
 applications of BPs in display devices~\cite{kikuchi,coleswidetrange,bpdevice}.
@@
 These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{domaingrowth,bp3,henrichfield}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}.
-Our methodology and simulation parameters are summarized in Supporting Information.
+Our methodology and simulation parameters are summarized in the Supporting Information.
 
-We show in Fig.~2 the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and and then dispersed colloidal particles within it.
-Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r=XXX$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface).
-For the latter we choose $w = 0.23, 2.3$ which are at the lower and upper end of the typical experimental range [CONFIRM, GIVE REF.].
+We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles within it.
+Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.10$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface).
+For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
 This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.
 
-For a single colloidal particle
-in a nematic fluid, $w$ controls the formation of a local topological defect at the particle surface~\cite{stark}, with weak distortion of the nearby fluid at small $w$.
+For a single colloidal particle in a nematic fluid, $w$ controls the formation of a local topological defect at the particle surface~\cite{stark}, with weak distortion of the nearby fluid at small $w$.
 In the colloid-BP system, $w$ also plays a determining role in the final composite structure.
-When $w$ is small (Fig.~ 2A,C), the attraction of particles to the disclination lattice causes this to become covered by particles, without disruption of its long-range order.
+When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes this to become covered by particles, with little disruption of its long-range order.
 Thus a templated colloidal crystal is created, as first reported in~\cite{miha}.
-Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which sets a typical spacing between them along the disclination lines.
-It seems likely that polymer-stabilised  blue phases~\cite{kikuchi}
-work by a similar principle to this small $w$ case, with a weak segregation of monomers towards the defect regions [IS THIS WHAT YOU MEANT?].
+Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
+1A).
+It seems likely that polymer-stabilised  blue phases~\cite{kikuchi} work by a similar principle to this small $w$ case, with a weak segregation of monomers towards the defect regions.
 
-For large $w$ (Fig.~2B,D), the picture changes dramatically.
-Each colloidal particle now disrupts significantly the order of the nearby fluid, with defect formation close to its surface~\cite{stark}. The resulting strong disruption of
-local liquid crystalline order around each particle leads now to a 
-strong interparticle attraction. This restructures the disclination
+For large $w$ (Fig.~1C,D), the picture changes dramatically.
+Each colloidal particle now disrupts significantly the order of the nearby fluid, with defect formation close to its surface~\cite{stark}.
+The resulting strong disruption of local liquid crystalline order around each particle leads now to
+a strong interparticle attraction. This restructures the disclination
 network completely, destroying the long range order of
 the original BPI topology. Most notably, particle aggregation favours
-the formation of defect junctions, where four disclinations meet. This structural motif is present in the unit cell of BPII rather than BPI (and also seen in the amorphous BPIII~\cite{bp3}). The 
-particle clusters are disjoint at low volume fractions ($\phi\simeq 1\%$) but 
-they interact elastically via the connecting disclinations.
+the formation of defect junctions, where four disclinations meet.
+This structural motif is present in the unit cell of BPII rather than BPI (and also seen in the amorphous BPIII~\cite{bp3}).
+The particle clusters are disjoint at low volume fractions ($\phi\simeq 1\%$) but they interact elastically via the connecting disclinations.
 At larger $\phi$, the aggregates join up to form a percolating colloidal cluster which essentially templates the disclinations rather than vice versa.
 
-All the structures in Fig.~2 should be soft solids, of nonzero elastic modulus $G$ at low frequency. This contrasts with pure BPs, where
+All the structures in Fig.~1A-D should be soft solids, of nonzero elastic 
+modulus $G$ at low frequency. This contrasts with pure BPs, where
 $G=0$ as the fluid can flow with a finite
 viscosity via permeation of molecules through a fixed defect lattice~\cite{permeation1,permeation2}. We expect the
-structures in Fig.~2A-C to be only weak gels, as here the 
+structures in Fig.~1A-C to be only weak gels, as here the 
 defect network percolates but particle contacts do not. This is broadly similar to the cholesteric-nanoparticle gel reported
-in~\cite{lubensky} (with $G<1$ Pa). In contrast, the ``cellular solid'' 
-structure in Fig.~2D should be much more resistant mechanically, 
+in~\cite{lubensky} (with $G<1$ Pa). In contrast, the colloidal gel
+structure in Fig.~1D should be much more resistant mechanically, 
 due to the formation of a thick percolating particle network. The 
@@
 colloid-nematic composites~\cite{tiffany} at much higher volume 
-fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa. [WHAT IS CELLULAR ABOUT THIS STRUCTURE?]
+fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa.
 
@@
 Here we know that, without particles, the kinetics can favour disordered (BPIII-like) intermediates which may be long lived or even metastable~\cite{domaingrowth}.
-Fig.~3 shows outcomes of such a quench protocol with the same parameter sets as in Fig.~2.
-For small $w$, the particles again decorate the disclination lines, but are templated onto an amorphous network rather than an ordered one.
-Increasing the volume fraction (Fig.~3B) leads to a more even coverage with
+Fig.~1E-H shows outcomes of such a quench protocol with the same parameter sets as in Fig.~1A-D.
+For small $w$, the particles again decorate the disclination lines, but are this time templated onto an amorphous network rather than an ordered one.
+Increasing the volume fraction (Fig.~1F) leads to a more even coverage with
 a better defined particle spacing. 
@@
 the self-assembly.
-At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~3C) whereas at higher $\phi$ a percolating network of this strands (the ``cellular solid, Fig.~3D) is again seen.
-Perhaps unsurprisingly, the free energy of these amorphous colloid-BP composites is considerably higher than for the decorated BPI structure (see Supporting Information, Fig.~S1).
-Nevertheless, all these structures
-are metastable, as the elastic forces keeping them together dwarf thermal
-energy from Brownian motion (see the scales of energies and energy
-differences in Fig.~S1).
-[TECHNICALLY IN THE METASTABLE STATE WITHOUT BROWNIAN TERMS THERE IS NO NET FORCE ON ANY PARTICLE.
-IS THIS ACTUALLY TRUE?
-IF NOT, I'M CONCERNED A REFEREE COULD JUST SAY WE NEED TO RUN FOR LONGER]
+At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloid strands (the gel in Fig.~1H) is again seen.
+The structures in Fig. 1A-H have been followed for about 20 ms 
+(see Supporting Information for the mapping between simulation and physical 
+time); they correspond to deep metastable minima (elastic interactions are of
+order $KR$ and dwarf thermal motion~\cite{stark,lavrentovich}).
 
 In display devices and other optical applications, it is typical for liquid crystals to be strongly confined (and subject to strong boundary effects), in contrast to the bulk geometries considered above.
-We therefore next describe
-how colloidal suspensions self assemble in thin sandwiches (thickness $h = XXX \lambda$), whose confining walls can favour either tangential (planar) or normal (homeotropic) orientation of the director field.
-For simplicity we here impose one or the other as a strict boundary condition [CONFIRM].
+We therefore next describe how colloidal suspensions self assemble in thin sandwiches (thickness $h = 3.5 \lambda$), whose confining walls can favour either tangential (planar degenerate) or normal (homeotropic) orientation of the director field.
+For simplicity we here impose one or the other as a strict boundary condition.
 
-Fig.~4 shows the case of normal wall anchoring on varying $\phi$ and $w$. We quench from the isotropic
-to the ordered phase as was done in Fig.~3, and again find a metastable, amorphous disclination network (see Supporting Information). The resulting colloid-cholesteric composites 
-are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~4A,C), 
-whereas for large $w$ we observe isolated
-clusters (Fig.~4B) and percolated cellular solids (Fig.~4D), for
-$\phi = 1\%$ and $5\%$ respectively. There are, however, significant 
-effects of the confining walls.
-First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls and also in the midplane (see side-views in Fig.~4).
-Second,
-the clusters formed at large $w$ are themselves anisotropic with an oblate habit.
+Fig.~2A-D shows the case of normal wall anchoring on varying $\phi$ and $w$.
+We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supporting Information).
+The resulting colloid-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $2\%$ respectively.
+There are, however, significant effects of the confining walls.
+First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls and also in the midplane (see side-views in Fig.~2A-D).
+Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit.
 These are similarly to those reported experimentally for large colloids in a cholesteric (but non-BP) host~\cite{niek}.
 
-Fig.~5 shows comparable simulations for planar wall anchoring. 
+Fig.~2E-H shows comparable simulations for planar wall anchoring. 
 Without colloids, planar anchoring results in a regular cholesteric helix,
@@
 dispersing particles with small $w$ arrests the loop shrinking process,
-stabilising twisted figure-of-eight rings (Fig.~5A) at small $\phi$, and plectonemic structures at larger $\phi$ (Fig.~5C).
+stabilising twisted rings (Fig.~2E) at small $\phi$, and more complicated 
+double-stranded structures at larger $\phi$ (Fig.~2F).
 For larger $w$, we observe thick colloidal ``ropes" about 5 colloids across
-(Fig.~5B,D). These ropes undulate to follow the chiral structure of the 
+(Fig.~2E,H). These ropes undulate to follow the chiral structure of the 
 underlying fluid.
-On increasing the density they develop branch points and finally again approach a ``cellular solid" morphology.
-Note that planar wall anchoring is incompatible with the normal anchoring on the colloids (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~5).
+On increasing the density they develop branch points and finally again approach a percolating gel morphology.
+Note that planar wall anchoring is incompatible with the normal anchoring on the colloids (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
 
-FIELD SWITCHING PARA HERE
+An interesting question, relevant for applications to devices, is whether 
+these structures can be modified, or switched, with electric fields.
+Fig.~3 addresses this question, starting again from an amorphous blue phase,
+this time at a higher chirality. Dispersing colloids in this networks
+results in the particle occupying the available nodes, where the order
+parameter is low (Fig.~3A). Applying a high enough field along the $z$ 
+direction leads to a switching to a new phase, where colloidal arrays fit 
+within a network with distorted hexagonal cells (Fig.~3B). Remarkably, upon 
+field removal this honeycomb structure does not find its way back to the 
+configuration before switching on the field, but gets stuck into a metastable 
+structure (Fig. S7), with residual anisotropy along the field switching 
+direction.
+The resulting multistability is indeed a must-have feature for new generation 
+switchable electrooptic devices for use in energy-saving electronics-free
+appliances, smart glass, and e-paper to cite just a few examples.
+
+%FIELD SWITCHING PARA HERE
 \if{The existence of several locally stable competing solids with 
@@
 
-The surprising wealth of colloid-BP composites obtained in Figs.~2-5 is the complex 
-consequence of a simple competition between two self-assembly principles. 
-First, blue phases provide a three-dimensional 
-template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}. Second, colloidal intrusions in a 
-liquid crystal raise the free energy locally; sharing this cost creates a 
-generic tendency to aggregate into clusters~\cite{tiffany}.
+The surprising wealth of colloid-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
+First, blue phases provide a three-dimensional template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}.
+Second, colloidal intrusions in a liquid crystal raise the free energy locally; sharing this cost creates a generic tendency to aggregate into clusters~\cite{tiffany}.
 Which of these factors dominates depends on $w$, a dimensionless measure of the anchoring strength of the director field at the colloidal surface.
@@
 Colloidal crystals, gels or twisted rings can then arise.
-For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating cellular solids, or helical colloidal ropes.
+For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating gels, or helical colloidal ropes.
 In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls.
-MORE ON FIELD SWITCH HERE
-This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid
-crystal devices, which are at the forefront of current technological 
-developments in this field.
-We have not explored in this work their switchability by flow, but this could again be inherited from the underlying BP dynamics \cite{flowswitch}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
+We have also shown that electric fields can be used to switch between metastable states, providing a route to future multistable energy-saving devices.
+%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
+We have not explored in this work their switchability by flow, but this could again be inherited from the underlying BP dynamics \cite{permeation2,adriano}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
 
 \begin{thebibliography}{99}
-\bibitem{stark} H. Stark, {\it Phys. Rep.} {\bf 351}, 387-474 (2001).
+\bibitem{stark} H. Stark. Physics of
+colloidal dispersions in nematic liquid crystals. 
+{\it Phys. Rep.} {\bf 351}, 387-474 (2001).
 \bibitem{mermin} Wright D.~C., Mermin, N.~D. Crystalline liquids: The blue phases. {\it Rev. Mod. Phys.} {\bf 61}, 385–432 (1989).
 \bibitem{miha} M. Ravnik, G.~P. Alexander, J.~M. Yeomans,  
-S. Zumer, {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188-5192 (2011).
-\bibitem{lavrentovich} O.~D. Lavrentovich, {\it Proc. Natl. Acad. Sci.
-USA} {\bf 108}, 5143-5144 (2011).
+S. Zumer.  Three-dimensional colloidal crystals in liquid crystalline 
+blue phases. {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188-5192 (2011).
+\bibitem{lavrentovich} O.~D. Lavrentovich.
+ Liquid crystals, photonic crystals, metamaterials and transformation optics. 
+{\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5143-5144 (2011).
 \bibitem{epaper} J. Heikenfeld, P. Drzaic,  J.~S. Yeo, T. Koch.
@@
 \bibitem{zumer} I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik,
-S. Zumer, {\it Science} {\bf 313}, 954-958 (2006).
+S. Zumer. Two-dimensional nematic colloidal crystals self-assembled by topological defects. {\it Science} {\bf 313}, 954-958 (2006).
 \bibitem{tiffany} T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo,
-W.~C.~K. Poon, {\it Science} {\bf 333}, 79-83 (2011).
-\bibitem{abbott} I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo, N.~L. Abbott, {\it Science} {\bf 332}, 1297-1300 (2011).
+W.~C.~K. Poon. A Self-Quenched Defect Glass in a Colloid-Nematic Liquid Crystal Composite. {\it Science} {\bf 333}, 79-83 (2011).
+\bibitem{abbott} I.~H.
+Lin, D.~S.
+Miller, P.~J.
+Bertics, C.~J.
+Murphy, J.~J.
+de Pablo, N.~L.
+Abbott.
+Endotoxin-Induced Structural Transformations in Liquid Crystalline Droplet.
+{\it Science} {\bf 332}, 1297-1300 (2011).
 \bibitem{colloiddevice} K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton,
-R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006). REPLACE WITH A SUITABLE REVIEW IF POSSIBLE?
-\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}
-{\bf 35}, 4419 (1987).
+R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
+\bibitem{choldevice} D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
+Bistable cholesteric reflective displays: Materials and Drive Schemes,
+{\it Annu. Rev. Mater. Sci.} {\bf 27}, 117-146 (1997).
+%\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}
+%{\bf 35}, 4419 (1987).
 \bibitem{kikuchi} H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama.
@@
 {\it Nature} {\bf 436}, 997-1000 (2005).
-\bibitem{bpdevice} Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
+\bibitem{bpdevice} Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu.
+ Electro-optics of polymer-stabilized blue phase liquid crystal displays.
 {\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
@@
 and smectic liquid crystals. {\it Phys. Rev. Lett.} {\bf 23}, 372-374 (1969).
-\bibitem{permeation2} A. Dupuis, D. Marenduzz, E. Orlandini, J.~M. Yeomans.
+\bibitem{permeation2} A. Dupuis, D. Marenduzzo, E. Orlandini, J.~M. Yeomans.
 Rheology of cholesteric blue phases. {\it Phys. Rev. Lett.} {\bf 95},
@@
 \bibitem{lubensky} M. Zapotocky, L. Ramos, P. Poulin, T.~C. Lubensky, 
-D.~A. Weitz, {\it Science} {\bf 283}, 209-212 (1999).
-\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles, {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
-\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates,
-{\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
-\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
+D.~A. Weitz.Particle-stabilized defect gel in cholesteric liquid crystals.  {\it Science} {\bf 283}, 209-212 (1999).
+\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles.
+Blue-phase templated fabrication of three-dimensional nanostructures for photonic applications. {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
+\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates.
+ Ordering dynamics of blue phases entails kinetic stabilization of amorphous networks. {\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
+\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo.
+ Structure of Blue Phase III of Cholesteric Liquid Crystals.
 {\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
-\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic, {\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
-\bibitem{tanaka} T. Araki, H. Tanaka, {\it Phys. Rev. Lett.} {\bf 97}, 
-127801 (2006).
-\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka,
-{\it Nat. Mat.} {\bf 10}, 303-309 (2011).
+\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic.  Entangled nematic colloidal dimers and wires. 
+{\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
+\bibitem{tanaka} T.
+Araki, H.
+Tanaka.
+Colloidal aggregation in a nematic liquid crystal: Topological arrest of particles by a single-stroke disclination line.
+{\it Phys.
+Rev.
+Lett.} {\bf 97}, 127801 (2006).
+\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
+Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
 \bibitem{adriano} A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini,
-F. Salvadore, {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
-\bibitem{fukuda} J.-I. Fukuda, S. Zumer, {\it Nat. Comm.} {\bf 2}, 246 (2011).
+F. Salvadore. Bistable Defect Structures In Blue Phase Devices. 
+ {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
+\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
+Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
+{\it Nat. Comm.} {\bf 2}, 246 (2011).
 \bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
@@
  Liquid Crystal. {\it Langmuir} {\bf 26}, 13502-13510 (2010).
-\bibitem{henrichfield} Henrich Field PRE 
-\bibitem{juho1} juho: microrheology of cholesterics
-\bibitem{juho2} juho: dimers in cholesterics
-\bibitem{flowswitch} Davide et al on flow reorganization of BPs
+\bibitem{henrichfield}  O. Henrich, D. Marenduzzo, K. Stratford, 
+M. E. Cates. Thermodynamics of blue phases in electric fields.
+{\it Phys. Rev. E} {\bf 81}, 031706 (2010).
+\bibitem{juho1} J. S. Lintuvuori, K. Stratford, M. E. Cates,
+D. Marenduzzo.   Colloids in Cholesterics: Size-Dependent Defects and Non-Stokesian Microrheology. {\it Phys. Rev. Lett.} {\bf 105}, 178302 (2010).
+\bibitem{juho2}  J. S. Lintuvuori, K. Stratford, M. E. Cates,
+D. Marenduzzo. 
+Self-Assembly and Nonlinear Dynamics of Dimeric Colloidal Rotors in 
+Cholesterics. {\it Phys. Rev. Lett.} {\bf 107}, 267802 (2011).
+%\bibitem{flowswitch} Davide et al on flow reorganization of BPs
 \bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang, Developing optofluidic technology through the fusion of microfluidics and optics {\it Nature} {\bf 442}, 381-386 (2006).
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{s0.png}
-\caption{Schematics of the molecular arrangement in a 
-cholesteric helix (A) and within a double twist cylinder (B, the
-axis of the cylinder is perpendicular to the plane of the page).
-The disclinations line arrangement in the unit cell of BPI is shown in
-(C), while a reference phase diagram (from Ref.~\cite{crooker}) is
-shown in (D). [I SEE NO REASON FOR INCLUDING PHASE DIAGRAM HERE. PUT IN SI IF NEEDED -- BUT SURELY IT IS VITAL TO APPLICATIONS THAT THIS IS *NOT* THE PHASE DIAGRAM FOR COLES OR KIKUCHI!]}
+\centerline{\includegraphics[width=15.cm]{Fig1.jpg}}
+\caption{Snapshots of the steady states obtained after dispersing
+a suspension of colloidal nanoparticles within a cholesteric liquid
+crystal in the BPI-forming region. (A-D) correspond to the case
+of a dispersion of particles in a pre-equilibrated BPI phase;
+(E-H) are obtained by dispersing colloids in an isotropic
+phase and then quenching into the range where BPI is stable, leading to formation of an amorphous, BPIII-like, disclination network.
+Structures correspond
+to (A,E) $w=0.23$ and $\phi=1\%$, 
+(B,F) $w=0.23$ and $\phi=5\%$, 
+(C,G) $w=2.3$ and $\phi=1\%$,
+(D,H) $w=2.3$ and $\phi=5\%$.
+The anchoring of the director field to the colloidal surface is normal.
+[For the full parameter list used to generate Figs.~1-3 see Supporting Information.] For clarity, only a portion of the simulation box (a quarter of
+each axis in size) is shown; the full structures are shown in Fig. S2 and S3.}
 \end{figure}
 
+%\begin{figure}
+%\includegraphics[width=\textwidth]{s2.jpg}
+%\caption{Steady states obtained by dispersing colloids in an isotropic phase and then quenching into the range where BPI is stable, leading to formation of an amorphous disclination network.
+Parameters otherwise as in Fig.~2.}
+%\end{figure}
+
 \begin{figure}
-\includegraphics[width=\textwidth]{s1.jpg}
-\caption{Snapshots of the steady states obtained after dispersing
-a suspension of colloidal nanoparticles within a pre-equilibrated  BPI phase.  Structures correspond
-to (A) $w=0.23$ and $\phi=1\%$, (B) $w=2.3$ and $\phi=1\%$,
-(C) $w=0.23$ and $\phi=5\%$, (D) $w=2.3$ and $\phi=5\%$.
-The anchoring of the director field to the colloidal surface is normal. [ISN'T THIS AUTOMATIC FOR $w>0$?]
-[For the full parameter list used to generate Figs.~2-5 see Supporting Information.] [FIX COLOURS AND VISIBILITY FOR ALL FIGURES.]}
+\centerline{\includegraphics[width=15.cm]{Fig2.jpg}}
+\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, 
+and then quenched into a regime where BPI is stable in the bulk. Both top and side views are provided.
+The anchoring of the director field at the walls is normal for (A-D) and planar for (E-H). 
+Structures correspond to (A,E) $w=0.23$ and $\phi=1\%$, 
+(B,F) $w=0.23$ and $\phi=2\%$, 
+(C,G) $w=2.3$ and $\phi=1\%$,
+(D,H) $w=2.3$ and $\phi=2\%$.
+As in Fig. 1, only a portion of the simulation box is shown for clarity;
+the full structures are shown in Fig. S5 and S6.}
 \end{figure}
 
-\begin{figure}
-\includegraphics[width=\textwidth]{s2.jpg}
-\caption{Steady states obtained by dispersing colloids in an isotropic phase and then quenching into the range where BPI is stable, leading to formation of an amorphous disclination network.
-Parameters otherwise as in Fig.~2.}
-\end{figure}
+%\begin{figure}
+%\includegraphics[width=\textwidth]{s4.jpg}
+%\caption{As Fig.~4, but with planar anchoring of the director 
+%field at the boundaries.}
+%\end{figure}
 
 \begin{figure}
-\includegraphics[width=\textwidth]{s3.jpg}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, with normal anchoring of the director field
-at the boundaries, and then quenched into a regime where BPI [CONFIRM] is stable in the bulk. Both top and side views are provided. 
-Structures correspond
-to (A) $w=XXX$ and $\phi=1\%$, (B) $w=XXX$ and $\phi=1\%$,
-(C) $w=XXX$ and $\phi=2\%$, (D) $w=XXX$ and $\phi=2\%$.
-}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{s4.jpg}
-\caption{As Fig.~4, but with planar anchoring of the director 
-field at the boundaries.}
+\centerline{\includegraphics[width=15.cm]{Fig3.png}}
+\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase, then quenched into a regime where BPIII is stable in the bulk (A), and finally subjected to an electric field along the vertical direction (B).
+The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.
+Structures correspond to $w=0.23$ and $\phi=1\%$ (see Supporting Information
+for full list of parameters). }
 \end{figure}

### 2013-10-02 trunk/papers/2012-sandwich/CholColloids.tex
Main text to include new figures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2155 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-10-02 papers/2012-sandwich/CholColloids.tex
@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  Main text for ``sandwich'' manuscript
+%
+%  Process with ``pdflatex CholColloids.tex''
+%
+%  The supporting material text is support.tex
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \documentclass[12pt]{article}
@@
 \begin{figure}
-\centerline{\includegraphics[width=15.cm]{Fig1.jpg}}
-\caption{Snapshots of the steady states obtained after dispersing
+
+\centerline{\includegraphics[width=\textwidth]{text-fig1.jpg}}
+\caption{Snapshots of the states obtained after dispersing
 a suspension of colloidal nanoparticles within a cholesteric liquid
@@
 
-%\begin{figure}
-%\includegraphics[width=\textwidth]{s2.jpg}
-%\caption{Steady states obtained by dispersing colloids in an isotropic phase and then quenching into the range where BPI is stable, leading to formation of an amorphous disclination network.
-Parameters otherwise as in Fig.~2.}
-%\end{figure}
-
 \begin{figure}
-\centerline{\includegraphics[width=15.cm]{Fig2.jpg}}
+\centerline{\includegraphics[width=\textwidth]{text-fig2.jpg}}
 \caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, 
@@
 
-%\begin{figure}
-%\includegraphics[width=\textwidth]{s4.jpg}
-%\caption{As Fig.~4, but with planar anchoring of the director 
-%field at the boundaries.}
-%\end{figure}
-
 \begin{figure}
-\centerline{\includegraphics[width=15.cm]{Fig3.png}}
+\centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
 \caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase, then quenched into a regime where BPIII is stable in the bulk (A), and finally subjected to an electric field along the vertical direction (B).

### 2013-10-02 papers/2012-sandwich/support.tex
Updates to supporting material
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2156 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \partial_\gamma Q_{\delta \beta} + 2 q_0 Q_{\alpha \beta})^2
-+ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2
++ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2.
 \label{free}
@@
 
-The remaining terms of the free energy density in Eq.~\ref{free} describe
+The remaining two terms of the free energy density in Eq.~\ref{free} describe
 distortions of the order parameter field. In theoretical work and when
@@
 $\kappa$, the chirality, and  $\tau$, the reduced temperature
-\cite{blue1}.  In terms of the parameters in the free energy density,
-these are:
+\cite{blue1} which can be used to characterise the system. In terms of
+the parameters in the free energy density, these are:
 \begin{eqnarray}\label{cntrl-param} 
@@
 \cite{blue1,oliver1}.
+
 In this work, the chirality and reduced temperature are always set to lie
-within the blue phase I region of the equilibrium phase diagram.
+within the blue phase I region of the equilibrium phase diagram. EXCEPT?
 
@@
 The effect of an electric field can be modelled by including the following
-term in the free energy density,
+term in the bulk free energy density:
 \begin{equation}
--(\epsilon_a/12\pi)E_{\alpha}Q_{\alpha\beta}E_{\beta}
+-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
 \end{equation} 
-where $E_{\beta}$ are the components of the electric field, and 
+where $E_{\alpha}$ are the components of the electric field, and 
 $\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
@@
 \begin{equation}
-{\cal E}^2=(27 \varepsilon_a E_\alpha E_\alpha)/(32 \pi A_0 \gamma).
+{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
 \end{equation}
@@
 degenerate planar anchoring is applied at the walls, where the
-preferred nematic director may lay in any orientation parallel
+preferred nematic director may adopt any orientation parallel
 to the surface. The preferred direction is determined explicitly
@@
 The parameters for the free energy are chosen to be representative
-of blue phase I: chirality $\kappa = 0.7$ and reduced temperature
-$\tau = 0$.
+of blue phase I: chirality $\kappa \sim 0.7$ and reduced temperature
+$\tau = -0.25$.
 The full free energy parameters are shown in Table~\ref{table:params}.
@@
 unit cells with fully periodic boundary conditions.
-Simulations at different solid volume fractions (1\% and 5\%), and different 
-surface anchoring strengths ($Wa/K = 0.23$ and $Wa/k = 2.3$) representing
-``weak'' and ``strong'' anchoring are run for 500,000 simulation
-time steps (and 700,000 time steps for 5\% solid volume fraction).
+Simulations at different solid volume fractions (1\% and 4\%), and different 
+surface anchoring strengths ($Wa/K = 0.23$ and $Wa/K = 2.3$) representing
+``weak'' and ``strong'' anchoring are run for two million simulation
+time steps.
 The initial distribution of the colloid positions is set at random
@@
 is performed in which the order parameter is initialised via a
-locally chosen random director (and using an equation of the form
-Eq.~\ref{eq:surf_q} with an initial amplitude of order $q = -0.2$).
-In order to avoid the mean-field spinodal point at $\gamma = 3.0$,
-the parameters are adjusted slightly while maintaining the same
+locally chosen random director, and
+Eq.~\ref{eq:surf_q} employed with a small amplitude $q^0 = 10^{-7}$ to
+provide an order parameter.
+We avoid the mean-field spinodal point at $\gamma = 3.0$:
+the parameters are the same as for the initial simulations with
 pitch $p = 32\sqrt{2}$. The chirality and reduced temperature (see
 Table~\ref{table:params}) remain appropriate for blue phase I.
-Colloids are added as before and the simulations run for 500,000 or 
-700,000 simulation steps.
+Colloids are added as before and the simulations run for 2 million
+simulation steps.
 
@@
 initialised randomly. The free energy parameters (see Table~\ref{table:params})
-are again appropriate for blue phase I (chirality $\kappa = 0.8$ and
+are again appropriate for (equilibrium) blue phase I
+(chirality $\kappa = 0.8$ and
 reduced temperature $\tau = -0.25$). The cholesteric pitch is set
-to be $p = 64$, providing a slight better resolution that the bulk
+to be $p = 64$, providing a slight better resolution than the bulk
 simulations.
@@
 1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
-in simulation units. All simulations are run for 1.6 million simulations
-time steps unless otherwise stated CHECK.
+in simulation units. All simulations are run for two million simulation
+time steps.
+
+{\bf 4.3 External electric field}
+
+OLIVER PLEASE ADD short description of protocol here. AND the
+parameters in the table.
 
@@
 \hline 
-Bulk BPI & 0.01 & 3.000 & 0.007061 & 32$\sqrt{2}$ & 0.7000 & 0.0000 &
+Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
 0.23, 2.3\\
 \hline
-Bulk Quench & 0.01 & 3.100 & 0.007061 & 32$\sqrt{2}$ & 0.6886 & -0.2903 &
+Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
 0.23, 2.3 \\
 \hline
-Confined & 0.01 & 3.0857 & 0.01897 & 64 & 0.8 & -0.25 & 0.23, 2.3\\
+Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
+\hline
+External field & a0& gamma & K & p & kappa & tau & Wa/K\\ 
 \hline
@@
 \end{center}
-\caption{Free energy parameters used in the simulations.}
+\caption{Free energy parameters used in the simulations. Bulk BPI
+correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
+geometries are shown in Fig.~2; and the external field case is
+relevant for Fig.~3.}
 \label{table:params}
@@
 largest eigenvalue of the local order parameter tensor. A low value
-of $q$ (typically 0.12) unambiguously identifies the disclinations.
+of the scalar order parameter $q$ (typically in the range 0.12--0.14)
+unambiguously identifies the disclinations.
 
@@
 
-To get from simulation to physical units, we need to calibrate scales of length,
-energy and time. For similar mappings, see e.g.~\cite{denniston2}.
+To get from simulation to physical units, we need to calibrate scales of
+length, energy and time in the simulations. For similar mappings see,
+e.g.~\cite{denniston2}.
 
-The length scale can be set by mapping our half pitch to realistic blue phase
-unit cell size, which are in the 100-500 nm range \cite{blue1}. 
-For bulk simulations (Fig. 1 and Fig. 3 in the main text), a suitable choice
-is that that one simulation unit (lattice site) corresponds to 10 nm. Therefore
-the colloidal size in Figs. 1 and 3 corresponds to $\sim$ 50 nm.
+The length scale can be set by mapping our half pitch to a realistic blue
+phase unit cell size, which are in the 100-500 nm range \cite{blue1}. 
+For bulk simulations (Fig.~1 and Fig.~3 in the main text), a suitable choice
+is one where one simulation unit (lattice site) corresponds to 10~nm. Therefore
+the colloidal size in Figs.~1 and~3 corresponds to $\sim$ 50~nm.
 
-To get an energy scale, we choose $A_0 \simeq 10^6$ Pa, which is 
+To obtain an energy scale, we choose $A_0 \simeq 10^6$~Pa, which is 
 reasonable from Ref.~\cite{blue1} (this is also the choice in
-Ref.~\cite{oliver2}). This leads to a simulation unit of
-free energy (or stress) equal to $10^{8}$Pa in Figs. 1 and 3.
+Ref.~\cite{oliver2}). This choice leads to a simulation unit of
+free energy (or stress) equal to $10^{8}$~Pa in Figs.~1 and~3.
 From the energy and length scales one can map elastic constants to
-physical units: for instance the simulation of Fig. 1 corresponds to
+physical units: for instance, the simulation in Fig.~1 corresponds to
 a liquid crystal with splay, bend and twist (Frank) elastic constants
-equal to 35 pN (or equivalently $K=$70 pN).
+equal to 35 pN (or equivalently $K=$70~pN). That is,
 \begin{equation}
@@
 \end{equation}
+WHAT ARE a/b???
 
-For the timescale calibration we use the formula (see e.g.~\cite{denniston1})
+For the timescale calibration we use the formula (see e.g.~\cite{denniston})
 \begin{equation}
-\gamma_1=\frac{2q^2}{\Gamma}\;.
+\gamma_1=\frac{2q^2}{\Gamma}
 \end{equation}
- which relates the `rotational viscosity' $\gamma_1$, to the ordering strength $q$ and the order parameter mobility $\Gamma$.
-In our simulations, we chose $\Gamma = 0.5$, while the value of $\gamma$ used in Fig. 1 leads to $q=1/2$.   
-For real liquid crystalline materials, $\gamma_1$ lies usually in range $10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness $\gamma_1 = 1$ Pa s (equivalently 10 poise).
-Given the previous mapping for free energy (or stress) units, this leads to a simulation unit of time equal to $10^{-8}$ s.
-The total simulated time in Figs.
-1-3 is therefore in the ms range.
+which relates the rotational viscosity $\gamma_1$, to the ordering strength
+$q$ and the order parameter mobility $\Gamma$.
+In our simulations, we choose $\Gamma = 0.5$, while the value of $\gamma$
+used in Fig.~1 leads to $q=1/2$.   
+For real liquid crystalline materials, $\gamma_1$ usually lies in range
+$10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness
+$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
+free energy (or stress) units, this leads to a simulation unit of time equal
+to $10^{-8}$~s. The total simulated time in Figs. 1-3 is therefore in the
+millisecond range.
 
@@
 dielectric permittivity of vacuum), we then obtain that an electric field
-of 5 V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.
+of 5~V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.
 
 Given these mappings, all quantities can be easily transformed from
-simulation to physical units and vice versa. For example Fig. 1 and 2
-use an isotropic viscosity $\eta \simeq$ 0.01 and 0.1 in simulation units,
-which translate into $0.01-0.1$ Pa s respectively. This is sensible (if 
+simulation to physical units and vice versa. For example Figs.~1 and~2
+use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
+which translate into $0.01-0.1$~Pa~s respectively. This is sensible (if 
 slightly low) for a molecular nematogen in the isotropic phase. 
@@
 in our model by the coupling to the order parameter.
-We note that the density is set to 1 in our simulations: this
+We note that the density is set to unity in our simulations: this
 corresponds to a fluid density which is about a thousand times larger
@@
 
-To summarize the above, our simulations represent BP-forming materials, with the interpretation of the LB units for length, time, and energy density are close to 10 nm, 10 ns, and 100 MPa respectively.
+To summarize, our simulations represent BP-forming materials, with the
+interpretation of the simulation units for length, time, and energy
+density being close to 10~nm, 10~ns, and 100~MPa respectively.
 
@@
 
-%ISSUE: no redshift?
-
 %ISSUE: do we want some free energy numbers?
 
-{\bf 6. Supporting Figures and Animations}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.35]{s1.jpg}
-\end{center}
-\caption{As for main manuscript Fig.~1A-E (initialised to equilibrium blue
-phase I), but for the entire simulation system of 128$^3$ lattice sites. (a) 
-solid volume fraction 1\% and weak anchoring; 
-(b) 1\% solid volume fraction and strong anchoring;
-(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
-volume fraction and strong anchoring. The view direction in the
-main figure is from the left here. There is a reference
-particle at the bottom left in each panel which does not take part
-in the simulation.}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.35]{s2.jpg}
-\caption{As for main manuscript Fig.~1E-H (initialised via a ``quench''
-to generate a disordered network), but for the entire simulation
-system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
-weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
-(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
-volume fraction and strong anchoring. Again, there is a reference
-particle at the bottom left in each case which does not take part
-in the simulation.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.42]{s3.jpg}
-\end{center}
-\caption{As for main manuscript Fig.~3 (confined geometry with normal
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (a) solid volume fraction 1\%
-and weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
-(c) 4\% solid volume fraction and weak anchoring; (d) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.42]{s4.jpg}
-\end{center}
-\caption{As for main manuscript Fig.~4 (confined geometry with planar
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (a) colloid solid volume fraction 1\%
-and weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
-(c) 4\% solid volume fraction and weak anchoring; (d) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.95\columnwidth]{s5.jpg}
-\end{center}
-\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
-(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
-(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
-\end{figure}
-
-\newpage
-
-{\bf Supporting References}
+\vfill
+\pagebreak
 
@@
 
+\bibitem{codef} M. E. Cates,  et al., 
+\textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).
+
 \end{thebibliography}
 
+\vfill\pagebreak
+
+{\bf Supporting Figures and Animations}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=0.8]{doubletwist.pdf}
+\end{center}
+\caption{
+(A) Schematic representation of the director field
+in a cholesteric (left) and within a double twist cylinder (right).
+(B) Snapshot of the disclination network in an equilibrated 
+blue phase I structure. (C) Same as (B) for an amorphous blue phase III
+network.}
+\end{figure}
+
+\vfill\pagebreak
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig2.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~1 A--D
+(with liquid crystal order parameter
+initialised to equilibrium blue phase I structure), but for the entire
+simulation system of 128$^3$ lattice sites. (A)  solid volume fraction~1\%
+and weak anchoring;  (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring. The view direction in the
+main figure is from the left here. There is a reference
+particle at the bottom left in each panel which does not take part
+in the simulation.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig3.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~1 E--H (initialised via a ``quench''
+to generate a disordered network), but for the entire simulation
+system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
+weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring. Again, there is a reference
+particle at the bottom left in each case which does not take part
+in the simulation.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig4.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~2 A--D (confined geometry with normal
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (A) solid volume fraction 1\%
+and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig5.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~2 E--H (confined geometry with planar
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (E) colloid solid volume fraction 1\%
+and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+%%\includegraphics[width=0.95\columnwidth]{s5.jpg}
+\end{center}
+\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
+(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
+(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
+\end{figure}
+
+%{\bf Captions for Supplementary Movies}
+
 \end{document}
-
-% MY NOTES
-
-KEVIN
-
-All particle radius $a = 2.3$.
-
-Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = 10\eta$;
-temperature $T =2.133333 10^{-9}$.
-
-Free energy:
-$A_0 = 0.01$; $\gamma = 3.085143$; $q_0 = 0.09817477$ (and $p = 64$);
-$\kappa_1 = \kappa_2 = 0.018971857$;
-Chirality $\bar{\kappa} = 0.8$; $\tau = -0.25$
-
-Amplitude order $ 0.0001$; Computed surface order $0.3509236$;
-Rotational diffusion $\Gamma = 0.5$
-Initialisation is 'random'.
-
-Wall surface $W = 0.018971957$, ie., $W/\kappa$ = 1 length
-
-Weak colloid anchoring $W = 0.0018971957$, ie., $W/\kappa = 0.1$ length
-and $Wa/\kappa = 0.23$.
-
-Strong colloid anchoring $W = 0.037943914$, ie., $W/\kappa = 2.0$ length
-or $Wa/\kappa = 4.6$.
-
-JUHO (BPI)
-
-Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = \eta$;
-Temperature: $T = 2.133333 10^{-9}$.
-
-Free energy: $A_0 = 0.01$; $\gamma = 3.0$; $q_0 = 0.13884$; (and $p = 45.25$
-is $32\sqrt{2}$);
-$\kappa_1 = \kappa_2 = 0.00706095$;
-Chirality $\bar{\kappa} = 0.7$ Reduced temperature $\tau = 0.0$.
-
-Amplitude of order $-0.2$; Computed surface order:
-Rotational diffusion $\Gamma = 0.5$
-
-Wall irrelevant
-
-Weak colloid anchoring $W = 0.000706095$, ie., $W/\kappa = 0.1$ length
-and $Wa / \kappa = 2.3$.
-
-Strong colloid anchoring $W = 0.00706095$, i.e., $W/\kappa = 1.0$
-and $Wa/\kappa = 1.0$.
-
-JUHO (QUENCH)
-
-Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = \eta$;
-Temperature: $T = 2.133333 10^{-9}$.
-
-Free energy: $A_0 = 0.01$; $\gamma = 3.1$; $q_0 = 0.13884$; (and $p = 45.25$
-is $32\sqrt{2}$);
-$\kappa_1 = \kappa_2 = 0.00706095$;
-Chirality $\bar{\kappa} = 0.68862$ Reduced temperature $\tau = -2.90322$.

### 2013-10-02 trunk/papers/2012-sandwich/support.tex
Added electric field protocol
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2159 40876441-3496-0410-a613-a2fca35b2ae2

@@
 1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
-in simulation units. All simulations are run for 1.6 million simulations
-time steps unless otherwise stated CHECK.
+in simulation units. All simulations are run for two million simulation
+time steps.
+
+{\bf 4.3 External electric field}
+
+The following protocol was used for the simulations of BPIII with
+colloidal particles: First, we started from a configuration of
+randomly oriented and positioned double twist cylinders from which an
+amorphous BPIII network emerged. The structure
+was equilibrated for $6\times10^{5}$ LB time steps until the relaxation came
+virtually to a standstill. Then the colloidal particles featuring week
+normal anchoring were randomly inserted and the composite system was
+equilibrated for another $1.2\times10^{6}$ time steps.
+At the end of the equilibration phase the homogenous electric field
+with reduced field strength $e=\sqrt{\frac{27\epsilon_a}{32\pi A_0\gamma}} E=0.8$ 
+was switched on along the z-direction for $1.2\times10^{6}$ time steps.
+This was found to be sufficiently long for the hexagonal structure to emerge and saturate.
+After the electric field has been switched off the system quickly relaxed
+to a metastable state which showed a residual anisotropy in the
+colloid distribution. The total length of this final relaxation phase
+was $1\times10^6$ LB time steps.
 
@@
 \hline 
-Bulk BPI & 0.01 & 3.000 & 0.007061 & 32$\sqrt{2}$ & 0.7000 & 0.0000 &
+Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
 0.23, 2.3\\
 \hline
-Bulk Quench & 0.01 & 3.100 & 0.007061 & 32$\sqrt{2}$ & 0.6886 & -0.2903 &
+Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
 0.23, 2.3 \\
 \hline
-Confined & 0.01 & 3.0857 & 0.01897 & 64 & 0.8 & -0.25 & 0.23, 2.3\\
+Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
+\hline
+External field & 0.004& 3.086 & 0.02 & 32 & -0.25 & 2.6 & 0.23\\ 
 \hline
@@
 \end{equation}
+WHAT ARE a/b???
 
@@
 \begin{equation}
-\gamma_1=\frac{2q^2}{\Gamma}\;.
+\gamma_1=\frac{2q^2}{\Gamma}
 \end{equation}
- which relates the `rotational viscosity' $\gamma_1$, to the ordering strength $q$ and the order parameter mobility $\Gamma$.
-In our simulations, we chose $\Gamma = 0.5$, while the value of $\gamma$ used in Fig. 1 leads to $q=1/2$.   
-For real liquid crystalline materials, $\gamma_1$ lies usually in range $10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness $\gamma_1 = 1$ Pa s (equivalently 10 poise).
-Given the previous mapping for free energy (or stress) units, this leads to a simulation unit of time equal to $10^{-8}$ s.
-The total simulated time in Figs.
-1-3 is therefore in the ms range.
+which relates the rotational viscosity $\gamma_1$, to the ordering strength
+$q$ and the order parameter mobility $\Gamma$.
+In our simulations, we choose $\Gamma = 0.5$, while the value of $\gamma$
+used in Fig.~1 leads to $q=1/2$.   
+For real liquid crystalline materials, $\gamma_1$ usually lies in range
+$10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness
+$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
+free energy (or stress) units, this leads to a simulation unit of time equal
+to $10^{-8}$~s. The total simulated time in Figs. 1-3 is therefore in the
+millisecond range.
 
@@
 
-%ISSUE: no redshift?
-
 %ISSUE: do we want some free energy numbers?
 
-{\bf 6. Supporting Figures and Animations}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.8]{SupplFig1.png}
-\end{center}
-\caption{(A) Schematic representation of the director field
-in a cholesteric (left) and within a double twist cylinder (right).
-(B) Snapshot of the disclination network in an equilibrated 
-blue phase I structure. (C) Same as (B) for an amorphous blue phase III
-network.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.35]{s1.jpg}
-\end{center}
-\caption{As for main manuscript Fig.~1A-E (initialised to equilibrium blue
-phase I), but for the entire simulation system of 128$^3$ lattice sites. (a) 
-solid volume fraction 1\% and weak anchoring; 
-(b) 1\% solid volume fraction and strong anchoring;
-(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
-volume fraction and strong anchoring. The view direction in the
-main figure is from the left here. There is a reference
-particle at the bottom left in each panel which does not take part
-in the simulation.}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.35]{s2.jpg}
-\caption{As for main manuscript Fig.~1E-H (initialised via a ``quench''
-to generate a disordered network), but for the entire simulation
-system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
-weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
-(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
-volume fraction and strong anchoring. Again, there is a reference
-particle at the bottom left in each case which does not take part
-in the simulation.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.42]{s3.jpg}
-\end{center}
-\caption{As for main manuscript Fig.~2A-D (confined geometry with normal
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (a) solid volume fraction 1\%
-and weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
-(c) 4\% solid volume fraction and weak anchoring; (d) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.42]{s4.jpg}
-\end{center}
-\caption{As for main manuscript Fig.~2E-H (confined geometry with planar
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (a) colloid solid volume fraction 1\%
-and weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
-(c) 4\% solid volume fraction and weak anchoring; (d) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=0.95\columnwidth]{s5.jpg}
-\end{center}
-\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
-(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
-(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
-\end{figure}
-
-\newpage
-
-{\bf Captions for Supplementary Movies} \\
-
-%{\bf Supporting References}
+\vfill
+\pagebreak
 
@@
 
+\vfill\pagebreak
+
+{\bf Supporting Figures and Animations}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=0.8]{doubletwist.pdf}
+\end{center}
+\caption{
+(A) Schematic representation of the director field
+in a cholesteric (left) and within a double twist cylinder (right).
+(B) Snapshot of the disclination network in an equilibrated 
+blue phase I structure. (C) Same as (B) for an amorphous blue phase III
+network.}
+\end{figure}
+
+\vfill\pagebreak
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig2.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~1 A--D
+(with liquid crystal order parameter
+initialised to equilibrium blue phase I structure), but for the entire
+simulation system of 128$^3$ lattice sites. (A)  solid volume fraction~1\%
+and weak anchoring;  (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring. The view direction in the
+main figure is from the left here. There is a reference
+particle at the bottom left in each panel which does not take part
+in the simulation.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig3.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~1 E--H (initialised via a ``quench''
+to generate a disordered network), but for the entire simulation
+system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
+weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring. Again, there is a reference
+particle at the bottom left in each case which does not take part
+in the simulation.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig4.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~2 A--D (confined geometry with normal
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (A) solid volume fraction 1\%
+and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig5.pdf}
+\end{center}
+\caption{As for main manuscript Fig.~2 E--H (confined geometry with planar
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (E) colloid solid volume fraction 1\%
+and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
+
+\begin{figure}[t]
+\begin{center}
+%%\includegraphics[width=0.95\columnwidth]{s5.jpg}
+\end{center}
+\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
+(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
+(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
+\end{figure}
+
+%{\bf Captions for Supplementary Movies}
+
 \end{document}
-
-% MY NOTES
-
-KEVIN
-
-All particle radius $a = 2.3$.
-
-Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = 10\eta$;
-temperature $T =2.133333 10^{-9}$.
-
-Free energy:
-$A_0 = 0.01$; $\gamma = 3.085143$; $q_0 = 0.09817477$ (and $p = 64$);
-$\kappa_1 = \kappa_2 = 0.018971857$;
-Chirality $\bar{\kappa} = 0.8$; $\tau = -0.25$
-
-Amplitude order $ 0.0001$; Computed surface order $0.3509236$;
-Rotational diffusion $\Gamma = 0.5$
-Initialisation is 'random'.
-
-Wall surface $W = 0.018971957$, ie., $W/\kappa$ = 1 length
-
-Weak colloid anchoring $W = 0.0018971957$, ie., $W/\kappa = 0.1$ length
-and $Wa/\kappa = 0.23$.
-
-Strong colloid anchoring $W = 0.037943914$, ie., $W/\kappa = 2.0$ length
-or $Wa/\kappa = 4.6$.
-
-JUHO (BPI)
-
-Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = \eta$;
-Temperature: $T = 2.133333 10^{-9}$.
-
-Free energy: $A_0 = 0.01$; $\gamma = 3.0$; $q_0 = 0.13884$; (and $p = 45.25$
-is $32\sqrt{2}$);
-$\kappa_1 = \kappa_2 = 0.00706095$;
-Chirality $\bar{\kappa} = 0.7$ Reduced temperature $\tau = 0.0$.
-
-Amplitude of order $-0.2$; Computed surface order:
-Rotational diffusion $\Gamma = 0.5$
-
-Wall irrelevant
-
-Weak colloid anchoring $W = 0.000706095$, ie., $W/\kappa = 0.1$ length
-and $Wa / \kappa = 2.3$.
-
-Strong colloid anchoring $W = 0.00706095$, i.e., $W/\kappa = 1.0$
-and $Wa/\kappa = 1.0$.
-
-JUHO (QUENCH)
-
-Fluid properties: $\rho_0 = 1$; $\eta = 0.01$; $\zeta = \eta$;
-Temperature: $T = 2.133333 10^{-9}$.
-
-Free energy: $A_0 = 0.01$; $\gamma = 3.1$; $q_0 = 0.13884$; (and $p = 45.25$
-is $32\sqrt{2}$);
-$\kappa_1 = \kappa_2 = 0.00706095$;
-Chirality $\bar{\kappa} = 0.68862$ Reduced temperature $\tau = -2.90322$.

### 2013-10-02 trunk/papers/2012-sandwich/support.tex
updated supporting material
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2160 40876441-3496-0410-a613-a2fca35b2ae2

@@
 In this work, the chirality and reduced temperature are always set to lie
-within the blue phase I region of the equilibrium phase diagram. EXCEPT?
+within the blue phase I region of the equilibrium phase diagram, except
+for Fig. 3 in the main manuscript, where the switching of colloids in
+blue phase III is studied.
 
@@
 reduced temperature $\tau = -0.25$). The cholesteric pitch is set
-to be $p = 64$, providing a slight better resolution than the bulk
+to be $p = 64$, providing a slightly better resolution than the bulk
 simulations.
@@
 \hline
-External field & 0.004& 3.086 & 0.02 & 32 & -0.25 & 2.6 & 0.23\\ 
+External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
 \hline
@@
 \end{table}
+
+{\bf 4.4 Disclination rendering}
 
@@
 equal to 35 pN (or equivalently $K=$70~pN). That is,
-\begin{equation}
-\frac{a}{b^2} = \frac{27}{2 A_0\gamma} \sim 2-5 \times 10^{-6}
-{\rm J}^{-1}\, {\rm m}^{3}.
-\end{equation}
 WHAT ARE a/b???
@@
 simulations with that of a hypothetical experiment.
-Given the previous value for $A_0$, and a dielectric anisotropy of 10
-(or equivalently $\epsilon_a=10\epsilon_0$, with $\epsilon_0$ the
+Given the previous value for $A_0$, and a dielectric anisotropy of 20
+(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
 dielectric permittivity of vacuum), we then obtain that an electric field
-of 5~V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.
+of 100 V/$\mu$m corresponds to a value of ${\cal E}\sim 0.4$ (considering
+a value of $\gamma=3.0857$ as in Fig. 3 of the main text.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[!h]
 \begin{center}
-\includegraphics[scale=0.8]{doubletwist.pdf}
+\includegraphics[scale=0.8]{support-fig1.png}
 \end{center}
@@
 
-\vfill\pagebreak
+\newpage
 
-\begin{figure}[h]
+%\vfill\pagebreak
+
+\begin{figure}[!h]
 \begin{center}
@@
 
-\begin{figure}[t]
+\newpage
+
+\begin{figure}[!h]
 \begin{center}
@@
 
-\begin{figure}[t]
+\newpage
+
+\begin{figure}[!h]
 \begin{center}
-\includegraphics[scale=0.42]{support-fig4.pdf}
+\includegraphics[scale=0.49]{support-fig4.png}
+\end{center}
+\caption{Plot of the free energy (per BP unit cell) versus time
+for the simulations corresponding to Fig. 1A-H in the main text
+(the labels correspond to those in Fig. 1).
+}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig5.pdf}
 \end{center}
@@
 
-\begin{figure}[t]
+\newpage
+
+\begin{figure}[!h]
 \begin{center}
-\includegraphics[scale=0.42]{support-fig5.pdf}
+\includegraphics[scale=0.42]{support-fig6.pdf}
 \end{center}
@@
 
-\begin{figure}[t]
+\newpage
+
+\if{\begin{figure}[!h]
 \begin{center}
@@
 \end{figure}
+}\fi
 
-%{\bf Captions for Supplementary Movies}
+%\newpage
+
+{\bf Captions for Supplementary Movies} \\
+
+{\bf Supplementary Movie 1:} 
+Final state of the end of the simulation in Fig.~1A. The movie
+shows a rotation of the configuration to better appreciate details.\\
+
+{\bf Supplementary Movie 2:} 
+Final state of the end of the simulation in Fig.~1B. The movie
+shows a rotation of the configuration to better appreciate details.\\
+
+{\bf Supplementary Movie 3:} 
+Final state of the end of the simulation in Fig.~1C. The movie
+shows a rotation of the configuration to better appreciate details. \\
+
+{\bf Supplementary Movie 4:} 
+Final state of the end of the simulation in Fig.~1D. The movie
+shows a rotation of the configuration to better appreciate details. \\
+
+{\bf Supplementary Movie 5:} 
+Final state at the end of the initial relaxation phase in Fig.~3. The movie
+shows a rotation of the configuration to better appreciate details.
+The distribution of the colloidal particles is isotropic with the particles forming clusters and having a tendency to occupy positions a the intersection of the BPIII disclination lines.\\
+
+{\bf Supplementary Movie 6:} 
+Configuration in a homogeneous electric field with strength
+corresponding to ${\cal E}=0.8$. The movie
+shows a rotation of the configuration to better appreciate details.
+A hexagonal pattern emerges where the particles form chains
+situated at the vertices of the honeycomb-like structure. \\
 

### 2013-10-02 trunk/papers/2012-sandwich/support.tex
Updated supporting text and figures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2164 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-In this work, the chirality and reduced temperature are always set to lie
-within the blue phase I region of the equilibrium phase diagram, except
-for Fig. 3 in the main manuscript, where the switching of colloids in
-blue phase III is studied.
+The effect of an electric field can be modelled by including the following
+term in the bulk free energy density:
+\begin{equation}
+-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
+\end{equation} 
+where $E_{\alpha}$ are the components of the electric field, and 
+$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
+crystal. The strength of the electric field is quantified via
+one further dimensionless number
+\begin{equation}
+{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
+\end{equation}
+The quantity ${\cal E}$ will be referred to as the reduced field strength.
 
-In addition to the fluid free density density $f(Q_{\alpha\beta})$, a surface
+{\bf 1.1 Surface free energy}
+
+In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
 free energy density is also present to represent the energetic cost of
@@
 fluid LC ordering.
-
-The effect of an electric field can be modelled by including the following
-term in the bulk free energy density:
-\begin{equation}
--\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
-\end{equation} 
-where $E_{\alpha}$ are the components of the electric field, and 
-$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
-crystal. The strength of the electric field is quantified via
-a further dimensionless number, 
-\begin{equation}
-{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
-\end{equation}
 
@@
 \rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
-=\partial_\beta P_{\alpha\beta}
+=\partial_\beta P_{\alpha\beta}.
 \end{equation}
@@
 \cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
-are set out in the following.
+are set out in \cite{juho}.
 
@@
 
-The following protocol was used for the simulations of BPIII with
-colloidal particles: First, we started from a configuration of
-randomly oriented and positioned double twist cylinders from which an
-amorphous BPIII network emerged. The structure
-was equilibrated for $6\times10^{5}$ LB time steps until the relaxation came
-virtually to a standstill. Then the colloidal particles featuring week
-normal anchoring were randomly inserted and the composite system was
-equilibrated for another $1.2\times10^{6}$ time steps.
-At the end of the equilibration phase the homogenous electric field
-with reduced field strength $e=\sqrt{\frac{27\epsilon_a}{32\pi A_0\gamma}} E=0.8$ 
-was switched on along the z-direction for $1.2\times10^{6}$ time steps.
-This was found to be sufficiently long for the hexagonal structure to emerge and saturate.
-After the electric field has been switched off the system quickly relaxed
-to a metastable state which showed a residual anisotropy in the
-colloid distribution. The total length of this final relaxation phase
-was $1\times10^6$ LB time steps.
+The following protocol is used for the simulations of BPIII with
+colloidal particles. An initial configuration of
+randomly oriented and positioned double twist cylinders is used,
+from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
+is equilibrated for $6\times10^{5}$ LB time steps until no significant
+further evolution is observed. The colloidal particles, with weak
+normal anchoring, are then inserted with random chosen positions,
+and the composite system is equilibrated for another $1.2\times10^{6}$
+time steps.
+At the end of the equilibration phase the uniform electric field
+with reduced field strength ${\cal E} = 0.8$ 
+is switched on (oriented along the $z$-direction), and the simulation
+run for a further $1.2\times10^{6}$ time steps.
+This is found to be sufficiently long for a hexagonal structure to
+emerge and saturate.
+If the electric field is then switched off, and the system quickly relaxes
+to a metastable state which shows a residual anisotropy in the
+colloidal distribution. The total length of this final relaxation phase
+is $1\times10^6$ LB time steps.
 
@@
 
-To get from simulation to physical units, we need to calibrate scales of
-length, energy and time in the simulations. For similar mappings see,
+To get from simulation to physical units, a calibration of scales of
+length, energy and time in the simulations is required. For similar
+mappings see,
 e.g.~\cite{denniston2}.
 
-The length scale can be set by mapping our half pitch to a realistic blue
-phase unit cell size, which are in the 100-500 nm range \cite{blue1}. 
+The length scale can be set by mapping the half pitch to a realistic blue
+phase unit cell size, which is in the 100--500~nm range \cite{blue1}. 
 For bulk simulations (Fig.~1 and Fig.~3 in the main text), a suitable choice
@@
 
-To obtain an energy scale, we choose $A_0 \simeq 10^6$~Pa, which is 
-reasonable from Ref.~\cite{blue1} (this is also the choice in
+To obtain an energy scale, it is possible to choose $A_0 \simeq 10^6$~Pa,
+which is reasonable following Ref.~\cite{blue1} (this is also the choice in
 Ref.~\cite{oliver2}). This choice leads to a simulation unit of
@@
 a liquid crystal with splay, bend and twist (Frank) elastic constants
-equal to 35 pN (or equivalently $K=$70~pN). That is,
-WHAT ARE a/b???
+equal to 35 pN (or equivalently $K=70$~pN).
 
-For the timescale calibration we use the formula (see e.g.~\cite{denniston})
+The timescale calibration may be obtained from (see e.g.~\cite{denniston})
 \begin{equation}
-\gamma_1=\frac{2q^2}{\Gamma}
+\gamma_1=\frac{2q^2}{\Gamma},
 \end{equation}
-which relates the rotational viscosity $\gamma_1$, to the ordering strength
+which relates the rotational viscosity $\gamma_1$ to the ordering strength
 $q$ and the order parameter mobility $\Gamma$.
-In our simulations, we choose $\Gamma = 0.5$, while the value of $\gamma$
-used in Fig.~1 leads to $q=1/2$.   
+In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
+for the simulation in Fig.~1 leads to $q=1/2$.   
 For real liquid crystalline materials, $\gamma_1$ usually lies in range
-$10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness
+$10^{-2}-1$ Pa s~\cite{deGennes}; for definiteness say
 $\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
@@
 To calibrate the electric field strength 
-we equate the value of the dimensionless parameter ${\cal E}$ in our 
+one can equate the value of the dimensionless parameter ${\cal E}$ in the 
 simulations with that of a hypothetical experiment.
@@
 (or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
-dielectric permittivity of vacuum), we then obtain that an electric field
-of 100 V/$\mu$m corresponds to a value of ${\cal E}\sim 0.4$ (considering
-a value of $\gamma=3.0857$ as in Fig. 3 of the main text.
+dielectric permittivity of vacuum), an electric field
+of 100 V/$\mu$m is obtained. This corresponds to a value of
+${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
+Fig.~3 of the main text).
 
@@
 in our model by the coupling to the order parameter.
-We note that the density is set to unity in our simulations: this
+Note that the density is set to unity in the simulations: this
 corresponds to a fluid density which is about a thousand times larger
-than in experiments. This however makes no difference in practice as
+than in experiments. This makes no difference in practice as
 the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
@@
 
-To summarize, our simulations represent BP-forming materials, with the
+To summarize, the simulations represent BP-forming materials, with the
 interpretation of the simulation units for length, time, and energy
@@
 %ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
-
-%ISSUE: do we want some free energy numbers?
 
@@
 
+\bibitem{juho}
+J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
+\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).
+
 \bibitem{denniston}
@@
 \textit{Europhys. Lett.} \textbf{72}, 403 (2005).
+
+\bibitem{oliver-bp3}
+O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
 
@@
 \newpage
-
-%\vfill\pagebreak
 
@@
 
-\if{\begin{figure}[!h]
+\begin{figure}[!h]
 \begin{center}
-%%\includegraphics[width=0.95\columnwidth]{s5.jpg}
+\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
 \end{center}
-\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
-(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
-(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
+\caption{Panels A and B as for Fig. 3, but with the disclination lines
+removed for clarity. Panel C shows the situation when the field is
+switched off: a residual anisotropic structure remains which reflects the
+structure induced by the electric field. The electric field direction
+when switched on (panel B) is into the plane of the paper.}
 \end{figure}
-}\fi
 
@@
 {\bf Supplementary Movie 1:} 
-Final state of the end of the simulation in Fig.~1A. The movie
-shows a rotation of the configuration to better appreciate details.\\
+The final state of the simulation as shown in Fig.~1A. The movie
+rotates the configuration to reveal more clearly its structure.\\
 
 {\bf Supplementary Movie 2:} 
-Final state of the end of the simulation in Fig.~1B. The movie
-shows a rotation of the configuration to better appreciate details.\\
+The final state of the simulation as shown in Fig.~1B. The movie
+rotates the configuration to reveal more clearly its structure.\\
 
 {\bf Supplementary Movie 3:} 
-Final state of the end of the simulation in Fig.~1C. The movie
-shows a rotation of the configuration to better appreciate details. \\
+The final state of the simulation as shown in Fig.~1C. The movie
+rotates the configuration to reveal more clearly its structure.\\
 
 {\bf Supplementary Movie 4:} 
-Final state of the end of the simulation in Fig.~1D. The movie
-shows a rotation of the configuration to better appreciate details. \\
+The final state of the simulation as shown in Fig.~1D. The movie
+rotates the configuration to reveal more clearly its structure.\\
 
 {\bf Supplementary Movie 5:} 
-Final state at the end of the initial relaxation phase in Fig.~3. The movie
-shows a rotation of the configuration to better appreciate details.
-The distribution of the colloidal particles is isotropic with the particles forming clusters and having a tendency to occupy positions a the intersection of the BPIII disclination lines.\\
+The configuration after the initial relaxation phase in Fig.~3. The movie
+shows a rotation of the configuration. The distribution of the colloidal
+particles is isotropic, with the particles forming clusters and having a
+tendency to occupy positions at the intersection of the BPIII disclination
+lines.\\
 
@@
 corresponding to ${\cal E}=0.8$. The movie
-shows a rotation of the configuration to better appreciate details.
+shows a rotation of the configuration.
 A hexagonal pattern emerges where the particles form chains

### 2013-10-02 papers/2012-sandwich/CholColloids.tex
Minor improvements to text
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2165 40876441-3496-0410-a613-a2fca35b2ae2

@@
 Remarkably, we find that many of these new materials are metastable: two or more structures can arise under identical thermodynamic conditions.
-The observed structure depends not only on the formulation protocol, but also on the field history.
+The observed structure depends not only on the formulation protocol,
+but also on the history of an applied electric field.  
 This new class of soft materials should thus be relevant to the design of switchable, multistable devices for optical technologies including, for example, e-paper~\cite{epaper}.}
@@
 they disrupt the orientational order in the fluid and create defects (disclination lines) in the nematic close to their surfaces.
-To minimize their free energy cost, it is generally advantageous for defects to be shared:
+To minimize the free energy cost, it is generally advantageous for defects to be shared:
 therefore the colloidal particles have a generic tendency to aggregate. Such aggregates may further 
@@
 as new-generation display devices as they can exploit the bistability
-inherent in cholesteric devices~\cite{colloiddevice,choldevice}.
+inherent in liquid crystal devices~\cite{colloiddevice,choldevice}.
 
@@
 In the colloid-BP system, $w$ also plays a determining role in the final composite structure.
-When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes this to become covered by particles, with little disruption of its long-range order.
+When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes
+ the lattice to become covered by particles, with little disruption of its
+ long-range order.
 Thus a templated colloidal crystal is created, as first reported in~\cite{miha}.
@@
 We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supporting Information).
-The resulting colloid-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $2\%$ respectively.
+The resulting colloid-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
 There are, however, significant effects of the confining walls.
@@
 Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit.
-These are similarly to those reported experimentally for large colloids in a cholesteric (but non-BP) host~\cite{niek}.
+These clusters
+are similar to those reported experimentally for large colloids in a
+cholesteric (but non-BP) host~\cite{niek}.
 
@@
 direction.
-The resulting multistability is indeed a must-have feature for new generation 
+The resulting multistability is indeed a prerequisite for new generation 
 switchable electrooptic devices for use in energy-saving electronics-free

### 2013-10-03 trunk/papers/2012-sandwich/CholColloids.tex
updated text
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2168 40876441-3496-0410-a613-a2fca35b2ae2

@@
 they offer exciting prospects for applications as biosensors~\cite{abbott}, or
-as new-generation display devices as they can exploit the bistability
-inherent in cholesteric devices~\cite{colloiddevice,choldevice}.
+as new-generation devices~\cite{colloiddevice,electrophoresis}.
 
@@
 We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles within it.
-Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.10$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface).
+Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface).
 For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
@@
 In the colloid-BP system, $w$ also plays a determining role in the final composite structure.
-When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes this to become covered by particles, with little disruption of its long-range order.
+When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes
+ the lattice to become covered by particles, with little disruption of its
+ long-range order.
 Thus a templated colloidal crystal is created, as first reported in~\cite{miha}.
-Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
+Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to very slow dynamics (Fig.
+S4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
 1A).
@@
 At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloid strands (the gel in Fig.~1H) is again seen.
-The structures in Fig. 1A-H have been followed for about 20 ms 
-(see Supporting Information for the mapping between simulation and physical 
-time); they correspond to deep metastable minima (elastic interactions are of
-order $KR$ and dwarf thermal motion~\cite{stark,lavrentovich}).
+The structures in Fig. 1A-H have been followed for $\sim$20 ms 
+(see Fig. S4); they correspond to either very slowly evolving states or 
+deep metastable minima (elastic interactions are of
+order $KR$ and dwarf thermal motion, see
+Fig. S4 and Ref.~\cite{stark,lavrentovich}).
 
 In display devices and other optical applications, it is typical for liquid crystals to be strongly confined (and subject to strong boundary effects), in contrast to the bulk geometries considered above.
-We therefore next describe how colloidal suspensions self assemble in thin sandwiches (thickness $h = 3.5 \lambda$), whose confining walls can favour either tangential (planar degenerate) or normal (homeotropic) orientation of the director field.
+We therefore next describe how colloidal suspensions self assemble in thin sandwiches (thickness $h = 1.75 \lambda$), whose confining walls can favour either tangential (planar degenerate) or normal (homeotropic) orientation of the director field.
 For simplicity we here impose one or the other as a strict boundary condition.
@@
 R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
-\bibitem{choldevice} D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
-Bistable cholesteric reflective displays: Materials and Drive Schemes,
-{\it Annu. Rev. Mater. Sci.} {\bf 27}, 117-146 (1997).
+\bibitem{electrophoresis}  O. D. Lavrentovich,	I. Lazo, O. P. Pishnyak.
+Nonlinear electrophoresis of dielectric and metal spheres in a nematic liquid 
+crystal. {\it Nature} {\bf 467}, 947-950 (2010).
+%D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
+%Bistable cholesteric reflective displays: Materials and Drive Schemes,
+%{\it Annu. Rev. Mater. Sci.} {\bf 27}, 117-146 (1997).
 %\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}

### 2013-10-03 papers/2012-sandwich/CholColloids.tex
Changed references and electric field part and captionCholColloids.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2170 40876441-3496-0410-a613-a2fca35b2ae2

@@
 they offer exciting prospects for applications as biosensors~\cite{abbott}, or
-as new-generation display devices as they can exploit the bistability
-inherent in liquid crystal devices~\cite{colloiddevice,choldevice}.
+as new-generation devices~\cite{colloiddevice,electrophoresis}.
 
@@
 However, recent advances in formulation have widened this stability range enormously (up to 250K) paving the way to
-applications of BPs in display devices~\cite{kikuchi,coleswidetrange,bpdevice}.
+applications of BPs in display devices~\cite{kikuchi,bplasers,coleswidetrange,bpdevice,coles}.
 
@@
 To explore such metastability, and switching strategies between states, we require multi-unit-cell, time-dependent simulations.
-These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{domaingrowth,bp3,henrichfield}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}.
+These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}.
 Our methodology and simulation parameters are summarized in the Supporting Information.
@@
 Thus a templated colloidal crystal is created, as first reported in~\cite{miha}.
-Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
+Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to very slow dynamics (Fig.
+S4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
 1A).
@@
 these structures can be modified, or switched, with electric fields.
-Fig.~3 addresses this question, starting again from an amorphous blue phase,
+Fig.~3 addresses this question, starting from amorphous Blue Phase III,
 this time at a higher chirality. Dispersing colloids in this networks
@@
 %This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
-We have not explored in this work their switchability by flow, but this could again be inherited from the underlying BP dynamics \cite{permeation2,adriano}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
+We have not explored in this work their switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
 
@@
 {\it J. Soc. Inf. Display} {\bf 19}, 129-156 (2011).
+\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic.  Entangled nematic colloidal dimers and wires. 
+{\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
 \bibitem{zumer} I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik,
 S. Zumer. Two-dimensional nematic colloidal crystals self-assembled by topological defects. {\it Science} {\bf 313}, 954-958 (2006).
+\bibitem{tanaka} T.
+Araki, H.
+Tanaka.
+Colloidal aggregation in a nematic liquid crystal: Topological arrest of particles by a single-stroke disclination line.
+{\it Phys.
+Rev.
+Lett.} {\bf 97}, 127801 (2006).
 \bibitem{tiffany} T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo,
@@
 R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
-\bibitem{choldevice} D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
-Bistable cholesteric reflective displays: Materials and Drive Schemes,
-{\it Annu. Rev. Mater. Sci.} {\bf 27}, 117-146 (1997).
+\bibitem{electrophoresis}  O. D. Lavrentovich,	I. Lazo, O. P. Pishnyak.
+Nonlinear electrophoresis of dielectric and metal spheres in a nematic liquid 
+crystal. {\it Nature} {\bf 467}, 947-950 (2010).
+%D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
+%Bistable cholesteric reflective displays: Materials and Drive Schemes,
+%{\it Annu. Rev. Mater. Sci.} {\bf 27}, 117-146 (1997).
 %\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}
 %{\bf 35}, 4419 (1987).
+\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo.
+ Structure of Blue Phase III of Cholesteric Liquid Crystals.
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
 \bibitem{kikuchi} H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama.
@@
 {\it Nat. Mat.} {\bf 1}, 64-68 (2002).
+\bibitem{bplasers} W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri.
+Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
+phase II. {\it Nat. Mat.} {\bf 1}, 111-113 (2002).
 \bibitem{coleswidetrange} H.~J. Coles, M.~N. Pivnenko.
@@
 {\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
-\bibitem{bplasers} W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri.
-Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
-phase II. {\it Nat. Mat.} {\bf 1}, 111-113 (2002).
+\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles.
+Blue-phase templated fabrication of three-dimensional nanostructures for photonic applications. {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
+\bibitem{adriano} A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini,
+F. Salvadore. Bistable Defect Structures In Blue Phase Devices. 
+ {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
+\bibitem{henrichfield}  O. Henrich, D. Marenduzzo, K. Stratford, 
+M. E. Cates. Thermodynamics of blue phases in electric fields.
+{\it Phys. Rev. E} {\bf 81}, 031706 (2010).
+\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates.
+ Ordering dynamics of blue phases entails kinetic stabilization of amorphous networks. {\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
+\bibitem{juho1} J. S. Lintuvuori, K. Stratford, M. E. Cates,
+D. Marenduzzo.   Colloids in Cholesterics: Size-Dependent Defects and Non-Stokesian Microrheology. {\it Phys. Rev. Lett.} {\bf 105}, 178302 (2010).
+\bibitem{juho2}  J. S. Lintuvuori, K. Stratford, M. E. Cates,
+D. Marenduzzo. 
+Self-Assembly and Nonlinear Dynamics of Dimeric Colloidal Rotors in 
+Cholesterics. {\it Phys. Rev. Lett.} {\bf 107}, 267802 (2011).
 \bibitem{permeation1} 	W. Helfrich. Capillary flow of cholesteric
@@
 D.~A. Weitz.Particle-stabilized defect gel in cholesteric liquid crystals.  {\it Science} {\bf 283}, 209-212 (1999).
-\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles.
-Blue-phase templated fabrication of three-dimensional nanostructures for photonic applications. {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
-\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates.
- Ordering dynamics of blue phases entails kinetic stabilization of amorphous networks. {\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
-\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo.
- Structure of Blue Phase III of Cholesteric Liquid Crystals.
-{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
-\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic.  Entangled nematic colloidal dimers and wires. 
-{\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
-\bibitem{tanaka} T.
-Araki, H.
-Tanaka.
-Colloidal aggregation in a nematic liquid crystal: Topological arrest of particles by a single-stroke disclination line.
-{\it Phys.
-Rev.
-Lett.} {\bf 97}, 127801 (2006).
-\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
-Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
-\bibitem{adriano} A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini,
-F. Salvadore. Bistable Defect Structures In Blue Phase Devices. 
- {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
-\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
-Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
-{\it Nat. Comm.} {\bf 2}, 246 (2011).
+%\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
+%Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
+%\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
+%Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
+%{\it Nat. Comm.} {\bf 2}, 246 (2011).
 \bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
@@
  Liquid Crystal. {\it Langmuir} {\bf 26}, 13502-13510 (2010).
-\bibitem{henrichfield}  O. Henrich, D. Marenduzzo, K. Stratford, 
-M. E. Cates. Thermodynamics of blue phases in electric fields.
-{\it Phys. Rev. E} {\bf 81}, 031706 (2010).
-\bibitem{juho1} J. S. Lintuvuori, K. Stratford, M. E. Cates,
-D. Marenduzzo.   Colloids in Cholesterics: Size-Dependent Defects and Non-Stokesian Microrheology. {\it Phys. Rev. Lett.} {\bf 105}, 178302 (2010).
-\bibitem{juho2}  J. S. Lintuvuori, K. Stratford, M. E. Cates,
-D. Marenduzzo. 
-Self-Assembly and Nonlinear Dynamics of Dimeric Colloidal Rotors in 
-Cholesterics. {\it Phys. Rev. Lett.} {\bf 107}, 267802 (2011).
-%\bibitem{flowswitch} Davide et al on flow reorganization of BPs
+\bibitem{flowswitch} O. Henrich, K. Stratford, P.V. Coveney, M.E. Cates, D. Marenduzzo,
+Rheology of cubic blue phases. {\it Soft Matter} (2013), DOI: 10.1039/C3SM50228G 
 \bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang, Developing optofluidic technology through the fusion of microfluidics and optics {\it Nature} {\bf 442}, 381-386 (2006).
@@
 \centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase, then quenched into a regime where BPIII is stable in the bulk (A), and finally subjected to an electric field along the vertical direction (B).
+\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A), and finally subjected to an electric field along the vertical direction (B).
 The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.

### 2013-10-03 trunk/papers/2012-sandwich/CholColloids.tex
Corrected packing fraction in caption of Fig.3
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2171 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %{\bf 35}, 4419 (1987).
+\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo.
+ Structure of Blue Phase III of Cholesteric Liquid Crystals.
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
 \bibitem{kikuchi} H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama.
@@
 \centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase, then quenched into a regime where BPIII is stable in the bulk (A), and finally subjected to an electric field along the vertical direction (B).
+\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A), and finally subjected to an electric field along the vertical direction (B).
 The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.
-Structures correspond to $w=0.23$ and $\phi=1\%$ (see Supporting Information
+Structures correspond to $w=0.23$ and $\phi=5\%$ (see Supporting Information
 for full list of parameters). }

### 2013-10-03 papers/2012-sandwich/support.tex
Updated supplement
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2172 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-In this work, the chirality and reduced temperature are always set to lie
-within the blue phase I region of the equilibrium phase diagram. EXCEPT?
+The effect of an electric field can be modelled by including the following
+term in the bulk free energy density:
+\begin{equation}
+-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
+\end{equation} 
+where $E_{\alpha}$ are the components of the electric field, and 
+$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
+crystal. The strength of the electric field is quantified via
+one further dimensionless number
+\begin{equation}
+{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
+\end{equation}
+The quantity ${\cal E}$ will be referred to as the reduced field strength.
 
-In addition to the fluid free density density $f(Q_{\alpha\beta})$, a surface
+{\bf 1.1 Surface free energy}
+
+In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
 free energy density is also present to represent the energetic cost of
@@
 The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
-discussed below. For a colloid of radius $a$, the strength of the surface
+discussed below. For a colloid of radius $R$, the strength of the surface
 anchoring compared which the bulk fluid elastic constant may be quantified
-by the dimensionless parameter $Wa/K$. For small values of this parameter,
+by the dimensionless parameter $WR/K$. For small values of this parameter,
 the presence of a colloid surface should have little impact on the local
 fluid LC ordering.
-
-The effect of an electric field can be modelled by including the following
-term in the bulk free energy density:
-\begin{equation}
--\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
-\end{equation} 
-where $E_{\alpha}$ are the components of the electric field, and 
-$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
-crystal. The strength of the electric field is quantified via
-a further dimensionless number, 
-\begin{equation}
-{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
-\end{equation}
 
@@
 
-In all simulations the colloidal (hard sphere) radius is $a = 2.3$. To
+In all simulations the colloidal (hard sphere) radius is $R = 2.3$. To
 prevent colloids overlapping at the hard sphere radius, an additional
@@
 Simulations at different solid volume fractions (1\% and 4\%), and different 
-surface anchoring strengths ($Wa/K = 0.23$ and $Wa/K = 2.3$) representing
+surface anchoring strengths ($WR/K = 0.23$ and $WR/K = 2.3$) representing
 ``weak'' and ``strong'' anchoring are run for two million simulation
@@
 pitch $p = 32\sqrt{2}$. The chirality and reduced temperature (see
-Table~\ref{table:params}) remain appropriate for blue phase I.
+Table~\ref{table:params}) remain appropriate for BPI.
 Colloids are added as before and the simulations run for 2 million
 simulation steps.
+
+For simplicity, we do not allow the size of the BPI unit cell to 
+dynamically readjust to minimize free energy (there is no 
+``redshift''~\cite{blue1}): this would lead to quantitatively different 
+values of the free energy, but qualitatively similar structures.
 
@@
 The surface anchoring for colloids is as before: always normal, but with
-$Wa/K = 0.23$ and $Wa/K = 2.3$. The colloid solid volume fractions are
+$WR/K = 0.23$ and $WR/K = 2.3$. The colloid solid volume fractions are
 1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
@@
 
-OLIVER PLEASE ADD short description of protocol here. AND the
-parameters in the table.
+For the simulations of BPIII with colloidal particles a system size of
+$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
+The simulations are
+performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
+An initial configuration of
+randomly oriented and positioned double twist cylinders is used,
+from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
+is equilibrated for $6\times10^{5}$ LB time steps until no significant
+further evolution is observed.
 
-\begin{table}
+The colloidal particles with weak
+normal anchoring ($WR / K= 0.23$), are then inserted with randomly chosen positions,
+and the composite system is equilibrated for another $1.2\times10^{6}$
+time steps.
+At the end of the equilibration phase the uniform electric field
+with reduced field strength ${\cal E} = 0.8$ 
+is switched on (oriented along the $z$-direction), and the simulation
+run for a further $1.2\times10^{6}$ time steps.
+This is found to be sufficiently long for a hexagonal structure to
+emerge and saturate.
+If the electric field is then switched off, and the system quickly relaxes
+to a metastable state which shows a residual anisotropy in the
+colloidal distribution. The total length of this final relaxation phase
+is $1\times10^6$ LB time steps.
+
+\begin{table}[h]
 \begin{center}
@@
 \hline
-         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $Wa/K$\\
+         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
 \hline 
@@
 \hline
-External field & a0& gamma & K & p & kappa & tau & Wa/K\\ 
+External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
 \hline
@@
 
+{\bf 4.4 Disclination rendering}
+
 In all cases visualisation of the defect lines is carried out via
@@
 a liquid crystal with splay, bend and twist (Frank) elastic constants
-equal to 35 pN (or equivalently $K=$70~pN). That is,
+equal to 35 pN (or equivalently $K=70$~pN).
+
+The timescale calibration may be obtained from (see e.g.~\cite{denniston})
 \begin{equation}
-\frac{a}{b^2} = \frac{27}{2 A_0\gamma} \sim 2-5 \times 10^{-6}
-{\rm J}^{-1}\, {\rm m}^{3}.
+\gamma_1=\frac{2q^2}{\Gamma},
 \end{equation}
-WHAT ARE a/b???
-
-For the timescale calibration we use the formula (see e.g.~\cite{denniston})
-\begin{equation}
-\gamma_1=\frac{2q^2}{\Gamma}
-\end{equation}
-which relates the rotational viscosity $\gamma_1$, to the ordering strength
+which relates the rotational viscosity $\gamma_1$ to the ordering strength
 $q$ and the order parameter mobility $\Gamma$.
-In our simulations, we choose $\Gamma = 0.5$, while the value of $\gamma$
-used in Fig.~1 leads to $q=1/2$.   
+In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
+for the simulation in Fig.~1 leads to $q=1/2$.   
 For real liquid crystalline materials, $\gamma_1$ usually lies in range
-$10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness
+$10^{-2}-1$ Pa s~\cite{deGennes}; for definiteness say
 $\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
@@
 To calibrate the electric field strength 
-we equate the value of the dimensionless parameter ${\cal E}$ in our 
+one can equate the value of the dimensionless parameter ${\cal E}$ in the 
 simulations with that of a hypothetical experiment.
-Given the previous value for $A_0$, and a dielectric anisotropy of 10
-(or equivalently $\epsilon_a=10\epsilon_0$, with $\epsilon_0$ the
-dielectric permittivity of vacuum), we then obtain that an electric field
-of 5~V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.
+Given the previous value for $A_0$, and a dielectric anisotropy of 20
+(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
+dielectric permittivity of vacuum), an electric field
+of 100 V/$\mu$m is obtained. This corresponds to a value of
+${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
+Fig.~3 of the main text).
 
@@
 in our model by the coupling to the order parameter.
-We note that the density is set to unity in our simulations: this
+Note that the density is set to unity in the simulations: this
 corresponds to a fluid density which is about a thousand times larger
-than in experiments. This however makes no difference in practice as
+than in experiments. This makes no difference in practice as
 the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
-velocity, in our simulations around $10^{-5}$ in simulation units) 
+velocity, in our simulations around $10^{-5}$ in lattice units, 
+and $\Lambda$ is a typical length scale, e.g. the size of a unit cell) 
 remains small enough~\cite{codef}.
 
-To summarize, our simulations represent BP-forming materials, with the
+To summarize, the simulations represent BP-forming materials, with the
 interpretation of the simulation units for length, time, and energy
@@
 {\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).
-
+\bibitem{deGennes}
+P.-G. de Gennes, J. Prost,
+{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).
 %\bibitem{marendu1}
@@
 %{\it Phys. Rev. E} {\bf 76}, 031921 (2007).
-
 %\bibitem{marendu2}
@@
 %{\it Comput. Math. Appl.}, doi:10.1016/j.camwa.2009.08.047[dx.doi.org] (2009).
-
 %\bibitem{marendu3}
@@
 %{\it Soft Matter} {\bf 5}, 3791-3800 (2009).
-
 \bibitem{oliver1}
@@
 \textit{Phys. Rev. E}, \textbf{81} 031706 (2010).
-
 \bibitem{beris}
@@
 (Oxford Univ. Press, New York, 1994).
-
-\bibitem{deGennes}
-P.-G. de Gennes, J. Prost,
-{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).
-
 \bibitem{ladd94}
@@
 311 (1994).
-
 \bibitem{nguyen2002}
@@
 \textit{Phys. Rev. E} \textbf{66} 046708 (2002).
-
+\bibitem{juho}
+J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
+\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).
 \bibitem{denniston}
@@
 \textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
-
+\bibitem{fournier2005}
+J.-B. Fournier and P. Galatola,
+\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
+\bibitem{oliver-bp3}
+O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
 \bibitem{denniston2}
@@
 \textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
-
-\bibitem{fournier2005}
-J.-B. Fournier and P. Galatola,
-\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
-
 \bibitem{oliver2}
@@
 \textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
-
 \bibitem{codef} M. E. Cates,  et al., 
@@
 
-\vfill\pagebreak
+\newpage
 
-\begin{figure}[h]
+\begin{figure}[!h]
 \begin{center}
@@
 
-\begin{figure}[t]
+\newpage
+
+\begin{figure}[!h]
 \begin{center}
-%%\includegraphics[width=0.95\columnwidth]{s5.jpg}
+\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
 \end{center}
-\caption{Pair correlation function for weak ($WR/K=0.23$) and strong ($WR/K=2.3$) anchoring cases considered in the manuscript.
-(A,B) BPI initialisation with A) volume fraction $\phi=1${\%} and B) $\phi=5${\%}.
-(C,D) ``quenched system'' with C) $\phi=1${\%} and D) $\phi=5${\%}.}
+\caption{Panels A and B as for Fig. 3, but with the disclination lines
+removed for clarity. Panel C shows the situation when the field is
+switched off: a residual anisotropic structure remains which reflects the
+structure induced by the electric field. The electric field direction
+when switched on (panel B) is into the plane of the paper.}
 \end{figure}
 
-%{\bf Captions for Supplementary Movies}
+%\newpage
+
+{\bf Captions for Supplementary Movies} \\
+
+{\bf Supplementary Movie 1:} 
+The final state of the simulation as shown in Fig.~1A. The movie
+rotates the configuration to reveal more clearly its structure.\\
+
+{\bf Supplementary Movie 2:} 
+The final state of the simulation as shown in Fig.~1B. The movie
+rotates the configuration to reveal more clearly its structure.\\
+
+{\bf Supplementary Movie 3:} 
+The final state of the simulation as shown in Fig.~1C. The movie
+rotates the configuration to reveal more clearly its structure.\\
+
+{\bf Supplementary Movie 4:} 
+The final state of the simulation as shown in Fig.~1D. The movie
+rotates the configuration to reveal more clearly its structure.\\
+
+{\bf Supplementary Movie 5:} 
+The configuration after the initial relaxation phase in Fig.~3. The movie
+shows a rotation of the configuration. The distribution of the colloidal
+particles is isotropic, with the particles forming clusters and having a
+tendency to occupy positions at the intersection of the BPIII disclination
+lines.\\
+
+{\bf Supplementary Movie 6:} 
+Configuration in a homogeneous electric field with strength
+corresponding to ${\cal E}=0.8$. The movie
+shows a rotation of the configuration.
+A hexagonal pattern emerges where the particles form chains
+situated at the vertices of the honeycomb-like structure. \\
 

### 2013-10-03 papers/2012-sandwich/CholColloids.tex
updated text
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2173 40876441-3496-0410-a613-a2fca35b2ae2

@@
 they offer exciting prospects for applications as biosensors~\cite{abbott}, or
-as new-generation devices~\cite{colloiddevice,electrophoresis}.
+as new-generation devices~\cite{colloiddevice,tanakanatmat}.
 
@@
 A second compelling reason to study colloid-BP composites is that, even without particles, BPs can show several competing metastable free energy minima,
-corresponding to different topologies of the defect network~\cite{adriano}, with a strong dependence on applied electric or magnetic fields~\cite{henrichfield}.
+corresponding to different topologies of the defect network~\cite{adriano,fukuda}, with a strong dependence on applied electric or magnetic fields~\cite{henrichfield}.
 % 
@@
 In the colloid-BP system, $w$ also plays a determining role in the final composite structure.
-When $w$ is small (Fig.~ 1A,B), the attraction of particles to the disclination lattice causes
+When $w$ is small (Fig.~ 1A,B; Suppl.
+Movie 1,2), the attraction of particles to the disclination lattice causes
  the lattice to become covered by particles, with little disruption of its
@@
 
-For large $w$ (Fig.~1C,D), the picture changes dramatically.
+For large $w$ (Fig.~1C,D; Suppl. Movie 3,4), the picture changes dramatically.
 Each colloidal particle now disrupts significantly the order of the nearby fluid, with defect formation close to its surface~\cite{stark}.
@@
 results in the particle occupying the available nodes, where the order
-parameter is low (Fig.~3A). Applying a high enough field along the $z$ 
+parameter is low (Fig.~3A; Suppl. Movie 5). Applying a high enough field along the $z$ 
 direction leads to a switching to a new phase, where colloidal arrays fit 
-within a network with distorted hexagonal cells (Fig.~3B). Remarkably, upon 
+within a network with distorted hexagonal cells (Fig.~3B; Suppl. Movie 6). Remarkably, upon 
 field removal this honeycomb structure does not find its way back to the 
@@
 R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
-\bibitem{electrophoresis}  O. D. Lavrentovich,	I. Lazo, O. P. Pishnyak.
-Nonlinear electrophoresis of dielectric and metal spheres in a nematic liquid 
-crystal. {\it Nature} {\bf 467}, 947-950 (2010).
+\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
+Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
+%\bibitem{electrophoresis}  O. D. Lavrentovich,	I. Lazo, O. P. Pishnyak.
+%Nonlinear electrophoresis of dielectric and metal spheres in a nematic liquid 
+%crystal. {\it Nature} {\bf 467}, 947-950 (2010).
 %D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
@@
  {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
+\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
+Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
+{\it Nat. Comm.} {\bf 2}, 246 (2011).
 \bibitem{henrichfield}  O. Henrich, D. Marenduzzo, K. Stratford, 
@@
 D.~A. Weitz.Particle-stabilized defect gel in cholesteric liquid crystals.  {\it Science} {\bf 283}, 209-212 (1999).
-%\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
-%Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
 %\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
@@
 \bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang, Developing optofluidic technology through the fusion of microfluidics and optics {\it Nature} {\bf 442}, 381-386 (2006).
-
+%\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
+%Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
 \end{thebibliography}
@@
 The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.
-Structures correspond to $w=0.23$ and $\phi=1\%$ (see Supporting Information
+Structures correspond to $w=0.23$ and $\phi=5\%$ (see Supporting Information
 for full list of parameters). }

### 2013-10-14 papers/2012-sandwich/support.tex
Added E_x-field pics and structure factor
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2176 40876441-3496-0410-a613-a2fca35b2ae2

@@
 structure induced by the electric field. The electric field direction
-when switched on (panel B) is into the plane of the paper.}
+when switched on (panel B) is into the plane of the paper and along z.}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.24\columnwidth]{support-fig8a.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8b.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8c.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8d.png}
+\end{center}
+\caption{Particle distribution for switched electric field direction: 
+The field is now oriented along x (it was oriented along z in Fig.7 and Fig.3 main article). 
+The leftmost panel shows the configuration with the disclination lines removed 
+for clarity. The centre left picture shows the Fourier transform of the colloids for   
+wavevectors $k_y, k_z\in [-3\pi/8,3\pi/8]$. The centre right panels depicts the residual 
+anisotropy after switch off.}
 \end{figure}

### 2013-10-15 trunk/papers/2012-sandwich/support.tex
updated supporting material
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2167 40876441-3496-0410-a613-a2fca35b2ae2
updated supporting material
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2169 40876441-3496-0410-a613-a2fca35b2ae2
Added structure factor plots and defintion
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2178 40876441-3496-0410-a613-a2fca35b2ae2

@@
 The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
-discussed below. For a colloid of radius $a$, the strength of the surface
+discussed below. For a colloid of radius $R$, the strength of the surface
 anchoring compared which the bulk fluid elastic constant may be quantified
-by the dimensionless parameter $Wa/K$. For small values of this parameter,
+by the dimensionless parameter $WR/K$. For small values of this parameter,
 the presence of a colloid surface should have little impact on the local
@@
 
-The following protocol is used for the simulations of BPIII with
-colloidal particles. An initial configuration of
+For the simulations of BPIII with colloidal particles a system size of
+$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
+The simulations are
+performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
+An initial configuration of
 randomly oriented and positioned double twist cylinders is used,
@@
 is equilibrated for $6\times10^{5}$ LB time steps until no significant
-further evolution is observed. The colloidal particles, with weak
-normal anchoring, are then inserted with random chosen positions,
+further evolution is observed.
+
+The colloidal particles with weak
+normal anchoring ($WR / K= 0.23$), are then inserted with randomly chosen positions,
 and the composite system is equilibrated for another $1.2\times10^{6}$
@@
 
-\begin{table}
+\begin{table}[h]
 \begin{center}
@@
 unambiguously identifies the disclinations.
+
+{\bf 4.4 Computation of the structure factor}
+
+In order to demonstrate structural information about the field-aligned 
+states and the residual anisotropy after switch off we followed a 
+procedure similar to the one described in \cite{oliver-bp3}.
+The structure factor in the current approach is defined via
+the Fourier transform of the colloid density:
+
+\begin{eqnarray}
+S({\bf k})&=&|F({\bf k})|^2\\
+F({\bf k})&=&\int_V d{\bf r} \rho({\bf r}) e^{i {\bf k r}}
+\end{eqnarray}
+
+The colloid density $\rho({\bf r})$ is a discretised density and taken to be
+$\rho({\bf r})=1$ if a lattice site at ${\bf r}$ is part of a colloid
+and $\rho({\bf r})=0$ otherwise. This definition, which deviates from the 
+standard definition of a structure factor and does not consider the colloids 
+as pointlike objects, allows to gain the same structural information
+in a simpler way.
 
@@
 the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
-velocity, in our simulations around $10^{-5}$ in simulation units) 
+velocity, in our simulations around $10^{-5}$ in lattice units, 
+and $\Lambda$ is a typical length scale, e.g. the size of a unit cell) 
 remains small enough~\cite{codef}.
@@
 \textit{Phys. Rev. E} \textbf{66} 046708 (2002).
-
 \bibitem{juho}
@@
 \textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).
-
 \bibitem{denniston}
@@
 \textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
-
-\bibitem{fournier2005}
-J.-B. Fournier and P. Galatola,
-\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
-
-\bibitem{oliver-bp3}
-O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
-{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
-
 \bibitem{oliver2}
@@
 structure induced by the electric field. The electric field direction
-when switched on (panel B) is into the plane of the paper.}
+when switched on (panel B) is into the plane of the paper and along z.}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.24\columnwidth]{support-fig8a.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8b.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8c.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8d.png}
+\end{center}
+\caption{Particle distribution during and after switched electric field direction: 
+The above images show the ordering with the electric field along the x-direction 
+after a previous run that had the field along z. For clarity disclination lines have
+been removed. The left panels show the configuration and the structure factor 
+while the field is on, whereas the righ panels show the same quantities after 
+switch-off. The structure factor data shows wave vectors $k_y, k_z \in [-3\pi/8 , 3\pi/8]$.}
 \end{figure}

### 2013-10-18 papers/2012-sandwich/support.tex
Added results of run with E_y-field
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2198 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-10-03 papers/2012-sandwich/support.tex
@@
 
+{\bf 4.4 Computation of the structure factor}
+
+In order to demonstrate structural information about the field-aligned 
+states and the residual anisotropy after switch off we followed a 
+procedure similar to the one described in \cite{oliver-bp3}.
+The structure factor in the current approach is defined via
+the Fourier transform of the colloid density:
+
+\begin{eqnarray}
+S({\bf k})&=&|F({\bf k})|^2\\
+F({\bf k})&=&\int_V d{\bf r} \rho({\bf r}) e^{i {\bf k r}}
+\end{eqnarray}
+
+The colloid density $\rho({\bf r})$ is a discretised density and taken to be
+$\rho({\bf r})=1$ if a lattice site at ${\bf r}$ is part of a colloid
+and $\rho({\bf r})=0$ otherwise. This definition, which deviates from the 
+standard definition of a structure factor and does not consider the colloids 
+as pointlike objects, allows to gain the same structural information
+in a simpler way.
+
 {\bf 5. Parameter mapping to physical units}
@@
 \includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
+\includegraphics[width=0.5\columnwidth]{support-fig7d.png}
 \end{center}
@@
 structure induced by the electric field. The electric field direction
-when switched on (panel B) is into the plane of the paper.}
+when switched on (panel B) is into the plane of the paper and along z.}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.24\columnwidth]{support-fig8a.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8b.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8c.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8d.png}\\
+\includegraphics[width=0.24\columnwidth]{support-fig8e.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8f.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8g.png}
+\includegraphics[width=0.24\columnwidth]{support-fig8h.png}
+\end{center}
+\caption{Particle distribution and structure factor in electric fields and after switch-off:
+The top images show the ordering of the particles in an electric field along the x-direction 
+after a previous run which had the field along z. For clarity all disclination lines have
+been removed and the view is with the field direction into the paper plane. 
+The bottom row depicts results for a consecutive run with the field now 
+oriented along the y-direction. In the left panels the field is on, whereas the right panels 
+show the situation after switch-off. 
+The structure factor data is for wave vectors $k_y, k_z \in [-3\pi/8 , 3\pi/8]$.}
 \end{figure}

### 2013-11-08 trunk/papers/2012-sandwich/CholColloids.tex
Added pics to Nat. Mater. draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2234 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A), and finally subjected to an electric field along the vertical direction (B).
-The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.
+\centerline{
+\includegraphics[width=0.35\textwidth]{col_disc_x_run1267.png}
+\begin{minipage}[t]{0.17\textwidth}
+\vspace*{-4.6cm}
+\includegraphics[width=\textwidth]{sq_x_run1267.png}
+\includegraphics[width=\textwidth]{sq_y_run1267.png}
+\end{minipage}
+\includegraphics[width=0.35\textwidth]{col_disc_x_run1341.png}
+\begin{minipage}[t]{0.17\textwidth}
+\vspace*{-4.6cm}
+\includegraphics[width=\textwidth]{sq_x_run1341.png}
+\includegraphics[width=\textwidth]{sq_y_run1341.png}
+\end{minipage}
+}
+\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A).
+The isotropic nature of the state is evident from the structure factor (B, $k_y-k_z$ plane) and (C, $k_x-k_z$ plane).
+Subjected to an electric field along the vertical x-direction a honeycomb structure forms with hexagonal ordering perpendicular to the field direction (D), (E, $k_y-k_z$ plane) and (F, $k_x-k_z$ plane).
 Structures correspond to $w=0.23$ and $\phi=5\%$ (see Supporting Information

### 2013-11-08 trunk/papers/2012-sandwich/support.tex
Added pics to Nat. Mater. draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2234 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \begin{center}
-\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1342.png}
+\includegraphics[width=0.32\columnwidth]{col_parallel_z_run1342.png}\\
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
 \end{center}
-\caption{Panels A and B as for Fig. 3, but with the disclination lines
-removed for clarity. Panel C shows the situation when the field is
-switched off: a residual anisotropic structure remains which reflects the
-structure induced by the electric field. The electric field direction
-when switched on (panel B) is into the plane of the paper and along z.}
+\caption{}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width=0.24\columnwidth]{support-fig8a.png}
-\includegraphics[width=0.24\columnwidth]{support-fig8b.png}
-\includegraphics[width=0.24\columnwidth]{support-fig8c.png}
-\includegraphics[width=0.24\columnwidth]{support-fig8d.png}
+\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}\\
+\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}\\
 \end{center}
-\caption{Particle distribution during and after switched electric field direction: 
-The above images show the ordering with the electric field along the x-direction 
-after a previous run that had the field along z. For clarity disclination lines have
-been removed. The left panels show the configuration and the structure factor 
-while the field is on, whereas the righ panels show the same quantities after 
-switch-off. The structure factor data shows wave vectors $k_y, k_z \in [-3\pi/8 , 3\pi/8]$.}
+\caption{}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1347.png}
+\end{center}
+\caption{}
 \end{figure}

### 2013-11-08 papers/2012-sandwich/CholColloids.tex
Updated text main paper and supplement
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2235 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A), and finally subjected to an electric field along the vertical direction (B).
-The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.
+\centerline{
+\includegraphics[width=0.35\textwidth]{col_disc_x_run1267.png}
+\begin{minipage}[t]{0.17\textwidth}
+\vspace*{-4.6cm}
+\includegraphics[width=\textwidth]{sq_x_run1267.png}
+\includegraphics[width=\textwidth]{sq_y_run1267.png}
+\end{minipage}
+\includegraphics[width=0.35\textwidth]{col_disc_x_run1341.png}
+\begin{minipage}[t]{0.17\textwidth}
+\vspace*{-4.6cm}
+\includegraphics[width=\textwidth]{sq_x_run1341.png}
+\includegraphics[width=\textwidth]{sq_y_run1341.png}
+\end{minipage}
+}
+\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A).
+The isotropic nature of the state is refected in structure factor data (B and C for a cut through the 3D structure factor at $k_x=0$ and $k_y=0$, respectively).
+Subjected to an electric field along the x-direction a honeycomb structure forms with hexagonal ordering perpendicular to the field direction (D).
+The hexagonal structure is also evident from structure factor data (E and F for cuts at $k_x=0$ and $k_y=0$, respectively).
 Structures correspond to $w=0.23$ and $\phi=5\%$ (see Supporting Information

### 2013-11-08 papers/2012-sandwich/support.tex
Updated text main paper and supplement
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2235 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-10-03 papers/2012-sandwich/support.tex
@@
 \begin{center}
-\includegraphics[width=0.95\columnwidth]{support-fig7.pdf}
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1342.png}
+\includegraphics[width=0.32\columnwidth]{col_parallel_z_run1342.png}\\
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
 \end{center}
-\caption{Panels A and B as for Fig. 3, but with the disclination lines
-removed for clarity. Panel C shows the situation when the field is
-switched off: a residual anisotropic structure remains which reflects the
-structure induced by the electric field. The electric field direction
-when switched on (panel B) is into the plane of the paper.}
+\caption{Particle distribution and cuts through the structure factor: Panel A shows state D of Fig.3 with field along x in parallel projection.
+The disclinations are not plotted for clarity.
+The distribution after switch-off is shown in panel D (along the former field direction x, and with the view along y and z, respectively.
+Panels E and F are cuts though the 3D structure factor of state (D/B/C) at $k_x=0$ and $k_y=0$.
+The structure factor data is for wave vectors $k{x,y,z} \in [-3\pi/8 , 3\pi/8]$.}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}\\
+\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}\\
+\end{center}
+\caption{Particle distribution in parallel projection and cuts through the structure factor: Panel A shows the distribution while the field along y is on after a run with the field along x (SI Fig.7).
+Cuts through the corresponding structure factor at $k_x=0$ and $k_y=0$ are shown in panel B and C.
+The situation after switch-off is depicted in panel D (panels E and F for cuts through the structure factor at $k_x=0$ and $k_y=0$).
+The structure factor data is for wave vectors $k{x,y,z} \in [-3\pi/8 , 3\pi/8]$.}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
+\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1347.png}
+\end{center}
+\caption{Particle distribution in parallel projection and cuts through the structure factor: Panel A shows the distribution while the field along x is on after a run with the field along y (SI Fig.8).
+Cuts through the corresponding structure factor at $k_x=0$ and $k_y=0$ are shown in panel B and C.
+The situation after switch-off is depicted in panel D (panels E and F for cuts through the structure factor at $k_x=0$ and $k_y=0$).
+The structure factor data is for wave vectors $k{x,y,z} \in [-3\pi/8 , 3\pi/8]$.}
 \end{figure}
@@
 %\newpage
+\clearpage
 

### 2013-11-15 papers/2012-sandwich/CholColloids.tex
Add acknowledgements and rejig figures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2240 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-10-03 papers/2012-sandwich/CholColloids.tex
@@
 they offer exciting prospects for applications as biosensors~\cite{abbott}, or
-as new-generation devices~\cite{colloiddevice,tanakanatmat}.
+as new devices~\cite{colloiddevice,tanakanatmat}.
 
@@
 These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}.
-Our methodology and simulation parameters are summarized in the Supporting Information.
+Our methodology and simulation parameters are summarized in the Supplementary Information.
 
@@
 Fig.~2A-D shows the case of normal wall anchoring on varying $\phi$ and $w$.
-We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supporting Information).
+We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supplementary Information).
 The resulting colloid-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
@@
 direction.
-The resulting multistability is indeed a prerequisite for new generation 
+The resulting multistability is a prerequisite for new generation 
 switchable electrooptic devices for use in energy-saving electronics-free
@@
 We have not explored in this work their switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
+
+\bigskip
+\textbf{Acknowledgements}
+KS, OH, and JSL are supported by UK EPSRC grants EP/J007404/1, EP/G036136/1,
+and EP/F054750/1, respectively. MEC is supported by the Royal Society.
+We thank both Hector and PRACE for computational resources.
 
@@
 The anchoring of the director field to the colloidal surface is normal.
-[For the full parameter list used to generate Figs.~1-3 see Supporting Information.] For clarity, only a portion of the simulation box (a quarter of
-each axis in size) is shown; the full structures are shown in Fig. S2 and S3.}
+[For the full parameter list used to generate Figs.~1-3 see Supplementary
+Information.] For clarity, only a portion of the simulation box (a quarter of
+each axis in size) is shown; the full structures are shown in Supplementary
+Fig. S2 and S3.}
 \end{figure}
@@
 As in Fig. 1, only a portion of the simulation box is shown for clarity;
-the full structures are shown in Fig. S5 and S6.}
+the full structures are shown in Supplementary Fig. S5 and S6.}
 \end{figure}
@@
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{text-fig3.jpg}}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A), and finally subjected to an electric field along the vertical direction (B).
-The electric field leads to the formation of a honeycomb structure with hexagonal ordering perpendicular to the field direction.
-Structures correspond to $w=0.23$ and $\phi=5\%$ (see Supporting Information
-for full list of parameters). }
+\begin{center}
+\includegraphics[width=0.45\textwidth]{col_disc_x_run1267.png}
+\includegraphics[width=0.45\textwidth]{sq_x_run1267.png}\\
+\includegraphics[width=0.45\textwidth]{col_disc_x_run1341.png}
+\includegraphics[width=0.44\textwidth]{sq_x_run1341.png}
+\end{center}
+\caption{Steady states obtained when a dispersion of colloids is inserted
+into a thermodynamically stable bulk BPIII with and without applied
+external electric field. Panel A shows the situation in zero field, and panel
+C the situation when the applied field is in the $x$-direction
+(perpendicular to the plane of the paper). Panels B and D show one
+plane of the three-dimensional structure factor $S(\mathbf{k})$ computed
+from the positions of the particles, with wavevector $k_x = 0$, corresponding
+to A and C respectively. With field, the disclinations form a honeycomb
+pattern with hexagonal ordering perpendicular to the field direction;
+this is reflected in the structure factor.
+Simulations use weakly anchoring particles $w=0.23$, and a volume
+fraction of $\phi=5\%$ (see Supplementary Information for full list of
+parameters). }
 \end{figure}

### 2013-11-15 papers/2012-sandwich/support.tex
Add acknowledgements and rejig figures
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2240 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-{\bf Supporting Information for: Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
+{\bf Supplementary Information for: Self-assembly of colloid-cholesteric
+composites: A route to switchable optical materials}
 
@@
 
-{\bf Supporting Figures and Animations}
+{\bf Supplementry Figures and Animations}
 
@@
 \includegraphics[width=0.32\columnwidth]{col_parallel_x_run1341.png}
-\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1342.png}
-\includegraphics[width=0.32\columnwidth]{col_parallel_z_run1342.png}\\
+\includegraphics[width=0.32\columnwidth]{sq_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
 \includegraphics[width=0.32\columnwidth]{col_parallel_x_run1342.png}
@@
 \end{center}
-\caption{Particle distribution and cuts through the structure factor: Panel A shows state D of Fig.3 with field along x in parallel projection.
-The disclinations are not plotted for clarity.
-The distribution after switch-off is shown in panel D (along the former field direction x, and with the view along y and z, respectively.
-Panels E and F are cuts though the 3D structure factor of state (D/B/C) at $k_x=0$ and $k_y=0$.
-The structure factor data is for wave vectors $k{x,y,z} \in [-3\pi/8 , 3\pi/8]$.}
+\caption{Structure with and without electric field, with disclination lines
+removed for clarity. Panels A and B reproduce the results Fig. 3 panels
+C and D with the applied field in the $x$-direction (into the plane of the
+paper). Also included is a cut of the structure factor with $k_y = 0$
+(panel C). Panels D, E, and F show the corresponding situation after the
+external field has been removed and the structure allowed to relax,
+and show residual hexagonal ordering. The structure factor data is for wave
+vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
+$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).
+[ADD PARTICLE Y-VIEW?]}
 \end{figure}
+
+\newpage
 
@@
 \includegraphics[width=0.32\columnwidth]{col_parallel_y_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}\\
+\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
 \includegraphics[width=0.32\columnwidth]{col_parallel_y_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}\\
+\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
 \end{center}
-\caption{Particle distribution in parallel projection and cuts through the structure factor: Panel A shows the distribution while the field along y is on after a run with the field along x (SI Fig.7).
-Cuts through the corresponding structure factor at $k_x=0$ and $k_y=0$ are shown in panel B and C.
-The situation after switch-off is depicted in panel D (panels E and F for cuts through the structure factor at $k_x=0$ and $k_y=0$).
-The structure factor data is for wave vectors $k{x,y,z} \in [-3\pi/8 , 3\pi/8]$.}
+\caption{Cycling the electric field. Starting from the final position of
+Supplementary Fig.~7 panel~C, the field is re-applied, this time in the
+$y$-direction. The corresponding view of the particle distribution is
+shown, again viewing along the field direction, in panel A.
+Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
+shown in panels B and C, respectively. The corresponding situation when
+the field is again switched off, and the structure allowed to relax,
+is shown in panels d, E, and F. The wave vector values used are the
+same as in Supplementary Fig.~7. [ADD PARTIVLE $y$-VIEW?]}
 \end{figure}
+
+\newpage
 
@@
 \end{center}
-\caption{Particle distribution in parallel projection and cuts through the structure factor: Panel A shows the distribution while the field along x is on after a run with the field along y (SI Fig.8).
-Cuts through the corresponding structure factor at $k_x=0$ and $k_y=0$ are shown in panel B and C.
-The situation after switch-off is depicted in panel D (panels E and F for cuts through the structure factor at $k_x=0$ and $k_y=0$).
-The structure factor data is for wave vectors $k{x,y,z} \in [-3\pi/8 , 3\pi/8]$.}
+\caption{Cycling the electric field further. Starting from the final
+position of Supplementary Fig.~8 panel~D, the field is re-applied
+again in the $x-$direction. Panel A shows the particle configuration
+viewed along the field direction, and panels B and C show the structure
+factor with $k_x = 0$ and $k_y = 0$ respectively. The corresponding
+situation when the field is removed and the structure allowed to relax
+is shown in panels D, E, and F. The wave vector values are again as
+in Supplementary Fig.~7. [DIFFERENCE CF 7?]}
 \end{figure}

### 2013-11-15 papers/2012-sandwich/support.tex
edited support file
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2241 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \caption{Structure with and without electric field, with disclination lines
-removed for clarity. Panels A and B reproduce the results Fig. 3 panels
-C and D with the applied field in the $x$-direction (into the plane of the
-paper). Also included is a cut of the structure factor with $k_y = 0$
-(panel C). Panels D, E, and F show the corresponding situation after the
+removed for clarity. Panels A and B reproduce the results in Fig. 3C and
+Fig. 3D respectively with the applied field in the $x$-direction 
+(out of the plane of the paper). Also included is a cut of the structure 
+factor with $k_y = 0$ (panel C). 
+Panels D, E, and F show the corresponding situation after the
 external field has been removed and the structure allowed to relax,
@@
 vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
-$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).
-[ADD PARTICLE Y-VIEW?]}
+$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
 \end{figure}
@@
 \caption{Cycling the electric field. Starting from the final position of
-Supplementary Fig.~7 panel~C, the field is re-applied, this time in the
+Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
 $y$-direction. The corresponding view of the particle distribution is
@@
 is shown in panels d, E, and F. The wave vector values used are the
-same as in Supplementary Fig.~7. [ADD PARTIVLE $y$-VIEW?]}
+same as in Supplementary Fig.~7.}
 \end{figure}
@@
 is shown in panels D, E, and F. The wave vector values are again as
-in Supplementary Fig.~7. [DIFFERENCE CF 7?]}
+in Supplementary Fig.~7. Note that the on and off states are 
+equivalent to those in Supplementary Fig.~7, up to a rotation.
+This is because all hexagonal structures perpendicular to the field
+direction are energetically degenerate,
+and the actual orientation of the hexagon emerges spontaneously during
+switching. }
 \end{figure}

### 2013-11-15 trunk/papers/2012-sandwich/CholColloids.tex
edited text
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2242 40876441-3496-0410-a613-a2fca35b2ae2

@@
 results in the particle occupying the available nodes, where the order
-parameter is low (Fig.~3A; Suppl. Movie 5). Applying a high enough field along the $z$ 
-direction leads to a switching to a new phase, where colloidal arrays fit 
-within a network with distorted hexagonal cells (Fig.~3B; Suppl. Movie 6). Remarkably, upon 
-field removal this honeycomb structure does not find its way back to the 
+parameter is low (Fig.~3A; Suppl. Movie 5). Applying a high enough field 
+along the $x$ direction leads to a switching to a new phase, where colloidal 
+arrays fit within a network with distorted hexagonal cells (Fig.~3B 
+and Fig. S7; Suppl. Movie 6). 
+Remarkably, upon field removal this honeycomb structure does not find its 
+way back to the 
 configuration before switching on the field, but gets stuck into a metastable 
 structure (Fig. S7), with residual anisotropy along the field switching 
-direction.
-The resulting multistability is indeed a prerequisite for new generation 
+direction. Cycling the field along the $x$ and $y$ directions leads to
+distinct metastable structures, which are retained after field removal
+(Fig. S8 and S9).
+The resulting multistability is a prerequisite for new generation 
 switchable electrooptic devices for use in energy-saving electronics-free
@@
 \begin{figure}
-\centerline{
-\includegraphics[width=0.35\textwidth]{col_disc_x_run1267.png}
-\begin{minipage}[t]{0.17\textwidth}
-\vspace*{-4.6cm}
-\includegraphics[width=\textwidth]{sq_x_run1267.png}
-\includegraphics[width=\textwidth]{sq_y_run1267.png}
-\end{minipage}
-\includegraphics[width=0.35\textwidth]{col_disc_x_run1341.png}
-\begin{minipage}[t]{0.17\textwidth}
-\vspace*{-4.6cm}
-\includegraphics[width=\textwidth]{sq_x_run1341.png}
-\includegraphics[width=\textwidth]{sq_y_run1341.png}
-\end{minipage}
-}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids is inserted into a thermodynamically stable bulk-BPIII (A).
-The isotropic nature of the state is evident from the structure factor (B, $k_y-k_z$ plane) and (C, $k_x-k_z$ plane).
-Subjected to an electric field along the vertical x-direction a honeycomb structure forms with hexagonal ordering perpendicular to the field direction (D), (E, $k_y-k_z$ plane) and (F, $k_x-k_z$ plane).
-Structures correspond to $w=0.23$ and $\phi=5\%$ (see Supporting Information
-for full list of parameters). }
+\begin{center}
+\includegraphics[width=0.45\textwidth]{col_disc_x_run1267.png}
+\includegraphics[width=0.45\textwidth]{sq_x_run1267.png}\\
+\includegraphics[width=0.45\textwidth]{col_disc_x_run1341.png}
+\includegraphics[width=0.44\textwidth]{sq_x_run1341.png}
+\end{center}
+\caption{Steady states obtained when a dispersion of colloids is inserted
+into a thermodynamically stable bulk BPIII with and without applied
+external electric field. Panel A shows the situation in zero field, and panel
+C the situation when the applied field is in the $x$-direction
+(perpendicular to the plane of the paper). Panels B and D show one
+plane of the three-dimensional structure factor $S(\mathbf{k})$ computed
+from the positions of the particles, with wavevector $k_x = 0$, corresponding
+to A and C respectively. With field, the disclinations form a honeycomb
+pattern with hexagonal ordering perpendicular to the field direction;
+this is reflected in the structure factor.
+Simulations use weakly anchoring particles $w=0.23$, and a volume
+fraction of $\phi=5\%$ (see Supplementary Information for full list of
+parameters). }
 \end{figure}

### 2013-11-15 papers/2012-sandwich/covercholcolloids.tex
cover letter
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2243 40876441-3496-0410-a613-a2fca35b2ae2

%\documentclass[aps,prl,groupedaddress,showpacs]{revtex4}
\documentclass[12pt]{article}

\usepackage{graphicx,subfigure}
\usepackage{color}

\pagestyle{empty}

\begin{document}

Dear Editor,\\

We would like to submit the attached manuscript, `Self-assembly of 
colloid-cholesteric composites: A route to switchable optical materials', 
for consideration for publication in {\it Nature Materials}. \\

We believe that our work merits publication in {\it Nature Materials} for the following reasons:\\

1. In scale and scope the work represents a major breakthrough in what 
can be achieved computationally in studying the phase behaviour of soft 
composite materials.

2. Here we give evidence that colloidal dispersions in chiral liquid 
crystals lead to a much larger range of structures than previously
expected. These new soft materials can be realised by tuning
the anchoring of the liquid crystal at the colloidal surface,
or at boundary planes, as well as the particle concentration. 
The geometries we consider are directly 
relevant to the physics of devices which could be built by
using the new materials we find in simulations. All our
structures can therefore be, in principle, found in the lab.

3.  The colloid-liquid crystal composite material is switchable
by a field. Remarkably, we have also found conditions which should
lead to the design of a multistable switchable device, which can
retain memory of several different states after field removal:
this is a must-have prerequisites for energy-saving electronics-free
liquid crystal devices (such as electronic paper and smart glass).\\

Given the substantial nature of the advancement we provide here, 
and its relevance for (i) our understanding of the fundamental behaviour
of multiscale composite materials and (ii) our capabilities to build new
devices based on them, we believe that {\it Nature Materials}
is the ideal avenue to communicate our results.\\

With best wishes,\\

K. Stratford, O. Henrich, J.~S. Lintuvuori, M.~E. Cates, D. Marenduzzo

\end{document}

J.J. de Pablo

T.C. Lubensky

H. Tanaka

H. Stark

J.C. Loudet

### 2013-11-21 trunk/papers/2012-sandwich/CholColloids.tex
Updated pictures Nat Mater draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2244 40876441-3496-0410-a613-a2fca35b2ae2
updated text (21 Nov, davide)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2256 40876441-3496-0410-a613-a2fca35b2ae2

@@
 These include regular crystals, glasses, percolating gels, isolated clusters, twisted rings and undulating colloidal ropes.
-The final structure can be tuned through the particle concentration, and by varying the surface interactions of the cholesteric host with both the colloids and any confining walls.
+The final structure can be tuned via the particle concentration, and by varying the surface interactions of the cholesteric host with both the colloids and any confining walls.
 Remarkably, we find that many of these new materials are metastable: two or more structures can arise under identical thermodynamic conditions.
@@
 but also on the history of an applied electric field.  
-This new class of soft materials should thus be relevant to the design of switchable, multistable devices for optical technologies including, for example, e-paper~\cite{epaper}.}
+This new class of soft materials should thus be relevant to the design of switchable, multistable devices for optical technologies including, for example, smart glass and e-paper~\cite{epaper}.}
 
@@
 Besides being of fundamental interest to materials science, these
-structures have tunable elastic and optical properties~\cite{stark}. Hence 
+structures may have tunable elastic and optical properties~\cite{stark}. Hence 
 they offer exciting prospects for applications as biosensors~\cite{abbott}, or
@@
 lead to geometric frustration because it is impossible to tile the whole 3D
-space with DTCs without also introducing disclination lines: singular topological defects on which the director field is undefined.
-In blue phases these disclination lines themselves either form a 3D periodic regular lattice (in so called BPI and BPII), or remain disordered (BPIII)~\cite{bp3}.
+space with DTCs without also introducing ``disclination lines'' 
+(These are singular topological defects on which the director field is 
+undefined.) In blue phases these disclination lines themselves either form a 3D periodic regular lattice (in so called BPI and BPII), or remain disordered (BPIII)~\cite{bp3}. 
 %
@@
 particles, they can potentially template colloidal self-assembly. 
-This was proved recently by simulations of BP-dispersed nanoparticles which interact only weakly with the molecules of a BP (to give a so-called `weak anchoring' regime)~\cite{miha}.
+This was proved recently by simulations of BP-dispersed nanoparticles which 
+exclude the surrounding liquid crystal but otherwise have negligible surface 
+interaction with it (to give a so-called `weak anchoring' regime)~\cite{miha}. 
 Such particles are attracted to the disclination network in BPs: by covering up the defect cores, the high elastic energy cost of those regions is avoided.
@@
 
-We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles within it.
-Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal (defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface).
-For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
+We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles at random within it.
+Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
+(This is defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface.) For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
 This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.
@@
 
-All the structures in Fig.~1A-D should be soft solids, of nonzero elastic 
+As colloid-nematic composites~\cite{tiffany}, all the structures in Fig.~1A-D 
+should be soft solids, of nonzero elastic 
 modulus $G$ at low frequency. This contrasts with pure BPs, where
@@
 
-A more practical protocol for including particles in a BPI phase is to disperse the particles in an isotropic phase, and then quenching this into the temperature range where BPI is stable.
+A more practical protocol for including particles in a BPI phase is to disperse the particles at random in an isotropic phase, and then quenching this into the temperature range where BPI is stable.
 Here we know that, without particles, the kinetics can favour disordered (BPIII-like) intermediates which may be long lived or even metastable~\cite{domaingrowth}.
@@
 There are, however, significant effects of the confining walls.
-First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls and also in the midplane (see side-views in Fig.~2A-D).
-Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit.
-These clusters
+First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls 
+(see side-views in Fig.~2A-D). Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit. These clusters
 are similar to those reported experimentally for large colloids in a
@@
 (Fig. S8 and S9).
-The resulting multistability is a prerequisite for new generation 
-switchable electrooptic devices for use in energy-saving electronics-free
-appliances, smart glass, and e-paper to cite just a few examples.
+%The resulting multistability is a prerequisite for new generation 
+%switchable electrooptic devices for use in energy-saving electronics-free
+%appliances, smart glass, and e-paper to cite just a few examples.
 
@@
 In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls.
-We have also shown that electric fields can be used to switch between metastable states, providing a route to future multistable energy-saving devices.
+We have also shown that electric fields can be used to switch between metastable states, providing a possible route to future multistable device applications.
 %This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
-We have not explored in this work their switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case they may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
+We have not explored in this work switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case the same materials may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
 
@@
 \textbf{Acknowledgements}
-KS, OH, and JSL are supported by UK EPSRC grants EP/J007404/1, EP/G036136/1,
-and EP/F054750/1, respectively. MEC is supported by the Royal Society.
+We acknowledge support by UK EPSRC grants EP/J007404/1 (KS and JSL), 
+EP/G036136/1 (OH), and EP/F054750/1 (KS), respectively. 
+MEC is supported by the Royal Society.
 We thank both Hector and PRACE for computational resources.
@@
 {\it Phys. Rep.} {\bf 351}, 387-474 (2001).
-\bibitem{mermin} Wright D.~C., Mermin, N.~D. Crystalline liquids: The blue phases. {\it Rev. Mod. Phys.} {\bf 61}, 385–432 (1989).
+\bibitem{mermin} D.~C. Wright, Mermin, N.~D. Crystalline liquids: The blue phases. {\it Rev. Mod. Phys.} {\bf 61}, 385–432 (1989).
 \bibitem{miha} M. Ravnik, G.~P. Alexander, J.~M. Yeomans,  
@@
 \begin{center}
-\includegraphics[width=0.45\textwidth]{col_disc_x_run1267.png}
+\includegraphics[width=0.5\textwidth]{col_disc_x_run1267.png}
 \includegraphics[width=0.45\textwidth]{sq_x_run1267.png}\\
-\includegraphics[width=0.45\textwidth]{col_disc_x_run1341.png}
-\includegraphics[width=0.44\textwidth]{sq_x_run1341.png}
+\includegraphics[width=0.5\textwidth]{col_disc_x_run1341.png}
+\includegraphics[width=0.45\textwidth]{sq_x_run1341.png}
 \end{center}

### 2013-11-21 trunk/papers/2012-sandwich/covercholcolloids.tex
updated cover letter (21 Nov, davide)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2257 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-11-15 papers/2012-sandwich/covercholcolloids.tex
@@
 
-1. In scale and scope the work represents a major breakthrough in what 
-can be achieved computationally in studying the phase behaviour of soft 
-composite materials.
-
-2. Here we give evidence that colloidal dispersions in chiral liquid 
+1. We give evidence that colloidal dispersions in chiral liquid 
 crystals lead to a much larger range of structures than previously
@@
 
-3.  The colloid-liquid crystal composite material is switchable
-by a field. Remarkably, we have also found conditions which should
-lead to the design of a multistable switchable device, which can
+2.  At least one of our new colloid-liquid crystal composite materials 
+is switchable by a field. Remarkably, we have also found conditions which 
+should lead to the design of a multistable switchable device, which can
 retain memory of several different states after field removal:
-this is a must-have prerequisites for energy-saving electronics-free
-liquid crystal devices (such as electronic paper and smart glass).\\
+this is a prerequisite for energy-saving electronics-free
+liquid crystal devices (such as electronic paper and smart glass).
+
+3. In scale and scope the work represents a major breakthrough in what 
+can be achieved computationally in studying the phase behaviour of soft 
+composite materials. \\
 
@@
 of multiscale composite materials and (ii) our capabilities to build new
-devices based on them, we believe that {\it Nature Materials}
-is the ideal avenue to communicate our results.\\
+devices based on them, we believe that publication in 
+{\it Nature Materials} is warranted.\\
 
@@
 
+R. D. Kamien
+
 H. Stark
 
-J.C. Loudet
+J.C. Loudet / Poulin

### 2013-11-21 papers/2012-sandwich/CholColloids.tex
updated ms (21 Nov, davide)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2258 40876441-3496-0410-a613-a2fca35b2ae2

@@
 results in the particle occupying the available nodes, where the order
-parameter is low (Fig.~3A; Suppl. Movie 5). Applying a high enough field along the $z$ 
-direction leads to a switching to a new phase, where colloidal arrays fit 
-within a network with distorted hexagonal cells (Fig.~3B; Suppl. Movie 6). Remarkably, upon 
-field removal this honeycomb structure does not find its way back to the 
+parameter is low (Fig.~3A; Suppl. Movie 5). Applying a high enough field 
+along the $x$ direction leads to a switching to a new phase, where colloidal 
+arrays fit within a network with distorted hexagonal cells (Fig.~3B 
+and Fig. S7; Suppl. Movie 6). 
+Remarkably, upon field removal this honeycomb structure does not find its 
+way back to the 
 configuration before switching on the field, but gets stuck into a metastable 
 structure (Fig. S7), with residual anisotropy along the field switching 
-direction.
-The resulting multistability is a prerequisite for new generation 
-switchable electrooptic devices for use in energy-saving electronics-free
-appliances, smart glass, and e-paper to cite just a few examples.
+direction. Cycling the field along the $x$ and $y$ directions leads to
+distinct metastable structures, which are retained after field removal
+(Fig. S8 and S9).
+%The resulting multistability is a prerequisite for new generation 
+%switchable electrooptic devices for use in energy-saving electronics-free
+%appliances, smart glass, and e-paper to cite just a few examples.
 
@@
 \textbf{Acknowledgements}
-KS, OH, and JSL are supported by UK EPSRC grants EP/J007404/1, EP/G036136/1,
-and EP/F054750/1, respectively. MEC is supported by the Royal Society.
+We acknowledge support by UK EPSRC grants EP/J007404/1 (KS and JSL), 
+EP/G036136/1 (OH), and EP/I030298/1 (KS and JSL).
+MEC is supported by the Royal Society.
 We thank both Hector and PRACE for computational resources.

### 2013-11-21 papers/2012-sandwich/CholColloids.tex
corrections
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2259 40876441-3496-0410-a613-a2fca35b2ae2

@@
 For larger $w$, we observe thick colloidal ``ropes" about 5 colloids across
-(Fig.~2E,H). These ropes undulate to follow the chiral structure of the 
+(Fig.~2G,H). These ropes undulate to follow the chiral structure of the 
 underlying fluid.
@@
 Structures correspond
-to (A,E) $w=0.23$ and $\phi=1\%$, 
-(B,F) $w=0.23$ and $\phi=5\%$, 
-(C,G) $w=2.3$ and $\phi=1\%$,
+to (A,E) $w=0.23$ and $\phi=1\%$;
+(B,F) $w=0.23$ and $\phi=5\%$; 
+(C,G) $w=2.3$ and $\phi=1\%$; and
 (D,H) $w=2.3$ and $\phi=5\%$.
@@
 [For the full parameter list used to generate Figs.~1-3 see Supplementary
-Information.] For clarity, only a portion of the simulation box (a quarter of
-each axis in size) is shown; the full structures are shown in Supplementary
-Fig. S2 and S3.}
+Information.] For clarity, only a portion of the simulation box is shown
+(one eighth of the total volume); the full structures are shown in
+Supplementary Fig. S2 and~S3.}
 \end{figure}
@@
 (D,H) $w=2.3$ and $\phi=2\%$.
-As in Fig. 1, only a portion of the simulation box is shown for clarity;
+As in Fig. 1, only one quarter of the simulation box is shown for clarity;
 the full structures are shown in Supplementary Fig. S5 and S6.}

### 2013-11-21 papers/2012-sandwich/support.tex
Updated pictures Nat Mater draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2244 40876441-3496-0410-a613-a2fca35b2ae2
corrections
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2259 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-{\bf Supplementary Information for: Self-assembly of colloid-cholesteric
-composites: A route to switchable optical materials}
+{\bf
+Self-assembly of colloid-cholesteric
+composites: A route to switchable optical materials.
+
+Supplementary Information}
 
@@
 order parameter ${\bf Q}$ and its gradients; ${\bf Q}$ is a traceless 
-and symmetric tensor which we will denote from now on in subscript
+and symmetric tensor which is denoted from now on in subscript
 notation as $Q_{\alpha\beta}$.
@@
 provide an order parameter.
-We avoid the mean-field spinodal point at $\gamma = 3.0$:
+The mean-field spinodal point at $\gamma = 3.0$ is avoided and
 the parameters are the same as for the initial simulations with
@@
 
-For simplicity, we do not allow the size of the BPI unit cell to 
-dynamically readjust to minimize free energy (there is no 
+For simplicity,  the size of the BPI unit cell is not allowed to 
+readjust dynamically to minimize free energy (there is no 
 ``redshift''~\cite{blue1}): this would lead to quantitatively different 
@@
 
-In order to demonstrate structural information about the field-aligned 
-states and the residual anisotropy after switch off we followed a 
-procedure similar to the one described in \cite{oliver-bp3}.
-The structure factor in the current approach is defined via
-the Fourier transform of the colloid density:
+To provide structural information about the field-aligned 
+states and the residual anisotropy after switch off a 
+procedure similar to the one described in \cite{oliver-bp3} is
+followed.
+The structure factor $S(\mathbf{k})$ in the current approach is
+defined via the Fourier transform of the colloid density $F(\mathbf{k})$:
 
-\begin{eqnarray}
-S({\bf k})&=&|F({\bf k})|^2\\
-F({\bf k})&=&\int_V d{\bf r} \rho({\bf r}) e^{i {\bf k r}}
-\end{eqnarray}
+%\begin{eqnarray}
+%S({\bf k})&=&|F({\bf k})|^2,\\
+%F({\bf k})&=&\int d^3{\bf r} \rho({\bf r}) e^{i {\bf k. r}}.
+%\end{eqnarray}
+
+\begin{equation}
+S({\bf k}) =  |F({\bf k})|^2, \mathrm{with}\
+F({\bf k}) = \int d^3{\bf r} \rho({\bf r}) e^{i {\bf k. r}}.
+\end{equation}
 
@@
 standard definition of a structure factor and does not consider the colloids 
-as pointlike objects, allows to gain the same structural information
-in a simpler way.
+as point-like objects, allows the same structural information
+to be obtained in a simple way.
 
@@
 \bibitem{blue1}
-D.C. Wright and N.D. Mermin,
+D.C. Wright and N.D. Morin,
 %  Crystalline liquids -- the blue phases.
@@
 \bibitem{deGennes}
-P.-G. de Gennes, J. Prost,
+P.-G. DEC Gunner's, J. Poorest,
 {\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).
@@
 
-{\bf Supplementry Figures and Animations}
+{\bf Supplementary Figures and Animations}
 
@@
 \begin{center}
-\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1341.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1341.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
 \includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
-\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1342.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
 \includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
@@
 \begin{center}
-\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1343.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
 \includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
 \includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
-\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1344.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
 \includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
@@
 \begin{center}
-\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1345.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1345.png}
 \includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
 \includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
-\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1347.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1347.png}
 \includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}

### 2013-12-12 trunk/htdocs/electrokinetic.tex
Added PETSc scaling
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2278 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-01-31 htdocs/electrokinetic.tex
@@
 
-Taking the gradient of the stress tensor gives
+Taking the divergence of the stress tensor gives
 
@@
 \end{figure}
+
+\subsubsection{Scaling with Krylov Subspace Solver}
+
+We use a Krylov subspace solver from the Portable Extendabl Toolkit for Scientific Computation (PETSc.
+
+\begin{figure}[htpb]
+\includegraphics[width=0.9\textwidth]{./pics/petsc_scaling.pdf}
+\caption{Scaling of a system of $128^3$ lattice sites on up to 4096 MPI tasks. 
+The scaling of both the Nernst Planck part as well as the Poisson solver are only limited by the overhead of communication to computation at very small decompositions.} 
+\label{fig6} 
+\end{figure}
 \clearpage

### 2014-02-20 trunk/papers/2012-sandwich/response.txt
Reviewers comments and repsonse
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2310 40876441-3496-0410-a613-a2fca35b2ae2

Date: Mon, 17 Feb 2014 04:37:57 -0500
From: c.huang@nature.com
To: dmarendu@ph.ed.ac.uk
Subject: Decision on manuscript NCOMMS-14-00180-T

** Please ensure you delete the link to your author home page in this e-mail
if you wish to forward it to your coauthors **

Dear Dr Marenduzzo,

Your manuscript entitled "Self-assembly of colloid-cholesteric composites: A
route to switchable optical materials" has now been seen by 2 referees,
whose comments are appended below. You
will see that, while they find your work of interest, they have raised
points that need to be addressed before we can make a decision on
publication.

We therefore invite you to revise and resubmit your manuscript, taking into
account the points raised. At the same time, we ask that you ensure your
manuscript complies with our format
requirements, which are summarized in the following checklist:
http://www.nature.com/ncomms/authors/ncomms_manuscript_checklist.pdf

Please use the following link to submit your revised manuscript, completed
checklist and point-by-point response to the referees' comments (which
should be in a separate document to any
cover letter):
http://mts-ncomms.nature.com/cgi-bin/main.plex?el=A6S2Lka4A1DfBL5I4A9v6k5dsV
e8yORLGQzsElBAZ

** This url links to your confidential home page and associated information
about manuscripts you may have submitted or be reviewing for us. If you wish
to forward this email to co-authors,
please delete the link to your homepage first **

We hope to receive your revised paper within two months; please let us know
if you aren't able to submit it within this time so that we can discuss the
best way forward. If we don't hear
from you, and the revision process takes significantly longer, we will close
your file. In this event, we will still be happy to reconsider your paper at
a later date, as long as nothing
similar has been accepted for publication at Nature Communications or
published elsewhere in the meantime.

We look forward to hearing from you soon.

Best regards,

Congcong Huang, PhD

Associate Editor
Nature Communications

Unit 10-11, 42F, The Center
989 Changle Road, Xuhui District
Shanghai, China, 200031
Tel: +86-21-2422-5011
Fax: +86-21-2422-5015
www.nature.com/naturecommunications

Reviewers' comments:

Reviewer #1 (Remarks to the Author):

The authors describe the results of a study of colloidal particles dispersed
in cholesteric blue phases (BP) focusing on the effects of particle
concentration and distribution. Additionally
effects of confinement and external fields are investigated. The use of the
lattice Boltzmann approach allows to follow dynamics after the thermal
quench and electric field application. The
study is an important step in modeling of BPs and should deserve a
publication in Nature Communications. Nevertheless there are some weak
points that need improvement before the publication.
Below is the list of remarks.

General remarks:

The term colloid is by the definition a dispersion of particles in a fluid.
Using the term also for the particles my lead to a confusion. In the
manuscript often also a more appropriate word
colloidal particle is used.

It would be important to add comments that would help experimentalist in
designing experiments with colloidal blue phases and perhaps stimulate
development of applications on the base of
these materials.

Particular remarks:

Page 2

In the discussion of the physical understanding of trapping of colloidal
particles in BP disclination networks it would be useful to refer to the
initial study by Ravnik et al. in Faraday
Discussions 2009 devoted to this topic.

250 K is more than the total mesophase stability range!

Page 3

The statement "This is defined such that W > 0 gives preferential
perpendicular orientation of the director field at the particle surface." is
not consistent with the interaction form
described in the Supplement.

For potential readers it would be useful to specify the particle radius in
nanometers in the main text. How it compares to the nematic correlation
length? How the radius influences the
structures? How tiny the particles can still be that the method based on the
Landau-de Gennes free energy is justified.

Is WR/K = 2.8 justified for nanoparticles used in the study?

Page 4:

Can the G elastic modulus be estimated for modeled structures?

Page 5:

Can annealing of the structures obtained with the thermal quench lead to
ordered colloidal structures?

Is also in case of the disordered structures observed a change in the
stability range of the colloidal BPs?

In the discussion of colloidal particles in BP thin layers it would be good
to comment the study of
Ravnik et al. in Soft Matter 2011 devoted the ordered structures of such a
system.

Page 6:

It would be good to comment the difference between structures in the bulk
and thin layer for particles with strong anchoring. Does this depend on the
visualization or is more behind?

Are the electric field induced changes irreversible only in case of
colloidal dispersions?

Page 7

Reference 2 has a mistake.

Supplement

Segment 4.2:

The statement "Wall anchoring is either normal or planar with W=K = 1 in
simulation units" is unclear.

References of the Supplement:
There are mistakes in references 1 and 2.

Reviewer #2 (Remarks to the Author):

This paper, "Self-assembly of colloid-cholesteric composities: A route to
switchable optical materials" reports simulations of the formation of
particle networks in a cholesteric host and in
particular in the blue phase of the cholesteric host. It reports the
influence of various physical parameters and in particular the surface
anchoring and concentration of the colloidal
particles. The paper also reports how external fields can be employed to
change the morphology of these colloidal composites and how this may be
employed in practical optical devices.

The paper builds on a body of previous research, both theoretical and
experimental that has explored the effects of dispersing particles in a
liquid crystal host. The novelty of the paper is
to model the effect of particles in the blue phases of a cholesteric host
ahd to postulate how these colloids may be employed in specific
appilcations.

I am not a theorteical physiscist and therefore will not comment on the
methodology but rather will take the perspective of an experimentalist and
how useful this paper is in understanding
these materials and more importantly how useful the paper is in suggesting
and designing experimental procedures and developing the actual materials.

The paper provides insight into the effectiveness of including particles
into the blue phase as a means of controlling and adjusting the structure
and stability of the materials. The paper
is clearly written and the figures, paricularly the included 3D videos,
provide clear perspective on the modeled phases of these materials. The
results are of interest and useful to
experimentalists working in the field. The paper would be even more useful
if it included references to actual materials, both the liquid crystal host
and the added particles and perhaps
even suggested a model system that would be a good starting point to study
experimentally. In a similar vein, it would also be useful to address the
practical issue of reproducibility, which
has been a major problem with these materials. The theoretical results
clearly indicate how important the formation conditions are to the final
result. What is the authors sense of how easy
it will be to reproducibly fabricate a device
with a specific morphology and optical properties. This seems a likely
practical issue particularly with the rich variety of structures reported in
the simulations.

** See NPG's author and referees' website at www.nature.com/authors for
information about policies, services and author benefits

This email has been sent through the NPG Manuscript Tracking System
NY-610A-NPG&MTS

Confidentiality Statement:

This e-mail is confidential and subject to copyright. Any unauthorised use
or disclosure of its contents is prohibited. If you have received this email
in error please notify our Manuscript
Tracking System Helpdesk team at http://npgapps.custhelp.com/app/ask.

Details of the confidentiality and pre-publicity policy may be found here
http://www.nature.com/authors/policies/confidentiality.html

Privacy Policy | Update Profile

### 2014-02-20 papers/2012-sandwich/reply.tex
Added draft reply
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2311 40876441-3496-0410-a613-a2fca35b2ae2

%\documentclass[aps,prl,groupedaddress,showpacs]{revtex4}
\documentclass[12pt]{article}

\usepackage{graphicx,subfigure}
\usepackage{color}

\pagestyle{empty}

\begin{document}

Dear Editor,\\

We would like to thank you and the referees for valuable comments to our 
manuscript, `Self-assembly of colloid-cholesteric composites: A 
route to switchable optical materials'. \\

Please find below a summary of the changes we applied and a detailed
response to the comments of the referees:\\

{\bf Referee\#1}\\

{\it General remarks}\\
- The term colloid has been replaced with colloidal particle throughout the main body of the manuscript and the supplement.\\
- Added comment about experimental realisations of BPs.\\

{\it Particular remarks}\\
- Ravnik et al Faraday Disc. 2009 has been cited.\\
- 250K and stability range of the mesophase\\
- inconsistency of the interaction form between supplement and main article\\
- specify particle radius and validity range of Landau-de Gennes FE functional\\
- WR/K=2.8 is strong anchoring regime\\
- elastic modulus G\\
- Annealing and thermal quench does not lead to ordered structures as can be seen from Fig.\\
- We did not determine the stability range of the disordered colloid-BP structures.\\
- A comment and reference to Ravnik et al. Soft Matter 2011 has been added.\\
- Comment on difference between structures in the bulk and thin layers with strong anchoring\\
- The changes induced by the electric field are only irreversible in the presence of colloidal particles with sufficiently strong anchoring.\\
- Reference 2 has been corrected\\

{\it Supplement}\\
- The comment about planar and normal wall anchoring has been changed and clarified.\\
- References 1 and 2 have been corrected.\\

{\bf Referee\#2}\\

{\it General remarks}\\
- References to actual materials, liquid crystal host materials and particles have been added.\\
- Comment about reproducibility of these materials has been added.\\

\end{document}

### 2014-02-21 papers/2012-sandwich/response.txt
First revision update
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2312 40876441-3496-0410-a613-a2fca35b2ae2

--- 2014-02-20 trunk/papers/2012-sandwich/response.txt
@@
-Date: Mon, 17 Feb 2014 04:37:57 -0500
-From: c.huang@nature.com
-To: dmarendu@ph.ed.ac.uk
-Subject: Decision on manuscript NCOMMS-14-00180-T
-
-** Please ensure you delete the link to your author home page in this e-mail
-if you wish to forward it to your coauthors **
-
-Dear Dr Marenduzzo,
-
-Your manuscript entitled "Self-assembly of colloid-cholesteric composites: A
-route to switchable optical materials" has now been seen by 2 referees,
-whose comments are appended below. You
-will see that, while they find your work of interest, they have raised
-points that need to be addressed before we can make a decision on
-publication.
-
-We therefore invite you to revise and resubmit your manuscript, taking into
-account the points raised. At the same time, we ask that you ensure your
-manuscript complies with our format
-requirements, which are summarized in the following checklist:
-http://www.nature.com/ncomms/authors/ncomms_manuscript_checklist.pdf
-
-Please use the following link to submit your revised manuscript, completed
-checklist and point-by-point response to the referees' comments (which
-should be in a separate document to any
-cover letter):
-http://mts-ncomms.nature.com/cgi-bin/main.plex?el=A6S2Lka4A1DfBL5I4A9v6k5dsV
-e8yORLGQzsElBAZ
-
-** This url links to your confidential home page and associated information
-about manuscripts you may have submitted or be reviewing for us. If you wish
-to forward this email to co-authors,
-please delete the link to your homepage first **
-
-We hope to receive your revised paper within two months; please let us know
-if you aren't able to submit it within this time so that we can discuss the
-best way forward. If we don't hear
-from you, and the revision process takes significantly longer, we will close
-your file. In this event, we will still be happy to reconsider your paper at
-a later date, as long as nothing
-similar has been accepted for publication at Nature Communications or
-published elsewhere in the meantime.
-
-We look forward to hearing from you soon.
-
-Best regards,
-
-Congcong Huang, PhD
-
-Associate Editor
-Nature Communications
-
-Unit 10-11, 42F, The Center
-989 Changle Road, Xuhui District
-Shanghai, China, 200031
-Tel: +86-21-2422-5011
-Fax: +86-21-2422-5015
-www.nature.com/naturecommunications
+Point by point response
 
@@
 
+RESPONSE: We have standardised on "colloidal particle(s)" or simply
+"particle(s)" where appropriate.
+
 It would be important to add comments that would help experimentalist in
@@
 
+RESPONSE: Citition for Ranvik et al Faraday Discuss. 2010 (sic) is added
+[current reference 16].
+
 250 K is more than the total mesophase stability range!
+
+RESPONSE: A stability range for templated blue phases of -125 to
++125 degrees Celsius is reported by Castles et al. Nat. Mat. 2012.
+We have made this connection more clear by rewriting the relevant
+sentence: "However, recent advances in formalation have widened
+this stability range enormously [11,12,13,14]. Most recently,
+templated blue phases with a stability range of -125 to 125$^\circ$C
+have been reported [Castles], paving the way to applications of BPs
+in operational display devices."
 
@@
 
+RESPONSE: We agree this statement is confusing. It has been replaced by
+a clearer statement: "The anchoring strength is defined such that
+W > 0 imparts a preferred orientation to the director field at the
+particle surface. This orientation can be either normal to the
+surface or in a plane tengential to the surface, a property related
+to surface chemistry in experimental systems but which can be imposed
+at will in simulations."
+
 For potential readers it would be useful to specify the particle radius in
@@
 Is WR/K = 2.8 justified for nanoparticles used in the study?
+
+RESPONSE: We have added in the main text (last para. introduction) a
+summary of the relevant
+dimensions in physical units: "The simulations represent a liquid crystal
+blue phase with unit cell size in the range of 100--500~nm \cite{mermin},
+and particles with a radius of around 50~nm."
+
+...
 
@@
 
+REESPONSE: A citition for Ravnik et al (2011) has been added [current
+reference 26].
+
 Page 6:
@@
 
+RESPONSE: The corrent punctuation in the page numbers has been
+inserted in the reference for Wright and Mermin [current reference 9].
+
 Supplement
@@
 
+RESPONSE: This text has been replaced by a statement which is
+transparent: "For both normal and planar anchoring at the walls,
+values of $W$ and $K$ are used which are the same as for the
+strongly anchoring particles."
+
 References of the Supplement:
 There are mistakes in references 1 and 2.
+
+REPSONSE: Reference 1, Wright and Mermin, was correct in
+the original supplement. Reference 2 has been updated to reflect
+the 2nd edition of de Gennes and Prost, which now appears in the
+main text [current reference 33].
 
@@
 
-** See NPG's author and referees' website at www.nature.com/authors for
-information about policies, services and author benefits
+* ADDITIONAL CHANGES TO MEET FORMAT REQUIREMENTS
 
-This email has been sent through the NPG Manuscript Tracking System
-NY-610A-NPG&MTS
+A number of other rearrangements have taken place to meet the
+format requirements as set out the manuscript checklist.
 
-Confidentiality Statement:
+Title
 
-This e-mail is confidential and subject to copyright. Any unauthorised use
-or disclosure of its contents is prohibited. If you have received this email
-in error please notify our Manuscript
-Tracking System Helpdesk team at http://npgapps.custhelp.com/app/ask.
+- The title has been adjusted to remove punctuation
 
-Details of the confidentiality and pre-publicity policy may be found here
-http://www.nature.com/authors/policies/confidentiality.html
+Authors
 
-Privacy Policy | Update Profile
+- The full postal address has been added
+- The corresponding author has been identified (Marenduzzo).
+
+Abstract
+
+- The abstract has been shorted to meet the 150 word limit.
+  Citations have been removed to the main text.
+
+Main text
+
+- Symbols and units have been adjusted in places to come into style.
+
+Results
+
+- Subheadings of < 60 characters have been added
+
+- References to supplementary items have been adjusted to come into style.
+
+Methods
+
+- All "methods" material (originally in supplementary information) is now
+contained within the main paper. This includes relevant references.
+
+- Dividing subheadings of < 60 characters have been added.
+
+References
+
+- Formatting is adjusted to come into style.
+
+End notes
+
+- A statement of author contributions has been added.
+
+- A statement on conflicts of interest has been added.
+
+Legends
+
+- Figure legends have been adjusted to include a brief title.
+
+Tables
+
+- A title has been included.

### 2014-02-21 papers/2012-sandwich/CholColloids.tex
First revision update
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2312 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %  The supporting material text is support.tex
+%
+%  Submitted version was SVN 2259.
 %
@@
 
-\title{Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
+\title{Self-assembly of colloid-cholesteric composites provides
+a route to switchable optical materials}
+
 \author{K. Stratford$^{*}$, O. Henrich$^{*}$, J.~S. Lintuvuori$^{*}$,\\
-M. E. Cates, D. Marenduzzo \\ 
-{\small SUPA, School of Physics and Astronomy, Edinburgh University,} \\
-\small{Edinburgh, EJ9 3JZ, UK}\\
-\small{$^{*}$ equally contributed to this work}}
+M. E. Cates, D. Marenduzzo$^\dagger$ \\ 
+SUPA, School of Physics and Astronomy, The King's Buildings,\\
+The University of Edinburgh, Edinburgh, EJ9 3JZ, UK\\
+\small{$^{*}$ equally contributed to this work}\\
+\small{$^\dagger$ Corresponding author: dmarendu@ph.ed.ac.uk}
+}
 \date{}
@@
 
-{\bf Colloidal particles dispersed in liquid crystals can form new materials with tunable elastic and electro-optic properties~\cite{stark}.
-In a host comprising  a periodic `blue phase' chiral liquid crystal~\cite{mermin}, the particles should template into colloidal crystals \cite{miha} with potential uses in photonics, metamaterials, and transformational optics~\cite{lavrentovich}.
-Here we show by computer simulation that colloid/cholesteric mixtures can give rise to a much wider range of soft-solid composites than previously anticipated.
-These include regular crystals, glasses, percolating gels, isolated clusters, twisted rings and undulating colloidal ropes.
-The final structure can be tuned via the particle concentration, and by varying the surface interactions of the cholesteric host with both the colloids and any confining walls.
-Remarkably, we find that many of these new materials are metastable: two or more structures can arise under identical thermodynamic conditions.
-The observed structure depends not only on the formulation protocol,
-but also on the history of an applied electric field.  
-This new class of soft materials should thus be relevant to the design of switchable, multistable devices for optical technologies including, for example, smart glass and e-paper~\cite{epaper}.}
+\noindent
 
+%% ABSTRACT 150 WORDS MAX NO REFERENCES
+\noindent
+{\bf
+Colloidal particles dispersed in liquid crystals can form new materials
+with tunable elastic and electro-optic properties. In a periodic `blue
+phase' host, particles should template into colloidal crystals with
+potential uses in photonics, metamaterials, and transformational optics.
+Here we show by computer simulation that colloid/cholesteric mixtures can
+give rise to regular crystals, glasses, percolating gels, isolated clusters,
+twisted rings and undulating colloidal ropes. This structure can be tuned
+via particle concentration, and by varying the surface interactions of the
+cholesteric host with both
+the particles and confining walls. Many of these new materials are metastable:
+two or more structures can arise under identical thermodynamic conditions.
+The observed structure depends not only on the formulation protocol, but also
+on the history of an applied electric field. This new class of soft materials
+should thus be relevant to design of switchable, multistable devices for
+optical technologies such as smart glass and e-paper.
+}
+%% EXACTLY 150
+
+\bigskip
+\noindent
+%% [Introduction: Max 1000 words; we have about 650]
+
+\noindent
 When spherical colloidal particles are mixed into a nematic liquid crystal, 
@@
 BPs arise because the director field can twist around more than one direction at once (see
-the cartoon in Fig. S1). Such double twist 
-regions (generally cylinders: DTCs) are energetically favoured at high enough molecular chirality, but
+the cartoon in Supplementary Fig.~1). Such double twist 
+regions (generally cylinders) are energetically favoured at high enough molecular chirality, but
 lead to geometric frustration because it is impossible to tile the whole 3D
-space with DTCs without also introducing ``disclination lines'' 
+space with double twist cylinders without also introducing ``disclination lines'' 
 (These are singular topological defects on which the director field is 
@@
 In conventional cholesteric materials~\cite{mermin} this balance is only achievable in a narrow temperature range of a few degrees, around the onset of liquid crystalline ordering.
-However, recent advances in formulation have widened this stability range enormously (up to 250K) paving the way to
-applications of BPs in display devices~\cite{kikuchi,bplasers,coleswidetrange,bpdevice,coles}.
+\textcolor{red}{
+However, recent advances in formulation have widened this stability
+range enormously~\cite{kikuchi,bplasers,coleswidetrange,bpdevice}. Most
+recently, templated BPs with a stability range of -125
+to~125$^\circ$C have been reported \cite{coles}, paving the way to
+applications of BPs in operational display devices.}
 
@@
 exclude the surrounding liquid crystal but otherwise have negligible surface 
-interaction with it (to give a so-called `weak anchoring' regime)~\cite{miha}. 
+interaction with it (to give a so-called `weak anchoring' regime)~\cite{extrareference1,miha}. 
 Such particles are attracted to the disclination network in BPs: by covering up the defect cores, the high elastic energy cost of those regions is avoided.
@@
 
-A second compelling reason to study colloid-BP composites is that, even without particles, BPs can show several competing metastable free energy minima,
+A second compelling reason to study colloidal-BP composites is that, even without particles, BPs can show several competing metastable free energy minima,
 corresponding to different topologies of the defect network~\cite{adriano,fukuda}, with a strong dependence on applied electric or magnetic fields~\cite{henrichfield}.
@@
 To explore such metastability, and switching strategies between states, we require multi-unit-cell, time-dependent simulations.
-These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}.
+These go beyond previous studies of colloidal-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloidal particles in cholesterics~\cite{juho1,juho2}.
 Our methodology and simulation parameters are summarized in the Supplementary Information.
 
+SHORT SUMMARY OF RESULTS AND CONCLUSIONS (PRESENT TENSE). To include a
+statement about confined systems with reference to \cite{extrareference2}.
+The simulations represent a liquid crystal
+blue phase with unit cell size in the range of 100--500~nm \cite{mermin},
+and particles with a radius of around 50~nm.
+
+\bigskip
+\noindent
+\textbf{\large Results} [Currently about 1400 words]
+
+\noindent
+\textbf{Structure in bulk BPI.}
 We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles at random within it.
-Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
-(This is defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface.) For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
+Among dimensionless control parameters are the particle solid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
+\textcolor{red}{
+The anchoring strength is defined such that $W > 0$ imparts a preferred
+orientation to the director field at the particle surface. This orientation
+can be either normal to the surface or in a plane tengential to the surface,
+a property related to surface chemistry in experimental systems but which
+can be imposed at will in simulations.
+We choose} $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
 This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.
@@
 For a single colloidal particle in a nematic fluid, $w$ controls the formation of a local topological defect at the particle surface~\cite{stark}, with weak distortion of the nearby fluid at small $w$.
-In the colloid-BP system, $w$ also plays a determining role in the final composite structure.
-When $w$ is small (Fig.~ 1A,B; Suppl.
-Movie 1,2), the attraction of particles to the disclination lattice causes
+In the colloidal-BP system, $w$ also plays a determining role in the final composite structure.
+When $w$ is small (Fig.~ 1A,B; Supplementary Movies~1 and~2), the attraction of particles to the disclination lattice causes
  the lattice to become covered by particles, with little disruption of its
@@
 Thus a templated colloidal crystal is created, as first reported in~\cite{miha}.
-Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloids which leads to very slow dynamics (Fig.
-S4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
+Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloidal particles which leads to very slow dynamics
+(Supplementary Fig.~4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
 1A).
@@
 
-For large $w$ (Fig.~1C,D; Suppl. Movie 3,4), the picture changes dramatically.
+For large $w$ (Fig.~1C,D; Supplementary Movies~3 and~4), the picture changes
+dramatically.
 Each colloidal particle now disrupts significantly the order of the nearby fluid, with defect formation close to its surface~\cite{stark}.
@@
 
-As colloid-nematic composites~\cite{tiffany}, all the structures in Fig.~1A-D 
+As colloidal-nematic composites~\cite{tiffany}, all the structures in
+Fig.~1A-D 
 should be soft solids, of nonzero elastic 
@@
 structure is now akin to that of the self-quenched glass found in 
-colloid-nematic composites~\cite{tiffany} at much higher volume 
+colloidal-nematic composites~\cite{tiffany} at much higher volume 
 fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa.
@@
 the self-assembly.
-At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloid strands (the gel in Fig.~1H) is again seen.
+At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloidal strands (the gel in Fig.~1H) is again seen.
 The structures in Fig. 1A-H have been followed for $\sim$20 ms 
-(see Fig. S4); they correspond to either very slowly evolving states or 
-deep metastable minima (elastic interactions are of
+(see Supplementary Fig.~4); they correspond to either very slowly evolving
+states or  deep metastable minima (elastic interactions are of
 order $KR$ and dwarf thermal motion, see
-Fig. S4 and Ref.~\cite{stark,lavrentovich}).
+Supplementary Fig.~4 and Ref.~\cite{stark,lavrentovich}).
 
+\medskip
+\noindent
+\textbf{Structures in confined BPI.}
 In display devices and other optical applications, it is typical for liquid crystals to be strongly confined (and subject to strong boundary effects), in contrast to the bulk geometries considered above.
@@
 We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supplementary Information).
-The resulting colloid-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
+The resulting colloidal-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
 There are, however, significant effects of the confining walls.
-First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls 
+First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloid particless at both walls 
 (see side-views in Fig.~2A-D). Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit. These clusters
-are similar to those reported experimentally for large colloids in a
+are similar to those reported experimentally for large colloid particles in a
 cholesteric (but non-BP) host~\cite{niek}.
@@
 Fig.~2E-H shows comparable simulations for planar wall anchoring. 
-Without colloids, planar anchoring results in a regular cholesteric helix,
+Without colloidal particles, planar anchoring results in a regular
+cholesteric helix,
 with axis perpendicular to the boundary walls. 
@@
 double-stranded structures at larger $\phi$ (Fig.~2F).
-For larger $w$, we observe thick colloidal ``ropes" about 5 colloids across
+For larger $w$, we observe thick colloidal ``ropes" about 5 particles across
 (Fig.~2G,H). These ropes undulate to follow the chiral structure of the 
@@
 On increasing the density they develop branch points and finally again approach a percolating gel morphology.
-Note that planar wall anchoring is incompatible with the normal anchoring on the colloids (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
+Note that planar wall anchoring is incompatible with the normal anchoring on the colloidal particles (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
 
+\medskip
+\noindent
+\textbf{Switching in electric field.}
 An interesting question, relevant for applications to devices, is whether 
@@
 Fig.~3 addresses this question, starting from amorphous Blue Phase III,
-this time at a higher chirality. Dispersing colloids in this networks
+this time at a higher chirality. Dispersing colloidal particles in this networks
 results in the particle occupying the available nodes, where the order
-parameter is low (Fig.~3A; Suppl. Movie 5). Applying a high enough field 
+parameter is low (Fig.~3A; Supplementary Movie~5). Applying a high enough
+field 
 along the $x$ direction leads to a switching to a new phase, where colloidal 
 arrays fit within a network with distorted hexagonal cells (Fig.~3B 
-and Fig. S7; Suppl. Movie 6). 
+and Supplementary Fig.~7; Supplementary Movie~6). 
 Remarkably, upon field removal this honeycomb structure does not find its 
@@
 configuration before switching on the field, but gets stuck into a metastable 
-structure (Fig. S7), with residual anisotropy along the field switching 
+structure (Supplementary Fig.~7), with residual anisotropy along the field
+switching 
 direction. Cycling the field along the $x$ and $y$ directions leads to
 distinct metastable structures, which are retained after field removal
-(Fig. S8 and S9).
-%The resulting multistability is a prerequisite for new generation 
-%switchable electrooptic devices for use in energy-saving electronics-free
-%appliances, smart glass, and e-paper to cite just a few examples.
+(Suplementary Fig.~8 and Fig.~9).
 
@@
 
-The surprising wealth of colloid-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
+\medskip
+\noindent
+\textbf{\large Discussion} [Currently about 260 words]
+
+\noindent
+The surprising wealth of colloidal-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
 First, blue phases provide a three-dimensional template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}.
@@
 For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating gels, or helical colloidal ropes.
-In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls.
+In both cases, control is offered by the process route, colloidal concentration, and the anchoring conditions at sample walls.
 We have also shown that electric fields can be used to switch between metastable states, providing a possible route to future multistable device applications.
-%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
+%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloid particles and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
 We have not explored in this work switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case the same materials may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
@@
 \bigskip
-\textbf{Acknowledgements}
+\noindent
+\textbf{\large Methods}
+
+INTRODUCTORY STATEMENT
+
+\noindent
+\textbf{Theromdynamics.}
+The thermodynamics of the cholesteric liquid crystal can be described
+by means of a  Landau-de Gennes free energy functional $\cal F$,
+whose density is written as $f$:
+\begin{eqnarray}
+{\cal F}[{\bf Q}]&=&\int d^3{\bf r} f({\bf Q}({\bf r})).
+\end{eqnarray}
+This free energy density $f = f({\bf Q})$ may be expanded in powers of the
+order parameter ${\bf Q}$ and its gradients; ${\bf Q}$ is a traceless 
+and symmetric tensor which is denoted from now on in subscript
+notation as $Q_{\alpha\beta}$.
+The largest eigenvalue $q$ and corresponding eigenvector $n_\alpha$
+of $Q_{\alpha\beta}$ describe the local strength and major orientation axis
+of molecular order.
+The theory based on the tensor $Q_{\alpha\beta}$, rather than a theory based
+solely
+on the director field $n_\alpha ({\bf r})$, allows treatment of disclinations
+(defect lines) in whose cores $n_\alpha$ is undefined.
+
+Explicitly, the free energy density is:
+\begin{eqnarray}
+f(Q_{\alpha\beta}) &=& {\textstyle \frac{1}{2}} A_0
+ \left(1- {\textstyle \frac{1}{3}}\gamma \right)Q^2_{\alpha \beta}
+-{\textstyle \frac{1}{3}}A_0\gamma Q_{\alpha \beta} Q_{\beta \gamma}Q_{\gamma \alpha} 
++ {\textstyle \frac{1}{4}} A_0\gamma (Q^2_{\alpha \beta})^2  \nonumber\\
+&+& {\textstyle \frac{1}{2}} K(\varepsilon_{\alpha \gamma \delta}
+\partial_\gamma Q_{\delta \beta} + 2 q_0 Q_{\alpha \beta})^2
++ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2.
+\label{free}
+\end{eqnarray}
+Here, repeated indices are summed over, while terms of the form
+$Q^2_{\alpha\beta}$ should be expanded to read $Q_{\alpha\beta}Q_{\alpha\beta}$;
+$\varepsilon_{\alpha\gamma\delta}$ is the permutation tensor.
+The first three terms are a bulk free energy density whose overall scale
+is set by $A_0$ (discussed further below); $\gamma$ is a control parameter
+related to a reduced temperature. Varying the latter in the absence of
+chiral terms ($q_0=0$) gives an isotropic-nematic transition at
+$\gamma = 2.7$ with a mean-field spinodal instability at $\gamma = 3$.
+
+The remaining two terms of the free energy density in Eq.~\ref{free} describe
+distortions of the order parameter field. In theoretical work and when
+describing a generic rather than a specific system, it is 
+conventional~\cite{mermin,deGennes} to
+assume that splay, bend and twist deformations of the director are equally
+costly, that is, there is a single elastic constant $K$. The parameter $q_0$
+is related to the helical pitch length $p$ via $q_0=2\pi/p$,
+describing one full turn of the director in the cholesteric phase.
+
+There are two dimensionless numbers, which are commonly referred to as
+$\kappa$, the chirality, and  $\tau$, the reduced temperature
+\cite{mermin} which can be used to characterise the system. In terms of
+the parameters in the free energy density, these are:
+\begin{eqnarray}\label{cntrl-param} 
+\tau&=&\frac{27(1-\gamma/3)}{\gamma}\label{tau}\\
+\kappa&=&\sqrt{\frac{108\ K\, q_0^2}{A_0\, \gamma}}\label{kappa}.
+\end{eqnarray}
+If the free energy density Eq.~\ref{free} is made dimensionless, $\tau$
+appears as prefactor of the term  quadratic in $Q_{\alpha\beta}$,
+whereas $\kappa$
+quantifies the ratio between bulk and  gradient free energy terms. The
+chirality and reduced temperature may be used to characterise an equilibrium
+phase diagram of the blue phases in bulk cholesteric liquid crystals
+\cite{mermin,henrichfield}.
+
+The effect of an electric field can be modelled by including the following
+term in the bulk free energy density:
+\begin{equation}
+-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
+\end{equation} 
+where $E_{\alpha}$ are the components of the electric field, and 
+$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
+crystal. The strength of the electric field is quantified via
+one further dimensionless number
+\begin{equation}
+{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
+\end{equation}
+The quantity ${\cal E}$ will be referred to as the reduced field strength.
+
+\medskip
+\noindent
+\textbf{Surface free energy.}
+In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
+free energy density is also present to represent the energetic cost of
+anchoring at a solid surface. If the preferred order parameter tensor
+at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
+density (per unit area) is
+\begin{equation}
+f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
+= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
+\end{equation}
+where $W$ is a constant determining the strength of the anchoring.
+The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
+discussed below. For a colloidal particle of radius $R$, the strength of
+the surface
+anchoring compared which the bulk fluid elastic constant may be quantified
+by the dimensionless parameter $WR/K$. For small values of this parameter,
+the presence of a particle surface should have little impact on the local
+fluid LC ordering.
+
+\medskip
+\noindent
+\textbf{Dynamics of the order parameter.}
+A framework for the dynamics of liquid crystals is provided by the 
+Beris-Edwards model \cite{beris}, in which the time evolution of the
+tensor order parameter obeys
+\begin{equation}
+\label{eom}
+\left(\partial_t+ u_\nu \partial_\nu \right) Q_{\alpha\beta} - S_{\alpha\beta}
+= \Gamma H_{\alpha\beta}.
+\end{equation}
+In the absence of flow, Eq.~\ref{eom} describes a relaxation towards
+equilibrium on a timescale determined by a collective rotational diffusion 
+constant $\Gamma$. This relaxation is driven by the molecular field
+$H_{\alpha\beta}$, which is the functional derivative of the free energy
+with respect to the order parameter \cite{beris}:
+\begin{equation}
+H_{\alpha\beta} = -\frac{\delta {\cal F}} {\delta Q_{\alpha\beta}} 
++ {\textstyle \frac{1}{3}} \delta_{\alpha\beta} 
+\mathrm{Tr}\left(\frac{\delta {\cal F}}{\delta Q_{\alpha\beta}}\right).
+\end{equation}
+
+The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
+derivative for rod-like molecules \cite{beris}. It couples the order
+parameter to the symmetric and antisymmetric parts of the velocity 
+gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
+The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
+$\Omega_{\alpha\beta}$ are defined as 
+\begin{equation}
+A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
+\end{equation}
+and
+\begin{equation}
+\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
+\end{equation}
+This full coupling term is then
+\begin{eqnarray}
+\label{coupling-term}
+S_{\alpha\beta}  &=&
+(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
+(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
++
+(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
+(\xi A_{\nu\beta} - \Omega_{\nu\beta})
+\nonumber\\ 
+&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
+Q_{\nu\nu} W_{\nu\nu}.
+\end{eqnarray}
+Here, $\xi$ is a material-dependent that controls the
+relative importance of rotational and elongational flow for molecular
+alignment, and determines in practice how the orientational order
+responds to a local shear flow. 
+In all cases, the value used here is $\xi = 0.7$, which is within
+the `flow aligning' regime, where molecules align at a fixed angle
+(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
+The value of the collective rotational diffusion used in all simulations is
+$\Gamma = 0.5$ in simulation units.
+
+\medskip
+\noindent
+\textbf{Hydrodynamics.}
+The momentum evolution obeys a Navier-Stokes equation driven by the divergence
+of a generalised stress $P_{\alpha\beta}$:
+\begin{equation}
+\label{nse}
+\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
+=\partial_\beta P_{\alpha\beta}.
+\end{equation}
+The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
+both viscous and thermodynamic components:
+\begin{eqnarray}
+P_{\alpha \beta}&=&
+- p_0 \delta_{\alpha \beta} 
++ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
+\nonumber\\
+&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
++ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
+-\xi \left(Q_{\alpha \gamma}
++ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
+\nonumber\\ 
+&+& 2 \xi  \left(Q_{\alpha \beta}
++ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
+-\partial_\alpha Q_{\gamma \nu}
+\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
+&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
+\end{eqnarray}
+A lattice Boltzmann (LB) flow solver is used, where the isotropic pressure
+$p_0$
+and viscous terms are managed directly by the solver (as in a simple Newtonian
+fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
+is treated as a local force density on the fluid. In all simulations
+reported the mean fluid density $\rho = 1$ and the viscosity
+$\eta = 0.01$ (simulations in confined geometry also used enhanced
+bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).
+
+\medskip
+\noindent
+\textbf{Surface boundary conditions for particles.}
+Hydrodynamic boundary conditions for solid objects are handled via
+the lattice Boltzmann solver. In particular, the standard method of
+bounce-back on links is used for both particles and confining walls
+\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
+are set out in \cite{juho}.
+
+In all the simulations reported here the preferred orientation of the
+director field at the particle surface is normal to the surface. The
+required nematic director at the particle surface in a given location
+$\hat{n}^0_\alpha$ may then be determined from geometry alone,
+and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
+\begin{equation}
+Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
+- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
+\label{eq:surf_q}
+\end{equation}
+The magnitude of surface order $q^0$ is set by
+\begin{equation}
+q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
++ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
+\label{eq:surf_q0}
+\end{equation}
+which corresponds to the bulk order in a defect-free nematically ordered 
+sample~\cite{denniston} (this is also very close to the value of the order 
+parameter in a cholesteric or in a blue phase away from disclinations).
+
+In all simulations the colloidal (hard sphere) radius is $R = 2.3$. To
+prevent particles overlapping at the hard sphere radius, an additional
+short range soft potential is included. This takes the form of
+$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
+The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
+units. Both potential and resulting force are smoothly matched to zero at
+a cut-off distance of $h_c = 0.25$ simulation units.
+
+\medskip
+\noindent
+\textbf{Surface boundary conditions for walls.}
+The confining solid walls, which are flat and stationary, have the
+same normal boundary condition as at the particle surface. In addition,
+degenerate planar anchoring is applied at the walls, where the
+preferred nematic director may adopt any orientation parallel
+to the surface. The preferred direction is determined explicitly
+by projecting the local fluid order parameter to the plane of the
+wall \cite{fournier2005}. Eq.~\ref{eq:surf_q} and Eq.~\ref{eq:surf_q0}
+are then used as before.
+
+To prevent the particles being forced into the wall, a correction
+to the lubrication force on the colloidal particle is added for very low
+particle
+surface to wall surface separations $h < 0.5$ lattice spacing. The
+correction is based on the analytic expression for the lubrication
+between a sphere and a flat surface.
+
+\medskip
+\noindent
+\textbf{Protocol and parameter details for bulk BPI.}
+The parameters for the free energy are chosen to be representative
+of blue phase I: chirality $\kappa \sim 0.7$ and reduced temperature
+$\tau = -0.25$.
+The full free energy parameters are shown in Table~\ref{table:params}.
+The order parameter tensor $Q_{\alpha\beta}$ for bulk blue phase I is
+initialised from an approximation in high chirality limit
+\cite{mermin,henrichfield}. Systems of 128$^3$ lattice sites are
+used with a pitch length of $p = 32\sqrt{2}$, which accommodates 4$^3$
+unit cells with fully periodic boundary conditions.
+Simulations at different solid volume fractions (1\% and 4\%), and different 
+surface anchoring strengths ($WR/K = 0.23$ and $WR/K = 2.3$) representing
+``weak'' and ``strong'' anchoring are run for two million simulation
+time steps.
+The initial distribution of the particle positions is set at random
+for the required solid volume fraction; the particles are initially
+at rest.
+
+\medskip
+\noindent
+\textbf{Parameter details for confined BPI.}
+Here, a narrow sandwich of fluid is placed between flat walls in
+perpendicular to the narrow coordinate direction, and with periodic
+boundaries in the other two directions. The system size used in all
+cases is 256$^2 \times$56 lattice sites. Each simulation is a ``quench''
+in a similar manner to that used in the bulk: the order parameter is
+initialised randomly. The free energy parameters (see Table~\ref{table:params})
+are again appropriate for (equilibrium) blue phase I
+(chirality $\kappa = 0.8$ and
+reduced temperature $\tau = -0.25$). The cholesteric pitch is set
+to be $p = 64$, providing a slightly better resolution than the bulk
+simulations.
+
+The surface anchoring for colloidal particles is as before: always normal,
+but with
+$WR/K = 0.23$ and $WR/K = 2.3$. The solid volume fractions are
+1\% and 4\%.
+\textcolor{red}{For both normal and planar anchoring at the walls,
+values of $W$ and $K$ are used which are the same as for the
+strongly anchoring particles.}
+All simulations are run for two million simulation
+time steps.
+
+\medskip
+\noindent
+\textbf{Parameters for external electric field.}
+For the simulations of BPIII with colloidal particles a system size of
+$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
+The simulations are
+performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
+An initial configuration of
+randomly oriented and positioned double twist cylinders is used,
+from which an amorphous BPIII network emerges \cite{bp3}. This structure
+is equilibrated for $6\times10^{5}$ LB time steps until no significant
+further evolution is observed.
+
+The colloidal particles with weak
+normal anchoring ($WR / K= 0.23$), are then inserted with randomly chosen positions,
+and the composite system is equilibrated for another $1.2\times10^{6}$
+time steps.
+At the end of the equilibration phase the uniform electric field
+with reduced field strength ${\cal E} = 0.8$ 
+is switched on (oriented along the $z$-direction), and the simulation
+run for a further $1.2\times10^{6}$ time steps.
+This is found to be sufficiently long for a hexagonal structure to
+emerge and saturate.
+If the electric field is then switched off, and the system quickly relaxes
+to a metastable state which shows a residual anisotropy in the
+colloidal distribution. The total length of this final relaxation phase
+is $1\times10^6$ LB time steps.
+
+\medskip
+\noindent
+\textbf{Disclination rendering.}
+In all cases visualisation of the defect lines is carried out via
+an isosurface of the scalar order parameter $q$ determined from the
+largest eigenvalue of the local order parameter tensor. A low value
+of the scalar order parameter $q$ (typically in the range 0.12--0.14)
+unambiguously identifies the disclinations.
+
+\medskip
+\noindent
+\textbf{Computation of the structure factor.}
+To provide structural information about the field-aligned 
+states and the residual anisotropy after switch off a 
+procedure similar to the one described in \cite{bp3} is
+followed.
+The structure factor $S(\mathbf{k})$ in the current approach is
+defined via the Fourier transform of the particle density $F(\mathbf{k})$:
+
+\begin{equation}
+S({\bf k}) =  |F({\bf k})|^2, \mathrm{with}\
+F({\bf k}) = \int d^3{\bf r} \rho({\bf r}) e^{i {\bf k. r}}.
+\end{equation}
+The colloidal density $\rho({\bf r})$ is a discretised density and taken to be
+$\rho({\bf r})=1$ if a lattice site at ${\bf r}$ is part of a solid particle
+and $\rho({\bf r})=0$ otherwise. This definition, which deviates from the 
+standard definition of a structure factor and does not consider the particles 
+as point-like objects, allows the same structural information
+to be obtained in a simple way.
+
+\medskip
+\noindent
+\textbf{Parameter mapping to physical units.}
+To get from simulation to physical units, a calibration of scales of
+length, energy and time in the simulations is required. For similar
+mappings see, e.g.~\cite{denniston2}.
+
+The length scale can be set by mapping the half pitch to a realistic blue
+phase unit cell size, which is in the 100--500~nm range \cite{mermin}. 
+For bulk simulations (Fig.~1 and Fig.~3 in the main text), a suitable choice
+is one where one simulation unit (lattice site) corresponds to 10~nm. Therefore
+the colloidal size in Figs.~1 and~3 corresponds to $\sim$ 50~nm.
+
+To obtain an energy scale, it is possible to choose $A_0 \simeq 10^6$~Pa,
+which is reasonable following Ref.~\cite{mermin} (this is also the choice in
+Ref.~\cite{domaingrowth}). This choice leads to a simulation unit of
+free energy (or stress) equal to $10^{8}$~Pa in Figs.~1 and~3.
+From the energy and length scales one can map elastic constants to
+physical units: for instance, the simulation in Fig.~1 corresponds to
+a liquid crystal with splay, bend and twist (Frank) elastic constants
+equal to 35 pN (or equivalently $K=70$~pN).
+
+The timescale calibration may be obtained from (see e.g.~\cite{denniston})
+\begin{equation}
+\gamma_1=\frac{2q^2}{\Gamma},
+\end{equation}
+which relates the rotational viscosity $\gamma_1$ to the ordering strength
+$q$ and the order parameter mobility $\Gamma$.
+In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
+for the simulation in Fig.~1 leads to $q=1/2$.   
+For real liquid crystalline materials, $\gamma_1$ usually lies in range
+$10^{-2}$--$1$~Pa~s~\cite{deGennes}; for definiteness say
+$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
+free energy (or stress) units, this leads to a simulation unit of time equal
+to $10^{-8}$~s. The total simulated time in Figs.~1--3 is therefore in the
+millisecond range.
+
+To calibrate the electric field strength 
+one can equate the value of the dimensionless parameter ${\cal E}$ in the 
+simulations with that of a hypothetical experiment.
+Given the previous value for $A_0$, and a dielectric anisotropy of 20
+(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
+dielectric permittivity of vacuum), an electric field
+of 100 V~$\mu$m$^{-1}$ is obtained. This corresponds to a value of
+${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
+Fig.~3 of the main text).
+
+Given these mappings, all quantities can be easily transformed from
+simulation to physical units and vice versa. For example Figs.~1 and~2
+use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
+which translate into $0.01$--$0.1$~Pa~s respectively. This is sensible (if 
+slightly low) for a molecular nematogen in the isotropic phase. 
+The effective viscosity in ordered phases is higher, but this is ensured
+in our model by the coupling to the order parameter.
+Note that the density is set to unity in the simulations: this
+corresponds to a fluid density which is about a thousand times larger
+than in experiments. This makes no difference in practice as
+the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
+velocity, in our simulations around $10^{-5}$ in lattice units, 
+and $\Lambda$ is a typical length scale, e.g. the size of a unit cell) 
+remains small enough~\cite{codef}.
+
+To summarise, the simulations represent BP-forming materials, with the
+interpretation of the simulation units for length, time, and energy
+density being close to 10~nm, 10~ns, and 100~MPa respectively.
+
+\pagebreak
+
+\begin{thebibliography}{99}
+
+\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj,
+D. Babic, N. Osterman, I. Musevic,
+Entangled nematic colloidal dimers and wires, 
+{\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
+
+\bibitem{zumer}
+I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik, S. Zumer,
+Two-dimensional nematic colloidal crystals self-assembled by topological
+defects,
+{\it Science} {\bf 313}, 954--958 (2006).
+
+\bibitem{tanaka} T. Araki, H. Tanaka,
+Colloidal aggregation in a nematic liquid crystal: Topological arrest of
+particles by a single-stroke disclination line,
+{\it Phys. Rev. Lett.} {\bf 97}, 127801 (2006).
+
+\bibitem{tiffany}
+T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo, W.~C.~K. Poon,
+A Self-Quenched Defect Glass in a Colloid-Nematic Liquid Crystal Composite,
+{\it Science} {\bf 333}, 79--83 (2011).
+
+\bibitem{stark} H. Stark,
+Physics of colloidal dispersions in nematic liquid crystals,
+{\it Phys. Rep.} {\bf 351}, 387--474 (2001).
+
+\bibitem{abbott}
+I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo,
+N.~L. Abbott,
+Endotoxin-Induced Structural Transformations in Liquid Crystalline Droplet,
+{\it Science} {\bf 332}, 1297--1300 (2011).
+
+\bibitem{colloiddevice}
+K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton, R. G. Capurso,
+Single-substrate cholesteric liquid crystal displays by colloidal
+self-assembly,
+{\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
+
+\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka,
+Memory and topological frustration in nematic liquid crystals confined
+in porous materials, {\it Nat. Mat.} {\bf 10}, 303--309 (2011).
+
+\bibitem{mermin} D.~C. Wright, Mermin, N.~D.,
+Crystalline liquids: The blue phases, {\it Rev. Mod. Phys.} {\bf 61},
+\textcolor{red}{385--432} (1989).
+
+\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
+ Structure of Blue Phase III of Cholesteric Liquid Crystals,
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
+
+\bibitem{kikuchi}
+H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama,
+Polymer-stabilized liquid crystal blue phases,
+{\it Nat. Mat.} {\bf 1}, 64--68 (2002).
+
+\bibitem{bplasers}
+W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri,
+Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
+phase II,
+{\it Nat. Mat.} {\bf 1}, 111--113 (2002).
+
+\bibitem{coleswidetrange}
+H.~J. Coles, M.~N. Pivnenko,
+Liquid crystal 'blue phases' with a wide temperature range,
+{\it Nature} {\bf 436}, 997--1000 (2005).
+
+\bibitem{bpdevice}
+Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
+ Electro-optics of polymer-stabilized blue phase liquid crystal displays,
+{\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
+
+\bibitem{coles}
+F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim,
+S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles,
+Blue-phase templated fabrication of three-dimensional nanostructures for
+photonic applications, {\it Nat. Mat.} {\bf 11}, 599--603 (2012).
+
+\bibitem{extrareference1}
+\textcolor{red}{M. Ravnik, G.P. Alexander, J.M. Yeomans, S. Zumer,
+Mesoscopic modelling of colloids in chiral nematics,
+{\it Faraday Discuss.} {\bf 144}, 159--169 (2010).}
+
+\bibitem{miha}
+M. Ravnik, G.~P. Alexander, J.~M. Yeomans, S. Zumer,
+Three-dimensional colloidal crystals in liquid crystalline 
+blue phases, {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188--5192 (2011).
+
+\bibitem{lavrentovich} O.~D. Lavrentovich,
+ Liquid crystals, photonic crystals, metamaterials and transformation optics, 
+{\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5143--5144 (2011).
+
+\bibitem{adriano}
+A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini, F. Salvadore,
+Bistable Defect Structures In Blue Phase Devices,
+ {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
+
+\bibitem{fukuda}
+J.-I. Fukuda, S. Zumer,
+Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal,
+{\it Nat. Commun.} {\bf 2}, doi:10.1038/ncomms1250 (2011).
+
+\bibitem{henrichfield} 
+O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates,
+Thermodynamics of blue phases in electric fields,
+{\it Phys. Rev. E} {\bf 81}, 031706 (2010).
+
+\bibitem{epaper} J. Heikenfeld, P. Drzaic,  J.~S. Yeo, T. Koch,
+A critical review of the present and future prospects for electronic paper,
+{\it J. Soc. Inf. Display} {\bf 19}, 129--156 (2011).
+
+\bibitem{domaingrowth}
+O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates,
+Ordering dynamics of blue phases entails kinetic stabilization of amorphous
+networks, {\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212--13215 (2010).
+
+\bibitem{juho1}
+J. S. Lintuvuori, K. Stratford, M. E. Cates, D. Marenduzzo,
+Colloids in cholesterics: size-dependent defects and non-Stokesian
+microrheology, {\it Phys. Rev. Lett.} {\bf 105}, 178302 (2010).
+
+\bibitem{juho2}
+J. S. Lintuvuori, K. Stratford, M. E. Cates, D. Marenduzzo,
+Self-assembly and nonlinear dynamics of dimeric colloidal rotors in 
+cholesterics, {\it Phys. Rev. Lett.} {\bf 107}, 267802 (2011).
+
+\bibitem{extrareference2}
+\textcolor{red}{M. Ravnik, J. Fukuda, J.M. Yeomans, S. Zumer,
+Confining blue phase colloids to thin layers,
+{\it Soft Matter} {\bf 7}, 10144--10150 (2011).}
+
+\bibitem{permeation1} 	W. Helfrich,
+Capillary flow of cholesteric and smectic liquid crystals,
+{\it Phys. Rev. Lett.} {\bf 23}, 372--374 (1969).
+
+\bibitem{permeation2}
+A. Dupuis, D. Marenduzzo, E. Orlandini, J.~M. Yeomans,
+Rheology of cholesteric blue phases,
+{\it Phys. Rev. Lett.} {\bf 95}, 097801 (2005).
+
+\bibitem{lubensky}
+M. Zapotocky, L. Ramos, P. Poulin, T.~C. Lubensky,  D.~A. Weitz,
+Particle-stabilized defect gel in cholesteric liquid crystals,
+{\it Science} {\bf 283}, 209--212 (1999).
+
+\bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
+Self-organization of particles with planar surface anchoring in a cholesteric
+liquid crystal, {\it Langmuir} {\bf 26}, 13502--13510 (2010).
+
+\bibitem{flowswitch}
+O. Henrich, K. Stratford, P.V. Coveney, M.E. Cates, D. Marenduzzo,
+Rheology of cubic blue phases,
+{\it Soft Matter} doi:10.1039/C3SM50228G (2013).
+
+\bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang,
+Developing optofluidic technology through the fusion of microfluidics and
+optics {\it Nature} {\bf 442}, 381--386 (2006).
+
+\bibitem{deGennes}
+\textcolor{red}{P.-G. de Gennes, J. Prost,
+{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).}
+
+\bibitem{beris}
+A. N. Beris, B. J. Edwards, 
+{\it Thermodynamics of Flowing Systems with Internal Microstructure}
+(Oxford University Press, 1994).
+
+\bibitem{ladd94}
+A.J.C. Ladd,
+\textit{J. Fluid Mech.} Numerical simulation of particulate suspensions via a
+discretized Boltzmann equation. Part 1. Theoretical foundation,
+\textbf{271}, 285--309 (1994).
+
+\bibitem{nguyen2002}
+N.-Q. Nguyen and A.J.C. Ladd,
+Lubrication corrections for lattice Boltzmann simulations of particle
+suspensions,
+\textit{Phys. Rev. E} \textbf{66} 046708 (2002).
+
+\bibitem{fournier2005}
+J.-B. Fournier and P. Galatola,
+Modeling planar degenerate wetting and anchoring in nematic liquid
+crystals,
+\textit{Europhys. Lett.} \textbf{72}, 403--409 (2005).
+
+\bibitem{juho}
+J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
+Colloids in liquid crystals: a lattice Boltzmann study,
+\textit{J. Mater. Chem.} \textbf{20}, 10547--10552 (2010).
+
+\bibitem{denniston}
+C. Denniston, E. Orlandini, J. M. Yeomans,
+Lattice Boltzmann simulations of liquid crystal hydrodynamics,
+\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
+
+\bibitem{denniston2}
+C. Denniston, D. Marenduzzo, E. Orlandini, J.M. Yeomans,
+Lattice Boltzmann algorithm for three-dimensional liquid-crystal
+hydrodynamics,
+\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745--1754 (2004).
+
+\bibitem{codef} M. E. Cates, K. Stratford, R. Adhikari, P. Stansell,
+J.-C. Desplat, I. Pagonabarraga, A.J. Wagner,
+Simulating colloid hydrodynamics with lattice Boltzmann methods, 
+\textit{J. Phys. Condens. Matter} \textbf{16},  3903--3915 (2004).
+
+\end{thebibliography}
+\noindent
+\textbf{\large Acknowledgements}
+
+\noindent
 We acknowledge support by UK EPSRC grants EP/J007404/1 (KS and JSL), 
@@
 
-\begin{thebibliography}{99}
-\bibitem{stark} H. Stark. Physics of
-colloidal dispersions in nematic liquid crystals. 
-{\it Phys. Rep.} {\bf 351}, 387-474 (2001).
-\bibitem{mermin} D.~C. Wright, Mermin, N.~D. Crystalline liquids: The blue phases. {\it Rev. Mod. Phys.} {\bf 61}, 385–432 (1989).
-\bibitem{miha} M. Ravnik, G.~P. Alexander, J.~M. Yeomans,  
-S. Zumer.  Three-dimensional colloidal crystals in liquid crystalline 
-blue phases. {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188-5192 (2011).
-\bibitem{lavrentovich} O.~D. Lavrentovich.
- Liquid crystals, photonic crystals, metamaterials and transformation optics. 
-{\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5143-5144 (2011).
-\bibitem{epaper} J. Heikenfeld, P. Drzaic,  J.~S. Yeo, T. Koch.
-A critical review of the present and future prospects for electronic paper.
-{\it J. Soc. Inf. Display} {\bf 19}, 129-156 (2011).
-\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj, D. Babic, N. Osterman, I. Musevic.  Entangled nematic colloidal dimers and wires. 
-{\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
-\bibitem{zumer} I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik,
-S. Zumer. Two-dimensional nematic colloidal crystals self-assembled by topological defects. {\it Science} {\bf 313}, 954-958 (2006).
-\bibitem{tanaka} T.
-Araki, H.
-Tanaka.
-Colloidal aggregation in a nematic liquid crystal: Topological arrest of particles by a single-stroke disclination line.
-{\it Phys.
-Rev.
-Lett.} {\bf 97}, 127801 (2006).
-\bibitem{tiffany} T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo,
-W.~C.~K. Poon. A Self-Quenched Defect Glass in a Colloid-Nematic Liquid Crystal Composite. {\it Science} {\bf 333}, 79-83 (2011).
-\bibitem{abbott} I.~H.
-Lin, D.~S.
-Miller, P.~J.
-Bertics, C.~J.
-Murphy, J.~J.
-de Pablo, N.~L.
-Abbott.
-Endotoxin-Induced Structural Transformations in Liquid Crystalline Droplet.
-{\it Science} {\bf 332}, 1297-1300 (2011).
-\bibitem{colloiddevice} K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton,
-R. G. Capurso. Single-substrate cholesteric liquid crystal displays by colloidal self-assembly. {\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
-\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
-Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
-%\bibitem{electrophoresis}  O. D. Lavrentovich,	I. Lazo, O. P. Pishnyak.
-%Nonlinear electrophoresis of dielectric and metal spheres in a nematic liquid 
-%crystal. {\it Nature} {\bf 467}, 947-950 (2010).
-%D.-K. Yang, X.-Y. Huang, Y.-M. Zhu. 
-%Bistable cholesteric reflective displays: Materials and Drive Schemes,
-%{\it Annu. Rev. Mater. Sci.} {\bf 27}, 117-146 (1997).
-%\bibitem{crooker} D. K. Yang and P. P. Crooker, {\it Phys. Rev. A}
-%{\bf 35}, 4419 (1987).
-\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo.
- Structure of Blue Phase III of Cholesteric Liquid Crystals.
-{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
-\bibitem{kikuchi} H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama.
-Polymer-stabilized liquid crystal blue phases.
-{\it Nat. Mat.} {\bf 1}, 64-68 (2002).
-\bibitem{bplasers} W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri.
-Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
-phase II. {\it Nat. Mat.} {\bf 1}, 111-113 (2002).
-\bibitem{coleswidetrange} H.~J. Coles, M.~N. Pivnenko.
-Liquid crystal 'blue phases' with a wide temperature range.
-{\it Nature} {\bf 436}, 997-1000 (2005).
-\bibitem{bpdevice} Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu.
- Electro-optics of polymer-stabilized blue phase liquid crystal displays.
-{\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
-\bibitem{coles} F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim, S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles.
-Blue-phase templated fabrication of three-dimensional nanostructures for photonic applications. {\it Nat. Mat.} {\bf 11}, 599-603 (2012).
-\bibitem{adriano} A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini,
-F. Salvadore. Bistable Defect Structures In Blue Phase Devices. 
- {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
-\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
-Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
-{\it Nat. Comm.} {\bf 2}, 246 (2011).
-\bibitem{henrichfield}  O. Henrich, D. Marenduzzo, K. Stratford, 
-M. E. Cates. Thermodynamics of blue phases in electric fields.
-{\it Phys. Rev. E} {\bf 81}, 031706 (2010).
-\bibitem{domaingrowth} O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates.
- Ordering dynamics of blue phases entails kinetic stabilization of amorphous networks. {\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212-13215 (2010).
-\bibitem{juho1} J. S. Lintuvuori, K. Stratford, M. E. Cates,
-D. Marenduzzo.   Colloids in Cholesterics: Size-Dependent Defects and Non-Stokesian Microrheology. {\it Phys. Rev. Lett.} {\bf 105}, 178302 (2010).
-\bibitem{juho2}  J. S. Lintuvuori, K. Stratford, M. E. Cates,
-D. Marenduzzo. 
-Self-Assembly and Nonlinear Dynamics of Dimeric Colloidal Rotors in 
-Cholesterics. {\it Phys. Rev. Lett.} {\bf 107}, 267802 (2011).
-\bibitem{permeation1} 	W. Helfrich. Capillary flow of cholesteric
-and smectic liquid crystals. {\it Phys. Rev. Lett.} {\bf 23}, 372-374 (1969).
-\bibitem{permeation2} A. Dupuis, D. Marenduzzo, E. Orlandini, J.~M. Yeomans.
-Rheology of cholesteric blue phases. {\it Phys. Rev. Lett.} {\bf 95},
-097801 (2005).
-\bibitem{lubensky} M. Zapotocky, L. Ramos, P. Poulin, T.~C. Lubensky, 
-D.~A. Weitz.Particle-stabilized defect gel in cholesteric liquid crystals.  {\it Science} {\bf 283}, 209-212 (1999).
-%\bibitem{fukuda} J.-I. Fukuda, S. Zumer.
-%Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal. 
-%{\it Nat. Comm.} {\bf 2}, 246 (2011).
-\bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
-Self-Organization of Particles with Planar Surface Anchoring in a Cholesteric
- Liquid Crystal. {\it Langmuir} {\bf 26}, 13502-13510 (2010).
-\bibitem{flowswitch} O. Henrich, K. Stratford, P.V. Coveney, M.E. Cates, D. Marenduzzo,
-Rheology of cubic blue phases. {\it Soft Matter} (2013), DOI: 10.1039/C3SM50228G 
-\bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang, Developing optofluidic technology through the fusion of microfluidics and optics {\it Nature} {\bf 442}, 381-386 (2006).
-%\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka.
-%Memory and topological frustration in nematic liquid crystals confined in porous materials. {\it Nat. Mat.} {\bf 10}, 303-309 (2011).
-\end{thebibliography}
+\bigskip
+\noindent
+\textbf{\large Author Contributions}
+
+\noindent
+KS, OH and JSL contributed to the code base for simulation and performed
+simulations. All authors helped to analyse the results and write the
+manuscript.
+
+\bigskip
+\noindent
+\textbf{\large Additional Information}
+
+\noindent
+\textbf{Competing financial interests:} The authors declare no competing
+financial interests.
 
@@
 \centerline{\includegraphics[width=\textwidth]{text-fig1.jpg}}
-\caption{Snapshots of the states obtained after dispersing
+\caption{\textbf{Bulk blue phase structures.} Snapshots of the states
+obtained after dispersing
 a suspension of colloidal nanoparticles within a cholesteric liquid
@@
 (one eighth of the total volume); the full structures are shown in
-Supplementary Fig. S2 and~S3.}
+Supplementary Figs.~2 and~3.}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{text-fig2.jpg}}
-\caption{Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, 
+\caption{\textbf{Confined blue phase structures.}
+Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, 
 and then quenched into a regime where BPI is stable in the bulk. Both top and side views are provided.
@@
 (D,H) $w=2.3$ and $\phi=2\%$.
-As in Fig. 1, only one quarter of the simulation box is shown for clarity;
-the full structures are shown in Supplementary Fig. S5 and S6.}
+As in Fig.~1, only one quarter of the simulation box is shown for clarity;
+the full structures are shown in Supplementary Figs.~5 and~6.}
 \end{figure}
@@
 \end{center}
-\caption{Steady states obtained when a dispersion of colloids is inserted
+\caption{\textbf{Structure in electric field.}
+Steady states obtained when a dispersion of colloids is inserted
 into a thermodynamically stable bulk BPIII with and without applied
@@
 
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\hline
+         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
+\hline 
+Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
+0.23, 2.3\\
+\hline
+Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
+0.23, 2.3 \\
+\hline
+Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
+\hline
+External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
+\hline
+\end{tabular}
+\end{center}
+\caption{\textbf{Free energy parameters used in the simulations.}
+Bulk BPI
+correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
+geometries are shown in Fig.~2; and the external field case is
+relevant for Fig.~3.}
+\label{table:params}
+\end{table}
+
 \end{document}

### 2014-02-24 papers/2012-sandwich/response.txt
Completed response and first stage of editing
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2316 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+RESPONSE: We added comments and references specifying experimental model systems
+(see also response to comment of Reviewer #2).
+
 Particular remarks:
@@
 
-RESPONSE: Citition for Ranvik et al Faraday Discuss. 2010 (sic) is added
+RESPONSE: Citition for Ravnik et al Faraday Discuss. 2010 (sic) is added
 [current reference 16].
@@
 particle surface. This orientation can be either normal to the
-surface or in a plane tengential to the surface, a property related
+surface or in a plane tangential to the surface, a property related
 to surface chemistry in experimental systems but which can be imposed
@@
 
+RESPONSE: We have added in the main text (last pararagraph introduction) a
+summary of the relevant dimensions in physical units: 
+"The simulations represent a liquid crystal blue phase with unit cell 
+size in the range of 100--500~nm \cite{mermin}, and particles with 
+a radius of around 50~nm." 
+A coarse-grained description of the mesophase like the Landau-de Gennes 
+theory is appropriate as long as the size of the colloidal particle is much 
+larger than the size of the mesogen, which are typically in the 
+range of 10^-9 m. Hence the Landau-de Gennes approach allows to 
+describe defects down to extremely small length scales.
+
 Is WR/K = 2.8 justified for nanoparticles used in the study?
 
-RESPONSE: We have added in the main text (last para. introduction) a
-summary of the relevant
-dimensions in physical units: "The simulations represent a liquid crystal
-blue phase with unit cell size in the range of 100--500~nm \cite{mermin},
-and particles with a radius of around 50~nm."
-
-...
+RESPONSE: The ration WR/K = 2.8 is justified for nanoparticles
+in the strong anchoring regime.
 
@@
 Can the G elastic modulus be estimated for modeled structures?
+
+RESPONSE: We added a comment that gives an estimate of the elastic modulus.
+Esitmates are in the kPa range.
 
@@
 
+RESPONSE: Our results suggest that this is difficult to achieve, but
+generally possible. It depends sensitively on the 
+anchoring strength and the size of the colloidal particles. 
+Strong anchoring conditions lead to a significant 
+distortion of the disclination lines, whereas in the weak anchoring 
+regime the forces that drive the particles towards the 
+disclination lines remain rather small.
+
 Is also in case of the disordered structures observed a change in the
 stability range of the colloidal BPs?
+
+RESPONSE: Determining the stability range of the composite
+phases was beyond the scope of this paper. This would have required 
+an optimisation of the pitch length, which requires a large number 
+of additional simulations for each individual parameter set.
 
@@
 
+RESPONSE: Due to the anchoring conditions the colloidal particles are 
+attracted or repelled from the walls in the thin layer geometry. This leads to a more
+two-dimensional structure compated to the situation without wall confinement.
+
 Are the electric field induced changes irreversible only in case of
 colloidal dispersions?
+
+RESPONSE: For the given parameters the field induced changes are only
+irreversible when colloids with sufficiently strong 
+anchoring conditions are present. The mesophase without colloids 
+undergoes a reversible ordering transition as described in reference [10].
 
@@
 the 2nd edition of de Gennes and Prost, which now appears in the
-main text [current reference 33].
+main text [current reference 34].
 
@@
 results are of interest and useful to
-experimentalists working in the field. The paper would be even more useful
-if it included references to actual materials, both the liquid crystal host
-and the added particles and perhaps
+experimentalists working in the field.
+
+The paper would be even more useful if it included references to actual 
+materials, both the liquid crystal host and the added particles and perhaps
 even suggested a model system that would be a good starting point to study
-experimentally. In a similar vein, it would also be useful to address the
-practical issue of reproducibility, which
-has been a major problem with these materials. The theoretical results
-clearly indicate how important the formation conditions are to the final
-result. What is the authors sense of how easy
-it will be to reproducibly fabricate a device
+experimentally.
+
+RESPONSE: We added some details and references of experimental
+realisations of these systems (see also comment from Reviewer #1).
+
+In a similar vein, it would also be useful to address the
+practical issue of reproducibility, which has been a major problem 
+with these materials. The theoretical results clearly indicate how important 
+the formation conditions are to the final result.
+
+What is the authors sense of how easy it will be to reproducibly fabricate a device
 with a specific morphology and optical properties. This seems a likely
@@
 the simulations.
+
+RESPONSE: Our feeling is that control mechanisms beyond simple anchoring 
+will be necessary for a device that features reproducable optical properties. 
+This could be for instance achieved with external electric fields as shown in 
+this publication. Alternative routes could be microfluidic flow or
+patterned, multidomain anchoring conditions.
 

### 2014-02-24 papers/2012-sandwich/CholColloids.tex
Completed response and first stage of editing
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2316 40876441-3496-0410-a613-a2fca35b2ae2

@@
 fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa.
+
+\textcolor{red}{An estimate of the low frequency shear modulus can be 
+obtained by comparing the mechanical properties of the network of defects 
+stabilised by colloidal particles to those of rubber materials~\cite{ramos}.
+With typical Frank elastic constants in the range of $K\simeq7\times10^{-11}$ N
+and unit cell sizes of the order $d=\simeq 1-5\times10^{-7}$ m this leads to
+a shear modulus of roughly $G\simeq K/d^2=0.3-7$ kPa.}
 
@@
 
+\medskip
+\noindent
+\textcolor{red}{
+\textbf{Experimental model systems.}
+Several experimental realisations of compostite
+materials consisting of nanoparticles in a liquid-crystalline 
+host material have been described in the literature 
+\cite{karatairi, rozic, dierking, cordoyiannis}. 
+The systems are generally rather delicate and several
+parameters have to be optimised in order to achieve a stabilisation 
+of the compound material. As a general rule the
+size of the colloidal particles has to be comparable to the size 
+of the defect cores. These are are usually in the range of several
+to a few dozen nanometers for typical host materials.
+The concentration of colloidal particles as well as the anchoring 
+strength should not bee too high as this would lead to rather 
+drastic changes in the defect structure 
+of the host material and render the composite material 
+thermodynamically unstable. 
+}
+
 \pagebreak
@@
 
+\bibitem{ramos} 
+\textcolor{red}{L. Ramos, M. Zapotocky, T.~C. Lubensky, D.~A. Weitz, Rheology of defect networks in cholesteric liquid crystals, {\it Phys. Rev. E} {\bf 66} 031711 (2002).}
+
 \bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
@@
 \textit{J. Phys. Condens. Matter} \textbf{16},  3903--3915 (2004).
+
+\textcolor{red}{
+\bibitem{karatairi} E. Karatairi et al.,
+Nanoparticle-induced widening of the temperature range of liquid-crystalline blue phases, 
+\textit{Phys. Rev. E} \textbf{81}, 041703 (2010).
+\bibitem{rozic} B. Ro\v zi\v c et al.,
+Theoretical and experimental study of the nanoparticle-driven blue phase stabilisation,
+\textit{Eur. Phys. J. E} \textbf{34}, 17 (2011).
+\bibitem{dierking} I. Dierking et al.,
+Stabilising liquid crystalline Blue Phases,
+\textit{Soft Matter} \textbf{8}, 4355--4362 (2012).
+\bibitem{cordoyiannis} G. Cordoyiannis et al.,
+Different modulated structures of topological defects stabilized by adaptive targeting nanoparticles, \textit{Soft Matter} \textbf{9}, 3956--3964 (2013).
+}
 

### 2014-03-03 papers/2012-sandwich/response.txt
Updated ms and response
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2323 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-RESPONSE: We added comments and references specifying experimental model systems
-(see also response to comment of Reviewer #2).
+RESPONSE: The issue of experimental guidance is addressed in detail
+in our reply to reviewer 2 below.
 
@@
 
-RESPONSE: Citition for Ravnik et al Faraday Discuss. 2010 (sic) is added
+RESPONSE: Citation for Ravnik et al Faraday Discuss. 2010 (sic) is added
 [current reference 16].
@@
 We have made this connection more clear by rewriting the relevant
-sentence: "However, recent advances in formalation have widened
+sentence: "However, recent advances in formulation have widened
 this stability range enormously [11,12,13,14]. Most recently,
@@
 
-RESPONSE: We have added in the main text (last pararagraph introduction) a
-summary of the relevant dimensions in physical units: 
-"The simulations represent a liquid crystal blue phase with unit cell 
-size in the range of 100--500~nm \cite{mermin}, and particles with 
-a radius of around 50~nm." 
-A coarse-grained description of the mesophase like the Landau-de Gennes 
-theory is appropriate as long as the size of the colloidal particle is much 
-larger than the size of the mesogen, which are typically in the 
-range of 10^-9 m. Hence the Landau-de Gennes approach allows to 
-describe defects down to extremely small length scales.
+RESPONSE: We have added in the main text (last paragraph introduction) a
+summary of the relevant dimensions in physical units:  "The simulations
+presented here are representative of a liquid crystal blue phase with unit
+cell size in the range of 100--500~nm and particles with a radius of
+around 50~nm." This new paragraph also serves the requirement for a brief
+summary of the results as set out in the manuscript checklist.
+
+RESPONSE (cont.): As a coarse-grained description of the mesophase, the
+Landau-de Gennes  theory is appropriate as long as the size of the
+colloidal particle is much larger than the size of the mesogen,
+the latter being typically in the range of 1--10 nanometres. Our
+simulations meet this criteria (particle size about 50 nm) and the
+use of Landau-de Gennes is therefore justified.
 
@@
 
-RESPONSE: The ration WR/K = 2.8 is justified for nanoparticles
-in the strong anchoring regime.
+RESPONSE: The choice of WR/K = 2.3 in the simulations (not 2.8)
+is used to separate clearly a strong anchoring regime from a
+weak anchoring regime (w = 0.23). While typical experimental
+values for particles of this size may mean the weak regime is
+more common, we do not see that this precludes interest in a
+strong anchoring regime.
 
@@
 
-RESPONSE: We added a comment that gives an estimate of the elastic modulus.
-Esitmates are in the kPa range.
+RESPONSE: We note we did have, in the original manuscript, a
+paragraph which discusses estimates of the elastic modulus G
+based on analogy with experimentally observed systems. To the
+existing paragraph we have appended:
+"This figure is consistent with an estimate obtained by comparing
+the mechanical properties of the network of defects stabilised by
+colloidal particles with those of rubber [Ramos et al 2002] for
+the relevant simulation parameters (see Supplementary Information)."
+To the supplementary information we append: "For a Frank elastic
+constant of $K = 70$~pN and unit cell sizes in the range
+$\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
+K/\lambda^2 =$0.3--7~kPa." The relevant reference is added.
 
@@
 
-REESPONSE: A citition for Ravnik et al (2011) has been added [current
+RESPONSE: A citation for Ravnik et al (2011) has been added [current
 reference 26].
@@
 
-RESPONSE: Due to the anchoring conditions the colloidal particles are 
-attracted or repelled from the walls in the thin layer geometry. This leads to a more
-two-dimensional structure compated to the situation without wall confinement.
+RESPONSE: We have added the following at the end of the section
+describing the confined geometry: "The degree of similarity between
+the bulk and confined structures is somewhat masked by the local
+change in particle density in the confined case. If one allows for
+this, Figs.~1 and~2 show the individual structural motifs are actually
+rather similar: templated lines for weak anchoring and clusters for
+strong anchoring."
+
+RESPONSE (cont.): One potential visualisation ambiguity is that the
+displayed box sizes are different in each of the figures (cf. particle
+size). An explicit statement has been added to each caption to make
+this more clear to the reader. ADD CAPTIONS?
 
@@
 undergoes a reversible ordering transition as described in reference [10].
+CHECK REFERENCE
 
@@
 
-RESPONSE: The corrent punctuation in the page numbers has been
+RESPONSE: The correct punctuation in the page numbers has been
 inserted in the reference for Wright and Mermin [current reference 9].
@@
 transparent: "For both normal and planar anchoring at the walls,
-values of $W$ and $K$ are used which are the same as for the
+the value of $W$ is the same as that used for the
 strongly anchoring particles."
@@
 
-REPSONSE: Reference 1, Wright and Mermin, was correct in
+RESPONSE: Reference 1, Wright and Mermin, was correct in
 the original supplement. Reference 2 has been updated to reflect
@@
 main text [current reference 34].
+CHECK REFERENCE
 
@@
 
-This paper, "Self-assembly of colloid-cholesteric composities: A route to
+This paper, "Self-assembly of colloid-cholesteric composites: A route to
 switchable optical materials" reports simulations of the formation of
@@
 ahd to postulate how these colloids may be employed in specific
-appilcations.
+applications.
 
@@
 
-RESPONSE: We added some details and references of experimental
-realisations of these systems (see also comment from Reviewer #1).
+RESPONSE: In a new paragraph in the discussion section, we include six
+new references specifically aimed at recent studies on experimental
+nanoparticle liquid crystal composites. These are currently [34-39].
+The paragraph also discusses the issue of a model system, and suggests
+properties such a system may have to help match experiment and simulation.
 
@@
 
-RESPONSE: Our feeling is that control mechanisms beyond simple anchoring 
-will be necessary for a device that features reproducable optical properties. 
-This could be for instance achieved with external electric fields as shown in 
-this publication. Alternative routes could be microfluidic flow or
-patterned, multidomain anchoring conditions.
+RESPONSE: Clearly the issue of experimental reproducibility is a difficult
+one to be addressed by, of necessity idealised, simulation. However, one
+point our simulations do show is that careful characterisation of not
+only the particles and host fluid is required, but also of possible
+confinement effects and associated anchoring conditions. This we also
+address in our updated discussion.
+
+The entire paragraph reads: ADD WHEN FINALISED
 
 * ADDITIONAL CHANGES TO MEET FORMAT REQUIREMENTS
+  cf Manuscript checklist
 

### 2014-03-03 papers/2012-sandwich/CholColloids.tex
Updated ms and response
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2323 40876441-3496-0410-a613-a2fca35b2ae2

--- 2014-02-21 papers/2012-sandwich/CholColloids.tex
@@
 SUPA, School of Physics and Astronomy, The King's Buildings,\\
-The University of Edinburgh, Edinburgh, EJ9 3JZ, UK\\
+The University of Edinburgh, Edinburgh, EH9 3JZ, UK\\
 \small{$^{*}$ equally contributed to this work}\\
@@
 These go beyond previous studies of colloidal-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloidal particles in cholesterics~\cite{juho1,juho2}.
-Our methodology and simulation parameters are summarized in the Supplementary Information.
 
-SHORT SUMMARY OF RESULTS AND CONCLUSIONS (PRESENT TENSE). To include a
-statement about confined systems with reference to \cite{extrareference2}.
-The simulations represent a liquid crystal
+\textcolor{red}{
+The simulations presented here are representative of a liquid crystal
 blue phase with unit cell size in the range of 100--500~nm \cite{mermin},
-and particles with a radius of around 50~nm.
+and particles with a diameter of around 50~nm. The simulations
+address both the bulk structure, and the structure which forms when
+the colloidal dispersion is confined in a narrow sandwich of width
+comparable to the unit cell size~\cite{extrareference2}. It is seen
+that a wide variety of structure is possible, and may be influenced
+by parameters including the solid volume fraction, and the details
+of the anchoring at the colloid surface (and the at the surface of the
+confining walls). Further, evidence is found for metstable switching of
+this structure in an applied electric field. Details of the
+computational approach are set out in the Methods section, while
+simulation parameters are reported in the Supplementary Information.
+}
 
@@
 \noindent
-\textbf{\large Results} [Currently about 1400 words]
+\textbf{\large Results}
 
@@
 fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa.
+\textcolor{red}{
+This figure is consistent with an estimate obtained by comparing
+the mechanical properties of the network of defects stabilised by
+colloidal particles with those of rubber~\cite{ramos} for
+the relevant simulation parameters (see Supplementary Information).}
 
@@
 There are, however, significant effects of the confining walls.
-First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloid particless at both walls 
+First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloid particles at both walls 
 (see side-views in Fig.~2A-D). Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit. These clusters
@@
 Note that planar wall anchoring is incompatible with the normal anchoring on the colloidal particles (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
+
+\textcolor{red}{
+The degree of similarity between
+the bulk and confined structures is somewhat masked by the local
+change in particle density in the confined case. If one allows for
+this, Figs.~1 and~2 show the individual structural motifs are actually
+rather similar: templated lines for weak anchoring and clusters for
+strong anchoring.
+}
 
@@
 distinct metastable structures, which are retained after field removal
-(Suplementary Fig.~8 and Fig.~9).
-
-%FIELD SWITCHING PARA HERE
-\if{The existence of several locally stable competing solids with 
-very different symmetry and, arguably, physical properties, 
-is of interest to technology, as it makes these materials ideal
-candidate to build multistable devices, which can retain memory of
-more than one metastable state in the absence of an applied field.
-Multistability is indeed a must-have feature for new generation switchable 
-electrooptic devices for use in energy-saving electronics-free
-appliances, smart glass, and e-paper to cite just a few examples.
-}\fi
+(Supplementary Fig.~8 and Fig.~9).
 
@@
 \noindent
-\textbf{\large Discussion} [Currently about 260 words]
+\textbf{\large Discussion}
 
 \noindent
-The surprising wealth of colloidal-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
-First, blue phases provide a three-dimensional template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}.
-Second, colloidal intrusions in a liquid crystal raise the free energy locally; sharing this cost creates a generic tendency to aggregate into clusters~\cite{tiffany}.
-Which of these factors dominates depends on $w$, a dimensionless measure of the anchoring strength of the director field at the colloidal surface.
-For $w\ll 1$, the templating principle dominates.
-Colloidal crystals, gels or twisted rings can then arise.
-For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating gels, or helical colloidal ropes.
-In both cases, control is offered by the process route, colloidal concentration, and the anchoring conditions at sample walls.
-We have also shown that electric fields can be used to switch between metastable states, providing a possible route to future multistable device applications.
-%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloid particles and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
+The surprising wealth of colloidal-BP composites obtained in Figs.~1-3 is
+the complex consequence of a simple competition between two self-assembly
+principles. 
+First, blue phases provide a three-dimensional template onto which
+colloidal particles are attracted so as to reduce the elastic stresses
+arising at the disclination cores~\cite{miha}. Second, colloidal intrusions
+in a liquid crystal raise the free energy locally; sharing this cost creates
+a generic tendency to aggregate into clusters~\cite{tiffany}. Which of these
+factors dominates depends on $w$, a dimensionless measure of the anchoring
+strength of the director field at the colloidal surface. For $w\ll 1$, the
+templating principle dominates. Colloidal crystals, gels or twisted rings
+can then arise. For $w \gg 1$, interparticle attractions defeat the
+templating tendency, and we predict disconnected aggregates, percolating
+gels, or helical colloidal ropes. In both cases, control is offered by the
+process route, colloidal concentration, and the anchoring conditions at
+sample walls. We have also shown that electric fields can be used to
+switch between metastable states, providing a possible route to future
+multistable device applications.
+
+\textcolor{red}{
+How do these simulations relate to experiment? Several authors have
+used nanoparticles in the context of stabilising BPs
+\cite{yoshida2009,karatairi}. These
+nanoparticles are typically a few nanometres in size, in which case
+stabilisation is explained by the mechanism of defect removal.
+Experiments suggest this mechanism is also
+relevant for larger particles in the range of 40--70~nm 
+\cite{dierking}, which is consistent with the current simulations.
+However, the mechanism is observed to be less efficient for
+particles above 100~nm, where particles may start to disrupt
+longer-range liquid crystal order \cite{dierking}. While the field of
+self-assembly with nanoparticles is relatively new \cite{draper},
+nanoparticles have also been used in both nematic \cite{milette2012}
+and chiral liquid crystals \cite{cordoyiannis} as host fluid templates.
+The variation of structure seen
+in our simulations suggest that some careful characterisation of parameters
+such as surface anchoring of particles (which are often stabilised, or even
+functionalised with surface ligands), together with those associated
+with any confining surfaces or interfaces, is required to understand the
+final structure and to be able to reproduce it with a given experimental
+protocol. The identification of a model system which could allow
+unambiguous comparison between experimental systems and simulation
+would be extremely useful. Such a system might adopt particles of
+dimeter in the range 50~nm or above, which allows a coarse-grained simulation
+approach of the type used here to be adopted, which can then capture the
+large-scale structure required to compare with experiment. In the same vein,
+detailed surface chemistry is difficult to represent in coarse-grained
+simulations, so heavily functionalised particles might be avoided.
+}
+
 We have not explored in this work switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case the same materials may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
@@
 
-INTRODUCTORY STATEMENT
+Here, the simulation method is described. Note that
+all quantities are expressed in simulation units. Details
+of the simulation parameters, and their mapping to physical units,
+may be found in the Supplementary Information.
 
+\medskip
 \noindent
-\textbf{Theromdynamics.}
+\textbf{Thermodynamics.}
 The thermodynamics of the cholesteric liquid crystal can be described
@@
 
-\medskip
-\noindent
-\textbf{Protocol and parameter details for bulk BPI.}
-The parameters for the free energy are chosen to be representative
-of blue phase I: chirality $\kappa \sim 0.7$ and reduced temperature
-$\tau = -0.25$.
-The full free energy parameters are shown in Table~\ref{table:params}.
-The order parameter tensor $Q_{\alpha\beta}$ for bulk blue phase I is
-initialised from an approximation in high chirality limit
-\cite{mermin,henrichfield}. Systems of 128$^3$ lattice sites are
-used with a pitch length of $p = 32\sqrt{2}$, which accommodates 4$^3$
-unit cells with fully periodic boundary conditions.
-Simulations at different solid volume fractions (1\% and 4\%), and different 
-surface anchoring strengths ($WR/K = 0.23$ and $WR/K = 2.3$) representing
-``weak'' and ``strong'' anchoring are run for two million simulation
-time steps.
-The initial distribution of the particle positions is set at random
-for the required solid volume fraction; the particles are initially
-at rest.
-
-\medskip
-\noindent
-\textbf{Parameter details for confined BPI.}
-Here, a narrow sandwich of fluid is placed between flat walls in
-perpendicular to the narrow coordinate direction, and with periodic
-boundaries in the other two directions. The system size used in all
-cases is 256$^2 \times$56 lattice sites. Each simulation is a ``quench''
-in a similar manner to that used in the bulk: the order parameter is
-initialised randomly. The free energy parameters (see Table~\ref{table:params})
-are again appropriate for (equilibrium) blue phase I
-(chirality $\kappa = 0.8$ and
-reduced temperature $\tau = -0.25$). The cholesteric pitch is set
-to be $p = 64$, providing a slightly better resolution than the bulk
-simulations.
-
-The surface anchoring for colloidal particles is as before: always normal,
-but with
-$WR/K = 0.23$ and $WR/K = 2.3$. The solid volume fractions are
-1\% and 4\%.
-\textcolor{red}{For both normal and planar anchoring at the walls,
-values of $W$ and $K$ are used which are the same as for the
-strongly anchoring particles.}
-All simulations are run for two million simulation
-time steps.
-
-\medskip
-\noindent
-\textbf{Parameters for external electric field.}
-For the simulations of BPIII with colloidal particles a system size of
-$128^3$ is used with a cholesteric pitch of $p=32$ in lattice units. 
-The simulations are
-performed at chirality $\kappa=2.6$ and reduced temperature $\tau=-0.25$.  
-An initial configuration of
-randomly oriented and positioned double twist cylinders is used,
-from which an amorphous BPIII network emerges \cite{bp3}. This structure
-is equilibrated for $6\times10^{5}$ LB time steps until no significant
-further evolution is observed.
-
-The colloidal particles with weak
-normal anchoring ($WR / K= 0.23$), are then inserted with randomly chosen positions,
-and the composite system is equilibrated for another $1.2\times10^{6}$
-time steps.
-At the end of the equilibration phase the uniform electric field
-with reduced field strength ${\cal E} = 0.8$ 
-is switched on (oriented along the $z$-direction), and the simulation
-run for a further $1.2\times10^{6}$ time steps.
-This is found to be sufficiently long for a hexagonal structure to
-emerge and saturate.
-If the electric field is then switched off, and the system quickly relaxes
-to a metastable state which shows a residual anisotropy in the
-colloidal distribution. The total length of this final relaxation phase
-is $1\times10^6$ LB time steps.
-
-\medskip
-\noindent
-\textbf{Disclination rendering.}
-In all cases visualisation of the defect lines is carried out via
-an isosurface of the scalar order parameter $q$ determined from the
-largest eigenvalue of the local order parameter tensor. A low value
-of the scalar order parameter $q$ (typically in the range 0.12--0.14)
-unambiguously identifies the disclinations.
-
-\medskip
-\noindent
-\textbf{Computation of the structure factor.}
-To provide structural information about the field-aligned 
-states and the residual anisotropy after switch off a 
-procedure similar to the one described in \cite{bp3} is
-followed.
-The structure factor $S(\mathbf{k})$ in the current approach is
-defined via the Fourier transform of the particle density $F(\mathbf{k})$:
-
-\begin{equation}
-S({\bf k}) =  |F({\bf k})|^2, \mathrm{with}\
-F({\bf k}) = \int d^3{\bf r} \rho({\bf r}) e^{i {\bf k. r}}.
-\end{equation}
-The colloidal density $\rho({\bf r})$ is a discretised density and taken to be
-$\rho({\bf r})=1$ if a lattice site at ${\bf r}$ is part of a solid particle
-and $\rho({\bf r})=0$ otherwise. This definition, which deviates from the 
-standard definition of a structure factor and does not consider the particles 
-as point-like objects, allows the same structural information
-to be obtained in a simple way.
-
-\medskip
-\noindent
-\textbf{Parameter mapping to physical units.}
-To get from simulation to physical units, a calibration of scales of
-length, energy and time in the simulations is required. For similar
-mappings see, e.g.~\cite{denniston2}.
-
-The length scale can be set by mapping the half pitch to a realistic blue
-phase unit cell size, which is in the 100--500~nm range \cite{mermin}. 
-For bulk simulations (Fig.~1 and Fig.~3 in the main text), a suitable choice
-is one where one simulation unit (lattice site) corresponds to 10~nm. Therefore
-the colloidal size in Figs.~1 and~3 corresponds to $\sim$ 50~nm.
-
-To obtain an energy scale, it is possible to choose $A_0 \simeq 10^6$~Pa,
-which is reasonable following Ref.~\cite{mermin} (this is also the choice in
-Ref.~\cite{domaingrowth}). This choice leads to a simulation unit of
-free energy (or stress) equal to $10^{8}$~Pa in Figs.~1 and~3.
-From the energy and length scales one can map elastic constants to
-physical units: for instance, the simulation in Fig.~1 corresponds to
-a liquid crystal with splay, bend and twist (Frank) elastic constants
-equal to 35 pN (or equivalently $K=70$~pN).
-
-The timescale calibration may be obtained from (see e.g.~\cite{denniston})
-\begin{equation}
-\gamma_1=\frac{2q^2}{\Gamma},
-\end{equation}
-which relates the rotational viscosity $\gamma_1$ to the ordering strength
-$q$ and the order parameter mobility $\Gamma$.
-In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
-for the simulation in Fig.~1 leads to $q=1/2$.   
-For real liquid crystalline materials, $\gamma_1$ usually lies in range
-$10^{-2}$--$1$~Pa~s~\cite{deGennes}; for definiteness say
-$\gamma_1 = 1$~Pa~s (equivalently 10 poise). Given the previous mapping for
-free energy (or stress) units, this leads to a simulation unit of time equal
-to $10^{-8}$~s. The total simulated time in Figs.~1--3 is therefore in the
-millisecond range.
-
-To calibrate the electric field strength 
-one can equate the value of the dimensionless parameter ${\cal E}$ in the 
-simulations with that of a hypothetical experiment.
-Given the previous value for $A_0$, and a dielectric anisotropy of 20
-(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
-dielectric permittivity of vacuum), an electric field
-of 100 V~$\mu$m$^{-1}$ is obtained. This corresponds to a value of
-${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
-Fig.~3 of the main text).
-
-Given these mappings, all quantities can be easily transformed from
-simulation to physical units and vice versa. For example Figs.~1 and~2
-use an isotropic viscosity $\eta \simeq$~0.01 and~0.1 in simulation units,
-which translate into $0.01$--$0.1$~Pa~s respectively. This is sensible (if 
-slightly low) for a molecular nematogen in the isotropic phase. 
-The effective viscosity in ordered phases is higher, but this is ensured
-in our model by the coupling to the order parameter.
-Note that the density is set to unity in the simulations: this
-corresponds to a fluid density which is about a thousand times larger
-than in experiments. This makes no difference in practice as
-the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
-velocity, in our simulations around $10^{-5}$ in lattice units, 
-and $\Lambda$ is a typical length scale, e.g. the size of a unit cell) 
-remains small enough~\cite{codef}.
-
-To summarise, the simulations represent BP-forming materials, with the
-interpretation of the simulation units for length, time, and energy
-density being close to 10~nm, 10~ns, and 100~MPa respectively.
-
 \pagebreak
@@
 
-\bibitem{wiresmiha} M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj,
-D. Babic, N. Osterman, I. Musevic,
+\bibitem{wiresmiha}
+Ravnik, M., et al.,
+%M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj,
+%D. Babic, N. Osterman, I. Musevic,
 Entangled nematic colloidal dimers and wires, 
@@
 \bibitem{zumer}
-I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik, S. Zumer,
+%I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik, S. Zumer,
+Musevic, I., Skarabot, M., Tkalec, U., Ravnik, M., and Zumer, S.,
 Two-dimensional nematic colloidal crystals self-assembled by topological
@@
 
-\bibitem{tanaka} T. Araki, H. Tanaka,
+\bibitem{tanaka}
+Araki, T., and Tanaka, H.,
 Colloidal aggregation in a nematic liquid crystal: Topological arrest of
@@
 \bibitem{tiffany}
-T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo, W.~C.~K. Poon,
+%T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo, W.~C.~K. Poon,
+Wood, T.A., Lintuvuori, J.S., Schofield, A.B, Marenduzzo, D., and
+Poon, W.C.K.,
 A Self-Quenched Defect Glass in a Colloid-Nematic Liquid Crystal Composite,
@@
 
-\bibitem{stark} H. Stark,
+\bibitem{stark} Stark, H.,
 Physics of colloidal dispersions in nematic liquid crystals,
@@
 \bibitem{abbott}
-I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo,
-N.~L. Abbott,
+%I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo,
+%N.~L. Abbott,
+Lin, I.H., et al.,
 Endotoxin-Induced Structural Transformations in Liquid Crystalline Droplet,
@@
 \bibitem{colloiddevice}
-K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton, R. G. Capurso,
+%K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton, R. G. Capurso,
+Chari, K., Rankin, C.M., Johnson, D.M, Blanton, T.N., and Capurso, R.G.,
 Single-substrate cholesteric liquid crystal displays by colloidal
@@
 
-\bibitem{tanakanatmat} T. Araki, M. Buscaglia, T. Bellini, H. Tanaka,
+\bibitem{tanakanatmat}
+Araki, T., Buscaglia, M., Bellini, T., and Tanaka, H.,
 Memory and topological frustration in nematic liquid crystals confined
@@
 
-\bibitem{mermin} D.~C. Wright, Mermin, N.~D.,
+\bibitem{mermin} Wright, D.C., and Mermin, N.D.,
 Crystalline liquids: The blue phases, {\it Rev. Mod. Phys.} {\bf 61},
@@
 
-\bibitem{bp3} O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
+\bibitem{bp3} Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
  Structure of Blue Phase III of Cholesteric Liquid Crystals,
@@
 \bibitem{kikuchi}
-H. Kikuchi, M. Yokota, Y. Hisakado, H. Yang, T. Kajiyama,
+Kikuchi, H., Yokota, M., Hisakado, Y., Yang, H., and Kajiyama, T.,
 Polymer-stabilized liquid crystal blue phases,
@@
 \bibitem{bplasers}
-W. Y. Cao, A. Munoz, P. Palffy-Muhoray, B. Taheri,
+Cao, W.Y., Munoz, A., Palffy-Muhoray, P., and  Taheri, B.,
 Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
@@
 \bibitem{coleswidetrange}
-H.~J. Coles, M.~N. Pivnenko,
+Coles, H.J., and  Pivnenko, M.N.,
 Liquid crystal 'blue phases' with a wide temperature range,
@@
 \bibitem{bpdevice}
-Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
+%Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
+Ge, Z.B., Gauza, S., Jiao, M.Z., Xianyu, H.Q., and  Wu, S.T.,
  Electro-optics of polymer-stabilized blue phase liquid crystal displays,
@@
 \bibitem{coles}
-F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim,
-S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles,
+%F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim,
+%S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles,
+Castles, F., et al.,
 Blue-phase templated fabrication of three-dimensional nanostructures for
@@
 \bibitem{extrareference1}
-\textcolor{red}{M. Ravnik, G.P. Alexander, J.M. Yeomans, S. Zumer,
+\textcolor{red}{Ravnik, M., Alexander, G.P., Yeomans, J.M., and Zumer, S.,
 Mesoscopic modelling of colloids in chiral nematics,
@@
 \bibitem{miha}
-M. Ravnik, G.~P. Alexander, J.~M. Yeomans, S. Zumer,
+Ravnik, M., Alexander, G.P.,  Yeomans, J.M., and  Zumer, S.,
 Three-dimensional colloidal crystals in liquid crystalline 
@@
 
-\bibitem{lavrentovich} O.~D. Lavrentovich,
+\bibitem{lavrentovich} Lavrentovich, O.D.,
  Liquid crystals, photonic crystals, metamaterials and transformation optics, 
@@
 \bibitem{adriano}
-A. Tiribocchi, G. Gonnella, D. Marenduzzo, E. Orlandini, F. Salvadore,
+Tiribocchi, A., Gonnella, G., Marenduzzo, D., Orlandini, E., and Salvadore, F.,
 Bistable Defect Structures In Blue Phase Devices,
@@
 \bibitem{fukuda}
-J.-I. Fukuda, S. Zumer,
+Fukuda, J.-I., and Zumer, S.,
 Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal,
@@
 \bibitem{henrichfield} 
-O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates,
+Henrich, O., Marenduzzo, D., Stratford, K., and  Cates, M.E.,
 Thermodynamics of blue phases in electric fields,
@@
 
-\bibitem{epaper} J. Heikenfeld, P. Drzaic,  J.~S. Yeo, T. Koch,
+\bibitem{epaper}
+Heikenfeld, J., Drzaic, P., Yeo, J.S., and Koch, T.,
 A critical review of the present and future prospects for electronic paper,
@@
 \bibitem{domaingrowth}
-O. Henrich, K. Stratford, D. Marenduzzo, M.~E. Cates,
+Henrich, O., Stratford, K., Marenduzzo, D., and  Cates, M.E.,
 Ordering dynamics of blue phases entails kinetic stabilization of amorphous
@@
 \bibitem{juho1}
-J. S. Lintuvuori, K. Stratford, M. E. Cates, D. Marenduzzo,
+Lintuvuori, J.S.,  Stratford, K., Cates, M.E., and Marenduzzo, D.,
 Colloids in cholesterics: size-dependent defects and non-Stokesian
@@
 \bibitem{juho2}
-J. S. Lintuvuori, K. Stratford, M. E. Cates, D. Marenduzzo,
+Lintuvuori, J.S.,  Stratford, K.,  Cates, M.E., and  Marenduzzo, D.,
 Self-assembly and nonlinear dynamics of dimeric colloidal rotors in 
@@
 \bibitem{extrareference2}
-\textcolor{red}{M. Ravnik, J. Fukuda, J.M. Yeomans, S. Zumer,
+\textcolor{red}{Ravnik, M., Fukuda, J.-I.,  Yeomans, J.M., and Zumer, S.,
 Confining blue phase colloids to thin layers,
@@
 
-\bibitem{permeation1} 	W. Helfrich,
+\bibitem{permeation1} 	Helfrich, W.,
 Capillary flow of cholesteric and smectic liquid crystals,
@@
 \bibitem{permeation2}
-A. Dupuis, D. Marenduzzo, E. Orlandini, J.~M. Yeomans,
+Dupuis, A.,  Marenduzzo, D.,  Orlandini, E., and Yeomans, J.M.,
 Rheology of cholesteric blue phases,
@@
 \bibitem{lubensky}
-M. Zapotocky, L. Ramos, P. Poulin, T.~C. Lubensky,  D.~A. Weitz,
+Zapotocky, M., Ramos, L., Poulin, P.,  Lubensky, T.C.,  and Weitz, D.A.,
 Particle-stabilized defect gel in cholesteric liquid crystals,
@@
 
-\bibitem{niek} N. Hijnen, T. A. Wood, D. Wilson, P. S. Clegg, 
+\bibitem{ramos} 
+\textcolor{red}{
+Ramos, L., Zapotocky, M.,  Lubensky, T.C., and  Weitz, D.A.,
+Rheology of defect networks in cholesteric liquid crystals,
+{\it Phys. Rev. E} {\bf 66} 031711 (2002).}
+
+\bibitem{niek} Hijnen, N., Wood, T.A., Wilson, D., and Clegg, P.S., 
 Self-organization of particles with planar surface anchoring in a cholesteric
@@
 \bibitem{flowswitch}
-O. Henrich, K. Stratford, P.V. Coveney, M.E. Cates, D. Marenduzzo,
+Henrich, O., Stratford, K., Coveney, P.V., Cates, M.E., and Marenduzzo, D.,
 Rheology of cubic blue phases,
@@
 
-\bibitem{optofluidics} D. Psaltis, S. R. Quake, C. Yang,
+\bibitem{optofluidics} Psaltis, D., Quake, S.R., Yang, C.,
 Developing optofluidic technology through the fusion of microfluidics and
@@
 
+\textcolor{red}{
+\bibitem{yoshida2009}
+Yoshida, H., et al.,
+Nanoparticle-stabilized cholesteric blue phases,
+\textit{Appl. Phys. Express}, \textbf{2}, 121501 (2009).
+%
+\bibitem{karatairi} Karatairi, E., et al.,
+% E. Karatairi, B. Ro\v{z}i\v{c}, Z. Kutnjak, V. Tzitzios, G. Nounesis,
+% G. Cordoyiannis, C. Glorieux, J. Thoen, S. Kralj,
+Nanoparticle-induced widening of the temperature range of
+liquid-crystalline blue phases, 
+\textit{Phys. Rev. E} \textbf{81}, 041703 (2010).
+%
+%\bibitem{rozic} B. Ro\v zi\v c,
+% B. Ro\v{z}i\v{c}, V. Tzitzios, E. Karatairi, U. Tkalec, G. Nounesis,
+% Z. Kutnjak, G. Cordoyiannis, R. Rosso, E.G. Virga, I. Mu\v{s}evi\{c},
+% S. Kralj
+%Theoretical and experimental study of the nanoparticle-driven blue phase
+%stabilisation,
+%\textit{Eur. Phys. J. E} \textbf{34}, 17 (2011).
+%
+\bibitem{dierking}
+Dierking, I., et al.,
+%I. Dierking, W. Blenkhorn, E. Credland, W. Drake, R. Kociuruba, B. Kayser,
+%T. Michael,
+Stabilising liquid crystalline blue phases,
+\textit{Soft Matter} \textbf{8}, 4355--4362 (2012).
+%
+\bibitem{draper}
+Draper, M., et al.
+Self-assembly and shape morphology of liquid crystalline gold metamaterials,
+\textit{Adv. Func. Mat.}, \textbf{21} 1260--1278 (2011).
+%
+\bibitem{milette2012}
+Milette, J., et al.,
+Reversible long-range network formation in gold nanoparticle-nematic
+liquid crystal composites,
+\textit{Soft Matter}, \textbf{8}, 173--179 (2012).
+%
+\bibitem{cordoyiannis} Cordoyiannis, G., et al.,
+% G. Cordoyiannis, V.S.R. Jampani, S. Kralj, S. Dhara, V. Tzitzios,
+% G. Basina, G. Nounesis, Z. Kutnjak, C.S.P. Tripathi, P. Losada-P\'{e}rez,
+% D. Jesenek, C. Glorieux, I. Mu\v{s}evi\v{c}, A. Zidan\v{s}ek,
+% H. Ameinitsch, J. Theon,
+Different modulated structures of topological defects stabilized by adaptive
+targeting nanoparticles,
+\textit{Soft Matter} \textbf{9}, 3956--3964 (2013).
+}
+
+%% METHODS
+
 \bibitem{deGennes}
-\textcolor{red}{P.-G. de Gennes, J. Prost,
+\textcolor{red}{de Gennes, P.-G., and  Prost, J.,
 {\it The Physics of Liquid Crystals} (Clarendon Press, 1995).}
@@
 \bibitem{beris}
-A. N. Beris, B. J. Edwards, 
+Beris, A.N., and  Edwards, B.J., 
 {\it Thermodynamics of Flowing Systems with Internal Microstructure}
@@
 \bibitem{ladd94}
-A.J.C. Ladd,
+Ladd, A.J.C.,
 \textit{J. Fluid Mech.} Numerical simulation of particulate suspensions via a
@@
 \bibitem{nguyen2002}
-N.-Q. Nguyen and A.J.C. Ladd,
+Nguyen, N.-Q., and Ladd, A.J.C.,
 Lubrication corrections for lattice Boltzmann simulations of particle
@@
 \bibitem{fournier2005}
-J.-B. Fournier and P. Galatola,
+Fournier, J.-B., and  Galatola, P.,
 Modeling planar degenerate wetting and anchoring in nematic liquid
@@
 \bibitem{juho}
-J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
+Lintuvuori, J.S., Marenduzzo, D., Stratford, K., and Cates, M.E.,
 Colloids in liquid crystals: a lattice Boltzmann study,
@@
 \bibitem{denniston}
-C. Denniston, E. Orlandini, J. M. Yeomans,
+Denniston, C.,  Orlandini, E., and  Yeomans, J.M.,
 Lattice Boltzmann simulations of liquid crystal hydrodynamics,
@@
 
-\bibitem{denniston2}
-C. Denniston, D. Marenduzzo, E. Orlandini, J.M. Yeomans,
-Lattice Boltzmann algorithm for three-dimensional liquid-crystal
-hydrodynamics,
-\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745--1754 (2004).
+%\bibitem{denniston2}
+%C. Denniston, D. Marenduzzo, E. Orlandini, J.M. Yeomans,
+%Lattice Boltzmann algorithm for three-dimensional liquid-crystal
+%hydrodynamics,
+%\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745--1754 (2004).
 
-\bibitem{codef} M. E. Cates, K. Stratford, R. Adhikari, P. Stansell,
-J.-C. Desplat, I. Pagonabarraga, A.J. Wagner,
-Simulating colloid hydrodynamics with lattice Boltzmann methods, 
-\textit{J. Phys. Condens. Matter} \textbf{16},  3903--3915 (2004).
+%\bibitem{codef} M. E. Cates, K. Stratford, R. Adhikari, P. Stansell,
+%J.-C. Desplat, I. Pagonabarraga, A.J. Wagner,
+%Simulating colloid hydrodynamics with lattice Boltzmann methods, 
+%\textit{J. Phys. Condens. Matter} \textbf{16},  3903--3915 (2004).
 
@@
 
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-\hline
-         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
-\hline 
-Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
-0.23, 2.3\\
-\hline
-Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
-0.23, 2.3 \\
-\hline
-Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
-\hline
-External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
-\hline
-\end{tabular}
-\end{center}
-\caption{\textbf{Free energy parameters used in the simulations.}
-Bulk BPI
-correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
-geometries are shown in Fig.~2; and the external field case is
-relevant for Fig.~3.}
-\label{table:params}
-\end{table}
-
 \end{document}

### 2014-03-04 trunk/papers/2012-sandwich/response.txt
updated response (Davide)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2324 40876441-3496-0410-a613-a2fca35b2ae2

@@
-Date: Mon, 17 Feb 2014 04:37:57 -0500
-From: c.huang@nature.com
-To: dmarendu@ph.ed.ac.uk
-Subject: Decision on manuscript NCOMMS-14-00180-T
+Dear Dr Huang,
 
-** Please ensure you delete the link to your author home page in this e-mail
-if you wish to forward it to your coauthors **
+many thanks for sending us the referee reports on our manuscript, 
+"Self-assembly of colloid-cholesteric composites: A route to switchable 
+optical materials". We are glad that both referees found our work interesting.
+We are also grateful for the suggestions for improvement, all of which
+we have now implemented in the revised version. We hope the revised
+manuscript can now be accepted for publication in Nature Communication.
 
-Dear Dr Marenduzzo,
+With best wishes,
 
-Your manuscript entitled "Self-assembly of colloid-cholesteric composites: A
-route to switchable optical materials" has now been seen by 2 referees,
-whose comments are appended below. You
-will see that, while they find your work of interest, they have raised
-points that need to be addressed before we can make a decision on
-publication.
+K. Stratford, O. Henrich, J. S. Lintuvuori, M. E. Cates and D. Marenduzzo
 
-We therefore invite you to revise and resubmit your manuscript, taking into
-account the points raised. At the same time, we ask that you ensure your
-manuscript complies with our format
-requirements, which are summarized in the following checklist:
-http://www.nature.com/ncomms/authors/ncomms_manuscript_checklist.pdf
+------------------------------------------------------------------
 
-Please use the following link to submit your revised manuscript, completed
-checklist and point-by-point response to the referees' comments (which
-should be in a separate document to any
-cover letter):
-http://mts-ncomms.nature.com/cgi-bin/main.plex?el=A6S2Lka4A1DfBL5I4A9v6k5dsV
-e8yORLGQzsElBAZ
+Point by point response to referees' comments
 
-** This url links to your confidential home page and associated information
-about manuscripts you may have submitted or be reviewing for us. If you wish
-to forward this email to co-authors,
-please delete the link to your homepage first **
+Reply to Reviewer #1:
 
-We hope to receive your revised paper within two months; please let us know
-if you aren't able to submit it within this time so that we can discuss the
-best way forward. If we don't hear
-from you, and the revision process takes significantly longer, we will close
-your file. In this event, we will still be happy to reconsider your paper at
-a later date, as long as nothing
-similar has been accepted for publication at Nature Communications or
-published elsewhere in the meantime.
-
-We look forward to hearing from you soon.
-
-Best regards,
-
-Congcong Huang, PhD
-
-Associate Editor
-Nature Communications
-
-Unit 10-11, 42F, The Center
-989 Changle Road, Xuhui District
-Shanghai, China, 200031
-Tel: +86-21-2422-5011
-Fax: +86-21-2422-5015
-www.nature.com/naturecommunications
-
-Reviewers' comments:
-
-Reviewer #1 (Remarks to the Author):
-
-The authors describe the results of a study of colloidal particles dispersed
+COMMENT:
+"The authors describe the results of a study of colloidal particles dispersed
 in cholesteric blue phases (BP) focusing on the effects of particle
@@
 points that need improvement before the publication.
-Below is the list of remarks.
+Below is the list of remarks."
 
-General remarks:
+RESPONSE:
+We thank the reviewer for her/his generally encouraging comments, and for
+recommending publication in Nature Communication subject to the resolution
+of some points. Below are our detailed reply to each of the reviewer's
+question, together with a description of how the comment has been dealt with
+in the revision.
 
-The term colloid is by the definition a dispersion of particles in a fluid.
+COMMENT:
+"The term colloid is by the definition a dispersion of particles in a fluid.
 Using the term also for the particles my lead to a confusion. In the
 manuscript often also a more appropriate word
-colloidal particle is used.
+colloidal particle is used."
 
-It would be important to add comments that would help experimentalist in
+RESPONSE: We have standardised on "colloidal particle(s)" or simply
+"particle(s)" where appropriate.
+
+COMMENT:
+"It would be important to add comments that would help experimentalist in
 designing experiments with colloidal blue phases and perhaps stimulate
 development of applications on the base of
-these materials.
+these materials."
 
-Particular remarks:
+RESPONSE: The issue of experimental guidance is addressed in detail
+in our reply to reviewer 2 below. In particular we have added a full
+paragraph in the discussion (page 3, highlighted in red), with explicit 
+reference to experimental studies on nanoparticle liquid crystal composites 
+(currently Refs. [34-39]).
 
-Page 2
+COMMENT:
+"Page 2: In the discussion of the physical understanding of trapping of colloidal particles in BP disclination networks it would be useful to refer to the
+initial study by Ravnik et al. in Faraday
+Discussions 2009 devoted to this topic."
 
-In the discussion of the physical understanding of trapping of colloidal
-particles in BP disclination networks it would be useful to refer to the
-initial study by Ravnik et al. in Faraday
-Discussions 2009 devoted to this topic.
+RESPONSE: Citation for Ravnik et al Faraday Discuss. 2010 (sic) is added
+[current reference 16].
 
-250 K is more than the total mesophase stability range!
+COMMENT: "250 K is more than the total mesophase stability range!"
 
-Page 3
+RESPONSE: A stability range for templated blue phases of -125 to
++125 degrees Celsius is reported by Castles et al. Nat. Mat. 2012.
+We have made this connection more clear by rewriting the relevant
+sentence, which now reads (see page 2, highlighted in red): 
+"However, recent advances in formulation have widened
+this stability range enormously [11,12,13,14]. Most recently,
+templated blue phases with a stability range of -125 to 125$^\circ$C
+have been reported [Castles], paving the way to applications of BPs
+in operational display devices."
 
-The statement "This is defined such that W > 0 gives preferential
+COMMENT: The statement "This is defined such that W > 0 gives preferential
 perpendicular orientation of the director field at the particle surface." is
@@
 
-For potential readers it would be useful to specify the particle radius in
+RESPONSE: We agree this statement was confusing. It has been replaced by
+a clearer statement (on page 4, highlighted in red): 
+"The anchoring strength is defined such that
+W > 0 imparts a preferred orientation to the director field at the
+particle surface. This orientation can be either normal to the
+surface or in a plane tangential to the surface, a property related
+to surface chemistry in experimental systems but which can be imposed
+at will in simulations."
+
+COMMENT:
+"For potential readers it would be useful to specify the particle radius in
 nanometers in the main text. How it compares to the nematic correlation
-length? How the radius influences the
-structures? How tiny the particles can still be that the method based on the
-Landau-de Gennes free energy is justified.
+length? How the radius influences the structures? How tiny the particles can 
+still be that the method based on the Landau-de Gennes free energy is 
+justified."
 
-Is WR/K = 2.8 justified for nanoparticles used in the study?
+RESPONSE: We have added in the main text (last paragraph introduction) a
+summary of the relevant dimensions in physical units:  "The simulations
+presented here are representative of a liquid crystal blue phase with unit
+cell size in the range of 100--500~nm and particles with a radius of
+around 50~nm." This new paragraph also serves the requirement for a brief
+summary of the results as set out in the manuscript checklist.
 
-Page 4:
+RESPONSE (cont.): As a coarse-grained description of the mesophase, the
+Landau-de Gennes  theory is appropriate as long as the size of the
+colloidal particle is much larger than the size of the mesogen,
+the latter being typically in the range of 1--10 nanometres. Our
+simulations meet this criteria (particle size about 50 nm) and the
+use of Landau-de Gennes is therefore justified.
 
-Can the G elastic modulus be estimated for modeled structures?
+COMMENT: "Is WR/K = 2.8 justified for nanoparticles used in the study?"
 
-Page 5:
+RESPONSE: The choice of WR/K = 2.3 in the simulations (not 2.8)
+is used to separate clearly a strong anchoring regime from a
+weak anchoring regime (w = 0.23). While typical experimental
+values for particles of this size may mean the weak regime is
+more common, we do not see that this precludes interest in a
+strong anchoring regime.
 
+COMMENT: "Page 4:
+Can the G elastic modulus be estimated for modeled structures?"
+
+RESPONSE: We note we did have, in the original manuscript, a
+paragraph which discusses estimates of the elastic modulus G
+based on analogy with experimentally observed systems. To the
+existing paragraph we have appended:
+"This figure is consistent with an estimate obtained by comparing
+the mechanical properties of the network of defects stabilised by
+colloidal particles with those of rubber [Ramos et al 2002] for
+the relevant simulation parameters (see Supplementary Information)."
+To the supplementary information we append: "For a Frank elastic
+constant of $K = 70$~pN and unit cell sizes in the range
+$\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
+K/\lambda^2 =$0.3--7~kPa." The relevant reference is added.
+
+COMMENT: "Page 5:
 Can annealing of the structures obtained with the thermal quench lead to
-ordered colloidal structures?
+ordered colloidal structures?"
 
-Is also in case of the disordered structures observed a change in the
-stability range of the colloidal BPs?
+RESPONSE: Our results suggest that this is probably not the case,
+although it is difficult to systematically explore all the possible
+quench protocol. That said, our data suggest that even starting with
+colloidal particles randomly dispersed in an *ordered* blue phase
+particles are able to distort the structure (see Fig. 1A, 1B); this
+suggests that time evolution if anything may lead to more disorder,
+rather than order.
 
-In the discussion of colloidal particles in BP thin layers it would be good
-to comment the study of
-Ravnik et al. in Soft Matter 2011 devoted the ordered structures of such a
-system.
+COMMENT: "Is also in case of the disordered structures observed a change in the
+stability range of the colloidal BPs?"
 
-Page 6:
+RESPONSE: Determining the stability range of the composite
+phases was beyond the scope of this paper. This would have required 
+an optimisation of the pitch length, which requires a large number 
+of additional simulations for each individual parameter set.
 
-It would be good to comment the difference between structures in the bulk
-and thin layer for particles with strong anchoring. Does this depend on the
-visualization or is more behind?
+COMMENT:
+"In the discussion of colloidal particles in BP thin layers it would be good
+to comment the study of Ravnik et al. in Soft Matter 2011 devoted 
+the ordered structures of such a system."
 
-Are the electric field induced changes irreversible only in case of
-colloidal dispersions?
+RESPONSE: A citation for Ravnik et al (2011) has been added [current
+reference 26].
 
-Page 7
+COMMENT:
+"Page 6: It would be good to comment the difference between structures in the 
+bulk and thin layer for particles with strong anchoring. Does this depend on the
+visualization or is more behind?"
 
-Reference 2 has a mistake.
+RESPONSE: We have added the following at the end of the section
+describing the confined geometry: "The degree of similarity between
+the bulk and confined structures is somewhat masked by the local
+change in particle density in the confined case. If one allows for
+this, Figs.~1 and~2 show the individual structural motifs are actually
+rather similar: templated lines for weak anchoring and clusters for
+strong anchoring."
 
-Supplement
+RESPONSE (cont.): One potential visualisation ambiguity is that the
+displayed box sizes are different in each of the figures (cf. particle
+size). An explicit statement has been added to each caption to make
+this more clear to the reader. ADD CAPTIONS?
 
-Segment 4.2:
+COMMENT:
+"Are the electric field induced changes irreversible only in case of
+colloidal dispersions?"
 
+RESPONSE: For the given parameters the field induced changes are only
+irreversible when colloids with sufficiently strong 
+anchoring conditions are present. The mesophase without colloids 
+undergoes a reversible ordering transition as described in reference [10].
+
+COMMENT: "Page 7 Reference 2 has a mistake."
+
+RESPONSE: The correct punctuation in the page numbers has been
+inserted in the reference for Wright and Mermin [current reference 9].
+
+COMMENT: "Supplement Segment 4.2:
 The statement "Wall anchoring is either normal or planar with W=K = 1 in
-simulation units" is unclear.
+simulation units" is unclear."
 
-References of the Supplement:
-There are mistakes in references 1 and 2.
+RESPONSE: This text has been replaced by a statement which is
+transparent: "For both normal and planar anchoring at the walls,
+the value of $W$ is the same as that used for the
+strongly anchoring particles."
 
-Reviewer #2 (Remarks to the Author):
+COMMENT:"References of the Supplement:
+There are mistakes in references 1 and 2."
 
-This paper, "Self-assembly of colloid-cholesteric composities: A route to
+RESPONSE: Reference 1, Wright and Mermin, was correct in
+the original supplement. Reference 2 has been updated to reflect
+the 2nd edition of de Gennes and Prost, which now appears in the
+main text [current reference 34].
+CHECK REFERENCE
+
+------------------------------------------------------------------------
+
+Reply to Reviewer 2
+
+COMMENT:
+"This paper, "Self-assembly of colloid-cholesteric composites: A route to
 switchable optical materials" reports simulations of the formation of
@@
 provide clear perspective on the modeled phases of these materials. The
-results are of interest and useful to
-experimentalists working in the field. The paper would be even more useful
-if it included references to actual materials, both the liquid crystal host
-and the added particles and perhaps
+results are of interest and useful to experimentalists working in the field. "
+
+RESPONSE:
+We are grateful to the reviewer for her/his overall positive view of our
+work. We also found her/his comments constructive and useful to improve
+our paper. We have addressed them as follows.
+
+COMMENT:
+"The paper would be even more useful if it included references to actual 
+materials, both the liquid crystal host and the added particles and perhaps
 even suggested a model system that would be a good starting point to study
-experimentally. In a similar vein, it would also be useful to address the
-practical issue of reproducibility, which
-has been a major problem with these materials. The theoretical results
-clearly indicate how important the formation conditions are to the final
-result. What is the authors sense of how easy
-it will be to reproducibly fabricate a device
-with a specific morphology and optical properties. This seems a likely
+experimentally. "
+
+RESPONSE: 
+"We have fully taken this point on board.
+In a new paragraph in the discussion section, we include six
+new references specifically aimed at recent studies on experimental
+nanoparticle liquid crystal composites. These are currently [34-39].
+The paragraph also discusses the issue of a model system, and suggests
+properties such a system may have to help match experiment and simulation."
+
+COMMENT:
+"In a similar vein, it would also be useful to address the
+practical issue of reproducibility, which has been a major problem 
+with these materials. The theoretical results clearly indicate how important 
+the formation conditions are to the final result.
+
+What is the authors sense of how easy it will be to reproducibly fabricate a 
+device with a specific morphology and optical properties. This seems a likely
 practical issue particularly with the rich variety of structures reported in
-the simulations.
+the simulations."
 
-** See NPG's author and referees' website at www.nature.com/authors for
-information about policies, services and author benefits
+RESPONSE: Clearly the issue of experimental reproducibility is a difficult
+one to be addressed by, of necessity idealised, simulation. However, one
+point our simulations do show is that careful characterisation of not
+only the particles and host fluid is required, but also of possible
+confinement effects and associated anchoring conditions. This we also
+address in our updated discussion, see the full paragraph highlighted in
+red on pages 7-8.
 
-This email has been sent through the NPG Manuscript Tracking System
-NY-610A-NPG&MTS
+Authors' note:
 
-Confidentiality Statement:
+* ADDITIONAL CHANGES TO MEET FORMAT REQUIREMENTS
+  cf Manuscript checklist
 
-This e-mail is confidential and subject to copyright. Any unauthorised use
-or disclosure of its contents is prohibited. If you have received this email
-in error please notify our Manuscript
-Tracking System Helpdesk team at http://npgapps.custhelp.com/app/ask.
+A number of other rearrangements have taken place to meet the
+format requirements as set out the manuscript checklist.
 
-Details of the confidentiality and pre-publicity policy may be found here
-http://www.nature.com/authors/policies/confidentiality.html
+Title
 
-Privacy Policy | Update Profile
+- The title has been adjusted to remove punctuation
+
+Authors
+
+- The full postal address has been added
+- The corresponding author has been identified (Marenduzzo).
+
+Abstract
+
+- The abstract has been shorted to meet the 150 word limit.
+  Citations have been removed to the main text.
+
+Main text
+
+- Symbols and units have been adjusted in places to come into style.
+
+Results
+
+- Subheadings of < 60 characters have been added
+
+- References to supplementary items have been adjusted to come into style.
+
+Methods
+
+- All "methods" material (originally in supplementary information) is now
+contained within the main paper. This includes relevant references.
+
+- Dividing subheadings of < 60 characters have been added.
+
+References
+
+- Formatting is adjusted to come into style.
+
+End notes
+
+- A statement of author contributions has been added.
+
+- A statement on conflicts of interest has been added.
+
+Legends
+
+- Figure legends have been adjusted to include a brief title.
+
+Tables
+
+- A title has been included.

### 2014-03-04 trunk/papers/2012-sandwich/CholColloids.tex
updated text (Davide)
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2325 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\title{Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
+\title{Self-assembly of colloid-cholesteric composites provides
+a route to switchable optical materials}
+
 \author{K. Stratford$^{*}$, O. Henrich$^{*}$, J.~S. Lintuvuori$^{*}$,\\
-M. E. Cates, D. Marenduzzo \\ 
-{\small SUPA, School of Physics and Astronomy, Edinburgh University,} \\
-\small{Edinburgh, EJ9 3JZ, UK}\\
-\small{$^{*}$ equally contributed to this work}}
+M. E. Cates, D. Marenduzzo$^\dagger$ \\ 
+SUPA, School of Physics and Astronomy, The King's Buildings,\\
+The University of Edinburgh, Edinburgh, EH9 3JZ, UK\\
+\small{$^{*}$ equally contributed to this work}\\
+\small{$^\dagger$ Corresponding author: dmarendu@ph.ed.ac.uk}
+}
 \date{}
@@
 To explore such metastability, and switching strategies between states, we require multi-unit-cell, time-dependent simulations.
-These go beyond previous studies of colloid-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloids in cholesterics~\cite{juho1,juho2}.
-Our methodology and simulation parameters are summarized in the Supplementary Information.
+These go beyond previous studies of colloidal-BP composites~\cite{miha} but follow comparable simulations of pure BPs~\cite{bp3,henrichfield,domaingrowth}, and isolated and dimeric colloidal particles in cholesterics~\cite{juho1,juho2}.
 
+\textcolor{red}{
+The simulations presented here are representative of a liquid crystal
+blue phase with unit cell size in the range of 100--500~nm \cite{mermin},
+and particles with a diameter of around 50~nm. The simulations
+address both the bulk structure, and the structure which forms when
+the colloidal dispersion is confined in a narrow sandwich of width
+comparable to the unit cell size~\cite{extrareference2}. It is seen
+that a wide variety of structure is possible, and may be influenced
+by parameters including the solid volume fraction, and the details
+of the anchoring at the colloid surface (and the at the surface of the
+confining walls). Further, evidence is found for metstable switching of
+this structure in an applied electric field. Details of the
+computational approach are set out in the Methods section, while
+simulation parameters are reported in the Supplementary Information.
+}
+
+\bigskip
+\noindent
+\textbf{\large Results}
+
+\noindent
+\textbf{Structure in bulk BPI.}
 We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles at random within it.
-Among dimensionless control parameters are the colloid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
-(This is defined such that $W>0$ gives preferential perpendicular orientation of the director field at the particle surface.) For the latter we choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
+Among dimensionless control parameters are the particle solid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
+\textcolor{red}{
+The anchoring strength is defined such that $W > 0$ imparts a preferred
+orientation to the director field at the particle surface. This orientation
+can be either normal to the surface or in a plane tangential to the surface,
+a property related to surface chemistry in experimental systems.% but which
+%can be imposed at will in simulations.
+We choose} $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
 This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.
@@
 structure is now akin to that of the self-quenched glass found in 
-colloid-nematic composites~\cite{tiffany} at much higher volume 
+colloidal-nematic composites~\cite{tiffany} at much higher volume 
 fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa.
+\textcolor{red}{
+This figure is consistent with an estimate obtained by comparing
+the mechanical properties of the network of defects stabilised by
+colloidal particles with those of rubber~\cite{ramos} for
+the relevant simulation parameters (see Supplementary Information).}
 
@@
 We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supplementary Information).
-The resulting colloid-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
+The resulting colloidal-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
 There are, however, significant effects of the confining walls.
-First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloids at both walls 
+First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloid particles at both walls 
 (see side-views in Fig.~2A-D). Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit. These clusters
-are similar to those reported experimentally for large colloids in a
+are similar to those reported experimentally for large colloid particles in a
 cholesteric (but non-BP) host~\cite{niek}.
@@
 double-stranded structures at larger $\phi$ (Fig.~2F).
-For larger $w$, we observe thick colloidal ``ropes" about 5 colloids across
-(Fig.~2E,H). These ropes undulate to follow the chiral structure of the 
+For larger $w$, we observe thick colloidal ``ropes" about 5 particles across
+(Fig.~2G,H). These ropes undulate to follow the chiral structure of the 
 underlying fluid.
 On increasing the density they develop branch points and finally again approach a percolating gel morphology.
-Note that planar wall anchoring is incompatible with the normal anchoring on the colloids (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
+Note that planar wall anchoring is incompatible with the normal anchoring on the colloidal particles (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
 
+\textcolor{red}{
+The degree of similarity between
+the bulk and confined structures is somewhat masked by the local
+change in particle density in the confined case. If one allows for
+this, Figs.~1 and~2 show the individual structural motifs are actually
+rather similar: templated lines for weak anchoring and clusters for
+strong anchoring.
+}
+
+\medskip
+\noindent
+\textbf{Switching in electric field.}
 An interesting question, relevant for applications to devices, is whether 
@@
 configuration before switching on the field, but gets stuck into a metastable 
-structure (Fig. S7), with residual anisotropy along the field switching 
+structure (Supplementary Fig.~7), with residual anisotropy along the field
+switching 
 direction. Cycling the field along the $x$ and $y$ directions leads to
 distinct metastable structures, which are retained after field removal
-(Fig. S8 and S9).
-%The resulting multistability is a prerequisite for new generation 
-%switchable electrooptic devices for use in energy-saving electronics-free
-%appliances, smart glass, and e-paper to cite just a few examples.
+(Supplementary Fig.~8 and Fig.~9).
 
-%FIELD SWITCHING PARA HERE
-\if{The existence of several locally stable competing solids with 
-very different symmetry and, arguably, physical properties, 
-is of interest to technology, as it makes these materials ideal
-candidate to build multistable devices, which can retain memory of
-more than one metastable state in the absence of an applied field.
-Multistability is indeed a must-have feature for new generation switchable 
-electrooptic devices for use in energy-saving electronics-free
-appliances, smart glass, and e-paper to cite just a few examples.
-}\fi
+\medskip
+\noindent
+\textbf{\large Discussion}
 
-The surprising wealth of colloid-BP composites obtained in Figs.~1-3 is the complex consequence of a simple competition between two self-assembly principles. 
-First, blue phases provide a three-dimensional template onto which colloidal particles are attracted so as to reduce the elastic stresses arising at the disclination cores~\cite{miha}.
-Second, colloidal intrusions in a liquid crystal raise the free energy locally; sharing this cost creates a generic tendency to aggregate into clusters~\cite{tiffany}.
-Which of these factors dominates depends on $w$, a dimensionless measure of the anchoring strength of the director field at the colloidal surface.
-For $w\ll 1$, the templating principle dominates.
-Colloidal crystals, gels or twisted rings can then arise.
-For $w \gg 1$, interparticle attractions defeat the templating tendency, and we predict disconnected aggregates, percolating gels, or helical colloidal ropes.
-In both cases, control is offered by the process route, colloid concentration, and the anchoring conditions at sample walls.
-We have also shown that electric fields can be used to switch between metastable states, providing a possible route to future multistable device applications.
-%This variety of metastable structures, with tunable properties that can combine the mechanical and optical features of colloids and liquid crystals~\cite{lavrentovich}, offers promise for the design of multistable liquid crystal devices, which are at the forefront of current technological developments in this field.
+\noindent
+The surprising wealth of colloidal-BP composites obtained in Figs.~1-3 is
+the complex consequence of a simple competition between two self-assembly
+principles. 
+First, blue phases provide a three-dimensional template onto which
+colloidal particles are attracted so as to reduce the elastic stresses
+arising at the disclination cores~\cite{miha}. Second, colloidal intrusions
+in a liquid crystal raise the free energy locally; sharing this cost creates
+a generic tendency to aggregate into clusters~\cite{tiffany}. Which of these
+factors dominates depends on $w$, a dimensionless measure of the anchoring
+strength of the director field at the colloidal surface. For $w\ll 1$, the
+templating principle dominates. Colloidal crystals, gels or twisted rings
+can then arise. For $w \gg 1$, interparticle attractions defeat the
+templating tendency, and we predict disconnected aggregates, percolating
+gels, or helical colloidal ropes. In both cases, control is offered by the
+process route, colloidal concentration, and the anchoring conditions at
+sample walls. We have also shown that electric fields can be used to
+switch between metastable states, providing a possible route to future
+multistable device applications.
+
+\textcolor{red}{
+How do these simulations relate to experiment? Several authors have
+used nanoparticles in the context of stabilising BPs
+\cite{yoshida2009,karatairi}. These
+nanoparticles are typically a few nanometres in size, in which case
+stabilisation is explained by the mechanism of defect removal.
+Experiments suggest this mechanism is also
+relevant for larger particles in the range of 40--70~nm 
+\cite{dierking}, which is consistent with the current simulations.
+However, the mechanism is observed to be less efficient for
+particles above 100~nm, where particles may start to disrupt
+longer-range liquid crystal order \cite{dierking}. While the field of
+self-assembly with nanoparticles is relatively new \cite{draper},
+nanoparticles have also been used in both nematic \cite{milette2012}
+and chiral liquid crystals \cite{cordoyiannis} as host fluid templates.
+The variation of structure seen
+in our simulations suggest that some careful characterisation of parameters
+such as surface anchoring of particles (which are often stabilised, or even
+functionalised with surface ligands), together with those associated
+with any confining surfaces or interfaces, is required to understand the
+final structure and to be able to reproduce it with a given experimental
+protocol. The identification of a model system which could allow
+unambiguous comparison between experimental systems and simulation
+would be extremely useful. Such a system might adopt particles of
+diameter in the range 50~nm or above, which allows a coarse-grained simulation
+approach of the type used here to be adopted, which can then capture the
+large-scale structure required to compare with experiment. In the same vein,
+detailed surface chemistry is difficult to represent in coarse-grained
+simulations, so heavily functionalised particles might be avoided.
+}
+
 We have not explored in this work switchability by flow \cite{flowswitch}, but this could again be inherited from the underlying BP dynamics \cite{adriano,permeation2}, in which case the same materials may also be relevant to the emerging technology of optofluidics~\cite{optofluidics}.
@@
 \bigskip
-\textbf{Acknowledgements}
+\noindent
+\textbf{\large Methods}
+
+Here, the simulation method is described. Note that
+all quantities are expressed in simulation units. Details
+of the simulation parameters, and their mapping to physical units,
+may be found in the Supplementary Information.
+
+\medskip
+\noindent
+\textbf{Thermodynamics.}
+The thermodynamics of the cholesteric liquid crystal can be described
+by means of a  Landau-de Gennes free energy functional $\cal F$,
+whose density is written as $f$:
+\begin{eqnarray}
+{\cal F}[{\bf Q}]&=&\int d^3{\bf r} f({\bf Q}({\bf r})).
+\end{eqnarray}
+This free energy density $f = f({\bf Q})$ may be expanded in powers of the
+order parameter ${\bf Q}$ and its gradients; ${\bf Q}$ is a traceless 
+and symmetric tensor which is denoted from now on in subscript
+notation as $Q_{\alpha\beta}$.
+The largest eigenvalue $q$ and corresponding eigenvector $n_\alpha$
+of $Q_{\alpha\beta}$ describe the local strength and major orientation axis
+of molecular order.
+The theory based on the tensor $Q_{\alpha\beta}$, rather than a theory based
+solely
+on the director field $n_\alpha ({\bf r})$, allows treatment of disclinations
+(defect lines) in whose cores $n_\alpha$ is undefined.
+
+Explicitly, the free energy density is:
+\begin{eqnarray}
+f(Q_{\alpha\beta}) &=& {\textstyle \frac{1}{2}} A_0
+ \left(1- {\textstyle \frac{1}{3}}\gamma \right)Q^2_{\alpha \beta}
+-{\textstyle \frac{1}{3}}A_0\gamma Q_{\alpha \beta} Q_{\beta \gamma}Q_{\gamma \alpha} 
++ {\textstyle \frac{1}{4}} A_0\gamma (Q^2_{\alpha \beta})^2  \nonumber\\
+&+& {\textstyle \frac{1}{2}} K(\varepsilon_{\alpha \gamma \delta}
+\partial_\gamma Q_{\delta \beta} + 2 q_0 Q_{\alpha \beta})^2
++ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2.
+\label{free}
+\end{eqnarray}
+Here, repeated indices are summed over, while terms of the form
+$Q^2_{\alpha\beta}$ should be expanded to read $Q_{\alpha\beta}Q_{\alpha\beta}$;
+$\varepsilon_{\alpha\gamma\delta}$ is the permutation tensor.
+The first three terms are a bulk free energy density whose overall scale
+is set by $A_0$ (discussed further below); $\gamma$ is a control parameter
+related to a reduced temperature. Varying the latter in the absence of
+chiral terms ($q_0=0$) gives an isotropic-nematic transition at
+$\gamma = 2.7$ with a mean-field spinodal instability at $\gamma = 3$.
+
+The remaining two terms of the free energy density in Eq.~\ref{free} describe
+distortions of the order parameter field. In theoretical work and when
+describing a generic rather than a specific system, it is 
+conventional~\cite{mermin,deGennes} to
+assume that splay, bend and twist deformations of the director are equally
+costly, that is, there is a single elastic constant $K$. The parameter $q_0$
+is related to the helical pitch length $p$ via $q_0=2\pi/p$,
+describing one full turn of the director in the cholesteric phase.
+
+There are two dimensionless numbers, which are commonly referred to as
+$\kappa$, the chirality, and  $\tau$, the reduced temperature
+\cite{mermin} which can be used to characterise the system. In terms of
+the parameters in the free energy density, these are:
+\begin{eqnarray}\label{cntrl-param} 
+\tau&=&\frac{27(1-\gamma/3)}{\gamma}\label{tau}\\
+\kappa&=&\sqrt{\frac{108\ K\, q_0^2}{A_0\, \gamma}}\label{kappa}.
+\end{eqnarray}
+If the free energy density Eq.~\ref{free} is made dimensionless, $\tau$
+appears as prefactor of the term  quadratic in $Q_{\alpha\beta}$,
+whereas $\kappa$
+quantifies the ratio between bulk and  gradient free energy terms. The
+chirality and reduced temperature may be used to characterise an equilibrium
+phase diagram of the blue phases in bulk cholesteric liquid crystals
+\cite{mermin,henrichfield}.
+
+The effect of an electric field can be modelled by including the following
+term in the bulk free energy density:
+\begin{equation}
+-\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
+\end{equation} 
+where $E_{\alpha}$ are the components of the electric field, and 
+$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
+crystal. The strength of the electric field is quantified via
+one further dimensionless number
+\begin{equation}
+{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
+\end{equation}
+The quantity ${\cal E}$ will be referred to as the reduced field strength.
+
+\medskip
+\noindent
+\textbf{Surface free energy.}
+In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
+free energy density is also present to represent the energetic cost of
+anchoring at a solid surface. If the preferred order parameter tensor
+at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
+density (per unit area) is
+\begin{equation}
+f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
+= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
+\end{equation}
+where $W$ is a constant determining the strength of the anchoring.
+The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
+discussed below. For a colloidal particle of radius $R$, the strength of
+the surface
+anchoring compared which the bulk fluid elastic constant may be quantified
+by the dimensionless parameter $WR/K$. For small values of this parameter,
+the presence of a particle surface should have little impact on the local
+fluid LC ordering.
+
+\medskip
+\noindent
+\textbf{Dynamics of the order parameter.}
+A framework for the dynamics of liquid crystals is provided by the 
+Beris-Edwards model \cite{beris}, in which the time evolution of the
+tensor order parameter obeys
+\begin{equation}
+\label{eom}
+\left(\partial_t+ u_\nu \partial_\nu \right) Q_{\alpha\beta} - S_{\alpha\beta}
+= \Gamma H_{\alpha\beta}.
+\end{equation}
+In the absence of flow, Eq.~\ref{eom} describes a relaxation towards
+equilibrium on a timescale determined by a collective rotational diffusion 
+constant $\Gamma$. This relaxation is driven by the molecular field
+$H_{\alpha\beta}$, which is the functional derivative of the free energy
+with respect to the order parameter \cite{beris}:
+\begin{equation}
+H_{\alpha\beta} = -\frac{\delta {\cal F}} {\delta Q_{\alpha\beta}} 
++ {\textstyle \frac{1}{3}} \delta_{\alpha\beta} 
+\mathrm{Tr}\left(\frac{\delta {\cal F}}{\delta Q_{\alpha\beta}}\right).
+\end{equation}
+
+The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
+derivative for rod-like molecules \cite{beris}. It couples the order
+parameter to the symmetric and antisymmetric parts of the velocity 
+gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
+The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
+$\Omega_{\alpha\beta}$ are defined as 
+\begin{equation}
+A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
+\end{equation}
+and
+\begin{equation}
+\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
+\end{equation}
+This full coupling term is then
+\begin{eqnarray}
+\label{coupling-term}
+S_{\alpha\beta}  &=&
+(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
+(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
++
+(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
+(\xi A_{\nu\beta} - \Omega_{\nu\beta})
+\nonumber\\ 
+&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
+Q_{\nu\nu} W_{\nu\nu}.
+\end{eqnarray}
+Here, $\xi$ is a material-dependent that controls the
+relative importance of rotational and elongational flow for molecular
+alignment, and determines in practice how the orientational order
+responds to a local shear flow. 
+In all cases, the value used here is $\xi = 0.7$, which is within
+the `flow aligning' regime, where molecules align at a fixed angle
+(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
+The value of the collective rotational diffusion used in all simulations is
+$\Gamma = 0.5$ in simulation units.
+
+\medskip
+\noindent
+\textbf{Hydrodynamics.}
+The momentum evolution obeys a Navier-Stokes equation driven by the divergence
+of a generalised stress $P_{\alpha\beta}$:
+\begin{equation}
+\label{nse}
+\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
+=\partial_\beta P_{\alpha\beta}.
+\end{equation}
+The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
+both viscous and thermodynamic components:
+\begin{eqnarray}
+P_{\alpha \beta}&=&
+- p_0 \delta_{\alpha \beta} 
++ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
+\nonumber\\
+&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
++ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
+-\xi \left(Q_{\alpha \gamma}
++ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
+\nonumber\\ 
+&+& 2 \xi  \left(Q_{\alpha \beta}
++ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
+-\partial_\alpha Q_{\gamma \nu}
+\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
+&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
+\end{eqnarray}
+A lattice Boltzmann (LB) flow solver is used, where the isotropic pressure
+$p_0$
+and viscous terms are managed directly by the solver (as in a simple Newtonian
+fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
+is treated as a local force density on the fluid. In all simulations
+reported the mean fluid density $\rho = 1$ and the viscosity
+$\eta = 0.01$ (simulations in confined geometry also used enhanced
+bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).
+
+\medskip
+\noindent
+\textbf{Surface boundary conditions for particles.}
+Hydrodynamic boundary conditions for solid objects are handled via
+the lattice Boltzmann solver. In particular, the standard method of
+bounce-back on links is used for both particles and confining walls
+\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
+are set out in \cite{juho}.
+
+In all the simulations reported here the preferred orientation of the
+director field at the particle surface is normal to the surface. The
+required nematic director at the particle surface in a given location
+$\hat{n}^0_\alpha$ may then be determined from geometry alone,
+and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
+\begin{equation}
+Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
+- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
+\label{eq:surf_q}
+\end{equation}
+The magnitude of surface order $q^0$ is set by
+\begin{equation}
+q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
++ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
+\label{eq:surf_q0}
+\end{equation}
+which corresponds to the bulk order in a defect-free nematically ordered 
+sample~\cite{denniston} (this is also very close to the value of the order 
+parameter in a cholesteric or in a blue phase away from disclinations).
+
+In all simulations the colloidal (hard sphere) radius is $R = 2.3$. To
+prevent particles overlapping at the hard sphere radius, an additional
+short range soft potential is included. This takes the form of
+$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
+The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
+units. Both potential and resulting force are smoothly matched to zero at
+a cut-off distance of $h_c = 0.25$ simulation units.
+
+\medskip
+\noindent
+\textbf{Surface boundary conditions for walls.}
+The confining solid walls, which are flat and stationary, have the
+same normal boundary condition as at the particle surface. In addition,
+degenerate planar anchoring is applied at the walls, where the
+preferred nematic director may adopt any orientation parallel
+to the surface. The preferred direction is determined explicitly
+by projecting the local fluid order parameter to the plane of the
+wall \cite{fournier2005}. Eq.~\ref{eq:surf_q} and Eq.~\ref{eq:surf_q0}
+are then used as before.
+
+To prevent the particles being forced into the wall, a correction
+to the lubrication force on the colloidal particle is added for very low
+particle
+surface to wall surface separations $h < 0.5$ lattice spacing. The
+correction is based on the analytic expression for the lubrication
+between a sphere and a flat surface.
+
+\pagebreak
+
+\begin{thebibliography}{99}
+
+\bibitem{wiresmiha}
+Ravnik, M., et al.,
+%M. Ravnik, M. Skarabot, S. Zumer, U. Tkalec, I. Poberaj,
+%D. Babic, N. Osterman, I. Musevic,
+Entangled nematic colloidal dimers and wires, 
+{\it Phys. Rev. Lett.} {\bf 99}, 247801 (2007).
+
+\bibitem{zumer}
+%I. Musevic, M. Skarabot, U. Tkalec, M. Ravnik, S. Zumer,
+Musevic, I., Skarabot, M., Tkalec, U., Ravnik, M., and Zumer, S.,
+Two-dimensional nematic colloidal crystals self-assembled by topological
+defects,
+{\it Science} {\bf 313}, 954--958 (2006).
+
+\bibitem{tanaka}
+Araki, T., and Tanaka, H.,
+Colloidal aggregation in a nematic liquid crystal: Topological arrest of
+particles by a single-stroke disclination line,
+{\it Phys. Rev. Lett.} {\bf 97}, 127801 (2006).
+
+\bibitem{tiffany}
+%T.~A. Wood, J.~S. Lintuvuori, A.~B. Schofield, D. Marenduzzo, W.~C.~K. Poon,
+Wood, T.A., Lintuvuori, J.S., Schofield, A.B, Marenduzzo, D., and
+Poon, W.C.K.,
+A Self-Quenched Defect Glass in a Colloid-Nematic Liquid Crystal Composite,
+{\it Science} {\bf 333}, 79--83 (2011).
+
+\bibitem{stark} Stark, H.,
+Physics of colloidal dispersions in nematic liquid crystals,
+{\it Phys. Rep.} {\bf 351}, 387--474 (2001).
+
+\bibitem{abbott}
+%I.~H. Lin, D.~S. Miller, P.~J. Bertics, C.~J. Murphy, J.~J. de Pablo,
+%N.~L. Abbott,
+Lin, I.H., et al.,
+Endotoxin-Induced Structural Transformations in Liquid Crystalline Droplet,
+{\it Science} {\bf 332}, 1297--1300 (2011).
+
+\bibitem{colloiddevice}
+%K. Chari, C. M. Rankin, D. M. Johnson, T. N. Blanton, R. G. Capurso,
+Chari, K., Rankin, C.M., Johnson, D.M, Blanton, T.N., and Capurso, R.G.,
+Single-substrate cholesteric liquid crystal displays by colloidal
+self-assembly,
+{\it Appl. Phys. Lett.} {\bf 88}, 043502 (2006).
+
+\bibitem{tanakanatmat}
+Araki, T., Buscaglia, M., Bellini, T., and Tanaka, H.,
+Memory and topological frustration in nematic liquid crystals confined
+in porous materials, {\it Nat. Mat.} {\bf 10}, 303--309 (2011).
+
+\bibitem{mermin} Wright, D.C., and Mermin, N.D.,
+Crystalline liquids: The blue phases, {\it Rev. Mod. Phys.} {\bf 61},
+\textcolor{red}{385--432} (1989).
+
+\bibitem{bp3} Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
+ Structure of Blue Phase III of Cholesteric Liquid Crystals,
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
+
+\bibitem{kikuchi}
+Kikuchi, H., Yokota, M., Hisakado, Y., Yang, H., and Kajiyama, T.,
+Polymer-stabilized liquid crystal blue phases,
+{\it Nat. Mat.} {\bf 1}, 64--68 (2002).
+
+\bibitem{bplasers}
+Cao, W.Y., Munoz, A., Palffy-Muhoray, P., and  Taheri, B.,
+Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
+phase II,
+{\it Nat. Mat.} {\bf 1}, 111--113 (2002).
+
+\bibitem{coleswidetrange}
+Coles, H.J., and  Pivnenko, M.N.,
+Liquid crystal 'blue phases' with a wide temperature range,
+{\it Nature} {\bf 436}, 997--1000 (2005).
+
+\bibitem{bpdevice}
+%Z.~B. Ge, S. Gauza, M.~Z. Jiao, H.~Q. Xianyu, S.~T. Wu,
+Ge, Z.B., Gauza, S., Jiao, M.Z., Xianyu, H.Q., and  Wu, S.T.,
+ Electro-optics of polymer-stabilized blue phase liquid crystal displays,
+{\it Appl. Phys. Lett.} {\bf 94}, 101104 (2009).
+
+\bibitem{coles}
+%F. Castles, F.~V. Day, S.~M. Morris, D-H. Ko, D.~J. Gardiner, M.~M. Qasim,
+%S. Nosheen, P.~J.~W. Hands, S.~S. Choi, R.~H. Friend, H.~J. Coles,
+Castles, F., et al.,
+Blue-phase templated fabrication of three-dimensional nanostructures for
+photonic applications, {\it Nat. Mat.} {\bf 11}, 599--603 (2012).
+
+\bibitem{extrareference1}
+\textcolor{red}{Ravnik, M., Alexander, G.P., Yeomans, J.M., and Zumer, S.,
+Mesoscopic modelling of colloids in chiral nematics,
+{\it Faraday Discuss.} {\bf 144}, 159--169 (2010).}
+
+\bibitem{miha}
+Ravnik, M., Alexander, G.P.,  Yeomans, J.M., and  Zumer, S.,
+Three-dimensional colloidal crystals in liquid crystalline 
+blue phases, {\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5188--5192 (2011).
+
+\bibitem{lavrentovich} Lavrentovich, O.D.,
+ Liquid crystals, photonic crystals, metamaterials and transformation optics, 
+{\it Proc. Natl. Acad. Sci. USA} {\bf 108}, 5143--5144 (2011).
+
+\bibitem{adriano}
+Tiribocchi, A., Gonnella, G., Marenduzzo, D., Orlandini, E., and Salvadore, F.,
+Bistable Defect Structures In Blue Phase Devices,
+ {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
+
+\bibitem{fukuda}
+Fukuda, J.-I., and Zumer, S.,
+Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal,
+{\it Nat. Commun.} {\bf 2}, doi:10.1038/ncomms1250 (2011).
+
+\bibitem{henrichfield} 
+Henrich, O., Marenduzzo, D., Stratford, K., and  Cates, M.E.,
+Thermodynamics of blue phases in electric fields,
+{\it Phys. Rev. E} {\bf 81}, 031706 (2010).
+
+\bibitem{epaper}
+Heikenfeld, J., Drzaic, P., Yeo, J.S., and Koch, T.,
+A critical review of the present and future prospects for electronic paper,
+{\it J. Soc. Inf. Display} {\bf 19}, 129--156 (2011).
+
+\bibitem{domaingrowth}
+Henrich, O., Stratford, K., Marenduzzo, D., and  Cates, M.E.,
+Ordering dynamics of blue phases entails kinetic stabilization of amorphous
+networks, {\it Proc. Natl. Acad. Sci. USA} {\bf 107}, 13212--13215 (2010).
+
+\bibitem{juho1}
+Lintuvuori, J.S.,  Stratford, K., Cates, M.E., and Marenduzzo, D.,
+Colloids in cholesterics: size-dependent defects and non-Stokesian
+microrheology, {\it Phys. Rev. Lett.} {\bf 105}, 178302 (2010).
+
+\bibitem{juho2}
+Lintuvuori, J.S.,  Stratford, K.,  Cates, M.E., and  Marenduzzo, D.,
+Self-assembly and nonlinear dynamics of dimeric colloidal rotors in 
+cholesterics, {\it Phys. Rev. Lett.} {\bf 107}, 267802 (2011).
+
+\bibitem{extrareference2}
+\textcolor{red}{Ravnik, M., Fukuda, J.-I.,  Yeomans, J.M., and Zumer, S.,
+Confining blue phase colloids to thin layers,
+{\it Soft Matter} {\bf 7}, 10144--10150 (2011).}
+
+\bibitem{permeation1} 	Helfrich, W.,
+Capillary flow of cholesteric and smectic liquid crystals,
+{\it Phys. Rev. Lett.} {\bf 23}, 372--374 (1969).
+
+\bibitem{permeation2}
+Dupuis, A.,  Marenduzzo, D.,  Orlandini, E., and Yeomans, J.M.,
+Rheology of cholesteric blue phases,
+{\it Phys. Rev. Lett.} {\bf 95}, 097801 (2005).
+
+\bibitem{lubensky}
+Zapotocky, M., Ramos, L., Poulin, P.,  Lubensky, T.C.,  and Weitz, D.A.,
+Particle-stabilized defect gel in cholesteric liquid crystals,
+{\it Science} {\bf 283}, 209--212 (1999).
+
+\bibitem{ramos} 
+\textcolor{red}{
+Ramos, L., Zapotocky, M.,  Lubensky, T.C., and  Weitz, D.A.,
+Rheology of defect networks in cholesteric liquid crystals,
+{\it Phys. Rev. E} {\bf 66} 031711 (2002).}
+
+\bibitem{niek} Hijnen, N., Wood, T.A., Wilson, D., and Clegg, P.S., 
+Self-organization of particles with planar surface anchoring in a cholesteric
+liquid crystal, {\it Langmuir} {\bf 26}, 13502--13510 (2010).
+
+\bibitem{flowswitch}
+Henrich, O., Stratford, K., Coveney, P.V., Cates, M.E., and Marenduzzo, D.,
+Rheology of cubic blue phases,
+{\it Soft Matter} doi:10.1039/C3SM50228G (2013).
+
+\bibitem{optofluidics} Psaltis, D., Quake, S.R., Yang, C.,
+Developing optofluidic technology through the fusion of microfluidics and
+optics {\it Nature} {\bf 442}, 381--386 (2006).
+
+\textcolor{red}{
+\bibitem{yoshida2009}
+Yoshida, H., et al.,
+Nanoparticle-stabilized cholesteric blue phases,
+\textit{Appl. Phys. Express}, \textbf{2}, 121501 (2009).
+%
+\bibitem{karatairi} Karatairi, E., et al.,
+% E. Karatairi, B. Ro\v{z}i\v{c}, Z. Kutnjak, V. Tzitzios, G. Nounesis,
+% G. Cordoyiannis, C. Glorieux, J. Thoen, S. Kralj,
+Nanoparticle-induced widening of the temperature range of
+liquid-crystalline blue phases, 
+\textit{Phys. Rev. E} \textbf{81}, 041703 (2010).
+%
+%\bibitem{rozic} B. Ro\v zi\v c,
+% B. Ro\v{z}i\v{c}, V. Tzitzios, E. Karatairi, U. Tkalec, G. Nounesis,
+% Z. Kutnjak, G. Cordoyiannis, R. Rosso, E.G. Virga, I. Mu\v{s}evi\{c},
+% S. Kralj
+%Theoretical and experimental study of the nanoparticle-driven blue phase
+%stabilisation,
+%\textit{Eur. Phys. J. E} \textbf{34}, 17 (2011).
+%
+\bibitem{dierking}
+Dierking, I., et al.,
+%I. Dierking, W. Blenkhorn, E. Credland, W. Drake, R. Kociuruba, B. Kayser,
+%T. Michael,
+Stabilising liquid crystalline blue phases,
+\textit{Soft Matter} \textbf{8}, 4355--4362 (2012).
+%
+\bibitem{draper}
+Draper, M., et al.
+Self-assembly and shape morphology of liquid crystalline gold metamaterials,
+\textit{Adv. Func. Mat.}, \textbf{21} 1260--1278 (2011).
+%
+\bibitem{milette2012}
+Milette, J., et al.,
+Reversible long-range network formation in gold nanoparticle-nematic
+liquid crystal composites,
+\textit{Soft Matter}, \textbf{8}, 173--179 (2012).
+%
+\bibitem{cordoyiannis} Cordoyiannis, G., et al.,
+% G. Cordoyiannis, V.S.R. Jampani, S. Kralj, S. Dhara, V. Tzitzios,
+% G. Basina, G. Nounesis, Z. Kutnjak, C.S.P. Tripathi, P. Losada-P\'{e}rez,
+% D. Jesenek, C. Glorieux, I. Mu\v{s}evi\v{c}, A. Zidan\v{s}ek,
+% H. Ameinitsch, J. Theon,
+Different modulated structures of topological defects stabilized by adaptive
+targeting nanoparticles,
+\textit{Soft Matter} \textbf{9}, 3956--3964 (2013).
+}
+
+%% METHODS
+
+\bibitem{deGennes}
+\textcolor{red}{de Gennes, P.-G., and  Prost, J.,
+{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).}
+
+\bibitem{beris}
+Beris, A.N., and  Edwards, B.J., 
+{\it Thermodynamics of Flowing Systems with Internal Microstructure}
+(Oxford University Press, 1994).
+
+\bibitem{ladd94}
+Ladd, A.J.C.,
+\textit{J. Fluid Mech.} Numerical simulation of particulate suspensions via a
+discretized Boltzmann equation. Part 1. Theoretical foundation,
+\textbf{271}, 285--309 (1994).
+
+\bibitem{nguyen2002}
+Nguyen, N.-Q., and Ladd, A.J.C.,
+Lubrication corrections for lattice Boltzmann simulations of particle
+suspensions,
+\textit{Phys. Rev. E} \textbf{66} 046708 (2002).
+
+\bibitem{fournier2005}
+Fournier, J.-B., and  Galatola, P.,
+Modeling planar degenerate wetting and anchoring in nematic liquid
+crystals,
+\textit{Europhys. Lett.} \textbf{72}, 403--409 (2005).
+
+\bibitem{juho}
+Lintuvuori, J.S., Marenduzzo, D., Stratford, K., and Cates, M.E.,
+Colloids in liquid crystals: a lattice Boltzmann study,
+\textit{J. Mater. Chem.} \textbf{20}, 10547--10552 (2010).
+
+\bibitem{denniston}
+Denniston, C.,  Orlandini, E., and  Yeomans, J.M.,
+Lattice Boltzmann simulations of liquid crystal hydrodynamics,
+\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
+
+%\bibitem{denniston2}
+%C. Denniston, D. Marenduzzo, E. Orlandini, J.M. Yeomans,
+%Lattice Boltzmann algorithm for three-dimensional liquid-crystal
+%hydrodynamics,
+%\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745--1754 (2004).
+
+%\bibitem{codef} M. E. Cates, K. Stratford, R. Adhikari, P. Stansell,
+%J.-C. Desplat, I. Pagonabarraga, A.J. Wagner,
+%Simulating colloid hydrodynamics with lattice Boltzmann methods, 
+%\textit{J. Phys. Condens. Matter} \textbf{16},  3903--3915 (2004).
+
+\end{thebibliography}
+\noindent
+\textbf{\large Acknowledgements}
+
+\noindent
 We acknowledge support by UK EPSRC grants EP/J007404/1 (KS and JSL), 
-EP/G036136/1 (OH), and EP/F054750/1 (KS), respectively. 
+EP/G036136/1 (OH), and EP/I030298/1 (KS and JSL).
 MEC is supported by the Royal Society.
@@
 [For the full parameter list used to generate Figs.~1-3 see Supplementary
-Information.] For clarity, only a portion of the simulation box (a quarter of
-each axis in size) is shown; the full structures are shown in Supplementary
-Fig. S2 and S3.}
+Information.] For clarity, only a portion of the simulation box is shown
+(one eighth of the total volume); the full structures are shown in
+Supplementary Figs.~2 and~3.}
 \end{figure}
@@
 (D,H) $w=2.3$ and $\phi=2\%$.
-As in Fig. 1, only a portion of the simulation box is shown for clarity;
-the full structures are shown in Supplementary Fig. S5 and S6.}
+As in Fig.~1, only one quarter of the simulation box is shown for clarity;
+the full structures are shown in Supplementary Figs.~5 and~6.}
 \end{figure}

### 2014-03-05 papers/2012-sandwich/response.txt
Removed editorial checks
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2326 40876441-3496-0410-a613-a2fca35b2ae2

@@
-Point by point response
+Dear Dr Huang,
 
-Reviewers' comments:
+many thanks for sending us the referee reports on our manuscript, 
+"Self-assembly of colloid-cholesteric composites: A route to switchable 
+optical materials". We are glad that both referees found our work interesting.
+We are also grateful for the suggestions for improvement, all of which
+we have now implemented in the revised version. We hope the revised
+manuscript can now be accepted for publication in Nature Communication.
 
-Reviewer #1 (Remarks to the Author):
+With best wishes,
 
-The authors describe the results of a study of colloidal particles dispersed
+K. Stratford, O. Henrich, J. S. Lintuvuori, M. E. Cates and D. Marenduzzo
+
+------------------------------------------------------------------
+
+Point by point response to referees' comments
+
+Reply to Reviewer #1:
+
+COMMENT:
+"The authors describe the results of a study of colloidal particles dispersed
 in cholesteric blue phases (BP) focusing on the effects of particle
@@
 points that need improvement before the publication.
-Below is the list of remarks.
+Below is the list of remarks."
 
-General remarks:
+RESPONSE:
+We thank the reviewer for her/his generally encouraging comments, and for
+recommending publication in Nature Communication subject to the resolution
+of some points. Below are our detailed reply to each of the reviewer's
+question, together with a description of how the comment has been dealt with
+in the revision.
 
-The term colloid is by the definition a dispersion of particles in a fluid.
+COMMENT:
+"The term colloid is by the definition a dispersion of particles in a fluid.
 Using the term also for the particles my lead to a confusion. In the
 manuscript often also a more appropriate word
-colloidal particle is used.
+colloidal particle is used."
 
@@
 
-It would be important to add comments that would help experimentalist in
+COMMENT:
+"It would be important to add comments that would help experimentalist in
 designing experiments with colloidal blue phases and perhaps stimulate
 development of applications on the base of
-these materials.
+these materials."
 
 RESPONSE: The issue of experimental guidance is addressed in detail
-in our reply to reviewer 2 below.
+in our reply to reviewer 2 below. In particular we have added a full
+paragraph in the discussion (page 3, highlighted in red), with explicit 
+reference to experimental studies on nanoparticle liquid crystal composites 
+(currently Refs. [34-39]).
 
-Particular remarks:
-
-Page 2
-
-In the discussion of the physical understanding of trapping of colloidal
-particles in BP disclination networks it would be useful to refer to the
+COMMENT:
+"Page 2: In the discussion of the physical understanding of trapping of colloidal particles in BP disclination networks it would be useful to refer to the
 initial study by Ravnik et al. in Faraday
-Discussions 2009 devoted to this topic.
+Discussions 2009 devoted to this topic."
 
@@
 
-250 K is more than the total mesophase stability range!
+COMMENT: "250 K is more than the total mesophase stability range!"
 
@@
 We have made this connection more clear by rewriting the relevant
-sentence: "However, recent advances in formulation have widened
+sentence, which now reads (see page 2, highlighted in red): 
+"However, recent advances in formulation have widened
 this stability range enormously [11,12,13,14]. Most recently,
@@
 
-Page 3
-
-The statement "This is defined such that W > 0 gives preferential
+COMMENT: The statement "This is defined such that W > 0 gives preferential
 perpendicular orientation of the director field at the particle surface." is
@@
 
-RESPONSE: We agree this statement is confusing. It has been replaced by
-a clearer statement: "The anchoring strength is defined such that
+RESPONSE: We agree this statement was confusing. It has been replaced by
+a clearer statement (on page 4, highlighted in red): 
+"The anchoring strength is defined such that
 W > 0 imparts a preferred orientation to the director field at the
@@
 
-For potential readers it would be useful to specify the particle radius in
+COMMENT:
+"For potential readers it would be useful to specify the particle radius in
 nanometers in the main text. How it compares to the nematic correlation
-length? How the radius influences the
-structures? How tiny the particles can still be that the method based on the
-Landau-de Gennes free energy is justified.
+length? How the radius influences the structures? How tiny the particles can 
+still be that the method based on the Landau-de Gennes free energy is 
+justified."
 
@@
 
-Is WR/K = 2.8 justified for nanoparticles used in the study?
+COMMENT: "Is WR/K = 2.8 justified for nanoparticles used in the study?"
 
@@
 
-Page 4:
-
-Can the G elastic modulus be estimated for modeled structures?
+COMMENT: "Page 4:
+Can the G elastic modulus be estimated for modeled structures?"
 
@@
 
-Page 5:
+COMMENT: "Page 5:
+Can annealing of the structures obtained with the thermal quench lead to
+ordered colloidal structures?"
 
-Can annealing of the structures obtained with the thermal quench lead to
-ordered colloidal structures?
+RESPONSE: Our results suggest that this is probably not the case,
+although it is difficult to systematically explore all the possible
+quench protocol. That said, our data suggest that even starting with
+colloidal particles randomly dispersed in an *ordered* blue phase
+particles are able to distort the structure (see Fig. 1A, 1B); this
+suggests that time evolution if anything may lead to more disorder,
+rather than order.
 
-RESPONSE: Our results suggest that this is difficult to achieve, but
-generally possible. It depends sensitively on the 
-anchoring strength and the size of the colloidal particles. 
-Strong anchoring conditions lead to a significant 
-distortion of the disclination lines, whereas in the weak anchoring 
-regime the forces that drive the particles towards the 
-disclination lines remain rather small.
-
-Is also in case of the disordered structures observed a change in the
-stability range of the colloidal BPs?
+COMMENT: "Is also in case of the disordered structures observed a change in the
+stability range of the colloidal BPs?"
 
@@
 
-In the discussion of colloidal particles in BP thin layers it would be good
-to comment the study of
-Ravnik et al. in Soft Matter 2011 devoted the ordered structures of such a
-system.
+COMMENT:
+"In the discussion of colloidal particles in BP thin layers it would be good
+to comment the study of Ravnik et al. in Soft Matter 2011 devoted 
+the ordered structures of such a system."
 
@@
 
-Page 6:
-
-It would be good to comment the difference between structures in the bulk
-and thin layer for particles with strong anchoring. Does this depend on the
-visualization or is more behind?
+COMMENT:
+"Page 6: It would be good to comment the difference between structures in the 
+bulk and thin layer for particles with strong anchoring. Does this depend on the
+visualization or is more behind?"
 
@@
 displayed box sizes are different in each of the figures (cf. particle
-size). An explicit statement has been added to each caption to make
-this more clear to the reader. ADD CAPTIONS?
+size). An explicit statement has been added to each caption in the main
+text to make this more clear to the reader.
 
-Are the electric field induced changes irreversible only in case of
-colloidal dispersions?
+COMMENT:
+"Are the electric field induced changes irreversible only in case of
+colloidal dispersions?"
 
@@
 undergoes a reversible ordering transition as described in reference [10].
-CHECK REFERENCE
 
-Page 7
-
-Reference 2 has a mistake.
+COMMENT: "Page 7 Reference 2 has a mistake."
 
@@
 
-Supplement
-
-Segment 4.2:
-
+COMMENT: "Supplement Segment 4.2:
 The statement "Wall anchoring is either normal or planar with W=K = 1 in
-simulation units" is unclear.
+simulation units" is unclear."
 
@@
 
-References of the Supplement:
-There are mistakes in references 1 and 2.
+COMMENT:"References of the Supplement:
+There are mistakes in references 1 and 2."
 
@@
 the 2nd edition of de Gennes and Prost, which now appears in the
-main text [current reference 34].
-CHECK REFERENCE
+main text [current reference 40].
 
-Reviewer #2 (Remarks to the Author):
+------------------------------------------------------------------------
 
-This paper, "Self-assembly of colloid-cholesteric composites: A route to
+Reply to Reviewer 2
+
+COMMENT:
+"This paper, "Self-assembly of colloid-cholesteric composites: A route to
 switchable optical materials" reports simulations of the formation of
@@
 provide clear perspective on the modeled phases of these materials. The
-results are of interest and useful to
-experimentalists working in the field.
+results are of interest and useful to experimentalists working in the field. "
 
-The paper would be even more useful if it included references to actual 
+RESPONSE:
+We are grateful to the reviewer for her/his overall positive view of our
+work. We also found her/his comments constructive and useful to improve
+our paper. We have addressed them as follows.
+
+COMMENT:
+"The paper would be even more useful if it included references to actual 
 materials, both the liquid crystal host and the added particles and perhaps
 even suggested a model system that would be a good starting point to study
-experimentally.
+experimentally. "
 
-RESPONSE: In a new paragraph in the discussion section, we include six
+RESPONSE: 
+"We have fully taken this point on board.
+In a new paragraph in the discussion section, we include six
 new references specifically aimed at recent studies on experimental
@@
 The paragraph also discusses the issue of a model system, and suggests
-properties such a system may have to help match experiment and simulation.
+properties such a system may have to help match experiment and simulation."
 
-In a similar vein, it would also be useful to address the
+COMMENT:
+"In a similar vein, it would also be useful to address the
 practical issue of reproducibility, which has been a major problem 
@@
 
-What is the authors sense of how easy it will be to reproducibly fabricate a device
-with a specific morphology and optical properties. This seems a likely
+What is the authors sense of how easy it will be to reproducibly fabricate a 
+device with a specific morphology and optical properties. This seems a likely
 practical issue particularly with the rich variety of structures reported in
-the simulations.
+the simulations."
 
@@
 confinement effects and associated anchoring conditions. This we also
-address in our updated discussion.
+address in our updated discussion, see the full paragraph highlighted in
+red on pages 7-8.
 
-The entire paragraph reads: ADD WHEN FINALISED
+Authors' note:
 

### 2014-03-05 papers/2012-sandwich/CholColloids.tex
Removed editorial checks
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2326 40876441-3496-0410-a613-a2fca35b2ae2

@@
 orientation to the director field at the particle surface. This orientation
-can be either normal to the surface or in a plane tengential to the surface,
-a property related to surface chemistry in experimental systems but which
-can be imposed at will in simulations.
+can be either normal to the surface or in a plane tangential to the surface,
+a property related to surface chemistry in experimental systems.% but which
+%can be imposed at will in simulations.
 We choose} $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
-This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.
+This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion~$KR$.
 
@@
 would be extremely useful. Such a system might adopt particles of
-dimeter in the range 50~nm or above, which allows a coarse-grained simulation
+diameter in the range 50~nm or above, which allows a coarse-grained simulation
 approach of the type used here to be adopted, which can then capture the
@@
 Information.] For clarity, only a portion of the simulation box is shown
-(one eighth of the total volume); the full structures are shown in
+\textcolor{red}{its extent being $\simeq 28R$}
+(one eighth of the total volume);
+the full structures are shown in
 Supplementary Figs.~2 and~3.}
@@
 (D,H) $w=2.3$ and $\phi=2\%$.
-As in Fig.~1, only one quarter of the simulation box is shown for clarity;
+As in Fig.~1, only one quarter of the simulation box is shown for clarity
+\textcolor{red}{(its horizontal extent being $\simeq 56R$ and vertical
+extent $\simeq 24R$ in the narrow direction)};
 the full structures are shown in Supplementary Figs.~5 and~6.}
@@
 fraction of $\phi=5\%$ (see Supplementary Information for full list of
-parameters). }
+parameters).
+\textcolor{red}{The full simulation with a box size of $\simeq 56R$ is shown.}}
 \end{figure}

### 2014-03-05 papers/2012-sandwich/response.txt
Final updates for response version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2327 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-many thanks for sending us the referee reports on our manuscript, 
+Many thanks for sending us the referee reports on our manuscript, 
 "Self-assembly of colloid-cholesteric composites: A route to switchable 
@@
 we have now implemented in the revised version. We hope the revised
-manuscript can now be accepted for publication in Nature Communication.
+manuscript can now be accepted for publication in Nature Communications.
 
@@
 K. Stratford, O. Henrich, J. S. Lintuvuori, M. E. Cates and D. Marenduzzo
+
+As per the manuscript checklist, we include the captions for
+Supplementary Movies here:
+
+Supplementary Movie 1:
+The final state of the simulation as shown in Fig.~1A. The movie
+rotates the configuration to reveal more clearly its structure.
+
+Supplementary Movie 2:
+The final state of the simulation as shown in Fig.~1B. The movie
+rotates the configuration to reveal more clearly its structure.
+
+Supplementary Movie 3:
+The final state of the simulation as shown in Fig.~1C. The movie
+rotates the configuration to reveal more clearly its structure.
+
+Supplementary Movie 4:
+The final state of the simulation as shown in Fig.~1D. The movie
+rotates the configuration to reveal more clearly its structure.
+
+Supplementary Movie 5:
+The configuration after the initial relaxation phase in Fig.~3. The movie
+shows a rotation of the configuration. The distribution of the colloidal
+particles is isotropic, with the particles forming clusters and having a
+tendency to occupy positions at the intersection of the BPIII disclination
+lines.
+
+Supplementary Movie 6:
+Configuration in a homogeneous electric field with strength
+corresponding to ${\cal E}=0.8$. The movie
+shows a rotation of the configuration.
+A hexagonal pattern emerges where the particles form chains
+situated at the vertices of the honeycomb-like structure. \\
 
@@
 RESPONSE: 
-"We have fully taken this point on board.
+We have fully taken this point on board.
 In a new paragraph in the discussion section, we include six
@@
 The paragraph also discusses the issue of a model system, and suggests
-properties such a system may have to help match experiment and simulation."
+properties such a system may have to help match experiment and simulation.
 

### 2014-03-05 trunk/papers/2012-sandwich/CholColloids.tex
Final updates for response version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2327 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \title{Self-assembly of colloid-cholesteric composites provides
-a route to switchable optical materials}
+a possible route to switchable optical materials}
 
@@
 of the anchoring at the colloid surface (and the at the surface of the
-confining walls). Further, evidence is found for metstable switching of
+confining walls). Further, evidence is found for metastable switching of
 this structure in an applied electric field. Details of the
@@
 We choose} $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
-This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion $KR$.
+This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion~$KR$.
 
@@
 Fig.~3 addresses this question, starting from amorphous Blue Phase III,
-this time at a higher chirality. Dispersing colloidal particles in this networks
+this time at a higher chirality. Dispersing colloidal particles in this
+network
 results in the particle occupying the available nodes, where the order
@@
 \end{equation}
-The quantity ${\cal E}$ will be referred to as the reduced field strength.
+The quantity ${\cal E}$ is known as the reduced field strength.
 
@@
 Araki, T., and Tanaka, H.,
-Colloidal aggregation in a nematic liquid crystal: Topological arrest of
+Colloidal aggregation in a nematic liquid crystal: topological arrest of
 particles by a single-stroke disclination line,
@@
 Poon, W.C.K.,
-A Self-Quenched Defect Glass in a Colloid-Nematic Liquid Crystal Composite,
+A self-quenched defect glass in a colloid-nematic liquid crystal composite,
 {\it Science} {\bf 333}, 79--83 (2011).
@@
 Lin, I.H., et al.,
-Endotoxin-Induced Structural Transformations in Liquid Crystalline Droplet,
+Endotoxin-induced structural transformations in liquid crystalline droplet,
 {\it Science} {\bf 332}, 1297--1300 (2011).
@@
 \bibitem{mermin} Wright, D.C., and Mermin, N.D.,
-Crystalline liquids: The blue phases, {\it Rev. Mod. Phys.} {\bf 61},
+Crystalline liquids: the blue phases, {\it Rev. Mod. Phys.} {\bf 61},
 \textcolor{red}{385--432} (1989).
@@
 \bibitem{bp3} Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
- Structure of Blue Phase III of Cholesteric Liquid Crystals,
+ Structure of blue phase III of cholesteric liquid crystals,
 {\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
@@
 Tiribocchi, A., Gonnella, G., Marenduzzo, D., Orlandini, E., and Salvadore, F.,
-Bistable Defect Structures In Blue Phase Devices,
+Bistable defect structures In blue phase devices,
  {\it Phys. Rev. Lett.} {\bf 107}, 237803 (2011).
@@
 
-\bibitem{flowswitch}
-Henrich, O., Stratford, K., Coveney, P.V., Cates, M.E., and Marenduzzo, D.,
-Rheology of cubic blue phases,
-{\it Soft Matter} doi:10.1039/C3SM50228G (2013).
-
-\bibitem{optofluidics} Psaltis, D., Quake, S.R., Yang, C.,
-Developing optofluidic technology through the fusion of microfluidics and
-optics {\it Nature} {\bf 442}, 381--386 (2006).
-
 \textcolor{red}{
@@
 
+\bibitem{flowswitch}
+Henrich, O., Stratford, K., Coveney, P.V., Cates, M.E., and Marenduzzo, D.,
+Rheology of cubic blue phases,
+{\it Soft Matter} doi:10.1039/C3SM50228G (2013).
+
+\bibitem{optofluidics} Psaltis, D., Quake, S.R., Yang, C.,
+Developing optofluidic technology through the fusion of microfluidics and
+optics {\it Nature} {\bf 442}, 381--386 (2006).
+
 %% METHODS
@@
 \textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
-
-%\bibitem{denniston2}
-%C. Denniston, D. Marenduzzo, E. Orlandini, J.M. Yeomans,
-%Lattice Boltzmann algorithm for three-dimensional liquid-crystal
-%hydrodynamics,
-%\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745--1754 (2004).
-
-%\bibitem{codef} M. E. Cates, K. Stratford, R. Adhikari, P. Stansell,
-%J.-C. Desplat, I. Pagonabarraga, A.J. Wagner,
-%Simulating colloid hydrodynamics with lattice Boltzmann methods, 
-%\textit{J. Phys. Condens. Matter} \textbf{16},  3903--3915 (2004).
 
@@
 Information.] For clarity, only a portion of the simulation box is shown
-(one eighth of the total volume); the full structures are shown in
+\textcolor{red}{its linear extent being $\simeq 28R$}
+(one eighth of the total volume);
+the full structures are shown in
 Supplementary Figs.~2 and~3.}
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{col_disc_x_run1267.png}
-\includegraphics[width=0.45\textwidth]{sq_x_run1267.png}\\
-\includegraphics[width=0.5\textwidth]{col_disc_x_run1341.png}
-\includegraphics[width=0.45\textwidth]{sq_x_run1341.png}
+\includegraphics[width=0.48\textwidth]{col_disc_x_run1267.png}
+\includegraphics[width=0.43\textwidth]{sq_x_run1267.png}\\
+\includegraphics[width=0.48\textwidth]{col_disc_x_run1341.png}
+\includegraphics[width=0.43\textwidth]{sq_x_run1341.png}
 \end{center}

### 2014-03-05 trunk/papers/2012-sandwich/support.tex
Final updates for response version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2327 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \documentclass[12pt,twoside]{article}
-\usepackage{graphicx}
+\usepackage{graphicx,color}
 
@@
 
-{\bf Supporting Information for: Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}
+{\bf
+Self-assembly of colloid-cholesteric
+composites provides a possible route to switchable optical materials.
+
+Supplementary Information}
 
@@
 
-{\bf 1. Thermodynamics}
-
-The thermodynamics of the cholesteric liquid crystal can be described
-by means of a  Landau-de Gennes free energy functional $\cal F$,
-whose density is written as $f$:
-\begin{eqnarray}
-{\cal F}[{\bf Q}]&=&\int d^3{\bf r} f({\bf Q}({\bf r})).
-\end{eqnarray}
-This free energy density $f = f({\bf Q})$ may be expanded in powers of the
-order parameter ${\bf Q}$ and its gradients; ${\bf Q}$ is a traceless 
-and symmetric tensor which we will denote from now on in subscript
-notation as $Q_{\alpha\beta}$.
-The largest eigenvalue $q$ and corresponding eigenvector $n_\alpha$
-of $Q_{\alpha\beta}$ describe the local strength and major orientation axis
-of molecular order.
-The theory based on the tensor $Q_{\alpha\beta}$, rather than a theory based
-solely
-on the director field $n_\alpha ({\bf r})$, allows treatment of disclinations
-(defect lines) in whose cores $n_\alpha$ is undefined.
-
-Explicitly, the free energy density is:
-\begin{eqnarray}
-f(Q_{\alpha\beta}) &=& {\textstyle \frac{1}{2}} A_0
- \left(1- {\textstyle \frac{1}{3}}\gamma \right)Q^2_{\alpha \beta}
--{\textstyle \frac{1}{3}}A_0\gamma Q_{\alpha \beta} Q_{\beta \gamma}Q_{\gamma \alpha} 
-+ {\textstyle \frac{1}{4}} A_0\gamma (Q^2_{\alpha \beta})^2  \nonumber\\
-&+& {\textstyle \frac{1}{2}} K(\varepsilon_{\alpha \gamma \delta}
-\partial_\gamma Q_{\delta \beta} + 2 q_0 Q_{\alpha \beta})^2
-+ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2.
-\label{free}
-\end{eqnarray}
-Here, repeated indices are summed over, while terms of the form
-$Q^2_{\alpha\beta}$ should be expanded to read $Q_{\alpha\beta}Q_{\alpha\beta}$;
-$\varepsilon_{\alpha\gamma\delta}$ is the permutation tensor.
-The first three terms are a bulk free energy density whose overall scale
-is set by $A_0$ (discussed further below); $\gamma$ is a control parameter
-related to a reduced temperature. Varying the latter in the absence of
-chiral terms ($q_0=0$) gives an isotropic-nematic transition at
-$\gamma = 2.7$ with a mean-field spinodal instability at $\gamma = 3$.
-
-The remaining two terms of the free energy density in Eq.~\ref{free} describe
-distortions of the order parameter field. In theoretical work and when
-describing a generic rather than a specific system, it is 
-conventional~\cite{blue1,deGennes} to
-assume that splay, bend and twist deformations of the director are equally
-costly, that is, there is a single elastic constant $K$. The parameter $q_0$
-is related to the helical pitch length $p$ via $q_0=2\pi/p$,
-describing one full turn of the director in the cholesteric phase.
-
-There are two dimensionless numbers, which are commonly referred to as
-$\kappa$, the chirality, and  $\tau$, the reduced temperature
-\cite{blue1} which can be used to characterise the system. In terms of
-the parameters in the free energy density, these are:
-\begin{eqnarray}\label{cntrl-param} 
-\tau&=&\frac{27(1-\gamma/3)}{\gamma}\label{tau}\\
-\kappa&=&\sqrt{\frac{108\ K\, q_0^2}{A_0\, \gamma}}\label{kappa}.
-\end{eqnarray}
-If the free energy density Eq.~\ref{free} is made dimensionless, $\tau$
-appears as prefactor of the term  quadratic in $Q_{\alpha\beta}$,
-whereas $\kappa$
-quantifies the ratio between bulk and  gradient free energy terms. The
-chirality and reduced temperature may be used to characterise an equilibrium
-phase diagram of the blue phases in bulk cholesteric liquid crystals
-\cite{blue1,oliver1}.
-
-The effect of an electric field can be modelled by including the following
-term in the bulk free energy density:
-\begin{equation}
--\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
-\end{equation} 
-where $E_{\alpha}$ are the components of the electric field, and 
-$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
-crystal. The strength of the electric field is quantified via
-one further dimensionless number
-\begin{equation}
-{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
-\end{equation}
-The quantity ${\cal E}$ will be referred to as the reduced field strength.
-
-{\bf 1.1 Surface free energy}
-
-In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
-free energy density is also present to represent the energetic cost of
-anchoring at a solid surface. If the preferred order parameter tensor
-at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
-density (per unit area) is
-\begin{equation}
-f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
-= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
-\end{equation}
-where $W$ is a constant determining the strength of the anchoring.
-The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
-discussed below. For a colloid of radius $R$, the strength of the surface
-anchoring compared which the bulk fluid elastic constant may be quantified
-by the dimensionless parameter $WR/K$. For small values of this parameter,
-the presence of a colloid surface should have little impact on the local
-fluid LC ordering.
-
-% NO REDSHIFT
-
-{\bf 2. Dynamics}
-
-{\bf 2.1 Order parameter}
-
-A framework for the dynamics of liquid crystals is provided by the 
-Beris-Edwards model \cite{beris}, in which the time evolution of the
-tensor order parameter obeys
-\begin{equation}
-\label{eom}
-\left(\partial_t+ u_\nu \partial_\nu \right) Q_{\alpha\beta} - S_{\alpha\beta}
-= \Gamma H_{\alpha\beta}.
-\end{equation}
-In the absence of flow, Eq.~\ref{eom} describes a relaxation towards
-equilibrium on a timescale determined by a collective rotational diffusion 
-constant $\Gamma$. This relaxation is driven by the molecular field
-$H_{\alpha\beta}$, which is the functional derivative of the free energy
-with respect to the order parameter \cite{beris}:
-\begin{equation}
-H_{\alpha\beta} = -\frac{\delta {\cal F}} {\delta Q_{\alpha\beta}} 
-+ {\textstyle \frac{1}{3}} \delta_{\alpha\beta} 
-\mathrm{Tr}\left(\frac{\delta {\cal F}}{\delta Q_{\alpha\beta}}\right).
-\end{equation}
-
-The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
-derivative for rod-like molecules \cite{beris}. It couples the order
-parameter to the symmetric and antisymmetric parts of the velocity 
-gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
-The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
-$\Omega_{\alpha\beta}$ are defined as 
-\begin{equation}
-A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
-\end{equation}
-and
-\begin{equation}
-\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
-\end{equation}
-This full coupling term is then
-\begin{eqnarray}
-\label{coupling-term}
-S_{\alpha\beta}  &=&
-(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
-(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
-+
-(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
-(\xi A_{\nu\beta} - \Omega_{\nu\beta})
-\nonumber\\ 
-&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
-Q_{\nu\nu} W_{\nu\nu}.
-\end{eqnarray}
-%CORRECT TRACE?!
-Here, $\xi$ is a material-dependent that controls the
-relative importance of rotational and elongational flow for molecular
-alignment, and determines in practice how the orientational order
-responds to a local shear flow. 
-In all cases, the value used here is $\xi = 0.7$, which is within
-the `flow aligning' regime, where molecules align at a fixed angle
-(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
-The value of the collective rotational diffusion used in all simulations is
-$\Gamma = 0.5$ in simulation units.
-
-{\bf 2.2 Hydrodynamics}
-
-The momentum evolution obeys a Navier-Stokes equation driven by the divergence
-of a generalised stress $P_{\alpha\beta}$:
-\begin{equation}
-\label{nse}
-\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
-=\partial_\beta P_{\alpha\beta}.
-\end{equation}
-The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
-both viscous and thermodynamic components:
-\begin{eqnarray}
-P_{\alpha \beta}&=&
-- p_0 \delta_{\alpha \beta} 
-+ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
-\nonumber\\
-&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
-+ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
--\xi \left(Q_{\alpha \gamma}
-+ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
-\nonumber\\ 
-&+& 2 \xi  \left(Q_{\alpha \beta}
-+ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
--\partial_\alpha Q_{\gamma \nu}
-\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
-&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
-\end{eqnarray}
-A lattice Boltzmann flow solver is used, where the isotropic pressure $p_0$
-and viscous terms are managed directly by the solver (as in a simple Newtonian
-fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
-is treated as a local force density on the fluid. In all simulations
-reported the mean fluid density $\rho = 1$ and the viscosity
-$\eta = 0.01$ (simulations in confined geometry also used enhanced
-bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).
-
-{\bf 3. Surface boundary conditions}
-
-Hydrodynamic boundary conditions for solid objects are handled via
-the lattice Boltzmann solver. In particular, the standard method of
-bounce-back on links is used for both colloids and confining walls
-\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
-are set out in \cite{juho}.
-
-{\bf 3.1 Colloids}
-
-In all the simulations reported here the preferred orientation of the
-director field at the colloid surface is normal to the surface. The
-required nematic director at the particle surface in a given location
-$\hat{n}^0_\alpha$ may then be determined from geometry alone,
-and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
-\begin{equation}
-Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
-- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
-\label{eq:surf_q}
-\end{equation}
-The magnitude of surface order $q^0$ is set by
-\begin{equation}
-q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
-+ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
-\label{eq:surf_q0}
-\end{equation}
-which corresponds to the bulk order in a defect-free nematically ordered 
-sample~\cite{denniston} (this is also very close to the value of the order 
-parameter in a cholesteric or in a blue phase away from disclinations).
-
-In all simulations the colloidal (hard sphere) radius is $R = 2.3$. To
-prevent colloids overlapping at the hard sphere radius, an additional
-short range soft potential is included. This takes the form of
-$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
-The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
-units. Both potential and resulting force are smoothly matched to zero at
-a cut-off distance of $h_c = 0.25$ simulation units.
-
-{\bf 3.2 Confining walls}
-
-The confining solid walls, which are flat and stationary, have the
-same normal boundary condition as at colloid surface. In addition,
-degenerate planar anchoring is applied at the walls, where the
-preferred nematic director may adopt any orientation parallel
-to the surface. The preferred direction is determined explicitly
-by projecting the local fluid order parameter to the plane of the
-wall \cite{fournier2005}. Eq.~\ref{eq:surf_q} and Eq.~\ref{eq:surf_q0}
-are then used as before.
-
-To prevent the colloids being forced into the wall, a correction
-to the lubrication force on the colloid is added for very low colloid
-surface to wall surface separations $h < 0.5$ lattice spacing. The
-correction is based on the analytic expression for the lubrication
-between a sphere and a flat surface.
-
-{\bf 4. Protocol and parameter details}
-
-{\bf 4.1 Bulk blue phase I}
-
+\noindent
+\textbf{Parameter details for bulk BPI.}
 The parameters for the free energy are chosen to be representative
@@
 locally chosen random director, and
-Eq.~\ref{eq:surf_q} employed with a small amplitude $q^0 = 10^{-7}$ to
+%Eq.~\ref{eq:surf_q} MAIN TEXT Q_0 = .... 
+Eq.~15
+employed with a small amplitude $q^0 = 10^{-7}$ to
 provide an order parameter.
-We avoid the mean-field spinodal point at $\gamma = 3.0$:
+The mean-field spinodal point at $\gamma = 3.0$ is avoided and
 the parameters are the same as for the initial simulations with
@@
 
-{\bf 4.2 Confined blue phase I}
-
+\noindent
+\textbf{Parameters for confined BPI.}
 Here, a narrow sandwich of fluid is placed between flat walls in
@@
 $WR/K = 0.23$ and $WR/K = 2.3$. The colloid solid volume fractions are
-1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
-in simulation units. All simulations are run for two million simulation
-time steps.
+1\% and 4\%.
+\textcolor{red}{For both normal and planar anchoring at the walls,
+the value of $W$ is the same as that used for the strongly anchoring
+particles.}
+All simulations are run for two million simulation time steps.
 
-{\bf 4.3 External electric field}
-
+\noindent
+\textbf{External electric field.}
 For the simulations of BPIII with colloidal particles a system size of
@@
 \end{center}
-\caption{Free energy parameters used in the simulations. Bulk BPI
+\caption{\textbf{Free energy parameters used in the simulations.} Bulk BPI
 correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
@@
 
-{\bf 4.4 Disclination rendering}
-
+\noindent
+\textbf{Disclination rendering.}
 In all cases visualisation of the defect lines is carried out via
@@
 
-{\bf 4.4 Computation of the structure factor}
-
-In order to demonstrate structural information about the field-aligned 
-states and the residual anisotropy after switch off we followed a 
-procedure similar to the one described in \cite{oliver-bp3}.
-The structure factor in the current approach is defined via
-the Fourier transform of the colloid density:
-
-\begin{eqnarray}
-S({\bf k})&=&|F({\bf k})|^2\\
-F({\bf k})&=&\int_V d{\bf r} \rho({\bf r}) e^{i {\bf k r}}
-\end{eqnarray}
-
+\noindent
+\textbf{Computation of the structure factor.}
+To provide structural information about the field-aligned 
+states and the residual anisotropy after switch off a 
+procedure similar to the one described in \cite{oliver-bp3} is
+followed.
+The structure factor $S(\mathbf{k})$ in the current approach is
+defined via the Fourier transform of the colloid density $F(\mathbf{k})$:
+\begin{equation}
+S({\bf k}) =  |F({\bf k})|^2, \mathrm{with}\
+F({\bf k}) = \int d^3{\bf r} \rho({\bf r}) e^{i {\bf k. r}}.
+\end{equation}
 The colloid density $\rho({\bf r})$ is a discretised density and taken to be
@@
 standard definition of a structure factor and does not consider the colloids 
-as pointlike objects, allows to gain the same structural information
-in a simpler way.
+as point-like objects, allows the same structural information
+to be obtained in a simple way.
 
-{\bf 5. Parameter mapping to physical units}
-
+\noindent
+\textbf{Parameter mapping to physical units.}
 To get from simulation to physical units, a calibration of scales of
@@
 dielectric permittivity of vacuum), an electric field
-of 100 V/$\mu$m is obtained. This corresponds to a value of
+of 100 V~$\mu$m$^{-1}$ is obtained. This corresponds to a value of
 ${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
@@
 
-To summarize, the simulations represent BP-forming materials, with the
+To summarise, the simulations represent BP-forming materials, with the
 interpretation of the simulation units for length, time, and energy
 density being close to 10~nm, 10~ns, and 100~MPa respectively.
-
-%ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
-
-\vfill
-\pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{blue1}
-D.C. Wright and N.D. Mermin,
-%  Crystalline liquids -- the blue phases.
-{\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).
-\bibitem{deGennes}
-P.-G. de Gennes, J. Prost,
-{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).
-%\bibitem{marendu1}
-%D. Marenduzzo, E.  Orlandini, M. E. Cates, J. M. Yeomans,
-%{\rm Steady-state hydrodynamic instabilities of active liquid crystals:
-%Hybrid lattice Boltzmann simulations.}
-%{\it Phys. Rev. E} {\bf 76}, 031921 (2007).
-%\bibitem{marendu2}
-%O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates, 
-%{\rm Domain growth in cholesteric blue phases:
-%hybrid lattice Boltzmann simulations}.
-%{\it Comput. Math. Appl.}, doi:10.1016/j.camwa.2009.08.047[dx.doi.org] (2009).
-%\bibitem{marendu3}
-%M. E. Cates, O. Henrich, D. Marenduzzo, K. Stratford,
-%{\rm  Lattice Boltzmann simulations of liquid crystalline fluids:
-%active gels and blue phases.}
-%{\it Soft Matter} {\bf 5}, 3791-3800 (2009).
-\bibitem{oliver1}
-O. Henrich, D. Marenduzzo, K. Stratford, and M.E. Cates,
-%Thermodynamics of blue phases in electric fields
-\textit{Phys. Rev. E}, \textbf{81} 031706 (2010).
-\bibitem{beris}
-A. N. Beris, B. J. Edwards, 
-{\it Thermodynamics of Flowing Systems with Internal Microstructure}
-(Oxford Univ. Press, New York, 1994).
-\bibitem{ladd94}
-A.J.C. Ladd,
-\textit{J. Fluid Mech.} \textbf{271}, 285 (1994); \textit{ibid.} \textbf{271}
-311 (1994).
-\bibitem{nguyen2002}
-N.-Q. Nguyen and A.J.C. Ladd,
-\textit{Phys. Rev. E} \textbf{66} 046708 (2002).
-\bibitem{juho}
-J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
-\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).
-\bibitem{denniston}
-C. Denniston, E. Orlandini, J. M. Yeomans, 
-\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
-\bibitem{fournier2005}
-J.-B. Fournier and P. Galatola,
-\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
-\bibitem{oliver-bp3}
-O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
-{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
-\bibitem{denniston2}
-C. Denniston, D. Marenduzzo, E. Orlandini, J. M.  Yeomans, 
-\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
-\bibitem{oliver2}
-O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
-\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
-\bibitem{codef} M. E. Cates,  et al., 
-\textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).
-
-\end{thebibliography}
+\textcolor{red}{For a Frank eleastic
+constant of $K = 70$~pN and unit cell sizes in the range
+$\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
+K/\lambda^2 =$0.3--7~kPa.}
 
@@
 
-{\bf Supporting Figures and Animations}
+{\bf Supplementary Figures and Animations}
 
@@
 \end{center}
-\caption{
+\caption{\textbf{Blue phase structures.}
 (A) Schematic representation of the director field
@@
 \end{center}
-\caption{As for main manuscript Fig.~1 A--D
+\caption{\textbf{Bulk blue phase structures.}
+As for main manuscript Fig.~1 A--D
 (with liquid crystal order parameter
@@
 \end{center}
-\caption{As for main manuscript Fig.~1 E--H (initialised via a ``quench''
+\caption{\textbf{Bulk blue phase structures.}
+As for main manuscript Fig.~1 E--H (initialised via a ``quench''
 to generate a disordered network), but for the entire simulation
@@
 \end{center}
-\caption{Plot of the free energy (per BP unit cell) versus time
+\caption{\textbf{Free energy evolution.}
+Plot of the free energy (per BP unit cell) versus time
 for the simulations corresponding to Fig. 1A-H in the main text
@@
 \end{center}
-\caption{As for main manuscript Fig.~2 A--D (confined geometry with normal
+\caption{\textbf{Confined blue phase structures.}
+As for main manuscript Fig.~2 A--D (confined geometry with normal
 anchoring walls at both top and bottom), but for the entire simulation
@@
 \end{center}
-\caption{As for main manuscript Fig.~2 E--H (confined geometry with planar
+\caption{\textbf{Confined blue phase structures.}
+As for main manuscript Fig.~2 E--H (confined geometry with planar
 anchoring walls at both top and bottom), but for the entire simulation
@@
 \begin{center}
-\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1341.png}
-\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1342.png}
-\includegraphics[width=0.32\columnwidth]{col_parallel_z_run1342.png}\\
-\includegraphics[width=0.32\columnwidth]{col_parallel_x_run1342.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
 \includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
@@
 \end{center}
-\caption{}
+\caption{\textbf{Structure with and without electric field.}
+(Disclination lines are 
+removed for clarity.) Panels A and B reproduce the results in Fig. 3C and
+Fig. 3D respectively with the applied field in the $x$-direction 
+(out of the plane of the paper). Also included is a cut of the structure 
+factor with $k_y = 0$ (panel C). 
+Panels D, E, and F show the corresponding situation after the
+external field has been removed and the structure allowed to relax,
+and show residual hexagonal ordering. The structure factor data is for wave
+vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
+$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
 \end{figure}
+
+\newpage
 
@@
 \begin{center}
-\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}\\
-\includegraphics[width=0.32\columnwidth]{col_parallel_y_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}\\
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
 \end{center}
-\caption{}
+\caption{\textbf{Cycling the electric field.}
+Starting from the final position of
+Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
+$y$-direction. The corresponding view of the particle distribution is
+shown, again viewing along the field direction, in panel A.
+Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
+shown in panels B and C, respectively. The corresponding situation when
+the field is again switched off, and the structure allowed to relax,
+is shown in panels D, E, and F. The wave vector values used are the
+same as in Supplementary Fig.~7.}
 \end{figure}
+
+\newpage
 
@@
 \end{center}
-\caption{}
+\caption{\textbf{Cycling the electric field further.}
+Starting from the final
+position of Supplementary Fig.~8 panel~D, the field is re-applied
+again in the $x-$direction. Panel A shows the particle configuration
+viewed along the field direction, and panels B and C show the structure
+factor with $k_x = 0$ and $k_y = 0$ respectively. The corresponding
+situation when the field is removed and the structure allowed to relax
+is shown in panels D, E, and F. The wave vector values are again as
+in Supplementary Fig.~7. Note that the on and off states are 
+equivalent to those in Supplementary Fig.~7, up to a rotation.
+This is because all hexagonal structures perpendicular to the field
+direction are energetically degenerate,
+and the actual orientation of the hexagon emerges spontaneously during
+switching. }
 \end{figure}
 
+\vfill
+\pagebreak
+
+\begin{thebibliography}{99}
+
+\bibitem{blue1}
+Wright, D.C., and Mermin, N.D.,
+Crystalline liquids: the blue phases,
+{\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).
+\bibitem{oliver1}
+Henrich, O., Marenduzzo, D., Stratford, K., and Cates, M.E.,
+Thermodynamics of blue phases in electric fields,
+\textit{Phys. Rev. E}, \textbf{81} 031706 (2010).
+\bibitem{oliver-bp3}
+Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
+Structure of blue phase III of cholesteric liquid crystals,
+{\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
+\bibitem{denniston2}
+Denniston, C., Marenduzzo, D., Orlandini, E., and  Yeomans, J.M.,
+Lattice Boltzmann algorithm for three-dimensional liquid-crystal
+hydrodynamics,
+\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
+\bibitem{oliver2}
+Henrich, O., Stratford, K.,  Marenduzzo, D., and Cates, M.E.,
+Ordering dynamics of blue phases entails kinetic stabilization of
+amorphous networks, 
+\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
+\bibitem{denniston}
+Denniston, C., Orlandini, E., and Yeomans, J.M.,
+Lattice Boltzmann simulattions of liquid crystal hydrodynamics,
+\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
+\bibitem{deGennes}
+\textcolor{red}{de Gennes, P.-G., and  Prost, J.,
+{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).}
+%\bibitem{fournier2005}
+%J.-B. Fournier and P. Galatola,
+%\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
+\bibitem{codef} Cates, M.E.,  et al.,
+Simulating colloid hydrodynamics with lattice Boltzmann,
+\textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).
+
+\end{thebibliography}
+
 %\newpage
+\clearpage
 
-{\bf Captions for Supplementary Movies} \\
+%{\bf Captions for Supplementary Movies} \\
+%
+%{\bf Supplementary Movie 1:} 
+%The final state of the simulation as shown in Fig.~1A. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 1:} 
-The final state of the simulation as shown in Fig.~1A. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 2:} 
+%The final state of the simulation as shown in Fig.~1B. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 2:} 
-The final state of the simulation as shown in Fig.~1B. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 3:} 
+%The final state of the simulation as shown in Fig.~1C. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 3:} 
-The final state of the simulation as shown in Fig.~1C. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 4:} 
+%The final state of the simulation as shown in Fig.~1D. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 4:} 
-The final state of the simulation as shown in Fig.~1D. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 5:} 
+%The configuration after the initial relaxation phase in Fig.~3. The movie
+%shows a rotation of the configuration. The distribution of the colloidal
+%particles is isotropic, with the particles forming clusters and having a
+%tendency to occupy positions at the intersection of the BPIII disclination
+%lines.\\
 
-{\bf Supplementary Movie 5:} 
-The configuration after the initial relaxation phase in Fig.~3. The movie
-shows a rotation of the configuration. The distribution of the colloidal
-particles is isotropic, with the particles forming clusters and having a
-tendency to occupy positions at the intersection of the BPIII disclination
-lines.\\
-
-{\bf Supplementary Movie 6:} 
-Configuration in a homogeneous electric field with strength
-corresponding to ${\cal E}=0.8$. The movie
-shows a rotation of the configuration.
-A hexagonal pattern emerges where the particles form chains
-situated at the vertices of the honeycomb-like structure. \\
+%{\bf Supplementary Movie 6:} 
+%Configuration in a homogeneous electric field with strength
+%corresponding to ${\cal E}=0.8$. The movie
+%shows a rotation of the configuration.
+%A hexagonal pattern emerges where the particles form chains
+%situated at the vertices of the honeycomb-like structure. \\
 

### 2014-03-05 papers/2012-sandwich/CholColloids.tex
Correct spelling
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2328 40876441-3496-0410-a613-a2fca35b2ae2

@@
 they disrupt the orientational order in the fluid and create defects (disclination lines) in the nematic close to their surfaces.
-To minimize the free energy cost, it is generally advantageous for defects to be shared:
+To minimise the free energy cost, it is generally advantageous for defects to be shared:
 therefore the colloidal particles have a generic tendency to aggregate. Such aggregates may further 
@@
 Information.] For clarity, only a portion of the simulation box is shown
-\textcolor{red}{its extent being $\simeq 28R$}
+\textcolor{red}{its linear extent being $\simeq 28R$}
 (one eighth of the total volume);

### 2014-03-05 trunk/papers/2012-sandwich/support.tex
Correct spelling
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2328 40876441-3496-0410-a613-a2fca35b2ae2

@@
 For simplicity,  the size of the BPI unit cell is not allowed to 
-readjust dynamically to minimize free energy (there is no 
+readjust dynamically to minimise free energy (there is no 
 ``redshift''~\cite{blue1}): this would lead to quantitatively different 

### 2014-03-21 papers/2012-sandwich/CholColloids.tex
Second reply
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2351 40876441-3496-0410-a613-a2fca35b2ae2

@@
 In conventional cholesteric materials~\cite{mermin} this balance is only achievable in a narrow temperature range of a few degrees, around the onset of liquid crystalline ordering.
-\textcolor{red}{
 However, recent advances in formulation have widened this stability
@@
 to~125$^\circ$C have been reported \cite{coles}, paving the way to
-applications of BPs in operational display devices.}
+applications of BPs in operational display devices.
 
@@
 
-\textcolor{red}{
 The simulations presented here are representative of a liquid crystal
@@
 simulation parameters are reported in the Supplementary Information.
-}
 
@@
 Among dimensionless control parameters are the particle solid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
-\textcolor{red}{
 The anchoring strength is defined such that $W > 0$ imparts a preferred
@@
 can be either normal to the surface or in a plane tangential to the surface,
-a property related to surface chemistry in experimental systems.% but which
-%can be imposed at will in simulations.
-We choose} $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
+a property related to surface chemistry in experimental systems.
+We choose $w = 0.23, 2.3$ which are within typical experimental range~\cite{tiffany}.
 This parameter can be viewed as the ratio of the anchoring energy, $WR^2$, to the elastic energy scale for distortion~$KR$.
@@
 fraction (over 20\% as opposed to below 5\% here), for which $G\sim 10^3$ Pa.
-\textcolor{red}{
 This figure is consistent with an estimate obtained by comparing
@@
 colloidal particles with those of rubber~\cite{ramos} for
-the relevant simulation parameters (see Supplementary Information).}
+the relevant simulation parameters (see Supplementary Information).
 
@@
 Supplementary Fig.~4 and Ref.~\cite{stark,lavrentovich}).
+
+\textcolor{red}{
+It may be possible to anneal the structures obtained via the quench route,
+although we have not attempted this in simulation. The results presented
+suggest that even when starting with colloidal particles randomly dispersed
+in an ordered blue phase, the particles are able to distort the structure
+(Fig.~1A-B). This suggests that it would be difficult to identify a quench
+protocol which would lead to less, rather than more, disorder.
+}
 
@@
 
-\textcolor{red}{
 The degree of similarity between
@@
 strong anchoring.
-}
 
@@
 (Supplementary Fig.~8 and Fig.~9).
+\textcolor{red}{
+We note that for the fluid parameters used here, the field induced
+changes are irreversible with the weakly anchoring colloid particles.
+The mesophase without colloid particles undergoes
+a reversible ordering transition~\cite{bp3}; the reversibility
+is retained in the
+presence of particles only if their anchoring is very weak.}
 
@@
 
-\textcolor{red}{
 How do these simulations relate to experiment? Several authors have
@@
 simulations, so heavily functionalised particles might be avoided.
-}
 
@@
 free energy density is also present to represent the energetic cost of
-anchoring at a solid surface. If the preferred order parameter tensor
-at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
-density (per unit area) is
+anchoring at a solid surface.
+\textcolor{red}{In the case of normal anchoring, the surface free energy
+density (per unit area) is}
 \begin{equation}
@@
 \end{equation}
-where $W$ is a constant determining the strength of the anchoring.
+\textcolor{red}{where $Q^0_{\alpha\beta}$ is the preferred order
+parameter tensor at the solid surface, and $W$ is a constant determining
+the strength of the anchoring. In the case of planar anchoring, a slightly
+more complicated expression is required to allow for degeneracy
+(see, e.g., \cite{fournier2005}.)}
 The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
@@
 Crystalline liquids: the blue phases, {\it Rev. Mod. Phys.} {\bf 61},
-\textcolor{red}{385--432} (1989).
+385--432 (1989).
 
@@
 \bibitem{extrareference1}
-\textcolor{red}{Ravnik, M., Alexander, G.P., Yeomans, J.M., and Zumer, S.,
+Ravnik, M., Alexander, G.P., Yeomans, J.M., and Zumer, S.,
 Mesoscopic modelling of colloids in chiral nematics,
-{\it Faraday Discuss.} {\bf 144}, 159--169 (2010).}
+{\it Faraday Discuss.} {\bf 144}, 159--169 (2010).
 
@@
 \bibitem{extrareference2}
-\textcolor{red}{Ravnik, M., Fukuda, J.-I.,  Yeomans, J.M., and Zumer, S.,
+Ravnik, M., Fukuda, J.-I.,  Yeomans, J.M., and Zumer, S.,
 Confining blue phase colloids to thin layers,
-{\it Soft Matter} {\bf 7}, 10144--10150 (2011).}
+{\it Soft Matter} {\bf 7}, 10144--10150 (2011).
 
@@
 \bibitem{ramos} 
-\textcolor{red}{
 Ramos, L., Zapotocky, M.,  Lubensky, T.C., and  Weitz, D.A.,
 Rheology of defect networks in cholesteric liquid crystals,
-{\it Phys. Rev. E} {\bf 66} 031711 (2002).}
+{\it Phys. Rev. E} {\bf 66} 031711 (2002).
 
@@
 
-\textcolor{red}{
 \bibitem{yoshida2009}
@@
 \textit{Soft Matter} \textbf{9}, 3956--3964 (2013).
-}
 
@@
 \bibitem{deGennes}
-\textcolor{red}{de Gennes, P.-G., and  Prost, J.,
-{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).}
+de Gennes, P.-G., and  Prost, J.,
+{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).
 
@@
 Information.] For clarity, only a portion of the simulation box is shown
-\textcolor{red}{its linear extent being $\simeq 28R$}
+its linear extent being $\simeq 28R$
 (one eighth of the total volume);
@@
 As in Fig.~1, only one quarter of the simulation box is shown for clarity
-\textcolor{red}{(its horizontal extent being $\simeq 56R$ and vertical
-extent $\simeq 24R$ in the narrow direction)};
+(its horizontal extent being $\simeq 56R$ and vertical
+extent $\simeq 24R$ in the narrow direction);
 the full structures are shown in Supplementary Figs.~5 and~6.}
@@
 parameters).
-\textcolor{red}{The full simulation with a box size of $\simeq 56R$ is shown.}}
+The full simulation with a box size of $\simeq 56R$ is shown.}
 \end{figure}

### 2014-03-21 trunk/papers/2012-sandwich/support.tex
Second reply
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2351 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \textcolor{red}{For both normal and planar anchoring at the walls,
-the value of $W$ is the same as that used for the strongly anchoring
-particles.}
+the strength is adjusted to correspond to that used for the strongly
+anchoring particles.}
 All simulations are run for two million simulation time steps.
@@
 density being close to 10~nm, 10~ns, and 100~MPa respectively.
-\textcolor{red}{For a Frank eleastic
+For a Frank eleastic
 constant of $K = 70$~pN and unit cell sizes in the range
 $\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
-K/\lambda^2 =$0.3--7~kPa.}
+K/\lambda^2 =$0.3--7~kPa.
 
@@
 \bibitem{deGennes}
-\textcolor{red}{de Gennes, P.-G., and  Prost, J.,
-{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).}
+de Gennes, P.-G., and  Prost, J.,
+{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).
 %\bibitem{fournier2005}

### 2014-04-16 papers/2012-sandwich/support.tex
updates following editorial comments plus cover art
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2363 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+% Prepend ``Supplementary'' Fig/Table caption headings
+\usepackage{caption}
+\captionsetup[table]{name=Supplementary Table}
+\captionsetup[figure]{name=Supplementary Figure}
+
+% Likewise...
+\renewcommand\refname{Supplementary References}
+
 \begin{document}
 
-{\bf
-Self-assembly of colloid-cholesteric
-composites: A route to switchable optical materials.
+{\bf Supplementary Figures}
 
-Supplementary Information}
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.8]{support-fig1.png}
+\end{center}
+\caption{\textbf{Blue phase structures.}
+(A) Schematic representation of the director field
+in a cholesteric (left) and within a double twist cylinder (right).
+(B) Snapshot of the disclination network in an equilibrated 
+blue phase I structure. (C) Same as (B) for an amorphous blue phase III
+network.}
+\end{figure}
 
-Here, we give more details of the simulation method, protocol, and parameters used.
-We also provide additional figures complementing the main text, and discuss the matching of simulation units to physical units.
+\newpage
 
-{\bf 1. Thermodynamics}
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig2.pdf}
+\end{center}
+\caption{\textbf{Bulk blue phase structures.}
+As for main manuscript Fig.~1 A--D
+(with liquid crystal order parameter
+initialised to equilibrium blue phase I structure), but for the entire
+simulation system of 128$^3$ lattice sites. (A)  solid volume fraction~1\%
+and weak anchoring;  (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring. The view direction in the
+main figure is from the left here. There is a reference
+particle at the bottom left in each panel which does not take part
+in the simulation.}
+\end{figure}
 
-The thermodynamics of the cholesteric liquid crystal can be described
-by means of a  Landau-de Gennes free energy functional $\cal F$,
-whose density is written as $f$:
-\begin{eqnarray}
-{\cal F}[{\bf Q}]&=&\int d^3{\bf r} f({\bf Q}({\bf r})).
-\end{eqnarray}
-This free energy density $f = f({\bf Q})$ may be expanded in powers of the
-order parameter ${\bf Q}$ and its gradients; ${\bf Q}$ is a traceless 
-and symmetric tensor which is denoted from now on in subscript
-notation as $Q_{\alpha\beta}$.
-The largest eigenvalue $q$ and corresponding eigenvector $n_\alpha$
-of $Q_{\alpha\beta}$ describe the local strength and major orientation axis
-of molecular order.
-The theory based on the tensor $Q_{\alpha\beta}$, rather than a theory based
-solely
-on the director field $n_\alpha ({\bf r})$, allows treatment of disclinations
-(defect lines) in whose cores $n_\alpha$ is undefined.
+\newpage
 
-Explicitly, the free energy density is:
-\begin{eqnarray}
-f(Q_{\alpha\beta}) &=& {\textstyle \frac{1}{2}} A_0
- \left(1- {\textstyle \frac{1}{3}}\gamma \right)Q^2_{\alpha \beta}
--{\textstyle \frac{1}{3}}A_0\gamma Q_{\alpha \beta} Q_{\beta \gamma}Q_{\gamma \alpha} 
-+ {\textstyle \frac{1}{4}} A_0\gamma (Q^2_{\alpha \beta})^2  \nonumber\\
-&+& {\textstyle \frac{1}{2}} K(\varepsilon_{\alpha \gamma \delta}
-\partial_\gamma Q_{\delta \beta} + 2 q_0 Q_{\alpha \beta})^2
-+ {\textstyle \frac{1}{2}} K (\partial_\beta Q_{\alpha \beta})^2.
-\label{free}
-\end{eqnarray}
-Here, repeated indices are summed over, while terms of the form
-$Q^2_{\alpha\beta}$ should be expanded to read $Q_{\alpha\beta}Q_{\alpha\beta}$;
-$\varepsilon_{\alpha\gamma\delta}$ is the permutation tensor.
-The first three terms are a bulk free energy density whose overall scale
-is set by $A_0$ (discussed further below); $\gamma$ is a control parameter
-related to a reduced temperature. Varying the latter in the absence of
-chiral terms ($q_0=0$) gives an isotropic-nematic transition at
-$\gamma = 2.7$ with a mean-field spinodal instability at $\gamma = 3$.
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig3.pdf}
+\end{center}
+\caption{\textbf{Bulk blue phase structures.}
+As for main manuscript Fig.~1 E--H (initialised via a ``quench''
+to generate a disordered network), but for the entire simulation
+system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
+weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring. Again, there is a reference
+particle at the bottom left in each case which does not take part
+in the simulation.}
+\end{figure}
 
-The remaining two terms of the free energy density in Eq.~\ref{free} describe
-distortions of the order parameter field. In theoretical work and when
-describing a generic rather than a specific system, it is 
-conventional~\cite{blue1,deGennes} to
-assume that splay, bend and twist deformations of the director are equally
-costly, that is, there is a single elastic constant $K$. The parameter $q_0$
-is related to the helical pitch length $p$ via $q_0=2\pi/p$,
-describing one full turn of the director in the cholesteric phase.
+\newpage
 
-There are two dimensionless numbers, which are commonly referred to as
-$\kappa$, the chirality, and  $\tau$, the reduced temperature
-\cite{blue1} which can be used to characterise the system. In terms of
-the parameters in the free energy density, these are:
-\begin{eqnarray}\label{cntrl-param} 
-\tau&=&\frac{27(1-\gamma/3)}{\gamma}\label{tau}\\
-\kappa&=&\sqrt{\frac{108\ K\, q_0^2}{A_0\, \gamma}}\label{kappa}.
-\end{eqnarray}
-If the free energy density Eq.~\ref{free} is made dimensionless, $\tau$
-appears as prefactor of the term  quadratic in $Q_{\alpha\beta}$,
-whereas $\kappa$
-quantifies the ratio between bulk and  gradient free energy terms. The
-chirality and reduced temperature may be used to characterise an equilibrium
-phase diagram of the blue phases in bulk cholesteric liquid crystals
-\cite{blue1,oliver1}.
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.49]{support-fig4.png}
+\end{center}
+\caption{\textbf{Free energy evolution.}
+Plot of the free energy (per BP unit cell) versus time
+for the simulations corresponding to Fig. 1A-H in the main text
+(the labels correspond to those in Fig. 1).
+}
+\end{figure}
 
-The effect of an electric field can be modelled by including the following
-term in the bulk free energy density:
-\begin{equation}
--\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
-\end{equation} 
-where $E_{\alpha}$ are the components of the electric field, and 
-$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
-crystal. The strength of the electric field is quantified via
-one further dimensionless number
-\begin{equation}
-{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
-\end{equation}
-The quantity ${\cal E}$ will be referred to as the reduced field strength.
+\newpage
 
-{\bf 1.1 Surface free energy}
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig5.pdf}
+\end{center}
+\caption{\textbf{Confined blue phase structures.}
+As for main manuscript Fig.~2 A--D (confined geometry with normal
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (A) solid volume fraction 1\%
+and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
 
-In addition to the fluid free energy density $f(Q_{\alpha\beta})$, a surface
-free energy density is also present to represent the energetic cost of
-anchoring at a solid surface. If the preferred order parameter tensor
-at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
-density (per unit area) is
-\begin{equation}
-f_s(Q_{\alpha\beta}, Q^0_{\alpha\beta})
-= {\textstyle \frac{1}{2}}W(Q_{\alpha\beta} - Q^0_{\alpha\beta})^2,
-\end{equation}
-where $W$ is a constant determining the strength of the anchoring.
-The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
-discussed below. For a colloid of radius $R$, the strength of the surface
-anchoring compared which the bulk fluid elastic constant may be quantified
-by the dimensionless parameter $WR/K$. For small values of this parameter,
-the presence of a colloid surface should have little impact on the local
-fluid LC ordering.
+\newpage
 
-% NO REDSHIFT
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig6.pdf}
+\end{center}
+\caption{\textbf{Confined blue phase structures.}
+As for main manuscript Fig.~2 E--H (confined geometry with planar
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (E) colloid solid volume fraction 1\%
+and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
 
-{\bf 2. Dynamics}
+\newpage
 
-{\bf 2.1 Order parameter}
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
+\end{center}
+\caption{\textbf{Structure with and without electric field.}
+(Disclination lines are 
+removed for clarity.) Panels A and B reproduce the results in Fig. 3C and
+Fig. 3D respectively with the applied field in the $x$-direction 
+(out of the plane of the paper). Also included is a cut of the structure 
+factor with $k_y = 0$ (panel C). 
+Panels D, E, and F show the corresponding situation after the
+external field has been removed and the structure allowed to relax,
+and show residual hexagonal ordering. The structure factor data is for wave
+vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
+$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
+\end{figure}
 
-A framework for the dynamics of liquid crystals is provided by the 
-Beris-Edwards model \cite{beris}, in which the time evolution of the
-tensor order parameter obeys
-\begin{equation}
-\label{eom}
-\left(\partial_t+ u_\nu \partial_\nu \right) Q_{\alpha\beta} - S_{\alpha\beta}
-= \Gamma H_{\alpha\beta}.
-\end{equation}
-In the absence of flow, Eq.~\ref{eom} describes a relaxation towards
-equilibrium on a timescale determined by a collective rotational diffusion 
-constant $\Gamma$. This relaxation is driven by the molecular field
-$H_{\alpha\beta}$, which is the functional derivative of the free energy
-with respect to the order parameter \cite{beris}:
-\begin{equation}
-H_{\alpha\beta} = -\frac{\delta {\cal F}} {\delta Q_{\alpha\beta}} 
-+ {\textstyle \frac{1}{3}} \delta_{\alpha\beta} 
-\mathrm{Tr}\left(\frac{\delta {\cal F}}{\delta Q_{\alpha\beta}}\right).
-\end{equation}
+\newpage
 
-The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
-derivative for rod-like molecules \cite{beris}. It couples the order
-parameter to the symmetric and antisymmetric parts of the velocity 
-gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
-The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
-$\Omega_{\alpha\beta}$ are defined as 
-\begin{equation}
-A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
-\end{equation}
-and
-\begin{equation}
-\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
-\end{equation}
-This full coupling term is then
-\begin{eqnarray}
-\label{coupling-term}
-S_{\alpha\beta}  &=&
-(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
-(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
-+
-(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
-(\xi A_{\nu\beta} - \Omega_{\nu\beta})
-\nonumber\\ 
-&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
-Q_{\nu\nu} W_{\nu\nu}.
-\end{eqnarray}
-%CORRECT TRACE?!
-Here, $\xi$ is a material-dependent that controls the
-relative importance of rotational and elongational flow for molecular
-alignment, and determines in practice how the orientational order
-responds to a local shear flow. 
-In all cases, the value used here is $\xi = 0.7$, which is within
-the `flow aligning' regime, where molecules align at a fixed angle
-(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
-The value of the collective rotational diffusion used in all simulations is
-$\Gamma = 0.5$ in simulation units.
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
+\end{center}
+\caption{\textbf{Cycling the electric field.}
+Starting from the final position of
+Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
+$y$-direction. The corresponding view of the particle distribution is
+shown, again viewing along the field direction, in panel A.
+Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
+shown in panels B and C, respectively. The corresponding situation when
+the field is again switched off, and the structure allowed to relax,
+is shown in panels D, E, and F. The wave vector values used are the
+same as in Supplementary Fig.~7.}
+\end{figure}
 
-{\bf 2.2 Hydrodynamics}
+\newpage
 
-The momentum evolution obeys a Navier-Stokes equation driven by the divergence
-of a generalised stress $P_{\alpha\beta}$:
-\begin{equation}
-\label{nse}
-\rho\,\partial_tu_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
-=\partial_\beta P_{\alpha\beta}.
-\end{equation}
-The pressure tensor $P_{\alpha\beta}$ is, in general, asymmetric and includes
-both viscous and thermodynamic components:
-\begin{eqnarray}
-P_{\alpha \beta}&=&
-- p_0 \delta_{\alpha \beta} 
-+ \eta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha\}
-\nonumber\\
-&-&  \xi H_{\alpha \gamma}\left(Q_{\gamma \beta}
-+ {\textstyle \frac{1}{3}} \delta_{\gamma \beta} \right)
--\xi \left(Q_{\alpha \gamma}
-+ { \textstyle \frac{1}{3}} \delta_{\alpha \gamma}\right) H_{\gamma \beta}
-\nonumber\\ 
-&+& 2 \xi  \left(Q_{\alpha \beta}
-+ {\textstyle \frac{1}{3}} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
--\partial_\alpha Q_{\gamma \nu}
-\frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
-&+& Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
-\end{eqnarray}
-A lattice Boltzmann flow solver is used, where the isotropic pressure $p_0$
-and viscous terms are managed directly by the solver (as in a simple Newtonian
-fluid, of viscosity $\eta$), whereas the divergence of the remaining terms
-is treated as a local force density on the fluid. In all simulations
-reported the mean fluid density $\rho = 1$ and the viscosity
-$\eta = 0.01$ (simulations in confined geometry also used enhanced
-bulk viscosity $\zeta = 10\eta$ to ensure numerical stability).
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1347.png}
+\end{center}
+\caption{\textbf{Cycling the electric field further.}
+Starting from the final
+position of Supplementary Fig.~8 panel~D, the field is re-applied
+again in the $x-$direction. Panel A shows the particle configuration
+viewed along the field direction, and panels B and C show the structure
+factor with $k_x = 0$ and $k_y = 0$ respectively. The corresponding
+situation when the field is removed and the structure allowed to relax
+is shown in panels D, E, and F. The wave vector values are again as
+in Supplementary Fig.~7. Note that the on and off states are 
+equivalent to those in Supplementary Fig.~7, up to a rotation.
+This is because all hexagonal structures perpendicular to the field
+direction are energetically degenerate,
+and the actual orientation of the hexagon emerges spontaneously during
+switching. }
+\end{figure}
 
-{\bf 3. Surface boundary conditions}
+\vfill\pagebreak
 
-Hydrodynamic boundary conditions for solid objects are handled via
-the lattice Boltzmann solver. In particular, the standard method of
-bounce-back on links is used for both colloids and confining walls
-\cite{ladd94,nguyen2002}. Boundary conditions for the order parameter
-are set out in \cite{juho}.
+{\bf Supplementary Table}
 
-{\bf 3.1 Colloids}
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\hline
+         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
+\hline 
+Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
+0.23, 2.3\\
+\hline
+Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
+0.23, 2.3 \\
+\hline
+Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
+\hline
+External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
+\hline
+\end{tabular}
+\end{center}
+\caption{\textbf{Free energy parameters used in the simulations.} Bulk BPI
+correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
+geometries are shown in Fig.~2; and the external field case is
+relevant for Fig.~3.}
+\label{table:params}
+\end{table}
 
-In all the simulations reported here the preferred orientation of the
-director field at the colloid surface is normal to the surface. The
-required nematic director at the particle surface in a given location
-$\hat{n}^0_\alpha$ may then be determined from geometry alone,
-and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
-\begin{equation}
-Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
-- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
-\label{eq:surf_q}
-\end{equation}
-The magnitude of surface order $q^0$ is set by
-\begin{equation}
-q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
-+ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
-\label{eq:surf_q0}
-\end{equation}
-which corresponds to the bulk order in a defect-free nematically ordered 
-sample~\cite{denniston} (this is also very close to the value of the order 
-parameter in a cholesteric or in a blue phase away from disclinations).
+\vfill
+\pagebreak
 
-In all simulations the colloidal (hard sphere) radius is $R = 2.3$. To
-prevent colloids overlapping at the hard sphere radius, an additional
-short range soft potential is included. This takes the form of
-$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
-The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
-units. Both potential and resulting force are smoothly matched to zero at
-a cut-off distance of $h_c = 0.25$ simulation units.
+{\bf Supplementary Methods}
 
-{\bf 3.2 Confining walls}
-
-The confining solid walls, which are flat and stationary, have the
-same normal boundary condition as at colloid surface. In addition,
-degenerate planar anchoring is applied at the walls, where the
-preferred nematic director may adopt any orientation parallel
-to the surface. The preferred direction is determined explicitly
-by projecting the local fluid order parameter to the plane of the
-wall \cite{fournier2005}. Eq.~\ref{eq:surf_q} and Eq.~\ref{eq:surf_q0}
-are then used as before.
-
-To prevent the colloids being forced into the wall, a correction
-to the lubrication force on the colloid is added for very low colloid
-surface to wall surface separations $h < 0.5$ lattice spacing. The
-correction is based on the analytic expression for the lubrication
-between a sphere and a flat surface.
-
-{\bf 4. Protocol and parameter details}
-
-{\bf 4.1 Bulk blue phase I}
-
+\noindent
+\textbf{Parameter details for bulk BPI.}
 The parameters for the free energy are chosen to be representative
@@
 locally chosen random director, and
-Eq.~\ref{eq:surf_q} employed with a small amplitude $q^0 = 10^{-7}$ to
+%Eq.~\ref{eq:surf_q} MAIN TEXT Q_0 = .... 
+Eq.~15
+employed with a small amplitude $q^0 = 10^{-7}$ to
 provide an order parameter.
@@
 $WR/K = 0.23$ and $WR/K = 2.3$. The colloid solid volume fractions are
-1\% and 4\%. Wall anchoring is either normal or planar with $W/K = 1$
-in simulation units. All simulations are run for two million simulation
-time steps.
+1\% and 4\%.
+For both normal and planar anchoring at the walls,
+the strength is adjusted to correspond to that used for the strongly
+anchoring particles.
+All simulations are run for two million simulation time steps.
 
-{\bf 4.3 External electric field}
-
+\noindent
+\textbf{External electric field.}
 For the simulations of BPIII with colloidal particles a system size of
@@
 from which an amorphous BPIII network emerges \cite{oliver-bp3}. This structure
-is equilibrated for $6\times10^{5}$ LB time steps until no significant
+is equilibrated for $6\times10^{5}$~LB time steps until no significant
 further evolution is observed.
@@
 
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-\hline
-         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
-\hline 
-Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
-0.23, 2.3\\
-\hline
-Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
-0.23, 2.3 \\
-\hline
-Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
-\hline
-External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
-\hline
-\end{tabular}
-\end{center}
-\caption{Free energy parameters used in the simulations. Bulk BPI
-correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
-geometries are shown in Fig.~2; and the external field case is
-relevant for Fig.~3.}
-\label{table:params}
-\end{table}
-
-{\bf 4.4 Disclination rendering}
-
+\noindent
+\textbf{Disclination rendering.}
 In all cases visualisation of the defect lines is carried out via
-an isosurface of the scalar order parameter $q$ determined from the
+an isosurface of the scalar order parameter~$q$ determined from the
 largest eigenvalue of the local order parameter tensor. A low value
-of the scalar order parameter $q$ (typically in the range 0.12--0.14)
+of the scalar order parameter~$q$ (typically in the range~0.12--0.14)
 unambiguously identifies the disclinations.
 
-{\bf 4.4 Computation of the structure factor}
-
+\noindent
+\textbf{Computation of the structure factor.}
 To provide structural information about the field-aligned 
@@
 defined via the Fourier transform of the colloid density $F(\mathbf{k})$:
-
-%\begin{eqnarray}
-%S({\bf k})&=&|F({\bf k})|^2,\\
-%F({\bf k})&=&\int d^3{\bf r} \rho({\bf r}) e^{i {\bf k. r}}.
-%\end{eqnarray}
-
 \begin{equation}
@@
 \end{equation}
-
 The colloid density $\rho({\bf r})$ is a discretised density and taken to be
@@
 
-{\bf 5. Parameter mapping to physical units}
-
+\noindent
+\textbf{Parameter mapping to physical units.}
 To get from simulation to physical units, a calibration of scales of
@@
 \begin{equation}
-\gamma_1=\frac{2q^2}{\Gamma},
+\gamma_1=\frac{2q^2}{\mathit{\Gamma}},
 \end{equation}
 which relates the rotational viscosity $\gamma_1$ to the ordering strength
-$q$ and the order parameter mobility $\Gamma$.
-In the present simulations, $\Gamma = 0.5$ and the value of $\gamma$
+$q$ and the order parameter mobility $\mathit{\Gamma}$.
+In the present simulations, $\mathit{\Gamma} = 0.5$ and the value of $\gamma$
 for the simulation in Fig.~1 leads to $q=1/2$.   
@@
 Given the previous value for $A_0$, and a dielectric anisotropy of 20
-(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
+(or equivalently $\epsilon_\mathrm{a}=20\epsilon_0$, with $\epsilon_0$ the
 dielectric permittivity of vacuum), an electric field
-of 100 V/$\mu$m is obtained. This corresponds to a value of
+of 100 V~$\mu$m$^{-1}$ is obtained. This corresponds to a value of
 ${\cal E}\sim 0.4$ (considering a value of $\gamma=3.0857$ as in
@@
 
-To summarize, the simulations represent BP-forming materials, with the
+To summarise, the simulations represent BP-forming materials, with the
 interpretation of the simulation units for length, time, and energy
 density being close to 10~nm, 10~ns, and 100~MPa respectively.
-
-%ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK
+For a Frank eleastic
+constant of $K = 70$~pN and unit cell sizes in the range
+$\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
+K/\lambda^2 =$ 0.3--7~kPa.
 
@@
 \bibitem{blue1}
-D.C. Wright and N.D. Morin,
-%  Crystalline liquids -- the blue phases.
+Wright, D.C., and Mermin, N.D.,
+Crystalline liquids: the blue phases,
 {\it Rev. Mod. Phys.} {\bf 61,} 385-432 (1989).
-\bibitem{deGennes}
-P.-G. DEC Gunner's, J. Poorest,
-{\it The Physics of Liquid Crystals} (Oxford Univ. Press, New York, 1993).
-%\bibitem{marendu1}
-%D. Marenduzzo, E.  Orlandini, M. E. Cates, J. M. Yeomans,
-%{\rm Steady-state hydrodynamic instabilities of active liquid crystals:
-%Hybrid lattice Boltzmann simulations.}
-%{\it Phys. Rev. E} {\bf 76}, 031921 (2007).
-%\bibitem{marendu2}
-%O. Henrich, D. Marenduzzo, K. Stratford, M. E. Cates, 
-%{\rm Domain growth in cholesteric blue phases:
-%hybrid lattice Boltzmann simulations}.
-%{\it Comput. Math. Appl.}, doi:10.1016/j.camwa.2009.08.047[dx.doi.org] (2009).
-%\bibitem{marendu3}
-%M. E. Cates, O. Henrich, D. Marenduzzo, K. Stratford,
-%{\rm  Lattice Boltzmann simulations of liquid crystalline fluids:
-%active gels and blue phases.}
-%{\it Soft Matter} {\bf 5}, 3791-3800 (2009).
 \bibitem{oliver1}
-O. Henrich, D. Marenduzzo, K. Stratford, and M.E. Cates,
-%Thermodynamics of blue phases in electric fields
+Henrich, O., Marenduzzo, D., Stratford, K., and Cates, M.E.,
+Thermodynamics of blue phases in electric fields,
 \textit{Phys. Rev. E}, \textbf{81} 031706 (2010).
-\bibitem{beris}
-A. N. Beris, B. J. Edwards, 
-{\it Thermodynamics of Flowing Systems with Internal Microstructure}
-(Oxford Univ. Press, New York, 1994).
-\bibitem{ladd94}
-A.J.C. Ladd,
-\textit{J. Fluid Mech.} \textbf{271}, 285 (1994); \textit{ibid.} \textbf{271}
-311 (1994).
-\bibitem{nguyen2002}
-N.-Q. Nguyen and A.J.C. Ladd,
-\textit{Phys. Rev. E} \textbf{66} 046708 (2002).
-\bibitem{juho}
-J.S. Lintuvuori, D. Marenduzzo, K. Stratford, and M.E. Cates,
-\textit{J. Mater. Chem.} \textbf{20}, 10547 (2010).
-\bibitem{denniston}
-C. Denniston, E. Orlandini, J. M. Yeomans, 
-\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
-\bibitem{fournier2005}
-J.-B. Fournier and P. Galatola,
-\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
 \bibitem{oliver-bp3}
-O. Henrich, K. Stratford, M.~E. Cates, D. Marenduzzo,
+Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
+Structure of blue phase III of cholesteric liquid crystals,
 {\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
 \bibitem{denniston2}
-C. Denniston, D. Marenduzzo, E. Orlandini, J. M.  Yeomans, 
+Denniston, C., Marenduzzo, D., Orlandini, E., and  Yeomans, J.M.,
+Lattice Boltzmann algorithm for three-dimensional liquid-crystal
+hydrodynamics,
 \textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).
 \bibitem{oliver2}
-O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
+Henrich, O., Stratford, K.,  Marenduzzo, D., and Cates, M.E.,
+Ordering dynamics of blue phases entails kinetic stabilization of
+amorphous networks, 
 \textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
-\bibitem{codef} M. E. Cates,  et al., 
+\bibitem{denniston}
+Denniston, C., Orlandini, E., and Yeomans, J.M.,
+Lattice Boltzmann simulattions of liquid crystal hydrodynamics,
+\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).
+\bibitem{deGennes}
+de Gennes, P.-G., and  Prost, J.,
+{\it The Physics of Liquid Crystals} (Clarendon Press, 1995).
+%\bibitem{fournier2005}
+%J.-B. Fournier and P. Galatola,
+%\textit{Europhys. Lett.} \textbf{72}, 403 (2005).
+\bibitem{codef} Cates, M.E.,  et al.,
+Simulating colloid hydrodynamics with lattice Boltzmann,
 \textit{J. Phys. Condens. Matter} \textbf{16},  3903 (2004).
@@
 
-\vfill\pagebreak
-
-{\bf Supplementary Figures and Animations}
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.8]{support-fig1.png}
-\end{center}
-\caption{
-(A) Schematic representation of the director field
-in a cholesteric (left) and within a double twist cylinder (right).
-(B) Snapshot of the disclination network in an equilibrated 
-blue phase I structure. (C) Same as (B) for an amorphous blue phase III
-network.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.35]{support-fig2.pdf}
-\end{center}
-\caption{As for main manuscript Fig.~1 A--D
-(with liquid crystal order parameter
-initialised to equilibrium blue phase I structure), but for the entire
-simulation system of 128$^3$ lattice sites. (A)  solid volume fraction~1\%
-and weak anchoring;  (B) 4\% solid volume fraction and weak anchoring;
-(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
-volume fraction and strong anchoring. The view direction in the
-main figure is from the left here. There is a reference
-particle at the bottom left in each panel which does not take part
-in the simulation.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.35]{support-fig3.pdf}
-\end{center}
-\caption{As for main manuscript Fig.~1 E--H (initialised via a ``quench''
-to generate a disordered network), but for the entire simulation
-system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
-weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
-(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
-volume fraction and strong anchoring. Again, there is a reference
-particle at the bottom left in each case which does not take part
-in the simulation.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.49]{support-fig4.png}
-\end{center}
-\caption{Plot of the free energy (per BP unit cell) versus time
-for the simulations corresponding to Fig. 1A-H in the main text
-(the labels correspond to those in Fig. 1).
-}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.42]{support-fig5.pdf}
-\end{center}
-\caption{As for main manuscript Fig.~2 A--D (confined geometry with normal
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (A) solid volume fraction 1\%
-and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
-(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.42]{support-fig6.pdf}
-\end{center}
-\caption{As for main manuscript Fig.~2 E--H (confined geometry with planar
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (E) colloid solid volume fraction 1\%
-and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
-(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
-\end{center}
-\caption{Structure with and without electric field, with disclination lines
-removed for clarity. Panels A and B reproduce the results in Fig. 3C and
-Fig. 3D respectively with the applied field in the $x$-direction 
-(out of the plane of the paper). Also included is a cut of the structure 
-factor with $k_y = 0$ (panel C). 
-Panels D, E, and F show the corresponding situation after the
-external field has been removed and the structure allowed to relax,
-and show residual hexagonal ordering. The structure factor data is for wave
-vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
-$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
-\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
-\end{center}
-\caption{Cycling the electric field. Starting from the final position of
-Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
-$y$-direction. The corresponding view of the particle distribution is
-shown, again viewing along the field direction, in panel A.
-Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
-shown in panels B and C, respectively. The corresponding situation when
-the field is again switched off, and the structure allowed to relax,
-is shown in panels d, E, and F. The wave vector values used are the
-same as in Supplementary Fig.~7.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1345.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1347.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1347.png}
-\end{center}
-\caption{Cycling the electric field further. Starting from the final
-position of Supplementary Fig.~8 panel~D, the field is re-applied
-again in the $x-$direction. Panel A shows the particle configuration
-viewed along the field direction, and panels B and C show the structure
-factor with $k_x = 0$ and $k_y = 0$ respectively. The corresponding
-situation when the field is removed and the structure allowed to relax
-is shown in panels D, E, and F. The wave vector values are again as
-in Supplementary Fig.~7. Note that the on and off states are 
-equivalent to those in Supplementary Fig.~7, up to a rotation.
-This is because all hexagonal structures perpendicular to the field
-direction are energetically degenerate,
-and the actual orientation of the hexagon emerges spontaneously during
-switching. }
-\end{figure}
-
 %\newpage
@@
 
-{\bf Captions for Supplementary Movies} \\
+%{\bf Captions for Supplementary Movies} \\
+%
+%{\bf Supplementary Movie 1:} 
+%The final state of the simulation as shown in Fig.~1A. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 1:} 
-The final state of the simulation as shown in Fig.~1A. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 2:} 
+%The final state of the simulation as shown in Fig.~1B. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 2:} 
-The final state of the simulation as shown in Fig.~1B. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 3:} 
+%The final state of the simulation as shown in Fig.~1C. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 3:} 
-The final state of the simulation as shown in Fig.~1C. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 4:} 
+%The final state of the simulation as shown in Fig.~1D. The movie
+%rotates the configuration to reveal more clearly its structure.\\
 
-{\bf Supplementary Movie 4:} 
-The final state of the simulation as shown in Fig.~1D. The movie
-rotates the configuration to reveal more clearly its structure.\\
+%{\bf Supplementary Movie 5:} 
+%The configuration after the initial relaxation phase in Fig.~3. The movie
+%shows a rotation of the configuration. The distribution of the colloidal
+%particles is isotropic, with the particles forming clusters and having a
+%tendency to occupy positions at the intersection of the BPIII disclination
+%lines.\\
 
-{\bf Supplementary Movie 5:} 
-The configuration after the initial relaxation phase in Fig.~3. The movie
-shows a rotation of the configuration. The distribution of the colloidal
-particles is isotropic, with the particles forming clusters and having a
-tendency to occupy positions at the intersection of the BPIII disclination
-lines.\\
-
-{\bf Supplementary Movie 6:} 
-Configuration in a homogeneous electric field with strength
-corresponding to ${\cal E}=0.8$. The movie
-shows a rotation of the configuration.
-A hexagonal pattern emerges where the particles form chains
-situated at the vertices of the honeycomb-like structure. \\
+%{\bf Supplementary Movie 6:} 
+%Configuration in a homogeneous electric field with strength
+%corresponding to ${\cal E}=0.8$. The movie
+%shows a rotation of the configuration.
+%A hexagonal pattern emerges where the particles form chains
+%situated at the vertices of the honeycomb-like structure. \\
 

### 2014-04-16 trunk/papers/2012-sandwich/CholColloids.tex
updates following editorial comments plus cover art
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2363 40876441-3496-0410-a613-a2fca35b2ae2

@@
 computational approach are set out in the Methods section, while
-simulation parameters are reported in the Supplementary Information.
-}
+simulation parameters are reported in the Supplementary Methods.
 
@@
 \textbf{Structure in bulk BPI.}
-We show in Fig.~1 A-D the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles at random within it.
+We show in Fig.~1~(\textbf{a}--\textbf{d})
+the final state structures of four simulations in which we first equilibrated a stable BPI disclination network in bulk (with periodic boundary conditions) and then dispersed colloidal particles at random within it.
 Among dimensionless control parameters are the particle solid volume fraction $\phi$  ($\phi = 1\%$ or $5\%$); the ratio $r = R/\lambda$ of particle radius to the BP lattice parameter $\lambda$ (here $r\simeq 0.14$) and the ratio $w = WR/K$ where $W$ is the anchoring strength of the cholesteric at the particle surface, and $K$ is the elastic constant of the liquid crystal.
-\textcolor{red}{
 The anchoring strength is defined such that $W > 0$ imparts a preferred
@@
 In the colloidal-BP system, $w$ also plays a determining role in the final composite structure.
-When $w$ is small (Fig.~ 1A,B; Supplementary Movies~1 and~2), the attraction of particles to the disclination lattice causes
+When $w$ is small
+(Fig.~1\textbf{a} and Fig.~1\textbf{b}; Supplementary Movies~1 and~2),
+the attraction of particles to the disclination lattice causes
  the lattice to become covered by particles, with little disruption of its
@@
 Interestingly, the weak but finite anchoring apparently creates an elastic force between the colloidal particles which leads to very slow dynamics
-(Supplementary Fig.~4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.
-1A).
+(Supplementary Fig.~4) and the formation of small colloidal lines within disclinations, especially for dilute samples (Fig.~1\textbf{a}).
 It seems likely that polymer-stabilised  blue phases~\cite{kikuchi} work by a similar principle to this small $w$ case, with a weak segregation of monomers towards the defect regions.
 
-For large $w$ (Fig.~1C,D; Supplementary Movies~3 and~4), the picture changes
+For large $w$ (Fig.~1\textbf{c} and Fig.~1\textbf{d};
+Supplementary Movies~3 and~4), the picture changes
 dramatically.
@@
 As colloidal-nematic composites~\cite{tiffany}, all the structures in
-Fig.~1A-D 
+Fig.~1~(\textbf{a}--\textbf{d}) 
 should be soft solids, of nonzero elastic 
@@
 viscosity via permeation of molecules through a fixed defect lattice~\cite{permeation1,permeation2}. We expect the
-structures in Fig.~1A-C to be only weak gels, as here the 
+structures in Fig.~1~(\textbf{a}--\textbf{c}) to be only weak gels,
+as here the 
 defect network percolates but particle contacts do not. This is broadly similar to the cholesteric-nanoparticle gel reported
 in~\cite{lubensky} (with $G<1$ Pa). In contrast, the colloidal gel
-structure in Fig.~1D should be much more resistant mechanically, 
+structure in Fig.~1\textbf{d} should be much more resistant mechanically, 
 due to the formation of a thick percolating particle network. The 
@@
 colloidal particles with those of rubber~\cite{ramos} for
-the relevant simulation parameters (see Supplementary Information).}
+the relevant simulation parameters (see Supplementary Table~1).
 
-A more practical protocol for including particles in a BPI phase is to disperse the particles at random in an isotropic phase, and then quenching this into the temperature range where BPI is stable.
-Here we know that, without particles, the kinetics can favour disordered (BPIII-like) intermediates which may be long lived or even metastable~\cite{domaingrowth}.
-Fig.~1E-H shows outcomes of such a quench protocol with the same parameter sets as in Fig.~1A-D.
-For small $w$, the particles again decorate the disclination lines, but are this time templated onto an amorphous network rather than an ordered one.
-Increasing the volume fraction (Fig.~1F) leads to a more even coverage with
-a better defined particle spacing. 
+A more practical protocol for including particles in a BPI phase is to
+disperse the particles at random in an isotropic phase, and then
+quenching this into the temperature range where BPI is stable.
+Here we know that, without particles, the kinetics can favour
+disordered (BPIII-like) intermediates which may be long lived
+or even metastable~\cite{domaingrowth}. Fig.~1~(\textbf{e}--\textbf{h})
+shows outcomes of such a quench protocol with the same parameter sets as
+in Fig.~1~(\textbf{a}--\textbf{d}). For small $w$, the particles again
+decorate the disclination lines, but are this time templated onto an
+amorphous network rather than an ordered one.
+Increasing the volume fraction (Fig.~1\textbf{f}) leads to a more even
+coverage with a better defined particle spacing. 
 For large $w$, on the other hand, strong interparticle forces again dictate
-the self-assembly.
-At low concentrations, discrete particle clusters decorate an amorphous defect lattice (Fig.~1G) whereas at higher $\phi$ a percolating network of colloidal strands (the gel in Fig.~1H) is again seen.
-The structures in Fig. 1A-H have been followed for $\sim$20 ms 
+the self-assembly. At low concentrations, discrete particle clusters
+decorate an amorphous defect lattice (Fig.~1\textbf{g}) whereas at
+higher $\phi$ a percolating network of colloidal strands
+(the gel in Fig.~1\textbf{h}) is again seen. 
+The structures in Fig.~1~(\textbf{a}--\textbf{h}) have been followed for
+$\sim$20 ms 
 (see Supplementary Fig.~4); they correspond to either very slowly evolving
@@
 Supplementary Fig.~4 and Ref.~\cite{stark,lavrentovich}).
+
+It may be possible to anneal the structures obtained via the quench route,
+although we have not attempted this in simulation. The results presented
+suggest that even when starting with colloidal particles randomly dispersed
+in an ordered blue phase, the particles are able to distort the structure
+(Fig.~1~\textbf{a},\textbf{b}). This suggests that it would be difficult
+to identify a quench
+protocol which would lead to less, rather than more, disorder.
 
@@
 
-Fig.~2A-D shows the case of normal wall anchoring on varying $\phi$ and $w$.
-We quench from the isotropic to the ordered phase as was done in Fig.~1E-H, and again find a metastable, amorphous disclination network (see Supplementary Information).
-The resulting colloidal-cholesteric composites are somewhat similar to those seen in the bulk: for small $w$ the particles are templated by the network (Fig.~2A,B), whereas for large $w$ we observe isolated clusters (Fig.~2C) and percolated colloidal gels (Fig.~2D), for $\phi = 1\%$ and $4\%$ respectively.
+Fig.~2~(\textbf{a}--\textbf{d}) shows the case of normal wall anchoring on
+varying $\phi$ and $w$. We quench from the isotropic to the ordered phase
+as was done in Fig.~1~(\textbf{e}--\textbf{h}), and again find a metastable,
+amorphous disclination network
+(see Supplementary Methods). The resulting colloidal-cholesteric composites
+are somewhat similar to those seen in the bulk: for small $w$ the particles
+are templated by the network (Fig.~2\textbf{a},\textbf{b}), whereas for
+large $w$ we observe isolated clusters (Fig.~2\textbf{c}) and percolated
+colloidal gels (Fig.~2\textbf{d}), for $\phi = 1\%$ and $4\%$ respectively.
 There are, however, significant effects of the confining walls.
-First, the normal anchoring at the wall recruits particles there,  giving (for the chosen $h$) a density enhancement of the colloid particles at both walls 
-(see side-views in Fig.~2A-D). Second, the clusters formed at large $w$ are themselves anisotropic with an oblate habit. These clusters
+First, the normal anchoring at the wall recruits particles there,  giving
+(for the chosen $h$) a density enhancement of the colloid particles at
+both walls  (see side-views in Fig.~2\textbf{a}--\textbf{d}). Second,
+the clusters formed at large $w$ are themselves anisotropic with an
+oblate habit. These clusters
 are similar to those reported experimentally for large colloid particles in a
@@
 
-Fig.~2E-H shows comparable simulations for planar wall anchoring. 
-Without colloidal particles, planar anchoring results in a regular
-cholesteric helix,
+Fig.~2~(\textbf{e}--\textbf{h}) shows comparable simulations for planar wall
+anchoring.  Without colloidal particles, planar anchoring results in a
+regular cholesteric helix,
 with axis perpendicular to the boundary walls. 
@@
 dispersing particles with small $w$ arrests the loop shrinking process,
-stabilising twisted rings (Fig.~2E) at small $\phi$, and more complicated 
-double-stranded structures at larger $\phi$ (Fig.~2F).
+stabilising twisted rings (Fig.~2\textbf{e}) at small $\phi$, and more
+complicated  double-stranded structures at larger $\phi$ (Fig.~2\textbf{f}).
 For larger $w$, we observe thick colloidal ``ropes" about 5 particles across
-(Fig.~2G,H). These ropes undulate to follow the chiral structure of the 
-underlying fluid.
-On increasing the density they develop branch points and finally again approach a percolating gel morphology.
-Note that planar wall anchoring is incompatible with the normal anchoring on the colloidal particles (for $w>0$), which are thus repelled from both walls (see the side views in Fig.~2E-H).
+(Fig.~2\textbf{g},\textbf{h}). These ropes undulate to follow the chiral
+structure of the 
+underlying fluid. On increasing the density they develop branch points and
+finally again approach a percolating gel morphology. Note that planar wall
+anchoring is incompatible with the normal anchoring on the colloidal
+particles (for $w>0$), which are thus repelled from both walls
+(see the side views in Fig.~2\textbf{e}--\textbf{h}).
 
-\textcolor{red}{
 The degree of similarity between
@@
 these structures can be modified, or switched, with electric fields.
-Fig.~3 addresses this question, starting from amorphous Blue Phase III,
+Fig.~3 addresses this question, starting from amorphous blue phase~III,
 this time at a higher chirality. Dispersing colloidal particles in this
@@
 results in the particle occupying the available nodes, where the order
-parameter is low (Fig.~3A; Supplementary Movie~5). Applying a high enough
-field 
+parameter is low (Fig.~3\textbf{a}; Supplementary Movie~5).
+Applying a high enough field 
 along the $x$ direction leads to a switching to a new phase, where colloidal 
-arrays fit within a network with distorted hexagonal cells (Fig.~3B 
+arrays fit within a network with distorted hexagonal cells (Fig.~3\textbf{b} 
 and Supplementary Fig.~7; Supplementary Movie~6). 
@@
 (Supplementary Fig.~8 and Fig.~9).
+We note that for the fluid parameters used here, the field induced
+changes are irreversible with the weakly anchoring colloid particles.
+The mesophase without colloid particles undergoes
+a reversible ordering transition~\cite{bp3}; the reversibility
+is retained in the
+presence of particles only if their anchoring is very weak.
 
@@
 of the simulation parameters, and their mapping to physical units,
-may be found in the Supplementary Information.
+may be found in Supplementary Table~1 and the Supplementary Methods.
 
@@
 \begin{equation}
--\frac{\epsilon_a}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
+-\frac{\epsilon_\mathrm{a}}{12\pi} E_{\alpha}Q_{\alpha\beta}E_{\beta},
 \end{equation} 
 where $E_{\alpha}$ are the components of the electric field, and 
-$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
+$\epsilon_\mathrm{a}$ (here assumed positive) is the dielectric anisotropy
+of the liquid
 crystal. The strength of the electric field is quantified via
@@
 \begin{equation}
-{\cal E}^2 = \frac{27 \epsilon_a}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
+{\cal E}^2 = \frac{27 \epsilon_\mathrm{a}}{32 \pi A_0 \gamma} E_\alpha E_\alpha.
 \end{equation}
@@
 free energy density is also present to represent the energetic cost of
-anchoring at a solid surface. If the preferred order parameter tensor
-at the solid surface is $Q^0_{\alpha\beta}$, then the surface free energy
+anchoring at a solid surface.
+In the case of normal anchoring, the surface free energy
 density (per unit area) is
@@
 \end{equation}
-where $W$ is a constant determining the strength of the anchoring.
+where $Q^0_{\alpha\beta}$ is the preferred order
+parameter tensor at the solid surface, and $W$ is a constant determining
+the strength of the anchoring. In the case of planar anchoring, a slightly
+more complicated expression is required to allow for degeneracy
+(see, e.g., \cite{fournier2005}).
 The determination of $Q^0_{\alpha\beta}$ for normal and planar anchoring is
@@
 \left(\partial_t+ u_\nu \partial_\nu \right) Q_{\alpha\beta} - S_{\alpha\beta}
-= \Gamma H_{\alpha\beta}.
+= \mathit{\Gamma} H_{\alpha\beta}.
 \end{equation}
@@
 equilibrium on a timescale determined by a collective rotational diffusion 
-constant $\Gamma$. This relaxation is driven by the molecular field
+constant $\mathit{\Gamma}$. This relaxation is driven by the molecular field
 $H_{\alpha\beta}$, which is the functional derivative of the free energy
@@
 The value of the collective rotational diffusion used in all simulations is
-$\Gamma = 0.5$ in simulation units.
+$\mathit{\Gamma} = 0.5$ in simulation units.
 
@@
 units. Both potential and resulting force are smoothly matched to zero at
-a cut-off distance of $h_c = 0.25$ simulation units.
+a cut-off distance of $h_\mathrm{c} = 0.25$ simulation units.
 
@@
 Quasi-two-dimensional Skyrmion lattices in a chiral nematic liquid crystal,
-{\it Nat. Commun.} {\bf 2}, doi:10.1038/ncomms1250 (2011).
+{\it Nat. Commun.} {\bf 2}, 246 (2011).
 
@@
 Rheology of cubic blue phases,
-{\it Soft Matter} doi:10.1039/C3SM50228G (2013).
+{\it Soft Matter} 10243--10256 (2013).
 
@@
 a suspension of colloidal nanoparticles within a cholesteric liquid
-crystal in the BPI-forming region. (A-D) correspond to the case
+crystal in the BPI-forming region. (\textbf{a})--(\textbf{d}) correspond
+to the case
 of a dispersion of particles in a pre-equilibrated BPI phase;
-(E-H) are obtained by dispersing colloids in an isotropic
-phase and then quenching into the range where BPI is stable, leading to formation of an amorphous, BPIII-like, disclination network.
-Structures correspond
-to (A,E) $w=0.23$ and $\phi=1\%$;
-(B,F) $w=0.23$ and $\phi=5\%$; 
-(C,G) $w=2.3$ and $\phi=1\%$; and
-(D,H) $w=2.3$ and $\phi=5\%$.
+(\textbf{e})--(\textbf{h}) are obtained by dispersing colloids in an isotropic
+phase and then quenching into the range where BPI is stable, leading to
+formation of an amorphous, BPIII-like, disclination network.
+Structures correspond to:
+(\textbf{a}) and (\textbf{e}) $w=0.23$ and $\phi=1\%$;
+(\textbf{b}) and (\textbf{f}) $w=0.23$ and $\phi=5\%$; 
+(\textbf{c}) and (\textbf{g}) $w=2.3$ and $\phi=1\%$;
+(\textbf{d}) and (\textbf{h}) $w=2.3$ and $\phi=5\%$.
 The anchoring of the director field to the colloidal surface is normal.
 [For the full parameter list used to generate Figs.~1-3 see Supplementary
-Information.] For clarity, only a portion of the simulation box is shown
-\textcolor{red}{its linear extent being $\simeq 28R$}
+Notes.] For clarity, only a portion of the simulation box is shown
+its linear extent being $\simeq 28R$
 (one eighth of the total volume);
@@
 \caption{\textbf{Confined blue phase structures.}
-Snapshots of the steady states obtained when a dispersion of colloids in the isotropic phase is placed in a sandwich geometry, 
-and then quenched into a regime where BPI is stable in the bulk. Both top and side views are provided.
-The anchoring of the director field at the walls is normal for (A-D) and planar for (E-H). 
-Structures correspond to (A,E) $w=0.23$ and $\phi=1\%$, 
-(B,F) $w=0.23$ and $\phi=2\%$, 
-(C,G) $w=2.3$ and $\phi=1\%$,
-(D,H) $w=2.3$ and $\phi=2\%$.
+Snapshots of the steady states obtained when a dispersion of colloids in
+the isotropic phase is placed in a sandwich geometry, 
+and then quenched into a regime where BPI is stable in the bulk.
+Both top and side views are provided.
+The anchoring of the director field at the walls is normal for
+(\textbf{a})--(\textbf{d}) and planar for (\textbf{e})--(\textbf{h}). 
+Structures correspond to:
+(\textbf{a}) and (\textbf{e}) $w=0.23$ and $\phi=1\%$; 
+(\textbf{b}) and (\textbf{f}) $w=0.23$ and $\phi=2\%$; 
+(\textbf{c}) and (\textbf{g}) $w=2.3$ and $\phi=1\%$;
+(\textbf{d}) and (\textbf{h}) $w=2.3$ and $\phi=2\%$.
 As in Fig.~1, only one quarter of the simulation box is shown for clarity
-\textcolor{red}{(its horizontal extent being $\simeq 56R$ and vertical
-extent $\simeq 24R$ in the narrow direction)};
+(its horizontal extent being $\simeq 56R$ and vertical
+extent $\simeq 24R$ in the narrow direction);
 the full structures are shown in Supplementary Figs.~5 and~6.}
@@
 into a thermodynamically stable bulk BPIII with and without applied
-external electric field. Panel A shows the situation in zero field, and panel
-C the situation when the applied field is in the $x$-direction
-(perpendicular to the plane of the paper). Panels B and D show one
+external electric field. Panel~(\textbf{a}) shows the situation in
+zero field, and panel~(\textbf{c})
+the situation when the applied field is in the $x$-direction
+(perpendicular to the plane of the paper). Panels (\textbf{b})
+and~(\textbf{d}) show one
 plane of the three-dimensional structure factor $S(\mathbf{k})$ computed
 from the positions of the particles, with wavevector $k_x = 0$, corresponding
-to A and C respectively. With field, the disclinations form a honeycomb
+to~(\textbf{a}) and~(\textbf{c}) respectively. With field, the disclinations
+form a honeycomb
 pattern with hexagonal ordering perpendicular to the field direction;
@@
 Simulations use weakly anchoring particles $w=0.23$, and a volume
-fraction of $\phi=5\%$ (see Supplementary Information for full list of
+fraction of $\phi=5\%$ (see Supplementary Notes for full list of
 parameters).
-\textcolor{red}{The full simulation with a box size of $\simeq 56R$ is shown.}}
+The full simulation with a box size of $\simeq 56R$ is shown.}
 \end{figure}

### 2014-04-17 trunk/papers/2012-sandwich/CholColloids.tex
Further minor adjustments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2364 40876441-3496-0410-a613-a2fca35b2ae2

@@
 Because this structure has a wavelength in the visible range, 
-the resulting material should inherit (and likely enhance) the incomplete photonic bandgap of the parent blue phase, as exploited for instance in laser devices~\cite{bplasers}; for related possibilities see \cite{lavrentovich}.
+the resulting material should inherit (and probably enhance) the incomplete photonic bandgap of the parent blue phase, as exploited for instance in laser devices~\cite{bplasers}; for related possibilities see \cite{lavrentovich}.
 
@@
 \bibitem{bp3} Henrich, O., Stratford, K., Cates, M.E., and Marenduzzo, D.,
- Structure of blue phase III of cholesteric liquid crystals,
+ Structure of blue phase~III of cholesteric liquid crystals,
 {\it Phys. Rev. Lett.} {\bf 106}, 107802 (2011).
@@
 Lasing in a three-dimensional photonic crystal of the liquid crystal blue 
-phase II,
+phase~II,
 {\it Nat. Mat.} {\bf 1}, 111--113 (2002).

### 2014-04-17 trunk/papers/2012-sandwich/support.tex
Further minor adjustments
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2364 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
+% Prepend ``Supplementary'' Fig/Table caption headings
+\usepackage{caption}
+\captionsetup[table]{name=Supplementary Table}
+\captionsetup[figure]{name=Supplementary Figure}
+
+% Likewise...
+\renewcommand\refname{Supplementary References}
+
 \begin{document}
 
-{\bf
-Self-assembly of colloid-cholesteric
-composites provides a possible route to switchable optical materials.
+{\bf Supplementary Figures}
 
-Supplementary Information}
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.8]{support-fig1.png}
+\end{center}
+\caption{\textbf{Blue phase structures.}
+(A) Schematic representation of the director field
+in a cholesteric (left) and within a double twist cylinder (right).
+(B) Snapshot of the disclination network in an equilibrated 
+blue phase I structure. (C) Same as (B) for an amorphous blue phase III
+network.}
+\end{figure}
 
-Here, we give more details of the simulation method, protocol, and parameters used.
-We also provide additional figures complementing the main text, and discuss the matching of simulation units to physical units.
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig2.pdf}
+\end{center}
+\caption{\textbf{Bulk blue phase structures.}
+As for main manuscript Fig.~1~\textbf{a}--\textbf{d}
+(with liquid crystal order parameter
+initialised to equilibrium blue phase~I structure), but for the entire
+simulation system of 128$^3$ lattice sites. (A)~solid volume fraction~1\%
+and weak anchoring;  (B)~4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring. The view direction in the
+main figure is from the left here. There is a reference
+particle at the bottom left in each panel which does not take part
+in the simulation.}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.35]{support-fig3.pdf}
+\end{center}
+\caption{\textbf{Bulk blue phase structures.}
+As for main manuscript Fig.~1 \textbf{e}--\textbf{h}
+(initialised via a ``quench''
+to generate a disordered network), but for the entire simulation
+system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
+weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring. Again, there is a reference
+particle at the bottom left in each case which does not take part
+in the simulation.}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.49]{support-fig4.png}
+\end{center}
+\caption{\textbf{Free energy evolution.}
+Plot of the free energy (per BP unit cell) versus time
+for the simulations corresponding to Fig.~1 (\textbf{a}--\textbf{h})
+in the main text.}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig5.pdf}
+\end{center}
+\caption{\textbf{Confined blue phase structures.}
+As for main manuscript Fig.~2 \textbf{a}--\textbf{d}
+(confined geometry with normal
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (A) solid volume fraction 1\%
+and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
+(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[scale=0.42]{support-fig6.pdf}
+\end{center}
+\caption{\textbf{Confined blue phase structures.}
+As for main manuscript Fig.~2 \textbf{e}--\textbf{h}
+(confined geometry with planar
+anchoring walls at both top and bottom), but for the entire simulation
+system of 256$^2\times$56 lattice sites. Each shows a top view and
+side view of the same simulation state. (E) colloid solid volume fraction 1\%
+and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
+(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
+volume fraction and strong anchoring.}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
+\end{center}
+\caption{\textbf{Structure with and without electric field.}
+(Disclination lines are 
+removed for clarity.) Panels~A and~B reproduce the results in Fig.~3\textbf{c}
+and
+Fig.~3\textbf{d} respectively with the applied field in the $x$-direction 
+(out of the plane of the paper). Also included is a cut of the structure 
+factor with $k_y = 0$ (panel C). 
+Panels D, E, and F show the corresponding situation after the
+external field has been removed and the structure allowed to relax,
+and show residual hexagonal ordering. The structure factor data is for wave
+vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
+$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
+\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
+\end{center}
+\caption{\textbf{Cycling the electric field.}
+Starting from the final position of
+Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
+$y$-direction. The corresponding view of the particle distribution is
+shown, again viewing along the field direction, in panel A.
+Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
+shown in panels~B and~C, respectively. The corresponding situation when
+the field is again switched off, and the structure allowed to relax,
+is shown in panels D, E, and~F. The wave vector values used are the
+same as in Supplementary Fig.~7.}
+\end{figure}
+
+\newpage
+
+\begin{figure}[!h]
+\begin{center}
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
+\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}
+\includegraphics[width=0.32\columnwidth]{sq_y_run1347.png}
+\end{center}
+\caption{\textbf{Cycling the electric field further.}
+Starting from the final
+position of Supplementary Fig.~8 panel~D, the field is re-applied
+again in the $x-$direction. Panel A shows the particle configuration
+viewed along the field direction, and panels B and C show the structure
+factor with $k_x = 0$ and $k_y = 0$ respectively. The corresponding
+situation when the field is removed and the structure allowed to relax
+is shown in panels D, E, and F. The wave vector values are again as
+in Supplementary Fig.~7. Note that the on and off states are 
+equivalent to those in Supplementary Fig.~7, up to a rotation.
+This is because all hexagonal structures perpendicular to the field
+direction are energetically degenerate,
+and the actual orientation of the hexagon emerges spontaneously during
+switching. }
+\end{figure}
+
+\vfill\pagebreak
+
+{\bf Supplementary Table}
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+\hline
+         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
+\hline 
+Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
+0.23, 2.3\\
+\hline
+Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
+0.23, 2.3 \\
+\hline
+Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
+\hline
+External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
+\hline
+\end{tabular}
+\end{center}
+\caption{\textbf{Free energy parameters used in the simulations.} Bulk BPI
+correspond to Fig.~1 (\textbf{a}--\textbf{d}), bulk quench to
+Fig.~1 (\textbf{e}--\textbf{h}); confined
+geometries are shown in Fig.~2; and the external field case is
+relevant for Fig.~3.}
+\label{table:params}
+\end{table}
+
+\vfill
+\pagebreak
+
+{\bf Supplementary Methods}
 
@@
 1\% and 4\%.
-\textcolor{red}{For both normal and planar anchoring at the walls,
+For both normal and planar anchoring at the walls,
 the strength is adjusted to correspond to that used for the strongly
-anchoring particles.}
+anchoring particles.
 All simulations are run for two million simulation time steps.
@@
 
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-\hline
-         & $A_0$ & $\gamma$ & $K$ & $p$ & $\kappa$ & $\tau$ & $WR/K$\\
-\hline 
-Bulk BPI & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
-0.23, 2.3\\
-\hline
-Bulk Quench & 0.01 & 3.086 & 0.007061 & 32$\sqrt{2}$ & 0.6902 & -0.2500 &
-0.23, 2.3 \\
-\hline
-Confined & 0.01 & 3.086 & 0.01897 & 64 & 0.8 & -0.2500 & 0.23, 2.3\\
-\hline
-External field & 0.004& 3.086 & 0.02 & 32 & 2.6 & -0.2500 & 0.23\\ 
-\hline
-\end{tabular}
-\end{center}
-\caption{\textbf{Free energy parameters used in the simulations.} Bulk BPI
-correspond to Fig.~1 (A-D), bulk quench to Fig.~1 (E--H); confined
-geometries are shown in Fig.~2; and the external field case is
-relevant for Fig.~3.}
-\label{table:params}
-\end{table}
-
 \noindent
@@
 In all cases visualisation of the defect lines is carried out via
-an isosurface of the scalar order parameter $q$ determined from the
+an isosurface of the scalar order parameter~$q$ determined from the
 largest eigenvalue of the local order parameter tensor. A low value
-of the scalar order parameter $q$ (typically in the range 0.12--0.14)
+of the scalar order parameter~$q$ (typically in the range~0.12--0.14)
 unambiguously identifies the disclinations.
@@
 Given the previous value for $A_0$, and a dielectric anisotropy of 20
-(or equivalently $\epsilon_a=20\epsilon_0$, with $\epsilon_0$ the
+(or equivalently $\epsilon_\mathrm{a}=20\epsilon_0$, with $\epsilon_0$ the
 dielectric permittivity of vacuum), an electric field
@@
 $\lambda =$100--500~nm, a corresponding shear modulus $G \simeq
-K/\lambda^2 =$0.3--7~kPa.
-
-\vfill\pagebreak
-
-{\bf Supplementary Figures and Animations}
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.8]{support-fig1.png}
-\end{center}
-\caption{\textbf{Blue phase structures.}
-(A) Schematic representation of the director field
-in a cholesteric (left) and within a double twist cylinder (right).
-(B) Snapshot of the disclination network in an equilibrated 
-blue phase I structure. (C) Same as (B) for an amorphous blue phase III
-network.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.35]{support-fig2.pdf}
-\end{center}
-\caption{\textbf{Bulk blue phase structures.}
-As for main manuscript Fig.~1 A--D
-(with liquid crystal order parameter
-initialised to equilibrium blue phase I structure), but for the entire
-simulation system of 128$^3$ lattice sites. (A)  solid volume fraction~1\%
-and weak anchoring;  (B) 4\% solid volume fraction and weak anchoring;
-(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
-volume fraction and strong anchoring. The view direction in the
-main figure is from the left here. There is a reference
-particle at the bottom left in each panel which does not take part
-in the simulation.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.35]{support-fig3.pdf}
-\end{center}
-\caption{\textbf{Bulk blue phase structures.}
-As for main manuscript Fig.~1 E--H (initialised via a ``quench''
-to generate a disordered network), but for the entire simulation
-system of 128$^3$ lattice sites. (E) solid volume fraction 1\% and
-weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
-(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
-volume fraction and strong anchoring. Again, there is a reference
-particle at the bottom left in each case which does not take part
-in the simulation.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.49]{support-fig4.png}
-\end{center}
-\caption{\textbf{Free energy evolution.}
-Plot of the free energy (per BP unit cell) versus time
-for the simulations corresponding to Fig. 1A-H in the main text
-(the labels correspond to those in Fig. 1).
-}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.42]{support-fig5.pdf}
-\end{center}
-\caption{\textbf{Confined blue phase structures.}
-As for main manuscript Fig.~2 A--D (confined geometry with normal
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (A) solid volume fraction 1\%
-and weak anchoring; (B) 4\% solid volume fraction and weak anchoring;
-(C) 1\% solid volume fraction and strong anchoring; (D) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[scale=0.42]{support-fig6.pdf}
-\end{center}
-\caption{\textbf{Confined blue phase structures.}
-As for main manuscript Fig.~2 E--H (confined geometry with planar
-anchoring walls at both top and bottom), but for the entire simulation
-system of 256$^2\times$56 lattice sites. Each shows a top view and
-side view of the same simulation state. (E) colloid solid volume fraction 1\%
-and weak anchoring; (F) 4\% solid volume fraction and weak anchoring;
-(G) 1\% solid volume fraction and strong anchoring; (H) 4\% solid
-volume fraction and strong anchoring.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1341.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1341_2.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1341.png}
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1342.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1342.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1342.png}
-\end{center}
-\caption{\textbf{Structure with and without electric field.}
-(Disclination lines are 
-removed for clarity.) Panels A and B reproduce the results in Fig. 3C and
-Fig. 3D respectively with the applied field in the $x$-direction 
-(out of the plane of the paper). Also included is a cut of the structure 
-factor with $k_y = 0$ (panel C). 
-Panels D, E, and F show the corresponding situation after the
-external field has been removed and the structure allowed to relax,
-and show residual hexagonal ordering. The structure factor data is for wave
-vectors $k_x = 0; (k_y, k_z) \in [-3\pi/8 , 3\pi/8]$ (panels B and E) and
-$k_y = 0; (k_x, k_z) \in [-3\pi/8, 3\pi/8]$ (panels C and F).}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1343.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1343.png}\\
-\includegraphics[width=0.34\columnwidth]{col_parallel_y_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1344.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1344.png}\\
-\end{center}
-\caption{\textbf{Cycling the electric field.}
-Starting from the final position of
-Supplementary Fig.~7 panel~D, the field is re-applied, this time in the
-$y$-direction. The corresponding view of the particle distribution is
-shown, again viewing along the field direction, in panel A.
-Corresponding structure factor cuts with $k_x = 0$ and $k_y = 0$ are
-shown in panels B and C, respectively. The corresponding situation when
-the field is again switched off, and the structure allowed to relax,
-is shown in panels D, E, and F. The wave vector values used are the
-same as in Supplementary Fig.~7.}
-\end{figure}
-
-\newpage
-
-\begin{figure}[!h]
-\begin{center}
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1345.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1345.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1345.png}\\
-\includegraphics[width=0.34\columnwidth]{col_parallel_x_run1347.png}
-\includegraphics[width=0.32\columnwidth]{sq_x_run1347.png}
-\includegraphics[width=0.32\columnwidth]{sq_y_run1347.png}
-\end{center}
-\caption{\textbf{Cycling the electric field further.}
-Starting from the final
-position of Supplementary Fig.~8 panel~D, the field is re-applied
-again in the $x-$direction. Panel A shows the particle configuration
-viewed along the field direction, and panels B and C show the structure
-factor with $k_x = 0$ and $k_y = 0$ respectively. The corresponding
-situation when the field is removed and the structure allowed to relax
-is shown in panels D, E, and F. The wave vector values are again as
-in Supplementary Fig.~7. Note that the on and off states are 
-equivalent to those in Supplementary Fig.~7, up to a rotation.
-This is because all hexagonal structures perpendicular to the field
-direction are energetically degenerate,
-and the actual orientation of the hexagon emerges spontaneously during
-switching. }
-\end{figure}
+K/\lambda^2 =$ 0.3--7~kPa.
 

### 2014-08-08 branches/charge/bbl.c
Remove old branches
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2419 40876441-3496-0410-a613-a2fca35b2ae2

@@ -1,800 +1,0 @@
-/*****************************************************************************
- *
- *  bbl.c
- *
- *  Bounce back on links.
- *
- *  $Id: bbl.c,v 1.11 2010-10-15 12:40:02 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <math.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "physics.h"
-#include "colloids.h"
-#include "colloid_sums.h"
-#include "model.h"
-#include "util.h"
-#include "wall.h"
-#include "phi.h"
-#include "bbl.h"
-
-static void bounce_back_pass1(void);
-static void bounce_back_pass2(void);
-static void mass_conservation_compute_force(void);
-static void update_colloids(void);
-
-static int bbl_active_ = 0;  /* Flag for active particles. */
-static double deltag_ = 0.0; /* Excess or deficit of phi between steps */
-static double stress_[3][3]; /* Surface stress */
-
-/*****************************************************************************
- *
- *  bounce_back_on_links
- *
- *  Driver routine for colloid bounce back on links.
- *
- *  The basic method is:
- *  Nguyen and Ladd [Phys. Rev. E {\bf 66}, 046708 (2002)].
- *
- *  The implicit velocity update requires two sweeps through the
- *  boundary nodes:
- *
- *  (1) Compute the velocity-independent force and torque on each
- *      colloid and the elements of the drag matrix for each colloid.
- *
- *  (2) Update the velocity of each colloid.
- *
- *  (3) Do the actual BBL on distributions with the updated colloid
- *      velocity.
- *
- *****************************************************************************/
-
-void bounce_back_on_links() {
-
[…]

### 2014-08-08 branches/charge/blue_phase_init.h

@@ -1,38 +1,0 @@
-/*****************************************************************************
- *
- *  blue_phase_init.h
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Oliver Henrich (o.henrich@ucl.ac.uk) wrote most of these.
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef BLUE_PHASE_INIT_H
-#define BLUE_PHASE_INIT_H
-
-void blue_phase_O8M_init(void);
-void blue_phase_O2_init(void);
-void blue_phase_O5_init(void);
-void blue_phase_H2D_init(void);
-void blue_phase_H3DA_init(void);
-void blue_phase_H3DB_init(void);
-void blue_phase_DTC_init(void);
-void blue_phase_BPIII_init(const double specs[3]);
-
-void blue_phase_nematic_init(const double n[3]);
-void blue_phase_twist_init(const int helical_axis);
-void blue_phase_chi_edge(int N, double z0, double x0);
-void blue_set_random_q_init(void);
-void blue_set_random_q_rectangle_init(const double xmin, const double xmax,
-				      const double ymin,const double ymax,
-				      const double zmin,const double zmax);
-void blue_phase_M_rot(double M[3][3], int dim, double alpha);
-void blue_phase_init_amplitude_set(const double amplitude);
-double blue_phase_init_amplitude(void);
-
-#endif

### 2014-08-08 branches/charge/colloids_Q_tensor.c

@@ -1,623 +1,0 @@
-/*****************************************************************************
- *
- *  colloids_Q_tensor.c
- *
- *  Routines to set the Q tensor inside a colloid to correspond
- *  to homeotropic (normal) or planar anchoring at the surface.
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Juho Lintuvuori (jlintuvu@ph.ed.ac.uk)
- *  (c) 2011 The University of Edinburgh
- *  
- *****************************************************************************/
-
-#include <stdio.h>
-#include <stdlib.h>
-#include <math.h>
-#include <assert.h>
-
-#include "pe.h"
-#include "build.h"
-#include "coords.h"
-#include "colloids.h"
-#include "phi.h"
-#include "io_harness.h"
-#include "lattice.h"
-#include "util.h"
-#include "model.h"
-#include "site_map.h"
-#include "blue_phase.h"
-#include "colloids_Q_tensor.h"
-
-struct io_info_t * io_info_scalar_q_;
-static int anchoring_coll_ = ANCHORING_NORMAL;
-static int anchoring_wall_ = ANCHORING_NORMAL;
-static int anchoring_method_ = ANCHORING_METHOD_NONE;
-static double w_surface_ = 0.0; /* Anchoring strength in free energy */
-static double w_surface_wall_ = 0.0; /* Anchoring strength in free energy */
-
-static int scalar_q_dir_write(FILE * fp, const int i, const int j, const int k);
-static int scalar_q_dir_write_ascii(FILE *, const int, const int, const int);
-static void scalar_biaxial_order_parameter_director(double q[3][3], double qs[5]);
-
-void COLL_set_Q(){
-
-  int ia;
-  int ic,jc,kc;
-  
-  colloid_t * p_colloid;
-
-  double rsite0[3];
-  double normal[3];
-  double dir[3];
-
-  colloid_t * colloid_at_site_index(int);
-
-  int nlocal[3],offset[3];
-  int index;
-
-  double q[3][3];
-  double qs[5];
-  double director[3];
-  double len_normal;
-  double rlen_normal;
-  double amplitude;
-  double rdotd;
-  double dir_len;
-
[…]

### 2014-08-08 branches/charge/model.c

@@ -1,930 +1,0 @@
-/*****************************************************************************
- *
- *  model.c
- *
- *  This encapsulates data/operations related to distributions.
- *  However, the implementation of the distribution is exposed
- *  for performance-critical operations ehich, for space
- *  considerations, are not included in this file.
- *
- *  The LB model is either D2Q9, D3Q15 or D3Q19, as included in model.h.
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "io_harness.h"
-#include "model.h"
-
-const double cs2  = (1.0/3.0);
-const double rcs2 = 3.0;
-
-double * f_;
-
-static int ndist_ = 1;
-static int nsite_ = 0;
-static int initialised_ = 0;
-static struct io_info_t * io_info_distribution_; 
-
-static MPI_Datatype plane_xy_full_;
-static MPI_Datatype plane_xz_full_;
-static MPI_Datatype plane_yz_full_;
-static MPI_Datatype plane_xy_reduced_[2];
-static MPI_Datatype plane_xz_reduced_[2];
-static MPI_Datatype plane_yz_reduced_[2];
-static MPI_Datatype plane_xy_[2];
-static MPI_Datatype plane_xz_[2];
-static MPI_Datatype plane_yz_[2];
-
-static MPI_Datatype site_x_[2];
-static MPI_Datatype site_y_[2];
-static MPI_Datatype site_z_[2];
-
-static void distribution_mpi_init(void);
-static void distribution_set_types(const int, MPI_Datatype *);
-static void distribution_set_blocks(const int, int *, const int, const int *);
-static void distribution_set_displacements(const int, MPI_Aint *, const int,
-					   const int *);
-static int distributions_read(FILE *, const int, const int, const int);
[…]

### 2014-08-08 branches/charge/psi.h

@@ -1,48 +1,0 @@
-/*****************************************************************************
- *
- *  psi.h
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2012 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef PSI_H
-#define PSI_H
-
-typedef struct psi_s psi_t;
-
-int psi_create(int nk, psi_t ** pobj);
-void psi_free(psi_t * obj);
-int psi_init_io_info(psi_t * obj, int grid[3]);
-
-int psi_nk(psi_t * obj, int * nk);
-int psi_valency(psi_t * obj, int n, int * iv);
-int psi_valency_set(psi_t * obj, int n, int iv);
-int psi_diffusivity(psi_t * obj, int n, double * diff);
-int psi_diffusivity_set(psi_t * obj, int n, double diff);
-int psi_halo_psi(psi_t * obj);
-int psi_halo_rho(psi_t * obj);
-
-int psi_rho(psi_t * obj, int index, int n, double * rho);
-int psi_rho_set(psi_t * obj, int index, int n, double rho);
-int psi_psi(psi_t * obj, int index, double * psi);
-int psi_psi_set(psi_t * obj, int index, double psi);
-int psi_rho_elec(psi_t * obj, int index, double * rho_elec);
-int psi_unit_charge(psi_t * obj, double * eunit);
-int psi_unit_charge_set(psi_t * obj, double eunit); 
-int psi_beta(psi_t * obj, double * beta);
-int psi_beta_set(psi_t * obj, double beta);
-int psi_epsilon(psi_t * obj, double * epsilon);
-int psi_epsilon_set(psi_t * obj, double epsilon);
-int psi_ionic_strength(psi_t * psi, int index, double * sion);
-int psi_bjerrum_length(psi_t * obj, double * lb);
-int psi_debye_length(psi_t * obj, double rho_b, double * ld);
-int psi_surface_potential(psi_t * obj, double sigma, double rho_b,
-			  double * sp);
-#endif

### 2014-08-08 branches/charge/psi_sor.h

@@ -1,14 +1,0 @@
-/*****************************************************************************
- *
- *  psi_sor.h
- *
- *****************************************************************************/
-
-#ifndef PSI_SOR_H
-#define PSI_SOR_H
-
-#include "psi.h"
-
-int psi_sor_poisson(psi_t * obj, double tol_abs, double tol_rel);
-
-#endif

### 2014-08-08 branches/d2008-6-1/src/collision.c

@@ -1,1038 +1,0 @@
-/*****************************************************************************
- *
- *  collision.c
- *
- *  Collision stage routines and associated data.
- *
- *  $Id: collision.c,v 1.19 2009-06-19 12:26:49 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2007 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <stdio.h>
-#include <math.h>
-#include <string.h> /* Remove strcmp */
-
-#include "pe.h"
-#include "ran.h"
-#include "timer.h"
-#include "coords.h"
-#include "physics.h"
-#include "runtime.h"
-#include "control.h"
-#include "free_energy.h"
-#include "phi.h"
-#include "phi_gradients.h"
-#include "phi_force.h"
-#include "phi_cahn_hilliard.h"
-#include "phi_stats.h"
-#include "lattice.h"
-
-#include "util.h"
-#include "utilities.h"
-#include "communicate.h"
-#include "leesedwards.h"
-#include "model.h"
-#include "collision.h"
-
-#include "site_map.h"
-#include "io_harness.h"
-
-extern Site * site;
-
-/* Variables (not static) */
-
-double    siteforce[3];
-
-/* Variables concerned with the collision */
-
-static int    nmodes_ = NVEL;  /* Modes to use in collsion stage */
-
-static double  noise0 = 0.1;   /* Initial noise amplitude    */
-static double rtau_shear;      /* Inverse relaxation time for shear modes */
-static double rtau_bulk;       /* Inverse relaxation time for bulk modes */
-static double rtau_ghost = 1.0;/* Inverse relaxation time for ghost modes */
-static double var_shear;       /* Variance for shear mode fluctuations */
-static double var_bulk;        /* Variance for bulk mode fluctuations */
-
-static double noise_var[NVEL]; /* Noise variances */
-
-void MODEL_collide_multirelaxation(void);
-void MODEL_collide_binary_lb(void);
-
[…]

### 2014-08-08 branches/d2008-6-1/src/phi.h

@@ -1,42 +1,0 @@
-/*****************************************************************************
- *
- *  phi.h
- *
- *  $Id: phi.h,v 1.3 2008-11-14 14:42:50 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group
- *  and Edinburgh Parallel Computing Centre
- *  (c) 2008 The University of Edinburgh
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *
- *****************************************************************************/
-
-#ifndef PHI_H
-#define PHI_H
-
-void phi_init(void);
-void phi_finish(void);
-void phi_halo(void);
-void phi_compute_phi_site(void);
- 
-double phi_get_phi_site(const int);
-void   phi_set_phi_site(const int, const double);
-double phi_get_delsq_phi_site(const int);
-void   phi_set_delsq_phi_site(const int, const double);
-void   phi_get_grad_phi_site(const int, double *);
-void   phi_set_grad_phi_site(const int, const double *);
-void   phi_get_grad_delsq_phi_site(const int, double *);
-double phi_get_delsq_delsq_phi_site(const int);
-void   phi_leesedwards_transformation(void);
-int    phi_is_finite_difference(void);
-void   phi_set_finite_difference(void);
-
-double phi_op_get_phi_site(const int, const int);
-double phi_op_get_delsq_phi_site(const int, const int);
-void   phi_op_get_grad_phi_site(const int, const int, double *);
-void   phi_op_set_phi_site(const int, const int, const double);
-
-extern struct io_info_t * io_info_phi;
-extern const int nop_;
-#endif

### 2014-08-08 branches/d2009-10-1/src/advection_bcs.c

@@ -1,137 +1,0 @@
-/****************************************************************************
- *
- *  advection_bcs.c
- *
- *  Advection boundary conditions.
- *
- *  $Id: advection_bcs.c,v 1.1.2.4 2010-04-02 07:50:28 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2009 The University of Edinburgh
- *
- ****************************************************************************/
-
-#include <assert.h>
-
-#include "pe.h"
-#include "wall.h"
-#include "coords.h"
-#include "site_map.h"
-#include "phi.h"
-
-/*****************************************************************************
- *
- *  advection_bcs_no_normal_fluxes
- *
- *  Set normal fluxes at solid fluid interfaces to zero.
- *
- *****************************************************************************/
-
-void advection_bcs_no_normal_flux(double * fluxe, double * fluxw,
-				  double * fluxy, double * fluxz) {
-
-  int nlocal[3];
-  int ic, jc, kc, index, n;
-  int nop;
-
-  double mask, maskw, maske, masky, maskz;
-
-  assert(fluxe);
-  assert(fluxw);
-  assert(fluxy);
-  assert(fluxz);
-
-  coords_nlocal(nlocal);
-  nop = phi_nop();
-
-  for (ic = 1; ic <= nlocal[X]; ic++) {
-    for (jc = 0; jc <= nlocal[Y]; jc++) {
-      for (kc = 0; kc <= nlocal[Z]; kc++) {
-
-	index = coords_index(ic, jc, kc);
-
-	mask  = (site_map_get_status_index(index)  == FLUID);
-	maske = (site_map_get_status(ic+1, jc, kc) == FLUID);
-	maskw = (site_map_get_status(ic-1, jc, kc) == FLUID);
-	masky = (site_map_get_status(ic, jc+1, kc) == FLUID);
-	maskz = (site_map_get_status(ic, jc, kc+1) == FLUID);
-
-	for (n = 0;  n < nop; n++) {
-	  fluxw[nop*index + n] *= mask*maskw;
-	  fluxe[nop*index + n] *= mask*maske;
-	  fluxy[nop*index + n] *= mask*masky;
-	  fluxz[nop*index + n] *= mask*maskz;
-	}
-
-      }
-    }
-  }
-
-  return;
-}
-
[…]

### 2014-08-08 branches/d2009-10-1/src/brazovskii_rt.h

@@ -1,20 +1,0 @@
-/****************************************************************************
- *
- *  brazovskii_rt.h
- *
- *  $Id: brazovskii_rt.h,v 1.1.2.1 2009-11-04 09:53:39 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group
- *  and Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) The University of Edinburgh (2009)
- *
- ****************************************************************************/
-
-#ifndef BRAZOVSKII_RT_H
-#define BRAZOVSKII_RT_H
-
-void brazovskii_run_time(void);
-
-#endif

### 2014-08-08 branches/d2009-10-1/src/brownian.c

@@ -1,511 +1,0 @@
-/*****************************************************************************
- *
- *  brownian.c
- *
- *  $Id: brownian.c,v 1.3.16.3 2010-09-30 18:02:35 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2007 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <math.h>
-
-#include "pe.h"
-#include "ran.h"
-#include "coords.h"
-#include "colloids.h"
-#include "colloids_halo.h"
-#include "physics.h"
-#include "util.h"
-
-void brownian_step_ermak_buckholz(void);
-void brownian_step_no_inertia(void);
-void brownian_step_no_inertia_test(void);
-void brownian_set_random(void);
-
-static double dt_ = 1.0;  /* Time step; fixed here to be 1 LB step */
-#define NTIME 16
-
-/*****************************************************************************
- *
- *  do_brownian_dynamics
- *
- *  Perform a number of Brownian Dynamics steps.
- *
- *****************************************************************************/
-
-void do_brownian_dynamics() {
-
-  int     ia;
-  int     ic, jc, kc;
-  colloid_t * p_colloid;
-  int i, n, nt, nmax = 100000;
-  double u[3][NTIME];
-  double r[3][NTIME];
-  double s[3][NTIME];
-  double correlator_uu[NTIME];
-  double correlator_ur[NTIME];
-  double correlator_rr[NTIME];
-  double correlator_ss[NTIME];
-  double mass, scaleu, scalex, beta;
-  double dr[3];
-
-  /* Brownian dynamics requires an update. */
-  assert(0);
-
-  brownian_step_no_inertia_test();
-  if (1) return;
-
-  mass  = (4.0/3.0)*pi_*pow(2.3, 3);
-  beta  = 6.0*pi_*get_eta_shear()*2.3/mass;
[…]

### 2014-08-08 branches/d2009-10-1/src/colloid.h

@@ -1,51 +1,0 @@
-/*****************************************************************************
- *
- *  colloid.h
- *
- *  The implementation is exposed for the time being.
- *
- *  $Id: colloid.h,v 1.1.2.1 2010-06-06 11:39:28 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef COLLOID_H
-#define COLLOID_H
-
-#include <stdio.h>
-
-typedef struct colloid_state_type colloid_state_t;
-
-struct colloid_state_type {
-
-  int    index;         /* Unique global index for colloid */
-  int    rebuild;       /* Rebuild flag */
-  double a0;            /* Input radius (lattice units) */
-  double ah;            /* Hydrodynamic radius (from calibration) */
-  double r[3];          /* Position */
-  double v[3];          /* Velocity */
-  double w[3];          /* Angular velocity omega */
-  double s[3];          /* Magnetic dipole, or spin */
-  double m[3];          /* Currect direction of motion vector (squirmer) */
-  double b1;	        /* squirmer active parameter b1 */
-  double b2;            /* squirmer active parameter b2 */
-  double c;             /* Wetting free energy parameter C */
-  double h;             /* Wetting free energy parameter H */
-  double dr[3];         /* r update (pending refactor of move/build process) */
-  double deltaphi;      /* order parameter bbl net; required to restart */
-  double spare1;        /* spare scalar */
-  double spare2[3];     /* spare vector */
-
-};
-
-int colloid_state_read_ascii(colloid_state_t * ps, FILE * fp);
-int colloid_state_read_binary(colloid_state_t * ps, FILE * fp);
-int colloid_state_write_ascii(colloid_state_t ps, FILE * fp);
-int colloid_state_write_binary(colloid_state_t ps, FILE * fp);
-
-#endif

### 2014-08-08 branches/d2009-10-1/src/distribution_rt.h

@@ -1,20 +1,0 @@
-/*****************************************************************************
- *
- *  distribution_rt.h
- *
- *  $Id: distribution_rt.h,v 1.1.2.1 2010-08-04 17:48:39 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef DISTRIBUTION_RT_H
-#define DISTRIBUTION_RT_H
-
-void distribution_run_time(void);
-
-#endif

### 2014-08-08 branches/d2009-10-1/src/free_energy_scalar.h

@@ -1,24 +1,0 @@
-/*****************************************************************************
- *
- *  free_energy_scalar.h
- *
- *  Abstract free energy for scalar order parameters.
- *
- *  $Id: free_energy_scalar.h,v 1.1.2.1 2010-03-21 13:43:23 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 the University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef FREE_ENERGY_SCALAR_H
-#define FREE_ENERGY_SCALAR_H
-
-#include "free_energy.h"
-
-/* Move chemical potential here */
-
-#endif

### 2014-08-08 branches/d2009-10-1/src/model.c

@@ -1,833 +1,0 @@
-/*****************************************************************************
- *
- *  model.c
- *
- *  This encapsulates data/operations related to distributions.
- *  However, the implementation of the distribution is exposed
- *  for performance-critical operations ehich, for space
- *  considerations, are not included in this file.
- *
- *  The LB model is either D2Q9, D3Q15 or D3Q19, as included in model.h.
- *
- *  $Id: model.c,v 1.17.4.11 2010-08-04 17:49:17 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "io_harness.h"
-#include "model.h"
-
-const double cs2  = (1.0/3.0);
-const double rcs2 = 3.0;
-
-double * f_;
-
-static int ndist_ = 1;
-static int nsite_ = 0;
-static int initialised_ = 0;
-static struct io_info_t * io_info_distribution_; 
-
-static MPI_Datatype plane_xy_full_;
-static MPI_Datatype plane_xz_full_;
-static MPI_Datatype plane_yz_full_;
-static MPI_Datatype plane_xy_reduced_[2];
-static MPI_Datatype plane_xz_reduced_[2];
-static MPI_Datatype plane_yz_reduced_[2];
-static MPI_Datatype plane_xy_[2];
-static MPI_Datatype plane_xz_[2];
-static MPI_Datatype plane_yz_[2];
-
-static MPI_Datatype site_x_[2];
-static MPI_Datatype site_y_[2];
-static MPI_Datatype site_z_[2];
-
-static void distribution_mpi_init(void);
-static void distribution_set_types(const int, MPI_Datatype *);
-static void distribution_set_blocks(const int, int *, const int, const int *);
-static void distribution_set_displacements(const int, MPI_Aint *, const int,
-					   const int *);
[…]

### 2014-08-08 branches/d2009-10-1/src/phi.c

@@ -1,761 +1,0 @@
-/*****************************************************************************
- *
- *  phi.c
- *
- *  Scalar, vector, tensor, order parameter.
- *
- *  $Id: phi.c,v 1.11.4.12 2010-08-06 17:44:05 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2008 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <stdlib.h>
-#include <math.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "control.h"
-#include "io_harness.h"
-#include "leesedwards.h"
-#include "phi.h"
-
-struct io_info_t * io_info_phi;
-
-double * phi_site;
-
-static int nop_ = 0;                    /* Number of order parameter fields */
-static int initialised_ = 0;
-static int phi_finite_difference_ = 0;  /* Default is LB for order parameter */
-static MPI_Datatype phi_xy_t_;
-static MPI_Datatype phi_xz_t_;
-static MPI_Datatype phi_yz_t_;
-
-static void phi_init_mpi(void);
-static void phi_init_io(void);
-static int  phi_read(FILE *, const int, const int, const int);
-static int  phi_write(FILE *, const int, const int, const int);
-static int  phi_read_ascii(FILE *, const int, const int, const int);
-static int  phi_write_ascii(FILE *, const int, const int, const int);
-static void phi_leesedwards_parallel(void);
-
-
-/****************************************************************************
- *
- *  phi_init
- *
- *  Allocate memory for the order parameter.
- *
- *  Space for buffers to hold Lees Edwards interpolated quantities
- *  is added to the main array.
- *
- ****************************************************************************/
-
-void phi_init() {
-
-  int nhalo;
-  int nsites;
-  int nbuffer;
-  int nlocal[3];
-
-  nhalo = coords_nhalo();
-  coords_nlocal(nlocal);
-  nbuffer = le_get_nxbuffer();
-
[…]

### 2014-08-08 branches/d2009-10-1/src/phi_fluctuations.h

@@ -1,21 +1,0 @@
-/*****************************************************************************
- *
- *  phi_fluctuations.h
- *
- *  $Id: phi_fluctuations.h,v 1.1.2.1 2010-04-05 03:42:18 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef PHI_FLUCTUATIONS_H
-#define PHI_FLUCTUATIONS_H
-
-void phi_fluctuations_random_flux(double * fluxw, double * fluxe,
-				  double * fluxy, double * fluxz);
-
-#endif

### 2014-08-08 branches/d2009-10-1/src/phi_force.h

@@ -1,21 +1,0 @@
-/*****************************************************************************
- *
- *  phi_force.h
- *
- *  $Id: phi_force.h,v 1.3.4.1 2010-03-30 14:21:38 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2008 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef PHI_FORCE_H
-#define PHI_FORCE_H
-
-void phi_force_calculation(void);
-void phi_force_required_set(const int flag);
-
-#endif

### 2014-08-08 branches/d2009-10-1/src/phi_force_colloid.c

@@ -1,498 +1,0 @@
-/*****************************************************************************
- *
- *  phi_force_colloid.c
- *
- *  The case of force from the thermodynamic sector on both fluid and
- *  colloid via the divergence of the chemical stress.
- *
- *  In the absence of solid, the force on the fluid is related
- *  to the divergence of the chemical stress
- *
- *  F_a = - d_b P_ab
- *
- *  Note that the stress is potentially antisymmetric, so this
- *  really is F_a = -d_b P_ab --- not F_a = -d_b P_ba.
- *
- *  The divergence is discretised as, e.g., in the x-direction,
- *  the difference between the interfacial values
- *
- *  d_x P_ab ~= P_ab(x+1/2) - P_ab(x-1/2)
- *
- *  and the interfacial values are based on linear interpolation
- *  P_ab(x+1/2) = 0.5 (P_ab(x) + P_ab(x+1))
- *
- *  etc (and likewise in the other directions).
- *
- *  The stress must be integrated over the colloid surface and the
- *  result added to the net force on a given colloid. Here the linear
- *  interpolation of the stress to the solid/fluid interface is again
- *  used. There are two options:
- *
- *  1) If the interior stress is available, use the value;
- *  2) If no interior stress is available, interpolate using zero.
- *
- *  The procedure ensures total momentum is conserved, ie., that
- *  leaving the fluid enters the colloid and vice versa.
- *
- *  $Id: phi_force_colloid.c,v 1.1.2.6 2010-09-30 18:02:35 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <stdlib.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "lattice.h"
[…]

### 2014-08-08 branches/d2009-10-1/src/symmetric.h

@@ -1,29 +1,0 @@
-/****************************************************************************
- *
- *  symmetric.h
- *
- *  $Id: symmetric.h,v 1.1.2.3 2010-05-12 17:08:16 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group
- *  and Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- ****************************************************************************/
-
-#ifndef SYMMETRIC_H
-#define SYMMETRIC_H
-
-#include "free_energy.h"
-
-void   symmetric_free_energy_parameters_set(double a, double b, double kappa);
-double symmetric_interfacial_tension(void);
-double symmetric_interfacial_width(void);
-double symmetric_free_energy_density(const int index);
-double symmetric_chemical_potential(const int index, const int nop);
-double symmetric_isotropic_pressure(const int index);
-void   symmetric_chemical_stress(const int index, double s[3][3]);
-
-#endif
-

### 2014-08-08 branches/d2009-10/src/ewald.h

@@ -1,22 +1,0 @@
-/*****************************************************************************
- *
- *  ewald.h
- *
- *  $Id: ewald.h,v 1.2 2007-12-05 17:56:12 kevin Exp $
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *
- *****************************************************************************/
-
-#ifndef EWALD_H_
-#define EWALD_H_
-
-void ewald_init(double, double);
-void ewald_sum(void);
-double ewald_real_space_energy(double [], double [], double []);
-double ewald_fourier_space_energy(void);
-double ewald_self_energy(void);
-
-void   ewald_test(void);
-void   ewald_total_energy(double *, double *, double *);
-#endif

### 2014-08-08 branches/d2009-10/src/interaction.c

@@ -1,1003 +1,0 @@
-/*****************************************************************************
- *
- *  interaction.c
- *
- *  Colloid potentials and colloid-colloid interactions.
- *
- *  Refactoring is in progress.
- *
- *  $Id: interaction.c,v 1.18 2009-10-30 18:05:03 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2008 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <string.h>
-#include <float.h>
-#include <math.h>
-
-#include "pe.h"
-#include "timer.h"
-#include "coords.h"
-#include "ran.h"
-#include "runtime.h"
-#include "free_energy.h"
-
-#include "active.h"
-#include "build.h"
-#include "physics.h"
-#include "potential.h"
-
-#include "colloids.h"
-#include "interaction.h"
-#include "communicate.h"
-#include "model.h"
-#include "site_map.h"
-#include "collision.h"
-#include "cio.h"
-#include "control.h"
-#include "subgrid.h"
-
-#include "ccomms.h"
-#include "ewald.h"
-
-extern int input_format;
-extern int output_format;
-
-static void    COLL_overlap(Colloid *, Colloid *);
-static void    COLL_set_fluid_gravity(void);
-static FVector COLL_lubrication(Colloid *, Colloid *, FVector, double);
-static void    COLL_init_colloids_test(void);
-static void    COLL_test_output(void);
-static void    coll_position_update(void);
-static double  coll_max_speed(void);
-
-struct lubrication_struct {
-  int corrections_on;
-  double cutoff_norm;  /* Normal */
-  double cutoff_tang;  /* Tangential */
-} lubrication;
-
-static double epotential_;
-
-/*****************************************************************************
- *
- *  COLL_update
- *
- *  The colloid positions have been updated. The following are required:
- *    (1) deal with processor/periodic boundaries
- *    (2) update the cell list if necessary and sort
[…]

### 2014-08-08 branches/d2009-10/src/model.h

@@ -1,61 +1,0 @@
-/*****************************************************************************
- *
- *  model.h
- *
- *  $Id: model.h,v 1.14 2009-08-20 16:26:09 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  (c) 2008 The University of Edinburgh
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *
- *****************************************************************************/
-
-#ifndef _MODEL_H
-#define _MODEL_H
-
-#include "d3q15.h"
-#include "d3q19.h"
-
-#if !defined (_D3Q15_) && !defined (_D3Q19_)
-#error "You must define either -D_D3Q15_ or -D_D3Q19_ in the Makeifle" 
-#endif
-
-/* Always three dimensions at the moment */
-enum {ND = 3};
-enum {NHYDRO = 10};
-
-/* Identify the hydrodynamic modes */
-enum {MRHO = 0,
-      MRUX = 1, MRUY = 2, MRUZ = 3,
-      MSXX = 4, MSXY = 5, MSXZ = 6, MSYY = 7, MSYZ = 8, MSZZ = 9};
-
-typedef struct {
-  double f[NVEL], g[NVEL];
-} Site;
-
-extern const double cs2;
-extern const double rcs2;
-extern struct io_info_t * io_info_distribution_;
-
-void   init_site(void);
-void   finish_site(void);
-void   halo_site(void);
-
-double get_f_at_site(const int, const int);
-double get_g_at_site(const int, const int);
-double get_rho_at_site(const int);
-double get_phi_at_site(const int);
-void   set_rho(const double, const int);
-void   set_phi(const double, const int);
-void   set_f_at_site(const int, const int, const double);
-void   set_g_at_site(const int, const int, const double);
-void   set_rho_u_at_site(const double, const double [], const int);
-void   get_momentum_at_site(const int, double[ND]);
-
-void distribution_get_stress_at_site(int index, double s[ND][ND]);
-void distribution_halo_set_complete(void);
-void distribution_halo_set_reduced(void);
-
-#endif

### 2014-08-08 branches/d2009-10/src/site_map.h

@@ -1,33 +1,0 @@
-/*****************************************************************************
- *
- *  site_map.h
- *
- *  $Id: site_map.h,v 1.2 2008-08-24 16:58:10 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2007 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef SITEMAP_H
-#define SITEMAP_H
-
-void site_map_init(void);
-void site_map_finish(void);
-void site_map_set_all(char);
-void site_map_halo(void);
-double site_map_volume(char);
-double site_map_get_C(int);
-double site_map_get_H(int);
-
-char site_map_get_status(int, int, int);
-char site_map_get_status_index(int);
-void site_map_set_status(int, int, int, char);
-void site_map_io_status_with_h(void);
-
-enum lattice_type { FLUID, BOUNDARY, SOLID, COLLOID };
-extern struct io_info_t * io_info_site_map;
-#endif

### 2014-08-08 branches/dev_src/src/coords.c

@@ -1,348 +1,0 @@
-/*****************************************************************************
- *
- *  coords.c
- *
- *  The physical coordinate system and the MPI Cartesian Communicator.
- *
- *  $Id: coords.c,v 1.2 2006-10-12 14:09:18 kevin Exp $
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *
- *****************************************************************************/
-
-#include <assert.h>
-
-#include "pe.h"
-#include "runtime.h"
-#include "coords.h"
-
-static int n_total[3] = {64, 64, 64};
-static int n_local[3];
-static int n_offset[3];
-
-static int periodic[3]                   = {1, 1, 1};
-static int pe_cartesian_rank             = 0;
-static int pe_cartesian_size[3]          = {1, 1, 1};
-static int pe_cartesian_coordinates[3]   = {0, 0, 0};
-static int pe_cartesian_neighbours[2][3] = {{0, 0, 0}, {0, 0, 0}};
-
-static double length[3] = {64.0, 64.0, 64.0};
-static double lmin[3] = {0.5, 0.5, 0.5};
-
-static void cart_init(void);
-static void default_decomposition(void);
-static int  is_ok_decomposition(void);
-
-#ifdef _MPI_
-static MPI_Comm cartesian_communicator;
-#endif
-
-/*****************************************************************************
- *
- *  coords_init
- *
- *  Set the physical lattice size, and periodic flags (always periodic
- *  by default.
- *
- *****************************************************************************/
-
-void coords_init() {
-
-  int n;
-
-  /* Look for "size" in the user input and set the lattice size. */
-
-  n = RUN_get_int_parameter_vector("size", n_total);
-
-  info((n == 0) ? "[Default] " : "[User   ] "); 
-  info("Lattice size is (%d, %d, %d)\n", n_total[X], n_total[Y], n_total[Z]);
-
-  for (n = 0; n < 3; n++) {
-    length[n] = (double) n_total[n];
-  }
-
-  /* Look for the "periodicity" in the user input. (This is
-   * not reported at the moment.) */
-
-  n = RUN_get_int_parameter_vector("periodicity", periodic);
-
-  info((n == 0) ? "[Default] " : "[User   ] "); 
-  info("periodic boundaries set to (%d, %d, %d)\n", periodic[X], periodic[Y],
-       periodic[Z]);
-
-  cart_init();
-
-  return;
-}
-
-/*****************************************************************************
- *
[…]

### 2014-08-08 branches/dev_src/src/ewald.h

@@ -1,22 +1,0 @@
-/*****************************************************************************
- *
- *  ewald.h
- *
- *  $Id: ewald.h,v 1.1.2.2 2007-05-02 13:42:41 kevin Exp $
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *
- *****************************************************************************/
-
-#ifndef EWALD_H_
-#define EWALD_H_
-
-void ewald_init(double, double);
-void ewald_sum(void);
-double ewald_real_space_energy(double [], double [], double []);
-double ewald_fourier_space_energy(void);
-double ewald_self_energy(void);
-
-void   ewald_test(void);
-void   ewald_total_energy(double *, double *, double *);
-#endif

### 2014-08-08 branches/dev_src/src/main.c

@@ -1,205 +1,0 @@
-/*****************************************************************************
- *
- *  Ludwig
- *
- *  A lattice Boltzmann code for complex fluids.
- *
- *****************************************************************************/
-
-#include <stdio.h>
-
-#include "pe.h"
-#include "runtime.h"
-#include "ran.h"
-#include "timer.h"
-#include "coords.h"
-#include "control.h"
-#include "free_energy.h"
-#include "model.h"
-#include "bbl.h"
-#include "subgrid.h"
-
-#include "colloids.h"
-#include "collision.h"
-#include "test.h"
-#include "wall.h"
-#include "communicate.h"
-#include "leesedwards.h"
-#include "interaction.h"
-#include "propagation.h"
-
-#include "lattice.h"
-#include "cio.h"
-#include "regsteer.h"
-
-static char rcsid[] = "$Id: main.c,v 1.10.2.2 2007-10-03 15:32:44 kevin Exp $";
-
-int main( int argc, char **argv )
-{
-  char    filename[FILENAME_MAX];
-  int     step;
-
-  /* Initialise the following:
-   *    - RealityGrid steering (if required)
-   *    - communications (MPI)
-   *    - random number generation (serial RNG and parallel fluctuations)
-   *    - model fields
-   *    - simple walls 
-   *    - colloidal particles */
-
-  REGS_init();
-
-  pe_init(argc, argv);
-  if (argc > 1) {
-    RUN_read_input_file(argv[1]);
-  }
-  else {
-    RUN_read_input_file("input");
-  }
-  coords_init();
-  init_control();
-
-  COM_init();
-
-  TIMER_init();
-  TIMER_start(TIMER_TOTAL);
-
-  ran_init();
-  RAND_init_fluctuations();
-  MODEL_init();
-  LE_init();
-  wall_init();
-  COLL_init();
-
-  init_free_energy();
-
-  /* Report initial statistics */
-
-  TEST_statistics();
-  TEST_momentum();
-
-  /* Main time stepping loop */
-
-  while (next_step()) {
-
-    TIMER_start(TIMER_STEPS);
-    step = get_step();
-
-    latt_zero_force();
-    COLL_update();
-    wall_update();
-
-    /* Collision stage */
-    collide();
-
-    LE_apply_LEBC();
-    halo_site();
-
-    /* Colloid bounce-back applied between collision and
-     * propagation steps. */
-
-#ifdef _SUBGRID_
[…]

### 2014-08-08 branches/dev_src/src/propagation.c

@@ -1,387 +1,0 @@
-/*****************************************************************************
- *
- *  propagation.c
- *
- *  Propagation schemes for the different models.
- *
- *  $Id: propagation.c,v 1.2.2.1 2007-04-30 15:05:03 kevin Exp $
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *
- *****************************************************************************/
-
-#include "pe.h"
-#include "timer.h"
-#include "coords.h"
-#include "model.h"
-
-extern Site * site;
-
-#ifdef _D3Q19_
-static void d3q19_propagate_single(void);
-static void d3q19_propagate_binary(void);
-#endif
-
-#ifdef _D3Q15_
-static void d3q15_propagate_single(void);
-static void d3q15_propagate_binary(void);
-#endif
-
-/*****************************************************************************
- *
- *  propagation
- *
- *  Driver routine for the propagation stage.
- *
- *****************************************************************************/
-
-void propagation() {
-
-#ifdef _D3Q19_
-
-  TIMER_start(TIMER_PROPAGATE);
-
-#ifdef _SINGLE_FLUID_
-  d3q19_propagate_single();
-#else
-  d3q19_propagate_binary();
-#endif
-
-  TIMER_stop(TIMER_PROPAGATE);
-
-#endif
-
-#ifdef _D3Q15_
-
-  TIMER_start(TIMER_PROPAGATE);
-
-#ifdef _SINGLE_FLUID_
-  d3q15_propagate_single();
-#else
-  d3q15_propagate_binary();
-#endif
-
-  TIMER_stop(TIMER_PROPAGATE);
-
-#endif
-
-  return;
-}
-
-#ifdef _D3Q15_
-
-/*****************************************************************************
- *
- *  d3q15_propagate_single
- *
- *****************************************************************************/
-
-static void d3q15_propagate_single() {
-
-  int i, j, k, ijk;
-  int xfac, yfac;
-  int N[3];
-
-  get_N_local(N);
-
-  yfac = (N[Z]+2);
-  xfac = (N[Y]+2)*yfac;
-
[…]

### 2014-08-08 branches/kevin-berk/brownian.c

@@ -1,511 +1,0 @@
-/*****************************************************************************
- *
- *  brownian.c
- *
- *  $Id: brownian.c,v 1.4 2010-10-15 12:40:02 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2007 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <math.h>
-
-#include "pe.h"
-#include "ran.h"
-#include "coords.h"
-#include "colloids.h"
-#include "colloids_halo.h"
-#include "physics.h"
-#include "util.h"
-
-void brownian_step_ermak_buckholz(void);
-void brownian_step_no_inertia(void);
-void brownian_step_no_inertia_test(void);
-void brownian_set_random(void);
-
-static double dt_ = 1.0;  /* Time step; fixed here to be 1 LB step */
-#define NTIME 16
-
-/*****************************************************************************
- *
- *  do_brownian_dynamics
- *
- *  Perform a number of Brownian Dynamics steps.
- *
- *****************************************************************************/
-
-void do_brownian_dynamics() {
-
-  int     ia;
-  int     ic, jc, kc;
-  colloid_t * p_colloid;
-  int i, n, nt, nmax = 100000;
-  double u[3][NTIME];
-  double r[3][NTIME];
-  double s[3][NTIME];
-  double correlator_uu[NTIME];
-  double correlator_ur[NTIME];
-  double correlator_rr[NTIME];
-  double correlator_ss[NTIME];
-  double mass, scaleu, scalex, beta;
-  double dr[3];
-
-  /* Brownian dynamics requires an update. */
-  assert(0);
-
-  brownian_step_no_inertia_test();
-  if (1) return;
-
-  mass  = (4.0/3.0)*pi_*pow(2.3, 3);
-  beta  = 6.0*pi_*get_eta_shear()*2.3/mass;
[…]

### 2014-08-08 branches/kevin-berk/d3q19.h

@@ -1,44 +1,0 @@
-/*****************************************************************************
- *
- *  d3q19.h
- *
- *  D3Q19 definitions.
- *
- *  $Id: d3q19.h,v 1.9 2010-10-15 12:40:02 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2008 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifdef _D3Q19_
-
-enum {NDIM = 3};
-enum {NVEL = 19};
-enum {CVXBLOCK = 1};
-enum {CVYBLOCK = 3};
-enum {CVZBLOCK = 5};
-
-extern const    int cv[NVEL][3];
-extern const double wv[NVEL];
-extern const double q_[NVEL][3][3];
-extern const double norm_[NVEL];
-extern const double ma_[NVEL][NVEL];
-extern const double mi_[NVEL][NVEL];
-
-extern const int xblocklen_cv[CVXBLOCK];
-extern const int xdisp_fwd_cv[CVXBLOCK];
-extern const int xdisp_bwd_cv[CVXBLOCK];
-
-extern const int yblocklen_cv[CVYBLOCK];
-extern const int ydisp_fwd_cv[CVYBLOCK];
-extern const int ydisp_bwd_cv[CVYBLOCK];
-
-extern const int zblocklen_cv[CVZBLOCK];
-extern const int zdisp_fwd_cv[CVZBLOCK];
-extern const int zdisp_bwd_cv[CVZBLOCK];
-
-#endif

### 2014-08-08 branches/kevin-berk/stats_colloid.h

@@ -1,21 +1,0 @@
-/*****************************************************************************
- *
- *  stats_colloid.h
- *
- *  $Id: stats_colloid.h,v 1.2 2010-10-15 12:40:03 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef STATS_COLLOID_H
-#define STATS_COLLOID_H
-
-void stats_colloid_momentum(double g[3]);
-void stats_colloid_velocity_minmax(void);
-
-#endif

### 2014-08-08 branches/kevin-field/src/advection_bcs.c

@@ -1,206 +1,0 @@
-/****************************************************************************
- *
- *  advection_bcs.c
- *
- *  Advection boundary conditions.
- *
- *  $Id: advection_bcs.c,v 1.2 2010-10-15 12:40:02 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2009 The University of Edinburgh
- *
- ****************************************************************************/
-
-#include <assert.h>
-
-#include "pe.h"
-#include "wall.h"
-#include "coords.h"
-#include "site_map.h"
-
-#ifdef OLD_PHI
-#include "phi.h"
-#else
-#include "field.h"
-#endif
-
-/*****************************************************************************
- *
- *  advection_bcs_no_normal_fluxes
- *
- *  Set normal fluxes at solid fluid interfaces to zero.
- *
- *****************************************************************************/
-
-void advection_bcs_no_normal_flux(int nf, double * fluxe, double * fluxw,
-				  double * fluxy, double * fluxz) {
-
-  int nlocal[3];
-  int ic, jc, kc, index, n;
-
-  double mask, maskw, maske, masky, maskz;
-
-  assert(fluxe);
-  assert(fluxw);
-  assert(fluxy);
-  assert(fluxz);
-
-  coords_nlocal(nlocal);
-
-  for (ic = 1; ic <= nlocal[X]; ic++) {
-    for (jc = 0; jc <= nlocal[Y]; jc++) {
-      for (kc = 0; kc <= nlocal[Z]; kc++) {
-
-	index = coords_index(ic, jc, kc);
-
-	mask  = (site_map_get_status_index(index)  == FLUID);
-	maske = (site_map_get_status(ic+1, jc, kc) == FLUID);
-	maskw = (site_map_get_status(ic-1, jc, kc) == FLUID);
-	masky = (site_map_get_status(ic, jc+1, kc) == FLUID);
-	maskz = (site_map_get_status(ic, jc, kc+1) == FLUID);
-
-	for (n = 0;  n < nf; n++) {
-	  fluxw[nf*index + n] *= mask*maskw;
-	  fluxe[nf*index + n] *= mask*maske;
-	  fluxy[nf*index + n] *= mask*masky;
[…]

### 2014-08-08 branches/kevin-field/src/blue_phase.h

@@ -1,71 +1,0 @@
-/*****************************************************************************
- *
- *  blue_phase.h
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef BLUEPHASE_H
-#define BLUEPHASE_H
-
-/* 'Extension' of free energy (pending free_energy_tensor.h) */
-
-#include "free_energy.h"
-
-void   blue_phase_set_free_energy_parameters(double, double, double, double);
-void   blue_phase_set_xi(double);
-void   blue_phase_set_zeta(double);
-void   blue_phase_set_gamma(double);
-
-double blue_phase_free_energy_density(const int);
-double blue_phase_compute_fed(double q[3][3], double dq[3][3][3]);
-
-void   blue_phase_molecular_field(const int, double h[3][3]);
-void   blue_phase_compute_h(double q[3][3], double dq[3][3][3],
-			    double dsq[3][3], double h[3][3]);
-void   blue_phase_compute_stress(double q[3][3], double dq[3][3][3],
-				 double h[3][3], double sth[3][3]);
-
-void   blue_phase_chemical_stress(const int, double sth[3][3]);
-void   blue_phase_redshift_set(const double redshift);
-void   blue_phase_electric_field_set(const double e[3]);
-void   blue_phase_dielectric_anisotropy_set(double e);
-double blue_phase_amplitude_compute(void);
-double blue_phase_get_xi(void);
-double blue_phase_get_zeta(void);
-double blue_phase_get_gamma(void);
-double blue_phase_chirality(void);
-double blue_phase_reduced_temperature(void);
-double blue_phase_redshift(void);
-double blue_phase_q0(void);
-double blue_phase_kappa0(void);
-double blue_phase_kappa1(void);
-double blue_phase_redshift(void);
-double blue_phase_rredshift(void);
-double blue_phase_a0(void);
-double blue_phase_gamma(void);
-double blue_phase_dimensionless_field_strength(void);
-
-void blue_phase_redshift_update_set(int onoff);
-void blue_phase_redshift_compute(void);
-
-void blue_phase_q_uniaxial(double amplitude, const double n[3], double q[3][3]);
-
[…]

### 2014-08-08 branches/kevin-field/src/collision.c

@@ -1,998 +1,0 @@
-/*****************************************************************************
- *
- *  collision.c
- *
- *  Collision stage routines and associated data.
- *
- *  Isothermal fluctuations following Adhikari et al., Europhys. Lett
- *  (2005).
- *
- *  The relaxation times can be set to give either 'm10', BGK, or
- *  'two-relaxation' time (TRT) models.
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2011 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <math.h>
-
-#include "pe.h"
-#include "util.h"
-#include "coords.h"
-#include "physics.h"
-#include "model.h"
-#include "site_map.h"
-#include "collision.h"
-#include "fluctuations.h"
-
-#include "phi.h"
-#include "free_energy.h"
-#include "phi_cahn_hilliard.h"
-
-#include "control.h"
-#include "propagation_ode.h"
-
-static int nmodes_ = NVEL;               /* Modes to use in collsion stage */
-static int nrelax_ = RELAXATION_M10;     /* Default is m10 */
-static int isothermal_fluctuations_ = 0; /* Flag for noise. */
-
-static double rtau_shear;       /* Inverse relaxation time for shear modes */
-static double rtau_bulk;        /* Inverse relaxation time for bulk modes */
-static double var_shear;        /* Variance for shear mode fluctuations */
-static double var_bulk;         /* Variance for bulk mode fluctuations */
-static double rtau_[NVEL];      /* Inverse relaxation times */
-static double noise_var[NVEL];  /* Noise variances */
-
-static fluctuations_t * fl_;
-
-static int collision_multirelaxation(hydro_t * hydro);
-static int collision_binary_lb(hydro_t * hydro);
-
-static void fluctuations_off(double shat[3][3], double ghat[NVEL]);
-       void collision_fluctuations(int index, double shat[3][3],
-				   double ghat[NVEL]);
-
[…]

### 2014-08-08 branches/kevin-field/src/gradient_3d_27pt_fluid.c

@@ -1,455 +1,0 @@
-/*****************************************************************************
- *
- *  gradient_3d_27pt_fluid.c
- *
- *  Gradient operations for equally-weighted 27-point stencil
- *  in three dimensions.
- *
- *        (ic-1, jc+1, kc) (ic, jc+1, kc) (ic+1, jc+1, kc)
- *        (ic-1, jc,   kc) (ic, jc,   kc) (ic+1, jc,   kc)
- *        (ic-1, jc-1, kc) (ic, jc-1, kc) (ic+1, jc,   kc)
- *
- *  ...and so in z-direction
- *
- *  d_x phi = [phi(ic+1, j, k) - phi(ic-1, j, k)] / 2*9
- *  for all j = jc-1,jc,jc+1, and k = kc-1,kc,kc+1
- * 
- *  d_y phi = [phi(i ,jc+1,k ) - phi(i ,jc-1,k )] / 2*9
- *  for all i = ic-1,ic,ic+1 and k = kc-1,kc,kc+1
- *
- *  d_z phi = [phi(i ,j ,kc+1) - phi(i ,j ,kc-1)] / 2*9
- *  for all i = ic-1,ic,ic+1 and j = jc-1,jc,jc+1
- *
- *  nabla^2 phi = phi(ic+1,jc,kc) + phi(ic-1,jc,kc)
- *              + phi(ic,jc+1,kc) + phi(ic,jc-1,kc)
- *              + phi(ic,jc,kc+1) + phi(ic,jc,kc-1)
- *              etc
- *              - 26 phi(ic,jc,kc)
- *
- *  Corrections for Lees-Edwards planes are included.
- *
- *  This scheme was fist instituted for the work of Kendon et al.
- *  JFM (2001).
- *
- *  $Id: gradient_3d_27pt_fluid.c,v 1.2 2010-10-15 12:40:03 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdlib.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "leesedwards.h"
-#include "wall.h"
-#include "gradient.h"
-#include "gradient_3d_27pt_fluid.h"
-
-static void gradient_3d_27pt_fluid_operator(const int nop,
[…]

### 2014-08-08 branches/kevin-field/src/nernst_planck.h

@@ -1,23 +1,0 @@
-/*****************************************************************************
- *
- *  nernst_planck.h
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2012 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef NERNST_PLANCK_H
-#define NERNST_PLANCK_H
-
-#include "psi.h"
-#include "hydro.h"
-
-int nernst_planck_driver(psi_t * psi, hydro_t * hydro);
-
-#endif

### 2014-08-08 branches/kevin-field/src/phi_lb_coupler.h

@@ -1,31 +1,0 @@
-/*****************************************************************************
- *
- *  phi_lb_coupler.h
- *
- *  $Id: phi_lb_coupler.h,v 1.4 2010-10-15 12:40:03 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) The University of Edinburgh (2009)
- *
- *****************************************************************************/
-
-#ifndef PHI_LB_COUPLER_H
-#define PHI_LB_COUPLER_H
-
-void phi_compute_phi_site(void);
-void phi_set_mean_phi(double phi0);
-void phi_init_drop(double r, double xi0);
-void phi_lb_init_drop(double r, double xi0);
-void phi_lb_coupler_phi_set(const int index, const double phi);
-
-#ifdef OLD_PHI
-#else
-#include "field.h"
-int phi_lb_to_field(field_t * phi);
-int phi_lb_from_field(field_t * phi);
-#endif
-
-#endif

### 2014-08-08 branches/kevin-field/src/site_map.c

@@ -1,607 +1,0 @@
-/*****************************************************************************
- *
- *  site_map.c
- *
- *  Keeps track of the solid/fluid status of the lattice.
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2007 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <stdio.h>
-#include <stdlib.h>
-
-#include "pe.h"
-#include "coords.h"
-#include "io_harness.h"
-#include "site_map.h"
-
-io_info_t * io_info_site_map;
-
-static void site_map_init_mpi(void);
-static void site_map_init_io(void);
-static int site_map_read(FILE *, const int, const int, const int);
-static int site_map_write(FILE *, const int, const int, const int);
-static int site_map_read_ascii(FILE *, const int, const int, const int);
-static int site_map_write_ascii(FILE *, const int, const int, const int);
-
-struct site_info_t {
-  char status;       /* Solid / fluid etc status flag */
-  double h;          /* Wetting free energy parameter H */
-                     /* Wetting free energy parameter C = 0 always */
-};
-
-static struct site_info_t * site_map;
-static int initialised_ = 0;
-static int io_type_ = 0;                 /* 0 = status only; 1 = status + h */
-static MPI_Datatype mpi_site_t_;
-static MPI_Datatype mpi_xy_t_;
-static MPI_Datatype mpi_xz_t_;
-static MPI_Datatype mpi_yz_t_;
-
-/*****************************************************************************
- *
- *  site_map_init
- *
- *  Allocate, and read site information from file if required.
- *
- *****************************************************************************/
-
-void site_map_init() {
-
-  int nsites;
-
-  nsites = coords_nsites();
-
-  site_map = (struct site_info_t *) malloc(nsites*sizeof(struct site_info_t));
-  if (site_map == NULL) fatal("malloc(site_map) failed\n");
-
-  site_map_init_io();
-  site_map_init_mpi();
-
-  initialised_ = 1;
-  site_map_set_all(FLUID);
-
-  return;
-}
-
[…]

### 2014-08-08 branches/kevin-field/src/site_map.h

@@ -1,39 +1,0 @@
-/*****************************************************************************
- *
- *  site_map.h
- *
- *  $Id: site_map.h,v 1.3 2010-10-15 12:40:03 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2007 The University of Edinburgh
- *
- *****************************************************************************/
-
-#ifndef SITEMAP_H
-#define SITEMAP_H
-
-#include "io_harness.h"
-
-void   site_map_init(void);
-void   site_map_finish(void);
-void   site_map_set_all(char);
-void   site_map_halo(void);
-double site_map_volume(char);
-double site_map_C(const int index);
-double site_map_H(const int index);
-void   site_map_C_set(const int index, const double c);
-void   site_map_H_set(const int index, const double h);
-void   site_map_set(const int index, const char status, const double c,
-		    const double h);
-
-char site_map_get_status(int, int, int);
-char site_map_get_status_index(int);
-void site_map_set_status(int, int, int, char);
-void site_map_io_status_with_h(void);
-
-enum lattice_type { FLUID, BOUNDARY, SOLID, COLLOID };
-extern io_info_t * io_info_site_map;
-#endif

### 2014-08-08 branches/kevin-field/src/stats_surfactant.c

@@ -1,103 +1,0 @@
-/*****************************************************************************
- *
- *  stats_surfactant.c
- *
- *  Some routines to perform analysis of the surfactant model.
- *
- *  $Id: stats_surfactant.c,v 1.2 2010-10-15 12:40:03 kevin Exp $
- *
- *  Edinburgh Soft Matter and Statistical Physics Group and
- *  Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2009 The University of Edinburgh
- *
- *****************************************************************************/
-
-#include <assert.h>
-#include <math.h>
-
-#include "pe.h"
-#include "coords.h"
-
-#ifdef OLD_PHI
-#include "phi.h"
-#else
-/* switch to field */
-#endif
-#include "control.h"
-#include "util.h"
-#include "surfactant.h"
-#include "stats_surfactant.h"
-
-/*****************************************************************************
- *
- *  stats_surfactant_1d
- *
- *  For a 1-d model in which we have a block initialisation of the
- *  composition, work out the profile of the free energy, and hence
- *  the current interfacial tension.
- *
- *****************************************************************************/
-
-void stats_surfactant_1d(void) {
-#ifdef OLD_PHI
-
-  int index;
-  int ic = 1, jc = 1, kc;
-  int nlocal[3];
-  double e, e0;
-  double psi_0, psi_b;
-  double sigma, sigma0;
-
-  /* This is not run in parallel, so assert it's serial.
-   * We also require surfactant */
-
-  assert(pe_size() == 1);
-  assert(phi_nop() == 2);
-
-  coords_nlocal(nlocal);
-
-  /* We assume z = 1 is a reasonable choice for the background
-   * free energy level, which we need to subtract to find the
-   * excess. */
-
-  kc = 1;
-  index = coords_index(ic, jc, kc);
-  e0 = surfactant_free_energy_density(index);
-  psi_b = phi_op_get_phi_site(index, 1);
-
-  /* To compute the surface tension, run through both interfaces
-   * and divide the final excess free energy by 2. We also record
-   * the maximum surfactant value (psi_0) along the way. */ 
-
-  sigma = 0.0;
-  psi_0 = 0.0;
-
-  for (kc = 1; kc <= nlocal[Z]; kc++) {
-
-    index = coords_index(ic, jc, kc);
-
[…]

### 2014-08-08 branches/kevin-field/src/symmetric.h

@@ -1,38 +1,0 @@
-/****************************************************************************
- *
- *  symmetric.h
- *
- *  $Id$
- *
- *  Edinburgh Soft Matter and Statistical Physics Group
- *  and Edinburgh Parallel Computing Centre
- *
- *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2010 The University of Edinburgh
- *
- ****************************************************************************/
-
-#ifndef SYMMETRIC_H
-#define SYMMETRIC_H
-
-#include "free_energy.h"
-
-void   symmetric_free_energy_parameters_set(double a, double b, double kappa);
-double symmetric_a(void);
-double symmetric_b(void);
-double symmetric_interfacial_tension(void);
-double symmetric_interfacial_width(void);
-double symmetric_free_energy_density(const int index);
-double symmetric_chemical_potential(const int index, const int nop);
-double symmetric_isotropic_pressure(const int index);
-void   symmetric_chemical_stress(const int index, double s[3][3]);
-
-#ifdef OLD_PHI
-#else
-#include "field.h"
-#include "field_grad.h"
-int symmetric_phi_set(field_t * phi, field_grad_t * dphi);
-#endif
-
-#endif
-

### 2014-08-08 branches/kevin-field/tests/trytests.sh

@@ -1,151 +1,0 @@
-##############################################################################
-#
-#  try.sh
-#
-#  This file runs the unit tests and the regression tests.
-#  It is intended for my nightly test, so may not be very
-#  portable.
-#
-#  $Id$
-#
-#  Edinburgh Soft Matter and Statistical Physics Group and
-#  Edinburgh Parallel Computing Centre
-#
-#  Kevin Stratford (kevin@epcc.ed.ac.uk)
-#  (c) 2010 The University of Edinburgh
-#
-##############################################################################
-#!/bin/bash
-
-# Test machine is currently martensite, where we need
-export PATH=$PATH:/opt/openmpi/bin
-
-# Serial unit tests
-
-cd ../mpi_s
-make clean
-make libc
-make testc
-
-cd ../src
-make clean
-make lib
-cd ../tests
-make clean
-make do_tests
-
-# Serial regressions
-
-cd ../src
-make serial
-
-echo Running autocorrelation test
-./Ludwig.exe ../tests/regression/test_auto1_input > /tmp/junk
-diff /tmp/junk ../tests/regression/test_auto1_srl_d3q19.ref
-
-echo Running LE1 test
-./Ludwig.exe ../tests/regression/test_le1_input > /tmp/junk
-diff /tmp/junk ../tests/regression/test_le1_srl_d3q19.ref
-
-echo Running LE2 test
-./Ludwig.exe ../tests/regression/test_le2_input > /tmp/junk
-diff /tmp/junk ../tests/regression/test_le2_srl_d3q19.ref1
-
-echo Running LE3 test
-./Ludwig.exe ../tests/regression/test_le3_input > /tmp/junk
-diff /tmp/junk ../tests/regression/test_le3_srl_d3q19.ref
-
-echo Running LE4 test
-./Ludwig.exe ../tests/regression/test_le4_input > /tmp/junk
-diff /tmp/junk ../tests/regression/test_le4_srl_d3q19.ref
-
-echo Running LE5 test
-./Ludwig.exe ../tests/regression/test_le5_input > /tmp/junk
-diff /tmp/junk ../tests/regression/test_le5_srl_d3q19.ref
-
-echo Running LE6 test
-./Ludwig.exe ../tests/regression/test_le6_input > /tmp/junk
[…]

### 2014-08-08 branches/porous-colloid/gradient_3d_7pt_solid.c

@@ -174,8 +174,8 @@ static void gradient_general6(const double * field, double * grad,
   double w2_coll;                           /* Second anchoring parameter */
   double w1_wall;
   double w2_wall;
-  double w1;
-  double w2;
+  double w1=0.0;
+  double w2=0.0;
   double q_0;                               /* Cholesteric pitch wavevector */
   double kappa0;                            /* Elastic constants */
   double kappa1;
@@ -965,7 +965,7 @@ static int gradient_no_iteration(const double * field, double * grad,
 	  }
 	  else if (nunknown == 2) {
 
-	    int ida, idb, na, nb;
+	    int idb, na, nb;
 	    double dq1[NOP];
 	    double bc1[6][NOP];
 	    double b2[2*6];
@@ -983,12 +983,11 @@ static int gradient_no_iteration(const double * field, double * grad,
 	    for (ia = 0; ia < nunknown; ia++) {
 
 	      na = normal[ia];    /* normal for unknown ia */
-	      ida = normal[ia]/2; /* coordinate direction for unknown ia */
 
 	      for (ib = 0; ib < nunknown; ib++) {
 
 		nb = normal[ib];
-		idb = normal[ib]/2;
+		idb = normal[ib]/2; /* coordinate direction of normal */
 
 		/* Compute 6x5 block (ia,ib) in system matrix */
 
@@ -1062,7 +1061,7 @@ static int gradient_no_iteration(const double * field, double * grad,
 	  }
 	  else if (nunknown == 3) {
 
-	    int ida, idb, na, nb;
+	    int idb, na, nb;
 	    double dq1[NOP];
 	    double bc1[6][NOP];
 	    double b3[3*6];
@@ -1076,12 +1075,11 @@ static int gradient_no_iteration(const double * field, double * grad,
 	    for (ia = 0; ia < nunknown; ia++) {
 
 	      na = normal[ia];    /* normal for unknown ia */
-	      ida = normal[ia]/2; /* coordinate direction for unknown ia */
 
 	      for (ib = 0; ib < nunknown; ib++) {
 
 		nb = normal[ib];
-		idb = normal[ib]/2;
+		idb = normal[ib]/2; /* coordinate direction of normal */
 
 		/* Compute 6x5 block (ia,ib) in system matrix */
 

### 2014-09-15 tags/2014-pre-release/htdocs/basicuser.tex
Tag for Sept 2014 trunk
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2463 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-05-10 htdocs/user.tex
@@
 %
-%  $Id$
+%  $Id: user.tex 1694 2012-06-28 09:44:12Z stratford $
 %
@@
 
-\section{Introduction}
+\section{Basic User Input}
 
-\subsection{Overview}
+\subsection{Parallel Performance}
 
-We aim to provide a robust and portable code, written in C, which
-can be used to perform serial and scalable parallel simulations of
-complex fluid systems based around hydrodynamics via the Lattice
-Boltzmann method. Complex fluids are introduced via a free energy,
-of which a number of different examples are available. The preferred
-method of dealing with the corresponding order parameter equations
-is now by using finite difference. However, for the case of a binary
-fluid, the `full LB' approach, using two distributions, is retained
-as an option for the time being.
+\subsubsection{General Comments}
 
-\subsection{Availability}
+The basic LB calculation should scale well in parallel, that is,
+the time taken for fixed problem size will decrease linearly
+with the number of MPI processes (strong scaling), or a larger
+problem can be run on a proportionally larger number processes
+in the same time (weak scaling). To ensure performance is retained
+in parallel, it is useful to understand some basic considerations.
+Some of these are discussed below.
 
-Under the auspices of the CCPs (Collaborative Computational Projects),
-and in particular CCP5, we have a CCPForge project account at
+\textit{Ludwig} has been run successfully on up to 131,072 MPI
+tasks with close to ideal scaling, and there is no problem in
+principle to run larger decompositions.
 
-\texttt{http://ccpforge.cse.rl.ac.uk/}
+\subsubsection{Things to Remember and Things to Avoid}
+\label{section-advice}
 
-This provides controlled access to known persons and uses version
-control via SVN, and includes bug
-tracking, and so on. The CCPForge SVN is not available to the public
-at the present time. You need to create an account at CCPForge, and
-let me or Juho know the username so you can be added to the access list.
+There are a number of simple issues to get right:
+\begin{itemize}
+\item
+Make sure the assertions are switched off (via preprocessor option
+\texttt{-DNDEBUG}) and appropriate optimisations
+for the current compiler are switched on for production runs (see Makefile).
+\item
+Do not use too small a local domain size. Typically, 16$^3$ to 32$^3$
+cubic local domains will give resaonble scaling. Anything smaller may
+be inefficient. The best size may depend on the exact nature of the
+calculation and the hardware. Note that this limitation will always
+ultimately limit strong scaling.
+\item
+Don't have diagnsotic output (freq\_statistics) more often than is
+necessary. This output requires a global communication which is slow.
+\item
+Use BINARY output options; ASCII output is slower, and produces larger
+files.
+\item
+Do not use serial I/O. This will have an increasingly high overhead as
+the number of processors is increases (it's probably reasonable to
+128--256 MPI tasks). This is because each MPI tasks writes its data,
+in turn, to a single file; this in inherently serial. To avoid this,
+you need to specify the I/O decomposition for the various output
+quantities of interest. This splits the system into regular blocks
+(in a similar approach to the MPI decomposition), the data for which
+are written to separate files in parallel. For example, if the MPI
+decomposition is 8\_8\_8 (512 MPI tasks), then a reasonable I/O
+decomposition may be 2\_2\_2 or 4\_4\_4. The only way to find out
+what works best is to try some tests for the problem size you want to
+address.
+\item
+Don't use the Ewald sum! This simply does not scale.
+An alternative algorithm is required.
+\end{itemize}
 
-To download the entire code, you need
-\begin{verbatim}
-svn checkout --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig
-\end{verbatim}
-where \texttt{<user>} is replaced by your CCPForge user name.
-Note that this will download all the existing tags and branches
-into a new directory named \texttt{ludwig}, as well as the main code!
-On success, you should see three directories
-\begin{verbatim}
-$ ls ludwig
-branches  tags      trunk
-\end{verbatim}
-The main code is in the \texttt{trunk} subdirectory. This documentation is
-found in the file \texttt{trunk/htdocs/ludwig.tex}, unit tests are in
-the directory \texttt{trunk/tests}, and the utilities are in
-\texttt{trunk/util}, and so on.
-
-If you do not want all the tags and branches (which take up a certain amount
-of space), you can specify that you
-want only the trunk, e.g.,
-\begin{verbatim}
-svn co --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk
-\end{verbatim}
-in which case you will get a new directory called \texttt{trunk} containing
-the source files. All the instructions below refer to paths relative to
-this trunk directory.
-
-If you have problems, it may be worth while looking at the issue
-tracker on the project web page. It would be useful if all issues
-could be reported via this mechanism. If you e-mail me, I will just
-insert your e-mail into the issue tracker!
-
-\subsection{Compilation}
-
-Locate your local C compiler. The following example uses \texttt{gcc}
-in serial, and \texttt{mpicc} in parallel. For MPI in particular,
-local details may vary.
-
-The other thing required by the Makefile is the utility
-\texttt{svnversion}, which allows the compile-time SVN version number to
-be included in the executable (and hence the output of the program). This
-is useful to remember which version you were using when looking at
-results at a later date. If
-\texttt{svnversion} is not available (and it
-should be if you have downloaded the code via SVN), you may get an
-error message, but the code should still compile.
-
-\subsubsection{Serial}
-
-First, compile the MPI stub library in the \texttt{mpi\_s}
-directory. Do this by editing the Makefile, and checking the compiler
-is appropriate. You should be able to build the library and run the
-tests using, e.g.,:
-
-\begin{verbatim}
-$ make libc
-gcc -Wall -I.  -c mpi_serial.c
-ar -cru libmpi.a mpi_serial.o
-$ make testc
-gcc -Wall   mpi_tests.c -L. -lmpi
-./a.out
-Running mpi_s tests...
-Finished mpi_s tests ok.
-\end{verbatim}
-
-Now compile the main code in the \texttt{src} directory. Again
-edit the Makefile to check the compiler. At this stage, you also
-need to choose the LB model using exactly one of the preprocessor
-options for D2Q9 (fluid only problems), D3Q15, or D3Q19. Now type,
-e.g., using D3Q19:
-
-\begin{verbatim}
-$ make serial
-gcc  -D SVN_REVISION='"'`svnversion`'"' -c svn.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d2q9.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d3q15.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c d3q19.c
-gcc  -D_D3Q19_ -O3 -Wall -I. -I../mpi_s -c model.c
-...
-\end{verbatim}
-
-This should provide an executable \texttt{Ludwig.exe} which is linked
-against the MPI stub library.
-
-\subsubsection{Parallel}
-
-Here, you do not need to compile the stub library. Simply compile
-the main code with, e.g.,:
-\begin{verbatim}
-$ make mpi
-mpicc -D_D3Q19_ -O3 -Wall -I. -c d2q9.c
-mpicc -D_D3Q19_ -O3 -Wall -I. -c d3q15.c
-mpicc -D_D3Q19_ -O3 -Wall -I. -c d3q19.c
-mpicc -D_D3Q19_ -O3 -Wall -I. -c model.c
-...
-\end{verbatim}
-
-Invoke the executable with the MPI launcher in the usual way for your
-system, e.g., \texttt{mpirun -np 8 ./Ludwig.exe} for 8 processors,
-and so on.
-
-\subsubsection{ANSI C}
-
-The code is ANSI C (1989), with the single exception of calls to the
-library function \texttt{erfc()} in the Ewald sum code. This
-can sometimes cause a problem, which can usually be solved with
-a compiler-specific work-around.
-
-\subsubsection{C assertions}
-
-The code makes quite a lot of use of standard C assertions, which
-are useful to prevent errors. They do result in a considerably
-slower execution in some instances, so production runs should
-switch off the assertions with the standard \texttt{NDEBUG}
-preprocessor flag.
-
-\section{User Input}
+\subsection{Setting up the input}
 
@@
 \end{verbatim}
+
 A reference input file \texttt{input.ref} is provided as a template.
@@
 (or commented out) in the input file.
-
-\subsection{Setting up the input}
 
@@
 
-\subsection{Symmetric binary fluid}
-
-\inputkey{free\_energy symmetric}
-
-The following applies for the binary fluid problem with
-compositional order parameter $\phi$ and free energy
-(excluding the term in the density $\rho$):
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
-\end{equation}
-
-This is described in some detail
-by Kendon et al \cite{viv}. The first two terms represent the bulk
-contribution, whereas the term in $\kappa$ penalises curvature in
-the interface.
-
-\inputkey{A} The parameter $A$. Note $A < 0$.
-
-\inputkey{B} The parameter $B$. Note $B = -A$ for common usage,
-although this is not enforced.
-
-\inputkey{K} The parameter $\kappa$, which is positive.
-
-The order parameter evolution is determined by a Cahn-Hilliard equation
-with mobility set by
-
-\inputkey{mobility} Sets mobility $M$ (uniform in space).
-
-The following parameters control the initialisation of the order
-parameter.
-
-\inputkey{phi\_initialisation}
-
-Determines how the compositional order parameter is initialised
-at the start of the run. If set to \texttt{spinodal} the value
-is set to $\phi_0$ as set above plus or minus a random noise, the
-magnitude of which is set by the value of the \texttt{noise} key.
-If set to \texttt{block} a one-dimensional profile is set up in
-the $z$-direction representing two blocks of fluid with
-$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
-$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
-appropriate width. The $\phi = -1$ section is in the middle.
-
-\inputkey{phi0}
-
-The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
-The default value is zero, i.e., a symmetric 50:50 mixture by volume.
-
-\inputkey{noise}
-
-The magnitude of the initial fluctuations in $\phi$ used to
-initiate spinodal decomposition.
-
-\subsubsection{Binary fluid using two distributions}
-
-\inputkey{symmetric\_lb}
-
-This is the special case where the composition is represented
-by a second LB distribution, and an appropriate lattice kinetic
-equation approximating the Cahn-Hilliard equation is solved.
-In this case the above parameters have the same meaning.
-
-\subsection{Brazovskii}
-
-\inputkey{free\_energy brazovskii}
-
-This is similar to the symmetric free energy, but with one extra term
-in a higher derivative of $\phi$.
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
-\end{equation}
-
-The parameters now include $C$. For $A<0$, phase separation occurs
-with a result depending on $\kappa$: ones get two symmetric phases
-for $\kappa >0$ (cf. symmetric) or a lamellar phase for
-$\kappa < 0$.
-
-\inputkey{A} Bulk parameter $A < 0$.
-
-\inputkey{B} Bulk parameter $B = -A$.
-
-\inputkey{K} Negative for lamellar phase.
-
-\inputkey{C} Positive.
-
-Other parameters, including the mobility, are set as for the
-symmetric free energy (see above).
-
-\subsection{Surfactant in binary fluid}
-
-\inputkey{free\_energy surfactant}
-
-Currently under reconstruction.
-
-This is for use with a two order parameter model the first of
-which is the composition, as for the symmetric free energy,
-while the second $\psi$ represents surfactant concentration.
-The free energy density is made up of a number of terms:
-
-\begin{equation}
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-\end{equation}
-is the standard symmetric term related to the composition.
-These parameters are represented in the input as
-
-\inputkey{surf\_A}
-\inputkey{surf\_B}
-\inputkey{surf\_kappa}
-
-\begin{equation}
-D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
-\end{equation}
-represents the energy of surfactant in the bulk phase with surfactant
-concentration $0 < \psi < 1$ with a single parameter $D$;
-\begin{equation}
--{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
--{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
-\end{equation}
-where the term in $\epsilon$ represents the energy reduction by
-adsorbing surfactant at the interface and the term in $\beta$
-provides an additional favourable term modelling the fact that
-the molecules like to line up; finally, there is an extra term
-\begin{equation}
-+{\textstyle \frac{1}{2}} W \psi \phi^2
-\end{equation}
-which is included to stabilise the interface \cite{theissengompper}.
-
-The corresponding keys for the input file are:
-
-\inputkey{D} is the bulk surfactant contribution parameter
-
-\inputkey{epsilon} is the parameter $\epsilon$
-
-\inputkey{W} is the parameter $W$
-
-\inputkey{beta} is the value of the parameter $\beta$.
-
-\inputkey{phi\_b} the initial, uniform, background surfactant
-concentration $\psi_b$.
-
-\inputkey{mobility\_psi} Sets the (uniform) mobility for the surfactant
-as appropriate for the Cahn-Hilliard equation.
-
-\subsection{Polar Active Gel}
-
-\inputkey{free\_energy polar\_active}
-
-This models a polar active gel with vector order parameter $P_\alpha$.
-The free energy density is:
-
-The corresponding input parameters are:
-
-\inputkey{polar\_active\_a}
-
-\inputkey{polar\_active\_b}
-
-\inputkey{polar\_active\_k}
-
-\inputkey{polar\_active\_klc}
-
-\inputkey{polar\_active\_zeta}
-
-\inputkey{polar\_active\_lambda}
-
-\inputkey{leslie\_ericksen\_gamma}
-
-\inputkey{leslie\_ericksen\_swim}
-
-\subsection{Liquid Crystal}
-
-\inputkey{free\_energy lc\_blue\_phase}
-
-Here, we have a tensor order parameter $Q_{\alpha\beta}$.
-The free energy density is
-\begin{eqnarray}
-f = {\textstyle\frac{1}{2}}A_0(1 - \gamma/3)Q^2_{\alpha\beta}
-  - {\textstyle\frac{1}{3}}A_0 \gamma
-Q_{\alpha\beta}Q_{\beta\delta}Q_{\delta\alpha}
- + {\textstyle\frac{1}{4}}A_0 \gamma (Q^2_{\alpha\beta})^2
-\nonumber
-\\
-+ {\textstyle\frac{1}{2}} \Big(
-\kappa_0 (\epsilon_{\alpha\delta\sigma} \partial_\delta Q_{\sigma\beta} +
-2q_0 Q_{\alpha\beta})^2 + \kappa_1(\partial_\alpha Q_{\alpha\beta})^2 \Big)
-\end{eqnarray}
-
-The corresponding input parameters are:
-
-\inputkey{lc\_a0} The bulk free energy parameter $A_0$
-
-\inputkey{lc\_gamma} The bulk free energy parameter $\gamma$
-
-\inputkey{lc\_q0} The pitch wavenumber $q_0 = 2\pi / p$, where p
-is the pitch.
-
-\inputkey{lc\_kappa0} Elastic constant $\kappa_0$ in distortion free energy
-
-\inputkey{lc\_kappa1} Elastic constant $\kappa_1$ in distortion free energy
-
-Note that the code currently enforces the `one elastic constant'
-approximation ($\kappa_0 = \kappa_1$), so both these values must
-be equal in the input. This constraint may be relaxed in the future.
-
-\inputkey{lc\_Gamma} The rotational diffusion constant appearing in
-the Beris Edwards equation (note \texttt{lc\_gamma} and \texttt{lc\_Gamma}
-are different).
-
-\inputkey{lc\_xi} Aspect ratio for rod lie molecules appearing in the
-stress tensor.
-
-\inputkey{lc\_active\_zeta} Apolar liquid crystal activity constant.
-
-\subsubsection{Initial LC order parameter $Q_{\alpha\beta}$}
-
-\inputkey{lc\_q\_initialisation}
-
-There are a number of useful choices here. The key value
-\texttt{twist} initialises a simple cholesteric with the
-helical axis in the $z-$direction;  \texttt{o8m}
-gives equilibrium BPI, \texttt{o2} gives BPII. The appropriate
-choices for the other parameters must be made correspondingly.
-
-An additional \texttt{lc\_q\_init\_amplitude} value sets the
-scalar order parameter  `amplitude' in each case.
-
-\subsubsection{The redshift}
-
-\inputkey{lc\_init\_redshift}
-
-Sets the initial value of the redshift parameter. For example,
-for the \texttt{o8m} initialisation, the redshift can be set to 0.83,
-and for the \texttt{o2} initialisation, the redshift can be set to 0.91.
-The default redshift is unity.
-
-\inputkey{lc\_redshift\_update [0|1]}
-
-Switches on (default is off) the dynamic computation of the redshift
-from the order parameter gradient terms in the free energy. The
-caluclation is performed at every time step, and override any
-initial value specified in the above. It is also robust to restarts.
-
-\subsection{Colloidal particles}
-
-The are a number of important options required for colloids. If
-none of these keys are present, no colloids will be used.
-
-\inputkey{colloid\_init}
-
-This determines how the code initialises particles (or not). As stated, the
-default is to have no particles (value \texttt{no\_colloids}). There
-are two other options at the moment.
-
-There are two options:
-
-\inputkey{colloid\_init    from\_file}
-
-requires a preprepared file of colloid information. To understand
-how to create such a file in the correct format,
-see the example program \texttt{colloid\_file.c} in the util
-directory.
-
-\inputkey{colloid\_init   random}
-
-This may be useful to initialise a limited number of colloids at
-low volume fraction. Note that there is no guarantee that the
-particles do not overlap, in which case the initialisation will
-terminate. The number of particles and their properties
-are determined by the value of the keys described below.
-
-\inputkey{colloid\_type}
-
-This controls the the type of particle. Appropriate values are
-\texttt{inactive}, giving a standard fully resolved LB colloid;
-\texttt{active} switches on the correction to BBL appropriate
-for active particles (expecting parameters $b_0$ and $b_1$ to be
-set); \texttt{subgrid} gives unresolved particles (radius $< 1$
-lattice unit) and no BBL.
-
-If more than one particle is used, all positions are set at random.
-
-\inputkey{colloid\_gravity  0.0\_0.0\_-0.0001}
-
-Sets a body force on each particle. Useful, for example, for
-sedimentation. For example, the above gives a constant force in
-the negative $z$-direction on each particle.
-The code automatically computes the compensating body force on fluid
-nodes required to give no net change in the total momentum of the
-system at each time step.
-
-\subsubsection{Interactions and the cell list}
-
-In each local domain, colloidal particles are stored in a cell list.
-This is a common structure in molecular dynamics-like problems. Here,
-it also serves as the basis for parallel communication of colloid
-information, so there are a number of constraints which it is important
-to understand. In particular, parallel communication imposes the
-constraint that there be at least 2 cells in the local domain in
-each coordinate direction.
-
-The size of the cells is computed automatically by the code based on
-the local domain size, particle size, and the cut-off distance of any
-pairwise interactions that are relevant. This will usually mean trying
-to maximise the number of cells (to reduce pairwise interaction
-calculation). The user must set the minimum allowable cell width.
-
-\inputkey{colloid\_cell\_min}
-
-This sets the minimum cell list width. This key must be set and there
-must be at least two cells per processor domain. If the minimum does
-not capture the specified interactions, a fatal error will result.
-For spherical particles, the minimum width will typically be
-$2a_h + h_c$ where $h_c$ is an interaction cut-off distance.
-
-In the case where very large particles are required, it can be useful
-to switch of the cell list. This is done via
-
-\inputkey{colloid\_cell\_list\_interactions no}
-
-The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
-there is effectively one cell in each local domain. This option
-should not be required in normal circumstances.
-
-\subsubsection{Initialisation from file}
-
-A file containing initial state properties for one or more colloids
-can be read at run file. This file must be of the correct from, but
-can be ASCII or binary. To help to create such a file, an example
-is given in \texttt{./util/colloid\_file.c}. This file will be
-read correctly in both serial and in parallel. Colloid positions
-should always be in terms of the global coordinates. By default,
-the input expected is \texttt{config.cds.init.001-001}.
-
-All colloid output is in serial (ie., a single file is produced)
-at the moment. Again, it may be ASCII or binary.
-
-\subsubsection{Initialisation at random}
-
-\inputkey{colloid\_random\_no} Sets the total number of particles
-
-\inputkey{colloid\_random\_a0}
-
-\inputkey{colloid\_random\_ah}
-
-\inputkey{colloid\_random\_dh}
-
-The parameter $a_0$ the nominal radius of the particle on the lattice.
-This is used to construct the links for BBL. The smallest acceptable LB
-particle is $a_0 = 1.25$ (approximately). There are 'magic' values
-including 1.25, 2.3, 3.7, 4.77 which minimise discretisation effects.
-The hydrodynamic radius of the particles is $a_h$. For all
-reasonable applications choose $a_h = a_0$. This is the radius for
-all physical purposes. The \texttt{colloid\_random\_dh} parameter
-specifies the smallest allowed surface-surface separation if two or
-more particles (or walls) are required. This is useful to prevent
-very close particles experiencing large and destabilising interactions
-at start time.
-
-The other state for the initial colloids can be set uniformly by
-means of the keys:
-
-\inputkey{colloid\_random\_v0} initial velocity vector
-
-\inputkey{colloid\_random\_w0} initial angular velocity vector
-
-\inputkey{colloid\_random\_s0} initial spin vector
-
-\inputkey{colloid\_random\_m0} initial direction of motion vector for active
-particles
-
-\inputkey{colloid\_random\_b1} active parameter $b_1$
-
-\inputkey{colloid\_random\_b2} active parameter $b_2$ (should be combined
-with the \texttt{active} particle type key as discussed above).
-
-There is one special case. If you want exactly one particle, you may
-set its initial position with
-
-\inputkey{colloid\_random\_r0} the initial position vector.
-
-\subsubsection{Restarting at random!}
-
-Be careful if you are restarting a simulation that used the random
-initialisation at $t=0$. You must change to
-
-\texttt{colloid\_init  from\_file}
-
-to avoid generating a completely new set of random positions
-(which will not be consistent with the density and order parameter
-fields at the restart time). This is almost certainly not what you
-intended!
-
-\subsubsection{Hydrodynamic radius}
-
-There is an important issue to understand about the (nominally)
-spherical particles in LB. The input radius $a_0$ is used to
-construct the link boundary conditions. However, discretisation
-errors in the boundary conditions means that the `real' particle
-size may be different. This real radius is the hydrodynamic radius $a_h$.
-As a further complication, the hydrodynamic radius can also
-be a function of viscosity for a given input radius. Some values
-are given in Table~X. This values are computed using a calibration
-procedure described by Nguyen and Ladd \cite{nguyen2001}.
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-\hline
-$\eta$ & \multicolumn{3}{c|}{$a_0$}\\ \cline{2-4} 
-       & 1.25 & 2.30 & 4.77 \\
-\hline
-1/6  & 1.09 (1.05) & 2.23 (2.20) & 4.76 \\
-1/100  & 1.40 (1.34) & 2.50 (2.46) & 5.01 \\
-1/1000  & 1.63 & 2.71 & 5.23 \\
-\hline
-\end{tabular}
-\end{center}
-\caption{A table of calibrated hydrodynamic radii as a function of input
-radius and fluid viscosity $\eta$. These figures are for D3Q19 (and D3Q15
-in brackets).}
-\end{table}
-
-If a calibration is required for a given radius and viscosity,
-use the \texttt{calibration on} key pair in the input file. Some
-examples are given in the \texttt{tests/calibration} directory.
-An apppropriate
-number of time steps must be used to allow the system to get over
-the initial transient (we use the momentum diffusion time for the
-system $L^2/\eta$) and to allow at least one
-particle Stokes time ($a/U$). The input radius and hydrodynamic
-radius should be set the same in the input file.
-
-% PENDING Ginzburg d'Humieres correct to ghost mode relaxation times
-
 \subsection{Boundary conditions}
@@
 
-\subsection{Lees-Edwards planes}
+\subsubsection{Lees-Edwards planes}
 
@@
 and results will be wrong.
-
-\section{Parallel Performance}
-
-\subsection{General Comments}
-
-The basic LB calculation should scale well in parallel, that is,
-the time taken for fixed problem size will decrease linearly
-with the number of MPI processes (strong scaling), or a larger
-problem can be run on a proportionally larger number processes
-in the same time (weak scaling). To ensure performance is retained
-in parallel, it is useful to understand some basic considerations.
-Some of these are discussed below.
-
-\textit{Ludwig} has been run successfully on up to 131,072 MPI
-tasks with close to ideal scaling, and there is no problem in
-principle to run larger decompositions.
-
-\subsection{Things to Remember and Things to Avoid}
-\ref{section-advice}
-
-There are a number of simple issues to get right:
-\begin{itemize}
-\item
-Make sure the assertions are switched off and appropriate optimisations
-are switched on for production runs (see Makefile).
-\item
-Do not use too small a local domain size. Typically, 16$^3$ to 32$^3$
-cubic local domains will give resaonble scaling. Anything smaller may
-be inefficient. The best size may depend on the exact nature of the
-calculation and the hardware. Note that this limitation will always
-ultimately limit strong scaling.
-\item
-Don't have diagnsotic output (freq\_statistics) more often than is
-necessary. This output requires a global communication which is slow.
-\item
-Use BINARY output options; ASCII output is slower, and produces larger
-files.
-\item
-Do not use serial I/O. This will have an increasingly high overhead as
-the number of processors is increases (it's probably reasonable to
-128--256 MPI tasks). This is because each MPI tasks writes its data,
-in turn, to a single file; this in inherently serial. To avoid this,
-you need to specify the I/O decomposition for the various output
-quantities of interest. This splits the system into regular blocks
-(in a similar approach to the MPI decomposition), the data for which
-are written to separate files in parallel. For example, if the MPI
-decomposition is 8\_8\_8 (512 MPI tasks), then a reasonable I/O
-decomposition may be 2\_2\_2 or 4\_4\_4. The only way to find out
-what works best is to try some tests for the problem size you want to
-address.
-\item
-Don't use the Ewald sum! This simply does not scale.
-An alternative algorithm is required.
-\end{itemize}

### 2014-09-15 tags/2014-pre-release/htdocs/intro.tex
Tag for Sept 2014 trunk
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2463 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-05-10 htdocs/user.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  user.tex
-%
-%  The 'user' section which holds an overview, and discusses
-%  compilation and running.
-%
-%  $Id$
-%
-%  Edinburgh Soft Matter and Statistical Physics Group and
-%  Edinburgh Parallel Computing Centre
-%
-%  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2010 The University of Edinburgh
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Introduction}
@@
 preprocessor flag.
-
-\section{User Input}
-
-The executable will look for an input file at run time.
-By default, this file is named \texttt{input} and should be in the
-current working directory. The alternative is to specify the name
-of the file explicitly on the command line:
-\begin{verbatim}
-$ ./Ludwig.exe input_file
-\end{verbatim}
-A reference input file \texttt{input.ref} is provided as a template.
-
-The contents of the input file are made up of \textit{key value}
-pairs with control the run-time behaviour. Blank lines and lines
-begining with \# are ignored as comments. The following describes
-the effect of various keys. Most keys have a default value which
-will be used by the code if the corresponding key is not present
-(or commented out) in the input file.
-
-\subsection{Setting up the input}
-
-Decide what free energy you want and check the relevant sections
-of this dcoumentation. If you want a simple fluid only, the free
-energy should be set to \texttt{none}.
-
-You should set at least the basic fluid parameters, the system
-size, and number of steps to be run.
-
-\textbf{IMPORTANT:} Although many incompatible choices of input
-parameters are trapped at run time, you must be careful in how
-the input file is set out. If you mistakenly set or unset some
-options in the input, erroneous results will surely result...
-
-\subsection{Basic run parameters}
-
-\inputkey{N\_cycles}
-
-The number of lattice Boltzmann time steps to execute. Default
-value: 0.
-
-\inputkey{N\_start}
-
-By default the code will start from time step $t = 0$. If you 
-wish to restart from a previously saved configuration, set
-the appropriate value for \texttt{N\_start}. The code will
-execute \texttt{N\_cycles} steps starting from this point.
-
-\inputkey{size}
-
-Controls the number of lattice points in the $(x, y, z)$ directions,
-respectively. Default value: 64\_64\_64. If a two-dimensional system
-is required, the $z$-direction can be set to 1.
-
-\inputkey{grid}
-
-In parallel, this sets the processor Cartesian decomposition, i.e.,
-the number of processors in each dimension of the Cartesian communicator.
-The default value may be implementation dependent as it is that returned
-by \texttt{MPI\_Dims\_create()}. For
-example:
-\begin{verbatim}
-size 64_64_64
-grid 4_2_1
-\end{verbatim}
-gives a total (physical) system size of 64 lattice sites in each
-direction, decomposed across 4 processors in the x-direction, 2 in
-the y-direction and 1 (no decomposition) in the z-direction. The
-total number of processors must therefore be 8. The local domain
-size per processor is then 16x32x64.
-
-\subsection{Fluid properties}
-
-\inputkey{free\_energy}
-
-This sets the free energy in use, and hence has many consequences.
-The default value is \texttt{none}, i.e., use a simple fluid only.
-The available choices are for symmetric binary fluids, Brazovskii
-smectics, polar active gels, and liquid crystals. The surfactant
-model is currently being rebuilt.
-
-\inputkey{viscosity}
-
-Sets the LB fluid shear viscosity $\eta$ (and related relaxation time).
-Safe values are roughly $0.2 > \eta > 0.0001 $. Default value is 1/6
-(i.e., relaxation time equal to unity).
-
-\inputkey{viscosity\_bulk}
-
-Sets the LB bulk viscosity $\zeta$ (and related relaxation time).
-This can be useful if you are concerned with incompressiblity
-violations, in which case $\zeta >> \eta$ can be useful. The
-default value is $\zeta = \eta$.
-
-\inputkey{isothermal\_fluctuations}
-
-This switches on the fluctuationing hydrodynamics. Default is off.
-
-\inputkey{temperature}
-
-The LB 'temperature' using fluctuating hydrodynamics. Safe values
-are $0.0001 > kT > 0$.
-
-\inputkey{ghost\_modes}
-
-Allows you to switch the ghost modes off in the collision stage.
-Default is on.
-
-\subsection{Symmetric binary fluid}
-
-\inputkey{free\_energy symmetric}
-
-The following applies for the binary fluid problem with
-compositional order parameter $\phi$ and free energy
-(excluding the term in the density $\rho$):
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
-\end{equation}
-
-This is described in some detail
-by Kendon et al \cite{viv}. The first two terms represent the bulk
-contribution, whereas the term in $\kappa$ penalises curvature in
-the interface.
-
-\inputkey{A} The parameter $A$. Note $A < 0$.
-
-\inputkey{B} The parameter $B$. Note $B = -A$ for common usage,
-although this is not enforced.
-
-\inputkey{K} The parameter $\kappa$, which is positive.
-
-The order parameter evolution is determined by a Cahn-Hilliard equation
-with mobility set by
-
-\inputkey{mobility} Sets mobility $M$ (uniform in space).
-
-The following parameters control the initialisation of the order
-parameter.
-
-\inputkey{phi\_initialisation}
-
-Determines how the compositional order parameter is initialised
-at the start of the run. If set to \texttt{spinodal} the value
-is set to $\phi_0$ as set above plus or minus a random noise, the
-magnitude of which is set by the value of the \texttt{noise} key.
-If set to \texttt{block} a one-dimensional profile is set up in
-the $z$-direction representing two blocks of fluid with
-$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
-$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
-appropriate width. The $\phi = -1$ section is in the middle.
-
-\inputkey{phi0}
-
-The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
-The default value is zero, i.e., a symmetric 50:50 mixture by volume.
-
-\inputkey{noise}
-
-The magnitude of the initial fluctuations in $\phi$ used to
-initiate spinodal decomposition.
-
-\subsubsection{Binary fluid using two distributions}
-
-\inputkey{symmetric\_lb}
-
-This is the special case where the composition is represented
-by a second LB distribution, and an appropriate lattice kinetic
-equation approximating the Cahn-Hilliard equation is solved.
-In this case the above parameters have the same meaning.
-
-\subsection{Brazovskii}
-
-\inputkey{free\_energy brazovskii}
-
-This is similar to the symmetric free energy, but with one extra term
-in a higher derivative of $\phi$.
-\begin{equation}
- F[\phi] = 
-\int dr \left(
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
-\end{equation}
-
-The parameters now include $C$. For $A<0$, phase separation occurs
-with a result depending on $\kappa$: ones get two symmetric phases
-for $\kappa >0$ (cf. symmetric) or a lamellar phase for
-$\kappa < 0$.
-
-\inputkey{A} Bulk parameter $A < 0$.
-
-\inputkey{B} Bulk parameter $B = -A$.
-
-\inputkey{K} Negative for lamellar phase.
-
-\inputkey{C} Positive.
-
-Other parameters, including the mobility, are set as for the
-symmetric free energy (see above).
-
-\subsection{Surfactant in binary fluid}
-
-\inputkey{free\_energy surfactant}
-
-Currently under reconstruction.
-
-This is for use with a two order parameter model the first of
-which is the composition, as for the symmetric free energy,
-while the second $\psi$ represents surfactant concentration.
-The free energy density is made up of a number of terms:
-
-\begin{equation}
-{\textstyle \frac{1}{2}}A\phi^2
-+ {\textstyle \frac{1}{4}}B\phi^4
-+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
-\end{equation}
-is the standard symmetric term related to the composition.
-These parameters are represented in the input as
-
-\inputkey{surf\_A}
-\inputkey{surf\_B}
-\inputkey{surf\_kappa}
-
-\begin{equation}
-D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
-\end{equation}
-represents the energy of surfactant in the bulk phase with surfactant
-concentration $0 < \psi < 1$ with a single parameter $D$;
-\begin{equation}
--{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
--{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
-\end{equation}
-where the term in $\epsilon$ represents the energy reduction by
-adsorbing surfactant at the interface and the term in $\beta$
-provides an additional favourable term modelling the fact that
-the molecules like to line up; finally, there is an extra term
-\begin{equation}
-+{\textstyle \frac{1}{2}} W \psi \phi^2
-\end{equation}
-which is included to stabilise the interface \cite{theissengompper}.
-
-The corresponding keys for the input file are:
-
-\inputkey{D} is the bulk surfactant contribution parameter
-
-\inputkey{epsilon} is the parameter $\epsilon$
-
-\inputkey{W} is the parameter $W$
-
-\inputkey{beta} is the value of the parameter $\beta$.
-
-\inputkey{phi\_b} the initial, uniform, background surfactant
-concentration $\psi_b$.
-
-\inputkey{mobility\_psi} Sets the (uniform) mobility for the surfactant
-as appropriate for the Cahn-Hilliard equation.
-
-\subsection{Polar Active Gel}
-
-\inputkey{free\_energy polar\_active}
-
-This models a polar active gel with vector order parameter $P_\alpha$.
-The free energy density is:
-
-The corresponding input parameters are:
-
-\inputkey{polar\_active\_a}
-
-\inputkey{polar\_active\_b}
-
-\inputkey{polar\_active\_k}
-
-\inputkey{polar\_active\_klc}
-
-\inputkey{polar\_active\_zeta}
-
-\inputkey{polar\_active\_lambda}
-
-\inputkey{leslie\_ericksen\_gamma}
-
-\inputkey{leslie\_ericksen\_swim}
-
-\subsection{Liquid Crystal}
-
-\inputkey{free\_energy lc\_blue\_phase}
-
-Here, we have a tensor order parameter $Q_{\alpha\beta}$.
-The free energy density is
-\begin{eqnarray}
-f = {\textstyle\frac{1}{2}}A_0(1 - \gamma/3)Q^2_{\alpha\beta}
-  - {\textstyle\frac{1}{3}}A_0 \gamma
-Q_{\alpha\beta}Q_{\beta\delta}Q_{\delta\alpha}
- + {\textstyle\frac{1}{4}}A_0 \gamma (Q^2_{\alpha\beta})^2
-\nonumber
-\\
-+ {\textstyle\frac{1}{2}} \Big(
-\kappa_0 (\epsilon_{\alpha\delta\sigma} \partial_\delta Q_{\sigma\beta} +
-2q_0 Q_{\alpha\beta})^2 + \kappa_1(\partial_\alpha Q_{\alpha\beta})^2 \Big)
-\end{eqnarray}
-
-The corresponding input parameters are:
-
-\inputkey{lc\_a0} The bulk free energy parameter $A_0$
-
-\inputkey{lc\_gamma} The bulk free energy parameter $\gamma$
-
-\inputkey{lc\_q0} The pitch wavenumber $q_0 = 2\pi / p$, where p
-is the pitch.
-
-\inputkey{lc\_kappa0} Elastic constant $\kappa_0$ in distortion free energy
-
-\inputkey{lc\_kappa1} Elastic constant $\kappa_1$ in distortion free energy
-
-Note that the code currently enforces the `one elastic constant'
-approximation ($\kappa_0 = \kappa_1$), so both these values must
-be equal in the input. This constraint may be relaxed in the future.
-
-\inputkey{lc\_Gamma} The rotational diffusion constant appearing in
-the Beris Edwards equation (note \texttt{lc\_gamma} and \texttt{lc\_Gamma}
-are different).
-
-\inputkey{lc\_xi} Aspect ratio for rod lie molecules appearing in the
-stress tensor.
-
-\inputkey{lc\_active\_zeta} Apolar liquid crystal activity constant.
-
-\subsubsection{Initial LC order parameter $Q_{\alpha\beta}$}
-
-\inputkey{lc\_q\_initialisation}
-
-There are a number of useful choices here. The key value
-\texttt{twist} initialises a simple cholesteric with the
-helical axis in the $z-$direction;  \texttt{o8m}
-gives equilibrium BPI, \texttt{o2} gives BPII. The appropriate
-choices for the other parameters must be made correspondingly.
-
-An additional \texttt{lc\_q\_init\_amplitude} value sets the
-scalar order parameter  `amplitude' in each case.
-
-\subsubsection{The redshift}
-
-\inputkey{lc\_init\_redshift}
-
-Sets the initial value of the redshift parameter. For example,
-for the \texttt{o8m} initialisation, the redshift can be set to 0.83,
-and for the \texttt{o2} initialisation, the redshift can be set to 0.91.
-The default redshift is unity.
-
-\inputkey{lc\_redshift\_update [0|1]}
-
-Switches on (default is off) the dynamic computation of the redshift
-from the order parameter gradient terms in the free energy. The
-caluclation is performed at every time step, and override any
-initial value specified in the above. It is also robust to restarts.
-
-\subsection{Colloidal particles}
-
-The are a number of important options required for colloids. If
-none of these keys are present, no colloids will be used.
-
-\inputkey{colloid\_init}
-
-This determines how the code initialises particles (or not). As stated, the
-default is to have no particles (value \texttt{no\_colloids}). There
-are two other options at the moment.
-
-There are two options:
-
-\inputkey{colloid\_init    from\_file}
-
-requires a preprepared file of colloid information. To understand
-how to create such a file in the correct format,
-see the example program \texttt{colloid\_file.c} in the util
-directory.
-
-\inputkey{colloid\_init   random}
-
-This may be useful to initialise a limited number of colloids at
-low volume fraction. Note that there is no guarantee that the
-particles do not overlap, in which case the initialisation will
-terminate. The number of particles and their properties
-are determined by the value of the keys described below.
-
-\inputkey{colloid\_type}
-
-This controls the the type of particle. Appropriate values are
-\texttt{inactive}, giving a standard fully resolved LB colloid;
-\texttt{active} switches on the correction to BBL appropriate
-for active particles (expecting parameters $b_0$ and $b_1$ to be
-set); \texttt{subgrid} gives unresolved particles (radius $< 1$
-lattice unit) and no BBL.
-
-If more than one particle is used, all positions are set at random.
-
-\inputkey{colloid\_gravity  0.0\_0.0\_-0.0001}
-
-Sets a body force on each particle. Useful, for example, for
-sedimentation. For example, the above gives a constant force in
-the negative $z$-direction on each particle.
-The code automatically computes the compensating body force on fluid
-nodes required to give no net change in the total momentum of the
-system at each time step.
-
-\subsubsection{Interactions and the cell list}
-
-In each local domain, colloidal particles are stored in a cell list.
-This is a common structure in molecular dynamics-like problems. Here,
-it also serves as the basis for parallel communication of colloid
-information, so there are a number of constraints which it is important
-to understand. In particular, parallel communication imposes the
-constraint that there be at least 2 cells in the local domain in
-each coordinate direction.
-
-The size of the cells is computed automatically by the code based on
-the local domain size, particle size, and the cut-off distance of any
-pairwise interactions that are relevant. This will usually mean trying
-to maximise the number of cells (to reduce pairwise interaction
-calculation). The user must set the minimum allowable cell width.
-
-\inputkey{colloid\_cell\_min}
-
-This sets the minimum cell list width. This key must be set and there
-must be at least two cells per processor domain. If the minimum does
-not capture the specified interactions, a fatal error will result.
-For spherical particles, the minimum width will typically be
-$2a_h + h_c$ where $h_c$ is an interaction cut-off distance.
-
-In the case where very large particles are required, it can be useful
-to switch of the cell list. This is done via
-
-\inputkey{colloid\_cell\_list\_interactions no}
-
-The constraint in this case is that $2a_h + h_c < L_{local}$, ie.,
-there is effectively one cell in each local domain. This option
-should not be required in normal circumstances.
-
-\subsubsection{Initialisation from file}
-
-A file containing initial state properties for one or more colloids
-can be read at run file. This file must be of the correct from, but
-can be ASCII or binary. To help to create such a file, an example
-is given in \texttt{./util/colloid\_file.c}. This file will be
-read correctly in both serial and in parallel. Colloid positions
-should always be in terms of the global coordinates. By default,
-the input expected is \texttt{config.cds.init.001-001}.
-
-All colloid output is in serial (ie., a single file is produced)
-at the moment. Again, it may be ASCII or binary.
-
-\subsubsection{Initialisation at random}
-
-\inputkey{colloid\_random\_no} Sets the total number of particles
-
-\inputkey{colloid\_random\_a0}
-
-\inputkey{colloid\_random\_ah}
-
-\inputkey{colloid\_random\_dh}
-
-The parameter $a_0$ the nominal radius of the particle on the lattice.
-This is used to construct the links for BBL. The smallest acceptable LB
-particle is $a_0 = 1.25$ (approximately). There are 'magic' values
-including 1.25, 2.3, 3.7, 4.77 which minimise discretisation effects.
-The hydrodynamic radius of the particles is $a_h$. For all
-reasonable applications choose $a_h = a_0$. This is the radius for
-all physical purposes. The \texttt{colloid\_random\_dh} parameter
-specifies the smallest allowed surface-surface separation if two or
-more particles (or walls) are required. This is useful to prevent
-very close particles experiencing large and destabilising interactions
-at start time.
-
-The other state for the initial colloids can be set uniformly by
-means of the keys:
-
-\inputkey{colloid\_random\_v0} initial velocity vector
-
-\inputkey{colloid\_random\_w0} initial angular velocity vector
-
-\inputkey{colloid\_random\_s0} initial spin vector
-
-\inputkey{colloid\_random\_m0} initial direction of motion vector for active
-particles
-
-\inputkey{colloid\_random\_b1} active parameter $b_1$
-
-\inputkey{colloid\_random\_b2} active parameter $b_2$ (should be combined
-with the \texttt{active} particle type key as discussed above).
-
-There is one special case. If you want exactly one particle, you may
-set its initial position with
-
-\inputkey{colloid\_random\_r0} the initial position vector.
-
-\subsubsection{Restarting at random!}
-
-Be careful if you are restarting a simulation that used the random
-initialisation at $t=0$. You must change to
-
-\texttt{colloid\_init  from\_file}
-
-to avoid generating a completely new set of random positions
-(which will not be consistent with the density and order parameter
-fields at the restart time). This is almost certainly not what you
-intended!
-
-\subsubsection{Hydrodynamic radius}
-
-There is an important issue to understand about the (nominally)
-spherical particles in LB. The input radius $a_0$ is used to
-construct the link boundary conditions. However, discretisation
-errors in the boundary conditions means that the `real' particle
-size may be different. This real radius is the hydrodynamic radius $a_h$.
-As a further complication, the hydrodynamic radius can also
-be a function of viscosity for a given input radius. Some values
-are given in Table~X. This values are computed using a calibration
-procedure described by Nguyen and Ladd \cite{nguyen2001}.
-
-\begin{table}[h]
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-\hline
-$\eta$ & \multicolumn{3}{c|}{$a_0$}\\ \cline{2-4} 
-       & 1.25 & 2.30 & 4.77 \\
-\hline
-1/6  & 1.09 (1.05) & 2.23 (2.20) & 4.76 \\
-1/100  & 1.40 (1.34) & 2.50 (2.46) & 5.01 \\
-1/1000  & 1.63 & 2.71 & 5.23 \\
-\hline
-\end{tabular}
-\end{center}
-\caption{A table of calibrated hydrodynamic radii as a function of input
-radius and fluid viscosity $\eta$. These figures are for D3Q19 (and D3Q15
-in brackets).}
-\end{table}
-
-If a calibration is required for a given radius and viscosity,
-use the \texttt{calibration on} key pair in the input file. Some
-examples are given in the \texttt{tests/calibration} directory.
-An apppropriate
-number of time steps must be used to allow the system to get over
-the initial transient (we use the momentum diffusion time for the
-system $L^2/\eta$) and to allow at least one
-particle Stokes time ($a/U$). The input radius and hydrodynamic
-radius should be set the same in the input file.
-
-% PENDING Ginzburg d'Humieres correct to ghost mode relaxation times
-
-\subsection{Boundary conditions}
-
-Solid planar boundary walls can be added at the extremities of the
-system (i.e., the volume of fluid is exactly that specified by
-the \texttt{size} keyword). The wall can be in one, two or all three
-coordinate directions, i.e., a 'slab' geometry, and 'duct' geometry
-or a 'box' geometry, respectively.
-
-The boundary walls are switched on via the keyword
-
-\inputkey{boundary\_walls}
-
-and with a complementatary specification of the key \texttt{periodicity}.
-
-For example, walls at $x = x_{\min}$  and $x = x_{\max}$  are specified
-by
-
-\inputkey{boundary\_walls 1\_0\_0}
-
-\inputkey{periodicity 0\_1\_1}
-
-Both must be present. Note that there are a number of constraints on
-valid choices, and the code will stop if there is an invalid choice, or
-the two keys above do not match. Common choices are:
-\texttt{boundary\_walls 0\_0\_1},
-which gives periodic boundaries in the $x-$ and $y-$directions, with
-walls at either edge in the $z-$direction; \texttt{0\_1\_1}
-gives periodic conditions in the $x-$ direction only, while
-\texttt{1\_1\_1} is a fully enclosed box.
-
-In the case of the slab geometry with walls at $z = z_{\min}$ and
-$z= z_{\max}$ (only),
-two further values are used so that shear may be imparted to the system:
-
-\inputkey{boundary\_speed\_bottom}
-
-\inputkey{boundary\_speed\_top}
-
-These values set the $x-$component of the wall velocity at
-$z = z_{\min}$ and $z = z_{\max}$, respectively. Like all
-velocities, these values must respect the low Mach number
-constraint $u_x < c_s$.
-
-There is a further option for use with the special case in which
-top and bottom walls in the $z-direction$ are used to drive a shear
-flow (typically with the top speed positive and the bottom speed
-negative). The key
-
-\inputkey{boundary\_shear\_init}
-
-may be switched on (set to 1) to provide initialisation of the
-LB distributions appropriate for the specified shear rate based
-on a linear velocity profile in the $z-$direction. The default
-is to initialise the fluid at rest.
-
-\subsection{Lees-Edwards planes}
-
-\inputkey{N\_LE\_plane}
-
-sets the total number of Lees-Edwards planes. Default is zero.
-The placing is as followings. The number of planes $n$ must
-divide the lattice size in the $x$-direction to give an integer
-$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
-
-\inputkey{LE\_plane\_vel}
-
-sets the velocity of each plane relative to the lattice. All planes
-have the same, constant, velocity.
-
-\inputkey{LE\_init\_profile}
-
-if set to 1, the fluid velocity is initialise to reflect a steady
-state shear flow appropriate for the number of planes at the
-given velocity. If set to zero, the fluid is initialised with
-zero velocity.
-
-Note that when the Lees Edwards boundaries are in place, output
-files containing lattice quantities must be `unrolled' to remove
-the time-dependent displacement of the planes. This is done using
-the extraction utility (see section on parallel I/O).
-
-The code works out the
-current displacement of the planes by computing $U_{LE} t$, where
-$t$ is the current time step. A shear run should then start from
-$t = 0$, otherwise an offset is required.
-
-\inputkey{LE\_time\_offset}
-
-It is often convenient to run an equilibration with no shear, and
-then to start an experiment after some number of steps. This
-key allows you to offset the start of the Lees-Edwards motion.
-It should then take the value of the start time (in time steps)
-corresponding to the restart at the end of the equilibration
-period. You also need to take this value into account when unrolling
-the output.
-
-\subsubsection{Lees-Edwards planes in parallel}
-
-There are a couple of additional constraints to use the Lees-Edwards
-planes in parallel. In particular, the planes cannot fall at a
-processor boundary in the $x$-direction. This means you should
-arrange an integer number of planes per process in the $x$-direction.
-(For example, use one plane per process; this will also ensure the number
-of planes
-still evenly divides the total system size.)
-This will interleave the planes with the processor decomposition.
-The $y$-direction and $z$-direction may be decomposed without
-further constraint.
-
-Note that this means a simulation with one plane will only work
-if there is one process in the $x$ decomposition.
-
-\subsection{Input/Output}
-
-\subsubsection{Serial I/O and standard output}
-
-A variety of information is printed to standard output during the
-course of the simulation. This should allow the user to keep track
-of basic progress in terms of integrated quantities, total mass,
-total momentum and so on. Frequency of output is controlled by
-
-\inputkey{freq\_statistics}
-
-which is an integer. Note that these global statistics require
-global communication, and so can affect performance in parallel.
-It is recommended that the frequency is kept to a minimum for
-large systems (e.g., every 500 or 1000 time steps). This
-consideration is true for most types of output (see section
-on parallel I/O).
-
-\subsubsection{Configurations}
-
-It may be appropriate to save entire model configurations from
-time to time, and certainly when a restart is wanted. For each lattice
-quantity, in addition to the data themselves, the program produces a
-single meta-data file (ending in
-\texttt{.meta}) which is used to describe these data. Colloid
-data does not require a meta-data file at the moment.
-
-\inputkey{freq\_config}
-
-This is an integer and sets the frequency of full configuration dumps.
-These may be large depending on the size of the system.
-Full configurations should consist of the LB distributions,
-order parameter (if present) and colloids (if present).
-
-\inputkey{config\_at\_end [yes | no]}
-
-forces (or switches off) a configuration dump at the end of the run.
-The default is to produce a configuration at the end of the run.
-
-\subsubsection{Other lattice quantities}
-
-A number of other lattice-based quantities are available as output.
-
-\inputkey{freq\_measure}
-
-sets the frequency of recording of colloid/order parameter output.
-This includes the combined scalar order parameter and director
-field output in the case of the liquid crystal free energy.
-
-\inputkey{freq\_phi}
-
-set frequency of order parameter output.
-
-\inputkey{freq\_vel}
-
-set frequency of velocity output. Note that all the hydrodynamic
-quantities can be reconstructed from the distribution output, if
-available. If only the velocity field is required, this method
-allows a considerable saving in storage.
-
-\subsubsection{I/O format}
-
-\inputkey{vel\_format}
-
-\inputkey{phi\_format}
-
-May be either \texttt{ASCII} or \texttt{BINARY}. Default is \texttt{BINARY}.
-Configuration I/O is always in binary. It is recommended to use
-binary output for speed and disk space considerations.
-
-\subsubsection{Parallel I/O}
-
-In serial, each output quantity appears in one file (\texttt{phi-}
-and so on). In parallel, this is also the case by default. However,
-this does not scale, owing to the fact that each process has to
-write to the same file in turn (ie., in serial). The solution is
-to make the I/O parallel. This is done by splitting the domain into
-different ``I/O grids'' each of which writes to a separate file.
-
-The default I/O grid can be set using, e.g.,
-
-\inputkey{default\_io\_grid 2\_2\_2}
-
-meaning that the domain is decomposed into a 2x2x2 Cartesian
-decomposition for I/O purposes. This will give rise to a total
-of 8 files per I/O event with extensions \texttt{.008-001} to
-\texttt{.008-008}. These files must be recombined if analysis
-is required. However, restarting can be performed with the
-parallel files.
-
-Individual lattice quantities can use separate I/O grids. This
-may be particularly useful for the distribution data, which are
-much larger than the other lattice quantities. This means a
-finer decomposition may be appropriate. Clearly, the size of the
-I/O decomposition cannot exceed the processor decomposition.
-See the advice in section \cite{section-advice} for guidelines.
-
-\subsection{Miscellaneous}
-
-\inputkey{random\_seed}
-
-sets the random number generator seed.
-
-\subsection{Dealing with the Parallel I/O files}
-
-When running in parallel, output of lattice-based (and particle)
-quantities takes place in parallel. As a consequence, the order
-of the output in the files is not in the same order as would be
-the case in serial. This means the parallel output must be
-manipulated into the correct order before any analysis.
-
-Parallel I/O subdivides the processors into one or more groups,
-each of which write data to a separate file. Information on the
-content of each of these files is contained in `meta-data' file
-for each quantity, which are produced automatically at the start
-of each run. The files are called, e.g., \texttt{phi.001-001.meta}.
-If there are 8 I/O groups, the files will be named \texttt{008-001}
-to \texttt{008-008} for the different groups.
-
-The actual data is stored for each relevant time step as, e.g.,
-\texttt{phi-000100.001-001} where the extension refers to the
-I/O. To recombined a single, or multiple, data files into a
-single file in the correct order, a utility program
-\texttt{extract.c} is provided.
-
-The executable takes two arguments, e.g.,
-
-\inputkey{./extract phi.001-001.meta phi-000100.001-001}
-
-where  the first is the first meta-data file in the relevant series,
-and the second is the first data file for the relevant time step. The
-program will combine the relevant quantities in the correct order
-and produce a single new file \texttt{phi-000100} (in the above case).
-
-\subsection{Restarting}
-
-When restarting in parallel, the processor decomposition must be
-preserved. An attempt to change the processor decomposition will
-cause the configuration files to be read in the incorrect order,
-and results will be wrong.
-
-\section{Parallel Performance}
-
-\subsection{General Comments}
-
-The basic LB calculation should scale well in parallel, that is,
-the time taken for fixed problem size will decrease linearly
-with the number of MPI processes (strong scaling), or a larger
-problem can be run on a proportionally larger number processes
-in the same time (weak scaling). To ensure performance is retained
-in parallel, it is useful to understand some basic considerations.
-Some of these are discussed below.
-
-\textit{Ludwig} has been run successfully on up to 131,072 MPI
-tasks with close to ideal scaling, and there is no problem in
-principle to run larger decompositions.
-
-\subsection{Things to Remember and Things to Avoid}
-\ref{section-advice}
-
-There are a number of simple issues to get right:
-\begin{itemize}
-\item
-Make sure the assertions are switched off and appropriate optimisations
-are switched on for production runs (see Makefile).
-\item
-Do not use too small a local domain size. Typically, 16$^3$ to 32$^3$
-cubic local domains will give resaonble scaling. Anything smaller may
-be inefficient. The best size may depend on the exact nature of the
-calculation and the hardware. Note that this limitation will always
-ultimately limit strong scaling.
-\item
-Don't have diagnsotic output (freq\_statistics) more often than is
-necessary. This output requires a global communication which is slow.
-\item
-Use BINARY output options; ASCII output is slower, and produces larger
-files.
-\item
-Do not use serial I/O. This will have an increasingly high overhead as
-the number of processors is increases (it's probably reasonable to
-128--256 MPI tasks). This is because each MPI tasks writes its data,
-in turn, to a single file; this in inherently serial. To avoid this,
-you need to specify the I/O decomposition for the various output
-quantities of interest. This splits the system into regular blocks
-(in a similar approach to the MPI decomposition), the data for which
-are written to separate files in parallel. For example, if the MPI
-decomposition is 8\_8\_8 (512 MPI tasks), then a reasonable I/O
-decomposition may be 2\_2\_2 or 4\_4\_4. The only way to find out
-what works best is to try some tests for the problem size you want to
-address.
-\item
-Don't use the Ewald sum! This simply does not scale.
-An alternative algorithm is required.
-\end{itemize}

### 2014-09-15 tags/2014-pre-release/htdocs/porous.tex
Tag for Sept 2014 trunk
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2463 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  porous.tex
%
%  Some considerations for porous media.
%
%  $Id$
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2011 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Porous Media}

Porous media calculations can be undertaken when appropriate
solid/fluid status information is supplied. The are a number
of switches available in the input file:

\inputkey{porous\_media\_file}

specifies the file stub name to be read at the start of execution.
(If the stub name is \texttt{file} then the code will expect to
find \texttt{file.001-001} in the current directory.

\inputkey{porous\_media\_format}

is either \texttt{ASCII} or \texttt{BINARY} as appropriate. Note that
in parallel, a single data file can be supplied, but it must be binary.
The default is \texttt{BINARY}.

\inputkey{porous\_media\_type}

is either \texttt{status\_only} (the default) or \texttt{status\_with\_h}.

\subsection{File format}

In all cases the status (fluid/solid) information is represented by a
single \texttt{char} (or integer), which must be supplied via the
porous media file. A single file should contain data matching the
current system size, and have the $z-$direction running fastest,
followed by the $y-$direction. Note that in non-periodic directions,
the structure must be 'closed', i.e., all the points at the edge
should be solid.

Fluid sites are designated by \texttt{0} and boundary or solid sites
by \texttt{1}. These data should be of type \texttt{char} in binary,
and may be integer in ASCII.

Where wetting information is required, the free energy parameter $H$
can be supplied by using the \texttt{status\_with\_h} switch. In this
case, the \texttt{char} status is augmented by a single \texttt{double}
value which is the local value of $H$. The order is then
$s_1,h_1, s_2, h_2, \ldots$.

An example of how to construct a porous media file is provided in
\texttt{util/capillary.c}, which builds an appropriate file for
a square or circular capillary tube. Please see the comments in
the file for further details. Note that the allowed
solid/fluid status values are defined in \texttt{src/site\_map.h}.
A solid boundary is \texttt{BOUNDARY}, while fluid is \texttt{FLUID}.

\subsection{Permeability calculations}

Single fluid permeability calculations for a given proous structure
can be undertaken by driving a flow via the fluid body force. Note
that the structure must be periodic in the direction of the force
(this may mean duplicating a 'reflected' version of a given sample
to create the correct input). The body force can be specified so
that, e.g., to drive a flow in the positive $x-$direction

\texttt{force 0.001\_0.000\_0.000}

The force should not be so large that the maximum velocity generated
threatens the Mach number constraint. To get a measurement of the
flow at equilibrium, the calculation should be run at least the
momentum diffusion time for the system ($L^2/\eta$ in LB time steps).

The net flow can be measured by combining the statistics for the
total momemtum and the total density (which is equal to the volume
with $\rho_0 = 1$).

Note that in the case of porous media with narrow channels at the grid
scale, the wall velocity can be dependent on the viscosity of the fluid.
This is an artefact of the bounce-back on links and will result in a
viscosity-dependent permeability (see, e.g., \cite{lipanmiller}).
To minimise this effect, e.g.,  Ginzburg and d'Humi\`eres \cite{ginzburg}
corrects the viscosity-dependence of the apparent boundary position
using a three relaxation time scheme. This is currently under
investigation.

\subsection{Circular and rectangular capillaries}

\label{section:exact_conductance}

For an infinite  capillary of circular cross section radius $a$,
the conductance is known analytically\cite{papanastasiou}. The
flow per unit area $J$ is given by
\begin{equation}
J = - \frac{1}{8\eta} \frac{\partial p}{ \partial x} a^2. 
\end{equation}
If the pressure gradient is replaced by a uniform body force
$-\partial p/ \partial x = \rho g$, then one can define a
viscosity-independent conductance $C$ via $ J = C \rho g / \eta$,
i.e., $C = a^2/8$. So, by measuring the volume flux at steady state
for a fixed applied body force, one can compute an estimate of the
conductance to compare with this result.

An analtyical expression is also available for the conductance
of a square or rectugular capillary. For an infinite capillary
of square cross section width $w \times h$ (where $h$ is the
longer),
the volume flux per unit area $J$ is expected to be
\cite{papanastasiou,edo1}
\begin{equation}
J = - \frac{1}{3\eta} \frac{\partial p}{\partial x} (h/2)^2
\left[
1 - 6(h/w) \sum_{k=1}^{\infty} \frac{\tanh(\alpha_k w/h)}{\alpha_k^5} 
\right]
\end{equation}
where $\alpha_k = (2k - 1)\pi/2$. The pressure gradient
$-\partial p / \partial x$ is replaced by the body force $\rho g$ and
one can difine the viscosity-independnet conductance $C$ via
$J = C\rho g / \eta$. The calculation procedes as above.

\subsection{Worked example}

\subsubsection{Setting up the structure}

We will work out the conductance of a simple one-dimensional capillary
of circular cross section. We can set up the structure using the
utility program found in \texttt{util/capillary.c}.

We set the system size to be $(10,10,32)$ in the $x-$, $y-$, and
$z-$directions, respectively. The cross section in $x-y$ will be
a discrete  approximation to a circle. The nominal radius is
$a = 4$ lattice units (allowing for solid sites at each edge).
In \texttt{capillary.c} we
set
\begin{verbatim}
const int xmax = 10;
const int ymax = 10;
const int zmax = 32;
\end{verbatim}
Note that the conductance should be independent of the length of the
capillary in the $z-$direction. We will check this later.

We choose a circular cross section via:
\begin{verbatim}
enum {CIRCLE, SQUARE};
const int xsection = CIRCLE;
\end{verbatim}
For this problem, the wetting parameters are irrelevant, and can be
ignored. In addition we set
\begin{verbatim}
enum {STATUS_ONLY, STATUS_WITH_H};
const int output_type = STATUS_ONLY;
\end{verbatim}
to specify that the output will contain only the structural information.
The output filename containing the structure is set via
\begin{verbatim}
const char * filename = "capillary.001-001";
\end{verbatim}

The program may be compiled and run via
\begin{verbatim}
$ gcc capillary.c -lm
$ ./a.out
\end{verbatim}
The output provides a simple representation of the cross section,
and a count of the number of solid sites, the number of fluid sites,
and the total. Again, the wetting parameters are irrelevant. For this
case we have
\begin{verbatim}
Cross section (0 = fluid, 1 = solid)
 1 1 1 1 1 1 1 1 1 1
 1 1 1 0 0 0 0 1 1 1
 1 1 0 0 0 0 0 0 1 1
 1 0 0 0 0 0 0 0 0 1
 1 0 0 0 0 0 0 0 0 1
 1 0 0 0 0 0 0 0 0 1
 1 0 0 0 0 0 0 0 0 1
 1 1 0 0 0 0 0 0 1 1
 1 1 1 0 0 0 0 1 1 1
 1 1 1 1 1 1 1 1 1 1
n = 3200 nsolid = 1536 nfluid = 1664
\end{verbatim}
Note that the outermost sites in each direction here are solid. There
should be a file \texttt{capillary.001-001} in the current directory.
This should be moved to the executable directory.

\subsubsection{Setting up the input}

Assuming we have compiled the serial code, we must now set up the
input file to be consistent with the capillary structure we want
to use. The \texttt{src/input.ref} file can be used as a template.
We should set the system size
\begin{verbatim}
size 10_10_32
\end{verbatim}
The location of the porous media file is specified using
\begin{verbatim}
porous_media_format BINARY
porous_media_file   capillary
porous_media_type   status_only
\end{verbatim}
Note that the extension \texttt{.001-001} of the filename has been
removed in the input file (it gets added back automatically by the
main code at run time).

As this is a single fluid calculation with no free energy involved
we must set
\begin{verbatim}
free_energy none
\end{verbatim}
What are the fluid parameters? We will set the viscosity to
$1/6$ in lattice units (and the bulk viscosity to the default
value by commenting it out; the default value is the same as
the shear viscosity):
\begin{verbatim}
viscosity 0.16666666666666666
#bulk_viscosity 0.1
\end{verbatim}
We expect that the number of time steps to reach a steady state
will be of order $t \approx a^2/\eta$, which we estimate using
$a=4$ so $t \approx 100$ LB time steps. We will try
\begin{verbatim}
N_cycles 200
\end{verbatim}
Note that we will need output on the total momentum of the system
as a function of time, so we set
\begin{verbatim}
freq_statistics 50
\end{verbatim}

Finally, we need to set a force in the $z-$direction to drive the
flow. Again, the final conductance should be indpendent of this
force providing the force is small enough that both the Mach number
and the Reynolds number are small compared to unity in steady stead.
We will set
\begin{verbatim}
force 0.00_0.00_0.0000001
\end{verbatim}

\subsubsection{Extracting the conductance}

With these parameters specified, we can run the code. The output
should reflect the parameters in the input file, and the time step
loop should start if the parameters, and the \texttt{capillary.001-001}
file has been read successfully. The code reports various fluid
properties. The density is \texttt{[rho]}, and we see that the
total is the same as the number of fluid sites reported by the
\texttt{capillary.c} program:
\begin{verbatim}
Scalars - total mean variance min max
[rho]    1664.00  1.00000000000-2.2204460e-16  1.00000000000 1.00000000000
\end{verbatim}
Note that the total density (ie., mass) should remain exactly unchanged
for the duration of the run; if not, there is something wrong.

The total momentum of the fluid in each coordinate direction is
reported, and we should see, as a function of time
\begin{verbatim}
Momentum - x y z
...
[fluid   ]  1.5321078e-14  1.8207658e-14  1.8529405e-03
[fluid   ] -8.2378548e-14  2.8865799e-15  1.9422398e-03
[fluid   ]  1.1302070e-13  6.1950445e-14  1.9467213e-03
[fluid   ]  2.4091840e-14  1.1102230e-16  1.9469462e-03
\end{verbatim}
Note as we have applied an external force in the $z-$direction,
the $z-$momentum increases with time while the momenutm in the
other two directions is constant (to machine accuracy, which is
around $10^{-16}$). Further, the $z-$momentum
is still changing slightly after 200 time steps, so we have
not run long enough to reach a steady state. If we run for
400 steps, we should see that the $z-$momentum is unchanged
over the last 100 time steps.
\begin{verbatim}
Momentum - x y z
...
[fluid   ]  1.0424994e-13  1.6098234e-14  1.9469580e-03
[fluid   ]  3.2973624e-14  3.2862602e-14  1.9469580e-03
\end{verbatim}
Finally, note the maximum velocity in the flow direction
is small compared with unity, and so both Mach number and
Reynolds number are also small in this case.

The figure we are interested in is the total momentum in the
$z-$direction which is $1.947\times 10^{-3}$. As the mean
density is unity, this is also the average total volume flux $J_v$.
The volume flux per unit area $J = J_v / a^2 L_z$, where $a^2 L_z$
is the volume of the discrete system (1664 in lattice units).
Following section \ref{section:exact_conductance},
we can write a viscosity independent conductance
$C = J\eta / \rho g$, with $g$ the force in the $z-$direction.
Putting all this together, we have $C = (1.947\times 10^{-3} / 1664)
\times (1/6) / (1.0 \times 1\times 10^{-7}) = 1.950$. The theoretical
figure is $C = 2$, so the simulation is correct to within a discretisation
error of about 3\%. Better accuracy may be acheived by increasing the
resolution of the circle; e.g., for $a = 8$, the result is 8.043 versus
an exact result of $C = 8$ --- an error of around 0.5\%.

\subsubsection{In parallel}

We can run the same calculation using the parallel code (recompile
with \texttt{make mpi}). In the input file we set
\begin{verbatim}
size 10_10_32
grid 1_1_2
\end{verbatim}
to set the parallel decomposition explicitly to two MPI tasks in the
$z-$direction. We also set
\begin{verbatim}
porous_media_format BINARY_SERIAL
porous_media_file   capillary
porous_media_type   status_only
\end{verbatim}
to ensure the structure file is read correctly in parallel.
We should run this on 2 MPI tasks, e.g.,
\begin{verbatim}
$ mpirun -np 2 ./Ludwig.exe input
\end{verbatim}
The final result should be exactly the same as the serial version
to machine accuracy. You may also be able to spot that the execution
time is approximately half that for the serial version (although this
problem is a little small for efficient parallelisation).

\subsubsection{General porous media}

For a general porous media, we can make use of Darcy's Law which
defines a permeability $k$ (with dimensions of area) via
\begin{equation}
J_v = - \frac{k A}{\eta} \frac{\partial P}{\partial z}
\end{equation}
where $J_v$ is the volume flux, $A$ is the cross-sectional area
of the sample, and $\partial P / \partial z$ is the pressure gradient
as before.

\subsubsection{Matching lattice and real units in porous media}

Following Succi \cite{succi} (Chapter~8) the following argument
can be made to match lattice quantities and real physical
quantities for a given system of interest.
Dimensionless lattice Boltzmann units set the lattice spacing
$\Delta x = 1$, the lattice time step $\Delta t = 1$ and the
mean density of the fluid to $\rho_0 = 1$. The speed of sound
in lattice units is $c_s = 1/ \sqrt{3}$. How do we match these
units to physical ones?

Suppose we have discretised an X-ray tomography image with a resolution
of 1 voxel equal to 1 $\mu$m. If we represent one voxel by one cubic
lattice with width $\Delta x$, then we can match $\Delta x = 1\mu$m. A
large data set might provide 1000 voxels on a side, giving
1000 lattice sites, which would represent 1 mm.
Similarly, if our real system is water at room temperature with density
$\rho = 1000$kg~m$^{-3}$, then we may equate the lattice density
$\rho_0 = 1$ to be equivalent to 1000~kg~m$^{-3}$. (This means the
mass of fluid at one lattice site of volume $\Delta x^3 = 1 \mu$m$^{3}$
is then 10$^{-15}$~kg, although this is not a very useful number.)

This leaves time. This is matched via the speed of sound. For the real
system, we take the speed of sound in water at 20$^o$C to be
1480 m~s$^{-1}$. So the speed of sound on the lattice
$(1/\sqrt{3}) \Delta x$ per $\Delta t$ matches 1480~m~s$^{-1}$,
or $\Delta t = 1/\sqrt{3} \times 10^{-6} / 1480 \approx 3.9 \times 10^{-10}$s.
Note that this is a small unit, i.e., it would require very many time
steps to reach a `macroscopic' time in any simulation.

We should also consider the Reynolds number $Re = \rho U L / \eta$.
Suppose our real structure has a characteristic pore size which is
just at the resolution of the X-ray tomography image: $L = h = 1\mu$m.
If the fluid (water) has viscosity $\eta = 10^{-3}$ Pa s, and
a typical flow is $U \sim 10^{-3}$ m~s$^{-1}$, then the pore scale
Reynolds number is $Re_h \sim 10^{-3}$. In the LB, if we choose a
lattice viscosity $\eta = 1/6$, and generate a typical flow in
lattice units of $10^{-4}$, then the Reynolds number in the
simulation (with $\rho_0 = 1$ and $L = \Delta x = 1$) is
$Re \sim 6 \times 10^{-4}$, which is similar.

There is one potential problem if we wish to study transport
phemonema where we might want to run a simulation long enough
for the flow to cross the whole sample. With $U \sim 10^{-4}$
in lattice units and a large sample size of, say,  1024$^3$ the
flow would take around 10$^7$ simulation time steps to propagate
information across the system. This is computationally expensive.
One solution is
to artificially raise the flow speed (and hence the Reynolds
number). This is acceptable as long as the Reynolds number
remains small compared with unity (all Reynolds numbers $< 0.1$
can be regarded as neglible \cite{batchelor}.) Here, for example,
if we raise the flow by two orders of magnitude to $10^{-2}$ in
lattice units,
the Reynolds number is still acceptable at $Re = 6\times 10^{-2}$,
and the number of simulation time steps is reduced to a more
managable $10^5$ (see also \cite{cates_scaling}).

### 2014-09-15 tags/2014-pre-release/src-alan/blue_phase_init.c

@@ -1,0 +1,1115 @@
+/*****************************************************************************
+ *
+ *  blue_phase_init.c
+ *
+ *  Various initial states for liquid crystal blue phases.
+ *
+ *  See, for example, Henrich et al, ...
+ *  and references therein.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Oliver Henrich (o.henrich@ucl.ac.uk) wrote most of these.
+ *  (c) 2012 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <math.h>
+#include <stdlib.h>
+
+#include "pe.h"
+#include "util.h"
+#include "coords.h"
+#include "phi.h"
+#include "blue_phase.h"
+#include "blue_phase_init.h"
+#include "ran.h"
+
+static double amplitude0_ = 0.0; /* Magnitude of order (initial) */
+
+/*****************************************************************************
+ *
+ *  blue_phase_init_amplitude
+ *
+ *****************************************************************************/
+
+double blue_phase_init_amplitude(void) {
+
+  return amplitude0_;
+}
+
+/*****************************************************************************
+ *
+ *  blue_phase_init_amplitude_set
+ *
+ *****************************************************************************/
+
+void blue_phase_init_amplitude_set(const double a) {
+
+  amplitude0_ = a;
+
+  return;
+}
+
+
+/*****************************************************************************
+ *
+ *  blue_phase_O8M_init
+ *
+ *  BP I using the current free energy parameter q0
+ *
[…]

### 2014-09-15 tags/2014-pre-release/src/phi_cahn_hilliard.c

@@ -1,0 +1,698 @@
+/*****************************************************************************
+ *
+ *  phi_cahn_hilliard.c
+ *
+ *  The time evolution of the order parameter phi is described
+ *  by the Cahn Hilliard equation
+ *
+ *     d_t phi + div (u phi - M grad mu - \hat{xi}) = 0.
+ *
+ *  The equation is solved here via finite difference. The velocity
+ *  field u is assumed known from the hydrodynamic sector. M is the
+ *  order parameter mobility. The chemical potential mu is set via
+ *  the choice of free energy.
+ *
+ *  Random fluxes \hat{xi} can be included if required via
+ *  phi_fluctuations.c with variance 2 M kT.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdlib.h>
+#include <math.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "leesedwards.h"
+#include "advection.h"
+#include "advection_bcs.h"
+#include "lattice.h"
+#include "free_energy.h"
+#include "phi_fluctuations.h"
+#include "phi.h"
+
+extern double * phi_site;
+static double * fluxe;
+static double * fluxw;
+static double * fluxy;
+static double * fluxz;
+
+static void phi_ch_diffusive_flux(void);
+static void phi_ch_update_forward_step(void);
+static void phi_ch_le_fix_fluxes(void);
+static void phi_ch_le_fix_fluxes_parallel(void);
+static void phi_ch_random_flux(void);
+
+static void phi_ch_diffusive_flux_noise(void);
+
+static double mobility_  = 0.0; /* Order parameter mobility */
+
+/*****************************************************************************
+ *
+ *  phi_cahn_hilliard
+ *
+ *  Compute the fluxes (advective/diffusive) and compute the update
+ *  to phi_site[].
+ *
[…]

### 2014-09-15 tags/2014-pre-release/src/stats_free_energy.c

@@ -1,0 +1,635 @@
+/****************************************************************************
+ *
+ *  stats_free_energy.c
+ *
+ *  Statistics for free energy density.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2011 The University of Edinburgh
+ *
+ ****************************************************************************/
+
+#include <assert.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "control.h"
+#include "colloids.h"
+#include "colloids_Q_tensor.h"
+#include "phi.h"
+#include "phi_gradients.h"
+#include "site_map.h"
+#include "wall.h"
+#include "free_energy.h"
+#include "stats_free_energy.h"
+#include "blue_phase.h"
+#include "util.h"
+
+static void stats_free_energy_wall(double * fs);
+static void stats_free_energy_wallx(double * fs);
+static void stats_free_energy_wally(double * fs);
+static void stats_free_energy_wallz(double * fs);
+static void stats_free_energy_colloid(double * fs);
+
+static int output_to_file_  = 1; /* To stdout or "free_energy.dat" */
+
+/****************************************************************************
+ *
+ *  stats_free_energy_density
+ *
+ *  Tots up the free energy density. The loop here totals the fluid,
+ *  and there is an additional calculation for different types of
+ *  solid surface.
+ *
+ ****************************************************************************/
+
+void stats_free_energy_density(void) {
+
+  int ic, jc, kc, index;
+  int nlocal[3];
+
+  double fed;
+  double fe_local[5];
+  double fe_total[5];
+  double rv;
+
+  double (* free_energy_density)(const int index);
+
+  coords_nlocal(nlocal);
+  free_energy_density = fe_density_function();
+
[…]

### 2014-09-24 branches/testcharge/htdocs/binary.tex
sync merge trunk to testcharge
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2478 40876441-3496-0410-a613-a2fca35b2ae2

--- 2010-11-29 htdocs/sigma.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  sigma.tex
-%
-%  A rather long section on the calibration of surface tension
-%  for the binary fluid models.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Binary Fluids}
 
-\section{Calibration of the Surface Tension}
+\subsection{The Cahn-Hilliard Equation}
 
-The binary fluid model introduces an interfacial tension $\sigma$ which
-can be predicted from the choice of free energy \cite{swift,kendon}.
-In particular, Kendon et al. 2000 described the process of calibrating
-the actual interfacial tension of the model fluid. This was based on
-a measurement of the Laplace pressure difference between the fluid
-inside and outside a spherical droplet via
-\begin{equation}
-\Delta p = 2 \sigma / R
-\end{equation}
-where $R$ is the radius of the spherical drop at equilibrium. This
-measurement appears quite difficult in practice owing to tendency of
-any initial drop to evaporate \cite{yue2007}.
+\subsection{The Choice of Free Energy}
 
-\subsection{New measurement}
+\subsection{A Second Distribution Function}
 
-Here, we prefer a measure based on the expected profile of the
-order parameter $\phi$ and the free energy density at a flat interface.
-In order to sample all orientations of the interface with respect to
-the lattice we again take a large droplet of fluid A initialised at
-rest in fluid B. We assume that the interface is locally flat and so
-the procedure holds.
+\subsection{The Collision}
 
-If a droplet with radius $r_0$ is constructed with an initial interfacial
-profile of $\phi(r) = \tanh{(r - r_0) / \xi_0}$
-with $r$ in the radial direction,  then the free energy density will be
-(with $-A = B$)
-\begin{equation}
- e(r) = -{\scriptstyle \frac{1}{4}} |A| [1 - \mathrm{sech}^4 (r-r_0)/\xi_0]
-+ {\scriptstyle \frac{1}{2}}
- (\kappa/\xi_0^2)\, \mathrm{sech}^4{(r-r_0)/\xi_0}.
-\end{equation}
-Obtaining a fit to $\xi_0$ from the measured interfacial profile of
-$\phi (r)$ and to $ e_0 = \kappa / 2 \xi_0^2$ from the peak in the measured
-profile of free energy
-density allows an estimate of the interfacial tension to be made:
-\begin{equation}
-\sigma = 2(8/9)^{1/2} \xi_0 |e_0|.
-\end{equation}
-This measurement is taken after the initial profile has been allowed
-to relax for at least 10 times the larger of the diffusion time
-$\xi_0^2 / M$ or the momentum diffusion time $\xi_0^2 / \eta$.
+\subsection{Bounce-Back on Links}
 
-We have measured the surface tension using the following approaches.
-
-\textit{Method 1}.
-
-LB for two distributions with relaxation of order parameter
-following Kendon etal. The force on the fluid is applied
-via the equilibrium stress following Swift et al.
-
-\textit{Method 2}.
-
-As method 1 but using the realxation of order parameter
-following Stratford etal.
-
-\textit{Method 3}.
-
-As Method 1 but with force via divergence of the chemical stress.
-
-\textit{Method 4}.
-
-As Method 3 but with force via divergence of the chemical stress.
-
-\textit{Method 5}.
-
-Using finite difference (1st order upwind) for the order parameter.
-
-\textit{Method 6}.
-
-Using finite difference (3rd order upwind).
-
-\subsection{Results}
-
-We follow the parameter sets in Kendon et al Table~2. These
-are reproduced here in Table~\ref{tab:r1}. Measured values for
-Run028 parameters are show in Table~\ref{tab:newr028}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{llllllll}
-\hline
-Run & $-A, B$ & $\kappa$ & $\eta$ & $M$ & $\sigma$  & $L_0$ & $T_0$ \\
-\hline
-Run028 & 0.083 & 0.053 & 1.41 & 0.05  & 0.0625 & 31.8 & 718 \\
-Run022 & 0.0625 & 0.04 & 0.5 & 0.25   & 0.0471 & 5.31 & 56.3 \\
-Run029 & 0.0625 & 0.04 & 0.2 & 0.15   & 0.0471 & 0.849 & 3.61 \\
-Run020 & 0.00625 & 0.004 & 0.025 & 2.00 & 0.00471 & 0.133 & 0.704 \\
-Run030 & 0.00625 & 0.004 & 0.0065 & 1.25 & 0.00471 & 0.00829 & 0.0124 \\
-Run019 & 0.003125 & 0.002 & 0.0014 & 4.00 & 0.00236 & 0.000831 & 0.000493 \\
-Run032 & 0.00125  & 0.0008 & 0.0005 & 5.0 & 0.000943 & 0.000265 & 0.000141\\
-\hline  
-\end{tabular}
-\end{center}
-\caption{Table showing common parameter values following Kendon et al.
-with the theoretical interfacial tension, $L_0$ and $T_0$. Note that
-the order parameter mobility is used here (cf $\tilde{M} = 2M$) in
-Kendon et al.}
-\label{tab:r1}
-\end{table}
-
-\begin{table}
-\begin{tabular}{lllllll}
-\hline
-Method & $\xi_0$ & $e_{min}$ & $e_{max}$ & $\sigma$  & $\phi_{min}$
-& $\phi_{max}$\\
-\hline
-\multicolumn{7}{c}{Run028}\\
-\hline
-Theory   & 1.13 & -0.0207 &  0.0207 & 0.0625 & -1.0 & 1.0 \\
-Method 1 & 1.00 & -0.0207 & 0.0128 & 0.063(3) & -1.0091 & 1.0028 \\
-Method 2 & 1.00 & -0.0207 & 0.0126 & 0.063(1) & -1.0113 & 0.9993 \\
-Method 3 & && & & & \\
-Method 4 & && & & & \\
-Method 5 & && & & & \\
-Method 6 & --& --& --& --& --& --\\
-\hline
-\multicolumn{7}{c}{Run029}\\
-\hline
-Theory & 1.13 & -0.0156 & 0.0156 & 0.0471  & -1.0 & 1.0\\
-Method 2 & 0.99 & -0.0156 & 0.010(?) & 0.048(4) & -1.0110 & 1.0003\\ 
-Method 6 & 1.02 & -0.0156 & 0.0099 & 0.049(0) & -1.0048 & 1.0002\\
-\hline
-\multicolumn{7}{c}{Run030}\\
-\hline
-Theory & 1.13 & -0.00156 & 0.00156 & 0.00471 & -1.0 & 1.0\\
-Method 2 & 1.04 & -0.00156 & 0.0008(?) & 0.0047(1) & -1.0167 & 1.0063\\
-Method 6 & 1.08 & -0.00156 & 0.0007(?) & 0.0048(0) & -1.0028 & 1.0031\\
-\hline
-\multicolumn{7}{c}{Run019}\\
-\hline
-Theory & 1.13 & -0.000781 & 0.000781 & 0.00236 & -1.0 & 1.0\\
-Method 2 & 1.29(?) & -0.000781 & 0.00015(?) & 0.0022(8) & -1.0312 & 1.0242\\
-Method 6 & 1.23(?) & 0.0010(?) &&  0.0024(?) & -1.0085 & 1.0019\\
-\hline 
-\multicolumn{7}{c}{Run032}\\
-\hline
-Theory   & 1.13 &  -0.000312 & 0.000312 & 0.000943  & -1.0 &1.0\\
-Method 2 & 1.07  & -0.000312& 0.000157 & 0.00094(7) & -1.0511 & 1.0311\\
-Method 6 & 1.6(5)& -0.000312 & not st. & 0.0010(7) &-1.0056 & 1.0010\\
-\hline
-\end{tabular}
-\caption{Results for the different methods for Run208 parameters.}
-\end{table}
-
-\section{Upwind Advection Schemes}
+\subsection{Upwind Advection Schemes}
 
@@
 
-\subsection{Uniformly third order scheme}
+\subsubsection{Uniformly third order scheme}
 
@@
 
-\subsection{Gory details}
+\subsubsection{Gory details}
 
@@
 
-\section{Lattice kinetic equation viewed as finite difference}
+\subsection{Lattice kinetic equation viewed as finite difference}
 
@@
 
-\subsection{Advective terms}
+\subsubsection{Advective terms}
 
@@
 
-\subsection{Dissipative terms}
+\subsubsection{Dissipative terms}
 
@@
 
-\subsection{Chemical potential term}
+\subsubsection{Chemical potential term}
 
@@
 properties. The final prefactor is $M$ (check).
+
+\subsection{User input}
+
+\inputkey{free\_energy symmetric}
+
+The following applies for the binary fluid problem with
+compositional order parameter $\phi$ and free energy
+(excluding the term in the density $\rho$):
+\begin{equation}
+ F[\phi] = 
+\int dr \left(
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
+\end{equation}
+
+This is described in some detail
+by Kendon et al \cite{viv}. The first two terms represent the bulk
+contribution, whereas the term in $\kappa$ penalises curvature in
+the interface.
+
+\inputkey{A} The parameter $A$. Note $A < 0$.
+
+\inputkey{B} The parameter $B$. Note $B = -A$ for common usage,
+although this is not enforced.
+
+\inputkey{K} The parameter $\kappa$, which is positive.
+
+The order parameter evolution is determined by a Cahn-Hilliard equation
+with mobility set by
+
+\inputkey{mobility} Sets mobility $M$ (uniform in space).
+
+The following parameters control the initialisation of the order
+parameter.
+
+\inputkey{phi\_initialisation}
+
+Determines how the compositional order parameter is initialised
+at the start of the run. If set to \texttt{spinodal} the value
+is set to $\phi_0$ as set above plus or minus a random noise, the
+magnitude of which is set by the value of the \texttt{noise} key.
+If set to \texttt{block} a one-dimensional profile is set up in
+the $z$-direction representing two blocks of fluid with
+$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
+$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
+appropriate width. The $\phi = -1$ section is in the middle.
+
+\inputkey{phi0}
+
+The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
+The default value is zero, i.e., a symmetric 50:50 mixture by volume.
+
+\inputkey{noise}
+
+The magnitude of the initial fluctuations in $\phi$ used to
+initiate spinodal decomposition.
+
+\subsubsection{Binary fluid using two distributions}
+
+\inputkey{symmetric\_lb}
+
+This is the special case where the composition is represented
+by a second LB distribution, and an appropriate lattice kinetic
+equation approximating the Cahn-Hilliard equation is solved.
+In this case the above parameters have the same meaning.
+
+Note that the relaxation time for the second distribution is
+related to the mobility by $tau_\phi = M\rho_0/\Delta t + 1/2$.
+
+\subsubsection{Brazovskii}
+
+\inputkey{free\_energy brazovskii}
+
+This is similar to the symmetric free energy, but with one extra term
+in a higher derivative of $\phi$.
+\begin{equation}
+ F[\phi] = 
+\int dr \left(
+{\textstyle \frac{1}{2}}A\phi^2
++ {\textstyle \frac{1}{4}}B\phi^4
++ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
++ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
+\end{equation}
+
+The parameters now include $C$. For $A<0$, phase separation occurs
+with a result depending on $\kappa$: ones get two symmetric phases
+for $\kappa >0$ (cf. symmetric) or a lamellar phase for
+$\kappa < 0$.
+
+\inputkey{A} Bulk parameter $A < 0$.
+
+\inputkey{B} Bulk parameter $B = -A$.
+
+\inputkey{K} Negative for lamellar phase.
+
+\inputkey{C} Positive.
+
+Other parameters, including the mobility, are set as for the
+symmetric free energy (see above).

### 2014-09-24 branches/testcharge/htdocs/lb.tex
sync merge trunk to testcharge
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2478 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-01-14 branches/CUDA/htdocs/sections.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  sections.tex
-%
-%  Content for Ludwig documentation.
-%
-%  $Id: sections.tex,v 1.15 2010-10-21 18:05:50 kevin Exp $
-%
-%  Edinburgh Soft Matter and Statistical Physics Group and
-%  Edinburgh Parallel Computing Centre
-%
-%  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{The Lattice Boltzmann Method}
@@
 
-\subsection{Colloids}
+\subsection{Model Details}
 
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
-\section{Polar Active Gel}
-
-Here we consider a polar vector field model to study active gels. 
-The order parameter is this time a vector $P_\alpha$
-(with variable magnitude) as there is no longer head-tail symmetry
-in the system as was the case for the ${\bf Q}$ tensor formulation. 
-The equation of motion as:
-\begin{equation}
-\partial_t P_\alpha  \partial_\beta (u_{\beta} + wP_{\beta}) P_{\alpha}
-= \lambda D_{\alpha\beta} P_{\beta} - \omega_{\alpha\beta} P_{\beta}
-+\Gamma' h_{\alpha}.
-\end{equation}
-
-In this equation, $w$ is an active term, due to swimming, which causes
-self-advection of the order parameter, while $\lambda$ is a material
-dependent constant -- positive for rod-like molecules. If $|\lambda|>1$
-the liquid crystalline passive phase is flow-aligning, otherwise it is
-flow-tumbling.
-
-Note that the convection for the Leslie-Ericksen is that the
-velocity gradient tensor should be writen $W_{\alpha\beta}
-= \partial_\alpha u_\beta$ (and not the other way round, which
-is usual).
-
-The molecular field is given by
-$h_{\alpha}=-\delta {\cal F}_{\rm pol}/\delta p_{\alpha}$
-where ${\cal F}_{\rm pol}$ is the free energy for a polar active nematic,
-whose density we can take to be:
-\begin{equation}
-f =  \frac{\alpha_2}{2} |{\mathbf P}|^2 + \frac{\alpha_4}{4}|{\mathbf P}|^4
-+ \frac{K}{2}\left(\partial_\alpha P_{\beta}\right)^2 + 
-\frac{K_{LC}}{2}\left[\partial_\alpha (P_{\beta}P_{\gamma})\right]^2 
-\end{equation}
-where $K$ and $K_{LC}$ are elastic constants, $\alpha_4>0$, and $\alpha_2$ 
-can be positive (isotropic phase) or negative (polar nematic phase).
-
-The molecular field is explicitly given by:
-\begin{equation}
-h_{\alpha}=-\alpha_2 P_{\alpha} -\alpha_4 P_{\alpha} P_{\beta}P_{\beta}+
-K \partial_{\beta}\partial_{\beta} P_{\alpha}+2 K_{LC}
-P_{\gamma}\partial_{\beta}\partial_{\beta} (P_{\gamma}P_{\alpha}).
-\end{equation}
-
-The Navier-Stokes equation is:
-\begin{equation}
-\rho \left(\partial_t+v_{\beta}\partial_{\beta}\right) v_{\alpha} =
-\partial_{\beta} \Pi_{\alpha\beta} 
-+\eta \partial_{\beta}\partial_{\beta} v_{\alpha}.
-\end{equation}
-
-The stress tensor this time is 
-\begin{eqnarray}
-\Pi_{\alpha\beta} =
-\frac{1}{2}\left(P_\alpha h_\beta - P_\beta h_\alpha \right)
-- \lambda\left( (P_\alpha h_\beta + P_\beta h_\alpha)/2 
-- (1/3)P_\gamma P_\gamma \delta_{\alpha\beta} \right)
-\\
--  \zeta ( P_\alpha P_\beta - (1/3)P_\gamma P_\gamma \delta_{\alpha\beta})
-- K \partial_\alpha P_\gamma \partial_\beta P_\gamma.
-\end{eqnarray}
-The active term in the stress tensor, proportional
-to $\zeta$, has the same form for apolar and polar active gels.
-
-\section{Code Design}
-
-\subsection{Some Comments}
-
-In designing a code in C the main object of which is to do high
-performance computing, there are some inevitable tensions with
-software engineering best-practice. In particular, complete
-encapsulation of data in the spirit of using abstract data types
-tends to cause problems of performance.
-
-For example, encapsulation of the lattice Boltzmann distribution
-$f_i$, together with appropriate ``get'' and ``set'' functions for
-access when approapriate, is possible. There could then be a
-complete separation of interface and implementation of the
-distribution (other than that each individual $f_i$ is represented
-by a \texttt{double}). However, experience suggests that
-this imposes severe impediment to optimisation and hence
-unacceptable performance degradation.
-
-Schemes do exist by which one can approximate the object
-oriented programming style in C, and so achieve a separation
-of interface and implementation. However, this tends to put
-a heavy burden on the use of \texttt{void} pointers, itself
-a recipe for possible disaster. A truly object-oriented
-language such as Java would be preferable.
-
-So, for the time being, some relaxation of encapsulation is required.
-Exposing a pointer to the distributions which allows them to be accessed
-directly where required. This can be done merely by declaring the
-necessary variables \texttt{extern}. In practice, such deviations
-from encapsulation can be minimised in the lattice sector of the
-code. In particlar, \texttt{extern} declarations should not be
-required in \texttt{main()}, i.e., it should be able to write a
-new application without worrying about details of implementation.
-
-\subsubsection{Objects}
-
-Somewhat more difficult is the part of the code involving
-colloids, which would really favour a truly object-oriented
-approach.
-
-\subsubsection{Serial and parallel}
-
-The design decission has been taken to support execution of the
-code in both serial and parallel. The default compilation
-assumes that a serial executable is required, while  MPI code
-is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).
-
-There is no shared memory inplementation at this time.
-
-\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}
-
-\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}
-
-\begin{figure}
-\begin{center}
-%\includegraphics{xfig/sketch.eps}
-\end{center}
-\caption{A scheme by which different parts of the code might be
-encapsulated and in such a way to allow reliable development and
-testing. As this is C, will do not adopt a truly ``object'' langauge.
-Broadly, different modular parts should only depend on stuff above
-in the diagram. So, at the top, there is the parallel environment
-where communication occurs within MPI\_COMM\_WORLD. Built on top
-of this is the Cartesian Communicator responsible for halo swaps
-and so forth. The LE code may have a separate Communicator again. 
-(There's an argument to say the diagram should be the other way up!)}
-\end{figure}
-
-\section{Code Implementation}
-
-\subsection{Stuff}
-
-\subsection{Coordinate System}
-
-\subsubsection{The global coordinate system}
-
-The user specifies the system size in lattice units at run time,
-which sets the extent of memory which must be allocated for the
-Cartesian lattice. The extent of the physical coordinate system
-then reflects the lattice size directly (see Figure \ref{fig_c1}). 
-Each lattice site, or node,  has coordinates corresponding to its
-integer position in the system and can be thought of as being
-at the centre of a control volume one lattice unit on a side.
-This means that the nominal edges of the
-system are offset from the lattice sites themselves by half a
-lattice unit in each direction.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_c1.eps}
-\end{center}
-\caption{The global Cartesian coordinate system reflects the choice
-of lattice size; lattices sites are located at
-$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
-(the Figure only shows the labels in the $y-$direction for simplicity).
-In the Figure, the lattice sites in the domain proper are the full
-circles; halo sites, which are used to effect periodic boundary
-conditions, are included as open cirles}
-\label{fig_c1}
-\end{figure}
-
-For particles, the centres of which can move continuously across the
-lattice, the extents of the global coordinate system are then
-$L_{min} < x < L + L_{min}$ and so on.
-
-\subsubsection{Addressing the lattice: the local coordinate system}
-
-\subsection{Solid walls}
-
-General solid objects which are fixed for the duration of
-the simulation, such as porous media, can be defined
-by assigning solid/fluid status to each lattice node. 
-The links required between solid and fluid nodes can
-then be identified for the given model.
-
-\textit{Comment.} In general, a solid object will not
-affect the periodic nature of the simulation. A general
-solid object cannot move, i.e., the boundary velocity
-must be zero.
-
-\textit{Special case.} In the case where the solid object
-represents a flat boundary wall at the edge of the system
-in one or more directions, the periodicity is broken in
-the corresponding directions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LE Implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Lees Edwards Sliding Period Boundary Conditions}
-
-\subsubsection{The LE module}
-
-\subsubsection{Distributions crossing the LE planes}
-
-With the planes conceptually half-way between lattice sites, the
-propagation moves some of the distributions (namely, those with
-$c_x = \pm1$) between different sliding blocks. While the collision
-stage is unaffected, action must be taken to adjust these
-distributions when the LE boundary conditions are active. This
-is done in a two-stage process described by Adhikari \textit{et al.}
-\cite{adhikari_desplat}, implemented between the collision and
-propagation.
-
-\textbf{a. Reprojection} The post-collision distributions with
-$c_x = \pm 1$ at sites adjacent to a boundary are modified to
-take account of the velocity jump $\pm u^{LE}_y$ between the sliding
-blocks. In terms of the hydrodynamic moments we have:
-\begin{eqnarray}
-\rho &\rightarrow& \rho', \\
-\rho u_\alpha &\rightarrow& (\rho u_\alpha)' \pm \rho' u^{LE}_\alpha, \\
-S_{\alpha\beta} &\rightarrow&
-S'_{\alpha\beta} \pm (\rho u_\alpha)' u^{LE}_\beta
-\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
-\end{eqnarray}
-If we work with the changes to the moemnts, then the density is
-unaffected $\delta\rho =  0$, the velocity is changed by
-$\delta u_\alpha = \pm u^{LE}_\alpha$, with an analogous expression for
-the change in the stress
-\begin{equation}
-\delta S_{\alpha\beta} =
-\pm (\rho u_\alpha)' u^{LE}_\beta
-\pm (\rho u_\beta)' u^{LE}_\alpha + \rho' u^{LE}_\alpha u^{LE}_\beta.
-\end{equation}
-We can then work out the change in the distributions by reprojecting
-via Eq.\ref{eq:}, i.e.,
-\begin{equation}
-f_i \rightarrow f_i' + w_i \Bigg(
-\frac{\rho \delta u_\alpha c_{i\alpha}}{c_s^2} +
-\frac{\delta S_{\alpha\beta}Q_{i\alpha\beta}}{2c_s^4} \Bigg).
-\end{equation}
-A similar set of expressions are given for the order parameter
-distributions in Adhikari \textit{et al.} \cite{adhikari_desplat}.
-
-\textbf{b. Interpolation of reprojected distributions}
-As the sliding blocks are displaced continuously with
-$\delta y = u_y^{LE}t$, which is not in general a whole number of
-lattice spacings, an interpolation is also required before
-propagation can take place. Consider the situation (Fig.~\ref{fig_le2}) from
-a stationary frame where the frame above has translated a small
-distance to the right ($\delta y < \Delta y$). In the stationary frame
-we must interpolate the ditributions with $c_x = 1$ to a
-`departure point' from which the propagation will move the
-distribution exactly to the appropriate lattice site in the moving
-frame. Schemeatically,
-\begin{equation}
-f_i'(x, y + \delta y, z) = 
- (1 - \delta y) f_i^\star(x, y, z) +
-\delta y f_i^\star(x, y + \Delta y, z),
-\end{equation}
-where $f_i'$ here represents the interpolated value and $f_i^\star$
-is the reprojected post-collision quantity.
-
-In the case where the displacement $\delta y > 1$, the relevant
-lattice sites involved in the interpolation are displaced to the right
-by the integral part of $\delta y$, while the relative weight given to
-the two sites involves
-the fractional part of $\delta y$. To take account of the periodic
-boundary conditions
-in the $y$-direction, all displacements are modulo $L_y$.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_le2.eps}
-\end{center}
-\caption{Schematic diagram showing interpolation of distributions in the
-stationary plane (below) in preparation for propagation into the moving
-frame (above). For a small displacement $\delta y = u^{LE} t$, the
-interpolation takes place between lattice sites in the stationary frame
-(green, left). Propogation from the interpolated positions (red) can
-then take place into the moving frame (right).}
-\label{fig_le2}
-\end{figure}
-
-In parallel, exactly the same calculation is undertaken. However,
-communication is required to take account of the processor
-decomposition in the $y$-direction. This requires the identification
-of the two processes which hold the relevant information as a function
-of displacement.
-
-Note that the interpolated values are stored in the appropriate lattice
-site in the rest frame that ensures the normal propagation stage moves all
-distributions to the correct destination in memory.
-
-\subsubsection{Order parameter gradients}
-
-When the order parameter gradient is computed at a given point
-near the LE boundaries, the stencil may extend out of the stationary
-frame. The process of interpolation is slightly different to that
-for the distributions.
-
-Here, we need to interpolate values in the moving frame to the position
-which is seen by the rest frame. Figure~\ref{fig_le3} shows a five-point
-stencil to calculate the gradient at the central point $(x,y,z)$ in the
-rest frame. An interpolation of the $\phi$-field of the form
-\begin{equation}
-\phi'(x + \Delta x, y - \delta y, z) =
-\delta y \phi(x + \Delta x,  y - \Delta y, z) + 
-(1 - \delta y) \phi(x + \Delta x, y, z)
-\end{equation}
-must then take place. In practice, the interpolated values are stored
-in the offset buffer. As the gradient calculation requires $\phi$
-values in the halo regions, interpolated values for the buffer halo
-region are also required.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_le3.eps}
-\end{center}
-\caption{Schematic diagram showing a five-point stencil to compute a
-gradient at the central point in the stationary frame. The upper point
-of the stencil is in the moving frame and so must be interpolated to
-the correct position (red sites).}
-\label{fig_le3}
-\end{figure}
-
-In parallel, the interpolation requires data from at most two adjacent
-processes in the along-plane $y-$direction as long as the $\phi$ halo
-regions are up-to-date. The rank of the processes involved is
-determined by the displacement $\delta y$ as a function of time.
-
-\subsubsection{Velocities and finite-difference fluxes}
-
-For the finite-difference implementation, the velocity field at
-the cell boundaries is required. At the planes, this means an
-interpolation which takes the same form as that for
-$\phi$ is needed.
-
-In computing the advective fluxes between cells, we must then
-allow for the presence of the planes. This is done by computing
-separately the 'east' and 'west' face fluxes in the $x$-direction
-(the flux that crosses the plane). Away from the planes, face-flux
-uniqueness $f_{ijk}^e = f_{i+1jk}^w$ ensures conservation of order
-parameter. However, at the planes, the non-linear conbination of
-velocity field and order parameter field does not in general give
-rise to consistent fluxes. In order to restore conservation, we
-reconcile east and west face fluxes at the plane by taking an
-average
-\begin{eqnarray}
-f_{ijk}^{e\star}&=&{\textstyle\frac{1}{2}}(f_{ijk}^e + f_{i+1j'k}^w),\quad 
-f_{i+1j'k}^w  = \delta y f_{i+1 j-1 k}^w + (1 - \delta y) f_{i+1jk}^w,\\
-f_{i+1 j k}^{w\star} &=& {\textstyle\frac{1}{2}}(f_{ij'k}^e + f_{i+1jk}^w),\quad
-f_{ij'k}^e = (1 - \delta y) f_{ijk}^e + \delta y f_{ij+1k}^e.
-\end{eqnarray}
-In this way, we average the east face flux in the rest frame $f_{ijk}^e$
-with the interpolated value of the west face flux in the moving frame
-$f_{i+1j'k}^w$, and vice-versa. One can then easily show that when
-integrated over the length of the plane, the average fluxes are
-consistent, i.e.,
-\begin{equation}
-\sum_j (f_{ijk}^{e\star} - f_{i+1jk}^{w\star}) = 0 \quad\forall k,
-\end{equation}
-thus ensuring order parameter conservation.
-
-A similar correction is required when computing the force on the fluid
-directly via the divergence of the thermodynamic stress
-$F_\alpha = \nabla_\beta P_{\alpha\beta}^{th}$. This is implementated
-by computing fluxes of momentum at cell faces in each direction, and
-then taking the divergence to compute the force. At the planes, an
-averaging procedure is again used to ensure conservation of momentum.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% End LE implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\vfill
-\pagebreak
-
-\section{Code Interface}
-
-\subsection{The Parallel Environment}
-
-\texttt{\#include ``pe.h''}
-
-As the code is to be available in both serial and parallel versions,
-it is desirable to abstract the basic environment to some extent.
-The parallel environment is intended to support MPI communication
-within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
-consistent picture in serial (i.e., rank zero is the single process).
-
-\texttt{void pe\_init(int argc, char ** argv)}\\
-Responsible for initialisation of MPI, and hence must be first
-execuatable statement. The arguments are those to \texttt{main}
-which will be passed through to \texttt{MPI\_Init()}.
-
-\texttt{int pe\_rank(void)}\\
-Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
-serial.
-
-\texttt{int pe\_size(void)}\\
-Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
-1 in serial.
-
-\texttt{void pe\_finalise(void)}\\
-Responsible for \texttt{MPI\_Finalize()}, and must be the last
-executable statement.
-
-\texttt{void info(const char * fmt, ...)}\\
-Behaves like \texttt{printf}, but only produces output on the
-root process.
-
-\texttt{void verbose(const char * fmt, ...)}\\
-Behaves as \texttt{printf} and reports the rank of the
-calling process in \texttt{MPI\_COMM\_WORLD}.
-
-\texttt{void fatal(const char * fmt, ...)}\\
-Prints an error message and identifies
-the rank of the calling process before terminating the program.
-
-\subsection{Run Time Input}
-
-\texttt{\#include ``runtime.h''}
-
-The ability to set a wide range of parameters at run time is
-very convenient, and is supported via an input file in which the
-user specifies parameters as key value pairs. The code maintains
-a list of key value pairs from which the user can retrieve the
-appropriate values for given keys.
-
-\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
-This reads the named input file and sets up the list of keys.
-
-\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The number
-of key matches found is returned (i.e., 0 or 1).
-
-\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The
-number of key matches is returned.
-
-\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
-Set \texttt{value} to the string assocaited with \texttt{key}, if
-present. Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
-If \texttt{key} is present, set the associated 3-vector of integers.
-Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
-double v[])}\\
-If \texttt{key} is present, set the associated 3-vector of doubles.
-Returns the number of key matches found.
-
-\textit{Comment:} We probably need a function to report unused keys, so
-the user can be alerted if a key has unexpectedly been ignored.
-
-\subsection{Random Number Generators}
-
-\texttt{\#include ``ran.h''}
-
-Note that the random number generation in parallel is
-decomposition dependent.
-
-\texttt{void rand\_init(void)}\\
-Responsible for initialising RNG state.
-
-\texttt{double rand\_serial\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$ from
-the serial generator.
-
-\texttt{double rand\_serial\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the serial generator.
-
-\textit{double rand\_parallel\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$
-from the serial generator.
-
-\texttt{double rand\_parallel\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the parallel generator.
-
-\textit{Comment:} The serial generator will produce the
-same sequence on any MPI process, while the parallel generator
-will produce different sequences. The latter is intended for
-generating, specifically, fluctuations, where the need for
-decomposition-independance is questionable.
-
-\subsection{Timers}
-
-\texttt{\#include ``timer.h''}
-
-Performance figures require timing for different sections
-of the code. The code provides a set of routines to time
-standard and user-defined sections of code. Timers are
-started and stopped by call to the appropriate routines.
-
-\texttt{TIMER\_init()}\\
-This initialises all the timers to zero (e.g., at the start of
-the program).
-
-\texttt{TIMER\_start(const int t\_id)}\\
-Start the given timer.
-
-\texttt{TIMER\_stop(const int t\_id)}\\
-Stop the given timer.
-
-\texttt{TIMER\_statistics(void)}\\
-Prints the current state of all the active timers to standard
-output on the root process in a digestible form. In serial,
-the minimum, maximum, and mean time for each timer is reported
-in seconds (based on the number of start/stop cycles). In
-parallel, the same figures are computed on each process and the
-result of the corresponding \texttt{MPI\_Reduce()} operation is
-printed on the root process.
-
-The code provides a number of timers which would time standard
-parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
-total execution time, and a number of spare timers for users to
-use at will.
-
-\textit{Comment:}
-No intra-run variance (i.e., between time steps) is provided, as the
-calculation required is considerably more complex to generalise, and
-requires memory proportional to the number of time steps used.
-
-\textit{Comment:}
-Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
-provided to insert calls to the above functions for optional timing
-of performance-sensitive regions
-of the code in conjunction with the preprocessor flag
-\texttt{-D\_TIMING\_MACRO\_ON\_}.
-
-\subsection{Model Time Control}
-
-This provides time step control, event frequency (?)
-
-\texttt{void init\_control(void)}\\
-Initialise time controls.
-
-\texttt{int get\_step(void)}\\
-Return the current model time step.
-
-\texttt{int next\_step(void)}\\
-Increments the current time step. Returns number of steps left
-to run (i.e., 0 when finished).
-
-\subsection{Coordinate System}
-
-\texttt{\#include ``coords.h''}
-
-A three-dimensional Cartesian coordinate system is used throughout,
-with the exact system size usually chosen by the user at run time.
-The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
-used to identify the coordinate directions.
-
-This also provides access to the MPI Cartesian Communicator which
-is responsible for most communication in the model.
-
-\texttt{int N\_total(const int dim)}\\
-Returns the total number of lattice sites in the given Cartesian direction.
-
-\texttt{double L(const int dim)}\\
-Returns the length of the system in the given Cartesian direction
-(which is the same as the number of lattice sites).
-
-\texttt{double Lmin(const int dim)}\\
-Returns coordinate position of left-hand edge of system in given
-Cartesian direction.
-
-\texttt{int is\_periodic(const int dim)}\\
-Returns 1 if given Cartesian direction is periodic, else 0.
-
-\texttt{void cart\_init(void)}\\
-Responsible for setting up the Cartesian decomposition.
-
-\texttt{int cart\_rank(void)}\\
-Returns rank in the Cartesian communicator.
-
-\texttt{MPI\_Comm cart\_comm(void)}\\
-Returns the MPI Communicator handle for the Cartesian communicator.
-
-\texttt{int cart\_size(const int dim)}\\
-Returns the size of the Cartesian communicator grid in the given
-coordinate direcxtion.
-
-\texttt{int cart\_coord(const int dim)}\\
-Returns the coordinate of the current process in the Cartesian
-Communicator.
-
-\texttt{int cart\_neigh(const int dir, const int dim)}\\
-Returns the rank of the neighbouring process in the
-Cartesian communicator.
-
-\subsection{Colloids}
-
-\texttt{\#include ``colloids.h''}
-
-These functions provide the basic interface for use
-of colloidal particles in the code.
-
-\subsubsection{Colloid structure}
-
-The basic colloid struct is defined ...
-
-\subsubsection{Colloid interface}
-
-\texttt{void colloids\_init(void)}\\
-Call once for initialisation. This initialises the cell list,
-so the coordinate system must be initialised via a call to
-\texttt{coords\_init()} first.
-
-\texttt{void colloids\_finish(void)}\\
-Call once for finalisation. Frees all colloid-related memory.
-
-\texttt{Colloid * allocate\_colloid(void)}\\
-Returns a pointer to a newly allocated \texttt{Colloid} object
-or fails gracefully.
-
-\texttt{void free\_colloid(Colloid * p\_colloid)}\\
-Deallocates all memory associated with \texttt{p\_colloid} or
-fails gracefully.
-
-\texttt{struct c\_link allocate\_boundary\_link(void)}\\
-Returns a pointer to a newly allocated colloid boundary
-link or fails gracefully.
-
-\texttt{void free\_boundary\_link(struct c\_link *)}\\
-Deallocates memory associated with a link or fails gracefully.
-
-\texttt{void colloids\_memory\_report(void)}\\
-The module keeps track of how much memory has been allocated in
-total for colloids and boundary links. This reports the current
-usage.
-
-\texttt{get\_Ncell\_total()}\\
-
-\texttt{get\_Ncell\_local()}\\
-
-\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
-This returns a pointer to the Colloid at the head of the list at
-cell list location (i, j, k) or \texttt{NULL} if there is none.
-
-\texttt{void cell\_insert(Colloid * p\_colloid)}\\
-This inserts a colloid at the head of the appropriate list
-depending upon its position. This is the only way a colloid
-can be added to the list.
-
-\texttt{update\_cells(void)}\\
-Performs the update of the cell list in response to updated
-particle positions.
-
-\texttt{IVector cell\_coords(FVector r)}\\
-
-\subsection{Speculation}
-
-\subsection{The Free Energy}
-
-It is imagined that a single file will encapsulate all the details
-of a given free energy. The interface for different choices of
-free energy should be the same.
-
-\texttt{FE\_init()}
-
-Responsible for querying the runtime input key value pair list to
-find appropriate user parameters for the free energy, or
-falling back to default values.
-
-\texttt{double FE\_get\_sigma()}
-
-Return the surface tension. There should be a similar routine
-for the interfacial thickness.
-
-\texttt{double FE\_get\_mu(const type location)}
-
-Return the chemical potential at the single lattice
-node \texttt{location}.
-
-\textit{Comment:} The computation of quantities such as the chemical
-potential will a) affect performance and b) require halo values of
-the order parameter (depending on the highest derivative
-in the order parameter present). Performance here should be the
-first consideration, so this needs to be handled with care.
-
-\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}
-
-Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
-given location.
-
-\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
-the number of dimensions. If we are to provide explicit support for
-D2Q9, \texttt{ND} will be 2, otherwise 3.
-
-\section{Implementation Notes}
-
-\subsection{Serial Version}
-
-\subsection{Efficiently Scaling Parallel Computing}
-
-We should be thinking about the 100,000 processor regime, and
-correspondingly large system sizes. To avoid embarrassment in this regime,
-some issues which don't usually arise are relevant, for example:
-
-\begin{enumerate}
-\item
-Large integer quantities such as
-\begin{verbatim}
-   N_total.x*N_total.y*N_total.z
-\end{verbatim}
-are likely to cause integer overflow (if default integers are 4 byte).
-\item
-Local data structures whose size is proportional
-to the number of processors are contraindicated, e.g.,
-anything like
-\begin{verbatim}
-p = (foo *) malloc(nprocs*sizeof(foo));
-\end{verbatim}
-is to be avoided. At the moment, we are in good shape on this
-(apart from some functions in \texttt{misc.c}).
-\item
-I/O is likely become a very important issue, so may need to think
-more about this.
-\item
-We should probably introduce a standard set of benchmarks which can
-be scaled up in some systematic way, so that we can keep track of
-performance over time.
-
-\end{enumerate}
-
-\section{Model Details}
-
-\subsection{D2Q9}
+\subsubsection{D2Q9}
 
@@
 
-\subsection{D3Q15}
+\subsubsection{D3Q15}
 
-These are the eigenvalues and eigenvectors of teh collision
+These are the eigenvalues and eigenvectors of the collision
 matrix used for D3Q15
@@
 
-\subsection{D3Q19}
+\subsubsection{D3Q19}
 
@@
 \pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{adhikari_desplat}
-Adhikari, R. J.-C. Desplat, and K. Stratford,
-Sliding periodic boundary conditions for lattice Boltzmann and lattice
-kinetic equations,
-\texttt{arXiv:cond-mat/0503175v1} (2005).
-
-\bibitem{adhikari2005}
-Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
-Fluctuating Lattice Boltzmann,
-\textit{Europhys. Lett.}, \textbf{71}, 473 2005.
-
-\bibitem{ald98}
-Aidun, C.K., Y. Lu, and E.-J. Ding,
-Direct analysis of particulate suspensions with inertia using the
-discrete Boltzmann equation,
-\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
-
-\bibitem{cates_scaling}
-M. E. Cates, J.-C. Desplat, P. Stansell, A.J. Wagner, K. Stratford,
-R. Adhikari, and I. Pagonabarraga,
-Physical and Computational Scaling Issues in Lattice Boltzmann
-Simulations of Binary Fluid Mixtures,
-\textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
-
-\bibitem{chunladd}
-B. Chun and A.J.C. Ladd,
-Interpolated boundary condition for lattice Boltzmann simulations in
-narrow gaps,
-\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
-
-\bibitem{cj98}
-Cichocki, B., and R.B. Jones,
-Image representation of a spherical particle near a hard wall,
-\textit{Physica A}, \textbf{258}, 273, 1998.
-
-\bibitem{chang}
-C.-H. Chang and E.I. Franses,
-Adsorption dynamics of surfactants at the air/water interface:
-a critical review of mathematical models, data, and mechanisms,
-\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
-
-\bibitem{cpc}
-Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
-LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
-\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
-
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
-\bibitem{diamant96}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid-fluid interfaces,
-\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
-
-\bibitem{eastoe}
-J. Eastoe and J.S. Dalton,
-Dynamic surface tension and adsorption mechanims of surfactants
-at the air-water interface,
-\textit{Advances in Colloid and Interface Science}, \textbf{85}
-13, 2000.
-
-\bibitem{ginzburg}
-I. Ginzburg and D. d'Humi\`eres,
-Multireflection boundary conditions for lattice Boltzmann models,
-\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
-
-\bibitem{h59}
-Hasimoto, H., On the periodic fundamental solutions of the Stokes
-equation and their application to viscous flow past a cubic array
-of spheres.
-\textit{J. Fluid Mech.}, \textbf{5}, 317.
-
-\bibitem{heemels}
-Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
-particles using the lattice-Boltzmann method,
-\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.
-
-\bibitem{jo84}
-Jeffrey, D.J., and Y. Onishi,
-Calculation of the resistance and mobility functions for the two
-unequal rigid spheres in low-Reynolds-number flow,
-\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.
-
-\bibitem{viv}
-Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
-P. Bladon,
-Inertial effects in three dimensional spinodal decomposition of
-a symmetric binary fluid mixture: A lattice Boltzmann study,
-\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).
-
-\bibitem{l94a}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 1. Theoretical foundation,
-\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.
-
-\bibitem{l94b}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 2. Numerical results,
-\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.
-
-\bibitem{l96a}
-Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
-spheres,
-\textit{Phys. Fluids}, \textbf{9}, 491. 1996.
-
-\bibitem{l96b}
-Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
-of non-Brownian spheres,
-\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.
-
-\bibitem{lv01}
-Ladd, A.J.C., and R. Verberg,
-Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
-
-\bibitem{lipanmiller}
-H. Li, C. Pan, and C.T. Miller,
-Pore-scale investigation of viscous coupling effects for two-phase
-flow in porous media,
-\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
-
-\bibitem{nl02}
-Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
-lattice-Boltzmann simulations of particle suspensions,
-\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.
-
-\bibitem{papanastasiou}
-T. paapnastasiou, G. Georgiou, and A. Alexandrou,
-\textit{Viscous Fluid Flow},
-CRC Press, Boca Raton, Florida, 2000.
-
-\bibitem{r95}
-Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
-Cambridge University Press, 1995.
-
-\bibitem{succi}
-S. Succi, \textit{The lattice Boltzmann equation and beyond},
-Oxford University Press, Oxford, 2001.
-
-\bibitem{edo1}
-M. Venuroli and E.S. Boek,
-Two-dimensional lattice-Boltzmann simulations of single phase
-flow in a pseudo two-dimensional micromodel,
-\textit{Physica A}, \textbf{362}, 23, 2006.
-
-\bibitem{vandergraaf}
-R.G.M. van der Sman and S. van der Graaf,
-Diffuse interface model of surfactant adsorption onto flat and
-droplet interfaces,
-\textit{Rheol. Acta} \textbf{46} 3 (2006).
-
-\bibitem{theissengompper}
-O. Theissen and G. Gompper,
-Lattice Boltzmann study of spontaneous emulsification,
-\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
-
-\bibitem{wardtordai}
-A.F.H. Ward and L. Tordai,
-\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
-
-\end{thebibliography}
-
-\end{document}

### 2014-09-24 branches/testcharge/htdocs/multigrid.tex
sync merge trunk to testcharge
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2478 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-08-01 htdocs/multigrid.tex
@@
 \begin{eqnarray*}
-k&\dots&\mbox{grid level with $k=1$ being the coarsest grid}\\
+k&\dots&\mbox{grid level $1\le k \le k_{max}$ with $k=1$ being the coarsest grid}\\
 \Omega_k &\dots& \mbox{regular grid with spacing $h(k)$}\\
@@
 This can have a detrimental effect on the overall-convergence of the algorithm as the lowest grid level is probably still too fine to achieve rapid convergence.
+It is also possible to use a fast, parallel iterative solver like conjugate-gradient on the coarstest grid.
 
-Another strategy is to use a fast, parallel iterative solver on the coarstest grid level like conjugate-gradient solver for instance.
-
-It is also possible to gather grid points that are initially allocated on different processes, solve only on a subset of processes with the others idling and scatter afterwards.
-Alternatively the data from all processes can be distributed over the entire topology by means of an MPI All-to-All call.
-The identical calculation can then be performed simulataneously on all processes, making further communication after each iteration step unnecessary.
-The grid can be also further coarsed down to the level where a direct solutin is possible.
+In a first approach we will opt for another strategy.
+On a coarse grid the data from all individual processes can be distributed over the entire topology by using an All-to-All MPI-call.
+An exactly identical calculation is then performed on all processes making communication after each iteration step unnecessary.
+The grid can be also further coarsened down to the level where a direct solution becomes feasible.
 
@@
 
-\begin{algorithm}
+\begin{algorithm}[H]
 \caption{Multigrid V-cycle for one iteration step $v_k^m\rightarrow v_k^{m+1}$}
@@
 in the \texttt{testcharge} directory, where we have SOR (both uniform
-permitivity and heterogenous permitivity) and should shortly have a
-Fourier solver (uniform permeativity) available.
+permittivity and heterogenous permittivity) and should shortly have a
+Fourier solver (uniform permittivity) available.
 
@@
 
-The existing code has a single coordinaite system which corresponds
-to the fine grid $\Omega_h$ in the multipgrid picture. Details of the
+The existing code has a single coordinate system which corresponds
+to the fine grid $\Omega_h$ in the multigrid picture. Details of the
 coordinate system (global and local extent; MPI communicators; 3d-to-1d
@@
 A second option would be to retain a single \texttt{coords\_t} object,
-but add functionality to recognise a ggrid level $k$ and perform the
-necessary arithmetic/logic assuning uniform refinement. (This is
+but add functionality to recognise a grid level $k$ and perform the
+necessary arithmetic/logic assuming uniform refinement. (This is
 probably a less elegant and flexible solution, if more compact.)
@@
 fine grid. This allows interpolation and restriction operators to be
-evalautated locally.
+evaluated locally.
 

### 2014-09-24 branches/testcharge/src-alan/blue_phase.h

@@ -1,0 +1,69 @@
+/*****************************************************************************
+ *
+ *  blue_phase.h
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef BLUEPHASE_H
+#define BLUEPHASE_H
+
+/* 'Extension' of free energy (pending free_energy_tensor.h) */
+
+#include "free_energy.h"
+
+void   blue_phase_set_free_energy_parameters(double, double, double, double);
+void   blue_phase_set_xi(double);
+void   blue_phase_set_zeta(double);
+void   blue_phase_set_gamma(double);
+
+double blue_phase_free_energy_density(const int);
+double blue_phase_compute_fed(double q[3][3], double dq[3][3][3]);
+double blue_phase_compute_bulk_fed(double q[3][3]);
+double blue_phase_compute_gradient_fed(double q[3][3], double dq[3][3][3]);
+
+void   blue_phase_molecular_field(const int, double h[3][3]);
+void   blue_phase_compute_h(double q[3][3], double dq[3][3][3],
+			    double dsq[3][3], double h[3][3]);
+void   blue_phase_compute_stress(double q[3][3], double dq[3][3][3],
+				 double h[3][3], double sth[3][3]);
+
+void   blue_phase_chemical_stress(const int, double sth[3][3]);
+void   blue_phase_redshift_set(const double redshift);
+void   blue_phase_electric_field_set(const double e[3]);
+void   blue_phase_dielectric_anisotropy_set(double e);
+double blue_phase_amplitude_compute(void);
+double blue_phase_get_xi(void);
+double blue_phase_get_zeta(void);
+double blue_phase_get_gamma(void);
+double blue_phase_chirality(void);
+double blue_phase_reduced_temperature(void);
+double blue_phase_redshift(void);
+double blue_phase_q0(void);
+double blue_phase_kappa0(void);
+double blue_phase_kappa1(void);
+double blue_phase_redshift(void);
+double blue_phase_rredshift(void);
+double blue_phase_a0(void);
+double blue_phase_gamma(void);
+double blue_phase_dimensionless_field_strength(void);
+
[…]

### 2014-11-14 branches/kevin/docs/devel.tex
docs directory
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2542 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  devel.tex
%
%  Information for users and delvelopers
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2014 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{General Information for Users and Developers}

This section contains information on the design of the code, along
with details of compilation an testing procedures
which may be of interest to both users and developers.
\textit{Ludwig} is named for Ludwig Boltzmann (1844--1906) to
reflect its background in lattice Boltzmann hydrodynamics.

\subsection{Manifest}

The top level ludwig directory should contain at least the following:
\begin{lstlisting}
$ ls
bash-3.2$ ls
LICENSE         docs            src             tests           version.h
README          mpi_s           targetDP        util
\end{lstlisting}
The code is released under a standard BSD license; the current version
number is found in \texttt{version.h}. The main source is found in
the \texttt{src} directory, with other relevant code in \texttt{tests}
and \texttt{util}. The \texttt{targetDP} directory holds code related
to the targetDP abstraction layer \cite{gray2013}. These documents are
found in \texttt{docs}.

\subsection{Code Overview}

The code is ANSI C (1999), and can be built without any external
dependencies, with the exception of the message passing interface
(MPI), which is used to provide domain-decomposition based parallelism.

\subsubsection{Design}

The \textit{Ludwig} code has evolved and expanded over a number of
years to its present state. Its original purpose was to investigate
specifically spinodal decomposition in binary fluids (see, e.g.,
\cite{kendon2001}). This work used a free-energy based formulation
combined with a two-distribution approach to the binary fluid problem
(with lattice Boltzmann model D3Q15 at that time). This approach  is
retained today, albeit in a somewhat updated form. The wetting problem
for binary fluid at solid surfaces has also been of consistent interest.
The code has always been developed with
parallel computing in mind, and has been run on a large number of
different parallel machines including the Cray T3E at Edinburgh. These
features, developed by J.-C. Desplat and others,  were reflected in the
early descriptive publication in \textit{Comp.\ Phys.\ Comm} in 2001
\cite{desplat2001}.

The expansion of the code to include a number of additional features
has occasioned significant alterations over time, and little of the
original code remains. However, the fundamental idea that the code
should essentially operate for high performance computers and be
implemented using ANSI C with message passing via MPI has remained
unchanged.

The code has a number of basic building blocks which are encapsulated
in individual files.

\subsubsection{Parallel environment}

The code is designed around message passing using MPI. A stub MPI
library is provided for platforms where a real MPI is not available
(an increasingly rare occurrence), in which case the code runs in
serial. The parallel environment (interface defined in \texttt{pe.h})
therefore underpins
the entire code and provides basic MPI infrastructure such as
\texttt{info()},
which provides \texttt{printf()} functionality at the root process
only. The parallel environment also provides information on version
number etc. MPI parallelism is implemented via standard domain
decomposition based on the computational lattice (discussed
further in the following section on the coordinate system).

\subsubsection{targetDP: threaded parallelism and vector parallelism}

To allow various threaded models to be included without duplicating
source code, we have developed a lightweight abstraction of the
thread level parallelism, This currently supports a standard single-threaded
model, OpenMP, or CUDA. targetDP (``target data parallel'') allows single
kernels to be written which can then be compiled for the different threaded
models which may be appropriate on different platforms. targetDP can also
be used as a convenient way to express vector parallelism
(typically SSE or AVX depending on processor architecture). targetDP allows
explicit specification of the vector length at compile time.

targetDP does not automatically identify parallelism; this must still be
added appropriately by the developer. It is merely a concise way to include
different threaded models. It is only available in the development version.

\subsubsection{Coordinate system}

It is important to understand the coordinate system used in the
computation. This is fundamentally a regular, 3-dimensional Cartesian
coordinate system . (Even if a D2Q9 LB model is employed, the code is
still fundamentally 3-dimensional, so it is perhaps not as efficient  
for 2-dimensional problems as it might be.)

The coordinate system is centred around the (LB) lattice, with lattice
spacings $\Delta x = \Delta y = \Delta z = 1$. We will refer, in general,
to the lattice spacing as $\Delta x$ throughout, its generalisation to
three dimensions $x,y,z$ being understood. Lattice sites in the
$x-$direction therefore have unit spacing and are located at
$x = 1, 2, \ldots, N_x$.
The length of the system $L_x = N_x$, with the limits of
the computational domain begin $x = 1/2$ and $x = L_x + 1/2$. This allows
us to specify a unit control volume centred on each lattice site $x_i$
being $i-1/2$ to $i + 1/2$. This will become particularly significant for
finite difference (finite volume) approaches discussed later.

Information on the coordinate system, system size and so on is
encapsulated in \texttt{coords.c}, which also deals with the
regular domain decomposition in parallel. Decompositions may be
explicitly requested by the user in the input. or computed by
the code itself at run time. \texttt{coords.h} also provides
basic functionality for message passing within a standard MPI
Cartesian communicator, specification of periodic boundary
conditions, and so on.

Three-dimensional fields are typically stored on the lattice,
but are addressed in compressed one-dimensional format. This
avoids use of multidimensional arrays in C. This addressing
must take account of the width of the halo region required at
the edge of each sub-domain required for exchanging information
in the domain decomposition. The extent of the halo region
varies depending on the application required, and is selected
automatically at run time.

\subsubsection{Lattice Boltzmann hydrodynamics}

The hydrodynamic core of the calculation is supplied by the lattice
Boltzmann method, which was central at the time of first development.
LB is also the basis for hydrodynamic solid-fluid interactions at
stationary walls and for moving spherical colloids. The LB approach
is described in more detail in Section~\ref{section:lb-hydrodynamics}.
For general
and historical references, the interested reader should consider,
e.g., Succi \cite{succi}.

\subsubsection{Free energy}

For complex fluids, hydrodynamics is augmented by the addition of
a free energy for the problem at hand, expressed in terms of an
appropriate order parameter. The order parameter may be a three
dimensional field, vector field, or tensor field. For each problem
type, appropriate tome evolution of the order parameter is supplied.

Coupling between the thermodynamic sector and the hydrodynamics is
abstracted so that the hydrodynamic core does not require alteration
for the different free energies. This is implemented via a series of
call back functions in the free energy which are set appropriately
at run time to correspond to that specified by the user in the
input.

Different (bulk) fluid free energy choices are complemented by
appropriate surface free energy contributions which typically
alter the computation of order parameter gradients at solid
surface. Specific gradient routines for the calculation of
order parameter gradients may be selected or added by the
user or developer.

Currently available free energies are:
\begin{itemize}
\item Symmetric binary fluid with scalar compositional order parameter
$\phi(\mathbf{r})$ and related Cahn-Hilliard equation;
\item Brazovskii smectics, again which scalar compositional order parameter
$\phi(\mathbf{r})$;
\item Polar (active) gels with vector order parameter $P_\alpha (\mathbf{r})$
and related Leslie-Erikson equation;
\item Landau-de Gennes liquid crystal free energy with tensor
orientational order parameter $Q_{\alpha\beta}(\mathbf{r})$, extended to
apolar active fluids and related Beris-Edwards equation;
\item a free energy appropriate for electrokinetics for charged fluids
and related Nernst-Planck equations (also requiring the solution of the
Poisson equation for the potential);
\item a coupled electrokinetic binary fluid model;
\item a liquid crystal emulsion free energy which couples a binary
composition to the liquid crystal.
\end{itemize}
See the relevant sections on each free energy for further details.

\subsection{Compilation}

Compilation of the main code is controlled by the \texttt{Makefile} in
the \texttt{./src} directory. This section discusses a number of issues
which influence compilation.

\subsubsection{Dependencies on third-party software:}
There is the option to use PETSC to solve the Poisson equation required in
the electrokinetic problem. A rather less efficient in-built method
can be used if PETSC is not available.
We suggest using PETSC
v3.4 or later available from Argonne National Laboratory
\texttt{http://www.mcs.anl.gov/petsc/}.

The tests use the lightweight implementation of exceptions provided
under the GPL Lesser General Public License by Guillermo Calvo. This
is included with the source.

\subsubsection{Makefile}

The \texttt{Makefile} provided in the source directory has a number of
targets to allow different models to be compiled. Before compilation,
the Makefile must be edited provide details of the local C compiler(s).
The relevant lines are usually limited to:
\begin{lstlisting}
CC = cc
MPICC = mpicc
CFLAGS = -O2
\end{lstlisting}
where \texttt{CC} is the compiler used for serial compilation,
\texttt{MPICC} is the compiler used for compilation against
MPI, and \texttt{CFLAGS} provides compiler switches (often
related to optimisation). Beyond these variables, changes to
the \texttt{Makefile} should not usually be required.

\subsubsection{C assertions}

The code makes quite a lot of use of standard C assertions, which
are useful to prevent errors. They do result in a considerably
slower execution in some instances, so production runs should
switch off the assertions with the standard \texttt{NDEBUG}
preprocessor flag. Add \texttt{-DNDEBUG} to \texttt{OPTS} in
the \texttt{Makefile}.

\subsubsection{Targets for serial and parallel code}

The code can be compiled with or without a true MPI library.
For serial execution, the MPI stub library must be compiled
first (see ``Quick Start for Users'' Section \ref{section:quick-serial}).
The appropriate target is:
\begin{lstlisting}
bash-3.2$ make serial
make serial-d3q19
make serial-model ``LB=-D_D3Q19_'' ``LBOBJ=d3q19.o''
...
\end{lstlisting}
from which it will be seen that the default LB model is D3Q19. This is
the same for the true MPI target
\begin{lstlisting}
bash-3.2$ make mpi
make mpi-d3q19
make mpi-model ``LB=-D_D3Q19_'' ``LBOBJ=d3q19.o''
...
\end{lstlisting}

\subsubsection{Targets for different LB models}

Specific targets are provided if an alternative LB model is wanted.
The available options are:
\begin{lstlisting}
make [ serial-d2q9 | serial-d3q15 | serial-d3q19 ]
make [ mpi-d2q9 | mpi-d3q15 | mpi-d3q19 ]
\end{lstlisting}
A further set of targets is supplied if `reverse' or structure-of-array
memory ordering is wanted (e.g., for GPU computing):
\begin{lstlisting}
make [ serial-d2q9r | serial-d3q15r | serial-d3q19r ]
make [ mpi-d2q9r | mpi-d3q15r | mpi-d3q19r ]
\end{lstlisting}
This will only influence efficiency of the code: results are unchanged.

\subsection{Tests}

A series of tests are provided in the \texttt{./tests} directory. These
are of two types. Unit tests are generally written when units of code
are first introduced to ensure basic operation is error free. Regression
tests are introduced and updated when physical results are validated.
Both are run regularly as the basis of a nightly test procedure, but
they can be run independently, e.g., after adding or changing code.

\subsubsection{Running the tests}

\subsection{Additional Notes for Developers}
\label{xref:developers}

Developers who wish to contribute code to the SVN repository
please consider the following pleas concerning standards.

\subsubsection{Coding standards}

While definitive statements on style are avoided, please try to
maintain some basic standards: (0) code should be strictly ANSI C99
standard; (1) code should compile without warnings when appropriate
compiler flags are set (e.g., \texttt{-Wall} under \texttt{gcc});
(2) avoid long lines of code for readability reasons;
(3) avoid confusing commented-out code; (4) avoid conditional pre-processor
directives if possible; (5) add meaningful and descriptive comments;
(6) use standard \texttt{assert()} to trap programming errors;
(7) explicitly trap possible run time errors (8) add appropriate
tests.

\subsubsection{Documentation standards}

Additions and alterations to the code need to be reflected in the
documentation. These should be checked in at the same time as the
code itself.

\subsubsection{Protocol}

Before checking in code, please follow procedure: (1) increment the
patch version number in \texttt{version.h} consistent with the SVN
(if a minor version increment is required, communication with
other developers must be considered); (2) run at least the unit
tests and the short regression tests; (3) add a note to the change
log; and (4) SVN update and commit.

\vfill
\pagebreak

### 2014-11-14 branches/kevin/docs/intro.tex
docs directory
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2542 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  intro.tex
%
%  Contains introduction and quick start for users.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2014 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\subsection{Overview}

This introduction provides a brief overview of what the \textit{Ludwig} code
does, how to obtain and build it, and how to run a sample problem.
It is assumed that the reader has at least a general knowledge
of Navier-Stokes hydrodynamics, complex fluids, and to some
extent statistical physics. This knowledge will be required to make sense
of the input and output involved in using the code. Those wanting
to work on or develop the code itself will need knowledge of ANSI C,
and perhaps message passing and CUDA. We assume the reader is using
a Unix-based system.

\subsubsection{Purpose of the code}

The \textit{Ludwig} code has been developed over a number of
years to address specific problems in complex fluids. The underlying
hydrodynamic model is based on the lattice Boltzmann equation
(LBE, or just `LB'). This itself may be used to study simple
(Newtonian) fluids in a number of different scenarios, including porous
media and particle suspensions. However, the
code is more generally suited to complex fluids, where a number of
options are available, among others: symmetric binary fluids and
Brazovskii smectics,
polar gels, liquid crystals, or charged fluid via a Poisson-Boltzmann
equation approach. These features are added in the framework of a
free energy approach, where specific compositional or orientational
order parameters are evolved according to the appropriate
coarse-grained dynamics, but also interact with the fluid in a
fully coupled fashion.

A number of other features are catered for is some or all situations.
These include fluctuating hydrodynamics, and appropriate fluctuating
versions of order parameter dynamics for binary solvents and liquid
crystals. Colloidal particles are available for simulation of
suspensions and dispersions, and with appropriate boundary
conditions binary mixtures, liquid crystals, and charged fluid. Shear
flow is available for fluid-only systems implemented via Lees-Edwards
sliding periodic boundaries. Not all these features play with all the
others: check the specific section of the document for details.

Broadly, the code is intended for complex fluid problems at low
Reynolds numbers, so there is no consideration of turbulence,
high Mach number flows, high density ratio flows, and so on.

\subsubsection{How the code works}

We aim to provide a robust and portable code, written in ANSI C, which
can be used to perform serial and scalable parallel simulations of
complex fluid systems based around hydrodynamics via the lattice
Boltzmann method. Time evolution of modelled quantities takes
place on a fixed regular discrete lattice. The preferred method of
dealing with the corresponding order parameter equations is by
using finite difference. However, for the case of a binary fluid,
a two-distribution lattice Boltzmann approach is also maintained
for historical reference.

Users control the operation of the code via a plain text input file;
output for various data are available. These data may be visualised
using appropriate third-party software (e.g., Paraview). Specific
diagnostic output may require alterations to the code.

Potential users should also note that the complex fluid simulations
enabled by \textit{Ludwig} can be time consuming, prone to instability,
and provide results which are difficult to
interpret. We make no apology for this: that's the way it is.

\subsubsection{Who should read what?}

This documentation is aimed largely at users, but includes information
that will also be relevant for developers. The documentation discusses
the underlying features of the coordinate system, hydrodynamics (LB), and
other generic features, and has a separate section for each free energy.
You will need to consult the section relevant to your problem.

\subsection{Obtaining the code}

Under the auspices of the CCPs (Collaborative Computational Projects),
and in particular CCP5, we have a CCPForge project account at
\texttt{http://ccpforge.cse.rl.ac.uk/}
which provides revision control via SVN, issue tracking, and so on.

\paragraph{Users:}

Code releases may be downloaded via anonymous download
\begin{lstlisting}[style=terminalverbatim]
http://ccpforge.cse.rl.ac.uk/gf/project/ludwig/frs/
\end{lstlisting}
The current release is available as \texttt{ludwig-0.4.0.tar.gz}.
Alternatively, registered users can obtain the code via svn checkout
in the same way as developers.

\paragraph{Developers:}

Developers need to create an account at CCPForge, and let one of the
project owners\footnote{Currently: \texttt{kevin@epcc.ed.ac.uk},
\texttt{ohenrich@epcc.ed.ac.uk}}
know the user name so you can be added to the access list.

To download the entire code, developers need
\begin{lstlisting}
svn checkout --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig
\end{lstlisting}
where \texttt{<user>} is replaced by your CCPForge user name. You will
see
\begin{lstlisting}
$ ls ludwig
branches  tags      trunk
\end{lstlisting}
which is the usual svn convention for trunk, branches and tags.
The process for committing code is described in section \ref{xref:developers}.

\vfill
\pagebreak

\section{Quick Start for Users}

This section provides a brief overview of compiling and running
the code, which should allow users to get off the ground.
The section assumes the code has been downloaded and you are in the
top level directory.

\subsection{Compilation}

Identify your local C compiler. The following example uses \texttt{gcc}
in serial, and \texttt{mpicc} in parallel. For MPI in particular,
local details may vary.

\subsubsection{Serial}
\label{section:quick-serial}

First, compile the MPI stub library in the \texttt{mpi\_s}
directory. Do this by editing the Makefile, and checking the compiler
is appropriate. You should be able to build the library and run the
tests using, e.g.,:

\begin{lstlisting}
$ make libc
gcc -Wall -I.  -c mpi_serial.c
ar -cru libmpi.a mpi_serial.o
$ make testc
gcc -Wall   mpi_tests.c -L. -lmpi
./a.out
Running mpi_s tests...
Finished mpi_s tests ok.
\end{lstlisting}

Now compile the main code in the \texttt{src} directory. Again
edit the Makefile to check the compiler. Compilation should
look like:

\begin{lstlisting}
$ make serial
make serial-d3q19
make serial-model "LB=-D_D3Q19_" "LBOBJ=d3q19.o"
make lib
make libar "INC=-I. -I../mpi_s" "LIBS= -L../mpi_s -lmpi -lm"
gcc -D_D3Q19_ -Wall -O2 -I. -I../mpi_s -c model.c
gcc -D_D3Q19_ -Wall -O2 -I. -I../mpi_s -c propagation.c
...
\end{lstlisting}
which shows that the default lattice Boltzmann model is D3Q19.
Successful compilation will provide an executable \texttt{Ludwig.exe}
which is linked against the MPI stub library.

\subsubsection{Parallel}

Here, you do not need to compile the stub library. Simply compile
the main code with, e.g.,:
\begin{lstlisting}
$ make mpi
make mpi-d3q19
make mpi-model "LB=-D_D3Q19_" "LBOBJ=d3q19.o"
make libmpi
make libar "CC=mpicc"
mpicc -D_D3Q19_ -Wall -O2 -I. -c model.c
mpicc -D_D3Q19_ -Wall -O2 -I. -c propagation.c
...
\end{lstlisting}

Again, this should produce an executable \texttt{./Ludwig.exe}
in the current directory, this time linked against the true MPI
library.

\subsection{Running an Example}

\subsubsection{Input file}

The behaviour of \textit{Ludwig} at run time is controlled by
a plain text input file which consists of a series for key value
pairs. Each key value pair controls a parameter or parameters
within the code, for example the system size and number of time steps,
the fluid properties density and viscosity, the free energy type
and associated parameters (if required), frequency and type of
output, parallel decomposition, and so on.

For most keys, the associated property has some default value which
will be used by the code if the relevant key is not present (or
commented out) in the input file. While such default values are
chosen to be at least sensible, users need to be aware that all
necessary keys need to be considered for a given problem. However,
many keys are irrelevant for any given problem.

A significant number of example input files are available as
part of the test suite, and these can form a useful starting
point for a given type of problem. We will consider one which
uses some of the common keys.

\subsubsection{Example input}

We will consider a simple example which computes the motion of
a single spherical colloidal particle in an initially stationary
fluid of given viscosity. The velocity may be measured as a
function of time. Assume we have an executable in the \texttt{src}
directory.

Make a copy of the input file
\begin{lstlisting}
$ cp ../tests/regression/d3q19/serial-auto-c01.inp .
\end{lstlisting}

Inspection of this file will reveal the following: blank lines
and comments --- lines beginning with \texttt{\#} --- may be included,
but are ignored at run time; other lines are parsed as key value
pairs, each pair on a separate line, and with key and value separated
by one or more spaces. Values may be characters or strings, a
scalar integer or 3-vector of integers, or a scalar or 3-vector
of floating point numbers. Values which are 3-vectors are delineated
by an underscore character (not a space), e.g., \texttt{1\_2\_3},
and always correspond to a Cartesian triple $(x,y,z)$.
A given value is parsed appropriately in the context of the associated key.

So, the first few lines of the above file are:
\lstinputlisting[firstline=1, lastline= 16]
{../tests/regression/d3q19/serial-auto-c01.inp}
Here, the first key value pair \texttt{N\_cycles 40} sets the number
of time steps to 40, while the second, \texttt{size 64\_64\_64}, sets
the system size to be 64 points in each of the three coordinate
directions. The \texttt{grid} key relates to the parallel
domain decomposition as is discussed in the following
section.

\subsubsection{Run time}

The executable takes a single argument on the command line which
is the name of the input file, which should be in the same
directory as the executable. If no input file name is supplied,
the executable will look for a default \texttt{input}. If no
input file is located at all, an error to that effect will be
reported.

\textbf{Serial: }
If a serial executable is run in the normal way with the copy of the
input file as the argument the code should take around 10--20 seconds
to execute 40 steps. The fist few lines of output should look like:

\begin{lstlisting}[belowskip=0pt]
$ ./Ludwig.exe ./serial-auto-c01.inp
\end{lstlisting}
\lstinputlisting[aboveskip=0pt, firstline=1, lastline= 16]
{../tests/regression/d3q19/serial-auto-c01.log}
This output shows that the appropriate input file has been read, and
the system size set correspondingly with a number of default settings
including periodic boundary conditions. ``No free energy'' tells us we
are using a single Newtonian fluid.

Normal termination of execution is accompanied by a report
of the time take by various parts of the code, and finally by
\begin{lstlisting}
...
Ludwig finished normally.
\end{lstlisting}

%\paragraph{Parallel: }
\textbf{Parallel: }
The executable compiled and linked against the MPI library can
be run with the MPI launcher for the local system, often
\texttt{mpirun}. For example, a run on 8 MPI tasks produces a
similar output to that seen in the serial case, but reports
details of the local domain decomposition:
\begin{lstlisting}
$ mpirun -np 8 ./Ludwig.exe ./serial-auto-c01.inp
Welcome to Ludwig v0.1.26 (MPI version running on 8 processes)
...
System details
--------------
System size:    64 64 64
Decomposition:  2 2 2
Local domain:   32 32 32
...
\end{lstlisting}
The decomposition is controlled by the \texttt{grid} key in
the input. Here, \texttt{grid 2\_2\_2} is consistent with the
8 MPI tasks specified on the command line, and the resulting
local decomposition is 32 lattice points in each coordinate direction.
If no grid is specified in the input, or a request is
make for a grid which cannot be met (e.g., the product
of the grid dimensions does not agree with the total number
of MPI tasks available)
the code will try to determine its own decomposition as best
it can. If no valid parallel decomposition is available at all,
the code will exit with a message to that effect.

Further details of various input key value pairs are given in
relevant sections of the documentation.

\subsubsection{Output}

The standard output of the running code produces a number of
aggregate quantities which allow a broad overview of the progress
of the computation to be seen by the user. These include, where
relevant, global statistics related to fluid density and momentum,
the integrated free energy, particle-related quantities, and so on.
The frequency of this information can be adjusted from the input
file (see, e.g., \texttt{freq\_statistics}).

Output for lattice-based quantities (for example, the velocity field
$u(\mathbf{r}; t)$) is via external file. This output is in serial
or in parallel according to how the model is run, and may be in
either ASCII or raw binary format as required. Output files are produced
with time step encoded in the file, and an extension which describes the
parallel output decomposition. Further, each quantity output in this way
is accompanied by a metadata description with \texttt{meta} extension.
For example, the two output files
\begin{lstlisting}
bash-3.2$ ls dist*
dist-00000020.001-001   dist.001-001.meta
\end{lstlisting}
contain the LB distributions for time step 20 (the file is 1 of a total
number of 1 in parallel), and the plain text metadata description,
respectively.

To ensure output for based-based quantities is in the correct order for
analysis, post-processing
may be required (always if the code is run in parallel). This uses a utility
provided for the purpose which required both the data and the metadata
description to recombine the parallel output. This utility is described
ELSEWHERE.

Output for colloidal particle data is again to file with a name encoding
the time step and parallel docomposition of the output. For example,
\begin{lstlisting}
bash-3.2$ ls config*
config.cds00000020.001-001
\end{lstlisting}
contains particle data for time step 20 in a format described ELSEWHERE.
These data may be requested in ASCII or raw binary format.

\subsubsection{Errors and run time failures}

The code attempts to provide meaningful diagnostic error messages
for common problems. Such errors include missing or incorrectly
formatted input files, inconsistent input values, and unacceptable
feature combinations. The code should also detect other run time
problems such as insufficient memory and errors writing output
files. These errors will result in termination.

Instability in the computation will often be manifested by numerically
absurd values in the statistical output (ultimately \texttt{NaN} in
many cases). Instability may or may not result in termination, depending
on the problem. Such instability is very often related to poor parameter
choices, of which there can be many combinations. Check the
relevant section of the documentation for advice on reasonable starting
parameters for different problems.

\subsection{Note on Units}

All computation in \textit{Ludwig} is undertaken in ``lattice
units,'' fundamentally related to the underlying LB fluid model
which expects discrete model space and time steps
$\Delta x = \Delta t = 1$. The natural way to approach problems
is then to ensure that appropriate dimensionless quantities are
reasonable. However, ``reasonable'' may not equate to ``matching
experimental values''. For example, typical flows in colloidal
suspensions my exhibit Reynolds numbers as low as $O(10^{-6})$
to $O(10^{-8})$.
Matching such a value in a computation may imply an impractically
small time step; a solution is to let the Reynolds number rise
artificially with the constraint that it remains small compared to $O(1)$.
Further discussion of the issue of units is provided in, e.g.,
\cite{cates_scaling}. Again, consult the relevant section of the documentation
for comments on specific problems.

\vfill
\pagebreak

### 2014-12-06 branches/kevin/docs/targetdp.tex
added targetdp notes
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2552 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  targetdp.tex
%
%  Target Data Parallel information
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2014 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Target Data Parallel}

\subsection{Aim}

\subsection{Target context and API}

\subsection{Memory}

There are three types of memory requirement we will consider.
(A fourth, constant host memory not suitable for target constant
memory is could also be handled, but is less likely to occur.)
An approach to each relevant type is set out in the three following
sections.

\subsubsection{\texttt{const} object on host, and suitable for
\texttt{\_\_target\_const\_\_} memory}

Objects declared \texttt{const} on the host are suitable for transfer
to target constant memory. In order to have a consistent reference
to the object, the following approach can be used. We declare two
objects which are initialised via a macro-expansion at file scope:

\begin{lstlisting}
struct const_object {
  ...
};

#define object_initialiser() { ... }

const              struct const_object host = object_initialiser();
__target_const__ struct const_object target = object_initialiser();
\end{lstlisting}
All references to the object in kernel-related code is then via the
variable \texttt{target}, to be declared \texttt{extern} the the
relevant object header if necessary.
By making the appropriate macro-expansions for \texttt{\_\_target\_const\_\_}
we allow for different implementations. An implementation with separate
address spaces must ensure the \texttt{host}
contents are copied to the \texttt{target} object before any kernel
related activity. We suggest that this is at a well-defined
\begin{lstlisting}
__host__ int object_commit(struct object * obj);
\end{lstlisting}
stage. There is no coherency issue as both objects are constant. In a
shared-memory implementation, constant data are shared by virtue of
being at file scope.

We note a slightly more elegant version of the above would be
\begin{lstlisting}
struct const_object {
  /* declare */
};

const struct const_object object = {
  /* initialise */
};

const              struct const_object * const host = &object;
__target_const__ struct const_object * const target = &object;
\end{lstlisting}
which does away with the need for repeated initialisation via
an error-prone macro expansion. However,
for CUDA implementation the expansion of \texttt{\_\_target\_const\_\_}
to \texttt{\_\_constant\_\_}
may not get past \texttt{nvcc} unless an appropriate cast can
be identified to make the \texttt{target} initialisation consistent.

\subsubsection{Non \texttt{const} object on host, but suitable for
\texttt{\_\_target\_const\_\_} memory}

Here, the situation is less clear. Consider a non-constant data item
on the host defined as part of a object, along with corresponding
\texttt{target}:
\begin{lstlisting}
struct object {
   struct kernel_const host;
   ...
};

__target_const__ struct kernel_const target;

\end{lstlisting}
Here, an implementation with either uniform or separate address
spaces must arrange a copy of the data at appropriate times so
that any updates to \texttt{host} are correctly reflected
in \texttt{target}. All kernel-related references are again through
\texttt{target}.
The slightly disjoint nature of this definition and increased pressure
on global name-space must be weighed against potential increase in kernel
performance and reduction in register pressure.

\subsubsection{Non \texttt{const} object in both address spaces}

The most general case may be handled in the following manner:
\begin{lstlisting}

struct object_kernel_data {
  ...
};

struct object {
   struct object_kernel_data * host;
   struct object_kernel_data * target;
   ...
};

\end{lstlisting}
All relevant kernel data may then be handled by references to
\texttt{target} in a consistent way. For implementations with
separate address spaces, the memory coherency issue
must be handled explicitly as both \texttt{host} and \texttt{target}
are subject to modification. However, an implementation with uniform
address space may simply alias the \texttt{target} to the \texttt{host}
pointer. No data reside at file scope.

\subsubsection{Aliasing}
Consider a kernel data structure:
\begin{lstlisting}
struct object_kernel_data {
  double * data1;
  double * data2;
};
\end{lstlisting}
An object of this type is to be passed into a kernel function via a pointer,
which itself may be declared in contract to be
\begin{lstlisting}
__target__ void kernel_function(struct object_kernel_data * __restrict__ obj);
\end{lstlisting}
Further instruction to the compiler in the kernel function itself would
be required to specify the status of individual data items, e.g.,
\begin{lstlisting}
__target__ void kernel_function(struct object_kernel_data * __restrict__ obj) {

  double * restrict data1 = obj->data1;
  double * restrict data2 = obj->data2;

  /* ... and reference via ``data1'' and ``data2'' */
\end{lstlisting}
Standard C99 is required for a host implementation.
A test of this for a CUDA implementation with \texttt{nvcc} is needed.

\subsection{Kernel Execution}

\subsubsection{Typical kernel in host code}

A typical kernel representing a loop over local domain lattice sites
might look like:
\begin{lstlisting}

int nextra = 1;  /* extent of kernel into halo regions */
int nlocal[3];    /* local domain size */

for (ic = 1 - nextra; ic <= nlocal[X] + nextra; ic++) {
  for (jc = 1 - nextra; jc <= nlocal[Y] + nextra; jc++) {
    for (kc = 1 - nextra; kc <= nlocal[Z] + nextra; kc++) {

       /* operations for site (ic, jc, kc) which may involve neighbours */
       index = coords_index(ic, jc, kc);
       ...
    }
  }
}

\end{lstlisting}
Indexing calculations for lattice data structures are required to
respect the total extent of the halo regions. Not all sites are
involved in a typical kernel as determined above by \texttt{nextra}.

\subsubsection{Typical kernel in target code}

In a target implementation, the schematic code block shown above is necessarily
split into two parts. The first is the scheduling of the required
iterations (taking the place of the nested loops), and the second
is the site-wise kernel (the code inside the loops).

The two parts might take the form, schematically:
\begin{lstlisting}
int nextra = 1;            /* extent of kernel into halo regions */
int nlocal[3];             /* local domain size */
kernel_block_t nblocks;   /* Kernel blocks */
kernel_block_t ntpb;       /* Number of threads per block */

target_launch(kernel_function, nblocks, ntpb, nextra, nlocal[3], ...);
\end{lstlisting}
to be followed by
\begin{lstlisting}
__target__ void kernel_function(int nextra, int nlocal[3], ...) {

  if (threadIndex < threadsRequired) {

    /* locate (ic, jc, kc) from  kernel thread index */
    ...
  }

  return;
}

\end{lstlisting}
A small number of utility functions would allow us to compute the actual
number of threads required dependent on \texttt{nextra}, and to perform
relevant indexing. See, e.g., functions in
\texttt{branches/CUDA/src/utilities\_gpu.cu}
for identifying colloid map on the target.

Note that a \texttt{target\_launch()} macro expansion (that looks
like a standard C syntax) would require something of the form
\begin{lstlisting}
#define target_launch(f, nblocks, ntpb, ...) f(__VA_ARGS__)
\end{lstlisting}
or, for a CUDA run-time API implementation
\begin{lstlisting}
#define target_launch(f, nblocks, ntpb, ...) f<<<nblocks, ntpb>>>(__VA_ARGS__)
\end{lstlisting}

\subsubsection{Kernel reductions}

Can we do reductions in a kernel in a sane fashion?

\subsubsection{Bespoke target implementations for a single kernel}

What does the code look like if we are forced to have separate kernel
implementations for different targets for some reason?

\vfill
\pagebreak

### 2014-12-10 branches/kevin/docs/targetdp.tex
branch update passing tests
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2558 40876441-3496-0410-a613-a2fca35b2ae2

@@
 on global name-space must be weighed against potential increase in kernel
-performance and reduction in register pressure.
+performance and reduction in register pressure. More seriously, thre is
+no way to write consistent \texttt{\_\_host\_\_ \_\_target\_\_} functions
+(we can see) except to move the host structure back to file scope to
+match the situation in the previous section.
 
@@
 pointer. No data reside at file scope.
+
+An attractive feature about this situation is that one can use the
+uniform picture in memory to write functions of contract
+\begin{lstlisting}
+__host__ __target__ int obj_function(struct obj_kernel_data * obj, ...);
+\end{lstlisting}
+which can be used in both host and kernel code.
 

### 2014-12-31 branches/kevin/docs/targetdp.tex
brnach update passin tests
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2574 40876441-3496-0410-a613-a2fca35b2ae2

--- 2014-12-06 branches/kevin/docs/targetdp.tex
@@
 
+\subsubsection{Error handling}
+
+The targetDP interface should provide a mechanism for errors to be
+handled by the application.
+
 \subsection{Memory}
 
-There are three types of memory requirement we will consider.
-(A fourth, constant host memory not suitable for target constant
-memory is could also be handled, but is less likely to occur.)
-An approach to each relevant type is set out in the three following
-sections.
+First, we consider the general problem of encapsulation of objects
+in the context of device memory. We then consider two special cases
+involving device constant memory.
+
+\subsubsection{General}
+
+It is useful to manage host and device memory in such a way that
+three is a uniform interface to related functionality, and that
+will not require substantially different code or API to handle the cases of
+unified and separate address spaces.
+
+The model used is to define host objects which include a reference
+to an object of the same type intended for device memory. A schematic
+example is
+\begin{lstlisting}
+struct object_s { 
+  ...                            /* data items (host) */
+  struct object_s * target;      /* device copy */
+};
+\end{lstlisting}
+The object created on the host is then responsible for the management
+of the appropriate device reference \texttt{target}. For implementations
+with a unified address space the host may simply alias the target
+reference to itself. For implementations with separate address spaces
+appropriate allocation of device memory must be arranged, along with
+necessary marshalling of data between host and device. The model
+allows consistent use of functions with contract
+\begin{lstlisting}
+__host__ __device__ int object_function(struct object_s * obj, ...)
+\end{lstlisting}
+Kernel functions may be declared with contract
+\begin{lstlisting}
+__target_entry__ void object_kernel(struct object_s * obj, ...)
+\end{lstlisting}
+and must take the \texttt{target} reference as actual argument when invoked
+from the host. Likewise, functions declared with \texttt{\_\_host\_\_}
+attribute may be
+invoked on the host with the
+host reference as actual argument.
+We note:
+\begin{enumerate}
+\item
+A number of data items only relevant on the host may be retained in
+the \texttt{target} reference redundantly.
+\item
+The \texttt{target} reference in the device copy is also redundant.
+\item
+The host does not maintain a permanent copy of all device memory
+references in this model. However, these may be recovered from the
+device if needed.
+\end{enumerate}
+I think the benefits this model, which allows a clearly uniform picture
+of host and device memory, and is largely transparent to details of the
+address space, justify the small number of slightly inelegant redundancies.
 
@@
 on global name-space must be weighed against potential increase in kernel
-performance and reduction in register pressure.
-
-\subsubsection{Non \texttt{const} object in both address spaces}
-
-The most general case may be handled in the following manner:
-\begin{lstlisting}
-
-struct object_kernel_data {
-  ...
-};
-
-struct object {
-   struct object_kernel_data * host;
-   struct object_kernel_data * target;
-   ...
-};
-
-\end{lstlisting}
-All relevant kernel data may then be handled by references to
-\texttt{target} in a consistent way. For implementations with
-separate address spaces, the memory coherency issue
-must be handled explicitly as both \texttt{host} and \texttt{target}
-are subject to modification. However, an implementation with uniform
-address space may simply alias the \texttt{target} to the \texttt{host}
-pointer. No data reside at file scope.
+performance and reduction in register pressure. More seriously, thre is
+no way to write consistent \texttt{\_\_host\_\_ \_\_target\_\_} functions
+(we can see) except to move the host structure back to file scope to
+match the situation in the previous section.
 

### 2014-12-31 branches/kevin/src/blue_phase_rt.c

@@ -335,17 +335,17 @@ int blue_phase_rt_initial_conditions(rt_t * rt, coords_t * cs, field_t * q) {
 
   if (strcmp(key1, "h2d") == 0) {
     info("Initialising Q_ab using H2D\n");
-    blue_phase_H2D_init(q);
+    blue_phase_H2D_init(cs, q);
   }
 
   if (strcmp(key1, "h3da") == 0) {
     info("Initialising Q_ab using H3DA\n");
-    blue_phase_H3DA_init(q);
+    blue_phase_H3DA_init(cs, q);
   }
 
   if (strcmp(key1, "h3db") == 0) {
     info("Initialising Q_ab using H3DB\n");
-    blue_phase_H3DB_init(q);
+    blue_phase_H3DB_init(cs, q);
   }
 
   if (strcmp(key1, "dtc") == 0) {
@@ -359,7 +359,7 @@ int blue_phase_rt_initial_conditions(rt_t * rt, coords_t * cs, field_t * q) {
     info("BPIII specifications: N_DTC=%g,  R_DTC=%g,  ", nhat2[0], nhat2[1]);
     if (nhat2[2] == 0) info("isotropic environment\n");
     if (nhat2[2] == 1) info("cholesteric environment\n");
-    blue_phase_BPIII_init(q, nhat2);
+    blue_phase_BPIII_init(cs, q, nhat2);
   }
 
   if (strcmp(key1, "cf1_x") == 0) {

### 2014-12-31 branches/kevin/src/field.c

@@ -260,6 +260,7 @@ int field_leesedwards(field_t * obj) {
   double dy;     /* Displacement for current ic->ib pair */
   double fr;     /* Fractional displacement */
   double t;      /* Time */
+  double ltot[3];
 
   const double r6 = (1.0/6.0);
 
@@ -269,6 +270,7 @@ int field_leesedwards(field_t * obj) {
   assert(obj);
   assert(obj->data);
 
+  coords_ltot(obj->cs, ltot);
   coords_cartsz(obj->cs, cartsz);
 
   if (cartsz[Y] > 1) {
@@ -291,7 +293,7 @@ int field_leesedwards(field_t * obj) {
 
       ic = le_index_buffer_to_real(ib);
       dy = le_buffer_displacement(ib, t);
-      dy = fmod(dy, L(Y));
+      dy = fmod(dy, ltot[Y]);
       jdy = floor(dy);
       fr  = 1.0 - (dy - jdy);
 
@@ -364,6 +366,7 @@ static int field_leesedwards_parallel(field_t * obj) {
   double fr;               /* Fractional displacement */
   double t;                /* Time */
   double * buffer;         /* Interpolation buffer */
+  double ltot[3];
   const double r6 = (1.0/6.0);
 
   int      nrank_s[3];     /* send ranks */
@@ -380,6 +383,8 @@ static int field_leesedwards_parallel(field_t * obj) {
   assert(obj);
 
   field_nf(obj, &nf);
+
+  coords_ltot(obj->cs, ltot);
   nhalo = coords_nhalo();
   coords_ntotal(ntotal);
   coords_nlocal(nlocal);
@@ -409,7 +414,7 @@ static int field_leesedwards_parallel(field_t * obj) {
     /* Work out the displacement-dependent quantities */
 
     dy = le_buffer_displacement(ib, t);
-    dy = fmod(dy, L(Y));
+    dy = fmod(dy, ltot[Y]);
     jdy = floor(dy);
     fr  = 1.0 - (dy - jdy);
     /* In the real system the first point we require is

### 2014-12-31 branches/kevin/src/hydro.c

@@ -358,6 +358,8 @@ int hydro_lees_edwards(hydro_t * obj) {
   double dy;     /* Displacement for current ic->ib pair */
   double fr;     /* Fractional displacement */
   double t;      /* Time */
+  double ltot[3];
+
   int jdy;       /* Integral part of displacement */
   int j1, j2;    /* j values in real system to interpolate between */
 
@@ -365,6 +367,7 @@ int hydro_lees_edwards(hydro_t * obj) {
 
   assert(obj);
 
+  coords_ltot(obj->cs, ltot);
   coords_cartsz(obj->cs, cartsz);
 
   if (cartsz[Y] > 1) {
@@ -394,7 +397,7 @@ int hydro_lees_edwards(hydro_t * obj) {
 
       ule[Y] = dy/t; /* STEADY SHEAR ONLY */
 
-      dy = fmod(dy, L(Y));
+      dy = fmod(dy, ltot[Y]);
       jdy = floor(dy);
       fr  = dy - jdy;
 
@@ -456,6 +459,7 @@ static int hydro_lees_edwards_parallel(hydro_t * obj) {
   int index, ia;
   int nf, nhalo;
   double ule[3];
+  double ltot[3];
 
   int      nrank_s[3];     /* send ranks */
   int      nrank_r[3];     /* recv ranks */
@@ -470,6 +474,7 @@ static int hydro_lees_edwards_parallel(hydro_t * obj) {
   assert(obj);
   nf = obj->nf;
 
+  coords_ltot(obj->cs, ltot);
   nhalo = coords_nhalo();
   coords_ntotal(ntotal);
   coords_nlocal(nlocal);
@@ -500,7 +505,7 @@ static int hydro_lees_edwards_parallel(hydro_t * obj) {
 
     dy = le_buffer_displacement(ib, t);
     ule[Y] = dy/t; /* STEADY SHEAR ONLY */
-    dy = fmod(dy, L(Y));
+    dy = fmod(dy, ltot[Y]);
     jdy = floor(dy);
     fr  = dy - jdy;
 

### 2014-12-31 branches/kevin/src/interaction.c

@@ -29,7 +29,6 @@
 
 #include "pe.h"
 #include "util.h"
-#include "coords.h"
 #include "physics.h"
 #include "control.h"
 #include "stats_colloid.h"
@@ -48,6 +47,7 @@ struct interact_s {
   void * abstr[INTERACT_MAX];        /* Abstract interaction types */
   compute_ft compute[INTERACT_MAX];  /* Corresponding compute functions */
   stat_ft stats[INTERACT_MAX];       /* Statisitics functions */
+  free_ft release[INTERACT_MAX];     /* Release child object call back */
 };
 
 /*****************************************************************************
@@ -76,13 +76,19 @@ int interact_create(interact_t ** pobj) {
  *
  *****************************************************************************/
 
-void interact_free(interact_t * obj) {
+int interact_free(interact_t * obj) {
+
+  int it;
 
   assert(obj);
 
+  for (it = 0; it < INTERACT_MAX; it++) {
+    if (obj->release[it]) obj->release[it](obj->abstr[it]);
+  }
+
   free(obj);
 
-  return;
+  return 0;
 }
 
 /*****************************************************************************
@@ -159,6 +165,26 @@ int interact_statistic_add(interact_t * obj, interact_enum_t it, void * pot,
 
   obj->abstr[it] = pot;
   obj->stats[it] = stats;
+
+  return 0;
+}
+
+/*****************************************************************************
+ *
+ *  interact_release_add
+ *
+ *****************************************************************************/
+
+int interact_release_add(interact_t * obj, interact_enum_t it, void * pot,
+			 free_ft release) {
+
+  assert(obj);
+  assert(it >= 0 && it <= INTERACT_MAX);
+  assert(pot);
+  assert(release);
+
+  obj->abstr[it] = pot;
+  obj->release[it] = release;
 
   return 0;
 }

### 2014-12-31 branches/kevin/src/pair_yukawa.h

@@ -16,11 +16,12 @@
 
 typedef struct pair_yukawa_s pair_yukawa_t;
 
+#include "coords.h"
 #include "colloids.h"
 #include "interaction.h"
 
-int pair_yukawa_create(pair_yukawa_t ** pobj);
-void pair_yukawa_free(pair_yukawa_t * obj);
+int pair_yukawa_create(coords_t * cs, pair_yukawa_t ** pobj);
+int pair_yukawa_free(pair_yukawa_t * obj);
 int pair_yukawa_info(pair_yukawa_t * obj);
 int pair_yukawa_param_set(pair_yukawa_t * obj, double epsilon, double kappa,
                           double rc);

### 2014-12-31 branches/kevin/src/polar_active_rt.c

@@ -10,7 +10,7 @@
  *  Edinburgh Parallel Computing Centre
  *
  *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2011 The University of Edinburgh
+ *  (c) 2011-2015 The University of Edinburgh
  *
  *****************************************************************************/
 
@@ -20,7 +20,6 @@
 #include <stdio.h>
 #include <string.h>
 
-#include "coords.h"
 #include "field.h"
 #include "field_grad.h"
 #include "io_harness.h"
@@ -28,7 +27,7 @@
 #include "polar_active.h"
 #include "polar_active_rt.h"
 
-static int polar_active_init_code(field_t * p);
+static int polar_active_init_code(coords_t * cs, field_t * p);
 
 /*****************************************************************************
  *
@@ -91,11 +90,12 @@ int polar_active_run_time(rt_t * rt) {
  *
  *****************************************************************************/
 
-int polar_active_rt_initial_conditions(rt_t * rt, field_t * p) {
+int polar_active_rt_initial_conditions(rt_t * rt, coords_t * cs, field_t * p) {
 
   char key[BUFSIZ];
 
   assert(rt);
+  assert(cs);
   assert(p);
 
   rt_string_parameter(rt, "polar_active_initialisation", key, BUFSIZ);
@@ -107,12 +107,12 @@ int polar_active_rt_initial_conditions(rt_t * rt, field_t * p) {
 
   if (strcmp(key, "from_code") == 0) {
     info("Initial polar order parameter from code\n");
-    polar_active_init_code(p);
+    polar_active_init_code(cs, p);
   }
 
   if (strcmp(key, "aster") == 0) {
     info("Initialise standard aster\n");
-    polar_active_init_aster(p);
+    polar_active_init_aster(cs, p);
   }
 
   return 0;
@@ -126,7 +126,7 @@ int polar_active_rt_initial_conditions(rt_t * rt, field_t * p) {
  *
  *****************************************************************************/
 
[…]

### 2014-12-31 branches/kevin/src/stats_free_energy.c

@@ -58,6 +58,7 @@ int stats_free_energy_density(coords_t * cs, field_t * q, map_t * map,
   double fed;
   double fe_local[NSTAT];
   double fe_total[NSTAT];
+  double ltot[3];
   double rv;
 
   double (* free_energy_density)(const int index);
@@ -67,6 +68,7 @@ int stats_free_energy_density(coords_t * cs, field_t * q, map_t * map,
 
   if (fe_set() == 0) return 0;
 
+  coords_ltot(cs, ltot);
   coords_nlocal(nlocal);
   free_energy_density = fe_density_function();
 
@@ -129,7 +131,7 @@ int stats_free_energy_density(coords_t * cs, field_t * q, map_t * map,
   }
   else {
     MPI_Reduce(fe_local, fe_total, 3, MPI_DOUBLE, MPI_SUM, 0, pe_comm());
-    rv = 1.0/(L(X)*L(Y)*L(Z));
+    rv = 1.0/(ltot[X]*ltot[Y]*ltot[Z]);
 
     info("\nFree energy density - timestep total fluid\n");
     info("[fed] %14d %17.10e %17.10e\n", get_step(), rv*fe_total[0],
@@ -514,6 +516,7 @@ int blue_phase_stats(coords_t * cs, field_t * qf, field_grad_t * dqf,
   double q2, q3, dq0, dq1, sum;
 
   double elocal[14], etotal[14];        /* Free energy contributions etc */
+  double ltot[3];
   double rv;
 
   FILE * fp_output;
@@ -524,8 +527,9 @@ int blue_phase_stats(coords_t * cs, field_t * qf, field_grad_t * dqf,
   assert(dqf);
   assert(map);
 
+  coords_ltot(cs, ltot);
   coords_nlocal(nlocal);
-  rv = 1.0/(L(X)*L(Y)*L(Z));
+  rv = 1.0/(ltot[X]*ltot[Y]*ltot[Z]);
 
   q0 = blue_phase_q0();
   kappa0 = blue_phase_kappa0();

### 2014-12-31 branches/kevin/src/stats_turbulent.c

@@ -228,6 +228,7 @@ int stats_turbulent_ubar_output(stats_turb_t * stat, const char * filename) {
   double * f1z;
   double raverage;
   double uy;
+  double ltot[3];
 
   MPI_Comm comm;
   MPI_Status status;
@@ -237,6 +238,7 @@ int stats_turbulent_ubar_output(stats_turb_t * stat, const char * filename) {
 
   assert(stat);
 
+  coords_ltot(stat->cs, ltot);
   coords_cart_comm(stat->cs, &comm);
   coords_cart_coords(stat->cs, cartcoords);
   coords_cartsz(stat->cs, cartsz);
@@ -251,7 +253,7 @@ int stats_turbulent_ubar_output(stats_turb_t * stat, const char * filename) {
 
   /* Set the averaging factor (if no data, set to zero) */
   raverage = 0.0;
-  if (stat->time_counter > 0) raverage = 1.0/(L(Y)*stat->time_counter); 
+  if (stat->time_counter > 0) raverage = 1.0/(ltot[Y]*stat->time_counter); 
 
   /* Take the sum in the y-direction and store in f1(x,z) */
 

### 2014-12-31 branches/kevin/version.h

@@ -14,6 +14,6 @@
 
 #define LUDWIG_MAJOR_VERSION 0
 #define LUDWIG_MINOR_VERSION 3
-#define LUDWIG_PATCH_VERSION 9
+#define LUDWIG_PATCH_VERSION 10
 
 #endif

### 2015-03-12 trunk/targetDP/doc/data_parallel_execution.tex
added targetDP specification
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2617 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Data Parallel Execution}

\newpage
\section{targetEntry}
\subsection{Description}

The \verb+__targetEntry__+ keyword is used in a function declaration
or definition to specify that the function should be compiled for the
target, and that it will be called directly from host code.

\subsection{Syntax}
\begin{verbatim}
__targetEntry__ functionReturnType functionName(...
\end{verbatim}

\begin{itemize}
\item \verb+functionName+: The name of the function to be compiled for the target.
\item \verb+functionReturnType+: The return type of the function.
\item \verb+...+ the remainder of the function declaration or definition.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalekernel:targetEntry} in Figure \ref{fig:scalekernel} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
Holds no value.
\subsubsection{CUDA}

\newpage
\section{target}
\subsection{Description}

The \verb+__target__+ keyword is used in a function declaration
or definition to specify that the function should be compiled for the
target, and that it will be called from a \verb+targetEntry+ or another \verb+target+ function.

\subsection{Syntax}
\begin{verbatim}
__target__ functionReturnType functionName(...
\end{verbatim}

\begin{itemize}
\item \verb+functionName+: The name of the function to be compiled for the target.
\item \verb+functionReturnType+: The return type of the function.
\item \verb+...+ the remainder of the function declaration or definition.
\end{itemize}

\subsection{Example}
Analogous to Line \ref{example:scalekernel:targetEntry} in Figure \ref{fig:scalekernel} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
Holds no value.
\subsubsection{CUDA}

\newpage
\section{targetHost}
\subsection{Description}

The \verb+__targetHost__+ keyword is used in a function declaration
or definition to specify that the function should be compiled for the
host.

\subsection{Syntax}
\begin{verbatim}
__targetHost__ functionReturnType functionName(...
\end{verbatim}

\begin{itemize}
\item \verb+functionName+: The name of the function to be compiled for the host.
\item \verb+functionReturnType+: The return type of the function.
\item \verb+...+ the remainder of the function declaration or definition.
\end{itemize}

\subsection{Example}
Analogous to Line \ref{example:scalekernel:targetEntry} in Figure \ref{fig:scalekernel} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
Holds no value.
\subsubsection{CUDA}
\verb+extern ``C'' __host__+

\newpage
\section{targetLaunch}

\subsection{Description}

The \verb+__targetLaunch__+ syntax is used to launch a function across a data parallel target architecture.

\subsection{Syntax}
\begin{verbatim}
functionName __targetLaunch__(size_t extent) \
                 (functionArgument1,functionArgument2,...);
\end{verbatim}

\begin{itemize}
\item \verb+functionName+: The name of the function to be launched. This function must be declared as \verb+__targetEntry__+ .
\item \verb+functionArguments+: The arguments to the function \verb+functionName+  
\item \verb+extent+: The total extent of data parallelism.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalelaunch:targetLaunch} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
Holds no value.
\subsubsection{CUDA}
CUDA \verb+<<<...>>>+ syntax.

\newpage
\section{targetSynchronize}

\subsection{Description}

The \verb+targetSynchronize+ function is used to block until the preceeding \verb+__targetLaunch__+ has completed.

\subsection{Syntax}
\begin{verbatim}
void targetSynchronize();
\end{verbatim}

\subsection{Example}
See Line \ref{example:scalelaunch:targetSynchronize} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.
\subsection{Implementation}
\subsubsection{C}
Dummy function.
\subsubsection{CUDA}
\verb+cudaThreadSynchronize+

%% \newpage
%% \section{targetErrorCheck}

%% \subsection{Description}

%% The \verb+__targetErrorCheck__+ function is used to query for errors resulting from operations involving the target. The

%% \subsection{Syntax}
%% \begin{verbatim}
%% __targetSynchronize__()
%% \end{verbatim}

%% \subsection{Example}

%% \subsection{Implementation}
%% \subsubsection{C}

%% \subsubsection{CUDA}

\newpage
\section{targetTLP}

\subsection{Description}

The \verb+__targetTLP__+ syntax is used, within a \verb+__targetEntry__+
function, to specify that the proceeding block of code should be
executed in parallel and mapped to thread level parallelism
(TLP). Note that he behaviour of this operation depends on the defined
virtual vector length (VVL), which controls the lower-level
Instruction Level Parallelsim (ILP) (see following section).

\subsection{Syntax}
\begin{verbatim}
__targetTLP__(int baseIndex, size_t extent) 
{
//code to be executed in parallel
}
\end{verbatim}

\begin{itemize}
\item \verb+extent+: The total extent of data parallelism, including
  both TLP and ILP
\item \verb+baseIndex+: the TLP index. This will vary from 0 to \verb+extent-VVL+ with stride VVL. This index should be combined with the ILP index to access shared arrays within the code block (see following section).

\end{itemize}

\subsection{Example}
See Line \ref{example:scalekernel:targetTLP} in Figure \ref{fig:scalekernel} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
OpenMP parallel loop.
\subsubsection{CUDA}
CUDA thread lookup.

\newpage
\section{targetILP}

\subsection{Description}

The \verb+__targetILP__+ syntax is used, within a \verb+__targetTLP__+
region, to specify that the proceeding block of code should be
executed in parallel and mapped to instruction level parallelism
(ILP), where the extent of te ILP is defined by the virtual vector length (VVL) in the targetDP implementation. 
\begin{comment}
need to document VVL in more detail somewhere and link.
\end{comment}

\subsection{Syntax}
\begin{verbatim}
__targetILP__(int vecIndex) 
{
//code to be executed in parallel
}
\end{verbatim}

\begin{itemize}
\item \verb+baseIndex+: the ILP index. This will vary from 0 to \verb+VVL-1+.  This index should be combined with the TLP index to access shared arrays within the code block (see previous section).
\end{itemize}

\subsection{Example}
See Line \ref{example:scalekernel:targetILP} in Figure \ref{fig:scalekernel} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
Short vectorizable loop.
\subsubsection{CUDA}
Short vectorizable loop.

\newpage
\section{targetCoords3D}

\subsection{Description}

The \verb+targetCoords3D+ function provides the 3D lattice coodinates corresponding to a specified linear index.

\subsection{Syntax}
\begin{verbatim}
void targetCoords3D(int coords3D[3], int extent3D[3], int index);
\end{verbatim}

\begin{itemize}
\item \verb+coords3D+ (output): an array of 3 integers to be populated with ther 3D coordinates.
\item \verb+extent3D+ (input): An array of 3 integers corresponding to the 3D dimensions of the lattice.
\item \verb+index+ (input): the linear index.

\end{itemize}

\subsection{Example}
**TO DO**
\subsection{Implementation}
\subsubsection{C}
**TO DO**
\subsubsection{CUDA}
**TO DO**

\newpage
\section{targetIndex3D}

\subsection{Description}

The \verb+targetIndex3D+ function returns the linear index corresponding to a specified set of 3D lattice coodinates.

\subsection{Syntax}
\begin{verbatim}
int targetIndex3D(int Xcoord,int Ycoord,int Zcoord,int extent3D[3]);
\end{verbatim}

\begin{itemize}
\item \verb+Xcoord+ (input): the specified coordinate in the X direction.
\item \verb+Ycoord+ (input): the specified coordinate in the Y direction.
\item \verb+Zcoord+ (input): the specified coordinate in the Z direction.
\item \verb+extent3D+ (input): an array of 3 integers corresponding to the 3D dimensions of the lattice.
\end{itemize}

\subsection{Example}
**TO DO**
\subsection{Implementation}
\subsubsection{C}
**TO DO**
\subsubsection{CUDA}
**TO DO**

### 2015-03-12 trunk/targetDP/doc/examples.tex
added targetDP specification
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2617 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Examples}\label{chapter:examples}
\begin{comment}
TO DO: describe this example. 
\end{comment}
\begin{figure}[h]
\begin{lstlisting}
__targetEntry__ void scale(double* t_field) { (*@\label{example:scalekernel:targetEntry}@*)

  int baseIndex;
  __targetTLP__(baseIndex, N) { (*@\label{example:scalekernel:targetTLP}@*)

    int iDim, vecIndex;
    for (iDim = 0; iDim < 3; iDim++) {

      __targetILP__(vecIndex)  \ (*@\label{example:scalekernel:targetILP}@*) 
         t_field[iDim*N + baseIndex + vecIndex] =  \
	 t_a*t_field[iDim*N + baseIndex + vecIndex];       	  
    }
  }
  return;
}
\end{lstlisting}
\caption{***}\label{fig:scalekernel}
\end{figure}

\begin{figure}[h]
\begin{lstlisting}
  targetMalloc((void **) &t_field, datasize); (*@\label{example:scalelaunch:targetMalloc}@*)

  copyToTarget(t_field, field, datasize); (*@\label{example:scalelaunch:copyToTarget}@*) 
  copyConstToTarget(&t_a, &a, sizeof(double)); (*@\label{example:scalelaunch:copyConstToTarget}@*)

  scale __targetLaunch__(N) (t_field); (*@\label{example:scalelaunch:targetLaunch}@*) 
  targetSynchronize();(*@\label{example:scalelaunch:targetSynchronize}@*)

  copyFromTarget(field, t_field, datasize); (*@\label{example:scalelaunch:copyFromTarget}@*) 
  targetFree(t_field); (*@\label{example:scalelaunch:targetFree}@*) 
\end{lstlisting}
\caption{***}\label{fig:scalelaunch}
\end{figure}

### 2015-03-12 trunk/targetDP/doc/introduction.tex
added targetDP specification
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2617 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Introduction}

\begin{comment}
TO DO.
\end{comment}
\section{Abstract}

**

\section{Motivation}

**

\section{Glossary}

\begin{itemize}

    \item \keyword{Accelerator}: ***
    \item \keyword{Host}: ***
    \item \keyword{Target}: ***
    \item \keyword{CUDA}: ***

\end{itemize}

### 2015-03-12 targetDP/doc/execution_model.tex
added targetDP specification
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2617 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Execution model}

The terminology ``host'' is used to refer to the CPU that is hosting
the execution of the application, and ``target'' to refer to the
device targeted for execution of data parallel operations. The target
can be the same CPU as the host, or it can be a separate device such
as an accelerator.

The targetDP API follows the fork-join model of parallel
execution. When the application initiates, a single thread executes
sequentially on the host, until it encounters a function to be
launched on the target. This function will be executed by a team of
threads on the target cooperating in a data-parallel manner (e.g. for
a structured grid problem each thread is responsible for a subset of
the grid).

To ensure that the target function has completed, a
\verb+targetSynchronize+ statement should follow the code location
where target function is launched. When the initial thread encounters
this statement, it will wait until the target region has completed. It
is possible for the initial thread to execute other instructions
(which do not depend on the results of the target function), after the
function launch but before the syncronisation call. This may result in
overlapping of host and target execution, and hence optimisation, in
some implementations. Once the target function has completed, the
initial thread will continue sequentially until another target
function launch is encountered.

Within each target function, each thread is given a unique index which
it uses to work in a data-parallel manner. Each thread works
independently from all others, but usually operating on a shared data
structure where the index is used to determine the portion of data
to process.

\begin{comment}
Possibly add targetSyncThreads() to
  allow communication within target functions?]
\end{comment}

### 2015-03-12 targetDP/doc/memory_management.tex
added targetDP specification
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2617 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Memory Management}

\newpage
\section{targetMalloc}

\subsection{Description}

The \verb+targetMalloc+ function allocates memory on the target.

\subsection{Syntax}
\begin{verbatim}
void targetMalloc(void **targetPtr, size_t n);
\end{verbatim}

\begin{itemize}
\item \verb+targetptr+: A pointer to the allocated memory.
\item \verb+n+: The number of bytes to be allocated.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalelaunch:targetMalloc} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.
\subsection{Implementation}
\subsubsection{C}
\verb+malloc+
\subsubsection{CUDA}
\verb+cudaMalloc+

\newpage
\section{targetCalloc}

\subsection{Description}

The \verb+targetCalloc+ function allocates, and initializes to zero, memory on the target.

\subsection{Syntax}
\begin{verbatim}
void targetCalloc(void **targetPtr, size_t n);
\end{verbatim}

\begin{itemize}
\item \verb+targetptr+: A pointer to the allocated memory.
\item \verb+n+: The number of bytes to be allocated.
\end{itemize}

\subsection{Example}
Analogous to Line \ref{example:scalelaunch:targetMalloc} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
\verb+calloc+
\subsubsection{CUDA}
\verb+cudaMalloc+ followed by \verb+cudaMemset+

\newpage
\section{targetFree}

\subsection{Description}

The \verb+targetFree+ function deallocates memory on the target.

\subsection{Syntax}
\begin{verbatim}
void targetFree(void *targetPtr);
\end{verbatim}

\begin{itemize}
\item \verb+targetPtr+: A pointer to the memory to be freed.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalelaunch:targetFree} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
\verb+free+
\subsubsection{CUDA}
\verb+cudaFree+

\newpage
\section{copyToTarget}

\subsection{Description}

The \verb+copyToTarget+ function copies data from the host to the target.

\subsection{Syntax}
\begin{verbatim}
void copyToTarget(void *targetData, const void *data, size_t n);
\end{verbatim}

\begin{itemize}
\item \verb+targetData+: A pointer to the destination array on the target.
\item \verb+data+: A pointer to the source array on the host.
\item \verb+n+: The number of bytes to be copied.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalelaunch:copyToTarget} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
\verb+memcpy+
\subsubsection{CUDA}
\verb+cudaMemcpy+

\newpage
\section{copyFromTarget}

\subsection{Description}

The \verb+copyFromTarget+ function copies data from the target to the host.

\subsection{Syntax}
\begin{verbatim}
void copyFromTarget(void *data, const void *targetData, size_t n);
\end{verbatim}

\begin{itemize}
\item \verb+data+: A pointer to the destination array on the host.
\item \verb+targetData+: A pointer to the source array on the target.
\item \verb+n+: The number of bytes to be copied.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalelaunch:copyFromTarget} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
\verb+memcpy+
\subsubsection{CUDA}
\verb+cudaMemcpy+

\newpage
\section{targetConst}
\subsection{Description}

The \verb+__targetConst__+ keyword is used in a variable or array declaration
to specify that the corresponding data can be treated as constant (read-only) on the target.

\subsection{Syntax}
\begin{verbatim}
__targetConst__ type variableName
\end{verbatim}

\begin{itemize}
\item \verb+variableName+: The name of the variable or array.
\item \verb+type+: The type of variabe or array.
\end{itemize}

\subsection{Example}
**TO DO**.
\subsection{Implementation}
\subsubsection{C}
Holds no value
\subsubsection{CUDA}
\verb+__constant__+

\newpage
\section{copyConstToTarget}

\subsection{Description}

The \verb+copyConstToTarget+ function copies data from the host to the target, where the data will remain constant (read-only) during the execution of functions on the target.

\subsection{Syntax}
\begin{verbatim}
void copyConstToTarget(void *targetData, const void *data, size_t n);
\end{verbatim}

\begin{itemize}
\item \verb+targetData+: A pointer to the destination array on the target. This must have been declared using the \verb+__targetConst__ + keyword.
\item \verb+data+: A pointer to the source array on the host.
\item \verb+n+: The number of bytes to be copied.
\end{itemize}

\subsection{Example}
See Line \ref{example:scalelaunch:copyConstToTarget} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.
\subsection{Implementation}
\subsubsection{C}
\verb+memcpy+
\subsubsection{CUDA}
\verb+cudaMemcpyToSymbol+

\newpage
\section{copyConstFromTarget}

\subsection{Description}

The \verb+copyConstFromTarget+ function copies data from a constant data location on the target to the host.

\subsection{Syntax}
\begin{verbatim}
void copyConstToTarget(void *targetData, const void *data, size_t n);
\end{verbatim}

\begin{itemize}
\item \verb+data+: A pointer to the destination array on the host.
\item \verb+targetData+: A pointer to the source array on the target. This must have been declared using the \verb+__targetConst__ + keyword.
\item \verb+n+: The number of bytes to be copied.
\end{itemize}

\subsection{Example}
Analogous to Line \ref{example:scalelaunch:copyConstToTarget} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.

\subsection{Implementation}
\subsubsection{C}
\verb+memcpy+
\subsubsection{CUDA}
\verb+cudaMemcpyFromSymbol+

\newpage
\section{targetConstAddress}

\subsection{Description}

The \verb+targetConstAddress+ function provides the target address for a constant object.

\subsection{Syntax}
\begin{verbatim}
void targetConstAddress(void **address, objectType object);
\end{verbatim}

\begin{itemize}
\item \verb+address+ (output): The pointer to the constant object on the target.
\item \verb+objectType+: The type of the object.
\item \verb+object+ (input): The constant object on the target. This should have been declared using the \verb+__targetConst__+ keyword.
\end{itemize}

\subsection{Example}
**TO DO**.

\subsection{Implementation}
\subsubsection{C}
Explicit copying of address.
\subsubsection{CUDA}
\verb+cudaGetSymbolAddress+

\newpage
\section{targetInit3D}

\subsection{Description}

The \verb+targetInit3D+ initialises the environment required to perform any of following the targetDP 3D lattice operations.

\subsection{Syntax}
\begin{verbatim}
void targetInit3D(size_t extent, size_t nFields);
\end{verbatim}

\begin{itemize}
\item \verb+extent+: The total extent of data parallelism (e.g. the number of lattice sites).
\item \verb+nFields+: The extent of data resident within each parallel partition (e.g. the number of fields per lattice site).
\end{itemize}

\subsection{Example}
**TO DO**.
\subsection{Implementation}
\subsubsection{C}
**TO DO**.
\subsubsection{CUDA}
**TO DO**.

\newpage
\section{targetFinalize3D}

\subsection{Description}

The \verb+targetFinalize3D+ finalizes the targetDP 3D environment.

\subsection{Syntax}
\begin{verbatim}
void targetFinalize3D();
\end{verbatim}

\subsection{Example}
**TO DO**.

\subsection{Implementation}
\subsubsection{C}
**TO DO**.
\subsubsection{CUDA}
**TO DO**.

\newpage
\section{copyToTargetBoundary3D}

\subsection{Description}

The \verb+copyToTargetBoundary3D+ function copies the data corresponding to the boundaries of a 3D lattice from the host to the target.

\subsection{Syntax}
\begin{verbatim}
void copyToTargetBoundary3D(void *targetData, const void *data, size_t extent3D[3], size_t nField, size_t offset, size_t depth);
\end{verbatim}

\begin{itemize}
\item \verb+targetData+: A pointer to the destination array on the target.
\item \verb+data+: A pointer to the source array on the host.
\item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
\item \verb+nFields+: The number of fields per lattice site.
\item \verb+offset+: The number of sites from the lattice edge at which each boundary face should start.
\item \verb+depth+: The depth of each boundary face.
\end{itemize}

\subsection{Example}
**TO DO**.
\subsection{Implementation}
\subsubsection{C}
**TO DO**.
\subsubsection{CUDA}
**TO DO**.

\newpage
\section{copyFromTargetBoundary3D}

\subsection{Description}

The \verb+copyFromTargetBoundary3D+ function copies the data corresponding to the boundaries of a 3D lattice from the target to the host.

\subsection{Syntax}
\begin{verbatim}
void copyFromTargetBoundary3D(void *data, const void *targetData, size_t extent3D[3], size_t nField, size_t offset, size_t depth);
\end{verbatim}

\begin{itemize}
\item \verb+data+: A pointer to the destination array on the host.
\item \verb+targetData+: A pointer to the source array on the target.
\item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
\item \verb+nFields+: The number of fields per lattice site.
\item \verb+offset+: The number of sites from the lattice edge at which each boundary face should start.
\item \verb+depth+: The depth of each boundary face.
\end{itemize}

\subsection{Example}
**TO DO**.
\subsection{Implementation}
\subsubsection{C}
**TO DO**.
\subsubsection{CUDA}
**TO DO**.
\newpage
\section{copyToTargetPointerMap3D}

\subsection{Description}

The \verb+copyToTargetPointerMap3D+ function copies a subset of lattice data from the host to the target. The sites to be included are defined using an array of pointers passed as input.

\subsection{Syntax}
\begin{verbatim}
void copyToTargetPointerMap3D(void *targetData, const void *data, 
           size_t extent3D[3], size_t nField, 
           int includeNeighbours, void** pointerArray);
\end{verbatim}

\begin{itemize}
\item \verb+targetData+: A pointer to the destination array on the target.
\item \verb+data+: A pointer to the source array on the host.
\item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
\item \verb+nField+: The number of fields per lattice site.
\item \verb+includeNeighbours+: A boolean switch to specify whether each included site should also have it's neighbours included (in the 19-point 3D stencil).
\item \verb+pointerArray+: An array of \verb+nSite+ pointers, where \verb+nSite+ is the total number of lattice sites. Each lattice site should be included unless the pointer corresponding to that site is \verb+NULL+.  
\end{itemize}

\subsection{Example}
**TO DO**.
\subsection{Implementation}
\subsubsection{C}
**TO DO**.
\subsubsection{CUDA}
**TO DO**.

\newpage
\section{copyFromTargetPointerMap3D}

\subsection{Description}

The \verb+copyFromTargetPointerMap3D+ function copies a subset of lattice data from the target to the host. The sites to be included are defined using an array of pointers passed as input.

\subsection{Syntax}
\begin{verbatim}
void copyFromTargetPointerMap3D(void *data, const void *targetData, 
           size_t extent3D[3], size_t nField, 
           int includeNeighbours, void** pointerArray);
\end{verbatim}

\begin{itemize}
\item \verb+data+: A pointer to the destination array on the host.
\item \verb+targetData+: A pointer to the source array on the target.
\item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
\item \verb+nField+: The number of fields per lattice site.
\item \verb+includeNeighbours+: A boolean switch to specify whether each included site should also have it's neighbours included (in the 19-point 3D stencil).
\item \verb+pointerArray+: An array of \verb+nSite+ pointers, where \verb+nSite+ is the total number of lattice sites. Each lattice site should be included unless the pointer corresponding to that site is \verb+NULL+.  
\end{itemize}

\subsection{Example}
**TO DO**.

\subsection{Implementation}
\subsubsection{C}
**TO DO**.
\subsubsection{CUDA}
**TO DO**.

### 2015-04-14 trunk/targetDP/doc/introduction.tex
more complete draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2632 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\begin{comment}
-TO DO.
-\end{comment}
+%\begin{comment}
+%TO DO.
+%\end{comment}
 \section{Abstract}
 
-**
+The targetDP API provides an abstraction layer which allows
+applications to \underline{t}arget \underline{D}ata
+\underline{P}arallel hardware in a platform agnostic manner, by
+abstracting the memory spaces and hierarchy of hardware
+parallelism. Applications written using targetDP syntax are
+performance portable: the same source code can be compiled for
+different targets (currently supported are accelerators such as GPUs
+and modern multi-core SIMD CPUs), without performance overheads. The
+targetDP API is aimed at the types of parallelism found in grid-based
+applications, but may be applicable to a wider class of problems. This
+document introduces the targetDP memory and execution models, and
+specifies the syntax and functionality of the
+interface. Implementation details are also provided, and a simple
+example is given to demonstrate usage.
 
 \section{Motivation}
-
-**
+It is becoming increasingly difficult for applications to exploit
+modern computers, which continue to increase in complexity and
+diversity with features including multiple multicore CPUs (often
+distributed across multiple nodes), vector floating point units,
+accelerators such as GPUs and non-uniform distributed memory spaces.
+From a scientist's perspective, it is not only imperative to achieve
+performance, but also to retain maintainability, sustainability and
+portability. The use of a simplistic, well structured and clearly
+defined abstraction layer such as targetDP can allow the programmer to
+express the scientific problems in a way that will automatically
+achieve good performance across the range of leading hardware
+solutions.
 
@@
 
-    \item \keyword{Accelerator}: ***
-    \item \keyword{Host}: ***
-    \item \keyword{Target}: ***
-    \item \keyword{CUDA}: ***
+    \item \keyword{CPU}: Central Processing Unit. The main computer chip used in a system, suitable for a wide variety of computational tasks.
+    \item \keyword{Accelerator}: A processing unit which is not used in isolation, but instead in tandem with the CPU, with the aim of improving the performance of key code sections. 
+    \item \keyword{GPU}: Graphics Processing Unit.
+A type of accelerator, originally evolved to render graphics content (particularly to satisfy demands of the gaming industry), but now widely used for general purpose computation.
+    \item \keyword{Host}: Another term for the CPU that ``hosts'' the application.
+    \item \keyword{Data Parallel}: The type of algorithmic parallelism involved where a single operation is performed to each element of a data set.
+The extent of parallelism is determined by the size of the data set.
+    \item \keyword{Target}: The device targeted for execution of data parallel operations.
+Depending on the underlying hardware available, the target could simply be the a CPU, or it could be a separate device such as an accelerator.
+    \item \keyword{CUDA}: Compute Unified Device Architecture. The parallel platform and model created by NVIDIA to allow general purpose programming of their GP architectures.
+    \item \keyword{TLP}: Thread Level Parallelism.
+    \item \keyword{ILP}: Instruction Level Parallelism.
 

### 2015-04-14 targetDP/doc/data_parallel_execution.tex
more complete draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2632 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-03-12 trunk/targetDP/doc/data_parallel_execution.tex
@@
 \subsubsection{CUDA}
+\verb+__global__+
 
@@
 \subsubsection{CUDA}
+\verb+__device__+
 
@@
 
-The \verb+targetSynchronize+ function is used to block until the preceeding \verb+__targetLaunch__+ has completed.
+The \verb+targetSynchronize+ function is used to block until the preceding \verb+__targetLaunch__+ has completed.
 
@@
 \newpage
-\section{targetTLP}
+\section{targetTLP}\label{sec:TLP}
 
@@
 virtual vector length (VVL), which controls the lower-level
-Instruction Level Parallelsim (ILP) (see following section).
+Instruction Level Parallelism (ILP) (see following section).
 
@@
 \newpage
-\section{targetILP}
+\section{targetILP}\label{sec:ILP}
 
@@
 executed in parallel and mapped to instruction level parallelism
-(ILP), where the extent of te ILP is defined by the virtual vector length (VVL) in the targetDP implementation. 
+(ILP), where the extent of the ILP is defined by the virtual vector length (VVL) in the targetDP implementation. 
 \begin{comment}
@@
 
-The \verb+targetCoords3D+ function provides the 3D lattice coodinates corresponding to a specified linear index.
+The \verb+targetCoords3D+ function provides the 3D lattice coordinates corresponding to a specified linear index.
 
@@
 \begin{itemize}
-\item \verb+coords3D+ (output): an array of 3 integers to be populated with ther 3D coordinates.
+\item \verb+coords3D+ (output): an array of 3 integers to be populated with the 3D coordinates.
 \item \verb+extent3D+ (input): An array of 3 integers corresponding to the 3D dimensions of the lattice.
@@
 
-The \verb+targetIndex3D+ function returns the linear index corresponding to a specified set of 3D lattice coodinates.
+The \verb+targetIndex3D+ function returns the linear index corresponding to a specified set of 3D lattice coordinates.
 

### 2015-05-14 incite/Milestone_Table_template.tex
adding incite bid templates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2645 40876441-3496-0410-a613-a2fca35b2ae2

%\documentclass[12pt,letterpaper,english]{article}
\documentclass[11pt,landscape]{article}
\usepackage[T1]{fontenc} % Standard package for selecting font encodingsamely titles, headers and contents
\usepackage{txfonts} % makes spacing between characters space correctly
\usepackage{xcolor} % Driver-independent color extensions for LaTeX and pdfLaTeX.
%\usepackage[pagestyles,raggedright]{titlesec} % related with sections—n
%\usepackage{blindtext} % To create text
\usepackage{fancyhdr} % header footer placement

\usepackage[top=1in, bottom=1in, left=1in, right=1in] {geometry} % Margins
\usepackage{graphicx}  % Essential for adding images to you document.

\usepackage{sectsty}
\sectionfont{\large}
\subsectionfont{\normalsize}
\subsubsectionfont{\normalsize \it}

\usepackage{caption}
\captionsetup{labelsep=period}

\renewcommand{\arraystretch}{1.25}

\pagestyle{fancy} % allows you to use the header and footer commands

\raggedright
\begin{document}

\setlength{\parindent}{0in} % Amount of indentation at the first line of a paragraph.

\pagestyle{fancy} \lhead{Title} \rhead{Lead PI} \renewcommand{%
\headrulewidth}{0.0pt}

\colorbox{yellow} {\bf {\emph{[Refer to the guidelines for instructions in preparing the proposal. Table does not count toward}}}
\colorbox{yellow} {\bf {\emph{[15-page project narrative count.]}}}

\bigskip

\textbf{Proposal Title (exactly as it appears on submission):} Text

\begin{table}[h]
%\centering
%\caption{Table title}
%\vspace{-.15in}
\label{Tab1}
\begin{tabular}{p{5cm} p{5cm} p{5cm}} \\ \hline 
\noalign{\smallskip}
\multicolumn{3}{l}{\textbf {Year 1}}\\ \hline 
\noalign{\smallskip}
\textbf{Milestone} & \textbf{Details} & \textbf{Date and Status} \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline 
\noalign{\smallskip}
\multicolumn{3}{l}{\textbf {Year 2 (if appropriate)}}\\ \hline
\noalign{\smallskip}
Insert Text & Insert Text & Insert Text \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline 
\noalign{\smallskip}
\multicolumn{3}{l}{\textbf {Year 3 (if appropriate)}}\\ \hline
\noalign{\smallskip}
Insert Text & Insert Text & Insert Text \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline 
Insert Text & Insert Text & Insert Text \\ \hline

\end{tabular}
\end{table}

\end{document}

### 2015-05-14 incite/Project_Narrative_template.tex
adding incite bid templates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2645 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[11pt,letterpaper,english]{article}
\usepackage[T1]{fontenc} % Standard package for selecting font encodings
\usepackage{txfonts} % makes spacing between characters space correctly
\usepackage{xcolor} % Driver-independent color extensions for LaTeX and pdfLaTeX.
\usepackage{hyperref}  %The ability to create hyperlinks within the document
%\usepackage{blindtext} % To create text
%\usepackage{mdwlist} % mdwlist for compact enumeration/list items
%\usepackage[pagestyles]{titlesec} % related with sections—namely titles, headers and contents
\usepackage{fancyhdr} % header footer placement

\usepackage[top=1in, bottom=1in, left=1in, right=1in] {geometry} % Margins
\usepackage{graphicx}   % Essential for adding images to you document.

\usepackage{sectsty}
\sectionfont{\normalsize}
\subsectionfont{\normalsize}
\subsubsectionfont{\normalsize \it}

\usepackage[font=bf]{caption}
\captionsetup{labelsep=period}

\setlength{\parskip}{\baselineskip}%
\setlength{\parindent}{0pt}%

\pagestyle{fancy} % allows you to use the header and footer commands

\raggedright
\begin{document}

\setlength{\parindent}{0in} % Amount of indentation at the first line of a paragraph.

\pagestyle{fancy} \lhead{Title} \rhead{Lead PI} \renewcommand{%
\headrulewidth}{0.0pt}

\begin{center}
{\bf PROJECT NARRATIVE (The narrative should not exceed 15 pages)} 
\end{center}

\vspace{-.15in}

%\colorbox{yellow} {\bf {\emph{[Refer to the guidelines for additional instructions for preparing the proposal.]}}}

%\vspace{.15in}

{\bf References and visual materials, such as charts, graphs, pictures, etc., are included in the 15 page limit.} References should be included at the end of the Project Narrative. URLs that provide information related to the proposal should not be included  {\bf The 15-page limit will be strictly enforced.}  The Project Narrative should address the following points:

\vspace{-.25in}
\section{SIGNIFICANCE OF RESEARCH}
\vspace{-.2in}
List any previous INCITE award(s) received and discuss the relationship to the work proposed. Explain what advances you expect to be enabled by an INCITE award that justifies an allocation of petascale resources (e.g., anticipated impact on community paradigms, valuable insights into or solving a long-standing challenge, etc.). Place the proposed research in the context of competing work in your discipline or business. The information should be sufficient for peer review in your area of research and also appropriate for general scientific review comparing your proposal with proposals in other disciplines. Potential impact is the predominant determinant for awards. This factor will be assessed by a peer-review panel. Also list any previous INCITE award(s) received and discuss the relationship to the work proposed. {\bf This section is typically about 4 pages.}

When you cite a reference, please insert a number in brackets, as shown here [1], to correspond to its number on the reference list at the end of this template. Please call out the references in numerical order and list them in the same way. The reference list will count toward the 15-page limit.

Insert paragraph(s).
\vspace{-.25in}
\subsection{Heading 2 (optional)}
\vspace{-.2in}

Insert paragraph(s).

\vspace{-.25in}
\subsubsection{Heading 3 (optional)}
\vspace{-.2in}

Insert paragraph(s).

\vspace{-.25in}
\section{RESEARCH OBJECTIVES AND MILESTONES }  
\vspace{-.2in}
Describe the proposed research, including its goals and milestones and the theoretical and computational methods it employs. Goals and milestones should articulate simulation and developmental objectives and be sufficiently detailed to assess the progress of the project for each year of any allocation granted. Milestones should correlate with those in the milestone table. It is especially important that you provide clear connections between the project's overarching milestones, the planned production simulations, and the compute time expected to be required for these simulations (e.g., should correlate with those in the ``Use of Resources Requested''). {\bf This section is typically about 6 pages.}

If included, call out equations, tables, figures, and references in numerical order in text, such as Eqs. (\ref{Eq. 1}) and (\ref{Eq. 2}), Table \ref{Tab1}, and Fig. \ref{Fig1} below.

\vspace{-.15in}
\begin{equation} \label{Eq. 1} 
\partial ,\phi +u\cdot \nabla \phi =\nabla ^{2} \phi +\frac{1}{\tau } R\left(\phi \right)\, \, ,
\end{equation}

\vspace{-.15in}
\begin{equation} \label{Eq. 2} 
\frac{\partial \phi }{\nabla \phi } =\frac{1}{2} \nabla ^{2} \phi \frac{e^{\frac{-R-R^{2} }{2u} } }{\left(2\tau \right)^{3N/2} } \sqrt{xyz} \, \, \sum 1+23\, \, \, . 
\end{equation}

\begin{table}[h]
\centering
\caption{Table title}
\vspace{-.15in}
\label{Tab1}
\begin{tabular}{llll} \\ \hline 
\textbf{Column one} & \textbf{Column two} & \textbf{Column three} & \textbf{Column four} \\ \hline 
 xxx & xxx\textit{${}^{a}$} & xxx & xxx \\ \hline 
xxx & xxx & xxx & xxx \\ \hline 
{\textit{$^{a}$}Footnote here.} \\
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\includegraphics[width=4.52in, height=3.08in, keepaspectratio=true]{image1.jpg}
\caption{Figure caption.}
\label{Fig1}
\end{figure}

\vspace{-.25in}
\subsection{Heading 2 (optional)}
\vspace{-.2in}

Insert paragraph(s).

\vspace{-.25in}
\subsubsection{Heading 3 (optional)}
\vspace{-.2in}

Insert paragraph(s).

\vspace{-.25in}
\section{COMPUTATIONAL READINESS}
\vspace{-.2in}

Proposals will be assessed on the need for, readiness to use, and reasonableness of the request for resources. {\bf This section is typically about 5 pages.}

Insert paragraph(s).

\vspace{-.25in}
\subsection{Use of Resources Requested}
\vspace{-.2in}

Describe your proposed production simulations and state how the runs are tied to each of your project's goals and milestones (Section 4, "Milestone Table"). For the simulations you plan to carry out during production runs, provide a
\begin{enumerate}
\item description of what jobs are going to be run and how they relate to the research/development objectives given above;
\item description of processor/core use for large runs (e.g., 10,000-hour run with 100 cores, or ten 10-hour runs with 10,000 cores, for a 1,000,000-hour allocation); 
\item clear, detailed explanation as to how you calculated the requested number of processor hours; and 
\item summary of your anticipated annual burn rate (e.g., linear or with periods of peak usage).\\
\vspace{.1in}
In addition, describe the data storage requirements required by your production simulations.  If at any point during your project the sum of your data storage needs in the scratch filesystems exceed 1 petabyte, specific justification is required. For your production simulations, provide a:

\item Description of any input data set required, its size, and how you anticipate transferring it to the LCF.
\item Estimate and breakdown of the anticipated cumulative size of stored data, in scratch and long-term archival storage, at the end of the requested award. 
\item What is the effective lifetime of your useful data?  If the lifetime varies, show the breakdown by the total size used.  Explain the reason for the lifetime. 
\item What types of tools for data storage, compression (reduction), movement, and analysis do you currently use? Are the tools and/or applications needed ready or are there new capabilities or features you are developing?
\item If intending to make any fraction of the data generated public, please specify:
\begin{enumerate}
\item How much data and the scientific purpose
\item What tool will be used to share the data
\item From where will the data be shared\\
\vspace{.1in}
NOTE: The LCF data management policies can be found at

OLCF:  {\href{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}}

ALCF:  {\href{http://www.alcf.anl.gov/user-guides/data-policy}{http://www.alcf.anl.gov/user-guides/data-policy}}
\end{enumerate}
\end{enumerate}

\vspace{-.25in}
\subsection{Leadership Classification}
\vspace{-.2in}

Summarize the requirement that best exemplifies the proposed computational work. Leadership targets in the INCITE program typically include one or both of the following categories:

\vspace{-.25in}
\begin{enumerate}
\item Use of 20 percent or more of the system for production simulations. Parameter sweeps, ensembles, design of experiments, and other statistical methods that require large numbers of discrete or loosely coupled simulations may be considered capability-class campaigns. See the FAQS for details and qualifiers.
\item Specific architectural needs that can only be met by the LCF.
\end{enumerate}
\vspace{-.2in}
Insert paragraph(s).

\vspace{-.25in}
\subsection{Parallel Performance}
\vspace{-.2in}

Provide direct evidence, {\bf including supporting quantitative data}, for your production application's parallel performance for the intended research simulations. Ideally, the proposing team will have generated the data. If you cite work by others, explain why it is applicable here. You should use the application code you intend for the production work, not a related code. Data for sample systems not related to the intended research is undesirable. Performance benchmarking should reflect all I/O requirements. Parallel performance data in either strong or weak scaling mode must be provided. Explain how the strong or weak scaling applies to the proposed work. See the examples at the end of this document.

NOTE: You may apply for a startup account at one of the centers to conduct performance studies. Applications are available at

ALCF: {\href{http://www.alcf.anl.gov/getting-started/apply-for-dd}{http://www.alcf.anl.gov/getting-started/apply-for-dd}}

OLCF: {\href{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}}

\vspace{-.25in}
\subsection{Computational Approach}
\vspace{-.2in}

Provide a detailed description of your computational approach, including a discussion of the state of the art in the field. The description should also mention:

\vspace{-.25in}
\begin{enumerate}
\item Particular libraries required by the production and analysis software, algorithms and numerical techniques employed (e.g., finite element, iterative solver), programming languages, and other software used.
\item Parallel programming model(s) used (e.g., MPI, OpenMP/Pthreads and QPX intrinsics for BG/Q; MPI, OpenMP/Pthreads, CUDA, OpenACC or AVX intrinsics for XK7).
\item Project workflow including the role of analysis and visualization; identify where the analysis will be done and any potential bottlenecks in the analysis process.  Describe any analysis and/or data reduction tools used.
\item Software workflow solution (e.g., pre- and postprocessing scripts that automate run management and analysis) to facilitate this volume of work.
\item I/O requirements (e.g., amount, size, bandwidth, etc.) for restart, analysis, and workflow. Highlight any exceptional I/O needs.
\end{enumerate}

\vspace{-.25in}
\subsection{Developmental Work}
\vspace{-.2in}

For the computational approach above, describe what, if any, development work has been carried out to date, especially on the architecture of the requested resource. Describe what development work will be executed, and when,  during the proposed INCITE campaign.

\vspace{-.15in}
\section{REFERENCES}
\vspace{-.15in}

References are optional and may be structured in accordance with any style. They {\bf \em {do}} count toward the 15-page limit.
\vspace{-.15in}
\begin{enumerate}\itemsep0pt
\item First Author, Second Author, and Third Author, ``An article in a journal,'' {\em Journal Name} {\bf 32}(4): 46--52.\\
\item First Author and Second Author, {\em Report Title}, Report Number, Publishing Organization or Agency, City, State, 2013.
\item First Author et al., ``Chapter Title,'' {\em Book Title}, Publisher, City, State, 2011.
\item Corporate or Agency Author, {\em Book Title}, Publisher, City, State, 2012.
\end{enumerate}

\end{document}

### 2015-05-14 incite/Project_Summary_template.tex
adding incite bid templates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2645 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-05-14 incite/Milestone_Table_template.tex
@@
-%\documentclass[12pt,letterpaper,english]{article}
-\documentclass[11pt,landscape]{article}
+\documentclass[11pt,letterpaper,english]{article}
 \usepackage[T1]{fontenc} % Standard package for selecting font encodingsamely titles, headers and contents
@@
 \usepackage{sectsty}
-\sectionfont{\large}
+\sectionfont{\normalsize}
 \subsectionfont{\normalsize}
@@
 \captionsetup{labelsep=period}
-
-\renewcommand{\arraystretch}{1.25}
 
@@
 
-\colorbox{yellow} {\bf {\emph{[Refer to the guidelines for instructions in preparing the proposal. Table does not count toward}}}
-\colorbox{yellow} {\bf {\emph{[15-page project narrative count.]}}}
+\begin{center}
+\bf {PROJECT EXECUTIVE SUMMARY} \\
+{\bf \small {\em (Must not exceed 1 page but does} not {\em count toward the 15-page project narrative limit.)}}
+\end{center}
 
+%\begin{flushleft}
 \bigskip
 
-\textbf{Proposal Title (exactly as it appears on submission):} Text
+\textbf{Title (\emph{80 characters max; strictly enforced})}: Text \smallskip
 
-\begin{table}[h]
-%\centering
-%\caption{Table title}
-%\vspace{-.15in}
-\label{Tab1}
-\begin{tabular}{p{5cm} p{5cm} p{5cm}} \\ \hline 
-\noalign{\smallskip}
-\multicolumn{3}{l}{\textbf {Year 1}}\\ \hline 
-\noalign{\smallskip}
-\textbf{Milestone} & \textbf{Details} & \textbf{Date and Status} \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline 
-\noalign{\smallskip}
-\multicolumn{3}{l}{\textbf {Year 2 (if appropriate)}}\\ \hline
-\noalign{\smallskip}
-Insert Text & Insert Text & Insert Text \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline 
-\noalign{\smallskip}
-\multicolumn{3}{l}{\textbf {Year 3 (if appropriate)}}\\ \hline
-\noalign{\smallskip}
-Insert Text & Insert Text & Insert Text \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline 
-Insert Text & Insert Text & Insert Text \\ \hline
+\textbf{PI and Co-PI(s)}: Text \smallskip
 
-\end{tabular}
-\end{table}
+\textbf{Applying Institution/Organization}: Text \smallskip
+
+\textbf{Number of Processor Hours Requested}: Text \smallskip
+
+\textbf{Amount of Storage Requested}: Text \smallskip
+
+\textbf{Executive Summary ({\emph{May use the remainder of page}}):} \\
+
+The executive summary should accurately describe the proposed research and the high-impact scientific or technical advances you will realize with the proposed INCITE allocation.
+Industry organizations should also summarize the potential economic or strategic business impact of the proposed research.\\
+\vspace{.15in}
+Insert paragraph(s).
+
+%\end{flushleft}
 

### 2015-05-14 incite/Publications_template.tex
adding incite bid templates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2645 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-05-14 incite/Project_Narrative_template.tex
@@
 \usepackage{xcolor} % Driver-independent color extensions for LaTeX and pdfLaTeX.
-\usepackage{hyperref}  %The ability to create hyperlinks within the document
+%\usepackage[pagestyles]{titlesec} % related with sections—namely titles, headers and contents
 %\usepackage{blindtext} % To create text
 %\usepackage{mdwlist} % mdwlist for compact enumeration/list items 
-%\usepackage[pagestyles]{titlesec} % related with sections—namely titles, headers and contents
 \usepackage{fancyhdr} % header footer placement
@@
 
-\usepackage[font=bf]{caption}
+\usepackage{caption}
 \captionsetup{labelsep=period}
@@
 \begin{center}
-{\bf PROJECT NARRATIVE (The narrative should not exceed 15 pages)} 
+\bf {PUBLICATIONS RESULTING FROM INCITE AWARDS} \\
+{\bf {\em (Does} not {\em count toward the page limit.)}}
 \end{center}
+\vspace{-.25in}
+Provide a list of publications resulting from previous INCITE awards to this project team for work related to the proposal under consideration.
+Only those publications that include an acknowledgement to INCITE and/or the LCF may be included.
+This list may {\bf \underline {not}} be used in lieu of references in the project narrative.
+If applicable, list the citation both here and in the project narrative.
 
-\vspace{-.15in}
+References must be single-column format, 11 point, Arial or Times New Roman.
 
-%\colorbox{yellow} {\bf {\emph{[Refer to the guidelines for additional instructions for preparing the proposal.]}}}
+%{\noindent \bf {REFERENCE EXAMPLES}}
 
-%\vspace{.15in}
-
-{\bf References and visual materials, such as charts, graphs, pictures, etc., are included in the 15 page limit.} References should be included at the end of the Project Narrative.
-URLs that provide information related to the proposal should not be included  {\bf The 15-page limit will be strictly enforced.}  The Project Narrative should address the following points:
-
-\vspace{-.25in}
-\section{SIGNIFICANCE OF RESEARCH}
-\vspace{-.2in}
-List any previous INCITE award(s) received and discuss the relationship to the work proposed.
-Explain what advances you expect to be enabled by an INCITE award that justifies an allocation of petascale resources (e.g., anticipated impact on community paradigms, valuable insights into or solving a long-standing challenge, etc.).
-Place the proposed research in the context of competing work in your discipline or business.
-The information should be sufficient for peer review in your area of research and also appropriate for general scientific review comparing your proposal with proposals in other disciplines.
-Potential impact is the predominant determinant for awards.
-This factor will be assessed by a peer-review panel.
-Also list any previous INCITE award(s) received and discuss the relationship to the work proposed.
-{\bf This section is typically about 4 pages.}
-
-When you cite a reference, please insert a number in brackets, as shown here [1], to correspond to its number on the reference list at the end of this template.
-Please call out the references in numerical order and list them in the same way.
-The reference list will count toward the 15-page limit.
-
-Insert paragraph(s).
-\vspace{-.25in}
-\subsection{Heading 2 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsubsection{Heading 3 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\section{RESEARCH OBJECTIVES AND MILESTONES }  
-\vspace{-.2in}
-Describe the proposed research, including its goals and milestones and the theoretical and computational methods it employs.
-Goals and milestones should articulate simulation and developmental objectives and be sufficiently detailed to assess the progress of the project for each year of any allocation granted.
-Milestones should correlate with those in the milestone table.
-It is especially important that you provide clear connections between the project's overarching milestones, the planned production simulations, and the compute time expected to be required for these simulations (e.g., should correlate with those in the ``Use of Resources Requested'').
-{\bf This section is typically about 6 pages.}
-
-If included, call out equations, tables, figures, and references in numerical order in text, such as Eqs. (\ref{Eq. 1}) and (\ref{Eq. 2}), Table \ref{Tab1}, and Fig. \ref{Fig1} below.
-
-\vspace{-.15in}
-\begin{equation} \label{Eq. 1} 
-\partial ,\phi +u\cdot \nabla \phi =\nabla ^{2} \phi +\frac{1}{\tau } R\left(\phi \right)\, \, ,
-\end{equation}
-
-\vspace{-.15in}
-\begin{equation} \label{Eq. 2} 
-\frac{\partial \phi }{\nabla \phi } =\frac{1}{2} \nabla ^{2} \phi \frac{e^{\frac{-R-R^{2} }{2u} } }{\left(2\tau \right)^{3N/2} } \sqrt{xyz} \, \, \sum 1+23\, \, \, . 
-\end{equation}
-
-\begin{table}[h]
-\centering
-\caption{Table title}
-\vspace{-.15in}
-\label{Tab1}
-\begin{tabular}{llll} \\ \hline 
-\textbf{Column one} & \textbf{Column two} & \textbf{Column three} & \textbf{Column four} \\ \hline 
- xxx & xxx\textit{${}^{a}$} & xxx & xxx \\ \hline 
-xxx & xxx & xxx & xxx \\ \hline 
-{\textit{$^{a}$}Footnote here.} \\
-\end{tabular}
-\end{table}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=4.52in, height=3.08in, keepaspectratio=true]{image1.jpg}
-\caption{Figure caption.}
-\label{Fig1}
-\end{figure}
-
-\vspace{-.25in}
-\subsection{Heading 2 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsubsection{Heading 3 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\section{COMPUTATIONAL READINESS}
-\vspace{-.2in}
-
-Proposals will be assessed on the need for, readiness to use, and reasonableness of the request for resources. {\bf This section is typically about 5 pages.}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsection{Use of Resources Requested}
-\vspace{-.2in}
-
-Describe your proposed production simulations and state how the runs are tied to each of your project's goals and milestones (Section 4, "Milestone Table").
-For the simulations you plan to carry out during production runs, provide a
-\begin{enumerate}
-\item description of what jobs are going to be run and how they relate to the research/development objectives given above;
-\item description of processor/core use for large runs (e.g., 10,000-hour run with 100 cores, or ten 10-hour runs with 10,000 cores, for a 1,000,000-hour allocation); 
-\item clear, detailed explanation as to how you calculated the requested number of processor hours; and 
-\item summary of your anticipated annual burn rate (e.g., linear or with periods of peak usage).\\
-\vspace{.1in}
-In addition, describe the data storage requirements required by your production simulations.
-If at any point during your project the sum of your data storage needs in the scratch filesystems exceed 1 petabyte, specific justification is required.
-For your production simulations, provide a:
-
-\item Description of any input data set required, its size, and how you anticipate transferring it to the LCF.
-\item Estimate and breakdown of the anticipated cumulative size of stored data, in scratch and long-term archival storage, at the end of the requested award. 
-\item What is the effective lifetime of your useful data?  If the lifetime varies, show the breakdown by the total size used.  Explain the reason for the lifetime. 
-\item What types of tools for data storage, compression (reduction), movement, and analysis do you currently use?
-Are the tools and/or applications needed ready or are there new capabilities or features you are developing?
-\item If intending to make any fraction of the data generated public, please specify:
-\begin{enumerate}
-\item How much data and the scientific purpose
-\item What tool will be used to share the data
-\item From where will the data be shared\\
-\vspace{.1in}
-NOTE: The LCF data management policies can be found at
-
-OLCF:  {\href{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}}
-
-ALCF:  {\href{http://www.alcf.anl.gov/user-guides/data-policy}{http://www.alcf.anl.gov/user-guides/data-policy}}
-\end{enumerate}
-\end{enumerate}
-
-\vspace{-.25in}
-\subsection{Leadership Classification}
-\vspace{-.2in}
-
-Summarize the requirement that best exemplifies the proposed computational work. Leadership targets in the INCITE program typically include one or both of the following categories:
-
-\vspace{-.25in}
-\begin{enumerate}
-\item Use of 20 percent or more of the system for production simulations.
-Parameter sweeps, ensembles, design of experiments, and other statistical methods that require large numbers of discrete or loosely coupled simulations may be considered capability-class campaigns.
-See the FAQS for details and qualifiers.
-\item Specific architectural needs that can only be met by the LCF.
-\end{enumerate}
-\vspace{-.2in}
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsection{Parallel Performance}
-\vspace{-.2in}
-
-Provide direct evidence, {\bf including supporting quantitative data}, for your production application's parallel performance for the intended research simulations.
-Ideally, the proposing team will have generated the data.
-If you cite work by others, explain why it is applicable here.
-You should use the application code you intend for the production work, not a related code.
-Data for sample systems not related to the intended research is undesirable.
-Performance benchmarking should reflect all I/O requirements.
-Parallel performance data in either strong or weak scaling mode must be provided.
-Explain how the strong or weak scaling applies to the proposed work.
-See the examples at the end of this document.
-
-NOTE: You may apply for a startup account at one of the centers to conduct performance studies. Applications are available at
-
-ALCF: {\href{http://www.alcf.anl.gov/getting-started/apply-for-dd}{http://www.alcf.anl.gov/getting-started/apply-for-dd}}
-
-OLCF: {\href{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}}
-
-\vspace{-.25in}
-\subsection{Computational Approach}
-\vspace{-.2in}
-
-Provide a detailed description of your computational approach, including a discussion of the state of the art in the field. The description should also mention:
-
-\vspace{-.25in}
-\begin{enumerate}
-\item Particular libraries required by the production and analysis software, algorithms and numerical techniques employed (e.g., finite element, iterative solver), programming languages, and other software used.
-\item Parallel programming model(s) used (e.g., MPI, OpenMP/Pthreads and QPX intrinsics for BG/Q; MPI, OpenMP/Pthreads, CUDA, OpenACC or AVX intrinsics for XK7).
-\item Project workflow including the role of analysis and visualization; identify where the analysis will be done and any potential bottlenecks in the analysis process.
-Describe any analysis and/or data reduction tools used.
-\item Software workflow solution (e.g., pre- and postprocessing scripts that automate run management and analysis) to facilitate this volume of work.
-\item I/O requirements (e.g., amount, size, bandwidth, etc.) for restart, analysis, and workflow. Highlight any exceptional I/O needs.
-\end{enumerate}
-
-\vspace{-.25in}
-\subsection{Developmental Work}
-\vspace{-.2in}
-
-For the computational approach above, describe what, if any, development work has been carried out to date, especially on the architecture of the requested resource.
-Describe what development work will be executed, and when,  during the proposed INCITE campaign.
-
-\vspace{-.15in}
-\section{REFERENCES}
-\vspace{-.15in}
-
-References are optional and may be structured in accordance with any style. They {\bf \em {do}} count toward the 15-page limit.
-\vspace{-.15in}
-\begin{enumerate}\itemsep0pt
-\item First Author, Second Author, and Third Author, ``An article in a journal,'' {\em Journal Name} {\bf 32}(4): 46--52.\\
-\item First Author and Second Author, {\em Report Title}, Report Number, Publishing Organization or Agency, City, State, 2013.
-\item First Author et al., ``Chapter Title,'' {\em Book Title}, Publisher, City, State, 2011.
-\item Corporate or Agency Author, {\em Book Title}, Publisher, City, State, 2012.
-\end{enumerate}
+%\leftskip 0.25in
+% \parindent -0.25in
+%{Author, First, Second Author, and Third Author, ``An article in a journal,'' {\em{Journal Name}} {\bf{32}}(4): 46--52.}\\
+%{Author, First, and Second Author, {\em{Report Title}}, Report Number, Publishing Organization or Agency, City, State, 2013.}\\
+%{Author, First, et al., ``Chapter Title,'' {\em{Book Title}}, Publisher, City, State, 2011.}\\
+%{Corporate or Agency Author, {\em{Book Title}}, Publisher, City, State, 2012.}
 

### 2015-05-14 trunk/incite/CV_template.tex
adding incite bid templates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2645 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-05-14 incite/Project_Narrative_template.tex
@@
 \usepackage{xcolor} % Driver-independent color extensions for LaTeX and pdfLaTeX.
-\usepackage{hyperref}  %The ability to create hyperlinks within the document
 %\usepackage{blindtext} % To create text
@@
 \usepackage{sectsty}
-\sectionfont{\normalsize}
+\sectionfont{\large}
 \subsectionfont{\normalsize}
@@
 
-\usepackage[font=bf]{caption}
+\usepackage{caption}
 \captionsetup{labelsep=period}
-
-\setlength{\parskip}{\baselineskip}%
-\setlength{\parindent}{0pt}%
 
@@
 
-\begin{center}
-{\bf PROJECT NARRATIVE (The narrative should not exceed 15 pages)} 
-\end{center}
+\centering {\bf Curriculum Vitae ({\emph{2-page limit}})}\\
+{\bf PI NAME}\\
+{\bf Contact Information} \smallskip
 
-\vspace{-.15in}
+\begin{flushleft} {\bf Professional Preparation}
+{\parindent 16pt
 
-%\colorbox{yellow} {\bf {\emph{[Refer to the guidelines for additional instructions for preparing the proposal.]}}}
+PhD \\ 
+MS \\ 
+BS \\ 
+}
 
-%\vspace{.15in}
+\vspace{.04in}
+{\bf Appointments}
+{\parindent 16pt
 
-{\bf References and visual materials, such as charts, graphs, pictures, etc., are included in the 15 page limit.} References should be included at the end of the Project Narrative.
-URLs that provide information related to the proposal should not be included  {\bf The 15-page limit will be strictly enforced.}  The Project Narrative should address the following points:
+2010--present \\ 
+2001--2010 \\ 
+1990--2001 \\ 
+}
 
-\vspace{-.25in}
-\section{SIGNIFICANCE OF RESEARCH}
-\vspace{-.2in}
-List any previous INCITE award(s) received and discuss the relationship to the work proposed.
-Explain what advances you expect to be enabled by an INCITE award that justifies an allocation of petascale resources (e.g., anticipated impact on community paradigms, valuable insights into or solving a long-standing challenge, etc.).
-Place the proposed research in the context of competing work in your discipline or business.
-The information should be sufficient for peer review in your area of research and also appropriate for general scientific review comparing your proposal with proposals in other disciplines.
-Potential impact is the predominant determinant for awards.
-This factor will be assessed by a peer-review panel.
-Also list any previous INCITE award(s) received and discuss the relationship to the work proposed.
-{\bf This section is typically about 4 pages.}
+\vspace{.04in}
+{\bf Five Publications Most Relevant to This Proposal}
+\vspace{-6pt}
+\begin{enumerate} \itemsep1pt \parskip0pt \parsep0pt
+\item Publication most relevant to this proposal \\ 
+\item Publication most relevant to this proposal \\ 
+\item Publication most relevant to this proposal \\ 
+\item Publication most relevant to this proposal \\ 
+\item Publication most relevant to this proposal \\ 
+\end{enumerate}
 
-When you cite a reference, please insert a number in brackets, as shown here [1], to correspond to its number on the reference list at the end of this template.
-Please call out the references in numerical order and list them in the same way.
-The reference list will count toward the 15-page limit.
+\vspace{-6pt}
+{\bf Research Interests and Expertise}
+{\parindent 16pt
 
 Insert paragraph(s).
-\vspace{-.25in}
-\subsection{Heading 2 (optional)}
-\vspace{-.2in}
+}
 
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsubsection{Heading 3 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\section{RESEARCH OBJECTIVES AND MILESTONES }  
-\vspace{-.2in}
-Describe the proposed research, including its goals and milestones and the theoretical and computational methods it employs.
-Goals and milestones should articulate simulation and developmental objectives and be sufficiently detailed to assess the progress of the project for each year of any allocation granted.
-Milestones should correlate with those in the milestone table.
-It is especially important that you provide clear connections between the project's overarching milestones, the planned production simulations, and the compute time expected to be required for these simulations (e.g., should correlate with those in the ``Use of Resources Requested'').
-{\bf This section is typically about 6 pages.}
-
-If included, call out equations, tables, figures, and references in numerical order in text, such as Eqs. (\ref{Eq. 1}) and (\ref{Eq. 2}), Table \ref{Tab1}, and Fig. \ref{Fig1} below.
-
-\vspace{-.15in}
-\begin{equation} \label{Eq. 1} 
-\partial ,\phi +u\cdot \nabla \phi =\nabla ^{2} \phi +\frac{1}{\tau } R\left(\phi \right)\, \, ,
-\end{equation}
-
-\vspace{-.15in}
-\begin{equation} \label{Eq. 2} 
-\frac{\partial \phi }{\nabla \phi } =\frac{1}{2} \nabla ^{2} \phi \frac{e^{\frac{-R-R^{2} }{2u} } }{\left(2\tau \right)^{3N/2} } \sqrt{xyz} \, \, \sum 1+23\, \, \, . 
-\end{equation}
-
-\begin{table}[h]
-\centering
-\caption{Table title}
-\vspace{-.15in}
-\label{Tab1}
-\begin{tabular}{llll} \\ \hline 
-\textbf{Column one} & \textbf{Column two} & \textbf{Column three} & \textbf{Column four} \\ \hline 
- xxx & xxx\textit{${}^{a}$} & xxx & xxx \\ \hline 
-xxx & xxx & xxx & xxx \\ \hline 
-{\textit{$^{a}$}Footnote here.} \\
-\end{tabular}
-\end{table}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=4.52in, height=3.08in, keepaspectratio=true]{image1.jpg}
-\caption{Figure caption.}
-\label{Fig1}
-\end{figure}
-
-\vspace{-.25in}
-\subsection{Heading 2 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsubsection{Heading 3 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\section{COMPUTATIONAL READINESS}
-\vspace{-.2in}
-
-Proposals will be assessed on the need for, readiness to use, and reasonableness of the request for resources. {\bf This section is typically about 5 pages.}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsection{Use of Resources Requested}
-\vspace{-.2in}
-
-Describe your proposed production simulations and state how the runs are tied to each of your project's goals and milestones (Section 4, "Milestone Table").
-For the simulations you plan to carry out during production runs, provide a
-\begin{enumerate}
-\item description of what jobs are going to be run and how they relate to the research/development objectives given above;
-\item description of processor/core use for large runs (e.g., 10,000-hour run with 100 cores, or ten 10-hour runs with 10,000 cores, for a 1,000,000-hour allocation); 
-\item clear, detailed explanation as to how you calculated the requested number of processor hours; and 
-\item summary of your anticipated annual burn rate (e.g., linear or with periods of peak usage).\\
-\vspace{.1in}
-In addition, describe the data storage requirements required by your production simulations.
-If at any point during your project the sum of your data storage needs in the scratch filesystems exceed 1 petabyte, specific justification is required.
-For your production simulations, provide a:
-
-\item Description of any input data set required, its size, and how you anticipate transferring it to the LCF.
-\item Estimate and breakdown of the anticipated cumulative size of stored data, in scratch and long-term archival storage, at the end of the requested award. 
-\item What is the effective lifetime of your useful data?  If the lifetime varies, show the breakdown by the total size used.  Explain the reason for the lifetime. 
-\item What types of tools for data storage, compression (reduction), movement, and analysis do you currently use?
-Are the tools and/or applications needed ready or are there new capabilities or features you are developing?
-\item If intending to make any fraction of the data generated public, please specify:
-\begin{enumerate}
-\item How much data and the scientific purpose
-\item What tool will be used to share the data
-\item From where will the data be shared\\
-\vspace{.1in}
-NOTE: The LCF data management policies can be found at
-
-OLCF:  {\href{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}}
-
-ALCF:  {\href{http://www.alcf.anl.gov/user-guides/data-policy}{http://www.alcf.anl.gov/user-guides/data-policy}}
-\end{enumerate}
+\vspace{.04in}
+{\bf Synergistic Activities}
+\vspace{-6pt}
+\begin{enumerate} \itemsep1pt \parskip0pt \parsep0pt
+\item Text \\ 
+\item Text \\ 
+\item Text \\ 
+\item Text \\ 
+\item Text \\ 
 \end{enumerate}
 
-\vspace{-.25in}
-\subsection{Leadership Classification}
-\vspace{-.2in}
+\vspace{-6pt}
+{\bf Collaborators ({\emph{past 5 years including name and current institution}})} 
+{\parindent 16pt
 
-Summarize the requirement that best exemplifies the proposed computational work. Leadership targets in the INCITE program typically include one or both of the following categories:
+Collaborator name, current institution \\
+Collaborator name, current institution \\
+}
 
-\vspace{-.25in}
-\begin{enumerate}
-\item Use of 20 percent or more of the system for production simulations.
-Parameter sweeps, ensembles, design of experiments, and other statistical methods that require large numbers of discrete or loosely coupled simulations may be considered capability-class campaigns.
-See the FAQS for details and qualifiers.
-\item Specific architectural needs that can only be met by the LCF.
-\end{enumerate}
-\vspace{-.2in}
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsection{Parallel Performance}
-\vspace{-.2in}
-
-Provide direct evidence, {\bf including supporting quantitative data}, for your production application's parallel performance for the intended research simulations.
-Ideally, the proposing team will have generated the data.
-If you cite work by others, explain why it is applicable here.
-You should use the application code you intend for the production work, not a related code.
-Data for sample systems not related to the intended research is undesirable.
-Performance benchmarking should reflect all I/O requirements.
-Parallel performance data in either strong or weak scaling mode must be provided.
-Explain how the strong or weak scaling applies to the proposed work.
-See the examples at the end of this document.
-
-NOTE: You may apply for a startup account at one of the centers to conduct performance studies. Applications are available at
-
-ALCF: {\href{http://www.alcf.anl.gov/getting-started/apply-for-dd}{http://www.alcf.anl.gov/getting-started/apply-for-dd}}
-
-OLCF: {\href{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}}
-
-\vspace{-.25in}
-\subsection{Computational Approach}
-\vspace{-.2in}
-
-Provide a detailed description of your computational approach, including a discussion of the state of the art in the field. The description should also mention:
-
-\vspace{-.25in}
-\begin{enumerate}
-\item Particular libraries required by the production and analysis software, algorithms and numerical techniques employed (e.g., finite element, iterative solver), programming languages, and other software used.
-\item Parallel programming model(s) used (e.g., MPI, OpenMP/Pthreads and QPX intrinsics for BG/Q; MPI, OpenMP/Pthreads, CUDA, OpenACC or AVX intrinsics for XK7).
-\item Project workflow including the role of analysis and visualization; identify where the analysis will be done and any potential bottlenecks in the analysis process.
-Describe any analysis and/or data reduction tools used.
-\item Software workflow solution (e.g., pre- and postprocessing scripts that automate run management and analysis) to facilitate this volume of work.
-\item I/O requirements (e.g., amount, size, bandwidth, etc.) for restart, analysis, and workflow. Highlight any exceptional I/O needs.
-\end{enumerate}
-
-\vspace{-.25in}
-\subsection{Developmental Work}
-\vspace{-.2in}
-
-For the computational approach above, describe what, if any, development work has been carried out to date, especially on the architecture of the requested resource.
-Describe what development work will be executed, and when,  during the proposed INCITE campaign.
-
-\vspace{-.15in}
-\section{REFERENCES}
-\vspace{-.15in}
-
-References are optional and may be structured in accordance with any style. They {\bf \em {do}} count toward the 15-page limit.
-\vspace{-.15in}
-\begin{enumerate}\itemsep0pt
-\item First Author, Second Author, and Third Author, ``An article in a journal,'' {\em Journal Name} {\bf 32}(4): 46--52.\\
-\item First Author and Second Author, {\em Report Title}, Report Number, Publishing Organization or Agency, City, State, 2013.
-\item First Author et al., ``Chapter Title,'' {\em Book Title}, Publisher, City, State, 2011.
-\item Corporate or Agency Author, {\em Book Title}, Publisher, City, State, 2012.
-\end{enumerate}
+\end{flushleft}
 

### 2015-05-14 trunk/incite/Personnel_Justification_template.tex
adding incite bid templates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2645 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-05-14 incite/Project_Narrative_template.tex
@@
 \begin{center}
-{\bf PROJECT NARRATIVE (The narrative should not exceed 15 pages)} 
+\bf {PERSONNEL JUSTIFICATION AND MANAGEMENT PLAN} \\
+{\bf  {\em (Does} not {\em count toward the 15-page project narrative limit.)}}
 \end{center}
 
-\vspace{-.15in}
+\vspace{-.25in}
+\begin{flushleft}
+{\noindent \bf  {PERSONNEL JUSTIFICATION}}
 
-%\colorbox{yellow} {\bf {\emph{[Refer to the guidelines for additional instructions for preparing the proposal.]}}}
-
-%\vspace{.15in}
-
-{\bf References and visual materials, such as charts, graphs, pictures, etc., are included in the 15 page limit.} References should be included at the end of the Project Narrative.
-URLs that provide information related to the proposal should not be included  {\bf The 15-page limit will be strictly enforced.}  The Project Narrative should address the following points:
-
-\vspace{-.25in}
-\section{SIGNIFICANCE OF RESEARCH}
-\vspace{-.2in}
-List any previous INCITE award(s) received and discuss the relationship to the work proposed.
-Explain what advances you expect to be enabled by an INCITE award that justifies an allocation of petascale resources (e.g., anticipated impact on community paradigms, valuable insights into or solving a long-standing challenge, etc.).
-Place the proposed research in the context of competing work in your discipline or business.
-The information should be sufficient for peer review in your area of research and also appropriate for general scientific review comparing your proposal with proposals in other disciplines.
-Potential impact is the predominant determinant for awards.
-This factor will be assessed by a peer-review panel.
-Also list any previous INCITE award(s) received and discuss the relationship to the work proposed.
-{\bf This section is typically about 4 pages.}
-
-When you cite a reference, please insert a number in brackets, as shown here [1], to correspond to its number on the reference list at the end of this template.
-Please call out the references in numerical order and list them in the same way.
-The reference list will count toward the 15-page limit.
+What personnel are already in place and what are their roles on the project?
+If applicable, describe personnel that will be hired on the project in the future and their responsibilities.
+The INCITE program welcomes proposals from individual PIs or teams of collaborators.
 
 Insert paragraph(s). 
-\vspace{-.25in}
-\subsection{Heading 2 (optional)}
-\vspace{-.2in}
+\vspace{-.15in}
+\begin{itemize}
+\item Text\\ 
+\item Text\\
+\end{itemize}
+
+{\noindent \bf  {MANAGEMENT PLAN}}
+
+Describe the project's leadership team and how decisions are made to allocate time to individuals or, for proposals with multiple collaborating teams, subgroups within the project.
+Describe the focus of each individual or subgroup.
+Successful proposals will include a management plan that conveys to reviewers the interrelationship between subgroups and how the sum of the parts will lead to scientific discovery or engineering solutions that are the overarching goal of the project.
+Also identify points of contact who will provide updates on the status of the work including publications, awards, and highlights of accomplishments.
 
 Insert paragraph(s).
-
-\vspace{-.25in}
-\subsubsection{Heading 3 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\section{RESEARCH OBJECTIVES AND MILESTONES }  
-\vspace{-.2in}
-Describe the proposed research, including its goals and milestones and the theoretical and computational methods it employs.
-Goals and milestones should articulate simulation and developmental objectives and be sufficiently detailed to assess the progress of the project for each year of any allocation granted.
-Milestones should correlate with those in the milestone table.
-It is especially important that you provide clear connections between the project's overarching milestones, the planned production simulations, and the compute time expected to be required for these simulations (e.g., should correlate with those in the ``Use of Resources Requested'').
-{\bf This section is typically about 6 pages.}
-
-If included, call out equations, tables, figures, and references in numerical order in text, such as Eqs. (\ref{Eq. 1}) and (\ref{Eq. 2}), Table \ref{Tab1}, and Fig. \ref{Fig1} below.
-
-\vspace{-.15in}
-\begin{equation} \label{Eq. 1} 
-\partial ,\phi +u\cdot \nabla \phi =\nabla ^{2} \phi +\frac{1}{\tau } R\left(\phi \right)\, \, ,
-\end{equation}
-
-\vspace{-.15in}
-\begin{equation} \label{Eq. 2} 
-\frac{\partial \phi }{\nabla \phi } =\frac{1}{2} \nabla ^{2} \phi \frac{e^{\frac{-R-R^{2} }{2u} } }{\left(2\tau \right)^{3N/2} } \sqrt{xyz} \, \, \sum 1+23\, \, \, . 
-\end{equation}
-
-\begin{table}[h]
-\centering
-\caption{Table title}
-\vspace{-.15in}
-\label{Tab1}
-\begin{tabular}{llll} \\ \hline 
-\textbf{Column one} & \textbf{Column two} & \textbf{Column three} & \textbf{Column four} \\ \hline 
- xxx & xxx\textit{${}^{a}$} & xxx & xxx \\ \hline 
-xxx & xxx & xxx & xxx \\ \hline 
-{\textit{$^{a}$}Footnote here.} \\
-\end{tabular}
-\end{table}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=4.52in, height=3.08in, keepaspectratio=true]{image1.jpg}
-\caption{Figure caption.}
-\label{Fig1}
-\end{figure}
-
-\vspace{-.25in}
-\subsection{Heading 2 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsubsection{Heading 3 (optional)}
-\vspace{-.2in}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\section{COMPUTATIONAL READINESS}
-\vspace{-.2in}
-
-Proposals will be assessed on the need for, readiness to use, and reasonableness of the request for resources. {\bf This section is typically about 5 pages.}
-
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsection{Use of Resources Requested}
-\vspace{-.2in}
-
-Describe your proposed production simulations and state how the runs are tied to each of your project's goals and milestones (Section 4, "Milestone Table").
-For the simulations you plan to carry out during production runs, provide a
-\begin{enumerate}
-\item description of what jobs are going to be run and how they relate to the research/development objectives given above;
-\item description of processor/core use for large runs (e.g., 10,000-hour run with 100 cores, or ten 10-hour runs with 10,000 cores, for a 1,000,000-hour allocation); 
-\item clear, detailed explanation as to how you calculated the requested number of processor hours; and 
-\item summary of your anticipated annual burn rate (e.g., linear or with periods of peak usage).\\
-\vspace{.1in}
-In addition, describe the data storage requirements required by your production simulations.
-If at any point during your project the sum of your data storage needs in the scratch filesystems exceed 1 petabyte, specific justification is required.
-For your production simulations, provide a:
-
-\item Description of any input data set required, its size, and how you anticipate transferring it to the LCF.
-\item Estimate and breakdown of the anticipated cumulative size of stored data, in scratch and long-term archival storage, at the end of the requested award. 
-\item What is the effective lifetime of your useful data?  If the lifetime varies, show the breakdown by the total size used.  Explain the reason for the lifetime. 
-\item What types of tools for data storage, compression (reduction), movement, and analysis do you currently use?
-Are the tools and/or applications needed ready or are there new capabilities or features you are developing?
-\item If intending to make any fraction of the data generated public, please specify:
-\begin{enumerate}
-\item How much data and the scientific purpose
-\item What tool will be used to share the data
-\item From where will the data be shared\\
-\vspace{.1in}
-NOTE: The LCF data management policies can be found at
-
-OLCF:  {\href{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}{https://www.olcf.ornl.gov/computing-resources/data-management/data-management-user-guide/}}
-
-ALCF:  {\href{http://www.alcf.anl.gov/user-guides/data-policy}{http://www.alcf.anl.gov/user-guides/data-policy}}
-\end{enumerate}
-\end{enumerate}
-
-\vspace{-.25in}
-\subsection{Leadership Classification}
-\vspace{-.2in}
-
-Summarize the requirement that best exemplifies the proposed computational work. Leadership targets in the INCITE program typically include one or both of the following categories:
-
-\vspace{-.25in}
-\begin{enumerate}
-\item Use of 20 percent or more of the system for production simulations.
-Parameter sweeps, ensembles, design of experiments, and other statistical methods that require large numbers of discrete or loosely coupled simulations may be considered capability-class campaigns.
-See the FAQS for details and qualifiers.
-\item Specific architectural needs that can only be met by the LCF.
-\end{enumerate}
-\vspace{-.2in}
-Insert paragraph(s).
-
-\vspace{-.25in}
-\subsection{Parallel Performance}
-\vspace{-.2in}
-
-Provide direct evidence, {\bf including supporting quantitative data}, for your production application's parallel performance for the intended research simulations.
-Ideally, the proposing team will have generated the data.
-If you cite work by others, explain why it is applicable here.
-You should use the application code you intend for the production work, not a related code.
-Data for sample systems not related to the intended research is undesirable.
-Performance benchmarking should reflect all I/O requirements.
-Parallel performance data in either strong or weak scaling mode must be provided.
-Explain how the strong or weak scaling applies to the proposed work.
-See the examples at the end of this document.
-
-NOTE: You may apply for a startup account at one of the centers to conduct performance studies. Applications are available at
-
-ALCF: {\href{http://www.alcf.anl.gov/getting-started/apply-for-dd}{http://www.alcf.anl.gov/getting-started/apply-for-dd}}
-
-OLCF: {\href{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}{http://www.olcf.ornl.gov/support/getting-started/olcf-director-discretion-project-application}}
-
-\vspace{-.25in}
-\subsection{Computational Approach}
-\vspace{-.2in}
-
-Provide a detailed description of your computational approach, including a discussion of the state of the art in the field. The description should also mention:
-
-\vspace{-.25in}
-\begin{enumerate}
-\item Particular libraries required by the production and analysis software, algorithms and numerical techniques employed (e.g., finite element, iterative solver), programming languages, and other software used.
-\item Parallel programming model(s) used (e.g., MPI, OpenMP/Pthreads and QPX intrinsics for BG/Q; MPI, OpenMP/Pthreads, CUDA, OpenACC or AVX intrinsics for XK7).
-\item Project workflow including the role of analysis and visualization; identify where the analysis will be done and any potential bottlenecks in the analysis process.
-Describe any analysis and/or data reduction tools used.
-\item Software workflow solution (e.g., pre- and postprocessing scripts that automate run management and analysis) to facilitate this volume of work.
-\item I/O requirements (e.g., amount, size, bandwidth, etc.) for restart, analysis, and workflow. Highlight any exceptional I/O needs.
-\end{enumerate}
-
-\vspace{-.25in}
-\subsection{Developmental Work}
-\vspace{-.2in}
-
-For the computational approach above, describe what, if any, development work has been carried out to date, especially on the architecture of the requested resource.
-Describe what development work will be executed, and when,  during the proposed INCITE campaign.
-
-\vspace{-.15in}
-\section{REFERENCES}
-\vspace{-.15in}
-
-References are optional and may be structured in accordance with any style. They {\bf \em {do}} count toward the 15-page limit.
-\vspace{-.15in}
-\begin{enumerate}\itemsep0pt
-\item First Author, Second Author, and Third Author, ``An article in a journal,'' {\em Journal Name} {\bf 32}(4): 46--52.\\
-\item First Author and Second Author, {\em Report Title}, Report Number, Publishing Organization or Agency, City, State, 2013.
-\item First Author et al., ``Chapter Title,'' {\em Book Title}, Publisher, City, State, 2011.
-\item Corporate or Agency Author, {\em Book Title}, Publisher, City, State, 2012.
-\end{enumerate}
+\end{flushleft}
 

### 2015-05-14 trunk/src/gradient_rt.h

@@ -18,6 +18,6 @@
 #include "field_grad.h"
 #include "map.h"
 
-int gradient_rt_init(field_grad_t * grad, map_t * map);
+__targetHost__ int gradient_rt_init(field_grad_t * grad, map_t * map);
 
 #endif

### 2015-07-02 doc/plan.tex
Add doc directory
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2701 40876441-3496-0410-a613-a2fca35b2ae2

--- 2011-06-21 htdocs/plan.tex
@@
 %
-%  $Id$
-%
 %  Edinburgh Soft Matter and Statistical Physics Group and
@@
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2011 The University of Edinburgh
+%  (c) 2011-2015 The University of Edinburgh
 %
@@
 
-% This is a section to be included in ludwig.tex
+\documentclass[11pt,twoside]{article}
+
+\setlength{\hoffset}{-1in}
+\setlength{\voffset}{-1in}
+
+\setlength{\topmargin}{2cm}
+\setlength{\evensidemargin}{3.1cm}
+\setlength{\oddsidemargin}{3.1cm}
+
+\setlength{\textwidth}{14.8cm}
+\setlength{\textheight}{23cm}
+
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{\smallskipamount}
+
+\newcommand{\inputkey}[1]{\framebox{\textbf{\texttt{#1}}}}
+\newcommand{\e}[1]{\cdot10^{#1}}
+\newcommand{\beq}{\begin{equation}}
+\newcommand{\eeq}{\end{equation}}
+\newcommand{\beqa}{\begin{eqnarray}}
+\newcommand{\eeqa}{\end{eqnarray}}
+\newcommand{\com}[1]{\textcolor{red}{#1}}
+\newcommand{\cur}[1]{{\textit{#1}}}
+
+\begin{document}
+
+\setcounter{page}{1}
+
+\tableofcontents
+
+\newpage
+
+\setcounter{page}{1}
 
@@
 This section is based on the IEEE standard for Software Configuration
-Management Plans IEEE 828-2005 \cite{ieee-208}, and follows the format
+Management Plans IEEE 828-2012 \cite{ieee-828-2012}, and follows the format
 set out therein.
@@
 scientific direction and development of the code and concomitant
-priorities, all
-the Software Configuration Management activities will be the
-responsibility of EPCC.
+priorities, all the Software Configuration Management activities will
+be the responsibility of EPCC.
 
@@
 
-\texttt{/trunk/current} a precursor code not under control
+\texttt{/trunk/doc} source files for this documentation;
 
-\texttt{/trunk/htdocs} source files for this documentation
-
-\texttt{/trunk/mpi\_s} MPI stub library (scheduled to move)
+\texttt{/trunk/mpi\_s} an MPI stub library;
 
 \texttt{/trunk/src} source code
+
+\texttt{/trunk/targetDP} thread parallelism abstraction layer, currently
+under development;
 
@@
 
-At present access is restricted. However, we have committed to making
-the code available to the public by October 2014. We expect this to
-be via the same repository, referred to as CCPForge.
-All members of the \textit{Ludwig} project are currently registered with
-a CCPForge user account, and fall into one of three categories:
-Administrator, Developer, and User. Only the first two categories have
+All access is via the CCPForge repository. Users can obtain the code
+via anonymous svn access at CPPForge.
+All developers of the \textit{Ludwig} project are currently registered with
+a CCPForge user account, and fall into one of two categories.
+Administrator or Developer. Only the account-holding developers have
 permission to change files in the repository. Users have read-only
-access. The exact mechanism for public access is yet to be decided,
-but may still require users to register with CCPForge.
+access.
 
@@
 complete dump of the repository itself to a local resource on a
-weekly basis for security.
+regular basis for security.
 
@@
 
-Can we account for the status of the code and tests?
+Can we account for the status of the code and tests.
 
@@
 will be in place to coincide with the public release.
+
+\newpage
+\section{Coordination Meetings}
+
+\begin{thebibliography}{99}
+\bibitem{ieee-828-2012} IEEE standard for Configuration Management
+\bibitem{mpi-standard} Message passing ineterface.
+\bibitem{paraview} Paraview is a standard visualisation package. 
+\end{thebibliography}
+
+\end{document}

### 2015-08-30 trunk/doc/plan.tex
Update to SCM plan
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2724 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-07-02 doc/plan.tex
@@
 the code is developed, how bugs are dealt with, the way in which
-new features may be added, and so on. It is therefore a part
+new features may be added, how correctness is ensured and tested, and so on.
+It is therefore a part
 of the process by which the code is maintained. The Software Configuration
@@
 based finite difference approach, which couples explicitly to the
-hydrodynamics. Examples of complex fluids include binary mixtures,
-surfactants, liquid crystals, and active gels. Solid objects such as
-porous media may be included, as well as moving particles (colloids).
+hydrodynamics.
 Components of \textit{Ludwig} include the main program, unit and
-regression tests, and a small number of utility programs used to prepare
+regression tests, and a small number support libraries, and a small
+number of utility programs used to prepare
 input and post-process output. The Plan is limited to these
@@
 relies on the Message Passing Interface \cite{mpi-standard}. The
-code will also run in serial, and is designed to be independent of
-the platform it is run on. The Plan is therefore not concerned with
+code will also run in serial, and is independent of
+the platform it is run on. The Plan is not concerned with
 hardware or system management activities.
 
-This document is meant to be fairly informal, but will be updated as
-the need arises. 
 \textit{Ludwig} is an on-going research project, and relies mainly
@@
 Council to provide support for staff time to work on maintenance and
-development. As such, we cannot commit to any and all user requests!
+development.
 
-\subsubsection{Definition of key terms}
+% IEEE section to be completed
+%\subsubsection{Definition of key terms}
 
@@
 The University of Edinburgh. This involves two groups: the Soft Matter
-Physics Group  under the leadership of Prof. Mike Cates FRS, and EPCC in
-the person of Kevin Stratford. We collaborate with a number of workers at
-different Universities around the world on different aspects of the code.
+Physics Group, and Edinburgh Parallel Computing Centre (EPCC). These
+groups collaborate with a number of workers at different Universities
+around the world on different aspects of the code.
 
@@
 
-\subsubsection{Procedures}
-
-There are currently no external constraints on The Plan.
+% IEEE section recheck what this actually means
+% \subsubsection{Procedures}
+% There are currently no external constraints on The Plan.
 
@@
 EPCC is responsible for the Software Configuration Management process.
-It is anticipated that the cost of this process will be negligible, and
-will not require monitoring. There is currently no independent
+It is anticipated that the cost of this process will be small as the
+project team can communicate informally on a regular basis. There is
+currently no independent
 surveillance of activities to ensure compliance with The Plan.
@@
 identified by its path in the repository as it exists at a given
-revision number. For example, SVN revision 1416 contains the
-following directories:
+revision number.
 
-\texttt{/trunk/doc} source files for this documentation;
+% Add pointer to standard manifest (README)
 
-\texttt{/trunk/mpi\_s} an MPI stub library;
+All access to the code is via the CCPForge repository. CCPForge identifies
+three different levels of access:
 
-\texttt{/trunk/src} source code
+\textit{Administrators:} allows administration of the site itself and
+registered members;
 
-\texttt{/trunk/targetDP} thread parallelism abstraction layer, currently
-under development;
+\textit{Developers:} members who have write access to the source code repository;
 
-\texttt{/trunk/tests} unit and regression test suite
+\textit{Users:} have read only access to the source code via SVN (either
+as a member or via anonymous download). Member users also have access to
+the tracker.
 
-\texttt{/trunk/util} utility programs
-
-All access is via the CCPForge repository. Users can obtain the code
-via anonymous svn access at CPPForge.
-All developers of the \textit{Ludwig} project are currently registered with
-a CCPForge user account, and fall into one of two categories.
-Administrator or Developer. Only the account-holding developers have
-permission to change files in the repository. Users have read-only
-access.
-
-As CCPForge is not physically under our control, we undertake a
-complete dump of the repository itself to a local resource on a
+As CCPForge is not physically under the control of EPCC, a
+complete dump of the repository to a local resource is undertaken on a
 regular basis for security.
@@
 the issue tracker at CCPForge. The project team will then decide
-whether the change is possible. If so, the implementation and
-testing of the change will take place, and the new code committed
-back to the repository.
+whether the change is possible and/or desirable. If so, the
+implementation and testing of the change will take place, and the
+new code committed back to the repository.
+
+\textit{Submitting a change request:} A description of the proposed change
+will be submitted to the CCPForge issue tracker as a change request to
+provide a record of the change process.
+
+\textit{Evaluating a change request:} The request, together with any
+additional information, will be evaluated by the SCM team. The request
+will be accepted, refined, or rejected as appropriate. The change owner
+will make necessary updates to the change record.
+
+\textit{Implementation of change:} Important changes to behaviour of code
+related to a change will be
+accompanied by a relevant descriptive record in the change request.
+
+\textit{Version control:} A \texttt{MAJOR.MINOR.PATCH} numerical version
+name scheme is used \cite{apacheAPR}.
+It is expected that bug fixes and relatively
+small changes will be accompanied by a unit increment in the patch level;
+more significant changes will be reflected at the minor version number;
+major reconstructions will occur rarely and incur a change of major version
+number.
 
@@
 
-Can we account for the status of the code and tests.
+% Privides tracability of changes, pending completed, etc
+% Configuration reports, release notes
 
-\subsubsection{Evaluation and review}
+The status of changes will be tracked at the CCPForge issue tracker
+with a named SCM team member owning the change. The relevant tracker
+item will be closed when successfully implemented and tested.
 
-How do we evaluate and audit the code?
+The tracker is available to CPPForge member users to provide a record
+of changes to the code. Anonymous users will be alerted to changes via
+release notes.
 
-\subsubsection{Interface control}
+%\subsubsection{Evaluation and review}
+%There are currently no formal mechna audit the code?
 
-At the moment, we do not explicitly interface to any code outside The
-Plan (except MPI). We could extend coverage to include data export
-to e.g., Paraview \cite{paraview}.
+% \subsubsection{Interface control}
+% This section discusses other organisations the developers must
+% ``interface'' with.
 
-\subsubsection{Subcontractor / vendor control}
-
-None required.
+% IEEE section not relevant
+%\subsubsection{Subcontractor / vendor control}
 
@@
 
-There is currently no formal release mechanism. If such a mechanism were
-introduced, management and  delivery would be via the SVN repository.
+There are currently no formal releases of the \textit{Ludwig} code.
+Any release management and  delivery will be via the SVN repository.
 
-\subsection{Schedules}
+% IEEE section. We don't really have any schedules.
+%\subsection{Schedules}
 
-Sequence of events for SCM process.
+%\subsection{Resources}
 
-\subsection{Resources}
-
-For each active, state what resources (infrastructure, software tools,
-personnel etc) are required.
+% Software resources (dependencies, etc)
+% Hardware resources
+% Human resources
 
@@
 
-The Plan is currently in a draft form. A complete version of The Plan
-will be in place to coincide with the public release.
+The Plan is currently under review to consider organisational changes
+affecting the responsibilities different stakeholders.
 
-\newpage
-\section{Coordination Meetings}
+%\newpage
+%\section{Coordination Meetings}
 
 \begin{thebibliography}{99}
-\bibitem{ieee-828-2012} IEEE standard for Configuration Management
-\bibitem{mpi-standard} Message passing ineterface.
-\bibitem{paraview} Paraview is a standard visualisation package. 
+\bibitem{ieee-828-2012} IEEE Standard 828-2011 ``IEEE Standard for
+Software Configuration Management Plans''
+\bibitem{mpi-standard} Message Passing Interface Forum. MPI: A Message
+Passing Interface Standard Version 1.3 (1998).
+%\bibitem{paraview} Paraview is a standard visualisation package. 
+\bibitem{apacheAPR} See, e.g., the Apache Portable Runtime Project ``APR's Version Numbering''  https://apr.apache.org/versioning.html (Accessed 2015).
 \end{thebibliography}
 
+\vfill\eject
+
 \end{document}

### 2015-10-16 branches/kevin-1407/targetDP/doc/execution_model.tex
1407 feature branch
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2755 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-03-12 targetDP/doc/execution_model.tex
@@
 device targeted for execution of data parallel operations. The target
-can be the same CPU as the host, or it can be a separate device such
-as an accelerator.
+could be the same CPU as the host, or it could be a separate device such
+as an accelerator (depending on the hardware available).
 
@@
 (which do not depend on the results of the target function), after the
-function launch but before the syncronisation call. This may result in
+function launch but before the synchronisation call. This may result in
 overlapping of host and target execution, and hence optimisation, in
@@
 \end{comment}
+
+\begin{comment}
+Some text on reductions needed.
+\end{comment}

### 2015-10-16 branches/kevin-1407/targetDP/doc/memory_model.tex
1407 feature branch
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2755 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Memory model}

\section{Model overview}
The targetDP model draws a distinction between the memory spaces
accessed by the host and target, such that code executed on the host
always accesses the host memory space, and code executed on the target
(i.e. within target functions) always accesses the target memory space. The
host memory space can be initialised using regular \verb+C/C+++
functionality, and the targetDP API provides the functionality
necessary to manage the target memory space and transfer data between
the host and target. For each data-parallel data structure, the
programmer should create both host and target copies, and should
update these from each other as and when required.

\subsection{Host memory model}
The sequential thread executing host code should always access host data
structures. The memory model is the same as one would expect from a
regular sequential application.

\subsection{Target memory model}
The team of threads performing the execution of target functions
should always act on target data structures. These can take one of 3
forms:
\begin{enumerate}
\item Those created using the targetDP memory allocation API
  functions. These are shared between all threads in the team, where
  each thread should use its unique index to access the portion of data
  for which it is responsible.
\item Those created using the targetDP constant data management
  functionality. These are read-only and normally used for relatively
  small amounts of constant data.
\item Those declared within the body of a target function. These are private
  to each thread in the team and should be used for temporary scratch structures.
\end{enumerate}

\begin{comment}
For 3, at the moment any data declared in the function but above the targetTLP keyword will be private for GPU threads but shared for CPU (since outside parallel region). So either need to make this clearer above, or somehow move OpenMP parallel region to target launch. 
\end{comment}

\section{Implementation}
\subsection{C}

The target memory structures will exist on the same physical memory as
the host structures. The implementation may either use separate target
copies (managed using regular C/C++ memory management functionality),
or use pointer aliasing for the target versions such that a reference
to any part of a target structure will correspond to exactly the same
physical address as that of the corresponding host structure.

\subsection{CUDA}

The target memory space will exist on the distinct GPU memory, i.e. in
a separate memory space from the host structures.

### 2015-10-16 branches/kevin-1407/src/bbl.h

@@ -1,0 +1,33 @@
+/*****************************************************************************
+ *
+ *  bbl.h
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010-2014 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef BBL_H
+#define BBL_H
+
+#include "colloids.h"
+#include "model.h"
+
+typedef struct bbl_s bbl_t;
+
+int bbl_create(lb_t * lb, bbl_t ** pobj);
+int bbl_free(bbl_t * obj);
+
+int bounce_back_on_links(bbl_t * bbl, lb_t * lb, colloids_info_t * cinfo);
+int bbl_pass0(bbl_t * bbl, lb_t * lb, colloids_info_t * cinfo);
+
+int bbl_active_set(bbl_t * bbl, colloids_info_t * cinfo);
+int bbl_update_colloids(bbl_t * bbl, colloids_info_t * cinfo);
+
+int bbl_surface_stress(bbl_t * bbl, double slocal[3][3]);
+int bbl_order_parameter_deficit(bbl_t * bbl, double * delta);
+
+#endif

### 2015-10-16 branches/kevin-1407/src/coords.h

@@ -1,0 +1,88 @@
+/*****************************************************************************
+ *
+ *  coords.h
+ *
+ *  $Id: coords.h,v 1.4 2010-10-15 12:40:02 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef COORDS_H
+#define COORDS_H
+
+#include <mpi.h>
+
+#include "targetDP.h"
+
+#define NSYMM 6      /* Elements for general symmetric tensor */
+
+enum cartesian_directions {X, Y, Z};
+enum cartesian_neighbours {FORWARD, BACKWARD};
+enum upper_triangle {XX, XY, XZ, YY, YZ, ZZ};
+
+__targetHost__ void   coords_init(void);
+__targetHost__ void   coords_finish(void);
+__targetHost__ void   coords_info(void);
+__targetHost__ int    N_total(const int);
+__targetHost__ int    is_periodic(const int);
+__targetHost__ double L(const int);
+__targetHost__ double Lmin(const int);
+__targetHost__ int    cart_rank(void);
+__targetHost__ int    cart_size(const int);
+__targetHost__ int    cart_coords(const int);
+__targetHost__ int    cart_neighb(const int direction, const int dimension);
+
+__targetHost__ MPI_Comm cart_comm(void);
+
+__targetHost__ void   coords_nlocal(int n[3]);
+__targetHost__ void   coords_nlocal_offset(int n[3]);
+__targetHost__ void   coords_nhalo_set(const int nhalo);
+__targetHost__ int    coords_nhalo(void);
+__targetHost__ int    coords_ntotal(int ntotal[3]);
+__targetHost__ void   coords_ntotal_set(const int n[3]);
+__targetHost__ void   coords_decomposition_set(const int p[3]);
+__targetHost__ void   coords_reorder_set(const int);
+__targetHost__ void   coords_periodicity_set(const int p[3]);
+__targetHost__ int    coords_nsites(void);
+__targetHost__ int    coords_index(const int ic, const int jc, const int kc);
[…]

### 2015-10-16 branches/kevin-1407/src/gradient_3d_7pt_solid.h

@@ -1,0 +1,23 @@
+/*****************************************************************************
+ *
+ *  gradient_3d_7pt_solid.h
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2011 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef GRADIENT_3D_7PT_SOLID_H
+#define GRADIENT_3D_7PT_SOLID_H
+
+#include "map.h"
+
+__targetHost__ int gradient_3d_7pt_solid_map_set(map_t * map);
+__targetHost__ int gradient_3d_7pt_solid_d2(const int nop, const double * field,double * t_field,
+			     double * grad,double * t_grad, double * delsq, double * t_delsq);
+#endif

### 2015-10-16 branches/kevin-1407/src/psi_force.c

@@ -1,0 +1,583 @@
+/*****************************************************************************
+ *
+ *  psi_force.c
+ *
+ *  Compute the force on the fluid originating with charge.
+ *
+ *  Edinburgh Soft Matter and Statisitical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  (c) 2013 The University of Edinburgh
+ *
+ *  Contributing authors:
+ *    Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *    Ignacio Pagonabarraga
+ *    Oliver Henrich
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <math.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "control.h"
+#include "physics.h"
+#include "psi_s.h"
+#include "fe_electro.h"
+#include "fe_electro_symmetric.h"
+#include "psi_force.h"
+#include "psi_gradients.h"
+
+static int psi_force_divergence_ = 1;
+
+
+/*****************************************************************************
+ *
+ *  psi_force_gradmu
+ *
+ *  This routine computes the force on the fluid via the gradient
+ *  of the chemical potential.
+ *  Note: The ionic solvation free energy difference is in units 
+ *        of kt and must be dressed for the force calculation.
+ *
+ *****************************************************************************/
+
+int psi_force_gradmu(psi_t * psi, field_t * phi, hydro_t * hydro,
+		map_t * map, colloids_info_t * cinfo) {
+
+  int ic, jc, kc;
+  int in, nk;
+  int ia;
+  int nlocal[3];
+  int index;
+  int xs, ys, zs;       /* Coordinate strides */
+  double rho, rho_elec; /* Species and electric charge density */
+  double e[3];          /* Total electric field */
+  double muphim1, muphip1, musm1, musp1;
+  double phi0;          /* Compositional order parameter */
+  double kt, eunit, reunit;
+  double force[3];
+  /* Cummulative forces for momentum correction */
+  double flocal[4] = {0.0, 0.0, 0.0, 0.0};
+  double fsum[4];
+
[…]

### 2015-10-16 branches/kevin-1407/src/symmetric.c

@@ -1,0 +1,492 @@
+/****************************************************************************
+ *
+ *  symmetric.c
+ *
+ *  Implementation of the symmetric \phi^4 free energy functional:
+ *
+ *  F[\phi] = (1/2) A \phi^2 + (1/4) B \phi^4 + (1/2) \kappa (\nabla\phi)^2
+ *
+ *  The first two terms represent the bulk free energy, while the
+ *  final term penalises curvature in the interface. For a complete
+ *  description see Kendon et al., J. Fluid Mech., 440, 147 (2001).
+ *
+ *  The usual mode of operation is to take a = -b < 0 and k > 0.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group
+ *  and Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2011 The University of Edinburgh
+ *
+ ****************************************************************************/
+
+#include <assert.h>
+#include <math.h>
+
+#include "util.h"
+#include "symmetric.h"
+#include "field_s.h"
+#include "field_grad_s.h"
+#include "coords.h"
+
+static double a_     = -0.003125;
+static double b_     = +0.003125;
+static double kappa_ = +0.002;
+
+static __targetConst__ double t_a_     = -0.003125;
+static __targetConst__ double t_b_     = +0.003125;
+static __targetConst__ double t_kappa_ = +0.002;
+
+
+static field_t * phi_ = NULL;
+static field_grad_t * grad_phi_ = NULL;
+
+
+// flag to track whHether this module has been initiated
+static char symmetric_flag=0;
+__targetHost__ char symmetric_in_use(){ return symmetric_flag; }
+
+
+
+/****************************************************************************
+ *
+ *  symmetric_phi_set
+ *
+ *  Attach a reference to the order parameter field object, and the
+ *  associated gradient object.
+ *
+ ****************************************************************************/
+
+__targetHost__ int symmetric_phi_set(field_t * phi, field_grad_t * dphi) {
+
+  assert(phi);
+  assert(dphi);
+
+  phi_ = phi;
+  grad_phi_ = dphi;
+
+  return 0;
+}
+
+
+
+
+/****************************************************************************
+ *
+ *  symmetric_free_energy_parameters_set
+ *
[…]

### 2015-10-16 branches/kevin-1407/src/wall_ss_cut.h

@@ -1,0 +1,31 @@
+/*****************************************************************************
+ *
+ *  wall_ss_cut.h
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  (c) The University of Edinburgh (2014)
+ *  Juho Lintuvuori (juho.lintuvuori@u-psud.fr)
+ *
+ *****************************************************************************/
+
+#ifndef WALL_SS_CUT_H
+#define WALL_SS_CUT_H
+
+#include "colloids.h"
+#include "interaction.h"
+
+typedef struct wall_ss_cut_s wall_ss_cut_t;
+
+int wall_ss_cut_create(wall_ss_cut_t ** pobj);
+void wall_ss_cut_free(wall_ss_cut_t * obj);
+int wall_ss_cut_info(wall_ss_cut_t * obj);
+int wall_ss_cut_register(wall_ss_cut_t * obj, interact_t * parent);
+int wall_ss_cut_compute(colloids_info_t * cinfo, void * self);
+int wall_ss_cut_stats(void * self, double * stats);
+int wall_ss_cut_single(wall_ss_cut_t * obj, double h, double * f, double * v);
+int wall_ss_cut_param_set(wall_ss_cut_t * obj, double epsilon, double sigma,
+			  int nu, double hc);
+
+#endif

### 2015-10-16 branches/kevin-1407/tests/test-all.sh

@@ -1,0 +1,28 @@
+##############################################################################
+#
+#  test-all.sh
+#
+#  Just a convenience to run all tests...
+#
+#  Edinburgh Soft Matter and Statistical Physics Group and
+#  Edinburgh Parallel Computing Centre
+#
+#  Kevin Stratford (kevin@epcc.ed.ac.uk)
+#  (c) 2014 The University of Edinburgh
+#
+##############################################################################
+#!/bin/bash
+
+. ./test-serial.sh -u d2q9
+. ./test-serial.sh -u d2q9r
+. ./test-serial.sh -u -r d3q15
+. ./test-serial.sh -u -r d3q15r
+. ./test-serial.sh -u -r d3q19
+. ./test-serial.sh -u -r d3q19r
+
+. ./test-mpix08.sh -u d3q15 pmpi08
+. ./test-mpix08.sh -u -r d3q19 pmpi08
+. ./test-mpix08.sh -u -r d3q19r pmpi08
+. ./test-mpix08.sh -u d3q19 long08
+
+. ./test-long64.sh -r d3q19 long64

### 2015-10-16 branches/kevin-1407/version.h

@@ -1,0 +1,19 @@
+/*****************************************************************************
+ *
+ *  version.h
+ *
+ *  The version is MAJOR.MINOR.PATCH
+ *  See, e.g., https://apr.apache.org/versioning.html
+ *
+ *  (c) 2014-2015 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef VERSION_H
+#define VERSION_H
+
+#define LUDWIG_MAJOR_VERSION 0
+#define LUDWIG_MINOR_VERSION 2
+#define LUDWIG_PATCH_VERSION 19
+
+#endif

### 2015-10-23 trunk/docs/devel.tex
Issue 1405: added new docs directory with content
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2764 40876441-3496-0410-a613-a2fca35b2ae2

--- 2014-11-14 branches/kevin/docs/devel.tex
@@
 are of two types. Unit tests are generally written when units of code
-are first introduced to ensure basic operation is error free. Regression
-tests are introduced and updated when physical results are validated.
+are first introduced to ensure basic operation is error free. The unit
+tests are encoded in an executable linked against the appropriate
+\textit{Ludwig} library.
+Regression
+tests are introduced and updated when physical results are (broadly)
+validated. The regression tests use the stand-alone executable for
+the appropriate model, and read input files which define the different
+tests.
 Both are run regularly as the basis of a nightly test procedure, but
@@
 \subsubsection{Running the tests}
+
+A number of different scripts are provided for running serial, parallel,
+or GPU tests. For each test script, there is the option to run either
+the relevant unit tests, or regression tests, or both. For example, to
+run both serial unit tests and serial regression tests for the D3Q19
+model, invoke
+\begin{lstlisting}
+./test-serial.sh -u -r d3q19
+\end{lstlisting}
+in the test directory.
+The model selection follows the naming scheme described for compilation
+in the previous section.
+See the individual scripts for further information on the command line
+options.
 

### 2015-10-23 trunk/docs/develop.tex
Issue 1405: added new docs directory with content
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2764 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  develop.tex
%
%  Further information for delvelopers
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2014-2015 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Further Information for Developers}

\subsection{Address models for 3-dimensional fields}

The code allows for different order of storage associated with 3-dimensional
fields (scalars, vectors, and so on). For historical reasons these are
referred to as `Model' and `Model~R' options, which correspond
to array-of-structures and structure-of-arrays layout, respectively. Early
versions of the code for CPU were written to favour summation of LB
distributions on a per lattice site basis in operations such as
$\rho(\mathbf{r}) = \sum_i f_i(\mathbf{r})$. This is array-of-structures,
where the $f_i$ are stored contiguously per site. Introduction of GPU
versions, where memory coalescing favours the opposite memory order,
were then referred to as `Model~R', the `R' standing for `reverse'.

The memory layouts are discussed below. In all cases, a 3-d field occupies
lattice sites with a unique spatial index determine by position, and
computed via \texttt{coords\_index()}. These position indices will be denoted
$r_0, r_1, \ldots, r_{n_s-1}$ where $n_s$ is the total number of lattice
sites (including halo points).

\subsubsection{Rank 1 objects (to include scalar fields)}

\textbf{ADDR\_MODEL}: The array-of-structures or Model order for an
$n-$vector field
with components $ v_\alpha = (v_0, v_1, \ldots, v_{n-1})$ is, schematically:
\[
\underbrace{ \boxed{v_0\vphantom{v_{n-1}}} \boxed{v_1\vphantom{v_{n-1}}}
\boxed{\ldots\vphantom{v_{n-1}}} \boxed{v_{n-1}} }_{r_0}
\underbrace{ \boxed{v_0\vphantom{v_{n-1}}} \boxed{v_1\vphantom{v_{n-1}}}
\boxed{\ldots\vphantom{v_{n-1}}} \boxed{v_{n-1}} }_{r_1}
\underbrace{ \boxed{v_0\vphantom{v_{n-1}}} \boxed{v_1\vphantom{v_{n-1}}}
\boxed{\ldots\vphantom{v_{n-1}}} \boxed{v_{n-1}} }_{r_2} \ldots
\underbrace{ \boxed{v_0\vphantom{v_{n-1}}} \boxed{v_1\vphantom{v_{n-1}}}
\boxed{\ldots\vphantom{v_{n-1}}} \boxed{v_{n-1}} }_{r_{n_s-1}}
\]

\textbf{ADDR\_MODEL\_R}: The Model~R version is:
\[
\overbrace{
\begin{array}{cc} \boxed{v_0} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{v_0} \\  _{r_1} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{v_0}} \\ \phantom{r_2}\end{array} \mkern-22mu
\begin{array}{cc} \boxed{v_0} \\  _{r_{n_s-1}} \end{array}
}^{component_0}
\overbrace{
\begin{array}{cc} \boxed{v_1} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{v_1} \\  _{r_1} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{v_0}} \\ \phantom{r_2}\end{array} \mkern-22mu
\begin{array}{cc} \boxed{v_1} \\  _{r_{n_s-1}} \end{array}
}^{component_1}
\begin{array}{cc} \ldots \\ \phantom{r_2} \end{array}
\overbrace{
\begin{array}{cc} \boxed{v_{n-1}} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{v_{n-1}} \\  _{r_1} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{v_{n-1}}} \\ \phantom{r_2}\end{array} \mkern-17mu
\begin{array}{cc} \boxed{v_{n-1}} \\  _{r_{n_s-1}} \end{array}
}^{component_{n-1}}
\]

A scalar field has $n=1$.

% ADDR_MACRO(n_s, n, i, \alpha) in this notation corresponds to
% ADDR_MACRO(nsites, nfield, index, ifld)

\subsubsection{Rank 2 objects (to include dyadic tensor fields)}

\textbf{ADDR\_MODEL}:
A general rank 2 tensor $t_{\alpha\beta}$ with components $(t_{0,0},
\ldots, t_{m-1,n-1})$ is stored as:
\begin{gather*}
\underbrace{
\boxed{t_{0,0}\vphantom{t_{m-1,n-1}}} \boxed{t_{1,0}\vphantom{t_{m-1,n-1}}}
\boxed{\ldots\vphantom{t_{m-1,n-1}}} \boxed{t_{m-1,0}\vphantom{t_{m-1,n-1}}}
\boxed{t_{0,1}\vphantom{t_{m-1,n-1}}} \boxed{t_{1,1}\vphantom{t_{m-1,n-1}}}
\boxed{\ldots\vphantom{t_{m-1,n-1}}} \boxed{t_{m-1,1}\vphantom{t_{m-1,n-1}}}
\ldots
\boxed{t_{0,n-1}\vphantom{t_{m-1,n-1}}}\boxed{t_{1,n-1} \vphantom{t_{m-1,n-1}}}
\boxed{\ldots\vphantom{t_{m-1,n-1}}} \boxed{t_{m-1,n-1}\vphantom{t_{m-1,n-1}}}
}_{r_0}
\ldots
\\
\ldots
\underbrace{
\boxed{t_{0,0}\vphantom{t_{m-1,n-1}}} \boxed{t_{1,0}\vphantom{t_{m-1,n-1}}}
\boxed{\ldots\vphantom{t_{m-1,n-1}}} \boxed{t_{m-1,0}\vphantom{t_{m-1,n-1}}}
\boxed{t_{0,1}\vphantom{t_{m-1,n-1}}} \boxed{t_{1,1}\vphantom{t_{m-1,n-1}}}
\boxed{\ldots\vphantom{t_{m-1,n-1}}} \boxed{t_{m-1,1}\vphantom{t_{m-1,n-1}}}
\ldots
\boxed{t_{0,n-1}\vphantom{t_{m-1,n-1}}}\boxed{t_{1,n-1} \vphantom{t_{m-1,n-1}}}
\boxed{\ldots\vphantom{t_{m-1,n-1}}} \boxed{t_{m-1,n-1}\vphantom{t_{m-1,n-1}}}
}_{r_{n_s-1}}
\end{gather*}
Dyadic tensors, for example the gradient of a vector field
$\partial_\alpha v_\beta$ in three dimensions, are stored in corresponding
fashion with $m=3$ and
$\partial_\alpha = (\partial_x, \partial_y, \partial_z)$.

\textbf{ADDR\_MODEL\_R}: The Model R version is
\[
\overbrace{
\begin{array}{cc} \boxed{t_{0,0}} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{t_{0,0}} \\  _{r_1} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{t_{0,0}}} \\ \phantom{r_2}\end{array} \mkern-18mu
\begin{array}{cc} \boxed{t_{0,0}} \\  _{r_{n_s-1}} \end{array}
}^{component_{0,0}}
\overbrace{
\begin{array}{cc} \boxed{t_{1,0}} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{t_{1,0}} \\  _{r_1} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{t_{0,0}}} \\ \phantom{r_2}\end{array} \mkern-18mu
\begin{array}{cc} \boxed{t_{1,0}} \\  _{r_{n_s-1}} \end{array}
}^{component_{1,0}}
\begin{array}{cc} \ldots\vphantom{t_{n-1}} \\ \phantom{r_2}\end{array}
\overbrace{
\begin{array}{cc} \boxed{t_{m-1,n-1}} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{t_{m-1,n-1}} \\  _{r_1} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{t_{0,0}}} \\ \phantom{r_2}\end{array} \mkern-17mu
\begin{array}{cc} \boxed{t_{m-1,n-1}} \\  _{r_{n_s-1}} \end{array}
}^{component_{m-1,n-1}}
\]

% ADDRRES_MACRO(n_s, m, n, i, alpha, beta) corresponds to, e.g.,
% ADDRESS_MACRO(nsites, NVECTOR, NQAB, index, ia, ib)

\subsubsection{Compressed rank 2 objects}

A symmetric tensor $S_{\alpha\beta}$ in three dimensions has six independent
components. It may be convenient to store this in compressed form as a
rank 1 vector $(S_{xx}, S_{xy}, S_{xz}, S_{yy}, S_{yz}, S_{zz})$ to eliminate
redundent storage.

A symmetric traceless rank 2 tensor --- for example, the Landau-de Gennes
liquid crystal order parameter $Q_{\alpha\beta}$ --- has five independent
components. This is stored as a rank 1 vector with five components
$(Q_{xx}, Q_{xy}, Q_{xz}, Q_{yy}, Q_{yz})$ to eliminate redundant
storage. API calls are provided to expand the compressed format to the
full rank-2 representation $Q_{\alpha\beta}$ and, conversely, to compress
the full representation to five components.

\subsubsection{Rank 3 objects (to include triadic tensor fields)}

The general rank 3 object $t_{\alpha\beta\gamma}$ with components
$(t_{0,0,0}, \ldots, t_{m-1,n-1,p-1})$ is stored in a manner which
generalises from the above, i.e., with the rightmost index running
fastest. Diagrams are omitted, but Model and Model R storage patterns
follow the same form as seen above.

A triadic tensor, for example the general second derivative of a vector
field $\partial_\alpha \partial_\beta v_\gamma$ may be stored as a rank
3 object.

\subsubsection{Compressed rank 3 objects}

Symmetry properties may be used to reduce the storage requirement
associated with rank 3 tensors. `For example, the gradient of the
liquid crystal order parameter $\partial_\gamma Q_{\alpha\beta}$
may be stored as a rank 2 object. The exact requirement will depend
on the construction of the tensor.

%
% The generalised macro would be
% ADDR_MACRO(nsites, m, n, p, i, alpha, beta, gamma)

\subsubsection{LB distribution data}

\textbf{ADDR\_MODEL}: For the LB distributions, up to two distinct
distributions can be accommodated to allow for a binary fluid
implementation, although the usual situation is to have only one.
The Model order for two $N$-velocity distributions $f$ and $g$ is:
\[
\underbrace{
\boxed{f_0\vphantom{f_{N-1}}}    \boxed{f_1\vphantom{f_{N-1}}}
\boxed{\ldots\vphantom{f_{N-1}}} \boxed{f_{N-1}} \,
\boxed{g_0\vphantom{f_{N-1}}}        \boxed{g_1\vphantom{f_{N-1}}}
\boxed{\ldots\vphantom{f_{N-1}}} \boxed{g_{N-1}\vphantom{f_{N-1}}}}_{r_0}
\ldots
\underbrace{
\boxed{f_0\vphantom{f_{N-1}}}    \boxed{f_1\vphantom{f_{N-1}}}
\boxed{\ldots\vphantom{f_{N-1}}} \boxed{f_{N-1}} \,
\boxed{g_0\vphantom{f_{N-1}}}        \boxed{g_1\vphantom{f_{N-1}}}
\boxed{\ldots\vphantom{f_{N-1}}} \boxed{g_{N-1}\vphantom{f_{N-1}}}}_{r_{n_s=1}}
\]

\textbf{ADDR\_MODEL\_R}: The Model R order is:

\[
\overbrace{
\overbrace{
\begin{array}{cc} \boxed{f_0} \\  _{r_0} \end{array} \mkern-18mu
\begin{array}{cc} \boxed{\ldots\vphantom{f_0}} \\ \phantom{r_2}\end{array} \mkern-22mu
\begin{array}{cc} \boxed{f_0} \\  _{r_{n_s-1}} \end{array}
}^{f_0}
\begin{array}{cc} \ldots \\ \phantom{r_2} \end{array}
\overbrace{
\begin{array}{cc} \boxed{f_{N-1}} \\  _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{f_{N-1}}} \\ \phantom{r_2}\end{array} \mkern-17mu
\begin{array}{cc} \boxed{f_{N-1}} \\  _{r_{n_s-1}} \end{array}
}^{f_{N-1}}
}^{f_i}
\overbrace{
\overbrace{
\begin{array}{cc} \boxed{g_0\vphantom{f_{N-1}}} \\ _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{f_{N-1}}} \\ \phantom{r_2}\end{array} \mkern-22mu
\begin{array}{cc} \boxed{g_0\vphantom{f_{N-1}}} \\  _{r_{n_s-1}} \end{array}
}^{g_0} \ldots
\overbrace{
\begin{array}{cc} \boxed{g_{N-1}\vphantom{f_{N-1}}} \\ _{r_0} \end{array} \mkern-17mu
\begin{array}{cc} \boxed{\ldots\vphantom{f_{N-1}}} \\ \phantom{r_2}\end{array} \mkern-17mu
\begin{array}{cc} \boxed{g_{N-1}\vphantom{f_{N-1}}} \\  _{r_{n_s-1}} \end{array}
}^{g_{N-1}\vphantom{f_{N-1}}}
}^{g_i} 
\]
For the single-distribution case, this is equivalent to the rank 1 vector
field with $n=N$.

\vfill
\pagebreak

### 2015-10-23 trunk/docs/lb.tex
Issue 1405: added new docs directory with content
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2764 40876441-3496-0410-a613-a2fca35b2ae2

--- 2014-09-24 branches/testcharge/htdocs/lb.tex
@@
-\section{The Lattice Boltzmann Method}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%  lb.tex
+%
+%  Section on lattice Boltzmann hydrodynamics
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-We review here the lattice Boltzmann method applied to single
-phase problems.
+\section{Lattice Boltzmann Hydrodynamics}
+\label{section:lb-hydrodynamics}
+
+We review here the lattice Boltzmann method applied to a simple
+Newtonian fluid with particular emphsis on the relevant
+implementation in \textit{Ludwig}.
 
@@
 \end{equation}
-Here, repeated indices are understood to be summed over.
+Here, repeated Greek indices are understood to be summed over.
 The conservation law is seem better if the forcing terms of the
@@
 where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
+equations in three dimensions have 10 degrees of freedom
+(or hydrodynamic modes) being
 $\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
@@
 
-\subsection{The Distribution Function and its Moments}
+\subsection{The Lattice Boltzmann Equation}
+
+The Navier-Stokes equation may be approximated in a discrete system
+by the lattice Boltzmann equation (LBE). A discrete density
+distribution function $f_i(\mathbf{r}; t)$ at lattice points $\mathbf{r}$
+and time $t$ evolves according to
+\begin{equation}
+f_i (\mathbf{r} + \mathbf{c}_i \Delta t; t + \Delta t) =
+f_i (\mathbf{r}; t) + \sum_j \mathcal{L}_{ij}
+\big( f_i(\mathbf{r};t) - f_i^{\mathrm{eq}}(\mathbf{r};t) \big)
+\end{equation}
+where $\mathbf{c}_i$ is the discrete velocity basis and $\Delta t$
+is the discrete time step. The collision operator
+$\mathcal{L}_{ij}$ provides the mechanism to compute a discrete
+update from the non-equilibrium distribution
+$f_i(\mathbf{r}; t)- f_i^\mathrm{eq}(\mathbf{r};t)$. Additional terms
+may be added to this equation to represent external body forces,
+thermal fluctuations, and so on. These additional terms are discussed
+in the following sections.
+
+\subsubsection{The distribution function and its moments}
 
 In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
+are complemented by the  distribution function
+$f_i(\mathbf{r}; t)$ defined with reference to the
+discrete velocity space $c_{i\alpha}$.
+It is possible to relate the hydrodynamic quatities to the distribution
+function via its moments, that is
 \begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
+\rho(\mathbf{r};t) = \sum_i f_i(\mathbf{r};t),  \quad
+\rho u_\alpha(\mathbf{r};t) = \sum_i f_i(\mathbf{r};t) c_{i\alpha},  \quad
+\Pi_{\alpha\beta}(\mathbf{r};t) =
+\sum_i f_i(\mathbf{r};t) c_{i\alpha} c_{i\beta}.
 \end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
+Here, the index of the summation is over the number of discrete
+velocities used as the basis, a number which will be denoted
+$N_\mathrm{vel}$. For example, in three dimensions
+$N_\mathrm{vel}$ is often 19 and the basis is referred to as D3Q19.
 
-In general, the number of modes supported in $n$ and these can
-be written as
+The number of moments, or modes, supported by a velocity
+set is exactly $N_\mathrm{vel}$, and these can be written in general as
 \begin{equation}
@@
 \end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
+where the $m_i$ are the eigenvectors of the collision matrix in the LBE.
+For example, in the case of the density, all the 
+$m_i^a = 1$ and the mode $M^a$ is the density $\rho = \sum_i f_i$. Note
+that the number of modes supported by a given basis will generally excceed the
+number of hydrodynamic modes; the excess modes have no direct physical
+interpretation and are variously referred to as non-hydrodynamic, kinetic,
+or ghost, modes.
+The ghost modes take no part in bulk hydrodynamics, but may become important
+in other contexts, such as thermal fluctuations and near boundaries.
+The distribution function can be related to the modes
+$M^a(\mathbf{r};t)$ via
 \begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
+f_i(\mathbf{r};t) = w_i \sum_a m_i^a N^a M^a(\mathbf{r};t).
 \end{equation}
+In this equation, $w_i$ are the standard LB weights appearing in the
+equilibrium distribution function, while the $N^a$ are a per-mode
+normalising factor uniquely determined by the orthogonality condition
+\begin{equation}
+N^a \sum_i w_i m_i^a m_i^b = \delta_{ab}.
+\end{equation}
+Writing the basis this way has the advantage that the equilibrium
+distribution projects directly into the hydrodynamic modes only.
+Putting it another way, we may write
+\begin{equation}
+f_i^\mathrm{eq} = w_i \big(\rho + \rho c_{i\alpha}u_\alpha / c_s^2
++ (c_{i\alpha} c_{i\beta} - c_s^2\delta_{\alpha\beta})
+(\Pi_{\alpha\beta}^\mathrm{eq} - p\delta_{\alpha\beta})/2c_s^4 \big)
+\end{equation}
+where only (equilibrium) hydrodynamic quantities appear on the right hand side.
 
-\subsubsection{Hydrodynamic modes}
+\subsubsection{Collision and relaxation times}
 
-\subsubsection{D3Q15 ghost modes}
+\subsection{Model Basis Descriptions}
 
-\subsubsection{D3Q19 ghost modes}
+\subsubsection{D2Q9}
 
-\subsection{The Lattice Boltzmann Equation}
+The D2Q9 model in two dimensions consists one zero vector (0,0), four
+vectors of length unity being $(\pm 1,0)$ and $(0, \pm 1)$, and four
+vectors of length $\sqrt{2}$ being $(\pm 1, \pm 1)$. The eigenvectors of
+the collision matrix, with associated weights and normalisers are shown
+in Table~\ref{table-d2q9-spec}.
 
-\subsection{The Collision}
-
-\subsection{External Body Forces}
+\begin{table}[t]
+\begin{center}
+\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
+\hline\hline
+$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
+\hline
+$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
+\hline
+$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
+\hline
+$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
+\hline
+$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
+& $c_{ix} c_{ix} - c_s^2$ \\
+\hline
+$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
+\hline
+$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
+& $c_{iy} c_{iy} - c_s^2$ \\
+\hline\hline
+$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
+\hline
+$J_{ix}$ & - & 0 &  4 & -2 & 4 & 0 & 0 & -4 & -2 & -4 & 3/8
+& $\chi^1 \rho c_{ix}$\\
+\hline
+$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
+& $\chi^1 \rho c_{iy}$\\
+\hline\hline
+$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
+\hline\hline
+\end{tabular}
+\end{center}
+\caption{TABLE CAPTION}
+\label{table-d2q9-spec}
+\end{table}
 
@@
 
-\subsubsection{Fluctuating ghost modes: D3Q15}
+\subsection{Hydrodynamic Boundary Conditions}
 
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
+\subsubsection{Bounce-Back on Links}
 
@@
 
-\subsubsection{Boundary walls}
+\subsubsection{Fixed objects}
 
-\subsection{Model Details}
-
-\subsubsection{D2Q9}
-
-These are the eigenvalues and eigenvectors of the collision
-matrix used for D2Q9
-
-\begin{table}
-\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
-\hline
-$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
-& $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
-& $c_{iy} c_{iy} - c_s^2$ \\
-\hline\hline
-$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - & 0 &  4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{iy}$\\
-\hline\hline
-$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
+\subsubsection{Moving objects}
 
@@
 \begin{table}
+\centering
+\tabcolsep=4pt
 \begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
@@
 \end{tabular}
+
+\caption{Table showing the details of the basis used for the D3Q15 model
+in three dimensions. The fifteen modes $M^a$ include two scalar kinetic
+modes $\chi^1$ and $\chi^3$, and one vector kinetic mode $J_{i\alpha}$.
+The weights in the equilibrium distribution are $w_i$ and the normaliser
+for each mode is $N^a$. The eigenvectors of the collision matrix are the
+colms of the transformation matrix $m_i^a$. The prefactor $p$ simply
+multiplies all elements of $m_i^a$ in that row as a convenience.}
 \end{table}
@@
 \begin{table}
+\centering
+\tabcolsep=4pt
 \begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
@@
 \end{tabular}
+\caption{Table showing the details of the basis used for the D3Q19 model in
+three dimensions. }
 \end{table}

### 2015-10-23 docs/plan.tex
Issue 1405: added new docs directory with content
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2764 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-08-30 trunk/doc/plan.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[11pt,twoside]{article}
-
-\setlength{\hoffset}{-1in}
-\setlength{\voffset}{-1in}
-
-\setlength{\topmargin}{2cm}
-\setlength{\evensidemargin}{3.1cm}
-\setlength{\oddsidemargin}{3.1cm}
-
-\setlength{\textwidth}{14.8cm}
-\setlength{\textheight}{23cm}
-
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{\smallskipamount}
-
-\newcommand{\inputkey}[1]{\framebox{\textbf{\texttt{#1}}}}
-\newcommand{\e}[1]{\cdot10^{#1}}
-\newcommand{\beq}{\begin{equation}}
-\newcommand{\eeq}{\end{equation}}
-\newcommand{\beqa}{\begin{eqnarray}}
-\newcommand{\eeqa}{\end{eqnarray}}
-\newcommand{\com}[1]{\textcolor{red}{#1}}
-\newcommand{\cur}[1]{{\textit{#1}}}
-
-\begin{document}
-
-\setcounter{page}{1}
-
-\tableofcontents
-
-\newpage
-
-\setcounter{page}{1}
 
@@
 
-%\newpage
-%\section{Coordination Meetings}
-
-\begin{thebibliography}{99}
-\bibitem{ieee-828-2012} IEEE Standard 828-2011 ``IEEE Standard for
-Software Configuration Management Plans''
-\bibitem{mpi-standard} Message Passing Interface Forum. MPI: A Message
-Passing Interface Standard Version 1.3 (1998).
+%\begin{thebibliography}{99}
+%\bibitem{ieee-828-2012} IEEE Standard 828-2011 ``IEEE Standard for
+%Software Configuration Management Plans''
+%\bibitem{mpi-standard} Message Passing Interface Forum. MPI: A Message
+%Passing Interface Standard Version 1.3 (1998).
 %\bibitem{paraview} Paraview is a standard visualisation package. 
-\bibitem{apacheAPR} See, e.g., the Apache Portable Runtime Project ``APR's Version Numbering''  https://apr.apache.org/versioning.html (Accessed 2015).
-\end{thebibliography}
-
-\vfill\eject
-
-\end{document}
+%\bibitem{apacheAPR} See, e.g., the Apache Portable Runtime Project ``APR's Version Numbering''  https://apr.apache.org/versioning.html (Accessed 2015).
+%\end{thebibliography}

### 2015-11-05 docs/references.tex
Issue 1405: added new docs directory with content
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2764 40876441-3496-0410-a613-a2fca35b2ae2
Additional notes on addressing model
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2765 40876441-3496-0410-a613-a2fca35b2ae2

--- 2013-01-31 htdocs/references.tex
@@
 %
-%  $Id$
+%  $Id: references.tex 1902 2013-01-31 15:11:10Z ohenrich $
 %
@@
 \texttt{arXiv:cond-mat/0503175v1} (2005).
+
+\bibitem{apacheAPR} 
+Apache Portable Runtime Project ``APR's Version Numbering''.
+See, e.g., \texttt{https://apr.apache.org/versioning.html)}
+(accessed November 2015).
 
@@
 
-\bibitem{cpc}
+\bibitem{desplat2001}
 Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
@@
 
+\bibitem{gray2013}
+A. Gray and K. Stratford,
+TargetDP reference.
+
 \bibitem{h59}
@@
 IEEE Standard 208-2005, IEEE Standard for Software Configuration Management
-Plans. See \texttt{http://ieeexplore.ieee.org/xpl/standards.jsp}
-(accessed 2011).
+Plans.
+See \texttt{http://ieeexplore.ieee.org/xpl/standards.jsp}
+(accessed November 2015).
+
+\bibitem{ieee-828-2012}
+IEEE Standard 828-2012. IEEE Standard for Configuration
+Management in Systems and Software Engineering. 
+See \texttt{http://ieeexplore.ieee.org/xpl/standards.jsp}
+(accessed November 2015).
 
@@
 
-\bibitem{viv}
+\bibitem{kendon2001}
 Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and

### 2015-11-05 trunk/docs/develop.tex
Additional notes on addressing model
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2765 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \subsubsection{Rank 1 objects (to include scalar fields)}
+\label{subsection:addressing-model-rank1}
 
@@
 Symmetry properties may be used to reduce the storage requirement
-associated with rank 3 tensors. `For example, the gradient of the
+associated with rank 3 tensors. For example, the gradient of the
 liquid crystal order parameter $\partial_\gamma Q_{\alpha\beta}$
@@
 
+\subsection{Generalised model for SIMD vectorisation}
+
+To promote parallelism at the instruction level, it is necessary to
+insure the innermost loop of any loop construct has an extent which
+is the vector length for the target architecture. The memory layout
+should therefore be adjusted accordingly. This means the MODEL format
+is generalised to a (rather clumsily named) array of structures of
+short vectors. The length of the short vectors is the SIMD vector
+length.
+
+The outermost loop in this context is always to be the loop over
+lattice sites, which is adjusted according to the vector length;
+see Section \ref{subsection:addressing-how-to} below for practical
+examples.
+
+The Model R picture, which targets coalescene, can be viewed as
+naturally supporting SIMD vectorisation by virtue of allowing
+contiguous access to individual quantities as a function of lattice
+index. (If SIMD vectorisation is wanted at all on GPU architecture,
+it can be as a means to releive register pressure.) Model R therefore
+remains unaltered and we concentrate on the Model picture.
+
+\subsubsection{Rank 1 objects}
+
+\textbf{VADDR\_MODEL}: The structure of short vectors is based on the
+SIMD vector length which we here denote $V$:
+\[
+\underbrace{
+\overbrace{
+\begin{array}{cc} \boxed{v_0} \\  _{r_0} \end{array} \mkern-17mu
+\begin{array}{cc} \boxed{v_0} \\  _{r_1} \end{array} \mkern-17mu
+\begin{array}{cc} \boxed{\ldots\vphantom{v_0}} \\ \phantom{r_2}\end{array} \mkern-20mu
+\begin{array}{cc} \boxed{v_0} \\  _{r_{V-1}} \end{array}
+}^{component_0}
+\overbrace{
+\begin{array}{cc} \boxed{v_1} \\  _{r_0} \end{array} \mkern-17mu
+\begin{array}{cc} \boxed{v_1} \\  _{r_1} \end{array} \mkern-17mu
+\begin{array}{cc} \boxed{\ldots\vphantom{v_0}} \\ \phantom{r_2}\end{array} \mkern-20mu
+\begin{array}{cc} \boxed{v_1} \\  _{r_{V-1}} \end{array}
+}^{component_1}
+%\begin{array}{cc} \ldots \\ \phantom{r_2} \end{array}
+\overbrace{
+\begin{array}{cc} \boxed{v_{n-1}} \\  _{r_0} \end{array} \mkern-17mu
+\begin{array}{cc} \boxed{v_{n-1}} \\  _{r_1} \end{array} \mkern-17mu
+\begin{array}{cc} \boxed{\ldots\vphantom{v_{n-1}}} \\ \phantom{r_2}\end{array}
+\mkern-17mu
+\begin{array}{cc} \boxed{v_{n-1}} \\  _{r_{V-1}} \end{array}
+}^{component_{n-1}}}_{\mathrm{SIMD\ BLOCK}}
+\quad\ldots
+\]
+Subsequent SIMD blocks involve lattice sites $r_V \ldots r_{2V-1}$,
+$r_{2V} \ldots r_{3V-1}$, and so on. If the SIMD vector length is unity,
+this collapses to the standard Model picture shown in
+Section~\ref{subsection:addressing-model-rank1}.
+
+The generalisation of this approach to rank 2 and rank 3 objects 
+follows the corresponding Model implementation.
+
+\subsection{Addressing: How to?}
+\label{subsection:addressing-how-to}
+
+To provide a transparent interface for addressing vector fields,
+a single API is provided which implements the addressing order
+selected at compile time. This interface is always based
+on a fixed order of indices which may include the
+vector index of the innermost SIMD loop if present. This
+allows both vectorised and non-vectorised loops to be constructed
+in a consistent manner.
+
+The usual model for constructing a loop involving all lattice sites
+(in this case without haloes) would be, in the absence of vectorisation:
+\begin{lstlisting}
+for (ic = 1; ic <= nlocal[X]; ic++) {
+  for (jc = 1; jc <= nlocal[Y]; jc++) {
+    for (kc = 1; kc <= nlocal[Z]; kc++) {
+
+      index = coords_index(ic, jc, kc);
+      /* Perform operation per lattice site ... */
+    }
+  }
+}
+\end{lstlisting}
+where final array indexing is based on the coordinate index and is
+specific to the object at hand. To allow transparent switching
+between Model and Model R versions, the indexing must be abstracted.
+As a concrete example, the following considers a rank 1 3-vector:
+\begin{lstlisting}
+for (ic = 1; ic <= nlocal[X]; ic++) {
+  for (jc = 1; jc <= nlocal[Y]; jc++) {
+    for (kc = 1; kc <= nlocal[Z]; kc++) {
+
+      index = coords_index(ic, jc, kc);
+      for (ia = 0; ia < 3; ia++) {
+        iref = addr_rank1(nsites, 3, index, ia);
+        array[iref] = ...;
+      }
+    }
+  }
+}
+\end{lstlisting}
+Here, the function \texttt{addr\_rank1()} performs the appropriate
+arithmetic to reference the correct 1-d array element in either
+address model. We note that
+this can be implemented to operate appropriately when the SIMD vector
+length is an arbitrary number.
+
+An equivalent vectorised version may be constructed as follows:
+\begin{lstlisting}
+for (n = 0; n < nsites; n += SIMDVL) {
+  index = coords_indexv(n, nextra);       /* Proposal */
+  for (ia = 0; ia < 3; ia++) {
+    for (iv = 0; iv < SIMDVL; iv++) {
+      iref = vaddr_rank1(nsites, 3, index, ia, iv);
+      array[iref] = ...;
+    }
+  }
+}
+\end{lstlisting}
+Note that a different function \texttt{vaddr\_rank1()} with an additional
+argument which is the index of the vector loop is used. Note also that
+vectorisaed versions must take into account the extent that the kernel
+extends into the halo region (here \texttt{nextra}), which involves
+logic to avoid lattice sites which are not required (not shown in this
+example).
+
+Note that, in practice, 
+the loop over lattice sites is further abstracted in the code to
+accommodate thread-level parallelism via targetDP interface.
+
 \vfill

### 2016-01-12 branches/kevin-1397/docs/devel.tex
Add eature branch
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2811 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-10-23 trunk/docs/devel.tex
@@
 bash-3.2$ ls
-LICENSE         docs            src             tests           version.h
-README          mpi_s           targetDP        util
+config  LICENSE      mpi_s   src       tests  version.h
+docs    Makefile.mk  README  targetDP  util
 \end{lstlisting}
@@
 (MPI), which is used to provide domain-decomposition based parallelism.
+Note that the code will also compile under NVIDIA \texttt{nvcc}, i.e.,
+it also meets the C++ standard.
 
@@
 
-Compilation of the main code is controlled by the \texttt{Makefile} in
-the \texttt{./src} directory. This section discusses a number of issues
+Compilation of the main code is controlled by the \texttt{config.mk} in
+the  top-level directory. A number of example \texttt{config.mk} files
+are provided in the \texttt{config} directory (which can be copied and
+adjusted as needed). This section discusses a number of issues
 which influence compilation.
@@
 
-The \texttt{Makefile} provided in the source directory has a number of
-targets to allow different models to be compiled. Before compilation,
-the Makefile must be edited provide details of the local C compiler(s).
-The relevant lines are usually limited to:
+The \texttt{Makefile.mk} and \texttt{config.mk} files in the top level
+directory control options for compilation. Before compilation,
+the \texttt{config.mk} file must be edited provide details of the local
+C compiler(s). The relevant lines are usually limited to, e.g.:
 \begin{lstlisting}
@@
 MPI, and \texttt{CFLAGS} provides compiler switches (often
-related to optimisation). Beyond these variables, changes to
-the \texttt{Makefile} should not usually be required.
+related to optimisation).
+
+The \texttt{Makefile} provided in each subdirectory includes
+the \texttt{config.mk} file via the top level \texttt{Makefile.mk}.
+Changes to the individual Makefiles should therefore not be required.
+
+If GPU compilation is required, the \texttt{config.mk} file should
+specify the local details concerning \texttt{nvcc}. This includes
+explicit information on the location of MPI headers and libraries
+if an MPI-parallel GPU version is required.
 
@@
 switch off the assertions with the standard \texttt{NDEBUG}
-preprocessor flag. Add \texttt{-DNDEBUG} to \texttt{OPTS} in
-the \texttt{Makefile}.
+preprocessor flag. Add \texttt{-DNDEBUG} to \texttt{CFLAGS} in
+the \texttt{config.mk}.
 
@@
 
-A number of different scripts are provided for running serial, parallel,
-or GPU tests. For each test script, there is the option to run either
+A number of different \texttt{Makefile} targets are provided for
+running serial, parallel,
+or GPU tests. For each test, there is the option to run either
 the relevant unit tests, or regression tests, or both. For example, to
@@
 \begin{lstlisting}
-./test-serial.sh -u -r d3q19
+$ cd tests
+$ make compile-run-serial
 \end{lstlisting}
@@
 in the previous section.
-See the individual scripts for further information on the command line
-options.
 

### 2016-01-12 branches/kevin-1397/targetDP/doc/examples.tex
Add eature branch
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2811 40876441-3496-0410-a613-a2fca35b2ae2

\chapter{Example}\label{chapter:examples}

Consider a simple example: the scaling of a 3-vector
field by a constant, as implemented in a sequential programming style in Figure \ref{fig:scaleorig}. On each lattice site exists a 3-vector (a collection of three values corresponding to the 3 spatial dimensions). The outer loop corresponds to lattice sites, and the inner loop to the 3 components within each lattice site. This is a simple example of operations on ``multi-valued'' data, a very common situation in scientific simulations.

The lattice-based parallelism corresponding to the outer loop can be
mapped to data parallel hardware using targetDP. We introduce targetDP
by replacing the sequential code with the function given in Figure
\ref{fig:scalekernel}. The \verb+t_+ syntax is used to identify target
data structures. The \verb+__targetEntry__+ syntax is used to specify
that this function is to be executed on the target, and it will be
called from host code.  We expose the lattice-based parallelism to
each of the TLP and ILP levels of hardware parallelism through use of
the combination \verb+__targetTLP__(baseIndex,N)+ and
\verb+__targetILP__(vecIndex)+ (See Sections \ref{sec:TLP} and
\ref{sec:ILP}). The former specifies that lattice-based parallelism
should be mapped to TLP, where each thread operates on a chunk of
lattice sites. The latter specifies that the sites within each chunk
should be mapped to ILP. It can be seen that the \verb+t_field+ array
is accessed by combining these indexes. The size of the chunk can be
set within the targetDP implementation, to give the best performance
for a particular architecture.

The \verb+scale+ function is called from host code as shown in Figure \ref{fig:scalelaunch}. The memory management facilities are used to allocate and transfer data to and from the target, as described in Chapter \ref{chap:memmanage}.

\begin{figure}[h]
\begin{lstlisting}
 for (idx = 0; idx < N; idx++) { //loop over lattice sites
   int iDim;
   for (iDim = 0; iDim < 3; iDim++) 
      field[iDim*N+idx] = a*field[iDim*N+idx];
 }

\end{lstlisting}
\caption{A sequential implementation of the scalar multiplication of each element of a lattice data structure.}\label{fig:scaleorig}
\end{figure}

\begin{figure}[h]
\begin{lstlisting}
__targetEntry__ void scale(double* t_field) { (*@\label{example:scalekernel:targetEntry}@*)

  int baseIndex;
  __targetTLP__(baseIndex, N) { (*@\label{example:scalekernel:targetTLP}@*)

    int iDim, vecIndex;
    for (iDim = 0; iDim < 3; iDim++) {

      __targetILP__(vecIndex)  \ (*@\label{example:scalekernel:targetILP}@*) 
         t_field[iDim*N + baseIndex + vecIndex] =  \
	 t_a*t_field[iDim*N + baseIndex + vecIndex];       	  
    }
  }
  return;
}
\end{lstlisting}
\caption{The targetDP implementation of the scalar multiplication kernel.}\label{fig:scalekernel}
\end{figure}

\begin{figure}[h]
\begin{lstlisting}
  targetMalloc((void **) &t_field, datasize); (*@\label{example:scalelaunch:targetMalloc}@*)

  copyToTarget(t_field, field, datasize); (*@\label{example:scalelaunch:copyToTarget}@*) 
  copyConstToTarget(&t_a, &a, sizeof(double)); (*@\label{example:scalelaunch:copyConstToTarget}@*)

  scale __targetLaunch__(N) (t_field); (*@\label{example:scalelaunch:targetLaunch}@*) 
  targetSynchronize();(*@\label{example:scalelaunch:targetSynchronize}@*)

  copyFromTarget(field, t_field, datasize); (*@\label{example:scalelaunch:copyFromTarget}@*)

  targetFree(t_field); (*@\label{example:scalelaunch:targetFree}@*) 
\end{lstlisting}
\caption{The host code used to invoke the targetDP scalar multiplication kernel.}\label{fig:scalelaunch}
\end{figure}

### 2016-01-12 branches/kevin-1397/targetDP/doc/memory_management.tex
Add eature branch
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2811 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-03-12 targetDP/doc/memory_management.tex
@@
-\chapter{Memory Management}
+\chapter{Memory Management}\label{chap:memmanage}
 
@@
 \begin{verbatim}
-void targetMalloc(void **targetPtr, size_t n);
+void targetMalloc(void** targetPtr, size_t n);
 \end{verbatim}
@@
 
-The \verb+targetCalloc+ function allocates, and initializes to zero, memory on the target.
+The \verb+targetCalloc+ function allocates, and initialises to zero, memory on the target.
 
@@
 \begin{verbatim}
-void targetCalloc(void **targetPtr, size_t n);
+void targetCalloc(void** targetPtr, size_t n);
 \end{verbatim}
@@
 \begin{verbatim}
-void targetFree(void *targetPtr);
+void targetFree(void* targetPtr);
 \end{verbatim}
@@
 \begin{verbatim}
-void copyToTarget(void *targetData, const void *data, size_t n);
+void copyToTarget(void* targetData, const void* data, size_t n);
 \end{verbatim}
@@
 \begin{verbatim}
-void copyFromTarget(void *data, const void *targetData, size_t n);
+void copyFromTarget(void* data, const void* targetData, size_t n);
 \end{verbatim}
@@
 \item \verb+variableName+: The name of the variable or array.
-\item \verb+type+: The type of variabe or array.
+\item \verb+type+: The type of variable or array.
 \end{itemize}
@@
 \begin{verbatim}
-void copyConstToTarget(void *targetData, const void *data, size_t n);
+void copyConstToTarget(void* targetData, const void* data, size_t n);
 \end{verbatim}
@@
 \begin{verbatim}
-void targetConstAddress(void **address, objectType object);
+void targetConstAddress(void** address, objectType object);
 \end{verbatim}
@@
 
-The \verb+targetInit3D+ initialises the environment required to perform any of following the targetDP 3D lattice operations.
+The \verb+targetInit3D+ initialises the environment required to perform any of the ``3D'' operations described in the rest of this chapter.
 
@@
 
-The \verb+targetFinalize3D+ finalizes the targetDP 3D environment.
+The \verb+targetFinalize3D+ finalises the targetDP 3D environment.
 
@@
 \begin{verbatim}
-void copyToTargetBoundary3D(void *targetData, const void *data, size_t extent3D[3], size_t nField, size_t offset, size_t depth);
+void copyToTargetBoundary3D(void* targetData, const void* data, 
+              size_t extent3D[3], size_t nField, 
+              size_t offset, size_t depth);
 \end{verbatim}
@@
 \begin{verbatim}
-void copyFromTargetBoundary3D(void *data, const void *targetData, size_t extent3D[3], size_t nField, size_t offset, size_t depth);
+void copyFromTargetBoundary3D(void* data, const void* targetData, 
+                 size_t extent3D[3], size_t nField, 
+                 size_t offset, size_t depth);
 \end{verbatim}
@@
 \begin{verbatim}
-void copyToTargetPointerMap3D(void *targetData, const void *data, 
+void copyToTargetPointerMap3D(void* targetData, const void* data, 
            size_t extent3D[3], size_t nField, 
@@
 \item \verb+nField+: The number of fields per lattice site.
-\item \verb+includeNeighbours+: A boolean switch to specify whether each included site should also have it's neighbours included (in the 19-point 3D stencil).
+\item \verb+includeNeighbours+: A Boolean switch to specify whether each included site should also have it's neighbours included (in the 19-point 3D stencil).
 \item \verb+pointerArray+: An array of \verb+nSite+ pointers, where \verb+nSite+ is the total number of lattice sites.
@@
 \begin{verbatim}
-void copyFromTargetPointerMap3D(void *data, const void *targetData, 
+void copyFromTargetPointerMap3D(void* data, const void* targetData, 
            size_t extent3D[3], size_t nField, 

### 2016-01-12 branches/kevin-1397/src/brownian.c

@@ -1,0 +1,511 @@
+/*****************************************************************************
+ *
+ *  brownian.c
+ *
+ *  $Id: brownian.c,v 1.4 2010-10-15 12:40:02 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2007 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <math.h>
+
+#include "pe.h"
+#include "ran.h"
+#include "coords.h"
+#include "colloids.h"
+#include "colloids_halo.h"
+#include "physics.h"
+#include "util.h"
+
+void brownian_step_ermak_buckholz(void);
+void brownian_step_no_inertia(void);
+void brownian_step_no_inertia_test(void);
+void brownian_set_random(void);
+
+static double dt_ = 1.0;  /* Time step; fixed here to be 1 LB step */
+#define NTIME 16
+
+/*****************************************************************************
+ *
+ *  do_brownian_dynamics
+ *
+ *  Perform a number of Brownian Dynamics steps.
+ *
+ *****************************************************************************/
+
+void do_brownian_dynamics() {
+
+  int     ia;
+  int     ic, jc, kc;
+  colloid_t * p_colloid;
+  int i, n, nt, nmax = 100000;
+  double u[3][NTIME];
+  double r[3][NTIME];
+  double s[3][NTIME];
+  double correlator_uu[NTIME];
+  double correlator_ur[NTIME];
+  double correlator_rr[NTIME];
+  double correlator_ss[NTIME];
+  double mass, scaleu, scalex, beta;
+  double dr[3];
+
+  /* Brownian dynamics requires an update. */
+  assert(0);
+
+  brownian_step_no_inertia_test();
+  if (1) return;
+
+  mass  = (4.0/3.0)*pi_*pow(2.3, 3);
+  beta  = 6.0*pi_*get_eta_shear()*2.3/mass;
[…]

### 2016-01-12 branches/kevin-1397/src/colloid.c

@@ -1,0 +1,219 @@
+/*****************************************************************************
+ *
+ *  colloid.c
+ *
+ *  State type for particles including bounce-back on links, wetting,
+ *  magnetic dipoles, and squirmers.
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010-2014 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdio.h>
+
+#include "colloid.h"
+
+/*****************************************************************************
+ *
+ *  colloid_state_read_ascii
+ *
+ *  Returns zero on success.
+ *
+ *****************************************************************************/
+
+int colloid_state_read_ascii(colloid_state_t * ps, FILE * fp) {
+
+  int n;
+  int nread = 0;
+  int ifail = 0;
+
+  const char * isformat = "%24d\n";
+  const char * sformat  = "%24le\n";
+  const char * vformat  = "%24le %24le %24le\n";
+
+  assert(ps);
+  assert(fp);
+
+  nread += fscanf(fp, isformat, &ps->index);
+  nread += fscanf(fp, isformat, &ps->rebuild);
+  nread += fscanf(fp, isformat, &ps->nbonds);
+  nread += fscanf(fp, isformat, &ps->nangles);
+  nread += fscanf(fp, isformat, &ps->isfixedr);
+  nread += fscanf(fp, isformat, &ps->isfixedv);
+  nread += fscanf(fp, isformat, &ps->isfixedw);
+  nread += fscanf(fp, isformat, &ps->isfixeds);
+  nread += fscanf(fp, isformat, &ps->type);
+
+  for (n = 0; n < NBOND_MAX; n++) {
+    nread += fscanf(fp, isformat, &ps->bond[n]);
+  }
+
+  nread += fscanf(fp, isformat, &ps->rng);
+
+  for (n = 0; n < NPAD_INT; n++) {
+    nread += fscanf(fp, isformat, &ps->intpad[n]);
+  }
+
+  nread += fscanf(fp, sformat, &ps->a0);
[…]

### 2016-01-12 branches/kevin-1397/src/field_s.h

@@ -1,0 +1,45 @@
+/*****************************************************************************
+ *
+ *  field_s.h
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2012 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef FIELD_S_H
+#define FIELD_S_H
+
+#include <mpi.h>
+
+#include "io_harness.h"
+#include "field.h"
+
+struct field_s {
+  int nf;                       /* Number of field components */
+  int nhcomm;                   /* Halo width required */
+  double * data;                /* Data on host */
+  double * t_data;              /* Data on target: DEPRECATED (see tcopy below)*/
+
+  MPI_Datatype halo[3];         /* Halo exchange data types */
+  io_info_t * info;             /* I/O Handler */
+  char * name;                  /* "phi", "p", "q" etc. */
+
+  field_t * tcopy;              /* copy of this structure on target */ 
+
+};
+
+
+#ifdef LB_DATA_SOA
+#define FLDADR ADDR_VECSITE_R
+#else
+#define FLDADR ADDR_VECSITE
+#endif
+
+
+#endif

### 2016-01-12 branches/kevin-1397/src/free_energy_tensor.h

@@ -1,0 +1,29 @@
+/*****************************************************************************
+ *
+ *  free_energy_tensor.h
+ *
+ *  Abstract free energy for tensor order parameters.
+ *
+ *  $Id: free_energy_tensor.h,v 1.2 2010-10-15 12:40:02 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 the University of Edinburgh
+ *
+ *****************************************************************************/
+
+#ifndef FREE_ENERGY_TENSOR_H
+#define FREE_ENERGY_TENSOR_H
+
+#include "targetDP.h"
+#include "free_energy.h"
+
+__targetHost__ void   fe_t_molecular_field_set(void (* f)(const int index, double h[3][3]));
+__targetHost__ void   (* fe_t_molecular_field(void))(const int index, double h[3][3]);
+
+__targetHost__ double fe_xi(void);
+__targetHost__ void   fe_xi_set(void);
+
+#endif

### 2016-01-12 branches/kevin-1397/src/gradient_3d_7pt_solid.c

@@ -1,0 +1,1670 @@
+/*****************************************************************************
+ *
+ *  gradient_3d_7pt_solid.c
+ *
+ *  Liquid Crystal tensor order parameter Q_ab.
+ *
+ *  Gradient operations for 3D seven point stencil.
+ *
+ *                        (ic, jc+1, kc)
+ *         (ic-1, jc, kc) (ic, jc  , kc) (ic+1, jc, kc)
+ *                        (ic, jc-1, kc)
+ *
+ *  ...and so in z-direction
+ *
+ *  d_x Q = [Q(ic+1,jc,kc) - Q(ic-1,jc,kc)] / 2
+ *  d_y Q = [Q(ic,jc+1,kc) - Q(ic,jc-1,kc)] / 2
+ *  d_z Q = [Q(ic,jc,kc+1) - Q(ic,jc,kc-1)] / 2
+ *
+ *  nabla^2 Q = Q(ic+1,jc,kc) + Q(ic-1,jc,kc)
+ *            + Q(ic,jc+1,kc) + Q(ic,jc-1,kc)
+ *            + Q(ic,jc,kc+1) + Q(ic,jc,kc-1) - 6 Q(ic,jc,kc)
+ *
+ *  The cholesteric anchoring boundary condition specifies the surface
+ *  free energy
+ *
+ *  f_s = w (Q_ab - Q^s_ab)^2
+ *
+ *  There is a correction related to the surface component of the
+ *  molecular field
+ *
+ *        w (Q_ab - Q^s_ab)
+ *
+ *  and for cholesterics, one related to the pitch wavenumber q0.
+ *  Anchoring is specified in colloids_Q_tensor.c at the moment.
+ *
+ *  A special treatment of edges and corners is required for colloids,
+ *  where an iterative approach is used to get two or three orthogonal
+ *  gradients at fluid points.
+ *
+ *  This will also cope with parallel boundaries separated by one fluid
+ *  points, whatever the solid involved.
+ * 
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2011 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <float.h>
+#include <math.h>
+#include <stdlib.h>
+
+#include "pe.h"
+#include "util.h"
[…]

### 2016-01-14 trunk/targetDP/doc/execution_model.tex
updating spec
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2815 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-10-16 branches/kevin-1407/targetDP/doc/execution_model.tex
@@
 this statement, it will wait until the target region has completed. It
-is possible for the initial thread to execute other instructions
+is possible, in principle, for the initial thread to execute other instructions
 (which do not depend on the results of the target function), after the
@@
 
-\begin{comment}
-Possibly add targetSyncThreads() to
-  allow communication within target functions?]
-\end{comment}
-
-\begin{comment}
-Some text on reductions needed.
-\end{comment}
+In applications, it is sometimes necessary to perform reductions,
+where multiple data values are combined in a certain way. For example,
+values existing on each grid point may be summed into a single total
+value. The targetDP model supports such operations in a simplistic
+way. It is the responsibility of the application to create the array
+of values (using standard targetDP functionality) to act as the input
+to the reduction operation. The application can then pass this array
+to the API function corresponding to the desired reduction operation
+(e.g. \verb+targetDoubleSum+ for the summation of an array of
+double-precision Values). If the required reduction operation does not yet
+exist, the user can simply extend the targetDP API using existing
+functionality as a template.

### 2016-01-14 targetDP/doc/introduction.tex
updating spec
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2815 40876441-3496-0410-a613-a2fca35b2ae2

--- 2015-04-14 trunk/targetDP/doc/introduction.tex
@@
 performance portable: the same source code can be compiled for
-different targets (currently supported are accelerators such as GPUs
-and modern multi-core SIMD CPUs), without performance overheads. The
-targetDP API is aimed at the types of parallelism found in grid-based
-applications, but may be applicable to a wider class of problems. This
-document introduces the targetDP memory and execution models, and
+different targets (where we currently support GPU accelerators and
+modern multicore or manycore SIMD CPUs), without performance
+overheads. The model is appropriate for abstracting the parallelism
+contained within each compute node, and can be combined with, e.g. MPI
+to allow use on systems containing multiple nodes. The targetDP API is
+primarily aimed at the types of parallelism found in grid-based
+applications, but may be applicable to a wider class of problems.
+This document introduces the targetDP memory and execution models, and
 specifies the syntax and functionality of the
@@
 modern computers, which continue to increase in complexity and
-diversity with features including multiple multicore CPUs (often
-distributed across multiple nodes), vector floating point units,
-accelerators such as GPUs and non-uniform distributed memory spaces.
-From a scientist's perspective, it is not only imperative to achieve
-performance, but also to retain maintainability, sustainability and
-portability. The use of a simplistic, well structured and clearly
-defined abstraction layer such as targetDP can allow the programmer to
-express the scientific problems in a way that will automatically
-achieve good performance across the range of leading hardware
-solutions.
+diversity with features including multicore/manycore CPUs, vector
+floating point units, accelerators such as GPUs and non-uniform
+distributed memory spaces.  From a scientist's perspective, it is not
+only imperative to achieve performance, but also to retain
+maintainability, sustainability and portability. The use of a
+simplistic, well structured and clearly defined abstraction layer such
+as targetDP can allow the programmer to express the scientific
+problems in a way that will automatically achieve good performance
+across the range of leading hardware solutions.
 
@@
 Depending on the underlying hardware available, the target could simply be the a CPU, or it could be a separate device such as an accelerator.
-    \item \keyword{CUDA}: Compute Unified Device Architecture. The parallel platform and model created by NVIDIA to allow general purpose programming of their GP architectures.
+    \item \keyword{CUDA}: Compute Unified Device Architecture. The parallel platform and model created by NVIDIA to allow general purpose programming of their GPU architectures.
     \item \keyword{TLP}: Thread Level Parallelism.

### 2016-01-15 targetDP/doc/execution_model.tex
release of Specification 0.1.0
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2817 40876441-3496-0410-a613-a2fca35b2ae2

@@
-\chapter{Execution model}
+\chapter{Execution model}\label{sec:exmodel}
 
 The terminology ``host'' is used to refer to the CPU that is hosting
-the execution of the application, and ``target'' to refer to the
+the execution of the application, and ``target'' refers to the
 device targeted for execution of data parallel operations. The target
-can be the same CPU as the host, or it can be a separate device such
-as an accelerator.
+could be the same CPU as the host, or it could be a separate device such
+as an accelerator (depending on the hardware available).
 
@@
 a structured grid problem each thread is responsible for a subset of
-the grid).
+the grid). Each thread in the team will have a unique index. For some
+architectures, it is important to expose not just thread-level
+parallelism (TLP), but also instruction-level parallelism (ILP). The
+targetDP model facilitates this by allowing striding at the thread
+level, such that each thread can operate on a chunk of data
+(e.g. multiple grid points) at the instruction level. The size of the
+chunk, which we call the ``Virtual Vector Length'' (VVL), can be tuned
+to the hardware.
 
@@
 this statement, it will wait until the target region has completed. It
-is possible for the initial thread to execute other instructions
+is possible, in principle, for the initial thread to execute other instructions
 (which do not depend on the results of the target function), after the
-function launch but before the syncronisation call. This may result in
+function launch but before the synchronisation call. This may result in
 overlapping of host and target execution, and hence optimisation, in
@@
 
-\begin{comment}
-Possibly add targetSyncThreads() to
-  allow communication within target functions?]
-\end{comment}
+In applications, it is sometimes necessary to perform reductions,
+where multiple data values are combined in a certain way. For example,
+values existing on each grid point may be summed into a single total
+value. The targetDP model supports such operations in a simplistic
+way. It is the responsibility of the application to create the array
+of values (using standard targetDP functionality) to act as the input
+to the reduction operation. The application can then pass this array
+to the API function corresponding to the desired reduction operation
+(e.g. \verb+targetDoubleSum+ for the summation of an array of
+double-precision Values). If the required reduction operation does not yet
+exist, the user can simply extend the targetDP API using existing
+functionality as a template.

### 2016-03-03 trunk/docs/input.tex
Updates to trunk/docs
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2870 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  input.tex
%
%  Instructions for contents of the user input file.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2016 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Input File}

\subsection{General}

By default, the run time expects to find user input in a file
\texttt{input} in the current working directory. If a different
file name is required, its name
should be provided as the sole command line argument, e.g.,
\begin{lstlisting}
./Ludwig.exe input_file_name
\end{lstlisting}
If the input file is not located in the current working directory
the code will terminate immediately with an error message.

When an input file is located, its content is read by a single MPI
task, and its contents then broadcast to all MPI relevant tasks.
The format of the file is plain ASCII text, and its contents are
parsed on a line by line basis. Lines may contain the following:
\begin{itemize}
\item comments introduced by \texttt{\#}.
\item key value pairs separated by white space.
\end{itemize}
Blank lines are treated as comments. The behaviour of the code is
determined by a set of key value pairs. Any given key may appear
only once in the input file; unused key value pairs are not reported.
If the key value pairs are not correctly formed, the code will terminate
with an error message and indicate the offending input line.

Key value pairs may be present in the input file, but have no effect for
any given run: they are merely ignored. Relevant control parameters for
given input are reported in the standard output.

\subsubsection{Key value pairs}

Key value pairs are made up of a key --- an alphanumeric string with no
white space --- and corresponding value following white space. Values
may take on the follow forms:
\begin{lstlisting}
key_string           value_string

key_integer_scalar  1
key_integer_vector  1_2_3

key_double_scalar    1.0
key_double_vector    1.0_2.0_3.0
\end{lstlisting}
Values which are strings should contain no white space. Scalar parameters
may be integer values, or floating point values with a decimal point
(scientific notation is also allowed).  Vector parameters are introduced
by a set of three values (to be interpreted as $x,y,z$ components of the
vector in Cartesian coordinates) separated by an underscore. The identity
of the key will specify what type of value is expected. Keys and (string)
values are case sensitive.

Most keys have an associated default value which will be used if
the key is not present. Some keys must be specified: an error will
occur if they are missing. The remainder of this part
of the guide details the various choices for key value pairs,
along with any default values, and any relevant constraints.

\subsection{The Free Energy}
\label{input-free-energy}

The choice of free energy is determined as follows:
\begin{lstlisting}
free_energy   none
\end{lstlisting}
The default value is \texttt{none}, i.e., a simple Newtonian fluid is used.
Possible values of the \texttt{free\_energy} key are:
\begin{lstlisting}
#  none                     Newtonian fluid [DEFAULT]
#  symmetric                Symmetric binary fluid (finite difference)
#  symmetric_lb             Symmetric binary fluid (two distributions)
#  brazovskii               Brazovskii smectics
#  surfactant               Surfactants
#  polar_active             Polar active gels
#  lc_blue_phase            Liquid crystal (nematics, cholesterics, BPs)
#  lc_droplet               Liquid crystal emulsions
#  fe_electro               Single fluid electrokinetics
#  fe_electro_symmetric     Binary fluid electrokinetics
\end{lstlisting}

The choice of free energy will control automatically a number of factors
related to choice of order parameter, the degree of parallel communication
required, and so on. Each free energy has a number of associated parameters
discussed in the following sections.

Details of general (Newtonian) fluid parameters, such as viscosity,
are discussed in Section~\ref{input-fluid-parameters}.

\subsubsection{Symmetric Binary Fluids}

We recall that the free energy density is, as a function of compositional
order $\phi$:
\[
{\textstyle \frac{1}{2}} A\phi^2 +
{\textstyle \frac{1}{4}} B\phi^4 +
{\textstyle \frac{1}{2}} \kappa (\mathbf{\nabla}\phi)^2.
\]

Parameters are introduced by (with default values):
\begin{lstlisting}
free_energy  symmetric
A            -0.0625                         # Default: -0.003125
B            +0.0625                         # Default: +0.003125
K            +0.04                           # Default: +0.002
\end{lstlisting}
Common usage has $A < 0$ and $B = -A$ so that $\phi^\star = \pm 1$.
The parameter $\kappa$
(key \texttt{K}) controls
the interfacial energy penalty and is usually positive.

\subsubsection{Brazovskii smectics}
\label{input-brazovskki-smectics}
The free energy density is:
\[
{\textstyle \frac{1}{2}} A\phi^2 +
{\textstyle \frac{1}{4}} B\phi^4 +
{\textstyle \frac{1}{2}} \kappa (\mathbf{\nabla}\phi)^2 +
{\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 
\]

Parameters are introduced via the keys:
\begin{lstlisting}
free_energy  brazovskii
A             -0.0005                        # Default: 0.0
B             +0.0005                        # Default: 0.0
K             -0.0006                        # Default: 0.0
C             +0.00076                       # Default: 0.0
\end{lstlisting}
For $A<0$, phase separation occurs with a result depending on $\kappa$:
one gets two symmetric phases for $\kappa >0$ (cf.\ the symmetric case)
or a lamellar phase for $\kappa < 0$. Typically, $B = -A$ and the
parameter in the highest derivative $C > 0$.

\subsubsection{Surfactants}
\label{input-surfactants}

The surfactant free energy should not be used at the present time.

\subsubsection{Polar active gels}
\label{input-polar-active-gels}

The free energy density is a function of vector order parameter $P_\alpha$:
\[
{\textstyle \frac{1}{2}} A P_\alpha P_\alpha +
{\textstyle \frac{1}{4}} B (P_\alpha P_\alpha)^2 +
{\textstyle \frac{1}{2}} \kappa (\partial_\alpha P_\beta)^2
\]

There are no default parameters:
\begin{lstlisting}
free_energy        polar_active
polar_active_a    -0.1                       # Default: 0.0
polar_active_b    +0.1                       # Default: 0.0
polar_active_k     0.01                      # Default: 0.0
\end{lstlisting}
It is usual to choose $B > 0$, in which case $A > 0$ gives
an isotropic phase, whereas $A < 0$ gives a polar nematic phase.
The elastic constant $\kappa$ (key \texttt{polar\_active\_k})
is positive.

\subsubsection{Liquid crystal}
\label{input-liquid-crystal}
The free energy density is a function of tensor order parameter
$Q_{\alpha\beta}$:
\begin{eqnarray}
{\textstyle\frac{1}{2}} A_0 (1 - \gamma/3)Q^2_{\alpha\beta} -
{\textstyle\frac{1}{3}} A_0 \gamma
                        Q_{\alpha\beta}Q_{\beta\delta}Q_{\delta\alpha} +
{\textstyle\frac{1}{4}} A_0 \gamma (Q^2_{\alpha\beta})^2
\nonumber \\
+ {\textstyle\frac{1}{2}} \Big(
\kappa_0 (\epsilon_{\alpha\delta\sigma} \partial_\delta Q_{\sigma\beta} +
2q_0 Q_{\alpha\beta})^2 + \kappa_1(\partial_\alpha Q_{\alpha\beta})^2 \Big)
\nonumber
\end{eqnarray}

The corresponding \texttt{free\_energy} value, despite its name, is
suitable for nematics and cholesterics, and not just blue phases:
\begin{lstlisting}
free_energy      lc_blue_phase
lc_a0            0.01                       # Deafult: 0.0
lc_gamma         3.0                        # Default: 0.0
lc_q0            0.19635                    # Default: 0.0
lc_kappa0        0.00648456                 # Default: 0.0
lc_kappa1        0.00648456                 # Default: 0.0
\end{lstlisting}
The bulk free energy parameter $A_0$ is positive and controls the
energy scale (key \texttt{lc\_a0}); $\gamma$ is positive and
influences the position in the phase diagram relative to the
isotropic/nematic transition (key \texttt{lc\_gamma}).
The two elastic constants must be equal, i.e., we enforce the
single elastic constant approximation (boths keys \texttt{lc\_kappa0} and
\texttt{lc\_kappa1} must be specified).

Other important parameters in the liquid crystal picture are:
\begin{lstlisting}
lc_xi            0.7                         # Default: 0.0
lc_Gamma         0.5                         # Default: 0.0
lc_active_zeta   0.0                         # Default: 0.0
\end{lstlisting}
The first is $\xi$ (key \texttt{lc\_xi}) is the effective molecular
aspect ratio and should be in the range $0< \xi< 1$. The rotational
diffusion constant is $\Gamma$ (key \texttt{lc\_Gamma}; not to be
confused with \texttt{lc\_gamma}). The (optional) apolar activity
parameter is $\zeta$ (key \texttt{lc\_active\_zeta}).

\subsubsection{Liquid crystal emulsion}
\label{input-liquid-crystal-emulsion}

This an interaction free energy which combines the symmetric and liquid
crystal free energies. The liquid crystal free energy constant $\gamma$
becomes a
function of composition via $\gamma(\phi) = \gamma_0 + \delta(1 + \phi)$,
and a coupling term is added to the free energy density:
\[
WQ_{\alpha\beta} \partial_\alpha \phi \partial_\beta \phi.
\]
Typically, we might choose $\gamma_0$ and $\delta$ so that
$\gamma(-\phi^\star) < 2.7$ and the $-\phi^\star$ phase is isotropic,
while $\gamma(+\phi^\star) > 2.7$ and the
$+\phi^\star$ phase is ordered (nematic, cholesteric, or blue phase).
Experience suggests that a suitable choice is $\gamma_0 = 2.5$ and
$\delta = 0.25$.

For anchoring constant $W > 0$, the liquid crystal anchoring at the
interface is planar, while for $W < 0$ the anchoring is normal. This
is set via key \texttt{lc\_droplet\_W}.

Relevant keys (with default values) are:
\begin{lstlisting}
free_energy            lc_droplet

A                      -0.0625
B                      +0.0625
K                      +0.053

lc_a0                   0.1
lc_q0                   0.19635
lc_kappa0               0.007
lc_kappa1               0.007

lc_droplet_gamma        2.586                # Default: 0.0
lc_droplet_delta        0.25                 # Default: 0.0
lc_droplet_W           -0.05                 # Default: 0.0
\end{lstlisting}
Note that key \texttt{lc\_gamma} is not set in this case.

\subsection{System Parameters}
\label{input-system-parameters}

Basic parameters controlling the number of time steps
and the system size are:
\begin{lstlisting}
N_start      0                              # Default: 0
N_cycles     100                            # Default: 0
size         128_128_1                      # Default: 64_64_64
\end{lstlisting}
A typical simulation will start from time zero (key \texttt{N\_start})
and run for a certain number of time steps (key \texttt{N\_cycles}).
The system size (key \texttt{size}) specifies the total number of
lattice sites in each dimension. If a two-dimensional system is
required, the extent in the $z$-direction must be set to unity, as
in the above example.

If a restart from a previous run is required, the choice of parameters
may be as follows:
\begin{lstlisting}
N_start      100
N_cycles     400
\end{lstlisting}
This will restart from data previously saved at time step 100, and
run a further 400 cycles, i.e., to time step 500.
% TODO Cross-reference to I/O.

\subsubsection{Parallel decomposition}

In parallel, the domain decompostion is closely related to the
system size, and is specified as follows:
\begin{lstlisting}
size         64_64_64
grid         4_2_1
\end{lstlisting}
The \texttt{grid} key specifies the number of MPI tasks required in
each coordinate direction. In the above example, the decomposition
is into 4 in the $x$-direction, into 2 in the $y$-direction, while
the $z$-direction is not decomposed. In this example, the local domain
size per MPI
task would then be $16\times32\times64$. The total number of MPI tasks
available must match the total implied by \texttt{grid} (8 in the
example).

The \texttt{grid} specifications must exactly divide the system size;
if no decomposition is possible, the code will terminate with an error
message.
If the requested decomposition is not valid, or \texttt{grid} is
omitted, the code will try to supply a decomposition based on
the number of MPI tasks available and \texttt{MPI\_Dims\_create()};
this may be implementation dependent.

\subsection{Fluid Parameters}
\label{input-fluid-parameters}

Control parameters for a Newtonian fluid include:
\begin{lstlisting}
fluid_rho0                 1.0
viscosity                  0.166666666666666
viscosity_bulk             0.166666666666666
isothermal_fluctuations    off
temperature                0.0
\end{lstlisting}
The mean fluid density is $\rho_0$ (key \texttt{fluid\_rho0}) which
defaults to unity in lattice units; it is not usually necessary to
change this. The shear viscosity is
\texttt{viscosity} and as default value 1/6 to correspond to
unit relaxation time in the lattice Boltzmann picture. Reasonable
values of the shear viscosity are $0.2 > \eta > 0.0001$ in lattice
units. Higher values move further into the over-relaxation region, and can
result in poor behaviour. Lower
values increase the Reynolds number and tend to cause
problems with stability. The bulk
viscosity has a default value which is equal to whatever shear
viscosity has been selected. Higher values of the bulk viscosity
may be set independently and can help to suppress large deviations
from incompressibility and maintain numerical stability
in certain situations.

If fluctuating hydrodynamics is wanted, set the value of
 \texttt{isothermal\_fluctualtions} to \texttt{on}. The associated
temperature is in lattice units: reasonable values (at $\rho_0 = 1$)
are $0 < kT < 0.0001$. If the temperature is too high, local
velocities will rapidly exceed the Mach number constraint and
the simulation will be unstable.

\vfill
\pagebreak

### 2016-03-03 trunk/docs/lb.tex
Updates to trunk/docs
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2870 40876441-3496-0410-a613-a2fca35b2ae2

@@
 discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic quatities to the distribution
+It is possible to relate the hydrodynamic quantities to the distribution
 function via its moments, that is
@@
 the collision matrix, with associated weights and normalisers are shown
-in Table~\ref{table-d2q9-spec}.
+in Table~\ref{table-d2q9-spec}. In two dimensions there are six hydrodynamic
+modes and a total of three kinetic modes, or ghost modes.
 
@@
 \hline\hline
-$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
+$M^a$ & $p$ & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
 \hline
@@
 \end{center}
-\caption{TABLE CAPTION}
+\caption{Table showing the details of the basis used for the D2Q9 model
+in two dimensions. The nine modes $M^a$ include six hydrodynamic modes,
+one scalar kinetic mode $\chi^1$, and one vector kinetic mode $J_{i\alpha}$.
+The weights in the equilibrium distribution function are $w_i$ and the
+normaliser for each mode is $N^a$. The eigenvectors of the collision
+matrix are the columns of the transformation matrix $m^a_i$. The prefactor
+$p$ (where present) multiplies all the elements to the right in that row.}
 \label{table-d2q9-spec}
@@
 
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsection{Hydrodynamic Boundary Conditions}
-
-\subsubsection{Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Fixed objects}
-
-\subsubsection{Moving objects}
-
 \subsubsection{D3Q15}
 
-These are the eigenvalues and eigenvectors of the collision
-matrix used for D3Q15
+The D3Q15 model in three dimensions consists of a set of vectors:
+one zero vector $(0,0,0)$, six vectors of length unity being
+$(\pm 1, 0, 0)$ cyclically permuted, and 8 vectors of length
+$\sqrt{3}$ being $(\pm 1, \pm 1, \pm 1)$.
+The eigenvalues and eigenvectors of the collision
+matrix used for D3Q15 are given in Table~\ref{table-d3q15-spec}.
 
-\begin{table}
+\begin{table}[t]
 \centering
@@
 \hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
+$M^a$ & $p$ & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
 \hline
@@
 for each mode is $N^a$. The eigenvectors of the collision matrix are the
-colms of the transformation matrix $m_i^a$. The prefactor $p$ simply
-multiplies all elements of $m_i^a$ in that row as a convenience.}
+columns of the transformation matrix $m_i^a$. The prefactor $p$ simply
+multiplies all elements of $m_i^a$ in that row as a convenience.
+\label{table-d3q15-spec}
+}
 \end{table}
@@
 
-\begin{table}
+The D3Q19 model in three dimensions is constructed with velocities:
+one zero vector $(0,0,0)$, three vectors of length unity being
+$(\pm 1, 0, 0)$ cyclically permuted, and twelve vectors of length
+$\sqrt{2}$ being $(\pm 1, \pm 1, 0)$ cyclically permuted. The
+details of the D3Q19 model are set out in Table~\ref{table-d3q19-spec}.
+
+\begin{table}[t]
 \centering
@@
 \caption{Table showing the details of the basis used for the D3Q19 model in
-three dimensions. }
+three dimensions. The nineteen modes $M^a$ include ten hydrodynamic modes,
+three scalr kinetic modes $\chi^1$, $\chi^2$, and $\chi^3$; there are also
+two vector kinetic modes $\chi^1 \rho c_{i\alpha}$
+and $\chi^2 \rho c_{i\alpha}$. The weights in the equilibrium distribution
+function are $w_i$, and the normaliser for each mode is $N^a$. The
+eigenvectors of the collision matrix are the columns of the transformation
+matrix $m^a_i$.
+\label{table-d3q19-spec}
+}
 \end{table}
 
+\subsection{Fluctuating LBE}
+
+It is possible \cite{adhikari2005} to simulate fluctuating
+hydrodynamics for an isothermal fluid via the inclusion of
+a fluctuating stress $\sigma_{\alpha\beta}$:
+\begin{equation}
+\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
++ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
+\end{equation}
+The fluctuation-dissipation theorem relates the magnitude of this
+random stress to the isothermal temperature and the viscosity.
+
+In the LBE, this translates to the addition of a random contribution
+$\xi_i$ to the distribution at the collision stage, so that
+\begin{equation}
+\ldots + \xi_i.
+\end{equation}
+
+For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
+i.e., those with dissipation, the fluctuating part may be written
+\begin{equation}
+\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
+\end{equation}
+where $\hat{\xi}^a$ is a noise termwhich has a variance determined
+by the relaxation time for given mode
+\begin{equation}
+\left< \hat{\xi}^a \hat{\xi}^b \right> =
+\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
+\left< \delta M^a \delta M^b \right>.
+\label{eq_fvar}
+\end{equation}
+
+\subsubsection{Fluctuating stress}
+
+For the stress, the random contribution to the distributions is
+\begin{equation}
+\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
+\end{equation}
+where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
+variates drawn from a Gaussian distribution with variance given
+by equation~\ref{eq_fvar}. In the case that the shear and bulk
+viscosities are the same, i.e., there is a single relaxation
+time, then the variances of the six independent components of
+the matrix are given by
+\begin{equation}
+\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
+\frac{2\tau + 1}{\tau^2}
+(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
+\end{equation}
+
+\subsection{Hydrodynamic Boundary Conditions}
+
+\subsubsection{Bounce-Back on Links}
+
+A very general method for the representation of solid objects
+within the LB approach was put forward by Ladd \cite{l94a, l94b}.
+Solid objects (of any shape) are defined by a boundary surface
+which intersects some of the velocity vectors $\mathbf{c}_i$
+joining lattice nodes. Sites inside are designated solid, while
+sites outside remain fluid. The correct boundary condition is
+defined by identifying \textit{links} between fluid and solid
+sites, which allows those elements of the distribution which would
+cross the boundary at the propagation step to be ``bounced-back''
+into the fluid. This bounce-back on links is an efficient method
+to obtain the  correct hydrodynamic interaction between solid
+and fluid.
+
+\subsubsection{Fixed objects}
+
+\subsubsection{Moving objects}
+
+\vfill
 \pagebreak

### 2016-06-06 branches/Anthony-nonblocking/docs/intro.tex
non-blocking halo exchange version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2897 40876441-3496-0410-a613-a2fca35b2ae2

--- 2014-11-14 branches/kevin/docs/intro.tex
@@
 %
-%  (c) 2014 The University of Edinburgh
+%  (c) 2016 The University of Edinburgh
 %
@@
 
+If you look in  the \texttt{trunk} directory, you should see
+\begin{lstlisting}
+$ ls trunk
+config  LICENSE      mpi_s   src       tests  version.h
+docs    Makefile.mk  README  targetDP  util
+\end{lstlisting}
+
 \vfill
@@
 
-\subsection{Compilation}
+\subsection{Configuration}
 
-Identify your local C compiler. The following example uses \texttt{gcc}
+You will need to copy one of the existing configuration files from
+the \texttt{config} directory to the top level directory. This file
+sets all the relevant local details for compilation and so on. For
+example:
+\begin{lstlisting}
+$ cp config/lunix-gcc-default.mk config.mk
+\end{lstlisting}
+Edit the new \texttt{config.mk} file to check, or adjust, the options
+as required. You will need to identify your local C compiler. The
+example uses \texttt{gcc}
 in serial, and \texttt{mpicc} in parallel. For MPI in particular,
-local details may vary.
+local details may vary. A quick way to compile the code is via
+\begin{lstlisting}
+$ cd tests
+$ make compile-mpi-d3q19
+\end{lstlisting}
+This integrates a number of separate stages which are described
+indivually in the following sections. The resulting executable
+will be \texttt{./src/Ludwig.exe} at the top level.
+
+\subsubsection{targetDP}
+
+A \texttt{targetDP} layer is required by the code in all cases. It
+allows compilation for both simple CPU systems and
+for a number of accelerator systems using a single source code. This
+will make the appropriate choices depending what has been specified
+in the top-level \texttt{config.mk} file. From the top-level directory
+\begin{lstlisting}
+$ cd targetDP
+$ make
+\end{lstlisting}
+should produce the appropriate \texttt{libtarget.a} library.
 
@@
 First, compile the MPI stub library in the \texttt{mpi\_s}
-directory. Do this by editing the Makefile, and checking the compiler
-is appropriate. You should be able to build the library and run the
-tests using, e.g.,:
+directory. You should be able to build the library and run the
+tests using, e.g.:
 
@@
 
-Now compile the main code in the \texttt{src} directory. Again
-edit the Makefile to check the compiler. Compilation should
-look like:
+Now compile the main code in the \texttt{src} directory.
+Compilation should look like:
 

### 2016-06-06 branches/Anthony-nonblocking/targetDP/doc/introduction.tex
non-blocking halo exchange version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@2897 40876441-3496-0410-a613-a2fca35b2ae2

--- 2016-01-14 targetDP/doc/introduction.tex
@@
 %\end{comment}
-\section{Abstract}
 
-The targetDP API provides an abstraction layer which allows
-applications to \underline{t}arget \underline{D}ata
-\underline{P}arallel hardware in a platform agnostic manner, by
-abstracting the memory spaces and hierarchy of hardware
-parallelism. Applications written using targetDP syntax are
-performance portable: the same source code can be compiled for
-different targets (where we currently support GPU accelerators and
-modern multicore or manycore SIMD CPUs), without performance
-overheads. The model is appropriate for abstracting the parallelism
-contained within each compute node, and can be combined with, e.g. MPI
-to allow use on systems containing multiple nodes. The targetDP API is
-primarily aimed at the types of parallelism found in grid-based
-applications, but may be applicable to a wider class of problems.
-This document introduces the targetDP memory and execution models, and
-specifies the syntax and functionality of the
-interface. Implementation details are also provided, and a simple
-example is given to demonstrate usage.
-
-\section{Motivation}
 It is becoming increasingly difficult for applications to exploit
@@
 
-\section{Glossary}
+The targetDP API (first introduced in \cite{hpcc}) provides an
+abstraction layer which allows applications to \underline{t}arget
+\underline{D}ata \underline{P}arallel hardware in a platform agnostic
+manner, by abstracting the memory spaces and hierarchy of hardware
+parallelism. Applications written using targetDP syntax are
+performance portable: the same source code can be compiled for
+different targets (where we currently support GPU accelerators and
+modern multicore or manycore SIMD CPUs), without performance
+overheads. The model is appropriate for abstracting the parallelism
+contained within each compute node, and can be combined with, e.g. MPI
+to allow use on systems containing multiple nodes. The targetDP API is
+primarily aimed at the types of parallelism found in grid-based
+applications, but may be applicable to a wider class of problems.
+
+The targetDP memory and execution models are described in Chapters \ref{sec:exmodel} and \ref{sec:memmodel} respectively, and the document goers on to specify the memory management functionality in Chapter \ref{chap:memmanage} and data parallel execution functionality in Chapter  \ref{sec:dpexecution}.
+Implementation details are also provided for the existing C and CUDA versions of targetDP throughout these chapters.
+Finally, a simple
+example is given to demonstrate usage in Chapter \ref{chapter:examples}.
+
+\section*{Glossary}
 

### 2016-06-06 branches/Anthony-nonblocking/src/hydro.c

@@ -1,0 +1,823 @@
+/*****************************************************************************
+ *
+ *  hydro.c
+ *
+ *  Hydrodynamic quantities: velocity, body force on fluid.
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  (c) 2012-2015 The University of Edinburgh
+ *  Contributing authors:
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  Alan Gray (alang@epcc.ed.ac.uk)
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <math.h>
+#include <stdlib.h> 
+
+#include "pe.h"
+#include "coords.h"
+#include "coords_field.h"
+#include "leesedwards.h"
+#include "io_harness.h"
+#include "util.h"
+#include "control.h" /* Can we move this into LE please */
+#include "hydro_s.h"
+#include "targetDP.h"
+
+static int hydro_lees_edwards_parallel(hydro_t * obj);
+static int hydro_u_write(FILE * fp, int index, void * self);
+static int hydro_u_write_ascii(FILE * fp, int index, void * self);
+static int hydro_u_read(FILE * fp, int index, void * self);
+
+/*****************************************************************************
+ *
+ *  hydro_create
+ *
+ *  We typically require a halo region for the velocity which is only
+ *  one lattice site in width, i.e., nhcomm = 1. This is independent
+ *  of the width of the halo region specified for coords object.
+ *
+ *****************************************************************************/
+
+int hydro_create(int nhcomm, hydro_t ** pobj) {
+
+  int nsites;
+  double * tmpptr;
+  hydro_t * obj = (hydro_t *) NULL;
+
+  assert(pobj);
+
+  obj = (hydro_t*) calloc(1, sizeof(hydro_t));
+  if (obj == NULL) fatal("calloc(hydro) failed\n");
+
+  obj->nf = 3; /* always for velocity, force */
+  obj->nhcomm = nhcomm;
+
+  nsites = le_nsites();
+  obj->u = (double*) calloc(obj->nf*nsites, sizeof(double));
+  if (obj->u == NULL) fatal("calloc(hydro->u) failed\n");
+
[…]

### 2016-06-06 branches/Anthony-nonblocking/src/polar_active_rt.c

@@ -1,0 +1,216 @@
+/*****************************************************************************
+ *
+ *  polar_active_rt.c
+ *
+ *  Run time initialisation for active gel free energy.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2011 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <float.h>
+#include <math.h>
+#include <stdio.h>
+#include <string.h>
+
+#include "pe.h"
+#include "runtime.h"
+#include "coords.h"
+#include "field.h"
+#include "field_grad.h"
+#include "io_harness.h"
+#include "free_energy_vector.h"
+#include "polar_active.h"
+#include "polar_active_rt.h"
+
+static int polar_active_init_code(field_t * p);
+
+/*****************************************************************************
+ *
+ *  polar_active_run_time
+ *
+ *  Sort out the active gel input parameters.
+ *
+ *****************************************************************************/
+
+void polar_active_run_time(void) {
+
+  double a;
+  double b;
+  double k1;
+  double delta;
+  double klc;
+  double zeta;
+  double lambda;
+
+  info("Polar active free energy selected.\n");
+
+  /* PARAMETERS */
+
+  RUN_get_double_parameter("polar_active_a", &a);
+  RUN_get_double_parameter("polar_active_b", &b);
+  RUN_get_double_parameter("polar_active_k", &k1);
+  RUN_get_double_parameter("polar_active_dk", &delta);
+  delta = 0.0; /* Pending molecular field */
+  RUN_get_double_parameter("polar_active_klc", &klc);
+  RUN_get_double_parameter("polar_active_zeta", &zeta);
+  RUN_get_double_parameter("polar_active_lambda", &lambda);
+
+  info("\n");
+
+  info("Parameters:\n");
+  info("Quadratic term a     = %14.7e\n", a);
+  info("Quartic term b       = %14.7e\n", b);
+  info("Elastic constant k   = %14.7e\n", k1);
[…]

### 2016-06-06 branches/Anthony-nonblocking/util/extract_colloids.c

@@ -1,0 +1,232 @@
+/*****************************************************************************
+ *
+ *  extract_colloids.c
+ *
+ *  Convert an output file to a csv file suitable for Paraview.
+ *  The csv file uses three extra particles at (xmax, 0, 0)
+ *  (0, ymax, 0) and (0, 0, zmax) to define the extent of the
+ *  system.
+ *
+ *  If you want a different set of colloid properties, you need
+ *  to arrange the header, and the output appropriately. The read
+ *  can be ascii or binary and is set by the switch below.
+ *
+ *  Compile against the ludwig library, e.g.,
+ *  $ cd ../targetDP
+ *  $ make
+ *  $ cd ../src
+ *  $ make serial
+ *  $ cd ../util
+ *  $ $(CC) -I../mpi_s -I../src -I../targetDP extract_colloids.c \
+ *          -L../mpi_s -lmpi -L../src -lludwig -lm
+ *
+ *  $ ./a.out <colloid file name> <nfile> <csv file name>
+ *
+ *  where the first argument is the file name stub
+ *        the second is the number of parallel files
+ *        and the third is the (single) ouput file name
+ *
+ *  If you have a set of files, try (eg. here with 4 parallel output files),
+ *
+ *  $ for f in config.cds*004-001; do g=`echo $f | sed s/.004-001//`; \
+ *  echo $g; ~/ludwig/trunk/util/extract_colloids $g 4 $g.csv; done
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2012 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <stdio.h>
+#include <stdlib.h>
+
+#include "colloid.h"
+
+#define NX 256
+#define NY 256
+#define NZ 256
+
+static const int    iread_ascii = 0;      /* Read ascii or binary */
+
+static const char * format3_    = "%10.5f, %10.5f, %10.5f,";
[…]

### 2016-06-07 targetDP/doc/examples.tex

--- 2016-01-12 branches/kevin-1397/targetDP/doc/examples.tex
@@
 \begin{lstlisting}
+
+//declare constant variable
+__targetConst__ double t_a; (*@\label{example:scalekernel:targetConst}@*)
+
 __targetEntry__ void scale(double* t_field) { (*@\label{example:scalekernel:targetEntry}@*)

### 2016-11-04 docs/binary.tex
trunk 4.9 updates to tests to reflect improved spelling cf svn 2976
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2988 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  binary.tex
%
%  Symmetric binary fluids and Brazovskii Smetics
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2016 The University of Edinburgh
%
%  Contributing authors:
%  Kevin Stratford (kevin@epcc.ed.ac.uk) and I've taken material from
%  Grace Kim's                           doctoral thesis (UoE 2009)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Binary Fluids}

In this section we discuss symmetric two-component fluids, and the closely
related subject of Brazovskii smectics.\ \textit{Ludwig} was first developed
to study issues such as spinodal decomposition in binary mixtures,
so we will introduce the background in that
context (see, e.g., Swift \textit{et al.} \cite{swift1996} and Kendon
\textit{et al.}, \cite{kendon2001} for pioneering work).

Early versions of \textit{Ludwig} used, exclusively, a two lattice
Boltzmann distribution scheme to introduce the relevant dynamical update
for the binary fluid \cite{desplat2001}.
While this approach is retained in the code
for historical reasons it is often advantageous, for a number of reasons
which are set out in the following, to use a coupled lattice
Boltzmann finite-difference approach. We describe both approaches in what
follows, but reccmmend using the finite-difference approach for all current
applications. First, we consider
the common theoretical background for the two-component binary fluid.

\subsection{The Symmetric Free Energy}

\subsubsection{Thermodynamics}

For a symmetric fluid of fixed content, we define a single coarse-grained
composition
variable, or order parameter, $\phi(\mathbf{r})$. At a finer level,
one can interpret the order parameter as
$\phi(\mathbf{r}) = (n_1 - n_2)/(n_1 + n_2)$
where the $n$ are number densities of the two components. In either case,
it is an important
feature of these systems that the order parameter is conserved so that
\begin{equation}
\int \mathrm{d}\mathbf{r} (\phi - \phi_0) = 0
\end{equation}
at all times, where $\phi_0$ is the mean order parameter. In general,
if the order parameter values associated with the two components are
$\phi_1$ and $\phi_2$, then the volume fraction of the first component is
$v_1 = (\phi_2 - \phi_0)/(\phi_2 - \phi_1)$, and $v_2 = 1 - v_1$.

For a uniform
state at constant volume and temperature we can write the density of the
(Helmholtz) free energy as $V(\phi)$. In mean field theory, it is predicted
that $V(\phi)$ has uniform positive curvature at high temperature, but
becomes concave below a critical temperature $T_c$. This is illustrated
in the left panel of Fig.~\ref{figure-symmetric-schematic}.
For temperatures larger than $T_c$, there is a homogeneous state with
$\phi = 0$ everywhere, i.e., the state is fully mixed. For temperatures
below $T_c$, the free energy is minimised by creating two separate
domains, each composed of a single constituent. This results in a
phase diagram of the sort 
shown in Fig.~1 (right panel) with mixed and
separated phases bounded by a co-existance curve.

In an initially mixed system spinodal decomposition occurs when a
temperature change leaves the system below the spinodal line in the
phase diagram. This leaves the system in a globally unstable state,
and the response is initially small changes in composition
everywhere which grow smoothly in time. These small changes
ultimately create interfaces between fluid domains;
minimisation of energy then translates to minimisation of interfacial
curvature which drives the fluid domains to grow. If the quench leaves
the system between the coexistance line and the spinoal line, the
system is in a metastable state, and the response is local changes
in composition which nucleate droplets.

Note that in some real fluid mixtures, the phase diagram is inverted,
and a quench into the coexistance region is upwards in temperature.
See, e.g., Chaikin and Lubensky \cite{chaikin-lubensky} for an extended
discusion of spinodal decomposition.

Quantitatively, fluid domains together with interfaces are described
by a free energy functional with two contributions. An appropriate
choice is the Ginzburg-Landau functional, sometimes referred to as the
squared gradient model:
\begin{equation}
F[\phi]  = \int \mathrm{d}\mathbf{r} [ V(\phi) + {\textstyle\frac{1}{2}}
\kappa (\partial_\alpha \phi)^2 ] 
\end{equation}
where the $V(\phi)$ is as before and the second term penalises gradients
in the composition with parameter $\kappa > 0$. The explicit form of
$V(\phi)$ is taken here from a general Landau expansion:
\begin{equation}
V(\phi) = {\textstyle\frac{1}{2}} A \phi^2
        + {\textstyle\frac{1}{4}} B \phi^4.
\end{equation}
The minima in the
potential (cf.\ Fig.~\ref{figure-symmetric-schematic}) can then be
identified by setting
$dV / d \phi = 0$, that is, $A\phi + B\phi^3 = \phi (A + B\phi^2) = 0$.
If we take $B>0$, this condition has two cases: we denote the
solutions $\phi^\star$ in both. The first is the mixed case
which has $A > 0$ so that we
must have $\phi^\star = 0$, and the second is the separated case
which has $A < 0$ and so 
$\phi^\star = \pm (-A/B)^{1/2}$.

The combination of parameters $A,B$ and $\kappa$ also determine the
interfacial tension and the interfacial width.
The stationary values of the free energy functional can be obtained
from the Euler-Lagrange equation for $\phi$. The relevant case is
where we have a functional the density of which is written
$f(\phi, \partial_\alpha \phi)$ and the corresponding Euler-Lagrange
equation is
\begin{equation}
\frac{\partial f}{\partial \phi}
- \partial_\alpha \frac{\partial f}{\partial\, \partial_\alpha \phi} = 0.
\end{equation}

\begin{figure}[t]

\includegraphics[width=0.9\textwidth]{figures/symmetric.eps}

\caption{Two panels showing (left) the form of the free energy density
$V(\phi)$ for the binary fluid for a number of different
temperatures. At temperatures above
the critical temperature, there is a single minimum (top curve), while
below $T_c$, two symmetric minima are present (lower curves). The related
phase diagram (right) in the temperature-composition plane shows mixed
and separated phases the boundary of which is the coexistence 
line. This (sometimes referred to as the binodal line) is the locus of
points which are the minima of $V(\phi)$ at different temperatures. The
spinodal line is formed by the locus of points which are the inflections
of $V(\phi)$.}
\label{figure-symmetric-schematic}
\end{figure}

If we consider a one-dimensional case, this is
\begin{equation}
A\phi + B\phi^3 - \kappa \frac{d^2 \phi}{dx^2} = 0.
\end{equation}
Note that the conservation of order parameter means this should be a
constrained minimisation, which can be handled by adding the
a Lagrange multiplier $\Lambda = \partial f / \partial \phi |_{\phi_0}$.
The resulting ordinary differential equation has a solution
$\phi(x) = \phi^\star \tanh(x/\xi)$ where $\xi$ is identified
as the interfacial width. It can be confirmed that this is a solution
if
\begin{equation}
\xi = (-2\kappa / A)^{1/2}.
\end{equation}
The one-dimensional case can also be used to determine the interfacial
tension, which is the excess energy associated with the interface.
If we denote the interfacial tension $\sigma$, then the excess
energy is
\begin{equation}
\sigma = \int_{-\infty}^{+\infty} \mathrm{d} x
\Big[
{\textstyle \frac{1}{2}} \kappa (d_x \phi)^2 + V(\phi) - V(\phi^\star)
\Big],
\end{equation}
which can be evaluated from the equilibrium profile with the interface
placed at the origin $\phi(x) = \phi^\star\tanh(x/\xi)$, along with a number
of standard integrals, to give
\begin{equation}
\sigma =  4\kappa\phi^{\star 2}/3\xi
 = \big(-8\kappa A^3/9B^2\big)^{1/2}.
\end{equation}
Further discussion of the practical choice of the free energy parameters,
and their
impact on interfacial width and tension is deferred until LATER.

\subsubsection{Dynamics}

In a purely diffusive regime, the time evolution of a conserved
order parameter may be expressed as the local divergence of a flux
$j_\alpha(\mathbf{r})$, that is,
$\partial_t \phi(\mathbf{r}) + \partial_\alpha j_\alpha(\mathbf{r}) = 0$.
In this picture, the system responds to small local changes in composition
by diffusion, which acts to reduce the energy. Formally, this
response of the free energy to a small change in composition is described
by the chemical potential
\begin{equation}
\mu = \frac{\delta F}{\delta \phi} = \frac{dV}{d\phi} - \kappa \partial_\alpha^2 \phi,
\end{equation}
being the functional derivative of the free energy with respect to the
order parameter. The dynamics is usually described in the context of the
model of Cahn and Hilliard \cite{cahn-hilliard-1958}, otherwise known as
Model B, where
\begin{equation}
\frac{\partial \phi}{\partial t} =
M\partial_\alpha^2 \frac{\delta F}{\delta \phi}.
\label{eq-cahn-hilliard}
\end{equation}
Here, the diffusive flux is related to the gradient of chemical potential
via $j_\alpha = -M \partial_\alpha \mu$, where $M$ is a mobilty
(taken to be uniform).

In the presence of a fluid with velocity field $u_\alpha(\mathbf{r})$,
the full equation for the time evolution of the order parameter is
\begin{equation}
\partial_t \phi + \partial_\alpha (\phi u_\alpha + M\partial_\alpha \mu) = 0.
\label{eq-cahn-hilliard-final}
\end{equation}
This form again stresses the nature of the dynamics as a conservation law,
involving the divergence of advective fluxes $\phi u_\alpha$ and diffusive
fluxes $j_\alpha = -M\partial_\alpha \mu$. It
also admits that the mobility may be a function of position, e.g., via
$M = M(\phi)$. Equation~\ref{eq-cahn-hilliard-final} will form the basis of
the numerical solution of the binary fluid problem, coupled to the
Navier-Stokes equation.

The presence of a interface, or non-uniform composition, can lead to
thermodynamic stresses on the fluid which we denote $P_{\alpha\beta}$,
in additional to the usual viscous
stresses. The divergence of this
additional stress represents a body force density in the Navier-Stokes
equation, which is related to the chemical potential via
$f_\alpha = -\phi\partial_\alpha \mu = -\partial_\beta P_{\alpha\beta}$.
The full form of the additional stress may be related to the order
parameter by
\begin{equation}
P_{\alpha\beta} = p_0 \delta_{\alpha\beta}
  + \kappa \partial_\alpha \phi \partial_\beta \phi
\end{equation}
with isotropic contribution
\begin{equation}
p_0 = \phi \frac{dV(\phi)}{d\phi} -  V(\phi)
-\kappa\phi\partial_\alpha^2 \phi - \textstyle{\frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
\end{equation}
Note that the stress $P_{\alpha\beta}$ is symmetric. At (local) equilibrium,
one expects gradients in the chemical potential to vanish,
and there will be neither diffusive fluxes ($j_\alpha = 0$) nor force on
the fluid ($f_\alpha = 0$).

\subsubsection{Surface (or wetting) free energy}

In the presence of a solid surface, the two component fluid gives
rise to the possiblity of a solid-fluid-fluid contact line. In the
symmetric case, the fluid-fluid interface will be at right angles to
a uniform flat surface. If one component wets the solid preferentially
(that is, it is energetically favourable for that component to be
in contact with the solid) the angle may move away from
90~degrees.
In general, the equilibrium contact angle $\theta$ is given by the
Young equation
\begin{equation}
\sigma_1 - \sigma_2 - \sigma \cos\theta = 0,
\end{equation}
where there are two solid-fluid surface tensions for components 1 and 2,
and the fluid-fluid interfacial tension is $\sigma$ as before.

This may be described in the free energy picture by adding a free
energy density per unit area assoicated with the surface:
\begin{equation}
f_s (\phi_s)= \textstyle{\frac{1}{2}} C \phi_s^2 + H \phi_s,
\end{equation}
where $\phi_s$ is the order parameter at the surface, and $C$ and $H$ are
constant parameters.

\subsection{Implementation}

\subsubsection{Lattice kinetic approach}

A second distribution function is introduced, denoted $g_i(\mathbf{r};t)$,
which describes the composition. The moments of this distribution, by
analogy with the hydrodynamic case with distribution
$f_i(\mathbf{r};t)$ given in
Eq.~\ref{lb-f-moments}, are written
\begin{equation}
\phi(\mathbf{r};t) = \sum_i g_i (\mathbf{r};t), \quad
j_\alpha(\mathbf{r};t) = \sum_i g_i(\mathbf{r};t) c_{i\alpha}, \quad
\Phi_{\alpha\beta}(\mathbf{r};t) = \sum_i g_i(\mathbf{r};t)
c_{i\alpha}c_{i\beta}.
\end{equation}
Again, the summation is over index $i$ which counts the number of
discrete velocity components for the relevant lattice Boltzmann
model $N_{\mathrm{vel}}$. However, for the Cahn-Hilliard equation,
the only relevant physical moments are the order parameter
$\phi(\mathbf{r};t)$ and the flux $j_\alpha(\mathbf{r};t)$, so
the quantity $\Phi_{\alpha\beta}(\mathbf{r};t)$ is counted among
the kinetic modes and has no precise physical interpretation. In
contrast to the previous section, here $j_\alpha(\mathbf{r};t)$
is used to represent the total flux (advective
plus diffusive) rather than the purely diffusive flux.

The lack of direct physical interpretation for the second moment
means there are a number of possible choices at the collision
stage for the $g_i(\mathbf{r};t)$ distribution.
\begin{enumerate}
\item
Following Stratford \textit{et al.} \cite{j-stat-phys-2005}:
relax $j_\alpha$ toward the equilibrium $\phi u_\alpha$
at a rate related to the mobility and fix $\Phi_{\alpha\beta} =
\phi u_\alpha u_\beta + \mu \delta_{\alpha\beta}$.
\item
Following Kendon \textit{et al.} \cite{kendon2001}: fix both
$j_\alpha = \phi$ and
$\Phi_{} = \phi u_\alpha u_\beta + M\mu \delta_{\alpha\beta}$ so the
mobilty enters explicitly with the chemical potential.
\end{enumerate}

In both cases a reprojection is used to recover the post-collision
distributions $g_i^\star(\mathbf{r};t)$. This explicitly eliminates
kinetic modes other than $\Phi_{\alpha\beta}$. Hence
\begin{equation}
g_i = w_i \big(\phi + \phi u_\alpha c_{i\alpha}/c_s^2 +
(\mu\delta_{\alpha\beta} + \phi u_\alpha u_\beta) Q_{i\alpha\beta}/2c_s^4\big).
\end{equation}
In practice, this is not stable, and we use
\begin{equation}
g_i = \phi\delta_{i0} + w_i \big( j_\alpha c_{i\alpha} / c_s^2
+ \Phi_{\alpha\beta}Q_{i\alpha\beta} / 2c_s^4\big)
\end{equation}
where the $\delta_{i0}$ has the effect of moving most of the order
parameter into the non-propagating rest distribution $g_0$. 
This reprojection approach has the advantage that it eliminates several
spurious anisotropic terms which are introduced by a single relaxation
time approach (used by Kendon \textit{et al.}).

In the relaxation approach, the form of the relaxation for the
three components of the order parameter flux is
\begin{equation}
j_\alpha^\star =
j_\alpha - \tau_\phi^{-1} (j_\alpha - \phi u_\alpha),
\end{equation}
where $j_\alpha^\star$ is the post-collision flux and the relaxation
time $\tau_\phi$ is related to the mobility  via
\begin{equation}
\tau = {\textstyle\frac{1}{2}} + M\rho_0/\Delta t.
\end{equation}

\subsubsection{Finite difference approach}

For a number of reasons, it is advantageous to replace the lattice
kinetic approach by a conventional finite difference/finite volume
treatment of the Cahn-Hilliard equation
\begin{equation}
\partial_t \phi + \partial_\alpha ( \phi u_\alpha - M \partial_\alpha \mu) = 0.
\end{equation}
(Note that the finite difference and finite volume terminology will
be used somewhat interchanably here.) Briefly, the advantages are:
\begin{enumerate}
\item
The ambiguity in interpretation of the kinetic modes is avoided,
along with the corresponding computational overhead in memory and
memory movement.
\item
Additional features may be added to the Cahn-Hilliard equation more
easily. These might include a heterogeneous mobility $M=M(\phi)$, and
order parameter fluctuations (see below).
\end{enumerate}

\vfill
\pagebreak

### 2016-12-21 docs/lb.tex
trunk 4.9 updates to tests to reflect improved spelling cf svn 2976
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2988 40876441-3496-0410-a613-a2fca35b2ae2
trunk documentation update
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3053 40876441-3496-0410-a613-a2fca35b2ae2

--- 2016-03-03 trunk/docs/lb.tex
@@
 \sum_i f_i(\mathbf{r};t) c_{i\alpha} c_{i\beta}.
+\label{lb-f-moments}
 \end{equation}
@@
 
+Colloidal particles are assumed to be spherical with a geometrical
+centre $\mathbf{r}_c$, which is also the centre of mass. The
+centre is allowed to move continuously across the lattice
+with velocity $\mathbf{U}$; the particle has an angular velocity
+$\mathbf{\Omega}$. The surface of the colloid is defined by an
+input radius, $a_0$, which determines which lattice nodes are
+inside or outside the colloid. (The hydrodynamic properties of
+the colloid are specified by a different radius $a_h$ --- more
+of this later.) The boundary links are then the set of vectors
+joining lattice nodes which intersect the spherical surface
+$\{\mathbf{c}_b\}$. Note that a lattice node exactly at
+the solid-fluid interface is defined to be outside the colloid.
+
+In the original approach of Ladd, fluid occupied nodes both inside
+and outside the particle. The effect of the ``internal fluid'' is
+known to be restricted to short time scales (compared to the
+characteristic time $a_0^2/\nu$), on which the fluid inside the
+particle relaxes to a solid body rotation \cite{heemels}. However,
+we use fully solid particles via the approach introduced by
+Nguyen and Ladd \cite{nguyen-ladd2002}.
+
+A boundary link is defined as joining a node $\mathbf{r}$
+inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
+If the post-collision distributions are denoted by $f^\ast$, then
+the distributions must be reflected at the solid surface so that
+\begin{equation}
+\label{eq:colloid_bbl1}
+f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
+- \frac{2w_{c_b} \rho_0 \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
+\end{equation}
+where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$.
+Note that the local density at the fluid site $\rho(\mathbf{r};t)$
+is replaced by
+the mean fluid density $\rho_0$. in the second term on the right-hand side.
+The velocity at the boundary is
+\begin{equation}
+\label{eq-colloid-ub}
+\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b.
+\end{equation}
+
+The force exerted on a
+single link is
+\begin{equation}
+\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
+t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
+\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
+\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
+\label{eq-colloid-fb}
+\end{equation}
+with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
+The total hydrodynamic force on the particle is then found by taking
+the sum of
+$\mathbf{F}_b$ over all the boundary links defining the particle.
+There is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
+which again is summed over all links to give the total torque on the colloid.
+Colloid dynamics is discussed in more detail in Section~\ref{section:colloids}.
+
+% End section
 \vfill

### 2016-12-21 branches/kevin-1397/docs/input.tex
branch documentation updates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@3054 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  input.tex
%
%  Instructions for contents of the user input file.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  (c) 2016-2017 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Input File}

\subsection{General}

By default, the run time expects to find user input in a file
\texttt{input} in the current working directory. If a different
file name is required, its name
should be provided as the sole command line argument, e.g.,
\begin{lstlisting}
./Ludwig.exe input_file_name
\end{lstlisting}
If the input file is not located in the current working directory
the code will terminate immediately with an error message.

When an input file is located, its content is read by a single MPI
task, and its contents then broadcast to all MPI relevant tasks.
The format of the file is plain ASCII text, and its contents are
parsed on a line by line basis. Lines may contain the following:
\begin{itemize}
\item comments introduced by \texttt{\#}.
\item key value pairs separated by white space.
\end{itemize}
Blank lines are treated as comments. The behaviour of the code is
determined by a set of key value pairs. Any given key may appear
only once in the input file; unused key value pairs are not reported.
If the key value pairs are not correctly formed, the code will terminate
with an error message and indicate the offending input line.

Key value pairs may be present in the input file, but have no effect for
any given run: they are merely ignored. Relevant control parameters for
given input are reported in the standard output.

\subsubsection{Key value pairs}

Key value pairs are made up of a key --- an alphanumeric string with no
white space --- and corresponding value following white space. Values
may take on the follow forms:
\begin{lstlisting}
key_string           value_string

key_integer_scalar  1
key_integer_vector  1_2_3

key_double_scalar    1.0
key_double_vector    1.0_2.0_3.0
\end{lstlisting}
Values which are strings should contain no white space. Scalar parameters
may be integer values, or floating point values with a decimal point
(scientific notation is also allowed).  Vector parameters are introduced
by a set of three values (to be interpreted as $x,y,z$ components of the
vector in Cartesian coordinates) separated by an underscore. The identity
of the key will specify what type of value is expected. Keys and (string)
values are case sensitive.

Most keys have an associated default value which will be used if
the key is not present. Some keys must be specified: an error will
occur if they are missing. The remainder of this part
of the guide details the various choices for key value pairs,
along with any default values, and any relevant constraints.

\subsection{The Free Energy}
\label{input-free-energy}

The choice of free energy is determined as follows:
\begin{lstlisting}
free_energy   none
\end{lstlisting}
The default value is \texttt{none}, i.e., a simple Newtonian fluid is used.
Possible values of the \texttt{free\_energy} key are:
\begin{lstlisting}
#  none                     Newtonian fluid [DEFAULT]
#  symmetric                Symmetric binary fluid (finite difference)
#  symmetric_lb             Symmetric binary fluid (two distributions)
#  brazovskii               Brazovskii smectics
#  surfactant               Surfactants
#  polar_active             Polar active gels
#  lc_blue_phase            Liquid crystal (nematics, cholesterics, BPs)
#  lc_droplet               Liquid crystal emulsions
#  fe_electro               Single fluid electrokinetics
#  fe_electro_symmetric     Binary fluid electrokinetics
\end{lstlisting}

The choice of free energy will control automatically a number of factors
related to choice of order parameter, the degree of parallel communication
required, and so on. Each free energy has a number of associated parameters
discussed in the following sections.

Details of general (Newtonian) fluid parameters, such as viscosity,
are discussed in Section~\ref{input-fluid-parameters}.

\subsubsection{Symmetric Binary Fluids}

We recall that the free energy density is, as a function of compositional
order $\phi$:
\[
{\textstyle \frac{1}{2}} A\phi^2 +
{\textstyle \frac{1}{4}} B\phi^4 +
{\textstyle \frac{1}{2}} \kappa (\mathbf{\nabla}\phi)^2.
\]

Parameters are introduced by (with default values):
\begin{lstlisting}
free_energy  symmetric
A            -0.0625                         # Default: -0.003125
B            +0.0625                         # Default: +0.003125
K            +0.04                           # Default: +0.002
\end{lstlisting}
Common usage has $A < 0$ and $B = -A$ so that $\phi^\star = \pm 1$.
The parameter $\kappa$
(key \texttt{K}) controls
the interfacial energy penalty and is usually positive.

\subsubsection{Brazovskii smectics}
\label{input-brazovskki-smectics}
The free energy density is:
\[
{\textstyle \frac{1}{2}} A\phi^2 +
{\textstyle \frac{1}{4}} B\phi^4 +
{\textstyle \frac{1}{2}} \kappa (\mathbf{\nabla}\phi)^2 +
{\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 
\]

Parameters are introduced via the keys:
\begin{lstlisting}
free_energy  brazovskii
A             -0.0005                        # Default: 0.0
B             +0.0005                        # Default: 0.0
K             -0.0006                        # Default: 0.0
C             +0.00076                       # Default: 0.0
\end{lstlisting}
For $A<0$, phase separation occurs with a result depending on $\kappa$:
one gets two symmetric phases for $\kappa >0$ (cf.\ the symmetric case)
or a lamellar phase for $\kappa < 0$. Typically, $B = -A$ and the
parameter in the highest derivative $C > 0$.

\subsubsection{Surfactants}
\label{input-surfactants}

The surfactant free energy should not be used at the present time.

\subsubsection{Polar active gels}
\label{input-polar-active-gels}

The free energy density is a function of vector order parameter $P_\alpha$:
\[
{\textstyle \frac{1}{2}} A P_\alpha P_\alpha +
{\textstyle \frac{1}{4}} B (P_\alpha P_\alpha)^2 +
{\textstyle \frac{1}{2}} \kappa (\partial_\alpha P_\beta)^2
\]

There are no default parameters:
\begin{lstlisting}
free_energy        polar_active
polar_active_a    -0.1                       # Default: 0.0
polar_active_b    +0.1                       # Default: 0.0
polar_active_k     0.01                      # Default: 0.0
\end{lstlisting}
It is usual to choose $B > 0$, in which case $A > 0$ gives
an isotropic phase, whereas $A < 0$ gives a polar nematic phase.
The elastic constant $\kappa$ (key \texttt{polar\_active\_k})
is positive.

\subsubsection{Liquid crystal}
\label{input-liquid-crystal}
The free energy density is a function of tensor order parameter
$Q_{\alpha\beta}$:
\begin{eqnarray}
{\textstyle\frac{1}{2}} A_0 (1 - \gamma/3)Q^2_{\alpha\beta} -
{\textstyle\frac{1}{3}} A_0 \gamma
                        Q_{\alpha\beta}Q_{\beta\delta}Q_{\delta\alpha} +
{\textstyle\frac{1}{4}} A_0 \gamma (Q^2_{\alpha\beta})^2
\nonumber \\
+ {\textstyle\frac{1}{2}} \Big(
\kappa_0 (\epsilon_{\alpha\delta\sigma} \partial_\delta Q_{\sigma\beta} +
2q_0 Q_{\alpha\beta})^2 + \kappa_1(\partial_\alpha Q_{\alpha\beta})^2 \Big)
\nonumber
\end{eqnarray}

The corresponding \texttt{free\_energy} value, despite its name, is
suitable for nematics and cholesterics, and not just blue phases:
\begin{lstlisting}
free_energy      lc_blue_phase
lc_a0            0.01                       # Deafult: 0.0
lc_gamma         3.0                        # Default: 0.0
lc_q0            0.19635                    # Default: 0.0
lc_kappa0        0.00648456                 # Default: 0.0
lc_kappa1        0.00648456                 # Default: 0.0
\end{lstlisting}
The bulk free energy parameter $A_0$ is positive and controls the
energy scale (key \texttt{lc\_a0}); $\gamma$ is positive and
influences the position in the phase diagram relative to the
isotropic/nematic transition (key \texttt{lc\_gamma}).
The two elastic constants must be equal, i.e., we enforce the
single elastic constant approximation (both keys \texttt{lc\_kappa0} and
\texttt{lc\_kappa1} must be specified).

Other important parameters in the liquid crystal picture are:
\begin{lstlisting}
lc_xi            0.7                         # Default: 0.0
lc_Gamma         0.5                         # Default: 0.0
lc_active_zeta   0.0                         # Default: 0.0
\end{lstlisting}
The first is $\xi$ (key \texttt{lc\_xi}) is the effective molecular
aspect ratio and should be in the range $0< \xi< 1$. The rotational
diffusion constant is $\Gamma$ (key \texttt{lc\_Gamma}; not to be
confused with \texttt{lc\_gamma}). The (optional) apolar activity
parameter is $\zeta$ (key \texttt{lc\_active\_zeta}).

\subsubsection{Liquid crystal anshoring}

Different types of anchoring are available at solid surfaces, with
one or two related free energy parameters depending on the type.
The type of anchoring may be set independently for stationary
boundaries (walls) and colloids.
\begin{lstlisting}
lc_anchoring_strength     0.01               # free energy parameter w1
lc_anchoring_strength_2   0.0                # free energy parameter w2
lc_wall_anchoring         normal             # ``normal'' or ``planar''
lc_coll_anchoring         normal             # ``normal'' or ``planar''
\end{lstlisting}
See section \ref{section-lc-anchoring} for details of surface anchoring.

\subsubsection{Liquid crystal emulsion}
\label{input-liquid-crystal-emulsion}

This an interaction free energy which combines the symmetric and liquid
crystal free energies. The liquid crystal free energy constant $\gamma$
becomes a
function of composition via $\gamma(\phi) = \gamma_0 + \delta(1 + \phi)$,
and a coupling term is added to the free energy density:
\[
WQ_{\alpha\beta} \partial_\alpha \phi \partial_\beta \phi.
\]
Typically, we might choose $\gamma_0$ and $\delta$ so that
$\gamma(-\phi^\star) < 2.7$ and the $-\phi^\star$ phase is isotropic,
while $\gamma(+\phi^\star) > 2.7$ and the
$+\phi^\star$ phase is ordered (nematic, cholesteric, or blue phase).
Experience suggests that a suitable choice is $\gamma_0 = 2.5$ and
$\delta = 0.25$.

For anchoring constant $W > 0$, the liquid crystal anchoring at the
interface is planar, while for $W < 0$ the anchoring is normal. This
is set via key \texttt{lc\_droplet\_W}.

Relevant keys (with default values) are:
\begin{lstlisting}
free_energy            lc_droplet

A                      -0.0625
B                      +0.0625
K                      +0.053

lc_a0                   0.1
lc_q0                   0.19635
lc_kappa0               0.007
lc_kappa1               0.007

lc_droplet_gamma        2.586                # Default: 0.0
lc_droplet_delta        0.25                 # Default: 0.0
lc_droplet_W           -0.05                 # Default: 0.0
\end{lstlisting}
Note that key \texttt{lc\_gamma} is not set in this case.

\subsection{System Parameters}
\label{input-system-parameters}

Basic parameters controlling the number of time steps
and the system size are:
\begin{lstlisting}
N_start      0                              # Default: 0
N_cycles     100                            # Default: 0
size         128_128_1                      # Default: 64_64_64
\end{lstlisting}
A typical simulation will start from time zero (key \texttt{N\_start})
and run for a certain number of time steps (key \texttt{N\_cycles}).
The system size (key \texttt{size}) specifies the total number of
lattice sites in each dimension. If a two-dimensional system is
required, the extent in the $z$-direction must be set to unity, as
in the above example.

If a restart from a previous run is required, the choice of parameters
may be as follows:
\begin{lstlisting}
N_start      100
N_cycles     400
\end{lstlisting}
This will restart from data previously saved at time step 100, and
run a further 400 cycles, i.e., to time step 500.
% TODO Cross-reference to I/O.

\subsubsection{Parallel decomposition}

In parallel, the domain decomposition is closely related to the
system size, and is specified as follows:
\begin{lstlisting}
size         64_64_64
grid         4_2_1
\end{lstlisting}
The \texttt{grid} key specifies the number of MPI tasks required in
each coordinate direction. In the above example, the decomposition
is into 4 in the $x$-direction, into 2 in the $y$-direction, while
the $z$-direction is not decomposed. In this example, the local domain
size per MPI
task would then be $16\times32\times64$. The total number of MPI tasks
available must match the total implied by \texttt{grid} (8 in the
example).

The \texttt{grid} specifications must exactly divide the system size;
if no decomposition is possible, the code will terminate with an error
message.
If the requested decomposition is not valid, or \texttt{grid} is
omitted, the code will try to supply a decomposition based on
the number of MPI tasks available and \texttt{MPI\_Dims\_create()};
this may be implementation dependent.

\subsection{Fluid Parameters}
\label{input-fluid-parameters}

Control parameters for a Newtonian fluid include:
\begin{lstlisting}
fluid_rho0                 1.0
viscosity                  0.166666666666666
viscosity_bulk             0.166666666666666
isothermal_fluctuations    off
temperature                0.0
\end{lstlisting}
The mean fluid density is $\rho_0$ (key \texttt{fluid\_rho0}) which
defaults to unity in lattice units; it is not usually necessary to
change this. The shear viscosity is
\texttt{viscosity} and as default value 1/6 to correspond to
unit relaxation time in the lattice Boltzmann picture. Reasonable
values of the shear viscosity are $0.2 > \eta > 0.0001$ in lattice
units. Higher values move further into the over-relaxation region, and can
result in poor behaviour. Lower
values increase the Reynolds number and tend to cause
problems with stability. The bulk
viscosity has a default value which is equal to whatever shear
viscosity has been selected. Higher values of the bulk viscosity
may be set independently and can help to suppress large deviations
from incompressibility and maintain numerical stability
in certain situations.

If fluctuating hydrodynamics is wanted, set the value of
 \texttt{isothermal\_fluctuations} to \texttt{on}. The associated
temperature is in lattice units: reasonable values (at $\rho_0 = 1$)
are $0 < kT < 0.0001$. If the temperature is too high, local
velocities will rapidly exceed the Mach number constraint and
the simulation will be unstable.

\subsection{Lees Edwards Planes}

Constant uniform shear may be introduced via a number of Lees Edwards
planes with given speed. This is appropriate for fluid-only
systems with periodic boundaries.
\begin{lstlisting}
N_LE_plane               2           # Number of planes (default: 0)
LE_plane_vel             0.05        # Constant plane speed
\end{lstlisting}
The placing of the planes in the system is as follows.
The number of planes $N$ must
divide evenly the lattice size in the $x$-direction to give an integer
$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
All planes have the same, constant, velocity jump associated with them:
this is positive in the positive $x$-direction. (This jump is usually
referred to as the plane speed.) The uniform shear rate
will be $\dot{\gamma} = N U_{LE} / L_x$ where $U_{LE}$ is the plane
speed (which is always in the $y$-direction).

The velocity gradient or shear direction is $x$, the flow
direction is $y$ and the vorticity direction is $z$.

The spacing between planes must not be less than twice the halo size
in lattice units; 8--16 lattice units may be the practical limit in
many cases. In additional, the speed of the planes must not cause a
violation of the
Mach number constraint in the fluid velocity on the lattice, which
will match the plane speed in magnitude directly either side of the
planes. A value of around 0.05 should be regarded as a maximum safe
limit for practical purposes.

Additional keys associated with the Lees Edwards planes are:
\begin{lstlisting}
LE_init_profile          1           # Initialise u(x) (off/on)
LE_time_offset           10000       # Offset time (default 0)
\end{lstlisting}
If \texttt{LE\_init\_profile} is set, the fluid velocity is initialised
to reflect a steady state shear flow appropriate for the number of
planes at the given speed (ie., shear rate). If set to zero (the default),
the fluid is initialised with zero velocity.

The code works out the current displacement of the planes by computing
$U_{LE} t$, where $t$ is the current time step. A shear run should then
start from $t = 0$, i.e. zero plane displacement.
It is often convenient to run an equilibration with no shear, and
then to start an experiment after some number of steps. This
key allows you to offset the start of the Lees-Edwards motion.
It should then take the value of the start time (in time steps)
corresponding to the restart at the end of the equilibration
period.

There are a couple of additional constraints to use the Lees-Edwards
planes in parallel. In particular, the planes cannot fall at a
processor boundary in the $x$-direction. This means you should
arrange an integer number of planes per process in the $x$-direction.
(For example, use one plane per process; this will also ensure the number
of planes
still evenly divides the total system size.)
This will interleave the planes with the processor decomposition.
The $y$-direction and $z$-direction may be decomposed without
further constraint.

Note that this means a simulation with one plane will only work
if there is one process in the $x$ decomposition.

\subsection{Colloids}
If no relevant key words are present, no colloids will be
expected at run time. The simulation will progress in the
usual fashion with fluid only.

If colloids are required, the \texttt{colloid\_init}
key word must be present to allow the code to determine where
colloid information will come from. The options for the
\texttt{colloid\_init} key word are summarised below:
\begin{lstlisting}
colloid_init             none

#  none                  no colloids [DEFAULT]
#  input_one             one colloid from input file
#  input_two             two colloids from input file
#  input_three           three colloids from input file
#  input_random          Small number at random
#  from_file             Read a separate file (including all restarts)
\end{lstlisting}

For idealised simulations which require 1, 2, or 3 colloids, relevant
initial state information 
for each particle can be included in the input file. In principle, most
of the colloid state as defined in the colloid
state structure in \texttt{colloid.h} may be specified. (Some state is
reserved for internal use only and cannot be set from the input file.)
Furthermore, not all the state is relevant in all simulations ---
quantities such as charge and wetting parameters may not be required,
in which case they a simply ignored.

A minimal initialisation is shown in the following example:
\begin{lstlisting}
colloid_init              input_one

colloid_one_a0            1.25
colloid_one_ah            1.25
colloid_one_r             12.0_12.0_12.0
\end{lstlisting}
This initialises a single colloid with an input radius $a_0=1.25$,
and a hydrodynamic radius $a_h=1.25$; in general both are required,
but they do not have to be equal.
A valid position is required within the extent of the system
$0.5 < x,y,z < L + 0.5$ as specified by the \texttt{size} key word.
State which is not explicitly defined is initialised to zero.

\subsubsection{General Initialisation}

A full list of colloid state-related key words is as follows. All
the quantities are floating point numbers unless explicitly stated
to be otherwise:
\begin{lstlisting}
# colloid_one_nbonds      (integer) number of bonds
#   colloid_one_bond1     (integer) index of bond partner 1
#   colloid_one_bond2     (integer) index of bond partner 2
# colloid_one_isfixedr    colloid has fixed position (integer 0 or 1)
# colloid_one_isfixedv    colloid has fixed velocity (integer 0 or 1)
# colloid_one_isfixedw    colloid has fixed angular velocity (0 or 1)
# colloid_one_isfixeds    colloid has fixed spin (0 or 1)
# colloid_one_type        ``default'' COLLOID_TYPE_DEFAULT
#                         ``active''  COLLOID_TYPE_ACTIVE
#                         ``subgrid'' COLLOID_TYPE_SUBGRID
# colloid_one_a0          input radius
# colloid_one_ah          hydrodynamic radius
# colloid_one_r           position vector
# colloid_one_v           velocity (vector)
# colloid_one_w           angular velocity (vector)
# colloid_one_s           spin (unit vector)
# colloid_one_m           direction of motion (unit) vector for swimmers 
\end{lstlisting}

Note that for magnetic particles, the appropriate initialisation involves
the spin key word \texttt{colloid\_one\_s} which relates to the dipole
moment $\mu \mathbf{s}_i$, while \texttt{colloid\_one\_m} relates to the
direction of motion vector. Do not confuse the two.
It is possible in principle to have magnetic active particles,
in which case the dipole direction or spin ($\mathbf{s}_i$) and the
direction of swimming motion $\mathbf{m}_i$ are allowed to be distinct.

\begin{lstlisting}
# colloid_one_b1          Squirmer parameter B_1
# colloid_one_b2          Squirmer parameter B_2
# colloid_one_rng         (integer) random number generator state
# colloid_one_q0          charge (charge species 0)
# colloid_one_q1          charge (charge species 1)
# colloid_one_epsilon     Permeativity
# colloid_one_c           Wetting parameter C
# colloid_one_h           Wetting parameter H
\end{lstlisting}

{\bf Example}: Single active (squirmer) particle

The following example shows a single active particle with initial
swimming direction along the $x$-axis.
\begin{lstlisting}
colloid_init              input_one

colloid_one_type          active
colloid_one_a0            7.25
colloid_one_ah            7.25
colloid_one_r             32.0_32.0_32.0
colloid_one_v             0.0_0.0_0.0
colloid_one_m             1.0_0.0_0.0
colloid_one_b1            0.05
colloid_one_b2            0.05
\end{lstlisting}

\subsubsection{Random initialisation}

For suspensions with more than few colloids, but still at
relatively low volume fraction (10--20\% by volume), it is
possible to request initialisation at random positions.

The additional key word value pair \texttt{colloid\_random\_no}
determines the total number of particles to be placed in
the system. To prevent particles being initialised very
close together, which can cause problems in the first few
time steps if strong potential interactions are present,
a ``grace'' distance or minimum surface-surface separation
may also be specified (\texttt{colloid\_random\_dh}).

The following example asks for 100 colloids to be initialised
at random positions, with a minimum separation of 0.5 lattice
spacing.

\begin{lstlisting}
colloid_init              input_random

colloid_random_no         100             # Total number of colloids
colloid_random_dh         0.5             # ``Grace'' distance

colloid_random_a0         2.30
colloid_random_ah         2.40
\end{lstlisting}

An input radius and hydrodynamic radius must be provided: these
are the same for all colloids.
If specific initialisations of the colloid state (excepting the
position) other than the radii are wanted, values should be provided
as for the single particle case in the preceding section, but using
\texttt{colloid\_random\_a0} in place of
\texttt{colloid\_one\_a0} and so on.

The code will try to initialise the requested number in the current
system size, but only makes a finite number of attempts to place
particles at random with no overlaps. (The initialisation will also
take into account the presence of any solid walls, using the same
grace distance.) If the the number of particles is too large, the
code will halt with a message to that effect.

In general, colloid information for a arbitrary configuration with many
particles should be read from a pre-prepared file. See the section on
File I/O for further information on reading files.

\subsubsection{Interactions}

Note that two-body pair-potential interactions are defined uniformly for
all colloids in a simulation. The same is true for lubrication corrections.
There are a number of constraints related to the computation of
interactions discussed below.

{\bf Boundary-colloid lubrication correction}.
Lubrication corrections (here the normal force) between a flat wall
(see section XREF) are
required to prevent overlap between colloid  and the wall.
The cutoff distance is set via the key word value pair
\begin{lstlisting}
boundary_lubrication_rcnormal    0.5
\end{lstlisting}
It is recommended that this is used in all cases involving walls.
A reasonable value is in the range $0.1 < r_c < 0.5$ in lattice
units, and should be calibrated for particle hydrodynamic radius
and fluid viscosity if exact results are important.

{\bf Colloid-colloid lubrication corrections}.
The key words to activate the calculation of lubrication corrections
are:
\begin{lstlisting}
lubrication_on                   1
lubrication_normal_cutoff        0.5
lubrication_tangential_cutoff    0.05
\end{lstlisting}

{\bf Soft sphere potential}.
A cut-and-shifted soft-sphere potential of the form
$v \sim \epsilon (\sigma/r)^\nu$ is
available. Some trial-and-error with the parameters may be required in
any given situation to ensure simulation stability in the long run. The
following gives an example of the relevant input key words:
\begin{lstlisting}
soft_sphere_on            1                 # integer 0/1 for off/on 
soft_sphere_epsilon       0.0004            # energy units
soft_sphere_sigma         1.0               # a length
soft_sphere_nu            1.0               # exponent is positive
soft_sphere_cutoff        2.25              # a surface-surface separation
\end{lstlisting}
See Section~\ref{section-colloids-soft-sphere} for a detailed description.

{\bf Lennard-Jones potential}.
The Lennard-Jones potential (Section~\ref{section-colloids-lennard-jones})
is controlled by the following key words:
\begin{lstlisting}
lennard_jones_on          1                 # integer 0/1 off/on
lj_epsilon                0.1               # energy units
lj_sigma                  2.6               # potential length scale
lj_cutoff                 8.0               # a centre-centre separation
\end{lstlisting}

{\bf Yukawa potential}.
A cut-and-shifted Yukawa potential of the form
$v \sim \epsilon \exp(-\kappa r)/r$ is
available using the following key word value pairs:

\begin{lstlisting}
yukawa_on                 1                 # integer 0/1 off/on
yukawa_epsilon            1.330             # energy units
yukawa_kappa              0.725             # an inverse length
yukawa_cutoff             16.0              # a centre-centre cutoff
\end{lstlisting}

{\bf Dipole-dipole interactions (Ewald sum)}.
The Ewald sum is completely specified in the input file
by the uniform dipole strength $\mu$ and the real-space cut off $r_c$.  
\begin{lstlisting}
ewald_sum                 1                 # integer 0/1 off/on
ewald_mu                  0.285             # dipole strength mu
ewald_rc                  16.0              # real space cut off
\end{lstlisting}

If short range interactions are required, particle information is stored
in a cell list, which allows efficient computation of the potentially
$N^2$ interactions present. This gives rise to a constraint that the
width of the cells must be large enough that all relevant interactions
are included. This generally means that the cells must be at least
$2a_h + h_c$ where $h_c$ is the largest relevant cut off distance.

The requirement for at least two cells per local domain in parallel
means that there is a associated minimum local domain size. This is
computed at run time on the basis of the input. If the local domain
is too small, the code will terminate with an error message. The
local domain size should be increased.

\subsubsection{External forces}

{\bf External body force (gravity)}.
The following example requests a uniform body force in the negative
$z$-direction on all particles.
\begin{lstlisting}
colloid_gravity           0.0_0.0_-0.001    # vector
\end{lstlisting}
The counterbalancing body force on the fluid which enforces the
constraint of momentum conservation for the system as a whole is
computed automatically by the code at each time step.

\subsection{Porous media}

Porous media calculations can be undertaken when appropriate
solid/fluid status information is supplied. The are a number
of switches available in the input file:
\begin{lstlisting}
porous_media_file
porous_media_format
porous_media_type
\end{lstlisting}

specifies the file stub name to be read at the start of execution.
(If the stub name is \texttt{file} then the code will expect to
find \texttt{file.001-001} in the current directory.

is either \texttt{ASCII} or \texttt{BINARY} as appropriate. Note that
in parallel, a single data file can be supplied, but it must be binary.
The default is \texttt{BINARY}.

is either \texttt{status\_only} (the default) or \texttt{status\_with\_h}.

\subsubsection{File format}

In all cases the status (fluid/solid) information is represented by a
single \texttt{char} (or integer), which must be supplied via the
porous media file. A single file should contain data matching the
current system size, and have the $z-$direction running fastest,
followed by the $y-$direction. Note that in non-periodic directions,
the structure must be 'closed', i.e., all the points at the edge
should be solid.

Fluid sites are designated by \texttt{0} and boundary or solid sites
by \texttt{1}. These data should be of type \texttt{char} in binary,
and may be integer in ASCII.

Where wetting information is required, the free energy parameter $H$
can be supplied by using the \texttt{status\_with\_h} switch. In this
case, the \texttt{char} status is augmented by a single \texttt{double}
value which is the local value of $H$. The order is then
$s_1,h_1, s_2, h_2, \ldots$.

An example of how to construct a porous media file is provided in
\texttt{util/capillary.c}, which builds an appropriate file for
a square or circular capillary tube. Please see the comments in
the file for further details. Note that the allowed
solid/fluid status values are defined in \texttt{src/site\_map.h}.
A solid boundary is \texttt{BOUNDARY}, while fluid is \texttt{FLUID}.

\subsection{Order parameter initialisations}

Free energy choices requiring one or more fluid order parameters can make
use of the following initial coniditions.

\subsubsection{Composition $\phi$}

The following initialisations are available.
\begin{lstlisting}
phi_initialisation          spinodal    # spinodal
phi0                        0.0         # mean
noise                       0.05        # noise amplitude
random_seed                 102839      # +ve integer
\end{lstlisting}
Suitable for initialising isothermal spinodal decomposition, the order
parameter may be set at random at each position via
$\phi = \phi_0 + A(\hat{\phi} - 1/2)$ with the random variate
$\hat\phi$ selected uniformly on the interval $[0,1)$. For symmetric
quenches (mean order parameter $\phi_0 = 0$ and $\phi^\star = \pm 1$),
a value of $A$ in the range 0.05-0.1 is usually appropriate.

For off-symmetric quenches, larger patches of fluid may be required to
initiate decomposition:
\begin{lstlisting}
phi_initialisation          patches     # patches of phi = +/- 1
phi_init_patch_size         2           # patch size
phi_init_patch_vol          0.1         # volume fraction phi = -1 phase
random_seed                 13          # +ve integer
\end{lstlisting}
The initialises cubics patches of fluid of given size with $\phi= \pm 1$
at random. The requested overall volume fractions may be met approximately.

A spherical drop can be initialised at the centre of the system.
\begin{lstlisting}
phi_initialisation          drop        # spherical droplet
phi_init_drop_radius        16.0        # radius
phi_init_drop_amplitude    -1.0         # phi value inside
\end{lstlisting}
The drop is initialised with a $\tanh(r/\xi)$ profile where the
interfacial width $\xi$ is computed via the current free energy
parameters.

% block
% bath

For restarted simulations, the default position is to read order
parameter information from file
\begin{lstlisting}
phi_initialisation          from_file
\end{lstlisting}
in which case a file or files for the appropriate time step should
be present in the working directory.

\subsubsection{Tensor order $Q_{\alpha\beta}$}

A number of different initialisations are available for the liquid
crystal order parameter $Q_{\alpha\beta}$. Some care may be required
to ensure consistency between the choice and the free energy
parameters, the system size, and so on (particularly for the blue phases).

A summmary of choices is:
\begin{lstlisting}
lc_q_initialisation   nematic          # uniform nematic...
lc_init_nematic       1.0_0.0_0.0      # ...with given director

lc_q_initialisation   cholesteric_x    # cholesteric with helical axis x
lc_q_initialisation   cholesteric_y    # cholesteric with helical axis y
lc_q_initialisation   cholesteric_z    # cholesteric with helical axis z

lc_q_initialisation   o8m              # BPI high chirality limit
lc_q_initialisation   o2               # BPII high chirality limit
lc_q_initialisation   o5
lc_q_initialisation   h2d              # 2d hexagonal
lc_q_initialisation   h3da             # 3d hexagonal BP A
lc_q_initialisation   h3db             # 3d hexagonal BP B
lc_q_initialisation   dtc              # double twist cylinders

lc_q_initialisation   bp3

lc_q_initialisation   cf1_x            # cholesteric ``finger'' axis x
lc_q_initialisation   cf1_y            # cholesteric ``finger'' axis y
lc_q_initialisation   cf1_z            # cholesteric ``finger'' axis z

lc_q_initialisation   cf1_fluc_x       # as cf1_x with random perterbations
lc_q_initialisation   cf1_fluc_y       # as cf1_y with random perturbations
lc_q_initialisation   cf1_flux_z       # as cf1_z with random perturbations

lc_q_initialistion    random           # with randomly chosen unit director
\end{lstlisting}
Note many of the initialiations require an initial amplitude of order,
which should be set via
\begin{lstlisting}
lc_q_init_amplitude   0.01             # initial amplitude of order A
\end{lstlisting}
For example, if an initial uniform nematic is requested with
unit director $n_\alpha$, the corresponding initial tensor will be
$$
Q_{\alpha\beta} = 
{\textstyle \frac{1}{2}} A (3 n_\alpha n_\beta - \delta_{\alpha\beta}).
$$

\vfill
\pagebreak

### 2016-12-21 branches/kevin-1397/docs/liquidcrystal.tex
branch documentation updates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@3054 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  liquidcrystal.tex
%
%  For nematics, cholesterics, blue phases based on the Landau
%  de Gennes approach.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Contributing authors:
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  Davide Marenduzzo was the source of the original LC approach.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Liquid Crystals}

\subsection{The Landau-de Gennes Approach}

\subsubsection{Tensor order parameter}

The liquid crystal order is described by a symmetric traceless tensor
$Q_{\alpha\beta}(\mathbf{r}; t)$. The largest eigenvalue and
associated eigenvector of $Q_{\alpha\beta}$ represent the magnitude
and local direction of liquid crystal molecular order. A theory based
on the tensor $Q_{\alpha\beta}$ has the advantage of being able to
describe disclinations where the local order vanishes and a director
is not defined. Readers are referred to, e.g., de Gennes and Prost
\cite{degennes-prost2002}, and Wright and Mermin \cite{wright-mermin}
for further information.

\subsubsection{Free Energy Density}

The free energy is a functional whose density
$f(Q_{\alpha\beta}, \partial_\gamma Q_{\alpha\beta})$
has bulk contributions depending on $Q_{\alpha\beta}$,
and elastic (distortion) contributions dependent on the
gradients of the order parameter
$\partial_\gamma Q_{\alpha\beta}$.

The bulk contributions are
\begin{equation}
f(Q_{\alpha\beta}) =
{\textstyle\frac{1}{2}} A_0 (1 - \gamma / 3) Q_{\alpha\beta}^2
-{\textstyle\frac{1}{3}} A_0 \gamma Q_{\alpha\beta} Q_{\beta\pi} Q_{\pi\alpha}
+ {\textstyle\frac{1}{4}} A_0 \gamma (Q_{\alpha\beta}^2)^2.
\end{equation}
Here, $A_0$ is a constant which sets the overall energy scale; $\gamma$
is a temperature-like parameter which controls the position in the
phase diagram relative to the isotropic-nematic transition.

The bend, splay and twist distortions making up the gradient free
energy are modelled with two elastic
constants $\kappa_0$ and $\kappa_1$ as
\begin{equation}
f(\partial_\gamma Q_{\alpha\beta}) = 
{\textstyle\frac{1}{2}} \kappa_0 (\partial_\alpha Q_{\alpha\beta})^2
+ {\textstyle\frac{1}{2}} \kappa_1
(\epsilon_{\alpha\mu\nu} \partial_\mu Q_{\nu\beta} + 2q_0 Q_{\alpha\beta})^2.
\label{equation-lc-gradient-fe}
\end{equation}
Here, $\epsilon_{\alpha\mu\nu}$ is the permutation tensor and
$q_0 = 2\pi/p$ is a wavevector related to the pitch~$p$ of the
cholesteric. Setting $q_0 = 0$ permits a nematic state.
The one-constant approximation makes the simplification that
$\kappa_0 = \kappa_1 = \kappa$; the distinction between
$\kappa_0$ and $\kappa_1$ will be kept
in what follows unless otherwise stated.

\subsubsection{Molecular Field}

The molecular field $H_{\alpha\beta}$ is defined as the functional
derivative of the
free energy density with respect to the order parameter which here
gives
\begin{equation}
H_{\alpha\beta} = -\Bigg[
  \frac{\partial f(Q_{\alpha\beta})}{\partial Q_{\alpha\beta}}
- \partial_\gamma \frac{\partial f(\partial_\gamma Q_{\alpha\beta})}
{\partial Q_{\alpha\beta,\gamma}} \Bigg].
\end{equation}
In the absence of flow, the molecular field determines how the
order parameter relaxes toward equilibrium.

As the final expression for $H_{\alpha\beta}$ is somewhat complex, it
is useful to 
account for the individual terms here. First, the terms from the bulk free
energy density give rise to:
\begin{equation}
-A_0 (1 - \gamma/3) Q_{\alpha\beta} + A_0 \gamma Q_{\alpha\pi} Q_{\pi\beta}
- A_0 \gamma Q_{\pi\sigma}^2 Q_{\alpha\beta}.
\nonumber
\end{equation}
The second term in this expression
(arising from the cubic term in the free energy) is
forced to be traceless by subtracting one third of
Tr$(Q_{\alpha\pi}Q_{\pi\beta}) = Q_{\alpha\beta}^2$
so that we have
\begin{equation}
- A_0 (1 - \gamma/3) Q_{\alpha\beta}
+ A_0 \gamma (Q_{\alpha\pi} Q_{\pi\beta}
               - {\textstyle\frac{1}{3}} Q_{\pi\sigma}^2\delta_{\alpha\beta})
- A_0 \gamma Q_{\pi\sigma}^2 Q_{\alpha\beta}.
\nonumber
\end{equation}

The gradient term in $\kappa_0$, along with the first term in $\kappa_1$
arising from the expansion of the brackets in
Equation~\ref{equation-lc-gradient-fe}, give
\begin{equation}
\kappa_0 \partial_\alpha \partial_\gamma Q_{\gamma\beta}.
+
\kappa_1 \partial_\gamma
( \partial_\gamma Q_{\alpha\beta} - \partial_\alpha Q_{\gamma\beta}).
\nonumber
\end{equation}
Note that in the one constant approximation, this term collapses
to $\kappa \partial^2 Q_{\alpha\beta}$. The cross term in the expansion
of the brackets in~\ref{equation-lc-gradient-fe}, having been symmetrised,
contributes
\begin{equation}
-2\kappa_1 q_0
( \epsilon_{\alpha\mu\nu} \partial_\mu Q_{\nu\beta}
+ \epsilon_{\beta\mu\nu} \partial_\mu Q_{\nu\alpha}),
\nonumber
\end{equation}
which must again be forced to be traceless to give
\begin{equation}
-2\kappa_1 q_0 \big\{
( \epsilon_{\alpha\mu\nu} \partial_\mu Q_{\nu\beta}
+ \epsilon_{\beta\mu\nu} \partial_\mu Q_{\nu\alpha})
- {\textstyle\frac{1}{3}}( 
   \epsilon_{\pi\mu\nu} \partial_\mu Q_{\nu\pi}
  +\epsilon_{\pi\mu\nu} \partial_\mu Q_{\nu\pi} )
   \delta_{\alpha\beta} \big\}.
\nonumber
\end{equation}
The final term in is
$-4 \kappa_1 q_0^2 Q_{\alpha\beta}$. The complete expression for the molecular
field is then
\begin{eqnarray}
H_{\alpha\beta} &=&
- A_0 (1 - \gamma/3) Q_{\alpha\beta}
+ A_0 \gamma (Q_{\alpha\pi} Q_{\pi\beta}
               - {\textstyle\frac{1}{3}} Q_{\pi\sigma}^2\delta_{\alpha\beta})
- A_0 \gamma Q_{\pi\sigma}^2 Q_{\alpha\beta}
\nonumber\\
&&
+\kappa_0 \partial_\alpha \partial_\gamma Q_{\gamma\beta}
+
\kappa_1 \partial_\gamma
( \partial_\gamma Q_{\alpha\beta} - \partial_\alpha Q_{\gamma\beta})
\nonumber\\
&&
-2\kappa_1 q_0 \big(
( \epsilon_{\alpha\mu\nu} \partial_\mu Q_{\nu\beta}
+ \epsilon_{\beta\mu\nu} \partial_\mu Q_{\nu\alpha})
- {\textstyle\frac{1}{3}}( 
   \epsilon_{\pi\mu\nu} \partial_\mu Q_{\nu\pi}
  +\epsilon_{\pi\mu\nu} \partial_\mu Q_{\nu\pi} )
   \delta_{\alpha\beta} \big)
\nonumber\\
&& - 4 \kappa_1 q_0^2 Q_{\alpha\beta}.
\end{eqnarray}

\subsection{Surface Anchoring}

\label{section-lc-anchoring}

The preferred orientation of the liquid crystal fluid at a solid surface,
usually referred to as the surface anchoring, is relevant for both
solid walls and colloids.

\subsubsection{General boundary condition}

To impose a suitable boundary condition for the order parameter tensor
at a solid-fluid interface, we consider again the gradient terms in
the free energy
\begin{equation}
f (\partial_\gamma Q_{\alpha\beta}) = 
{\textstyle\frac{1}{2}} \kappa_0 (\partial_\beta Q_{\alpha\beta})^2
+ {\textstyle\frac{1}{2}}
 \kappa_1 (\epsilon_{\alpha\gamma\sigma} \partial_\gamma
Q_{\sigma\beta} + 2q_0 Q_{\alpha\beta})^2
\end{equation}
where we retain
two elastic constants $\kappa_0$ and $\kappa_1$;
the cholesteric pitch $p = 2\pi/q_0$.

Also relevant is a surface free energy (an area density), the exact form
of which is determined by the type of anchoring required. In general, we
expect this to depend on the order parameter, but not its gradients, so
write
\begin{equation}
f_s = f_s (Q_{\alpha\beta}, Q^0_{\alpha\beta})
\end{equation}
where $Q^0_{\alpha\beta}$ is some preferred order parameter configuration
at the surface (to be either normal or planar as discussed in the following
sections).

The boundary condition we wish to apply is derived from the Euler-Lagrange
equations, and for the fluid and surface terms gives rise to the equation
\begin{equation}
n_\gamma \frac{\partial f}{\partial Q_{\alpha\beta,\gamma}}
+ \frac{\partial f_s}{\partial Q_{\alpha\beta}} = 0,
\label{equation-lc-general-bc}
\end{equation}
where $n_\gamma$ is the outward unit normal at the surface (pointing into
the fluid). Note that the first term in the derivative of the fluid free
energy term with respect to
$Q_{\alpha\beta,\gamma}$ can be expended as
\begin{equation}
\kappa_0 n_\beta \partial_\gamma Q_{\alpha\gamma}
+ \kappa_1 n_\gamma
(\partial_\gamma Q_{\alpha\beta} - \partial_\alpha Q_{\gamma\beta})
- 2\kappa_1 q_0 n_\gamma \epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}.
\nonumber
\end{equation}
However, we should use a symmetric form (the derivative with respect to
$Q_{\beta\alpha,\gamma}$ is just as good) so we write this as:
\begin{eqnarray}
&
{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
+ n_\beta \partial_\gamma Q_{\alpha\gamma})
+ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
+ \partial_\beta Q_{\gamma\alpha})
\nonumber
\\
&
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha}).
\nonumber
\end{eqnarray}
The exact boundary condition will contain this term plus one
depending on the exact choice of the surface free energy which
describes a given anchoring condition.

\subsubsection{Normal or homeotropic anchoring}
The preferred direction of the surface order here is, as the name
suggests, normal to surface. We can write
\begin{equation}
f_s = {\textstyle\frac{1}{2}} w (Q_{\alpha\beta} - Q_{\alpha\beta}^0)^2
\end{equation}
where $w$ is a constant.
The preferred orientation $Q^0_{\alpha\beta}$ is based on the unit normal
at the surface $n_\gamma$, and is computed via the uniaxial approximation:
\begin{equation}
Q^0_{\alpha\beta}
= {\textstyle \frac{1}{2}} A (3n_\alpha n_\beta - \delta_{\alpha\beta}).
\end{equation}
The amplitude $A$ is provided by Equation~\ref{equation-lc-amplitude}. The
full boundary condition for the gradient of the tensor order parameter
at the solid-fluid boundary from Equation~\ref{equation-lc-general-bc}
is then:
\begin{eqnarray}
{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
+ n_\beta \partial_\gamma Q_{\alpha\gamma})
+ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
+ \partial_\beta Q_{\gamma\alpha})
\nonumber
\\
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
- w(Q_{\alpha\beta} - Q_{\alpha\beta}^0) = 0.
\label{equation-lc-bc-normal}
\end{eqnarray}

\subsubsection{Planar (degenerate) anchoring}

For planar anchoring, the preferred orientation is in the local tangent
plane at the surface: this is a degenerate case as any orientation in
the plane is energetically equivalent. An appropriate boundary
condition is described by Fournier and Galatola \cite{fournier2005}
which we write as:
\begin{equation}
f_s =
{\textstyle\frac{1}{2}} w_1 (\tilde{Q}_{\alpha\beta} - Q^\perp_{\alpha\beta})^2
+ {\textstyle\frac{1}{2}} w_2 (\tilde{Q}_{\alpha\beta}^2 - A^2)^2.
\end{equation}
Here again, the amplitude $A$ is provided by
Equation~\ref{equation-lc-amplitude}. To compute this term we take
the local fluid order parameter $Q_{\alpha\beta}$, form the quantity
$$
\tilde{Q}_{\alpha\beta}
= Q_{\alpha\beta} + {\textstyle \frac{1}{2}A\delta_{\alpha\beta} }
$$
which is then projected onto the tangent plane via
$
\tilde{Q}^\perp_{\alpha\beta}
= P_{\alpha\gamma} \tilde{Q}_{\gamma\sigma} P_{\sigma\beta}
$
with the local surface normal entering through
$P_{\alpha\beta} = \delta_{\alpha\beta} - n_\alpha n_\beta$.
The full boundary condition resulting from
Equation~\ref{equation-lc-general-bc} is then
\begin{eqnarray}
{\textstyle\frac{1}{2}} \kappa_0 (n_\alpha \partial_\gamma Q_{\beta\gamma}
+ n_\beta \partial_\gamma Q_{\alpha\gamma})
+ \kappa_1 n_\gamma \partial_\gamma Q_{\alpha\beta}
- {\textstyle\frac{1}{2}} \kappa_1 n_\gamma ( \partial_\alpha Q_{\gamma\beta}
+ \partial_\beta Q_{\gamma\alpha})
\nonumber
\\
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
- w_1 (\tilde{Q}_{\alpha\beta} - \tilde{Q}_{\alpha\beta}^\perp)
- 2w_2(\tilde{Q}_{\alpha\beta}^2 - A^2)\tilde{Q}_{\alpha\beta} = 0.
\label{equation-lc-bc-planar}
\end{eqnarray}

\subsection{Dynamics}

The time evolution of the orientational order parameter $Q_{\alpha\beta}$
in the presence of flow can be described by the Beris-Edwards equation
\cite{beris-edwards}.

\begin{equation}
\partial_t Q_{\alpha\beta} + \partial_\gamma (u_\gamma Q_{\alpha\beta})
+ S_{\alpha\beta}(W_{\alpha\beta}, Q_{\alpha\beta}) = -\Gamma  H_{\alpha\beta}.
\label{equation-lc-beris-edwards}
\end{equation}
This relates the time rate of change of the order parameter to terms
involving advection, the response to shear
$S_{\alpha\beta}(W_{\alpha\beta},Q_{\alpha\beta})$, and the molecular
field.

The advective term involves the fluid velocity $u_\alpha$ and the shear
term involves the velocity gradient tensor
$W_{\alpha\beta}= \partial_\beta u_\alpha$. The full definition of the
shear term is:
\begin{eqnarray}
S_{\alpha\beta} (W_{\alpha\beta}, Q_{\alpha\beta})
& = 
(\xi D_{\alpha\pi} + \Omega_{\alpha\pi})
(Q_{\pi\beta} + {\textstyle \frac{1}{3}} \delta_{\pi\beta})
+
(Q_{\alpha\pi} + {\textstyle \frac{1}{3}} \delta_{\alpha\pi})
(\xi D_{\pi\beta} - \Omega_{\pi\beta})
\nonumber\\
&-
2\xi(Q_{\alpha\beta} + {\textstyle\frac{1}{3}}\delta_{\alpha\beta})
Q_{\pi\sigma}W_{\sigma\pi}
\nonumber
\end{eqnarray}
where $D_{\alpha\beta} = \frac{1}{2}(W_{\alpha\beta} + W_{\beta\alpha})$ and
 $\Omega_{\alpha\beta} = \frac{1}{2}(W_{\alpha\beta} - W_{\beta\alpha})$
are the symmetric and antisymmetric contributions to the velocity gradient
tensor and $\xi$ is a material constant representing an effective
molecular aspect ratio.

The thermodynamic contribution to the stress on the fluid can be viewed as
the sum of three parts:
\begin{equation}
\Pi_{\alpha\beta} = \sigma_{\alpha\beta} + \tau_{\alpha\beta}
-
\partial_\alpha Q_{\pi\nu}
\frac{\delta {\cal F}}{ \delta \partial_\beta Q_{\pi\nu}}
\label{equation-lc-stress}
\end{equation}
where $\sigma_{\alpha\beta}$ and $\tau_{\alpha\beta}$ are the symmetric
and antisymmetric contributions from the liquid crystal order:
\begin{equation}
\sigma_{\alpha\beta} =
-p_0 \delta_{\alpha\beta} 
- \xi H_{\alpha\pi}(Q_{\pi\beta} + {\textstyle\frac{1}{3}}\delta_{\pi\beta})
- \xi (Q_{\alpha\pi} + {\textstyle\frac{1}{3}}\delta_{\alpha\pi}) H_{\pi\beta}
+ 2\xi (Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta} )
Q_{\pi\nu} H_{\pi\nu},
\end{equation}
where $p_0$ is the isotropic pressure, and
\begin{equation}
\tau_{\alpha\beta} = Q_{\alpha\pi} H_{\pi\beta} - H_{\alpha\pi} Q_{\pi\beta}. 
\end{equation}
The final term in Equation~\ref{equation-lc-stress} is expanded as
$$
-\kappa_0 \partial_\alpha Q_{\pi\beta} \partial_\nu Q_{\pi\nu}
-\kappa_1 \partial_\alpha Q_{\pi\nu}
\left[ \partial_\beta Q_{\pi\nu} - \partial_\pi Q_{\nu\beta}
+ 2q_0 \epsilon_{\gamma\beta\pi} Q_{\gamma\nu} \right].
$$

\subsubsection{Active stress}

An additional term may be added to the stress to model an active
apolar fluid. This term is
\begin{equation}
\Pi_{\alpha\beta}^a = -\zeta
(Q_{\alpha\beta} - {\textstyle\frac{1}{3}}\delta_{\alpha\beta})
\end{equation}
where $\zeta$ is an activity parameter: $\zeta < 0$ gives
contractile behaviour and $\zeta > 0$ gives extensile behaviour.

\subsection{Implementation}

\subsubsection{Fluid}

The time evolution of the order parameter is computed by a hybrid
approach where the hydrodynamics is in the LB sector, and the
Beris-Edwards Equation is solved in a finite difference approach.
The hydrodynamic quantities and the order parameter share the same
regular lattice. The velocity field, computed via LB, supplies
advective fluxes and the velocity gradient tensor using an appropriate
finite difference stencil.

Coupling to the Navier-Stokes equations is via a body force computed
locally at each lattice site via the divergence of the stress in
Equation~\ref{equation-lc-stress}.

It is also possible to compute solely relaxational dynamics where there
is no flow, and evolution only depends on the molecular field, and the
rotational diffusion constant.

\subsubsection{Solid}

First, we note that the assignment of solid and fluid lattice
nodes for the order parameter follows that for the density:
inside and outside are distinguished using the nominal radius
of the colloid $a_0$ and its position. It is useful, in addition,
to think about a series of control volumes surrounding each lattice
node whose faces are aligned with the lattice
(see Figure~\ref{figure-lc-hybrid}). A set of these
faces constitute the solid-fluid boundary in the hybrid picture.
Advective fluxes of order parameter are computed at the faces
of the control volumes, and the boundary condition is zero
normal flux at solid-fluid interfaces. Note that the colloid
is assumed to be stationary in assigning these fluxes.

\begin{figure}[h]
\begin{center}
\input{hybrid2.epic}
\end{center}
\caption{The hybrid picture. In the lattice
Boltzmann picture (left) a surface is defined by a set of links
$f_b$, which involve discrete vectors $\mathbf{c}_b \Delta t$ which
connect fluid and solid sites. For the order parameter (right), the
colloid is represented by the set of faces, e.g., that between sites
$i,j$ and $i+1,j$ with unit normal $\hat{n}_x$. This is a rather small
colloid.}
\label{figure-lc-hybrid}
\end{figure}

The net hydrodynamic force on the colloid is computed via BBL
in the usual way. As discussed above, the force on the fluid
originating from the order parameter is computed via the
discrete divergence of the stress $\Pi_{\alpha\beta}$. In the
fluid, this is implemented by interpolating $\Pi_{\alpha\beta}$
to the control volume faces and taking differences between faces
in each direction. This method has the advantage that, with the
introduction of solid faces, an extrapolation of $\Pi_{\alpha\beta}$
to the solid-fluid boundary is possible. This allows one to compute
the divergence of the stress at fluid nodes adjacent to the colloid
in the normal way. At the same time, the discrete equivalent of
\begin{equation}
F_\alpha^\mathrm{coll} = \int \Pi_{\alpha\beta} \hat{n}_\beta dS
\end{equation}
may be found by summing $\Pi_{\alpha\beta}$ over the relevant solid-fluid
control volume faces. By construction, this ensures that momentum lost by
the fluid is gained by the colloid, i.e., global momentum is
conserved.

Finally, movement of the colloid across the lattice is accompanied
by changes in its discrete shape. When fluid sites are destroyed,
corresponding order parameter information is also lost. If new
fluid sites are created, new order parameter information may be
added locally either by interpolation from nearby fluid sites, or
from geometrical information from the local surface anchoring.

\subsubsection{Boundary conditions}

\begin{table}[t]
\centering
\tabcolsep=4pt
\begin{tabular}{|c|cccccc|}
\hline
&
$Q_{xx,x}$ & $Q_{xy,x}$ & $Q_{xz,x}$ & $Q_{yy,x}$ & $Q_{yz,x}$ & $Q_{zz,x}$\\
\hline
$Q_{xx}$ &
$\kappa_0 n_x$ & $-\kappa_1 n_y$ & $-\kappa_1 n_z$ & & &\\
$Q_{xy}$ &
$\kappa_0 n_y$ & $\kappa' n_x$ & & $-\kappa_1 n_y$  & $-\kappa_1 n_z$ & \\
$Q_{xz}$ &
$\kappa_0 n_z$ & & $\kappa' n_x$ & & $-\kappa_1 n_y$ &$ -\kappa_1 n_z$\\
$Q_{yy}$ &
 & $\kappa_0 n_y$ & & $\kappa_1 n_x$ & &\\
$Q_{yz}$ &
 & $\kappa_0 n_z$ & $\kappa_0 n_y$ & & $2\kappa_1 n_x$ & \\
$Q_{zz}$ &
 & & $\kappa_0 n_z$ & & & $\kappa_1 n_x$\\
\hline
\hline
&
$Q_{xx,y}$ & $Q_{xy,y}$ & $Q_{xz,y}$ & $Q_{yy,y}$ & $Q_{yz,y}$ & $Q_{zz,y}$\\
\hline
$Q_{xx}$ &
$\kappa_1 n_y$ & $\kappa_0 n_x$ & & & &\\
$Q_{xy}$ &
$-\kappa_1 n_x$ & $\kappa' n_y$ & $-\kappa_1 n_z$ & $\kappa_0 n_x$ & &\\
$Q_{xz}$ &
 & $\kappa_0 n_z$ & $2\kappa_1 n_y$ & & $\kappa_0 n_x$ & \\
$Q_{yy}$ &
 & $-\kappa_1 n_x$ & & $\kappa_0 n_y$ & $-\kappa_1 n_z$ & \\
$Q_{yz}$ &
 & & $-\kappa_1 n_x$ & $\kappa_0 n_z$ & $\kappa' n_y$ & $-\kappa_1 n_z$\\
$Q_{zz}$ &
 & & & & $\kappa_0 n_z$ & $\kappa_1 n_y$\\
\hline
\hline
&
$Q_{xx,z}$ & $Q_{xy,z}$ & $Q_{xz,z}$ & $Q_{yy,z}$ & $Q_{yz,z}$ & $Q_{zz,z}$\\
\hline
$Q_{xx}$ &
$\kappa_1 n_z$ & & $\kappa_0 n_x$ & & & \\
$Q_{xy}$ &
 & $2\kappa_1 n_z$ & $\kappa_0 n_y$ & & $\kappa_0 n_x$ & \\
$Q_{xz}$ &
$-\kappa_1 n_x$ & $-\kappa_1 n_y$ & $\kappa' n_z$ & & & $\kappa_0 n_x$  \\
$Q_{yy}$ &
 & & & $\kappa_1 n_z$ & $\kappa_0 n_y$ & \\
$Q_{yz}$ &
 & $-\kappa_1 n_x$ & & $-\kappa_1 n_y$ & $\kappa' n_z$ & $\kappa_0 n_y$ \\
$Q_{zz}$ &
 & & $-\kappa_1 n_x$ & & $-\kappa_1 n_y $ & $\kappa_0 n_z$\\
\hline
\end{tabular}
\caption{Coefficients of the various derivatives of the order parameter
tensor appearing in six equations for the elements of the
order parameter (including $Q_{zz}$).
Note $\kappa_0 + \kappa_1 = \kappa'$ and all the
coefficients have been multiplied by a factor of 2 in the off-diagonal
equations.}
\label{table-lc-bc}
\end{table}

In three dimensions, the boundary condition Eq.~\ref{equation-lc-general-bc}
provides six equations containing (potentially) 18 unknown derivatives
$\partial_\gamma Q_{\alpha\beta}$
corresponding to the 6 elements of the order parameter
tensor $Q_{xx}$, $Q_{xy}$, $Q_{xz}$, $Q_{yy}$, $Q_{yz}$, $Q_{zz}$.
While the equation corresponding to $Q_{zz}$ must appear to
retain isotropy, $Q_{zz}$ itself, and its derivatives, may be
replaced via the
constraint on the trace of $Q_{\alpha\beta}$. We can therefore either
solve a fully determined system including $Q_{zz}$, and then impose
tracelessness on the result, or replace $Q_{zz}$ and solve six
equations for five unknowns, with the sixth equation acting as the
constraint. These methods provide the same answer for cases where
the surface normal is along the coordinate directions.

In general, Equation~\ref{equation-lc-general-bc} is slightly
cumbersome: the coefficients of the various derivatives for each
of the six equations are shown in Table~\ref{table-lc-bc}. As an
concrete example,
at a flat surface with normal anchoring and, e.g., $n = (1, 0, 0)$,
the number of unknowns (six) are the gradients $\partial_x Q_{\alpha\beta}$
at the boundary --- we assume the tangential gradients
$\partial_y Q_{\alpha\beta}$
and $\partial_z Q_{\alpha\beta}$ can be approximated
using the standard differencing method involving only fluid values of
$Q_{\alpha\beta}$. We proceed by
computing the constant terms relevant for normal anchoring
$$
- \kappa_1 q_0 n_\gamma (\epsilon_{\alpha\gamma\sigma} Q_{\sigma\beta}
+ \epsilon_{\beta\gamma\sigma}Q_{\sigma\alpha})
- w(Q_{\alpha\beta} - Q_{\alpha\beta}^0)
$$
using $Q_{\alpha\beta}$ from the adjacent fluid site. To these
constant terms are added the tangential gradients.
The gradients at the surface are then computed by solving a 6x6
linear algebra problem for $\partial_x Q_{\alpha\beta}$. This
allows the full gradient at the adjacent fluid site to be constructed.

At concave edges or corners, where it is not possible to compute the
tangential gradients from the usual stencil as for a flat interface,
a different approach is required. Here, either a 12$\times$12 or
18$\times$18 system of equations is solved containing the relevant
unknown coefficients from Table~\ref{table-lc-bc} and the relevant
constant terms computed as before.

\vfill
\pagebreak

### 2016-12-21 docs/references.tex
branch documentation updates
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/branches/kevin-1397@3054 40876441-3496-0410-a613-a2fca35b2ae2

@@
 
-\bibitem{adhikari_desplat}
+\bibitem{adhikari-desplat2005}
 Adhikari, R. J.-C. Desplat, and K. Stratford,
@@
 
+\bibitem{allen-tildesley}
+M.P. Allen and D.J. Tildesley,
+\textit{Computer simulation of liquids},
+Oxford University Press (1987).
+
 \bibitem{batchelor}
@@
 
+\bibitem{beris-edwards}
+Beris, A.N., and B.J. Edwards,
+\textit{Thermodynamics of flowing systems with internal microstructure},
+Oxford University Press (1994).
+
 \bibitem{blake}
@@
 \textit{J. Fluid Mech.}, \textbf{46}, 199, 1971.
+
+\bibitem{cahn-hilliard-1958}
+J.W. Cahn and J.E. Hilliard,
+Free energy of a nonuniform system. I. Interfacial free energy,
+\textit{J. Chem. Phys.} \textbf{28} 258--267 (1958).
 
@@
 \textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
+
+\bibitem{chaikin-lubensky}
+P.M. Chaikin and T.C. Lubensky,
+\textit{Principles of condensed matter physics}, Cambridge University
+Press (1995).
 
@@
 \textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
+
+\bibitem{degennes-prost2002}
+de Gennes, J.-G., and J. Prost,
+\textit{The physics of liquid crystals},
+Oxford University Press (2002).
 
@@
 
+\bibitem{lees-edwards1972}
+A.W. Lees and S.F. Edwards,
+The computer study of transport processes under extreme conditions,
+\textit{J. Phys C} \textbf{5} 1921--1929 (1972).
+
 \bibitem{lipanmiller}
@@
 
-\bibitem{r95}
+\bibitem{rapaport1995}
 Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
 Cambridge University Press, 1995.
+
+\bibitem{j-stat-phys-2005}
+K. Stratford, R. Adhikari, I. Pagonabarraga, and J.-C. Desplat,
+Lattice Boltzmann for binary fluids with suspended colloids,
+\textit{J. Stat. Phys.} \textbf{121} 163--178 (2005).
+
+\bibitem{swift1996}
+M.R. Swift, E. Orlandini, W.R. Osborn, and J.M. Yeomans,
+Lattice Boltzmann simulations of liquid-gas and binary fluid systems,
+\textit{Phys. Rev. E} \textbf{54} 5041--5052 (1996).
 
@@
 
+\bibitem{wagner-pagonabarraga2002}
+A.J. Wagner and I. Pagonabarraga,
+Lees-Edwards boundary conditions of lattice Boltzmann,
+\textit{J. Stat. Phys.} \textbf{107} 521--537 (2002).
+
 \bibitem{wardtordai}
@@
 
-\bibitem{wright}
-D.C. Wright and N.D. Mermin,
-\textit{Rev. Mod. Phys.} \textbf{61}, 385 (1989).
+\bibitem{wright-mermin}
+Wright, D.C> and N.D. Mermin,
+\textit{Rev. Mod. Phys.} \textbf{61}, 385--432 (1989).
 

### 2016-12-21 tags/2016-v0.4.12/src/blue_phase_rt.h
Tag of final v0.4.12
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@3055 40876441-3496-0410-a613-a2fca35b2ae2

@@ -1,0 +1,23 @@
+/*****************************************************************************
+ *
+ *  blue_phase_rt.h
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2011 The University of Edinbrugh
+ *
+ *****************************************************************************/
+
+#ifndef BLUE_PHASE_RT_H
+#define BLUE_PHASE_RT_H
+
+#include "field.h"
+
+int blue_phase_rt_initial_conditions(field_t * q);
+void blue_phase_run_time(void);
+
+#endif

### 2016-12-21 tags/2016-v0.4.12/src/coords.c

@@ -1,0 +1,669 @@
+/*****************************************************************************
+ *
+ *  coords.c
+ *
+ *  The physical coordinate system and the MPI Cartesian Communicator.
+ *
+ *  $Id: coords.c,v 1.4 2010-10-15 12:40:02 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2010 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <float.h>
+
+#include "pe.h"
+#include "coords.h"
+
+/* The effective state here is, with default values: */
+
+static int nhalo_ = 1;
+static int ntotal_[3]                  = {64, 64, 64};
+static int periodic[3]                 = {1, 1, 1};
+static int pe_cartesian_size[3]        = {1, 1, 1};
+static MPI_Comm cartesian_communicator = MPI_COMM_NULL;
+static MPI_Comm cart_periodic          = MPI_COMM_NULL;
+static int reorder_                    = 1;
+static int initialised_                = 0;
+
+/* The following lookups will be set by a call to coords_init(), for
+ * convenience, based on the current state: */
+
+static int xfac_;
+static int yfac_;
+static int n_local[3];
+
+static int pe_cartesian_rank             = 0;
+static int pe_cartesian_coordinates[3]   = {0, 0, 0};
+static int pe_cartesian_neighbours[2][3] = {{0, 0, 0}, {0, 0, 0}};
+
+/* Lmin is fixed for all current use. */
+
+static double lmin[3] = {0.5, 0.5, 0.5};
+
+static double radius_ = FLT_MAX;
+
+static void default_decomposition(void);
+static int  is_ok_decomposition(void);
+
+__targetConst__ int tc_nSites;
+__targetConst__ int tc_noffset[3]; 
+__targetConst__ int tc_Nall[3];
+__targetConst__ int tc_ntotal[3];
+__targetConst__ int tc_periodic[3];
+__targetConst__ int tc_nhalo;
+__targetConst__ int tc_nextra;
+
+/*****************************************************************************
+ *
+ *  coords_init
+ *
+ *  Set up the Cartesian communicator for the current state.
+ *  We must check here that the decomposition is allowable.
+ *
[…]

### 2016-12-21 tags/2016-v0.4.12/src/io_harness.c

@@ -1,0 +1,870 @@
+/*****************************************************************************
+ *
+ *  io_harness.c
+ *
+ *  Drivers for serial/parallel IO for lattice quantities.
+ *
+ *  Each quantity (e.g., distributions, order parameter) stored on
+ *  the lattice should set up an io_info struct which tells the
+ *  io_harness how to actually do the read/write.
+ *
+ *  Actual read and writes can be initiated with a call io_read() or
+ *  io_write() with the appropriate io_info struct.
+ *
+ *  Parallel IO takes place by taking a Cartesian decomposition of
+ *  the system which can be the same, or coarser than that of the
+ *  lattice Cartesian communicator. Each IO communicator group so
+ *  defined then deals with its own file.
+ *
+ *  $Id$
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2007-2014 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+
+#include "pe.h"
+#include "util.h"
+#include "coords.h"
+#include "leesedwards.h"
+#include "io_harness.h"
+
+struct io_decomposition_t {
+  int n_io;         /* Total number of I/O groups (files) in decomposition */
+  int index;        /* Index of this I/O group {0, 1, ...} */
+  MPI_Comm comm;    /* MPI communicator for this group */
+  int rank;         /* Rank of this process in communicator */
+  int size;         /* Size of this group in processes */
+  int ngroup[3];    /* Global I/O group topology XYZ */
+  int coords[3];    /* Coordinates of this group in I/O topology XYZ */
+  int nsite[3];     /* Size of file in lattice sites */
+  int offset[3];    /* Offset of the file on the lattice */
+};
+
+struct io_info_s {
+  struct io_decomposition_t * io_comm;
+  size_t bytesize;
+  int metadata_written;
+  int processor_independent;
+  int single_file_read;
+  char metadata_stub[FILENAME_MAX];
+  char name[FILENAME_MAX];
+
+  int (* write_function)   (FILE *, const int, const int, const int);
+  int (* read_function)    (FILE *, const int, const int, const int);
+  int (* write_function_a) (FILE *, const int, const int, const int);
[…]

### 2016-12-21 tags/2016-v0.4.12/src/phi_stats.c

@@ -1,0 +1,236 @@
+/*****************************************************************************
+ *
+ *  phi_stats.c
+ *
+ *  Order parameter statistics.
+ *
+ *  $Id: phi_stats.c,v 1.9 2010-10-15 12:40:03 kevin Exp $
+ *
+ *  Edinburgh Soft Matter and Statistical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *  (c) 2008 The University of Edinburgh
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <float.h>
+#include <math.h>
+#include <stdlib.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "util.h"
+#include "phi_stats.h"
+
+/*****************************************************************************
+ *
+ *  stats_field_info
+ *
+ *****************************************************************************/
+
+int stats_field_info(field_t * obj, map_t * map) {
+
+  int n, nf;
+  MPI_Comm comm;
+
+  double fmin[NQAB];
+  double fmax[NQAB];
+  double fsum[NQAB];
+  double fvar[NQAB];
+  double fvol, rvol;
+  double fbar, f2;
+
+  assert(obj);
+  assert(map);
+
+  field_nf(obj, &nf);
+  assert(nf <= NQAB);
+
+  comm = pe_comm();
+  stats_field_reduce(obj, map, fmin, fmax, fsum, fvar, &fvol, 0, comm);
+
+  rvol = 1.0 / fvol;
+
+  for (n = 0; n < nf; n++) {
+
+    fbar = rvol*fsum[n];                 /* mean */
+    f2   = rvol*fvar[n]  - fbar*fbar;    /* variance */
+
+    info("[phi] %14.7e %14.7e%14.7e %14.7e%14.7e\n", fsum[n], fbar, f2,
+	 fmin[n], fmax[n]);
+  }
+
+  return 0;
+}
+
+/*****************************************************************************
+ *
+ *  stats_field_info_bbl
+ *
+ *  With correction for bbl for conserved order parameters (largely
+ *  for binary fluid).
+ *
+ *****************************************************************************/
+
[…]

### 2016-12-21 tags/2016-v0.4.12/src/psi_force.c

@@ -1,0 +1,583 @@
+/*****************************************************************************
+ *
+ *  psi_force.c
+ *
+ *  Compute the force on the fluid originating with charge.
+ *
+ *  Edinburgh Soft Matter and Statisitical Physics Group and
+ *  Edinburgh Parallel Computing Centre
+ *
+ *  (c) 2013 The University of Edinburgh
+ *
+ *  Contributing authors:
+ *    Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *    Ignacio Pagonabarraga
+ *    Oliver Henrich
+ *
+ *****************************************************************************/
+
+#include <assert.h>
+#include <math.h>
+
+#include "pe.h"
+#include "coords.h"
+#include "control.h"
+#include "physics.h"
+#include "psi_s.h"
+#include "fe_electro.h"
+#include "fe_electro_symmetric.h"
+#include "psi_force.h"
+#include "psi_gradients.h"
+
+static int psi_force_divergence_ = 1;
+
+
+/*****************************************************************************
+ *
+ *  psi_force_gradmu
+ *
+ *  This routine computes the force on the fluid via the gradient
+ *  of the chemical potential.
+ *  Note: The ionic solvation free energy difference is in units 
+ *        of kt and must be dressed for the force calculation.
+ *
+ *****************************************************************************/
+
+int psi_force_gradmu(psi_t * psi, field_t * phi, hydro_t * hydro,
+		map_t * map, colloids_info_t * cinfo) {
+
+  int ic, jc, kc;
+  int in, nk;
+  int ia;
+  int nlocal[3];
+  int index;
+  int xs, ys, zs;       /* Coordinate strides */
+  double rho, rho_elec; /* Species and electric charge density */
+  double e[3];          /* Total electric field */
+  double muphim1, muphip1, musm1, musp1;
+  double phi0;          /* Compositional order parameter */
+  double kt, eunit, reunit;
+  double force[3];
+  /* Cummulative forces for momentum correction */
+  double flocal[4] = {0.0, 0.0, 0.0, 0.0};
+  double fsum[4];
+
[…]

### 2016-12-23 trunk/docs/intro.tex
VERSION 0.6.0
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@3067 40876441-3496-0410-a613-a2fca35b2ae2

--- 2016-06-06 branches/Anthony-nonblocking/docs/intro.tex
@@
 %
-%  (c) 2016 The University of Edinburgh
+%  (c) 2016-2017 The University of Edinburgh
 %
@@
 \end{lstlisting}
-The current release is available as \texttt{ludwig-0.4.0.tar.gz}.
+The current release is available as \texttt{ludwig-0.6.0.tar.gz}.
 Alternatively, registered users can obtain the code via svn checkout
@@
 
-%\paragraph{Parallel: }
 \textbf{Parallel: }
@@
 Output for colloidal particle data is again to file with a name encoding
-the time step and parallel docomposition of the output. For example,
+the time step and parallel decomposition of the output. For example,
 \begin{lstlisting}
@@
 \end{lstlisting}
-contains particle data for time step 20 in a format described ELSEWHERE.
+contains particle data for time step 20 in a format described in
+Section~\ref{examples}.
 These data may be requested in ASCII or raw binary format.

### 2016-12-23 trunk/src/coords_s.h

@@ -1,0 +1,41 @@
+
+#ifndef LUDWIG_COORDS_S_H
+#define LUDWIG_COORDS_S_H
+
+#include "coords.h"
+
+typedef struct coords_param_s cs_param_t;
+
+struct coords_param_s {
+  int nhalo;                       /* Width of halo region */
+  int nsites;                      /* Total sites (incl. halo) */
+  int ntotal[3];                   /* System (physical) size */
+  int nlocal[3];                   /* Local system size */
+  int noffset[3];                  /* Local system offset */
+  int str[3];                      /* Memory strides */
+  int periodic[3];                 /* Periodic boundary (non-periodic = 0) */
+
+  int mpi_cartsz[3];               /* Cartesian size */
+  int mpi_cartcoords[3];           /* Cartesian coordinates lookup */
+
+  double lmin[3];                  /* System L_min */
+};
+
+struct coords_s {
+  pe_t * pe;                       /* Retain a reference to pe */
+  int nref;                        /* Reference count */
+
+  cs_param_t * param;              /* Constants */
+
+  /* Host data */
+  int mpi_cartrank;                /* MPI Cartesian rank */
+  int reorder;                     /* MPI reorder flag */
+  int mpi_cart_neighbours[2][3];   /* Ranks of Cartesian neighbours lookup */
+
+  MPI_Comm commcart;               /* Cartesian communicator */
+  MPI_Comm commperiodic;           /* Cartesian periodic communicator */
+
+  cs_t * target;                   /* Host pointer to target memory */
+};
+
+#endif

### 2016-12-23 trunk/src/field_grad.c

@@ -4,13 +4,14 @@
  *
  *  Gradients (not just "grad" in the mathematical sense) of a field.
  *
- *  $Id$
- *
  *  Edinburgh Soft Matter and Statistical Physics Group and
  *  Edinburgh Parallel Computing Centre
  *
+ *  (c) 2012-2016 The University of Edinburgh
+ *
+ *  Contributing authors:
  *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2012 The University of Edinburgh
+ *  Alan Gray (alang@epcc.ed.ac.uk)
  *
  *****************************************************************************/
 
@@ -32,9 +33,9 @@ static int field_grad_init(field_grad_t * obj);
  *
  *****************************************************************************/
 
-int field_grad_create(field_t * f, int level, field_grad_t ** pobj) {
+__host__ int field_grad_create(field_t * f, int level, field_grad_t ** pobj) {
 
-  field_grad_t * obj =  (field_grad_t*) NULL;
+  field_grad_t * obj =  (field_grad_t *) NULL;
 
   assert(f);
   assert(pobj);
@@ -62,43 +63,54 @@ int field_grad_create(field_t * f, int level, field_grad_t ** pobj) {
 
 static int field_grad_init(field_grad_t * obj) {
 
+  int ndevice;
   int nsites;
+  double * tmp;
 
   assert(obj);
 
-  nsites = le_nsites();
+  nsites = obj->field->nsites;
+  obj->nsite = nsites;
+
+  targetGetDeviceCount(&ndevice);
+
+  if (ndevice == 0) {
+    obj->target = obj;
+  }
+  else {
+    targetCalloc((void **) &obj->target, sizeof(field_grad_t));
+    copyToTarget(&obj->target->nf, &obj->nf, sizeof(int));
+    copyToTarget(&obj->target->nsite, &obj->nsite, sizeof(int));
+    copyToTarget(&obj->target->field, &obj->field->target, sizeof(field_t *));
+  }
 
   if (obj->level >= 2) {
-    obj->grad = (double*) calloc(NVECTOR*obj->nf*nsites, sizeof(double));
-    obj->delsq = (double*) calloc(obj->nf*nsites, sizeof(double));
+
[…]

### 2016-12-23 trunk/src/gradient_2d_tomita_fluid.h

@@ -8,18 +8,16 @@
  *  Edinburgh Parallel Computing Centre
  *
  *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2011 The University of Edinburgh
+ *  (c) 2011-2016 The University of Edinburgh
  *
  *****************************************************************************/
 
 #ifndef GRADIENT_2D_TOMITA_FLUID_H
 #define GRADIENT_2D_TOMITA_FLUID_H
 
-int gradient_2d_tomita_fluid_d2(const int nop, const double * field,double * t_field,
-				double * grad,double * t_grad, double * delsq, double * t_delsq);
+#include "field_grad.h"
 
-int gradient_2d_tomita_fluid_d4(const int nop, const double * field,double * t_field,
-				double * grad,double * t_grad, double * delsq, double * t_delsq);
-
+__host__ int grad_2d_tomita_fluid_d2(field_grad_t * fgrad);
+__host__ int grad_2d_tomita_fluid_d4(field_grad_t * fgrad);
 
 #endif

### 2016-12-23 trunk/src/phi_force.c

@@ -2,6 +2,13 @@
  *
  *  phi_force.c
  *
+ *  TODO:
+ *  This file largely concerns computing the divergence of the
+ *  stress when LE planes are present.
+ *  Other material needs to be refactored. In particular, the
+ *  only routine specific to phi is the phi grad mu calculation.
+ *
+ *
  *  Computes the force on the fluid from the thermodynamic sector
  *  via the divergence of the chemical stress. Its calculation as
  *  a divergence ensures momentum is conserved.
@@ -14,7 +21,7 @@
  *  Edinburgh Parallel Computing Centre
  *
  *  Kevin Stratford (kevin@epcc.ed.ac.uk)
- *  (c) 2011 The University of Edinburgh
+ *  (c) 2011-2016 The University of Edinburgh
  *
  *****************************************************************************/
 
@@ -22,78 +29,32 @@
 #include <stdlib.h>
 #include <math.h>
 
-#include "pe.h"
-#include "coords.h"
-#include "hydro.h"
-#include "field.h"
-#include "leesedwards.h"
-#include "free_energy.h"
-#include "wall.h"
+#include "kernel.h"
 #include "field_grad_s.h"
 #include "field_s.h"
 #include "hydro_s.h"
-#include "phi_force_stress.h"
+#include "pth_s.h"
 #include "timer.h"
+#include "phi_force.h"
+#include "phi_force_colloid.h"
 
-static int phi_force_calculation_fluid(field_t* q, field_grad_t* q_grad, hydro_t * hydro);
-
-static int phi_force_compute_fluxes(double * fe, double * fw, double * fy,
-				    double * fz);
-static int phi_force_flux_divergence(hydro_t * hydro, double * fe,
+static int phi_force_compute_fluxes(lees_edw_t * le, fe_t * fe, double * fxe,
+				    double * fxw,
+				    double * fxy,
+				    double * fxz);
+static int phi_force_flux_divergence(cs_t * cs, hydro_t * hydro, double * fe,
 				     double * fw, double * fy, double * fz);
-static int phi_force_flux_fix_local(double * fluxe, double * fluxw);
[…]

### 2016-12-23 trunk/src/phi_force_colloid.h

@@ -7,18 +7,26 @@
  *  Edinburgh Soft Matter and Statistical Physics Group and
  *  Edinburgh Parallel Computing Centre
  *
- *  (c) The University of Edinburgh (2009)
  *  Kevin Stratford (kevin@epcc.ed.ac.uk)
+ *
+ *  (c) 2009-2016 The University of Edinburgh
  *
  *****************************************************************************/
 
-#ifndef PHI_FORCE_COLLOID_H
-#define PHI_FORCE_COLLOID_H
+#ifndef LUDWIG_PHI_FORCE_COLLOID_H
+#define LUDWIG_PHI_FORCE_COLLOID_H
 
+#include "phi_force_stress.h"
+#include "free_energy.h"
 #include "colloids.h"
 #include "hydro.h"
 #include "map.h"
+#include "wall.h"
 
-__targetHost__ int phi_force_colloid(colloids_info_t * cinfo, field_t* q, field_grad_t* q_grad, hydro_t * hydro, map_t * map);
+__host__ int pth_force_fluid_driver(pth_t * pth, hydro_t * hydro);
+__host__ int pth_force_fluid_wall_driver(pth_t * pth, hydro_t * hydro,
+					 map_t * map, wall_t * wall);
+__host__ int pth_force_colloid(pth_t * pth, fe_t * fe, colloids_info_t * cinfo,
+			       hydro_t * hydro, map_t * map, wall_t * wall);
 
 #endif

### 2016-12-23 trunk/src/stats_symmetric.c

@@ -9,7 +9,7 @@
  *  Edinburgh Soft Matter and Statistical Physics Group and
  *  Edinburgh Parallel Computing Centre
  *
- *  (c) 2011-2013 The University of Edinburgh
+ *  (c) 2011-2016 The University of Edinburgh
  *
  *  Contributing authors:
  *    Kevin Stratford (kevin@epcc.ed.ac.uk)
@@ -37,7 +37,8 @@
  *
  *****************************************************************************/
 
-int stats_symmetric_length(field_grad_t * phi_grad, map_t * map,
+int stats_symmetric_length(fe_symm_t * fe, field_grad_t * phi_grad,
+			   map_t * map,
 			   int timestep) {
 
 #define NSTAT 7
@@ -46,7 +47,6 @@ int stats_symmetric_length(field_grad_t * phi_grad, map_t * map,
   int nlocal[3];
   int status;
 
-  double a, b;                          /* Free energy parameters */
   double xi0;                           /* interfacial width */
 
   double dphi[3];                       /* grad phi */
@@ -61,6 +61,7 @@ int stats_symmetric_length(field_grad_t * phi_grad, map_t * map,
   double eigenvals[3], eigenvecs[3][3];
   double rvolume;
 
+  fe_symm_param_t param;
   MPI_Comm comm;
 
   assert(phi_grad);
@@ -69,9 +70,8 @@ int stats_symmetric_length(field_grad_t * phi_grad, map_t * map,
   coords_nlocal(nlocal);
   comm = pe_comm();
 
-  a = symmetric_a();
-  b = symmetric_b();
-  xi0 = symmetric_interfacial_width();
+  fe_symm_param(fe, &param);
+  fe_symm_interfacial_width(fe, &xi0);
 
   for (ia = 0; ia < NSTAT; ia++) {
     dphi_local[ia] = 0.0;
@@ -118,13 +118,13 @@ int stats_symmetric_length(field_grad_t * phi_grad, map_t * map,
   /* Length scales in coordinate directions, and natural directions */
 
   for (ia = 0; ia < 3; ia++) {
-    lcoordinate[ia] = -4.0*a/(3.0*b*xi0*dphiab[ia][ia]);
+    lcoordinate[ia] = -4.0*param.a/(3.0*param.b*xi0*dphiab[ia][ia]);
   }
 
[…]

### 2016-12-23 docs/binary.tex

@@
 $f_i(\mathbf{r};t)$ given in
-Eq.~\ref{lb-f-moments}, are written
+Eq.~\ref{equation-lb-f-moments}, are written
 \begin{equation}

### 2017-02-14 docs/colloid.tex
Added binary, colloid and EK parts to documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/branches/kevin-1397@3038 40876441-3496-0410-a613-a2fca35b2ae2

--- 2012-12-10 trunk/htdocs/colloid.tex
@@
 %
+%  Largely colloid hydrodynamics and colloid-colloid interactions
+%
+%  Edinburgh Soft Matter and Statistical Physics Group and
+%  Edinburgh Parallel Computing Centre
+%
+%  Contributing authors:
+%  Kevin Stratford (kevin@epcc.ed.ac.uk)
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Colloid Examples}
+\section{Colloid Dynamics}
 
-\subsection{Initialisation from file}
+\label{section:colloids}
 
-We present a number of examples taken from the regression tests, the inputs
-for which are found in
-\begin{verbatim}
-trunk/tests/regression
-\end{verbatim}
-For each test there is an input file, and an ASCII file containing the
-details of the initial colloid state which is read at run time by means
-of the key value pair
-\begin{verbatim}
-colloid_init    from_file
-\end{verbatim}
+\subsection{Hydrodynamics}
 
-A file of colloid information may be created with the utility
-\begin{verbatim}
-util/colloid_file.c
-\end{verbatim}
-As a minimum, this file must specify:
+\label{section-colloid-hydrodynamics}
+
+Having computed the total force and torque on an individual colloid
+by the process of bounce-back-on-links,
+is is possible to update the the linear velocities
+\begin{equation}
+m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
+\end{equation}
+where the mass of the colloid is related to the input radius by
+$m_0 = (4/3)\pi\rho_0 a_0^3$ and the angular
+velocity
+\begin{equation}
+I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
+\mathbf{T}(t),
+\end{equation}
+where the moment of inertia is $I_0 = (2/5) m_0 a_0^2$.
+However, this explicit update is generally found to have poor stability
+properties \cite{l94b, nguyen-ladd2002}. The alternative is to use a
+velocity update
+which is implicit \cite{heemels,nguyen-ladd2002}.
+
+The total force and torque on a colloid can be split into a velocity-dependent
+and a velocity-independent part by combining Equations~(\ref{eq-colloid-ub})
+and~(\ref{eq-colloid-fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
+In this way, one can write
+\begin{equation}
+\left( \begin{array}{rr} \mathbf{F} \\ \mathbf{T}
+\end{array} \right) =
+\left( \begin{array}{rr} \mathbf{F}_0 \\ \mathbf{T}_0
+\end{array} \right) -
+\left( \begin{array}{rr}
+\mathbf{\boldsymbol{\zeta}}^{FU} & \mathbf{\boldsymbol{\zeta}}^{F\Omega} \\
+\mathbf{\boldsymbol{\zeta}}^{TU} & \mathbf{\boldsymbol{\zeta}}^{T\Omega}
+\end{array} \right)
+\left( \begin{array}{rr} \mathbf{U} \\ \mathbf{\Omega}
+\end{array} \right).
+\end{equation}
+%%
+Here, the velocity independent parts of the force and the torque
+(appropriate for a colloid at rest) are
+\begin{eqnarray}
+\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
+\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
+- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
+\\
+\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
+\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
+- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
+(\mathbf{r}_b\times \mathbf{c}_b),
+\end{eqnarray}
+and the matrices $\boldsymbol{\zeta}$ are interpreted as drag
+coefficients and can be written as
+\begin{eqnarray}
+\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
+\\
+\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
+\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
+\\
+\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
+\\
+\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
+\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
+(\mathbf{r}_b \times \mathbf{c}_b).
+\end{eqnarray}
+
+The velocity updates can now be rewritten using an
+implicit (or self-consistent) approach, where the velocity
+at the new time step is used in computing the velocity-dependent
+terms:
+\begin{equation}
+\label{eq-colloid-implicit-update}
+\left( \begin{array}{rr}
+m_0 \mathbf{U}(t + \Delta t) \\ I_0\mathbf{T}(t + \Delta t)
+\end{array} \right) =
+\left( \begin{array}{rr} m_0 \mathbf{U}(t) \\ I_0\mathbf{T}(t)
+\end{array} \right) +
+%%
+\Delta t
+\bigg[
+\left( \begin{array}{rr} \mathbf{F}_0 \\ \mathbf{T}_0
+\end{array} \right) -
+\left( \begin{array}{rr}
+\mathbf{\boldsymbol{\zeta}}^{FU} & \mathbf{\boldsymbol{\zeta}}^{F\Omega} \\
+\mathbf{\boldsymbol{\zeta}}^{TU} & \mathbf{\boldsymbol{\zeta}}^{T\Omega}
+\end{array} \right)
+\left( \begin{array}{rr}
+\mathbf{U}(t + \Delta t) \\ \mathbf{\Omega}(t + \Delta t)
+\end{array} \right)
+\bigg].
+\end{equation}
+
+If the particle has a symmetric distribution of boundary links (the
+special case where the centre of mass is on a symmetry point of the
+lattice) the $\boldsymbol{\zeta}^{FU}$ and
+$\boldsymbol{\zeta}^{T\Omega}$ matrices are
+diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
+and $\boldsymbol{\zeta}^{TU}$ matrices are zero. In general, however,
+this is not the case, and the full drag matrix must be used to form
+the update of the dynamic quantities in which case
+Equation~\ref{eq-colloid-implicit-update} is solved
+as a 6$\times$6 matrix inversion.
+
+Finally, the position of the particle can be updated: an Euler
+forward step is taken using the mean of the old and updated velocity
+\begin{equation}
+\label{eq-colloid-position-update}
+\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + (1/2)\Delta t
+\big[ \mathbf{U}(t) + \mathbf{U}(t + \Delta t) \big].
+\end{equation}
+
+\subsubsection{Rotational motions}
+
+For a number of applications it is useful to update not only
+position, as in Equation~\ref{eq-colloid-position-update},
+but also orientation. This is relevant for magnetic
+dipoles, swimming directions,  and so on. A convenient way to
+implement the required rotations is by considering quaternions.
+
+A quaternion can be thought of as an extended complex number
+\begin{equation}
+q = w + xi + yj + zk
+\end{equation}
+where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
+of a quaternion is
+\begin{equation}
+N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
+\end{equation}
+Addition and subtraction of quaternions proceeds as normal, but
+care is required for multiplication, which is associative but not
+commutative. If the quaternion is written as $q = (w, {\bf v})$ where
+$\bf{v}$ is a normal 3-vector, then multiplication can be written
+\begin{equation}
+q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
++ {\bf v}_1 \times {\bf v}_2),
+\end{equation}
+where the usual scalar and vector products are used. Note that the
+conjugate of a quaternion $q^\star = (w, -{\bf v})$ so that
+$qq^\star = N^2(q)$.
+It can be shown that a rotation of an arbitrary vector though an
+angle $\theta$ around the unit axis of rotation $\hat{\bf w}$ can
+be represented by the quaternion
+\begin{equation}
+q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
+\end{equation}
+which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
+then the rotated vector ${\bf v}^{'}$ can be written as
+\begin{equation}
+(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
+\end{equation}
+which can be expanded as
+\begin{equation}
+{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
+{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
+\end{equation}
+This equation is used to update appropriate vector quantities at
+each time step by a rotation
+$\mathbf{w} = \Delta t \mathbf{\Omega}(t +\Delta t)$,
+where the angular velocity is that determined by the implicit
+procedure described in Section~\ref{section-colloid-hydrodynamics}.
+
+\subsubsection{Changes in particle shape}
+
+One consequence is that the discrete shape of the particle fluctuates
+as the colloid centre moves relative to the
+lattice. This means that the size of the particle as seen by the
+fluid changes, despite the fact that the input radius of the colloid
+is fixed.  However, these fluctuations are small for input radii
+greater than about 5 lattice units \cite{l96a}. Furthermore, the
+fluctuations are greater for some input radii than others
+(leading to preferred values of $a_0$ 1.25, 2.33, and so on).
+
+In the original approach of Ladd \cite{l96a}, changes in the map of
+boundary links are accommodated by the internal fluid. If the particle
+shape changes so that an internal node is exposed, the internal fluid
+at that site rejoins the fluid proper with the solid body momentum
+(assuming the internal fluid is relaxed to solid-body rotation).
+If a fluid node is covered by particle movement, then it simply
+joins the internal fluid, and will relax to solid-body rotation.
+Without internal fluid, changes in particle shape in which fluid nodes
+are either covered or uncovered must be accompanied by the removal or
+addition of fluid with appropriate properties at the nodes in question.
+It is important that this is done in a way which minimises the
+perturbation to the fluid flow around the particle.
+
+If a fluid node at $\mathbf{r}$ is covered by the movement of a
+solid particle, the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$,
+of which an excess $\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$
+must be replaced explicitly so that the overall mean fluid density is
+unchanged. At the same time, the colloid must assume the momentum lost
+by the fluid
+$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
+
+Similarly, when a fluid node is exposed by particle movement,
+fluid must be replaced with the appropriate distribution, density
+and velocity. If the
+new density is $\rho$ (to be determined by some method) then
+the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
+\Delta x^3$ which must be balanced elsewhere. If the new fluid
+is assumed to have velocity $\mathbf{u}$ then the particle must
+give up momentum $\Delta x^3 \rho\mathbf{u}$.
+
+Following Nguyen and Ladd \cite{nguyen-ladd2002}, these changes owing
+to change in particle shape are implemented by adding a small
+correction to the bounce-back at the following time step.
+The excess mass from a covered fluid site is redistributed over all the
+boundary nodes by redefined the momentum transfer at bounce-back
+\begin{equation}
+f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
+- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
++ \frac{w_{c_b} \rho_0\Delta M_c }{A},
+\label{eq:q1}
+\end{equation}
+where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
+force on the particle owing to the change in shape is then
+that lost by the fluid plus the contributions from the final term
+in (\ref{eq:q1}) summed over all the boundary links
+\begin{equation}
+\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
++ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
+\mathbf{c}_b.
+\label{eq:q2}
+\end{equation}
+
+Note that for a closed surface,
+$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
+side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
+does not contribute to the net force and torque on the particle.
+The corresponding change in torque is
+\begin{equation}
+\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
+\times \sum_i f_i \mathbf{c}_i
++ \frac{\Delta x^3}{\Delta t}
+\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
+\label{eq:q2a}
+\end{equation}
+where $\mathbf{r}_c$ is the boundary vector at the position where the
+fluid has been removed. Again, for a closed surface, the
+redistribution of mass does not contribute to the net torque
+on the particle.
+
+Likewise, the contribution to the bounce-back from newly uncovered nodes
+is
+\begin{equation}
+- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
+\label{eq:q3}
+\end{equation}
+leading to a change in force of
+\begin{equation}
+\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
+\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
+\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
+\label{eq:q4}
+\end{equation}
+with a corresponding change in the torque.
+If more than one lattice node is either covered or uncovered by
+the movement of the particle, then these contributions add in a simple
+way. The overall contributions can be added to the
+velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
+Equation~\ref{eq-colloid-implicit-update}.
+
+\subsubsection{Particles near contact}
+
+Particles near contact may not possess a full set of boundary
+links. This leads to a potential
+non-conservation of mass associated with the particle motion
+which must be corrected.
+
+Again following Nguyen and Ladd \cite{nguyen-ladd2002}, mass conservation is
+enforced explicitly using the following procedure. There is a mass
+transfer associated with the bounce-back at each link of
+$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
+The net mass transport into the particle is the sum of this
+quantity over all the links, and can be written
+\begin{equation}
+\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
+\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
+  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
+\end{equation}
+For any particle with a full set of boundary links, both
+$\sum_b w_{c_b} \mathbf{c}_b$
+and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
+However, if the set of boundary links is incomplete
+$\Delta M_s$ is not zero and
+a correction is required. This is again made by redistributing the
+excess or deficit of mass over the other existing boundary nodes.
+The additional contribution to the bounce-back here is
+$-w_{c_b} \rho_0 \Delta M_s/A$, where
+$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
+
+This contribution
+now adds to the velocity-dependent part of the force and torque and
+leads to a slightly different form of the drag matrices
+\begin{eqnarray}
+\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
+\\
+\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
+\overline{\mathbf{r}_b\times\mathbf{c}_b}),
+\\
+\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
+(\mathbf{c}_b - \overline{\mathbf{c}_b}),
+\\
+\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
+\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
+(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
+\end{eqnarray}
+The mean quantities
+\begin{equation}
+\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
+\end{equation}
+and
+\begin{equation}
+\overline{\mathbf{r}_b\times\mathbf{c}_b} =
+\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
+\end{equation}
+A little algebra will show that both occurrences of $\mathbf{c}_b$ and/or
+$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
+can be replaced by their deviation from the mean. This provides a
+convenient way to compute the drag matrices (maintaining symmetry)
+and computing the correct force and torque on the particle when close
+to contact.
+
+\subsection{Colloid-Colloid interactions}
+
+Surface-surface separation between two colloids is denoted
+$h = r_{12} - a_1 - a_2$ where $r_{12}$ is the centre-centre
+separation, and is based on the hydrodynamic radius $a_h$.
+For ease of notation all references to $a$ in this section refer to
+the hydrodynamic radius $a_h$.
+
+\subsubsection{Lubrication corrections}
+
+Hydrodynamic lubrication interactions between two particles may
+not be fully resolved on the lattice if the particles are close
+to contact. The degree to which this is the case depends on the
+size of the particles, and their exact discrete position on the
+lattice. There will be a cut-off separation $h_c$ associated
+with corrections which must be determined by calibration
+\cite{nguyen-ladd2002}; the cut-off is different for the
+different contributions to the interaction.
+
+The force and torque between the pair may be corrected
+by adding back a contribution taken from the known theoretical
+expression for two spheres. Two contributions are handled:
+
 \begin{enumerate}
+\item {\bf Normal force}
+
+The appropriate correction term between colloids with hydrodynamic
+radii $a_1$ and $a_2$, and having separation $h < h_c$, is
+\begin{equation}
+\mathbf{f}_{12} = -6\pi\eta \frac{ a_1^2 a_2^2}{(a_1 + a_2)^2}
+\left[ \frac{1}{h} - \frac{1}{h_c} \right]
+\hat{\mathbf{r}}_{12}. ( \mathbf{U}_1 - \mathbf{U}_2)\,
+\hat{\mathbf{r}}_{12}.
+\end{equation}
+
 \item
-A unique integer id for each particle;
-\item
-a radius and hydrodynamic radius $a_0$ and $a_h$ (use the same value
-if unsure);
-\item
-an initial position ($x_{min} < x < x_{max}$) with $x_{min} = 0.5$
-and $x_{max} = L_x + 0.5$ etc, where $L_x$ is the appropriate system
-size for the problem at hand;
-\item
-the initial velocity etc may be safely initialised to zero.
+{\bf Tangential force}
+
+The corresponding tangential component is a somewhat complex
+expression:
+\begin{equation}
+\mathbf{f}_{12} =
+-(24/15) \pi\eta A
+\big[
+(\mathbf{U}_1 - \mathbf{U}_2)
+-  \hat{\mathbf{r}}_{12} . (\mathbf{U}_1 - \mathbf{U}_2) \hat{\mathbf{r}}_{12}
+\big],
+\end{equation}
+where
+\begin{equation}
+A = \frac{a_1 a_2 (2a_1^2 + a_1 a_2 + 2a_2^2)}{(a_1 + a_2)^3}
+\left[
+\ln\!\left(\frac{a_1 + a_2}{2h}\right)
+- \ln\!\left(\frac{a_1 + a_2}{2h_c}\right)
+\right].
+\nonumber
+\end{equation}
 \end{enumerate}
-Note that is an inter-particle potential is to be specified at run time,
-the initial position of the particles should not be so close that a
-large force is experienced at time $t=0$. This can destabilise the
-dynamics.
 
-A single file of colloid information is produced in either ASCII or binary
-format, which can be read in by specifying the appropriate
-\begin{verbatim}
-colloid_io_format_input   ASCII_SERIAL
-\end{verbatim}
-or \texttt{BINARY\_SERIAL} key value in both serial and parallel.
+In the presence of isothermal fluctuations, there is an additional
+consideration related to the dissipation related to the correction
+itself. This adds an additional random force with a variance
+$(2k_BT |\mathbf{f}_{12}|)^{1/2}$ to each lubrication component.
 
-\subsubsection{Very short range potential}
+We do not implement the collective implicit update required for
+stability when many particles are in mutual lubrication contact
+\cite{nguyen-ladd2002}. This treats the velocity-dependent
+corrections as effectively velocity independent and not as part
+of the implicit update.
 
-See
-\begin{verbatim}
-tests/regression/test_spin_solid2_input
-tests/regression/config.cds.init.001-001
-tests/regression/test_spin_solid2_d3q19.ref*
-\end{verbatim}
+This means that mutual contact must be prevented
+via a repulsive potential of sufficient strength to keep particles
+from close approach (the update may become unstable if not). For
+many purposes it may then be preferable to ignore the lubrication
+corrections completely. Long-range lubrication ($h > h_c$) will
+still be well represented by the lattice fluid.
 
-An example of a moderate volume fraction of particles is given for
-a binary fluid undergoing spinodal decomposition. Here,
-the capillary interaction between the neutrally wetting colloids
-causes a strong effective attraction between particles. It is therefore
-necessary to ensure the particles do not collide to the point of
-overlapping at their hard-sphere radius. A counterbalancing short-range
-soft-sphere potential is then defined in the input file:
-\begin{verbatim}
-soft_sphere_on 1
-soft_sphere_epsilon 0.0004
-soft_sphere_sigma 0.1
-soft_sphere_nu 1.0
-soft_sphere_cutoff 0.25
-\end{verbatim}
-where, following Eq.~\ref{eq_ss_shift}, we have
-$v(r) \sim \epsilon (\sigma /r)^\nu$ ``cut-and-shifted'' so that
-both potential and force smoothly match to zero at the cut-off
-distance $r_c$ (here $0.25\Delta x$). The energy scale $\epsilon$
-will need to be adjusted depending on the exact problem at hand
-(here it will be related to the fluid-fluid interfacial tension).
+\subsubsection{Hard sphere interaction}
 
-Other relevant parameters here are
-\begin{verbatim}
-colloid_cell_min 8.0
-lubrication_on 0
-\end{verbatim}
-which control, respectively, the minimum cell list width used to
-help identify interactions, and the lubrication correction (here
-switched off).
+Particles do not overlap, i.e., there is always
+a hard-sphere interaction which is a function of the separation $h$:
+\begin{equation}
+v^{hs}(h) = \left\{
+\begin{array}{ll}
+\infty & h \leq 0,\\
+0   & h > 0.
+\end{array} \right.
+\end{equation}
+The hard-sphere interaction does not give rise to a force, but gives rise
+to program termination if $h<0$.
 
-\subsubsection{Short-range potential}
+\subsubsection{Soft-sphere interaction}
 
-See
-\begin{verbatim}
-tests/regression/test_yukawa_input
-tests/regression/test_yukawa_cds.001-001
-tests/regression/test_yukawa_d3q10.ref1
-\end{verbatim}
+\label{section-colloids-soft-sphere}
 
-This example involves a number of interacting particles in a simple
-fluid including Brownian motion. The potential is Yukawa-like,
-representative of a screened Coulomb interaction.
+A simple soft-sphere interaction is available, the basic form
+of which is:
+\begin{equation}
+v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
+\end{equation}
+where $\epsilon$ sets the energy scale, and $\sigma$ sets the
+characteristic width. The steepness of the potential is set by
+the exponent $\nu (> 0)$.
 
-We have the parameters:
-\begin{verbatim}
-yukawa_on 1
-yukawa_epsilon  1.330764285
-yukawa_kappa 0.72463768115
-yukawa_cutoff 16.0
-\end{verbatim}
-where the potential is $v(r) \sim \epsilon (-\kappa r) / r$. Again,
-both potential and force are smoothly matched to zero at the
-cut-off distance $r_c$ (here $r_c = 16\Delta x$). The relatively
-long cut-off distance means that the minimum cell list cell width
-must be correspondingly large (16 lattice units) which limits the
-minimum parallel domain size.
+To prevent the need for calculation of long-range interactions,
+the soft-sphere potential is truncated in a ``cut-and-shift''
+approach. This is done in such a way as to smoothly match
+both the potential and the force at the cut-off distance
+$r_c$. For $r < r_c$ the potential is then
+\begin{equation}
+v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
+\label{eq_ss_shift}
+\end{equation}
+Clearly, this potential is not exactly what we first thought of as
+soft-sphere. Matching the potential smoothly ensures conservation
+of energy, while matching the force smoothly prevents potential
+instabilities in the molecular dynamics update.
 
-This example also includes Brownian motion by means of fluctuating
-hydrodynamics. The relevant parameters in the input are:
-\begin{verbatim}
-isothermal_fluctuations on 
-temperature 0.0002133333
-\end{verbatim}
-Note that the \texttt{temperature} parameter here specifies
-$k_BT = \langle c_x^2\rangle = \langle c_y^2 \rangle = \langle c_z^2 \rangle$
-in three dimensions (at equilibrium);
-this is reported in the output of the code at run time.
+The soft-sphere potential can be useful as a mechanism to
+keep the particles from touching (which risks hard-sphere interactions),
+in which case is is possible to compute the interaction as a function
+of the separation $h$, instead of $r$. The force between
+two particles is computed via the derivative of equation~\ref{eq_ss_shift}
+with respect to $r$:
+\begin{equation}
+\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
+\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
+\end{equation}
+where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
+particle $i$ to the centre of $j$.
 
-\subsubsection{Long range forces}
+\subsubsection{Lennard-Jones interaction}
 
-An Ewald sum is available for magnetic dipoles.
+\label{section-colloids-lennard-jones}
+
+A cut-and-shifted Lennard-Jones interaction is available based on the
+centre-centre separtion of pairs. The basic form
+of the potential is
+\begin{equation}
+v^{lj}(r) = 4\epsilon
+\left[ \left(\frac{\sigma}{r}\right)^{12} - 
+\left( \frac{\sigma}{r} \right)^6 \right]
+\end{equation}
+where $\epsilon$ sets the energy scale, and $\sigma$ is a characteristic
+width. The cut-and-shift of both the potential and the associated force
+is as described in the previous section.
+
+If using the Lennard-Jones potential, some care may be required to
+ensure that no hard-sphere overlaps occur at the level of the
+hydrodynamic radius $a_h$.
+
+\subsubsection{Dipole-dipole interactions}
+
+For particles with magnetic dipoles $\mathbf{m}_i$, long-range
+dipole-dipole interactions may be computed using a standard
+Ewald sum for a periodic system
+(e.g., \cite{allen-tildesley}, \cite{rapaport1995}).
+We describe this here only briefly.
+
+We consider particles with uniform dipoles
+$\mathbf{m}_i = \mu \mathbf{s}_i$ where $\mu$ is the dipole
+strength and $\mathbf{s}_i$ is
+a unit vector describing the orientation of the dipole. The
+basic dipole-dipole interaction is then
+\begin{equation}
+v^{dd} (r) = \frac{\mu_0 \mu^2}{4\pi}
+\big( \mathbf{s}_i . \mathbf{s}_j
+- 3(\mathbf{s}_i.\mathbf{r}_{ij})( \mathbf{s}_j .\mathbf{r}_{ij})
+\big),
+\end{equation}
+where $\mu_0$ is the permeability. The Ewald sum decomposes this
+into a short-range part contributing forces and torques computed
+in real space and reciprocal-space part computed as a sum over
+reciprocal space vectors. The choice of the real-space cut-off
+$r_c$ determines the number of reciprocal space vectors required
+in the sum.
+
+A relevant parameter for magnetic colloid systems is the
+dipolar coupling constant which measures the strength of
+the dipole-dipole interaction compared with the thermal
+energy $k_bT$ (which may be introduced via isothermal
+fluctuations in the fluid). If we take $\mu_0 / 4\pi$ to be
+unity in lattice units, the the dipolar coupling constant
+is
+\begin{equation}
+\lambda = \mu^2 / 8k_bT a^3.
+\end{equation} 
+\subsection{External Forces}
+
+\subsubsection{Gravity}
+
+For the purposes of sedimentation and other driven motions, the code
+allows a uniform gravitational acceleration on colloids which
+may be written as $\mathbf{g} = (g_x, g_y, g_z)$. We will consider the
+simple situation where $\mathbf{g} = (0, 0, -g)$, i.e., the
+negative $z$-direction is ``downwards.''
+
+A colloid of density $\rho$ sedimenting in this gravitational field
+experiences a net buoyancy force density $(\rho-\rho_0)g = \Delta\rho g$,
+where $\rho_0$ is the fluid density. (If the colloid and fluid density
+are equal, one can still think in the limit that $\Delta\rho \rightarrow 0$
+and $g \rightarrow \infty$, but where the product $\Delta\rho g $ is finite.)
+
+In a fully periodic system, the application of the force on the colloids
+will cause a continuous and unboundaed acceleration. To prevent this, and
+to balance the momentum budget,
+an equal an opposite body force density is applied uniformly to fluid
+sites at each time step. In a system with a ``bottom'' (for example, a
+flat wall perpendicular to the gravitational force) no such correction
+is required.
+
+\subsubsection{Magnetic field}
+
+The the presence of a uniform external magnetic field $\mathbf{B}$,
+a particle with dipole $\mathbf{s}$ experiences a torque
+$\mathbf{s} \times \mathbf{B}$.
+
+\subsubsection{Electric Field}
+
+In the presence of a uniform external electric field $\mathbf{E}$, a
+particle with charge $q$ experiences a body force $q\mathbf{E}$.
+
+% TODO
+% Yukawa
+% Bonds, angles, etc
+% Squirmer
+% Sub grid dynamics
+% solid-solid links
+
+% End section
+\vfill
+\pagebreak

### 2017-02-14 docs/devel.tex

--- 2016-01-12 branches/kevin-1397/docs/devel.tex
@@
 %
-%  (c) 2014 The University of Edinburgh
+%  (c) 2014-2017 The University of Edinburgh
 %
@@
 different threaded models. It is only available in the development version.
+
+\subsubsection{Array address models}
+
+In a related issue, the code allows different data layout to be used
+to improve performance. The default layout is array of structures (AOS)
+which is suitable for most CPU architectures. (A blocked version AOSOA
+is also available which may promote vectorisation.) For GPU machines, a
+structure of arrays (SOA) format can be selected at compile time.
 
@@
 
+\subsubsection{Address models}
+
+The default memory layout is array of structures (AOS). Compiler
+preprocessor macros for blocked SOA (\texttt{-DADDR\_AOSOA}) or
+structure of arrays (\texttt{-DADDR\_SOA})
+may be appropriate for different platforms.
+
+Add the relevant flags to the \texttt{config.mk} file before
+compilation to effect the given address model.
+
 \subsubsection{Targets for serial and parallel code}

### 2017-02-14 docs/develop.tex

--- 2015-11-05 trunk/docs/develop.tex
@@
 %
-%  (c) 2014-2015 The University of Edinburgh
+%  (c) 2014-2017 The University of Edinburgh
+%
+%  Contributing authors:
+%  Kevin Stratford (kevin@epcc.ed.ac.uk)
 %
@@
 The code allows for different order of storage associated with 3-dimensional
-fields (scalars, vectors, and so on). For historical reasons these are
-referred to as `Model' and `Model~R' options, which correspond
+fields (scalars, vectors, and so on). These storage schemes are usually
+referred to as array of structures (AOS), structure of arrays (SOA), and
+for the blocked version array of structures of (short) arrays (AOSOA).
+For historical reasons these are
+also sometimes referred to as `Model' and `Model~R' options, which correspond
 to array-of-structures and structure-of-arrays layout, respectively. Early
@@
 
-\textbf{ADDR\_MODEL}: The array-of-structures or Model order for an
+\textbf{ADDR\_AOS}: The array-of-structures order for an
 $n-$vector field
@@
 
-\textbf{ADDR\_MODEL\_R}: The Model~R version is:
+\textbf{ADDR\_SOA}: The structure-of-arrays version is:
 \[
@@
 
-\textbf{ADDR\_MODEL}:
+\textbf{ADDR\_AOS}:
 A general rank 2 tensor $t_{\alpha\beta}$ with components $(t_{0,0},
@@
 
-\textbf{ADDR\_MODEL\_R}: The Model R version is
+\textbf{ADDR\_SOA}: The structure-of-arrays version is
 \[
@@
 generalises from the above, i.e., with the rightmost index running
-fastest. Diagrams are omitted, but Model and Model R storage patterns
+fastest. Diagrams are omitted, but AOS and SOA storage patterns
 follow the same form as seen above.
@@
 
-\textbf{ADDR\_MODEL}: For the LB distributions, up to two distinct
+\textbf{ADDR\_AOS}: For the LB distributions, up to two distinct
 distributions can be accommodated to allow for a binary fluid
 implementation, although the usual situation is to have only one.
-The Model order for two $N$-velocity distributions $f$ and $g$ is:
+The AOS order for two $N$-velocity distributions $f$ and $g$ is:
 \[
@@
 
-\textbf{ADDR\_MODEL\_R}: The Model R order is:
+\textbf{ADDR\_SOA}: The strucuture-of-arrays order is:
 
@@
 
-The Model R picture, which targets coalescene, can be viewed as
+The SOA picture, which targets coalescence, can be viewed as
 naturally supporting SIMD vectorisation by virtue of allowing
@@
 index. (If SIMD vectorisation is wanted at all on GPU architecture,
-it can be as a means to releive register pressure.) Model R therefore
-remains unaltered and we concentrate on the Model picture.
+it can be as a means to relieve register pressure.) SOA therefore
+remains unaltered and we concentrate on the AOS picture.
 
@@
 
-\textbf{VADDR\_MODEL}: The structure of short vectors is based on the
+\textbf{ADDR\_AOSOA}: The structure of short vectors is based on the
 SIMD vector length which we here denote $V$:
@@
 
+\subsubsection{Compilation}
+
+A number of options exist to control at compilation time the choice
+of the address model. If no options are specified the default
+addressing model is AOS. Two additional C preprocessor directives
+exist:
+
+\texttt{-DADDR\_SOA} selects structure-of-arrays order (e.g., for
+coallescing on GPU platforms).
+
+\texttt{-DADDR\_AOSOA} block SOA pattern (usually combined with
+a non-unit vector length).
+
+For development purposes, the address functions allow the array indices
+to be checked against the correct range. For production runs, the
+relevant functions are replaced by macros if \texttt{-DNDEBUG} is
+specified.
+
+The targetDP vector length is specified via, e.g.,  \texttt{-DVVL=4}.
+
+\subsubsection{Kernels}
+
 To provide a transparent interface for addressing vector fields,
@@
 specific to the object at hand. To allow transparent switching
-between Model and Model R versions, the indexing must be abstracted.
+between different addressing models, the indexing must be abstracted.
+The abstraction is provided in \texttt{memory.h}, where functions
+(or function-like macros) are provided to make the appropriate
+address computation.
 As a concrete example, the following considers a rank 1 3-vector:
@@
 
-An equivalent vectorised version may be constructed as follows:
+This will usually be carried out in computational kernels, where the
+\texttt{kernel.h} interface can be used to help to manage the
+one-dimensional loop structure required the the targetDP threading
+model. In this way, the kernel launch stage will specify the limits
+of the kernel in terms of local $x$-coordinates, $y$-coordinates, and
+$z$-coordinates in the usual way. The kernel itself is then written
+in terms of a one-dimensions kernel index.
+
+In the following schematic, the launch stage defines a kernel context
+to be passed to the kernel along with the other actual arguments:
 \begin{lstlisting}
-for (n = 0; n < nsites; n += SIMDVL) {
-  index = coords_indexv(n, nextra);       /* Proposal */
+  kernel_info_t limits;
+  kernel_ctxt_t * ctxt;
+
+  limits.xmin = 1; limits.xmax = nlocal[X];
+  limits.ymin = 1; limits.ymax = nlocal[Y];
+  limits.zmin = 1; limits,zmax = nlocal[Z];
+
+  kernel_ctxt_create(NSIMDVL, limits, &ctxt);
+
+  __host_launch(kernel_function, nblk, ntpb, ktx, ...);
+\end{lstlisting}
+
+An suitable  vectorised kernel may be constructed as follows with the
+appropriate base index for the loop computed via the utility
+\texttt{kernel\_baseindex()}:
+\begin{lstlisting}
+for (kindex = 0; kindex < kiterations; kindex += SIMDVL) {
+  index = kernel_baseindex(ctxt, kindex);
   for (ia = 0; ia < 3; ia++) {
     for (iv = 0; iv < SIMDVL; iv++) {
-      iref = vaddr_rank1(nsites, 3, index, ia, iv);
+      iref = addr_rank1(nsites, 3, index + iv, ia);
       array[iref] = ...;
@@
 \end{lstlisting}
-Note that a different function \texttt{vaddr\_rank1()} with an additional
+Note that the same function \texttt{addr\_rank1()} with the 
 argument which is the index of the vector loop is used. Note also that
-vectorisaed versions must take into account the extent that the kernel
-extends into the halo region (here \texttt{nextra}), which involves
-logic to avoid lattice sites which are not required (not shown in this
+vectorised versions must take into account the extent that the kernel
+extends into the halo region, which involves
+logic to avoid lattice sites which are not required (this type of
+masking operation is not shown in this
 example).
-
-Note that, in practice, 
-the loop over lattice sites is further abstracted in the code to
-accommodate thread-level parallelism via targetDP interface.
 

### 2017-02-14 docs/electrokinetic.tex

--- 2013-01-31 htdocs/electrokinetic.tex
@@
 \clearpage
-
-\subsubsection{Scaling of the SOR}
-
-We have run some simple scaling tests for the electrokinetic problem
-with zero fluid velocity. A system of $128^3$ lattice sites was used
-as representative of a small production system. The code was run on
-the Mare Nostrum machine in Barcelona for a small number of time
-steps to assess performance on up to 512 MPI tasks. The results are
-shown in Fig. \ref{fig5}.
-
-While the scaling for the Nernst Planck part of
-the calulcation are reasonable, it can be seen that the SOR routine
-used to solve the Poisson equation is not really acceptable. This
-was not unexpected. Some improvements might be possible, but the
-algorithm is essentailly unsuitable for large systems.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.9\textwidth]{./pics/scaling.pdf}
-\caption{Scaling of a system of $128^3$ lattice sites on up to 512 MPI tasks. 
-While the scaling for the Nernst Planck part of the calulcation are reasonable, 
-it can be seen that the SOR routine used to solve the Poisson equation deviates
-strongly from ideal behaviour.} 
-\label{fig5} 
-\end{figure}
-\clearpage
-
-\subsection{User input}

### 2017-02-14 docs/examples.tex

--- 2014-09-15 tags/2014-pre-release/htdocs/porous.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
-%  porous.tex
+%  examples.tex
 %
-%  Some considerations for porous media.
-%
-%  $Id$
+%  Some worked examples:
+%  1. Conductance in simple porous media
 %
@@
 %
+%  (c) 2010-2017 The University of Edinburgh
+%
+%  Contributing authors:
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2011 The University of Edinburgh
 %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Porous Media}
+\section{Examples}
 
-Porous media calculations can be undertaken when appropriate
-solid/fluid status information is supplied. The are a number
-of switches available in the input file:
+\subsection{Permeability calculations in simple porous media}
 
-\inputkey{porous\_media\_file}
-
-specifies the file stub name to be read at the start of execution.
-(If the stub name is \texttt{file} then the code will expect to
-find \texttt{file.001-001} in the current directory.
-
-\inputkey{porous\_media\_format}
-
-is either \texttt{ASCII} or \texttt{BINARY} as appropriate. Note that
-in parallel, a single data file can be supplied, but it must be binary.
-The default is \texttt{BINARY}.
-
-\inputkey{porous\_media\_type}
-
-is either \texttt{status\_only} (the default) or \texttt{status\_with\_h}.
-
-\subsection{File format}
-
-In all cases the status (fluid/solid) information is represented by a
-single \texttt{char} (or integer), which must be supplied via the
-porous media file. A single file should contain data matching the
-current system size, and have the $z-$direction running fastest,
-followed by the $y-$direction. Note that in non-periodic directions,
-the structure must be 'closed', i.e., all the points at the edge
-should be solid.
-
-Fluid sites are designated by \texttt{0} and boundary or solid sites
-by \texttt{1}. These data should be of type \texttt{char} in binary,
-and may be integer in ASCII.
-
-Where wetting information is required, the free energy parameter $H$
-can be supplied by using the \texttt{status\_with\_h} switch. In this
-case, the \texttt{char} status is augmented by a single \texttt{double}
-value which is the local value of $H$. The order is then
-$s_1,h_1, s_2, h_2, \ldots$.
-
-An example of how to construct a porous media file is provided in
-\texttt{util/capillary.c}, which builds an appropriate file for
-a square or circular capillary tube. Please see the comments in
-the file for further details. Note that the allowed
-solid/fluid status values are defined in \texttt{src/site\_map.h}.
-A solid boundary is \texttt{BOUNDARY}, while fluid is \texttt{FLUID}.
-
-\subsection{Permeability calculations}
-
-Single fluid permeability calculations for a given proous structure
+Single fluid permeability calculations for a given porous structure
 can be undertaken by driving a flow via the fluid body force. Note
@@
 that, e.g., to drive a flow in the positive $x-$direction
-
-\texttt{force 0.001\_0.000\_0.000}
-
+\begin{lstlisting}
+force 0.001_0.000_0.000
+\end{lstlisting}
 The force should not be so large that the maximum velocity generated
@@
 The net flow can be measured by combining the statistics for the
-total momemtum and the total density (which is equal to the volume
-with $\rho_0 = 1$).
+total momemnum and the total density (which is equal to the volume
+with $\rho_0 = 1$). We will see an example of this later in the
+section.
 
@@
 corrects the viscosity-dependence of the apparent boundary position
-using a three relaxation time scheme. This is currently under
-investigation.
+using a three relaxation time scheme.
 
-\subsection{Circular and rectangular capillaries}
+Before considering a concrete example, it's worth reviewing the basic
+calculation of the conductance in ideal geometries.
 
-\label{section:exact_conductance}
+\subsubsection{Circular and rectangular capillaries}
+
+\label{section-examples-exact-conductance}
 
@@
 
-An analtyical expression is also available for the conductance
-of a square or rectugular capillary. For an infinite capillary
+An analytical expression is also available for the conductance
+of a square or rectangular capillary. For an infinite capillary
 of square cross section width $w \times h$ (where $h$ is the
@@
 $-\partial p / \partial x$ is replaced by the body force $\rho g$ and
-one can difine the viscosity-independnet conductance $C$ via
-$J = C\rho g / \eta$. The calculation procedes as above.
+one can difine the viscosity-independent conductance $C$ via
+$J = C\rho g / \eta$. The calculation proceeds as above.
 
-\subsection{Worked example}
-
-\subsubsection{Setting up the structure}
+\subsubsection{Setting up a structure}
 
@@
 set
-\begin{verbatim}
+\begin{lstlisting}
 const int xmax = 10;
@@
 const int zmax = 32;
-\end{verbatim}
+\end{lstlisting}
 Note that the conductance should be independent of the length of the
@@
 We choose a circular cross section via:
-\begin{verbatim}
-enum {CIRCLE, SQUARE};
+\begin{lstlisting}
+enum {CIRCLE, SQUARE, XWALL, YWALL, ZWALL, XWALL_OBSTACLES};
 const int xsection = CIRCLE;
-\end{verbatim}
+\end{lstlisting}
 For this problem, the wetting parameters are irrelevant, and can be
 ignored. In addition we set
-\begin{verbatim}
-enum {STATUS_ONLY, STATUS_WITH_H};
+\begin{lstlisting}
+enum {STATUS_ONLY, STATUS_WITH_H, STATUS_WITH_SIGMA};
 const int output_type = STATUS_ONLY;
-\end{verbatim}
+\end{lstlisting}
 to specify that the output will contain only the structural information.
 The output filename containing the structure is set via
-\begin{verbatim}
+\begin{lstlisting}
 const char * filename = "capillary.001-001";
-\end{verbatim}
+\end{lstlisting}
 
 The program may be compiled and run via
-\begin{verbatim}
-$ gcc capillary.c -lm
-$ ./a.out
-\end{verbatim}
+\begin{lstlisting}
+$ make capillary.c -lm
+$ ./capillary
+\end{lstlisting}
 The output provides a simple representation of the cross section,
@@
 case we have
-\begin{verbatim}
+\begin{lstlisting}
 Cross section (0 = fluid, 1 = solid)
@@
 n = 3200 nsolid = 1536 nfluid = 1664
-\end{verbatim}
+\end{lstlisting}
 Note that the outermost sites in each direction here are solid. There
@@
 We should set the system size
-\begin{verbatim}
+\begin{lstlisting}
 size 10_10_32
-\end{verbatim}
+\end{lstlisting}
 The location of the porous media file is specified using
-\begin{verbatim}
+\begin{lstlisting}
 porous_media_format BINARY
@@
 porous_media_type   status_only
-\end{verbatim}
+\end{lstlisting}
 Note that the extension \texttt{.001-001} of the filename has been
@@
 we must set
-\begin{verbatim}
+\begin{lstlisting}
 free_energy none
-\end{verbatim}
+\end{lstlisting}
 What are the fluid parameters? We will set the viscosity to
@@
 the shear viscosity):
-\begin{verbatim}
+\begin{lstlisting}
 viscosity 0.16666666666666666
 #bulk_viscosity 0.1
-\end{verbatim}
+\end{lstlisting}
 We expect that the number of time steps to reach a steady state
@@
 $a=4$ so $t \approx 100$ LB time steps. We will try
-\begin{verbatim}
+\begin{lstlisting}
 N_cycles 200
-\end{verbatim}
-Note that we will need output on the total momentum of the system
+\end{lstlisting}
+Note that we will need output on the total momentum and the volume
+flux of the system
 as a function of time, so we set
-\begin{verbatim}
+\begin{lstlisting}
 freq_statistics 50
-\end{verbatim}
+\end{lstlisting}
 
-Finally, we need to set a force in the $z-$direction to drive the
-flow. Again, the final conductance should be indpendent of this
+Finally, we need to set a force in the $z$-direction to drive the
+flow. Again, the final conductance should be independent of this
 force providing the force is small enough that both the Mach number
@@
 We will set
-\begin{verbatim}
+\begin{lstlisting}
 force 0.00_0.00_0.0000001
-\end{verbatim}
+\end{lstlisting}
 
@@
 \texttt{capillary.c} program:
-\begin{verbatim}
+\begin{lstlisting}
 Scalars - total mean variance min max
 [rho]    1664.00  1.00000000000-2.2204460e-16  1.00000000000 1.00000000000
-\end{verbatim}
+\end{lstlisting}
 Note that the total density (ie., mass) should remain exactly unchanged
@@
 
-The total momentum of the fluid in each coordinate direction is
+The integrated volume flux of the fluid in each coordinate direction is
 reported, and we should see, as a function of time
-\begin{verbatim}
-Momentum - x y z
+\begin{lstlisting}
+Velocity - x y z
 ...
-[fluid   ]  1.5321078e-14  1.8207658e-14  1.8529405e-03
-[fluid   ] -8.2378548e-14  2.8865799e-15  1.9422398e-03
-[fluid   ]  1.1302070e-13  6.1950445e-14  1.9467213e-03
-[fluid   ]  2.4091840e-14  1.1102230e-16  1.9469462e-03
-\end{verbatim}
+[vol flux]  3.1086245e-15  2.4868996e-14  1.9303426e-03
+[vol flux] -6.8278716e-14  4.8849813e-15  2.0251488e-03
+[vol flux]  8.4265928e-14  5.5178084e-14  2.0299067e-03
+[vol flux]  1.6209256e-14  1.2212453e-14  2.0301454e-03
+\end{lstlisting}
 Note as we have applied an external force in the $z-$direction,
-the $z-$momentum increases with time while the momenutm in the
-other two directions is constant (to machine accuracy, which is
-around $10^{-16}$). Further, the $z-$momentum
+the $z$-flux increases with time while the volume flux in the
+other two directions is constant (zero to machine accuracy, which is
+around $10^{-16}$). Further, the $z$-volume flux
 is still changing slightly after 200 time steps, so we have
 not run long enough to reach a steady state. If we run for
-400 steps, we should see that the $z-$momentum is unchanged
+400 steps, we should see that the $z$-flux is unchanged
 over the last 100 time steps.
-\begin{verbatim}
-Momentum - x y z
+\begin{lstlisting}
+Velocity - x y z
 ...
-[fluid   ]  1.0424994e-13  1.6098234e-14  1.9469580e-03
-[fluid   ]  3.2973624e-14  3.2862602e-14  1.9469580e-03
-\end{verbatim}
+[vol flux]  6.6391337e-14 -2.2204460e-15  2.0301580e-03
+[vol flux]  3.4638958e-14  3.6415315e-14  2.0301580e-03
+\end{lstlisting}
 Finally, note the maximum velocity in the flow direction
@@
 
-The figure we are interested in is the total momentum in the
-$z-$direction which is $1.947\times 10^{-3}$. As the mean
-density is unity, this is also the average total volume flux $J_v$.
+The figure we are interested in is the volume flux in the
+$z$-direction which is $2.030\times 10^{-3}$. As the mean
+density is unity, this is also the integrated volume flux $J_v$.
 The volume flux per unit area $J = J_v / a^2 L_z$, where $a^2 L_z$
 is the volume of the discrete system (1664 in lattice units).
-Following section \ref{section:exact_conductance},
+Following section \ref{section-examples-exact-conductance},
 we can write a viscosity independent conductance
 $C = J\eta / \rho g$, with $g$ the force in the $z-$direction.
-Putting all this together, we have $C = (1.947\times 10^{-3} / 1664)
-\times (1/6) / (1.0 \times 1\times 10^{-7}) = 1.950$. The theoretical
+Putting all this together, we have $C = (2.030\times 10^{-3} / 1664)
+\times (1/6) / (1.0 \times 1\times 10^{-7}) = 2.033$. The theoretical
 figure is $C = 2$, so the simulation is correct to within a discretisation
-error of about 3\%. Better accuracy may be acheived by increasing the
-resolution of the circle; e.g., for $a = 8$, the result is 8.043 versus
-an exact result of $C = 8$ --- an error of around 0.5\%.
+error of about 2\%.
 
@@
 with \texttt{make mpi}). In the input file we set
-\begin{verbatim}
+\begin{lstlisting}
 size 10_10_32
 grid 1_1_2
-\end{verbatim}
+\end{lstlisting}
 to set the parallel decomposition explicitly to two MPI tasks in the
-$z-$direction. We also set
-\begin{verbatim}
+$z$-direction. We also set
+\begin{lstlisting}
 porous_media_format BINARY_SERIAL
@@
 porous_media_type   status_only
-\end{verbatim}
+\end{lstlisting}
 to ensure the structure file is read correctly in parallel.
 We should run this on 2 MPI tasks, e.g.,
-\begin{verbatim}
+\begin{lstlisting}
 $ mpirun -np 2 ./Ludwig.exe input
-\end{verbatim}
+\end{lstlisting}
 The final result should be exactly the same as the serial version
@@
 There is one potential problem if we wish to study transport
-phemonema where we might want to run a simulation long enough
+phenomena where we might want to run a simulation long enough
 for the flow to cross the whole sample. With $U \sim 10^{-4}$
@@
 remains small compared with unity (all Reynolds numbers $< 0.1$
-can be regarded as neglible \cite{batchelor}.) Here, for example,
+can be regarded as negligible \cite{batchelor}.) Here, for example,
 if we raise the flow by two orders of magnitude to $10^{-2}$ in
@@
 managable $10^5$ (see also \cite{cates_scaling}).
+
+% End examples
+\vfill
+\pagebreak

### 2017-02-14 docs/leesedwards.tex

--- 2011-01-14 branches/CUDA/htdocs/sections.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
-%  sections.tex
+%  leesedwards.tex
 %
-%  Content for Ludwig documentation.
-%
-%  $Id: sections.tex,v 1.15 2010-10-21 18:05:50 kevin Exp $
+%  Lees Edwards Sliding Perioidic Boundary Conditions
 %
@@
 %
+%  (c) 2016 The University of Edinburgh
+%
+%  Contributing authors:
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2008 The University of Edinburgh
 %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{The Lattice Boltzmann Method}
+\section{Lees Edwards Sliding Periodic Boundary Conditions}
 
-We review here the lattice Boltzmann method applied to single
-phase problems.
+\subsection{Background}
 
-\subsection{The Navier Stokes Equation}
+The idea of introducing a Galilean transformation in a periodic
+computation to model uniform shear was introduced by Lees and
+Edwards in 1972 \cite{lees-edwards1972}. It was first adapted
+to the lattice Boltzmann picture by Wagner and Pagonabarraga
+in 2000 \cite{wagner-pagonabarraga2002}. The current
+implementation follows the description of Adhikari
+et~al.\ \cite{adhikari-desplat2005}.
 
-We seek to solve the isothermal Navier-Stokes equations which, often
-written in vector form, express mass conservation
-\begin{equation}
-\partial_t \rho + \boldsymbol{\nabla}.(\rho\mathbf{u}) = 0
-\label{eq_mass1}
-\end{equation}
-and the conservation of momentum
-\begin{equation}
-\partial_t (\rho\mathbf{u}) + \boldsymbol{\nabla}.(\mathbf{\rho uu}) =
--\boldsymbol{\nabla}p + \eta \nabla^2 \mathbf{u}
-+\zeta \boldsymbol{\nabla}(\boldsymbol{\nabla}.\mathbf{u}).
-\label{eq_momentum1}
-\end{equation}
-Equation~\ref{eq_mass1} expresses the local rate of change of the
-density $\rho(\mathbf{r}; t)$ as the divergence of the flux of
-mass associated with the velocity field $\mathbf{u}(\mathbf{r}; t)$.
-Equation~\ref{eq_momentum1} expresses Newton's second law for
-momentum, where the terms on the right hand side represent the
-force on the fluid.
-
-For this work, it is more convenient to rewrite these equations
-in tensor notation, where Cartesian coordinates ${x,y,z}$ are
-represented by indices $\alpha$ and $\beta$, viz
-\begin{equation}
-\partial_t \rho + \nabla_\alpha (\rho u_\alpha) = 0
-\end{equation}
-and
-\begin{equation}
-\partial_t (\rho u_\alpha) + \nabla_\beta (\rho u_\alpha u_\beta)
-= -\nabla_\alpha p
-+  \eta \nabla_\beta (u_\alpha \nabla_\beta + \nabla_\alpha u_\beta)
-+ \zeta \nabla_\alpha (\nabla_\gamma u_\gamma).
-\end{equation}
-Here, repeated indices are understood to be summed over.
-The conservation law is seem better if the forcing terms of the
-right hand side are combined in the fluid stress $\Pi_{\alpha\beta}$
-so that
-\begin{equation}
-\partial_t (\rho u_\alpha) +\nabla_\beta \Pi_{\alpha\beta} = 0.
-\end{equation}
-In this case the expanded expression for the stress tensor is
-\begin{equation}
-\Pi_{\alpha\beta} = p \delta_{\alpha\beta} + \rho u_\alpha u_\beta 
-+ \eta \nabla_\alpha v_\beta + \zeta (\nabla_\gamma v_\gamma)\delta_{\alpha\beta}
-\end{equation}
-where $\delta_{\alpha\beta}$ is that of Kroneker. Th Navier-Stokes
-equations have 10 degrees of freedom (or hydrodynamic modes) being
-$\rho$, three components of the mass flux $\rho u_\alpha$, and 6 independent
-modes from the (symmetric) stress tensor $\Pi_{\alpha\beta}$.
-
-\subsection{The Distribution Function and its Moments}
-
-In lattice Boltzmann, the density and velocity of the continuum fluid
-are replaced by their discrete counterparrts, the distribution function
-$f_i(\mathbf{r}; t)$ and the discrete velocity space $c_{i\alpha}$.
-It is possible to relate the hydrodynamic modes to distribution
-function via its moments, so
-\begin{equation}
-\rho = \sum_i f_i,  \,
-\rho u_\alpha = \sum_i f_i c_{i\alpha},  \,
-\Pi_{\alpha\beta} = \sum_i f_i c_{i\alpha} c_{i\beta.}
-\end{equation}
-here, the index of the summation is over the number of discrete
-velocities used $n$, which in three dimensions is generally
-denoted by D3Q$n$. We restrict our discussion to D3Q15 and
-D3Q19 (see below).
-
-In general, the number of modes supported in $n$ and these can
-be written as
-\begin{equation}
-M^a(\mathbf{r};t) = \sum_i m_i^a f_i(\mathbf{r};t),
-\end{equation}
-where $m_i^a$ is the corresponding eigenvector of the collision
-maxtrix (see section blah). For example, if $m_i^a = 1$ then the
-mode $M^a$ is the density. Correspondingly, the distribution
-function can be related to the modes $M^a$ via
-\begin{equation}
-f_i() = \sum_a w_i m_i^a M^a() N^a
-\end{equation}
-
-\subsubsection{Hydrodynamic modes}
-
-\subsubsection{D3Q15 ghost modes}
-
-\subsubsection{D3Q19 ghost modes}
-
-\subsection{The Lattice Boltzmann Equation}
-
-\subsection{The Collision}
-
-\subsection{External Body Forces}
-
-\subsection{Fluctuating LBE}
-
-It is possible \cite{adhikari2005} to simulate fluctuating
-hydrodynamics for an isothermal fluid via the inclusion of
-a fluctuating stress $\sigma_{\alpha\beta}$:
-\begin{equation}
-\Pi_{\alpha\beta} = p\delta_{\alpha\beta} + \rho u_\alpha u_\beta
-+ \eta_{\alpha\beta\gamma\delta} \nabla_\gamma u_\delta + \sigma_{\alpha\beta}.
-\end{equation}
-The fluctuation-dissipation theorem relates the magnitude of this
-random stress to the isothermal temperature and the viscosity.
-
-In the LBE, this translates to the addition of a random contribution
-$\xi_i$ to the distribution at the collision stage, so that
-\begin{equation}
-\ldots + \xi_i.
-\end{equation}
-
-For the conserved modes $\xi_i = 0$. For all the non-conserved modes,
-i.e., those with dissipation, the fluctuating part may be written
-\begin{equation}
-\xi_i (\mathbf{r}; t) = w_i m_i^a \hat{\xi}^a (\mathbf{r}; t) N^a
-\end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
-by the relaxation time for given mode
-\begin{equation}
-\left< \hat{\xi}^a \hat{\xi}^b \right> =
-\frac{\tau_a + \tau_b + 1}{\tau_a \tau_b}
-\left< \delta M^a \delta M^b \right>.
-\label{eq_fvar}
-\end{equation}
-
-\subsubsection{Fluctuating stress}
-
-For the stress, the random contribution to the distributions is
-\begin{equation}
-\xi_i = w_i \frac{Q_{i\alpha\beta} \hat{\sigma}_{\alpha\beta}}{4c_s^2}
-\end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
-variates drawn from a Gaussian distribution with variance given
-by equation~\ref{eq_fvar}. In the case that the shear and bulk
-viscosities are the same, i.e., there is a single relaxation
-time, then the variances of the six independent components of
-the matrix are given by
-\begin{equation}
-\left< \hat{\sigma}_{\alpha\beta} \hat{\sigma}_{\mu\nu} \right> =
-\frac{2\tau + 1}{\tau^2}
-(\delta_{\alpha\mu}\delta_{\beta\nu} + \delta_{\alpha\nu} \delta_{\beta\mu}).
-\end{equation}
-
-\subsubsection{Fluctuating ghost modes: D3Q15}
-
-\subsubsection{Fluctuating ghost modes: D3Q19}
-
-\subsection{The Propagation}
-
-\subsection{Solid bodies and Bounce-Back on Links}
-
-A very general method for the representation of solid objects
-within the LB approach was put forward by Ladd \cite{l94a, l94b}.
-Solid objects (of any shape) are defined by a boundary surface
-which intersects some of the velocity vectors $\mathbf{c}_i$
-joining lattice nodes. Sites inside are designated solid, while
-sites outside remain fluid. The correct boundary condition is
-defined by identifying \textit{links} between fluid and solid
-sites, which allows those elements of the distribution which would
-cross the boundary at the propagation step to be ``bounced-back''
-into the fluid. This bounce-back on links is an efficient method
-to obtain the  correct hydrodynamic interaction between solid
-and fluid.
-
-\subsubsection{Boundary walls}
-
-\subsection{Colloids}
-
-\subsubsection{General}
-
-Colloidal particles are assumed to be spherical with a geometrical
-centre $\mathbf{r}_c$, which is also the centre of mass. The
-centre is allowed to move continuously across the lattice
-with velocity $\mathbf{U}$; the particle has an angular velocity
-$\mathbf{\Omega}$. The surface of the colloid is defined by an
-input radius, $a_0$, which determines which lattice nodes are
-inside or outside the colloid. (The hydrodynamic properties of
-the colloid are specified by a different radius $a_h$ --- more
-of this later.) The boundary links are then the set of vectors
-joining lattice nodes which intersect the spherical surface
-$\{\mathbf{c}_b\}$. A schematic picture is presented in
-Figure~\ref{fig_coll2}. Note that a lattice node exactly at
-the solid-fluid interface is defined to be outside the colloid.
-
-In the original approach of Ladd, fluid occupied nodes both inside
-and outside the particle. The effect of the ``internal fluid'' is
-known to be restricted to short time scales (compared to the
-characteristic time $a_0^2/\nu$), on which the fluid inside the
-particle relaxes to a solid body rotation \cite{heemels}. There
-are a number of problems related to the internal fluid, so we
-use fully solid particles.
-
-Nguyen and Ladd.
-
-For each boundary link, it is useful to define a boundary vector,
-$\mathbf{r}_b$, which connects the centre of the colloid to the
-appropriate boundary node (note that the length of this vector is
-not necessarily equal to the input radius of the colloid).
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidhalflinksnew.eps}
-\end{center}
-\caption{Boundary links for a colloidal particle in two dimensions
-with D2Q9 (half the particle is shown). Links join fluid sites (red)
-to solid sites (green) and
-intersect the circular shell radius $a_0$. Boundary nodes (circles)
-lie exactly half way between pairs of lattice nodes. The generalisation
-to three dimensions is straightforward.}
-\label{fig_coll2}
-\end{figure}
-
-A \textit{boundary node} halfway between fluid nodes which are joined
-by a boundary link. The position of the boundary node is always
-exactly halfway along the boundary link ($\mathbf{r} + \frac{1}{2}\mathbf{c}_b
-\Delta t$) regardless of the actual position of the intersection
-of the colloid surface and the boundary link.
-The colloid therefore has a discrete representation which becomes a
-better approximation to the sphere as the input radius becomes larger;
-this approximation is known to be reasonably good for $a_0 > 5\Delta x$.
-
-\subsubsection{Bounce-back on links}
-
-The standard boundary condition required for the solid-fluid
-interface of a moving particle is described by Ladd \cite{l94b},
-and is generally refered to as bounce-back on links (BBL).
-A boundary link is defined as joining a node $\mathbf{r}$
-inside the particle to one outside at $\mathbf{r} + \mathbf{c}_b \Delta t$.
-If the post-collision distributions are denoted by $f^\ast$, then
-the distributions must be reflected at the solid surface so that
-\begin{equation}
-\label{eq:bbl1}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2w_{c_b} \rho \mathbf{u}_b.\mathbf{c}_b}{c_s^2}
-\end{equation}
-where the boundary link $\mathbf{c}_{b'} = -\mathbf{c}_b$. The velocity
-at the boundary
-\begin{equation}
-\label{eq:ub}
-\mathbf{u}_b = \mathbf{U} + \mathbf{\Omega}\times\mathbf{r}_b
-\end{equation}
-is determined by the particle linear velocity $\mathbf{U}$ and angular
-velocity $\mathbf{\Omega}$. The change in momentum described by
-
-The force exerted on a
-single link is
-\begin{equation}
-\mathbf{F}_b(\mathbf{r} + {\scriptstyle\frac{1}{2}}\mathbf{c}_b\Delta t;
-t + {\scriptstyle\frac{1}{2}}\Delta t) = \frac{\Delta x^3}{\Delta t}
-\Big[ 2f_b^\ast(\mathbf{r}; t) - \frac{2w_{c_b}\rho_0 \mathbf{u}_b .
-\mathbf{c}_b}{c_s^2} \Big] \mathbf{c}_b,
-\label{eq:fb}
-\end{equation}
-with corresponding torque $\mathbf{T}_b = \mathbf{r}_b \times \mathbf{F}_b$.
-
-The total hydrodynamic force on the particle is then found by taking
-the sum of
-$\mathbf{F}_b$ over all the boundary links defining the particle.
-The is an associated torque on each link of $\mathbf{r}_b\times\mathbf{F}_b$,
-which again is summed over all links to give the total torque on the colloid.
-
-Note that $\rho$ in the above equations is the density of the fluid
-at the appropriate fluid node for the link. As the density fluctuations
-in the fluid are small compared with the mean density $\rho_0$, the
-density in the correction to the bounce-back can be replaced by the
-mean $\rho_0$.
-
-\subsubsection{Dynamics}
-Having computed the total force and torque on an individual colloid,
-is is possible to update the the linear velocities
-\begin{equation}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t \mathbf{F} (t),
-\end{equation}
-where the mass of the colloid is related to the input radius by
-$m_0 = {\scriptstyle\frac{4}{3}}\pi\rho_0 a_0^3$ and the angular
-velocity
-\begin{equation}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\mathbf{T}(t),
-\end{equation}
-where the moment of inertia is $I_0 = {\scriptstyle\frac{2}{5}}m_o a_o^2$.
-However, this explicit update is generally found to have poor stability
-properties \cite{l94b, nl02}. The alternative is to use a velocity update
-which is implicit \cite{heemels,nl02}.
-
-The total force and torque on a colloid can be split into velocity-dependent
-and -independent parts by combining Equations~(\ref{eq:ub})
-and~(\ref{eq:fb}) to eliminate the boundary velocity $\mathbf{u}_b$.
-In this way, the decomposition is
-\begin{eqnarray}
-\mathbf{F} = \mathbf{F}_0 - \boldsymbol{\zeta}^{FU}.\mathbf{U}
--\boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega},
-\label{eq:d3}
-\\
-\mathbf{T} = \mathbf{T}_0 - \boldsymbol{\zeta}^{TU}.\mathbf{U}
--\boldsymbol{\zeta}^{T\Omega}.\mathbf{\Omega}.
-\label{eq:d4}
-\end{eqnarray}
-The velocity independent parts of the force and the torque
-(appropriate for a colloid at rest) are
-\begin{eqnarray}
-\mathbf{F}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big] \mathbf{c}_b,
-\\
-\mathbf{T}_0(t + {\scriptstyle\frac{1}{2}}\Delta t) =
-\frac{2\Delta x^3}{\Delta t} \sum_b \big[ f_b^\ast(\mathbf{r}; t)
-- f_{b'}^\ast(\mathbf{r} + \mathbf{c}_i \Delta t; t) \big]
-(\mathbf{r}_b\times \mathbf{c}_b)
-\end{eqnarray}
-The matrices $\boldsymbol{\zeta}$ are interpreted as drag
-coefficients and can be written as
-\begin{eqnarray}
-\mathbf{\boldsymbol{\zeta}}^{FU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{F\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b),
-\\
-\mathbf{\boldsymbol{\zeta}}^{TU} &=& \frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b) \mathbf{c}_b,
-\\
-\mathbf{\boldsymbol{\zeta}}^{T\Omega} &=&
-\frac{4\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b \times \mathbf{c}_b)
-(\mathbf{r}_b \times \mathbf{c}_b).
-\end{eqnarray}
-As an example, the full form of the $\boldsymbol{\zeta}^{F\Omega}$ matrix is
-
-\begin{equation}
-\label{eq:k14}
-\left( \begin{array}{rrr}
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bx} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{by} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_x &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_y &
-\sum_b c_{bz} (\mathbf{r}_b \times \mathbf{c}_b)_z \\
-
-\end{array} \right).
-\end{equation}
-If the particle has a symmetric distribution of boundary links (the
-special case where the centre of mass is on a symmetry point of the
-lattice) the $\boldsymbol{\zeta}^{FU}$ and
-$\boldsymbol{\zeta}^{T\Omega}$ matrices are
-diagonal, while the $\boldsymbol{\zeta}^{F\Omega}$
-and $\boldsymbol{\zeta}^{TU}$ matrices are zero.
-
-The velocity updates can now be rewritten using an
-implicit update, which results in
-\begin{eqnarray}
-\label{eq:k15}
-m_0 \mathbf{U}(t + \Delta t) = m_0 \mathbf{U}(t) + \Delta t
-\big[ \mathbf{F}_0 (t + {\scriptstyle\frac{1}{2}}\Delta t)
-- \boldsymbol{\zeta}^{FU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{F\Omega}.\mathbf{\Omega}(t + \Delta t) \big],
-\\
-\label{eq:k16}
-I_0 \mathbf{\Omega} (t + \Delta t) = I_0 \mathbf{\Omega}(t) + \Delta t
-\big[ \mathbf{T}_0(t + {\scriptstyle\frac{1}{2}} \Delta t)
-- \boldsymbol{\zeta}^{TU} . \mathbf{U}(t + \Delta t)
-- \boldsymbol{\zeta}^{T\Omega}. \mathbf{\Omega}(t + \Delta t) \big].
-\end{eqnarray}
-In general all the matrix elements are non-zero, in which case
-equations~(\ref{eq:k15}) and~(\ref{eq:k16}) can be solved via a
-6$\times$6 matrix inversion.
-
-Finally, the position of the particle can be updated: an Euler
-forward step is taken using the updated velocity
-\begin{equation}
-\mathbf{r} (t + \Delta t) = \mathbf{r} (t) + \Delta t\mathbf{U}(t + \Delta t)
-\end{equation}
-
-\subsubsection{Changes in particle shape}
-
-One consequence is that the discrete shape of the particle fluctuates
-as the colloid centre moves relative to the
-lattice. This means that the size of the particle as seen by the
-fluid changes despite the fact that the input radius of the colloid
-is fixed.  However, these fluctuations are small for input radii
-greater than about 5 lattice units \cite{l96a}. Furthermore, the
-fluctuations may be greater for some input radii than others
-\cite{nl02}.
-
-In the standard approach, changes in the map of boundary links
-are accomodated by the internal fluid. If the particle shape
-changes so that an internal node is exposed, the internal fluid
-at that site rejoins the fluid proper with the solid body momentum
-(assuming the internal fluid is relaxed to solid-body rotation).
-If a fluid node is covered by particle movement, then it simply
-joins the internal fluid, and will relax to solid-body rotation.
-Without internal fluid, changes in particle shape
-in which fluid nodes are either covered or uncovered must be
-accompanied by the removal or addition
-of fluid with appropriate properties at the nodes in question.
-It is essential that this
-is done in a way which minimises the perturbation to the fluid
-flow.
-
-If a fluid node at $\mathbf{r}$ is covered by the movement of a
-solid particle,
-the fluid loses a mass $\rho(\mathbf{r}) \Delta x^3$, of which an excess
-$\Delta M_c = (\rho(\mathbf{r}) - \rho_0)\Delta x^3$ must be replaced
-explicitly so that
-the overall mass density is unchanged. At the same time, the
-colloid must assume the momentum lost by the fluid
-$\Delta x^3 \sum_i f_i(\mathbf{r}) \mathbf{c}_i$.
-
-Similarly, when a fluid node is exposed by particle movement,
-fluid must be replaced with the appropriate distribution, density
-and velocity. If the
-new density is $\rho$ (to be determined by some method) then
-the fluid gains an excess of mass $\Delta M_u = (\rho - \rho_0)
-\Delta x^3$ which must be balanced elsewhere. If the new fluid
-is assumed to have velocity $\mathbf{u}$ then the particle must
-give up momentum $\Delta x^3 \rho\mathbf{u}$.
-
-Following Nguyen and Ladd \cite{nl02}, these changes owing
-to change in particle shape are implemented by added a small
-correction to the bounce-back at the following time step.
-The excess mass from a covered fluid site is redistributed over all the
-boundary nodes by redefined the mometum transfer at bounce-back
-\begin{equation}
-f_{b'}(\mathbf{r}; t + \Delta t) = f_b^\ast (\mathbf{r}; t)
-- \frac{2 w_{c_b} \rho_0 \mathbf{u}_b . \mathbf{c}_b}{c_s^2}
-+ \frac{w_{c_b} \rho_0\Delta M_c }{A},
-\label{eq:q1}
-\end{equation}
-where $A = \rho_0 \Delta x^3 \sum_b w_{c_b}$. The accompanying
-force on the particle owing to the change in shape is then
-that lost by the fluid plus the contributions from the final term
-in (\ref{eq:q1}) summed over all the boundary links
-\begin{equation}
-\Delta \mathbf{F}_c = \frac{\Delta x^3}{\Delta t}\sum_{i} f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t} \sum_b \frac{w_{c_b} \rho_0 \Delta M_c}{A}
-\mathbf{c}_b.
-\label{eq:q2}
-\end{equation}
-
-Note that for a closed surface,
-$\sum_b w_{c_b}\mathbf{c}_b = 0$, and the second term on the right-hand
-side of Eq.~(\ref{eq:q2}) is zero, i.e., the redistribution of mass
-does not contribute to the net force and torque on the particle.
-The corresponding change in torque is
-\begin{equation}
-\Delta \mathbf{T}_c = \frac{\Delta x^3}{\Delta t} \mathbf{r}_c
-\times \sum_i f_i \mathbf{c}_i
-+ \frac{\Delta x^3}{\Delta t}
-\sum_b \frac{w_{c_b}\rho_0 \Delta M_c}{A} \mathbf{r}_b \times \mathbf{c}_b,
-\label{eq:q2a}
-\end{equation}
-where $\mathbf{r}_c$ is the boundary vector at the position where the
-fluid has been removed. Again, for a closed surface, the
-redistribution of mass does not contribute to the net torque
-on the particle.
-
-Likewise, the contribution to the bounce-back from newly uncovered nodes
-is
-\begin{equation}
-- \frac{w_{c_b} \rho_0 \Delta M_u}{A}
-\label{eq:q3}
-\end{equation}
-leading to a change in force of
-\begin{equation}
-\Delta \mathbf{F}_u = -\frac{\Delta x^3 \rho(\mathbf{r})
-\mathbf{u}(\mathbf{r})}{\Delta t} - \frac{\Delta x^3}{\Delta t} \sum_b
-\frac{w_{c_b} \rho_0 \Delta M_u}{A} \mathbf{c}_b,
-\label{eq:q4}
-\end{equation}
-with a corresponding change in the torque.
-If more than one lattice node is either covered or uncovered by
-the movement of the particle, then these contributions add in a simple
-way. The overall contributions can be added to the
-velocity-independent terms $\mathbf{F}_0$ and $\mathbf{T}_0$ appearing in
-Equations~(\ref{eq:d3}) and~(\ref{eq:d4}).
-
-\subsubsection{Particles near contact}
-
-Particles near contact may not pocess a full set of boundary
-links (see Figure~\ref{fig:f5}). This leads to a potential
-non-conservation of mass associated with the particle motion
-which must be corrected.
-
-\begin{figure}[tb]
-\begin{center}
-\includegraphics{xfig/colloidclose.eps}
-\end{center}
-\caption{Two colloids close to contact are missing links in the
-region of closest approach owing to a lack of fluid nodes in the
-interstice.}
-\label{fig:f5}
-\end{figure}
-Again following Nguyen and Ladd \cite{nl02}, mass conservation is
-enforced explicitly using the following procedure. There is a mass
-transfer associated with the bounce-back at each link of
-$(2w_{c_b}\rho_0 \mathbf{u}_b . \mathbf{c}_b / c_s^2) \Delta x^3$.
-The net mass transport into the particle is the sum of this
-quantity over all the links, and can be written
-\begin{equation}
-\Delta M_s = - \frac{2\Delta x^3 \rho_0}{c_s^2}
-\Big[ \mathbf{U}.\sum_b w_{c_b} \mathbf{c}_b +
-  \mathbf{\Omega} . \sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b \Big].
-\end{equation}
-For any partilce with a full set of boundary links, both
-$\sum_b w_{c_b} \mathbf{c}_b$
-and $\sum_b w_{c_b} \mathbf{r}_b \times \mathbf{c}_b$ are zero.
-However, if the set of boundary links is imcomplete
-(as in Figure~\ref{fig:f5}) $\Delta M_s$ is not zero and
-a correction is required. This is again made by redistributing the
-excess or deficit of mass over the other existing boundary nodes.
-The additional contribution to the bounce-back here is
-$-w_{c_b} \rho_0 \Delta M_s/A$, where
-$A = \rho_0 \Delta x^3 \sum_b w_{c_b}$ (cf.\ Equation~\ref{eq:q1}).
-
-This contribution
-now adds to the velocity-dependent part of the force and torque and
-leads to a slightly different form of the drag matrices
-\begin{eqnarray}
-\boldsymbol{\zeta}^{FU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{c}_b - \overline{\mathbf{c}_b}) \mathbf{c}_b
-\\
-\boldsymbol{\zeta}^{F\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} \mathbf{c}_b (\mathbf{r}_b \times \mathbf{c}_b -
-\overline{\mathbf{r}_b\times\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{TU} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{c}_b - \overline{\mathbf{c}_b}),
-\\
-\boldsymbol{\zeta}^{T\Omega} &=& \frac{-2\rho_0 \Delta x^3}{c_s^2 \Delta t}
-\sum_b w_{c_b} (\mathbf{r}_b\times\mathbf{c}_b)
-(\mathbf{r}_b\times\mathbf{c}_b - \overline{\mathbf{r}_b\times\mathbf{c}_b}).
-\end{eqnarray}
-The mean quantities
-\begin{equation}
-\overline{\mathbf{c}_b} = \frac{\sum_b w_{c_b} \mathbf{c}_b}{ \sum_b w_{c_b}}
-\end{equation}
-and
-\begin{equation}
-\overline{\mathbf{r}_b\times\mathbf{c}_b} =
-\frac{\sum_b w_{c_b} \mathbf{r}_b\times\mathbf{c}_b}{\sum_b w_{c_b}}.
-\end{equation}
-A little algebra will show that both occurances of $\mathbf{c}_b$ and/or
-$\mathbf{r}_b\times\mathbf{c}_b$ in the definition of the drag matrices
-can be replaced by their deviation from the mean. This provides a
-convenient way to compute the drag matrices (maintaining symmetry)
-and computing the correct force and torque on the particle when close
-to contact.
-
-\subsubsection{Rotational motions}
-
-For a number of applications it is useful to update not only
-the psoition of a spherical particle, but its orientation. This
-is important, e.g., when considering magnetic particles. The
-best way to implementment the required rotations is by
-considering quaternions.
-
-A quaternion can be thought of as an extended complex number
-\begin{equation}
-q = w + xi + yj + zk
-\end{equation}
-where $i^2 = j^2 = k^2 = -1$ and $ijk = -1$. The norm, or length
-of a quaternion is
-\begin{equation}
-N(q) = (w^2 + x^2 + y^2 + z^2)^{1/2}.
-\end{equation}
-Addition and subtraction of quaternions proceeds as normal, but
-care is required for multiplication, which is associative but not
-commutative. If the quaternion is writen as $q = (w, {\bf v})$ where
-$\bf{v}$ is a normal 3-vector, then multiplication can be written
-\begin{equation}
-q_1 q_2 = (w_1 w_2 - {\bf v}_1 . {\bf v}_2, w_1 {\bf v}_2 + w_2 {\bf v}_1
-+ {\bf v}_1 \times {\bf v}_2),
-\end{equation}
-where the usual scalar and vector products are used. Note that the
-conjugate of a quaterion $q^\star = (w, -{\bf v})$ so that
-$qq^\star = N^2(q)$.
-It can be shown that a rotation of an arbitrary vector though an
-angle $\theta$ around the unit axis of rotaion $\hat{\bf w}$ can
-be represented by the quarternion
-\begin{equation}
-q = (\cos(\theta/2), \hat{\bf w} \sin(\theta/2))
-\end{equation}
-which ensures that $N(q) = 1$. If the arbitrary vector is (0, {\bf v}),
-then the rotated vector ${\bf v}^{'}$ can be written as
-\begin{equation}
-(0, {\bf v}^{'}) = q (0, {\bf v}) q^\star
-\end{equation}
-which can be expanded as
-\begin{equation}
-{\bf v}^{'} = (1 - \cos\theta)({\bf v}.{\hat{\bf w}) \hat{\bf w} +
-{\bf v}\cos\theta} + (\hat{\bf w} \times {\bf v})\sin\theta.  
-\end{equation}
-
-\subsection{Colloid-Colloid interactions}
-
-\subsubsection{Lubrication corrections}
-
-\subsubsection{Hard sphere interaction}
-
-Particles do not overlap, i.e., there is always
-a hard-sphere interaction which is a function of the separation $h$:
-\begin{equation}
-v^{hs}(h) = \left\{
-\begin{array}{ll}
-\infty & h \leq 0,\\
-0   & h > 0.
-\end{array} \right.
-\end{equation}
-The hard-sphere interaction does not give rise to a force.
-
-\subsubsection{Soft-sphere interaction}
-
-A simple soft-sphere interaction is available, the basic form
-of which is:
-\begin{equation}
-v^{ss}(r) = \epsilon (\sigma / r)^{\nu},
-\end{equation}
-where $\epsilon$ sets the energy scale, and $\sigma$ sets the
-characteristic width. The steepness of the potential is set by
-the exponent $\nu (> 0)$.
-
-To prevent the need for calculation of long-range interactions,
-the soft-sphere potential is truncated in a ``cut-and-shift''
-approach. This is done in such a way as to smoothly match
-both the potential and the force at the cut-off distance
-$r_c$. For $r < r_c$ the potential is then
-\begin{equation}
-v^{ss}(r) - v^{ss}(r_c) - (r - r_c) \left.\frac{d v^{ss}}{dr}\right|_{r=r_c}.
-\label{eq_ss_shift}
-\end{equation}
-Clearly, this potential is not exactly what we first thought of as
-soft-sphere. Matching the potential smoothly ensures conservation
-of energy, while matching the force smoothly prevents potential
-instabilities in the molecular dynamics update.
-
-The soft-sphere potential can be useful as a mechanism to
-keep the particles from touching (which risks hard-sphere interactions),
-in which case is is possible to compute the interaction as a function
-of the separation $h$, instead of $r$. The force between
-two particles is computed via the derivative of equation~\ref{eq_ss_shift}
-with respect to $r$:
-\begin{equation}
-\mathbf{F}_{ij}(r) = -\left\{\frac{d v^{ss}(r)}{dr} -
-\left. \frac{d v^{ss}(r)}{dr}\right|_{r=r_c} \right\} \mathbf{\hat{r}}_{ij}
-\end{equation}
-where $\mathbf{\hat{r}}_{ij}$ is the unit vector joining the centre of
-particle $i$ to the centre of $j$.
-
-\subsubsection{Leonard-Jones interaction}
-
-\section{Lattice Boltzmann for a Binary Fluid}
-
-\subsection{The Cahn-Hilliard Equation}
-
-\subsection{The Choice of Free Energy}
-
-\subsection{A Second Distribution Function}
-
-\subsection{The Collision}
-
-\subsection{Bounce-Back on Links}
-
-\section{Surfactants}
-
-This section describes the approach taken to study surfactants. Here,
-we always adopt the hydrid method, i.e., LB for the hydrodynmic
-equations and finite difference for the associated Cahn-Hilliard
-equations. Coupling between the two sectors is via the velocity
-field and the external force on the fluid computed via the
-divergence of the `chemical stress'.
-
-In this model the surfactant appears as a concentration field
-$\psi(\mathbf{r};t)$, alongside the compositional order parameter
-$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
-two corresponding Cahn-Hilliard equations: first,
-
-\begin{equation}
-\partial_t \phi + \partial_\alpha \phi u_\alpha = 
-\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
-\end{equation}
-and second
-\begin{equation}
-\partial_t \psi + \partial_\alpha \psi u_\alpha = 
-\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
-\end{equation}
-
-\subsection{Free Energy}
-
-A number of different free energies have been used to study
-surfactants [cite Yeomans]. Here, the free energy is based on
-that used by van der Graff and van der Sman and others [cite].
-There are a number of different contributions to the free
-energy density:
-\begin{equation}
-f = f_\phi + f_\psi +f_1 + f_{ex}.
-\end{equation}
-The first is exactly that used for the binary fluid, i.e.,
-\begin{equation}
-f_\phi (\phi, \partial_\alpha\phi) =
-{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
-+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
-\end{equation}
-The surface tension and the interfacial width are determined from the
-choice of parameters $A,B,\kappa$ as before.
-The second term in the free energy density
-represents the entropic
-energy of mixing of the surfactant with the bulk phases
-\begin{equation}
-f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
-\end{equation}
-where the parameter $kT$ sets the energy scale for the whole system.
-The surfactant order parameter, $\psi$, is unity for a saturated
-interface and zero for a plain interface.
-
-The surfactant lowers the energy of the interface where it is adsorbed
-there, represented by the term
-\begin{equation}
-f_1(\psi, \partial_\alpha \phi) 
-= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
-  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
-\end{equation}
-where $\epsilon$ and $\beta$ are parameters. Note that only the
-first term appears in \cite{vandergraaf}, while the second term is
-based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
-the surfactant profile at the interface is symmetric (so-called
-Langmuir adsoption) whereas the term in $\beta$ induces an
-asymmetry (Frumpkin adsorption).
-Finally,  $f_{ex}$ is [van der Sman claims]
-an enthalpic contribution introduced 
-to stabilise the interface \cite{theissengompper}
-\begin{equation}
-f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
-\end{equation}
-with $W$ a parameter.
-
-\subsection{Chemical potentials}
-
-The respective chemical potentials are obtained from the functional
-derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
-$\mu_\psi = \delta F / \delta \psi$. We obtain:
-\begin{equation}
-\mu_\phi =
-A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
-+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
-+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
-+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
-+ \beta \psi^2 \partial_\alpha^2 \phi,
-\end{equation}
-and
-\begin{equation}
-\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
-+ {\textstyle \frac{1}{2}}W\phi^2
-- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
-- \beta\psi (\partial_\alpha \phi)^2.
-\end{equation}
-As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
-+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
-\cite{theissengompper})
-\begin{equation}
-p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
-\end{equation}
-There is no closed form for the tensor part, which must be chosen to
-comply with the constraint that in mechanical equilibrium
-\begin{equation}
-\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
-\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
-\end{equation}
-In this way we obtain for the isotropic contribution
-\begin{eqnarray}
-p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
-- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
-- \kappa \phi \partial_\alpha^2 \phi
-- kT \ln(1-\psi) + W\psi\phi^2
-\nonumber\\
-+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
-+ \epsilon\phi\psi \partial_\alpha^2 \phi
-+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
-+ \beta \phi \psi^2 \partial_\alpha^2 \phi
-- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
-\end{eqnarray}
-and for the tensor part we can have
-\begin{equation}
-\tilde{P}_{\alpha\beta} =
-(\kappa - \epsilon \psi - \beta\psi^2) 
-\partial_\alpha \phi \partial_\beta \phi.
-\end{equation}
-
-\section{Polar Active Gel}
-
-Here we consider a polar vector field model to study active gels. 
-The order parameter is this time a vector $P_\alpha$
-(with variable magnitude) as there is no longer head-tail symmetry
-in the system as was the case for the ${\bf Q}$ tensor formulation. 
-The equation of motion as:
-\begin{equation}
-\partial_t P_\alpha  \partial_\beta (u_{\beta} + wP_{\beta}) P_{\alpha}
-= \lambda D_{\alpha\beta} P_{\beta} - \omega_{\alpha\beta} P_{\beta}
-+\Gamma' h_{\alpha}.
-\end{equation}
-
-In this equation, $w$ is an active term, due to swimming, which causes
-self-advection of the order parameter, while $\lambda$ is a material
-dependent constant -- positive for rod-like molecules. If $|\lambda|>1$
-the liquid crystalline passive phase is flow-aligning, otherwise it is
-flow-tumbling.
-
-Note that the convection for the Leslie-Ericksen is that the
-velocity gradient tensor should be writen $W_{\alpha\beta}
-= \partial_\alpha u_\beta$ (and not the other way round, which
-is usual).
-
-The molecular field is given by
-$h_{\alpha}=-\delta {\cal F}_{\rm pol}/\delta p_{\alpha}$
-where ${\cal F}_{\rm pol}$ is the free energy for a polar active nematic,
-whose density we can take to be:
-\begin{equation}
-f =  \frac{\alpha_2}{2} |{\mathbf P}|^2 + \frac{\alpha_4}{4}|{\mathbf P}|^4
-+ \frac{K}{2}\left(\partial_\alpha P_{\beta}\right)^2 + 
-\frac{K_{LC}}{2}\left[\partial_\alpha (P_{\beta}P_{\gamma})\right]^2 
-\end{equation}
-where $K$ and $K_{LC}$ are elastic constants, $\alpha_4>0$, and $\alpha_2$ 
-can be positive (isotropic phase) or negative (polar nematic phase).
-
-The molecular field is explicitly given by:
-\begin{equation}
-h_{\alpha}=-\alpha_2 P_{\alpha} -\alpha_4 P_{\alpha} P_{\beta}P_{\beta}+
-K \partial_{\beta}\partial_{\beta} P_{\alpha}+2 K_{LC}
-P_{\gamma}\partial_{\beta}\partial_{\beta} (P_{\gamma}P_{\alpha}).
-\end{equation}
-
-The Navier-Stokes equation is:
-\begin{equation}
-\rho \left(\partial_t+v_{\beta}\partial_{\beta}\right) v_{\alpha} =
-\partial_{\beta} \Pi_{\alpha\beta} 
-+\eta \partial_{\beta}\partial_{\beta} v_{\alpha}.
-\end{equation}
-
-The stress tensor this time is 
-\begin{eqnarray}
-\Pi_{\alpha\beta} =
-\frac{1}{2}\left(P_\alpha h_\beta - P_\beta h_\alpha \right)
-- \lambda\left( (P_\alpha h_\beta + P_\beta h_\alpha)/2 
-- (1/3)P_\gamma P_\gamma \delta_{\alpha\beta} \right)
-\\
--  \zeta ( P_\alpha P_\beta - (1/3)P_\gamma P_\gamma \delta_{\alpha\beta})
-- K \partial_\alpha P_\gamma \partial_\beta P_\gamma.
-\end{eqnarray}
-The active term in the stress tensor, proportional
-to $\zeta$, has the same form for apolar and polar active gels.
-
-\section{Code Design}
-
-\subsection{Some Comments}
-
-In designing a code in C the main object of which is to do high
-performance computing, there are some inevitable tensions with
-software engineering best-practice. In particular, complete
-encapsulation of data in the spirit of using abstract data types
-tends to cause problems of performance.
-
-For example, encapsulation of the lattice Boltzmann distribution
-$f_i$, together with appropriate ``get'' and ``set'' functions for
-access when approapriate, is possible. There could then be a
-complete separation of interface and implementation of the
-distribution (other than that each individual $f_i$ is represented
-by a \texttt{double}). However, experience suggests that
-this imposes severe impediment to optimisation and hence
-unacceptable performance degradation.
-
-Schemes do exist by which one can approximate the object
-oriented programming style in C, and so achieve a separation
-of interface and implementation. However, this tends to put
-a heavy burden on the use of \texttt{void} pointers, itself
-a recipe for possible disaster. A truly object-oriented
-language such as Java would be preferable.
-
-So, for the time being, some relaxation of encapsulation is required.
-Exposing a pointer to the distributions which allows them to be accessed
-directly where required. This can be done merely by declaring the
-necessary variables \texttt{extern}. In practice, such deviations
-from encapsulation can be minimised in the lattice sector of the
-code. In particlar, \texttt{extern} declarations should not be
-required in \texttt{main()}, i.e., it should be able to write a
-new application without worrying about details of implementation.
-
-\subsubsection{Objects}
-
-Somewhat more difficult is the part of the code involving
-colloids, which would really favour a truly object-oriented
-approach.
-
-\subsubsection{Serial and parallel}
-
-The design decission has been taken to support execution of the
-code in both serial and parallel. The default compilation
-assumes that a serial executable is required, while  MPI code
-is introduced via C-preprocessor directives (\texttt{-D\_MPI\_}).
-
-There is no shared memory inplementation at this time.
-
-\subsection{Structural Components: \texttt{MPI\_COMM\_WORLD}}
-
-\subsection{Coordinate System: \texttt{MPI\_CARTESIAN\_COMMUNICATOR}}
-
-\begin{figure}
-\begin{center}
-%\includegraphics{xfig/sketch.eps}
-\end{center}
-\caption{A scheme by which different parts of the code might be
-encapsulated and in such a way to allow reliable development and
-testing. As this is C, will do not adopt a truly ``object'' langauge.
-Broadly, different modular parts should only depend on stuff above
-in the diagram. So, at the top, there is the parallel environment
-where communication occurs within MPI\_COMM\_WORLD. Built on top
-of this is the Cartesian Communicator responsible for halo swaps
-and so forth. The LE code may have a separate Communicator again. 
-(There's an argument to say the diagram should be the other way up!)}
-\end{figure}
-
-\section{Code Implementation}
-
-\subsection{Stuff}
-
-\subsection{Coordinate System}
-
-\subsubsection{The global coordinate system}
-
-The user specifies the system size in lattice units at run time,
-which sets the extent of memory which must be allocated for the
-Cartesian lattice. The extent of the physical coordinate system
-then reflects the lattice size directly (see Figure \ref{fig_c1}). 
-Each lattice site, or node,  has coordinates corresponding to its
-integer position in the system and can be thought of as being
-at the centre of a control volume one lattice unit on a side.
-This means that the nominal edges of the
-system are offset from the lattice sites themselves by half a
-lattice unit in each direction.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_c1.eps}
-\end{center}
-\caption{The global Cartesian coordinate system reflects the choice
-of lattice size; lattices sites are located at
-$x=1, \ldots, x = N_{total}$ and so on in each coordinate direction
-(the Figure only shows the labels in the $y-$direction for simplicity).
-In the Figure, the lattice sites in the domain proper are the full
-circles; halo sites, which are used to effect periodic boundary
-conditions, are included as open cirles}
-\label{fig_c1}
-\end{figure}
-
-For particles, the centres of which can move continuously across the
-lattice, the extents of the global coordinate system are then
-$L_{min} < x < L + L_{min}$ and so on.
-
-\subsubsection{Addressing the lattice: the local coordinate system}
-
-\subsection{Solid walls}
-
-General solid objects which are fixed for the duration of
-the simulation, such as porous media, can be defined
-by assigning solid/fluid status to each lattice node. 
-The links required between solid and fluid nodes can
-then be identified for the given model.
-
-\textit{Comment.} In general, a solid object will not
-affect the periodic nature of the simulation. A general
-solid object cannot move, i.e., the boundary velocity
-must be zero.
-
-\textit{Special case.} In the case where the solid object
-represents a flat boundary wall at the edge of the system
-in one or more directions, the periodicity is broken in
-the corresponding directions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% LE Implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Lees Edwards Sliding Period Boundary Conditions}
-
-\subsubsection{The LE module}
-
-\subsubsection{Distributions crossing the LE planes}
+\subsection{Distributions crossing the LE planes}
 
@@
 distributions when the LE boundary conditions are active. This
-is done in a two-stage process described by Adhikari \textit{et al.}
-\cite{adhikari_desplat}, implemented between the collision and
-propagation.
+is done in a two-stage process of reprojection and interpolation
+implemented between the collision and propagation.
 
-\textbf{a. Reprojection} The post-collision distributions with
+\subsubsection{Reprojection}
+The post-collision distributions with
 $c_x = \pm 1$ at sites adjacent to a boundary are modified to
@@
 \end{eqnarray}
-If we work with the changes to the moemnts, then the density is
+If we work with the changes to the moments, then the density is
 unaffected $\delta\rho =  0$, the velocity is changed by
@@
 A similar set of expressions are given for the order parameter
-distributions in Adhikari \textit{et al.} \cite{adhikari_desplat}.
+distributions in Adhikari \textit{et al.} \cite{adhikari-desplat2005}.
 
-\textbf{b. Interpolation of reprojected distributions}
+\subsubsection{Interpolation of reprojected distributions}
 As the sliding blocks are displaced continuously with
@@
 lattice spacings, an interpolation is also required before
-propagation can take place. Consider the situation (Fig.~\ref{fig_le2}) from
+propagation can take place. Consider the situation  from
 a stationary frame where the frame above has translated a small
 distance to the right ($\delta y < \Delta y$). In the stationary frame
-we must interpolate the ditributions with $c_x = 1$ to a
+we must interpolate the distributions with $c_x = 1$ to a
 `departure point' from which the propagation will move the
 distribution exactly to the appropriate lattice site in the moving
-frame. Schemeatically,
+frame. Schematically,
 \begin{equation}
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_le2.eps}
-\end{center}
-\caption{Schematic diagram showing interpolation of distributions in the
-stationary plane (below) in preparation for propagation into the moving
-frame (above). For a small displacement $\delta y = u^{LE} t$, the
-interpolation takes place between lattice sites in the stationary frame
-(green, left). Propogation from the interpolated positions (red) can
-then take place into the moving frame (right).}
-\label{fig_le2}
-\end{figure}
-
-In parallel, exactly the same calculation is undertaken. However,
-communication is required to take account of the processor
-decomposition in the $y$-direction. This requires the identification
-of the two processes which hold the relevant information as a function
-of displacement.
-
-Note that the interpolated values are stored in the appropriate lattice
-site in the rest frame that ensures the normal propagation stage moves all
-distributions to the correct destination in memory.
-
-\subsubsection{Order parameter gradients}
+\subsection{Order parameter gradients}
 
@@
 Here, we need to interpolate values in the moving frame to the position
-which is seen by the rest frame. Figure~\ref{fig_le3} shows a five-point
-stencil to calculate the gradient at the central point $(x,y,z)$ in the
+which is seen by the rest frame. If a five-point
+stencil is used to calculate the gradient at the central point
+$(x,y,z)$ in the
 rest frame. An interpolation of the $\phi$-field of the form
@@
 
-\begin{figure}[h]
-\begin{center}
-\includegraphics{xfig/fig_le3.eps}
-\end{center}
-\caption{Schematic diagram showing a five-point stencil to compute a
-gradient at the central point in the stationary frame. The upper point
-of the stencil is in the moving frame and so must be interpolated to
-the correct position (red sites).}
-\label{fig_le3}
-\end{figure}
-
 In parallel, the interpolation requires data from at most two adjacent
@@
 
-\subsubsection{Velocities and finite-difference fluxes}
+\subsection{Velocities and finite-difference fluxes}
 
@@
 uniqueness $f_{ijk}^e = f_{i+1jk}^w$ ensures conservation of order
-parameter. However, at the planes, the non-linear conbination of
+parameter. However, at the planes, the non-linear combination of
 velocity field and order parameter field does not in general give
@@
 directly via the divergence of the thermodynamic stress
-$F_\alpha = \nabla_\beta P_{\alpha\beta}^{th}$. This is implementated
+$F_\alpha = \nabla_\beta P_{\alpha\beta}^{th}$. This is implemented
 by computing fluxes of momentum at cell faces in each direction, and
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% End LE implementation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+% End section
 \vfill
 \pagebreak
-
-\section{Code Interface}
-
-\subsection{The Parallel Environment}
-
-\texttt{\#include ``pe.h''}
-
-As the code is to be available in both serial and parallel versions,
-it is desirable to abstract the basic environment to some extent.
-The parallel environment is intended to support MPI communication
-within \texttt{MPI\_COMM\_WORLD}, or fall back to a minimally
-consistent picture in serial (i.e., rank zero is the single process).
-
-\texttt{void pe\_init(int argc, char ** argv)}\\
-Responsible for initialisation of MPI, and hence must be first
-execuatable statement. The arguments are those to \texttt{main}
-which will be passed through to \texttt{MPI\_Init()}.
-
-\texttt{int pe\_rank(void)}\\
-Returns the process rank in \texttt{MPI\_COMM\_WORLD}, or 0 in
-serial.
-
-\texttt{int pe\_size(void)}\\
-Returns the number of processes in \texttt{MPI\_COMM\_WORLD}, or
-1 in serial.
-
-\texttt{void pe\_finalise(void)}\\
-Responsible for \texttt{MPI\_Finalize()}, and must be the last
-executable statement.
-
-\texttt{void info(const char * fmt, ...)}\\
-Behaves like \texttt{printf}, but only produces output on the
-root process.
-
-\texttt{void verbose(const char * fmt, ...)}\\
-Behaves as \texttt{printf} and reports the rank of the
-calling process in \texttt{MPI\_COMM\_WORLD}.
-
-\texttt{void fatal(const char * fmt, ...)}\\
-Prints an error message and identifies
-the rank of the calling process before terminating the program.
-
-\subsection{Run Time Input}
-
-\texttt{\#include ``runtime.h''}
-
-The ability to set a wide range of parameters at run time is
-very convenient, and is supported via an input file in which the
-user specifies parameters as key value pairs. The code maintains
-a list of key value pairs from which the user can retrieve the
-appropriate values for given keys.
-
-\texttt{void RUN\_read\_input\_file(const char * input\_file\_name)}\\
-This reads the named input file and sets up the list of keys.
-
-\texttt{int RUN\_get\_double\_parameter(const char *key, double * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The number
-of key matches found is returned (i.e., 0 or 1).
-
-\texttt{int RUN\_get\_int\_parameter(const char * key, int * value)}\\
-Set the \texttt{value} associated with \texttt{key}, if present. The
-number of key matches is returned.
-
-\texttt{int  RUN\_get\_string\_parameter(const char * key, char * value)}\\
-Set \texttt{value} to the string assocaited with \texttt{key}, if
-present. Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_int\_parameter\_vector(const char * key, int v[])}\\
-If \texttt{key} is present, set the associated 3-vector of integers.
-Returns the number of key matches found.
-
-\texttt{int  RUN\_get\_double\_parameter\_vector(const char * key,
-double v[])}\\
-If \texttt{key} is present, set the associated 3-vector of doubles.
-Returns the number of key matches found.
-
-\textit{Comment:} We probably need a function to report unused keys, so
-the user can be alerted if a key has unexpectedly been ignored.
-
-\subsection{Random Number Generators}
-
-\texttt{\#include ``ran.h''}
-
-Note that the random number generation in parallel is
-decomposition dependent.
-
-\texttt{void rand\_init(void)}\\
-Responsible for initialising RNG state.
-
-\texttt{double rand\_serial\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$ from
-the serial generator.
-
-\texttt{double rand\_serial\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the serial generator.
-
-\textit{double rand\_parallel\_uniform(void)}\\
-Return a single variate uniformly distributed on $[0,1]$
-from the serial generator.
-
-\texttt{double rand\_parallel\_gaussian(void)}\\
-Return a single variate with Gaussian distribution about zero
-and having unit variance from the parallel generator.
-
-\textit{Comment:} The serial generator will produce the
-same sequence on any MPI process, while the parallel generator
-will produce different sequences. The latter is intended for
-generating, specifically, fluctuations, where the need for
-decomposition-independance is questionable.
-
-\subsection{Timers}
-
-\texttt{\#include ``timer.h''}
-
-Performance figures require timing for different sections
-of the code. The code provides a set of routines to time
-standard and user-defined sections of code. Timers are
-started and stopped by call to the appropriate routines.
-
-\texttt{TIMER\_init()}\\
-This initialises all the timers to zero (e.g., at the start of
-the program).
-
-\texttt{TIMER\_start(const int t\_id)}\\
-Start the given timer.
-
-\texttt{TIMER\_stop(const int t\_id)}\\
-Stop the given timer.
-
-\texttt{TIMER\_statistics(void)}\\
-Prints the current state of all the active timers to standard
-output on the root process in a digestible form. In serial,
-the minimum, maximum, and mean time for each timer is reported
-in seconds (based on the number of start/stop cycles). In
-parallel, the same figures are computed on each process and the
-result of the corresponding \texttt{MPI\_Reduce()} operation is
-printed on the root process.
-
-The code provides a number of timers which would time standard
-parts of the code, e.g., \texttt{t\_id = TIMER\_TOTAL} for the
-total execution time, and a number of spare timers for users to
-use at will.
-
-\textit{Comment:}
-No intra-run variance (i.e., between time steps) is provided, as the
-calculation required is considerably more complex to generalise, and
-requires memory proportional to the number of time steps used.
-
-\textit{Comment:}
-Two macros, \texttt{TIMER\_START()} and \texttt{TIMER\_STOP()} are
-provided to insert calls to the above functions for optional timing
-of performance-sensitive regions
-of the code in conjunction with the preprocessor flag
-\texttt{-D\_TIMING\_MACRO\_ON\_}.
-
-\subsection{Model Time Control}
-
-This provides time step control, event frequency (?)
-
-\texttt{void init\_control(void)}\\
-Initialise time controls.
-
-\texttt{int get\_step(void)}\\
-Return the current model time step.
-
-\texttt{int next\_step(void)}\\
-Increments the current time step. Returns number of steps left
-to run (i.e., 0 when finished).
-
-\subsection{Coordinate System}
-
-\texttt{\#include ``coords.h''}
-
-A three-dimensional Cartesian coordinate system is used throughout,
-with the exact system size usually chosen by the user at run time.
-The symbolic constants \texttt{X}, \texttt{Y}, and \texttt{Z} are
-used to identify the coordinate directions.
-
-This also provides access to the MPI Cartesian Communicator which
-is responsible for most communication in the model.
-
-\texttt{int N\_total(const int dim)}\\
-Returns the total number of lattice sites in the given Cartesian direction.
-
-\texttt{double L(const int dim)}\\
-Returns the length of the system in the given Cartesian direction
-(which is the same as the number of lattice sites).
-
-\texttt{double Lmin(const int dim)}\\
-Returns coordinate position of left-hand edge of system in given
-Cartesian direction.
-
-\texttt{int is\_periodic(const int dim)}\\
-Returns 1 if given Cartesian direction is periodic, else 0.
-
-\texttt{void cart\_init(void)}\\
-Responsible for setting up the Cartesian decomposition.
-
-\texttt{int cart\_rank(void)}\\
-Returns rank in the Cartesian communicator.
-
-\texttt{MPI\_Comm cart\_comm(void)}\\
-Returns the MPI Communicator handle for the Cartesian communicator.
-
-\texttt{int cart\_size(const int dim)}\\
-Returns the size of the Cartesian communicator grid in the given
-coordinate direcxtion.
-
-\texttt{int cart\_coord(const int dim)}\\
-Returns the coordinate of the current process in the Cartesian
-Communicator.
-
-\texttt{int cart\_neigh(const int dir, const int dim)}\\
-Returns the rank of the neighbouring process in the
-Cartesian communicator.
-
-\subsection{Colloids}
-
-\texttt{\#include ``colloids.h''}
-
-These functions provide the basic interface for use
-of colloidal particles in the code.
-
-\subsubsection{Colloid structure}
-
-The basic colloid struct is defined ...
-
-\subsubsection{Colloid interface}
-
-\texttt{void colloids\_init(void)}\\
-Call once for initialisation. This initialises the cell list,
-so the coordinate system must be initialised via a call to
-\texttt{coords\_init()} first.
-
-\texttt{void colloids\_finish(void)}\\
-Call once for finalisation. Frees all colloid-related memory.
-
-\texttt{Colloid * allocate\_colloid(void)}\\
-Returns a pointer to a newly allocated \texttt{Colloid} object
-or fails gracefully.
-
-\texttt{void free\_colloid(Colloid * p\_colloid)}\\
-Deallocates all memory associated with \texttt{p\_colloid} or
-fails gracefully.
-
-\texttt{struct c\_link allocate\_boundary\_link(void)}\\
-Returns a pointer to a newly allocated colloid boundary
-link or fails gracefully.
-
-\texttt{void free\_boundary\_link(struct c\_link *)}\\
-Deallocates memory associated with a link or fails gracefully.
-
-\texttt{void colloids\_memory\_report(void)}\\
-The module keeps track of how much memory has been allocated in
-total for colloids and boundary links. This reports the current
-usage.
-
-\texttt{get\_Ncell\_total()}\\
-
-\texttt{get\_Ncell\_local()}\\
-
-\texttt{Colloid * cell\_get\_colloid(int i, int j, int k)}\\
-This returns a pointer to the Colloid at the head of the list at
-cell list location (i, j, k) or \texttt{NULL} if there is none.
-
-\texttt{void cell\_insert(Colloid * p\_colloid)}\\
-This inserts a colloid at the head of the appropriate list
-depending upon its position. This is the only way a colloid
-can be added to the list.
-
-\texttt{update\_cells(void)}\\
-Performs the update of the cell list in response to updated
-particle positions.
-
-\texttt{IVector cell\_coords(FVector r)}\\
-
-\subsection{Speculation}
-
-\subsection{The Free Energy}
-
-It is imagined that a single file will encapsulate all the details
-of a given free energy. The interface for different choices of
-free energy should be the same.
-
-\texttt{FE\_init()}
-
-Responsible for querying the runtime input key value pair list to
-find appropriate user parameters for the free energy, or
-falling back to default values.
-
-\texttt{double FE\_get\_sigma()}
-
-Return the surface tension. There should be a similar routine
-for the interfacial thickness.
-
-\texttt{double FE\_get\_mu(const type location)}
-
-Return the chemical potential at the single lattice
-node \texttt{location}.
-
-\textit{Comment:} The computation of quantities such as the chemical
-potential will a) affect performance and b) require halo values of
-the order parameter (depending on the highest derivative
-in the order parameter present). Performance here should be the
-first consideration, so this needs to be handled with care.
-
-\texttt{double pchem[ND][ND] FE\_get\_pchem(const type location)}
-
-Return the chemical stress ${\cal P}_{\alpha\beta}$ at the
-given location.
-
-\textit{Comment:} This is an $nd \times nd$ matrix, where $nd$ is
-the number of dimensions. If we are to provide explicit support for
-D2Q9, \texttt{ND} will be 2, otherwise 3.
-
-\section{Implementation Notes}
-
-\subsection{Serial Version}
-
-\subsection{Efficiently Scaling Parallel Computing}
-
-We should be thinking about the 100,000 processor regime, and
-correspondingly large system sizes. To avoid embarrassment in this regime,
-some issues which don't usually arise are relevant, for example:
-
-\begin{enumerate}
-\item
-Large integer quantities such as
-\begin{verbatim}
-   N_total.x*N_total.y*N_total.z
-\end{verbatim}
-are likely to cause integer overflow (if default integers are 4 byte).
-\item
-Local data structures whose size is proportional
-to the number of processors are contraindicated, e.g.,
-anything like
-\begin{verbatim}
-p = (foo *) malloc(nprocs*sizeof(foo));
-\end{verbatim}
-is to be avoided. At the moment, we are in good shape on this
-(apart from some functions in \texttt{misc.c}).
-\item
-I/O is likely become a very important issue, so may need to think
-more about this.
-\item
-We should probably introduce a standard set of benchmarks which can
-be scaled up in some systematic way, so that we can keep track of
-performance over time.
-
-\end{enumerate}
-
-\section{Model Details}
-
-\subsection{D2Q9}
-
-These are the eigenvalues and eigenvectors of the collision
-matrix used for D2Q9
-
-\begin{table}
-\begin{tabular}{|l|r|rrrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{9}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - & 1 &  1 &  1 &  1 &  1 &  1 &  1 &   1 &  1 & 1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - & 0 &  1 &  1 & 1 & 0 &  0 & -1 &  1 & -1 & 3 & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - & 0 & 1 &  0 &  -1 &  1 &  -1 & 1 & 0 & -1 & 3  &$c_{iy}$ \\
-\hline
-$Q_{xx}$ & 1/3 & -1 &  2 &  2 & 2 & -1 & -1 & 2 & 2 & 2 & 9/2 
-& $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - & 0 &  1 & 0 & -1 & 0 & 0 & -1 & 0 & 1 & 9 & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yy}$ & 1/3 & -1 &  2 & -1 & 2 & 2 & 2 & 2 & -1 & 2 & 9/2
-& $c_{iy} c_{iy} - c_s^2$ \\
-\hline\hline
-$\chi^1$ & - &  1 & 4 & -2 & 4 & -2 & -2 & 4 & -2 & 4 & 1/4 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - & 0 &  4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - & 0 & 4 & 0 & -4 & -2 & 2 & 4 & 0 & -4 & 3/8
-& $\chi^1 \rho c_{iy}$\\
-\hline\hline
-$w_i$ & 1/36 & 16 & 1 & 4 & 1 & 4 & 4 & 1 & 4 & 1 & & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q15}
-
-These are the eigenvalues and eigenvectors of teh collision
-matrix used for D3Q15
-
-\begin{table}
-\begin{tabular}{|l|r|r|rrrrrr|rrrrrrrr|r|l|}
-\hline\hline
-$M^a$ & & \multicolumn{15}{c|}{$m_i^a$} & $N^a$  &\\
-\hline
-$\rho$ & - &
- 1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &  1 &
-1 &$\mathbf{1}$ \\
-\hline
-$\rho c_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 &  1 & -1 &  1 & -1 &
-3  & $c_{ix}$ \\
-\hline
-$\rho c_{iy}$ & - &
- 0 &  0 &  0 &  1 & -1 &  0 &  0 &  1 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &
-3  &$c_{iy}$ \\
-\hline
-$\rho c_{iz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 &  1 &  1 &  1 & -1 & -1 & -1 & -1 &
-3  & $c_{iz}$ \\
-\hline
-$Q_{xx}$ & 1/3 &
--1 &  2 &  2 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
-9/2  & $c_{ix} c_{ix} - c_s^2$ \\
-\hline
-$Q_{yy}$ & 1/3 &
--1 & -1 & -1 &  2 &  2 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iy} c_{iy} - c_s^2$ \\
-\hline
-$Q_{zz}$ & 1/3 &
--1 & -1 & -1 & -1 & -1 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &  2 &
- 9/2 & $c_{iz} c_{iz} - c_s^2$ \\
-\hline
-$Q_{xy}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 &  1 & -1 & -1 &  1 &
-9  & $c_{ix} c_{iy}$ \\
-\hline
-$Q_{yz}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 &  1 & -1 & -1 & -1 & -1 &  1 &  1 &
-9  & $c_{iy} c_{iz}$ \\
-\hline
-$Q_{zx}$ & - &
- 0 &  0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 &  1 & -1 & -1 &  1 & -1 &  1 &
-9  & $c_{iz} c_{ix}$ \\
-\hline\hline
-$\chi^1$ & - &
--2 &  1 &  1 &  1 &  1 &  1 &  1 & -2 & -2 & -2 & -2 & -2 & -2 & -2 & -2 &
-1/2 & $\chi^1$ \\
-\hline
-$J_{ix}$ & - &
- 0 &  1 & -1 &  0 &  0 &  0 &  0 & -2 &  2 & -2 &  2 & -2 &  2 & -2 &  2 &
-3/2 & $\chi^1 \rho c_{ix}$\\
-\hline
-$J_{iy}$ & - &
- 0 &   0 &  0 &  1 & -1 &  0 &  0 & -2 & -2 &  2 &  2 & -2 & -2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iy}$\\
-\hline
-$J_{iz}$ & - &
- 0 &   0 &  0 &  0 &  0 &  1 & -1 & -2 & -2 & -2 & -2 &  2 &  2 &  2 &  2 &
-3/2 & $\chi^1 \rho c_{iz}$\\
-\hline
-$\chi^3$ & - &
- 0 &   0 &  0 &  0 &  0 &  0 &  0 &  1 & -1 & -1 &  1 & -1 &  1 &  1 & -1 &
-9 & $c_{ix} c_{iy} c_{iz}$ \\
-\hline\hline
-$w_i$ & 1/72 &
-$16$ & 8 & 8 & 8 & 8 & 8 & 8 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 &
- & $w_i$\\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\subsection{D3Q19}
-
-\begin{table}
-\begin{tabular}{|l||r|rrrrrr|rrrr|rrrr|rrrr|r||}
-\hline\hline
-$M^a$ & \multicolumn{19}{c||}{$m_i^a$} & $N^a$\\
-\hline
-$\rho $ & 1 &  1 &  1 &  1 &  1 &  1 &  1 & 
-         1 &  1 &  1 &   1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1
-& 1\\
-\hline
-$\rho c_{ix}$ & 0 &  1 &  -1 &  0 &  0 &  0 &  0 & 
-         1 &  1 &  -1 &   -1 &  1 &  1 &  -1 & -1 & 0 & 0 & 0 & 0
-& 3 \\
-\hline
-$\rho c_{iy}$ & 0 &  0 &  0 &  1 &  -1 &  0 &  0 & 
-         1 &  -1 &  1 &   -1 &  0 &  0 &  0 & 0 & 1 & 1 & -1 & -1
-& 3\\
-\hline
-$\rho c_{iz}$ & 0 &  0 &  0 &  0 &  0 &  1 &  -1 & 
-         0 &  0 &  0 &   0 &  1 &  -1 &  1 & -1 & 1 & -1 & 1 & -1
-& 3\\
-\hline
-$Q_{ixx}$ & -1 &  2 &  2 &  -1&  -1 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  2 &  2 &  2 & 2 & -1 & -1 & -1 & -1
-& 9/2\\
-\hline
-$Q_{iyy}$ & -1 &  -1 &  -1 &  2&  2 &  -1 &  -1 & 
-         2 &  2 &  2 &   2 &  -1 &  -1 &  -1 & -1 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{izz}$ & -1 &  -1 &  -1 &  -1&  -1 &  2 &  2 & 
-         -1 &  -1 &  -1 &   -1 &  2 &  2 & 2 & 2 & 2 & 2 & 2 & 2
-& 9/2\\
-\hline
-$Q_{ixy}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          1 &  -1 &  -1 &    1 &  0 &  0 & 0 & 0 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{ixz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  1 & -1 & -1 & 1 & 0 & 0 & 0 & 0
-& 9\\
-\hline
-$Q_{iyz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-          0 &   0 &   0 &   0 &  0 & 0 & 0 & 0 & 1 & -1 & -1 & 1
-& 9\\
-\hline\hline
-$\chi^1$ & 0 &  1 &  1 &  1 &  1 &  -2 &  -2 & 
-         -2 &  -2 &  -2 &  -2 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 3/4\\
-\hline
-$\chi^1 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         -2 &  -2 &  2 &  2 &  1 &  1 & -1 & -1 & 0 & 0 & 0 & 0
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iy}$ & 0 &  0 &  0 &  1&  -1 &  0 &  0 & 
-         -2 &  2 &  -2 &  2 &  0 &  0 & 0 & 0 & 1 & 1 & -1 & -1
-& 3/2\\
-\hline
-$\chi^1 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  -2 &  2 & 
-         0 &  0 &  0 &  0 &  1 &  -1 & 1 & -1 & 1 & -1 & 1 & -1
-& 3/2\\
-\hline
-$\chi^2$ & 0 &  1 &  1 &  -1&  -1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & -1 & -1 & 1 & 1 & 1 & 1
-& 9/4\\
-\hline
-$\chi^2 \rho c_{ix}$ & 0 &  1 &  -1 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  -1 & 1 & 1 & 0 & 0 & 0 & 0
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iy}$ & 0 &  0 &  0 & -1&   1 &  0 &  0 & 
-         0 &  0 &  0 &  0 &   0 &  0 & 0 & 0 & 1 &  1 & -1 & -1
-& 9/2\\
-\hline
-$\chi^2 \rho c_{iz}$ & 0 &  0 &  0 &  0&  0 &  0 &  0 & 
-         0 &  0 &  0 &  0 &  -1 &  1 & -1 & 1 & 1 & -1 & 1 & -1
-& 9/2\\
-\hline
-$\chi^3$ & 1 &  -2 &  -2 &  -2&  -2 &  -2 &  -2 & 
-         1 &  1 &  1 &  1 &  1 &  1 & 1 & 1 & 1 & 1 & 1 & 1
-& 1/2\\
-\hline\hline
-$w_i$ & 12 & 2 & 2 & 2 & 2 & 2 & 2 & 
-1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1
-& \\
-\hline\hline
-\end{tabular}
-\end{table}
-
-\pagebreak
-
-\begin{thebibliography}{99}
-
-\bibitem{adhikari_desplat}
-Adhikari, R. J.-C. Desplat, and K. Stratford,
-Sliding periodic boundary conditions for lattice Boltzmann and lattice
-kinetic equations,
-\texttt{arXiv:cond-mat/0503175v1} (2005).
-
-\bibitem{adhikari2005}
-Adhikari, R., K. Stratford. M.E. Cates, and A.J. Wagner,
-Fluctuating Lattice Boltzmann,
-\textit{Europhys. Lett.}, \textbf{71}, 473 2005.
-
-\bibitem{ald98}
-Aidun, C.K., Y. Lu, and E.-J. Ding,
-Direct analysis of particulate suspensions with inertia using the
-discrete Boltzmann equation,
-\textit{J. Fluid Mech.}, \textbf{373}, 287, 1998.
-
-\bibitem{cates_scaling}
-M. E. Cates, J.-C. Desplat, P. Stansell, A.J. Wagner, K. Stratford,
-R. Adhikari, and I. Pagonabarraga,
-Physical and Computational Scaling Issues in Lattice Boltzmann
-Simulations of Binary Fluid Mixtures,
-\textit{Phil. Trans. Roy. Soc. A}, \textbf{363}, 1917 (2005).
-
-\bibitem{chunladd}
-B. Chun and A.J.C. Ladd,
-Interpolated boundary condition for lattice Boltzmann simulations in
-narrow gaps,
-\textit{Phys. Rev. E}, \textbf{75}, 066705, 2007.
-
-\bibitem{cj98}
-Cichocki, B., and R.B. Jones,
-Image representation of a spherical particle near a hard wall,
-\textit{Physica A}, \textbf{258}, 273, 1998.
-
-\bibitem{chang}
-C.-H. Chang and E.I. Franses,
-Adsorption dynamics of surfactants at the air/water interface:
-a critical review of mathematical models, data, and mechanisms,
-\textit{Colloids and Surfaces A}, \textbf{100} 1, 1995.
-
-\bibitem{cpc}
-Desplat, J.-C., I. Pagonabarraga, and P. Bladon,
-LUDWIG: A parallel lattice-Boltzmann code for complex fluids.
-\textit{Comput. Phys. Comms.}, \textbf{134}, 273, 2001.
-
-\bibitem{diamant}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid/fluid
-interfaces: non-ionic surfactants,
-\textit{Europhys. Lett.} \textbf{34}, 575 (1996).
-
-\bibitem{diamant96}
-H. Diamant and D. Andelman,
-Kinetics of surfactant adsorption at fluid-fluid interfaces,
-\textit{J. Phys. Chem.}, \textbf{100} 13732, 1996.
-
-\bibitem{eastoe}
-J. Eastoe and J.S. Dalton,
-Dynamic surface tension and adsorption mechanims of surfactants
-at the air-water interface,
-\textit{Advances in Colloid and Interface Science}, \textbf{85}
-13, 2000.
-
-\bibitem{ginzburg}
-I. Ginzburg and D. d'Humi\`eres,
-Multireflection boundary conditions for lattice Boltzmann models,
-\textit{Phys. Rev. E}, \textbf{68}, 066614, 2003.
-
-\bibitem{h59}
-Hasimoto, H., On the periodic fundamental solutions of the Stokes
-equation and their application to viscous flow past a cubic array
-of spheres.
-\textit{J. Fluid Mech.}, \textbf{5}, 317.
-
-\bibitem{heemels}
-Heemels, M.W., M.H.J. Hagen, and C.P. Lowe, Simulating solid colloidal
-particles using the lattice-Boltzmann method,
-\textit{J. Comp. Phys.}, \textbf{164}, 48, 2000.
-
-\bibitem{jo84}
-Jeffrey, D.J., and Y. Onishi,
-Calculation of the resistance and mobility functions for the two
-unequal rigid spheres in low-Reynolds-number flow,
-\textit{J. Fluid Mech.}, \textbf{139}, 261, 1984.
-
-\bibitem{viv}
-Kendon, V.M., M.E. Cates, I. Pagonabarraga, J.-C. Desplat, and
-P. Bladon,
-Inertial effects in three dimensional spinodal decomposition of
-a symmetric binary fluid mixture: A lattice Boltzmann study,
-\textit{J. Fluid Mech.}, \textbf{440}, 147 (2001).
-
-\bibitem{l94a}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 1. Theoretical foundation,
-\textit{J. Fluid. Mech.}, \textbf{271}, 285, 1994.
-
-\bibitem{l94b}
-Ladd, A.J.C., Numerical simulations of particulate suspensions
-via a discretised Boltzmann equation. Part 2. Numerical results,
-\textit{J. Fluid. Mech.}, \textbf{271}, 311, 1994.
-
-\bibitem{l96a}
-Ladd, A.J.C., Sedimentation of homogenous suspensions of non-Brownian
-spheres,
-\textit{Phys. Fluids}, \textbf{9}, 491. 1996.
-
-\bibitem{l96b}
-Ladd, A.J.C., Hydrodynamic screening in sedimentating suspensions
-of non-Brownian spheres,
-\textit{Phys. Rev. Lett.}, \textbf{76}, 1392, 1996.
-
-\bibitem{lv01}
-Ladd, A.J.C., and R. Verberg,
-Lattice-Boltzmann simulations of particle-fluid suspensions,
-\textit{J. Stat. Phys.}, \textbf{104}, 1191, 2001.
-
-\bibitem{lipanmiller}
-H. Li, C. Pan, and C.T. Miller,
-Pore-scale investigation of viscous coupling effects for two-phase
-flow in porous media,
-\textit{Phys. Rev. E}, \textbf{72}, 026705, 2005.
-
-\bibitem{nl02}
-Nguyen, N.-Q., and A.J.C. Ladd, Lubrication corrections for
-lattice-Boltzmann simulations of particle suspensions,
-\textit{Phys. Rev. E}, \textbf{66}, 046708, 2002.
-
-\bibitem{papanastasiou}
-T. paapnastasiou, G. Georgiou, and A. Alexandrou,
-\textit{Viscous Fluid Flow},
-CRC Press, Boca Raton, Florida, 2000.
-
-\bibitem{r95}
-Rapaport, D.C., \textit{The Art of Molecular Dynamics Simulation},
-Cambridge University Press, 1995.
-
-\bibitem{succi}
-S. Succi, \textit{The lattice Boltzmann equation and beyond},
-Oxford University Press, Oxford, 2001.
-
-\bibitem{edo1}
-M. Venuroli and E.S. Boek,
-Two-dimensional lattice-Boltzmann simulations of single phase
-flow in a pseudo two-dimensional micromodel,
-\textit{Physica A}, \textbf{362}, 23, 2006.
-
-\bibitem{vandergraaf}
-R.G.M. van der Sman and S. van der Graaf,
-Diffuse interface model of surfactant adsorption onto flat and
-droplet interfaces,
-\textit{Rheol. Acta} \textbf{46} 3 (2006).
-
-\bibitem{theissengompper}
-O. Theissen and G. Gompper,
-Lattice Boltzmann study of spontaneous emulsification,
-\textit{Eur. Phys. J. B}, \textbf{11} 91 (1999).
-
-\bibitem{wardtordai}
-A.F.H. Ward and L. Tordai,
-\textit{J. Chem. Phys.} \textbf{14} 453, 1946.
-
-\end{thebibliography}
-
-\end{document}

### 2017-02-14 docs/plan.tex
Additional notes on addressing model
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2765 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
+%  See, for example, Annex D of IEEE 828-2012
+%  IEEE Standard for Configuration Management in Systems and Software
+%  Engineering
+%
 %  Edinburgh Soft Matter and Statistical Physics Group and
@@
 %
+%  (c) 2011-2015 The University of Edinburgh
+%  Contributing authors:
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2011-2015 The University of Edinburgh
 %
@@
 
-The \textit{Ludwig} code is designed to study the hydrodynamic properties
-of simple and complex fluids based around the numerical solution of the
-Navier-Stokes equations. Complex fluids are dealt with via a free-energy
-based finite difference approach, which couples explicitly to the
-hydrodynamics.
+The \textit{Ludwig} code is designed to study simple and complex fluids
+based around the numerical solution of the Navier-Stokes equations.
 Components of \textit{Ludwig} include the main program, unit and
@@
 
-\textit{Ludwig} is specifically designed for parallel computers and
-relies on the Message Passing Interface \cite{mpi-standard}. The
-code will also run in serial, and is independent of
-the platform it is run on. The Plan is not concerned with
-hardware or system management activities.
+% IEEE section to be completed
+\subsubsection{Relationship to organisation and other projects}
 
@@
 
-This section is based on the IEEE standard for Software Configuration
-Management Plans IEEE 828-2012 \cite{ieee-828-2012}, and follows the format
+This section is based on the IEEE standard for Configuration
+Management IEEE 828-2012 \cite{ieee-828-2012}, and follows the format
 set out therein.
@@
 section at the end of the document.
+%
 
-\subsection{Management}
+% Criteria for identication
 
-\subsubsection{Organisations}
+\subsection{Identification}
 
-\textit{Ludwig} is developed by a team based in the School of Physics at
-The University of Edinburgh. This involves two groups: the Soft Matter
-Physics Group, and Edinburgh Parallel Computing Centre (EPCC). These
-groups collaborate with a number of workers at different Universities
-around the world on different aspects of the code.
-
-\subsubsection{Responsibilities}
-
-While the Soft Matter Physics Group is responsible for the overall
-scientific direction and development of the code and concomitant
-priorities, all the Software Configuration Management activities will
-be the responsibility of EPCC.
-
-% IEEE section recheck what this actually means
-% \subsubsection{Procedures}
-% There are currently no external constraints on The Plan.
-
-\subsubsection{Process Management}
-
-EPCC is responsible for the Software Configuration Management process.
-It is anticipated that the cost of this process will be small as the
-project team can communicate informally on a regular basis. There is
-currently no independent
-surveillance of activities to ensure compliance with The Plan.
-
-\subsection{Activities}
-
-\subsubsection{Identification}
-
-The material under control of the project includes three sets of files:
-this documentation, the source code, and the test suite which is used
-to monitor the status of the code.
+The material under control of the project includes a number of sets
+of files: this documentation, the source code, the build and test
+suite which is used to monitor the status of the code, and so on.
 Control of project files is via a single on-line repository which
@@
 
+Specifically, configuration items are taken to include files in the
+trunk of the repository
+
+\texttt{http://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk}
+
+Other branches in the repository are not considered to be 
+under configuration management.
+
+% IEEE section
+% Limitations and assumptions
+% Deliberately empty at the moment.
+
+\subsection{Responsibility and authority}
+
+While the Soft Matter Physics Group is responsible for the overall
+scientific direction and development of the code and concomitant
+priorities, all Configuration Management activities will
+be the responsibility of EPCC at The University of Edinburgh.
+
+\subsection{Project organisation}
+
+\subsubsection{Organisations}
+
+\textit{Ludwig} is developed by a team based in the School of Physics at
+The University of Edinburgh. This involves two groups: the Soft Matter
+Physics Group, and Edinburgh Parallel Computing Centre (EPCC). These
+groups collaborate with a number of workers at different Universities
+around the world on different aspects of the code.
+
+\subsubsection{Process Management}
+
+EPCC is responsible for the Software Configuration Management process.
+It is anticipated that the cost of this process will be small as the
+project team can communicate informally on a regular basis. There is
+currently no independent
+surveillance of activities to ensure compliance with The Plan.
+
 % Add pointer to standard manifest (README)
 
-All access to the code is via the CCPForge repository. CCPForge identifies
-three different levels of access:
+%All access to the code is via the CCPForge repository. CCPForge identifies
+%three different levels of access:
 
-\textit{Administrators:} allows administration of the site itself and
-registered members;
+%\textit{Administrators:} allows administration of the site itself and
+%registered members;
 
-\textit{Developers:} members who have write access to the source code repository;
+%\textit{Developers:} members who have write access to the source code repository;
 
-\textit{Users:} have read only access to the source code via SVN (either
-as a member or via anonymous download). Member users also have access to
-the tracker.
+%\textit{Users:} have read only access to the source code via SVN (either
+%as a member or via anonymous download). Member users also have access to
+%the tracker.
 
-As CCPForge is not physically under the control of EPCC, a
-complete dump of the repository to a local resource is undertaken on a
-regular basis for security.
+%As CCPForge is not physically under the control of EPCC, a
+%complete dump of the repository to a local resource is undertaken on a
+%regular basis for security.
 
@@
 %\begin{thebibliography}{99}
-%\bibitem{ieee-828-2012} IEEE Standard 828-2011 ``IEEE Standard for
+
 %Software Configuration Management Plans''
@@
 %\bibitem{paraview} Paraview is a standard visualisation package.
-%\bibitem{apacheAPR} See, e.g., the Apache Portable Runtime Project ``APR's Version Numbering''  https://apr.apache.org/versioning.html (Accessed 2015).
-%\end{thebibliography}

### 2017-02-14 targetDP/doc/data_parallel_execution.tex

@@
-\chapter{Data Parallel Execution}
+\chapter{Data Parallel Execution}\label{sec:dpexecution}
+
+This chapter specifices the data parallel execution functionality in targetDP.
 
@@
 executed in parallel and mapped to instruction level parallelism
-(ILP), where the extent of the ILP is defined by the virtual vector length (VVL) in the targetDP implementation. 
-\begin{comment}
-need to document VVL in more detail somewhere and link.
-\end{comment}
+(ILP), where the extent of the ILP is defined by the virtual vector length (VVL) in the targetDP implementation (see \ref{sec:exmodel}).
 
@@
 
-\subsection{Example}
-**TO DO**
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**
-\subsubsection{CUDA}
-**TO DO**
-
 \newpage
@@
 \end{itemize}
-
-\subsection{Example}
-**TO DO**
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**
-\subsubsection{CUDA}
-**TO DO**

### 2017-02-14 targetDP/doc/memory_management.tex

@@
-\chapter{Memory Management}
+\chapter{Memory Management}\label{chap:memmanage}
+
+This chapter specifices the memory management functionality in targetDP.
 
@@
 \newpage
+
+\section{targetMallocUnified}
+
+\subsection{Description}
+
+The \verb+targetMallocUnified+ function allocates unified memory that can be accessed on the host or the target.
+This should be used with caution since it may result in poor performance relative to use of \verb+targetMalloc+.
+
+\subsection{Syntax}
+\begin{verbatim}
+void targetMallocUnified(void** targetPtr, size_t n);
+\end{verbatim}
+
+\begin{itemize}
+\item \verb+targetptr+: A pointer to the allocated memory.
+\item \verb+n+: The number of bytes to be allocated.
+\end{itemize}
+
+\subsection{Example}
+Analogous to Line \ref{example:scalelaunch:targetMalloc} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.
+\subsection{Implementation}
+\subsubsection{C}
+\verb+malloc+
+\subsubsection{CUDA}
+\verb+cudaMallocManaged+
+
+\newpage
+\section{targetCallocUnified}
+
+\subsection{Description}
+
+The \verb+targetCallocUnified+ function allocates, and initialises to zero, unified memory that can be accessed on the host or the target.
+This should be used with caution since it may result in poor performance relative to use of \verb+targetCalloc+.
+
+\subsection{Syntax}
+\begin{verbatim}
+void targetCallocUnified(void** targetPtr, size_t n);
+\end{verbatim}
+
+\begin{itemize}
+\item \verb+targetptr+: A pointer to the allocated memory.
+\item \verb+n+: The number of bytes to be allocated.
+\end{itemize}
+
+\subsection{Example}
+Analogous to Line \ref{example:scalelaunch:targetMalloc} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.
+
+\subsection{Implementation}
+\subsubsection{C}
+\verb+calloc+
+\subsubsection{CUDA}
+\verb+cudaMallocManaged+ followed by \verb+cudaMemset+
+
+\newpage
 \section{targetFree}
@@
 \newpage
+\section{copyDeepDoubleArrayToTarget}
+
+\subsection{Description}
+
+The \verb+copyDeepDoubleArrayToTarget+ function copies an array of double precision values from the host to the target, where the array is contained within another object.
+
+\subsection{Syntax}
+\begin{verbatim}
+void copyDeepDoubleArrayToTarget(void* targetObjectAddress,
+    void* hostObjectAddress,void* hostComponentAddress,size_t n);
+\end{verbatim}
+
+\begin{itemize}
+\item \verb+targetObjectAddress+: A pointer to the target copy of the object that contains the data array.
+\item \verb+hostObjectAddress+: A pointer to the host copy of the object that contains the data array.
+\item \verb+hostComponentAddress+: A pointer to the host copy of the start of the array contained within the object.
+\item \verb+n+: The number of elements to be copied.
+\end{itemize}
+
+\subsection{Implementation}
+\subsubsection{C}
+Pointer arithmetic to determine memory locations, followed by \verb+memcpy+
+\subsubsection{CUDA}
+Pointer arithmetic to determine memory locations, followed by \verb+cudaMemcpy+
+
+\newpage
+\section{copyDeepDoubleArrayFromTarget}
+
+\subsection{Description}
+
+The \verb+copyDeepDoubleArrayTFromTarget+ function copies an array of double precision values from the target to the host, where the array is contained within another object.
+
+\subsection{Syntax}
+\begin{verbatim}
+void copyDeepDoubleArrayFromTarget(void* hostObjectAddress,
+   void* targetObjectAddress,void* hostComponentAddress,size_t n);
+\end{verbatim}
+
+\begin{itemize}
+\item \verb+hostObjectAddress+: A pointer to the host copy of the object that contains the data array.
+\item \verb+targetObjectAddress+: A pointer to the target copy of the object that contains the data array.
+\item \verb+hostComponentAddress+: A pointer to the host copy of the start of the array contained within the object.
+\item \verb+n+: The number of elements to be copied.
+\end{itemize}
+
+\subsection{Implementation}
+\subsubsection{C}
+Pointer arithmetic to determine memory locations, followed by \verb+memcpy+
+\subsubsection{CUDA}
+Pointer arithmetic to determine memory locations, followed by \verb+cudaMemcpy+
+
+\newpage
+\section{targetZero}
+
+\subsection{Description}
+
+The \verb+targetZero+ function sets a (double precision) array on the target to zero.
+
+\subsection{Syntax}
+\begin{verbatim}
+void targetZero(double* targetData, size_t n);
+\end{verbatim}
+
+\begin{itemize}
+\item \verb+targetData+: A pointer to the array on the target.
+\item \verb+n+: The number of elements in the array.
+\end{itemize}
+
+\subsection{Implementation}
+\subsubsection{C}
+A loop to zero each element.
+\subsubsection{CUDA}
+A kernel to zero each element.
+
+\newpage
+\section{targetSetConstant}
+
+\subsection{Description}
+
+The \verb+targetSetConstant+ function sets each element of a (double precision) array on the target to the specified constant value.
+
+\subsection{Syntax}
+\begin{verbatim}
+void targetSetConstant(double* targetData, double value, size_t n);
+\end{verbatim}
+
+\begin{itemize}
+\item \verb+targetData+: A pointer to the array on the target.
+\item \verb+value+: The value.
+\item \verb+n+: The number of elements in the array.
+\end{itemize}
+
+\subsection{Implementation}
+\subsubsection{C}
+A loop to set each element.
+\subsubsection{CUDA}
+A kernel to set each element.
+
+\newpage
 \section{targetConst}
@@
 \subsection{Example}
-**TO DO**.
+See Line \ref{example:scalekernel:targetConst} in Figure \ref{fig:scalelaunch} in Section \ref{chapter:examples}.
 \subsection{Implementation}
@@
 \end{itemize}
-
-\subsection{Example}
-**TO DO**.
 
@@
 
-\subsection{Example}
-**TO DO**.
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**.
-\subsubsection{CUDA}
-**TO DO**.
-
 \newpage
@@
 
-\subsection{Example}
-**TO DO**.
+%% \newpage
+%% \section{copyToTargetBoundary3D}
 
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**.
-\subsubsection{CUDA}
-**TO DO**.
+%% \subsection{Description}
 
-\newpage
-\section{copyToTargetBoundary3D}
+%% The \verb+copyToTargetBoundary3D+ function copies the data corresponding to the boundaries of a 3D lattice from the host to the target.
 
-\subsection{Description}
+%% \subsection{Syntax}
+%% \begin{verbatim}
+%% void copyToTargetBoundary3D(void* targetData, const void* data, 
+%%               size_t extent3D[3], size_t nField, 
+%%               size_t offset, size_t depth);
+%% \end{verbatim}
 
-The \verb+copyToTargetBoundary3D+ function copies the data corresponding to the boundaries of a 3D lattice from the host to the target.
+%% \begin{itemize}
+%% \item \verb+targetData+: A pointer to the destination array on the target.
+%% \item \verb+data+: A pointer to the source array on the host.
+%% \item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
+%% \item \verb+nFields+: The number of fields per lattice site.
+%% \item \verb+offset+: The number of sites from the lattice edge at which each boundary face should start.
+%% \item \verb+depth+: The depth of each boundary face.
+%% \end{itemize}
 
-\subsection{Syntax}
-\begin{verbatim}
-void copyToTargetBoundary3D(void *targetData, const void *data, size_t extent3D[3], size_t nField, size_t offset, size_t depth);
-\end{verbatim}
+%% \newpage
+%% \section{copyFromTargetBoundary3D}
 
-\begin{itemize}
-\item \verb+targetData+: A pointer to the destination array on the target.
-\item \verb+data+: A pointer to the source array on the host.
-\item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
-\item \verb+nFields+: The number of fields per lattice site.
-\item \verb+offset+: The number of sites from the lattice edge at which each boundary face should start.
-\item \verb+depth+: The depth of each boundary face.
-\end{itemize}
+%% \subsection{Description}
 
-\subsection{Example}
-**TO DO**.
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**.
-\subsubsection{CUDA}
-**TO DO**.
+%% The \verb+copyFromTargetBoundary3D+ function copies the data corresponding to the boundaries of a 3D lattice from the target to the host.
 
-\newpage
-\section{copyFromTargetBoundary3D}
+%% \subsection{Syntax}
+%% \begin{verbatim}
+%% void copyFromTargetBoundary3D(void* data, const void* targetData, 
+%%                  size_t extent3D[3], size_t nField, 
+%%                  size_t offset, size_t depth);
+%% \end{verbatim}
 
-\subsection{Description}
+%% \begin{itemize}
+%% \item \verb+data+: A pointer to the destination array on the host.
+%% \item \verb+targetData+: A pointer to the source array on the target.
+%% \item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
+%% \item \verb+nFields+: The number of fields per lattice site.
+%% \item \verb+offset+: The number of sites from the lattice edge at which each boundary face should start.
+%% \item \verb+depth+: The depth of each boundary face.
+%% \end{itemize}
 
-The \verb+copyFromTargetBoundary3D+ function copies the data corresponding to the boundaries of a 3D lattice from the target to the host.
-
-\subsection{Syntax}
-\begin{verbatim}
-void copyFromTargetBoundary3D(void *data, const void *targetData, size_t extent3D[3], size_t nField, size_t offset, size_t depth);
-\end{verbatim}
-
-\begin{itemize}
-\item \verb+data+: A pointer to the destination array on the host.
-\item \verb+targetData+: A pointer to the source array on the target.
-\item \verb+extent3D+: An array of 3 integers corresponding to the 3D dimensions of the lattice.
-\item \verb+nFields+: The number of fields per lattice site.
-\item \verb+offset+: The number of sites from the lattice edge at which each boundary face should start.
-\item \verb+depth+: The depth of each boundary face.
-\end{itemize}
-
-\subsection{Example}
-**TO DO**.
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**.
-\subsubsection{CUDA}
-**TO DO**.
 \newpage
@@
 \end{itemize}
-
-\subsection{Example}
-**TO DO**.
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**.
-\subsubsection{CUDA}
-**TO DO**.
 
@@
 \end{itemize}
-
-\subsection{Example}
-**TO DO**.
-
-\subsection{Implementation}
-\subsubsection{C}
-**TO DO**.
-\subsubsection{CUDA}
-**TO DO**.

### 2017-02-14 targetDP/doc/memory_model.tex

--- 2015-10-16 branches/kevin-1407/targetDP/doc/memory_model.tex
@@
-\chapter{Memory model}
+\chapter{Memory model}\label{sec:memmodel}
 
 \section{Model overview}
-The targetDP model draws a distinction between the memory spaces
-accessed by the host and target, such that code executed on the host
-always accesses the host memory space, and code executed on the target
-(i.e. within target functions) always accesses the target memory space. The
-host memory space can be initialised using regular \verb+C/C+++
-functionality, and the targetDP API provides the functionality
-necessary to manage the target memory space and transfer data between
-the host and target. For each data-parallel data structure, the
-programmer should create both host and target copies, and should
-update these from each other as and when required.
+The targetDP model draws a distinction between the memory space
+accessed by the host and that accessed by the target. Even although
+there is a trend towards ``unified'' address spaces, on which this
+distinction is not strictly required for applications to run successfully, such
+visibility at the application level is often crucial to allow good
+performance when running on those target architectures that have
+associated high-bandwidth memory systems (such as GPU and Intel Xeon
+Phi architectures). In the targetDP model, it is assumed that code
+executed on the host always accesses the host memory space, and code
+executed on the target (i.e. within target functions) always accesses
+the target memory space. The host memory space can be initialised
+using regular \verb+C/C+++ functionality, and the targetDP API
+provides the functionality necessary to manage the target memory space
+and transfer data between the host and target. For each data-parallel
+data structure, the programmer should create both host and target
+copies, and should update these from each other as and when required.
 
@@
 The team of threads performing the execution of target functions
-should always act on target data structures. These can take one of 3
-forms:
+should always act on target data structures. These data structures can
+take one of 3 forms:
 \begin{enumerate}
@@
 
-\begin{comment}
-For 3, at the moment any data declared in the function but above the targetTLP keyword will be private for GPU threads but shared for CPU (since outside parallel region).
+%\begin{comment}
+%For 3, at the moment any data declared in the function but above the targetTLP keyword will be private for GPU threads but shared for CPU (since outside parallel region).
 So either need to make this clearer above, or somehow move OpenMP parallel region to target launch.
-\end{comment}
+%\end{comment}
 

### 2017-02-21 docs/binary.tex
Trunk add LC emulsion document and relevant changes
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3079 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \section{Binary Fluids}
+\label{section-binary}
 
@@
 
-\includegraphics[width=0.9\textwidth]{figures/symmetric.eps}
+\includegraphics[width=0.9\textwidth]{figures/symmetric}
 
@@
 and their
-impact on interfacial width and tension is deferred until LATER.
+impact on interfacial width and tension is deferred until
+Section~\ref{section-examples}.
+
+\subsubsection{The free energy density in practice}
+
+Writing the free energy density in full, we have
+\begin{equation}
+\label{eq-binary-fed}
+f(\phi, \partial_\alpha \phi) =
+{\textstyle\frac{1}{2}} A \phi^2 +
+{\textstyle\frac{1}{4}} B \phi^4 +
+{\textstyle\frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
+\end{equation}
+Note that some authors define $A$ to be a positive constant in which case
+a negtive sign appears in the quadratic term above. We retain $A$ negative
+to describe the separated case.
 

### 2017-02-21 docs/lb.tex
Trunk add LC emulsion document and relevant changes
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3079 40876441-3496-0410-a613-a2fca35b2ae2

@@
 We review here the lattice Boltzmann method applied to a simple
-Newtonian fluid with particular emphsis on the relevant
+Newtonian fluid with particular emphasis on the relevant
 implementation in \textit{Ludwig}.
@@
 \sum_i f_i(\mathbf{r};t) c_{i\alpha} c_{i\beta}.
-\label{lb-f-moments}
+\label{equation-lb-f-moments}
 \end{equation}
@@
 $m_i^a = 1$ and the mode $M^a$ is the density $\rho = \sum_i f_i$. Note
-that the number of modes supported by a given basis will generally excceed the
+that the number of modes supported by a given basis will generally exceed the
 number of hydrodynamic modes; the excess modes have no direct physical
@@
 normaliser for each mode is $N^a$. The eigenvectors of the collision
-matrix are the columns of the transformation matrix $m^a_i$. The prefactor
+matrix are the columns of the transformation matrix $m^a_i$. The pre-factor
 $p$ (where present) multiplies all the elements to the right in that row.}
@@
 for each mode is $N^a$. The eigenvectors of the collision matrix are the
-columns of the transformation matrix $m_i^a$. The prefactor $p$ simply
+columns of the transformation matrix $m_i^a$. The pre-factor $p$ simply
 multiplies all elements of $m_i^a$ in that row as a convenience.
@@
 three dimensions. The nineteen modes $M^a$ include ten hydrodynamic modes,
-three scalr kinetic modes $\chi^1$, $\chi^2$, and $\chi^3$; there are also
+three scalar kinetic modes $\chi^1$, $\chi^2$, and $\chi^3$; there are also
 two vector kinetic modes $\chi^1 \rho c_{i\alpha}$
@@
 \end{equation}
-where $\hat{\xi}^a$ is a noise termwhich has a variance determined
+where $\hat{\xi}^a$ is a noise term which has a variance determined
 by the relaxation time for given mode
@@
 \end{equation}
-where $\hat{\sigma}_{\alpha\beta}$ is a symmtric matrix of random
+where $\hat{\sigma}_{\alpha\beta}$ is a symmetric matrix of random
 variates drawn from a Gaussian distribution with variance given

### 2017-02-21 docs/lc-emulsion.tex
Trunk add LC emulsion document and relevant changes
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3079 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  lc-emulsion.tex
%
%  Liquid Crystal Emulsions.
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Contributing Authors:
%
%  Adriano Tiribocchi made the original draft.
%  Juho lintuvuori (juho.lintuvuori@u-bordeaux.fr)
%  Kevin Stratforrd (kevin@epcc.ed.ac.uk)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Liquid Crystal Emulsions}

This section describes the approach used to model liquid crystal
emulsions which might include, for example, droplets of liquid
crystal in water. This combines elements of the binary fluid model
discussed in Section~\ref{section-binary} with the liquid crystal
approach of Section~\ref{section-lc}.

\subsection{Combined Approach}

\subsubsection{Order parameters}

To describe a liquid crystal emulsion, two order parameters are used:
the composition $\phi({\mathbf r}; t)$ and the orientational order
$Q_{\alpha\beta}({\mathbf r}; t)$.

\subsubsection{Free energy densities}

The free energy density is here the sum of three contributions:
\begin{equation}
f(\phi, Q_{\alpha\beta}) = f_{\rm c}(\phi) + f_{\rm lc}(Q_{\alpha\beta})
                         + f_{\rm anchoring}(\phi, Q_{\alpha\beta}).
\end{equation}

The first contribution relates to the composition using the standard
symmetric approach
\begin{equation}
f_{\rm c}(\phi) =
 {\textstyle\frac{1}{2}}A\phi^2
+ {\textstyle\frac{1}{4}}B\phi^4
+ {\textstyle\frac{1}{2}}\kappa (\partial_\alpha \phi)^2
\end{equation}
(cf.\ Equation~\ref{eq-binary-fed}),
where we recall that in the separated regime $A$ is a negative constant,
$B$ is a positive constant, and $\kappa$ is a positive constant which
determines the interfacial tension. The extremal values of the
composition are $\phi^\star = (-A/B)^{1/2}$ while
uniformly mixed components in equal proportion have $\phi = 0$.

The second contribution is the liquid crystal bulk free energy density
(Equation~\ref{eq-lc-fed-bulk}):
\begin{equation}
f_{\rm lc} =
  {\textstyle\frac{1}{2}}A_0 \left(1-\gamma(\phi)/3\right) Q_{\alpha\beta}^2
- {\textstyle\frac{1}{3}}A_0 \gamma(\phi)
  Q_{\alpha\beta} Q_{\beta\pi} Q_{\pi\alpha}
+ {\textstyle\frac{1}{4}} A_0 \gamma(\phi) (Q_{\alpha\beta})^2,
\end{equation}
where the composition enters parametrically through the control
parameter $\gamma = \gamma(\phi)$. This will be discussion further
below. The distortion terms in the
free energy related to the elastic constants $\kappa_0$ and $\kappa_1$
are unchanged (and are as appear in Equation~\ref{equation-lc-gradient-fe}).

Finally, there is a coupling between the liquid crystal and the
composition which determines the anchoring (normal or planar) at
the boundary of the droplet:
\begin{equation}
f_{\rm anchoring}(\phi,Q_{\alpha\beta}) =
W Q_{\alpha\beta}\partial_{\alpha}\phi\partial_{\beta}\phi.
\end{equation}
For $W>0$ the anchoring is planar, and for $W<0$ the anchoring is normal.

\subsubsection{Control of the isotropic/nematic transition}
The function $\gamma(\phi)$ allows control of the liquid crystal
ordering as the composition changes. An appropriate choice of function
for the case of $\phi^\star = 1$ is
\begin{equation}
\gamma(\phi)=\gamma_0+\delta\left(1+\phi\right)
\end{equation}
so that $\gamma(\phi=-1)=\gamma_0$, and $\gamma(\phi=1)=\gamma_0+2\delta$.
We should choose $\gamma_0$ and $\delta$ such that $\gamma(\phi=-1)$ 
is smaller than 2.7 (so that for $\phi=-1$ the fluid is isotropic)
and $\gamma(\phi=+1)$ is larger than 2.7 (so that when $\phi=1$ the
liquid crystal is in the ordered phase, nematic or cholesteric etc.).
Experience suggests that for example 
$\gamma_0=2.5$ and $\delta=0.25$ is a suitable choice.

\subsubsection{Chemical potential}

The chemical potential $\mu(\phi,Q_{\alpha\beta})$ resulting from the
combined free energy density is the functional derivative of the
total energy with respect to the composition~$\phi$. In the
case that $\gamma(\phi) \sim \delta\phi$ we generate terms
\begin{equation}
A\phi + B\phi^3 -\kappa \partial_{\gamma}\partial_{\gamma}\phi
- \textstyle{\frac{1}{6}} A_0 \delta Q_{\alpha\beta}^2
- \textstyle{\frac{1}{3}} A_0 \delta
                          Q_{\alpha\beta} Q_{\beta\pi} Q_{\pi\alpha}
+ \textstyle{\frac{1}{4}} A_0 \delta (Q_{\alpha\beta}^2)^2
\end{equation}
from $f_{\rm c}(\phi)$ and $f_{\rm lc}(Q_{\alpha\beta})$.
The interfacial or anchoring contribution to the chemical potential is
\begin{equation} 
-W \left(
 \partial_{\alpha}\phi\partial_{\beta}Q_{\alpha\beta}
+\partial_{\beta}\phi\partial_{\alpha}Q_{\alpha\beta}\right)
- 2W Q_{\alpha\beta} \partial_\alpha\partial_\beta\phi.
\end{equation}

\subsubsection{Molecular field}

The terms in the molecular field resulting from the bulk free energy
density are:
\begin{equation}
-A_0 (1 - \gamma(\phi)/3) Q_{\alpha\beta}
+A_0 \gamma(\phi)\left[ Q_{\alpha\pi} Q_{\pi\beta}
     - {\textstyle\frac{1}{3}} \delta_{\alpha\beta} Q_{\pi\sigma}^2\right]
-A_0 \gamma (\phi)Q_{\pi\nu}^2 Q_{\alpha\beta} \\ \nonumber
\end{equation}
where the terms related to the elastic contants $\kappa_0$ and $\kappa_1$
have been omitted (cf. Equation~\ref{eq-lc-h-full}). The interfacial
contribution to the molecular field is
\begin{equation}
-W \left( \partial_{\alpha}\phi\partial_{\beta}\phi
- {\textstyle\frac{1}{3}}
  \partial_{\pi}\phi \partial_{\pi}\phi \delta_{\alpha\beta} \right).
\end{equation}
As before, relevant contributions to the molecular field have
been rendered traceless.

\subsubsection{Stress and the force on the fluid}
The thermodynamic contribution to the pressure tensor is
\begin{equation}
\label{eq-lc-emulsion-stress}
\Pi_{\alpha\beta} = \sigma_{\alpha\beta} + \tau_{\alpha\beta} + P_{\alpha\beta}
\end{equation}
where $\sigma_{\alpha\beta}$ and $\tau_{\alpha\beta}$ are the symmetric and
antisymmetric contributions
arising from the liquid crystalline order: 
\begin{equation}
\sigma_{\alpha\beta} =
-p_0\delta_{\alpha\beta}
- \xi H_{\alpha\pi}
  \left(Q_{\pi\beta}+{\textstyle \frac{1}{3}} \delta_{\alpha\beta}\right)
- \xi
  \left(Q_{\alpha\pi} + {\textstyle\frac{1}{3}}\delta_{\alpha\pi}\right)
  H_{\pi\beta}
+ 2\xi \left(Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}
\right) Q_{\pi\nu} H_{\pi\nu}
\nonumber
\end{equation}
and
\begin{equation}
\tau_{\alpha\beta} = Q_{\alpha\pi} H_{\pi\beta} - H_{\alpha\pi}Q_{\pi\beta}.
\nonumber
\end{equation}

The term $P_{\alpha\beta}$ includes the binary fluid and coupling
(between LC and binary fluid) contributions and is somewhat more
complicated than in the bare liquid crystal case:
\begin{eqnarray}
\label{eq-lc-emulsion-pab}
P_{\alpha\beta} =
-\left( \phi\frac{\delta  F}{\delta\phi} - F \right) \delta_{\alpha\beta}
-\frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\alpha}\phi
-\frac{\delta F}{\delta\left(\partial_{\beta}Q_{\gamma\nu}\right)}\partial_{\alpha}Q_{\gamma\nu}.
\end{eqnarray}
where we have re-introduced the full free energy functional $F$.
This term is discussed in the following section.

\subsection{Implementation}

It is convenient \cite{sulaiman2006} to rewrite a divergence of the
stress  $-\partial_\beta P_{\alpha\beta}$ with the stress defined in
Equation~\ref{eq-lc-emulsion-pab}. This force can be expanded in full as

\begin{equation}
\partial_{\alpha}\left(\phi\mu - F\right)
+\partial_{\beta}\frac{\delta F}{\delta Q_{\pi\nu,\beta}}
 \partial_{\alpha} Q_{\pi\nu}
+\frac{\delta F}{\delta Q_{\pi\nu,\beta}} \partial_{\alpha}\partial_{\beta}
 Q_{\pi\nu}
+\partial_{\beta} \frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\alpha}\phi
+\frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\alpha}\partial_{\beta}\phi.
\nonumber
\end{equation}
With some manipulation this can be recast as 
\begin{equation}
-\left( \frac{\partial F}{\partial Q_{\pi\nu}} 
       - \partial_{\beta}\frac{\delta F}{\delta Q_{\pi\nu,\beta}}
 \right) \partial_{\alpha}Q_{\pi\nu}
- \left(\frac{\partial F}{\partial\phi} 
- \partial_{\beta} \frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\right)\partial_{\alpha}\phi + \partial_{\alpha}\left(\mu\phi\right)
\nonumber
\end{equation}
which is a local force on the fluid
\begin{equation}
\label{eq-lc-emulsion-hgradq}
H_{\pi\nu} \partial_{\alpha} Q_{\pi\nu} + \phi \partial_{\alpha}\mu.
\end{equation}

The total force on the fluid locally is the divergence of the stress,
where the total stress is $\Pi_{\alpha\beta}$
given by equation~(\ref{eq-lc-emulsion-stress}).
It is sometimes useful to compute the various contributions to the
force in different ways. The contribution from the symmetric and
antisymmetric stress $\sigma_{\alpha\beta}$ and  $\tau_{\alpha\beta}$
can be calculated numerically as a divergence. The contribution from
$P_{\alpha\beta}$ can be computed numerically as a divergence, or via the
analytical form of the divergence Equation~\ref{eq-lc-emulsion-hgradq}.
The  last option is
at the expense of global conservation of momentum.

\vfill
\pagebreak

### 2017-02-21 docs/lc-emulsion.tex
Attempt to clean up lc-droplet stress tensor
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3081 40876441-3496-0410-a613-a2fca35b2ae2

@@
 (Equation~\ref{eq-lc-fed-bulk}):
-\begin{equation}
-f_{\rm lc} =
+\begin{eqnarray}
+f_{\rm lc} &=&
   {\textstyle\frac{1}{2}}A_0 \left(1-\gamma(\phi)/3\right) Q_{\alpha\beta}^2
@@
   Q_{\alpha\beta} Q_{\beta\pi} Q_{\pi\alpha}
-+ {\textstyle\frac{1}{4}} A_0 \gamma(\phi) (Q_{\alpha\beta})^2,
-\end{equation}
-where the composition enters parametrically through the control
++ {\textstyle\frac{1}{4}} A_0 \gamma(\phi) (Q_{\alpha\beta})^2\nonumber\\
+&+&{\textstyle\frac{1}{2}} \kappa_0 (\partial_\alpha Q_{\alpha\beta})^2
++ {\textstyle\frac{1}{2}} \kappa_1
+(\epsilon_{\alpha\mu\nu} \partial_\mu Q_{\nu\beta} + 2q_0 Q_{\alpha\beta})^2,
+\label{eq-lc-fed}
+\end{eqnarray}
+where in the first line of Equation~\ref{eq-lc-fed} the composition enters parametrically through the control
 parameter $\gamma = \gamma(\phi)$. This will be discussion further
 below. The distortion terms in the
-free energy related to the elastic constants $\kappa_0$ and $\kappa_1$
+second line of Equation~\ref{eq-lc-fed} related to the elastic constants $\kappa_0$ and $\kappa_1$
 are unchanged (and are as appear in Equation~\ref{equation-lc-gradient-fe}).
@@
 combined free energy density is the functional derivative of the
-total energy with respect to the composition~$\phi$. In the
+total energy with respect to the composition~$\phi$,
+
+\begin{equation}
+\mu = \mu_{\rm c} + \mu_{\rm lc} + \mu_{\rm anchoring} = \frac{\delta F}{\delta \phi} 
+= \frac{\partial F}{\partial \phi} - \partial_\alpha \frac{\partial F}{\partial (\partial_\alpha \phi)}  
+\end{equation}
+
+In the
 case that $\gamma(\phi) \sim \delta\phi$ we generate terms
 \begin{equation}
-A\phi + B\phi^3 -\kappa \partial_{\gamma}\partial_{\gamma}\phi
+\mu_{\rm c} + \mu_{\rm lc} = A\phi + B\phi^3 -\kappa \partial_{\gamma}\partial_{\gamma}\phi
 - \textstyle{\frac{1}{6}} A_0 \delta Q_{\alpha\beta}^2
@@
 \begin{equation} 
--W \left(
+\mu_{\rm anchoring}=-W \left(
  \partial_{\alpha}\phi\partial_{\beta}Q_{\alpha\beta}
@@
 
+The molecular field resulting from the combined free energy is the functional
+derivative of the total free energy with respect to the tensor order parameter $Q_{\alpha\beta}$,
+
+\begin{equation}
+H_{\alpha\beta} = H_{\alpha\beta,{\rm lc}} + H_{\alpha\beta, {\rm anchoring}}  = \frac{\delta F}{\delta Q_{\alpha\beta}} 
+= \frac{\partial F}{\partial Q_{\alpha\beta}} - \partial_\gamma \frac{\partial F}{\partial (\partial_\gamma Q_{\alpha\beta})}  
+\end{equation}
+
 The terms in the molecular field resulting from the bulk free energy
-density are:
+density of the liquid crystalline part are:
 \begin{equation}
--A_0 (1 - \gamma(\phi)/3) Q_{\alpha\beta}
+H_{\alpha\beta,{\rm lc}} = -A_0 (1 - \gamma(\phi)/3) Q_{\alpha\beta}
 +A_0 \gamma(\phi)\left[ Q_{\alpha\pi} Q_{\pi\beta}
@@
 \begin{equation}
--W \left( \partial_{\alpha}\phi\partial_{\beta}\phi
+H_{\alpha\beta, {\rm anchoring}}  = -W \left( \partial_{\alpha}\phi\partial_{\beta}\phi
 - {\textstyle\frac{1}{3}}
@@
 -\frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\alpha}\phi
--\frac{\delta F}{\delta\left(\partial_{\beta}Q_{\gamma\nu}\right)}\partial_{\alpha}Q_{\gamma\nu}.
+-\frac{\delta F}{\delta\left(\partial_{\beta}Q_{\gamma\nu}\right)}\partial_{\alpha}Q_{\gamma\nu},
 \end{eqnarray}
@@
 \partial_{\alpha}\left(\phi\mu - F\right)
-+\partial_{\beta}\frac{\delta F}{\delta Q_{\pi\nu,\beta}}
++\partial_{\beta} \left(\frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\right)\partial_{\alpha}\phi
++\frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\beta}\left(\partial_{\alpha}\phi\right)
++\partial_{\beta}\left(\frac{\delta F}{\delta Q_{\pi\nu,\beta}}\right)
  \partial_{\alpha} Q_{\pi\nu}
-+\frac{\delta F}{\delta Q_{\pi\nu,\beta}} \partial_{\alpha}\partial_{\beta}
- Q_{\pi\nu}
-+\partial_{\beta} \frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\alpha}\phi
-+\frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\partial_{\alpha}\partial_{\beta}\phi.
++\frac{\delta F}{\delta Q_{\pi\nu,\beta}} \partial_{\beta}\left(\partial_{\alpha} Q_{\pi\nu}\right).
 \nonumber
 \end{equation}
-With some manipulation this can be recast as 
-\begin{equation}
--\left( \frac{\partial F}{\partial Q_{\pi\nu}} 
-       - \partial_{\beta}\frac{\delta F}{\delta Q_{\pi\nu,\beta}}
+With some manipulation and using $$\frac{\partial F}{\partial (\partial_\alpha\partial_\beta\phi)} = \frac{\partial F}{\partial (\partial_\alpha Q_{\mu\nu,\beta})}=0$$
+this can be recast as 
+\begin{eqnarray}
+-\partial_\beta P_{\alpha\beta} &=& -\left( \frac{\partial F}{\partial Q_{\pi\nu}} 
+       - \partial_{\beta}\frac{\partial F}{\partial Q_{\pi\nu,\beta}}
  \right) \partial_{\alpha}Q_{\pi\nu}
 - \left(\frac{\partial F}{\partial\phi} 
-- \partial_{\beta} \frac{\delta F}{\delta \left(\partial_{\beta}\phi\right)}\right)\partial_{\alpha}\phi + \partial_{\alpha}\left(\mu\phi\right)
-\nonumber
-\end{equation}
-which is a local force on the fluid
-\begin{equation}
+- \partial_{\beta} \frac{\partial F}{\partial \left(\partial_{\beta}\phi\right)}\right)\partial_{\alpha}\phi + \partial_{\alpha}\left(\mu\phi\right)\nonumber\\
+&=& -\left(\frac{\delta F}{\delta Q_{\pi\nu}}\right) \partial_{\alpha}Q_{\pi\nu} - \left(\frac{\delta F}{\delta \phi}\right) \partial_{\alpha} \phi + \partial_{\alpha}\left(\mu\phi\right)\nonumber\\ 
+&=& H_{\pi\nu}\; \partial_{\alpha} Q_{\pi\nu} + \phi \;\partial_{\alpha}\mu,
 \label{eq-lc-emulsion-hgradq}
-H_{\pi\nu} \partial_{\alpha} Q_{\pi\nu} + \phi \partial_{\alpha}\mu.
-\end{equation}
-
+\end{eqnarray}
+which is a local force on the fluid.
 The total force on the fluid locally is the divergence of the stress,

### 2017-02-22 docs/liquidcrystal.tex
Trunk add LC emulsion document and relevant changes
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3079 40876441-3496-0410-a613-a2fca35b2ae2
Various layout and text improvements
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3082 40876441-3496-0410-a613-a2fca35b2ae2

--- 2016-12-21 branches/kevin-1397/docs/liquidcrystal.tex
@@
 \section{Liquid Crystals}
+\label{section-lc}
 
@@
 \begin{equation}
+\label{eq-lc-fed-bulk}
 f(Q_{\alpha\beta}) =
@@
 \begin{eqnarray}
+\label{eq-lc-h-full}
 H_{\alpha\beta} &=&
@@
 \end{equation}
-The amplitude $A$ is provided by Equation~\ref{equation-lc-amplitude}. The
-full boundary condition for the gradient of the tensor order parameter
+The amplitude $A$ is provided by 
+\begin{equation}
+\label{equation-lc-amplitude}
+A=\frac{2}{3}*(\frac{1}{4}+\frac{3}{4}*\sqrt{(1 - 8/(3\gamma)})).
+\end{equation}
+The full boundary condition for the gradient of the tensor order parameter
 at the solid-fluid boundary from Equation~\ref{equation-lc-general-bc}
@@
 \begin{center}
-\input{hybrid2.epic}
+\input{figures/hybrid.tex}
 \end{center}

### 2017-06-08 docs/tutorial/tutorial.tex
Moved tutorial into separate directory
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3128 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  ludwigTutorial.tex
%
%  Introductory tutorial for new users of Ludwig. 
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre and
%  Department of Physics, University of Strathclyde
%
%  Contributing authors:
%  Oliver Henrich (oliver.henrich@strath.ac.uk)
%  Fraser Mackay (s1026487@sms.ed.ac.uk)
%
%  (c) 2008-2017 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[11pt,twoside,a4paper]{article}

\usepackage{amsmath}
\usepackage{moreverb}
\usepackage{lscape}
\usepackage{epic}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{color}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{enumerate}

\usepackage{geometry}
 \geometry{
 a4paper,
 left=3.1cm,
 right=3.1cm,
 top=2.25cm,
 bottom=2.25cm 
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{\smallskipamount}

\newcommand{\inputkey}[1]{\framebox{\textbf{\texttt{#1}}}}
\newcommand{\e}[1]{\cdot10^{#1}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqa}{\begin{eqnarray}}
\newcommand{\eeqa}{\end{eqnarray}}
\newcommand{\com}[1]{\textcolor{red}{#1}}
\newcommand{\plag}[1]{\textcolor{green}{#1}}
\newcommand{\cur}[1]{{\textit{#1}}}

\definecolor{terminalcolour}{gray}{0.96}

\lstdefinestyle{terminalverbatim}{
  basicstyle=\small\ttfamily,
  columns=flexible,
  backgroundcolor=\color{terminalcolour},
  xleftmargin=0pt
}

\begin{document}

\setcounter{page}{1}

% We're not including the table of contents just at the moment
%\tableofcontents

\newpage
\lstset{style=terminalverbatim}

\setcounter{page}{1}

% These redefinitions are just compressing the spacing a little.

\makeatletter
\renewcommand*{\section}{%
\@startsection {section}{1}{\z@}%
  {-1.75ex \@plus -0.5ex \@minus -.1ex}%
  {1.15ex \@plus.1ex}%
  {\normalfont\Large\bfseries}%
}
\renewcommand*{\subsection}{%
\@startsection {subsection}{2}{\z@}%
  {-1.75ex \@plus -0.5ex \@minus -.1ex}%
  {1.15ex \@plus.1ex}%
  {\normalfont\large\bfseries}%
}
\renewcommand*{\subsubsection}{%
\@startsection {subsubsection}{1}{\z@}%
  {-1.75ex \@plus -0.5ex \@minus -.1ex}%
  {1.15ex \@plus.1ex}%
  {\normalfont\normalsize\bfseries}%
}

% Table of Contents
\pagenumbering{roman}
\title{Ludwig User Tutorial}
\author{Fraser Mackay and Oliver Henrich}
\maketitle 
\tableofcontents
\clearpage

% Sections
\pagenumbering{arabic}

\section{Introduction}

This tutorial document takes new users of the \texttt{Ludwig} code through the process of how to obtain 
and build the code before outlining some short tutorial exercises where simulations 
of simple example cases will be run and visualised.
It is assumed that the reader is familar with UNIX-based operationg systems and has 
a basic knowledge of hydrodynamics, complex fluids, and to some extent statistical physics. 
This knowledge will be required to make sense of the input and output involved in using the code.

The system requirements are an up-to-date version of a C compiler as well as working installations 
of Apache's Subversion (SVN) version control system, the visualisation tool
ParaView (\hyperref[ParaView]{https://www.paraview.org}) and the MPI-library 
for compiling the parallel version of the code.

\section{Access and Compilation}

\subsection{Obtaining the Code}
\label{sec:getCode}

The \texttt{Ludwig} project is currently hosted at our central repository 
at CCPForge:\\
(\hyperref[CCPForge]{https://ccpforge.cse.rl.ac.uk}).\\
For full access to the repository please open a CCPForge account and request
to join the project \texttt{Ludwig}. The following command grants anonymous access:

\begin{lstlisting}[style=terminalverbatim]
$ svn checkout https://ccpforge.cse.rl.ac.uk/svn/ludwig
\end{lstlisting}

A copy of the \texttt{Ludwig} repository will now be available in your current directory. 
You should see:

\begin{lstlisting}
$ ls ludwig
branches  tags  trunk
\end{lstlisting}
which is the usual svn layout for trunk, branches and tagged versions.

A full documentation of \texttt{Ludwig} can be obtained by running 
the command \texttt{make pdf} twice in the \texttt{/trunk/docs} directory:

\begin{lstlisting}[style=terminalverbatim]
$ cd /ludwig/trunk/docs/
$ make pdf
\end{lstlisting}

\subsection{Configuration and Build}

The following section outlines the configuration and compilation 
procedure. The configuration step is based on the GNU Compiler Collection gcc.

\begin{enumerate}
\item Go to the directory \texttt{/ludwig/trunk/}: \\
\begin{lstlisting}
$ cd ludwig/trunk
\end{lstlisting}
\item Copy the configuration file \texttt{lunix-gcc-default.mk} to the \texttt{/trunk} directory 
and rename to \texttt{config.mk}: \\
\begin{lstlisting}
$ cp config/lunix-gcc-default.mk ./config.mk
\end{lstlisting}
\item Issue \texttt{make} in \texttt{/targetDP}: \\
\begin{lstlisting}
$ cd targetDP/
$ make 
\end{lstlisting}
This creates dummy functions for calls to the targetDP library.
\item Issue \texttt{make} in \texttt{/mpi\_s}: \\
\begin{lstlisting}
$ cd ../mpi_s/
$ make 
\end{lstlisting}
This creates dummy functions for calls to the MPI-library,
which is required for serial compilation.
\item Edit \texttt{Makefile} in \texttt{/src}. To switch off 
the assertions set \texttt{OPTS = -DNP\_D3Q6 -DNDEBUG}. This is
important for production runs (see also comment in \texttt{Makefile}): \\
\begin{lstlisting}
$ cd ../src/
$ vim Makefile
...

include ../Makefile.mk

MAIN = main
EXECUTABLE = Ludwig.exe
LIBRARY = libludwig.a

OPTS = -DNP_D3Q6 -DNDEBUG
LIBS = -L../targetDP -ltarget -lm
INC = -I. -I ../targetDP
...
\end{lstlisting}
\item To compile the code in serial issue: \\
\begin{lstlisting}
$ make serial
\end{lstlisting} 
To compile the parallel version of the code issue: \\
\begin{lstlisting}
$ make mpi
\end{lstlisting} 
This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}.
\end{enumerate}

%\clearpage
%\vfill\pagebreak
%\input{references.tex}

\end{document}

### 2017-06-16 docs/tutorial/tutorial.tex
Added test 1 to ludwig tutorial
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3135 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
-%  ludwigTutorial.tex
+%  tutorial.tex
 %
@@
 \usepackage{enumerate}
+\graphicspath{{./tutorialFigs/}}
 
@@
 
+\section{Tutorial Test Exercises}
+
+\subsection{Test 1: Poiseuille Flow of a Simple Fluid in a Cavity}
+
+In this section, the first test exercise example, the Poiseuille flow of a simple fluid in a cavity, is outlined.
+This example gives a step-by-step guide example for performing a short run ($10,000$ time-steps) of the \texttt{Ludwig} code for a simple system and provides a short introduction to the post-processing and data analysis procedure.
+Before starting, the code should be compiled in parallel and the executable \texttt{Ludwig.exe} should be copied to your working directory.
+
+\begin{enumerate}
+\item Copy the file \texttt{input} located in the directory \texttt{ludwig/docs/tutorial/test1} to your working directory.
+Note that this input file is a modified version of the reference input file \texttt{input.ref} located in \texttt{/src} containing the required \com{variables??} for this run.
+\item Run the code in your working directory by issuing the command:
+\begin{lstlisting}
+$ mpirun -np 4 ./Ludwig.exe input
+\end{lstlisting}
+\item You should now see output files with names of the form \texttt{vel-XXXXXXXX.001-001} where \texttt{XXXXXXXX} is replaced by the time-step number in increments of $1000$.
+These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation.
+\item The first step to the post-processing of these data is to copy the files \texttt{vtk\_extract.c} and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util} into your working directory.
+\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
+\begin{enumerate}
+\item \texttt{input\_binary\_=0} \\ i.e. The data in the velocity files (\texttt{vel-XXXXXXXX.001-001}) are in ASCII format.
+\item \texttt{output\_binary\_=0} \\ i.e. The output should also be in ASCII format.
+\item \texttt{output\_index=1} \\ i.e. Produce the output data with lattice identifiers in the text files.
+\item \texttt{vtk\_header=0} \\ i.e. The data files should not have column headers (these will be required later for visualisation in \textit{ParaView})
+\item \texttt{output\_cmf=0} \\ i.e. The data should not be in column-major format (also required for \textit{ParaView}).
+\end{enumerate}
+\item Edit \texttt{vtk\_extract\_script.py}. Set:
+\begin{enumerate}
+\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and increments of the velocity data.
+\item \texttt{vel=1} for velocity post-processing. 
+\end{enumerate}
+\item You should now see files named \texttt{vel-XXXXXXXX.vtk}.
+The first three columns of these text files are the spacial lattice indices followed by the components of the velocity vectors in Cartesian coordinates.
+To plot the velocity profile of the system, you can plot the $y$-component of the velocity against the spacial $x$-coordinate on the lattice with the command:
+\begin{lstlisting}
+gnuplot> p`vel-XXXXXXXX.vtk' u 1:5
+\end{lstlisting} 
+By plotting the data in each file, you should be able to see the velocity profile converging to the final profile as shown in Fig.
+\ref{fig:velocityProfile}.
+For reference, an reference example of the final output at $10,000$ time-steps can be found, in addition to a file containing the standard terminal output of the code in the directory \texttt{/trunk/docs/tutorial/test1/}.
+\end{enumerate}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{velProf.png}
+  \caption{Plot of the parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps}
+  \label{fig:velocityProfile}
+  \end{center}
+\end{figure}
+
 %\clearpage

### 2017-06-20 docs/tutorial/tutorial.tex
Edited tutorial section for test1
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3139 40876441-3496-0410-a613-a2fca35b2ae2

--- 2017-06-08 docs/tutorial/tutorial.tex
@@
 of Apache's Subversion (SVN) version control system, the visualisation tool
-ParaView (\hyperref[ParaView]{https://www.paraview.org}) and the MPI-library 
-for compiling the parallel version of the code.
+ParaView (\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
+for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.
 
@@
 
+\section{Tutorial Test Exercises}
+
+\subsection{Test 1: Poiseuille Flow of a Simple Fluid in a Cavity}
+
+In this section the Poiseuille flow of a simple fluid in a cavity, is outlined. 
+The example is a step-by-step guide for performing a short run ($10,000$ time-steps) 
+with \texttt{Ludwig} code, followed by a short introduction to the post-processing 
+and data analysis procedures. Before starting, the code should be compiled in 
+parallel and the executable \texttt{Ludwig.exe} should be copied into your working directory.
+
+\begin{enumerate}
+\item Copy the input file \texttt{input} located in the directory \texttt{/trunk/docs/tutorial/test1} 
+to your working directory. Note that this input file is a modified version of the 
+reference input file \texttt{input.ref} located in \texttt{/trunk/src} with a limited set of parameters
+for this run. 
+\item Run the code in your working directory by issuing the command:
+\begin{lstlisting}
+$ mpirun -np 4 ./Ludwig.exe input
+\end{lstlisting}
+\item You should now see statistics being reported on your standard output and 
+output files being created with names of the form \texttt{vel-XXXXXXXX.001-001}. 
+These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
+\item The first step to the post-processing of these data is to copy the files \texttt{vtk\_extract.c} 
+and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util} into your working directory.
+\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
+\begin{enumerate}
+\item \texttt{input\_binary\_=0} \\ if the velocity data is in ASCII format (see flag in input file),
+\item \texttt{output\_binary\_=0} \\ as the output of the postprocessing routines should also be in ASCII format,
+\item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k}, 
+\item \texttt{vtk\_header=0} \\ to omit headers (these will be required later for visualisation in \textit{ParaView})
+\item \texttt{output\_cmf=0} to produce data in row-major format (column-major format is required for \textit{ParaView}).
+\end{enumerate}
+\item Edit \texttt{vtk\_extract\_script.py}. Set:
+\begin{enumerate}
+\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and increments of the velocity data
+to be processed.
+\item \texttt{vel=1} for velocity post-processing, 
+\item The other flags should be set to zero, e.g. \texttt{phi=0}. 
+\end{enumerate}
+\item You should now see files named \texttt{vel-XXXXXXXX.vtk} (Note: The extension \texttt{.vtk} is only used here for simplicity.
+It does not mean the generated data is in vtk-format). 
+The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.\\
+
+To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
+the spatial $x$-coordinate on the lattice with the command:
+\begin{lstlisting}
+gnuplot> p`vel-XXXXXXXX.vtk' u 1:5
+\end{lstlisting} 
+By plotting the data in each file, you should be able to see the velocity profile converging to the final profile 
+as shown in Fig. \ref{fig:velocityProfile}. In the directory \texttt{/trunk/docs/tutorial/test1} a reference file of the final 
+output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
+\end{enumerate}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{velProf.png}
+  \caption{The parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps}.
+  \label{fig:velocityProfile}
+  \end{center}
+\end{figure}
+
 %\clearpage

### 2017-06-28 docs/tutorial/tutorial.tex
Edited tutorial section for test1
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3140 40876441-3496-0410-a613-a2fca35b2ae2
Added tutorial section for test2
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3144 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \usepackage{enumerate}
+\usepackage{float}
 \graphicspath{{./tutorialFigs/}}
@@
 \item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k}, 
-\item \texttt{vtk\_header=0} \\ to omit headers (these will be required later for visualisation in \textit{ParaView})
+\item \texttt{vtk\_header=0} \\ to omit headers (these will be required later for visualisation in \textit{ParaView}),
 \item \texttt{output\_cmf=0} to produce data in row-major format (column-major format is required for \textit{ParaView}).
@@
 \end{enumerate}
+\item Issue the command:
+\begin{lstlisting}
+$ python vtk_extract_script.py
+\end{lstlisting}
 \item You should now see files named \texttt{vel-XXXXXXXX.vtk} (Note: The extension \texttt{.vtk} is only used here for simplicity.
@@
 By plotting the data in each file, you should be able to see the velocity profile converging to the final profile 
-as shown in Fig. \ref{fig:velocityProfile}. In the directory \texttt{/trunk/docs/tutorial/test1} a reference file of the final 
+as shown in Fig. \ref{fig:velocityProfile}.
+
+In the directory \texttt{/trunk/docs/tutorial/test1} a reference file of the final 
 output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
@@
 \includegraphics[width=0.6\linewidth]{velProf.png}
-  \caption{The parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps}.
+  \caption{The parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps.}
   \label{fig:velocityProfile}
@@
 \end{figure}
+
+\subsection{Test 1: Spinodal Decomposition of a Binary Fluid}
+
+This section outlines the second example, the spinodal decomposition of a binary fluid in 
+three-dimensions.
+The example describes the post-processing procedure for this system and a brief guide to 
+the visualisation of a 3D system using ParaView.
+As in the previous example, the code should be compiled in parallel and the executable 
+\texttt{Ludwig.exe} should be copied into your working directory.
+
+\begin{enumerate}
+\item Copy the input file \texttt{input} located in the directory 
+\texttt{/trunk/docs/tutorial/test2} 
+to your working directory. This input file contains the parameters required for this run. 
+\item Run the code in your working directory by issuing the command:
+\begin{lstlisting}
+$ mpirun -np 8 ./Ludwig.exe input
+\end{lstlisting}
+\item As before, you should now see statistics being reported on your standard output but 
+in this case two sets of output files being created containing the velocity and compositional
+order parameter ($\phi$) data at each lattice point in the simulation. In this example, these 
+data are in binary format.
+\item The post-processing of these data again requires the files \texttt{vtk\_extract.c} 
+and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util}.
+\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
+\begin{enumerate}
+\item \texttt{input\_binary\_=1} \\ if the velocity and $\phi$ data are in binary format (see 
+flag in input file),
+\item \texttt{output\_binary\_=0} \\ as the output of the post-processing routines should be 
+in ASCII format,
+\item \texttt{output\_index=0} \\ as Cartesian lattice indices are not required, 
+\item \texttt{vtk\_header=1} \\ to include headers,
+\item \texttt{output\_cmf=1} to produce data in column-major format.
+\end{enumerate}
+\item Edit \texttt{vtk\_extract\_script.py}. Set:
+\begin{enumerate}
+\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and 
+increments of the velocity data
+to be processed.
+\item \texttt{vel=1} for velocity post-processing, 
+\item \texttt{phi=1} for $\phi$ post-processing,
+\item The other flags should be set to zero. 
+\end{enumerate}
+\item Issue the command:
+\begin{lstlisting}
+$ python vtk_extract_script.py
+\end{lstlisting}
+\item You should now see files with the extension \texttt{.vtk} corresponding to the velocity 
+and $\phi$ data ready to be visualised using ParaView.
+\begin{enumerate}
+\item Open the files containing the $\phi$ data.
+\item Click on the group of files in the pipeline browser and apply a contour map using 
+Filters/Common/Contour or in the tool bar:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\linewidth]{contour.png}
+  \caption{The contour button in the toolbar.}
+  \label{fig:contour}
+  \end{center}
+\end{figure}
+
+You should now see a three-dimensional rendering of the binary fluid.
+\item Open the files containing the velocity data.
+\item Click on the group of files in the pipeline browser and apply the glyph filter using 
+Filters/Common/Glyph or in the tool bar:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\linewidth]{glyph.png}
+  \caption{The glyph button in the toolbar}
+  \label{fig:glyph}
+  \end{center}
+\end{figure}
+
+\item Change the colour code to GlyphVector Magnitude.
+\item In the pipeline browser, you should see:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{pipeline.png}
+  \caption{The complete pipline browser for the visualisation of spinodal decomposition.}
+  \label{fig:pipeline}
+  \end{center}
+\end{figure}
+
+A 3D visualisation of the system should also now be shown:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{system.png}
+  \caption{A visualisation of a separating binary fluid using ParaView.}
+  \label{fig:sysVisualisation}
+  \end{center}
+\end{figure}
+
+\item You can move through different frames of the simulation with the arrow buttons:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.4\linewidth]{arrows.png}
+  \caption{A visualisation of a separating binary fluid using ParaView.}
+  \label{fig:sysVisualisation}
+  \end{center}
+\end{figure}
+
+as you do this, you should see the coarsening of the binary fluid.
+
+\item In the File menu you can save a screen shot of the visualisation or save the state of the system in order to reload the entire visualisation.
+\end{enumerate}
+\end{enumerate}
 

### 2017-10-03 docs/tutorial/tutorial.tex
added test3 to tutorial.tex
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3165 40876441-3496-0410-a613-a2fca35b2ae2

@@
 \end{lstlisting}
+Note that you may need to specify the path to the MPI wrapper compiler in this file.
 \item Issue \texttt{make} in \texttt{/targetDP}: \\
@@
 
-\subsection{Test 1: Spinodal Decomposition of a Binary Fluid}
+\subsection{Test 2: Spinodal Decomposition of a Binary Fluid}
 
@@
 \includegraphics[width=0.8\linewidth]{contour.png}
-  \caption{The contour button in the toolbar.}
+  \caption{The contour button in the tool bar.}
   \label{fig:contour}
@@
 \includegraphics[width=0.8\linewidth]{glyph.png}
-  \caption{The glyph button in the toolbar}
+  \caption{The glyph button in the tool bar}
   \label{fig:glyph}
@@
   \caption{A visualisation of a separating binary fluid using ParaView.}
-  \label{fig:sysVisualisation}
+  \label{fig:changeFrame}
   \end{center}
@@
 
+\subsection{Test 3: Colloids Dispersed in a Liquid Crystalline Fluid}
+
+For this third example, the system being used is a that of colloidal particles suspended in 
+a liquid crystalline fluid.
+The example illustrates the post-processing procedure for a colloidal system in addition to 
+demonstrating the ways to visualise liquid crystals in ParaView. 
+As before, the code should be compiled in parallel and the executable 
+\texttt{Ludwig.exe} should be copied into your working directory.
+
+\begin{enumerate}
+\item Copy the input file \texttt{input} located in the directory 
+\texttt{/trunk/docs/tutorial/test3} to your working directory. We see in this input file that 
+there are many more parameters 
+than in the previous examples. Specifically, we note that \texttt{colloid\_cell\_min} 
+must be set if colloids are present and must be at least $2r + \delta$ where $r$ is colloid 
+radius and $\delta$ - here set to 0.5 - catches any colloid-colloid interactions. This is shown for our example - where this value is set to 8.0 - in Fig. \ref{fig:colloid_int}.
+
+ \begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.5\linewidth]{colloidInit.png}
+  \caption{A diagrammatic representation of a colloid in a 32$^3$ system showing that the value, \texttt{colloid\_cell\_min} must be set such that it can contain the entire colloid plus colloid-colloid interactions.}
+  \label{fig:colloid_int}
+  \end{center}
+\end{figure}
+
+\item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids.
+Firstly compile the code in serial and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}.
+These are the file required for the initialisation and visualisation of colloids respectively.
+\item In the file, \texttt{colloid\_init.c} set the flags for:
+\begin{enumerate}
+\item \texttt{NMC = 0},
+\item \texttt{periodic[3] = {1,1,1}} \\ for our fully periodic system,
+\item \texttt{file\_format=ASCII} \\ resulting in ASCII output,
+\item \texttt{a0 = 3.5} \\ which sets the radius of the colloids,
+\item \texttt{ah = 3.5} \\ which sets the hydrodynamic radius of the colloid,
+\item \texttt{vf = 0.02} \\ this sets the volume fraction of colloids to 2\%, for our system of size 32$^2$ with colloids of radius 3.5, this will result in 3 colloids,
+\item \texttt{dh = 0.5} \\ to set the grace distance for interactions (note that this means that one colloid plus this distance is less than \texttt{colloid\_cell\_min} which was set in the input file.
+\end{enumerate}
+\item Now issue: \\
+\begin{lstlisting}
+$ make colloid_init
+$ ./colloid_init
+\end{lstlisting}
+this produces a file, \texttt{config.cds.init.001-001} which should be copied into your working directory.
+\item In the file, \texttt{extract\_colloids.c} set:
+\begin{enumerate}
+\item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
+\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
+\item \texttt{cds\_with\_v  = 1} \\ which will output the velocities and positions of the colloids,
+\end{enumerate}
+\item Now issue: \\
+\begin{lstlisting}
+$ make extract_colloids
+\end{lstlisting}
+and copy the executable \texttt{extract\_colloids} into your working directory along with the 
+usual post-processing scripts, \texttt{vtk\_extract.c} and \texttt{vtk\_extract\_script.py} 
+scripts. 
+\item Next, run the code in your working directory by issuing the command:
+\begin{lstlisting}
+$ mpirun -np 8 ./Ludwig.exe input
+\end{lstlisting}
+\item For the post-processing of this system, in \texttt{vtk\_extract.c}, set:
+\begin{enumerate}
+\item \texttt{input\_binary\_  = 1} \\ due to binary output from code,
+\item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
+\item \texttt{output\_lc\_op\_   = 1} \\ to visualise the liquid crystal order parameter,
+\item \texttt{output\_lc\_dir\_  = 1} \\ to visualise the liquid crystal director field,
+\item \texttt{output\_lc\_biop\_ = 0} \\ since the liquid crystals are not biaxial,
+\end{enumerate}
+\item and in \texttt{vtk\_extract\_script.py},
+\begin{enumerate}
+\item \texttt{vel = 1}, \\ to visualise the velocity field,
+\item \texttt{q = 1}, \\ since there are liquid crystals present,
+\item \texttt{colcds = 0} and 
+\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
+\end{enumerate}
+\item After running the python script as before, \texttt{.csv} files with the colloid data should 
+be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
+\item The system can now be visualised, again using ParaView:
+\begin{enumerate}
+\item The liquid crystal, scalar order parameter can be visualised by opening the files: 
+\texttt{op-q-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
+minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\linewidth]{thresh.png}
+  \caption{The threshold button in the tool bar.}
+  \label{fig:thresh}
+  \end{center}
+\end{figure}
+
+\item For the director field open the \texttt{dir-q-XXXXXXXX.vtk} and apply the `glyphvector' filter. In this case, change the glyph 
+type from `arrow' to `line' and scale the colour of the lines by the $x$-component of the 
+vector.
+\item The velocity field can be visualised using arrows as before with the `glyphvector' filter.
+\item To visualise the colloids, open the files \texttt{col-cdsXXXXXXXX.csv}. This will open 
+a table, which can be closed. Next apply the filter `Table to Points' found in 
+Filters/Alphabetical/Table to Points and set the X, Y and Z Columns in the `Properties' tab 
+below the Pipeline Browser to be the $x$, $y$ and $z$-coordinates in the table via the 
+drop down menus:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.5\linewidth]{colloidCoords.png}
+  \caption{Setting the coordinates of the colloid locations in the properties tab.}
+  \label{fig:collCoords}
+  \end{center}
+\end{figure}
+
+ You should now see points in the render window at the location of the centres of each of the colloids.
+\item To fully visualise the colloids, apply the `glyphvector' filter and set glyph type to 
+`sphere.' In the properties tab, set the radius to 3.5, scale mode `off' and scale factor to 
+1.0 in the properties. In order to access these settings, you will need to click the `cog' icon shown here:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.5\linewidth]{cog.png}
+  \caption{Location of the `cog' icon in the properties tab.}
+  \label{fig:cog}
+  \end{center}
+\end{figure}
+
+\item Next, glyph mode must be set to `all points' and you should set the colour of teh spheres to `Solid Color'.
+Finally, you can increase the resolution of the spheres by increasing the `Theta Resolution' and `Phi Resolution' quantities.
+\item In the pipeline browser, should now see:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{colloidsPipeline.png}
+  \caption{The pipeline browser for the colloidal suspension in liquid crystals.}
+  \label{fig:collPipe}
+  \end{center}
+\end{figure}
+
+The 3D visualisation of this system should appear as:
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.6\linewidth]{colloidSystem.png}
+  \caption{The visualisation of the colloidal suspension in liquid crystals.}
+  \label{fig:collVis}
+  \end{center}
+\end{figure}
+
+\end{enumerate}
+\end{enumerate}
+
 %\clearpage

### 2017-10-20 docs/tutorial/tutorial.tex
Minor edits in tutorial
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3170 40876441-3496-0410-a613-a2fca35b2ae2

@@
 of simple example cases will be run and visualised.
-It is assumed that the reader is familar with UNIX-based operationg systems and has 
+It is assumed that the reader is familar with UNIX-based operating systems and has 
 a basic knowledge of hydrodynamics, complex fluids, and to some extent statistical physics. 
@@
 The system requirements are an up-to-date version of a C compiler as well as working installations 
-of Apache's Subversion (SVN) version control system, the visualisation tool
+of a text editor like \texttt{vi} or \texttt{emacs}, Apache's Subversion (SVN) version control system, the visualisation tool
 ParaView (\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
@@
 
-\subsection{Configuration and Build}
+\subsection{Configuration and Build}\label{configbuild}
 
@@
 \item Copy the configuration file \texttt{lunix-gcc-default.mk} to the \texttt{/trunk} directory 
-and rename to \texttt{config.mk}: \\
+and rename to \texttt{config.mk} (both steps exectued below at once): \\
 \begin{lstlisting}
@@
 This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}. 
+Before going from a serial to a parallel compilation or vice versa
+you should clean the directory by issuing
+\begin{lstlisting}
+$ make clean
+\end{lstlisting}
+
 \end{enumerate}
@@
 \item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids.
-Firstly compile the code in serial and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}.
+Firstly compile the code in serial (see section \ref{configbuild})and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}.
 These are the file required for the initialisation and visualisation of colloids respectively.

### 2018-05-04 docs/lc-emulsion.tex

@@
 \begin{equation}
-\gamma(\phi)=\gamma_0+\delta\left(1+\phi\right)
+\gamma(\phi)=\gamma_0+\Delta\left(1+\phi\right)
 \end{equation}
-so that $\gamma(\phi=-1)=\gamma_0$, and $\gamma(\phi=1)=\gamma_0+2\delta$.
-We should choose $\gamma_0$ and $\delta$ such that $\gamma(\phi=-1)$ 
+so that $\gamma(\phi=-1)=\gamma_0$, and $\gamma(\phi=1)=\gamma_0+2\Delta$.
+We should choose $\gamma_0$ and $\Delta$ such that $\gamma(\phi=-1)$ 
 is smaller than 2.7 (so that for $\phi=-1$ the fluid is isotropic)
@@
 Experience suggests that for example 
-$\gamma_0=2.5$ and $\delta=0.25$ is a suitable choice.
+$\gamma_0=2.5$ and $\Delta=0.25$ is a suitable choice.
 
@@
 In the
-case that $\gamma(\phi) \sim \delta\phi$ we generate terms
+case that $\gamma(\phi) \sim \Delta\cdot\phi$ we generate terms
 \begin{equation}
 \mu_{\rm c} + \mu_{\rm lc} = A\phi + B\phi^3 -\kappa \partial_{\gamma}\partial_{\gamma}\phi
-- \textstyle{\frac{1}{6}} A_0 \delta Q_{\alpha\beta}^2
-- \textstyle{\frac{1}{3}} A_0 \delta
+- \textstyle{\frac{1}{6}} A_0 \Delta Q_{\alpha\beta}^2
+- \textstyle{\frac{1}{3}} A_0 \Delta
                           Q_{\alpha\beta} Q_{\beta\pi} Q_{\pi\alpha}
-+ \textstyle{\frac{1}{4}} A_0 \delta (Q_{\alpha\beta}^2)^2
++ \textstyle{\frac{1}{4}} A_0 \Delta (Q_{\alpha\beta}^2)^2
 \end{equation}
@@
 Equation~\ref{eq-lc-emulsion-pab}. This force can be expanded in full as
-
 \begin{equation}
@@
 +\frac{\delta F}{\delta Q_{\pi\nu,\beta}} \partial_{\beta}\left(\partial_{\alpha} Q_{\pi\nu}\right).
-\nonumber
+\label{eq-lc-emulsion-force}
 \end{equation}
-With some manipulation and using $$\frac{\partial F}{\partial (\partial_\alpha\partial_\beta\phi)} = \frac{\partial F}{\partial (\partial_\alpha Q_{\mu\nu,\beta})}=0$$
-this can be recast as 
+Applying the chain rule on the first bracket in Equation \ref{eq-lc-emulsion-force} leads to
+\begin{equation}
+\partial_{\alpha}(\phi\mu) 
+- \frac{\partial F}{\partial \phi} \partial_\alpha \phi
+- \frac{\partial F}{\partial(\partial_\beta \phi)} \partial_\alpha (\partial_\beta\phi)
+- \frac{\partial F}{\partial Q_{\pi\nu}} \partial_\alpha Q_{\pi\nu}  
+- \frac{\partial F}{\partial Q_{\pi\nu,\beta}} \partial_\alpha Q_{\pi\nu,\beta}\nonumber.
+\end{equation}
+The expansion of the functional derivatives in Equation \ref{eq-lc-emulsion-force} gives
+\begin{eqnarray}
+&&\partial_{\beta} \left(\frac{\partial F}{\partial (\partial_{\beta}\phi)} - \partial_\gamma \frac{\partial F}{\partial (\partial_\gamma\partial_\beta\phi)}\right) \partial_{\alpha}\phi
++\left(\frac{\partial F}{\partial (\partial_{\beta}\phi)} - \partial_\gamma \frac{\partial F}{\partial (\partial_\gamma\partial_\beta\phi)}\right) \partial_\beta\left(\partial_{\alpha}\phi\right)\nonumber\\
+&&\partial_{\beta} \left(\frac{\partial F}{\partial Q_{\pi\nu,\beta}} - \partial_\gamma \frac{\partial F}{\partial (\partial_\gamma Q_{\pi\nu,\beta)}}\right) \partial_{\alpha}Q_{\pi\nu}
++ \left(\frac{\partial F}{\partial Q_{\pi\nu,\beta}} - \partial_\gamma \frac{\partial F}{\partial (\partial_\gamma Q_{\pi\nu,\beta)}}\right) \partial_\beta Q_{\pi\nu,\alpha}\nonumber.
+\end{eqnarray}
+Using the symmetry of partial derivatives and $$\frac{\partial F}{\partial (\partial_\gamma\partial_\beta\phi)} = \frac{\partial F}{\partial (\partial_\gamma Q_{\mu\nu,\beta})}=0$$
+this can be recast with some manipulation as 
 \begin{eqnarray}
@@
 - \left(\frac{\partial F}{\partial\phi} 
-- \partial_{\beta} \frac{\partial F}{\partial \left(\partial_{\beta}\phi\right)}\right)\partial_{\alpha}\phi + \partial_{\alpha}\left(\mu\phi\right)\nonumber\\
-&=& -\left(\frac{\delta F}{\delta Q_{\pi\nu}}\right) \partial_{\alpha}Q_{\pi\nu} - \left(\frac{\delta F}{\delta \phi}\right) \partial_{\alpha} \phi + \partial_{\alpha}\left(\mu\phi\right)\nonumber\\ 
+- \partial_{\beta} \frac{\partial F}{\partial \left(\partial_{\beta}\phi\right)}\right)\partial_{\alpha}\phi + \partial_{\alpha}\left(\phi\mu\right)\nonumber\\
+&=& -\left(\frac{\delta F}{\delta Q_{\pi\nu}}\right) \partial_{\alpha}Q_{\pi\nu} - \left(\frac{\delta F}{\delta \phi}\right) \partial_{\alpha} \phi + \partial_{\alpha}\left(\phi\mu\right)\nonumber\\ 
 &=& H_{\pi\nu}\; \partial_{\alpha} Q_{\pi\nu} + \phi \;\partial_{\alpha}\mu,

### 2018-05-04 docs/references.tex
Trunk add LC emulsion document and relevant changes
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3079 40876441-3496-0410-a613-a2fca35b2ae2

@@
 %
+%  (c) 2011-2016 The University of Edinburgh
 %  Kevin Stratford (kevin@epcc.ed.ac.uk)
-%  (c) 2011 The University of Edinburgh
 %
@@
 
-\bibitem{swift1996}
-M.R. Swift, E. Orlandini, W.R. Osborn, and J.M. Yeomans,
-Lattice Boltzmann simulations of liquid-gas and binary fluid systems,
-\textit{Phys. Rev. E} \textbf{54} 5041--5052 (1996).
-
 \bibitem{succi}
@@
 Oxford University Press, Oxford, 2001.
+
+\bibitem{sulaiman2006}
+N. Sulaiman, D. Marenduzzo, and J.M. Yeomans,
+Lattice Boltzmann algorithm to simulate isotropic-nematic emulsions,
+\textit{Phys. Rev. E} \textbf{74} 041708 (2006).
+
+\bibitem{swift1996}
+M.R. Swift, E. Orlandini, W.R. Osborn, and J.M. Yeomans,
+Lattice Boltzmann simulation of liquid-gas and binary fluid systems,
+\textit{Phys. Rev. E}, \textbf{54} 5041 (1996).
 
@@
 \bibitem{Lyklema} J. Lyklema {\em Fundamentals of Interface and Colloid Science} Academic Press (1995).
-\bibitem{Mafe} S. Maf\'e, J.A. Manzanares, J. Pellicer, {\textit J. Electroanal. Chem.} {\textbf 241}, 5    7-77 (1988).
-\bibitem{Capuani} F. Capuani, I. Pagonabarraga, D. Frenkel, {\textit J. Chem. Phys.} {\textbf 121}, 973-    986 (2004).
+\bibitem{Mafe} S. Maf\'e, J.A. Manzanares, J. Pellicer, {\textit J. Electroanal. Chem.} {\textbf 241}, 57-77 (1988).
+\bibitem{Capuani} F. Capuani, I. Pagonabarraga, D. Frenkel, {\textit J. Chem. Phys.} {\textbf 121}, 973-986 (2004).
 \bibitem{Rotenberg} B. Rotenberg, I. Pagonabarraga, D. Frenkel, {\textit Farad. Discuss.} {\textbf 144}, 223-243 (2010).
-\bibitem{Landau-ED} L.D. Landau, E.M. Lifshitz, {\textit Electrodynamics of Continuous Media}, \S 15    , 2nd ed., Pergamon Press, Oxford, UK (1984).
+\bibitem{Landau-ED} L.D. Landau, E.M. Lifshitz, {\textit Electrodynamics of Continuous Media}, \S 15, 2nd ed., Pergamon Press, Oxford, UK (1984).
 \bibitem{Melcher} J.R. Melcher, {\textit Continuum Electromechanics}, \S 3.10, MIT Press, Cambridge, MA, USA (1981).\\
@@
 \url{http://ocw.mit.edu/ans7870/resources/melcher/resized/cem_811.pdf}
-\bibitem{Landau-EL} L.D. Landau, E.M. Lifshitz, {\textit Theory of Elasticity}, \S 3 \& \S 16, 3rd e    d., Butterworth-Heinemann, Oxford, UK (1986).
+\bibitem{Landau-EL} L.D. Landau, E.M. Lifshitz, {\textit Theory of Elasticity}, \S 3 \& \S 16, 3rd ed., Butterworth-Heinemann, Oxford, UK (1986).
 

### 2018-07-06 docs/intro.tex
VERSION 0.8.0

--- 2016-12-23 trunk/docs/intro.tex
@@
 %
-%  (c) 2016-2017 The University of Edinburgh
+%  (c) 2016-2018 The University of Edinburgh
 %
@@
 
-Under the auspices of the CCPs (Collaborative Computational Projects),
-and in particular CCP5, we have a CCPForge project account at
-\texttt{http://ccpforge.cse.rl.ac.uk/}
-which provides revision control via SVN, issue tracking, and so on.
+The code is publicly available from
+\texttt{http://github/ludwig-cf/ludwig/}
+which provides revision control via git, issue tracking, and so on.
 
-\paragraph{Users:}
-
-Code releases may be downloaded via anonymous download
-\begin{lstlisting}[style=terminalverbatim]
-http://ccpforge.cse.rl.ac.uk/gf/project/ludwig/frs/
-\end{lstlisting}
-The current release is available as \texttt{ludwig-0.6.0.tar.gz}.
-Alternatively, registered users can obtain the code via svn checkout
-in the same way as developers.
-
-\paragraph{Developers:}
-
-Developers need to create an account at CCPForge, and let one of the
-project owners\footnote{Currently: \texttt{kevin@epcc.ed.ac.uk},
-\texttt{ohenrich@epcc.ed.ac.uk}}
-know the user name so you can be added to the access list.
-
-To download the entire code, developers need
-\begin{lstlisting}
-svn checkout --username <user> http://ccpforge.cse.rl.ac.uk/svn/ludwig
-\end{lstlisting}
-where \texttt{<user>} is replaced by your CCPForge user name. You will
-see
-\begin{lstlisting}
-$ ls ludwig
-branches  tags      trunk
-\end{lstlisting}
-which is the usual svn convention for trunk, branches and tags.
-The process for committing code is described in section \ref{xref:developers}.
-
-If you look in  the \texttt{trunk} directory, you should see
-\begin{lstlisting}
-$ ls trunk
-config  LICENSE      mpi_s   src       tests  version.h
-docs    Makefile.mk  README  targetDP  util
-\end{lstlisting}
+The current stable release is available as the master branch.
 
@@
 contains particle data for time step 20 in a format described in
-Section~\ref{examples}.
+Section~\ref{section-examples}.
 These data may be requested in ASCII or raw binary format.

### 2018-07-06 targetDP/targetDP.h

@@ -1,281 +1,0 @@
-/*
- * targetDP.h: definitions, macros and declarations for targetDP.
- * Alan Gray
- * 
- * Copyright 2015 The University of Edinburgh
- *
- * Licensed under the Apache License, Version 2.0 (the "License");
- * you may not use this file except in compliance with the License.
- * You may obtain a copy of the License at
- *
- * http://www.apache.org/licenses/LICENSE-2.0
- *
- * Unless required by applicable law or agreed to in writing, software
- * distributed under the License is distributed on an "AS IS" BASIS,
- * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
- * See the License for the specific language governing permissions and
- * limitations under the License. 
- */
-
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-
-#ifndef TDP_INCLUDED_
-#define TDP_INCLUDED_
-
-/* KS. Additions */
-
-#include "cuda_stub_api.h" /* To be merged with target_api.h */
-#include "target_api.h"
-
-/* KS. End additions */
-
-
-/* Main settings */
-
-#ifndef VVL
-
-#define VVL_CUDA 1 /* virtual vector length for TARGETDP CUDA (usually 1) */
-#define VVL_C 1 /* virtual vector length for TARGETDP C (usually 1 for AoS */
-              /*  or a small multiple of the hardware vector length for SoA)*/
-
-#else /* allow this to be overwritten with compilation option -DVVL=N */
-
-#define VVL_CUDA VVL
-#define VVL_C VVL
-
-#endif
-
-/* End main settings */
-
-
-#ifdef __NVCC__ /* CUDA */
-
-
-/*
- * CUDA Settings 
- */
-
-#define DEFAULT_TPB 128 /* default threads per block */
-
-/* Instruction-level-parallelism vector length  - to be tuned to hardware*/
-#ifndef VVL
-#define VVL VVL_CUDA
-#endif
-
-/*
- * Language Extensions 
- */
-
-/* The __targetLaunch__ syntax is used to launch a function across 
- * a data parallel target architecture. */
-#define __targetLaunch__(extent) \
-  <<<((extent/VVL)+DEFAULT_TPB-1)/DEFAULT_TPB,DEFAULT_TPB>>>
-
-/* as above but with stride of 1 */
-#define __targetLaunchNoStride__(extent) \
-  <<<((extent)+DEFAULT_TPB-1)/DEFAULT_TPB,DEFAULT_TPB>>>
-  
-
[…]

### 2018-07-06 targetDP/target_cuda.h

@@ -1,32 +1,0 @@
-/****************************************************************************
- *
- *  target_cuda.h
- *
- ****************************************************************************/
-
-#ifndef TARGET_CUDA_H
-#define TARGET_CUDA_H
-
-#define CUDA_MAX_THREADS_PER_BLOCK 128
-
-#define tdp_cuda_simt_for_all(index, ndata, stride) \
-  index = (stride)*(blockIdx.x*blockDim.x + threadIdx.x);
-
-#define tdp_cuda_simt_for(index, ndata, stride) \
-  tdp_cuda_simt_for_all(index, ndata, stride) \
-  if (index < (ndata))
-
-
-#define tdp_cuda_simt_parallel_for(index, ndata, stride) \
-  tdp_cuda_simt_for(index, ndata, stride)
-/*
-#define tdp_cuda_launch(kernel_function, nblocks, ntpb, ...) \
-  kernel_function<<<nblocks, ntpb>>>(__VA_ARGS__)
-
-#define tdp_cuda_launch4s(kernel_function, nblocks, ntpb, shmem, stream, ...) \
-  kernel_function<<<nblocks, ntpb, shmem, stream>>>(__VA_ARGS__)
-*/
-#define tdp_cuda_simd_for(iv, nsimdvl) \
-  for (iv = 0; iv < (nsimdvl); iv++)
-
-#endif

### 2018-09-25 docs/tutorial/tutorial.tex
Updated access information

@@
 The system requirements are an up-to-date version of a C compiler as well as working installations 
-of a text editor like \texttt{vi} or \texttt{emacs}, Apache's Subversion (SVN) version control system, the visualisation tool
+of a text editor like \texttt{vi} or \texttt{emacs}, the visualisation tool
 ParaView (\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
@@
 
-The \texttt{Ludwig} project is currently hosted at our central repository 
-at CCPForge:\\
-(\hyperref[CCPForge]{https://ccpforge.cse.rl.ac.uk}).\\
-For full access to the repository please open a CCPForge account and request
-to join the project \texttt{Ludwig}. The following command grants anonymous access:
+The \texttt{Ludwig} project is currently hosted at our central repository at GitHub:\\
+(\hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}).\\
+
+\smallskip
+You can either download or clone the repository. 
+The following command clones the repository into \texttt{ludwig} and requires a working 
+git installation on your local system:
 
 \begin{lstlisting}[style=terminalverbatim]
-$ svn checkout https://ccpforge.cse.rl.ac.uk/svn/ludwig
+$ git clone https://github.com/ludwig-cf/ludwig
 \end{lstlisting}

### 2018-10-05 docs/tutorial/tutorial.tex

@@
 of a text editor like \texttt{vi} or \texttt{emacs}, the visualisation tool
-ParaView (\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
+ParaView \\
+(\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
 for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.
@@
 The \texttt{Ludwig} project is currently hosted at our central repository at GitHub:\\
-(\hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}).\\
+(\hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}).
+\smallskip
 
-\smallskip
 You can either download or clone the repository. 
-The following command clones the repository into \texttt{ludwig} and requires a working 
+The following command clones the repository and requires a working 
 git installation on your local system:
@@
 $ ls ludwig
-branches  tags  trunk
+LICENSE Makefile.mk config docs mpi_s src target tests util version.h
 \end{lstlisting}
-which is the usual svn layout for trunk, branches and tagged versions.
 
 A full documentation of \texttt{Ludwig} can be obtained by running 
-the command \texttt{make pdf} twice in the \texttt{/trunk/docs} directory:
+the command \texttt{make pdf} twice in the \texttt{/docs} directory:
 
 \begin{lstlisting}[style=terminalverbatim]
-$ cd /ludwig/trunk/docs/
+$ cd /ludwig/docs/
 $ make pdf
@@
 \begin{enumerate}
-\item Go to the directory \texttt{/ludwig/trunk/}: \\
+\item Go to the top level directory \texttt{/ludwig}: \\
 \begin{lstlisting}
-$ cd ludwig/trunk
+$ cd ludwig
 \end{lstlisting}
-\item Copy the configuration file \texttt{lunix-gcc-default.mk} to the \texttt{/trunk} directory 
+\item Copy the configuration file \texttt{lunix-gcc-default.mk} to top level directory 
 and rename to \texttt{config.mk} (both steps exectued below at once): \\
@@
 Note that you may need to specify the path to the MPI wrapper compiler in this file.
-\item Issue \texttt{make} in \texttt{/targetDP}: \\
+\item Issue \texttt{make} in \texttt{/target}: \\
 \begin{lstlisting}
-$ cd targetDP/
+$ cd target/
 $ make 
@@
 OPTS = -DNP_D3Q6 -DNDEBUG
-LIBS = -L../targetDP -ltarget -lm
-INC = -I. -I ../targetDP
+LIBS = -L../target -ltarget -lm
+INC = -I. -I ../target
 ...

### 2018-10-22 docs/tutorial/tutorial.tex
Amended tutorial to reflect latest changes in post-processing

@@
 \usepackage{float}
+\usepackage{hyperref}
 \graphicspath{{./tutorialFigs/}}
@@
 ParaView \\
-(\hyperref[ParaView]{https://www.paraview.org}), the MPI-library 
+({\color{blue}\hyperref[ParaView]{https://www.paraview.org}}), the MPI-library 
 for compiling the parallel version of the code and \texttt{gnuplot} for simple visualisations.
+The instructions for compilation of this tutorial and the full documentation assume a LaTeX distribution
+and \texttt{pdflatex} are installed.
 
@@
 The \texttt{Ludwig} project is currently hosted at our central repository at GitHub:\\
-(\hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}).
+({\color{blue} \hyperref[GitHubLudwig]{https://github.com/ludwig-cf/ludwig}}).
 \smallskip
@@
 A full documentation of \texttt{Ludwig} can be obtained by running 
-the command \texttt{make pdf} twice in the \texttt{/docs} directory:
+the command \texttt{make pdf} twice (for references) in the \texttt{/docs} directory:
 
@@
 $ make pdf
+\end{lstlisting}
+
+This tutorial and directories with input files and sample output is contained in 
+a subdirectory of \texttt{/docs}. It can be compiled in the following way:
+
+\begin{lstlisting}[style=terminalverbatim]
+$ cd /ludwig/docs/tutorial
+$ make
 \end{lstlisting}
@@
 This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}. 
+Note that more than one thread can be used in the compilation process,
+which can lead to a considerable speedup  
+\begin{lstlisting}
+$ make -j4 mpi
+\end{lstlisting} 
 Before going from a serial to a parallel compilation or vice versa
@@
 parallel and the executable \texttt{Ludwig.exe} should be copied into your working directory.
-
 \begin{enumerate}
-\item Copy the input file \texttt{input} located in the directory \texttt{/trunk/docs/tutorial/test1} 
-to your working directory. Note that this input file is a modified version of the 
-reference input file \texttt{input.ref} located in \texttt{/trunk/src} with a limited set of parameters
-for this run. 
+\item Copy the input file \texttt{input} located in the directory \texttt{/docs/tutorial/test1} 
+to your working directory. Note that the input files in this tutorial are modified versions of the 
+reference input file \texttt{input.ref} in \texttt{/src} with a limited set of parameters
+for each run.
 \item Run the code in your working directory by issuing the command:
@@
 \end{lstlisting}
+
 \item You should now see statistics being reported on your standard output and 
@@
 These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
-\item The first step to the post-processing of these data is to copy the files \texttt{vtk\_extract.c} 
-and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util} into your working directory.
-\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
+\item The first step to the post-processing of the data is to compile the \texttt{extract.c} 
+routine in \texttt{/util}. 
+\item Edit \texttt{extract.c}. Specifically, set the flags for:
 \begin{enumerate}
-\item \texttt{input\_binary\_=0} \\ if the velocity data is in ASCII format (see flag in input file),
-\item \texttt{output\_binary\_=0} \\ as the output of the postprocessing routines should also be in ASCII format,
-\item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k}, 
-\item \texttt{vtk\_header=0} \\ to omit headers (these will be required later for visualisation in \textit{ParaView}),
-\item \texttt{output\_cmf=0} to produce data in row-major format (column-major format is required for \textit{ParaView}).
+\item \texttt{input\_binary\_=0} \\ if the velocity data is in ASCII format (see flag in input file for \texttt{test1}),
+\item \texttt{output\_binary\_=0} \\ as the output of the postprocessing routines should also be in ASCII format for this tutorial,
+\item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k} for direct visualisation with \texttt{gnuplot}, 
 \end{enumerate}
-\item Edit \texttt{vtk\_extract\_script.py}. Set:
-\begin{enumerate}
-\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and increments of the velocity data
-to be processed.
-\item \texttt{vel=1} for velocity post-processing, 
-\item The other flags should be set to zero, e.g. \texttt{phi=0}. 
-\end{enumerate}
-\item Issue the command:
+\item Compile with 
 \begin{lstlisting}
-$ python vtk_extract_script.py
+$ make extract 
 \end{lstlisting} 
-\item You should now see files named \texttt{vel-XXXXXXXX.vtk} (Note: The extension \texttt{.vtk} is only used here for simplicity.
-It does not mean the generated data is in vtk-format). 
+and copy the executable \texttt{extract} into your working directory.
+\item Post-process the raw data by issuing 
+\begin{lstlisting}
+$ ./extract meta_file_name data_file_name_stub
+\end{lstlisting} 
+replacing the filenames with those of the present case (i.e. \texttt{vel.001-001.meta} and e.g. \texttt{vel-00010000}).
+\item You should now see files named \texttt{vel-XXXXXXXX}.
 The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.\\
-
-To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
+\item To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
 the spatial $x$-coordinate on the lattice with the command:
 \begin{lstlisting}
-gnuplot> p`vel-XXXXXXXX.vtk' u 1:5
+gnuplot> p `vel-XXXXXXXX' u 1:5
 \end{lstlisting} 
-By plotting the data in each file, you should be able to see the velocity profile converging to the final profile 
-as shown in Fig. \ref{fig:velocityProfile}.
+By plotting the data in each file, you should be able to see the velocity profile converging to the final profile as shown in Fig. \ref{fig:velocityProfile}.
 
-In the directory \texttt{/trunk/docs/tutorial/test1} a reference file of the final 
+In the directory \texttt{/docs/tutorial/test1} a reference file of the final 
 output at time step $10,000$ can be found, in addition to a file containing the standard output of the code.
@@
 \item Copy the input file \texttt{input} located in the directory 
-\texttt{/trunk/docs/tutorial/test2} 
+\texttt{/docs/tutorial/test2} 
 to your working directory. This input file contains the parameters required for this run. 
@@
 \end{lstlisting}
-\item As before, you should now see statistics being reported on your standard output but 
-in this case two sets of output files being created containing the velocity and compositional
-order parameter ($\phi$) data at each lattice point in the simulation. In this example, these 
-data are in binary format.
-\item The post-processing of these data again requires the files \texttt{vtk\_extract.c} 
-and \texttt{vtk\_extract\_script.py} from \texttt{/trunk/util}.
-\item Edit \texttt{vtk\_extract.c}. Specifically, set the flags for:
+\item As before, you should now see statistics being reported on your standard output. But 
+in this case two sets of output files are being created, one containing the velocity and 
+the other compositional order parameter ($\phi$) data at each lattice point in the simulation. 
+In this example, these data are in binary format.
+\item The post-processing of the data requires again a compiled exectuable of \texttt{extract.c} in
+\texttt{/util}, but this time with slightly different settings. Specifically, set the flags for:
 \begin{enumerate}
-\item \texttt{input\_binary\_=1} \\ if the velocity and $\phi$ data are in binary format (see 
-flag in input file),
+\item \texttt{input\_binary\_=1} \\ if the scalar order parameter and velocity data are in binary format (see 
+flag in input file for \texttt{test2}),
 \item \texttt{output\_binary\_=0} \\ as the output of the post-processing routines should be 
@@
 \item \texttt{output\_index=0} \\ as Cartesian lattice indices are not required, 
-\item \texttt{vtk\_header=1} \\ to include headers,
-\item \texttt{output\_cmf=1} to produce data in column-major format.
+\item \texttt{output\_cmf=1} to produce data in column-major format, which is required for the correct 
+visualisation with ParaView.
 \end{enumerate}
-\item Edit \texttt{vtk\_extract\_script.py}. Set:
+\item Compile with 
+\begin{lstlisting}
+make extract
+\end{lstlisting} 
+and copy the executable \texttt{extract} into your working directory.
+\item A python script \texttt{extract.py}
+can be used for convenient post-processing of multiple raw data files and time steps.
+ Edit \texttt{extract.py} and set:
 \begin{enumerate}
@@
 \begin{lstlisting}
-$ python vtk_extract_script.py
+$ python extract.py
 \end{lstlisting}
@@
 \begin{enumerate}
-\item Open the files containing the $\phi$ data.
+\item In Paraview open the files containing the $\phi$ data.
 \item Click on the group of files in the pipeline browser and apply a contour map using 
@@
 \item Copy the input file \texttt{input} located in the directory 
-\texttt{/trunk/docs/tutorial/test3} to your working directory. We see in this input file that 
+\texttt{/docs/tutorial/test3} to your working directory. We see in this input file that 
 there are many more parameters 
@@
 \item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids.
-Firstly compile the code in serial (see section \ref{configbuild})and go to the directory \texttt{ludwig/trunk/util} here you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}.
-These are the file required for the initialisation and visualisation of colloids respectively.
+Firstly, compile the \texttt{Ludwig} code in serial (see section \ref{configbuild}) and go to the directory \texttt{ludwig/util}.
+There you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}.
+These are the files required for the initialisation and post-processing of colloids, respectively.
 \item In the file, \texttt{colloid\_init.c} set the flags for:
@@
 \end{lstlisting}
-this produces a file, \texttt{config.cds.init.001-001} which should be copied into your working directory.
+this produces a file \texttt{config.cds.init.001-001}, which you have to copy into your working directory.
 \item In the file, \texttt{extract\_colloids.c} set:
@@
 \item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
-\item \texttt{cds\_with\_v  = 1} \\ which will output the velocities and positions of the colloids,
+\item \texttt{cds\_with\_v  = 1} \\ which will output the positions and velocity data of the colloids,
 \end{enumerate}
@@
 \end{lstlisting}
-and copy the executable \texttt{extract\_colloids} into your working directory along with the 
-usual post-processing scripts, \texttt{vtk\_extract.c} and \texttt{vtk\_extract\_script.py} 
-scripts. 
-\item Next, run the code in your working directory by issuing the command:
-\begin{lstlisting}
-$ mpirun -np 8 ./Ludwig.exe input
-\end{lstlisting}
-\item For the post-processing of this system, in \texttt{vtk\_extract.c}, set:
+and copy the executable \texttt{extract\_colloids} into your working directory.
+\item For the post-processing of this system, in \texttt{extract.c}, set:
 \begin{enumerate}
@@
 \item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
-\item \texttt{output\_lc\_op\_   = 1} \\ to visualise the liquid crystal order parameter,
-\item \texttt{output\_lc\_dir\_  = 1} \\ to visualise the liquid crystal director field,
-\item \texttt{output\_lc\_biop\_ = 0} \\ since the liquid crystals are not biaxial,
 \end{enumerate}
-\item and in \texttt{vtk\_extract\_script.py},
+compile in the usual manner and copy the post-processing executable \texttt{extract} and 
+the script \texttt{extract.py} as well into your working directory. 
+\item In \texttt{extract.py},
 \begin{enumerate}
@@
 \end{enumerate}
-\item After running the python script as before, \texttt{.csv} files with the colloid data should 
+
+\item Next, run the code in your working directory by issuing the command:
+\begin{lstlisting}
+$ mpirun -np 8 ./Ludwig.exe input
+\end{lstlisting}
+
+\item Run the post-processing routine as before using the python script. 
+\texttt{.csv} files with the colloid data should 
 be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
@@
 \item The liquid crystal, scalar order parameter can be visualised by opening the files: 
-\texttt{op-q-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
+\texttt{lcs-XXXXXXXX.vtk} and applying the `contour' filter. By setting a threshold 
 minimum $1 \times 10^{-4}$ the order parameter in the colloids can be ignored, this can be done using the tool bar:
@@
 
-\item For the director field open the \texttt{dir-q-XXXXXXXX.vtk} and apply the `glyphvector' filter. In this case, change the glyph 
+\item For the director field open the \texttt{lcd-XXXXXXXX.vtk} and apply the `glyphvector' filter. In this case, change the glyph 
 type from `arrow' to `line' and scale the colour of the lines by the $x$-component of the 

### 2018-10-22 docs/tutorial/tutorial.tex
More edits on tutorial

@@
 \item The first step to the post-processing of the data is to compile the \texttt{extract.c} 
-routine in \texttt{/util}. 
+routine in \texttt{/util}. This requires the \texttt{Ludwig} code to be compiled in serial 
+(see section \ref{configbuild}).
+
 \item Edit \texttt{extract.c}. Specifically, set the flags for:
@@
 \item The post-processing of the data requires again a compiled exectuable of \texttt{extract.c} in
-\texttt{/util}, but this time with slightly different settings. Specifically, set the flags for:
+\texttt{/util}, but this time with slightly different settings. 
+Again, this requires the \texttt{Ludwig} code to be compiled in serial 
+(see section \ref{configbuild}).
+Specifically, set the flags for:
 \begin{enumerate}
@@
 \item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids.
-Firstly, compile the \texttt{Ludwig} code in serial (see section \ref{configbuild}) and go to the directory \texttt{ludwig/util}.
+Firstly, and as before, compile the \texttt{Ludwig} code in serial (see section \ref{configbuild}) and go to the directory \texttt{ludwig/util}.
 There you should see the files \texttt{colloid\_init.c} and \texttt{extract\_colloid.c}.

### 2018-10-22 docs/tutorial/tutorial.tex
More edits in tutorial, removed flag settings

@@
 This tutorial and directories with input files and sample output is contained in 
-a subdirectory of \texttt{/docs}. It can be compiled in the following way:
+a subdirectory of \texttt{/docs}. It can be compiled in the following way (issue twice for TOC):
 
@@
 routine in \texttt{/util}. This requires the \texttt{Ludwig} code to be compiled in serial 
-(see section \ref{configbuild}).
-
-\item Edit \texttt{extract.c}. Specifically, set the flags for:
-\begin{enumerate}
-\item \texttt{input\_binary\_=0} \\ if the velocity data is in ASCII format (see flag in input file for \texttt{test1}),
-\item \texttt{output\_binary\_=0} \\ as the output of the postprocessing routines should also be in ASCII format for this tutorial,
-\item \texttt{output\_index=1} \\ to obtain Cartesian lattice indices \textit{i, j, k} for direct visualisation with \texttt{gnuplot}, 
-\end{enumerate}
-\item Compile with 
+(see section \ref{configbuild}). Compile with 
 \begin{lstlisting}
@@
 \begin{lstlisting}
-$ ./extract meta_file_name data_file_name_stub
+$ ./extract -a -i meta_file_name data_file_name_stub
 \end{lstlisting} 
 replacing the filenames with those of the present case (i.e. \texttt{vel.001-001.meta} and e.g. \texttt{vel-00010000}).
+The command line options '\texttt{-a}' and '\texttt{-i}' give output in ASCII format and a lattice index 
+for easy visualisation in \texttt{gnuplot}.
+
 \item You should now see files named \texttt{vel-XXXXXXXX}.
@@
 \begin{center}
-\includegraphics[width=0.6\linewidth]{velProf.png}
+\includegraphics[width=0.8\linewidth]{velProf.png}
   \caption{The parabolic velocity profile of a simple fluid in a cavity after $10,000$ time-steps.}
@@
 In this example, these data are in binary format.
-\item The post-processing of the data requires again a compiled exectuable of \texttt{extract.c} in
-\texttt{/util}, but this time with slightly different settings. 
-Again, this requires the \texttt{Ludwig} code to be compiled in serial 
-(see section \ref{configbuild}).
-Specifically, set the flags for:
-\begin{enumerate}
-\item \texttt{input\_binary\_=1} \\ if the scalar order parameter and velocity data are in binary format (see 
-flag in input file for \texttt{test2}),
-\item \texttt{output\_binary\_=0} \\ as the output of the post-processing routines should be 
-in ASCII format,
-\item \texttt{output\_index=0} \\ as Cartesian lattice indices are not required, 
-\item \texttt{output\_cmf=1} to produce data in column-major format, which is required for the correct 
-visualisation with ParaView.
-\end{enumerate}
-\item Compile with 
+\item It should be possible to use the same exectuable \texttt{extract} from the previous 
+example for the  post-processing. Alternatively, compile with 
 \begin{lstlisting}
@@
 and copy the executable \texttt{extract} into your working directory.
+Remember that this requires the \texttt{Ludwig} code to be compiled in serial 
+(see section \ref{configbuild}).
 \item A python script \texttt{extract.py}
@@
 \begin{center}
-\includegraphics[width=0.6\linewidth]{system.png}
+\includegraphics[width=0.8\linewidth]{system.png}
   \caption{A visualisation of a separating binary fluid using ParaView.}
@@
 
-\item You can move through different frames of the simulation with the arrow buttons:
+\item You can move through different frames of the simulation with the arrow buttons.
+As you do this, you should see the coarsening of the binary fluid.
 
@@
 \includegraphics[width=0.4\linewidth]{arrows.png}
-  \caption{A visualisation of a separating binary fluid using ParaView.}
+  \caption{Control buttons for the visualisation in ParaView.}
   \label{fig:changeFrame}
@@
 \end{figure}
-
-as you do this, you should see the coarsening of the binary fluid.
 
@@
 this produces a file \texttt{config.cds.init.001-001}, which you have to copy into your working directory.
+
+\item Next, run the code in your working directory by issuing the command:
+\begin{lstlisting}
+$ mpirun -np 8 ./Ludwig.exe input
+\end{lstlisting}
+
 \item In the file, \texttt{extract\_colloids.c} set:
@@
 and copy the executable \texttt{extract\_colloids} into your working directory.
-\item For the post-processing of this system, in \texttt{extract.c}, set:
-\begin{enumerate}
-\item \texttt{input\_binary\_  = 1} \\ due to binary output from code,
-\item \texttt{output\_binary\_ = 0} \\ for output to be visualised in ASCII,
-\end{enumerate}
-compile in the usual manner and copy the post-processing executable \texttt{extract} and 
-the script \texttt{extract.py} as well into your working directory. 
-\item In \texttt{extract.py},
+\item For the post-processing of other data either copy the existing executable \texttt{extract} from 
+previous examples or compile in \texttt{/util} in the usual manner.
+Remember that this requires the \texttt{Ludwig} code to be compiled in serial 
+(see section \ref{configbuild}).
+
+\item Copy the script \texttt{extract.py} as well into your working directory and set:
 \begin{enumerate}
@@
 \end{enumerate}
-
-\item Next, run the code in your working directory by issuing the command:
-\begin{lstlisting}
-$ mpirun -np 8 ./Ludwig.exe input
-\end{lstlisting}
 
@@
 \begin{center}
-\includegraphics[width=0.6\linewidth]{colloidSystem.png}
+\includegraphics[width=0.8\linewidth]{colloidSystem.png}
   \caption{The visualisation of the colloidal suspension in liquid crystals.}

### 2019-08-14 docs/intro.tex

@@
 %
-%  (c) 2016-2018 The University of Edinburgh
+%  (c) 2016-2019 The University of Edinburgh
 %
@@
 
-This section provides a brief overview of compiling and running
-the code, which should allow users to get off the ground.
-The section assumes the code has been downloaded and you are in the
-top level directory.
-
-\subsection{Configuration}
-
-You will need to copy one of the existing configuration files from
-the \texttt{config} directory to the top level directory. This file
-sets all the relevant local details for compilation and so on. For
-example:
-\begin{lstlisting}
-$ cp config/lunix-gcc-default.mk config.mk
-\end{lstlisting}
-Edit the new \texttt{config.mk} file to check, or adjust, the options
-as required. You will need to identify your local C compiler. The
-example uses \texttt{gcc}
-in serial, and \texttt{mpicc} in parallel. For MPI in particular,
-local details may vary. A quick way to compile the code is via
-\begin{lstlisting}
-$ cd tests
-$ make compile-mpi-d3q19
-\end{lstlisting}
-This integrates a number of separate stages which are described
-indivually in the following sections. The resulting executable
-will be \texttt{./src/Ludwig.exe} at the top level.
-
-\subsubsection{targetDP}
-
-A \texttt{targetDP} layer is required by the code in all cases. It
-allows compilation for both simple CPU systems and
-for a number of accelerator systems using a single source code. This
-will make the appropriate choices depending what has been specified
-in the top-level \texttt{config.mk} file. From the top-level directory
-\begin{lstlisting}
-$ cd targetDP
-$ make
-\end{lstlisting}
-should produce the appropriate \texttt{libtarget.a} library.
-
-\subsubsection{Serial}
-\label{section:quick-serial}
-
-First, compile the MPI stub library in the \texttt{mpi\_s}
-directory. You should be able to build the library and run the
-tests using, e.g.:
-
-\begin{lstlisting}
-$ make libc
-gcc -Wall -I.  -c mpi_serial.c
-ar -cru libmpi.a mpi_serial.o
-$ make testc
-gcc -Wall   mpi_tests.c -L. -lmpi
-./a.out
-Running mpi_s tests...
-Finished mpi_s tests ok.
-\end{lstlisting}
-
-Now compile the main code in the \texttt{src} directory.
-Compilation should look like:
-
-\begin{lstlisting}
-$ make serial
-make serial-d3q19
-make serial-model "LB=-D_D3Q19_" "LBOBJ=d3q19.o"
-make lib
-make libar "INC=-I. -I../mpi_s" "LIBS= -L../mpi_s -lmpi -lm"
-gcc -D_D3Q19_ -Wall -O2 -I. -I../mpi_s -c model.c
-gcc -D_D3Q19_ -Wall -O2 -I. -I../mpi_s -c propagation.c
-...
-\end{lstlisting}
-which shows that the default lattice Boltzmann model is D3Q19.
-Successful compilation will provide an executable \texttt{Ludwig.exe}
-which is linked against the MPI stub library.
-
-\subsubsection{Parallel}
-
-Here, you do not need to compile the stub library. Simply compile
-the main code with, e.g.,:
-\begin{lstlisting}
-$ make mpi
-make mpi-d3q19
-make mpi-model "LB=-D_D3Q19_" "LBOBJ=d3q19.o"
-make libmpi
-make libar "CC=mpicc"
-mpicc -D_D3Q19_ -Wall -O2 -I. -c model.c
-mpicc -D_D3Q19_ -Wall -O2 -I. -c propagation.c
-...
-\end{lstlisting}
-
-Again, this should produce an executable \texttt{./Ludwig.exe}
-in the current directory, this time linked against the true MPI
-library.
-
 \subsection{Running an Example}
@@
 \begin{lstlisting}
-$ cp ../tests/regression/d3q19/serial-auto-c01.inp .
+$ cp ../tests/regression/d3q19-extra/serial-auto-c01.inp .
 \end{lstlisting}
@@
 \lstinputlisting[firstline=1, lastline= 16]
-{../tests/regression/d3q19/serial-auto-c01.inp}
+{../tests/regression/d3q19-extra/serial-auto-c01.inp}
 Here, the first key value pair \texttt{N\_cycles 40} sets the number
@@
 \lstinputlisting[aboveskip=0pt, firstline=1, lastline= 16]
-{../tests/regression/d3q19/serial-auto-c01.log}
+{../tests/regression/d3q19-extra/serial-auto-c01.log}
 This output shows that the appropriate input file has been read, and

### 2019-08-14 docs/tutorial/tutorial.tex

@@
 %
-%  (c) 2008-2017 The University of Edinburgh
+%  (c) 2008-2019 The University of Edinburgh
 %
@@
 $ ls ludwig
-LICENSE Makefile.mk config docs mpi_s src target tests util version.h
+CHANGES.md      Makefile        config          src             util
+CONTRIBUTING.md Makefile.mk     docs            target          version.h
+LICENSE         README.md       mpi_s           tests
 \end{lstlisting}
@@
 \begin{lstlisting}[style=terminalverbatim]
-$ cd /ludwig/docs/
+$ cd ./ludwig/docs/
 $ make pdf
@@
 \begin{lstlisting}[style=terminalverbatim]
-$ cd /ludwig/docs/tutorial
+$ cd ./ludwig/docs/tutorial
 $ make
@@
 \end{lstlisting}
-\item Copy the configuration file \texttt{lunix-gcc-default.mk} to top level directory 
+\item Copy the (parallel) configuration file \texttt{lunix-mpicc-default.mk}
+to top level directory 
 and rename to \texttt{config.mk} (both steps exectued below at once): \\
 \begin{lstlisting}
-$ cp config/lunix-gcc-default.mk ./config.mk
+$ cp config/unix-mpicc-default.mk ./config.mk
 \end{lstlisting}
-Note that you may need to specify the path to the MPI wrapper compiler in this file.
-\item Issue \texttt{make} in \texttt{/target}: \\
+Note that you may need to specify the correct path to the MPI wrapper compiler
+in this file (assumed to be \texttt{mpicc}).
+\item Issue \texttt{make} in the top level directory: \\
 \begin{lstlisting}
-$ cd target/
 $ make 
 \end{lstlisting}
-This creates dummy functions for calls to the targetDP library.
-\item Issue \texttt{make} in \texttt{/mpi\_s}: \\
-\begin{lstlisting}
-$ cd ../mpi_s/
-$ make 
-\end{lstlisting}
-This creates dummy functions for calls to the MPI-library,
-which is required for serial compilation.
-\item Edit \texttt{Makefile} in \texttt{/src}. To switch off 
-the assertions set \texttt{OPTS = -DNP\_D3Q6 -DNDEBUG}. This is
-important for production runs (see also comment in \texttt{Makefile}): \\
-\begin{lstlisting}
-$ cd ../src/
-$ vim Makefile
-...
-
-include ../Makefile.mk
-
-MAIN = main
-EXECUTABLE = Ludwig.exe
-LIBRARY = libludwig.a
-
-OPTS = -DNP_D3Q6 -DNDEBUG
-LIBS = -L../target -ltarget -lm
-INC = -I. -I ../target
-...
-\end{lstlisting}
-\item To compile the code in serial issue: \\
-\begin{lstlisting}
-$ make serial
-\end{lstlisting} 
-To compile the parallel version of the code issue: \\
-\begin{lstlisting}
-$ make mpi
-\end{lstlisting} 
-This creates the executable file \texttt{Ludwig.exe} in \texttt{/src}. 
+This creates the executable file \texttt{./Ludwig.exe} in \texttt{/src}. 
 Note that more than one thread can be used in the compilation process,
@@
 \begin{lstlisting}
-$ make -j4 mpi
+$ make -j 4
 \end{lstlisting} 
-Before going from a serial to a parallel compilation or vice versa
-you should clean the directory by issuing
+If you want a serial build, use the conifguation file
+\texttt{config/unix-gcc-default.mk} and include the extra step
 \begin{lstlisting}
-$ make clean
+$ make serial
+$ make
 \end{lstlisting}

### 2020-02-02 docs/input.tex
trunk 0.6.6 removed stray output in physics
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@3151 40876441-3496-0410-a613-a2fca35b2ae2

--- 2016-12-21 branches/kevin-1397/docs/input.tex
@@
 
-\subsubsection{Liquid crystal anshoring}
+\subsubsection{Liquid crystal anchoring}
 
@@
 
-\subsection{Porous media}
-
-Porous media calculations can be undertaken when appropriate
-solid/fluid status information is supplied. The are a number
-of switches available in the input file:
-\begin{lstlisting}
-porous_media_file
-porous_media_format
-porous_media_type
-\end{lstlisting}
-
-specifies the file stub name to be read at the start of execution.
-(If the stub name is \texttt{file} then the code will expect to
-find \texttt{file.001-001} in the current directory.
-
-is either \texttt{ASCII} or \texttt{BINARY} as appropriate. Note that
-in parallel, a single data file can be supplied, but it must be binary.
-The default is \texttt{BINARY}.
-
-is either \texttt{status\_only} (the default) or \texttt{status\_with\_h}.
-
-\subsubsection{File format}
-
-In all cases the status (fluid/solid) information is represented by a
-single \texttt{char} (or integer), which must be supplied via the
-porous media file. A single file should contain data matching the
-current system size, and have the $z-$direction running fastest,
-followed by the $y-$direction. Note that in non-periodic directions,
-the structure must be 'closed', i.e., all the points at the edge
-should be solid.
-
-Fluid sites are designated by \texttt{0} and boundary or solid sites
-by \texttt{1}. These data should be of type \texttt{char} in binary,
-and may be integer in ASCII.
-
-Where wetting information is required, the free energy parameter $H$
-can be supplied by using the \texttt{status\_with\_h} switch. In this
-case, the \texttt{char} status is augmented by a single \texttt{double}
-value which is the local value of $H$. The order is then
-$s_1,h_1, s_2, h_2, \ldots$.
-
-An example of how to construct a porous media file is provided in
-\texttt{util/capillary.c}, which builds an appropriate file for
-a square or circular capillary tube. Please see the comments in
-the file for further details. Note that the allowed
-solid/fluid status values are defined in \texttt{src/site\_map.h}.
-A solid boundary is \texttt{BOUNDARY}, while fluid is \texttt{FLUID}.
-
 \subsection{Order parameter initialisations}

### 2020-04-14 docs/devel.tex

@@
 
-The tests use the lightweight implementation of exceptions provided
-under the GPL Lesser General Public License by Guillermo Calvo. This
-is included with the source.
-
-\subsubsection{Makefile}
-
-The \texttt{Makefile.mk} and \texttt{config.mk} files in the top level
-directory control options for compilation. Before compilation,
-the \texttt{config.mk} file must be edited provide details of the local
-C compiler(s). The relevant lines are usually limited to, e.g.:
-\begin{lstlisting}
-CC = cc
-MPICC = mpicc
-CFLAGS = -O2
-\end{lstlisting}
-where \texttt{CC} is the compiler used for serial compilation,
-\texttt{MPICC} is the compiler used for compilation against
-MPI, and \texttt{CFLAGS} provides compiler switches (often
-related to optimisation).
-
-The \texttt{Makefile} provided in each subdirectory includes
-the \texttt{config.mk} file via the top level \texttt{Makefile.mk}.
-Changes to the individual Makefiles should therefore not be required.
-
-If GPU compilation is required, the \texttt{config.mk} file should
-specify the local details concerning \texttt{nvcc}. This includes
-explicit information on the location of MPI headers and libraries
-if an MPI-parallel GPU version is required.
-
-\subsubsection{C assertions}
-
-The code makes quite a lot of use of standard C assertions, which
-are useful to prevent errors. They do result in a considerably
-slower execution in some instances, so production runs should
-switch off the assertions with the standard \texttt{NDEBUG}
-preprocessor flag. Add \texttt{-DNDEBUG} to \texttt{CFLAGS} in
-the \texttt{config.mk}.
-
-\subsubsection{Address models}
-
-The default memory layout is array of structures (AOS). Compiler
-preprocessor macros for blocked SOA (\texttt{-DADDR\_AOSOA}) or
-structure of arrays (\texttt{-DADDR\_SOA})
-may be appropriate for different platforms.
-
-Add the relevant flags to the \texttt{config.mk} file before
-compilation to effect the given address model.
-
-\subsubsection{Targets for serial and parallel code}
-
-The code can be compiled with or without a true MPI library.
-For serial execution, the MPI stub library must be compiled
-first (see ``Quick Start for Users'' Section \ref{section:quick-serial}).
-The appropriate target is:
-\begin{lstlisting}
-bash-3.2$ make serial
-make serial-d3q19
-make serial-model ``LB=-D_D3Q19_'' ``LBOBJ=d3q19.o''
-...
-\end{lstlisting}
-from which it will be seen that the default LB model is D3Q19. This is
-the same for the true MPI target
-\begin{lstlisting}
-bash-3.2$ make mpi
-make mpi-d3q19
-make mpi-model ``LB=-D_D3Q19_'' ``LBOBJ=d3q19.o''
-...
-\end{lstlisting}
-
-\subsubsection{Targets for different LB models}
-
-Specific targets are provided if an alternative LB model is wanted.
-The available options are:
-\begin{lstlisting}
-make [ serial-d2q9 | serial-d3q15 | serial-d3q19 ]
-make [ mpi-d2q9 | mpi-d3q15 | mpi-d3q19 ]
-\end{lstlisting}
-A further set of targets is supplied if `reverse' or structure-of-array
-memory ordering is wanted (e.g., for GPU computing):
-\begin{lstlisting}
-make [ serial-d2q9r | serial-d3q15r | serial-d3q19r ]
-make [ mpi-d2q9r | mpi-d3q15r | mpi-d3q19r ]
-\end{lstlisting}
-This will only influence efficiency of the code: results are unchanged.
-
-\subsection{Tests}
-
-A series of tests are provided in the \texttt{./tests} directory. These
-are of two types. Unit tests are generally written when units of code
-are first introduced to ensure basic operation is error free. The unit
-tests are encoded in an executable linked against the appropriate
-\textit{Ludwig} library.
-Regression
-tests are introduced and updated when physical results are (broadly)
-validated. The regression tests use the stand-alone executable for
-the appropriate model, and read input files which define the different
-tests.
-Both are run regularly as the basis of a nightly test procedure, but
-they can be run independently, e.g., after adding or changing code.
-
-\subsubsection{Running the tests}
-
-A number of different \texttt{Makefile} targets are provided for
-running serial, parallel,
-or GPU tests. For each test, there is the option to run either
-the relevant unit tests, or regression tests, or both. For example, to
-run both serial unit tests and serial regression tests for the D3Q19
-model, invoke
-\begin{lstlisting}
-$ cd tests
-$ make compile-run-serial
-\end{lstlisting}
-in the test directory.
-The model selection follows the naming scheme described for compilation
-in the previous section.
-
-\subsection{Additional Notes for Developers}
-\label{xref:developers}
-
-Developers who wish to contribute code to the SVN repository
-please consider the following pleas concerning standards.
-
-\subsubsection{Coding standards}
-
-While definitive statements on style are avoided, please try to
-maintain some basic standards: (0) code should be strictly ANSI C99
-standard; (1) code should compile without warnings when appropriate
-compiler flags are set (e.g., \texttt{-Wall} under \texttt{gcc});
-(2) avoid long lines of code for readability reasons;
-(3) avoid confusing commented-out code; (4) avoid conditional pre-processor
-directives if possible; (5) add meaningful and descriptive comments;
-(6) use standard \texttt{assert()} to trap programming errors;
-(7) explicitly trap possible run time errors (8) add appropriate
-tests.
-
-\subsubsection{Documentation standards}
-
-Additions and alterations to the code need to be reflected in the
-documentation. These should be checked in at the same time as the
-code itself.
-
-\subsubsection{Protocol}
-
-Before checking in code, please follow procedure: (1) increment the
-patch version number in \texttt{version.h} consistent with the SVN
-(if a minor version increment is required, communication with
-other developers must be considered); (2) run at least the unit
-tests and the short regression tests; (3) add a note to the change
-log; and (4) SVN update and commit.
-
 \vfill

### 2022-04-08 docs/tutorial/tutorial.tex

@@
 
-\setcounter{page}{1}
-
-% We're not including the table of contents just at the moment
-%\tableofcontents
-
-\newpage
 \lstset{style=terminalverbatim}
@@
 procedure. The configuration step is based on the GNU Compiler Collection gcc.
-
 \begin{enumerate}
-\item Go to the top level directory \texttt{/ludwig}: \\
+\item Go to the top level directory \texttt{/ludwig}:
 \begin{lstlisting}
@@
 to top level directory 
-and rename to \texttt{config.mk} (both steps exectued below at once): \\
+and rename to \texttt{config.mk} (both steps exectued below at once):
 \begin{lstlisting}
@@
 in this file (assumed to be \texttt{mpicc}).
-\item Issue \texttt{make} in the top level directory: \\
+
+For production runs you may want to use thie compiler flag below instead of the default one:
+\begin{lstlisting}
+CFLAGS  = -O3 -g -DNDEBUG -DNSIMDVL=4
+\end{lstlisting}
+\item Issue \texttt{make} in the top level directory:
 \begin{lstlisting}
@@
 \end{lstlisting} 
-If you want a serial build, use the conifguation file
-\texttt{config/unix-gcc-default.mk} and include the extra step
+If you want a serial build, clean your directory tree with
+\begin{lstlisting}
+$ make clean
+\end{lstlisting}
+use the conifguation file
+\texttt{config/unix-gcc-default.mk} and include the extra keyword \texttt{serial}:
 \begin{lstlisting}
 $ make serial
-$ make
 \end{lstlisting}
-
 \end{enumerate}
@@
 \end{lstlisting}
-
 \item You should now see statistics being reported on your standard output and 
@@
 These files contain the components of the velocity vectors in Cartesian coordinates at each lattice point in the simulation. 
-\item The first step to the post-processing of the data is to compile the \texttt{extract.c} 
-routine in \texttt{/util}. This requires the \texttt{Ludwig} code to be compiled in serial 
-(see section \ref{configbuild}). Compile with 
-\begin{lstlisting}
-$ make extract 
-\end{lstlisting} 
-and copy the executable \texttt{extract} into your working directory.
+\item The first step to the post-processing is to get hold of the \texttt{extract} executable in \texttt{/util},
+which is created during the main compilation of \texttt{Ludwig}, and copy it into your tutorial working directory. 
 \item Post-process the raw data by issuing 
@@
 for easy visualisation in \texttt{gnuplot}.
-
-\item You should now see files named \texttt{vel-XXXXXXXX}.
-The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.\\
+\item You should now see files named \texttt{vel-XXXXXXXX}. The first three columns of these ASCII files are the spatial lattice indices followed by the Cartesian components of the velocity vectors.
 \item To plot the velocity profile of the system, you can plot the $y$-component of the velocity against 
@@
 \end{enumerate} 
-
 \begin{figure}[h]
@@
 In this example, these data are in binary format.
-\item It should be possible to use the same exectuable \texttt{extract} from the previous 
-example for the  post-processing. Alternatively, compile with 
+\item You are able to use the same exectuable \texttt{extract} from the previous 
+example for the  post-processing with different arguments and the \texttt{-k} flag to produce
+files in VTK-format:
 \begin{lstlisting}
-make extract
+$ ./extract -k vel.001-001.meta vel-00010000.001-001
+$ ./extract -k phi.001-001.meta phi-00010000.001-001
 \end{lstlisting}
-and copy the executable \texttt{extract} into your working directory.
-Remember that this requires the \texttt{Ludwig} code to be compiled in serial 
-(see section \ref{configbuild}).
-\item A python script \texttt{extract.py}
-can be used for convenient post-processing of multiple raw data files and time steps.
- Edit \texttt{extract.py} and set:
-\begin{enumerate}
-\item \texttt{nstart=1000}, \texttt{nint=1000} and \texttt{nend=10000} to set the range and 
-increments of the velocity data
-to be processed.
-\item \texttt{vel=1} for velocity post-processing, 
-\item \texttt{phi=1} for $\phi$ post-processing,
-\item The other flags should be set to zero. 
-\end{enumerate}
-\item Issue the command:
+\item A bash script like this one below can be used for convenient post-processing of multiple 
+raw data files and time steps (with appropriate start, increment and end in the \texttt{for} loop):
 \begin{lstlisting}
-$ python extract.py
+#!/bin/bash
+for i in `seq 1000 1000 10000`;
+do
+  tstep=$(printf "%08d" $i)
+  ./extract -k vel.001-001.meta vel-${tstep}.001-001  
+  ./extract -k phi.001-001.meta phi-${tstep}.001-001
+done
 \end{lstlisting}
@@
 \texttt{Ludwig.exe} should be copied into your working directory.
-
 \begin{enumerate}
@@
 radius and $\delta$ - here set to 0.5 - catches any colloid-colloid interactions. This is shown for our example - where this value is set to 8.0 - in Fig. \ref{fig:colloid_int}.
-
  \begin{figure}[H]
@@
 \end{figure}
-
 \item The following steps outline the preliminary stages required in order to initialise the colloidal system and for the visualisation of colloids.
@@
 \end{enumerate}
-\item Now issue: \\
+\item Now issue
 \begin{lstlisting}
@@
 this produces a file \texttt{config.cds.init.001-001}, which you have to copy into your working directory.
-
 \item Next, run the code in your working directory by issuing the command:
@@
 \end{lstlisting}
-
-\item In the file, \texttt{extract\_colloids.c} set:
+\item An \texttt{extract\_colloids} executable is produced in \texttt{/util} during the main compilation of \texttt{Ludwig}. 
+For our problem we need to set potentially different parameters. In the file, \texttt{extract\_colloids.c} set:
 \begin{enumerate}
 \item The system size, \texttt{NX},  \texttt{NY},  \texttt{NZ} all = 32,
-\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII,
+\item \texttt{iread\_ascii = 1} \\ since the output is in ASCII format,
 \item \texttt{cds\_with\_v  = 1} \\ which will output the positions and velocity data of the colloids,
 \end{enumerate}
-\item Now issue: \\
+\item Now issue
 \begin{lstlisting}
@@
 previous examples or compile in \texttt{/util} in the usual manner.
-Remember that this requires the \texttt{Ludwig} code to be compiled in serial 
-(see section \ref{configbuild}).
-
-\item Copy the script \texttt{extract.py} as well into your working directory and set:
-\begin{enumerate}
-\item \texttt{vel = 1}, \\ to visualise the velocity field,
-\item \texttt{q = 1}, \\ since there are liquid crystals present,
-\item \texttt{colcds = 0} and 
-\item \texttt{colcdsvel = 1} so that the colloid position and velocity data are both available.
-\end{enumerate}
-
-\item Run the post-processing routine as before using the python script. 
+\item Run the post-processing routines as before using different arguments:
+\begin{lstlisting}
+$ ./extract -k vel.001-001.meta vel-00010000.001-001
+$ ./extract -k -s -d q.001-001.meta q-00010000.001-001
+$ ./extract_colloids config.cds00010000 1 col-cdsvel-00010000.csv
+\end{lstlisting}
+The \texttt{-s} and \texttt{-d} flags produce scalar order parameter and director field output.
 \texttt{.csv} files with the colloid data should 
 be seen in addition to the \texttt{.vtk} files for the velocity, order parameter and director field.
+\item Again, a bash script like this one below can be used for convenient post-processing of multiple raw data files and time steps (with appropriate start, increment and end in the \texttt{for} loop):
+\begin{lstlisting}
+#!/bin/bash
+for i in `seq 1000 1000 10000`;
+do
+  tstep=$(printf "%08d" $i)
+  ./extract -k vel.001-001.meta vel-${tstep}.001-001  
+  ./extract -k -s -d q.001-001.meta q-${tstep}.001-001
+  ./extract_colloids config.cds${tstep} 1 col-cdsvel-${tstep}.csv
+done
+\end{lstlisting}
 \item The system can now be visualised, again using ParaView:

### 2022-08-15 docs/tutorial/tutorial.tex

@@
 #!/bin/bash
-for i in `seq 1000 1000 10000`;
+for i in $(seq 1000 1000 10000);
 do

### 2023-07-07 docs/intro.tex

@@
 %
-%  (c) 2016-2019 The University of Edinburgh
+%  (c) 2016-2023 The University of Edinburgh
+%
+%  This material is being migrated to https://ludwig.epcc.ed.ac.uk/index.html
 %
@@
 
-\vfill
-\pagebreak
-
-\section{Quick Start for Users}
-
-\subsection{Running an Example}
-
-\subsubsection{Input file}
-
-The behaviour of \textit{Ludwig} at run time is controlled by
-a plain text input file which consists of a series for key value
-pairs. Each key value pair controls a parameter or parameters
-within the code, for example the system size and number of time steps,
-the fluid properties density and viscosity, the free energy type
-and associated parameters (if required), frequency and type of
-output, parallel decomposition, and so on.
-
-For most keys, the associated property has some default value which
-will be used by the code if the relevant key is not present (or
-commented out) in the input file. While such default values are
-chosen to be at least sensible, users need to be aware that all
-necessary keys need to be considered for a given problem. However,
-many keys are irrelevant for any given problem.
-
-A significant number of example input files are available as
-part of the test suite, and these can form a useful starting
-point for a given type of problem. We will consider one which
-uses some of the common keys.
-
-\subsubsection{Example input}
-
-We will consider a simple example which computes the motion of
-a single spherical colloidal particle in an initially stationary
-fluid of given viscosity. The velocity may be measured as a
-function of time. Assume we have an executable in the \texttt{src}
-directory.
-
-Make a copy of the input file
-\begin{lstlisting}
-$ cp ../tests/regression/d3q19-extra/serial-auto-c01.inp .
-\end{lstlisting}
-
-Inspection of this file will reveal the following: blank lines
-and comments --- lines beginning with \texttt{\#} --- may be included,
-but are ignored at run time; other lines are parsed as key value
-pairs, each pair on a separate line, and with key and value separated
-by one or more spaces. Values may be characters or strings, a
-scalar integer or 3-vector of integers, or a scalar or 3-vector
-of floating point numbers. Values which are 3-vectors are delineated
-by an underscore character (not a space), e.g., \texttt{1\_2\_3},
-and always correspond to a Cartesian triple $(x,y,z)$.
-A given value is parsed appropriately in the context of the associated key.
-
-So, the first few lines of the above file are:
-\lstinputlisting[firstline=1, lastline= 16]
-{../tests/regression/d3q19-extra/serial-auto-c01.inp}
-Here, the first key value pair \texttt{N\_cycles 40} sets the number
-of time steps to 40, while the second, \texttt{size 64\_64\_64}, sets
-the system size to be 64 points in each of the three coordinate
-directions. The \texttt{grid} key relates to the parallel
-domain decomposition as is discussed in the following
-section.
-
-\subsubsection{Run time}
-
-The executable takes a single argument on the command line which
-is the name of the input file, which should be in the same
-directory as the executable. If no input file name is supplied,
-the executable will look for a default \texttt{input}. If no
-input file is located at all, an error to that effect will be
-reported.
-
-\textbf{Serial: }
-If a serial executable is run in the normal way with the copy of the
-input file as the argument the code should take around 10--20 seconds
-to execute 40 steps. The fist few lines of output should look like:
-
-\begin{lstlisting}[belowskip=0pt]
-$ ./Ludwig.exe ./serial-auto-c01.inp
-\end{lstlisting}
-\lstinputlisting[aboveskip=0pt, firstline=1, lastline= 16]
-{../tests/regression/d3q19-extra/serial-auto-c01.log}
-This output shows that the appropriate input file has been read, and
-the system size set correspondingly with a number of default settings
-including periodic boundary conditions. ``No free energy'' tells us we
-are using a single Newtonian fluid.
-
-Normal termination of execution is accompanied by a report
-of the time take by various parts of the code, and finally by
-\begin{lstlisting}
-...
-Ludwig finished normally.
-\end{lstlisting}
-
-\textbf{Parallel: }
-The executable compiled and linked against the MPI library can
-be run with the MPI launcher for the local system, often
-\texttt{mpirun}. For example, a run on 8 MPI tasks produces a
-similar output to that seen in the serial case, but reports
-details of the local domain decomposition:
-\begin{lstlisting}
-$ mpirun -np 8 ./Ludwig.exe ./serial-auto-c01.inp
-Welcome to Ludwig v0.1.26 (MPI version running on 8 processes)
-...
-System details
---------------
-System size:    64 64 64
-Decomposition:  2 2 2
-Local domain:   32 32 32
-...
-\end{lstlisting}
-The decomposition is controlled by the \texttt{grid} key in
-the input. Here, \texttt{grid 2\_2\_2} is consistent with the
-8 MPI tasks specified on the command line, and the resulting
-local decomposition is 32 lattice points in each coordinate direction.
-If no grid is specified in the input, or a request is
-make for a grid which cannot be met (e.g., the product
-of the grid dimensions does not agree with the total number
-of MPI tasks available)
-the code will try to determine its own decomposition as best
-it can. If no valid parallel decomposition is available at all,
-the code will exit with a message to that effect.
-
-Further details of various input key value pairs are given in
-relevant sections of the documentation.
-
-\subsubsection{Output}
-
-The standard output of the running code produces a number of
-aggregate quantities which allow a broad overview of the progress
-of the computation to be seen by the user. These include, where
-relevant, global statistics related to fluid density and momentum,
-the integrated free energy, particle-related quantities, and so on.
-The frequency of this information can be adjusted from the input
-file (see, e.g., \texttt{freq\_statistics}).
-
-Output for lattice-based quantities (for example, the velocity field
-$u(\mathbf{r}; t)$) is via external file. This output is in serial
-or in parallel according to how the model is run, and may be in
-either ASCII or raw binary format as required. Output files are produced
-with time step encoded in the file, and an extension which describes the
-parallel output decomposition. Further, each quantity output in this way
-is accompanied by a metadata description with \texttt{meta} extension.
-For example, the two output files
-\begin{lstlisting}
-bash-3.2$ ls dist*
-dist-00000020.001-001   dist.001-001.meta
-\end{lstlisting}
-contain the LB distributions for time step 20 (the file is 1 of a total
-number of 1 in parallel), and the plain text metadata description,
-respectively.
-
-To ensure output for based-based quantities is in the correct order for
-analysis, post-processing
-may be required (always if the code is run in parallel). This uses a utility
-provided for the purpose which required both the data and the metadata
-description to recombine the parallel output. This utility is described
-ELSEWHERE.
-
-Output for colloidal particle data is again to file with a name encoding
-the time step and parallel decomposition of the output. For example,
-\begin{lstlisting}
-bash-3.2$ ls config*
-config.cds00000020.001-001
-\end{lstlisting}
-contains particle data for time step 20 in a format described in
-Section~\ref{section-examples}.
-These data may be requested in ASCII or raw binary format.
-
-\subsubsection{Errors and run time failures}
-
-The code attempts to provide meaningful diagnostic error messages
-for common problems. Such errors include missing or incorrectly
-formatted input files, inconsistent input values, and unacceptable
-feature combinations. The code should also detect other run time
-problems such as insufficient memory and errors writing output
-files. These errors will result in termination.
-
-Instability in the computation will often be manifested by numerically
-absurd values in the statistical output (ultimately \texttt{NaN} in
-many cases). Instability may or may not result in termination, depending
-on the problem. Such instability is very often related to poor parameter
-choices, of which there can be many combinations. Check the
-relevant section of the documentation for advice on reasonable starting
-parameters for different problems.
-
 \subsection{Note on Units}

### 2023-07-07 docs/input.tex

@@
 
-\subsection{General}
-
-By default, the run time expects to find user input in a file
-\texttt{input} in the current working directory. If a different
-file name is required, its name
-should be provided as the sole command line argument, e.g.,
-\begin{lstlisting}
-./Ludwig.exe input_file_name
-\end{lstlisting}
-If the input file is not located in the current working directory
-the code will terminate immediately with an error message.
-
-When an input file is located, its content is read by a single MPI
-task, and its contents then broadcast to all MPI relevant tasks.
-The format of the file is plain ASCII text, and its contents are
-parsed on a line by line basis. Lines may contain the following:
-\begin{itemize}
-\item comments introduced by \texttt{\#}.
-\item key value pairs separated by white space.
-\end{itemize}
-Blank lines are treated as comments. The behaviour of the code is
-determined by a set of key value pairs. Any given key may appear
-only once in the input file; unused key value pairs are not reported.
-If the key value pairs are not correctly formed, the code will terminate
-with an error message and indicate the offending input line.
-
-Key value pairs may be present in the input file, but have no effect for
-any given run: they are merely ignored. Relevant control parameters for
-given input are reported in the standard output.
-
-\subsubsection{Key value pairs}
-
-Key value pairs are made up of a key --- an alphanumeric string with no
-white space --- and corresponding value following white space. Values
-may take on the follow forms:
-\begin{lstlisting}
-key_string           value_string
-
-key_integer_scalar  1
-key_integer_vector  1_2_3
-
-key_double_scalar    1.0
-key_double_vector    1.0_2.0_3.0
-\end{lstlisting}
-Values which are strings should contain no white space. Scalar parameters
-may be integer values, or floating point values with a decimal point
-(scientific notation is also allowed).  Vector parameters are introduced
-by a set of three values (to be interpreted as $x,y,z$ components of the
-vector in Cartesian coordinates) separated by an underscore. The identity
-of the key will specify what type of value is expected. Keys and (string)
-values are case sensitive.
-
-Most keys have an associated default value which will be used if
-the key is not present. Some keys must be specified: an error will
-occur if they are missing. The remainder of this part
-of the guide details the various choices for key value pairs,
-along with any default values, and any relevant constraints.
-
-\subsection{The Free Energy}
-\label{input-free-energy}
-
-The choice of free energy is determined as follows:
-\begin{lstlisting}
-free_energy   none
-\end{lstlisting}
-The default value is \texttt{none}, i.e., a simple Newtonian fluid is used.
-Possible values of the \texttt{free\_energy} key are:
-\begin{lstlisting}
-#  none                     Newtonian fluid [DEFAULT]
-#  symmetric                Symmetric binary fluid (finite difference)
-#  symmetric_lb             Symmetric binary fluid (two distributions)
-#  brazovskii               Brazovskii smectics
-#  surfactant               Surfactants
-#  polar_active             Polar active gels
-#  lc_blue_phase            Liquid crystal (nematics, cholesterics, BPs)
-#  lc_droplet               Liquid crystal emulsions
-#  fe_electro               Single fluid electrokinetics
-#  fe_electro_symmetric     Binary fluid electrokinetics
-\end{lstlisting}
-
-The choice of free energy will control automatically a number of factors
-related to choice of order parameter, the degree of parallel communication
-required, and so on. Each free energy has a number of associated parameters
-discussed in the following sections.
-
-Details of general (Newtonian) fluid parameters, such as viscosity,
-are discussed in Section~\ref{input-fluid-parameters}.
-
-\subsubsection{Symmetric Binary Fluids}
-
-We recall that the free energy density is, as a function of compositional
-order $\phi$:
-\[
-{\textstyle \frac{1}{2}} A\phi^2 +
-{\textstyle \frac{1}{4}} B\phi^4 +
-{\textstyle \frac{1}{2}} \kappa (\mathbf{\nabla}\phi)^2.
-\]
-
-Parameters are introduced by (with default values):
-\begin{lstlisting}
-free_energy  symmetric
-A            -0.0625                         # Default: -0.003125
-B            +0.0625                         # Default: +0.003125
-K            +0.04                           # Default: +0.002
-\end{lstlisting}
-Common usage has $A < 0$ and $B = -A$ so that $\phi^\star = \pm 1$.
-The parameter $\kappa$
-(key \texttt{K}) controls
-the interfacial energy penalty and is usually positive.
-
-\subsubsection{Brazovskii smectics}
-\label{input-brazovskki-smectics}
-The free energy density is:
-\[
-{\textstyle \frac{1}{2}} A\phi^2 +
-{\textstyle \frac{1}{4}} B\phi^4 +
-{\textstyle \frac{1}{2}} \kappa (\mathbf{\nabla}\phi)^2 +
-{\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 
-\]
-
-Parameters are introduced via the keys:
-\begin{lstlisting}
-free_energy  brazovskii
-A             -0.0005                        # Default: 0.0
-B             +0.0005                        # Default: 0.0
-K             -0.0006                        # Default: 0.0
-C             +0.00076                       # Default: 0.0
-\end{lstlisting}
-For $A<0$, phase separation occurs with a result depending on $\kappa$:
-one gets two symmetric phases for $\kappa >0$ (cf.\ the symmetric case)
-or a lamellar phase for $\kappa < 0$. Typically, $B = -A$ and the
-parameter in the highest derivative $C > 0$.
-
-\subsubsection{Surfactants}
-\label{input-surfactants}
-
-The surfactant free energy should not be used at the present time.
-
-\subsubsection{Polar active gels}
-\label{input-polar-active-gels}
-
-The free energy density is a function of vector order parameter $P_\alpha$:
-\[
-{\textstyle \frac{1}{2}} A P_\alpha P_\alpha +
-{\textstyle \frac{1}{4}} B (P_\alpha P_\alpha)^2 +
-{\textstyle \frac{1}{2}} \kappa (\partial_\alpha P_\beta)^2
-\]
-
-There are no default parameters:
-\begin{lstlisting}
-free_energy        polar_active
-polar_active_a    -0.1                       # Default: 0.0
-polar_active_b    +0.1                       # Default: 0.0
-polar_active_k     0.01                      # Default: 0.0
-\end{lstlisting}
-It is usual to choose $B > 0$, in which case $A > 0$ gives
-an isotropic phase, whereas $A < 0$ gives a polar nematic phase.
-The elastic constant $\kappa$ (key \texttt{polar\_active\_k})
-is positive.
-
-\subsubsection{Liquid crystal}
-\label{input-liquid-crystal}
-The free energy density is a function of tensor order parameter
-$Q_{\alpha\beta}$:
-\begin{eqnarray}
-{\textstyle\frac{1}{2}} A_0 (1 - \gamma/3)Q^2_{\alpha\beta} -
-{\textstyle\frac{1}{3}} A_0 \gamma
-                        Q_{\alpha\beta}Q_{\beta\delta}Q_{\delta\alpha} +
-{\textstyle\frac{1}{4}} A_0 \gamma (Q^2_{\alpha\beta})^2
-\nonumber \\
-+ {\textstyle\frac{1}{2}} \Big(
-\kappa_0 (\epsilon_{\alpha\delta\sigma} \partial_\delta Q_{\sigma\beta} +
-2q_0 Q_{\alpha\beta})^2 + \kappa_1(\partial_\alpha Q_{\alpha\beta})^2 \Big)
-\nonumber
-\end{eqnarray}
-
-The corresponding \texttt{free\_energy} value, despite its name, is
-suitable for nematics and cholesterics, and not just blue phases:
-\begin{lstlisting}
-free_energy      lc_blue_phase
-lc_a0            0.01                       # Deafult: 0.0
-lc_gamma         3.0                        # Default: 0.0
-lc_q0            0.19635                    # Default: 0.0
-lc_kappa0        0.00648456                 # Default: 0.0
-lc_kappa1        0.00648456                 # Default: 0.0
-\end{lstlisting}
-The bulk free energy parameter $A_0$ is positive and controls the
-energy scale (key \texttt{lc\_a0}); $\gamma$ is positive and
-influences the position in the phase diagram relative to the
-isotropic/nematic transition (key \texttt{lc\_gamma}).
-The two elastic constants must be equal, i.e., we enforce the
-single elastic constant approximation (both keys \texttt{lc\_kappa0} and
-\texttt{lc\_kappa1} must be specified).
-
-Other important parameters in the liquid crystal picture are:
-\begin{lstlisting}
-lc_xi            0.7                         # Default: 0.0
-lc_Gamma         0.5                         # Default: 0.0
-lc_active_zeta   0.0                         # Default: 0.0
-\end{lstlisting}
-The first is $\xi$ (key \texttt{lc\_xi}) is the effective molecular
-aspect ratio and should be in the range $0< \xi< 1$. The rotational
-diffusion constant is $\Gamma$ (key \texttt{lc\_Gamma}; not to be
-confused with \texttt{lc\_gamma}). The (optional) apolar activity
-parameter is $\zeta$ (key \texttt{lc\_active\_zeta}).
-
-\subsubsection{Liquid crystal anchoring}
-
-Different types of anchoring are available at solid surfaces, with
-one or two related free energy parameters depending on the type.
-The type of anchoring may be set independently for stationary
-boundaries (walls) and colloids.
-\begin{lstlisting}
-lc_anchoring_strength     0.01               # free energy parameter w1
-lc_anchoring_strength_2   0.0                # free energy parameter w2
-lc_wall_anchoring         normal             # ``normal'' or ``planar''
-lc_coll_anchoring         normal             # ``normal'' or ``planar''
-\end{lstlisting}
-See section \ref{section-lc-anchoring} for details of surface anchoring.
-
-\subsubsection{Liquid crystal emulsion}
-\label{input-liquid-crystal-emulsion}
-
-This an interaction free energy which combines the symmetric and liquid
-crystal free energies. The liquid crystal free energy constant $\gamma$
-becomes a
-function of composition via $\gamma(\phi) = \gamma_0 + \delta(1 + \phi)$,
-and a coupling term is added to the free energy density:
-\[
-WQ_{\alpha\beta} \partial_\alpha \phi \partial_\beta \phi.
-\]
-Typically, we might choose $\gamma_0$ and $\delta$ so that
-$\gamma(-\phi^\star) < 2.7$ and the $-\phi^\star$ phase is isotropic,
-while $\gamma(+\phi^\star) > 2.7$ and the
-$+\phi^\star$ phase is ordered (nematic, cholesteric, or blue phase).
-Experience suggests that a suitable choice is $\gamma_0 = 2.5$ and
-$\delta = 0.25$.
-
-For anchoring constant $W > 0$, the liquid crystal anchoring at the
-interface is planar, while for $W < 0$ the anchoring is normal. This
-is set via key \texttt{lc\_droplet\_W}.
-
-Relevant keys (with default values) are:
-\begin{lstlisting}
-free_energy            lc_droplet
-
-A                      -0.0625
-B                      +0.0625
-K                      +0.053
-
-lc_a0                   0.1
-lc_q0                   0.19635
-lc_kappa0               0.007
-lc_kappa1               0.007
-
-lc_droplet_gamma        2.586                # Default: 0.0
-lc_droplet_delta        0.25                 # Default: 0.0
-lc_droplet_W           -0.05                 # Default: 0.0
-\end{lstlisting}
-Note that key \texttt{lc\_gamma} is not set in this case.
-
-\subsection{System Parameters}
-\label{input-system-parameters}
-
-Basic parameters controlling the number of time steps
-and the system size are:
-\begin{lstlisting}
-N_start      0                              # Default: 0
-N_cycles     100                            # Default: 0
-size         128_128_1                      # Default: 64_64_64
-\end{lstlisting}
-A typical simulation will start from time zero (key \texttt{N\_start})
-and run for a certain number of time steps (key \texttt{N\_cycles}).
-The system size (key \texttt{size}) specifies the total number of
-lattice sites in each dimension. If a two-dimensional system is
-required, the extent in the $z$-direction must be set to unity, as
-in the above example.
-
-If a restart from a previous run is required, the choice of parameters
-may be as follows:
-\begin{lstlisting}
-N_start      100
-N_cycles     400
-\end{lstlisting}
-This will restart from data previously saved at time step 100, and
-run a further 400 cycles, i.e., to time step 500.
-% TODO Cross-reference to I/O.
-
-\subsubsection{Parallel decomposition}
-
-In parallel, the domain decomposition is closely related to the
-system size, and is specified as follows:
-\begin{lstlisting}
-size         64_64_64
-grid         4_2_1
-\end{lstlisting}
-The \texttt{grid} key specifies the number of MPI tasks required in
-each coordinate direction. In the above example, the decomposition
-is into 4 in the $x$-direction, into 2 in the $y$-direction, while
-the $z$-direction is not decomposed. In this example, the local domain
-size per MPI
-task would then be $16\times32\times64$. The total number of MPI tasks
-available must match the total implied by \texttt{grid} (8 in the
-example).
-
-The \texttt{grid} specifications must exactly divide the system size;
-if no decomposition is possible, the code will terminate with an error
-message.
-If the requested decomposition is not valid, or \texttt{grid} is
-omitted, the code will try to supply a decomposition based on
-the number of MPI tasks available and \texttt{MPI\_Dims\_create()};
-this may be implementation dependent.
-
-\subsection{Fluid Parameters}
-\label{input-fluid-parameters}
-
-Control parameters for a Newtonian fluid include:
-\begin{lstlisting}
-fluid_rho0                 1.0
-viscosity                  0.166666666666666
-viscosity_bulk             0.166666666666666
-isothermal_fluctuations    off
-temperature                0.0
-\end{lstlisting}
-The mean fluid density is $\rho_0$ (key \texttt{fluid\_rho0}) which
-defaults to unity in lattice units; it is not usually necessary to
-change this. The shear viscosity is
-\texttt{viscosity} and as default value 1/6 to correspond to
-unit relaxation time in the lattice Boltzmann picture. Reasonable
-values of the shear viscosity are $0.2 > \eta > 0.0001$ in lattice
-units. Higher values move further into the over-relaxation region, and can
-result in poor behaviour. Lower
-values increase the Reynolds number and tend to cause
-problems with stability. The bulk
-viscosity has a default value which is equal to whatever shear
-viscosity has been selected. Higher values of the bulk viscosity
-may be set independently and can help to suppress large deviations
-from incompressibility and maintain numerical stability
-in certain situations.
-
-If fluctuating hydrodynamics is wanted, set the value of
- \texttt{isothermal\_fluctuations} to \texttt{on}. The associated
-temperature is in lattice units: reasonable values (at $\rho_0 = 1$)
-are $0 < kT < 0.0001$. If the temperature is too high, local
-velocities will rapidly exceed the Mach number constraint and
-the simulation will be unstable.
-
-\subsection{Lees Edwards Planes}
-
-Constant uniform shear may be introduced via a number of Lees Edwards
-planes with given speed. This is appropriate for fluid-only
-systems with periodic boundaries.
-\begin{lstlisting}
-N_LE_plane               2           # Number of planes (default: 0)
-LE_plane_vel             0.05        # Constant plane speed
-\end{lstlisting}
-The placing of the planes in the system is as follows.
-The number of planes $N$ must
-divide evenly the lattice size in the $x$-direction to give an integer
-$\delta x$. Planes are then placed at $\delta x / 2, 3\delta x/2, \ldots$.
-All planes have the same, constant, velocity jump associated with them:
-this is positive in the positive $x$-direction. (This jump is usually
-referred to as the plane speed.) The uniform shear rate
-will be $\dot{\gamma} = N U_{LE} / L_x$ where $U_{LE}$ is the plane
-speed (which is always in the $y$-direction).
-
-The velocity gradient or shear direction is $x$, the flow
-direction is $y$ and the vorticity direction is $z$.
-
-The spacing between planes must not be less than twice the halo size
-in lattice units; 8--16 lattice units may be the practical limit in
-many cases. In additional, the speed of the planes must not cause a
-violation of the
-Mach number constraint in the fluid velocity on the lattice, which
-will match the plane speed in magnitude directly either side of the
-planes. A value of around 0.05 should be regarded as a maximum safe
-limit for practical purposes.
-
-Additional keys associated with the Lees Edwards planes are:
-\begin{lstlisting}
-LE_init_profile          1           # Initialise u(x) (off/on)
-LE_time_offset           10000       # Offset time (default 0)
-\end{lstlisting}
-If \texttt{LE\_init\_profile} is set, the fluid velocity is initialised
-to reflect a steady state shear flow appropriate for the number of
-planes at the given speed (ie., shear rate). If set to zero (the default),
-the fluid is initialised with zero velocity.
-
-The code works out the current displacement of the planes by computing
-$U_{LE} t$, where $t$ is the current time step. A shear run should then
-start from $t = 0$, i.e. zero plane displacement.
-It is often convenient to run an equilibration with no shear, and
-then to start an experiment after some number of steps. This
-key allows you to offset the start of the Lees-Edwards motion.
-It should then take the value of the start time (in time steps)
-corresponding to the restart at the end of the equilibration
-period.
-
-There are a couple of additional constraints to use the Lees-Edwards
-planes in parallel. In particular, the planes cannot fall at a
-processor boundary in the $x$-direction. This means you should
-arrange an integer number of planes per process in the $x$-direction.
-(For example, use one plane per process; this will also ensure the number
-of planes
-still evenly divides the total system size.)
-This will interleave the planes with the processor decomposition.
-The $y$-direction and $z$-direction may be decomposed without
-further constraint.
-
-Note that this means a simulation with one plane will only work
-if there is one process in the $x$ decomposition.
-
-\subsection{Colloids}
-If no relevant key words are present, no colloids will be
-expected at run time. The simulation will progress in the
-usual fashion with fluid only.
-
-If colloids are required, the \texttt{colloid\_init}
-key word must be present to allow the code to determine where
-colloid information will come from. The options for the
-\texttt{colloid\_init} key word are summarised below:
-\begin{lstlisting}
-colloid_init             none
-
-#  none                  no colloids [DEFAULT]
-#  input_one             one colloid from input file
-#  input_two             two colloids from input file
-#  input_three           three colloids from input file
-#  input_random          Small number at random
-#  from_file             Read a separate file (including all restarts)
-\end{lstlisting}
-
-For idealised simulations which require 1, 2, or 3 colloids, relevant
-initial state information 
-for each particle can be included in the input file. In principle, most
-of the colloid state as defined in the colloid
-state structure in \texttt{colloid.h} may be specified. (Some state is
-reserved for internal use only and cannot be set from the input file.)
-Furthermore, not all the state is relevant in all simulations ---
-quantities such as charge and wetting parameters may not be required,
-in which case they a simply ignored.
-
-A minimal initialisation is shown in the following example:
-\begin{lstlisting}
-colloid_init              input_one
-
-colloid_one_a0            1.25
-colloid_one_ah            1.25
-colloid_one_r             12.0_12.0_12.0
-\end{lstlisting}
-This initialises a single colloid with an input radius $a_0=1.25$,
-and a hydrodynamic radius $a_h=1.25$; in general both are required,
-but they do not have to be equal.
-A valid position is required within the extent of the system
-$0.5 < x,y,z < L + 0.5$ as specified by the \texttt{size} key word.
-State which is not explicitly defined is initialised to zero.
-
-\subsubsection{General Initialisation}
-
-A full list of colloid state-related key words is as follows. All
-the quantities are floating point numbers unless explicitly stated
-to be otherwise:
-\begin{lstlisting}
-# colloid_one_nbonds      (integer) number of bonds
-#   colloid_one_bond1     (integer) index of bond partner 1
-#   colloid_one_bond2     (integer) index of bond partner 2
-# colloid_one_isfixedr    colloid has fixed position (integer 0 or 1)
-# colloid_one_isfixedv    colloid has fixed velocity (integer 0 or 1)
-# colloid_one_isfixedw    colloid has fixed angular velocity (0 or 1)
-# colloid_one_isfixeds    colloid has fixed spin (0 or 1)
-# colloid_one_type        ``default'' COLLOID_TYPE_DEFAULT
-#                         ``active''  COLLOID_TYPE_ACTIVE
-#                         ``subgrid'' COLLOID_TYPE_SUBGRID
-# colloid_one_a0          input radius
-# colloid_one_ah          hydrodynamic radius
-# colloid_one_r           position vector
-# colloid_one_v           velocity (vector)
-# colloid_one_w           angular velocity (vector)
-# colloid_one_s           spin (unit vector)
-# colloid_one_m           direction of motion (unit) vector for swimmers 
-\end{lstlisting}
-
-Note that for magnetic particles, the appropriate initialisation involves
-the spin key word \texttt{colloid\_one\_s} which relates to the dipole
-moment $\mu \mathbf{s}_i$, while \texttt{colloid\_one\_m} relates to the
-direction of motion vector. Do not confuse the two.
-It is possible in principle to have magnetic active particles,
-in which case the dipole direction or spin ($\mathbf{s}_i$) and the
-direction of swimming motion $\mathbf{m}_i$ are allowed to be distinct.
-
-\begin{lstlisting}
-# colloid_one_b1          Squirmer parameter B_1
-# colloid_one_b2          Squirmer parameter B_2
-# colloid_one_rng         (integer) random number generator state
-# colloid_one_q0          charge (charge species 0)
-# colloid_one_q1          charge (charge species 1)
-# colloid_one_epsilon     Permeativity
-# colloid_one_c           Wetting parameter C
-# colloid_one_h           Wetting parameter H
-\end{lstlisting}
-
-{\bf Example}: Single active (squirmer) particle
-
-The following example shows a single active particle with initial
-swimming direction along the $x$-axis.
-\begin{lstlisting}
-colloid_init              input_one
-
-colloid_one_type          active
-colloid_one_a0            7.25
-colloid_one_ah            7.25
-colloid_one_r             32.0_32.0_32.0
-colloid_one_v             0.0_0.0_0.0
-colloid_one_m             1.0_0.0_0.0
-colloid_one_b1            0.05
-colloid_one_b2            0.05
-\end{lstlisting}
-
-\subsubsection{Random initialisation}
-
-For suspensions with more than few colloids, but still at
-relatively low volume fraction (10--20\% by volume), it is
-possible to request initialisation at random positions.
-
-The additional key word value pair \texttt{colloid\_random\_no}
-determines the total number of particles to be placed in
-the system. To prevent particles being initialised very
-close together, which can cause problems in the first few
-time steps if strong potential interactions are present,
-a ``grace'' distance or minimum surface-surface separation
-may also be specified (\texttt{colloid\_random\_dh}).
-
-The following example asks for 100 colloids to be initialised
-at random positions, with a minimum separation of 0.5 lattice
-spacing.
-
-\begin{lstlisting}
-colloid_init              input_random
-
-colloid_random_no         100             # Total number of colloids
-colloid_random_dh         0.5             # ``Grace'' distance
-
-colloid_random_a0         2.30
-colloid_random_ah         2.40
-\end{lstlisting}
-
-An input radius and hydrodynamic radius must be provided: these
-are the same for all colloids.
-If specific initialisations of the colloid state (excepting the
-position) other than the radii are wanted, values should be provided
-as for the single particle case in the preceding section, but using
-\texttt{colloid\_random\_a0} in place of
-\texttt{colloid\_one\_a0} and so on.
-
-The code will try to initialise the requested number in the current
-system size, but only makes a finite number of attempts to place
-particles at random with no overlaps. (The initialisation will also
-take into account the presence of any solid walls, using the same
-grace distance.) If the the number of particles is too large, the
-code will halt with a message to that effect.
-
-In general, colloid information for a arbitrary configuration with many
-particles should be read from a pre-prepared file. See the section on
-File I/O for further information on reading files.
-
-\subsubsection{Interactions}
-
-Note that two-body pair-potential interactions are defined uniformly for
-all colloids in a simulation. The same is true for lubrication corrections.
-There are a number of constraints related to the computation of
-interactions discussed below.
-
-{\bf Boundary-colloid lubrication correction}.
-Lubrication corrections (here the normal force) between a flat wall
-(see section XREF) are
-required to prevent overlap between colloid  and the wall.
-The cutoff distance is set via the key word value pair
-\begin{lstlisting}
-boundary_lubrication_rcnormal    0.5
-\end{lstlisting}
-It is recommended that this is used in all cases involving walls.
-A reasonable value is in the range $0.1 < r_c < 0.5$ in lattice
-units, and should be calibrated for particle hydrodynamic radius
-and fluid viscosity if exact results are important.
-
-{\bf Colloid-colloid lubrication corrections}.
-The key words to activate the calculation of lubrication corrections
-are:
-\begin{lstlisting}
-lubrication_on                   1
-lubrication_normal_cutoff        0.5
-lubrication_tangential_cutoff    0.05
-\end{lstlisting}
-
-{\bf Soft sphere potential}.
-A cut-and-shifted soft-sphere potential of the form
-$v \sim \epsilon (\sigma/r)^\nu$ is
-available. Some trial-and-error with the parameters may be required in
-any given situation to ensure simulation stability in the long run. The
-following gives an example of the relevant input key words:
-\begin{lstlisting}
-soft_sphere_on            1                 # integer 0/1 for off/on 
-soft_sphere_epsilon       0.0004            # energy units
-soft_sphere_sigma         1.0               # a length
-soft_sphere_nu            1.0               # exponent is positive
-soft_sphere_cutoff        2.25              # a surface-surface separation
-\end{lstlisting}
-See Section~\ref{section-colloids-soft-sphere} for a detailed description.
-
-{\bf Lennard-Jones potential}.
-The Lennard-Jones potential (Section~\ref{section-colloids-lennard-jones})
-is controlled by the following key words:
-\begin{lstlisting}
-lennard_jones_on          1                 # integer 0/1 off/on
-lj_epsilon                0.1               # energy units
-lj_sigma                  2.6               # potential length scale
-lj_cutoff                 8.0               # a centre-centre separation
-\end{lstlisting}
-
-{\bf Yukawa potential}.
-A cut-and-shifted Yukawa potential of the form
-$v \sim \epsilon \exp(-\kappa r)/r$ is
-available using the following key word value pairs:
-
-\begin{lstlisting}
-yukawa_on                 1                 # integer 0/1 off/on
-yukawa_epsilon            1.330             # energy units
-yukawa_kappa              0.725             # an inverse length
-yukawa_cutoff             16.0              # a centre-centre cutoff
-\end{lstlisting}
-
-{\bf Dipole-dipole interactions (Ewald sum)}.
-The Ewald sum is completely specified in the input file
-by the uniform dipole strength $\mu$ and the real-space cut off $r_c$.  
-\begin{lstlisting}
-ewald_sum                 1                 # integer 0/1 off/on
-ewald_mu                  0.285             # dipole strength mu
-ewald_rc                  16.0              # real space cut off
-\end{lstlisting}
-
-If short range interactions are required, particle information is stored
-in a cell list, which allows efficient computation of the potentially
-$N^2$ interactions present. This gives rise to a constraint that the
-width of the cells must be large enough that all relevant interactions
-are included. This generally means that the cells must be at least
-$2a_h + h_c$ where $h_c$ is the largest relevant cut off distance.
-
-The requirement for at least two cells per local domain in parallel
-means that there is a associated minimum local domain size. This is
-computed at run time on the basis of the input. If the local domain
-is too small, the code will terminate with an error message. The
-local domain size should be increased.
-
-\subsubsection{External forces}
-
-{\bf External body force (gravity)}.
-The following example requests a uniform body force in the negative
-$z$-direction on all particles.
-\begin{lstlisting}
-colloid_gravity           0.0_0.0_-0.001    # vector
-\end{lstlisting}
-The counterbalancing body force on the fluid which enforces the
-constraint of momentum conservation for the system as a whole is
-computed automatically by the code at each time step.
-
-\subsection{Order parameter initialisations}
-
-Free energy choices requiring one or more fluid order parameters can make
-use of the following initial coniditions.
-
-\subsubsection{Composition $\phi$}
-
-The following initialisations are available.
-\begin{lstlisting}
-phi_initialisation          spinodal    # spinodal
-phi0                        0.0         # mean
-noise                       0.05        # noise amplitude
-random_seed                 102839      # +ve integer
-\end{lstlisting}
-Suitable for initialising isothermal spinodal decomposition, the order
-parameter may be set at random at each position via
-$\phi = \phi_0 + A(\hat{\phi} - 1/2)$ with the random variate
-$\hat\phi$ selected uniformly on the interval $[0,1)$. For symmetric
-quenches (mean order parameter $\phi_0 = 0$ and $\phi^\star = \pm 1$),
-a value of $A$ in the range 0.05-0.1 is usually appropriate.
-
-For off-symmetric quenches, larger patches of fluid may be required to
-initiate decomposition:
-\begin{lstlisting}
-phi_initialisation          patches     # patches of phi = +/- 1
-phi_init_patch_size         2           # patch size
-phi_init_patch_vol          0.1         # volume fraction phi = -1 phase
-random_seed                 13          # +ve integer
-\end{lstlisting}
-The initialises cubics patches of fluid of given size with $\phi= \pm 1$
-at random. The requested overall volume fractions may be met approximately.
-
-A spherical drop can be initialised at the centre of the system.
-\begin{lstlisting}
-phi_initialisation          drop        # spherical droplet
-phi_init_drop_radius        16.0        # radius
-phi_init_drop_amplitude    -1.0         # phi value inside
-\end{lstlisting}
-The drop is initialised with a $\tanh(r/\xi)$ profile where the
-interfacial width $\xi$ is computed via the current free energy
-parameters.
-
-% block
-% bath
-
-For restarted simulations, the default position is to read order
-parameter information from file
-\begin{lstlisting}
-phi_initialisation          from_file
-\end{lstlisting}
-in which case a file or files for the appropriate time step should
-be present in the working directory.
-
-\subsubsection{Tensor order $Q_{\alpha\beta}$}
-
-A number of different initialisations are available for the liquid
-crystal order parameter $Q_{\alpha\beta}$. Some care may be required
-to ensure consistency between the choice and the free energy
-parameters, the system size, and so on (particularly for the blue phases).
-
-A summmary of choices is:
-\begin{lstlisting}
-lc_q_initialisation   nematic          # uniform nematic...
-lc_init_nematic       1.0_0.0_0.0      # ...with given director
-
-lc_q_initialisation   cholesteric_x    # cholesteric with helical axis x
-lc_q_initialisation   cholesteric_y    # cholesteric with helical axis y
-lc_q_initialisation   cholesteric_z    # cholesteric with helical axis z
-
-lc_q_initialisation   o8m              # BPI high chirality limit
-lc_q_initialisation   o2               # BPII high chirality limit
-lc_q_initialisation   o5
-lc_q_initialisation   h2d              # 2d hexagonal
-lc_q_initialisation   h3da             # 3d hexagonal BP A
-lc_q_initialisation   h3db             # 3d hexagonal BP B
-lc_q_initialisation   dtc              # double twist cylinders
-
-lc_q_initialisation   bp3
-
-lc_q_initialisation   cf1_x            # cholesteric ``finger'' axis x
-lc_q_initialisation   cf1_y            # cholesteric ``finger'' axis y
-lc_q_initialisation   cf1_z            # cholesteric ``finger'' axis z
-
-lc_q_initialisation   cf1_fluc_x       # as cf1_x with random perterbations
-lc_q_initialisation   cf1_fluc_y       # as cf1_y with random perturbations
-lc_q_initialisation   cf1_flux_z       # as cf1_z with random perturbations
-
-lc_q_initialistion    random           # with randomly chosen unit director
-\end{lstlisting}
-Note many of the initialiations require an initial amplitude of order,
-which should be set via
-\begin{lstlisting}
-lc_q_init_amplitude   0.01             # initial amplitude of order A
-\end{lstlisting}
-For example, if an initial uniform nematic is requested with
-unit director $n_\alpha$, the corresponding initial tensor will be
-$$
-Q_{\alpha\beta} = 
-{\textstyle \frac{1}{2}} A (3 n_\alpha n_\beta - \delta_{\alpha\beta}).
-$$
+The up-to-date version of this material now appears at
+\texttt{https://ludwig.epcc.ed.ac.uk/inputs/index.html}
 

### 2023-07-20 docs/electrokinetic.tex

@@
 
-\subsubsection{Liquid-junction potential}
-
-The liquid junction potential 
-is a charge separation process that 
-occurs when electrolytes with slightly different concentrations
-whose species have different diffusivities are brought into contact.
-Charges from the regions of higher concentration diffuse   
-into the parts with lower concentration. Due to the difference 
-in diffusivity they migrate at different speeds, leaving parts of
-the system charged. This leads to a build-up of a potential
-which balances the diffusive flux.
-
-After the initial build-up phase the potential decreases slowly 
-again until the charge concentration has become homogeneous throughout 
-the system. Both timescales of emergence and decay of the potential
-can be separated by chosing a sufficiently large system size.
-
-This problem allowed us to verify the correct temporal 
-behaviour of the Nernst-Planck equation solver by resolving the transient 
-dynamics without having to account for advective terms.
-
-\begin{figure}[h!t]
-\includegraphics[width=0.495\textwidth]{./pics/test_lj_zoom1.pdf}
-\includegraphics[width=0.495\textwidth]{./pics/test_lj_zoom3.pdf}
-\caption{Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue).
-The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$.
-Deviations can be seen which are presumably due
-to the approximate nature of the analytical solution and the fact that we
-have a different asymptotic flux close to the boundary - the theoretical 
-analysis assumes no flux at a boundary which is infinitely far away from
-the diffusive region.} 
-\label{fig3} 
-\end{figure}
-
-For simplicity we considered systems of size 
-$L_x\times L_y\times Lz=128\times4\times4$ and 
-$L_x\times L_y\times Lz=256\times4\times4$ with
-periodic boundary conditions at either end.
-The two halfs were electroneutral and had ionic concentrations 
-$\rho_{L,\pm}=\rho_{0,\pm} + \delta\rho$ and 
-$\rho_{R,\pm}=\rho_{0,\pm} - \delta\rho$ 
-with $\rho_{0,\pm}=1\e{-2}$ and $\delta\rho = 0.01$.
-
-The potential difference between both sides during the build-up 
-obeys approximately
-
-\begin{equation}
-\Delta\Psi(t)\simeq\Delta\Psi_P \left\{1-\exp\left(-\frac{t}{\tau_e}\right)\right\}\\
-\end{equation}
-
-with
-
-\begin{equation}
-\Delta\Psi_P=\frac{(D_+ D_-)}{\beta e (D_+ + D_-)} \frac{\delta\rho}{\rho_0}\\ 
-\end{equation}
-
-as saturation value of the potential difference.
-The saturation time scale is given by
-
-\begin{equation}
-\tau_e=\frac{\varepsilon}{\beta \, e^2 (D_+ + D_-) \rho_0}.
-\end{equation}
-
-\begin{figure}[htpb]
-\includegraphics[width=0.9\textwidth]{./pics/test_lj_decay1.pdf}
-\caption{Rescaled plot of the decay: Times have been Time evolution of the liquid junction potential for $L_x=128$ (blue), $L_x=256$ (green) and $L_x=512$ (blue).
-The dashed black curves represent the approximate solution in the limit $l_D/L_x<<1$.
-Deviations can be seen which are due
-to the approximate nature of the analytical solution and the diffusive fluxes close to the boundary.} 
-\label{fig4} 
-\end{figure}
-
-A more exact solution can be derived in the limit $l_D/L_x<<1, N\to\infty$:
-
-\begin{eqnarray}
-\Delta\Psi(t)&=&\Delta\Psi_P \left\{1-\exp\left(-\frac{t}{\tau_e}\right)\right\}\frac{4}{\pi}\left\{\sum_{m=1}^N \frac{\sin^3(m\pi/2)}{m} \exp\left(-\frac{ m^2\, t}{\tau_d}\right)\right\}\\
-\tau_d&=&\frac{L^2}{2\pi^2 (D_+ + D_-)}.\label{taud}
-\end{eqnarray}
-
-It contains as well the dependence on the decay timescale $\tau_d$.
-Only odd indices $m$ contribute to the sum:
-
-\begin{eqnarray}
-\sum_{m=1}^N \frac{\sin^3(m\pi/2)}{m} \exp\left(-\frac{m^2\, t}{\tau_d}\right)&=&\nonumber\\
-&&\hspace*{-4cm} \exp\left(-\frac{t}{\tau_d}\right)-\frac{1}{3} \exp\left(-\frac{9 t}{\tau_d}\right)+\frac{1}{5}\exp\left(-\frac{25t}{\tau_d}\right)-\frac{1}{9}\exp\left(-\frac{81 t}{\tau_d}\right)+\ldots
-\end{eqnarray}
-
-A complete discussion of the solution can be found in \cite{Mafe}. 
-There, the upper limit of significant modes has been also estimated as $N_{max} = L/\pi l_D$.
-Note the factor 2 difference between Eq. \ref{taud} and the corresponding expression in \cite{Mafe}.
-
-The following parameters were used:
-dielectric permittivity $\epsilon=3.3\e{3}$, temperature $\beta^{-1}=3.333\e{-5}$, unit charge $e=1$, valency $z_\pm=\pm1$, diffusivities $D_+=0.0125$ and $D_-=0.0075$.
-We obtained
-$\Delta\Psi_P=1.6667\e{-7}$, $\tau_e=550$, $\tau_d=41501.2\, (L_x=128)$, $\tau_d=166004.6\, (L_x=256)$ and $\tau_d=664018.5 (L_x=512)$.
-The results for the potential difference over time are shown in Fig. \ref{fig3}.
-
-Fig. \ref{fig4} shows results with times rescaled to the decay 
-time scale $\tau_d$ (cf. Eq. \ref{taud}). Obviously the 
-deviations we observe are not due to the limited system size 
-and have a more systematic origin.
-
-The curves coincide if 
-the theoretic limit for $\tau_d$ is rescaled by a factor $1.067$,
-suggesting the effective system length for this sort of setup is
-actually about 3\% larger than the numerical value.
-
-A reason for this might be the approximate nature of the analytical solution 
-and the fact that it was gained
-for an infinitely large system with constant charge concentrations,
-vanishing currents at both ends and finite diffusive zone of size $L_x$.
-In our situation the entire system is within the diffusive zone.
-This may lead to smaller effective diffusivities or larger effective
-system sizes.
-Interestingly, all runs with solid walls at both ends resulted in 
-oscillatory behaviour and an effective system size of $2L_x$.
-
-\subsubsection{Electroosmotic Flow}
-
-To test the implementation with all couplings to external and 
-internal forces we consider a forced charged fluid in a slit
-of size $L_z$. An electrostatic field $E_{||}$ is allied
-parallel to the walls. The entire system is electroneutral with 
-each wall having the surface charge density $\sigma$ 
-and compensationg counterions with total charge $2 \sigma A_{wall}$
-in the fluid.
-
-In equilibrium the charge density at a distance $x$ from the wall obeys
-
-\begin{equation}
-\rho(x)=\frac{\rho_0}{\cos^2(K\,x)}
-\end{equation}
-
-with
-
-\begin{equation}
-\rho_0=K^2/2\pi l_B
-\end{equation}
-
-and
-
-\begin{eqnarray}
-K \,L_x \tan\left(\frac{K\, L_x}{2}\right)&=&\pi\, l_B\, L_x\, 2\sigma\label{kex} \\
-K \,L_x&\simeq&\sqrt{4\pi \,l_B\,L_x\,2\sigma}\label{klin}.
-\end{eqnarray}
-
-The liniarised version Eq. \ref{klin} has only a limited range of applicabilty.
-We solved Eq. \ref{kex} numerically and found solutions 
-$K=0.01959\; (\sigma=0.003125)$ and $K=0.03311\; (\sigma=0.00125)$, 
-which is reasonably far away from the 
-theoretical limit $K_{max}=\pi/L_x$ set by the tangent.
-
-Note the factor 2 difference on the lhs of Eq. \ref{kex} with respect 
-to \cite{Capuani, Rotenberg}. There is also a factor $L_x$ missing on 
-the rhs of Eq. \ref{klin}.
-
-The steady state velocity of the fluid can be derived from the 
-force balance of the gradient of the stresses and the electrostatic
-forces:
-
-\begin{eqnarray}
-v_y(x)&=&\hat{v} \ln\left(\frac{\cos(K\,x)}{\cos(K\,L_x/2)}\right)\label{vy}\\
-\hat{v}&=&\frac{e \,E_{||}\rho_o}{\eta\, K^2}=\frac{e \,E_{||}}{2\pi\eta l_B}\label{vhat}
-\end{eqnarray}
-
-The result for two different charge densities is shown in Fig. \ref{fig6}.
-The accuracy is acceptable with deviations for high surface 
-charged potentially being caused by the chosen discretisation or 
-by the numerical solution of Eq. \ref{kex} approaching the limit 
-of $\pi/L_z\simeq0.049$.
-
-\begin{figure}[htpb]
-\includegraphics[width=0.9\textwidth]{./pics/test_eo.pdf}
-\caption{Steady state flow profile across a slit of width $L_x=63$ for an applied field of magnitude $e \beta E L_x=1.89$ for two different charge densities $\sigma=0.003125$ (red) and $\sigma=0.0125$ (blue), Bjerrum length $l_B=0.7234$, viscosity $\eta=0.1$ and unit charge $e=1$.
-The dashed black lines correspond to theoretical prediction according to Eqs.
-\ref{vy} and \ref{vhat}.}
-\label{fig6} 
-\end{figure}
-
 \subsubsection{Debye-H\"uckel Theory}
